### 2017-09-15 paper/casebase_jss.Rmd
Initial commit for JSS paper

---
author:
  - name: Sahir Bhatnagar *
    affiliation: McGill Univeristy
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon *
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{maxime.turgeon@mail.mcgill.ca}
    url: http://maxturgeon.ca/
  - name: James Hanley
    affiliation: McGill Univeristy
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. * joint co-authors
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
---

# Code formatting

Don't use markdown, instead use the more precise latex commands:

* \proglang{Java}
* \pkg{plyr}
* \code{print("abc")}

# Introduction

  - Motivation
    + Flexible 
    + Flexible
    + Flexible

# Theoretical details

# Implementation details

  1. Population-tim plots
  2. Sampling
  3. Fitting
  4. Absolute Risks

# Population-time plots

# Case study 1: Veteran data (or ERSPC if we can)

  - First example
  - Show how we can test for non-proportional hazard?

# Case study 2: Bone-marrow transplant

  - Competing risk example
  - Comparison with Cox and Fine-Gray

# Case study 3: Vaccination study (recurrent events)

  - Give a more complex example of sampling; time-dependent exposure
    + Sampling needs to be done manually, but fitting function can still be used

# Discussion

### 2017-09-15 paper/casebase_jss.tex
Initial commit for JSS paper

\documentclass[article]{jss}
\usepackage[utf8]{inputenc}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\author{
Sahir Bhatnagar *\\McGill Univeristy \And Maxime Turgeon *\\McGill University \And James Hanley\\McGill Univeristy \And Olli Saarela\\University of Toronto
}
\title{\pkg{casebase}: An Alternative Framework For Survival Analysis}

\Plainauthor{Sahir Bhatnagar *, Maxime Turgeon *, James Hanley, Olli Saarela}
\Plaintitle{casebase: An Alternative Framework For Survival Analysis}
\Shorttitle{\pkg{casebase}: An Alternative Framework For Survival Analysis}

\Abstract{
The abstract of the article. * joint co-authors
}

\Keywords{keywords, not capitalized, \proglang{Java}}
\Plainkeywords{keywords, not capitalized, Java}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{}
%% \Acceptdate{2012-06-04}

\Address{
    Sahir Bhatnagar *\\
  McGill Univeristy\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{sahir.bhatnagar@mail.mcgill.ca}\\
  URL: \url{http://sahirbhatnagar.com/}\\~\\
      Maxime Turgeon *\\
  McGill University\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{maxime.turgeon@mail.mcgill.ca}\\
  URL: \url{http://maxturgeon.ca/}\\~\\
      James Hanley\\
  McGill Univeristy\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{james.hanley@mcgill.ca}\\
  URL: \url{http://www.medicine.mcgill.ca/epidemiology/hanley/}\\~\\
      Olli Saarela\\
  University of Toronto\\
  Dalla Lana School of Public Health, 155 College Street, 6th floor,
  Toronto, Ontario M5T 3M7, Canada\\
  E-mail: \email{olli.saarela@utoronto.ca}\\
  URL: \url{http://individual.utoronto.ca/osaarela/}\\~\\
  }

% Pandoc header

\usepackage{amsmath}

\begin{document}

\section{Code formatting}\label{code-formatting}

Don't use markdown, instead use the more precise latex commands:

\begin{itemize}
\tightlist
\item
  \proglang{Java}
\item
  \pkg{plyr}
\item
  \code{print("abc")}
\end{itemize}

\section{Introduction}\label{introduction}

\begin{itemize}
\tightlist
\item
  Motivation

  \begin{itemize}
  \tightlist
  \item
    Flexible
  \item
    Flexible
  \item
    Flexible
  \end{itemize}
\end{itemize}

\section{Theoretical details}\label{theoretical-details}

\section{Implementation details}\label{implementation-details}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Population-tim plots
\item
  Sampling
\item
  Fitting
\item
  Absolute Risks
\end{enumerate}

\section{Population-time plots}\label{population-time-plots}

\section{Case study 1: Veteran data (or ERSPC if we
can)}\label{case-study-1-veteran-data-or-erspc-if-we-can}

\begin{itemize}
\tightlist
\item
  First example
\item
  Show how we can test for non-proportional hazard?
\end{itemize}

\section{Case study 2: Bone-marrow
transplant}\label{case-study-2-bone-marrow-transplant}

\begin{itemize}
\tightlist
\item
  Competing risk example
\item
  Comparison with Cox and Fine-Gray
\end{itemize}

\section{Case study 3: Vaccination study (recurrent
events)}\label{case-study-3-vaccination-study-recurrent-events}

\begin{itemize}
\tightlist
\item
  Give a more complex example of sampling; time-dependent exposure

  \begin{itemize}
  \tightlist
  \item
    Sampling needs to be done manually, but fitting function can still
    be used
  \end{itemize}
\end{itemize}

\end{document}

### 2017-09-15 paper/casebase_jss.Rmd
add bibtex file for references
add bibtex file for references

@@
   plain:     [keywords, not capitalized, Java]
+bibliography: casebase_references.bib
 preamble: >

### 2017-09-16 paper/casebase_jss.Rmd
change plots to poptime

--- 2017-09-15 paper/casebase_jss.Rmd
@@
   plain:     [keywords, not capitalized, Java]
+bibliography: casebase_references.bib
+csl: jss.csl
 preamble: >
@@
 
-  - Competing risk example
-  - Comparison with Cox and Fine-Gray
+We will use the same data that was used in Scrucca *et al* [@scrucca2010regression].
+The data is available on the main author's [website](http://www.stat.unipg.it/luca/R/); it is also available as part of this package.
+
+```{r}
+library(casebase)
+library(magrittr)
+library(tidyverse)
+data(bmtcrr)
+head(bmtcrr) %>% knitr::kable(format = "latex")
+```
+
+We will perform a competing risk analysis on data from 177 patients who received a stem cell transplant for acute leukemia.
+The event of interest in relapse, but other competing causes (e.g.
+transplant-related death) need to be taken into account.
+We also want to take into account the effect of several covariates such as Sex, Disease (lymphoblastic or myeloblastic leukemia, abbreviated as ALL and AML, respectively), Phase at transplant (Relapse, CR1, CR2, CR3), Source of stem cells (bone marrow and peripheral blood, coded as BM+PB, or peripheral blood, coded as PB), and Age.
+Below, we reproduce their Table 1:
+
+```{r}
+table1 <- tibble::tribble(
+    ~Variable, ~Description, ~`Statistical summary`,
+    "Sex", "Sex", "M=Male (100)",
+    "", "", "F=Female (77)",
+    "D", "Disease", "ALL (73)",
+    "", "", "AML (104)",
+    "Phase", "Phase", "CR1 (47)", 
+    "", "", "CR2 (45)", 
+    "", "", "CR3 (12)", 
+    "", "", "Relapse (73)",
+    "Source", "Type of transplant", "BM+PB (21)", 
+    "", "", "PB (156)",
+    "Age", "Age of patient (years)", "4–62", 
+    "", "", "30.47 (13.04)",
+    "Ftime", "Failure time (months)", "0.13–131.77",
+    "", "", "20.28 (30.78)",
+    "Status", "Status indicator", "0=censored (46)", 
+    "", "", "1=relapse (56)", 
+    "", "", "2=competing event (75)"
+)
+
+knitr::kable(table1, format = "latex")
+```
+
+<!-- | Variable | Description            | Statistical summary    | -->
+<!-- | -------- | ---------------------- | ---------------------- | -->
+<!-- | Sex      | Sex                    | M=Male (100) <br> F=Female (77) | -->
+<!-- | D        | Disease                | ALL (73) <br> AML (104) | -->
+<!-- | Phase    | Phase                  | CR1 (47) <br> CR2 (45) <br> CR3 (12) <br> Relapse (73) | -->
+<!-- | Source   | Type of transplant     | BM+PB (21) <br> PB (156) | -->
+<!-- | Age      | Age of patient (years) | 4–62 <br> 30.47 (13.04) | -->
+<!-- | Ftime    | Failure time (months)  | 0.13–131.77 <br> 20.28 (30.78) | -->
+<!-- | Status   | Status indicator       | 0=censored (46) <br> 1=relapse (56) <br> 2=competing event (75) | -->
+
+The statistical summary is generated differently for continuous and categorical variables:
+
+ - For continuous variables, we are given the range, followed by the mean and standard deviation.
+
+ - For categorical variables, we are given the counts for each category.
+
+Note that failure time can also correspond to censoring.
+
+In order to try and visualize the incidence density of relapse, we can look at a population-time plot: on the X-axis we have time, and on the Y-axis we have the size of the risk set at a particular time point.
+Failure times associated to the event of interest can then be highlighted on the plot using red dots.
+
+```{r}
+pt_object <- casebase::popTime(bmtcrr, event = "Status", time = "ftime")
+plot(pt_object)
+```
+
+```{r}
+pt_object_comp <- bmtcrr %>% 
+    mutate(Status = case_when(Status == 1 ~ 2L, 
+                              Status == 2 ~ 1L, 
+                              TRUE ~ Status)) %>% 
+    casebase::popTime(event = "Status", time = "ftime")
+plot(pt_object_comp, point.colour = "blue")
+```
+
+From this last plot, we can see that there is no censoring during the first 10 months.
+Moreover, we see that the last competing event occurs around 20 months.
+Putting all this information together, we have evidence of two types of patients: very sick patients who either relapse or have a competing event early on, and healthier patients who are eventually lost to follow-up.
+
+We now turn to the analysis of this dataset.
+The population-time plots above give evidence of non-constant hazard; therefore, we will explicitely include time in the model.
+Note that we also include all other variables as possible confounders.
+First, we include time as a linear term:
+
+```{r warning=FALSE}
+model1 <- fitSmoothHazard(Status ~ ftime + Sex + D + Phase + Source + Age, 
+                          data = bmtcrr, 
+                          ratio = 100,
+                          time = "ftime")
+summary(model1)
+```
+
+Because of the results in Turgeon *et al* [@turgeonCompRisk], the standard errors we obtain from the multinomial logit fit are asymptotically correct, and therefore can be used to construct asymptotic confidence intervals.
+
+From this summary, we see that time is indeed significant, as is Phase (only relapse vs.
+CR1).
+Interestingly, we see that the type of disease is only significant for the event of interest, whereas the type of transplant and the age of the patient are only significant for the competing event.
+
+Next, we include the logarithm of time in the model (which leads to a Weibull hazard):
+
+```{r warning = FALSE}
+model2 <- fitSmoothHazard(Status ~ log(ftime) + Sex + D + Phase + Source + Age, 
+                          data = bmtcrr, 
+                          ratio = 100, 
+                          time = "ftime")
+summary(model2)
+```
+
+As we can see, the results are similar to the ones with a Gompertz hazard, although Sex is now significant for the competing event.
+
+Finally, using splines, we can be quite flexible about the way the hazard depends on time:
+
+```{r warning = FALSE}
+model3 <- fitSmoothHazard(
+    Status ~ splines::bs(ftime) + Sex + D + Phase + Source + Age, 
+    data = bmtcrr, 
+    ratio = 100, 
+    time = "ftime")
+summary(model3)
+```
+
+Again, we see that the results are quite similar for this third model.
+
+We now look at the 2-year risk of relapse:
+
+```{r absRisk, warning = FALSE}
+linearRisk <- absoluteRisk(object = model1, time = 24, newdata = bmtcrr[1:10,])
+logRisk <- absoluteRisk(object = model2, time = 24, newdata = bmtcrr[1:10,])
+splineRisk <- absoluteRisk(object = model3, time = 24, newdata = bmtcrr[1:10,])
+```
+
+```{r absRiskPlot}
+plot(linearRisk, logRisk,
+     xlab = "Linear", ylab = "Log/Spline", pch = 19,
+     xlim = c(0,1), ylim = c(0,1), col = 'red')
+points(linearRisk, splineRisk,
+       col = 'blue', pch = 19)
+abline(a = 0, b = 1, lty = 2, lwd = 2)
+legend("topleft", legend = c("Log", "Spline"),
+       pch = 19, col = c("red", "blue"))
+```
+
+We can also estimate the mean absolute risk for the entire dataset:
+```{r}
+mean(linearRisk)
+mean(logRisk)
+mean(splineRisk)
+```
+
+```{r warning = FALSE, cache = TRUE}
+# Absolute risks----
+library(tidyverse)
+library(magrittr)
+library(splines)
+library(survival)
+library(timereg)
+
+bmtcrr %<>% filter(Age >= 16)
+
+bmtcrr %>% select(Sex) %>% table
+bmtcrr %>% select(D) %>% table
+bmtcrr %>% select(Phase) %>% table
+bmtcrr %>% select(Source) %>% table
+bmtcrr %>% select(Status) %>% table
+
+model_cb <- fitSmoothHazard(Status ~ bs(ftime, df = 5) + Sex + D + Phase + Source + Age,
+                            data = bmtcrr, time = "ftime")
+model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) + 
+                          const(Phase) + const(Source) + const(Age), 
+                      data = bmtcrr, cause = 1, model = "fg")
+model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
+                   data = bmtcrr)
+
+time_points <- (0:100)*60/100
+
+# ALL vs. AML
+newdata <- data.frame("Sex" = factor(c("F", "F"), 
+                                     levels = c("F", "M")),
+                      "D" = c("ALL", "AML"),
+                      "Phase" = factor(c("Relapse", "Relapse"), 
+                                       levels = c("CR1", "CR2", "CR3", "Relapse")),
+                      "Age" = c(35, 35),
+                      "Source" = factor(c("PB", "PB"), 
+                                        levels = c("BM+PB", "PB")))
+
+risk_cb <- absoluteRisk(object = model_cb, time = time_points,
+                        method = "montecarlo", newdata = newdata)
+risk_cb <- bind_rows(data.frame(Time = time_points, Method = "Case-base", Risk = risk_cb[,2], Disease = "ALL",
+                                stringsAsFactors = FALSE),
+                     data.frame(Time = time_points, Method = "Case-base", Risk = risk_cb[,3], Disease = "AML",
+                                stringsAsFactors = FALSE))
+
+risk_fg <- predict(model_fg, newdata, times = time_points)
+risk_fg <- bind_rows(data.frame(Time = time_points, Method = "Fine-Gray", Risk = risk_fg$P1[1,], Disease = "ALL",
+                                stringsAsFactors = FALSE),
+                     data.frame(Time = time_points, Method = "Fine-Gray", Risk = risk_fg$P1[2,], Disease = "AML",
+                                stringsAsFactors = FALSE))
+risk_km <- map_df(c("ALL", "AML"), function(disease) {
+    foo <- bmtcrr %>% 
+        filter(D == disease) %>% 
+        survfit(Surv(ftime,Status == 1) ~ 1, data = .)
+    data.frame(Time = foo$time, Method = "Kaplan-Meier", Risk = 1 - foo$surv, Disease = disease,
+               stringsAsFactors = FALSE) %>%
+        filter(Time <= 60)
+})
+
+disease_names <- c("ALL" = "Acute Lymphoid Leukemia",
+                   "AML" = "Acute Myeloid Leukemia")
+risk_cb %>%
+    bind_rows(risk_fg,
+              risk_km) %>% 
+    arrange(Disease, Risk) %>% 
+    ggplot(aes(Time, Risk, colour = Method)) + 
+    geom_line(data = . %>% filter(Method == "Case-base")) + 
+    geom_step(data = . %>% filter(Method != "Case-base")) + 
+    facet_grid(Disease ~ ., labeller = as_labeller(disease_names)) + ylim(c(0,1)) +
+    theme(legend.position = "top") +
+    xlab("Time (in Months)") + ylab("Relapse risk")
+```
+
+```{r}
+# Table of coefficients----
+z_value <- qnorm(0.975)
+table_cb <- summary(model_cb)@coef3[seq(13, 25, by = 2), 1:2]
+table_cb <- cbind(table_cb[,1], 
+                  table_cb[,1] - z_value * table_cb[,2],
+                  table_cb[,1] + z_value * table_cb[,2])
+table_cb <- round(exp(table_cb), 2)
+
+table_cox <- round(summary(model_cox)$conf.int[,-2], 2)
+```
 
@@
 # Discussion
+
+# References {-}
+<!-- placeholder for References in toc --!>

### 2017-09-16 paper/casebase_jss.tex
add bibtex file for references
add bibtex file for references
change plots to poptime

@@
 
+We will use the same data that was used in Scrucca \emph{et al}
+\citep{scrucca2010regression}. The data is available on the main
+author's \href{http://www.stat.unipg.it/luca/R/}{website}; it is also
+available as part of this package.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> library(casebase)
+\end{CodeInput}
+
+\begin{CodeOutput}
+See example usage at http://sahirbhatnagar.com/casebase/
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> library(magrittr)
+R> library(tidyverse)
+\end{CodeInput}
+
+\begin{CodeOutput}
+Loading tidyverse: ggplot2
+Loading tidyverse: tibble
+Loading tidyverse: tidyr
+Loading tidyverse: readr
+Loading tidyverse: purrr
+Loading tidyverse: dplyr
+\end{CodeOutput}
+
+\begin{CodeOutput}
+Conflicts with tidy packages ----------------------------------------------
+\end{CodeOutput}
+
+\begin{CodeOutput}
+filter(): dplyr, stats
+lag():    dplyr, stats
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> data(bmtcrr)
+R> head(bmtcrr) %>% knitr::kable(format = "latex")
+\end{CodeInput}
+
+\begin{tabular}{l|l|l|r|r|l|r}
+\hline
+Sex & D & Phase & Age & Status & Source & ftime\\
+\hline
+M & ALL & Relapse & 48 & 2 & BM+PB & 0.67\\
+\hline
+F & AML & CR2 & 23 & 1 & BM+PB & 9.50\\
+\hline
+M & ALL & CR3 & 7 & 0 & BM+PB & 131.77\\
+\hline
+F & ALL & CR2 & 26 & 2 & BM+PB & 24.03\\
+\hline
+F & ALL & CR2 & 36 & 2 & BM+PB & 1.47\\
+\hline
+M & ALL & Relapse & 17 & 2 & BM+PB & 2.23\\
+\hline
+\end{tabular}
+
+\end{CodeChunk}
+
+We will perform a competing risk analysis on data from 177 patients who
+received a stem cell transplant for acute leukemia. The event of
+interest in relapse, but other competing causes (e.g.~transplant-related
+death) need to be taken into account. We also want to take into account
+the effect of several covariates such as Sex, Disease (lymphoblastic or
+myeloblastic leukemia, abbreviated as ALL and AML, respectively), Phase
+at transplant (Relapse, CR1, CR2, CR3), Source of stem cells (bone
+marrow and peripheral blood, coded as BM+PB, or peripheral blood, coded
+as PB), and Age. Below, we reproduce their Table 1:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> table1 <- tibble::tribble(
+R+     ~Variable, ~Description, ~`Statistical summary`,
+R+     "Sex", "Sex", "M=Male (100)",
+R+     "", "", "F=Female (77)",
+R+     "D", "Disease", "ALL (73)",
+R+     "", "", "AML (104)",
+R+     "Phase", "Phase", "CR1 (47)", 
+R+     "", "", "CR2 (45)", 
+R+     "", "", "CR3 (12)", 
+R+     "", "", "Relapse (73)",
+R+     "Source", "Type of transplant", "BM+PB (21)", 
+R+     "", "", "PB (156)",
+R+     "Age", "Age of patient (years)", "4–62", 
+R+     "", "", "30.47 (13.04)",
+R+     "Ftime", "Failure time (months)", "0.13–131.77",
+R+     "", "", "20.28 (30.78)",
+R+     "Status", "Status indicator", "0=censored (46)", 
+R+     "", "", "1=relapse (56)", 
+R+     "", "", "2=competing event (75)"
+R+ )
+R> 
+R> knitr::kable(table1, format = "latex")
+\end{CodeInput}
+
+\begin{tabular}{l|l|l}
+\hline
+Variable & Description & Statistical summary\\
+\hline
+Sex & Sex & M=Male (100)\\
+\hline
+ &  & F=Female (77)\\
+\hline
+D & Disease & ALL (73)\\
+\hline
+ &  & AML (104)\\
+\hline
+Phase & Phase & CR1 (47)\\
+\hline
+ &  & CR2 (45)\\
+\hline
+ &  & CR3 (12)\\
+\hline
+ &  & Relapse (73)\\
+\hline
+Source & Type of transplant & BM+PB (21)\\
+\hline
+ &  & PB (156)\\
+\hline
+Age & Age of patient (years) & 4–62\\
+\hline
+ &  & 30.47 (13.04)\\
+\hline
+Ftime & Failure time (months) & 0.13–131.77\\
+\hline
+ &  & 20.28 (30.78)\\
+\hline
+Status & Status indicator & 0=censored (46)\\
+\hline
+ &  & 1=relapse (56)\\
+\hline
+ &  & 2=competing event (75)\\
+\hline
+\end{tabular}
+
+\end{CodeChunk}
+
+The statistical summary is generated differently for continuous and
+categorical variables:
+
 \begin{itemize}
-\tightlist
 \item
-  Competing risk example
+  For continuous variables, we are given the range, followed by the mean
+  and standard deviation.
 \item
-  Comparison with Cox and Fine-Gray
+  For categorical variables, we are given the counts for each category.
 \end{itemize}
+
+Note that failure time can also correspond to censoring.
+
+In order to try and visualize the incidence density of relapse, we can
+look at a population-time plot: on the X-axis we have time, and on the
+Y-axis we have the size of the risk set at a particular time point.
+Failure times associated to the event of interest can then be
+highlighted on the plot using red dots.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> pt_object <- casebase::popTime(bmtcrr, event = "Status", time = "ftime")
+R> plot(pt_object)
+\end{CodeInput}
+
+\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-3-1} \end{center}
+
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> pt_object_comp <- bmtcrr %>% 
+R+     mutate(Status = case_when(Status == 1 ~ 2L, 
+R+                               Status == 2 ~ 1L, 
+R+                               TRUE ~ Status)) %>% 
+R+     casebase::popTime(event = "Status", time = "ftime")
+R> plot(pt_object_comp, point.colour = "blue")
+\end{CodeInput}
+
+\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-4-1} \end{center}
+
+\end{CodeChunk}
+
+From this last plot, we can see that there is no censoring during the
+first 10 months. Moreover, we see that the last competing event occurs
+around 20 months. Putting all this information together, we have
+evidence of two types of patients: very sick patients who either relapse
+or have a competing event early on, and healthier patients who are
+eventually lost to follow-up.
+
+We now turn to the analysis of this dataset. The population-time plots
+above give evidence of non-constant hazard; therefore, we will
+explicitely include time in the model. Note that we also include all
+other variables as possible confounders. First, we include time as a
+linear term:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> model1 <- fitSmoothHazard(Status ~ ftime + Sex + D + Phase + Source + Age, 
+R+                           data = bmtcrr, 
+R+                           ratio = 100,
+R+                           time = "ftime")
+R> summary(model1)
+\end{CodeInput}
+
+\begin{CodeOutput}
+
+Call:
+vglm(formula = formula, family = multinomial(refLevel = 1), data = sampleData)
+
+Pearson residuals:
+                       Min       1Q   Median        3Q   Max
+log(mu[,2]/mu[,1]) -0.2324 -0.06980 -0.03900 -0.014615 45.18
+log(mu[,3]/mu[,1]) -0.3028 -0.09008 -0.03748 -0.008052 19.79
+
+Coefficients: 
+                Estimate Std. Error z value Pr(>|z|)    
+(Intercept):1  -3.473159   0.684472  -5.074 3.89e-07 ***
+(Intercept):2  -2.483719   0.465134  -5.340 9.31e-08 ***
+ftime:1        -0.069515   0.014754  -4.712 2.46e-06 ***
+ftime:2        -0.104482   0.018274  -5.717 1.08e-08 ***
+SexM:1         -0.298211   0.282266  -1.056 0.290744    
+SexM:2         -0.443015   0.236317  -1.875 0.060839 .  
+DAML:1         -0.580429   0.301271  -1.927 0.054029 .  
+DAML:2         -0.114156   0.274610  -0.416 0.677629    
+PhaseCR2:1      0.119138   0.466642   0.255 0.798485    
+PhaseCR2:2      0.213409   0.331729   0.643 0.520015    
+PhaseCR3:1      0.446436   0.690259   0.647 0.517783    
+PhaseCR3:2      0.151515   0.526646   0.288 0.773577    
+PhaseRelapse:1  1.361336   0.390162   3.489 0.000485 ***
+PhaseRelapse:2  0.657054   0.307296   2.138 0.032502 *  
+SourcePB:1      0.570967   0.564054   1.012 0.311415    
+SourcePB:2     -0.941242   0.352134  -2.673 0.007518 ** 
+Age:1          -0.008575   0.011883  -0.722 0.470504    
+Age:2           0.024955   0.009868   2.529 0.011442 *  
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+
+Number of linear predictors:  2
+
+Names of linear predictors: log(mu[,2]/mu[,1]), log(mu[,3]/mu[,1])
+
+Residual deviance: 1412.245 on 26444 degrees of freedom
+
+Log-likelihood: -706.1222 on 26444 degrees of freedom
+
+Number of iterations: 10
+
+Reference group is level  1  of the response
+\end{CodeOutput}
+\end{CodeChunk}
+
+Because of the results in Turgeon \emph{et al} \citep{turgeonCompRisk},
+the standard errors we obtain from the multinomial logit fit are
+asymptotically correct, and therefore can be used to construct
+asymptotic confidence intervals.
+
+From this summary, we see that time is indeed significant, as is Phase
+(only relapse vs.~CR1). Interestingly, we see that the type of disease
+is only significant for the event of interest, whereas the type of
+transplant and the age of the patient are only significant for the
+competing event.
+
+Next, we include the logarithm of time in the model (which leads to a
+Weibull hazard):
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> model2 <- fitSmoothHazard(Status ~ log(ftime) + Sex + D + Phase + Source + Age, 
+R+                           data = bmtcrr, 
+R+                           ratio = 100, 
+R+                           time = "ftime")
+R> summary(model2)
+\end{CodeInput}
+
+\begin{CodeOutput}
+
+Call:
+vglm(formula = formula, family = multinomial(refLevel = 1), data = sampleData)
+
+Pearson residuals:
+                       Min       1Q   Median       3Q   Max
+log(mu[,2]/mu[,1]) -0.4301 -0.06839 -0.04673 -0.03529 29.47
+log(mu[,3]/mu[,1]) -0.6313 -0.07728 -0.05488 -0.04460 22.07
+
+Coefficients: 
+                Estimate Std. Error z value Pr(>|z|)    
+(Intercept):1  -3.867645   0.712859  -5.426 5.78e-08 ***
+(Intercept):2  -2.915708   0.470010  -6.203 5.52e-10 ***
+log(ftime):1   -0.345600   0.071298  -4.847 1.25e-06 ***
+log(ftime):2   -0.426302   0.057776  -7.379 1.60e-13 ***
+SexM:1         -0.410838   0.293247  -1.401 0.161215    
+SexM:2         -0.482041   0.243193  -1.982 0.047465 *  
+DAML:1         -0.689708   0.302498  -2.280 0.022605 *  
+DAML:2         -0.165629   0.285741  -0.580 0.562152    
+PhaseCR2:1      0.264372   0.466874   0.566 0.571216    
+PhaseCR2:2      0.389075   0.330864   1.176 0.239620    
+PhaseCR3:1      0.416963   0.717569   0.581 0.561189    
+PhaseCR3:2      0.035184   0.539420   0.065 0.947994    
+PhaseRelapse:1  1.497807   0.391712   3.824 0.000131 ***
+PhaseRelapse:2  0.900626   0.306158   2.942 0.003264 ** 
+SourcePB:1      0.525555   0.606178   0.867 0.385942    
+SourcePB:2     -1.129794   0.370857  -3.046 0.002316 ** 
+Age:1          -0.002345   0.011604  -0.202 0.839836    
+Age:2           0.029056   0.009815   2.960 0.003072 ** 
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+
+Number of linear predictors:  2
+
+Names of linear predictors: log(mu[,2]/mu[,1]), log(mu[,3]/mu[,1])
+
+Residual deviance: 1502.709 on 26444 degrees of freedom
+
+Log-likelihood: -751.3548 on 26444 degrees of freedom
+
+Number of iterations: 8
+
+Reference group is level  1  of the response
+\end{CodeOutput}
+\end{CodeChunk}
+
+As we can see, the results are similar to the ones with a Gompertz
+hazard, although Sex is now significant for the competing event.
+
+Finally, using splines, we can be quite flexible about the way the
+hazard depends on time:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> model3 <- fitSmoothHazard(
+R+     Status ~ splines::bs(ftime) + Sex + D + Phase + Source + Age, 
+R+     data = bmtcrr, 
+R+     ratio = 100, 
+R+     time = "ftime")
+R> summary(model3)
+\end{CodeInput}
+
+\begin{CodeOutput}
+
+Call:
+vglm(formula = formula, family = multinomial(refLevel = 1), data = sampleData)
+
+Pearson residuals:
+                       Min       1Q   Median         3Q   Max
+log(mu[,2]/mu[,1]) -0.2032 -0.07093 -0.03950 -6.478e-03 57.42
+log(mu[,3]/mu[,1]) -0.2714 -0.09462 -0.01215 -3.633e-06 37.16
+
+Coefficients: 
+                        Estimate Std. Error z value Pr(>|z|)    
+(Intercept):1          -3.800875   0.713833  -5.325 1.01e-07 ***
+(Intercept):2          -3.255436   0.515685  -6.313 2.74e-10 ***
+splines::bs(ftime)1:1   0.603866   2.311386   0.261 0.793894    
+splines::bs(ftime)1:2   8.064772   3.745888   2.153 0.031321 *  
+splines::bs(ftime)2:1 -17.868972   8.390698  -2.130 0.033203 *  
+splines::bs(ftime)2:2 -82.148252  26.046430  -3.154 0.001611 ** 
+splines::bs(ftime)3:1  -2.012019   9.340686  -0.215 0.829453    
+splines::bs(ftime)3:2  -2.790995  22.345501  -0.125 0.900601    
+SexM:1                 -0.277399   0.284388  -0.975 0.329349    
+SexM:2                 -0.418228   0.237149  -1.764 0.077805 .  
+DAML:1                 -0.649732   0.304770  -2.132 0.033017 *  
+DAML:2                 -0.143387   0.278769  -0.514 0.607001    
+PhaseCR2:1              0.176145   0.467658   0.377 0.706431    
+PhaseCR2:2              0.340090   0.332192   1.024 0.305941    
+PhaseCR3:1              0.540857   0.697533   0.775 0.438112    
+PhaseCR3:2              0.241321   0.529642   0.456 0.648656    
+PhaseRelapse:1          1.503814   0.395120   3.806 0.000141 ***
+PhaseRelapse:2          0.874907   0.314308   2.784 0.005376 ** 
+SourcePB:1              0.444475   0.576583   0.771 0.440779    
+SourcePB:2             -1.174262   0.360405  -3.258 0.001121 ** 
+Age:1                  -0.005234   0.011979  -0.437 0.662179    
+Age:2                   0.028974   0.010077   2.875 0.004039 ** 
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+
+Number of linear predictors:  2
+
+Names of linear predictors: log(mu[,2]/mu[,1]), log(mu[,3]/mu[,1])
+
+Residual deviance: 1393.424 on 26440 degrees of freedom
+
+Log-likelihood: -696.7122 on 26440 degrees of freedom
+
+Number of iterations: 16
+
+Reference group is level  1  of the response
+\end{CodeOutput}
+\end{CodeChunk}
+
+Again, we see that the results are quite similar for this third model.
+
+We now look at the 2-year risk of relapse:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> linearRisk <- absoluteRisk(object = model1, time = 24, newdata = bmtcrr[1:10,])
+R> logRisk <- absoluteRisk(object = model2, time = 24, newdata = bmtcrr[1:10,])
+R> splineRisk <- absoluteRisk(object = model3, time = 24, newdata = bmtcrr[1:10,])
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> plot(linearRisk, logRisk,
+R+      xlab = "Linear", ylab = "Log/Spline", pch = 19,
+R+      xlim = c(0,1), ylim = c(0,1), col = 'red')
+R> points(linearRisk, splineRisk,
+R+        col = 'blue', pch = 19)
+R> abline(a = 0, b = 1, lty = 2, lwd = 2)
+R> legend("topleft", legend = c("Log", "Spline"),
+R+        pch = 19, col = c("red", "blue"))
+\end{CodeInput}
+
+\begin{center}\includegraphics{casebase_jss_files/figure-latex/absRiskPlot-1} \end{center}
+
+\end{CodeChunk}
+
+We can also estimate the mean absolute risk for the entire dataset:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> mean(linearRisk)
+\end{CodeInput}
+
+\begin{CodeOutput}
+[1] 0.1991813
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> mean(logRisk)
+\end{CodeInput}
+
+\begin{CodeOutput}
+[1] 0.1560315
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> mean(splineRisk)
+\end{CodeInput}
+
+\begin{CodeOutput}
+[1] 0.1087245
+\end{CodeOutput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Absolute risks----
+R> library(tidyverse)
+R> library(magrittr)
+R> library(splines)
+R> library(survival)
+R> library(timereg)
+R> 
+R> bmtcrr %<>% filter(Age >= 16)
+R> 
+R> bmtcrr %>% select(Sex) %>% table
+\end{CodeInput}
+
+\begin{CodeOutput}
+.
+ F  M 
+72 87 
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> bmtcrr %>% select(D) %>% table
+\end{CodeInput}
+
+\begin{CodeOutput}
+.
+ALL AML 
+ 59 100 
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> bmtcrr %>% select(Phase) %>% table
+\end{CodeInput}
+
+\begin{CodeOutput}
+.
+    CR1     CR2     CR3 Relapse 
+     44      40      10      65 
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> bmtcrr %>% select(Source) %>% table
+\end{CodeInput}
+
+\begin{CodeOutput}
+.
+BM+PB    PB 
+   15   144 
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> bmtcrr %>% select(Status) %>% table
+\end{CodeInput}
+
+\begin{CodeOutput}
+.
+ 0  1  2 
+40 49 70 
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> model_cb <- fitSmoothHazard(Status ~ bs(ftime, df = 5) + Sex + D + Phase + Source + Age,
+R+                             data = bmtcrr, time = "ftime")
+R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) + 
+R+                           const(Phase) + const(Source) + const(Age), 
+R+                       data = bmtcrr, cause = 1, model = "fg")
+R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
+R+                    data = bmtcrr)
+R> 
+R> time_points <- (0:100)*60/100
+R> 
+R> # ALL vs. AML
+R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
+R+                                      levels = c("F", "M")),
+R+                       "D" = c("ALL", "AML"),
+R+                       "Phase" = factor(c("Relapse", "Relapse"), 
+R+                                        levels = c("CR1", "CR2", "CR3", "Relapse")),
+R+                       "Age" = c(35, 35),
+R+                       "Source" = factor(c("PB", "PB"), 
+R+                                         levels = c("BM+PB", "PB")))
+R> 
+R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
+R+                         method = "montecarlo", newdata = newdata)
+R> risk_cb <- bind_rows(data.frame(Time = time_points, Method = "Case-base", Risk = risk_cb[,2], Disease = "ALL",
+R+                                 stringsAsFactors = FALSE),
+R+                      data.frame(Time = time_points, Method = "Case-base", Risk = risk_cb[,3], Disease = "AML",
+R+                                 stringsAsFactors = FALSE))
+R> 
+R> risk_fg <- predict(model_fg, newdata, times = time_points)
+R> risk_fg <- bind_rows(data.frame(Time = time_points, Method = "Fine-Gray", Risk = risk_fg$P1[1,], Disease = "ALL",
+R+                                 stringsAsFactors = FALSE),
+R+                      data.frame(Time = time_points, Method = "Fine-Gray", Risk = risk_fg$P1[2,], Disease = "AML",
+R+                                 stringsAsFactors = FALSE))
+R> risk_km <- map_df(c("ALL", "AML"), function(disease) {
+R+     foo <- bmtcrr %>% 
+R+         filter(D == disease) %>% 
+R+         survfit(Surv(ftime,Status == 1) ~ 1, data = .)
+R+     data.frame(Time = foo$time, Method = "Kaplan-Meier", Risk = 1 - foo$surv, Disease = disease,
+R+                stringsAsFactors = FALSE) %>%
+R+         filter(Time <= 60)
+R+ })
+R> 
+R> disease_names <- c("ALL" = "Acute Lymphoid Leukemia",
+R+                    "AML" = "Acute Myeloid Leukemia")
+R> risk_cb %>%
+R+     bind_rows(risk_fg,
+R+               risk_km) %>% 
+R+     arrange(Disease, Risk) %>% 
+R+     ggplot(aes(Time, Risk, colour = Method)) + 
+R+     geom_line(data = . %>% filter(Method == "Case-base")) + 
+R+     geom_step(data = . %>% filter(Method != "Case-base")) + 
+R+     facet_grid(Disease ~ ., labeller = as_labeller(disease_names)) + ylim(c(0,1)) +
+R+     theme(legend.position = "top") +
+R+     xlab("Time (in Months)") + ylab("Relapse risk")
+\end{CodeInput}
+
+\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-9-1} \end{center}
+
+\end{CodeChunk}
+
+\textbackslash{}begin\{CodeChunk\}
+
+\textbackslash{}begin\{CodeInput\} R\textgreater{} \# Table of
+coefficients---- R\textgreater{} z\_value \textless{}- qnorm(0.975)
+R\textgreater{} table\_cb \textless{}- summary(model\_cb)\citet[seq(13,
+25, by = 2), 1:2]{coef3} R\textgreater{} table\_cb \textless{}-
+cbind(table\_cb{[},1{]}, R+ table\_cb{[},1{]} - z\_value *
+table\_cb{[},2{]}, R+ table\_cb{[},1{]} + z\_value * table\_cb{[},2{]})
+R\textgreater{} table\_cb \textless{}- round(exp(table\_cb), 2)
+R\textgreater{} R\textgreater{} table\_cox \textless{}-
+round(summary(model\_cox)\$conf.int{[},-2{]}, 2)
+\textbackslash{}end\{CodeInput\} \textbackslash{}end\{CodeChunk\}
 
@@
 
+\section{Discussion}\label{discussion}
+
+\section*{References}\label{references}
+\addcontentsline{toc}{section}{References}
+
+\bibliography{casebase_references.bib}
+
 \end{document}

### 2017-09-18 paper/casebase_jss.tex
combine vignette with talk

--- 2017-09-15 paper/casebase_jss.tex
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Competing risk example
-\item
-  Comparison with Cox and Fine-Gray
-\end{itemize}
+The next example shows how case-base sampling can also be used in the
+context of a competing risk analysis. For illustrative purposes, we will
+use the same data that was used in Scrucca \emph{et al}
+\citeyearpar{scrucca2010regression}. The data was downloaded from the
+main author's website, and it is also available as part of the
+\pkg{casebase} package.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> library(casebase)
+R> data(bmtcrr)
+\end{CodeInput}
+\end{CodeChunk}
+
+The data contains information on 177 patients who received a stem-cell
+transplant for acute leukemia. The event of interest is relapse, but
+other competing causes (e.g.~transplant-related death) were also
+recorded Several covariates were also captured at baseline: sex, disease
+type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL
+and AML, respectively), disease phase at transplant (Relapse, CR1, CR2,
+CR3), source of stem cells (bone marrow and peripheral blood, coded as
+BM+PB, or only peripheral blood, coded as PB), and age. A summary of
+these baseline characteristics appear in Table \ref{tab:table1bmtcrr}.
+We note that the statistical summaries were generated differently for
+different variable types: for continuous variables, we gave the range,
+followed by the mean and standard deviation; for categorical variables,
+we gave the counts for each category.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{lll}
+  \hline
+Variable & Description & Statistical summary \\ 
+  \hline
+Sex & Sex & M=Male (100) \\ 
+   &  & F=Female (77) \\ 
+  D & Disease & ALL (73) \\ 
+   &  & AML (104) \\ 
+  Phase & Phase & CR1 (47) \\ 
+   &  & CR2 (45) \\ 
+   &  & CR3 (12) \\ 
+   &  & Relapse (73) \\ 
+  Source & Type of transplant & BM+PB (21) \\ 
+   &  & PB (156) \\ 
+  Age & Age of patient (years) & 4–62 \\ 
+   &  & 30.47 (13.04) \\ 
+  Ftime & Failure time (months) & 0.13–131.77 \\ 
+   &  & 20.28 (30.78) \\ 
+  Status & Status indicator & 0=censored (46) \\ 
+   &  & 1=relapse (56) \\ 
+   &  & 2=competing event (75) \\ 
+   \hline
+\end{tabular}
+\caption{Baseline characteristics of patients in the stem-cell transplant study.}
+\label{tab:table1bmtcrr}
+\end{table}
+
+In order to try and visualize the incidence density of relapse, we can
+look at the corresponding population-time plot. In Figure
+\ref{fig:compPop1}, failure times associated with relapse are
+highlighted on the plot using red points, while Figure
+\ref{fig:compPop2} provides a similar population-time plot for competing
+events.
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{casebase_jss_files/figure-latex/compPop1-1}
+
+}
+
+\caption{\label{fig:compPop1}Population-time plot for the stem-cell transplant study. The points represent the event of interest (i.e., relapse).}\label{fig:compPop1}
+\end{figure}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{casebase_jss_files/figure-latex/compPop2-1}
+
+}
+
+\caption{\label{fig:compPop2}Population-time plot for the stem-cell transplant study. The points represent the competing events.}\label{fig:compPop2}
+\end{figure}
+\end{CodeChunk}
+
+Our main objective is to compute the absolute risk of relapse for a
+given set of covariates. First, we fit a smooth hazard to the data; for
+the sake of this example, we opted for a linear term for time:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> model_cb <- fitSmoothHazard(
+R+     Status ~ ftime + Sex + D + Phase + Source + Age, 
+R+     data = bmtcrr, 
+R+     ratio = 100, 
+R+     time = "ftime")
+\end{CodeInput}
+\end{CodeChunk}
+
+From the fit object, we can extract both the hazard ratios and their
+corresponding confidence intervals:
+
+\begin{CodeChunk}
+
+\begin{tabular}{l|r|l}
+\hline
+Covariates & HR & 95\% CI\\
+\hline
+Sex & 0.71 & (0.41, 1.24)\\
+\hline
+Disease & 0.52 & (0.28, 0.93)\\
+\hline
+Phase (CR2 vs. CR1) & 1.17 & (0.47, 2.93)\\
+\hline
+Phase (CR3 vs. CR1) & 1.78 & (0.46, 6.91)\\
+\hline
+Phase (Relapse vs. CR1) & 4.45 & (2.07, 9.57)\\
+\hline
+Source & 1.64 & (0.54, 5)\\
+\hline
+Age & 0.99 & (0.97, 1.02)\\
+\hline
+\end{tabular}
+
+\end{CodeChunk}
+
+As we can see, the only significant hazard ratio is the one associated
+with the phase of the disease at transplant. More precisely, being in
+relapse at transplant is associated with a hazard ratio of 3.92 when
+compared to CR1.
+
+Given our estimate of the hazard function, we can compute the absolute
+risk curve for a fixed covariate profile. We performed this computation
+for a 35 year old woman who received a stem-cell transplant from
+peripheral blood at relapse. We compared the absolute risk curve for
+such a woman with acute lymphoblastic leukemia with that for a similar
+woman with acute myeloblastic leukemia. Figure \ref{fig:compAbsrisk}
+shows these two curves as a function of time. This figure also shows the
+Kaplan-Meier estimate fitted to the two disease groups (ignoring the
+other covariates).
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Pick 100 equidistant points between 0 and 60 months
+R> time_points <- seq(0, 60, length.out = 100)
+R> 
+R> # Data.frame containing risk profile
+R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
+R+                                      levels = levels(bmtcrr[,"Sex"])),
+R+                       "D" = c("ALL", "AML"),
+R+                       "Phase" = factor(c("Relapse", "Relapse"), 
+R+                                        levels = levels(bmtcrr[,"Phase"])),
+R+                       "Age" = c(35, 35),
+R+                       "Source" = factor(c("PB", "PB"), 
+R+                                         levels = levels(bmtcrr[,"Source"])))
+R> 
+R> # Estimate absolute risk curve
+R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
+R+                         method = "montecarlo", newdata = newdata)
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-6-1}
+
+}
+
+\caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups.
+The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:unnamed-chunk-6}
+\end{figure}
+\end{CodeChunk}
 
@@
 
+\section{Discussion}\label{discussion}
+
+\bibliography{casebase_references.bib}
+
 \end{document}

### 2017-09-18 paper/casebase_jss.Rmd

--- 2017-09-15 paper/casebase_jss.Rmd
@@
   plain:     [keywords, not capitalized, Java]
+bibliography: casebase_references.bib
+biblio-style: jss
+csl: jss.csl
 preamble: >
   \usepackage{amsmath}
-output: rticles::jss_article
+output: 
+  rticles::jss_article:
+    fig_caption: true
 ---
@@
 * \code{print("abc")}
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
+library(tidyverse)
+library(magrittr)
+```
 
@@
 
-  - Competing risk example
-  - Comparison with Cox and Fine-Gray
+The next example shows how case-base sampling can also be used in the context of a competing risk analysis.
+For illustrative purposes, we will use the same data that was used in Scrucca *et al* [-@scrucca2010regression].
+The data was downloaded from the main author's website, and it is also available as part of the \pkg{casebase} package.
+
+```{r message = FALSE}
+library(casebase)
+data(bmtcrr)
+```
+
+The data contains information on 177 patients who received a stem-cell transplant for acute leukemia.
+The event of interest is relapse, but other competing causes (e.g.
+transplant-related death) were also recorded Several covariates were also captured at baseline: sex, disease type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL and AML, respectively), disease phase at transplant (Relapse, CR1, CR2, CR3), source of stem cells (bone marrow and peripheral blood, coded as BM+PB, or only peripheral blood, coded as PB), and age.
+A summary of these baseline characteristics appear in Table \ref{tab:table1bmtcrr}.
+We note that the statistical summaries were generated differently for different variable types: for continuous variables, we gave the range, followed by the mean and standard deviation; for categorical variables, we gave the counts for each category.
+
+```{r echo=FALSE, eval = FALSE}
+table1 <- tibble::tribble(
+    ~Variable, ~Description, ~`Statistical summary`,
+    "Sex", "Sex", "M=Male (100)",
+    "", "", "F=Female (77)",
+    "D", "Disease", "ALL (73)",
+    "", "", "AML (104)",
+    "Phase", "Phase", "CR1 (47)", 
+    "", "", "CR2 (45)", 
+    "", "", "CR3 (12)", 
+    "", "", "Relapse (73)",
+    "Source", "Type of transplant", "BM+PB (21)", 
+    "", "", "PB (156)",
+    "Age", "Age of patient (years)", "4–62", 
+    "", "", "30.47 (13.04)",
+    "Ftime", "Failure time (months)", "0.13–131.77",
+    "", "", "20.28 (30.78)",
+    "Status", "Status indicator", "0=censored (46)", 
+    "", "", "1=relapse (56)", 
+    "", "", "2=competing event (75)"
+)
+
+xtable::xtable(table1) %>%
+    print(include.rownames = FALSE)
+```
+
+<!-- latex table generated in R 3.4.1 by xtable 1.8-2 package -->
+<!-- Sun Sep 17 16:03:22 2017 -->
+\begin{table}[ht]
+\centering
+\begin{tabular}{lll}
+  \hline
+Variable & Description & Statistical summary \\ 
+  \hline
+Sex & Sex & M=Male (100) \\ 
+   &  & F=Female (77) \\ 
+  D & Disease & ALL (73) \\ 
+   &  & AML (104) \\ 
+  Phase & Phase & CR1 (47) \\ 
+   &  & CR2 (45) \\ 
+   &  & CR3 (12) \\ 
+   &  & Relapse (73) \\ 
+  Source & Type of transplant & BM+PB (21) \\ 
+   &  & PB (156) \\ 
+  Age & Age of patient (years) & 4–62 \\ 
+   &  & 30.47 (13.04) \\ 
+  Ftime & Failure time (months) & 0.13–131.77 \\ 
+   &  & 20.28 (30.78) \\ 
+  Status & Status indicator & 0=censored (46) \\ 
+   &  & 1=relapse (56) \\ 
+   &  & 2=competing event (75) \\ 
+   \hline
+\end{tabular}
+\caption{Baseline characteristics of patients in the stem-cell transplant study.}
+\label{tab:table1bmtcrr}
+\end{table}
+
+In order to try and visualize the incidence density of relapse, we can look at the corresponding population-time plot.
+In Figure \ref{fig:compPop1}, failure times associated with relapse are highlighted on the plot using red points, while Figure \ref{fig:compPop2} provides a similar population-time plot for competing events.
+
+```{r compPop1, fig.cap="\\label{fig:compPop1}Population-time plot for the stem-cell transplant study. The points represent the event of interest (i.e., relapse).", echo=FALSE}
+pt_object <- casebase::popTime(bmtcrr, event = "Status", time = "ftime")
+plot(pt_object)
+```
+
+```{r compPop2, fig.cap="\\label{fig:compPop2}Population-time plot for the stem-cell transplant study. The points represent the competing events.", echo=FALSE}
+pt_object_comp <- bmtcrr %>% 
+    mutate(Status = case_when(Status == 1 ~ 2L, 
+                              Status == 2 ~ 1L, 
+                              TRUE ~ Status)) %>% 
+    casebase::popTime(event = "Status", time = "ftime")
+plot(pt_object_comp, point.colour = "blue")
+```
+
+Our main objective is to compute the absolute risk of relapse for a given set of covariates.
+First, we fit a smooth hazard to the data; for the sake of this example, we opted for a linear term for time:
+
+```{r warning = FALSE}
+model_cb <- fitSmoothHazard(
+    Status ~ ftime + Sex + D + Phase + Source + Age, 
+    data = bmtcrr, 
+    ratio = 100, 
+    time = "ftime")
+```
+
+```{r, echo = FALSE, eval = TRUE}
+library(survival)
+# Treat competing event as censoring
+model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
+                   data = bmtcrr)
+```
+
+From the fit object, we can extract both the hazard ratios and their corresponding confidence intervals:
+
+```{r, echo=FALSE}
+# Table of coefficients----
+z_value <- qnorm(0.975)
+foo <- summary(model_cb)@coef3
+table_cb <- foo[!grepl("Intercept", rownames(foo)) & 
+                    !grepl("ftime", rownames(foo)) &
+                    grepl(":1", rownames(foo)), 
+                1:2]
+table_cb <- cbind(table_cb[,1], 
+                  table_cb[,1] - z_value * table_cb[,2],
+                  table_cb[,1] + z_value * table_cb[,2])
+table_cb <- round(exp(table_cb), 2)
+
+table_cox <- round(summary(model_cox)$conf.int[,-2], 2)
+rownames(table_cb) <- rownames(table_cox)
+colnames(table_cb) <- colnames(table_cox)
+
+table_cb %<>% as.data.frame %>% 
+    mutate(Covariates = rownames(.)) %>% 
+    mutate(`95% CI` = glue::glue_data(., "({`lower .95`}, {`upper .95`})")) %>% 
+    rename(HR = `exp(coef)`) %>% 
+    select(Covariates, HR, `95% CI`)
+
+# table_cox %<>% as.data.frame %>% 
+#     mutate(Covariates = rownames(.)) %>% 
+#     mutate(`CI (Cox)` = glue::glue_data(., "({`lower .95`}, {`upper .95`})")) %>% 
+#     rename(`HR (Cox)` = `exp(coef)`) %>% 
+#     select(Covariates, `HR (Cox)`, `CI (Cox)`)
+
+table_cb %>% 
+# inner_join(table_cb, table_cox, by = "Covariates") %>%
+    mutate(Covariates = case_when(
+        Covariates == "SexM" ~ "Sex",
+        Covariates == "DAML" ~ "Disease",
+        Covariates == "PhaseCR2" ~ "Phase (CR2 vs. CR1)",
+        Covariates == "PhaseCR3" ~ "Phase (CR3 vs. CR1)",
+        Covariates == "PhaseRelapse" ~ "Phase (Relapse vs. CR1)",
+        Covariates == "SourcePB" ~ "Source",
+        TRUE ~ Covariates
+    )) %>% 
+    knitr::kable(format = "latex")
+```
+
+As we can see, the only significant hazard ratio is the one associated with the phase of the disease at transplant.
+More precisely, being in relapse at transplant is associated with a hazard ratio of 3.92 when compared to CR1.
+
+Given our estimate of the hazard function, we can compute the absolute risk curve for a fixed covariate profile.
+We performed this computation for a 35 year old woman who received a stem-cell transplant from peripheral blood at relapse.
+We compared the absolute risk curve for such a woman with acute lymphoblastic leukemia with that for a similar woman with acute myeloblastic leukemia.
+Figure \ref{fig:compAbsrisk} shows these two curves as a function of time.
+This figure also shows the Kaplan-Meier estimate fitted to the two disease groups (ignoring the other covariates).
+
+```{r cb_risk, warning = FALSE, cache = TRUE}
+# Pick 100 equidistant points between 0 and 60 months
+time_points <- seq(0, 60, length.out = 100)
+
+# Data.frame containing risk profile
+newdata <- data.frame("Sex" = factor(c("F", "F"), 
+                                     levels = levels(bmtcrr[,"Sex"])),
+                      "D" = c("ALL", "AML"),
+                      "Phase" = factor(c("Relapse", "Relapse"), 
+                                       levels = levels(bmtcrr[,"Phase"])),
+                      "Age" = c(35, 35),
+                      "Source" = factor(c("PB", "PB"), 
+                                        levels = levels(bmtcrr[,"Source"])))
+
+# Estimate absolute risk curve
+risk_cb <- absoluteRisk(object = model_cb, time = time_points,
+                        method = "montecarlo", newdata = newdata)
+```
+
+```{r fg_risk, eval = FALSE, echo = FALSE}
+library(timereg)
+model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) + 
+                          const(Phase) + const(Source) + const(Age), 
+                      data = bmtcrr, cause = 1, model = "fg")
+
+# Estimate absolute risk curve
+risk_fg <- predict(model_fg, newdata, times = time_points)
+```
+
+```{r echo = FALSE, fig.cap="\\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups.
+The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate."}
+risk_cb <- bind_rows(
+    data.frame(Time = time_points, 
+               Method = "Case-base", 
+               Risk = risk_cb[,2], 
+               Disease = "ALL",
+               stringsAsFactors = FALSE),
+    data.frame(Time = time_points, 
+               Method = "Case-base", 
+               Risk = risk_cb[,3], 
+               Disease = "AML",
+               stringsAsFactors = FALSE)
+    )
+
+# risk_fg <- bind_rows(
+#     data.frame(Time = time_points, 
+#                Method = "Fine-Gray", 
+#                Risk = risk_fg$P1[1,], 
+#                Disease = "ALL",
+#                stringsAsFactors = FALSE),
+#     data.frame(Time = time_points, 
+#                Method = "Fine-Gray", 
+#                Risk = risk_fg$P1[2,], 
+#                Disease = "AML",
+#                stringsAsFactors = FALSE)
+#     )
+
+risk_km <- purrr::map_df(c("ALL", "AML"), 
+                         function(disease) {
+    tmp <- bmtcrr %>% 
+        filter(D == disease) %>% 
+        survival::survfit(survival::Surv(ftime,Status == 1) ~ 1, 
+                data = .)
+    data.frame(Time = tmp$time, 
+               Method = "Kaplan-Meier", 
+               Risk = 1 - tmp$surv, 
+               Disease = disease,
+               stringsAsFactors = FALSE) %>%
+        filter(Time <= 60)
+})
+
+risk_all <- bind_rows(
+    risk_cb,
+    risk_km
+    ) %>% arrange(Disease, Risk)
+
+ggplot(risk_all, aes(x = Time, y = Risk, colour = Method)) + 
+    # geom_line for smooth curve
+    geom_line(data = filter(risk_all, Method == "Case-base")) + 
+    # geom_step for step function
+    geom_step(data = filter(risk_all, Method != "Case-base")) + 
+    facet_grid(Disease ~ .) +
+    ylim(c(0,1)) +
+    theme(legend.position = "top") +
+    xlab("Time (in Months)") + ylab("Relapse risk")
+```
 

### 2017-09-18 paper/casebase_jss.Rmd
add implementation details

@@
 
-  1. Population-tim plots
-  2. Sampling
-  3. Fitting
-  4. Absolute Risks
+The functions in the casebase package can be divided into two categories: 1) data visualization, in the form of population-time plots; and 2) data analysis.
 
-# Population-time plots
+We explicitely aimed at being compatible with both \code{data.frame}s and \code{data.table}s. This is evident in some of the coding choices we made, and it is also reflected in our testing units.
+
+## Population-time plots
+
+## Data analysis
+
+The data analysis step was separated into three parts: 1) case-base sampling; 2) estimation of the smooth hazard function; and 3) calculation of the risk function.
+By separating the sampling and estimation functions, we allowed the possibility of users implementing more complex sampling scheme, as described in Saarela.
+
+The sampling scheme selected for \code{sampleCaseBase} was described in Hanley and Miettinen: we first sample along the ``person'' axis, proportional to each individual's total follow-up time, and then we sample a moment uniformly over their follow-up time.
+This sampling scheme is equivalent to the following picture: imagine representing the total follow-up time of all individuals in the study along a single dimension, where the follow-up time of the next individual would start exactly when the follow-up time of the previous individual ends.
+Then the base series could be sampled uniformly from this one-dimensional representation of the overall follow-up time.
+In any case, the output is a dataset of the same class as the input, where each row corresponds to a person-moment.
+The covariate profile for each such person-moment is retained, and an offset term is added to the dataset.
+This output could then be used to fit a smooth hazard function, or for visualization of the base series.
+
+The fitting function \code{fitSmoothHazard} starts by looking at the class of the dataset: if it was generated from \code{sampleCaseBase}, it automatically inherited the class \code{cbData}.
+If the dataset supplied to \code{fitSmoothHazard} does not inherit from \code{cbData}, then the fitting function starts by calling \code{sampleCaseBase} to generate the base series.
+In other words, the occasional user can bypass \code{sampleCaseBase} altogether and only worry about the fitting function \code{fitSmoothHazard}.
+
+The fitting function retains the familiar formula interface of \code{glm}.
+The left-hand side of the formula should be the name of the column corresponding to the event type.
+The right-hand side can be any combination of the covariates, along with an explicit functional form for the time variable.
+Note that non-proportional hazard models can be achieved at this stage by adding an interaction term involving time.
+The offset term does not need to be specified by the user, as it is automatically added to the formula.
+
+Finally, the hazard function is fitted to the data using the function \code{glm}, unless there is more than one type event (i.e.
+in a competing-risk analysis), in which case it uses the multinomical regression capabilities of the \pkg{VGAM} package.
+This package was selected for its ability to fit multinomial regression models with an offset.
+
+Once a model-fit object has been returned by \code{fitSmoothHazard}, all the familiar summary and diagnostic functions are available: \code{print}, \code{summary}, \code{predict}, \code{plot}, etc.
+Our package provides one more functionality: it computes risk functions from the model fit.
+For the case of a single event, it uses the familiar identity
+$$S(t) = \exp\left(-\int_0^t h(u;X) du\right).$$
+The integral is computed using either the \code{stats::integrate} function or Monte-Carlo integration.
+For the case of a competing-event analysis, the event-specific risk is computed using a nested double integral; in this setting, Monte-Carlo integration is faster and more flexible than \code{stats::integrate}.
+This is due to the fact that the computation involves a double loop over the selected time points; Monte-Carlo integration performs this step by using the vectorized \code{rowSums} and \code{colSums} functions.
+
+To decide between a single-event and a competing-event analysis, we created \code{absoluteRisk} as an \code{S3} generic, with methods for both \code{glm} and \code{CompRisk} objects (the latter inherits from \code{vglm} as well).
+The method dispatch system of \proglang{R} then takes care of matching the correct input to the correct methodology, without the user's intervention.
 

### 2017-09-18 paper/casebase_jss.Rmd

@@
 The data analysis step was separated into three parts: 1) case-base sampling; 2) estimation of the smooth hazard function; and 3) calculation of the risk function.
-By separating the sampling and estimation functions, we allowed the possibility of users implementing more complex sampling scheme, as described in Saarela.
+By separating the sampling and estimation functions, we allowed the possibility of users implementing more complex sampling scheme, as described in Saarela [-@saarela2016case].
 
-The sampling scheme selected for \code{sampleCaseBase} was described in Hanley and Miettinen: we first sample along the ``person'' axis, proportional to each individual's total follow-up time, and then we sample a moment uniformly over their follow-up time.
+The sampling scheme selected for \code{sampleCaseBase} was described in Hanley and Miettinen [-@hanley2009fitting]: we first sample along the ``person'' axis, proportional to each individual's total follow-up time, and then we sample a moment uniformly over their follow-up time.
 This sampling scheme is equivalent to the following picture: imagine representing the total follow-up time of all individuals in the study along a single dimension, where the follow-up time of the next individual would start exactly when the follow-up time of the previous individual ends.

### 2017-09-18 paper/casebase_jss.tex

@@
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Population-tim plots
-\item
-  Sampling
-\item
-  Fitting
-\item
-  Absolute Risks
-\end{enumerate}
+The functions in the casebase package can be divided into two
+categories: 1) data visualization, in the form of population-time plots;
+and 2) data analysis.
 
-\section{Population-time plots}\label{population-time-plots}
+We explicitely aimed at being compatible with both \code{data.frame}s
+and \code{data.table}s. This is evident in some of the coding choices we
+made, and it is also reflected in our testing units.
+
+\subsection{Population-time plots}\label{population-time-plots}
+
+\subsection{Data analysis}\label{data-analysis}
+
+The data analysis step was separated into three parts: 1) case-base
+sampling; 2) estimation of the smooth hazard function; and 3)
+calculation of the risk function. By separating the sampling and
+estimation functions, we allowed the possibility of users implementing
+more complex sampling scheme, as described in Saarela
+\citeyearpar{saarela2016case}.
+
+The sampling scheme selected for \code{sampleCaseBase} was described in
+Hanley and Miettinen \citeyearpar{hanley2009fitting}: we first sample
+along the ``person'' axis, proportional to each individual's total
+follow-up time, and then we sample a moment uniformly over their
+follow-up time. This sampling scheme is equivalent to the following
+picture: imagine representing the total follow-up time of all
+individuals in the study along a single dimension, where the follow-up
+time of the next individual would start exactly when the follow-up time
+of the previous individual ends. Then the base series could be sampled
+uniformly from this one-dimensional representation of the overall
+follow-up time. In any case, the output is a dataset of the same class
+as the input, where each row corresponds to a person-moment. The
+covariate profile for each such person-moment is retained, and an offset
+term is added to the dataset. This output could then be used to fit a
+smooth hazard function, or for visualization of the base series.
+
+The fitting function \code{fitSmoothHazard} starts by looking at the
+class of the dataset: if it was generated from \code{sampleCaseBase}, it
+automatically inherited the class \code{cbData}. If the dataset supplied
+to \code{fitSmoothHazard} does not inherit from \code{cbData}, then the
+fitting function starts by calling \code{sampleCaseBase} to generate the
+base series. In other words, the occasional user can bypass
+\code{sampleCaseBase} altogether and only worry about the fitting
+function \code{fitSmoothHazard}.
+
+The fitting function retains the familiar formula interface of
+\code{glm}. The left-hand side of the formula should be the name of the
+column corresponding to the event type. The right-hand side can be any
+combination of the covariates, along with an explicit functional form
+for the time variable. Note that non-proportional hazard models can be
+achieved at this stage by adding an interaction term involving time. The
+offset term does not need to be specified by the user, as it is
+automatically added to the formula.
+
+Finally, the hazard function is fitted to the data using the function
+\code{glm}, unless there is more than one type event (i.e.~in a
+competing-risk analysis), in which case it uses the multinomical
+regression capabilities of the \pkg{VGAM} package. This package was
+selected for its ability to fit multinomial regression models with an
+offset.
+
+Once a model-fit object has been returned by \code{fitSmoothHazard}, all
+the familiar summary and diagnostic functions are available:
+\code{print}, \code{summary}, \code{predict}, \code{plot}, etc. Our
+package provides one more functionality: it computes risk functions from
+the model fit. For the case of a single event, it uses the familiar
+identity \[S(t) = \exp\left(-\int_0^t h(u;X) du\right).\] The integral
+is computed using either the \code{stats::integrate} function or
+Monte-Carlo integration. For the case of a competing-event analysis, the
+event-specific risk is computed using a nested double integral; in this
+setting, Monte-Carlo integration is faster and more flexible than
+\code{stats::integrate}. This is due to the fact that the computation
+involves a double loop over the selected time points; Monte-Carlo
+integration performs this step by using the vectorized \code{rowSums}
+and \code{colSums} functions.
+
+To decide between a single-event and a competing-event analysis, we
+created \code{absoluteRisk} as an \code{S3} generic, with methods for
+both \code{glm} and \code{CompRisk} objects (the latter inherits from
+\code{vglm} as well). The method dispatch system of \proglang{R} then
+takes care of matching the correct input to the correct methodology,
+without the user's intervention.
 
@@
 \hline
-Sex & 0.71 & (0.41, 1.24)\\
+Sex & 0.81 & (0.47, 1.41)\\
 \hline
-Disease & 0.52 & (0.28, 0.93)\\
+Disease & 0.57 & (0.31, 1.03)\\
 \hline
-Phase (CR2 vs. CR1) & 1.17 & (0.47, 2.93)\\
+Phase (CR2 vs. CR1) & 1.21 & (0.48, 3.01)\\
 \hline
-Phase (CR3 vs. CR1) & 1.78 & (0.46, 6.91)\\
+Phase (CR3 vs. CR1) & 1.75 & (0.45, 6.8)\\
 \hline
-Phase (Relapse vs. CR1) & 4.45 & (2.07, 9.57)\\
+Phase (Relapse vs. CR1) & 4.23 & (1.97, 9.09)\\
 \hline
-Source & 1.64 & (0.54, 5)\\
+Source & 1.60 & (0.53, 4.88)\\
 \hline
-Age & 0.99 & (0.97, 1.02)\\
+Age & 1.00 & (0.97, 1.02)\\
 \hline
@@
 
-\section{Discussion}\label{discussion}
-
+\renewcommand\refname{Discussion}
 \bibliography{casebase_references.bib}

### 2017-09-25 paper/casebase_jss.Rmd
add theoretical details

@@
 # Theoretical details
+
+As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to discretize the study base into an infinite amount of *person moments*.
+These person moments are indexed by both an individual in the study and a time point, and therefore each person moment has a covariate profile, an exposure status and an outcome status attached to it.
+We note that there is only a finite number of person moments associated with the event of interest (what Hanley & Miettinen call the *case series*).
+The case-base sampling refers to the sampling from the base of a representative finite sample called the *base series*.
+
+Popular survival analysis methods, like Kaplan-Meier and Cox regression, rely on the notion of risk-set sampling. Case-base sampling can be seen as an alternative.
+
+As shown by Saarela & Arjas [-@saarela2015non] (and further expanded in Saarela [-@saarela2016case]), writing the likelihood arising from this data-generating mechanism using the framework of non-homogenous Poisson processes, we eventually reach an expression where each person-moment's contribution is of the form
+$$\frac{\lambda(t)^{dN(t)}}{\rho(t) + \lambda(t)},$$
+where $N(t)$ is the counting process associated with the event of interest, $\lambda(t)$ is the corresponding hazard function, and $\rho(t)$ is the hazard function for the Poisson process associated with case-base sampling.
+This parametric form suggests that we can readily estimate log-hazards of the form $\log(\lambda(t)) = g(t; X)$ using logistic regression, where each observation corresponds to a person moment, the function $g(t; X)$ is linear in a finite number of parameters, and where we treat $\log(\rho(t))$ as an offset.
+
+In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base samping *uniformly*, i.e.
+to sample the base series uniformly from the study base.
+In terms of Poisson processes, this sampling strategy corresponds to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base.
+More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism  are proportional to the cardiovascular disease event rate given by the Framingham score.
 

### 2017-09-25 paper/casebase_jss.tex
add theoretical details

@@
 \section{Theoretical details}\label{theoretical-details}
+
+As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
+key idea behind case-base sampling is to discretize the study base into
+an infinite amount of \emph{person moments}. These person moments are
+indexed by both an individual in the study and a time point, and
+therefore each person moment has a covariate profile, an exposure status
+and an outcome status attached to it. We note that there is only a
+finite number of person moments associated with the event of interest
+(what Hanley \& Miettinen call the \emph{case series}). The case-base
+sampling refers to the sampling from the base of a representative finite
+sample called the \emph{base series}.
+
+Popular survival analysis methods, like Kaplan-Meier and Cox regression,
+rely on the notion of risk-set sampling. Case-base sampling can be seen
+as an alternative.
+
+As shown by Saarela \& Arjas \citeyearpar{saarela2015non} (and further
+expanded in Saarela \citeyearpar{saarela2016case}), writing the
+likelihood arising from this data-generating mechanism using the
+framework of non-homogenous Poisson processes, we eventually reach an
+expression where each person-moment's contribution is of the form
+\[\frac{\lambda(t)^{dN(t)}}{\rho(t) + \lambda(t)},\] where \(N(t)\) is
+the counting process associated with the event of interest,
+\(\lambda(t)\) is the corresponding hazard function, and \(\rho(t)\) is
+the hazard function for the Poisson process associated with case-base
+sampling. This parametric form suggests that we can readily estimate
+log-hazards of the form \(\log(\lambda(t)) = g(t; X)\) using logistic
+regression, where each observation corresponds to a person moment, the
+function \(g(t; X)\) is linear in a finite number of parameters, and
+where we treat \(\log(\rho(t))\) as an offset.
+
+In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
+suggest performing case-base samping \emph{uniformly}, i.e.~to sample
+the base series uniformly from the study base. In terms of Poisson
+processes, this sampling strategy corresponds to a time-homogeneous
+Poisson process with intensity equal to \(b/B\), where \(b\) is the
+number of sampled observations in the base series, and \(B\) is the
+total population-time for the study base. More complex examples are also
+available; see for example Saarela \& Arjas
+\citeyearpar{saarela2015non}, where the probabilities of the sampling
+mechanism are proportional to the cardiovascular disease event rate
+given by the Framingham score.
 
@@
 \hline
-Sex & 0.81 & (0.47, 1.41)\\
+Sex & 0.82 & (0.46, 1.48)\\
 \hline
-Disease & 0.57 & (0.31, 1.03)\\
+Disease & 0.48 & (0.25, 0.92)\\
 \hline
-Phase (CR2 vs. CR1) & 1.21 & (0.48, 3.01)\\
+Phase (CR2 vs. CR1) & 1.43 & (0.56, 3.66)\\
 \hline
-Phase (CR3 vs. CR1) & 1.75 & (0.45, 6.8)\\
+Phase (CR3 vs. CR1) & 1.63 & (0.41, 6.45)\\
 \hline
-Phase (Relapse vs. CR1) & 4.23 & (1.97, 9.09)\\
+Phase (Relapse vs. CR1) & 4.73 & (2.13, 10.48)\\
 \hline
-Source & 1.60 & (0.53, 4.88)\\
+Source & 1.51 & (0.49, 4.69)\\
 \hline
-Age & 1.00 & (0.97, 1.02)\\
+Age & 0.99 & (0.97, 1.02)\\
 \hline

### 2017-10-08 paper/casebase_jss.Rmd
add intro vignette and fix comp errors

@@
 library(magrittr)
+library(survival)
+library(casebase)
+library(splines)
+options(digits = 2, scipen = 999)
 ```
@@
 
-# Case study 1: Veteran data (or ERSPC if we can)
+# Case study 1--European Randomized Study of Prostate Cancer Screening Data
 
-  - First example
-  - Show how we can test for non-proportional hazard?
+This vignette introduces the main functions in the `casebase` package.
+The methods implemented in this package are based on the method developped in [Fitting Smooth-in-Time Prognostic Risk Functions via Logistic Regression (Hanley and Miettinen, 2009)](https://github.com/sahirbhatnagar/casebase/blob/master/references/Hanley_Miettinen-2009-Inter_J_of_Biostats.pdf).
+A rigorous treatment of the theory is developed in [A case-base sampling method for estimating recurrent
+event intensities (Saarela, 2015)](https://github.com/sahirbhatnagar/casebase/blob/master/references/Saarela-2015-Lifetime_Data_Analysis.pdf) and [Non-parametric Bayesian Hazard Regression for Chronic Disease Risk Assessment](https://github.com/sahirbhatnagar/casebase/blob/master/references/Saarela_et_al-2015-Scandinavian_Journal_of_Statistics.pdf).
+The motivation for this work is nicely summarised by Cox:
 
-# Case study 2: Bone-marrow transplant
+![](figures/reid_cox.png)
+
+The purpose of the `casebase` package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence (CI)) of an event, for a particular patient.
+The following points should be noted:
+
+1. Time matching/risk set sampling (including Cox partial likelihood) eliminates the baseline hazard from the likelihood expression for the
+hazard ratios
+2. If, however, the absolute risks are of interest, they have to be recovered using the semi-parametric Breslow estimator
+3.
+Alternative approaches for fitting flexible hazard models for estimating absolute risks, not requiring this two-step approach?
+Yes!
+[Hanley and Miettinen, 2009](https://github.com/sahirbhatnagar/casebase/blob/master/references/Hanley_Miettinen-2009-Inter_J_of_Biostats.pdf)
+
+> [Hanley and Miettinen, 2009](https://github.com/sahirbhatnagar/casebase/blob/master/references/Hanley_Miettinen-2009-Inter_J_of_Biostats.pdf) propose a fully parametric hazard model that can be fit via logistic regression.
+From the fitted hazard function, cumulative incidence and, thus, risk functions of time, treatment and profile can be easily derived.
+
+The `casebase` package fits the family of hazard functions of the form
+
+$$ h(x,t) = exp[g(x,t)] $$
+where $t$ denotes the numerical value (number of units) of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any).
+Different functions of $t$ lead to different parametric hazard models.
+
+The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$.
+
+$$ h(x,t) = exp(\beta_0 + \beta_1 x) $$
+
+The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been on study a.k.a the memoryless property (Kalbfleisch and Prentice, 2002).
+
+The Gompertz hazard model is given by including a linear term for time:
+
+$$ h(x,t)  = exp(\beta_0 + \beta_1 t + \beta_2 x) $$
+
+Use of $log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
+
+$$ h(x, t)  = exp(\beta_0 + \beta_1 \log(t) + \beta_2 x) $$
+
+Recall that the relative risk model (Cox, 1972)
+
+$$ \lambda(t;x) = \lambda_0(t) exp(\mathbf{X}\boldsymbol{\beta})  $$
+where $\lambda_0(\cdot)$ is an arbitrary unspecified baseline hazard function for continous time.
+
+In the following table we provide a comparison between the Cox model and case-base sampling:
+
+```{r, echo=FALSE, eval = FALSE, results='asis'}
+knitr::kable(data.frame(`feature` = c("model type", "time", "cumulative incidence", "non-proportional hazards","model testing","competing risks", "prediction"),
+           `Cox` = c("semi-parametric","left hand side of the equation", "step function", "interaction of covariates with time"," ", "difficult", "Kaplan-Meier-based"),
+           `Case Base Sampling` = c("fully parametric (logistic/multinomial regression)", "right hand side - allows flexible modeling of time",
+           "smooth-in-time curve", "interaction of covariates with time", "make use of GLM framework (LRT, AIC, BIC)",
+           "cause-specific cumulative incidence functions (CIFs) directly obtained via multinomial regression", "ROC, AUC, risk reclassification probabilities")), escape = FALSE)
+```
+
+[Slides from Olli Saarela](https://www.fields.utoronto.ca/programs/scientific/14-15/biomarker/slides/saarela.pdf)
+
+We fist load the required packages:
+
+Throughout this vignette, we make use of the European Randomized Study of Prostate Cancer Screening data which ships with the `casebase` package:
+
+```{r}
+data(ERSPC)
+head(ERSPC)
+ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
+                       levels = c(0,1), 
+                       labels = c("Control group", "Screening group"))
+```
+
+The results of this study were published by [Schroder FH, et al.
+N Engl J Med 2009](https://github.com/sahirbhatnagar/casebase/blob/master/references/Schroder_et_al-2009-NEJM.pdf).
+There's a really interesting story on how this data was obtained.
+See `help(ERSPC)` and [Liu Z, Rich B, Hanley JA, Recovering the raw data behind a non-parametric survival curve.
+Systematic Reviews 2014](https://github.com/sahirbhatnagar/casebase/blob/master/references/Liu_et_al-2015-Systematic_Reviews.pdf) for further details.
+
+Population time plots can be extremely informative graphical displays of survival data.
+They should be the first step in your exploratory data analyses.
+We facilitate this task in the `casebase` package using the `popTime` function.
+We first create the necessary dataset for producing the population time plots:
+
+```{r}
+pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
+```
+
+We can see its contents and its class:
+
+```{r}
+head(pt_object)
+
+class(pt_object)
+```
+
+The `casebase` package has a `plot` method for objects of class `popTime`:
+
+```{r}
+plot(pt_object)
+```
+
+We can also create exposure stratified plots by specifying the `exposure` argument in the `popTime` function:
+
+```{r}
+pt_object_strat <- casebase::popTime(ERSPC, 
+                                     event = "DeadOfPrCa", 
+                                     exposure = "ScrArm")
+```
+
+We can see its contents and its class:
+
+```{r}
+head(pt_object_strat)
+
+class(pt_object_strat)
+```
+
+The `casebase` package also has a `plot` method for objects of class `popTimeExposure`:
+
+```{r}
+plot(pt_object_strat)
+```
+
+We can also plot them side-by-side using the `ncol` argument:
+
+```{r}
+plot(pt_object_strat, ncol = 2)
+```
+
+We first fit a Cox model, examine the hazard ratio for the screening group (relative to the control group), and plot the cumulative incidence function (CIF).
+
+```{r}
+cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
+                             data = ERSPC)
+(sum_cox_model <- summary(cox_model))
+```
+
+We can plot the CIF for each group:
+
+```{r}
+new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
+                       ignore = 99)
+
+plot(survfit(cox_model, newdata = new_data),
+     xlab = "Years since Randomization", 
+     ylab = "Cumulative Incidence", 
+     fun = "event",
+     xlim = c(0,15), conf.int = F, col = c("red","blue"), 
+     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+                    exp(coef(cox_model)), 
+                    exp(confint(cox_model))[1], 
+                    exp(confint(cox_model))[2]))
+legend("topleft", 
+       legend = c("Control group", "Screening group"), 
+       col = c("red","blue"),
+       lty = c(1, 1), 
+       bg = "gray90")
+```
+
+We compare it to the figure in [Schroder FH, et al.
+N Engl J Med 2009](https://github.com/sahirbhatnagar/casebase/blob/master/references/Schroder_et_al-2009-NEJM.pdf) and see that the plots are very similar, as is the hazard ratio and 95% confidence interval:
+
+![](figures/schroderNEJM_figure2.png)
+
+Next we fit several models using case-base sampling. The models we fit differ in how we choose to model time.
+
+The `fitSmoothHazard` function provides an estimate of the hazard function $h(x, t)$ is the hazard function, $t$) denotes the numerical value (number of units) of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any).
+
+```{r}
+# set the seed for reproducible output
+set.seed(1234)
+
+casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
+                                                  data = ERSPC, 
+                                                  ratio = 100)
+
+summary(casebase_exponential)
+exp(coef(casebase_exponential)[2])
+exp(confint(casebase_exponential)[2,])
+```
+
+The `absoluteRisk` function provides an estimate of the cumulative incidence curves for a specific risk profile using the following equation:
+
+$$ CI(x, t) = 1 - exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$
+
+In the plot below, we overlay the estimated CIF from the casebase exponential model on the Cox model CIF:
+
+```{r}
+smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
+                                          time = seq(0,15,0.1), 
+                                          newdata = new_data)
+
+plot(survfit(cox_model, newdata = new_data),
+     xlab = "Years since Randomization", 
+     ylab = "Cumulative Incidence", 
+     fun = "event",
+     xlim = c(0,15), conf.int = F, col = c("red","blue"), 
+     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+                    exp(coef(cox_model)), 
+                    exp(confint(cox_model))[1], 
+                    exp(confint(cox_model))[2]))
+lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
+lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
+
+legend("topleft", 
+       legend = c("Control group (Cox)","Control group (Casebase)",
+                  "Screening group (Cox)", "Screening group (Casebase)"), 
+       col = c("red","red", "blue","blue"),
+       lty = c(1, 2, 1, 2), 
+       bg = "gray90")
+
+```
+
+As we can see, the exponential model is not a good fit.
+Based on what we observed in the population time plot, where more events are observed later on in time, this poor fit is expected.
+A constant hazard model would overestimate the cumulative incidence earlier on in time, and underestimate it later on, which is what we see in the cumulative incidence plot.
+This example demonstrates the benefits of population time plots as an exploratory analysis tool.
+
+Next we enter time linearly into the model:
+
+```{r}
+casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
+                                 data = ERSPC, 
+                                 ratio = 100)
+
+summary(casebase_time)
+exp(coef(casebase_time))
+exp(confint(casebase_time))
+```
+
+```{r}
+smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
+                                          time = seq(0,15,0.1), 
+                                          newdata = new_data)
+
+plot(survfit(cox_model, newdata = new_data),
+     xlab = "Years since Randomization", 
+     ylab = "Cumulative Incidence", 
+     fun = "event",
+     xlim = c(0,15), conf.int = F, col = c("red","blue"), 
+     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+                    exp(coef(cox_model)), 
+                    exp(confint(cox_model))[1], 
+                    exp(confint(cox_model))[2]))
+lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
+lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
+
+legend("topleft", 
+       legend = c("Control group (Cox)","Control group (Casebase)",
+                  "Screening group (Cox)", "Screening group (Casebase)"), 
+       col = c("red","red", "blue","blue"),
+       lty = c(1, 2, 1, 2), 
+       bg = "gray90")
+
+```
+
+We see that the Weibull model leads to a better fit.
+
+Next we try to enter a smooth function of time into the model using the `splines` package
+
+```{r}
+casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
+                                    data = ERSPC, 
+                                    ratio = 100)
+
+summary(casebase_splines)
+exp(coef(casebase_splines))
+exp(confint(casebase_splines))
+```
+
+```{r}
+smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
+                                    time = seq(0,15,0.1), 
+                                    newdata = new_data)
+
+plot(survfit(cox_model, newdata = new_data),
+     xlab = "Years since Randomization", 
+     ylab = "Cumulative Incidence", 
+     fun = "event",
+     xlim = c(0,15), conf.int = F, col = c("red","blue"), 
+     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+                    exp(coef(cox_model)), 
+                    exp(confint(cox_model))[1], 
+                    exp(confint(cox_model))[2]))
+lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
+lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
+
+legend("topleft", 
+       legend = c("Control group (Cox)","Control group (Casebase)",
+                  "Screening group (Cox)", "Screening group (Casebase)"), 
+       col = c("red","red", "blue","blue"),
+       lty = c(1, 2, 1, 2), 
+       bg = "gray90")
+
+```
+
+It looks like the best fit.
+
+Since we are in the GLM framework, we can easily test for which model better fits the data using a Likelihood ratio test (LRT).
+The null hypothesis here is that the linear model is just as good as the larger (in terms of number of parameters) splines model.
+
+```{r}
+anova(casebase_time, casebase_splines, test = "LRT")
+```
+
+We see that splines model is the better fit.
+
+# Case study 2--Bone-marrow transplant
 
@@
 
-# Case study 3: Vaccination study (recurrent events)
+# Case study 3
 
-  - Give a more complex example of sampling; time-dependent exposure
-    + Sampling needs to be done manually, but fitting function can still be used
+  - Glmnet, gam, and gbm
 
 # Discussion
+
+ - Discuss

### 2017-10-08 paper/casebase_jss.tex
add intro vignette and fix comp errors

@@
 
-\section{Case study 1: Veteran data (or ERSPC if we
-can)}\label{case-study-1-veteran-data-or-erspc-if-we-can}
+\section{Case study 1--European Randomized Study of Prostate Cancer
+Screening
+Data}\label{case-study-1european-randomized-study-of-prostate-cancer-screening-data}
 
-\begin{itemize}
+This vignette introduces the main functions in the \texttt{casebase}
+package. The methods implemented in this package are based on the method
+developped in
+\href{https://github.com/sahirbhatnagar/casebase/blob/master/references/Hanley_Miettinen-2009-Inter_J_of_Biostats.pdf}{Fitting
+Smooth-in-Time Prognostic Risk Functions via Logistic Regression (Hanley
+and Miettinen, 2009)}. A rigorous treatment of the theory is developed
+in
+\href{https://github.com/sahirbhatnagar/casebase/blob/master/references/Saarela-2015-Lifetime_Data_Analysis.pdf}{A
+case-base sampling method for estimating recurrent event intensities
+(Saarela, 2015)} and
+\href{https://github.com/sahirbhatnagar/casebase/blob/master/references/Saarela_et_al-2015-Scandinavian_Journal_of_Statistics.pdf}{Non-parametric
+Bayesian Hazard Regression for Chronic Disease Risk Assessment}. The
+motivation for this work is nicely summarised by Cox:
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{figures/reid_cox.png}
+\caption{}
+\end{figure}
+
+The purpose of the \texttt{casebase} package is to provide practitioners
+with an easy-to-use software tool to predict the risk (or cumulative
+incidence (CI)) of an event, for a particular patient. The following
+points should be noted:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
 \tightlist
 \item
-  First example
+  Time matching/risk set sampling (including Cox partial likelihood)
+  eliminates the baseline hazard from the likelihood expression for the
+  hazard ratios
 \item
-  Show how we can test for non-proportional hazard?
-\end{itemize}
+  If, however, the absolute risks are of interest, they have to be
+  recovered using the semi-parametric Breslow estimator
+\item
+  Alternative approaches for fitting flexible hazard models for
+  estimating absolute risks, not requiring this two-step approach? Yes!
+  \href{https://github.com/sahirbhatnagar/casebase/blob/master/references/Hanley_Miettinen-2009-Inter_J_of_Biostats.pdf}{Hanley
+  and Miettinen, 2009}
+\end{enumerate}
 
-\section{Case study 2: Bone-marrow
-transplant}\label{case-study-2-bone-marrow-transplant}
+\begin{quote}
+\href{https://github.com/sahirbhatnagar/casebase/blob/master/references/Hanley_Miettinen-2009-Inter_J_of_Biostats.pdf}{Hanley
+and Miettinen, 2009} propose a fully parametric hazard model that can be
+fit via logistic regression. From the fitted hazard function, cumulative
+incidence and, thus, risk functions of time, treatment and profile can
+be easily derived.
+\end{quote}
+
+The \texttt{casebase} package fits the family of hazard functions of the
+form
+
+\[ h(x,t) = exp[g(x,t)] \] where \(t\) denotes the numerical value
+(number of units) of a point in prognostic/prospective time and \(x\) is
+the realization of the vector \(X\) of variates based on the patient's
+profile and intervention (if any). Different functions of \(t\) lead to
+different parametric hazard models.
+
+The simplest of these models is the one-parameter exponential
+distribution which is obtained by taking the hazard function to be
+constant over the range of \(t\).
+
+\[ h(x,t) = exp(\beta_0 + \beta_1 x) \]
+
+The instantaneous failure rate is independent of \(t\), so that the
+conditional chance of failure in a time interval of specified length is
+the same regardless of how long the individual has been on study a.k.a
+the memoryless property (Kalbfleisch and Prentice, 2002).
+
+The Gompertz hazard model is given by including a linear term for time:
+
+\[ h(x,t)  = exp(\beta_0 + \beta_1 t + \beta_2 x) \]
+
+Use of \(log(t)\) yields the Weibull hazard which allows for a power
+dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
+
+\[ h(x, t)  = exp(\beta_0 + \beta_1 \log(t) + \beta_2 x) \]
+
+Recall that the relative risk model (Cox, 1972)
+
+\[ \lambda(t;x) = \lambda_0(t) exp(\mathbf{X}\boldsymbol{\beta})  \]
+where \(\lambda_0(\cdot)\) is an arbitrary unspecified baseline hazard
+function for continous time.
+
+In the following table we provide a comparison between the Cox model and
+case-base sampling:
+
+\href{https://www.fields.utoronto.ca/programs/scientific/14-15/biomarker/slides/saarela.pdf}{Slides
+from Olli Saarela}
+
+We fist load the required packages:
+
+Throughout this vignette, we make use of the European Randomized Study
+of Prostate Cancer Screening data which ships with the \texttt{casebase}
+package:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> data(ERSPC)
+R> head(ERSPC)
+\end{CodeInput}
+
+\begin{CodeOutput}
+  ScrArm Follow.Up.Time DeadOfPrCa
+1      1         0.0027          0
+2      1         0.0027          0
+3      1         0.0027          0
+4      0         0.0027          0
+5      0         0.0027          0
+6      0         0.0027          0
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
+R+                        levels = c(0,1), 
+R+                        labels = c("Control group", "Screening group"))
+\end{CodeInput}
+\end{CodeChunk}
+
+The results of this study were published by
+\href{https://github.com/sahirbhatnagar/casebase/blob/master/references/Schroder_et_al-2009-NEJM.pdf}{Schroder
+FH, et al. N Engl J Med 2009}. There's a really interesting story on how
+this data was obtained. See \texttt{help(ERSPC)} and
+\href{https://github.com/sahirbhatnagar/casebase/blob/master/references/Liu_et_al-2015-Systematic_Reviews.pdf}{Liu
+Z, Rich B, Hanley JA, Recovering the raw data behind a non-parametric
+survival curve. Systematic Reviews 2014} for further details.
+
+Population time plots can be extremely informative graphical displays of
+survival data. They should be the first step in your exploratory data
+analyses. We facilitate this task in the \texttt{casebase} package using
+the \texttt{popTime} function. We first create the necessary dataset for
+producing the population time plots:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
+\end{CodeInput}
+
+\begin{CodeOutput}
+'Follow.Up.Time' will be used as the time variable
+\end{CodeOutput}
+\end{CodeChunk}
+
+We can see its contents and its class:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> head(pt_object)
+\end{CodeInput}
+
+\begin{CodeOutput}
+            ScrArm   time event original.time original.event event status
+1: Screening group 0.0027     0        0.0027              0     censored
+2: Screening group 0.0027     0        0.0027              0     censored
+3: Screening group 0.0027     0        0.0027              0     censored
+4:   Control group 0.0027     0        0.0027              0     censored
+5:   Control group 0.0027     0        0.0027              0     censored
+6:   Control group 0.0027     0        0.0027              0     censored
+   ycoord yc n_available
+1: 159893  0           0
+2: 159892  0           0
+3: 159891  0           0
+4: 159890  0           0
+5: 159889  0           0
+6: 159888  0           0
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> class(pt_object)
+\end{CodeInput}
+
+\begin{CodeOutput}
+[1] "popTime"    "data.table" "data.frame"
+\end{CodeOutput}
+\end{CodeChunk}
+
+The \texttt{casebase} package has a \texttt{plot} method for objects of
+class \texttt{popTime}:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> plot(pt_object)
+\end{CodeInput}
+
+\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-5-1} \end{center}
+
+\end{CodeChunk}
+
+We can also create exposure stratified plots by specifying the
+\texttt{exposure} argument in the \texttt{popTime} function:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> pt_object_strat <- casebase::popTime(ERSPC, 
+R+                                      event = "DeadOfPrCa", 
+R+                                      exposure = "ScrArm")
+\end{CodeInput}
+
+\begin{CodeOutput}
+'Follow.Up.Time' will be used as the time variable
+\end{CodeOutput}
+\end{CodeChunk}
+
+We can see its contents and its class:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> head(pt_object_strat)
+\end{CodeInput}
+
+\begin{CodeOutput}
+$data
+                 ScrArm    time event original.time original.event
+     1:   Control group  0.0027     0        0.0027              0
+     2:   Control group  0.0027     0        0.0027              0
+     3:   Control group  0.0027     0        0.0027              0
+     4:   Control group  0.0027     0        0.0027              0
+     5:   Control group  0.0137     0        0.0137              0
+    ---                                                           
+159889: Screening group 14.9405     0       14.9405              0
+159890: Screening group 14.9405     0       14.9405              0
+159891: Screening group 14.9405     0       14.9405              0
+159892: Screening group 14.9405     0       14.9405              0
+159893: Screening group 14.9405     0       14.9405              0
+        event status ycoord yc n_available
+     1:     censored  88232  0           0
+     2:     censored  88231  0           0
+     3:     censored  88230  0           0
+     4:     censored  88229  0           0
+     5:     censored  88228  0           0
+    ---                                   
+159889:     censored      5  0           0
+159890:     censored      4  0           0
+159891:     censored      3  0           0
+159892:     censored      2  0           0
+159893:     censored      1  0           0
+
+$exposure
+[1] "ScrArm"
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> class(pt_object_strat)
+\end{CodeInput}
+
+\begin{CodeOutput}
+[1] "popTimeExposure" "list"           
+\end{CodeOutput}
+\end{CodeChunk}
+
+The \texttt{casebase} package also has a \texttt{plot} method for
+objects of class \texttt{popTimeExposure}:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> plot(pt_object_strat)
+\end{CodeInput}
+
+\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-8-1} \end{center}
+
+\end{CodeChunk}
+
+We can also plot them side-by-side using the \texttt{ncol} argument:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> plot(pt_object_strat, ncol = 2)
+\end{CodeInput}
+
+\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-9-1} \end{center}
+
+\end{CodeChunk}
+
+We first fit a Cox model, examine the hazard ratio for the screening
+group (relative to the control group), and plot the cumulative incidence
+function (CIF).
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
+R+                              data = ERSPC)
+R> (sum_cox_model <- summary(cox_model))
+\end{CodeInput}
+
+\begin{CodeOutput}
+Call:
+survival::coxph(formula = Surv(Follow.Up.Time, DeadOfPrCa) ~ 
+    ScrArm, data = ERSPC)
+
+  n= 159893, number of events= 540
+
+                        coef exp(coef) se(coef)     z Pr(>|z|)  
+ScrArmScreening group -0.222     0.801    0.088 -2.52    0.012 *
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+
+                      exp(coef) exp(-coef) lower .95 upper .95
+ScrArmScreening group     0.801       1.25     0.674     0.952
+
+Concordance= 0.519  (se = 0.011 )
+Rsquare= 0   (max possible= 0.075 )
+Likelihood ratio test= 6.45  on 1 df,   p=0.0111
+Wald test            = 6.37  on 1 df,   p=0.0116
+Score (logrank) test = 6.39  on 1 df,   p=0.0115
+\end{CodeOutput}
+\end{CodeChunk}
+
+We can plot the CIF for each group:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
+R+                        ignore = 99)
+R> 
+R> plot(survfit(cox_model, newdata = new_data),
+R+      xlab = "Years since Randomization", 
+R+      ylab = "Cumulative Incidence", 
+R+      fun = "event",
+R+      xlim = c(0,15), conf.int = F, col = c("red","blue"), 
+R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+                     exp(coef(cox_model)), 
+R+                     exp(confint(cox_model))[1], 
+R+                     exp(confint(cox_model))[2]))
+R> legend("topleft", 
+R+        legend = c("Control group", "Screening group"), 
+R+        col = c("red","blue"),
+R+        lty = c(1, 1), 
+R+        bg = "gray90")
+\end{CodeInput}
+
+\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-11-1} \end{center}
+
+\end{CodeChunk}
+
+We compare it to the figure in
+\href{https://github.com/sahirbhatnagar/casebase/blob/master/references/Schroder_et_al-2009-NEJM.pdf}{Schroder
+FH, et al. N Engl J Med 2009} and see that the plots are very similar,
+as is the hazard ratio and 95\% confidence interval:
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{figures/schroderNEJM_figure2.png}
+\caption{}
+\end{figure}
+
+Next we fit several models using case-base sampling. The models we fit
+differ in how we choose to model time.
+
+The \texttt{fitSmoothHazard} function provides an estimate of the hazard
+function \(h(x, t)\) is the hazard function, \(t\)) denotes the
+numerical value (number of units) of a point in prognostic/prospective
+time and \(x\) is the realization of the vector \(X\) of variates based
+on the patient's profile and intervention (if any).
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # set the seed for reproducible output
+R> set.seed(1234)
+R> 
+R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
+R+                                                   data = ERSPC, 
+R+                                                   ratio = 100)
+\end{CodeInput}
+
+\begin{CodeOutput}
+'Follow.Up.Time' will be used as the time variable
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> summary(casebase_exponential)
+\end{CodeInput}
+
+\begin{CodeOutput}
+
+Call:
+glm(formula = formula, family = binomial, data = sampleData)
+
+Deviance Residuals: 
+   Min      1Q  Median      3Q     Max  
+-0.458  -0.458  -0.409  -0.409   2.246
+
+Coefficients:
+                      Estimate Std. Error z value            Pr(>|z|)    
+(Intercept)            -7.7553     0.0584 -132.87 <0.0000000000000002 ***
+ScrArmScreening group  -0.2382     0.0921   -2.59              0.0097 ** 
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+
+(Dispersion parameter for binomial family taken to be 1)
+
+    Null deviance: 3619.1  on 5939  degrees of freedom
+Residual deviance: 3612.3  on 5938  degrees of freedom
+AIC: 3616
+
+Number of Fisher Scoring iterations: 5
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> exp(coef(casebase_exponential)[2])
+\end{CodeInput}
+
+\begin{CodeOutput}
+ScrArmScreening group 
+                 0.79 
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> exp(confint(casebase_exponential)[2,])
+\end{CodeInput}
+
+\begin{CodeOutput}
+Waiting for profiling to be done...
+\end{CodeOutput}
+
+\begin{CodeOutput}
+ 2.5 % 97.5 % 
+  0.66   0.94 
+\end{CodeOutput}
+\end{CodeChunk}
+
+The \texttt{absoluteRisk} function provides an estimate of the
+cumulative incidence curves for a specific risk profile using the
+following equation:
+
+\[ CI(x, t) = 1 - exp\left( - \int_0^t h(x, u) \textrm{d}u \right) \]
+
+In the plot below, we overlay the estimated CIF from the casebase
+exponential model on the Cox model CIF:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
+R+                                           time = seq(0,15,0.1), 
+R+                                           newdata = new_data)
+R> 
+R> plot(survfit(cox_model, newdata = new_data),
+R+      xlab = "Years since Randomization", 
+R+      ylab = "Cumulative Incidence", 
+R+      fun = "event",
+R+      xlim = c(0,15), conf.int = F, col = c("red","blue"), 
+R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+                     exp(coef(cox_model)), 
+R+                     exp(confint(cox_model))[1], 
+R+                     exp(confint(cox_model))[2]))
+R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
+R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
+R> 
+R> 
+R> legend("topleft", 
+R+        legend = c("Control group (Cox)","Control group (Casebase)",
+R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R+        col = c("red","red", "blue","blue"),
+R+        lty = c(1, 2, 1, 2), 
+R+        bg = "gray90")
+\end{CodeInput}
+
+\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-13-1} \end{center}
+
+\end{CodeChunk}
+
+As we can see, the exponential model is not a good fit. Based on what we
+observed in the population time plot, where more events are observed
+later on in time, this poor fit is expected. A constant hazard model
+would overestimate the cumulative incidence earlier on in time, and
+underestimate it later on, which is what we see in the cumulative
+incidence plot. This example demonstrates the benefits of population
+time plots as an exploratory analysis tool.
+
+Next we enter time linearly into the model:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
+R+                                  data = ERSPC, 
+R+                                  ratio = 100)
+\end{CodeInput}
+
+\begin{CodeOutput}
+'Follow.Up.Time' will be used as the time variable
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> summary(casebase_time)
+\end{CodeInput}
+
+\begin{CodeOutput}
+
+Call:
+glm(formula = formula, family = binomial, data = sampleData)
+
+Deviance Residuals: 
+   Min      1Q  Median      3Q     Max  
+-1.043  -0.487  -0.365  -0.275   2.744
+
+Coefficients:
+                      Estimate Std. Error z value            Pr(>|z|)    
+(Intercept)            -9.0421     0.1183  -76.45 <0.0000000000000002 ***
+Follow.Up.Time          0.2254     0.0158   14.24 <0.0000000000000002 ***
+ScrArmScreening group  -0.2611     0.0940   -2.78              0.0055 ** 
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+
+(Dispersion parameter for binomial family taken to be 1)
+
+    Null deviance: 3619.1  on 5939  degrees of freedom
+Residual deviance: 3394.9  on 5937  degrees of freedom
+AIC: 3401
+
+Number of Fisher Scoring iterations: 5
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> exp(coef(casebase_time))
+\end{CodeInput}
+
+\begin{CodeOutput}
+          (Intercept)        Follow.Up.Time ScrArmScreening group 
+              0.00012               1.25285               0.77023 
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> exp(confint(casebase_time))
+\end{CodeInput}
+
+\begin{CodeOutput}
+Waiting for profiling to be done...
+\end{CodeOutput}
+
+\begin{CodeOutput}
+                         2.5 %  97.5 %
+(Intercept)           0.000093 0.00015
+Follow.Up.Time        1.214840 1.29263
+ScrArmScreening group 0.639934 0.92534
+\end{CodeOutput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
+R+                                           time = seq(0,15,0.1), 
+R+                                           newdata = new_data)
+R> 
+R> plot(survfit(cox_model, newdata = new_data),
+R+      xlab = "Years since Randomization", 
+R+      ylab = "Cumulative Incidence", 
+R+      fun = "event",
+R+      xlim = c(0,15), conf.int = F, col = c("red","blue"), 
+R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+                     exp(coef(cox_model)), 
+R+                     exp(confint(cox_model))[1], 
+R+                     exp(confint(cox_model))[2]))
+R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
+R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
+R> 
+R> legend("topleft", 
+R+        legend = c("Control group (Cox)","Control group (Casebase)",
+R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R+        col = c("red","red", "blue","blue"),
+R+        lty = c(1, 2, 1, 2), 
+R+        bg = "gray90")
+\end{CodeInput}
+
+\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-15-1} \end{center}
+
+\end{CodeChunk}
+
+We see that the Weibull model leads to a better fit.
+
+Next we try to enter a smooth function of time into the model using the
+\texttt{splines} package
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
+R+                                     data = ERSPC, 
+R+                                     ratio = 100)
+\end{CodeInput}
+
+\begin{CodeOutput}
+'Follow.Up.Time' will be used as the time variable
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> summary(casebase_splines)
+\end{CodeInput}
+
+\begin{CodeOutput}
+
+Call:
+glm(formula = formula, family = binomial, data = sampleData)
+
+Deviance Residuals: 
+   Min      1Q  Median      3Q     Max  
+-1.197  -0.522  -0.398  -0.212   3.187
+
+Coefficients:
+                      Estimate Std. Error z value             Pr(>|z|)    
+(Intercept)           -10.5045     0.3342  -31.43 < 0.0000000000000002 ***
+bs(Follow.Up.Time)1     4.8348     0.8400    5.76     0.00000000862087 ***
+bs(Follow.Up.Time)2     1.8367     0.4859    3.78              0.00016 ***
+bs(Follow.Up.Time)3     5.0478     0.6998    7.21     0.00000000000055 ***
+ScrArmScreening group  -0.1503     0.0941   -1.60              0.11013    
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+
+(Dispersion parameter for binomial family taken to be 1)
+
+    Null deviance: 3619.1  on 5939  degrees of freedom
+Residual deviance: 3356.5  on 5935  degrees of freedom
+AIC: 3367
+
+Number of Fisher Scoring iterations: 6
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> exp(coef(casebase_splines))
+\end{CodeInput}
+
+\begin{CodeOutput}
+          (Intercept)   bs(Follow.Up.Time)1   bs(Follow.Up.Time)2 
+             0.000027            125.811140              6.275672 
+  bs(Follow.Up.Time)3 ScrArmScreening group 
+           155.673063              0.860457 
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> exp(confint(casebase_splines))
+\end{CodeInput}
+
+\begin{CodeOutput}
+Waiting for profiling to be done...
+\end{CodeOutput}
+
+\begin{CodeOutput}
+                          2.5 %     97.5 %
+(Intercept)            0.000014   0.000051
+bs(Follow.Up.Time)1   25.404219 687.891148
+bs(Follow.Up.Time)2    2.425559  16.399777
+bs(Follow.Up.Time)3   39.667300 626.079069
+ScrArmScreening group  0.714859   1.033843
+\end{CodeOutput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
+R+                                     time = seq(0,15,0.1), 
+R+                                     newdata = new_data)
+R> 
+R> plot(survfit(cox_model, newdata = new_data),
+R+      xlab = "Years since Randomization", 
+R+      ylab = "Cumulative Incidence", 
+R+      fun = "event",
+R+      xlim = c(0,15), conf.int = F, col = c("red","blue"), 
+R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+                     exp(coef(cox_model)), 
+R+                     exp(confint(cox_model))[1], 
+R+                     exp(confint(cox_model))[2]))
+R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
+R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
+R> 
+R> legend("topleft", 
+R+        legend = c("Control group (Cox)","Control group (Casebase)",
+R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R+        col = c("red","red", "blue","blue"),
+R+        lty = c(1, 2, 1, 2), 
+R+        bg = "gray90")
+\end{CodeInput}
+
+\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-17-1} \end{center}
+
+\end{CodeChunk}
+
+It looks like the best fit.
+
+Since we are in the GLM framework, we can easily test for which model
+better fits the data using a Likelihood ratio test (LRT). The null
+hypothesis here is that the linear model is just as good as the larger
+(in terms of number of parameters) splines model.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> anova(casebase_time, casebase_splines, test = "LRT")
+\end{CodeInput}
+
+\begin{CodeOutput}
+Analysis of Deviance Table
+
+Model 1: DeadOfPrCa ~ Follow.Up.Time + ScrArm + offset(offset)
+Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
+  Resid. Df Resid. Dev Df Deviance     Pr(>Chi)    
+1      5937       3395                             
+2      5935       3357  2     38.4 0.0000000046 ***
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{CodeOutput}
+\end{CodeChunk}
+
+We see that splines model is the better fit.
+
+\section{Case study 2--Bone-marrow
+transplant}\label{case-study-2bone-marrow-transplant}
 
@@
 \hline
-Sex & 0.82 & (0.46, 1.48)\\
+Sex & 0.78 & (0.43, 1.41)\\
 \hline
-Disease & 0.48 & (0.25, 0.92)\\
+Disease & 0.41 & (0.22, 0.77)\\
 \hline
-Phase (CR2 vs. CR1) & 1.43 & (0.56, 3.66)\\
+Phase (CR2 vs. CR1) & 1.27 & (0.5, 3.23)\\
 \hline
-Phase (CR3 vs. CR1) & 1.63 & (0.41, 6.45)\\
+Phase (CR3 vs. CR1) & 1.77 & (0.44, 7.09)\\
 \hline
-Phase (Relapse vs. CR1) & 4.73 & (2.13, 10.48)\\
+Phase (Relapse vs. CR1) & 5.17 & (2.34, 11.43)\\
 \hline
-Source & 1.51 & (0.49, 4.69)\\
+Source & 1.87 & (0.59, 5.88)\\
 \hline
-Age & 0.99 & (0.97, 1.02)\\
+Age & 1.00 & (0.97, 1.02)\\
 \hline
@@
 
-{\centering \includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-6-1}
+{\centering \includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-24-1}
 
@@
 \caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups.
-The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:unnamed-chunk-6}
+The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:unnamed-chunk-24}
 \end{figure}
@@
 
-\section{Case study 3: Vaccination study (recurrent
-events)}\label{case-study-3-vaccination-study-recurrent-events}
+\section{Case study 3}\label{case-study-3}
 
@@
 \item
-  Give a more complex example of sampling; time-dependent exposure
+  Glmnet, gam, and gbm
+\end{itemize}
+
+\section{Discussion}\label{discussion}
 
@@
 \item
-    Sampling needs to be done manually, but fitting function can still
-    be used
-  \end{itemize}
+  Discuss
 \end{itemize}
 
-\renewcommand\refname{Discussion}
 \bibliography{casebase_references.bib}

### 2017-10-11 paper/casebase_jss.Rmd
condense case study 1

--- 2017-09-25 paper/casebase_jss.Rmd
@@
 library(magrittr)
+library(survival)
+library(casebase)
+library(splines)
+options(digits = 2, scipen = 999)
+# set the seed for reproducible output
+set.seed(1234)
 ```
@@
 
-  - Motivation
-    + Flexible 
-    + Flexible
-    + Flexible
+The purpose of the `casebase` package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence (CI)) of an event, for a particular patient.
+The following points should be noted:
+
+1. Time matching/risk set sampling (including Cox partial likelihood) eliminates the baseline hazard from the likelihood expression for the
+hazard ratios
+2. If, however, the absolute risks are of interest, they have to be recovered using the semi-parametric Breslow estimator
+3. Alternative approaches for fitting flexible hazard models for estimating absolute risks, not requiring this two-step approach? Yes! [@hanley2009fitting]
+
+[@hanley2009fitting] propose a fully parametric hazard model that can be fit via logistic regression.
+From the fitted hazard function, cumulative incidence and, thus, risk functions of time, treatment and profile can be easily derived.
 
@@
 More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism  are proportional to the cardiovascular disease event rate given by the Framingham score.
+
+The `casebase` package fits the family of hazard functions of the form
+
+$$ h(x,t) = exp[g(x,t)] $$
+where $t$ denotes the numerical value (number of units) of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any).
+Different functions of $t$ lead to different parametric hazard models.
+
+The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$.
+
+$$ h(x,t) = exp(\beta_0 + \beta_1 x) $$
+
+The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been on study a.k.a the memoryless property (Kalbfleisch and Prentice, 2002).
+
+The Gompertz hazard model is given by including a linear term for time:
+
+$$ h(x,t)  = exp(\beta_0 + \beta_1 t + \beta_2 x) $$
+
+Use of $log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
+
+$$ h(x, t)  = exp(\beta_0 + \beta_1 \log(t) + \beta_2 x) $$
 
@@
 
-# Case study 1: Veteran data (or ERSPC if we can)
+# Case study 1--European Randomized Study of Prostate Cancer Screening
 
-  - First example
-  - Show how we can test for non-proportional hazard?
+To introduce the different features available, we make use of the European Randomized Study of Prostate Cancer Screening data; this dataset is available through the \pkg{casebase} package:
 
-# Case study 2: Bone-marrow transplant
+```{r}
+data(ERSPC)
+ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
+                       levels = c(0,1), 
+                       labels = c("Control group", "Screening group"))
+```
+
+The results of this study were published by [schroder2009screening]. This data was obtained using the approach described in [liu2014recovering].
+
+Population time plots can be extremely informative graphical displays of survival data.
+They should be the first step in an exploratory data analysis.
+We facilitate this task in the \pkg{casebase} package using the \code{popTime} function.
+We first create the necessary dataset for producing the population time plots, and we can produce the plot by using the corresponding \code{plot} method:
+
+```{r}
+pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
+plot(pt_object)
+```
+
+We can also create exposure stratified plots by specifying the \code{exposure} argument in the \code{popTime} function:
+
+```{r}
+pt_object_strat <- casebase::popTime(ERSPC, 
+                                     event = "DeadOfPrCa", 
+                                     exposure = "ScrArm")
+plot(pt_object_strat)
+```
+
+We can also plot them side-by-side using the \code{ncol} argument:
+
+```{r}
+plot(pt_object_strat, ncol = 2)
+```
+
+First, we fit a Cox model to the data, examine the hazard ratio for the screening group (relative to the control group), and plot the cumulative incidence function (CIF).
+
+```{r}
+cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
+                             data = ERSPC)
+summary(cox_model)
+```
+
+We can plot the CIF for each group:
+
+```{r}
+new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
+                       ignore = 99)
+
+plot(survfit(cox_model, newdata = new_data),
+     xlab = "Years since Randomization", 
+     ylab = "Cumulative Incidence", 
+     fun = "event",
+     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+                    exp(coef(cox_model)), 
+                    exp(confint(cox_model))[1], 
+                    exp(confint(cox_model))[2]))
+legend("topleft", 
+       legend = c("Control group", "Screening group"), 
+       col = c("red","blue"),
+       lty = c(1, 1), 
+       bg = "gray90")
+```
+
+Next we fit several models using case-base sampling. The models we fit differ in how we choose to model time.
+
+The \code{fitSmoothHazard} function provides an estimate of the hazard function $h(x, t)$ is the hazard function, where $t$ denotes the numerical value of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any).
+
+```{r}
+casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
+                                                  data = ERSPC, 
+                                                  ratio = 100)
+
+summary(casebase_exponential)
+exp(coef(casebase_exponential)[2])
+exp(confint(casebase_exponential)[2,])
+```
+
+The \code{absoluteRisk} function provides an estimate of the cumulative incidence curves for a specific risk profile using the following equation:
+
+$$ CI(x, t) = 1 - exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$
+
+In the plot below, we overlay the estimated CIF from the casebase exponential model on the Cox model CIF:
+
+```{r}
+smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
+                                          time = seq(0,15,0.1), 
+                                          newdata = new_data)
+
+plot(survfit(cox_model, newdata = new_data),
+     xlab = "Years since Randomization", 
+     ylab = "Cumulative Incidence", 
+     fun = "event",
+     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+                    exp(coef(cox_model)), 
+                    exp(confint(cox_model))[1], 
+                    exp(confint(cox_model))[2]))
+lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
+lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
+
+legend("topleft", 
+       legend = c("Control group (Cox)","Control group (Casebase)",
+                  "Screening group (Cox)", "Screening group (Casebase)"), 
+       col = c("red","red", "blue","blue"),
+       lty = c(1, 2, 1, 2), 
+       bg = "gray90")
+```
+
+As we can see, the exponential model is not a good fit.
+Based on what we observed in the population time plot, where more events are observed later on in time, this poor fit is expected.
+A constant hazard model would overestimate the cumulative incidence earlier on in time, and underestimate it later on; this is what we see on the cumulative incidence plot.
+This example demonstrates the benefits of population time plots as an exploratory analysis tool.
+
+Next we enter time linearly into the model:
+
+```{r}
+casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
+                                 data = ERSPC, 
+                                 ratio = 100)
+
+summary(casebase_time)
+exp(coef(casebase_time))
+exp(confint(casebase_time))
+```
+
+```{r}
+smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
+                                          time = seq(0,15,0.1), 
+                                          newdata = new_data)
+
+plot(survfit(cox_model, newdata = new_data),
+     xlab = "Years since Randomization", 
+     ylab = "Cumulative Incidence", 
+     fun = "event",
+     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+                    exp(coef(cox_model)), 
+                    exp(confint(cox_model))[1], 
+                    exp(confint(cox_model))[2]))
+lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
+lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
+
+legend("topleft", 
+       legend = c("Control group (Cox)","Control group (Casebase)",
+                  "Screening group (Cox)", "Screening group (Casebase)"), 
+       col = c("red","red", "blue","blue"),
+       lty = c(1, 2, 1, 2), 
+       bg = "gray90")
+```
+
+We see that the Weibull model leads to a better fit.
+
+Next we try to enter a smooth function of time into the model using the \code{splines} package:
+
+```{r}
+casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ splines::bs(Follow.Up.Time) + ScrArm, 
+                                    data = ERSPC, 
+                                    ratio = 100)
+
+summary(casebase_splines)
+exp(coef(casebase_splines))
+exp(confint(casebase_splines))
+```
+
+```{r}
+smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
+                                    time = seq(0,15,0.1), 
+                                    newdata = new_data)
+
+plot(survfit(cox_model, newdata = new_data),
+     xlab = "Years since Randomization", 
+     ylab = "Cumulative Incidence", 
+     fun = "event",
+     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+                    exp(coef(cox_model)), 
+                    exp(confint(cox_model))[1], 
+                    exp(confint(cox_model))[2]))
+lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
+lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
+
+legend("topleft", 
+       legend = c("Control group (Cox)","Control group (Casebase)",
+                  "Screening group (Cox)", "Screening group (Casebase)"), 
+       col = c("red","red", "blue","blue"),
+       lty = c(1, 2, 1, 2), 
+       bg = "gray90")
+```
+
+It looks like the best fit.
+
+Since we are within the GLM framework, we can easily test for which model better fits the data using a Likelihood Ratio Test (LRT).
+The null hypothesis here is that the linear model is just as good as the larger (in terms of number of parameters) splines model.
+
+```{r}
+anova(casebase_time, casebase_splines, test = "LRT")
+```
+
+As espected, we see that splines model provides a better fit.
+
+# Case study 2--Bone-marrow transplant
 
@@
 
-# Case study 3: Vaccination study (recurrent events)
+# Case study 3
 
-  - Give a more complex example of sampling; time-dependent exposure
-    + Sampling needs to be done manually, but fitting function can still be used
+  - Glmnet, gam, and gbm
 
 # Discussion
+
+In the following table we provide a comparison between the Cox model and case-base sampling:
+
+```{r, echo=FALSE, eval = FALSE, results='asis'}
+knitr::kable(data.frame(`feature` = c("model type", "time", "cumulative incidence", "non-proportional hazards","model testing","competing risks", "prediction"),
+           `Cox` = c("semi-parametric","left hand side of the equation", "step function", "interaction of covariates with time"," ", "difficult", "Kaplan-Meier-based"),
+           `Case Base Sampling` = c("fully parametric (logistic/multinomial regression)", "right hand side - allows flexible modeling of time",
+           "smooth-in-time curve", "interaction of covariates with time", "make use of GLM framework (LRT, AIC, BIC)",
+           "cause-specific cumulative incidence functions (CIFs) directly obtained via multinomial regression", "ROC, AUC, risk reclassification probabilities")), escape = FALSE)
+```

### 2017-10-11 paper/casebase_jss.tex
condense case study 1

--- 2017-09-25 paper/casebase_jss.tex
@@
 
-\begin{itemize}
+The purpose of the \texttt{casebase} package is to provide practitioners
+with an easy-to-use software tool to predict the risk (or cumulative
+incidence (CI)) of an event, for a particular patient. The following
+points should be noted:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
 \tightlist
 \item
-  Motivation
+  Time matching/risk set sampling (including Cox partial likelihood)
+  eliminates the baseline hazard from the likelihood expression for the
+  hazard ratios
+\item
+  If, however, the absolute risks are of interest, they have to be
+  recovered using the semi-parametric Breslow estimator
+\item
+  Alternative approaches for fitting flexible hazard models for
+  estimating absolute risks, not requiring this two-step approach? Yes!
+  \citep{hanley2009fitting}
+\end{enumerate}
 
-  \begin{itemize}
-  \tightlist
-  \item
-    Flexible
-  \item
-    Flexible
-  \item
-    Flexible
-  \end{itemize}
-\end{itemize}
+\citep{hanley2009fitting} propose a fully parametric hazard model that
+can be fit via logistic regression. From the fitted hazard function,
+cumulative incidence and, thus, risk functions of time, treatment and
+profile can be easily derived.
 
@@
 given by the Framingham score.
+
+The \texttt{casebase} package fits the family of hazard functions of the
+form
+
+\[ h(x,t) = exp[g(x,t)] \] where \(t\) denotes the numerical value
+(number of units) of a point in prognostic/prospective time and \(x\) is
+the realization of the vector \(X\) of variates based on the patient's
+profile and intervention (if any). Different functions of \(t\) lead to
+different parametric hazard models.
+
+The simplest of these models is the one-parameter exponential
+distribution which is obtained by taking the hazard function to be
+constant over the range of \(t\).
+
+\[ h(x,t) = exp(\beta_0 + \beta_1 x) \]
+
+The instantaneous failure rate is independent of \(t\), so that the
+conditional chance of failure in a time interval of specified length is
+the same regardless of how long the individual has been on study a.k.a
+the memoryless property (Kalbfleisch and Prentice, 2002).
+
+The Gompertz hazard model is given by including a linear term for time:
+
+\[ h(x,t)  = exp(\beta_0 + \beta_1 t + \beta_2 x) \]
+
+Use of \(log(t)\) yields the Weibull hazard which allows for a power
+dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
+
+\[ h(x, t)  = exp(\beta_0 + \beta_1 \log(t) + \beta_2 x) \]
 
@@
 
-\section{Case study 1: Veteran data (or ERSPC if we
-can)}\label{case-study-1-veteran-data-or-erspc-if-we-can}
+\section{Case study 1--European Randomized Study of Prostate Cancer
+Screening}\label{case-study-1european-randomized-study-of-prostate-cancer-screening}
 
-\begin{itemize}
-\tightlist
-\item
-  First example
-\item
-  Show how we can test for non-proportional hazard?
-\end{itemize}
+To introduce the different features available, we make use of the
+European Randomized Study of Prostate Cancer Screening data; this
+dataset is available through the \pkg{casebase} package:
 
-\section{Case study 2: Bone-marrow
-transplant}\label{case-study-2-bone-marrow-transplant}
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> data(ERSPC)
+R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
+R+                        levels = c(0,1), 
+R+                        labels = c("Control group", "Screening group"))
+\end{CodeInput}
+\end{CodeChunk}
+
+The results of this study were published by {[}schroder2009screening{]}.
+This data was obtained using the approach described in
+{[}liu2014recovering{]}.
+
+Population time plots can be extremely informative graphical displays of
+survival data. They should be the first step in an exploratory data
+analysis. We facilitate this task in the \pkg{casebase} package using
+the \code{popTime} function. We first create the necessary dataset for
+producing the population time plots, and we can produce the plot by
+using the corresponding \code{plot} method:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
+\end{CodeInput}
+
+\begin{CodeOutput}
+'Follow.Up.Time' will be used as the time variable
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> plot(pt_object)
+\end{CodeInput}
+
+\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-2-1} \end{center}
+
+\end{CodeChunk}
+
+We can also create exposure stratified plots by specifying the
+\code{exposure} argument in the \code{popTime} function:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> pt_object_strat <- casebase::popTime(ERSPC, 
+R+                                      event = "DeadOfPrCa", 
+R+                                      exposure = "ScrArm")
+\end{CodeInput}
+
+\begin{CodeOutput}
+'Follow.Up.Time' will be used as the time variable
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> plot(pt_object_strat)
+\end{CodeInput}
+
+\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-3-1} \end{center}
+
+\end{CodeChunk}
+
+We can also plot them side-by-side using the \code{ncol} argument:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> plot(pt_object_strat, ncol = 2)
+\end{CodeInput}
+
+\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-4-1} \end{center}
+
+\end{CodeChunk}
+
+First, we fit a Cox model to the data, examine the hazard ratio for the
+screening group (relative to the control group), and plot the cumulative
+incidence function (CIF).
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
+R+                              data = ERSPC)
+R> summary(cox_model)
+\end{CodeInput}
+
+\begin{CodeOutput}
+Call:
+survival::coxph(formula = Surv(Follow.Up.Time, DeadOfPrCa) ~ 
+    ScrArm, data = ERSPC)
+
+  n= 159893, number of events= 540
+
+                        coef exp(coef) se(coef)     z Pr(>|z|)  
+ScrArmScreening group -0.222     0.801    0.088 -2.52    0.012 *
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+
+                      exp(coef) exp(-coef) lower .95 upper .95
+ScrArmScreening group     0.801       1.25     0.674     0.952
+
+Concordance= 0.519  (se = 0.011 )
+Rsquare= 0   (max possible= 0.075 )
+Likelihood ratio test= 6.45  on 1 df,   p=0.0111
+Wald test            = 6.37  on 1 df,   p=0.0116
+Score (logrank) test = 6.39  on 1 df,   p=0.0115
+\end{CodeOutput}
+\end{CodeChunk}
+
+We can plot the CIF for each group:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
+R+                        ignore = 99)
+R> 
+R> plot(survfit(cox_model, newdata = new_data),
+R+      xlab = "Years since Randomization", 
+R+      ylab = "Cumulative Incidence", 
+R+      fun = "event",
+R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+                     exp(coef(cox_model)), 
+R+                     exp(confint(cox_model))[1], 
+R+                     exp(confint(cox_model))[2]))
+R> legend("topleft", 
+R+        legend = c("Control group", "Screening group"), 
+R+        col = c("red","blue"),
+R+        lty = c(1, 1), 
+R+        bg = "gray90")
+\end{CodeInput}
+
+\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-6-1} \end{center}
+
+\end{CodeChunk}
+
+Next we fit several models using case-base sampling. The models we fit
+differ in how we choose to model time.
+
+The \code{fitSmoothHazard} function provides an estimate of the hazard
+function \(h(x, t)\) is the hazard function, where \(t\) denotes the
+numerical value of a point in prognostic/prospective time and \(x\) is
+the realization of the vector \(X\) of variates based on the patient's
+profile and intervention (if any).
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
+R+                                                   data = ERSPC, 
+R+                                                   ratio = 100)
+\end{CodeInput}
+
+\begin{CodeOutput}
+'Follow.Up.Time' will be used as the time variable
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> summary(casebase_exponential)
+\end{CodeInput}
+
+\begin{CodeOutput}
+
+Call:
+glm(formula = formula, family = binomial, data = sampleData)
+
+Deviance Residuals: 
+   Min      1Q  Median      3Q     Max  
+-0.458  -0.458  -0.409  -0.409   2.247
+
+Coefficients:
+                      Estimate Std. Error z value            Pr(>|z|)    
+(Intercept)            -7.7546     0.0584  -132.8 <0.0000000000000002 ***
+ScrArmScreening group  -0.2397     0.0921    -2.6              0.0093 ** 
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+
+(Dispersion parameter for binomial family taken to be 1)
+
+    Null deviance: 3619.1  on 5939  degrees of freedom
+Residual deviance: 3612.2  on 5938  degrees of freedom
+AIC: 3616
+
+Number of Fisher Scoring iterations: 5
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> exp(coef(casebase_exponential)[2])
+\end{CodeInput}
+
+\begin{CodeOutput}
+ScrArmScreening group 
+                 0.79 
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> exp(confint(casebase_exponential)[2,])
+\end{CodeInput}
+
+\begin{CodeOutput}
+Waiting for profiling to be done...
+\end{CodeOutput}
+
+\begin{CodeOutput}
+ 2.5 % 97.5 % 
+  0.66   0.94 
+\end{CodeOutput}
+\end{CodeChunk}
+
+The \code{absoluteRisk} function provides an estimate of the cumulative
+incidence curves for a specific risk profile using the following
+equation:
+
+\[ CI(x, t) = 1 - exp\left( - \int_0^t h(x, u) \textrm{d}u \right) \]
+
+In the plot below, we overlay the estimated CIF from the casebase
+exponential model on the Cox model CIF:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
+R+                                           time = seq(0,15,0.1), 
+R+                                           newdata = new_data)
+R> 
+R> plot(survfit(cox_model, newdata = new_data),
+R+      xlab = "Years since Randomization", 
+R+      ylab = "Cumulative Incidence", 
+R+      fun = "event",
+R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+                     exp(coef(cox_model)), 
+R+                     exp(confint(cox_model))[1], 
+R+                     exp(confint(cox_model))[2]))
+R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
+R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
+R> 
+R> 
+R> legend("topleft", 
+R+        legend = c("Control group (Cox)","Control group (Casebase)",
+R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R+        col = c("red","red", "blue","blue"),
+R+        lty = c(1, 2, 1, 2), 
+R+        bg = "gray90")
+\end{CodeInput}
+
+\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-8-1} \end{center}
+
+\end{CodeChunk}
+
+As we can see, the exponential model is not a good fit. Based on what we
+observed in the population time plot, where more events are observed
+later on in time, this poor fit is expected. A constant hazard model
+would overestimate the cumulative incidence earlier on in time, and
+underestimate it later on; this is what we see on the cumulative
+incidence plot. This example demonstrates the benefits of population
+time plots as an exploratory analysis tool.
+
+Next we enter time linearly into the model:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
+R+                                  data = ERSPC, 
+R+                                  ratio = 100)
+\end{CodeInput}
+
+\begin{CodeOutput}
+'Follow.Up.Time' will be used as the time variable
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> summary(casebase_time)
+\end{CodeInput}
+
+\begin{CodeOutput}
+
+Call:
+glm(formula = formula, family = binomial, data = sampleData)
+
+Deviance Residuals: 
+   Min      1Q  Median      3Q     Max  
+-1.030  -0.489  -0.367  -0.278   2.740
+
+Coefficients:
+                      Estimate Std. Error z value            Pr(>|z|)    
+(Intercept)            -9.0228     0.1186  -76.07 <0.0000000000000002 ***
+Follow.Up.Time          0.2217     0.0159   13.97 <0.0000000000000002 ***
+ScrArmScreening group  -0.2680     0.0939   -2.85              0.0043 ** 
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+
+(Dispersion parameter for binomial family taken to be 1)
+
+    Null deviance: 3619.1  on 5939  degrees of freedom
+Residual deviance: 3402.7  on 5937  degrees of freedom
+AIC: 3409
+
+Number of Fisher Scoring iterations: 5
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> exp(coef(casebase_time))
+\end{CodeInput}
+
+\begin{CodeOutput}
+          (Intercept)        Follow.Up.Time ScrArmScreening group 
+              0.00012               1.24824               0.76490 
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> exp(confint(casebase_time))
+\end{CodeInput}
+
+\begin{CodeOutput}
+Waiting for profiling to be done...
+\end{CodeOutput}
+
+\begin{CodeOutput}
+                         2.5 %  97.5 %
+(Intercept)           0.000095 0.00015
+Follow.Up.Time        1.210257 1.28799
+ScrArmScreening group 0.635606 0.91877
+\end{CodeOutput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
+R+                                           time = seq(0,15,0.1), 
+R+                                           newdata = new_data)
+R> 
+R> plot(survfit(cox_model, newdata = new_data),
+R+      xlab = "Years since Randomization", 
+R+      ylab = "Cumulative Incidence", 
+R+      fun = "event",
+R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+                     exp(coef(cox_model)), 
+R+                     exp(confint(cox_model))[1], 
+R+                     exp(confint(cox_model))[2]))
+R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
+R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
+R> 
+R> legend("topleft", 
+R+        legend = c("Control group (Cox)","Control group (Casebase)",
+R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R+        col = c("red","red", "blue","blue"),
+R+        lty = c(1, 2, 1, 2), 
+R+        bg = "gray90")
+\end{CodeInput}
+
+\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-10-1} \end{center}
+
+\end{CodeChunk}
+
+We see that the Weibull model leads to a better fit.
+
+Next we try to enter a smooth function of time into the model using the
+\code{splines} package:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ splines::bs(Follow.Up.Time) + ScrArm, 
+R+                                     data = ERSPC, 
+R+                                     ratio = 100)
+\end{CodeInput}
+
+\begin{CodeOutput}
+'Follow.Up.Time' will be used as the time variable
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> summary(casebase_splines)
+\end{CodeInput}
+
+\begin{CodeOutput}
+
+Call:
+glm(formula = formula, family = binomial, data = sampleData)
+
+Deviance Residuals: 
+   Min      1Q  Median      3Q     Max  
+-1.245  -0.521  -0.397  -0.211   3.208
+
+Coefficients:
+                             Estimate Std. Error z value
+(Intercept)                  -10.5377     0.3384  -31.14
+splines::bs(Follow.Up.Time)1   4.9852     0.8551    5.83
+splines::bs(Follow.Up.Time)2   1.7658     0.4963    3.56
+splines::bs(Follow.Up.Time)3   5.1969     0.7203    7.22
+ScrArmScreening group         -0.1850     0.0941   -1.97
+                                         Pr(>|z|)    
+(Intercept)                  < 0.0000000000000002 ***
+splines::bs(Follow.Up.Time)1     0.00000000553602 ***
+splines::bs(Follow.Up.Time)2              0.00037 ***
+splines::bs(Follow.Up.Time)3     0.00000000000054 ***
+ScrArmScreening group                     0.04937 *  
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+
+(Dispersion parameter for binomial family taken to be 1)
+
+    Null deviance: 3619.1  on 5939  degrees of freedom
+Residual deviance: 3352.4  on 5935  degrees of freedom
+AIC: 3362
+
+Number of Fisher Scoring iterations: 6
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> exp(coef(casebase_splines))
+\end{CodeInput}
+
+\begin{CodeOutput}
+                 (Intercept) splines::bs(Follow.Up.Time)1 
+                    0.000027                   146.229029 
+splines::bs(Follow.Up.Time)2 splines::bs(Follow.Up.Time)3 
+                    5.846329                   180.718284 
+       ScrArmScreening group 
+                    0.831083 
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> exp(confint(casebase_splines))
+\end{CodeInput}
+
+\begin{CodeOutput}
+Waiting for profiling to be done...
+\end{CodeOutput}
+
+\begin{CodeOutput}
+                                 2.5 %    97.5 %
+(Intercept)                   0.000013   0.00005
+splines::bs(Follow.Up.Time)1 28.719353 825.32637
+splines::bs(Follow.Up.Time)2  2.210676  15.57574
+splines::bs(Follow.Up.Time)3 44.370102 759.24168
+ScrArmScreening group         0.690354   0.99867
+\end{CodeOutput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
+R+                                     time = seq(0,15,0.1), 
+R+                                     newdata = new_data)
+R> 
+R> plot(survfit(cox_model, newdata = new_data),
+R+      xlab = "Years since Randomization", 
+R+      ylab = "Cumulative Incidence", 
+R+      fun = "event",
+R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+                     exp(coef(cox_model)), 
+R+                     exp(confint(cox_model))[1], 
+R+                     exp(confint(cox_model))[2]))
+R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
+R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
+R> 
+R> legend("topleft", 
+R+        legend = c("Control group (Cox)","Control group (Casebase)",
+R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R+        col = c("red","red", "blue","blue"),
+R+        lty = c(1, 2, 1, 2), 
+R+        bg = "gray90")
+\end{CodeInput}
+
+\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-12-1} \end{center}
+
+\end{CodeChunk}
+
+It looks like the best fit.
+
+Since we are within the GLM framework, we can easily test for which
+model better fits the data using a Likelihood Ratio Test (LRT). The null
+hypothesis here is that the linear model is just as good as the larger
+(in terms of number of parameters) splines model.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> anova(casebase_time, casebase_splines, test = "LRT")
+\end{CodeInput}
+
+\begin{CodeOutput}
+Analysis of Deviance Table
+
+Model 1: DeadOfPrCa ~ Follow.Up.Time + ScrArm + offset(offset)
+Model 2: DeadOfPrCa ~ splines::bs(Follow.Up.Time) + ScrArm + offset(offset)
+  Resid. Df Resid. Dev Df Deviance       Pr(>Chi)    
+1      5937       3403                               
+2      5935       3352  2     50.3 0.000000000012 ***
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{CodeOutput}
+\end{CodeChunk}
+
+As espected, we see that splines model provides a better fit.
+
+\section{Case study 2--Bone-marrow
+transplant}\label{case-study-2bone-marrow-transplant}
 
@@
 \hline
-Sex & 0.82 & (0.46, 1.48)\\
+Sex & 0.72 & (0.4, 1.29)\\
 \hline
-Disease & 0.48 & (0.25, 0.92)\\
+Disease & 0.61 & (0.33, 1.15)\\
 \hline
-Phase (CR2 vs. CR1) & 1.43 & (0.56, 3.66)\\
+Phase (CR2 vs. CR1) & 1.31 & (0.52, 3.31)\\
 \hline
-Phase (CR3 vs. CR1) & 1.63 & (0.41, 6.45)\\
+Phase (CR3 vs. CR1) & 1.58 & (0.39, 6.32)\\
 \hline
-Phase (Relapse vs. CR1) & 4.73 & (2.13, 10.48)\\
+Phase (Relapse vs. CR1) & 4.50 & (2.07, 9.77)\\
 \hline
-Source & 1.51 & (0.49, 4.69)\\
+Source & 1.42 & (0.46, 4.41)\\
 \hline
-Age & 0.99 & (0.97, 1.02)\\
+Age & 1.00 & (0.97, 1.02)\\
 \hline
@@
 
-{\centering \includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-6-1}
+{\centering \includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-19-1}
 
@@
 \caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups.
-The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:unnamed-chunk-6}
+The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:unnamed-chunk-19}
 \end{figure}
@@
 \item
-  Give a more complex example of sampling; time-dependent exposure
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Sampling needs to be done manually, but fitting function can still
-    be used
-  \end{itemize}
+  Glmnet, gam, and gbm
 \end{itemize}
 
-\renewcommand\refname{Discussion}
+\section{Discussion}\label{discussion}
+
+In the following table we provide a comparison between the Cox model and
+case-base sampling:
+
 \bibliography{casebase_references.bib}

### 2019-07-22 README.Rmd
cleaning up repo into a more streamlined process so that Travis builds th
the paper. Using rrtools package for setup. Also fixed bug in rticles JSS template

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# cbpaper

[![Travis-CI Build Status](https://travis-ci.org/sahirbhatnagar/cbpaper.svg?branch=master)](https://travis-ci.org/sahirbhatnagar/cbpaper)

Source code for Casebase paper

for template, use

```R 
devtools::install_github("rstudio/rticles")
```

# Outline

1. Introduction
2. Theory
3. Implementation Details (Population time plots, Data analysis)
4. Case study 1 ERPSC (Single Event)
5. Case study 2 Bone Marrow Transplant (Competing risk)
6. Variable Selection (see https://cran.r-project.org/web/packages/TCGA2STAT/vignettes/TCGA2STAT.html for HD survival Data) 
7. Discussion

# To-DO (July 15)

1) Max: implement multinomial glmnet
2) Jesse: tests single and competing risk variable selection on TCGA. Look at variable selection litterature for competing risks. Find data
3) Sahir: Implementation details (population time plots)
4) Sahir: Review existing literature. What exists in terms of package. Look at Hanley and Miettinen. CRAN task view.

# To-DO (July 22)

1) Max: theory text
2) Jesse: tests single event variable selection on TCGA. Look at variable selection litterature for single. Plot KM curve with casebase+glmnet, and glmnet+cox
3) Sahir: Implementation details (population time plots), check issue.
4) Sahir: Review existing literature. What exists in terms of package. Look at Hanley and Miettinen. CRAN task view.

| Package | on CRAN     |  documentation |  Published | Description | Function call |
| :---    |  :---:      |    :---:        |    :---:   |  :---     |  :---:        |
| [flexsurv](https://cran.r-project.org/package=flexsurv)   | :heavy_check_mark:    | [Vignette](https://cran.r-project.org/web/packages/flexsurv/vignettes/flexsurv.pdf)    | [Jackson, C. JSS 2016](https://www.jstatsoft.org/article/view/v070i08) | Fully-parametric.  Any parametric time-to-event distribution may be fitted if the user supplies a probability density or hazard function, and ideally also their cumulative versions. Standard survival distributions are built in, including the three and four-parameter generalized gamma and Fdistributions. Any parameter of any distribution can be modelled as a linear or log-linear function of covariates. The package also includes the spline model of Royston and Parmar (2002), in which both baseline survival and covariate effects can be arbitrarily flexible parametric functions of time. See Table 1 for full list of distributions.  |  `flexsurvreg(Surv(recyrs, censrec) ~ group, data = bc, dist = "gengamma")`   `flexsurvspline(Surv(recyrs, censrec) ~ group, data = bc, k = 1, scale = "odds")`  |
| git diff     | git diff       | git diff      | dd | 44 | dd |

# Casebase paper

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to download or install

You can download the compendium as a zip from from this URL: </archive/master.zip>

Or you can install this compendium as an R package, pap, from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
remotes::install_github("/")
```

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2019-07-22 analysis/paper/paper.Rmd
cleaning up repo into a more streamlined process so that Travis builds th
the paper. Using rrtools package for setup. Also fixed bug in rticles JSS template

--- 2017-10-11 paper/casebase_jss.Rmd
@@
 ---
+documentclass: jss
 author:
@@
   - name: Maxime Turgeon *
+    affiliation: University of Manitoba
+    address: >
+      318 Machray Hall
+      Winnipeg, MB, Canada R3T 2N2
+    email: \email{max.turgeon@umanitoba.ca}
+    url: http://maxturgeon.ca/
+  - name: Jesse Islam 
     affiliation: McGill University
@@
       Montreal, QC, Canada H3A 1A2
-    email: \email{maxime.turgeon@mail.mcgill.ca}
-    url: http://maxturgeon.ca/
+    email: \email{jesse.islam@mail.mcgill.ca}
   - name: James Hanley
@@
   plain:     [keywords, not capitalized, Java]
-bibliography: casebase_references.bib
-biblio-style: jss
-csl: jss.csl
 preamble: >
   \usepackage{amsmath}
+  \usepackage{longtable}
 output: 
   rticles::jss_article:
+    number_sections: TRUE     #added argument option 
+    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
+biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
+bibliography: references.bib  #Also not included in template by default. 
+fig_width: 7
+fig_height: 6
 fig_caption: true
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  warning = FALSE,
+  message = FALSE,
+  echo = TRUE,
+  comment = "#>",
+  fig.path = "../figures/"
+)
+
 library(tidyverse)
@@
 set.seed(1234)
+
+eval_introduction <- TRUE
+eval_theory <- FALSE
+eval_implementation <- TRUE
+eval_cs1 <- FALSE
+eval_cs2 <- FALSE
+eval_cs3 <- FALSE
 ```
@@
 2. If, however, the absolute risks are of interest, they have to be recovered using the semi-parametric Breslow estimator
-3. Alternative approaches for fitting flexible hazard models for estimating absolute risks, not requiring this two-step approach? Yes! [@hanley2009fitting]
+3. Alternative approaches for fitting flexible hazard models for estimating absolute risks, not requiring this two-step approach? Yes!
+[@hanley2009fitting]
 
@@
 From the fitted hazard function, cumulative incidence and, thus, risk functions of time, treatment and profile can be easily derived.
+
+\begin{table}
+  \begin{tabular}{p{1.8in}lll}
+\hline
+    &  Parameters &  Density \proglang{R} function & \code{dist}\\
+    & {\footnotesize{(location in \code{\color{red}{red}})}} & & \\
+\hline
+    Exponential & \code{\color{red}{rate}}             & \code{dexp}   & \code{"exp"} \\
+    Weibull (accelerated failure time)     & \code{shape, {\color{red}{scale}}}     & \code{dweibull} & \code{"weibull"} \\
+    Weibull (proportional hazards)     & \code{shape, {\color{red}{scale}}}     & \code{dweibullPH} & \code{"weibullPH"} \\
+    Gamma       & \code{shape, \color{red}{rate}}      & \code{dgamma} & \code{"gamma"}\\
+    Log-normal  & \code{{\color{red}{meanlog}}, sdlog}   & \code{dlnorm} & \code{"lnorm"}\\
+    Gompertz    & \code{shape, {\color{red}{rate}}}      & \code{dgompertz} & \code{"gompertz"} \\
+    Log-logistic & \code{shape, {\color{red}{scale}}}   & \code{dllogis} & \code{"llogis"}\\
+    Generalized gamma (Prentice 1975)   & \code{{\color{red}{mu}}, sigma, Q} & \code{dgengamma} & \code{"gengamma"} \\
+    Generalized gamma (Stacy 1962)& \code{shape, {\color{red}{scale}}, k} & \code{dgengamma.orig} & \code{"gengamma.orig"} \\
+    Generalized F     (stable)    & \code{{\color{red}{mu}}, sigma, Q, P} & \code{dgenf} & \code{"genf"} \\
+    Generalized F     (original)  & \code{{\color{red}{mu}}, sigma, s1, s2} & \code{dgenf.orig} & \code{"genf.orig"} \\
+\hline
+  \end{tabular}
+  \caption{Built-in parametric survival distributions in \pkg{flexsurv}.}
+  \label{tab:dists}
+\end{table}
 
@@
 
-```{r}
+```{r erspc-data, eval = eval_cs1}
 data(ERSPC)
@@
 
-```{r}
+```{r plot-erspc-data, eval = eval_cs1}
 pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
@@
 
-```{r}
+```{r plot-stratified-erspc-data, eval = eval_cs1}
 pt_object_strat <- casebase::popTime(ERSPC, 
@@
 
-```{r}
+```{r plot-stratified-erspc-data-side, eval = eval_cs1}
 plot(pt_object_strat, ncol = 2)
@@
 
-```{r}
+```{r erspc-cox, eval = eval_cs1}
 cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
@@
 
-```{r}
+```{r erspc-cox-cif, eval = eval_cs1}
 new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
@@
 
-```{r}
+```{r erspc-casebase-exponential, eval = eval_cs1}
 casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
@@
 
-```{r}
+```{r erspc-casebase-cif, eval = eval_cs1}
 smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
@@
 
-```{r}
+```{r erspc-casebase-weibull, eval = eval_cs1}
 casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
@@
 
-```{r}
+```{r erspc-casebase-weibull-cif, eval = eval_cs1}
 smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
@@
 
-```{r}
+```{r erspc-casebase-splines, eval = eval_cs1}
 casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ splines::bs(Follow.Up.Time) + ScrArm, 
@@
 
-```{r}
+```{r erspc-casebase-splines-cif, eval = eval_cs1}
 smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
@@
 
-```{r}
+```{r erscp-compare, eval = eval_cs1}
 anova(casebase_time, casebase_splines, test = "LRT")
@@
 
-```{r message = FALSE}
+```{r bmtcrr-data, eval = eval_cs2, message = FALSE, echo=2}
 library(casebase)
@@
 
-```{r compPop1, fig.cap="\\label{fig:compPop1}Population-time plot for the stem-cell transplant study. The points represent the event of interest (i.e., relapse).", echo=FALSE}
+```{r compPop1, fig.cap="\\label{fig:compPop1}Population-time plot for the stem-cell transplant study. The points represent the event of interest (i.e., relapse).", echo=FALSE, eval = eval_cs2}
 pt_object <- casebase::popTime(bmtcrr, event = "Status", time = "ftime")
@@
 
-```{r compPop2, fig.cap="\\label{fig:compPop2}Population-time plot for the stem-cell transplant study. The points represent the competing events.", echo=FALSE}
+```{r compPop2, fig.cap="\\label{fig:compPop2}Population-time plot for the stem-cell transplant study. The points represent the competing events.", echo=FALSE, eval = eval_cs2}
 pt_object_comp <- bmtcrr %>% 
@@
 
-```{r warning = FALSE}
+```{r bmtcrr-casebase-weibull, warning = FALSE, eval = eval_cs2}
 model_cb <- fitSmoothHazard(
@@
 
-```{r, echo = FALSE, eval = TRUE}
+```{r bmtcrr-cox, echo = FALSE, eval = eval_cs2}
 library(survival)
@@
 
-```{r, echo=FALSE}
+```{r, echo=FALSE, eval = eval_cs2}
 # Table of coefficients----
@@
 
-```{r cb_risk, warning = FALSE, cache = TRUE}
+```{r cb_risk, warning = FALSE, cache = FALSE, eval = eval_cs2}
 # Pick 100 equidistant points between 0 and 60 months
-time_points <- seq(0, 60, length.out = 100)
+time_points <- seq(0, 60, length.out = 50)
 
@@
 risk_cb <- absoluteRisk(object = model_cb, time = time_points,
-                        method = "montecarlo", newdata = newdata)
+                        method = "numerical", newdata = newdata)
 ```
@@
 ```{r echo = FALSE, fig.cap="\\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups.
-The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate."}
+The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.", eval = eval_cs2}
 risk_cb <- bind_rows(
@@
 ```
+
+# Environment Details
+
+This report was generated on `r Sys.time()` using the following computational environment and dependencies:
+
+```{r colophon, cache = FALSE}
+# which R packages and versions?
+devtools::session_info()
+```
+
+The current Git commit details are:
+
+```{r}
+# what commit is this file at? 
+git2r::repository(here::here())
+```

### 2019-07-22 analysis/paper/paper.tex

--- 2017-09-25 paper/casebase_jss.tex
@@
-\documentclass[article]{jss}
+\documentclass[
+]{jss}
+
 \usepackage[utf8]{inputenc}
@@
 
-\usepackage{amsmath}
+\usepackage{amsmath} \usepackage{longtable}
 
@@
 
-\section{Code formatting}\label{code-formatting}
+\hypertarget{code-formatting}{%
+\section{Code formatting}\label{code-formatting}}
 
@@
 \begin{itemize}
-\tightlist
 \item
@@
 
-\section{Introduction}\label{introduction}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
-\begin{itemize}
+The purpose of the \texttt{casebase} package is to provide practitioners
+with an easy-to-use software tool to predict the risk (or cumulative
+incidence (CI)) of an event, for a particular patient. The following
+points should be noted:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
 \tightlist
 \item
-  Motivation
+  Time matching/risk set sampling (including Cox partial likelihood)
+  eliminates the baseline hazard from the likelihood expression for the
+  hazard ratios
+\item
+  If, however, the absolute risks are of interest, they have to be
+  recovered using the semi-parametric Breslow estimator
+\item
+  Alternative approaches for fitting flexible hazard models for
+  estimating absolute risks, not requiring this two-step approach? Yes!
+  \citep{hanley2009fitting}
+\end{enumerate}
 
-  \begin{itemize}
-  \tightlist
-  \item
-    Flexible
-  \item
-    Flexible
-  \item
-    Flexible
-  \end{itemize}
-\end{itemize}
+\citep{hanley2009fitting} propose a fully parametric hazard model that
+can be fit via logistic regression. From the fitted hazard function,
+cumulative incidence and, thus, risk functions of time, treatment and
+profile can be easily derived.
 
-\section{Theoretical details}\label{theoretical-details}
+\begin{table}
+  \begin{tabular}{p{1.8in}lll}
+\hline
+    &  Parameters &  Density \proglang{R} function & \code{dist}\\
+    & {\footnotesize{(location in \code{\color{red}{red}})}} & & \\
+\hline
+    Exponential & \code{\color{red}{rate}}             & \code{dexp}   & \code{"exp"} \\
+    Weibull (accelerated failure time)     & \code{shape, {\color{red}{scale}}}     & \code{dweibull} & \code{"weibull"} \\
+    Weibull (proportional hazards)     & \code{shape, {\color{red}{scale}}}     & \code{dweibullPH} & \code{"weibullPH"} \\
+    Gamma       & \code{shape, \color{red}{rate}}      & \code{dgamma} & \code{"gamma"}\\
+    Log-normal  & \code{{\color{red}{meanlog}}, sdlog}   & \code{dlnorm} & \code{"lnorm"}\\
+    Gompertz    & \code{shape, {\color{red}{rate}}}      & \code{dgompertz} & \code{"gompertz"} \\
+    Log-logistic & \code{shape, {\color{red}{scale}}}   & \code{dllogis} & \code{"llogis"}\\
+    Generalized gamma (Prentice 1975)   & \code{{\color{red}{mu}}, sigma, Q} & \code{dgengamma} & \code{"gengamma"} \\
+    Generalized gamma (Stacy 1962)& \code{shape, {\color{red}{scale}}, k} & \code{dgengamma.orig} & \code{"gengamma.orig"} \\
+    Generalized F     (stable)    & \code{{\color{red}{mu}}, sigma, Q, P} & \code{dgenf} & \code{"genf"} \\
+    Generalized F     (original)  & \code{{\color{red}{mu}}, sigma, s1, s2} & \code{dgenf.orig} & \code{"genf.orig"} \\
+\hline
+  \end{tabular}
+  \caption{Built-in parametric survival distributions in \pkg{flexsurv}.}
+  \label{tab:dists}
+\end{table}
+
+\hypertarget{theoretical-details}{%
+\section{Theoretical details}\label{theoretical-details}}
 
@@
 
-\section{Implementation details}\label{implementation-details}
+The \texttt{casebase} package fits the family of hazard functions of the
+form
+
+\[ h(x,t) = exp[g(x,t)] \] where \(t\) denotes the numerical value
+(number of units) of a point in prognostic/prospective time and \(x\) is
+the realization of the vector \(X\) of variates based on the patient's
+profile and intervention (if any). Different functions of \(t\) lead to
+different parametric hazard models.
+
+The simplest of these models is the one-parameter exponential
+distribution which is obtained by taking the hazard function to be
+constant over the range of \(t\).
+
+\[ h(x,t) = exp(\beta_0 + \beta_1 x) \]
+
+The instantaneous failure rate is independent of \(t\), so that the
+conditional chance of failure in a time interval of specified length is
+the same regardless of how long the individual has been on study a.k.a
+the memoryless property (Kalbfleisch and Prentice, 2002).
+
+The Gompertz hazard model is given by including a linear term for time:
+
+\[ h(x,t)  = exp(\beta_0 + \beta_1 t + \beta_2 x) \]
+
+Use of \(log(t)\) yields the Weibull hazard which allows for a power
+dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
+
+\[ h(x, t)  = exp(\beta_0 + \beta_1 \log(t) + \beta_2 x) \]
+
+\hypertarget{implementation-details}{%
+\section{Implementation details}\label{implementation-details}}
 
@@
 
-\subsection{Population-time plots}\label{population-time-plots}
+\hypertarget{population-time-plots}{%
+\subsection{Population-time plots}\label{population-time-plots}}
 
-\subsection{Data analysis}\label{data-analysis}
+\hypertarget{data-analysis}{%
+\subsection{Data analysis}\label{data-analysis}}
 
@@
 
-\section{Case study 1: Veteran data (or ERSPC if we
-can)}\label{case-study-1-veteran-data-or-erspc-if-we-can}
+\hypertarget{case-study-1european-randomized-study-of-prostate-cancer-screening}{%
+\section{Case study 1--European Randomized Study of Prostate Cancer
+Screening}\label{case-study-1european-randomized-study-of-prostate-cancer-screening}}
 
-\begin{itemize}
-\tightlist
-\item
-  First example
-\item
-  Show how we can test for non-proportional hazard?
-\end{itemize}
+To introduce the different features available, we make use of the
+European Randomized Study of Prostate Cancer Screening data; this
+dataset is available through the \pkg{casebase} package:
 
-\section{Case study 2: Bone-marrow
-transplant}\label{case-study-2-bone-marrow-transplant}
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> data(ERSPC)
+R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
+R>                        levels = c(0,1), 
+R>                        labels = c("Control group", "Screening group"))
+\end{CodeInput}
+\end{CodeChunk}
+
+The results of this study were published by {[}schroder2009screening{]}.
+This data was obtained using the approach described in
+{[}liu2014recovering{]}.
+
+Population time plots can be extremely informative graphical displays of
+survival data. They should be the first step in an exploratory data
+analysis. We facilitate this task in the \pkg{casebase} package using
+the \code{popTime} function. We first create the necessary dataset for
+producing the population time plots, and we can produce the plot by
+using the corresponding \code{plot} method:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
+R> plot(pt_object)
+\end{CodeInput}
+\end{CodeChunk}
+
+We can also create exposure stratified plots by specifying the
+\code{exposure} argument in the \code{popTime} function:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> pt_object_strat <- casebase::popTime(ERSPC, 
+R>                                      event = "DeadOfPrCa", 
+R>                                      exposure = "ScrArm")
+R> plot(pt_object_strat)
+\end{CodeInput}
+\end{CodeChunk}
+
+We can also plot them side-by-side using the \code{ncol} argument:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> plot(pt_object_strat, ncol = 2)
+\end{CodeInput}
+\end{CodeChunk}
+
+First, we fit a Cox model to the data, examine the hazard ratio for the
+screening group (relative to the control group), and plot the cumulative
+incidence function (CIF).
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
+R>                              data = ERSPC)
+R> summary(cox_model)
+\end{CodeInput}
+\end{CodeChunk}
+
+We can plot the CIF for each group:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
+R>                        ignore = 99)
+R> 
+R> plot(survfit(cox_model, newdata = new_data),
+R>      xlab = "Years since Randomization", 
+R>      ylab = "Cumulative Incidence", 
+R>      fun = "event",
+R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R>                     exp(coef(cox_model)), 
+R>                     exp(confint(cox_model))[1], 
+R>                     exp(confint(cox_model))[2]))
+R> legend("topleft", 
+R>        legend = c("Control group", "Screening group"), 
+R>        col = c("red","blue"),
+R>        lty = c(1, 1), 
+R>        bg = "gray90")
+\end{CodeInput}
+\end{CodeChunk}
+
+Next we fit several models using case-base sampling. The models we fit
+differ in how we choose to model time.
+
+The \code{fitSmoothHazard} function provides an estimate of the hazard
+function \(h(x, t)\) is the hazard function, where \(t\) denotes the
+numerical value of a point in prognostic/prospective time and \(x\) is
+the realization of the vector \(X\) of variates based on the patient's
+profile and intervention (if any).
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
+R>                                                   data = ERSPC, 
+R>                                                   ratio = 100)
+R> 
+R> summary(casebase_exponential)
+R> exp(coef(casebase_exponential)[2])
+R> exp(confint(casebase_exponential)[2,])
+\end{CodeInput}
+\end{CodeChunk}
+
+The \code{absoluteRisk} function provides an estimate of the cumulative
+incidence curves for a specific risk profile using the following
+equation:
+
+\[ CI(x, t) = 1 - exp\left( - \int_0^t h(x, u) \textrm{d}u \right) \]
+
+In the plot below, we overlay the estimated CIF from the casebase
+exponential model on the Cox model CIF:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
+R>                                           time = seq(0,15,0.1), 
+R>                                           newdata = new_data)
+R> 
+R> plot(survfit(cox_model, newdata = new_data),
+R>      xlab = "Years since Randomization", 
+R>      ylab = "Cumulative Incidence", 
+R>      fun = "event",
+R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R>                     exp(coef(cox_model)), 
+R>                     exp(confint(cox_model))[1], 
+R>                     exp(confint(cox_model))[2]))
+R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
+R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
+R> 
+R> 
+R> legend("topleft", 
+R>        legend = c("Control group (Cox)","Control group (Casebase)",
+R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R>        col = c("red","red", "blue","blue"),
+R>        lty = c(1, 2, 1, 2), 
+R>        bg = "gray90")
+\end{CodeInput}
+\end{CodeChunk}
+
+As we can see, the exponential model is not a good fit. Based on what we
+observed in the population time plot, where more events are observed
+later on in time, this poor fit is expected. A constant hazard model
+would overestimate the cumulative incidence earlier on in time, and
+underestimate it later on; this is what we see on the cumulative
+incidence plot. This example demonstrates the benefits of population
+time plots as an exploratory analysis tool.
+
+Next we enter time linearly into the model:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
+R>                                  data = ERSPC, 
+R>                                  ratio = 100)
+R> 
+R> summary(casebase_time)
+R> exp(coef(casebase_time))
+R> exp(confint(casebase_time))
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
+R>                                           time = seq(0,15,0.1), 
+R>                                           newdata = new_data)
+R> 
+R> plot(survfit(cox_model, newdata = new_data),
+R>      xlab = "Years since Randomization", 
+R>      ylab = "Cumulative Incidence", 
+R>      fun = "event",
+R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R>                     exp(coef(cox_model)), 
+R>                     exp(confint(cox_model))[1], 
+R>                     exp(confint(cox_model))[2]))
+R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
+R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
+R> 
+R> legend("topleft", 
+R>        legend = c("Control group (Cox)","Control group (Casebase)",
+R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R>        col = c("red","red", "blue","blue"),
+R>        lty = c(1, 2, 1, 2), 
+R>        bg = "gray90")
+\end{CodeInput}
+\end{CodeChunk}
+
+We see that the Weibull model leads to a better fit.
+
+Next we try to enter a smooth function of time into the model using the
+\code{splines} package:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ splines::bs(Follow.Up.Time) + ScrArm, 
+R>                                     data = ERSPC, 
+R>                                     ratio = 100)
+R> 
+R> summary(casebase_splines)
+R> exp(coef(casebase_splines))
+R> exp(confint(casebase_splines))
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
+R>                                     time = seq(0,15,0.1), 
+R>                                     newdata = new_data)
+R> 
+R> plot(survfit(cox_model, newdata = new_data),
+R>      xlab = "Years since Randomization", 
+R>      ylab = "Cumulative Incidence", 
+R>      fun = "event",
+R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R>                     exp(coef(cox_model)), 
+R>                     exp(confint(cox_model))[1], 
+R>                     exp(confint(cox_model))[2]))
+R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
+R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
+R> 
+R> legend("topleft", 
+R>        legend = c("Control group (Cox)","Control group (Casebase)",
+R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R>        col = c("red","red", "blue","blue"),
+R>        lty = c(1, 2, 1, 2), 
+R>        bg = "gray90")
+\end{CodeInput}
+\end{CodeChunk}
+
+It looks like the best fit.
+
+Since we are within the GLM framework, we can easily test for which
+model better fits the data using a Likelihood Ratio Test (LRT). The null
+hypothesis here is that the linear model is just as good as the larger
+(in terms of number of parameters) splines model.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> anova(casebase_time, casebase_splines, test = "LRT")
+\end{CodeInput}
+\end{CodeChunk}
+
+As espected, we see that splines model provides a better fit.
+
+\hypertarget{case-study-2bone-marrow-transplant}{%
+\section{Case study 2--Bone-marrow
+transplant}\label{case-study-2bone-marrow-transplant}}
 
@@
 \begin{CodeInput}
-R> library(casebase)
 R> data(bmtcrr)
@@
 
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{casebase_jss_files/figure-latex/compPop1-1}
-
-}
-
-\caption{\label{fig:compPop1}Population-time plot for the stem-cell transplant study. The points represent the event of interest (i.e., relapse).}\label{fig:compPop1}
-\end{figure}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{casebase_jss_files/figure-latex/compPop2-1}
-
-}
-
-\caption{\label{fig:compPop2}Population-time plot for the stem-cell transplant study. The points represent the competing events.}\label{fig:compPop2}
-\end{figure}
-\end{CodeChunk}
-
 Our main objective is to compute the absolute risk of relapse for a
@@
 R> model_cb <- fitSmoothHazard(
-R+     Status ~ ftime + Sex + D + Phase + Source + Age, 
-R+     data = bmtcrr, 
-R+     ratio = 100, 
-R+     time = "ftime")
+R>     Status ~ ftime + Sex + D + Phase + Source + Age, 
+R>     data = bmtcrr, 
+R>     ratio = 100, 
+R>     time = "ftime")
 \end{CodeInput}
@@
 corresponding confidence intervals:
-
-\begin{CodeChunk}
-
-\begin{tabular}{l|r|l}
-\hline
-Covariates & HR & 95\% CI\\
-\hline
-Sex & 0.82 & (0.46, 1.48)\\
-\hline
-Disease & 0.48 & (0.25, 0.92)\\
-\hline
-Phase (CR2 vs. CR1) & 1.43 & (0.56, 3.66)\\
-\hline
-Phase (CR3 vs. CR1) & 1.63 & (0.41, 6.45)\\
-\hline
-Phase (Relapse vs. CR1) & 4.73 & (2.13, 10.48)\\
-\hline
-Source & 1.51 & (0.49, 4.69)\\
-\hline
-Age & 0.99 & (0.97, 1.02)\\
-\hline
-\end{tabular}
-
-\end{CodeChunk}
 
@@
 R> # Pick 100 equidistant points between 0 and 60 months
-R> time_points <- seq(0, 60, length.out = 100)
+R> time_points <- seq(0, 60, length.out = 50)
 R> 
@@
 R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
-R+                                      levels = levels(bmtcrr[,"Sex"])),
-R+                       "D" = c("ALL", "AML"),
-R+                       "Phase" = factor(c("Relapse", "Relapse"), 
-R+                                        levels = levels(bmtcrr[,"Phase"])),
-R+                       "Age" = c(35, 35),
-R+                       "Source" = factor(c("PB", "PB"), 
-R+                                         levels = levels(bmtcrr[,"Source"])))
+R>                                      levels = levels(bmtcrr[,"Sex"])),
+R>                       "D" = c("ALL", "AML"),
+R>                       "Phase" = factor(c("Relapse", "Relapse"), 
+R>                                        levels = levels(bmtcrr[,"Phase"])),
+R>                       "Age" = c(35, 35),
+R>                       "Source" = factor(c("PB", "PB"), 
+R>                                         levels = levels(bmtcrr[,"Source"])))
 R> 
@@
 R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
-R+                         method = "montecarlo", newdata = newdata)
+R>                         method = "numerical", newdata = newdata)
 \end{CodeInput}
@@
 
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-6-1}
-
-}
-
-\caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups.
-The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:unnamed-chunk-6}
-\end{figure}
-\end{CodeChunk}
-
-\section{Case study 3: Vaccination study (recurrent
-events)}\label{case-study-3-vaccination-study-recurrent-events}
+\hypertarget{case-study-3}{%
+\section{Case study 3}\label{case-study-3}}
 
@@
 \item
-  Give a more complex example of sampling; time-dependent exposure
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Sampling needs to be done manually, but fitting function can still
-    be used
-  \end{itemize}
+  Glmnet, gam, and gbm
 \end{itemize}
 
-\renewcommand\refname{Discussion}
-\bibliography{casebase_references.bib}
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+In the following table we provide a comparison between the Cox model and
+case-base sampling:
+
+\hypertarget{environment-details}{%
+\section{Environment Details}\label{environment-details}}
+
+This report was generated on 2019-07-22 17:17:57 using the following
+computational environment and dependencies:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # which R packages and versions?
+R> devtools::session_info()
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> - Session info ----------------------------------------------------------
+#>  setting  value                       
+#>  version  R version 3.6.0 (2019-04-26)
+#>  os       Pop!_OS 18.10               
+#>  system   x86_64, linux-gnu           
+#>  ui       X11                         
+#>  language en_CA:en                    
+#>  collate  en_CA.UTF-8                 
+#>  ctype    en_CA.UTF-8                 
+#>  tz       America/Toronto             
+#>  date     2019-07-22                  
+#> 
+#> - Packages --------------------------------------------------------------
+#>  package     * version    date       lib
+#>  assertthat    0.2.1      2019-03-21 [1]
+#>  backports     1.1.4      2019-04-10 [1]
+#>  broom         0.5.1      2018-12-05 [1]
+#>  callr         3.2.0      2019-03-15 [1]
+#>  casebase    * 0.1.1.9000 2019-07-22 [1]
+#>  cellranger    1.1.0      2016-07-27 [1]
+#>  cli           1.1.0      2019-03-19 [1]
+#>  colorspace    1.4-1      2019-03-18 [1]
+#>  crayon        1.3.4      2017-09-16 [1]
+#>  data.table    1.12.2     2019-04-07 [1]
+#>  desc          1.2.0      2018-05-01 [1]
+#>  devtools      2.0.1      2018-10-26 [1]
+#>  digest        0.6.20     2019-07-04 [1]
+#>  dplyr       * 0.8.0.1    2019-02-15 [1]
+#>  evaluate      0.14       2019-05-28 [1]
+#>  forcats     * 0.4.0      2019-02-17 [1]
+#>  fs            1.3.1      2019-05-06 [1]
+#>  generics      0.0.2      2018-11-29 [1]
+#>  ggplot2     * 3.2.0      2019-06-16 [1]
+#>  glue          1.3.1      2019-03-12 [1]
+#>  gtable        0.3.0      2019-03-25 [1]
+#>  haven         2.1.0      2019-02-19 [1]
+#>  hms           0.4.2      2018-03-10 [1]
+#>  htmltools     0.3.6      2017-04-28 [1]
+#>  httr          1.4.0      2018-12-11 [1]
+#>  jsonlite      1.6        2018-12-07 [1]
+#>  knitr         1.23       2019-05-18 [1]
+#>  lattice       0.20-38    2018-11-04 [5]
+#>  lazyeval      0.2.2      2019-03-15 [1]
+#>  lubridate     1.7.4      2018-04-11 [1]
+#>  magrittr    * 1.5        2014-11-22 [1]
+#>  Matrix        1.2-17     2019-03-22 [5]
+#>  memoise       1.1.0      2017-04-21 [1]
+#>  mgcv          1.8-28     2019-03-21 [5]
+#>  modelr        0.1.4      2019-02-18 [1]
+#>  munsell       0.5.0      2018-06-12 [1]
+#>  nlme          3.1-140    2019-05-12 [5]
+#>  pillar        1.4.2      2019-06-29 [1]
+#>  pkgbuild      1.0.3      2019-03-20 [1]
+#>  pkgconfig     2.0.2      2018-08-16 [1]
+#>  pkgload       1.0.2      2018-10-29 [1]
+#>  prettyunits   1.0.2      2015-07-13 [1]
+#>  processx      3.3.1      2019-05-08 [1]
+#>  ps            1.3.0      2018-12-21 [1]
+#>  purrr       * 0.3.2      2019-03-15 [1]
+#>  R6            2.4.0      2019-02-14 [1]
+#>  Rcpp          1.0.1      2019-03-17 [1]
+#>  readr       * 1.3.1      2018-12-21 [1]
+#>  readxl        1.3.0      2019-02-15 [1]
+#>  remotes       2.0.2      2018-10-30 [1]
+#>  rlang         0.4.0      2019-06-25 [1]
+#>  rmarkdown     1.14       2019-07-12 [1]
+#>  rprojroot     1.3-2      2018-01-03 [1]
+#>  rstudioapi    0.10       2019-03-19 [1]
+#>  rticles       0.9.1      2019-07-22 [1]
+#>  rvest         0.3.2      2016-06-17 [1]
+#>  scales        1.0.0      2018-08-09 [1]
+#>  sessioninfo   1.1.1      2018-11-05 [1]
+#>  stringi       1.4.3      2019-03-12 [1]
+#>  stringr     * 1.4.0      2019-02-10 [1]
+#>  survival    * 2.43-3     2018-11-26 [5]
+#>  testthat      2.1.1      2019-04-23 [1]
+#>  tibble      * 2.1.3      2019-06-06 [1]
+#>  tidyr       * 0.8.2      2018-10-28 [1]
+#>  tidyselect    0.2.5      2018-10-11 [1]
+#>  tidyverse   * 1.2.1      2017-11-14 [1]
+#>  usethis       1.5.0.9000 2019-06-24 [1]
+#>  VGAM          1.1-1      2019-02-18 [1]
+#>  withr         2.1.2      2018-03-15 [1]
+#>  xfun          0.8        2019-06-25 [1]
+#>  xml2          1.2.0      2018-01-24 [1]
+#>  yaml          2.2.0      2018-07-25 [1]
+#>  source                                  
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.0)                          
+#>  Github (sahirbhatnagar/casebase@669a0eb)
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.3)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.3)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  Github (rstudio/rticles@8d56fc6)        
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  Github (r-lib/usethis@d4eabb9)          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.5.1)                          
+#> 
+#> [1] /home/sahir/R/x86_64-pc-linux-gnu-library/3.5
+#> [2] /home/sahir/R/x86_64-pc-linux-gnu-library/3.6
+#> [3] /usr/local/lib/R/site-library
+#> [4] /usr/lib/R/site-library
+#> [5] /usr/lib/R/library
+\end{CodeOutput}
+\end{CodeChunk}
+
+The current Git commit details are:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # what commit is this file at? 
+R> git2r::repository(here::here())
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> Local:    litreview /home/sahir/git_repositories/cbpaper
+#> Head:     [38153a4] 2019-07-22: Update README.md
+\end{CodeOutput}
+\end{CodeChunk}
+
+\bibliography{references.bib}
 

### 2019-07-22 analysis/paper/paper.Rmd
Merge with master following the change to a new directory structure

@@
     address: >
-      318 Machray Hall
-      Winnipeg, MB, Canada R3T 2N2
+      186 Dysart Road
+      Winnipeg, MB, Canada R3B 0S8
     email: \email{max.turgeon@umanitoba.ca}
-    url: http://maxturgeon.ca/
+    url: https://maxturgeon.ca/
   - name: Jesse Islam 
@@
 ```{r erspc-casebase-splines, eval = eval_cs1}
-casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ splines::bs(Follow.Up.Time) + ScrArm, 
+library(splines)
+casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm,
                                     data = ERSPC, 

### 2019-07-22 analysis/paper/paper.tex
Merge with master following the change to a new directory structure

@@
 \author{
-Sahir Bhatnagar *\\McGill Univeristy \And Maxime Turgeon *\\McGill University \And James Hanley\\McGill Univeristy \And Olli Saarela\\University of Toronto
+Sahir Bhatnagar *\\McGill Univeristy \And Maxime Turgeon *\\University of Manitoba \And Jesse Islam\\McGill University \And James Hanley\\McGill Univeristy \And Olli Saarela\\University of Toronto
 }
@@
 
-\Plainauthor{Sahir Bhatnagar *, Maxime Turgeon *, James Hanley, Olli Saarela}
+\Plainauthor{Sahir Bhatnagar *, Maxime Turgeon *, Jesse Islam, James Hanley, Olli Saarela}
 \Plaintitle{casebase: An Alternative Framework For Survival Analysis}
@@
       Maxime Turgeon *\\
+  University of Manitoba\\
+  186 Dysart Road Winnipeg, MB, Canada R3B 0S8\\
+  E-mail: \email{max.turgeon@umanitoba.ca}\\
+  URL: \url{https://maxturgeon.ca/}\\~\\
+      Jesse Islam\\
   McGill University\\
   1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
-  E-mail: \email{maxime.turgeon@mail.mcgill.ca}\\
-  URL: \url{http://maxturgeon.ca/}\\~\\
+  E-mail: \email{jesse.islam@mail.mcgill.ca}\\
+
       James Hanley\\
@@
 \begin{CodeInput}
-R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ splines::bs(Follow.Up.Time) + ScrArm, 
+R> library(splines)
+R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm,
 R>                                     data = ERSPC, 
@@
 
-This report was generated on 2019-07-22 17:17:57 using the following
+This report was generated on 2019-07-22 16:50:40 using the following
 computational environment and dependencies:
@@
 #>  setting  value                       
-#>  version  R version 3.6.0 (2019-04-26)
-#>  os       Pop!_OS 18.10               
+#>  version  R version 3.6.1 (2019-07-05)
+#>  os       Ubuntu 18.04.2 LTS          
 #>  system   x86_64, linux-gnu           
 #>  ui       X11                         
-#>  language en_CA:en                    
+#>  language (EN)                        
 #>  collate  en_CA.UTF-8                 
 #>  ctype    en_CA.UTF-8                 
-#>  tz       America/Toronto             
+#>  tz       America/Winnipeg            
 #>  date     2019-07-22                  
@@
 #> - Packages --------------------------------------------------------------
-#>  package     * version    date       lib
-#>  assertthat    0.2.1      2019-03-21 [1]
-#>  backports     1.1.4      2019-04-10 [1]
-#>  broom         0.5.1      2018-12-05 [1]
-#>  callr         3.2.0      2019-03-15 [1]
-#>  casebase    * 0.1.1.9000 2019-07-22 [1]
-#>  cellranger    1.1.0      2016-07-27 [1]
-#>  cli           1.1.0      2019-03-19 [1]
-#>  colorspace    1.4-1      2019-03-18 [1]
-#>  crayon        1.3.4      2017-09-16 [1]
-#>  data.table    1.12.2     2019-04-07 [1]
-#>  desc          1.2.0      2018-05-01 [1]
-#>  devtools      2.0.1      2018-10-26 [1]
-#>  digest        0.6.20     2019-07-04 [1]
-#>  dplyr       * 0.8.0.1    2019-02-15 [1]
-#>  evaluate      0.14       2019-05-28 [1]
-#>  forcats     * 0.4.0      2019-02-17 [1]
-#>  fs            1.3.1      2019-05-06 [1]
-#>  generics      0.0.2      2018-11-29 [1]
-#>  ggplot2     * 3.2.0      2019-06-16 [1]
-#>  glue          1.3.1      2019-03-12 [1]
-#>  gtable        0.3.0      2019-03-25 [1]
-#>  haven         2.1.0      2019-02-19 [1]
-#>  hms           0.4.2      2018-03-10 [1]
-#>  htmltools     0.3.6      2017-04-28 [1]
-#>  httr          1.4.0      2018-12-11 [1]
-#>  jsonlite      1.6        2018-12-07 [1]
-#>  knitr         1.23       2019-05-18 [1]
-#>  lattice       0.20-38    2018-11-04 [5]
-#>  lazyeval      0.2.2      2019-03-15 [1]
-#>  lubridate     1.7.4      2018-04-11 [1]
-#>  magrittr    * 1.5        2014-11-22 [1]
-#>  Matrix        1.2-17     2019-03-22 [5]
-#>  memoise       1.1.0      2017-04-21 [1]
-#>  mgcv          1.8-28     2019-03-21 [5]
-#>  modelr        0.1.4      2019-02-18 [1]
-#>  munsell       0.5.0      2018-06-12 [1]
-#>  nlme          3.1-140    2019-05-12 [5]
-#>  pillar        1.4.2      2019-06-29 [1]
-#>  pkgbuild      1.0.3      2019-03-20 [1]
-#>  pkgconfig     2.0.2      2018-08-16 [1]
-#>  pkgload       1.0.2      2018-10-29 [1]
-#>  prettyunits   1.0.2      2015-07-13 [1]
-#>  processx      3.3.1      2019-05-08 [1]
-#>  ps            1.3.0      2018-12-21 [1]
-#>  purrr       * 0.3.2      2019-03-15 [1]
-#>  R6            2.4.0      2019-02-14 [1]
-#>  Rcpp          1.0.1      2019-03-17 [1]
-#>  readr       * 1.3.1      2018-12-21 [1]
-#>  readxl        1.3.0      2019-02-15 [1]
-#>  remotes       2.0.2      2018-10-30 [1]
-#>  rlang         0.4.0      2019-06-25 [1]
-#>  rmarkdown     1.14       2019-07-12 [1]
-#>  rprojroot     1.3-2      2018-01-03 [1]
-#>  rstudioapi    0.10       2019-03-19 [1]
-#>  rticles       0.9.1      2019-07-22 [1]
-#>  rvest         0.3.2      2016-06-17 [1]
-#>  scales        1.0.0      2018-08-09 [1]
-#>  sessioninfo   1.1.1      2018-11-05 [1]
-#>  stringi       1.4.3      2019-03-12 [1]
-#>  stringr     * 1.4.0      2019-02-10 [1]
-#>  survival    * 2.43-3     2018-11-26 [5]
-#>  testthat      2.1.1      2019-04-23 [1]
-#>  tibble      * 2.1.3      2019-06-06 [1]
-#>  tidyr       * 0.8.2      2018-10-28 [1]
-#>  tidyselect    0.2.5      2018-10-11 [1]
-#>  tidyverse   * 1.2.1      2017-11-14 [1]
-#>  usethis       1.5.0.9000 2019-06-24 [1]
-#>  VGAM          1.1-1      2019-02-18 [1]
-#>  withr         2.1.2      2018-03-15 [1]
-#>  xfun          0.8        2019-06-25 [1]
-#>  xml2          1.2.0      2018-01-24 [1]
-#>  yaml          2.2.0      2018-07-25 [1]
-#>  source                                  
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.0)                          
-#>  Github (sahirbhatnagar/casebase@669a0eb)
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.3)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.3)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  Github (rstudio/rticles@8d56fc6)        
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  Github (r-lib/usethis@d4eabb9)          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.5.1)                          
+#>  package     * version    date       lib source                          
+#>  assertthat    0.2.0      2017-04-11 [1] CRAN (R 3.5.0)                  
+#>  backports     1.1.2      2017-12-13 [1] CRAN (R 3.5.0)                  
+#>  broom         0.4.4      2018-03-29 [1] CRAN (R 3.5.0)                  
+#>  callr         3.3.0      2019-07-04 [1] CRAN (R 3.6.1)                  
+#>  casebase    * 0.1.0.9000 2019-07-19 [1] local                           
+#>  cellranger    1.1.0      2016-07-27 [1] CRAN (R 3.5.0)                  
+#>  cli           1.1.0      2019-03-19 [1] CRAN (R 3.6.1)                  
+#>  colorspace    1.3-2      2016-12-14 [1] CRAN (R 3.5.0)                  
+#>  crayon        1.3.4      2017-09-16 [1] CRAN (R 3.5.0)                  
+#>  data.table    1.11.2     2018-05-08 [1] CRAN (R 3.5.0)                  
+#>  desc          1.2.0      2018-05-01 [1] CRAN (R 3.5.0)                  
+#>  devtools      2.1.0      2019-07-06 [1] CRAN (R 3.6.1)                  
+#>  digest        0.6.20     2019-07-04 [1] CRAN (R 3.6.1)                  
+#>  dplyr       * 0.8.0.1    2019-02-15 [1] CRAN (R 3.5.2)                  
+#>  evaluate      0.14       2019-05-28 [1] CRAN (R 3.6.1)                  
+#>  forcats     * 0.3.0      2018-02-19 [1] CRAN (R 3.5.0)                  
+#>  foreign       0.8-70     2018-04-23 [1] CRAN (R 3.5.0)                  
+#>  fs            1.3.1      2019-05-06 [1] CRAN (R 3.6.1)                  
+#>  ggplot2     * 3.2.0      2019-06-16 [1] CRAN (R 3.6.1)                  
+#>  glue          1.3.1      2019-03-12 [1] CRAN (R 3.6.1)                  
+#>  gtable        0.2.0      2016-02-26 [1] CRAN (R 3.5.0)                  
+#>  haven         1.1.1      2018-01-18 [1] CRAN (R 3.5.0)                  
+#>  hms           0.4.2      2018-03-10 [1] CRAN (R 3.5.0)                  
+#>  htmltools     0.3.6      2017-04-28 [1] CRAN (R 3.5.0)                  
+#>  httr          1.3.1      2017-08-20 [1] CRAN (R 3.5.0)                  
+#>  jsonlite      1.6        2018-12-07 [1] CRAN (R 3.6.1)                  
+#>  knitr         1.23       2019-05-18 [1] CRAN (R 3.6.1)                  
+#>  lattice       0.20-38    2018-11-04 [4] CRAN (R 3.5.1)                  
+#>  lazyeval      0.2.1      2017-10-29 [1] CRAN (R 3.5.0)                  
+#>  lubridate     1.7.4      2018-04-11 [1] CRAN (R 3.5.0)                  
+#>  magrittr    * 1.5        2014-11-22 [1] CRAN (R 3.5.0)                  
+#>  Matrix        1.2-17     2019-03-22 [4] CRAN (R 3.5.3)                  
+#>  memoise       1.1.0      2017-04-21 [1] CRAN (R 3.5.0)                  
+#>  mgcv          1.8-24     2018-06-18 [1] CRAN (R 3.5.1)                  
+#>  mnormt        1.5-5      2016-10-15 [1] CRAN (R 3.5.0)                  
+#>  modelr        0.1.2      2018-05-11 [1] CRAN (R 3.5.0)                  
+#>  munsell       0.4.3      2016-02-13 [1] CRAN (R 3.5.0)                  
+#>  nlme          3.1-140    2019-05-12 [4] CRAN (R 3.6.0)                  
+#>  pillar        1.3.1      2018-12-15 [1] CRAN (R 3.5.2)                  
+#>  pkgbuild      1.0.3      2019-03-20 [1] CRAN (R 3.6.1)                  
+#>  pkgconfig     2.0.2      2018-08-16 [1] CRAN (R 3.5.2)                  
+#>  pkgload       1.0.2      2018-10-29 [1] CRAN (R 3.6.1)                  
+#>  plyr          1.8.4      2016-06-08 [1] CRAN (R 3.5.0)                  
+#>  prettyunits   1.0.2      2015-07-13 [1] CRAN (R 3.5.2)                  
+#>  processx      3.4.0      2019-07-03 [1] CRAN (R 3.6.1)                  
+#>  ps            1.3.0      2018-12-21 [1] CRAN (R 3.6.1)                  
+#>  psych         1.8.4      2018-05-06 [1] CRAN (R 3.5.0)                  
+#>  purrr       * 0.2.4      2017-10-18 [1] CRAN (R 3.5.0)                  
+#>  R6            2.2.2      2017-06-17 [1] CRAN (R 3.5.0)                  
+#>  Rcpp          1.0.1      2019-03-17 [1] CRAN (R 3.6.1)                  
+#>  readr       * 1.1.1      2017-05-16 [1] CRAN (R 3.5.0)                  
+#>  readxl        1.1.0      2018-04-20 [1] CRAN (R 3.5.0)                  
+#>  remotes       2.1.0      2019-06-24 [1] CRAN (R 3.6.1)                  
+#>  reshape2      1.4.3      2017-12-11 [1] CRAN (R 3.5.0)                  
+#>  rlang         0.4.0      2019-06-25 [1] CRAN (R 3.6.1)                  
+#>  rmarkdown     1.14       2019-07-12 [1] CRAN (R 3.6.1)                  
+#>  rprojroot     1.3-2      2018-01-03 [1] CRAN (R 3.5.0)                  
+#>  rstudioapi    0.7        2017-09-07 [1] CRAN (R 3.5.0)                  
+#>  rticles       0.9.1      2019-07-22 [1] Github (rstudio/rticles@8d56fc6)
+#>  rvest         0.3.2      2016-06-17 [1] CRAN (R 3.5.0)                  
+#>  scales        0.5.0      2017-08-24 [1] CRAN (R 3.5.0)                  
+#>  sessioninfo   1.1.1      2018-11-05 [1] CRAN (R 3.6.1)                  
+#>  stringi       1.4.3      2019-03-12 [1] CRAN (R 3.6.1)                  
+#>  stringr     * 1.4.0      2019-02-10 [1] CRAN (R 3.6.1)                  
+#>  survival    * 2.43-3     2018-11-26 [4] CRAN (R 3.5.1)                  
+#>  testthat      2.1.1      2019-04-23 [1] CRAN (R 3.6.1)                  
+#>  tibble      * 2.0.1      2019-01-12 [1] CRAN (R 3.5.2)                  
+#>  tidyr       * 0.8.1      2018-05-18 [1] CRAN (R 3.5.0)                  
+#>  tidyselect    0.2.5      2018-10-11 [1] CRAN (R 3.5.2)                  
+#>  tidyverse   * 1.2.1      2017-11-14 [1] CRAN (R 3.5.0)                  
+#>  usethis       1.5.1      2019-07-04 [1] CRAN (R 3.6.1)                  
+#>  VGAM          1.1-1      2019-02-18 [1] CRAN (R 3.6.1)                  
+#>  withr         2.1.2      2018-03-15 [1] CRAN (R 3.5.0)                  
+#>  xfun          0.8        2019-06-25 [1] CRAN (R 3.6.1)                  
+#>  xml2          1.2.0      2018-01-24 [1] CRAN (R 3.5.0)                  
+#>  yaml          2.2.0      2018-07-25 [1] CRAN (R 3.6.1)                  
 #> 
-#> [1] /home/sahir/R/x86_64-pc-linux-gnu-library/3.5
-#> [2] /home/sahir/R/x86_64-pc-linux-gnu-library/3.6
-#> [3] /usr/local/lib/R/site-library
-#> [4] /usr/lib/R/site-library
-#> [5] /usr/lib/R/library
+#> [1] /home/mturgeon/Rlibs
+#> [2] /usr/local/lib/R/site-library
+#> [3] /usr/lib/R/site-library
+#> [4] /usr/lib/R/library
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> Local:    litreview /home/sahir/git_repositories/cbpaper
-#> Head:     [38153a4] 2019-07-22: Update README.md
+#> Local:    review-max /home/mturgeon/Documents/git_repositories/cbpaper
+#> Head:     [c8ac771] 2019-07-22: Merge pull request #5 from sahirbhatnagar/litreview
 \end{CodeOutput}

### 2019-07-23 README.Rmd

@@
 
+[![Binder](http://mybinder.org/badge.svg)](http://beta.mybinder.org/v2/gh/sahirbhatnagar/cbpaper/master?urlpath=rstudio)  
 [![Travis-CI Build Status](https://travis-ci.org/sahirbhatnagar/cbpaper.svg?branch=master)](https://travis-ci.org/sahirbhatnagar/cbpaper)

### 2019-07-25 analysis/paper/paper.Rmd
review theory and implemntation sections

--- 2019-07-22 analysis/paper/paper.Rmd
@@
 
-# Code formatting
+# Code formatting {.unnumbered}
 
@@
 
-Popular survival analysis methods, like Kaplan-Meier and Cox regression, rely on the notion of risk-set sampling. Case-base sampling can be seen as an alternative.
+<!-- Popular survival analysis methods, like Kaplan-Meier and Cox regression, rely on the notion of risk-set sampling. Case-base sampling can be seen as an alternative. -->
 
 As shown by Saarela & Arjas [-@saarela2015non] (and further expanded in Saarela [-@saarela2016case]), writing the likelihood arising from this data-generating mechanism using the framework of non-homogenous Poisson processes, we eventually reach an expression where each person-moment's contribution is of the form
-$$\frac{\lambda(t)^{dN(t)}}{\rho(t) + \lambda(t)},$$
-where $N(t)$ is the counting process associated with the event of interest, $\lambda(t)$ is the corresponding hazard function, and $\rho(t)$ is the hazard function for the Poisson process associated with case-base sampling.
-This parametric form suggests that we can readily estimate log-hazards of the form $\log(\lambda(t)) = g(t; X)$ using logistic regression, where each observation corresponds to a person moment, the function $g(t; X)$ is linear in a finite number of parameters, and where we treat $\log(\rho(t))$ as an offset.
+$$\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},$$
+where $N(t)$ is the counting process associated with the event of interest, $h(t)$ is the corresponding hazard function, and $\rho(t)$ is the hazard function for the Poisson process associated with case-base sampling.
+This parametric form suggests that we can readily estimate log-hazards of the form $\log(h(t)) = g(t; X)$ using logistic regression, where each observation corresponds to a person moment, the function $g(t; X)$ is linear in a finite number of parameters, and where we treat $-\log(\rho(t))$ as an offset.
 
@@
 to sample the base series uniformly from the study base.
-In terms of Poisson processes, this sampling strategy corresponds to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base.
+In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base.
 More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism  are proportional to the cardiovascular disease event rate given by the Framingham score.
@@
 
-$$ h(x,t) = exp[g(x,t)] $$
-where $t$ denotes the numerical value (number of units) of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any).
+$$ h(t;X) = \exp[g(t;X)] $$
+where $t$ denotes time and $X$, the individual's covariate profile.
 Different functions of $t$ lead to different parametric hazard models.
-
 The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$.
 
-$$ h(x,t) = exp(\beta_0 + \beta_1 x) $$
+$$ h(t;X) = \exp(\beta_0 + \beta_1 X) $$
 
-The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been on study a.k.a the memoryless property (Kalbfleisch and Prentice, 2002).
+The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
+This is also known as the *memoryless property* (Kalbfleisch and Prentice, 2002).
 
@@
 
-$$ h(x,t)  = exp(\beta_0 + \beta_1 t + \beta_2 x) $$
+$$ h(t;X)  = \exp(\beta_0 + \beta_1 t + \beta_2 X) $$
 
-Use of $log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
+Use of $\log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
 
-$$ h(x, t)  = exp(\beta_0 + \beta_1 \log(t) + \beta_2 x) $$
+$$ h(t;X)  = \exp(\beta_0 + \beta_1 \log(t) + \beta_2 X) $$
+
+For competing-risk analyses with $J$ possible events, we can show that each person-moment's contribution of the likelihood is of the form
+
+$$\frac{h_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^Jh_j(t)},$$
+
+where $N_j(t)$ is the counting process associated with the event of type $j$ and $h_j(t)$ is the corresponding hazard function.
+As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1\end{align*}}
 
@@
 
-The functions in the casebase package can be divided into two categories: 1) data visualization, in the form of population-time plots; and 2) data analysis.
-
-We explicitely aimed at being compatible with both \code{data.frame}s and \code{data.table}s. This is evident in some of the coding choices we made, and it is also reflected in our testing units.
+The functions in the casebase package can be divided into two categories: 1) data visualization, in the form of population-time plots; and 2) parametric modeling.
+We explicitely aimed at being compatible with both \code{data.frame}s and \code{data.table}s.
+This is evident in some of the coding choices we made, and it is also reflected in our unit tests.
 
@@
 
-## Data analysis
+## Parametric modeling
 
-The data analysis step was separated into three parts: 1) case-base sampling; 2) estimation of the smooth hazard function; and 3) calculation of the risk function.
+The parametric modeling step was separated into three parts:
+
+  1. case-base sampling; 
+  2. estimation of the smooth hazard function; 
+  3. calculation of the risk function.
+
 By separating the sampling and estimation functions, we allowed the possibility of users implementing more complex sampling scheme, as described in Saarela [-@saarela2016case].
@@
 
-Finally, the hazard function is fitted to the data using the function \code{glm}, unless there is more than one type event (i.e.
-in a competing-risk analysis), in which case it uses the multinomical regression capabilities of the \pkg{VGAM} package.
+To fit the hazard function, we provide several approaches that are available via the \code{family} parameter. These approaches are:
+
+  - \code{glm}: This is the familiar logistic regression.
+  - \code{glmnet}: This option allows for variable selection using Lasso or elastic-net. This functionality is provided through the \code{glmnet} package [@friedman2010jss].
+  - \code{gam}: This option provides support for *Generalized Additive Models* via the \code{gam} package [@hastie1987generalized].
+  - \code{gbm}: This option provides support for *Gradient Boosted Trees* via the \code{gbm} package. This feature is still experimental.
+
+In the case of multiple events, the hazard is fitted via multinomial regression as performed by the \pkg{VGAM} package.
 This package was selected for its ability to fit multinomial regression models with an offset.
@@
 $$S(t) = \exp\left(-\int_0^t h(u;X) du\right).$$
-The integral is computed using either the \code{stats::integrate} function or Monte-Carlo integration.
-For the case of a competing-event analysis, the event-specific risk is computed using a nested double integral; in this setting, Monte-Carlo integration is faster and more flexible than \code{stats::integrate}.
-This is due to the fact that the computation involves a double loop over the selected time points; Monte-Carlo integration performs this step by using the vectorized \code{rowSums} and \code{colSums} functions.
+The integral is computed using either the \code{stats::integrate} function or Monte-Carlo integration. The risk function (or cumulative incidence function) is then defined as
 
-To decide between a single-event and a competing-event analysis, we created \code{absoluteRisk} as an \code{S3} generic, with methods for both \code{glm} and \code{CompRisk} objects (the latter inherits from \code{vglm} as well).
-The method dispatch system of \proglang{R} then takes care of matching the correct input to the correct methodology, without the user's intervention.
+$$ CI(t) = 1 - S(t).$$
+
+For the case of a competing-event analysis, the event-specific risk is computed using the following procedure: first, we compute the overall survival function (i.e. for all event types):
+
+$$ S(t) = \exp\left(-\int_0^t H(u;X) du\right),\qquad H(t;X) = \sum_{j=1}^J h_j(t;X).$$
+From this, we can derive the event-specific subdensities:
+
+$$ f_j(t) = h_j(t)S(t).$$
+
+Finally, by integrating these subdensities, we obtain the event-specific cumulative incidence functions:
+
+$$ CI_j(t) = \int_0^t f_j(u)du.$$
+
+We created \code{absoluteRisk} as an \code{S3} generic, with methods for the different types of outputs of \code{fitSmoothHazard}.
+The method dispatch system of \proglang{R} then takes care of matching the correct output to the correct methodology for calculating the cumulative incidence function, without the user's intervention.
+
+In the following sections, we illustrate these functionalities in the context of three case studies.
 
@@
 
-$$ CI(x, t) = 1 - exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$
+$$ CI(x, t) = 1 - \exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$
 
@@
 ```{r erspc-casebase-splines, eval = eval_cs1}
-casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ splines::bs(Follow.Up.Time) + ScrArm, 
+casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
                                     data = ERSPC, 
@@
 
-# Case study 3
+# Case study 3--The Cancer Genome Atlas
 
-  - Glmnet, gam, and gbm
+  - Glmnet and TCGA data
 

### 2019-07-25 analysis/paper/paper.tex
review theory and implemntation sections

@@
   University of Manitoba\\
-  186 Dysart Road Winnipeg, MB, Canada R3B 0S8\\
+  318 Machray Hall Winnipeg, MB, Canada R3T 2N2\\
   E-mail: \email{max.turgeon@umanitoba.ca}\\
-  URL: \url{https://maxturgeon.ca/}\\~\\
+  URL: \url{http://maxturgeon.ca/}\\~\\
       Jesse Islam\\
@@
 \hypertarget{code-formatting}{%
-\section{Code formatting}\label{code-formatting}}
+\section*{Code formatting}\label{code-formatting}}
+\addcontentsline{toc}{section}{Code formatting}
 
@@
 
-Popular survival analysis methods, like Kaplan-Meier and Cox regression,
-rely on the notion of risk-set sampling. Case-base sampling can be seen
-as an alternative.
-
 As shown by Saarela \& Arjas \citeyearpar{saarela2015non} (and further
@@
 expression where each person-moment's contribution is of the form
-\[\frac{\lambda(t)^{dN(t)}}{\rho(t) + \lambda(t)},\] where \(N(t)\) is
-the counting process associated with the event of interest,
-\(\lambda(t)\) is the corresponding hazard function, and \(\rho(t)\) is
-the hazard function for the Poisson process associated with case-base
-sampling. This parametric form suggests that we can readily estimate
-log-hazards of the form \(\log(\lambda(t)) = g(t; X)\) using logistic
-regression, where each observation corresponds to a person moment, the
-function \(g(t; X)\) is linear in a finite number of parameters, and
-where we treat \(\log(\rho(t))\) as an offset.
+\[\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},\] where \(N(t)\) is the counting
+process associated with the event of interest, \(h(t)\) is the
+corresponding hazard function, and \(\rho(t)\) is the hazard function
+for the Poisson process associated with case-base sampling. This
+parametric form suggests that we can readily estimate log-hazards of the
+form \(\log(h(t)) = g(t; X)\) using logistic regression, where each
+observation corresponds to a person moment, the function \(g(t; X)\) is
+linear in a finite number of parameters, and where we treat
+\(-\log(\rho(t))\) as an offset.
 
@@
 the base series uniformly from the study base. In terms of Poisson
-processes, this sampling strategy corresponds to a time-homogeneous
-Poisson process with intensity equal to \(b/B\), where \(b\) is the
-number of sampled observations in the base series, and \(B\) is the
-total population-time for the study base. More complex examples are also
-available; see for example Saarela \& Arjas
+processes, this sampling strategy corresponds essentially to a
+time-homogeneous Poisson process with intensity equal to \(b/B\), where
+\(b\) is the number of sampled observations in the base series, and
+\(B\) is the total population-time for the study base. More complex
+examples are also available; see for example Saarela \& Arjas
 \citeyearpar{saarela2015non}, where the probabilities of the sampling
@@
 
-\[ h(x,t) = exp[g(x,t)] \] where \(t\) denotes the numerical value
-(number of units) of a point in prognostic/prospective time and \(x\) is
-the realization of the vector \(X\) of variates based on the patient's
-profile and intervention (if any). Different functions of \(t\) lead to
-different parametric hazard models.
+\[ h(t;X) = \exp[g(t;X)] \] where \(t\) denotes time and \(X\), the
+individual's covariate profile. Different functions of \(t\) lead to
+different parametric hazard models. The simplest of these models is the
+one-parameter exponential distribution which is obtained by taking the
+hazard function to be constant over the range of \(t\).
 
-The simplest of these models is the one-parameter exponential
-distribution which is obtained by taking the hazard function to be
-constant over the range of \(t\).
-
-\[ h(x,t) = exp(\beta_0 + \beta_1 x) \]
+\[ h(t;X) = \exp(\beta_0 + \beta_1 X) \]
 
@@
 conditional chance of failure in a time interval of specified length is
-the same regardless of how long the individual has been on study a.k.a
-the memoryless property (Kalbfleisch and Prentice, 2002).
+the same regardless of how long the individual has been in the study.
+This is also known as the \emph{memoryless property} (Kalbfleisch and
+Prentice, 2002).
 
@@
 
-\[ h(x,t)  = exp(\beta_0 + \beta_1 t + \beta_2 x) \]
+\[ h(t;X)  = \exp(\beta_0 + \beta_1 t + \beta_2 X) \]
 
-Use of \(log(t)\) yields the Weibull hazard which allows for a power
+Use of \(\log(t)\) yields the Weibull hazard which allows for a power
 dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
 
-\[ h(x, t)  = exp(\beta_0 + \beta_1 \log(t) + \beta_2 x) \]
+\[ h(t;X)  = \exp(\beta_0 + \beta_1 \log(t) + \beta_2 X) \]
+
+For competing-risk analyses with \(J\) possible events, we can show that
+each person-moment's contribution of the likelihood is of the form
+
+\[\frac{h_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^Jh_j(t)},\]
+
+where \(N_j(t)\) is the counting process associated with the event of
+type \(j\) and \(h_j(t)\) is the corresponding hazard function. As may
+be expected, this functional form is similar to the terms appearing in
+the likelihood function for multinomial
+regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1\end{align*}}
 
@@
 categories: 1) data visualization, in the form of population-time plots;
-and 2) data analysis.
-
-We explicitely aimed at being compatible with both \code{data.frame}s
-and \code{data.table}s. This is evident in some of the coding choices we
-made, and it is also reflected in our testing units.
+and 2) parametric modeling. We explicitely aimed at being compatible
+with both \code{data.frame}s and \code{data.table}s. This is evident in
+some of the coding choices we made, and it is also reflected in our unit
+tests.
 
@@
 
-\hypertarget{data-analysis}{%
-\subsection{Data analysis}\label{data-analysis}}
+\hypertarget{parametric-modeling}{%
+\subsection{Parametric modeling}\label{parametric-modeling}}
 
-The data analysis step was separated into three parts: 1) case-base
-sampling; 2) estimation of the smooth hazard function; and 3)
-calculation of the risk function. By separating the sampling and
-estimation functions, we allowed the possibility of users implementing
-more complex sampling scheme, as described in Saarela
-\citeyearpar{saarela2016case}.
+The parametric modeling step was separated into three parts:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  case-base sampling;
+\item
+  estimation of the smooth hazard function;
+\item
+  calculation of the risk function.
+\end{enumerate}
+
+By separating the sampling and estimation functions, we allowed the
+possibility of users implementing more complex sampling scheme, as
+described in Saarela \citeyearpar{saarela2016case}.
 
@@
 
-Finally, the hazard function is fitted to the data using the function
-\code{glm}, unless there is more than one type event (i.e.~in a
-competing-risk analysis), in which case it uses the multinomical
-regression capabilities of the \pkg{VGAM} package. This package was
+To fit the hazard function, we provide several approaches that are
+available via the \code{family} parameter. These approaches are:
+
+\begin{itemize}
+\tightlist
+\item
+  \code{glm}: This is the familiar logistic regression.
+\item
+  \code{glmnet}: This option allows for variable selection using Lasso
+  or elastic-net. This functionality is provided through the
+  \code{glmnet} package \citep{friedman2010jss}.
+\item
+  \code{gam}: This option provides support for \emph{Generalized
+  Additive Models} via the \code{gam} package
+  \citep{hastie1987generalized}.
+\item
+  \code{gbm}: This option provides support for \emph{Gradient Boosted
+  Trees} via the \code{gbm} package. This feature is still experimental.
+\end{itemize}
+
+In the case of multiple events, the hazard is fitted via multinomial
+regression as performed by the \pkg{VGAM} package. This package was
 selected for its ability to fit multinomial regression models with an
@@
 is computed using either the \code{stats::integrate} function or
-Monte-Carlo integration. For the case of a competing-event analysis, the
-event-specific risk is computed using a nested double integral; in this
-setting, Monte-Carlo integration is faster and more flexible than
-\code{stats::integrate}. This is due to the fact that the computation
-involves a double loop over the selected time points; Monte-Carlo
-integration performs this step by using the vectorized \code{rowSums}
-and \code{colSums} functions.
+Monte-Carlo integration. The risk function (or cumulative incidence
+function) is then defined as
 
-To decide between a single-event and a competing-event analysis, we
-created \code{absoluteRisk} as an \code{S3} generic, with methods for
-both \code{glm} and \code{CompRisk} objects (the latter inherits from
-\code{vglm} as well). The method dispatch system of \proglang{R} then
-takes care of matching the correct input to the correct methodology,
-without the user's intervention.
+\[ CI(t) = 1 - S(t).\]
+
+For the case of a competing-event analysis, the event-specific risk is
+computed using the following procedure: first, we compute the overall
+survival function (i.e.~for all event types):
+
+\[ S(t) = \exp\left(-\int_0^t H(u;X) du\right),\qquad H(t;X) = \sum_{j=1}^J h_j(t;X).\]
+From this, we can derive the event-specific subdensities:
+
+\[ f_j(t) = h_j(t)S(t).\]
+
+Finally, by integrating these subdensities, we obtain the event-specific
+cumulative incidence functions:
+
+\[ CI_j(t) = \int_0^t f_j(u)du.\]
+
+We created \code{absoluteRisk} as an \code{S3} generic, with methods for
+the different types of outputs of \code{fitSmoothHazard}. The method
+dispatch system of \proglang{R} then takes care of matching the correct
+output to the correct methodology for calculating the cumulative
+incidence function, without the user's intervention.
+
+In the following sections, we illustrate these functionalities in the
+context of three case studies.
 
@@
 
-\[ CI(x, t) = 1 - exp\left( - \int_0^t h(x, u) \textrm{d}u \right) \]
+\[ CI(x, t) = 1 - \exp\left( - \int_0^t h(x, u) \textrm{d}u \right) \]
 
@@
 \begin{CodeInput}
-R> library(splines)
 R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
@@
 
-\hypertarget{case-study-3}{%
-\section{Case study 3}\label{case-study-3}}
+\hypertarget{case-study-3the-cancer-genome-atlas}{%
+\section{Case study 3--The Cancer Genome
+Atlas}\label{case-study-3the-cancer-genome-atlas}}
 
@@
 \item
-  Glmnet, gam, and gbm
+  Glmnet and TCGA data
 \end{itemize}
@@
 
-This report was generated on 2019-07-22 16:50:40 using the following
+This report was generated on 2019-07-25 16:51:47 using the following
 computational environment and dependencies:
@@
 #>  ui       X11                         
-#>  language (EN)                        
+#>  language en_CA:en                    
 #>  collate  en_CA.UTF-8                 
@@
 #>  tz       America/Winnipeg            
-#>  date     2019-07-22                  
+#>  date     2019-07-25                  
 #> 
@@
 #>  package     * version    date       lib source        
-#>  assertthat    0.2.0      2017-04-11 [1] CRAN (R 3.5.0)                  
-#>  backports     1.1.2      2017-12-13 [1] CRAN (R 3.5.0)                  
-#>  broom         0.4.4      2018-03-29 [1] CRAN (R 3.5.0)                  
-#>  callr         3.3.0      2019-07-04 [1] CRAN (R 3.6.1)                  
-#>  casebase    * 0.1.0.9000 2019-07-19 [1] local                           
-#>  cellranger    1.1.0      2016-07-27 [1] CRAN (R 3.5.0)                  
+#>  assertthat    0.2.1      2019-03-21 [1] CRAN (R 3.6.1)
+#>  backports     1.1.4      2019-04-10 [1] CRAN (R 3.6.1)
+#>  broom         0.5.2      2019-04-07 [1] CRAN (R 3.6.1)
+#>  callr         3.3.1      2019-07-18 [1] CRAN (R 3.6.1)
+#>  casebase    * 0.2.1.9000 2019-07-25 [1] local         
+#>  cellranger    1.1.0      2016-07-27 [1] CRAN (R 3.6.1)
 #>  cli           1.1.0      2019-03-19 [1] CRAN (R 3.6.1)
-#>  colorspace    1.3-2      2016-12-14 [1] CRAN (R 3.5.0)                  
-#>  crayon        1.3.4      2017-09-16 [1] CRAN (R 3.5.0)                  
-#>  data.table    1.11.2     2018-05-08 [1] CRAN (R 3.5.0)                  
-#>  desc          1.2.0      2018-05-01 [1] CRAN (R 3.5.0)                  
+#>  colorspace    1.4-1      2019-03-18 [1] CRAN (R 3.6.1)
+#>  crayon        1.3.4      2017-09-16 [1] CRAN (R 3.6.1)
+#>  data.table    1.12.2     2019-04-07 [1] CRAN (R 3.6.1)
+#>  desc          1.2.0      2018-05-01 [1] CRAN (R 3.6.1)
 #>  devtools      2.1.0      2019-07-06 [1] CRAN (R 3.6.1)
 #>  digest        0.6.20     2019-07-04 [1] CRAN (R 3.6.1)
-#>  dplyr       * 0.8.0.1    2019-02-15 [1] CRAN (R 3.5.2)                  
+#>  dplyr       * 0.8.3      2019-07-04 [1] CRAN (R 3.6.1)
 #>  evaluate      0.14       2019-05-28 [1] CRAN (R 3.6.1)
-#>  forcats     * 0.3.0      2018-02-19 [1] CRAN (R 3.5.0)                  
-#>  foreign       0.8-70     2018-04-23 [1] CRAN (R 3.5.0)                  
+#>  forcats     * 0.4.0      2019-02-17 [1] CRAN (R 3.6.1)
 #>  fs            1.3.1      2019-05-06 [1] CRAN (R 3.6.1)
+#>  generics      0.0.2      2018-11-29 [1] CRAN (R 3.6.1)
 #>  ggplot2     * 3.2.0      2019-06-16 [1] CRAN (R 3.6.1)
 #>  glue          1.3.1      2019-03-12 [1] CRAN (R 3.6.1)
-#>  gtable        0.2.0      2016-02-26 [1] CRAN (R 3.5.0)                  
-#>  haven         1.1.1      2018-01-18 [1] CRAN (R 3.5.0)                  
-#>  hms           0.4.2      2018-03-10 [1] CRAN (R 3.5.0)                  
-#>  htmltools     0.3.6      2017-04-28 [1] CRAN (R 3.5.0)                  
-#>  httr          1.3.1      2017-08-20 [1] CRAN (R 3.5.0)                  
+#>  gtable        0.3.0      2019-03-25 [1] CRAN (R 3.6.1)
+#>  haven         2.1.1      2019-07-04 [1] CRAN (R 3.6.1)
+#>  hms           0.5.0      2019-07-09 [1] CRAN (R 3.6.1)
+#>  htmltools     0.3.6      2017-04-28 [1] CRAN (R 3.6.1)
+#>  httr          1.4.0      2018-12-11 [1] CRAN (R 3.6.1)
 #>  jsonlite      1.6        2018-12-07 [1] CRAN (R 3.6.1)
@@
 #>  lattice       0.20-38    2018-11-04 [4] CRAN (R 3.5.1)
-#>  lazyeval      0.2.1      2017-10-29 [1] CRAN (R 3.5.0)                  
-#>  lubridate     1.7.4      2018-04-11 [1] CRAN (R 3.5.0)                  
-#>  magrittr    * 1.5        2014-11-22 [1] CRAN (R 3.5.0)                  
+#>  lazyeval      0.2.2      2019-03-15 [1] CRAN (R 3.6.1)
+#>  lubridate     1.7.4      2018-04-11 [1] CRAN (R 3.6.1)
+#>  magrittr    * 1.5        2014-11-22 [1] CRAN (R 3.6.1)
 #>  Matrix        1.2-17     2019-03-22 [4] CRAN (R 3.5.3)
-#>  memoise       1.1.0      2017-04-21 [1] CRAN (R 3.5.0)                  
-#>  mgcv          1.8-24     2018-06-18 [1] CRAN (R 3.5.1)                  
-#>  mnormt        1.5-5      2016-10-15 [1] CRAN (R 3.5.0)                  
-#>  modelr        0.1.2      2018-05-11 [1] CRAN (R 3.5.0)                  
-#>  munsell       0.4.3      2016-02-13 [1] CRAN (R 3.5.0)                  
+#>  memoise       1.1.0      2017-04-21 [1] CRAN (R 3.6.1)
+#>  mgcv          1.8-28     2019-03-21 [4] CRAN (R 3.5.3)
+#>  modelr        0.1.4      2019-02-18 [1] CRAN (R 3.6.1)
+#>  munsell       0.5.0      2018-06-12 [1] CRAN (R 3.6.1)
 #>  nlme          3.1-140    2019-05-12 [4] CRAN (R 3.6.0)
-#>  pillar        1.3.1      2018-12-15 [1] CRAN (R 3.5.2)                  
+#>  pillar        1.4.2      2019-06-29 [1] CRAN (R 3.6.1)
 #>  pkgbuild      1.0.3      2019-03-20 [1] CRAN (R 3.6.1)
-#>  pkgconfig     2.0.2      2018-08-16 [1] CRAN (R 3.5.2)                  
+#>  pkgconfig     2.0.2      2018-08-16 [1] CRAN (R 3.6.1)
 #>  pkgload       1.0.2      2018-10-29 [1] CRAN (R 3.6.1)
-#>  plyr          1.8.4      2016-06-08 [1] CRAN (R 3.5.0)                  
-#>  prettyunits   1.0.2      2015-07-13 [1] CRAN (R 3.5.2)                  
-#>  processx      3.4.0      2019-07-03 [1] CRAN (R 3.6.1)                  
+#>  prettyunits   1.0.2      2015-07-13 [1] CRAN (R 3.6.1)
+#>  processx      3.4.1      2019-07-18 [1] CRAN (R 3.6.1)
 #>  ps            1.3.0      2018-12-21 [1] CRAN (R 3.6.1)
-#>  psych         1.8.4      2018-05-06 [1] CRAN (R 3.5.0)                  
-#>  purrr       * 0.2.4      2017-10-18 [1] CRAN (R 3.5.0)                  
-#>  R6            2.2.2      2017-06-17 [1] CRAN (R 3.5.0)                  
+#>  purrr       * 0.3.2      2019-03-15 [1] CRAN (R 3.6.1)
+#>  R6            2.4.0      2019-02-14 [1] CRAN (R 3.6.1)
 #>  Rcpp          1.0.1      2019-03-17 [1] CRAN (R 3.6.1)
-#>  readr       * 1.1.1      2017-05-16 [1] CRAN (R 3.5.0)                  
-#>  readxl        1.1.0      2018-04-20 [1] CRAN (R 3.5.0)                  
+#>  readr       * 1.3.1      2018-12-21 [1] CRAN (R 3.6.1)
+#>  readxl        1.3.1      2019-03-13 [1] CRAN (R 3.6.1)
 #>  remotes       2.1.0      2019-06-24 [1] CRAN (R 3.6.1)
-#>  reshape2      1.4.3      2017-12-11 [1] CRAN (R 3.5.0)                  
 #>  rlang         0.4.0      2019-06-25 [1] CRAN (R 3.6.1)
 #>  rmarkdown     1.14       2019-07-12 [1] CRAN (R 3.6.1)
-#>  rprojroot     1.3-2      2018-01-03 [1] CRAN (R 3.5.0)                  
-#>  rstudioapi    0.7        2017-09-07 [1] CRAN (R 3.5.0)                  
-#>  rticles       0.9.1      2019-07-22 [1] Github (rstudio/rticles@8d56fc6)
-#>  rvest         0.3.2      2016-06-17 [1] CRAN (R 3.5.0)                  
-#>  scales        0.5.0      2017-08-24 [1] CRAN (R 3.5.0)                  
+#>  rprojroot     1.3-2      2018-01-03 [1] CRAN (R 3.6.1)
+#>  rstudioapi    0.10       2019-03-19 [1] CRAN (R 3.6.1)
+#>  rticles       0.9        2019-07-12 [1] CRAN (R 3.6.1)
+#>  rvest         0.3.4      2019-05-15 [1] CRAN (R 3.6.1)
+#>  scales        1.0.0      2018-08-09 [1] CRAN (R 3.6.1)
 #>  sessioninfo   1.1.1      2018-11-05 [1] CRAN (R 3.6.1)
@@
 #>  survival    * 2.43-3     2018-11-26 [4] CRAN (R 3.5.1)
-#>  testthat      2.1.1      2019-04-23 [1] CRAN (R 3.6.1)                  
-#>  tibble      * 2.0.1      2019-01-12 [1] CRAN (R 3.5.2)                  
-#>  tidyr       * 0.8.1      2018-05-18 [1] CRAN (R 3.5.0)                  
-#>  tidyselect    0.2.5      2018-10-11 [1] CRAN (R 3.5.2)                  
-#>  tidyverse   * 1.2.1      2017-11-14 [1] CRAN (R 3.5.0)                  
+#>  testthat      2.2.0      2019-07-22 [1] CRAN (R 3.6.1)
+#>  tibble      * 2.1.3      2019-06-06 [1] CRAN (R 3.6.1)
+#>  tidyr       * 0.8.3      2019-03-01 [1] CRAN (R 3.6.1)
+#>  tidyselect    0.2.5      2018-10-11 [1] CRAN (R 3.6.1)
+#>  tidyverse   * 1.2.1      2017-11-14 [1] CRAN (R 3.6.1)
 #>  usethis       1.5.1      2019-07-04 [1] CRAN (R 3.6.1)
+#>  vctrs         0.2.0      2019-07-05 [1] CRAN (R 3.6.1)
 #>  VGAM          1.1-1      2019-02-18 [1] CRAN (R 3.6.1)
-#>  withr         2.1.2      2018-03-15 [1] CRAN (R 3.5.0)                  
+#>  withr         2.1.2      2018-03-15 [1] CRAN (R 3.6.1)
 #>  xfun          0.8        2019-06-25 [1] CRAN (R 3.6.1)
-#>  xml2          1.2.0      2018-01-24 [1] CRAN (R 3.5.0)                  
+#>  xml2          1.2.0      2018-01-24 [1] CRAN (R 3.6.1)
 #>  yaml          2.2.0      2018-07-25 [1] CRAN (R 3.6.1)
+#>  zeallot       0.1.0      2018-01-28 [1] CRAN (R 3.6.1)
 #> 
-#> [1] /home/mturgeon/Rlibs
+#> [1] /home/turgeonm/R/x86_64-pc-linux-gnu-library/3.6
 #> [2] /usr/local/lib/R/site-library
@@
 \begin{CodeOutput}
-#> Local:    review-max /home/mturgeon/Documents/git_repositories/cbpaper
-#> Head:     [c8ac771] 2019-07-22: Merge pull request #5 from sahirbhatnagar/litreview
+#> Local:    review-max /home/turgeonm/Documents/git_repos/cbpaper
+#> Head:     [c0cb26d] 2019-07-23: Merge pull request #6 from sahirbhatnagar/fixdocker
 \end{CodeOutput}

### 2019-07-29 analysis/paper/paper.Rmd
Added table of existing work for smooth hazard and variable selection

--- 2019-07-22 analysis/paper/paper.Rmd
@@
   \usepackage{longtable}
+  \usepackage{booktabs}
 output: 
@@
 
-\begin{table}
-  \begin{tabular}{p{1.8in}lll}
+\begin{longtable}{lp{1in}p{2in}p{1.1in}}
+%  \begin{tabular}{lp{1in}p{2in}p{1.1in}}
 \hline
-    &  Parameters &  Density \proglang{R} function & \code{dist}\\
-    & {\footnotesize{(location in \code{\color{red}{red}})}} & & \\
+    Package &    documentation  & Type & Function Call\\
+    % & {\footnotesize{(location in \code{\color{red}{red}})}} & & \\
 \hline
-    Exponential & \code{\color{red}{rate}}             & \code{dexp}   & \code{"exp"} \\
-    Weibull (accelerated failure time)     & \code{shape, {\color{red}{scale}}}     & \code{dweibull} & \code{"weibull"} \\
-    Weibull (proportional hazards)     & \code{shape, {\color{red}{scale}}}     & \code{dweibullPH} & \code{"weibullPH"} \\
-    Gamma       & \code{shape, \color{red}{rate}}      & \code{dgamma} & \code{"gamma"}\\
-    Log-normal  & \code{{\color{red}{meanlog}}, sdlog}   & \code{dlnorm} & \code{"lnorm"}\\
-    Gompertz    & \code{shape, {\color{red}{rate}}}      & \code{dgompertz} & \code{"gompertz"} \\
-    Log-logistic & \code{shape, {\color{red}{scale}}}   & \code{dllogis} & \code{"llogis"}\\
-    Generalized gamma (Prentice 1975)   & \code{{\color{red}{mu}}, sigma, Q} & \code{dgengamma} & \code{"gengamma"} \\
-    Generalized gamma (Stacy 1962)& \code{shape, {\color{red}{scale}}, k} & \code{dgengamma.orig} & \code{"gengamma.orig"} \\
-    Generalized F     (stable)    & \code{{\color{red}{mu}}, sigma, Q, P} & \code{dgenf} & \code{"genf"} \\
-    Generalized F     (original)  & \code{{\color{red}{mu}}, sigma, s1, s2} & \code{dgenf.orig} & \code{"genf.orig"} \\
+    \href{https://cran.r-project.org/package=flexsurv}{flexsurv}  & \href{https://cran.r-project.org/web/packages/flexsurv/vignettes/flexsurv.pdf}{Vignette}, \href{https://www.jstatsoft.org/article/view/v070i08}{Jackson, C.
+JSS 2016}  & Fully-parametric.
+Also includes the spline model of Royston and Parmar (2002), in which both baseline survival and covariate effects can be arbitrarily flexible parametric functions of time.
+&   \texttt{flexsurvreg(Surv(recyrs, censrec) ~ group, data = bc, dist = "gengamma")}   \texttt{flexsurvspline(Surv(recyrs, censrec) ~ group, data = bc, k = 1, scale = "odds")}  \\
+    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% \\
+    \href{https://cran.r-project.org/package=AdapEnetClass}{AdapEnetClass}  & \href{https://link.springer.com/article/10.1007/s11222-015-9555-8}{Khan et al.
+Statistics and Computing 2016}  & Variable selection in AFT models using elastic net.
+No absolute risk estimates.
+&   \texttt{cv.AWEnet(x, y, delta, weight=wt, lambda2=0.001, maxit=10, plot.it = T, AEnet=T)}  \\
+    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% \\
+    \href{https://cran.r-project.org/package=eha}{eha}  & \href{https://cran.r-project.org/web/packages/eha/vignettes/parametric.pdf}{Vignette}  & AFT model.
+Automatic produces plot of estimated smooth hazard function.
+&   \texttt{phreg(Surv(enter + 40, exit + 40, event) ~ I(birthdate - 1810) + strata(ses), data = mort, dist = "gompertz")}  \\
+        %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% \\
+    \href{https://cran.r-project.org/package=smoothSurv}{smoothSurv}  & \href{https://cran.r-project.org/web/packages/smoothSurv/smoothSurv.pdf}{R manual}  & An AFT model with an error distribution assumed to be a mixture of G-splines    &   \texttt{smoothSurvReg(surv ~ x1 * x2, logscale = ~1, info = FALSE, lambda = exp(2:(-1)))}  \\
 \hline
-  \end{tabular}
+%  \end{tabular}
   \caption{Built-in parametric survival distributions in \pkg{flexsurv}.}
   \label{tab:dists}
-\end{table}
+\end{longtable}
 

### 2019-07-29 analysis/paper/paper.tex
Added table of existing work for smooth hazard and variable selection

--- 2019-07-22 analysis/paper/paper.tex
@@
       Maxime Turgeon *\\
+  University of Manitoba\\
+  318 Machray Hall Winnipeg, MB, Canada R3T 2N2\\
+  E-mail: \email{max.turgeon@umanitoba.ca}\\
+  URL: \url{http://maxturgeon.ca/}\\~\\
+      Jesse Islam\\
   McGill University\\
   1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
-  E-mail: \email{maxime.turgeon@mail.mcgill.ca}\\
-  URL: \url{http://maxturgeon.ca/}\\~\\
+  E-mail: \email{jesse.islam@mail.mcgill.ca}\\
+
       James Hanley\\
@@
 
-\usepackage{amsmath} \usepackage{longtable}
+\usepackage{amsmath} \usepackage{longtable} \usepackage{booktabs}
 
@@
 
-This report was generated on 2019-07-22 17:17:57 using the following
+This report was generated on 2019-07-29 10:55:18 using the following
 computational environment and dependencies:
@@
 #>  setting  value                       
-#>  version  R version 3.6.0 (2019-04-26)
+#>  version  R version 3.6.1 (2019-07-05)
 #>  os       Pop!_OS 18.10               
@@
 #>  tz       America/Toronto             
-#>  date     2019-07-22                  
+#>  date     2019-07-29                  
 #> 
@@
 \begin{CodeOutput}
-#> Local:    litreview /home/sahir/git_repositories/cbpaper
-#> Head:     [38153a4] 2019-07-22: Update README.md
+#> Local:    litreview-sahir /home/sahir/git_repositories/cbpaper
+#> Remote:   litreview-sahir @ origin (https://github.com/sahirbhatnagar/cbpaper)
+#> Head:     [c0cb26d] 2019-07-23: Merge pull request #6 from sahirbhatnagar/fixdocker
 \end{CodeOutput}

### 2019-08-02 analysis/paper/paper.Rmd
fix overlapping author names

--- 2019-07-25 analysis/paper/paper.Rmd
@@
   - name: Maxime Turgeon *
-    affiliation: University of Manitoba
+    affiliation: 'University of Manitoba \AND'
     address: >
-      318 Machray Hall
+      186 Dysart Road
       Winnipeg, MB, Canada R3T 2N2
@@
 eval_introduction <- TRUE
-eval_theory <- FALSE
+eval_theory <- TRUE
 eval_implementation <- TRUE
-eval_cs1 <- FALSE
-eval_cs2 <- FALSE
+eval_cs1 <- TRUE
+eval_cs2 <- TRUE
 eval_cs3 <- FALSE
@@
 
-The purpose of the `casebase` package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence (CI)) of an event, for a particular patient.
+The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence (CI)) of an event, for a particular patient.
 The following points should be noted:
@@
 
-```{r compPop1, fig.cap="\\label{fig:compPop1}Population-time plot for the stem-cell transplant study. The points represent the event of interest (i.e., relapse).", echo=FALSE, eval = eval_cs2}
+```{r compPop1, fig.cap="Population-time plot for the stem-cell transplant study. The points represent the event of interest (i.e., relapse).", echo=FALSE, eval = eval_cs2}
 pt_object <- casebase::popTime(bmtcrr, event = "Status", time = "ftime")
@@
 
-```{r compPop2, fig.cap="\\label{fig:compPop2}Population-time plot for the stem-cell transplant study. The points represent the competing events.", echo=FALSE, eval = eval_cs2}
+```{r compPop2, fig.cap="Population-time plot for the stem-cell transplant study. The points represent the competing events.", echo=FALSE, eval = eval_cs2}
 pt_object_comp <- bmtcrr %>% 
@@
 
+In this article, we presented the \proglang{R} package \pkg{casebase} that provides functions to fit smooth parametric hazards and estimate cumulative incidence functions using case-base sampling.
+We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through three case studies.
+
 In the following table we provide a comparison between the Cox model and case-base sampling:

### 2019-08-02 analysis/paper/paper.tex
fix overlapping author names

--- 2019-07-25 analysis/paper/paper.tex
@@
 \author{
-Sahir Bhatnagar *\\McGill Univeristy \And Maxime Turgeon *\\University of Manitoba \And Jesse Islam\\McGill University \And James Hanley\\McGill Univeristy \And Olli Saarela\\University of Toronto
+Sahir Bhatnagar *\\McGill Univeristy \And Maxime Turgeon *\\University of Manitoba \AND Jesse Islam\\McGill University \And James Hanley\\McGill Univeristy \And Olli Saarela\\University of Toronto
 }
@@
 R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
-R>                        levels = c(0,1), 
-R>                        labels = c("Control group", "Screening group"))
+R+                        levels = c(0,1), 
+R+                        labels = c("Control group", "Screening group"))
 \end{CodeInput}
@@
 \end{CodeInput}
+
+\begin{center}\includegraphics{../figures/plot-erspc-data-1} \end{center}
+
 \end{CodeChunk}
@@
 R> pt_object_strat <- casebase::popTime(ERSPC, 
-R>                                      event = "DeadOfPrCa", 
-R>                                      exposure = "ScrArm")
+R+                                      event = "DeadOfPrCa", 
+R+                                      exposure = "ScrArm")
 R> plot(pt_object_strat)
 \end{CodeInput}
+
+\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-1} \end{center}
+
 \end{CodeChunk}
@@
 \end{CodeInput}
+
+\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-side-1} \end{center}
+
 \end{CodeChunk}
@@
 R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
-R>                              data = ERSPC)
+R+                              data = ERSPC)
 R> summary(cox_model)
 \end{CodeInput}
+
+\begin{CodeOutput}
+#> Call:
+#> survival::coxph(formula = Surv(Follow.Up.Time, DeadOfPrCa) ~ 
+#>     ScrArm, data = ERSPC)
+#> 
+#>   n= 159893, number of events= 540 
+#> 
+#>                         coef exp(coef) se(coef)     z Pr(>|z|)  
+#> ScrArmScreening group -0.222     0.801    0.088 -2.52    0.012 *
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#>                       exp(coef) exp(-coef) lower .95 upper .95
+#> ScrArmScreening group     0.801       1.25     0.674     0.952
+#> 
+#> Concordance= 0.519  (se = 0.011 )
+#> Rsquare= 0   (max possible= 0.075 )
+#> Likelihood ratio test= 6.45  on 1 df,   p=0.01
+#> Wald test            = 6.37  on 1 df,   p=0.01
+#> Score (logrank) test = 6.39  on 1 df,   p=0.01
+\end{CodeOutput}
 \end{CodeChunk}
@@
 R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
-R>                        ignore = 99)
+R+                        ignore = 99)
 R> 
 R> plot(survfit(cox_model, newdata = new_data),
-R>      xlab = "Years since Randomization", 
-R>      ylab = "Cumulative Incidence", 
-R>      fun = "event",
-R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>                     exp(coef(cox_model)), 
-R>                     exp(confint(cox_model))[1], 
-R>                     exp(confint(cox_model))[2]))
+R+      xlab = "Years since Randomization", 
+R+      ylab = "Cumulative Incidence", 
+R+      fun = "event",
+R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+                     exp(coef(cox_model)), 
+R+                     exp(confint(cox_model))[1], 
+R+                     exp(confint(cox_model))[2]))
 R> legend("topleft", 
-R>        legend = c("Control group", "Screening group"), 
-R>        col = c("red","blue"),
-R>        lty = c(1, 1), 
-R>        bg = "gray90")
+R+        legend = c("Control group", "Screening group"), 
+R+        col = c("red","blue"),
+R+        lty = c(1, 1), 
+R+        bg = "gray90")
 \end{CodeInput}
+
+\begin{center}\includegraphics{../figures/erspc-cox-cif-1} \end{center}
+
 \end{CodeChunk}
@@
 R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
-R>                                                   data = ERSPC, 
-R>                                                   ratio = 100)
+R+                                                   data = ERSPC, 
+R+                                                   ratio = 100)
 R> 
 R> summary(casebase_exponential)
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> 
+#> Call:
+#> glm(formula = formula, family = binomial, data = sampleData)
+#> 
+#> Deviance Residuals: 
+#>    Min      1Q  Median      3Q     Max  
+#> -0.148  -0.148  -0.148  -0.133   3.078  
+#> 
+#> Coefficients:
+#>                       Estimate Std. Error z value            Pr(>|z|)    
+#> (Intercept)            -7.7659     0.0557 -139.46 <0.0000000000000002 ***
+#> ScrArmScreening group  -0.2147     0.0884   -2.43               0.015 *  
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#> (Dispersion parameter for binomial family taken to be 1)
+#> 
+#>     Null deviance: 6059  on 54539  degrees of freedom
+#> Residual deviance: 6053  on 54538  degrees of freedom
+#> AIC: 6057
+#> 
+#> Number of Fisher Scoring iterations: 7
+\end{CodeOutput}
+
+\begin{CodeInput}
 R> exp(coef(casebase_exponential)[2])
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> ScrArmScreening group 
+#>                  0.81
+\end{CodeOutput}
+
+\begin{CodeInput}
 R> exp(confint(casebase_exponential)[2,])
 \end{CodeInput}
+
+\begin{CodeOutput}
+#>  2.5 % 97.5 % 
+#>   0.68   0.96
+\end{CodeOutput}
 \end{CodeChunk}
@@
 R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
-R>                                           time = seq(0,15,0.1), 
-R>                                           newdata = new_data)
+R+                                           time = seq(0,15,0.1), 
+R+                                           newdata = new_data)
 R> 
 R> plot(survfit(cox_model, newdata = new_data),
-R>      xlab = "Years since Randomization", 
-R>      ylab = "Cumulative Incidence", 
-R>      fun = "event",
-R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>                     exp(coef(cox_model)), 
-R>                     exp(confint(cox_model))[1], 
-R>                     exp(confint(cox_model))[2]))
+R+      xlab = "Years since Randomization", 
+R+      ylab = "Cumulative Incidence", 
+R+      fun = "event",
+R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+                     exp(coef(cox_model)), 
+R+                     exp(confint(cox_model))[1], 
+R+                     exp(confint(cox_model))[2]))
 R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
@@
 R> legend("topleft", 
-R>        legend = c("Control group (Cox)","Control group (Casebase)",
-R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R>        col = c("red","red", "blue","blue"),
-R>        lty = c(1, 2, 1, 2), 
-R>        bg = "gray90")
+R+        legend = c("Control group (Cox)","Control group (Casebase)",
+R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R+        col = c("red","red", "blue","blue"),
+R+        lty = c(1, 2, 1, 2), 
+R+        bg = "gray90")
 \end{CodeInput}
+
+\begin{center}\includegraphics{../figures/erspc-casebase-cif-1} \end{center}
+
 \end{CodeChunk}
@@
 R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
-R>                                  data = ERSPC, 
-R>                                  ratio = 100)
+R+                                  data = ERSPC, 
+R+                                  ratio = 100)
 R> 
 R> summary(casebase_time)
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> 
+#> Call:
+#> glm(formula = formula, family = binomial, data = sampleData)
+#> 
+#> Deviance Residuals: 
+#>    Min      1Q  Median      3Q     Max  
+#> -0.401  -0.162  -0.122  -0.094   3.466  
+#> 
+#> Coefficients:
+#>                       Estimate Std. Error z value            Pr(>|z|)    
+#> (Intercept)            -9.0314     0.1121  -80.53 <0.0000000000000002 ***
+#> Follow.Up.Time          0.2222     0.0145   15.29 <0.0000000000000002 ***
+#> ScrArmScreening group  -0.2320     0.0886   -2.62              0.0089 ** 
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#> (Dispersion parameter for binomial family taken to be 1)
+#> 
+#>     Null deviance: 6059  on 54539  degrees of freedom
+#> Residual deviance: 5811  on 54537  degrees of freedom
+#> AIC: 5817
+#> 
+#> Number of Fisher Scoring iterations: 8
+\end{CodeOutput}
+
+\begin{CodeInput}
 R> exp(coef(casebase_time))
+\end{CodeInput}
+
+\begin{CodeOutput}
+#>           (Intercept)        Follow.Up.Time ScrArmScreening group 
+#>               0.00012               1.24882               0.79297
+\end{CodeOutput}
+
+\begin{CodeInput}
 R> exp(confint(casebase_time))
 \end{CodeInput}
+
+\begin{CodeOutput}
+#>                          2.5 %  97.5 %
+#> (Intercept)           0.000096 0.00015
+#> Follow.Up.Time        1.213876 1.28507
+#> ScrArmScreening group 0.665769 0.94256
+\end{CodeOutput}
 \end{CodeChunk}
@@
 R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
-R>                                           time = seq(0,15,0.1), 
-R>                                           newdata = new_data)
+R+                                           time = seq(0,15,0.1), 
+R+                                           newdata = new_data)
 R> 
 R> plot(survfit(cox_model, newdata = new_data),
-R>      xlab = "Years since Randomization", 
-R>      ylab = "Cumulative Incidence", 
-R>      fun = "event",
-R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>                     exp(coef(cox_model)), 
-R>                     exp(confint(cox_model))[1], 
-R>                     exp(confint(cox_model))[2]))
+R+      xlab = "Years since Randomization", 
+R+      ylab = "Cumulative Incidence", 
+R+      fun = "event",
+R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+                     exp(coef(cox_model)), 
+R+                     exp(confint(cox_model))[1], 
+R+                     exp(confint(cox_model))[2]))
 R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
@@
 R> legend("topleft", 
-R>        legend = c("Control group (Cox)","Control group (Casebase)",
-R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R>        col = c("red","red", "blue","blue"),
-R>        lty = c(1, 2, 1, 2), 
-R>        bg = "gray90")
+R+        legend = c("Control group (Cox)","Control group (Casebase)",
+R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R+        col = c("red","red", "blue","blue"),
+R+        lty = c(1, 2, 1, 2), 
+R+        bg = "gray90")
 \end{CodeInput}
+
+\begin{center}\includegraphics{../figures/erspc-casebase-weibull-cif-1} \end{center}
+
 \end{CodeChunk}
@@
 R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
-R>                                     data = ERSPC, 
-R>                                     ratio = 100)
+R+                                     data = ERSPC, 
+R+                                     ratio = 100)
 R> 
 R> summary(casebase_splines)
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> 
+#> Call:
+#> glm(formula = formula, family = binomial, data = sampleData)
+#> 
+#> Deviance Residuals: 
+#>    Min      1Q  Median      3Q     Max  
+#> -0.405  -0.173  -0.136  -0.085   3.796  
+#> 
+#> Coefficients:
+#>                       Estimate Std. Error z value             Pr(>|z|)    
+#> (Intercept)           -10.2643     0.3159  -32.49 < 0.0000000000000002 ***
+#> bs(Follow.Up.Time)1     4.2983     0.8005    5.37        0.00000007893 ***
+#> bs(Follow.Up.Time)2     2.0945     0.4824    4.34        0.00001412028 ***
+#> bs(Follow.Up.Time)3     4.5544     0.7067    6.44        0.00000000012 ***
+#> ScrArmScreening group  -0.2169     0.0886   -2.45                0.014 *  
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#> (Dispersion parameter for binomial family taken to be 1)
+#> 
+#>     Null deviance: 6059.0  on 54539  degrees of freedom
+#> Residual deviance: 5790.5  on 54535  degrees of freedom
+#> AIC: 5800
+#> 
+#> Number of Fisher Scoring iterations: 8
+\end{CodeOutput}
+
+\begin{CodeInput}
 R> exp(coef(casebase_splines))
+\end{CodeInput}
+
+\begin{CodeOutput}
+#>           (Intercept)   bs(Follow.Up.Time)1   bs(Follow.Up.Time)2 
+#>              0.000035             73.578176              8.121029 
+#>   bs(Follow.Up.Time)3 ScrArmScreening group 
+#>             95.053534              0.805017
+\end{CodeOutput}
+
+\begin{CodeInput}
 R> exp(confint(casebase_splines))
 \end{CodeInput}
+
+\begin{CodeOutput}
+#>                           2.5 %     97.5 %
+#> (Intercept)            0.000018   0.000063
+#> bs(Follow.Up.Time)1   15.835008 367.061126
+#> bs(Follow.Up.Time)2    3.252368  21.652009
+#> bs(Follow.Up.Time)3   22.537699 364.141700
+#> ScrArmScreening group  0.675920   0.956823
+\end{CodeOutput}
 \end{CodeChunk}
@@
 R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
-R>                                     time = seq(0,15,0.1), 
-R>                                     newdata = new_data)
+R+                                     time = seq(0,15,0.1), 
+R+                                     newdata = new_data)
 R> 
 R> plot(survfit(cox_model, newdata = new_data),
-R>      xlab = "Years since Randomization", 
-R>      ylab = "Cumulative Incidence", 
-R>      fun = "event",
-R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>                     exp(coef(cox_model)), 
-R>                     exp(confint(cox_model))[1], 
-R>                     exp(confint(cox_model))[2]))
+R+      xlab = "Years since Randomization", 
+R+      ylab = "Cumulative Incidence", 
+R+      fun = "event",
+R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+                     exp(coef(cox_model)), 
+R+                     exp(confint(cox_model))[1], 
+R+                     exp(confint(cox_model))[2]))
 R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
@@
 R> legend("topleft", 
-R>        legend = c("Control group (Cox)","Control group (Casebase)",
-R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R>        col = c("red","red", "blue","blue"),
-R>        lty = c(1, 2, 1, 2), 
-R>        bg = "gray90")
+R+        legend = c("Control group (Cox)","Control group (Casebase)",
+R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R+        col = c("red","red", "blue","blue"),
+R+        lty = c(1, 2, 1, 2), 
+R+        bg = "gray90")
 \end{CodeInput}
+
+\begin{center}\includegraphics{../figures/erspc-casebase-splines-cif-1} \end{center}
+
 \end{CodeChunk}
@@
 \end{CodeInput}
+
+\begin{CodeOutput}
+#> Analysis of Deviance Table
+#> 
+#> Model 1: DeadOfPrCa ~ Follow.Up.Time + ScrArm + offset(offset)
+#> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
+#>   Resid. Df Resid. Dev Df Deviance Pr(>Chi)    
+#> 1     54537       5811                         
+#> 2     54535       5790  2     20.5 0.000035 ***
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{CodeOutput}
 \end{CodeChunk}
@@
 
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/compPop1-1}
+
+}
+
+\caption[Population-time plot for the stem-cell transplant study]{Population-time plot for the stem-cell transplant study.
+The points represent the event of interest (i.e., relapse).}\label{fig:compPop1}
+\end{figure}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/compPop2-1}
+
+}
+
+\caption[Population-time plot for the stem-cell transplant study]{Population-time plot for the stem-cell transplant study. The points represent the competing events.}\label{fig:compPop2}
+\end{figure}
+\end{CodeChunk}
+
 Our main objective is to compute the absolute risk of relapse for a
@@
 R> model_cb <- fitSmoothHazard(
-R>     Status ~ ftime + Sex + D + Phase + Source + Age, 
-R>     data = bmtcrr, 
-R>     ratio = 100, 
-R>     time = "ftime")
+R+     Status ~ ftime + Sex + D + Phase + Source + Age, 
+R+     data = bmtcrr, 
+R+     ratio = 100, 
+R+     time = "ftime")
 \end{CodeInput}
@@
 corresponding confidence intervals:
+
+\begin{CodeChunk}
+
+\begin{tabular}{l|r|l}
+\hline
+Covariates & HR & 95\% CI\\
+\hline
+Sex & 0.71 & (0.41, 1.24)\\
+\hline
+Disease & 0.55 & (0.31, 1)\\
+\hline
+Phase (CR2 vs. CR1) & 1.18 & (0.47, 2.93)\\
+\hline
+Phase (CR3 vs. CR1) & 1.47 & (0.38, 5.63)\\
+\hline
+Phase (Relapse vs. CR1) & 4.30 & (2, 9.27)\\
+\hline
+Source & 1.56 & (0.52, 4.7)\\
+\hline
+Age & 0.99 & (0.97, 1.02)\\
+\hline
+\end{tabular}
+
+\end{CodeChunk}
 
@@
 R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
-R>                                      levels = levels(bmtcrr[,"Sex"])),
-R>                       "D" = c("ALL", "AML"),
-R>                       "Phase" = factor(c("Relapse", "Relapse"), 
-R>                                        levels = levels(bmtcrr[,"Phase"])),
-R>                       "Age" = c(35, 35),
-R>                       "Source" = factor(c("PB", "PB"), 
-R>                                         levels = levels(bmtcrr[,"Source"])))
+R+                                      levels = levels(bmtcrr[,"Sex"])),
+R+                       "D" = c("ALL", "AML"),
+R+                       "Phase" = factor(c("Relapse", "Relapse"), 
+R+                                        levels = levels(bmtcrr[,"Phase"])),
+R+                       "Age" = c(35, 35),
+R+                       "Source" = factor(c("PB", "PB"), 
+R+                                         levels = levels(bmtcrr[,"Source"])))
 R> 
@@
 R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
-R>                         method = "numerical", newdata = newdata)
+R+                         method = "numerical", newdata = newdata)
 \end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/unnamed-chunk-3-1}
+
+}
+
+\caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups.
+The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:unnamed-chunk-3}
+\end{figure}
 \end{CodeChunk}
@@
 
-This report was generated on 2019-07-25 16:51:47 using the following
+This report was generated on 2019-08-02 15:11:12 using the following
 computational environment and dependencies:
@@
 #>  tz       America/Winnipeg            
-#>  date     2019-07-25                  
+#>  date     2019-08-02                  
 #> 
@@
 #>  knitr         1.23       2019-05-18 [1] CRAN (R 3.6.1)
+#>  labeling      0.3        2014-08-23 [1] CRAN (R 3.6.1)
 #>  lattice       0.20-38    2018-11-04 [4] CRAN (R 3.5.1)
@@
 #>  magrittr    * 1.5        2014-11-22 [1] CRAN (R 3.6.1)
+#>  MASS          7.3-51.1   2018-11-01 [4] CRAN (R 3.5.1)
 #>  Matrix        1.2-17     2019-03-22 [4] CRAN (R 3.5.3)
@@
 #>  pkgload       1.0.2      2018-10-29 [1] CRAN (R 3.6.1)
+#>  plyr          1.8.4      2016-06-08 [1] CRAN (R 3.6.1)
 #>  prettyunits   1.0.2      2015-07-13 [1] CRAN (R 3.6.1)
@@
 #>  remotes       2.1.0      2019-06-24 [1] CRAN (R 3.6.1)
+#>  reshape2      1.4.3      2017-12-11 [1] CRAN (R 3.6.1)
 #>  rlang         0.4.0      2019-06-25 [1] CRAN (R 3.6.1)
@@
 #> Local:    review-max /home/turgeonm/Documents/git_repos/cbpaper
-#> Head:     [c0cb26d] 2019-07-23: Merge pull request #6 from sahirbhatnagar/fixdocker
+#> Head:     [ed2a3bc] 2019-07-25: review theory and implemntation sections
 \end{CodeOutput}

### 2019-12-05 analysis/paper/paper.Rmd
add analysis of stanford data

@@
   - name: Sahir Bhatnagar *
-    affiliation: McGill Univeristy
+    affiliation: McGill University
     address: >
@@
   - name: James Hanley
-    affiliation: McGill Univeristy
+    affiliation: McGill University
     address: >
@@
 
-# Code formatting {.unnumbered}
+<!-- # Code formatting {.unnumbered} -->
 
-Don't use markdown, instead use the more precise latex commands:
+<!-- Don't use markdown, instead use the more precise latex commands: -->
 
-* \proglang{Java}
-* \pkg{plyr}
-* \code{print("abc")}
+<!-- * \proglang{Java} -->
+<!-- * \pkg{plyr} -->
+<!-- * \code{print("abc")} -->
 
@@
 eval_implementation <- TRUE
-eval_cs1 <- TRUE
-eval_cs2 <- TRUE
+eval_cs1 <- FALSE
+eval_cs2 <- FALSE
 eval_cs3 <- FALSE
+eval_cs4 <- TRUE
 ```
@@
 
-As shown by Saarela & Arjas [-@saarela2015non] (and further expanded in Saarela [-@saarela2016case]), writing the likelihood arising from this data-generating mechanism using the framework of non-homogenous Poisson processes, we eventually reach an expression where each person-moment's contribution is of the form
+As shown by Saarela & Arjas [-@saarela2015non] (and further expanded in Saarela [-@saarela2016case]), writing the likelihood arising from this data-generating mechanism using the framework of non-homogeneous Poisson processes, we eventually reach an expression where each person-moment's contribution is of the form
 $$\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},$$
@@
 
-In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base samping *uniformly*, i.e.
+In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base sampling *uniformly*, i.e.
 to sample the base series uniformly from the study base.
@@
 The functions in the casebase package can be divided into two categories: 1) data visualization, in the form of population-time plots; and 2) parametric modeling.
-We explicitely aimed at being compatible with both \code{data.frame}s and \code{data.table}s.
+We explicitly aimed at being compatible with both \code{data.frame}s and \code{data.table}s.
 This is evident in some of the coding choices we made, and it is also reflected in our unit tests.
@@
 
-As espected, we see that splines model provides a better fit.
+As expected, we see that splines model provides a better fit.
 
@@
 
+# Case study 4--Stanford Heart Transplant Data
+
+Although the previous case studies provide a broad overview of the capabilities of \pkg{casebase}, they all have two properties in common:
+\begin{itemize}
+  \item All covariates are fixed, i.e. they do not change over time;
+  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
+\end{itemize}
+
+The purpose of the next case study is to show how \pkg{casebase} can also be used for survival analysis problems that do not share these properties.
+To this end, we will use the Stanford Heart Transplant Data [@clark1971cardiac,@crowley1977covariance]; we will use the version available in the \pkg{survival} package.
+
+```{r stanford-poptime, eval = eval_cs4}
+library(survival)
+library(casebase)
+
+stanford_popTime <- popTime(jasa, time = "futime", 
+                            event = "fustat")
+plot(stanford_popTime)
+```
+
+Since the exposure is time dependent, we need to manually define the exposure variable *after* case-base sampling and *before* fitting the hazard function.
+For this reason, we will use the `sampleCaseBase` function directly.
+
+```{r, eval = eval_cs4}
+library(tidyverse)
+library(lubridate)
+
+cb_data <- sampleCaseBase(jasa, time = "futime", 
+                          event = "fustat", ratio = 10)
+```
+
+Next, we will compute the number of days from acceptance into the program to transplant, and we use this variable to determine whether each population-moment is exposed or not.
+
+```{r, eval = eval_cs4}
+# Define exposure variable
+cb_data <- mutate(cb_data,
+                  txtime = time_length(accept.dt %--% tx.date, 
+                                       unit = "days"),
+                  exposure = case_when(
+                    is.na(txtime) ~ 0L,
+                    txtime > futime ~ 0L,
+                    txtime <= futime ~ 1L
+                  ))
+```
+
+Finally, we can fit the hazard using various linear predictors.
+
+```{r, eval = eval_cs4, warning = FALSE}
+library(splines)
+# Fit several models
+fit1 <- fitSmoothHazard(fustat ~ exposure,
+                        data = cb_data, time = "futime")
+fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
+                        data = cb_data, time = "futime")
+fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
+                        data = cb_data, time = "futime")
+fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
+                        data = cb_data, time = "futime")
+```
+
+Note that the fourth model (i.e.
+`fit4`) includes an interaction term between exposure and follow-up time.
+In other words, this model no longer exhibit proportional hazards.
+We can then compare the goodness of fit of these four models using the Akaike Information Criterion (AIC).
+
+```{r, eval = eval_cs4}
+# Compute AIC
+c("Model1" = AIC(fit1),
+  "Model2" = AIC(fit2),
+  "Model3" = AIC(fit3),
+  "Model4" = AIC(fit4))
+```
+
+As we can, the best fit is the fourth model. By visualizing the hazard functions for both exposed and unexposed individuals, we can more clearly see how the hazards are no longer proportional.
+
+```{r stanford-hazard, eval = eval_cs4}
+# Compute hazards---
+# First, create a list of time points for both exposure status
+hazard_data <- expand.grid(exposure = c(0, 1),
+                           futime = seq(0, 1000,
+                                        length.out = 100))
+# Set the offset to zero
+hazard_data$offset <- 0 
+# Use predict to get the fitted values, and exponentiate to 
+# transform to the right scale
+hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
+                                 type = "link"))
+# Add labels for plots
+hazard_data$Status = factor(hazard_data$exposure,
+                            labels = c("NoTrans", "Trans"))
+
+ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
+    geom_line() +
+    theme_minimal() +
+    theme(legend.position = 'top') +
+    ylab('Hazard') + xlab('Follow-up time')
+```
+
+The non-proportionality seems to be more pronounced at the beginning of follow-up than the end.
+Finally, we can turn these estimate of the hazard function into estimates of the cumulative incidence functions.
+
+```{r stanford-risk, eval = eval_cs4}
+# Compute absolute risk curves
+newdata <- data.frame(exposure = c(0, 1))
+absrisk <- absoluteRisk(fit4, newdata = newdata, 
+                        time = seq(0, 1000, length.out = 100))
+
+colnames(absrisk) <- c("Time", "NoTrans", "Trans")
+
+# Rearrange the data
+absrisk <- gather(as.data.frame(absrisk),
+                  "Status", "Risk", -Time)
+
+ggplot(absrisk, aes(Time, Risk, colour = Status)) +
+  geom_line() +
+  theme_minimal() +
+  theme(legend.position = 'top') +
+  expand_limits(y = 0.5) +
+  xlab('Follow-up time') + ylab('Cum. Incidence')
+```
+
+Note that we can easily adapt the code above to the situation where a patient receives a heart transplant at a point in time of interest, for example after 30 days.
+
+```{r stanford-1y-haz, eval = eval_cs4}
+# Compute hazards---
+# First, create a list of time points for both exposure status
+one_yr_haz <- data.frame(futime = seq(0, 365,
+                                      length.out = 100))
+one_yr_haz <- mutate(one_yr_haz,
+                     offset = 0,
+                     exposure = if_else(futime < 30, 0, 1))
+one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
+                                type = "link"))
+ggplot(one_yr_haz, aes(futime, hazard)) +
+  geom_line() +
+  theme_minimal() +
+  theme(legend.position = 'top') +
+  ylab('Hazard') + xlab('Follow-up time') +
+  geom_vline(xintercept = 30, linetype = 'dashed')
+```
+
+We can then compare the 1-year mortality risk without transplant and with transplant at 30 days.
+
+```{r stanford-1y-risk, eval = eval_cs4}
+absoluteRisk(fit4, newdata = data.frame(exposure = 0),
+             time = 365)
+
+# Use the trapezoidal rule to estimate the cumulative hazard
+min_hazard <- min(one_yr_haz$hazard)
+max_hazard <- max(one_yr_haz$hazard)
+increment <- 365/99
+cumhaz_est <- 0.5*increment*(sum(2*one_yr_haz$hazard) - min_hazard - max_hazard)
+1 - exp(-cumhaz_est)
+```
+
+As we can see, the risk estimate at 1-year is about 30% lower if the patient receives a heart transplant at 30 days.
+
 # Discussion

### 2019-12-05 analysis/paper/paper.tex
add analysis of stanford data

--- 2019-07-25 analysis/paper/paper.tex
@@
 \author{
-Sahir Bhatnagar *\\McGill Univeristy \And Maxime Turgeon *\\University of Manitoba \And Jesse Islam\\McGill University \And James Hanley\\McGill Univeristy \And Olli Saarela\\University of Toronto
+Sahir Bhatnagar *\\McGill University \And Maxime Turgeon *\\University of Manitoba \AND Jesse Islam\\McGill University \And James Hanley\\McGill University \And Olli Saarela\\University of Toronto
 }
@@
     Sahir Bhatnagar *\\
-  McGill Univeristy\\
+  McGill University\\
   1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
@@
   University of Manitoba\\
-  318 Machray Hall Winnipeg, MB, Canada R3T 2N2\\
+  186 Dysart Road Winnipeg, MB, Canada R3T 2N2\\
   E-mail: \email{max.turgeon@umanitoba.ca}\\
@@
       James Hanley\\
-  McGill Univeristy\\
+  McGill University\\
   1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
@@
 
-\hypertarget{code-formatting}{%
-\section*{Code formatting}\label{code-formatting}}
-\addcontentsline{toc}{section}{Code formatting}
-
-Don't use markdown, instead use the more precise latex commands:
-
-\begin{itemize}
-\item
-  \proglang{Java}
-\item
-  \pkg{plyr}
-\item
-  \code{print("abc")}
-\end{itemize}
-
 \hypertarget{introduction}{%
@@
 
-The purpose of the \texttt{casebase} package is to provide practitioners
+The purpose of the \pkg{casebase} package is to provide practitioners
 with an easy-to-use software tool to predict the risk (or cumulative
@@
 likelihood arising from this data-generating mechanism using the
-framework of non-homogenous Poisson processes, we eventually reach an
+framework of non-homogeneous Poisson processes, we eventually reach an
 expression where each person-moment's contribution is of the form
@@
 In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
-suggest performing case-base samping \emph{uniformly}, i.e.~to sample
+suggest performing case-base sampling \emph{uniformly}, i.e.~to sample
 the base series uniformly from the study base. In terms of Poisson
@@
 categories: 1) data visualization, in the form of population-time plots;
-and 2) parametric modeling. We explicitely aimed at being compatible
-with both \code{data.frame}s and \code{data.table}s. This is evident in
-some of the coding choices we made, and it is also reflected in our unit
+and 2) parametric modeling. We explicitly aimed at being compatible with
+both \code{data.frame}s and \code{data.table}s. This is evident in some
+of the coding choices we made, and it is also reflected in our unit
 tests.
@@
 
+\hypertarget{case-study-4stanford-heart-transplant-data}{%
+\section{Case study 4--Stanford Heart Transplant
+Data}\label{case-study-4stanford-heart-transplant-data}}
+
+Although the previous case studies provide a broad overview of the
+capabilities of \pkg{casebase}, they all have two properties in common:
+
+\begin{itemize}
+  \item All covariates are fixed, i.e. they do not change over time;
+  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
+\end{itemize}
+
+The purpose of the next case study is to show how \pkg{casebase} can
+also be used for survival analysis problems that do not share these
+properties. To this end, we will use the Stanford Heart Transplant Data
+\citep[\citet{crowley1977covariance}]{clark1971cardiac}; we will use the
+version available in the \pkg{survival} package.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> library(survival)
+R> library(casebase)
+R> 
+R> stanford_popTime <- popTime(jasa, time = "futime", 
+R+                             event = "fustat")
+R> plot(stanford_popTime)
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/stanford-poptime-1} \end{center}
+
+\end{CodeChunk}
+
+Since the exposure is time dependent, we need to manually define the
+exposure variable \emph{after} case-base sampling and \emph{before}
+fitting the hazard function. For this reason, we will use the
+\texttt{sampleCaseBase} function directly.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> library(tidyverse)
+R> library(lubridate)
+R> 
+R> cb_data <- sampleCaseBase(jasa, time = "futime", 
+R+                           event = "fustat", ratio = 10)
+\end{CodeInput}
+\end{CodeChunk}
+
+Next, we will compute the number of days from acceptance into the
+program to transplant, and we use this variable to determine whether
+each population-moment is exposed or not.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Define exposure variable
+R> cb_data <- mutate(cb_data,
+R+                   txtime = time_length(accept.dt %--% tx.date, 
+R+                                        unit = "days"),
+R+                   exposure = case_when(
+R+                     is.na(txtime) ~ 0L,
+R+                     txtime > futime ~ 0L,
+R+                     txtime <= futime ~ 1L
+R+                   ))
+\end{CodeInput}
+\end{CodeChunk}
+
+Finally, we can fit the hazard using various linear predictors.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> library(splines)
+R> # Fit several models
+R> fit1 <- fitSmoothHazard(fustat ~ exposure,
+R+                         data = cb_data, time = "futime")
+R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
+R+                         data = cb_data, time = "futime")
+R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
+R+                         data = cb_data, time = "futime")
+R> fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
+R+                         data = cb_data, time = "futime")
+\end{CodeInput}
+\end{CodeChunk}
+
+Note that the fourth model (i.e. \texttt{fit4}) includes an interaction
+term between exposure and follow-up time. In other words, this model no
+longer exhibit proportional hazards. We can then compare the goodness of
+fit of these four models using the Akaike Information Criterion (AIC).
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Compute AIC
+R> c("Model1" = AIC(fit1),
+R+   "Model2" = AIC(fit2),
+R+   "Model3" = AIC(fit3),
+R+   "Model4" = AIC(fit4))
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> Model1 Model2 Model3 Model4 
+#>    796    760    730    728
+\end{CodeOutput}
+\end{CodeChunk}
+
+As we can, the best fit is the fourth model. By visualizing the hazard
+functions for both exposed and unexposed individuals, we can more
+clearly see how the hazards are no longer proportional.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Compute hazards---
+R> # First, create a list of time points for both exposure status
+R> hazard_data <- expand.grid(exposure = c(0, 1),
+R+                            futime = seq(0, 1000,
+R+                                         length.out = 100))
+R> # Set the offset to zero
+R> hazard_data$offset <- 0 
+R> # Use predict to get the fitted values, and exponentiate to 
+R> # transform to the right scale
+R> hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
+R+                                  type = "link"))
+R> # Add labels for plots
+R> hazard_data$Status = factor(hazard_data$exposure,
+R+                             labels = c("NoTrans", "Trans"))
+R> 
+R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
+R+     geom_line() +
+R+     theme_minimal() +
+R+     theme(legend.position = 'top') +
+R+     ylab('Hazard') + xlab('Follow-up time')
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/stanford-hazard-1} \end{center}
+
+\end{CodeChunk}
+
+The non-proportionality seems to be more pronounced at the beginning of
+follow-up than the end. Finally, we can turn these estimate of the
+hazard function into estimates of the cumulative incidence functions.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Compute absolute risk curves
+R> newdata <- data.frame(exposure = c(0, 1))
+R> absrisk <- absoluteRisk(fit4, newdata = newdata, 
+R+                         time = seq(0, 1000, length.out = 100))
+R> 
+R> colnames(absrisk) <- c("Time", "NoTrans", "Trans")
+R> 
+R> # Rearrange the data
+R> absrisk <- gather(as.data.frame(absrisk),
+R+                   "Status", "Risk", -Time)
+R>  
+R> ggplot(absrisk, aes(Time, Risk, colour = Status)) +
+R+   geom_line() +
+R+   theme_minimal() +
+R+   theme(legend.position = 'top') +
+R+   expand_limits(y = 0.5) +
+R+   xlab('Follow-up time') + ylab('Cum. Incidence')
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/stanford-risk-1} \end{center}
+
+\end{CodeChunk}
+
+Note that we can easily adapt the code above to the situation where a
+patient receives a heart transplant at a point in time of interest, for
+example after 30 days.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Compute hazards---
+R> # First, create a list of time points for both exposure status
+R> one_yr_haz <- data.frame(futime = seq(0, 365,
+R+                                       length.out = 100))
+R> one_yr_haz <- mutate(one_yr_haz,
+R+                      offset = 0,
+R+                      exposure = if_else(futime < 30, 0, 1))
+R> one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
+R+                                 type = "link"))
+R> ggplot(one_yr_haz, aes(futime, hazard)) +
+R+   geom_line() +
+R+   theme_minimal() +
+R+   theme(legend.position = 'top') +
+R+   ylab('Hazard') + xlab('Follow-up time') +
+R+   geom_vline(xintercept = 30, linetype = 'dashed')
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/stanford-1y-haz-1} \end{center}
+
+\end{CodeChunk}
+
+We can then compare the 1-year mortality risk without transplant and
+with transplant at 30 days.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> absoluteRisk(fit4, newdata = data.frame(exposure = 0),
+R+              time = 365)
+\end{CodeInput}
+
+\begin{CodeOutput}
+#>         
+#> 365 0.27
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> 
+R> # Use the trapezoidal rule to estimate the cumulative hazard
+R> min_hazard <- min(one_yr_haz$hazard)
+R> max_hazard <- max(one_yr_haz$hazard)
+R> increment <- 365/99
+R> cumhaz_est <- 0.5*increment*(sum(2*one_yr_haz$hazard) - min_hazard - max_hazard)
+R> 1 - exp(-cumhaz_est)
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> [1] 0.18
+\end{CodeOutput}
+\end{CodeChunk}
+
+As we can see, the risk estimate at 1-year is about 30\% lower if the
+patient receives a heart transplant at 30 days.
+
 \hypertarget{discussion}{%
@@
 
+In this article, we presented the \proglang{R} package \pkg{casebase}
+that provides functions to fit smooth parametric hazards and estimate
+cumulative incidence functions using case-base sampling. We outlined the
+theoretical underpinnings of the approach, we provided details about our
+implementation, and we illustrated the merits of the approach and the
+package through three case studies.
+
 In the following table we provide a comparison between the Cox model and
@@
 
-This report was generated on 2019-07-25 16:51:47 using the following
+This report was generated on 2019-12-05 17:15:04 using the following
 computational environment and dependencies:
@@
 \begin{CodeOutput}
-#> - Session info ----------------------------------------------------------
+#> - Session info ---------------------------------------------------------------
 #>  setting  value                       
 #>  version  R version 3.6.1 (2019-07-05)
-#>  os       Ubuntu 18.04.2 LTS          
+#>  os       Ubuntu 18.04.3 LTS          
 #>  system   x86_64, linux-gnu           
@@
 #>  tz       America/Winnipeg            
-#>  date     2019-07-25                  
+#>  date     2019-12-05                  
 #> 
-#> - Packages --------------------------------------------------------------
-#>  package     * version    date       lib source        
-#>  assertthat    0.2.1      2019-03-21 [1] CRAN (R 3.6.1)
-#>  backports     1.1.4      2019-04-10 [1] CRAN (R 3.6.1)
-#>  broom         0.5.2      2019-04-07 [1] CRAN (R 3.6.1)
-#>  callr         3.3.1      2019-07-18 [1] CRAN (R 3.6.1)
-#>  casebase    * 0.2.1.9000 2019-07-25 [1] local         
-#>  cellranger    1.1.0      2016-07-27 [1] CRAN (R 3.6.1)
-#>  cli           1.1.0      2019-03-19 [1] CRAN (R 3.6.1)
-#>  colorspace    1.4-1      2019-03-18 [1] CRAN (R 3.6.1)
-#>  crayon        1.3.4      2017-09-16 [1] CRAN (R 3.6.1)
-#>  data.table    1.12.2     2019-04-07 [1] CRAN (R 3.6.1)
-#>  desc          1.2.0      2018-05-01 [1] CRAN (R 3.6.1)
-#>  devtools      2.1.0      2019-07-06 [1] CRAN (R 3.6.1)
-#>  digest        0.6.20     2019-07-04 [1] CRAN (R 3.6.1)
-#>  dplyr       * 0.8.3      2019-07-04 [1] CRAN (R 3.6.1)
-#>  evaluate      0.14       2019-05-28 [1] CRAN (R 3.6.1)
-#>  forcats     * 0.4.0      2019-02-17 [1] CRAN (R 3.6.1)
-#>  fs            1.3.1      2019-05-06 [1] CRAN (R 3.6.1)
-#>  generics      0.0.2      2018-11-29 [1] CRAN (R 3.6.1)
-#>  ggplot2     * 3.2.0      2019-06-16 [1] CRAN (R 3.6.1)
-#>  glue          1.3.1      2019-03-12 [1] CRAN (R 3.6.1)
-#>  gtable        0.3.0      2019-03-25 [1] CRAN (R 3.6.1)
-#>  haven         2.1.1      2019-07-04 [1] CRAN (R 3.6.1)
-#>  hms           0.5.0      2019-07-09 [1] CRAN (R 3.6.1)
-#>  htmltools     0.3.6      2017-04-28 [1] CRAN (R 3.6.1)
-#>  httr          1.4.0      2018-12-11 [1] CRAN (R 3.6.1)
-#>  jsonlite      1.6        2018-12-07 [1] CRAN (R 3.6.1)
-#>  knitr         1.23       2019-05-18 [1] CRAN (R 3.6.1)
-#>  lattice       0.20-38    2018-11-04 [4] CRAN (R 3.5.1)
-#>  lazyeval      0.2.2      2019-03-15 [1] CRAN (R 3.6.1)
-#>  lubridate     1.7.4      2018-04-11 [1] CRAN (R 3.6.1)
-#>  magrittr    * 1.5        2014-11-22 [1] CRAN (R 3.6.1)
-#>  Matrix        1.2-17     2019-03-22 [4] CRAN (R 3.5.3)
-#>  memoise       1.1.0      2017-04-21 [1] CRAN (R 3.6.1)
-#>  mgcv          1.8-28     2019-03-21 [4] CRAN (R 3.5.3)
-#>  modelr        0.1.4      2019-02-18 [1] CRAN (R 3.6.1)
-#>  munsell       0.5.0      2018-06-12 [1] CRAN (R 3.6.1)
-#>  nlme          3.1-140    2019-05-12 [4] CRAN (R 3.6.0)
-#>  pillar        1.4.2      2019-06-29 [1] CRAN (R 3.6.1)
-#>  pkgbuild      1.0.3      2019-03-20 [1] CRAN (R 3.6.1)
-#>  pkgconfig     2.0.2      2018-08-16 [1] CRAN (R 3.6.1)
-#>  pkgload       1.0.2      2018-10-29 [1] CRAN (R 3.6.1)
-#>  prettyunits   1.0.2      2015-07-13 [1] CRAN (R 3.6.1)
-#>  processx      3.4.1      2019-07-18 [1] CRAN (R 3.6.1)
-#>  ps            1.3.0      2018-12-21 [1] CRAN (R 3.6.1)
-#>  purrr       * 0.3.2      2019-03-15 [1] CRAN (R 3.6.1)
-#>  R6            2.4.0      2019-02-14 [1] CRAN (R 3.6.1)
-#>  Rcpp          1.0.1      2019-03-17 [1] CRAN (R 3.6.1)
-#>  readr       * 1.3.1      2018-12-21 [1] CRAN (R 3.6.1)
-#>  readxl        1.3.1      2019-03-13 [1] CRAN (R 3.6.1)
-#>  remotes       2.1.0      2019-06-24 [1] CRAN (R 3.6.1)
-#>  rlang         0.4.0      2019-06-25 [1] CRAN (R 3.6.1)
-#>  rmarkdown     1.14       2019-07-12 [1] CRAN (R 3.6.1)
-#>  rprojroot     1.3-2      2018-01-03 [1] CRAN (R 3.6.1)
-#>  rstudioapi    0.10       2019-03-19 [1] CRAN (R 3.6.1)
-#>  rticles       0.9        2019-07-12 [1] CRAN (R 3.6.1)
-#>  rvest         0.3.4      2019-05-15 [1] CRAN (R 3.6.1)
-#>  scales        1.0.0      2018-08-09 [1] CRAN (R 3.6.1)
-#>  sessioninfo   1.1.1      2018-11-05 [1] CRAN (R 3.6.1)
-#>  stringi       1.4.3      2019-03-12 [1] CRAN (R 3.6.1)
-#>  stringr     * 1.4.0      2019-02-10 [1] CRAN (R 3.6.1)
-#>  survival    * 2.43-3     2018-11-26 [4] CRAN (R 3.5.1)
-#>  testthat      2.2.0      2019-07-22 [1] CRAN (R 3.6.1)
-#>  tibble      * 2.1.3      2019-06-06 [1] CRAN (R 3.6.1)
-#>  tidyr       * 0.8.3      2019-03-01 [1] CRAN (R 3.6.1)
-#>  tidyselect    0.2.5      2018-10-11 [1] CRAN (R 3.6.1)
-#>  tidyverse   * 1.2.1      2017-11-14 [1] CRAN (R 3.6.1)
-#>  usethis       1.5.1      2019-07-04 [1] CRAN (R 3.6.1)
-#>  vctrs         0.2.0      2019-07-05 [1] CRAN (R 3.6.1)
-#>  VGAM          1.1-1      2019-02-18 [1] CRAN (R 3.6.1)
-#>  withr         2.1.2      2018-03-15 [1] CRAN (R 3.6.1)
-#>  xfun          0.8        2019-06-25 [1] CRAN (R 3.6.1)
-#>  xml2          1.2.0      2018-01-24 [1] CRAN (R 3.6.1)
-#>  yaml          2.2.0      2018-07-25 [1] CRAN (R 3.6.1)
-#>  zeallot       0.1.0      2018-01-28 [1] CRAN (R 3.6.1)
+#> - Packages -------------------------------------------------------------------
+#>  package     * version    date       lib
+#>  assertthat    0.2.1      2019-03-21 [1]
+#>  backports     1.1.5      2019-10-02 [1]
+#>  broom         0.5.2      2019-04-07 [1]
+#>  callr         3.3.2      2019-09-22 [1]
+#>  casebase    * 0.2.1.9001 2019-08-13 [1]
+#>  cellranger    1.1.0      2016-07-27 [1]
+#>  cli           1.1.0      2019-03-19 [1]
+#>  colorspace    1.4-1      2019-03-18 [1]
+#>  crayon        1.3.4      2017-09-16 [1]
+#>  data.table    1.12.6     2019-10-18 [1]
+#>  DBI           1.0.0      2018-05-02 [1]
+#>  dbplyr        1.4.2      2019-06-17 [1]
+#>  desc          1.2.0      2018-05-01 [1]
+#>  devtools      2.2.1      2019-09-24 [1]
+#>  digest        0.6.23     2019-11-23 [1]
+#>  dplyr       * 0.8.3      2019-07-04 [1]
+#>  ellipsis      0.3.0      2019-09-20 [1]
+#>  evaluate      0.14       2019-05-28 [1]
+#>  farver        2.0.1      2019-11-13 [1]
+#>  forcats     * 0.4.0      2019-02-17 [1]
+#>  fs            1.3.1      2019-05-06 [1]
+#>  generics      0.0.2      2018-11-29 [1]
+#>  ggplot2     * 3.2.1      2019-08-10 [1]
+#>  glue          1.3.1      2019-03-12 [1]
+#>  gtable        0.3.0      2019-03-25 [1]
+#>  haven         2.2.0      2019-11-08 [1]
+#>  hms           0.5.2      2019-10-30 [1]
+#>  htmltools     0.4.0      2019-10-04 [1]
+#>  httr          1.4.1      2019-08-05 [1]
+#>  jsonlite      1.6        2018-12-07 [1]
+#>  knitr         1.26       2019-11-12 [1]
+#>  labeling      0.3        2014-08-23 [1]
+#>  lattice       0.20-38    2018-11-04 [4]
+#>  lazyeval      0.2.2      2019-03-15 [1]
+#>  lifecycle     0.1.0      2019-08-01 [1]
+#>  lubridate   * 1.7.4      2018-04-11 [1]
+#>  magrittr    * 1.5        2014-11-22 [1]
+#>  Matrix        1.2-17     2019-03-22 [4]
+#>  memoise       1.1.0      2017-04-21 [1]
+#>  mgcv          1.8-31     2019-11-09 [4]
+#>  modelr        0.1.5      2019-08-08 [1]
+#>  munsell       0.5.0      2018-06-12 [1]
+#>  nlme          3.1-142    2019-11-07 [4]
+#>  pillar        1.4.2      2019-06-29 [1]
+#>  pkgbuild      1.0.6      2019-10-09 [1]
+#>  pkgconfig     2.0.3      2019-09-22 [1]
+#>  pkgload       1.0.2      2018-10-29 [1]
+#>  prettyunits   1.0.2      2015-07-13 [1]
+#>  processx      3.4.1      2019-07-18 [1]
+#>  ps            1.3.0      2018-12-21 [1]
+#>  purrr       * 0.3.3      2019-10-18 [1]
+#>  R6            2.4.1      2019-11-12 [1]
+#>  Rcpp          1.0.3      2019-11-08 [1]
+#>  readr       * 1.3.1      2018-12-21 [1]
+#>  readxl        1.3.1      2019-03-13 [1]
+#>  remotes       2.1.0      2019-06-24 [1]
+#>  reprex        0.3.0      2019-05-16 [1]
+#>  rlang         0.4.2      2019-11-23 [1]
+#>  rmarkdown     1.18       2019-11-27 [1]
+#>  rprojroot     1.3-2      2018-01-03 [1]
+#>  rstudioapi    0.10       2019-03-19 [1]
+#>  rticles       0.12       2019-11-12 [1]
+#>  rvest         0.3.5      2019-11-08 [1]
+#>  scales        1.1.0      2019-11-18 [1]
+#>  sessioninfo   1.1.1      2018-11-05 [1]
+#>  stringi       1.4.3      2019-03-12 [1]
+#>  stringr     * 1.4.0      2019-02-10 [1]
+#>  survival    * 2.44-1.1   2019-04-01 [4]
+#>  testthat      2.3.1      2019-12-01 [1]
+#>  tibble      * 2.1.3      2019-06-06 [1]
+#>  tidyr       * 1.0.0      2019-09-11 [1]
+#>  tidyselect    0.2.5      2018-10-11 [1]
+#>  tidyverse   * 1.3.0      2019-11-21 [1]
+#>  usethis       1.5.1      2019-07-04 [1]
+#>  vctrs         0.2.0      2019-07-05 [1]
+#>  VGAM          1.1-2      2019-11-21 [1]
+#>  withr         2.1.2      2018-03-15 [1]
+#>  xfun          0.11       2019-11-12 [1]
+#>  xml2          1.2.2      2019-08-09 [1]
+#>  yaml          2.2.0      2018-07-25 [1]
+#>  zeallot       0.1.0      2018-01-28 [1]
+#>  source                                  
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  Github (sahirbhatnagar/casebase@aecb816)
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
 #> 
@@
 #> Local:    review-max /home/turgeonm/Documents/git_repos/cbpaper
-#> Head:     [c0cb26d] 2019-07-23: Merge pull request #6 from sahirbhatnagar/fixdocker
+#> Head:     [77f5e41] 2019-08-02: fix overlapping author names
 \end{CodeOutput}

### 2019-12-06 analysis/paper/paper.Rmd
add some details on stanford data

@@
 
-The purpose of the next case study is to show how \pkg{casebase} can also be used for survival analysis problems that do not share these properties.
+Case-base sampling has been used in the literature to study vaccination safety, where the exposure period was defined as the week following vaccination [@saarela2015case].
+Hence, the main covariate of interest, i.e.
+exposure to the vaccine, was changing over time.
+In this context, case-base sampling offers an efficient alternative to nested case-control designs or self-matching.
+
+The purpose of the next case study is to show how \pkg{casebase} can also be used for survival analysis problems that do not share the properties above.
 To this end, we will use the Stanford Heart Transplant Data [@clark1971cardiac,@crowley1977covariance]; we will use the version available in the \pkg{survival} package.
+
+Recall the setting of this study: patients were admitted to the Stanford program after meeting with their physician and determining that they were unlikely to respond to other forms of treatment.
+After enrollment, the program searches for a suitable donor for the patient, which can take anywhere between a few days to almost a year.
+We are interested in the effect of a heart transplant on survival; therefore, the patient is considered exposed only after the transplant has occured.
+
+As before, we can look at the population-time plot for a graphical summary of the event incidence.
+As we can see, most events occur early during the follow-up period, and therefore we do not expect the hazard to be constant.
 
@@
 
-Since the exposure is time dependent, we need to manually define the exposure variable *after* case-base sampling and *before* fitting the hazard function.
+Since the exposure is time-dependent, we need to manually define the exposure variable *after* case-base sampling and *before* fitting the hazard function.
 For this reason, we will use the `sampleCaseBase` function directly.
@@
 In other words, this model no longer exhibit proportional hazards.
+The evidence of non-proportionality of hazards in the Stanford Heart Transplant data has been widely discussed [@arjas1988graphical].
+
 We can then compare the goodness of fit of these four models using the Akaike Information Criterion (AIC).

### 2019-12-06 analysis/paper/paper.tex
add some details on stanford data

@@
 
+Case-base sampling has been used in the literature to study vaccination
+safety, where the exposure period was defined as the week following
+vaccination \citep{saarela2015case}. Hence, the main covariate of
+interest, i.e.~exposure to the vaccine, was changing over time. In this
+context, case-base sampling offers an efficient alternative to nested
+case-control designs or self-matching.
+
 The purpose of the next case study is to show how \pkg{casebase} can
-also be used for survival analysis problems that do not share these
-properties. To this end, we will use the Stanford Heart Transplant Data
-\citep[\citet{crowley1977covariance}]{clark1971cardiac}; we will use the
-version available in the \pkg{survival} package.
+also be used for survival analysis problems that do not share the
+properties above. To this end, we will use the Stanford Heart Transplant
+Data \citep[\citet{crowley1977covariance}]{clark1971cardiac}; we will
+use the version available in the \pkg{survival} package.
+
+Recall the setting of this study: patients were admitted to the Stanford
+program after meeting with their physician and determining that they
+were unlikely to respond to other forms of treatment. After enrollment,
+the program searches for a suitable donor for the patient, which can
+take anywhere between a few days to almost a year. We are interested in
+the effect of a heart transplant on survival; therefore, the patient is
+considered exposed only after the transplant has occured.
+
+As before, we can look at the population-time plot for a graphical
+summary of the event incidence. As we can see, most events occur early
+during the follow-up period, and therefore we do not expect the hazard
+to be constant.
 
@@
 
-Since the exposure is time dependent, we need to manually define the
+Since the exposure is time-dependent, we need to manually define the
 exposure variable \emph{after} case-base sampling and \emph{before}
@@
 term between exposure and follow-up time. In other words, this model no
-longer exhibit proportional hazards. We can then compare the goodness of
-fit of these four models using the Akaike Information Criterion (AIC).
+longer exhibit proportional hazards. The evidence of non-proportionality
+of hazards in the Stanford Heart Transplant data has been widely
+discussed \citep{arjas1988graphical}.
+
+We can then compare the goodness of fit of these four models using the
+Akaike Information Criterion (AIC).
 
@@
 
-This report was generated on 2019-12-05 17:15:04 using the following
+This report was generated on 2019-12-06 10:11:23 using the following
 computational environment and dependencies:
@@
 #>  tz       America/Winnipeg            
-#>  date     2019-12-05                  
+#>  date     2019-12-06                  
 #> 
@@
 #> Local:    review-max /home/turgeonm/Documents/git_repos/cbpaper
-#> Head:     [77f5e41] 2019-08-02: fix overlapping author names
+#> Head:     [d0d7b8a] 2019-12-05: add analysis of stanford data
 \end{CodeOutput}

### 2019-12-06 analysis/paper/paper.Rmd
currentDraftEdits
added lit-review and placeholder analysis (this will change once casebase is properly updated)

--- 2019-08-02 analysis/paper/paper.Rmd
@@
   - name: Maxime Turgeon *
-    affiliation: 'University of Manitoba \AND'
+    affiliation: University of Manitoba
     address: >
-      186 Dysart Road
+      318 Machray Hall
       Winnipeg, MB, Canada R3T 2N2
@@
   \usepackage{longtable}
+  \usepackage{graphicx}
 output: 
@@
 
-# Code formatting {.unnumbered}
+# Code formatting
 
@@
 * \code{print("abc")}
-
 ```{r setup, include=FALSE}
@@
 library(splines)
+library(readr)
+library(mice)
+library(VIM)
+library(tibble)
+library(glmnet)
+library(ggplot2)
+devtools::load_all()
 options(digits = 2, scipen = 999)
 # set the seed for reproducible output
-set.seed(1234)
+set.seed(2)
 
 eval_introduction <- TRUE
-eval_theory <- TRUE
+eval_theory <- FALSE
 eval_implementation <- TRUE
-eval_cs1 <- TRUE
-eval_cs2 <- TRUE
-eval_cs3 <- FALSE
+eval_cs1 <- FALSE
+eval_cs2 <- FALSE
+eval_cs3 <- TRUE
 ```
@@
 
-The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence (CI)) of an event, for a particular patient.
+The purpose of the `casebase` package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence (CI)) of an event, for a particular patient.
 The following points should be noted:
 
-1. Time matching/risk set sampling (including Cox partial likelihood) eliminates the baseline hazard from the likelihood expression for the
 hazard ratios
@@
 
+# Literature review
+To conduct our review on current methods available in survival analysis, a scraper was coded to examine all packages in the survival task view on CRAN (ref).
+There was a total of 259 packages, many of which are not relevant to the problems casebase addresses.
+To filter these packages, a scraper was used to search for competing risks, cumulative incidence functions, non- proportional hazards and penalization.
+This reduced our set of 259 packages down to 60.
+From here, we manually examined each package to determine which ones are comparable to casebase, with either a different approach to the same problems or future features we would like to add to casebase.
+
+These packages can belong to three main categories: non-proportional hazards, regularization and competing risks.
+Besides these three main categories, there are extra features of interest as well.
+Table XXX summarizes the findings.
+
+Many packages exist that implement semi-parametric cox models.
+Of those, glmnet, glmpath, penalized, RiskRegression introduce penalization in their model fitting.
+CoxRidge, rstpm2 and survival include penalization, but they implement the extended cox model permitting intuitive non proportional hazard modeling.
+In terms of penalization, the benefit of casebase comes from its flexibility.
+Linear models have a wide tool set, considerably larger than those made for cox regression (reference for number of packages between their task views, regression is spread out across Multivariate, SocialSciences, Robust among others.
+The focus of this task view is on Optimization Infrastructure Packages , General Purpose Continuous Solvers , Mathematical Programming Solvers , and Specific Applications in Optimization , or Multi Objective Optimization ).
+Casebase permits many regularization tools developed for linear regression to be repurposed within a survival context.
+
+Parametric models permit non-proportional hazards and are implemented in a handful of packages.
+Of interest are CFC, flexsurv, SmoothHazard, Rsptm2 and Survival.
+Smooth hazard is limited to weibull distributions, where as flexsurv and survival contain implementations for users to supply any distribution they would like.
+Flexsurv, smooth hazard, mets and rstpm2 can use splines, permitting a flexible fit over time.
+Though casebase is limited to 3 distributions at the moment, it also uses splines allowing for distributions that may not be consistent over time.
+
+ Of the methods mentioned so far, only CFC, survival, mets and flexsurv contain implementations for competing risks(how is it done for parametric models?).
+The differentiating factor between these packages and casebase is that casebase handles competing risks through multiple logistic regression.
+
+A feature in Smooth hazard, survival and Rstpm2 are they they can handle interval censoring, where as casebase currently can not.
+
+Of all the packages listed in the table, CFC, Flexsurv, mets, RiskRegression, rstpm2 and survival have implementations for cumulative incidence.
+Since casebase is parametric, the baseline hazard is known and requires integration to retrieve the cumulative incidence.
+This is similar to the approaches of CFC, flexsurv,
+\begin{table}[]
+\centering
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{ccccccccc}
+\hline
+\textbf{Package}        & \textbf{Competing Risks} & \textbf{Non-proportional} & \textbf{Penalization} & \textbf{Splines} & \textbf{Parametric} & \textbf{Semi-parametric} & \textbf{Interval/left Censoring} & \textbf{Absolute Risk}           \\
+\hline
+\textbf{casebase}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
+\textbf{CFC}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
+\textbf{coxRidge}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{crrp}           & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
+\textbf{fastcox}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{Flexrsurv}      &                          & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
+\\
+\textbf{Flexsurv}       & x                        & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
+\\
+\textbf{glmnet}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{glmpath}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{mets}           & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
+\textbf{penalized}      &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{RiskRegression} &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
+\textbf{Rstpm2}         &                          & x                         & x                     & x                & x                   & x                        & x                                & Cumm haz                         \\
+\textbf{SmoothHazard}   &                          & x                         &                       & x                & x                   &                          & x                                & Cumulative transition intensity?
+\\
+\textbf{Survival}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x
+\end{tabular}%
+}
+\caption{Different features of interest in various survival packages.}
+\label{tab:my-table}
+\end{table}
+
 # Theoretical details
@@
 
-<!-- Popular survival analysis methods, like Kaplan-Meier and Cox regression, rely on the notion of risk-set sampling. Case-base sampling can be seen as an alternative. -->
+Popular survival analysis methods, like Kaplan-Meier and Cox regression, rely on the notion of risk-set sampling. Case-base sampling can be seen as an alternative.
 
 As shown by Saarela & Arjas [-@saarela2015non] (and further expanded in Saarela [-@saarela2016case]), writing the likelihood arising from this data-generating mechanism using the framework of non-homogenous Poisson processes, we eventually reach an expression where each person-moment's contribution is of the form
-$$\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},$$
-where $N(t)$ is the counting process associated with the event of interest, $h(t)$ is the corresponding hazard function, and $\rho(t)$ is the hazard function for the Poisson process associated with case-base sampling.
-This parametric form suggests that we can readily estimate log-hazards of the form $\log(h(t)) = g(t; X)$ using logistic regression, where each observation corresponds to a person moment, the function $g(t; X)$ is linear in a finite number of parameters, and where we treat $-\log(\rho(t))$ as an offset.
+$$\frac{\lambda(t)^{dN(t)}}{\rho(t) + \lambda(t)},$$
+where $N(t)$ is the counting process associated with the event of interest, $\lambda(t)$ is the corresponding hazard function, and $\rho(t)$ is the hazard function for the Poisson process associated with case-base sampling.
+This parametric form suggests that we can readily estimate log-hazards of the form $\log(\lambda(t)) = g(t; X)$ using logistic regression, where each observation corresponds to a person moment, the function $g(t; X)$ is linear in a finite number of parameters, and where we treat $\log(\rho(t))$ as an offset.
 
@@
 to sample the base series uniformly from the study base.
-In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base.
+In terms of Poisson processes, this sampling strategy corresponds to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base.
 More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism  are proportional to the cardiovascular disease event rate given by the Framingham score.
@@
 
-$$ h(t;X) = \exp[g(t;X)] $$
-where $t$ denotes time and $X$, the individual's covariate profile.
+$$ h(x,t) = exp[g(x,t)] $$
+where $t$ denotes the numerical value (number of units) of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any).
 Different functions of $t$ lead to different parametric hazard models.
+
 The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$.
 
-$$ h(t;X) = \exp(\beta_0 + \beta_1 X) $$
+$$ h(x,t) = exp(\beta_0 + \beta_1 x) $$
 
-The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
-This is also known as the *memoryless property* (Kalbfleisch and Prentice, 2002).
+The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been on study a.k.a the memoryless property (Kalbfleisch and Prentice, 2002).
 
@@
 
-$$ h(t;X)  = \exp(\beta_0 + \beta_1 t + \beta_2 X) $$
+$$ h(x,t)  = exp(\beta_0 + \beta_1 t + \beta_2 x) $$
 
-Use of $\log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
+Use of $log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
 
-$$ h(t;X)  = \exp(\beta_0 + \beta_1 \log(t) + \beta_2 X) $$
-
-For competing-risk analyses with $J$ possible events, we can show that each person-moment's contribution of the likelihood is of the form
-
-$$\frac{h_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^Jh_j(t)},$$
-
-where $N_j(t)$ is the counting process associated with the event of type $j$ and $h_j(t)$ is the corresponding hazard function.
-As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1\end{align*}}
+$$ h(x, t)  = exp(\beta_0 + \beta_1 \log(t) + \beta_2 x) $$
 
@@
 
-The functions in the casebase package can be divided into two categories: 1) data visualization, in the form of population-time plots; and 2) parametric modeling.
-We explicitely aimed at being compatible with both \code{data.frame}s and \code{data.table}s.
-This is evident in some of the coding choices we made, and it is also reflected in our unit tests.
+The functions in the casebase package can be divided into two categories: 1) data visualization, in the form of population-time plots; and 2) data analysis.
+
+We explicitely aimed at being compatible with both \code{data.frame}s and \code{data.table}s. This is evident in some of the coding choices we made, and it is also reflected in our testing units.
 
@@
 
-## Parametric modeling
+## Data analysis
 
-The parametric modeling step was separated into three parts:
-
-  1. case-base sampling; 
-  2. estimation of the smooth hazard function; 
-  3. calculation of the risk function.
-
+The data analysis step was separated into three parts: 1) case-base sampling; 2) estimation of the smooth hazard function; and 3) calculation of the risk function.
 By separating the sampling and estimation functions, we allowed the possibility of users implementing more complex sampling scheme, as described in Saarela [-@saarela2016case].
@@
 
-To fit the hazard function, we provide several approaches that are available via the \code{family} parameter. These approaches are:
-
-  - \code{glm}: This is the familiar logistic regression.
-  - \code{glmnet}: This option allows for variable selection using Lasso or elastic-net. This functionality is provided through the \code{glmnet} package [@friedman2010jss].
-  - \code{gam}: This option provides support for *Generalized Additive Models* via the \code{gam} package [@hastie1987generalized].
-  - \code{gbm}: This option provides support for *Gradient Boosted Trees* via the \code{gbm} package. This feature is still experimental.
-
-In the case of multiple events, the hazard is fitted via multinomial regression as performed by the \pkg{VGAM} package.
+Finally, the hazard function is fitted to the data using the function \code{glm}, unless there is more than one type event (i.e.
+in a competing-risk analysis), in which case it uses the multinomical regression capabilities of the \pkg{VGAM} package.
 This package was selected for its ability to fit multinomial regression models with an offset.
@@
 $$S(t) = \exp\left(-\int_0^t h(u;X) du\right).$$
-The integral is computed using either the \code{stats::integrate} function or Monte-Carlo integration. The risk function (or cumulative incidence function) is then defined as
+The integral is computed using either the \code{stats::integrate} function or Monte-Carlo integration.
+For the case of a competing-event analysis, the event-specific risk is computed using a nested double integral; in this setting, Monte-Carlo integration is faster and more flexible than \code{stats::integrate}.
+This is due to the fact that the computation involves a double loop over the selected time points; Monte-Carlo integration performs this step by using the vectorized \code{rowSums} and \code{colSums} functions.
 
-$$ CI(t) = 1 - S(t).$$
-
-For the case of a competing-event analysis, the event-specific risk is computed using the following procedure: first, we compute the overall survival function (i.e. for all event types):
-
-$$ S(t) = \exp\left(-\int_0^t H(u;X) du\right),\qquad H(t;X) = \sum_{j=1}^J h_j(t;X).$$
-From this, we can derive the event-specific subdensities:
-
-$$ f_j(t) = h_j(t)S(t).$$
-
-Finally, by integrating these subdensities, we obtain the event-specific cumulative incidence functions:
-
-$$ CI_j(t) = \int_0^t f_j(u)du.$$
-
-We created \code{absoluteRisk} as an \code{S3} generic, with methods for the different types of outputs of \code{fitSmoothHazard}.
-The method dispatch system of \proglang{R} then takes care of matching the correct output to the correct methodology for calculating the cumulative incidence function, without the user's intervention.
-
-In the following sections, we illustrate these functionalities in the context of three case studies.
+To decide between a single-event and a competing-event analysis, we created \code{absoluteRisk} as an \code{S3} generic, with methods for both \code{glm} and \code{CompRisk} objects (the latter inherits from \code{vglm} as well).
+The method dispatch system of \proglang{R} then takes care of matching the correct input to the correct methodology, without the user's intervention.
 
@@
 
-$$ CI(x, t) = 1 - \exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$
+$$ CI(x, t) = 1 - exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$
 
@@
 ```{r erspc-casebase-splines, eval = eval_cs1}
-casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
+casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ splines::bs(Follow.Up.Time) + ScrArm, 
                                     data = ERSPC, 
@@
 \centering
-\begin{tabular}{lll}
+\begin{tabular}{ccc}
   \hline
@@
 
-```{r compPop1, fig.cap="Population-time plot for the stem-cell transplant study. The points represent the event of interest (i.e., relapse).", echo=FALSE, eval = eval_cs2}
+```{r compPop1, fig.cap="\\label{fig:compPop1}Population-time plot for the stem-cell transplant study. The points represent the event of interest (i.e., relapse).", echo=FALSE, eval = eval_cs2}
 pt_object <- casebase::popTime(bmtcrr, event = "Status", time = "ftime")
@@
 
-```{r compPop2, fig.cap="Population-time plot for the stem-cell transplant study. The points represent the competing events.", echo=FALSE, eval = eval_cs2}
+```{r compPop2, fig.cap="\\label{fig:compPop2}Population-time plot for the stem-cell transplant study. The points represent the competing events.", echo=FALSE, eval = eval_cs2}
 pt_object_comp <- bmtcrr %>% 
@@
 
-# Case study 3--The Cancer Genome Atlas
+# Case study 3--Study to Understand Prognoses Preferences Outcomes and Risks of Treatment
 
-  - Glmnet and TCGA data
+We examined the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset (REF from main site).
+the SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital.
+Imputation was conducted using the default methods in the mice package in R, available on CRAN.
+Before imputation, there was a total of 1000 individuals and 35 variables.
+After imputation, we have 474 observations where 67.5% of individuals in the study experience the event of interest.
+The SUPPORT dataset will be used to demonstrate regularization within casebase (REF), while comparing their absolute risk preditions for a new individual.
+A description of each variable can be found in table XXX1 and a breakdown of each categorical variable in table XXX2.
+The data was downloaded from the main author's website, and it is also available as part of the casebase package.
 
+\begin{table}[]
+\centering
+\begin{tabular}{ccccc}
+\hline
+\textbf{Name} & \textbf{Labels}                          & \textbf{Levels} & \textbf{Storage} & \textbf{NAs} \\
+\hline
+age           & Age                                      &                 & double           & 0            \\
+death         & Death at any time up to NDI date:31DEC94 &                 & double           & 0            \\
+sex           &                                          & 2               & integer          & 0            \\
+hospdead      & Death in Hospital                        &                 & double           & 0            \\
+slos          & Days from Study Entry to Discharge       &                 & double           & 0            \\
+d.time        & Days of Follow-Up                        &                 & double           & 0            \\
+dzgroup       &                                          & 8               & integer          & 0            \\
+dzclass       &                                          & 4               & integer          & 0            \\
+num.co        & number of comorbidities                  &                 & double           & 0            \\
+edu           & Years of Education                       &                 & double           & 202          \\
+income        &                                          & 4               & integer          & 349          \\
+scoma         & SUPPORT Coma Score based on Glasgow D3   &                 & double           & 0            \\
+charges       & Hospital Charges                         &                 & double           & 25           \\
+totcst        & Total RCC cost                           &                 & double           & 105          \\
+totmcst       & Total micro-cost                         &                 & double           & 372          \\
+avtisst       & Average TISS, Days 3-25                  &                 & double           & 6            \\
+race          &                                          & 5               & integer          & 5            \\
+meanbp        & Mean Arterial Blood Pressure Day 3       &                 & double           & 0            \\
+wblc          & White Blood Cell Count Day 3             &                 & double           & 24           \\
+hrt           & Heart Rate Day 3                         &                 & double           & 0            \\
+resp          & Respiration Rate Day 3                   &                 & double           & 0            \\
+temp          & Temperature (celcius) Day 3              &                 & double           & 0            \\
+pafi          & PaO2/(.01*FiO2) Day 3                    &                 & double           & 253          \\
+alb           & Serum Albumin Day 3                      &                 & double           & 378          \\
+bili          & Bilirubin Day 3                          &                 & double           & 297          \\
+crea          & Serum creatinine Day 3                   &                 & double           & 3            \\
+sod           & Serum sodium Day 3                       &                 & double           & 0            \\
+ph            & Serum pH (arterial) Day 3                &                 & double           & 250          \\
+glucose       & Glucose Day 3                            &                 & double           & 470          \\
+bun           & BUN Day 3                                &                 & double           & 455          \\
+urine         & Urine Output Day 3                       &                 & double           & 517          \\
+adlp          & ADL Patient Day 3                        &                 & double           & 634          \\
+adls          & ADL Surrogate Day 3                      &                 & double           & 310          \\
+sfdm2         &                                          & 5               & integer          & 159          \\
+adlsc         & Imputed ADL Calibrated to Surrogate      &                 & double           & 0           
+\end{tabular}
+\caption{A description of each variable in the SUPPORT dataset, taken from (REF).}
+\label{tab:my-table}
+\end{table}
+
+\begin{table}[]
+\centering
+\begin{tabular}{cc}
+\hline
+\textbf{Variable} & \textbf{Levels}                      \\
+\hline
+sex      & female                                        \\
+         & male                                          \\
+dzgroup  & ARF/MOSF w/Sepsis                             \\
+         & COPD                                          \\
+         & CHF                                           \\
+         & Cirrhosis                                     \\
+         & Coma                                          \\
+         & Colon Cancer                                  \\
+         & Lung Cancer                                   \\
+         & MOSF w/Malig                                  \\
+dzclass  & ARF/MOSF                                      \\
+         & COPD/CHF/Cirrhosis                            \\
+         & Coma                                          \\
+         & Cancer                                        \\
+income   & under \$11k                                   \\
+         & $11-$25k                                      \\
+         & $25-$50k                                      \\
+         & \textgreater{}\$50k                           \\
+race     & white                                         \\
+         & black                                         \\
+         & asian                                         \\
+         & other                                         \\
+         & hispanic                                      \\
+sfdm2    & no(M2 and SIP pres)                           \\
+         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
+         & SIP\textgreater{}=30                          \\
+         & Coma or Intub                                 \\
+         & \textless{}2 mo. follow-up                   
+\end{tabular}
+\caption{A description of each level within each categorical variable in the dataset, taken from REF.}
+\label{tab:my-table}
+\end{table}
+
+Hospital death as a covariate was removed, as this is directly informative of death.
+The glmnet package(REF) on CRAN was used to introduce elastic net, lasso and ridge regressions to casebase.
+To visualize the effect of imputation on incidence density, we can refer to the pre-imputation population-time plot in Figure \ref{fig:ptPreI} and post-imputation population time plot in Figure \ref{fig:ptPostI}.
+The incidence density is consistent between imputed and complete datasets.
+
+```{r supportSetUp, echo = FALSE, eval = eval_cs3}
+knitr::opts_chunk$set(echo = TRUE)
+#library(casebase)
+
+set.seed(2)#AT RISK INDIVIDUAL
+#set.seed(3)#less risk individual
+#read and remove ID column
+support <- read_tsv("support.tsv") 
+#impute without death, to avoid correlations
+sup.impute <-mice(support[,-c(2)], m = 1, maxit = 10,printFlag=FALSE)
+#take a complete dataset, Only one is being made and used here.
+#The logic being we only want to compare between models. 
+#Any Bias introduced is introduced to all models.
+completeData<-complete(sup.impute,1)
+#put death back into the complete dataset, as it was removed.
+completeData$death<-support$death
+#make everything into factors and remove observations without
+#imputed values.
+completeData<-na.omit(as.data.frame(unclass(completeData)))
+completeData<-completeData[,-c(3)]
+#keep one individual out, at random, for which we will develop an 
+#absolute risk curve.
+sam=sample(1:nrow(completeData), 1)
+#Set workingData to the remaining individuals in the data
+#workingData=completeData[-c(sam),]
+workingCompleteData=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))[,-c(1)]
+#Create u and xCox, which will be used when fitting Cox with glmnet.
+newData=workingCompleteData[sam,]
+x=workingCompleteData[-c(sam),]
+#x and y will be used to fit the casebase model under glmnet.
+#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
+y=data.matrix(completeData[-c(sam),c(34,4)])
+```
+
+```{r poptimePreImputation, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:ptPreI} Population-time plot before imputation.
+Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study."}
+#demonstrating incidence density before imputation
+plot(popTime(as.data.frame(support),time = "d.time",event = "death"))
+```
+
+```{r poptimePostImputation, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:ptPostI} Population-time plot before imputation.
+Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study."}
+#demonstrating incidence density after imputation
+plot(popTime(as.data.frame(workingCompleteData),time = "d.time",event = "death"))
+```
+Our main objective is to compute the absolute risk of death for a given set of covariates, using regularization when fitting our model.
+First, we fit a smooth hazard to the data; for the sake of this example, we opted for a linear term for time.
+As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model.
+For the SUPPORT dataset, all factor variables were converted into dummy variables.
+Here, we used lasso penalization by setting alpha to 1.
+```{r CBhazard, echo = TRUE, eval = eval_cs3}
+#Using casebase's glmnet implementation, using splines and a case to base ratio of 100,
+#to determine the hazard function
+cb.Model=fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
+                             formula_time = ~d.time,alpha=1,ratio=100)
+```
+Then, we take our model and use the absoluteRisk function to integrate and retrieve our absolute risk function.
+```{r CBabsolute, echo = TRUE, eval = eval_cs3}
+#Estimating the absolute risk curve using the newData parameter.
+casebaseAbsolute=absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) ,
+                              s="lambda.1se",method=c("numerical"))
+```
+We will compare this curve to a regularized version of cox regression, and a kaplan-meier survival curve.
+The regularized cox regression absolute risk curve required two extra steps to retrive.
+Coxnet from the glmnet package has tools to fit a regularized cox model, but requires that a survival object skeleton of the selected variables, so that the survival package tools can be used to retrive the absolute risk.
+We handle this by fitting a regularized cox model with the glmnet package, and a cox model from the survival package with the selected variables from the regularized fit.
+This second model serves as a skeleton that permits the use of survival package functions.
+The coefficients from the regularized model will replace the coeficients in our skeleton.
+the resulting model will have the correct coefficients when integrating, but will have an incorrect standard error.
+For the purposes of this case study, we are only interested in the absolute risk curve itself and not the standard error.
+
+```{r coxHazAbsolute, echo = TRUE, eval = eval_cs3}
+#Create u and xCox, which will be used when fitting Cox with glmnet.
+u=Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
+xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))[-c(sam),]
+#hazard for cox using glmnet
+coxglmFit=cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
+#convergence demonstrated in plot
+plot(coxglmFit)
+#taking the coefficient estimates for later use
+nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
+nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
+#creating a new dataset that only contains the covariates chosen through glmnet.
+#cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
+cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
+#newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
+#fitting a cox model using regular estimation, however we will not keep it.
+#this is used more as an object place holder.
+coxFit <- coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
+#The coefficients of this object will be replaced with the estimates from coxglmFit.
+#Doing so makes it so that everything is invalid aside from the coefficients.
+#In this case, all we need to estimate the absolute risk is the coefficients.
+#Std. error would be incorrect here, if we were to draw error bars.
+coxFit$coefficients<-nonzero_coef_cox@x
+#Fitting absolute risk curve for cox+glmnet
+newDataCox=newData[nonzero_covariate_cox$X1-1]
+
+abCoxFit<-survfit(coxFit,newdata=as.data.frame(t(newDataCox)),time = seq(0,max(completeData$d.time), 1),type="breslow")
+```
+We used the survival package to calculate the kaplan-meier absolute risk function. 
+```{r fitKM+abRisk, echo = TRUE, eval = eval_cs3}
+#creating a surv object to be used to fit an unadjusted absolute risk curve.
+# (Kaplan-Meier risk curve)
+km <- Surv(time = completeData$d.time, event = completeData$death)
+abKm<-survfit(km~1,type='kaplan-meier',conf.type='log')
+```
+
+Now that our three absolute risk functions have been calculated, we compare them all in Figure \ref{fig:abSC}.
+Both coxnet and casebase decrease the absolute risk by the end of the study, in comparison to kaplan-meier.
+```{r abSupportComparison, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
+# A plot to compare all three Absolute risk (Commulative Incidence)
+plot(casebaseAbsolute,type='l',col="black",lwd=2,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
+lines(abCoxFit ,col="red",fun="event",lwd=3,conf.int = FALSE)
+lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
+legend("bottom", 
+       legend = c( "Casebase (Lasso+linear)","semi-parametric (Cox)+glmnet","KM curve"), 
+       col = c("black","red","Blue"),
+       lty = c(1, 1, 1), 
+       bg = "gray90")
+```
 # Discussion
-
-In this article, we presented the \proglang{R} package \pkg{casebase} that provides functions to fit smooth parametric hazards and estimate cumulative incidence functions using case-base sampling.
-We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through three case studies.
 
@@
 The current Git commit details are:
-
 ```{r}

### 2019-12-06 analysis/paper/paper.tex
currentDraftEdits
added lit-review and placeholder analysis (this will change once casebase is properly updated)

--- 2019-12-05 analysis/paper/paper.tex
@@
 \author{
-Sahir Bhatnagar *\\McGill University \And Maxime Turgeon *\\University of Manitoba \AND Jesse Islam\\McGill University \And James Hanley\\McGill University \And Olli Saarela\\University of Toronto
+Sahir Bhatnagar *\\McGill Univeristy \And Maxime Turgeon *\\University of Manitoba \And Jesse Islam\\McGill University \And James Hanley\\McGill Univeristy \And Olli Saarela\\University of Toronto
 }
@@
     Sahir Bhatnagar *\\
-  McGill University\\
+  McGill Univeristy\\
   1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
@@
   University of Manitoba\\
-  186 Dysart Road Winnipeg, MB, Canada R3T 2N2\\
+  318 Machray Hall Winnipeg, MB, Canada R3T 2N2\\
   E-mail: \email{max.turgeon@umanitoba.ca}\\
@@
       James Hanley\\
-  McGill University\\
+  McGill Univeristy\\
   1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
@@
 
-\usepackage{amsmath} \usepackage{longtable}
+\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx}
 
 \begin{document}
+
+\hypertarget{code-formatting}{%
+\section{Code formatting}\label{code-formatting}}
+
+Don't use markdown, instead use the more precise latex commands:
+
+\begin{itemize}
+\item
+  \proglang{Java}
+\item
+  \pkg{plyr}
+\item
+  \code{print("abc")}
+\end{itemize}
 
@@
 
-The purpose of the \pkg{casebase} package is to provide practitioners
+The purpose of the \texttt{casebase} package is to provide practitioners
 with an easy-to-use software tool to predict the risk (or cumulative
@@
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Time matching/risk set sampling (including Cox partial likelihood)
-  eliminates the baseline hazard from the likelihood expression for the
-  hazard ratios
-\item
-  If, however, the absolute risks are of interest, they have to be
-  recovered using the semi-parametric Breslow estimator
-\item
-  Alternative approaches for fitting flexible hazard models for
-  estimating absolute risks, not requiring this two-step approach? Yes!
+hazard ratios 2. If, however, the absolute risks are of interest, they
+have to be recovered using the semi-parametric Breslow estimator 3.
+Alternative approaches for fitting flexible hazard models for estimating
+absolute risks, not requiring this two-step approach? Yes!
 \citep{hanley2009fitting}
-\end{enumerate}
 
@@
 
+\hypertarget{literature-review}{%
+\section{Literature review}\label{literature-review}}
+
+To conduct our review on current methods available in survival analysis,
+a scraper was coded to examine all packages in the survival task view on
+CRAN (ref). There was a total of 259 packages, many of which are not
+relevant to the problems casebase addresses. To filter these packages, a
+scraper was used to search for competing risks, cumulative incidence
+functions, non- proportional hazards and penalization. This reduced our
+set of 259 packages down to 60. From here, we manually examined each
+package to determine which ones are comparable to casebase, with either
+a different approach to the same problems or future features we would
+like to add to casebase.
+
+These packages can belong to three main categories: non-proportional
+hazards, regularization and competing risks. Besides these three main
+categories, there are extra features of interest as well. Table XXX
+summarizes the findings.
+
+Many packages exist that implement semi-parametric cox models. Of those,
+glmnet, glmpath, penalized, RiskRegression introduce penalization in
+their model fitting. CoxRidge, rstpm2 and survival include penalization,
+but they implement the extended cox model permitting intuitive non
+proportional hazard modeling. In terms of penalization, the benefit of
+casebase comes from its flexibility. Linear models have a wide tool set,
+considerably larger than those made for cox regression (reference for
+number of packages between their task views, regression is spread out
+across Multivariate, SocialSciences, Robust among others. The focus of
+this task view is on Optimization Infrastructure Packages , General
+Purpose Continuous Solvers , Mathematical Programming Solvers , and
+Specific Applications in Optimization , or Multi Objective Optimization
+). Casebase permits many regularization tools developed for linear
+regression to be repurposed within a survival context.
+
+Parametric models permit non-proportional hazards and are implemented in
+a handful of packages. Of interest are CFC, flexsurv, SmoothHazard,
+Rsptm2 and Survival. Smooth hazard is limited to weibull distributions,
+where as flexsurv and survival contain implementations for users to
+supply any distribution they would like. Flexsurv, smooth hazard, mets
+and rstpm2 can use splines, permitting a flexible fit over time. Though
+casebase is limited to 3 distributions at the moment, it also uses
+splines allowing for distributions that may not be consistent over time.
+
+Of the methods mentioned so far, only CFC, survival, mets and flexsurv
+contain implementations for competing risks(how is it done for
+parametric models?). The differentiating factor between these packages
+and casebase is that casebase handles competing risks through multiple
+logistic regression.
+
+A feature in Smooth hazard, survival and Rstpm2 are they they can handle
+interval censoring, where as casebase currently can not.
+
+Of all the packages listed in the table, CFC, Flexsurv, mets,
+RiskRegression, rstpm2 and survival have implementations for cumulative
+incidence. Since casebase is parametric, the baseline hazard is known
+and requires integration to retrieve the cumulative incidence. This is
+similar to the approaches of CFC, flexsurv,\\
+
+\begin{table}[]
+\centering
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{ccccccccc}
+\hline
+\textbf{Package}        & \textbf{Competing Risks} & \textbf{Non-proportional} & \textbf{Penalization} & \textbf{Splines} & \textbf{Parametric} & \textbf{Semi-parametric} & \textbf{Interval/left Censoring} & \textbf{Absolute Risk}           \\
+\hline
+\textbf{casebase}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
+\textbf{CFC}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
+\textbf{coxRidge}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{crrp}           & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
+\textbf{fastcox}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{Flexrsurv}      &                          & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
+\\
+\textbf{Flexsurv}       & x                        & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
+\\
+\textbf{glmnet}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{glmpath}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{mets}           & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
+\textbf{penalized}      &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{RiskRegression} &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
+\textbf{Rstpm2}         &                          & x                         & x                     & x                & x                   & x                        & x                                & Cumm haz                         \\
+\textbf{SmoothHazard}   &                          & x                         &                       & x                & x                   &                          & x                                & Cumulative transition intensity?
+\\
+\textbf{Survival}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x
+\end{tabular}%
+}
+\caption{Different features of interest in various survival packages.}
+\label{tab:my-table}
+\end{table}
+
 \hypertarget{theoretical-details}{%
@@
 
+Popular survival analysis methods, like Kaplan-Meier and Cox regression,
+rely on the notion of risk-set sampling. Case-base sampling can be seen
+as an alternative.
+
 As shown by Saarela \& Arjas \citeyearpar{saarela2015non} (and further
@@
 likelihood arising from this data-generating mechanism using the
-framework of non-homogeneous Poisson processes, we eventually reach an
+framework of non-homogenous Poisson processes, we eventually reach an
 expression where each person-moment's contribution is of the form
-\[\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},\] where \(N(t)\) is the counting
-process associated with the event of interest, \(h(t)\) is the
-corresponding hazard function, and \(\rho(t)\) is the hazard function
-for the Poisson process associated with case-base sampling. This
-parametric form suggests that we can readily estimate log-hazards of the
-form \(\log(h(t)) = g(t; X)\) using logistic regression, where each
-observation corresponds to a person moment, the function \(g(t; X)\) is
-linear in a finite number of parameters, and where we treat
-\(-\log(\rho(t))\) as an offset.
+\[\frac{\lambda(t)^{dN(t)}}{\rho(t) + \lambda(t)},\] where \(N(t)\) is
+the counting process associated with the event of interest,
+\(\lambda(t)\) is the corresponding hazard function, and \(\rho(t)\) is
+the hazard function for the Poisson process associated with case-base
+sampling. This parametric form suggests that we can readily estimate
+log-hazards of the form \(\log(\lambda(t)) = g(t; X)\) using logistic
+regression, where each observation corresponds to a person moment, the
+function \(g(t; X)\) is linear in a finite number of parameters, and
+where we treat \(\log(\rho(t))\) as an offset.
 
 In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
-suggest performing case-base sampling \emph{uniformly}, i.e.~to sample
+suggest performing case-base samping \emph{uniformly}, i.e.~to sample
 the base series uniformly from the study base. In terms of Poisson
-processes, this sampling strategy corresponds essentially to a
-time-homogeneous Poisson process with intensity equal to \(b/B\), where
-\(b\) is the number of sampled observations in the base series, and
-\(B\) is the total population-time for the study base. More complex
-examples are also available; see for example Saarela \& Arjas
+processes, this sampling strategy corresponds to a time-homogeneous
+Poisson process with intensity equal to \(b/B\), where \(b\) is the
+number of sampled observations in the base series, and \(B\) is the
+total population-time for the study base. More complex examples are also
+available; see for example Saarela \& Arjas
 \citeyearpar{saarela2015non}, where the probabilities of the sampling
@@
 
-\[ h(t;X) = \exp[g(t;X)] \] where \(t\) denotes time and \(X\), the
-individual's covariate profile. Different functions of \(t\) lead to
-different parametric hazard models. The simplest of these models is the
-one-parameter exponential distribution which is obtained by taking the
-hazard function to be constant over the range of \(t\).
-
-\[ h(t;X) = \exp(\beta_0 + \beta_1 X) \]
+\[ h(x,t) = exp[g(x,t)] \] where \(t\) denotes the numerical value
+(number of units) of a point in prognostic/prospective time and \(x\) is
+the realization of the vector \(X\) of variates based on the patient's
+profile and intervention (if any). Different functions of \(t\) lead to
+different parametric hazard models.
+
+The simplest of these models is the one-parameter exponential
+distribution which is obtained by taking the hazard function to be
+constant over the range of \(t\).
+
+\[ h(x,t) = exp(\beta_0 + \beta_1 x) \]
 
@@
 conditional chance of failure in a time interval of specified length is
-the same regardless of how long the individual has been in the study.
-This is also known as the \emph{memoryless property} (Kalbfleisch and
-Prentice, 2002).
+the same regardless of how long the individual has been on study a.k.a
+the memoryless property (Kalbfleisch and Prentice, 2002).
 
@@
 
-\[ h(t;X)  = \exp(\beta_0 + \beta_1 t + \beta_2 X) \]
-
-Use of \(\log(t)\) yields the Weibull hazard which allows for a power
+\[ h(x,t)  = exp(\beta_0 + \beta_1 t + \beta_2 x) \]
+
+Use of \(log(t)\) yields the Weibull hazard which allows for a power
 dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
 
-\[ h(t;X)  = \exp(\beta_0 + \beta_1 \log(t) + \beta_2 X) \]
-
-For competing-risk analyses with \(J\) possible events, we can show that
-each person-moment's contribution of the likelihood is of the form
-
-\[\frac{h_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^Jh_j(t)},\]
-
-where \(N_j(t)\) is the counting process associated with the event of
-type \(j\) and \(h_j(t)\) is the corresponding hazard function. As may
-be expected, this functional form is similar to the terms appearing in
-the likelihood function for multinomial
-regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1\end{align*}}
+\[ h(x, t)  = exp(\beta_0 + \beta_1 \log(t) + \beta_2 x) \]
 
@@
 categories: 1) data visualization, in the form of population-time plots;
-and 2) parametric modeling. We explicitly aimed at being compatible with
-both \code{data.frame}s and \code{data.table}s. This is evident in some
-of the coding choices we made, and it is also reflected in our unit
-tests.
+and 2) data analysis.
+
+We explicitely aimed at being compatible with both \code{data.frame}s
+and \code{data.table}s. This is evident in some of the coding choices we
+made, and it is also reflected in our testing units.
 
@@
 
-\hypertarget{parametric-modeling}{%
-\subsection{Parametric modeling}\label{parametric-modeling}}
-
-The parametric modeling step was separated into three parts:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  case-base sampling;
-\item
-  estimation of the smooth hazard function;
-\item
-  calculation of the risk function.
-\end{enumerate}
-
-By separating the sampling and estimation functions, we allowed the
-possibility of users implementing more complex sampling scheme, as
-described in Saarela \citeyearpar{saarela2016case}.
+\hypertarget{data-analysis}{%
+\subsection{Data analysis}\label{data-analysis}}
+
+The data analysis step was separated into three parts: 1) case-base
+sampling; 2) estimation of the smooth hazard function; and 3)
+calculation of the risk function. By separating the sampling and
+estimation functions, we allowed the possibility of users implementing
+more complex sampling scheme, as described in Saarela
+\citeyearpar{saarela2016case}.
 
@@
 
-To fit the hazard function, we provide several approaches that are
-available via the \code{family} parameter. These approaches are:
-
-\begin{itemize}
-\tightlist
-\item
-  \code{glm}: This is the familiar logistic regression.
-\item
-  \code{glmnet}: This option allows for variable selection using Lasso
-  or elastic-net. This functionality is provided through the
-  \code{glmnet} package \citep{friedman2010jss}.
-\item
-  \code{gam}: This option provides support for \emph{Generalized
-  Additive Models} via the \code{gam} package
-  \citep{hastie1987generalized}.
-\item
-  \code{gbm}: This option provides support for \emph{Gradient Boosted
-  Trees} via the \code{gbm} package. This feature is still experimental.
-\end{itemize}
-
-In the case of multiple events, the hazard is fitted via multinomial
-regression as performed by the \pkg{VGAM} package. This package was
+Finally, the hazard function is fitted to the data using the function
+\code{glm}, unless there is more than one type event (i.e.~in a
+competing-risk analysis), in which case it uses the multinomical
+regression capabilities of the \pkg{VGAM} package. This package was
 selected for its ability to fit multinomial regression models with an
@@
 is computed using either the \code{stats::integrate} function or
-Monte-Carlo integration. The risk function (or cumulative incidence
-function) is then defined as
-
-\[ CI(t) = 1 - S(t).\]
-
-For the case of a competing-event analysis, the event-specific risk is
-computed using the following procedure: first, we compute the overall
-survival function (i.e.~for all event types):
-
-\[ S(t) = \exp\left(-\int_0^t H(u;X) du\right),\qquad H(t;X) = \sum_{j=1}^J h_j(t;X).\]
-From this, we can derive the event-specific subdensities:
-
-\[ f_j(t) = h_j(t)S(t).\]
-
-Finally, by integrating these subdensities, we obtain the event-specific
-cumulative incidence functions:
-
-\[ CI_j(t) = \int_0^t f_j(u)du.\]
-
-We created \code{absoluteRisk} as an \code{S3} generic, with methods for
-the different types of outputs of \code{fitSmoothHazard}. The method
-dispatch system of \proglang{R} then takes care of matching the correct
-output to the correct methodology for calculating the cumulative
-incidence function, without the user's intervention.
-
-In the following sections, we illustrate these functionalities in the
-context of three case studies.
+Monte-Carlo integration. For the case of a competing-event analysis, the
+event-specific risk is computed using a nested double integral; in this
+setting, Monte-Carlo integration is faster and more flexible than
+\code{stats::integrate}. This is due to the fact that the computation
+involves a double loop over the selected time points; Monte-Carlo
+integration performs this step by using the vectorized \code{rowSums}
+and \code{colSums} functions.
+
+To decide between a single-event and a competing-event analysis, we
+created \code{absoluteRisk} as an \code{S3} generic, with methods for
+both \code{glm} and \code{CompRisk} objects (the latter inherits from
+\code{vglm} as well). The method dispatch system of \proglang{R} then
+takes care of matching the correct input to the correct methodology,
+without the user's intervention.
 
@@
 
-\[ CI(x, t) = 1 - \exp\left( - \int_0^t h(x, u) \textrm{d}u \right) \]
+\[ CI(x, t) = 1 - exp\left( - \int_0^t h(x, u) \textrm{d}u \right) \]
 
@@
 \begin{CodeInput}
-R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
+R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ splines::bs(Follow.Up.Time) + ScrArm, 
 R>                                     data = ERSPC, 
@@
 
-As expected, we see that splines model provides a better fit.
+As espected, we see that splines model provides a better fit.
 
@@
 
-\hypertarget{case-study-3the-cancer-genome-atlas}{%
-\section{Case study 3--The Cancer Genome
-Atlas}\label{case-study-3the-cancer-genome-atlas}}
-
-\begin{itemize}
-\tightlist
-\item
-  Glmnet and TCGA data
-\end{itemize}
-
-\hypertarget{case-study-4stanford-heart-transplant-data}{%
-\section{Case study 4--Stanford Heart Transplant
-Data}\label{case-study-4stanford-heart-transplant-data}}
-
-Although the previous case studies provide a broad overview of the
-capabilities of \pkg{casebase}, they all have two properties in common:
-
-\begin{itemize}
-  \item All covariates are fixed, i.e. they do not change over time;
-  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
-\end{itemize}
-
-The purpose of the next case study is to show how \pkg{casebase} can
-also be used for survival analysis problems that do not share these
-properties. To this end, we will use the Stanford Heart Transplant Data
-\citep[\citet{crowley1977covariance}]{clark1971cardiac}; we will use the
-version available in the \pkg{survival} package.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(survival)
-R> library(casebase)
+\hypertarget{case-study-3study-to-understand-prognoses-preferences-outcomes-and-risks-of-treatment}{%
+\section{Case study 3--Study to Understand Prognoses Preferences
+Outcomes and Risks of
+Treatment}\label{case-study-3study-to-understand-prognoses-preferences-outcomes-and-risks-of-treatment}}
+
+We examined the Study to Understand Prognoses Preferences Outcomes and
+Risks of Treatment (SUPPORT) dataset (REF from main site). the SUPPORT
+dataset tracks death for individuals who are considered seriously ill
+within a hospital. Imputation was conducted using the default methods in
+the mice package in R, available on CRAN. Before imputation, there was a
+total of 1000 individuals and 35 variables. After imputation, we have
+474 observations where 67.5\% of individuals in the study experience the
+event of interest. The SUPPORT dataset will be used to demonstrate
+regularization within casebase (REF), while comparing their absolute
+risk preditions for a new individual. A description of each variable can
+be found in table XXX1 and a breakdown of each categorical variable in
+table XXX2. The data was downloaded from the main author's website, and
+it is also available as part of the casebase package.
+
+\begin{table}[]
+\centering
+\begin{tabular}{ccccc}
+\hline
+\textbf{Name} & \textbf{Labels}                          & \textbf{Levels} & \textbf{Storage} & \textbf{NAs} \\
+\hline
+age           & Age                                      &                 & double           & 0            \\
+death         & Death at any time up to NDI date:31DEC94 &                 & double           & 0            \\
+sex           &                                          & 2               & integer          & 0            \\
+hospdead      & Death in Hospital                        &                 & double           & 0            \\
+slos          & Days from Study Entry to Discharge       &                 & double           & 0            \\
+d.time        & Days of Follow-Up                        &                 & double           & 0            \\
+dzgroup       &                                          & 8               & integer          & 0            \\
+dzclass       &                                          & 4               & integer          & 0            \\
+num.co        & number of comorbidities                  &                 & double           & 0            \\
+edu           & Years of Education                       &                 & double           & 202          \\
+income        &                                          & 4               & integer          & 349          \\
+scoma         & SUPPORT Coma Score based on Glasgow D3   &                 & double           & 0            \\
+charges       & Hospital Charges                         &                 & double           & 25           \\
+totcst        & Total RCC cost                           &                 & double           & 105          \\
+totmcst       & Total micro-cost                         &                 & double           & 372          \\
+avtisst       & Average TISS, Days 3-25                  &                 & double           & 6            \\
+race          &                                          & 5               & integer          & 5            \\
+meanbp        & Mean Arterial Blood Pressure Day 3       &                 & double           & 0            \\
+wblc          & White Blood Cell Count Day 3             &                 & double           & 24           \\
+hrt           & Heart Rate Day 3                         &                 & double           & 0            \\
+resp          & Respiration Rate Day 3                   &                 & double           & 0            \\
+temp          & Temperature (celcius) Day 3              &                 & double           & 0            \\
+pafi          & PaO2/(.01*FiO2) Day 3                    &                 & double           & 253          \\
+alb           & Serum Albumin Day 3                      &                 & double           & 378          \\
+bili          & Bilirubin Day 3                          &                 & double           & 297          \\
+crea          & Serum creatinine Day 3                   &                 & double           & 3            \\
+sod           & Serum sodium Day 3                       &                 & double           & 0            \\
+ph            & Serum pH (arterial) Day 3                &                 & double           & 250          \\
+glucose       & Glucose Day 3                            &                 & double           & 470          \\
+bun           & BUN Day 3                                &                 & double           & 455          \\
+urine         & Urine Output Day 3                       &                 & double           & 517          \\
+adlp          & ADL Patient Day 3                        &                 & double           & 634          \\
+adls          & ADL Surrogate Day 3                      &                 & double           & 310          \\
+sfdm2         &                                          & 5               & integer          & 159          \\
+adlsc         & Imputed ADL Calibrated to Surrogate      &                 & double           & 0           
+\end{tabular}
+\caption{A description of each variable in the SUPPORT dataset, taken from (REF).}
+\label{tab:my-table}
+\end{table}
+
+\begin{table}[]
+\centering
+\begin{tabular}{cc}
+\hline
+\textbf{Variable} & \textbf{Levels}                      \\
+\hline
+sex      & female                                        \\
+         & male                                          \\
+dzgroup  & ARF/MOSF w/Sepsis                             \\
+         & COPD                                          \\
+         & CHF                                           \\
+         & Cirrhosis                                     \\
+         & Coma                                          \\
+         & Colon Cancer                                  \\
+         & Lung Cancer                                   \\
+         & MOSF w/Malig                                  \\
+dzclass  & ARF/MOSF                                      \\
+         & COPD/CHF/Cirrhosis                            \\
+         & Coma                                          \\
+         & Cancer                                        \\
+income   & under \$11k                                   \\
+         & $11-$25k                                      \\
+         & $25-$50k                                      \\
+         & \textgreater{}\$50k                           \\
+race     & white                                         \\
+         & black                                         \\
+         & asian                                         \\
+         & other                                         \\
+         & hispanic                                      \\
+sfdm2    & no(M2 and SIP pres)                           \\
+         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
+         & SIP\textgreater{}=30                          \\
+         & Coma or Intub                                 \\
+         & \textless{}2 mo. follow-up                   
+\end{tabular}
+\caption{A description of each level within each categorical variable in the dataset, taken from REF.}
+\label{tab:my-table}
+\end{table}
+
+Hospital death as a covariate was removed, as this is directly
+informative of death. The glmnet package(REF) on CRAN was used to
+introduce elastic net, lasso and ridge regressions to casebase. To
+visualize the effect of imputation on incidence density, we can refer to
+the pre-imputation population-time plot in Figure \ref{fig:ptPreI} and
+post-imputation population time plot in Figure \ref{fig:ptPostI}. The
+incidence density is consistent between imputed and complete datasets.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #demonstrating incidence density before imputation
+R> plot(popTime(as.data.frame(support),time = "d.time",event = "death"))
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #demonstrating incidence density after imputation
+R> plot(popTime(as.data.frame(workingCompleteData),time = "d.time",event = "death"))
+\end{CodeInput}
+\end{CodeChunk}
+
+Our main objective is to compute the absolute risk of death for a given
+set of covariates, using regularization when fitting our model. First,
+we fit a smooth hazard to the data; for the sake of this example, we
+opted for a linear term for time. As casebase makes use of the glmnet
+package, we interact with the fitsmoothhazard.fit function using a
+matrix interface, where y contains the time and event variables and x
+contains all other variables we would like to include in the model. For
+the SUPPORT dataset, all factor variables were converted into dummy
+variables. Here, we used lasso penalization by setting alpha to 1.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #Using casebase's glmnet implementation, using splines and a case to base ratio of 100,
+R> #to determine the hazard function
+R> cb.Model=fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
+R>                              formula_time = ~d.time,alpha=1,ratio=100)
+\end{CodeInput}
+\end{CodeChunk}
+
+Then, we take our model and use the absoluteRisk function to integrate
+and retrieve our absolute risk function.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #Estimating the absolute risk curve using the newData parameter.
+R> casebaseAbsolute=absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) ,
+R>                               s="lambda.1se",method=c("numerical"))
+\end{CodeInput}
+\end{CodeChunk}
+
+We will compare this curve to a regularized version of cox regression,
+and a kaplan-meier survival curve. The regularized cox regression
+absolute risk curve required two extra steps to retrive. Coxnet from the
+glmnet package has tools to fit a regularized cox model, but requires
+that a survival object skeleton of the selected variables, so that the
+survival package tools can be used to retrive the absolute risk. We
+handle this by fitting a regularized cox model with the glmnet package,
+and a cox model from the survival package with the selected variables
+from the regularized fit. This second model serves as a skeleton that
+permits the use of survival package functions. The coefficients from the
+regularized model will replace the coeficients in our skeleton. the
+resulting model will have the correct coefficients when integrating, but
+will have an incorrect standard error. For the purposes of this case
+study, we are only interested in the absolute risk curve itself and not
+the standard error.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #Create u and xCox, which will be used when fitting Cox with glmnet.
+R> u=Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
+R> xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))[-c(sam),]
+R> #hazard for cox using glmnet
+R> coxglmFit=cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
+R> #convergence demonstrated in plot
+R> plot(coxglmFit)
+R> #taking the coefficient estimates for later use
+R> nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
+R> nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
+R> #creating a new dataset that only contains the covariates chosen through glmnet.
+R> #cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
+R> cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
+R> #newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
+R> #fitting a cox model using regular estimation, however we will not keep it.
+R> #this is used more as an object place holder.
+R> coxFit <- coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
+R> #The coefficients of this object will be replaced with the estimates from coxglmFit.
+R> #Doing so makes it so that everything is invalid aside from the coefficients.
+R> #In this case, all we need to estimate the absolute risk is the coefficients.
+R> #Std. error would be incorrect here, if we were to draw error bars.
+R> coxFit$coefficients<-nonzero_coef_cox@x
+R> #Fitting absolute risk curve for cox+glmnet
+R> newDataCox=newData[nonzero_covariate_cox$X1-1]
 R> 
-R> stanford_popTime <- popTime(jasa, time = "futime", 
-R+                             event = "fustat")
-R> plot(stanford_popTime)
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/stanford-poptime-1} \end{center}
-
-\end{CodeChunk}
-
-Since the exposure is time dependent, we need to manually define the
-exposure variable \emph{after} case-base sampling and \emph{before}
-fitting the hazard function. For this reason, we will use the
-\texttt{sampleCaseBase} function directly.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(tidyverse)
-R> library(lubridate)
-R> 
-R> cb_data <- sampleCaseBase(jasa, time = "futime", 
-R+                           event = "fustat", ratio = 10)
-\end{CodeInput}
-\end{CodeChunk}
-
-Next, we will compute the number of days from acceptance into the
-program to transplant, and we use this variable to determine whether
-each population-moment is exposed or not.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Define exposure variable
-R> cb_data <- mutate(cb_data,
-R+                   txtime = time_length(accept.dt %--% tx.date, 
-R+                                        unit = "days"),
-R+                   exposure = case_when(
-R+                     is.na(txtime) ~ 0L,
-R+                     txtime > futime ~ 0L,
-R+                     txtime <= futime ~ 1L
-R+                   ))
-\end{CodeInput}
-\end{CodeChunk}
-
-Finally, we can fit the hazard using various linear predictors.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(splines)
-R> # Fit several models
-R> fit1 <- fitSmoothHazard(fustat ~ exposure,
-R+                         data = cb_data, time = "futime")
-R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
-R+                         data = cb_data, time = "futime")
-R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
-R+                         data = cb_data, time = "futime")
-R> fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
-R+                         data = cb_data, time = "futime")
-\end{CodeInput}
-\end{CodeChunk}
-
-Note that the fourth model (i.e. \texttt{fit4}) includes an interaction
-term between exposure and follow-up time. In other words, this model no
-longer exhibit proportional hazards. We can then compare the goodness of
-fit of these four models using the Akaike Information Criterion (AIC).
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute AIC
-R> c("Model1" = AIC(fit1),
-R+   "Model2" = AIC(fit2),
-R+   "Model3" = AIC(fit3),
-R+   "Model4" = AIC(fit4))
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> Model1 Model2 Model3 Model4 
-#>    796    760    730    728
-\end{CodeOutput}
-\end{CodeChunk}
-
-As we can, the best fit is the fourth model. By visualizing the hazard
-functions for both exposed and unexposed individuals, we can more
-clearly see how the hazards are no longer proportional.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute hazards---
-R> # First, create a list of time points for both exposure status
-R> hazard_data <- expand.grid(exposure = c(0, 1),
-R+                            futime = seq(0, 1000,
-R+                                         length.out = 100))
-R> # Set the offset to zero
-R> hazard_data$offset <- 0 
-R> # Use predict to get the fitted values, and exponentiate to 
-R> # transform to the right scale
-R> hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
-R+                                  type = "link"))
-R> # Add labels for plots
-R> hazard_data$Status = factor(hazard_data$exposure,
-R+                             labels = c("NoTrans", "Trans"))
-R> 
-R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
-R+     geom_line() +
-R+     theme_minimal() +
-R+     theme(legend.position = 'top') +
-R+     ylab('Hazard') + xlab('Follow-up time')
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/stanford-hazard-1} \end{center}
-
-\end{CodeChunk}
-
-The non-proportionality seems to be more pronounced at the beginning of
-follow-up than the end. Finally, we can turn these estimate of the
-hazard function into estimates of the cumulative incidence functions.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute absolute risk curves
-R> newdata <- data.frame(exposure = c(0, 1))
-R> absrisk <- absoluteRisk(fit4, newdata = newdata, 
-R+                         time = seq(0, 1000, length.out = 100))
-R> 
-R> colnames(absrisk) <- c("Time", "NoTrans", "Trans")
-R> 
-R> # Rearrange the data
-R> absrisk <- gather(as.data.frame(absrisk),
-R+                   "Status", "Risk", -Time)
-R>  
-R> ggplot(absrisk, aes(Time, Risk, colour = Status)) +
-R+   geom_line() +
-R+   theme_minimal() +
-R+   theme(legend.position = 'top') +
-R+   expand_limits(y = 0.5) +
-R+   xlab('Follow-up time') + ylab('Cum. Incidence')
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/stanford-risk-1} \end{center}
-
-\end{CodeChunk}
-
-Note that we can easily adapt the code above to the situation where a
-patient receives a heart transplant at a point in time of interest, for
-example after 30 days.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute hazards---
-R> # First, create a list of time points for both exposure status
-R> one_yr_haz <- data.frame(futime = seq(0, 365,
-R+                                       length.out = 100))
-R> one_yr_haz <- mutate(one_yr_haz,
-R+                      offset = 0,
-R+                      exposure = if_else(futime < 30, 0, 1))
-R> one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
-R+                                 type = "link"))
-R> ggplot(one_yr_haz, aes(futime, hazard)) +
-R+   geom_line() +
-R+   theme_minimal() +
-R+   theme(legend.position = 'top') +
-R+   ylab('Hazard') + xlab('Follow-up time') +
-R+   geom_vline(xintercept = 30, linetype = 'dashed')
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/stanford-1y-haz-1} \end{center}
-
-\end{CodeChunk}
-
-We can then compare the 1-year mortality risk without transplant and
-with transplant at 30 days.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> absoluteRisk(fit4, newdata = data.frame(exposure = 0),
-R+              time = 365)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#>         
-#> 365 0.27
-\end{CodeOutput}
-
-\begin{CodeInput}
-R> 
-R> # Use the trapezoidal rule to estimate the cumulative hazard
-R> min_hazard <- min(one_yr_haz$hazard)
-R> max_hazard <- max(one_yr_haz$hazard)
-R> increment <- 365/99
-R> cumhaz_est <- 0.5*increment*(sum(2*one_yr_haz$hazard) - min_hazard - max_hazard)
-R> 1 - exp(-cumhaz_est)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> [1] 0.18
-\end{CodeOutput}
-\end{CodeChunk}
-
-As we can see, the risk estimate at 1-year is about 30\% lower if the
-patient receives a heart transplant at 30 days.
+R> abCoxFit<-survfit(coxFit,newdata=as.data.frame(t(newDataCox)),time = seq(0,max(completeData$d.time), 1),type="breslow")
+\end{CodeInput}
+\end{CodeChunk}
+
+We used the survival package to calculate the kaplan-meier absolute risk
+function.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #creating a surv object to be used to fit an unadjusted absolute risk curve.
+R> # (Kaplan-Meier risk curve)
+R> km <- Surv(time = completeData$d.time, event = completeData$death)
+R> abKm<-survfit(km~1,type='kaplan-meier',conf.type='log')
+\end{CodeInput}
+\end{CodeChunk}
+
+Now that our three absolute risk functions have been calculated, we
+compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase
+decrease the absolute risk by the end of the study, in comparison to
+kaplan-meier.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # A plot to compare all three Absolute risk (Commulative Incidence)
+R> plot(casebaseAbsolute,type='l',col="black",lwd=2,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
+R> lines(abCoxFit ,col="red",fun="event",lwd=3,conf.int = FALSE)
+R> lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
+R> legend("bottom", 
+R>        legend = c( "Casebase (Lasso+linear)","semi-parametric (Cox)+glmnet","KM curve"), 
+R>        col = c("black","red","Blue"),
+R>        lty = c(1, 1, 1), 
+R>        bg = "gray90")
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 
-In this article, we presented the \proglang{R} package \pkg{casebase}
-that provides functions to fit smooth parametric hazards and estimate
-cumulative incidence functions using case-base sampling. We outlined the
-theoretical underpinnings of the approach, we provided details about our
-implementation, and we illustrated the merits of the approach and the
-package through three case studies.
-
 In the following table we provide a comparison between the Cox model and
@@
 
-This report was generated on 2019-12-05 17:15:04 using the following
+This report was generated on 2019-12-04 13:59:29 using the following
 computational environment and dependencies:
@@
 #>  setting  value                       
-#>  version  R version 3.6.1 (2019-07-05)
-#>  os       Ubuntu 18.04.3 LTS          
-#>  system   x86_64, linux-gnu           
+#>  version  R version 3.6.0 (2019-04-26)
+#>  os       macOS High Sierra 10.13.6   
+#>  system   x86_64, darwin15.6.0        
 #>  ui       X11                         
-#>  language en_CA:en                    
+#>  language (EN)                        
 #>  collate  en_CA.UTF-8                 
 #>  ctype    en_CA.UTF-8                 
-#>  tz       America/Winnipeg            
-#>  date     2019-12-05                  
+#>  tz       America/Montreal            
+#>  date     2019-12-04                  
 #> 
 #> - Packages -------------------------------------------------------------------
-#>  package     * version    date       lib
+#>  ! package       * version    date       lib
+#>    abind           1.4-5      2016-07-21 [1]
 #>    assertthat      0.2.1      2019-03-21 [1]
 #>    backports       1.1.5      2019-10-02 [1]
+#>    bookdown        0.16       2019-11-22 [1]
+#>    boot            1.3-23     2019-07-05 [1]
 #>    broom           0.5.2      2019-04-07 [1]
 #>    callr           3.3.2      2019-09-22 [1]
-#>  casebase    * 0.2.1.9001 2019-08-13 [1]
+#>    car             3.0-5      2019-11-15 [1]
+#>    carData         3.0-3      2019-11-16 [1]
+#>    casebase      * 0.2.1.9001 2019-08-20 [1]
+#>  R casebasepaper * 0.0.0.9000 <NA>       [?]
 #>    cellranger      1.1.0      2016-07-27 [1]
+#>    class           7.3-15     2019-01-01 [1]
 #>    cli             1.1.0      2019-03-19 [1]
-#>  colorspace    1.4-1      2019-03-18 [1]
+#>    codetools       0.2-16     2018-12-24 [1]
+#>    colorspace    * 1.4-1      2019-03-18 [1]
 #>    crayon          1.3.4      2017-09-16 [1]
-#>  data.table    1.12.6     2019-10-18 [1]
+#>    curl            4.2        2019-09-24 [1]
+#>    data.table    * 1.12.6     2019-10-18 [1]
 #>    DBI             1.0.0      2018-05-02 [1]
 #>    dbplyr          1.4.2      2019-06-17 [1]
+#>    DEoptimR        1.0-8      2016-11-19 [1]
 #>    desc            1.2.0      2018-05-01 [1]
@@
 #>    dplyr         * 0.8.3      2019-07-04 [1]
+#>    e1071           1.7-2      2019-06-05 [1]
 #>    ellipsis        0.3.0      2019-09-20 [1]
 #>    evaluate        0.14       2019-05-28 [1]
-#>  farver        2.0.1      2019-11-13 [1]
 #>    forcats       * 0.4.0      2019-02-17 [1]
+#>    foreach         1.4.7      2019-07-27 [1]
+#>    foreign         0.8-72     2019-08-02 [1]
 #>    fs              1.3.1      2019-05-06 [1]
@@
 #>    ggplot2       * 3.2.1      2019-08-10 [1]
+#>    git2r           0.26.1     2019-06-29 [1]
+#>    glmnet        * 3.0-1      2019-11-15 [1]
 #>    glue            1.3.1      2019-03-12 [1]
@@
 #>    haven           2.2.0      2019-11-08 [1]
+#>    here            0.1        2017-05-28 [1]
 #>    hms             0.5.2      2019-10-30 [1]
@@
 #>    httr            1.4.1      2019-08-05 [1]
+#>    iterators       1.0.12     2019-07-26 [1]
+#>    jomo            2.6-10     2019-10-22 [1]
 #>    jsonlite        1.6        2018-12-07 [1]
 #>    knitr           1.26       2019-11-12 [1]
-#>  labeling      0.3        2014-08-23 [1]
-#>  lattice       0.20-38    2018-11-04 [4]
+#>    laeken          0.5.0      2019-01-10 [1]
+#>    lattice       * 0.20-38    2018-11-04 [1]
+#>    lava            1.6.6      2019-08-01 [1]
 #>    lazyeval        0.2.2      2019-03-15 [1]
 #>    lifecycle       0.1.0      2019-08-01 [1]
-#>  lubridate   * 1.7.4      2018-04-11 [1]
+#>    lme4            1.1-21     2019-03-05 [1]
+#>    lmtest          0.9-37     2019-04-30 [1]
+#>    lubridate       1.7.4      2018-04-11 [1]
 #>    magrittr      * 1.5        2014-11-22 [1]
-#>  Matrix        1.2-17     2019-03-22 [4]
+#>    MASS            7.3-51.4   2019-03-31 [1]
+#>    Matrix        * 1.2-17     2019-03-22 [1]
 #>    memoise         1.1.0      2017-04-21 [1]
-#>  mgcv          1.8-31     2019-11-09 [4]
+#>    mgcv            1.8-31     2019-11-09 [1]
+#>    mice          * 3.6.0      2019-07-10 [1]
+#>    minqa           1.2.4      2014-10-09 [1]
+#>    mitml           0.3-7      2019-01-07 [1]
 #>    modelr          0.1.5      2019-08-08 [1]
 #>    munsell         0.5.0      2018-06-12 [1]
-#>  nlme          3.1-142    2019-11-07 [4]
+#>    nlme            3.1-142    2019-11-07 [1]
+#>    nloptr          1.2.1      2018-10-03 [1]
+#>    nnet            7.3-12     2016-02-02 [1]
+#>    numDeriv        2016.8-1.1 2019-06-06 [1]
+#>    openxlsx        4.1.3      2019-11-07 [1]
+#>    pan             1.6        2018-06-29 [1]
 #>    pillar          1.4.2      2019-06-29 [1]
@@
 #>    R6              2.4.1      2019-11-12 [1]
+#>    ranger          0.11.2     2019-03-07 [1]
 #>    Rcpp            1.0.3      2019-11-08 [1]
@@
 #>    reprex          0.3.0      2019-05-16 [1]
+#>    rio             0.5.16     2018-11-26 [1]
 #>    rlang           0.4.2      2019-11-23 [1]
-#>  rmarkdown     1.18       2019-11-27 [1]
+#>    rmarkdown       1.17       2019-11-13 [1]
+#>    robustbase      0.93-5     2019-05-12 [1]
+#>    rpart           4.1-15     2019-04-12 [1]
 #>    rprojroot       1.3-2      2018-01-03 [1]
@@
 #>    sessioninfo     1.1.1      2018-11-05 [1]
+#>    shape           1.4.4      2018-02-07 [1]
+#>    sp              1.3-2      2019-11-07 [1]
 #>    stringi         1.4.3      2019-03-12 [1]
 #>    stringr       * 1.4.0      2019-02-10 [1]
-#>  survival    * 2.44-1.1   2019-04-01 [4]
-#>  testthat      2.3.1      2019-12-01 [1]
+#>    survival      * 3.1-7      2019-11-09 [1]
+#>    testthat        2.3.0      2019-11-05 [1]
 #>    tibble        * 2.1.3      2019-06-06 [1]
@@
 #>    tidyverse     * 1.3.0      2019-11-21 [1]
+#>    timereg         1.9.4      2019-07-30 [1]
 #>    usethis         1.5.1      2019-07-04 [1]
+#>    vcd             1.4-4      2017-12-06 [1]
 #>    vctrs           0.2.0      2019-07-05 [1]
 #>    VGAM            1.1-2      2019-11-21 [1]
+#>    VIM           * 4.8.0      2019-02-11 [1]
 #>    withr           2.1.2      2018-03-15 [1]
@@
 #>    xml2            1.2.2      2019-08-09 [1]
+#>    xtable          1.8-4      2019-04-21 [1]
 #>    yaml            2.2.0      2018-07-25 [1]
 #>    zeallot         0.1.0      2018-01-28 [1]
+#>    zip             2.0.4      2019-09-01 [1]
+#>    zoo             1.8-6      2019-05-28 [1]
 #>  source                                  
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
 #>  Github (sahirbhatnagar/casebase@aecb816)
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
+#>  <NA>                                    
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
 #> 
-#> [1] /home/turgeonm/R/x86_64-pc-linux-gnu-library/3.6
-#> [2] /usr/local/lib/R/site-library
-#> [3] /usr/lib/R/site-library
-#> [4] /usr/lib/R/library
+#> [1] /Library/Frameworks/R.framework/Versions/3.6/Resources/library
+#> 
+#>  R -- Package was removed from disk.
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> Local:    review-max /home/turgeonm/Documents/git_repos/cbpaper
-#> Head:     [77f5e41] 2019-08-02: fix overlapping author names
+#> Local:    master /Users/jislam/Desktop/Gits/cbpaper
+#> Remote:   master @ origin (https://github.com/sahirbhatnagar/cbpaper.git)
+#> Head:     [5022c53] 2019-12-04: rough litreview and casestudy 3 added
 \end{CodeOutput}

### 2019-12-06 analysis/paper/paper.Rmd
Lit-Review Edits
TODO: fix table references in lit review

debug case study 3 after casebase pull request complete

@@
 library(splines)
-library(readr)
-library(mice)
-library(VIM)
-library(tibble)
+#library(readr)
+#library(tibble)
 library(glmnet)
-library(ggplot2)
-devtools::load_all()
 options(digits = 2, scipen = 999)
@@
 eval_cs2 <- FALSE
-eval_cs3 <- TRUE
+eval_cs3 <- FALSE
 ```
@@
 # Literature review
-To conduct our review on current methods available in survival analysis, a scraper was coded to examine all packages in the survival task view on CRAN (ref).
+To conduct our review on current methods available in survival analysis, a scraper was coded to examine all packages in the survival task view on CRAN [-@survTaskView].
 There was a total of 259 packages, many of which are not relevant to the problems casebase addresses.
@@
 Many packages exist that implement semi-parametric cox models.
-Of those, glmnet, glmpath, penalized, RiskRegression introduce penalization in their model fitting.
-CoxRidge, rstpm2 and survival include penalization, but they implement the extended cox model permitting intuitive non proportional hazard modeling.
+Of those, glmnet [-@regpathcox]???, glmpath [-@park_hastie], penalized [-@l1penal], RiskRegression [-@gerds_blanche] introduce penalization in their model fitting.
+CoxRidge [-@perperoglou] , rstpm2 [-@clements_liu] and survival [-@survival-package] include penalization, but they implement the extended cox model permitting intuitive non proportional hazard modeling while the previous four do not.
 In terms of penalization, the benefit of casebase comes from its flexibility.
-Linear models have a wide tool set, considerably larger than those made for cox regression (reference for number of packages between their task views, regression is spread out across Multivariate, SocialSciences, Robust among others.
-The focus of this task view is on Optimization Infrastructure Packages , General Purpose Continuous Solvers , Mathematical Programming Solvers , and Specific Applications in Optimization , or Multi Objective Optimization ).
+Linear models have a wide tool set, considerably larger than those made for cox regression (*reference for number of packages between their task views, regression is spread out across Multivariate, SocialSciences, Robust among others.
+The focus of this task view is on Optimization Infrastructure Packages , General Purpose Continuous Solvers , Mathematical Programming Solvers , and Specific Applications in Optimization , or Multi Objective Optimization*).
 Casebase permits many regularization tools developed for linear regression to be repurposed within a survival context.
@@
 Parametric models permit non-proportional hazards and are implemented in a handful of packages.
-Of interest are CFC, flexsurv, SmoothHazard, Rsptm2 and Survival.
-Smooth hazard is limited to weibull distributions, where as flexsurv and survival contain implementations for users to supply any distribution they would like.
-Flexsurv, smooth hazard, mets and rstpm2 can use splines, permitting a flexible fit over time.
-Though casebase is limited to 3 distributions at the moment, it also uses splines allowing for distributions that may not be consistent over time.
+Of interest are CFC [-@sharabiani_mahani_2019], flexsurv [-@flexsurv] , SmoothHazard [-@smoothHazard], Rsptm2 [-@clements_liu] and Survival [-@survival-package].
+Smooth hazard  is limited to weibull distributions[-@smoothHazard], where as flexsurv and survival contain implementations for users to supply any distribution they would like [-@survival-package], [-@flexsurv].
+Flexsurv, smooth hazard, mets and rstpm2 can use splines, permitting a flexible fit over time [-@flexsurv], [-@smoothHazard], [-@scheike2014estimating], [-@clements_liu] .
+Though casebase is limited to 3 distributions at the moment, it also uses splines allowing for distributions that may not be consistent over survival time.
 
- Of the methods mentioned so far, only CFC, survival, mets and flexsurv contain implementations for competing risks(how is it done for parametric models?).
+Of the methods mentioned so far, only CFC,flexsurv, mets and survival contain implementations for competing risks [-@sharabiani_mahani_2019], [-@flexsurv],[-@scheike2014estimating] ,[-@survival-package] (*how is it done for parametric models?*).
 The differentiating factor between these packages and casebase is that casebase handles competing risks through multiple logistic regression.
+Smooth hazard, survival and Rstpm2 can handle interval censoring, where as casebase currently can not.
+Interval censoring is simple feature to implement and is a potential improvement to add to casebase.
+(There's also multistate models, which I haven't mentioned in this paper)
 
-A feature in Smooth hazard, survival and Rstpm2 are they they can handle interval censoring, where as casebase currently can not.
+Of all the packages listed in the table, CFC , Flexsurv, mets, RiskRegression, rstpm2 and survival have implementations for cumulative incidence [-@sharabiani_mahani_2019], [-@flexsurv], [-@scheike2014estimating], [-@gerds_blanche], [-@clements_liu], [-@survival-package].
+Since casebase is parametric, the baseline hazard is known and requires integration to retrieve the cumulative incidence.
+This is similar to the approaches of CFC and flexsurv [-@sharabiani_mahani_2019], [-@flexsurv], while the other packages use semi-parametric approaches to retrieve the baseline after fitting [-@scheike2014estimating], [-@gerds_blanche].
+rstpm2 and survival use both for their respective parametric and semi-parametric model implementations [-@clements_liu], [-@survival-package].
 
-Of all the packages listed in the table, CFC, Flexsurv, mets, RiskRegression, rstpm2 and survival have implementations for cumulative incidence.
-Since casebase is parametric, the baseline hazard is known and requires integration to retrieve the cumulative incidence.
-This is similar to the approaches of CFC, flexsurv,
 \begin{table}[]
@@
 \hline
-\textbf{casebase}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
-\textbf{CFC}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
-\textbf{coxRidge}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{crrp}           & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
-\textbf{fastcox}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{Flexrsurv}      &                          & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
+\textbf{casebase} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
+\textbf{CFC} \cite{sharabiani_mahani_2019}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
+\textbf{coxRidge} \cite{perperoglou}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{crrp} \cite{fu}           & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
+\textbf{fastcox} \cite{yi_zou}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{Flexrsurv} \cite{flexrsurv}      &                          & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
 \\
-\textbf{Flexsurv}       & x                        & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
+\textbf{Flexsurv} \cite{flexsurv}       & x                        & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
 \\
-\textbf{glmnet}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{glmpath}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{mets}           & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
-\textbf{penalized}      &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{RiskRegression} &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
-\textbf{Rstpm2}         &                          & x                         & x                     & x                & x                   & x                        & x                                & Cumm haz                         \\
-\textbf{SmoothHazard}   &                          & x                         &                       & x                & x                   &                          & x                                & Cumulative transition intensity?
-\\
-\textbf{Survival}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x
+\textbf{glmnet} \cite{regpathcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{glmpath} \cite{park_hastie}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{mets} \cite{scheike2014estimating}           & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
+\textbf{penalized} \cite{goeman_meijer2019}      &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{RiskRegression} \cite{gerds_blanche} &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
+\textbf{Rstpm2} \cite{clements_liu}         &                          & x                         & x                     & x                & x                   & x                        & x                                & Cumm haz                         \\
+\textbf{SmoothHazard} \cite{smoothHazard}   &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
+\textbf{Survival} \cite{survival-package}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x
 \end{tabular}%
@@
 ```{r supportSetUp, echo = FALSE, eval = eval_cs3}
-knitr::opts_chunk$set(echo = TRUE)
-#library(casebase)
-
-set.seed(2)#AT RISK INDIVIDUAL
-#set.seed(3)#less risk individual
-#read and remove ID column
-support <- read_tsv("support.tsv") 
-#impute without death, to avoid correlations
-sup.impute <-mice(support[,-c(2)], m = 1, maxit = 10,printFlag=FALSE)
-#take a complete dataset, Only one is being made and used here.
-#The logic being we only want to compare between models. 
-#Any Bias introduced is introduced to all models.
-completeData<-complete(sup.impute,1)
-#put death back into the complete dataset, as it was removed.
-completeData$death<-support$death
-#make everything into factors and remove observations without
-#imputed values.
-completeData<-na.omit(as.data.frame(unclass(completeData)))
-completeData<-completeData[,-c(3)]
+completeData=casebase::support
 #keep one individual out, at random, for which we will develop an 
@@
 #demonstrating incidence density before imputation
-plot(popTime(as.data.frame(support),time = "d.time",event = "death"))
+plot(casebase::popTime(as.data.frame(support),time = "d.time",event = "death"))
 ```
@@
 #demonstrating incidence density after imputation
-plot(popTime(as.data.frame(workingCompleteData),time = "d.time",event = "death"))
+plot(casebase::popTime(as.data.frame(workingCompleteData),time = "d.time",event = "death"))
 ```
@@
 #to determine the hazard function
-cb.Model=fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
+cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
                              formula_time = ~d.time,alpha=1,ratio=100)
@@
 #Estimating the absolute risk curve using the newData parameter.
-casebaseAbsolute=absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) ,
+casebaseAbsolute=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) ,
                               s="lambda.1se",method=c("numerical"))
@@
 #Create u and xCox, which will be used when fitting Cox with glmnet.
-u=Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
+u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
 xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))[-c(sam),]
 #hazard for cox using glmnet
-coxglmFit=cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
+coxglmFit=glmnet::cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
 #convergence demonstrated in plot
@@
 #this is used more as an object place holder.
-coxFit <- coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
+coxFit <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
 #The coefficients of this object will be replaced with the estimates from coxglmFit.
@@
 
-abCoxFit<-survfit(coxFit,newdata=as.data.frame(t(newDataCox)),time = seq(0,max(completeData$d.time), 1),type="breslow")
+abCoxFit<-survival::survfit(coxFit,newdata=as.data.frame(t(newDataCox)),time = seq(0,max(completeData$d.time), 1),type="breslow")
 ```
@@
 # (Kaplan-Meier risk curve)
-km <- Surv(time = completeData$d.time, event = completeData$death)
-abKm<-survfit(km~1,type='kaplan-meier',conf.type='log')
+km <- survival::Surv(time = completeData$d.time, event = completeData$death)
+abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
 ```

### 2019-12-06 analysis/paper/paper.tex
Lit-Review Edits
TODO: fix table references in lit review

debug case study 3 after casebase pull request complete

@@
 a scraper was coded to examine all packages in the survival task view on
-CRAN (ref). There was a total of 259 packages, many of which are not
-relevant to the problems casebase addresses. To filter these packages, a
-scraper was used to search for competing risks, cumulative incidence
-functions, non- proportional hazards and penalization. This reduced our
-set of 259 packages down to 60. From here, we manually examined each
-package to determine which ones are comparable to casebase, with either
-a different approach to the same problems or future features we would
-like to add to casebase.
+CRAN \citeyearpar{survTaskView}. There was a total of 259 packages, many
+of which are not relevant to the problems casebase addresses. To filter
+these packages, a scraper was used to search for competing risks,
+cumulative incidence functions, non- proportional hazards and
+penalization. This reduced our set of 259 packages down to 60. From
+here, we manually examined each package to determine which ones are
+comparable to casebase, with either a different approach to the same
+problems or future features we would like to add to casebase.
 
@@
 Many packages exist that implement semi-parametric cox models. Of those,
-glmnet, glmpath, penalized, RiskRegression introduce penalization in
-their model fitting. CoxRidge, rstpm2 and survival include penalization,
-but they implement the extended cox model permitting intuitive non
-proportional hazard modeling. In terms of penalization, the benefit of
-casebase comes from its flexibility. Linear models have a wide tool set,
-considerably larger than those made for cox regression (reference for
-number of packages between their task views, regression is spread out
-across Multivariate, SocialSciences, Robust among others. The focus of
-this task view is on Optimization Infrastructure Packages , General
-Purpose Continuous Solvers , Mathematical Programming Solvers , and
-Specific Applications in Optimization , or Multi Objective Optimization
-). Casebase permits many regularization tools developed for linear
+glmnet \citeyearpar{regpathcox}???, glmpath \citeyearpar{park_hastie},
+penalized \citeyearpar{l1penal}, RiskRegression
+\citeyearpar{gerds_blanche} introduce penalization in their model
+fitting. CoxRidge \citeyearpar{perperoglou} , rstpm2
+\citeyearpar{clements_liu} and survival \citeyearpar{survival-package}
+include penalization, but they implement the extended cox model
+permitting intuitive non proportional hazard modeling while the previous
+four do not. In terms of penalization, the benefit of casebase comes
+from its flexibility. Linear models have a wide tool set, considerably
+larger than those made for cox regression (\emph{reference for number of
+packages between their task views, regression is spread out across
+Multivariate, SocialSciences, Robust among others. The focus of this
+task view is on Optimization Infrastructure Packages , General Purpose
+Continuous Solvers , Mathematical Programming Solvers , and Specific
+Applications in Optimization , or Multi Objective Optimization}).
+Casebase permits many regularization tools developed for linear
 regression to be repurposed within a survival context.
@@
 Parametric models permit non-proportional hazards and are implemented in
-a handful of packages. Of interest are CFC, flexsurv, SmoothHazard,
-Rsptm2 and Survival. Smooth hazard is limited to weibull distributions,
-where as flexsurv and survival contain implementations for users to
-supply any distribution they would like. Flexsurv, smooth hazard, mets
-and rstpm2 can use splines, permitting a flexible fit over time. Though
+a handful of packages. Of interest are CFC
+\citeyearpar{sharabiani_mahani_2019}, flexsurv \citeyearpar{flexsurv} ,
+SmoothHazard \citeyearpar{smoothHazard}, Rsptm2
+\citeyearpar{clements_liu} and Survival \citeyearpar{survival-package}.
+Smooth hazard is limited to weibull
+distributions\citeyearpar{smoothHazard}, where as flexsurv and survival
+contain implementations for users to supply any distribution they would
+like \citeyearpar{survival-package}, \citeyearpar{flexsurv}. Flexsurv,
+smooth hazard, mets and rstpm2 can use splines, permitting a flexible
+fit over time \citeyearpar{flexsurv}, \citeyearpar{smoothHazard},
+\citeyearpar{scheike2014estimating}, \citeyearpar{clements_liu} . Though
 casebase is limited to 3 distributions at the moment, it also uses
-splines allowing for distributions that may not be consistent over time.
-
-Of the methods mentioned so far, only CFC, survival, mets and flexsurv
-contain implementations for competing risks(how is it done for
-parametric models?). The differentiating factor between these packages
-and casebase is that casebase handles competing risks through multiple
-logistic regression.
-
-A feature in Smooth hazard, survival and Rstpm2 are they they can handle
-interval censoring, where as casebase currently can not.
-
-Of all the packages listed in the table, CFC, Flexsurv, mets,
+splines allowing for distributions that may not be consistent over
+survival time.
+
+Of the methods mentioned so far, only CFC,flexsurv, mets and survival
+contain implementations for competing risks
+\citeyearpar{sharabiani_mahani_2019},
+\citeyearpar{flexsurv},\citeyearpar{scheike2014estimating}
+,\citeyearpar{survival-package} (\emph{how is it done for parametric
+models?}). The differentiating factor between these packages and
+casebase is that casebase handles competing risks through multiple
+logistic regression. Smooth hazard, survival and Rstpm2 can handle
+interval censoring, where as casebase currently can not. Interval
+censoring is simple feature to implement and is a potential improvement
+to add to casebase. (There's also multistate models, which I haven't
+mentioned in this paper)
+
+Of all the packages listed in the table, CFC , Flexsurv, mets,
 RiskRegression, rstpm2 and survival have implementations for cumulative
-incidence. Since casebase is parametric, the baseline hazard is known
-and requires integration to retrieve the cumulative incidence. This is
-similar to the approaches of CFC, flexsurv,\\
+incidence \citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
+\citeyearpar{scheike2014estimating}, \citeyearpar{gerds_blanche},
+\citeyearpar{clements_liu}, \citeyearpar{survival-package}. Since
+casebase is parametric, the baseline hazard is known and requires
+integration to retrieve the cumulative incidence. This is similar to the
+approaches of CFC and flexsurv \citeyearpar{sharabiani_mahani_2019},
+\citeyearpar{flexsurv}, while the other packages use semi-parametric
+approaches to retrieve the baseline after fitting
+\citeyearpar{scheike2014estimating}, \citeyearpar{gerds_blanche}. rstpm2
+and survival use both for their respective parametric and
+semi-parametric model implementations \citeyearpar{clements_liu},
+\citeyearpar{survival-package}.
 
@@
 R> #demonstrating incidence density before imputation
-R> plot(popTime(as.data.frame(support),time = "d.time",event = "death"))
+R> plot(casebase::popTime(as.data.frame(support),time = "d.time",event = "death"))
 \end{CodeInput}
@@
 R> #demonstrating incidence density after imputation
-R> plot(popTime(as.data.frame(workingCompleteData),time = "d.time",event = "death"))
+R> plot(casebase::popTime(as.data.frame(workingCompleteData),time = "d.time",event = "death"))
 \end{CodeInput}
@@
 R> #to determine the hazard function
-R> cb.Model=fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
+R> cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
 R>                              formula_time = ~d.time,alpha=1,ratio=100)
@@
 R> #Estimating the absolute risk curve using the newData parameter.
-R> casebaseAbsolute=absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) ,
+R> casebaseAbsolute=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) ,
 R>                               s="lambda.1se",method=c("numerical"))
@@
 R> #Create u and xCox, which will be used when fitting Cox with glmnet.
-R> u=Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
+R> u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
 R> xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))[-c(sam),]
 R> #hazard for cox using glmnet
-R> coxglmFit=cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
+R> coxglmFit=glmnet::cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
 R> #convergence demonstrated in plot
@@
 R> #this is used more as an object place holder.
-R> coxFit <- coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
+R> coxFit <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
 R> #The coefficients of this object will be replaced with the estimates from coxglmFit.
@@
 R> 
-R> abCoxFit<-survfit(coxFit,newdata=as.data.frame(t(newDataCox)),time = seq(0,max(completeData$d.time), 1),type="breslow")
+R> abCoxFit<-survival::survfit(coxFit,newdata=as.data.frame(t(newDataCox)),time = seq(0,max(completeData$d.time), 1),type="breslow")
 \end{CodeInput}
@@
 R> # (Kaplan-Meier risk curve)
-R> km <- Surv(time = completeData$d.time, event = completeData$death)
-R> abKm<-survfit(km~1,type='kaplan-meier',conf.type='log')
+R> km <- survival::Surv(time = completeData$d.time, event = completeData$death)
+R> abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
 \end{CodeInput}
@@
 
-This report was generated on 2019-12-04 13:59:29 using the following
+This report was generated on 2019-12-06 16:16:07 using the following
 computational environment and dependencies:
@@
 #>  tz       America/Montreal            
-#>  date     2019-12-04                  
+#>  date     2019-12-06                  
 #> 
 #> - Packages -------------------------------------------------------------------
-#>  ! package       * version    date       lib
-#>    abind           1.4-5      2016-07-21 [1]
+#>  package     * version    date       lib
 #>  assertthat    0.2.1      2019-03-21 [1]
 #>  backports     1.1.5      2019-10-02 [1]
-#>    bookdown        0.16       2019-11-22 [1]
-#>    boot            1.3-23     2019-07-05 [1]
 #>  broom         0.5.2      2019-04-07 [1]
 #>  callr         3.3.2      2019-09-22 [1]
-#>    car             3.0-5      2019-11-15 [1]
-#>    carData         3.0-3      2019-11-16 [1]
 #>  casebase    * 0.2.1.9001 2019-08-20 [1]
-#>  R casebasepaper * 0.0.0.9000 <NA>       [?]
 #>  cellranger    1.1.0      2016-07-27 [1]
-#>    class           7.3-15     2019-01-01 [1]
 #>  cli           1.1.0      2019-03-19 [1]
 #>  codetools     0.2-16     2018-12-24 [1]
-#>    colorspace    * 1.4-1      2019-03-18 [1]
+#>  colorspace    1.4-1      2019-03-18 [1]
 #>  crayon        1.3.4      2017-09-16 [1]
-#>    curl            4.2        2019-09-24 [1]
-#>    data.table    * 1.12.6     2019-10-18 [1]
+#>  data.table    1.12.6     2019-10-18 [1]
 #>  DBI           1.0.0      2018-05-02 [1]
 #>  dbplyr        1.4.2      2019-06-17 [1]
-#>    DEoptimR        1.0-8      2016-11-19 [1]
 #>  desc          1.2.0      2018-05-01 [1]
@@
 #>  dplyr       * 0.8.3      2019-07-04 [1]
-#>    e1071           1.7-2      2019-06-05 [1]
 #>  ellipsis      0.3.0      2019-09-20 [1]
@@
 #>  foreach       1.4.7      2019-07-27 [1]
-#>    foreign         0.8-72     2019-08-02 [1]
 #>  fs            1.3.1      2019-05-06 [1]
@@
 #>  ggplot2     * 3.2.1      2019-08-10 [1]
-#>    git2r           0.26.1     2019-06-29 [1]
 #>  glmnet      * 3.0-1      2019-11-15 [1]
@@
 #>  haven         2.2.0      2019-11-08 [1]
-#>    here            0.1        2017-05-28 [1]
 #>  hms           0.5.2      2019-10-30 [1]
@@
 #>  iterators     1.0.12     2019-07-26 [1]
-#>    jomo            2.6-10     2019-10-22 [1]
 #>  jsonlite      1.6        2018-12-07 [1]
 #>  knitr         1.26       2019-11-12 [1]
-#>    laeken          0.5.0      2019-01-10 [1]
-#>    lattice       * 0.20-38    2018-11-04 [1]
-#>    lava            1.6.6      2019-08-01 [1]
+#>  lattice       0.20-38    2018-11-04 [1]
 #>  lazyeval      0.2.2      2019-03-15 [1]
 #>  lifecycle     0.1.0      2019-08-01 [1]
-#>    lme4            1.1-21     2019-03-05 [1]
-#>    lmtest          0.9-37     2019-04-30 [1]
 #>  lubridate     1.7.4      2018-04-11 [1]
 #>  magrittr    * 1.5        2014-11-22 [1]
-#>    MASS            7.3-51.4   2019-03-31 [1]
 #>  Matrix      * 1.2-17     2019-03-22 [1]
@@
 #>  mgcv          1.8-31     2019-11-09 [1]
-#>    mice          * 3.6.0      2019-07-10 [1]
-#>    minqa           1.2.4      2014-10-09 [1]
-#>    mitml           0.3-7      2019-01-07 [1]
 #>  modelr        0.1.5      2019-08-08 [1]
@@
 #>  nlme          3.1-142    2019-11-07 [1]
-#>    nloptr          1.2.1      2018-10-03 [1]
-#>    nnet            7.3-12     2016-02-02 [1]
-#>    numDeriv        2016.8-1.1 2019-06-06 [1]
-#>    openxlsx        4.1.3      2019-11-07 [1]
-#>    pan             1.6        2018-06-29 [1]
 #>  pillar        1.4.2      2019-06-29 [1]
@@
 #>  R6            2.4.1      2019-11-12 [1]
-#>    ranger          0.11.2     2019-03-07 [1]
 #>  Rcpp          1.0.3      2019-11-08 [1]
@@
 #>  reprex        0.3.0      2019-05-16 [1]
-#>    rio             0.5.16     2018-11-26 [1]
 #>  rlang         0.4.2      2019-11-23 [1]
 #>  rmarkdown     1.17       2019-11-13 [1]
-#>    robustbase      0.93-5     2019-05-12 [1]
-#>    rpart           4.1-15     2019-04-12 [1]
 #>  rprojroot     1.3-2      2018-01-03 [1]
@@
 #>  shape         1.4.4      2018-02-07 [1]
-#>    sp              1.3-2      2019-11-07 [1]
 #>  stringi       1.4.3      2019-03-12 [1]
@@
 #>  tidyverse   * 1.3.0      2019-11-21 [1]
-#>    timereg         1.9.4      2019-07-30 [1]
 #>  usethis       1.5.1      2019-07-04 [1]
-#>    vcd             1.4-4      2017-12-06 [1]
 #>  vctrs         0.2.0      2019-07-05 [1]
 #>  VGAM          1.1-2      2019-11-21 [1]
-#>    VIM           * 4.8.0      2019-02-11 [1]
 #>  withr         2.1.2      2018-03-15 [1]
@@
 #>  xml2          1.2.2      2019-08-09 [1]
-#>    xtable          1.8-4      2019-04-21 [1]
 #>  yaml          2.2.0      2018-07-25 [1]
 #>  zeallot       0.1.0      2018-01-28 [1]
-#>    zip             2.0.4      2019-09-01 [1]
-#>    zoo             1.8-6      2019-05-28 [1]
 #>  source                                  
@@
 #>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
 #>  Github (sahirbhatnagar/casebase@aecb816)
-#>  <NA>                                    
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
 #>  CRAN (R 3.6.0)                          
@@
 #> [1] /Library/Frameworks/R.framework/Versions/3.6/Resources/library
-#> 
-#>  R -- Package was removed from disk.
 \end{CodeOutput}
@@
 #> Local:    master /Users/jislam/Desktop/Gits/cbpaper
-#> Remote:   master @ origin (https://github.com/sahirbhatnagar/cbpaper.git)
-#> Head:     [5022c53] 2019-12-04: rough litreview and casestudy 3 added
+#> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
+#> Head:     [22bfed0] 2019-12-06: currentDraftEdits
 \end{CodeOutput}

### 2019-12-08 analysis/paper/paper.Rmd

--- 2019-12-06 analysis/paper/paper.Rmd
@@
   - name: Sahir Bhatnagar *
-    affiliation: McGill University
+    affiliation: McGill Univeristy
     address: >
@@
     email: \email{max.turgeon@umanitoba.ca}
-    url: http://maxturgeon.ca/
+    url: https://maxturgeon.ca/
   - name: Jesse Islam 
@@
   - name: James Hanley
-    affiliation: McGill University
+    affiliation: McGill Univeristy
     address: >
@@
 
-<!-- # Code formatting {.unnumbered} -->
+# Code formatting {.unnumbered}
 
-<!-- Don't use markdown, instead use the more precise latex commands: -->
+Don't use markdown, instead use the more precise latex commands:
 
-<!-- * \proglang{Java} -->
-<!-- * \pkg{plyr} -->
-<!-- * \code{print("abc")} -->
+* \proglang{Java}
+* \pkg{plyr}
+* \code{print("abc")}
 
@@
 eval_implementation <- TRUE
-eval_cs1 <- FALSE
-eval_cs2 <- FALSE
+eval_cs1 <- TRUE
+eval_cs2 <- TRUE
 eval_cs3 <- FALSE
-eval_cs4 <- TRUE
 ```
@@
 
-As shown by Saarela & Arjas [-@saarela2015non] (and further expanded in Saarela [-@saarela2016case]), writing the likelihood arising from this data-generating mechanism using the framework of non-homogeneous Poisson processes, we eventually reach an expression where each person-moment's contribution is of the form
+As shown by Saarela & Arjas [-@saarela2015non] (and further expanded in Saarela [-@saarela2016case]), writing the likelihood arising from this data-generating mechanism using the framework of non-homogenous Poisson processes, we eventually reach an expression where each person-moment's contribution is of the form
 $$\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},$$
@@
 
-In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base sampling *uniformly*, i.e.
+In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base samping *uniformly*, i.e.
 to sample the base series uniformly from the study base.
@@
 The functions in the casebase package can be divided into two categories: 1) data visualization, in the form of population-time plots; and 2) parametric modeling.
-We explicitly aimed at being compatible with both \code{data.frame}s and \code{data.table}s.
+We explicitely aimed at being compatible with both \code{data.frame}s and \code{data.table}s.
 This is evident in some of the coding choices we made, and it is also reflected in our unit tests.
@@
 
-# Case study 4--Stanford Heart Transplant Data
-
-Although the previous case studies provide a broad overview of the capabilities of \pkg{casebase}, they all have two properties in common:
-\begin{itemize}
-  \item All covariates are fixed, i.e. they do not change over time;
-  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
-\end{itemize}
-
-Case-base sampling has been used in the literature to study vaccination safety, where the exposure period was defined as the week following vaccination [@saarela2015case].
-Hence, the main covariate of interest, i.e.
-exposure to the vaccine, was changing over time.
-In this context, case-base sampling offers an efficient alternative to nested case-control designs or self-matching.
-
-The purpose of the next case study is to show how \pkg{casebase} can also be used for survival analysis problems that do not share the properties above.
-To this end, we will use the Stanford Heart Transplant Data [@clark1971cardiac,@crowley1977covariance]; we will use the version available in the \pkg{survival} package.
-
-Recall the setting of this study: patients were admitted to the Stanford program after meeting with their physician and determining that they were unlikely to respond to other forms of treatment.
-After enrollment, the program searches for a suitable donor for the patient, which can take anywhere between a few days to almost a year.
-We are interested in the effect of a heart transplant on survival; therefore, the patient is considered exposed only after the transplant has occured.
-
-As before, we can look at the population-time plot for a graphical summary of the event incidence.
-As we can see, most events occur early during the follow-up period, and therefore we do not expect the hazard to be constant.
-
-```{r stanford-poptime, eval = eval_cs4}
-library(survival)
-library(casebase)
-
-stanford_popTime <- popTime(jasa, time = "futime", 
-                            event = "fustat")
-plot(stanford_popTime)
-```
-
-Since the exposure is time-dependent, we need to manually define the exposure variable *after* case-base sampling and *before* fitting the hazard function.
-For this reason, we will use the `sampleCaseBase` function directly.
-
-```{r, eval = eval_cs4}
-library(tidyverse)
-library(lubridate)
-
-cb_data <- sampleCaseBase(jasa, time = "futime", 
-                          event = "fustat", ratio = 10)
-```
-
-Next, we will compute the number of days from acceptance into the program to transplant, and we use this variable to determine whether each population-moment is exposed or not.
-
-```{r, eval = eval_cs4}
-# Define exposure variable
-cb_data <- mutate(cb_data,
-                  txtime = time_length(accept.dt %--% tx.date, 
-                                       unit = "days"),
-                  exposure = case_when(
-                    is.na(txtime) ~ 0L,
-                    txtime > futime ~ 0L,
-                    txtime <= futime ~ 1L
-                  ))
-```
-
-Finally, we can fit the hazard using various linear predictors.
-
-```{r, eval = eval_cs4, warning = FALSE}
-library(splines)
-# Fit several models
-fit1 <- fitSmoothHazard(fustat ~ exposure,
-                        data = cb_data, time = "futime")
-fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
-                        data = cb_data, time = "futime")
-fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
-                        data = cb_data, time = "futime")
-fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
-                        data = cb_data, time = "futime")
-```
-
-Note that the fourth model (i.e.
-`fit4`) includes an interaction term between exposure and follow-up time.
-In other words, this model no longer exhibit proportional hazards.
-The evidence of non-proportionality of hazards in the Stanford Heart Transplant data has been widely discussed [@arjas1988graphical].
-
-We can then compare the goodness of fit of these four models using the Akaike Information Criterion (AIC).
-
-```{r, eval = eval_cs4}
-# Compute AIC
-c("Model1" = AIC(fit1),
-  "Model2" = AIC(fit2),
-  "Model3" = AIC(fit3),
-  "Model4" = AIC(fit4))
-```
-
-As we can, the best fit is the fourth model. By visualizing the hazard functions for both exposed and unexposed individuals, we can more clearly see how the hazards are no longer proportional.
-
-```{r stanford-hazard, eval = eval_cs4}
-# Compute hazards---
-# First, create a list of time points for both exposure status
-hazard_data <- expand.grid(exposure = c(0, 1),
-                           futime = seq(0, 1000,
-                                        length.out = 100))
-# Set the offset to zero
-hazard_data$offset <- 0 
-# Use predict to get the fitted values, and exponentiate to 
-# transform to the right scale
-hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
-                                 type = "link"))
-# Add labels for plots
-hazard_data$Status = factor(hazard_data$exposure,
-                            labels = c("NoTrans", "Trans"))
-
-ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
-    geom_line() +
-    theme_minimal() +
-    theme(legend.position = 'top') +
-    ylab('Hazard') + xlab('Follow-up time')
-```
-
-The non-proportionality seems to be more pronounced at the beginning of follow-up than the end.
-Finally, we can turn these estimate of the hazard function into estimates of the cumulative incidence functions.
-
-```{r stanford-risk, eval = eval_cs4}
-# Compute absolute risk curves
-newdata <- data.frame(exposure = c(0, 1))
-absrisk <- absoluteRisk(fit4, newdata = newdata, 
-                        time = seq(0, 1000, length.out = 100))
-
-colnames(absrisk) <- c("Time", "NoTrans", "Trans")
-
-# Rearrange the data
-absrisk <- gather(as.data.frame(absrisk),
-                  "Status", "Risk", -Time)
-
-ggplot(absrisk, aes(Time, Risk, colour = Status)) +
-  geom_line() +
-  theme_minimal() +
-  theme(legend.position = 'top') +
-  expand_limits(y = 0.5) +
-  xlab('Follow-up time') + ylab('Cum. Incidence')
-```
-
-Note that we can easily adapt the code above to the situation where a patient receives a heart transplant at a point in time of interest, for example after 30 days.
-
-```{r stanford-1y-haz, eval = eval_cs4}
-# Compute hazards---
-# First, create a list of time points for both exposure status
-one_yr_haz <- data.frame(futime = seq(0, 365,
-                                      length.out = 100))
-one_yr_haz <- mutate(one_yr_haz,
-                     offset = 0,
-                     exposure = if_else(futime < 30, 0, 1))
-one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
-                                type = "link"))
-ggplot(one_yr_haz, aes(futime, hazard)) +
-  geom_line() +
-  theme_minimal() +
-  theme(legend.position = 'top') +
-  ylab('Hazard') + xlab('Follow-up time') +
-  geom_vline(xintercept = 30, linetype = 'dashed')
-```
-
-We can then compare the 1-year mortality risk without transplant and with transplant at 30 days.
-
-```{r stanford-1y-risk, eval = eval_cs4}
-absoluteRisk(fit4, newdata = data.frame(exposure = 0),
-             time = 365)
-
-# Use the trapezoidal rule to estimate the cumulative hazard
-min_hazard <- min(one_yr_haz$hazard)
-max_hazard <- max(one_yr_haz$hazard)
-increment <- 365/99
-cumhaz_est <- 0.5*increment*(sum(2*one_yr_haz$hazard) - min_hazard - max_hazard)
-1 - exp(-cumhaz_est)
-```
-
-As we can see, the risk estimate at 1-year is about 30% lower if the patient receives a heart transplant at 30 days.
-
 # Discussion

### 2019-12-08 analysis/paper/paper.Rmd

@@
 
+# Case study 4--Stanford Heart Transplant Data
+
+Although the previous case studies provide a broad overview of the capabilities of \pkg{casebase}, they all have two properties in common:
+\begin{itemize}
+  \item All covariates are fixed, i.e. they do not change over time;
+  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
+\end{itemize}
+
+Case-base sampling has been used in the literature to study vaccination safety, where the exposure period was defined as the week following vaccination [@saarela2015case].
+Hence, the main covariate of interest, i.e.
+exposure to the vaccine, was changing over time.
+In this context, case-base sampling offers an efficient alternative to nested case-control designs or self-matching.
+
+The purpose of the next case study is to show how \pkg{casebase} can also be used for survival analysis problems that do not share the properties above.
+To this end, we will use the Stanford Heart Transplant Data [@clark1971cardiac,@crowley1977covariance]; we will use the version available in the \pkg{survival} package.
+
+Recall the setting of this study: patients were admitted to the Stanford program after meeting with their physician and determining that they were unlikely to respond to other forms of treatment.
+After enrollment, the program searches for a suitable donor for the patient, which can take anywhere between a few days to almost a year.
+We are interested in the effect of a heart transplant on survival; therefore, the patient is considered exposed only after the transplant has occured.
+
+As before, we can look at the population-time plot for a graphical summary of the event incidence.
+As we can see, most events occur early during the follow-up period, and therefore we do not expect the hazard to be constant.
+
+```{r stanford-poptime, eval = eval_cs4}
+library(survival)
+library(casebase)
+
+stanford_popTime <- popTime(jasa, time = "futime", 
+                            event = "fustat")
+plot(stanford_popTime)
+```
+
+Since the exposure is time-dependent, we need to manually define the exposure variable *after* case-base sampling and *before* fitting the hazard function.
+For this reason, we will use the `sampleCaseBase` function directly.
+
+```{r, eval = eval_cs4}
+library(tidyverse)
+library(lubridate)
+
+cb_data <- sampleCaseBase(jasa, time = "futime", 
+                          event = "fustat", ratio = 10)
+```
+
+Next, we will compute the number of days from acceptance into the program to transplant, and we use this variable to determine whether each population-moment is exposed or not.
+
+```{r, eval = eval_cs4}
+# Define exposure variable
+cb_data <- mutate(cb_data,
+                  txtime = time_length(accept.dt %--% tx.date, 
+                                       unit = "days"),
+                  exposure = case_when(
+                    is.na(txtime) ~ 0L,
+                    txtime > futime ~ 0L,
+                    txtime <= futime ~ 1L
+                  ))
+```
+
+Finally, we can fit the hazard using various linear predictors.
+
+```{r, eval = eval_cs4, warning = FALSE}
+library(splines)
+# Fit several models
+fit1 <- fitSmoothHazard(fustat ~ exposure,
+                        data = cb_data, time = "futime")
+fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
+                        data = cb_data, time = "futime")
+fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
+                        data = cb_data, time = "futime")
+fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
+                        data = cb_data, time = "futime")
+```
+
+Note that the fourth model (i.e.
+`fit4`) includes an interaction term between exposure and follow-up time.
+In other words, this model no longer exhibit proportional hazards.
+The evidence of non-proportionality of hazards in the Stanford Heart Transplant data has been widely discussed [@arjas1988graphical].
+
+We can then compare the goodness of fit of these four models using the Akaike Information Criterion (AIC).
+
+```{r, eval = eval_cs4}
+# Compute AIC
+c("Model1" = AIC(fit1),
+  "Model2" = AIC(fit2),
+  "Model3" = AIC(fit3),
+  "Model4" = AIC(fit4))
+```
+
+As we can, the best fit is the fourth model. By visualizing the hazard functions for both exposed and unexposed individuals, we can more clearly see how the hazards are no longer proportional.
+
+```{r stanford-hazard, eval = eval_cs4}
+# Compute hazards---
+# First, create a list of time points for both exposure status
+hazard_data <- expand.grid(exposure = c(0, 1),
+                           futime = seq(0, 1000,
+                                        length.out = 100))
+# Set the offset to zero
+hazard_data$offset <- 0 
+# Use predict to get the fitted values, and exponentiate to 
+# transform to the right scale
+hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
+                                 type = "link"))
+# Add labels for plots
+hazard_data$Status = factor(hazard_data$exposure,
+                            labels = c("NoTrans", "Trans"))
+
+ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
+    geom_line() +
+    theme_minimal() +
+    theme(legend.position = 'top') +
+    ylab('Hazard') + xlab('Follow-up time')
+```
+
+The non-proportionality seems to be more pronounced at the beginning of follow-up than the end.
+Finally, we can turn these estimate of the hazard function into estimates of the cumulative incidence functions.
+
+```{r stanford-risk, eval = eval_cs4}
+# Compute absolute risk curves
+newdata <- data.frame(exposure = c(0, 1))
+absrisk <- absoluteRisk(fit4, newdata = newdata, 
+                        time = seq(0, 1000, length.out = 100))
+
+colnames(absrisk) <- c("Time", "NoTrans", "Trans")
+
+# Rearrange the data
+absrisk <- gather(as.data.frame(absrisk),
+                  "Status", "Risk", -Time)
+
+ggplot(absrisk, aes(Time, Risk, colour = Status)) +
+  geom_line() +
+  theme_minimal() +
+  theme(legend.position = 'top') +
+  expand_limits(y = 0.5) +
+  xlab('Follow-up time') + ylab('Cum. Incidence')
+```
+
+Note that we can easily adapt the code above to the situation where a patient receives a heart transplant at a point in time of interest, for example after 30 days.
+
+```{r stanford-1y-haz, eval = eval_cs4}
+# Compute hazards---
+# First, create a list of time points for both exposure status
+one_yr_haz <- data.frame(futime = seq(0, 365,
+                                      length.out = 100))
+one_yr_haz <- mutate(one_yr_haz,
+                     offset = 0,
+                     exposure = if_else(futime < 30, 0, 1))
+one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
+                                type = "link"))
+ggplot(one_yr_haz, aes(futime, hazard)) +
+  geom_line() +
+  theme_minimal() +
+  theme(legend.position = 'top') +
+  ylab('Hazard') + xlab('Follow-up time') +
+  geom_vline(xintercept = 30, linetype = 'dashed')
+```
+
+We can then compare the 1-year mortality risk without transplant and with transplant at 30 days.
+
+```{r stanford-1y-risk, eval = eval_cs4}
+absoluteRisk(fit4, newdata = data.frame(exposure = 0),
+             time = 365)
+
+# Use the trapezoidal rule to estimate the cumulative hazard
+min_hazard <- min(one_yr_haz$hazard)
+max_hazard <- max(one_yr_haz$hazard)
+increment <- 365/99
+cumhaz_est <- 0.5*increment*(sum(2*one_yr_haz$hazard) - min_hazard - max_hazard)
+1 - exp(-cumhaz_est)
+```
+
+As we can see, the risk estimate at 1-year is about 30% lower if the patient receives a heart transplant at 30 days.
+
 # Discussion

### 2019-12-08 analysis/paper/paper.tex

--- 2019-12-06 analysis/paper/paper.tex
@@
   E-mail: \email{max.turgeon@umanitoba.ca}\\
-  URL: \url{http://maxturgeon.ca/}\\~\\
+  URL: \url{https://maxturgeon.ca/}\\~\\
       Jesse Islam\\
@@
 
-Note that the fourth model (i.e. \texttt{fit4}) includes an interaction
+Note that the fourth model (i.e.~\texttt{fit4}) includes an interaction
 term between exposure and follow-up time. In other words, this model no
@@
 
-This report was generated on 2019-12-06 10:11:23 using the following
+This report was generated on 2019-12-08 11:49:58 using the following
 computational environment and dependencies:
@@
 \begin{CodeOutput}
-#> - Session info ---------------------------------------------------------------
+#> - Session info ----------------------------------------------------------
 #>  setting  value                       
@@
 #>  ui       X11                         
-#>  language en_CA:en                    
+#>  language (EN)                        
 #>  collate  en_CA.UTF-8                 
@@
 #>  tz       America/Winnipeg            
-#>  date     2019-12-06                  
+#>  date     2019-12-08                  
 #> 
-#> - Packages -------------------------------------------------------------------
-#>  package     * version    date       lib
-#>  assertthat    0.2.1      2019-03-21 [1]
-#>  backports     1.1.5      2019-10-02 [1]
-#>  broom         0.5.2      2019-04-07 [1]
-#>  callr         3.3.2      2019-09-22 [1]
-#>  casebase    * 0.2.1.9001 2019-08-13 [1]
-#>  cellranger    1.1.0      2016-07-27 [1]
-#>  cli           1.1.0      2019-03-19 [1]
-#>  colorspace    1.4-1      2019-03-18 [1]
-#>  crayon        1.3.4      2017-09-16 [1]
-#>  data.table    1.12.6     2019-10-18 [1]
-#>  DBI           1.0.0      2018-05-02 [1]
-#>  dbplyr        1.4.2      2019-06-17 [1]
-#>  desc          1.2.0      2018-05-01 [1]
-#>  devtools      2.2.1      2019-09-24 [1]
-#>  digest        0.6.23     2019-11-23 [1]
-#>  dplyr       * 0.8.3      2019-07-04 [1]
-#>  ellipsis      0.3.0      2019-09-20 [1]
-#>  evaluate      0.14       2019-05-28 [1]
-#>  farver        2.0.1      2019-11-13 [1]
-#>  forcats     * 0.4.0      2019-02-17 [1]
-#>  fs            1.3.1      2019-05-06 [1]
-#>  generics      0.0.2      2018-11-29 [1]
-#>  ggplot2     * 3.2.1      2019-08-10 [1]
-#>  glue          1.3.1      2019-03-12 [1]
-#>  gtable        0.3.0      2019-03-25 [1]
-#>  haven         2.2.0      2019-11-08 [1]
-#>  hms           0.5.2      2019-10-30 [1]
-#>  htmltools     0.4.0      2019-10-04 [1]
-#>  httr          1.4.1      2019-08-05 [1]
-#>  jsonlite      1.6        2018-12-07 [1]
-#>  knitr         1.26       2019-11-12 [1]
-#>  labeling      0.3        2014-08-23 [1]
-#>  lattice       0.20-38    2018-11-04 [4]
-#>  lazyeval      0.2.2      2019-03-15 [1]
-#>  lifecycle     0.1.0      2019-08-01 [1]
-#>  lubridate   * 1.7.4      2018-04-11 [1]
-#>  magrittr    * 1.5        2014-11-22 [1]
-#>  Matrix        1.2-17     2019-03-22 [4]
-#>  memoise       1.1.0      2017-04-21 [1]
-#>  mgcv          1.8-31     2019-11-09 [4]
-#>  modelr        0.1.5      2019-08-08 [1]
-#>  munsell       0.5.0      2018-06-12 [1]
-#>  nlme          3.1-142    2019-11-07 [4]
-#>  pillar        1.4.2      2019-06-29 [1]
-#>  pkgbuild      1.0.6      2019-10-09 [1]
-#>  pkgconfig     2.0.3      2019-09-22 [1]
-#>  pkgload       1.0.2      2018-10-29 [1]
-#>  prettyunits   1.0.2      2015-07-13 [1]
-#>  processx      3.4.1      2019-07-18 [1]
-#>  ps            1.3.0      2018-12-21 [1]
-#>  purrr       * 0.3.3      2019-10-18 [1]
-#>  R6            2.4.1      2019-11-12 [1]
-#>  Rcpp          1.0.3      2019-11-08 [1]
-#>  readr       * 1.3.1      2018-12-21 [1]
-#>  readxl        1.3.1      2019-03-13 [1]
-#>  remotes       2.1.0      2019-06-24 [1]
-#>  reprex        0.3.0      2019-05-16 [1]
-#>  rlang         0.4.2      2019-11-23 [1]
-#>  rmarkdown     1.18       2019-11-27 [1]
-#>  rprojroot     1.3-2      2018-01-03 [1]
-#>  rstudioapi    0.10       2019-03-19 [1]
-#>  rticles       0.12       2019-11-12 [1]
-#>  rvest         0.3.5      2019-11-08 [1]
-#>  scales        1.1.0      2019-11-18 [1]
-#>  sessioninfo   1.1.1      2018-11-05 [1]
-#>  stringi       1.4.3      2019-03-12 [1]
-#>  stringr     * 1.4.0      2019-02-10 [1]
-#>  survival    * 2.44-1.1   2019-04-01 [4]
-#>  testthat      2.3.1      2019-12-01 [1]
-#>  tibble      * 2.1.3      2019-06-06 [1]
-#>  tidyr       * 1.0.0      2019-09-11 [1]
-#>  tidyselect    0.2.5      2018-10-11 [1]
-#>  tidyverse   * 1.3.0      2019-11-21 [1]
-#>  usethis       1.5.1      2019-07-04 [1]
-#>  vctrs         0.2.0      2019-07-05 [1]
-#>  VGAM          1.1-2      2019-11-21 [1]
-#>  withr         2.1.2      2018-03-15 [1]
-#>  xfun          0.11       2019-11-12 [1]
-#>  xml2          1.2.2      2019-08-09 [1]
-#>  yaml          2.2.0      2018-07-25 [1]
-#>  zeallot       0.1.0      2018-01-28 [1]
-#>  source                                  
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  Github (sahirbhatnagar/casebase@aecb816)
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
+#> - Packages --------------------------------------------------------------
+#>  package     * version    date       lib source                          
+#>  assertthat    0.2.1      2019-03-21 [1] CRAN (R 3.6.1)                  
+#>  backports     1.1.5      2019-10-02 [1] CRAN (R 3.6.1)                  
+#>  broom         0.5.2      2019-04-07 [1] CRAN (R 3.6.1)                  
+#>  callr         3.3.1      2019-07-18 [1] CRAN (R 3.6.1)                  
+#>  casebase    * 0.2.1.9001 2019-07-29 [1] local                           
+#>  cellranger    1.1.0      2016-07-27 [1] CRAN (R 3.5.0)                  
+#>  cli           1.1.0      2019-03-19 [1] CRAN (R 3.6.1)                  
+#>  colorspace    1.4-1      2019-03-18 [1] CRAN (R 3.6.1)                  
+#>  crayon        1.3.4      2017-09-16 [1] CRAN (R 3.5.0)                  
+#>  data.table    1.12.2     2019-04-07 [1] CRAN (R 3.6.1)                  
+#>  desc          1.2.0      2018-05-01 [1] CRAN (R 3.5.0)                  
+#>  devtools      2.1.0      2019-07-06 [1] CRAN (R 3.6.1)                  
+#>  digest        0.6.22     2019-10-21 [1] CRAN (R 3.6.1)                  
+#>  dplyr       * 0.8.3      2019-07-04 [1] CRAN (R 3.6.1)                  
+#>  ellipsis      0.3.0      2019-09-20 [1] CRAN (R 3.6.1)                  
+#>  evaluate      0.14       2019-05-28 [1] CRAN (R 3.6.1)                  
+#>  forcats     * 0.4.0      2019-02-17 [1] CRAN (R 3.6.1)                  
+#>  fs            1.3.1      2019-05-06 [1] CRAN (R 3.6.1)                  
+#>  generics      0.0.2      2018-11-29 [1] CRAN (R 3.6.1)                  
+#>  ggplot2     * 3.2.1      2019-08-10 [1] CRAN (R 3.6.1)                  
+#>  glue          1.3.1      2019-03-12 [1] CRAN (R 3.6.1)                  
+#>  gtable        0.3.0      2019-03-25 [1] CRAN (R 3.6.1)                  
+#>  haven         2.1.1      2019-07-04 [1] CRAN (R 3.6.1)                  
+#>  hms           0.5.2      2019-10-30 [1] CRAN (R 3.6.1)                  
+#>  htmltools     0.3.6      2017-04-28 [1] CRAN (R 3.5.0)                  
+#>  httr          1.4.1      2019-08-05 [1] CRAN (R 3.6.1)                  
+#>  jsonlite      1.6        2018-12-07 [1] CRAN (R 3.6.1)                  
+#>  knitr         1.23       2019-05-18 [1] CRAN (R 3.6.1)                  
+#>  labeling      0.3        2014-08-23 [1] CRAN (R 3.5.0)                  
+#>  lattice       0.20-38    2018-11-04 [4] CRAN (R 3.5.1)                  
+#>  lazyeval      0.2.2      2019-03-15 [1] CRAN (R 3.6.1)                  
+#>  lifecycle     0.1.0      2019-08-01 [1] CRAN (R 3.6.1)                  
+#>  lubridate   * 1.7.4      2018-04-11 [1] CRAN (R 3.5.0)                  
+#>  magrittr    * 1.5        2014-11-22 [1] CRAN (R 3.5.0)                  
+#>  Matrix        1.2-18     2019-11-27 [4] CRAN (R 3.6.1)                  
+#>  memoise       1.1.0      2017-04-21 [1] CRAN (R 3.5.0)                  
+#>  mgcv          1.8-24     2018-06-18 [1] CRAN (R 3.5.1)                  
+#>  modelr        0.1.5      2019-08-08 [1] CRAN (R 3.6.1)                  
+#>  munsell       0.5.0      2018-06-12 [1] CRAN (R 3.6.1)                  
+#>  nlme          3.1-142    2019-11-07 [4] CRAN (R 3.6.1)                  
+#>  pillar        1.4.2      2019-06-29 [1] CRAN (R 3.6.1)                  
+#>  pkgbuild      1.0.3      2019-03-20 [1] CRAN (R 3.6.1)                  
+#>  pkgconfig     2.0.3      2019-09-22 [1] CRAN (R 3.6.1)                  
+#>  pkgload       1.0.2      2018-10-29 [1] CRAN (R 3.6.1)                  
+#>  prettyunits   1.0.2      2015-07-13 [1] CRAN (R 3.5.2)                  
+#>  processx      3.4.1      2019-07-18 [1] CRAN (R 3.6.1)                  
+#>  ps            1.3.0      2018-12-21 [1] CRAN (R 3.6.1)                  
+#>  purrr       * 0.3.3      2019-10-18 [1] CRAN (R 3.6.1)                  
+#>  R6            2.4.1      2019-11-12 [1] CRAN (R 3.6.1)                  
+#>  Rcpp          1.0.3      2019-11-08 [1] CRAN (R 3.6.1)                  
+#>  readr       * 1.3.1      2018-12-21 [1] CRAN (R 3.6.1)                  
+#>  readxl        1.3.1      2019-03-13 [1] CRAN (R 3.6.1)                  
+#>  remotes       2.1.0      2019-06-24 [1] CRAN (R 3.6.1)                  
+#>  rlang         0.4.1      2019-10-24 [1] CRAN (R 3.6.1)                  
+#>  rmarkdown     1.16       2019-10-01 [1] CRAN (R 3.6.1)                  
+#>  rprojroot     1.3-2      2018-01-03 [1] CRAN (R 3.5.0)                  
+#>  rstudioapi    0.10       2019-03-19 [1] CRAN (R 3.6.1)                  
+#>  rticles       0.9.1      2019-07-22 [1] Github (rstudio/rticles@8d56fc6)
+#>  rvest         0.3.5      2019-11-08 [1] CRAN (R 3.6.1)                  
+#>  scales        1.0.0      2018-08-09 [1] CRAN (R 3.6.1)                  
+#>  sessioninfo   1.1.1      2018-11-05 [1] CRAN (R 3.6.1)                  
+#>  stringi       1.4.3      2019-03-12 [1] CRAN (R 3.6.1)                  
+#>  stringr     * 1.4.0      2019-02-10 [1] CRAN (R 3.6.1)                  
+#>  survival    * 2.44-1.1   2019-04-01 [4] CRAN (R 3.6.1)                  
+#>  testthat      2.2.0      2019-07-22 [1] CRAN (R 3.6.1)                  
+#>  tibble      * 2.1.3      2019-06-06 [1] CRAN (R 3.6.1)                  
+#>  tidyr       * 1.0.0      2019-09-11 [1] CRAN (R 3.6.1)                  
+#>  tidyselect    0.2.5      2018-10-11 [1] CRAN (R 3.5.2)                  
+#>  tidyverse   * 1.2.1      2017-11-14 [1] CRAN (R 3.5.0)                  
+#>  usethis       1.5.1      2019-07-04 [1] CRAN (R 3.6.1)                  
+#>  vctrs         0.2.0      2019-07-05 [1] CRAN (R 3.6.1)                  
+#>  VGAM          1.1-1      2019-02-18 [1] CRAN (R 3.6.1)                  
+#>  withr         2.1.2      2018-03-15 [1] CRAN (R 3.5.0)                  
+#>  xfun          0.8        2019-06-25 [1] CRAN (R 3.6.1)                  
+#>  xml2          1.2.2      2019-08-09 [1] CRAN (R 3.6.1)                  
+#>  yaml          2.2.0      2018-07-25 [1] CRAN (R 3.6.1)                  
+#>  zeallot       0.1.0      2018-01-28 [1] CRAN (R 3.5.1)                  
 #> 
-#> [1] /home/turgeonm/R/x86_64-pc-linux-gnu-library/3.6
+#> [1] /home/mturgeon/Rlibs
 #> [2] /usr/local/lib/R/site-library
@@
 \begin{CodeOutput}
-#> Local:    review-max /home/turgeonm/Documents/git_repos/cbpaper
-#> Head:     [d0d7b8a] 2019-12-05: add analysis of stanford data
+#> Local:    review-max /home/mturgeon/Documents/git_repositories/cbpaper
+#> Head:     [f601653] 2019-12-08: Merge branch 'review-max' of https://github.com/sahirbhatnagar/cbpaper into review-max
 \end{CodeOutput}

### 2019-12-08 analysis/paper/paper.Rmd
fix table captions

--- 2019-12-06 analysis/paper/paper.Rmd
@@
 
-# Code formatting
+<!-- # Code formatting -->
 
-Don't use markdown, instead use the more precise latex commands:
+<!-- Don't use markdown, instead use the more precise latex commands: -->
 
-* \proglang{Java}
-* \pkg{plyr}
-* \code{print("abc")}
+<!-- * \proglang{Java} -->
+<!-- * \pkg{plyr} -->
+<!-- * \code{print("abc")} -->
+
 ```{r setup, include=FALSE}
@@
 
-hazard ratios
+  1. hazard ratios
   2. If, however, the absolute risks are of interest, they have to be recovered using the semi-parametric Breslow estimator
-3. Alternative approaches for fitting flexible hazard models for estimating absolute risks, not requiring this two-step approach? Yes!
-[@hanley2009fitting]
+  3. Alternative approaches for fitting flexible hazard models for estimating absolute risks, not requiring this two-step approach? Yes! [@hanley2009fitting]
 
@@
 Besides these three main categories, there are extra features of interest as well.
-Table XXX summarizes the findings.
+Table \ref{tab:surv-pkgs} summarizes the findings.
 
-Many packages exist that implement semi-parametric cox models.
-Of those, glmnet [-@regpathcox]???, glmpath [-@park_hastie], penalized [-@l1penal], RiskRegression [-@gerds_blanche] introduce penalization in their model fitting.
-CoxRidge [-@perperoglou] , rstpm2 [-@clements_liu] and survival [-@survival-package] include penalization, but they implement the extended cox model permitting intuitive non proportional hazard modeling while the previous four do not.
-In terms of penalization, the benefit of casebase comes from its flexibility.
-Linear models have a wide tool set, considerably larger than those made for cox regression (*reference for number of packages between their task views, regression is spread out across Multivariate, SocialSciences, Robust among others.
-The focus of this task view is on Optimization Infrastructure Packages , General Purpose Continuous Solvers , Mathematical Programming Solvers , and Specific Applications in Optimization , or Multi Objective Optimization*).
-Casebase permits many regularization tools developed for linear regression to be repurposed within a survival context.
+Many packages exist that implement semi-parametric Cox models.
+Of those, \pkg{glmnet} [-@regpathcox]???, \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche] introduce penalization in their model fitting.
+\pkg{CoxRidge} [-@perperoglou], \pkg{rstpm2} [-@clements_liu] and \pkg{survival} [-@survival-package] include penalization, but they implement the extended cox model permitting intuitive non proportional hazard modeling while the previous four do not.
+In terms of penalization, the benefits of \pkg{casebase} comes from its flexibility.
+Linear models have a wide tool set, considerably larger than those made for Cox regression.
+<!--(*reference for number of packages between their task views, regression is spread out across Multivariate, SocialSciences, Robust among others.
+The focus of this task view is on Optimization Infrastructure Packages , General Purpose Continuous Solvers , Mathematical Programming Solvers , and Specific Applications in Optimization , or Multi Objective Optimization*)--> \pkg{casebase} permits many regularization tools developed for linear regression to be repurposed within a survival context.
 
 Parametric models permit non-proportional hazards and are implemented in a handful of packages.
-Of interest are CFC [-@sharabiani_mahani_2019], flexsurv [-@flexsurv] , SmoothHazard [-@smoothHazard], Rsptm2 [-@clements_liu] and Survival [-@survival-package].
-Smooth hazard  is limited to weibull distributions[-@smoothHazard], where as flexsurv and survival contain implementations for users to supply any distribution they would like [-@survival-package], [-@flexsurv].
-Flexsurv, smooth hazard, mets and rstpm2 can use splines, permitting a flexible fit over time [-@flexsurv], [-@smoothHazard], [-@scheike2014estimating], [-@clements_liu] .
-Though casebase is limited to 3 distributions at the moment, it also uses splines allowing for distributions that may not be consistent over survival time.
+Of interest are \pkg{CFC} [-@sharabiani_mahani_2019], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu] and \pkg{survival} [-@survival-package].
+\pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas \pkg{flexsurv} and \pkg{survival} contain implementations for users to supply any distribution they would like [-@survival-package], [-@flexsurv].
+\pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} can use splines, permitting a flexible fit over time [-@flexsurv], [-@smoothHazard], [-@scheike2014estimating], [-@clements_liu].
+Though \pkg{casebase} is limited to three distributions at the moment, it also uses splines allowing for distributions that may not be consistent over survival time.
 
-Of the methods mentioned so far, only CFC,flexsurv, mets and survival contain implementations for competing risks [-@sharabiani_mahani_2019], [-@flexsurv],[-@scheike2014estimating] ,[-@survival-package] (*how is it done for parametric models?*).
-The differentiating factor between these packages and casebase is that casebase handles competing risks through multiple logistic regression.
-Smooth hazard, survival and Rstpm2 can handle interval censoring, where as casebase currently can not.
-Interval censoring is simple feature to implement and is a potential improvement to add to casebase.
-(There's also multistate models, which I haven't mentioned in this paper)
+Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets} and \pkg{survival} contain implementations for competing risks [-@sharabiani_mahani_2019], [-@flexsurv], [-@scheike2014estimating], [-@survival-package].
+The differentiating factor between these packages and \pkg{casebase} is that \pkg{casebase} handles competing risks through multiple logistic regression.
+\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval censoring, whereas \pkg{casebase} currently cannot.
+Interval censoring is a simple feature to implement and is a potential improvement to add to \pkg{casebase}.
 
-Of all the packages listed in the table, CFC , Flexsurv, mets, RiskRegression, rstpm2 and survival have implementations for cumulative incidence [-@sharabiani_mahani_2019], [-@flexsurv], [-@scheike2014estimating], [-@gerds_blanche], [-@clements_liu], [-@survival-package].
-Since casebase is parametric, the baseline hazard is known and requires integration to retrieve the cumulative incidence.
-This is similar to the approaches of CFC and flexsurv [-@sharabiani_mahani_2019], [-@flexsurv], while the other packages use semi-parametric approaches to retrieve the baseline after fitting [-@scheike2014estimating], [-@gerds_blanche].
-rstpm2 and survival use both for their respective parametric and semi-parametric model implementations [-@clements_liu], [-@survival-package].
+Of all the packages listed in Table \ref{tab:surv-pkgs}, \pkg{CFC}, \pkg{Flexsurv}, \pkg{mets}, \pkg{RiskRegression}, \pkg{rstpm2} and \pkg{survival} have implementations for cumulative incidence [-@sharabiani_mahani_2019], [-@flexsurv], [-@scheike2014estimating], [-@gerds_blanche], [-@clements_liu], [-@survival-package].
+Since \pkg{casebase} is parametric, the baseline hazard is known and requires integration to retrieve the cumulative incidence.
+This is similar to the approaches of \pkg{CFC} and \pkg{flexsurv} [-@sharabiani_mahani_2019], [-@flexsurv], while the other packages use semi-parametric approaches to retrieve the baseline after fitting [-@scheike2014estimating], [-@gerds_blanche].
+\pkg{rstpm2} and \pkg{survival} use both for their respective parametric and semi-parametric model implementations [-@clements_liu], [-@survival-package].
 
-\begin{table}[]
+\begin{table}[ht]
 \centering
@@
 \caption{Different features of interest in various survival packages.}
-\label{tab:my-table}
+\label{tab:surv-pkgs}
 \end{table}
@@
 The SUPPORT dataset will be used to demonstrate regularization within casebase (REF), while comparing their absolute risk preditions for a new individual.
-A description of each variable can be found in table XXX1 and a breakdown of each categorical variable in table XXX2.
+A description of each variable can be found in Table~\ref{tab:support1} and a breakdown of each categorical variable in Table~\ref{tab:support2}.
 The data was downloaded from the main author's website, and it is also available as part of the casebase package.
 
-\begin{table}[]
+\begin{table}[ht]
 \centering
@@
 \caption{A description of each variable in the SUPPORT dataset, taken from (REF).}
-\label{tab:my-table}
+\label{tab:support1}
 \end{table}
 
-\begin{table}[]
+\begin{table}[ht]
 \centering
@@
 \caption{A description of each level within each categorical variable in the dataset, taken from REF.}
-\label{tab:my-table}
+\label{tab:support2}
 \end{table}
@@
 ```
+
 # Discussion

### 2019-12-08 analysis/paper/paper.tex
fix table captions

--- 2019-12-06 analysis/paper/paper.tex
@@
 
-\hypertarget{code-formatting}{%
-\section{Code formatting}\label{code-formatting}}
-
-Don't use markdown, instead use the more precise latex commands:
-
-\begin{itemize}
-\item
-  \proglang{Java}
-\item
-  \pkg{plyr}
-\item
-  \code{print("abc")}
-\end{itemize}
-
 \hypertarget{introduction}{%
@@
 
-hazard ratios 2. If, however, the absolute risks are of interest, they
-have to be recovered using the semi-parametric Breslow estimator 3.
-Alternative approaches for fitting flexible hazard models for estimating
-absolute risks, not requiring this two-step approach? Yes!
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  hazard ratios
+\item
+  If, however, the absolute risks are of interest, they have to be
+  recovered using the semi-parametric Breslow estimator
+\item
+  Alternative approaches for fitting flexible hazard models for
+  estimating absolute risks, not requiring this two-step approach? Yes!
   \citep{hanley2009fitting}
+\end{enumerate}
 
@@
 hazards, regularization and competing risks. Besides these three main
-categories, there are extra features of interest as well. Table XXX
-summarizes the findings.
-
-Many packages exist that implement semi-parametric cox models. Of those,
-glmnet \citeyearpar{regpathcox}???, glmpath \citeyearpar{park_hastie},
-penalized \citeyearpar{l1penal}, RiskRegression
-\citeyearpar{gerds_blanche} introduce penalization in their model
-fitting. CoxRidge \citeyearpar{perperoglou} , rstpm2
-\citeyearpar{clements_liu} and survival \citeyearpar{survival-package}
-include penalization, but they implement the extended cox model
-permitting intuitive non proportional hazard modeling while the previous
-four do not. In terms of penalization, the benefit of casebase comes
-from its flexibility. Linear models have a wide tool set, considerably
-larger than those made for cox regression (\emph{reference for number of
-packages between their task views, regression is spread out across
-Multivariate, SocialSciences, Robust among others. The focus of this
-task view is on Optimization Infrastructure Packages , General Purpose
-Continuous Solvers , Mathematical Programming Solvers , and Specific
-Applications in Optimization , or Multi Objective Optimization}).
-Casebase permits many regularization tools developed for linear
-regression to be repurposed within a survival context.
+categories, there are extra features of interest as well. Table
+\ref{tab:surv-pkgs} summarizes the findings.
+
+Many packages exist that implement semi-parametric Cox models. Of those,
+\pkg{glmnet} \citeyearpar{regpathcox}???, \pkg{glmpath}
+\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
+\pkg{RiskRegression} \citeyearpar{gerds_blanche} introduce penalization
+in their model fitting. \pkg{CoxRidge} \citeyearpar{perperoglou},
+\pkg{rstpm2} \citeyearpar{clements_liu} and \pkg{survival}
+\citeyearpar{survival-package} include penalization, but they implement
+the extended cox model permitting intuitive non proportional hazard
+modeling while the previous four do not. In terms of penalization, the
+benefits of \pkg{casebase} comes from its flexibility. Linear models
+have a wide tool set, considerably larger than those made for Cox
+regression. \pkg{casebase} permits many regularization tools developed
+for linear regression to be repurposed within a survival context.
 
 Parametric models permit non-proportional hazards and are implemented in
-a handful of packages. Of interest are CFC
-\citeyearpar{sharabiani_mahani_2019}, flexsurv \citeyearpar{flexsurv} ,
-SmoothHazard \citeyearpar{smoothHazard}, Rsptm2
-\citeyearpar{clements_liu} and Survival \citeyearpar{survival-package}.
-Smooth hazard is limited to weibull
-distributions\citeyearpar{smoothHazard}, where as flexsurv and survival
-contain implementations for users to supply any distribution they would
-like \citeyearpar{survival-package}, \citeyearpar{flexsurv}. Flexsurv,
-smooth hazard, mets and rstpm2 can use splines, permitting a flexible
-fit over time \citeyearpar{flexsurv}, \citeyearpar{smoothHazard},
-\citeyearpar{scheike2014estimating}, \citeyearpar{clements_liu} . Though
-casebase is limited to 3 distributions at the moment, it also uses
-splines allowing for distributions that may not be consistent over
+a handful of packages. Of interest are \pkg{CFC}
+\citeyearpar{sharabiani_mahani_2019}, \pkg{flexsurv}
+\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
+\pkg{rsptm2} \citeyearpar{clements_liu} and \pkg{survival}
+\citeyearpar{survival-package}. \pkg{SmoothHazard} is limited to Weibull
+distributions \citeyearpar{smoothHazard}, whereas \pkg{flexsurv} and
+\pkg{survival} contain implementations for users to supply any
+distribution they would like \citeyearpar{survival-package},
+\citeyearpar{flexsurv}. \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets}
+and \pkg{rstpm2} can use splines, permitting a flexible fit over time
+\citeyearpar{flexsurv}, \citeyearpar{smoothHazard},
+\citeyearpar{scheike2014estimating}, \citeyearpar{clements_liu}. Though
+\pkg{casebase} is limited to three distributions at the moment, it also
+uses splines allowing for distributions that may not be consistent over
 survival time.
 
-Of the methods mentioned so far, only CFC,flexsurv, mets and survival
-contain implementations for competing risks
-\citeyearpar{sharabiani_mahani_2019},
-\citeyearpar{flexsurv},\citeyearpar{scheike2014estimating}
-,\citeyearpar{survival-package} (\emph{how is it done for parametric
-models?}). The differentiating factor between these packages and
-casebase is that casebase handles competing risks through multiple
-logistic regression. Smooth hazard, survival and Rstpm2 can handle
-interval censoring, where as casebase currently can not. Interval
-censoring is simple feature to implement and is a potential improvement
-to add to casebase. (There's also multistate models, which I haven't
-mentioned in this paper)
-
-Of all the packages listed in the table, CFC , Flexsurv, mets,
-RiskRegression, rstpm2 and survival have implementations for cumulative
-incidence \citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
+Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
+\pkg{mets} and \pkg{survival} contain implementations for competing
+risks \citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
+\citeyearpar{scheike2014estimating}, \citeyearpar{survival-package}. The
+differentiating factor between these packages and \pkg{casebase} is that
+\pkg{casebase} handles competing risks through multiple logistic
+regression. \pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can
+handle interval censoring, whereas \pkg{casebase} currently cannot.
+Interval censoring is a simple feature to implement and is a potential
+improvement to add to \pkg{casebase}.
+
+Of all the packages listed in Table \ref{tab:surv-pkgs}, \pkg{CFC},
+\pkg{Flexsurv}, \pkg{mets}, \pkg{RiskRegression}, \pkg{rstpm2} and
+\pkg{survival} have implementations for cumulative incidence
+\citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
 \citeyearpar{scheike2014estimating}, \citeyearpar{gerds_blanche},
 \citeyearpar{clements_liu}, \citeyearpar{survival-package}. Since
-casebase is parametric, the baseline hazard is known and requires
+\pkg{casebase} is parametric, the baseline hazard is known and requires
 integration to retrieve the cumulative incidence. This is similar to the
-approaches of CFC and flexsurv \citeyearpar{sharabiani_mahani_2019},
-\citeyearpar{flexsurv}, while the other packages use semi-parametric
-approaches to retrieve the baseline after fitting
-\citeyearpar{scheike2014estimating}, \citeyearpar{gerds_blanche}. rstpm2
-and survival use both for their respective parametric and
-semi-parametric model implementations \citeyearpar{clements_liu},
+approaches of \pkg{CFC} and \pkg{flexsurv}
+\citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv}, while the
+other packages use semi-parametric approaches to retrieve the baseline
+after fitting \citeyearpar{scheike2014estimating},
+\citeyearpar{gerds_blanche}. \pkg{rstpm2} and \pkg{survival} use both
+for their respective parametric and semi-parametric model
+implementations \citeyearpar{clements_liu},
 \citeyearpar{survival-package}.
 
-\begin{table}[]
+\begin{table}[ht]
 \centering
@@
 risk preditions for a new individual. A description of each variable can
-be found in table XXX1 and a breakdown of each categorical variable in
-table XXX2. The data was downloaded from the main author's website, and
-it is also available as part of the casebase package.
-
-\begin{table}[]
+be found in Table\textasciitilde{}\ref{tab:support1} and a breakdown of
+each categorical variable in Table\textasciitilde{}\ref{tab:support2}.
+The data was downloaded from the main author's website, and it is also
+available as part of the casebase package.
+
+\begin{table}[ht]
 \centering
@@
 
-This report was generated on 2019-12-06 16:16:07 using the following
+This report was generated on 2019-12-08 12:21:42 using the following
 computational environment and dependencies:
@@
 \begin{CodeOutput}
-#> - Session info ---------------------------------------------------------------
+#> - Session info ----------------------------------------------------------
 #>  setting  value                       
-#>  version  R version 3.6.0 (2019-04-26)
-#>  os       macOS High Sierra 10.13.6   
-#>  system   x86_64, darwin15.6.0        
+#>  version  R version 3.6.1 (2019-07-05)
+#>  os       Ubuntu 18.04.3 LTS          
+#>  system   x86_64, linux-gnu           
 #>  ui       X11                         
@@
 #>  ctype    en_CA.UTF-8                 
-#>  tz       America/Montreal            
-#>  date     2019-12-06                  
+#>  tz       America/Winnipeg            
+#>  date     2019-12-08                  
 #> 
-#> - Packages -------------------------------------------------------------------
-#>  package     * version    date       lib
-#>  assertthat    0.2.1      2019-03-21 [1]
-#>  backports     1.1.5      2019-10-02 [1]
-#>  broom         0.5.2      2019-04-07 [1]
-#>  callr         3.3.2      2019-09-22 [1]
-#>  casebase    * 0.2.1.9001 2019-08-20 [1]
-#>  cellranger    1.1.0      2016-07-27 [1]
-#>  cli           1.1.0      2019-03-19 [1]
-#>  codetools     0.2-16     2018-12-24 [1]
-#>  colorspace    1.4-1      2019-03-18 [1]
-#>  crayon        1.3.4      2017-09-16 [1]
-#>  data.table    1.12.6     2019-10-18 [1]
-#>  DBI           1.0.0      2018-05-02 [1]
-#>  dbplyr        1.4.2      2019-06-17 [1]
-#>  desc          1.2.0      2018-05-01 [1]
-#>  devtools      2.2.1      2019-09-24 [1]
-#>  digest        0.6.23     2019-11-23 [1]
-#>  dplyr       * 0.8.3      2019-07-04 [1]
-#>  ellipsis      0.3.0      2019-09-20 [1]
-#>  evaluate      0.14       2019-05-28 [1]
-#>  forcats     * 0.4.0      2019-02-17 [1]
-#>  foreach       1.4.7      2019-07-27 [1]
-#>  fs            1.3.1      2019-05-06 [1]
-#>  generics      0.0.2      2018-11-29 [1]
-#>  ggplot2     * 3.2.1      2019-08-10 [1]
-#>  glmnet      * 3.0-1      2019-11-15 [1]
-#>  glue          1.3.1      2019-03-12 [1]
-#>  gtable        0.3.0      2019-03-25 [1]
-#>  haven         2.2.0      2019-11-08 [1]
-#>  hms           0.5.2      2019-10-30 [1]
-#>  htmltools     0.4.0      2019-10-04 [1]
-#>  httr          1.4.1      2019-08-05 [1]
-#>  iterators     1.0.12     2019-07-26 [1]
-#>  jsonlite      1.6        2018-12-07 [1]
-#>  knitr         1.26       2019-11-12 [1]
-#>  lattice       0.20-38    2018-11-04 [1]
-#>  lazyeval      0.2.2      2019-03-15 [1]
-#>  lifecycle     0.1.0      2019-08-01 [1]
-#>  lubridate     1.7.4      2018-04-11 [1]
-#>  magrittr    * 1.5        2014-11-22 [1]
-#>  Matrix      * 1.2-17     2019-03-22 [1]
-#>  memoise       1.1.0      2017-04-21 [1]
-#>  mgcv          1.8-31     2019-11-09 [1]
-#>  modelr        0.1.5      2019-08-08 [1]
-#>  munsell       0.5.0      2018-06-12 [1]
-#>  nlme          3.1-142    2019-11-07 [1]
-#>  pillar        1.4.2      2019-06-29 [1]
-#>  pkgbuild      1.0.6      2019-10-09 [1]
-#>  pkgconfig     2.0.3      2019-09-22 [1]
-#>  pkgload       1.0.2      2018-10-29 [1]
-#>  prettyunits   1.0.2      2015-07-13 [1]
-#>  processx      3.4.1      2019-07-18 [1]
-#>  ps            1.3.0      2018-12-21 [1]
-#>  purrr       * 0.3.3      2019-10-18 [1]
-#>  R6            2.4.1      2019-11-12 [1]
-#>  Rcpp          1.0.3      2019-11-08 [1]
-#>  readr       * 1.3.1      2018-12-21 [1]
-#>  readxl        1.3.1      2019-03-13 [1]
-#>  remotes       2.1.0      2019-06-24 [1]
-#>  reprex        0.3.0      2019-05-16 [1]
-#>  rlang         0.4.2      2019-11-23 [1]
-#>  rmarkdown     1.17       2019-11-13 [1]
-#>  rprojroot     1.3-2      2018-01-03 [1]
-#>  rstudioapi    0.10       2019-03-19 [1]
-#>  rticles       0.12       2019-11-12 [1]
-#>  rvest         0.3.5      2019-11-08 [1]
-#>  scales        1.1.0      2019-11-18 [1]
-#>  sessioninfo   1.1.1      2018-11-05 [1]
-#>  shape         1.4.4      2018-02-07 [1]
-#>  stringi       1.4.3      2019-03-12 [1]
-#>  stringr     * 1.4.0      2019-02-10 [1]
-#>  survival    * 3.1-7      2019-11-09 [1]
-#>  testthat      2.3.0      2019-11-05 [1]
-#>  tibble      * 2.1.3      2019-06-06 [1]
-#>  tidyr       * 1.0.0      2019-09-11 [1]
-#>  tidyselect    0.2.5      2018-10-11 [1]
-#>  tidyverse   * 1.3.0      2019-11-21 [1]
-#>  usethis       1.5.1      2019-07-04 [1]
-#>  vctrs         0.2.0      2019-07-05 [1]
-#>  VGAM          1.1-2      2019-11-21 [1]
-#>  withr         2.1.2      2018-03-15 [1]
-#>  xfun          0.11       2019-11-12 [1]
-#>  xml2          1.2.2      2019-08-09 [1]
-#>  yaml          2.2.0      2018-07-25 [1]
-#>  zeallot       0.1.0      2018-01-28 [1]
-#>  source                                  
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  Github (sahirbhatnagar/casebase@aecb816)
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
+#> - Packages --------------------------------------------------------------
+#>  package     * version    date       lib source                          
+#>  assertthat    0.2.1      2019-03-21 [1] CRAN (R 3.6.1)                  
+#>  backports     1.1.5      2019-10-02 [1] CRAN (R 3.6.1)                  
+#>  broom         0.5.2      2019-04-07 [1] CRAN (R 3.6.1)                  
+#>  callr         3.3.1      2019-07-18 [1] CRAN (R 3.6.1)                  
+#>  casebase    * 0.2.1.9001 2019-07-29 [1] local                           
+#>  cellranger    1.1.0      2016-07-27 [1] CRAN (R 3.5.0)                  
+#>  cli           1.1.0      2019-03-19 [1] CRAN (R 3.6.1)                  
+#>  codetools     0.2-16     2018-12-24 [4] CRAN (R 3.5.2)                  
+#>  colorspace    1.4-1      2019-03-18 [1] CRAN (R 3.6.1)                  
+#>  crayon        1.3.4      2017-09-16 [1] CRAN (R 3.5.0)                  
+#>  data.table    1.12.2     2019-04-07 [1] CRAN (R 3.6.1)                  
+#>  desc          1.2.0      2018-05-01 [1] CRAN (R 3.5.0)                  
+#>  devtools      2.1.0      2019-07-06 [1] CRAN (R 3.6.1)                  
+#>  digest        0.6.22     2019-10-21 [1] CRAN (R 3.6.1)                  
+#>  dplyr       * 0.8.3      2019-07-04 [1] CRAN (R 3.6.1)                  
+#>  evaluate      0.14       2019-05-28 [1] CRAN (R 3.6.1)                  
+#>  forcats     * 0.4.0      2019-02-17 [1] CRAN (R 3.6.1)                  
+#>  foreach     * 1.4.4      2017-12-12 [1] CRAN (R 3.5.0)                  
+#>  fs            1.3.1      2019-05-06 [1] CRAN (R 3.6.1)                  
+#>  generics      0.0.2      2018-11-29 [1] CRAN (R 3.6.1)                  
+#>  ggplot2     * 3.2.1      2019-08-10 [1] CRAN (R 3.6.1)                  
+#>  glmnet      * 2.0-18     2019-05-20 [1] CRAN (R 3.6.1)                  
+#>  glue          1.3.1      2019-03-12 [1] CRAN (R 3.6.1)                  
+#>  gtable        0.3.0      2019-03-25 [1] CRAN (R 3.6.1)                  
+#>  haven         2.1.1      2019-07-04 [1] CRAN (R 3.6.1)                  
+#>  hms           0.5.2      2019-10-30 [1] CRAN (R 3.6.1)                  
+#>  htmltools     0.3.6      2017-04-28 [1] CRAN (R 3.5.0)                  
+#>  httr          1.4.1      2019-08-05 [1] CRAN (R 3.6.1)                  
+#>  iterators     1.0.9      2017-12-12 [1] CRAN (R 3.5.0)                  
+#>  jsonlite      1.6        2018-12-07 [1] CRAN (R 3.6.1)                  
+#>  knitr         1.23       2019-05-18 [1] CRAN (R 3.6.1)                  
+#>  lattice       0.20-38    2018-11-04 [4] CRAN (R 3.5.1)                  
+#>  lazyeval      0.2.2      2019-03-15 [1] CRAN (R 3.6.1)                  
+#>  lifecycle     0.1.0      2019-08-01 [1] CRAN (R 3.6.1)                  
+#>  lubridate     1.7.4      2018-04-11 [1] CRAN (R 3.5.0)                  
+#>  magrittr    * 1.5        2014-11-22 [1] CRAN (R 3.5.0)                  
+#>  Matrix      * 1.2-18     2019-11-27 [4] CRAN (R 3.6.1)                  
+#>  memoise       1.1.0      2017-04-21 [1] CRAN (R 3.5.0)                  
+#>  mgcv          1.8-24     2018-06-18 [1] CRAN (R 3.5.1)                  
+#>  modelr        0.1.5      2019-08-08 [1] CRAN (R 3.6.1)                  
+#>  munsell       0.5.0      2018-06-12 [1] CRAN (R 3.6.1)                  
+#>  nlme          3.1-142    2019-11-07 [4] CRAN (R 3.6.1)                  
+#>  pillar        1.4.2      2019-06-29 [1] CRAN (R 3.6.1)                  
+#>  pkgbuild      1.0.3      2019-03-20 [1] CRAN (R 3.6.1)                  
+#>  pkgconfig     2.0.3      2019-09-22 [1] CRAN (R 3.6.1)                  
+#>  pkgload       1.0.2      2018-10-29 [1] CRAN (R 3.6.1)                  
+#>  prettyunits   1.0.2      2015-07-13 [1] CRAN (R 3.5.2)                  
+#>  processx      3.4.1      2019-07-18 [1] CRAN (R 3.6.1)                  
+#>  ps            1.3.0      2018-12-21 [1] CRAN (R 3.6.1)                  
+#>  purrr       * 0.3.3      2019-10-18 [1] CRAN (R 3.6.1)                  
+#>  R6            2.4.1      2019-11-12 [1] CRAN (R 3.6.1)                  
+#>  Rcpp          1.0.3      2019-11-08 [1] CRAN (R 3.6.1)                  
+#>  readr       * 1.3.1      2018-12-21 [1] CRAN (R 3.6.1)                  
+#>  readxl        1.3.1      2019-03-13 [1] CRAN (R 3.6.1)                  
+#>  remotes       2.1.0      2019-06-24 [1] CRAN (R 3.6.1)                  
+#>  rlang         0.4.1      2019-10-24 [1] CRAN (R 3.6.1)                  
+#>  rmarkdown     1.16       2019-10-01 [1] CRAN (R 3.6.1)                  
+#>  rprojroot     1.3-2      2018-01-03 [1] CRAN (R 3.5.0)                  
+#>  rstudioapi    0.10       2019-03-19 [1] CRAN (R 3.6.1)                  
+#>  rticles       0.9.1      2019-07-22 [1] Github (rstudio/rticles@8d56fc6)
+#>  rvest         0.3.5      2019-11-08 [1] CRAN (R 3.6.1)                  
+#>  scales        1.0.0      2018-08-09 [1] CRAN (R 3.6.1)                  
+#>  sessioninfo   1.1.1      2018-11-05 [1] CRAN (R 3.6.1)                  
+#>  stringi       1.4.3      2019-03-12 [1] CRAN (R 3.6.1)                  
+#>  stringr     * 1.4.0      2019-02-10 [1] CRAN (R 3.6.1)                  
+#>  survival    * 2.44-1.1   2019-04-01 [4] CRAN (R 3.6.1)                  
+#>  testthat      2.2.0      2019-07-22 [1] CRAN (R 3.6.1)                  
+#>  tibble      * 2.1.3      2019-06-06 [1] CRAN (R 3.6.1)                  
+#>  tidyr       * 1.0.0      2019-09-11 [1] CRAN (R 3.6.1)                  
+#>  tidyselect    0.2.5      2018-10-11 [1] CRAN (R 3.5.2)                  
+#>  tidyverse   * 1.2.1      2017-11-14 [1] CRAN (R 3.5.0)                  
+#>  usethis       1.5.1      2019-07-04 [1] CRAN (R 3.6.1)                  
+#>  vctrs         0.2.0      2019-07-05 [1] CRAN (R 3.6.1)                  
+#>  VGAM          1.1-1      2019-02-18 [1] CRAN (R 3.6.1)                  
+#>  withr         2.1.2      2018-03-15 [1] CRAN (R 3.5.0)                  
+#>  xfun          0.8        2019-06-25 [1] CRAN (R 3.6.1)                  
+#>  xml2          1.2.2      2019-08-09 [1] CRAN (R 3.6.1)                  
+#>  yaml          2.2.0      2018-07-25 [1] CRAN (R 3.6.1)                  
+#>  zeallot       0.1.0      2018-01-28 [1] CRAN (R 3.5.1)                  
 #> 
-#> [1] /Library/Frameworks/R.framework/Versions/3.6/Resources/library
+#> [1] /home/mturgeon/Rlibs
+#> [2] /usr/local/lib/R/site-library
+#> [3] /usr/lib/R/site-library
+#> [4] /usr/lib/R/library
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> Local:    master /Users/jislam/Desktop/Gits/cbpaper
-#> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
-#> Head:     [22bfed0] 2019-12-06: currentDraftEdits
+#> Local:    Jesse-Islam-master /home/mturgeon/Documents/git_repositories/cbpaper
+#> Head:     [71a5a9c] 2019-12-06: Lit-Review Edits
 \end{CodeOutput}

### 2019-12-08 analysis/paper/paper.tex

--- 2019-12-08 analysis/paper/paper.tex
@@
 
-\usepackage{amsmath} \usepackage{longtable}
+\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx}
 
@@
 \item
-  Time matching/risk set sampling (including Cox partial likelihood)
-  eliminates the baseline hazard from the likelihood expression for the
   hazard ratios
@@
 
+\hypertarget{literature-review}{%
+\section{Literature review}\label{literature-review}}
+
+To conduct our review on current methods available in survival analysis,
+a scraper was coded to examine all packages in the survival task view on
+CRAN \citeyearpar{survTaskView}. There was a total of 259 packages, many
+of which are not relevant to the problems casebase addresses. To filter
+these packages, a scraper was used to search for competing risks,
+cumulative incidence functions, non- proportional hazards and
+penalization. This reduced our set of 259 packages down to 60. From
+here, we manually examined each package to determine which ones are
+comparable to casebase, with either a different approach to the same
+problems or future features we would like to add to casebase.
+
+These packages can belong to three main categories: non-proportional
+hazards, regularization and competing risks. Besides these three main
+categories, there are extra features of interest as well. Table
+\ref{tab:surv-pkgs} summarizes the findings.
+
+Many packages exist that implement semi-parametric Cox models. Of those,
+\pkg{glmnet} \citeyearpar{regpathcox}???, \pkg{glmpath}
+\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
+\pkg{RiskRegression} \citeyearpar{gerds_blanche} introduce penalization
+in their model fitting. \pkg{CoxRidge} \citeyearpar{perperoglou},
+\pkg{rstpm2} \citeyearpar{clements_liu} and \pkg{survival}
+\citeyearpar{survival-package} include penalization, but they implement
+the extended cox model permitting intuitive non proportional hazard
+modeling while the previous four do not. In terms of penalization, the
+benefits of \pkg{casebase} comes from its flexibility. Linear models
+have a wide tool set, considerably larger than those made for Cox
+regression. \pkg{casebase} permits many regularization tools developed
+for linear regression to be repurposed within a survival context.
+
+Parametric models permit non-proportional hazards and are implemented in
+a handful of packages. Of interest are \pkg{CFC}
+\citeyearpar{sharabiani_mahani_2019}, \pkg{flexsurv}
+\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
+\pkg{rsptm2} \citeyearpar{clements_liu} and \pkg{survival}
+\citeyearpar{survival-package}. \pkg{SmoothHazard} is limited to Weibull
+distributions \citeyearpar{smoothHazard}, whereas \pkg{flexsurv} and
+\pkg{survival} contain implementations for users to supply any
+distribution they would like \citeyearpar{survival-package},
+\citeyearpar{flexsurv}. \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets}
+and \pkg{rstpm2} can use splines, permitting a flexible fit over time
+\citeyearpar{flexsurv}, \citeyearpar{smoothHazard},
+\citeyearpar{scheike2014estimating}, \citeyearpar{clements_liu}. Though
+\pkg{casebase} is limited to three distributions at the moment, it also
+uses splines allowing for distributions that may not be consistent over
+survival time.
+
+Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
+\pkg{mets} and \pkg{survival} contain implementations for competing
+risks \citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
+\citeyearpar{scheike2014estimating}, \citeyearpar{survival-package}. The
+differentiating factor between these packages and \pkg{casebase} is that
+\pkg{casebase} handles competing risks through multiple logistic
+regression. \pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can
+handle interval censoring, whereas \pkg{casebase} currently cannot.
+Interval censoring is a simple feature to implement and is a potential
+improvement to add to \pkg{casebase}.
+
+Of all the packages listed in Table \ref{tab:surv-pkgs}, \pkg{CFC},
+\pkg{Flexsurv}, \pkg{mets}, \pkg{RiskRegression}, \pkg{rstpm2} and
+\pkg{survival} have implementations for cumulative incidence
+\citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
+\citeyearpar{scheike2014estimating}, \citeyearpar{gerds_blanche},
+\citeyearpar{clements_liu}, \citeyearpar{survival-package}. Since
+\pkg{casebase} is parametric, the baseline hazard is known and requires
+integration to retrieve the cumulative incidence. This is similar to the
+approaches of \pkg{CFC} and \pkg{flexsurv}
+\citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv}, while the
+other packages use semi-parametric approaches to retrieve the baseline
+after fitting \citeyearpar{scheike2014estimating},
+\citeyearpar{gerds_blanche}. \pkg{rstpm2} and \pkg{survival} use both
+for their respective parametric and semi-parametric model
+implementations \citeyearpar{clements_liu},
+\citeyearpar{survival-package}.
+
+\begin{table}[ht]
+\centering
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{ccccccccc}
+\hline
+\textbf{Package}        & \textbf{Competing Risks} & \textbf{Non-proportional} & \textbf{Penalization} & \textbf{Splines} & \textbf{Parametric} & \textbf{Semi-parametric} & \textbf{Interval/left Censoring} & \textbf{Absolute Risk}           \\
+\hline
+\textbf{casebase} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
+\textbf{CFC} \cite{sharabiani_mahani_2019}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
+\textbf{coxRidge} \cite{perperoglou}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{crrp} \cite{fu}           & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
+\textbf{fastcox} \cite{yi_zou}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{Flexrsurv} \cite{flexrsurv}      &                          & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
+\\
+\textbf{Flexsurv} \cite{flexsurv}       & x                        & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
+\\
+\textbf{glmnet} \cite{regpathcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{glmpath} \cite{park_hastie}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{mets} \cite{scheike2014estimating}           & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
+\textbf{penalized} \cite{goeman_meijer2019}      &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{RiskRegression} \cite{gerds_blanche} &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
+\textbf{Rstpm2} \cite{clements_liu}         &                          & x                         & x                     & x                & x                   & x                        & x                                & Cumm haz                         \\
+\textbf{SmoothHazard} \cite{smoothHazard}   &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
+\textbf{Survival} \cite{survival-package}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x
+\end{tabular}%
+}
+\caption{Different features of interest in various survival packages.}
+\label{tab:surv-pkgs}
+\end{table}
+
 \hypertarget{theoretical-details}{%
@@
 
-\hypertarget{case-study-3the-cancer-genome-atlas}{%
-\section{Case study 3--The Cancer Genome
-Atlas}\label{case-study-3the-cancer-genome-atlas}}
-
-\begin{itemize}
-\tightlist
-\item
-  Glmnet and TCGA data
-\end{itemize}
+\hypertarget{case-study-3study-to-understand-prognoses-preferences-outcomes-and-risks-of-treatment}{%
+\section{Case study 3--Study to Understand Prognoses Preferences
+Outcomes and Risks of
+Treatment}\label{case-study-3study-to-understand-prognoses-preferences-outcomes-and-risks-of-treatment}}
+
+We examined the Study to Understand Prognoses Preferences Outcomes and
+Risks of Treatment (SUPPORT) dataset (REF from main site). the SUPPORT
+dataset tracks death for individuals who are considered seriously ill
+within a hospital. Imputation was conducted using the default methods in
+the mice package in R, available on CRAN. Before imputation, there was a
+total of 1000 individuals and 35 variables. After imputation, we have
+474 observations where 67.5\% of individuals in the study experience the
+event of interest. The SUPPORT dataset will be used to demonstrate
+regularization within casebase (REF), while comparing their absolute
+risk preditions for a new individual. A description of each variable can
+be found in Table\textasciitilde{}\ref{tab:support1} and a breakdown of
+each categorical variable in Table\textasciitilde{}\ref{tab:support2}.
+The data was downloaded from the main author's website, and it is also
+available as part of the casebase package.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{ccccc}
+\hline
+\textbf{Name} & \textbf{Labels}                          & \textbf{Levels} & \textbf{Storage} & \textbf{NAs} \\
+\hline
+age           & Age                                      &                 & double           & 0            \\
+death         & Death at any time up to NDI date:31DEC94 &                 & double           & 0            \\
+sex           &                                          & 2               & integer          & 0            \\
+hospdead      & Death in Hospital                        &                 & double           & 0            \\
+slos          & Days from Study Entry to Discharge       &                 & double           & 0            \\
+d.time        & Days of Follow-Up                        &                 & double           & 0            \\
+dzgroup       &                                          & 8               & integer          & 0            \\
+dzclass       &                                          & 4               & integer          & 0            \\
+num.co        & number of comorbidities                  &                 & double           & 0            \\
+edu           & Years of Education                       &                 & double           & 202          \\
+income        &                                          & 4               & integer          & 349          \\
+scoma         & SUPPORT Coma Score based on Glasgow D3   &                 & double           & 0            \\
+charges       & Hospital Charges                         &                 & double           & 25           \\
+totcst        & Total RCC cost                           &                 & double           & 105          \\
+totmcst       & Total micro-cost                         &                 & double           & 372          \\
+avtisst       & Average TISS, Days 3-25                  &                 & double           & 6            \\
+race          &                                          & 5               & integer          & 5            \\
+meanbp        & Mean Arterial Blood Pressure Day 3       &                 & double           & 0            \\
+wblc          & White Blood Cell Count Day 3             &                 & double           & 24           \\
+hrt           & Heart Rate Day 3                         &                 & double           & 0            \\
+resp          & Respiration Rate Day 3                   &                 & double           & 0            \\
+temp          & Temperature (celcius) Day 3              &                 & double           & 0            \\
+pafi          & PaO2/(.01*FiO2) Day 3                    &                 & double           & 253          \\
+alb           & Serum Albumin Day 3                      &                 & double           & 378          \\
+bili          & Bilirubin Day 3                          &                 & double           & 297          \\
+crea          & Serum creatinine Day 3                   &                 & double           & 3            \\
+sod           & Serum sodium Day 3                       &                 & double           & 0            \\
+ph            & Serum pH (arterial) Day 3                &                 & double           & 250          \\
+glucose       & Glucose Day 3                            &                 & double           & 470          \\
+bun           & BUN Day 3                                &                 & double           & 455          \\
+urine         & Urine Output Day 3                       &                 & double           & 517          \\
+adlp          & ADL Patient Day 3                        &                 & double           & 634          \\
+adls          & ADL Surrogate Day 3                      &                 & double           & 310          \\
+sfdm2         &                                          & 5               & integer          & 159          \\
+adlsc         & Imputed ADL Calibrated to Surrogate      &                 & double           & 0           
+\end{tabular}
+\caption{A description of each variable in the SUPPORT dataset, taken from (REF).}
+\label{tab:support1}
+\end{table}
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{cc}
+\hline
+\textbf{Variable} & \textbf{Levels}                      \\
+\hline
+sex      & female                                        \\
+         & male                                          \\
+dzgroup  & ARF/MOSF w/Sepsis                             \\
+         & COPD                                          \\
+         & CHF                                           \\
+         & Cirrhosis                                     \\
+         & Coma                                          \\
+         & Colon Cancer                                  \\
+         & Lung Cancer                                   \\
+         & MOSF w/Malig                                  \\
+dzclass  & ARF/MOSF                                      \\
+         & COPD/CHF/Cirrhosis                            \\
+         & Coma                                          \\
+         & Cancer                                        \\
+income   & under \$11k                                   \\
+         & $11-$25k                                      \\
+         & $25-$50k                                      \\
+         & \textgreater{}\$50k                           \\
+race     & white                                         \\
+         & black                                         \\
+         & asian                                         \\
+         & other                                         \\
+         & hispanic                                      \\
+sfdm2    & no(M2 and SIP pres)                           \\
+         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
+         & SIP\textgreater{}=30                          \\
+         & Coma or Intub                                 \\
+         & \textless{}2 mo. follow-up                   
+\end{tabular}
+\caption{A description of each level within each categorical variable in the dataset, taken from REF.}
+\label{tab:support2}
+\end{table}
+
+Hospital death as a covariate was removed, as this is directly
+informative of death. The glmnet package(REF) on CRAN was used to
+introduce elastic net, lasso and ridge regressions to casebase. To
+visualize the effect of imputation on incidence density, we can refer to
+the pre-imputation population-time plot in Figure \ref{fig:ptPreI} and
+post-imputation population time plot in Figure \ref{fig:ptPostI}. The
+incidence density is consistent between imputed and complete datasets.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #demonstrating incidence density before imputation
+R> plot(casebase::popTime(as.data.frame(support),time = "d.time",event = "death"))
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #demonstrating incidence density after imputation
+R> plot(casebase::popTime(as.data.frame(workingCompleteData),time = "d.time",event = "death"))
+\end{CodeInput}
+\end{CodeChunk}
+
+Our main objective is to compute the absolute risk of death for a given
+set of covariates, using regularization when fitting our model. First,
+we fit a smooth hazard to the data; for the sake of this example, we
+opted for a linear term for time. As casebase makes use of the glmnet
+package, we interact with the fitsmoothhazard.fit function using a
+matrix interface, where y contains the time and event variables and x
+contains all other variables we would like to include in the model. For
+the SUPPORT dataset, all factor variables were converted into dummy
+variables. Here, we used lasso penalization by setting alpha to 1.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #Using casebase's glmnet implementation, using splines and a case to base ratio of 100,
+R> #to determine the hazard function
+R> cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
+R>                              formula_time = ~d.time,alpha=1,ratio=100)
+\end{CodeInput}
+\end{CodeChunk}
+
+Then, we take our model and use the absoluteRisk function to integrate
+and retrieve our absolute risk function.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #Estimating the absolute risk curve using the newData parameter.
+R> casebaseAbsolute=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) ,
+R>                               s="lambda.1se",method=c("numerical"))
+\end{CodeInput}
+\end{CodeChunk}
+
+We will compare this curve to a regularized version of cox regression,
+and a kaplan-meier survival curve. The regularized cox regression
+absolute risk curve required two extra steps to retrive. Coxnet from the
+glmnet package has tools to fit a regularized cox model, but requires
+that a survival object skeleton of the selected variables, so that the
+survival package tools can be used to retrive the absolute risk. We
+handle this by fitting a regularized cox model with the glmnet package,
+and a cox model from the survival package with the selected variables
+from the regularized fit. This second model serves as a skeleton that
+permits the use of survival package functions. The coefficients from the
+regularized model will replace the coeficients in our skeleton. the
+resulting model will have the correct coefficients when integrating, but
+will have an incorrect standard error. For the purposes of this case
+study, we are only interested in the absolute risk curve itself and not
+the standard error.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #Create u and xCox, which will be used when fitting Cox with glmnet.
+R> u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
+R> xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))[-c(sam),]
+R> #hazard for cox using glmnet
+R> coxglmFit=glmnet::cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
+R> #convergence demonstrated in plot
+R> plot(coxglmFit)
+R> #taking the coefficient estimates for later use
+R> nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
+R> nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
+R> #creating a new dataset that only contains the covariates chosen through glmnet.
+R> #cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
+R> cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
+R> #newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
+R> #fitting a cox model using regular estimation, however we will not keep it.
+R> #this is used more as an object place holder.
+R> coxFit <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
+R> #The coefficients of this object will be replaced with the estimates from coxglmFit.
+R> #Doing so makes it so that everything is invalid aside from the coefficients.
+R> #In this case, all we need to estimate the absolute risk is the coefficients.
+R> #Std. error would be incorrect here, if we were to draw error bars.
+R> coxFit$coefficients<-nonzero_coef_cox@x
+R> #Fitting absolute risk curve for cox+glmnet
+R> newDataCox=newData[nonzero_covariate_cox$X1-1]
+R> 
+R> abCoxFit<-survival::survfit(coxFit,newdata=as.data.frame(t(newDataCox)),time = seq(0,max(completeData$d.time), 1),type="breslow")
+\end{CodeInput}
+\end{CodeChunk}
+
+We used the survival package to calculate the kaplan-meier absolute risk
+function.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #creating a surv object to be used to fit an unadjusted absolute risk curve.
+R> # (Kaplan-Meier risk curve)
+R> km <- survival::Surv(time = completeData$d.time, event = completeData$death)
+R> abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
+\end{CodeInput}
+\end{CodeChunk}
+
+Now that our three absolute risk functions have been calculated, we
+compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase
+decrease the absolute risk by the end of the study, in comparison to
+kaplan-meier.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # A plot to compare all three Absolute risk (Commulative Incidence)
+R> plot(casebaseAbsolute,type='l',col="black",lwd=2,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
+R> lines(abCoxFit ,col="red",fun="event",lwd=3,conf.int = FALSE)
+R> lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
+R> legend("bottom", 
+R>        legend = c( "Casebase (Lasso+linear)","semi-parametric (Cox)+glmnet","KM curve"), 
+R>        col = c("black","red","Blue"),
+R>        lty = c(1, 1, 1), 
+R>        bg = "gray90")
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 #> Model1 Model2 Model3 Model4 
-#>    796    760    730    728
+#>    798    759    737    708
 \end{CodeOutput}
@@
 #>         
-#> 365 0.27
+#> 365 0.32
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> [1] 0.18
+#> [1] 0.24
 \end{CodeOutput}
@@
 
-This report was generated on 2019-12-08 11:49:58 using the following
+This report was generated on 2019-12-08 12:31:41 using the following
 computational environment and dependencies:
@@
 #>  cli           1.1.0      2019-03-19 [1] CRAN (R 3.6.1)                  
+#>  codetools     0.2-16     2018-12-24 [4] CRAN (R 3.5.2)                  
 #>  colorspace    1.4-1      2019-03-18 [1] CRAN (R 3.6.1)                  
@@
 #>  forcats     * 0.4.0      2019-02-17 [1] CRAN (R 3.6.1)                  
+#>  foreach     * 1.4.4      2017-12-12 [1] CRAN (R 3.5.0)                  
 #>  fs            1.3.1      2019-05-06 [1] CRAN (R 3.6.1)                  
@@
 #>  ggplot2     * 3.2.1      2019-08-10 [1] CRAN (R 3.6.1)                  
+#>  glmnet      * 2.0-18     2019-05-20 [1] CRAN (R 3.6.1)                  
 #>  glue          1.3.1      2019-03-12 [1] CRAN (R 3.6.1)                  
@@
 #>  httr          1.4.1      2019-08-05 [1] CRAN (R 3.6.1)                  
+#>  iterators     1.0.9      2017-12-12 [1] CRAN (R 3.5.0)                  
 #>  jsonlite      1.6        2018-12-07 [1] CRAN (R 3.6.1)                  
@@
 #>  magrittr    * 1.5        2014-11-22 [1] CRAN (R 3.5.0)                  
-#>  Matrix        1.2-18     2019-11-27 [4] CRAN (R 3.6.1)                  
+#>  Matrix      * 1.2-18     2019-11-27 [4] CRAN (R 3.6.1)                  
 #>  memoise       1.1.0      2017-04-21 [1] CRAN (R 3.5.0)                  
@@
 #> Local:    review-max /home/mturgeon/Documents/git_repositories/cbpaper
-#> Head:     [f601653] 2019-12-08: Merge branch 'review-max' of https://github.com/sahirbhatnagar/cbpaper into review-max
+#> Head:     [69663bb] 2019-12-08: Merge branch 'review-max' of https://github.com/sahirbhatnagar/cbpaper into review-max
 \end{CodeOutput}

### 2019-12-08 analysis/paper/paper.Rmd

--- 2019-12-08 analysis/paper/paper.Rmd
@@
 
-<!-- # Code formatting {.unnumbered} -->
+<!-- # Code formatting -->
 
@@
 library(splines)
+#library(readr)
+#library(tibble)
+library(glmnet)
 options(digits = 2, scipen = 999)
 # set the seed for reproducible output
-set.seed(1234)
+set.seed(2)
 
@@
 
-1. Time matching/risk set sampling (including Cox partial likelihood) eliminates the baseline hazard from the likelihood expression for the
-hazard ratios
+  1. hazard ratios
   2. If, however, the absolute risks are of interest, they have to be recovered using the semi-parametric Breslow estimator
-3. Alternative approaches for fitting flexible hazard models for estimating absolute risks, not requiring this two-step approach? Yes!
-[@hanley2009fitting]
+  3. Alternative approaches for fitting flexible hazard models for estimating absolute risks, not requiring this two-step approach? Yes! [@hanley2009fitting]
 
@@
 
+# Literature review
+To conduct our review on current methods available in survival analysis, a scraper was coded to examine all packages in the survival task view on CRAN [-@survTaskView].
+There was a total of 259 packages, many of which are not relevant to the problems casebase addresses.
+To filter these packages, a scraper was used to search for competing risks, cumulative incidence functions, non- proportional hazards and penalization.
+This reduced our set of 259 packages down to 60.
+From here, we manually examined each package to determine which ones are comparable to casebase, with either a different approach to the same problems or future features we would like to add to casebase.
+
+These packages can belong to three main categories: non-proportional hazards, regularization and competing risks.
+Besides these three main categories, there are extra features of interest as well.
+Table \ref{tab:surv-pkgs} summarizes the findings.
+
+Many packages exist that implement semi-parametric Cox models.
+Of those, \pkg{glmnet} [-@regpathcox]???, \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche] introduce penalization in their model fitting.
+\pkg{CoxRidge} [-@perperoglou], \pkg{rstpm2} [-@clements_liu] and \pkg{survival} [-@survival-package] include penalization, but they implement the extended cox model permitting intuitive non proportional hazard modeling while the previous four do not.
+In terms of penalization, the benefits of \pkg{casebase} comes from its flexibility.
+Linear models have a wide tool set, considerably larger than those made for Cox regression.
+<!--(*reference for number of packages between their task views, regression is spread out across Multivariate, SocialSciences, Robust among others.
+The focus of this task view is on Optimization Infrastructure Packages , General Purpose Continuous Solvers , Mathematical Programming Solvers , and Specific Applications in Optimization , or Multi Objective Optimization*)--> \pkg{casebase} permits many regularization tools developed for linear regression to be repurposed within a survival context.
+
+Parametric models permit non-proportional hazards and are implemented in a handful of packages.
+Of interest are \pkg{CFC} [-@sharabiani_mahani_2019], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu] and \pkg{survival} [-@survival-package].
+\pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas \pkg{flexsurv} and \pkg{survival} contain implementations for users to supply any distribution they would like [-@survival-package], [-@flexsurv].
+\pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} can use splines, permitting a flexible fit over time [-@flexsurv], [-@smoothHazard], [-@scheike2014estimating], [-@clements_liu].
+Though \pkg{casebase} is limited to three distributions at the moment, it also uses splines allowing for distributions that may not be consistent over survival time.
+
+Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets} and \pkg{survival} contain implementations for competing risks [-@sharabiani_mahani_2019], [-@flexsurv], [-@scheike2014estimating], [-@survival-package].
+The differentiating factor between these packages and \pkg{casebase} is that \pkg{casebase} handles competing risks through multiple logistic regression.
+\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval censoring, whereas \pkg{casebase} currently cannot.
+Interval censoring is a simple feature to implement and is a potential improvement to add to \pkg{casebase}.
+
+Of all the packages listed in Table \ref{tab:surv-pkgs}, \pkg{CFC}, \pkg{Flexsurv}, \pkg{mets}, \pkg{RiskRegression}, \pkg{rstpm2} and \pkg{survival} have implementations for cumulative incidence [-@sharabiani_mahani_2019], [-@flexsurv], [-@scheike2014estimating], [-@gerds_blanche], [-@clements_liu], [-@survival-package].
+Since \pkg{casebase} is parametric, the baseline hazard is known and requires integration to retrieve the cumulative incidence.
+This is similar to the approaches of \pkg{CFC} and \pkg{flexsurv} [-@sharabiani_mahani_2019], [-@flexsurv], while the other packages use semi-parametric approaches to retrieve the baseline after fitting [-@scheike2014estimating], [-@gerds_blanche].
+\pkg{rstpm2} and \pkg{survival} use both for their respective parametric and semi-parametric model implementations [-@clements_liu], [-@survival-package].
+
+\begin{table}[ht]
+\centering
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{ccccccccc}
+\hline
+\textbf{Package}        & \textbf{Competing Risks} & \textbf{Non-proportional} & \textbf{Penalization} & \textbf{Splines} & \textbf{Parametric} & \textbf{Semi-parametric} & \textbf{Interval/left Censoring} & \textbf{Absolute Risk}           \\
+\hline
+\textbf{casebase} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
+\textbf{CFC} \cite{sharabiani_mahani_2019}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
+\textbf{coxRidge} \cite{perperoglou}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{crrp} \cite{fu}           & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
+\textbf{fastcox} \cite{yi_zou}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{Flexrsurv} \cite{flexrsurv}      &                          & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
+\\
+\textbf{Flexsurv} \cite{flexsurv}       & x                        & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
+\\
+\textbf{glmnet} \cite{regpathcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{glmpath} \cite{park_hastie}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{mets} \cite{scheike2014estimating}           & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
+\textbf{penalized} \cite{goeman_meijer2019}      &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{RiskRegression} \cite{gerds_blanche} &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
+\textbf{Rstpm2} \cite{clements_liu}         &                          & x                         & x                     & x                & x                   & x                        & x                                & Cumm haz                         \\
+\textbf{SmoothHazard} \cite{smoothHazard}   &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
+\textbf{Survival} \cite{survival-package}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x
+\end{tabular}%
+}
+\caption{Different features of interest in various survival packages.}
+\label{tab:surv-pkgs}
+\end{table}
+
 # Theoretical details
@@
 
-# Case study 3--The Cancer Genome Atlas
-
-  - Glmnet and TCGA data
+# Case study 3--Study to Understand Prognoses Preferences Outcomes and Risks of Treatment
+
+We examined the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset (REF from main site).
+the SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital.
+Imputation was conducted using the default methods in the mice package in R, available on CRAN.
+Before imputation, there was a total of 1000 individuals and 35 variables.
+After imputation, we have 474 observations where 67.5% of individuals in the study experience the event of interest.
+The SUPPORT dataset will be used to demonstrate regularization within casebase (REF), while comparing their absolute risk preditions for a new individual.
+A description of each variable can be found in Table~\ref{tab:support1} and a breakdown of each categorical variable in Table~\ref{tab:support2}.
+The data was downloaded from the main author's website, and it is also available as part of the casebase package.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{ccccc}
+\hline
+\textbf{Name} & \textbf{Labels}                          & \textbf{Levels} & \textbf{Storage} & \textbf{NAs} \\
+\hline
+age           & Age                                      &                 & double           & 0            \\
+death         & Death at any time up to NDI date:31DEC94 &                 & double           & 0            \\
+sex           &                                          & 2               & integer          & 0            \\
+hospdead      & Death in Hospital                        &                 & double           & 0            \\
+slos          & Days from Study Entry to Discharge       &                 & double           & 0            \\
+d.time        & Days of Follow-Up                        &                 & double           & 0            \\
+dzgroup       &                                          & 8               & integer          & 0            \\
+dzclass       &                                          & 4               & integer          & 0            \\
+num.co        & number of comorbidities                  &                 & double           & 0            \\
+edu           & Years of Education                       &                 & double           & 202          \\
+income        &                                          & 4               & integer          & 349          \\
+scoma         & SUPPORT Coma Score based on Glasgow D3   &                 & double           & 0            \\
+charges       & Hospital Charges                         &                 & double           & 25           \\
+totcst        & Total RCC cost                           &                 & double           & 105          \\
+totmcst       & Total micro-cost                         &                 & double           & 372          \\
+avtisst       & Average TISS, Days 3-25                  &                 & double           & 6            \\
+race          &                                          & 5               & integer          & 5            \\
+meanbp        & Mean Arterial Blood Pressure Day 3       &                 & double           & 0            \\
+wblc          & White Blood Cell Count Day 3             &                 & double           & 24           \\
+hrt           & Heart Rate Day 3                         &                 & double           & 0            \\
+resp          & Respiration Rate Day 3                   &                 & double           & 0            \\
+temp          & Temperature (celcius) Day 3              &                 & double           & 0            \\
+pafi          & PaO2/(.01*FiO2) Day 3                    &                 & double           & 253          \\
+alb           & Serum Albumin Day 3                      &                 & double           & 378          \\
+bili          & Bilirubin Day 3                          &                 & double           & 297          \\
+crea          & Serum creatinine Day 3                   &                 & double           & 3            \\
+sod           & Serum sodium Day 3                       &                 & double           & 0            \\
+ph            & Serum pH (arterial) Day 3                &                 & double           & 250          \\
+glucose       & Glucose Day 3                            &                 & double           & 470          \\
+bun           & BUN Day 3                                &                 & double           & 455          \\
+urine         & Urine Output Day 3                       &                 & double           & 517          \\
+adlp          & ADL Patient Day 3                        &                 & double           & 634          \\
+adls          & ADL Surrogate Day 3                      &                 & double           & 310          \\
+sfdm2         &                                          & 5               & integer          & 159          \\
+adlsc         & Imputed ADL Calibrated to Surrogate      &                 & double           & 0           
+\end{tabular}
+\caption{A description of each variable in the SUPPORT dataset, taken from (REF).}
+\label{tab:support1}
+\end{table}
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{cc}
+\hline
+\textbf{Variable} & \textbf{Levels}                      \\
+\hline
+sex      & female                                        \\
+         & male                                          \\
+dzgroup  & ARF/MOSF w/Sepsis                             \\
+         & COPD                                          \\
+         & CHF                                           \\
+         & Cirrhosis                                     \\
+         & Coma                                          \\
+         & Colon Cancer                                  \\
+         & Lung Cancer                                   \\
+         & MOSF w/Malig                                  \\
+dzclass  & ARF/MOSF                                      \\
+         & COPD/CHF/Cirrhosis                            \\
+         & Coma                                          \\
+         & Cancer                                        \\
+income   & under \$11k                                   \\
+         & $11-$25k                                      \\
+         & $25-$50k                                      \\
+         & \textgreater{}\$50k                           \\
+race     & white                                         \\
+         & black                                         \\
+         & asian                                         \\
+         & other                                         \\
+         & hispanic                                      \\
+sfdm2    & no(M2 and SIP pres)                           \\
+         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
+         & SIP\textgreater{}=30                          \\
+         & Coma or Intub                                 \\
+         & \textless{}2 mo. follow-up                   
+\end{tabular}
+\caption{A description of each level within each categorical variable in the dataset, taken from REF.}
+\label{tab:support2}
+\end{table}
+
+Hospital death as a covariate was removed, as this is directly informative of death.
+The glmnet package(REF) on CRAN was used to introduce elastic net, lasso and ridge regressions to casebase.
+To visualize the effect of imputation on incidence density, we can refer to the pre-imputation population-time plot in Figure \ref{fig:ptPreI} and post-imputation population time plot in Figure \ref{fig:ptPostI}.
+The incidence density is consistent between imputed and complete datasets.
+
+```{r supportSetUp, echo = FALSE, eval = eval_cs3}
+completeData=casebase::support
+#keep one individual out, at random, for which we will develop an 
+#absolute risk curve.
+sam=sample(1:nrow(completeData), 1)
+#Set workingData to the remaining individuals in the data
+#workingData=completeData[-c(sam),]
+workingCompleteData=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))[,-c(1)]
+#Create u and xCox, which will be used when fitting Cox with glmnet.
+newData=workingCompleteData[sam,]
+x=workingCompleteData[-c(sam),]
+#x and y will be used to fit the casebase model under glmnet.
+#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
+y=data.matrix(completeData[-c(sam),c(34,4)])
+```
+
+```{r poptimePreImputation, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:ptPreI} Population-time plot before imputation.
+Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study."}
+#demonstrating incidence density before imputation
+plot(casebase::popTime(as.data.frame(support),time = "d.time",event = "death"))
+```
+
+```{r poptimePostImputation, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:ptPostI} Population-time plot before imputation.
+Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study."}
+#demonstrating incidence density after imputation
+plot(casebase::popTime(as.data.frame(workingCompleteData),time = "d.time",event = "death"))
+```
+Our main objective is to compute the absolute risk of death for a given set of covariates, using regularization when fitting our model.
+First, we fit a smooth hazard to the data; for the sake of this example, we opted for a linear term for time.
+As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model.
+For the SUPPORT dataset, all factor variables were converted into dummy variables.
+Here, we used lasso penalization by setting alpha to 1.
+```{r CBhazard, echo = TRUE, eval = eval_cs3}
+#Using casebase's glmnet implementation, using splines and a case to base ratio of 100,
+#to determine the hazard function
+cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
+                             formula_time = ~d.time,alpha=1,ratio=100)
+```
+Then, we take our model and use the absoluteRisk function to integrate and retrieve our absolute risk function.
+```{r CBabsolute, echo = TRUE, eval = eval_cs3}
+#Estimating the absolute risk curve using the newData parameter.
+casebaseAbsolute=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) ,
+                              s="lambda.1se",method=c("numerical"))
+```
+We will compare this curve to a regularized version of cox regression, and a kaplan-meier survival curve.
+The regularized cox regression absolute risk curve required two extra steps to retrive.
+Coxnet from the glmnet package has tools to fit a regularized cox model, but requires that a survival object skeleton of the selected variables, so that the survival package tools can be used to retrive the absolute risk.
+We handle this by fitting a regularized cox model with the glmnet package, and a cox model from the survival package with the selected variables from the regularized fit.
+This second model serves as a skeleton that permits the use of survival package functions.
+The coefficients from the regularized model will replace the coeficients in our skeleton.
+the resulting model will have the correct coefficients when integrating, but will have an incorrect standard error.
+For the purposes of this case study, we are only interested in the absolute risk curve itself and not the standard error.
+
+```{r coxHazAbsolute, echo = TRUE, eval = eval_cs3}
+#Create u and xCox, which will be used when fitting Cox with glmnet.
+u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
+xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))[-c(sam),]
+#hazard for cox using glmnet
+coxglmFit=glmnet::cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
+#convergence demonstrated in plot
+plot(coxglmFit)
+#taking the coefficient estimates for later use
+nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
+nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
+#creating a new dataset that only contains the covariates chosen through glmnet.
+#cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
+cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
+#newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
+#fitting a cox model using regular estimation, however we will not keep it.
+#this is used more as an object place holder.
+coxFit <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
+#The coefficients of this object will be replaced with the estimates from coxglmFit.
+#Doing so makes it so that everything is invalid aside from the coefficients.
+#In this case, all we need to estimate the absolute risk is the coefficients.
+#Std. error would be incorrect here, if we were to draw error bars.
+coxFit$coefficients<-nonzero_coef_cox@x
+#Fitting absolute risk curve for cox+glmnet
+newDataCox=newData[nonzero_covariate_cox$X1-1]
+
+abCoxFit<-survival::survfit(coxFit,newdata=as.data.frame(t(newDataCox)),time = seq(0,max(completeData$d.time), 1),type="breslow")
+```
+We used the survival package to calculate the kaplan-meier absolute risk function. 
+```{r fitKM+abRisk, echo = TRUE, eval = eval_cs3}
+#creating a surv object to be used to fit an unadjusted absolute risk curve.
+# (Kaplan-Meier risk curve)
+km <- survival::Surv(time = completeData$d.time, event = completeData$death)
+abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
+```
+
+Now that our three absolute risk functions have been calculated, we compare them all in Figure \ref{fig:abSC}.
+Both coxnet and casebase decrease the absolute risk by the end of the study, in comparison to kaplan-meier.
+```{r abSupportComparison, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
+# A plot to compare all three Absolute risk (Commulative Incidence)
+plot(casebaseAbsolute,type='l',col="black",lwd=2,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
+lines(abCoxFit ,col="red",fun="event",lwd=3,conf.int = FALSE)
+lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
+legend("bottom", 
+       legend = c( "Casebase (Lasso+linear)","semi-parametric (Cox)+glmnet","KM curve"), 
+       col = c("black","red","Blue"),
+       lty = c(1, 1, 1), 
+       bg = "gray90")
+```
 

### 2019-12-10 analysis/paper/paper.Rmd
edit litt review

@@
 
+# Theoretical details
+
+As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to discretize the study base into an infinite amount of *person moments*.
+These person moments are indexed by both an individual in the study and a time point, and therefore each person moment has a covariate profile, an exposure status and an outcome status attached to it.
+We note that there is only a finite number of person moments associated with the event of interest (what Hanley & Miettinen call the *case series*).
+The case-base sampling refers to the sampling from the base of a representative finite sample called the *base series*.
+
+<!-- Popular survival analysis methods, like Kaplan-Meier and Cox regression, rely on the notion of risk-set sampling. Case-base sampling can be seen as an alternative. -->
+
+As shown by Saarela & Arjas [-@saarela2015non] (and further expanded in Saarela [-@saarela2016case]), writing the likelihood arising from this data-generating mechanism using the framework of non-homogeneous Poisson processes, we eventually reach an expression where each person-moment's contribution is of the form
+$$\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},$$
+where $N(t)$ is the counting process associated with the event of interest, $h(t)$ is the corresponding hazard function, and $\rho(t)$ is the hazard function for the Poisson process associated with case-base sampling.
+This parametric form suggests that we can readily estimate log-hazards of the form $\log(h(t)) = g(t; X)$ using logistic regression, where each observation corresponds to a person moment, the function $g(t; X)$ is linear in a finite number of parameters, and where we treat $-\log(\rho(t))$ as an offset.
+
+In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base sampling *uniformly*, i.e.
+to sample the base series uniformly from the study base.
+In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base.
+More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism  are proportional to the cardiovascular disease event rate given by the Framingham score.
+
+The `casebase` package fits the family of hazard functions of the form
+
+$$ h(t;X) = \exp[g(t;X)] $$
+where $t$ denotes time and $X$, the individual's covariate profile.
+Different functions of $t$ lead to different parametric hazard models.
+The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$.
+
+$$ h(t;X) = \exp(\beta_0 + \beta_1 X) $$
+
+The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
+This is also known as the *memoryless property* (Kalbfleisch and Prentice, 2002).
+
+The Gompertz hazard model is given by including a linear term for time:
+
+$$ h(t;X)  = \exp(\beta_0 + \beta_1 t + \beta_2 X) $$
+
+Use of $\log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
+
+$$ h(t;X)  = \exp(\beta_0 + \beta_1 \log(t) + \beta_2 X) $$
+
+For competing-risk analyses with $J$ possible events, we can show that each person-moment's contribution of the likelihood is of the form
+
+$$\frac{h_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^Jh_j(t)},$$
+
+where $N_j(t)$ is the counting process associated with the event of type $j$ and $h_j(t)$ is the corresponding hazard function.
+As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1\end{align*}}
+
+# Existing packages
+
+Survival analysis is an important branch of applied statistics and epidemiology.
+Accordingly, there is a vast ecosystem of \code{R} packages implementing different methods.
+In this section, we describe how the functionalities of \pkg{casebase} compare to these packages.
+
+At the time of writing, a cursory examination of CRAN's task view on survival analysis reveals that there are over 250 packages related to survival analysis [-@survTaskView].
+For the purposes of this article, we restricted our description to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and cumulative incidence curve estimation.
+By searching for appropriate keywords in the \code{DESCRIPTION} file of these packages, we found 60 relevant packages.
+These 60 packages were further reduced through manual curation to a subset of 14 packages.
+These packages appears in Table \ref{tab:surv-pkgs}, along with some of the functionalities they offer.
+\textcolor{red}{JESSE: It's not clear how you selected this subset, going from 60 to 14 is a big drop.}
+
+Several packages implement penalized estimation for the Cox model: \pkg{glmnet} [-@regpathcox], \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche].
+Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: \pkg{CoxRidge} [-@perperoglou], \pkg{rstpm2} [-@clements_liu] and \pkg{survival} [-@survival-package].
+On the other hand, \pkg{casebase} provides penalized estimation in the context of parametric hazards.
+To our knowledge, this is the only package to offer this functionality.
+\textcolor{red}{JESSE: Is this true?
+Because it seems to be a logical conclusion from your paragraph.}
+
+Parametric survival models are implemented in a handful of packages: \pkg{CFC} [-@sharabiani_mahani_2019], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu], \pkg{mets} [-@scheike2014estimating], and \pkg{survival}.
+The types of models they allow vary for each package.
+For example, \pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \pkg{flexsurv} and \pkg{survival} allow users to supply any distribution of their choice.
+Also, \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of time using splines, which allows flexible modeling of the hazard function.
+Moreover, \pkg{flexsurv} has the ability to estimate both scale and shape parameters for a variety of parametric families (see Table \ref{tab:dists}).
+As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear model of covariates (including time).
+Therefore, our package allows the user to model the effect of time using splines, and through interaction terms involving covariates and time, it also allows user to fit time-varying coefficient models.
+However, we do not explicitly model any shape parameter, unlike \pkg{flexsurv}.
+
+Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets} and \pkg{survival} contain implementations for competing risks.
+The differentiating factor between these packages and \pkg{casebase} is that \pkg{casebase} handles competing risks through multiple logistic regression.
+\textcolor{red}{JESSE: There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}.
+Could you incorporate some of their discussion, with proper citations?
+The biggest omission right now is \pkg{cmprsk}, which we actually use below.}
+
+Finally, several packages include functions to estimate the cumulative incidence function.
+The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard.
+The first category broadly corresponds to parametric survival models, where the full hazard is explictly modeled.
+Using this estimate, the survival function and the cumulative incidence function can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below).
+Packages including this functionality include \pkg{CFC}, \pkg{Flexsurv}, \pkg{mets}, and \pkg{survival}.
+Our package \pkg{casebase} also follows this approach for both single-event and competing-event analyses.
+The second category outlined above broadly corresponds to Cox models.
+These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function.
+This is achieved using semi-parametric estimators (e.g.
+Breslow's estimator).
+Packages that implement this approach include \pkg{RiskRegression}, \pkg{rstpm2}, and \pkg{survival}.
+As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category produces estimates in the form of step-wise functions.
+This was one of the main motivations for introducting case-base sampling in survival analysis.
+
+\begin{table}[ht]
+\centering
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{ccccccccc}
+\hline
+\textbf{Package}        & \textbf{Competing Risks} & \textbf{Non-proportional} & \textbf{Penalization} & \textbf{Splines} & \textbf{Parametric} & \textbf{Semi-parametric} & \textbf{Interval/left Censoring} & \textbf{Absolute Risk}           \\
+\hline
+\textbf{casebasee} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
+\textbf{CFC} \cite{sharabiani_mahani_2019}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
+\textbf{coxRidge} \cite{perperoglou}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{crrp} \cite{fu}           & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
+\textbf{fastcox} \cite{yi_zou}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{Flexrsurv} \cite{flexrsurv}      &                          & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
+\\
+\textbf{Flexsurv} \cite{flexsurv}       & x                        & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
+\\
+\textbf{glmnet} \cite{regpathcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{glmpath} \cite{park_hastie}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{mets} \cite{scheike2014estimating}           & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
+\textbf{penalized} \cite{goeman_meijer2019}      &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{RiskRegression} \cite{gerds_blanche} &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
+\textbf{Rstpm2} \cite{clements_liu}         &                          & x                         & x                     & x                & x                   & x                        & x                                & Cumm haz                         \\
+\textbf{SmoothHazard} \cite{smoothHazard}   &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
+\textbf{Survival} \cite{survival-package}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x
+\end{tabular}%
+}
+\caption{Different features of interest in various survival packages.}
+\label{tab:surv-pkgs}
+\end{table}
+
 \begin{table}
@@
 
-# Literature review
-To conduct our review on current methods available in survival analysis, a scraper was coded to examine all packages in the survival task view on CRAN [-@survTaskView].
-There was a total of 259 packages, many of which are not relevant to the problems casebase addresses.
-To filter these packages, a scraper was used to search for competing risks, cumulative incidence functions, non- proportional hazards and penalization.
-This reduced our set of 259 packages down to 60.
-From here, we manually examined each package to determine which ones are comparable to casebase, with either a different approach to the same problems or future features we would like to add to casebase.
-
-These packages can belong to three main categories: non-proportional hazards, regularization and competing risks.
-Besides these three main categories, there are extra features of interest as well.
-Table \ref{tab:surv-pkgs} summarizes the findings.
-
-Many packages exist that implement semi-parametric Cox models.
-Of those, \pkg{glmnet} [-@regpathcox]???, \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche] introduce penalization in their model fitting.
-\pkg{CoxRidge} [-@perperoglou], \pkg{rstpm2} [-@clements_liu] and \pkg{survival} [-@survival-package] include penalization, but they implement the extended cox model permitting intuitive non proportional hazard modeling while the previous four do not.
-In terms of penalization, the benefits of \pkg{casebase} comes from its flexibility.
-Linear models have a wide tool set, considerably larger than those made for Cox regression.
-<!--(*reference for number of packages between their task views, regression is spread out across Multivariate, SocialSciences, Robust among others.
-The focus of this task view is on Optimization Infrastructure Packages , General Purpose Continuous Solvers , Mathematical Programming Solvers , and Specific Applications in Optimization , or Multi Objective Optimization*)--> \pkg{casebase} permits many regularization tools developed for linear regression to be repurposed within a survival context.
-
-Parametric models permit non-proportional hazards and are implemented in a handful of packages.
-Of interest are \pkg{CFC} [-@sharabiani_mahani_2019], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu] and \pkg{survival} [-@survival-package].
-\pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas \pkg{flexsurv} and \pkg{survival} contain implementations for users to supply any distribution they would like [-@survival-package], [-@flexsurv].
-\pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} can use splines, permitting a flexible fit over time [-@flexsurv], [-@smoothHazard], [-@scheike2014estimating], [-@clements_liu].
-Though \pkg{casebase} is limited to three distributions at the moment, it also uses splines allowing for distributions that may not be consistent over survival time.
-
-Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets} and \pkg{survival} contain implementations for competing risks [-@sharabiani_mahani_2019], [-@flexsurv], [-@scheike2014estimating], [-@survival-package].
-The differentiating factor between these packages and \pkg{casebase} is that \pkg{casebase} handles competing risks through multiple logistic regression.
-\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval censoring, whereas \pkg{casebase} currently cannot.
-Interval censoring is a simple feature to implement and is a potential improvement to add to \pkg{casebase}.
-
-Of all the packages listed in Table \ref{tab:surv-pkgs}, \pkg{CFC}, \pkg{Flexsurv}, \pkg{mets}, \pkg{RiskRegression}, \pkg{rstpm2} and \pkg{survival} have implementations for cumulative incidence [-@sharabiani_mahani_2019], [-@flexsurv], [-@scheike2014estimating], [-@gerds_blanche], [-@clements_liu], [-@survival-package].
-Since \pkg{casebase} is parametric, the baseline hazard is known and requires integration to retrieve the cumulative incidence.
-This is similar to the approaches of \pkg{CFC} and \pkg{flexsurv} [-@sharabiani_mahani_2019], [-@flexsurv], while the other packages use semi-parametric approaches to retrieve the baseline after fitting [-@scheike2014estimating], [-@gerds_blanche].
-\pkg{rstpm2} and \pkg{survival} use both for their respective parametric and semi-parametric model implementations [-@clements_liu], [-@survival-package].
-
-\begin{table}[ht]
-\centering
-\resizebox{\textwidth}{!}{%
-\begin{tabular}{ccccccccc}
-\hline
-\textbf{Package}        & \textbf{Competing Risks} & \textbf{Non-proportional} & \textbf{Penalization} & \textbf{Splines} & \textbf{Parametric} & \textbf{Semi-parametric} & \textbf{Interval/left Censoring} & \textbf{Absolute Risk}           \\
-\hline
-\textbf{casebase} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
-\textbf{CFC} \cite{sharabiani_mahani_2019}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
-\textbf{coxRidge} \cite{perperoglou}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{crrp} \cite{fu}           & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
-\textbf{fastcox} \cite{yi_zou}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{Flexrsurv} \cite{flexrsurv}      &                          & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
-\\
-\textbf{Flexsurv} \cite{flexsurv}       & x                        & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
-\\
-\textbf{glmnet} \cite{regpathcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{glmpath} \cite{park_hastie}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{mets} \cite{scheike2014estimating}           & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
-\textbf{penalized} \cite{goeman_meijer2019}      &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{RiskRegression} \cite{gerds_blanche} &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
-\textbf{Rstpm2} \cite{clements_liu}         &                          & x                         & x                     & x                & x                   & x                        & x                                & Cumm haz                         \\
-\textbf{SmoothHazard} \cite{smoothHazard}   &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
-\textbf{Survival} \cite{survival-package}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x
-\end{tabular}%
-}
-\caption{Different features of interest in various survival packages.}
-\label{tab:surv-pkgs}
-\end{table}
-
-# Theoretical details
-
-As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to discretize the study base into an infinite amount of *person moments*.
-These person moments are indexed by both an individual in the study and a time point, and therefore each person moment has a covariate profile, an exposure status and an outcome status attached to it.
-We note that there is only a finite number of person moments associated with the event of interest (what Hanley & Miettinen call the *case series*).
-The case-base sampling refers to the sampling from the base of a representative finite sample called the *base series*.
-
-<!-- Popular survival analysis methods, like Kaplan-Meier and Cox regression, rely on the notion of risk-set sampling. Case-base sampling can be seen as an alternative. -->
-
-As shown by Saarela & Arjas [-@saarela2015non] (and further expanded in Saarela [-@saarela2016case]), writing the likelihood arising from this data-generating mechanism using the framework of non-homogeneous Poisson processes, we eventually reach an expression where each person-moment's contribution is of the form
-$$\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},$$
-where $N(t)$ is the counting process associated with the event of interest, $h(t)$ is the corresponding hazard function, and $\rho(t)$ is the hazard function for the Poisson process associated with case-base sampling.
-This parametric form suggests that we can readily estimate log-hazards of the form $\log(h(t)) = g(t; X)$ using logistic regression, where each observation corresponds to a person moment, the function $g(t; X)$ is linear in a finite number of parameters, and where we treat $-\log(\rho(t))$ as an offset.
-
-In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base sampling *uniformly*, i.e.
-to sample the base series uniformly from the study base.
-In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base.
-More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism  are proportional to the cardiovascular disease event rate given by the Framingham score.
-
-The `casebase` package fits the family of hazard functions of the form
-
-$$ h(t;X) = \exp[g(t;X)] $$
-where $t$ denotes time and $X$, the individual's covariate profile.
-Different functions of $t$ lead to different parametric hazard models.
-The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$.
-
-$$ h(t;X) = \exp(\beta_0 + \beta_1 X) $$
-
-The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
-This is also known as the *memoryless property* (Kalbfleisch and Prentice, 2002).
-
-The Gompertz hazard model is given by including a linear term for time:
-
-$$ h(t;X)  = \exp(\beta_0 + \beta_1 t + \beta_2 X) $$
-
-Use of $\log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
-
-$$ h(t;X)  = \exp(\beta_0 + \beta_1 \log(t) + \beta_2 X) $$
-
-For competing-risk analyses with $J$ possible events, we can show that each person-moment's contribution of the likelihood is of the form
-
-$$\frac{h_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^Jh_j(t)},$$
-
-where $N_j(t)$ is the counting process associated with the event of type $j$ and $h_j(t)$ is the corresponding hazard function.
-As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1\end{align*}}
-
 # Implementation details
@@
 For the case of a single event, it uses the familiar identity
-$$S(t) = \exp\left(-\int_0^t h(u;X) du\right).$$
+\begin{equation}\label{eqn:surv}
+S(t) = \exp\left(-\int_0^t h(u;X) du\right).
+\end{equation}
 The integral is computed using either the \code{stats::integrate} function or Monte-Carlo integration. The risk function (or cumulative incidence function) is then defined as
-
-$$ CI(t) = 1 - S(t).$$
+\begin{equation}\label{eqn:CI}
+CI(t) = 1 - S(t).
+\end{equation}
 
@@
 After imputation, we have 474 observations where 67.5% of individuals in the study experience the event of interest.
-The SUPPORT dataset will be used to demonstrate regularization within casebase (REF), while comparing their absolute risk preditions for a new individual.
+The SUPPORT dataset will be used to demonstrate regularization within casebase (REF), while comparing their absolute risk predictions for a new individual.
 A description of each variable can be found in Table~\ref{tab:support1} and a breakdown of each categorical variable in Table~\ref{tab:support2}.
@@
 resp          & Respiration Rate Day 3                   &                 & double           & 0            \\
-temp          & Temperature (celcius) Day 3              &                 & double           & 0            \\
+temp          & Temperature (Celsius) Day 3              &                 & double           & 0            \\
 pafi          & PaO2/(.01*FiO2) Day 3                    &                 & double           & 253          \\
@@
 We will compare this curve to a regularized version of cox regression, and a kaplan-meier survival curve.
-The regularized cox regression absolute risk curve required two extra steps to retrive.
-Coxnet from the glmnet package has tools to fit a regularized cox model, but requires that a survival object skeleton of the selected variables, so that the survival package tools can be used to retrive the absolute risk.
+The regularized cox regression absolute risk curve required two extra steps to retrieve.
+Coxnet from the glmnet package has tools to fit a regularized cox model, but requires that a survival object skeleton of the selected variables, so that the survival package tools can be used to retrieve the absolute risk.
 We handle this by fitting a regularized cox model with the glmnet package, and a cox model from the survival package with the selected variables from the regularized fit.
 This second model serves as a skeleton that permits the use of survival package functions.
-The coefficients from the regularized model will replace the coeficients in our skeleton.
+The coefficients from the regularized model will replace the coefficients in our skeleton.
 the resulting model will have the correct coefficients when integrating, but will have an incorrect standard error.
@@
 After enrollment, the program searches for a suitable donor for the patient, which can take anywhere between a few days to almost a year.
-We are interested in the effect of a heart transplant on survival; therefore, the patient is considered exposed only after the transplant has occured.
+We are interested in the effect of a heart transplant on survival; therefore, the patient is considered exposed only after the transplant has occurred.
 
@@
 
+\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval censoring, whereas \pkg{casebase} currently cannot.
+Interval censoring is a simple feature to implement and is a potential improvement to add to \pkg{casebase}.
+
 # Environment Details

### 2019-12-10 analysis/paper/paper.tex
edit litt review

@@
 profile can be easily derived.
+
+\hypertarget{theoretical-details}{%
+\section{Theoretical details}\label{theoretical-details}}
+
+As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
+key idea behind case-base sampling is to discretize the study base into
+an infinite amount of \emph{person moments}. These person moments are
+indexed by both an individual in the study and a time point, and
+therefore each person moment has a covariate profile, an exposure status
+and an outcome status attached to it. We note that there is only a
+finite number of person moments associated with the event of interest
+(what Hanley \& Miettinen call the \emph{case series}). The case-base
+sampling refers to the sampling from the base of a representative finite
+sample called the \emph{base series}.
+
+As shown by Saarela \& Arjas \citeyearpar{saarela2015non} (and further
+expanded in Saarela \citeyearpar{saarela2016case}), writing the
+likelihood arising from this data-generating mechanism using the
+framework of non-homogeneous Poisson processes, we eventually reach an
+expression where each person-moment's contribution is of the form
+\[\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},\] where \(N(t)\) is the counting
+process associated with the event of interest, \(h(t)\) is the
+corresponding hazard function, and \(\rho(t)\) is the hazard function
+for the Poisson process associated with case-base sampling. This
+parametric form suggests that we can readily estimate log-hazards of the
+form \(\log(h(t)) = g(t; X)\) using logistic regression, where each
+observation corresponds to a person moment, the function \(g(t; X)\) is
+linear in a finite number of parameters, and where we treat
+\(-\log(\rho(t))\) as an offset.
+
+In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
+suggest performing case-base sampling \emph{uniformly}, i.e.~to sample
+the base series uniformly from the study base. In terms of Poisson
+processes, this sampling strategy corresponds essentially to a
+time-homogeneous Poisson process with intensity equal to \(b/B\), where
+\(b\) is the number of sampled observations in the base series, and
+\(B\) is the total population-time for the study base. More complex
+examples are also available; see for example Saarela \& Arjas
+\citeyearpar{saarela2015non}, where the probabilities of the sampling
+mechanism are proportional to the cardiovascular disease event rate
+given by the Framingham score.
+
+The \texttt{casebase} package fits the family of hazard functions of the
+form
+
+\[ h(t;X) = \exp[g(t;X)] \] where \(t\) denotes time and \(X\), the
+individual's covariate profile. Different functions of \(t\) lead to
+different parametric hazard models. The simplest of these models is the
+one-parameter exponential distribution which is obtained by taking the
+hazard function to be constant over the range of \(t\).
+
+\[ h(t;X) = \exp(\beta_0 + \beta_1 X) \]
+
+The instantaneous failure rate is independent of \(t\), so that the
+conditional chance of failure in a time interval of specified length is
+the same regardless of how long the individual has been in the study.
+This is also known as the \emph{memoryless property} (Kalbfleisch and
+Prentice, 2002).
+
+The Gompertz hazard model is given by including a linear term for time:
+
+\[ h(t;X)  = \exp(\beta_0 + \beta_1 t + \beta_2 X) \]
+
+Use of \(\log(t)\) yields the Weibull hazard which allows for a power
+dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
+
+\[ h(t;X)  = \exp(\beta_0 + \beta_1 \log(t) + \beta_2 X) \]
+
+For competing-risk analyses with \(J\) possible events, we can show that
+each person-moment's contribution of the likelihood is of the form
+
+\[\frac{h_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^Jh_j(t)},\]
+
+where \(N_j(t)\) is the counting process associated with the event of
+type \(j\) and \(h_j(t)\) is the corresponding hazard function. As may
+be expected, this functional form is similar to the terms appearing in
+the likelihood function for multinomial
+regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1\end{align*}}
+
+\hypertarget{existing-packages}{%
+\section{Existing packages}\label{existing-packages}}
+
+Survival analysis is an important branch of applied statistics and
+epidemiology. Accordingly, there is a vast ecosystem of \code{R}
+packages implementing different methods. In this section, we describe
+how the functionalities of \pkg{casebase} compare to these packages.
+
+At the time of writing, a cursory examination of CRAN's task view on
+survival analysis reveals that there are over 250 packages related to
+survival analysis \citeyearpar{survTaskView}. For the purposes of this
+article, we restricted our description to packages that implement at
+least one of the following features: parametric modeling,
+non-proportional hazard models, competing risk analysis, penalized
+estimation, and cumulative incidence curve estimation. By searching for
+appropriate keywords in the \code{DESCRIPTION} file of these packages,
+we found 60 relevant packages. These 60 packages were further reduced
+through manual curation to a subset of 14 packages. These packages
+appears in Table \ref{tab:surv-pkgs}, along with some of the
+functionalities they offer.
+\textcolor{red}{JESSE: It's not clear how you selected this subset, going from 60 to 14 is a big drop.}
+
+Several packages implement penalized estimation for the Cox model:
+\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
+\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
+\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
+packages also include penalized estimation in the context of Cox models
+with time-varying coefficients: \pkg{CoxRidge}
+\citeyearpar{perperoglou}, \pkg{rstpm2} \citeyearpar{clements_liu} and
+\pkg{survival} \citeyearpar{survival-package}. On the other hand,
+\pkg{casebase} provides penalized estimation in the context of
+parametric hazards. To our knowledge, this is the only package to offer
+this functionality.
+\textcolor{red}{JESSE: Is this true? Because it seems to be a logical conclusion from your paragraph.}
+
+Parametric survival models are implemented in a handful of packages:
+\pkg{CFC} \citeyearpar{sharabiani_mahani_2019}, \pkg{flexsurv}
+\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
+\pkg{rsptm2} \citeyearpar{clements_liu}, \pkg{mets}
+\citeyearpar{scheike2014estimating}, and \pkg{survival}. The types of
+models they allow vary for each package. For example, \pkg{SmoothHazard}
+is limited to Weibull distributions \citeyearpar{smoothHazard}, whereas
+both \pkg{flexsurv} and \pkg{survival} allow users to supply any
+distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard},
+\pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of
+time using splines, which allows flexible modeling of the hazard
+function. Moreover, \pkg{flexsurv} has the ability to estimate both
+scale and shape parameters for a variety of parametric families (see
+Table \ref{tab:dists}). As discussed above, \pkg{casebase} can model any
+parametric family whose log-hazard can be expressed as a linear model of
+covariates (including time). Therefore, our package allows the user to
+model the effect of time using splines, and through interaction terms
+involving covariates and time, it also allows user to fit time-varying
+coefficient models. However, we do not explicitly model any shape
+parameter, unlike \pkg{flexsurv}.
+
+Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
+\pkg{mets} and \pkg{survival} contain implementations for competing
+risks. The differentiating factor between these packages and
+\pkg{casebase} is that \pkg{casebase} handles competing risks through
+multiple logistic regression.
+\textcolor{red}{JESSE: There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}.
+Could you incorporate some of their discussion, with proper citations?
+The biggest omission right now is \pkg{cmprsk}, which we actually use below.}
+
+Finally, several packages include functions to estimate the cumulative
+incidence function. The corresponding methods generally fall into two
+categories: transformation of the estimated hazard function, and
+semi-parametric estimation of the baseline hazard. The first category
+broadly corresponds to parametric survival models, where the full hazard
+is explictly modeled. Using this estimate, the survival function and the
+cumulative incidence function can be obtained using their functional
+relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below).
+Packages including this functionality include \pkg{CFC}, \pkg{Flexsurv},
+\pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows
+this approach for both single-event and competing-event analyses. The
+second category outlined above broadly corresponds to Cox models. These
+models do not model the full hazard function, and therefore the baseline
+hazard needs to be estimated separately in order to estimate the
+survival function. This is achieved using semi-parametric estimators
+(e.g.~Breslow's estimator). Packages that implement this approach
+include \pkg{RiskRegression}, \pkg{rstpm2}, and \pkg{survival}. As
+mentioned in the introduction, a key distinguishing factor between these
+two approaches is that the first category leads to smooth estimates of
+the cumulative incidence function, whereas the second category produces
+estimates in the form of step-wise functions. This was one of the main
+motivations for introducting case-base sampling in survival analysis.
+
+\begin{table}[ht]
+\centering
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{ccccccccc}
+\hline
+\textbf{Package}        & \textbf{Competing Risks} & \textbf{Non-proportional} & \textbf{Penalization} & \textbf{Splines} & \textbf{Parametric} & \textbf{Semi-parametric} & \textbf{Interval/left Censoring} & \textbf{Absolute Risk}           \\
+\hline
+\textbf{casebasee} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
+\textbf{CFC} \cite{sharabiani_mahani_2019}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
+\textbf{coxRidge} \cite{perperoglou}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{crrp} \cite{fu}           & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
+\textbf{fastcox} \cite{yi_zou}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{Flexrsurv} \cite{flexrsurv}      &                          & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
+\\
+\textbf{Flexsurv} \cite{flexsurv}       & x                        & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
+\\
+\textbf{glmnet} \cite{regpathcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{glmpath} \cite{park_hastie}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{mets} \cite{scheike2014estimating}           & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
+\textbf{penalized} \cite{goeman_meijer2019}      &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{RiskRegression} \cite{gerds_blanche} &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
+\textbf{Rstpm2} \cite{clements_liu}         &                          & x                         & x                     & x                & x                   & x                        & x                                & Cumm haz                         \\
+\textbf{SmoothHazard} \cite{smoothHazard}   &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
+\textbf{Survival} \cite{survival-package}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x
+\end{tabular}%
+}
+\caption{Different features of interest in various survival packages.}
+\label{tab:surv-pkgs}
+\end{table}
 
@@
 
-\hypertarget{literature-review}{%
-\section{Literature review}\label{literature-review}}
-
-To conduct our review on current methods available in survival analysis,
-a scraper was coded to examine all packages in the survival task view on
-CRAN \citeyearpar{survTaskView}. There was a total of 259 packages, many
-of which are not relevant to the problems casebase addresses. To filter
-these packages, a scraper was used to search for competing risks,
-cumulative incidence functions, non- proportional hazards and
-penalization. This reduced our set of 259 packages down to 60. From
-here, we manually examined each package to determine which ones are
-comparable to casebase, with either a different approach to the same
-problems or future features we would like to add to casebase.
-
-These packages can belong to three main categories: non-proportional
-hazards, regularization and competing risks. Besides these three main
-categories, there are extra features of interest as well. Table
-\ref{tab:surv-pkgs} summarizes the findings.
-
-Many packages exist that implement semi-parametric Cox models. Of those,
-\pkg{glmnet} \citeyearpar{regpathcox}???, \pkg{glmpath}
-\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
-\pkg{RiskRegression} \citeyearpar{gerds_blanche} introduce penalization
-in their model fitting. \pkg{CoxRidge} \citeyearpar{perperoglou},
-\pkg{rstpm2} \citeyearpar{clements_liu} and \pkg{survival}
-\citeyearpar{survival-package} include penalization, but they implement
-the extended cox model permitting intuitive non proportional hazard
-modeling while the previous four do not. In terms of penalization, the
-benefits of \pkg{casebase} comes from its flexibility. Linear models
-have a wide tool set, considerably larger than those made for Cox
-regression. \pkg{casebase} permits many regularization tools developed
-for linear regression to be repurposed within a survival context.
-
-Parametric models permit non-proportional hazards and are implemented in
-a handful of packages. Of interest are \pkg{CFC}
-\citeyearpar{sharabiani_mahani_2019}, \pkg{flexsurv}
-\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
-\pkg{rsptm2} \citeyearpar{clements_liu} and \pkg{survival}
-\citeyearpar{survival-package}. \pkg{SmoothHazard} is limited to Weibull
-distributions \citeyearpar{smoothHazard}, whereas \pkg{flexsurv} and
-\pkg{survival} contain implementations for users to supply any
-distribution they would like \citeyearpar{survival-package},
-\citeyearpar{flexsurv}. \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets}
-and \pkg{rstpm2} can use splines, permitting a flexible fit over time
-\citeyearpar{flexsurv}, \citeyearpar{smoothHazard},
-\citeyearpar{scheike2014estimating}, \citeyearpar{clements_liu}. Though
-\pkg{casebase} is limited to three distributions at the moment, it also
-uses splines allowing for distributions that may not be consistent over
-survival time.
-
-Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
-\pkg{mets} and \pkg{survival} contain implementations for competing
-risks \citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
-\citeyearpar{scheike2014estimating}, \citeyearpar{survival-package}. The
-differentiating factor between these packages and \pkg{casebase} is that
-\pkg{casebase} handles competing risks through multiple logistic
-regression. \pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can
-handle interval censoring, whereas \pkg{casebase} currently cannot.
-Interval censoring is a simple feature to implement and is a potential
-improvement to add to \pkg{casebase}.
-
-Of all the packages listed in Table \ref{tab:surv-pkgs}, \pkg{CFC},
-\pkg{Flexsurv}, \pkg{mets}, \pkg{RiskRegression}, \pkg{rstpm2} and
-\pkg{survival} have implementations for cumulative incidence
-\citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
-\citeyearpar{scheike2014estimating}, \citeyearpar{gerds_blanche},
-\citeyearpar{clements_liu}, \citeyearpar{survival-package}. Since
-\pkg{casebase} is parametric, the baseline hazard is known and requires
-integration to retrieve the cumulative incidence. This is similar to the
-approaches of \pkg{CFC} and \pkg{flexsurv}
-\citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv}, while the
-other packages use semi-parametric approaches to retrieve the baseline
-after fitting \citeyearpar{scheike2014estimating},
-\citeyearpar{gerds_blanche}. \pkg{rstpm2} and \pkg{survival} use both
-for their respective parametric and semi-parametric model
-implementations \citeyearpar{clements_liu},
-\citeyearpar{survival-package}.
-
-\begin{table}[ht]
-\centering
-\resizebox{\textwidth}{!}{%
-\begin{tabular}{ccccccccc}
-\hline
-\textbf{Package}        & \textbf{Competing Risks} & \textbf{Non-proportional} & \textbf{Penalization} & \textbf{Splines} & \textbf{Parametric} & \textbf{Semi-parametric} & \textbf{Interval/left Censoring} & \textbf{Absolute Risk}           \\
-\hline
-\textbf{casebase} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
-\textbf{CFC} \cite{sharabiani_mahani_2019}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
-\textbf{coxRidge} \cite{perperoglou}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{crrp} \cite{fu}           & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
-\textbf{fastcox} \cite{yi_zou}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{Flexrsurv} \cite{flexrsurv}      &                          & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
-\\
-\textbf{Flexsurv} \cite{flexsurv}       & x                        & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
-\\
-\textbf{glmnet} \cite{regpathcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{glmpath} \cite{park_hastie}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{mets} \cite{scheike2014estimating}           & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
-\textbf{penalized} \cite{goeman_meijer2019}      &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{RiskRegression} \cite{gerds_blanche} &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
-\textbf{Rstpm2} \cite{clements_liu}         &                          & x                         & x                     & x                & x                   & x                        & x                                & Cumm haz                         \\
-\textbf{SmoothHazard} \cite{smoothHazard}   &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
-\textbf{Survival} \cite{survival-package}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x
-\end{tabular}%
-}
-\caption{Different features of interest in various survival packages.}
-\label{tab:surv-pkgs}
-\end{table}
-
-\hypertarget{theoretical-details}{%
-\section{Theoretical details}\label{theoretical-details}}
-
-As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
-key idea behind case-base sampling is to discretize the study base into
-an infinite amount of \emph{person moments}. These person moments are
-indexed by both an individual in the study and a time point, and
-therefore each person moment has a covariate profile, an exposure status
-and an outcome status attached to it. We note that there is only a
-finite number of person moments associated with the event of interest
-(what Hanley \& Miettinen call the \emph{case series}). The case-base
-sampling refers to the sampling from the base of a representative finite
-sample called the \emph{base series}.
-
-As shown by Saarela \& Arjas \citeyearpar{saarela2015non} (and further
-expanded in Saarela \citeyearpar{saarela2016case}), writing the
-likelihood arising from this data-generating mechanism using the
-framework of non-homogeneous Poisson processes, we eventually reach an
-expression where each person-moment's contribution is of the form
-\[\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},\] where \(N(t)\) is the counting
-process associated with the event of interest, \(h(t)\) is the
-corresponding hazard function, and \(\rho(t)\) is the hazard function
-for the Poisson process associated with case-base sampling. This
-parametric form suggests that we can readily estimate log-hazards of the
-form \(\log(h(t)) = g(t; X)\) using logistic regression, where each
-observation corresponds to a person moment, the function \(g(t; X)\) is
-linear in a finite number of parameters, and where we treat
-\(-\log(\rho(t))\) as an offset.
-
-In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
-suggest performing case-base sampling \emph{uniformly}, i.e.~to sample
-the base series uniformly from the study base. In terms of Poisson
-processes, this sampling strategy corresponds essentially to a
-time-homogeneous Poisson process with intensity equal to \(b/B\), where
-\(b\) is the number of sampled observations in the base series, and
-\(B\) is the total population-time for the study base. More complex
-examples are also available; see for example Saarela \& Arjas
-\citeyearpar{saarela2015non}, where the probabilities of the sampling
-mechanism are proportional to the cardiovascular disease event rate
-given by the Framingham score.
-
-The \texttt{casebase} package fits the family of hazard functions of the
-form
-
-\[ h(t;X) = \exp[g(t;X)] \] where \(t\) denotes time and \(X\), the
-individual's covariate profile. Different functions of \(t\) lead to
-different parametric hazard models. The simplest of these models is the
-one-parameter exponential distribution which is obtained by taking the
-hazard function to be constant over the range of \(t\).
-
-\[ h(t;X) = \exp(\beta_0 + \beta_1 X) \]
-
-The instantaneous failure rate is independent of \(t\), so that the
-conditional chance of failure in a time interval of specified length is
-the same regardless of how long the individual has been in the study.
-This is also known as the \emph{memoryless property} (Kalbfleisch and
-Prentice, 2002).
-
-The Gompertz hazard model is given by including a linear term for time:
-
-\[ h(t;X)  = \exp(\beta_0 + \beta_1 t + \beta_2 X) \]
-
-Use of \(\log(t)\) yields the Weibull hazard which allows for a power
-dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
-
-\[ h(t;X)  = \exp(\beta_0 + \beta_1 \log(t) + \beta_2 X) \]
-
-For competing-risk analyses with \(J\) possible events, we can show that
-each person-moment's contribution of the likelihood is of the form
-
-\[\frac{h_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^Jh_j(t)},\]
-
-where \(N_j(t)\) is the counting process associated with the event of
-type \(j\) and \(h_j(t)\) is the corresponding hazard function. As may
-be expected, this functional form is similar to the terms appearing in
-the likelihood function for multinomial
-regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1\end{align*}}
-
 \hypertarget{implementation-details}{%
@@
 the model fit. For the case of a single event, it uses the familiar
-identity \[S(t) = \exp\left(-\int_0^t h(u;X) du\right).\] The integral
-is computed using either the \code{stats::integrate} function or
-Monte-Carlo integration. The risk function (or cumulative incidence
-function) is then defined as
-
-\[ CI(t) = 1 - S(t).\]
+identity \begin{equation}\label{eqn:surv}
+S(t) = \exp\left(-\int_0^t h(u;X) du\right).
+\end{equation} The integral is computed using either the
+\code{stats::integrate} function or Monte-Carlo integration. The risk
+function (or cumulative incidence function) is then defined as
+\begin{equation}\label{eqn:CI}
+CI(t) = 1 - S(t).
+\end{equation}
 
@@
 regularization within casebase (REF), while comparing their absolute
-risk preditions for a new individual. A description of each variable can
-be found in Table\textasciitilde{}\ref{tab:support1} and a breakdown of
-each categorical variable in Table\textasciitilde{}\ref{tab:support2}.
-The data was downloaded from the main author's website, and it is also
-available as part of the casebase package.
+risk predictions for a new individual. A description of each variable
+can be found in Table\textasciitilde{}\ref{tab:support1} and a breakdown
+of each categorical variable in
+Table\textasciitilde{}\ref{tab:support2}. The data was downloaded from
+the main author's website, and it is also available as part of the
+casebase package.
 
@@
 and a kaplan-meier survival curve. The regularized cox regression
-absolute risk curve required two extra steps to retrive. Coxnet from the
-glmnet package has tools to fit a regularized cox model, but requires
-that a survival object skeleton of the selected variables, so that the
-survival package tools can be used to retrive the absolute risk. We
-handle this by fitting a regularized cox model with the glmnet package,
-and a cox model from the survival package with the selected variables
-from the regularized fit. This second model serves as a skeleton that
-permits the use of survival package functions. The coefficients from the
-regularized model will replace the coeficients in our skeleton. the
-resulting model will have the correct coefficients when integrating, but
-will have an incorrect standard error. For the purposes of this case
-study, we are only interested in the absolute risk curve itself and not
-the standard error.
+absolute risk curve required two extra steps to retrieve. Coxnet from
+the glmnet package has tools to fit a regularized cox model, but
+requires that a survival object skeleton of the selected variables, so
+that the survival package tools can be used to retrieve the absolute
+risk. We handle this by fitting a regularized cox model with the glmnet
+package, and a cox model from the survival package with the selected
+variables from the regularized fit. This second model serves as a
+skeleton that permits the use of survival package functions. The
+coefficients from the regularized model will replace the coefficients in
+our skeleton. the resulting model will have the correct coefficients
+when integrating, but will have an incorrect standard error. For the
+purposes of this case study, we are only interested in the absolute risk
+curve itself and not the standard error.
 
@@
 the effect of a heart transplant on survival; therefore, the patient is
-considered exposed only after the transplant has occured.
+considered exposed only after the transplant has occurred.
 
@@
 
+\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval
+censoring, whereas \pkg{casebase} currently cannot. Interval censoring
+is a simple feature to implement and is a potential improvement to add
+to \pkg{casebase}.
+
 \hypertarget{environment-details}{%
@@
 
-This report was generated on 2019-12-08 12:31:41 using the following
+This report was generated on 2019-12-10 12:28:40 using the following
 computational environment and dependencies:
@@
 #>  tz       America/Winnipeg            
-#>  date     2019-12-08                  
+#>  date     2019-12-10                  
 #> 
@@
 \begin{CodeOutput}
-#> Local:    review-max /home/mturgeon/Documents/git_repositories/cbpaper
-#> Head:     [69663bb] 2019-12-08: Merge branch 'review-max' of https://github.com/sahirbhatnagar/cbpaper into review-max
+#> Local:    review /home/mturgeon/Documents/git_repositories/cbpaper
+#> Head:     [4435e71] 2019-12-08: Merge branch 'review-max'
 \end{CodeOutput}

### 2019-12-11 analysis/paper/paper.Rmd
edit third case study

@@
   \usepackage{graphicx}
+  \DeclareUnicodeCharacter{2500}{-}
 output: 
@@
     "", "", "PB (156)",
-    "Age", "Age of patient (years)", "4–62", 
+    "Age", "Age of patient (years)", "4--62", 
     "", "", "30.47 (13.04)",
-    "Ftime", "Failure time (months)", "0.13–131.77",
+    "Ftime", "Failure time (months)", "0.13--131.77",
     "", "", "20.28 (30.78)",
@@
    &  & PB (156) \\ 
-  Age & Age of patient (years) & 4–62 \\ 
+  Age & Age of patient (years) & 4--62 \\ 
    &  & 30.47 (13.04) \\ 
-  Ftime & Failure time (months) & 0.13–131.77 \\ 
+  Ftime & Failure time (months) & 0.13--131.77 \\ 
    &  & 20.28 (30.78) \\ 
@@
 
-# Case study 3--Study to Understand Prognoses Preferences Outcomes and Risks of Treatment
-
-We examined the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset (REF from main site).
-the SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital.
-Imputation was conducted using the default methods in the mice package in R, available on CRAN.
-Before imputation, there was a total of 1000 individuals and 35 variables.
+# Case study 3--SUPPORT Data
+
+We examined the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [-@knaus1995support].
+The SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital.
+Imputation was conducted using the default methods in the \pkg{mice} package in \proglang{R}, available on CRAN.
+\textcolor{red}{Why was this method used?
+The Vanderbilt University website lists describes another method for imputation that we could have used.} Before imputation, there was a total of 1000 individuals and 35 variables.
 After imputation, we have 474 observations where 67.5% of individuals in the study experience the event of interest.
-The SUPPORT dataset will be used to demonstrate regularization within casebase (REF), while comparing their absolute risk predictions for a new individual.
+The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.
 A description of each variable can be found in Table~\ref{tab:support1} and a breakdown of each categorical variable in Table~\ref{tab:support2}.
-The data was downloaded from the main author's website, and it is also available as part of the casebase package.
+The data was downloaded from the main author's website, and it is also available as part of the \pkg{casebase} package.
 
@@
 \end{tabular}
-\caption{A description of each variable in the SUPPORT dataset, taken from (REF).}
+\caption{A description of each variable in the SUPPORT dataset.}
 \label{tab:support1}
@@
 \end{tabular}
-\caption{A description of each level within each categorical variable in the dataset, taken from REF.}
+\caption{A description of each level within each categorical variable in the dataset.}
 \label{tab:support2}
@@
 Hospital death as a covariate was removed, as this is directly informative of death.
-The glmnet package(REF) on CRAN was used to introduce elastic net, lasso and ridge regressions to casebase.
+The \pkg{glmnet} package on CRAN was used to introduce elastic net, lasso and ridge regressions to \pkg{casebase}.
 To visualize the effect of imputation on incidence density, we can refer to the pre-imputation population-time plot in Figure \ref{fig:ptPreI} and post-imputation population time plot in Figure \ref{fig:ptPostI}.
@@
 ```{r supportSetUp, echo = FALSE, eval = eval_cs3}
-completeData=casebase::support
+completeData <- casebase::support
 #keep one individual out, at random, for which we will develop an 
 #absolute risk curve.
-sam=sample(1:nrow(completeData), 1)
+sam <- sample(1:nrow(completeData), 1)
 #Set workingData to the remaining individuals in the data
 #workingData=completeData[-c(sam),]
-workingCompleteData=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))[,-c(1)]
+workingCompleteData <- as.matrix(sparse.model.matrix(death~ . - d.time, data = completeData))[,-c(1)]
 #Create u and xCox, which will be used when fitting Cox with glmnet.
-newData=workingCompleteData[sam,]
-x=workingCompleteData[-c(sam),]
+newData <- workingCompleteData[sam,]
+x <- workingCompleteData[-c(sam),]
 #x and y will be used to fit the casebase model under glmnet.
 #x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
-y=data.matrix(completeData[-c(sam),c(34,4)])
+y <- data.matrix(completeData[-c(sam), c(34, 4)])
 ```
@@
 Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study."}
-#demonstrating incidence density before imputation
-plot(casebase::popTime(as.data.frame(support),time = "d.time",event = "death"))
+popTime_pre <- popTime(as.data.frame(support),
+                       time = "d.time", event = "death")
+plot(popTime_pre)
 ```
@@
 Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study."}
-#demonstrating incidence density after imputation
-plot(casebase::popTime(as.data.frame(workingCompleteData),time = "d.time",event = "death"))
-```
-Our main objective is to compute the absolute risk of death for a given set of covariates, using regularization when fitting our model.
-First, we fit a smooth hazard to the data; for the sake of this example, we opted for a linear term for time.
-As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model.
+popTime_post <- popTime(as.data.frame(workingCompleteData),
+                        time = "d.time", event = "death")
+plot(popTime_post)
+```
+
+Our main objective is to estimate the cumulative incidence of death for a given covariate profile.
+Given th number of covariates available, we will do this using regularized estimation.
+First, we fit the hazard function; for the sake of simplicity, we opted for a linear term for time.
+As \pkg{casebase} makes use of the \pkg{glmnet} package, we will use the function \code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}.
+Specifically, we input a matrix \code{y}, which contains the time and event variables, and a matrix \code{x}, which contains all other variables we would like to include in the model.
 For the SUPPORT dataset, all factor variables were converted into dummy variables.
-Here, we used lasso penalization by setting alpha to 1.
+Here, we used lasso penalization by setting \code{alpha} to 1.
+
 ```{r CBhazard, echo = TRUE, eval = eval_cs3}
-#Using casebase's glmnet implementation, using splines and a case to base ratio of 100,
-#to determine the hazard function
-cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
-                             formula_time = ~d.time,alpha=1,ratio=100)
-```
-Then, we take our model and use the absoluteRisk function to integrate and retrieve our absolute risk function.
+cb.Model <- fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", 
+                                event = "death", formula_time = ~ d.time, 
+                                alpha = 1, ratio = 100)
+```
+
+Then, we take our model object and use the \code{absoluteRisk} function to compute our estimate of the cumulative incidence function.
+
 ```{r CBabsolute, echo = TRUE, eval = eval_cs3}
-#Estimating the absolute risk curve using the newData parameter.
-casebaseAbsolute=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) ,
-                              s="lambda.1se",method=c("numerical"))
-```
-We will compare this curve to a regularized version of cox regression, and a kaplan-meier survival curve.
-The regularized cox regression absolute risk curve required two extra steps to retrieve.
-Coxnet from the glmnet package has tools to fit a regularized cox model, but requires that a survival object skeleton of the selected variables, so that the survival package tools can be used to retrieve the absolute risk.
-We handle this by fitting a regularized cox model with the glmnet package, and a cox model from the survival package with the selected variables from the regularized fit.
-This second model serves as a skeleton that permits the use of survival package functions.
-The coefficients from the regularized model will replace the coefficients in our skeleton.
-the resulting model will have the correct coefficients when integrating, but will have an incorrect standard error.
-For the purposes of this case study, we are only interested in the absolute risk curve itself and not the standard error.
+casebaseAbsolute <- absoluteRisk(cb.Model,
+                                 time = seq(1, max(completeData$d.time), 0.5),
+                                 newdata = t(newData),
+                                 s = "lambda.1se", method = "numerical")
+```
+
+We will compare this curve to a regularized version of Cox regression, and a Kaplan-Meier survival curve.
+Using the \pkg{glmnet} package, we can fit the Cox regression model using regularized estimation.
+However, \pkg{glmnet} does not provide any functionality for estimating the baseline hazard, and the functionalities in the \pkg{survival} package do not work on the output from \code{cv.glmnet}.
+To circumvent this issue and estimate the baseline hazard, we followed the following steps.
+First, we fitted a non-regularized Cox model using only the variables selected by \pkg{glmnet}; the baseline hazard will be estimated using this object.
+Then, we replaced the estimates of the regression coefficients by the estimates from \pkg{glmnet}.
+Finally, this modified \code{coxph} object is fed to the \code{survival::survfit} function to obtain the estimated survival function.
 
 ```{r coxHazAbsolute, echo = TRUE, eval = eval_cs3}
-#Create u and xCox, which will be used when fitting Cox with glmnet.
-u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
-xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))[-c(sam),]
-#hazard for cox using glmnet
-coxglmFit=glmnet::cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
-#convergence demonstrated in plot
-plot(coxglmFit)
+# Create u and xCox, which will be used when fitting Cox with glmnet.
+u <- survival::Surv(time = as.numeric(y[,2]), 
+                    event = as.factor(y[,1]))
+xCox <- as.matrix(sparse.model.matrix(death~ . - d.time,
+                                      data = completeData))[-c(sam),]
+# hazard for cox using glmnet
+coxglmFit <- glmnet::cv.glmnet(x = xCox, y = u,
+                               family = "cox", alpha = 1)
+# convergence demonstrated in plot
+# plot(coxglmFit)
 #taking the coefficient estimates for later use
@@
 nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
-#creating a new dataset that only contains the covariates chosen through glmnet.
-#cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
-cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
-#newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
-#fitting a cox model using regular estimation, however we will not keep it.
-#this is used more as an object place holder.
-coxFit <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
-#The coefficients of this object will be replaced with the estimates from coxglmFit.
-#Doing so makes it so that everything is invalid aside from the coefficients.
-#In this case, all we need to estimate the absolute risk is the coefficients.
-#Std. error would be incorrect here, if we were to draw error bars.
-coxFit$coefficients<-nonzero_coef_cox@x
-#Fitting absolute risk curve for cox+glmnet
-newDataCox=newData[nonzero_covariate_cox$X1-1]
-
-abCoxFit<-survival::survfit(coxFit,newdata=as.data.frame(t(newDataCox)),time = seq(0,max(completeData$d.time), 1),type="breslow")
-```
-We used the survival package to calculate the kaplan-meier absolute risk function. 
+# creating a new dataset that only contains the covariates chosen through glmnet.
+cleanCoxData <- as.data.frame(cbind(as.numeric(y[,2]),
+                                    as.numeric(y[,1]),
+                                    xCox[, nonzero_covariate_cox$X1]))
+# fitting a cox model using regular estimation, however we will not keep it.
+# this is used more as an object place holder.
+coxFit <- survival::coxph(Surv(time = V1, event = V2) ~ ., 
+                          data = cleanCoxData)
+# The coefficients of this object will be replaced 
+# with the estimates from coxglmFit.
+coxFit$coefficients <- nonzero_coef_cox@x
+# Fitting CI curve for cox+glmnet
+newDataCox <- newData[nonzero_covariate_cox$X1 - 1]
+
+abCoxFit <- survival::survfit(coxFit, newdata = as.data.frame(t(newDataCox)),
+                              time = seq(0, max(completeData$d.time), 1),
+                              type = "breslow")
+```
+
+We used the \pkg{survival} package to calculate the Kaplan-Meier estimate of the survival function.
+
 ```{r fitKM+abRisk, echo = TRUE, eval = eval_cs3}
-#creating a surv object to be used to fit an unadjusted absolute risk curve.
-# (Kaplan-Meier risk curve)
-km <- survival::Surv(time = completeData$d.time, event = completeData$death)
-abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
-```
-
-Now that our three absolute risk functions have been calculated, we compare them all in Figure \ref{fig:abSC}.
-Both coxnet and casebase decrease the absolute risk by the end of the study, in comparison to kaplan-meier.
+km <- survival::Surv(time = completeData$d.time, 
+                     event = completeData$death)
+abKm <- survival::survfit(km ~ 1,
+                          type = 'kaplan-meier')
+```
+
+Now that our three cumulative incidence functions have been calculated, we compare them all in Figure \ref{fig:abSC}.
+Both \pkg{glmnet} and \pkg{casebase} have a lower estimated cumulative incidence by the end of the study, when compared to Kaplan-Meier.
+
 ```{r abSupportComparison, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
-# A plot to compare all three Absolute risk (Commulative Incidence)
-plot(casebaseAbsolute,type='l',col="black",lwd=2,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
-lines(abCoxFit ,col="red",fun="event",lwd=3,conf.int = FALSE)
-lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
+plot(casebaseAbsolute, type='l', col = "black", lwd = 2, 
+     main = "Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves", 
+     xlab = "Survival-Time", ylab = "Cumulative Incidence", ylim = c(0,1))
+lines(abCoxFit, col = "red", fun = "event", lwd = 3, conf.int = FALSE)
+lines(abKm, col = "blue", fun = "event", lwd = 3, conf.int = FALSE)
 legend("bottom", 
-       legend = c( "Casebase (Lasso+linear)","semi-parametric (Cox)+glmnet","KM curve"), 
-       col = c("black","red","Blue"),
-       lty = c(1, 1, 1), 
-       bg = "gray90")
+       legend = c("casebase", "Cox+glmnet", "KM curve"), 
+       col = c("black", "red", "blue"), lty = c(1, 1, 1), bg = "gray90")
 ```
@@
 The purpose of the next case study is to show how \pkg{casebase} can also be used for survival analysis problems that do not share the properties above.
-To this end, we will use the Stanford Heart Transplant Data [@clark1971cardiac,@crowley1977covariance]; we will use the version available in the \pkg{survival} package.
+To this end, we will use the Stanford Heart Transplant Data [@clark1971cardiac,@crowley1977covariance]; we will use the version available in the \pkg{surivival} package.
 
@@
 
-In the following table we provide a comparison between the Cox model and case-base sampling:
+<!-- In the following table we provide a comparison between the Cox model and case-base sampling: -->
 

### 2019-12-11 analysis/paper/paper.tex
edit third case study

@@
 
-\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx}
+\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx} \DeclareUnicodeCharacter{2500}{-}
 
@@
 
-\hypertarget{case-study-3study-to-understand-prognoses-preferences-outcomes-and-risks-of-treatment}{%
-\section{Case study 3--Study to Understand Prognoses Preferences
-Outcomes and Risks of
-Treatment}\label{case-study-3study-to-understand-prognoses-preferences-outcomes-and-risks-of-treatment}}
+\hypertarget{case-study-3support-data}{%
+\section{Case study 3--SUPPORT Data}\label{case-study-3support-data}}
 
 We examined the Study to Understand Prognoses Preferences Outcomes and
-Risks of Treatment (SUPPORT) dataset (REF from main site). the SUPPORT
-dataset tracks death for individuals who are considered seriously ill
-within a hospital. Imputation was conducted using the default methods in
-the mice package in R, available on CRAN. Before imputation, there was a
-total of 1000 individuals and 35 variables. After imputation, we have
-474 observations where 67.5\% of individuals in the study experience the
-event of interest. The SUPPORT dataset will be used to demonstrate
-regularization within casebase (REF), while comparing their absolute
-risk predictions for a new individual. A description of each variable
-can be found in Table\textasciitilde{}\ref{tab:support1} and a breakdown
-of each categorical variable in
-Table\textasciitilde{}\ref{tab:support2}. The data was downloaded from
-the main author's website, and it is also available as part of the
-casebase package.
+Risks of Treatment (SUPPORT) dataset \citeyearpar{knaus1995support}. The
+SUPPORT dataset tracks death for individuals who are considered
+seriously ill within a hospital. Imputation was conducted using the
+default methods in the \pkg{mice} package in \proglang{R}, available on
+CRAN.
+\textcolor{red}{Why was this method used? The Vanderbilt University website lists describes another method for imputation that we could have used.}
+Before imputation, there was a total of 1000 individuals and 35
+variables. After imputation, we have 474 observations where 67.5\% of
+individuals in the study experience the event of interest. The SUPPORT
+dataset will be used to demonstrate regularization within casebase,
+while comparing their absolute risk predictions for a new individual. A
+description of each variable can be found in
+Table\textasciitilde{}\ref{tab:support1} and a breakdown of each
+categorical variable in Table\textasciitilde{}\ref{tab:support2}. The
+data was downloaded from the main author's website, and it is also
+available as part of the \pkg{casebase} package.
 
@@
 Hospital death as a covariate was removed, as this is directly
-informative of death. The glmnet package(REF) on CRAN was used to
-introduce elastic net, lasso and ridge regressions to casebase. To
+informative of death. The \pkg{glmnet} package on CRAN was used to
+introduce elastic net, lasso and ridge regressions to \pkg{casebase}. To
 visualize the effect of imputation on incidence density, we can refer to
@@
 \begin{CodeInput}
-R> #demonstrating incidence density before imputation
-R> plot(casebase::popTime(as.data.frame(support),time = "d.time",event = "death"))
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> #demonstrating incidence density after imputation
-R> plot(casebase::popTime(as.data.frame(workingCompleteData),time = "d.time",event = "death"))
-\end{CodeInput}
-\end{CodeChunk}
-
-Our main objective is to compute the absolute risk of death for a given
-set of covariates, using regularization when fitting our model. First,
-we fit a smooth hazard to the data; for the sake of this example, we
-opted for a linear term for time. As casebase makes use of the glmnet
-package, we interact with the fitsmoothhazard.fit function using a
-matrix interface, where y contains the time and event variables and x
+R> popTime_pre <- popTime(as.data.frame(support),
+R>                        time = "d.time", event = "death")
+R> plot(popTime_pre)
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> popTime_post <- popTime(as.data.frame(workingCompleteData),
+R>                         time = "d.time", event = "death")
+R> plot(popTime_post)
+\end{CodeInput}
+\end{CodeChunk}
+
+Our main objective is to estimate the cumulative incidence of death for
+a given covariate profile. Given th number of covariates available, we
+will do this using regularized estimation. First, we fit the hazard
+function; for the sake of simplicity, we opted for a linear term for
+time. As \pkg{casebase} makes use of the \pkg{glmnet} package, we will
+use the function \code{fitsmoothhazard.fit}, which follows the syntax
+from \code{glmnet}. Specifically, we input a matrix \code{y}, which
+contains the time and event variables, and a matrix \code{x}, which
 contains all other variables we would like to include in the model. For
 the SUPPORT dataset, all factor variables were converted into dummy
-variables. Here, we used lasso penalization by setting alpha to 1.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> #Using casebase's glmnet implementation, using splines and a case to base ratio of 100,
-R> #to determine the hazard function
-R> cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
-R>                              formula_time = ~d.time,alpha=1,ratio=100)
-\end{CodeInput}
-\end{CodeChunk}
-
-Then, we take our model and use the absoluteRisk function to integrate
-and retrieve our absolute risk function.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> #Estimating the absolute risk curve using the newData parameter.
-R> casebaseAbsolute=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) ,
-R>                               s="lambda.1se",method=c("numerical"))
-\end{CodeInput}
-\end{CodeChunk}
-
-We will compare this curve to a regularized version of cox regression,
-and a kaplan-meier survival curve. The regularized cox regression
-absolute risk curve required two extra steps to retrieve. Coxnet from
-the glmnet package has tools to fit a regularized cox model, but
-requires that a survival object skeleton of the selected variables, so
-that the survival package tools can be used to retrieve the absolute
-risk. We handle this by fitting a regularized cox model with the glmnet
-package, and a cox model from the survival package with the selected
-variables from the regularized fit. This second model serves as a
-skeleton that permits the use of survival package functions. The
-coefficients from the regularized model will replace the coefficients in
-our skeleton. the resulting model will have the correct coefficients
-when integrating, but will have an incorrect standard error. For the
-purposes of this case study, we are only interested in the absolute risk
-curve itself and not the standard error.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> #Create u and xCox, which will be used when fitting Cox with glmnet.
-R> u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
-R> xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))[-c(sam),]
-R> #hazard for cox using glmnet
-R> coxglmFit=glmnet::cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
-R> #convergence demonstrated in plot
-R> plot(coxglmFit)
+variables. Here, we used lasso penalization by setting \code{alpha} to
+1.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> cb.Model <- fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", 
+R>                                 event = "death", formula_time = ~ d.time, 
+R>                                 alpha = 1, ratio = 100)
+\end{CodeInput}
+\end{CodeChunk}
+
+Then, we take our model object and use the \code{absoluteRisk} function
+to compute our estimate of the cumulative incidence function.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> casebaseAbsolute <- absoluteRisk(cb.Model,
+R>                                  time = seq(1, max(completeData$d.time), 0.5),
+R>                                  newdata = t(newData),
+R>                                  s = "lambda.1se", method = "numerical")
+\end{CodeInput}
+\end{CodeChunk}
+
+We will compare this curve to a regularized version of Cox regression,
+and a Kaplan-Meier survival curve. Using the \pkg{glmnet} package, we
+can fit the Cox regression model using regularized estimation. However,
+\pkg{glmnet} does not provide any functionality for estimating the
+baseline hazard, and the functionalities in the \pkg{survival} package
+do not work on the output from \code{cv.glmnet}. To circumvent this
+issue and estimate the baseline hazard, we followed the following steps.
+First, we fitted a non-regularized Cox model using only the variables
+selected by \pkg{glmnet}; the baseline hazard will be estimated using
+this object. Then, we replaced the estimates of the regression
+coefficients by the estimates from \pkg{glmnet}. Finally, this modified
+\code{coxph} object is fed to the \code{survival::survfit} function to
+obtain the estimated survival function.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Create u and xCox, which will be used when fitting Cox with glmnet.
+R> u <- survival::Surv(time = as.numeric(y[,2]), 
+R>                     event = as.factor(y[,1]))
+R> xCox <- as.matrix(sparse.model.matrix(death~ . - d.time,
+R>                                       data = completeData))[-c(sam),]
+R> # hazard for cox using glmnet
+R> coxglmFit <- glmnet::cv.glmnet(x = xCox, y = u,
+R>                                family = "cox", alpha = 1)
+R> # convergence demonstrated in plot
+R> # plot(coxglmFit)
 R> #taking the coefficient estimates for later use
@@
 R> nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
-R> #creating a new dataset that only contains the covariates chosen through glmnet.
-R> #cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
-R> cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
-R> #newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
-R> #fitting a cox model using regular estimation, however we will not keep it.
-R> #this is used more as an object place holder.
-R> coxFit <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
-R> #The coefficients of this object will be replaced with the estimates from coxglmFit.
-R> #Doing so makes it so that everything is invalid aside from the coefficients.
-R> #In this case, all we need to estimate the absolute risk is the coefficients.
-R> #Std. error would be incorrect here, if we were to draw error bars.
-R> coxFit$coefficients<-nonzero_coef_cox@x
-R> #Fitting absolute risk curve for cox+glmnet
-R> newDataCox=newData[nonzero_covariate_cox$X1-1]
-R> 
-R> abCoxFit<-survival::survfit(coxFit,newdata=as.data.frame(t(newDataCox)),time = seq(0,max(completeData$d.time), 1),type="breslow")
-\end{CodeInput}
-\end{CodeChunk}
-
-We used the survival package to calculate the kaplan-meier absolute risk
-function.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> #creating a surv object to be used to fit an unadjusted absolute risk curve.
-R> # (Kaplan-Meier risk curve)
-R> km <- survival::Surv(time = completeData$d.time, event = completeData$death)
-R> abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
-\end{CodeInput}
-\end{CodeChunk}
-
-Now that our three absolute risk functions have been calculated, we
-compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase
-decrease the absolute risk by the end of the study, in comparison to
-kaplan-meier.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # A plot to compare all three Absolute risk (Commulative Incidence)
-R> plot(casebaseAbsolute,type='l',col="black",lwd=2,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
-R> lines(abCoxFit ,col="red",fun="event",lwd=3,conf.int = FALSE)
-R> lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
+R> # creating a new dataset that only contains the covariates chosen through glmnet.
+R> cleanCoxData <- as.data.frame(cbind(as.numeric(y[,2]),
+R>                                     as.numeric(y[,1]),
+R>                                     xCox[, nonzero_covariate_cox$X1]))
+R> # fitting a cox model using regular estimation, however we will not keep it.
+R> # this is used more as an object place holder.
+R> coxFit <- survival::coxph(Surv(time = V1, event = V2) ~ ., 
+R>                           data = cleanCoxData)
+R> # The coefficients of this object will be replaced 
+R> # with the estimates from coxglmFit.
+R> coxFit$coefficients <- nonzero_coef_cox@x
+R> # Fitting CI curve for cox+glmnet
+R> newDataCox <- newData[nonzero_covariate_cox$X1 - 1]
+R> 
+R> abCoxFit <- survival::survfit(coxFit, newdata = as.data.frame(t(newDataCox)),
+R>                               time = seq(0, max(completeData$d.time), 1),
+R>                               type = "breslow")
+\end{CodeInput}
+\end{CodeChunk}
+
+We used the \pkg{survival} package to calculate the Kaplan-Meier
+estimate of the survival function.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> km <- survival::Surv(time = completeData$d.time, 
+R>                      event = completeData$death)
+R> abKm <- survival::survfit(km ~ 1,
+R>                           type = 'kaplan-meier')
+\end{CodeInput}
+\end{CodeChunk}
+
+Now that our three cumulative incidence functions have been calculated,
+we compare them all in Figure \ref{fig:abSC}. Both \pkg{glmnet} and
+\pkg{casebase} have a lower estimated cumulative incidence by the end of
+the study, when compared to Kaplan-Meier.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> plot(casebaseAbsolute, type='l', col = "black", lwd = 2, 
+R>      main = "Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves", 
+R>      xlab = "Survival-Time", ylab = "Cumulative Incidence", ylim = c(0,1))
+R> lines(abCoxFit, col = "red", fun = "event", lwd = 3, conf.int = FALSE)
+R> lines(abKm, col = "blue", fun = "event", lwd = 3, conf.int = FALSE)
 R> legend("bottom", 
-R>        legend = c( "Casebase (Lasso+linear)","semi-parametric (Cox)+glmnet","KM curve"), 
-R>        col = c("black","red","Blue"),
-R>        lty = c(1, 1, 1), 
-R>        bg = "gray90")
+R>        legend = c("casebase", "Cox+glmnet", "KM curve"), 
+R>        col = c("black", "red", "blue"), lty = c(1, 1, 1), bg = "gray90")
 \end{CodeInput}
@@
 Data \citep[\citet{crowley1977covariance}]{clark1971cardiac}; we will
-use the version available in the \pkg{survival} package.
+use the version available in the \pkg{surivival} package.
 
@@
 
-In the following table we provide a comparison between the Cox model and
-case-base sampling:
-
 \pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval
@@
 
-This report was generated on 2019-12-10 12:28:40 using the following
+This report was generated on 2019-12-11 10:31:00 using the following
 computational environment and dependencies:
@@
 \begin{CodeOutput}
-#> - Session info ----------------------------------------------------------
+#> ─ Session info ──────────────────────────────────────────────────────────
 #>  setting  value                       
@@
 #>  tz       America/Winnipeg            
-#>  date     2019-12-10                  
+#>  date     2019-12-11                  
 #> 
-#> - Packages --------------------------------------------------------------
-#>  package     * version    date       lib source                          
-#>  assertthat    0.2.1      2019-03-21 [1] CRAN (R 3.6.1)                  
-#>  backports     1.1.5      2019-10-02 [1] CRAN (R 3.6.1)                  
-#>  broom         0.5.2      2019-04-07 [1] CRAN (R 3.6.1)                  
-#>  callr         3.3.1      2019-07-18 [1] CRAN (R 3.6.1)                  
-#>  casebase    * 0.2.1.9001 2019-07-29 [1] local                           
-#>  cellranger    1.1.0      2016-07-27 [1] CRAN (R 3.5.0)                  
-#>  cli           1.1.0      2019-03-19 [1] CRAN (R 3.6.1)                  
-#>  codetools     0.2-16     2018-12-24 [4] CRAN (R 3.5.2)                  
-#>  colorspace    1.4-1      2019-03-18 [1] CRAN (R 3.6.1)                  
-#>  crayon        1.3.4      2017-09-16 [1] CRAN (R 3.5.0)                  
-#>  data.table    1.12.2     2019-04-07 [1] CRAN (R 3.6.1)                  
-#>  desc          1.2.0      2018-05-01 [1] CRAN (R 3.5.0)                  
-#>  devtools      2.1.0      2019-07-06 [1] CRAN (R 3.6.1)                  
-#>  digest        0.6.22     2019-10-21 [1] CRAN (R 3.6.1)                  
-#>  dplyr       * 0.8.3      2019-07-04 [1] CRAN (R 3.6.1)                  
-#>  ellipsis      0.3.0      2019-09-20 [1] CRAN (R 3.6.1)                  
-#>  evaluate      0.14       2019-05-28 [1] CRAN (R 3.6.1)                  
-#>  forcats     * 0.4.0      2019-02-17 [1] CRAN (R 3.6.1)                  
-#>  foreach     * 1.4.4      2017-12-12 [1] CRAN (R 3.5.0)                  
-#>  fs            1.3.1      2019-05-06 [1] CRAN (R 3.6.1)                  
-#>  generics      0.0.2      2018-11-29 [1] CRAN (R 3.6.1)                  
-#>  ggplot2     * 3.2.1      2019-08-10 [1] CRAN (R 3.6.1)                  
-#>  glmnet      * 2.0-18     2019-05-20 [1] CRAN (R 3.6.1)                  
-#>  glue          1.3.1      2019-03-12 [1] CRAN (R 3.6.1)                  
-#>  gtable        0.3.0      2019-03-25 [1] CRAN (R 3.6.1)                  
-#>  haven         2.1.1      2019-07-04 [1] CRAN (R 3.6.1)                  
-#>  hms           0.5.2      2019-10-30 [1] CRAN (R 3.6.1)                  
-#>  htmltools     0.3.6      2017-04-28 [1] CRAN (R 3.5.0)                  
-#>  httr          1.4.1      2019-08-05 [1] CRAN (R 3.6.1)                  
-#>  iterators     1.0.9      2017-12-12 [1] CRAN (R 3.5.0)                  
-#>  jsonlite      1.6        2018-12-07 [1] CRAN (R 3.6.1)                  
-#>  knitr         1.23       2019-05-18 [1] CRAN (R 3.6.1)                  
-#>  labeling      0.3        2014-08-23 [1] CRAN (R 3.5.0)                  
-#>  lattice       0.20-38    2018-11-04 [4] CRAN (R 3.5.1)                  
-#>  lazyeval      0.2.2      2019-03-15 [1] CRAN (R 3.6.1)                  
-#>  lifecycle     0.1.0      2019-08-01 [1] CRAN (R 3.6.1)                  
-#>  lubridate   * 1.7.4      2018-04-11 [1] CRAN (R 3.5.0)                  
-#>  magrittr    * 1.5        2014-11-22 [1] CRAN (R 3.5.0)                  
-#>  Matrix      * 1.2-18     2019-11-27 [4] CRAN (R 3.6.1)                  
-#>  memoise       1.1.0      2017-04-21 [1] CRAN (R 3.5.0)                  
-#>  mgcv          1.8-24     2018-06-18 [1] CRAN (R 3.5.1)                  
-#>  modelr        0.1.5      2019-08-08 [1] CRAN (R 3.6.1)                  
-#>  munsell       0.5.0      2018-06-12 [1] CRAN (R 3.6.1)                  
-#>  nlme          3.1-142    2019-11-07 [4] CRAN (R 3.6.1)                  
-#>  pillar        1.4.2      2019-06-29 [1] CRAN (R 3.6.1)                  
-#>  pkgbuild      1.0.3      2019-03-20 [1] CRAN (R 3.6.1)                  
-#>  pkgconfig     2.0.3      2019-09-22 [1] CRAN (R 3.6.1)                  
-#>  pkgload       1.0.2      2018-10-29 [1] CRAN (R 3.6.1)                  
-#>  prettyunits   1.0.2      2015-07-13 [1] CRAN (R 3.5.2)                  
-#>  processx      3.4.1      2019-07-18 [1] CRAN (R 3.6.1)                  
-#>  ps            1.3.0      2018-12-21 [1] CRAN (R 3.6.1)                  
-#>  purrr       * 0.3.3      2019-10-18 [1] CRAN (R 3.6.1)                  
-#>  R6            2.4.1      2019-11-12 [1] CRAN (R 3.6.1)                  
-#>  Rcpp          1.0.3      2019-11-08 [1] CRAN (R 3.6.1)                  
-#>  readr       * 1.3.1      2018-12-21 [1] CRAN (R 3.6.1)                  
-#>  readxl        1.3.1      2019-03-13 [1] CRAN (R 3.6.1)                  
-#>  remotes       2.1.0      2019-06-24 [1] CRAN (R 3.6.1)                  
-#>  rlang         0.4.1      2019-10-24 [1] CRAN (R 3.6.1)                  
-#>  rmarkdown     1.16       2019-10-01 [1] CRAN (R 3.6.1)                  
-#>  rprojroot     1.3-2      2018-01-03 [1] CRAN (R 3.5.0)                  
-#>  rstudioapi    0.10       2019-03-19 [1] CRAN (R 3.6.1)                  
-#>  rticles       0.9.1      2019-07-22 [1] Github (rstudio/rticles@8d56fc6)
-#>  rvest         0.3.5      2019-11-08 [1] CRAN (R 3.6.1)                  
-#>  scales        1.0.0      2018-08-09 [1] CRAN (R 3.6.1)                  
-#>  sessioninfo   1.1.1      2018-11-05 [1] CRAN (R 3.6.1)                  
-#>  stringi       1.4.3      2019-03-12 [1] CRAN (R 3.6.1)                  
-#>  stringr     * 1.4.0      2019-02-10 [1] CRAN (R 3.6.1)                  
-#>  survival    * 2.44-1.1   2019-04-01 [4] CRAN (R 3.6.1)                  
-#>  testthat      2.2.0      2019-07-22 [1] CRAN (R 3.6.1)                  
-#>  tibble      * 2.1.3      2019-06-06 [1] CRAN (R 3.6.1)                  
-#>  tidyr       * 1.0.0      2019-09-11 [1] CRAN (R 3.6.1)                  
-#>  tidyselect    0.2.5      2018-10-11 [1] CRAN (R 3.5.2)                  
-#>  tidyverse   * 1.2.1      2017-11-14 [1] CRAN (R 3.5.0)                  
-#>  usethis       1.5.1      2019-07-04 [1] CRAN (R 3.6.1)                  
-#>  vctrs         0.2.0      2019-07-05 [1] CRAN (R 3.6.1)                  
-#>  VGAM          1.1-1      2019-02-18 [1] CRAN (R 3.6.1)                  
-#>  withr         2.1.2      2018-03-15 [1] CRAN (R 3.5.0)                  
-#>  xfun          0.8        2019-06-25 [1] CRAN (R 3.6.1)                  
-#>  xml2          1.2.2      2019-08-09 [1] CRAN (R 3.6.1)                  
-#>  yaml          2.2.0      2018-07-25 [1] CRAN (R 3.6.1)                  
-#>  zeallot       0.1.0      2018-01-28 [1] CRAN (R 3.5.1)                  
+#> ─ Packages ──────────────────────────────────────────────────────────────
+#>  package     * version    date       lib
+#>  assertthat    0.2.1      2019-03-21 [1]
+#>  backports     1.1.5      2019-10-02 [1]
+#>  broom         0.5.2      2019-04-07 [1]
+#>  callr         3.3.1      2019-07-18 [1]
+#>  casebase    * 0.2.1.9001 2019-12-11 [1]
+#>  cellranger    1.1.0      2016-07-27 [1]
+#>  cli           2.0.0      2019-12-09 [1]
+#>  codetools     0.2-16     2018-12-24 [4]
+#>  colorspace    1.4-1      2019-03-18 [1]
+#>  crayon        1.3.4      2017-09-16 [1]
+#>  data.table    1.12.8     2019-12-09 [1]
+#>  desc          1.2.0      2018-05-01 [1]
+#>  devtools      2.1.0      2019-07-06 [1]
+#>  digest        0.6.23     2019-11-23 [1]
+#>  dplyr       * 0.8.3      2019-07-04 [1]
+#>  ellipsis      0.3.0      2019-09-20 [1]
+#>  evaluate      0.14       2019-05-28 [1]
+#>  fansi         0.4.0      2018-10-05 [1]
+#>  farver        2.0.1      2019-11-13 [1]
+#>  forcats     * 0.4.0      2019-02-17 [1]
+#>  foreach     * 1.4.4      2017-12-12 [1]
+#>  fs            1.3.1      2019-05-06 [1]
+#>  generics      0.0.2      2018-11-29 [1]
+#>  ggplot2     * 3.2.1      2019-08-10 [1]
+#>  glmnet      * 2.0-18     2019-05-20 [1]
+#>  glue          1.3.1      2019-03-12 [1]
+#>  gtable        0.3.0      2019-03-25 [1]
+#>  haven         2.1.1      2019-07-04 [1]
+#>  hms           0.5.2      2019-10-30 [1]
+#>  htmltools     0.3.6      2017-04-28 [1]
+#>  httr          1.4.1      2019-08-05 [1]
+#>  iterators     1.0.9      2017-12-12 [1]
+#>  jsonlite      1.6        2018-12-07 [1]
+#>  knitr         1.23       2019-05-18 [1]
+#>  labeling      0.3        2014-08-23 [1]
+#>  lattice       0.20-38    2018-11-04 [4]
+#>  lazyeval      0.2.2      2019-03-15 [1]
+#>  lifecycle     0.1.0      2019-08-01 [1]
+#>  lubridate   * 1.7.4      2018-04-11 [1]
+#>  magrittr    * 1.5        2014-11-22 [1]
+#>  Matrix      * 1.2-18     2019-11-27 [4]
+#>  memoise       1.1.0      2017-04-21 [1]
+#>  mgcv          1.8-24     2018-06-18 [1]
+#>  modelr        0.1.5      2019-08-08 [1]
+#>  munsell       0.5.0      2018-06-12 [1]
+#>  nlme          3.1-142    2019-11-07 [4]
+#>  pillar        1.4.2      2019-06-29 [1]
+#>  pkgbuild      1.0.3      2019-03-20 [1]
+#>  pkgconfig     2.0.3      2019-09-22 [1]
+#>  pkgload       1.0.2      2018-10-29 [1]
+#>  prettyunits   1.0.2      2015-07-13 [1]
+#>  processx      3.4.1      2019-07-18 [1]
+#>  ps            1.3.0      2018-12-21 [1]
+#>  purrr       * 0.3.3      2019-10-18 [1]
+#>  R6            2.4.1      2019-11-12 [1]
+#>  Rcpp          1.0.3      2019-11-08 [1]
+#>  readr       * 1.3.1      2018-12-21 [1]
+#>  readxl        1.3.1      2019-03-13 [1]
+#>  remotes       2.1.0      2019-06-24 [1]
+#>  rlang         0.4.2      2019-11-23 [1]
+#>  rmarkdown     1.16       2019-10-01 [1]
+#>  rprojroot     1.3-2      2018-01-03 [1]
+#>  rstudioapi    0.10       2019-03-19 [1]
+#>  rticles       0.9.1      2019-07-22 [1]
+#>  rvest         0.3.5      2019-11-08 [1]
+#>  scales        1.1.0      2019-11-18 [1]
+#>  sessioninfo   1.1.1      2018-11-05 [1]
+#>  stringi       1.4.3      2019-03-12 [1]
+#>  stringr     * 1.4.0      2019-02-10 [1]
+#>  survival    * 2.44-1.1   2019-04-01 [4]
+#>  testthat      2.2.0      2019-07-22 [1]
+#>  tibble      * 2.1.3      2019-06-06 [1]
+#>  tidyr       * 1.0.0      2019-09-11 [1]
+#>  tidyselect    0.2.5      2018-10-11 [1]
+#>  tidyverse   * 1.2.1      2017-11-14 [1]
+#>  usethis       1.5.1      2019-07-04 [1]
+#>  vctrs         0.2.0      2019-07-05 [1]
+#>  VGAM          1.1-2      2019-11-21 [1]
+#>  withr         2.1.2      2018-03-15 [1]
+#>  xfun          0.8        2019-06-25 [1]
+#>  xml2          1.2.2      2019-08-09 [1]
+#>  yaml          2.2.0      2018-07-25 [1]
+#>  zeallot       0.1.0      2018-01-28 [1]
+#>  source                                  
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  Github (sahirbhatnagar/casebase@e2e73fe)
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.2)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.2)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.2)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  Github (rstudio/rticles@8d56fc6)        
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.2)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.1)                          
 #> 
@@
 \begin{CodeOutput}
-#> Local:    review /home/mturgeon/Documents/git_repositories/cbpaper
-#> Head:     [4435e71] 2019-12-08: Merge branch 'review-max'
+#> Local:    review-glmnet /home/mturgeon/Documents/git_repositories/cbpaper
+#> Head:     [b7e23a6] 2019-12-10: Merge branch 'master' into review-glmnet
 \end{CodeOutput}

### 2019-12-11 analysis/paper/paper.Rmd
Bib+Why 60->14+packageCapitalization+cs3
fixed some bib issues, put a short description of why we went from 60 to 14, fixed package name capitalization issues and made case study 3 work with the attached support data.

--- 2019-12-08 analysis/paper/paper.Rmd
@@
   - name: Maxime Turgeon *
-    affiliation: 'University of Manitoba \AND'
+    affiliation: University of Manitoba
     address: >
-      186 Dysart Road
+      318 Machray Hall
       Winnipeg, MB, Canada R3T 2N2
     email: \email{max.turgeon@umanitoba.ca}
-    url: https://maxturgeon.ca/
+    url: http://maxturgeon.ca/
   - name: Jesse Islam 
@@
   \usepackage{graphicx}
+  \DeclareUnicodeCharacter{2500}{}
+  \usepackage{tabularx}
 output: 
@@
 
-<!-- # Code formatting -->
-
-<!-- Don't use markdown, instead use the more precise latex commands: -->
-
-<!-- * \proglang{Java} -->
-<!-- * \pkg{plyr} -->
-<!-- * \code{print("abc")} -->
-
+# Code formatting
+
+Don't use markdown, instead use the more precise latex commands:
+
+* \proglang{Java}
+* \pkg{plyr}
+* \code{print("abc")}
 ```{r setup, include=FALSE}
@@
 library(splines)
-#library(readr)
-#library(tibble)
+library(tibble)
 library(glmnet)
@@
 # set the seed for reproducible output
-set.seed(2)
+set.seed(1234)
 
 eval_introduction <- TRUE
-eval_theory <- TRUE
-eval_implementation <- TRUE
+eval_theory <- FALSE
+eval_implementation <- FALSE
 eval_cs1 <- FALSE
@@
 eval_cs3 <- FALSE
-eval_cs4 <- TRUE
 ```
@@
 
-The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence (CI)) of an event, for a particular patient.
+The purpose of the `casebase` package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence (CI)) of an event, for a particular patient.
 The following points should be noted:
 
-  1. hazard ratios
+hazard ratios
 2. If, however, the absolute risks are of interest, they have to be recovered using the semi-parametric Breslow estimator
-  3. Alternative approaches for fitting flexible hazard models for estimating absolute risks, not requiring this two-step approach? Yes! [@hanley2009fitting]
+3. Alternative approaches for fitting flexible hazard models for estimating absolute risks, not requiring this two-step approach? Yes!
+[@hanley2009fitting]
 
@@
 From here, we manually examined each package to determine which ones are comparable to casebase, with either a different approach to the same problems or future features we would like to add to casebase.
+In particular, we excluded packages that were focused on a different set of problems, such as frailty and multi-state models, as casebase is currently focused on absolute-risk, competing events and single event analysis.
 
@@
 Besides these three main categories, there are extra features of interest as well.
-Table \ref{tab:surv-pkgs} summarizes the findings.
-
-Many packages exist that implement semi-parametric Cox models.
-Of those, \pkg{glmnet} [-@regpathcox]???, \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche] introduce penalization in their model fitting.
-\pkg{CoxRidge} [-@perperoglou], \pkg{rstpm2} [-@clements_liu] and \pkg{survival} [-@survival-package] include penalization, but they implement the extended cox model permitting intuitive non proportional hazard modeling while the previous four do not.
-In terms of penalization, the benefits of \pkg{casebase} comes from its flexibility.
-Linear models have a wide tool set, considerably larger than those made for Cox regression.
-<!--(*reference for number of packages between their task views, regression is spread out across Multivariate, SocialSciences, Robust among others.
-The focus of this task view is on Optimization Infrastructure Packages , General Purpose Continuous Solvers , Mathematical Programming Solvers , and Specific Applications in Optimization , or Multi Objective Optimization*)--> \pkg{casebase} permits many regularization tools developed for linear regression to be repurposed within a survival context.
+Table XXX summarizes the findings.
+
+Many packages exist that implement semi-parametric cox models.
+Of those, glmnet [-@regpathcox]???, glmpath [-@park_hastie], penalized [-@l1penal], RiskRegression [-@gerds_blanche] introduce penalization in their model fitting.
+CoxRidge [-@perperoglou] , rstpm2 [-@clements_liu] and survival [-@survival-package] include penalization, but they implement the extended cox model permitting intuitive non proportional hazard modeling while the previous four do not.
+In terms of penalization, the benefit of casebase comes from its flexibility.
+Linear models have a wide tool set, considerably larger than those made for cox regression (*reference for number of packages between their task views, regression is spread out across Multivariate, SocialSciences, Robust among others.
+The focus of this task view is on Optimization Infrastructure Packages , General Purpose Continuous Solvers , Mathematical Programming Solvers , and Specific Applications in Optimization , or Multi Objective Optimization*).
+Casebase permits many regularization tools developed for linear regression to be repurposed within a survival context.
 
 Parametric models permit non-proportional hazards and are implemented in a handful of packages.
-Of interest are \pkg{CFC} [-@sharabiani_mahani_2019], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu] and \pkg{survival} [-@survival-package].
-\pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas \pkg{flexsurv} and \pkg{survival} contain implementations for users to supply any distribution they would like [-@survival-package], [-@flexsurv].
-\pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} can use splines, permitting a flexible fit over time [-@flexsurv], [-@smoothHazard], [-@scheike2014estimating], [-@clements_liu].
-Though \pkg{casebase} is limited to three distributions at the moment, it also uses splines allowing for distributions that may not be consistent over survival time.
-
-Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets} and \pkg{survival} contain implementations for competing risks [-@sharabiani_mahani_2019], [-@flexsurv], [-@scheike2014estimating], [-@survival-package].
-The differentiating factor between these packages and \pkg{casebase} is that \pkg{casebase} handles competing risks through multiple logistic regression.
-\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval censoring, whereas \pkg{casebase} currently cannot.
-Interval censoring is a simple feature to implement and is a potential improvement to add to \pkg{casebase}.
-
-Of all the packages listed in Table \ref{tab:surv-pkgs}, \pkg{CFC}, \pkg{Flexsurv}, \pkg{mets}, \pkg{RiskRegression}, \pkg{rstpm2} and \pkg{survival} have implementations for cumulative incidence [-@sharabiani_mahani_2019], [-@flexsurv], [-@scheike2014estimating], [-@gerds_blanche], [-@clements_liu], [-@survival-package].
-Since \pkg{casebase} is parametric, the baseline hazard is known and requires integration to retrieve the cumulative incidence.
-This is similar to the approaches of \pkg{CFC} and \pkg{flexsurv} [-@sharabiani_mahani_2019], [-@flexsurv], while the other packages use semi-parametric approaches to retrieve the baseline after fitting [-@scheike2014estimating], [-@gerds_blanche].
-\pkg{rstpm2} and \pkg{survival} use both for their respective parametric and semi-parametric model implementations [-@clements_liu], [-@survival-package].
-
-\begin{table}[ht]
+Of interest are CFC [-@sharabiani_mahani_2019], flexsurv [-@flexsurv] , SmoothHazard [-@smoothHazard], Rsptm2 [-@clements_liu] and Survival [-@survival-package].
+Smooth hazard  is limited to weibull distributions[-@smoothHazard], where as flexsurv and survival contain implementations for users to supply any distribution they would like [-@survival-package], [-@flexsurv].
+Flexsurv, smooth hazard, mets and rstpm2 can use splines, permitting a flexible fit over time [-@flexsurv], [-@smoothHazard], [-@scheike2014estimating], [-@clements_liu] .
+Though casebase is limited to 3 distributions at the moment, it also uses splines allowing for distributions that may not be consistent over survival time.
+
+Of the methods mentioned so far, only CFC,flexsurv, mets and survival contain implementations for competing risks [-@sharabiani_mahani_2019], [-@flexsurv],[-@scheike2014estimating] ,[-@survival-package] (*how is it done for parametric models?*).
+The differentiating factor between these packages and casebase is that casebase handles competing risks through multiple logistic regression.
+Smooth hazard, survival and Rstpm2 can handle interval censoring, where as casebase currently can not.
+Interval censoring is simple feature to implement and is a potential improvement to add to casebase.
+(There's also multistate models, which I haven't mentioned in this paper)
+
+Of all the packages listed in the table, CFC , Flexsurv, mets, RiskRegression, rstpm2 and survival have implementations for cumulative incidence [-@sharabiani_mahani_2019], [-@flexsurv], [-@scheike2014estimating], [-@gerds_blanche], [-@clements_liu], [-@survival-package].
+Since casebase is parametric, the baseline hazard is known and requires integration to retrieve the cumulative incidence.
+This is similar to the approaches of CFC and flexsurv [-@sharabiani_mahani_2019], [-@flexsurv], while the other packages use semi-parametric approaches to retrieve the baseline after fitting [-@scheike2014estimating], [-@gerds_blanche].
+rstpm2 and survival use both for their respective parametric and semi-parametric model implementations [-@clements_liu], [-@survival-package].
+
+\begin{table}[]
 \centering
@@
 \hline
-\textbf{Package}        & \textbf{Competing Risks} & \textbf{Non-proportional} & \textbf{Penalization} & \textbf{Splines} & \textbf{Parametric} & \textbf{Semi-parametric} & \textbf{Interval/left Censoring} & \textbf{Absolute Risk}           \\
+\textbf{Package}        & \vtop{\hbox{\strut \textbf{Competing}}\hbox{\strut \textbf{risks}}}  & \textbf{Non-proportional} & \textbf{Penalization} & \textbf{Splines} & \textbf{Parametric} & \textbf{Semi-parametric} & \vtop{\hbox{\strut \textbf{Interval/left}}\hbox{\strut \textbf{censoring}}} & \vtop{\hbox{\strut \textbf{Absolute}}\hbox{\strut \textbf{risk}}}           \\
 \hline
@@
 \textbf{fastcox} \cite{yi_zou}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{Flexrsurv} \cite{flexrsurv}      &                          & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
-\\
-\textbf{Flexsurv} \cite{flexsurv}       & x                        & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?
-\\
+\textbf{Flexrsurv} \cite{flexrsurv}      &                          & x                         &                       & x                & x                   &                          &                                  & x               \\
+\textbf{Flexsurv} \cite{flexsurv}       & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\
 \textbf{glmnet} \cite{regpathcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
@@
 \textbf{RiskRegression} \cite{gerds_blanche} &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
-\textbf{Rstpm2} \cite{clements_liu}         &                          & x                         & x                     & x                & x                   & x                        & x                                & Cumm haz                         \\
+\textbf{Rstpm2} \cite{clements_liu}         &                          & x                         & x                     & x                & x                   & x                        & x                                & x                         \\
 \textbf{SmoothHazard} \cite{smoothHazard}   &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
@@
 \caption{Different features of interest in various survival packages.}
-\label{tab:surv-pkgs}
+\label{tab:my-table}
 \end{table}
@@
 
-<!-- Popular survival analysis methods, like Kaplan-Meier and Cox regression, rely on the notion of risk-set sampling. Case-base sampling can be seen as an alternative. -->
-
-As shown by Saarela & Arjas [-@saarela2015non] (and further expanded in Saarela [-@saarela2016case]), writing the likelihood arising from this data-generating mechanism using the framework of non-homogeneous Poisson processes, we eventually reach an expression where each person-moment's contribution is of the form
-$$\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},$$
-where $N(t)$ is the counting process associated with the event of interest, $h(t)$ is the corresponding hazard function, and $\rho(t)$ is the hazard function for the Poisson process associated with case-base sampling.
-This parametric form suggests that we can readily estimate log-hazards of the form $\log(h(t)) = g(t; X)$ using logistic regression, where each observation corresponds to a person moment, the function $g(t; X)$ is linear in a finite number of parameters, and where we treat $-\log(\rho(t))$ as an offset.
-
-In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base sampling *uniformly*, i.e.
+Popular survival analysis methods, like Kaplan-Meier and Cox regression, rely on the notion of risk-set sampling. Case-base sampling can be seen as an alternative.
+
+As shown by Saarela & Arjas [-@saarela2015non] (and further expanded in Saarela [-@saarela2016case]), writing the likelihood arising from this data-generating mechanism using the framework of non-homogenous Poisson processes, we eventually reach an expression where each person-moment's contribution is of the form
+$$\frac{\lambda(t)^{dN(t)}}{\rho(t) + \lambda(t)},$$
+where $N(t)$ is the counting process associated with the event of interest, $\lambda(t)$ is the corresponding hazard function, and $\rho(t)$ is the hazard function for the Poisson process associated with case-base sampling.
+This parametric form suggests that we can readily estimate log-hazards of the form $\log(\lambda(t)) = g(t; X)$ using logistic regression, where each observation corresponds to a person moment, the function $g(t; X)$ is linear in a finite number of parameters, and where we treat $\log(\rho(t))$ as an offset.
+
+In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base samping *uniformly*, i.e.
 to sample the base series uniformly from the study base.
-In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base.
+In terms of Poisson processes, this sampling strategy corresponds to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base.
 More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism  are proportional to the cardiovascular disease event rate given by the Framingham score.
@@
 
-$$ h(t;X) = \exp[g(t;X)] $$
-where $t$ denotes time and $X$, the individual's covariate profile.
+$$ h(x,t) = exp[g(x,t)] $$
+where $t$ denotes the numerical value (number of units) of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any).
 Different functions of $t$ lead to different parametric hazard models.
+
 The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$.
 
-$$ h(t;X) = \exp(\beta_0 + \beta_1 X) $$
-
-The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
-This is also known as the *memoryless property* (Kalbfleisch and Prentice, 2002).
+$$ h(x,t) = exp(\beta_0 + \beta_1 x) $$
+
+The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been on study a.k.a the memoryless property (Kalbfleisch and Prentice, 2002).
 
@@
 
-$$ h(t;X)  = \exp(\beta_0 + \beta_1 t + \beta_2 X) $$
-
-Use of $\log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
-
-$$ h(t;X)  = \exp(\beta_0 + \beta_1 \log(t) + \beta_2 X) $$
-
-For competing-risk analyses with $J$ possible events, we can show that each person-moment's contribution of the likelihood is of the form
-
-$$\frac{h_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^Jh_j(t)},$$
-
-where $N_j(t)$ is the counting process associated with the event of type $j$ and $h_j(t)$ is the corresponding hazard function.
-As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1\end{align*}}
+$$ h(x,t)  = exp(\beta_0 + \beta_1 t + \beta_2 x) $$
+
+Use of $log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
+
+$$ h(x, t)  = exp(\beta_0 + \beta_1 \log(t) + \beta_2 x) $$
 
@@
 
-The functions in the casebase package can be divided into two categories: 1) data visualization, in the form of population-time plots; and 2) parametric modeling.
-We explicitly aimed at being compatible with both \code{data.frame}s and \code{data.table}s.
-This is evident in some of the coding choices we made, and it is also reflected in our unit tests.
+The functions in the casebase package can be divided into two categories: 1) data visualization, in the form of population-time plots; and 2) data analysis.
+
+We explicitely aimed at being compatible with both \code{data.frame}s and \code{data.table}s. This is evident in some of the coding choices we made, and it is also reflected in our testing units.
 
@@
 
-## Parametric modeling
-
-The parametric modeling step was separated into three parts:
-
-  1. case-base sampling; 
-  2. estimation of the smooth hazard function; 
-  3. calculation of the risk function.
-
+## Data analysis
+
+The data analysis step was separated into three parts: 1) case-base sampling; 2) estimation of the smooth hazard function; and 3) calculation of the risk function.
 By separating the sampling and estimation functions, we allowed the possibility of users implementing more complex sampling scheme, as described in Saarela [-@saarela2016case].
@@
 $$S(t) = \exp\left(-\int_0^t h(u;X) du\right).$$
-The integral is computed using either the \code{stats::integrate} function or Monte-Carlo integration. The risk function (or cumulative incidence function) is then defined as
-
-$$ CI(t) = 1 - S(t).$$
-
-For the case of a competing-event analysis, the event-specific risk is computed using the following procedure: first, we compute the overall survival function (i.e. for all event types):
-
-$$ S(t) = \exp\left(-\int_0^t H(u;X) du\right),\qquad H(t;X) = \sum_{j=1}^J h_j(t;X).$$
-From this, we can derive the event-specific subdensities:
-
-$$ f_j(t) = h_j(t)S(t).$$
-
-Finally, by integrating these subdensities, we obtain the event-specific cumulative incidence functions:
-
-$$ CI_j(t) = \int_0^t f_j(u)du.$$
-
-We created \code{absoluteRisk} as an \code{S3} generic, with methods for the different types of outputs of \code{fitSmoothHazard}.
-The method dispatch system of \proglang{R} then takes care of matching the correct output to the correct methodology for calculating the cumulative incidence function, without the user's intervention.
-
-In the following sections, we illustrate these functionalities in the context of three case studies.
+The integral is computed using either the \code{stats::integrate} function or Monte-Carlo integration.
+For the case of a competing-event analysis, the event-specific risk is computed using a nested double integral; in this setting, Monte-Carlo integration is faster and more flexible than \code{stats::integrate}.
+This is due to the fact that the computation involves a double loop over the selected time points; Monte-Carlo integration performs this step by using the vectorized \code{rowSums} and \code{colSums} functions.
+
+To decide between a single-event and a competing-event analysis, we created \code{absoluteRisk} as an \code{S3} generic, with methods for both \code{glm} and \code{CompRisk} objects (the latter inherits from \code{vglm} as well).
+The method dispatch system of \proglang{R} then takes care of matching the correct input to the correct methodology, without the user's intervention.
 
@@
 The SUPPORT dataset will be used to demonstrate regularization within casebase (REF), while comparing their absolute risk preditions for a new individual.
-A description of each variable can be found in Table~\ref{tab:support1} and a breakdown of each categorical variable in Table~\ref{tab:support2}.
+A description of each variable can be found in table XXX1 and a breakdown of each categorical variable in table XXX2.
 The data was downloaded from the main author's website, and it is also available as part of the casebase package.
 
-\begin{table}[ht]
+\begin{table}[]
 \centering
@@
 \caption{A description of each variable in the SUPPORT dataset, taken from (REF).}
-\label{tab:support1}
+\label{tab:my-table}
 \end{table}
 
-\begin{table}[ht]
+\begin{table}[]
 \centering
@@
 \caption{A description of each level within each categorical variable in the dataset, taken from REF.}
-\label{tab:support2}
+\label{tab:my-table}
 \end{table}
@@
 The glmnet package(REF) on CRAN was used to introduce elastic net, lasso and ridge regressions to casebase.
-To visualize the effect of imputation on incidence density, we can refer to the pre-imputation population-time plot in Figure \ref{fig:ptPreI} and post-imputation population time plot in Figure \ref{fig:ptPostI}.
-The incidence density is consistent between imputed and complete datasets.
+To visualize the incidence density, we can refer to the post-imputation population-time plot in Figure \ref{fig:ptPostI}.
 
@@
 completeData=casebase::support
+completeData<-completeData[,-c(3)]
 #keep one individual out, at random, for which we will develop an 
@@
 
-```{r poptimePreImputation, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:ptPreI} Population-time plot before imputation.
-Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study."}
-#demonstrating incidence density before imputation
-plot(casebase::popTime(as.data.frame(support),time = "d.time",event = "death"))
-```
-
 ```{r poptimePostImputation, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:ptPostI} Population-time plot before imputation.
@@
 #demonstrating incidence density after imputation
-plot(casebase::popTime(as.data.frame(workingCompleteData),time = "d.time",event = "death"))
+plot(casebase::popTime(as.data.frame(completeData),time = "d.time",event = "death"))
 ```
@@
 #Estimating the absolute risk curve using the newData parameter.
-casebaseAbsolute=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) ,
-                              s="lambda.1se",method=c("numerical"))
+casebaseAbsolute=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) , s="lambda.1se",method=c("numerical"))
 ```
@@
 ```
-
-# Case study 4--Stanford Heart Transplant Data
-
-Although the previous case studies provide a broad overview of the capabilities of \pkg{casebase}, they all have two properties in common:
-\begin{itemize}
-  \item All covariates are fixed, i.e. they do not change over time;
-  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
-\end{itemize}
-
-Case-base sampling has been used in the literature to study vaccination safety, where the exposure period was defined as the week following vaccination [@saarela2015case].
-Hence, the main covariate of interest, i.e.
-exposure to the vaccine, was changing over time.
-In this context, case-base sampling offers an efficient alternative to nested case-control designs or self-matching.
-
-The purpose of the next case study is to show how \pkg{casebase} can also be used for survival analysis problems that do not share the properties above.
-To this end, we will use the Stanford Heart Transplant Data [@clark1971cardiac,@crowley1977covariance]; we will use the version available in the \pkg{survival} package.
-
-Recall the setting of this study: patients were admitted to the Stanford program after meeting with their physician and determining that they were unlikely to respond to other forms of treatment.
-After enrollment, the program searches for a suitable donor for the patient, which can take anywhere between a few days to almost a year.
-We are interested in the effect of a heart transplant on survival; therefore, the patient is considered exposed only after the transplant has occured.
-
-As before, we can look at the population-time plot for a graphical summary of the event incidence.
-As we can see, most events occur early during the follow-up period, and therefore we do not expect the hazard to be constant.
-
-```{r stanford-poptime, eval = eval_cs4}
-library(survival)
-library(casebase)
-
-stanford_popTime <- popTime(jasa, time = "futime", 
-                            event = "fustat")
-plot(stanford_popTime)
-```
-
-Since the exposure is time-dependent, we need to manually define the exposure variable *after* case-base sampling and *before* fitting the hazard function.
-For this reason, we will use the `sampleCaseBase` function directly.
-
-```{r, eval = eval_cs4}
-library(tidyverse)
-library(lubridate)
-
-cb_data <- sampleCaseBase(jasa, time = "futime", 
-                          event = "fustat", ratio = 10)
-```
-
-Next, we will compute the number of days from acceptance into the program to transplant, and we use this variable to determine whether each population-moment is exposed or not.
-
-```{r, eval = eval_cs4}
-# Define exposure variable
-cb_data <- mutate(cb_data,
-                  txtime = time_length(accept.dt %--% tx.date, 
-                                       unit = "days"),
-                  exposure = case_when(
-                    is.na(txtime) ~ 0L,
-                    txtime > futime ~ 0L,
-                    txtime <= futime ~ 1L
-                  ))
-```
-
-Finally, we can fit the hazard using various linear predictors.
-
-```{r, eval = eval_cs4, warning = FALSE}
-library(splines)
-# Fit several models
-fit1 <- fitSmoothHazard(fustat ~ exposure,
-                        data = cb_data, time = "futime")
-fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
-                        data = cb_data, time = "futime")
-fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
-                        data = cb_data, time = "futime")
-fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
-                        data = cb_data, time = "futime")
-```
-
-Note that the fourth model (i.e.
-`fit4`) includes an interaction term between exposure and follow-up time.
-In other words, this model no longer exhibit proportional hazards.
-The evidence of non-proportionality of hazards in the Stanford Heart Transplant data has been widely discussed [@arjas1988graphical].
-
-We can then compare the goodness of fit of these four models using the Akaike Information Criterion (AIC).
-
-```{r, eval = eval_cs4}
-# Compute AIC
-c("Model1" = AIC(fit1),
-  "Model2" = AIC(fit2),
-  "Model3" = AIC(fit3),
-  "Model4" = AIC(fit4))
-```
-
-As we can, the best fit is the fourth model. By visualizing the hazard functions for both exposed and unexposed individuals, we can more clearly see how the hazards are no longer proportional.
-
-```{r stanford-hazard, eval = eval_cs4}
-# Compute hazards---
-# First, create a list of time points for both exposure status
-hazard_data <- expand.grid(exposure = c(0, 1),
-                           futime = seq(0, 1000,
-                                        length.out = 100))
-# Set the offset to zero
-hazard_data$offset <- 0 
-# Use predict to get the fitted values, and exponentiate to 
-# transform to the right scale
-hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
-                                 type = "link"))
-# Add labels for plots
-hazard_data$Status = factor(hazard_data$exposure,
-                            labels = c("NoTrans", "Trans"))
-
-ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
-    geom_line() +
-    theme_minimal() +
-    theme(legend.position = 'top') +
-    ylab('Hazard') + xlab('Follow-up time')
-```
-
-The non-proportionality seems to be more pronounced at the beginning of follow-up than the end.
-Finally, we can turn these estimate of the hazard function into estimates of the cumulative incidence functions.
-
-```{r stanford-risk, eval = eval_cs4}
-# Compute absolute risk curves
-newdata <- data.frame(exposure = c(0, 1))
-absrisk <- absoluteRisk(fit4, newdata = newdata, 
-                        time = seq(0, 1000, length.out = 100))
-
-colnames(absrisk) <- c("Time", "NoTrans", "Trans")
-
-# Rearrange the data
-absrisk <- gather(as.data.frame(absrisk),
-                  "Status", "Risk", -Time)
-
-ggplot(absrisk, aes(Time, Risk, colour = Status)) +
-  geom_line() +
-  theme_minimal() +
-  theme(legend.position = 'top') +
-  expand_limits(y = 0.5) +
-  xlab('Follow-up time') + ylab('Cum. Incidence')
-```
-
-Note that we can easily adapt the code above to the situation where a patient receives a heart transplant at a point in time of interest, for example after 30 days.
-
-```{r stanford-1y-haz, eval = eval_cs4}
-# Compute hazards---
-# First, create a list of time points for both exposure status
-one_yr_haz <- data.frame(futime = seq(0, 365,
-                                      length.out = 100))
-one_yr_haz <- mutate(one_yr_haz,
-                     offset = 0,
-                     exposure = if_else(futime < 30, 0, 1))
-one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
-                                type = "link"))
-ggplot(one_yr_haz, aes(futime, hazard)) +
-  geom_line() +
-  theme_minimal() +
-  theme(legend.position = 'top') +
-  ylab('Hazard') + xlab('Follow-up time') +
-  geom_vline(xintercept = 30, linetype = 'dashed')
-```
-
-We can then compare the 1-year mortality risk without transplant and with transplant at 30 days.
-
-```{r stanford-1y-risk, eval = eval_cs4}
-absoluteRisk(fit4, newdata = data.frame(exposure = 0),
-             time = 365)
-
-# Use the trapezoidal rule to estimate the cumulative hazard
-min_hazard <- min(one_yr_haz$hazard)
-max_hazard <- max(one_yr_haz$hazard)
-increment <- 365/99
-cumhaz_est <- 0.5*increment*(sum(2*one_yr_haz$hazard) - min_hazard - max_hazard)
-1 - exp(-cumhaz_est)
-```
-
-As we can see, the risk estimate at 1-year is about 30% lower if the patient receives a heart transplant at 30 days.
-
 # Discussion
-
-In this article, we presented the \proglang{R} package \pkg{casebase} that provides functions to fit smooth parametric hazards and estimate cumulative incidence functions using case-base sampling.
-We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through three case studies.
 

### 2019-12-11 analysis/paper/paper.tex
Bib+Why 60->14+packageCapitalization+cs3
fixed some bib issues, put a short description of why we went from 60 to 14, fixed package name capitalization issues and made case study 3 work with the attached support data.

--- 2019-12-08 analysis/paper/paper.tex
@@
   University of Manitoba\\
-  186 Dysart Road Winnipeg, MB, Canada R3T 2N2\\
+  318 Machray Hall Winnipeg, MB, Canada R3T 2N2\\
   E-mail: \email{max.turgeon@umanitoba.ca}\\
-  URL: \url{https://maxturgeon.ca/}\\~\\
+  URL: \url{http://maxturgeon.ca/}\\~\\
       Jesse Islam\\
@@
 
+\DeclareUnicodeCharacter{2500}{}
+
+\usepackage{tabularx}
+
 \begin{document}
+
+\hypertarget{code-formatting}{%
+\section{Code formatting}\label{code-formatting}}
+
+Don't use markdown, instead use the more precise latex commands:
+
+\begin{itemize}
+\item
+  \proglang{Java}
+\item
+  \pkg{plyr}
+\item
+  \code{print("abc")}
+\end{itemize}
 
@@
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  hazard ratios
-\item
-  If, however, the absolute risks are of interest, they have to be
-  recovered using the semi-parametric Breslow estimator
-\item
-  Alternative approaches for fitting flexible hazard models for
-  estimating absolute risks, not requiring this two-step approach? Yes!
+hazard ratios 2. If, however, the absolute risks are of interest, they
+have to be recovered using the semi-parametric Breslow estimator 3.
+Alternative approaches for fitting flexible hazard models for estimating
+absolute risks, not requiring this two-step approach? Yes!
 \citep{hanley2009fitting}
-\end{enumerate}
 
@@
 hazards, regularization and competing risks. Besides these three main
-categories, there are extra features of interest as well. Table
-\ref{tab:surv-pkgs} summarizes the findings.
-
-Many packages exist that implement semi-parametric Cox models. Of those,
-\pkg{glmnet} \citeyearpar{regpathcox}???, \pkg{glmpath}
-\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
-\pkg{RiskRegression} \citeyearpar{gerds_blanche} introduce penalization
-in their model fitting. \pkg{CoxRidge} \citeyearpar{perperoglou},
-\pkg{rstpm2} \citeyearpar{clements_liu} and \pkg{survival}
-\citeyearpar{survival-package} include penalization, but they implement
-the extended cox model permitting intuitive non proportional hazard
-modeling while the previous four do not. In terms of penalization, the
-benefits of \pkg{casebase} comes from its flexibility. Linear models
-have a wide tool set, considerably larger than those made for Cox
-regression. \pkg{casebase} permits many regularization tools developed
-for linear regression to be repurposed within a survival context.
+categories, there are extra features of interest as well. Table XXX
+summarizes the findings.
+
+Many packages exist that implement semi-parametric cox models. Of those,
+glmnet \citeyearpar{regpathcox}???, glmpath \citeyearpar{park_hastie},
+penalized \citeyearpar{l1penal}, RiskRegression
+\citeyearpar{gerds_blanche} introduce penalization in their model
+fitting. CoxRidge \citeyearpar{perperoglou} , rstpm2
+\citeyearpar{clements_liu} and survival \citeyearpar{survival-package}
+include penalization, but they implement the extended cox model
+permitting intuitive non proportional hazard modeling while the previous
+four do not. In terms of penalization, the benefit of casebase comes
+from its flexibility. Linear models have a wide tool set, considerably
+larger than those made for cox regression (\emph{reference for number of
+packages between their task views, regression is spread out across
+Multivariate, SocialSciences, Robust among others. The focus of this
+task view is on Optimization Infrastructure Packages , General Purpose
+Continuous Solvers , Mathematical Programming Solvers , and Specific
+Applications in Optimization , or Multi Objective Optimization}).
+Casebase permits many regularization tools developed for linear
+regression to be repurposed within a survival context.
 
 Parametric models permit non-proportional hazards and are implemented in
-a handful of packages. Of interest are \pkg{CFC}
-\citeyearpar{sharabiani_mahani_2019}, \pkg{flexsurv}
-\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
-\pkg{rsptm2} \citeyearpar{clements_liu} and \pkg{survival}
-\citeyearpar{survival-package}. \pkg{SmoothHazard} is limited to Weibull
-distributions \citeyearpar{smoothHazard}, whereas \pkg{flexsurv} and
-\pkg{survival} contain implementations for users to supply any
-distribution they would like \citeyearpar{survival-package},
-\citeyearpar{flexsurv}. \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets}
-and \pkg{rstpm2} can use splines, permitting a flexible fit over time
-\citeyearpar{flexsurv}, \citeyearpar{smoothHazard},
-\citeyearpar{scheike2014estimating}, \citeyearpar{clements_liu}. Though
-\pkg{casebase} is limited to three distributions at the moment, it also
-uses splines allowing for distributions that may not be consistent over
+a handful of packages. Of interest are CFC
+\citeyearpar{sharabiani_mahani_2019}, flexsurv \citeyearpar{flexsurv} ,
+SmoothHazard \citeyearpar{smoothHazard}, Rsptm2
+\citeyearpar{clements_liu} and Survival \citeyearpar{survival-package}.
+Smooth hazard is limited to weibull
+distributions\citeyearpar{smoothHazard}, where as flexsurv and survival
+contain implementations for users to supply any distribution they would
+like \citeyearpar{survival-package}, \citeyearpar{flexsurv}. Flexsurv,
+smooth hazard, mets and rstpm2 can use splines, permitting a flexible
+fit over time \citeyearpar{flexsurv}, \citeyearpar{smoothHazard},
+\citeyearpar{scheike2014estimating}, \citeyearpar{clements_liu} . Though
+casebase is limited to 3 distributions at the moment, it also uses
+splines allowing for distributions that may not be consistent over
 survival time.
 
-Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
-\pkg{mets} and \pkg{survival} contain implementations for competing
-risks \citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
-\citeyearpar{scheike2014estimating}, \citeyearpar{survival-package}. The
-differentiating factor between these packages and \pkg{casebase} is that
-\pkg{casebase} handles competing risks through multiple logistic
-regression. \pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can
-handle interval censoring, whereas \pkg{casebase} currently cannot.
-Interval censoring is a simple feature to implement and is a potential
-improvement to add to \pkg{casebase}.
-
-Of all the packages listed in Table \ref{tab:surv-pkgs}, \pkg{CFC},
-\pkg{Flexsurv}, \pkg{mets}, \pkg{RiskRegression}, \pkg{rstpm2} and
-\pkg{survival} have implementations for cumulative incidence
-\citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
+Of the methods mentioned so far, only CFC,flexsurv, mets and survival
+contain implementations for competing risks
+\citeyearpar{sharabiani_mahani_2019},
+\citeyearpar{flexsurv},\citeyearpar{scheike2014estimating}
+,\citeyearpar{survival-package} (\emph{how is it done for parametric
+models?}). The differentiating factor between these packages and
+casebase is that casebase handles competing risks through multiple
+logistic regression. Smooth hazard, survival and Rstpm2 can handle
+interval censoring, where as casebase currently can not. Interval
+censoring is simple feature to implement and is a potential improvement
+to add to casebase. (There's also multistate models, which I haven't
+mentioned in this paper)
+
+Of all the packages listed in the table, CFC , Flexsurv, mets,
+RiskRegression, rstpm2 and survival have implementations for cumulative
+incidence \citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
 \citeyearpar{scheike2014estimating}, \citeyearpar{gerds_blanche},
 \citeyearpar{clements_liu}, \citeyearpar{survival-package}. Since
-\pkg{casebase} is parametric, the baseline hazard is known and requires
+casebase is parametric, the baseline hazard is known and requires
 integration to retrieve the cumulative incidence. This is similar to the
-approaches of \pkg{CFC} and \pkg{flexsurv}
-\citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv}, while the
-other packages use semi-parametric approaches to retrieve the baseline
-after fitting \citeyearpar{scheike2014estimating},
-\citeyearpar{gerds_blanche}. \pkg{rstpm2} and \pkg{survival} use both
-for their respective parametric and semi-parametric model
-implementations \citeyearpar{clements_liu},
+approaches of CFC and flexsurv \citeyearpar{sharabiani_mahani_2019},
+\citeyearpar{flexsurv}, while the other packages use semi-parametric
+approaches to retrieve the baseline after fitting
+\citeyearpar{scheike2014estimating}, \citeyearpar{gerds_blanche}. rstpm2
+and survival use both for their respective parametric and
+semi-parametric model implementations \citeyearpar{clements_liu},
 \citeyearpar{survival-package}.
 
-\begin{table}[ht]
+\begin{table}[]
 \centering
@@
 risk preditions for a new individual. A description of each variable can
-be found in Table\textasciitilde{}\ref{tab:support1} and a breakdown of
-each categorical variable in Table\textasciitilde{}\ref{tab:support2}.
-The data was downloaded from the main author's website, and it is also
-available as part of the casebase package.
-
-\begin{table}[ht]
+be found in table XXX1 and a breakdown of each categorical variable in
+table XXX2. The data was downloaded from the main author's website, and
+it is also available as part of the casebase package.
+
+\begin{table}[]
 \centering
@@
 introduce elastic net, lasso and ridge regressions to casebase. To
-visualize the effect of imputation on incidence density, we can refer to
-the pre-imputation population-time plot in Figure \ref{fig:ptPreI} and
-post-imputation population time plot in Figure \ref{fig:ptPostI}. The
-incidence density is consistent between imputed and complete datasets.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> #demonstrating incidence density before imputation
-R> plot(casebase::popTime(as.data.frame(support),time = "d.time",event = "death"))
-\end{CodeInput}
-\end{CodeChunk}
+visualize the incidence density, we can refer to the post-imputation
+population-time plot in Figure \ref{fig:ptPostI}.
 
@@
 R> #demonstrating incidence density after imputation
-R> plot(casebase::popTime(as.data.frame(workingCompleteData),time = "d.time",event = "death"))
+R> plot(casebase::popTime(as.data.frame(completeData),time = "d.time",event = "death"))
 \end{CodeInput}
@@
 R> #Estimating the absolute risk curve using the newData parameter.
-R> casebaseAbsolute=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) ,
-R>                               s="lambda.1se",method=c("numerical"))
+R> casebaseAbsolute=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) , s="lambda.1se",method=c("numerical"))
 \end{CodeInput}
@@
 
-\hypertarget{case-study-4stanford-heart-transplant-data}{%
-\section{Case study 4--Stanford Heart Transplant
-Data}\label{case-study-4stanford-heart-transplant-data}}
-
-Although the previous case studies provide a broad overview of the
-capabilities of \pkg{casebase}, they all have two properties in common:
-
-\begin{itemize}
-  \item All covariates are fixed, i.e. they do not change over time;
-  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
-\end{itemize}
-
-Case-base sampling has been used in the literature to study vaccination
-safety, where the exposure period was defined as the week following
-vaccination \citep{saarela2015case}. Hence, the main covariate of
-interest, i.e.~exposure to the vaccine, was changing over time. In this
-context, case-base sampling offers an efficient alternative to nested
-case-control designs or self-matching.
-
-The purpose of the next case study is to show how \pkg{casebase} can
-also be used for survival analysis problems that do not share the
-properties above. To this end, we will use the Stanford Heart Transplant
-Data \citep[\citet{crowley1977covariance}]{clark1971cardiac}; we will
-use the version available in the \pkg{survival} package.
-
-Recall the setting of this study: patients were admitted to the Stanford
-program after meeting with their physician and determining that they
-were unlikely to respond to other forms of treatment. After enrollment,
-the program searches for a suitable donor for the patient, which can
-take anywhere between a few days to almost a year. We are interested in
-the effect of a heart transplant on survival; therefore, the patient is
-considered exposed only after the transplant has occured.
-
-As before, we can look at the population-time plot for a graphical
-summary of the event incidence. As we can see, most events occur early
-during the follow-up period, and therefore we do not expect the hazard
-to be constant.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(survival)
-R> library(casebase)
-R> 
-R> stanford_popTime <- popTime(jasa, time = "futime", 
-R+                             event = "fustat")
-R> plot(stanford_popTime)
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/stanford-poptime-1} \end{center}
-
-\end{CodeChunk}
-
-Since the exposure is time-dependent, we need to manually define the
-exposure variable \emph{after} case-base sampling and \emph{before}
-fitting the hazard function. For this reason, we will use the
-\texttt{sampleCaseBase} function directly.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(tidyverse)
-R> library(lubridate)
-R> 
-R> cb_data <- sampleCaseBase(jasa, time = "futime", 
-R+                           event = "fustat", ratio = 10)
-\end{CodeInput}
-\end{CodeChunk}
-
-Next, we will compute the number of days from acceptance into the
-program to transplant, and we use this variable to determine whether
-each population-moment is exposed or not.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Define exposure variable
-R> cb_data <- mutate(cb_data,
-R+                   txtime = time_length(accept.dt %--% tx.date, 
-R+                                        unit = "days"),
-R+                   exposure = case_when(
-R+                     is.na(txtime) ~ 0L,
-R+                     txtime > futime ~ 0L,
-R+                     txtime <= futime ~ 1L
-R+                   ))
-\end{CodeInput}
-\end{CodeChunk}
-
-Finally, we can fit the hazard using various linear predictors.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(splines)
-R> # Fit several models
-R> fit1 <- fitSmoothHazard(fustat ~ exposure,
-R+                         data = cb_data, time = "futime")
-R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
-R+                         data = cb_data, time = "futime")
-R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
-R+                         data = cb_data, time = "futime")
-R> fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
-R+                         data = cb_data, time = "futime")
-\end{CodeInput}
-\end{CodeChunk}
-
-Note that the fourth model (i.e.~\texttt{fit4}) includes an interaction
-term between exposure and follow-up time. In other words, this model no
-longer exhibit proportional hazards. The evidence of non-proportionality
-of hazards in the Stanford Heart Transplant data has been widely
-discussed \citep{arjas1988graphical}.
-
-We can then compare the goodness of fit of these four models using the
-Akaike Information Criterion (AIC).
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute AIC
-R> c("Model1" = AIC(fit1),
-R+   "Model2" = AIC(fit2),
-R+   "Model3" = AIC(fit3),
-R+   "Model4" = AIC(fit4))
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> Model1 Model2 Model3 Model4 
-#>    798    759    737    708
-\end{CodeOutput}
-\end{CodeChunk}
-
-As we can, the best fit is the fourth model. By visualizing the hazard
-functions for both exposed and unexposed individuals, we can more
-clearly see how the hazards are no longer proportional.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute hazards---
-R> # First, create a list of time points for both exposure status
-R> hazard_data <- expand.grid(exposure = c(0, 1),
-R+                            futime = seq(0, 1000,
-R+                                         length.out = 100))
-R> # Set the offset to zero
-R> hazard_data$offset <- 0 
-R> # Use predict to get the fitted values, and exponentiate to 
-R> # transform to the right scale
-R> hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
-R+                                  type = "link"))
-R> # Add labels for plots
-R> hazard_data$Status = factor(hazard_data$exposure,
-R+                             labels = c("NoTrans", "Trans"))
-R> 
-R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
-R+     geom_line() +
-R+     theme_minimal() +
-R+     theme(legend.position = 'top') +
-R+     ylab('Hazard') + xlab('Follow-up time')
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/stanford-hazard-1} \end{center}
-
-\end{CodeChunk}
-
-The non-proportionality seems to be more pronounced at the beginning of
-follow-up than the end. Finally, we can turn these estimate of the
-hazard function into estimates of the cumulative incidence functions.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute absolute risk curves
-R> newdata <- data.frame(exposure = c(0, 1))
-R> absrisk <- absoluteRisk(fit4, newdata = newdata, 
-R+                         time = seq(0, 1000, length.out = 100))
-R> 
-R> colnames(absrisk) <- c("Time", "NoTrans", "Trans")
-R> 
-R> # Rearrange the data
-R> absrisk <- gather(as.data.frame(absrisk),
-R+                   "Status", "Risk", -Time)
-R>  
-R> ggplot(absrisk, aes(Time, Risk, colour = Status)) +
-R+   geom_line() +
-R+   theme_minimal() +
-R+   theme(legend.position = 'top') +
-R+   expand_limits(y = 0.5) +
-R+   xlab('Follow-up time') + ylab('Cum. Incidence')
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/stanford-risk-1} \end{center}
-
-\end{CodeChunk}
-
-Note that we can easily adapt the code above to the situation where a
-patient receives a heart transplant at a point in time of interest, for
-example after 30 days.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute hazards---
-R> # First, create a list of time points for both exposure status
-R> one_yr_haz <- data.frame(futime = seq(0, 365,
-R+                                       length.out = 100))
-R> one_yr_haz <- mutate(one_yr_haz,
-R+                      offset = 0,
-R+                      exposure = if_else(futime < 30, 0, 1))
-R> one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
-R+                                 type = "link"))
-R> ggplot(one_yr_haz, aes(futime, hazard)) +
-R+   geom_line() +
-R+   theme_minimal() +
-R+   theme(legend.position = 'top') +
-R+   ylab('Hazard') + xlab('Follow-up time') +
-R+   geom_vline(xintercept = 30, linetype = 'dashed')
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/stanford-1y-haz-1} \end{center}
-
-\end{CodeChunk}
-
-We can then compare the 1-year mortality risk without transplant and
-with transplant at 30 days.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> absoluteRisk(fit4, newdata = data.frame(exposure = 0),
-R+              time = 365)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#>         
-#> 365 0.32
-\end{CodeOutput}
-
-\begin{CodeInput}
-R> 
-R> # Use the trapezoidal rule to estimate the cumulative hazard
-R> min_hazard <- min(one_yr_haz$hazard)
-R> max_hazard <- max(one_yr_haz$hazard)
-R> increment <- 365/99
-R> cumhaz_est <- 0.5*increment*(sum(2*one_yr_haz$hazard) - min_hazard - max_hazard)
-R> 1 - exp(-cumhaz_est)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> [1] 0.24
-\end{CodeOutput}
-\end{CodeChunk}
-
-As we can see, the risk estimate at 1-year is about 30\% lower if the
-patient receives a heart transplant at 30 days.
-
 \hypertarget{discussion}{%
@@
 
-This report was generated on 2019-12-08 12:31:41 using the following
+This report was generated on 2019-12-11 12:28:29 using the following
 computational environment and dependencies:
@@
 \begin{CodeOutput}
-#> - Session info ----------------------------------------------------------
+#> ─ Session info ───────────────────────────────────────────────────────────────
 #>  setting  value                       
-#>  version  R version 3.6.1 (2019-07-05)
-#>  os       Ubuntu 18.04.3 LTS          
-#>  system   x86_64, linux-gnu           
+#>  version  R version 3.6.0 (2019-04-26)
+#>  os       macOS High Sierra 10.13.6   
+#>  system   x86_64, darwin15.6.0        
 #>  ui       X11                         
@@
 #>  ctype    en_CA.UTF-8                 
-#>  tz       America/Winnipeg            
-#>  date     2019-12-08                  
+#>  tz       America/Montreal            
+#>  date     2019-12-11                  
 #> 
-#> - Packages --------------------------------------------------------------
-#>  package     * version    date       lib source                          
-#>  assertthat    0.2.1      2019-03-21 [1] CRAN (R 3.6.1)                  
-#>  backports     1.1.5      2019-10-02 [1] CRAN (R 3.6.1)                  
-#>  broom         0.5.2      2019-04-07 [1] CRAN (R 3.6.1)                  
-#>  callr         3.3.1      2019-07-18 [1] CRAN (R 3.6.1)                  
-#>  casebase    * 0.2.1.9001 2019-07-29 [1] local                           
-#>  cellranger    1.1.0      2016-07-27 [1] CRAN (R 3.5.0)                  
-#>  cli           1.1.0      2019-03-19 [1] CRAN (R 3.6.1)                  
-#>  codetools     0.2-16     2018-12-24 [4] CRAN (R 3.5.2)                  
-#>  colorspace    1.4-1      2019-03-18 [1] CRAN (R 3.6.1)                  
-#>  crayon        1.3.4      2017-09-16 [1] CRAN (R 3.5.0)                  
-#>  data.table    1.12.2     2019-04-07 [1] CRAN (R 3.6.1)                  
-#>  desc          1.2.0      2018-05-01 [1] CRAN (R 3.5.0)                  
-#>  devtools      2.1.0      2019-07-06 [1] CRAN (R 3.6.1)                  
-#>  digest        0.6.22     2019-10-21 [1] CRAN (R 3.6.1)                  
-#>  dplyr       * 0.8.3      2019-07-04 [1] CRAN (R 3.6.1)                  
-#>  ellipsis      0.3.0      2019-09-20 [1] CRAN (R 3.6.1)                  
-#>  evaluate      0.14       2019-05-28 [1] CRAN (R 3.6.1)                  
-#>  forcats     * 0.4.0      2019-02-17 [1] CRAN (R 3.6.1)                  
-#>  foreach     * 1.4.4      2017-12-12 [1] CRAN (R 3.5.0)                  
-#>  fs            1.3.1      2019-05-06 [1] CRAN (R 3.6.1)                  
-#>  generics      0.0.2      2018-11-29 [1] CRAN (R 3.6.1)                  
-#>  ggplot2     * 3.2.1      2019-08-10 [1] CRAN (R 3.6.1)                  
-#>  glmnet      * 2.0-18     2019-05-20 [1] CRAN (R 3.6.1)                  
-#>  glue          1.3.1      2019-03-12 [1] CRAN (R 3.6.1)                  
-#>  gtable        0.3.0      2019-03-25 [1] CRAN (R 3.6.1)                  
-#>  haven         2.1.1      2019-07-04 [1] CRAN (R 3.6.1)                  
-#>  hms           0.5.2      2019-10-30 [1] CRAN (R 3.6.1)                  
-#>  htmltools     0.3.6      2017-04-28 [1] CRAN (R 3.5.0)                  
-#>  httr          1.4.1      2019-08-05 [1] CRAN (R 3.6.1)                  
-#>  iterators     1.0.9      2017-12-12 [1] CRAN (R 3.5.0)                  
-#>  jsonlite      1.6        2018-12-07 [1] CRAN (R 3.6.1)                  
-#>  knitr         1.23       2019-05-18 [1] CRAN (R 3.6.1)                  
-#>  labeling      0.3        2014-08-23 [1] CRAN (R 3.5.0)                  
-#>  lattice       0.20-38    2018-11-04 [4] CRAN (R 3.5.1)                  
-#>  lazyeval      0.2.2      2019-03-15 [1] CRAN (R 3.6.1)                  
-#>  lifecycle     0.1.0      2019-08-01 [1] CRAN (R 3.6.1)                  
-#>  lubridate   * 1.7.4      2018-04-11 [1] CRAN (R 3.5.0)                  
-#>  magrittr    * 1.5        2014-11-22 [1] CRAN (R 3.5.0)                  
-#>  Matrix      * 1.2-18     2019-11-27 [4] CRAN (R 3.6.1)                  
-#>  memoise       1.1.0      2017-04-21 [1] CRAN (R 3.5.0)                  
-#>  mgcv          1.8-24     2018-06-18 [1] CRAN (R 3.5.1)                  
-#>  modelr        0.1.5      2019-08-08 [1] CRAN (R 3.6.1)                  
-#>  munsell       0.5.0      2018-06-12 [1] CRAN (R 3.6.1)                  
-#>  nlme          3.1-142    2019-11-07 [4] CRAN (R 3.6.1)                  
-#>  pillar        1.4.2      2019-06-29 [1] CRAN (R 3.6.1)                  
-#>  pkgbuild      1.0.3      2019-03-20 [1] CRAN (R 3.6.1)                  
-#>  pkgconfig     2.0.3      2019-09-22 [1] CRAN (R 3.6.1)                  
-#>  pkgload       1.0.2      2018-10-29 [1] CRAN (R 3.6.1)                  
-#>  prettyunits   1.0.2      2015-07-13 [1] CRAN (R 3.5.2)                  
-#>  processx      3.4.1      2019-07-18 [1] CRAN (R 3.6.1)                  
-#>  ps            1.3.0      2018-12-21 [1] CRAN (R 3.6.1)                  
-#>  purrr       * 0.3.3      2019-10-18 [1] CRAN (R 3.6.1)                  
-#>  R6            2.4.1      2019-11-12 [1] CRAN (R 3.6.1)                  
-#>  Rcpp          1.0.3      2019-11-08 [1] CRAN (R 3.6.1)                  
-#>  readr       * 1.3.1      2018-12-21 [1] CRAN (R 3.6.1)                  
-#>  readxl        1.3.1      2019-03-13 [1] CRAN (R 3.6.1)                  
-#>  remotes       2.1.0      2019-06-24 [1] CRAN (R 3.6.1)                  
-#>  rlang         0.4.1      2019-10-24 [1] CRAN (R 3.6.1)                  
-#>  rmarkdown     1.16       2019-10-01 [1] CRAN (R 3.6.1)                  
-#>  rprojroot     1.3-2      2018-01-03 [1] CRAN (R 3.5.0)                  
-#>  rstudioapi    0.10       2019-03-19 [1] CRAN (R 3.6.1)                  
-#>  rticles       0.9.1      2019-07-22 [1] Github (rstudio/rticles@8d56fc6)
-#>  rvest         0.3.5      2019-11-08 [1] CRAN (R 3.6.1)                  
-#>  scales        1.0.0      2018-08-09 [1] CRAN (R 3.6.1)                  
-#>  sessioninfo   1.1.1      2018-11-05 [1] CRAN (R 3.6.1)                  
-#>  stringi       1.4.3      2019-03-12 [1] CRAN (R 3.6.1)                  
-#>  stringr     * 1.4.0      2019-02-10 [1] CRAN (R 3.6.1)                  
-#>  survival    * 2.44-1.1   2019-04-01 [4] CRAN (R 3.6.1)                  
-#>  testthat      2.2.0      2019-07-22 [1] CRAN (R 3.6.1)                  
-#>  tibble      * 2.1.3      2019-06-06 [1] CRAN (R 3.6.1)                  
-#>  tidyr       * 1.0.0      2019-09-11 [1] CRAN (R 3.6.1)                  
-#>  tidyselect    0.2.5      2018-10-11 [1] CRAN (R 3.5.2)                  
-#>  tidyverse   * 1.2.1      2017-11-14 [1] CRAN (R 3.5.0)                  
-#>  usethis       1.5.1      2019-07-04 [1] CRAN (R 3.6.1)                  
-#>  vctrs         0.2.0      2019-07-05 [1] CRAN (R 3.6.1)                  
-#>  VGAM          1.1-1      2019-02-18 [1] CRAN (R 3.6.1)                  
-#>  withr         2.1.2      2018-03-15 [1] CRAN (R 3.5.0)                  
-#>  xfun          0.8        2019-06-25 [1] CRAN (R 3.6.1)                  
-#>  xml2          1.2.2      2019-08-09 [1] CRAN (R 3.6.1)                  
-#>  yaml          2.2.0      2018-07-25 [1] CRAN (R 3.6.1)                  
-#>  zeallot       0.1.0      2018-01-28 [1] CRAN (R 3.5.1)                  
+#> ─ Packages ───────────────────────────────────────────────────────────────────
+#>  package     * version    date       lib
+#>  assertthat    0.2.1      2019-03-21 [1]
+#>  backports     1.1.5      2019-10-02 [1]
+#>  broom         0.5.2      2019-04-07 [1]
+#>  callr         3.4.0      2019-12-09 [1]
+#>  casebase    * 0.2.1.9001 2019-12-11 [1]
+#>  cellranger    1.1.0      2016-07-27 [1]
+#>  cli           2.0.0      2019-12-09 [1]
+#>  codetools     0.2-16     2018-12-24 [1]
+#>  colorspace    1.4-1      2019-03-18 [1]
+#>  crayon        1.3.4      2017-09-16 [1]
+#>  data.table    1.12.8     2019-12-09 [1]
+#>  DBI           1.0.0      2018-05-02 [1]
+#>  dbplyr        1.4.2      2019-06-17 [1]
+#>  desc          1.2.0      2018-05-01 [1]
+#>  devtools      2.2.1      2019-09-24 [1]
+#>  digest        0.6.23     2019-11-23 [1]
+#>  dplyr       * 0.8.3      2019-07-04 [1]
+#>  ellipsis      0.3.0      2019-09-20 [1]
+#>  evaluate      0.14       2019-05-28 [1]
+#>  fansi         0.4.0      2018-10-05 [1]
+#>  forcats     * 0.4.0      2019-02-17 [1]
+#>  foreach       1.4.7      2019-07-27 [1]
+#>  fs            1.3.1      2019-05-06 [1]
+#>  generics      0.0.2      2018-11-29 [1]
+#>  ggplot2     * 3.2.1      2019-08-10 [1]
+#>  glmnet      * 3.0-1      2019-11-15 [1]
+#>  glue          1.3.1      2019-03-12 [1]
+#>  gtable        0.3.0      2019-03-25 [1]
+#>  haven         2.2.0      2019-11-08 [1]
+#>  hms           0.5.2      2019-10-30 [1]
+#>  htmltools     0.4.0      2019-10-04 [1]
+#>  httr          1.4.1      2019-08-05 [1]
+#>  iterators     1.0.12     2019-07-26 [1]
+#>  jsonlite      1.6        2018-12-07 [1]
+#>  knitr         1.26       2019-11-12 [1]
+#>  lattice       0.20-38    2018-11-04 [1]
+#>  lazyeval      0.2.2      2019-03-15 [1]
+#>  lifecycle     0.1.0      2019-08-01 [1]
+#>  lubridate     1.7.4      2018-04-11 [1]
+#>  magrittr    * 1.5        2014-11-22 [1]
+#>  Matrix      * 1.2-17     2019-03-22 [1]
+#>  memoise       1.1.0      2017-04-21 [1]
+#>  mgcv          1.8-31     2019-11-09 [1]
+#>  modelr        0.1.5      2019-08-08 [1]
+#>  munsell       0.5.0      2018-06-12 [1]
+#>  nlme          3.1-142    2019-11-07 [1]
+#>  pillar        1.4.2      2019-06-29 [1]
+#>  pkgbuild      1.0.6      2019-10-09 [1]
+#>  pkgconfig     2.0.3      2019-09-22 [1]
+#>  pkgload       1.0.2      2018-10-29 [1]
+#>  prettyunits   1.0.2      2015-07-13 [1]
+#>  processx      3.4.1      2019-07-18 [1]
+#>  ps            1.3.0      2018-12-21 [1]
+#>  purrr       * 0.3.3      2019-10-18 [1]
+#>  R6            2.4.1      2019-11-12 [1]
+#>  Rcpp          1.0.3      2019-11-08 [1]
+#>  readr       * 1.3.1      2018-12-21 [1]
+#>  readxl        1.3.1      2019-03-13 [1]
+#>  remotes       2.1.0      2019-06-24 [1]
+#>  reprex        0.3.0      2019-05-16 [1]
+#>  rlang         0.4.2      2019-11-23 [1]
+#>  rmarkdown     1.18       2019-11-27 [1]
+#>  rprojroot     1.3-2      2018-01-03 [1]
+#>  rstudioapi    0.10       2019-03-19 [1]
+#>  rticles       0.12       2019-11-12 [1]
+#>  rvest         0.3.5      2019-11-08 [1]
+#>  scales        1.1.0      2019-11-18 [1]
+#>  sessioninfo   1.1.1      2018-11-05 [1]
+#>  shape         1.4.4      2018-02-07 [1]
+#>  stringi       1.4.3      2019-03-12 [1]
+#>  stringr     * 1.4.0      2019-02-10 [1]
+#>  survival    * 3.1-7      2019-11-09 [1]
+#>  testthat      2.3.1      2019-12-01 [1]
+#>  tibble      * 2.1.3      2019-06-06 [1]
+#>  tidyr       * 1.0.0      2019-09-11 [1]
+#>  tidyselect    0.2.5      2018-10-11 [1]
+#>  tidyverse   * 1.3.0      2019-11-21 [1]
+#>  usethis       1.5.1      2019-07-04 [1]
+#>  vctrs         0.2.0      2019-07-05 [1]
+#>  VGAM          1.1-2      2019-11-21 [1]
+#>  withr         2.1.2      2018-03-15 [1]
+#>  xfun          0.11       2019-11-12 [1]
+#>  xml2          1.2.2      2019-08-09 [1]
+#>  yaml          2.2.0      2018-07-25 [1]
+#>  zeallot       0.1.0      2018-01-28 [1]
+#>  source                                  
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  Github (sahirbhatnagar/casebase@e2e73fe)
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
 #> 
-#> [1] /home/mturgeon/Rlibs
-#> [2] /usr/local/lib/R/site-library
-#> [3] /usr/lib/R/site-library
-#> [4] /usr/lib/R/library
+#> [1] /Library/Frameworks/R.framework/Versions/3.6/Resources/library
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> Local:    review-max /home/mturgeon/Documents/git_repositories/cbpaper
-#> Head:     [69663bb] 2019-12-08: Merge branch 'review-max' of https://github.com/sahirbhatnagar/cbpaper into review-max
+#> Local:    master /Users/jislam/Desktop/Gits/cbpaper
+#> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
+#> Head:     [71a5a9c] 2019-12-06: Lit-Review Edits
 \end{CodeOutput}

### 2019-12-11 analysis/paper/paper.Rmd
edit first case study

--- 2019-12-11 analysis/paper/paper.Rmd
@@
 eval_implementation <- TRUE
-eval_cs1 <- FALSE
+eval_cs1 <- TRUE
 eval_cs2 <- FALSE
@@
 
-To introduce the different features available, we make use of the European Randomized Study of Prostate Cancer Screening data; this dataset is available through the \pkg{casebase} package:
+For the first case study, we will use of the European Randomized Study of Prostate Cancer Screening data. This dataset is available through the \pkg{casebase} package:
 
@@
 ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
-                       levels = c(0,1), 
+                       levels = c(0, 1), 
                        labels = c("Control group", "Screening group"))
@@
 
-The results of this study were published by [schroder2009screening]. This data was obtained using the approach described in [liu2014recovering].
+The results of this study were published by [@schroder2009screening], and the dataset itself was obtained using the approach described in [@liu2014recovering].
 
@@
 We facilitate this task in the \pkg{casebase} package using the \code{popTime} function.
-We first create the necessary dataset for producing the population time plots, and we can produce the plot by using the corresponding \code{plot} method:
+We first create the necessary dataset for producing the population time plots, and we can generate the plot by using the corresponding \code{plot} method:
 
@@
 
-First, we fit a Cox model to the data, examine the hazard ratio for the screening group (relative to the control group), and plot the cumulative incidence function (CIF).
+ADD PARAGRAPH ABOUT WHAT WE CAN CONCLUDE FROM THESE GRAPHS.
+
+Next, we investigate the differences between the control and the screening arms.
+A common choice for this type of analysis is to use a Cox regression model and estimate the hazard ratio for the screening group (relative to the control group).
+In \proglang{R}, it can be done as follows:
 
@@
 
-We can plot the CIF for each group:
+We can also plot the cumulative incidence function (CIF) for each group. However, this involves estimating the baseline hazard, which Cox regression treated as a nuisance parameter.
 
@@
      xlab = "Years since Randomization", 
-     ylab = "Cumulative Incidence", 
+     ylab = "Cumulative Incidence (%)", 
      fun = "event",
@@
                     exp(confint(cox_model))[1], 
-                    exp(confint(cox_model))[2]))
+                    exp(confint(cox_model))[2]),
+     ylim = c(0, 2), yscale = 100)
 legend("topleft", 
@@
 
-Next we fit several models using case-base sampling. The models we fit differ in how we choose to model time.
-
-The \code{fitSmoothHazard} function provides an estimate of the hazard function $h(x, t)$ is the hazard function, where $t$ denotes the numerical value of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any).
+Next, we show how case-base sampling can be used to carry out a similar analysis. We fit several models that differ in how we choose to model time.
+
+<!-- The \code{fitSmoothHazard} function provides an estimate of the hazard function $h(x, t)$ is the hazard function, where $t$ denotes the numerical value of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any).
+-->
+
+First, we will fit an exponential model. Recall that this corresponds to excluding time from the linear predictor.
 
@@
 
-The \code{absoluteRisk} function provides an estimate of the cumulative incidence curves for a specific risk profile using the following equation:
-
-$$ CI(x, t) = 1 - \exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$
-
-In the plot below, we overlay the estimated CIF from the casebase exponential model on the Cox model CIF:
+<!-- The \code{absoluteRisk} function provides an estimate of the cumulative incidence curves for a specific risk profile using the following equation: -->
+
+<!-- $$ CI(x, t) = 1 - \exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$ -->
+
+We can then use the \code{absoluteRisk} function to get an estimate of the cumulative incidence curve for a specific covariate profile.
+In the plot below, we overlay the estimated CIF from the exponential model on the Cox model CIF:
 
@@
                     exp(confint(cox_model))[1], 
-                    exp(confint(cox_model))[2]))
+                    exp(confint(cox_model))[2]),
+     ylim = c(0, 2), yscale = 100)
 lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
@@
 
-As we can see, the exponential model is not a good fit.
-Based on what we observed in the population time plot, where more events are observed later on in time, this poor fit is expected.
+As we can see, the exponential model gives us an estimate of the hazard ratio that is similar to the one obtained using Cox regression.
+However, the CIF estimates are quite different.
+Based on what we observed in the population time plot, where more events are observed later on in time, we do not expect a constant hazard would be a good description for this data.
+In other words, this poor fit for the exponential hazard is expected.
 A constant hazard model would overestimate the cumulative incidence earlier on in time, and underestimate it later on; this is what we see on the cumulative incidence plot.
@@
 
-Next we enter time linearly into the model:
+For the next model, we include time as a linear term. Recall that this corresponds to a Weibull model.
 
@@
 
+Again, the estimate of the hazard ratio is similar to that obtained from Cox regression. We then look at the estimate of the CIF:
+
 ```{r erspc-casebase-weibull-cif, eval = eval_cs1}
@@
                     exp(confint(cox_model))[1], 
-                    exp(confint(cox_model))[2]))
+                    exp(confint(cox_model))[2]),
+     ylim = c(0, 2), yscale = 100)
 lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
@@
 
-We see that the Weibull model leads to a better fit.
-
-Next we try to enter a smooth function of time into the model using the \code{splines} package:
+We see that the Weibull model leads to a better fit of the CIF than the Exponential model, when compared with the semi-parametric approach.
+
+Finally, we can model the hazard as a smooth function of time using the \code{splines} package:
 
@@
 
+Once again, we can see that the estimate of the hazard ratio is similar to that from the Cox regression. We then look at the estimate of the CIF:
+
 ```{r erspc-casebase-splines-cif, eval = eval_cs1}
@@
                     exp(confint(cox_model))[1], 
-                    exp(confint(cox_model))[2]))
+                    exp(confint(cox_model))[2]),
+     ylim = c(0, 2), yscale = 100)
 lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
@@
 
-It looks like the best fit.
-
-Since we are within the GLM framework, we can easily test for which model better fits the data using a Likelihood Ratio Test (LRT).
-The null hypothesis here is that the linear model is just as good as the larger (in terms of number of parameters) splines model.
+Qualitatively, the combination of splines and case-base sampling produces the parametric model that most closely resembles the Cox model.
+In the following table, we see that the confidence intervals are also similar across all four models.
+In other words, this reinforces the idea that, under proportional hazards, we do not need to model the full hazard to obtain reliable estimates of the hazard ratio.
+Of course, different parametric models for the hazards give rise to qualitatively different estimates for the CIF.
+
+```{r echo = FALSE}
+CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
+CI_exp <- c(exp(coef(casebase_exponential))["ScrArmScreening group"], exp(confint(casebase_exponential))["ScrArmScreening group",])
+CI_time <- c(exp(coef(casebase_time))["ScrArmScreening group"], exp(confint(casebase_time))["ScrArmScreening group",])
+CI_splines <- c(exp(coef(casebase_splines))["ScrArmScreening group"], exp(confint(casebase_splines))["ScrArmScreening group",])
+
+table_ci <- rbind(CI_cox, CI_exp, CI_time, CI_splines)
+colnames(table_ci) <- c("HR", "LB", "UB")
+
+data.frame(Model = c("Cox", "Exponential", "Weibull", "Splines")) %>% 
+  bind_cols(as.data.frame(table_ci)) %>% 
+  mutate(CI = paste0("(", round(LB, 2), ", ",
+                     round(UB, 2), ")"),
+         Width = UB - LB) %>% 
+  dplyr::select(-LB, -UB) %>% 
+  knitr::kable(format = "latex")
+```
+
+As noted above, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models.
+Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) splines model.
 
 ```{r erscp-compare, eval = eval_cs1}
-anova(casebase_time, casebase_splines, test = "LRT")
-```
-
-As expected, we see that splines model provides a better fit.
+anova(casebase_exponential, casebase_splines, test = "LRT")
+```
+
+As expected, we see that splines model provides a better fit. Similarly, we can compare the AIC for all three case-base models:
+
+```{r erscp-aic, eval = eval_cs1}
+AIC(casebase_exponential)
+AIC(casebase_time)
+AIC(casebase_splines)
+```
+
+Once again, we have evidence that the splines model provides the best fit.
+
+With this first case study, we saw how population-time plots can provide a useful summary of the incidence for our event of interests.
+In particular, we saw how it can provide evidence against an exponential hazard model.
+We then explored how different parametric models can provide different estimates of the CIF, even though they all give similar estimates for the hazard ratios.
 

### 2019-12-11 analysis/paper/paper.tex
edit first case study

--- 2019-12-11 analysis/paper/paper.tex
@@
 
-To introduce the different features available, we make use of the
-European Randomized Study of Prostate Cancer Screening data; this
-dataset is available through the \pkg{casebase} package:
+For the first case study, we will use of the European Randomized Study
+of Prostate Cancer Screening data. This dataset is available through the
+\pkg{casebase} package:
 
@@
 R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
-R>                        levels = c(0,1), 
-R>                        labels = c("Control group", "Screening group"))
-\end{CodeInput}
-\end{CodeChunk}
-
-The results of this study were published by {[}schroder2009screening{]}.
-This data was obtained using the approach described in
-{[}liu2014recovering{]}.
+R+                        levels = c(0, 1), 
+R+                        labels = c("Control group", "Screening group"))
+\end{CodeInput}
+\end{CodeChunk}
+
+The results of this study were published by
+\citep{schroder2009screening}, and the dataset itself was obtained using
+the approach described in \citep{liu2014recovering}.
 
@@
 the \code{popTime} function. We first create the necessary dataset for
-producing the population time plots, and we can produce the plot by
+producing the population time plots, and we can generate the plot by
 using the corresponding \code{plot} method:
@@
 \end{CodeInput}
-\end{CodeChunk}
-
-First, we fit a Cox model to the data, examine the hazard ratio for the
-screening group (relative to the control group), and plot the cumulative
-incidence function (CIF).
+
+\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-side-1} \end{center}
+
+\end{CodeChunk}
+
+ADD PARAGRAPH ABOUT WHAT WE CAN CONCLUDE FROM THESE GRAPHS.
+
+Next, we investigate the differences between the control and the
+screening arms. A common choice for this type of analysis is to use a
+Cox regression model and estimate the hazard ratio for the screening
+group (relative to the control group). In \proglang{R}, it can be done
+as follows:
 
@@
 R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
-R>                              data = ERSPC)
+R+                              data = ERSPC)
 R> summary(cox_model)
 \end{CodeInput}
-\end{CodeChunk}
-
-We can plot the CIF for each group:
+
+\begin{CodeOutput}
+#> Call:
+#> survival::coxph(formula = Surv(Follow.Up.Time, DeadOfPrCa) ~ 
+#>     ScrArm, data = ERSPC)
+#> 
+#>   n= 159893, number of events= 540 
+#> 
+#>                         coef exp(coef) se(coef)     z Pr(>|z|)  
+#> ScrArmScreening group -0.222     0.801    0.088 -2.52    0.012 *
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#>                       exp(coef) exp(-coef) lower .95 upper .95
+#> ScrArmScreening group     0.801       1.25     0.674     0.952
+#> 
+#> Concordance= 0.519  (se = 0.011 )
+#> Likelihood ratio test= 6.45  on 1 df,   p=0.01
+#> Wald test            = 6.37  on 1 df,   p=0.01
+#> Score (logrank) test = 6.39  on 1 df,   p=0.01
+\end{CodeOutput}
+\end{CodeChunk}
+
+We can also plot the cumulative incidence function (CIF) for each group.
+However, this involves estimating the baseline hazard, which Cox
+regression treated as a nuisance parameter.
 
@@
 R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
-R>                        ignore = 99)
+R+                        ignore = 99)
 R> 
 R> plot(survfit(cox_model, newdata = new_data),
-R>      xlab = "Years since Randomization", 
-R>      ylab = "Cumulative Incidence", 
-R>      fun = "event",
-R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>                     exp(coef(cox_model)), 
-R>                     exp(confint(cox_model))[1], 
-R>                     exp(confint(cox_model))[2]))
+R+      xlab = "Years since Randomization", 
+R+      ylab = "Cumulative Incidence (%)", 
+R+      fun = "event",
+R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+                     exp(coef(cox_model)), 
+R+                     exp(confint(cox_model))[1], 
+R+                     exp(confint(cox_model))[2]),
+R+      ylim = c(0, 2), yscale = 100)
 R> legend("topleft", 
-R>        legend = c("Control group", "Screening group"), 
-R>        col = c("red","blue"),
-R>        lty = c(1, 1), 
-R>        bg = "gray90")
-\end{CodeInput}
-\end{CodeChunk}
-
-Next we fit several models using case-base sampling. The models we fit
-differ in how we choose to model time.
-
-The \code{fitSmoothHazard} function provides an estimate of the hazard
-function \(h(x, t)\) is the hazard function, where \(t\) denotes the
-numerical value of a point in prognostic/prospective time and \(x\) is
-the realization of the vector \(X\) of variates based on the patient's
-profile and intervention (if any).
+R+        legend = c("Control group", "Screening group"), 
+R+        col = c("red","blue"),
+R+        lty = c(1, 1), 
+R+        bg = "gray90")
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/erspc-cox-cif-1} \end{center}
+
+\end{CodeChunk}
+
+Next, we show how case-base sampling can be used to carry out a similar
+analysis. We fit several models that differ in how we choose to model
+time.
+
+First, we will fit an exponential model. Recall that this corresponds to
+excluding time from the linear predictor.
 
@@
 R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
-R>                                                   data = ERSPC, 
-R>                                                   ratio = 100)
+R+                                                   data = ERSPC, 
+R+                                                   ratio = 100)
 R> 
 R> summary(casebase_exponential)
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> 
+#> Call:
+#> glm(formula = formula, family = binomial, data = sampleData)
+#> 
+#> Deviance Residuals: 
+#>    Min      1Q  Median      3Q     Max  
+#> -0.148  -0.148  -0.148  -0.133   3.079  
+#> 
+#> Coefficients:
+#>                       Estimate Std. Error z value            Pr(>|z|)    
+#> (Intercept)            -7.7650     0.0557 -139.44 <0.0000000000000002 ***
+#> ScrArmScreening group  -0.2166     0.0884   -2.45               0.014 *  
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#> (Dispersion parameter for binomial family taken to be 1)
+#> 
+#>     Null deviance: 6059.0  on 54539  degrees of freedom
+#> Residual deviance: 6052.9  on 54538  degrees of freedom
+#> AIC: 6057
+#> 
+#> Number of Fisher Scoring iterations: 7
+\end{CodeOutput}
+
+\begin{CodeInput}
 R> exp(coef(casebase_exponential)[2])
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> ScrArmScreening group 
+#>                  0.81
+\end{CodeOutput}
+
+\begin{CodeInput}
 R> exp(confint(casebase_exponential)[2,])
 \end{CodeInput}
-\end{CodeChunk}
-
-The \code{absoluteRisk} function provides an estimate of the cumulative
-incidence curves for a specific risk profile using the following
-equation:
-
-\[ CI(x, t) = 1 - \exp\left( - \int_0^t h(x, u) \textrm{d}u \right) \]
-
-In the plot below, we overlay the estimated CIF from the casebase
-exponential model on the Cox model CIF:
+
+\begin{CodeOutput}
+#>  2.5 % 97.5 % 
+#>   0.68   0.96
+\end{CodeOutput}
+\end{CodeChunk}
+
+We can then use the \code{absoluteRisk} function to get an estimate of
+the cumulative incidence curve for a specific covariate profile. In the
+plot below, we overlay the estimated CIF from the exponential model on
+the Cox model CIF:
 
@@
 R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
-R>                                           time = seq(0,15,0.1), 
-R>                                           newdata = new_data)
+R+                                           time = seq(0,15,0.1), 
+R+                                           newdata = new_data)
 R> 
 R> plot(survfit(cox_model, newdata = new_data),
-R>      xlab = "Years since Randomization", 
-R>      ylab = "Cumulative Incidence", 
-R>      fun = "event",
-R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>                     exp(coef(cox_model)), 
-R>                     exp(confint(cox_model))[1], 
-R>                     exp(confint(cox_model))[2]))
+R+      xlab = "Years since Randomization", 
+R+      ylab = "Cumulative Incidence (%)", 
+R+      fun = "event",
+R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+                     exp(coef(cox_model)), 
+R+                     exp(confint(cox_model))[1], 
+R+                     exp(confint(cox_model))[2]),
+R+      ylim = c(0, 2), yscale = 100)
 R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
@@
 R> legend("topleft", 
-R>        legend = c("Control group (Cox)","Control group (Casebase)",
-R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R>        col = c("red","red", "blue","blue"),
-R>        lty = c(1, 2, 1, 2), 
-R>        bg = "gray90")
-\end{CodeInput}
-\end{CodeChunk}
-
-As we can see, the exponential model is not a good fit. Based on what we
-observed in the population time plot, where more events are observed
-later on in time, this poor fit is expected. A constant hazard model
-would overestimate the cumulative incidence earlier on in time, and
-underestimate it later on; this is what we see on the cumulative
-incidence plot. This example demonstrates the benefits of population
-time plots as an exploratory analysis tool.
-
-Next we enter time linearly into the model:
+R+        legend = c("Control group (Cox)","Control group (Casebase)",
+R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R+        col = c("red","red", "blue","blue"),
+R+        lty = c(1, 2, 1, 2), 
+R+        bg = "gray90")
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/erspc-casebase-cif-1} \end{center}
+
+\end{CodeChunk}
+
+As we can see, the exponential model gives us an estimate of the hazard
+ratio that is similar to the one obtained using Cox regression. However,
+the CIF estimates are quite different. Based on what we observed in the
+population time plot, where more events are observed later on in time,
+we do not expect a constant hazard would be a good description for this
+data. In other words, this poor fit for the exponential hazard is
+expected. A constant hazard model would overestimate the cumulative
+incidence earlier on in time, and underestimate it later on; this is
+what we see on the cumulative incidence plot. This example demonstrates
+the benefits of population time plots as an exploratory analysis tool.
+
+For the next model, we include time as a linear term. Recall that this
+corresponds to a Weibull model.
 
@@
 R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
-R>                                  data = ERSPC, 
-R>                                  ratio = 100)
+R+                                  data = ERSPC, 
+R+                                  ratio = 100)
 R> 
 R> summary(casebase_time)
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> 
+#> Call:
+#> glm(formula = formula, family = binomial, data = sampleData)
+#> 
+#> Deviance Residuals: 
+#>    Min      1Q  Median      3Q     Max  
+#> -0.389  -0.161  -0.122  -0.095   3.464  
+#> 
+#> Coefficients:
+#>                       Estimate Std. Error z value            Pr(>|z|)    
+#> (Intercept)            -9.0257     0.1122  -80.42 <0.0000000000000002 ***
+#> Follow.Up.Time          0.2207     0.0145   15.18 <0.0000000000000002 ***
+#> ScrArmScreening group  -0.2311     0.0886   -2.61              0.0091 ** 
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#> (Dispersion parameter for binomial family taken to be 1)
+#> 
+#>     Null deviance: 6059  on 54539  degrees of freedom
+#> Residual deviance: 5815  on 54537  degrees of freedom
+#> AIC: 5821
+#> 
+#> Number of Fisher Scoring iterations: 8
+\end{CodeOutput}
+
+\begin{CodeInput}
 R> exp(coef(casebase_time))
+\end{CodeInput}
+
+\begin{CodeOutput}
+#>           (Intercept)        Follow.Up.Time ScrArmScreening group 
+#>               0.00012               1.24701               0.79370
+\end{CodeOutput}
+
+\begin{CodeInput}
 R> exp(confint(casebase_time))
 \end{CodeInput}
-\end{CodeChunk}
+
+\begin{CodeOutput}
+#>                          2.5 %  97.5 %
+#> (Intercept)           0.000096 0.00015
+#> Follow.Up.Time        1.212086 1.28321
+#> ScrArmScreening group 0.666385 0.94341
+\end{CodeOutput}
+\end{CodeChunk}
+
+Again, the estimate of the hazard ratio is similar to that obtained from
+Cox regression. We then look at the estimate of the CIF:
 
@@
 R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
-R>                                           time = seq(0,15,0.1), 
-R>                                           newdata = new_data)
+R+                                           time = seq(0,15,0.1), 
+R+                                           newdata = new_data)
 R> 
 R> plot(survfit(cox_model, newdata = new_data),
-R>      xlab = "Years since Randomization", 
-R>      ylab = "Cumulative Incidence", 
-R>      fun = "event",
-R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>                     exp(coef(cox_model)), 
-R>                     exp(confint(cox_model))[1], 
-R>                     exp(confint(cox_model))[2]))
+R+      xlab = "Years since Randomization", 
+R+      ylab = "Cumulative Incidence (%)", 
+R+      fun = "event",
+R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+                     exp(coef(cox_model)), 
+R+                     exp(confint(cox_model))[1], 
+R+                     exp(confint(cox_model))[2]),
+R+      ylim = c(0, 2), yscale = 100)
 R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
@@
 R> legend("topleft", 
-R>        legend = c("Control group (Cox)","Control group (Casebase)",
-R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R>        col = c("red","red", "blue","blue"),
-R>        lty = c(1, 2, 1, 2), 
-R>        bg = "gray90")
-\end{CodeInput}
-\end{CodeChunk}
-
-We see that the Weibull model leads to a better fit.
-
-Next we try to enter a smooth function of time into the model using the
+R+        legend = c("Control group (Cox)","Control group (Casebase)",
+R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R+        col = c("red","red", "blue","blue"),
+R+        lty = c(1, 2, 1, 2), 
+R+        bg = "gray90")
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/erspc-casebase-weibull-cif-1} \end{center}
+
+\end{CodeChunk}
+
+We see that the Weibull model leads to a better fit of the CIF than the
+Exponential model, when compared with the semi-parametric approach.
+
+Finally, we can model the hazard as a smooth function of time using the
 \code{splines} package:
@@
 R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
-R>                                     data = ERSPC, 
-R>                                     ratio = 100)
+R+                                     data = ERSPC, 
+R+                                     ratio = 100)
 R> 
 R> summary(casebase_splines)
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> 
+#> Call:
+#> glm(formula = formula, family = binomial, data = sampleData)
+#> 
+#> Deviance Residuals: 
+#>    Min      1Q  Median      3Q     Max  
+#> -0.397  -0.173  -0.136  -0.085   3.797  
+#> 
+#> Coefficients:
+#>                       Estimate Std. Error z value             Pr(>|z|)    
+#> (Intercept)           -10.2787     0.3175  -32.38 < 0.0000000000000002 ***
+#> bs(Follow.Up.Time)1     4.3191     0.8062    5.36        0.00000008456 ***
+#> bs(Follow.Up.Time)2     2.1061     0.4878    4.32        0.00001580161 ***
+#> bs(Follow.Up.Time)3     4.5310     0.7160    6.33        0.00000000025 ***
+#> ScrArmScreening group  -0.2068     0.0886   -2.33                 0.02 *  
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#> (Dispersion parameter for binomial family taken to be 1)
+#> 
+#>     Null deviance: 6059.0  on 54539  degrees of freedom
+#> Residual deviance: 5792.2  on 54535  degrees of freedom
+#> AIC: 5802
+#> 
+#> Number of Fisher Scoring iterations: 9
+\end{CodeOutput}
+
+\begin{CodeInput}
 R> exp(coef(casebase_splines))
+\end{CodeInput}
+
+\begin{CodeOutput}
+#>           (Intercept)   bs(Follow.Up.Time)1   bs(Follow.Up.Time)2 
+#>              0.000034             75.119228              8.216108 
+#>   bs(Follow.Up.Time)3 ScrArmScreening group 
+#>             92.847641              0.813203
+\end{CodeOutput}
+
+\begin{CodeInput}
 R> exp(confint(casebase_splines))
 \end{CodeInput}
-\end{CodeChunk}
+
+\begin{CodeOutput}
+#>                           2.5 %     97.5 %
+#> (Intercept)            0.000018   0.000062
+#> bs(Follow.Up.Time)1   15.997506 378.737706
+#> bs(Follow.Up.Time)2    3.256838  22.143147
+#> bs(Follow.Up.Time)3   21.638298 362.098155
+#> ScrArmScreening group  0.682786   0.966562
+\end{CodeOutput}
+\end{CodeChunk}
+
+Once again, we can see that the estimate of the hazard ratio is similar
+to that from the Cox regression. We then look at the estimate of the
+CIF:
 
@@
 R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
-R>                                     time = seq(0,15,0.1), 
-R>                                     newdata = new_data)
+R+                                     time = seq(0,15,0.1), 
+R+                                     newdata = new_data)
 R> 
 R> plot(survfit(cox_model, newdata = new_data),
-R>      xlab = "Years since Randomization", 
-R>      ylab = "Cumulative Incidence", 
-R>      fun = "event",
-R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>                     exp(coef(cox_model)), 
-R>                     exp(confint(cox_model))[1], 
-R>                     exp(confint(cox_model))[2]))
+R+      xlab = "Years since Randomization", 
+R+      ylab = "Cumulative Incidence (%)", 
+R+      fun = "event",
+R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+                     exp(coef(cox_model)), 
+R+                     exp(confint(cox_model))[1], 
+R+                     exp(confint(cox_model))[2]),
+R+      ylim = c(0, 2), yscale = 100)
 R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
@@
 R> legend("topleft", 
-R>        legend = c("Control group (Cox)","Control group (Casebase)",
-R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R>        col = c("red","red", "blue","blue"),
-R>        lty = c(1, 2, 1, 2), 
-R>        bg = "gray90")
-\end{CodeInput}
-\end{CodeChunk}
-
-It looks like the best fit.
-
-Since we are within the GLM framework, we can easily test for which
-model better fits the data using a Likelihood Ratio Test (LRT). The null
-hypothesis here is that the linear model is just as good as the larger
-(in terms of number of parameters) splines model.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> anova(casebase_time, casebase_splines, test = "LRT")
-\end{CodeInput}
-\end{CodeChunk}
-
-As expected, we see that splines model provides a better fit.
+R+        legend = c("Control group (Cox)","Control group (Casebase)",
+R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R+        col = c("red","red", "blue","blue"),
+R+        lty = c(1, 2, 1, 2), 
+R+        bg = "gray90")
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/erspc-casebase-splines-cif-1} \end{center}
+
+\end{CodeChunk}
+
+Qualitatively, the combination of splines and case-base sampling
+produces the parametric model that most closely resembles the Cox model.
+In the following table, we see that the confidence intervals are also
+similar across all four models. In other words, this reinforces the idea
+that, under proportional hazards, we do not need to model the full
+hazard to obtain reliable estimates of the hazard ratio. Of course,
+different parametric models for the hazards give rise to qualitatively
+different estimates for the CIF.
+
+\begin{CodeChunk}
+
+\begin{tabular}{l|r|l|r}
+\hline
+Model & HR & CI & Width\\
+\hline
+Cox & 0.80 & (0.67, 0.95) & 0.28\\
+\hline
+Exponential & 0.81 & (0.68, 0.96) & 0.28\\
+\hline
+Weibull & 0.79 & (0.67, 0.94) & 0.28\\
+\hline
+Splines & 0.81 & (0.68, 0.97) & 0.28\\
+\hline
+\end{tabular}
+
+\end{CodeChunk}
+
+As noted above, the usual asymptotic results hold for likelihood ratio
+tests built using case-base sampling models. Therefore, we can easily
+test the null hypothesis that the exponential model is just as good as
+the larger (in terms of number of parameters) splines model.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> anova(casebase_exponential, casebase_splines, test = "LRT")
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> Analysis of Deviance Table
+#> 
+#> Model 1: DeadOfPrCa ~ ScrArm + offset(offset)
+#> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
+#>   Resid. Df Resid. Dev Df Deviance            Pr(>Chi)    
+#> 1     54538       6053                                    
+#> 2     54535       5792  3      261 <0.0000000000000002 ***
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{CodeOutput}
+\end{CodeChunk}
+
+As expected, we see that splines model provides a better fit. Similarly,
+we can compare the AIC for all three case-base models:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> AIC(casebase_exponential)
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> [1] 6057
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> AIC(casebase_time)
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> [1] 5821
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> AIC(casebase_splines)
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> [1] 5802
+\end{CodeOutput}
+\end{CodeChunk}
+
+Once again, we have evidence that the splines model provides the best
+fit.
+
+With this first case study, we saw how population-time plots can provide
+a useful summary of the incidence for our event of interests. In
+particular, we saw how it can provide evidence against an exponential
+hazard model. We then explored how different parametric models can
+provide different estimates of the CIF, even though they all give
+similar estimates for the hazard ratios.
 
@@
 #> Model1 Model2 Model3 Model4 
-#>    798    759    737    708
+#>    810    772    741    722
 \end{CodeOutput}
@@
 #>         
-#> 365 0.32
+#> 365 0.25
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> [1] 0.24
+#> [1] 0.2
 \end{CodeOutput}
@@
 
-This report was generated on 2019-12-11 10:31:00 using the following
+This report was generated on 2019-12-11 15:19:04 using the following
 computational environment and dependencies:
@@
 #>  magrittr    * 1.5        2014-11-22 [1]
+#>  MASS          7.3-51.4   2019-04-26 [4]
 #>  Matrix      * 1.2-18     2019-11-27 [4]
@@
 #>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
 #>  CRAN (R 3.5.0)                          
@@
 #> Local:    review-glmnet /home/mturgeon/Documents/git_repositories/cbpaper
-#> Head:     [b7e23a6] 2019-12-10: Merge branch 'master' into review-glmnet
+#> Head:     [491ef55] 2019-12-11: edit third case study
 \end{CodeOutput}

### 2019-12-11 analysis/paper/paper.tex

--- 2019-12-11 analysis/paper/paper.tex
@@
 comparable to casebase, with either a different approach to the same
-problems or future features we would like to add to casebase.
+problems or future features we would like to add to casebase. In
+particular, we excluded packages that were focused on a different set of
+problems, such as frailty and multi-state models, as casebase is
+currently focused on absolute-risk, competing events and single event
+analysis.
 
@@
 
-This report was generated on 2019-12-11 12:28:29 using the following
+This report was generated on 2019-12-11 15:31:05 using the following
 computational environment and dependencies:
@@
 \begin{CodeOutput}
-#> ─ Session info ───────────────────────────────────────────────────────────────
+#> ─ Session info ──────────────────────────────────────────────────────────
 #>  setting  value                       
-#>  version  R version 3.6.0 (2019-04-26)
-#>  os       macOS High Sierra 10.13.6   
-#>  system   x86_64, darwin15.6.0        
+#>  version  R version 3.6.1 (2019-07-05)
+#>  os       Ubuntu 18.04.3 LTS          
+#>  system   x86_64, linux-gnu           
 #>  ui       X11                         
@@
 #>  ctype    en_CA.UTF-8                 
-#>  tz       America/Montreal            
+#>  tz       America/Winnipeg            
 #>  date     2019-12-11                  
 #> 
-#> ─ Packages ───────────────────────────────────────────────────────────────────
+#> ─ Packages ──────────────────────────────────────────────────────────────
 #>  package     * version    date       lib
@@
 #>  broom         0.5.2      2019-04-07 [1]
-#>  callr         3.4.0      2019-12-09 [1]
+#>  callr         3.3.1      2019-07-18 [1]
 #>  casebase    * 0.2.1.9001 2019-12-11 [1]
@@
 #>  cli           2.0.0      2019-12-09 [1]
-#>  codetools     0.2-16     2018-12-24 [1]
+#>  codetools     0.2-16     2018-12-24 [4]
 #>  colorspace    1.4-1      2019-03-18 [1]
@@
 #>  data.table    1.12.8     2019-12-09 [1]
-#>  DBI           1.0.0      2018-05-02 [1]
-#>  dbplyr        1.4.2      2019-06-17 [1]
 #>  desc          1.2.0      2018-05-01 [1]
-#>  devtools      2.2.1      2019-09-24 [1]
+#>  devtools      2.1.0      2019-07-06 [1]
 #>  digest        0.6.23     2019-11-23 [1]
 #>  dplyr       * 0.8.3      2019-07-04 [1]
-#>  ellipsis      0.3.0      2019-09-20 [1]
 #>  evaluate      0.14       2019-05-28 [1]
@@
 #>  forcats     * 0.4.0      2019-02-17 [1]
-#>  foreach       1.4.7      2019-07-27 [1]
+#>  foreach     * 1.4.4      2017-12-12 [1]
 #>  fs            1.3.1      2019-05-06 [1]
@@
 #>  ggplot2     * 3.2.1      2019-08-10 [1]
-#>  glmnet      * 3.0-1      2019-11-15 [1]
+#>  glmnet      * 2.0-18     2019-05-20 [1]
 #>  glue          1.3.1      2019-03-12 [1]
 #>  gtable        0.3.0      2019-03-25 [1]
-#>  haven         2.2.0      2019-11-08 [1]
+#>  haven         2.1.1      2019-07-04 [1]
 #>  hms           0.5.2      2019-10-30 [1]
-#>  htmltools     0.4.0      2019-10-04 [1]
+#>  htmltools     0.3.6      2017-04-28 [1]
 #>  httr          1.4.1      2019-08-05 [1]
-#>  iterators     1.0.12     2019-07-26 [1]
+#>  iterators     1.0.9      2017-12-12 [1]
 #>  jsonlite      1.6        2018-12-07 [1]
-#>  knitr         1.26       2019-11-12 [1]
-#>  lattice       0.20-38    2018-11-04 [1]
+#>  knitr         1.23       2019-05-18 [1]
+#>  lattice       0.20-38    2018-11-04 [4]
 #>  lazyeval      0.2.2      2019-03-15 [1]
@@
 #>  magrittr    * 1.5        2014-11-22 [1]
-#>  Matrix      * 1.2-17     2019-03-22 [1]
+#>  Matrix      * 1.2-18     2019-11-27 [4]
 #>  memoise       1.1.0      2017-04-21 [1]
-#>  mgcv          1.8-31     2019-11-09 [1]
+#>  mgcv          1.8-24     2018-06-18 [1]
 #>  modelr        0.1.5      2019-08-08 [1]
 #>  munsell       0.5.0      2018-06-12 [1]
-#>  nlme          3.1-142    2019-11-07 [1]
+#>  nlme          3.1-142    2019-11-07 [4]
 #>  pillar        1.4.2      2019-06-29 [1]
-#>  pkgbuild      1.0.6      2019-10-09 [1]
+#>  pkgbuild      1.0.3      2019-03-20 [1]
 #>  pkgconfig     2.0.3      2019-09-22 [1]
@@
 #>  remotes       2.1.0      2019-06-24 [1]
-#>  reprex        0.3.0      2019-05-16 [1]
 #>  rlang         0.4.2      2019-11-23 [1]
-#>  rmarkdown     1.18       2019-11-27 [1]
+#>  rmarkdown     1.16       2019-10-01 [1]
 #>  rprojroot     1.3-2      2018-01-03 [1]
 #>  rstudioapi    0.10       2019-03-19 [1]
-#>  rticles       0.12       2019-11-12 [1]
+#>  rticles       0.9.1      2019-07-22 [1]
 #>  rvest         0.3.5      2019-11-08 [1]
@@
 #>  sessioninfo   1.1.1      2018-11-05 [1]
-#>  shape         1.4.4      2018-02-07 [1]
 #>  stringi       1.4.3      2019-03-12 [1]
 #>  stringr     * 1.4.0      2019-02-10 [1]
-#>  survival    * 3.1-7      2019-11-09 [1]
-#>  testthat      2.3.1      2019-12-01 [1]
+#>  survival    * 2.44-1.1   2019-04-01 [4]
+#>  testthat      2.2.0      2019-07-22 [1]
 #>  tibble      * 2.1.3      2019-06-06 [1]
@@
 #>  tidyselect    0.2.5      2018-10-11 [1]
-#>  tidyverse   * 1.3.0      2019-11-21 [1]
+#>  tidyverse   * 1.2.1      2017-11-14 [1]
 #>  usethis       1.5.1      2019-07-04 [1]
@@
 #>  withr         2.1.2      2018-03-15 [1]
-#>  xfun          0.11       2019-11-12 [1]
+#>  xfun          0.8        2019-06-25 [1]
 #>  xml2          1.2.2      2019-08-09 [1]
@@
 #>  source                                  
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
 #>  Github (sahirbhatnagar/casebase@e2e73fe)
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.2)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.2)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.2)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  Github (rstudio/rticles@8d56fc6)        
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.2)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.1)                          
 #> 
-#> [1] /Library/Frameworks/R.framework/Versions/3.6/Resources/library
+#> [1] /home/mturgeon/Rlibs
+#> [2] /usr/local/lib/R/site-library
+#> [3] /usr/lib/R/site-library
+#> [4] /usr/lib/R/library
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> Local:    master /Users/jislam/Desktop/Gits/cbpaper
-#> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
-#> Head:     [71a5a9c] 2019-12-06: Lit-Review Edits
+#> Local:    Jesse-Islam-master /home/mturgeon/Documents/git_repositories/cbpaper
+#> Head:     [32c45a7] 2019-12-08: fix table captions
 \end{CodeOutput}

### 2019-12-11 analysis/paper/paper.Rmd

@@
   - name: Maxime Turgeon *
-    affiliation: University of Manitoba
+    affiliation: 'University of Manitoba \AND'
     address: >
-      318 Machray Hall
+      186 Dysart Road
       Winnipeg, MB, Canada R3T 2N2
     email: \email{max.turgeon@umanitoba.ca}
-    url: http://maxturgeon.ca/
+    url: https://maxturgeon.ca/
   - name: Jesse Islam 
@@
 eval_cs2 <- FALSE
-eval_cs3 <- FALSE
+eval_cs3 <- TRUE
+eval_cs4 <- TRUE
 ```
@@
 From the fitted hazard function, cumulative incidence and, thus, risk functions of time, treatment and profile can be easily derived.
+
+# Theoretical details
+
+As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to discretize the study base into an infinite amount of *person moments*.
+These person moments are indexed by both an individual in the study and a time point, and therefore each person moment has a covariate profile, an exposure status and an outcome status attached to it.
+We note that there is only a finite number of person moments associated with the event of interest (what Hanley & Miettinen call the *case series*).
+The case-base sampling refers to the sampling from the base of a representative finite sample called the *base series*.
+
+<!-- Popular survival analysis methods, like Kaplan-Meier and Cox regression, rely on the notion of risk-set sampling. Case-base sampling can be seen as an alternative. -->
+
+As shown by Saarela & Arjas [-@saarela2015non] (and further expanded in Saarela [-@saarela2016case]), writing the likelihood arising from this data-generating mechanism using the framework of non-homogeneous Poisson processes, we eventually reach an expression where each person-moment's contribution is of the form
+$$\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},$$
+where $N(t)$ is the counting process associated with the event of interest, $h(t)$ is the corresponding hazard function, and $\rho(t)$ is the hazard function for the Poisson process associated with case-base sampling.
+This parametric form suggests that we can readily estimate log-hazards of the form $\log(h(t)) = g(t; X)$ using logistic regression, where each observation corresponds to a person moment, the function $g(t; X)$ is linear in a finite number of parameters, and where we treat $-\log(\rho(t))$ as an offset.
+
+In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base sampling *uniformly*, i.e.
+to sample the base series uniformly from the study base.
+In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base.
+More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism  are proportional to the cardiovascular disease event rate given by the Framingham score.
+
+The `casebase` package fits the family of hazard functions of the form
+
+$$ h(t;X) = \exp[g(t;X)] $$
+where $t$ denotes time and $X$, the individual's covariate profile.
+Different functions of $t$ lead to different parametric hazard models.
+The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$.
+
+$$ h(t;X) = \exp(\beta_0 + \beta_1 X) $$
+
+The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
+This is also known as the *memoryless property* (Kalbfleisch and Prentice, 2002).
+
+The Gompertz hazard model is given by including a linear term for time:
+
+$$ h(t;X)  = \exp(\beta_0 + \beta_1 t + \beta_2 X) $$
+
+Use of $\log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
+
+$$ h(t;X)  = \exp(\beta_0 + \beta_1 \log(t) + \beta_2 X) $$
+
+For competing-risk analyses with $J$ possible events, we can show that each person-moment's contribution of the likelihood is of the form
+
+$$\frac{h_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^Jh_j(t)},$$
+
+where $N_j(t)$ is the counting process associated with the event of type $j$ and $h_j(t)$ is the corresponding hazard function.
+As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1\end{align*}}
+
+# Existing packages
+
+Survival analysis is an important branch of applied statistics and epidemiology.
+Accordingly, there is a vast ecosystem of \code{R} packages implementing different methods.
+In this section, we describe how the functionalities of \pkg{casebase} compare to these packages.
+
+At the time of writing, a cursory examination of CRAN's task view on survival analysis reveals that there are over 250 packages related to survival analysis [-@survTaskView].
+For the purposes of this article, we restricted our description to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and cumulative incidence curve estimation.
+By searching for appropriate keywords in the \code{DESCRIPTION} file of these packages, we found 60 relevant packages.
+These 60 packages were then manually examined to determine which ones are comparable to \pkg{casebase}.
+In particular, we excluded packages that were focused on a different set of problems, such as frailty and multi-state models.
+The remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with a description of some of the functionalities they offer.
+
+Several packages implement penalized estimation for the Cox model: \pkg{glmnet} [-@regpathcox], \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche].
+Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: \pkg{CoxRidge} [-@perperoglou], \pkg{rstpm2} [-@clements_liu] and \pkg{survival} [-@survival-package].
+On the other hand, \pkg{casebase} provides penalized estimation in the context of parametric hazards.
+To our knowledge, this is the only package to offer this functionality.
+\textcolor{red}{Is this true?
+Because it seems to be a logical conclusion from your paragraph.}
+
+Parametric survival models are implemented in a handful of packages: \pkg{CFC} [-@sharabiani_mahani_2019], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu], \pkg{mets} [-@scheike2014estimating], and \pkg{survival}.
+The types of models they allow vary for each package.
+For example, \pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \pkg{flexsurv} and \pkg{survival} allow users to supply any distribution of their choice.
+Also, \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of time using splines, which allows flexible modeling of the hazard function.
+Moreover, \pkg{flexsurv} has the ability to estimate both scale and shape parameters for a variety of parametric families (see Table \ref{tab:dists}).
+As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear model of covariates (including time).
+Therefore, our package allows the user to model the effect of time using splines, and through interaction terms involving covariates and time, it also allows user to fit time-varying coefficient models.
+However, we do not explicitly model any shape parameter, unlike \pkg{flexsurv}.
+
+Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets} and \pkg{survival} contain implementations for competing risks.
+The differentiating factor between these packages and \pkg{casebase} is that \pkg{casebase} handles competing risks through multiple logistic regression.
+\textcolor{red}{There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}.
+Could you incorporate some of their discussion, with proper citations?
+The biggest omission right now is \pkg{cmprsk}, which we actually use below.}
+
+Finally, several packages include functions to estimate the cumulative incidence function.
+The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard.
+The first category broadly corresponds to parametric survival models, where the full hazard is explictly modeled.
+Using this estimate, the survival function and the cumulative incidence function can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below).
+Packages including this functionality include \pkg{CFC}, \pkg{Flexsurv}, \pkg{mets}, and \pkg{survival}.
+Our package \pkg{casebase} also follows this approach for both single-event and competing-event analyses.
+The second category outlined above broadly corresponds to Cox models.
+These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function.
+This is achieved using semi-parametric estimators (e.g.
+Breslow's estimator).
+Packages that implement this approach include \pkg{RiskRegression}, \pkg{rstpm2}, and \pkg{survival}.
+As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category produces estimates in the form of step-wise functions.
+This was one of the main motivations for introducting case-base sampling in survival analysis.
+
+\begin{table}[ht]
+\centering
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{ccccccccc}
+\hline
+\textbf{Package}        & \vtop{\hbox{\strut \textbf{Competing}}\hbox{\strut \textbf{risks}}}  & \textbf{Non-proportional} & \textbf{Penalization} & \textbf{Splines} & \textbf{Parametric} & \textbf{Semi-parametric} & \vtop{\hbox{\strut \textbf{Interval/left}}\hbox{\strut \textbf{censoring}}} & \vtop{\hbox{\strut \textbf{Absolute}}\hbox{\strut \textbf{risk}}}           \\
+\hline
+\textbf{casebasee} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
+\textbf{CFC} \cite{sharabiani_mahani_2019}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
+\textbf{coxRidge} \cite{perperoglou}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{crrp} \cite{fu}           & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
+\textbf{fastcox} \cite{yi_zou}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{Flexrsurv} \cite{flexrsurv}      &                          & x                         &                       & x                & x                   &                          &                                  & x               \\
+\textbf{Flexsurv} \cite{flexsurv}       & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\
+\textbf{glmnet} \cite{regpathcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{glmpath} \cite{park_hastie}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{mets} \cite{scheike2014estimating}           & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
+\textbf{penalized} \cite{goeman_meijer2019}      &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{RiskRegression} \cite{gerds_blanche} &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
+\textbf{Rstpm2} \cite{clements_liu}         &                          & x                         & x                     & x                & x                   & x                        & x                                & x                         \\
+\textbf{SmoothHazard} \cite{smoothHazard}   &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
+\textbf{Survival} \cite{survival-package}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x
+\end{tabular}%
+}
+\caption{Different features of interest in various survival packages.}
+\label{tab:surv-pkgs}
+\end{table}
 
@@
 
-# Literature review
-To conduct our review on current methods available in survival analysis, a scraper was coded to examine all packages in the survival task view on CRAN [-@survTaskView].
-There was a total of 259 packages, many of which are not relevant to the problems casebase addresses.
-To filter these packages, a scraper was used to search for competing risks, cumulative incidence functions, non- proportional hazards and penalization.
-This reduced our set of 259 packages down to 60.
-From here, we manually examined each package to determine which ones are comparable to casebase, with either a different approach to the same problems or future features we would like to add to casebase.
-In particular, we excluded packages that were focused on a different set of problems, such as frailty and multi-state models, as casebase is currently focused on absolute-risk, competing events and single event analysis.
-
-These packages can belong to three main categories: non-proportional hazards, regularization and competing risks.
-Besides these three main categories, there are extra features of interest as well.
-Table \ref{tab:surv-pkgs} summarizes the findings.
-
-Many packages exist that implement semi-parametric Cox models.
-Of those, \pkg{glmnet} [-@regpathcox]???, \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche] introduce penalization in their model fitting.
-\pkg{CoxRidge} [-@perperoglou], \pkg{rstpm2} [-@clements_liu] and \pkg{survival} [-@survival-package] include penalization, but they implement the extended cox model permitting intuitive non proportional hazard modeling while the previous four do not.
-In terms of penalization, the benefits of \pkg{casebase} comes from its flexibility.
-Linear models have a wide tool set, considerably larger than those made for Cox regression.
-<!--(*reference for number of packages between their task views, regression is spread out across Multivariate, SocialSciences, Robust among others.
-The focus of this task view is on Optimization Infrastructure Packages , General Purpose Continuous Solvers , Mathematical Programming Solvers , and Specific Applications in Optimization , or Multi Objective Optimization*)--> \pkg{casebase} permits many regularization tools developed for linear regression to be repurposed within a survival context.
-
-Parametric models permit non-proportional hazards and are implemented in a handful of packages.
-Of interest are \pkg{CFC} [-@sharabiani_mahani_2019], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu] and \pkg{survival} [-@survival-package].
-\pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas \pkg{flexsurv} and \pkg{survival} contain implementations for users to supply any distribution they would like [-@survival-package], [-@flexsurv].
-\pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} can use splines, permitting a flexible fit over time [-@flexsurv], [-@smoothHazard], [-@scheike2014estimating], [-@clements_liu].
-Though \pkg{casebase} is limited to three distributions at the moment, it also uses splines allowing for distributions that may not be consistent over survival time.
-
-Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets} and \pkg{survival} contain implementations for competing risks [-@sharabiani_mahani_2019], [-@flexsurv], [-@scheike2014estimating], [-@survival-package].
-The differentiating factor between these packages and \pkg{casebase} is that \pkg{casebase} handles competing risks through multiple logistic regression.
-\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval censoring, whereas \pkg{casebase} currently cannot.
-Interval censoring is a simple feature to implement and is a potential improvement to add to \pkg{casebase}.
-
-Of all the packages listed in Table \ref{tab:surv-pkgs}, \pkg{CFC}, \pkg{Flexsurv}, \pkg{mets}, \pkg{RiskRegression}, \pkg{rstpm2} and \pkg{survival} have implementations for cumulative incidence [-@sharabiani_mahani_2019], [-@flexsurv], [-@scheike2014estimating], [-@gerds_blanche], [-@clements_liu], [-@survival-package].
-Since \pkg{casebase} is parametric, the baseline hazard is known and requires integration to retrieve the cumulative incidence.
-This is similar to the approaches of \pkg{CFC} and \pkg{flexsurv} [-@sharabiani_mahani_2019], [-@flexsurv], while the other packages use semi-parametric approaches to retrieve the baseline after fitting [-@scheike2014estimating], [-@gerds_blanche].
-\pkg{rstpm2} and \pkg{survival} use both for their respective parametric and semi-parametric model implementations [-@clements_liu], [-@survival-package].
-
-\begin{table}[ht]
-\centering
-\resizebox{\textwidth}{!}{%
-\begin{tabular}{ccccccccc}
-\hline
-\textbf{Package}        & \vtop{\hbox{\strut \textbf{Competing}}\hbox{\strut \textbf{risks}}}  & \textbf{Non-proportional} & \textbf{Penalization} & \textbf{Splines} & \textbf{Parametric} & \textbf{Semi-parametric} & \vtop{\hbox{\strut \textbf{Interval/left}}\hbox{\strut \textbf{censoring}}} & \vtop{\hbox{\strut \textbf{Absolute}}\hbox{\strut \textbf{risk}}}           \\
-\hline
-\textbf{casebase} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
-\textbf{CFC} \cite{sharabiani_mahani_2019}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
-\textbf{coxRidge} \cite{perperoglou}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{crrp} \cite{fu}           & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
-\textbf{fastcox} \cite{yi_zou}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{Flexrsurv} \cite{flexrsurv}      &                          & x                         &                       & x                & x                   &                          &                                  & x               \\
-\textbf{Flexsurv} \cite{flexsurv}       & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\
-\textbf{glmnet} \cite{regpathcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{glmpath} \cite{park_hastie}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{mets} \cite{scheike2014estimating}           & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
-\textbf{penalized} \cite{goeman_meijer2019}      &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{RiskRegression} \cite{gerds_blanche} &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
-\textbf{Rstpm2} \cite{clements_liu}         &                          & x                         & x                     & x                & x                   & x                        & x                                & x                         \\
-\textbf{SmoothHazard} \cite{smoothHazard}   &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
-\textbf{Survival} \cite{survival-package}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x
-\end{tabular}%
-}
-\caption{Different features of interest in various survival packages.}
-\label{tab:surv-pkgs}
-\end{table}
-
-# Theoretical details
-
-As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to discretize the study base into an infinite amount of *person moments*.
-These person moments are indexed by both an individual in the study and a time point, and therefore each person moment has a covariate profile, an exposure status and an outcome status attached to it.
-We note that there is only a finite number of person moments associated with the event of interest (what Hanley & Miettinen call the *case series*).
-The case-base sampling refers to the sampling from the base of a representative finite sample called the *base series*.
-
-Popular survival analysis methods, like Kaplan-Meier and Cox regression, rely on the notion of risk-set sampling. Case-base sampling can be seen as an alternative.
-
-As shown by Saarela & Arjas [-@saarela2015non] (and further expanded in Saarela [-@saarela2016case]), writing the likelihood arising from this data-generating mechanism using the framework of non-homogenous Poisson processes, we eventually reach an expression where each person-moment's contribution is of the form
-$$\frac{\lambda(t)^{dN(t)}}{\rho(t) + \lambda(t)},$$
-where $N(t)$ is the counting process associated with the event of interest, $\lambda(t)$ is the corresponding hazard function, and $\rho(t)$ is the hazard function for the Poisson process associated with case-base sampling.
-This parametric form suggests that we can readily estimate log-hazards of the form $\log(\lambda(t)) = g(t; X)$ using logistic regression, where each observation corresponds to a person moment, the function $g(t; X)$ is linear in a finite number of parameters, and where we treat $\log(\rho(t))$ as an offset.
-
-In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base samping *uniformly*, i.e.
-to sample the base series uniformly from the study base.
-In terms of Poisson processes, this sampling strategy corresponds to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base.
-More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism  are proportional to the cardiovascular disease event rate given by the Framingham score.
-
-The `casebase` package fits the family of hazard functions of the form
-
-$$ h(x,t) = exp[g(x,t)] $$
-where $t$ denotes the numerical value (number of units) of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any).
-Different functions of $t$ lead to different parametric hazard models.
-
-The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$.
-
-$$ h(x,t) = exp(\beta_0 + \beta_1 x) $$
-
-The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been on study a.k.a the memoryless property (Kalbfleisch and Prentice, 2002).
-
-The Gompertz hazard model is given by including a linear term for time:
-
-$$ h(x,t)  = exp(\beta_0 + \beta_1 t + \beta_2 x) $$
-
-Use of $log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
-
-$$ h(x, t)  = exp(\beta_0 + \beta_1 \log(t) + \beta_2 x) $$
-
 # Implementation details
 
-The functions in the casebase package can be divided into two categories: 1) data visualization, in the form of population-time plots; and 2) data analysis.
-
-We explicitely aimed at being compatible with both \code{data.frame}s and \code{data.table}s. This is evident in some of the coding choices we made, and it is also reflected in our testing units.
+The functions in the casebase package can be divided into two categories: 1) data visualization, in the form of population-time plots; and 2) parametric modeling.
+We explicitly aimed at being compatible with both \code{data.frame}s and \code{data.table}s.
+This is evident in some of the coding choices we made, and it is also reflected in our unit tests.
 
@@
 
-## Data analysis
-
-The data analysis step was separated into three parts: 1) case-base sampling; 2) estimation of the smooth hazard function; and 3) calculation of the risk function.
+## Parametric modeling
+
+The parametric modeling step was separated into three parts:
+
+  1. case-base sampling; 
+  2. estimation of the smooth hazard function; 
+  3. calculation of the risk function.
+
 By separating the sampling and estimation functions, we allowed the possibility of users implementing more complex sampling scheme, as described in Saarela [-@saarela2016case].
@@
 For the case of a single event, it uses the familiar identity
-$$S(t) = \exp\left(-\int_0^t h(u;X) du\right).$$
-The integral is computed using either the \code{stats::integrate} function or Monte-Carlo integration.
-For the case of a competing-event analysis, the event-specific risk is computed using a nested double integral; in this setting, Monte-Carlo integration is faster and more flexible than \code{stats::integrate}.
-This is due to the fact that the computation involves a double loop over the selected time points; Monte-Carlo integration performs this step by using the vectorized \code{rowSums} and \code{colSums} functions.
-
-To decide between a single-event and a competing-event analysis, we created \code{absoluteRisk} as an \code{S3} generic, with methods for both \code{glm} and \code{CompRisk} objects (the latter inherits from \code{vglm} as well).
-The method dispatch system of \proglang{R} then takes care of matching the correct input to the correct methodology, without the user's intervention.
+\begin{equation}\label{eqn:surv}
+S(t) = \exp\left(-\int_0^t h(u;X) du\right).
+\end{equation}
+The integral is computed using either the \code{stats::integrate} function or Monte-Carlo integration. The risk function (or cumulative incidence function) is then defined as
+\begin{equation}\label{eqn:CI}
+CI(t) = 1 - S(t).
+\end{equation}
+
+For the case of a competing-event analysis, the event-specific risk is computed using the following procedure: first, we compute the overall survival function (i.e. for all event types):
+
+$$ S(t) = \exp\left(-\int_0^t H(u;X) du\right),\qquad H(t;X) = \sum_{j=1}^J h_j(t;X).$$
+From this, we can derive the event-specific subdensities:
+
+$$ f_j(t) = h_j(t)S(t).$$
+
+Finally, by integrating these subdensities, we obtain the event-specific cumulative incidence functions:
+
+$$ CI_j(t) = \int_0^t f_j(u)du.$$
+
+We created \code{absoluteRisk} as an \code{S3} generic, with methods for the different types of outputs of \code{fitSmoothHazard}.
+The method dispatch system of \proglang{R} then takes care of matching the correct output to the correct methodology for calculating the cumulative incidence function, without the user's intervention.
+
+In the following sections, we illustrate these functionalities in the context of three case studies.
 
@@
 
+# Case study 4--Stanford Heart Transplant Data
+
+Although the previous case studies provide a broad overview of the capabilities of \pkg{casebase}, they all have two properties in common:
+\begin{itemize}
+  \item All covariates are fixed, i.e. they do not change over time;
+  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
+\end{itemize}
+
+Case-base sampling has been used in the literature to study vaccination safety, where the exposure period was defined as the week following vaccination [@saarela2015case].
+Hence, the main covariate of interest, i.e.
+exposure to the vaccine, was changing over time.
+In this context, case-base sampling offers an efficient alternative to nested case-control designs or self-matching.
+
+The purpose of the next case study is to show how \pkg{casebase} can also be used for survival analysis problems that do not share the properties above.
+To this end, we will use the Stanford Heart Transplant Data [@clark1971cardiac,@crowley1977covariance]; we will use the version available in the \pkg{survival} package.
+
+Recall the setting of this study: patients were admitted to the Stanford program after meeting with their physician and determining that they were unlikely to respond to other forms of treatment.
+After enrollment, the program searches for a suitable donor for the patient, which can take anywhere between a few days to almost a year.
+We are interested in the effect of a heart transplant on survival; therefore, the patient is considered exposed only after the transplant has occurred.
+
+As before, we can look at the population-time plot for a graphical summary of the event incidence.
+As we can see, most events occur early during the follow-up period, and therefore we do not expect the hazard to be constant.
+
+```{r stanford-poptime, eval = eval_cs4}
+library(survival)
+library(casebase)
+
+stanford_popTime <- popTime(jasa, time = "futime", 
+                            event = "fustat")
+plot(stanford_popTime)
+```
+
+Since the exposure is time-dependent, we need to manually define the exposure variable *after* case-base sampling and *before* fitting the hazard function.
+For this reason, we will use the `sampleCaseBase` function directly.
+
+```{r, eval = eval_cs4}
+library(tidyverse)
+library(lubridate)
+
+cb_data <- sampleCaseBase(jasa, time = "futime", 
+                          event = "fustat", ratio = 10)
+```
+
+Next, we will compute the number of days from acceptance into the program to transplant, and we use this variable to determine whether each population-moment is exposed or not.
+
+```{r, eval = eval_cs4}
+# Define exposure variable
+cb_data <- mutate(cb_data,
+                  txtime = time_length(accept.dt %--% tx.date, 
+                                       unit = "days"),
+                  exposure = case_when(
+                    is.na(txtime) ~ 0L,
+                    txtime > futime ~ 0L,
+                    txtime <= futime ~ 1L
+                  ))
+```
+
+Finally, we can fit the hazard using various linear predictors.
+
+```{r, eval = eval_cs4, warning = FALSE}
+library(splines)
+# Fit several models
+fit1 <- fitSmoothHazard(fustat ~ exposure,
+                        data = cb_data, time = "futime")
+fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
+                        data = cb_data, time = "futime")
+fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
+                        data = cb_data, time = "futime")
+fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
+                        data = cb_data, time = "futime")
+```
+
+Note that the fourth model (i.e.
+`fit4`) includes an interaction term between exposure and follow-up time.
+In other words, this model no longer exhibit proportional hazards.
+The evidence of non-proportionality of hazards in the Stanford Heart Transplant data has been widely discussed [@arjas1988graphical].
+
+We can then compare the goodness of fit of these four models using the Akaike Information Criterion (AIC).
+
+```{r, eval = eval_cs4}
+# Compute AIC
+c("Model1" = AIC(fit1),
+  "Model2" = AIC(fit2),
+  "Model3" = AIC(fit3),
+  "Model4" = AIC(fit4))
+```
+
+As we can, the best fit is the fourth model. By visualizing the hazard functions for both exposed and unexposed individuals, we can more clearly see how the hazards are no longer proportional.
+
+```{r stanford-hazard, eval = eval_cs4}
+# Compute hazards---
+# First, create a list of time points for both exposure status
+hazard_data <- expand.grid(exposure = c(0, 1),
+                           futime = seq(0, 1000,
+                                        length.out = 100))
+# Set the offset to zero
+hazard_data$offset <- 0 
+# Use predict to get the fitted values, and exponentiate to 
+# transform to the right scale
+hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
+                                 type = "link"))
+# Add labels for plots
+hazard_data$Status = factor(hazard_data$exposure,
+                            labels = c("NoTrans", "Trans"))
+
+ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
+    geom_line() +
+    theme_minimal() +
+    theme(legend.position = 'top') +
+    ylab('Hazard') + xlab('Follow-up time')
+```
+
+The non-proportionality seems to be more pronounced at the beginning of follow-up than the end.
+Finally, we can turn these estimate of the hazard function into estimates of the cumulative incidence functions.
+
+```{r stanford-risk, eval = eval_cs4}
+# Compute absolute risk curves
+newdata <- data.frame(exposure = c(0, 1))
+absrisk <- absoluteRisk(fit4, newdata = newdata, 
+                        time = seq(0, 1000, length.out = 100))
+
+colnames(absrisk) <- c("Time", "NoTrans", "Trans")
+
+# Rearrange the data
+absrisk <- gather(as.data.frame(absrisk),
+                  "Status", "Risk", -Time)
+
+ggplot(absrisk, aes(Time, Risk, colour = Status)) +
+  geom_line() +
+  theme_minimal() +
+  theme(legend.position = 'top') +
+  expand_limits(y = 0.5) +
+  xlab('Follow-up time') + ylab('Cum. Incidence')
+```
+
+Note that we can easily adapt the code above to the situation where a patient receives a heart transplant at a point in time of interest, for example after 30 days.
+
+```{r stanford-1y-haz, eval = eval_cs4}
+# Compute hazards---
+# First, create a list of time points for both exposure status
+one_yr_haz <- data.frame(futime = seq(0, 365,
+                                      length.out = 100))
+one_yr_haz <- mutate(one_yr_haz,
+                     offset = 0,
+                     exposure = if_else(futime < 30, 0, 1))
+one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
+                                type = "link"))
+ggplot(one_yr_haz, aes(futime, hazard)) +
+  geom_line() +
+  theme_minimal() +
+  theme(legend.position = 'top') +
+  ylab('Hazard') + xlab('Follow-up time') +
+  geom_vline(xintercept = 30, linetype = 'dashed')
+```
+
+We can then compare the 1-year mortality risk without transplant and with transplant at 30 days.
+
+```{r stanford-1y-risk, eval = eval_cs4}
+absoluteRisk(fit4, newdata = data.frame(exposure = 0),
+             time = 365)
+
+# Use the trapezoidal rule to estimate the cumulative hazard
+min_hazard <- min(one_yr_haz$hazard)
+max_hazard <- max(one_yr_haz$hazard)
+increment <- 365/99
+cumhaz_est <- 0.5*increment*(sum(2*one_yr_haz$hazard) - min_hazard - max_hazard)
+1 - exp(-cumhaz_est)
+```
+
+As we can see, the risk estimate at 1-year is about 30% lower if the patient receives a heart transplant at 30 days.
+
 # Discussion
+
+In this article, we presented the \proglang{R} package \pkg{casebase} that provides functions to fit smooth parametric hazards and estimate cumulative incidence functions using case-base sampling.
+We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through three case studies.
 

### 2019-12-11 analysis/paper/paper.tex

@@
   University of Manitoba\\
-  318 Machray Hall Winnipeg, MB, Canada R3T 2N2\\
+  186 Dysart Road Winnipeg, MB, Canada R3T 2N2\\
   E-mail: \email{max.turgeon@umanitoba.ca}\\
-  URL: \url{http://maxturgeon.ca/}\\~\\
+  URL: \url{https://maxturgeon.ca/}\\~\\
       Jesse Islam\\
@@
 profile can be easily derived.
+
+\hypertarget{theoretical-details}{%
+\section{Theoretical details}\label{theoretical-details}}
+
+As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
+key idea behind case-base sampling is to discretize the study base into
+an infinite amount of \emph{person moments}. These person moments are
+indexed by both an individual in the study and a time point, and
+therefore each person moment has a covariate profile, an exposure status
+and an outcome status attached to it. We note that there is only a
+finite number of person moments associated with the event of interest
+(what Hanley \& Miettinen call the \emph{case series}). The case-base
+sampling refers to the sampling from the base of a representative finite
+sample called the \emph{base series}.
+
+As shown by Saarela \& Arjas \citeyearpar{saarela2015non} (and further
+expanded in Saarela \citeyearpar{saarela2016case}), writing the
+likelihood arising from this data-generating mechanism using the
+framework of non-homogeneous Poisson processes, we eventually reach an
+expression where each person-moment's contribution is of the form
+\[\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},\] where \(N(t)\) is the counting
+process associated with the event of interest, \(h(t)\) is the
+corresponding hazard function, and \(\rho(t)\) is the hazard function
+for the Poisson process associated with case-base sampling. This
+parametric form suggests that we can readily estimate log-hazards of the
+form \(\log(h(t)) = g(t; X)\) using logistic regression, where each
+observation corresponds to a person moment, the function \(g(t; X)\) is
+linear in a finite number of parameters, and where we treat
+\(-\log(\rho(t))\) as an offset.
+
+In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
+suggest performing case-base sampling \emph{uniformly}, i.e.~to sample
+the base series uniformly from the study base. In terms of Poisson
+processes, this sampling strategy corresponds essentially to a
+time-homogeneous Poisson process with intensity equal to \(b/B\), where
+\(b\) is the number of sampled observations in the base series, and
+\(B\) is the total population-time for the study base. More complex
+examples are also available; see for example Saarela \& Arjas
+\citeyearpar{saarela2015non}, where the probabilities of the sampling
+mechanism are proportional to the cardiovascular disease event rate
+given by the Framingham score.
+
+The \texttt{casebase} package fits the family of hazard functions of the
+form
+
+\[ h(t;X) = \exp[g(t;X)] \] where \(t\) denotes time and \(X\), the
+individual's covariate profile. Different functions of \(t\) lead to
+different parametric hazard models. The simplest of these models is the
+one-parameter exponential distribution which is obtained by taking the
+hazard function to be constant over the range of \(t\).
+
+\[ h(t;X) = \exp(\beta_0 + \beta_1 X) \]
+
+The instantaneous failure rate is independent of \(t\), so that the
+conditional chance of failure in a time interval of specified length is
+the same regardless of how long the individual has been in the study.
+This is also known as the \emph{memoryless property} (Kalbfleisch and
+Prentice, 2002).
+
+The Gompertz hazard model is given by including a linear term for time:
+
+\[ h(t;X)  = \exp(\beta_0 + \beta_1 t + \beta_2 X) \]
+
+Use of \(\log(t)\) yields the Weibull hazard which allows for a power
+dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
+
+\[ h(t;X)  = \exp(\beta_0 + \beta_1 \log(t) + \beta_2 X) \]
+
+For competing-risk analyses with \(J\) possible events, we can show that
+each person-moment's contribution of the likelihood is of the form
+
+\[\frac{h_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^Jh_j(t)},\]
+
+where \(N_j(t)\) is the counting process associated with the event of
+type \(j\) and \(h_j(t)\) is the corresponding hazard function. As may
+be expected, this functional form is similar to the terms appearing in
+the likelihood function for multinomial
+regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1\end{align*}}
+
+\hypertarget{existing-packages}{%
+\section{Existing packages}\label{existing-packages}}
+
+Survival analysis is an important branch of applied statistics and
+epidemiology. Accordingly, there is a vast ecosystem of \code{R}
+packages implementing different methods. In this section, we describe
+how the functionalities of \pkg{casebase} compare to these packages.
+
+At the time of writing, a cursory examination of CRAN's task view on
+survival analysis reveals that there are over 250 packages related to
+survival analysis \citeyearpar{survTaskView}. For the purposes of this
+article, we restricted our description to packages that implement at
+least one of the following features: parametric modeling,
+non-proportional hazard models, competing risk analysis, penalized
+estimation, and cumulative incidence curve estimation. By searching for
+appropriate keywords in the \code{DESCRIPTION} file of these packages,
+we found 60 relevant packages. These 60 packages were then manually
+examined to determine which ones are comparable to \pkg{casebase}. In
+particular, we excluded packages that were focused on a different set of
+problems, such as frailty and multi-state models. The remaining 14
+packages appear in Table \ref{tab:surv-pkgs}, along with a description
+of some of the functionalities they offer.
+
+Several packages implement penalized estimation for the Cox model:
+\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
+\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
+\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
+packages also include penalized estimation in the context of Cox models
+with time-varying coefficients: \pkg{CoxRidge}
+\citeyearpar{perperoglou}, \pkg{rstpm2} \citeyearpar{clements_liu} and
+\pkg{survival} \citeyearpar{survival-package}. On the other hand,
+\pkg{casebase} provides penalized estimation in the context of
+parametric hazards. To our knowledge, this is the only package to offer
+this functionality.
+\textcolor{red}{Is this true? Because it seems to be a logical conclusion from your paragraph.}
+
+Parametric survival models are implemented in a handful of packages:
+\pkg{CFC} \citeyearpar{sharabiani_mahani_2019}, \pkg{flexsurv}
+\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
+\pkg{rsptm2} \citeyearpar{clements_liu}, \pkg{mets}
+\citeyearpar{scheike2014estimating}, and \pkg{survival}. The types of
+models they allow vary for each package. For example, \pkg{SmoothHazard}
+is limited to Weibull distributions \citeyearpar{smoothHazard}, whereas
+both \pkg{flexsurv} and \pkg{survival} allow users to supply any
+distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard},
+\pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of
+time using splines, which allows flexible modeling of the hazard
+function. Moreover, \pkg{flexsurv} has the ability to estimate both
+scale and shape parameters for a variety of parametric families (see
+Table \ref{tab:dists}). As discussed above, \pkg{casebase} can model any
+parametric family whose log-hazard can be expressed as a linear model of
+covariates (including time). Therefore, our package allows the user to
+model the effect of time using splines, and through interaction terms
+involving covariates and time, it also allows user to fit time-varying
+coefficient models. However, we do not explicitly model any shape
+parameter, unlike \pkg{flexsurv}.
+
+Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
+\pkg{mets} and \pkg{survival} contain implementations for competing
+risks. The differentiating factor between these packages and
+\pkg{casebase} is that \pkg{casebase} handles competing risks through
+multiple logistic regression.
+\textcolor{red}{There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}.
+Could you incorporate some of their discussion, with proper citations?
+The biggest omission right now is \pkg{cmprsk}, which we actually use below.}
+
+Finally, several packages include functions to estimate the cumulative
+incidence function. The corresponding methods generally fall into two
+categories: transformation of the estimated hazard function, and
+semi-parametric estimation of the baseline hazard. The first category
+broadly corresponds to parametric survival models, where the full hazard
+is explictly modeled. Using this estimate, the survival function and the
+cumulative incidence function can be obtained using their functional
+relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below).
+Packages including this functionality include \pkg{CFC}, \pkg{Flexsurv},
+\pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows
+this approach for both single-event and competing-event analyses. The
+second category outlined above broadly corresponds to Cox models. These
+models do not model the full hazard function, and therefore the baseline
+hazard needs to be estimated separately in order to estimate the
+survival function. This is achieved using semi-parametric estimators
+(e.g.~Breslow's estimator). Packages that implement this approach
+include \pkg{RiskRegression}, \pkg{rstpm2}, and \pkg{survival}. As
+mentioned in the introduction, a key distinguishing factor between these
+two approaches is that the first category leads to smooth estimates of
+the cumulative incidence function, whereas the second category produces
+estimates in the form of step-wise functions. This was one of the main
+motivations for introducting case-base sampling in survival analysis.
+
+\begin{table}[ht]
+\centering
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{ccccccccc}
+\hline
+\textbf{Package}        & \vtop{\hbox{\strut \textbf{Competing}}\hbox{\strut \textbf{risks}}}  & \textbf{Non-proportional} & \textbf{Penalization} & \textbf{Splines} & \textbf{Parametric} & \textbf{Semi-parametric} & \vtop{\hbox{\strut \textbf{Interval/left}}\hbox{\strut \textbf{censoring}}} & \vtop{\hbox{\strut \textbf{Absolute}}\hbox{\strut \textbf{risk}}}           \\
+\hline
+\textbf{casebasee} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
+\textbf{CFC} \cite{sharabiani_mahani_2019}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
+\textbf{coxRidge} \cite{perperoglou}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{crrp} \cite{fu}           & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
+\textbf{fastcox} \cite{yi_zou}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{Flexrsurv} \cite{flexrsurv}      &                          & x                         &                       & x                & x                   &                          &                                  & x               \\
+\textbf{Flexsurv} \cite{flexsurv}       & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\
+\textbf{glmnet} \cite{regpathcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{glmpath} \cite{park_hastie}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{mets} \cite{scheike2014estimating}           & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
+\textbf{penalized} \cite{goeman_meijer2019}      &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{RiskRegression} \cite{gerds_blanche} &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
+\textbf{Rstpm2} \cite{clements_liu}         &                          & x                         & x                     & x                & x                   & x                        & x                                & x                         \\
+\textbf{SmoothHazard} \cite{smoothHazard}   &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
+\textbf{Survival} \cite{survival-package}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x
+\end{tabular}%
+}
+\caption{Different features of interest in various survival packages.}
+\label{tab:surv-pkgs}
+\end{table}
 
@@
 
-\hypertarget{literature-review}{%
-\section{Literature review}\label{literature-review}}
-
-To conduct our review on current methods available in survival analysis,
-a scraper was coded to examine all packages in the survival task view on
-CRAN \citeyearpar{survTaskView}. There was a total of 259 packages, many
-of which are not relevant to the problems casebase addresses. To filter
-these packages, a scraper was used to search for competing risks,
-cumulative incidence functions, non- proportional hazards and
-penalization. This reduced our set of 259 packages down to 60. From
-here, we manually examined each package to determine which ones are
-comparable to casebase, with either a different approach to the same
-problems or future features we would like to add to casebase. In
-particular, we excluded packages that were focused on a different set of
-problems, such as frailty and multi-state models, as casebase is
-currently focused on absolute-risk, competing events and single event
-analysis.
-
-These packages can belong to three main categories: non-proportional
-hazards, regularization and competing risks. Besides these three main
-categories, there are extra features of interest as well. Table
-\ref{tab:surv-pkgs} summarizes the findings.
-
-Many packages exist that implement semi-parametric Cox models. Of those,
-\pkg{glmnet} \citeyearpar{regpathcox}???, \pkg{glmpath}
-\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
-\pkg{RiskRegression} \citeyearpar{gerds_blanche} introduce penalization
-in their model fitting. \pkg{CoxRidge} \citeyearpar{perperoglou},
-\pkg{rstpm2} \citeyearpar{clements_liu} and \pkg{survival}
-\citeyearpar{survival-package} include penalization, but they implement
-the extended cox model permitting intuitive non proportional hazard
-modeling while the previous four do not. In terms of penalization, the
-benefits of \pkg{casebase} comes from its flexibility. Linear models
-have a wide tool set, considerably larger than those made for Cox
-regression. \pkg{casebase} permits many regularization tools developed
-for linear regression to be repurposed within a survival context.
-
-Parametric models permit non-proportional hazards and are implemented in
-a handful of packages. Of interest are \pkg{CFC}
-\citeyearpar{sharabiani_mahani_2019}, \pkg{flexsurv}
-\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
-\pkg{rsptm2} \citeyearpar{clements_liu} and \pkg{survival}
-\citeyearpar{survival-package}. \pkg{SmoothHazard} is limited to Weibull
-distributions \citeyearpar{smoothHazard}, whereas \pkg{flexsurv} and
-\pkg{survival} contain implementations for users to supply any
-distribution they would like \citeyearpar{survival-package},
-\citeyearpar{flexsurv}. \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets}
-and \pkg{rstpm2} can use splines, permitting a flexible fit over time
-\citeyearpar{flexsurv}, \citeyearpar{smoothHazard},
-\citeyearpar{scheike2014estimating}, \citeyearpar{clements_liu}. Though
-\pkg{casebase} is limited to three distributions at the moment, it also
-uses splines allowing for distributions that may not be consistent over
-survival time.
-
-Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
-\pkg{mets} and \pkg{survival} contain implementations for competing
-risks \citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
-\citeyearpar{scheike2014estimating}, \citeyearpar{survival-package}. The
-differentiating factor between these packages and \pkg{casebase} is that
-\pkg{casebase} handles competing risks through multiple logistic
-regression. \pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can
-handle interval censoring, whereas \pkg{casebase} currently cannot.
-Interval censoring is a simple feature to implement and is a potential
-improvement to add to \pkg{casebase}.
-
-Of all the packages listed in Table \ref{tab:surv-pkgs}, \pkg{CFC},
-\pkg{Flexsurv}, \pkg{mets}, \pkg{RiskRegression}, \pkg{rstpm2} and
-\pkg{survival} have implementations for cumulative incidence
-\citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
-\citeyearpar{scheike2014estimating}, \citeyearpar{gerds_blanche},
-\citeyearpar{clements_liu}, \citeyearpar{survival-package}. Since
-\pkg{casebase} is parametric, the baseline hazard is known and requires
-integration to retrieve the cumulative incidence. This is similar to the
-approaches of \pkg{CFC} and \pkg{flexsurv}
-\citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv}, while the
-other packages use semi-parametric approaches to retrieve the baseline
-after fitting \citeyearpar{scheike2014estimating},
-\citeyearpar{gerds_blanche}. \pkg{rstpm2} and \pkg{survival} use both
-for their respective parametric and semi-parametric model
-implementations \citeyearpar{clements_liu},
-\citeyearpar{survival-package}.
-
-\begin{table}[ht]
-\centering
-\resizebox{\textwidth}{!}{%
-\begin{tabular}{ccccccccc}
-\hline
-\textbf{Package}        & \vtop{\hbox{\strut \textbf{Competing}}\hbox{\strut \textbf{risks}}}  & \textbf{Non-proportional} & \textbf{Penalization} & \textbf{Splines} & \textbf{Parametric} & \textbf{Semi-parametric} & \vtop{\hbox{\strut \textbf{Interval/left}}\hbox{\strut \textbf{censoring}}} & \vtop{\hbox{\strut \textbf{Absolute}}\hbox{\strut \textbf{risk}}}           \\
-\hline
-\textbf{casebase} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
-\textbf{CFC} \cite{sharabiani_mahani_2019}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
-\textbf{coxRidge} \cite{perperoglou}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{crrp} \cite{fu}           & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
-\textbf{fastcox} \cite{yi_zou}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{Flexrsurv} \cite{flexrsurv}      &                          & x                         &                       & x                & x                   &                          &                                  & x               \\
-\textbf{Flexsurv} \cite{flexsurv}       & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\
-\textbf{glmnet} \cite{regpathcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{glmpath} \cite{park_hastie}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{mets} \cite{scheike2014estimating}           & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
-\textbf{penalized} \cite{goeman_meijer2019}      &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{RiskRegression} \cite{gerds_blanche} &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
-\textbf{Rstpm2} \cite{clements_liu}         &                          & x                         & x                     & x                & x                   & x                        & x                                & x                         \\
-\textbf{SmoothHazard} \cite{smoothHazard}   &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
-\textbf{Survival} \cite{survival-package}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x
-\end{tabular}%
-}
-\caption{Different features of interest in various survival packages.}
-\label{tab:surv-pkgs}
-\end{table}
-
-\hypertarget{theoretical-details}{%
-\section{Theoretical details}\label{theoretical-details}}
-
-As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
-key idea behind case-base sampling is to discretize the study base into
-an infinite amount of \emph{person moments}. These person moments are
-indexed by both an individual in the study and a time point, and
-therefore each person moment has a covariate profile, an exposure status
-and an outcome status attached to it. We note that there is only a
-finite number of person moments associated with the event of interest
-(what Hanley \& Miettinen call the \emph{case series}). The case-base
-sampling refers to the sampling from the base of a representative finite
-sample called the \emph{base series}.
-
-Popular survival analysis methods, like Kaplan-Meier and Cox regression,
-rely on the notion of risk-set sampling. Case-base sampling can be seen
-as an alternative.
-
-As shown by Saarela \& Arjas \citeyearpar{saarela2015non} (and further
-expanded in Saarela \citeyearpar{saarela2016case}), writing the
-likelihood arising from this data-generating mechanism using the
-framework of non-homogenous Poisson processes, we eventually reach an
-expression where each person-moment's contribution is of the form
-\[\frac{\lambda(t)^{dN(t)}}{\rho(t) + \lambda(t)},\] where \(N(t)\) is
-the counting process associated with the event of interest,
-\(\lambda(t)\) is the corresponding hazard function, and \(\rho(t)\) is
-the hazard function for the Poisson process associated with case-base
-sampling. This parametric form suggests that we can readily estimate
-log-hazards of the form \(\log(\lambda(t)) = g(t; X)\) using logistic
-regression, where each observation corresponds to a person moment, the
-function \(g(t; X)\) is linear in a finite number of parameters, and
-where we treat \(\log(\rho(t))\) as an offset.
-
-In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
-suggest performing case-base samping \emph{uniformly}, i.e.~to sample
-the base series uniformly from the study base. In terms of Poisson
-processes, this sampling strategy corresponds to a time-homogeneous
-Poisson process with intensity equal to \(b/B\), where \(b\) is the
-number of sampled observations in the base series, and \(B\) is the
-total population-time for the study base. More complex examples are also
-available; see for example Saarela \& Arjas
-\citeyearpar{saarela2015non}, where the probabilities of the sampling
-mechanism are proportional to the cardiovascular disease event rate
-given by the Framingham score.
-
-The \texttt{casebase} package fits the family of hazard functions of the
-form
-
-\[ h(x,t) = exp[g(x,t)] \] where \(t\) denotes the numerical value
-(number of units) of a point in prognostic/prospective time and \(x\) is
-the realization of the vector \(X\) of variates based on the patient's
-profile and intervention (if any). Different functions of \(t\) lead to
-different parametric hazard models.
-
-The simplest of these models is the one-parameter exponential
-distribution which is obtained by taking the hazard function to be
-constant over the range of \(t\).
-
-\[ h(x,t) = exp(\beta_0 + \beta_1 x) \]
-
-The instantaneous failure rate is independent of \(t\), so that the
-conditional chance of failure in a time interval of specified length is
-the same regardless of how long the individual has been on study a.k.a
-the memoryless property (Kalbfleisch and Prentice, 2002).
-
-The Gompertz hazard model is given by including a linear term for time:
-
-\[ h(x,t)  = exp(\beta_0 + \beta_1 t + \beta_2 x) \]
-
-Use of \(log(t)\) yields the Weibull hazard which allows for a power
-dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
-
-\[ h(x, t)  = exp(\beta_0 + \beta_1 \log(t) + \beta_2 x) \]
-
 \hypertarget{implementation-details}{%
@@
 categories: 1) data visualization, in the form of population-time plots;
-and 2) data analysis.
-
-We explicitely aimed at being compatible with both \code{data.frame}s
-and \code{data.table}s. This is evident in some of the coding choices we
-made, and it is also reflected in our testing units.
+and 2) parametric modeling. We explicitly aimed at being compatible with
+both \code{data.frame}s and \code{data.table}s. This is evident in some
+of the coding choices we made, and it is also reflected in our unit
+tests.
 
@@
 
-\hypertarget{data-analysis}{%
-\subsection{Data analysis}\label{data-analysis}}
-
-The data analysis step was separated into three parts: 1) case-base
-sampling; 2) estimation of the smooth hazard function; and 3)
-calculation of the risk function. By separating the sampling and
-estimation functions, we allowed the possibility of users implementing
-more complex sampling scheme, as described in Saarela
-\citeyearpar{saarela2016case}.
+\hypertarget{parametric-modeling}{%
+\subsection{Parametric modeling}\label{parametric-modeling}}
+
+The parametric modeling step was separated into three parts:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  case-base sampling;
+\item
+  estimation of the smooth hazard function;
+\item
+  calculation of the risk function.
+\end{enumerate}
+
+By separating the sampling and estimation functions, we allowed the
+possibility of users implementing more complex sampling scheme, as
+described in Saarela \citeyearpar{saarela2016case}.
 
@@
 the model fit. For the case of a single event, it uses the familiar
-identity \[S(t) = \exp\left(-\int_0^t h(u;X) du\right).\] The integral
-is computed using either the \code{stats::integrate} function or
-Monte-Carlo integration. For the case of a competing-event analysis, the
-event-specific risk is computed using a nested double integral; in this
-setting, Monte-Carlo integration is faster and more flexible than
-\code{stats::integrate}. This is due to the fact that the computation
-involves a double loop over the selected time points; Monte-Carlo
-integration performs this step by using the vectorized \code{rowSums}
-and \code{colSums} functions.
-
-To decide between a single-event and a competing-event analysis, we
-created \code{absoluteRisk} as an \code{S3} generic, with methods for
-both \code{glm} and \code{CompRisk} objects (the latter inherits from
-\code{vglm} as well). The method dispatch system of \proglang{R} then
-takes care of matching the correct input to the correct methodology,
-without the user's intervention.
+identity \begin{equation}\label{eqn:surv}
+S(t) = \exp\left(-\int_0^t h(u;X) du\right).
+\end{equation} The integral is computed using either the
+\code{stats::integrate} function or Monte-Carlo integration. The risk
+function (or cumulative incidence function) is then defined as
+\begin{equation}\label{eqn:CI}
+CI(t) = 1 - S(t).
+\end{equation}
+
+For the case of a competing-event analysis, the event-specific risk is
+computed using the following procedure: first, we compute the overall
+survival function (i.e.~for all event types):
+
+\[ S(t) = \exp\left(-\int_0^t H(u;X) du\right),\qquad H(t;X) = \sum_{j=1}^J h_j(t;X).\]
+From this, we can derive the event-specific subdensities:
+
+\[ f_j(t) = h_j(t)S(t).\]
+
+Finally, by integrating these subdensities, we obtain the event-specific
+cumulative incidence functions:
+
+\[ CI_j(t) = \int_0^t f_j(u)du.\]
+
+We created \code{absoluteRisk} as an \code{S3} generic, with methods for
+the different types of outputs of \code{fitSmoothHazard}. The method
+dispatch system of \proglang{R} then takes care of matching the correct
+output to the correct methodology for calculating the cumulative
+incidence function, without the user's intervention.
+
+In the following sections, we illustrate these functionalities in the
+context of three case studies.
 
@@
 \begin{CodeInput}
-R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ splines::bs(Follow.Up.Time) + ScrArm, 
+R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
 R>                                     data = ERSPC, 
@@
 \end{CodeInput}
+\begin{figure}
+
+{\centering \includegraphics{../figures/poptimePostImputation-1}
+
+}
+
+\caption{\label{fig:ptPostI} Population-time plot before imputation.
+Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study.}\label{fig:poptimePostImputation}
+\end{figure}
 \end{CodeChunk}
@@
 R> cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
-R>                              formula_time = ~d.time,alpha=1,ratio=100)
+R+                              formula_time = ~d.time,alpha=1,ratio=100)
 \end{CodeInput}
@@
 R> plot(coxglmFit)
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/coxHazAbsolute-1} \end{center}
+
+\begin{CodeInput}
 R> #taking the coefficient estimates for later use
@@
 R> legend("bottom", 
-R>        legend = c( "Casebase (Lasso+linear)","semi-parametric (Cox)+glmnet","KM curve"), 
-R>        col = c("black","red","Blue"),
-R>        lty = c(1, 1, 1), 
-R>        bg = "gray90")
-\end{CodeInput}
-\end{CodeChunk}
+R+        legend = c( "Casebase (Lasso+linear)","semi-parametric (Cox)+glmnet","KM curve"), 
+R+        col = c("black","red","Blue"),
+R+        lty = c(1, 1, 1), 
+R+        bg = "gray90")
+\end{CodeInput}
+\begin{figure}
+
+{\centering \includegraphics{../figures/abSupportComparison-1}
+
+}
+
+\caption{\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue.}\label{fig:abSupportComparison}
+\end{figure}
+\end{CodeChunk}
+
+\hypertarget{case-study-4stanford-heart-transplant-data}{%
+\section{Case study 4--Stanford Heart Transplant
+Data}\label{case-study-4stanford-heart-transplant-data}}
+
+Although the previous case studies provide a broad overview of the
+capabilities of \pkg{casebase}, they all have two properties in common:
+
+\begin{itemize}
+  \item All covariates are fixed, i.e. they do not change over time;
+  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
+\end{itemize}
+
+Case-base sampling has been used in the literature to study vaccination
+safety, where the exposure period was defined as the week following
+vaccination \citep{saarela2015case}. Hence, the main covariate of
+interest, i.e.~exposure to the vaccine, was changing over time. In this
+context, case-base sampling offers an efficient alternative to nested
+case-control designs or self-matching.
+
+The purpose of the next case study is to show how \pkg{casebase} can
+also be used for survival analysis problems that do not share the
+properties above. To this end, we will use the Stanford Heart Transplant
+Data \citep[\citet{crowley1977covariance}]{clark1971cardiac}; we will
+use the version available in the \pkg{survival} package.
+
+Recall the setting of this study: patients were admitted to the Stanford
+program after meeting with their physician and determining that they
+were unlikely to respond to other forms of treatment. After enrollment,
+the program searches for a suitable donor for the patient, which can
+take anywhere between a few days to almost a year. We are interested in
+the effect of a heart transplant on survival; therefore, the patient is
+considered exposed only after the transplant has occurred.
+
+As before, we can look at the population-time plot for a graphical
+summary of the event incidence. As we can see, most events occur early
+during the follow-up period, and therefore we do not expect the hazard
+to be constant.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> library(survival)
+R> library(casebase)
+R> 
+R> stanford_popTime <- popTime(jasa, time = "futime", 
+R+                             event = "fustat")
+R> plot(stanford_popTime)
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/stanford-poptime-1} \end{center}
+
+\end{CodeChunk}
+
+Since the exposure is time-dependent, we need to manually define the
+exposure variable \emph{after} case-base sampling and \emph{before}
+fitting the hazard function. For this reason, we will use the
+\texttt{sampleCaseBase} function directly.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> library(tidyverse)
+R> library(lubridate)
+R> 
+R> cb_data <- sampleCaseBase(jasa, time = "futime", 
+R+                           event = "fustat", ratio = 10)
+\end{CodeInput}
+\end{CodeChunk}
+
+Next, we will compute the number of days from acceptance into the
+program to transplant, and we use this variable to determine whether
+each population-moment is exposed or not.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Define exposure variable
+R> cb_data <- mutate(cb_data,
+R+                   txtime = time_length(accept.dt %--% tx.date, 
+R+                                        unit = "days"),
+R+                   exposure = case_when(
+R+                     is.na(txtime) ~ 0L,
+R+                     txtime > futime ~ 0L,
+R+                     txtime <= futime ~ 1L
+R+                   ))
+\end{CodeInput}
+\end{CodeChunk}
+
+Finally, we can fit the hazard using various linear predictors.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> library(splines)
+R> # Fit several models
+R> fit1 <- fitSmoothHazard(fustat ~ exposure,
+R+                         data = cb_data, time = "futime")
+R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
+R+                         data = cb_data, time = "futime")
+R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
+R+                         data = cb_data, time = "futime")
+R> fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
+R+                         data = cb_data, time = "futime")
+\end{CodeInput}
+\end{CodeChunk}
+
+Note that the fourth model (i.e.~\texttt{fit4}) includes an interaction
+term between exposure and follow-up time. In other words, this model no
+longer exhibit proportional hazards. The evidence of non-proportionality
+of hazards in the Stanford Heart Transplant data has been widely
+discussed \citep{arjas1988graphical}.
+
+We can then compare the goodness of fit of these four models using the
+Akaike Information Criterion (AIC).
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Compute AIC
+R> c("Model1" = AIC(fit1),
+R+   "Model2" = AIC(fit2),
+R+   "Model3" = AIC(fit3),
+R+   "Model4" = AIC(fit4))
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> Model1 Model2 Model3 Model4 
+#>    803    763    731    723
+\end{CodeOutput}
+\end{CodeChunk}
+
+As we can, the best fit is the fourth model. By visualizing the hazard
+functions for both exposed and unexposed individuals, we can more
+clearly see how the hazards are no longer proportional.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Compute hazards---
+R> # First, create a list of time points for both exposure status
+R> hazard_data <- expand.grid(exposure = c(0, 1),
+R+                            futime = seq(0, 1000,
+R+                                         length.out = 100))
+R> # Set the offset to zero
+R> hazard_data$offset <- 0 
+R> # Use predict to get the fitted values, and exponentiate to 
+R> # transform to the right scale
+R> hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
+R+                                  type = "link"))
+R> # Add labels for plots
+R> hazard_data$Status = factor(hazard_data$exposure,
+R+                             labels = c("NoTrans", "Trans"))
+R> 
+R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
+R+     geom_line() +
+R+     theme_minimal() +
+R+     theme(legend.position = 'top') +
+R+     ylab('Hazard') + xlab('Follow-up time')
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/stanford-hazard-1} \end{center}
+
+\end{CodeChunk}
+
+The non-proportionality seems to be more pronounced at the beginning of
+follow-up than the end. Finally, we can turn these estimate of the
+hazard function into estimates of the cumulative incidence functions.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Compute absolute risk curves
+R> newdata <- data.frame(exposure = c(0, 1))
+R> absrisk <- absoluteRisk(fit4, newdata = newdata, 
+R+                         time = seq(0, 1000, length.out = 100))
+R> 
+R> colnames(absrisk) <- c("Time", "NoTrans", "Trans")
+R> 
+R> # Rearrange the data
+R> absrisk <- gather(as.data.frame(absrisk),
+R+                   "Status", "Risk", -Time)
+R>  
+R> ggplot(absrisk, aes(Time, Risk, colour = Status)) +
+R+   geom_line() +
+R+   theme_minimal() +
+R+   theme(legend.position = 'top') +
+R+   expand_limits(y = 0.5) +
+R+   xlab('Follow-up time') + ylab('Cum. Incidence')
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/stanford-risk-1} \end{center}
+
+\end{CodeChunk}
+
+Note that we can easily adapt the code above to the situation where a
+patient receives a heart transplant at a point in time of interest, for
+example after 30 days.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Compute hazards---
+R> # First, create a list of time points for both exposure status
+R> one_yr_haz <- data.frame(futime = seq(0, 365,
+R+                                       length.out = 100))
+R> one_yr_haz <- mutate(one_yr_haz,
+R+                      offset = 0,
+R+                      exposure = if_else(futime < 30, 0, 1))
+R> one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
+R+                                 type = "link"))
+R> ggplot(one_yr_haz, aes(futime, hazard)) +
+R+   geom_line() +
+R+   theme_minimal() +
+R+   theme(legend.position = 'top') +
+R+   ylab('Hazard') + xlab('Follow-up time') +
+R+   geom_vline(xintercept = 30, linetype = 'dashed')
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/stanford-1y-haz-1} \end{center}
+
+\end{CodeChunk}
+
+We can then compare the 1-year mortality risk without transplant and
+with transplant at 30 days.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> absoluteRisk(fit4, newdata = data.frame(exposure = 0),
+R+              time = 365)
+\end{CodeInput}
+
+\begin{CodeOutput}
+#>         
+#> 365 0.28
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> 
+R> # Use the trapezoidal rule to estimate the cumulative hazard
+R> min_hazard <- min(one_yr_haz$hazard)
+R> max_hazard <- max(one_yr_haz$hazard)
+R> increment <- 365/99
+R> cumhaz_est <- 0.5*increment*(sum(2*one_yr_haz$hazard) - min_hazard - max_hazard)
+R> 1 - exp(-cumhaz_est)
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> [1] 0.19
+\end{CodeOutput}
+\end{CodeChunk}
+
+As we can see, the risk estimate at 1-year is about 30\% lower if the
+patient receives a heart transplant at 30 days.
 
@@
 
-This report was generated on 2019-12-11 15:31:05 using the following
+This report was generated on 2019-12-11 15:42:03 using the following
 computational environment and dependencies:
@@
 #>  dplyr       * 0.8.3      2019-07-04 [1]
+#>  ellipsis      0.3.0      2019-09-20 [1]
 #>  evaluate      0.14       2019-05-28 [1]
 #>  fansi         0.4.0      2018-10-05 [1]
+#>  farver        2.0.1      2019-11-13 [1]
 #>  forcats     * 0.4.0      2019-02-17 [1]
@@
 #>  knitr         1.23       2019-05-18 [1]
+#>  labeling      0.3        2014-08-23 [1]
 #>  lattice       0.20-38    2018-11-04 [4]
@@
 #>  lifecycle     0.1.0      2019-08-01 [1]
-#>  lubridate     1.7.4      2018-04-11 [1]
+#>  lubridate   * 1.7.4      2018-04-11 [1]
 #>  magrittr    * 1.5        2014-11-22 [1]
@@
 #>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
 #>  CRAN (R 3.5.2)                          
 #>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
 #>  CRAN (R 3.5.0)                          
@@
 #>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
 #>  CRAN (R 3.5.1)                          
@@
 \begin{CodeOutput}
-#> Local:    Jesse-Islam-master /home/mturgeon/Documents/git_repositories/cbpaper
-#> Head:     [32c45a7] 2019-12-08: fix table captions
+#> Local:    review /home/mturgeon/Documents/git_repositories/cbpaper
+#> Head:     [ff898bb] 2019-12-10: edit litt review
 \end{CodeOutput}

### 2019-12-11 analysis/paper/paper.Rmd

--- 2019-12-11 analysis/paper/paper.Rmd
@@
   \usepackage{graphicx}
+  \usepackage{tabularx}
   \DeclareUnicodeCharacter{2500}{-}
@@
 # set the seed for reproducible output
-set.seed(2)
+set.seed(1234)
 
@@
 eval_implementation <- TRUE
-eval_cs1 <- TRUE
+eval_cs1 <- FALSE
 eval_cs2 <- FALSE
 eval_cs3 <- FALSE
-eval_cs4 <- TRUE
+eval_cs4 <- FALSE
 ```
@@
 By searching for appropriate keywords in the \code{DESCRIPTION} file of these packages, we found 60 relevant packages.
-These 60 packages were further reduced through manual curation to a subset of 14 packages.
-These packages appears in Table \ref{tab:surv-pkgs}, along with some of the functionalities they offer.
-\textcolor{red}{JESSE: It's not clear how you selected this subset, going from 60 to 14 is a big drop.}
+These 60 packages were then manually examined to determine which ones are comparable to \pkg{casebase}.
+In particular, we excluded packages that were focused on a different set of problems, such as frailty and multi-state models.
+The remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with a description of some of the functionalities they offer.
 
@@
 To our knowledge, this is the only package to offer this functionality.
-\textcolor{red}{JESSE: Is this true?
+\textcolor{red}{Is this true?
 Because it seems to be a logical conclusion from your paragraph.}
@@
 The differentiating factor between these packages and \pkg{casebase} is that \pkg{casebase} handles competing risks through multiple logistic regression.
-\textcolor{red}{JESSE: There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}.
+\textcolor{red}{There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}.
 Could you incorporate some of their discussion, with proper citations?
@@
 
-```{r echo = FALSE}
+```{r erspc-95ci, echo = FALSE, eval = eval_cs1}
 CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
@@
 Imputation was conducted using the default methods in the \pkg{mice} package in \proglang{R}, available on CRAN.
-\textcolor{red}{Why was this method used?
-The Vanderbilt University website lists describes another method for imputation that we could have used.} Before imputation, there was a total of 1000 individuals and 35 variables.
+Before imputation, there was a total of 1000 individuals and 35 variables.
 After imputation, we have 474 observations where 67.5% of individuals in the study experience the event of interest.
 The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.
-A description of each variable can be found in Table~\ref{tab:support1} and a breakdown of each categorical variable in Table~\ref{tab:support2}.
+A description of each variable can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}.
 The data was downloaded from the main author's website, and it is also available as part of the \pkg{casebase} package.
@@
 Hospital death as a covariate was removed, as this is directly informative of death.
-The \pkg{glmnet} package on CRAN was used to introduce elastic net, lasso and ridge regressions to \pkg{casebase}.
-To visualize the effect of imputation on incidence density, we can refer to the pre-imputation population-time plot in Figure \ref{fig:ptPreI} and post-imputation population time plot in Figure \ref{fig:ptPostI}.
-The incidence density is consistent between imputed and complete datasets.
+The glmnet package(REF) on CRAN was used to introduce elastic net, lasso and ridge regressions to casebase.
+To visualize the incidence density, we can refer to the post-imputation population-time plot in Figure \ref{fig:ptPostI}.
 
 ```{r supportSetUp, echo = FALSE, eval = eval_cs3}
-completeData <- casebase::support
+completeData=casebase::support
+completeData<-completeData[,-c(3)]
 #keep one individual out, at random, for which we will develop an 
@@
 Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study."}
-popTime_post <- popTime(as.data.frame(workingCompleteData),
+popTime_post <- popTime(as.data.frame(completeData),
                         time = "d.time", event = "death")
@@
 ```{r CBabsolute, echo = TRUE, eval = eval_cs3}
+
 casebaseAbsolute <- absoluteRisk(cb.Model,

### 2019-12-11 analysis/paper/paper.tex

--- 2019-12-11 analysis/paper/paper.tex
@@
 
-\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx} \DeclareUnicodeCharacter{2500}{-}
+\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx} \usepackage{tabularx} \DeclareUnicodeCharacter{2500}{-}
 
@@
 appropriate keywords in the \code{DESCRIPTION} file of these packages,
-we found 60 relevant packages. These 60 packages were further reduced
-through manual curation to a subset of 14 packages. These packages
-appears in Table \ref{tab:surv-pkgs}, along with some of the
-functionalities they offer.
-\textcolor{red}{JESSE: It's not clear how you selected this subset, going from 60 to 14 is a big drop.}
+we found 60 relevant packages. These 60 packages were then manually
+examined to determine which ones are comparable to \pkg{casebase}. In
+particular, we excluded packages that were focused on a different set of
+problems, such as frailty and multi-state models. The remaining 14
+packages appear in Table \ref{tab:surv-pkgs}, along with a description
+of some of the functionalities they offer.
 
@@
 this functionality.
-\textcolor{red}{JESSE: Is this true? Because it seems to be a logical conclusion from your paragraph.}
+\textcolor{red}{Is this true? Because it seems to be a logical conclusion from your paragraph.}
 
@@
 multiple logistic regression.
-\textcolor{red}{JESSE: There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}.
+\textcolor{red}{There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}.
 Could you incorporate some of their discussion, with proper citations?
@@
 R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
-R+                        levels = c(0, 1), 
-R+                        labels = c("Control group", "Screening group"))
+R>                        levels = c(0, 1), 
+R>                        labels = c("Control group", "Screening group"))
 \end{CodeInput}
@@
 \end{CodeInput}
-
-\begin{center}\includegraphics{../figures/plot-erspc-data-1} \end{center}
-
 \end{CodeChunk}
@@
 R> pt_object_strat <- casebase::popTime(ERSPC, 
-R+                                      event = "DeadOfPrCa", 
-R+                                      exposure = "ScrArm")
+R>                                      event = "DeadOfPrCa", 
+R>                                      exposure = "ScrArm")
 R> plot(pt_object_strat)
 \end{CodeInput}
-
-\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-1} \end{center}
-
 \end{CodeChunk}
@@
 \end{CodeInput}
-
-\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-side-1} \end{center}
-
 \end{CodeChunk}
@@
 R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
-R+                              data = ERSPC)
+R>                              data = ERSPC)
 R> summary(cox_model)
 \end{CodeInput}
-
-\begin{CodeOutput}
-#> Call:
-#> survival::coxph(formula = Surv(Follow.Up.Time, DeadOfPrCa) ~ 
-#>     ScrArm, data = ERSPC)
-#> 
-#>   n= 159893, number of events= 540 
-#> 
-#>                         coef exp(coef) se(coef)     z Pr(>|z|)  
-#> ScrArmScreening group -0.222     0.801    0.088 -2.52    0.012 *
-#> ---
-#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-#> 
-#>                       exp(coef) exp(-coef) lower .95 upper .95
-#> ScrArmScreening group     0.801       1.25     0.674     0.952
-#> 
-#> Concordance= 0.519  (se = 0.011 )
-#> Likelihood ratio test= 6.45  on 1 df,   p=0.01
-#> Wald test            = 6.37  on 1 df,   p=0.01
-#> Score (logrank) test = 6.39  on 1 df,   p=0.01
-\end{CodeOutput}
 \end{CodeChunk}
@@
 R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
-R+                        ignore = 99)
+R>                        ignore = 99)
 R> 
 R> plot(survfit(cox_model, newdata = new_data),
-R+      xlab = "Years since Randomization", 
-R+      ylab = "Cumulative Incidence (%)", 
-R+      fun = "event",
-R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R+                     exp(coef(cox_model)), 
-R+                     exp(confint(cox_model))[1], 
-R+                     exp(confint(cox_model))[2]),
-R+      ylim = c(0, 2), yscale = 100)
+R>      xlab = "Years since Randomization", 
+R>      ylab = "Cumulative Incidence (%)", 
+R>      fun = "event",
+R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R>                     exp(coef(cox_model)), 
+R>                     exp(confint(cox_model))[1], 
+R>                     exp(confint(cox_model))[2]),
+R>      ylim = c(0, 2), yscale = 100)
 R> legend("topleft", 
-R+        legend = c("Control group", "Screening group"), 
-R+        col = c("red","blue"),
-R+        lty = c(1, 1), 
-R+        bg = "gray90")
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/erspc-cox-cif-1} \end{center}
-
+R>        legend = c("Control group", "Screening group"), 
+R>        col = c("red","blue"),
+R>        lty = c(1, 1), 
+R>        bg = "gray90")
+\end{CodeInput}
 \end{CodeChunk}
@@
 R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
-R+                                                   data = ERSPC, 
-R+                                                   ratio = 100)
+R>                                                   data = ERSPC, 
+R>                                                   ratio = 100)
 R> 
 R> summary(casebase_exponential)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> 
-#> Call:
-#> glm(formula = formula, family = binomial, data = sampleData)
-#> 
-#> Deviance Residuals: 
-#>    Min      1Q  Median      3Q     Max  
-#> -0.148  -0.148  -0.148  -0.133   3.079  
-#> 
-#> Coefficients:
-#>                       Estimate Std. Error z value            Pr(>|z|)    
-#> (Intercept)            -7.7650     0.0557 -139.44 <0.0000000000000002 ***
-#> ScrArmScreening group  -0.2166     0.0884   -2.45               0.014 *  
-#> ---
-#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-#> 
-#> (Dispersion parameter for binomial family taken to be 1)
-#> 
-#>     Null deviance: 6059.0  on 54539  degrees of freedom
-#> Residual deviance: 6052.9  on 54538  degrees of freedom
-#> AIC: 6057
-#> 
-#> Number of Fisher Scoring iterations: 7
-\end{CodeOutput}
-
-\begin{CodeInput}
 R> exp(coef(casebase_exponential)[2])
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> ScrArmScreening group 
-#>                  0.81
-\end{CodeOutput}
-
-\begin{CodeInput}
 R> exp(confint(casebase_exponential)[2,])
 \end{CodeInput}
-
-\begin{CodeOutput}
-#>  2.5 % 97.5 % 
-#>   0.68   0.96
-\end{CodeOutput}
 \end{CodeChunk}
@@
 R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
-R+                                           time = seq(0,15,0.1), 
-R+                                           newdata = new_data)
+R>                                           time = seq(0,15,0.1), 
+R>                                           newdata = new_data)
 R> 
 R> plot(survfit(cox_model, newdata = new_data),
-R+      xlab = "Years since Randomization", 
-R+      ylab = "Cumulative Incidence (%)", 
-R+      fun = "event",
-R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R+                     exp(coef(cox_model)), 
-R+                     exp(confint(cox_model))[1], 
-R+                     exp(confint(cox_model))[2]),
-R+      ylim = c(0, 2), yscale = 100)
+R>      xlab = "Years since Randomization", 
+R>      ylab = "Cumulative Incidence (%)", 
+R>      fun = "event",
+R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R>                     exp(coef(cox_model)), 
+R>                     exp(confint(cox_model))[1], 
+R>                     exp(confint(cox_model))[2]),
+R>      ylim = c(0, 2), yscale = 100)
 R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
@@
 R> legend("topleft", 
-R+        legend = c("Control group (Cox)","Control group (Casebase)",
-R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R+        col = c("red","red", "blue","blue"),
-R+        lty = c(1, 2, 1, 2), 
-R+        bg = "gray90")
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/erspc-casebase-cif-1} \end{center}
-
+R>        legend = c("Control group (Cox)","Control group (Casebase)",
+R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R>        col = c("red","red", "blue","blue"),
+R>        lty = c(1, 2, 1, 2), 
+R>        bg = "gray90")
+\end{CodeInput}
 \end{CodeChunk}
@@
 R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
-R+                                  data = ERSPC, 
-R+                                  ratio = 100)
+R>                                  data = ERSPC, 
+R>                                  ratio = 100)
 R> 
 R> summary(casebase_time)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> 
-#> Call:
-#> glm(formula = formula, family = binomial, data = sampleData)
-#> 
-#> Deviance Residuals: 
-#>    Min      1Q  Median      3Q     Max  
-#> -0.389  -0.161  -0.122  -0.095   3.464  
-#> 
-#> Coefficients:
-#>                       Estimate Std. Error z value            Pr(>|z|)    
-#> (Intercept)            -9.0257     0.1122  -80.42 <0.0000000000000002 ***
-#> Follow.Up.Time          0.2207     0.0145   15.18 <0.0000000000000002 ***
-#> ScrArmScreening group  -0.2311     0.0886   -2.61              0.0091 ** 
-#> ---
-#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-#> 
-#> (Dispersion parameter for binomial family taken to be 1)
-#> 
-#>     Null deviance: 6059  on 54539  degrees of freedom
-#> Residual deviance: 5815  on 54537  degrees of freedom
-#> AIC: 5821
-#> 
-#> Number of Fisher Scoring iterations: 8
-\end{CodeOutput}
-
-\begin{CodeInput}
 R> exp(coef(casebase_time))
-\end{CodeInput}
-
-\begin{CodeOutput}
-#>           (Intercept)        Follow.Up.Time ScrArmScreening group 
-#>               0.00012               1.24701               0.79370
-\end{CodeOutput}
-
-\begin{CodeInput}
 R> exp(confint(casebase_time))
 \end{CodeInput}
-
-\begin{CodeOutput}
-#>                          2.5 %  97.5 %
-#> (Intercept)           0.000096 0.00015
-#> Follow.Up.Time        1.212086 1.28321
-#> ScrArmScreening group 0.666385 0.94341
-\end{CodeOutput}
 \end{CodeChunk}
@@
 R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
-R+                                           time = seq(0,15,0.1), 
-R+                                           newdata = new_data)
+R>                                           time = seq(0,15,0.1), 
+R>                                           newdata = new_data)
 R> 
 R> plot(survfit(cox_model, newdata = new_data),
-R+      xlab = "Years since Randomization", 
-R+      ylab = "Cumulative Incidence (%)", 
-R+      fun = "event",
-R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R+                     exp(coef(cox_model)), 
-R+                     exp(confint(cox_model))[1], 
-R+                     exp(confint(cox_model))[2]),
-R+      ylim = c(0, 2), yscale = 100)
+R>      xlab = "Years since Randomization", 
+R>      ylab = "Cumulative Incidence (%)", 
+R>      fun = "event",
+R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R>                     exp(coef(cox_model)), 
+R>                     exp(confint(cox_model))[1], 
+R>                     exp(confint(cox_model))[2]),
+R>      ylim = c(0, 2), yscale = 100)
 R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
@@
 R> legend("topleft", 
-R+        legend = c("Control group (Cox)","Control group (Casebase)",
-R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R+        col = c("red","red", "blue","blue"),
-R+        lty = c(1, 2, 1, 2), 
-R+        bg = "gray90")
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/erspc-casebase-weibull-cif-1} \end{center}
-
+R>        legend = c("Control group (Cox)","Control group (Casebase)",
+R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R>        col = c("red","red", "blue","blue"),
+R>        lty = c(1, 2, 1, 2), 
+R>        bg = "gray90")
+\end{CodeInput}
 \end{CodeChunk}
@@
 R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
-R+                                     data = ERSPC, 
-R+                                     ratio = 100)
+R>                                     data = ERSPC, 
+R>                                     ratio = 100)
 R> 
 R> summary(casebase_splines)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> 
-#> Call:
-#> glm(formula = formula, family = binomial, data = sampleData)
-#> 
-#> Deviance Residuals: 
-#>    Min      1Q  Median      3Q     Max  
-#> -0.397  -0.173  -0.136  -0.085   3.797  
-#> 
-#> Coefficients:
-#>                       Estimate Std. Error z value             Pr(>|z|)    
-#> (Intercept)           -10.2787     0.3175  -32.38 < 0.0000000000000002 ***
-#> bs(Follow.Up.Time)1     4.3191     0.8062    5.36        0.00000008456 ***
-#> bs(Follow.Up.Time)2     2.1061     0.4878    4.32        0.00001580161 ***
-#> bs(Follow.Up.Time)3     4.5310     0.7160    6.33        0.00000000025 ***
-#> ScrArmScreening group  -0.2068     0.0886   -2.33                 0.02 *  
-#> ---
-#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-#> 
-#> (Dispersion parameter for binomial family taken to be 1)
-#> 
-#>     Null deviance: 6059.0  on 54539  degrees of freedom
-#> Residual deviance: 5792.2  on 54535  degrees of freedom
-#> AIC: 5802
-#> 
-#> Number of Fisher Scoring iterations: 9
-\end{CodeOutput}
-
-\begin{CodeInput}
 R> exp(coef(casebase_splines))
-\end{CodeInput}
-
-\begin{CodeOutput}
-#>           (Intercept)   bs(Follow.Up.Time)1   bs(Follow.Up.Time)2 
-#>              0.000034             75.119228              8.216108 
-#>   bs(Follow.Up.Time)3 ScrArmScreening group 
-#>             92.847641              0.813203
-\end{CodeOutput}
-
-\begin{CodeInput}
 R> exp(confint(casebase_splines))
 \end{CodeInput}
-
-\begin{CodeOutput}
-#>                           2.5 %     97.5 %
-#> (Intercept)            0.000018   0.000062
-#> bs(Follow.Up.Time)1   15.997506 378.737706
-#> bs(Follow.Up.Time)2    3.256838  22.143147
-#> bs(Follow.Up.Time)3   21.638298 362.098155
-#> ScrArmScreening group  0.682786   0.966562
-\end{CodeOutput}
 \end{CodeChunk}
@@
 R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
-R+                                     time = seq(0,15,0.1), 
-R+                                     newdata = new_data)
+R>                                     time = seq(0,15,0.1), 
+R>                                     newdata = new_data)
 R> 
 R> plot(survfit(cox_model, newdata = new_data),
-R+      xlab = "Years since Randomization", 
-R+      ylab = "Cumulative Incidence (%)", 
-R+      fun = "event",
-R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R+                     exp(coef(cox_model)), 
-R+                     exp(confint(cox_model))[1], 
-R+                     exp(confint(cox_model))[2]),
-R+      ylim = c(0, 2), yscale = 100)
+R>      xlab = "Years since Randomization", 
+R>      ylab = "Cumulative Incidence (%)", 
+R>      fun = "event",
+R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R>                     exp(coef(cox_model)), 
+R>                     exp(confint(cox_model))[1], 
+R>                     exp(confint(cox_model))[2]),
+R>      ylim = c(0, 2), yscale = 100)
 R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
@@
 R> legend("topleft", 
-R+        legend = c("Control group (Cox)","Control group (Casebase)",
-R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R+        col = c("red","red", "blue","blue"),
-R+        lty = c(1, 2, 1, 2), 
-R+        bg = "gray90")
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/erspc-casebase-splines-cif-1} \end{center}
-
+R>        legend = c("Control group (Cox)","Control group (Casebase)",
+R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R>        col = c("red","red", "blue","blue"),
+R>        lty = c(1, 2, 1, 2), 
+R>        bg = "gray90")
+\end{CodeInput}
 \end{CodeChunk}
@@
 
-\begin{CodeChunk}
-
-\begin{tabular}{l|r|l|r}
-\hline
-Model & HR & CI & Width\\
-\hline
-Cox & 0.80 & (0.67, 0.95) & 0.28\\
-\hline
-Exponential & 0.81 & (0.68, 0.96) & 0.28\\
-\hline
-Weibull & 0.79 & (0.67, 0.94) & 0.28\\
-\hline
-Splines & 0.81 & (0.68, 0.97) & 0.28\\
-\hline
-\end{tabular}
-
-\end{CodeChunk}
-
 As noted above, the usual asymptotic results hold for likelihood ratio
@@
 \end{CodeInput}
-
-\begin{CodeOutput}
-#> Analysis of Deviance Table
-#> 
-#> Model 1: DeadOfPrCa ~ ScrArm + offset(offset)
-#> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
-#>   Resid. Df Resid. Dev Df Deviance            Pr(>Chi)    
-#> 1     54538       6053                                    
-#> 2     54535       5792  3      261 <0.0000000000000002 ***
-#> ---
-#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{CodeOutput}
 \end{CodeChunk}
@@
 R> AIC(casebase_exponential)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> [1] 6057
-\end{CodeOutput}
-
-\begin{CodeInput}
 R> AIC(casebase_time)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> [1] 5821
-\end{CodeOutput}
-
-\begin{CodeInput}
 R> AIC(casebase_splines)
 \end{CodeInput}
-
-\begin{CodeOutput}
-#> [1] 5802
-\end{CodeOutput}
 \end{CodeChunk}
@@
 default methods in the \pkg{mice} package in \proglang{R}, available on
-CRAN.
-\textcolor{red}{Why was this method used? The Vanderbilt University website lists describes another method for imputation that we could have used.}
-Before imputation, there was a total of 1000 individuals and 35
+CRAN. Before imputation, there was a total of 1000 individuals and 35
 variables. After imputation, we have 474 observations where 67.5\% of
@@
 while comparing their absolute risk predictions for a new individual. A
-description of each variable can be found in
-Table\textasciitilde{}\ref{tab:support1} and a breakdown of each
-categorical variable in Table\textasciitilde{}\ref{tab:support2}. The
-data was downloaded from the main author's website, and it is also
-available as part of the \pkg{casebase} package.
+description of each variable can be found in Table \ref{tab:support1}
+and a breakdown of each categorical variable in Table
+\ref{tab:support2}. The data was downloaded from the main author's
+website, and it is also available as part of the \pkg{casebase} package.
 
@@
 Hospital death as a covariate was removed, as this is directly
-informative of death. The \pkg{glmnet} package on CRAN was used to
-introduce elastic net, lasso and ridge regressions to \pkg{casebase}. To
-visualize the effect of imputation on incidence density, we can refer to
-the pre-imputation population-time plot in Figure \ref{fig:ptPreI} and
-post-imputation population time plot in Figure \ref{fig:ptPostI}. The
-incidence density is consistent between imputed and complete datasets.
+informative of death. The glmnet package(REF) on CRAN was used to
+introduce elastic net, lasso and ridge regressions to casebase. To
+visualize the incidence density, we can refer to the post-imputation
+population-time plot in Figure \ref{fig:ptPostI}.
 
@@
 \begin{CodeInput}
-R> popTime_post <- popTime(as.data.frame(workingCompleteData),
+R> popTime_post <- popTime(as.data.frame(completeData),
 R>                         time = "d.time", event = "death")
@@
 \begin{CodeInput}
+R> 
 R> casebaseAbsolute <- absoluteRisk(cb.Model,
@@
 R> stanford_popTime <- popTime(jasa, time = "futime", 
-R+                             event = "fustat")
+R>                             event = "fustat")
 R> plot(stanford_popTime)
 \end{CodeInput}
-
-\begin{center}\includegraphics{../figures/stanford-poptime-1} \end{center}
-
 \end{CodeChunk}
@@
 R> cb_data <- sampleCaseBase(jasa, time = "futime", 
-R+                           event = "fustat", ratio = 10)
+R>                           event = "fustat", ratio = 10)
 \end{CodeInput}
@@
 R> cb_data <- mutate(cb_data,
-R+                   txtime = time_length(accept.dt %--% tx.date, 
-R+                                        unit = "days"),
-R+                   exposure = case_when(
-R+                     is.na(txtime) ~ 0L,
-R+                     txtime > futime ~ 0L,
-R+                     txtime <= futime ~ 1L
-R+                   ))
+R>                   txtime = time_length(accept.dt %--% tx.date, 
+R>                                        unit = "days"),
+R>                   exposure = case_when(
+R>                     is.na(txtime) ~ 0L,
+R>                     txtime > futime ~ 0L,
+R>                     txtime <= futime ~ 1L
+R>                   ))
 \end{CodeInput}
@@
 R> fit1 <- fitSmoothHazard(fustat ~ exposure,
-R+                         data = cb_data, time = "futime")
+R>                         data = cb_data, time = "futime")
 R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
-R+                         data = cb_data, time = "futime")
+R>                         data = cb_data, time = "futime")
 R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
-R+                         data = cb_data, time = "futime")
+R>                         data = cb_data, time = "futime")
 R> fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
-R+                         data = cb_data, time = "futime")
+R>                         data = cb_data, time = "futime")
 \end{CodeInput}
@@
 R> c("Model1" = AIC(fit1),
-R+   "Model2" = AIC(fit2),
-R+   "Model3" = AIC(fit3),
-R+   "Model4" = AIC(fit4))
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> Model1 Model2 Model3 Model4 
-#>    810    772    741    722
-\end{CodeOutput}
+R>   "Model2" = AIC(fit2),
+R>   "Model3" = AIC(fit3),
+R>   "Model4" = AIC(fit4))
+\end{CodeInput}
 \end{CodeChunk}
@@
 R> hazard_data <- expand.grid(exposure = c(0, 1),
-R+                            futime = seq(0, 1000,
-R+                                         length.out = 100))
+R>                            futime = seq(0, 1000,
+R>                                         length.out = 100))
 R> # Set the offset to zero
@@
 R> hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
-R+                                  type = "link"))
+R>                                  type = "link"))
 R> # Add labels for plots
 R> hazard_data$Status = factor(hazard_data$exposure,
-R+                             labels = c("NoTrans", "Trans"))
+R>                             labels = c("NoTrans", "Trans"))
 R> 
 R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
-R+     geom_line() +
-R+     theme_minimal() +
-R+     theme(legend.position = 'top') +
-R+     ylab('Hazard') + xlab('Follow-up time')
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/stanford-hazard-1} \end{center}
-
+R>     geom_line() +
+R>     theme_minimal() +
+R>     theme(legend.position = 'top') +
+R>     ylab('Hazard') + xlab('Follow-up time')
+\end{CodeInput}
 \end{CodeChunk}
@@
 R> absrisk <- absoluteRisk(fit4, newdata = newdata, 
-R+                         time = seq(0, 1000, length.out = 100))
+R>                         time = seq(0, 1000, length.out = 100))
 R> 
@@
 R> absrisk <- gather(as.data.frame(absrisk),
-R+                   "Status", "Risk", -Time)
+R>                   "Status", "Risk", -Time)
 R>  
 R> ggplot(absrisk, aes(Time, Risk, colour = Status)) +
-R+   geom_line() +
-R+   theme_minimal() +
-R+   theme(legend.position = 'top') +
-R+   expand_limits(y = 0.5) +
-R+   xlab('Follow-up time') + ylab('Cum. Incidence')
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/stanford-risk-1} \end{center}
-
+R>   geom_line() +
+R>   theme_minimal() +
+R>   theme(legend.position = 'top') +
+R>   expand_limits(y = 0.5) +
+R>   xlab('Follow-up time') + ylab('Cum. Incidence')
+\end{CodeInput}
 \end{CodeChunk}
@@
 R> one_yr_haz <- data.frame(futime = seq(0, 365,
-R+                                       length.out = 100))
+R>                                       length.out = 100))
 R> one_yr_haz <- mutate(one_yr_haz,
-R+                      offset = 0,
-R+                      exposure = if_else(futime < 30, 0, 1))
+R>                      offset = 0,
+R>                      exposure = if_else(futime < 30, 0, 1))
 R> one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
-R+                                 type = "link"))
+R>                                 type = "link"))
 R> ggplot(one_yr_haz, aes(futime, hazard)) +
-R+   geom_line() +
-R+   theme_minimal() +
-R+   theme(legend.position = 'top') +
-R+   ylab('Hazard') + xlab('Follow-up time') +
-R+   geom_vline(xintercept = 30, linetype = 'dashed')
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/stanford-1y-haz-1} \end{center}
-
+R>   geom_line() +
+R>   theme_minimal() +
+R>   theme(legend.position = 'top') +
+R>   ylab('Hazard') + xlab('Follow-up time') +
+R>   geom_vline(xintercept = 30, linetype = 'dashed')
+\end{CodeInput}
 \end{CodeChunk}
@@
 R> absoluteRisk(fit4, newdata = data.frame(exposure = 0),
-R+              time = 365)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#>         
-#> 365 0.25
-\end{CodeOutput}
-
-\begin{CodeInput}
+R>              time = 365)
 R> 
@@
 \end{CodeInput}
-
-\begin{CodeOutput}
-#> [1] 0.2
-\end{CodeOutput}
 \end{CodeChunk}
@@
 
-This report was generated on 2019-12-11 15:19:04 using the following
+This report was generated on 2019-12-11 15:59:22 using the following
 computational environment and dependencies:
@@
 #>  dplyr       * 0.8.3      2019-07-04 [1]
-#>  ellipsis      0.3.0      2019-09-20 [1]
 #>  evaluate      0.14       2019-05-28 [1]
 #>  fansi         0.4.0      2018-10-05 [1]
-#>  farver        2.0.1      2019-11-13 [1]
 #>  forcats     * 0.4.0      2019-02-17 [1]
@@
 #>  knitr         1.23       2019-05-18 [1]
-#>  labeling      0.3        2014-08-23 [1]
 #>  lattice       0.20-38    2018-11-04 [4]
@@
 #>  lifecycle     0.1.0      2019-08-01 [1]
-#>  lubridate   * 1.7.4      2018-04-11 [1]
+#>  lubridate     1.7.4      2018-04-11 [1]
 #>  magrittr    * 1.5        2014-11-22 [1]
-#>  MASS          7.3-51.4   2019-04-26 [4]
 #>  Matrix      * 1.2-18     2019-11-27 [4]
@@
 #>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
 #>  CRAN (R 3.5.2)                          
 #>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
 #>  CRAN (R 3.5.0)                          
@@
 #>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
 #>  CRAN (R 3.5.1)                          
@@
 #>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
 #>  CRAN (R 3.6.1)                          
@@
 \begin{CodeOutput}
-#> Local:    review-glmnet /home/mturgeon/Documents/git_repositories/cbpaper
-#> Head:     [491ef55] 2019-12-11: edit third case study
+#> Local:    review /home/mturgeon/Documents/git_repositories/cbpaper
+#> Head:     [bc801c3] 2019-12-11: Merge branch 'Jesse-Islam-master' into review
 \end{CodeOutput}

### 2019-12-12 analysis/paper/paper.Rmd
review 2nd case study

@@
   \usepackage{tabularx}
+  \usepackage{float}
+  \usepackage{booktabs}
+  \usepackage{makecell}
   \DeclareUnicodeCharacter{2500}{-}
@@
 )
-
+options(kableExtra.latex.load_packages = FALSE)
 library(tidyverse)
@@
 library(glmnet)
+library(kableExtra)
 options(digits = 2, scipen = 999)
@@
 eval_cs1 <- FALSE
-eval_cs2 <- FALSE
+eval_cs2 <- TRUE
 eval_cs3 <- FALSE
@@
   dplyr::select(-LB, -UB) %>% 
-  knitr::kable(format = "latex")
+  knitr::kable(format = "latex", booktabs = TRUE)  %>%
+  kable_styling()
 ```
@@
 ```{r erscp-aic, eval = eval_cs1}
-AIC(casebase_exponential)
-AIC(casebase_time)
-AIC(casebase_splines)
+c("Exp." = AIC(casebase_exponential),
+  "Weib." = AIC(casebase_time),
+  "Splines" = AIC(casebase_splines))
 ```
@@
 
-The next example shows how case-base sampling can also be used in the context of a competing risk analysis.
+In the next case study, we show how case-base sampling can also be used in the context of a competing risk analysis.
 For illustrative purposes, we will use the same data that was used in Scrucca *et al* [-@scrucca2010regression].
@@
 
-```{r bmtcrr-data, eval = eval_cs2, message = FALSE, echo=2}
+```{r bmtcrr-data, eval = eval_cs2, message = FALSE}
 library(casebase)
@@
 The event of interest is relapse, but other competing causes (e.g.
-transplant-related death) were also recorded Several covariates were also captured at baseline: sex, disease type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL and AML, respectively), disease phase at transplant (Relapse, CR1, CR2, CR3), source of stem cells (bone marrow and peripheral blood, coded as BM+PB, or only peripheral blood, coded as PB), and age.
+transplant-related death) were also recorded.
+Several covariates were captured at baseline: sex, disease type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL and AML, respectively), disease phase at transplant (Relapse, CR1, CR2, CR3), source of stem cells (bone marrow and peripheral blood, coded as BM+PB, or only peripheral blood, coded as PB), and age.
 A summary of these baseline characteristics appear in Table \ref{tab:table1bmtcrr}.
@@
 
-```{r echo=FALSE, eval = FALSE}
-table1 <- tibble::tribble(
+```{r echo = FALSE, eval = eval_cs2}
+tibble::tribble(
     ~Variable, ~Description, ~`Statistical summary`,
-    "Sex", "Sex", "M=Male (100)",
+    "\\code{Sex}", "Sex", "M=Male (100)",
     "", "", "F=Female (77)",
-    "D", "Disease", "ALL (73)",
+    "\\code{D}", "Disease", "ALL (73)",
     "", "", "AML (104)",
-    "Phase", "Phase", "CR1 (47)", 
+    "\\code{Phase}", "Phase", "CR1 (47)", 
     "", "", "CR2 (45)", 
@@
     "", "", "Relapse (73)",
-    "Source", "Type of transplant", "BM+PB (21)", 
+    "\\code{Source}", "Type of transplant", "BM+PB (21)", 
     "", "", "PB (156)",
-    "Age", "Age of patient (years)", "4--62", 
+    "\\code{Age}", "Age of patient (years)", "4--62", 
     "", "", "30.47 (13.04)",
-    "Ftime", "Failure time (months)", "0.13--131.77",
+    "\\code{ftime}", "Failure time (months)", "0.13--131.77",
     "", "", "20.28 (30.78)",
-    "Status", "Status indicator", "0=censored (46)", 
+    "\\code{Status}", "Status indicator", "0=censored (46)", 
     "", "", "1=relapse (56)", 
     "", "", "2=competing event (75)"
-)
-
-xtable::xtable(table1) %>%
-    print(include.rownames = FALSE)
-```
-
-<!-- latex table generated in R 3.4.1 by xtable 1.8-2 package -->
-<!-- Sun Sep 17 16:03:22 2017 -->
-\begin{table}[ht]
-\centering
-\begin{tabular}{ccc}
-  \hline
-Variable & Description & Statistical summary \\ 
-  \hline
-Sex & Sex & M=Male (100) \\ 
-   &  & F=Female (77) \\ 
-  D & Disease & ALL (73) \\ 
-   &  & AML (104) \\ 
-  Phase & Phase & CR1 (47) \\ 
-   &  & CR2 (45) \\ 
-   &  & CR3 (12) \\ 
-   &  & Relapse (73) \\ 
-  Source & Type of transplant & BM+PB (21) \\ 
-   &  & PB (156) \\ 
-  Age & Age of patient (years) & 4--62 \\ 
-   &  & 30.47 (13.04) \\ 
-  Ftime & Failure time (months) & 0.13--131.77 \\ 
-   &  & 20.28 (30.78) \\ 
-  Status & Status indicator & 0=censored (46) \\ 
-   &  & 1=relapse (56) \\ 
-   &  & 2=competing event (75) \\ 
-   \hline
-\end{tabular}
-\caption{Baseline characteristics of patients in the stem-cell transplant study.}
-\label{tab:table1bmtcrr}
-\end{table}
-
-In order to try and visualize the incidence density of relapse, we can look at the corresponding population-time plot.
-In Figure \ref{fig:compPop1}, failure times associated with relapse are highlighted on the plot using red points, while Figure \ref{fig:compPop2} provides a similar population-time plot for competing events.
-
-```{r compPop1, fig.cap="Population-time plot for the stem-cell transplant study. The points represent the event of interest (i.e., relapse).", echo=FALSE, eval = eval_cs2}
+) %>% 
+  knitr::kable(format = "latex", booktabs = TRUE,
+               caption = "Baseline characteristics of patients in the stem-cell transplant study.", 
+               label = "table1bmtcrr",
+               escape = FALSE) %>%
+  kable_styling()
+```
+
+First, we can look at a population-time plot to visualize the incidence density of both relapse and the competing events.
+In Figure \ref{fig:compPop}, failure times are highlighted on the plot using red points for the event of interest (panel A) and blue dots for competing events (panel B).
+In both panels, we see evidence of a non-constant hazard function: the density of points is larger at the beginning of follow-up than at the end.
+
+```{r compPop, fig.cap="Population-time plot for the stem-cell transplant study.
+A) The red points represent relapse events.
+B) The blue points represent the competing events.", echo=FALSE, eval = eval_cs2}
+library(cowplot)
 pt_object <- casebase::popTime(bmtcrr, event = "Status", time = "ftime")
-plot(pt_object)
-```
-
-```{r compPop2, fig.cap="Population-time plot for the stem-cell transplant study. The points represent the competing events.", echo=FALSE, eval = eval_cs2}
+
 pt_object_comp <- bmtcrr %>% 
@@
     casebase::popTime(event = "Status", time = "ftime")
-plot(pt_object_comp, point.colour = "blue")
-```
-
-Our main objective is to compute the absolute risk of relapse for a given set of covariates.
-First, we fit a smooth hazard to the data; for the sake of this example, we opted for a linear term for time:
+
+plot_grid(plot(pt_object) + ggtitle("Relapse"), 
+          plot(pt_object_comp, point.colour = "blue") +
+            ggtitle("Competing events"), 
+          labels = c('A', 'B'))
+```
+
+Our main objective is to compute the absolute risk of relapse for a given set of covariates. First, we fit a smooth hazard to the data using a linear term for time:
 
@@
 
-```{r bmtcrr-cox, echo = FALSE, eval = eval_cs2}
+We will compare our hazard ratio estimates to that obtained from a Cox regression. To do so, we need to treat the competing event as censoring.
+
+```{r bmtcrr-cox, echo = TRUE, eval = eval_cs2}
 library(survival)
@@
 
-From the fit object, we can extract both the hazard ratios and their corresponding confidence intervals:
-
-```{r, echo=FALSE, eval = eval_cs2}
+```{r bmtcrr-cis, echo=FALSE, eval = eval_cs2}
 # Table of coefficients----
+library(glue)
 z_value <- qnorm(0.975)
@@
 
-table_cb %<>% as.data.frame %>% 
-    mutate(Covariates = rownames(.)) %>% 
-    mutate(`95% CI` = glue::glue_data(., "({`lower .95`}, {`upper .95`})")) %>% 
+table_cb <- as.data.frame(table_cb) %>%
+  tibble::rownames_to_column("Covariates") %>% 
+    mutate(CI = glue::glue_data(., "({`lower .95`}, {`upper .95`})")) %>% 
     rename(HR = `exp(coef)`) %>% 
-    select(Covariates, HR, `95% CI`)
-
-# table_cox %<>% as.data.frame %>% 
-#     mutate(Covariates = rownames(.)) %>% 
-#     mutate(`CI (Cox)` = glue::glue_data(., "({`lower .95`}, {`upper .95`})")) %>% 
-#     rename(`HR (Cox)` = `exp(coef)`) %>% 
-#     select(Covariates, `HR (Cox)`, `CI (Cox)`)
+    select(Covariates, HR, CI)
+
+table_cox <- as.data.frame(table_cox) %>%
+  tibble::rownames_to_column("Covariates") %>% 
+  dplyr::mutate(CI_Cox = glue::glue_data(., "({`lower .95`}, {`upper .95`})")) %>%
+  dplyr::rename(HR_Cox = `exp(coef)`) %>%
+  dplyr::select(Covariates, HR_Cox, CI_Cox)
 
 table_cb %>% 
-# inner_join(table_cb, table_cox, by = "Covariates") %>%
-    mutate(Covariates = case_when(
+  dplyr::inner_join(table_cox, by = "Covariates") %>%
+  dplyr::mutate(Covariates = dplyr::case_when(
     Covariates == "SexM" ~ "Sex",
@@
     )) %>% 
-    knitr::kable(format = "latex")
-```
-
-As we can see, the only significant hazard ratio is the one associated with the phase of the disease at transplant.
-More precisely, being in relapse at transplant is associated with a hazard ratio of 3.92 when compared to CR1.
-
-Given our estimate of the hazard function, we can compute the absolute risk curve for a fixed covariate profile.
-We performed this computation for a 35 year old woman who received a stem-cell transplant from peripheral blood at relapse.
+  knitr::kable(format = "latex", booktabs = TRUE,
+               col.names = c("Covariates", "HR", "95% CI", "HR", "95% CI"),
+               caption = "ADD CAPTION") %>% 
+  kable_styling() %>% 
+  add_header_above(c(" " = 1, "Case-Base" = 2, "Cox" = 2))
+```
+
+From the fit object, we can extract both the hazard ratios and their corresponding confidence intervals.
+These quantities appear in Table \ref{tab:bmtcrr-cis}.
+As we can see, the only significant hazard ratio identified by case-base sampling is the one associated with the phase of the disease at transplant.
+More precisely, being in relapse at transplant is associated with a hazard ratio of 3.89 when compared to CR1.
+
+Given the estimate of the hazard function obtained using case-base sampling, we can compute the absolute risk curve for a fixed covariate profile.
+We perform this computation for a 35 year old woman who received a stem-cell transplant from peripheral blood at relapse.
 We compared the absolute risk curve for such a woman with acute lymphoblastic leukemia with that for a similar woman with acute myeloblastic leukemia.
-Figure \ref{fig:compAbsrisk} shows these two curves as a function of time.
-This figure also shows the Kaplan-Meier estimate fitted to the two disease groups (ignoring the other covariates).
+We will estimate the curve from 0 to 60 months.
 
@@
                                      levels = levels(bmtcrr[,"Sex"])),
-                      "D" = c("ALL", "AML"),
+                      "D" = c("ALL", "AML"), # Both diseases
                       "Phase" = factor(c("Relapse", "Relapse"), 
@@
 
-```{r fg_risk, eval = FALSE, echo = FALSE}
+We will compare our estimates to that obtained from a corresponding Fine-Gray model [@fine1999proportional].
+The Fine-Gray model is a semiparametric model for the cause-specific *subdistribution*, i.e.
+the function $f_k(t)$ such that
+$$CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),$$
+where $CI_k(t)$ is the cause-specific cumulative indicence.
+The Fine-Gray model allows to directly assess the effect of a covariate on the subdistribution, as opposed to the hazard.
+For the computation, we will use the \pkg{timereg} package [@timereg].
+
+```{r fg_risk, eval = eval_cs2, echo = TRUE}
 library(timereg)
@@
 
-```{r echo = FALSE, fig.cap="\\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups.
+```{r bmtcrr-risk, echo = FALSE, fig.cap="\\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups.
 The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.", eval = eval_cs2}
-risk_cb <- bind_rows(
+library(ggplot2)
+risk_all <- dplyr::bind_rows(
     data.frame(Time = time_points, 
@@
                Disease = "AML",
+               stringsAsFactors = FALSE),
+    data.frame(Time = time_points,
+               Method = "Fine-Gray",
+               Risk = risk_fg$P1[1,],
+               Disease = "ALL",
+               stringsAsFactors = FALSE),
+    data.frame(Time = time_points,
+               Method = "Fine-Gray",
+               Risk = risk_fg$P1[2,],
+               Disease = "AML",
                stringsAsFactors = FALSE)
@@
 
-# risk_fg <- bind_rows(
-#     data.frame(Time = time_points, 
-#                Method = "Fine-Gray", 
-#                Risk = risk_fg$P1[1,], 
-#                Disease = "ALL",
-#                stringsAsFactors = FALSE),
-#     data.frame(Time = time_points, 
-#                Method = "Fine-Gray", 
-#                Risk = risk_fg$P1[2,], 
-#                Disease = "AML",
-#                stringsAsFactors = FALSE)
-#     )
-
-risk_km <- purrr::map_df(c("ALL", "AML"), 
-                         function(disease) {
-    tmp <- bmtcrr %>% 
-        filter(D == disease) %>% 
-        survival::survfit(survival::Surv(ftime,Status == 1) ~ 1, 
-                data = .)
-    data.frame(Time = tmp$time, 
-               Method = "Kaplan-Meier", 
-               Risk = 1 - tmp$surv, 
-               Disease = disease,
-               stringsAsFactors = FALSE) %>%
-        filter(Time <= 60)
-})
-
-risk_all <- bind_rows(
-    risk_cb,
-    risk_km
-    ) %>% arrange(Disease, Risk)
-
 ggplot(risk_all, aes(x = Time, y = Risk, colour = Method)) + 
   # geom_line for smooth curve
-    geom_line(data = filter(risk_all, Method == "Case-base")) + 
+  geom_line(data = dplyr::filter(risk_all, Method == "Case-base")) + 
   # geom_step for step function
-    geom_step(data = filter(risk_all, Method != "Case-base")) + 
+  geom_step(data = dplyr::filter(risk_all, Method != "Case-base")) + 
   facet_grid(Disease ~ .) +
   ylim(c(0,1)) +
-    theme(legend.position = "top") +
+  theme_minimal() +
+  theme(legend.position = 'top') +    
   xlab("Time (in Months)") + ylab("Relapse risk")
 ```
+
+Figure \ref{fig:compAbsrisk} shows the absolute risk curves for both case-base sampling and the Fine-Gray model.
+As we can see, the two approaches agree quite well for acute myeloblastic leukemia; however, there seems to be a difference of about 5% between the two curves for acute lymphoblastic leukemia.
+This difference does not appear to be significant: the curve from case-base sampling is contained within a 95% confidence band around the Fine-Gray absolute risk curve (figure not shown).
 
@@
 The SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital.
-Imputation was conducted using the default methods in the \pkg{mice} package in \proglang{R}, available on CRAN.
+Imputation was conducted using predictive mean matching and using the \pkg{mice} package in \proglang{R}.
 Before imputation, there was a total of 1000 individuals and 35 variables.
 After imputation, we have 474 observations where 67.5% of individuals in the study experience the event of interest.
-The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.
+\textcolor{red}{How come we lose individuals during imputation?} The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.
 A description of each variable can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}.
@@
 Hospital death as a covariate was removed, as this is directly informative of death.
-The glmnet package(REF) on CRAN was used to introduce elastic net, lasso and ridge regressions to casebase.
+The glmnet package on CRAN was used to introduce elastic net, lasso and ridge regressions to casebase.
 To visualize the incidence density, we can refer to the post-imputation population-time plot in Figure \ref{fig:ptPostI}.

### 2019-12-12 analysis/paper/paper.tex
review 2nd case study

@@
 
-\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx} \usepackage{tabularx} \DeclareUnicodeCharacter{2500}{-}
+\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx} \usepackage{tabularx} \usepackage{float} \usepackage{booktabs} \usepackage{makecell} \DeclareUnicodeCharacter{2500}{-}
 
@@
 \begin{CodeInput}
-R> AIC(casebase_exponential)
-R> AIC(casebase_time)
-R> AIC(casebase_splines)
+R> c("Exp." = AIC(casebase_exponential),
+R>   "Weib." = AIC(casebase_time),
+R>   "Splines" = AIC(casebase_splines))
 \end{CodeInput}
@@
 
-The next example shows how case-base sampling can also be used in the
-context of a competing risk analysis. For illustrative purposes, we will
-use the same data that was used in Scrucca \emph{et al}
+In the next case study, we show how case-base sampling can also be used
+in the context of a competing risk analysis. For illustrative purposes,
+we will use the same data that was used in Scrucca \emph{et al}
 \citeyearpar{scrucca2010regression}. The data was downloaded from the
@@
 \begin{CodeInput}
+R> library(casebase)
 R> data(bmtcrr)
@@
 other competing causes (e.g.~transplant-related death) were also
-recorded Several covariates were also captured at baseline: sex, disease
+recorded. Several covariates were captured at baseline: sex, disease
 type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL
@@
 
-\begin{table}[ht]
+\begin{CodeChunk}
+\begin{table}[t]
+
+\caption{\label{tab:table1bmtcrr}Baseline characteristics of patients in the stem-cell transplant study.}
 \centering
-\begin{tabular}{ccc}
-  \hline
-Variable & Description & Statistical summary \\ 
-  \hline
-Sex & Sex & M=Male (100) \\ 
-   &  & F=Female (77) \\ 
-  D & Disease & ALL (73) \\ 
-   &  & AML (104) \\ 
-  Phase & Phase & CR1 (47) \\ 
-   &  & CR2 (45) \\ 
-   &  & CR3 (12) \\ 
-   &  & Relapse (73) \\ 
-  Source & Type of transplant & BM+PB (21) \\ 
-   &  & PB (156) \\ 
-  Age & Age of patient (years) & 4--62 \\ 
-   &  & 30.47 (13.04) \\ 
-  Ftime & Failure time (months) & 0.13--131.77 \\ 
-   &  & 20.28 (30.78) \\ 
-  Status & Status indicator & 0=censored (46) \\ 
-   &  & 1=relapse (56) \\ 
-   &  & 2=competing event (75) \\ 
-   \hline
+\begin{tabular}{lll}
+\toprule
+Variable & Description & Statistical summary\\
+\midrule
+\code{Sex} & Sex & M=Male (100)\\
+ &  & F=Female (77)\\
+\code{D} & Disease & ALL (73)\\
+ &  & AML (104)\\
+\code{Phase} & Phase & CR1 (47)\\
+\addlinespace
+ &  & CR2 (45)\\
+ &  & CR3 (12)\\
+ &  & Relapse (73)\\
+\code{Source} & Type of transplant & BM+PB (21)\\
+ &  & PB (156)\\
+\addlinespace
+\code{Age} & Age of patient (years) & 4--62\\
+ &  & 30.47 (13.04)\\
+\code{ftime} & Failure time (months) & 0.13--131.77\\
+ &  & 20.28 (30.78)\\
+\code{Status} & Status indicator & 0=censored (46)\\
+\addlinespace
+ &  & 1=relapse (56)\\
+ &  & 2=competing event (75)\\
+\bottomrule
 \end{tabular}
-\caption{Baseline characteristics of patients in the stem-cell transplant study.}
-\label{tab:table1bmtcrr}
 \end{table}
 
-In order to try and visualize the incidence density of relapse, we can
-look at the corresponding population-time plot. In Figure
-\ref{fig:compPop1}, failure times associated with relapse are
-highlighted on the plot using red points, while Figure
-\ref{fig:compPop2} provides a similar population-time plot for competing
-events.
+\end{CodeChunk}
+
+First, we can look at a population-time plot to visualize the incidence
+density of both relapse and the competing events. In Figure
+\ref{fig:compPop}, failure times are highlighted on the plot using red
+points for the event of interest (panel A) and blue dots for competing
+events (panel B). In both panels, we see evidence of a non-constant
+hazard function: the density of points is larger at the beginning of
+follow-up than at the end.
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/compPop-1}
+
+}
+
+\caption[Population-time plot for the stem-cell transplant study]{Population-time plot for the stem-cell transplant study.
+A) The red points represent relapse events.
+B) The blue points represent the competing events.}\label{fig:compPop}
+\end{figure}
+\end{CodeChunk}
 
 Our main objective is to compute the absolute risk of relapse for a
-given set of covariates. First, we fit a smooth hazard to the data; for
-the sake of this example, we opted for a linear term for time:
+given set of covariates. First, we fit a smooth hazard to the data using
+a linear term for time:
 
@@
 R> model_cb <- fitSmoothHazard(
-R>     Status ~ ftime + Sex + D + Phase + Source + Age, 
-R>     data = bmtcrr, 
-R>     ratio = 100, 
-R>     time = "ftime")
-\end{CodeInput}
+R+     Status ~ ftime + Sex + D + Phase + Source + Age, 
+R+     data = bmtcrr, 
+R+     ratio = 100, 
+R+     time = "ftime")
+\end{CodeInput}
+\end{CodeChunk}
+
+We will compare our hazard ratio estimates to that obtained from a Cox
+regression. To do so, we need to treat the competing event as censoring.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> library(survival)
+R> # Treat competing event as censoring
+R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
+R+                    data = bmtcrr)
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{table}[t]
+
+\caption{\label{tab:bmtcrr-cis}ADD CAPTION}
+\centering
+\begin{tabular}{lrlrl}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
+\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
+Covariates & HR & 95\% CI & HR & 95\% CI\\
+\midrule
+Sex & 0.77 & (0.45, 1.35) & 0.68 & (0.39, 1.19)\\
+Disease & 0.58 & (0.32, 1.04) & 0.52 & (0.29, 0.93)\\
+Phase (CR2 vs. CR1) & 1.11 & (0.45, 2.76) & 1.21 & (0.48, 3.02)\\
+Phase (CR3 vs. CR1) & 1.58 & (0.41, 6.04) & 1.67 & (0.43, 6.5)\\
+Phase (Relapse vs. CR1) & 3.89 & (1.81, 8.35) & 4.55 & (2.09, 9.9)\\
+\addlinespace
+Source & 1.48 & (0.49, 4.46) & 1.46 & (0.47, 4.54)\\
+Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
+\bottomrule
+\end{tabular}
+\end{table}
+
 \end{CodeChunk}
@@
 From the fit object, we can extract both the hazard ratios and their
-corresponding confidence intervals:
-
-As we can see, the only significant hazard ratio is the one associated
-with the phase of the disease at transplant. More precisely, being in
-relapse at transplant is associated with a hazard ratio of 3.92 when
-compared to CR1.
-
-Given our estimate of the hazard function, we can compute the absolute
-risk curve for a fixed covariate profile. We performed this computation
-for a 35 year old woman who received a stem-cell transplant from
-peripheral blood at relapse. We compared the absolute risk curve for
-such a woman with acute lymphoblastic leukemia with that for a similar
-woman with acute myeloblastic leukemia. Figure \ref{fig:compAbsrisk}
-shows these two curves as a function of time. This figure also shows the
-Kaplan-Meier estimate fitted to the two disease groups (ignoring the
-other covariates).
+corresponding confidence intervals. These quantities appear in Table
+\ref{tab:bmtcrr-cis}. As we can see, the only significant hazard ratio
+identified by case-base sampling is the one associated with the phase of
+the disease at transplant. More precisely, being in relapse at
+transplant is associated with a hazard ratio of 3.89 when compared to
+CR1.
+
+Given the estimate of the hazard function obtained using case-base
+sampling, we can compute the absolute risk curve for a fixed covariate
+profile. We perform this computation for a 35 year old woman who
+received a stem-cell transplant from peripheral blood at relapse. We
+compared the absolute risk curve for such a woman with acute
+lymphoblastic leukemia with that for a similar woman with acute
+myeloblastic leukemia. We will estimate the curve from 0 to 60 months.
 
@@
 R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
-R>                                      levels = levels(bmtcrr[,"Sex"])),
-R>                       "D" = c("ALL", "AML"),
-R>                       "Phase" = factor(c("Relapse", "Relapse"), 
-R>                                        levels = levels(bmtcrr[,"Phase"])),
-R>                       "Age" = c(35, 35),
-R>                       "Source" = factor(c("PB", "PB"), 
-R>                                         levels = levels(bmtcrr[,"Source"])))
+R+                                      levels = levels(bmtcrr[,"Sex"])),
+R+                       "D" = c("ALL", "AML"), # Both diseases
+R+                       "Phase" = factor(c("Relapse", "Relapse"), 
+R+                                        levels = levels(bmtcrr[,"Phase"])),
+R+                       "Age" = c(35, 35),
+R+                       "Source" = factor(c("PB", "PB"), 
+R+                                         levels = levels(bmtcrr[,"Source"])))
 R> 
@@
 R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
-R>                         method = "numerical", newdata = newdata)
-\end{CodeInput}
-\end{CodeChunk}
+R+                         method = "numerical", newdata = newdata)
+\end{CodeInput}
+\end{CodeChunk}
+
+We will compare our estimates to that obtained from a corresponding
+Fine-Gray model \citep{fine1999proportional}. The Fine-Gray model is a
+semiparametric model for the cause-specific \emph{subdistribution},
+i.e.~the function \(f_k(t)\) such that
+\[CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),\] where
+\(CI_k(t)\) is the cause-specific cumulative indicence. The Fine-Gray
+model allows to directly assess the effect of a covariate on the
+subdistribution, as opposed to the hazard. For the computation, we will
+use the \pkg{timereg} package \citep{timereg}.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> library(timereg)
+R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) + 
+R+                           const(Phase) + const(Source) + const(Age), 
+R+                       data = bmtcrr, cause = 1, model = "fg")
+R> 
+R> # Estimate absolute risk curve
+R> risk_fg <- predict(model_fg, newdata, times = time_points)
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/bmtcrr-risk-1}
+
+}
+
+\caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups.
+The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:bmtcrr-risk}
+\end{figure}
+\end{CodeChunk}
+
+Figure \ref{fig:compAbsrisk} shows the absolute risk curves for both
+case-base sampling and the Fine-Gray model. As we can see, the two
+approaches agree quite well for acute myeloblastic leukemia; however,
+there seems to be a difference of about 5\% between the two curves for
+acute lymphoblastic leukemia. This difference does not appear to be
+significant: the curve from case-base sampling is contained within a
+95\% confidence band around the Fine-Gray absolute risk curve (figure
+not shown).
 
@@
 SUPPORT dataset tracks death for individuals who are considered
-seriously ill within a hospital. Imputation was conducted using the
-default methods in the \pkg{mice} package in \proglang{R}, available on
-CRAN. Before imputation, there was a total of 1000 individuals and 35
-variables. After imputation, we have 474 observations where 67.5\% of
-individuals in the study experience the event of interest. The SUPPORT
-dataset will be used to demonstrate regularization within casebase,
-while comparing their absolute risk predictions for a new individual. A
-description of each variable can be found in Table \ref{tab:support1}
-and a breakdown of each categorical variable in Table
+seriously ill within a hospital. Imputation was conducted using
+predictive mean matching and using the \pkg{mice} package in
+\proglang{R}. Before imputation, there was a total of 1000 individuals
+and 35 variables. After imputation, we have 474 observations where
+67.5\% of individuals in the study experience the event of interest.
+\textcolor{red}{How come we lose individuals during imputation?} The
+SUPPORT dataset will be used to demonstrate regularization within
+casebase, while comparing their absolute risk predictions for a new
+individual. A description of each variable can be found in Table
+\ref{tab:support1} and a breakdown of each categorical variable in Table
 \ref{tab:support2}. The data was downloaded from the main author's
@@
 Hospital death as a covariate was removed, as this is directly
-informative of death. The glmnet package(REF) on CRAN was used to
-introduce elastic net, lasso and ridge regressions to casebase. To
-visualize the incidence density, we can refer to the post-imputation
-population-time plot in Figure \ref{fig:ptPostI}.
+informative of death. The glmnet package on CRAN was used to introduce
+elastic net, lasso and ridge regressions to casebase. To visualize the
+incidence density, we can refer to the post-imputation population-time
+plot in Figure \ref{fig:ptPostI}.
 
@@
 
-This report was generated on 2019-12-11 15:59:22 using the following
+This report was generated on 2019-12-12 13:11:14 using the following
 computational environment and dependencies:
@@
 #>  tz       America/Winnipeg            
-#>  date     2019-12-11                  
+#>  date     2019-12-12                  
 #> 
@@
 #>  colorspace    1.4-1      2019-03-18 [1]
+#>  cowplot     * 1.0.0      2019-07-11 [1]
 #>  crayon        1.3.4      2017-09-16 [1]
@@
 #>  fansi         0.4.0      2018-10-05 [1]
+#>  farver        2.0.1      2019-11-13 [1]
 #>  forcats     * 0.4.0      2019-02-17 [1]
@@
 #>  glmnet      * 2.0-18     2019-05-20 [1]
-#>  glue          1.3.1      2019-03-12 [1]
+#>  glue        * 1.3.1      2019-03-12 [1]
 #>  gtable        0.3.0      2019-03-25 [1]
@@
 #>  jsonlite      1.6        2018-12-07 [1]
+#>  kableExtra  * 1.1.0      2019-03-16 [1]
 #>  knitr         1.23       2019-05-18 [1]
+#>  labeling      0.3        2014-08-23 [1]
 #>  lattice       0.20-38    2018-11-04 [4]
+#>  lava          1.6.5      2019-02-12 [1]
 #>  lazyeval      0.2.2      2019-03-15 [1]
@@
 #>  nlme          3.1-142    2019-11-07 [4]
+#>  numDeriv      2016.8-1   2016-08-27 [1]
 #>  pillar        1.4.2      2019-06-29 [1]
@@
 #>  pkgload       1.0.2      2018-10-29 [1]
+#>  plyr          1.8.5      2019-12-10 [1]
 #>  prettyunits   1.0.2      2015-07-13 [1]
@@
 #>  remotes       2.1.0      2019-06-24 [1]
+#>  reshape2      1.4.3      2017-12-11 [1]
 #>  rlang         0.4.2      2019-11-23 [1]
@@
 #>  tidyverse   * 1.2.1      2017-11-14 [1]
+#>  timereg     * 1.9.4      2019-07-30 [1]
 #>  usethis       1.5.1      2019-07-04 [1]
@@
 #>  VGAM          1.1-2      2019-11-21 [1]
+#>  viridisLite   0.3.0      2018-02-01 [1]
+#>  webshot       0.5.1      2018-09-28 [1]
 #>  withr         2.1.2      2018-03-15 [1]
@@
 #>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
 #>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.1)                          
 #>  CRAN (R 3.6.1)                          
@@
 #>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
 #>  CRAN (R 3.5.2)                          
@@
 #>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.2)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
 #>  CRAN (R 3.6.1)                          
@@
 #>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.5.0)                          
 #>  CRAN (R 3.6.1)                          
@@
 #> Local:    review /home/mturgeon/Documents/git_repositories/cbpaper
-#> Head:     [bc801c3] 2019-12-11: Merge branch 'Jesse-Islam-master' into review
+#> Head:     [c4f42a9] 2019-12-11: Merge branch 'review-glmnet' into review
 \end{CodeOutput}

### 2019-12-12 analysis/paper/paper.Rmd

@@
 eval_cs1 <- FALSE
-eval_cs2 <- TRUE
-eval_cs3 <- FALSE
+eval_cs2 <- FALSE
+eval_cs3 <- TRUE
 eval_cs4 <- FALSE
@@
 For illustrative purposes, we will use the same data that was used in Scrucca *et al* [-@scrucca2010regression].
-The data was downloaded from the main author's website, and it is also available as part of the \pkg{casebase} package.
+The data was downloaded from the first author's website, and it is also available as part of the \pkg{casebase} package.
 
@@
 
-We examined the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [-@knaus1995support].
+In the first two case studies, we described the basic functionalities of the \pkg{casebase} package: creating population-time plots, fitting parametric models for hazard functions, and estimating the corresponding cumulative incidence curves.
+In the next two case studies, we explore more advanced features.
+
+For the third case study, we will look at regularized estimation for the hazard function.
+This can be performed using a combination of case-base sampling and regularized logistic regression.
+To demonstrate this capability, we examine the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [@knaus1995support].
 The SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital.
-Imputation was conducted using predictive mean matching and using the \pkg{mice} package in \proglang{R}.
+Imputation was conducted using predictive mean matching and using the \pkg{mice} package in \proglang{R} [@mice].
 Before imputation, there was a total of 1000 individuals and 35 variables.
-After imputation, we have 474 observations where 67.5% of individuals in the study experience the event of interest.
-\textcolor{red}{How come we lose individuals during imputation?} The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.
-A description of each variable can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}.
-The data was downloaded from the main author's website, and it is also available as part of the \pkg{casebase} package.
+After imputation, we have 474 individuals.
+\textcolor{red}{How come we lose individuals during imputation?} Of these, 67.5% experienced the event of interest.
+<!--The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.--> A description of each variable can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}.
+The original data was obtained from the website of the Department of Biostatistics at Vanderbilt University.
+The imputed dataset is available as part of the \pkg{casebase} package.
 
@@
 
-Hospital death as a covariate was removed, as this is directly informative of death.
-The glmnet package on CRAN was used to introduce elastic net, lasso and ridge regressions to casebase.
-To visualize the incidence density, we can refer to the post-imputation population-time plot in Figure \ref{fig:ptPostI}.
-
-```{r supportSetUp, echo = FALSE, eval = eval_cs3}
-completeData=casebase::support
-completeData<-completeData[,-c(3)]
-#keep one individual out, at random, for which we will develop an 
-#absolute risk curve.
-sam <- sample(1:nrow(completeData), 1)
-#Set workingData to the remaining individuals in the data
-#workingData=completeData[-c(sam),]
-workingCompleteData <- as.matrix(sparse.model.matrix(death~ . - d.time, data = completeData))[,-c(1)]
-#Create u and xCox, which will be used when fitting Cox with glmnet.
-newData <- workingCompleteData[sam,]
-x <- workingCompleteData[-c(sam),]
-#x and y will be used to fit the casebase model under glmnet.
-#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
-y <- data.matrix(completeData[-c(sam), c(34, 4)])
-```
-
-```{r poptimePreImputation, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:ptPreI} Population-time plot before imputation.
-Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study."}
-popTime_pre <- popTime(as.data.frame(support),
+<!--Hospital death as a covariate was removed, as this is directly informative of death.
+The glmnet package on CRAN was used to introduce elastic net, lasso and ridge regressions to casebase.--> As before, the first step is to visualize the incidence density.
+The population-time plot for the SUPPORT data appears in Figure \ref{fig:support-poptime}.
+\textcolor{red}{This population-time plot is strange.
+The red dots never seem to go above 200.}
+
+```{r support-poptime, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for tje SUPPORT data."}
+data(support)
+support_popTime <- popTime(support,
                            time = "d.time", event = "death")
-plot(popTime_pre)
-```
-
-```{r poptimePostImputation, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:ptPostI} Population-time plot before imputation.
-Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study."}
-popTime_post <- popTime(as.data.frame(completeData),
-                        time = "d.time", event = "death")
-plot(popTime_post)
+plot(support_popTime)
+```
+
+Before we proceed with the analysis, we first remove hospital death as a covariate, as this is directly informative of death and cannot be used as a predictor.
+Then we select one observation at random.
+We will estimate the cumulative incidence curve for this individual.
+The other observations will be used to fit the hazard function.
+
+```{r supportSetUp, echo = TRUE, eval = eval_cs3}
+support <- dplyr::select(support, -hospdead)
+
+# Randomly select one observation
+indiv <- sample(1:nrow(support), size = 1)
 ```
@@
 Our main objective is to estimate the cumulative incidence of death for a given covariate profile.
-Given th number of covariates available, we will do this using regularized estimation.
-First, we fit the hazard function; for the sake of simplicity, we opted for a linear term for time.
+Given the number of covariates available, we will do this using regularized estimation.
 As \pkg{casebase} makes use of the \pkg{glmnet} package, we will use the function \code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}.
@@
 For the SUPPORT dataset, all factor variables were converted into dummy variables.
-Here, we used lasso penalization by setting \code{alpha} to 1.
+
+```{r support-design, eval = eval_cs3}
+# Create design matrix without the time variable
+support_matrix <- model.matrix(death ~ . - d.time, 
+                               data = support)
+# Remove intercept term
+support_matrix <- support_matrix[,colnames(support_matrix) != "(Intercept)"]
+
+# Create test and train dataset
+support_test <- support_matrix[indiv,,drop = FALSE]
+
+support_train <- support_matrix[-indiv,]
+y <- as.matrix(support[-indiv, c("death", "d.time")])
+```
+
+We are now ready to estimate the hazard function.
+We use lasso penalization; this is achieved by setting \code{alpha} to 1.
+Also, note that the argument \code{formula_time} can be used to specify the function form for time; in our example, we model time linearly.
 
 ```{r CBhazard, echo = TRUE, eval = eval_cs3}
-cb.Model <- fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", 
-                                event = "death", formula_time = ~ d.time, 
+support_cb <- fitSmoothHazard.fit(support_train, y, 
+                                  formula_time = ~ d.time, 
+                                  family = "glmnet", 
+                                  event = "death", time = "d.time", 
                                   alpha = 1, ratio = 100)
@@
 Then, we take our model object and use the \code{absoluteRisk} function to compute our estimate of the cumulative incidence function.
+Since \code{glmnet} fitted the penalized logistic regression for a range of penalty terms, we need to specify which one we want to use for our prediction.
+We select \code{s = "lambda.1se"}, as recommended.
 
 ```{r CBabsolute, echo = TRUE, eval = eval_cs3}
-
-casebaseAbsolute <- absoluteRisk(cb.Model,
-                                 time = seq(1, max(completeData$d.time), 0.5),
-                                 newdata = t(newData),
+time_points <- seq(1, max(support$d.time), by = 0.5)
+support_risk <- absoluteRisk(support_cb,
+                             time = time_points,
+                             newdata = support_test,
                              s = "lambda.1se", method = "numerical")
@@
 Using the \pkg{glmnet} package, we can fit the Cox regression model using regularized estimation.
+
+```{r coxHaz, echo = TRUE, eval = eval_cs3}
+# Create survival object for glmnet
+surv_glmnet <- survival::Surv(time = y[, "d.time"], 
+                              event = y[, "death"])
+
+support_cox <- glmnet::cv.glmnet(x = support_train, 
+                                 y = surv_glmnet,
+                                 family = "cox", alpha = 1)
+```
+
 However, \pkg{glmnet} does not provide any functionality for estimating the baseline hazard, and the functionalities in the \pkg{survival} package do not work on the output from \code{cv.glmnet}.
-To circumvent this issue and estimate the baseline hazard, we followed the following steps.
+To circumvent this issue and get an estimate of the baseline hazard, we followed the following steps.
 First, we fitted a non-regularized Cox model using only the variables selected by \pkg{glmnet}; the baseline hazard will be estimated using this object.
@@
 
-```{r coxHazAbsolute, echo = TRUE, eval = eval_cs3}
-# Create u and xCox, which will be used when fitting Cox with glmnet.
-u <- survival::Surv(time = as.numeric(y[,2]), 
-                    event = as.factor(y[,1]))
-xCox <- as.matrix(sparse.model.matrix(death~ . - d.time,
-                                      data = completeData))[-c(sam),]
-# hazard for cox using glmnet
-coxglmFit <- glmnet::cv.glmnet(x = xCox, y = u,
-                               family = "cox", alpha = 1)
-# convergence demonstrated in plot
-# plot(coxglmFit)
-#taking the coefficient estimates for later use
-nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
-nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
-# creating a new dataset that only contains the covariates chosen through glmnet.
-cleanCoxData <- as.data.frame(cbind(as.numeric(y[,2]),
-                                    as.numeric(y[,1]),
-                                    xCox[, nonzero_covariate_cox$X1]))
-# fitting a cox model using regular estimation, however we will not keep it.
-# this is used more as an object place holder.
-coxFit <- survival::coxph(Surv(time = V1, event = V2) ~ ., 
-                          data = cleanCoxData)
-# The coefficients of this object will be replaced 
-# with the estimates from coxglmFit.
-coxFit$coefficients <- nonzero_coef_cox@x
+```{r coxAbsolute, echo = TRUE, eval = eval_cs3}
+# Identifying selected covariates
+nonzero_covariates_cox <- predict(support_cox, type = "nonzero", 
+                                  s = "lambda.1se")
+nonzero_coef_cox <- coef(support_cox, s = "lambda.1se")
+# Creating a new dataset that only contains the covariates selected by glmnet
+support_cox_data <- as.data.frame(cbind(y,
+                                        support_train[, nonzero_covariates_cox$X1]))
+# Fitting a cox model
+support_cox <- survival::coxph(Surv(time = d.time, event = death) ~ ., 
+                          data = support_cox_data)
+# Replace the estimate with those obtained by glmnet
+support_cox$coefficients <- nonzero_coef_cox@x
+
 # Fitting CI curve for cox+glmnet
-newDataCox <- newData[nonzero_covariate_cox$X1 - 1]
-
-abCoxFit <- survival::survfit(coxFit, newdata = as.data.frame(t(newDataCox)),
-                              time = seq(0, max(completeData$d.time), 1),
+support_cox_risk <- survival::survfit(support_cox,
+                                      newdata = as.data.frame(support_test),
+                                      time = time_points,
                                       type = "breslow")
@@
 ```{r fitKM+abRisk, echo = TRUE, eval = eval_cs3}
-km <- survival::Surv(time = completeData$d.time, 
-                     event = completeData$death)
+km <- survival::Surv(time = support$d.time, 
+                     event = support$death)
 abKm <- survival::survfit(km ~ 1,
@@
 Both \pkg{glmnet} and \pkg{casebase} have a lower estimated cumulative incidence by the end of the study, when compared to Kaplan-Meier.
-
-```{r abSupportComparison, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
-plot(casebaseAbsolute, type='l', col = "black", lwd = 2, 
+\textcolor{red}{It would be good to have an idea of why the case-base and Coxnet curves are so different.
+At the very least, I think we should compare the select covariates.}
+
+```{r abSupportComparison, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Comparison of cumulative incidence curves between regularized casebase in black, regularized Cox regression in red and Kaplan-Meier in blue."}
+plot(support_risk, type='l', col = "black", lwd = 2, 
      main = "Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves", 
      xlab = "Survival-Time", ylab = "Cumulative Incidence", ylim = c(0,1))
-lines(abCoxFit, col = "red", fun = "event", lwd = 3, conf.int = FALSE)
+lines(support_cox_risk, col = "red", fun = "event", lwd = 3, conf.int = FALSE)
 lines(abKm, col = "blue", fun = "event", lwd = 3, conf.int = FALSE)

### 2019-12-12 analysis/paper/paper.tex

@@
 \citeyearpar{scrucca2010regression}. The data was downloaded from the
-main author's website, and it is also available as part of the
+first author's website, and it is also available as part of the
 \pkg{casebase} package.
@@
 
-\begin{CodeChunk}
-\begin{table}[t]
-
-\caption{\label{tab:table1bmtcrr}Baseline characteristics of patients in the stem-cell transplant study.}
-\centering
-\begin{tabular}{lll}
-\toprule
-Variable & Description & Statistical summary\\
-\midrule
-\code{Sex} & Sex & M=Male (100)\\
- &  & F=Female (77)\\
-\code{D} & Disease & ALL (73)\\
- &  & AML (104)\\
-\code{Phase} & Phase & CR1 (47)\\
-\addlinespace
- &  & CR2 (45)\\
- &  & CR3 (12)\\
- &  & Relapse (73)\\
-\code{Source} & Type of transplant & BM+PB (21)\\
- &  & PB (156)\\
-\addlinespace
-\code{Age} & Age of patient (years) & 4--62\\
- &  & 30.47 (13.04)\\
-\code{ftime} & Failure time (months) & 0.13--131.77\\
- &  & 20.28 (30.78)\\
-\code{Status} & Status indicator & 0=censored (46)\\
-\addlinespace
- &  & 1=relapse (56)\\
- &  & 2=competing event (75)\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\end{CodeChunk}
-
 First, we can look at a population-time plot to visualize the incidence
@@
 
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/compPop-1}
-
-}
-
-\caption[Population-time plot for the stem-cell transplant study]{Population-time plot for the stem-cell transplant study.
-A) The red points represent relapse events.
-B) The blue points represent the competing events.}\label{fig:compPop}
-\end{figure}
-\end{CodeChunk}
-
 Our main objective is to compute the absolute risk of relapse for a
@@
 R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-R+                    data = bmtcrr)
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{table}[t]
-
-\caption{\label{tab:bmtcrr-cis}ADD CAPTION}
-\centering
-\begin{tabular}{lrlrl}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
-Covariates & HR & 95\% CI & HR & 95\% CI\\
-\midrule
-Sex & 0.77 & (0.45, 1.35) & 0.68 & (0.39, 1.19)\\
-Disease & 0.58 & (0.32, 1.04) & 0.52 & (0.29, 0.93)\\
-Phase (CR2 vs. CR1) & 1.11 & (0.45, 2.76) & 1.21 & (0.48, 3.02)\\
-Phase (CR3 vs. CR1) & 1.58 & (0.41, 6.04) & 1.67 & (0.43, 6.5)\\
-Phase (Relapse vs. CR1) & 3.89 & (1.81, 8.35) & 4.55 & (2.09, 9.9)\\
-\addlinespace
-Source & 1.48 & (0.49, 4.46) & 1.46 & (0.47, 4.54)\\
-Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
-\bottomrule
-\end{tabular}
-\end{table}
-
+R>                    data = bmtcrr)
+\end{CodeInput}
 \end{CodeChunk}
@@
 R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
-R+                                      levels = levels(bmtcrr[,"Sex"])),
-R+                       "D" = c("ALL", "AML"), # Both diseases
-R+                       "Phase" = factor(c("Relapse", "Relapse"), 
-R+                                        levels = levels(bmtcrr[,"Phase"])),
-R+                       "Age" = c(35, 35),
-R+                       "Source" = factor(c("PB", "PB"), 
-R+                                         levels = levels(bmtcrr[,"Source"])))
+R>                                      levels = levels(bmtcrr[,"Sex"])),
+R>                       "D" = c("ALL", "AML"), # Both diseases
+R>                       "Phase" = factor(c("Relapse", "Relapse"), 
+R>                                        levels = levels(bmtcrr[,"Phase"])),
+R>                       "Age" = c(35, 35),
+R>                       "Source" = factor(c("PB", "PB"), 
+R>                                         levels = levels(bmtcrr[,"Source"])))
 R> 
@@
 R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
-R+                         method = "numerical", newdata = newdata)
+R>                         method = "numerical", newdata = newdata)
 \end{CodeInput}
@@
 R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) + 
-R+                           const(Phase) + const(Source) + const(Age), 
-R+                       data = bmtcrr, cause = 1, model = "fg")
+R>                           const(Phase) + const(Source) + const(Age), 
+R>                       data = bmtcrr, cause = 1, model = "fg")
 R> 
@@
 \end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/bmtcrr-risk-1}
-
-}
-
-\caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups.
-The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:bmtcrr-risk}
-\end{figure}
 \end{CodeChunk}
@@
 
-We examined the Study to Understand Prognoses Preferences Outcomes and
-Risks of Treatment (SUPPORT) dataset \citeyearpar{knaus1995support}. The
-SUPPORT dataset tracks death for individuals who are considered
-seriously ill within a hospital. Imputation was conducted using
-predictive mean matching and using the \pkg{mice} package in
-\proglang{R}. Before imputation, there was a total of 1000 individuals
-and 35 variables. After imputation, we have 474 observations where
-67.5\% of individuals in the study experience the event of interest.
-\textcolor{red}{How come we lose individuals during imputation?} The
-SUPPORT dataset will be used to demonstrate regularization within
-casebase, while comparing their absolute risk predictions for a new
-individual. A description of each variable can be found in Table
-\ref{tab:support1} and a breakdown of each categorical variable in Table
-\ref{tab:support2}. The data was downloaded from the main author's
-website, and it is also available as part of the \pkg{casebase} package.
+In the first two case studies, we described the basic functionalities of
+the \pkg{casebase} package: creating population-time plots, fitting
+parametric models for hazard functions, and estimating the corresponding
+cumulative incidence curves. In the next two case studies, we explore
+more advanced features.
+
+For the third case study, we will look at regularized estimation for the
+hazard function. This can be performed using a combination of case-base
+sampling and regularized logistic regression. To demonstrate this
+capability, we examine the Study to Understand Prognoses Preferences
+Outcomes and Risks of Treatment (SUPPORT) dataset
+\citep{knaus1995support}. The SUPPORT dataset tracks death for
+individuals who are considered seriously ill within a hospital.
+Imputation was conducted using predictive mean matching and using the
+\pkg{mice} package in \proglang{R} \citep{mice}. Before imputation,
+there was a total of 1000 individuals and 35 variables. After
+imputation, we have 474 individuals.
+\textcolor{red}{How come we lose individuals during imputation?} Of
+these, 67.5\% experienced the event of interest. A description of each
+variable can be found in Table \ref{tab:support1} and a breakdown of
+each categorical variable in Table \ref{tab:support2}. The original data
+was obtained from the website of the Department of Biostatistics at
+Vanderbilt University. The imputed dataset is available as part of the
+\pkg{casebase} package.
 
@@
 
-Hospital death as a covariate was removed, as this is directly
-informative of death. The glmnet package on CRAN was used to introduce
-elastic net, lasso and ridge regressions to casebase. To visualize the
-incidence density, we can refer to the post-imputation population-time
-plot in Figure \ref{fig:ptPostI}.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> popTime_pre <- popTime(as.data.frame(support),
-R>                        time = "d.time", event = "death")
-R> plot(popTime_pre)
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> popTime_post <- popTime(as.data.frame(completeData),
-R>                         time = "d.time", event = "death")
-R> plot(popTime_post)
+As before, the first step is to visualize the incidence density. The
+population-time plot for the SUPPORT data appears in Figure
+\ref{fig:support-poptime}.
+\textcolor{red}{This population-time plot is strange. The red dots never seem to go above 200.}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> data(support)
+R> support_popTime <- popTime(support,
+R+                            time = "d.time", event = "death")
+R> plot(support_popTime)
+\end{CodeInput}
+\begin{figure}
+
+{\centering \includegraphics{../figures/support-poptime-1}
+
+}
+
+\caption{\label{fig:support-poptime} Population-time plot for tje SUPPORT data.}\label{fig:support-poptime}
+\end{figure}
+\end{CodeChunk}
+
+Before we proceed with the analysis, we first remove hospital death as a
+covariate, as this is directly informative of death and cannot be used
+as a predictor. Then we select one observation at random. We will
+estimate the cumulative incidence curve for this individual. The other
+observations will be used to fit the hazard function.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> support <- dplyr::select(support, -hospdead)
+R> 
+R> # Randomly select one observation
+R> indiv <- sample(1:nrow(support), size = 1)
 \end{CodeInput}
@@
 Our main objective is to estimate the cumulative incidence of death for
-a given covariate profile. Given th number of covariates available, we
-will do this using regularized estimation. First, we fit the hazard
-function; for the sake of simplicity, we opted for a linear term for
-time. As \pkg{casebase} makes use of the \pkg{glmnet} package, we will
-use the function \code{fitsmoothhazard.fit}, which follows the syntax
-from \code{glmnet}. Specifically, we input a matrix \code{y}, which
-contains the time and event variables, and a matrix \code{x}, which
-contains all other variables we would like to include in the model. For
-the SUPPORT dataset, all factor variables were converted into dummy
-variables. Here, we used lasso penalization by setting \code{alpha} to
-1.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> cb.Model <- fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", 
-R>                                 event = "death", formula_time = ~ d.time, 
-R>                                 alpha = 1, ratio = 100)
+a given covariate profile. Given the number of covariates available, we
+will do this using regularized estimation. As \pkg{casebase} makes use
+of the \pkg{glmnet} package, we will use the function
+\code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}.
+Specifically, we input a matrix \code{y}, which contains the time and
+event variables, and a matrix \code{x}, which contains all other
+variables we would like to include in the model. For the SUPPORT
+dataset, all factor variables were converted into dummy variables.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Create design matrix without the time variable
+R> support_matrix <- model.matrix(death ~ . - d.time, 
+R+                                data = support)
+R> # Remove intercept term
+R> support_matrix <- support_matrix[,colnames(support_matrix) != "(Intercept)"]
+R> 
+R> # Create test and train dataset
+R> support_test <- support_matrix[indiv,,drop = FALSE]
+R> 
+R> support_train <- support_matrix[-indiv,]
+R> y <- as.matrix(support[-indiv, c("death", "d.time")])
+\end{CodeInput}
+\end{CodeChunk}
+
+We are now ready to estimate the hazard function. We use lasso
+penalization; this is achieved by setting \code{alpha} to 1. Also, note
+that the argument \code{formula_time} can be used to specify the
+function form for time; in our example, we model time linearly.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> support_cb <- fitSmoothHazard.fit(support_train, y, 
+R+                                   formula_time = ~ d.time, 
+R+                                   family = "glmnet", 
+R+                                   event = "death", time = "d.time", 
+R+                                   alpha = 1, ratio = 100)
 \end{CodeInput}
@@
 Then, we take our model object and use the \code{absoluteRisk} function
-to compute our estimate of the cumulative incidence function.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> 
-R> casebaseAbsolute <- absoluteRisk(cb.Model,
-R>                                  time = seq(1, max(completeData$d.time), 0.5),
-R>                                  newdata = t(newData),
-R>                                  s = "lambda.1se", method = "numerical")
+to compute our estimate of the cumulative incidence function. Since
+\code{glmnet} fitted the penalized logistic regression for a range of
+penalty terms, we need to specify which one we want to use for our
+prediction. We select \code{s = "lambda.1se"}, as recommended.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> time_points <- seq(1, max(support$d.time), by = 0.5)
+R> support_risk <- absoluteRisk(support_cb,
+R+                              time = time_points,
+R+                              newdata = support_test,
+R+                              s = "lambda.1se", method = "numerical")
 \end{CodeInput}
@@
 and a Kaplan-Meier survival curve. Using the \pkg{glmnet} package, we
-can fit the Cox regression model using regularized estimation. However,
-\pkg{glmnet} does not provide any functionality for estimating the
-baseline hazard, and the functionalities in the \pkg{survival} package
-do not work on the output from \code{cv.glmnet}. To circumvent this
-issue and estimate the baseline hazard, we followed the following steps.
-First, we fitted a non-regularized Cox model using only the variables
-selected by \pkg{glmnet}; the baseline hazard will be estimated using
-this object. Then, we replaced the estimates of the regression
-coefficients by the estimates from \pkg{glmnet}. Finally, this modified
-\code{coxph} object is fed to the \code{survival::survfit} function to
-obtain the estimated survival function.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Create u and xCox, which will be used when fitting Cox with glmnet.
-R> u <- survival::Surv(time = as.numeric(y[,2]), 
-R>                     event = as.factor(y[,1]))
-R> xCox <- as.matrix(sparse.model.matrix(death~ . - d.time,
-R>                                       data = completeData))[-c(sam),]
-R> # hazard for cox using glmnet
-R> coxglmFit <- glmnet::cv.glmnet(x = xCox, y = u,
-R>                                family = "cox", alpha = 1)
-R> # convergence demonstrated in plot
-R> # plot(coxglmFit)
-R> #taking the coefficient estimates for later use
-R> nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
-R> nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
-R> # creating a new dataset that only contains the covariates chosen through glmnet.
-R> cleanCoxData <- as.data.frame(cbind(as.numeric(y[,2]),
-R>                                     as.numeric(y[,1]),
-R>                                     xCox[, nonzero_covariate_cox$X1]))
-R> # fitting a cox model using regular estimation, however we will not keep it.
-R> # this is used more as an object place holder.
-R> coxFit <- survival::coxph(Surv(time = V1, event = V2) ~ ., 
-R>                           data = cleanCoxData)
-R> # The coefficients of this object will be replaced 
-R> # with the estimates from coxglmFit.
-R> coxFit$coefficients <- nonzero_coef_cox@x
+can fit the Cox regression model using regularized estimation.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Create survival object for glmnet
+R> surv_glmnet <- survival::Surv(time = y[, "d.time"], 
+R+                               event = y[, "death"])
+R> 
+R> support_cox <- glmnet::cv.glmnet(x = support_train, 
+R+                                  y = surv_glmnet,
+R+                                  family = "cox", alpha = 1)
+\end{CodeInput}
+\end{CodeChunk}
+
+However, \pkg{glmnet} does not provide any functionality for estimating
+the baseline hazard, and the functionalities in the \pkg{survival}
+package do not work on the output from \code{cv.glmnet}. To circumvent
+this issue and get an estimate of the baseline hazard, we followed the
+following steps. First, we fitted a non-regularized Cox model using only
+the variables selected by \pkg{glmnet}; the baseline hazard will be
+estimated using this object. Then, we replaced the estimates of the
+regression coefficients by the estimates from \pkg{glmnet}. Finally,
+this modified \code{coxph} object is fed to the \code{survival::survfit}
+function to obtain the estimated survival function.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Identifying selected covariates
+R> nonzero_covariates_cox <- predict(support_cox, type = "nonzero", 
+R+                                   s = "lambda.1se")
+R> nonzero_coef_cox <- coef(support_cox, s = "lambda.1se")
+R> # Creating a new dataset that only contains the covariates selected by glmnet
+R> support_cox_data <- as.data.frame(cbind(y,
+R+                                         support_train[, nonzero_covariates_cox$X1]))
+R> # Fitting a cox model
+R> support_cox <- survival::coxph(Surv(time = d.time, event = death) ~ ., 
+R+                           data = support_cox_data)
+R> # Replace the estimate with those obtained by glmnet
+R> support_cox$coefficients <- nonzero_coef_cox@x
+R> 
 R> # Fitting CI curve for cox+glmnet
-R> newDataCox <- newData[nonzero_covariate_cox$X1 - 1]
-R> 
-R> abCoxFit <- survival::survfit(coxFit, newdata = as.data.frame(t(newDataCox)),
-R>                               time = seq(0, max(completeData$d.time), 1),
-R>                               type = "breslow")
+R> support_cox_risk <- survival::survfit(support_cox,
+R+                                       newdata = as.data.frame(support_test),
+R+                                       time = time_points,
+R+                                       type = "breslow")
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> km <- survival::Surv(time = completeData$d.time, 
-R>                      event = completeData$death)
+R> km <- survival::Surv(time = support$d.time, 
+R+                      event = support$death)
 R> abKm <- survival::survfit(km ~ 1,
-R>                           type = 'kaplan-meier')
+R+                           type = 'kaplan-meier')
 \end{CodeInput}
@@
 the study, when compared to Kaplan-Meier.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> plot(casebaseAbsolute, type='l', col = "black", lwd = 2, 
-R>      main = "Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves", 
-R>      xlab = "Survival-Time", ylab = "Cumulative Incidence", ylim = c(0,1))
-R> lines(abCoxFit, col = "red", fun = "event", lwd = 3, conf.int = FALSE)
-R> lines(abKm, col = "blue", fun = "event", lwd = 3, conf.int = FALSE)
-R> legend("bottom", 
-R>        legend = c("casebase", "Cox+glmnet", "KM curve"), 
-R>        col = c("black", "red", "blue"), lty = c(1, 1, 1), bg = "gray90")
-\end{CodeInput}
+\textcolor{red}{It would be good to have an idea of why the case-base and Coxnet curves are so different. At the very least, I think we should compare the select covariates.}
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/abSupportComparison-1}
+
+}
+
+\caption{\label{fig:abSC} Comparison of cumulative incidence curves between regularized casebase in black, regularized Cox regression in red and Kaplan-Meier in blue.}\label{fig:abSupportComparison}
+\end{figure}
 \end{CodeChunk}
@@
 
-This report was generated on 2019-12-12 13:11:14 using the following
+This report was generated on 2019-12-12 15:16:14 using the following
 computational environment and dependencies:
@@
 #>  colorspace    1.4-1      2019-03-18 [1]
-#>  cowplot     * 1.0.0      2019-07-11 [1]
 #>  crayon        1.3.4      2017-09-16 [1]
@@
 #>  glmnet      * 2.0-18     2019-05-20 [1]
-#>  glue        * 1.3.1      2019-03-12 [1]
+#>  glue          1.3.1      2019-03-12 [1]
 #>  gtable        0.3.0      2019-03-25 [1]
@@
 #>  lattice       0.20-38    2018-11-04 [4]
-#>  lava          1.6.5      2019-02-12 [1]
 #>  lazyeval      0.2.2      2019-03-15 [1]
@@
 #>  nlme          3.1-142    2019-11-07 [4]
-#>  numDeriv      2016.8-1   2016-08-27 [1]
 #>  pillar        1.4.2      2019-06-29 [1]
@@
 #>  pkgload       1.0.2      2018-10-29 [1]
-#>  plyr          1.8.5      2019-12-10 [1]
 #>  prettyunits   1.0.2      2015-07-13 [1]
@@
 #>  remotes       2.1.0      2019-06-24 [1]
-#>  reshape2      1.4.3      2017-12-11 [1]
 #>  rlang         0.4.2      2019-11-23 [1]
@@
 #>  tidyverse   * 1.2.1      2017-11-14 [1]
-#>  timereg     * 1.9.4      2019-07-30 [1]
 #>  usethis       1.5.1      2019-07-04 [1]
@@
 #>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
 #>  CRAN (R 3.5.0)                          
@@
 #>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
 #>  CRAN (R 3.5.2)                          
@@
 #>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
 #>  CRAN (R 3.6.1)                          
@@
 #> Local:    review /home/mturgeon/Documents/git_repositories/cbpaper
-#> Head:     [c4f42a9] 2019-12-11: Merge branch 'review-glmnet' into review
+#> Head:     [c095709] 2019-12-12: review 2nd case study
 \end{CodeOutput}

### 2020-01-16 analysis/paper/paper.Rmd
fixedMaxComments, added questions, averaged absolute risks
I fixed a few of Max's comments, but some I was not certain about. particularly what to do with CFC's own package methods exploration. they dove into aspects I did not attempt to cover in the casebase lit review. NOTE: cmprsk was found in initial scraper, but was ignored in the manual pass through.

The final plot no longer uses a single individual but averages all individuals in our dataset. still uses our work around. Next I will attempt to use riskRegression and compare.

need to add timereg (riskregression is a wrapper for timereg)

@@
 library(survival)
-library(casebase)
+#library(casebase)
+devtools::load_all('../../../casebase')#testing
 library(splines)
@@
 In particular, we excluded packages that were focused on a different set of problems, such as frailty and multi-state models.
-The remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with a description of some of the functionalities they offer.
+The remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with some of the functionalities they offer.
 
 Several packages implement penalized estimation for the Cox model: \pkg{glmnet} [-@regpathcox], \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche].
-Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: \pkg{CoxRidge} [-@perperoglou], \pkg{rstpm2} [-@clements_liu] and \pkg{survival} [-@survival-package].
-On the other hand, \pkg{casebase} provides penalized estimation in the context of parametric hazards.
-To our knowledge, this is the only package to offer this functionality.
-\textcolor{red}{Is this true?
-Because it seems to be a logical conclusion from your paragraph.}
-
-Parametric survival models are implemented in a handful of packages: \pkg{CFC} [-@sharabiani_mahani_2019], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu], \pkg{mets} [-@scheike2014estimating], and \pkg{survival}.
+Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \pkg{CoxRidge} [-@perperoglou] and \pkg{rstpm2} [-@clements_liu], while \pkg{survival} [-@survival-package] has an implementation for ridge.
+\pkg{casebase} provides penalized estimation in the context of parametric hazards.
+To our knowledge, \pkg{casebase} and \pkg{rsptm2} are the only packages to offer this functionality.
+\textcolor{red}{I can dive deeper to determine if this is a fair statement}
+
+Parametric survival models are implemented in a handful of packages: \pkg{CFC} [-@mahani2015bayesian], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu], \pkg{mets} [-@scheike2014estimating], and \pkg{survival}.
 The types of models they allow vary for each package.
@@
 
-Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets} and \pkg{survival} contain implementations for competing risks.
+Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, \pkg{cmprsk} and \pkg{survival} contain implementations for competing risks.
 The differentiating factor between these packages and \pkg{casebase} is that \pkg{casebase} handles competing risks through multiple logistic regression.
-\textcolor{red}{There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}.
-Could you incorporate some of their discussion, with proper citations?
-The biggest omission right now is \pkg{cmprsk}, which we actually use below.}
+\textcolor{red}{The discussion for \pkg{CFC} discusses aspects that I'm not certain are important for our argument.
+Mixture models, Vertical modelling,Psuedo-observations, machine learning models.
+Should i include these?}
 
@@
 \textbf{casebasee} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
-\textbf{CFC} \cite{sharabiani_mahani_2019}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
+\textbf{CFC} \cite{mahani2015bayesian}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
+textbf{cmprsk} \cite{gray_2019}            & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\
 \textbf{coxRidge} \cite{perperoglou}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
@@
 ```{r bmtcrr-data, eval = eval_cs2, message = FALSE}
-library(casebase)
+
 data(bmtcrr)
@@
 
-```{r bmtcrr-risk, echo = FALSE, fig.cap="\\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups.
+```{r bmtcrr-risk, echo = FALSE, fig.cap="\\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups.
 The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.", eval = eval_cs2}
@@
 The SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital.
-Imputation was conducted using predictive mean matching and using the \pkg{mice} package in \proglang{R} [@mice].
-Before imputation, there was a total of 1000 individuals and 35 variables.
-After imputation, we have 474 individuals.
-\textcolor{red}{How come we lose individuals during imputation?} Of these, 67.5% experienced the event of interest.
+Imputation was conducted using the \pkg{mice} package in \proglang{R} with default settings [@mice].
+Before imputation, there were a total of 1000 individuals and 35 variables.
+After imputation, we have 690 individuals.
+\textcolor{red}{we lose individuals due to coliniarity.
+If a pair of covariates are deemed colinear to a certain degree, they are excluded from any imputations.
+here, only edu  income  avtisst*  race  ph*  glucose*  adlp* were imputed.} Of these, 66.09% experienced the event of interest.
 <!--The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.--> A description of each variable can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}.
-The original data was obtained from the website of the Department of Biostatistics at Vanderbilt University.
+The original data was obtained from the Department of Biostatistics at Vanderbilt University \textcolor{red}{do we need to reference the dataset itself?
+Or is the paper enough?
+In my code for generating the data in casebase, i link to the original data online and pull from there}.
 The imputed dataset is available as part of the \pkg{casebase} package.
@@
 The population-time plot for the SUPPORT data appears in Figure \ref{fig:support-poptime}.
-\textcolor{red}{This population-time plot is strange.
-The red dots never seem to go above 200.}
-
-```{r support-poptime, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for tje SUPPORT data."}
+\textcolor{red}{red dots now go all over.
+percentile_number introduced as a parameter in popTime}
+
+```{r support-poptime, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data."}
 data(support)
 support_popTime <- popTime(support,
-                           time = "d.time", event = "death")
+                           time = "d.time", event = "death",percentile_number = 0.31)
 plot(support_popTime)
@@
 
-```{r supportSetUp, echo = TRUE, eval = eval_cs3}
-support <- dplyr::select(support, -hospdead)
-
-# Randomly select one observation
-indiv <- sample(1:nrow(support), size = 1)
-```
-
-Our main objective is to estimate the cumulative incidence of death for a given covariate profile.
-Given the number of covariates available, we will do this using regularized estimation.
-As \pkg{casebase} makes use of the \pkg{glmnet} package, we will use the function \code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}.
-Specifically, we input a matrix \code{y}, which contains the time and event variables, and a matrix \code{x}, which contains all other variables we would like to include in the model.
+```{r supportSetUp, echo = FALSE, eval = eval_cs3}
+completeData=casebase::support
+#remove hospital dead
+completeData<-completeData[,-c(4)]
+#keep one individual out, at random, for which we will develop an 
+#absolute risk curve.
+#sam=sample(1:nrow(completeData), 1)
+#Set workingData to the remaining individuals in the data
+#workingData=completeData[-c(sam),]
+workingCompleteData=model.matrix(death~ .-d.time,data=completeData)[,-c(1)]#remove intercept
+#Create u and xCox, which will be used when fitting Cox with glmnet.
+newData=workingCompleteData
+x=workingCompleteData
+#x and y will be used to fit the casebase model under glmnet.
+#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
+y=data.matrix(completeData[,c(2,5)])
+```
+
+Our main objective is to compute the absolute risk of death for a given set of covariates, using regularization when fitting our model.
+First, we fit a smooth hazard to the data; for the sake of this example, we opted for a linear term for time.
+As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model.
 For the SUPPORT dataset, all factor variables were converted into dummy variables.
-
-```{r support-design, eval = eval_cs3}
-# Create design matrix without the time variable
-support_matrix <- model.matrix(death ~ . - d.time, 
-                               data = support)
-# Remove intercept term
-support_matrix <- support_matrix[,colnames(support_matrix) != "(Intercept)"]
-
-# Create test and train dataset
-support_test <- support_matrix[indiv,,drop = FALSE]
-
-support_train <- support_matrix[-indiv,]
-y <- as.matrix(support[-indiv, c("death", "d.time")])
-```
-
-We are now ready to estimate the hazard function.
-We use lasso penalization; this is achieved by setting \code{alpha} to 1.
-Also, note that the argument \code{formula_time} can be used to specify the function form for time; in our example, we model time linearly.
-
+Here, we used lasso penalization by setting alpha to 1.
 ```{r CBhazard, echo = TRUE, eval = eval_cs3}
-support_cb <- fitSmoothHazard.fit(support_train, y, 
-                                  formula_time = ~ d.time, 
-                                  family = "glmnet", 
-                                  event = "death", time = "d.time", 
-                                  alpha = 1, ratio = 100)
-```
-
-Then, we take our model object and use the \code{absoluteRisk} function to compute our estimate of the cumulative incidence function.
-Since \code{glmnet} fitted the penalized logistic regression for a range of penalty terms, we need to specify which one we want to use for our prediction.
-We select \code{s = "lambda.1se"}, as recommended.
-
+#Using casebase's glmnet implementation, using splines and a case to base ratio of 100,
+#to determine the hazard function
+cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
+                                       formula_time = ~d.time,alpha=1,ratio=100)
+```
+Then, we take our model and use the absoluteRisk function to integrate and retrieve our absolute risk function.
 ```{r CBabsolute, echo = TRUE, eval = eval_cs3}
-time_points <- seq(1, max(support$d.time), by = 0.5)
-support_risk <- absoluteRisk(support_cb,
-                             time = time_points,
-                             newdata = support_test,
-                             s = "lambda.1se", method = "numerical")
-```
-
-We will compare this curve to a regularized version of Cox regression, and a Kaplan-Meier survival curve.
-Using the \pkg{glmnet} package, we can fit the Cox regression model using regularized estimation.
-
-```{r coxHaz, echo = TRUE, eval = eval_cs3}
-# Create survival object for glmnet
-surv_glmnet <- survival::Surv(time = y[, "d.time"], 
-                              event = y[, "death"])
-
-support_cox <- glmnet::cv.glmnet(x = support_train, 
-                                 y = surv_glmnet,
-                                 family = "cox", alpha = 1)
-```
-
-However, \pkg{glmnet} does not provide any functionality for estimating the baseline hazard, and the functionalities in the \pkg{survival} package do not work on the output from \code{cv.glmnet}.
-To circumvent this issue and get an estimate of the baseline hazard, we followed the following steps.
-First, we fitted a non-regularized Cox model using only the variables selected by \pkg{glmnet}; the baseline hazard will be estimated using this object.
-Then, we replaced the estimates of the regression coefficients by the estimates from \pkg{glmnet}.
-Finally, this modified \code{coxph} object is fed to the \code{survival::survfit} function to obtain the estimated survival function.
-
-```{r coxAbsolute, echo = TRUE, eval = eval_cs3}
-# Identifying selected covariates
-nonzero_covariates_cox <- predict(support_cox, type = "nonzero", 
-                                  s = "lambda.1se")
-nonzero_coef_cox <- coef(support_cox, s = "lambda.1se")
-# Creating a new dataset that only contains the covariates selected by glmnet
-support_cox_data <- as.data.frame(cbind(y,
-                                        support_train[, nonzero_covariates_cox$X1]))
-# Fitting a cox model
-support_cox <- survival::coxph(Surv(time = d.time, event = death) ~ ., 
-                          data = support_cox_data)
-# Replace the estimate with those obtained by glmnet
-support_cox$coefficients <- nonzero_coef_cox@x
-
-# Fitting CI curve for cox+glmnet
-support_cox_risk <- survival::survfit(support_cox,
-                                      newdata = as.data.frame(support_test),
-                                      time = time_points,
-                                      type = "breslow")
-```
-
-We used the \pkg{survival} package to calculate the Kaplan-Meier estimate of the survival function.
-
+#Estimating the absolute risk curve using the newData parameter.
+casebaseAbsolute=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 50),newdata = newData , s="lambda.1se",method=c("numerical"))
+cab=data.frame(casebaseAbsolute[,1],rowMeans(casebaseAbsolute[,-c(1)]))
+```
+We will compare this curve to a regularized version of cox regression, and a kaplan-meier survival curve.
+The cox regression absolute risk curve with lasso penalization required two extra steps to retrieve.
+Coxnet from the glmnet package has tools to fit a regularized cox model, but requires that a survival object skeleton of the selected variables, so that the survival package tools can be used to retrive the absolute risk.
+We handle this by fitting a regularized cox model with the glmnet package, and a cox model from the survival package with the selected variables from the regularized fit.
+This second model serves as a skeleton that permits the use of survival package functions.
+The coefficients from the regularized model will replace the coeficients in our skeleton.
+the resulting model will have the correct coefficients when integrating, but will have an incorrect standard error.
+For the purposes of this case study, we are only interested in the absolute risk curve itself and not the standard error.
+
+```{r coxHazAbsolute, echo = TRUE, eval = eval_cs3}
+#Create u and xCox, which will be used when fitting Cox with glmnet.
+u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
+xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))
+#hazard for cox using glmnet
+coxglmFit=glmnet::cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
+#convergence demonstrated in plot
+plot(coxglmFit)
+#taking the coefficient estimates for later use
+nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
+nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
+#creating a new dataset that only contains the covariates chosen through glmnet.
+#cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
+cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
+#newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
+#fitting a cox model using regular estimation, however we will not keep it.
+#this is used more as an object place holder.
+coxFit <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
+#The coefficients of this object will be replaced with the estimates from coxglmFit.
+#Doing so makes it so that everything is invalid aside from the coefficients.
+#In this case, all we need to estimate the absolute risk is the coefficients.
+#Std. error would be incorrect here, if we were to draw error bars.
+coxFit$coefficients<-nonzero_coef_cox@x
+#Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
+newDataCox=newData[,nonzero_covariate_cox$X1-1]
+
+abCoxFit<-survival::survfit(coxFit,newdata=as.data.frame(newDataCox),time = seq(0,max(completeData$d.time), 1),type="breslow")
+
+abCF<-abCoxFit
+abCF$surv<-rowMeans(abCF$surv)
+```
+We used the survival package to calculate the kaplan-meier absolute risk function. 
 ```{r fitKM+abRisk, echo = TRUE, eval = eval_cs3}
-km <- survival::Surv(time = support$d.time, 
-                     event = support$death)
-abKm <- survival::survfit(km ~ 1,
-                          type = 'kaplan-meier')
-```
-
-Now that our three cumulative incidence functions have been calculated, we compare them all in Figure \ref{fig:abSC}.
-Both \pkg{glmnet} and \pkg{casebase} have a lower estimated cumulative incidence by the end of the study, when compared to Kaplan-Meier.
-\textcolor{red}{It would be good to have an idea of why the case-base and Coxnet curves are so different.
-At the very least, I think we should compare the select covariates.}
-
-```{r abSupportComparison, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Comparison of cumulative incidence curves between regularized casebase in black, regularized Cox regression in red and Kaplan-Meier in blue."}
-plot(support_risk, type='l', col = "black", lwd = 2, 
-     main = "Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves", 
-     xlab = "Survival-Time", ylab = "Cumulative Incidence", ylim = c(0,1))
-lines(support_cox_risk, col = "red", fun = "event", lwd = 3, conf.int = FALSE)
-lines(abKm, col = "blue", fun = "event", lwd = 3, conf.int = FALSE)
-legend("bottom", 
-       legend = c("casebase", "Cox+glmnet", "KM curve"), 
-       col = c("black", "red", "blue"), lty = c(1, 1, 1), bg = "gray90")
+#creating a surv object to be used to fit an unadjusted absolute risk curve.
+# (Kaplan-Meier risk curve)
+km <- survival::Surv(time = completeData$d.time, event = completeData$death)
+abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
+
+```
+
+Now that our three absolute risk functions have been calculated, we compare them all in Figure \ref{fig:abSC}.
+Both coxnet and casebase decrease the absolute risk by the end of the study, in comparison to kaplan-meier.
+```{r abSupportComparison, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
+# A plot to compare all three Absolute risk (Commulative Incidence)
+plot(cab[,1],cab[,2],type='l',col="black",lwd=3,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
+lines(abCF,col="red",fun="event",lwd=3,conf.int = FALSE)
+lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
+legend("bottomright", 
+       legend = c( "Casebase (Lasso+linear)","semi-parametric (Cox)+glmnet","KM curve"), 
+       col = c("black","red","Blue"),
+       lty = c(1, 1, 1), 
+       bg = "gray90")
 ```
@@
 The purpose of the next case study is to show how \pkg{casebase} can also be used for survival analysis problems that do not share the properties above.
-To this end, we will use the Stanford Heart Transplant Data [@clark1971cardiac,@crowley1977covariance]; we will use the version available in the \pkg{surivival} package.
+To this end, we will use the Stanford Heart Transplant Data [@clark1971cardiac,@crowley1977covariance]; we will use the version available in the \pkg{survival} package.
 

### 2020-01-16 analysis/paper/paper.tex
fixedMaxComments, added questions, averaged absolute risks
I fixed a few of Max's comments, but some I was not certain about. particularly what to do with CFC's own package methods exploration. they dove into aspects I did not attempt to cover in the casebase lit review. NOTE: cmprsk was found in initial scraper, but was ignored in the manual pass through.

The final plot no longer uses a single individual but averages all individuals in our dataset. still uses our work around. Next I will attempt to use riskRegression and compare.

need to add timereg (riskregression is a wrapper for timereg)

@@
 problems, such as frailty and multi-state models. The remaining 14
-packages appear in Table \ref{tab:surv-pkgs}, along with a description
-of some of the functionalities they offer.
+packages appear in Table \ref{tab:surv-pkgs}, along with some of the
+functionalities they offer.
 
@@
 packages also include penalized estimation in the context of Cox models
-with time-varying coefficients: \pkg{CoxRidge}
-\citeyearpar{perperoglou}, \pkg{rstpm2} \citeyearpar{clements_liu} and
-\pkg{survival} \citeyearpar{survival-package}. On the other hand,
+with time-varying coefficients: elastic-net penalization with
+\pkg{CoxRidge} \citeyearpar{perperoglou} and \pkg{rstpm2}
+\citeyearpar{clements_liu}, while \pkg{survival}
+\citeyearpar{survival-package} has an implementation for ridge.
 \pkg{casebase} provides penalized estimation in the context of
-parametric hazards. To our knowledge, this is the only package to offer
-this functionality.
-\textcolor{red}{Is this true? Because it seems to be a logical conclusion from your paragraph.}
+parametric hazards. To our knowledge, \pkg{casebase} and \pkg{rsptm2}
+are the only packages to offer this functionality.
+\textcolor{red}{I can dive deeper to determine if this is a fair statement}
 
 Parametric survival models are implemented in a handful of packages:
-\pkg{CFC} \citeyearpar{sharabiani_mahani_2019}, \pkg{flexsurv}
+\pkg{CFC} \citeyearpar{mahani2015bayesian}, \pkg{flexsurv}
 \citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
@@
 Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
-\pkg{mets} and \pkg{survival} contain implementations for competing
-risks. The differentiating factor between these packages and
+\pkg{mets}, \pkg{cmprsk} and \pkg{survival} contain implementations for
+competing risks. The differentiating factor between these packages and
 \pkg{casebase} is that \pkg{casebase} handles competing risks through
 multiple logistic regression.
-\textcolor{red}{There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}.
-Could you incorporate some of their discussion, with proper citations?
-The biggest omission right now is \pkg{cmprsk}, which we actually use below.}
+\textcolor{red}{The discussion for \pkg{CFC} discusses aspects that I'm not certain are important for our argument.
+Mixture models, Vertical modelling,Psuedo-observations, machine learning models.
+Should i include these?}
 
@@
 \begin{CodeInput}
-R> library(casebase)
+R> 
 R> data(bmtcrr)
@@
 individuals who are considered seriously ill within a hospital.
-Imputation was conducted using predictive mean matching and using the
-\pkg{mice} package in \proglang{R} \citep{mice}. Before imputation,
-there was a total of 1000 individuals and 35 variables. After
-imputation, we have 474 individuals.
-\textcolor{red}{How come we lose individuals during imputation?} Of
-these, 67.5\% experienced the event of interest. A description of each
-variable can be found in Table \ref{tab:support1} and a breakdown of
-each categorical variable in Table \ref{tab:support2}. The original data
-was obtained from the website of the Department of Biostatistics at
-Vanderbilt University. The imputed dataset is available as part of the
-\pkg{casebase} package.
+Imputation was conducted using the \pkg{mice} package in \proglang{R}
+with default settings \citep{mice}. Before imputation, there were a
+total of 1000 individuals and 35 variables. After imputation, we have
+690 individuals.
+\textcolor{red}{we lose individuals due to coliniarity.
+If a pair of covariates are deemed colinear to a certain degree, they are excluded from any imputations.
+here, only edu  income  avtisst*  race  ph*  glucose*  adlp* were imputed.}
+Of these, 66.09\% experienced the event of interest. A description of
+each variable can be found in Table \ref{tab:support1} and a breakdown
+of each categorical variable in Table \ref{tab:support2}. The original
+data was obtained from the Department of Biostatistics at Vanderbilt
+University
+\textcolor{red}{do we need to reference the dataset itself? Or is the paper enough? In my code for generating the data in casebase, i link to the original data online and pull from there}.
+The imputed dataset is available as part of the \pkg{casebase} package.
 
@@
 population-time plot for the SUPPORT data appears in Figure
-\ref{fig:support-poptime}.
-\textcolor{red}{This population-time plot is strange. The red dots never seem to go above 200.}
+\ref{fig:support-poptime}. \textbackslash{}textcolor\{red\}\{red dots
+now go all over. percentile\_number introduced as a parameter in
+popTime\}
 
@@
 R> support_popTime <- popTime(support,
-R+                            time = "d.time", event = "death")
+R+                            time = "d.time", event = "death",percentile_number = 0.31)
 R> plot(support_popTime)
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> support <- dplyr::select(support, -hospdead)
-R> 
-R> # Randomly select one observation
-R> indiv <- sample(1:nrow(support), size = 1)
-\end{CodeInput}
-\end{CodeChunk}
-
-Our main objective is to estimate the cumulative incidence of death for
-a given covariate profile. Given the number of covariates available, we
-will do this using regularized estimation. As \pkg{casebase} makes use
-of the \pkg{glmnet} package, we will use the function
-\code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}.
-Specifically, we input a matrix \code{y}, which contains the time and
-event variables, and a matrix \code{x}, which contains all other
-variables we would like to include in the model. For the SUPPORT
-dataset, all factor variables were converted into dummy variables.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Create design matrix without the time variable
-R> support_matrix <- model.matrix(death ~ . - d.time, 
-R+                                data = support)
-R> # Remove intercept term
-R> support_matrix <- support_matrix[,colnames(support_matrix) != "(Intercept)"]
-R> 
-R> # Create test and train dataset
-R> support_test <- support_matrix[indiv,,drop = FALSE]
-R> 
-R> support_train <- support_matrix[-indiv,]
-R> y <- as.matrix(support[-indiv, c("death", "d.time")])
-\end{CodeInput}
-\end{CodeChunk}
-
-We are now ready to estimate the hazard function. We use lasso
-penalization; this is achieved by setting \code{alpha} to 1. Also, note
-that the argument \code{formula_time} can be used to specify the
-function form for time; in our example, we model time linearly.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> support_cb <- fitSmoothHazard.fit(support_train, y, 
-R+                                   formula_time = ~ d.time, 
-R+                                   family = "glmnet", 
-R+                                   event = "death", time = "d.time", 
-R+                                   alpha = 1, ratio = 100)
-\end{CodeInput}
-\end{CodeChunk}
-
-Then, we take our model object and use the \code{absoluteRisk} function
-to compute our estimate of the cumulative incidence function. Since
-\code{glmnet} fitted the penalized logistic regression for a range of
-penalty terms, we need to specify which one we want to use for our
-prediction. We select \code{s = "lambda.1se"}, as recommended.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> time_points <- seq(1, max(support$d.time), by = 0.5)
-R> support_risk <- absoluteRisk(support_cb,
-R+                              time = time_points,
-R+                              newdata = support_test,
-R+                              s = "lambda.1se", method = "numerical")
-\end{CodeInput}
-\end{CodeChunk}
-
-We will compare this curve to a regularized version of Cox regression,
-and a Kaplan-Meier survival curve. Using the \pkg{glmnet} package, we
-can fit the Cox regression model using regularized estimation.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Create survival object for glmnet
-R> surv_glmnet <- survival::Surv(time = y[, "d.time"], 
-R+                               event = y[, "death"])
-R> 
-R> support_cox <- glmnet::cv.glmnet(x = support_train, 
-R+                                  y = surv_glmnet,
-R+                                  family = "cox", alpha = 1)
-\end{CodeInput}
-\end{CodeChunk}
-
-However, \pkg{glmnet} does not provide any functionality for estimating
-the baseline hazard, and the functionalities in the \pkg{survival}
-package do not work on the output from \code{cv.glmnet}. To circumvent
-this issue and get an estimate of the baseline hazard, we followed the
-following steps. First, we fitted a non-regularized Cox model using only
-the variables selected by \pkg{glmnet}; the baseline hazard will be
-estimated using this object. Then, we replaced the estimates of the
-regression coefficients by the estimates from \pkg{glmnet}. Finally,
-this modified \code{coxph} object is fed to the \code{survival::survfit}
-function to obtain the estimated survival function.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Identifying selected covariates
-R> nonzero_covariates_cox <- predict(support_cox, type = "nonzero", 
-R+                                   s = "lambda.1se")
-R> nonzero_coef_cox <- coef(support_cox, s = "lambda.1se")
-R> # Creating a new dataset that only contains the covariates selected by glmnet
-R> support_cox_data <- as.data.frame(cbind(y,
-R+                                         support_train[, nonzero_covariates_cox$X1]))
-R> # Fitting a cox model
-R> support_cox <- survival::coxph(Surv(time = d.time, event = death) ~ ., 
-R+                           data = support_cox_data)
-R> # Replace the estimate with those obtained by glmnet
-R> support_cox$coefficients <- nonzero_coef_cox@x
-R> 
-R> # Fitting CI curve for cox+glmnet
-R> support_cox_risk <- survival::survfit(support_cox,
-R+                                       newdata = as.data.frame(support_test),
-R+                                       time = time_points,
-R+                                       type = "breslow")
-\end{CodeInput}
-\end{CodeChunk}
-
-We used the \pkg{survival} package to calculate the Kaplan-Meier
-estimate of the survival function.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> km <- survival::Surv(time = support$d.time, 
-R+                      event = support$death)
-R> abKm <- survival::survfit(km ~ 1,
-R+                           type = 'kaplan-meier')
-\end{CodeInput}
-\end{CodeChunk}
-
-Now that our three cumulative incidence functions have been calculated,
-we compare them all in Figure \ref{fig:abSC}. Both \pkg{glmnet} and
-\pkg{casebase} have a lower estimated cumulative incidence by the end of
-the study, when compared to Kaplan-Meier.
-\textcolor{red}{It would be good to have an idea of why the case-base and Coxnet curves are so different. At the very least, I think we should compare the select covariates.}
-
-\begin{CodeChunk}
+Our main objective is to compute the absolute risk of death for a given
+set of covariates, using regularization when fitting our model. First,
+we fit a smooth hazard to the data; for the sake of this example, we
+opted for a linear term for time. As casebase makes use of the glmnet
+package, we interact with the fitsmoothhazard.fit function using a
+matrix interface, where y contains the time and event variables and x
+contains all other variables we would like to include in the model. For
+the SUPPORT dataset, all factor variables were converted into dummy
+variables. Here, we used lasso penalization by setting alpha to 1.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #Using casebase's glmnet implementation, using splines and a case to base ratio of 100,
+R> #to determine the hazard function
+R> cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
+R+                                        formula_time = ~d.time,alpha=1,ratio=100)
+\end{CodeInput}
+\end{CodeChunk}
+
+Then, we take our model and use the absoluteRisk function to integrate
+and retrieve our absolute risk function.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #Estimating the absolute risk curve using the newData parameter.
+R> casebaseAbsolute=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 50),newdata = newData , s="lambda.1se",method=c("numerical"))
+R> cab=data.frame(casebaseAbsolute[,1],rowMeans(casebaseAbsolute[,-c(1)]))
+\end{CodeInput}
+\end{CodeChunk}
+
+We will compare this curve to a regularized version of cox regression,
+and a kaplan-meier survival curve. The cox regression absolute risk
+curve with lasso penalization required two extra steps to retrieve.
+Coxnet from the glmnet package has tools to fit a regularized cox model,
+but requires that a survival object skeleton of the selected variables,
+so that the survival package tools can be used to retrive the absolute
+risk. We handle this by fitting a regularized cox model with the glmnet
+package, and a cox model from the survival package with the selected
+variables from the regularized fit. This second model serves as a
+skeleton that permits the use of survival package functions. The
+coefficients from the regularized model will replace the coeficients in
+our skeleton. the resulting model will have the correct coefficients
+when integrating, but will have an incorrect standard error. For the
+purposes of this case study, we are only interested in the absolute risk
+curve itself and not the standard error.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #Create u and xCox, which will be used when fitting Cox with glmnet.
+R> u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
+R> xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))
+R> #hazard for cox using glmnet
+R> coxglmFit=glmnet::cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
+R> #convergence demonstrated in plot
+R> plot(coxglmFit)
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/coxHazAbsolute-1} \end{center}
+
+\begin{CodeInput}
+R> #taking the coefficient estimates for later use
+R> nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
+R> nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
+R> #creating a new dataset that only contains the covariates chosen through glmnet.
+R> #cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
+R> cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
+R> #newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
+R> #fitting a cox model using regular estimation, however we will not keep it.
+R> #this is used more as an object place holder.
+R> coxFit <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
+R> #The coefficients of this object will be replaced with the estimates from coxglmFit.
+R> #Doing so makes it so that everything is invalid aside from the coefficients.
+R> #In this case, all we need to estimate the absolute risk is the coefficients.
+R> #Std. error would be incorrect here, if we were to draw error bars.
+R> coxFit$coefficients<-nonzero_coef_cox@x
+R> #Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
+R> newDataCox=newData[,nonzero_covariate_cox$X1-1]
+R> 
+R> abCoxFit<-survival::survfit(coxFit,newdata=as.data.frame(newDataCox),time = seq(0,max(completeData$d.time), 1),type="breslow")
+R> 
+R> abCF<-abCoxFit
+R> abCF$surv<-rowMeans(abCF$surv)
+\end{CodeInput}
+\end{CodeChunk}
+
+We used the survival package to calculate the kaplan-meier absolute risk
+function.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #creating a surv object to be used to fit an unadjusted absolute risk curve.
+R> # (Kaplan-Meier risk curve)
+R> km <- survival::Surv(time = completeData$d.time, event = completeData$death)
+R> abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
+\end{CodeInput}
+\end{CodeChunk}
+
+Now that our three absolute risk functions have been calculated, we
+compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase
+decrease the absolute risk by the end of the study, in comparison to
+kaplan-meier.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # A plot to compare all three Absolute risk (Commulative Incidence)
+R> plot(cab[,1],cab[,2],type='l',col="black",lwd=3,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
+R> lines(abCF,col="red",fun="event",lwd=3,conf.int = FALSE)
+R> lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
+R> legend("bottomright", 
+R+        legend = c( "Casebase (Lasso+linear)","semi-parametric (Cox)+glmnet","KM curve"), 
+R+        col = c("black","red","Blue"),
+R+        lty = c(1, 1, 1), 
+R+        bg = "gray90")
+\end{CodeInput}
 \begin{figure}
@@
 
-\caption{\label{fig:abSC} Comparison of cumulative incidence curves between regularized casebase in black, regularized Cox regression in red and Kaplan-Meier in blue.}\label{fig:abSupportComparison}
+\caption{\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue.}\label{fig:abSupportComparison}
 \end{figure}
@@
 
-Note that the fourth model (i.e.~\texttt{fit4}) includes an interaction
+Note that the fourth model (i.e. \texttt{fit4}) includes an interaction
 term between exposure and follow-up time. In other words, this model no
@@
 
-This report was generated on 2019-12-12 15:16:14 using the following
+This report was generated on 2020-01-16 13:51:03 using the following
 computational environment and dependencies:
@@
 \begin{CodeOutput}
-#> ─ Session info ──────────────────────────────────────────────────────────
+#> - Session info ---------------------------------------------------------------
 #>  setting  value                       
-#>  version  R version 3.6.1 (2019-07-05)
-#>  os       Ubuntu 18.04.3 LTS          
-#>  system   x86_64, linux-gnu           
+#>  version  R version 3.6.0 (2019-04-26)
+#>  os       macOS High Sierra 10.13.6   
+#>  system   x86_64, darwin15.6.0        
 #>  ui       X11                         
@@
 #>  ctype    en_CA.UTF-8                 
-#>  tz       America/Winnipeg            
-#>  date     2019-12-12                  
+#>  tz       America/Montreal            
+#>  date     2020-01-16                  
 #> 
-#> ─ Packages ──────────────────────────────────────────────────────────────
-#>  package     * version    date       lib
+#> - Packages -------------------------------------------------------------------
+#>  ! package     * version    date       lib
 #>    assertthat    0.2.1      2019-03-21 [1]
@@
 #>    broom         0.5.2      2019-04-07 [1]
-#>  callr         3.3.1      2019-07-18 [1]
-#>  casebase    * 0.2.1.9001 2019-12-11 [1]
+#>    callr         3.4.0      2019-12-09 [1]
+#>  P casebase    * 0.2.1.9001 2019-12-11 [?]
 #>    cellranger    1.1.0      2016-07-27 [1]
-#>  cli           2.0.0      2019-12-09 [1]
-#>  codetools     0.2-16     2018-12-24 [4]
+#>    cli           2.0.1      2020-01-08 [1]
+#>    codetools     0.2-16     2018-12-24 [1]
 #>    colorspace    1.4-1      2019-03-18 [1]
@@
 #>    data.table    1.12.8     2019-12-09 [1]
+#>    DBI           1.0.0      2018-05-02 [1]
+#>    dbplyr        1.4.2      2019-06-17 [1]
 #>    desc          1.2.0      2018-05-01 [1]
-#>  devtools      2.1.0      2019-07-06 [1]
+#>    devtools      2.2.1      2019-09-24 [1]
 #>    digest        0.6.23     2019-11-23 [1]
 #>    dplyr       * 0.8.3      2019-07-04 [1]
+#>    ellipsis      0.3.0      2019-09-20 [1]
 #>    evaluate      0.14       2019-05-28 [1]
-#>  fansi         0.4.0      2018-10-05 [1]
+#>    fansi         0.4.1      2020-01-08 [1]
 #>    farver        2.0.1      2019-11-13 [1]
 #>    forcats     * 0.4.0      2019-02-17 [1]
-#>  foreach     * 1.4.4      2017-12-12 [1]
+#>    foreach       1.4.7      2019-07-27 [1]
 #>    fs            1.3.1      2019-05-06 [1]
@@
 #>    ggplot2     * 3.2.1      2019-08-10 [1]
-#>  glmnet      * 2.0-18     2019-05-20 [1]
+#>    glmnet      * 3.0-1      2019-11-15 [1]
 #>    glue          1.3.1      2019-03-12 [1]
 #>    gtable        0.3.0      2019-03-25 [1]
-#>  haven         2.1.1      2019-07-04 [1]
+#>    haven         2.2.0      2019-11-08 [1]
 #>    hms           0.5.2      2019-10-30 [1]
-#>  htmltools     0.3.6      2017-04-28 [1]
+#>    htmltools     0.4.0      2019-10-04 [1]
 #>    httr          1.4.1      2019-08-05 [1]
-#>  iterators     1.0.9      2017-12-12 [1]
+#>    iterators     1.0.12     2019-07-26 [1]
 #>    jsonlite      1.6        2018-12-07 [1]
 #>    kableExtra  * 1.1.0      2019-03-16 [1]
-#>  knitr         1.23       2019-05-18 [1]
+#>    knitr         1.26       2019-11-12 [1]
 #>    labeling      0.3        2014-08-23 [1]
-#>  lattice       0.20-38    2018-11-04 [4]
+#>    lattice       0.20-38    2018-11-04 [1]
 #>    lazyeval      0.2.2      2019-03-15 [1]
@@
 #>    magrittr    * 1.5        2014-11-22 [1]
-#>  Matrix      * 1.2-18     2019-11-27 [4]
+#>    Matrix      * 1.2-17     2019-03-22 [1]
 #>    memoise       1.1.0      2017-04-21 [1]
-#>  mgcv          1.8-24     2018-06-18 [1]
+#>    mgcv          1.8-31     2019-11-09 [1]
 #>    modelr        0.1.5      2019-08-08 [1]
 #>    munsell       0.5.0      2018-06-12 [1]
-#>  nlme          3.1-142    2019-11-07 [4]
-#>  pillar        1.4.2      2019-06-29 [1]
-#>  pkgbuild      1.0.3      2019-03-20 [1]
+#>    nlme          3.1-142    2019-11-07 [1]
+#>    pillar        1.4.3      2019-12-20 [1]
+#>    pkgbuild      1.0.6      2019-10-09 [1]
 #>    pkgconfig     2.0.3      2019-09-22 [1]
@@
 #>    remotes       2.1.0      2019-06-24 [1]
+#>    reprex        0.3.0      2019-05-16 [1]
 #>    rlang         0.4.2      2019-11-23 [1]
-#>  rmarkdown     1.16       2019-10-01 [1]
+#>    rmarkdown     1.18       2019-11-27 [1]
 #>    rprojroot     1.3-2      2018-01-03 [1]
 #>    rstudioapi    0.10       2019-03-19 [1]
-#>  rticles       0.9.1      2019-07-22 [1]
+#>    rticles       0.12       2019-11-12 [1]
 #>    rvest         0.3.5      2019-11-08 [1]
@@
 #>    sessioninfo   1.1.1      2018-11-05 [1]
-#>  stringi       1.4.3      2019-03-12 [1]
+#>    shape         1.4.4      2018-02-07 [1]
+#>    stringi       1.4.5      2020-01-11 [1]
 #>    stringr     * 1.4.0      2019-02-10 [1]
-#>  survival    * 2.44-1.1   2019-04-01 [4]
-#>  testthat      2.2.0      2019-07-22 [1]
+#>    survival    * 3.1-7      2019-11-09 [1]
+#>    testthat    * 2.3.1      2019-12-01 [1]
 #>    tibble      * 2.1.3      2019-06-06 [1]
@@
 #>    tidyselect    0.2.5      2018-10-11 [1]
-#>  tidyverse   * 1.2.1      2017-11-14 [1]
+#>    tidyverse   * 1.3.0      2019-11-21 [1]
 #>    usethis       1.5.1      2019-07-04 [1]
-#>  vctrs         0.2.0      2019-07-05 [1]
+#>    vctrs         0.2.1      2019-12-17 [1]
 #>    VGAM          1.1-2      2019-11-21 [1]
 #>    viridisLite   0.3.0      2018-02-01 [1]
-#>  webshot       0.5.1      2018-09-28 [1]
+#>    webshot       0.5.2      2019-11-22 [1]
 #>    withr         2.1.2      2018-03-15 [1]
-#>  xfun          0.8        2019-06-25 [1]
+#>    xfun          0.11       2019-11-12 [1]
 #>    xml2          1.2.2      2019-08-09 [1]
@@
 #>  source                                  
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
 #>  Github (sahirbhatnagar/casebase@e2e73fe)
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
-#>  Github (rstudio/rticles@8d56fc6)        
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.2)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.1)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
 #> 
-#> [1] /home/mturgeon/Rlibs
-#> [2] /usr/local/lib/R/site-library
-#> [3] /usr/lib/R/site-library
-#> [4] /usr/lib/R/library
+#> [1] /Library/Frameworks/R.framework/Versions/3.6/Resources/library
+#> 
+#>  P -- Loaded and on-disk path mismatch.
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> Local:    review /home/mturgeon/Documents/git_repositories/cbpaper
-#> Head:     [c095709] 2019-12-12: review 2nd case study
+#> Local:    master /Users/jislam/Desktop/Gits/cbpaper
+#> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
+#> Head:     [a528002] 2019-12-12: Merge pull request #12 from sahirbhatnagar/review
 \end{CodeOutput}

### 2020-01-16 analysis/paper/paper.Rmd
mostly Done
I fixed a few of Max's comments, but some I was not certain about. particularly what to do with CFC's own package methods exploration. they dove into aspects I did not attempt to cover in the casebase lit review. NOTE: cmprsk was found in initial scraper, but was ignored in the manual pass through.

The final plot no longer uses a single individual but averages all individuals in our dataset. still uses our work around. Next I will attempt to use riskRegression and compare.

need to add timereg (riskregression is a wrapper for timereg)

@@
 library(survival)
-#library(casebase)
-devtools::load_all('../../../casebase')#testing
+library(casebase)
+#devtools::load_all('../../../casebase')#testing
 library(splines)

### 2020-01-21 analysis/paper/paper.Rmd
write discussion and conclusion

--- 2019-12-12 analysis/paper/paper.Rmd
@@
 eval_implementation <- TRUE
-eval_cs1 <- FALSE
-eval_cs2 <- FALSE
+eval_cs1 <- TRUE
+eval_cs2 <- TRUE
 eval_cs3 <- TRUE
-eval_cs4 <- FALSE
+eval_cs4 <- TRUE
 ```
@@
 
-```{r support-poptime, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for tje SUPPORT data."}
+```{r support-poptime, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data."}
 data(support)
@@
 # Use the trapezoidal rule to estimate the cumulative hazard
-min_hazard <- min(one_yr_haz$hazard)
-max_hazard <- max(one_yr_haz$hazard)
-increment <- 365/99
-cumhaz_est <- 0.5*increment*(sum(2*one_yr_haz$hazard) - min_hazard - max_hazard)
+cumhaz_est <- pracma::trapz(one_yr_haz$futime, one_yr_haz$hazard)
 1 - exp(-cumhaz_est)
@@
 In this article, we presented the \proglang{R} package \pkg{casebase} that provides functions to fit smooth parametric hazards and estimate cumulative incidence functions using case-base sampling.
-We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through three case studies.
-
-<!-- In the following table we provide a comparison between the Cox model and case-base sampling: -->
-
-```{r, echo=FALSE, eval = FALSE, results='asis'}
+We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through four case studies.
+
+As a methodological framework, case-base sampling is very flexible.
+This flexibility has been explored before in the literature: for example, Saarela and Hanley [-@saarela2015case] used case-base sampling to model a time-dependent exposure variable in a vaccine safety study.
+As another example, Saarela and Arjas [-@saarela2015non] combined case-base sampling and a Bayesian non-parametric framework to compute individualized risk assessments for chronic diseases.
+In the case studies above, we explored this flexibility along two fronts.
+On the one hand, we showed how splines could be used as part of the linear predictor to model the effect of time on the hazard.
+This strategy yielded estimates of the survival function that were qualitatively similar to semiparametric estimates derived from Cox regression; however, case-base sampling led to estimates that *varying smoothly in time*.
+On the other hand, we also displayed the flexibility of case-base sampling by showing how it could be combined with penalized logistic regression to perform variable selection.
+Similarly, our package \pkg{casebase} provides the user with the functionality of Generalized Additive Models.
+Moreover, case-base sampling can be combined with quasi-likelihood estimation to fit survival models that potentially account for the presence of over-dispersion.
+All of the examples illustrate how the case-base sampling framework in general, and the package \pkg{casebase} in particular, allows the user to fit a broad family of survival functions.
+
+As presented in Hanley & Miettinen [-@hanley2009fitting], case-base sampling is comprised of three steps: 1) sampling a case series and a base series from the study; 2) fit the log-hazard as a linear function of predictors (including time); and 3) use the fitted hazard to estimate the cumulative incidence curve.
+Accordingly, our package provides functions for each step.
+Moreover, the simple interface of the \code{fittingSmoothHazard} function resembles the \code{glm} interface, which should look familiar to new users.
+Our modular approach also provides a convenient way to extend our package for new sampling or fitting strategies.
+
+In the case studies above, we compared the performance of case-base sampling with that of Cox regression and Fine-Gray models.
+In terms of function interface, \pkg{casebase} uses a formula interface that is closer to that of \code{glm}, in that the event variable is the only variable appearing on the left-hand side of the formula.
+By contrast, both \code{survival::coxph} and \code{timereg::comp.risk} use arrays as the left-hand side that captures both the event type and time.
+Both approaches to modeling yield user-friendly code.
+In terms of output, both approaches differ significantly.
+Case-base sampling produces smooth hazard and smooth cumulative incidence curves, whereas Cox regression and Fine-Gray models produce step-wise cumulative incidence functions and never explicitly model the hazard function.
+Qualitatively, we showed that by using splines in the linear predictor, all three models yielded similar curves.
+However, the smooth nature of the output of \pkg{casebase} improves interpretation for consumers of these predictions.
+In Table \ref{tab:compCBvsCox}, we provide a side-by-side comparison between the Cox model and case-base sampling.
+
+\begin{table}
+\caption{\label{tab:compCBvsCox}Comparison between the Cox model and case-base sampling}
+\centering
+\begin{tabular}[t]{llp{5cm}}
+\toprule
+Feature & Cox model & Case-base sampling\\
+\midrule
+Model type & Semi-parametric & Fully parametric\\
+Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\
+Cumulative incidence & Step function & Smooth-in-time curve\\
+Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\
+Model testing &  & Use GLM framework \newline (e.g.\ LRT, AIC, BIC)\\
+\addlinespace
+Competing risks & Difficult & Cause-specific CIFs\\
+% Prediction & Kaplan-Meier-based & ROC, AUC, risk reclassification probabilities\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+Our choice of modeling the log-hazard as a linear function of covariates allows us to develop a simple computational scheme for estimation.
+However, as a downside, it does not allow us to model location and scale parameters separately like the package \pkg{flexsurv}.
+For example, if we look at the Weibull distribution as parametrised in \code{stats::pweibull}, the log-hazard function is given by
+$$ \log h(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,$$
+where $\alpha,\beta$ are shape and scale parameters, respectively.
+Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} allows the user to model the scale parameter as a function of covariates.
+Of course, this added flexibility comes at the cost of interpretability: by modeling the log-hazard directly, the parameter estimates from \pkg{casebase} can be interpreted as estimates of log-hazard ratios.
+To improve the flexibility of \pkg{casebase} at capturing the scale of a parametric family, we could replace the logistic regression with its quasi-likelihood counterpart and therefore model over- and under-dispersion with respect to the logistic likelihood.
+We defer the study of the properties and performance of such a model to a future article.
+
+Future work will look at some of the methodological extensions of case-base sampling.
+First, to assess the quality of the model fit, we may want to study the properties of the residuals (e.g.
+Cox-Snell, martingale).
+More work needs to be done to understand these residuals in the context of the partial likelihood underlying case-base sampling.
+The resulting diagnostic tools would then be integrated in this package.
+Also, we are interested in extending case-base sampling to account for interval censoring.
+This type of censoring is very common in longitudinal studies, and many packages (e.g.
+\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2}) provide functions to account for it.
+Again, we hope to include any resulting methodology as part of this package.
+
+In future versions of the package, we also want to increase the complement of diagnostic and inferential tools that are currently available.
+For example, we would like to include the ability to compute confidence intervals for the cumulative incidence curve.
+The delta method or parametric bootstrap are two different strategies we can use to construct approximate confidence intervals.
+Furthermore, we would like to include functions to compute calibration and discrimination statistics (e.g.
+Brier score, AUC) for our models.
+Saarela and Arjas [-@saarela2015non] also describe how to obtain a posterior distribution for the AUC from their model.
+
+In conclusion, we presented the \proglang{R} package \pkg{casebase} which implements case-base sampling for fitting parametric survival models and for estimating smooth cumulative incidence functions using the framework of generalized linear models.
+We strongly believe that its flexibility and its foundation on the familiar logistic regression model will make it appealing to new and established practioners.
+
+```{r, echo=FALSE, eval = FALSE}
 knitr::kable(data.frame(`feature` = c("model type", "time", "cumulative incidence", "non-proportional hazards","model testing","competing risks", "prediction"),
@@
            "smooth-in-time curve", "interaction of covariates with time", "make use of GLM framework (LRT, AIC, BIC)",
-           "cause-specific cumulative incidence functions (CIFs) directly obtained via multinomial regression", "ROC, AUC, risk reclassification probabilities")), escape = FALSE)
-```
-
-\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval censoring, whereas \pkg{casebase} currently cannot.
-Interval censoring is a simple feature to implement and is a potential improvement to add to \pkg{casebase}.
+           "cause-specific cumulative incidence functions (CIFs) directly obtained via multinomial regression", "ROC, AUC, risk reclassification probabilities")), 
+           col.names = c("Feature", "Cox model", "Case-base sampling"),
+           format = "latex", booktabs = TRUE,
+           caption = "Comparison between the Cox model and case-base sampling", 
+           escape = FALSE)
+```
 

### 2020-01-21 analysis/paper/paper.Rmd

@@
 library(kableExtra)
+library(pracma)
 options(digits = 2, scipen = 999)

### 2020-01-21 analysis/paper/paper.tex

--- 2019-12-12 analysis/paper/paper.tex
@@
 R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
-R>                        levels = c(0, 1), 
-R>                        labels = c("Control group", "Screening group"))
+R+                        levels = c(0, 1), 
+R+                        labels = c("Control group", "Screening group"))
 \end{CodeInput}
@@
 R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
-R>                              data = ERSPC)
+R+                              data = ERSPC)
 R> summary(cox_model)
 \end{CodeInput}
+
+\begin{CodeOutput}
+#> Call:
+#> survival::coxph(formula = Surv(Follow.Up.Time, DeadOfPrCa) ~ 
+#>     ScrArm, data = ERSPC)
+#> 
+#>   n= 159893, number of events= 540 
+#> 
+#>                         coef exp(coef) se(coef)     z Pr(>|z|)  
+#> ScrArmScreening group -0.222     0.801    0.088 -2.52    0.012 *
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#>                       exp(coef) exp(-coef) lower .95 upper .95
+#> ScrArmScreening group     0.801       1.25     0.674     0.952
+#> 
+#> Concordance= 0.519  (se = 0.011 )
+#> Likelihood ratio test= 6.45  on 1 df,   p=0.01
+#> Wald test            = 6.37  on 1 df,   p=0.01
+#> Score (logrank) test = 6.39  on 1 df,   p=0.01
+\end{CodeOutput}
 \end{CodeChunk}
@@
 R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
-R>                        ignore = 99)
+R+                        ignore = 99)
 R> 
 R> plot(survfit(cox_model, newdata = new_data),
-R>      xlab = "Years since Randomization", 
-R>      ylab = "Cumulative Incidence (%)", 
-R>      fun = "event",
-R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>                     exp(coef(cox_model)), 
-R>                     exp(confint(cox_model))[1], 
-R>                     exp(confint(cox_model))[2]),
-R>      ylim = c(0, 2), yscale = 100)
+R+      xlab = "Years since Randomization", 
+R+      ylab = "Cumulative Incidence (%)", 
+R+      fun = "event",
+R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+                     exp(coef(cox_model)), 
+R+                     exp(confint(cox_model))[1], 
+R+                     exp(confint(cox_model))[2]),
+R+      ylim = c(0, 2), yscale = 100)
 R> legend("topleft", 
-R>        legend = c("Control group", "Screening group"), 
-R>        col = c("red","blue"),
-R>        lty = c(1, 1), 
-R>        bg = "gray90")
-\end{CodeInput}
+R+        legend = c("Control group", "Screening group"), 
+R+        col = c("red","blue"),
+R+        lty = c(1, 1), 
+R+        bg = "gray90")
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/erspc-cox-cif-1} \end{center}
+
 \end{CodeChunk}
@@
 R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
-R>                                           time = seq(0,15,0.1), 
-R>                                           newdata = new_data)
+R+                                           time = seq(0,15,0.1), 
+R+                                           newdata = new_data)
 R> 
 R> plot(survfit(cox_model, newdata = new_data),
-R>      xlab = "Years since Randomization", 
-R>      ylab = "Cumulative Incidence (%)", 
-R>      fun = "event",
-R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>                     exp(coef(cox_model)), 
-R>                     exp(confint(cox_model))[1], 
-R>                     exp(confint(cox_model))[2]),
-R>      ylim = c(0, 2), yscale = 100)
+R+      xlab = "Years since Randomization", 
+R+      ylab = "Cumulative Incidence (%)", 
+R+      fun = "event",
+R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+                     exp(coef(cox_model)), 
+R+                     exp(confint(cox_model))[1], 
+R+                     exp(confint(cox_model))[2]),
+R+      ylim = c(0, 2), yscale = 100)
 R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
@@
 R> legend("topleft", 
-R>        legend = c("Control group (Cox)","Control group (Casebase)",
-R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R>        col = c("red","red", "blue","blue"),
-R>        lty = c(1, 2, 1, 2), 
-R>        bg = "gray90")
-\end{CodeInput}
+R+        legend = c("Control group (Cox)","Control group (Casebase)",
+R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R+        col = c("red","red", "blue","blue"),
+R+        lty = c(1, 2, 1, 2), 
+R+        bg = "gray90")
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/erspc-casebase-cif-1} \end{center}
+
 \end{CodeChunk}
@@
 R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
-R>                                           time = seq(0,15,0.1), 
-R>                                           newdata = new_data)
+R+                                           time = seq(0,15,0.1), 
+R+                                           newdata = new_data)
 R> 
 R> plot(survfit(cox_model, newdata = new_data),
-R>      xlab = "Years since Randomization", 
-R>      ylab = "Cumulative Incidence (%)", 
-R>      fun = "event",
-R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>                     exp(coef(cox_model)), 
-R>                     exp(confint(cox_model))[1], 
-R>                     exp(confint(cox_model))[2]),
-R>      ylim = c(0, 2), yscale = 100)
+R+      xlab = "Years since Randomization", 
+R+      ylab = "Cumulative Incidence (%)", 
+R+      fun = "event",
+R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+                     exp(coef(cox_model)), 
+R+                     exp(confint(cox_model))[1], 
+R+                     exp(confint(cox_model))[2]),
+R+      ylim = c(0, 2), yscale = 100)
 R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
@@
 R> legend("topleft", 
-R>        legend = c("Control group (Cox)","Control group (Casebase)",
-R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R>        col = c("red","red", "blue","blue"),
-R>        lty = c(1, 2, 1, 2), 
-R>        bg = "gray90")
-\end{CodeInput}
+R+        legend = c("Control group (Cox)","Control group (Casebase)",
+R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R+        col = c("red","red", "blue","blue"),
+R+        lty = c(1, 2, 1, 2), 
+R+        bg = "gray90")
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/erspc-casebase-weibull-cif-1} \end{center}
+
 \end{CodeChunk}
@@
 R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
-R>                                     time = seq(0,15,0.1), 
-R>                                     newdata = new_data)
+R+                                     time = seq(0,15,0.1), 
+R+                                     newdata = new_data)
 R> 
 R> plot(survfit(cox_model, newdata = new_data),
-R>      xlab = "Years since Randomization", 
-R>      ylab = "Cumulative Incidence (%)", 
-R>      fun = "event",
-R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>                     exp(coef(cox_model)), 
-R>                     exp(confint(cox_model))[1], 
-R>                     exp(confint(cox_model))[2]),
-R>      ylim = c(0, 2), yscale = 100)
+R+      xlab = "Years since Randomization", 
+R+      ylab = "Cumulative Incidence (%)", 
+R+      fun = "event",
+R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+                     exp(coef(cox_model)), 
+R+                     exp(confint(cox_model))[1], 
+R+                     exp(confint(cox_model))[2]),
+R+      ylim = c(0, 2), yscale = 100)
 R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
@@
 R> legend("topleft", 
-R>        legend = c("Control group (Cox)","Control group (Casebase)",
-R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R>        col = c("red","red", "blue","blue"),
-R>        lty = c(1, 2, 1, 2), 
-R>        bg = "gray90")
-\end{CodeInput}
+R+        legend = c("Control group (Cox)","Control group (Casebase)",
+R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R+        col = c("red","red", "blue","blue"),
+R+        lty = c(1, 2, 1, 2), 
+R+        bg = "gray90")
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/erspc-casebase-splines-cif-1} \end{center}
+
 \end{CodeChunk}
@@
 
+\begin{CodeChunk}
+\begin{table}[H]
+\centering
+\begin{tabular}{lrlr}
+\toprule
+Model & HR & CI & Width\\
+\midrule
+Cox & 0.80 & (0.67, 0.95) & 0.28\\
+Exponential & 0.81 & (0.68, 0.96) & 0.28\\
+Weibull & 0.79 & (0.67, 0.94) & 0.28\\
+Splines & 0.81 & (0.68, 0.96) & 0.28\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\end{CodeChunk}
+
 As noted above, the usual asymptotic results hold for likelihood ratio
@@
 \end{CodeInput}
+
+\begin{CodeOutput}
+#> Analysis of Deviance Table
+#> 
+#> Model 1: DeadOfPrCa ~ ScrArm + offset(offset)
+#> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
+#>   Resid. Df Resid. Dev Df Deviance            Pr(>Chi)    
+#> 1     54538       6053                                    
+#> 2     54535       5790  3      262 <0.0000000000000002 ***
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{CodeOutput}
 \end{CodeChunk}
@@
 R> c("Exp." = AIC(casebase_exponential),
-R>   "Weib." = AIC(casebase_time),
-R>   "Splines" = AIC(casebase_splines))
-\end{CodeInput}
+R+   "Weib." = AIC(casebase_time),
+R+   "Splines" = AIC(casebase_splines))
+\end{CodeInput}
+
+\begin{CodeOutput}
+#>    Exp.   Weib. Splines 
+#>    6057    5817    5800
+\end{CodeOutput}
 \end{CodeChunk}
@@
 
+\begin{CodeChunk}
+\begin{table}
+
+\caption{\label{tab:table1bmtcrr}Baseline characteristics of patients in the stem-cell transplant study.}
+\centering
+\begin{tabular}[t]{lll}
+\toprule
+Variable & Description & Statistical summary\\
+\midrule
+\code{Sex} & Sex & M=Male (100)\\
+ &  & F=Female (77)\\
+\code{D} & Disease & ALL (73)\\
+ &  & AML (104)\\
+\code{Phase} & Phase & CR1 (47)\\
+\addlinespace
+ &  & CR2 (45)\\
+ &  & CR3 (12)\\
+ &  & Relapse (73)\\
+\code{Source} & Type of transplant & BM+PB (21)\\
+ &  & PB (156)\\
+\addlinespace
+\code{Age} & Age of patient (years) & 4--62\\
+ &  & 30.47 (13.04)\\
+\code{ftime} & Failure time (months) & 0.13--131.77\\
+ &  & 20.28 (30.78)\\
+\code{Status} & Status indicator & 0=censored (46)\\
+\addlinespace
+ &  & 1=relapse (56)\\
+ &  & 2=competing event (75)\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\end{CodeChunk}
+
 First, we can look at a population-time plot to visualize the incidence
@@
 
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/compPop-1}
+
+}
+
+\caption[Population-time plot for the stem-cell transplant study]{Population-time plot for the stem-cell transplant study.
+A) The red points represent relapse events.
+B) The blue points represent the competing events.}\label{fig:compPop}
+\end{figure}
+\end{CodeChunk}
+
 Our main objective is to compute the absolute risk of relapse for a
@@
 R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-R>                    data = bmtcrr)
-\end{CodeInput}
+R+                    data = bmtcrr)
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{table}
+
+\caption{\label{tab:bmtcrr-cis}ADD CAPTION}
+\centering
+\begin{tabular}[t]{lrlrl}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
+\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
+Covariates & HR & 95\% CI & HR & 95\% CI\\
+\midrule
+Sex & 0.71 & (0.41, 1.24) & 0.68 & (0.39, 1.19)\\
+Disease & 0.55 & (0.31, 1) & 0.52 & (0.29, 0.93)\\
+Phase (CR2 vs. CR1) & 1.18 & (0.47, 2.93) & 1.21 & (0.48, 3.02)\\
+Phase (CR3 vs. CR1) & 1.47 & (0.38, 5.63) & 1.67 & (0.43, 6.5)\\
+Phase (Relapse vs. CR1) & 4.30 & (2, 9.27) & 4.55 & (2.09, 9.9)\\
+\addlinespace
+Source & 1.56 & (0.52, 4.7) & 1.46 & (0.47, 4.54)\\
+Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
+\bottomrule
+\end{tabular}
+\end{table}
+
 \end{CodeChunk}
@@
 R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
-R>                                      levels = levels(bmtcrr[,"Sex"])),
-R>                       "D" = c("ALL", "AML"), # Both diseases
-R>                       "Phase" = factor(c("Relapse", "Relapse"), 
-R>                                        levels = levels(bmtcrr[,"Phase"])),
-R>                       "Age" = c(35, 35),
-R>                       "Source" = factor(c("PB", "PB"), 
-R>                                         levels = levels(bmtcrr[,"Source"])))
+R+                                      levels = levels(bmtcrr[,"Sex"])),
+R+                       "D" = c("ALL", "AML"), # Both diseases
+R+                       "Phase" = factor(c("Relapse", "Relapse"), 
+R+                                        levels = levels(bmtcrr[,"Phase"])),
+R+                       "Age" = c(35, 35),
+R+                       "Source" = factor(c("PB", "PB"), 
+R+                                         levels = levels(bmtcrr[,"Source"])))
 R> 
@@
 R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
-R>                         method = "numerical", newdata = newdata)
+R+                         method = "numerical", newdata = newdata)
 \end{CodeInput}
@@
 R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) + 
-R>                           const(Phase) + const(Source) + const(Age), 
-R>                       data = bmtcrr, cause = 1, model = "fg")
+R+                           const(Phase) + const(Source) + const(Age), 
+R+                       data = bmtcrr, cause = 1, model = "fg")
 R> 
@@
 \end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/bmtcrr-risk-1}
+
+}
+
+\caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups.
+The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:bmtcrr-risk}
+\end{figure}
 \end{CodeChunk}
@@
 
-\caption{\label{fig:support-poptime} Population-time plot for tje SUPPORT data.}\label{fig:support-poptime}
+\caption{\label{fig:support-poptime} Population-time plot for the SUPPORT data.}\label{fig:support-poptime}
 \end{figure}
@@
 R> stanford_popTime <- popTime(jasa, time = "futime", 
-R>                             event = "fustat")
+R+                             event = "fustat")
 R> plot(stanford_popTime)
 \end{CodeInput}
+
+\begin{center}\includegraphics{../figures/stanford-poptime-1} \end{center}
+
 \end{CodeChunk}
@@
 R> cb_data <- sampleCaseBase(jasa, time = "futime", 
-R>                           event = "fustat", ratio = 10)
+R+                           event = "fustat", ratio = 10)
 \end{CodeInput}
@@
 R> cb_data <- mutate(cb_data,
-R>                   txtime = time_length(accept.dt %--% tx.date, 
-R>                                        unit = "days"),
-R>                   exposure = case_when(
-R>                     is.na(txtime) ~ 0L,
-R>                     txtime > futime ~ 0L,
-R>                     txtime <= futime ~ 1L
-R>                   ))
+R+                   txtime = time_length(accept.dt %--% tx.date, 
+R+                                        unit = "days"),
+R+                   exposure = case_when(
+R+                     is.na(txtime) ~ 0L,
+R+                     txtime > futime ~ 0L,
+R+                     txtime <= futime ~ 1L
+R+                   ))
 \end{CodeInput}
@@
 R> fit1 <- fitSmoothHazard(fustat ~ exposure,
-R>                         data = cb_data, time = "futime")
+R+                         data = cb_data, time = "futime")
 R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
-R>                         data = cb_data, time = "futime")
+R+                         data = cb_data, time = "futime")
 R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
-R>                         data = cb_data, time = "futime")
+R+                         data = cb_data, time = "futime")
 R> fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
-R>                         data = cb_data, time = "futime")
-\end{CodeInput}
-\end{CodeChunk}
-
-Note that the fourth model (i.e.~\texttt{fit4}) includes an interaction
+R+                         data = cb_data, time = "futime")
+\end{CodeInput}
+\end{CodeChunk}
+
+Note that the fourth model (i.e. \texttt{fit4}) includes an interaction
 term between exposure and follow-up time. In other words, this model no
@@
 R> c("Model1" = AIC(fit1),
-R>   "Model2" = AIC(fit2),
-R>   "Model3" = AIC(fit3),
-R>   "Model4" = AIC(fit4))
-\end{CodeInput}
+R+   "Model2" = AIC(fit2),
+R+   "Model3" = AIC(fit3),
+R+   "Model4" = AIC(fit4))
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> Model1 Model2 Model3 Model4 
+#>    795    760    729    724
+\end{CodeOutput}
 \end{CodeChunk}
@@
 R> hazard_data <- expand.grid(exposure = c(0, 1),
-R>                            futime = seq(0, 1000,
-R>                                         length.out = 100))
+R+                            futime = seq(0, 1000,
+R+                                         length.out = 100))
 R> # Set the offset to zero
@@
 R> hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
-R>                                  type = "link"))
+R+                                  type = "link"))
 R> # Add labels for plots
 R> hazard_data$Status = factor(hazard_data$exposure,
-R>                             labels = c("NoTrans", "Trans"))
+R+                             labels = c("NoTrans", "Trans"))
 R> 
 R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
-R>     geom_line() +
-R>     theme_minimal() +
-R>     theme(legend.position = 'top') +
-R>     ylab('Hazard') + xlab('Follow-up time')
-\end{CodeInput}
+R+     geom_line() +
+R+     theme_minimal() +
+R+     theme(legend.position = 'top') +
+R+     ylab('Hazard') + xlab('Follow-up time')
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/stanford-hazard-1} \end{center}
+
 \end{CodeChunk}
@@
 R> absrisk <- absoluteRisk(fit4, newdata = newdata, 
-R>                         time = seq(0, 1000, length.out = 100))
+R+                         time = seq(0, 1000, length.out = 100))
 R> 
@@
 R> absrisk <- gather(as.data.frame(absrisk),
-R>                   "Status", "Risk", -Time)
+R+                   "Status", "Risk", -Time)
 R>  
 R> ggplot(absrisk, aes(Time, Risk, colour = Status)) +
-R>   geom_line() +
-R>   theme_minimal() +
-R>   theme(legend.position = 'top') +
-R>   expand_limits(y = 0.5) +
-R>   xlab('Follow-up time') + ylab('Cum. Incidence')
-\end{CodeInput}
+R+   geom_line() +
+R+   theme_minimal() +
+R+   theme(legend.position = 'top') +
+R+   expand_limits(y = 0.5) +
+R+   xlab('Follow-up time') + ylab('Cum. Incidence')
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/stanford-risk-1} \end{center}
+
 \end{CodeChunk}
@@
 R> one_yr_haz <- data.frame(futime = seq(0, 365,
-R>                                       length.out = 100))
+R+                                       length.out = 100))
 R> one_yr_haz <- mutate(one_yr_haz,
-R>                      offset = 0,
-R>                      exposure = if_else(futime < 30, 0, 1))
+R+                      offset = 0,
+R+                      exposure = if_else(futime < 30, 0, 1))
 R> one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
-R>                                 type = "link"))
+R+                                 type = "link"))
 R> ggplot(one_yr_haz, aes(futime, hazard)) +
-R>   geom_line() +
-R>   theme_minimal() +
-R>   theme(legend.position = 'top') +
-R>   ylab('Hazard') + xlab('Follow-up time') +
-R>   geom_vline(xintercept = 30, linetype = 'dashed')
-\end{CodeInput}
+R+   geom_line() +
+R+   theme_minimal() +
+R+   theme(legend.position = 'top') +
+R+   ylab('Hazard') + xlab('Follow-up time') +
+R+   geom_vline(xintercept = 30, linetype = 'dashed')
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/stanford-1y-haz-1} \end{center}
+
 \end{CodeChunk}
@@
 R> absoluteRisk(fit4, newdata = data.frame(exposure = 0),
-R>              time = 365)
+R+              time = 365)
+\end{CodeInput}
+
+\begin{CodeOutput}
+#>         
+#> 365 0.27
+\end{CodeOutput}
+
+\begin{CodeInput}
 R> 
 R> # Use the trapezoidal rule to estimate the cumulative hazard
-R> min_hazard <- min(one_yr_haz$hazard)
-R> max_hazard <- max(one_yr_haz$hazard)
-R> increment <- 365/99
-R> cumhaz_est <- 0.5*increment*(sum(2*one_yr_haz$hazard) - min_hazard - max_hazard)
+R> cumhaz_est <- pracma::trapz(one_yr_haz$futime, one_yr_haz$hazard)
 R> 1 - exp(-cumhaz_est)
 \end{CodeInput}
+
+\begin{CodeOutput}
+#> [1] 0.19
+\end{CodeOutput}
 \end{CodeChunk}
@@
 implementation, and we illustrated the merits of the approach and the
-package through three case studies.
-
-\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval
-censoring, whereas \pkg{casebase} currently cannot. Interval censoring
-is a simple feature to implement and is a potential improvement to add
-to \pkg{casebase}.
+package through four case studies.
+
+As a methodological framework, case-base sampling is very flexible. This
+flexibility has been explored before in the literature: for example,
+Saarela and Hanley \citeyearpar{saarela2015case} used case-base sampling
+to model a time-dependent exposure variable in a vaccine safety study.
+As another example, Saarela and Arjas \citeyearpar{saarela2015non}
+combined case-base sampling and a Bayesian non-parametric framework to
+compute individualized risk assessments for chronic diseases. In the
+case studies above, we explored this flexibility along two fronts. On
+the one hand, we showed how splines could be used as part of the linear
+predictor to model the effect of time on the hazard. This strategy
+yielded estimates of the survival function that were qualitatively
+similar to semiparametric estimates derived from Cox regression;
+however, case-base sampling led to estimates that \emph{varying smoothly
+in time}. On the other hand, we also displayed the flexibility of
+case-base sampling by showing how it could be combined with penalized
+logistic regression to perform variable selection. Similarly, our
+package \pkg{casebase} provides the user with the functionality of
+Generalized Additive Models. Moreover, case-base sampling can be
+combined with quasi-likelihood estimation to fit survival models that
+potentially account for the presence of over-dispersion. All of the
+examples illustrate how the case-base sampling framework in general, and
+the package \pkg{casebase} in particular, allows the user to fit a broad
+family of survival functions.
+
+As presented in Hanley \& Miettinen \citeyearpar{hanley2009fitting},
+case-base sampling is comprised of three steps: 1) sampling a case
+series and a base series from the study; 2) fit the log-hazard as a
+linear function of predictors (including time); and 3) use the fitted
+hazard to estimate the cumulative incidence curve. Accordingly, our
+package provides functions for each step. Moreover, the simple interface
+of the \code{fittingSmoothHazard} function resembles the \code{glm}
+interface, which should look familiar to new users. Our modular approach
+also provides a convenient way to extend our package for new sampling or
+fitting strategies.
+
+In the case studies above, we compared the performance of case-base
+sampling with that of Cox regression and Fine-Gray models. In terms of
+function interface, \pkg{casebase} uses a formula interface that is
+closer to that of \code{glm}, in that the event variable is the only
+variable appearing on the left-hand side of the formula. By contrast,
+both \code{survival::coxph} and \code{timereg::comp.risk} use arrays as
+the left-hand side that captures both the event type and time. Both
+approaches to modeling yield user-friendly code. In terms of output,
+both approaches differ significantly. Case-base sampling produces smooth
+hazard and smooth cumulative incidence curves, whereas Cox regression
+and Fine-Gray models produce step-wise cumulative incidence functions
+and never explicitly model the hazard function. Qualitatively, we showed
+that by using splines in the linear predictor, all three models yielded
+similar curves. However, the smooth nature of the output of
+\pkg{casebase} improves interpretation for consumers of these
+predictions. In Table \ref{tab:compCBvsCox}, we provide a side-by-side
+comparison between the Cox model and case-base sampling.
+
+\begin{table}
+\caption{\label{tab:compCBvsCox}Comparison between the Cox model and case-base sampling}
+\centering
+\begin{tabular}[t]{llp{5cm}}
+\toprule
+Feature & Cox model & Case-base sampling\\
+\midrule
+Model type & Semi-parametric & Fully parametric\\
+Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\
+Cumulative incidence & Step function & Smooth-in-time curve\\
+Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\
+Model testing &  & Use GLM framework \newline (e.g.\ LRT, AIC, BIC)\\
+\addlinespace
+Competing risks & Difficult & Cause-specific CIFs\\
+% Prediction & Kaplan-Meier-based & ROC, AUC, risk reclassification probabilities\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+Our choice of modeling the log-hazard as a linear function of covariates
+allows us to develop a simple computational scheme for estimation.
+However, as a downside, it does not allow us to model location and scale
+parameters separately like the package \pkg{flexsurv}. For example, if
+we look at the Weibull distribution as parametrised in
+\code{stats::pweibull}, the log-hazard function is given by
+\[ \log h(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,\]
+where \(\alpha,\beta\) are shape and scale parameters, respectively.
+Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} allows the
+user to model the scale parameter as a function of covariates. Of
+course, this added flexibility comes at the cost of interpretability: by
+modeling the log-hazard directly, the parameter estimates from
+\pkg{casebase} can be interpreted as estimates of log-hazard ratios. To
+improve the flexibility of \pkg{casebase} at capturing the scale of a
+parametric family, we could replace the logistic regression with its
+quasi-likelihood counterpart and therefore model over- and
+under-dispersion with respect to the logistic likelihood. We defer the
+study of the properties and performance of such a model to a future
+article.
+
+Future work will look at some of the methodological extensions of
+case-base sampling. First, to assess the quality of the model fit, we
+may want to study the properties of the residuals (e.g.~Cox-Snell,
+martingale). More work needs to be done to understand these residuals in
+the context of the partial likelihood underlying case-base sampling. The
+resulting diagnostic tools would then be integrated in this package.
+Also, we are interested in extending case-base sampling to account for
+interval censoring. This type of censoring is very common in
+longitudinal studies, and many packages (e.g. \pkg{SmoothHazard},
+\pkg{survival} and \pkg{Rstpm2}) provide functions to account for it.
+Again, we hope to include any resulting methodology as part of this
+package.
+
+In future versions of the package, we also want to increase the
+complement of diagnostic and inferential tools that are currently
+available. For example, we would like to include the ability to compute
+confidence intervals for the cumulative incidence curve. The delta
+method or parametric bootstrap are two different strategies we can use
+to construct approximate confidence intervals. Furthermore, we would
+like to include functions to compute calibration and discrimination
+statistics (e.g.~Brier score, AUC) for our models. Saarela and Arjas
+\citeyearpar{saarela2015non} also describe how to obtain a posterior
+distribution for the AUC from their model.
+
+In conclusion, we presented the \proglang{R} package \pkg{casebase}
+which implements case-base sampling for fitting parametric survival
+models and for estimating smooth cumulative incidence functions using
+the framework of generalized linear models. We strongly believe that its
+flexibility and its foundation on the familiar logistic regression model
+will make it appealing to new and established practioners.
 
@@
 
-This report was generated on 2019-12-12 15:16:14 using the following
+This report was generated on 2020-01-21 11:25:53 using the following
 computational environment and dependencies:
@@
 \begin{CodeOutput}
-#> ─ Session info ──────────────────────────────────────────────────────────
+#> - Session info ---------------------------------------------------------------
 #>  setting  value                       
-#>  version  R version 3.6.1 (2019-07-05)
+#>  version  R version 3.6.2 (2019-12-12)
 #>  os       Ubuntu 18.04.3 LTS          
@@
 #>  tz       America/Winnipeg            
-#>  date     2019-12-12                  
-#> 
-#> ─ Packages ──────────────────────────────────────────────────────────────
+#>  date     2020-01-21                  
+#> 
+#> - Packages -------------------------------------------------------------------
 #>  package     * version    date       lib
@@
 #>  broom         0.5.2      2019-04-07 [1]
-#>  callr         3.3.1      2019-07-18 [1]
-#>  casebase    * 0.2.1.9001 2019-12-11 [1]
+#>  callr         3.3.2      2019-09-22 [1]
+#>  casebase    * 0.2.1.9001 2020-01-10 [1]
 #>  cellranger    1.1.0      2016-07-27 [1]
-#>  cli           2.0.0      2019-12-09 [1]
+#>  cli           2.0.1      2020-01-08 [1]
 #>  codetools     0.2-16     2018-12-24 [4]
 #>  colorspace    1.4-1      2019-03-18 [1]
+#>  cowplot     * 1.0.0      2019-07-11 [1]
 #>  crayon        1.3.4      2017-09-16 [1]
 #>  data.table    1.12.8     2019-12-09 [1]
+#>  DBI           1.0.0      2018-05-02 [1]
+#>  dbplyr        1.4.2      2019-06-17 [1]
 #>  desc          1.2.0      2018-05-01 [1]
-#>  devtools      2.1.0      2019-07-06 [1]
+#>  devtools      2.2.1      2019-09-24 [1]
 #>  digest        0.6.23     2019-11-23 [1]
 #>  dplyr       * 0.8.3      2019-07-04 [1]
+#>  ellipsis      0.3.0      2019-09-20 [1]
 #>  evaluate      0.14       2019-05-28 [1]
-#>  fansi         0.4.0      2018-10-05 [1]
+#>  fansi         0.4.1      2020-01-08 [1]
 #>  farver        2.0.1      2019-11-13 [1]
 #>  forcats     * 0.4.0      2019-02-17 [1]
-#>  foreach     * 1.4.4      2017-12-12 [1]
+#>  foreach       1.4.7      2019-07-27 [1]
 #>  fs            1.3.1      2019-05-06 [1]
@@
 #>  ggplot2     * 3.2.1      2019-08-10 [1]
-#>  glmnet      * 2.0-18     2019-05-20 [1]
-#>  glue          1.3.1      2019-03-12 [1]
+#>  glmnet      * 3.0-1      2019-11-15 [1]
+#>  glue        * 1.3.1      2019-03-12 [1]
 #>  gtable        0.3.0      2019-03-25 [1]
-#>  haven         2.1.1      2019-07-04 [1]
+#>  haven         2.2.0      2019-11-08 [1]
 #>  hms           0.5.2      2019-10-30 [1]
-#>  htmltools     0.3.6      2017-04-28 [1]
+#>  htmltools     0.4.0      2019-10-04 [1]
 #>  httr          1.4.1      2019-08-05 [1]
-#>  iterators     1.0.9      2017-12-12 [1]
+#>  iterators     1.0.12     2019-07-26 [1]
 #>  jsonlite      1.6        2018-12-07 [1]
 #>  kableExtra  * 1.1.0      2019-03-16 [1]
-#>  knitr         1.23       2019-05-18 [1]
+#>  knitr         1.26       2019-11-12 [1]
 #>  labeling      0.3        2014-08-23 [1]
 #>  lattice       0.20-38    2018-11-04 [4]
+#>  lava          1.6.6      2019-08-01 [1]
 #>  lazyeval      0.2.2      2019-03-15 [1]
 #>  lifecycle     0.1.0      2019-08-01 [1]
-#>  lubridate     1.7.4      2018-04-11 [1]
+#>  lubridate   * 1.7.4      2018-04-11 [1]
 #>  magrittr    * 1.5        2014-11-22 [1]
+#>  MASS          7.3-51.5   2019-12-20 [4]
 #>  Matrix      * 1.2-18     2019-11-27 [4]
 #>  memoise       1.1.0      2017-04-21 [1]
-#>  mgcv          1.8-24     2018-06-18 [1]
+#>  mgcv          1.8-31     2019-11-09 [4]
 #>  modelr        0.1.5      2019-08-08 [1]
 #>  munsell       0.5.0      2018-06-12 [1]
-#>  nlme          3.1-142    2019-11-07 [4]
-#>  pillar        1.4.2      2019-06-29 [1]
-#>  pkgbuild      1.0.3      2019-03-20 [1]
+#>  nlme          3.1-143    2019-12-10 [4]
+#>  numDeriv      2016.8-1.1 2019-06-06 [1]
+#>  pillar        1.4.3      2019-12-20 [1]
+#>  pkgbuild      1.0.6      2019-10-09 [1]
 #>  pkgconfig     2.0.3      2019-09-22 [1]
 #>  pkgload       1.0.2      2018-10-29 [1]
+#>  plyr          1.8.5      2019-12-10 [1]
+#>  pracma        2.2.9      2019-12-15 [1]
 #>  prettyunits   1.0.2      2015-07-13 [1]
@@
 #>  remotes       2.1.0      2019-06-24 [1]
+#>  reprex        0.3.0      2019-05-16 [1]
+#>  reshape2      1.4.3      2017-12-11 [1]
 #>  rlang         0.4.2      2019-11-23 [1]
-#>  rmarkdown     1.16       2019-10-01 [1]
+#>  rmarkdown     1.18       2019-11-27 [1]
 #>  rprojroot     1.3-2      2018-01-03 [1]
 #>  rstudioapi    0.10       2019-03-19 [1]
-#>  rticles       0.9.1      2019-07-22 [1]
+#>  rticles       0.12       2019-11-12 [1]
 #>  rvest         0.3.5      2019-11-08 [1]
@@
 #>  sessioninfo   1.1.1      2018-11-05 [1]
-#>  stringi       1.4.3      2019-03-12 [1]
+#>  shape         1.4.4      2018-02-07 [1]
+#>  stringi       1.4.4      2020-01-09 [1]
 #>  stringr     * 1.4.0      2019-02-10 [1]
-#>  survival    * 2.44-1.1   2019-04-01 [4]
-#>  testthat      2.2.0      2019-07-22 [1]
+#>  survival    * 3.1-8      2019-12-03 [4]
+#>  testthat      2.3.1      2019-12-01 [1]
 #>  tibble      * 2.1.3      2019-06-06 [1]
@@
 #>  tidyselect    0.2.5      2018-10-11 [1]
-#>  tidyverse   * 1.2.1      2017-11-14 [1]
+#>  tidyverse   * 1.3.0      2019-11-21 [1]
+#>  timereg     * 1.9.4      2019-07-30 [1]
 #>  usethis       1.5.1      2019-07-04 [1]
-#>  vctrs         0.2.0      2019-07-05 [1]
+#>  vctrs         0.2.1      2019-12-17 [1]
 #>  VGAM          1.1-2      2019-11-21 [1]
 #>  viridisLite   0.3.0      2018-02-01 [1]
-#>  webshot       0.5.1      2018-09-28 [1]
+#>  webshot       0.5.2      2019-11-22 [1]
 #>  withr         2.1.2      2018-03-15 [1]
-#>  xfun          0.8        2019-06-25 [1]
+#>  xfun          0.11       2019-11-12 [1]
 #>  xml2          1.2.2      2019-08-09 [1]
@@
 #>  CRAN (R 3.6.1)                          
-#>  Github (sahirbhatnagar/casebase@e2e73fe)
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
+#>  Github (sahirbhatnagar/casebase@ae42a9f)
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.2)                          
 #>  CRAN (R 3.5.2)                          
 #>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
 #>  CRAN (R 3.5.1)                          
@@
 #>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
-#>  Github (rstudio/rticles@8d56fc6)        
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.2)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.0)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.1)                          
-#> 
-#> [1] /home/mturgeon/Rlibs
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.1)                          
+#> 
+#> [1] /home/turgeonm/R/x86_64-pc-linux-gnu-library/3.6
 #> [2] /usr/local/lib/R/site-library
@@
 \begin{CodeOutput}
-#> Local:    review /home/mturgeon/Documents/git_repositories/cbpaper
-#> Head:     [c095709] 2019-12-12: review 2nd case study
+#> Local:    review /home/turgeonm/Documents/git_repos/cbpaper
+#> Head:     [db3b3b1] 2019-12-12: review of cs 3 with new version of package
 \end{CodeOutput}

### 2020-01-28 analysis/paper/paper.Rmd
cleaning up case study 1. merging into one figure. and echo=FALSE defaults for
whole document

--- 2020-01-21 analysis/paper/paper.Rmd
@@
 title:
+  # If you use tex in the formatted title, also supply version without
+  # For running headers, if needed
   formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
-  # If you use tex in the formatted title, also supply version without
   plain:     "casebase: An Alternative Framework For Survival Analysis"
-  # For running headers, if needed
   short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
@@
 fig_caption: true
+editor_options: 
+  chunk_output_type: console
 ---
@@
   message = FALSE,
-  echo = TRUE,
+  echo = FALSE,
   comment = "#>",
@@
                     exp(confint(cox_model))[2]), 
-     ylim = c(0, 2), yscale = 100)
+     yscale = 100)
 legend("topleft", 
@@
        bg = "gray90")
-```
-
-Next, we show how case-base sampling can be used to carry out a similar analysis. We fit several models that differ in how we choose to model time.
-
-<!-- The \code{fitSmoothHazard} function provides an estimate of the hazard function $h(x, t)$ is the hazard function, where $t$ denotes the numerical value of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any).
--->
-
-First, we will fit an exponential model. Recall that this corresponds to excluding time from the linear predictor.
-
-```{r erspc-casebase-exponential, eval = eval_cs1}
+
 casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
@@
                                                   ratio = 100)
-
-summary(casebase_exponential)
-exp(coef(casebase_exponential)[2])
-exp(confint(casebase_exponential)[2,])
-```
-
-<!-- The \code{absoluteRisk} function provides an estimate of the cumulative incidence curves for a specific risk profile using the following equation: -->
-
-<!-- $$ CI(x, t) = 1 - \exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$ -->
-
-We can then use the \code{absoluteRisk} function to get an estimate of the cumulative incidence curve for a specific covariate profile.
-In the plot below, we overlay the estimated CIF from the exponential model on the Cox model CIF:
-
-```{r erspc-casebase-cif, eval = eval_cs1}
 smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
@@
                                           newdata = new_data)
+lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
+lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
+
+casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
+                                 data = ERSPC, 
+                                 ratio = 100)
+smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
+                                          time = seq(0,15,0.1), 
+                                          newdata = new_data)
+lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
+lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
+
+casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
+                                    data = ERSPC, 
+                                    ratio = 100)
+lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
+lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
+
+legend("topleft", 
+       legend = c("Control group (Cox)","Control group (Casebase)",
+                  "Screening group (Cox)", "Screening group (Casebase)"), 
+       col = c("red","red", "blue","blue"),
+       lty = c(1, 2, 1, 2), 
+       bg = "gray90")
+
+```
+
+Next, we show how case-base sampling can be used to carry out a similar analysis. We fit several models that differ in how we choose to model time.
+
+<!-- The \code{fitSmoothHazard} function provides an estimate of the hazard function $h(x, t)$ is the hazard function, where $t$ denotes the numerical value of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any).
+-->
+
+First, we will fit an exponential model. Recall that this corresponds to excluding time from the linear predictor.
+
+<!-- The \code{absoluteRisk} function provides an estimate of the cumulative incidence curves for a specific risk profile using the following equation: -->
+
+<!-- $$ CI(x, t) = 1 - \exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$ -->
+
+We can then use the \code{absoluteRisk} function to get an estimate of the cumulative incidence curve for a specific covariate profile.
+In the plot below, we overlay the estimated CIF from the exponential model on the Cox model CIF:
+
+```{r erspc-casebase-cif, eval = eval_cs1}
 
@@
                     exp(confint(cox_model))[2]),
-     ylim = c(0, 2), yscale = 100)
-lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
-lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
-
-legend("topleft", 
-       legend = c("Control group (Cox)","Control group (Casebase)",
-                  "Screening group (Cox)", "Screening group (Casebase)"), 
-       col = c("red","red", "blue","blue"),
-       lty = c(1, 2, 1, 2), 
-       bg = "gray90")
+     yscale = 100)
+
 ```
@@
 ```{r erspc-casebase-weibull, eval = eval_cs1}
-casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
-                                 data = ERSPC, 
-                                 ratio = 100)
 
@@
 ```{r erspc-casebase-weibull-cif, eval = eval_cs1}
-smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
-                                          time = seq(0,15,0.1), 
-                                          newdata = new_data)
 
@@
                     exp(confint(cox_model))[2]),
-     ylim = c(0, 2), yscale = 100)
-lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
-lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
+     yscale = 100)
 
@@
 ```{r erspc-casebase-splines, eval = eval_cs1}
-casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
-                                    data = ERSPC, 
-                                    ratio = 100)
 
@@
                     exp(confint(cox_model))[2]),
-     ylim = c(0, 2), yscale = 100)
-lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
-lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
+     yscale = 100)
 

### 2020-01-28 analysis/paper/paper.tex
cleaning up case study 1. merging into one figure. and echo=FALSE defaults for
whole document

@@
 R+                     exp(confint(cox_model))[2]), 
-R+      ylim = c(0, 2), yscale = 100)
+R+      yscale = 100)
 R> legend("topleft", 
@@
 R+                     exp(confint(cox_model))[2]),
-R+      ylim = c(0, 2), yscale = 100)
+R+      yscale = 100)
 R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
@@
 R+                     exp(confint(cox_model))[2]),
-R+      ylim = c(0, 2), yscale = 100)
+R+      yscale = 100)
 R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
@@
 R+                     exp(confint(cox_model))[2]),
-R+      ylim = c(0, 2), yscale = 100)
+R+      yscale = 100)
 R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
@@
 #> Model1 Model2 Model3 Model4 
-#>    795    760    729    724
+#>    803    760    733    716
 \end{CodeOutput}
@@
 #>         
-#> 365 0.27
+#> 365 0.29
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> [1] 0.19
+#> [1] 0.21
 \end{CodeOutput}
@@
 
-This report was generated on 2020-01-21 11:25:53 using the following
+This report was generated on 2020-01-28 14:29:58 using the following
 computational environment and dependencies:
@@
 #>  version  R version 3.6.2 (2019-12-12)
-#>  os       Ubuntu 18.04.3 LTS          
+#>  os       Pop!_OS 19.10               
 #>  system   x86_64, linux-gnu           
 #>  ui       X11                         
-#>  language en_CA:en                    
-#>  collate  en_CA.UTF-8                 
-#>  ctype    en_CA.UTF-8                 
-#>  tz       America/Winnipeg            
-#>  date     2020-01-21                  
+#>  language en_US:en                    
+#>  collate  en_US.UTF-8                 
+#>  ctype    en_US.UTF-8                 
+#>  tz       America/Toronto             
+#>  date     2020-01-28                  
 #> 
@@
 #>  backports     1.1.5      2019-10-02 [1]
-#>  broom         0.5.2      2019-04-07 [1]
-#>  callr         3.3.2      2019-09-22 [1]
-#>  casebase    * 0.2.1.9001 2020-01-10 [1]
+#>  broom         0.5.4      2020-01-27 [1]
+#>  callr         3.4.1      2020-01-24 [1]
+#>  casebase    * 0.2.1.9001 2020-01-27 [1]
 #>  cellranger    1.1.0      2016-07-27 [1]
@@
 #>  data.table    1.12.8     2019-12-09 [1]
-#>  DBI           1.0.0      2018-05-02 [1]
+#>  DBI           1.1.0      2019-12-15 [1]
 #>  dbplyr        1.4.2      2019-06-17 [1]
@@
 #>  fansi         0.4.1      2020-01-08 [1]
-#>  farver        2.0.1      2019-11-13 [1]
+#>  farver        2.0.3      2020-01-16 [1]
 #>  forcats     * 0.4.0      2019-02-17 [1]
@@
 #>  ggplot2     * 3.2.1      2019-08-10 [1]
-#>  glmnet      * 3.0-1      2019-11-15 [1]
+#>  glmnet      * 3.0-2      2019-12-11 [1]
 #>  glue        * 1.3.1      2019-03-12 [1]
@@
 #>  haven         2.2.0      2019-11-08 [1]
-#>  hms           0.5.2      2019-10-30 [1]
+#>  hms           0.5.3      2020-01-08 [1]
 #>  htmltools     0.4.0      2019-10-04 [1]
@@
 #>  kableExtra  * 1.1.0      2019-03-16 [1]
-#>  knitr         1.26       2019-11-12 [1]
+#>  knitr         1.27       2020-01-16 [1]
 #>  labeling      0.3        2014-08-23 [1]
@@
 #>  plyr          1.8.5      2019-12-10 [1]
-#>  pracma        2.2.9      2019-12-15 [1]
-#>  prettyunits   1.0.2      2015-07-13 [1]
+#>  pracma      * 2.2.9      2019-12-15 [1]
+#>  prettyunits   1.1.1      2020-01-24 [1]
 #>  processx      3.4.1      2019-07-18 [1]
@@
 #>  reshape2      1.4.3      2017-12-11 [1]
-#>  rlang         0.4.2      2019-11-23 [1]
-#>  rmarkdown     1.18       2019-11-27 [1]
+#>  rlang         0.4.3      2020-01-24 [1]
+#>  rmarkdown     2.1        2020-01-20 [1]
 #>  rprojroot     1.3-2      2018-01-03 [1]
 #>  rstudioapi    0.10       2019-03-19 [1]
-#>  rticles       0.12       2019-11-12 [1]
+#>  rticles       0.13       2019-12-19 [1]
 #>  rvest         0.3.5      2019-11-08 [1]
@@
 #>  shape         1.4.4      2018-02-07 [1]
-#>  stringi       1.4.4      2020-01-09 [1]
+#>  stringi       1.4.5      2020-01-11 [1]
 #>  stringr     * 1.4.0      2019-02-10 [1]
@@
 #>  tibble      * 2.1.3      2019-06-06 [1]
-#>  tidyr       * 1.0.0      2019-09-11 [1]
+#>  tidyr       * 1.0.2      2020-01-24 [1]
 #>  tidyselect    0.2.5      2018-10-11 [1]
@@
 #>  usethis       1.5.1      2019-07-04 [1]
-#>  vctrs         0.2.1      2019-12-17 [1]
+#>  vctrs         0.2.2      2020-01-24 [1]
 #>  VGAM          1.1-2      2019-11-21 [1]
@@
 #>  withr         2.1.2      2018-03-15 [1]
-#>  xfun          0.11       2019-11-12 [1]
+#>  xfun          0.12       2020-01-13 [1]
 #>  xml2          1.2.2      2019-08-09 [1]
 #>  yaml          2.2.0      2018-07-25 [1]
-#>  zeallot       0.1.0      2018-01-28 [1]
 #>  source                                  
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  Github (sahirbhatnagar/casebase@24359be)
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
 #>  CRAN (R 3.6.1)                          
+#>  CRAN (R 3.6.2)                          
 #>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  Github (sahirbhatnagar/casebase@ae42a9f)
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.5.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.5.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.1)                          
-#> 
-#> [1] /home/turgeonm/R/x86_64-pc-linux-gnu-library/3.6
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.2)                          
+#> 
+#> [1] /home/sahir/R/x86_64-pc-linux-gnu-library/3.6
 #> [2] /usr/local/lib/R/site-library
@@
 \begin{CodeOutput}
-#> Local:    review /home/turgeonm/Documents/git_repos/cbpaper
-#> Head:     [db3b3b1] 2019-12-12: review of cs 3 with new version of package
+#> Local:    write-popTime /home/sahir/git_repositories/cbpaper
+#> Head:     [c926a9e] 2020-01-21: Merge pull request #15 from sahirbhatnagar/fix-casebase-dependency
 \end{CodeOutput}

### 2020-01-30 README.Rmd
cleaned up ERSCP CIF's in 2x2 grid. Need to show
equivalence between survreg and casebase in comparison of coefficients

@@
 
-Source code for Casebase paper
-
-for template, use
+Source code for casebase paper. Install the following packages from GitHub
 
 ```R 
-devtools::install_github("rstudio/rticles")
+# install.packages("devtools")
+
+devtools::install_github("rstudio/rticles") # for template
+devtools::install_github("benmarwick/rrtools") # for building the research compendium/docker container
 ```
+
+# To contribute
+
+1. Modify the `analysis/paper/paper.Rmd` file.
+2. Run `rrtools::add_dependencies_to_description()`. This command will ensure all used packages are added to the DESCRIPTION file.
+3. Compile the `paper.Rmd` file
+4. Commit and push your changes to GitHub
 

### 2020-01-30 analysis/paper/paper.Rmd
cleaned up ERSCP CIF's in 2x2 grid. Need to show
equivalence between survreg and casebase in comparison of coefficients

@@
 
+<!--
 \begin{table}
@@
 \end{table}
+-->
 
@@
 ```{r plot-erspc-data, eval = eval_cs1}
-pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
-plot(pt_object)
+# pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
+# plot(pt_object)
 ```
@@
                                      exposure = "ScrArm")
-plot(pt_object_strat)
+# plot(pt_object_strat)
 ```
@@
 
-```{r erspc-cox-cif, eval = eval_cs1}
-new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
-                       ignore = 99)
-
+```{r erspc-cox-cif, eval = eval_cs1, fig.width=8, fig.height=6, fig.cap="Cumulative incidence functions for control and screening groups in the ERSPC data.
+In each of the panels, we plot the CIF from the Cox model (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression.
+(A) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution.
+(B) Linear function of time which is equivalent to a Gompertz distribution.
+(C) The natural logarithm of time which is equivalent to a Weibull distribution.
+(D) Cubic b-spline expansion of time."}
+cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+myCols <- cbPalette[c(4,7)]
+new_data <- data.frame(ScrArm = c("Control group","Screening group"), ignore = 99)
+
+fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
+             gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
+             weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
+             splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
+
+fits <- lapply(fmla, function(i) {
+  casebase::fitSmoothHazard(i, data = ERSPC, ratio = 100)
+})
+
+risks <- lapply(fits, function(i) {
+  casebase::absoluteRisk(object = i, time = seq(0,15,0.1), newdata = new_data)
+})
+
+# casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
+#                                                   data = ERSPC, 
+#                                                   ratio = 100)
+# smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
+#                                           time = seq(0,15,0.1), 
+#                                           newdata = new_data)
+# 
+# casebase_weibull <- casebase::fitSmoothHazard(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm, 
+#                                            data = ERSPC, 
+#                                            ratio = 100)
+# smooth_risk_weibull <- casebase::absoluteRisk(object = casebase_weibull, 
+#                                            time = seq(0,15,0.1), 
+#                                            newdata = new_data)
+# casebase_splines <- casebase::fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
+#                                               data = ERSPC, 
+#                                               ratio = 100)
+# smooth_risk_splines <- casebase::absoluteRisk(object = casebase_splines, 
+#                                               time = seq(0,15,0.1), 
+#                                               newdata = new_data)
+# 
+# casebase_interaction <- casebase::fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time*ScrArm, 
+#                                                   data = ERSPC, 
+#                                                   ratio = 100)
+# smooth_risk_interaction <- casebase::absoluteRisk(object = casebase_interaction, 
+#                                                   time = seq(0,15,0.1), 
+#                                                   newdata = new_data)
+
+# par(mfrow = c(2, 2))
+# casebase::hazardPlot(casebase_interaction, newdata = new_data[2,], ci = FALSE)
+# casebase::hazardPlot(casebase_interaction, newdata = new_data[1,], ci = FALSE, add = TRUE, lty = 2)
+# 
+# casebase::hazardPlot(casebase_splines, newdata = new_data[2,], ci = FALSE)
+# casebase::hazardPlot(casebase_splines, newdata = new_data[1,], ci = FALSE, add = TRUE, lty = 2)
+# 
+# casebase::hazardPlot(casebase_time, newdata = new_data[2,], ci = FALSE)
+# casebase::hazardPlot(casebase_time, newdata = new_data[1,], add = TRUE, lty = 2)
+# 
+# casebase::hazardPlot(casebase_exponential, newdata = new_data[2,], ci = FALSE)
+# casebase::hazardPlot(casebase_exponential, newdata = new_data[1,], add = TRUE, lty = 2)
+
+# dev.off()
+layout(matrix(1:4, ncol = 2, byrow = FALSE), respect = FALSE)
+par(mar = c(0, 4.1, 4.1, 0))
 plot(survfit(cox_model, newdata = new_data),
-     xlab = "Years since Randomization", 
-     ylab = "Cumulative Incidence (%)", 
+     ylab = "",
+     xlab = "",
      fun = "event",
-     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-                    exp(coef(cox_model)), 
-                    exp(confint(cox_model))[1], 
-                    exp(confint(cox_model))[2]), 
+     xlim = c(0,15), conf.int = FALSE, col = myCols, 
+     # main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+     #                Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+     #                exp(coef(cox_model)), 
+     #                exp(confint(cox_model))[1], 
+     #                exp(confint(cox_model))[2]), 
+     lwd = 2,
+     yscale = 100,
+     xaxt = 'n')
+legend("topleft", 
+       legend = c("Control (Cox)","Control (Casebase)",
+                  "Screening (Cox)", "Screening (Casebase)"), 
+       col = rep(myCols, each = 2),
+       lty = c(1, 2, 1, 2), 
+       lwd = 2,
+       cex = 1.1,
+       bg = "gray90")
+lines(risks$exponential[,1], risks$exponential[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risks$exponential[,1], risks$exponential[,3], col = myCols[2], lty = 2, lwd = 2)
+text(x = 15*0.97, y = 0.05/100, labels = "(A)", cex = 1.5)
+
+par(mar = c(4.1, 4.1, 0, 0))
+plot(survfit(cox_model, newdata = new_data),
+     ylab = "", xlab = "",
+     fun = "event",
+     xlim = c(0,15), conf.int = FALSE, col = myCols, 
+     lwd = 2,
      yscale = 100)
-legend("topleft", 
-       legend = c("Control group", "Screening group"), 
-       col = c("red","blue"),
-       lty = c(1, 1), 
-       bg = "gray90")
+lines(risks$weibull[,1], risks$weibull[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risks$weibull[,1], risks$weibull[,3], col = myCols[2], lty = 2, lwd = 2)  
+text(x = 15*0.97, y = 0.05/100, labels = "(C)", cex = 1.5)
+mtext("Cumulative incidence (%)", side = 2, line = -1.5, outer = TRUE)
+
+par(mar = c(0, 0, 4.1, 3.9))
+plot(survfit(cox_model, newdata = new_data),
+     ylab = "",xlab = "",
+     fun = "event",
+     xaxt = 'n', yaxt = 'n',
+     xlim = c(0,15), conf.int = FALSE, col = myCols, 
+     lwd = 2,
+     yscale = 100)
+lines(risks$gompertz[,1], risks$gompertz[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risks$gompertz[,1], risks$gompertz[,3], col = myCols[2], lty = 2, lwd = 2)
+text(x = 15*0.97, y = 0.05/100, labels = "(B)", cex = 1.5)
+
+par(mar = c(4.1, 0, 0, 3.9))
+plot(survfit(cox_model, newdata = new_data),
+     ylab = "",xlab = "",
+     fun = "event",
+     yaxt = 'n',
+     xlim = c(0,15), conf.int = FALSE, col = myCols, 
+     lwd = 2,
+     yscale = 100)
+lines(risks$splines[,1], risks$splines[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risks$splines[,1], risks$splines[,3], col = myCols[2], lty = 2, lwd = 2)
+text(x = 15*0.97, y = 0.05/100, labels = "(D)", cex = 1.5)
+mtext("Years since randomization", side = 1, line = -1.5, outer = TRUE)
+```
+
+## Equivalence between casebase and AFT
+
+```{r}
+# exponential -------------------------------------------------------------
 
@@
                                                   ratio = 100)
-smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
-                                          time = seq(0,15,0.1), 
-                                          newdata = new_data)
-lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
-lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
-
-casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
+aft_exponential <- survreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
+                           data = ERSPC, dist = "exp")
+
+-coef(aft_exponential)
+coef(casebase_exponential)
+
+# weibull -----------------------------------------------------------------
+
+casebase_weibull <- casebase::fitSmoothHazard(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm, 
                                            data = ERSPC, 
                                            ratio = 100)
-smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
-                                          time = seq(0,15,0.1), 
-                                          newdata = new_data)
-lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
-lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
-
-casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
-                                    data = ERSPC, 
-                                    ratio = 100)
-lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
-lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
-
-legend("topleft", 
-       legend = c("Control group (Cox)","Control group (Casebase)",
-                  "Screening group (Cox)", "Screening group (Casebase)"), 
-       col = c("red","red", "blue","blue"),
-       lty = c(1, 2, 1, 2), 
-       bg = "gray90")
-
+aft_weibull <- survreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
+                       data = ERSPC, dist = "weibull")
+
+-coef(aft_weibull)/aft_weibull$scale
+coef(casebase_weibull)
+```
+
+```{r, eval = F}
+knitr::knit_exit()
 ```
@@
 In the plot below, we overlay the estimated CIF from the exponential model on the Cox model CIF:
-
-```{r erspc-casebase-cif, eval = eval_cs1}
-
-plot(survfit(cox_model, newdata = new_data),
-     xlab = "Years since Randomization", 
-     ylab = "Cumulative Incidence (%)", 
-     fun = "event",
-     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-                    exp(coef(cox_model)), 
-                    exp(confint(cox_model))[1], 
-                    exp(confint(cox_model))[2]),
-     yscale = 100)
-
-```
 
@@
 
-```{r erspc-casebase-weibull, eval = eval_cs1}
-
-summary(casebase_time)
-exp(coef(casebase_time))
-exp(confint(casebase_time))
-```
-
 Again, the estimate of the hazard ratio is similar to that obtained from Cox regression. We then look at the estimate of the CIF:
 
-```{r erspc-casebase-weibull-cif, eval = eval_cs1}
-
-plot(survfit(cox_model, newdata = new_data),
-     xlab = "Years since Randomization", 
-     ylab = "Cumulative Incidence (%)", 
-     fun = "event",
-     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-                    exp(coef(cox_model)), 
-                    exp(confint(cox_model))[1], 
-                    exp(confint(cox_model))[2]),
-     yscale = 100)
-
-legend("topleft", 
-       legend = c("Control group (Cox)","Control group (Casebase)",
-                  "Screening group (Cox)", "Screening group (Casebase)"), 
-       col = c("red","red", "blue","blue"),
-       lty = c(1, 2, 1, 2), 
-       bg = "gray90")
-```
-
 We see that the Weibull model leads to a better fit of the CIF than the Exponential model, when compared with the semi-parametric approach.
@@
 
-```{r erspc-casebase-splines, eval = eval_cs1}
-
-summary(casebase_splines)
-exp(coef(casebase_splines))
-exp(confint(casebase_splines))
-```
-
 Once again, we can see that the estimate of the hazard ratio is similar to that from the Cox regression. We then look at the estimate of the CIF:
-
-```{r erspc-casebase-splines-cif, eval = eval_cs1}
-smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
-                                    time = seq(0,15,0.1), 
-                                    newdata = new_data)
-
-plot(survfit(cox_model, newdata = new_data),
-     xlab = "Years since Randomization", 
-     ylab = "Cumulative Incidence (%)", 
-     fun = "event",
-     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-                    exp(coef(cox_model)), 
-                    exp(confint(cox_model))[1], 
-                    exp(confint(cox_model))[2]),
-     yscale = 100)
-
-legend("topleft", 
-       legend = c("Control group (Cox)","Control group (Casebase)",
-                  "Screening group (Cox)", "Screening group (Casebase)"), 
-       col = c("red","red", "blue","blue"),
-       lty = c(1, 2, 1, 2), 
-       bg = "gray90")
-```
 
@@
 CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
-CI_exp <- c(exp(coef(casebase_exponential))["ScrArmScreening group"], exp(confint(casebase_exponential))["ScrArmScreening group",])
-CI_time <- c(exp(coef(casebase_time))["ScrArmScreening group"], exp(confint(casebase_time))["ScrArmScreening group",])
-CI_splines <- c(exp(coef(casebase_splines))["ScrArmScreening group"], exp(confint(casebase_splines))["ScrArmScreening group",])
-
-table_ci <- rbind(CI_cox, CI_exp, CI_time, CI_splines)
+CI_exp <- c(exp(coef(fits$exponential))["ScrArmScreening group"], exp(confint(fits$exponential))["ScrArmScreening group",])
+CI_gompertz <- c(exp(coef(fits$gompertz))["ScrArmScreening group"], exp(confint(fits$gompertz))["ScrArmScreening group",])
+CI_weibull <- c(exp(coef(fits$weibull))["ScrArmScreening group"], exp(confint(fits$weibull))["ScrArmScreening group",])
+CI_splines <- c(exp(coef(fits$splines))["ScrArmScreening group"], exp(confint(fits$splines))["ScrArmScreening group",])
+
+table_ci <- rbind(CI_cox, CI_exp, CI_gompertz, CI_weibull, CI_splines)
 colnames(table_ci) <- c("HR", "LB", "UB")
 
-data.frame(Model = c("Cox", "Exponential", "Weibull", "Splines")) %>% 
+data.frame(Model = c("Cox", "Exponential", "Gompertz", "Weibull", "Splines")) %>% 
   bind_cols(as.data.frame(table_ci)) %>% 
@@
 ```{r erscp-compare, eval = eval_cs1}
-anova(casebase_exponential, casebase_splines, test = "LRT")
+anova(fits$exponential, fits$splines, test = "LRT")
 ```
@@
 ```{r erscp-aic, eval = eval_cs1}
-c("Exp." = AIC(casebase_exponential),
-  "Weib." = AIC(casebase_time),
-  "Splines" = AIC(casebase_splines))
+c("Exp." = AIC(fits$exponential),
+  "Gomp." = AIC(fits$gompertz),
+  "Weib." = AIC(fits$exponential),
+  "Splines" = AIC(fits$splines))
 ```

### 2020-01-30 analysis/paper/paper.tex
cleaned up ERSCP CIF's in 2x2 grid. Need to show
equivalence between survreg and casebase in comparison of coefficients

@@
 
-\begin{table}
-  \begin{tabular}{p{1.8in}lll}
-\hline
-    &  Parameters &  Density \proglang{R} function & \code{dist}\\
-    & {\footnotesize{(location in \code{\color{red}{red}})}} & & \\
-\hline
-    Exponential & \code{\color{red}{rate}}             & \code{dexp}   & \code{"exp"} \\
-    Weibull (accelerated failure time)     & \code{shape, {\color{red}{scale}}}     & \code{dweibull} & \code{"weibull"} \\
-    Weibull (proportional hazards)     & \code{shape, {\color{red}{scale}}}     & \code{dweibullPH} & \code{"weibullPH"} \\
-    Gamma       & \code{shape, \color{red}{rate}}      & \code{dgamma} & \code{"gamma"}\\
-    Log-normal  & \code{{\color{red}{meanlog}}, sdlog}   & \code{dlnorm} & \code{"lnorm"}\\
-    Gompertz    & \code{shape, {\color{red}{rate}}}      & \code{dgompertz} & \code{"gompertz"} \\
-    Log-logistic & \code{shape, {\color{red}{scale}}}   & \code{dllogis} & \code{"llogis"}\\
-    Generalized gamma (Prentice 1975)   & \code{{\color{red}{mu}}, sigma, Q} & \code{dgengamma} & \code{"gengamma"} \\
-    Generalized gamma (Stacy 1962)& \code{shape, {\color{red}{scale}}, k} & \code{dgengamma.orig} & \code{"gengamma.orig"} \\
-    Generalized F     (stable)    & \code{{\color{red}{mu}}, sigma, Q, P} & \code{dgenf} & \code{"genf"} \\
-    Generalized F     (original)  & \code{{\color{red}{mu}}, sigma, s1, s2} & \code{dgenf.orig} & \code{"genf.orig"} \\
-\hline
-  \end{tabular}
-  \caption{Built-in parametric survival distributions in \pkg{flexsurv}.}
-  \label{tab:dists}
-\end{table}
-
 \hypertarget{implementation-details}{%
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> data(ERSPC)
-R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
-R+                        levels = c(0, 1), 
-R+                        labels = c("Control group", "Screening group"))
-\end{CodeInput}
-\end{CodeChunk}
-
 The results of this study were published by
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
-R> plot(pt_object)
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/plot-erspc-data-1} \end{center}
-
-\end{CodeChunk}
-
 We can also create exposure stratified plots by specifying the
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> pt_object_strat <- casebase::popTime(ERSPC, 
-R+                                      event = "DeadOfPrCa", 
-R+                                      exposure = "ScrArm")
-R> plot(pt_object_strat)
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-1} \end{center}
-
-\end{CodeChunk}
-
 We can also plot them side-by-side using the \code{ncol} argument:
@@
 \begin{CodeChunk}
-
-\begin{CodeInput}
-R> plot(pt_object_strat, ncol = 2)
-\end{CodeInput}
 
@@
 \begin{CodeChunk}
-
-\begin{CodeInput}
-R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
-R+                              data = ERSPC)
-R> summary(cox_model)
-\end{CodeInput}
 
@@
 \begin{CodeChunk}
-
-\begin{CodeInput}
-R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
-R+                        ignore = 99)
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R+      xlab = "Years since Randomization", 
-R+      ylab = "Cumulative Incidence (%)", 
-R+      fun = "event",
-R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R+                     exp(coef(cox_model)), 
-R+                     exp(confint(cox_model))[1], 
-R+                     exp(confint(cox_model))[2]), 
-R+      yscale = 100)
-R> legend("topleft", 
-R+        legend = c("Control group", "Screening group"), 
-R+        col = c("red","blue"),
-R+        lty = c(1, 1), 
-R+        bg = "gray90")
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/erspc-cox-cif-1} \end{center}
-
+\begin{figure}
+
+{\centering \includegraphics{../figures/erspc-cox-cif-1}
+
+}
+
+\caption[Cumulative incidence functions for control and screening groups in the ERSPC data]{Cumulative incidence functions for control and screening groups in the ERSPC data.
+In each of the panels, we plot the CIF from the Cox model (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression.
+(A) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution.
+(B) Linear function of time which is equivalent to a Gompertz distribution.
+(C) The natural logarithm of time which is equivalent to a Weibull distribution.
+(D) Cubic b-spline expansion of time.}\label{fig:erspc-cox-cif}
+\end{figure}
+\end{CodeChunk}
+
+\hypertarget{equivalence-between-casebase-and-aft}{%
+\subsection{Equivalence between casebase and
+AFT}\label{equivalence-between-casebase-and-aft}}
+
+\begin{CodeChunk}
+
+\begin{CodeOutput}
+#>           (Intercept) ScrArmScreening group 
+#>                 -7.77                 -0.22
+\end{CodeOutput}
+
+\begin{CodeOutput}
+#>           (Intercept) ScrArmScreening group 
+#>                 -7.76                 -0.23
+\end{CodeOutput}
+
+\begin{CodeOutput}
+#>           (Intercept) ScrArmScreening group 
+#>                -10.17                 -0.22
+\end{CodeOutput}
+
+\begin{CodeOutput}
+#>           (Intercept)   log(Follow.Up.Time) ScrArmScreening group 
+#>                 -9.44                  1.07                 -0.23
+\end{CodeOutput}
 \end{CodeChunk}
@@
 excluding time from the linear predictor.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
-R+                                                   data = ERSPC, 
-R+                                                   ratio = 100)
-R> 
-R> summary(casebase_exponential)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> 
-#> Call:
-#> glm(formula = formula, family = binomial, data = sampleData)
-#> 
-#> Deviance Residuals: 
-#>    Min      1Q  Median      3Q     Max  
-#> -0.148  -0.148  -0.148  -0.133   3.078  
-#> 
-#> Coefficients:
-#>                       Estimate Std. Error z value            Pr(>|z|)    
-#> (Intercept)            -7.7659     0.0557 -139.46 <0.0000000000000002 ***
-#> ScrArmScreening group  -0.2147     0.0884   -2.43               0.015 *  
-#> ---
-#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-#> 
-#> (Dispersion parameter for binomial family taken to be 1)
-#> 
-#>     Null deviance: 6059  on 54539  degrees of freedom
-#> Residual deviance: 6053  on 54538  degrees of freedom
-#> AIC: 6057
-#> 
-#> Number of Fisher Scoring iterations: 7
-\end{CodeOutput}
-
-\begin{CodeInput}
-R> exp(coef(casebase_exponential)[2])
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> ScrArmScreening group 
-#>                  0.81
-\end{CodeOutput}
-
-\begin{CodeInput}
-R> exp(confint(casebase_exponential)[2,])
-\end{CodeInput}
-
-\begin{CodeOutput}
-#>  2.5 % 97.5 % 
-#>   0.68   0.96
-\end{CodeOutput}
-\end{CodeChunk}
 
@@
 the Cox model CIF:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
-R+                                           time = seq(0,15,0.1), 
-R+                                           newdata = new_data)
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R+      xlab = "Years since Randomization", 
-R+      ylab = "Cumulative Incidence (%)", 
-R+      fun = "event",
-R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R+                     exp(coef(cox_model)), 
-R+                     exp(confint(cox_model))[1], 
-R+                     exp(confint(cox_model))[2]),
-R+      yscale = 100)
-R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
-R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
-R> 
-R> 
-R> legend("topleft", 
-R+        legend = c("Control group (Cox)","Control group (Casebase)",
-R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R+        col = c("red","red", "blue","blue"),
-R+        lty = c(1, 2, 1, 2), 
-R+        bg = "gray90")
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/erspc-casebase-cif-1} \end{center}
-
-\end{CodeChunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
-R+                                  data = ERSPC, 
-R+                                  ratio = 100)
-R> 
-R> summary(casebase_time)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> 
-#> Call:
-#> glm(formula = formula, family = binomial, data = sampleData)
-#> 
-#> Deviance Residuals: 
-#>    Min      1Q  Median      3Q     Max  
-#> -0.401  -0.162  -0.122  -0.094   3.466  
-#> 
-#> Coefficients:
-#>                       Estimate Std. Error z value            Pr(>|z|)    
-#> (Intercept)            -9.0314     0.1121  -80.53 <0.0000000000000002 ***
-#> Follow.Up.Time          0.2222     0.0145   15.29 <0.0000000000000002 ***
-#> ScrArmScreening group  -0.2320     0.0886   -2.62              0.0089 ** 
-#> ---
-#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-#> 
-#> (Dispersion parameter for binomial family taken to be 1)
-#> 
-#>     Null deviance: 6059  on 54539  degrees of freedom
-#> Residual deviance: 5811  on 54537  degrees of freedom
-#> AIC: 5817
-#> 
-#> Number of Fisher Scoring iterations: 8
-\end{CodeOutput}
-
-\begin{CodeInput}
-R> exp(coef(casebase_time))
-\end{CodeInput}
-
-\begin{CodeOutput}
-#>           (Intercept)        Follow.Up.Time ScrArmScreening group 
-#>               0.00012               1.24882               0.79297
-\end{CodeOutput}
-
-\begin{CodeInput}
-R> exp(confint(casebase_time))
-\end{CodeInput}
-
-\begin{CodeOutput}
-#>                          2.5 %  97.5 %
-#> (Intercept)           0.000096 0.00015
-#> Follow.Up.Time        1.213876 1.28507
-#> ScrArmScreening group 0.665769 0.94256
-\end{CodeOutput}
-\end{CodeChunk}
-
 Again, the estimate of the hazard ratio is similar to that obtained from
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
-R+                                           time = seq(0,15,0.1), 
-R+                                           newdata = new_data)
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R+      xlab = "Years since Randomization", 
-R+      ylab = "Cumulative Incidence (%)", 
-R+      fun = "event",
-R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R+                     exp(coef(cox_model)), 
-R+                     exp(confint(cox_model))[1], 
-R+                     exp(confint(cox_model))[2]),
-R+      yscale = 100)
-R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
-R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
-R> 
-R> legend("topleft", 
-R+        legend = c("Control group (Cox)","Control group (Casebase)",
-R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R+        col = c("red","red", "blue","blue"),
-R+        lty = c(1, 2, 1, 2), 
-R+        bg = "gray90")
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/erspc-casebase-weibull-cif-1} \end{center}
-
-\end{CodeChunk}
-
 We see that the Weibull model leads to a better fit of the CIF than the
@@
 \code{splines} package:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
-R+                                     data = ERSPC, 
-R+                                     ratio = 100)
-R> 
-R> summary(casebase_splines)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> 
-#> Call:
-#> glm(formula = formula, family = binomial, data = sampleData)
-#> 
-#> Deviance Residuals: 
-#>    Min      1Q  Median      3Q     Max  
-#> -0.405  -0.173  -0.136  -0.085   3.796  
-#> 
-#> Coefficients:
-#>                       Estimate Std. Error z value             Pr(>|z|)    
-#> (Intercept)           -10.2643     0.3159  -32.49 < 0.0000000000000002 ***
-#> bs(Follow.Up.Time)1     4.2983     0.8005    5.37        0.00000007893 ***
-#> bs(Follow.Up.Time)2     2.0945     0.4824    4.34        0.00001412028 ***
-#> bs(Follow.Up.Time)3     4.5544     0.7067    6.44        0.00000000012 ***
-#> ScrArmScreening group  -0.2169     0.0886   -2.45                0.014 *  
-#> ---
-#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-#> 
-#> (Dispersion parameter for binomial family taken to be 1)
-#> 
-#>     Null deviance: 6059.0  on 54539  degrees of freedom
-#> Residual deviance: 5790.5  on 54535  degrees of freedom
-#> AIC: 5800
-#> 
-#> Number of Fisher Scoring iterations: 8
-\end{CodeOutput}
-
-\begin{CodeInput}
-R> exp(coef(casebase_splines))
-\end{CodeInput}
-
-\begin{CodeOutput}
-#>           (Intercept)   bs(Follow.Up.Time)1   bs(Follow.Up.Time)2 
-#>              0.000035             73.578176              8.121029 
-#>   bs(Follow.Up.Time)3 ScrArmScreening group 
-#>             95.053534              0.805017
-\end{CodeOutput}
-
-\begin{CodeInput}
-R> exp(confint(casebase_splines))
-\end{CodeInput}
-
-\begin{CodeOutput}
-#>                           2.5 %     97.5 %
-#> (Intercept)            0.000018   0.000063
-#> bs(Follow.Up.Time)1   15.835008 367.061126
-#> bs(Follow.Up.Time)2    3.252368  21.652009
-#> bs(Follow.Up.Time)3   22.537699 364.141700
-#> ScrArmScreening group  0.675920   0.956823
-\end{CodeOutput}
-\end{CodeChunk}
 
@@
 CIF:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
-R+                                     time = seq(0,15,0.1), 
-R+                                     newdata = new_data)
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R+      xlab = "Years since Randomization", 
-R+      ylab = "Cumulative Incidence (%)", 
-R+      fun = "event",
-R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R+                     exp(coef(cox_model)), 
-R+                     exp(confint(cox_model))[1], 
-R+                     exp(confint(cox_model))[2]),
-R+      yscale = 100)
-R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
-R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
-R> 
-R> legend("topleft", 
-R+        legend = c("Control group (Cox)","Control group (Casebase)",
-R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R+        col = c("red","red", "blue","blue"),
-R+        lty = c(1, 2, 1, 2), 
-R+        bg = "gray90")
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/erspc-casebase-splines-cif-1} \end{center}
-
-\end{CodeChunk}
 
@@
 Exponential & 0.81 & (0.68, 0.96) & 0.28\\
-Weibull & 0.79 & (0.67, 0.94) & 0.28\\
+Gompertz & 0.78 & (0.66, 0.93) & 0.27\\
+Weibull & 0.80 & (0.67, 0.95) & 0.28\\
 Splines & 0.81 & (0.68, 0.96) & 0.28\\
@@
 \begin{CodeChunk}
-
-\begin{CodeInput}
-R> anova(casebase_exponential, casebase_splines, test = "LRT")
-\end{CodeInput}
 
@@
 #> 1     54538       6053                                    
-#> 2     54535       5790  3      262 <0.0000000000000002 ***
+#> 2     54535       5786  3      267 <0.0000000000000002 ***
 #> ---
@@
 
-\begin{CodeInput}
-R> c("Exp." = AIC(casebase_exponential),
-R+   "Weib." = AIC(casebase_time),
-R+   "Splines" = AIC(casebase_splines))
-\end{CodeInput}
-
 \begin{CodeOutput}
-#>    Exp.   Weib. Splines 
-#>    6057    5817    5800
+#>    Exp.   Gomp.   Weib. Splines 
+#>    6057    5816    6057    5796
 \end{CodeOutput}
@@
 \pkg{casebase} package.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(casebase)
-R> data(bmtcrr)
-\end{CodeInput}
-\end{CodeChunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> model_cb <- fitSmoothHazard(
-R+     Status ~ ftime + Sex + D + Phase + Source + Age, 
-R+     data = bmtcrr, 
-R+     ratio = 100, 
-R+     time = "ftime")
-\end{CodeInput}
-\end{CodeChunk}
-
 We will compare our hazard ratio estimates to that obtained from a Cox
@@
 \midrule
-Sex & 0.71 & (0.41, 1.24) & 0.68 & (0.39, 1.19)\\
-Disease & 0.55 & (0.31, 1) & 0.52 & (0.29, 0.93)\\
-Phase (CR2 vs. CR1) & 1.18 & (0.47, 2.93) & 1.21 & (0.48, 3.02)\\
-Phase (CR3 vs. CR1) & 1.47 & (0.38, 5.63) & 1.67 & (0.43, 6.5)\\
-Phase (Relapse vs. CR1) & 4.30 & (2, 9.27) & 4.55 & (2.09, 9.9)\\
+Sex & 0.80 & (0.46, 1.39) & 0.68 & (0.39, 1.19)\\
+Disease & 0.58 & (0.32, 1.03) & 0.52 & (0.29, 0.93)\\
+Phase (CR2 vs. CR1) & 1.16 & (0.46, 2.89) & 1.21 & (0.48, 3.02)\\
+Phase (CR3 vs. CR1) & 1.73 & (0.44, 6.72) & 1.67 & (0.43, 6.5)\\
+Phase (Relapse vs. CR1) & 3.93 & (1.82, 8.49) & 4.55 & (2.09, 9.9)\\
 \addlinespace
-Source & 1.56 & (0.52, 4.7) & 1.46 & (0.47, 4.54)\\
+Source & 1.65 & (0.54, 5.03) & 1.46 & (0.47, 4.54)\\
 Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
@@
 myeloblastic leukemia. We will estimate the curve from 0 to 60 months.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Pick 100 equidistant points between 0 and 60 months
-R> time_points <- seq(0, 60, length.out = 50)
-R> 
-R> # Data.frame containing risk profile
-R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
-R+                                      levels = levels(bmtcrr[,"Sex"])),
-R+                       "D" = c("ALL", "AML"), # Both diseases
-R+                       "Phase" = factor(c("Relapse", "Relapse"), 
-R+                                        levels = levels(bmtcrr[,"Phase"])),
-R+                       "Age" = c(35, 35),
-R+                       "Source" = factor(c("PB", "PB"), 
-R+                                         levels = levels(bmtcrr[,"Source"])))
-R> 
-R> # Estimate absolute risk curve
-R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
-R+                         method = "numerical", newdata = newdata)
-\end{CodeInput}
-\end{CodeChunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Create design matrix without the time variable
-R> support_matrix <- model.matrix(death ~ . - d.time, 
-R+                                data = support)
-R> # Remove intercept term
-R> support_matrix <- support_matrix[,colnames(support_matrix) != "(Intercept)"]
-R> 
-R> # Create test and train dataset
-R> support_test <- support_matrix[indiv,,drop = FALSE]
-R> 
-R> support_train <- support_matrix[-indiv,]
-R> y <- as.matrix(support[-indiv, c("death", "d.time")])
-\end{CodeInput}
-\end{CodeChunk}
-
 We are now ready to estimate the hazard function. We use lasso
@@
 
-\begin{CodeInput}
-R> library(survival)
-R> library(casebase)
-R> 
-R> stanford_popTime <- popTime(jasa, time = "futime", 
-R+                             event = "fustat")
-R> plot(stanford_popTime)
-\end{CodeInput}
-
 \begin{center}\includegraphics{../figures/stanford-poptime-1} \end{center}
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(tidyverse)
-R> library(lubridate)
-R> 
-R> cb_data <- sampleCaseBase(jasa, time = "futime", 
-R+                           event = "fustat", ratio = 10)
-\end{CodeInput}
-\end{CodeChunk}
-
 Next, we will compute the number of days from acceptance into the
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Define exposure variable
-R> cb_data <- mutate(cb_data,
-R+                   txtime = time_length(accept.dt %--% tx.date, 
-R+                                        unit = "days"),
-R+                   exposure = case_when(
-R+                     is.na(txtime) ~ 0L,
-R+                     txtime > futime ~ 0L,
-R+                     txtime <= futime ~ 1L
-R+                   ))
-\end{CodeInput}
-\end{CodeChunk}
-
 Finally, we can fit the hazard using various linear predictors.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(splines)
-R> # Fit several models
-R> fit1 <- fitSmoothHazard(fustat ~ exposure,
-R+                         data = cb_data, time = "futime")
-R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
-R+                         data = cb_data, time = "futime")
-R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
-R+                         data = cb_data, time = "futime")
-R> fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
-R+                         data = cb_data, time = "futime")
-\end{CodeInput}
-\end{CodeChunk}
 
@@
 
-\begin{CodeInput}
-R> # Compute AIC
-R> c("Model1" = AIC(fit1),
-R+   "Model2" = AIC(fit2),
-R+   "Model3" = AIC(fit3),
-R+   "Model4" = AIC(fit4))
-\end{CodeInput}
-
 \begin{CodeOutput}
 #> Model1 Model2 Model3 Model4 
-#>    803    760    733    716
+#>    794    756    728    715
 \end{CodeOutput}
@@
 
-\begin{CodeInput}
-R> # Compute hazards---
-R> # First, create a list of time points for both exposure status
-R> hazard_data <- expand.grid(exposure = c(0, 1),
-R+                            futime = seq(0, 1000,
-R+                                         length.out = 100))
-R> # Set the offset to zero
-R> hazard_data$offset <- 0 
-R> # Use predict to get the fitted values, and exponentiate to 
-R> # transform to the right scale
-R> hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
-R+                                  type = "link"))
-R> # Add labels for plots
-R> hazard_data$Status = factor(hazard_data$exposure,
-R+                             labels = c("NoTrans", "Trans"))
-R> 
-R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
-R+     geom_line() +
-R+     theme_minimal() +
-R+     theme(legend.position = 'top') +
-R+     ylab('Hazard') + xlab('Follow-up time')
-\end{CodeInput}
-
 \begin{center}\includegraphics{../figures/stanford-hazard-1} \end{center}
@@
 
-\begin{CodeInput}
-R> # Compute absolute risk curves
-R> newdata <- data.frame(exposure = c(0, 1))
-R> absrisk <- absoluteRisk(fit4, newdata = newdata, 
-R+                         time = seq(0, 1000, length.out = 100))
-R> 
-R> colnames(absrisk) <- c("Time", "NoTrans", "Trans")
-R> 
-R> # Rearrange the data
-R> absrisk <- gather(as.data.frame(absrisk),
-R+                   "Status", "Risk", -Time)
-R>  
-R> ggplot(absrisk, aes(Time, Risk, colour = Status)) +
-R+   geom_line() +
-R+   theme_minimal() +
-R+   theme(legend.position = 'top') +
-R+   expand_limits(y = 0.5) +
-R+   xlab('Follow-up time') + ylab('Cum. Incidence')
-\end{CodeInput}
-
 \begin{center}\includegraphics{../figures/stanford-risk-1} \end{center}
@@
 
-\begin{CodeInput}
-R> # Compute hazards---
-R> # First, create a list of time points for both exposure status
-R> one_yr_haz <- data.frame(futime = seq(0, 365,
-R+                                       length.out = 100))
-R> one_yr_haz <- mutate(one_yr_haz,
-R+                      offset = 0,
-R+                      exposure = if_else(futime < 30, 0, 1))
-R> one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
-R+                                 type = "link"))
-R> ggplot(one_yr_haz, aes(futime, hazard)) +
-R+   geom_line() +
-R+   theme_minimal() +
-R+   theme(legend.position = 'top') +
-R+   ylab('Hazard') + xlab('Follow-up time') +
-R+   geom_vline(xintercept = 30, linetype = 'dashed')
-\end{CodeInput}
-
 \begin{center}\includegraphics{../figures/stanford-1y-haz-1} \end{center}
@@
 
-\begin{CodeInput}
-R> absoluteRisk(fit4, newdata = data.frame(exposure = 0),
-R+              time = 365)
-\end{CodeInput}
-
 \begin{CodeOutput}
 #>         
-#> 365 0.29
+#> 365 0.33
 \end{CodeOutput}
 
-\begin{CodeInput}
-R> 
-R> # Use the trapezoidal rule to estimate the cumulative hazard
-R> cumhaz_est <- pracma::trapz(one_yr_haz$futime, one_yr_haz$hazard)
-R> 1 - exp(-cumhaz_est)
-\end{CodeInput}
-
 \begin{CodeOutput}
-#> [1] 0.21
+#> [1] 0.22
 \end{CodeOutput}
@@
 
-This report was generated on 2020-01-28 14:29:58 using the following
+This report was generated on 2020-01-29 23:33:21 using the following
 computational environment and dependencies:
@@
 \begin{CodeChunk}
-
-\begin{CodeInput}
-R> # which R packages and versions?
-R> devtools::session_info()
-\end{CodeInput}
 
@@
 #>  tz       America/Toronto             
-#>  date     2020-01-28                  
+#>  date     2020-01-29                  
 #> 
@@
 
-\begin{CodeInput}
-R> # what commit is this file at? 
-R> git2r::repository(here::here())
-\end{CodeInput}
-
 \begin{CodeOutput}
 #> Local:    write-popTime /home/sahir/git_repositories/cbpaper
-#> Head:     [c926a9e] 2020-01-21: Merge pull request #15 from sahirbhatnagar/fix-casebase-dependency
+#> Remote:   write-popTime @ origin (https://github.com/sahirbhatnagar/cbpaper.git)
+#> Head:     [d990874] 2020-01-28: cleaning up case study 1. merging into one figure. and echo=FALSE defaults for
 \end{CodeOutput}

### 2020-01-31 analysis/paper/paper.Rmd
cbPaper draft for covars and interactions
draft paper for co-variates and interactions

--- 2020-01-21 analysis/paper/paper.Rmd
@@
 library(casebase)
+#devtools::load_all('../../../casebase')#testing
 library(splines)
@@
 library(kableExtra)
-library(pracma)
 options(digits = 2, scipen = 999)
@@
 
-eval_introduction <- TRUE
+eval_introduction <- FALSE
 eval_theory <- TRUE
-eval_implementation <- TRUE
-eval_cs1 <- TRUE
-eval_cs2 <- TRUE
+eval_implementation <- FALSE
+eval_cs1 <- FALSE
+eval_cs2 <- FALSE
 eval_cs3 <- TRUE
-eval_cs4 <- TRUE
+eval_cs4 <- FALSE
 ```
@@
 
-Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets} and \pkg{survival} contain implementations for competing risks.
+Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, \pkg{cmprsk} and \pkg{survival} contain implementations for competing risks.
 The differentiating factor between these packages and \pkg{casebase} is that \pkg{casebase} handles competing risks through multiple logistic regression.
-\textcolor{red}{There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}.
-Could you incorporate some of their discussion, with proper citations?
-The biggest omission right now is \pkg{cmprsk}, which we actually use below.}
+Note that \pkg{CFC} has extra features outside the scope of the metthods discussed in this paper.
 
@@
 The SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital.
-Imputation was conducted using predictive mean matching and using the \pkg{mice} package in \proglang{R} [@mice].
-Before imputation, there was a total of 1000 individuals and 35 variables.
-After imputation, we have 474 individuals.
-\textcolor{red}{How come we lose individuals during imputation?} Of these, 67.5% experienced the event of interest.
+Imputation was conducted using the \pkg{mice} package in \proglang{R} with default settings [@mice].
+Before imputation, there were a total of 1000 individuals and 35 variables.
+After imputation, we have 690 individuals.
+We lose individuals due to coliniarity.
+If a pair of covariates are deemed colinear to a certain degree, they are excluded from any imputations.
+Here, only edu  income  avtisst*  race  ph*  glucose*  adlp* were imputed.
+Of these, 66.09% experienced the event of interest.
 <!--The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.--> A description of each variable can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}.
-The original data was obtained from the website of the Department of Biostatistics at Vanderbilt University.
+The original data was obtained from the Department of Biostatistics at Vanderbilt University.
 The imputed dataset is available as part of the \pkg{casebase} package.
@@
 The population-time plot for the SUPPORT data appears in Figure \ref{fig:support-poptime}.
-\textcolor{red}{This population-time plot is strange.
-The red dots never seem to go above 200.}
+\textcolor{red}{red dots now go all over.
+percentile_number introduced as a parameter in popTime}
 
@@
 support_popTime <- popTime(support,
-                           time = "d.time", event = "death")
+                           time = "d.time", event = "death",percentile_number = 0.31)
 plot(support_popTime)
@@
 
-```{r supportSetUp, echo = TRUE, eval = eval_cs3}
-support <- dplyr::select(support, -hospdead)
-
-# Randomly select one observation
-indiv <- sample(1:nrow(support), size = 1)
-```
-
-Our main objective is to estimate the cumulative incidence of death for a given covariate profile.
-Given the number of covariates available, we will do this using regularized estimation.
-As \pkg{casebase} makes use of the \pkg{glmnet} package, we will use the function \code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}.
-Specifically, we input a matrix \code{y}, which contains the time and event variables, and a matrix \code{x}, which contains all other variables we would like to include in the model.
+```{r supportSetUp, echo = FALSE, eval = eval_cs3}
+completeData=casebase::support
+#remove hospital dead
+completeData<-completeData[,-c(4)]
+rat=1  #universal ratio for eval_cs3
+#keep one individual out, at random, for which we will develop an 
+#absolute risk curve.
+#sam=sample(1:nrow(completeData), 1)
+#Set workingData to the remaining individuals in the data
+#workingData=completeData[-c(sam),]
+workingCompleteData=model.matrix(death~ .-d.time,data=completeData)[,-c(1)]#remove intercept
+#Create u and xCox, which will be used when fitting Cox with glmnet.
+newData=workingCompleteData
+x=workingCompleteData
+#x and y will be used to fit the casebase model under glmnet.
+#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
+y=data.matrix(completeData[,c(2,5)])
+```
+
+Our main objective is to compute the absolute risk of death for a given set of covariates, using regularization when fitting our model.
+First, we fit a smooth hazard to the data; for the sake of this example, we opted for a linear term for time.
+As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model.
 For the SUPPORT dataset, all factor variables were converted into dummy variables.
-
-```{r support-design, eval = eval_cs3}
-# Create design matrix without the time variable
-support_matrix <- model.matrix(death ~ . - d.time, 
-                               data = support)
-# Remove intercept term
-support_matrix <- support_matrix[,colnames(support_matrix) != "(Intercept)"]
-
-# Create test and train dataset
-support_test <- support_matrix[indiv,,drop = FALSE]
-
-support_train <- support_matrix[-indiv,]
-y <- as.matrix(support[-indiv, c("death", "d.time")])
-```
-
-We are now ready to estimate the hazard function.
-We use lasso penalization; this is achieved by setting \code{alpha} to 1.
-Also, note that the argument \code{formula_time} can be used to specify the function form for time; in our example, we model time linearly.
-
-```{r CBhazard, echo = TRUE, eval = eval_cs3}
-support_cb <- fitSmoothHazard.fit(support_train, y, 
-                                  formula_time = ~ d.time, 
-                                  family = "glmnet", 
-                                  event = "death", time = "d.time", 
-                                  alpha = 1, ratio = 100)
-```
-
-Then, we take our model object and use the \code{absoluteRisk} function to compute our estimate of the cumulative incidence function.
-Since \code{glmnet} fitted the penalized logistic regression for a range of penalty terms, we need to specify which one we want to use for our prediction.
-We select \code{s = "lambda.1se"}, as recommended.
-
-```{r CBabsolute, echo = TRUE, eval = eval_cs3}
-time_points <- seq(1, max(support$d.time), by = 0.5)
-support_risk <- absoluteRisk(support_cb,
-                             time = time_points,
-                             newdata = support_test,
-                             s = "lambda.1se", method = "numerical")
-```
-
-We will compare this curve to a regularized version of Cox regression, and a Kaplan-Meier survival curve.
-Using the \pkg{glmnet} package, we can fit the Cox regression model using regularized estimation.
-
-```{r coxHaz, echo = TRUE, eval = eval_cs3}
-# Create survival object for glmnet
-surv_glmnet <- survival::Surv(time = y[, "d.time"], 
-                              event = y[, "death"])
-
-support_cox <- glmnet::cv.glmnet(x = support_train, 
-                                 y = surv_glmnet,
-                                 family = "cox", alpha = 1)
-```
-
-However, \pkg{glmnet} does not provide any functionality for estimating the baseline hazard, and the functionalities in the \pkg{survival} package do not work on the output from \code{cv.glmnet}.
-To circumvent this issue and get an estimate of the baseline hazard, we followed the following steps.
-First, we fitted a non-regularized Cox model using only the variables selected by \pkg{glmnet}; the baseline hazard will be estimated using this object.
-Then, we replaced the estimates of the regression coefficients by the estimates from \pkg{glmnet}.
-Finally, this modified \code{coxph} object is fed to the \code{survival::survfit} function to obtain the estimated survival function.
-
-```{r coxAbsolute, echo = TRUE, eval = eval_cs3}
-# Identifying selected covariates
-nonzero_covariates_cox <- predict(support_cox, type = "nonzero", 
-                                  s = "lambda.1se")
-nonzero_coef_cox <- coef(support_cox, s = "lambda.1se")
-# Creating a new dataset that only contains the covariates selected by glmnet
-support_cox_data <- as.data.frame(cbind(y,
-                                        support_train[, nonzero_covariates_cox$X1]))
-# Fitting a cox model
-support_cox <- survival::coxph(Surv(time = d.time, event = death) ~ ., 
-                          data = support_cox_data)
-# Replace the estimate with those obtained by glmnet
-support_cox$coefficients <- nonzero_coef_cox@x
-
-# Fitting CI curve for cox+glmnet
-support_cox_risk <- survival::survfit(support_cox,
-                                      newdata = as.data.frame(support_test),
-                                      time = time_points,
-                                      type = "breslow")
-```
-
-We used the \pkg{survival} package to calculate the Kaplan-Meier estimate of the survival function.
+Here, we used lasso penalization by setting alpha to 1.
+
+```{r cbHazardInteractions}, echo = TRUE, eval = eval_cs3}
+xInteractions=x*y[,2]
+colnames(xInteractions) <- paste("d.time*", colnames(xInteractions), sep = "")
+xFull=cbind(x,xInteractions)
+cb.ModelInter=casebase::fitSmoothHazard.fit(xFull,y,family="glmnet",time="d.time",event="death",
+                                       formula_time = ~d.time,alpha=1,ratio=rat)
+```
+Then, we take our model and use the absoluteRisk function to integrate and retrieve our absolute risk function.
+
+```{r CBabsoluteInteracts, echo = TRUE, eval = eval_cs3}
+#Estimating the absolute risk curve using the newData parameter.
+casebaseAbsoluteInter=casebase::absoluteRisk(cb.ModelInter,time = seq(1,max(completeData$d.time), 400),newdata = xFull , s="lambda.1se",method=c("numerical"))
+
+cabI=data.frame(casebaseAbsoluteInter[,1],rowMeans(casebaseAbsoluteInter[,-c(1)]))
+```
+We will compare this curve to a regularized version of cox regression, and a kaplan-meier survival curve.
+The cox regression absolute risk curve with lasso penalization required two extra steps to retrieve.
+Coxnet from the glmnet package has tools to fit a regularized cox model, but requires that a survival object skeleton of the selected variables, so that the survival package tools can be used to retrive the absolute risk.
+We handle this by fitting a regularized cox model with the glmnet package, and a cox model from the survival package with the selected variables from the regularized fit.
+This second model serves as a skeleton that permits the use of survival package functions.
+The coefficients from the regularized model will replace the coeficients in our skeleton.
+the resulting model will have the correct coefficients when integrating, but will have an incorrect standard error.
+For the purposes of this case study, we are only interested in the absolute risk curve itself and not the standard error.
+
+```{r coxHazAbsoluteInteracts, echo = TRUE, eval = eval_cs3}
+#Create u and xCox, which will be used when fitting Cox with glmnet.
+u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
+
+xCoxI=as.matrix(cbind(data.frame("(Intercept)"=rep(1,each=690)),xFull))
+colnames(xCoxI)[1]="(Intercept)"
+
+#hazard for cox using glmnet
+coxNetI=glmnet::cv.glmnet(x=xCoxI,y=u, family="cox",alpha=1)
+#convergence demonstrated in plot
+#plot(coxNet)
+#taking the coefficient estimates for later use
+nonzero_covariate_cox <- predict(coxNetI, type = "nonzero", s = "lambda.1se")
+nonzero_coef_cox <- coef(coxNetI, s = "lambda.1se")
+#creating a new dataset that only contains the covariates chosen through glmnet.
+#cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
+cleanCoxDataI<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,nonzero_covariate_cox$X1]))
+#newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
+#fitting a cox model using regular estimation, however we will not keep it.
+#this is used more as an object place holder.
+coxNetFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxDataI)
+#The coefficients of this object will be replaced with the estimates from coxNet.
+#Doing so makes it so that everything is invalid aside from the coefficients.
+#In this case, all we need to estimate the absolute risk is the coefficients.
+#Std. error would be incorrect here, if we were to draw error bars.
+coxNetFitI$coefficients<-nonzero_coef_cox@x
+#Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
+#posCox=nonzero_covariate_cox$X1
+#newDataCoxI=cleanCoxDataI[,posCox]
+
+abcoxNetI<-survival::survfit(coxNetFitI,type="breslow",newdata=cleanCoxDataI)
+
+abCNI<-abcoxNetI
+abCNI$surv=rowMeans(abCNI$surv)
+```
+
+```{r unpenalizedCox, echo = TRUE, eval = eval_cs3}
+fullDataCox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,-c(1)]))
+coxFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = fullDataCox)
+abcoxFitI<-survival::survfit(coxFitI,newdata=fullDataCox)
+abCFI<-abcoxFitI
+abCFI$surv<-rowMeans(abCFI$surv)
+#cb.ModelNorm=casebase::fitSmoothHazard(V2~.,time="V1",event="V2",
+       #                                ratio=rat,data=fullDataCox)
+#summary(cb.ModelNorm)#need to figure out how to deal with NAs
+#casebaseAbsoluteNorm=casebase::absoluteRisk(cb.ModelNorm,time = seq(1,max(completeData$d.time), 200),newdata = newData , s="lambda.1se",method=c("numerical"))
+#cabn=data.frame(casebaseAbsoluteNorm[,1],rowMeans(casebaseAbsoluteNorm[,-c(1)]))
+
+```
+We used the survival package to calculate the kaplan-meier absolute risk function.
 
 ```{r fitKM+abRisk, echo = TRUE, eval = eval_cs3}
-km <- survival::Surv(time = support$d.time, 
-                     event = support$death)
-abKm <- survival::survfit(km ~ 1,
-                          type = 'kaplan-meier')
-```
-
-Now that our three cumulative incidence functions have been calculated, we compare them all in Figure \ref{fig:abSC}.
-Both \pkg{glmnet} and \pkg{casebase} have a lower estimated cumulative incidence by the end of the study, when compared to Kaplan-Meier.
-\textcolor{red}{It would be good to have an idea of why the case-base and Coxnet curves are so different.
-At the very least, I think we should compare the select covariates.}
-
-```{r abSupportComparison, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Comparison of cumulative incidence curves between regularized casebase in black, regularized Cox regression in red and Kaplan-Meier in blue."}
-plot(support_risk, type='l', col = "black", lwd = 2, 
-     main = "Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves", 
-     xlab = "Survival-Time", ylab = "Cumulative Incidence", ylim = c(0,1))
-lines(support_cox_risk, col = "red", fun = "event", lwd = 3, conf.int = FALSE)
-lines(abKm, col = "blue", fun = "event", lwd = 3, conf.int = FALSE)
-legend("bottom", 
-       legend = c("casebase", "Cox+glmnet", "KM curve"), 
-       col = c("black", "red", "blue"), lty = c(1, 1, 1), bg = "gray90")
+#creating a surv object to be used to fit an unadjusted absolute risk curve.
+# (Kaplan-Meier risk curve)
+km <- survival::Surv(time = completeData$d.time, event = completeData$death)
+abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
+
+```
+
+Now that our three absolute risk functions have been calculated, we compare them all in Figure \ref{fig:abSC}.
+Both coxnet and casebase decrease the absolute risk by the end of the study, in comparison to kaplan-meier.
+```{r abSupportComparison, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
+# A plot to compare all three Absolute risk (Commulative Incidence)
+plot(cabI[,1],cabI[,2],type='l',col="black",lwd=3,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
+lines(abCFI,col="red",fun="event",lwd=3,conf.int = FALSE)
+lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
+lines(abCNI,col="purple",fun="event",lwd=3,conf.int = FALSE)
+legend("bottomright", 
+       legend = c( "Casebase (Lasso+linear)","semi-parametric Coxnet","KM curve","cox"), 
+       col = c("black","red","Blue","purple"),
+       lty = c(1, 1, 1), 
+       bg = "gray90")
+```
+
+```{r covariatesRegularizedLollipopPlot,eval=FALSE}
+
+#get covariates excluding d.time
+estimatescb=setDT(as.data.frame(coef(cb.Model)[-c(1,2),1]), keep.rownames = TRUE)[]
+estimatescb$Model="casebase"
+colnames(estimatescb)<-c("Coefs","Estimate","Model")
+estimatescox=setDT(as.data.frame(coef(coxNet)[-c(1),1]), keep.rownames = TRUE)[]
+estimatescox$Model="Cox"
+colnames(estimatescox)<-c("Coefs","Estimate","Model")
+estimates=rbind(estimatescb, estimatescox)
+estimates$Estimate[estimates$Estimate==0]=NA# make it so unchosen covariates appear empty
+library(ggstance)# for vertical dodge
+  ggplot(estimates, aes(x = Estimate, y = Coefs, color = Model)) +
+        geom_segment(aes(x = 0, y = Coefs, xend = Estimate, yend = Coefs),size=0.5 ) +
+        geom_point(position=ggstance::position_dodgev(height=0.2))
+
+```
+
+```{r covariatesLolliInteractions,eval=eval_cs3}
+#get covariates excluding d.time
+library(data.table)
+estimatescb=setDT(as.data.frame(coef(cb.ModelInter)[-c(1,2),1]), keep.rownames = TRUE)[]
+estimatescb$Model="casebase"
+colnames(estimatescb)<-c("Coefs","Estimate","Model")
+estimatescox=setDT(as.data.frame(coef(coxNetI)[-c(1),1]), keep.rownames = TRUE)[]
+estimatescox$Model="Cox"
+colnames(estimatescox)<-c("Coefs","Estimate","Model")
+estimatesI=rbind(estimatescb, estimatescox)
+estimatesI$Estimate[estimatesI$Estimate==0]=NA# make it so unchosen covariates appear empty
+library(ggstance)# for vertical dodge
+  ggplot(estimatesI, aes(x = Coefs, y = Estimate , color = Model)) +
+        geom_segment(aes(y = 0, x = Coefs, yend = Estimate, xend = Coefs),size=0.5, position = position_dodge(width = 0.5)) +
+        geom_point(position = position_dodge(width = 0.5))
+
 ```
@@
 # Use the trapezoidal rule to estimate the cumulative hazard
-cumhaz_est <- pracma::trapz(one_yr_haz$futime, one_yr_haz$hazard)
+min_hazard <- min(one_yr_haz$hazard)
+max_hazard <- max(one_yr_haz$hazard)
+increment <- 365/99
+cumhaz_est <- 0.5*increment*(sum(2*one_yr_haz$hazard) - min_hazard - max_hazard)
 1 - exp(-cumhaz_est)
@@
 In this article, we presented the \proglang{R} package \pkg{casebase} that provides functions to fit smooth parametric hazards and estimate cumulative incidence functions using case-base sampling.
-We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through four case studies.
-
-As a methodological framework, case-base sampling is very flexible.
-This flexibility has been explored before in the literature: for example, Saarela and Hanley [-@saarela2015case] used case-base sampling to model a time-dependent exposure variable in a vaccine safety study.
-As another example, Saarela and Arjas [-@saarela2015non] combined case-base sampling and a Bayesian non-parametric framework to compute individualized risk assessments for chronic diseases.
-In the case studies above, we explored this flexibility along two fronts.
-On the one hand, we showed how splines could be used as part of the linear predictor to model the effect of time on the hazard.
-This strategy yielded estimates of the survival function that were qualitatively similar to semiparametric estimates derived from Cox regression; however, case-base sampling led to estimates that *varying smoothly in time*.
-On the other hand, we also displayed the flexibility of case-base sampling by showing how it could be combined with penalized logistic regression to perform variable selection.
-Similarly, our package \pkg{casebase} provides the user with the functionality of Generalized Additive Models.
-Moreover, case-base sampling can be combined with quasi-likelihood estimation to fit survival models that potentially account for the presence of over-dispersion.
-All of the examples illustrate how the case-base sampling framework in general, and the package \pkg{casebase} in particular, allows the user to fit a broad family of survival functions.
-
-As presented in Hanley & Miettinen [-@hanley2009fitting], case-base sampling is comprised of three steps: 1) sampling a case series and a base series from the study; 2) fit the log-hazard as a linear function of predictors (including time); and 3) use the fitted hazard to estimate the cumulative incidence curve.
-Accordingly, our package provides functions for each step.
-Moreover, the simple interface of the \code{fittingSmoothHazard} function resembles the \code{glm} interface, which should look familiar to new users.
-Our modular approach also provides a convenient way to extend our package for new sampling or fitting strategies.
-
-In the case studies above, we compared the performance of case-base sampling with that of Cox regression and Fine-Gray models.
-In terms of function interface, \pkg{casebase} uses a formula interface that is closer to that of \code{glm}, in that the event variable is the only variable appearing on the left-hand side of the formula.
-By contrast, both \code{survival::coxph} and \code{timereg::comp.risk} use arrays as the left-hand side that captures both the event type and time.
-Both approaches to modeling yield user-friendly code.
-In terms of output, both approaches differ significantly.
-Case-base sampling produces smooth hazard and smooth cumulative incidence curves, whereas Cox regression and Fine-Gray models produce step-wise cumulative incidence functions and never explicitly model the hazard function.
-Qualitatively, we showed that by using splines in the linear predictor, all three models yielded similar curves.
-However, the smooth nature of the output of \pkg{casebase} improves interpretation for consumers of these predictions.
-In Table \ref{tab:compCBvsCox}, we provide a side-by-side comparison between the Cox model and case-base sampling.
-
-\begin{table}
-\caption{\label{tab:compCBvsCox}Comparison between the Cox model and case-base sampling}
-\centering
-\begin{tabular}[t]{llp{5cm}}
-\toprule
-Feature & Cox model & Case-base sampling\\
-\midrule
-Model type & Semi-parametric & Fully parametric\\
-Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\
-Cumulative incidence & Step function & Smooth-in-time curve\\
-Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\
-Model testing &  & Use GLM framework \newline (e.g.\ LRT, AIC, BIC)\\
-\addlinespace
-Competing risks & Difficult & Cause-specific CIFs\\
-% Prediction & Kaplan-Meier-based & ROC, AUC, risk reclassification probabilities\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-Our choice of modeling the log-hazard as a linear function of covariates allows us to develop a simple computational scheme for estimation.
-However, as a downside, it does not allow us to model location and scale parameters separately like the package \pkg{flexsurv}.
-For example, if we look at the Weibull distribution as parametrised in \code{stats::pweibull}, the log-hazard function is given by
-$$ \log h(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,$$
-where $\alpha,\beta$ are shape and scale parameters, respectively.
-Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} allows the user to model the scale parameter as a function of covariates.
-Of course, this added flexibility comes at the cost of interpretability: by modeling the log-hazard directly, the parameter estimates from \pkg{casebase} can be interpreted as estimates of log-hazard ratios.
-To improve the flexibility of \pkg{casebase} at capturing the scale of a parametric family, we could replace the logistic regression with its quasi-likelihood counterpart and therefore model over- and under-dispersion with respect to the logistic likelihood.
-We defer the study of the properties and performance of such a model to a future article.
-
-Future work will look at some of the methodological extensions of case-base sampling.
-First, to assess the quality of the model fit, we may want to study the properties of the residuals (e.g.
-Cox-Snell, martingale).
-More work needs to be done to understand these residuals in the context of the partial likelihood underlying case-base sampling.
-The resulting diagnostic tools would then be integrated in this package.
-Also, we are interested in extending case-base sampling to account for interval censoring.
-This type of censoring is very common in longitudinal studies, and many packages (e.g.
-\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2}) provide functions to account for it.
-Again, we hope to include any resulting methodology as part of this package.
-
-In future versions of the package, we also want to increase the complement of diagnostic and inferential tools that are currently available.
-For example, we would like to include the ability to compute confidence intervals for the cumulative incidence curve.
-The delta method or parametric bootstrap are two different strategies we can use to construct approximate confidence intervals.
-Furthermore, we would like to include functions to compute calibration and discrimination statistics (e.g.
-Brier score, AUC) for our models.
-Saarela and Arjas [-@saarela2015non] also describe how to obtain a posterior distribution for the AUC from their model.
-
-In conclusion, we presented the \proglang{R} package \pkg{casebase} which implements case-base sampling for fitting parametric survival models and for estimating smooth cumulative incidence functions using the framework of generalized linear models.
-We strongly believe that its flexibility and its foundation on the familiar logistic regression model will make it appealing to new and established practioners.
-
-```{r, echo=FALSE, eval = FALSE}
+We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through three case studies.
+
+<!-- In the following table we provide a comparison between the Cox model and case-base sampling: -->
+
+```{r, echo=FALSE, eval = FALSE, results='asis'}
 knitr::kable(data.frame(`feature` = c("model type", "time", "cumulative incidence", "non-proportional hazards","model testing","competing risks", "prediction"),
@@
            "smooth-in-time curve", "interaction of covariates with time", "make use of GLM framework (LRT, AIC, BIC)",
-           "cause-specific cumulative incidence functions (CIFs) directly obtained via multinomial regression", "ROC, AUC, risk reclassification probabilities")), 
-           col.names = c("Feature", "Cox model", "Case-base sampling"),
-           format = "latex", booktabs = TRUE,
-           caption = "Comparison between the Cox model and case-base sampling", 
-           escape = FALSE)
-```
+           "cause-specific cumulative incidence functions (CIFs) directly obtained via multinomial regression", "ROC, AUC, risk reclassification probabilities")), escape = FALSE)
+```
+
+\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval censoring, whereas \pkg{casebase} currently cannot.
+Interval censoring is a simple feature to implement and is a potential improvement to add to \pkg{casebase}.
 

### 2020-01-31 analysis/paper/paper.tex
cbPaper draft for covars and interactions
draft paper for co-variates and interactions

@@
 Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
-\pkg{mets} and \pkg{survival} contain implementations for competing
-risks. The differentiating factor between these packages and
+\pkg{mets}, \pkg{cmprsk} and \pkg{survival} contain implementations for
+competing risks. The differentiating factor between these packages and
 \pkg{casebase} is that \pkg{casebase} handles competing risks through
-multiple logistic regression.
-\textcolor{red}{There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}.
-Could you incorporate some of their discussion, with proper citations?
-The biggest omission right now is \pkg{cmprsk}, which we actually use below.}
+multiple logistic regression. Note that \pkg{CFC} has extra features
+outside the scope of the metthods discussed in this paper.
 
@@
 
+\begin{table}
+  \begin{tabular}{p{1.8in}lll}
+\hline
+    &  Parameters &  Density \proglang{R} function & \code{dist}\\
+    & {\footnotesize{(location in \code{\color{red}{red}})}} & & \\
+\hline
+    Exponential & \code{\color{red}{rate}}             & \code{dexp}   & \code{"exp"} \\
+    Weibull (accelerated failure time)     & \code{shape, {\color{red}{scale}}}     & \code{dweibull} & \code{"weibull"} \\
+    Weibull (proportional hazards)     & \code{shape, {\color{red}{scale}}}     & \code{dweibullPH} & \code{"weibullPH"} \\
+    Gamma       & \code{shape, \color{red}{rate}}      & \code{dgamma} & \code{"gamma"}\\
+    Log-normal  & \code{{\color{red}{meanlog}}, sdlog}   & \code{dlnorm} & \code{"lnorm"}\\
+    Gompertz    & \code{shape, {\color{red}{rate}}}      & \code{dgompertz} & \code{"gompertz"} \\
+    Log-logistic & \code{shape, {\color{red}{scale}}}   & \code{dllogis} & \code{"llogis"}\\
+    Generalized gamma (Prentice 1975)   & \code{{\color{red}{mu}}, sigma, Q} & \code{dgengamma} & \code{"gengamma"} \\
+    Generalized gamma (Stacy 1962)& \code{shape, {\color{red}{scale}}, k} & \code{dgengamma.orig} & \code{"gengamma.orig"} \\
+    Generalized F     (stable)    & \code{{\color{red}{mu}}, sigma, Q, P} & \code{dgenf} & \code{"genf"} \\
+    Generalized F     (original)  & \code{{\color{red}{mu}}, sigma, s1, s2} & \code{dgenf.orig} & \code{"genf.orig"} \\
+\hline
+  \end{tabular}
+  \caption{Built-in parametric survival distributions in \pkg{flexsurv}.}
+  \label{tab:dists}
+\end{table}
+
 \hypertarget{implementation-details}{%
@@
 
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> data(ERSPC)
+R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
+R>                        levels = c(0, 1), 
+R>                        labels = c("Control group", "Screening group"))
+\end{CodeInput}
+\end{CodeChunk}
+
 The results of this study were published by
@@
 
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
+R> plot(pt_object)
+\end{CodeInput}
+\end{CodeChunk}
+
 We can also create exposure stratified plots by specifying the
@@
 
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> pt_object_strat <- casebase::popTime(ERSPC, 
+R>                                      event = "DeadOfPrCa", 
+R>                                      exposure = "ScrArm")
+R> plot(pt_object_strat)
+\end{CodeInput}
+\end{CodeChunk}
+
 We can also plot them side-by-side using the \code{ncol} argument:
@@
 
-\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-side-1} \end{center}
-
+\begin{CodeInput}
+R> plot(pt_object_strat, ncol = 2)
+\end{CodeInput}
 \end{CodeChunk}
@@
 
-\begin{CodeOutput}
-#> Call:
-#> survival::coxph(formula = Surv(Follow.Up.Time, DeadOfPrCa) ~ 
-#>     ScrArm, data = ERSPC)
-#> 
-#>   n= 159893, number of events= 540 
-#> 
-#>                         coef exp(coef) se(coef)     z Pr(>|z|)  
-#> ScrArmScreening group -0.222     0.801    0.088 -2.52    0.012 *
-#> ---
-#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-#> 
-#>                       exp(coef) exp(-coef) lower .95 upper .95
-#> ScrArmScreening group     0.801       1.25     0.674     0.952
-#> 
-#> Concordance= 0.519  (se = 0.011 )
-#> Likelihood ratio test= 6.45  on 1 df,   p=0.01
-#> Wald test            = 6.37  on 1 df,   p=0.01
-#> Score (logrank) test = 6.39  on 1 df,   p=0.01
-\end{CodeOutput}
+\begin{CodeInput}
+R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
+R>                              data = ERSPC)
+R> summary(cox_model)
+\end{CodeInput}
 \end{CodeChunk}
@@
 \begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/erspc-cox-cif-1}
-
-}
-
-\caption[Cumulative incidence functions for control and screening groups in the ERSPC data]{Cumulative incidence functions for control and screening groups in the ERSPC data.
-In each of the panels, we plot the CIF from the Cox model (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression.
-(A) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution.
-(B) Linear function of time which is equivalent to a Gompertz distribution.
-(C) The natural logarithm of time which is equivalent to a Weibull distribution.
-(D) Cubic b-spline expansion of time.}\label{fig:erspc-cox-cif}
-\end{figure}
-\end{CodeChunk}
-
-\hypertarget{equivalence-between-casebase-and-aft}{%
-\subsection{Equivalence between casebase and
-AFT}\label{equivalence-between-casebase-and-aft}}
-
-\begin{CodeChunk}
-
-\begin{CodeOutput}
-#>           (Intercept) ScrArmScreening group 
-#>                 -7.77                 -0.22
-\end{CodeOutput}
-
-\begin{CodeOutput}
-#>           (Intercept) ScrArmScreening group 
-#>                 -7.76                 -0.23
-\end{CodeOutput}
-
-\begin{CodeOutput}
-#>           (Intercept) ScrArmScreening group 
-#>                -10.17                 -0.22
-\end{CodeOutput}
-
-\begin{CodeOutput}
-#>           (Intercept)   log(Follow.Up.Time) ScrArmScreening group 
-#>                 -9.44                  1.07                 -0.23
-\end{CodeOutput}
+
+\begin{CodeInput}
+R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
+R>                        ignore = 99)
+R> 
+R> plot(survfit(cox_model, newdata = new_data),
+R>      xlab = "Years since Randomization", 
+R>      ylab = "Cumulative Incidence (%)", 
+R>      fun = "event",
+R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R>                     exp(coef(cox_model)), 
+R>                     exp(confint(cox_model))[1], 
+R>                     exp(confint(cox_model))[2]),
+R>      ylim = c(0, 2), yscale = 100)
+R> legend("topleft", 
+R>        legend = c("Control group", "Screening group"), 
+R>        col = c("red","blue"),
+R>        lty = c(1, 1), 
+R>        bg = "gray90")
+\end{CodeInput}
 \end{CodeChunk}
@@
 excluding time from the linear predictor.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
+R>                                                   data = ERSPC, 
+R>                                                   ratio = 100)
+R> 
+R> summary(casebase_exponential)
+R> exp(coef(casebase_exponential)[2])
+R> exp(confint(casebase_exponential)[2,])
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 the Cox model CIF:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
+R>                                           time = seq(0,15,0.1), 
+R>                                           newdata = new_data)
+R> 
+R> plot(survfit(cox_model, newdata = new_data),
+R>      xlab = "Years since Randomization", 
+R>      ylab = "Cumulative Incidence (%)", 
+R>      fun = "event",
+R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R>                     exp(coef(cox_model)), 
+R>                     exp(confint(cox_model))[1], 
+R>                     exp(confint(cox_model))[2]),
+R>      ylim = c(0, 2), yscale = 100)
+R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
+R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
+R> 
+R> 
+R> legend("topleft", 
+R>        legend = c("Control group (Cox)","Control group (Casebase)",
+R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R>        col = c("red","red", "blue","blue"),
+R>        lty = c(1, 2, 1, 2), 
+R>        bg = "gray90")
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
+R>                                  data = ERSPC, 
+R>                                  ratio = 100)
+R> 
+R> summary(casebase_time)
+R> exp(coef(casebase_time))
+R> exp(confint(casebase_time))
+\end{CodeInput}
+\end{CodeChunk}
+
 Again, the estimate of the hazard ratio is similar to that obtained from
@@
 
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
+R>                                           time = seq(0,15,0.1), 
+R>                                           newdata = new_data)
+R> 
+R> plot(survfit(cox_model, newdata = new_data),
+R>      xlab = "Years since Randomization", 
+R>      ylab = "Cumulative Incidence (%)", 
+R>      fun = "event",
+R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R>                     exp(coef(cox_model)), 
+R>                     exp(confint(cox_model))[1], 
+R>                     exp(confint(cox_model))[2]),
+R>      ylim = c(0, 2), yscale = 100)
+R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
+R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
+R> 
+R> legend("topleft", 
+R>        legend = c("Control group (Cox)","Control group (Casebase)",
+R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R>        col = c("red","red", "blue","blue"),
+R>        lty = c(1, 2, 1, 2), 
+R>        bg = "gray90")
+\end{CodeInput}
+\end{CodeChunk}
+
 We see that the Weibull model leads to a better fit of the CIF than the
@@
 \code{splines} package:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
+R>                                     data = ERSPC, 
+R>                                     ratio = 100)
+R> 
+R> summary(casebase_splines)
+R> exp(coef(casebase_splines))
+R> exp(confint(casebase_splines))
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 CIF:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
+R>                                     time = seq(0,15,0.1), 
+R>                                     newdata = new_data)
+R> 
+R> plot(survfit(cox_model, newdata = new_data),
+R>      xlab = "Years since Randomization", 
+R>      ylab = "Cumulative Incidence (%)", 
+R>      fun = "event",
+R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R>                     exp(coef(cox_model)), 
+R>                     exp(confint(cox_model))[1], 
+R>                     exp(confint(cox_model))[2]),
+R>      ylim = c(0, 2), yscale = 100)
+R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
+R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
+R> 
+R> legend("topleft", 
+R>        legend = c("Control group (Cox)","Control group (Casebase)",
+R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R>        col = c("red","red", "blue","blue"),
+R>        lty = c(1, 2, 1, 2), 
+R>        bg = "gray90")
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 
-\begin{CodeChunk}
-\begin{table}[H]
-\centering
-\begin{tabular}{lrlr}
-\toprule
-Model & HR & CI & Width\\
-\midrule
-Cox & 0.80 & (0.67, 0.95) & 0.28\\
-Exponential & 0.81 & (0.68, 0.96) & 0.28\\
-Gompertz & 0.78 & (0.66, 0.93) & 0.27\\
-Weibull & 0.80 & (0.67, 0.95) & 0.28\\
-Splines & 0.81 & (0.68, 0.96) & 0.28\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\end{CodeChunk}
-
 As noted above, the usual asymptotic results hold for likelihood ratio
@@
 
-\begin{CodeOutput}
-#> Analysis of Deviance Table
-#> 
-#> Model 1: DeadOfPrCa ~ ScrArm + offset(offset)
-#> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
-#>   Resid. Df Resid. Dev Df Deviance            Pr(>Chi)    
-#> 1     54538       6053                                    
-#> 2     54535       5786  3      267 <0.0000000000000002 ***
-#> ---
-#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{CodeOutput}
+\begin{CodeInput}
+R> anova(casebase_exponential, casebase_splines, test = "LRT")
+\end{CodeInput}
 \end{CodeChunk}
@@
 
-\begin{CodeOutput}
-#>    Exp.   Gomp.   Weib. Splines 
-#>    6057    5816    6057    5796
-\end{CodeOutput}
+\begin{CodeInput}
+R> c("Exp." = AIC(casebase_exponential),
+R>   "Weib." = AIC(casebase_time),
+R>   "Splines" = AIC(casebase_splines))
+\end{CodeInput}
 \end{CodeChunk}
@@
 \pkg{casebase} package.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> 
+R> data(bmtcrr)
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 
-\begin{CodeChunk}
-\begin{table}
-
-\caption{\label{tab:table1bmtcrr}Baseline characteristics of patients in the stem-cell transplant study.}
-\centering
-\begin{tabular}[t]{lll}
-\toprule
-Variable & Description & Statistical summary\\
-\midrule
-\code{Sex} & Sex & M=Male (100)\\
- &  & F=Female (77)\\
-\code{D} & Disease & ALL (73)\\
- &  & AML (104)\\
-\code{Phase} & Phase & CR1 (47)\\
-\addlinespace
- &  & CR2 (45)\\
- &  & CR3 (12)\\
- &  & Relapse (73)\\
-\code{Source} & Type of transplant & BM+PB (21)\\
- &  & PB (156)\\
-\addlinespace
-\code{Age} & Age of patient (years) & 4--62\\
- &  & 30.47 (13.04)\\
-\code{ftime} & Failure time (months) & 0.13--131.77\\
- &  & 20.28 (30.78)\\
-\code{Status} & Status indicator & 0=censored (46)\\
-\addlinespace
- &  & 1=relapse (56)\\
- &  & 2=competing event (75)\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\end{CodeChunk}
-
 First, we can look at a population-time plot to visualize the incidence
@@
 
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> model_cb <- fitSmoothHazard(
+R>     Status ~ ftime + Sex + D + Phase + Source + Age, 
+R>     data = bmtcrr, 
+R>     ratio = 100, 
+R>     time = "ftime")
+\end{CodeInput}
+\end{CodeChunk}
+
 We will compare our hazard ratio estimates to that obtained from a Cox
@@
 R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-R+                    data = bmtcrr)
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{table}
-
-\caption{\label{tab:bmtcrr-cis}ADD CAPTION}
-\centering
-\begin{tabular}[t]{lrlrl}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
-Covariates & HR & 95\% CI & HR & 95\% CI\\
-\midrule
-Sex & 0.80 & (0.46, 1.39) & 0.68 & (0.39, 1.19)\\
-Disease & 0.58 & (0.32, 1.03) & 0.52 & (0.29, 0.93)\\
-Phase (CR2 vs. CR1) & 1.16 & (0.46, 2.89) & 1.21 & (0.48, 3.02)\\
-Phase (CR3 vs. CR1) & 1.73 & (0.44, 6.72) & 1.67 & (0.43, 6.5)\\
-Phase (Relapse vs. CR1) & 3.93 & (1.82, 8.49) & 4.55 & (2.09, 9.9)\\
-\addlinespace
-Source & 1.65 & (0.54, 5.03) & 1.46 & (0.47, 4.54)\\
-Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
-\bottomrule
-\end{tabular}
-\end{table}
-
+R>                    data = bmtcrr)
+\end{CodeInput}
 \end{CodeChunk}
@@
 myeloblastic leukemia. We will estimate the curve from 0 to 60 months.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Pick 100 equidistant points between 0 and 60 months
+R> time_points <- seq(0, 60, length.out = 50)
+R> 
+R> # Data.frame containing risk profile
+R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
+R>                                      levels = levels(bmtcrr[,"Sex"])),
+R>                       "D" = c("ALL", "AML"), # Both diseases
+R>                       "Phase" = factor(c("Relapse", "Relapse"), 
+R>                                        levels = levels(bmtcrr[,"Phase"])),
+R>                       "Age" = c(35, 35),
+R>                       "Source" = factor(c("PB", "PB"), 
+R>                                         levels = levels(bmtcrr[,"Source"])))
+R> 
+R> # Estimate absolute risk curve
+R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
+R>                         method = "numerical", newdata = newdata)
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 individuals who are considered seriously ill within a hospital.
-Imputation was conducted using predictive mean matching and using the
-\pkg{mice} package in \proglang{R} \citep{mice}. Before imputation,
-there was a total of 1000 individuals and 35 variables. After
-imputation, we have 474 individuals.
-\textcolor{red}{How come we lose individuals during imputation?} Of
-these, 67.5\% experienced the event of interest. A description of each
-variable can be found in Table \ref{tab:support1} and a breakdown of
-each categorical variable in Table \ref{tab:support2}. The original data
-was obtained from the website of the Department of Biostatistics at
-Vanderbilt University. The imputed dataset is available as part of the
-\pkg{casebase} package.
+Imputation was conducted using the \pkg{mice} package in \proglang{R}
+with default settings \citep{mice}. Before imputation, there were a
+total of 1000 individuals and 35 variables. After imputation, we have
+690 individuals. We lose individuals due to coliniarity. If a pair of
+covariates are deemed colinear to a certain degree, they are excluded
+from any imputations. Here, only edu income avtisst* race ph* glucose*
+adlp* were imputed. Of these, 66.09\% experienced the event of interest.
+A description of each variable can be found in Table \ref{tab:support1}
+and a breakdown of each categorical variable in Table
+\ref{tab:support2}. The original data was obtained from the Department
+of Biostatistics at Vanderbilt University. The imputed dataset is
+available as part of the \pkg{casebase} package.
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> support <- dplyr::select(support, -hospdead)
-R> 
-R> # Randomly select one observation
-R> indiv <- sample(1:nrow(support), size = 1)
-\end{CodeInput}
-\end{CodeChunk}
-
-Our main objective is to estimate the cumulative incidence of death for
-a given covariate profile. Given the number of covariates available, we
-will do this using regularized estimation. As \pkg{casebase} makes use
-of the \pkg{glmnet} package, we will use the function
-\code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}.
-Specifically, we input a matrix \code{y}, which contains the time and
-event variables, and a matrix \code{x}, which contains all other
-variables we would like to include in the model. For the SUPPORT
-dataset, all factor variables were converted into dummy variables.
-
-We are now ready to estimate the hazard function. We use lasso
-penalization; this is achieved by setting \code{alpha} to 1. Also, note
-that the argument \code{formula_time} can be used to specify the
-function form for time; in our example, we model time linearly.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> support_cb <- fitSmoothHazard.fit(support_train, y, 
-R+                                   formula_time = ~ d.time, 
-R+                                   family = "glmnet", 
-R+                                   event = "death", time = "d.time", 
-R+                                   alpha = 1, ratio = 100)
-\end{CodeInput}
-\end{CodeChunk}
-
-Then, we take our model object and use the \code{absoluteRisk} function
-to compute our estimate of the cumulative incidence function. Since
-\code{glmnet} fitted the penalized logistic regression for a range of
-penalty terms, we need to specify which one we want to use for our
-prediction. We select \code{s = "lambda.1se"}, as recommended.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> time_points <- seq(1, max(support$d.time), by = 0.5)
-R> support_risk <- absoluteRisk(support_cb,
-R+                              time = time_points,
-R+                              newdata = support_test,
-R+                              s = "lambda.1se", method = "numerical")
-\end{CodeInput}
-\end{CodeChunk}
-
-We will compare this curve to a regularized version of Cox regression,
-and a Kaplan-Meier survival curve. Using the \pkg{glmnet} package, we
-can fit the Cox regression model using regularized estimation.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Create survival object for glmnet
-R> surv_glmnet <- survival::Surv(time = y[, "d.time"], 
-R+                               event = y[, "death"])
-R> 
-R> support_cox <- glmnet::cv.glmnet(x = support_train, 
-R+                                  y = surv_glmnet,
-R+                                  family = "cox", alpha = 1)
-\end{CodeInput}
-\end{CodeChunk}
-
-However, \pkg{glmnet} does not provide any functionality for estimating
-the baseline hazard, and the functionalities in the \pkg{survival}
-package do not work on the output from \code{cv.glmnet}. To circumvent
-this issue and get an estimate of the baseline hazard, we followed the
-following steps. First, we fitted a non-regularized Cox model using only
-the variables selected by \pkg{glmnet}; the baseline hazard will be
-estimated using this object. Then, we replaced the estimates of the
-regression coefficients by the estimates from \pkg{glmnet}. Finally,
-this modified \code{coxph} object is fed to the \code{survival::survfit}
-function to obtain the estimated survival function.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Identifying selected covariates
-R> nonzero_covariates_cox <- predict(support_cox, type = "nonzero", 
-R+                                   s = "lambda.1se")
-R> nonzero_coef_cox <- coef(support_cox, s = "lambda.1se")
-R> # Creating a new dataset that only contains the covariates selected by glmnet
-R> support_cox_data <- as.data.frame(cbind(y,
-R+                                         support_train[, nonzero_covariates_cox$X1]))
-R> # Fitting a cox model
-R> support_cox <- survival::coxph(Surv(time = d.time, event = death) ~ ., 
-R+                           data = support_cox_data)
-R> # Replace the estimate with those obtained by glmnet
-R> support_cox$coefficients <- nonzero_coef_cox@x
-R> 
-R> # Fitting CI curve for cox+glmnet
-R> support_cox_risk <- survival::survfit(support_cox,
-R+                                       newdata = as.data.frame(support_test),
-R+                                       time = time_points,
-R+                                       type = "breslow")
-\end{CodeInput}
-\end{CodeChunk}
-
-We used the \pkg{survival} package to calculate the Kaplan-Meier
-estimate of the survival function.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> km <- survival::Surv(time = support$d.time, 
-R+                      event = support$death)
-R> abKm <- survival::survfit(km ~ 1,
-R+                           type = 'kaplan-meier')
-\end{CodeInput}
-\end{CodeChunk}
-
-Now that our three cumulative incidence functions have been calculated,
-we compare them all in Figure \ref{fig:abSC}. Both \pkg{glmnet} and
-\pkg{casebase} have a lower estimated cumulative incidence by the end of
-the study, when compared to Kaplan-Meier.
-\textcolor{red}{It would be good to have an idea of why the case-base and Coxnet curves are so different. At the very least, I think we should compare the select covariates.}
-
-\begin{CodeChunk}
+Our main objective is to compute the absolute risk of death for a given
+set of covariates, using regularization when fitting our model. First,
+we fit a smooth hazard to the data; for the sake of this example, we
+opted for a linear term for time. As casebase makes use of the glmnet
+package, we interact with the fitsmoothhazard.fit function using a
+matrix interface, where y contains the time and event variables and x
+contains all other variables we would like to include in the model. For
+the SUPPORT dataset, all factor variables were converted into dummy
+variables. Here, we used lasso penalization by setting alpha to 1.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> xInteractions=x*y[,2]
+R> colnames(xInteractions) <- paste("d.time*", colnames(xInteractions), sep = "")
+R> xFull=cbind(x,xInteractions)
+R> cb.ModelInter=casebase::fitSmoothHazard.fit(xFull,y,family="glmnet",time="d.time",event="death",
+R+                                        formula_time = ~d.time,alpha=1,ratio=rat)
+\end{CodeInput}
+\end{CodeChunk}
+
+Then, we take our model and use the absoluteRisk function to integrate
+and retrieve our absolute risk function.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #Estimating the absolute risk curve using the newData parameter.
+R> casebaseAbsoluteInter=casebase::absoluteRisk(cb.ModelInter,time = seq(1,max(completeData$d.time), 400),newdata = xFull , s="lambda.1se",method=c("numerical"))
+R> 
+R> cabI=data.frame(casebaseAbsoluteInter[,1],rowMeans(casebaseAbsoluteInter[,-c(1)]))
+\end{CodeInput}
+\end{CodeChunk}
+
+We will compare this curve to a regularized version of cox regression,
+and a kaplan-meier survival curve. The cox regression absolute risk
+curve with lasso penalization required two extra steps to retrieve.
+Coxnet from the glmnet package has tools to fit a regularized cox model,
+but requires that a survival object skeleton of the selected variables,
+so that the survival package tools can be used to retrive the absolute
+risk. We handle this by fitting a regularized cox model with the glmnet
+package, and a cox model from the survival package with the selected
+variables from the regularized fit. This second model serves as a
+skeleton that permits the use of survival package functions. The
+coefficients from the regularized model will replace the coeficients in
+our skeleton. the resulting model will have the correct coefficients
+when integrating, but will have an incorrect standard error. For the
+purposes of this case study, we are only interested in the absolute risk
+curve itself and not the standard error.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #Create u and xCox, which will be used when fitting Cox with glmnet.
+R> u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
+R> 
+R> xCoxI=as.matrix(cbind(data.frame("(Intercept)"=rep(1,each=690)),xFull))
+R> colnames(xCoxI)[1]="(Intercept)"
+R> 
+R> #hazard for cox using glmnet
+R> coxNetI=glmnet::cv.glmnet(x=xCoxI,y=u, family="cox",alpha=1)
+R> #convergence demonstrated in plot
+R> #plot(coxNet)
+R> #taking the coefficient estimates for later use
+R> nonzero_covariate_cox <- predict(coxNetI, type = "nonzero", s = "lambda.1se")
+R> nonzero_coef_cox <- coef(coxNetI, s = "lambda.1se")
+R> #creating a new dataset that only contains the covariates chosen through glmnet.
+R> #cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
+R> cleanCoxDataI<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,nonzero_covariate_cox$X1]))
+R> #newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
+R> #fitting a cox model using regular estimation, however we will not keep it.
+R> #this is used more as an object place holder.
+R> coxNetFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxDataI)
+R> #The coefficients of this object will be replaced with the estimates from coxNet.
+R> #Doing so makes it so that everything is invalid aside from the coefficients.
+R> #In this case, all we need to estimate the absolute risk is the coefficients.
+R> #Std. error would be incorrect here, if we were to draw error bars.
+R> coxNetFitI$coefficients<-nonzero_coef_cox@x
+R> #Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
+R> #posCox=nonzero_covariate_cox$X1
+R> #newDataCoxI=cleanCoxDataI[,posCox]
+R> 
+R> abcoxNetI<-survival::survfit(coxNetFitI,type="breslow",newdata=cleanCoxDataI)
+R> 
+R> abCNI<-abcoxNetI
+R> abCNI$surv=rowMeans(abCNI$surv)
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> fullDataCox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,-c(1)]))
+R> coxFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = fullDataCox)
+R> abcoxFitI<-survival::survfit(coxFitI,newdata=fullDataCox)
+R> abCFI<-abcoxFitI
+R> abCFI$surv<-rowMeans(abCFI$surv)
+R> #cb.ModelNorm=casebase::fitSmoothHazard(V2~.,time="V1",event="V2",
+R>        #                                ratio=rat,data=fullDataCox)
+R> #summary(cb.ModelNorm)#need to figure out how to deal with NAs
+R> #casebaseAbsoluteNorm=casebase::absoluteRisk(cb.ModelNorm,time = seq(1,max(completeData$d.time), 200),newdata = newData , s="lambda.1se",method=c("numerical"))
+R> #cabn=data.frame(casebaseAbsoluteNorm[,1],rowMeans(casebaseAbsoluteNorm[,-c(1)]))
+\end{CodeInput}
+\end{CodeChunk}
+
+We used the survival package to calculate the kaplan-meier absolute risk
+function.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #creating a surv object to be used to fit an unadjusted absolute risk curve.
+R> # (Kaplan-Meier risk curve)
+R> km <- survival::Surv(time = completeData$d.time, event = completeData$death)
+R> abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
+\end{CodeInput}
+\end{CodeChunk}
+
+Now that our three absolute risk functions have been calculated, we
+compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase
+decrease the absolute risk by the end of the study, in comparison to
+kaplan-meier.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # A plot to compare all three Absolute risk (Commulative Incidence)
+R> plot(cabI[,1],cabI[,2],type='l',col="black",lwd=3,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
+R> lines(abCFI,col="red",fun="event",lwd=3,conf.int = FALSE)
+R> lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
+R> lines(abCNI,col="purple",fun="event",lwd=3,conf.int = FALSE)
+R> legend("bottomright", 
+R+        legend = c( "Casebase (Lasso+linear)","semi-parametric Coxnet","KM curve","cox"), 
+R+        col = c("black","red","Blue","purple"),
+R+        lty = c(1, 1, 1), 
+R+        bg = "gray90")
+\end{CodeInput}
 \begin{figure}
@@
 
-\caption{\label{fig:abSC} Comparison of cumulative incidence curves between regularized casebase in black, regularized Cox regression in red and Kaplan-Meier in blue.}\label{fig:abSupportComparison}
+\caption{\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue.}\label{fig:abSupportComparison}
 \end{figure}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> 
+R> #get covariates excluding d.time
+R> estimatescb=setDT(as.data.frame(coef(cb.Model)[-c(1,2),1]), keep.rownames = TRUE)[]
+R> estimatescb$Model="casebase"
+R> colnames(estimatescb)<-c("Coefs","Estimate","Model")
+R> estimatescox=setDT(as.data.frame(coef(coxNet)[-c(1),1]), keep.rownames = TRUE)[]
+R> estimatescox$Model="Cox"
+R> colnames(estimatescox)<-c("Coefs","Estimate","Model")
+R> estimates=rbind(estimatescb, estimatescox)
+R> estimates$Estimate[estimates$Estimate==0]=NA# make it so unchosen covariates appear empty
+R> library(ggstance)# for vertical dodge
+R>   ggplot(estimates, aes(x = Estimate, y = Coefs, color = Model)) +
+R>         geom_segment(aes(x = 0, y = Coefs, xend = Estimate, yend = Coefs),size=0.5 ) +
+R>         geom_point(position=ggstance::position_dodgev(height=0.2))
+R> 
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #get covariates excluding d.time
+R> library(data.table)
+R> estimatescb=setDT(as.data.frame(coef(cb.ModelInter)[-c(1,2),1]), keep.rownames = TRUE)[]
+R> estimatescb$Model="casebase"
+R> colnames(estimatescb)<-c("Coefs","Estimate","Model")
+R> estimatescox=setDT(as.data.frame(coef(coxNetI)[-c(1),1]), keep.rownames = TRUE)[]
+R> estimatescox$Model="Cox"
+R> colnames(estimatescox)<-c("Coefs","Estimate","Model")
+R> estimatesI=rbind(estimatescb, estimatescox)
+R> estimatesI$Estimate[estimatesI$Estimate==0]=NA# make it so unchosen covariates appear empty
+R> library(ggstance)# for vertical dodge
+R>   ggplot(estimatesI, aes(x = Coefs, y = Estimate , color = Model)) +
+R+         geom_segment(aes(y = 0, x = Coefs, yend = Estimate, xend = Coefs),size=0.5, position = position_dodge(width = 0.5)) +
+R+         geom_point(position = position_dodge(width = 0.5))
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/covariatesLolliInteractions-1} \end{center}
+
 \end{CodeChunk}
@@
 Data \citep[\citet{crowley1977covariance}]{clark1971cardiac}; we will
-use the version available in the \pkg{surivival} package.
+use the version available in the \pkg{survival} package.
 
@@
 
-\begin{center}\includegraphics{../figures/stanford-poptime-1} \end{center}
-
+\begin{CodeInput}
+R> library(survival)
+R> library(casebase)
+R> 
+R> stanford_popTime <- popTime(jasa, time = "futime", 
+R>                             event = "fustat")
+R> plot(stanford_popTime)
+\end{CodeInput}
 \end{CodeChunk}
@@
 
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> library(tidyverse)
+R> library(lubridate)
+R> 
+R> cb_data <- sampleCaseBase(jasa, time = "futime", 
+R>                           event = "fustat", ratio = 10)
+\end{CodeInput}
+\end{CodeChunk}
+
 Next, we will compute the number of days from acceptance into the
@@
 
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Define exposure variable
+R> cb_data <- mutate(cb_data,
+R>                   txtime = time_length(accept.dt %--% tx.date, 
+R>                                        unit = "days"),
+R>                   exposure = case_when(
+R>                     is.na(txtime) ~ 0L,
+R>                     txtime > futime ~ 0L,
+R>                     txtime <= futime ~ 1L
+R>                   ))
+\end{CodeInput}
+\end{CodeChunk}
+
 Finally, we can fit the hazard using various linear predictors.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> library(splines)
+R> # Fit several models
+R> fit1 <- fitSmoothHazard(fustat ~ exposure,
+R>                         data = cb_data, time = "futime")
+R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
+R>                         data = cb_data, time = "futime")
+R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
+R>                         data = cb_data, time = "futime")
+R> fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
+R>                         data = cb_data, time = "futime")
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 
-\begin{CodeOutput}
-#> Model1 Model2 Model3 Model4 
-#>    794    756    728    715
-\end{CodeOutput}
+\begin{CodeInput}
+R> # Compute AIC
+R> c("Model1" = AIC(fit1),
+R>   "Model2" = AIC(fit2),
+R>   "Model3" = AIC(fit3),
+R>   "Model4" = AIC(fit4))
+\end{CodeInput}
 \end{CodeChunk}
@@
 
-\begin{center}\includegraphics{../figures/stanford-hazard-1} \end{center}
-
+\begin{CodeInput}
+R> # Compute hazards---
+R> # First, create a list of time points for both exposure status
+R> hazard_data <- expand.grid(exposure = c(0, 1),
+R>                            futime = seq(0, 1000,
+R>                                         length.out = 100))
+R> # Set the offset to zero
+R> hazard_data$offset <- 0 
+R> # Use predict to get the fitted values, and exponentiate to 
+R> # transform to the right scale
+R> hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
+R>                                  type = "link"))
+R> # Add labels for plots
+R> hazard_data$Status = factor(hazard_data$exposure,
+R>                             labels = c("NoTrans", "Trans"))
+R> 
+R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
+R>     geom_line() +
+R>     theme_minimal() +
+R>     theme(legend.position = 'top') +
+R>     ylab('Hazard') + xlab('Follow-up time')
+\end{CodeInput}
 \end{CodeChunk}
@@
 
-\begin{center}\includegraphics{../figures/stanford-risk-1} \end{center}
-
+\begin{CodeInput}
+R> # Compute absolute risk curves
+R> newdata <- data.frame(exposure = c(0, 1))
+R> absrisk <- absoluteRisk(fit4, newdata = newdata, 
+R>                         time = seq(0, 1000, length.out = 100))
+R> 
+R> colnames(absrisk) <- c("Time", "NoTrans", "Trans")
+R> 
+R> # Rearrange the data
+R> absrisk <- gather(as.data.frame(absrisk),
+R>                   "Status", "Risk", -Time)
+R>  
+R> ggplot(absrisk, aes(Time, Risk, colour = Status)) +
+R>   geom_line() +
+R>   theme_minimal() +
+R>   theme(legend.position = 'top') +
+R>   expand_limits(y = 0.5) +
+R>   xlab('Follow-up time') + ylab('Cum. Incidence')
+\end{CodeInput}
 \end{CodeChunk}
@@
 
-\begin{center}\includegraphics{../figures/stanford-1y-haz-1} \end{center}
-
+\begin{CodeInput}
+R> # Compute hazards---
+R> # First, create a list of time points for both exposure status
+R> one_yr_haz <- data.frame(futime = seq(0, 365,
+R>                                       length.out = 100))
+R> one_yr_haz <- mutate(one_yr_haz,
+R>                      offset = 0,
+R>                      exposure = if_else(futime < 30, 0, 1))
+R> one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
+R>                                 type = "link"))
+R> ggplot(one_yr_haz, aes(futime, hazard)) +
+R>   geom_line() +
+R>   theme_minimal() +
+R>   theme(legend.position = 'top') +
+R>   ylab('Hazard') + xlab('Follow-up time') +
+R>   geom_vline(xintercept = 30, linetype = 'dashed')
+\end{CodeInput}
 \end{CodeChunk}
@@
 
-\begin{CodeOutput}
-#>         
-#> 365 0.33
-\end{CodeOutput}
-
-\begin{CodeOutput}
-#> [1] 0.22
-\end{CodeOutput}
+\begin{CodeInput}
+R> absoluteRisk(fit4, newdata = data.frame(exposure = 0),
+R>              time = 365)
+R> 
+R> # Use the trapezoidal rule to estimate the cumulative hazard
+R> min_hazard <- min(one_yr_haz$hazard)
+R> max_hazard <- max(one_yr_haz$hazard)
+R> increment <- 365/99
+R> cumhaz_est <- 0.5*increment*(sum(2*one_yr_haz$hazard) - min_hazard - max_hazard)
+R> 1 - exp(-cumhaz_est)
+\end{CodeInput}
 \end{CodeChunk}
@@
 implementation, and we illustrated the merits of the approach and the
-package through four case studies.
-
-As a methodological framework, case-base sampling is very flexible. This
-flexibility has been explored before in the literature: for example,
-Saarela and Hanley \citeyearpar{saarela2015case} used case-base sampling
-to model a time-dependent exposure variable in a vaccine safety study.
-As another example, Saarela and Arjas \citeyearpar{saarela2015non}
-combined case-base sampling and a Bayesian non-parametric framework to
-compute individualized risk assessments for chronic diseases. In the
-case studies above, we explored this flexibility along two fronts. On
-the one hand, we showed how splines could be used as part of the linear
-predictor to model the effect of time on the hazard. This strategy
-yielded estimates of the survival function that were qualitatively
-similar to semiparametric estimates derived from Cox regression;
-however, case-base sampling led to estimates that \emph{varying smoothly
-in time}. On the other hand, we also displayed the flexibility of
-case-base sampling by showing how it could be combined with penalized
-logistic regression to perform variable selection. Similarly, our
-package \pkg{casebase} provides the user with the functionality of
-Generalized Additive Models. Moreover, case-base sampling can be
-combined with quasi-likelihood estimation to fit survival models that
-potentially account for the presence of over-dispersion. All of the
-examples illustrate how the case-base sampling framework in general, and
-the package \pkg{casebase} in particular, allows the user to fit a broad
-family of survival functions.
-
-As presented in Hanley \& Miettinen \citeyearpar{hanley2009fitting},
-case-base sampling is comprised of three steps: 1) sampling a case
-series and a base series from the study; 2) fit the log-hazard as a
-linear function of predictors (including time); and 3) use the fitted
-hazard to estimate the cumulative incidence curve. Accordingly, our
-package provides functions for each step. Moreover, the simple interface
-of the \code{fittingSmoothHazard} function resembles the \code{glm}
-interface, which should look familiar to new users. Our modular approach
-also provides a convenient way to extend our package for new sampling or
-fitting strategies.
-
-In the case studies above, we compared the performance of case-base
-sampling with that of Cox regression and Fine-Gray models. In terms of
-function interface, \pkg{casebase} uses a formula interface that is
-closer to that of \code{glm}, in that the event variable is the only
-variable appearing on the left-hand side of the formula. By contrast,
-both \code{survival::coxph} and \code{timereg::comp.risk} use arrays as
-the left-hand side that captures both the event type and time. Both
-approaches to modeling yield user-friendly code. In terms of output,
-both approaches differ significantly. Case-base sampling produces smooth
-hazard and smooth cumulative incidence curves, whereas Cox regression
-and Fine-Gray models produce step-wise cumulative incidence functions
-and never explicitly model the hazard function. Qualitatively, we showed
-that by using splines in the linear predictor, all three models yielded
-similar curves. However, the smooth nature of the output of
-\pkg{casebase} improves interpretation for consumers of these
-predictions. In Table \ref{tab:compCBvsCox}, we provide a side-by-side
-comparison between the Cox model and case-base sampling.
-
-\begin{table}
-\caption{\label{tab:compCBvsCox}Comparison between the Cox model and case-base sampling}
-\centering
-\begin{tabular}[t]{llp{5cm}}
-\toprule
-Feature & Cox model & Case-base sampling\\
-\midrule
-Model type & Semi-parametric & Fully parametric\\
-Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\
-Cumulative incidence & Step function & Smooth-in-time curve\\
-Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\
-Model testing &  & Use GLM framework \newline (e.g.\ LRT, AIC, BIC)\\
-\addlinespace
-Competing risks & Difficult & Cause-specific CIFs\\
-% Prediction & Kaplan-Meier-based & ROC, AUC, risk reclassification probabilities\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-Our choice of modeling the log-hazard as a linear function of covariates
-allows us to develop a simple computational scheme for estimation.
-However, as a downside, it does not allow us to model location and scale
-parameters separately like the package \pkg{flexsurv}. For example, if
-we look at the Weibull distribution as parametrised in
-\code{stats::pweibull}, the log-hazard function is given by
-\[ \log h(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,\]
-where \(\alpha,\beta\) are shape and scale parameters, respectively.
-Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} allows the
-user to model the scale parameter as a function of covariates. Of
-course, this added flexibility comes at the cost of interpretability: by
-modeling the log-hazard directly, the parameter estimates from
-\pkg{casebase} can be interpreted as estimates of log-hazard ratios. To
-improve the flexibility of \pkg{casebase} at capturing the scale of a
-parametric family, we could replace the logistic regression with its
-quasi-likelihood counterpart and therefore model over- and
-under-dispersion with respect to the logistic likelihood. We defer the
-study of the properties and performance of such a model to a future
-article.
-
-Future work will look at some of the methodological extensions of
-case-base sampling. First, to assess the quality of the model fit, we
-may want to study the properties of the residuals (e.g.~Cox-Snell,
-martingale). More work needs to be done to understand these residuals in
-the context of the partial likelihood underlying case-base sampling. The
-resulting diagnostic tools would then be integrated in this package.
-Also, we are interested in extending case-base sampling to account for
-interval censoring. This type of censoring is very common in
-longitudinal studies, and many packages (e.g. \pkg{SmoothHazard},
-\pkg{survival} and \pkg{Rstpm2}) provide functions to account for it.
-Again, we hope to include any resulting methodology as part of this
-package.
-
-In future versions of the package, we also want to increase the
-complement of diagnostic and inferential tools that are currently
-available. For example, we would like to include the ability to compute
-confidence intervals for the cumulative incidence curve. The delta
-method or parametric bootstrap are two different strategies we can use
-to construct approximate confidence intervals. Furthermore, we would
-like to include functions to compute calibration and discrimination
-statistics (e.g.~Brier score, AUC) for our models. Saarela and Arjas
-\citeyearpar{saarela2015non} also describe how to obtain a posterior
-distribution for the AUC from their model.
-
-In conclusion, we presented the \proglang{R} package \pkg{casebase}
-which implements case-base sampling for fitting parametric survival
-models and for estimating smooth cumulative incidence functions using
-the framework of generalized linear models. We strongly believe that its
-flexibility and its foundation on the familiar logistic regression model
-will make it appealing to new and established practioners.
+package through three case studies.
+
+\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval
+censoring, whereas \pkg{casebase} currently cannot. Interval censoring
+is a simple feature to implement and is a potential improvement to add
+to \pkg{casebase}.
 
@@
 
-This report was generated on 2020-01-29 23:33:21 using the following
+This report was generated on 2020-01-30 20:17:21 using the following
 computational environment and dependencies:
@@
 \begin{CodeChunk}
+
+\begin{CodeInput}
+R> # which R packages and versions?
+R> devtools::session_info()
+\end{CodeInput}
 
@@
 #>  setting  value                       
-#>  version  R version 3.6.2 (2019-12-12)
-#>  os       Pop!_OS 19.10               
-#>  system   x86_64, linux-gnu           
+#>  version  R version 3.6.0 (2019-04-26)
+#>  os       macOS High Sierra 10.13.6   
+#>  system   x86_64, darwin15.6.0        
 #>  ui       X11                         
-#>  language en_US:en                    
-#>  collate  en_US.UTF-8                 
-#>  ctype    en_US.UTF-8                 
-#>  tz       America/Toronto             
-#>  date     2020-01-29                  
+#>  language (EN)                        
+#>  collate  en_CA.UTF-8                 
+#>  ctype    en_CA.UTF-8                 
+#>  tz       America/Montreal            
+#>  date     2020-01-30                  
 #> 
@@
 #>  backports     1.1.5      2019-10-02 [1]
-#>  broom         0.5.4      2020-01-27 [1]
+#>  broom         0.5.2      2019-04-07 [1]
 #>  callr         3.4.1      2020-01-24 [1]
-#>  casebase    * 0.2.1.9001 2020-01-27 [1]
+#>  casebase    * 0.2.1.9001 2020-01-30 [1]
 #>  cellranger    1.1.0      2016-07-27 [1]
 #>  cli           2.0.1      2020-01-08 [1]
-#>  codetools     0.2-16     2018-12-24 [4]
+#>  codetools     0.2-16     2018-12-24 [1]
 #>  colorspace    1.4-1      2019-03-18 [1]
-#>  cowplot     * 1.0.0      2019-07-11 [1]
 #>  crayon        1.3.4      2017-09-16 [1]
-#>  data.table    1.12.8     2019-12-09 [1]
-#>  DBI           1.1.0      2019-12-15 [1]
+#>  data.table  * 1.12.8     2019-12-09 [1]
+#>  DBI           1.0.0      2018-05-02 [1]
 #>  dbplyr        1.4.2      2019-06-17 [1]
@@
 #>  ggplot2     * 3.2.1      2019-08-10 [1]
+#>  ggstance    * 0.3.3      2019-08-19 [1]
 #>  glmnet      * 3.0-2      2019-12-11 [1]
-#>  glue        * 1.3.1      2019-03-12 [1]
+#>  glue          1.3.1      2019-03-12 [1]
 #>  gtable        0.3.0      2019-03-25 [1]
 #>  haven         2.2.0      2019-11-08 [1]
-#>  hms           0.5.3      2020-01-08 [1]
+#>  hms           0.5.2      2019-10-30 [1]
 #>  htmltools     0.4.0      2019-10-04 [1]
@@
 #>  labeling      0.3        2014-08-23 [1]
-#>  lattice       0.20-38    2018-11-04 [4]
-#>  lava          1.6.6      2019-08-01 [1]
+#>  lattice       0.20-38    2018-11-04 [1]
 #>  lazyeval      0.2.2      2019-03-15 [1]
 #>  lifecycle     0.1.0      2019-08-01 [1]
-#>  lubridate   * 1.7.4      2018-04-11 [1]
+#>  lubridate     1.7.4      2018-04-11 [1]
 #>  magrittr    * 1.5        2014-11-22 [1]
-#>  MASS          7.3-51.5   2019-12-20 [4]
-#>  Matrix      * 1.2-18     2019-11-27 [4]
+#>  Matrix      * 1.2-17     2019-03-22 [1]
 #>  memoise       1.1.0      2017-04-21 [1]
-#>  mgcv          1.8-31     2019-11-09 [4]
+#>  mgcv          1.8-31     2019-11-09 [1]
 #>  modelr        0.1.5      2019-08-08 [1]
 #>  munsell       0.5.0      2018-06-12 [1]
-#>  nlme          3.1-143    2019-12-10 [4]
-#>  numDeriv      2016.8-1.1 2019-06-06 [1]
+#>  nlme          3.1-142    2019-11-07 [1]
 #>  pillar        1.4.3      2019-12-20 [1]
@@
 #>  pkgload       1.0.2      2018-10-29 [1]
-#>  plyr          1.8.5      2019-12-10 [1]
-#>  pracma      * 2.2.9      2019-12-15 [1]
 #>  prettyunits   1.1.1      2020-01-24 [1]
@@
 #>  reprex        0.3.0      2019-05-16 [1]
-#>  reshape2      1.4.3      2017-12-11 [1]
-#>  rlang         0.4.3      2020-01-24 [1]
+#>  rlang         0.4.4      2020-01-28 [1]
 #>  rmarkdown     2.1        2020-01-20 [1]
@@
 #>  rstudioapi    0.10       2019-03-19 [1]
-#>  rticles       0.13       2019-12-19 [1]
+#>  rticles       0.12       2019-11-12 [1]
 #>  rvest         0.3.5      2019-11-08 [1]
@@
 #>  stringr     * 1.4.0      2019-02-10 [1]
-#>  survival    * 3.1-8      2019-12-03 [4]
+#>  survival    * 3.1-7      2019-11-09 [1]
 #>  testthat      2.3.1      2019-12-01 [1]
 #>  tibble      * 2.1.3      2019-06-06 [1]
-#>  tidyr       * 1.0.2      2020-01-24 [1]
+#>  tidyr       * 1.0.0      2019-09-11 [1]
 #>  tidyselect    0.2.5      2018-10-11 [1]
 #>  tidyverse   * 1.3.0      2019-11-21 [1]
-#>  timereg     * 1.9.4      2019-07-30 [1]
 #>  usethis       1.5.1      2019-07-04 [1]
@@
 #>  source                                  
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
 #>  Github (sahirbhatnagar/casebase@24359be)
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
 #> 
-#> [1] /home/sahir/R/x86_64-pc-linux-gnu-library/3.6
-#> [2] /usr/local/lib/R/site-library
-#> [3] /usr/lib/R/site-library
-#> [4] /usr/lib/R/library
+#> [1] /Library/Frameworks/R.framework/Versions/3.6/Resources/library
 \end{CodeOutput}
@@
 
+\begin{CodeInput}
+R> # what commit is this file at? 
+R> git2r::repository(here::here())
+\end{CodeInput}
+
 \begin{CodeOutput}
-#> Local:    write-popTime /home/sahir/git_repositories/cbpaper
-#> Remote:   write-popTime @ origin (https://github.com/sahirbhatnagar/cbpaper.git)
-#> Head:     [d990874] 2020-01-28: cleaning up case study 1. merging into one figure. and echo=FALSE defaults for
+#> Local:    master /Users/jislam/Desktop/Gits/cbpaper
+#> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
+#> Head:     [b0db01f] 2020-01-16: mostly Done
 \end{CodeOutput}

### 2020-02-05 analysis/paper/paper.Rmd
add intro and edit some of the sections

--- 2020-01-28 analysis/paper/paper.Rmd
@@
 title:
+  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
   # If you use tex in the formatted title, also supply version without
+  plain:     "casebase: An Alternative Framework For Survival Analysis"
   # For running headers, if needed
-  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
-  plain:     "casebase: An Alternative Framework For Survival Analysis"
   short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
@@
 fig_caption: true
-editor_options: 
-  chunk_output_type: console
 ---
@@
   message = FALSE,
-  echo = FALSE,
+  echo = TRUE,
   comment = "#>",
@@
 
-The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence (CI)) of an event, for a particular patient.
-The following points should be noted:
-
-  1. hazard ratios
-  2. If, however, the absolute risks are of interest, they have to be recovered using the semi-parametric Breslow estimator
-  3. Alternative approaches for fitting flexible hazard models for estimating absolute risks, not requiring this two-step approach? Yes! [@hanley2009fitting]
-
-[@hanley2009fitting] propose a fully parametric hazard model that can be fit via logistic regression.
-From the fitted hazard function, cumulative incidence and, thus, risk functions of time, treatment and profile can be easily derived.
+Survival analysis has been greatly influenced over the last 50 years by the partial likelihood approach of the Cox proportional hazard model [@cox1972regression].
+This approach provides a flexible way of assessing the influence of covariates on the hazard function, without the need to specify a parametric survival model.
+This flexibility comes at the cost of decoupling the baseline hazard from the effect of the covariates.
+To recover the whole survival curve---or the cumulative incidence function (CIF)---we then need to separately estimate the baseline hazard [@breslow1972discussion].
+This in turn often leads to stepwise estimates of the survival function.
+
+From the perspective of clinicians and their patients, the most relevant quantity is often the 5- or 10-year risk of having a certain event given the patient's particular circumstances, and not the hazard ratio between a treatment and control group.
+Therefore, to make sound clinical decisions, it is important to accurately estimate the *full* hazard function, which can then be used to estimate the cumulative incidence function (CIF).
+From this perspective, the CIF and the survival function estimates are now smooth functions of time.
+
+With the goal of fitting smooth-in-time hazard functions, Hanley & Miettinen [-@hanley2009fitting] proposed a general framework for estimating fully parametric hazard models via logistic regression.
+Their approach provides users that are familiar with generalized linear models with a natural way of fitting parametric survival models.
+Moreover, their framework is very flexible: general functions of time can be estimated (e.g.
+using splines or general additive models), and hence these models retain some of the flexibility of partial likelihood approaches.
+
+In this article, we present an \proglang{R} package that combines the ideas of Hanley & Miettinen into a simple interface.
+The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile.
+Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function.
+
+In what follows, we first recall some theoretical details on case-base sampling and its use for estimating parameteric hazard functions.
+We then give a short review of existing \proglang{R} packages that implement comparable features as \pkg{casebase}.
+Next, we provide some details about the implementation of case-base sampling in our package, and we give a brief survey of its main functions.
+This is followed by four case studies that illustrate the flexibility and capabilities of \pkg{casebase}.
+We show how the same framework can be used for competing risk analyses, penalized estimation, and for studies with time-dependent exposures.
+Finally, we end the article with a discussion of the results and of future directions.
 
@@
 
-In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base sampling *uniformly*, i.e.
-to sample the base series uniformly from the study base.
+In Hanley & Miettinen [-@hanley2009fitting], the authors suggest sampling the base series *uniformly* from the study base.
 In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base.
@@
 
-The `casebase` package fits the family of hazard functions of the form
-
-$$ h(t;X) = \exp[g(t;X)] $$
-where $t$ denotes time and $X$, the individual's covariate profile.
+<!-- The `casebase` package fits the family of hazard functions of the form -->
+
+<!-- $$ h(t;X) = \exp[g(t;X)] $$ -->
+<!-- where $t$ denotes time and $X$, the individual's covariate profile.  -->
+
+Let $g(t; X)$ be the linear predictor such that $\log(h(t;X)) = g(t; X)$.
 Different functions of $t$ lead to different parametric hazard models.
-The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$.
-
-$$ h(t;X) = \exp(\beta_0 + \beta_1 X) $$
-
-The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
-This is also known as the *memoryless property* (Kalbfleisch and Prentice, 2002).
+The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$:
+
+$$ \log(h(t; X)) = \beta_0 + \beta_1 X.$$
+
+The instantaneous failure rate is independent of $t$.\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
+This is also known as the \textit{memoryless property} \citep{kalbfleisch2011statistical}.}
 
@@
 
-$$ h(t;X)  = \exp(\beta_0 + \beta_1 t + \beta_2 X) $$
-
-Use of $\log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
-
-$$ h(t;X)  = \exp(\beta_0 + \beta_1 \log(t) + \beta_2 X) $$
+$$ \log(h(t; X)) = \beta_0 + \beta_1 t + \beta_2 X.$$
+
+Use of $\log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time [@kalbfleisch2011statistical]:
+
+$$ \log(h(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. $$
 
@@
 where $N_j(t)$ is the counting process associated with the event of type $j$ and $h_j(t)$ is the corresponding hazard function.
-As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1\end{align*}}
+As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}
 
@@
                     exp(confint(cox_model))[2]),
-     yscale = 100)
+     ylim = c(0, 2), yscale = 100)
 legend("topleft", 
@@
        bg = "gray90")
-
+```
+
+Next, we show how case-base sampling can be used to carry out a similar analysis. We fit several models that differ in how we choose to model time.
+
+<!-- The \code{fitSmoothHazard} function provides an estimate of the hazard function $h(x, t)$ is the hazard function, where $t$ denotes the numerical value of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any).
+-->
+
+First, we will fit an exponential model. Recall that this corresponds to excluding time from the linear predictor.
+
+```{r erspc-casebase-exponential, eval = eval_cs1}
 casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
@@
                                                   ratio = 100)
+
+summary(casebase_exponential)
+exp(coef(casebase_exponential)[2])
+exp(confint(casebase_exponential)[2,])
+```
+
+<!-- The \code{absoluteRisk} function provides an estimate of the cumulative incidence curves for a specific risk profile using the following equation: -->
+
+<!-- $$ CI(x, t) = 1 - \exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$ -->
+
+We can then use the \code{absoluteRisk} function to get an estimate of the cumulative incidence curve for a specific covariate profile.
+In the plot below, we overlay the estimated CIF from the exponential model on the Cox model CIF:
+
+```{r erspc-casebase-cif, eval = eval_cs1}
 smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
@@
                                           newdata = new_data)
-lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
-lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
-
-casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
-                                 data = ERSPC, 
-                                 ratio = 100)
-smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
-                                          time = seq(0,15,0.1), 
-                                          newdata = new_data)
-lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
-lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
-
-casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
-                                    data = ERSPC, 
-                                    ratio = 100)
-lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
-lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
-
-legend("topleft", 
-       legend = c("Control group (Cox)","Control group (Casebase)",
-                  "Screening group (Cox)", "Screening group (Casebase)"), 
-       col = c("red","red", "blue","blue"),
-       lty = c(1, 2, 1, 2), 
-       bg = "gray90")
-
-```
-
-Next, we show how case-base sampling can be used to carry out a similar analysis. We fit several models that differ in how we choose to model time.
-
-<!-- The \code{fitSmoothHazard} function provides an estimate of the hazard function $h(x, t)$ is the hazard function, where $t$ denotes the numerical value of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any).
--->
-
-First, we will fit an exponential model. Recall that this corresponds to excluding time from the linear predictor.
-
-<!-- The \code{absoluteRisk} function provides an estimate of the cumulative incidence curves for a specific risk profile using the following equation: -->
-
-<!-- $$ CI(x, t) = 1 - \exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$ -->
-
-We can then use the \code{absoluteRisk} function to get an estimate of the cumulative incidence curve for a specific covariate profile.
-In the plot below, we overlay the estimated CIF from the exponential model on the Cox model CIF:
-
-```{r erspc-casebase-cif, eval = eval_cs1}
 
@@
                     exp(confint(cox_model))[2]),
-     yscale = 100)
-
+     ylim = c(0, 2), yscale = 100)
+lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
+lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
+
+legend("topleft", 
+       legend = c("Control group (Cox)","Control group (Casebase)",
+                  "Screening group (Cox)", "Screening group (Casebase)"), 
+       col = c("red","red", "blue","blue"),
+       lty = c(1, 2, 1, 2), 
+       bg = "gray90")
 ```
@@
 ```{r erspc-casebase-weibull, eval = eval_cs1}
+casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
+                                 data = ERSPC, 
+                                 ratio = 100)
 
@@
 ```{r erspc-casebase-weibull-cif, eval = eval_cs1}
+smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
+                                          time = seq(0,15,0.1), 
+                                          newdata = new_data)
 
@@
                     exp(confint(cox_model))[2]),
-     yscale = 100)
+     ylim = c(0, 2), yscale = 100)
+lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
+lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
 
@@
 ```{r erspc-casebase-splines, eval = eval_cs1}
+casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
+                                    data = ERSPC, 
+                                    ratio = 100)
 
@@
                     exp(confint(cox_model))[2]),
-     yscale = 100)
+     ylim = c(0, 2), yscale = 100)
+lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
+lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
 
@@
 
-In the next case study, we show how case-base sampling can also be used in the context of a competing risk analysis.
+In the next case study, we show how case-base sampling can  be used in the context of a competing risk analysis.
 For illustrative purposes, we will use the same data that was used in Scrucca *et al* [-@scrucca2010regression].
@@
 First, we can look at a population-time plot to visualize the incidence density of both relapse and the competing events.
-In Figure \ref{fig:compPop}, failure times are highlighted on the plot using red points for the event of interest (panel A) and blue dots for competing events (panel B).
+In Figure \ref{fig:compPop}, failure times are highlighted on the plot using red dots for the event of interest (panel A) and blue dots for competing events (panel B).
 In both panels, we see evidence of a non-constant hazard function: the density of points is larger at the beginning of follow-up than at the end.
@@
 
-Our main objective is to compute the absolute risk of relapse for a given set of covariates. First, we fit a smooth hazard to the data using a linear term for time:
+Our main objective is to compute the absolute risk of relapse for a given set of covariates. We start by fitting a smooth hazard to the data using a linear term for time:
 
@@
                col.names = c("Covariates", "HR", "95% CI", "HR", "95% CI"),
-               caption = "ADD CAPTION") %>% 
+               caption = "Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.") %>% 
   kable_styling() %>% 
@@
 
-We will compare our estimates to that obtained from a corresponding Fine-Gray model [@fine1999proportional].
+We will compare our estimates to that obtained from a corresponding Fine-Gray model [-@fine1999proportional].
 The Fine-Gray model is a semiparametric model for the cause-specific *subdistribution*, i.e.
@@
 The Fine-Gray model allows to directly assess the effect of a covariate on the subdistribution, as opposed to the hazard.
-For the computation, we will use the \pkg{timereg} package [@timereg].
+For the computation, we will use the \pkg{timereg} package [@timereg]:
 
@@
 
-Although the previous case studies provide a broad overview of the capabilities of \pkg{casebase}, they all have two properties in common:
-\begin{itemize}
-  \item All covariates are fixed, i.e. they do not change over time;
-  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
-\end{itemize}
-
-Case-base sampling has been used in the literature to study vaccination safety, where the exposure period was defined as the week following vaccination [@saarela2015case].
+In the previous case studies, we only considered covariates that were fixed at baseline.
+In this next case study, we will use the Stanford Heart Transplant data [@clark1971cardiac,@crowley1977covariance] to show how case-base sampling can also be used in the context of time-varying covariates.
+As an example that already appeared in the literature, case-base sampling was to study vaccination safety, where the exposure period was defined as the week following vaccination [@saarela2015case].
 Hence, the main covariate of interest, i.e.
@@
 
-The purpose of the next case study is to show how \pkg{casebase} can also be used for survival analysis problems that do not share the properties above.
-To this end, we will use the Stanford Heart Transplant Data [@clark1971cardiac,@crowley1977covariance]; we will use the version available in the \pkg{surivival} package.
-
-Recall the setting of this study: patients were admitted to the Stanford program after meeting with their physician and determining that they were unlikely to respond to other forms of treatment.
-After enrollment, the program searches for a suitable donor for the patient, which can take anywhere between a few days to almost a year.
+Recall the setting of Stanford Heart Transplant study: patients were admitted to the Stanford program after meeting with their physician and determining that they were unlikely to respond to other forms of treatment.
+After enrollment, the program searched for a suitable donor for the patient, which could take anywhere between a few days to almost a year.
 We are interested in the effect of a heart transplant on survival; therefore, the patient is considered exposed only after the transplant has occurred.
@@
 The non-proportionality seems to be more pronounced at the beginning of follow-up than the end.
-Finally, we can turn these estimate of the hazard function into estimates of the cumulative incidence functions.
+Finally, we can turn these estimates of the hazard function into estimates of the cumulative incidence functions.
 
@@
 
-In this article, we presented the \proglang{R} package \pkg{casebase} that provides functions to fit smooth parametric hazards and estimate cumulative incidence functions using case-base sampling.
+In this article, we presented the \proglang{R} package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating cumulative incidence functions using case-base sampling.
 We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through four case studies.
@@
 On the one hand, we showed how splines could be used as part of the linear predictor to model the effect of time on the hazard.
-This strategy yielded estimates of the survival function that were qualitatively similar to semiparametric estimates derived from Cox regression; however, case-base sampling led to estimates that *varying smoothly in time*.
+This strategy yielded estimates of the survival function that were qualitatively similar to semiparametric estimates derived from Cox regression; however, case-base sampling led to estimates of the survival function that *vary smoothly in time*.
 On the other hand, we also displayed the flexibility of case-base sampling by showing how it could be combined with penalized logistic regression to perform variable selection.
-Similarly, our package \pkg{casebase} provides the user with the functionality of Generalized Additive Models.
-Moreover, case-base sampling can be combined with quasi-likelihood estimation to fit survival models that potentially account for the presence of over-dispersion.
-All of the examples illustrate how the case-base sampling framework in general, and the package \pkg{casebase} in particular, allows the user to fit a broad family of survival functions.
+Even though we did not illustrate it in this article, case-base sampling can also be combined with the framework of *generalized additive models*.
+This functionality has been implemented in our package \pkg{casebase}.
+Similarly, case-base sampling can also be combined with quasi-likelihood estimation to fit survival models that can account for the presence of over-dispersion.
+All of these examples illustrate how the case-base sampling framework in general, and the package \pkg{casebase} in particular, allows the user to fit a broad and flexible family of survival functions.
 
@@
 Accordingly, our package provides functions for each step.
-Moreover, the simple interface of the \code{fittingSmoothHazard} function resembles the \code{glm} interface, which should look familiar to new users.
+Moreover, the simple interface of the \code{fittingSmoothHazard} function resembles the \code{glm} interface.
+This interface should look familiar to new users.
 Our modular approach also provides a convenient way to extend our package for new sampling or fitting strategies.
@@
 In terms of function interface, \pkg{casebase} uses a formula interface that is closer to that of \code{glm}, in that the event variable is the only variable appearing on the left-hand side of the formula.
-By contrast, both \code{survival::coxph} and \code{timereg::comp.risk} use arrays as the left-hand side that captures both the event type and time.
+By contrast, both \code{survival::coxph} and \code{timereg::comp.risk} use arrays that captures both the event type and time.
 Both approaches to modeling yield user-friendly code.
-In terms of output, both approaches differ significantly.
-Case-base sampling produces smooth hazard and smooth cumulative incidence curves, whereas Cox regression and Fine-Gray models produce step-wise cumulative incidence functions and never explicitly model the hazard function.
+However, in terms of output, both approaches differ significantly.
+Case-base sampling produces smooth hazards and smooth cumulative incidence curves, whereas Cox regression and Fine-Gray models produce step-wise cumulative incidence functions and never explicitly model the hazard function.
 Qualitatively, we showed that by using splines in the linear predictor, all three models yielded similar curves.
-However, the smooth nature of the output of \pkg{casebase} improves interpretation for consumers of these predictions.
+However, the smooth nature of the output of \pkg{casebase} provides a more intuitive interpretation for consumers of these predictions.
 In Table \ref{tab:compCBvsCox}, we provide a side-by-side comparison between the Cox model and case-base sampling.
@@
 Future work will look at some of the methodological extensions of case-base sampling.
-First, to assess the quality of the model fit, we may want to study the properties of the residuals (e.g.
+First, to assess the quality of the model fit, we want to study the properties of the residuals (e.g.
 Cox-Snell, martingale).
@@
 Saarela and Arjas [-@saarela2015non] also describe how to obtain a posterior distribution for the AUC from their model.
+Their approach could also be included in \pkg{casebase}.
 
@@
 # which R packages and versions?
-devtools::session_info()
+# devtools::session_info()
+sessionInfo()
 ```

### 2020-02-05 analysis/paper/paper.tex
add intro and edit some of the sections

--- 2020-01-28 analysis/paper/paper.tex
@@
 
-The purpose of the \pkg{casebase} package is to provide practitioners
-with an easy-to-use software tool to predict the risk (or cumulative
-incidence (CI)) of an event, for a particular patient. The following
-points should be noted:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  hazard ratios
-\item
-  If, however, the absolute risks are of interest, they have to be
-  recovered using the semi-parametric Breslow estimator
-\item
-  Alternative approaches for fitting flexible hazard models for
-  estimating absolute risks, not requiring this two-step approach? Yes!
-  \citep{hanley2009fitting}
-\end{enumerate}
-
-\citep{hanley2009fitting} propose a fully parametric hazard model that
-can be fit via logistic regression. From the fitted hazard function,
-cumulative incidence and, thus, risk functions of time, treatment and
-profile can be easily derived.
+Survival analysis has been greatly influenced over the last 50 years by
+the partial likelihood approach of the Cox proportional hazard model
+\citep{cox1972regression}. This approach provides a flexible way of
+assessing the influence of covariates on the hazard function, without
+the need to specify a parametric survival model. This flexibility comes
+at the cost of decoupling the baseline hazard from the effect of the
+covariates. To recover the whole survival curve---or the cumulative
+incidence function (CIF)---we then need to separately estimate the
+baseline hazard \citep{breslow1972discussion}. This in turn often leads
+to stepwise estimates of the survival function.
+
+From the perspective of clinicians and their patients, the most relevant
+quantity is often the 5- or 10-year risk of having a certain event given
+the patient's particular circumstances, and not the hazard ratio between
+a treatment and control group. Therefore, to make sound clinical
+decisions, it is important to accurately estimate the \emph{full} hazard
+function, which can then be used to estimate the cumulative incidence
+function (CIF). From this perspective, the CIF and the survival function
+estimates are now smooth functions of time.
+
+With the goal of fitting smooth-in-time hazard functions, Hanley \&
+Miettinen \citeyearpar{hanley2009fitting} proposed a general framework
+for estimating fully parametric hazard models via logistic regression.
+Their approach provides users that are familiar with generalized linear
+models with a natural way of fitting parametric survival models.
+Moreover, their framework is very flexible: general functions of time
+can be estimated (e.g.~using splines or general additive models), and
+hence these models retain some of the flexibility of partial likelihood
+approaches.
+
+In this article, we present an \proglang{R} package that combines the
+ideas of Hanley \& Miettinen into a simple interface. The purpose of the
+\pkg{casebase} package is to provide practitioners with an easy-to-use
+software tool to predict the risk (or cumulative incidence) of an event,
+conditional on a particular patient's covariate profile. Our package
+retains the flexibility of case-base sampling and the familiar interface
+of the \code{glm} function.
+
+In what follows, we first recall some theoretical details on case-base
+sampling and its use for estimating parameteric hazard functions. We
+then give a short review of existing \proglang{R} packages that
+implement comparable features as \pkg{casebase}. Next, we provide some
+details about the implementation of case-base sampling in our package,
+and we give a brief survey of its main functions. This is followed by
+four case studies that illustrate the flexibility and capabilities of
+\pkg{casebase}. We show how the same framework can be used for competing
+risk analyses, penalized estimation, and for studies with time-dependent
+exposures. Finally, we end the article with a discussion of the results
+and of future directions.
 
@@
 In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
-suggest performing case-base sampling \emph{uniformly}, i.e.~to sample
-the base series uniformly from the study base. In terms of Poisson
-processes, this sampling strategy corresponds essentially to a
-time-homogeneous Poisson process with intensity equal to \(b/B\), where
-\(b\) is the number of sampled observations in the base series, and
-\(B\) is the total population-time for the study base. More complex
-examples are also available; see for example Saarela \& Arjas
-\citeyearpar{saarela2015non}, where the probabilities of the sampling
-mechanism are proportional to the cardiovascular disease event rate
-given by the Framingham score.
-
-The \texttt{casebase} package fits the family of hazard functions of the
-form
-
-\[ h(t;X) = \exp[g(t;X)] \] where \(t\) denotes time and \(X\), the
-individual's covariate profile. Different functions of \(t\) lead to
+suggest sampling the base series \emph{uniformly} from the study base.
+In terms of Poisson processes, this sampling strategy corresponds
+essentially to a time-homogeneous Poisson process with intensity equal
+to \(b/B\), where \(b\) is the number of sampled observations in the
+base series, and \(B\) is the total population-time for the study base.
+More complex examples are also available; see for example Saarela \&
+Arjas \citeyearpar{saarela2015non}, where the probabilities of the
+sampling mechanism are proportional to the cardiovascular disease event
+rate given by the Framingham score.
+
+Let \(g(t; X)\) be the linear predictor such that
+\(\log(h(t;X)) = g(t; X)\). Different functions of \(t\) lead to
 different parametric hazard models. The simplest of these models is the
 one-parameter exponential distribution which is obtained by taking the
-hazard function to be constant over the range of \(t\).
-
-\[ h(t;X) = \exp(\beta_0 + \beta_1 X) \]
-
-The instantaneous failure rate is independent of \(t\), so that the
-conditional chance of failure in a time interval of specified length is
-the same regardless of how long the individual has been in the study.
-This is also known as the \emph{memoryless property} (Kalbfleisch and
-Prentice, 2002).
+hazard function to be constant over the range of \(t\):
+
+\[ \log(h(t; X)) = \beta_0 + \beta_1 X.\]
+
+The instantaneous failure rate is independent of
+\(t\).\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
+This is also known as the \textit{memoryless property} \citep{kalbfleisch2011statistical}.}
 
@@
 
-\[ h(t;X)  = \exp(\beta_0 + \beta_1 t + \beta_2 X) \]
+\[ \log(h(t; X)) = \beta_0 + \beta_1 t + \beta_2 X.\]
 
 Use of \(\log(t)\) yields the Weibull hazard which allows for a power
-dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
-
-\[ h(t;X)  = \exp(\beta_0 + \beta_1 \log(t) + \beta_2 X) \]
+dependence of the hazard on time \citep{kalbfleisch2011statistical}:
+
+\[ \log(h(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. \]
 
@@
 the likelihood function for multinomial
-regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1\end{align*}}
+regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}
 
@@
 R+                     exp(confint(cox_model))[2]),
-R+      yscale = 100)
+R+      ylim = c(0, 2), yscale = 100)
 R> legend("topleft", 
@@
 R+                     exp(confint(cox_model))[2]),
-R+      yscale = 100)
+R+      ylim = c(0, 2), yscale = 100)
 R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
@@
 R+                     exp(confint(cox_model))[2]),
-R+      yscale = 100)
+R+      ylim = c(0, 2), yscale = 100)
 R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
@@
 R+                     exp(confint(cox_model))[2]),
-R+      yscale = 100)
+R+      ylim = c(0, 2), yscale = 100)
 R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
@@
 
-In the next case study, we show how case-base sampling can also be used
-in the context of a competing risk analysis. For illustrative purposes,
-we will use the same data that was used in Scrucca \emph{et al}
+In the next case study, we show how case-base sampling can be used in
+the context of a competing risk analysis. For illustrative purposes, we
+will use the same data that was used in Scrucca \emph{et al}
 \citeyearpar{scrucca2010regression}. The data was downloaded from the
@@
 \ref{fig:compPop}, failure times are highlighted on the plot using red
-points for the event of interest (panel A) and blue dots for competing
+dots for the event of interest (panel A) and blue dots for competing
 events (panel B). In both panels, we see evidence of a non-constant
@@
 Our main objective is to compute the absolute risk of relapse for a
-given set of covariates. First, we fit a smooth hazard to the data using
-a linear term for time:
+given set of covariates. We start by fitting a smooth hazard to the data
+using a linear term for time:
 
@@
 
-\caption{\label{tab:bmtcrr-cis}ADD CAPTION}
+\caption{\label{tab:bmtcrr-cis}Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.}
 \centering
@@
 We will compare our estimates to that obtained from a corresponding
-Fine-Gray model \citep{fine1999proportional}. The Fine-Gray model is a
-semiparametric model for the cause-specific \emph{subdistribution},
+Fine-Gray model \citeyearpar{fine1999proportional}. The Fine-Gray model
+is a semiparametric model for the cause-specific \emph{subdistribution},
 i.e.~the function \(f_k(t)\) such that
@@
 subdistribution, as opposed to the hazard. For the computation, we will
-use the \pkg{timereg} package \citep{timereg}.
+use the \pkg{timereg} package \citep{timereg}:
 
@@
 
-Although the previous case studies provide a broad overview of the
-capabilities of \pkg{casebase}, they all have two properties in common:
-
-\begin{itemize}
-  \item All covariates are fixed, i.e. they do not change over time;
-  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
-\end{itemize}
-
-Case-base sampling has been used in the literature to study vaccination
-safety, where the exposure period was defined as the week following
-vaccination \citep{saarela2015case}. Hence, the main covariate of
-interest, i.e.~exposure to the vaccine, was changing over time. In this
-context, case-base sampling offers an efficient alternative to nested
+In the previous case studies, we only considered covariates that were
+fixed at baseline. In this next case study, we will use the Stanford
+Heart Transplant data
+\citep[\citet{crowley1977covariance}]{clark1971cardiac} to show how
+case-base sampling can also be used in the context of time-varying
+covariates. As an example that already appeared in the literature,
+case-base sampling was to study vaccination safety, where the exposure
+period was defined as the week following vaccination
+\citep{saarela2015case}. Hence, the main covariate of interest,
+i.e.~exposure to the vaccine, was changing over time. In this context,
+case-base sampling offers an efficient alternative to nested
 case-control designs or self-matching.
 
-The purpose of the next case study is to show how \pkg{casebase} can
-also be used for survival analysis problems that do not share the
-properties above. To this end, we will use the Stanford Heart Transplant
-Data \citep[\citet{crowley1977covariance}]{clark1971cardiac}; we will
-use the version available in the \pkg{surivival} package.
-
-Recall the setting of this study: patients were admitted to the Stanford
-program after meeting with their physician and determining that they
-were unlikely to respond to other forms of treatment. After enrollment,
-the program searches for a suitable donor for the patient, which can
-take anywhere between a few days to almost a year. We are interested in
-the effect of a heart transplant on survival; therefore, the patient is
-considered exposed only after the transplant has occurred.
+Recall the setting of Stanford Heart Transplant study: patients were
+admitted to the Stanford program after meeting with their physician and
+determining that they were unlikely to respond to other forms of
+treatment. After enrollment, the program searched for a suitable donor
+for the patient, which could take anywhere between a few days to almost
+a year. We are interested in the effect of a heart transplant on
+survival; therefore, the patient is considered exposed only after the
+transplant has occurred.
 
@@
 #> Model1 Model2 Model3 Model4 
-#>    803    760    733    716
+#>    795    760    729    724
 \end{CodeOutput}
@@
 The non-proportionality seems to be more pronounced at the beginning of
-follow-up than the end. Finally, we can turn these estimate of the
+follow-up than the end. Finally, we can turn these estimates of the
 hazard function into estimates of the cumulative incidence functions.
@@
 #>         
-#> 365 0.29
+#> 365 0.27
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> [1] 0.21
+#> [1] 0.19
 \end{CodeOutput}
@@
 
-In this article, we presented the \proglang{R} package \pkg{casebase}
-that provides functions to fit smooth parametric hazards and estimate
-cumulative incidence functions using case-base sampling. We outlined the
-theoretical underpinnings of the approach, we provided details about our
-implementation, and we illustrated the merits of the approach and the
-package through four case studies.
+In this article, we presented the \proglang{R} package \pkg{casebase},
+which provides functions for fitting smooth parametric hazards and
+estimating cumulative incidence functions using case-base sampling. We
+outlined the theoretical underpinnings of the approach, we provided
+details about our implementation, and we illustrated the merits of the
+approach and the package through four case studies.
 
@@
 similar to semiparametric estimates derived from Cox regression;
-however, case-base sampling led to estimates that \emph{varying smoothly
-in time}. On the other hand, we also displayed the flexibility of
-case-base sampling by showing how it could be combined with penalized
-logistic regression to perform variable selection. Similarly, our
-package \pkg{casebase} provides the user with the functionality of
-Generalized Additive Models. Moreover, case-base sampling can be
-combined with quasi-likelihood estimation to fit survival models that
-potentially account for the presence of over-dispersion. All of the
-examples illustrate how the case-base sampling framework in general, and
-the package \pkg{casebase} in particular, allows the user to fit a broad
-family of survival functions.
+however, case-base sampling led to estimates of the survival function
+that \emph{vary smoothly in time}. On the other hand, we also displayed
+the flexibility of case-base sampling by showing how it could be
+combined with penalized logistic regression to perform variable
+selection. Even though we did not illustrate it in this article,
+case-base sampling can also be combined with the framework of
+\emph{generalized additive models}. This functionality has been
+implemented in our package \pkg{casebase}. Similarly, case-base sampling
+can also be combined with quasi-likelihood estimation to fit survival
+models that can account for the presence of over-dispersion. All of
+these examples illustrate how the case-base sampling framework in
+general, and the package \pkg{casebase} in particular, allows the user
+to fit a broad and flexible family of survival functions.
 
@@
 of the \code{fittingSmoothHazard} function resembles the \code{glm}
-interface, which should look familiar to new users. Our modular approach
-also provides a convenient way to extend our package for new sampling or
-fitting strategies.
+interface. This interface should look familiar to new users. Our modular
+approach also provides a convenient way to extend our package for new
+sampling or fitting strategies.
 
@@
 variable appearing on the left-hand side of the formula. By contrast,
-both \code{survival::coxph} and \code{timereg::comp.risk} use arrays as
-the left-hand side that captures both the event type and time. Both
-approaches to modeling yield user-friendly code. In terms of output,
-both approaches differ significantly. Case-base sampling produces smooth
-hazard and smooth cumulative incidence curves, whereas Cox regression
-and Fine-Gray models produce step-wise cumulative incidence functions
-and never explicitly model the hazard function. Qualitatively, we showed
-that by using splines in the linear predictor, all three models yielded
-similar curves. However, the smooth nature of the output of
-\pkg{casebase} improves interpretation for consumers of these
+both \code{survival::coxph} and \code{timereg::comp.risk} use arrays
+that captures both the event type and time. Both approaches to modeling
+yield user-friendly code. However, in terms of output, both approaches
+differ significantly. Case-base sampling produces smooth hazards and
+smooth cumulative incidence curves, whereas Cox regression and Fine-Gray
+models produce step-wise cumulative incidence functions and never
+explicitly model the hazard function. Qualitatively, we showed that by
+using splines in the linear predictor, all three models yielded similar
+curves. However, the smooth nature of the output of \pkg{casebase}
+provides a more intuitive interpretation for consumers of these
 predictions. In Table \ref{tab:compCBvsCox}, we provide a side-by-side
@@
 case-base sampling. First, to assess the quality of the model fit, we
-may want to study the properties of the residuals (e.g.~Cox-Snell,
+want to study the properties of the residuals (e.g.~Cox-Snell,
 martingale). More work needs to be done to understand these residuals in
@@
 \citeyearpar{saarela2015non} also describe how to obtain a posterior
-distribution for the AUC from their model.
+distribution for the AUC from their model. Their approach could also be
+included in \pkg{casebase}.
 
@@
 
-This report was generated on 2020-01-28 14:29:58 using the following
+This report was generated on 2020-02-05 17:43:41 using the following
 computational environment and dependencies:
@@
 R> # which R packages and versions?
-R> devtools::session_info()
+R> # devtools::session_info()
+R> sessionInfo()
 \end{CodeInput}
@@
 \begin{CodeOutput}
-#> - Session info ---------------------------------------------------------------
-#>  setting  value                       
-#>  version  R version 3.6.2 (2019-12-12)
-#>  os       Pop!_OS 19.10               
-#>  system   x86_64, linux-gnu           
-#>  ui       X11                         
-#>  language en_US:en                    
-#>  collate  en_US.UTF-8                 
-#>  ctype    en_US.UTF-8                 
-#>  tz       America/Toronto             
-#>  date     2020-01-28                  
-#> 
-#> - Packages -------------------------------------------------------------------
-#>  package     * version    date       lib
-#>  assertthat    0.2.1      2019-03-21 [1]
-#>  backports     1.1.5      2019-10-02 [1]
-#>  broom         0.5.4      2020-01-27 [1]
-#>  callr         3.4.1      2020-01-24 [1]
-#>  casebase    * 0.2.1.9001 2020-01-27 [1]
-#>  cellranger    1.1.0      2016-07-27 [1]
-#>  cli           2.0.1      2020-01-08 [1]
-#>  codetools     0.2-16     2018-12-24 [4]
-#>  colorspace    1.4-1      2019-03-18 [1]
-#>  cowplot     * 1.0.0      2019-07-11 [1]
-#>  crayon        1.3.4      2017-09-16 [1]
-#>  data.table    1.12.8     2019-12-09 [1]
-#>  DBI           1.1.0      2019-12-15 [1]
-#>  dbplyr        1.4.2      2019-06-17 [1]
-#>  desc          1.2.0      2018-05-01 [1]
-#>  devtools      2.2.1      2019-09-24 [1]
-#>  digest        0.6.23     2019-11-23 [1]
-#>  dplyr       * 0.8.3      2019-07-04 [1]
-#>  ellipsis      0.3.0      2019-09-20 [1]
-#>  evaluate      0.14       2019-05-28 [1]
-#>  fansi         0.4.1      2020-01-08 [1]
-#>  farver        2.0.3      2020-01-16 [1]
-#>  forcats     * 0.4.0      2019-02-17 [1]
-#>  foreach       1.4.7      2019-07-27 [1]
-#>  fs            1.3.1      2019-05-06 [1]
-#>  generics      0.0.2      2018-11-29 [1]
-#>  ggplot2     * 3.2.1      2019-08-10 [1]
-#>  glmnet      * 3.0-2      2019-12-11 [1]
-#>  glue        * 1.3.1      2019-03-12 [1]
-#>  gtable        0.3.0      2019-03-25 [1]
-#>  haven         2.2.0      2019-11-08 [1]
-#>  hms           0.5.3      2020-01-08 [1]
-#>  htmltools     0.4.0      2019-10-04 [1]
-#>  httr          1.4.1      2019-08-05 [1]
-#>  iterators     1.0.12     2019-07-26 [1]
-#>  jsonlite      1.6        2018-12-07 [1]
-#>  kableExtra  * 1.1.0      2019-03-16 [1]
-#>  knitr         1.27       2020-01-16 [1]
-#>  labeling      0.3        2014-08-23 [1]
-#>  lattice       0.20-38    2018-11-04 [4]
-#>  lava          1.6.6      2019-08-01 [1]
-#>  lazyeval      0.2.2      2019-03-15 [1]
-#>  lifecycle     0.1.0      2019-08-01 [1]
-#>  lubridate   * 1.7.4      2018-04-11 [1]
-#>  magrittr    * 1.5        2014-11-22 [1]
-#>  MASS          7.3-51.5   2019-12-20 [4]
-#>  Matrix      * 1.2-18     2019-11-27 [4]
-#>  memoise       1.1.0      2017-04-21 [1]
-#>  mgcv          1.8-31     2019-11-09 [4]
-#>  modelr        0.1.5      2019-08-08 [1]
-#>  munsell       0.5.0      2018-06-12 [1]
-#>  nlme          3.1-143    2019-12-10 [4]
-#>  numDeriv      2016.8-1.1 2019-06-06 [1]
-#>  pillar        1.4.3      2019-12-20 [1]
-#>  pkgbuild      1.0.6      2019-10-09 [1]
-#>  pkgconfig     2.0.3      2019-09-22 [1]
-#>  pkgload       1.0.2      2018-10-29 [1]
-#>  plyr          1.8.5      2019-12-10 [1]
-#>  pracma      * 2.2.9      2019-12-15 [1]
-#>  prettyunits   1.1.1      2020-01-24 [1]
-#>  processx      3.4.1      2019-07-18 [1]
-#>  ps            1.3.0      2018-12-21 [1]
-#>  purrr       * 0.3.3      2019-10-18 [1]
-#>  R6            2.4.1      2019-11-12 [1]
-#>  Rcpp          1.0.3      2019-11-08 [1]
-#>  readr       * 1.3.1      2018-12-21 [1]
-#>  readxl        1.3.1      2019-03-13 [1]
-#>  remotes       2.1.0      2019-06-24 [1]
-#>  reprex        0.3.0      2019-05-16 [1]
-#>  reshape2      1.4.3      2017-12-11 [1]
-#>  rlang         0.4.3      2020-01-24 [1]
-#>  rmarkdown     2.1        2020-01-20 [1]
-#>  rprojroot     1.3-2      2018-01-03 [1]
-#>  rstudioapi    0.10       2019-03-19 [1]
-#>  rticles       0.13       2019-12-19 [1]
-#>  rvest         0.3.5      2019-11-08 [1]
-#>  scales        1.1.0      2019-11-18 [1]
-#>  sessioninfo   1.1.1      2018-11-05 [1]
-#>  shape         1.4.4      2018-02-07 [1]
-#>  stringi       1.4.5      2020-01-11 [1]
-#>  stringr     * 1.4.0      2019-02-10 [1]
-#>  survival    * 3.1-8      2019-12-03 [4]
-#>  testthat      2.3.1      2019-12-01 [1]
-#>  tibble      * 2.1.3      2019-06-06 [1]
-#>  tidyr       * 1.0.2      2020-01-24 [1]
-#>  tidyselect    0.2.5      2018-10-11 [1]
-#>  tidyverse   * 1.3.0      2019-11-21 [1]
-#>  timereg     * 1.9.4      2019-07-30 [1]
-#>  usethis       1.5.1      2019-07-04 [1]
-#>  vctrs         0.2.2      2020-01-24 [1]
-#>  VGAM          1.1-2      2019-11-21 [1]
-#>  viridisLite   0.3.0      2018-02-01 [1]
-#>  webshot       0.5.2      2019-11-22 [1]
-#>  withr         2.1.2      2018-03-15 [1]
-#>  xfun          0.12       2020-01-13 [1]
-#>  xml2          1.2.2      2019-08-09 [1]
-#>  yaml          2.2.0      2018-07-25 [1]
-#>  source                                  
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  Github (sahirbhatnagar/casebase@24359be)
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#> 
-#> [1] /home/sahir/R/x86_64-pc-linux-gnu-library/3.6
-#> [2] /usr/local/lib/R/site-library
-#> [3] /usr/lib/R/site-library
-#> [4] /usr/lib/R/library
+#> R version 3.6.2 (2019-12-12)
+#> Platform: x86_64-pc-linux-gnu (64-bit)
+#> Running under: Ubuntu 18.04.4 LTS
+#> 
+#> Matrix products: default
+#> BLAS:   /usr/lib/x86_64-linux-gnu/openblas/libblas.so.3
+#> LAPACK: /usr/lib/x86_64-linux-gnu/libopenblasp-r0.2.20.so
+#> 
+#> locale:
+#>  [1] LC_CTYPE=en_CA.UTF-8       LC_NUMERIC=C              
+#>  [3] LC_TIME=en_CA.UTF-8        LC_COLLATE=en_CA.UTF-8    
+#>  [5] LC_MONETARY=en_CA.UTF-8    LC_MESSAGES=en_CA.UTF-8   
+#>  [7] LC_PAPER=en_CA.UTF-8       LC_NAME=C                 
+#>  [9] LC_ADDRESS=C               LC_TELEPHONE=C            
+#> [11] LC_MEASUREMENT=en_CA.UTF-8 LC_IDENTIFICATION=C       
+#> 
+#> attached base packages:
+#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
+#> [8] base     
+#> 
+#> other attached packages:
+#>  [1] lubridate_1.7.4     timereg_1.9.4       glue_1.3.1         
+#>  [4] cowplot_1.0.0       pracma_2.2.9        kableExtra_1.1.0   
+#>  [7] glmnet_3.0-1        Matrix_1.2-18       casebase_0.2.1.9001
+#> [10] survival_3.1-8      magrittr_1.5        forcats_0.4.0      
+#> [13] stringr_1.4.0       dplyr_0.8.3         purrr_0.3.3        
+#> [16] readr_1.3.1         tidyr_1.0.0         tibble_2.1.3       
+#> [19] ggplot2_3.2.1       tidyverse_1.3.0    
+#> 
+#> loaded via a namespace (and not attached):
+#>  [1] httr_1.4.1          VGAM_1.1-2          jsonlite_1.6       
+#>  [4] viridisLite_0.3.0   foreach_1.4.7       modelr_0.1.5       
+#>  [7] assertthat_0.2.1    stats4_3.6.2        cellranger_1.1.0   
+#> [10] yaml_2.2.0          numDeriv_2016.8-1.1 pillar_1.4.3       
+#> [13] backports_1.1.5     lattice_0.20-38     digest_0.6.23      
+#> [16] rvest_0.3.5         colorspace_1.4-1    plyr_1.8.5         
+#> [19] htmltools_0.4.0     pkgconfig_2.0.3     broom_0.5.2        
+#> [22] haven_2.2.0         scales_1.1.0        webshot_0.5.2      
+#> [25] lava_1.6.6          mgcv_1.8-31         rticles_0.12       
+#> [28] generics_0.0.2      farver_2.0.1        ellipsis_0.3.0     
+#> [31] withr_2.1.2         lazyeval_0.2.2      cli_2.0.1          
+#> [34] crayon_1.3.4        readxl_1.3.1        evaluate_0.14      
+#> [37] fs_1.3.1            fansi_0.4.1         MASS_7.3-51.5      
+#> [40] nlme_3.1-143        xml2_1.2.2          tools_3.6.2        
+#> [43] data.table_1.12.8   hms_0.5.2           lifecycle_0.1.0    
+#> [46] munsell_0.5.0       reprex_0.3.0        compiler_3.6.2     
+#> [49] rlang_0.4.2         grid_3.6.2          iterators_1.0.12   
+#> [52] rstudioapi_0.10     labeling_0.3        rmarkdown_1.18     
+#> [55] gtable_0.3.0        codetools_0.2-16    DBI_1.0.0          
+#> [58] reshape2_1.4.3      R6_2.4.1            knitr_1.26         
+#> [61] zeallot_0.1.0       shape_1.4.4         stringi_1.4.4      
+#> [64] Rcpp_1.0.3          vctrs_0.2.1         dbplyr_1.4.2       
+#> [67] tidyselect_0.2.5    xfun_0.11
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> Local:    write-popTime /home/sahir/git_repositories/cbpaper
+#> Local:    intro /home/turgeonm/Documents/git_repos/cbpaper
 #> Head:     [c926a9e] 2020-01-21: Merge pull request #15 from sahirbhatnagar/fix-casebase-dependency

### 2020-02-06 README.Rmd
comparison of casebase to survreg for case study 1

@@
 4. Commit and push your changes to GitHub
+
+# Docker Image
+
+```{bash, eval = FALSE, echo = TRUE}
+# https://github.com/rocker-org/rocker/wiki/Using-the-RStudio-image
+# https://www.digitalocean.com/community/tutorials/how-to-install-and-use-docker-on-ubuntu-16-04
+# https://docs.docker.com/engine/reference/run/ which explains the run tags
+docker pull sahirbhatnagar/cbpaper:latest # pulls the image locally
+docker images # see list of images
+docker ps -a # also see list of images
+docker run -d -p 8787:8787 -e PASSWORD=<YOUR_PASS> --name cbpaper sahirbhatnagar/cbpaper
+# then go to http://localhost:8787
+# username is rstudio, password is what you specified
+# in R do: setwd('/cbpaper/') and then you should see the folder with all the materials in the folder RStudio pane
+docker stop cbpaper # this can be what you supplied to --name in the above command or the container ID
+```
+
+# File structure of repo
+
+```
+analysis/
+|
+├── paper/
+│   ├── paper.Rmd       # this is the main document to edit. just testing code for now
+│   └── references.bib  # this contains the reference list information
+├── figures/            # location of the figures produced by the Rmd
+|
+├── data/
+    ├── raw_data/       # data obtained from elsewhere
+    └── derived_data/   # data generated during the analysis
+
+```
 

### 2020-02-06 analysis/paper/paper.Rmd
comparison of casebase to survreg for case study 1

--- 2020-01-30 analysis/paper/paper.Rmd
@@
   \usepackage{makecell}
+  \usepackage{tabu}
   \DeclareUnicodeCharacter{2500}{-}
@@
 library(pracma)
+library(flexsurv) # for fitting gompertz AFT
 options(digits = 2, scipen = 999)
@@
 
-# casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
-#                                                   data = ERSPC, 
-#                                                   ratio = 100)
-# smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
-#                                           time = seq(0,15,0.1), 
-#                                           newdata = new_data)
-# 
-# casebase_weibull <- casebase::fitSmoothHazard(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm, 
-#                                            data = ERSPC, 
-#                                            ratio = 100)
-# smooth_risk_weibull <- casebase::absoluteRisk(object = casebase_weibull, 
-#                                            time = seq(0,15,0.1), 
-#                                            newdata = new_data)
-# casebase_splines <- casebase::fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
-#                                               data = ERSPC, 
-#                                               ratio = 100)
-# smooth_risk_splines <- casebase::absoluteRisk(object = casebase_splines, 
-#                                               time = seq(0,15,0.1), 
-#                                               newdata = new_data)
-# 
-# casebase_interaction <- casebase::fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time*ScrArm, 
-#                                                   data = ERSPC, 
-#                                                   ratio = 100)
-# smooth_risk_interaction <- casebase::absoluteRisk(object = casebase_interaction, 
-#                                                   time = seq(0,15,0.1), 
-#                                                   newdata = new_data)
-
-# par(mfrow = c(2, 2))
-# casebase::hazardPlot(casebase_interaction, newdata = new_data[2,], ci = FALSE)
-# casebase::hazardPlot(casebase_interaction, newdata = new_data[1,], ci = FALSE, add = TRUE, lty = 2)
-# 
-# casebase::hazardPlot(casebase_splines, newdata = new_data[2,], ci = FALSE)
-# casebase::hazardPlot(casebase_splines, newdata = new_data[1,], ci = FALSE, add = TRUE, lty = 2)
-# 
-# casebase::hazardPlot(casebase_time, newdata = new_data[2,], ci = FALSE)
-# casebase::hazardPlot(casebase_time, newdata = new_data[1,], add = TRUE, lty = 2)
-# 
-# casebase::hazardPlot(casebase_exponential, newdata = new_data[2,], ci = FALSE)
-# casebase::hazardPlot(casebase_exponential, newdata = new_data[1,], add = TRUE, lty = 2)
+aft_fits <- lapply(list("exp","gompertz","weibull"), function(i) 
+  flexsurv::flexsurvreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
+                           data = ERSPC, dist = i)
+  )
+
+names(aft_fits) <- list("exponential","gompertz","weibull")
 
@@
 
+```{r, results='hide'}
+par(mfrow = c(2, 2))
+lapply(fits, function(i) {
+  casebase::hazardPlot(i, newdata = new_data[2,], ci = FALSE)
+  casebase::hazardPlot(i, newdata = new_data[1,], ci = FALSE, add = TRUE, lty = 2)  
+})
+```
+
 ## Equivalence between casebase and AFT
 
-```{r}
+```{r, eval = FALSE}
 # exponential -------------------------------------------------------------
@@
 -coef(aft_weibull)/aft_weibull$scale
+
 coef(casebase_weibull)
@@
 ```{r erspc-95ci, echo = FALSE, eval = eval_cs1}
-CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
-CI_exp <- c(exp(coef(fits$exponential))["ScrArmScreening group"], exp(confint(fits$exponential))["ScrArmScreening group",])
-CI_gompertz <- c(exp(coef(fits$gompertz))["ScrArmScreening group"], exp(confint(fits$gompertz))["ScrArmScreening group",])
-CI_weibull <- c(exp(coef(fits$weibull))["ScrArmScreening group"], exp(confint(fits$weibull))["ScrArmScreening group",])
-CI_splines <- c(exp(coef(fits$splines))["ScrArmScreening group"], exp(confint(fits$splines))["ScrArmScreening group",])
-
-table_ci <- rbind(CI_cox, CI_exp, CI_gompertz, CI_weibull, CI_splines)
-colnames(table_ci) <- c("HR", "LB", "UB")
-
-data.frame(Model = c("Cox", "Exponential", "Gompertz", "Weibull", "Splines")) %>% 
-  bind_cols(as.data.frame(table_ci)) %>% 
-  mutate(CI = paste0("(", round(LB, 2), ", ",
-                     round(UB, 2), ")"),
-         Width = UB - LB) %>% 
-  dplyr::select(-LB, -UB) %>% 
-  knitr::kable(format = "latex", booktabs = TRUE)  %>%
-  kable_styling()
+# CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
+# CI_exp <- c(exp(coef(fits$exponential))["ScrArmScreening group"], exp(confint(fits$exponential))["ScrArmScreening group",])
+# CI_gompertz <- c(exp(coef(fits$gompertz))["ScrArmScreening group"], exp(confint(fits$gompertz))["ScrArmScreening group",])
+# CI_weibull <- c(exp(coef(fits$weibull))["ScrArmScreening group"], exp(confint(fits$weibull))["ScrArmScreening group",])
+# CI_splines <- c(exp(coef(fits$splines))["ScrArmScreening group"], exp(confint(fits$splines))["ScrArmScreening group",])
+
+# function to exponentiate HR and CI and return a HR (lower, Upper) character
+exp_est_ci <- function(model = c("exponential","gompertz","weibull","splines"),
+                       fit, 
+                       covariate = "ScrArmScreening group", 
+                       type = c("casebase", "survreg")) {
+
+  type <- match.arg(type)
+  covariate <- match.arg(covariate)
+
+  if (type == "casebase"){
+
+    cis <- confint(fit[[model]])
+
+    sprintf("%0.2f (%0.2f, %0.2f)",
+            exp(coef(fit[[model]]))[[covariate]],
+            exp(cis)[covariate,1],
+            exp(cis)[covariate,2])
+  } else {
+
+    if (model == "weibull") {
+
+      # The maximised log-likelihoods are the same between survreg and flexsurvreg, 
+      # however the parameterisation is different: the first coefficient (Intercept) 
+      # reported by survreg is log(mu), and survreg's "scale" is
+      # dweibull’s (thus flexsurvreg)'s 1 / shape
+      res <- -fit[[model]]$res[covariate,] / (1 / fit[[model]]$res["shape",])
+
+      sprintf("%0.2f (%0.2f, %0.2f)",
+              exp(res[["est"]]),
+              exp(res[["U95%"]]),
+              exp(res[["L95%"]]))
+
+    } else if (model == "splines") {
+
+      c("--")
+
+    } else {
+
+      res <- fit[[model]]$res[covariate,]
+
+      sprintf("%0.2f (%0.2f, %0.2f)",
+              exp(res[["est"]]),
+              exp(res[["L95%"]]),
+              exp(res[["U95%"]]))
+    }
+  }
+}
+
+casebase_HR_CI <- lapply(list("exponential","gompertz","weibull","splines"), 
+                         exp_est_ci, 
+                         fit = fits,
+                         covariate = "ScrArmScreening group", 
+                         type = "casebase")
+
+survreg_HR_CI <- lapply(list("exponential","gompertz","weibull","splines"), 
+                         exp_est_ci, 
+                         fit = aft_fits,
+                         covariate = "ScrArmScreening group", 
+                         type = "survreg")
+
+pt <- cbind(list("Exponential","Gompertz","Weibull","Splines"),
+            do.call(rbind, casebase_HR_CI), do.call(rbind, survreg_HR_CI))
+
+# table_ci <- rbind(CI_cox, CI_exp, CI_gompertz, CI_weibull, CI_splines)
+# colnames(table_ci) <- c("HR", "LB", "UB")
+# 
+# data.frame(Model = c("Cox", "Exponential", "Gompertz", "Weibull", "Splines")) %>% 
+#   bind_cols(as.data.frame(table_ci)) %>% 
+#   mutate(CI = paste0("(", round(LB, 2), ", ",
+#                      round(UB, 2), ")"),
+#          Width = UB - LB) %>% 
+#   dplyr::select(-LB, -UB) %>% 
+#   knitr::kable(format = "latex", booktabs = TRUE)  %>%
+#   kable_styling()
+```
+
+```{r, eval = TRUE}
+## ---- print-sim-table ----
+
+kable(pt, "latex", booktabs = T, align = c("l","c","c"),
+      caption = c("Mean "),
+      col.names = c("Model","casebase","survreg")) %>%
+  kable_styling() %>%
+  # add_header_above(c(" "," ","No overlap" = 2, "All causal SNPs\nin kinship" = 2, "No overlap" = 2, "All causal SNPs\nin kinship" = 2)) %>%
+  # add_header_above(c(" "," ","Null model" = 4, "1% Causal SNPs" = 4)) %>%
+  column_spec(1, bold=T) %>%
+  collapse_rows(columns = 1, latex_hline = "major", valign = "middle") %>%
+  footnote(general = "Median (Inter-quartile range) is given for Model Size.",
+           general_title = "Note:") #%>%
+  # footnote(general = "Model Size is the number of selected variables using the high-dimensional BIC for \texttt{ggmix} and 10-fold cross validation for \texttt{lasso} and \texttt{twostep}.")
+  #          # number = c("", "Footnote 2; "),
+  #          # alphabet = c("Footnote A; ", "Footnote B; "),
+  #          # symbol = c("Footnote Symbol 1; ", "Footnote Symbol 2"),
+  #          general_title = "Note: ", number_title = "Type I: ",
+  #          alphabet_title = "Type II: ", symbol_title = "Type III: ",
+  #          footnote_as_chunk = T, title_format = c("italic", "underline")
+  # )
 ```
@@
 Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) splines model.
+
+```{r}
+knitr::knit_exit()
+```
 

### 2020-02-06 analysis/paper/paper.tex
comparison of casebase to survreg for case study 1

--- 2020-01-30 analysis/paper/paper.tex
@@
 
-\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx} \usepackage{tabularx} \usepackage{float} \usepackage{booktabs} \usepackage{makecell} \DeclareUnicodeCharacter{2500}{-}
+\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx} \usepackage{tabularx} \usepackage{float} \usepackage{booktabs} \usepackage{makecell} \usepackage{tabu} \DeclareUnicodeCharacter{2500}{-}
 
@@
 
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{../figures/unnamed-chunk-1-1} \end{center}
+
+\end{CodeChunk}
+
 \hypertarget{equivalence-between-casebase-and-aft}{%
@@
 AFT}\label{equivalence-between-casebase-and-aft}}
-
-\begin{CodeChunk}
-
-\begin{CodeOutput}
-#>           (Intercept) ScrArmScreening group 
-#>                 -7.77                 -0.22
-\end{CodeOutput}
-
-\begin{CodeOutput}
-#>           (Intercept) ScrArmScreening group 
-#>                 -7.76                 -0.23
-\end{CodeOutput}
-
-\begin{CodeOutput}
-#>           (Intercept) ScrArmScreening group 
-#>                -10.17                 -0.22
-\end{CodeOutput}
-
-\begin{CodeOutput}
-#>           (Intercept)   log(Follow.Up.Time) ScrArmScreening group 
-#>                 -9.44                  1.07                 -0.23
-\end{CodeOutput}
-\end{CodeChunk}
 
@@
 \begin{CodeChunk}
-\begin{table}[H]
+\begin{table}
+
+\caption{\label{tab:unnamed-chunk-4}Mean }
 \centering
-\begin{tabular}{lrlr}
+\begin{tabular}[t]{>{\bfseries}lcc}
 \toprule
-Model & HR & CI & Width\\
+Model & casebase & survreg\\
 \midrule
-Cox & 0.80 & (0.67, 0.95) & 0.28\\
-Exponential & 0.81 & (0.68, 0.96) & 0.28\\
-Gompertz & 0.78 & (0.66, 0.93) & 0.27\\
-Weibull & 0.80 & (0.67, 0.95) & 0.28\\
-Splines & 0.81 & (0.68, 0.96) & 0.28\\
+Exponential & 0.81 (0.68, 0.96) & 0.81 (0.68, 0.96)\\
+\cmidrule{1-3}
+Gompertz & 0.78 (0.66, 0.93) & 0.80 (0.67, 0.95)\\
+\cmidrule{1-3}
+Weibull & 0.80 (0.67, 0.95) & 0.80 (0.65, 0.96)\\
+\cmidrule{1-3}
+Splines & 0.81 (0.68, 0.96) & --\\
 \bottomrule
+\multicolumn{3}{l}{\textit{Note:}}\\
+\multicolumn{3}{l}{Median (Inter-quartile range) is given for Model Size.}\\
 \end{tabular}
@@
 
-\begin{CodeChunk}
-
-\begin{CodeOutput}
-#> Analysis of Deviance Table
-#> 
-#> Model 1: DeadOfPrCa ~ ScrArm + offset(offset)
-#> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
-#>   Resid. Df Resid. Dev Df Deviance            Pr(>Chi)    
-#> 1     54538       6053                                    
-#> 2     54535       5786  3      267 <0.0000000000000002 ***
-#> ---
-#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{CodeOutput}
-\end{CodeChunk}
-
-As expected, we see that splines model provides a better fit. Similarly,
-we can compare the AIC for all three case-base models:
-
-\begin{CodeChunk}
-
-\begin{CodeOutput}
-#>    Exp.   Gomp.   Weib. Splines 
-#>    6057    5816    6057    5796
-\end{CodeOutput}
-\end{CodeChunk}
-
-Once again, we have evidence that the splines model provides the best
-fit.
-
-With this first case study, we saw how population-time plots can provide
-a useful summary of the incidence for our event of interests. In
-particular, we saw how it can provide evidence against an exponential
-hazard model. We then explored how different parametric models can
-provide different estimates of the CIF, even though they all give
-similar estimates for the hazard ratios.
-
-\hypertarget{case-study-2bone-marrow-transplant}{%
-\section{Case study 2--Bone-marrow
-transplant}\label{case-study-2bone-marrow-transplant}}
-
-In the next case study, we show how case-base sampling can also be used
-in the context of a competing risk analysis. For illustrative purposes,
-we will use the same data that was used in Scrucca \emph{et al}
-\citeyearpar{scrucca2010regression}. The data was downloaded from the
-first author's website, and it is also available as part of the
-\pkg{casebase} package.
-
-The data contains information on 177 patients who received a stem-cell
-transplant for acute leukemia. The event of interest is relapse, but
-other competing causes (e.g.~transplant-related death) were also
-recorded. Several covariates were captured at baseline: sex, disease
-type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL
-and AML, respectively), disease phase at transplant (Relapse, CR1, CR2,
-CR3), source of stem cells (bone marrow and peripheral blood, coded as
-BM+PB, or only peripheral blood, coded as PB), and age. A summary of
-these baseline characteristics appear in Table \ref{tab:table1bmtcrr}.
-We note that the statistical summaries were generated differently for
-different variable types: for continuous variables, we gave the range,
-followed by the mean and standard deviation; for categorical variables,
-we gave the counts for each category.
-
-\begin{CodeChunk}
-\begin{table}
-
-\caption{\label{tab:table1bmtcrr}Baseline characteristics of patients in the stem-cell transplant study.}
-\centering
-\begin{tabular}[t]{lll}
-\toprule
-Variable & Description & Statistical summary\\
-\midrule
-\code{Sex} & Sex & M=Male (100)\\
- &  & F=Female (77)\\
-\code{D} & Disease & ALL (73)\\
- &  & AML (104)\\
-\code{Phase} & Phase & CR1 (47)\\
-\addlinespace
- &  & CR2 (45)\\
- &  & CR3 (12)\\
- &  & Relapse (73)\\
-\code{Source} & Type of transplant & BM+PB (21)\\
- &  & PB (156)\\
-\addlinespace
-\code{Age} & Age of patient (years) & 4--62\\
- &  & 30.47 (13.04)\\
-\code{ftime} & Failure time (months) & 0.13--131.77\\
- &  & 20.28 (30.78)\\
-\code{Status} & Status indicator & 0=censored (46)\\
-\addlinespace
- &  & 1=relapse (56)\\
- &  & 2=competing event (75)\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\end{CodeChunk}
-
-First, we can look at a population-time plot to visualize the incidence
-density of both relapse and the competing events. In Figure
-\ref{fig:compPop}, failure times are highlighted on the plot using red
-points for the event of interest (panel A) and blue dots for competing
-events (panel B). In both panels, we see evidence of a non-constant
-hazard function: the density of points is larger at the beginning of
-follow-up than at the end.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/compPop-1}
-
-}
-
-\caption[Population-time plot for the stem-cell transplant study]{Population-time plot for the stem-cell transplant study.
-A) The red points represent relapse events.
-B) The blue points represent the competing events.}\label{fig:compPop}
-\end{figure}
-\end{CodeChunk}
-
-Our main objective is to compute the absolute risk of relapse for a
-given set of covariates. First, we fit a smooth hazard to the data using
-a linear term for time:
-
-We will compare our hazard ratio estimates to that obtained from a Cox
-regression. To do so, we need to treat the competing event as censoring.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(survival)
-R> # Treat competing event as censoring
-R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-R+                    data = bmtcrr)
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{table}
-
-\caption{\label{tab:bmtcrr-cis}ADD CAPTION}
-\centering
-\begin{tabular}[t]{lrlrl}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
-Covariates & HR & 95\% CI & HR & 95\% CI\\
-\midrule
-Sex & 0.80 & (0.46, 1.39) & 0.68 & (0.39, 1.19)\\
-Disease & 0.58 & (0.32, 1.03) & 0.52 & (0.29, 0.93)\\
-Phase (CR2 vs. CR1) & 1.16 & (0.46, 2.89) & 1.21 & (0.48, 3.02)\\
-Phase (CR3 vs. CR1) & 1.73 & (0.44, 6.72) & 1.67 & (0.43, 6.5)\\
-Phase (Relapse vs. CR1) & 3.93 & (1.82, 8.49) & 4.55 & (2.09, 9.9)\\
-\addlinespace
-Source & 1.65 & (0.54, 5.03) & 1.46 & (0.47, 4.54)\\
-Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\end{CodeChunk}
-
-From the fit object, we can extract both the hazard ratios and their
-corresponding confidence intervals. These quantities appear in Table
-\ref{tab:bmtcrr-cis}. As we can see, the only significant hazard ratio
-identified by case-base sampling is the one associated with the phase of
-the disease at transplant. More precisely, being in relapse at
-transplant is associated with a hazard ratio of 3.89 when compared to
-CR1.
-
-Given the estimate of the hazard function obtained using case-base
-sampling, we can compute the absolute risk curve for a fixed covariate
-profile. We perform this computation for a 35 year old woman who
-received a stem-cell transplant from peripheral blood at relapse. We
-compared the absolute risk curve for such a woman with acute
-lymphoblastic leukemia with that for a similar woman with acute
-myeloblastic leukemia. We will estimate the curve from 0 to 60 months.
-
-We will compare our estimates to that obtained from a corresponding
-Fine-Gray model \citep{fine1999proportional}. The Fine-Gray model is a
-semiparametric model for the cause-specific \emph{subdistribution},
-i.e.~the function \(f_k(t)\) such that
-\[CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),\] where
-\(CI_k(t)\) is the cause-specific cumulative indicence. The Fine-Gray
-model allows to directly assess the effect of a covariate on the
-subdistribution, as opposed to the hazard. For the computation, we will
-use the \pkg{timereg} package \citep{timereg}.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(timereg)
-R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) + 
-R+                           const(Phase) + const(Source) + const(Age), 
-R+                       data = bmtcrr, cause = 1, model = "fg")
-R> 
-R> # Estimate absolute risk curve
-R> risk_fg <- predict(model_fg, newdata, times = time_points)
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/bmtcrr-risk-1}
-
-}
-
-\caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups.
-The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:bmtcrr-risk}
-\end{figure}
-\end{CodeChunk}
-
-Figure \ref{fig:compAbsrisk} shows the absolute risk curves for both
-case-base sampling and the Fine-Gray model. As we can see, the two
-approaches agree quite well for acute myeloblastic leukemia; however,
-there seems to be a difference of about 5\% between the two curves for
-acute lymphoblastic leukemia. This difference does not appear to be
-significant: the curve from case-base sampling is contained within a
-95\% confidence band around the Fine-Gray absolute risk curve (figure
-not shown).
-
-\hypertarget{case-study-3support-data}{%
-\section{Case study 3--SUPPORT Data}\label{case-study-3support-data}}
-
-In the first two case studies, we described the basic functionalities of
-the \pkg{casebase} package: creating population-time plots, fitting
-parametric models for hazard functions, and estimating the corresponding
-cumulative incidence curves. In the next two case studies, we explore
-more advanced features.
-
-For the third case study, we will look at regularized estimation for the
-hazard function. This can be performed using a combination of case-base
-sampling and regularized logistic regression. To demonstrate this
-capability, we examine the Study to Understand Prognoses Preferences
-Outcomes and Risks of Treatment (SUPPORT) dataset
-\citep{knaus1995support}. The SUPPORT dataset tracks death for
-individuals who are considered seriously ill within a hospital.
-Imputation was conducted using predictive mean matching and using the
-\pkg{mice} package in \proglang{R} \citep{mice}. Before imputation,
-there was a total of 1000 individuals and 35 variables. After
-imputation, we have 474 individuals.
-\textcolor{red}{How come we lose individuals during imputation?} Of
-these, 67.5\% experienced the event of interest. A description of each
-variable can be found in Table \ref{tab:support1} and a breakdown of
-each categorical variable in Table \ref{tab:support2}. The original data
-was obtained from the website of the Department of Biostatistics at
-Vanderbilt University. The imputed dataset is available as part of the
-\pkg{casebase} package.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ccccc}
-\hline
-\textbf{Name} & \textbf{Labels}                          & \textbf{Levels} & \textbf{Storage} & \textbf{NAs} \\
-\hline
-age           & Age                                      &                 & double           & 0            \\
-death         & Death at any time up to NDI date:31DEC94 &                 & double           & 0            \\
-sex           &                                          & 2               & integer          & 0            \\
-hospdead      & Death in Hospital                        &                 & double           & 0            \\
-slos          & Days from Study Entry to Discharge       &                 & double           & 0            \\
-d.time        & Days of Follow-Up                        &                 & double           & 0            \\
-dzgroup       &                                          & 8               & integer          & 0            \\
-dzclass       &                                          & 4               & integer          & 0            \\
-num.co        & number of comorbidities                  &                 & double           & 0            \\
-edu           & Years of Education                       &                 & double           & 202          \\
-income        &                                          & 4               & integer          & 349          \\
-scoma         & SUPPORT Coma Score based on Glasgow D3   &                 & double           & 0            \\
-charges       & Hospital Charges                         &                 & double           & 25           \\
-totcst        & Total RCC cost                           &                 & double           & 105          \\
-totmcst       & Total micro-cost                         &                 & double           & 372          \\
-avtisst       & Average TISS, Days 3-25                  &                 & double           & 6            \\
-race          &                                          & 5               & integer          & 5            \\
-meanbp        & Mean Arterial Blood Pressure Day 3       &                 & double           & 0            \\
-wblc          & White Blood Cell Count Day 3             &                 & double           & 24           \\
-hrt           & Heart Rate Day 3                         &                 & double           & 0            \\
-resp          & Respiration Rate Day 3                   &                 & double           & 0            \\
-temp          & Temperature (Celsius) Day 3              &                 & double           & 0            \\
-pafi          & PaO2/(.01*FiO2) Day 3                    &                 & double           & 253          \\
-alb           & Serum Albumin Day 3                      &                 & double           & 378          \\
-bili          & Bilirubin Day 3                          &                 & double           & 297          \\
-crea          & Serum creatinine Day 3                   &                 & double           & 3            \\
-sod           & Serum sodium Day 3                       &                 & double           & 0            \\
-ph            & Serum pH (arterial) Day 3                &                 & double           & 250          \\
-glucose       & Glucose Day 3                            &                 & double           & 470          \\
-bun           & BUN Day 3                                &                 & double           & 455          \\
-urine         & Urine Output Day 3                       &                 & double           & 517          \\
-adlp          & ADL Patient Day 3                        &                 & double           & 634          \\
-adls          & ADL Surrogate Day 3                      &                 & double           & 310          \\
-sfdm2         &                                          & 5               & integer          & 159          \\
-adlsc         & Imputed ADL Calibrated to Surrogate      &                 & double           & 0           
-\end{tabular}
-\caption{A description of each variable in the SUPPORT dataset.}
-\label{tab:support1}
-\end{table}
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{cc}
-\hline
-\textbf{Variable} & \textbf{Levels}                      \\
-\hline
-sex      & female                                        \\
-         & male                                          \\
-dzgroup  & ARF/MOSF w/Sepsis                             \\
-         & COPD                                          \\
-         & CHF                                           \\
-         & Cirrhosis                                     \\
-         & Coma                                          \\
-         & Colon Cancer                                  \\
-         & Lung Cancer                                   \\
-         & MOSF w/Malig                                  \\
-dzclass  & ARF/MOSF                                      \\
-         & COPD/CHF/Cirrhosis                            \\
-         & Coma                                          \\
-         & Cancer                                        \\
-income   & under \$11k                                   \\
-         & $11-$25k                                      \\
-         & $25-$50k                                      \\
-         & \textgreater{}\$50k                           \\
-race     & white                                         \\
-         & black                                         \\
-         & asian                                         \\
-         & other                                         \\
-         & hispanic                                      \\
-sfdm2    & no(M2 and SIP pres)                           \\
-         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
-         & SIP\textgreater{}=30                          \\
-         & Coma or Intub                                 \\
-         & \textless{}2 mo. follow-up                   
-\end{tabular}
-\caption{A description of each level within each categorical variable in the dataset.}
-\label{tab:support2}
-\end{table}
-
-As before, the first step is to visualize the incidence density. The
-population-time plot for the SUPPORT data appears in Figure
-\ref{fig:support-poptime}.
-\textcolor{red}{This population-time plot is strange. The red dots never seem to go above 200.}
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> data(support)
-R> support_popTime <- popTime(support,
-R+                            time = "d.time", event = "death")
-R> plot(support_popTime)
-\end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics{../figures/support-poptime-1}
-
-}
-
-\caption{\label{fig:support-poptime} Population-time plot for the SUPPORT data.}\label{fig:support-poptime}
-\end{figure}
-\end{CodeChunk}
-
-Before we proceed with the analysis, we first remove hospital death as a
-covariate, as this is directly informative of death and cannot be used
-as a predictor. Then we select one observation at random. We will
-estimate the cumulative incidence curve for this individual. The other
-observations will be used to fit the hazard function.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> support <- dplyr::select(support, -hospdead)
-R> 
-R> # Randomly select one observation
-R> indiv <- sample(1:nrow(support), size = 1)
-\end{CodeInput}
-\end{CodeChunk}
-
-Our main objective is to estimate the cumulative incidence of death for
-a given covariate profile. Given the number of covariates available, we
-will do this using regularized estimation. As \pkg{casebase} makes use
-of the \pkg{glmnet} package, we will use the function
-\code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}.
-Specifically, we input a matrix \code{y}, which contains the time and
-event variables, and a matrix \code{x}, which contains all other
-variables we would like to include in the model. For the SUPPORT
-dataset, all factor variables were converted into dummy variables.
-
-We are now ready to estimate the hazard function. We use lasso
-penalization; this is achieved by setting \code{alpha} to 1. Also, note
-that the argument \code{formula_time} can be used to specify the
-function form for time; in our example, we model time linearly.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> support_cb <- fitSmoothHazard.fit(support_train, y, 
-R+                                   formula_time = ~ d.time, 
-R+                                   family = "glmnet", 
-R+                                   event = "death", time = "d.time", 
-R+                                   alpha = 1, ratio = 100)
-\end{CodeInput}
-\end{CodeChunk}
-
-Then, we take our model object and use the \code{absoluteRisk} function
-to compute our estimate of the cumulative incidence function. Since
-\code{glmnet} fitted the penalized logistic regression for a range of
-penalty terms, we need to specify which one we want to use for our
-prediction. We select \code{s = "lambda.1se"}, as recommended.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> time_points <- seq(1, max(support$d.time), by = 0.5)
-R> support_risk <- absoluteRisk(support_cb,
-R+                              time = time_points,
-R+                              newdata = support_test,
-R+                              s = "lambda.1se", method = "numerical")
-\end{CodeInput}
-\end{CodeChunk}
-
-We will compare this curve to a regularized version of Cox regression,
-and a Kaplan-Meier survival curve. Using the \pkg{glmnet} package, we
-can fit the Cox regression model using regularized estimation.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Create survival object for glmnet
-R> surv_glmnet <- survival::Surv(time = y[, "d.time"], 
-R+                               event = y[, "death"])
-R> 
-R> support_cox <- glmnet::cv.glmnet(x = support_train, 
-R+                                  y = surv_glmnet,
-R+                                  family = "cox", alpha = 1)
-\end{CodeInput}
-\end{CodeChunk}
-
-However, \pkg{glmnet} does not provide any functionality for estimating
-the baseline hazard, and the functionalities in the \pkg{survival}
-package do not work on the output from \code{cv.glmnet}. To circumvent
-this issue and get an estimate of the baseline hazard, we followed the
-following steps. First, we fitted a non-regularized Cox model using only
-the variables selected by \pkg{glmnet}; the baseline hazard will be
-estimated using this object. Then, we replaced the estimates of the
-regression coefficients by the estimates from \pkg{glmnet}. Finally,
-this modified \code{coxph} object is fed to the \code{survival::survfit}
-function to obtain the estimated survival function.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Identifying selected covariates
-R> nonzero_covariates_cox <- predict(support_cox, type = "nonzero", 
-R+                                   s = "lambda.1se")
-R> nonzero_coef_cox <- coef(support_cox, s = "lambda.1se")
-R> # Creating a new dataset that only contains the covariates selected by glmnet
-R> support_cox_data <- as.data.frame(cbind(y,
-R+                                         support_train[, nonzero_covariates_cox$X1]))
-R> # Fitting a cox model
-R> support_cox <- survival::coxph(Surv(time = d.time, event = death) ~ ., 
-R+                           data = support_cox_data)
-R> # Replace the estimate with those obtained by glmnet
-R> support_cox$coefficients <- nonzero_coef_cox@x
-R> 
-R> # Fitting CI curve for cox+glmnet
-R> support_cox_risk <- survival::survfit(support_cox,
-R+                                       newdata = as.data.frame(support_test),
-R+                                       time = time_points,
-R+                                       type = "breslow")
-\end{CodeInput}
-\end{CodeChunk}
-
-We used the \pkg{survival} package to calculate the Kaplan-Meier
-estimate of the survival function.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> km <- survival::Surv(time = support$d.time, 
-R+                      event = support$death)
-R> abKm <- survival::survfit(km ~ 1,
-R+                           type = 'kaplan-meier')
-\end{CodeInput}
-\end{CodeChunk}
-
-Now that our three cumulative incidence functions have been calculated,
-we compare them all in Figure \ref{fig:abSC}. Both \pkg{glmnet} and
-\pkg{casebase} have a lower estimated cumulative incidence by the end of
-the study, when compared to Kaplan-Meier.
-\textcolor{red}{It would be good to have an idea of why the case-base and Coxnet curves are so different. At the very least, I think we should compare the select covariates.}
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/abSupportComparison-1}
-
-}
-
-\caption{\label{fig:abSC} Comparison of cumulative incidence curves between regularized casebase in black, regularized Cox regression in red and Kaplan-Meier in blue.}\label{fig:abSupportComparison}
-\end{figure}
-\end{CodeChunk}
-
-\hypertarget{case-study-4stanford-heart-transplant-data}{%
-\section{Case study 4--Stanford Heart Transplant
-Data}\label{case-study-4stanford-heart-transplant-data}}
-
-Although the previous case studies provide a broad overview of the
-capabilities of \pkg{casebase}, they all have two properties in common:
-
-\begin{itemize}
-  \item All covariates are fixed, i.e. they do not change over time;
-  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
-\end{itemize}
-
-Case-base sampling has been used in the literature to study vaccination
-safety, where the exposure period was defined as the week following
-vaccination \citep{saarela2015case}. Hence, the main covariate of
-interest, i.e.~exposure to the vaccine, was changing over time. In this
-context, case-base sampling offers an efficient alternative to nested
-case-control designs or self-matching.
-
-The purpose of the next case study is to show how \pkg{casebase} can
-also be used for survival analysis problems that do not share the
-properties above. To this end, we will use the Stanford Heart Transplant
-Data \citep[\citet{crowley1977covariance}]{clark1971cardiac}; we will
-use the version available in the \pkg{surivival} package.
-
-Recall the setting of this study: patients were admitted to the Stanford
-program after meeting with their physician and determining that they
-were unlikely to respond to other forms of treatment. After enrollment,
-the program searches for a suitable donor for the patient, which can
-take anywhere between a few days to almost a year. We are interested in
-the effect of a heart transplant on survival; therefore, the patient is
-considered exposed only after the transplant has occurred.
-
-As before, we can look at the population-time plot for a graphical
-summary of the event incidence. As we can see, most events occur early
-during the follow-up period, and therefore we do not expect the hazard
-to be constant.
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics{../figures/stanford-poptime-1} \end{center}
-
-\end{CodeChunk}
-
-Since the exposure is time-dependent, we need to manually define the
-exposure variable \emph{after} case-base sampling and \emph{before}
-fitting the hazard function. For this reason, we will use the
-\texttt{sampleCaseBase} function directly.
-
-Next, we will compute the number of days from acceptance into the
-program to transplant, and we use this variable to determine whether
-each population-moment is exposed or not.
-
-Finally, we can fit the hazard using various linear predictors.
-
-Note that the fourth model (i.e. \texttt{fit4}) includes an interaction
-term between exposure and follow-up time. In other words, this model no
-longer exhibit proportional hazards. The evidence of non-proportionality
-of hazards in the Stanford Heart Transplant data has been widely
-discussed \citep{arjas1988graphical}.
-
-We can then compare the goodness of fit of these four models using the
-Akaike Information Criterion (AIC).
-
-\begin{CodeChunk}
-
-\begin{CodeOutput}
-#> Model1 Model2 Model3 Model4 
-#>    794    756    728    715
-\end{CodeOutput}
-\end{CodeChunk}
-
-As we can, the best fit is the fourth model. By visualizing the hazard
-functions for both exposed and unexposed individuals, we can more
-clearly see how the hazards are no longer proportional.
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics{../figures/stanford-hazard-1} \end{center}
-
-\end{CodeChunk}
-
-The non-proportionality seems to be more pronounced at the beginning of
-follow-up than the end. Finally, we can turn these estimate of the
-hazard function into estimates of the cumulative incidence functions.
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics{../figures/stanford-risk-1} \end{center}
-
-\end{CodeChunk}
-
-Note that we can easily adapt the code above to the situation where a
-patient receives a heart transplant at a point in time of interest, for
-example after 30 days.
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics{../figures/stanford-1y-haz-1} \end{center}
-
-\end{CodeChunk}
-
-We can then compare the 1-year mortality risk without transplant and
-with transplant at 30 days.
-
-\begin{CodeChunk}
-
-\begin{CodeOutput}
-#>         
-#> 365 0.33
-\end{CodeOutput}
-
-\begin{CodeOutput}
-#> [1] 0.22
-\end{CodeOutput}
-\end{CodeChunk}
-
-As we can see, the risk estimate at 1-year is about 30\% lower if the
-patient receives a heart transplant at 30 days.
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-In this article, we presented the \proglang{R} package \pkg{casebase}
-that provides functions to fit smooth parametric hazards and estimate
-cumulative incidence functions using case-base sampling. We outlined the
-theoretical underpinnings of the approach, we provided details about our
-implementation, and we illustrated the merits of the approach and the
-package through four case studies.
-
-As a methodological framework, case-base sampling is very flexible. This
-flexibility has been explored before in the literature: for example,
-Saarela and Hanley \citeyearpar{saarela2015case} used case-base sampling
-to model a time-dependent exposure variable in a vaccine safety study.
-As another example, Saarela and Arjas \citeyearpar{saarela2015non}
-combined case-base sampling and a Bayesian non-parametric framework to
-compute individualized risk assessments for chronic diseases. In the
-case studies above, we explored this flexibility along two fronts. On
-the one hand, we showed how splines could be used as part of the linear
-predictor to model the effect of time on the hazard. This strategy
-yielded estimates of the survival function that were qualitatively
-similar to semiparametric estimates derived from Cox regression;
-however, case-base sampling led to estimates that \emph{varying smoothly
-in time}. On the other hand, we also displayed the flexibility of
-case-base sampling by showing how it could be combined with penalized
-logistic regression to perform variable selection. Similarly, our
-package \pkg{casebase} provides the user with the functionality of
-Generalized Additive Models. Moreover, case-base sampling can be
-combined with quasi-likelihood estimation to fit survival models that
-potentially account for the presence of over-dispersion. All of the
-examples illustrate how the case-base sampling framework in general, and
-the package \pkg{casebase} in particular, allows the user to fit a broad
-family of survival functions.
-
-As presented in Hanley \& Miettinen \citeyearpar{hanley2009fitting},
-case-base sampling is comprised of three steps: 1) sampling a case
-series and a base series from the study; 2) fit the log-hazard as a
-linear function of predictors (including time); and 3) use the fitted
-hazard to estimate the cumulative incidence curve. Accordingly, our
-package provides functions for each step. Moreover, the simple interface
-of the \code{fittingSmoothHazard} function resembles the \code{glm}
-interface, which should look familiar to new users. Our modular approach
-also provides a convenient way to extend our package for new sampling or
-fitting strategies.
-
-In the case studies above, we compared the performance of case-base
-sampling with that of Cox regression and Fine-Gray models. In terms of
-function interface, \pkg{casebase} uses a formula interface that is
-closer to that of \code{glm}, in that the event variable is the only
-variable appearing on the left-hand side of the formula. By contrast,
-both \code{survival::coxph} and \code{timereg::comp.risk} use arrays as
-the left-hand side that captures both the event type and time. Both
-approaches to modeling yield user-friendly code. In terms of output,
-both approaches differ significantly. Case-base sampling produces smooth
-hazard and smooth cumulative incidence curves, whereas Cox regression
-and Fine-Gray models produce step-wise cumulative incidence functions
-and never explicitly model the hazard function. Qualitatively, we showed
-that by using splines in the linear predictor, all three models yielded
-similar curves. However, the smooth nature of the output of
-\pkg{casebase} improves interpretation for consumers of these
-predictions. In Table \ref{tab:compCBvsCox}, we provide a side-by-side
-comparison between the Cox model and case-base sampling.
-
-\begin{table}
-\caption{\label{tab:compCBvsCox}Comparison between the Cox model and case-base sampling}
-\centering
-\begin{tabular}[t]{llp{5cm}}
-\toprule
-Feature & Cox model & Case-base sampling\\
-\midrule
-Model type & Semi-parametric & Fully parametric\\
-Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\
-Cumulative incidence & Step function & Smooth-in-time curve\\
-Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\
-Model testing &  & Use GLM framework \newline (e.g.\ LRT, AIC, BIC)\\
-\addlinespace
-Competing risks & Difficult & Cause-specific CIFs\\
-% Prediction & Kaplan-Meier-based & ROC, AUC, risk reclassification probabilities\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-Our choice of modeling the log-hazard as a linear function of covariates
-allows us to develop a simple computational scheme for estimation.
-However, as a downside, it does not allow us to model location and scale
-parameters separately like the package \pkg{flexsurv}. For example, if
-we look at the Weibull distribution as parametrised in
-\code{stats::pweibull}, the log-hazard function is given by
-\[ \log h(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,\]
-where \(\alpha,\beta\) are shape and scale parameters, respectively.
-Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} allows the
-user to model the scale parameter as a function of covariates. Of
-course, this added flexibility comes at the cost of interpretability: by
-modeling the log-hazard directly, the parameter estimates from
-\pkg{casebase} can be interpreted as estimates of log-hazard ratios. To
-improve the flexibility of \pkg{casebase} at capturing the scale of a
-parametric family, we could replace the logistic regression with its
-quasi-likelihood counterpart and therefore model over- and
-under-dispersion with respect to the logistic likelihood. We defer the
-study of the properties and performance of such a model to a future
-article.
-
-Future work will look at some of the methodological extensions of
-case-base sampling. First, to assess the quality of the model fit, we
-may want to study the properties of the residuals (e.g.~Cox-Snell,
-martingale). More work needs to be done to understand these residuals in
-the context of the partial likelihood underlying case-base sampling. The
-resulting diagnostic tools would then be integrated in this package.
-Also, we are interested in extending case-base sampling to account for
-interval censoring. This type of censoring is very common in
-longitudinal studies, and many packages (e.g. \pkg{SmoothHazard},
-\pkg{survival} and \pkg{Rstpm2}) provide functions to account for it.
-Again, we hope to include any resulting methodology as part of this
-package.
-
-In future versions of the package, we also want to increase the
-complement of diagnostic and inferential tools that are currently
-available. For example, we would like to include the ability to compute
-confidence intervals for the cumulative incidence curve. The delta
-method or parametric bootstrap are two different strategies we can use
-to construct approximate confidence intervals. Furthermore, we would
-like to include functions to compute calibration and discrimination
-statistics (e.g.~Brier score, AUC) for our models. Saarela and Arjas
-\citeyearpar{saarela2015non} also describe how to obtain a posterior
-distribution for the AUC from their model.
-
-In conclusion, we presented the \proglang{R} package \pkg{casebase}
-which implements case-base sampling for fitting parametric survival
-models and for estimating smooth cumulative incidence functions using
-the framework of generalized linear models. We strongly believe that its
-flexibility and its foundation on the familiar logistic regression model
-will make it appealing to new and established practioners.
-
-\hypertarget{environment-details}{%
-\section{Environment Details}\label{environment-details}}
-
-This report was generated on 2020-01-29 23:33:21 using the following
-computational environment and dependencies:
-
-\begin{CodeChunk}
-
-\begin{CodeOutput}
-#> - Session info ---------------------------------------------------------------
-#>  setting  value                       
-#>  version  R version 3.6.2 (2019-12-12)
-#>  os       Pop!_OS 19.10               
-#>  system   x86_64, linux-gnu           
-#>  ui       X11                         
-#>  language en_US:en                    
-#>  collate  en_US.UTF-8                 
-#>  ctype    en_US.UTF-8                 
-#>  tz       America/Toronto             
-#>  date     2020-01-29                  
-#> 
-#> - Packages -------------------------------------------------------------------
-#>  package     * version    date       lib
-#>  assertthat    0.2.1      2019-03-21 [1]
-#>  backports     1.1.5      2019-10-02 [1]
-#>  broom         0.5.4      2020-01-27 [1]
-#>  callr         3.4.1      2020-01-24 [1]
-#>  casebase    * 0.2.1.9001 2020-01-27 [1]
-#>  cellranger    1.1.0      2016-07-27 [1]
-#>  cli           2.0.1      2020-01-08 [1]
-#>  codetools     0.2-16     2018-12-24 [4]
-#>  colorspace    1.4-1      2019-03-18 [1]
-#>  cowplot     * 1.0.0      2019-07-11 [1]
-#>  crayon        1.3.4      2017-09-16 [1]
-#>  data.table    1.12.8     2019-12-09 [1]
-#>  DBI           1.1.0      2019-12-15 [1]
-#>  dbplyr        1.4.2      2019-06-17 [1]
-#>  desc          1.2.0      2018-05-01 [1]
-#>  devtools      2.2.1      2019-09-24 [1]
-#>  digest        0.6.23     2019-11-23 [1]
-#>  dplyr       * 0.8.3      2019-07-04 [1]
-#>  ellipsis      0.3.0      2019-09-20 [1]
-#>  evaluate      0.14       2019-05-28 [1]
-#>  fansi         0.4.1      2020-01-08 [1]
-#>  farver        2.0.3      2020-01-16 [1]
-#>  forcats     * 0.4.0      2019-02-17 [1]
-#>  foreach       1.4.7      2019-07-27 [1]
-#>  fs            1.3.1      2019-05-06 [1]
-#>  generics      0.0.2      2018-11-29 [1]
-#>  ggplot2     * 3.2.1      2019-08-10 [1]
-#>  glmnet      * 3.0-2      2019-12-11 [1]
-#>  glue        * 1.3.1      2019-03-12 [1]
-#>  gtable        0.3.0      2019-03-25 [1]
-#>  haven         2.2.0      2019-11-08 [1]
-#>  hms           0.5.3      2020-01-08 [1]
-#>  htmltools     0.4.0      2019-10-04 [1]
-#>  httr          1.4.1      2019-08-05 [1]
-#>  iterators     1.0.12     2019-07-26 [1]
-#>  jsonlite      1.6        2018-12-07 [1]
-#>  kableExtra  * 1.1.0      2019-03-16 [1]
-#>  knitr         1.27       2020-01-16 [1]
-#>  labeling      0.3        2014-08-23 [1]
-#>  lattice       0.20-38    2018-11-04 [4]
-#>  lava          1.6.6      2019-08-01 [1]
-#>  lazyeval      0.2.2      2019-03-15 [1]
-#>  lifecycle     0.1.0      2019-08-01 [1]
-#>  lubridate   * 1.7.4      2018-04-11 [1]
-#>  magrittr    * 1.5        2014-11-22 [1]
-#>  MASS          7.3-51.5   2019-12-20 [4]
-#>  Matrix      * 1.2-18     2019-11-27 [4]
-#>  memoise       1.1.0      2017-04-21 [1]
-#>  mgcv          1.8-31     2019-11-09 [4]
-#>  modelr        0.1.5      2019-08-08 [1]
-#>  munsell       0.5.0      2018-06-12 [1]
-#>  nlme          3.1-143    2019-12-10 [4]
-#>  numDeriv      2016.8-1.1 2019-06-06 [1]
-#>  pillar        1.4.3      2019-12-20 [1]
-#>  pkgbuild      1.0.6      2019-10-09 [1]
-#>  pkgconfig     2.0.3      2019-09-22 [1]
-#>  pkgload       1.0.2      2018-10-29 [1]
-#>  plyr          1.8.5      2019-12-10 [1]
-#>  pracma      * 2.2.9      2019-12-15 [1]
-#>  prettyunits   1.1.1      2020-01-24 [1]
-#>  processx      3.4.1      2019-07-18 [1]
-#>  ps            1.3.0      2018-12-21 [1]
-#>  purrr       * 0.3.3      2019-10-18 [1]
-#>  R6            2.4.1      2019-11-12 [1]
-#>  Rcpp          1.0.3      2019-11-08 [1]
-#>  readr       * 1.3.1      2018-12-21 [1]
-#>  readxl        1.3.1      2019-03-13 [1]
-#>  remotes       2.1.0      2019-06-24 [1]
-#>  reprex        0.3.0      2019-05-16 [1]
-#>  reshape2      1.4.3      2017-12-11 [1]
-#>  rlang         0.4.3      2020-01-24 [1]
-#>  rmarkdown     2.1        2020-01-20 [1]
-#>  rprojroot     1.3-2      2018-01-03 [1]
-#>  rstudioapi    0.10       2019-03-19 [1]
-#>  rticles       0.13       2019-12-19 [1]
-#>  rvest         0.3.5      2019-11-08 [1]
-#>  scales        1.1.0      2019-11-18 [1]
-#>  sessioninfo   1.1.1      2018-11-05 [1]
-#>  shape         1.4.4      2018-02-07 [1]
-#>  stringi       1.4.5      2020-01-11 [1]
-#>  stringr     * 1.4.0      2019-02-10 [1]
-#>  survival    * 3.1-8      2019-12-03 [4]
-#>  testthat      2.3.1      2019-12-01 [1]
-#>  tibble      * 2.1.3      2019-06-06 [1]
-#>  tidyr       * 1.0.2      2020-01-24 [1]
-#>  tidyselect    0.2.5      2018-10-11 [1]
-#>  tidyverse   * 1.3.0      2019-11-21 [1]
-#>  timereg     * 1.9.4      2019-07-30 [1]
-#>  usethis       1.5.1      2019-07-04 [1]
-#>  vctrs         0.2.2      2020-01-24 [1]
-#>  VGAM          1.1-2      2019-11-21 [1]
-#>  viridisLite   0.3.0      2018-02-01 [1]
-#>  webshot       0.5.2      2019-11-22 [1]
-#>  withr         2.1.2      2018-03-15 [1]
-#>  xfun          0.12       2020-01-13 [1]
-#>  xml2          1.2.2      2019-08-09 [1]
-#>  yaml          2.2.0      2018-07-25 [1]
-#>  source                                  
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  Github (sahirbhatnagar/casebase@24359be)
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.1)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#>  CRAN (R 3.6.2)                          
-#> 
-#> [1] /home/sahir/R/x86_64-pc-linux-gnu-library/3.6
-#> [2] /usr/local/lib/R/site-library
-#> [3] /usr/lib/R/site-library
-#> [4] /usr/lib/R/library
-\end{CodeOutput}
-\end{CodeChunk}
-
-The current Git commit details are:
-
-\begin{CodeChunk}
-
-\begin{CodeOutput}
-#> Local:    write-popTime /home/sahir/git_repositories/cbpaper
-#> Remote:   write-popTime @ origin (https://github.com/sahirbhatnagar/cbpaper.git)
-#> Head:     [d990874] 2020-01-28: cleaning up case study 1. merging into one figure. and echo=FALSE defaults for
-\end{CodeOutput}
-\end{CodeChunk}
-
 \bibliography{references.bib}

### 2020-02-06 analysis/paper/paper.Rmd
add discussion of flexible sampling

--- 2020-02-05 analysis/paper/paper.Rmd
@@
 Their approach could also be included in \pkg{casebase}.
+Finally, we want to provide more flexibility in how the case-base sampling is performed.
+This could be achieved by adding a \code{hazard} argument to the function \code{sampleCaseBase}.
+In this way, users could specify their own sampling mechanism.
+For example, they could provide a hazard that gives sampling probabilities that are proportional to the cardiovascular disease event rate given by the Framingham score [@saarela2015non].
 

### 2020-02-06 analysis/paper/paper.tex
add discussion of flexible sampling

--- 2020-02-05 analysis/paper/paper.tex
@@
 distribution for the AUC from their model. Their approach could also be
-included in \pkg{casebase}.
+included in \pkg{casebase}. Finally, we want to provide more flexibility
+in how the case-base sampling is performed. This could be achieved by
+adding a \code{hazard} argument to the function \code{sampleCaseBase}.
+In this way, users could specify their own sampling mechanism. For
+example, they could provide a hazard that gives sampling probabilities
+that are proportional to the cardiovascular disease event rate given by
+the Framingham score \citep{saarela2015non}.
 
@@
 
-This report was generated on 2020-02-05 17:43:41 using the following
+This report was generated on 2020-02-06 13:51:22 using the following
 computational environment and dependencies:
@@
 #> Local:    intro /home/turgeonm/Documents/git_repos/cbpaper
-#> Head:     [c926a9e] 2020-01-21: Merge pull request #15 from sahirbhatnagar/fix-casebase-dependency
+#> Head:     [92cad64] 2020-02-05: add intro and edit some of the sections
 \end{CodeOutput}

### 2020-02-06 analysis/paper/paper.tex
add discussion of flexible sampling

@@
 sampling mechanism are proportional to the cardiovascular disease event
-rate given by the Framingham score.
+rate given by the Framingham score. With this sampling mechanism,
+Saarela \& Arjas are able to increase the efficiency of their
+estimators, when compared to uniform sampling.
 
@@
 
-This report was generated on 2020-02-06 13:51:22 using the following
+This report was generated on 2020-02-07 10:31:46 using the following
 computational environment and dependencies:
@@
 #> Local:    intro /home/turgeonm/Documents/git_repos/cbpaper
-#> Head:     [92cad64] 2020-02-05: add intro and edit some of the sections
+#> Head:     [d5e7d09] 2020-02-07: add discussion of flexible sampling
 \end{CodeOutput}

### 2020-02-06 Untitled.Rmd
Switched to non interaction model
note that I only ran CS3 so the pdf is not complete. fixes include all red text

Jesse Still needs to clean up the text and deliberate his write up in the case study. Code will also be cleaned to be more readable.

---
title: "ipareport"
author: "Jesse"
date: "30/01/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r run}
library(casebase)
library(survival)
library(riskRegression)
data(veteran)

new_data <- data.frame(trt = "test",
                       celltype = "adeno",
                       karno = median(veteran$karno),
                       diagtime = median(veteran$diagtime),
                       age = median(veteran$age),
                       prior = "no")

modelKM <- coxph(Surv(time, status) ~ 1,
                  data = veteran, y=TRUE, x = TRUE)
abKM<- survfit(modelKM)

modelCox <- coxph(Surv(time, status) ~ karno + diagtime + age + prior + celltype + trt,
                data = veteran, y=TRUE, x = TRUE)
abCox<- survfit(modelCox)

modelCB <- fitSmoothHazard(status ~ time + karno + diagtime + age + prior +
                              celltype + trt, data = veteran, ratio = 100)
abCB <- absoluteRisk(object = modelCB, time = seq(0,11, sep=99)) ##takes long to run

plot(abCB[,1],rowMeans(abCB[,-c(1)]),type='l',col="black",lwd=3,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
lines(abCox,col="red",fun="event",lwd=3,conf.int = FALSE)
lines(abKM ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
legend("bottomright",
       legend = c( "Casebase (Lasso+linear)","semi-parametric (Cox)+glmnet","KM curve"),
       col = c("black","red","Blue"),
       lty = c(1, 1, 1),
       bg = "gray90")

s=Score(list("KM"=modelKM,"Cox"=modelCox),
        formula=Surv(time,status)~1,data=veteran,conf.int=FALSE,times=c(5,8))

BrierNoCensoring=sum((1-rowMeans(abCB[,-c(1)]))^2)/length(abCB[,1])

fit=modelCox
brierErrorCox <- function(fit, Time,data) {
    cfit <- survfit(Surv(time, status==0)~1, data) #KM curve
    sfit <- survfit(fit, newdata = data)  # model curve for fit2
    S <- summary(sfit, Time)$surv
    cProb <- summary(cfit, Time)$surv
    Score <- matrix(NA, nrow(S), ncol(S))
    for (i in 1:length(Time)) {
        CensBefore <- fit$y[,2]==0 & fit$y[,1] < Time[i]
        y <- fit$y[,1] > Time[i]
            Score[i,] <- (y - S[i,])^2

        Score[i, CensBefore] <- 0
    }
    Err <- sweep(Score, 1, cProb, '/')
    apply(Err, 1, mean)
}
fit=modelCB
data=veteran
brierErrorCB <- function(fit, Time,data) {
    cfit <- survfit(Surv(time, status==0)~1, data) #KM curve
    sfit <- absoluteRisk(object = modelCB, time=Time)  # model curve for fit2
    S <- 1-sfit[,-c(1)]
    cProb <- summary(cfit, Time)$surv
    Score <- matrix(NA, nrow(S), ncol(S))
    for (i in 1:length(Time)) {
        CensBefore <- data$status==0 & fit$originalData$time < Time[i]
        y <- fit$originalData$time > Time[i]
        Score[i,] <- (y - S[i,])^2
        Score[i, CensBefore] <- 0
    }

    Err <- sweep(Score, 1, cProb, '/')
    Err=rowMeans(Err)

}

Time <- seq(0, 300, len=200)
errCox <- cbind(brierErrorCox(modelCox, Time,veteran))
errCB<- cbind(brierErrorCB(modelCB,Time,veteran))
errors<-data.frame(coxPredictionError=errCox,cbPredictionError=errCB)
library(ggplot2)
library(dplyr)
library(tidyr)
errors=na.omit(errors)  %>%
    gather(key,value, coxPredictionError, cbPredictionError)
errors$time=Time

errors%>%
    ggplot(aes(x=time,y=value,col=key,group=key))+geom_line(lwd=1)+

    labs(x = "time",y="Brier prediction error")

```

### 2020-02-06 analysis/paper/paper.Rmd
Switched to non interaction model
note that I only ran CS3 so the pdf is not complete. fixes include all red text

Jesse Still needs to clean up the text and deliberate his write up in the case study. Code will also be cleaned to be more readable.

--- 2020-02-05 analysis/paper/paper.Rmd
@@
 library(glmnet)
+library(data.table)
 library(kableExtra)
-library(pracma)
 options(digits = 2, scipen = 999)
@@
 
-eval_introduction <- TRUE
-eval_theory <- TRUE
-eval_implementation <- TRUE
-eval_cs1 <- TRUE
-eval_cs2 <- TRUE
+eval_introduction <- FALSE
+eval_theory <- FALSE
+eval_implementation <- FALSE
+eval_cs1 <- FALSE
+eval_cs2 <- FALSE
 eval_cs3 <- TRUE
-eval_cs4 <- TRUE
+eval_cs4 <- FALSE
 ```
@@
 
-Survival analysis has been greatly influenced over the last 50 years by the partial likelihood approach of the Cox proportional hazard model [@cox1972regression].
-This approach provides a flexible way of assessing the influence of covariates on the hazard function, without the need to specify a parametric survival model.
-This flexibility comes at the cost of decoupling the baseline hazard from the effect of the covariates.
-To recover the whole survival curve---or the cumulative incidence function (CIF)---we then need to separately estimate the baseline hazard [@breslow1972discussion].
-This in turn often leads to stepwise estimates of the survival function.
-
-From the perspective of clinicians and their patients, the most relevant quantity is often the 5- or 10-year risk of having a certain event given the patient's particular circumstances, and not the hazard ratio between a treatment and control group.
-Therefore, to make sound clinical decisions, it is important to accurately estimate the *full* hazard function, which can then be used to estimate the cumulative incidence function (CIF).
-From this perspective, the CIF and the survival function estimates are now smooth functions of time.
-
-With the goal of fitting smooth-in-time hazard functions, Hanley & Miettinen [-@hanley2009fitting] proposed a general framework for estimating fully parametric hazard models via logistic regression.
-Their approach provides users that are familiar with generalized linear models with a natural way of fitting parametric survival models.
-Moreover, their framework is very flexible: general functions of time can be estimated (e.g.
-using splines or general additive models), and hence these models retain some of the flexibility of partial likelihood approaches.
-
-In this article, we present an \proglang{R} package that combines the ideas of Hanley & Miettinen into a simple interface.
-The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile.
-Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function.
-
-In what follows, we first recall some theoretical details on case-base sampling and its use for estimating parameteric hazard functions.
-We then give a short review of existing \proglang{R} packages that implement comparable features as \pkg{casebase}.
-Next, we provide some details about the implementation of case-base sampling in our package, and we give a brief survey of its main functions.
-This is followed by four case studies that illustrate the flexibility and capabilities of \pkg{casebase}.
-We show how the same framework can be used for competing risk analyses, penalized estimation, and for studies with time-dependent exposures.
-Finally, we end the article with a discussion of the results and of future directions.
+The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence (CI)) of an event, for a particular patient.
+The following points should be noted:
+
+  1. hazard ratios
+  2. If, however, the absolute risks are of interest, they have to be recovered using the semi-parametric Breslow estimator
+  3. Alternative approaches for fitting flexible hazard models for estimating absolute risks, not requiring this two-step approach? Yes! [@hanley2009fitting]
+
+[@hanley2009fitting] propose a fully parametric hazard model that can be fit via logistic regression.
+From the fitted hazard function, cumulative incidence and, thus, risk functions of time, treatment and profile can be easily derived.
 
@@
 
-In Hanley & Miettinen [-@hanley2009fitting], the authors suggest sampling the base series *uniformly* from the study base.
+In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base sampling *uniformly*, i.e.
+to sample the base series uniformly from the study base.
 In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base.
@@
 
-<!-- The `casebase` package fits the family of hazard functions of the form -->
-
-<!-- $$ h(t;X) = \exp[g(t;X)] $$ -->
-<!-- where $t$ denotes time and $X$, the individual's covariate profile.  -->
-
-Let $g(t; X)$ be the linear predictor such that $\log(h(t;X)) = g(t; X)$.
+The `casebase` package fits the family of hazard functions of the form
+
+$$ h(t;X) = \exp[g(t;X)] $$
+where $t$ denotes time and $X$, the individual's covariate profile.
 Different functions of $t$ lead to different parametric hazard models.
-The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$:
-
-$$ \log(h(t; X)) = \beta_0 + \beta_1 X.$$
-
-The instantaneous failure rate is independent of $t$.\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
-This is also known as the \textit{memoryless property} \citep{kalbfleisch2011statistical}.}
+The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$.
+
+$$ h(t;X) = \exp(\beta_0 + \beta_1 X) $$
+
+The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
+This is also known as the *memoryless property* (Kalbfleisch and Prentice, 2002).
 
@@
 
-$$ \log(h(t; X)) = \beta_0 + \beta_1 t + \beta_2 X.$$
-
-Use of $\log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time [@kalbfleisch2011statistical]:
-
-$$ \log(h(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. $$
+$$ h(t;X)  = \exp(\beta_0 + \beta_1 t + \beta_2 X) $$
+
+Use of $\log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
+
+$$ h(t;X)  = \exp(\beta_0 + \beta_1 \log(t) + \beta_2 X) $$
 
@@
 where $N_j(t)$ is the counting process associated with the event of type $j$ and $h_j(t)$ is the corresponding hazard function.
-As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}
+As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1\end{align*}}
 
@@
 
-In the next case study, we show how case-base sampling can  be used in the context of a competing risk analysis.
+In the next case study, we show how case-base sampling can also be used in the context of a competing risk analysis.
 For illustrative purposes, we will use the same data that was used in Scrucca *et al* [-@scrucca2010regression].
@@
 First, we can look at a population-time plot to visualize the incidence density of both relapse and the competing events.
-In Figure \ref{fig:compPop}, failure times are highlighted on the plot using red dots for the event of interest (panel A) and blue dots for competing events (panel B).
+In Figure \ref{fig:compPop}, failure times are highlighted on the plot using red points for the event of interest (panel A) and blue dots for competing events (panel B).
 In both panels, we see evidence of a non-constant hazard function: the density of points is larger at the beginning of follow-up than at the end.
@@
 
-Our main objective is to compute the absolute risk of relapse for a given set of covariates. We start by fitting a smooth hazard to the data using a linear term for time:
+Our main objective is to compute the absolute risk of relapse for a given set of covariates. First, we fit a smooth hazard to the data using a linear term for time:
 
@@
                col.names = c("Covariates", "HR", "95% CI", "HR", "95% CI"),
-               caption = "Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.") %>% 
+               caption = "ADD CAPTION") %>% 
   kable_styling() %>% 
@@
 
-We will compare our estimates to that obtained from a corresponding Fine-Gray model [-@fine1999proportional].
+We will compare our estimates to that obtained from a corresponding Fine-Gray model [@fine1999proportional].
 The Fine-Gray model is a semiparametric model for the cause-specific *subdistribution*, i.e.
@@
 The Fine-Gray model allows to directly assess the effect of a covariate on the subdistribution, as opposed to the hazard.
-For the computation, we will use the \pkg{timereg} package [@timereg]:
+For the computation, we will use the \pkg{timereg} package [@timereg].
 
@@
 
-For the third case study, we will look at regularized estimation for the hazard function.
+For the third case study, we will introduce regularization into the estimation of the hazard function.
 This can be performed using a combination of case-base sampling and regularized logistic regression.
@@
 The SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital.
-Imputation was conducted using predictive mean matching and using the \pkg{mice} package in \proglang{R} [@mice].
-Before imputation, there was a total of 1000 individuals and 35 variables.
-After imputation, we have 474 individuals.
-\textcolor{red}{How come we lose individuals during imputation?} Of these, 67.5% experienced the event of interest.
+Imputation was conducted using the \pkg{mice} package in \proglang{R} with default settings [@mice].
+Before imputation, there were a total of 1000 individuals and 35 variables.
+charges, Totcst, totmcst and sfdm2 were removed as they are response variables we are not interested in for the sake of this analysis.
+After imputation, we have 690 individuals.
+We lose individuals due to coliniarity between variables, preventing imputation.
+If a pair of covariates are deemed colinear to a certain degree, they are excluded from any imputations.
+Here, only edu  income  avtisst*  race  ph*  glucose*  adlp* were imputed.
+Of these 690 individuals, 66.09% experienced the event of interest.
 <!--The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.--> A description of each variable can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}.
-The original data was obtained from the website of the Department of Biostatistics at Vanderbilt University.
+The original data was obtained from the Department of Biostatistics at Vanderbilt University.
 The imputed dataset is available as part of the \pkg{casebase} package.
@@
 The population-time plot for the SUPPORT data appears in Figure \ref{fig:support-poptime}.
-\textcolor{red}{This population-time plot is strange.
-The red dots never seem to go above 200.}
-
-```{r support-poptime, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data."}
+
+```{r support-poptime, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data.
+Red points represent the case series.
+The gray space demonstrates the base series."}
 data(support)
 support_popTime <- popTime(support,
-                           time = "d.time", event = "death")
+                           time = "d.time", event = "death",percentile_number = 0.31)
 plot(support_popTime)
@@
 
-Before we proceed with the analysis, we first remove hospital death as a covariate, as this is directly informative of death and cannot be used as a predictor.
-Then we select one observation at random.
-We will estimate the cumulative incidence curve for this individual.
-The other observations will be used to fit the hazard function.
-
-```{r supportSetUp, echo = TRUE, eval = eval_cs3}
-support <- dplyr::select(support, -hospdead)
-
-# Randomly select one observation
-indiv <- sample(1:nrow(support), size = 1)
-```
-
-Our main objective is to estimate the cumulative incidence of death for a given covariate profile.
-Given the number of covariates available, we will do this using regularized estimation.
-As \pkg{casebase} makes use of the \pkg{glmnet} package, we will use the function \code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}.
-Specifically, we input a matrix \code{y}, which contains the time and event variables, and a matrix \code{x}, which contains all other variables we would like to include in the model.
-For the SUPPORT dataset, all factor variables were converted into dummy variables.
-
-```{r support-design, eval = eval_cs3}
-# Create design matrix without the time variable
-support_matrix <- model.matrix(death ~ . - d.time, 
-                               data = support)
-# Remove intercept term
-support_matrix <- support_matrix[,colnames(support_matrix) != "(Intercept)"]
-
-# Create test and train dataset
-support_test <- support_matrix[indiv,,drop = FALSE]
-
-support_train <- support_matrix[-indiv,]
-y <- as.matrix(support[-indiv, c("death", "d.time")])
-```
-
-We are now ready to estimate the hazard function.
-We use lasso penalization; this is achieved by setting \code{alpha} to 1.
-Also, note that the argument \code{formula_time} can be used to specify the function form for time; in our example, we model time linearly.
-
-```{r CBhazard, echo = TRUE, eval = eval_cs3}
-support_cb <- fitSmoothHazard.fit(support_train, y, 
-                                  formula_time = ~ d.time, 
-                                  family = "glmnet", 
-                                  event = "death", time = "d.time", 
-                                  alpha = 1, ratio = 100)
-```
-
-Then, we take our model object and use the \code{absoluteRisk} function to compute our estimate of the cumulative incidence function.
-Since \code{glmnet} fitted the penalized logistic regression for a range of penalty terms, we need to specify which one we want to use for our prediction.
-We select \code{s = "lambda.1se"}, as recommended.
-
-```{r CBabsolute, echo = TRUE, eval = eval_cs3}
-time_points <- seq(1, max(support$d.time), by = 0.5)
-support_risk <- absoluteRisk(support_cb,
-                             time = time_points,
-                             newdata = support_test,
-                             s = "lambda.1se", method = "numerical")
-```
-
-We will compare this curve to a regularized version of Cox regression, and a Kaplan-Meier survival curve.
-Using the \pkg{glmnet} package, we can fit the Cox regression model using regularized estimation.
-
-```{r coxHaz, echo = TRUE, eval = eval_cs3}
-# Create survival object for glmnet
-surv_glmnet <- survival::Surv(time = y[, "d.time"], 
-                              event = y[, "death"])
-
-support_cox <- glmnet::cv.glmnet(x = support_train, 
-                                 y = surv_glmnet,
-                                 family = "cox", alpha = 1)
-```
-
-However, \pkg{glmnet} does not provide any functionality for estimating the baseline hazard, and the functionalities in the \pkg{survival} package do not work on the output from \code{cv.glmnet}.
-To circumvent this issue and get an estimate of the baseline hazard, we followed the following steps.
-First, we fitted a non-regularized Cox model using only the variables selected by \pkg{glmnet}; the baseline hazard will be estimated using this object.
-Then, we replaced the estimates of the regression coefficients by the estimates from \pkg{glmnet}.
-Finally, this modified \code{coxph} object is fed to the \code{survival::survfit} function to obtain the estimated survival function.
-
-```{r coxAbsolute, echo = TRUE, eval = eval_cs3}
-# Identifying selected covariates
-nonzero_covariates_cox <- predict(support_cox, type = "nonzero", 
-                                  s = "lambda.1se")
-nonzero_coef_cox <- coef(support_cox, s = "lambda.1se")
-# Creating a new dataset that only contains the covariates selected by glmnet
-support_cox_data <- as.data.frame(cbind(y,
-                                        support_train[, nonzero_covariates_cox$X1]))
-# Fitting a cox model
-support_cox <- survival::coxph(Surv(time = d.time, event = death) ~ ., 
-                          data = support_cox_data)
-# Replace the estimate with those obtained by glmnet
-support_cox$coefficients <- nonzero_coef_cox@x
-
-# Fitting CI curve for cox+glmnet
-support_cox_risk <- survival::survfit(support_cox,
-                                      newdata = as.data.frame(support_test),
-                                      time = time_points,
-                                      type = "breslow")
-```
-
-We used the \pkg{survival} package to calculate the Kaplan-Meier estimate of the survival function.
+Before we proceed with the analysis, we first remove hospital death as a covariate, as this is directly informative of death.
+All covariates will be used to estimate the hazard function.
+The hazard function will be used to estimate the cumulative incidence of each individual in the study.
+Then, the average cumulative incidience curve for all observations will be used for our comparison.
+
+```{r supportSetUp, echo = FALSE, eval = eval_cs3}
+completeData=casebase::support
+#remove hospital dead
+completeData<-completeData[,-c(4)]
+ratio=100  #universal ratio for eval_cs3
+#keep one individual out, at random, for which we will develop an 
+#absolute risk curve.
+#sam=sample(1:nrow(completeData), 1)
+#Set workingData to the remaining individuals in the data
+#workingData=completeData[-c(sam),]
+workingCompleteData=model.matrix(death~ .-d.time,data=completeData)[,-c(1)]#remove intercept
+#Create u and xCox, which will be used when fitting Cox with glmnet.
+newData=workingCompleteData
+x=workingCompleteData
+#x and y will be used to fit the casebase model under glmnet.
+#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
+y=data.matrix(completeData[,c(2,5)])
+```
+
+Our main objective is to compute the absolute risk of death for a given set of covariates, using regularization when fitting our model.
+First, we fit a smooth hazard to the data using a weibull distribution.
+As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model.
+For the SUPPORT dataset, all factor variables were converted into indicator variables.
+Here, we used lasso penalization by setting alpha to 1.
+
+```{r cbHazardInteractions}, echo = TRUE, eval = eval_cs3}
+
+cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
+                                       formula_time = ~log(d.time),alpha=1,ratio=ratio)
+```
+Then, we take our model and use the `absoluteRisk` function to integrate and retrieve our cumulative incidence curves.
+
+```{r CBabsolute, echo = FALSE, eval = eval_cs3}
+#Estimating the absolute risk curve using the newData parameter.
+casebaseCI=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 20),newdata = x , s="lambda.1se",method=c("numerical"))
+
+cbCIAve=data.frame(casebaseCI[,1],rowMeans(casebaseCI[,-c(1)]))
+```
+We will compare this curve to a regularized version of cox regression, and a kaplan-meier survival curve.
+The cox regression absolute risk curve with lasso penalization required two extra steps to retrieve.
+Coxnet from the glmnet package has tools to fit a regularized cox model, but requires a survival object skeleton of the selected variables, so that the survival package tools can be used to retrive the absolute risk.
+We handle this by fitting a regularized cox model with the glmnet package, and a cox model from the survival package with the selected variables from the regularized fit.
+This second model serves as a skeleton that permits the use of survival package functions.
+The coefficients from the regularized model will replace the coefficients in our skeleton.
+the resulting model will have the correct coefficients when integrating, but will have an incorrect standard error.
+For the purposes of this case study, we are only interested in the absolute risk curve itself and not the standard error.
+
+```{r coxHazAbsolute, echo = FALSE, eval = eval_cs3}
+#Create u and xCox, which will be used when fitting Cox with glmnet.
+u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
+xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))
+#hazard for cox using glmnet
+coxglmFit=glmnet::cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
+#convergence demonstrated in plot
+#plot(coxglmFit)
+#taking the coefficient estimates for later use
+nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
+nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
+#creating a new dataset that only contains the covariates chosen through glmnet.
+#cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
+cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
+#newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
+#fitting a cox model using regular estimation, however we will not keep it.
+#this is used more as an object place holder.
+coxFit <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
+#The coefficients of this object will be replaced with the estimates from coxglmFit.
+#Doing so makes it so that everything is invalid aside from the coefficients.
+#In this case, all we need to estimate the absolute risk is the coefficients.
+#Std. error would be incorrect here, if we were to draw error bars.
+coxFit$coefficients<-nonzero_coef_cox@x
+#Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
+newDataCox=newData[,nonzero_covariate_cox$X1-1]
+abCoxFit<-survival::survfit(coxFit,newdata=as.data.frame(newDataCox),time = seq(0,max(completeData$d.time), 1),type="breslow")
+abCF<-abCoxFit
+abCF$surv<-rowMeans(abCF$surv)
+```
+We used the survival package to calculate cox regression's cumulative incidience without penalization and the kaplan-meier curve for incidience.
+
+```{r unpenalizedCox, echo = FALSE, eval = eval_cs3}
+Cox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,-c(1)]))
+#coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
+#abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)
+#abCFup<-abcoxFitup
+#abCFup$surv<-rowMeans(abCFup$surv)
+#cb.ModelNorm=casebase::fitSmoothHazard(V2~.,time="V1",event="V2",
+       #                                ratio=rat,data=fullDataCox)
+#summary(cb.ModelNorm)#need to figure out how to deal with NAs
+#casebaseAbsoluteNorm=casebase::absoluteRisk(cb.ModelNorm,time = seq(1,max(completeData$d.time), 200),newdata = newData , s="lambda.1se",method=c("numerical"))
+#cabn=data.frame(casebaseAbsoluteNorm[,1],rowMeans(casebaseAbsoluteNorm[,-c(1)]))
+```
 
 ```{r fitKM+abRisk, echo = TRUE, eval = eval_cs3}
-km <- survival::Surv(time = support$d.time, 
-                     event = support$death)
-abKm <- survival::survfit(km ~ 1,
-                          type = 'kaplan-meier')
-```
-
-Now that our three cumulative incidence functions have been calculated, we compare them all in Figure \ref{fig:abSC}.
-Both \pkg{glmnet} and \pkg{casebase} have a lower estimated cumulative incidence by the end of the study, when compared to Kaplan-Meier.
-\textcolor{red}{It would be good to have an idea of why the case-base and Coxnet curves are so different.
-At the very least, I think we should compare the select covariates.}
-
-```{r abSupportComparison, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Comparison of cumulative incidence curves between regularized casebase in black, regularized Cox regression in red and Kaplan-Meier in blue."}
-plot(support_risk, type='l', col = "black", lwd = 2, 
-     main = "Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves", 
-     xlab = "Survival-Time", ylab = "Cumulative Incidence", ylim = c(0,1))
-lines(support_cox_risk, col = "red", fun = "event", lwd = 3, conf.int = FALSE)
-lines(abKm, col = "blue", fun = "event", lwd = 3, conf.int = FALSE)
-legend("bottom", 
-       legend = c("casebase", "Cox+glmnet", "KM curve"), 
-       col = c("black", "red", "blue"), lty = c(1, 1, 1), bg = "gray90")
+
+coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
+abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)
+
+#creating a surv object to be used to fit an unadjusted absolute risk curve.
+# (Kaplan-Meier risk curve)
+km <- survival::Surv(time = completeData$d.time, event = completeData$death)
+abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
+```
+
+Now that our four cumulative incidence estimations have been calculated, we compare them all in Figure \ref{fig:abSC}.
+Both coxnet and casebase decrease the absolute risk by the end of the study, in comparison to kaplan-meier.
+```{r abSupportComparison, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
+abCFup<-abcoxFitup
+abCFup$surv<-rowMeans(abCFup$surv)
+# A plot to compare all three Absolute risk (Commulative Incidence)
+plot(cbCIAve[,1],cbCIAve[,2],type='l',col="black",lwd=3,main="Cumulative Incidence comparison",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
+lines(abCF,col="red",fun="event",lwd=3,conf.int = FALSE)
+lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
+lines(abCFup ,col="purple",fun="event",lwd=3,conf.int = FALSE)
+legend("bottomright", 
+       legend = c( "Penalized Weibull with casebase","Coxnet","Cox","Kaplan-Meier"), 
+       col = c("black","red","Blue","purple"),
+       lty = c(1, 1, 1), 
+       bg = "gray90")
+```
+
+```{r covariatesRegularizedLollipopPlot,eval=FALSE}
+
+#get covariates excluding d.time
+estimatescb=setDT(as.data.frame(coef(cb.Model)[-c(1,2),1]), keep.rownames = TRUE)[]
+estimatescb$Model="casebase"
+colnames(estimatescb)<-c("Coefs","Estimate","Model")
+estimatescox=setDT(as.data.frame(coef(coxglmFit)[-c(1),1]), keep.rownames = TRUE)[]
+estimatescox$Model="Cox"
+colnames(estimatescox)<-c("Coefs","Estimate","Model")
+estimates=rbind(estimatescb, estimatescox)
+estimates$Estimate[estimates$Estimate==0]=NA# make it so unchosen covariates appear empty
+library(ggstance)# for vertical dodge
+  ggplot(estimates, aes(x = Estimate, y = Coefs, color = Model)) +
+        geom_segment(aes(x = 0, y = Coefs, xend = Estimate, yend = Coefs),size=0.5 ) +
+        geom_point(position=ggstance::position_dodgev(height=0.2))
+
 ```
@@
 
-In the previous case studies, we only considered covariates that were fixed at baseline.
-In this next case study, we will use the Stanford Heart Transplant data [@clark1971cardiac,@crowley1977covariance] to show how case-base sampling can also be used in the context of time-varying covariates.
-As an example that already appeared in the literature, case-base sampling was to study vaccination safety, where the exposure period was defined as the week following vaccination [@saarela2015case].
+Although the previous case studies provide a broad overview of the capabilities of \pkg{casebase}, they all have two properties in common:
+\begin{itemize}
+  \item All covariates are fixed, i.e. they do not change over time;
+  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
+\end{itemize}
+
+Case-base sampling has been used in the literature to study vaccination safety, where the exposure period was defined as the week following vaccination [@saarela2015case].
 Hence, the main covariate of interest, i.e.
@@
 
-Recall the setting of Stanford Heart Transplant study: patients were admitted to the Stanford program after meeting with their physician and determining that they were unlikely to respond to other forms of treatment.
-After enrollment, the program searched for a suitable donor for the patient, which could take anywhere between a few days to almost a year.
+The purpose of the next case study is to show how \pkg{casebase} can also be used for survival analysis problems that do not share the properties above.
+To this end, we will use the Stanford Heart Transplant Data [@clark1971cardiac,@crowley1977covariance]; we will use the version available in the \pkg{survival} package.
+
+Recall the setting of this study: patients were admitted to the Stanford program after meeting with their physician and determining that they were unlikely to respond to other forms of treatment.
+After enrollment, the program searches for a suitable donor for the patient, which can take anywhere between a few days to almost a year.
 We are interested in the effect of a heart transplant on survival; therefore, the patient is considered exposed only after the transplant has occurred.
@@
 The non-proportionality seems to be more pronounced at the beginning of follow-up than the end.
-Finally, we can turn these estimates of the hazard function into estimates of the cumulative incidence functions.
+Finally, we can turn these estimate of the hazard function into estimates of the cumulative incidence functions.
 
@@
 
-In this article, we presented the \proglang{R} package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating cumulative incidence functions using case-base sampling.
-We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through four case studies.
-
-As a methodological framework, case-base sampling is very flexible.
-This flexibility has been explored before in the literature: for example, Saarela and Hanley [-@saarela2015case] used case-base sampling to model a time-dependent exposure variable in a vaccine safety study.
-As another example, Saarela and Arjas [-@saarela2015non] combined case-base sampling and a Bayesian non-parametric framework to compute individualized risk assessments for chronic diseases.
-In the case studies above, we explored this flexibility along two fronts.
-On the one hand, we showed how splines could be used as part of the linear predictor to model the effect of time on the hazard.
-This strategy yielded estimates of the survival function that were qualitatively similar to semiparametric estimates derived from Cox regression; however, case-base sampling led to estimates of the survival function that *vary smoothly in time*.
-On the other hand, we also displayed the flexibility of case-base sampling by showing how it could be combined with penalized logistic regression to perform variable selection.
-Even though we did not illustrate it in this article, case-base sampling can also be combined with the framework of *generalized additive models*.
-This functionality has been implemented in our package \pkg{casebase}.
-Similarly, case-base sampling can also be combined with quasi-likelihood estimation to fit survival models that can account for the presence of over-dispersion.
-All of these examples illustrate how the case-base sampling framework in general, and the package \pkg{casebase} in particular, allows the user to fit a broad and flexible family of survival functions.
-
-As presented in Hanley & Miettinen [-@hanley2009fitting], case-base sampling is comprised of three steps: 1) sampling a case series and a base series from the study; 2) fit the log-hazard as a linear function of predictors (including time); and 3) use the fitted hazard to estimate the cumulative incidence curve.
-Accordingly, our package provides functions for each step.
-Moreover, the simple interface of the \code{fittingSmoothHazard} function resembles the \code{glm} interface.
-This interface should look familiar to new users.
-Our modular approach also provides a convenient way to extend our package for new sampling or fitting strategies.
-
-In the case studies above, we compared the performance of case-base sampling with that of Cox regression and Fine-Gray models.
-In terms of function interface, \pkg{casebase} uses a formula interface that is closer to that of \code{glm}, in that the event variable is the only variable appearing on the left-hand side of the formula.
-By contrast, both \code{survival::coxph} and \code{timereg::comp.risk} use arrays that captures both the event type and time.
-Both approaches to modeling yield user-friendly code.
-However, in terms of output, both approaches differ significantly.
-Case-base sampling produces smooth hazards and smooth cumulative incidence curves, whereas Cox regression and Fine-Gray models produce step-wise cumulative incidence functions and never explicitly model the hazard function.
-Qualitatively, we showed that by using splines in the linear predictor, all three models yielded similar curves.
-However, the smooth nature of the output of \pkg{casebase} provides a more intuitive interpretation for consumers of these predictions.
-In Table \ref{tab:compCBvsCox}, we provide a side-by-side comparison between the Cox model and case-base sampling.
-
-\begin{table}
-\caption{\label{tab:compCBvsCox}Comparison between the Cox model and case-base sampling}
-\centering
-\begin{tabular}[t]{llp{5cm}}
-\toprule
-Feature & Cox model & Case-base sampling\\
-\midrule
-Model type & Semi-parametric & Fully parametric\\
-Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\
-Cumulative incidence & Step function & Smooth-in-time curve\\
-Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\
-Model testing &  & Use GLM framework \newline (e.g.\ LRT, AIC, BIC)\\
-\addlinespace
-Competing risks & Difficult & Cause-specific CIFs\\
-% Prediction & Kaplan-Meier-based & ROC, AUC, risk reclassification probabilities\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-Our choice of modeling the log-hazard as a linear function of covariates allows us to develop a simple computational scheme for estimation.
-However, as a downside, it does not allow us to model location and scale parameters separately like the package \pkg{flexsurv}.
-For example, if we look at the Weibull distribution as parametrised in \code{stats::pweibull}, the log-hazard function is given by
-$$ \log h(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,$$
-where $\alpha,\beta$ are shape and scale parameters, respectively.
-Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} allows the user to model the scale parameter as a function of covariates.
-Of course, this added flexibility comes at the cost of interpretability: by modeling the log-hazard directly, the parameter estimates from \pkg{casebase} can be interpreted as estimates of log-hazard ratios.
-To improve the flexibility of \pkg{casebase} at capturing the scale of a parametric family, we could replace the logistic regression with its quasi-likelihood counterpart and therefore model over- and under-dispersion with respect to the logistic likelihood.
-We defer the study of the properties and performance of such a model to a future article.
-
-Future work will look at some of the methodological extensions of case-base sampling.
-First, to assess the quality of the model fit, we want to study the properties of the residuals (e.g.
-Cox-Snell, martingale).
-More work needs to be done to understand these residuals in the context of the partial likelihood underlying case-base sampling.
-The resulting diagnostic tools would then be integrated in this package.
-Also, we are interested in extending case-base sampling to account for interval censoring.
-This type of censoring is very common in longitudinal studies, and many packages (e.g.
-\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2}) provide functions to account for it.
-Again, we hope to include any resulting methodology as part of this package.
-
-In future versions of the package, we also want to increase the complement of diagnostic and inferential tools that are currently available.
-For example, we would like to include the ability to compute confidence intervals for the cumulative incidence curve.
-The delta method or parametric bootstrap are two different strategies we can use to construct approximate confidence intervals.
-Furthermore, we would like to include functions to compute calibration and discrimination statistics (e.g.
-Brier score, AUC) for our models.
-Saarela and Arjas [-@saarela2015non] also describe how to obtain a posterior distribution for the AUC from their model.
-Their approach could also be included in \pkg{casebase}.
-
-In conclusion, we presented the \proglang{R} package \pkg{casebase} which implements case-base sampling for fitting parametric survival models and for estimating smooth cumulative incidence functions using the framework of generalized linear models.
-We strongly believe that its flexibility and its foundation on the familiar logistic regression model will make it appealing to new and established practioners.
-
-```{r, echo=FALSE, eval = FALSE}
+In this article, we presented the \proglang{R} package \pkg{casebase} that provides functions to fit smooth parametric hazards and estimate cumulative incidence functions using case-base sampling.
+We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through three case studies.
+
+<!-- In the following table we provide a comparison between the Cox model and case-base sampling: -->
+
+```{r, echo=FALSE, eval = FALSE, results='asis'}
 knitr::kable(data.frame(`feature` = c("model type", "time", "cumulative incidence", "non-proportional hazards","model testing","competing risks", "prediction"),
@@
 # which R packages and versions?
-# devtools::session_info()
-sessionInfo()
+devtools::session_info()
 ```

### 2020-02-06 analysis/paper/paper.tex
Switched to non interaction model
note that I only ran CS3 so the pdf is not complete. fixes include all red text

Jesse Still needs to clean up the text and deliberate his write up in the case study. Code will also be cleaned to be more readable.

--- 2020-02-06 analysis/paper/paper.tex
@@
 
-\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx} \usepackage{tabularx} \usepackage{float} \usepackage{booktabs} \usepackage{makecell} \usepackage{tabu} \DeclareUnicodeCharacter{2500}{-}
+\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx} \usepackage{tabularx} \usepackage{float} \usepackage{booktabs} \usepackage{makecell} \DeclareUnicodeCharacter{2500}{-}
 
@@
 \label{tab:surv-pkgs}
+\end{table}
+
+\begin{table}
+  \begin{tabular}{p{1.8in}lll}
+\hline
+    &  Parameters &  Density \proglang{R} function & \code{dist}\\
+    & {\footnotesize{(location in \code{\color{red}{red}})}} & & \\
+\hline
+    Exponential & \code{\color{red}{rate}}             & \code{dexp}   & \code{"exp"} \\
+    Weibull (accelerated failure time)     & \code{shape, {\color{red}{scale}}}     & \code{dweibull} & \code{"weibull"} \\
+    Weibull (proportional hazards)     & \code{shape, {\color{red}{scale}}}     & \code{dweibullPH} & \code{"weibullPH"} \\
+    Gamma       & \code{shape, \color{red}{rate}}      & \code{dgamma} & \code{"gamma"}\\
+    Log-normal  & \code{{\color{red}{meanlog}}, sdlog}   & \code{dlnorm} & \code{"lnorm"}\\
+    Gompertz    & \code{shape, {\color{red}{rate}}}      & \code{dgompertz} & \code{"gompertz"} \\
+    Log-logistic & \code{shape, {\color{red}{scale}}}   & \code{dllogis} & \code{"llogis"}\\
+    Generalized gamma (Prentice 1975)   & \code{{\color{red}{mu}}, sigma, Q} & \code{dgengamma} & \code{"gengamma"} \\
+    Generalized gamma (Stacy 1962)& \code{shape, {\color{red}{scale}}, k} & \code{dgengamma.orig} & \code{"gengamma.orig"} \\
+    Generalized F     (stable)    & \code{{\color{red}{mu}}, sigma, Q, P} & \code{dgenf} & \code{"genf"} \\
+    Generalized F     (original)  & \code{{\color{red}{mu}}, sigma, s1, s2} & \code{dgenf.orig} & \code{"genf.orig"} \\
+\hline
+  \end{tabular}
+  \caption{Built-in parametric survival distributions in \pkg{flexsurv}.}
+  \label{tab:dists}
 \end{table}
@@
 
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
+R> plot(pt_object)
+\end{CodeInput}
+\end{CodeChunk}
+
 We can also create exposure stratified plots by specifying the
 \code{exposure} argument in the \code{popTime} function:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> pt_object_strat <- casebase::popTime(ERSPC, 
+R>                                      event = "DeadOfPrCa", 
+R>                                      exposure = "ScrArm")
+R> plot(pt_object_strat)
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 \begin{CodeChunk}
-\begin{figure}
 
-{\centering \includegraphics{../figures/erspc-cox-cif-1}
-
-}
-
-\caption[Cumulative incidence functions for control and screening groups in the ERSPC data]{Cumulative incidence functions for control and screening groups in the ERSPC data.
-In each of the panels, we plot the CIF from the Cox model (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression.
-(A) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution.
-(B) Linear function of time which is equivalent to a Gompertz distribution.
-(C) The natural logarithm of time which is equivalent to a Weibull distribution.
-(D) Cubic b-spline expansion of time.}\label{fig:erspc-cox-cif}
-\end{figure}
+\begin{CodeInput}
+R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
+R>                        ignore = 99)
+R> 
+R> plot(survfit(cox_model, newdata = new_data),
+R>      xlab = "Years since Randomization", 
+R>      ylab = "Cumulative Incidence (%)", 
+R>      fun = "event",
+R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R>                     exp(coef(cox_model)), 
+R>                     exp(confint(cox_model))[1], 
+R>                     exp(confint(cox_model))[2]),
+R>      ylim = c(0, 2), yscale = 100)
+R> legend("topleft", 
+R>        legend = c("Control group", "Screening group"), 
+R>        col = c("red","blue"),
+R>        lty = c(1, 1), 
+R>        bg = "gray90")
+\end{CodeInput}
 \end{CodeChunk}
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics{../figures/unnamed-chunk-1-1} \end{center}
-
-\end{CodeChunk}
-
-\hypertarget{equivalence-between-casebase-and-aft}{%
-\subsection{Equivalence between casebase and
-AFT}\label{equivalence-between-casebase-and-aft}}
 
@@
 
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
+R>                                                   data = ERSPC, 
+R>                                                   ratio = 100)
+R> 
+R> summary(casebase_exponential)
+R> exp(coef(casebase_exponential)[2])
+R> exp(confint(casebase_exponential)[2,])
+\end{CodeInput}
+\end{CodeChunk}
+
 We can then use the \code{absoluteRisk} function to get an estimate of
@@
 
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
+R>                                  data = ERSPC, 
+R>                                  ratio = 100)
+R> 
+R> summary(casebase_time)
+R> exp(coef(casebase_time))
+R> exp(confint(casebase_time))
+\end{CodeInput}
+\end{CodeChunk}
+
 Again, the estimate of the hazard ratio is similar to that obtained from
 Cox regression. We then look at the estimate of the CIF:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
+R>                                           time = seq(0,15,0.1), 
+R>                                           newdata = new_data)
+R> 
+R> plot(survfit(cox_model, newdata = new_data),
+R>      xlab = "Years since Randomization", 
+R>      ylab = "Cumulative Incidence (%)", 
+R>      fun = "event",
+R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
+R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R>                     exp(coef(cox_model)), 
+R>                     exp(confint(cox_model))[1], 
+R>                     exp(confint(cox_model))[2]),
+R>      ylim = c(0, 2), yscale = 100)
+R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
+R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
+R> 
+R> legend("topleft", 
+R>        legend = c("Control group (Cox)","Control group (Casebase)",
+R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
+R>        col = c("red","red", "blue","blue"),
+R>        lty = c(1, 2, 1, 2), 
+R>        bg = "gray90")
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
+R>                                     data = ERSPC, 
+R>                                     ratio = 100)
+R> 
+R> summary(casebase_splines)
+R> exp(coef(casebase_splines))
+R> exp(confint(casebase_splines))
+\end{CodeInput}
+\end{CodeChunk}
+
 Once again, we can see that the estimate of the hazard ratio is similar
@@
 
-\begin{CodeChunk}
-\begin{table}
-
-\caption{\label{tab:unnamed-chunk-4}Mean }
-\centering
-\begin{tabular}[t]{>{\bfseries}lcc}
-\toprule
-Model & casebase & survreg\\
-\midrule
-Exponential & 0.81 (0.68, 0.96) & 0.81 (0.68, 0.96)\\
-\cmidrule{1-3}
-Gompertz & 0.78 (0.66, 0.93) & 0.80 (0.67, 0.95)\\
-\cmidrule{1-3}
-Weibull & 0.80 (0.67, 0.95) & 0.80 (0.65, 0.96)\\
-\cmidrule{1-3}
-Splines & 0.81 (0.68, 0.96) & --\\
-\bottomrule
-\multicolumn{3}{l}{\textit{Note:}}\\
-\multicolumn{3}{l}{Median (Inter-quartile range) is given for Model Size.}\\
-\end{tabular}
-\end{table}
-
-\end{CodeChunk}
-
 As noted above, the usual asymptotic results hold for likelihood ratio
@@
 
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> anova(casebase_exponential, casebase_splines, test = "LRT")
+\end{CodeInput}
+\end{CodeChunk}
+
+As expected, we see that splines model provides a better fit. Similarly,
+we can compare the AIC for all three case-base models:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> c("Exp." = AIC(casebase_exponential),
+R>   "Weib." = AIC(casebase_time),
+R>   "Splines" = AIC(casebase_splines))
+\end{CodeInput}
+\end{CodeChunk}
+
+Once again, we have evidence that the splines model provides the best
+fit.
+
+With this first case study, we saw how population-time plots can provide
+a useful summary of the incidence for our event of interests. In
+particular, we saw how it can provide evidence against an exponential
+hazard model. We then explored how different parametric models can
+provide different estimates of the CIF, even though they all give
+similar estimates for the hazard ratios.
+
+\hypertarget{case-study-2bone-marrow-transplant}{%
+\section{Case study 2--Bone-marrow
+transplant}\label{case-study-2bone-marrow-transplant}}
+
+In the next case study, we show how case-base sampling can also be used
+in the context of a competing risk analysis. For illustrative purposes,
+we will use the same data that was used in Scrucca \emph{et al}
+\citeyearpar{scrucca2010regression}. The data was downloaded from the
+first author's website, and it is also available as part of the
+\pkg{casebase} package.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> 
+R> data(bmtcrr)
+\end{CodeInput}
+\end{CodeChunk}
+
+The data contains information on 177 patients who received a stem-cell
+transplant for acute leukemia. The event of interest is relapse, but
+other competing causes (e.g.~transplant-related death) were also
+recorded. Several covariates were captured at baseline: sex, disease
+type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL
+and AML, respectively), disease phase at transplant (Relapse, CR1, CR2,
+CR3), source of stem cells (bone marrow and peripheral blood, coded as
+BM+PB, or only peripheral blood, coded as PB), and age. A summary of
+these baseline characteristics appear in Table \ref{tab:table1bmtcrr}.
+We note that the statistical summaries were generated differently for
+different variable types: for continuous variables, we gave the range,
+followed by the mean and standard deviation; for categorical variables,
+we gave the counts for each category.
+
+First, we can look at a population-time plot to visualize the incidence
+density of both relapse and the competing events. In Figure
+\ref{fig:compPop}, failure times are highlighted on the plot using red
+points for the event of interest (panel A) and blue dots for competing
+events (panel B). In both panels, we see evidence of a non-constant
+hazard function: the density of points is larger at the beginning of
+follow-up than at the end.
+
+Our main objective is to compute the absolute risk of relapse for a
+given set of covariates. First, we fit a smooth hazard to the data using
+a linear term for time:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> model_cb <- fitSmoothHazard(
+R>     Status ~ ftime + Sex + D + Phase + Source + Age, 
+R>     data = bmtcrr, 
+R>     ratio = 100, 
+R>     time = "ftime")
+\end{CodeInput}
+\end{CodeChunk}
+
+We will compare our hazard ratio estimates to that obtained from a Cox
+regression. To do so, we need to treat the competing event as censoring.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> library(survival)
+R> # Treat competing event as censoring
+R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
+R>                    data = bmtcrr)
+\end{CodeInput}
+\end{CodeChunk}
+
+From the fit object, we can extract both the hazard ratios and their
+corresponding confidence intervals. These quantities appear in Table
+\ref{tab:bmtcrr-cis}. As we can see, the only significant hazard ratio
+identified by case-base sampling is the one associated with the phase of
+the disease at transplant. More precisely, being in relapse at
+transplant is associated with a hazard ratio of 3.89 when compared to
+CR1.
+
+Given the estimate of the hazard function obtained using case-base
+sampling, we can compute the absolute risk curve for a fixed covariate
+profile. We perform this computation for a 35 year old woman who
+received a stem-cell transplant from peripheral blood at relapse. We
+compared the absolute risk curve for such a woman with acute
+lymphoblastic leukemia with that for a similar woman with acute
+myeloblastic leukemia. We will estimate the curve from 0 to 60 months.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Pick 100 equidistant points between 0 and 60 months
+R> time_points <- seq(0, 60, length.out = 50)
+R> 
+R> # Data.frame containing risk profile
+R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
+R>                                      levels = levels(bmtcrr[,"Sex"])),
+R>                       "D" = c("ALL", "AML"), # Both diseases
+R>                       "Phase" = factor(c("Relapse", "Relapse"), 
+R>                                        levels = levels(bmtcrr[,"Phase"])),
+R>                       "Age" = c(35, 35),
+R>                       "Source" = factor(c("PB", "PB"), 
+R>                                         levels = levels(bmtcrr[,"Source"])))
+R> 
+R> # Estimate absolute risk curve
+R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
+R>                         method = "numerical", newdata = newdata)
+\end{CodeInput}
+\end{CodeChunk}
+
+We will compare our estimates to that obtained from a corresponding
+Fine-Gray model \citep{fine1999proportional}. The Fine-Gray model is a
+semiparametric model for the cause-specific \emph{subdistribution},
+i.e.~the function \(f_k(t)\) such that
+\[CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),\] where
+\(CI_k(t)\) is the cause-specific cumulative indicence. The Fine-Gray
+model allows to directly assess the effect of a covariate on the
+subdistribution, as opposed to the hazard. For the computation, we will
+use the \pkg{timereg} package \citep{timereg}.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> library(timereg)
+R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) + 
+R>                           const(Phase) + const(Source) + const(Age), 
+R>                       data = bmtcrr, cause = 1, model = "fg")
+R> 
+R> # Estimate absolute risk curve
+R> risk_fg <- predict(model_fg, newdata, times = time_points)
+\end{CodeInput}
+\end{CodeChunk}
+
+Figure \ref{fig:compAbsrisk} shows the absolute risk curves for both
+case-base sampling and the Fine-Gray model. As we can see, the two
+approaches agree quite well for acute myeloblastic leukemia; however,
+there seems to be a difference of about 5\% between the two curves for
+acute lymphoblastic leukemia. This difference does not appear to be
+significant: the curve from case-base sampling is contained within a
+95\% confidence band around the Fine-Gray absolute risk curve (figure
+not shown).
+
+\hypertarget{case-study-3support-data}{%
+\section{Case study 3--SUPPORT Data}\label{case-study-3support-data}}
+
+In the first two case studies, we described the basic functionalities of
+the \pkg{casebase} package: creating population-time plots, fitting
+parametric models for hazard functions, and estimating the corresponding
+cumulative incidence curves. In the next two case studies, we explore
+more advanced features.
+
+For the third case study, we will introduce regularization into the
+estimation of the hazard function. This can be performed using a
+combination of case-base sampling and regularized logistic regression.
+To demonstrate this capability, we examine the Study to Understand
+Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset
+\citep{knaus1995support}. The SUPPORT dataset tracks death for
+individuals who are considered seriously ill within a hospital.
+Imputation was conducted using the \pkg{mice} package in \proglang{R}
+with default settings \citep{mice}. Before imputation, there were a
+total of 1000 individuals and 35 variables. charges, Totcst, totmcst and
+sfdm2 were removed as they are response variables we are not interested
+in for the sake of this analysis. After imputation, we have 690
+individuals. We lose individuals due to coliniarity between variables,
+preventing imputation. If a pair of covariates are deemed colinear to a
+certain degree, they are excluded from any imputations. Here, only edu
+income avtisst* race ph* glucose* adlp* were imputed. Of these 690
+individuals, 66.09\% experienced the event of interest. A description of
+each variable can be found in Table \ref{tab:support1} and a breakdown
+of each categorical variable in Table \ref{tab:support2}. The original
+data was obtained from the Department of Biostatistics at Vanderbilt
+University. The imputed dataset is available as part of the
+\pkg{casebase} package.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{ccccc}
+\hline
+\textbf{Name} & \textbf{Labels}                          & \textbf{Levels} & \textbf{Storage} & \textbf{NAs} \\
+\hline
+age           & Age                                      &                 & double           & 0            \\
+death         & Death at any time up to NDI date:31DEC94 &                 & double           & 0            \\
+sex           &                                          & 2               & integer          & 0            \\
+hospdead      & Death in Hospital                        &                 & double           & 0            \\
+slos          & Days from Study Entry to Discharge       &                 & double           & 0            \\
+d.time        & Days of Follow-Up                        &                 & double           & 0            \\
+dzgroup       &                                          & 8               & integer          & 0            \\
+dzclass       &                                          & 4               & integer          & 0            \\
+num.co        & number of comorbidities                  &                 & double           & 0            \\
+edu           & Years of Education                       &                 & double           & 202          \\
+income        &                                          & 4               & integer          & 349          \\
+scoma         & SUPPORT Coma Score based on Glasgow D3   &                 & double           & 0            \\
+charges       & Hospital Charges                         &                 & double           & 25           \\
+totcst        & Total RCC cost                           &                 & double           & 105          \\
+totmcst       & Total micro-cost                         &                 & double           & 372          \\
+avtisst       & Average TISS, Days 3-25                  &                 & double           & 6            \\
+race          &                                          & 5               & integer          & 5            \\
+meanbp        & Mean Arterial Blood Pressure Day 3       &                 & double           & 0            \\
+wblc          & White Blood Cell Count Day 3             &                 & double           & 24           \\
+hrt           & Heart Rate Day 3                         &                 & double           & 0            \\
+resp          & Respiration Rate Day 3                   &                 & double           & 0            \\
+temp          & Temperature (Celsius) Day 3              &                 & double           & 0            \\
+pafi          & PaO2/(.01*FiO2) Day 3                    &                 & double           & 253          \\
+alb           & Serum Albumin Day 3                      &                 & double           & 378          \\
+bili          & Bilirubin Day 3                          &                 & double           & 297          \\
+crea          & Serum creatinine Day 3                   &                 & double           & 3            \\
+sod           & Serum sodium Day 3                       &                 & double           & 0            \\
+ph            & Serum pH (arterial) Day 3                &                 & double           & 250          \\
+glucose       & Glucose Day 3                            &                 & double           & 470          \\
+bun           & BUN Day 3                                &                 & double           & 455          \\
+urine         & Urine Output Day 3                       &                 & double           & 517          \\
+adlp          & ADL Patient Day 3                        &                 & double           & 634          \\
+adls          & ADL Surrogate Day 3                      &                 & double           & 310          \\
+sfdm2         &                                          & 5               & integer          & 159          \\
+adlsc         & Imputed ADL Calibrated to Surrogate      &                 & double           & 0           
+\end{tabular}
+\caption{A description of each variable in the SUPPORT dataset.}
+\label{tab:support1}
+\end{table}
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{cc}
+\hline
+\textbf{Variable} & \textbf{Levels}                      \\
+\hline
+sex      & female                                        \\
+         & male                                          \\
+dzgroup  & ARF/MOSF w/Sepsis                             \\
+         & COPD                                          \\
+         & CHF                                           \\
+         & Cirrhosis                                     \\
+         & Coma                                          \\
+         & Colon Cancer                                  \\
+         & Lung Cancer                                   \\
+         & MOSF w/Malig                                  \\
+dzclass  & ARF/MOSF                                      \\
+         & COPD/CHF/Cirrhosis                            \\
+         & Coma                                          \\
+         & Cancer                                        \\
+income   & under \$11k                                   \\
+         & $11-$25k                                      \\
+         & $25-$50k                                      \\
+         & \textgreater{}\$50k                           \\
+race     & white                                         \\
+         & black                                         \\
+         & asian                                         \\
+         & other                                         \\
+         & hispanic                                      \\
+sfdm2    & no(M2 and SIP pres)                           \\
+         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
+         & SIP\textgreater{}=30                          \\
+         & Coma or Intub                                 \\
+         & \textless{}2 mo. follow-up                   
+\end{tabular}
+\caption{A description of each level within each categorical variable in the dataset.}
+\label{tab:support2}
+\end{table}
+
+As before, the first step is to visualize the incidence density. The
+population-time plot for the SUPPORT data appears in Figure
+\ref{fig:support-poptime}.
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/support-poptime-1}
+
+}
+
+\caption{\label{fig:support-poptime} Population-time plot for the SUPPORT data. Red points represent the case series. The gray space demonstrates the base series.}\label{fig:support-poptime}
+\end{figure}
+\end{CodeChunk}
+
+Before we proceed with the analysis, we first remove hospital death as a
+covariate, as this is directly informative of death. All covariates will
+be used to estimate the hazard function. The hazard function will be
+used to estimate the cumulative incidence of each individual in the
+study. Then, the average cumulative incidience curve for all
+observations will be used for our comparison.
+
+Our main objective is to compute the absolute risk of death for a given
+set of covariates, using regularization when fitting our model. First,
+we fit a smooth hazard to the data using a weibull distribution. As
+casebase makes use of the glmnet package, we interact with the
+fitsmoothhazard.fit function using a matrix interface, where y contains
+the time and event variables and x contains all other variables we would
+like to include in the model. For the SUPPORT dataset, all factor
+variables were converted into indicator variables. Here, we used lasso
+penalization by setting alpha to 1.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> 
+R> cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
+R+                                        formula_time = ~log(d.time),alpha=1,ratio=ratio)
+\end{CodeInput}
+\end{CodeChunk}
+
+Then, we take our model and use the \texttt{absoluteRisk} function to
+integrate and retrieve our cumulative incidence curves.
+
+We will compare this curve to a regularized version of cox regression,
+and a kaplan-meier survival curve. The cox regression absolute risk
+curve with lasso penalization required two extra steps to retrieve.
+Coxnet from the glmnet package has tools to fit a regularized cox model,
+but requires a survival object skeleton of the selected variables, so
+that the survival package tools can be used to retrive the absolute
+risk. We handle this by fitting a regularized cox model with the glmnet
+package, and a cox model from the survival package with the selected
+variables from the regularized fit. This second model serves as a
+skeleton that permits the use of survival package functions. The
+coefficients from the regularized model will replace the coefficients in
+our skeleton. the resulting model will have the correct coefficients
+when integrating, but will have an incorrect standard error. For the
+purposes of this case study, we are only interested in the absolute risk
+curve itself and not the standard error.
+
+We used the survival package to calculate cox regression's cumulative
+incidience without penalization and the kaplan-meier curve for
+incidience.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> 
+R> coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
+R> abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)
+R> 
+R> #creating a surv object to be used to fit an unadjusted absolute risk curve.
+R> # (Kaplan-Meier risk curve)
+R> km <- survival::Surv(time = completeData$d.time, event = completeData$death)
+R> abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
+\end{CodeInput}
+\end{CodeChunk}
+
+Now that our four cumulative incidence estimations have been calculated,
+we compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase
+decrease the absolute risk by the end of the study, in comparison to
+kaplan-meier.
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/abSupportComparison-1}
+
+}
+
+\caption{\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue.}\label{fig:abSupportComparison}
+\end{figure}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> 
+R> #get covariates excluding d.time
+R> estimatescb=setDT(as.data.frame(coef(cb.Model)[-c(1,2),1]), keep.rownames = TRUE)[]
+R> estimatescb$Model="casebase"
+R> colnames(estimatescb)<-c("Coefs","Estimate","Model")
+R> estimatescox=setDT(as.data.frame(coef(coxglmFit)[-c(1),1]), keep.rownames = TRUE)[]
+R> estimatescox$Model="Cox"
+R> colnames(estimatescox)<-c("Coefs","Estimate","Model")
+R> estimates=rbind(estimatescb, estimatescox)
+R> estimates$Estimate[estimates$Estimate==0]=NA# make it so unchosen covariates appear empty
+R> library(ggstance)# for vertical dodge
+R>   ggplot(estimates, aes(x = Estimate, y = Coefs, color = Model)) +
+R>         geom_segment(aes(x = 0, y = Coefs, xend = Estimate, yend = Coefs),size=0.5 ) +
+R>         geom_point(position=ggstance::position_dodgev(height=0.2))
+R> 
+\end{CodeInput}
+\end{CodeChunk}
+
+\hypertarget{case-study-4stanford-heart-transplant-data}{%
+\section{Case study 4--Stanford Heart Transplant
+Data}\label{case-study-4stanford-heart-transplant-data}}
+
+Although the previous case studies provide a broad overview of the
+capabilities of \pkg{casebase}, they all have two properties in common:
+
+\begin{itemize}
+  \item All covariates are fixed, i.e. they do not change over time;
+  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
+\end{itemize}
+
+Case-base sampling has been used in the literature to study vaccination
+safety, where the exposure period was defined as the week following
+vaccination \citep{saarela2015case}. Hence, the main covariate of
+interest, i.e.~exposure to the vaccine, was changing over time. In this
+context, case-base sampling offers an efficient alternative to nested
+case-control designs or self-matching.
+
+The purpose of the next case study is to show how \pkg{casebase} can
+also be used for survival analysis problems that do not share the
+properties above. To this end, we will use the Stanford Heart Transplant
+Data \citep[\citet{crowley1977covariance}]{clark1971cardiac}; we will
+use the version available in the \pkg{survival} package.
+
+Recall the setting of this study: patients were admitted to the Stanford
+program after meeting with their physician and determining that they
+were unlikely to respond to other forms of treatment. After enrollment,
+the program searches for a suitable donor for the patient, which can
+take anywhere between a few days to almost a year. We are interested in
+the effect of a heart transplant on survival; therefore, the patient is
+considered exposed only after the transplant has occurred.
+
+As before, we can look at the population-time plot for a graphical
+summary of the event incidence. As we can see, most events occur early
+during the follow-up period, and therefore we do not expect the hazard
+to be constant.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> library(survival)
+R> library(casebase)
+R> 
+R> stanford_popTime <- popTime(jasa, time = "futime", 
+R>                             event = "fustat")
+R> plot(stanford_popTime)
+\end{CodeInput}
+\end{CodeChunk}
+
+Since the exposure is time-dependent, we need to manually define the
+exposure variable \emph{after} case-base sampling and \emph{before}
+fitting the hazard function. For this reason, we will use the
+\texttt{sampleCaseBase} function directly.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> library(tidyverse)
+R> library(lubridate)
+R> 
+R> cb_data <- sampleCaseBase(jasa, time = "futime", 
+R>                           event = "fustat", ratio = 10)
+\end{CodeInput}
+\end{CodeChunk}
+
+Next, we will compute the number of days from acceptance into the
+program to transplant, and we use this variable to determine whether
+each population-moment is exposed or not.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Define exposure variable
+R> cb_data <- mutate(cb_data,
+R>                   txtime = time_length(accept.dt %--% tx.date, 
+R>                                        unit = "days"),
+R>                   exposure = case_when(
+R>                     is.na(txtime) ~ 0L,
+R>                     txtime > futime ~ 0L,
+R>                     txtime <= futime ~ 1L
+R>                   ))
+\end{CodeInput}
+\end{CodeChunk}
+
+Finally, we can fit the hazard using various linear predictors.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> library(splines)
+R> # Fit several models
+R> fit1 <- fitSmoothHazard(fustat ~ exposure,
+R>                         data = cb_data, time = "futime")
+R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
+R>                         data = cb_data, time = "futime")
+R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
+R>                         data = cb_data, time = "futime")
+R> fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
+R>                         data = cb_data, time = "futime")
+\end{CodeInput}
+\end{CodeChunk}
+
+Note that the fourth model (i.e. \texttt{fit4}) includes an interaction
+term between exposure and follow-up time. In other words, this model no
+longer exhibit proportional hazards. The evidence of non-proportionality
+of hazards in the Stanford Heart Transplant data has been widely
+discussed \citep{arjas1988graphical}.
+
+We can then compare the goodness of fit of these four models using the
+Akaike Information Criterion (AIC).
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Compute AIC
+R> c("Model1" = AIC(fit1),
+R>   "Model2" = AIC(fit2),
+R>   "Model3" = AIC(fit3),
+R>   "Model4" = AIC(fit4))
+\end{CodeInput}
+\end{CodeChunk}
+
+As we can, the best fit is the fourth model. By visualizing the hazard
+functions for both exposed and unexposed individuals, we can more
+clearly see how the hazards are no longer proportional.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Compute hazards---
+R> # First, create a list of time points for both exposure status
+R> hazard_data <- expand.grid(exposure = c(0, 1),
+R>                            futime = seq(0, 1000,
+R>                                         length.out = 100))
+R> # Set the offset to zero
+R> hazard_data$offset <- 0 
+R> # Use predict to get the fitted values, and exponentiate to 
+R> # transform to the right scale
+R> hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
+R>                                  type = "link"))
+R> # Add labels for plots
+R> hazard_data$Status = factor(hazard_data$exposure,
+R>                             labels = c("NoTrans", "Trans"))
+R> 
+R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
+R>     geom_line() +
+R>     theme_minimal() +
+R>     theme(legend.position = 'top') +
+R>     ylab('Hazard') + xlab('Follow-up time')
+\end{CodeInput}
+\end{CodeChunk}
+
+The non-proportionality seems to be more pronounced at the beginning of
+follow-up than the end. Finally, we can turn these estimate of the
+hazard function into estimates of the cumulative incidence functions.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Compute absolute risk curves
+R> newdata <- data.frame(exposure = c(0, 1))
+R> absrisk <- absoluteRisk(fit4, newdata = newdata, 
+R>                         time = seq(0, 1000, length.out = 100))
+R> 
+R> colnames(absrisk) <- c("Time", "NoTrans", "Trans")
+R> 
+R> # Rearrange the data
+R> absrisk <- gather(as.data.frame(absrisk),
+R>                   "Status", "Risk", -Time)
+R>  
+R> ggplot(absrisk, aes(Time, Risk, colour = Status)) +
+R>   geom_line() +
+R>   theme_minimal() +
+R>   theme(legend.position = 'top') +
+R>   expand_limits(y = 0.5) +
+R>   xlab('Follow-up time') + ylab('Cum. Incidence')
+\end{CodeInput}
+\end{CodeChunk}
+
+Note that we can easily adapt the code above to the situation where a
+patient receives a heart transplant at a point in time of interest, for
+example after 30 days.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Compute hazards---
+R> # First, create a list of time points for both exposure status
+R> one_yr_haz <- data.frame(futime = seq(0, 365,
+R>                                       length.out = 100))
+R> one_yr_haz <- mutate(one_yr_haz,
+R>                      offset = 0,
+R>                      exposure = if_else(futime < 30, 0, 1))
+R> one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
+R>                                 type = "link"))
+R> ggplot(one_yr_haz, aes(futime, hazard)) +
+R>   geom_line() +
+R>   theme_minimal() +
+R>   theme(legend.position = 'top') +
+R>   ylab('Hazard') + xlab('Follow-up time') +
+R>   geom_vline(xintercept = 30, linetype = 'dashed')
+\end{CodeInput}
+\end{CodeChunk}
+
+We can then compare the 1-year mortality risk without transplant and
+with transplant at 30 days.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> absoluteRisk(fit4, newdata = data.frame(exposure = 0),
+R>              time = 365)
+R> 
+R> # Use the trapezoidal rule to estimate the cumulative hazard
+R> min_hazard <- min(one_yr_haz$hazard)
+R> max_hazard <- max(one_yr_haz$hazard)
+R> increment <- 365/99
+R> cumhaz_est <- 0.5*increment*(sum(2*one_yr_haz$hazard) - min_hazard - max_hazard)
+R> 1 - exp(-cumhaz_est)
+\end{CodeInput}
+\end{CodeChunk}
+
+As we can see, the risk estimate at 1-year is about 30\% lower if the
+patient receives a heart transplant at 30 days.
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+In this article, we presented the \proglang{R} package \pkg{casebase}
+that provides functions to fit smooth parametric hazards and estimate
+cumulative incidence functions using case-base sampling. We outlined the
+theoretical underpinnings of the approach, we provided details about our
+implementation, and we illustrated the merits of the approach and the
+package through three case studies.
+
+\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval
+censoring, whereas \pkg{casebase} currently cannot. Interval censoring
+is a simple feature to implement and is a potential improvement to add
+to \pkg{casebase}.
+
+\hypertarget{environment-details}{%
+\section{Environment Details}\label{environment-details}}
+
+This report was generated on 2020-02-06 17:40:56 using the following
+computational environment and dependencies:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # which R packages and versions?
+R> devtools::session_info()
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> - Session info ---------------------------------------------------------------
+#>  setting  value                       
+#>  version  R version 3.6.0 (2019-04-26)
+#>  os       macOS High Sierra 10.13.6   
+#>  system   x86_64, darwin15.6.0        
+#>  ui       X11                         
+#>  language (EN)                        
+#>  collate  en_CA.UTF-8                 
+#>  ctype    en_CA.UTF-8                 
+#>  tz       America/Montreal            
+#>  date     2020-02-06                  
+#> 
+#> - Packages -------------------------------------------------------------------
+#>  package     * version    date       lib
+#>  assertthat    0.2.1      2019-03-21 [1]
+#>  backports     1.1.5      2019-10-02 [1]
+#>  broom         0.5.2      2019-04-07 [1]
+#>  callr         3.4.1      2020-01-24 [1]
+#>  casebase    * 0.2.1.9001 2020-01-30 [1]
+#>  cellranger    1.1.0      2016-07-27 [1]
+#>  cli           2.0.1      2020-01-08 [1]
+#>  codetools     0.2-16     2018-12-24 [1]
+#>  colorspace    1.4-1      2019-03-18 [1]
+#>  crayon        1.3.4      2017-09-16 [1]
+#>  data.table  * 1.12.8     2019-12-09 [1]
+#>  DBI           1.0.0      2018-05-02 [1]
+#>  dbplyr        1.4.2      2019-06-17 [1]
+#>  desc          1.2.0      2018-05-01 [1]
+#>  devtools      2.2.1      2019-09-24 [1]
+#>  digest        0.6.23     2019-11-23 [1]
+#>  dplyr       * 0.8.3      2019-07-04 [1]
+#>  ellipsis      0.3.0      2019-09-20 [1]
+#>  evaluate      0.14       2019-05-28 [1]
+#>  fansi         0.4.1      2020-01-08 [1]
+#>  farver        2.0.3      2020-01-16 [1]
+#>  forcats     * 0.4.0      2019-02-17 [1]
+#>  foreach       1.4.7      2019-07-27 [1]
+#>  fs            1.3.1      2019-05-06 [1]
+#>  generics      0.0.2      2018-11-29 [1]
+#>  ggplot2     * 3.2.1      2019-08-10 [1]
+#>  glmnet      * 3.0-2      2019-12-11 [1]
+#>  glue          1.3.1      2019-03-12 [1]
+#>  gtable        0.3.0      2019-03-25 [1]
+#>  haven         2.2.0      2019-11-08 [1]
+#>  hms           0.5.2      2019-10-30 [1]
+#>  htmltools     0.4.0      2019-10-04 [1]
+#>  httr          1.4.1      2019-08-05 [1]
+#>  iterators     1.0.12     2019-07-26 [1]
+#>  jsonlite      1.6        2018-12-07 [1]
+#>  kableExtra  * 1.1.0      2019-03-16 [1]
+#>  knitr         1.27       2020-01-16 [1]
+#>  labeling      0.3        2014-08-23 [1]
+#>  lattice       0.20-38    2018-11-04 [1]
+#>  lazyeval      0.2.2      2019-03-15 [1]
+#>  lifecycle     0.1.0      2019-08-01 [1]
+#>  lubridate     1.7.4      2018-04-11 [1]
+#>  magrittr    * 1.5        2014-11-22 [1]
+#>  Matrix      * 1.2-17     2019-03-22 [1]
+#>  memoise       1.1.0      2017-04-21 [1]
+#>  mgcv          1.8-31     2019-11-09 [1]
+#>  modelr        0.1.5      2019-08-08 [1]
+#>  munsell       0.5.0      2018-06-12 [1]
+#>  nlme          3.1-142    2019-11-07 [1]
+#>  pillar        1.4.3      2019-12-20 [1]
+#>  pkgbuild      1.0.6      2019-10-09 [1]
+#>  pkgconfig     2.0.3      2019-09-22 [1]
+#>  pkgload       1.0.2      2018-10-29 [1]
+#>  prettyunits   1.1.1      2020-01-24 [1]
+#>  processx      3.4.1      2019-07-18 [1]
+#>  ps            1.3.0      2018-12-21 [1]
+#>  purrr       * 0.3.3      2019-10-18 [1]
+#>  R6            2.4.1      2019-11-12 [1]
+#>  Rcpp          1.0.3      2019-11-08 [1]
+#>  readr       * 1.3.1      2018-12-21 [1]
+#>  readxl        1.3.1      2019-03-13 [1]
+#>  remotes       2.1.0      2019-06-24 [1]
+#>  reprex        0.3.0      2019-05-16 [1]
+#>  rlang         0.4.4      2020-01-28 [1]
+#>  rmarkdown     2.1        2020-01-20 [1]
+#>  rprojroot     1.3-2      2018-01-03 [1]
+#>  rstudioapi    0.10       2019-03-19 [1]
+#>  rticles       0.12       2019-11-12 [1]
+#>  rvest         0.3.5      2019-11-08 [1]
+#>  scales        1.1.0      2019-11-18 [1]
+#>  sessioninfo   1.1.1      2018-11-05 [1]
+#>  shape         1.4.4      2018-02-07 [1]
+#>  stringi       1.4.5      2020-01-11 [1]
+#>  stringr     * 1.4.0      2019-02-10 [1]
+#>  survival    * 3.1-7      2019-11-09 [1]
+#>  testthat      2.3.1      2019-12-01 [1]
+#>  tibble      * 2.1.3      2019-06-06 [1]
+#>  tidyr       * 1.0.0      2019-09-11 [1]
+#>  tidyselect    0.2.5      2018-10-11 [1]
+#>  tidyverse   * 1.3.0      2019-11-21 [1]
+#>  usethis       1.5.1      2019-07-04 [1]
+#>  vctrs         0.2.2      2020-01-24 [1]
+#>  VGAM          1.1-2      2019-11-21 [1]
+#>  viridisLite   0.3.0      2018-02-01 [1]
+#>  webshot       0.5.2      2019-11-22 [1]
+#>  withr         2.1.2      2018-03-15 [1]
+#>  xfun          0.12       2020-01-13 [1]
+#>  xml2          1.2.2      2019-08-09 [1]
+#>  yaml          2.2.0      2018-07-25 [1]
+#>  source                                  
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  Github (sahirbhatnagar/casebase@24359be)
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#>  CRAN (R 3.6.0)                          
+#> 
+#> [1] /Library/Frameworks/R.framework/Versions/3.6/Resources/library
+\end{CodeOutput}
+\end{CodeChunk}
+
+The current Git commit details are:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # what commit is this file at? 
+R> git2r::repository(here::here())
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> Local:    master /Users/jislam/Desktop/Gits/cbpaper
+#> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
+#> Head:     [a04834a] 2020-01-31: cbPaper draft for covars and interactions
+\end{CodeOutput}
+\end{CodeChunk}
+
 \bibliography{references.bib}

### 2020-02-06 cleanedCS3.Rmd
Switched to non interaction model
note that I only ran CS3 so the pdf is not complete. fixes include all red text

Jesse Still needs to clean up the text and deliberate his write up in the case study. Code will also be cleaned to be more readable.

--- 2020-02-06 analysis/paper/paper.Rmd
@@
 library(glmnet)
+library(kableExtra)
 library(data.table)
-library(kableExtra)
 options(digits = 2, scipen = 999)
@@
 eval_introduction <- FALSE
-eval_theory <- FALSE
+eval_theory <- TRUE
 eval_implementation <- FALSE
@@
 ```
-
-# Introduction
-
-The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence (CI)) of an event, for a particular patient.
-The following points should be noted:
-
-  1. hazard ratios
-  2. If, however, the absolute risks are of interest, they have to be recovered using the semi-parametric Breslow estimator
-  3. Alternative approaches for fitting flexible hazard models for estimating absolute risks, not requiring this two-step approach? Yes! [@hanley2009fitting]
-
-[@hanley2009fitting] propose a fully parametric hazard model that can be fit via logistic regression.
-From the fitted hazard function, cumulative incidence and, thus, risk functions of time, treatment and profile can be easily derived.
-
-# Theoretical details
-
-As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to discretize the study base into an infinite amount of *person moments*.
-These person moments are indexed by both an individual in the study and a time point, and therefore each person moment has a covariate profile, an exposure status and an outcome status attached to it.
-We note that there is only a finite number of person moments associated with the event of interest (what Hanley & Miettinen call the *case series*).
-The case-base sampling refers to the sampling from the base of a representative finite sample called the *base series*.
-
-<!-- Popular survival analysis methods, like Kaplan-Meier and Cox regression, rely on the notion of risk-set sampling. Case-base sampling can be seen as an alternative. -->
-
-As shown by Saarela & Arjas [-@saarela2015non] (and further expanded in Saarela [-@saarela2016case]), writing the likelihood arising from this data-generating mechanism using the framework of non-homogeneous Poisson processes, we eventually reach an expression where each person-moment's contribution is of the form
-$$\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},$$
-where $N(t)$ is the counting process associated with the event of interest, $h(t)$ is the corresponding hazard function, and $\rho(t)$ is the hazard function for the Poisson process associated with case-base sampling.
-This parametric form suggests that we can readily estimate log-hazards of the form $\log(h(t)) = g(t; X)$ using logistic regression, where each observation corresponds to a person moment, the function $g(t; X)$ is linear in a finite number of parameters, and where we treat $-\log(\rho(t))$ as an offset.
-
-In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base sampling *uniformly*, i.e.
-to sample the base series uniformly from the study base.
-In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base.
-More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism  are proportional to the cardiovascular disease event rate given by the Framingham score.
-
-The `casebase` package fits the family of hazard functions of the form
-
-$$ h(t;X) = \exp[g(t;X)] $$
-where $t$ denotes time and $X$, the individual's covariate profile.
-Different functions of $t$ lead to different parametric hazard models.
-The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$.
-
-$$ h(t;X) = \exp(\beta_0 + \beta_1 X) $$
-
-The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
-This is also known as the *memoryless property* (Kalbfleisch and Prentice, 2002).
-
-The Gompertz hazard model is given by including a linear term for time:
-
-$$ h(t;X)  = \exp(\beta_0 + \beta_1 t + \beta_2 X) $$
-
-Use of $\log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
-
-$$ h(t;X)  = \exp(\beta_0 + \beta_1 \log(t) + \beta_2 X) $$
-
-For competing-risk analyses with $J$ possible events, we can show that each person-moment's contribution of the likelihood is of the form
-
-$$\frac{h_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^Jh_j(t)},$$
-
-where $N_j(t)$ is the counting process associated with the event of type $j$ and $h_j(t)$ is the corresponding hazard function.
-As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1\end{align*}}
-
-# Existing packages
-
-Survival analysis is an important branch of applied statistics and epidemiology.
-Accordingly, there is a vast ecosystem of \code{R} packages implementing different methods.
-In this section, we describe how the functionalities of \pkg{casebase} compare to these packages.
-
-At the time of writing, a cursory examination of CRAN's task view on survival analysis reveals that there are over 250 packages related to survival analysis [-@survTaskView].
-For the purposes of this article, we restricted our description to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and cumulative incidence curve estimation.
-By searching for appropriate keywords in the \code{DESCRIPTION} file of these packages, we found 60 relevant packages.
-These 60 packages were then manually examined to determine which ones are comparable to \pkg{casebase}.
-In particular, we excluded packages that were focused on a different set of problems, such as frailty and multi-state models.
-The remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with some of the functionalities they offer.
-
-Several packages implement penalized estimation for the Cox model: \pkg{glmnet} [-@regpathcox], \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche].
-Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \pkg{CoxRidge} [-@perperoglou] and \pkg{rstpm2} [-@clements_liu], while \pkg{survival} [-@survival-package] has an implementation for ridge.
-\pkg{casebase} provides penalized estimation in the context of parametric hazards.
-To our knowledge, \pkg{casebase} and \pkg{rsptm2} are the only packages to offer this functionality.
-\textcolor{red}{I can dive deeper to determine if this is a fair statement}
-
-Parametric survival models are implemented in a handful of packages: \pkg{CFC} [-@mahani2015bayesian], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu], \pkg{mets} [-@scheike2014estimating], and \pkg{survival}.
-The types of models they allow vary for each package.
-For example, \pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \pkg{flexsurv} and \pkg{survival} allow users to supply any distribution of their choice.
-Also, \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of time using splines, which allows flexible modeling of the hazard function.
-Moreover, \pkg{flexsurv} has the ability to estimate both scale and shape parameters for a variety of parametric families (see Table \ref{tab:dists}).
-As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear model of covariates (including time).
-Therefore, our package allows the user to model the effect of time using splines, and through interaction terms involving covariates and time, it also allows user to fit time-varying coefficient models.
-However, we do not explicitly model any shape parameter, unlike \pkg{flexsurv}.
-
-Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, \pkg{cmprsk} and \pkg{survival} contain implementations for competing risks.
-The differentiating factor between these packages and \pkg{casebase} is that \pkg{casebase} handles competing risks through multiple logistic regression.
-Note that \pkg{CFC} has extra features outside the scope of the metthods discussed in this paper.
-
-Finally, several packages include functions to estimate the cumulative incidence function.
-The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard.
-The first category broadly corresponds to parametric survival models, where the full hazard is explictly modeled.
-Using this estimate, the survival function and the cumulative incidence function can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below).
-Packages including this functionality include \pkg{CFC}, \pkg{Flexsurv}, \pkg{mets}, and \pkg{survival}.
-Our package \pkg{casebase} also follows this approach for both single-event and competing-event analyses.
-The second category outlined above broadly corresponds to Cox models.
-These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function.
-This is achieved using semi-parametric estimators (e.g.
-Breslow's estimator).
-Packages that implement this approach include \pkg{RiskRegression}, \pkg{rstpm2}, and \pkg{survival}.
-As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category produces estimates in the form of step-wise functions.
-This was one of the main motivations for introducting case-base sampling in survival analysis.
-
-\begin{table}[ht]
-\centering
-\resizebox{\textwidth}{!}{%
-\begin{tabular}{ccccccccc}
-\hline
-\textbf{Package}        & \vtop{\hbox{\strut \textbf{Competing}}\hbox{\strut \textbf{risks}}}  & \textbf{Non-proportional} & \textbf{Penalization} & \textbf{Splines} & \textbf{Parametric} & \textbf{Semi-parametric} & \vtop{\hbox{\strut \textbf{Interval/left}}\hbox{\strut \textbf{censoring}}} & \vtop{\hbox{\strut \textbf{Absolute}}\hbox{\strut \textbf{risk}}}           \\
-\hline
-\textbf{casebasee} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
-\textbf{CFC} \cite{mahani2015bayesian}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
-textbf{cmprsk} \cite{gray_2019}            & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\
-\textbf{coxRidge} \cite{perperoglou}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{crrp} \cite{fu}           & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
-\textbf{fastcox} \cite{yi_zou}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{Flexrsurv} \cite{flexrsurv}      &                          & x                         &                       & x                & x                   &                          &                                  & x               \\
-\textbf{Flexsurv} \cite{flexsurv}       & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\
-\textbf{glmnet} \cite{regpathcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{glmpath} \cite{park_hastie}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{mets} \cite{scheike2014estimating}           & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
-\textbf{penalized} \cite{goeman_meijer2019}      &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{RiskRegression} \cite{gerds_blanche} &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
-\textbf{Rstpm2} \cite{clements_liu}         &                          & x                         & x                     & x                & x                   & x                        & x                                & x                         \\
-\textbf{SmoothHazard} \cite{smoothHazard}   &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
-\textbf{Survival} \cite{survival-package}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x
-\end{tabular}%
-}
-\caption{Different features of interest in various survival packages.}
-\label{tab:surv-pkgs}
-\end{table}
-
-\begin{table}
-  \begin{tabular}{p{1.8in}lll}
-\hline
-    &  Parameters &  Density \proglang{R} function & \code{dist}\\
-    & {\footnotesize{(location in \code{\color{red}{red}})}} & & \\
-\hline
-    Exponential & \code{\color{red}{rate}}             & \code{dexp}   & \code{"exp"} \\
-    Weibull (accelerated failure time)     & \code{shape, {\color{red}{scale}}}     & \code{dweibull} & \code{"weibull"} \\
-    Weibull (proportional hazards)     & \code{shape, {\color{red}{scale}}}     & \code{dweibullPH} & \code{"weibullPH"} \\
-    Gamma       & \code{shape, \color{red}{rate}}      & \code{dgamma} & \code{"gamma"}\\
-    Log-normal  & \code{{\color{red}{meanlog}}, sdlog}   & \code{dlnorm} & \code{"lnorm"}\\
-    Gompertz    & \code{shape, {\color{red}{rate}}}      & \code{dgompertz} & \code{"gompertz"} \\
-    Log-logistic & \code{shape, {\color{red}{scale}}}   & \code{dllogis} & \code{"llogis"}\\
-    Generalized gamma (Prentice 1975)   & \code{{\color{red}{mu}}, sigma, Q} & \code{dgengamma} & \code{"gengamma"} \\
-    Generalized gamma (Stacy 1962)& \code{shape, {\color{red}{scale}}, k} & \code{dgengamma.orig} & \code{"gengamma.orig"} \\
-    Generalized F     (stable)    & \code{{\color{red}{mu}}, sigma, Q, P} & \code{dgenf} & \code{"genf"} \\
-    Generalized F     (original)  & \code{{\color{red}{mu}}, sigma, s1, s2} & \code{dgenf.orig} & \code{"genf.orig"} \\
-\hline
-  \end{tabular}
-  \caption{Built-in parametric survival distributions in \pkg{flexsurv}.}
-  \label{tab:dists}
-\end{table}
-
-# Implementation details
-
-The functions in the casebase package can be divided into two categories: 1) data visualization, in the form of population-time plots; and 2) parametric modeling.
-We explicitly aimed at being compatible with both \code{data.frame}s and \code{data.table}s.
-This is evident in some of the coding choices we made, and it is also reflected in our unit tests.
-
-## Population-time plots
-
-## Parametric modeling
-
-The parametric modeling step was separated into three parts:
-
-  1. case-base sampling; 
-  2. estimation of the smooth hazard function; 
-  3. calculation of the risk function.
-
-By separating the sampling and estimation functions, we allowed the possibility of users implementing more complex sampling scheme, as described in Saarela [-@saarela2016case].
-
-The sampling scheme selected for \code{sampleCaseBase} was described in Hanley and Miettinen [-@hanley2009fitting]: we first sample along the ``person'' axis, proportional to each individual's total follow-up time, and then we sample a moment uniformly over their follow-up time.
-This sampling scheme is equivalent to the following picture: imagine representing the total follow-up time of all individuals in the study along a single dimension, where the follow-up time of the next individual would start exactly when the follow-up time of the previous individual ends.
-Then the base series could be sampled uniformly from this one-dimensional representation of the overall follow-up time.
-In any case, the output is a dataset of the same class as the input, where each row corresponds to a person-moment.
-The covariate profile for each such person-moment is retained, and an offset term is added to the dataset.
-This output could then be used to fit a smooth hazard function, or for visualization of the base series.
-
-The fitting function \code{fitSmoothHazard} starts by looking at the class of the dataset: if it was generated from \code{sampleCaseBase}, it automatically inherited the class \code{cbData}.
-If the dataset supplied to \code{fitSmoothHazard} does not inherit from \code{cbData}, then the fitting function starts by calling \code{sampleCaseBase} to generate the base series.
-In other words, the occasional user can bypass \code{sampleCaseBase} altogether and only worry about the fitting function \code{fitSmoothHazard}.
-
-The fitting function retains the familiar formula interface of \code{glm}.
-The left-hand side of the formula should be the name of the column corresponding to the event type.
-The right-hand side can be any combination of the covariates, along with an explicit functional form for the time variable.
-Note that non-proportional hazard models can be achieved at this stage by adding an interaction term involving time.
-The offset term does not need to be specified by the user, as it is automatically added to the formula.
-
-To fit the hazard function, we provide several approaches that are available via the \code{family} parameter. These approaches are:
-
-  - \code{glm}: This is the familiar logistic regression.
-  - \code{glmnet}: This option allows for variable selection using Lasso or elastic-net. This functionality is provided through the \code{glmnet} package [@friedman2010jss].
-  - \code{gam}: This option provides support for *Generalized Additive Models* via the \code{gam} package [@hastie1987generalized].
-  - \code{gbm}: This option provides support for *Gradient Boosted Trees* via the \code{gbm} package. This feature is still experimental.
-
-In the case of multiple events, the hazard is fitted via multinomial regression as performed by the \pkg{VGAM} package.
-This package was selected for its ability to fit multinomial regression models with an offset.
-
-Once a model-fit object has been returned by \code{fitSmoothHazard}, all the familiar summary and diagnostic functions are available: \code{print}, \code{summary}, \code{predict}, \code{plot}, etc.
-Our package provides one more functionality: it computes risk functions from the model fit.
-For the case of a single event, it uses the familiar identity
-\begin{equation}\label{eqn:surv}
-S(t) = \exp\left(-\int_0^t h(u;X) du\right).
-\end{equation}
-The integral is computed using either the \code{stats::integrate} function or Monte-Carlo integration. The risk function (or cumulative incidence function) is then defined as
-\begin{equation}\label{eqn:CI}
-CI(t) = 1 - S(t).
-\end{equation}
-
-For the case of a competing-event analysis, the event-specific risk is computed using the following procedure: first, we compute the overall survival function (i.e. for all event types):
-
-$$ S(t) = \exp\left(-\int_0^t H(u;X) du\right),\qquad H(t;X) = \sum_{j=1}^J h_j(t;X).$$
-From this, we can derive the event-specific subdensities:
-
-$$ f_j(t) = h_j(t)S(t).$$
-
-Finally, by integrating these subdensities, we obtain the event-specific cumulative incidence functions:
-
-$$ CI_j(t) = \int_0^t f_j(u)du.$$
-
-We created \code{absoluteRisk} as an \code{S3} generic, with methods for the different types of outputs of \code{fitSmoothHazard}.
-The method dispatch system of \proglang{R} then takes care of matching the correct output to the correct methodology for calculating the cumulative incidence function, without the user's intervention.
-
-In the following sections, we illustrate these functionalities in the context of three case studies.
-
-# Case study 1--European Randomized Study of Prostate Cancer Screening
-
-For the first case study, we will use of the European Randomized Study of Prostate Cancer Screening data. This dataset is available through the \pkg{casebase} package:
-
-```{r erspc-data, eval = eval_cs1}
-data(ERSPC)
-ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
-                       levels = c(0, 1), 
-                       labels = c("Control group", "Screening group"))
-```
-
-The results of this study were published by [@schroder2009screening], and the dataset itself was obtained using the approach described in [@liu2014recovering].
-
-Population time plots can be extremely informative graphical displays of survival data.
-They should be the first step in an exploratory data analysis.
-We facilitate this task in the \pkg{casebase} package using the \code{popTime} function.
-We first create the necessary dataset for producing the population time plots, and we can generate the plot by using the corresponding \code{plot} method:
-
-```{r plot-erspc-data, eval = eval_cs1}
-pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
-plot(pt_object)
-```
-
-We can also create exposure stratified plots by specifying the \code{exposure} argument in the \code{popTime} function:
-
-```{r plot-stratified-erspc-data, eval = eval_cs1}
-pt_object_strat <- casebase::popTime(ERSPC, 
-                                     event = "DeadOfPrCa", 
-                                     exposure = "ScrArm")
-plot(pt_object_strat)
-```
-
-We can also plot them side-by-side using the \code{ncol} argument:
-
-```{r plot-stratified-erspc-data-side, eval = eval_cs1}
-plot(pt_object_strat, ncol = 2)
-```
-
-ADD PARAGRAPH ABOUT WHAT WE CAN CONCLUDE FROM THESE GRAPHS.
-
-Next, we investigate the differences between the control and the screening arms.
-A common choice for this type of analysis is to use a Cox regression model and estimate the hazard ratio for the screening group (relative to the control group).
-In \proglang{R}, it can be done as follows:
-
-```{r erspc-cox, eval = eval_cs1}
-cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
-                             data = ERSPC)
-summary(cox_model)
-```
-
-We can also plot the cumulative incidence function (CIF) for each group. However, this involves estimating the baseline hazard, which Cox regression treated as a nuisance parameter.
-
-```{r erspc-cox-cif, eval = eval_cs1}
-new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
-                       ignore = 99)
-
-plot(survfit(cox_model, newdata = new_data),
-     xlab = "Years since Randomization", 
-     ylab = "Cumulative Incidence (%)", 
-     fun = "event",
-     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-                    exp(coef(cox_model)), 
-                    exp(confint(cox_model))[1], 
-                    exp(confint(cox_model))[2]),
-     ylim = c(0, 2), yscale = 100)
-legend("topleft", 
-       legend = c("Control group", "Screening group"), 
-       col = c("red","blue"),
-       lty = c(1, 1), 
-       bg = "gray90")
-```
-
-Next, we show how case-base sampling can be used to carry out a similar analysis. We fit several models that differ in how we choose to model time.
-
-<!-- The \code{fitSmoothHazard} function provides an estimate of the hazard function $h(x, t)$ is the hazard function, where $t$ denotes the numerical value of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any).
--->
-
-First, we will fit an exponential model. Recall that this corresponds to excluding time from the linear predictor.
-
-```{r erspc-casebase-exponential, eval = eval_cs1}
-casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
-                                                  data = ERSPC, 
-                                                  ratio = 100)
-
-summary(casebase_exponential)
-exp(coef(casebase_exponential)[2])
-exp(confint(casebase_exponential)[2,])
-```
-
-<!-- The \code{absoluteRisk} function provides an estimate of the cumulative incidence curves for a specific risk profile using the following equation: -->
-
-<!-- $$ CI(x, t) = 1 - \exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$ -->
-
-We can then use the \code{absoluteRisk} function to get an estimate of the cumulative incidence curve for a specific covariate profile.
-In the plot below, we overlay the estimated CIF from the exponential model on the Cox model CIF:
-
-```{r erspc-casebase-cif, eval = eval_cs1}
-smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
-                                          time = seq(0,15,0.1), 
-                                          newdata = new_data)
-
-plot(survfit(cox_model, newdata = new_data),
-     xlab = "Years since Randomization", 
-     ylab = "Cumulative Incidence (%)", 
-     fun = "event",
-     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-                    exp(coef(cox_model)), 
-                    exp(confint(cox_model))[1], 
-                    exp(confint(cox_model))[2]),
-     ylim = c(0, 2), yscale = 100)
-lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
-lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
-
-legend("topleft", 
-       legend = c("Control group (Cox)","Control group (Casebase)",
-                  "Screening group (Cox)", "Screening group (Casebase)"), 
-       col = c("red","red", "blue","blue"),
-       lty = c(1, 2, 1, 2), 
-       bg = "gray90")
-```
-
-As we can see, the exponential model gives us an estimate of the hazard ratio that is similar to the one obtained using Cox regression.
-However, the CIF estimates are quite different.
-Based on what we observed in the population time plot, where more events are observed later on in time, we do not expect a constant hazard would be a good description for this data.
-In other words, this poor fit for the exponential hazard is expected.
-A constant hazard model would overestimate the cumulative incidence earlier on in time, and underestimate it later on; this is what we see on the cumulative incidence plot.
-This example demonstrates the benefits of population time plots as an exploratory analysis tool.
-
-For the next model, we include time as a linear term. Recall that this corresponds to a Weibull model.
-
-```{r erspc-casebase-weibull, eval = eval_cs1}
-casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
-                                 data = ERSPC, 
-                                 ratio = 100)
-
-summary(casebase_time)
-exp(coef(casebase_time))
-exp(confint(casebase_time))
-```
-
-Again, the estimate of the hazard ratio is similar to that obtained from Cox regression. We then look at the estimate of the CIF:
-
-```{r erspc-casebase-weibull-cif, eval = eval_cs1}
-smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
-                                          time = seq(0,15,0.1), 
-                                          newdata = new_data)
-
-plot(survfit(cox_model, newdata = new_data),
-     xlab = "Years since Randomization", 
-     ylab = "Cumulative Incidence (%)", 
-     fun = "event",
-     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-                    exp(coef(cox_model)), 
-                    exp(confint(cox_model))[1], 
-                    exp(confint(cox_model))[2]),
-     ylim = c(0, 2), yscale = 100)
-lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
-lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
-
-legend("topleft", 
-       legend = c("Control group (Cox)","Control group (Casebase)",
-                  "Screening group (Cox)", "Screening group (Casebase)"), 
-       col = c("red","red", "blue","blue"),
-       lty = c(1, 2, 1, 2), 
-       bg = "gray90")
-```
-
-We see that the Weibull model leads to a better fit of the CIF than the Exponential model, when compared with the semi-parametric approach.
-
-Finally, we can model the hazard as a smooth function of time using the \code{splines} package:
-
-```{r erspc-casebase-splines, eval = eval_cs1}
-casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
-                                    data = ERSPC, 
-                                    ratio = 100)
-
-summary(casebase_splines)
-exp(coef(casebase_splines))
-exp(confint(casebase_splines))
-```
-
-Once again, we can see that the estimate of the hazard ratio is similar to that from the Cox regression. We then look at the estimate of the CIF:
-
-```{r erspc-casebase-splines-cif, eval = eval_cs1}
-smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
-                                    time = seq(0,15,0.1), 
-                                    newdata = new_data)
-
-plot(survfit(cox_model, newdata = new_data),
-     xlab = "Years since Randomization", 
-     ylab = "Cumulative Incidence (%)", 
-     fun = "event",
-     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-                    exp(coef(cox_model)), 
-                    exp(confint(cox_model))[1], 
-                    exp(confint(cox_model))[2]),
-     ylim = c(0, 2), yscale = 100)
-lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
-lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
-
-legend("topleft", 
-       legend = c("Control group (Cox)","Control group (Casebase)",
-                  "Screening group (Cox)", "Screening group (Casebase)"), 
-       col = c("red","red", "blue","blue"),
-       lty = c(1, 2, 1, 2), 
-       bg = "gray90")
-```
-
-Qualitatively, the combination of splines and case-base sampling produces the parametric model that most closely resembles the Cox model.
-In the following table, we see that the confidence intervals are also similar across all four models.
-In other words, this reinforces the idea that, under proportional hazards, we do not need to model the full hazard to obtain reliable estimates of the hazard ratio.
-Of course, different parametric models for the hazards give rise to qualitatively different estimates for the CIF.
-
-```{r erspc-95ci, echo = FALSE, eval = eval_cs1}
-CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
-CI_exp <- c(exp(coef(casebase_exponential))["ScrArmScreening group"], exp(confint(casebase_exponential))["ScrArmScreening group",])
-CI_time <- c(exp(coef(casebase_time))["ScrArmScreening group"], exp(confint(casebase_time))["ScrArmScreening group",])
-CI_splines <- c(exp(coef(casebase_splines))["ScrArmScreening group"], exp(confint(casebase_splines))["ScrArmScreening group",])
-
-table_ci <- rbind(CI_cox, CI_exp, CI_time, CI_splines)
-colnames(table_ci) <- c("HR", "LB", "UB")
-
-data.frame(Model = c("Cox", "Exponential", "Weibull", "Splines")) %>% 
-  bind_cols(as.data.frame(table_ci)) %>% 
-  mutate(CI = paste0("(", round(LB, 2), ", ",
-                     round(UB, 2), ")"),
-         Width = UB - LB) %>% 
-  dplyr::select(-LB, -UB) %>% 
-  knitr::kable(format = "latex", booktabs = TRUE)  %>%
-  kable_styling()
-```
-
-As noted above, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models.
-Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) splines model.
-
-```{r erscp-compare, eval = eval_cs1}
-anova(casebase_exponential, casebase_splines, test = "LRT")
-```
-
-As expected, we see that splines model provides a better fit. Similarly, we can compare the AIC for all three case-base models:
-
-```{r erscp-aic, eval = eval_cs1}
-c("Exp." = AIC(casebase_exponential),
-  "Weib." = AIC(casebase_time),
-  "Splines" = AIC(casebase_splines))
-```
-
-Once again, we have evidence that the splines model provides the best fit.
-
-With this first case study, we saw how population-time plots can provide a useful summary of the incidence for our event of interests.
-In particular, we saw how it can provide evidence against an exponential hazard model.
-We then explored how different parametric models can provide different estimates of the CIF, even though they all give similar estimates for the hazard ratios.
-
-# Case study 2--Bone-marrow transplant
-
-In the next case study, we show how case-base sampling can also be used in the context of a competing risk analysis.
-For illustrative purposes, we will use the same data that was used in Scrucca *et al* [-@scrucca2010regression].
-The data was downloaded from the first author's website, and it is also available as part of the \pkg{casebase} package.
-
-```{r bmtcrr-data, eval = eval_cs2, message = FALSE}
-
-data(bmtcrr)
-```
-
-The data contains information on 177 patients who received a stem-cell transplant for acute leukemia.
-The event of interest is relapse, but other competing causes (e.g.
-transplant-related death) were also recorded.
-Several covariates were captured at baseline: sex, disease type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL and AML, respectively), disease phase at transplant (Relapse, CR1, CR2, CR3), source of stem cells (bone marrow and peripheral blood, coded as BM+PB, or only peripheral blood, coded as PB), and age.
-A summary of these baseline characteristics appear in Table \ref{tab:table1bmtcrr}.
-We note that the statistical summaries were generated differently for different variable types: for continuous variables, we gave the range, followed by the mean and standard deviation; for categorical variables, we gave the counts for each category.
-
-```{r echo = FALSE, eval = eval_cs2}
-tibble::tribble(
-    ~Variable, ~Description, ~`Statistical summary`,
-    "\\code{Sex}", "Sex", "M=Male (100)",
-    "", "", "F=Female (77)",
-    "\\code{D}", "Disease", "ALL (73)",
-    "", "", "AML (104)",
-    "\\code{Phase}", "Phase", "CR1 (47)", 
-    "", "", "CR2 (45)", 
-    "", "", "CR3 (12)", 
-    "", "", "Relapse (73)",
-    "\\code{Source}", "Type of transplant", "BM+PB (21)", 
-    "", "", "PB (156)",
-    "\\code{Age}", "Age of patient (years)", "4--62", 
-    "", "", "30.47 (13.04)",
-    "\\code{ftime}", "Failure time (months)", "0.13--131.77",
-    "", "", "20.28 (30.78)",
-    "\\code{Status}", "Status indicator", "0=censored (46)", 
-    "", "", "1=relapse (56)", 
-    "", "", "2=competing event (75)"
-) %>% 
-  knitr::kable(format = "latex", booktabs = TRUE,
-               caption = "Baseline characteristics of patients in the stem-cell transplant study.", 
-               label = "table1bmtcrr",
-               escape = FALSE) %>%
-  kable_styling()
-```
-
-First, we can look at a population-time plot to visualize the incidence density of both relapse and the competing events.
-In Figure \ref{fig:compPop}, failure times are highlighted on the plot using red points for the event of interest (panel A) and blue dots for competing events (panel B).
-In both panels, we see evidence of a non-constant hazard function: the density of points is larger at the beginning of follow-up than at the end.
-
-```{r compPop, fig.cap="Population-time plot for the stem-cell transplant study.
-A) The red points represent relapse events.
-B) The blue points represent the competing events.", echo=FALSE, eval = eval_cs2}
-library(cowplot)
-pt_object <- casebase::popTime(bmtcrr, event = "Status", time = "ftime")
-
-pt_object_comp <- bmtcrr %>% 
-    mutate(Status = case_when(Status == 1 ~ 2L, 
-                              Status == 2 ~ 1L, 
-                              TRUE ~ Status)) %>% 
-    casebase::popTime(event = "Status", time = "ftime")
-
-plot_grid(plot(pt_object) + ggtitle("Relapse"), 
-          plot(pt_object_comp, point.colour = "blue") +
-            ggtitle("Competing events"), 
-          labels = c('A', 'B'))
-```
-
-Our main objective is to compute the absolute risk of relapse for a given set of covariates. First, we fit a smooth hazard to the data using a linear term for time:
-
-```{r bmtcrr-casebase-weibull, warning = FALSE, eval = eval_cs2}
-model_cb <- fitSmoothHazard(
-    Status ~ ftime + Sex + D + Phase + Source + Age, 
-    data = bmtcrr, 
-    ratio = 100, 
-    time = "ftime")
-```
-
-We will compare our hazard ratio estimates to that obtained from a Cox regression. To do so, we need to treat the competing event as censoring.
-
-```{r bmtcrr-cox, echo = TRUE, eval = eval_cs2}
-library(survival)
-# Treat competing event as censoring
-model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-                   data = bmtcrr)
-```
-
-```{r bmtcrr-cis, echo=FALSE, eval = eval_cs2}
-# Table of coefficients----
-library(glue)
-z_value <- qnorm(0.975)
-foo <- summary(model_cb)@coef3
-table_cb <- foo[!grepl("Intercept", rownames(foo)) & 
-                    !grepl("ftime", rownames(foo)) &
-                    grepl(":1", rownames(foo)), 
-                1:2]
-table_cb <- cbind(table_cb[,1], 
-                  table_cb[,1] - z_value * table_cb[,2],
-                  table_cb[,1] + z_value * table_cb[,2])
-table_cb <- round(exp(table_cb), 2)
-
-table_cox <- round(summary(model_cox)$conf.int[,-2], 2)
-rownames(table_cb) <- rownames(table_cox)
-colnames(table_cb) <- colnames(table_cox)
-
-table_cb <- as.data.frame(table_cb) %>%
-  tibble::rownames_to_column("Covariates") %>% 
-    mutate(CI = glue::glue_data(., "({`lower .95`}, {`upper .95`})")) %>% 
-    rename(HR = `exp(coef)`) %>% 
-    select(Covariates, HR, CI)
-
-table_cox <- as.data.frame(table_cox) %>%
-  tibble::rownames_to_column("Covariates") %>% 
-  dplyr::mutate(CI_Cox = glue::glue_data(., "({`lower .95`}, {`upper .95`})")) %>%
-  dplyr::rename(HR_Cox = `exp(coef)`) %>%
-  dplyr::select(Covariates, HR_Cox, CI_Cox)
-
-table_cb %>% 
-  dplyr::inner_join(table_cox, by = "Covariates") %>%
-  dplyr::mutate(Covariates = dplyr::case_when(
-    Covariates == "SexM" ~ "Sex",
-    Covariates == "DAML" ~ "Disease",
-    Covariates == "PhaseCR2" ~ "Phase (CR2 vs. CR1)",
-    Covariates == "PhaseCR3" ~ "Phase (CR3 vs. CR1)",
-    Covariates == "PhaseRelapse" ~ "Phase (Relapse vs. CR1)",
-    Covariates == "SourcePB" ~ "Source",
-    TRUE ~ Covariates
-    )) %>% 
-  knitr::kable(format = "latex", booktabs = TRUE,
-               col.names = c("Covariates", "HR", "95% CI", "HR", "95% CI"),
-               caption = "ADD CAPTION") %>% 
-  kable_styling() %>% 
-  add_header_above(c(" " = 1, "Case-Base" = 2, "Cox" = 2))
-```
-
-From the fit object, we can extract both the hazard ratios and their corresponding confidence intervals.
-These quantities appear in Table \ref{tab:bmtcrr-cis}.
-As we can see, the only significant hazard ratio identified by case-base sampling is the one associated with the phase of the disease at transplant.
-More precisely, being in relapse at transplant is associated with a hazard ratio of 3.89 when compared to CR1.
-
-Given the estimate of the hazard function obtained using case-base sampling, we can compute the absolute risk curve for a fixed covariate profile.
-We perform this computation for a 35 year old woman who received a stem-cell transplant from peripheral blood at relapse.
-We compared the absolute risk curve for such a woman with acute lymphoblastic leukemia with that for a similar woman with acute myeloblastic leukemia.
-We will estimate the curve from 0 to 60 months.
-
-```{r cb_risk, warning = FALSE, cache = FALSE, eval = eval_cs2}
-# Pick 100 equidistant points between 0 and 60 months
-time_points <- seq(0, 60, length.out = 50)
-
-# Data.frame containing risk profile
-newdata <- data.frame("Sex" = factor(c("F", "F"), 
-                                     levels = levels(bmtcrr[,"Sex"])),
-                      "D" = c("ALL", "AML"), # Both diseases
-                      "Phase" = factor(c("Relapse", "Relapse"), 
-                                       levels = levels(bmtcrr[,"Phase"])),
-                      "Age" = c(35, 35),
-                      "Source" = factor(c("PB", "PB"), 
-                                        levels = levels(bmtcrr[,"Source"])))
-
-# Estimate absolute risk curve
-risk_cb <- absoluteRisk(object = model_cb, time = time_points,
-                        method = "numerical", newdata = newdata)
-```
-
-We will compare our estimates to that obtained from a corresponding Fine-Gray model [@fine1999proportional].
-The Fine-Gray model is a semiparametric model for the cause-specific *subdistribution*, i.e.
-the function $f_k(t)$ such that
-$$CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),$$
-where $CI_k(t)$ is the cause-specific cumulative indicence.
-The Fine-Gray model allows to directly assess the effect of a covariate on the subdistribution, as opposed to the hazard.
-For the computation, we will use the \pkg{timereg} package [@timereg].
-
-```{r fg_risk, eval = eval_cs2, echo = TRUE}
-library(timereg)
-model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) + 
-                          const(Phase) + const(Source) + const(Age), 
-                      data = bmtcrr, cause = 1, model = "fg")
-
-# Estimate absolute risk curve
-risk_fg <- predict(model_fg, newdata, times = time_points)
-```
-
-```{r bmtcrr-risk, echo = FALSE, fig.cap="\\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups.
-The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.", eval = eval_cs2}
-library(ggplot2)
-risk_all <- dplyr::bind_rows(
-    data.frame(Time = time_points, 
-               Method = "Case-base", 
-               Risk = risk_cb[,2], 
-               Disease = "ALL",
-               stringsAsFactors = FALSE),
-    data.frame(Time = time_points, 
-               Method = "Case-base", 
-               Risk = risk_cb[,3], 
-               Disease = "AML",
-               stringsAsFactors = FALSE),
-    data.frame(Time = time_points,
-               Method = "Fine-Gray",
-               Risk = risk_fg$P1[1,],
-               Disease = "ALL",
-               stringsAsFactors = FALSE),
-    data.frame(Time = time_points,
-               Method = "Fine-Gray",
-               Risk = risk_fg$P1[2,],
-               Disease = "AML",
-               stringsAsFactors = FALSE)
-    )
-
-ggplot(risk_all, aes(x = Time, y = Risk, colour = Method)) + 
-  # geom_line for smooth curve
-  geom_line(data = dplyr::filter(risk_all, Method == "Case-base")) + 
-  # geom_step for step function
-  geom_step(data = dplyr::filter(risk_all, Method != "Case-base")) + 
-  facet_grid(Disease ~ .) +
-  ylim(c(0,1)) +
-  theme_minimal() +
-  theme(legend.position = 'top') +    
-  xlab("Time (in Months)") + ylab("Relapse risk")
-```
-
-Figure \ref{fig:compAbsrisk} shows the absolute risk curves for both case-base sampling and the Fine-Gray model.
-As we can see, the two approaches agree quite well for acute myeloblastic leukemia; however, there seems to be a difference of about 5% between the two curves for acute lymphoblastic leukemia.
-This difference does not appear to be significant: the curve from case-base sampling is contained within a 95% confidence band around the Fine-Gray absolute risk curve (figure not shown).
 
@@
 
-cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
+cb.ModelInter=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
                                        formula_time = ~log(d.time),alpha=1,ratio=ratio)
@@
 #Estimating the absolute risk curve using the newData parameter.
-casebaseCI=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 20),newdata = x , s="lambda.1se",method=c("numerical"))
+casebaseCIAll=casebase::absoluteRisk(cb.ModelInter,time = seq(1,max(completeData$d.time), 400),newdata = x , s="lambda.1se",method=c("numerical"))
 
-cbCIAve=data.frame(casebaseCI[,1],rowMeans(casebaseCI[,-c(1)]))
+cbCIAve=data.frame(casebaseCIAll[,1],rowMeans(casebaseCIAll[,-c(1)]))
 ```
@@
 ```
-We used the survival package to calculate cox regression's cumulative incidience without penalization and the kaplan-meier curve for incidience.
-
-```{r unpenalizedCox, echo = FALSE, eval = eval_cs3}
-Cox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,-c(1)]))
-#coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
-#abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)
-#abCFup<-abcoxFitup
-#abCFup$surv<-rowMeans(abCFup$surv)
-#cb.ModelNorm=casebase::fitSmoothHazard(V2~.,time="V1",event="V2",
-       #                                ratio=rat,data=fullDataCox)
-#summary(cb.ModelNorm)#need to figure out how to deal with NAs
-#casebaseAbsoluteNorm=casebase::absoluteRisk(cb.ModelNorm,time = seq(1,max(completeData$d.time), 200),newdata = newData , s="lambda.1se",method=c("numerical"))
-#cabn=data.frame(casebaseAbsoluteNorm[,1],rowMeans(casebaseAbsoluteNorm[,-c(1)]))
-```
+We used the survival package to calculate the kaplan-meier absolute risk function.
 
 ```{r fitKM+abRisk, echo = TRUE, eval = eval_cs3}
-
-coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
-abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)
-
 #creating a surv object to be used to fit an unadjusted absolute risk curve.
@@
 
-Now that our four cumulative incidence estimations have been calculated, we compare them all in Figure \ref{fig:abSC}.
+Now that our three absolute risk functions have been calculated, we compare them all in Figure \ref{fig:abSC}.
 Both coxnet and casebase decrease the absolute risk by the end of the study, in comparison to kaplan-meier.
-```{r abSupportComparison, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
-abCFup<-abcoxFitup
-abCFup$surv<-rowMeans(abCFup$surv)
+```{r abSupportComparison, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
 # A plot to compare all three Absolute risk (Commulative Incidence)
-plot(cbCIAve[,1],cbCIAve[,2],type='l',col="black",lwd=3,main="Cumulative Incidence comparison",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
+plot(cbCIAve[,1],cbCIAve[,2],type='l',col="black",lwd=3,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
 lines(abCF,col="red",fun="event",lwd=3,conf.int = FALSE)
 lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
-lines(abCFup ,col="purple",fun="event",lwd=3,conf.int = FALSE)
 legend("bottomright", 
-       legend = c( "Penalized Weibull with casebase","Coxnet","Cox","Kaplan-Meier"), 
-       col = c("black","red","Blue","purple"),
+       legend = c( "Casebase (Lasso+linear)","semi-parametric (Cox)+glmnet","KM curve"), 
+       col = c("black","red","Blue"),
        lty = c(1, 1, 1), 
@@
 colnames(estimatescb)<-c("Coefs","Estimate","Model")
-estimatescox=setDT(as.data.frame(coef(coxglmFit)[-c(1),1]), keep.rownames = TRUE)[]
+estimatescox=setDT(as.data.frame(coef(coxNet)[-c(1),1]), keep.rownames = TRUE)[]
 estimatescox$Model="Cox"
@@
 ```
-
-# Case study 4--Stanford Heart Transplant Data
-
-Although the previous case studies provide a broad overview of the capabilities of \pkg{casebase}, they all have two properties in common:
-\begin{itemize}
-  \item All covariates are fixed, i.e. they do not change over time;
-  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
-\end{itemize}
-
-Case-base sampling has been used in the literature to study vaccination safety, where the exposure period was defined as the week following vaccination [@saarela2015case].
-Hence, the main covariate of interest, i.e.
-exposure to the vaccine, was changing over time.
-In this context, case-base sampling offers an efficient alternative to nested case-control designs or self-matching.
-
-The purpose of the next case study is to show how \pkg{casebase} can also be used for survival analysis problems that do not share the properties above.
-To this end, we will use the Stanford Heart Transplant Data [@clark1971cardiac,@crowley1977covariance]; we will use the version available in the \pkg{survival} package.
-
-Recall the setting of this study: patients were admitted to the Stanford program after meeting with their physician and determining that they were unlikely to respond to other forms of treatment.
-After enrollment, the program searches for a suitable donor for the patient, which can take anywhere between a few days to almost a year.
-We are interested in the effect of a heart transplant on survival; therefore, the patient is considered exposed only after the transplant has occurred.
-
-As before, we can look at the population-time plot for a graphical summary of the event incidence.
-As we can see, most events occur early during the follow-up period, and therefore we do not expect the hazard to be constant.
-
-```{r stanford-poptime, eval = eval_cs4}
-library(survival)
-library(casebase)
-
-stanford_popTime <- popTime(jasa, time = "futime", 
-                            event = "fustat")
-plot(stanford_popTime)
-```
-
-Since the exposure is time-dependent, we need to manually define the exposure variable *after* case-base sampling and *before* fitting the hazard function.
-For this reason, we will use the `sampleCaseBase` function directly.
-
-```{r, eval = eval_cs4}
-library(tidyverse)
-library(lubridate)
-
-cb_data <- sampleCaseBase(jasa, time = "futime", 
-                          event = "fustat", ratio = 10)
-```
-
-Next, we will compute the number of days from acceptance into the program to transplant, and we use this variable to determine whether each population-moment is exposed or not.
-
-```{r, eval = eval_cs4}
-# Define exposure variable
-cb_data <- mutate(cb_data,
-                  txtime = time_length(accept.dt %--% tx.date, 
-                                       unit = "days"),
-                  exposure = case_when(
-                    is.na(txtime) ~ 0L,
-                    txtime > futime ~ 0L,
-                    txtime <= futime ~ 1L
-                  ))
-```
-
-Finally, we can fit the hazard using various linear predictors.
-
-```{r, eval = eval_cs4, warning = FALSE}
-library(splines)
-# Fit several models
-fit1 <- fitSmoothHazard(fustat ~ exposure,
-                        data = cb_data, time = "futime")
-fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
-                        data = cb_data, time = "futime")
-fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
-                        data = cb_data, time = "futime")
-fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
-                        data = cb_data, time = "futime")
-```
-
-Note that the fourth model (i.e.
-`fit4`) includes an interaction term between exposure and follow-up time.
-In other words, this model no longer exhibit proportional hazards.
-The evidence of non-proportionality of hazards in the Stanford Heart Transplant data has been widely discussed [@arjas1988graphical].
-
-We can then compare the goodness of fit of these four models using the Akaike Information Criterion (AIC).
-
-```{r, eval = eval_cs4}
-# Compute AIC
-c("Model1" = AIC(fit1),
-  "Model2" = AIC(fit2),
-  "Model3" = AIC(fit3),
-  "Model4" = AIC(fit4))
-```
-
-As we can, the best fit is the fourth model. By visualizing the hazard functions for both exposed and unexposed individuals, we can more clearly see how the hazards are no longer proportional.
-
-```{r stanford-hazard, eval = eval_cs4}
-# Compute hazards---
-# First, create a list of time points for both exposure status
-hazard_data <- expand.grid(exposure = c(0, 1),
-                           futime = seq(0, 1000,
-                                        length.out = 100))
-# Set the offset to zero
-hazard_data$offset <- 0 
-# Use predict to get the fitted values, and exponentiate to 
-# transform to the right scale
-hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
-                                 type = "link"))
-# Add labels for plots
-hazard_data$Status = factor(hazard_data$exposure,
-                            labels = c("NoTrans", "Trans"))
-
-ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
-    geom_line() +
-    theme_minimal() +
-    theme(legend.position = 'top') +
-    ylab('Hazard') + xlab('Follow-up time')
-```
-
-The non-proportionality seems to be more pronounced at the beginning of follow-up than the end.
-Finally, we can turn these estimate of the hazard function into estimates of the cumulative incidence functions.
-
-```{r stanford-risk, eval = eval_cs4}
-# Compute absolute risk curves
-newdata <- data.frame(exposure = c(0, 1))
-absrisk <- absoluteRisk(fit4, newdata = newdata, 
-                        time = seq(0, 1000, length.out = 100))
-
-colnames(absrisk) <- c("Time", "NoTrans", "Trans")
-
-# Rearrange the data
-absrisk <- gather(as.data.frame(absrisk),
-                  "Status", "Risk", -Time)
-
-ggplot(absrisk, aes(Time, Risk, colour = Status)) +
-  geom_line() +
-  theme_minimal() +
-  theme(legend.position = 'top') +
-  expand_limits(y = 0.5) +
-  xlab('Follow-up time') + ylab('Cum. Incidence')
-```
-
-Note that we can easily adapt the code above to the situation where a patient receives a heart transplant at a point in time of interest, for example after 30 days.
-
-```{r stanford-1y-haz, eval = eval_cs4}
-# Compute hazards---
-# First, create a list of time points for both exposure status
-one_yr_haz <- data.frame(futime = seq(0, 365,
-                                      length.out = 100))
-one_yr_haz <- mutate(one_yr_haz,
-                     offset = 0,
-                     exposure = if_else(futime < 30, 0, 1))
-one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
-                                type = "link"))
-ggplot(one_yr_haz, aes(futime, hazard)) +
-  geom_line() +
-  theme_minimal() +
-  theme(legend.position = 'top') +
-  ylab('Hazard') + xlab('Follow-up time') +
-  geom_vline(xintercept = 30, linetype = 'dashed')
-```
-
-We can then compare the 1-year mortality risk without transplant and with transplant at 30 days.
-
-```{r stanford-1y-risk, eval = eval_cs4}
-absoluteRisk(fit4, newdata = data.frame(exposure = 0),
-             time = 365)
-
-# Use the trapezoidal rule to estimate the cumulative hazard
-min_hazard <- min(one_yr_haz$hazard)
-max_hazard <- max(one_yr_haz$hazard)
-increment <- 365/99
-cumhaz_est <- 0.5*increment*(sum(2*one_yr_haz$hazard) - min_hazard - max_hazard)
-1 - exp(-cumhaz_est)
-```
-
-As we can see, the risk estimate at 1-year is about 30% lower if the patient receives a heart transplant at 30 days.
-
-# Discussion
-
-In this article, we presented the \proglang{R} package \pkg{casebase} that provides functions to fit smooth parametric hazards and estimate cumulative incidence functions using case-base sampling.
-We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through three case studies.
-
-<!-- In the following table we provide a comparison between the Cox model and case-base sampling: -->
-
-```{r, echo=FALSE, eval = FALSE, results='asis'}
-knitr::kable(data.frame(`feature` = c("model type", "time", "cumulative incidence", "non-proportional hazards","model testing","competing risks", "prediction"),
-           `Cox` = c("semi-parametric","left hand side of the equation", "step function", "interaction of covariates with time"," ", "difficult", "Kaplan-Meier-based"),
-           `Case Base Sampling` = c("fully parametric (logistic/multinomial regression)", "right hand side - allows flexible modeling of time",
-           "smooth-in-time curve", "interaction of covariates with time", "make use of GLM framework (LRT, AIC, BIC)",
-           "cause-specific cumulative incidence functions (CIFs) directly obtained via multinomial regression", "ROC, AUC, risk reclassification probabilities")), escape = FALSE)
-```
-
-\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval censoring, whereas \pkg{casebase} currently cannot.
-Interval censoring is a simple feature to implement and is a potential improvement to add to \pkg{casebase}.
-
-# Environment Details
-
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
-
-```{r colophon, cache = FALSE}
-# which R packages and versions?
-devtools::session_info()
-```
-
-The current Git commit details are:
-```{r}
-# what commit is this file at? 
-git2r::repository(here::here())
-```

### 2020-02-06 cleanedCS3.tex
Switched to non interaction model
note that I only ran CS3 so the pdf is not complete. fixes include all red text

Jesse Still needs to clean up the text and deliberate his write up in the case study. Code will also be cleaned to be more readable.

--- 2020-02-06 analysis/paper/paper.tex
@@
 \begin{document}
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-The purpose of the \pkg{casebase} package is to provide practitioners
-with an easy-to-use software tool to predict the risk (or cumulative
-incidence (CI)) of an event, for a particular patient. The following
-points should be noted:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  hazard ratios
-\item
-  If, however, the absolute risks are of interest, they have to be
-  recovered using the semi-parametric Breslow estimator
-\item
-  Alternative approaches for fitting flexible hazard models for
-  estimating absolute risks, not requiring this two-step approach? Yes!
-  \citep{hanley2009fitting}
-\end{enumerate}
-
-\citep{hanley2009fitting} propose a fully parametric hazard model that
-can be fit via logistic regression. From the fitted hazard function,
-cumulative incidence and, thus, risk functions of time, treatment and
-profile can be easily derived.
-
-\hypertarget{theoretical-details}{%
-\section{Theoretical details}\label{theoretical-details}}
-
-As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
-key idea behind case-base sampling is to discretize the study base into
-an infinite amount of \emph{person moments}. These person moments are
-indexed by both an individual in the study and a time point, and
-therefore each person moment has a covariate profile, an exposure status
-and an outcome status attached to it. We note that there is only a
-finite number of person moments associated with the event of interest
-(what Hanley \& Miettinen call the \emph{case series}). The case-base
-sampling refers to the sampling from the base of a representative finite
-sample called the \emph{base series}.
-
-As shown by Saarela \& Arjas \citeyearpar{saarela2015non} (and further
-expanded in Saarela \citeyearpar{saarela2016case}), writing the
-likelihood arising from this data-generating mechanism using the
-framework of non-homogeneous Poisson processes, we eventually reach an
-expression where each person-moment's contribution is of the form
-\[\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},\] where \(N(t)\) is the counting
-process associated with the event of interest, \(h(t)\) is the
-corresponding hazard function, and \(\rho(t)\) is the hazard function
-for the Poisson process associated with case-base sampling. This
-parametric form suggests that we can readily estimate log-hazards of the
-form \(\log(h(t)) = g(t; X)\) using logistic regression, where each
-observation corresponds to a person moment, the function \(g(t; X)\) is
-linear in a finite number of parameters, and where we treat
-\(-\log(\rho(t))\) as an offset.
-
-In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
-suggest performing case-base sampling \emph{uniformly}, i.e.~to sample
-the base series uniformly from the study base. In terms of Poisson
-processes, this sampling strategy corresponds essentially to a
-time-homogeneous Poisson process with intensity equal to \(b/B\), where
-\(b\) is the number of sampled observations in the base series, and
-\(B\) is the total population-time for the study base. More complex
-examples are also available; see for example Saarela \& Arjas
-\citeyearpar{saarela2015non}, where the probabilities of the sampling
-mechanism are proportional to the cardiovascular disease event rate
-given by the Framingham score.
-
-The \texttt{casebase} package fits the family of hazard functions of the
-form
-
-\[ h(t;X) = \exp[g(t;X)] \] where \(t\) denotes time and \(X\), the
-individual's covariate profile. Different functions of \(t\) lead to
-different parametric hazard models. The simplest of these models is the
-one-parameter exponential distribution which is obtained by taking the
-hazard function to be constant over the range of \(t\).
-
-\[ h(t;X) = \exp(\beta_0 + \beta_1 X) \]
-
-The instantaneous failure rate is independent of \(t\), so that the
-conditional chance of failure in a time interval of specified length is
-the same regardless of how long the individual has been in the study.
-This is also known as the \emph{memoryless property} (Kalbfleisch and
-Prentice, 2002).
-
-The Gompertz hazard model is given by including a linear term for time:
-
-\[ h(t;X)  = \exp(\beta_0 + \beta_1 t + \beta_2 X) \]
-
-Use of \(\log(t)\) yields the Weibull hazard which allows for a power
-dependence of the hazard on time (Kalbfleisch and Prentice, 2002):
-
-\[ h(t;X)  = \exp(\beta_0 + \beta_1 \log(t) + \beta_2 X) \]
-
-For competing-risk analyses with \(J\) possible events, we can show that
-each person-moment's contribution of the likelihood is of the form
-
-\[\frac{h_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^Jh_j(t)},\]
-
-where \(N_j(t)\) is the counting process associated with the event of
-type \(j\) and \(h_j(t)\) is the corresponding hazard function. As may
-be expected, this functional form is similar to the terms appearing in
-the likelihood function for multinomial
-regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1\end{align*}}
-
-\hypertarget{existing-packages}{%
-\section{Existing packages}\label{existing-packages}}
-
-Survival analysis is an important branch of applied statistics and
-epidemiology. Accordingly, there is a vast ecosystem of \code{R}
-packages implementing different methods. In this section, we describe
-how the functionalities of \pkg{casebase} compare to these packages.
-
-At the time of writing, a cursory examination of CRAN's task view on
-survival analysis reveals that there are over 250 packages related to
-survival analysis \citeyearpar{survTaskView}. For the purposes of this
-article, we restricted our description to packages that implement at
-least one of the following features: parametric modeling,
-non-proportional hazard models, competing risk analysis, penalized
-estimation, and cumulative incidence curve estimation. By searching for
-appropriate keywords in the \code{DESCRIPTION} file of these packages,
-we found 60 relevant packages. These 60 packages were then manually
-examined to determine which ones are comparable to \pkg{casebase}. In
-particular, we excluded packages that were focused on a different set of
-problems, such as frailty and multi-state models. The remaining 14
-packages appear in Table \ref{tab:surv-pkgs}, along with some of the
-functionalities they offer.
-
-Several packages implement penalized estimation for the Cox model:
-\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
-\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
-\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
-packages also include penalized estimation in the context of Cox models
-with time-varying coefficients: elastic-net penalization with
-\pkg{CoxRidge} \citeyearpar{perperoglou} and \pkg{rstpm2}
-\citeyearpar{clements_liu}, while \pkg{survival}
-\citeyearpar{survival-package} has an implementation for ridge.
-\pkg{casebase} provides penalized estimation in the context of
-parametric hazards. To our knowledge, \pkg{casebase} and \pkg{rsptm2}
-are the only packages to offer this functionality.
-\textcolor{red}{I can dive deeper to determine if this is a fair statement}
-
-Parametric survival models are implemented in a handful of packages:
-\pkg{CFC} \citeyearpar{mahani2015bayesian}, \pkg{flexsurv}
-\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
-\pkg{rsptm2} \citeyearpar{clements_liu}, \pkg{mets}
-\citeyearpar{scheike2014estimating}, and \pkg{survival}. The types of
-models they allow vary for each package. For example, \pkg{SmoothHazard}
-is limited to Weibull distributions \citeyearpar{smoothHazard}, whereas
-both \pkg{flexsurv} and \pkg{survival} allow users to supply any
-distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard},
-\pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of
-time using splines, which allows flexible modeling of the hazard
-function. Moreover, \pkg{flexsurv} has the ability to estimate both
-scale and shape parameters for a variety of parametric families (see
-Table \ref{tab:dists}). As discussed above, \pkg{casebase} can model any
-parametric family whose log-hazard can be expressed as a linear model of
-covariates (including time). Therefore, our package allows the user to
-model the effect of time using splines, and through interaction terms
-involving covariates and time, it also allows user to fit time-varying
-coefficient models. However, we do not explicitly model any shape
-parameter, unlike \pkg{flexsurv}.
-
-Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
-\pkg{mets}, \pkg{cmprsk} and \pkg{survival} contain implementations for
-competing risks. The differentiating factor between these packages and
-\pkg{casebase} is that \pkg{casebase} handles competing risks through
-multiple logistic regression. Note that \pkg{CFC} has extra features
-outside the scope of the metthods discussed in this paper.
-
-Finally, several packages include functions to estimate the cumulative
-incidence function. The corresponding methods generally fall into two
-categories: transformation of the estimated hazard function, and
-semi-parametric estimation of the baseline hazard. The first category
-broadly corresponds to parametric survival models, where the full hazard
-is explictly modeled. Using this estimate, the survival function and the
-cumulative incidence function can be obtained using their functional
-relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below).
-Packages including this functionality include \pkg{CFC}, \pkg{Flexsurv},
-\pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows
-this approach for both single-event and competing-event analyses. The
-second category outlined above broadly corresponds to Cox models. These
-models do not model the full hazard function, and therefore the baseline
-hazard needs to be estimated separately in order to estimate the
-survival function. This is achieved using semi-parametric estimators
-(e.g.~Breslow's estimator). Packages that implement this approach
-include \pkg{RiskRegression}, \pkg{rstpm2}, and \pkg{survival}. As
-mentioned in the introduction, a key distinguishing factor between these
-two approaches is that the first category leads to smooth estimates of
-the cumulative incidence function, whereas the second category produces
-estimates in the form of step-wise functions. This was one of the main
-motivations for introducting case-base sampling in survival analysis.
-
-\begin{table}[ht]
-\centering
-\resizebox{\textwidth}{!}{%
-\begin{tabular}{ccccccccc}
-\hline
-\textbf{Package}        & \vtop{\hbox{\strut \textbf{Competing}}\hbox{\strut \textbf{risks}}}  & \textbf{Non-proportional} & \textbf{Penalization} & \textbf{Splines} & \textbf{Parametric} & \textbf{Semi-parametric} & \vtop{\hbox{\strut \textbf{Interval/left}}\hbox{\strut \textbf{censoring}}} & \vtop{\hbox{\strut \textbf{Absolute}}\hbox{\strut \textbf{risk}}}           \\
-\hline
-\textbf{casebasee} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
-\textbf{CFC} \cite{mahani2015bayesian}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
-textbf{cmprsk} \cite{gray_2019}            & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\
-\textbf{coxRidge} \cite{perperoglou}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{crrp} \cite{fu}           & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
-\textbf{fastcox} \cite{yi_zou}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{Flexrsurv} \cite{flexrsurv}      &                          & x                         &                       & x                & x                   &                          &                                  & x               \\
-\textbf{Flexsurv} \cite{flexsurv}       & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\
-\textbf{glmnet} \cite{regpathcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{glmpath} \cite{park_hastie}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{mets} \cite{scheike2014estimating}           & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
-\textbf{penalized} \cite{goeman_meijer2019}      &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{RiskRegression} \cite{gerds_blanche} &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
-\textbf{Rstpm2} \cite{clements_liu}         &                          & x                         & x                     & x                & x                   & x                        & x                                & x                         \\
-\textbf{SmoothHazard} \cite{smoothHazard}   &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
-\textbf{Survival} \cite{survival-package}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x
-\end{tabular}%
-}
-\caption{Different features of interest in various survival packages.}
-\label{tab:surv-pkgs}
-\end{table}
-
-\begin{table}
-  \begin{tabular}{p{1.8in}lll}
-\hline
-    &  Parameters &  Density \proglang{R} function & \code{dist}\\
-    & {\footnotesize{(location in \code{\color{red}{red}})}} & & \\
-\hline
-    Exponential & \code{\color{red}{rate}}             & \code{dexp}   & \code{"exp"} \\
-    Weibull (accelerated failure time)     & \code{shape, {\color{red}{scale}}}     & \code{dweibull} & \code{"weibull"} \\
-    Weibull (proportional hazards)     & \code{shape, {\color{red}{scale}}}     & \code{dweibullPH} & \code{"weibullPH"} \\
-    Gamma       & \code{shape, \color{red}{rate}}      & \code{dgamma} & \code{"gamma"}\\
-    Log-normal  & \code{{\color{red}{meanlog}}, sdlog}   & \code{dlnorm} & \code{"lnorm"}\\
-    Gompertz    & \code{shape, {\color{red}{rate}}}      & \code{dgompertz} & \code{"gompertz"} \\
-    Log-logistic & \code{shape, {\color{red}{scale}}}   & \code{dllogis} & \code{"llogis"}\\
-    Generalized gamma (Prentice 1975)   & \code{{\color{red}{mu}}, sigma, Q} & \code{dgengamma} & \code{"gengamma"} \\
-    Generalized gamma (Stacy 1962)& \code{shape, {\color{red}{scale}}, k} & \code{dgengamma.orig} & \code{"gengamma.orig"} \\
-    Generalized F     (stable)    & \code{{\color{red}{mu}}, sigma, Q, P} & \code{dgenf} & \code{"genf"} \\
-    Generalized F     (original)  & \code{{\color{red}{mu}}, sigma, s1, s2} & \code{dgenf.orig} & \code{"genf.orig"} \\
-\hline
-  \end{tabular}
-  \caption{Built-in parametric survival distributions in \pkg{flexsurv}.}
-  \label{tab:dists}
-\end{table}
-
-\hypertarget{implementation-details}{%
-\section{Implementation details}\label{implementation-details}}
-
-The functions in the casebase package can be divided into two
-categories: 1) data visualization, in the form of population-time plots;
-and 2) parametric modeling. We explicitly aimed at being compatible with
-both \code{data.frame}s and \code{data.table}s. This is evident in some
-of the coding choices we made, and it is also reflected in our unit
-tests.
-
-\hypertarget{population-time-plots}{%
-\subsection{Population-time plots}\label{population-time-plots}}
-
-\hypertarget{parametric-modeling}{%
-\subsection{Parametric modeling}\label{parametric-modeling}}
-
-The parametric modeling step was separated into three parts:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  case-base sampling;
-\item
-  estimation of the smooth hazard function;
-\item
-  calculation of the risk function.
-\end{enumerate}
-
-By separating the sampling and estimation functions, we allowed the
-possibility of users implementing more complex sampling scheme, as
-described in Saarela \citeyearpar{saarela2016case}.
-
-The sampling scheme selected for \code{sampleCaseBase} was described in
-Hanley and Miettinen \citeyearpar{hanley2009fitting}: we first sample
-along the ``person'' axis, proportional to each individual's total
-follow-up time, and then we sample a moment uniformly over their
-follow-up time. This sampling scheme is equivalent to the following
-picture: imagine representing the total follow-up time of all
-individuals in the study along a single dimension, where the follow-up
-time of the next individual would start exactly when the follow-up time
-of the previous individual ends. Then the base series could be sampled
-uniformly from this one-dimensional representation of the overall
-follow-up time. In any case, the output is a dataset of the same class
-as the input, where each row corresponds to a person-moment. The
-covariate profile for each such person-moment is retained, and an offset
-term is added to the dataset. This output could then be used to fit a
-smooth hazard function, or for visualization of the base series.
-
-The fitting function \code{fitSmoothHazard} starts by looking at the
-class of the dataset: if it was generated from \code{sampleCaseBase}, it
-automatically inherited the class \code{cbData}. If the dataset supplied
-to \code{fitSmoothHazard} does not inherit from \code{cbData}, then the
-fitting function starts by calling \code{sampleCaseBase} to generate the
-base series. In other words, the occasional user can bypass
-\code{sampleCaseBase} altogether and only worry about the fitting
-function \code{fitSmoothHazard}.
-
-The fitting function retains the familiar formula interface of
-\code{glm}. The left-hand side of the formula should be the name of the
-column corresponding to the event type. The right-hand side can be any
-combination of the covariates, along with an explicit functional form
-for the time variable. Note that non-proportional hazard models can be
-achieved at this stage by adding an interaction term involving time. The
-offset term does not need to be specified by the user, as it is
-automatically added to the formula.
-
-To fit the hazard function, we provide several approaches that are
-available via the \code{family} parameter. These approaches are:
-
-\begin{itemize}
-\tightlist
-\item
-  \code{glm}: This is the familiar logistic regression.
-\item
-  \code{glmnet}: This option allows for variable selection using Lasso
-  or elastic-net. This functionality is provided through the
-  \code{glmnet} package \citep{friedman2010jss}.
-\item
-  \code{gam}: This option provides support for \emph{Generalized
-  Additive Models} via the \code{gam} package
-  \citep{hastie1987generalized}.
-\item
-  \code{gbm}: This option provides support for \emph{Gradient Boosted
-  Trees} via the \code{gbm} package. This feature is still experimental.
-\end{itemize}
-
-In the case of multiple events, the hazard is fitted via multinomial
-regression as performed by the \pkg{VGAM} package. This package was
-selected for its ability to fit multinomial regression models with an
-offset.
-
-Once a model-fit object has been returned by \code{fitSmoothHazard}, all
-the familiar summary and diagnostic functions are available:
-\code{print}, \code{summary}, \code{predict}, \code{plot}, etc. Our
-package provides one more functionality: it computes risk functions from
-the model fit. For the case of a single event, it uses the familiar
-identity \begin{equation}\label{eqn:surv}
-S(t) = \exp\left(-\int_0^t h(u;X) du\right).
-\end{equation} The integral is computed using either the
-\code{stats::integrate} function or Monte-Carlo integration. The risk
-function (or cumulative incidence function) is then defined as
-\begin{equation}\label{eqn:CI}
-CI(t) = 1 - S(t).
-\end{equation}
-
-For the case of a competing-event analysis, the event-specific risk is
-computed using the following procedure: first, we compute the overall
-survival function (i.e.~for all event types):
-
-\[ S(t) = \exp\left(-\int_0^t H(u;X) du\right),\qquad H(t;X) = \sum_{j=1}^J h_j(t;X).\]
-From this, we can derive the event-specific subdensities:
-
-\[ f_j(t) = h_j(t)S(t).\]
-
-Finally, by integrating these subdensities, we obtain the event-specific
-cumulative incidence functions:
-
-\[ CI_j(t) = \int_0^t f_j(u)du.\]
-
-We created \code{absoluteRisk} as an \code{S3} generic, with methods for
-the different types of outputs of \code{fitSmoothHazard}. The method
-dispatch system of \proglang{R} then takes care of matching the correct
-output to the correct methodology for calculating the cumulative
-incidence function, without the user's intervention.
-
-In the following sections, we illustrate these functionalities in the
-context of three case studies.
-
-\hypertarget{case-study-1european-randomized-study-of-prostate-cancer-screening}{%
-\section{Case study 1--European Randomized Study of Prostate Cancer
-Screening}\label{case-study-1european-randomized-study-of-prostate-cancer-screening}}
-
-For the first case study, we will use of the European Randomized Study
-of Prostate Cancer Screening data. This dataset is available through the
-\pkg{casebase} package:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> data(ERSPC)
-R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
-R>                        levels = c(0, 1), 
-R>                        labels = c("Control group", "Screening group"))
-\end{CodeInput}
-\end{CodeChunk}
-
-The results of this study were published by
-\citep{schroder2009screening}, and the dataset itself was obtained using
-the approach described in \citep{liu2014recovering}.
-
-Population time plots can be extremely informative graphical displays of
-survival data. They should be the first step in an exploratory data
-analysis. We facilitate this task in the \pkg{casebase} package using
-the \code{popTime} function. We first create the necessary dataset for
-producing the population time plots, and we can generate the plot by
-using the corresponding \code{plot} method:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
-R> plot(pt_object)
-\end{CodeInput}
-\end{CodeChunk}
-
-We can also create exposure stratified plots by specifying the
-\code{exposure} argument in the \code{popTime} function:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> pt_object_strat <- casebase::popTime(ERSPC, 
-R>                                      event = "DeadOfPrCa", 
-R>                                      exposure = "ScrArm")
-R> plot(pt_object_strat)
-\end{CodeInput}
-\end{CodeChunk}
-
-We can also plot them side-by-side using the \code{ncol} argument:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> plot(pt_object_strat, ncol = 2)
-\end{CodeInput}
-\end{CodeChunk}
-
-ADD PARAGRAPH ABOUT WHAT WE CAN CONCLUDE FROM THESE GRAPHS.
-
-Next, we investigate the differences between the control and the
-screening arms. A common choice for this type of analysis is to use a
-Cox regression model and estimate the hazard ratio for the screening
-group (relative to the control group). In \proglang{R}, it can be done
-as follows:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
-R>                              data = ERSPC)
-R> summary(cox_model)
-\end{CodeInput}
-\end{CodeChunk}
-
-We can also plot the cumulative incidence function (CIF) for each group.
-However, this involves estimating the baseline hazard, which Cox
-regression treated as a nuisance parameter.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
-R>                        ignore = 99)
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R>      xlab = "Years since Randomization", 
-R>      ylab = "Cumulative Incidence (%)", 
-R>      fun = "event",
-R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>                     exp(coef(cox_model)), 
-R>                     exp(confint(cox_model))[1], 
-R>                     exp(confint(cox_model))[2]),
-R>      ylim = c(0, 2), yscale = 100)
-R> legend("topleft", 
-R>        legend = c("Control group", "Screening group"), 
-R>        col = c("red","blue"),
-R>        lty = c(1, 1), 
-R>        bg = "gray90")
-\end{CodeInput}
-\end{CodeChunk}
-
-Next, we show how case-base sampling can be used to carry out a similar
-analysis. We fit several models that differ in how we choose to model
-time.
-
-First, we will fit an exponential model. Recall that this corresponds to
-excluding time from the linear predictor.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
-R>                                                   data = ERSPC, 
-R>                                                   ratio = 100)
-R> 
-R> summary(casebase_exponential)
-R> exp(coef(casebase_exponential)[2])
-R> exp(confint(casebase_exponential)[2,])
-\end{CodeInput}
-\end{CodeChunk}
-
-We can then use the \code{absoluteRisk} function to get an estimate of
-the cumulative incidence curve for a specific covariate profile. In the
-plot below, we overlay the estimated CIF from the exponential model on
-the Cox model CIF:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
-R>                                           time = seq(0,15,0.1), 
-R>                                           newdata = new_data)
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R>      xlab = "Years since Randomization", 
-R>      ylab = "Cumulative Incidence (%)", 
-R>      fun = "event",
-R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>                     exp(coef(cox_model)), 
-R>                     exp(confint(cox_model))[1], 
-R>                     exp(confint(cox_model))[2]),
-R>      ylim = c(0, 2), yscale = 100)
-R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
-R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
-R> 
-R> 
-R> legend("topleft", 
-R>        legend = c("Control group (Cox)","Control group (Casebase)",
-R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R>        col = c("red","red", "blue","blue"),
-R>        lty = c(1, 2, 1, 2), 
-R>        bg = "gray90")
-\end{CodeInput}
-\end{CodeChunk}
-
-As we can see, the exponential model gives us an estimate of the hazard
-ratio that is similar to the one obtained using Cox regression. However,
-the CIF estimates are quite different. Based on what we observed in the
-population time plot, where more events are observed later on in time,
-we do not expect a constant hazard would be a good description for this
-data. In other words, this poor fit for the exponential hazard is
-expected. A constant hazard model would overestimate the cumulative
-incidence earlier on in time, and underestimate it later on; this is
-what we see on the cumulative incidence plot. This example demonstrates
-the benefits of population time plots as an exploratory analysis tool.
-
-For the next model, we include time as a linear term. Recall that this
-corresponds to a Weibull model.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
-R>                                  data = ERSPC, 
-R>                                  ratio = 100)
-R> 
-R> summary(casebase_time)
-R> exp(coef(casebase_time))
-R> exp(confint(casebase_time))
-\end{CodeInput}
-\end{CodeChunk}
-
-Again, the estimate of the hazard ratio is similar to that obtained from
-Cox regression. We then look at the estimate of the CIF:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
-R>                                           time = seq(0,15,0.1), 
-R>                                           newdata = new_data)
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R>      xlab = "Years since Randomization", 
-R>      ylab = "Cumulative Incidence (%)", 
-R>      fun = "event",
-R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>                     exp(coef(cox_model)), 
-R>                     exp(confint(cox_model))[1], 
-R>                     exp(confint(cox_model))[2]),
-R>      ylim = c(0, 2), yscale = 100)
-R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
-R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
-R> 
-R> legend("topleft", 
-R>        legend = c("Control group (Cox)","Control group (Casebase)",
-R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R>        col = c("red","red", "blue","blue"),
-R>        lty = c(1, 2, 1, 2), 
-R>        bg = "gray90")
-\end{CodeInput}
-\end{CodeChunk}
-
-We see that the Weibull model leads to a better fit of the CIF than the
-Exponential model, when compared with the semi-parametric approach.
-
-Finally, we can model the hazard as a smooth function of time using the
-\code{splines} package:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
-R>                                     data = ERSPC, 
-R>                                     ratio = 100)
-R> 
-R> summary(casebase_splines)
-R> exp(coef(casebase_splines))
-R> exp(confint(casebase_splines))
-\end{CodeInput}
-\end{CodeChunk}
-
-Once again, we can see that the estimate of the hazard ratio is similar
-to that from the Cox regression. We then look at the estimate of the
-CIF:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
-R>                                     time = seq(0,15,0.1), 
-R>                                     newdata = new_data)
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R>      xlab = "Years since Randomization", 
-R>      ylab = "Cumulative Incidence (%)", 
-R>      fun = "event",
-R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>                     exp(coef(cox_model)), 
-R>                     exp(confint(cox_model))[1], 
-R>                     exp(confint(cox_model))[2]),
-R>      ylim = c(0, 2), yscale = 100)
-R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
-R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
-R> 
-R> legend("topleft", 
-R>        legend = c("Control group (Cox)","Control group (Casebase)",
-R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R>        col = c("red","red", "blue","blue"),
-R>        lty = c(1, 2, 1, 2), 
-R>        bg = "gray90")
-\end{CodeInput}
-\end{CodeChunk}
-
-Qualitatively, the combination of splines and case-base sampling
-produces the parametric model that most closely resembles the Cox model.
-In the following table, we see that the confidence intervals are also
-similar across all four models. In other words, this reinforces the idea
-that, under proportional hazards, we do not need to model the full
-hazard to obtain reliable estimates of the hazard ratio. Of course,
-different parametric models for the hazards give rise to qualitatively
-different estimates for the CIF.
-
-As noted above, the usual asymptotic results hold for likelihood ratio
-tests built using case-base sampling models. Therefore, we can easily
-test the null hypothesis that the exponential model is just as good as
-the larger (in terms of number of parameters) splines model.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> anova(casebase_exponential, casebase_splines, test = "LRT")
-\end{CodeInput}
-\end{CodeChunk}
-
-As expected, we see that splines model provides a better fit. Similarly,
-we can compare the AIC for all three case-base models:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> c("Exp." = AIC(casebase_exponential),
-R>   "Weib." = AIC(casebase_time),
-R>   "Splines" = AIC(casebase_splines))
-\end{CodeInput}
-\end{CodeChunk}
-
-Once again, we have evidence that the splines model provides the best
-fit.
-
-With this first case study, we saw how population-time plots can provide
-a useful summary of the incidence for our event of interests. In
-particular, we saw how it can provide evidence against an exponential
-hazard model. We then explored how different parametric models can
-provide different estimates of the CIF, even though they all give
-similar estimates for the hazard ratios.
-
-\hypertarget{case-study-2bone-marrow-transplant}{%
-\section{Case study 2--Bone-marrow
-transplant}\label{case-study-2bone-marrow-transplant}}
-
-In the next case study, we show how case-base sampling can also be used
-in the context of a competing risk analysis. For illustrative purposes,
-we will use the same data that was used in Scrucca \emph{et al}
-\citeyearpar{scrucca2010regression}. The data was downloaded from the
-first author's website, and it is also available as part of the
-\pkg{casebase} package.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> 
-R> data(bmtcrr)
-\end{CodeInput}
-\end{CodeChunk}
-
-The data contains information on 177 patients who received a stem-cell
-transplant for acute leukemia. The event of interest is relapse, but
-other competing causes (e.g.~transplant-related death) were also
-recorded. Several covariates were captured at baseline: sex, disease
-type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL
-and AML, respectively), disease phase at transplant (Relapse, CR1, CR2,
-CR3), source of stem cells (bone marrow and peripheral blood, coded as
-BM+PB, or only peripheral blood, coded as PB), and age. A summary of
-these baseline characteristics appear in Table \ref{tab:table1bmtcrr}.
-We note that the statistical summaries were generated differently for
-different variable types: for continuous variables, we gave the range,
-followed by the mean and standard deviation; for categorical variables,
-we gave the counts for each category.
-
-First, we can look at a population-time plot to visualize the incidence
-density of both relapse and the competing events. In Figure
-\ref{fig:compPop}, failure times are highlighted on the plot using red
-points for the event of interest (panel A) and blue dots for competing
-events (panel B). In both panels, we see evidence of a non-constant
-hazard function: the density of points is larger at the beginning of
-follow-up than at the end.
-
-Our main objective is to compute the absolute risk of relapse for a
-given set of covariates. First, we fit a smooth hazard to the data using
-a linear term for time:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> model_cb <- fitSmoothHazard(
-R>     Status ~ ftime + Sex + D + Phase + Source + Age, 
-R>     data = bmtcrr, 
-R>     ratio = 100, 
-R>     time = "ftime")
-\end{CodeInput}
-\end{CodeChunk}
-
-We will compare our hazard ratio estimates to that obtained from a Cox
-regression. To do so, we need to treat the competing event as censoring.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(survival)
-R> # Treat competing event as censoring
-R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-R>                    data = bmtcrr)
-\end{CodeInput}
-\end{CodeChunk}
-
-From the fit object, we can extract both the hazard ratios and their
-corresponding confidence intervals. These quantities appear in Table
-\ref{tab:bmtcrr-cis}. As we can see, the only significant hazard ratio
-identified by case-base sampling is the one associated with the phase of
-the disease at transplant. More precisely, being in relapse at
-transplant is associated with a hazard ratio of 3.89 when compared to
-CR1.
-
-Given the estimate of the hazard function obtained using case-base
-sampling, we can compute the absolute risk curve for a fixed covariate
-profile. We perform this computation for a 35 year old woman who
-received a stem-cell transplant from peripheral blood at relapse. We
-compared the absolute risk curve for such a woman with acute
-lymphoblastic leukemia with that for a similar woman with acute
-myeloblastic leukemia. We will estimate the curve from 0 to 60 months.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Pick 100 equidistant points between 0 and 60 months
-R> time_points <- seq(0, 60, length.out = 50)
-R> 
-R> # Data.frame containing risk profile
-R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
-R>                                      levels = levels(bmtcrr[,"Sex"])),
-R>                       "D" = c("ALL", "AML"), # Both diseases
-R>                       "Phase" = factor(c("Relapse", "Relapse"), 
-R>                                        levels = levels(bmtcrr[,"Phase"])),
-R>                       "Age" = c(35, 35),
-R>                       "Source" = factor(c("PB", "PB"), 
-R>                                         levels = levels(bmtcrr[,"Source"])))
-R> 
-R> # Estimate absolute risk curve
-R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
-R>                         method = "numerical", newdata = newdata)
-\end{CodeInput}
-\end{CodeChunk}
-
-We will compare our estimates to that obtained from a corresponding
-Fine-Gray model \citep{fine1999proportional}. The Fine-Gray model is a
-semiparametric model for the cause-specific \emph{subdistribution},
-i.e.~the function \(f_k(t)\) such that
-\[CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),\] where
-\(CI_k(t)\) is the cause-specific cumulative indicence. The Fine-Gray
-model allows to directly assess the effect of a covariate on the
-subdistribution, as opposed to the hazard. For the computation, we will
-use the \pkg{timereg} package \citep{timereg}.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(timereg)
-R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) + 
-R>                           const(Phase) + const(Source) + const(Age), 
-R>                       data = bmtcrr, cause = 1, model = "fg")
-R> 
-R> # Estimate absolute risk curve
-R> risk_fg <- predict(model_fg, newdata, times = time_points)
-\end{CodeInput}
-\end{CodeChunk}
-
-Figure \ref{fig:compAbsrisk} shows the absolute risk curves for both
-case-base sampling and the Fine-Gray model. As we can see, the two
-approaches agree quite well for acute myeloblastic leukemia; however,
-there seems to be a difference of about 5\% between the two curves for
-acute lymphoblastic leukemia. This difference does not appear to be
-significant: the curve from case-base sampling is contained within a
-95\% confidence band around the Fine-Gray absolute risk curve (figure
-not shown).
 
@@
 R> 
-R> cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
+R> cb.ModelInter=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
 R+                                        formula_time = ~log(d.time),alpha=1,ratio=ratio)
@@
 
-We used the survival package to calculate cox regression's cumulative
-incidience without penalization and the kaplan-meier curve for
-incidience.
+We used the survival package to calculate the kaplan-meier absolute risk
+function.
 
@@
 \begin{CodeInput}
-R> 
-R> coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
-R> abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)
-R> 
 R> #creating a surv object to be used to fit an unadjusted absolute risk curve.
@@
 
-Now that our four cumulative incidence estimations have been calculated,
-we compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase
+Now that our three absolute risk functions have been calculated, we
+compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase
 decrease the absolute risk by the end of the study, in comparison to
@@
 \begin{CodeChunk}
+
+\begin{CodeInput}
+R> # A plot to compare all three Absolute risk (Commulative Incidence)
+R> plot(cbCIAve[,1],cbCIAve[,2],type='l',col="black",lwd=3,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
+R> lines(abCF,col="red",fun="event",lwd=3,conf.int = FALSE)
+R> lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
+R> legend("bottomright", 
+R+        legend = c( "Casebase (Lasso+linear)","semi-parametric (Cox)+glmnet","KM curve"), 
+R+        col = c("black","red","Blue"),
+R+        lty = c(1, 1, 1), 
+R+        bg = "gray90")
+\end{CodeInput}
 \begin{figure}
@@
 R> colnames(estimatescb)<-c("Coefs","Estimate","Model")
-R> estimatescox=setDT(as.data.frame(coef(coxglmFit)[-c(1),1]), keep.rownames = TRUE)[]
+R> estimatescox=setDT(as.data.frame(coef(coxNet)[-c(1),1]), keep.rownames = TRUE)[]
 R> estimatescox$Model="Cox"
@@
 
-\hypertarget{case-study-4stanford-heart-transplant-data}{%
-\section{Case study 4--Stanford Heart Transplant
-Data}\label{case-study-4stanford-heart-transplant-data}}
-
-Although the previous case studies provide a broad overview of the
-capabilities of \pkg{casebase}, they all have two properties in common:
-
-\begin{itemize}
-  \item All covariates are fixed, i.e. they do not change over time;
-  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
-\end{itemize}
-
-Case-base sampling has been used in the literature to study vaccination
-safety, where the exposure period was defined as the week following
-vaccination \citep{saarela2015case}. Hence, the main covariate of
-interest, i.e.~exposure to the vaccine, was changing over time. In this
-context, case-base sampling offers an efficient alternative to nested
-case-control designs or self-matching.
-
-The purpose of the next case study is to show how \pkg{casebase} can
-also be used for survival analysis problems that do not share the
-properties above. To this end, we will use the Stanford Heart Transplant
-Data \citep[\citet{crowley1977covariance}]{clark1971cardiac}; we will
-use the version available in the \pkg{survival} package.
-
-Recall the setting of this study: patients were admitted to the Stanford
-program after meeting with their physician and determining that they
-were unlikely to respond to other forms of treatment. After enrollment,
-the program searches for a suitable donor for the patient, which can
-take anywhere between a few days to almost a year. We are interested in
-the effect of a heart transplant on survival; therefore, the patient is
-considered exposed only after the transplant has occurred.
-
-As before, we can look at the population-time plot for a graphical
-summary of the event incidence. As we can see, most events occur early
-during the follow-up period, and therefore we do not expect the hazard
-to be constant.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(survival)
-R> library(casebase)
-R> 
-R> stanford_popTime <- popTime(jasa, time = "futime", 
-R>                             event = "fustat")
-R> plot(stanford_popTime)
-\end{CodeInput}
-\end{CodeChunk}
-
-Since the exposure is time-dependent, we need to manually define the
-exposure variable \emph{after} case-base sampling and \emph{before}
-fitting the hazard function. For this reason, we will use the
-\texttt{sampleCaseBase} function directly.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(tidyverse)
-R> library(lubridate)
-R> 
-R> cb_data <- sampleCaseBase(jasa, time = "futime", 
-R>                           event = "fustat", ratio = 10)
-\end{CodeInput}
-\end{CodeChunk}
-
-Next, we will compute the number of days from acceptance into the
-program to transplant, and we use this variable to determine whether
-each population-moment is exposed or not.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Define exposure variable
-R> cb_data <- mutate(cb_data,
-R>                   txtime = time_length(accept.dt %--% tx.date, 
-R>                                        unit = "days"),
-R>                   exposure = case_when(
-R>                     is.na(txtime) ~ 0L,
-R>                     txtime > futime ~ 0L,
-R>                     txtime <= futime ~ 1L
-R>                   ))
-\end{CodeInput}
-\end{CodeChunk}
-
-Finally, we can fit the hazard using various linear predictors.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(splines)
-R> # Fit several models
-R> fit1 <- fitSmoothHazard(fustat ~ exposure,
-R>                         data = cb_data, time = "futime")
-R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
-R>                         data = cb_data, time = "futime")
-R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
-R>                         data = cb_data, time = "futime")
-R> fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
-R>                         data = cb_data, time = "futime")
-\end{CodeInput}
-\end{CodeChunk}
-
-Note that the fourth model (i.e. \texttt{fit4}) includes an interaction
-term between exposure and follow-up time. In other words, this model no
-longer exhibit proportional hazards. The evidence of non-proportionality
-of hazards in the Stanford Heart Transplant data has been widely
-discussed \citep{arjas1988graphical}.
-
-We can then compare the goodness of fit of these four models using the
-Akaike Information Criterion (AIC).
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute AIC
-R> c("Model1" = AIC(fit1),
-R>   "Model2" = AIC(fit2),
-R>   "Model3" = AIC(fit3),
-R>   "Model4" = AIC(fit4))
-\end{CodeInput}
-\end{CodeChunk}
-
-As we can, the best fit is the fourth model. By visualizing the hazard
-functions for both exposed and unexposed individuals, we can more
-clearly see how the hazards are no longer proportional.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute hazards---
-R> # First, create a list of time points for both exposure status
-R> hazard_data <- expand.grid(exposure = c(0, 1),
-R>                            futime = seq(0, 1000,
-R>                                         length.out = 100))
-R> # Set the offset to zero
-R> hazard_data$offset <- 0 
-R> # Use predict to get the fitted values, and exponentiate to 
-R> # transform to the right scale
-R> hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
-R>                                  type = "link"))
-R> # Add labels for plots
-R> hazard_data$Status = factor(hazard_data$exposure,
-R>                             labels = c("NoTrans", "Trans"))
-R> 
-R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
-R>     geom_line() +
-R>     theme_minimal() +
-R>     theme(legend.position = 'top') +
-R>     ylab('Hazard') + xlab('Follow-up time')
-\end{CodeInput}
-\end{CodeChunk}
-
-The non-proportionality seems to be more pronounced at the beginning of
-follow-up than the end. Finally, we can turn these estimate of the
-hazard function into estimates of the cumulative incidence functions.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute absolute risk curves
-R> newdata <- data.frame(exposure = c(0, 1))
-R> absrisk <- absoluteRisk(fit4, newdata = newdata, 
-R>                         time = seq(0, 1000, length.out = 100))
-R> 
-R> colnames(absrisk) <- c("Time", "NoTrans", "Trans")
-R> 
-R> # Rearrange the data
-R> absrisk <- gather(as.data.frame(absrisk),
-R>                   "Status", "Risk", -Time)
-R>  
-R> ggplot(absrisk, aes(Time, Risk, colour = Status)) +
-R>   geom_line() +
-R>   theme_minimal() +
-R>   theme(legend.position = 'top') +
-R>   expand_limits(y = 0.5) +
-R>   xlab('Follow-up time') + ylab('Cum. Incidence')
-\end{CodeInput}
-\end{CodeChunk}
-
-Note that we can easily adapt the code above to the situation where a
-patient receives a heart transplant at a point in time of interest, for
-example after 30 days.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute hazards---
-R> # First, create a list of time points for both exposure status
-R> one_yr_haz <- data.frame(futime = seq(0, 365,
-R>                                       length.out = 100))
-R> one_yr_haz <- mutate(one_yr_haz,
-R>                      offset = 0,
-R>                      exposure = if_else(futime < 30, 0, 1))
-R> one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
-R>                                 type = "link"))
-R> ggplot(one_yr_haz, aes(futime, hazard)) +
-R>   geom_line() +
-R>   theme_minimal() +
-R>   theme(legend.position = 'top') +
-R>   ylab('Hazard') + xlab('Follow-up time') +
-R>   geom_vline(xintercept = 30, linetype = 'dashed')
-\end{CodeInput}
-\end{CodeChunk}
-
-We can then compare the 1-year mortality risk without transplant and
-with transplant at 30 days.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> absoluteRisk(fit4, newdata = data.frame(exposure = 0),
-R>              time = 365)
-R> 
-R> # Use the trapezoidal rule to estimate the cumulative hazard
-R> min_hazard <- min(one_yr_haz$hazard)
-R> max_hazard <- max(one_yr_haz$hazard)
-R> increment <- 365/99
-R> cumhaz_est <- 0.5*increment*(sum(2*one_yr_haz$hazard) - min_hazard - max_hazard)
-R> 1 - exp(-cumhaz_est)
-\end{CodeInput}
-\end{CodeChunk}
-
-As we can see, the risk estimate at 1-year is about 30\% lower if the
-patient receives a heart transplant at 30 days.
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-In this article, we presented the \proglang{R} package \pkg{casebase}
-that provides functions to fit smooth parametric hazards and estimate
-cumulative incidence functions using case-base sampling. We outlined the
-theoretical underpinnings of the approach, we provided details about our
-implementation, and we illustrated the merits of the approach and the
-package through three case studies.
-
-\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval
-censoring, whereas \pkg{casebase} currently cannot. Interval censoring
-is a simple feature to implement and is a potential improvement to add
-to \pkg{casebase}.
-
-\hypertarget{environment-details}{%
-\section{Environment Details}\label{environment-details}}
-
-This report was generated on 2020-02-06 17:40:56 using the following
-computational environment and dependencies:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # which R packages and versions?
-R> devtools::session_info()
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> - Session info ---------------------------------------------------------------
-#>  setting  value                       
-#>  version  R version 3.6.0 (2019-04-26)
-#>  os       macOS High Sierra 10.13.6   
-#>  system   x86_64, darwin15.6.0        
-#>  ui       X11                         
-#>  language (EN)                        
-#>  collate  en_CA.UTF-8                 
-#>  ctype    en_CA.UTF-8                 
-#>  tz       America/Montreal            
-#>  date     2020-02-06                  
-#> 
-#> - Packages -------------------------------------------------------------------
-#>  package     * version    date       lib
-#>  assertthat    0.2.1      2019-03-21 [1]
-#>  backports     1.1.5      2019-10-02 [1]
-#>  broom         0.5.2      2019-04-07 [1]
-#>  callr         3.4.1      2020-01-24 [1]
-#>  casebase    * 0.2.1.9001 2020-01-30 [1]
-#>  cellranger    1.1.0      2016-07-27 [1]
-#>  cli           2.0.1      2020-01-08 [1]
-#>  codetools     0.2-16     2018-12-24 [1]
-#>  colorspace    1.4-1      2019-03-18 [1]
-#>  crayon        1.3.4      2017-09-16 [1]
-#>  data.table  * 1.12.8     2019-12-09 [1]
-#>  DBI           1.0.0      2018-05-02 [1]
-#>  dbplyr        1.4.2      2019-06-17 [1]
-#>  desc          1.2.0      2018-05-01 [1]
-#>  devtools      2.2.1      2019-09-24 [1]
-#>  digest        0.6.23     2019-11-23 [1]
-#>  dplyr       * 0.8.3      2019-07-04 [1]
-#>  ellipsis      0.3.0      2019-09-20 [1]
-#>  evaluate      0.14       2019-05-28 [1]
-#>  fansi         0.4.1      2020-01-08 [1]
-#>  farver        2.0.3      2020-01-16 [1]
-#>  forcats     * 0.4.0      2019-02-17 [1]
-#>  foreach       1.4.7      2019-07-27 [1]
-#>  fs            1.3.1      2019-05-06 [1]
-#>  generics      0.0.2      2018-11-29 [1]
-#>  ggplot2     * 3.2.1      2019-08-10 [1]
-#>  glmnet      * 3.0-2      2019-12-11 [1]
-#>  glue          1.3.1      2019-03-12 [1]
-#>  gtable        0.3.0      2019-03-25 [1]
-#>  haven         2.2.0      2019-11-08 [1]
-#>  hms           0.5.2      2019-10-30 [1]
-#>  htmltools     0.4.0      2019-10-04 [1]
-#>  httr          1.4.1      2019-08-05 [1]
-#>  iterators     1.0.12     2019-07-26 [1]
-#>  jsonlite      1.6        2018-12-07 [1]
-#>  kableExtra  * 1.1.0      2019-03-16 [1]
-#>  knitr         1.27       2020-01-16 [1]
-#>  labeling      0.3        2014-08-23 [1]
-#>  lattice       0.20-38    2018-11-04 [1]
-#>  lazyeval      0.2.2      2019-03-15 [1]
-#>  lifecycle     0.1.0      2019-08-01 [1]
-#>  lubridate     1.7.4      2018-04-11 [1]
-#>  magrittr    * 1.5        2014-11-22 [1]
-#>  Matrix      * 1.2-17     2019-03-22 [1]
-#>  memoise       1.1.0      2017-04-21 [1]
-#>  mgcv          1.8-31     2019-11-09 [1]
-#>  modelr        0.1.5      2019-08-08 [1]
-#>  munsell       0.5.0      2018-06-12 [1]
-#>  nlme          3.1-142    2019-11-07 [1]
-#>  pillar        1.4.3      2019-12-20 [1]
-#>  pkgbuild      1.0.6      2019-10-09 [1]
-#>  pkgconfig     2.0.3      2019-09-22 [1]
-#>  pkgload       1.0.2      2018-10-29 [1]
-#>  prettyunits   1.1.1      2020-01-24 [1]
-#>  processx      3.4.1      2019-07-18 [1]
-#>  ps            1.3.0      2018-12-21 [1]
-#>  purrr       * 0.3.3      2019-10-18 [1]
-#>  R6            2.4.1      2019-11-12 [1]
-#>  Rcpp          1.0.3      2019-11-08 [1]
-#>  readr       * 1.3.1      2018-12-21 [1]
-#>  readxl        1.3.1      2019-03-13 [1]
-#>  remotes       2.1.0      2019-06-24 [1]
-#>  reprex        0.3.0      2019-05-16 [1]
-#>  rlang         0.4.4      2020-01-28 [1]
-#>  rmarkdown     2.1        2020-01-20 [1]
-#>  rprojroot     1.3-2      2018-01-03 [1]
-#>  rstudioapi    0.10       2019-03-19 [1]
-#>  rticles       0.12       2019-11-12 [1]
-#>  rvest         0.3.5      2019-11-08 [1]
-#>  scales        1.1.0      2019-11-18 [1]
-#>  sessioninfo   1.1.1      2018-11-05 [1]
-#>  shape         1.4.4      2018-02-07 [1]
-#>  stringi       1.4.5      2020-01-11 [1]
-#>  stringr     * 1.4.0      2019-02-10 [1]
-#>  survival    * 3.1-7      2019-11-09 [1]
-#>  testthat      2.3.1      2019-12-01 [1]
-#>  tibble      * 2.1.3      2019-06-06 [1]
-#>  tidyr       * 1.0.0      2019-09-11 [1]
-#>  tidyselect    0.2.5      2018-10-11 [1]
-#>  tidyverse   * 1.3.0      2019-11-21 [1]
-#>  usethis       1.5.1      2019-07-04 [1]
-#>  vctrs         0.2.2      2020-01-24 [1]
-#>  VGAM          1.1-2      2019-11-21 [1]
-#>  viridisLite   0.3.0      2018-02-01 [1]
-#>  webshot       0.5.2      2019-11-22 [1]
-#>  withr         2.1.2      2018-03-15 [1]
-#>  xfun          0.12       2020-01-13 [1]
-#>  xml2          1.2.2      2019-08-09 [1]
-#>  yaml          2.2.0      2018-07-25 [1]
-#>  source                                  
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  Github (sahirbhatnagar/casebase@24359be)
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#> 
-#> [1] /Library/Frameworks/R.framework/Versions/3.6/Resources/library
-\end{CodeOutput}
-\end{CodeChunk}
-
-The current Git commit details are:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # what commit is this file at? 
-R> git2r::repository(here::here())
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> Local:    master /Users/jislam/Desktop/Gits/cbpaper
-#> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
-#> Head:     [a04834a] 2020-01-31: cbPaper draft for covars and interactions
-\end{CodeOutput}
-\end{CodeChunk}
-
 \bibliography{references.bib}

### 2020-02-06 analysis/paper/paper.Rmd

@@
 
-For the third case study, we will introduce regularization into the estimation of the hazard function.
+For the third case study, we will look at regularized estimation for the hazard function.
 This can be performed using a combination of case-base sampling and regularized logistic regression.
@@
 Before imputation, there were a total of 1000 individuals and 35 variables.
-charges, Totcst, totmcst and sfdm2 were removed as they are response variables we are not interested in for the sake of this analysis.
 After imputation, we have 690 individuals.
-We lose individuals due to coliniarity between variables, preventing imputation.
+We lose individuals due to coliniarity.
 If a pair of covariates are deemed colinear to a certain degree, they are excluded from any imputations.
 Here, only edu  income  avtisst*  race  ph*  glucose*  adlp* were imputed.
-Of these 690 individuals, 66.09% experienced the event of interest.
+Of these, 66.09% experienced the event of interest.
 <!--The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.--> A description of each variable can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}.
@@
 
-```{r support-poptime, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data.
-Red points represent the case series.
-The gray space demonstrates the base series."}
+```{r support-poptime, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data."}
 data(support)
@@
 completeData<-completeData[,-c(4)]
-ratio=100  #universal ratio for eval_cs3
+rat=1  #universal ratio for eval_cs3
 #keep one individual out, at random, for which we will develop an 
@@
 Our main objective is to compute the absolute risk of death for a given set of covariates, using regularization when fitting our model.
-First, we fit a smooth hazard to the data using a weibull distribution.
+First, we fit a smooth hazard to the data; for the sake of this example, we opted for a linear term for time.
 As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model.
-For the SUPPORT dataset, all factor variables were converted into indicator variables.
+For the SUPPORT dataset, all factor variables were converted into dummy variables.
 Here, we used lasso penalization by setting alpha to 1.
@@
 ```{r cbHazardInteractions}, echo = TRUE, eval = eval_cs3}
-
-cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
-                                       formula_time = ~log(d.time),alpha=1,ratio=ratio)
-```
-Then, we take our model and use the `absoluteRisk` function to integrate and retrieve our cumulative incidence curves.
-
-```{r CBabsolute, echo = FALSE, eval = eval_cs3}
+xInteractions=x*y[,2]
+colnames(xInteractions) <- paste("d.time*", colnames(xInteractions), sep = "")
+xFull=cbind(x,xInteractions)
+cb.ModelInter=casebase::fitSmoothHazard.fit(xFull,y,family="glmnet",time="d.time",event="death",
+                                       formula_time = ~d.time,alpha=1,ratio=rat)
+```
+Then, we take our model and use the absoluteRisk function to integrate and retrieve our absolute risk function.
+
+```{r CBabsoluteInteracts, echo = TRUE, eval = eval_cs3}
 #Estimating the absolute risk curve using the newData parameter.
-casebaseCI=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 20),newdata = x , s="lambda.1se",method=c("numerical"))
-
-cbCIAve=data.frame(casebaseCI[,1],rowMeans(casebaseCI[,-c(1)]))
+casebaseAbsoluteInter=casebase::absoluteRisk(cb.ModelInter,time = seq(1,max(completeData$d.time), 400),newdata = xFull , s="lambda.1se",method=c("numerical"))
+
+cabI=data.frame(casebaseAbsoluteInter[,1],rowMeans(casebaseAbsoluteInter[,-c(1)]))
 ```
@@
 
-```{r coxHazAbsolute, echo = FALSE, eval = eval_cs3}
+```{r coxHazAbsoluteInteracts, echo = TRUE, eval = eval_cs3}
 #Create u and xCox, which will be used when fitting Cox with glmnet.
 u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
-xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))
+
+xCoxI=as.matrix(cbind(data.frame("(Intercept)"=rep(1,each=690)),xFull))
+colnames(xCoxI)[1]="(Intercept)"
+
 #hazard for cox using glmnet
-coxglmFit=glmnet::cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
+coxNetI=glmnet::cv.glmnet(x=xCoxI,y=u, family="cox",alpha=1)
 #convergence demonstrated in plot
-#plot(coxglmFit)
+#plot(coxNet)
 #taking the coefficient estimates for later use
-nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
-nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
+nonzero_covariate_cox <- predict(coxNetI, type = "nonzero", s = "lambda.1se")
+nonzero_coef_cox <- coef(coxNetI, s = "lambda.1se")
 #creating a new dataset that only contains the covariates chosen through glmnet.
 #cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
-cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
+cleanCoxDataI<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,nonzero_covariate_cox$X1]))
 #newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
@@
 #this is used more as an object place holder.
-coxFit <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
-#The coefficients of this object will be replaced with the estimates from coxglmFit.
+coxNetFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxDataI)
+#The coefficients of this object will be replaced with the estimates from coxNet.
 #Doing so makes it so that everything is invalid aside from the coefficients.
@@
 #Std. error would be incorrect here, if we were to draw error bars.
-coxFit$coefficients<-nonzero_coef_cox@x
+coxNetFitI$coefficients<-nonzero_coef_cox@x
 #Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
-newDataCox=newData[,nonzero_covariate_cox$X1-1]
-abCoxFit<-survival::survfit(coxFit,newdata=as.data.frame(newDataCox),time = seq(0,max(completeData$d.time), 1),type="breslow")
-abCF<-abCoxFit
-abCF$surv<-rowMeans(abCF$surv)
-```
-We used the survival package to calculate cox regression's cumulative incidience without penalization and the kaplan-meier curve for incidience.
-
-```{r unpenalizedCox, echo = FALSE, eval = eval_cs3}
-Cox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,-c(1)]))
-#coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
-#abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)
-#abCFup<-abcoxFitup
-#abCFup$surv<-rowMeans(abCFup$surv)
+#posCox=nonzero_covariate_cox$X1
+#newDataCoxI=cleanCoxDataI[,posCox]
+
+abcoxNetI<-survival::survfit(coxNetFitI,type="breslow",newdata=cleanCoxDataI)
+
+abCNI<-abcoxNetI
+abCNI$surv=rowMeans(abCNI$surv)
+```
+
+```{r unpenalizedCox, echo = TRUE, eval = eval_cs3}
+fullDataCox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,-c(1)]))
+coxFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = fullDataCox)
+abcoxFitI<-survival::survfit(coxFitI,newdata=fullDataCox)
+abCFI<-abcoxFitI
+abCFI$surv<-rowMeans(abCFI$surv)
 #cb.ModelNorm=casebase::fitSmoothHazard(V2~.,time="V1",event="V2",
@@
 #cabn=data.frame(casebaseAbsoluteNorm[,1],rowMeans(casebaseAbsoluteNorm[,-c(1)]))
-```
+
+```
+We used the survival package to calculate the kaplan-meier absolute risk function.
 
@@
 # A plot to compare all three Absolute risk (Commulative Incidence)
-plot(cbCIAve[,1],cbCIAve[,2],type='l',col="black",lwd=3,main="Cumulative Incidence comparison",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
-lines(abCF,col="red",fun="event",lwd=3,conf.int = FALSE)
+plot(cabI[,1],cabI[,2],type='l',col="black",lwd=3,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
+lines(abCFI,col="red",fun="event",lwd=3,conf.int = FALSE)
 lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
-lines(abCFup ,col="purple",fun="event",lwd=3,conf.int = FALSE)
+lines(abCNI,col="purple",fun="event",lwd=3,conf.int = FALSE)
 legend("bottomright", 
-       legend = c( "Penalized Weibull with casebase","Coxnet","Cox","Kaplan-Meier"), 
+       legend = c( "Casebase (Lasso+linear)","semi-parametric Coxnet","KM curve","cox"), 
        col = c("black","red","Blue","purple"),
@@
         geom_point(position=ggstance::position_dodgev(height=0.2))
+
+```
+
+```{r covariatesLolliInteractions,eval=eval_cs3}
+#get covariates excluding d.time
+library(data.table)
+estimatescb=setDT(as.data.frame(coef(cb.ModelInter)[-c(1,2),1]), keep.rownames = TRUE)[]
+estimatescb$Model="casebase"
+colnames(estimatescb)<-c("Coefs","Estimate","Model")
+estimatescox=setDT(as.data.frame(coef(coxNetI)[-c(1),1]), keep.rownames = TRUE)[]
+estimatescox$Model="Cox"
+colnames(estimatescox)<-c("Coefs","Estimate","Model")
+estimatesI=rbind(estimatescb, estimatescox)
+estimatesI$Estimate[estimatesI$Estimate==0]=NA# make it so unchosen covariates appear empty
+library(ggstance)# for vertical dodge
+  ggplot(estimatesI, aes(x = Coefs, y = Estimate , color = Model)) +
+        geom_segment(aes(y = 0, x = Coefs, yend = Estimate, xend = Coefs),size=0.5, position = position_dodge(width = 0.5)) +
+        geom_point(position = position_dodge(width = 0.5))
 

### 2020-02-07 analysis/paper/paper.Rmd
code clean up to fix conflits

@@
 completeData<-completeData[,-c(4)]
-rat=1  #universal ratio for eval_cs3
+ratio=10  #universal ratio for eval_cs3
 #keep one individual out, at random, for which we will develop an 

### 2020-02-07 analysis/paper/paper.Rmd

--- 2020-02-06 analysis/paper/paper.Rmd
@@
 library(casebase)
-#devtools::load_all('../../../casebase')#testing
 library(splines)
@@
 library(glmnet)
-library(data.table)
 library(kableExtra)
+library(pracma)
 options(digits = 2, scipen = 999)
@@
 
-eval_introduction <- FALSE
-eval_theory <- FALSE
-eval_implementation <- FALSE
-eval_cs1 <- FALSE
-eval_cs2 <- FALSE
+eval_introduction <- TRUE
+eval_theory <- TRUE
+eval_implementation <- TRUE
+eval_cs1 <- TRUE
+eval_cs2 <- TRUE
 eval_cs3 <- TRUE
-eval_cs4 <- FALSE
+eval_cs4 <- TRUE
 ```
@@
 
-In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base sampling *uniformly*, i.e.
-to sample the base series uniformly from the study base.
+In Hanley & Miettinen [-@hanley2009fitting], the authors suggest sampling the base series *uniformly* from the study base.
 In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base.
 More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism are proportional to the cardiovascular disease event rate given by the Framingham score.
-
-The `casebase` package fits the family of hazard functions of the form
-
-$$ h(t;X) = \exp[g(t;X)] $$
-where $t$ denotes time and $X$, the individual's covariate profile.
+With this sampling mechanism, Saarela & Arjas are able to increase the efficiency of their estimators, when compared to uniform sampling.
+
+<!-- The `casebase` package fits the family of hazard functions of the form -->
+
+<!-- $$ h(t;X) = \exp[g(t;X)] $$ -->
+<!-- where $t$ denotes time and $X$, the individual's covariate profile.  -->
+
+Let $g(t; X)$ be the linear predictor such that $\log(h(t;X)) = g(t; X)$.
 Different functions of $t$ lead to different parametric hazard models.
-The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$.
-
-$$ h(t;X) = \exp(\beta_0 + \beta_1 X) $$
-
-The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
-This is also known as the *memoryless property* (Kalbfleisch and Prentice, 2002).
+The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$:
+
+$$ \log(h(t; X)) = \beta_0 + \beta_1 X.$$
+
+The instantaneous failure rate is independent of $t$.\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
+This is also known as the \textit{memoryless property} \citep{kalbfleisch2011statistical}.}
 
@@
 In particular, we excluded packages that were focused on a different set of problems, such as frailty and multi-state models.
-The remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with some of the functionalities they offer.
+The remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with a description of some of the functionalities they offer.
 
 Several packages implement penalized estimation for the Cox model: \pkg{glmnet} [-@regpathcox], \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche].
-Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \pkg{CoxRidge} [-@perperoglou] and \pkg{rstpm2} [-@clements_liu], while \pkg{survival} [-@survival-package] has an implementation for ridge.
-\pkg{casebase} provides penalized estimation in the context of parametric hazards.
-To our knowledge, \pkg{casebase} and \pkg{rsptm2} are the only packages to offer this functionality.
-\textcolor{red}{I can dive deeper to determine if this is a fair statement}
-
-Parametric survival models are implemented in a handful of packages: \pkg{CFC} [-@mahani2015bayesian], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu], \pkg{mets} [-@scheike2014estimating], and \pkg{survival}.
+Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: \pkg{CoxRidge} [-@perperoglou], \pkg{rstpm2} [-@clements_liu] and \pkg{survival} [-@survival-package].
+On the other hand, \pkg{casebase} provides penalized estimation in the context of parametric hazards.
+To our knowledge, this is the only package to offer this functionality.
+\textcolor{red}{Is this true?
+Because it seems to be a logical conclusion from your paragraph.}
+
+Parametric survival models are implemented in a handful of packages: \pkg{CFC} [-@sharabiani_mahani_2019], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu], \pkg{mets} [-@scheike2014estimating], and \pkg{survival}.
 The types of models they allow vary for each package.
@@
 
-Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, \pkg{cmprsk} and \pkg{survival} contain implementations for competing risks.
+Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets} and \pkg{survival} contain implementations for competing risks.
 The differentiating factor between these packages and \pkg{casebase} is that \pkg{casebase} handles competing risks through multiple logistic regression.
-Note that \pkg{CFC} has extra features outside the scope of the metthods discussed in this paper.
+\textcolor{red}{There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}.
+Could you incorporate some of their discussion, with proper citations?
+The biggest omission right now is \pkg{cmprsk}, which we actually use below.}
 
@@
 \textbf{casebasee} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
-\textbf{CFC} \cite{mahani2015bayesian}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
-textbf{cmprsk} \cite{gray_2019}            & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\
+\textbf{CFC} \cite{sharabiani_mahani_2019}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
 \textbf{coxRidge} \cite{perperoglou}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
@@
 ```{r bmtcrr-data, eval = eval_cs2, message = FALSE}
-
+library(casebase)
 data(bmtcrr)
@@
 
-```{r bmtcrr-risk, echo = FALSE, fig.cap="\\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups.
+```{r bmtcrr-risk, echo = FALSE, fig.cap="\\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups.
 The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.", eval = eval_cs2}
@@
 The SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital.
-Imputation was conducted using the \pkg{mice} package in \proglang{R} with default settings [@mice].
-Before imputation, there were a total of 1000 individuals and 35 variables.
-charges, Totcst, totmcst and sfdm2 were removed as they are response variables we are not interested in for the sake of this analysis.
-After imputation, we have 690 individuals.
-We lose individuals due to coliniarity between variables, preventing imputation.
-If a pair of covariates are deemed colinear to a certain degree, they are excluded from any imputations.
-Here, only edu  income  avtisst*  race  ph*  glucose*  adlp* were imputed.
-Of these 690 individuals, 66.09% experienced the event of interest.
+Imputation was conducted using predictive mean matching and using the \pkg{mice} package in \proglang{R} [@mice].
+Before imputation, there was a total of 1000 individuals and 35 variables.
+After imputation, we have 474 individuals.
+\textcolor{red}{How come we lose individuals during imputation?} Of these, 67.5% experienced the event of interest.
 <!--The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.--> A description of each variable can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}.
-The original data was obtained from the Department of Biostatistics at Vanderbilt University.
+The original data was obtained from the website of the Department of Biostatistics at Vanderbilt University.
 The imputed dataset is available as part of the \pkg{casebase} package.
@@
 The population-time plot for the SUPPORT data appears in Figure \ref{fig:support-poptime}.
-
-```{r support-poptime, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data.
-Red points represent the case series.
-The gray space demonstrates the base series."}
+\textcolor{red}{This population-time plot is strange.
+The red dots never seem to go above 200.}
+
+```{r support-poptime, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data."}
 data(support)
 support_popTime <- popTime(support,
-                           time = "d.time", event = "death",percentile_number = 0.31)
+                           time = "d.time", event = "death")
 plot(support_popTime)
@@
 
-Before we proceed with the analysis, we first remove hospital death as a covariate, as this is directly informative of death.
-All covariates will be used to estimate the hazard function.
-The hazard function will be used to estimate the cumulative incidence of each individual in the study.
-Then, the average cumulative incidience curve for all observations will be used for our comparison.
-
-```{r supportSetUp, echo = FALSE, eval = eval_cs3}
-completeData=casebase::support
-#remove hospital dead
-completeData<-completeData[,-c(4)]
-ratio=100  #universal ratio for eval_cs3
-#keep one individual out, at random, for which we will develop an 
-#absolute risk curve.
-#sam=sample(1:nrow(completeData), 1)
-#Set workingData to the remaining individuals in the data
-#workingData=completeData[-c(sam),]
-workingCompleteData=model.matrix(death~ .-d.time,data=completeData)[,-c(1)]#remove intercept
-#Create u and xCox, which will be used when fitting Cox with glmnet.
-newData=workingCompleteData
-x=workingCompleteData
-#x and y will be used to fit the casebase model under glmnet.
-#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
-y=data.matrix(completeData[,c(2,5)])
-```
-
-Our main objective is to compute the absolute risk of death for a given set of covariates, using regularization when fitting our model.
-First, we fit a smooth hazard to the data using a weibull distribution.
-As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model.
-For the SUPPORT dataset, all factor variables were converted into indicator variables.
-Here, we used lasso penalization by setting alpha to 1.
-
-```{r cbHazardInteractions}, echo = TRUE, eval = eval_cs3}
-
-cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
-                                       formula_time = ~log(d.time),alpha=1,ratio=ratio)
-```
-Then, we take our model and use the `absoluteRisk` function to integrate and retrieve our cumulative incidence curves.
-
-```{r CBabsolute, echo = FALSE, eval = eval_cs3}
-#Estimating the absolute risk curve using the newData parameter.
-casebaseCI=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 20),newdata = x , s="lambda.1se",method=c("numerical"))
-
-cbCIAve=data.frame(casebaseCI[,1],rowMeans(casebaseCI[,-c(1)]))
-```
-We will compare this curve to a regularized version of cox regression, and a kaplan-meier survival curve.
-The cox regression absolute risk curve with lasso penalization required two extra steps to retrieve.
-Coxnet from the glmnet package has tools to fit a regularized cox model, but requires a survival object skeleton of the selected variables, so that the survival package tools can be used to retrive the absolute risk.
-We handle this by fitting a regularized cox model with the glmnet package, and a cox model from the survival package with the selected variables from the regularized fit.
-This second model serves as a skeleton that permits the use of survival package functions.
-The coefficients from the regularized model will replace the coefficients in our skeleton.
-the resulting model will have the correct coefficients when integrating, but will have an incorrect standard error.
-For the purposes of this case study, we are only interested in the absolute risk curve itself and not the standard error.
-
-```{r coxHazAbsolute, echo = FALSE, eval = eval_cs3}
-#Create u and xCox, which will be used when fitting Cox with glmnet.
-u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
-xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))
-#hazard for cox using glmnet
-coxglmFit=glmnet::cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
-#convergence demonstrated in plot
-#plot(coxglmFit)
-#taking the coefficient estimates for later use
-nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
-nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
-#creating a new dataset that only contains the covariates chosen through glmnet.
-#cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
-cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
-#newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
-#fitting a cox model using regular estimation, however we will not keep it.
-#this is used more as an object place holder.
-coxFit <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
-#The coefficients of this object will be replaced with the estimates from coxglmFit.
-#Doing so makes it so that everything is invalid aside from the coefficients.
-#In this case, all we need to estimate the absolute risk is the coefficients.
-#Std. error would be incorrect here, if we were to draw error bars.
-coxFit$coefficients<-nonzero_coef_cox@x
-#Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
-newDataCox=newData[,nonzero_covariate_cox$X1-1]
-abCoxFit<-survival::survfit(coxFit,newdata=as.data.frame(newDataCox),time = seq(0,max(completeData$d.time), 1),type="breslow")
-abCF<-abCoxFit
-abCF$surv<-rowMeans(abCF$surv)
-```
-We used the survival package to calculate cox regression's cumulative incidience without penalization and the kaplan-meier curve for incidience.
-
-```{r unpenalizedCox, echo = FALSE, eval = eval_cs3}
-Cox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,-c(1)]))
-#coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
-#abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)
-#abCFup<-abcoxFitup
-#abCFup$surv<-rowMeans(abCFup$surv)
-#cb.ModelNorm=casebase::fitSmoothHazard(V2~.,time="V1",event="V2",
-       #                                ratio=rat,data=fullDataCox)
-#summary(cb.ModelNorm)#need to figure out how to deal with NAs
-#casebaseAbsoluteNorm=casebase::absoluteRisk(cb.ModelNorm,time = seq(1,max(completeData$d.time), 200),newdata = newData , s="lambda.1se",method=c("numerical"))
-#cabn=data.frame(casebaseAbsoluteNorm[,1],rowMeans(casebaseAbsoluteNorm[,-c(1)]))
-```
+Before we proceed with the analysis, we first remove hospital death as a covariate, as this is directly informative of death and cannot be used as a predictor.
+Then we select one observation at random.
+We will estimate the cumulative incidence curve for this individual.
+The other observations will be used to fit the hazard function.
+
+```{r supportSetUp, echo = TRUE, eval = eval_cs3}
+support <- dplyr::select(support, -hospdead)
+
+# Randomly select one observation
+indiv <- sample(1:nrow(support), size = 1)
+```
+
+Our main objective is to estimate the cumulative incidence of death for a given covariate profile.
+Given the number of covariates available, we will do this using regularized estimation.
+As \pkg{casebase} makes use of the \pkg{glmnet} package, we will use the function \code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}.
+Specifically, we input a matrix \code{y}, which contains the time and event variables, and a matrix \code{x}, which contains all other variables we would like to include in the model.
+For the SUPPORT dataset, all factor variables were converted into dummy variables.
+
+```{r support-design, eval = eval_cs3}
+# Create design matrix without the time variable
+support_matrix <- model.matrix(death ~ . - d.time, 
+                               data = support)
+# Remove intercept term
+support_matrix <- support_matrix[,colnames(support_matrix) != "(Intercept)"]
+
+# Create test and train dataset
+support_test <- support_matrix[indiv,,drop = FALSE]
+
+support_train <- support_matrix[-indiv,]
+y <- as.matrix(support[-indiv, c("death", "d.time")])
+```
+
+We are now ready to estimate the hazard function.
+We use lasso penalization; this is achieved by setting \code{alpha} to 1.
+Also, note that the argument \code{formula_time} can be used to specify the function form for time; in our example, we model time linearly.
+
+```{r CBhazard, echo = TRUE, eval = eval_cs3}
+support_cb <- fitSmoothHazard.fit(support_train, y, 
+                                  formula_time = ~ d.time, 
+                                  family = "glmnet", 
+                                  event = "death", time = "d.time", 
+                                  alpha = 1, ratio = 100)
+```
+
+Then, we take our model object and use the \code{absoluteRisk} function to compute our estimate of the cumulative incidence function.
+Since \code{glmnet} fitted the penalized logistic regression for a range of penalty terms, we need to specify which one we want to use for our prediction.
+We select \code{s = "lambda.1se"}, as recommended.
+
+```{r CBabsolute, echo = TRUE, eval = eval_cs3}
+time_points <- seq(1, max(support$d.time), by = 0.5)
+support_risk <- absoluteRisk(support_cb,
+                             time = time_points,
+                             newdata = support_test,
+                             s = "lambda.1se", method = "numerical")
+```
+
+We will compare this curve to a regularized version of Cox regression, and a Kaplan-Meier survival curve.
+Using the \pkg{glmnet} package, we can fit the Cox regression model using regularized estimation.
+
+```{r coxHaz, echo = TRUE, eval = eval_cs3}
+# Create survival object for glmnet
+surv_glmnet <- survival::Surv(time = y[, "d.time"], 
+                              event = y[, "death"])
+
+support_cox <- glmnet::cv.glmnet(x = support_train, 
+                                 y = surv_glmnet,
+                                 family = "cox", alpha = 1)
+```
+
+However, \pkg{glmnet} does not provide any functionality for estimating the baseline hazard, and the functionalities in the \pkg{survival} package do not work on the output from \code{cv.glmnet}.
+To circumvent this issue and get an estimate of the baseline hazard, we followed the following steps.
+First, we fitted a non-regularized Cox model using only the variables selected by \pkg{glmnet}; the baseline hazard will be estimated using this object.
+Then, we replaced the estimates of the regression coefficients by the estimates from \pkg{glmnet}.
+Finally, this modified \code{coxph} object is fed to the \code{survival::survfit} function to obtain the estimated survival function.
+
+```{r coxAbsolute, echo = TRUE, eval = eval_cs3}
+# Identifying selected covariates
+nonzero_covariates_cox <- predict(support_cox, type = "nonzero", 
+                                  s = "lambda.1se")
+nonzero_coef_cox <- coef(support_cox, s = "lambda.1se")
+# Creating a new dataset that only contains the covariates selected by glmnet
+support_cox_data <- as.data.frame(cbind(y,
+                                        support_train[, nonzero_covariates_cox$X1]))
+# Fitting a cox model
+support_cox <- survival::coxph(Surv(time = d.time, event = death) ~ ., 
+                          data = support_cox_data)
+# Replace the estimate with those obtained by glmnet
+support_cox$coefficients <- nonzero_coef_cox@x
+
+# Fitting CI curve for cox+glmnet
+support_cox_risk <- survival::survfit(support_cox,
+                                      newdata = as.data.frame(support_test),
+                                      time = time_points,
+                                      type = "breslow")
+```
+
+We used the \pkg{survival} package to calculate the Kaplan-Meier estimate of the survival function.
 
 ```{r fitKM+abRisk, echo = TRUE, eval = eval_cs3}
-
-coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
-abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)
-
-#creating a surv object to be used to fit an unadjusted absolute risk curve.
-# (Kaplan-Meier risk curve)
-km <- survival::Surv(time = completeData$d.time, event = completeData$death)
-abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
-```
-
-Now that our four cumulative incidence estimations have been calculated, we compare them all in Figure \ref{fig:abSC}.
-Both coxnet and casebase decrease the absolute risk by the end of the study, in comparison to kaplan-meier.
-```{r abSupportComparison, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
-abCFup<-abcoxFitup
-abCFup$surv<-rowMeans(abCFup$surv)
-# A plot to compare all three Absolute risk (Commulative Incidence)
-plot(cbCIAve[,1],cbCIAve[,2],type='l',col="black",lwd=3,main="Cumulative Incidence comparison",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
-lines(abCF,col="red",fun="event",lwd=3,conf.int = FALSE)
-lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
-lines(abCFup ,col="purple",fun="event",lwd=3,conf.int = FALSE)
-legend("bottomright", 
-       legend = c( "Penalized Weibull with casebase","Coxnet","Cox","Kaplan-Meier"), 
-       col = c("black","red","Blue","purple"),
-       lty = c(1, 1, 1), 
-       bg = "gray90")
-```
-
-```{r covariatesRegularizedLollipopPlot,eval=FALSE}
-
-#get covariates excluding d.time
-estimatescb=setDT(as.data.frame(coef(cb.Model)[-c(1,2),1]), keep.rownames = TRUE)[]
-estimatescb$Model="casebase"
-colnames(estimatescb)<-c("Coefs","Estimate","Model")
-estimatescox=setDT(as.data.frame(coef(coxglmFit)[-c(1),1]), keep.rownames = TRUE)[]
-estimatescox$Model="Cox"
-colnames(estimatescox)<-c("Coefs","Estimate","Model")
-estimates=rbind(estimatescb, estimatescox)
-estimates$Estimate[estimates$Estimate==0]=NA# make it so unchosen covariates appear empty
-library(ggstance)# for vertical dodge
-  ggplot(estimates, aes(x = Estimate, y = Coefs, color = Model)) +
-        geom_segment(aes(x = 0, y = Coefs, xend = Estimate, yend = Coefs),size=0.5 ) +
-        geom_point(position=ggstance::position_dodgev(height=0.2))
-
+km <- survival::Surv(time = support$d.time, 
+                     event = support$death)
+abKm <- survival::survfit(km ~ 1,
+                          type = 'kaplan-meier')
+```
+
+Now that our three cumulative incidence functions have been calculated, we compare them all in Figure \ref{fig:abSC}.
+Both \pkg{glmnet} and \pkg{casebase} have a lower estimated cumulative incidence by the end of the study, when compared to Kaplan-Meier.
+\textcolor{red}{It would be good to have an idea of why the case-base and Coxnet curves are so different.
+At the very least, I think we should compare the select covariates.}
+
+```{r abSupportComparison, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Comparison of cumulative incidence curves between regularized casebase in black, regularized Cox regression in red and Kaplan-Meier in blue."}
+plot(support_risk, type='l', col = "black", lwd = 2, 
+     main = "Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves", 
+     xlab = "Survival-Time", ylab = "Cumulative Incidence", ylim = c(0,1))
+lines(support_cox_risk, col = "red", fun = "event", lwd = 3, conf.int = FALSE)
+lines(abKm, col = "blue", fun = "event", lwd = 3, conf.int = FALSE)
+legend("bottom", 
+       legend = c("casebase", "Cox+glmnet", "KM curve"), 
+       col = c("black", "red", "blue"), lty = c(1, 1, 1), bg = "gray90")
 ```
@@
 
-The purpose of the next case study is to show how \pkg{casebase} can also be used for survival analysis problems that do not share the properties above.
-To this end, we will use the Stanford Heart Transplant Data [@clark1971cardiac,@crowley1977covariance]; we will use the version available in the \pkg{survival} package.
-
-Recall the setting of this study: patients were admitted to the Stanford program after meeting with their physician and determining that they were unlikely to respond to other forms of treatment.
-After enrollment, the program searches for a suitable donor for the patient, which can take anywhere between a few days to almost a year.
+Recall the setting of Stanford Heart Transplant study: patients were admitted to the Stanford program after meeting with their physician and determining that they were unlikely to respond to other forms of treatment.
+After enrollment, the program searched for a suitable donor for the patient, which could take anywhere between a few days to almost a year.
 We are interested in the effect of a heart transplant on survival; therefore, the patient is considered exposed only after the transplant has occurred.
@@
 
-In this article, we presented the \proglang{R} package \pkg{casebase} that provides functions to fit smooth parametric hazards and estimate cumulative incidence functions using case-base sampling.
-We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through three case studies.
-
-<!-- In the following table we provide a comparison between the Cox model and case-base sampling: -->
-
-```{r, echo=FALSE, eval = FALSE, results='asis'}
+In this article, we presented the \proglang{R} package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating cumulative incidence functions using case-base sampling.
+We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through four case studies.
+
+As a methodological framework, case-base sampling is very flexible.
+This flexibility has been explored before in the literature: for example, Saarela and Hanley [-@saarela2015case] used case-base sampling to model a time-dependent exposure variable in a vaccine safety study.
+As another example, Saarela and Arjas [-@saarela2015non] combined case-base sampling and a Bayesian non-parametric framework to compute individualized risk assessments for chronic diseases.
+In the case studies above, we explored this flexibility along two fronts.
+On the one hand, we showed how splines could be used as part of the linear predictor to model the effect of time on the hazard.
+This strategy yielded estimates of the survival function that were qualitatively similar to semiparametric estimates derived from Cox regression; however, case-base sampling led to estimates of the survival function that *vary smoothly in time*.
+On the other hand, we also displayed the flexibility of case-base sampling by showing how it could be combined with penalized logistic regression to perform variable selection.
+Even though we did not illustrate it in this article, case-base sampling can also be combined with the framework of *generalized additive models*.
+This functionality has been implemented in our package \pkg{casebase}.
+Similarly, case-base sampling can also be combined with quasi-likelihood estimation to fit survival models that can account for the presence of over-dispersion.
+All of these examples illustrate how the case-base sampling framework in general, and the package \pkg{casebase} in particular, allows the user to fit a broad and flexible family of survival functions.
+
+As presented in Hanley & Miettinen [-@hanley2009fitting], case-base sampling is comprised of three steps: 1) sampling a case series and a base series from the study; 2) fit the log-hazard as a linear function of predictors (including time); and 3) use the fitted hazard to estimate the cumulative incidence curve.
+Accordingly, our package provides functions for each step.
+Moreover, the simple interface of the \code{fittingSmoothHazard} function resembles the \code{glm} interface.
+This interface should look familiar to new users.
+Our modular approach also provides a convenient way to extend our package for new sampling or fitting strategies.
+
+In the case studies above, we compared the performance of case-base sampling with that of Cox regression and Fine-Gray models.
+In terms of function interface, \pkg{casebase} uses a formula interface that is closer to that of \code{glm}, in that the event variable is the only variable appearing on the left-hand side of the formula.
+By contrast, both \code{survival::coxph} and \code{timereg::comp.risk} use arrays that captures both the event type and time.
+Both approaches to modeling yield user-friendly code.
+However, in terms of output, both approaches differ significantly.
+Case-base sampling produces smooth hazards and smooth cumulative incidence curves, whereas Cox regression and Fine-Gray models produce step-wise cumulative incidence functions and never explicitly model the hazard function.
+Qualitatively, we showed that by using splines in the linear predictor, all three models yielded similar curves.
+However, the smooth nature of the output of \pkg{casebase} provides a more intuitive interpretation for consumers of these predictions.
+In Table \ref{tab:compCBvsCox}, we provide a side-by-side comparison between the Cox model and case-base sampling.
+
+\begin{table}
+\caption{\label{tab:compCBvsCox}Comparison between the Cox model and case-base sampling}
+\centering
+\begin{tabular}[t]{llp{5cm}}
+\toprule
+Feature & Cox model & Case-base sampling\\
+\midrule
+Model type & Semi-parametric & Fully parametric\\
+Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\
+Cumulative incidence & Step function & Smooth-in-time curve\\
+Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\
+Model testing &  & Use GLM framework \newline (e.g.\ LRT, AIC, BIC)\\
+\addlinespace
+Competing risks & Difficult & Cause-specific CIFs\\
+% Prediction & Kaplan-Meier-based & ROC, AUC, risk reclassification probabilities\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+Our choice of modeling the log-hazard as a linear function of covariates allows us to develop a simple computational scheme for estimation.
+However, as a downside, it does not allow us to model location and scale parameters separately like the package \pkg{flexsurv}.
+For example, if we look at the Weibull distribution as parametrised in \code{stats::pweibull}, the log-hazard function is given by
+$$ \log h(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,$$
+where $\alpha,\beta$ are shape and scale parameters, respectively.
+Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} allows the user to model the scale parameter as a function of covariates.
+Of course, this added flexibility comes at the cost of interpretability: by modeling the log-hazard directly, the parameter estimates from \pkg{casebase} can be interpreted as estimates of log-hazard ratios.
+To improve the flexibility of \pkg{casebase} at capturing the scale of a parametric family, we could replace the logistic regression with its quasi-likelihood counterpart and therefore model over- and under-dispersion with respect to the logistic likelihood.
+We defer the study of the properties and performance of such a model to a future article.
+
+Future work will look at some of the methodological extensions of case-base sampling.
+First, to assess the quality of the model fit, we want to study the properties of the residuals (e.g.
+Cox-Snell, martingale).
+More work needs to be done to understand these residuals in the context of the partial likelihood underlying case-base sampling.
+The resulting diagnostic tools would then be integrated in this package.
+Also, we are interested in extending case-base sampling to account for interval censoring.
+This type of censoring is very common in longitudinal studies, and many packages (e.g.
+\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2}) provide functions to account for it.
+Again, we hope to include any resulting methodology as part of this package.
+
+In future versions of the package, we also want to increase the complement of diagnostic and inferential tools that are currently available.
+For example, we would like to include the ability to compute confidence intervals for the cumulative incidence curve.
+The delta method or parametric bootstrap are two different strategies we can use to construct approximate confidence intervals.
+Furthermore, we would like to include functions to compute calibration and discrimination statistics (e.g.
+Brier score, AUC) for our models.
+Saarela and Arjas [-@saarela2015non] also describe how to obtain a posterior distribution for the AUC from their model.
+Their approach could also be included in \pkg{casebase}.
+Finally, we want to provide more flexibility in how the case-base sampling is performed.
+This could be achieved by adding a \code{hazard} argument to the function \code{sampleCaseBase}.
+In this way, users could specify their own sampling mechanism.
+For example, they could provide a hazard that gives sampling probabilities that are proportional to the cardiovascular disease event rate given by the Framingham score [@saarela2015non].
+
+In conclusion, we presented the \proglang{R} package \pkg{casebase} which implements case-base sampling for fitting parametric survival models and for estimating smooth cumulative incidence functions using the framework of generalized linear models.
+We strongly believe that its flexibility and its foundation on the familiar logistic regression model will make it appealing to new and established practioners.
+
+```{r, echo=FALSE, eval = FALSE}
 knitr::kable(data.frame(`feature` = c("model type", "time", "cumulative incidence", "non-proportional hazards","model testing","competing risks", "prediction"),

### 2020-02-07 analysis/paper/paper.tex

--- 2020-02-06 analysis/paper/paper.tex
@@
 
-This report was generated on 2020-02-07 10:31:46 using the following
+This report was generated on 2020-02-07 11:07:26 using the following
 computational environment and dependencies:
@@
 #> Local:    intro /home/turgeonm/Documents/git_repos/cbpaper
-#> Head:     [d5e7d09] 2020-02-07: add discussion of flexible sampling
+#> Head:     [f1e32e8] 2020-02-07: add discussion of flexible sampling
 \end{CodeOutput}

### 2020-02-07 analysis/paper/paper.Rmd
still working on popTime plots

@@
 
-Survival analysis has been greatly influenced over the last 50 years by the partial likelihood approach of the Cox proportional hazard model [@cox1972regression].
-This approach provides a flexible way of assessing the influence of covariates on the hazard function, without the need to specify a parametric survival model.
-This flexibility comes at the cost of decoupling the baseline hazard from the effect of the covariates.
-To recover the whole survival curve---or the cumulative incidence function (CIF)---we then need to separately estimate the baseline hazard [@breslow1972discussion].
-This in turn often leads to stepwise estimates of the survival function.
-
-From the perspective of clinicians and their patients, the most relevant quantity is often the 5- or 10-year risk of having a certain event given the patient's particular circumstances, and not the hazard ratio between a treatment and control group.
-Therefore, to make sound clinical decisions, it is important to accurately estimate the *full* hazard function, which can then be used to estimate the cumulative incidence function (CIF).
-From this perspective, the CIF and the survival function estimates are now smooth functions of time.
-
-With the goal of fitting smooth-in-time hazard functions, Hanley & Miettinen [-@hanley2009fitting] proposed a general framework for estimating fully parametric hazard models via logistic regression.
-Their approach provides users that are familiar with generalized linear models with a natural way of fitting parametric survival models.
-Moreover, their framework is very flexible: general functions of time can be estimated (e.g.
-using splines or general additive models), and hence these models retain some of the flexibility of partial likelihood approaches.
-
-In this article, we present an \proglang{R} package that combines the ideas of Hanley & Miettinen into a simple interface.
-The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile.
-Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function.
-
-In what follows, we first recall some theoretical details on case-base sampling and its use for estimating parameteric hazard functions.
-We then give a short review of existing \proglang{R} packages that implement comparable features as \pkg{casebase}.
-Next, we provide some details about the implementation of case-base sampling in our package, and we give a brief survey of its main functions.
-This is followed by four case studies that illustrate the flexibility and capabilities of \pkg{casebase}.
-We show how the same framework can be used for competing risk analyses, penalized estimation, and for studies with time-dependent exposures.
-Finally, we end the article with a discussion of the results and of future directions.
-
-# Theoretical details
+The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence (CI)) of an event, for a particular patient.
+The following points should be noted:
+
+  1. hazard ratios
+  2. If, however, the absolute risks are of interest, they have to be recovered using the semi-parametric Breslow estimator
+  3. Alternative approaches for fitting flexible hazard models for estimating absolute risks, not requiring this two-step approach? Yes! [@hanley2009fitting]
+
+[@hanley2009fitting] propose a fully parametric hazard model that can be fit via logistic regression.
+From the fitted hazard function, cumulative incidence and, thus, risk functions of time, treatment and profile can be easily derived.
+
+# Theoretical details {#theory}
 
@@
 
-In Hanley & Miettinen [-@hanley2009fitting], the authors suggest sampling the base series *uniformly* from the study base.
+In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base sampling *uniformly*, i.e.
+to sample the base series uniformly from the study base.
 In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base.
 More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism  are proportional to the cardiovascular disease event rate given by the Framingham score.
-With this sampling mechanism, Saarela & Arjas are able to increase the efficiency of their estimators, when compared to uniform sampling.
-
-<!-- The `casebase` package fits the family of hazard functions of the form -->
-
-<!-- $$ h(t;X) = \exp[g(t;X)] $$ -->
-<!-- where $t$ denotes time and $X$, the individual's covariate profile.  -->
-
-Let $g(t; X)$ be the linear predictor such that $\log(h(t;X)) = g(t; X)$.
+
+The `casebase` package fits the family of hazard functions of the form
+
+$$ h(t;X) = \exp[g(t;X)] $$
+where $t$ denotes time and $X$, the individual's covariate profile.
 Different functions of $t$ lead to different parametric hazard models.
-The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$:
-
-$$ \log(h(t; X)) = \beta_0 + \beta_1 X.$$
-
-The instantaneous failure rate is independent of $t$.\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
-This is also known as the \textit{memoryless property} \citep{kalbfleisch2011statistical}.}
+The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$.
+
+$$ h(t;X) = \exp(\beta_0 + \beta_1 X) $$
+
+The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
+This is also known as the *memoryless property* (Kalbfleisch and Prentice, 2002).
 
@@
 
-The functions in the casebase package can be divided into two categories: 1) data visualization, in the form of population-time plots; and 2) parametric modeling.
+The functions in the casebase package can be divided into two categories: 1) data visualization, in the form of population-time plots and hazard curves; and 2) parametric modeling.
 We explicitly aimed at being compatible with both \code{data.frame}s and \code{data.table}s.
@@
 ## Population-time plots
+
+As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to discretize the study base into an infinite amount of *person moments*.
+These person moments are indexed by both an individual in the study and a time point, and therefore each person moment has a covariate profile, an exposure status and an outcome status attached to it.
+We note that there is only a finite number of person moments associated with the event of interest (what Hanley & Miettinen call the *case series*).
+The case-base sampling refers to the sampling from the base of a representative finite sample called the *base series*.
+
+The case-base sampling approach described in Section \ref{theory} can be visualized in the form of a population time plot.
+They are extremely informative graphical displays of survival data and should be one of the first steps in an exploratory data analysis.
+We facilitate this task in the \pkg{casebase} package using the \code{popTime} function.
+This is done in two steps:
+\begin{enumerate}
+\item Pass your dataset to the \code{casebase::popTime} function and specify the column names corresponding to the timescale, the event status and an exposure group of interest (optional).
+This will create an object of class \code{popTime} (or \code{popTimeExposure} if you specify a value for the exposure argument)
+
+Pass the object of class popTime (or popTimeExposure) to the plot function
+\end{enumerate}
+
+We first create the necessary dataset for producing the population time plots, and we can generate the plot by using the corresponding \code{plot} method:
+
+```{r plot-erspc-data, eval = eval_cs1}
+# pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
+# plot(pt_object)
+```
+
+We can also create exposure stratified plots by specifying the \code{exposure} argument in the \code{popTime} function:
+
+```{r plot-stratified-erspc-data, eval = eval_cs1}
+pt_object_strat <- casebase::popTime(ERSPC, 
+                                     event = "DeadOfPrCa", 
+                                     exposure = "ScrArm")
+# plot(pt_object_strat)
+```
+
+We can also plot them side-by-side using the \code{ncol} argument:
+
+```{r plot-stratified-erspc-data-side, eval = eval_cs1}
+plot(pt_object_strat, ncol = 2)
+```
+
+ADD PARAGRAPH ABOUT WHAT WE CAN CONCLUDE FROM THESE GRAPHS.
 
@@
 
-Population time plots can be extremely informative graphical displays of survival data.
-They should be the first step in an exploratory data analysis.
-We facilitate this task in the \pkg{casebase} package using the \code{popTime} function.
-We first create the necessary dataset for producing the population time plots, and we can generate the plot by using the corresponding \code{plot} method:
-
-```{r plot-erspc-data, eval = eval_cs1}
-pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
-plot(pt_object)
-```
-
-We can also create exposure stratified plots by specifying the \code{exposure} argument in the \code{popTime} function:
-
-```{r plot-stratified-erspc-data, eval = eval_cs1}
-pt_object_strat <- casebase::popTime(ERSPC, 
-                                     event = "DeadOfPrCa", 
-                                     exposure = "ScrArm")
-plot(pt_object_strat)
-```
-
-We can also plot them side-by-side using the \code{ncol} argument:
-
-```{r plot-stratified-erspc-data-side, eval = eval_cs1}
-plot(pt_object_strat, ncol = 2)
-```
-
-ADD PARAGRAPH ABOUT WHAT WE CAN CONCLUDE FROM THESE GRAPHS.
-
 Next, we investigate the differences between the control and the screening arms.
@@
 
-```{r erspc-cox-cif, eval = eval_cs1}
-new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
-                       ignore = 99)
-
+```{r erspc-cox-cif, eval = eval_cs1, fig.width=8, fig.height=6, fig.cap="Cumulative incidence functions for control and screening groups in the ERSPC data.
+In each of the panels, we plot the CIF from the Cox model (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression.
+(A) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution.
+(B) Linear function of time which is equivalent to a Gompertz distribution.
+(C) The natural logarithm of time which is equivalent to a Weibull distribution.
+(D) Cubic b-spline expansion of time."}
+cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+myCols <- cbPalette[c(4,7)]
+new_data <- data.frame(ScrArm = c("Control group","Screening group"), ignore = 99)
+
+fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
+             gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
+             weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
+             splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
+
+fits <- lapply(fmla, function(i) {
+  casebase::fitSmoothHazard(i, data = ERSPC, ratio = 100)
+})
+
+risks <- lapply(fits, function(i) {
+  casebase::absoluteRisk(object = i, time = seq(0,15,0.1), newdata = new_data)
+})
+
+aft_fits <- lapply(list("exp","gompertz","weibull"), function(i) 
+  flexsurv::flexsurvreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
+                           data = ERSPC, dist = i)
+  )
+
+names(aft_fits) <- list("exponential","gompertz","weibull")
+
+# dev.off()
+layout(matrix(1:4, ncol = 2, byrow = FALSE), respect = FALSE)
+par(mar = c(0, 4.1, 4.1, 0))
 plot(survfit(cox_model, newdata = new_data),
-     xlab = "Years since Randomization", 
-     ylab = "Cumulative Incidence (%)", 
+     ylab = "",
+     xlab = "",
      fun = "event",
-     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-                    exp(coef(cox_model)), 
-                    exp(confint(cox_model))[1], 
-                    exp(confint(cox_model))[2]),
-     ylim = c(0, 2), yscale = 100)
+     xlim = c(0,15), conf.int = FALSE, col = myCols, 
+     # main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+     #                Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+     #                exp(coef(cox_model)), 
+     #                exp(confint(cox_model))[1], 
+     #                exp(confint(cox_model))[2]), 
+     lwd = 2,
+     yscale = 100,
+     xaxt = 'n')
 legend("topleft", 
-       legend = c("Control group", "Screening group"), 
-       col = c("red","blue"),
-       lty = c(1, 1), 
+       legend = c("Control (Cox)","Control (Casebase)",
+                  "Screening (Cox)", "Screening (Casebase)"), 
+       col = rep(myCols, each = 2),
+       lty = c(1, 2, 1, 2), 
+       lwd = 2,
+       cex = 1.1,
        bg = "gray90")
-```
-
-Next, we show how case-base sampling can be used to carry out a similar analysis. We fit several models that differ in how we choose to model time.
-
-<!-- The \code{fitSmoothHazard} function provides an estimate of the hazard function $h(x, t)$ is the hazard function, where $t$ denotes the numerical value of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any).
--->
-
-First, we will fit an exponential model. Recall that this corresponds to excluding time from the linear predictor.
-
-```{r erspc-casebase-exponential, eval = eval_cs1}
+lines(risks$exponential[,1], risks$exponential[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risks$exponential[,1], risks$exponential[,3], col = myCols[2], lty = 2, lwd = 2)
+text(x = 15*0.97, y = 0.05/100, labels = "(A)", cex = 1.5)
+
+par(mar = c(4.1, 4.1, 0, 0))
+plot(survfit(cox_model, newdata = new_data),
+     ylab = "", xlab = "",
+     fun = "event",
+     xlim = c(0,15), conf.int = FALSE, col = myCols, 
+     lwd = 2,
+     yscale = 100)
+lines(risks$weibull[,1], risks$weibull[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risks$weibull[,1], risks$weibull[,3], col = myCols[2], lty = 2, lwd = 2)  
+text(x = 15*0.97, y = 0.05/100, labels = "(C)", cex = 1.5)
+mtext("Cumulative incidence (%)", side = 2, line = -1.5, outer = TRUE)
+
+par(mar = c(0, 0, 4.1, 3.9))
+plot(survfit(cox_model, newdata = new_data),
+     ylab = "",xlab = "",
+     fun = "event",
+     xaxt = 'n', yaxt = 'n',
+     xlim = c(0,15), conf.int = FALSE, col = myCols, 
+     lwd = 2,
+     yscale = 100)
+lines(risks$gompertz[,1], risks$gompertz[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risks$gompertz[,1], risks$gompertz[,3], col = myCols[2], lty = 2, lwd = 2)
+text(x = 15*0.97, y = 0.05/100, labels = "(B)", cex = 1.5)
+
+par(mar = c(4.1, 0, 0, 3.9))
+plot(survfit(cox_model, newdata = new_data),
+     ylab = "",xlab = "",
+     fun = "event",
+     yaxt = 'n',
+     xlim = c(0,15), conf.int = FALSE, col = myCols, 
+     lwd = 2,
+     yscale = 100)
+lines(risks$splines[,1], risks$splines[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risks$splines[,1], risks$splines[,3], col = myCols[2], lty = 2, lwd = 2)
+text(x = 15*0.97, y = 0.05/100, labels = "(D)", cex = 1.5)
+mtext("Years since randomization", side = 1, line = -1.5, outer = TRUE)
+```
+
+```{r, results='hide'}
+par(mfrow = c(2, 2))
+lapply(fits, function(i) {
+  casebase::hazardPlot(i, newdata = new_data[2,], ci = FALSE)
+  casebase::hazardPlot(i, newdata = new_data[1,], ci = FALSE, add = TRUE, lty = 2)  
+})
+```
+
+## Equivalence between casebase and AFT
+
+```{r, eval = FALSE}
+# exponential -------------------------------------------------------------
+
 casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
@@
                                                   ratio = 100)
-
-summary(casebase_exponential)
-exp(coef(casebase_exponential)[2])
-exp(confint(casebase_exponential)[2,])
-```
+aft_exponential <- survreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
+                           data = ERSPC, dist = "exp")
+
+-coef(aft_exponential)
+coef(casebase_exponential)
+
+# weibull -----------------------------------------------------------------
+
+casebase_weibull <- casebase::fitSmoothHazard(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm, 
+                                           data = ERSPC, 
+                                           ratio = 100)
+aft_weibull <- survreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
+                       data = ERSPC, dist = "weibull")
+
+-coef(aft_weibull)/aft_weibull$scale
+
+coef(casebase_weibull)
+```
+
+```{r, eval = F}
+knitr::knit_exit()
+```
+
+Next, we show how case-base sampling can be used to carry out a similar analysis. We fit several models that differ in how we choose to model time.
+
+<!-- The \code{fitSmoothHazard} function provides an estimate of the hazard function $h(x, t)$ is the hazard function, where $t$ denotes the numerical value of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any).
+-->
+
+First, we will fit an exponential model. Recall that this corresponds to excluding time from the linear predictor.
 
@@
 In the plot below, we overlay the estimated CIF from the exponential model on the Cox model CIF:
-
-```{r erspc-casebase-cif, eval = eval_cs1}
-smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
-                                          time = seq(0,15,0.1), 
-                                          newdata = new_data)
-
-plot(survfit(cox_model, newdata = new_data),
-     xlab = "Years since Randomization", 
-     ylab = "Cumulative Incidence (%)", 
-     fun = "event",
-     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-                    exp(coef(cox_model)), 
-                    exp(confint(cox_model))[1], 
-                    exp(confint(cox_model))[2]),
-     ylim = c(0, 2), yscale = 100)
-lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
-lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
-
-legend("topleft", 
-       legend = c("Control group (Cox)","Control group (Casebase)",
-                  "Screening group (Cox)", "Screening group (Casebase)"), 
-       col = c("red","red", "blue","blue"),
-       lty = c(1, 2, 1, 2), 
-       bg = "gray90")
-```
 
@@
 
-```{r erspc-casebase-weibull, eval = eval_cs1}
-casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
-                                 data = ERSPC, 
-                                 ratio = 100)
-
-summary(casebase_time)
-exp(coef(casebase_time))
-exp(confint(casebase_time))
-```
-
 Again, the estimate of the hazard ratio is similar to that obtained from Cox regression. We then look at the estimate of the CIF:
 
-```{r erspc-casebase-weibull-cif, eval = eval_cs1}
-smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
-                                          time = seq(0,15,0.1), 
-                                          newdata = new_data)
-
-plot(survfit(cox_model, newdata = new_data),
-     xlab = "Years since Randomization", 
-     ylab = "Cumulative Incidence (%)", 
-     fun = "event",
-     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-                    exp(coef(cox_model)), 
-                    exp(confint(cox_model))[1], 
-                    exp(confint(cox_model))[2]),
-     ylim = c(0, 2), yscale = 100)
-lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
-lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
-
-legend("topleft", 
-       legend = c("Control group (Cox)","Control group (Casebase)",
-                  "Screening group (Cox)", "Screening group (Casebase)"), 
-       col = c("red","red", "blue","blue"),
-       lty = c(1, 2, 1, 2), 
-       bg = "gray90")
-```
-
 We see that the Weibull model leads to a better fit of the CIF than the Exponential model, when compared with the semi-parametric approach.
@@
 
-```{r erspc-casebase-splines, eval = eval_cs1}
-casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
-                                    data = ERSPC, 
-                                    ratio = 100)
-
-summary(casebase_splines)
-exp(coef(casebase_splines))
-exp(confint(casebase_splines))
-```
-
 Once again, we can see that the estimate of the hazard ratio is similar to that from the Cox regression. We then look at the estimate of the CIF:
-
-```{r erspc-casebase-splines-cif, eval = eval_cs1}
-smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
-                                    time = seq(0,15,0.1), 
-                                    newdata = new_data)
-
-plot(survfit(cox_model, newdata = new_data),
-     xlab = "Years since Randomization", 
-     ylab = "Cumulative Incidence (%)", 
-     fun = "event",
-     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-                    exp(coef(cox_model)), 
-                    exp(confint(cox_model))[1], 
-                    exp(confint(cox_model))[2]),
-     ylim = c(0, 2), yscale = 100)
-lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
-lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
-
-legend("topleft", 
-       legend = c("Control group (Cox)","Control group (Casebase)",
-                  "Screening group (Cox)", "Screening group (Casebase)"), 
-       col = c("red","red", "blue","blue"),
-       lty = c(1, 2, 1, 2), 
-       bg = "gray90")
-```
 
@@
 ```{r erspc-95ci, echo = FALSE, eval = eval_cs1}
-CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
-CI_exp <- c(exp(coef(casebase_exponential))["ScrArmScreening group"], exp(confint(casebase_exponential))["ScrArmScreening group",])
-CI_time <- c(exp(coef(casebase_time))["ScrArmScreening group"], exp(confint(casebase_time))["ScrArmScreening group",])
-CI_splines <- c(exp(coef(casebase_splines))["ScrArmScreening group"], exp(confint(casebase_splines))["ScrArmScreening group",])
-
-table_ci <- rbind(CI_cox, CI_exp, CI_time, CI_splines)
-colnames(table_ci) <- c("HR", "LB", "UB")
-
-data.frame(Model = c("Cox", "Exponential", "Weibull", "Splines")) %>% 
-  bind_cols(as.data.frame(table_ci)) %>% 
-  mutate(CI = paste0("(", round(LB, 2), ", ",
-                     round(UB, 2), ")"),
-         Width = UB - LB) %>% 
-  dplyr::select(-LB, -UB) %>% 
-  knitr::kable(format = "latex", booktabs = TRUE)  %>%
-  kable_styling()
+# CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
+# CI_exp <- c(exp(coef(fits$exponential))["ScrArmScreening group"], exp(confint(fits$exponential))["ScrArmScreening group",])
+# CI_gompertz <- c(exp(coef(fits$gompertz))["ScrArmScreening group"], exp(confint(fits$gompertz))["ScrArmScreening group",])
+# CI_weibull <- c(exp(coef(fits$weibull))["ScrArmScreening group"], exp(confint(fits$weibull))["ScrArmScreening group",])
+# CI_splines <- c(exp(coef(fits$splines))["ScrArmScreening group"], exp(confint(fits$splines))["ScrArmScreening group",])
+
+# function to exponentiate HR and CI and return a HR (lower, Upper) character
+exp_est_ci <- function(model = c("exponential","gompertz","weibull","splines"),
+                       fit, 
+                       covariate = "ScrArmScreening group", 
+                       type = c("casebase", "survreg")) {
+
+  type <- match.arg(type)
+  covariate <- match.arg(covariate)
+
+  if (type == "casebase"){
+
+    cis <- confint(fit[[model]])
+
+    sprintf("%0.2f (%0.2f, %0.2f)",
+            exp(coef(fit[[model]]))[[covariate]],
+            exp(cis)[covariate,1],
+            exp(cis)[covariate,2])
+  } else {
+
+    if (model == "weibull") {
+
+      # The maximised log-likelihoods are the same between survreg and flexsurvreg, 
+      # however the parameterisation is different: the first coefficient (Intercept) 
+      # reported by survreg is log(mu), and survreg's "scale" is
+      # dweibull’s (thus flexsurvreg)'s 1 / shape
+      res <- -fit[[model]]$res[covariate,] / (1 / fit[[model]]$res["shape",])
+
+      sprintf("%0.2f (%0.2f, %0.2f)",
+              exp(res[["est"]]),
+              exp(res[["U95%"]]),
+              exp(res[["L95%"]]))
+
+    } else if (model == "splines") {
+
+      c("--")
+
+    } else {
+
+      res <- fit[[model]]$res[covariate,]
+
+      sprintf("%0.2f (%0.2f, %0.2f)",
+              exp(res[["est"]]),
+              exp(res[["L95%"]]),
+              exp(res[["U95%"]]))
+    }
+  }
+}
+
+casebase_HR_CI <- lapply(list("exponential","gompertz","weibull","splines"), 
+                         exp_est_ci, 
+                         fit = fits,
+                         covariate = "ScrArmScreening group", 
+                         type = "casebase")
+
+survreg_HR_CI <- lapply(list("exponential","gompertz","weibull","splines"), 
+                         exp_est_ci, 
+                         fit = aft_fits,
+                         covariate = "ScrArmScreening group", 
+                         type = "survreg")
+
+pt <- cbind(list("Exponential","Gompertz","Weibull","Splines"),
+            do.call(rbind, casebase_HR_CI), do.call(rbind, survreg_HR_CI))
+
+# table_ci <- rbind(CI_cox, CI_exp, CI_gompertz, CI_weibull, CI_splines)
+# colnames(table_ci) <- c("HR", "LB", "UB")
+# 
+# data.frame(Model = c("Cox", "Exponential", "Gompertz", "Weibull", "Splines")) %>% 
+#   bind_cols(as.data.frame(table_ci)) %>% 
+#   mutate(CI = paste0("(", round(LB, 2), ", ",
+#                      round(UB, 2), ")"),
+#          Width = UB - LB) %>% 
+#   dplyr::select(-LB, -UB) %>% 
+#   knitr::kable(format = "latex", booktabs = TRUE)  %>%
+#   kable_styling()
+```
+
+```{r, eval = TRUE}
+## ---- print-sim-table ----
+
+kable(pt, "latex", booktabs = T, align = c("l","c","c"),
+      caption = c("Mean "),
+      col.names = c("Model","casebase","survreg")) %>%
+  kable_styling() %>%
+  # add_header_above(c(" "," ","No overlap" = 2, "All causal SNPs\nin kinship" = 2, "No overlap" = 2, "All causal SNPs\nin kinship" = 2)) %>%
+  # add_header_above(c(" "," ","Null model" = 4, "1% Causal SNPs" = 4)) %>%
+  column_spec(1, bold=T) %>%
+  collapse_rows(columns = 1, latex_hline = "major", valign = "middle") %>%
+  footnote(general = "Median (Inter-quartile range) is given for Model Size.",
+           general_title = "Note:") #%>%
+  # footnote(general = "Model Size is the number of selected variables using the high-dimensional BIC for \texttt{ggmix} and 10-fold cross validation for \texttt{lasso} and \texttt{twostep}.")
+  #          # number = c("", "Footnote 2; "),
+  #          # alphabet = c("Footnote A; ", "Footnote B; "),
+  #          # symbol = c("Footnote Symbol 1; ", "Footnote Symbol 2"),
+  #          general_title = "Note: ", number_title = "Type I: ",
+  #          alphabet_title = "Type II: ", symbol_title = "Type III: ",
+  #          footnote_as_chunk = T, title_format = c("italic", "underline")
+  # )
 ```
@@
 
+```{r}
+knitr::knit_exit()
+```
+
 ```{r erscp-compare, eval = eval_cs1}
-anova(casebase_exponential, casebase_splines, test = "LRT")
+anova(fits$exponential, fits$splines, test = "LRT")
 ```
@@
 
-Recall the setting of Stanford Heart Transplant study: patients were admitted to the Stanford program after meeting with their physician and determining that they were unlikely to respond to other forms of treatment.
-After enrollment, the program searched for a suitable donor for the patient, which could take anywhere between a few days to almost a year.
+The purpose of the next case study is to show how \pkg{casebase} can also be used for survival analysis problems that do not share the properties above.
+To this end, we will use the Stanford Heart Transplant Data [@clark1971cardiac,@crowley1977covariance]; we will use the version available in the \pkg{surivival} package.
+
+Recall the setting of this study: patients were admitted to the Stanford program after meeting with their physician and determining that they were unlikely to respond to other forms of treatment.
+After enrollment, the program searches for a suitable donor for the patient, which can take anywhere between a few days to almost a year.
 We are interested in the effect of a heart transplant on survival; therefore, the patient is considered exposed only after the transplant has occurred.
@@
 library(casebase)
+
+colnames(jasa)[which(colnames(jasa)=="futime")] <- "futime_old"
+colnames(jasa)[which(colnames(jasa)=="age")] <- "futime"
 
@@
 
-In this article, we presented the \proglang{R} package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating cumulative incidence functions using case-base sampling.
+In this article, we presented the \proglang{R} package \pkg{casebase} that provides functions to fit smooth parametric hazards and estimate cumulative incidence functions using case-base sampling.
 We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through four case studies.
@@
 On the one hand, we showed how splines could be used as part of the linear predictor to model the effect of time on the hazard.
-This strategy yielded estimates of the survival function that were qualitatively similar to semiparametric estimates derived from Cox regression; however, case-base sampling led to estimates of the survival function that *vary smoothly in time*.
+This strategy yielded estimates of the survival function that were qualitatively similar to semiparametric estimates derived from Cox regression; however, case-base sampling led to estimates that *varying smoothly in time*.
 On the other hand, we also displayed the flexibility of case-base sampling by showing how it could be combined with penalized logistic regression to perform variable selection.
-Even though we did not illustrate it in this article, case-base sampling can also be combined with the framework of *generalized additive models*.
-This functionality has been implemented in our package \pkg{casebase}.
-Similarly, case-base sampling can also be combined with quasi-likelihood estimation to fit survival models that can account for the presence of over-dispersion.
-All of these examples illustrate how the case-base sampling framework in general, and the package \pkg{casebase} in particular, allows the user to fit a broad and flexible family of survival functions.
+Similarly, our package \pkg{casebase} provides the user with the functionality of Generalized Additive Models.
+Moreover, case-base sampling can be combined with quasi-likelihood estimation to fit survival models that potentially account for the presence of over-dispersion.
+All of the examples illustrate how the case-base sampling framework in general, and the package \pkg{casebase} in particular, allows the user to fit a broad family of survival functions.
 
@@
 Accordingly, our package provides functions for each step.
-Moreover, the simple interface of the \code{fittingSmoothHazard} function resembles the \code{glm} interface.
-This interface should look familiar to new users.
+Moreover, the simple interface of the \code{fittingSmoothHazard} function resembles the \code{glm} interface, which should look familiar to new users.
 Our modular approach also provides a convenient way to extend our package for new sampling or fitting strategies.
@@
 In terms of function interface, \pkg{casebase} uses a formula interface that is closer to that of \code{glm}, in that the event variable is the only variable appearing on the left-hand side of the formula.
-By contrast, both \code{survival::coxph} and \code{timereg::comp.risk} use arrays that captures both the event type and time.
+By contrast, both \code{survival::coxph} and \code{timereg::comp.risk} use arrays as the left-hand side that captures both the event type and time.
 Both approaches to modeling yield user-friendly code.
-However, in terms of output, both approaches differ significantly.
-Case-base sampling produces smooth hazards and smooth cumulative incidence curves, whereas Cox regression and Fine-Gray models produce step-wise cumulative incidence functions and never explicitly model the hazard function.
+In terms of output, both approaches differ significantly.
+Case-base sampling produces smooth hazard and smooth cumulative incidence curves, whereas Cox regression and Fine-Gray models produce step-wise cumulative incidence functions and never explicitly model the hazard function.
 Qualitatively, we showed that by using splines in the linear predictor, all three models yielded similar curves.
-However, the smooth nature of the output of \pkg{casebase} provides a more intuitive interpretation for consumers of these predictions.
+However, the smooth nature of the output of \pkg{casebase} improves interpretation for consumers of these predictions.
 In Table \ref{tab:compCBvsCox}, we provide a side-by-side comparison between the Cox model and case-base sampling.
@@
 Future work will look at some of the methodological extensions of case-base sampling.
-First, to assess the quality of the model fit, we want to study the properties of the residuals (e.g.
+First, to assess the quality of the model fit, we may want to study the properties of the residuals (e.g.
 Cox-Snell, martingale).
@@
 Saarela and Arjas [-@saarela2015non] also describe how to obtain a posterior distribution for the AUC from their model.
-Their approach could also be included in \pkg{casebase}.
-Finally, we want to provide more flexibility in how the case-base sampling is performed.
-This could be achieved by adding a \code{hazard} argument to the function \code{sampleCaseBase}.
-In this way, users could specify their own sampling mechanism.
-For example, they could provide a hazard that gives sampling probabilities that are proportional to the cardiovascular disease event rate given by the Framingham score [@saarela2015non].
 

### 2020-02-07 analysis/paper/paper.tex
still working on popTime plots

--- 2020-02-06 analysis/paper/paper.tex
@@
 
-\hypertarget{theoretical-details}{%
-\section{Theoretical details}\label{theoretical-details}}
+\hypertarget{theory}{%
+\section{Theoretical details}\label{theory}}
 
@@
 problems, such as frailty and multi-state models. The remaining 14
-packages appear in Table \ref{tab:surv-pkgs}, along with some of the
-functionalities they offer.
+packages appear in Table \ref{tab:surv-pkgs}, along with a description
+of some of the functionalities they offer.
 
@@
 packages also include penalized estimation in the context of Cox models
-with time-varying coefficients: elastic-net penalization with
-\pkg{CoxRidge} \citeyearpar{perperoglou} and \pkg{rstpm2}
-\citeyearpar{clements_liu}, while \pkg{survival}
-\citeyearpar{survival-package} has an implementation for ridge.
+with time-varying coefficients: \pkg{CoxRidge}
+\citeyearpar{perperoglou}, \pkg{rstpm2} \citeyearpar{clements_liu} and
+\pkg{survival} \citeyearpar{survival-package}. On the other hand,
 \pkg{casebase} provides penalized estimation in the context of
-parametric hazards. To our knowledge, \pkg{casebase} and \pkg{rsptm2}
-are the only packages to offer this functionality.
-\textcolor{red}{I can dive deeper to determine if this is a fair statement}
+parametric hazards. To our knowledge, this is the only package to offer
+this functionality.
+\textcolor{red}{Is this true? Because it seems to be a logical conclusion from your paragraph.}
 
 Parametric survival models are implemented in a handful of packages:
-\pkg{CFC} \citeyearpar{mahani2015bayesian}, \pkg{flexsurv}
+\pkg{CFC} \citeyearpar{sharabiani_mahani_2019}, \pkg{flexsurv}
 \citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
@@
 Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
-\pkg{mets}, \pkg{cmprsk} and \pkg{survival} contain implementations for
-competing risks. The differentiating factor between these packages and
+\pkg{mets} and \pkg{survival} contain implementations for competing
+risks. The differentiating factor between these packages and
 \pkg{casebase} is that \pkg{casebase} handles competing risks through
-multiple logistic regression. Note that \pkg{CFC} has extra features
-outside the scope of the metthods discussed in this paper.
+multiple logistic regression.
+\textcolor{red}{There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}.
+Could you incorporate some of their discussion, with proper citations?
+The biggest omission right now is \pkg{cmprsk}, which we actually use below.}
 
@@
 The functions in the casebase package can be divided into two
-categories: 1) data visualization, in the form of population-time plots;
-and 2) parametric modeling. We explicitly aimed at being compatible with
-both \code{data.frame}s and \code{data.table}s. This is evident in some
-of the coding choices we made, and it is also reflected in our unit
-tests.
+categories: 1) data visualization, in the form of population-time plots
+and hazard curves; and 2) parametric modeling. We explicitly aimed at
+being compatible with both \code{data.frame}s and \code{data.table}s.
+This is evident in some of the coding choices we made, and it is also
+reflected in our unit tests.
 
@@
 \subsection{Population-time plots}\label{population-time-plots}}
+
+As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
+key idea behind case-base sampling is to discretize the study base into
+an infinite amount of \emph{person moments}. These person moments are
+indexed by both an individual in the study and a time point, and
+therefore each person moment has a covariate profile, an exposure status
+and an outcome status attached to it. We note that there is only a
+finite number of person moments associated with the event of interest
+(what Hanley \& Miettinen call the \emph{case series}). The case-base
+sampling refers to the sampling from the base of a representative finite
+sample called the \emph{base series}.
+
+The case-base sampling approach described in Section \ref{theory} can be
+visualized in the form of a population time plot. They are extremely
+informative graphical displays of survival data and should be one of the
+first steps in an exploratory data analysis. We facilitate this task in
+the \pkg{casebase} package using the \code{popTime} function. This is
+done in two steps:
+
+\begin{enumerate}
+\item Pass your dataset to the casebase::popTime function and specify the column names corresponding to
+
+the time of event
+the event status
+exposure (optional)
+
+This will create an object of class popTime (or popTimeExposure if you specify a value for the exposure argument)
+
+Pass the object of class popTime (or popTimeExposure) to the plot function
+\end{enumerate}
+
+We first create the necessary dataset for producing the population time
+plots, and we can generate the plot by using the corresponding
+\code{plot} method:
+
+We can also create exposure stratified plots by specifying the
+\code{exposure} argument in the \code{popTime} function:
+
+We can also plot them side-by-side using the \code{ncol} argument:
+
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-side-1} \end{center}
+
+\end{CodeChunk}
+
+ADD PARAGRAPH ABOUT WHAT WE CAN CONCLUDE FROM THESE GRAPHS.
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> data(ERSPC)
-R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
-R>                        levels = c(0, 1), 
-R>                        labels = c("Control group", "Screening group"))
-\end{CodeInput}
-\end{CodeChunk}
-
 The results of this study were published by
@@
 the approach described in \citep{liu2014recovering}.
-
-Population time plots can be extremely informative graphical displays of
-survival data. They should be the first step in an exploratory data
-analysis. We facilitate this task in the \pkg{casebase} package using
-the \code{popTime} function. We first create the necessary dataset for
-producing the population time plots, and we can generate the plot by
-using the corresponding \code{plot} method:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
-R> plot(pt_object)
-\end{CodeInput}
-\end{CodeChunk}
-
-We can also create exposure stratified plots by specifying the
-\code{exposure} argument in the \code{popTime} function:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> pt_object_strat <- casebase::popTime(ERSPC, 
-R>                                      event = "DeadOfPrCa", 
-R>                                      exposure = "ScrArm")
-R> plot(pt_object_strat)
-\end{CodeInput}
-\end{CodeChunk}
-
-We can also plot them side-by-side using the \code{ncol} argument:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> plot(pt_object_strat, ncol = 2)
-\end{CodeInput}
-\end{CodeChunk}
-
-ADD PARAGRAPH ABOUT WHAT WE CAN CONCLUDE FROM THESE GRAPHS.
 
@@
 
-\begin{CodeInput}
-R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
-R>                              data = ERSPC)
-R> summary(cox_model)
-\end{CodeInput}
+\begin{CodeOutput}
+#> Call:
+#> survival::coxph(formula = Surv(Follow.Up.Time, DeadOfPrCa) ~ 
+#>     ScrArm, data = ERSPC)
+#> 
+#>   n= 159893, number of events= 540 
+#> 
+#>                         coef exp(coef) se(coef)     z Pr(>|z|)  
+#> ScrArmScreening group -0.222     0.801    0.088 -2.52    0.012 *
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#>                       exp(coef) exp(-coef) lower .95 upper .95
+#> ScrArmScreening group     0.801       1.25     0.674     0.952
+#> 
+#> Concordance= 0.519  (se = 0.011 )
+#> Likelihood ratio test= 6.45  on 1 df,   p=0.01
+#> Wald test            = 6.37  on 1 df,   p=0.01
+#> Score (logrank) test = 6.39  on 1 df,   p=0.01
+\end{CodeOutput}
 \end{CodeChunk}
@@
 \begin{CodeChunk}
+\begin{figure}
 
-\begin{CodeInput}
-R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
-R>                        ignore = 99)
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R>      xlab = "Years since Randomization", 
-R>      ylab = "Cumulative Incidence (%)", 
-R>      fun = "event",
-R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>                     exp(coef(cox_model)), 
-R>                     exp(confint(cox_model))[1], 
-R>                     exp(confint(cox_model))[2]),
-R>      ylim = c(0, 2), yscale = 100)
-R> legend("topleft", 
-R>        legend = c("Control group", "Screening group"), 
-R>        col = c("red","blue"),
-R>        lty = c(1, 1), 
-R>        bg = "gray90")
-\end{CodeInput}
+{\centering \includegraphics{../figures/erspc-cox-cif-1}
+
+}
+
+\caption[Cumulative incidence functions for control and screening groups in the ERSPC data]{Cumulative incidence functions for control and screening groups in the ERSPC data.
+In each of the panels, we plot the CIF from the Cox model (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression.
+(A) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution.
+(B) Linear function of time which is equivalent to a Gompertz distribution.
+(C) The natural logarithm of time which is equivalent to a Weibull distribution.
+(D) Cubic b-spline expansion of time.}\label{fig:erspc-cox-cif}
+\end{figure}
 \end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{../figures/unnamed-chunk-1-1} \end{center}
+
+\end{CodeChunk}
+
+\hypertarget{equivalence-between-casebase-and-aft}{%
+\subsection{Equivalence between casebase and
+AFT}\label{equivalence-between-casebase-and-aft}}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
-R>                                                   data = ERSPC, 
-R>                                                   ratio = 100)
-R> 
-R> summary(casebase_exponential)
-R> exp(coef(casebase_exponential)[2])
-R> exp(confint(casebase_exponential)[2,])
-\end{CodeInput}
-\end{CodeChunk}
-
 We can then use the \code{absoluteRisk} function to get an estimate of
@@
 the Cox model CIF:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
-R>                                           time = seq(0,15,0.1), 
-R>                                           newdata = new_data)
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R>      xlab = "Years since Randomization", 
-R>      ylab = "Cumulative Incidence (%)", 
-R>      fun = "event",
-R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>                     exp(coef(cox_model)), 
-R>                     exp(confint(cox_model))[1], 
-R>                     exp(confint(cox_model))[2]),
-R>      ylim = c(0, 2), yscale = 100)
-R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
-R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
-R> 
-R> 
-R> legend("topleft", 
-R>        legend = c("Control group (Cox)","Control group (Casebase)",
-R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R>        col = c("red","red", "blue","blue"),
-R>        lty = c(1, 2, 1, 2), 
-R>        bg = "gray90")
-\end{CodeInput}
-\end{CodeChunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
-R>                                  data = ERSPC, 
-R>                                  ratio = 100)
-R> 
-R> summary(casebase_time)
-R> exp(coef(casebase_time))
-R> exp(confint(casebase_time))
-\end{CodeInput}
-\end{CodeChunk}
-
 Again, the estimate of the hazard ratio is similar to that obtained from
 Cox regression. We then look at the estimate of the CIF:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
-R>                                           time = seq(0,15,0.1), 
-R>                                           newdata = new_data)
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R>      xlab = "Years since Randomization", 
-R>      ylab = "Cumulative Incidence (%)", 
-R>      fun = "event",
-R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>                     exp(coef(cox_model)), 
-R>                     exp(confint(cox_model))[1], 
-R>                     exp(confint(cox_model))[2]),
-R>      ylim = c(0, 2), yscale = 100)
-R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
-R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
-R> 
-R> legend("topleft", 
-R>        legend = c("Control group (Cox)","Control group (Casebase)",
-R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R>        col = c("red","red", "blue","blue"),
-R>        lty = c(1, 2, 1, 2), 
-R>        bg = "gray90")
-\end{CodeInput}
-\end{CodeChunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
-R>                                     data = ERSPC, 
-R>                                     ratio = 100)
-R> 
-R> summary(casebase_splines)
-R> exp(coef(casebase_splines))
-R> exp(confint(casebase_splines))
-\end{CodeInput}
-\end{CodeChunk}
-
 Once again, we can see that the estimate of the hazard ratio is similar
@@
 CIF:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
-R>                                     time = seq(0,15,0.1), 
-R>                                     newdata = new_data)
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R>      xlab = "Years since Randomization", 
-R>      ylab = "Cumulative Incidence (%)", 
-R>      fun = "event",
-R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>                     exp(coef(cox_model)), 
-R>                     exp(confint(cox_model))[1], 
-R>                     exp(confint(cox_model))[2]),
-R>      ylim = c(0, 2), yscale = 100)
-R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
-R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
-R> 
-R> legend("topleft", 
-R>        legend = c("Control group (Cox)","Control group (Casebase)",
-R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R>        col = c("red","red", "blue","blue"),
-R>        lty = c(1, 2, 1, 2), 
-R>        bg = "gray90")
-\end{CodeInput}
-\end{CodeChunk}
 
@@
 
+\begin{CodeChunk}
+\begin{table}
+
+\caption{\label{tab:unnamed-chunk-4}Mean }
+\centering
+\begin{tabular}[t]{>{\bfseries}lcc}
+\toprule
+Model & casebase & survreg\\
+\midrule
+Exponential & 0.81 (0.68, 0.96) & 0.81 (0.68, 0.96)\\
+\cmidrule{1-3}
+Gompertz & 0.78 (0.66, 0.93) & 0.80 (0.67, 0.95)\\
+\cmidrule{1-3}
+Weibull & 0.80 (0.67, 0.95) & 0.80 (0.65, 0.96)\\
+\cmidrule{1-3}
+Splines & 0.81 (0.68, 0.96) & --\\
+\bottomrule
+\multicolumn{3}{l}{\textit{Note:}}\\
+\multicolumn{3}{l}{Median (Inter-quartile range) is given for Model Size.}\\
+\end{tabular}
+\end{table}
+
+\end{CodeChunk}
+
 As noted above, the usual asymptotic results hold for likelihood ratio
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> anova(casebase_exponential, casebase_splines, test = "LRT")
-\end{CodeInput}
-\end{CodeChunk}
-
-As expected, we see that splines model provides a better fit. Similarly,
-we can compare the AIC for all three case-base models:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> c("Exp." = AIC(casebase_exponential),
-R>   "Weib." = AIC(casebase_time),
-R>   "Splines" = AIC(casebase_splines))
-\end{CodeInput}
-\end{CodeChunk}
-
-Once again, we have evidence that the splines model provides the best
-fit.
-
-With this first case study, we saw how population-time plots can provide
-a useful summary of the incidence for our event of interests. In
-particular, we saw how it can provide evidence against an exponential
-hazard model. We then explored how different parametric models can
-provide different estimates of the CIF, even though they all give
-similar estimates for the hazard ratios.
-
-\hypertarget{case-study-2bone-marrow-transplant}{%
-\section{Case study 2--Bone-marrow
-transplant}\label{case-study-2bone-marrow-transplant}}
-
-In the next case study, we show how case-base sampling can also be used
-in the context of a competing risk analysis. For illustrative purposes,
-we will use the same data that was used in Scrucca \emph{et al}
-\citeyearpar{scrucca2010regression}. The data was downloaded from the
-first author's website, and it is also available as part of the
-\pkg{casebase} package.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> 
-R> data(bmtcrr)
-\end{CodeInput}
-\end{CodeChunk}
-
-The data contains information on 177 patients who received a stem-cell
-transplant for acute leukemia. The event of interest is relapse, but
-other competing causes (e.g.~transplant-related death) were also
-recorded. Several covariates were captured at baseline: sex, disease
-type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL
-and AML, respectively), disease phase at transplant (Relapse, CR1, CR2,
-CR3), source of stem cells (bone marrow and peripheral blood, coded as
-BM+PB, or only peripheral blood, coded as PB), and age. A summary of
-these baseline characteristics appear in Table \ref{tab:table1bmtcrr}.
-We note that the statistical summaries were generated differently for
-different variable types: for continuous variables, we gave the range,
-followed by the mean and standard deviation; for categorical variables,
-we gave the counts for each category.
-
-First, we can look at a population-time plot to visualize the incidence
-density of both relapse and the competing events. In Figure
-\ref{fig:compPop}, failure times are highlighted on the plot using red
-points for the event of interest (panel A) and blue dots for competing
-events (panel B). In both panels, we see evidence of a non-constant
-hazard function: the density of points is larger at the beginning of
-follow-up than at the end.
-
-Our main objective is to compute the absolute risk of relapse for a
-given set of covariates. First, we fit a smooth hazard to the data using
-a linear term for time:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> model_cb <- fitSmoothHazard(
-R>     Status ~ ftime + Sex + D + Phase + Source + Age, 
-R>     data = bmtcrr, 
-R>     ratio = 100, 
-R>     time = "ftime")
-\end{CodeInput}
-\end{CodeChunk}
-
-We will compare our hazard ratio estimates to that obtained from a Cox
-regression. To do so, we need to treat the competing event as censoring.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(survival)
-R> # Treat competing event as censoring
-R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-R>                    data = bmtcrr)
-\end{CodeInput}
-\end{CodeChunk}
-
-From the fit object, we can extract both the hazard ratios and their
-corresponding confidence intervals. These quantities appear in Table
-\ref{tab:bmtcrr-cis}. As we can see, the only significant hazard ratio
-identified by case-base sampling is the one associated with the phase of
-the disease at transplant. More precisely, being in relapse at
-transplant is associated with a hazard ratio of 3.89 when compared to
-CR1.
-
-Given the estimate of the hazard function obtained using case-base
-sampling, we can compute the absolute risk curve for a fixed covariate
-profile. We perform this computation for a 35 year old woman who
-received a stem-cell transplant from peripheral blood at relapse. We
-compared the absolute risk curve for such a woman with acute
-lymphoblastic leukemia with that for a similar woman with acute
-myeloblastic leukemia. We will estimate the curve from 0 to 60 months.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Pick 100 equidistant points between 0 and 60 months
-R> time_points <- seq(0, 60, length.out = 50)
-R> 
-R> # Data.frame containing risk profile
-R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
-R>                                      levels = levels(bmtcrr[,"Sex"])),
-R>                       "D" = c("ALL", "AML"), # Both diseases
-R>                       "Phase" = factor(c("Relapse", "Relapse"), 
-R>                                        levels = levels(bmtcrr[,"Phase"])),
-R>                       "Age" = c(35, 35),
-R>                       "Source" = factor(c("PB", "PB"), 
-R>                                         levels = levels(bmtcrr[,"Source"])))
-R> 
-R> # Estimate absolute risk curve
-R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
-R>                         method = "numerical", newdata = newdata)
-\end{CodeInput}
-\end{CodeChunk}
-
-We will compare our estimates to that obtained from a corresponding
-Fine-Gray model \citep{fine1999proportional}. The Fine-Gray model is a
-semiparametric model for the cause-specific \emph{subdistribution},
-i.e.~the function \(f_k(t)\) such that
-\[CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),\] where
-\(CI_k(t)\) is the cause-specific cumulative indicence. The Fine-Gray
-model allows to directly assess the effect of a covariate on the
-subdistribution, as opposed to the hazard. For the computation, we will
-use the \pkg{timereg} package \citep{timereg}.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(timereg)
-R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) + 
-R>                           const(Phase) + const(Source) + const(Age), 
-R>                       data = bmtcrr, cause = 1, model = "fg")
-R> 
-R> # Estimate absolute risk curve
-R> risk_fg <- predict(model_fg, newdata, times = time_points)
-\end{CodeInput}
-\end{CodeChunk}
-
-Figure \ref{fig:compAbsrisk} shows the absolute risk curves for both
-case-base sampling and the Fine-Gray model. As we can see, the two
-approaches agree quite well for acute myeloblastic leukemia; however,
-there seems to be a difference of about 5\% between the two curves for
-acute lymphoblastic leukemia. This difference does not appear to be
-significant: the curve from case-base sampling is contained within a
-95\% confidence band around the Fine-Gray absolute risk curve (figure
-not shown).
-
-\hypertarget{case-study-3support-data}{%
-\section{Case study 3--SUPPORT Data}\label{case-study-3support-data}}
-
-In the first two case studies, we described the basic functionalities of
-the \pkg{casebase} package: creating population-time plots, fitting
-parametric models for hazard functions, and estimating the corresponding
-cumulative incidence curves. In the next two case studies, we explore
-more advanced features.
-
-For the third case study, we will introduce regularization into the
-estimation of the hazard function. This can be performed using a
-combination of case-base sampling and regularized logistic regression.
-To demonstrate this capability, we examine the Study to Understand
-Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset
-\citep{knaus1995support}. The SUPPORT dataset tracks death for
-individuals who are considered seriously ill within a hospital.
-Imputation was conducted using the \pkg{mice} package in \proglang{R}
-with default settings \citep{mice}. Before imputation, there were a
-total of 1000 individuals and 35 variables. charges, Totcst, totmcst and
-sfdm2 were removed as they are response variables we are not interested
-in for the sake of this analysis. After imputation, we have 690
-individuals. We lose individuals due to coliniarity between variables,
-preventing imputation. If a pair of covariates are deemed colinear to a
-certain degree, they are excluded from any imputations. Here, only edu
-income avtisst* race ph* glucose* adlp* were imputed. Of these 690
-individuals, 66.09\% experienced the event of interest. A description of
-each variable can be found in Table \ref{tab:support1} and a breakdown
-of each categorical variable in Table \ref{tab:support2}. The original
-data was obtained from the Department of Biostatistics at Vanderbilt
-University. The imputed dataset is available as part of the
-\pkg{casebase} package.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ccccc}
-\hline
-\textbf{Name} & \textbf{Labels}                          & \textbf{Levels} & \textbf{Storage} & \textbf{NAs} \\
-\hline
-age           & Age                                      &                 & double           & 0            \\
-death         & Death at any time up to NDI date:31DEC94 &                 & double           & 0            \\
-sex           &                                          & 2               & integer          & 0            \\
-hospdead      & Death in Hospital                        &                 & double           & 0            \\
-slos          & Days from Study Entry to Discharge       &                 & double           & 0            \\
-d.time        & Days of Follow-Up                        &                 & double           & 0            \\
-dzgroup       &                                          & 8               & integer          & 0            \\
-dzclass       &                                          & 4               & integer          & 0            \\
-num.co        & number of comorbidities                  &                 & double           & 0            \\
-edu           & Years of Education                       &                 & double           & 202          \\
-income        &                                          & 4               & integer          & 349          \\
-scoma         & SUPPORT Coma Score based on Glasgow D3   &                 & double           & 0            \\
-charges       & Hospital Charges                         &                 & double           & 25           \\
-totcst        & Total RCC cost                           &                 & double           & 105          \\
-totmcst       & Total micro-cost                         &                 & double           & 372          \\
-avtisst       & Average TISS, Days 3-25                  &                 & double           & 6            \\
-race          &                                          & 5               & integer          & 5            \\
-meanbp        & Mean Arterial Blood Pressure Day 3       &                 & double           & 0            \\
-wblc          & White Blood Cell Count Day 3             &                 & double           & 24           \\
-hrt           & Heart Rate Day 3                         &                 & double           & 0            \\
-resp          & Respiration Rate Day 3                   &                 & double           & 0            \\
-temp          & Temperature (Celsius) Day 3              &                 & double           & 0            \\
-pafi          & PaO2/(.01*FiO2) Day 3                    &                 & double           & 253          \\
-alb           & Serum Albumin Day 3                      &                 & double           & 378          \\
-bili          & Bilirubin Day 3                          &                 & double           & 297          \\
-crea          & Serum creatinine Day 3                   &                 & double           & 3            \\
-sod           & Serum sodium Day 3                       &                 & double           & 0            \\
-ph            & Serum pH (arterial) Day 3                &                 & double           & 250          \\
-glucose       & Glucose Day 3                            &                 & double           & 470          \\
-bun           & BUN Day 3                                &                 & double           & 455          \\
-urine         & Urine Output Day 3                       &                 & double           & 517          \\
-adlp          & ADL Patient Day 3                        &                 & double           & 634          \\
-adls          & ADL Surrogate Day 3                      &                 & double           & 310          \\
-sfdm2         &                                          & 5               & integer          & 159          \\
-adlsc         & Imputed ADL Calibrated to Surrogate      &                 & double           & 0           
-\end{tabular}
-\caption{A description of each variable in the SUPPORT dataset.}
-\label{tab:support1}
-\end{table}
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{cc}
-\hline
-\textbf{Variable} & \textbf{Levels}                      \\
-\hline
-sex      & female                                        \\
-         & male                                          \\
-dzgroup  & ARF/MOSF w/Sepsis                             \\
-         & COPD                                          \\
-         & CHF                                           \\
-         & Cirrhosis                                     \\
-         & Coma                                          \\
-         & Colon Cancer                                  \\
-         & Lung Cancer                                   \\
-         & MOSF w/Malig                                  \\
-dzclass  & ARF/MOSF                                      \\
-         & COPD/CHF/Cirrhosis                            \\
-         & Coma                                          \\
-         & Cancer                                        \\
-income   & under \$11k                                   \\
-         & $11-$25k                                      \\
-         & $25-$50k                                      \\
-         & \textgreater{}\$50k                           \\
-race     & white                                         \\
-         & black                                         \\
-         & asian                                         \\
-         & other                                         \\
-         & hispanic                                      \\
-sfdm2    & no(M2 and SIP pres)                           \\
-         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
-         & SIP\textgreater{}=30                          \\
-         & Coma or Intub                                 \\
-         & \textless{}2 mo. follow-up                   
-\end{tabular}
-\caption{A description of each level within each categorical variable in the dataset.}
-\label{tab:support2}
-\end{table}
-
-As before, the first step is to visualize the incidence density. The
-population-time plot for the SUPPORT data appears in Figure
-\ref{fig:support-poptime}.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/support-poptime-1}
-
-}
-
-\caption{\label{fig:support-poptime} Population-time plot for the SUPPORT data. Red points represent the case series. The gray space demonstrates the base series.}\label{fig:support-poptime}
-\end{figure}
-\end{CodeChunk}
-
-Before we proceed with the analysis, we first remove hospital death as a
-covariate, as this is directly informative of death. All covariates will
-be used to estimate the hazard function. The hazard function will be
-used to estimate the cumulative incidence of each individual in the
-study. Then, the average cumulative incidience curve for all
-observations will be used for our comparison.
-
-Our main objective is to compute the absolute risk of death for a given
-set of covariates, using regularization when fitting our model. First,
-we fit a smooth hazard to the data using a weibull distribution. As
-casebase makes use of the glmnet package, we interact with the
-fitsmoothhazard.fit function using a matrix interface, where y contains
-the time and event variables and x contains all other variables we would
-like to include in the model. For the SUPPORT dataset, all factor
-variables were converted into indicator variables. Here, we used lasso
-penalization by setting alpha to 1.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> 
-R> cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
-R+                                        formula_time = ~log(d.time),alpha=1,ratio=ratio)
-\end{CodeInput}
-\end{CodeChunk}
-
-Then, we take our model and use the \texttt{absoluteRisk} function to
-integrate and retrieve our cumulative incidence curves.
-
-We will compare this curve to a regularized version of cox regression,
-and a kaplan-meier survival curve. The cox regression absolute risk
-curve with lasso penalization required two extra steps to retrieve.
-Coxnet from the glmnet package has tools to fit a regularized cox model,
-but requires a survival object skeleton of the selected variables, so
-that the survival package tools can be used to retrive the absolute
-risk. We handle this by fitting a regularized cox model with the glmnet
-package, and a cox model from the survival package with the selected
-variables from the regularized fit. This second model serves as a
-skeleton that permits the use of survival package functions. The
-coefficients from the regularized model will replace the coefficients in
-our skeleton. the resulting model will have the correct coefficients
-when integrating, but will have an incorrect standard error. For the
-purposes of this case study, we are only interested in the absolute risk
-curve itself and not the standard error.
-
-We used the survival package to calculate cox regression's cumulative
-incidience without penalization and the kaplan-meier curve for
-incidience.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> 
-R> coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
-R> abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)
-R> 
-R> #creating a surv object to be used to fit an unadjusted absolute risk curve.
-R> # (Kaplan-Meier risk curve)
-R> km <- survival::Surv(time = completeData$d.time, event = completeData$death)
-R> abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
-\end{CodeInput}
-\end{CodeChunk}
-
-Now that our four cumulative incidence estimations have been calculated,
-we compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase
-decrease the absolute risk by the end of the study, in comparison to
-kaplan-meier.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/abSupportComparison-1}
-
-}
-
-\caption{\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue.}\label{fig:abSupportComparison}
-\end{figure}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> 
-R> #get covariates excluding d.time
-R> estimatescb=setDT(as.data.frame(coef(cb.Model)[-c(1,2),1]), keep.rownames = TRUE)[]
-R> estimatescb$Model="casebase"
-R> colnames(estimatescb)<-c("Coefs","Estimate","Model")
-R> estimatescox=setDT(as.data.frame(coef(coxglmFit)[-c(1),1]), keep.rownames = TRUE)[]
-R> estimatescox$Model="Cox"
-R> colnames(estimatescox)<-c("Coefs","Estimate","Model")
-R> estimates=rbind(estimatescb, estimatescox)
-R> estimates$Estimate[estimates$Estimate==0]=NA# make it so unchosen covariates appear empty
-R> library(ggstance)# for vertical dodge
-R>   ggplot(estimates, aes(x = Estimate, y = Coefs, color = Model)) +
-R>         geom_segment(aes(x = 0, y = Coefs, xend = Estimate, yend = Coefs),size=0.5 ) +
-R>         geom_point(position=ggstance::position_dodgev(height=0.2))
-R> 
-\end{CodeInput}
-\end{CodeChunk}
-
-\hypertarget{case-study-4stanford-heart-transplant-data}{%
-\section{Case study 4--Stanford Heart Transplant
-Data}\label{case-study-4stanford-heart-transplant-data}}
-
-Although the previous case studies provide a broad overview of the
-capabilities of \pkg{casebase}, they all have two properties in common:
-
-\begin{itemize}
-  \item All covariates are fixed, i.e. they do not change over time;
-  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
-\end{itemize}
-
-Case-base sampling has been used in the literature to study vaccination
-safety, where the exposure period was defined as the week following
-vaccination \citep{saarela2015case}. Hence, the main covariate of
-interest, i.e.~exposure to the vaccine, was changing over time. In this
-context, case-base sampling offers an efficient alternative to nested
-case-control designs or self-matching.
-
-The purpose of the next case study is to show how \pkg{casebase} can
-also be used for survival analysis problems that do not share the
-properties above. To this end, we will use the Stanford Heart Transplant
-Data \citep[\citet{crowley1977covariance}]{clark1971cardiac}; we will
-use the version available in the \pkg{survival} package.
-
-Recall the setting of this study: patients were admitted to the Stanford
-program after meeting with their physician and determining that they
-were unlikely to respond to other forms of treatment. After enrollment,
-the program searches for a suitable donor for the patient, which can
-take anywhere between a few days to almost a year. We are interested in
-the effect of a heart transplant on survival; therefore, the patient is
-considered exposed only after the transplant has occurred.
-
-As before, we can look at the population-time plot for a graphical
-summary of the event incidence. As we can see, most events occur early
-during the follow-up period, and therefore we do not expect the hazard
-to be constant.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(survival)
-R> library(casebase)
-R> 
-R> stanford_popTime <- popTime(jasa, time = "futime", 
-R>                             event = "fustat")
-R> plot(stanford_popTime)
-\end{CodeInput}
-\end{CodeChunk}
-
-Since the exposure is time-dependent, we need to manually define the
-exposure variable \emph{after} case-base sampling and \emph{before}
-fitting the hazard function. For this reason, we will use the
-\texttt{sampleCaseBase} function directly.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(tidyverse)
-R> library(lubridate)
-R> 
-R> cb_data <- sampleCaseBase(jasa, time = "futime", 
-R>                           event = "fustat", ratio = 10)
-\end{CodeInput}
-\end{CodeChunk}
-
-Next, we will compute the number of days from acceptance into the
-program to transplant, and we use this variable to determine whether
-each population-moment is exposed or not.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Define exposure variable
-R> cb_data <- mutate(cb_data,
-R>                   txtime = time_length(accept.dt %--% tx.date, 
-R>                                        unit = "days"),
-R>                   exposure = case_when(
-R>                     is.na(txtime) ~ 0L,
-R>                     txtime > futime ~ 0L,
-R>                     txtime <= futime ~ 1L
-R>                   ))
-\end{CodeInput}
-\end{CodeChunk}
-
-Finally, we can fit the hazard using various linear predictors.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(splines)
-R> # Fit several models
-R> fit1 <- fitSmoothHazard(fustat ~ exposure,
-R>                         data = cb_data, time = "futime")
-R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
-R>                         data = cb_data, time = "futime")
-R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
-R>                         data = cb_data, time = "futime")
-R> fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
-R>                         data = cb_data, time = "futime")
-\end{CodeInput}
-\end{CodeChunk}
-
-Note that the fourth model (i.e. \texttt{fit4}) includes an interaction
-term between exposure and follow-up time. In other words, this model no
-longer exhibit proportional hazards. The evidence of non-proportionality
-of hazards in the Stanford Heart Transplant data has been widely
-discussed \citep{arjas1988graphical}.
-
-We can then compare the goodness of fit of these four models using the
-Akaike Information Criterion (AIC).
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute AIC
-R> c("Model1" = AIC(fit1),
-R>   "Model2" = AIC(fit2),
-R>   "Model3" = AIC(fit3),
-R>   "Model4" = AIC(fit4))
-\end{CodeInput}
-\end{CodeChunk}
-
-As we can, the best fit is the fourth model. By visualizing the hazard
-functions for both exposed and unexposed individuals, we can more
-clearly see how the hazards are no longer proportional.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute hazards---
-R> # First, create a list of time points for both exposure status
-R> hazard_data <- expand.grid(exposure = c(0, 1),
-R>                            futime = seq(0, 1000,
-R>                                         length.out = 100))
-R> # Set the offset to zero
-R> hazard_data$offset <- 0 
-R> # Use predict to get the fitted values, and exponentiate to 
-R> # transform to the right scale
-R> hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
-R>                                  type = "link"))
-R> # Add labels for plots
-R> hazard_data$Status = factor(hazard_data$exposure,
-R>                             labels = c("NoTrans", "Trans"))
-R> 
-R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
-R>     geom_line() +
-R>     theme_minimal() +
-R>     theme(legend.position = 'top') +
-R>     ylab('Hazard') + xlab('Follow-up time')
-\end{CodeInput}
-\end{CodeChunk}
-
-The non-proportionality seems to be more pronounced at the beginning of
-follow-up than the end. Finally, we can turn these estimate of the
-hazard function into estimates of the cumulative incidence functions.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute absolute risk curves
-R> newdata <- data.frame(exposure = c(0, 1))
-R> absrisk <- absoluteRisk(fit4, newdata = newdata, 
-R>                         time = seq(0, 1000, length.out = 100))
-R> 
-R> colnames(absrisk) <- c("Time", "NoTrans", "Trans")
-R> 
-R> # Rearrange the data
-R> absrisk <- gather(as.data.frame(absrisk),
-R>                   "Status", "Risk", -Time)
-R>  
-R> ggplot(absrisk, aes(Time, Risk, colour = Status)) +
-R>   geom_line() +
-R>   theme_minimal() +
-R>   theme(legend.position = 'top') +
-R>   expand_limits(y = 0.5) +
-R>   xlab('Follow-up time') + ylab('Cum. Incidence')
-\end{CodeInput}
-\end{CodeChunk}
-
-Note that we can easily adapt the code above to the situation where a
-patient receives a heart transplant at a point in time of interest, for
-example after 30 days.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute hazards---
-R> # First, create a list of time points for both exposure status
-R> one_yr_haz <- data.frame(futime = seq(0, 365,
-R>                                       length.out = 100))
-R> one_yr_haz <- mutate(one_yr_haz,
-R>                      offset = 0,
-R>                      exposure = if_else(futime < 30, 0, 1))
-R> one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
-R>                                 type = "link"))
-R> ggplot(one_yr_haz, aes(futime, hazard)) +
-R>   geom_line() +
-R>   theme_minimal() +
-R>   theme(legend.position = 'top') +
-R>   ylab('Hazard') + xlab('Follow-up time') +
-R>   geom_vline(xintercept = 30, linetype = 'dashed')
-\end{CodeInput}
-\end{CodeChunk}
-
-We can then compare the 1-year mortality risk without transplant and
-with transplant at 30 days.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> absoluteRisk(fit4, newdata = data.frame(exposure = 0),
-R>              time = 365)
-R> 
-R> # Use the trapezoidal rule to estimate the cumulative hazard
-R> min_hazard <- min(one_yr_haz$hazard)
-R> max_hazard <- max(one_yr_haz$hazard)
-R> increment <- 365/99
-R> cumhaz_est <- 0.5*increment*(sum(2*one_yr_haz$hazard) - min_hazard - max_hazard)
-R> 1 - exp(-cumhaz_est)
-\end{CodeInput}
-\end{CodeChunk}
-
-As we can see, the risk estimate at 1-year is about 30\% lower if the
-patient receives a heart transplant at 30 days.
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-In this article, we presented the \proglang{R} package \pkg{casebase}
-that provides functions to fit smooth parametric hazards and estimate
-cumulative incidence functions using case-base sampling. We outlined the
-theoretical underpinnings of the approach, we provided details about our
-implementation, and we illustrated the merits of the approach and the
-package through three case studies.
-
-\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval
-censoring, whereas \pkg{casebase} currently cannot. Interval censoring
-is a simple feature to implement and is a potential improvement to add
-to \pkg{casebase}.
-
-\hypertarget{environment-details}{%
-\section{Environment Details}\label{environment-details}}
-
-This report was generated on 2020-02-06 17:40:56 using the following
-computational environment and dependencies:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # which R packages and versions?
-R> devtools::session_info()
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> - Session info ---------------------------------------------------------------
-#>  setting  value                       
-#>  version  R version 3.6.0 (2019-04-26)
-#>  os       macOS High Sierra 10.13.6   
-#>  system   x86_64, darwin15.6.0        
-#>  ui       X11                         
-#>  language (EN)                        
-#>  collate  en_CA.UTF-8                 
-#>  ctype    en_CA.UTF-8                 
-#>  tz       America/Montreal            
-#>  date     2020-02-06                  
-#> 
-#> - Packages -------------------------------------------------------------------
-#>  package     * version    date       lib
-#>  assertthat    0.2.1      2019-03-21 [1]
-#>  backports     1.1.5      2019-10-02 [1]
-#>  broom         0.5.2      2019-04-07 [1]
-#>  callr         3.4.1      2020-01-24 [1]
-#>  casebase    * 0.2.1.9001 2020-01-30 [1]
-#>  cellranger    1.1.0      2016-07-27 [1]
-#>  cli           2.0.1      2020-01-08 [1]
-#>  codetools     0.2-16     2018-12-24 [1]
-#>  colorspace    1.4-1      2019-03-18 [1]
-#>  crayon        1.3.4      2017-09-16 [1]
-#>  data.table  * 1.12.8     2019-12-09 [1]
-#>  DBI           1.0.0      2018-05-02 [1]
-#>  dbplyr        1.4.2      2019-06-17 [1]
-#>  desc          1.2.0      2018-05-01 [1]
-#>  devtools      2.2.1      2019-09-24 [1]
-#>  digest        0.6.23     2019-11-23 [1]
-#>  dplyr       * 0.8.3      2019-07-04 [1]
-#>  ellipsis      0.3.0      2019-09-20 [1]
-#>  evaluate      0.14       2019-05-28 [1]
-#>  fansi         0.4.1      2020-01-08 [1]
-#>  farver        2.0.3      2020-01-16 [1]
-#>  forcats     * 0.4.0      2019-02-17 [1]
-#>  foreach       1.4.7      2019-07-27 [1]
-#>  fs            1.3.1      2019-05-06 [1]
-#>  generics      0.0.2      2018-11-29 [1]
-#>  ggplot2     * 3.2.1      2019-08-10 [1]
-#>  glmnet      * 3.0-2      2019-12-11 [1]
-#>  glue          1.3.1      2019-03-12 [1]
-#>  gtable        0.3.0      2019-03-25 [1]
-#>  haven         2.2.0      2019-11-08 [1]
-#>  hms           0.5.2      2019-10-30 [1]
-#>  htmltools     0.4.0      2019-10-04 [1]
-#>  httr          1.4.1      2019-08-05 [1]
-#>  iterators     1.0.12     2019-07-26 [1]
-#>  jsonlite      1.6        2018-12-07 [1]
-#>  kableExtra  * 1.1.0      2019-03-16 [1]
-#>  knitr         1.27       2020-01-16 [1]
-#>  labeling      0.3        2014-08-23 [1]
-#>  lattice       0.20-38    2018-11-04 [1]
-#>  lazyeval      0.2.2      2019-03-15 [1]
-#>  lifecycle     0.1.0      2019-08-01 [1]
-#>  lubridate     1.7.4      2018-04-11 [1]
-#>  magrittr    * 1.5        2014-11-22 [1]
-#>  Matrix      * 1.2-17     2019-03-22 [1]
-#>  memoise       1.1.0      2017-04-21 [1]
-#>  mgcv          1.8-31     2019-11-09 [1]
-#>  modelr        0.1.5      2019-08-08 [1]
-#>  munsell       0.5.0      2018-06-12 [1]
-#>  nlme          3.1-142    2019-11-07 [1]
-#>  pillar        1.4.3      2019-12-20 [1]
-#>  pkgbuild      1.0.6      2019-10-09 [1]
-#>  pkgconfig     2.0.3      2019-09-22 [1]
-#>  pkgload       1.0.2      2018-10-29 [1]
-#>  prettyunits   1.1.1      2020-01-24 [1]
-#>  processx      3.4.1      2019-07-18 [1]
-#>  ps            1.3.0      2018-12-21 [1]
-#>  purrr       * 0.3.3      2019-10-18 [1]
-#>  R6            2.4.1      2019-11-12 [1]
-#>  Rcpp          1.0.3      2019-11-08 [1]
-#>  readr       * 1.3.1      2018-12-21 [1]
-#>  readxl        1.3.1      2019-03-13 [1]
-#>  remotes       2.1.0      2019-06-24 [1]
-#>  reprex        0.3.0      2019-05-16 [1]
-#>  rlang         0.4.4      2020-01-28 [1]
-#>  rmarkdown     2.1        2020-01-20 [1]
-#>  rprojroot     1.3-2      2018-01-03 [1]
-#>  rstudioapi    0.10       2019-03-19 [1]
-#>  rticles       0.12       2019-11-12 [1]
-#>  rvest         0.3.5      2019-11-08 [1]
-#>  scales        1.1.0      2019-11-18 [1]
-#>  sessioninfo   1.1.1      2018-11-05 [1]
-#>  shape         1.4.4      2018-02-07 [1]
-#>  stringi       1.4.5      2020-01-11 [1]
-#>  stringr     * 1.4.0      2019-02-10 [1]
-#>  survival    * 3.1-7      2019-11-09 [1]
-#>  testthat      2.3.1      2019-12-01 [1]
-#>  tibble      * 2.1.3      2019-06-06 [1]
-#>  tidyr       * 1.0.0      2019-09-11 [1]
-#>  tidyselect    0.2.5      2018-10-11 [1]
-#>  tidyverse   * 1.3.0      2019-11-21 [1]
-#>  usethis       1.5.1      2019-07-04 [1]
-#>  vctrs         0.2.2      2020-01-24 [1]
-#>  VGAM          1.1-2      2019-11-21 [1]
-#>  viridisLite   0.3.0      2018-02-01 [1]
-#>  webshot       0.5.2      2019-11-22 [1]
-#>  withr         2.1.2      2018-03-15 [1]
-#>  xfun          0.12       2020-01-13 [1]
-#>  xml2          1.2.2      2019-08-09 [1]
-#>  yaml          2.2.0      2018-07-25 [1]
-#>  source                                  
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  Github (sahirbhatnagar/casebase@24359be)
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#>  CRAN (R 3.6.0)                          
-#> 
-#> [1] /Library/Frameworks/R.framework/Versions/3.6/Resources/library
-\end{CodeOutput}
-\end{CodeChunk}
-
-The current Git commit details are:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # what commit is this file at? 
-R> git2r::repository(here::here())
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> Local:    master /Users/jislam/Desktop/Gits/cbpaper
-#> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
-#> Head:     [a04834a] 2020-01-31: cbPaper draft for covars and interactions
-\end{CodeOutput}
-\end{CodeChunk}
-
 \bibliography{references.bib}

### 2020-02-07 analysis/paper/paper.Rmd

@@
 Saarela and Arjas [-@saarela2015non] also describe how to obtain a posterior distribution for the AUC from their model.
+Their approach could also be included in \pkg{casebase}.
+Finally, we want to provide more flexibility in how the case-base sampling is performed.
+This could be achieved by adding a \code{hazard} argument to the function \code{sampleCaseBase}.
+In this way, users could specify their own sampling mechanism.
+For example, they could provide a hazard that gives sampling probabilities that are proportional to the cardiovascular disease event rate given by the Framingham score [@saarela2015non].
 

### 2020-02-07 analysis/paper/paper.tex

--- 2020-02-07 analysis/paper/paper.tex
@@
 the approach described in \citep{liu2014recovering}.
-
-Population time plots can be extremely informative graphical displays of
-survival data. They should be the first step in an exploratory data
-analysis. We facilitate this task in the \pkg{casebase} package using
-the \code{popTime} function. We first create the necessary dataset for
-producing the population time plots, and we can generate the plot by
-using the corresponding \code{plot} method:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
-R> plot(pt_object)
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/plot-erspc-data-1} \end{center}
-
-\end{CodeChunk}
-
-We can also create exposure stratified plots by specifying the
-\code{exposure} argument in the \code{popTime} function:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> pt_object_strat <- casebase::popTime(ERSPC, 
-R+                                      event = "DeadOfPrCa", 
-R+                                      exposure = "ScrArm")
-R> plot(pt_object_strat)
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-1} \end{center}
-
-\end{CodeChunk}
-
-We can also plot them side-by-side using the \code{ncol} argument:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> plot(pt_object_strat, ncol = 2)
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-side-1} \end{center}
-
-\end{CodeChunk}
-
-ADD PARAGRAPH ABOUT WHAT WE CAN CONCLUDE FROM THESE GRAPHS.
 
@@
 \begin{CodeChunk}
-
-\begin{CodeInput}
-R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
-R+                        ignore = 99)
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R+      xlab = "Years since Randomization", 
-R+      ylab = "Cumulative Incidence (%)", 
-R+      fun = "event",
-R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R+                     exp(coef(cox_model)), 
-R+                     exp(confint(cox_model))[1], 
-R+                     exp(confint(cox_model))[2]),
-R+      ylim = c(0, 2), yscale = 100)
-R> legend("topleft", 
-R+        legend = c("Control group", "Screening group"), 
-R+        col = c("red","blue"),
-R+        lty = c(1, 1), 
-R+        bg = "gray90")
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/erspc-cox-cif-1} \end{center}
-
-\end{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/erspc-cox-cif-1}
+
+}
+
+\caption[Cumulative incidence functions for control and screening groups in the ERSPC data]{Cumulative incidence functions for control and screening groups in the ERSPC data.
+In each of the panels, we plot the CIF from the Cox model (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression.
+(A) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution.
+(B) Linear function of time which is equivalent to a Gompertz distribution.
+(C) The natural logarithm of time which is equivalent to a Weibull distribution.
+(D) Cubic b-spline expansion of time.}\label{fig:erspc-cox-cif}
+\end{figure}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{../figures/unnamed-chunk-1-1} \end{center}
+
+\end{CodeChunk}
+
+\hypertarget{equivalence-between-casebase-and-aft}{%
+\subsection{Equivalence between casebase and
+AFT}\label{equivalence-between-casebase-and-aft}}
 
@@
 the Cox model CIF:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
-R+                                           time = seq(0,15,0.1), 
-R+                                           newdata = new_data)
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R+      xlab = "Years since Randomization", 
-R+      ylab = "Cumulative Incidence (%)", 
-R+      fun = "event",
-R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R+                     exp(coef(cox_model)), 
-R+                     exp(confint(cox_model))[1], 
-R+                     exp(confint(cox_model))[2]),
-R+      ylim = c(0, 2), yscale = 100)
-R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
-R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
-R> 
-R> 
-R> legend("topleft", 
-R+        legend = c("Control group (Cox)","Control group (Casebase)",
-R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R+        col = c("red","red", "blue","blue"),
-R+        lty = c(1, 2, 1, 2), 
-R+        bg = "gray90")
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/erspc-casebase-cif-1} \end{center}
-
-\end{CodeChunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
-R+                                           time = seq(0,15,0.1), 
-R+                                           newdata = new_data)
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R+      xlab = "Years since Randomization", 
-R+      ylab = "Cumulative Incidence (%)", 
-R+      fun = "event",
-R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R+                     exp(coef(cox_model)), 
-R+                     exp(confint(cox_model))[1], 
-R+                     exp(confint(cox_model))[2]),
-R+      ylim = c(0, 2), yscale = 100)
-R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
-R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
-R> 
-R> legend("topleft", 
-R+        legend = c("Control group (Cox)","Control group (Casebase)",
-R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R+        col = c("red","red", "blue","blue"),
-R+        lty = c(1, 2, 1, 2), 
-R+        bg = "gray90")
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/erspc-casebase-weibull-cif-1} \end{center}
-
-\end{CodeChunk}
-
 We see that the Weibull model leads to a better fit of the CIF than the
@@
 CIF:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
-R+                                     time = seq(0,15,0.1), 
-R+                                     newdata = new_data)
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R+      xlab = "Years since Randomization", 
-R+      ylab = "Cumulative Incidence (%)", 
-R+      fun = "event",
-R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R+                     exp(coef(cox_model)), 
-R+                     exp(confint(cox_model))[1], 
-R+                     exp(confint(cox_model))[2]),
-R+      ylim = c(0, 2), yscale = 100)
-R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
-R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
-R> 
-R> legend("topleft", 
-R+        legend = c("Control group (Cox)","Control group (Casebase)",
-R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R+        col = c("red","red", "blue","blue"),
-R+        lty = c(1, 2, 1, 2), 
-R+        bg = "gray90")
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/erspc-casebase-splines-cif-1} \end{center}
-
-\end{CodeChunk}
 
@@
 \begin{CodeChunk}
-\begin{table}[H]
+\begin{table}
+
+\caption{\label{tab:unnamed-chunk-4}Mean }
 \centering
-\begin{tabular}{lrlr}
+\begin{tabular}[t]{>{\bfseries}lcc}
 \toprule
-Model & HR & CI & Width\\
+Model & casebase & survreg\\
 \midrule
-Cox & 0.80 & (0.67, 0.95) & 0.28\\
-Exponential & 0.81 & (0.68, 0.96) & 0.28\\
-Weibull & 0.79 & (0.67, 0.94) & 0.28\\
-Splines & 0.81 & (0.68, 0.96) & 0.28\\
+Exponential & 0.81 (0.68, 0.96) & 0.81 (0.68, 0.96)\\
+\cmidrule{1-3}
+Gompertz & 0.78 (0.66, 0.93) & 0.80 (0.67, 0.95)\\
+\cmidrule{1-3}
+Weibull & 0.80 (0.67, 0.95) & 0.80 (0.65, 0.96)\\
+\cmidrule{1-3}
+Splines & 0.81 (0.68, 0.96) & --\\
 \bottomrule
+\multicolumn{3}{l}{\textit{Note:}}\\
+\multicolumn{3}{l}{Median (Inter-quartile range) is given for Model Size.}\\
 \end{tabular}

### 2020-02-21 analysis/paper/paper.Rmd

--- 2020-02-07 analysis/paper/paper.Rmd
@@
 library(glmnet)
+library(data.table)
 library(kableExtra)
@@
 
-eval_introduction <- TRUE
+eval_introduction <- FALSE
 eval_theory <- TRUE
@@
 eval_cs2 <- TRUE
-eval_cs3 <- TRUE
+eval_cs3 <- FALSE
 eval_cs4 <- TRUE
@@
 
-Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets} and \pkg{survival} contain implementations for competing risks.
-The differentiating factor between these packages and \pkg{casebase} is that \pkg{casebase} handles competing risks through multiple logistic regression.
-\textcolor{red}{There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}.
-Could you incorporate some of their discussion, with proper citations?
-The biggest omission right now is \pkg{cmprsk}, which we actually use below.}
+Several \proglang{R} packages implement methodologies for competing risk analysis; for a different perspective, see Mahani & Sharabiani [-@mahani2015bayesian].
+The package \pkg{cmprsk} provides methods for cause-specific subdistributions, such as in the Fine-Gray model [-@fine1999proportional].
+On the other hand, the package \pkg{CFC} estimates cause-specific cumulative incidence functions from unadjusted, non-parametric survival functions.
+Our package \pkg{casebase} also provides functionalities for competing risk analysis by estimating the cause-specific hazards, and from these quantities, we can derive the cause-specific cumulative incidence functions.
 
@@
 \hline
-\textbf{casebasee} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
-\textbf{CFC} \cite{sharabiani_mahani_2019}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
+\textbf{casebase} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
+\textbf{CFC} \cite{mahani2015bayesian}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
+textbf{cmprsk} \cite{gray_2019}            & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\
 \textbf{coxRidge} \cite{perperoglou}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
@@
 The population-time plot for the SUPPORT data appears in Figure \ref{fig:support-poptime}.
-\textcolor{red}{This population-time plot is strange.
-The red dots never seem to go above 200.}
 
@@
 
-Before we proceed with the analysis, we first remove hospital death as a covariate, as this is directly informative of death and cannot be used as a predictor.
-Then we select one observation at random.
-We will estimate the cumulative incidence curve for this individual.
-The other observations will be used to fit the hazard function.
-
-```{r supportSetUp, echo = TRUE, eval = eval_cs3}
-support <- dplyr::select(support, -hospdead)
-
-# Randomly select one observation
-indiv <- sample(1:nrow(support), size = 1)
-```
-
-Our main objective is to estimate the cumulative incidence of death for a given covariate profile.
-Given the number of covariates available, we will do this using regularized estimation.
-As \pkg{casebase} makes use of the \pkg{glmnet} package, we will use the function \code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}.
-Specifically, we input a matrix \code{y}, which contains the time and event variables, and a matrix \code{x}, which contains all other variables we would like to include in the model.
+Before we proceed with the analysis, we first remove hospital death as a covariate, as this is directly informative of death.
+All covariates will be used to estimate the hazard function.
+The hazard function will be used to estimate the cumulative incidence of each individual in the study.
+Then, the average cumulative incidience curve for all observations will be used for our comparison.
+
+```{r supportSetUp, echo = FALSE, eval = eval_cs3}
+completeData=casebase::support
+#remove hospital dead
+completeData<-completeData[,-c(4)]
+ratio=10  #universal ratio for eval_cs3
+#keep one individual out, at random, for which we will develop an 
+#absolute risk curve.
+#sam=sample(1:nrow(completeData), 1)
+#Set workingData to the remaining individuals in the data
+#workingData=completeData[-c(sam),]
+workingCompleteData=model.matrix(death~ .-d.time,data=completeData)[,-c(1)]#remove intercept
+#Create u and xCox, which will be used when fitting Cox with glmnet.
+newData=workingCompleteData
+x=workingCompleteData
+#x and y will be used to fit the casebase model under glmnet.
+#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
+y=data.matrix(completeData[,c(2,5)])
+```
+
+Our main objective is to compute the absolute risk of death for a given set of covariates, using regularization when fitting our model.
+First, we fit a smooth hazard to the data; for the sake of this example, we opted for a linear term for time.
+As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model.
 For the SUPPORT dataset, all factor variables were converted into dummy variables.
-
-```{r support-design, eval = eval_cs3}
-# Create design matrix without the time variable
-support_matrix <- model.matrix(death ~ . - d.time, 
-                               data = support)
-# Remove intercept term
-support_matrix <- support_matrix[,colnames(support_matrix) != "(Intercept)"]
-
-# Create test and train dataset
-support_test <- support_matrix[indiv,,drop = FALSE]
-
-support_train <- support_matrix[-indiv,]
-y <- as.matrix(support[-indiv, c("death", "d.time")])
-```
-
-We are now ready to estimate the hazard function.
-We use lasso penalization; this is achieved by setting \code{alpha} to 1.
-Also, note that the argument \code{formula_time} can be used to specify the function form for time; in our example, we model time linearly.
-
-```{r CBhazard, echo = TRUE, eval = eval_cs3}
-support_cb <- fitSmoothHazard.fit(support_train, y, 
-                                  formula_time = ~ d.time, 
-                                  family = "glmnet", 
-                                  event = "death", time = "d.time", 
-                                  alpha = 1, ratio = 100)
-```
-
-Then, we take our model object and use the \code{absoluteRisk} function to compute our estimate of the cumulative incidence function.
-Since \code{glmnet} fitted the penalized logistic regression for a range of penalty terms, we need to specify which one we want to use for our prediction.
-We select \code{s = "lambda.1se"}, as recommended.
-
-```{r CBabsolute, echo = TRUE, eval = eval_cs3}
-time_points <- seq(1, max(support$d.time), by = 0.5)
-support_risk <- absoluteRisk(support_cb,
-                             time = time_points,
-                             newdata = support_test,
-                             s = "lambda.1se", method = "numerical")
-```
-
-We will compare this curve to a regularized version of Cox regression, and a Kaplan-Meier survival curve.
-Using the \pkg{glmnet} package, we can fit the Cox regression model using regularized estimation.
-
-```{r coxHaz, echo = TRUE, eval = eval_cs3}
-# Create survival object for glmnet
-surv_glmnet <- survival::Surv(time = y[, "d.time"], 
-                              event = y[, "death"])
-
-support_cox <- glmnet::cv.glmnet(x = support_train, 
-                                 y = surv_glmnet,
-                                 family = "cox", alpha = 1)
-```
-
-However, \pkg{glmnet} does not provide any functionality for estimating the baseline hazard, and the functionalities in the \pkg{survival} package do not work on the output from \code{cv.glmnet}.
-To circumvent this issue and get an estimate of the baseline hazard, we followed the following steps.
-First, we fitted a non-regularized Cox model using only the variables selected by \pkg{glmnet}; the baseline hazard will be estimated using this object.
-Then, we replaced the estimates of the regression coefficients by the estimates from \pkg{glmnet}.
-Finally, this modified \code{coxph} object is fed to the \code{survival::survfit} function to obtain the estimated survival function.
-
-```{r coxAbsolute, echo = TRUE, eval = eval_cs3}
-# Identifying selected covariates
-nonzero_covariates_cox <- predict(support_cox, type = "nonzero", 
-                                  s = "lambda.1se")
-nonzero_coef_cox <- coef(support_cox, s = "lambda.1se")
-# Creating a new dataset that only contains the covariates selected by glmnet
-support_cox_data <- as.data.frame(cbind(y,
-                                        support_train[, nonzero_covariates_cox$X1]))
-# Fitting a cox model
-support_cox <- survival::coxph(Surv(time = d.time, event = death) ~ ., 
-                          data = support_cox_data)
-# Replace the estimate with those obtained by glmnet
-support_cox$coefficients <- nonzero_coef_cox@x
-
-# Fitting CI curve for cox+glmnet
-support_cox_risk <- survival::survfit(support_cox,
-                                      newdata = as.data.frame(support_test),
-                                      time = time_points,
-                                      type = "breslow")
-```
-
-We used the \pkg{survival} package to calculate the Kaplan-Meier estimate of the survival function.
+Here, we used lasso penalization by setting alpha to 1.
+
+```{r cbHazardInteractions}, echo = TRUE, eval = eval_cs3}
+xInteractions=x*y[,2]
+colnames(xInteractions) <- paste("d.time*", colnames(xInteractions), sep = "")
+xFull=cbind(x,xInteractions)
+cb.ModelInter=casebase::fitSmoothHazard.fit(xFull,y,family="glmnet",time="d.time",event="death",
+                                       formula_time = ~d.time,alpha=1,ratio=rat)
+```
+Then, we take our model and use the absoluteRisk function to integrate and retrieve our absolute risk function.
+
+```{r CBabsoluteInteracts, echo = TRUE, eval = eval_cs3}
+#Estimating the absolute risk curve using the newData parameter.
+casebaseAbsoluteInter=casebase::absoluteRisk(cb.ModelInter,time = seq(1,max(completeData$d.time), 400),newdata = xFull , s="lambda.1se",method=c("numerical"))
+
+cabI=data.frame(casebaseAbsoluteInter[,1],rowMeans(casebaseAbsoluteInter[,-c(1)]))
+```
+We will compare this curve to a regularized version of cox regression, and a kaplan-meier survival curve.
+The cox regression absolute risk curve with lasso penalization required two extra steps to retrieve.
+Coxnet from the glmnet package has tools to fit a regularized cox model, but requires a survival object skeleton of the selected variables, so that the survival package tools can be used to retrive the absolute risk.
+We handle this by fitting a regularized cox model with the glmnet package, and a cox model from the survival package with the selected variables from the regularized fit.
+This second model serves as a skeleton that permits the use of survival package functions.
+The coefficients from the regularized model will replace the coefficients in our skeleton.
+the resulting model will have the correct coefficients when integrating, but will have an incorrect standard error.
+For the purposes of this case study, we are only interested in the absolute risk curve itself and not the standard error.
+
+```{r coxHazAbsoluteInteracts, echo = TRUE, eval = eval_cs3}
+#Create u and xCox, which will be used when fitting Cox with glmnet.
+u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
+
+xCoxI=as.matrix(cbind(data.frame("(Intercept)"=rep(1,each=690)),xFull))
+colnames(xCoxI)[1]="(Intercept)"
+
+#hazard for cox using glmnet
+coxNetI=glmnet::cv.glmnet(x=xCoxI,y=u, family="cox",alpha=1)
+#convergence demonstrated in plot
+#plot(coxNet)
+#taking the coefficient estimates for later use
+nonzero_covariate_cox <- predict(coxNetI, type = "nonzero", s = "lambda.1se")
+nonzero_coef_cox <- coef(coxNetI, s = "lambda.1se")
+#creating a new dataset that only contains the covariates chosen through glmnet.
+#cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
+cleanCoxDataI<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,nonzero_covariate_cox$X1]))
+#newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
+#fitting a cox model using regular estimation, however we will not keep it.
+#this is used more as an object place holder.
+coxNetFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxDataI)
+#The coefficients of this object will be replaced with the estimates from coxNet.
+#Doing so makes it so that everything is invalid aside from the coefficients.
+#In this case, all we need to estimate the absolute risk is the coefficients.
+#Std. error would be incorrect here, if we were to draw error bars.
+coxNetFitI$coefficients<-nonzero_coef_cox@x
+#Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
+#posCox=nonzero_covariate_cox$X1
+#newDataCoxI=cleanCoxDataI[,posCox]
+
+abcoxNetI<-survival::survfit(coxNetFitI,type="breslow",newdata=cleanCoxDataI)
+
+abCNI<-abcoxNetI
+abCNI$surv=rowMeans(abCNI$surv)
+```
+
+```{r unpenalizedCox, echo = TRUE, eval = eval_cs3}
+fullDataCox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,-c(1)]))
+coxFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = fullDataCox)
+abcoxFitI<-survival::survfit(coxFitI,newdata=fullDataCox)
+abCFI<-abcoxFitI
+abCFI$surv<-rowMeans(abCFI$surv)
+#cb.ModelNorm=casebase::fitSmoothHazard(V2~.,time="V1",event="V2",
+       #                                ratio=rat,data=fullDataCox)
+#summary(cb.ModelNorm)#need to figure out how to deal with NAs
+#casebaseAbsoluteNorm=casebase::absoluteRisk(cb.ModelNorm,time = seq(1,max(completeData$d.time), 200),newdata = newData , s="lambda.1se",method=c("numerical"))
+#cabn=data.frame(casebaseAbsoluteNorm[,1],rowMeans(casebaseAbsoluteNorm[,-c(1)]))
+
+```
+We used the survival package to calculate the kaplan-meier absolute risk function.
 
 ```{r fitKM+abRisk, echo = TRUE, eval = eval_cs3}
-km <- survival::Surv(time = support$d.time, 
-                     event = support$death)
-abKm <- survival::survfit(km ~ 1,
-                          type = 'kaplan-meier')
-```
-
-Now that our three cumulative incidence functions have been calculated, we compare them all in Figure \ref{fig:abSC}.
-Both \pkg{glmnet} and \pkg{casebase} have a lower estimated cumulative incidence by the end of the study, when compared to Kaplan-Meier.
-\textcolor{red}{It would be good to have an idea of why the case-base and Coxnet curves are so different.
-At the very least, I think we should compare the select covariates.}
-
-```{r abSupportComparison, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Comparison of cumulative incidence curves between regularized casebase in black, regularized Cox regression in red and Kaplan-Meier in blue."}
-plot(support_risk, type='l', col = "black", lwd = 2, 
-     main = "Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves", 
-     xlab = "Survival-Time", ylab = "Cumulative Incidence", ylim = c(0,1))
-lines(support_cox_risk, col = "red", fun = "event", lwd = 3, conf.int = FALSE)
-lines(abKm, col = "blue", fun = "event", lwd = 3, conf.int = FALSE)
-legend("bottom", 
-       legend = c("casebase", "Cox+glmnet", "KM curve"), 
-       col = c("black", "red", "blue"), lty = c(1, 1, 1), bg = "gray90")
+
+coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
+abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)
+
+#creating a surv object to be used to fit an unadjusted absolute risk curve.
+# (Kaplan-Meier risk curve)
+km <- survival::Surv(time = completeData$d.time, event = completeData$death)
+abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
+```
+
+Now that our four cumulative incidence estimations have been calculated, we compare them all in Figure \ref{fig:abSC}.
+Both coxnet and casebase decrease the absolute risk by the end of the study, in comparison to kaplan-meier.
+```{r abSupportComparison, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
+abCFup<-abcoxFitup
+abCFup$surv<-rowMeans(abCFup$surv)
+# A plot to compare all three Absolute risk (Commulative Incidence)
+plot(cabI[,1],cabI[,2],type='l',col="black",lwd=3,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
+lines(abCFI,col="red",fun="event",lwd=3,conf.int = FALSE)
+lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
+lines(abCNI,col="purple",fun="event",lwd=3,conf.int = FALSE)
+legend("bottomright", 
+       legend = c( "Casebase (Lasso+linear)","semi-parametric Coxnet","KM curve","cox"), 
+       col = c("black","red","Blue","purple"),
+       lty = c(1, 1, 1), 
+       bg = "gray90")
+```
+
+```{r covariatesRegularizedLollipopPlot,eval=FALSE}
+
+#get covariates excluding d.time
+estimatescb=setDT(as.data.frame(coef(cb.Model)[-c(1,2),1]), keep.rownames = TRUE)[]
+estimatescb$Model="casebase"
+colnames(estimatescb)<-c("Coefs","Estimate","Model")
+estimatescox=setDT(as.data.frame(coef(coxNet)[-c(1),1]), keep.rownames = TRUE)[]
+estimatescox$Model="Cox"
+colnames(estimatescox)<-c("Coefs","Estimate","Model")
+estimates=rbind(estimatescb, estimatescox)
+estimates$Estimate[estimates$Estimate==0]=NA# make it so unchosen covariates appear empty
+library(ggstance)# for vertical dodge
+  ggplot(estimates, aes(x = Estimate, y = Coefs, color = Model)) +
+        geom_segment(aes(x = 0, y = Coefs, xend = Estimate, yend = Coefs),size=0.5 ) +
+        geom_point(position=ggstance::position_dodgev(height=0.2))
+
+```
+
+```{r covariatesLolliInteractions,eval=eval_cs3}
+#get covariates excluding d.time
+library(data.table)
+estimatescb=setDT(as.data.frame(coef(cb.ModelInter)[-c(1,2),1]), keep.rownames = TRUE)[]
+estimatescb$Model="casebase"
+colnames(estimatescb)<-c("Coefs","Estimate","Model")
+estimatescox=setDT(as.data.frame(coef(coxNetI)[-c(1),1]), keep.rownames = TRUE)[]
+estimatescox$Model="Cox"
+colnames(estimatescox)<-c("Coefs","Estimate","Model")
+estimatesI=rbind(estimatescb, estimatescox)
+estimatesI$Estimate[estimatesI$Estimate==0]=NA# make it so unchosen covariates appear empty
+library(ggstance)# for vertical dodge
+  ggplot(estimatesI, aes(x = Coefs, y = Estimate , color = Model)) +
+        geom_segment(aes(y = 0, x = Coefs, yend = Estimate, xend = Coefs),size=0.5, position = position_dodge(width = 0.5)) +
+        geom_point(position = position_dodge(width = 0.5))
+
 ```
@@
 In this next case study, we will use the Stanford Heart Transplant data [@clark1971cardiac,@crowley1977covariance] to show how case-base sampling can also be used in the context of time-varying covariates.
-As an example that already appeared in the literature, case-base sampling was to study vaccination safety, where the exposure period was defined as the week following vaccination [@saarela2015case].
+As an example that already appeared in the literature, case-base sampling was used to study vaccination safety, where the exposure period was defined as the week following vaccination [@saarela2015case].
 Hence, the main covariate of interest, i.e.

### 2020-02-21 analysis/paper/paper.tex

--- 2020-02-07 analysis/paper/paper.tex
@@
 
-Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
-\pkg{mets} and \pkg{survival} contain implementations for competing
-risks. The differentiating factor between these packages and
-\pkg{casebase} is that \pkg{casebase} handles competing risks through
-multiple logistic regression.
-\textcolor{red}{There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}.
-Could you incorporate some of their discussion, with proper citations?
-The biggest omission right now is \pkg{cmprsk}, which we actually use below.}
+Several \proglang{R} packages implement methodologies for competing risk
+analysis; for a different perspective, see Mahani \& Sharabiani
+\citeyearpar{mahani2015bayesian}. The package \pkg{cmprsk} provides
+methods for cause-specific subdistributions, such as in the Fine-Gray
+model \citeyearpar{fine1999proportional}. On the other hand, the package
+\pkg{CFC} estimates cause-specific cumulative incidence functions from
+unadjusted, non-parametric survival functions. Our package
+\pkg{casebase} also provides functionalities for competing risk analysis
+by estimating the cause-specific hazards, and from these quantities, we
+can derive the cause-specific cumulative incidence functions.
 
@@
 
-\caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups.
+\caption{\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups.
 The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:bmtcrr-risk}
@@
 \ref{fig:support-poptime}.
-\textcolor{red}{This population-time plot is strange. The red dots never seem to go above 200.}
 
@@
 R> support_popTime <- popTime(support,
-R+                            time = "d.time", event = "death")
+R>                            time = "d.time", event = "death",percentile_number = 0.31)
 R> plot(support_popTime)
 \end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics{../figures/support-poptime-1}
-
-}
-
-\caption{\label{fig:support-poptime} Population-time plot for the SUPPORT data.}\label{fig:support-poptime}
-\end{figure}
 \end{CodeChunk}
@@
 Before we proceed with the analysis, we first remove hospital death as a
-covariate, as this is directly informative of death and cannot be used
-as a predictor. Then we select one observation at random. We will
-estimate the cumulative incidence curve for this individual. The other
-observations will be used to fit the hazard function.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> support <- dplyr::select(support, -hospdead)
-R> 
-R> # Randomly select one observation
-R> indiv <- sample(1:nrow(support), size = 1)
-\end{CodeInput}
-\end{CodeChunk}
-
-Our main objective is to estimate the cumulative incidence of death for
-a given covariate profile. Given the number of covariates available, we
-will do this using regularized estimation. As \pkg{casebase} makes use
-of the \pkg{glmnet} package, we will use the function
-\code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}.
-Specifically, we input a matrix \code{y}, which contains the time and
-event variables, and a matrix \code{x}, which contains all other
-variables we would like to include in the model. For the SUPPORT
-dataset, all factor variables were converted into dummy variables.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Create design matrix without the time variable
-R> support_matrix <- model.matrix(death ~ . - d.time, 
-R+                                data = support)
-R> # Remove intercept term
-R> support_matrix <- support_matrix[,colnames(support_matrix) != "(Intercept)"]
-R> 
-R> # Create test and train dataset
-R> support_test <- support_matrix[indiv,,drop = FALSE]
-R> 
-R> support_train <- support_matrix[-indiv,]
-R> y <- as.matrix(support[-indiv, c("death", "d.time")])
-\end{CodeInput}
-\end{CodeChunk}
-
-We are now ready to estimate the hazard function. We use lasso
-penalization; this is achieved by setting \code{alpha} to 1. Also, note
-that the argument \code{formula_time} can be used to specify the
-function form for time; in our example, we model time linearly.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> support_cb <- fitSmoothHazard.fit(support_train, y, 
-R+                                   formula_time = ~ d.time, 
-R+                                   family = "glmnet", 
-R+                                   event = "death", time = "d.time", 
-R+                                   alpha = 1, ratio = 100)
-\end{CodeInput}
-\end{CodeChunk}
-
-Then, we take our model object and use the \code{absoluteRisk} function
-to compute our estimate of the cumulative incidence function. Since
-\code{glmnet} fitted the penalized logistic regression for a range of
-penalty terms, we need to specify which one we want to use for our
-prediction. We select \code{s = "lambda.1se"}, as recommended.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> time_points <- seq(1, max(support$d.time), by = 0.5)
-R> support_risk <- absoluteRisk(support_cb,
-R+                              time = time_points,
-R+                              newdata = support_test,
-R+                              s = "lambda.1se", method = "numerical")
-\end{CodeInput}
-\end{CodeChunk}
-
-We will compare this curve to a regularized version of Cox regression,
-and a Kaplan-Meier survival curve. Using the \pkg{glmnet} package, we
-can fit the Cox regression model using regularized estimation.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Create survival object for glmnet
-R> surv_glmnet <- survival::Surv(time = y[, "d.time"], 
-R+                               event = y[, "death"])
-R> 
-R> support_cox <- glmnet::cv.glmnet(x = support_train, 
-R+                                  y = surv_glmnet,
-R+                                  family = "cox", alpha = 1)
-\end{CodeInput}
-\end{CodeChunk}
-
-However, \pkg{glmnet} does not provide any functionality for estimating
-the baseline hazard, and the functionalities in the \pkg{survival}
-package do not work on the output from \code{cv.glmnet}. To circumvent
-this issue and get an estimate of the baseline hazard, we followed the
-following steps. First, we fitted a non-regularized Cox model using only
-the variables selected by \pkg{glmnet}; the baseline hazard will be
-estimated using this object. Then, we replaced the estimates of the
-regression coefficients by the estimates from \pkg{glmnet}. Finally,
-this modified \code{coxph} object is fed to the \code{survival::survfit}
-function to obtain the estimated survival function.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Identifying selected covariates
-R> nonzero_covariates_cox <- predict(support_cox, type = "nonzero", 
-R+                                   s = "lambda.1se")
-R> nonzero_coef_cox <- coef(support_cox, s = "lambda.1se")
-R> # Creating a new dataset that only contains the covariates selected by glmnet
-R> support_cox_data <- as.data.frame(cbind(y,
-R+                                         support_train[, nonzero_covariates_cox$X1]))
-R> # Fitting a cox model
-R> support_cox <- survival::coxph(Surv(time = d.time, event = death) ~ ., 
-R+                           data = support_cox_data)
-R> # Replace the estimate with those obtained by glmnet
-R> support_cox$coefficients <- nonzero_coef_cox@x
-R> 
-R> # Fitting CI curve for cox+glmnet
-R> support_cox_risk <- survival::survfit(support_cox,
-R+                                       newdata = as.data.frame(support_test),
-R+                                       time = time_points,
-R+                                       type = "breslow")
-\end{CodeInput}
-\end{CodeChunk}
-
-We used the \pkg{survival} package to calculate the Kaplan-Meier
-estimate of the survival function.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> km <- survival::Surv(time = support$d.time, 
-R+                      event = support$death)
-R> abKm <- survival::survfit(km ~ 1,
-R+                           type = 'kaplan-meier')
-\end{CodeInput}
-\end{CodeChunk}
-
-Now that our three cumulative incidence functions have been calculated,
-we compare them all in Figure \ref{fig:abSC}. Both \pkg{glmnet} and
-\pkg{casebase} have a lower estimated cumulative incidence by the end of
-the study, when compared to Kaplan-Meier.
-\textcolor{red}{It would be good to have an idea of why the case-base and Coxnet curves are so different. At the very least, I think we should compare the select covariates.}
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/abSupportComparison-1}
-
-}
-
-\caption{\label{fig:abSC} Comparison of cumulative incidence curves between regularized casebase in black, regularized Cox regression in red and Kaplan-Meier in blue.}\label{fig:abSupportComparison}
-\end{figure}
+covariate, as this is directly informative of death. All covariates will
+be used to estimate the hazard function. The hazard function will be
+used to estimate the cumulative incidence of each individual in the
+study. Then, the average cumulative incidience curve for all
+observations will be used for our comparison.
+
+Our main objective is to compute the absolute risk of death for a given
+set of covariates, using regularization when fitting our model. First,
+we fit a smooth hazard to the data; for the sake of this example, we
+opted for a linear term for time. As casebase makes use of the glmnet
+package, we interact with the fitsmoothhazard.fit function using a
+matrix interface, where y contains the time and event variables and x
+contains all other variables we would like to include in the model. For
+the SUPPORT dataset, all factor variables were converted into dummy
+variables. Here, we used lasso penalization by setting alpha to 1.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> xInteractions=x*y[,2]
+R> colnames(xInteractions) <- paste("d.time*", colnames(xInteractions), sep = "")
+R> xFull=cbind(x,xInteractions)
+R> cb.ModelInter=casebase::fitSmoothHazard.fit(xFull,y,family="glmnet",time="d.time",event="death",
+R>                                        formula_time = ~d.time,alpha=1,ratio=rat)
+\end{CodeInput}
+\end{CodeChunk}
+
+Then, we take our model and use the absoluteRisk function to integrate
+and retrieve our absolute risk function.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #Estimating the absolute risk curve using the newData parameter.
+R> casebaseAbsoluteInter=casebase::absoluteRisk(cb.ModelInter,time = seq(1,max(completeData$d.time), 400),newdata = xFull , s="lambda.1se",method=c("numerical"))
+R> 
+R> cabI=data.frame(casebaseAbsoluteInter[,1],rowMeans(casebaseAbsoluteInter[,-c(1)]))
+\end{CodeInput}
+\end{CodeChunk}
+
+We will compare this curve to a regularized version of cox regression,
+and a kaplan-meier survival curve. The cox regression absolute risk
+curve with lasso penalization required two extra steps to retrieve.
+Coxnet from the glmnet package has tools to fit a regularized cox model,
+but requires a survival object skeleton of the selected variables, so
+that the survival package tools can be used to retrive the absolute
+risk. We handle this by fitting a regularized cox model with the glmnet
+package, and a cox model from the survival package with the selected
+variables from the regularized fit. This second model serves as a
+skeleton that permits the use of survival package functions. The
+coefficients from the regularized model will replace the coefficients in
+our skeleton. the resulting model will have the correct coefficients
+when integrating, but will have an incorrect standard error. For the
+purposes of this case study, we are only interested in the absolute risk
+curve itself and not the standard error.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #Create u and xCox, which will be used when fitting Cox with glmnet.
+R> u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
+R> 
+R> xCoxI=as.matrix(cbind(data.frame("(Intercept)"=rep(1,each=690)),xFull))
+R> colnames(xCoxI)[1]="(Intercept)"
+R> 
+R> #hazard for cox using glmnet
+R> coxNetI=glmnet::cv.glmnet(x=xCoxI,y=u, family="cox",alpha=1)
+R> #convergence demonstrated in plot
+R> #plot(coxNet)
+R> #taking the coefficient estimates for later use
+R> nonzero_covariate_cox <- predict(coxNetI, type = "nonzero", s = "lambda.1se")
+R> nonzero_coef_cox <- coef(coxNetI, s = "lambda.1se")
+R> #creating a new dataset that only contains the covariates chosen through glmnet.
+R> #cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
+R> cleanCoxDataI<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,nonzero_covariate_cox$X1]))
+R> #newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
+R> #fitting a cox model using regular estimation, however we will not keep it.
+R> #this is used more as an object place holder.
+R> coxNetFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxDataI)
+R> #The coefficients of this object will be replaced with the estimates from coxNet.
+R> #Doing so makes it so that everything is invalid aside from the coefficients.
+R> #In this case, all we need to estimate the absolute risk is the coefficients.
+R> #Std. error would be incorrect here, if we were to draw error bars.
+R> coxNetFitI$coefficients<-nonzero_coef_cox@x
+R> #Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
+R> #posCox=nonzero_covariate_cox$X1
+R> #newDataCoxI=cleanCoxDataI[,posCox]
+R> 
+R> abcoxNetI<-survival::survfit(coxNetFitI,type="breslow",newdata=cleanCoxDataI)
+R> 
+R> abCNI<-abcoxNetI
+R> abCNI$surv=rowMeans(abCNI$surv)
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> fullDataCox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,-c(1)]))
+R> coxFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = fullDataCox)
+R> abcoxFitI<-survival::survfit(coxFitI,newdata=fullDataCox)
+R> abCFI<-abcoxFitI
+R> abCFI$surv<-rowMeans(abCFI$surv)
+R> #cb.ModelNorm=casebase::fitSmoothHazard(V2~.,time="V1",event="V2",
+R>        #                                ratio=rat,data=fullDataCox)
+R> #summary(cb.ModelNorm)#need to figure out how to deal with NAs
+R> #casebaseAbsoluteNorm=casebase::absoluteRisk(cb.ModelNorm,time = seq(1,max(completeData$d.time), 200),newdata = newData , s="lambda.1se",method=c("numerical"))
+R> #cabn=data.frame(casebaseAbsoluteNorm[,1],rowMeans(casebaseAbsoluteNorm[,-c(1)]))
+R> 
+\end{CodeInput}
+\end{CodeChunk}
+
+We used the survival package to calculate the kaplan-meier absolute risk
+function.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> 
+R> coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
+R> abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)
+R> 
+R> #creating a surv object to be used to fit an unadjusted absolute risk curve.
+R> # (Kaplan-Meier risk curve)
+R> km <- survival::Surv(time = completeData$d.time, event = completeData$death)
+R> abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
+\end{CodeInput}
+\end{CodeChunk}
+
+Now that our four cumulative incidence estimations have been calculated,
+we compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase
+decrease the absolute risk by the end of the study, in comparison to
+kaplan-meier.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> 
+R> #get covariates excluding d.time
+R> estimatescb=setDT(as.data.frame(coef(cb.Model)[-c(1,2),1]), keep.rownames = TRUE)[]
+R> estimatescb$Model="casebase"
+R> colnames(estimatescb)<-c("Coefs","Estimate","Model")
+R> estimatescox=setDT(as.data.frame(coef(coxNet)[-c(1),1]), keep.rownames = TRUE)[]
+R> estimatescox$Model="Cox"
+R> colnames(estimatescox)<-c("Coefs","Estimate","Model")
+R> estimates=rbind(estimatescb, estimatescox)
+R> estimates$Estimate[estimates$Estimate==0]=NA# make it so unchosen covariates appear empty
+R> library(ggstance)# for vertical dodge
+R>   ggplot(estimates, aes(x = Estimate, y = Coefs, color = Model)) +
+R>         geom_segment(aes(x = 0, y = Coefs, xend = Estimate, yend = Coefs),size=0.5 ) +
+R>         geom_point(position=ggstance::position_dodgev(height=0.2))
+R> 
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #get covariates excluding d.time
+R> library(data.table)
+R> estimatescb=setDT(as.data.frame(coef(cb.ModelInter)[-c(1,2),1]), keep.rownames = TRUE)[]
+R> estimatescb$Model="casebase"
+R> colnames(estimatescb)<-c("Coefs","Estimate","Model")
+R> estimatescox=setDT(as.data.frame(coef(coxNetI)[-c(1),1]), keep.rownames = TRUE)[]
+R> estimatescox$Model="Cox"
+R> colnames(estimatescox)<-c("Coefs","Estimate","Model")
+R> estimatesI=rbind(estimatescb, estimatescox)
+R> estimatesI$Estimate[estimatesI$Estimate==0]=NA# make it so unchosen covariates appear empty
+R> library(ggstance)# for vertical dodge
+R>   ggplot(estimatesI, aes(x = Coefs, y = Estimate , color = Model)) +
+R>         geom_segment(aes(y = 0, x = Coefs, yend = Estimate, xend = Coefs),size=0.5, position = position_dodge(width = 0.5)) +
+R>         geom_point(position = position_dodge(width = 0.5))
+R>   
+R>     
+\end{CodeInput}
 \end{CodeChunk}
@@
 covariates. As an example that already appeared in the literature,
-case-base sampling was to study vaccination safety, where the exposure
-period was defined as the week following vaccination
+case-base sampling was used to study vaccination safety, where the
+exposure period was defined as the week following vaccination
 \citep{saarela2015case}. Hence, the main covariate of interest,
@@
 #> Model1 Model2 Model3 Model4 
-#>    795    760    729    724
+#>    791    760    727    724
 \end{CodeOutput}
@@
 #>         
-#> 365 0.27
+#> 365 0.28
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> [1] 0.19
+#> [1] 0.18
 \end{CodeOutput}
@@
 
-This report was generated on 2020-02-07 11:07:26 using the following
+This report was generated on 2020-02-21 13:47:19 using the following
 computational environment and dependencies:
@@
 #>  [4] cowplot_1.0.0       pracma_2.2.9        kableExtra_1.1.0   
-#>  [7] glmnet_3.0-1        Matrix_1.2-18       casebase_0.2.1.9001
-#> [10] survival_3.1-8      magrittr_1.5        forcats_0.4.0      
-#> [13] stringr_1.4.0       dplyr_0.8.3         purrr_0.3.3        
-#> [16] readr_1.3.1         tidyr_1.0.0         tibble_2.1.3       
-#> [19] ggplot2_3.2.1       tidyverse_1.3.0    
+#>  [7] data.table_1.12.8   glmnet_3.0-1        Matrix_1.2-18      
+#> [10] casebase_0.2.1.9001 survival_3.1-8      magrittr_1.5       
+#> [13] forcats_0.4.0       stringr_1.4.0       dplyr_0.8.3        
+#> [16] purrr_0.3.3         readr_1.3.1         tidyr_1.0.0        
+#> [19] tibble_2.1.3        ggplot2_3.2.1       tidyverse_1.3.0    
 #> 
@@
 #> [10] yaml_2.2.0          numDeriv_2016.8-1.1 pillar_1.4.3       
-#> [13] backports_1.1.5     lattice_0.20-38     digest_0.6.23      
+#> [13] backports_1.1.5     lattice_0.20-38     digest_0.6.24      
 #> [16] rvest_0.3.5         colorspace_1.4-1    plyr_1.8.5         
@@
 #> [25] lava_1.6.6          mgcv_1.8-31         rticles_0.12       
-#> [28] generics_0.0.2      farver_2.0.1        ellipsis_0.3.0     
+#> [28] generics_0.0.2      farver_2.0.3        ellipsis_0.3.0     
 #> [31] withr_2.1.2         lazyeval_0.2.2      cli_2.0.1          
@@
 #> [37] fs_1.3.1            fansi_0.4.1         MASS_7.3-51.5      
-#> [40] nlme_3.1-143        xml2_1.2.2          tools_3.6.2        
-#> [43] data.table_1.12.8   hms_0.5.2           lifecycle_0.1.0    
-#> [46] munsell_0.5.0       reprex_0.3.0        compiler_3.6.2     
-#> [49] rlang_0.4.2         grid_3.6.2          iterators_1.0.12   
-#> [52] rstudioapi_0.10     labeling_0.3        rmarkdown_1.18     
-#> [55] gtable_0.3.0        codetools_0.2-16    DBI_1.0.0          
-#> [58] reshape2_1.4.3      R6_2.4.1            knitr_1.26         
-#> [61] zeallot_0.1.0       shape_1.4.4         stringi_1.4.4      
-#> [64] Rcpp_1.0.3          vctrs_0.2.1         dbplyr_1.4.2       
-#> [67] tidyselect_0.2.5    xfun_0.11
+#> [40] nlme_3.1-144        xml2_1.2.2          tools_3.6.2        
+#> [43] hms_0.5.2           lifecycle_0.1.0     munsell_0.5.0      
+#> [46] reprex_0.3.0        compiler_3.6.2      rlang_0.4.4        
+#> [49] grid_3.6.2          iterators_1.0.12    rstudioapi_0.10    
+#> [52] labeling_0.3        rmarkdown_1.18      gtable_0.3.0       
+#> [55] codetools_0.2-16    DBI_1.0.0           reshape2_1.4.3     
+#> [58] R6_2.4.1            knitr_1.26          shape_1.4.4        
+#> [61] stringi_1.4.6       Rcpp_1.0.3          vctrs_0.2.3        
+#> [64] dbplyr_1.4.2        tidyselect_0.2.5    xfun_0.11
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> Local:    intro /home/turgeonm/Documents/git_repos/cbpaper
-#> Head:     [f1e32e8] 2020-02-07: add discussion of flexible sampling
+#> Local:    Jesse-Islam-master /home/turgeonm/Documents/git_repos/cbpaper
+#> Head:     [3ec29a1] 2020-02-07: Merge pull request #18 from sahirbhatnagar/intro
 \end{CodeOutput}

### 2020-03-03 analysis/paper/paper.Rmd
expand theoretical details

@@
 eval_implementation <- TRUE
-eval_cs1 <- TRUE
-eval_cs2 <- TRUE
+eval_cs1 <- FALSE
+eval_cs2 <- FALSE
 eval_cs3 <- FALSE
-eval_cs4 <- TRUE
+eval_cs4 <- FALSE
 ```
@@
 
-<!-- Popular survival analysis methods, like Kaplan-Meier and Cox regression, rely on the notion of risk-set sampling. Case-base sampling can be seen as an alternative. -->
-
-As shown by Saarela & Arjas [-@saarela2015non] (and further expanded in Saarela [-@saarela2016case]), writing the likelihood arising from this data-generating mechanism using the framework of non-homogeneous Poisson processes, we eventually reach an expression where each person-moment's contribution is of the form
-$$\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},$$
-where $N(t)$ is the counting process associated with the event of interest, $h(t)$ is the corresponding hazard function, and $\rho(t)$ is the hazard function for the Poisson process associated with case-base sampling.
-This parametric form suggests that we can readily estimate log-hazards of the form $\log(h(t)) = g(t; X)$ using logistic regression, where each observation corresponds to a person moment, the function $g(t; X)$ is linear in a finite number of parameters, and where we treat $-\log(\rho(t))$ as an offset.
+To describe the theoretical foundations of case-base sampling, we use the framework of counting processes.
+In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras; the interested reader can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case].
+First, let $N_{i}(t) \in \{0, 1\}$ be counting processes corresponding to the event of interest for individual $i=1, \ldots,n$.
+For simplicity, we will consider Type I censoring due to the end of follow-up at time $\tau$ (the general case of non-informative censoring is treated in Saarela [-@saarela2016case]).
+We assume a continuous time model, which implies that the counting process jumps are less than or equal to one.
+We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
+$$\lambda_{i}(t) dt = E[dN_{i}(t)].$$
+
+Next, we model the base series sampling mechanism using non-homogeneous Poisson processes $R_i(t) \in \{0, 1, 2, \ldots\}$, with the person-moments where $dR_i(t) = 1$ constituting the base series.
+The process $Q_{i}(t) = R_i(t) + N_{i}(t)$ then counts both the case and base series person-moments contributed by individual $i$.
+This process is typically defined by the user via its hazard function $\rho_i(t)$.
+The process $Q_{i}(t)$ is characterized by $E[dQ_{i}(t)] = \lambda_{i}(t)dt + \rho_i(t)dt$.
+
+If the hazard function $\lambda_{i}(t; \theta)$ is parametrized in terms of $\theta$, we could define an estimator $\hat{\theta}$ by maximization of the likelihood expression
+$$L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^\tau \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)}\left(1 - \lambda_{i}(t;\theta)\right)^{1 - dN_{i}(t)},$$
+where $\prod_{t\in[0,u)}$ represents a product integral from $0$ to $u$. However, the integral over time makes the computation and maximisation of $L_0(\theta)$ challenging.
+
+Case-base sampling allows us to avoid this integral. By conditioning on a sampled person-moment, we get individual contributions of the form
+$$P(dN_{i}(t) \mid dQ_{i}(t) = 1) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.$$
+Therefore, we can define an estimating equation for $\theta$ as follows:
+$$L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.$$
+When a logarithmic link function is used for modeling the hazard function, the above expression is of a logistic regression form with an offset term $\log(1/\rho_i(t))$.
+Note that the sampling units selected in the case-base sampling mechanism are person-moments, rather than individuals, and the parameters to be estimated are hazards or hazard ratios rather than odds or odds ratios.
+Generally, an individual can contribute more than one person-moment, and thus the terms in the product integral are not independent.
+Nonetheless, Saarela [-@saarela2016case] showed that the logarithm of this estimating function has mean zero at the true value $\theta=\theta_0$, and that the resulting estimator $\hat{\theta}$ is asymptotically normally distributed.
 
 In Hanley & Miettinen [-@hanley2009fitting], the authors suggest sampling the base series *uniformly* from the study base.
-In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base.
+In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with hazard equal to $\rho_i(t) = b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base.
 More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism are proportional to the cardiovascular disease event rate given by the Framingham score.
@@
 
-<!-- The `casebase` package fits the family of hazard functions of the form -->
-
-<!-- $$ h(t;X) = \exp[g(t;X)] $$ -->
-<!-- where $t$ denotes time and $X$, the individual's covariate profile.  -->
-
-Let $g(t; X)$ be the linear predictor such that $\log(h(t;X)) = g(t; X)$.
+Let $g(t; X)$ be the linear predictor such that $\log(\lambda(t;X)) = g(t; X)$.
 Different functions of $t$ lead to different parametric hazard models.
@@
 
-$$ \log(h(t; X)) = \beta_0 + \beta_1 X.$$
+$$ \log(\lambda(t; X)) = \beta_0 + \beta_1 X.$$
 
@@
 
-$$ \log(h(t; X)) = \beta_0 + \beta_1 t + \beta_2 X.$$
+$$ \log(\lambda(t; X)) = \beta_0 + \beta_1 t + \beta_2 X.$$
 
@@
 
-$$ \log(h(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. $$
+$$ \log(\lambda(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. $$
 
@@
 
-$$\frac{h_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^Jh_j(t)},$$
-
-where $N_j(t)$ is the counting process associated with the event of type $j$ and $h_j(t)$ is the corresponding hazard function.
+$$\frac{\lambda_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},$$
+
+where $N_j(t)$ is the counting process associated with the event of type $j$ and $\lambda_j(t)$ is the corresponding hazard function.
 As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}

### 2020-03-03 analysis/paper/paper.tex
expand theoretical details

@@
 
-As shown by Saarela \& Arjas \citeyearpar{saarela2015non} (and further
-expanded in Saarela \citeyearpar{saarela2016case}), writing the
-likelihood arising from this data-generating mechanism using the
-framework of non-homogeneous Poisson processes, we eventually reach an
-expression where each person-moment's contribution is of the form
-\[\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},\] where \(N(t)\) is the counting
-process associated with the event of interest, \(h(t)\) is the
-corresponding hazard function, and \(\rho(t)\) is the hazard function
-for the Poisson process associated with case-base sampling. This
-parametric form suggests that we can readily estimate log-hazards of the
-form \(\log(h(t)) = g(t; X)\) using logistic regression, where each
-observation corresponds to a person moment, the function \(g(t; X)\) is
-linear in a finite number of parameters, and where we treat
-\(-\log(\rho(t))\) as an offset.
+To describe the theoretical foundations of case-base sampling, we use
+the framework of counting processes. In what follows, we abuse notation
+slightly and omit any mention of \(\sigma\)-algebras; the interested
+reader can refer to Saarela \& Arjas \citeyearpar{saarela2015non} and
+Saarela \citeyearpar{saarela2016case}. First, let
+\(N_{i}(t) \in \{0, 1\}\) be counting processes corresponding to the
+event of interest for individual \(i=1, \ldots,n\). For simplicity, we
+will consider Type I censoring due to the end of follow-up at time
+\(\tau\) (the general case of non-informative censoring is treated in
+Saarela \citeyearpar{saarela2016case}). We assume a continuous time
+model, which implies that the counting process jumps are less than or
+equal to one. We are interested in modeling the hazard functions
+\(\lambda_{i}(t)\) of the processes \(N_i(t)\), and which satisfy
+\[\lambda_{i}(t) dt = E[dN_{i}(t)].\]
+
+Next, we model the base series sampling mechanism using non-homogeneous
+Poisson processes \(R_i(t) \in \{0, 1, 2, \ldots\}\), with the
+person-moments where \(dR_i(t) = 1\) constituting the base series. The
+process \(Q_{i}(t) = R_i(t) + N_{i}(t)\) then counts both the case and
+base series person-moments contributed by individual \(i\). This process
+is typically defined by the user via its hazard function \(\rho_i(t)\).
+The process \(Q_{i}(t)\) is characterized by
+\(E[dQ_{i}(t)] = \lambda_{i}(t)dt + \rho_i(t)dt\).
+
+If the hazard function \(\lambda_{i}(t; \theta)\) is parametrized in
+terms of \(\theta\), we could define an estimator \(\hat{\theta}\) by
+maximization of the likelihood expression
+\[L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^\tau \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)}\left(1 - \lambda_{i}(t;\theta)\right)^{1 - dN_{i}(t)},\]
+where \(\prod_{t\in[0,u)}\) represents a product integral from \(0\) to
+\(u\). However, the integral over time makes the computation and
+maximisation of \(L_0(\theta)\) challenging.
+
+Case-base sampling allows us to avoid this integral. By conditioning on
+a sampled person-moment, we get individual contributions of the form
+\[P(dN_{i}(t) \mid dQ_{i}(t) = 1) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.\]
+Therefore, we can define an estimating equation for \(\theta\) as
+follows:
+\[L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.\]
+When a logarithmic link function is used for modeling the hazard
+function, the above expression is of a logistic regression form with an
+offset term \(\log(1/\rho_i(t))\). Note that the sampling units selected
+in the case-base sampling mechanism are person-moments, rather than
+individuals, and the parameters to be estimated are hazards or hazard
+ratios rather than odds or odds ratios. Generally, an individual can
+contribute more than one person-moment, and thus the terms in the
+product integral are not independent. Nonetheless, Saarela
+\citeyearpar{saarela2016case} showed that the logarithm of this
+estimating function has mean zero at the true value \(\theta=\theta_0\),
+and that the resulting estimator \(\hat{\theta}\) is asymptotically
+normally distributed.
 
@@
 In terms of Poisson processes, this sampling strategy corresponds
-essentially to a time-homogeneous Poisson process with intensity equal
-to \(b/B\), where \(b\) is the number of sampled observations in the
-base series, and \(B\) is the total population-time for the study base.
-More complex examples are also available; see for example Saarela \&
-Arjas \citeyearpar{saarela2015non}, where the probabilities of the
+essentially to a time-homogeneous Poisson process with hazard equal to
+\(\rho_i(t) = b/B\), where \(b\) is the number of sampled observations
+in the base series, and \(B\) is the total population-time for the study
+base. More complex examples are also available; see for example Saarela
+\& Arjas \citeyearpar{saarela2015non}, where the probabilities of the
 sampling mechanism are proportional to the cardiovascular disease event
@@
 Let \(g(t; X)\) be the linear predictor such that
-\(\log(h(t;X)) = g(t; X)\). Different functions of \(t\) lead to
+\(\log(\lambda(t;X)) = g(t; X)\). Different functions of \(t\) lead to
 different parametric hazard models. The simplest of these models is the
@@
 
-\[ \log(h(t; X)) = \beta_0 + \beta_1 X.\]
+\[ \log(\lambda(t; X)) = \beta_0 + \beta_1 X.\]
 
@@
 
-\[ \log(h(t; X)) = \beta_0 + \beta_1 t + \beta_2 X.\]
+\[ \log(\lambda(t; X)) = \beta_0 + \beta_1 t + \beta_2 X.\]
 
@@
 
-\[ \log(h(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. \]
+\[ \log(\lambda(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. \]
 
@@
 
-\[\frac{h_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^Jh_j(t)},\]
+\[\frac{\lambda_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},\]
 
 where \(N_j(t)\) is the counting process associated with the event of
-type \(j\) and \(h_j(t)\) is the corresponding hazard function. As may
-be expected, this functional form is similar to the terms appearing in
-the likelihood function for multinomial
+type \(j\) and \(\lambda_j(t)\) is the corresponding hazard function. As
+may be expected, this functional form is similar to the terms appearing
+in the likelihood function for multinomial
 regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}
@@
 R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
-R+                                                   data = ERSPC, 
-R+                                                   ratio = 100)
+R>                                                   data = ERSPC, 
+R>                                                   ratio = 100)
 R> 
 R> summary(casebase_exponential)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> 
-#> Call:
-#> glm(formula = formula, family = binomial, data = sampleData)
-#> 
-#> Deviance Residuals: 
-#>    Min      1Q  Median      3Q     Max  
-#> -0.148  -0.148  -0.148  -0.133   3.078  
-#> 
-#> Coefficients:
-#>                       Estimate Std. Error z value            Pr(>|z|)    
-#> (Intercept)            -7.7659     0.0557 -139.46 <0.0000000000000002 ***
-#> ScrArmScreening group  -0.2147     0.0884   -2.43               0.015 *  
-#> ---
-#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-#> 
-#> (Dispersion parameter for binomial family taken to be 1)
-#> 
-#>     Null deviance: 6059  on 54539  degrees of freedom
-#> Residual deviance: 6053  on 54538  degrees of freedom
-#> AIC: 6057
-#> 
-#> Number of Fisher Scoring iterations: 7
-\end{CodeOutput}
-
-\begin{CodeInput}
 R> exp(coef(casebase_exponential)[2])
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> ScrArmScreening group 
-#>                  0.81
-\end{CodeOutput}
-
-\begin{CodeInput}
 R> exp(confint(casebase_exponential)[2,])
 \end{CodeInput}
-
-\begin{CodeOutput}
-#>  2.5 % 97.5 % 
-#>   0.68   0.96
-\end{CodeOutput}
 \end{CodeChunk}
@@
 R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
-R+                                  data = ERSPC, 
-R+                                  ratio = 100)
+R>                                  data = ERSPC, 
+R>                                  ratio = 100)
 R> 
 R> summary(casebase_time)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> 
-#> Call:
-#> glm(formula = formula, family = binomial, data = sampleData)
-#> 
-#> Deviance Residuals: 
-#>    Min      1Q  Median      3Q     Max  
-#> -0.401  -0.162  -0.122  -0.094   3.466  
-#> 
-#> Coefficients:
-#>                       Estimate Std. Error z value            Pr(>|z|)    
-#> (Intercept)            -9.0314     0.1121  -80.53 <0.0000000000000002 ***
-#> Follow.Up.Time          0.2222     0.0145   15.29 <0.0000000000000002 ***
-#> ScrArmScreening group  -0.2320     0.0886   -2.62              0.0089 ** 
-#> ---
-#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-#> 
-#> (Dispersion parameter for binomial family taken to be 1)
-#> 
-#>     Null deviance: 6059  on 54539  degrees of freedom
-#> Residual deviance: 5811  on 54537  degrees of freedom
-#> AIC: 5817
-#> 
-#> Number of Fisher Scoring iterations: 8
-\end{CodeOutput}
-
-\begin{CodeInput}
 R> exp(coef(casebase_time))
-\end{CodeInput}
-
-\begin{CodeOutput}
-#>           (Intercept)        Follow.Up.Time ScrArmScreening group 
-#>               0.00012               1.24882               0.79297
-\end{CodeOutput}
-
-\begin{CodeInput}
 R> exp(confint(casebase_time))
 \end{CodeInput}
-
-\begin{CodeOutput}
-#>                          2.5 %  97.5 %
-#> (Intercept)           0.000096 0.00015
-#> Follow.Up.Time        1.213876 1.28507
-#> ScrArmScreening group 0.665769 0.94256
-\end{CodeOutput}
 \end{CodeChunk}
@@
 R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
-R+                                     data = ERSPC, 
-R+                                     ratio = 100)
+R>                                     data = ERSPC, 
+R>                                     ratio = 100)
 R> 
 R> summary(casebase_splines)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> 
-#> Call:
-#> glm(formula = formula, family = binomial, data = sampleData)
-#> 
-#> Deviance Residuals: 
-#>    Min      1Q  Median      3Q     Max  
-#> -0.405  -0.173  -0.136  -0.085   3.796  
-#> 
-#> Coefficients:
-#>                       Estimate Std. Error z value             Pr(>|z|)    
-#> (Intercept)           -10.2643     0.3159  -32.49 < 0.0000000000000002 ***
-#> bs(Follow.Up.Time)1     4.2983     0.8005    5.37        0.00000007893 ***
-#> bs(Follow.Up.Time)2     2.0945     0.4824    4.34        0.00001412028 ***
-#> bs(Follow.Up.Time)3     4.5544     0.7067    6.44        0.00000000012 ***
-#> ScrArmScreening group  -0.2169     0.0886   -2.45                0.014 *  
-#> ---
-#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-#> 
-#> (Dispersion parameter for binomial family taken to be 1)
-#> 
-#>     Null deviance: 6059.0  on 54539  degrees of freedom
-#> Residual deviance: 5790.5  on 54535  degrees of freedom
-#> AIC: 5800
-#> 
-#> Number of Fisher Scoring iterations: 8
-\end{CodeOutput}
-
-\begin{CodeInput}
 R> exp(coef(casebase_splines))
-\end{CodeInput}
-
-\begin{CodeOutput}
-#>           (Intercept)   bs(Follow.Up.Time)1   bs(Follow.Up.Time)2 
-#>              0.000035             73.578176              8.121029 
-#>   bs(Follow.Up.Time)3 ScrArmScreening group 
-#>             95.053534              0.805017
-\end{CodeOutput}
-
-\begin{CodeInput}
 R> exp(confint(casebase_splines))
 \end{CodeInput}
-
-\begin{CodeOutput}
-#>                           2.5 %     97.5 %
-#> (Intercept)            0.000018   0.000063
-#> bs(Follow.Up.Time)1   15.835008 367.061126
-#> bs(Follow.Up.Time)2    3.252368  21.652009
-#> bs(Follow.Up.Time)3   22.537699 364.141700
-#> ScrArmScreening group  0.675920   0.956823
-\end{CodeOutput}
 \end{CodeChunk}
@@
 
-\begin{CodeChunk}
-\begin{table}[H]
-\centering
-\begin{tabular}{lrlr}
-\toprule
-Model & HR & CI & Width\\
-\midrule
-Cox & 0.80 & (0.67, 0.95) & 0.28\\
-Exponential & 0.81 & (0.68, 0.96) & 0.28\\
-Weibull & 0.79 & (0.67, 0.94) & 0.28\\
-Splines & 0.81 & (0.68, 0.96) & 0.28\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\end{CodeChunk}
-
 As noted above, the usual asymptotic results hold for likelihood ratio
@@
 \end{CodeInput}
-
-\begin{CodeOutput}
-#> Analysis of Deviance Table
-#> 
-#> Model 1: DeadOfPrCa ~ ScrArm + offset(offset)
-#> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
-#>   Resid. Df Resid. Dev Df Deviance            Pr(>Chi)    
-#> 1     54538       6053                                    
-#> 2     54535       5790  3      262 <0.0000000000000002 ***
-#> ---
-#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{CodeOutput}
 \end{CodeChunk}
@@
 R> c("Exp." = AIC(casebase_exponential),
-R+   "Weib." = AIC(casebase_time),
-R+   "Splines" = AIC(casebase_splines))
-\end{CodeInput}
-
-\begin{CodeOutput}
-#>    Exp.   Weib. Splines 
-#>    6057    5817    5800
-\end{CodeOutput}
+R>   "Weib." = AIC(casebase_time),
+R>   "Splines" = AIC(casebase_splines))
+\end{CodeInput}
 \end{CodeChunk}
@@
 R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-R+                    data = bmtcrr)
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{table}
-
-\caption{\label{tab:bmtcrr-cis}Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.}
-\centering
-\begin{tabular}[t]{lrlrl}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
-Covariates & HR & 95\% CI & HR & 95\% CI\\
-\midrule
-Sex & 0.71 & (0.41, 1.24) & 0.68 & (0.39, 1.19)\\
-Disease & 0.55 & (0.31, 1) & 0.52 & (0.29, 0.93)\\
-Phase (CR2 vs. CR1) & 1.18 & (0.47, 2.93) & 1.21 & (0.48, 3.02)\\
-Phase (CR3 vs. CR1) & 1.47 & (0.38, 5.63) & 1.67 & (0.43, 6.5)\\
-Phase (Relapse vs. CR1) & 4.30 & (2, 9.27) & 4.55 & (2.09, 9.9)\\
-\addlinespace
-Source & 1.56 & (0.52, 4.7) & 1.46 & (0.47, 4.54)\\
-Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
-\bottomrule
-\end{tabular}
-\end{table}
-
+R>                    data = bmtcrr)
+\end{CodeInput}
 \end{CodeChunk}
@@
 \end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/bmtcrr-risk-1}
-
-}
-
-\caption{\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups.
-The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:bmtcrr-risk}
-\end{figure}
 \end{CodeChunk}
@@
 R> c("Model1" = AIC(fit1),
-R+   "Model2" = AIC(fit2),
-R+   "Model3" = AIC(fit3),
-R+   "Model4" = AIC(fit4))
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> Model1 Model2 Model3 Model4 
-#>    791    760    727    724
-\end{CodeOutput}
+R>   "Model2" = AIC(fit2),
+R>   "Model3" = AIC(fit3),
+R>   "Model4" = AIC(fit4))
+\end{CodeInput}
 \end{CodeChunk}
@@
 R> absoluteRisk(fit4, newdata = data.frame(exposure = 0),
-R+              time = 365)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#>         
-#> 365 0.28
-\end{CodeOutput}
-
-\begin{CodeInput}
+R>              time = 365)
 R> 
@@
 \end{CodeInput}
-
-\begin{CodeOutput}
-#> [1] 0.18
-\end{CodeOutput}
 \end{CodeChunk}
@@
 
-This report was generated on 2020-02-21 13:47:19 using the following
+This report was generated on 2020-03-03 13:15:36 using the following
 computational environment and dependencies:
@@
 #> other attached packages:
-#>  [1] lubridate_1.7.4     timereg_1.9.4       glue_1.3.1         
-#>  [4] cowplot_1.0.0       pracma_2.2.9        kableExtra_1.1.0   
-#>  [7] data.table_1.12.8   glmnet_3.0-1        Matrix_1.2-18      
-#> [10] casebase_0.2.1.9001 survival_3.1-8      magrittr_1.5       
-#> [13] forcats_0.4.0       stringr_1.4.0       dplyr_0.8.3        
-#> [16] purrr_0.3.3         readr_1.3.1         tidyr_1.0.0        
-#> [19] tibble_2.1.3        ggplot2_3.2.1       tidyverse_1.3.0    
+#>  [1] pracma_2.2.9        kableExtra_1.1.0    data.table_1.12.8  
+#>  [4] glmnet_3.0-1        Matrix_1.2-18       casebase_0.2.1.9001
+#>  [7] survival_3.1-8      magrittr_1.5        forcats_0.4.0      
+#> [10] stringr_1.4.0       dplyr_0.8.3         purrr_0.3.3        
+#> [13] readr_1.3.1         tidyr_1.0.0         tibble_2.1.3       
+#> [16] ggplot2_3.2.1       tidyverse_1.3.0    
 #> 
 #> loaded via a namespace (and not attached):
-#>  [1] httr_1.4.1          VGAM_1.1-2          jsonlite_1.6       
-#>  [4] viridisLite_0.3.0   foreach_1.4.7       modelr_0.1.5       
-#>  [7] assertthat_0.2.1    stats4_3.6.2        cellranger_1.1.0   
-#> [10] yaml_2.2.0          numDeriv_2016.8-1.1 pillar_1.4.3       
-#> [13] backports_1.1.5     lattice_0.20-38     digest_0.6.24      
-#> [16] rvest_0.3.5         colorspace_1.4-1    plyr_1.8.5         
-#> [19] htmltools_0.4.0     pkgconfig_2.0.3     broom_0.5.2        
-#> [22] haven_2.2.0         scales_1.1.0        webshot_0.5.2      
-#> [25] lava_1.6.6          mgcv_1.8-31         rticles_0.12       
-#> [28] generics_0.0.2      farver_2.0.3        ellipsis_0.3.0     
-#> [31] withr_2.1.2         lazyeval_0.2.2      cli_2.0.1          
-#> [34] crayon_1.3.4        readxl_1.3.1        evaluate_0.14      
-#> [37] fs_1.3.1            fansi_0.4.1         MASS_7.3-51.5      
-#> [40] nlme_3.1-144        xml2_1.2.2          tools_3.6.2        
-#> [43] hms_0.5.2           lifecycle_0.1.0     munsell_0.5.0      
-#> [46] reprex_0.3.0        compiler_3.6.2      rlang_0.4.4        
-#> [49] grid_3.6.2          iterators_1.0.12    rstudioapi_0.10    
-#> [52] labeling_0.3        rmarkdown_1.18      gtable_0.3.0       
-#> [55] codetools_0.2-16    DBI_1.0.0           reshape2_1.4.3     
-#> [58] R6_2.4.1            knitr_1.26          shape_1.4.4        
-#> [61] stringi_1.4.6       Rcpp_1.0.3          vctrs_0.2.3        
-#> [64] dbplyr_1.4.2        tidyselect_0.2.5    xfun_0.11
+#>  [1] Rcpp_1.0.3        lubridate_1.7.4   lattice_0.20-40   assertthat_0.2.1 
+#>  [5] digest_0.6.25     foreach_1.4.7     R6_2.4.1          cellranger_1.1.0 
+#>  [9] backports_1.1.5   reprex_0.3.0      stats4_3.6.2      evaluate_0.14    
+#> [13] httr_1.4.1        pillar_1.4.3      rlang_0.4.5       lazyeval_0.2.2   
+#> [17] readxl_1.3.1      rstudioapi_0.10   rticles_0.12      rmarkdown_1.18   
+#> [21] webshot_0.5.2     munsell_0.5.0     broom_0.5.2       compiler_3.6.2   
+#> [25] modelr_0.1.5      xfun_0.11         pkgconfig_2.0.3   shape_1.4.4      
+#> [29] mgcv_1.8-31       htmltools_0.4.0   tidyselect_0.2.5  codetools_0.2-16 
+#> [33] viridisLite_0.3.0 fansi_0.4.1       crayon_1.3.4      dbplyr_1.4.2     
+#> [37] withr_2.1.2       grid_3.6.2        nlme_3.1-144      jsonlite_1.6.1   
+#> [41] gtable_0.3.0      lifecycle_0.1.0   DBI_1.0.0         scales_1.1.0     
+#> [45] cli_2.0.2         stringi_1.4.6     fs_1.3.1          xml2_1.2.2       
+#> [49] generics_0.0.2    vctrs_0.2.3       iterators_1.0.12  tools_3.6.2      
+#> [53] glue_1.3.1        hms_0.5.2         yaml_2.2.1        colorspace_1.4-1 
+#> [57] rvest_0.3.5       VGAM_1.1-2        knitr_1.26        haven_2.2.0
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> Local:    Jesse-Islam-master /home/turgeonm/Documents/git_repos/cbpaper
-#> Head:     [3ec29a1] 2020-02-07: Merge pull request #18 from sahirbhatnagar/intro
+#> Local:    review /home/turgeonm/Documents/git_repos/cbpaper
+#> Head:     [8804f68] 2020-02-21: Merge branch 'Jesse-Islam-master'
 \end{CodeOutput}

### 2020-03-03 analysis/paper/paper.Rmd
update implementation details

@@
 
-eval_introduction <- FALSE
+eval_introduction <- TRUE
 eval_theory <- TRUE
 eval_implementation <- TRUE
-eval_cs1 <- FALSE
-eval_cs2 <- FALSE
+eval_cs1 <- TRUE
+eval_cs2 <- TRUE
 eval_cs3 <- FALSE
-eval_cs4 <- FALSE
+eval_cs4 <- TRUE
 ```
@@
 
-The sampling scheme selected for \code{sampleCaseBase} was described in Hanley and Miettinen [-@hanley2009fitting]: we first sample along the ``person'' axis, proportional to each individual's total follow-up time, and then we sample a moment uniformly over their follow-up time.
+The sampling scheme selected for \code{sampleCaseBase} was described in Hanley & Miettinen [-@hanley2009fitting]: we first sample along the ``person'' axis, proportional to each individual's total follow-up time, and then we sample a moment uniformly over their follow-up time.
 This sampling scheme is equivalent to the following picture: imagine representing the total follow-up time of all individuals in the study along a single dimension, where the follow-up time of the next individual would start exactly when the follow-up time of the previous individual ends.
@@
 If the dataset supplied to \code{fitSmoothHazard} does not inherit from \code{cbData}, then the fitting function starts by calling \code{sampleCaseBase} to generate the base series.
-In other words, the occasional user can bypass \code{sampleCaseBase} altogether and only worry about the fitting function \code{fitSmoothHazard}.
+In other words, users can bypass \code{sampleCaseBase} altogether and only worry about the fitting function \code{fitSmoothHazard}.
 
@@
 The right-hand side can be any combination of the covariates, along with an explicit functional form for the time variable.
-Note that non-proportional hazard models can be achieved at this stage by adding an interaction term involving time.
-The offset term does not need to be specified by the user, as it is automatically added to the formula.
+Note that non-proportional hazard models can be achieved at this stage by adding an interaction term involving time (cf.
+Case Study 4 below).
+The offset term does not need to be specified by the user, as it is automatically added to the formula before calling \code{glm}.
 
@@
 In the case of multiple events, the hazard is fitted via multinomial regression as performed by the \pkg{VGAM} package.
-This package was selected for its ability to fit multinomial regression models with an offset.
+We selected this package for its ability to fit multinomial regression models with an offset.
 
@@
 \begin{equation}\label{eqn:surv}
-S(t) = \exp\left(-\int_0^t h(u;X) du\right).
+S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right).
 \end{equation}
@@
 
-$$ S(t) = \exp\left(-\int_0^t H(u;X) du\right),\qquad H(t;X) = \sum_{j=1}^J h_j(t;X).$$
+$$ S(t) = \exp\left(-\int_0^t \Lambda(u;X) du\right),\qquad \Lambda(t;X) = \sum_{j=1}^J \lambda_j(t;X).$$
 From this, we can derive the event-specific subdensities:
 
-$$ f_j(t) = h_j(t)S(t).$$
+$$ f_j(t) = \lambda_j(t)S(t).$$
 
@@
 $$ CI_j(t) = \int_0^t f_j(u)du.$$
-
-We created \code{absoluteRisk} as an \code{S3} generic, with methods for the different types of outputs of \code{fitSmoothHazard}.
-The method dispatch system of \proglang{R} then takes care of matching the correct output to the correct methodology for calculating the cumulative incidence function, without the user's intervention.
-
-In the following sections, we illustrate these functionalities in the context of three case studies.
+The integrals are computed using either numerical integration (via the trapezoidal rule) or Monte Carlo integration.
+This option is controlled by the argument \code{method} of the \code{absoluteRisk} function.
+
+We created \code{absoluteRisk} as an \code{S3} generic, with methods that match the different types of outputs of \code{fitSmoothHazard}.
+The method dispatch system of \proglang{R} then takes care of matching the correct output to the correct methodology for calculating the cumulative incidence function, without requiring any intervention from the user.
+
+In the following sections, we illustrate these functionalities in the context of four case studies.
 
@@
 For example, if we look at the Weibull distribution as parametrised in \code{stats::pweibull}, the log-hazard function is given by
-$$ \log h(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,$$
+$$ \log \lambda(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,$$
 where $\alpha,\beta$ are shape and scale parameters, respectively.

### 2020-03-03 analysis/paper/paper.tex

--- 2020-02-21 analysis/paper/paper.tex
@@
 The sampling scheme selected for \code{sampleCaseBase} was described in
-Hanley and Miettinen \citeyearpar{hanley2009fitting}: we first sample
+Hanley \& Miettinen \citeyearpar{hanley2009fitting}: we first sample
 along the ``person'' axis, proportional to each individual's total
@@
 fitting function starts by calling \code{sampleCaseBase} to generate the
-base series. In other words, the occasional user can bypass
-\code{sampleCaseBase} altogether and only worry about the fitting
-function \code{fitSmoothHazard}.
+base series. In other words, users can bypass \code{sampleCaseBase}
+altogether and only worry about the fitting function
+\code{fitSmoothHazard}.
 
@@
 for the time variable. Note that non-proportional hazard models can be
-achieved at this stage by adding an interaction term involving time. The
-offset term does not need to be specified by the user, as it is
-automatically added to the formula.
+achieved at this stage by adding an interaction term involving time
+(cf.~Case Study 4 below). The offset term does not need to be specified
+by the user, as it is automatically added to the formula before calling
+\code{glm}.
 
@@
 In the case of multiple events, the hazard is fitted via multinomial
-regression as performed by the \pkg{VGAM} package. This package was
-selected for its ability to fit multinomial regression models with an
+regression as performed by the \pkg{VGAM} package. We selected this
+package for its ability to fit multinomial regression models with an
 offset.
@@
 identity \begin{equation}\label{eqn:surv}
-S(t) = \exp\left(-\int_0^t h(u;X) du\right).
+S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right).
 \end{equation} The integral is computed using either the
@@
 
-\[ S(t) = \exp\left(-\int_0^t H(u;X) du\right),\qquad H(t;X) = \sum_{j=1}^J h_j(t;X).\]
+\[ S(t) = \exp\left(-\int_0^t \Lambda(u;X) du\right),\qquad \Lambda(t;X) = \sum_{j=1}^J \lambda_j(t;X).\]
 From this, we can derive the event-specific subdensities:
 
-\[ f_j(t) = h_j(t)S(t).\]
+\[ f_j(t) = \lambda_j(t)S(t).\]
 
@@
 
-\[ CI_j(t) = \int_0^t f_j(u)du.\]
-
-We created \code{absoluteRisk} as an \code{S3} generic, with methods for
-the different types of outputs of \code{fitSmoothHazard}. The method
-dispatch system of \proglang{R} then takes care of matching the correct
-output to the correct methodology for calculating the cumulative
-incidence function, without the user's intervention.
+\[ CI_j(t) = \int_0^t f_j(u)du.\] The integrals are computed using
+either numerical integration (via the trapezoidal rule) or Monte Carlo
+integration. This option is controlled by the argument \code{method} of
+the \code{absoluteRisk} function.
+
+We created \code{absoluteRisk} as an \code{S3} generic, with methods
+that match the different types of outputs of \code{fitSmoothHazard}. The
+method dispatch system of \proglang{R} then takes care of matching the
+correct output to the correct methodology for calculating the cumulative
+incidence function, without requiring any intervention from the user.
 
 In the following sections, we illustrate these functionalities in the
-context of three case studies.
+context of four case studies.
 
@@
 \code{stats::pweibull}, the log-hazard function is given by
-\[ \log h(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,\]
+\[ \log \lambda(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,\]
 where \(\alpha,\beta\) are shape and scale parameters, respectively.
@@
 
-This report was generated on 2020-02-21 13:47:19 using the following
+This report was generated on 2020-03-03 13:34:41 using the following
 computational environment and dependencies:
@@
 #> loaded via a namespace (and not attached):
-#>  [1] httr_1.4.1          VGAM_1.1-2          jsonlite_1.6       
+#>  [1] httr_1.4.1          VGAM_1.1-2          jsonlite_1.6.1     
 #>  [4] viridisLite_0.3.0   foreach_1.4.7       modelr_0.1.5       
 #>  [7] assertthat_0.2.1    stats4_3.6.2        cellranger_1.1.0   
-#> [10] yaml_2.2.0          numDeriv_2016.8-1.1 pillar_1.4.3       
-#> [13] backports_1.1.5     lattice_0.20-38     digest_0.6.24      
-#> [16] rvest_0.3.5         colorspace_1.4-1    plyr_1.8.5         
+#> [10] yaml_2.2.1          numDeriv_2016.8-1.1 pillar_1.4.3       
+#> [13] backports_1.1.5     lattice_0.20-40     digest_0.6.25      
+#> [16] rvest_0.3.5         colorspace_1.4-1    plyr_1.8.6         
 #> [19] htmltools_0.4.0     pkgconfig_2.0.3     broom_0.5.2        
@@
 #> [28] generics_0.0.2      farver_2.0.3        ellipsis_0.3.0     
-#> [31] withr_2.1.2         lazyeval_0.2.2      cli_2.0.1          
+#> [31] withr_2.1.2         lazyeval_0.2.2      cli_2.0.2          
 #> [34] crayon_1.3.4        readxl_1.3.1        evaluate_0.14      
@@
 #> [43] hms_0.5.2           lifecycle_0.1.0     munsell_0.5.0      
-#> [46] reprex_0.3.0        compiler_3.6.2      rlang_0.4.4        
+#> [46] reprex_0.3.0        compiler_3.6.2      rlang_0.4.5        
 #> [49] grid_3.6.2          iterators_1.0.12    rstudioapi_0.10    
@@
 \begin{CodeOutput}
-#> Local:    Jesse-Islam-master /home/turgeonm/Documents/git_repos/cbpaper
-#> Head:     [3ec29a1] 2020-02-07: Merge pull request #18 from sahirbhatnagar/intro
+#> Local:    review /home/turgeonm/Documents/git_repos/cbpaper
+#> Head:     [96bfdb4] 2020-03-03: expand theoretical details
 \end{CodeOutput}

### 2020-03-05 analysis/paper/paper.Rmd
Intermediate step before I restucture and remove things from tables
CS3

@@
 To demonstrate this capability, we examine the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [@knaus1995support].
-The SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital.
+The SUPPORT dataset tracks death for individuals who are considered seriously ill within 5 hospitals in the United States of America.
 Imputation was conducted using the \pkg{mice} package in \proglang{R} with default settings [@mice].
-Before imputation, there were a total of 1000 individuals and 35 variables.
-After imputation, we have 690 individuals.
-We lose individuals due to coliniarity.
-If a pair of covariates are deemed colinear to a certain degree, they are excluded from any imputations.
-Here, only edu  income  avtisst*  race  ph*  glucose*  adlp* were imputed.
-Of these, 66.09% experienced the event of interest.
-<!--The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.--> A description of each variable can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}.
-The original data was obtained from the Department of Biostatistics at Vanderbilt University.
+Before imputation, there were a total of 9105 individuals and 47 variables.
+Of those variables, a few were removed before imputation.
+We removed three response variables: hospital charges, patient ratio of costs to charges and patient micro-costs.
+We also removed functional disability and income as they are ordinal covariates, because ordinal variables are treated as factors in our current model.
+Finally, we removed 8 covariates relating to the results of previous findings: we removed SUPPORT day 3 physiology score (sps), 	APACHE III day 3 physiology score (aps),	SUPPORT model 2-month survival estimate, SUPPORT model 6-month survival estimate, Physician's 2-month survival estimate for pt., Physician's 6-month survival estimate for pt., Patient had Do Not Resuscitate (DNR) order and Day of DNR order (<0 if before study).
+Of these, sps and aps were added on after imputation, as they were missing only 1 observation and permitted us to compare findings between our model and their's.
+
+First we impute manually using recommended normal values for physiological measures [@harrell_2020].
+Next, we impute a single dataset using \pkg{mice} with default settings.
+After imputation, we noted that the covariate for surrogate activities of daily living was not imputed.
+This is due to colinearity between the other two covariates for activities of daily living.
+As such, surrogate activites of daily living was removed from our model.
+After these imputations, we have a total of 9104 observations and 30 variables, two of which are sps and aps, which will work as a baseline for our comparison to previous findings.
+This dataset is available  Its worth noting that these scores were developed for a period of 180 days, which makes our comparison over 5.56 years exploratory.
+Also, our single imputation introduces bias in the data.
+As the goal here is purely demonstrative, this bias will be consistent with any of the following models.
+As such, their comparison is still valid, though any interpretation requires an awareness of said bias.
+<!--The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.--> A description of used variables can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}.
+The original data was obtained from the Department of Biostatistics at Vanderbilt University [@harrell_2020].
 The imputed dataset is available as part of the \pkg{casebase} package.
@@
 race          &                                          & 5               & integer          & 5            \\
+sps 	        & support physiology score day 3           &	               & double           &	1            \\
+aps           & APS III no coma, imp bun,uout for ph1,D3 &	               & double	          & 1            \\
+hday	        & Day in Hospital at Study Admit	         &                 & double           & 0            \\	
+diabetes      &	Diabetes (Com 27-28, Dx 73)	             &                 & double	          & 0            \\
+dementia      & Dementia (Comorbidity 6)                 &	               & double	          & 0            \\
+ca		        & Cancer State                             & 3               & integer	        & 0            \\
 meanbp        & Mean Arterial Blood Pressure Day 3       &                 & double           & 0            \\
@@
 urine         & Urine Output Day 3                       &                 & double           & 517          \\
-adlp          & ADL Patient Day 3                        &                 & double           & 634          \\
+adlp          & Activiites of Daily Living (ADL) Patient Day 3                        &                 & double           & 634          \\
 adls          & ADL Surrogate Day 3                      &                 & double           & 310          \\
@@
          & hispanic                                      \\
+ca       & Yes                                           \\
+         & No                                            \\
+         & Metastatic                                    \\
 sfdm2    & no(M2 and SIP pres)                           \\

### 2020-03-05 analysis/paper/paper.Rmd
In process of changing my results
loaded data, now ready to process

@@
 We removed three response variables: hospital charges, patient ratio of costs to charges and patient micro-costs.
+Next, we removed hospital death as it was directly informative of our event of interest, death.
 We also removed functional disability and income as they are ordinal covariates, because ordinal variables are treated as factors in our current model.
@@
 sex           &                                          & 2               & integer          & 0            \\
-hospdead      & Death in Hospital                        &                 & double           & 0            \\
 slos          & Days from Study Entry to Discharge       &                 & double           & 0            \\
@@
 scoma         & SUPPORT Coma Score based on Glasgow D3   &                 & double           & 0            \\
-charges       & Hospital Charges                         &                 & double           & 25           \\
-totcst        & Total RCC cost                           &                 & double           & 105          \\
-totmcst       & Total micro-cost                         &                 & double           & 372          \\
 avtisst       & Average TISS, Days 3-25                  &                 & double           & 6            \\
@@
 adlp          & Activiites of Daily Living (ADL) Patient Day 3 &           & double           & 634          \\
-adls          & ADL Surrogate Day 3                      &                 & double           & 310          \\
-sfdm2         &                                          & 5               & integer          & 159          \\
 adlsc         & Imputed ADL Calibrated to Surrogate      &                 & double           & 0           
@@
          & Cancer                                        \\
-income   & under \$11k                                   \\
-         & $11-$25k                                      \\
-         & $25-$50k                                      \\
          & \textgreater{}\$50k                           \\
@@
          & Metastatic                                    \\
-sfdm2    & no(M2 and SIP pres)                           \\
-         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
-         & SIP\textgreater{}=30                          \\
-         & Coma or Intub                                 \\
-         & \textless{}2 mo. follow-up                   
 \end{tabular}
@@
 The population-time plot for the SUPPORT data appears in Figure \ref{fig:support-poptime}.
+As we can see, the incidence density is not constant.
 
 ```{r support-poptime, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data."}
-data(support)
-support_popTime <- popTime(support,
-                           time = "d.time", event = "death",percentile_number = 0.31)
-plot(support_popTime)
-```
-
-Before we proceed with the analysis, we first remove hospital death as a covariate, as this is directly informative of death.
-All covariates will be used to estimate the hazard function.
-The hazard function will be used to estimate the cumulative incidence of each individual in the study.
-Then, the average cumulative incidience curve for all observations will be used for our comparison.
-
-```{r supportSetUp, echo = FALSE, eval = eval_cs3}
-completeData=casebase::support
-#remove hospital dead
-completeData<-completeData[,-c(4)]
-ratio=10  #universal ratio for eval_cs3
-#keep one individual out, at random, for which we will develop an 
-#absolute risk curve.
-#sam=sample(1:nrow(completeData), 1)
-#Set workingData to the remaining individuals in the data
-#workingData=completeData[-c(sam),]
+
+completeData<-casebase::support
 workingCompleteData=model.matrix(death~ .-d.time,data=completeData)[,-c(1)]#remove intercept
@@
 y=data.matrix(completeData[,c(2,5)])
-```
+
+support_popTime <- popTime(support,
+                           time = "d.time", event = "death",percentile_number = 0.31)
+plot(support_popTime)
+```
+
+Our next step is to get the hazard function.
+All covariates will be used to estimate the hazard function.
+The hazard function will be used to estimate the cumulative incidence of each individual in the study.
+Then, the average cumulative incidience curve for all observations will be used for our comparison.
 
@@
 Here, we used lasso penalization by setting alpha to 1.
-
-```{r cbHazardInteractions}, echo = TRUE, eval = eval_cs3}
-xInteractions=x*y[,2]
-colnames(xInteractions) <- paste("d.time*", colnames(xInteractions), sep = "")
-xFull=cbind(x,xInteractions)
-cb.ModelInter=casebase::fitSmoothHazard.fit(xFull,y,family="glmnet",time="d.time",event="death",
-                                       formula_time = ~d.time,alpha=1,ratio=rat)
-```
-Then, we take our model and use the absoluteRisk function to integrate and retrieve our absolute risk function.
-
-```{r CBabsoluteInteracts, echo = TRUE, eval = eval_cs3}
-#Estimating the absolute risk curve using the newData parameter.
-casebaseAbsoluteInter=casebase::absoluteRisk(cb.ModelInter,time = seq(1,max(completeData$d.time), 400),newdata = xFull , s="lambda.1se",method=c("numerical"))
-
-cabI=data.frame(casebaseAbsoluteInter[,1],rowMeans(casebaseAbsoluteInter[,-c(1)]))
-```
-We will compare this curve to a regularized version of cox regression, and a kaplan-meier survival curve.
-The cox regression absolute risk curve with lasso penalization required two extra steps to retrieve.
-Coxnet from the glmnet package has tools to fit a regularized cox model, but requires a survival object skeleton of the selected variables, so that the survival package tools can be used to retrive the absolute risk.
-We handle this by fitting a regularized cox model with the glmnet package, and a cox model from the survival package with the selected variables from the regularized fit.
-This second model serves as a skeleton that permits the use of survival package functions.
-The coefficients from the regularized model will replace the coefficients in our skeleton.
-the resulting model will have the correct coefficients when integrating, but will have an incorrect standard error.
-For the purposes of this case study, we are only interested in the absolute risk curve itself and not the standard error.
-
-```{r coxHazAbsoluteInteracts, echo = TRUE, eval = eval_cs3}
-#Create u and xCox, which will be used when fitting Cox with glmnet.
-u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
-
-xCoxI=as.matrix(cbind(data.frame("(Intercept)"=rep(1,each=690)),xFull))
-colnames(xCoxI)[1]="(Intercept)"
-
-#hazard for cox using glmnet
-coxNetI=glmnet::cv.glmnet(x=xCoxI,y=u, family="cox",alpha=1)
-#convergence demonstrated in plot
-#plot(coxNet)
-#taking the coefficient estimates for later use
-nonzero_covariate_cox <- predict(coxNetI, type = "nonzero", s = "lambda.1se")
-nonzero_coef_cox <- coef(coxNetI, s = "lambda.1se")
-#creating a new dataset that only contains the covariates chosen through glmnet.
-#cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
-cleanCoxDataI<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,nonzero_covariate_cox$X1]))
-#newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
-#fitting a cox model using regular estimation, however we will not keep it.
-#this is used more as an object place holder.
-coxNetFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxDataI)
-#The coefficients of this object will be replaced with the estimates from coxNet.
-#Doing so makes it so that everything is invalid aside from the coefficients.
-#In this case, all we need to estimate the absolute risk is the coefficients.
-#Std. error would be incorrect here, if we were to draw error bars.
-coxNetFitI$coefficients<-nonzero_coef_cox@x
-#Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
-#posCox=nonzero_covariate_cox$X1
-#newDataCoxI=cleanCoxDataI[,posCox]
-
-abcoxNetI<-survival::survfit(coxNetFitI,type="breslow",newdata=cleanCoxDataI)
-
-abCNI<-abcoxNetI
-abCNI$surv=rowMeans(abCNI$surv)
-```
-
-```{r unpenalizedCox, echo = TRUE, eval = eval_cs3}
-fullDataCox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,-c(1)]))
-coxFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = fullDataCox)
-abcoxFitI<-survival::survfit(coxFitI,newdata=fullDataCox)
-abCFI<-abcoxFitI
-abCFI$surv<-rowMeans(abCFI$surv)
-#cb.ModelNorm=casebase::fitSmoothHazard(V2~.,time="V1",event="V2",
-       #                                ratio=rat,data=fullDataCox)
-#summary(cb.ModelNorm)#need to figure out how to deal with NAs
-#casebaseAbsoluteNorm=casebase::absoluteRisk(cb.ModelNorm,time = seq(1,max(completeData$d.time), 200),newdata = newData , s="lambda.1se",method=c("numerical"))
-#cabn=data.frame(casebaseAbsoluteNorm[,1],rowMeans(casebaseAbsoluteNorm[,-c(1)]))
-
-```
-We used the survival package to calculate the kaplan-meier absolute risk function.
-
-```{r fitKM+abRisk, echo = TRUE, eval = eval_cs3}
-
-coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
-abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)
-
-#creating a surv object to be used to fit an unadjusted absolute risk curve.
-# (Kaplan-Meier risk curve)
-km <- survival::Surv(time = completeData$d.time, event = completeData$death)
-abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
-```
-
-Now that our four cumulative incidence estimations have been calculated, we compare them all in Figure \ref{fig:abSC}.
-Both coxnet and casebase decrease the absolute risk by the end of the study, in comparison to kaplan-meier.
-```{r abSupportComparison, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
-abCFup<-abcoxFitup
-abCFup$surv<-rowMeans(abCFup$surv)
-# A plot to compare all three Absolute risk (Commulative Incidence)
-plot(cabI[,1],cabI[,2],type='l',col="black",lwd=3,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
-lines(abCFI,col="red",fun="event",lwd=3,conf.int = FALSE)
-lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
-lines(abCNI,col="purple",fun="event",lwd=3,conf.int = FALSE)
-legend("bottomright", 
-       legend = c( "Casebase (Lasso+linear)","semi-parametric Coxnet","KM curve","cox"), 
-       col = c("black","red","Blue","purple"),
-       lty = c(1, 1, 1), 
-       bg = "gray90")
-```
-
-```{r covariatesRegularizedLollipopPlot,eval=FALSE}
-
-#get covariates excluding d.time
-estimatescb=setDT(as.data.frame(coef(cb.Model)[-c(1,2),1]), keep.rownames = TRUE)[]
-estimatescb$Model="casebase"
-colnames(estimatescb)<-c("Coefs","Estimate","Model")
-estimatescox=setDT(as.data.frame(coef(coxNet)[-c(1),1]), keep.rownames = TRUE)[]
-estimatescox$Model="Cox"
-colnames(estimatescox)<-c("Coefs","Estimate","Model")
-estimates=rbind(estimatescb, estimatescox)
-estimates$Estimate[estimates$Estimate==0]=NA# make it so unchosen covariates appear empty
-library(ggstance)# for vertical dodge
-  ggplot(estimates, aes(x = Estimate, y = Coefs, color = Model)) +
-        geom_segment(aes(x = 0, y = Coefs, xend = Estimate, yend = Coefs),size=0.5 ) +
-        geom_point(position=ggstance::position_dodgev(height=0.2))
-
-```
-
-```{r covariatesLolliInteractions,eval=eval_cs3}
-#get covariates excluding d.time
-library(data.table)
-estimatescb=setDT(as.data.frame(coef(cb.ModelInter)[-c(1,2),1]), keep.rownames = TRUE)[]
-estimatescb$Model="casebase"
-colnames(estimatescb)<-c("Coefs","Estimate","Model")
-estimatescox=setDT(as.data.frame(coef(coxNetI)[-c(1),1]), keep.rownames = TRUE)[]
-estimatescox$Model="Cox"
-colnames(estimatescox)<-c("Coefs","Estimate","Model")
-estimatesI=rbind(estimatescb, estimatescox)
-estimatesI$Estimate[estimatesI$Estimate==0]=NA# make it so unchosen covariates appear empty
-library(ggstance)# for vertical dodge
-  ggplot(estimatesI, aes(x = Coefs, y = Estimate , color = Model)) +
-        geom_segment(aes(y = 0, x = Coefs, yend = Estimate, xend = Coefs),size=0.5, position = position_dodge(width = 0.5)) +
-        geom_point(position = position_dodge(width = 0.5))
-
-```
 

### 2020-03-06 analysis/paper/paper.Rmd
NOT COMPLETE
to do:
fix casebase absoluterisk for glmnet. 
fix ??? tags
add lines to dot whisker plot
Need to removeRcolorBrewer palletes

@@
 library(pracma)
+library(RColorBrewer)
 options(digits = 2, scipen = 999)
@@
 
-eval_introduction <- TRUE
-eval_theory <- TRUE
-eval_implementation <- TRUE
-eval_cs1 <- TRUE
-eval_cs2 <- TRUE
-eval_cs3 <- FALSE
-eval_cs4 <- TRUE
+eval_introduction <- FALSE
+eval_theory <- FALSE
+eval_implementation <- FALSE
+eval_cs1 <- FALSE
+eval_cs2 <- FALSE
+eval_cs3 <- TRUE
+eval_cs4 <- FALSE
 ```
@@
 ```{r support-poptime, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data."}
-
+support<-casebase::support
+
+support_popTime <- popTime(support,
+                           time = "d.time", event = "death",percentile_number = 0.31)
+plot(support_popTime)
+```
+
+Before generating an absolute risk curve using regularized logistic regression, we will explore three options as a baseline comparison, beyond a simple Kaplan-meier curve.
+For all the following analyses in this case study, we will use 80% of the observations to train our models, while 20% will be used to generate absolute risk curves, all of which will be averaged to demonstrate each model's predictive ability.
+We will then compare each modeling using the IPA score (???NEED REFERENCE HERE???) We will examine three cox regression scenarios: sps only, aps only and all covariates excluding sps and aps.
+The original study compared these scores over a span of 180 days, which may not be comparable over the full 5.56 years.
+Over 5.56 years, \ref{fig:baselines} demonstrates that ???BEST MODEL BASED ON IPA???
+has the best calibration and discrimination with an IPA score of ???INSERT ACTUAL BEST SCORE???
+.
+??BEST MODEL??
+will be used in the following comparison to regularized cox, regularized logistic regression and the unadjusted Kaplan-Meier curve.
+
+```{r baselinePlotsPrep, eval=eval_cs3}
 completeData<-casebase::support
@@
 
-support_popTime <- popTime(support,
-                           time = "d.time", event = "death",percentile_number = 0.31)
-plot(support_popTime)
-```
-
-Our next step is to get the hazard function.
-All covariates will be used to estimate the hazard function.
-The hazard function will be used to estimate the cumulative incidence of each individual in the study.
-Then, the average cumulative incidience curve for all observations will be used for our comparison.
-
-Our main objective is to compute the absolute risk of death for a given set of covariates, using regularization when fitting our model.
-First, we fit a smooth hazard to the data; for the sake of this example, we opted for a linear term for time.
+fullDataCox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),x))
+train_index <- sample(1:nrow(fullDataCox), 0.8 * nrow(fullDataCox))
+test_index <- setdiff(1:nrow(fullDataCox), train_index)
+
+train<-fullDataCox[train_index,]
+test<-fullDataCox[test_index,]
+coxSPS<- survival::coxph(Surv(time=V1,event=V2) ~ sps, data = train)
+abcoxSPS<-survival::survfit(coxSPS,newdata = test)
+
+coxAPS<- survival::coxph(Surv(time=V1,event=V2) ~ aps, data = train)
+abcoxAPS<-survival::survfit(coxAPS,newdata = test)
+
+coxFull<- survival::coxph(Surv(time=V1,event=V2) ~ .-sps -aps, data = train)
+abcoxFull<-survival::survfit(coxAPS,newdata = test)
+#library(prodlim)
+
+KM<- survival::coxph(Surv(time=V1,event=V2) ~ 1, data = test)
+abKM<-survival::survfit(KM)
+#abKM <- prodlim(Hist(V1,V2)~1,data=test) minimal difference
+abKMcompare<-cbind(abKM$time[abKM$time<=1631],1-abKM$surv[abKM$time<=1631])
+colnames(abKMcompare)<-c("Time","KM")
+
+baseCoxComparisonData<-cbind(abcoxSPS$time[abcoxSPS$time<=1631],1-rowMeans(abcoxSPS$surv[abcoxSPS$time<=1631,]),1-rowMeans(abcoxAPS$surv[abcoxSPS$time<=1631,]),1-abKM$surv,1-rowMeans(abcoxFull$surv[abcoxSPS$time<=1631,]))
+colnames(baseCoxComparisonData)<-c("Time","SPS","APS","Unadjusted","Full")
+myColors <- brewer.pal(5,"Dark2")
+FigureWidth='90%'
+
+```
+
+```{r baselines, echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:baselines} AbsoluteRisk curves from hazard models.
+Pink: All covariates without SPS nor APS.
+Green: APS only model.
+Blue: SPS only model.
+Purple: unadjusted curve.
+All three models underestimate the risk in comparison to the unadjusted curve."}
+ggplot(as.data.frame(baseCoxComparisonData), aes(Time)) + 
+  geom_line(aes(y = SPS, colour = "SPS"),lwd=1) + 
+  geom_line(aes(y = APS, colour = "APS"),lwd=1) +
+  geom_step(data=as.data.frame(abKMcompare),mapping=aes(x=Time,y=KM,color="Unadjusted"),lwd=1)+
+  geom_line(aes(y = Full, col = "All Covariates without SPS APS"),lwd=1)+
+  labs(y= "Probability of Death", x = "Survival time (Days)")
+```
+
+To get the absolute risk following regularized cox regression, this required that we work around limitations in existing packages.
+To make use of the \pkg{survival} package, we first get our hazard coefficients using \pkg{glmnet}'s coxnet implementation for lasso penalization [@regpathcox].
+Next, we fit a cox model using the survival package with the selected variables only, from our previous fit.
+Finally, we replaced the coefficients in the survival object with those from our glmnet fit, allowing us to make use of the absolute risk tools available in the \pkg{survival} package.
+
+For our penalized logistic regression model, we opted for a logistic function of time, resulting in a Weibull distribution assumption.
 As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model.
 For the SUPPORT dataset, all factor variables were converted into dummy variables.
-Here, we used lasso penalization by setting alpha to 1.
+Here, we used lasso penalization by setting alpha to 1, leading to our smooth hazard function, depicted in the ???
+curve in \ref{fig:cs3Final}.
+??? I AM STILL PENALIZING TIME ???
+
+We then take our hazard and use it to predict the absolute risk of each individual in our test set, over time.
+These individuals are then average resulting in the ???curve in \ref{fig:cs3Final}.
+???SAME BUG IS BACK???
+
+```{r supportSetUp, echo = FALSE, eval = eval_cs3}
+
+ratio=10  #universal ratio for TRUE
+#keep one individual out, at random, for which we will develop an 
+#absolute risk curve.
+#sam=sample(1:nrow(completeData), )
+#Set workingData to the remaining individuals in the data
+#workingData=completeData[-c(sam),]
+
+#x and y will be used to fit the casebase model under glmnet.
+#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
+y=data.matrix(train[,c(2,1)])
+colnames(y)<-c("death","d.time")
+x=data.matrix(train[,-c(2,1)])
+newData=data.matrix(test[,-c(2,1)])
+```
+
+```{r coxHazAbsolute, echo = FALSE,eval= eval_cs3 }
+#Create u and xCox, which will be used when fitting Cox with glmnet.
+
+u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
+
+xCox=as.matrix(cbind(data.frame("(Intercept)"=rep(1,each=length(x[,1]))),x))
+colnames(xCox)[2]="(Intercept)"
+
+#hazard for cox using glmnet
+coxNet=glmnet::cv.glmnet(x=x,y=u, family="cox",alpha=1,standardize = TRUE)
+#convergence demonstrated in plot
+#plot(coxNet)
+#taking the coefficient estimates for later use
+nonzero_covariate_cox <- predict(coxNet, type = "nonzero", s = "lambda.1se")
+nonzero_coef_cox <- coef(coxNet, s = "lambda.1se")
+#creating a new dataset that only contains the covariates chosen through glmnet.
+#cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
+cleanCoxData<-as.data.frame(cbind(d.time=as.numeric(y[,2]),death=as.numeric(y[,1]),x[,nonzero_covariate_cox$X1]))
+#newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
+#fitting a cox model using regular estimation, however we will not keep it.
+#this is used more as an object place holder.
+coxNet <- survival::coxph(Surv(time=d.time,event=death) ~ ., data = cleanCoxData)
+
+#extract coefficients for lollipop plot later
+estimatescoxNet=setDT(as.data.frame(coef(coxNet)), keep.rownames = TRUE)[]
+#estimatescoxNet$std.error<-0
+estimatescoxNet$Model="CoxNet"
+colnames(estimatescoxNet)<-c("term","estimate","model")
+
+#The coefficients of this object will be replaced with the estimates from coxNet.
+#Doing so makes it so that everything is invalid aside from the coefficients.
+#In this case, all we need to estimate the absolute risk is the coefficients.
+#Std. error would be incorrect here, if we were to draw error bars.
+coxNet$coefficients<-nonzero_coef_cox@x
+#Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
+#posCox=nonzero_covariate_cox$X1
+#newDataCoxI=cleanCoxDataI[,posCox]
+
+abcoxNet<-survival::survfit(coxNet,type="breslow",newdata=as.data.frame(newData[,nonzero_covariate_cox$X1]))
+
+abCN=as.data.frame(cbind(abcoxNet$time, 1-rowMeans(abcoxNet$surv)))
+colnames(abCN)<-c("Time","Absolute risk")
+abCN$Model="Coxnet"
+abCN=abCN[abCN$Time<=max(abKMcompare[,1]),]
+#plot(abCN[,1],1-abCN[,2])
+```
+
+```{r cbHazard}, echo = FALSE, eval = eval_cs3}
+
+cbFull=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",formula_time = ~log(d.time),alpha=1,ratio=ratio,standardize = TRUE)
+```
+
+```{r CBabsolute, echo = FALSE, eval = eval_cs3}
+#Estimating the absolute risk curve using the newData parameter.
+abCbFull=casebase::absoluteRisk(cbFull,time = seq(1,max(abKMcompare[,1]), 25),newdata = newData , s="lambda.1se",method=c("numerical"))
+abCB=data.frame(abCbFull[,1],rowMeans(abCbFull[,-c(1)]))
+
+colnames(abCB)<-c("Time","Absolute risk")
+abCB$Model="Penalized logistic"
+
+ab=rbind(abCB,abCN)
+
+```
+
+```{r include=TRUE,eval=eval_cs3,fig.cap="\\label{fig:baselines} HERES SOME TEXT ABT IT"}
+ggplot(as.data.frame(ab), aes(Time)) + 
+    geom_line(data=as.data.frame(baseCoxComparisonData),mapping=aes(y = Full, colour = "Cox"),lwd=2)+
+  geom_line(aes(y = `Absolute risk`, colour = Model),lwd=2)+
+  geom_step(data=as.data.frame(abKMcompare),mapping=aes(x=Time,y=KM,color="Unadjusted"),lwd=2)+
+  scale_colour_manual(name = "Models",values = myColors[c(4,1,3,2)])+
+  labs(y= "Probability of Death", x = "Survival time (Days)")
+```
+
+???SHORT SPIEL ABOUT IPA BETWEEN SEMI_PARAMETRIC AND PARAMETRIC which leads to conclusion of best overall model. This refers to the COEFPLOT in a following figure???
+
+```{r coefplots, eval=eval_cs3}
+
+library(dotwhisker)
+library(broom)
+library(dplyr)
+
+estimatescb=setDT(as.data.frame(coef(cbFull)[-c(1,2),1]), keep.rownames = TRUE)[]
+#estimatescb$std.error<-0
+estimatescb$Model="casebase"
+colnames(estimatescb)<-c("term","estimate","model")
+
+estimatescoxFull=setDT(as.data.frame(coef(coxFull)), keep.rownames = TRUE)[]
+#estimatescoxFull$std.error<-0
+estimatescoxFull$Model="CoxFull"
+colnames(estimatescoxFull)<-c("term","estimate","model")
+
+lolliplotDots<-rbind(estimatescb,estimatescoxNet,estimatescoxFull)
+lolliplotDots$conf.low<-NA
+lolliplotDots$conf.high<-NA
+lolliplotDots$conf.high[lolliplotDots$estimate< 0]<-0
+
+lolliplotDots$conf.low[lolliplotDots$estimate> 0]<-0
+
+dwplot(lolliplotDots)
+
+```
 

### 2020-03-06 analysis/paper/paper.tex
NOT COMPLETE
to do:
fix casebase absoluterisk for glmnet. 
fix ??? tags
add lines to dot whisker plot
Need to removeRcolorBrewer palletes

--- 2020-03-03 analysis/paper/paper.tex
@@
 \citep{knaus1995support}. The SUPPORT dataset tracks death for
-individuals who are considered seriously ill within a hospital.
-Imputation was conducted using the \pkg{mice} package in \proglang{R}
-with default settings \citep{mice}. Before imputation, there were a
-total of 1000 individuals and 35 variables. After imputation, we have
-690 individuals. We lose individuals due to coliniarity. If a pair of
-covariates are deemed colinear to a certain degree, they are excluded
-from any imputations. Here, only edu income avtisst* race ph* glucose*
-adlp* were imputed. Of these, 66.09\% experienced the event of interest.
-A description of each variable can be found in Table \ref{tab:support1}
-and a breakdown of each categorical variable in Table
+individuals who are considered seriously ill within 5 hospitals in the
+United States of America. Imputation was conducted using the \pkg{mice}
+package in \proglang{R} with default settings \citep{mice}. Before
+imputation, there were a total of 9105 individuals and 47 variables. Of
+those variables, a few were removed before imputation. We removed three
+response variables: hospital charges, patient ratio of costs to charges
+and patient micro-costs. Next, we removed hospital death as it was
+directly informative of our event of interest, death. We also removed
+functional disability and income as they are ordinal covariates, because
+ordinal variables are treated as factors in our current model. Finally,
+we removed 8 covariates relating to the results of previous findings: we
+removed SUPPORT day 3 physiology score (sps), APACHE III day 3
+physiology score (aps), SUPPORT model 2-month survival estimate, SUPPORT
+model 6-month survival estimate, Physician's 2-month survival estimate
+for pt., Physician's 6-month survival estimate for pt., Patient had Do
+Not Resuscitate (DNR) order and Day of DNR order (\textless{}0 if before
+study). Of these, sps and aps were added on after imputation, as they
+were missing only 1 observation and permitted us to compare findings
+between our model and their's.
+
+First we impute manually using recommended normal values for
+physiological measures \citep{harrell_2020}. Next, we impute a single
+dataset using \pkg{mice} with default settings. After imputation, we
+noted that the covariate for surrogate activities of daily living was
+not imputed. This is due to colinearity between the other two covariates
+for activities of daily living. As such, surrogate activites of daily
+living was removed from our model. After these imputations, we have a
+total of 9104 observations and 30 variables, two of which are sps and
+aps, which will work as a baseline for our comparison to previous
+findings. This dataset is available Its worth noting that these scores
+were developed for a period of 180 days, which makes our comparison over
+5.56 years exploratory. Also, our single imputation introduces bias in
+the data. As the goal here is purely demonstrative, this bias will be
+consistent with any of the following models. As such, their comparison
+is still valid, though any interpretation requires an awareness of said
+bias. A description of used variables can be found in Table
+\ref{tab:support1} and a breakdown of each categorical variable in Table
 \ref{tab:support2}. The original data was obtained from the Department
-of Biostatistics at Vanderbilt University. The imputed dataset is
-available as part of the \pkg{casebase} package.
+of Biostatistics at Vanderbilt University \citep{harrell_2020}. The
+imputed dataset is available as part of the \pkg{casebase} package.
 
@@
 scoma         & SUPPORT Coma Score based on Glasgow D3   &                 & double           & 0            \\
-charges       & Hospital Charges                         &                 & double           & 25           \\
-totcst        & Total RCC cost                           &                 & double           & 105          \\
-totmcst       & Total micro-cost                         &                 & double           & 372          \\
 avtisst       & Average TISS, Days 3-25                  &                 & double           & 6            \\
 race          &                                          & 5               & integer          & 5            \\
+sps             & support physiology score day 3           &                   & double           & 1            \\
+aps           & APS III no coma, imp bun,uout for ph1,D3 &                 & double           & 1            \\
+hday            & Day in Hospital at Study Admit             &                 & double           & 0            \\ 
+diabetes      & Diabetes (Com 27-28, Dx 73)              &                 & double           & 0            \\
+dementia      & Dementia (Comorbidity 6)                 &                 & double           & 0            \\
+ca              & Cancer State                             & 3               & integer          & 0            \\
 meanbp        & Mean Arterial Blood Pressure Day 3       &                 & double           & 0            \\
@@
 urine         & Urine Output Day 3                       &                 & double           & 517          \\
-adlp          & ADL Patient Day 3                        &                 & double           & 634          \\
-adls          & ADL Surrogate Day 3                      &                 & double           & 310          \\
-sfdm2         &                                          & 5               & integer          & 159          \\
+adlp          & Activiites of Daily Living (ADL) Patient Day 3 &           & double           & 634          \\
 adlsc         & Imputed ADL Calibrated to Surrogate      &                 & double           & 0           
@@
          & hispanic                                      \\
-sfdm2    & no(M2 and SIP pres)                           \\
-         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
-         & SIP\textgreater{}=30                          \\
-         & Coma or Intub                                 \\
-         & \textless{}2 mo. follow-up                   
+ca       & Yes                                           \\
+         & No                                            \\
+         & Metastatic                                    \\
 \end{tabular}
@@
 population-time plot for the SUPPORT data appears in Figure
-\ref{fig:support-poptime}.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> data(support)
+\ref{fig:support-poptime}. As we can see, the incidence density is not
+constant.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> support<-casebase::support
+R> 
 R> support_popTime <- popTime(support,
-R>                            time = "d.time", event = "death",percentile_number = 0.31)
+R+                            time = "d.time", event = "death",percentile_number = 0.31)
 R> plot(support_popTime)
 \end{CodeInput}
-\end{CodeChunk}
-
-Before we proceed with the analysis, we first remove hospital death as a
-covariate, as this is directly informative of death. All covariates will
-be used to estimate the hazard function. The hazard function will be
-used to estimate the cumulative incidence of each individual in the
-study. Then, the average cumulative incidience curve for all
-observations will be used for our comparison.
-
-Our main objective is to compute the absolute risk of death for a given
-set of covariates, using regularization when fitting our model. First,
-we fit a smooth hazard to the data; for the sake of this example, we
-opted for a linear term for time. As casebase makes use of the glmnet
-package, we interact with the fitsmoothhazard.fit function using a
-matrix interface, where y contains the time and event variables and x
-contains all other variables we would like to include in the model. For
-the SUPPORT dataset, all factor variables were converted into dummy
-variables. Here, we used lasso penalization by setting alpha to 1.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> xInteractions=x*y[,2]
-R> colnames(xInteractions) <- paste("d.time*", colnames(xInteractions), sep = "")
-R> xFull=cbind(x,xInteractions)
-R> cb.ModelInter=casebase::fitSmoothHazard.fit(xFull,y,family="glmnet",time="d.time",event="death",
-R>                                        formula_time = ~d.time,alpha=1,ratio=rat)
-\end{CodeInput}
-\end{CodeChunk}
-
-Then, we take our model and use the absoluteRisk function to integrate
-and retrieve our absolute risk function.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> #Estimating the absolute risk curve using the newData parameter.
-R> casebaseAbsoluteInter=casebase::absoluteRisk(cb.ModelInter,time = seq(1,max(completeData$d.time), 400),newdata = xFull , s="lambda.1se",method=c("numerical"))
-R> 
-R> cabI=data.frame(casebaseAbsoluteInter[,1],rowMeans(casebaseAbsoluteInter[,-c(1)]))
-\end{CodeInput}
-\end{CodeChunk}
-
-We will compare this curve to a regularized version of cox regression,
-and a kaplan-meier survival curve. The cox regression absolute risk
-curve with lasso penalization required two extra steps to retrieve.
-Coxnet from the glmnet package has tools to fit a regularized cox model,
-but requires a survival object skeleton of the selected variables, so
-that the survival package tools can be used to retrive the absolute
-risk. We handle this by fitting a regularized cox model with the glmnet
-package, and a cox model from the survival package with the selected
-variables from the regularized fit. This second model serves as a
-skeleton that permits the use of survival package functions. The
-coefficients from the regularized model will replace the coefficients in
-our skeleton. the resulting model will have the correct coefficients
-when integrating, but will have an incorrect standard error. For the
-purposes of this case study, we are only interested in the absolute risk
-curve itself and not the standard error.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
+\begin{figure}
+
+{\centering \includegraphics{../figures/support-poptime-1}
+
+}
+
+\caption{\label{fig:support-poptime} Population-time plot for the SUPPORT data.}\label{fig:support-poptime}
+\end{figure}
+\end{CodeChunk}
+
+Before generating an absolute risk curve using regularized logistic
+regression, we will explore three options as a baseline comparison,
+beyond a simple Kaplan-meier curve. For all the following analyses in
+this case study, we will use 80\% of the observations to train our
+models, while 20\% will be used to generate absolute risk curves, all of
+which will be averaged to demonstrate each model's predictive ability.
+We will then compare each modeling using the IPA score (???NEED
+REFERENCE HERE???) We will examine three cox regression scenarios: sps
+only, aps only and all covariates excluding sps and aps. The original
+study compared these scores over a span of 180 days, which may not be
+comparable over the full 5.56 years. Over 5.56 years,
+\ref{fig:baselines} demonstrates that ???BEST MODEL BASED ON IPA??? has
+the best calibration and discrimination with an IPA score of ???INSERT
+ACTUAL BEST SCORE??? . ??BEST MODEL?? will be used in the following
+comparison to regularized cox, regularized logistic regression and the
+unadjusted Kaplan-Meier curve.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> completeData<-casebase::support
+R> workingCompleteData=model.matrix(death~ .-d.time,data=completeData)[,-c(1)]#remove intercept
 R> #Create u and xCox, which will be used when fitting Cox with glmnet.
-R> u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
-R> 
-R> xCoxI=as.matrix(cbind(data.frame("(Intercept)"=rep(1,each=690)),xFull))
-R> colnames(xCoxI)[1]="(Intercept)"
-R> 
-R> #hazard for cox using glmnet
-R> coxNetI=glmnet::cv.glmnet(x=xCoxI,y=u, family="cox",alpha=1)
-R> #convergence demonstrated in plot
-R> #plot(coxNet)
-R> #taking the coefficient estimates for later use
-R> nonzero_covariate_cox <- predict(coxNetI, type = "nonzero", s = "lambda.1se")
-R> nonzero_coef_cox <- coef(coxNetI, s = "lambda.1se")
-R> #creating a new dataset that only contains the covariates chosen through glmnet.
-R> #cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
-R> cleanCoxDataI<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,nonzero_covariate_cox$X1]))
-R> #newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
-R> #fitting a cox model using regular estimation, however we will not keep it.
-R> #this is used more as an object place holder.
-R> coxNetFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxDataI)
-R> #The coefficients of this object will be replaced with the estimates from coxNet.
-R> #Doing so makes it so that everything is invalid aside from the coefficients.
-R> #In this case, all we need to estimate the absolute risk is the coefficients.
-R> #Std. error would be incorrect here, if we were to draw error bars.
-R> coxNetFitI$coefficients<-nonzero_coef_cox@x
-R> #Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
-R> #posCox=nonzero_covariate_cox$X1
-R> #newDataCoxI=cleanCoxDataI[,posCox]
-R> 
-R> abcoxNetI<-survival::survfit(coxNetFitI,type="breslow",newdata=cleanCoxDataI)
-R> 
-R> abCNI<-abcoxNetI
-R> abCNI$surv=rowMeans(abCNI$surv)
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> fullDataCox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,-c(1)]))
-R> coxFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = fullDataCox)
-R> abcoxFitI<-survival::survfit(coxFitI,newdata=fullDataCox)
-R> abCFI<-abcoxFitI
-R> abCFI$surv<-rowMeans(abCFI$surv)
-R> #cb.ModelNorm=casebase::fitSmoothHazard(V2~.,time="V1",event="V2",
-R>        #                                ratio=rat,data=fullDataCox)
-R> #summary(cb.ModelNorm)#need to figure out how to deal with NAs
-R> #casebaseAbsoluteNorm=casebase::absoluteRisk(cb.ModelNorm,time = seq(1,max(completeData$d.time), 200),newdata = newData , s="lambda.1se",method=c("numerical"))
-R> #cabn=data.frame(casebaseAbsoluteNorm[,1],rowMeans(casebaseAbsoluteNorm[,-c(1)]))
-R> 
-\end{CodeInput}
-\end{CodeChunk}
-
-We used the survival package to calculate the kaplan-meier absolute risk
-function.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> 
-R> coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
-R> abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)
-R> 
-R> #creating a surv object to be used to fit an unadjusted absolute risk curve.
-R> # (Kaplan-Meier risk curve)
-R> km <- survival::Surv(time = completeData$d.time, event = completeData$death)
-R> abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
-\end{CodeInput}
-\end{CodeChunk}
-
-Now that our four cumulative incidence estimations have been calculated,
-we compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase
-decrease the absolute risk by the end of the study, in comparison to
-kaplan-meier.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> 
-R> #get covariates excluding d.time
-R> estimatescb=setDT(as.data.frame(coef(cb.Model)[-c(1,2),1]), keep.rownames = TRUE)[]
+R> newData=workingCompleteData
+R> x=workingCompleteData
+R> #x and y will be used to fit the casebase model under glmnet.
+R> #x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
+R> y=data.matrix(completeData[,c(2,5)])
+R> 
+R> fullDataCox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),x))
+R> train_index <- sample(1:nrow(fullDataCox), 0.8 * nrow(fullDataCox))
+R> test_index <- setdiff(1:nrow(fullDataCox), train_index)
+R> 
+R> train<-fullDataCox[train_index,]
+R> test<-fullDataCox[test_index,]
+R> coxSPS<- survival::coxph(Surv(time=V1,event=V2) ~ sps, data = train)
+R> abcoxSPS<-survival::survfit(coxSPS,newdata = test)
+R> 
+R> coxAPS<- survival::coxph(Surv(time=V1,event=V2) ~ aps, data = train)
+R> abcoxAPS<-survival::survfit(coxAPS,newdata = test)
+R> 
+R> coxFull<- survival::coxph(Surv(time=V1,event=V2) ~ .-sps -aps, data = train)
+R> abcoxFull<-survival::survfit(coxAPS,newdata = test)
+R> #library(prodlim)
+R> 
+R> 
+R> KM<- survival::coxph(Surv(time=V1,event=V2) ~ 1, data = test)
+R> abKM<-survival::survfit(KM)
+R> #abKM <- prodlim(Hist(V1,V2)~1,data=test) minimal difference
+R> abKMcompare<-cbind(abKM$time[abKM$time<=1631],1-abKM$surv[abKM$time<=1631])
+R> colnames(abKMcompare)<-c("Time","KM")
+R> 
+R> baseCoxComparisonData<-cbind(abcoxSPS$time[abcoxSPS$time<=1631],1-rowMeans(abcoxSPS$surv[abcoxSPS$time<=1631,]),1-rowMeans(abcoxAPS$surv[abcoxSPS$time<=1631,]),1-abKM$surv,1-rowMeans(abcoxFull$surv[abcoxSPS$time<=1631,]))
+R> colnames(baseCoxComparisonData)<-c("Time","SPS","APS","Unadjusted","Full")
+R> myColors <- brewer.pal(5,"Dark2")
+R> FigureWidth='90%'
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/baselines-1}
+
+}
+
+\caption{\label{fig:baselines} AbsoluteRisk curves from hazard models.
+Pink: All covariates without SPS nor APS.
+Green: APS only model.
+Blue: SPS only model.
+Purple: unadjusted curve.
+All three models underestimate the risk in comparison to the unadjusted curve.}\label{fig:baselines}
+\end{figure}
+\end{CodeChunk}
+
+To get the absolute risk following regularized cox regression, this
+required that we work around limitations in existing packages. To make
+use of the \pkg{survival} package, we first get our hazard coefficients
+using \pkg{glmnet}'s coxnet implementation for lasso penalization
+\citep{regpathcox}. Next, we fit a cox model using the survival package
+with the selected variables only, from our previous fit. Finally, we
+replaced the coefficients in the survival object with those from our
+glmnet fit, allowing us to make use of the absolute risk tools available
+in the \pkg{survival} package.
+
+For our penalized logistic regression model, we opted for a logistic
+function of time, resulting in a Weibull distribution assumption. As
+casebase makes use of the glmnet package, we interact with the
+fitsmoothhazard.fit function using a matrix interface, where y contains
+the time and event variables and x contains all other variables we would
+like to include in the model. For the SUPPORT dataset, all factor
+variables were converted into dummy variables. Here, we used lasso
+penalization by setting alpha to 1, leading to our smooth hazard
+function, depicted in the ??? curve in \ref{fig:cs3Final}. ??? I AM
+STILL PENALIZING TIME ???
+
+We then take our hazard and use it to predict the absolute risk of each
+individual in our test set, over time. These individuals are then
+average resulting in the ???curve in \ref{fig:cs3Final}. ???SAME BUG IS
+BACK???
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> ggplot(as.data.frame(ab), aes(Time)) + 
+R+     geom_line(data=as.data.frame(baseCoxComparisonData),mapping=aes(y = Full, colour = "Cox"),lwd=2)+
+R+   geom_line(aes(y = `Absolute risk`, colour = Model),lwd=2)+
+R+   geom_step(data=as.data.frame(abKMcompare),mapping=aes(x=Time,y=KM,color="Unadjusted"),lwd=2)+
+R+   scale_colour_manual(name = "Models",values = myColors[c(4,1,3,2)])+
+R+   labs(y= "Probability of Death", x = "Survival time (Days)")
+\end{CodeInput}
+\begin{figure}
+
+{\centering \includegraphics{../figures/unnamed-chunk-2-1}
+
+}
+
+\caption{\label{fig:baselines} HERES SOME TEXT ABT IT}\label{fig:unnamed-chunk-2}
+\end{figure}
+\end{CodeChunk}
+
+???SHORT SPIEL ABOUT IPA BETWEEN SEMI\_PARAMETRIC AND PARAMETRIC which
+leads to conclusion of best overall model. This refers to the COEFPLOT
+in a following figure???
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> 
+R> library(dotwhisker)
+R> library(broom)
+R> library(dplyr)
+R> 
+R> estimatescb=setDT(as.data.frame(coef(cbFull)[-c(1,2),1]), keep.rownames = TRUE)[]
+R> #estimatescb$std.error<-0
 R> estimatescb$Model="casebase"
-R> colnames(estimatescb)<-c("Coefs","Estimate","Model")
-R> estimatescox=setDT(as.data.frame(coef(coxNet)[-c(1),1]), keep.rownames = TRUE)[]
-R> estimatescox$Model="Cox"
-R> colnames(estimatescox)<-c("Coefs","Estimate","Model")
-R> estimates=rbind(estimatescb, estimatescox)
-R> estimates$Estimate[estimates$Estimate==0]=NA# make it so unchosen covariates appear empty
-R> library(ggstance)# for vertical dodge
-R>   ggplot(estimates, aes(x = Estimate, y = Coefs, color = Model)) +
-R>         geom_segment(aes(x = 0, y = Coefs, xend = Estimate, yend = Coefs),size=0.5 ) +
-R>         geom_point(position=ggstance::position_dodgev(height=0.2))
-R> 
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> #get covariates excluding d.time
-R> library(data.table)
-R> estimatescb=setDT(as.data.frame(coef(cb.ModelInter)[-c(1,2),1]), keep.rownames = TRUE)[]
-R> estimatescb$Model="casebase"
-R> colnames(estimatescb)<-c("Coefs","Estimate","Model")
-R> estimatescox=setDT(as.data.frame(coef(coxNetI)[-c(1),1]), keep.rownames = TRUE)[]
-R> estimatescox$Model="Cox"
-R> colnames(estimatescox)<-c("Coefs","Estimate","Model")
-R> estimatesI=rbind(estimatescb, estimatescox)
-R> estimatesI$Estimate[estimatesI$Estimate==0]=NA# make it so unchosen covariates appear empty
-R> library(ggstance)# for vertical dodge
-R>   ggplot(estimatesI, aes(x = Coefs, y = Estimate , color = Model)) +
-R>         geom_segment(aes(y = 0, x = Coefs, yend = Estimate, xend = Coefs),size=0.5, position = position_dodge(width = 0.5)) +
-R>         geom_point(position = position_dodge(width = 0.5))
-R>   
-R>     
-\end{CodeInput}
+R> colnames(estimatescb)<-c("term","estimate","model")
+R> 
+R> estimatescoxFull=setDT(as.data.frame(coef(coxFull)), keep.rownames = TRUE)[]
+R> #estimatescoxFull$std.error<-0
+R> estimatescoxFull$Model="CoxFull"
+R> colnames(estimatescoxFull)<-c("term","estimate","model")
+R> 
+R> lolliplotDots<-rbind(estimatescb,estimatescoxNet,estimatescoxFull)
+R> lolliplotDots$conf.low<-NA
+R> lolliplotDots$conf.high<-NA
+R> lolliplotDots$conf.high[lolliplotDots$estimate< 0]<-0
+R> 
+R> lolliplotDots$conf.low[lolliplotDots$estimate> 0]<-0
+R> 
+R> dwplot(lolliplotDots)
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/coefplots-1} \end{center}
+
 \end{CodeChunk}
@@
 interval censoring. This type of censoring is very common in
-longitudinal studies, and many packages (e.g. \pkg{SmoothHazard},
+longitudinal studies, and many packages (e.g.~\pkg{SmoothHazard},
 \pkg{survival} and \pkg{Rstpm2}) provide functions to account for it.
@@
 
-This report was generated on 2020-03-03 13:15:36 using the following
+This report was generated on 2020-03-05 23:20:03 using the following
 computational environment and dependencies:
@@
 \begin{CodeOutput}
-#> R version 3.6.2 (2019-12-12)
-#> Platform: x86_64-pc-linux-gnu (64-bit)
-#> Running under: Ubuntu 18.04.4 LTS
+#> R version 3.6.1 (2019-07-05)
+#> Platform: x86_64-w64-mingw32/x64 (64-bit)
+#> Running under: Windows 10 x64 (build 18363)
 #> 
 #> Matrix products: default
-#> BLAS:   /usr/lib/x86_64-linux-gnu/openblas/libblas.so.3
-#> LAPACK: /usr/lib/x86_64-linux-gnu/libopenblasp-r0.2.20.so
 #> 
 #> locale:
-#>  [1] LC_CTYPE=en_CA.UTF-8       LC_NUMERIC=C              
-#>  [3] LC_TIME=en_CA.UTF-8        LC_COLLATE=en_CA.UTF-8    
-#>  [5] LC_MONETARY=en_CA.UTF-8    LC_MESSAGES=en_CA.UTF-8   
-#>  [7] LC_PAPER=en_CA.UTF-8       LC_NAME=C                 
-#>  [9] LC_ADDRESS=C               LC_TELEPHONE=C            
-#> [11] LC_MEASUREMENT=en_CA.UTF-8 LC_IDENTIFICATION=C       
+#> [1] LC_COLLATE=English_United States.1252 
+#> [2] LC_CTYPE=English_United States.1252   
+#> [3] LC_MONETARY=English_United States.1252
+#> [4] LC_NUMERIC=C                          
+#> [5] LC_TIME=English_United States.1252    
 #> 
@@
 #> other attached packages:
-#>  [1] pracma_2.2.9        kableExtra_1.1.0    data.table_1.12.8  
-#>  [4] glmnet_3.0-1        Matrix_1.2-18       casebase_0.2.1.9001
-#>  [7] survival_3.1-8      magrittr_1.5        forcats_0.4.0      
-#> [10] stringr_1.4.0       dplyr_0.8.3         purrr_0.3.3        
-#> [13] readr_1.3.1         tidyr_1.0.0         tibble_2.1.3       
-#> [16] ggplot2_3.2.1       tidyverse_1.3.0    
+#>  [1] broom_0.5.5         dotwhisker_0.5.0    RColorBrewer_1.1-2 
+#>  [4] pracma_2.2.9        kableExtra_1.1.0    data.table_1.12.8  
+#>  [7] glmnet_3.0-2        Matrix_1.2-17       casebase_0.2.1.9001
+#> [10] survival_3.1-8      magrittr_1.5        forcats_0.5.0      
+#> [13] stringr_1.4.0       dplyr_0.8.3         purrr_0.3.3        
+#> [16] readr_1.3.1         tidyr_1.0.2         tibble_2.1.3       
+#> [19] ggplot2_3.3.0       tidyverse_1.3.0    
 #> 
 #> loaded via a namespace (and not attached):
-#>  [1] Rcpp_1.0.3        lubridate_1.7.4   lattice_0.20-40   assertthat_0.2.1 
-#>  [5] digest_0.6.25     foreach_1.4.7     R6_2.4.1          cellranger_1.1.0 
-#>  [9] backports_1.1.5   reprex_0.3.0      stats4_3.6.2      evaluate_0.14    
-#> [13] httr_1.4.1        pillar_1.4.3      rlang_0.4.5       lazyeval_0.2.2   
-#> [17] readxl_1.3.1      rstudioapi_0.10   rticles_0.12      rmarkdown_1.18   
-#> [21] webshot_0.5.2     munsell_0.5.0     broom_0.5.2       compiler_3.6.2   
-#> [25] modelr_0.1.5      xfun_0.11         pkgconfig_2.0.3   shape_1.4.4      
-#> [29] mgcv_1.8-31       htmltools_0.4.0   tidyselect_0.2.5  codetools_0.2-16 
-#> [33] viridisLite_0.3.0 fansi_0.4.1       crayon_1.3.4      dbplyr_1.4.2     
-#> [37] withr_2.1.2       grid_3.6.2        nlme_3.1-144      jsonlite_1.6.1   
-#> [41] gtable_0.3.0      lifecycle_0.1.0   DBI_1.0.0         scales_1.1.0     
-#> [45] cli_2.0.2         stringi_1.4.6     fs_1.3.1          xml2_1.2.2       
-#> [49] generics_0.0.2    vctrs_0.2.3       iterators_1.0.12  tools_3.6.2      
-#> [53] glue_1.3.1        hms_0.5.2         yaml_2.2.1        colorspace_1.4-1 
-#> [57] rvest_0.3.5       VGAM_1.1-2        knitr_1.26        haven_2.2.0
+#>  [1] httr_1.4.1        VGAM_1.1-2        jsonlite_1.6.1    viridisLite_0.3.0
+#>  [5] foreach_1.4.8     modelr_0.1.6      assertthat_0.2.1  stats4_3.6.1     
+#>  [9] ggstance_0.3.3    cellranger_1.1.0  yaml_2.2.1        pillar_1.4.3     
+#> [13] backports_1.1.5   lattice_0.20-38   glue_1.3.1        digest_0.6.25    
+#> [17] rvest_0.3.5       colorspace_1.4-1  plyr_1.8.6        htmltools_0.4.0  
+#> [21] pkgconfig_2.0.3   haven_2.2.0       scales_1.1.0      webshot_0.5.2    
+#> [25] mgcv_1.8-28       rticles_0.14      generics_0.0.2    farver_2.0.3     
+#> [29] withr_2.1.2       cli_2.0.2         crayon_1.3.4      readxl_1.3.1     
+#> [33] evaluate_0.14     fs_1.3.2          fansi_0.4.1       nlme_3.1-140     
+#> [37] xml2_1.2.2        tools_3.6.1       hms_0.5.3         lifecycle_0.1.0  
+#> [41] munsell_0.5.0     reprex_0.3.0      compiler_3.6.1    rlang_0.4.5      
+#> [45] grid_3.6.1        iterators_1.0.12  rstudioapi_0.11   labeling_0.3     
+#> [49] rmarkdown_2.1     gtable_0.3.0      codetools_0.2-16  DBI_1.1.0        
+#> [53] R6_2.4.1          lubridate_1.7.4   knitr_1.28        shape_1.4.4      
+#> [57] stringi_1.4.6     Rcpp_1.0.3        vctrs_0.2.3       dbplyr_1.4.2     
+#> [61] tidyselect_1.0.0  xfun_0.12
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> Local:    review /home/turgeonm/Documents/git_repos/cbpaper
-#> Head:     [8804f68] 2020-02-21: Merge branch 'Jesse-Islam-master'
+#> Local:    master C:/Users/jesse/Documents/gits/cbpaper
+#> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
+#> Head:     [843365a] 2020-03-05: In process of changing my results
 \end{CodeOutput}

### 2020-03-06 analysis/paper/paper.Rmd
Fixed lollipop plot
to do:
fix casebase absoluterisk for glmnet. 
fix ??? tags

--- 2020-03-05 analysis/paper/paper.Rmd
@@
 bibliography: references.bib  #Also not included in template by default. 
-fig_width: 7
-fig_height: 6
+fig_width: 12
+fig_height: 9
 fig_caption: true
@@
 library(casebase)
-#devtools::load_all('../../../casebase')#testing
+# devtools::load_all('../../../casebase')#testing
 library(splines)
@@
   levels = c(0, 1),
-                       labels = c("Control group", "Screening group"))
+  labels = c("Control group", "Screening group")
+)
 ```
@@
   event = "DeadOfPrCa",
-                                     exposure = "ScrArm")
+  exposure = "ScrArm"
+)
 plot(pt_object_strat)
@@
 cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm,
-                             data = ERSPC)
+  data = ERSPC
+)
 summary(cox_model)
@@
 ```{r erspc-cox-cif, eval = eval_cs1}
-new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
-                       ignore = 99)
+new_data <- data.frame(
+  ScrArm = c("Control group", "Screening group"),
+  ignore = 99
+)
 
@@
   fun = "event",
-     xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-     main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+  xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
+  main = sprintf(
+    "Estimated Cumulative Incidence (risk) of Death from Prostate 
                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
@@
     exp(confint(cox_model))[1],
-                    exp(confint(cox_model))[2]),
-     ylim = c(0, 2), yscale = 100)
+    exp(confint(cox_model))[2]
+  ),
+  ylim = c(0, 2), yscale = 100
+)
 legend("topleft",
   legend = c("Control group", "Screening group"),
-       col = c("red","blue"),
+  col = c("red", "blue"),
   lty = c(1, 1),
-       bg = "gray90")
+  bg = "gray90"
+)
 ```
@@
   data = ERSPC,
-                                                  ratio = 100)
+  ratio = 100
+)
 
@@
 exp(coef(casebase_exponential)[2])
-exp(confint(casebase_exponential)[2,])
+exp(confint(casebase_exponential)[2, ])
 ```
@@
 ```{r erspc-casebase-cif, eval = eval_cs1}
-smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
-                                          time = seq(0,15,0.1), 
-                                          newdata = new_data)
+smooth_risk_exp <- casebase::absoluteRisk(
+  object = casebase_exponential,
+  time = seq(0, 15, 0.1),
+  newdata = new_data
+)
 
@@
     exp(confint(cox_model))[1],
-                    exp(confint(cox_model))[2]),
-     ylim = c(0, 2), yscale = 100)
-lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
-lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
+    exp(confint(cox_model))[2]
+  ),
+  ylim = c(0, 2), yscale = 100
+)
+lines(smooth_risk_exp[, 1], smooth_risk_exp[, 2], col = "red", lty = 2)
+lines(smooth_risk_exp[, 1], smooth_risk_exp[, 3], col = "blue", lty = 2)
 
 legend("topleft",
-       legend = c("Control group (Cox)","Control group (Casebase)",
-                  "Screening group (Cox)", "Screening group (Casebase)"), 
-       col = c("red","red", "blue","blue"),
+  legend = c(
+    "Control group (Cox)", "Control group (Casebase)",
+    "Screening group (Cox)", "Screening group (Casebase)"
+  ),
+  col = c("red", "red", "blue", "blue"),
   lty = c(1, 2, 1, 2),
-       bg = "gray90")
+  bg = "gray90"
+)
 ```
@@
   data = ERSPC,
-                                 ratio = 100)
+  ratio = 100
+)
 
@@
 ```{r erspc-casebase-weibull-cif, eval = eval_cs1}
-smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
-                                          time = seq(0,15,0.1), 
-                                          newdata = new_data)
+smooth_risk_time <- casebase::absoluteRisk(
+  object = casebase_time,
+  time = seq(0, 15, 0.1),
+  newdata = new_data
+)
 
@@
     exp(confint(cox_model))[1],
-                    exp(confint(cox_model))[2]),
-     ylim = c(0, 2), yscale = 100)
-lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
-lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
+    exp(confint(cox_model))[2]
+  ),
+  ylim = c(0, 2), yscale = 100
+)
+lines(smooth_risk_time[, 1], smooth_risk_time[, 2], col = "red", lty = 2)
+lines(smooth_risk_time[, 1], smooth_risk_time[, 3], col = "blue", lty = 2)
 
 legend("topleft",
-       legend = c("Control group (Cox)","Control group (Casebase)",
-                  "Screening group (Cox)", "Screening group (Casebase)"), 
-       col = c("red","red", "blue","blue"),
+  legend = c(
+    "Control group (Cox)", "Control group (Casebase)",
+    "Screening group (Cox)", "Screening group (Casebase)"
+  ),
+  col = c("red", "red", "blue", "blue"),
   lty = c(1, 2, 1, 2),
-       bg = "gray90")
+  bg = "gray90"
+)
 ```
@@
   data = ERSPC,
-                                    ratio = 100)
+  ratio = 100
+)
 
@@
 ```{r erspc-casebase-splines-cif, eval = eval_cs1}
-smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
-                                    time = seq(0,15,0.1), 
-                                    newdata = new_data)
+smooth_risk_splines <- absoluteRisk(
+  object = casebase_splines,
+  time = seq(0, 15, 0.1),
+  newdata = new_data
+)
 
@@
     exp(confint(cox_model))[1],
-                    exp(confint(cox_model))[2]),
-     ylim = c(0, 2), yscale = 100)
-lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
-lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
+    exp(confint(cox_model))[2]
+  ),
+  ylim = c(0, 2), yscale = 100
+)
+lines(smooth_risk_splines[, 1], smooth_risk_splines[, 2], col = "red", lty = 2)
+lines(smooth_risk_splines[, 1], smooth_risk_splines[, 3], col = "blue", lty = 2)
 
 legend("topleft",
-       legend = c("Control group (Cox)","Control group (Casebase)",
-                  "Screening group (Cox)", "Screening group (Casebase)"), 
-       col = c("red","red", "blue","blue"),
+  legend = c(
+    "Control group (Cox)", "Control group (Casebase)",
+    "Screening group (Cox)", "Screening group (Casebase)"
+  ),
+  col = c("red", "red", "blue", "blue"),
   lty = c(1, 2, 1, 2),
-       bg = "gray90")
+  bg = "gray90"
+)
 ```
@@
 CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
-CI_exp <- c(exp(coef(casebase_exponential))["ScrArmScreening group"], exp(confint(casebase_exponential))["ScrArmScreening group",])
-CI_time <- c(exp(coef(casebase_time))["ScrArmScreening group"], exp(confint(casebase_time))["ScrArmScreening group",])
-CI_splines <- c(exp(coef(casebase_splines))["ScrArmScreening group"], exp(confint(casebase_splines))["ScrArmScreening group",])
+CI_exp <- c(exp(coef(casebase_exponential))["ScrArmScreening group"], exp(confint(casebase_exponential))["ScrArmScreening group", ])
+CI_time <- c(exp(coef(casebase_time))["ScrArmScreening group"], exp(confint(casebase_time))["ScrArmScreening group", ])
+CI_splines <- c(exp(coef(casebase_splines))["ScrArmScreening group"], exp(confint(casebase_splines))["ScrArmScreening group", ])
 
@@
   bind_cols(as.data.frame(table_ci)) %>%
-  mutate(CI = paste0("(", round(LB, 2), ", ",
-                     round(UB, 2), ")"),
-         Width = UB - LB) %>% 
+  mutate(
+    CI = paste0(
+      "(", round(LB, 2), ", ",
+      round(UB, 2), ")"
+    ),
+    Width = UB - LB
+  ) %>%
   dplyr::select(-LB, -UB) %>%
@@
 ```{r erscp-aic, eval = eval_cs1}
-c("Exp." = AIC(casebase_exponential),
+c(
+  "Exp." = AIC(casebase_exponential),
   "Weib." = AIC(casebase_time),
-  "Splines" = AIC(casebase_splines))
+  "Splines" = AIC(casebase_splines)
+)
 ```
@@
 ) %>%
-  knitr::kable(format = "latex", booktabs = TRUE,
+  knitr::kable(
+    format = "latex", booktabs = TRUE,
     caption = "Baseline characteristics of patients in the stem-cell transplant study.",
     label = "table1bmtcrr",
-               escape = FALSE) %>%
+    escape = FALSE
+  ) %>%
   kable_styling()
@@
 pt_object_comp <- bmtcrr %>%
-    mutate(Status = case_when(Status == 1 ~ 2L, 
+  mutate(Status = case_when(
+    Status == 1 ~ 2L,
     Status == 2 ~ 1L,
-                              TRUE ~ Status)) %>% 
+    TRUE ~ Status
+  )) %>%
   casebase::popTime(event = "Status", time = "ftime")
@@
     ggtitle("Competing events"),
-          labels = c('A', 'B'))
+  labels = c("A", "B")
+)
 ```
@@
   ratio = 100,
-    time = "ftime")
+  time = "ftime"
+)
 ```
@@
 model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-                   data = bmtcrr)
+  data = bmtcrr
+)
 ```
@@
 foo <- summary(model_cb)@coef3
-table_cb <- foo[!grepl("Intercept", rownames(foo)) & 
+table_cb <- foo[
+  !grepl("Intercept", rownames(foo)) &
     !grepl("ftime", rownames(foo)) &
     grepl(":1", rownames(foo)),
-                1:2]
-table_cb <- cbind(table_cb[,1], 
-                  table_cb[,1] - z_value * table_cb[,2],
-                  table_cb[,1] + z_value * table_cb[,2])
+  1:2
+]
+table_cb <- cbind(
+  table_cb[, 1],
+  table_cb[, 1] - z_value * table_cb[, 2],
+  table_cb[, 1] + z_value * table_cb[, 2]
+)
 table_cb <- round(exp(table_cb), 2)
 
-table_cox <- round(summary(model_cox)$conf.int[,-2], 2)
+table_cox <- round(summary(model_cox)$conf.int[, -2], 2)
 rownames(table_cb) <- rownames(table_cox)
@@
   )) %>%
-  knitr::kable(format = "latex", booktabs = TRUE,
+  knitr::kable(
+    format = "latex", booktabs = TRUE,
     col.names = c("Covariates", "HR", "95% CI", "HR", "95% CI"),
-               caption = "Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.") %>% 
+    caption = "Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented."
+  ) %>%
   kable_styling() %>%
@@
 # Data.frame containing risk profile
-newdata <- data.frame("Sex" = factor(c("F", "F"), 
-                                     levels = levels(bmtcrr[,"Sex"])),
+newdata <- data.frame(
+  "Sex" = factor(c("F", "F"),
+    levels = levels(bmtcrr[, "Sex"])
+  ),
   "D" = c("ALL", "AML"), # Both diseases
   "Phase" = factor(c("Relapse", "Relapse"),
-                                       levels = levels(bmtcrr[,"Phase"])),
+    levels = levels(bmtcrr[, "Phase"])
+  ),
   "Age" = c(35, 35),
   "Source" = factor(c("PB", "PB"),
-                                        levels = levels(bmtcrr[,"Source"])))
+    levels = levels(bmtcrr[, "Source"])
+  )
+)
 
 # Estimate absolute risk curve
-risk_cb <- absoluteRisk(object = model_cb, time = time_points,
-                        method = "numerical", newdata = newdata)
+risk_cb <- absoluteRisk(
+  object = model_cb, time = time_points,
+  method = "numerical", newdata = newdata
+)
 ```
@@
   const(Phase) + const(Source) + const(Age),
-                      data = bmtcrr, cause = 1, model = "fg")
+data = bmtcrr, cause = 1, model = "fg"
+)
 
@@
 risk_all <- dplyr::bind_rows(
-    data.frame(Time = time_points, 
+  data.frame(
+    Time = time_points,
     Method = "Case-base",
-               Risk = risk_cb[,2], 
+    Risk = risk_cb[, 2],
     Disease = "ALL",
-               stringsAsFactors = FALSE),
-    data.frame(Time = time_points, 
+    stringsAsFactors = FALSE
+  ),
+  data.frame(
+    Time = time_points,
     Method = "Case-base",
-               Risk = risk_cb[,3], 
+    Risk = risk_cb[, 3],
     Disease = "AML",
-               stringsAsFactors = FALSE),
-    data.frame(Time = time_points,
+    stringsAsFactors = FALSE
+  ),
+  data.frame(
+    Time = time_points,
     Method = "Fine-Gray",
-               Risk = risk_fg$P1[1,],
+    Risk = risk_fg$P1[1, ],
     Disease = "ALL",
-               stringsAsFactors = FALSE),
-    data.frame(Time = time_points,
+    stringsAsFactors = FALSE
+  ),
+  data.frame(
+    Time = time_points,
     Method = "Fine-Gray",
-               Risk = risk_fg$P1[2,],
+    Risk = risk_fg$P1[2, ],
     Disease = "AML",
-               stringsAsFactors = FALSE)
+    stringsAsFactors = FALSE
+  )
 )
@@
   facet_grid(Disease ~ .) +
-  ylim(c(0,1)) +
+  ylim(c(0, 1)) +
   theme_minimal() +
-  theme(legend.position = 'top') +    
-  xlab("Time (in Months)") + ylab("Relapse risk")
+  theme(legend.position = "top") +
+  xlab("Time (in Months)") +
+  ylab("Relapse risk")
 ```
@@
 
-```{r support-poptime, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data."}
-
-completeData<-casebase::support
-workingCompleteData=model.matrix(death~ .-d.time,data=completeData)[,-c(1)]#remove intercept
-#Create u and xCox, which will be used when fitting Cox with glmnet.
-newData=workingCompleteData
-x=workingCompleteData
-#x and y will be used to fit the casebase model under glmnet.
-#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
-y=data.matrix(completeData[,c(2,5)])
+```{r support-poptime, echo=FALSE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data."}
+support <- casebase::support
 
 support_popTime <- popTime(support,
-                           time = "d.time", event = "death",percentile_number = 0.31)
+  time = "d.time", event = "death", percentile_number = 0.31
+)
 plot(support_popTime)
@@
 
-Our next step is to get the hazard function.
-All covariates will be used to estimate the hazard function.
-The hazard function will be used to estimate the cumulative incidence of each individual in the study.
-Then, the average cumulative incidience curve for all observations will be used for our comparison.
-
-Our main objective is to compute the absolute risk of death for a given set of covariates, using regularization when fitting our model.
-First, we fit a smooth hazard to the data; for the sake of this example, we opted for a linear term for time.
+Before generating an absolute risk curve using regularized logistic regression, we will explore three options as a baseline comparison, beyond a simple Kaplan-meier curve.
+For all the following analyses in this case study, we will use 80% of the observations to train our models, while 20% will be used to generate absolute risk curves, all of which will be averaged to demonstrate each model's predictive ability.
+We will then compare each modeling using the IPA score (???NEED REFERENCE HERE???) We will examine three cox regression scenarios: sps only, aps only and all covariates excluding sps and aps.
+The original study compared these scores over a span of 180 days, which may not be comparable over the full 5.56 years.
+Over 5.56 years, \ref{fig:baselines} demonstrates that ???BEST MODEL BASED ON IPA???
+has the best calibration and discrimination with an IPA score of ???INSERT ACTUAL BEST SCORE???
+.
+??BEST MODEL??
+will be used in the following comparison to regularized cox, regularized logistic regression and the unadjusted Kaplan-Meier curve.
+
+```{r baselinePlotsPrep, eval=eval_cs3,echo=FALSE}
+completeData <- casebase::support
+workingCompleteData <- model.matrix(death ~ . - d.time, data = completeData)[, -c(1)] # remove intercept
+# Create u and xCox, which will be used when fitting Cox with glmnet.
+newData <- workingCompleteData
+x <- workingCompleteData
+# x and y will be used to fit the casebase model under glmnet.
+# x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
+y <- data.matrix(completeData[, c(2, 5)])
+
+fullDataCox <- as.data.frame(cbind(as.numeric(y[, 2]), as.numeric(y[, 1]), x))
+train_index <- sample(1:nrow(fullDataCox), 0.8 * nrow(fullDataCox))
+test_index <- setdiff(1:nrow(fullDataCox), train_index)
+
+train <- fullDataCox[train_index, ]
+test <- fullDataCox[test_index, ]
+coxSPS <- survival::coxph(Surv(time = V1, event = V2) ~ sps, data = train)
+abcoxSPS <- survival::survfit(coxSPS, newdata = test)
+
+coxAPS <- survival::coxph(Surv(time = V1, event = V2) ~ aps, data = train)
+abcoxAPS <- survival::survfit(coxAPS, newdata = test)
+
+coxFull <- survival::coxph(Surv(time = V1, event = V2) ~ . - sps - aps, data = train)
+abcoxFull <- survival::survfit(coxAPS, newdata = test)
+# library(prodlim)
+
+KM <- survival::coxph(Surv(time = V1, event = V2) ~ 1, data = test)
+abKM <- survival::survfit(KM)
+# abKM <- prodlim(Hist(V1,V2)~1,data=test) minimal difference
+abKMcompare <- cbind(abKM$time[abKM$time <= 1631], 1 - abKM$surv[abKM$time <= 1631])
+colnames(abKMcompare) <- c("Time", "KM")
+
+baseCoxComparisonData <- cbind(abcoxSPS$time[abcoxSPS$time <= 1631], 1 - rowMeans(abcoxSPS$surv[abcoxSPS$time <= 1631, ]), 1 - rowMeans(abcoxAPS$surv[abcoxSPS$time <= 1631, ]), 1 - abKM$surv, 1 - rowMeans(abcoxFull$surv[abcoxSPS$time <= 1631, ]))
+colnames(baseCoxComparisonData) <- c("Time", "SPS", "APS", "Unadjusted", "Full")
+```
+
+```{r baselines, echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:baselines} AbsoluteRisk curves from hazard models.
+Pink: All covariates without SPS nor APS.
+Green: APS only model.
+Blue: SPS only model.
+Purple: unadjusted curve.
+All three models underestimate the risk in comparison to the unadjusted curve.",echo=FALSE}
+ggplot(as.data.frame(baseCoxComparisonData), aes(Time)) +
+  geom_line(aes(y = SPS, colour = "SPS"), lwd = 1) +
+  geom_line(aes(y = APS, colour = "APS"), lwd = 1) +
+  geom_step(data = as.data.frame(abKMcompare), mapping = aes(x = Time, y = KM, color = "Unadjusted"), lwd = 1) +
+  geom_line(aes(y = Full, col = "All Covariates without SPS APS"), lwd = 1) +
+  labs(y = "Probability of Death", x = "Survival time (Days)")
+```
+
+To get the absolute risk following regularized cox regression, this required that we work around limitations in existing packages.
+To make use of the \pkg{survival} package, we first get our hazard coefficients using \pkg{glmnet}'s coxnet implementation for lasso penalization [@regpathcox].
+Next, we fit a cox model using the survival package with the selected variables only, from our previous fit.
+Finally, we replaced the coefficients in the survival object with those from our glmnet fit, allowing us to make use of the absolute risk tools available in the \pkg{survival} package.
+
+For our penalized logistic regression model, we opted for a logistic function of time, resulting in a Weibull distribution assumption.
 As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model.
 For the SUPPORT dataset, all factor variables were converted into dummy variables.
-Here, we used lasso penalization by setting alpha to 1.
+Here, we used lasso penalization by setting alpha to 1, leading to our smooth hazard function, depicted in the ???
+curve in \ref{fig:cs3Final}.
+??? I AM STILL PENALIZING TIME ???
+
+We then take our hazard and use it to predict the absolute risk of each individual in our test set, over time.
+These individuals are then average resulting in the ???curve in \ref{fig:cs3Final}.
+???SAME BUG IS BACK???
+
+```{r supportSetUp, echo = FALSE, eval = eval_cs3}
+
+#ratio <- 100 # universal ratio for TRUE
+
+# x and y will be used to fit the casebase model under glmnet.
+y <- data.matrix(train[, c(2, 1)])
+colnames(y) <- c("death", "d.time")
+x <- data.matrix(train[, -c(2, 1)])
+newData <- data.matrix(test[, -c(2, 1)])
+```
+
+```{r coxHazAbsolute, echo = FALSE,eval= eval_cs3 }
+# Create u and xCox, which will be used when fitting Cox with glmnet.
+
+u <- survival::Surv(time = as.numeric(y[, 2]), event = as.factor(y[, 1]))
+
+xCox <- as.matrix(cbind(data.frame("(Intercept)" = rep(1, each = length(x[, 1]))), x))
+colnames(xCox)[2] <- "(Intercept)"
+
+# hazard for cox using glmnet
+coxNet <- glmnet::cv.glmnet(x = x, y = u, family = "cox", alpha = 1, standardize = TRUE)
+# convergence demonstrated in plot
+# plot(coxNet)
+# taking the coefficient estimates for later use
+nonzero_covariate_cox <- predict(coxNet, type = "nonzero", s = "lambda.1se")
+nonzero_coef_cox <- coef(coxNet, s = "lambda.1se")
+# creating a new dataset that only contains the covariates chosen through glmnet.
+cleanCoxData <- as.data.frame(cbind(d.time = as.numeric(y[, 2]), death = as.numeric(y[, 1]), x[, nonzero_covariate_cox$X1]))
+
+# fitting a cox model using regular estimation, however we will not keep it.
+# this is used more as an object place holder.
+coxNet <- survival::coxph(Surv(time = d.time, event = death) ~ ., data = cleanCoxData)
+
+# extract coefficients for lollipop plot later
+estimatescoxNet <- setDT(as.data.frame(coef(coxNet)), keep.rownames = TRUE)[]
+
+estimatescoxNet$Model <- "CoxNet"
+colnames(estimatescoxNet) <- c("term", "estimate", "model")
+
+# The coefficients of this object will be replaced with the estimates from coxNet.
+# Doing so makes it so that everything is invalid aside from the coefficients.
+# In this case, all we need to estimate the absolute risk is the coefficients.
+# Std. error would be incorrect here, if we were to draw error bars.
+coxNet$coefficients <- nonzero_coef_cox@x
+# Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
+
+abcoxNet <- survival::survfit(coxNet, type = "breslow", newdata = as.data.frame(newData[, nonzero_covariate_cox$X1]))
+
+abCN <- as.data.frame(cbind(abcoxNet$time, 1 - rowMeans(abcoxNet$surv)))
+colnames(abCN) <- c("Time", "Absolute risk")
+abCN$Model <- "Coxnet"
+abCN <- abCN[abCN$Time <= max(abKMcompare[, 1]), ]
+# plot(abCN[,1],1-abCN[,2])
+```
+
+```{r cbHazard}, echo = TRUE, eval = eval_cs3}
+
+cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = 10, standardize = TRUE)
+```
+
+```{r CBabsolute, echo = TRUE, eval = eval_cs3}
+# Estimating the absolute risk curve using the newData parameter.
+abCbFull <- casebase::absoluteRisk(cbFull, time = seq(1, max(abKMcompare[, 1]), 25), newdata = newData, s = "lambda.1se", method = c("numerical"))
+```
+
+```{r cbAbProcessing,echo=FALSE,eval=eval_cs3}
+abCB <- data.frame(abCbFull[, 1], rowMeans(abCbFull[, -c(1)]))
+
+colnames(abCB) <- c("Time", "Absolute risk")
+abCB$Model <- "Penalized logistic"
+
+ab <- rbind(abCB, abCN)
+```
+
+```{r include=TRUE,eval=eval_cs3,fig.cap="\\label{fig:baselines} HERES SOME TEXT ABT IT"}
+ggplot(as.data.frame(ab), aes(Time)) +
+  geom_line(data = as.data.frame(baseCoxComparisonData), mapping = aes(y = Full, colour = "Cox"), lwd = 2) +
+  geom_line(aes(y = `Absolute risk`, colour = Model), lwd = 2) +
+  geom_step(data = as.data.frame(abKMcompare), mapping = aes(x = Time, y = KM, color = "Unadjusted"), lwd = 2) +
+  labs(y = "Probability of Death", x = "Survival time (Days)")
+```
+
+???SHORT SPIEL ABOUT IPA BETWEEN SEMI_PARAMETRIC AND PARAMETRIC which leads to conclusion of best overall model. This refers to the COEFPLOT in a following figure???
+
+```{r coefplots, eval=eval_cs3}
+
+library(dotwhisker)
+library(broom)
+library(dplyr)
+
+estimatescb <- setDT(as.data.frame(coef(cbFull)[-c(1, 2), 1]), keep.rownames = TRUE)[]
+# estimatescb$std.error<-0
+estimatescb$Model <- "casebase"
+colnames(estimatescb) <- c("term", "estimate", "model")
+
+estimatescoxFull <- setDT(as.data.frame(coef(coxFull)), keep.rownames = TRUE)[]
+# estimatescoxFull$std.error<-0
+estimatescoxFull$Model <- "CoxFull"
+colnames(estimatescoxFull) <- c("term", "estimate", "model")
+
+lolliplotDots <- rbind(estimatescb, estimatescoxNet, estimatescoxFull)
+lolliplotDots$conf.low <- lolliplotDots$estimate
+lolliplotDots$conf.high <- lolliplotDots$estimate
+lolliplotDots$conf.high[lolliplotDots$estimate < 0] <- 0
+lolliplotDots$conf.low[lolliplotDots$estimate > 0] <- 0
+
+dwplot(lolliplotDots)
+```
 
@@
 
-stanford_popTime <- popTime(jasa, time = "futime", 
-                            event = "fustat")
+stanford_popTime <- popTime(jasa,
+  time = "futime",
+  event = "fustat"
+)
 plot(stanford_popTime)
@@
 
-cb_data <- sampleCaseBase(jasa, time = "futime", 
-                          event = "fustat", ratio = 10)
+cb_data <- sampleCaseBase(jasa,
+  time = "futime",
+  event = "fustat", ratio = 10
+)
 ```
@@
   txtime = time_length(accept.dt %--% tx.date,
-                                       unit = "days"),
+    unit = "days"
+  ),
   exposure = case_when(
@@
     txtime <= futime ~ 1L
-                  ))
+  )
+)
 ```
@@
 fit1 <- fitSmoothHazard(fustat ~ exposure,
-                        data = cb_data, time = "futime")
+  data = cb_data, time = "futime"
+)
 fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
-                        data = cb_data, time = "futime")
+  data = cb_data, time = "futime"
+)
 fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
-                        data = cb_data, time = "futime")
-fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
-                        data = cb_data, time = "futime")
+  data = cb_data, time = "futime"
+)
+fit4 <- fitSmoothHazard(fustat ~ exposure * bs(futime),
+  data = cb_data, time = "futime"
+)
 ```
@@
 # Compute AIC
-c("Model1" = AIC(fit1),
+c(
+  "Model1" = AIC(fit1),
   "Model2" = AIC(fit2),
   "Model3" = AIC(fit3),
-  "Model4" = AIC(fit4))
+  "Model4" = AIC(fit4)
+)
 ```
@@
 # First, create a list of time points for both exposure status
-hazard_data <- expand.grid(exposure = c(0, 1),
+hazard_data <- expand.grid(
+  exposure = c(0, 1),
   futime = seq(0, 1000,
-                                        length.out = 100))
+    length.out = 100
+  )
+)
 # Set the offset to zero
@@
 # transform to the right scale
-hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
-                                 type = "link"))
+hazard_data$hazard <- exp(predict(fit4,
+  newdata = hazard_data,
+  type = "link"
+))
 # Add labels for plots
-hazard_data$Status = factor(hazard_data$exposure,
-                            labels = c("NoTrans", "Trans"))
+hazard_data$Status <- factor(hazard_data$exposure,
+  labels = c("NoTrans", "Trans")
+)
 
@@
   theme_minimal() +
-    theme(legend.position = 'top') +
-    ylab('Hazard') + xlab('Follow-up time')
+  theme(legend.position = "top") +
+  ylab("Hazard") +
+  xlab("Follow-up time")
 ```
@@
 newdata <- data.frame(exposure = c(0, 1))
-absrisk <- absoluteRisk(fit4, newdata = newdata, 
-                        time = seq(0, 1000, length.out = 100))
+absrisk <- absoluteRisk(fit4,
+  newdata = newdata,
+  time = seq(0, 1000, length.out = 100)
+)
 
@@
 # Rearrange the data
-absrisk <- gather(as.data.frame(absrisk),
-                  "Status", "Risk", -Time)
+absrisk <- gather(
+  as.data.frame(absrisk),
+  "Status", "Risk", -Time
+)
 
@@
   theme_minimal() +
-  theme(legend.position = 'top') +
+  theme(legend.position = "top") +
   expand_limits(y = 0.5) +
-  xlab('Follow-up time') + ylab('Cum. Incidence')
+  xlab("Follow-up time") +
+  ylab("Cum. Incidence")
 ```
@@
 one_yr_haz <- data.frame(futime = seq(0, 365,
-                                      length.out = 100))
+  length.out = 100
+))
 one_yr_haz <- mutate(one_yr_haz,
   offset = 0,
-                     exposure = if_else(futime < 30, 0, 1))
-one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
-                                type = "link"))
+  exposure = if_else(futime < 30, 0, 1)
+)
+one_yr_haz$hazard <- exp(predict(fit4,
+  newdata = one_yr_haz,
+  type = "link"
+))
 ggplot(one_yr_haz, aes(futime, hazard)) +
@@
   theme_minimal() +
-  theme(legend.position = 'top') +
-  ylab('Hazard') + xlab('Follow-up time') +
-  geom_vline(xintercept = 30, linetype = 'dashed')
+  theme(legend.position = "top") +
+  ylab("Hazard") +
+  xlab("Follow-up time") +
+  geom_vline(xintercept = 30, linetype = "dashed")
 ```
@@
 ```{r stanford-1y-risk, eval = eval_cs4}
-absoluteRisk(fit4, newdata = data.frame(exposure = 0),
-             time = 365)
+absoluteRisk(fit4,
+  newdata = data.frame(exposure = 0),
+  time = 365
+)
 
@@
 ```{r, echo=FALSE, eval = FALSE}
-knitr::kable(data.frame(`feature` = c("model type", "time", "cumulative incidence", "non-proportional hazards","model testing","competing risks", "prediction"),
-           `Cox` = c("semi-parametric","left hand side of the equation", "step function", "interaction of covariates with time"," ", "difficult", "Kaplan-Meier-based"),
-           `Case Base Sampling` = c("fully parametric (logistic/multinomial regression)", "right hand side - allows flexible modeling of time",
+knitr::kable(data.frame(
+  `feature` = c("model type", "time", "cumulative incidence", "non-proportional hazards", "model testing", "competing risks", "prediction"),
+  `Cox` = c("semi-parametric", "left hand side of the equation", "step function", "interaction of covariates with time", " ", "difficult", "Kaplan-Meier-based"),
+  `Case Base Sampling` = c(
+    "fully parametric (logistic/multinomial regression)", "right hand side - allows flexible modeling of time",
     "smooth-in-time curve", "interaction of covariates with time", "make use of GLM framework (LRT, AIC, BIC)",
-           "cause-specific cumulative incidence functions (CIFs) directly obtained via multinomial regression", "ROC, AUC, risk reclassification probabilities")), 
+    "cause-specific cumulative incidence functions (CIFs) directly obtained via multinomial regression", "ROC, AUC, risk reclassification probabilities"
+  )
+),
 col.names = c("Feature", "Cox model", "Case-base sampling"),
@@
 caption = "Comparison between the Cox model and case-base sampling",
-           escape = FALSE)
+escape = FALSE
+)
 ```

### 2020-03-06 analysis/paper/paper.tex
Fixed lollipop plot
to do:
fix casebase absoluterisk for glmnet. 
fix ??? tags

@@
 R>   levels = c(0, 1),
-R>                        labels = c("Control group", "Screening group"))
+R>   labels = c("Control group", "Screening group")
+R> )
 \end{CodeInput}
@@
 R>   event = "DeadOfPrCa",
-R>                                      exposure = "ScrArm")
+R>   exposure = "ScrArm"
+R> )
 R> plot(pt_object_strat)
@@
 R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm,
-R>                              data = ERSPC)
+R>   data = ERSPC
+R> )
 R> summary(cox_model)
@@
 \begin{CodeInput}
-R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
-R>                        ignore = 99)
+R> new_data <- data.frame(
+R>   ScrArm = c("Control group", "Screening group"),
+R>   ignore = 99
+R> )
 R> 
@@
 R>   fun = "event",
-R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
-R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+R>   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
+R>   main = sprintf(
+R>     "Estimated Cumulative Incidence (risk) of Death from Prostate 
 R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
@@
 R>     exp(confint(cox_model))[1],
-R>                     exp(confint(cox_model))[2]),
-R>      ylim = c(0, 2), yscale = 100)
+R>     exp(confint(cox_model))[2]
+R>   ),
+R>   ylim = c(0, 2), yscale = 100
+R> )
 R> legend("topleft",
 R>   legend = c("Control group", "Screening group"),
-R>        col = c("red","blue"),
+R>   col = c("red", "blue"),
 R>   lty = c(1, 1),
-R>        bg = "gray90")
+R>   bg = "gray90"
+R> )
 \end{CodeInput}
@@
 R>   data = ERSPC,
-R>                                                   ratio = 100)
+R>   ratio = 100
+R> )
 R> 
@@
 R> exp(coef(casebase_exponential)[2])
-R> exp(confint(casebase_exponential)[2,])
+R> exp(confint(casebase_exponential)[2, ])
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
-R>                                           time = seq(0,15,0.1), 
-R>                                           newdata = new_data)
+R> smooth_risk_exp <- casebase::absoluteRisk(
+R>   object = casebase_exponential,
+R>   time = seq(0, 15, 0.1),
+R>   newdata = new_data
+R> )
 R> 
@@
 R>     exp(confint(cox_model))[1],
-R>                     exp(confint(cox_model))[2]),
-R>      ylim = c(0, 2), yscale = 100)
-R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
-R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
+R>     exp(confint(cox_model))[2]
+R>   ),
+R>   ylim = c(0, 2), yscale = 100
+R> )
+R> lines(smooth_risk_exp[, 1], smooth_risk_exp[, 2], col = "red", lty = 2)
+R> lines(smooth_risk_exp[, 1], smooth_risk_exp[, 3], col = "blue", lty = 2)
 R> 
@@
 R> legend("topleft",
-R>        legend = c("Control group (Cox)","Control group (Casebase)",
-R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R>        col = c("red","red", "blue","blue"),
+R>   legend = c(
+R>     "Control group (Cox)", "Control group (Casebase)",
+R>     "Screening group (Cox)", "Screening group (Casebase)"
+R>   ),
+R>   col = c("red", "red", "blue", "blue"),
 R>   lty = c(1, 2, 1, 2),
-R>        bg = "gray90")
+R>   bg = "gray90"
+R> )
 \end{CodeInput}
@@
 R>   data = ERSPC,
-R>                                  ratio = 100)
+R>   ratio = 100
+R> )
 R> 
@@
 \begin{CodeInput}
-R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
-R>                                           time = seq(0,15,0.1), 
-R>                                           newdata = new_data)
+R> smooth_risk_time <- casebase::absoluteRisk(
+R>   object = casebase_time,
+R>   time = seq(0, 15, 0.1),
+R>   newdata = new_data
+R> )
 R> 
@@
 R>     exp(confint(cox_model))[1],
-R>                     exp(confint(cox_model))[2]),
-R>      ylim = c(0, 2), yscale = 100)
-R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
-R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
+R>     exp(confint(cox_model))[2]
+R>   ),
+R>   ylim = c(0, 2), yscale = 100
+R> )
+R> lines(smooth_risk_time[, 1], smooth_risk_time[, 2], col = "red", lty = 2)
+R> lines(smooth_risk_time[, 1], smooth_risk_time[, 3], col = "blue", lty = 2)
 R> 
 R> legend("topleft",
-R>        legend = c("Control group (Cox)","Control group (Casebase)",
-R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R>        col = c("red","red", "blue","blue"),
+R>   legend = c(
+R>     "Control group (Cox)", "Control group (Casebase)",
+R>     "Screening group (Cox)", "Screening group (Casebase)"
+R>   ),
+R>   col = c("red", "red", "blue", "blue"),
 R>   lty = c(1, 2, 1, 2),
-R>        bg = "gray90")
+R>   bg = "gray90"
+R> )
 \end{CodeInput}
@@
 R>   data = ERSPC,
-R>                                     ratio = 100)
+R>   ratio = 100
+R> )
 R> 
@@
 \begin{CodeInput}
-R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
-R>                                     time = seq(0,15,0.1), 
-R>                                     newdata = new_data)
+R> smooth_risk_splines <- absoluteRisk(
+R>   object = casebase_splines,
+R>   time = seq(0, 15, 0.1),
+R>   newdata = new_data
+R> )
 R> 
@@
 R>     exp(confint(cox_model))[1],
-R>                     exp(confint(cox_model))[2]),
-R>      ylim = c(0, 2), yscale = 100)
-R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
-R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
+R>     exp(confint(cox_model))[2]
+R>   ),
+R>   ylim = c(0, 2), yscale = 100
+R> )
+R> lines(smooth_risk_splines[, 1], smooth_risk_splines[, 2], col = "red", lty = 2)
+R> lines(smooth_risk_splines[, 1], smooth_risk_splines[, 3], col = "blue", lty = 2)
 R> 
 R> legend("topleft",
-R>        legend = c("Control group (Cox)","Control group (Casebase)",
-R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
-R>        col = c("red","red", "blue","blue"),
+R>   legend = c(
+R>     "Control group (Cox)", "Control group (Casebase)",
+R>     "Screening group (Cox)", "Screening group (Casebase)"
+R>   ),
+R>   col = c("red", "red", "blue", "blue"),
 R>   lty = c(1, 2, 1, 2),
-R>        bg = "gray90")
+R>   bg = "gray90"
+R> )
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> c("Exp." = AIC(casebase_exponential),
+R> c(
+R>   "Exp." = AIC(casebase_exponential),
 R>   "Weib." = AIC(casebase_time),
-R>   "Splines" = AIC(casebase_splines))
+R>   "Splines" = AIC(casebase_splines)
+R> )
 \end{CodeInput}
@@
 R>   ratio = 100,
-R>     time = "ftime")
+R>   time = "ftime"
+R> )
 \end{CodeInput}
@@
 R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-R>                    data = bmtcrr)
+R>   data = bmtcrr
+R> )
 \end{CodeInput}
@@
 R> # Data.frame containing risk profile
-R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
-R>                                      levels = levels(bmtcrr[,"Sex"])),
+R> newdata <- data.frame(
+R>   "Sex" = factor(c("F", "F"),
+R>     levels = levels(bmtcrr[, "Sex"])
+R>   ),
 R>   "D" = c("ALL", "AML"), # Both diseases
 R>   "Phase" = factor(c("Relapse", "Relapse"),
-R>                                        levels = levels(bmtcrr[,"Phase"])),
+R>     levels = levels(bmtcrr[, "Phase"])
+R>   ),
 R>   "Age" = c(35, 35),
 R>   "Source" = factor(c("PB", "PB"),
-R>                                         levels = levels(bmtcrr[,"Source"])))
+R>     levels = levels(bmtcrr[, "Source"])
+R>   )
+R> )
 R> 
 R> # Estimate absolute risk curve
-R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
-R>                         method = "numerical", newdata = newdata)
+R> risk_cb <- absoluteRisk(
+R>   object = model_cb, time = time_points,
+R>   method = "numerical", newdata = newdata
+R> )
 \end{CodeInput}
@@
 R>   const(Phase) + const(Source) + const(Age),
-R>                       data = bmtcrr, cause = 1, model = "fg")
+R> data = bmtcrr, cause = 1, model = "fg"
+R> )
 R> 
@@
 \begin{CodeChunk}
-
-\begin{CodeInput}
-R> support<-casebase::support
-R> 
-R> support_popTime <- popTime(support,
-R+                            time = "d.time", event = "death",percentile_number = 0.31)
-R> plot(support_popTime)
-\end{CodeInput}
 \begin{figure}
@@
 \begin{CodeChunk}
-
-\begin{CodeInput}
-R> completeData<-casebase::support
-R> workingCompleteData=model.matrix(death~ .-d.time,data=completeData)[,-c(1)]#remove intercept
-R> #Create u and xCox, which will be used when fitting Cox with glmnet.
-R> newData=workingCompleteData
-R> x=workingCompleteData
-R> #x and y will be used to fit the casebase model under glmnet.
-R> #x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
-R> y=data.matrix(completeData[,c(2,5)])
-R> 
-R> fullDataCox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),x))
-R> train_index <- sample(1:nrow(fullDataCox), 0.8 * nrow(fullDataCox))
-R> test_index <- setdiff(1:nrow(fullDataCox), train_index)
-R> 
-R> train<-fullDataCox[train_index,]
-R> test<-fullDataCox[test_index,]
-R> coxSPS<- survival::coxph(Surv(time=V1,event=V2) ~ sps, data = train)
-R> abcoxSPS<-survival::survfit(coxSPS,newdata = test)
-R> 
-R> coxAPS<- survival::coxph(Surv(time=V1,event=V2) ~ aps, data = train)
-R> abcoxAPS<-survival::survfit(coxAPS,newdata = test)
-R> 
-R> coxFull<- survival::coxph(Surv(time=V1,event=V2) ~ .-sps -aps, data = train)
-R> abcoxFull<-survival::survfit(coxAPS,newdata = test)
-R> #library(prodlim)
-R> 
-R> 
-R> KM<- survival::coxph(Surv(time=V1,event=V2) ~ 1, data = test)
-R> abKM<-survival::survfit(KM)
-R> #abKM <- prodlim(Hist(V1,V2)~1,data=test) minimal difference
-R> abKMcompare<-cbind(abKM$time[abKM$time<=1631],1-abKM$surv[abKM$time<=1631])
-R> colnames(abKMcompare)<-c("Time","KM")
-R> 
-R> baseCoxComparisonData<-cbind(abcoxSPS$time[abcoxSPS$time<=1631],1-rowMeans(abcoxSPS$surv[abcoxSPS$time<=1631,]),1-rowMeans(abcoxAPS$surv[abcoxSPS$time<=1631,]),1-abKM$surv,1-rowMeans(abcoxFull$surv[abcoxSPS$time<=1631,]))
-R> colnames(baseCoxComparisonData)<-c("Time","SPS","APS","Unadjusted","Full")
-R> myColors <- brewer.pal(5,"Dark2")
-R> FigureWidth='90%'
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
 \begin{figure}
@@
 \begin{CodeInput}
+R> 
+R> cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = ratio, standardize = TRUE)
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Estimating the absolute risk curve using the newData parameter.
+R> abCbFull <- casebase::absoluteRisk(cbFull, time = seq(1, max(abKMcompare[, 1]), 25), newdata = newData, s = "lambda.1se", method = c("numerical"))
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
 R> ggplot(as.data.frame(ab), aes(Time)) +
-R+     geom_line(data=as.data.frame(baseCoxComparisonData),mapping=aes(y = Full, colour = "Cox"),lwd=2)+
-R+   geom_line(aes(y = `Absolute risk`, colour = Model),lwd=2)+
-R+   geom_step(data=as.data.frame(abKMcompare),mapping=aes(x=Time,y=KM,color="Unadjusted"),lwd=2)+
-R+   scale_colour_manual(name = "Models",values = myColors[c(4,1,3,2)])+
-R+   labs(y= "Probability of Death", x = "Survival time (Days)")
+R+   geom_line(data = as.data.frame(baseCoxComparisonData), mapping = aes(y = Full, colour = "Cox"), lwd = 2) +
+R+   geom_line(aes(y = `Absolute risk`, colour = Model), lwd = 2) +
+R+   geom_step(data = as.data.frame(abKMcompare), mapping = aes(x = Time, y = KM, color = "Unadjusted"), lwd = 2) +
+R+   labs(y = "Probability of Death", x = "Survival time (Days)")
 \end{CodeInput}
@@
 R> 
-R> estimatescb=setDT(as.data.frame(coef(cbFull)[-c(1,2),1]), keep.rownames = TRUE)[]
-R> #estimatescb$std.error<-0
-R> estimatescb$Model="casebase"
-R> colnames(estimatescb)<-c("term","estimate","model")
-R> 
-R> estimatescoxFull=setDT(as.data.frame(coef(coxFull)), keep.rownames = TRUE)[]
-R> #estimatescoxFull$std.error<-0
-R> estimatescoxFull$Model="CoxFull"
-R> colnames(estimatescoxFull)<-c("term","estimate","model")
-R> 
-R> lolliplotDots<-rbind(estimatescb,estimatescoxNet,estimatescoxFull)
-R> lolliplotDots$conf.low<-NA
-R> lolliplotDots$conf.high<-NA
-R> lolliplotDots$conf.high[lolliplotDots$estimate< 0]<-0
-R> 
-R> lolliplotDots$conf.low[lolliplotDots$estimate> 0]<-0
+R> estimatescb <- setDT(as.data.frame(coef(cbFull)[-c(1, 2), 1]), keep.rownames = TRUE)[]
+R> # estimatescb$std.error<-0
+R> estimatescb$Model <- "casebase"
+R> colnames(estimatescb) <- c("term", "estimate", "model")
+R> 
+R> estimatescoxFull <- setDT(as.data.frame(coef(coxFull)), keep.rownames = TRUE)[]
+R> # estimatescoxFull$std.error<-0
+R> estimatescoxFull$Model <- "CoxFull"
+R> colnames(estimatescoxFull) <- c("term", "estimate", "model")
+R> 
+R> lolliplotDots <- rbind(estimatescb, estimatescoxNet, estimatescoxFull)
+R> lolliplotDots$conf.low <- lolliplotDots$estimate
+R> lolliplotDots$conf.high <- lolliplotDots$estimate
+R> lolliplotDots$conf.high[lolliplotDots$estimate < 0] <- 0
+R> lolliplotDots$conf.low[lolliplotDots$estimate > 0] <- 0
 R> 
@@
 R> 
-R> stanford_popTime <- popTime(jasa, time = "futime", 
-R>                             event = "fustat")
+R> stanford_popTime <- popTime(jasa,
+R>   time = "futime",
+R>   event = "fustat"
+R> )
 R> plot(stanford_popTime)
@@
 R> 
-R> cb_data <- sampleCaseBase(jasa, time = "futime", 
-R>                           event = "fustat", ratio = 10)
+R> cb_data <- sampleCaseBase(jasa,
+R>   time = "futime",
+R>   event = "fustat", ratio = 10
+R> )
 \end{CodeInput}
@@
 R>   txtime = time_length(accept.dt %--% tx.date,
-R>                                        unit = "days"),
+R>     unit = "days"
+R>   ),
 R>   exposure = case_when(
@@
 R>     txtime <= futime ~ 1L
-R>                   ))
+R>   )
+R> )
 \end{CodeInput}
@@
 R> fit1 <- fitSmoothHazard(fustat ~ exposure,
-R>                         data = cb_data, time = "futime")
+R>   data = cb_data, time = "futime"
+R> )
 R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
-R>                         data = cb_data, time = "futime")
+R>   data = cb_data, time = "futime"
+R> )
 R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
-R>                         data = cb_data, time = "futime")
-R> fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
-R>                         data = cb_data, time = "futime")
+R>   data = cb_data, time = "futime"
+R> )
+R> fit4 <- fitSmoothHazard(fustat ~ exposure * bs(futime),
+R>   data = cb_data, time = "futime"
+R> )
 \end{CodeInput}
@@
 R> # Compute AIC
-R> c("Model1" = AIC(fit1),
+R> c(
+R>   "Model1" = AIC(fit1),
 R>   "Model2" = AIC(fit2),
 R>   "Model3" = AIC(fit3),
-R>   "Model4" = AIC(fit4))
+R>   "Model4" = AIC(fit4)
+R> )
 \end{CodeInput}
@@
 R> # First, create a list of time points for both exposure status
-R> hazard_data <- expand.grid(exposure = c(0, 1),
+R> hazard_data <- expand.grid(
+R>   exposure = c(0, 1),
 R>   futime = seq(0, 1000,
-R>                                         length.out = 100))
+R>     length.out = 100
+R>   )
+R> )
 R> # Set the offset to zero
@@
 R> # transform to the right scale
-R> hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
-R>                                  type = "link"))
+R> hazard_data$hazard <- exp(predict(fit4,
+R>   newdata = hazard_data,
+R>   type = "link"
+R> ))
 R> # Add labels for plots
-R> hazard_data$Status = factor(hazard_data$exposure,
-R>                             labels = c("NoTrans", "Trans"))
+R> hazard_data$Status <- factor(hazard_data$exposure,
+R>   labels = c("NoTrans", "Trans")
+R> )
 R> 
@@
 R>   theme_minimal() +
-R>     theme(legend.position = 'top') +
-R>     ylab('Hazard') + xlab('Follow-up time')
+R>   theme(legend.position = "top") +
+R>   ylab("Hazard") +
+R>   xlab("Follow-up time")
 \end{CodeInput}
@@
 R> newdata <- data.frame(exposure = c(0, 1))
-R> absrisk <- absoluteRisk(fit4, newdata = newdata, 
-R>                         time = seq(0, 1000, length.out = 100))
+R> absrisk <- absoluteRisk(fit4,
+R>   newdata = newdata,
+R>   time = seq(0, 1000, length.out = 100)
+R> )
 R> 
@@
 R> # Rearrange the data
-R> absrisk <- gather(as.data.frame(absrisk),
-R>                   "Status", "Risk", -Time)
+R> absrisk <- gather(
+R>   as.data.frame(absrisk),
+R>   "Status", "Risk", -Time
+R> )
 R> 
@@
 R>   theme_minimal() +
-R>   theme(legend.position = 'top') +
+R>   theme(legend.position = "top") +
 R>   expand_limits(y = 0.5) +
-R>   xlab('Follow-up time') + ylab('Cum. Incidence')
+R>   xlab("Follow-up time") +
+R>   ylab("Cum. Incidence")
 \end{CodeInput}
@@
 R> one_yr_haz <- data.frame(futime = seq(0, 365,
-R>                                       length.out = 100))
+R>   length.out = 100
+R> ))
 R> one_yr_haz <- mutate(one_yr_haz,
 R>   offset = 0,
-R>                      exposure = if_else(futime < 30, 0, 1))
-R> one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
-R>                                 type = "link"))
+R>   exposure = if_else(futime < 30, 0, 1)
+R> )
+R> one_yr_haz$hazard <- exp(predict(fit4,
+R>   newdata = one_yr_haz,
+R>   type = "link"
+R> ))
 R> ggplot(one_yr_haz, aes(futime, hazard)) +
@@
 R>   theme_minimal() +
-R>   theme(legend.position = 'top') +
-R>   ylab('Hazard') + xlab('Follow-up time') +
-R>   geom_vline(xintercept = 30, linetype = 'dashed')
+R>   theme(legend.position = "top") +
+R>   ylab("Hazard") +
+R>   xlab("Follow-up time") +
+R>   geom_vline(xintercept = 30, linetype = "dashed")
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> absoluteRisk(fit4, newdata = data.frame(exposure = 0),
-R>              time = 365)
+R> absoluteRisk(fit4,
+R>   newdata = data.frame(exposure = 0),
+R>   time = 365
+R> )
 R> 
@@
 
-This report was generated on 2020-03-05 23:20:03 using the following
+This report was generated on 2020-03-06 08:01:48 using the following
 computational environment and dependencies:
@@
 #> other attached packages:
-#>  [1] broom_0.5.5         dotwhisker_0.5.0    RColorBrewer_1.1-2 
-#>  [4] pracma_2.2.9        kableExtra_1.1.0    data.table_1.12.8  
-#>  [7] glmnet_3.0-2        Matrix_1.2-17       casebase_0.2.1.9001
-#> [10] survival_3.1-8      magrittr_1.5        forcats_0.5.0      
-#> [13] stringr_1.4.0       dplyr_0.8.3         purrr_0.3.3        
-#> [16] readr_1.3.1         tidyr_1.0.2         tibble_2.1.3       
-#> [19] ggplot2_3.3.0       tidyverse_1.3.0    
+#>  [1] broom_0.5.5         dotwhisker_0.5.0    pracma_2.2.9       
+#>  [4] kableExtra_1.1.0    data.table_1.12.8   glmnet_3.0-2       
+#>  [7] Matrix_1.2-17       casebase_0.2.1.9001 survival_3.1-8     
+#> [10] magrittr_1.5        forcats_0.5.0       stringr_1.4.0      
+#> [13] dplyr_0.8.3         purrr_0.3.3         readr_1.3.1        
+#> [16] tidyr_1.0.2         tibble_2.1.3        ggplot2_3.3.0      
+#> [19] tidyverse_1.3.0    
 #> 
@@
 #> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
-#> Head:     [843365a] 2020-03-05: In process of changing my results
+#> Head:     [6682dfb] 2020-03-06: NOT COMPLETE
 \end{CodeOutput}

### 2020-03-20 analysis/paper/paper.Rmd
still working on poptime section

--- 2020-03-05 analysis/paper/paper.Rmd
@@
 library(glmnet)
-library(data.table)
 library(kableExtra)
 library(pracma)
+library(flexsurv) # for fitting gompertz AFT
 options(digits = 2, scipen = 999)
@@
 eval_cs2 <- TRUE
-eval_cs3 <- FALSE
+eval_cs3 <- TRUE
 eval_cs4 <- TRUE
@@
 
-# Theoretical details
+# Theoretical details {#theory}
 
@@
 
-To describe the theoretical foundations of case-base sampling, we use the framework of counting processes.
-In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras; the interested reader can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case].
-First, let $N_{i}(t) \in \{0, 1\}$ be counting processes corresponding to the event of interest for individual $i=1, \ldots,n$.
-For simplicity, we will consider Type I censoring due to the end of follow-up at time $\tau$ (the general case of non-informative censoring is treated in Saarela [-@saarela2016case]).
-We assume a continuous time model, which implies that the counting process jumps are less than or equal to one.
-We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
-$$\lambda_{i}(t) dt = E[dN_{i}(t)].$$
-
-Next, we model the base series sampling mechanism using non-homogeneous Poisson processes $R_i(t) \in \{0, 1, 2, \ldots\}$, with the person-moments where $dR_i(t) = 1$ constituting the base series.
-The process $Q_{i}(t) = R_i(t) + N_{i}(t)$ then counts both the case and base series person-moments contributed by individual $i$.
-This process is typically defined by the user via its hazard function $\rho_i(t)$.
-The process $Q_{i}(t)$ is characterized by $E[dQ_{i}(t)] = \lambda_{i}(t)dt + \rho_i(t)dt$.
-
-If the hazard function $\lambda_{i}(t; \theta)$ is parametrized in terms of $\theta$, we could define an estimator $\hat{\theta}$ by maximization of the likelihood expression
-$$L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^\tau \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)}\left(1 - \lambda_{i}(t;\theta)\right)^{1 - dN_{i}(t)},$$
-where $\prod_{t\in[0,u)}$ represents a product integral from $0$ to $u$. However, the integral over time makes the computation and maximisation of $L_0(\theta)$ challenging.
-
-Case-base sampling allows us to avoid this integral. By conditioning on a sampled person-moment, we get individual contributions of the form
-$$P(dN_{i}(t) \mid dQ_{i}(t) = 1) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.$$
-Therefore, we can define an estimating equation for $\theta$ as follows:
-$$L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.$$
-When a logarithmic link function is used for modeling the hazard function, the above expression is of a logistic regression form with an offset term $\log(1/\rho_i(t))$.
-Note that the sampling units selected in the case-base sampling mechanism are person-moments, rather than individuals, and the parameters to be estimated are hazards or hazard ratios rather than odds or odds ratios.
-Generally, an individual can contribute more than one person-moment, and thus the terms in the product integral are not independent.
-Nonetheless, Saarela [-@saarela2016case] showed that the logarithm of this estimating function has mean zero at the true value $\theta=\theta_0$, and that the resulting estimator $\hat{\theta}$ is asymptotically normally distributed.
+<!-- Popular survival analysis methods, like Kaplan-Meier and Cox regression, rely on the notion of risk-set sampling. Case-base sampling can be seen as an alternative. -->
+
+As shown by Saarela & Arjas [-@saarela2015non] (and further expanded in Saarela [-@saarela2016case]), writing the likelihood arising from this data-generating mechanism using the framework of non-homogeneous Poisson processes, we eventually reach an expression where each person-moment's contribution is of the form
+$$\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},$$
+where $N(t)$ is the counting process associated with the event of interest, $h(t)$ is the corresponding hazard function, and $\rho(t)$ is the hazard function for the Poisson process associated with case-base sampling.
+This parametric form suggests that we can readily estimate log-hazards of the form $\log(h(t)) = g(t; X)$ using logistic regression, where each observation corresponds to a person moment, the function $g(t; X)$ is linear in a finite number of parameters, and where we treat $-\log(\rho(t))$ as an offset.
 
 In Hanley & Miettinen [-@hanley2009fitting], the authors suggest sampling the base series *uniformly* from the study base.
-In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with hazard equal to $\rho_i(t) = b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base.
+In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base.
 More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism are proportional to the cardiovascular disease event rate given by the Framingham score.
@@
 
-Let $g(t; X)$ be the linear predictor such that $\log(\lambda(t;X)) = g(t; X)$.
+<!-- The `casebase` package fits the family of hazard functions of the form -->
+
+<!-- $$ h(t;X) = \exp[g(t;X)] $$ -->
+<!-- where $t$ denotes time and $X$, the individual's covariate profile.  -->
+
+Let $g(t; X)$ be the linear predictor such that $\log(h(t;X)) = g(t; X)$.
 Different functions of $t$ lead to different parametric hazard models.
@@
 
-$$ \log(\lambda(t; X)) = \beta_0 + \beta_1 X.$$
+$$ \log(h(t; X)) = \beta_0 + \beta_1 X.$$
 
@@
 
-$$ \log(\lambda(t; X)) = \beta_0 + \beta_1 t + \beta_2 X.$$
+$$ \log(h(t; X)) = \beta_0 + \beta_1 t + \beta_2 X.$$
 
@@
 
-$$ \log(\lambda(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. $$
+$$ \log(h(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. $$
 
@@
 
-$$\frac{\lambda_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},$$
-
-where $N_j(t)$ is the counting process associated with the event of type $j$ and $\lambda_j(t)$ is the corresponding hazard function.
+$$\frac{h_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^Jh_j(t)},$$
+
+where $N_j(t)$ is the counting process associated with the event of type $j$ and $h_j(t)$ is the corresponding hazard function.
 As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}
@@
 
-Several \proglang{R} packages implement methodologies for competing risk analysis; for a different perspective, see Mahani & Sharabiani [-@mahani2015bayesian].
-The package \pkg{cmprsk} provides methods for cause-specific subdistributions, such as in the Fine-Gray model [-@fine1999proportional].
-On the other hand, the package \pkg{CFC} estimates cause-specific cumulative incidence functions from unadjusted, non-parametric survival functions.
-Our package \pkg{casebase} also provides functionalities for competing risk analysis by estimating the cause-specific hazards, and from these quantities, we can derive the cause-specific cumulative incidence functions.
+Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets} and \pkg{survival} contain implementations for competing risks.
+The differentiating factor between these packages and \pkg{casebase} is that \pkg{casebase} handles competing risks through multiple logistic regression.
+\textcolor{red}{There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}.
+Could you incorporate some of their discussion, with proper citations?
+The biggest omission right now is \pkg{cmprsk}, which we actually use below.}
 
@@
 \hline
-\textbf{casebase} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
-\textbf{CFC} \cite{mahani2015bayesian}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
-textbf{cmprsk} \cite{gray_2019}            & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\
+\textbf{casebasee} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
+\textbf{CFC} \cite{sharabiani_mahani_2019}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
 \textbf{coxRidge} \cite{perperoglou}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
@@
 ## Population-time plots
+
+The case-base sampling approach described in Section \ref{theory} can be visualized in the form of a population time plot.
+They are extremely informative graphical displays of survival data and should be one of the first steps in an exploratory data analysis.
+We facilitate this task in the \pkg{casebase} package using the \code{popTime} function and \code{plot} method as follows:
+
+1.
+The \code{casebase::popTime} function takes as input the original dataset along with the column names corresponding to the timescale, the event status and an exposure group of interest (optional).
+This will create an object of class \code{popTime} (or \code{popTimeExposure} if you specify a value for the exposure argument).
+2. The corresponding \code{plot} methods for the object created in Step 1 can be called to create the population time plot with several options for customizing the aesthetics.
+
+To illustrate these functions we will use the European Randomized Study of Prostate Cancer Screening data which is available through the \pkg{casebase} package.
+The results of this study were published by [@schroder2009screening], and the dataset itself was obtained using the approach described in [@liu2014recovering].
+
+We first create the necessary dataset for producing the population time plots using the \code{casebase::popTime} function.
+In this example we stratify by the treatment group.
+Note that the resulting object stores the exposure variable as an attribute:
+
+```{r plot-erspc-data, eval = eval_cs1}
+data("ERSPC")
+ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
+                       levels = c(0,1), 
+                       labels = c("Control group", "Screening group"))
+pt_object <- casebase::popTime(ERSPC, 
+                               time = "Follow.Up.Time",
+                               event = "DeadOfPrCa",
+                               exposure = "ScrArm")
+attr(pt_object, "exposure")
+```
+
+<!--The \code{pt_object} is a \code{data.table} containing the original dataset augmented with-->
+We then pass this object to the \code{plot} function which leverages the \pkg{ggplot2} package [@ggplot2] to build population time plots.
+It builds the plot by sequentially adding layers, starting with a layer for the area representing the population time followed by points to show the casebase sampling mechanism.
+This function gives user the flexibility to add any combination of the case.series, base.series and competing events.
+The case series and competing events are sampled at random vertically on the plot in order to visualise the incidence density using the popTime function.
+That is, imagine we draw a vertical line at a specific event time.
+We then plot the point at a randomly sampled y-coordinate along this vertical line.
+This is done to avoid having all points along the upper edge of the plot (because the subjects with the least amount of observation time are plotted at the top of the y-axis).
+By randomly distributing them, we can get a better sense of the inicidence density.
+The base series is sampled horizontally on the plot using the sampleCaseBase function.
+
+One benefit from these plots is that it allows you to see the incidence density.
+This can be seen from the distribution of the red dots in the above plot.
+We can see that more events are observed later on in time.
+Therefore a constant hazard model would not be appropriate in this instance as it would overestimate the cumulative incidence earlier on in time, and underestimate it later on.
+
+The unique _step shape_ of the population time plot is due to the randomization of the Finnish cohorts which were carried out on January 1 of each of the 4 years 1996 to 1999.
+This, coupled with the uniform December 31 2006 censoring date, lead to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up.
+
+It is important to note that the red points are random distributed across the grey area for each time of event.
+That is, imagine we draw a vertical line at a specific event time.
+We then plot the red point at a randomly sampled y-coordinate along this vertical line.
+This is done to avoid having all red points along the upper edge of the plot (because the subjects with the least amount of observation time are plotted at the top of the y-axis).
+By randomly distributing them, we can get a better sense of the inicidence density.
+
+We can also create exposure stratified plots by specifying the \code{exposure} argument in the \code{popTime} function:
+
+```{r plot-stratified-erspc-data, eval = eval_cs1}
+pt_object_strat <- casebase::popTime(ERSPC, 
+                                     event = "DeadOfPrCa", 
+                                     exposure = "ScrArm")
+# plot(pt_object_strat)
+```
+
+We can also plot them side-by-side using the \code{ncol} argument:
+
+```{r plot-stratified-erspc-data-side, eval = eval_cs1}
+plot(pt_object_strat, ncol = 2)
+```
+
+ADD PARAGRAPH ABOUT WHAT WE CAN CONCLUDE FROM THESE GRAPHS.
+
+For a more detailed description of how to modify the aesthetics of a population time plot, please refer to the [Customizing Population Time Plots](http://sahirbhatnagar.com/casebase/articles/customizingpopTime.html) vignette.
 
@@
 
-The sampling scheme selected for \code{sampleCaseBase} was described in Hanley & Miettinen [-@hanley2009fitting]: we first sample along the ``person'' axis, proportional to each individual's total follow-up time, and then we sample a moment uniformly over their follow-up time.
+The sampling scheme selected for \code{sampleCaseBase} was described in Hanley and Miettinen [-@hanley2009fitting]: we first sample along the ``person'' axis, proportional to each individual's total follow-up time, and then we sample a moment uniformly over their follow-up time.
 This sampling scheme is equivalent to the following picture: imagine representing the total follow-up time of all individuals in the study along a single dimension, where the follow-up time of the next individual would start exactly when the follow-up time of the previous individual ends.
@@
 If the dataset supplied to \code{fitSmoothHazard} does not inherit from \code{cbData}, then the fitting function starts by calling \code{sampleCaseBase} to generate the base series.
-In other words, users can bypass \code{sampleCaseBase} altogether and only worry about the fitting function \code{fitSmoothHazard}.
+In other words, the occasional user can bypass \code{sampleCaseBase} altogether and only worry about the fitting function \code{fitSmoothHazard}.
 
@@
 The right-hand side can be any combination of the covariates, along with an explicit functional form for the time variable.
-Note that non-proportional hazard models can be achieved at this stage by adding an interaction term involving time (cf.
-Case Study 4 below).
-The offset term does not need to be specified by the user, as it is automatically added to the formula before calling \code{glm}.
+Note that non-proportional hazard models can be achieved at this stage by adding an interaction term involving time.
+The offset term does not need to be specified by the user, as it is automatically added to the formula.
 
@@
 In the case of multiple events, the hazard is fitted via multinomial regression as performed by the \pkg{VGAM} package.
-We selected this package for its ability to fit multinomial regression models with an offset.
+This package was selected for its ability to fit multinomial regression models with an offset.
 
@@
 \begin{equation}\label{eqn:surv}
-S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right).
+S(t) = \exp\left(-\int_0^t h(u;X) du\right).
 \end{equation}
@@
 
-$$ S(t) = \exp\left(-\int_0^t \Lambda(u;X) du\right),\qquad \Lambda(t;X) = \sum_{j=1}^J \lambda_j(t;X).$$
+$$ S(t) = \exp\left(-\int_0^t H(u;X) du\right),\qquad H(t;X) = \sum_{j=1}^J h_j(t;X).$$
 From this, we can derive the event-specific subdensities:
 
-$$ f_j(t) = \lambda_j(t)S(t).$$
+$$ f_j(t) = h_j(t)S(t).$$
 
@@
 $$ CI_j(t) = \int_0^t f_j(u)du.$$
-The integrals are computed using either numerical integration (via the trapezoidal rule) or Monte Carlo integration.
-This option is controlled by the argument \code{method} of the \code{absoluteRisk} function.
-
-We created \code{absoluteRisk} as an \code{S3} generic, with methods that match the different types of outputs of \code{fitSmoothHazard}.
-The method dispatch system of \proglang{R} then takes care of matching the correct output to the correct methodology for calculating the cumulative incidence function, without requiring any intervention from the user.
-
-In the following sections, we illustrate these functionalities in the context of four case studies.
+
+We created \code{absoluteRisk} as an \code{S3} generic, with methods for the different types of outputs of \code{fitSmoothHazard}.
+The method dispatch system of \proglang{R} then takes care of matching the correct output to the correct methodology for calculating the cumulative incidence function, without the user's intervention.
+
+In the following sections, we illustrate these functionalities in the context of three case studies.
 
 # Case study 1--European Randomized Study of Prostate Cancer Screening
-
-For the first case study, we will use of the European Randomized Study of Prostate Cancer Screening data. This dataset is available through the \pkg{casebase} package:
-
-```{r erspc-data, eval = eval_cs1}
-data(ERSPC)
-ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
-                       levels = c(0, 1), 
-                       labels = c("Control group", "Screening group"))
-```
-
-The results of this study were published by [@schroder2009screening], and the dataset itself was obtained using the approach described in [@liu2014recovering].
-
-Population time plots can be extremely informative graphical displays of survival data.
-They should be the first step in an exploratory data analysis.
-We facilitate this task in the \pkg{casebase} package using the \code{popTime} function.
-We first create the necessary dataset for producing the population time plots, and we can generate the plot by using the corresponding \code{plot} method:
-
-```{r plot-erspc-data, eval = eval_cs1}
-pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
-plot(pt_object)
-```
-
-We can also create exposure stratified plots by specifying the \code{exposure} argument in the \code{popTime} function:
-
-```{r plot-stratified-erspc-data, eval = eval_cs1}
-pt_object_strat <- casebase::popTime(ERSPC, 
-                                     event = "DeadOfPrCa", 
-                                     exposure = "ScrArm")
-plot(pt_object_strat)
-```
-
-We can also plot them side-by-side using the \code{ncol} argument:
-
-```{r plot-stratified-erspc-data-side, eval = eval_cs1}
-plot(pt_object_strat, ncol = 2)
-```
-
-ADD PARAGRAPH ABOUT WHAT WE CAN CONCLUDE FROM THESE GRAPHS.
 
@@
 To demonstrate this capability, we examine the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [@knaus1995support].
-The SUPPORT dataset tracks death for individuals who are considered seriously ill within 5 hospitals in the United States of America.
-Imputation was conducted using the \pkg{mice} package in \proglang{R} with default settings [@mice].
-Before imputation, there were a total of 9105 individuals and 47 variables.
-Of those variables, a few were removed before imputation.
-We removed three response variables: hospital charges, patient ratio of costs to charges and patient micro-costs.
-Next, we removed hospital death as it was directly informative of our event of interest, death.
-We also removed functional disability and income as they are ordinal covariates, because ordinal variables are treated as factors in our current model.
-Finally, we removed 8 covariates relating to the results of previous findings: we removed SUPPORT day 3 physiology score (sps), 	APACHE III day 3 physiology score (aps),	SUPPORT model 2-month survival estimate, SUPPORT model 6-month survival estimate, Physician's 2-month survival estimate for pt., Physician's 6-month survival estimate for pt., Patient had Do Not Resuscitate (DNR) order and Day of DNR order (<0 if before study).
-Of these, sps and aps were added on after imputation, as they were missing only 1 observation and permitted us to compare findings between our model and their's.
-
-First we impute manually using recommended normal values for physiological measures [@harrell_2020].
-Next, we impute a single dataset using \pkg{mice} with default settings.
-After imputation, we noted that the covariate for surrogate activities of daily living was not imputed.
-This is due to colinearity between the other two covariates for activities of daily living.
-As such, surrogate activites of daily living was removed from our model.
-After these imputations, we have a total of 9104 observations and 30 variables, two of which are sps and aps, which will work as a baseline for our comparison to previous findings.
-This dataset is available  Its worth noting that these scores were developed for a period of 180 days, which makes our comparison over 5.56 years exploratory.
-Also, our single imputation introduces bias in the data.
-As the goal here is purely demonstrative, this bias will be consistent with any of the following models.
-As such, their comparison is still valid, though any interpretation requires an awareness of said bias.
-<!--The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.--> A description of used variables can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}.
-The original data was obtained from the Department of Biostatistics at Vanderbilt University [@harrell_2020].
+The SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital.
+Imputation was conducted using predictive mean matching and using the \pkg{mice} package in \proglang{R} [@mice].
+Before imputation, there was a total of 1000 individuals and 35 variables.
+After imputation, we have 474 individuals.
+\textcolor{red}{How come we lose individuals during imputation?} Of these, 67.5% experienced the event of interest.
+<!--The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.--> A description of each variable can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}.
+The original data was obtained from the website of the Department of Biostatistics at Vanderbilt University.
 The imputed dataset is available as part of the \pkg{casebase} package.
@@
 sex           &                                          & 2               & integer          & 0            \\
+hospdead      & Death in Hospital                        &                 & double           & 0            \\
 slos          & Days from Study Entry to Discharge       &                 & double           & 0            \\
@@
 scoma         & SUPPORT Coma Score based on Glasgow D3   &                 & double           & 0            \\
+charges       & Hospital Charges                         &                 & double           & 25           \\
+totcst        & Total RCC cost                           &                 & double           & 105          \\
+totmcst       & Total micro-cost                         &                 & double           & 372          \\
 avtisst       & Average TISS, Days 3-25                  &                 & double           & 6            \\
 race          &                                          & 5               & integer          & 5            \\
-sps 	        & support physiology score day 3           &	               & double           &	1            \\
-aps           & APS III no coma, imp bun,uout for ph1,D3 &	               & double	          & 1            \\
-hday	        & Day in Hospital at Study Admit	         &                 & double           & 0            \\	
-diabetes      &	Diabetes (Com 27-28, Dx 73)	             &                 & double	          & 0            \\
-dementia      & Dementia (Comorbidity 6)                 &	               & double	          & 0            \\
-ca		        & Cancer State                             & 3               & integer	        & 0            \\
 meanbp        & Mean Arterial Blood Pressure Day 3       &                 & double           & 0            \\
@@
 urine         & Urine Output Day 3                       &                 & double           & 517          \\
-adlp          & Activiites of Daily Living (ADL) Patient Day 3 &           & double           & 634          \\
+adlp          & ADL Patient Day 3                        &                 & double           & 634          \\
+adls          & ADL Surrogate Day 3                      &                 & double           & 310          \\
+sfdm2         &                                          & 5               & integer          & 159          \\
 adlsc         & Imputed ADL Calibrated to Surrogate      &                 & double           & 0           
@@
          & Cancer                                        \\
+income   & under \$11k                                   \\
+         & $11-$25k                                      \\
+         & $25-$50k                                      \\
          & \textgreater{}\$50k                           \\
@@
          & hispanic                                      \\
-ca       & Yes                                           \\
-         & No                                            \\
-         & Metastatic                                    \\
+sfdm2    & no(M2 and SIP pres)                           \\
+         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
+         & SIP\textgreater{}=30                          \\
+         & Coma or Intub                                 \\
+         & \textless{}2 mo. follow-up                   
 \end{tabular}
@@
 The population-time plot for the SUPPORT data appears in Figure \ref{fig:support-poptime}.
-As we can see, the incidence density is not constant.
+\textcolor{red}{This population-time plot is strange.
+The red dots never seem to go above 200.}
 
 ```{r support-poptime, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data."}
-
-completeData<-casebase::support
-workingCompleteData=model.matrix(death~ .-d.time,data=completeData)[,-c(1)]#remove intercept
-#Create u and xCox, which will be used when fitting Cox with glmnet.
-newData=workingCompleteData
-x=workingCompleteData
-#x and y will be used to fit the casebase model under glmnet.
-#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
-y=data.matrix(completeData[,c(2,5)])
-
+data(support)
 support_popTime <- popTime(support,
-                           time = "d.time", event = "death",percentile_number = 0.31)
+                           time = "d.time", event = "death")
 plot(support_popTime)
@@
 
-Our next step is to get the hazard function.
-All covariates will be used to estimate the hazard function.
-The hazard function will be used to estimate the cumulative incidence of each individual in the study.
-Then, the average cumulative incidience curve for all observations will be used for our comparison.
-
-Our main objective is to compute the absolute risk of death for a given set of covariates, using regularization when fitting our model.
-First, we fit a smooth hazard to the data; for the sake of this example, we opted for a linear term for time.
-As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model.
+Before we proceed with the analysis, we first remove hospital death as a covariate, as this is directly informative of death and cannot be used as a predictor.
+Then we select one observation at random.
+We will estimate the cumulative incidence curve for this individual.
+The other observations will be used to fit the hazard function.
+
+```{r supportSetUp, echo = TRUE, eval = eval_cs3}
+support <- dplyr::select(support, -hospdead)
+
+# Randomly select one observation
+indiv <- sample(1:nrow(support), size = 1)
+```
+
+Our main objective is to estimate the cumulative incidence of death for a given covariate profile.
+Given the number of covariates available, we will do this using regularized estimation.
+As \pkg{casebase} makes use of the \pkg{glmnet} package, we will use the function \code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}.
+Specifically, we input a matrix \code{y}, which contains the time and event variables, and a matrix \code{x}, which contains all other variables we would like to include in the model.
 For the SUPPORT dataset, all factor variables were converted into dummy variables.
-Here, we used lasso penalization by setting alpha to 1.
+
+```{r support-design, eval = eval_cs3}
+# Create design matrix without the time variable
+support_matrix <- model.matrix(death ~ . - d.time, 
+                               data = support)
+# Remove intercept term
+support_matrix <- support_matrix[,colnames(support_matrix) != "(Intercept)"]
+
+# Create test and train dataset
+support_test <- support_matrix[indiv,,drop = FALSE]
+
+support_train <- support_matrix[-indiv,]
+y <- as.matrix(support[-indiv, c("death", "d.time")])
+```
+
+We are now ready to estimate the hazard function.
+We use lasso penalization; this is achieved by setting \code{alpha} to 1.
+Also, note that the argument \code{formula_time} can be used to specify the function form for time; in our example, we model time linearly.
+
+```{r CBhazard, echo = TRUE, eval = eval_cs3}
+support_cb <- fitSmoothHazard.fit(support_train, y, 
+                                  formula_time = ~ d.time, 
+                                  family = "glmnet", 
+                                  event = "death", time = "d.time", 
+                                  alpha = 1, ratio = 100)
+```
+
+Then, we take our model object and use the \code{absoluteRisk} function to compute our estimate of the cumulative incidence function.
+Since \code{glmnet} fitted the penalized logistic regression for a range of penalty terms, we need to specify which one we want to use for our prediction.
+We select \code{s = "lambda.1se"}, as recommended.
+
+```{r CBabsolute, echo = TRUE, eval = eval_cs3}
+time_points <- seq(1, max(support$d.time), by = 0.5)
+support_risk <- absoluteRisk(support_cb,
+                             time = time_points,
+                             newdata = support_test,
+                             s = "lambda.1se", method = "numerical")
+```
+
+We will compare this curve to a regularized version of Cox regression, and a Kaplan-Meier survival curve.
+Using the \pkg{glmnet} package, we can fit the Cox regression model using regularized estimation.
+
+```{r coxHaz, echo = TRUE, eval = eval_cs3}
+# Create survival object for glmnet
+surv_glmnet <- survival::Surv(time = y[, "d.time"], 
+                              event = y[, "death"])
+
+support_cox <- glmnet::cv.glmnet(x = support_train, 
+                                 y = surv_glmnet,
+                                 family = "cox", alpha = 1)
+```
+
+However, \pkg{glmnet} does not provide any functionality for estimating the baseline hazard, and the functionalities in the \pkg{survival} package do not work on the output from \code{cv.glmnet}.
+To circumvent this issue and get an estimate of the baseline hazard, we followed the following steps.
+First, we fitted a non-regularized Cox model using only the variables selected by \pkg{glmnet}; the baseline hazard will be estimated using this object.
+Then, we replaced the estimates of the regression coefficients by the estimates from \pkg{glmnet}.
+Finally, this modified \code{coxph} object is fed to the \code{survival::survfit} function to obtain the estimated survival function.
+
+```{r coxAbsolute, echo = TRUE, eval = eval_cs3}
+# Identifying selected covariates
+nonzero_covariates_cox <- predict(support_cox, type = "nonzero", 
+                                  s = "lambda.1se")
+nonzero_coef_cox <- coef(support_cox, s = "lambda.1se")
+# Creating a new dataset that only contains the covariates selected by glmnet
+support_cox_data <- as.data.frame(cbind(y,
+                                        support_train[, nonzero_covariates_cox$X1]))
+# Fitting a cox model
+support_cox <- survival::coxph(Surv(time = d.time, event = death) ~ ., 
+                          data = support_cox_data)
+# Replace the estimate with those obtained by glmnet
+support_cox$coefficients <- nonzero_coef_cox@x
+
+# Fitting CI curve for cox+glmnet
+support_cox_risk <- survival::survfit(support_cox,
+                                      newdata = as.data.frame(support_test),
+                                      time = time_points,
+                                      type = "breslow")
+```
+
+We used the \pkg{survival} package to calculate the Kaplan-Meier estimate of the survival function.
+
+```{r fitKM+abRisk, echo = TRUE, eval = eval_cs3}
+km <- survival::Surv(time = support$d.time, 
+                     event = support$death)
+abKm <- survival::survfit(km ~ 1,
+                          type = 'kaplan-meier')
+```
+
+Now that our three cumulative incidence functions have been calculated, we compare them all in Figure \ref{fig:abSC}.
+Both \pkg{glmnet} and \pkg{casebase} have a lower estimated cumulative incidence by the end of the study, when compared to Kaplan-Meier.
+\textcolor{red}{It would be good to have an idea of why the case-base and Coxnet curves are so different.
+At the very least, I think we should compare the select covariates.}
+
+```{r abSupportComparison, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Comparison of cumulative incidence curves between regularized casebase in black, regularized Cox regression in red and Kaplan-Meier in blue."}
+plot(support_risk, type='l', col = "black", lwd = 2, 
+     main = "Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves", 
+     xlab = "Survival-Time", ylab = "Cumulative Incidence", ylim = c(0,1))
+lines(support_cox_risk, col = "red", fun = "event", lwd = 3, conf.int = FALSE)
+lines(abKm, col = "blue", fun = "event", lwd = 3, conf.int = FALSE)
+legend("bottom", 
+       legend = c("casebase", "Cox+glmnet", "KM curve"), 
+       col = c("black", "red", "blue"), lty = c(1, 1, 1), bg = "gray90")
+```
 
@@
 In this next case study, we will use the Stanford Heart Transplant data [@clark1971cardiac,@crowley1977covariance] to show how case-base sampling can also be used in the context of time-varying covariates.
-As an example that already appeared in the literature, case-base sampling was used to study vaccination safety, where the exposure period was defined as the week following vaccination [@saarela2015case].
+As an example that already appeared in the literature, case-base sampling was to study vaccination safety, where the exposure period was defined as the week following vaccination [@saarela2015case].
 Hence, the main covariate of interest, i.e.
@@
 For example, if we look at the Weibull distribution as parametrised in \code{stats::pweibull}, the log-hazard function is given by
-$$ \log \lambda(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,$$
+$$ \log h(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,$$
 where $\alpha,\beta$ are shape and scale parameters, respectively.

### 2020-03-20 analysis/paper/paper.tex
still working on poptime section

@@
 
-To describe the theoretical foundations of case-base sampling, we use
-the framework of counting processes. In what follows, we abuse notation
-slightly and omit any mention of \(\sigma\)-algebras; the interested
-reader can refer to Saarela \& Arjas \citeyearpar{saarela2015non} and
-Saarela \citeyearpar{saarela2016case}. First, let
-\(N_{i}(t) \in \{0, 1\}\) be counting processes corresponding to the
-event of interest for individual \(i=1, \ldots,n\). For simplicity, we
-will consider Type I censoring due to the end of follow-up at time
-\(\tau\) (the general case of non-informative censoring is treated in
-Saarela \citeyearpar{saarela2016case}). We assume a continuous time
-model, which implies that the counting process jumps are less than or
-equal to one. We are interested in modeling the hazard functions
-\(\lambda_{i}(t)\) of the processes \(N_i(t)\), and which satisfy
-\[\lambda_{i}(t) dt = E[dN_{i}(t)].\]
-
-Next, we model the base series sampling mechanism using non-homogeneous
-Poisson processes \(R_i(t) \in \{0, 1, 2, \ldots\}\), with the
-person-moments where \(dR_i(t) = 1\) constituting the base series. The
-process \(Q_{i}(t) = R_i(t) + N_{i}(t)\) then counts both the case and
-base series person-moments contributed by individual \(i\). This process
-is typically defined by the user via its hazard function \(\rho_i(t)\).
-The process \(Q_{i}(t)\) is characterized by
-\(E[dQ_{i}(t)] = \lambda_{i}(t)dt + \rho_i(t)dt\).
-
-If the hazard function \(\lambda_{i}(t; \theta)\) is parametrized in
-terms of \(\theta\), we could define an estimator \(\hat{\theta}\) by
-maximization of the likelihood expression
-\[L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^\tau \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)}\left(1 - \lambda_{i}(t;\theta)\right)^{1 - dN_{i}(t)},\]
-where \(\prod_{t\in[0,u)}\) represents a product integral from \(0\) to
-\(u\). However, the integral over time makes the computation and
-maximisation of \(L_0(\theta)\) challenging.
-
-Case-base sampling allows us to avoid this integral. By conditioning on
-a sampled person-moment, we get individual contributions of the form
-\[P(dN_{i}(t) \mid dQ_{i}(t) = 1) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.\]
-Therefore, we can define an estimating equation for \(\theta\) as
-follows:
-\[L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.\]
-When a logarithmic link function is used for modeling the hazard
-function, the above expression is of a logistic regression form with an
-offset term \(\log(1/\rho_i(t))\). Note that the sampling units selected
-in the case-base sampling mechanism are person-moments, rather than
-individuals, and the parameters to be estimated are hazards or hazard
-ratios rather than odds or odds ratios. Generally, an individual can
-contribute more than one person-moment, and thus the terms in the
-product integral are not independent. Nonetheless, Saarela
-\citeyearpar{saarela2016case} showed that the logarithm of this
-estimating function has mean zero at the true value \(\theta=\theta_0\),
-and that the resulting estimator \(\hat{\theta}\) is asymptotically
-normally distributed.
+As shown by Saarela \& Arjas \citeyearpar{saarela2015non} (and further
+expanded in Saarela \citeyearpar{saarela2016case}), writing the
+likelihood arising from this data-generating mechanism using the
+framework of non-homogeneous Poisson processes, we eventually reach an
+expression where each person-moment's contribution is of the form
+\[\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},\] where \(N(t)\) is the counting
+process associated with the event of interest, \(h(t)\) is the
+corresponding hazard function, and \(\rho(t)\) is the hazard function
+for the Poisson process associated with case-base sampling. This
+parametric form suggests that we can readily estimate log-hazards of the
+form \(\log(h(t)) = g(t; X)\) using logistic regression, where each
+observation corresponds to a person moment, the function \(g(t; X)\) is
+linear in a finite number of parameters, and where we treat
+\(-\log(\rho(t))\) as an offset.
 
@@
 In terms of Poisson processes, this sampling strategy corresponds
-essentially to a time-homogeneous Poisson process with hazard equal to
-\(\rho_i(t) = b/B\), where \(b\) is the number of sampled observations
-in the base series, and \(B\) is the total population-time for the study
-base. More complex examples are also available; see for example Saarela
-\& Arjas \citeyearpar{saarela2015non}, where the probabilities of the
+essentially to a time-homogeneous Poisson process with intensity equal
+to \(b/B\), where \(b\) is the number of sampled observations in the
+base series, and \(B\) is the total population-time for the study base.
+More complex examples are also available; see for example Saarela \&
+Arjas \citeyearpar{saarela2015non}, where the probabilities of the
 sampling mechanism are proportional to the cardiovascular disease event
@@
 Let \(g(t; X)\) be the linear predictor such that
-\(\log(\lambda(t;X)) = g(t; X)\). Different functions of \(t\) lead to
+\(\log(h(t;X)) = g(t; X)\). Different functions of \(t\) lead to
 different parametric hazard models. The simplest of these models is the
@@
 
-\[ \log(\lambda(t; X)) = \beta_0 + \beta_1 X.\]
+\[ \log(h(t; X)) = \beta_0 + \beta_1 X.\]
 
@@
 
-\[ \log(\lambda(t; X)) = \beta_0 + \beta_1 t + \beta_2 X.\]
+\[ \log(h(t; X)) = \beta_0 + \beta_1 t + \beta_2 X.\]
 
@@
 
-\[ \log(\lambda(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. \]
+\[ \log(h(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. \]
 
@@
 
-\[\frac{\lambda_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},\]
+\[\frac{h_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^Jh_j(t)},\]
 
 where \(N_j(t)\) is the counting process associated with the event of
-type \(j\) and \(\lambda_j(t)\) is the corresponding hazard function. As
-may be expected, this functional form is similar to the terms appearing
-in the likelihood function for multinomial
+type \(j\) and \(h_j(t)\) is the corresponding hazard function. As may
+be expected, this functional form is similar to the terms appearing in
+the likelihood function for multinomial
 regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}
@@
 
-Several \proglang{R} packages implement methodologies for competing risk
-analysis; for a different perspective, see Mahani \& Sharabiani
-\citeyearpar{mahani2015bayesian}. The package \pkg{cmprsk} provides
-methods for cause-specific subdistributions, such as in the Fine-Gray
-model \citeyearpar{fine1999proportional}. On the other hand, the package
-\pkg{CFC} estimates cause-specific cumulative incidence functions from
-unadjusted, non-parametric survival functions. Our package
-\pkg{casebase} also provides functionalities for competing risk analysis
-by estimating the cause-specific hazards, and from these quantities, we
-can derive the cause-specific cumulative incidence functions.
+Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
+\pkg{mets} and \pkg{survival} contain implementations for competing
+risks. The differentiating factor between these packages and
+\pkg{casebase} is that \pkg{casebase} handles competing risks through
+multiple logistic regression.
+\textcolor{red}{There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}.
+Could you incorporate some of their discussion, with proper citations?
+The biggest omission right now is \pkg{cmprsk}, which we actually use below.}
 
@@
 
-\hypertarget{parametric-modeling}{%
-\subsection{Parametric modeling}\label{parametric-modeling}}
-
-The parametric modeling step was separated into three parts:
+The case-base sampling approach described in Section \ref{theory} can be
+visualized in the form of a population time plot. They are extremely
+informative graphical displays of survival data and should be one of the
+first steps in an exploratory data analysis. We facilitate this task in
+the \pkg{casebase} package using the \code{popTime} function and
+\code{plot} method as follows:
 
@@
 \item
+  The \code{casebase::popTime} function takes as input the original
+  dataset along with the column names corresponding to the timescale,
+  the event status and an exposure group of interest (optional). This
+  will create an object of class \code{popTime} (or
+  \code{popTimeExposure} if you specify a value for the exposure
+  argument).\\
+\item
+  The corresponding \code{plot} methods for the object created in Step 1
+  can be called to create the population time plot with several options
+  for customizing the aesthetics.
+\end{enumerate}
+
+To illustrate these functions we will use the European Randomized Study
+of Prostate Cancer Screening data which is available through the
+\pkg{casebase} package. The results of this study were published by
+\citep{schroder2009screening}, and the dataset itself was obtained using
+the approach described in \citep{liu2014recovering}.
+
+We first create the necessary dataset for producing the population time
+plots using the \code{casebase::popTime} function. In this example we
+stratify by the treatment group. Note that the resulting object stores
+the exposure variable as an attribute:
+
+\begin{CodeChunk}
+
+\begin{CodeOutput}
+#> [1] "ScrArm"
+\end{CodeOutput}
+\end{CodeChunk}
+
+We then pass this object to the \code{plot} function which leverages the
+\pkg{ggplot2} package \citep{ggplot2} to build population time plots. It
+builds the plot by sequentially adding layers, starting with a layer for
+the area representing the population time followed by points to show the
+casebase sampling mechanism. This function gives user the flexibility to
+add any combination of the case.series, base.series and competing
+events. The case series and competing events are sampled at random
+vertically on the plot in order to visualise the incidence density using
+the popTime function. That is, imagine we draw a vertical line at a
+specific event time. We then plot the point at a randomly sampled
+y-coordinate along this vertical line. This is done to avoid having all
+points along the upper edge of the plot (because the subjects with the
+least amount of observation time are plotted at the top of the y-axis).
+By randomly distributing them, we can get a better sense of the
+inicidence density. The base series is sampled horizontally on the plot
+using the sampleCaseBase function.
+
+One benefit from these plots is that it allows you to see the incidence
+density. This can be seen from the distribution of the red dots in the
+above plot. We can see that more events are observed later on in time.
+Therefore a constant hazard model would not be appropriate in this
+instance as it would overestimate the cumulative incidence earlier on in
+time, and underestimate it later on.
+
+The unique \emph{step shape} of the population time plot is due to the
+randomization of the Finnish cohorts which were carried out on January 1
+of each of the 4 years 1996 to 1999. This, coupled with the uniform
+December 31 2006 censoring date, lead to large numbers of men with
+exactly 11, 10, 9 or 8 years of follow-up.
+
+It is important to note that the red points are random distributed
+across the grey area for each time of event. That is, imagine we draw a
+vertical line at a specific event time. We then plot the red point at a
+randomly sampled y-coordinate along this vertical line. This is done to
+avoid having all red points along the upper edge of the plot (because
+the subjects with the least amount of observation time are plotted at
+the top of the y-axis). By randomly distributing them, we can get a
+better sense of the inicidence density.
+
+We can also create exposure stratified plots by specifying the
+\code{exposure} argument in the \code{popTime} function:
+
+We can also plot them side-by-side using the \code{ncol} argument:
+
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-side-1} \end{center}
+
+\end{CodeChunk}
+
+ADD PARAGRAPH ABOUT WHAT WE CAN CONCLUDE FROM THESE GRAPHS.
+
+For a more detailed description of how to modify the aesthetics of a
+population time plot, please refer to the
+\href{http://sahirbhatnagar.com/casebase/articles/customizingpopTime.html}{Customizing
+Population Time Plots} vignette.
+
+\hypertarget{parametric-modeling}{%
+\subsection{Parametric modeling}\label{parametric-modeling}}
+
+The parametric modeling step was separated into three parts:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
   case-base sampling;
@@
 The sampling scheme selected for \code{sampleCaseBase} was described in
-Hanley \& Miettinen \citeyearpar{hanley2009fitting}: we first sample
+Hanley and Miettinen \citeyearpar{hanley2009fitting}: we first sample
 along the ``person'' axis, proportional to each individual's total
@@
 fitting function starts by calling \code{sampleCaseBase} to generate the
-base series. In other words, users can bypass \code{sampleCaseBase}
-altogether and only worry about the fitting function
-\code{fitSmoothHazard}.
+base series. In other words, the occasional user can bypass
+\code{sampleCaseBase} altogether and only worry about the fitting
+function \code{fitSmoothHazard}.
 
@@
 for the time variable. Note that non-proportional hazard models can be
-achieved at this stage by adding an interaction term involving time
-(cf.~Case Study 4 below). The offset term does not need to be specified
-by the user, as it is automatically added to the formula before calling
-\code{glm}.
+achieved at this stage by adding an interaction term involving time. The
+offset term does not need to be specified by the user, as it is
+automatically added to the formula.
 
@@
 In the case of multiple events, the hazard is fitted via multinomial
-regression as performed by the \pkg{VGAM} package. We selected this
-package for its ability to fit multinomial regression models with an
+regression as performed by the \pkg{VGAM} package. This package was
+selected for its ability to fit multinomial regression models with an
 offset.
@@
 identity \begin{equation}\label{eqn:surv}
-S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right).
+S(t) = \exp\left(-\int_0^t h(u;X) du\right).
 \end{equation} The integral is computed using either the
@@
 
-\[ S(t) = \exp\left(-\int_0^t \Lambda(u;X) du\right),\qquad \Lambda(t;X) = \sum_{j=1}^J \lambda_j(t;X).\]
+\[ S(t) = \exp\left(-\int_0^t H(u;X) du\right),\qquad H(t;X) = \sum_{j=1}^J h_j(t;X).\]
 From this, we can derive the event-specific subdensities:
 
-\[ f_j(t) = \lambda_j(t)S(t).\]
+\[ f_j(t) = h_j(t)S(t).\]
 
@@
 
-\[ CI_j(t) = \int_0^t f_j(u)du.\] The integrals are computed using
-either numerical integration (via the trapezoidal rule) or Monte Carlo
-integration. This option is controlled by the argument \code{method} of
-the \code{absoluteRisk} function.
-
-We created \code{absoluteRisk} as an \code{S3} generic, with methods
-that match the different types of outputs of \code{fitSmoothHazard}. The
-method dispatch system of \proglang{R} then takes care of matching the
-correct output to the correct methodology for calculating the cumulative
-incidence function, without requiring any intervention from the user.
+\[ CI_j(t) = \int_0^t f_j(u)du.\]
+
+We created \code{absoluteRisk} as an \code{S3} generic, with methods for
+the different types of outputs of \code{fitSmoothHazard}. The method
+dispatch system of \proglang{R} then takes care of matching the correct
+output to the correct methodology for calculating the cumulative
+incidence function, without the user's intervention.
 
 In the following sections, we illustrate these functionalities in the
-context of four case studies.
+context of three case studies.
 
@@
 Screening}\label{case-study-1european-randomized-study-of-prostate-cancer-screening}}
-
-For the first case study, we will use of the European Randomized Study
-of Prostate Cancer Screening data. This dataset is available through the
-\pkg{casebase} package:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> data(ERSPC)
-R> ERSPC$ScrArm <- factor(ERSPC$ScrArm,
-R>   levels = c(0, 1),
-R>   labels = c("Control group", "Screening group")
-R> )
-\end{CodeInput}
-\end{CodeChunk}
-
-The results of this study were published by
-\citep{schroder2009screening}, and the dataset itself was obtained using
-the approach described in \citep{liu2014recovering}.
-
-Population time plots can be extremely informative graphical displays of
-survival data. They should be the first step in an exploratory data
-analysis. We facilitate this task in the \pkg{casebase} package using
-the \code{popTime} function. We first create the necessary dataset for
-producing the population time plots, and we can generate the plot by
-using the corresponding \code{plot} method:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
-R> plot(pt_object)
-\end{CodeInput}
-\end{CodeChunk}
-
-We can also create exposure stratified plots by specifying the
-\code{exposure} argument in the \code{popTime} function:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> pt_object_strat <- casebase::popTime(ERSPC,
-R>   event = "DeadOfPrCa",
-R>   exposure = "ScrArm"
-R> )
-R> plot(pt_object_strat)
-\end{CodeInput}
-\end{CodeChunk}
-
-We can also plot them side-by-side using the \code{ncol} argument:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> plot(pt_object_strat, ncol = 2)
-\end{CodeInput}
-\end{CodeChunk}
-
-ADD PARAGRAPH ABOUT WHAT WE CAN CONCLUDE FROM THESE GRAPHS.
 
@@
 
-\begin{CodeInput}
-R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm,
-R>   data = ERSPC
-R> )
-R> summary(cox_model)
-\end{CodeInput}
+\begin{CodeOutput}
+#> Call:
+#> survival::coxph(formula = Surv(Follow.Up.Time, DeadOfPrCa) ~ 
+#>     ScrArm, data = ERSPC)
+#> 
+#>   n= 159893, number of events= 540 
+#> 
+#>                         coef exp(coef) se(coef)     z Pr(>|z|)  
+#> ScrArmScreening group -0.222     0.801    0.088 -2.52    0.012 *
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#>                       exp(coef) exp(-coef) lower .95 upper .95
+#> ScrArmScreening group     0.801       1.25     0.674     0.952
+#> 
+#> Concordance= 0.519  (se = 0.011 )
+#> Likelihood ratio test= 6.45  on 1 df,   p=0.01
+#> Wald test            = 6.37  on 1 df,   p=0.01
+#> Score (logrank) test = 6.39  on 1 df,   p=0.01
+\end{CodeOutput}
 \end{CodeChunk}
@@
 \begin{CodeChunk}
-
-\begin{CodeInput}
-R> new_data <- data.frame(
-R>   ScrArm = c("Control group", "Screening group"),
-R>   ignore = 99
-R> )
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R>   xlab = "Years since Randomization",
-R>   ylab = "Cumulative Incidence (%)",
-R>   fun = "event",
-R>   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
-R>   main = sprintf(
-R>     "Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>     exp(coef(cox_model)),
-R>     exp(confint(cox_model))[1],
-R>     exp(confint(cox_model))[2]
-R>   ),
-R>   ylim = c(0, 2), yscale = 100
-R> )
-R> legend("topleft",
-R>   legend = c("Control group", "Screening group"),
-R>   col = c("red", "blue"),
-R>   lty = c(1, 1),
-R>   bg = "gray90"
-R> )
-\end{CodeInput}
+\begin{figure}
+
+{\centering \includegraphics{../figures/erspc-cox-cif-1}
+
+}
+
+\caption[Cumulative incidence functions for control and screening groups in the ERSPC data]{Cumulative incidence functions for control and screening groups in the ERSPC data.
+In each of the panels, we plot the CIF from the Cox model (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression.
+(A) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution.
+(B) Linear function of time which is equivalent to a Gompertz distribution.
+(C) The natural logarithm of time which is equivalent to a Weibull distribution.
+(D) Cubic b-spline expansion of time.}\label{fig:erspc-cox-cif}
+\end{figure}
 \end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{../figures/unnamed-chunk-1-1} \end{center}
+
+\end{CodeChunk}
+
+\hypertarget{equivalence-between-casebase-and-aft}{%
+\subsection{Equivalence between casebase and
+AFT}\label{equivalence-between-casebase-and-aft}}
 
@@
 excluding time from the linear predictor.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm,
-R>   data = ERSPC,
-R>   ratio = 100
-R> )
-R> 
-R> summary(casebase_exponential)
-R> exp(coef(casebase_exponential)[2])
-R> exp(confint(casebase_exponential)[2, ])
-\end{CodeInput}
-\end{CodeChunk}
 
@@
 the Cox model CIF:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> smooth_risk_exp <- casebase::absoluteRisk(
-R>   object = casebase_exponential,
-R>   time = seq(0, 15, 0.1),
-R>   newdata = new_data
-R> )
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R>   xlab = "Years since Randomization",
-R>   ylab = "Cumulative Incidence (%)",
-R>   fun = "event",
-R>   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
-R>   main = sprintf(
-R>     "Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>     exp(coef(cox_model)),
-R>     exp(confint(cox_model))[1],
-R>     exp(confint(cox_model))[2]
-R>   ),
-R>   ylim = c(0, 2), yscale = 100
-R> )
-R> lines(smooth_risk_exp[, 1], smooth_risk_exp[, 2], col = "red", lty = 2)
-R> lines(smooth_risk_exp[, 1], smooth_risk_exp[, 3], col = "blue", lty = 2)
-R> 
-R> 
-R> legend("topleft",
-R>   legend = c(
-R>     "Control group (Cox)", "Control group (Casebase)",
-R>     "Screening group (Cox)", "Screening group (Casebase)"
-R>   ),
-R>   col = c("red", "red", "blue", "blue"),
-R>   lty = c(1, 2, 1, 2),
-R>   bg = "gray90"
-R> )
-\end{CodeInput}
-\end{CodeChunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm,
-R>   data = ERSPC,
-R>   ratio = 100
-R> )
-R> 
-R> summary(casebase_time)
-R> exp(coef(casebase_time))
-R> exp(confint(casebase_time))
-\end{CodeInput}
-\end{CodeChunk}
-
 Again, the estimate of the hazard ratio is similar to that obtained from
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> smooth_risk_time <- casebase::absoluteRisk(
-R>   object = casebase_time,
-R>   time = seq(0, 15, 0.1),
-R>   newdata = new_data
-R> )
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R>   xlab = "Years since Randomization",
-R>   ylab = "Cumulative Incidence (%)",
-R>   fun = "event",
-R>   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
-R>   main = sprintf(
-R>     "Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>     exp(coef(cox_model)),
-R>     exp(confint(cox_model))[1],
-R>     exp(confint(cox_model))[2]
-R>   ),
-R>   ylim = c(0, 2), yscale = 100
-R> )
-R> lines(smooth_risk_time[, 1], smooth_risk_time[, 2], col = "red", lty = 2)
-R> lines(smooth_risk_time[, 1], smooth_risk_time[, 3], col = "blue", lty = 2)
-R> 
-R> legend("topleft",
-R>   legend = c(
-R>     "Control group (Cox)", "Control group (Casebase)",
-R>     "Screening group (Cox)", "Screening group (Casebase)"
-R>   ),
-R>   col = c("red", "red", "blue", "blue"),
-R>   lty = c(1, 2, 1, 2),
-R>   bg = "gray90"
-R> )
-\end{CodeInput}
-\end{CodeChunk}
-
 We see that the Weibull model leads to a better fit of the CIF than the
@@
 \code{splines} package:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm,
-R>   data = ERSPC,
-R>   ratio = 100
-R> )
-R> 
-R> summary(casebase_splines)
-R> exp(coef(casebase_splines))
-R> exp(confint(casebase_splines))
-\end{CodeInput}
-\end{CodeChunk}
 
@@
 CIF:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> smooth_risk_splines <- absoluteRisk(
-R>   object = casebase_splines,
-R>   time = seq(0, 15, 0.1),
-R>   newdata = new_data
-R> )
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R>   xlab = "Years since Randomization",
-R>   ylab = "Cumulative Incidence (%)",
-R>   fun = "event",
-R>   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
-R>   main = sprintf(
-R>     "Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>     exp(coef(cox_model)),
-R>     exp(confint(cox_model))[1],
-R>     exp(confint(cox_model))[2]
-R>   ),
-R>   ylim = c(0, 2), yscale = 100
-R> )
-R> lines(smooth_risk_splines[, 1], smooth_risk_splines[, 2], col = "red", lty = 2)
-R> lines(smooth_risk_splines[, 1], smooth_risk_splines[, 3], col = "blue", lty = 2)
-R> 
-R> legend("topleft",
-R>   legend = c(
-R>     "Control group (Cox)", "Control group (Casebase)",
-R>     "Screening group (Cox)", "Screening group (Casebase)"
-R>   ),
-R>   col = c("red", "red", "blue", "blue"),
-R>   lty = c(1, 2, 1, 2),
-R>   bg = "gray90"
-R> )
-\end{CodeInput}
-\end{CodeChunk}
 
@@
 
+\begin{CodeChunk}
+\begin{table}
+
+\caption{\label{tab:unnamed-chunk-4}Mean }
+\centering
+\begin{tabular}[t]{>{\bfseries}lcc}
+\toprule
+Model & casebase & survreg\\
+\midrule
+Exponential & 0.81 (0.68, 0.96) & 0.81 (0.68, 0.96)\\
+\cmidrule{1-3}
+Gompertz & 0.79 (0.66, 0.94) & 0.80 (0.67, 0.95)\\
+\cmidrule{1-3}
+Weibull & 0.80 (0.67, 0.95) & 0.80 (0.65, 0.96)\\
+\cmidrule{1-3}
+Splines & 0.80 (0.67, 0.95) & --\\
+\bottomrule
+\multicolumn{3}{l}{\textit{Note:}}\\
+\multicolumn{3}{l}{Median (Inter-quartile range) is given for Model Size.}\\
+\end{tabular}
+\end{table}
+
+\end{CodeChunk}
+
 As noted above, the usual asymptotic results hold for likelihood ratio
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> anova(casebase_exponential, casebase_splines, test = "LRT")
-\end{CodeInput}
-\end{CodeChunk}
-
-As expected, we see that splines model provides a better fit. Similarly,
-we can compare the AIC for all three case-base models:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> c(
-R>   "Exp." = AIC(casebase_exponential),
-R>   "Weib." = AIC(casebase_time),
-R>   "Splines" = AIC(casebase_splines)
-R> )
-\end{CodeInput}
-\end{CodeChunk}
-
-Once again, we have evidence that the splines model provides the best
-fit.
-
-With this first case study, we saw how population-time plots can provide
-a useful summary of the incidence for our event of interests. In
-particular, we saw how it can provide evidence against an exponential
-hazard model. We then explored how different parametric models can
-provide different estimates of the CIF, even though they all give
-similar estimates for the hazard ratios.
-
-\hypertarget{case-study-2bone-marrow-transplant}{%
-\section{Case study 2--Bone-marrow
-transplant}\label{case-study-2bone-marrow-transplant}}
-
-In the next case study, we show how case-base sampling can be used in
-the context of a competing risk analysis. For illustrative purposes, we
-will use the same data that was used in Scrucca \emph{et al}
-\citeyearpar{scrucca2010regression}. The data was downloaded from the
-first author's website, and it is also available as part of the
-\pkg{casebase} package.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> 
-R> data(bmtcrr)
-\end{CodeInput}
-\end{CodeChunk}
-
-The data contains information on 177 patients who received a stem-cell
-transplant for acute leukemia. The event of interest is relapse, but
-other competing causes (e.g.~transplant-related death) were also
-recorded. Several covariates were captured at baseline: sex, disease
-type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL
-and AML, respectively), disease phase at transplant (Relapse, CR1, CR2,
-CR3), source of stem cells (bone marrow and peripheral blood, coded as
-BM+PB, or only peripheral blood, coded as PB), and age. A summary of
-these baseline characteristics appear in Table \ref{tab:table1bmtcrr}.
-We note that the statistical summaries were generated differently for
-different variable types: for continuous variables, we gave the range,
-followed by the mean and standard deviation; for categorical variables,
-we gave the counts for each category.
-
-First, we can look at a population-time plot to visualize the incidence
-density of both relapse and the competing events. In Figure
-\ref{fig:compPop}, failure times are highlighted on the plot using red
-dots for the event of interest (panel A) and blue dots for competing
-events (panel B). In both panels, we see evidence of a non-constant
-hazard function: the density of points is larger at the beginning of
-follow-up than at the end.
-
-Our main objective is to compute the absolute risk of relapse for a
-given set of covariates. We start by fitting a smooth hazard to the data
-using a linear term for time:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> model_cb <- fitSmoothHazard(
-R>   Status ~ ftime + Sex + D + Phase + Source + Age,
-R>   data = bmtcrr,
-R>   ratio = 100,
-R>   time = "ftime"
-R> )
-\end{CodeInput}
-\end{CodeChunk}
-
-We will compare our hazard ratio estimates to that obtained from a Cox
-regression. To do so, we need to treat the competing event as censoring.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(survival)
-R> # Treat competing event as censoring
-R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-R>   data = bmtcrr
-R> )
-\end{CodeInput}
-\end{CodeChunk}
-
-From the fit object, we can extract both the hazard ratios and their
-corresponding confidence intervals. These quantities appear in Table
-\ref{tab:bmtcrr-cis}. As we can see, the only significant hazard ratio
-identified by case-base sampling is the one associated with the phase of
-the disease at transplant. More precisely, being in relapse at
-transplant is associated with a hazard ratio of 3.89 when compared to
-CR1.
-
-Given the estimate of the hazard function obtained using case-base
-sampling, we can compute the absolute risk curve for a fixed covariate
-profile. We perform this computation for a 35 year old woman who
-received a stem-cell transplant from peripheral blood at relapse. We
-compared the absolute risk curve for such a woman with acute
-lymphoblastic leukemia with that for a similar woman with acute
-myeloblastic leukemia. We will estimate the curve from 0 to 60 months.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Pick 100 equidistant points between 0 and 60 months
-R> time_points <- seq(0, 60, length.out = 50)
-R> 
-R> # Data.frame containing risk profile
-R> newdata <- data.frame(
-R>   "Sex" = factor(c("F", "F"),
-R>     levels = levels(bmtcrr[, "Sex"])
-R>   ),
-R>   "D" = c("ALL", "AML"), # Both diseases
-R>   "Phase" = factor(c("Relapse", "Relapse"),
-R>     levels = levels(bmtcrr[, "Phase"])
-R>   ),
-R>   "Age" = c(35, 35),
-R>   "Source" = factor(c("PB", "PB"),
-R>     levels = levels(bmtcrr[, "Source"])
-R>   )
-R> )
-R> 
-R> # Estimate absolute risk curve
-R> risk_cb <- absoluteRisk(
-R>   object = model_cb, time = time_points,
-R>   method = "numerical", newdata = newdata
-R> )
-\end{CodeInput}
-\end{CodeChunk}
-
-We will compare our estimates to that obtained from a corresponding
-Fine-Gray model \citeyearpar{fine1999proportional}. The Fine-Gray model
-is a semiparametric model for the cause-specific \emph{subdistribution},
-i.e.~the function \(f_k(t)\) such that
-\[CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),\] where
-\(CI_k(t)\) is the cause-specific cumulative indicence. The Fine-Gray
-model allows to directly assess the effect of a covariate on the
-subdistribution, as opposed to the hazard. For the computation, we will
-use the \pkg{timereg} package \citep{timereg}:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(timereg)
-R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) +
-R>   const(Phase) + const(Source) + const(Age),
-R> data = bmtcrr, cause = 1, model = "fg"
-R> )
-R> 
-R> # Estimate absolute risk curve
-R> risk_fg <- predict(model_fg, newdata, times = time_points)
-\end{CodeInput}
-\end{CodeChunk}
-
-Figure \ref{fig:compAbsrisk} shows the absolute risk curves for both
-case-base sampling and the Fine-Gray model. As we can see, the two
-approaches agree quite well for acute myeloblastic leukemia; however,
-there seems to be a difference of about 5\% between the two curves for
-acute lymphoblastic leukemia. This difference does not appear to be
-significant: the curve from case-base sampling is contained within a
-95\% confidence band around the Fine-Gray absolute risk curve (figure
-not shown).
-
-\hypertarget{case-study-3support-data}{%
-\section{Case study 3--SUPPORT Data}\label{case-study-3support-data}}
-
-In the first two case studies, we described the basic functionalities of
-the \pkg{casebase} package: creating population-time plots, fitting
-parametric models for hazard functions, and estimating the corresponding
-cumulative incidence curves. In the next two case studies, we explore
-more advanced features.
-
-For the third case study, we will look at regularized estimation for the
-hazard function. This can be performed using a combination of case-base
-sampling and regularized logistic regression. To demonstrate this
-capability, we examine the Study to Understand Prognoses Preferences
-Outcomes and Risks of Treatment (SUPPORT) dataset
-\citep{knaus1995support}. The SUPPORT dataset tracks death for
-individuals who are considered seriously ill within 5 hospitals in the
-United States of America. Imputation was conducted using the \pkg{mice}
-package in \proglang{R} with default settings \citep{mice}. Before
-imputation, there were a total of 9105 individuals and 47 variables. Of
-those variables, a few were removed before imputation. We removed three
-response variables: hospital charges, patient ratio of costs to charges
-and patient micro-costs. Next, we removed hospital death as it was
-directly informative of our event of interest, death. We also removed
-functional disability and income as they are ordinal covariates, because
-ordinal variables are treated as factors in our current model. Finally,
-we removed 8 covariates relating to the results of previous findings: we
-removed SUPPORT day 3 physiology score (sps), APACHE III day 3
-physiology score (aps), SUPPORT model 2-month survival estimate, SUPPORT
-model 6-month survival estimate, Physician's 2-month survival estimate
-for pt., Physician's 6-month survival estimate for pt., Patient had Do
-Not Resuscitate (DNR) order and Day of DNR order (\textless{}0 if before
-study). Of these, sps and aps were added on after imputation, as they
-were missing only 1 observation and permitted us to compare findings
-between our model and their's.
-
-First we impute manually using recommended normal values for
-physiological measures \citep{harrell_2020}. Next, we impute a single
-dataset using \pkg{mice} with default settings. After imputation, we
-noted that the covariate for surrogate activities of daily living was
-not imputed. This is due to colinearity between the other two covariates
-for activities of daily living. As such, surrogate activites of daily
-living was removed from our model. After these imputations, we have a
-total of 9104 observations and 30 variables, two of which are sps and
-aps, which will work as a baseline for our comparison to previous
-findings. This dataset is available Its worth noting that these scores
-were developed for a period of 180 days, which makes our comparison over
-5.56 years exploratory. Also, our single imputation introduces bias in
-the data. As the goal here is purely demonstrative, this bias will be
-consistent with any of the following models. As such, their comparison
-is still valid, though any interpretation requires an awareness of said
-bias. A description of used variables can be found in Table
-\ref{tab:support1} and a breakdown of each categorical variable in Table
-\ref{tab:support2}. The original data was obtained from the Department
-of Biostatistics at Vanderbilt University \citep{harrell_2020}. The
-imputed dataset is available as part of the \pkg{casebase} package.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ccccc}
-\hline
-\textbf{Name} & \textbf{Labels}                          & \textbf{Levels} & \textbf{Storage} & \textbf{NAs} \\
-\hline
-age           & Age                                      &                 & double           & 0            \\
-death         & Death at any time up to NDI date:31DEC94 &                 & double           & 0            \\
-sex           &                                          & 2               & integer          & 0            \\
-slos          & Days from Study Entry to Discharge       &                 & double           & 0            \\
-d.time        & Days of Follow-Up                        &                 & double           & 0            \\
-dzgroup       &                                          & 8               & integer          & 0            \\
-dzclass       &                                          & 4               & integer          & 0            \\
-num.co        & number of comorbidities                  &                 & double           & 0            \\
-edu           & Years of Education                       &                 & double           & 202          \\
-income        &                                          & 4               & integer          & 349          \\
-scoma         & SUPPORT Coma Score based on Glasgow D3   &                 & double           & 0            \\
-avtisst       & Average TISS, Days 3-25                  &                 & double           & 6            \\
-race          &                                          & 5               & integer          & 5            \\
-sps             & support physiology score day 3           &                   & double           & 1            \\
-aps           & APS III no coma, imp bun,uout for ph1,D3 &                 & double           & 1            \\
-hday            & Day in Hospital at Study Admit             &                 & double           & 0            \\ 
-diabetes      & Diabetes (Com 27-28, Dx 73)              &                 & double           & 0            \\
-dementia      & Dementia (Comorbidity 6)                 &                 & double           & 0            \\
-ca              & Cancer State                             & 3               & integer          & 0            \\
-meanbp        & Mean Arterial Blood Pressure Day 3       &                 & double           & 0            \\
-wblc          & White Blood Cell Count Day 3             &                 & double           & 24           \\
-hrt           & Heart Rate Day 3                         &                 & double           & 0            \\
-resp          & Respiration Rate Day 3                   &                 & double           & 0            \\
-temp          & Temperature (Celsius) Day 3              &                 & double           & 0            \\
-pafi          & PaO2/(.01*FiO2) Day 3                    &                 & double           & 253          \\
-alb           & Serum Albumin Day 3                      &                 & double           & 378          \\
-bili          & Bilirubin Day 3                          &                 & double           & 297          \\
-crea          & Serum creatinine Day 3                   &                 & double           & 3            \\
-sod           & Serum sodium Day 3                       &                 & double           & 0            \\
-ph            & Serum pH (arterial) Day 3                &                 & double           & 250          \\
-glucose       & Glucose Day 3                            &                 & double           & 470          \\
-bun           & BUN Day 3                                &                 & double           & 455          \\
-urine         & Urine Output Day 3                       &                 & double           & 517          \\
-adlp          & Activiites of Daily Living (ADL) Patient Day 3 &           & double           & 634          \\
-adlsc         & Imputed ADL Calibrated to Surrogate      &                 & double           & 0           
-\end{tabular}
-\caption{A description of each variable in the SUPPORT dataset.}
-\label{tab:support1}
-\end{table}
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{cc}
-\hline
-\textbf{Variable} & \textbf{Levels}                      \\
-\hline
-sex      & female                                        \\
-         & male                                          \\
-dzgroup  & ARF/MOSF w/Sepsis                             \\
-         & COPD                                          \\
-         & CHF                                           \\
-         & Cirrhosis                                     \\
-         & Coma                                          \\
-         & Colon Cancer                                  \\
-         & Lung Cancer                                   \\
-         & MOSF w/Malig                                  \\
-dzclass  & ARF/MOSF                                      \\
-         & COPD/CHF/Cirrhosis                            \\
-         & Coma                                          \\
-         & Cancer                                        \\
-         & \textgreater{}\$50k                           \\
-race     & white                                         \\
-         & black                                         \\
-         & asian                                         \\
-         & other                                         \\
-         & hispanic                                      \\
-ca       & Yes                                           \\
-         & No                                            \\
-         & Metastatic                                    \\
-\end{tabular}
-\caption{A description of each level within each categorical variable in the dataset.}
-\label{tab:support2}
-\end{table}
-
-As before, the first step is to visualize the incidence density. The
-population-time plot for the SUPPORT data appears in Figure
-\ref{fig:support-poptime}. As we can see, the incidence density is not
-constant.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/support-poptime-1}
-
-}
-
-\caption{\label{fig:support-poptime} Population-time plot for the SUPPORT data.}\label{fig:support-poptime}
-\end{figure}
-\end{CodeChunk}
-
-Before generating an absolute risk curve using regularized logistic
-regression, we will explore three options as a baseline comparison,
-beyond a simple Kaplan-meier curve. For all the following analyses in
-this case study, we will use 80\% of the observations to train our
-models, while 20\% will be used to generate absolute risk curves, all of
-which will be averaged to demonstrate each model's predictive ability.
-We will then compare each modeling using the IPA score (???NEED
-REFERENCE HERE???) We will examine three cox regression scenarios: sps
-only, aps only and all covariates excluding sps and aps. The original
-study compared these scores over a span of 180 days, which may not be
-comparable over the full 5.56 years. Over 5.56 years,
-\ref{fig:baselines} demonstrates that ???BEST MODEL BASED ON IPA??? has
-the best calibration and discrimination with an IPA score of ???INSERT
-ACTUAL BEST SCORE??? . ??BEST MODEL?? will be used in the following
-comparison to regularized cox, regularized logistic regression and the
-unadjusted Kaplan-Meier curve.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/baselines-1}
-
-}
-
-\caption{\label{fig:baselines} AbsoluteRisk curves from hazard models.
-Pink: All covariates without SPS nor APS.
-Green: APS only model.
-Blue: SPS only model.
-Purple: unadjusted curve.
-All three models underestimate the risk in comparison to the unadjusted curve.}\label{fig:baselines}
-\end{figure}
-\end{CodeChunk}
-
-To get the absolute risk following regularized cox regression, this
-required that we work around limitations in existing packages. To make
-use of the \pkg{survival} package, we first get our hazard coefficients
-using \pkg{glmnet}'s coxnet implementation for lasso penalization
-\citep{regpathcox}. Next, we fit a cox model using the survival package
-with the selected variables only, from our previous fit. Finally, we
-replaced the coefficients in the survival object with those from our
-glmnet fit, allowing us to make use of the absolute risk tools available
-in the \pkg{survival} package.
-
-For our penalized logistic regression model, we opted for a logistic
-function of time, resulting in a Weibull distribution assumption. As
-casebase makes use of the glmnet package, we interact with the
-fitsmoothhazard.fit function using a matrix interface, where y contains
-the time and event variables and x contains all other variables we would
-like to include in the model. For the SUPPORT dataset, all factor
-variables were converted into dummy variables. Here, we used lasso
-penalization by setting alpha to 1, leading to our smooth hazard
-function, depicted in the ??? curve in \ref{fig:cs3Final}. ??? I AM
-STILL PENALIZING TIME ???
-
-We then take our hazard and use it to predict the absolute risk of each
-individual in our test set, over time. These individuals are then
-average resulting in the ???curve in \ref{fig:cs3Final}. ???SAME BUG IS
-BACK???
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> 
-R> cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = ratio, standardize = TRUE)
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Estimating the absolute risk curve using the newData parameter.
-R> abCbFull <- casebase::absoluteRisk(cbFull, time = seq(1, max(abKMcompare[, 1]), 25), newdata = newData, s = "lambda.1se", method = c("numerical"))
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> ggplot(as.data.frame(ab), aes(Time)) +
-R+   geom_line(data = as.data.frame(baseCoxComparisonData), mapping = aes(y = Full, colour = "Cox"), lwd = 2) +
-R+   geom_line(aes(y = `Absolute risk`, colour = Model), lwd = 2) +
-R+   geom_step(data = as.data.frame(abKMcompare), mapping = aes(x = Time, y = KM, color = "Unadjusted"), lwd = 2) +
-R+   labs(y = "Probability of Death", x = "Survival time (Days)")
-\end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics{../figures/unnamed-chunk-2-1}
-
-}
-
-\caption{\label{fig:baselines} HERES SOME TEXT ABT IT}\label{fig:unnamed-chunk-2}
-\end{figure}
-\end{CodeChunk}
-
-???SHORT SPIEL ABOUT IPA BETWEEN SEMI\_PARAMETRIC AND PARAMETRIC which
-leads to conclusion of best overall model. This refers to the COEFPLOT
-in a following figure???
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> 
-R> library(dotwhisker)
-R> library(broom)
-R> library(dplyr)
-R> 
-R> estimatescb <- setDT(as.data.frame(coef(cbFull)[-c(1, 2), 1]), keep.rownames = TRUE)[]
-R> # estimatescb$std.error<-0
-R> estimatescb$Model <- "casebase"
-R> colnames(estimatescb) <- c("term", "estimate", "model")
-R> 
-R> estimatescoxFull <- setDT(as.data.frame(coef(coxFull)), keep.rownames = TRUE)[]
-R> # estimatescoxFull$std.error<-0
-R> estimatescoxFull$Model <- "CoxFull"
-R> colnames(estimatescoxFull) <- c("term", "estimate", "model")
-R> 
-R> lolliplotDots <- rbind(estimatescb, estimatescoxNet, estimatescoxFull)
-R> lolliplotDots$conf.low <- lolliplotDots$estimate
-R> lolliplotDots$conf.high <- lolliplotDots$estimate
-R> lolliplotDots$conf.high[lolliplotDots$estimate < 0] <- 0
-R> lolliplotDots$conf.low[lolliplotDots$estimate > 0] <- 0
-R> 
-R> dwplot(lolliplotDots)
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/coefplots-1} \end{center}
-
-\end{CodeChunk}
-
-\hypertarget{case-study-4stanford-heart-transplant-data}{%
-\section{Case study 4--Stanford Heart Transplant
-Data}\label{case-study-4stanford-heart-transplant-data}}
-
-In the previous case studies, we only considered covariates that were
-fixed at baseline. In this next case study, we will use the Stanford
-Heart Transplant data
-\citep[\citet{crowley1977covariance}]{clark1971cardiac} to show how
-case-base sampling can also be used in the context of time-varying
-covariates. As an example that already appeared in the literature,
-case-base sampling was used to study vaccination safety, where the
-exposure period was defined as the week following vaccination
-\citep{saarela2015case}. Hence, the main covariate of interest,
-i.e.~exposure to the vaccine, was changing over time. In this context,
-case-base sampling offers an efficient alternative to nested
-case-control designs or self-matching.
-
-Recall the setting of Stanford Heart Transplant study: patients were
-admitted to the Stanford program after meeting with their physician and
-determining that they were unlikely to respond to other forms of
-treatment. After enrollment, the program searched for a suitable donor
-for the patient, which could take anywhere between a few days to almost
-a year. We are interested in the effect of a heart transplant on
-survival; therefore, the patient is considered exposed only after the
-transplant has occurred.
-
-As before, we can look at the population-time plot for a graphical
-summary of the event incidence. As we can see, most events occur early
-during the follow-up period, and therefore we do not expect the hazard
-to be constant.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(survival)
-R> library(casebase)
-R> 
-R> stanford_popTime <- popTime(jasa,
-R>   time = "futime",
-R>   event = "fustat"
-R> )
-R> plot(stanford_popTime)
-\end{CodeInput}
-\end{CodeChunk}
-
-Since the exposure is time-dependent, we need to manually define the
-exposure variable \emph{after} case-base sampling and \emph{before}
-fitting the hazard function. For this reason, we will use the
-\texttt{sampleCaseBase} function directly.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(tidyverse)
-R> library(lubridate)
-R> 
-R> cb_data <- sampleCaseBase(jasa,
-R>   time = "futime",
-R>   event = "fustat", ratio = 10
-R> )
-\end{CodeInput}
-\end{CodeChunk}
-
-Next, we will compute the number of days from acceptance into the
-program to transplant, and we use this variable to determine whether
-each population-moment is exposed or not.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Define exposure variable
-R> cb_data <- mutate(cb_data,
-R>   txtime = time_length(accept.dt %--% tx.date,
-R>     unit = "days"
-R>   ),
-R>   exposure = case_when(
-R>     is.na(txtime) ~ 0L,
-R>     txtime > futime ~ 0L,
-R>     txtime <= futime ~ 1L
-R>   )
-R> )
-\end{CodeInput}
-\end{CodeChunk}
-
-Finally, we can fit the hazard using various linear predictors.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(splines)
-R> # Fit several models
-R> fit1 <- fitSmoothHazard(fustat ~ exposure,
-R>   data = cb_data, time = "futime"
-R> )
-R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
-R>   data = cb_data, time = "futime"
-R> )
-R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
-R>   data = cb_data, time = "futime"
-R> )
-R> fit4 <- fitSmoothHazard(fustat ~ exposure * bs(futime),
-R>   data = cb_data, time = "futime"
-R> )
-\end{CodeInput}
-\end{CodeChunk}
-
-Note that the fourth model (i.e.~\texttt{fit4}) includes an interaction
-term between exposure and follow-up time. In other words, this model no
-longer exhibit proportional hazards. The evidence of non-proportionality
-of hazards in the Stanford Heart Transplant data has been widely
-discussed \citep{arjas1988graphical}.
-
-We can then compare the goodness of fit of these four models using the
-Akaike Information Criterion (AIC).
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute AIC
-R> c(
-R>   "Model1" = AIC(fit1),
-R>   "Model2" = AIC(fit2),
-R>   "Model3" = AIC(fit3),
-R>   "Model4" = AIC(fit4)
-R> )
-\end{CodeInput}
-\end{CodeChunk}
-
-As we can, the best fit is the fourth model. By visualizing the hazard
-functions for both exposed and unexposed individuals, we can more
-clearly see how the hazards are no longer proportional.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute hazards---
-R> # First, create a list of time points for both exposure status
-R> hazard_data <- expand.grid(
-R>   exposure = c(0, 1),
-R>   futime = seq(0, 1000,
-R>     length.out = 100
-R>   )
-R> )
-R> # Set the offset to zero
-R> hazard_data$offset <- 0
-R> # Use predict to get the fitted values, and exponentiate to
-R> # transform to the right scale
-R> hazard_data$hazard <- exp(predict(fit4,
-R>   newdata = hazard_data,
-R>   type = "link"
-R> ))
-R> # Add labels for plots
-R> hazard_data$Status <- factor(hazard_data$exposure,
-R>   labels = c("NoTrans", "Trans")
-R> )
-R> 
-R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
-R>   geom_line() +
-R>   theme_minimal() +
-R>   theme(legend.position = "top") +
-R>   ylab("Hazard") +
-R>   xlab("Follow-up time")
-\end{CodeInput}
-\end{CodeChunk}
-
-The non-proportionality seems to be more pronounced at the beginning of
-follow-up than the end. Finally, we can turn these estimates of the
-hazard function into estimates of the cumulative incidence functions.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute absolute risk curves
-R> newdata <- data.frame(exposure = c(0, 1))
-R> absrisk <- absoluteRisk(fit4,
-R>   newdata = newdata,
-R>   time = seq(0, 1000, length.out = 100)
-R> )
-R> 
-R> colnames(absrisk) <- c("Time", "NoTrans", "Trans")
-R> 
-R> # Rearrange the data
-R> absrisk <- gather(
-R>   as.data.frame(absrisk),
-R>   "Status", "Risk", -Time
-R> )
-R> 
-R> ggplot(absrisk, aes(Time, Risk, colour = Status)) +
-R>   geom_line() +
-R>   theme_minimal() +
-R>   theme(legend.position = "top") +
-R>   expand_limits(y = 0.5) +
-R>   xlab("Follow-up time") +
-R>   ylab("Cum. Incidence")
-\end{CodeInput}
-\end{CodeChunk}
-
-Note that we can easily adapt the code above to the situation where a
-patient receives a heart transplant at a point in time of interest, for
-example after 30 days.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute hazards---
-R> # First, create a list of time points for both exposure status
-R> one_yr_haz <- data.frame(futime = seq(0, 365,
-R>   length.out = 100
-R> ))
-R> one_yr_haz <- mutate(one_yr_haz,
-R>   offset = 0,
-R>   exposure = if_else(futime < 30, 0, 1)
-R> )
-R> one_yr_haz$hazard <- exp(predict(fit4,
-R>   newdata = one_yr_haz,
-R>   type = "link"
-R> ))
-R> ggplot(one_yr_haz, aes(futime, hazard)) +
-R>   geom_line() +
-R>   theme_minimal() +
-R>   theme(legend.position = "top") +
-R>   ylab("Hazard") +
-R>   xlab("Follow-up time") +
-R>   geom_vline(xintercept = 30, linetype = "dashed")
-\end{CodeInput}
-\end{CodeChunk}
-
-We can then compare the 1-year mortality risk without transplant and
-with transplant at 30 days.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> absoluteRisk(fit4,
-R>   newdata = data.frame(exposure = 0),
-R>   time = 365
-R> )
-R> 
-R> # Use the trapezoidal rule to estimate the cumulative hazard
-R> cumhaz_est <- pracma::trapz(one_yr_haz$futime, one_yr_haz$hazard)
-R> 1 - exp(-cumhaz_est)
-\end{CodeInput}
-\end{CodeChunk}
-
-As we can see, the risk estimate at 1-year is about 30\% lower if the
-patient receives a heart transplant at 30 days.
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-In this article, we presented the \proglang{R} package \pkg{casebase},
-which provides functions for fitting smooth parametric hazards and
-estimating cumulative incidence functions using case-base sampling. We
-outlined the theoretical underpinnings of the approach, we provided
-details about our implementation, and we illustrated the merits of the
-approach and the package through four case studies.
-
-As a methodological framework, case-base sampling is very flexible. This
-flexibility has been explored before in the literature: for example,
-Saarela and Hanley \citeyearpar{saarela2015case} used case-base sampling
-to model a time-dependent exposure variable in a vaccine safety study.
-As another example, Saarela and Arjas \citeyearpar{saarela2015non}
-combined case-base sampling and a Bayesian non-parametric framework to
-compute individualized risk assessments for chronic diseases. In the
-case studies above, we explored this flexibility along two fronts. On
-the one hand, we showed how splines could be used as part of the linear
-predictor to model the effect of time on the hazard. This strategy
-yielded estimates of the survival function that were qualitatively
-similar to semiparametric estimates derived from Cox regression;
-however, case-base sampling led to estimates of the survival function
-that \emph{vary smoothly in time}. On the other hand, we also displayed
-the flexibility of case-base sampling by showing how it could be
-combined with penalized logistic regression to perform variable
-selection. Even though we did not illustrate it in this article,
-case-base sampling can also be combined with the framework of
-\emph{generalized additive models}. This functionality has been
-implemented in our package \pkg{casebase}. Similarly, case-base sampling
-can also be combined with quasi-likelihood estimation to fit survival
-models that can account for the presence of over-dispersion. All of
-these examples illustrate how the case-base sampling framework in
-general, and the package \pkg{casebase} in particular, allows the user
-to fit a broad and flexible family of survival functions.
-
-As presented in Hanley \& Miettinen \citeyearpar{hanley2009fitting},
-case-base sampling is comprised of three steps: 1) sampling a case
-series and a base series from the study; 2) fit the log-hazard as a
-linear function of predictors (including time); and 3) use the fitted
-hazard to estimate the cumulative incidence curve. Accordingly, our
-package provides functions for each step. Moreover, the simple interface
-of the \code{fittingSmoothHazard} function resembles the \code{glm}
-interface. This interface should look familiar to new users. Our modular
-approach also provides a convenient way to extend our package for new
-sampling or fitting strategies.
-
-In the case studies above, we compared the performance of case-base
-sampling with that of Cox regression and Fine-Gray models. In terms of
-function interface, \pkg{casebase} uses a formula interface that is
-closer to that of \code{glm}, in that the event variable is the only
-variable appearing on the left-hand side of the formula. By contrast,
-both \code{survival::coxph} and \code{timereg::comp.risk} use arrays
-that captures both the event type and time. Both approaches to modeling
-yield user-friendly code. However, in terms of output, both approaches
-differ significantly. Case-base sampling produces smooth hazards and
-smooth cumulative incidence curves, whereas Cox regression and Fine-Gray
-models produce step-wise cumulative incidence functions and never
-explicitly model the hazard function. Qualitatively, we showed that by
-using splines in the linear predictor, all three models yielded similar
-curves. However, the smooth nature of the output of \pkg{casebase}
-provides a more intuitive interpretation for consumers of these
-predictions. In Table \ref{tab:compCBvsCox}, we provide a side-by-side
-comparison between the Cox model and case-base sampling.
-
-\begin{table}
-\caption{\label{tab:compCBvsCox}Comparison between the Cox model and case-base sampling}
-\centering
-\begin{tabular}[t]{llp{5cm}}
-\toprule
-Feature & Cox model & Case-base sampling\\
-\midrule
-Model type & Semi-parametric & Fully parametric\\
-Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\
-Cumulative incidence & Step function & Smooth-in-time curve\\
-Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\
-Model testing &  & Use GLM framework \newline (e.g.\ LRT, AIC, BIC)\\
-\addlinespace
-Competing risks & Difficult & Cause-specific CIFs\\
-% Prediction & Kaplan-Meier-based & ROC, AUC, risk reclassification probabilities\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-Our choice of modeling the log-hazard as a linear function of covariates
-allows us to develop a simple computational scheme for estimation.
-However, as a downside, it does not allow us to model location and scale
-parameters separately like the package \pkg{flexsurv}. For example, if
-we look at the Weibull distribution as parametrised in
-\code{stats::pweibull}, the log-hazard function is given by
-\[ \log \lambda(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,\]
-where \(\alpha,\beta\) are shape and scale parameters, respectively.
-Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} allows the
-user to model the scale parameter as a function of covariates. Of
-course, this added flexibility comes at the cost of interpretability: by
-modeling the log-hazard directly, the parameter estimates from
-\pkg{casebase} can be interpreted as estimates of log-hazard ratios. To
-improve the flexibility of \pkg{casebase} at capturing the scale of a
-parametric family, we could replace the logistic regression with its
-quasi-likelihood counterpart and therefore model over- and
-under-dispersion with respect to the logistic likelihood. We defer the
-study of the properties and performance of such a model to a future
-article.
-
-Future work will look at some of the methodological extensions of
-case-base sampling. First, to assess the quality of the model fit, we
-want to study the properties of the residuals (e.g.~Cox-Snell,
-martingale). More work needs to be done to understand these residuals in
-the context of the partial likelihood underlying case-base sampling. The
-resulting diagnostic tools would then be integrated in this package.
-Also, we are interested in extending case-base sampling to account for
-interval censoring. This type of censoring is very common in
-longitudinal studies, and many packages (e.g.~\pkg{SmoothHazard},
-\pkg{survival} and \pkg{Rstpm2}) provide functions to account for it.
-Again, we hope to include any resulting methodology as part of this
-package.
-
-In future versions of the package, we also want to increase the
-complement of diagnostic and inferential tools that are currently
-available. For example, we would like to include the ability to compute
-confidence intervals for the cumulative incidence curve. The delta
-method or parametric bootstrap are two different strategies we can use
-to construct approximate confidence intervals. Furthermore, we would
-like to include functions to compute calibration and discrimination
-statistics (e.g.~Brier score, AUC) for our models. Saarela and Arjas
-\citeyearpar{saarela2015non} also describe how to obtain a posterior
-distribution for the AUC from their model. Their approach could also be
-included in \pkg{casebase}. Finally, we want to provide more flexibility
-in how the case-base sampling is performed. This could be achieved by
-adding a \code{hazard} argument to the function \code{sampleCaseBase}.
-In this way, users could specify their own sampling mechanism. For
-example, they could provide a hazard that gives sampling probabilities
-that are proportional to the cardiovascular disease event rate given by
-the Framingham score \citep{saarela2015non}.
-
-In conclusion, we presented the \proglang{R} package \pkg{casebase}
-which implements case-base sampling for fitting parametric survival
-models and for estimating smooth cumulative incidence functions using
-the framework of generalized linear models. We strongly believe that its
-flexibility and its foundation on the familiar logistic regression model
-will make it appealing to new and established practioners.
-
-\hypertarget{environment-details}{%
-\section{Environment Details}\label{environment-details}}
-
-This report was generated on 2020-03-06 08:01:48 using the following
-computational environment and dependencies:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # which R packages and versions?
-R> # devtools::session_info()
-R> sessionInfo()
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> R version 3.6.1 (2019-07-05)
-#> Platform: x86_64-w64-mingw32/x64 (64-bit)
-#> Running under: Windows 10 x64 (build 18363)
-#> 
-#> Matrix products: default
-#> 
-#> locale:
-#> [1] LC_COLLATE=English_United States.1252 
-#> [2] LC_CTYPE=English_United States.1252   
-#> [3] LC_MONETARY=English_United States.1252
-#> [4] LC_NUMERIC=C                          
-#> [5] LC_TIME=English_United States.1252    
-#> 
-#> attached base packages:
-#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
-#> [8] base     
-#> 
-#> other attached packages:
-#>  [1] broom_0.5.5         dotwhisker_0.5.0    pracma_2.2.9       
-#>  [4] kableExtra_1.1.0    data.table_1.12.8   glmnet_3.0-2       
-#>  [7] Matrix_1.2-17       casebase_0.2.1.9001 survival_3.1-8     
-#> [10] magrittr_1.5        forcats_0.5.0       stringr_1.4.0      
-#> [13] dplyr_0.8.3         purrr_0.3.3         readr_1.3.1        
-#> [16] tidyr_1.0.2         tibble_2.1.3        ggplot2_3.3.0      
-#> [19] tidyverse_1.3.0    
-#> 
-#> loaded via a namespace (and not attached):
-#>  [1] httr_1.4.1        VGAM_1.1-2        jsonlite_1.6.1    viridisLite_0.3.0
-#>  [5] foreach_1.4.8     modelr_0.1.6      assertthat_0.2.1  stats4_3.6.1     
-#>  [9] ggstance_0.3.3    cellranger_1.1.0  yaml_2.2.1        pillar_1.4.3     
-#> [13] backports_1.1.5   lattice_0.20-38   glue_1.3.1        digest_0.6.25    
-#> [17] rvest_0.3.5       colorspace_1.4-1  plyr_1.8.6        htmltools_0.4.0  
-#> [21] pkgconfig_2.0.3   haven_2.2.0       scales_1.1.0      webshot_0.5.2    
-#> [25] mgcv_1.8-28       rticles_0.14      generics_0.0.2    farver_2.0.3     
-#> [29] withr_2.1.2       cli_2.0.2         crayon_1.3.4      readxl_1.3.1     
-#> [33] evaluate_0.14     fs_1.3.2          fansi_0.4.1       nlme_3.1-140     
-#> [37] xml2_1.2.2        tools_3.6.1       hms_0.5.3         lifecycle_0.1.0  
-#> [41] munsell_0.5.0     reprex_0.3.0      compiler_3.6.1    rlang_0.4.5      
-#> [45] grid_3.6.1        iterators_1.0.12  rstudioapi_0.11   labeling_0.3     
-#> [49] rmarkdown_2.1     gtable_0.3.0      codetools_0.2-16  DBI_1.1.0        
-#> [53] R6_2.4.1          lubridate_1.7.4   knitr_1.28        shape_1.4.4      
-#> [57] stringi_1.4.6     Rcpp_1.0.3        vctrs_0.2.3       dbplyr_1.4.2     
-#> [61] tidyselect_1.0.0  xfun_0.12
-\end{CodeOutput}
-\end{CodeChunk}
-
-The current Git commit details are:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # what commit is this file at?
-R> git2r::repository(here::here())
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> Local:    master C:/Users/jesse/Documents/gits/cbpaper
-#> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
-#> Head:     [6682dfb] 2020-03-06: NOT COMPLETE
-\end{CodeOutput}
-\end{CodeChunk}
-
 \bibliography{references.bib}

### 2020-03-20 analysis/paper/paper.Rmd
Auto stash before merge of "write-popTime" and "origin/master"

@@
 
+The splitting of these tasks gives the user the flexibility to create population-time plots with the graphics system of their choice using the object created in Step 1, while the object created in Step 2 allows one to leverage the \pkg{ggplot2} family of functions for further customization.
 1.
@@
 
+We first pass this dataset to the \code{popTime} function.
+If you do not specify the time and event arguments, this function will try to guess them using regular expressions.
+We then generate the plot by using the corresponding \code{plot} method:
 ```{r plot-erspc-data, eval = eval_cs1}
@@
 
+The randomization of the Finnish cohorts were carried out on January 1 of each of the 4 years 1996 to 1999.
+This, coupled with the uniform December 31 2006 censoring date, lead to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up.
+Tracked backwards in time (i.e.
+from right to left), the population-time plot shows the recruitment pattern from its beginning in 1991, and in particular the Jan 1 entries in successive years.
+Tracked forwards in time (i.e.
+from left to right), the plot for the first three years shows attrition due entirely to death (mainly from other causes).
+Since the Swedish and Belgian centres were the last to close their recruitment - in December 2003 - the minimum potential follow-up is three years.
+Tracked further forwards in time (i.e.
+after year 3) the attrition is
 We can also create exposure stratified plots by specifying the \code{exposure} argument in the \code{popTime} function:
@@
                                      exposure = "ScrArm")
-# plot(pt_object_strat)
+# class(pt_object_strat)
+# pt_object
+plot(pt_object_strat)
+# dev.off()
+# library(magrittr)
+# pt <- plot(pt_object_strat)
+# pt %>% class
+
+# library(ggplot2)
+# pt + theme_minimal()
+
 ```
@@
 
-```{r, eval = F}
+```{r, eval = T}
 knitr::knit_exit()

### 2020-03-26 analysis/paper/paper.Rmd
AddedBrier
Still need to do:
- Formatting is a little off
- some plots are not optimized yet
-interpreting the brier score, casebase is pretty bad!

--- 2020-03-06 analysis/paper/paper.Rmd
@@
 library(pracma)
+library(riskRegression)
+library(prodlim)
 options(digits = 2, scipen = 999)
@@
 For all the following analyses in this case study, we will use 80% of the observations to train our models, while 20% will be used to generate absolute risk curves, all of which will be averaged to demonstrate each model's predictive ability.
-We will then compare each modeling using the IPA score (???NEED REFERENCE HERE???) We will examine three cox regression scenarios: sps only, aps only and all covariates excluding sps and aps.
+We will then compare each modeling using the brier score [@graf1999ass] We will examine three cox regression scenarios: sps only, aps only and all covariates excluding sps and aps.
+
 The original study compared these scores over a span of 180 days, which may not be comparable over the full 5.56 years.
-Over 5.56 years, \ref{fig:baselines} demonstrates that ???BEST MODEL BASED ON IPA???
-has the best calibration and discrimination with an IPA score of ???INSERT ACTUAL BEST SCORE???
-.
-??BEST MODEL??
-will be used in the following comparison to regularized cox, regularized logistic regression and the unadjusted Kaplan-Meier curve.
+Over 5.56 years, \ref{fig:baselines} demonstrates that all the cox models have similar, if not identical average predictions.
+Figure \ref{fig:brier1} examines the brier scores as they change over time.
+As the full model has the lowest brier score over all, it will be used in the following comparison to regularized cox, regularized logistic regression and the unadjusted Kaplan-Meier curve.
 
@@
 fullDataCox <- as.data.frame(cbind(as.numeric(y[, 2]), as.numeric(y[, 1]), x))
+names(fullDataCox)[names(fullDataCox) == 'V1'] <- 'time'
+names(fullDataCox)[names(fullDataCox) == 'V2'] <- 'event'
 train_index <- sample(1:nrow(fullDataCox), 0.8 * nrow(fullDataCox))
@@
 test <- fullDataCox[test_index, ]
-coxSPS <- survival::coxph(Surv(time = V1, event = V2) ~ sps, data = train)
+coxSPS <- survival::coxph(Surv(time =time , event = event) ~ sps, data = train, x = TRUE)
 abcoxSPS <- survival::survfit(coxSPS, newdata = test)
 
-coxAPS <- survival::coxph(Surv(time = V1, event = V2) ~ aps, data = train)
+coxAPS <- survival::coxph(Surv(time =time , event = event) ~ aps, data = train, x = TRUE)
 abcoxAPS <- survival::survfit(coxAPS, newdata = test)
-
-coxFull <- survival::coxph(Surv(time = V1, event = V2) ~ . - sps - aps, data = train)
-abcoxFull <- survival::survfit(coxAPS, newdata = test)
+coxfullForumla<-as.formula(paste("Surv(time = time, event = event) ~ age + sexmale + slos + dzgroupCHF + 
+    dzgroupCirrhosis + `dzgroupColon Cancer` + dzgroupComa + 
+    dzgroupCOPD + `dzgroupLung Cancer` + `dzgroupMOSF w/Malig` + 
+    dzclassCancer + dzclassComa + `dzclassCOPD/CHF/Cirrhosis` + 
+    num.co + edu + scoma + avtisst + raceasian + raceblack + 
+    racehispanic + raceother + racewhite + hday + diabetes + 
+    dementia + cano + cayes + meanbp + wblc + hrt + resp + temp + 
+    pafi + alb + bili + crea + sod + ph + glucose + bun + urine + 
+    adlp + adlsc "))
+coxFull <- survival::coxph(coxfullForumla, data = train, x = TRUE)
+abcoxFull <- survival::survfit(coxFull, newdata = test)
 # library(prodlim)
 
-KM <- survival::coxph(Surv(time = V1, event = V2) ~ 1, data = test)
+KM <- survival::coxph(Surv(time = time, event = event) ~ 1, data = test)
 abKM <- survival::survfit(KM)
-# abKM <- prodlim(Hist(V1,V2)~1,data=test) minimal difference
+# abKM <- prodlim(Hist(,event)~1,data=test) minimal difference
 abKMcompare <- cbind(abKM$time[abKM$time <= 1631], 1 - abKM$surv[abKM$time <= 1631])
@@
 
-baseCoxComparisonData <- cbind(abcoxSPS$time[abcoxSPS$time <= 1631], 1 - rowMeans(abcoxSPS$surv[abcoxSPS$time <= 1631, ]), 1 - rowMeans(abcoxAPS$surv[abcoxSPS$time <= 1631, ]), 1 - abKM$surv, 1 - rowMeans(abcoxFull$surv[abcoxSPS$time <= 1631, ]))
-colnames(baseCoxComparisonData) <- c("Time", "SPS", "APS", "Unadjusted", "Full")
+baseCoxComparisonData <- cbind(abcoxSPS$time[abcoxSPS$time <= 1631], 1 - rowMeans(abcoxSPS$surv[abcoxSPS$time <= 1631, ]), 1 - rowMeans(abcoxAPS$surv[abcoxAPS$time <= 1631, ]), 1 - rowMeans(abcoxFull$surv[abcoxFull$time <= 1631, ]))
+
+colnames(baseCoxComparisonData) <- c("Time", "SPS", "APS","Full")
 ```
@@
 
+```{r brierpt1, echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:brier1} Brier scores over time. Full model performs best.",echo=FALSE}
+
+times <- sort(unique(test$time))
+brierAPS <- Score(list("PredictionModel" = coxAPS), data = test, formula = Hist(time, event != 0) ~ 1, summary = "ipa", se.fit = 0L, metrics = "brier", contrasts = FALSE, times = times)
+brierSPS <- Score(list("PredictionModel" = coxSPS), data = test, formula = Hist(time, event != 0) ~ 1, summary = "ipa", se.fit = 0L, metrics = "brier", contrasts = FALSE, times = times)
+
+coxFull$coefficients[which(is.na(coxFull$coefficients))]=0
+brierFull <- Score(list("PredictionModel" = coxFull), data = test, formula = Surv(time, event != 0) ~ 1, summary = "ipa", se.fit = 0L, metrics = "brier", contrasts = FALSE, times = times)
+
+brierPlotter<-rbind(
+      data.frame(brierScore = brierFull$Brier$score$Brier[brierFull$Brier$score$model == "PredictionModel"],times=times,model="cox Full"),
+      data.frame(brierScore = brierAPS$Brier$score$Brier[brierAPS$Brier$score$model == "PredictionModel"],times=times,model="cox APS only"),
+            data.frame(brierScore = brierSPS$Brier$score$Brier[brierSPS$Brier$score$model == "PredictionModel"],times=times,model="cox SPS only")
+      )
+
+ggplot(data = brierPlotter, mapping = aes(x = times, y = brierScore, col = model)) +
+  geom_line() +
+  xlab("Times") +
+  ylab("Brier-Score")+
+  ggtitle("Brier scores over time")
+```
+
 To get the absolute risk following regularized cox regression, this required that we work around limitations in existing packages.
@@
 For the SUPPORT dataset, all factor variables were converted into dummy variables.
-Here, we used lasso penalization by setting alpha to 1, leading to our smooth hazard function, depicted in the ???
-curve in \ref{fig:cs3Final}.
-??? I AM STILL PENALIZING TIME ???
-
+Here, we used lasso penalization by setting alpha to 1, leading to our smooth hazard function.
  We then take our hazard and use it to predict the absolute risk of each individual in our test set, over time.
-These individuals are then average resulting in the ???curve in \ref{fig:cs3Final}.
-???SAME BUG IS BACK???
+These individuals are then average resulting in the smooth absolute risk curve in Figure \ref{fig:cs3Final}.
+
+**note that I am still penalizing time, there's a bug I did not want to deal with when I try to use `penalty.factor` in fitsmoothhazard . If this is still here, i have not fixed it.**
 
@@
 # this is used more as an object place holder.
-coxNet <- survival::coxph(Surv(time = d.time, event = death) ~ ., data = cleanCoxData)
+coxNet <- survival::coxph(Surv(time = d.time, event = death) ~ ., data = cleanCoxData,x=TRUE)
 
@@
 
-estimatescoxNet$Model <- "CoxNet"
+estimatescoxNet$Model <- "Penalized Cox"
 colnames(estimatescoxNet) <- c("term", "estimate", "model")
@@
 
+brierCoxNet <- Score(list("PredictionModel" = coxNet), data = as.data.frame(cbind(test[,c(1,2)],newData[, nonzero_covariate_cox$X1])), formula = Hist(time, event != 0) ~ 1, summary = "ipa", se.fit = 0L, metrics = "brier", contrasts = FALSE, times = times)
+
 abCN <- as.data.frame(cbind(abcoxNet$time, 1 - rowMeans(abcoxNet$surv)))
@@
 
-cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = 10, standardize = TRUE)
+cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = 100, standardize = TRUE)
 ```
@@
 # Estimating the absolute risk curve using the newData parameter.
-abCbFull <- casebase::absoluteRisk(cbFull, time = seq(1, max(abKMcompare[, 1]), 25), newdata = newData, s = "lambda.1se", method = c("numerical"))
+abCbFull <- casebase::absoluteRisk(cbFull, time = times, newdata = newData, s = "lambda.1se", method = c("numerical"))
 ```
@@
 
-```{r include=TRUE,eval=eval_cs3,fig.cap="\\label{fig:baselines} HERES SOME TEXT ABT IT"}
+```{r include=TRUE,echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:cs3Final} HERES SOME TEXT ABT IT"}
 ggplot(as.data.frame(ab), aes(Time)) +
@@
 
-```{r coefplots, eval=eval_cs3}
+```{r finalBrierplot, echo=FALSE, eval=eval_cs3}
+brierScoreKMCens <- function(predSurvs, times, newData) {
+  # inverse probability censoring weights
+  # Note that we take probabilities right before the drops.
+
+  #DIFFERENCES START HERE
+  fitCens <- prodlim::prodlim(Hist(time, event != 0) ~ 1, newData, reverse = TRUE)
+  IPCW.subject.times <- prodlim::predictSurvIndividual(fitCens, lag = 1) # G(t-|X)
+  #DIFFERENCES END HERE
+
+  # Empty matrix that will be filled in with the following loop
+  Score <- matrix(NA, nrow(predSurvs), ncol(predSurvs))
+  IPCWMatrix <- matrix(NA, nrow(predSurvs), ncol(predSurvs))
+  # for each point in time we have predicted
+  for (i in 1:length(times)) {
+    # get number of censored individuals so long as their survival time is less than times[i]
+    # these individuals do not have an effect on right-censored brier score.
+    CensBefore <- newData$event == 0 & newData$time < times[i]
+    # y encompasses all survival times larger than t[i] with a 1, 0 otherwise
+    y <- drop(t(newData$time > times[i]))
+    # above permits the two parts of right-censored brier score to be calculated, without IPCW, in one line
+    Score[i, ] <- (y - predSurvs[i, ])^2
+
+    # Generate IPCWMatrix
+    IPCWMatrix[i, y == 0] <- IPCW.subject.times[y == 0] # G(t-|X) filled in corresponding positions
+    #DIFFERENCES START HERE
+    IPCW.time <- predict(fitCens, newdata = newData, times = times[i], level.chaos = 1, mode = "matrix", type = "surv", lag = 1)
+    #DIFFERENCES END HERE
+    IPCWMatrix[i, y == 1] <- IPCW.time # G(t) filled, same value, for remaining positions.
+    # above calculated individuals who don't have an effect on brier score at specfied time. they're scores are set to 0.
+    Score[i, CensBefore] <- 0
+  }
+
+  # apply IPCW to all scores
+  Err <- Score / IPCWMatrix
+
+  # Average curve demonstrating right-censored brier averaged over test-set, for each time of interest.
+  Err <- apply(Err, 1, mean)
+  return(Err)
+}
+absRiskcb <-abCbFull
+absRiskcb[, -c(1)] <- 1 - absRiskcb[, -c(1)] # make it survival probabilities
+absRiskcb <- absRiskcb[-c(1), -c(1)] # remove extra time at 0. Also remove the first column (times) as it matches perfectly now
+
+kmCensoredBrierScores <- brierScoreKMCens(predSurvs = absRiskcb, times = times, newData = data.frame(time=test$time,event=test$event))
+
+finalBrierDataToPlot<-rbind(
+  data.frame(brierScore = brierCoxNet$Brier$score$Brier[brierCoxNet$Brier$score$model == "PredictionModel"],times=times,model="Penalized Cox"),
+   data.frame(brierScore = brierFull$Brier$score$Brier[brierFull$Brier$score$model == "PredictionModel"],times=times,model="cox"),
+      rbind(data.frame(brierScore=kmCensoredBrierScores,times=times,model="Penalized logistic"))
+
+  )
+
+ggplot(data = finalBrierDataToPlot, mapping = aes(x = times, y = brierScore, col = model)) +
+  geom_line() +
+  xlab("Times") +
+  ylab("Brier-Score")+
+  ggtitle("Brier scores over time")
+
+```
+
+```{r coefplots,echo=FALSE, eval=eval_cs3}
 
@@
 # estimatescb$std.error<-0
-estimatescb$Model <- "casebase"
+estimatescb$Model <- "Penalized logistic"
 colnames(estimatescb) <- c("term", "estimate", "model")
@@
 # estimatescoxFull$std.error<-0
-estimatescoxFull$Model <- "CoxFull"
+estimatescoxFull$Model <- "Cox"
 colnames(estimatescoxFull) <- c("term", "estimate", "model")
@@
 lolliplotDots$conf.low[lolliplotDots$estimate > 0] <- 0
-
-dwplot(lolliplotDots)
+lolliplotDots$term=gsub( '\`', "", as.character(lolliplotDots$term))
+dwplot(lolliplotDots)+
+  theme(axis.text.y = element_text(size = 7.5, angle = 0, hjust = 1, vjust = 0),
+        strip.text.x = element_blank(),
+      strip.background = element_rect(colour="white", fill="white"),
+      legend.position=c(.855,0.15),legend.text=element_text(size=7.5)
+      )
 ```

### 2020-03-26 analysis/paper/paper.tex
Auto stash before merge of "write-popTime" and "origin/master"
AddedBrier
Still need to do:
- Formatting is a little off
- some plots are not optimized yet
-interpreting the brier score, casebase is pretty bad!

--- 2020-03-06 analysis/paper/paper.tex
@@
 which will be averaged to demonstrate each model's predictive ability.
-We will then compare each modeling using the IPA score (???NEED
-REFERENCE HERE???) We will examine three cox regression scenarios: sps
-only, aps only and all covariates excluding sps and aps. The original
-study compared these scores over a span of 180 days, which may not be
-comparable over the full 5.56 years. Over 5.56 years,
-\ref{fig:baselines} demonstrates that ???BEST MODEL BASED ON IPA??? has
-the best calibration and discrimination with an IPA score of ???INSERT
-ACTUAL BEST SCORE??? . ??BEST MODEL?? will be used in the following
-comparison to regularized cox, regularized logistic regression and the
-unadjusted Kaplan-Meier curve.
+We will then compare each modeling using the brier score
+\citep{graf1999ass} We will examine three cox regression scenarios: sps
+only, aps only and all covariates excluding sps and aps.
+
+The original study compared these scores over a span of 180 days, which
+may not be comparable over the full 5.56 years. Over 5.56 years,
+\ref{fig:baselines} demonstrates that all the cox models have similar,
+if not identical average predictions. Figure \ref{fig:brier1} examines
+the brier scores as they change over time. As the full model has the
+lowest brier score over all, it will be used in the following comparison
+to regularized cox, regularized logistic regression and the unadjusted
+Kaplan-Meier curve.
 
@@
 All three models underestimate the risk in comparison to the unadjusted curve.}\label{fig:baselines}
+\end{figure}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/brierpt1-1}
+
+}
+
+\caption{\label{fig:brier1} Brier scores over time. Full model performs best.}\label{fig:brierpt1}
 \end{figure}
@@
 penalization by setting alpha to 1, leading to our smooth hazard
-function, depicted in the ??? curve in \ref{fig:cs3Final}. ??? I AM
-STILL PENALIZING TIME ???
-
-We then take our hazard and use it to predict the absolute risk of each
-individual in our test set, over time. These individuals are then
-average resulting in the ???curve in \ref{fig:cs3Final}. ???SAME BUG IS
-BACK???
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> 
-R> cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = ratio, standardize = TRUE)
+function. We then take our hazard and use it to predict the absolute
+risk of each individual in our test set, over time. These individuals
+are then average resulting in the smooth absolute risk curve in Figure
+\ref{fig:cs3Final}.
+
+\textbf{note that I am still penalizing time, there's a bug I did not
+want to deal with when I try to use \texttt{penalty.factor} in
+fitsmoothhazard . If this is still here, i have not fixed it.}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> 
+R> cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = 100, standardize = TRUE)
 \end{CodeInput}
@@
 R> # Estimating the absolute risk curve using the newData parameter.
-R> abCbFull <- casebase::absoluteRisk(cbFull, time = seq(1, max(abKMcompare[, 1]), 25), newdata = newData, s = "lambda.1se", method = c("numerical"))
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> ggplot(as.data.frame(ab), aes(Time)) +
-R+   geom_line(data = as.data.frame(baseCoxComparisonData), mapping = aes(y = Full, colour = "Cox"), lwd = 2) +
-R+   geom_line(aes(y = `Absolute risk`, colour = Model), lwd = 2) +
-R+   geom_step(data = as.data.frame(abKMcompare), mapping = aes(x = Time, y = KM, color = "Unadjusted"), lwd = 2) +
-R+   labs(y = "Probability of Death", x = "Survival time (Days)")
-\end{CodeInput}
+R> abCbFull <- casebase::absoluteRisk(cbFull, time = times, newdata = newData, s = "lambda.1se", method = c("numerical"))
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
 \begin{figure}
@@
 
-\caption{\label{fig:baselines} HERES SOME TEXT ABT IT}\label{fig:unnamed-chunk-2}
+\caption{\label{fig:cs3Final} HERES SOME TEXT ABT IT}\label{fig:unnamed-chunk-2}
 \end{figure}
@@
 
-\begin{CodeInput}
-R> 
-R> library(dotwhisker)
-R> library(broom)
-R> library(dplyr)
-R> 
-R> estimatescb <- setDT(as.data.frame(coef(cbFull)[-c(1, 2), 1]), keep.rownames = TRUE)[]
-R> # estimatescb$std.error<-0
-R> estimatescb$Model <- "casebase"
-R> colnames(estimatescb) <- c("term", "estimate", "model")
-R> 
-R> estimatescoxFull <- setDT(as.data.frame(coef(coxFull)), keep.rownames = TRUE)[]
-R> # estimatescoxFull$std.error<-0
-R> estimatescoxFull$Model <- "CoxFull"
-R> colnames(estimatescoxFull) <- c("term", "estimate", "model")
-R> 
-R> lolliplotDots <- rbind(estimatescb, estimatescoxNet, estimatescoxFull)
-R> lolliplotDots$conf.low <- lolliplotDots$estimate
-R> lolliplotDots$conf.high <- lolliplotDots$estimate
-R> lolliplotDots$conf.high[lolliplotDots$estimate < 0] <- 0
-R> lolliplotDots$conf.low[lolliplotDots$estimate > 0] <- 0
-R> 
-R> dwplot(lolliplotDots)
-\end{CodeInput}
+\begin{center}\includegraphics{../figures/finalBrierplot-1} \end{center}
+
+\end{CodeChunk}
+
+\begin{CodeChunk}
 
@@
 
-This report was generated on 2020-03-06 08:01:48 using the following
+This report was generated on 2020-03-26 16:26:24 using the following
 computational environment and dependencies:
@@
 #> other attached packages:
-#>  [1] broom_0.5.5         dotwhisker_0.5.0    pracma_2.2.9       
-#>  [4] kableExtra_1.1.0    data.table_1.12.8   glmnet_3.0-2       
-#>  [7] Matrix_1.2-17       casebase_0.2.1.9001 survival_3.1-8     
-#> [10] magrittr_1.5        forcats_0.5.0       stringr_1.4.0      
-#> [13] dplyr_0.8.3         purrr_0.3.3         readr_1.3.1        
-#> [16] tidyr_1.0.2         tibble_2.1.3        ggplot2_3.3.0      
-#> [19] tidyverse_1.3.0    
+#>  [1] broom_0.5.5               dotwhisker_0.5.0         
+#>  [3] prodlim_2019.11.13        riskRegression_2020.02.05
+#>  [5] pracma_2.2.9              kableExtra_1.1.0         
+#>  [7] data.table_1.12.8         glmnet_3.0-2             
+#>  [9] Matrix_1.2-17             casebase_0.2.1.9001      
+#> [11] survival_3.1-8            magrittr_1.5             
+#> [13] forcats_0.5.0             stringr_1.4.0            
+#> [15] dplyr_0.8.3               purrr_0.3.3              
+#> [17] readr_1.3.1               tidyr_1.0.2              
+#> [19] tibble_2.1.3              ggplot2_3.3.0            
+#> [21] tidyverse_1.3.0          
 #> 
 #> loaded via a namespace (and not attached):
-#>  [1] httr_1.4.1        VGAM_1.1-2        jsonlite_1.6.1    viridisLite_0.3.0
-#>  [5] foreach_1.4.8     modelr_0.1.6      assertthat_0.2.1  stats4_3.6.1     
-#>  [9] ggstance_0.3.3    cellranger_1.1.0  yaml_2.2.1        pillar_1.4.3     
-#> [13] backports_1.1.5   lattice_0.20-38   glue_1.3.1        digest_0.6.25    
-#> [17] rvest_0.3.5       colorspace_1.4-1  plyr_1.8.6        htmltools_0.4.0  
-#> [21] pkgconfig_2.0.3   haven_2.2.0       scales_1.1.0      webshot_0.5.2    
-#> [25] mgcv_1.8-28       rticles_0.14      generics_0.0.2    farver_2.0.3     
-#> [29] withr_2.1.2       cli_2.0.2         crayon_1.3.4      readxl_1.3.1     
-#> [33] evaluate_0.14     fs_1.3.2          fansi_0.4.1       nlme_3.1-140     
-#> [37] xml2_1.2.2        tools_3.6.1       hms_0.5.3         lifecycle_0.1.0  
-#> [41] munsell_0.5.0     reprex_0.3.0      compiler_3.6.1    rlang_0.4.5      
-#> [45] grid_3.6.1        iterators_1.0.12  rstudioapi_0.11   labeling_0.3     
-#> [49] rmarkdown_2.1     gtable_0.3.0      codetools_0.2-16  DBI_1.1.0        
-#> [53] R6_2.4.1          lubridate_1.7.4   knitr_1.28        shape_1.4.4      
-#> [57] stringi_1.4.6     Rcpp_1.0.3        vctrs_0.2.3       dbplyr_1.4.2     
-#> [61] tidyselect_1.0.0  xfun_0.12
+#>  [1] nlme_3.1-140        fs_1.3.2            cmprsk_2.2-9       
+#>  [4] lubridate_1.7.4     rms_5.1-4           webshot_0.5.2      
+#>  [7] RColorBrewer_1.1-2  httr_1.4.1          numDeriv_2016.8-1.1
+#> [10] tools_3.6.1         backports_1.1.5     R6_2.4.1           
+#> [13] rpart_4.1-15        Hmisc_4.3-1         DBI_1.1.0          
+#> [16] mgcv_1.8-28         colorspace_1.4-1    nnet_7.3-12        
+#> [19] withr_2.1.2         gridExtra_2.3       tidyselect_1.0.0   
+#> [22] compiler_3.6.1      quantreg_5.54       cli_2.0.2          
+#> [25] rvest_0.3.5         htmlTable_1.13.3    SparseM_1.78       
+#> [28] xml2_1.2.5          sandwich_2.5-1      labeling_0.3       
+#> [31] checkmate_2.0.0     scales_1.1.0        mvtnorm_1.1-0      
+#> [34] polspline_1.1.17    digest_0.6.25       foreign_0.8-71     
+#> [37] rmarkdown_2.1       base64enc_0.1-3     rticles_0.14       
+#> [40] jpeg_0.1-8.1        pkgconfig_2.0.3     htmltools_0.4.0    
+#> [43] dbplyr_1.4.2        htmlwidgets_1.5.1   rlang_0.4.5        
+#> [46] readxl_1.3.1        rstudioapi_0.11     VGAM_1.1-2         
+#> [49] farver_2.0.3        shape_1.4.4         generics_0.0.2     
+#> [52] zoo_1.8-7           jsonlite_1.6.1      acepack_1.4.1      
+#> [55] Formula_1.2-3       Rcpp_1.0.3          munsell_0.5.0      
+#> [58] fansi_0.4.1         lifecycle_0.2.0     multcomp_1.4-12    
+#> [61] stringi_1.4.6       yaml_2.2.1          MASS_7.3-51.4      
+#> [64] plyr_1.8.6          ggstance_0.3.3      grid_3.6.1         
+#> [67] crayon_1.3.4        lattice_0.20-38     haven_2.2.0        
+#> [70] hms_0.5.3           timereg_1.9.4       knitr_1.28         
+#> [73] pillar_1.4.3        codetools_0.2-16    stats4_3.6.1       
+#> [76] reprex_0.3.0        glue_1.3.1          evaluate_0.14      
+#> [79] latticeExtra_0.6-29 modelr_0.1.6        vctrs_0.2.4        
+#> [82] png_0.1-7           foreach_1.4.8       MatrixModels_0.4-1 
+#> [85] cellranger_1.1.0    gtable_0.3.0        assertthat_0.2.1   
+#> [88] xfun_0.12           viridisLite_0.3.0   iterators_1.0.12   
+#> [91] cluster_2.1.0       lava_1.6.7          TH.data_1.0-10
 \end{CodeOutput}
@@
 #> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
-#> Head:     [6682dfb] 2020-03-06: NOT COMPLETE
+#> Head:     [a01a2a8] 2020-03-06: Fixed lollipop plot
 \end{CodeOutput}

### 2020-03-31 analysis/paper/paper.Rmd
Fixed penal. on time and Brier
-Time is now penalized according to issue #99.
- Brier score is using the riskRegression package now. This uses KM as the censoring distribution... prodlim::prodlim() instead of survival::survfit()

TODO: describe Coefficient plot,

@@
 
-Before generating an absolute risk curve using regularized logistic regression, we will explore three options as a baseline comparison, beyond a simple Kaplan-meier curve.
+Before generating an absolute risk curve using regularized logistic regression, we will explore three options as a baseline comparison, beyond a Kaplan-meier curve.
+The original study compared these scores over a span of 180 days, which may not be comparable over the full 5.56 years.
+Therefore, We will examine three cox regression scenarios: sps only, aps only and all covariates excluding sps and aps.
+Next, we will compare each modeling using the Brier score [@graf1999ass].
+The Brier score is able to demonstrate both discrimination and calibration in a single model [@graf1999ass] and provide a comparable metric between semi-parametric and parametric methods in the survival context.
 For all the following analyses in this case study, we will use 80% of the observations to train our models, while 20% will be used to generate absolute risk curves, all of which will be averaged to demonstrate each model's predictive ability.
-We will then compare each modeling using the brier score [@graf1999ass] We will examine three cox regression scenarios: sps only, aps only and all covariates excluding sps and aps.
-
-The original study compared these scores over a span of 180 days, which may not be comparable over the full 5.56 years.
 Over 5.56 years, \ref{fig:baselines} demonstrates that all the cox models have similar, if not identical average predictions.
 Figure \ref{fig:brier1} examines the brier scores as they change over time.
-As the full model has the lowest brier score over all, it will be used in the following comparison to regularized cox, regularized logistic regression and the unadjusted Kaplan-Meier curve.
+As the full model has the lowest brier score on overall, it will be used in the following comparison to regularized cox, regularized logistic regression and the unadjusted Kaplan-Meier curve.
 
@@
 completeData <- casebase::support
+
+completeData$d.time<-completeData$d.time/365
+
 workingCompleteData <- model.matrix(death ~ . - d.time, data = completeData)[, -c(1)] # remove intercept
+
 # Create u and xCox, which will be used when fitting Cox with glmnet.
@@
 names(fullDataCox)[names(fullDataCox) == 'V2'] <- 'event'
+
+#here we specify train and test size
 train_index <- sample(1:nrow(fullDataCox), 0.8 * nrow(fullDataCox))
@@
   geom_line(aes(y = Full, col = "All Covariates without SPS APS"), lwd = 1) +
-  labs(y = "Probability of Death", x = "Survival time (Days)")
-```
-
-```{r brierpt1, echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:brier1} Brier scores over time. Full model performs best.",echo=FALSE}
+  labs(y = "Probability of Death", x = "Survival time (Years)")
+```
+
+```{r brierpt1, echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:brier1} Brier scores over time.
+Full model (Cox Full) performs best, while the other two (APS only and SPS only) are comparable.",echo=FALSE}
 
@@
  We then take our hazard and use it to predict the absolute risk of each individual in our test set, over time.
-These individuals are then average resulting in the smooth absolute risk curve in Figure \ref{fig:cs3Final}.
-
-**note that I am still penalizing time, there's a bug I did not want to deal with when I try to use `penalty.factor` in fitsmoothhazard . If this is still here, i have not fixed it.**
+The probabilites over time for each individual are averaged, resulting in the smooth absolute risk curve in figure \ref{fig:cs3Final}.
+On average, the Brier score is lower in our penalized logistic regression model, compared to the other models (fig.
+\ref{fig:cs3FinalBrier}).
+
+This case study serves to demonstrate how regularization can be used in a survival context, using logistic regression.
+Our approach performs comparably to Cox regression, demonstrating its validity in an applied setting.
 
@@
 colnames(abCN) <- c("Time", "Absolute risk")
-abCN$Model <- "Coxnet"
+abCN$Model <- "Penalized Cox"
 abCN <- abCN[abCN$Time <= max(abKMcompare[, 1]), ]
@@
 
-cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = 100, standardize = TRUE)
+cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = 10, standardize = TRUE,penalty.facto=c(0,rep(1,ncol(x))))
 ```
@@
 
-```{r include=TRUE,echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:cs3Final} HERES SOME TEXT ABT IT"}
+```{r include=TRUE,echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:cs3Final} Four absolute risk curves, integrated from the following hazard models: Cox (pink), penalized Cox (green), penalized logistic (blue) and Kaplan-Meier (purple)."}
 ggplot(as.data.frame(ab), aes(Time)) +
-  geom_line(data = as.data.frame(baseCoxComparisonData), mapping = aes(y = Full, colour = "Cox"), lwd = 2) +
-  geom_line(aes(y = `Absolute risk`, colour = Model), lwd = 2) +
-  geom_step(data = as.data.frame(abKMcompare), mapping = aes(x = Time, y = KM, color = "Unadjusted"), lwd = 2) +
-  labs(y = "Probability of Death", x = "Survival time (Days)")
-```
-
-???SHORT SPIEL ABOUT IPA BETWEEN SEMI_PARAMETRIC AND PARAMETRIC which leads to conclusion of best overall model. This refers to the COEFPLOT in a following figure???
-
-```{r finalBrierplot, echo=FALSE, eval=eval_cs3}
-brierScoreKMCens <- function(predSurvs, times, newData) {
-  # inverse probability censoring weights
-  # Note that we take probabilities right before the drops.
-
-  #DIFFERENCES START HERE
-  fitCens <- prodlim::prodlim(Hist(time, event != 0) ~ 1, newData, reverse = TRUE)
-  IPCW.subject.times <- prodlim::predictSurvIndividual(fitCens, lag = 1) # G(t-|X)
-  #DIFFERENCES END HERE
-
-  # Empty matrix that will be filled in with the following loop
-  Score <- matrix(NA, nrow(predSurvs), ncol(predSurvs))
-  IPCWMatrix <- matrix(NA, nrow(predSurvs), ncol(predSurvs))
-  # for each point in time we have predicted
-  for (i in 1:length(times)) {
-    # get number of censored individuals so long as their survival time is less than times[i]
-    # these individuals do not have an effect on right-censored brier score.
-    CensBefore <- newData$event == 0 & newData$time < times[i]
-    # y encompasses all survival times larger than t[i] with a 1, 0 otherwise
-    y <- drop(t(newData$time > times[i]))
-    # above permits the two parts of right-censored brier score to be calculated, without IPCW, in one line
-    Score[i, ] <- (y - predSurvs[i, ])^2
-
-    # Generate IPCWMatrix
-    IPCWMatrix[i, y == 0] <- IPCW.subject.times[y == 0] # G(t-|X) filled in corresponding positions
-    #DIFFERENCES START HERE
-    IPCW.time <- predict(fitCens, newdata = newData, times = times[i], level.chaos = 1, mode = "matrix", type = "surv", lag = 1)
-    #DIFFERENCES END HERE
-    IPCWMatrix[i, y == 1] <- IPCW.time # G(t) filled, same value, for remaining positions.
-    # above calculated individuals who don't have an effect on brier score at specfied time. they're scores are set to 0.
-    Score[i, CensBefore] <- 0
+  geom_line(data = as.data.frame(baseCoxComparisonData), mapping = aes(y = Full, colour = "Cox"), lwd = 1) +
+  geom_line(aes(y = `Absolute risk`, colour = Model), lwd = 1) +
+  geom_step(data = as.data.frame(abKMcompare), mapping = aes(x = Time, y = KM, color = "Unadjusted"), lwd = 1) +
+  labs(y = "Probability of Death", x = "Survival time (Years)")
+```
+
+```{r riskregressionBrier,echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:cs3FinalBrier} 4 curves comparing the brier scores as they change over time between a null model (Kaplan-Meier, black), Cox regression (yellow), penalized Cox regression (blue) and penalized logistic regression (green)."}
+
+predictRisk.singleRisk <- function(object, newdata, times, cause, ...) {
+  if (object$call[[1]]!="glm"){
+    #get all covariates excluding intercept and time
+    coVars=colnames(object$originalData$x)
+    #coVars is used in lines 44 and 50
+    newdata=data.matrix(drop(subset(newdata, select=coVars)))
   }
 
-  # apply IPCW to all scores
-  Err <- Score / IPCWMatrix
-
-  # Average curve demonstrating right-censored brier averaged over test-set, for each time of interest.
-  Err <- apply(Err, 1, mean)
-  return(Err)
+  # if (missing(cause)) stop("Argument cause should be the event type for which we predict the absolute risk.")
+  # the output of absoluteRisk is an array with dimension dependening on the length of the requested times:
+  # case 1: the number of time points is 1
+  #         dim(array) =  (length(time), NROW(newdata), number of causes in the data)
+  if (length(times) == 1) {
+    a <- absoluteRisk(object, newdata = newdata, time = times)
+    p <- matrix(a, ncol = 1)
+  } else {
+    # case 2 a) zero is included in the number of time points
+    if (0 %in% times) {
+      # dim(array) =  (length(time)+1, NROW(newdata)+1, number of causes in the data)
+      a <- absoluteRisk(object, newdata = newdata, time = times)
+      p <- t(a)
+    } else {
+      # case 2 b) zero is not included in the number of time points (but the absoluteRisk function adds it)
+      a <- absoluteRisk(object, newdata = newdata, time = times)
+      ### we need to invert the plot because, by default, we get cumulative incidence
+      #a[, -c(1)] <- 1 - a[, -c(1)]
+      ### we remove time 0 for everyone, and remove the time column
+      a <- a[-c(1), -c(1)] ### a[-c(1), ] to keep times column, but remove time 0 probabilities
+      # now we transpose the matrix because in riskRegression we work with number of
+      # observations in rows and time points in columns
+      p <- t(a)
     }
-absRiskcb <-abCbFull
-absRiskcb[, -c(1)] <- 1 - absRiskcb[, -c(1)] # make it survival probabilities
-absRiskcb <- absRiskcb[-c(1), -c(1)] # remove extra time at 0. Also remove the first column (times) as it matches perfectly now
-
-kmCensoredBrierScores <- brierScoreKMCens(predSurvs = absRiskcb, times = times, newData = data.frame(time=test$time,event=test$event))
-
-finalBrierDataToPlot<-rbind(
-  data.frame(brierScore = brierCoxNet$Brier$score$Brier[brierCoxNet$Brier$score$model == "PredictionModel"],times=times,model="Penalized Cox"),
-   data.frame(brierScore = brierFull$Brier$score$Brier[brierFull$Brier$score$model == "PredictionModel"],times=times,model="cox"),
-      rbind(data.frame(brierScore=kmCensoredBrierScores,times=times,model="Penalized logistic"))
-
-  )
-
-ggplot(data = finalBrierDataToPlot, mapping = aes(x = times, y = brierScore, col = model)) +
+  }
+  if (NROW(p) != NROW(newdata) || NCOL(p) != length(times)) {
+    stop(paste("\nPrediction matrix has wrong dimensions:\nRequested newdata x times: ", NROW(newdata), " x ", length(times), "\nProvided prediction matrix: ", NROW(p), " x ", NCOL(p), "\n\n", sep = ""))
+  }
+  p
+}
+
+class(cbFull) <- c("singleRisk",class(cbFull))
+brierFinalResults <- Score(list("Cox" = coxFull,"Penalized Cox"=coxNet,"Penalized Logistic"=cbFull), data = test, formula = Hist(time, event != 0) ~ 1, summary = "ipa", se.fit = 0L, metrics = "brier", contrasts = FALSE, times =times)
+
+levels(brierFinalResults$Brier$score$model)[levels(brierFinalResults$Brier$score$model)=="Null model"] <- "Kaplan-Meier"
+ggplot(data = brierFinalResults$Brier$score, mapping = aes(x = times, y = Brier, col = model)) +
   geom_line() +

### 2020-03-31 analysis/paper/paper.tex
Fixed penal. on time and Brier
-Time is now penalized according to issue #99.
- Brier score is using the riskRegression package now. This uses KM as the censoring distribution... prodlim::prodlim() instead of survival::survfit()

TODO: describe Coefficient plot,

@@
 regression, we will explore three options as a baseline comparison,
-beyond a simple Kaplan-meier curve. For all the following analyses in
-this case study, we will use 80\% of the observations to train our
-models, while 20\% will be used to generate absolute risk curves, all of
-which will be averaged to demonstrate each model's predictive ability.
-We will then compare each modeling using the brier score
-\citep{graf1999ass} We will examine three cox regression scenarios: sps
-only, aps only and all covariates excluding sps and aps.
-
-The original study compared these scores over a span of 180 days, which
-may not be comparable over the full 5.56 years. Over 5.56 years,
+beyond a Kaplan-meier curve. The original study compared these scores
+over a span of 180 days, which may not be comparable over the full 5.56
+years. Therefore, We will examine three cox regression scenarios: sps
+only, aps only and all covariates excluding sps and aps. Next, we will
+compare each modeling using the Brier score \citep{graf1999ass}. The
+Brier score is able to demonstrate both discrimination and calibration
+in a single model \citep{graf1999ass} and provide a comparable metric
+between semi-parametric and parametric methods in the survival context.
+For all the following analyses in this case study, we will use 80\% of
+the observations to train our models, while 20\% will be used to
+generate absolute risk curves, all of which will be averaged to
+demonstrate each model's predictive ability. Over 5.56 years,
 \ref{fig:baselines} demonstrates that all the cox models have similar,
@@
 the brier scores as they change over time. As the full model has the
-lowest brier score over all, it will be used in the following comparison
-to regularized cox, regularized logistic regression and the unadjusted
-Kaplan-Meier curve.
+lowest brier score on overall, it will be used in the following
+comparison to regularized cox, regularized logistic regression and the
+unadjusted Kaplan-Meier curve.
 
@@
 
-\caption{\label{fig:brier1} Brier scores over time. Full model performs best.}\label{fig:brierpt1}
+\caption{\label{fig:brier1} Brier scores over time. Full model (Cox Full) performs best, while the other two (APS only and SPS only) are comparable.}\label{fig:brierpt1}
 \end{figure}
@@
 function. We then take our hazard and use it to predict the absolute
-risk of each individual in our test set, over time. These individuals
-are then average resulting in the smooth absolute risk curve in Figure
-\ref{fig:cs3Final}.
-
-\textbf{note that I am still penalizing time, there's a bug I did not
-want to deal with when I try to use \texttt{penalty.factor} in
-fitsmoothhazard . If this is still here, i have not fixed it.}
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> 
-R> cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = 100, standardize = TRUE)
+risk of each individual in our test set, over time. The probabilites
+over time for each individual are averaged, resulting in the smooth
+absolute risk curve in figure \ref{fig:cs3Final}. On average, the Brier
+score is lower in our penalized logistic regression model, compared to
+the other models (fig.~\ref{fig:cs3FinalBrier}).
+
+This case study serves to demonstrate how regularization can be used in
+a survival context, using logistic regression. Our approach performs
+comparably to Cox regression, demonstrating its validity in an applied
+setting.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> 
+R> cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = 10, standardize = TRUE,penalty.facto=c(0,rep(1,ncol(x))))
 \end{CodeInput}
@@
 
-\caption{\label{fig:cs3Final} HERES SOME TEXT ABT IT}\label{fig:unnamed-chunk-2}
+\caption{\label{fig:cs3Final} Four absolute risk curves, integrated from the following hazard models: Cox (pink), penalized Cox (green), penalized logistic (blue) and Kaplan-Meier (purple).}\label{fig:unnamed-chunk-2}
 \end{figure}
@@
 
-???SHORT SPIEL ABOUT IPA BETWEEN SEMI\_PARAMETRIC AND PARAMETRIC which
-leads to conclusion of best overall model. This refers to the COEFPLOT
-in a following figure???
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics{../figures/finalBrierplot-1} \end{center}
-
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/riskregressionBrier-1}
+
+}
+
+\caption{\label{fig:cs3FinalBrier} 4 curves comparing the brier scores as they change over time between a null model (Kaplan-Meier, black), Cox regression (yellow), penalized Cox regression (blue) and penalized logistic regression (green).}\label{fig:riskregressionBrier}
+\end{figure}
 \end{CodeChunk}
@@
 
-This report was generated on 2020-03-26 16:26:24 using the following
+This report was generated on 2020-03-31 18:19:20 using the following
 computational environment and dependencies:
@@
 #> [13] rpart_4.1-15        Hmisc_4.3-1         DBI_1.1.0          
-#> [16] mgcv_1.8-28         colorspace_1.4-1    nnet_7.3-12        
+#> [16] mgcv_1.8-28         colorspace_1.4-1    nnet_7.3-13        
 #> [19] withr_2.1.2         gridExtra_2.3       tidyselect_1.0.0   
@@
 #> [52] zoo_1.8-7           jsonlite_1.6.1      acepack_1.4.1      
-#> [55] Formula_1.2-3       Rcpp_1.0.3          munsell_0.5.0      
+#> [55] Formula_1.2-3       Rcpp_1.0.4          munsell_0.5.0      
 #> [58] fansi_0.4.1         lifecycle_0.2.0     multcomp_1.4-12    
@@
 #> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
-#> Head:     [a01a2a8] 2020-03-06: Fixed lollipop plot
+#> Head:     [5f22217] 2020-03-26: AddedBrier
 \end{CodeOutput}

### 2020-04-09 analysis/paper/paper.Rmd
Uses development version of Jesse-Islam/riskRegression
It seems to be working with my version of riskRegression, with the function acting as a part of the package.

The issue with 1- absolute risk I was worried about is not longer an issue. Turns out It was wrong before anyway. predictRisk attempts to output the probability of the event occuring

@@
 
-predictRisk.singleRisk <- function(object, newdata, times, cause, ...) {
-  if (object$call[[1]]!="glm"){
-    #get all covariates excluding intercept and time
-    coVars=colnames(object$originalData$x)
-    #coVars is used in lines 44 and 50
-    newdata=data.matrix(drop(subset(newdata, select=coVars)))
-  }
-
-  # if (missing(cause)) stop("Argument cause should be the event type for which we predict the absolute risk.")
-  # the output of absoluteRisk is an array with dimension dependening on the length of the requested times:
-  # case 1: the number of time points is 1
-  #         dim(array) =  (length(time), NROW(newdata), number of causes in the data)
-  if (length(times) == 1) {
-    a <- absoluteRisk(object, newdata = newdata, time = times)
-    p <- matrix(a, ncol = 1)
-  } else {
-    # case 2 a) zero is included in the number of time points
-    if (0 %in% times) {
-      # dim(array) =  (length(time)+1, NROW(newdata)+1, number of causes in the data)
-      a <- absoluteRisk(object, newdata = newdata, time = times)
-      p <- t(a)
-    } else {
-      # case 2 b) zero is not included in the number of time points (but the absoluteRisk function adds it)
-      a <- absoluteRisk(object, newdata = newdata, time = times)
-      ### we need to invert the plot because, by default, we get cumulative incidence
-      #a[, -c(1)] <- 1 - a[, -c(1)]
-      ### we remove time 0 for everyone, and remove the time column
-      a <- a[-c(1), -c(1)] ### a[-c(1), ] to keep times column, but remove time 0 probabilities
-      # now we transpose the matrix because in riskRegression we work with number of
-      # observations in rows and time points in columns
-      p <- t(a)
-    }
-  }
-  if (NROW(p) != NROW(newdata) || NCOL(p) != length(times)) {
-    stop(paste("\nPrediction matrix has wrong dimensions:\nRequested newdata x times: ", NROW(newdata), " x ", length(times), "\nProvided prediction matrix: ", NROW(p), " x ", NCOL(p), "\n\n", sep = ""))
-  }
-  p
-}
-
-class(cbFull) <- c("singleRisk",class(cbFull))
-brierFinalResults <- Score(list("Cox" = coxFull,"Penalized Cox"=coxNet,"Penalized Logistic"=cbFull), data = test, formula = Hist(time, event != 0) ~ 1, summary = "ipa", se.fit = 0L, metrics = "brier", contrasts = FALSE, times =times)
+class(cbFull) <- c("singleEventCB",class(cbFull))
+brierFinalResults <- Score(list("Cox" = coxFull,"Penalized Cox"=coxNet,"Penalized Logistic"=cbFull), data = test, formula = Hist(time, event != 0) ~ 1, summary = "ibs", se.fit = 0L, metrics = "brier", contrasts = FALSE, times =times)
 
@@
 
-```{r coefplots,echo=FALSE, eval=eval_cs3}
+```{r coefplots,echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} A lollipop plot demonstrating the estimated coefficients between the following three models: Cox (blue), penalized Cox (Green), penalized logisitic (red).
+If a coefficient is estimated to be 0, it is removed from the plot"}
 
@@
 lolliplotDots$term=gsub( '\`', "", as.character(lolliplotDots$term))
+
+lolliplotDots[which(lolliplotDots$estimate==0),c(2,4,5)]<-NA
+
 dwplot(lolliplotDots)+

### 2020-04-09 analysis/paper/paper.tex
Uses development version of Jesse-Islam/riskRegression
It seems to be working with my version of riskRegression, with the function acting as a part of the package.

The issue with 1- absolute risk I was worried about is not longer an issue. Turns out It was wrong before anyway. predictRisk attempts to output the probability of the event occuring

@@
 \begin{CodeChunk}
-
-\begin{center}\includegraphics{../figures/coefplots-1} \end{center}
-
+\begin{figure}
+
+{\centering \includegraphics{../figures/coefplots-1}
+
+}
+
+\caption{\label{fig:cs3FinalBrier} A lollipop plot demonstrating the estimated coefficients between the following three models: Cox (blue), penalized Cox (Green), penalized logisitic (red).
+If a coefficient is estimated to be 0, it is removed from the plot}\label{fig:coefplots}
+\end{figure}
 \end{CodeChunk}
@@
 
-This report was generated on 2020-03-31 18:19:20 using the following
+This report was generated on 2020-04-09 13:28:46 using the following
 computational environment and dependencies:
@@
 \begin{CodeOutput}
-#> R version 3.6.1 (2019-07-05)
+#> R version 3.6.3 (2020-02-29)
 #> Platform: x86_64-w64-mingw32/x64 (64-bit)
@@
 #>  [1] broom_0.5.5               dotwhisker_0.5.0         
-#>  [3] prodlim_2019.11.13        riskRegression_2020.02.05
+#>  [3] prodlim_2019.11.13        riskRegression_2020.03.27
 #>  [5] pracma_2.2.9              kableExtra_1.1.0         
 #>  [7] data.table_1.12.8         glmnet_3.0-2             
-#>  [9] Matrix_1.2-17             casebase_0.2.1.9001      
-#> [11] survival_3.1-8            magrittr_1.5             
+#>  [9] Matrix_1.2-18             casebase_0.2.1.9001      
+#> [11] survival_3.1-11           magrittr_1.5             
 #> [13] forcats_0.5.0             stringr_1.4.0            
-#> [15] dplyr_0.8.3               purrr_0.3.3              
+#> [15] dplyr_0.8.5               purrr_0.3.3              
 #> [17] readr_1.3.1               tidyr_1.0.2              
-#> [19] tibble_2.1.3              ggplot2_3.3.0            
+#> [19] tibble_3.0.0              ggplot2_3.3.0            
 #> [21] tidyverse_1.3.0          
@@
 #> loaded via a namespace (and not attached):
-#>  [1] nlme_3.1-140        fs_1.3.2            cmprsk_2.2-9       
-#>  [4] lubridate_1.7.4     rms_5.1-4           webshot_0.5.2      
-#>  [7] RColorBrewer_1.1-2  httr_1.4.1          numDeriv_2016.8-1.1
-#> [10] tools_3.6.1         backports_1.1.5     R6_2.4.1           
-#> [13] rpart_4.1-15        Hmisc_4.3-1         DBI_1.1.0          
-#> [16] mgcv_1.8-28         colorspace_1.4-1    nnet_7.3-13        
-#> [19] withr_2.1.2         gridExtra_2.3       tidyselect_1.0.0   
-#> [22] compiler_3.6.1      quantreg_5.54       cli_2.0.2          
-#> [25] rvest_0.3.5         htmlTable_1.13.3    SparseM_1.78       
-#> [28] xml2_1.2.5          sandwich_2.5-1      labeling_0.3       
-#> [31] checkmate_2.0.0     scales_1.1.0        mvtnorm_1.1-0      
-#> [34] polspline_1.1.17    digest_0.6.25       foreign_0.8-71     
-#> [37] rmarkdown_2.1       base64enc_0.1-3     rticles_0.14       
-#> [40] jpeg_0.1-8.1        pkgconfig_2.0.3     htmltools_0.4.0    
-#> [43] dbplyr_1.4.2        htmlwidgets_1.5.1   rlang_0.4.5        
-#> [46] readxl_1.3.1        rstudioapi_0.11     VGAM_1.1-2         
-#> [49] farver_2.0.3        shape_1.4.4         generics_0.0.2     
-#> [52] zoo_1.8-7           jsonlite_1.6.1      acepack_1.4.1      
-#> [55] Formula_1.2-3       Rcpp_1.0.4          munsell_0.5.0      
-#> [58] fansi_0.4.1         lifecycle_0.2.0     multcomp_1.4-12    
-#> [61] stringi_1.4.6       yaml_2.2.1          MASS_7.3-51.4      
-#> [64] plyr_1.8.6          ggstance_0.3.3      grid_3.6.1         
-#> [67] crayon_1.3.4        lattice_0.20-38     haven_2.2.0        
-#> [70] hms_0.5.3           timereg_1.9.4       knitr_1.28         
-#> [73] pillar_1.4.3        codetools_0.2-16    stats4_3.6.1       
-#> [76] reprex_0.3.0        glue_1.3.1          evaluate_0.14      
-#> [79] latticeExtra_0.6-29 modelr_0.1.6        vctrs_0.2.4        
-#> [82] png_0.1-7           foreach_1.4.8       MatrixModels_0.4-1 
-#> [85] cellranger_1.1.0    gtable_0.3.0        assertthat_0.2.1   
-#> [88] xfun_0.12           viridisLite_0.3.0   iterators_1.0.12   
-#> [91] cluster_2.1.0       lava_1.6.7          TH.data_1.0-10
+#>  [1] TH.data_1.0-10      VGAM_1.1-2          colorspace_1.4-1   
+#>  [4] ellipsis_0.3.0      ggstance_0.3.4      htmlTable_1.13.3   
+#>  [7] base64enc_0.1-3     fs_1.4.1            rstudioapi_0.11    
+#> [10] farver_2.0.3        MatrixModels_0.4-1  fansi_0.4.1        
+#> [13] mvtnorm_1.1-0       lubridate_1.7.8     xml2_1.3.1         
+#> [16] codetools_0.2-16    knitr_1.28          Formula_1.2-3      
+#> [19] jsonlite_1.6.1      cluster_2.1.0       dbplyr_1.4.2       
+#> [22] png_0.1-7           compiler_3.6.3      httr_1.4.1         
+#> [25] backports_1.1.6     assertthat_0.2.1    cli_2.0.2          
+#> [28] acepack_1.4.1       htmltools_0.4.0     quantreg_5.55      
+#> [31] tools_3.6.3         gtable_0.3.0        glue_1.4.0         
+#> [34] Rcpp_1.0.4.6        cellranger_1.1.0    vctrs_0.2.4        
+#> [37] nlme_3.1-145        iterators_1.0.12    xfun_0.12          
+#> [40] rvest_0.3.5         lifecycle_0.2.0     polspline_1.1.17   
+#> [43] MASS_7.3-51.5       zoo_1.8-7           scales_1.1.0       
+#> [46] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
+#> [49] RColorBrewer_1.1-2  rticles_0.14        yaml_2.2.1         
+#> [52] gridExtra_2.3       rms_5.1-4           rpart_4.1-15       
+#> [55] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
+#> [58] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
+#> [61] mets_1.2.7.1        rlang_0.4.5         pkgconfig_2.0.3    
+#> [64] evaluate_0.14       lattice_0.20-41     htmlwidgets_1.5.1  
+#> [67] labeling_0.3        cmprsk_2.2-9        tidyselect_1.0.0   
+#> [70] plyr_1.8.6          R6_2.4.1            generics_0.0.2     
+#> [73] Hmisc_4.4-0         multcomp_1.4-13     DBI_1.1.0          
+#> [76] pillar_1.4.3        haven_2.2.0         foreign_0.8-76     
+#> [79] withr_2.1.2         mgcv_1.8-31         nnet_7.3-13        
+#> [82] modelr_0.1.6        crayon_1.3.4        rmarkdown_2.1      
+#> [85] timereg_1.9.4       jpeg_0.1-8.1        grid_3.6.3         
+#> [88] readxl_1.3.1        reprex_0.3.0        digest_0.6.25      
+#> [91] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_3.6.3       
+#> [94] munsell_0.5.0       viridisLite_0.3.0
 \end{CodeOutput}
@@
 #> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
-#> Head:     [5f22217] 2020-03-26: AddedBrier
+#> Head:     [b31e195] 2020-03-31: Fixed penal. on time and Brier
 \end{CodeOutput}

### 2020-04-16 analysis/paper/paper.Rmd
Write up and plots fixed. plot descriptions fixed
All tasks complete. I will need to do a last grammar/writing check

@@
 library(prodlim)
+library(colorspace)
 options(digits = 2, scipen = 999)
@@
 To our knowledge, \pkg{casebase} and \pkg{rsptm2} are the only packages to offer this functionality.
-\textcolor{red}{I can dive deeper to determine if this is a fair statement}
 
@@
 In the first two case studies, we described the basic functionalities of the \pkg{casebase} package: creating population-time plots, fitting parametric models for hazard functions, and estimating the corresponding cumulative incidence curves.
-In the next two case studies, we explore more advanced features.
-
-For the third case study, we will look at regularized estimation for the hazard function.
-This can be performed using a combination of case-base sampling and regularized logistic regression.
-To demonstrate this capability, we examine the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [@knaus1995support].
+In the next two case studies, we explore advanced features.
+
+For the third case study, we will demonstrate regularization during hazard estimation.
+Since we are using logistic regression, existing packages, like \pkg{glmnet} can be leveraged.
+To demonstrate this feature, we examine the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [@knaus1995support].
 The SUPPORT dataset tracks death for individuals who are considered seriously ill within 5 hospitals in the United States of America.
 Imputation was conducted using the \pkg{mice} package in \proglang{R} with default settings [@mice].
-Before imputation, there were a total of 9105 individuals and 47 variables.
-Of those variables, a few were removed before imputation.
-We removed three response variables: hospital charges, patient ratio of costs to charges and patient micro-costs.
-Next, we removed hospital death as it was directly informative of our event of interest, death.
-We also removed functional disability and income as they are ordinal covariates, because ordinal variables are treated as factors in our current model.
-Finally, we removed 8 covariates relating to the results of previous findings: we removed SUPPORT day 3 physiology score (sps), 	APACHE III day 3 physiology score (aps),	SUPPORT model 2-month survival estimate, SUPPORT model 6-month survival estimate, Physician's 2-month survival estimate for pt., Physician's 6-month survival estimate for pt., Patient had Do Not Resuscitate (DNR) order and Day of DNR order (<0 if before study).
-Of these, sps and aps were added on after imputation, as they were missing only 1 observation and permitted us to compare findings between our model and their's.
-
-First we impute manually using recommended normal values for physiological measures [@harrell_2020].
-Next, we impute a single dataset using \pkg{mice} with default settings.
-After imputation, we noted that the covariate for surrogate activities of daily living was not imputed.
-This is due to colinearity between the other two covariates for activities of daily living.
-As such, surrogate activites of daily living was removed from our model.
-After these imputations, we have a total of 9104 observations and 30 variables, two of which are sps and aps, which will work as a baseline for our comparison to previous findings.
-This dataset is available  Its worth noting that these scores were developed for a period of 180 days, which makes our comparison over 5.56 years exploratory.
+The imputed dataset is available as part of the \pkg{casebase} package.
+For more information please see the SUPPORT data documentation in the package.
+
+we have a total of 9104 observations and 30 variables, two of which are SUPPORT day 3 physiology score (sps) and 	APACHE III day 3 physiology score (aps).
+These scores were developed for a period of 180 days, which makes our comparison over 5.56 years exploratory.
+Therefore, we will assess these two models before using either as our baseline, by comparing them to a full model using all variables.
 Also, our single imputation introduces bias in the data.
@@
 The original data was obtained from the Department of Biostatistics at Vanderbilt University [@harrell_2020].
-The imputed dataset is available as part of the \pkg{casebase} package.
 
@@
 
-<!--Hospital death as a covariate was removed, as this is directly informative of death.
-The glmnet package on CRAN was used to introduce elastic net, lasso and ridge regressions to casebase.--> As before, the first step is to visualize the incidence density.
-The population-time plot for the SUPPORT data appears in Figure \ref{fig:support-poptime}.
-As we can see, the incidence density is not constant.
-
-```{r support-poptime, echo=FALSE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data."}
-support <- casebase::support
-
-support_popTime <- popTime(support,
-  time = "d.time", event = "death", percentile_number = 0.31
-)
-plot(support_popTime)
-```
-
 Before generating an absolute risk curve using regularized logistic regression, we will explore three options as a baseline comparison, beyond a Kaplan-meier curve.
@@
 Figure \ref{fig:brier1} examines the brier scores as they change over time.
-As the full model has the lowest brier score on overall, it will be used in the following comparison to regularized cox, regularized logistic regression and the unadjusted Kaplan-Meier curve.
+As the full model has the lowest brier score on overall, it will be used in the following absolute risk comparison to regularized cox, regularized logistic regression and the unadjusted Kaplan-Meier curve.
 
 ```{r baselinePlotsPrep, eval=eval_cs3,echo=FALSE}
+q4 <- qualitative_hcl(4, palette = "Dark 3")
 completeData <- casebase::support
@@
 
-```{r baselines, echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:baselines} AbsoluteRisk curves from hazard models.
-Pink: All covariates without SPS nor APS.
-Green: APS only model.
-Blue: SPS only model.
-Purple: unadjusted curve.
-All three models underestimate the risk in comparison to the unadjusted curve.",echo=FALSE}
+```{r baselines, echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:baselines} Absolute risk curves estimated from hazard models.
+Red: APS only hazard was used to generate the absolute risk.
+Green: All covariates aside from APS and SPS.
+Blue: SPS only.
+Purple: unadjusted Kaplan-Meier curve.",echo=FALSE}
 ggplot(as.data.frame(baseCoxComparisonData), aes(Time)) +
+  geom_line(aes(y = APS, colour = "APS"), lwd = 1) +
+  geom_line(aes(y = Full, col = "Full"), lwd = 1) +
   geom_line(aes(y = SPS, colour = "SPS"), lwd = 1) +
-  geom_line(aes(y = APS, colour = "APS"), lwd = 1) +
   geom_step(data = as.data.frame(abKMcompare), mapping = aes(x = Time, y = KM, color = "Unadjusted"), lwd = 1) +
-  geom_line(aes(y = Full, col = "All Covariates without SPS APS"), lwd = 1) +
-  labs(y = "Probability of Death", x = "Survival time (Years)")
-```
-
-```{r brierpt1, echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:brier1} Brier scores over time.
-Full model (Cox Full) performs best, while the other two (APS only and SPS only) are comparable.",echo=FALSE}
+  labs(y = "Probability of Death", x = "Survival time (Years)",color = "Models")+
+   scale_color_manual(values=q4)
+```
+
+```{r brierpt1, echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:brier1} Brier scores over survival time.
+Full model (Full) performs best, while the other two (APS only and SPS only) are comparable.
+Red: APS only model was used to generate the brier scores.
+Green: all covariates excluding APS and SPS.
+Blue: SPS only model.",echo=FALSE}
 
@@
 brierPlotter<-rbind(
-      data.frame(brierScore = brierFull$Brier$score$Brier[brierFull$Brier$score$model == "PredictionModel"],times=times,model="cox Full"),
-      data.frame(brierScore = brierAPS$Brier$score$Brier[brierAPS$Brier$score$model == "PredictionModel"],times=times,model="cox APS only"),
-            data.frame(brierScore = brierSPS$Brier$score$Brier[brierSPS$Brier$score$model == "PredictionModel"],times=times,model="cox SPS only")
+      data.frame(brierScore = brierAPS$Brier$score$Brier[brierAPS$Brier$score$model == "PredictionModel"],times=times,model="APS"),
+      data.frame(brierScore = brierFull$Brier$score$Brier[brierFull$Brier$score$model == "PredictionModel"],times=times,model="Full"),
+            data.frame(brierScore = brierSPS$Brier$score$Brier[brierSPS$Brier$score$model == "PredictionModel"],times=times,model="SPS")
       )
@@
   geom_line() +
-  xlab("Times") +
+  xlab("Survival time (Years)") +
   ylab("Brier-Score")+
-  ggtitle("Brier scores over time")
-```
-
-To get the absolute risk following regularized cox regression, this required that we work around limitations in existing packages.
+   labs(color = "Models")+
+  ggtitle("Brier scores over time")+
+   scale_color_manual(values=q4[1:3])
+```
+
+To get the absolute risk following regularized cox regression, we had to work around limitations in existing packages.
 To make use of the \pkg{survival} package, we first get our hazard coefficients using \pkg{glmnet}'s coxnet implementation for lasso penalization [@regpathcox].
@@
 
-For our penalized logistic regression model, we opted for a logistic function of time, resulting in a Weibull distribution assumption.
+```{r coxNetCode,eval=FALSE,echo=TRUE}
+# hazard for cox using glmnet
+coxNet <- glmnet::cv.glmnet(x = x, y = y, family = "cox", alpha = 1, standardize = TRUE)
+#getting all non-zero covariates
+nonzero_covariate_cox <- predict(coxNet, type = "nonzero", s = "lambda.1se")
+#keeping coefficients for later
+nonzero_coef_cox <- coef(coxNet, s = "lambda.1se")
+# cutting dataset to only include nonzero covariates
+cleanCoxData <- as.data.frame(cbind(y, x[, nonzero_covariate_cox]))
+#create a skeleton object
+coxNet <- survival::coxph(Surv(time = d.time, event = death) ~ ., data = cleanCoxData,x=TRUE)
+#replace coefficients with regularized ones
+coxNet$coefficients <- nonzero_coef_cox@x
+```
+
+For our penalized logistic regression model, we opted for the natural log of time, which assumes a Weibull distribution for our event.
 As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model.
@@
 Here, we used lasso penalization by setting alpha to 1, leading to our smooth hazard function.
- We then take our hazard and use it to predict the absolute risk of each individual in our test set, over time.
-The probabilites over time for each individual are averaged, resulting in the smooth absolute risk curve in figure \ref{fig:cs3Final}.
-On average, the Brier score is lower in our penalized logistic regression model, compared to the other models (fig.
-\ref{fig:cs3FinalBrier}).
-
-This case study serves to demonstrate how regularization can be used in a survival context, using logistic regression.
-Our approach performs comparably to Cox regression, demonstrating its validity in an applied setting.
+
+```{r casebaseDisplayCode, eval=FALSE,echo=TRUE}
+#regularized logistic regression to estimate hazard
+cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet",
+                      time = "d.time", event = "death",
+                      formula_time = ~ log(d.time), alpha = 1,
+                      ratio = 100, standardize = TRUE, penalty.factor=c(0,rep(1,ncol(x))))
+
+# Estimating the absolute risk curve using the newData parameter.
+abCbFull <- casebase::absoluteRisk(cbFull, time = times, newdata = newData, s = "lambda.1se", method = c("numerical"))
+
+```
 
@@
 
-#ratio <- 100 # universal ratio for TRUE
-
+#remove apsSPS
+train<-train[,-c(46,47)]
+test<-test[,-c(46,47)]
 # x and y will be used to fit the casebase model under glmnet.
@@
 
-cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = 10, standardize = TRUE,penalty.facto=c(0,rep(1,ncol(x))))
+cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = 10, standardize = TRUE,penalty.factor=c(0,rep(1,ncol(x))))
 ```
@@
 
-```{r include=TRUE,echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:cs3Final} Four absolute risk curves, integrated from the following hazard models: Cox (pink), penalized Cox (green), penalized logistic (blue) and Kaplan-Meier (purple)."}
-ggplot(as.data.frame(ab), aes(Time)) +
-  geom_line(data = as.data.frame(baseCoxComparisonData), mapping = aes(y = Full, colour = "Cox"), lwd = 1) +
-  geom_line(aes(y = `Absolute risk`, colour = Model), lwd = 1) +
-  geom_step(data = as.data.frame(abKMcompare), mapping = aes(x = Time, y = KM, color = "Unadjusted"), lwd = 1) +
-  labs(y = "Probability of Death", x = "Survival time (Years)")
-```
-
-```{r riskregressionBrier,echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:cs3FinalBrier} 4 curves comparing the brier scores as they change over time between a null model (Kaplan-Meier, black), Cox regression (yellow), penalized Cox regression (blue) and penalized logistic regression (green)."}
-
-class(cbFull) <- c("singleEventCB",class(cbFull))
-brierFinalResults <- Score(list("Cox" = coxFull,"Penalized Cox"=coxNet,"Penalized Logistic"=cbFull), data = test, formula = Hist(time, event != 0) ~ 1, summary = "ibs", se.fit = 0L, metrics = "brier", contrasts = FALSE, times =times)
-
-levels(brierFinalResults$Brier$score$model)[levels(brierFinalResults$Brier$score$model)=="Null model"] <- "Kaplan-Meier"
-ggplot(data = brierFinalResults$Brier$score, mapping = aes(x = times, y = Brier, col = model)) +
-  geom_line() +
-  xlab("Times") +
-  ylab("Brier-Score")+
-  ggtitle("Brier scores over time")
-
-```
-
-```{r coefplots,echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} A lollipop plot demonstrating the estimated coefficients between the following three models: Cox (blue), penalized Cox (Green), penalized logisitic (red).
+```{r coefplots,echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} A lollipop plot demonstrating the estimated coefficients between the following three models: Cox (green), penalized Cox (red), penalized logisitic (blue).
 If a coefficient is estimated to be 0, it is removed from the plot"}
@@
 
-lolliplotDots <- rbind(estimatescb, estimatescoxNet, estimatescoxFull)
+lolliplotDots <- rbind(estimatescoxFull, estimatescoxNet,estimatescb)
 lolliplotDots$conf.low <- lolliplotDots$estimate
@@
 lolliplotDots[which(lolliplotDots$estimate==0),c(2,4,5)]<-NA
-
+unChosen<- unique(lolliplotDots[which(is.na(lolliplotDots$estimate)),1])
+
+lolliplotDots[which(lolliplotDots$term %in% unChosen$term),c(2,4,5)]<-NA
+
+lolliplotDots<-lolliplotDots[-which(is.na(lolliplotDots$estimate)),]
 dwplot(lolliplotDots)+
-  theme(axis.text.y = element_text(size = 7.5, angle = 0, hjust = 1, vjust = 0),
+  theme(axis.text.y = element_text(size = 11, angle = 0, hjust = 1, vjust = 0),
         strip.text.x = element_blank(),
       strip.background = element_rect(colour="white", fill="white"),
-      legend.position=c(.855,0.15),legend.text=element_text(size=7.5)
-      )
-```
+      legend.position=c(.855,0.15),legend.text=element_text(size=11)
+      )+
+  ggtitle("Coefficient plot")+
+   labs(color = "Models")+
+   scale_color_manual(values=q4[c(2,1,3)])
+```
+
+```{r include=TRUE,echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:cs3Final} Four absolute risk curves, integrated from the following hazard models: Cox (green), penalized Cox (red), penalized logistic (blue) and unadjusted (Kaplan-Meier, purple)."}
+ggplot(as.data.frame(ab), aes(Time)) +
+  geom_line(data = as.data.frame(baseCoxComparisonData), mapping = aes(y = Full, colour = "Cox"), lwd = 1) +
+  geom_line(aes(y = `Absolute risk`, colour = Model), lwd = 1) +
+  geom_step(data = as.data.frame(abKMcompare), mapping = aes(x = Time, y = KM, color = "Unadjusted"), lwd = 1) +
+  labs(y = "Probability of Death", x = "Survival time (Years)",color = "Models")+
+  ggtitle("Absolute risk curves")+
+   scale_color_manual(values=q4[c(2,1,3,4)])
+```
+
+```{r riskregressionBrier,echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:cs3FinalBrier} 4 curves comparing the brier scores as they change over time between Cox regression (green), penalized Cox regression (red) and penalized logistic regression (blue) and a unadjusted curve (Kaplan-Meier, purple).
+Penalized logistic regression produced the lowest curve."}
+
+class(cbFull) <- c("singleEventCB",class(cbFull))
+brierFinalResults <- Score(list("Cox" = coxFull,"Penalized Cox"=coxNet,"Penalized Logistic"=cbFull), data = test, formula = Hist(time, event != 0) ~ 1, summary = NULL, se.fit = 0L, metrics = "brier", contrasts = FALSE, times =times)
+
+levels(brierFinalResults$Brier$score$model)[levels(brierFinalResults$Brier$score$model)=="Null model"] <- "Unadjusted"
+b=brierFinalResults
+
+brierFinalResults$Brier$score$model<- factor(brierFinalResults$Brier$score$model, levels = c("Cox","Penalized Cox","Penalized Logistic", "Unadjusted"))
+
+ggplot(data = brierFinalResults$Brier$score, mapping = aes(x = times, y = Brier, col = model)) +
+  geom_line() +
+  xlab("Survival time (Years)") +
+  ylab("Brier-Score")+
+  ggtitle("Brier scores over time")+
+  labs(color = "Models")+
+  scale_color_manual(values=q4[c(2,1,3,4)])
+
+```
+
+Our first comparison plot examines the covariates that were selected between all three hazard functions.
+Note that Cox regression without penalization selected more covariates than the other two, but only the intersection of cofficients with values greater than 0 for all models are displayed in \ref{fig:cs3lolliPlot}.
+A visible tendency is for Cox regression to have larger estimated coefficients.
+This is expected, as there are no restrictions on the size of coefficients.
+The smallest coefficient estimates are found in penalized logistic model, aside from cancer state (cano).
+These values do not inform us of model performance.
+To do so, we take our hazard functions and use them to predict the absolute risk of each individual in our test set, over time.
+The probabilites over time for each individual are averaged, resulting in the absolute risk curves in figure \ref{fig:cs3Final}.
+There is separation between the curves, but it is minimal.
+To better demonstrate the differences between the models, we compare brier scores over time.
+\ref{fig:cs3FinalBrier} the Brier score is lower in our penalized logistic regression model, compared to the other models.
+This difference, however, is minimal.
+We can say with relative certainty that any of the hazard models outperform an unadjusted curve, while penalized logistic regression performs the best on this testing dataset.
 

### 2020-04-16 analysis/paper/paper.tex
Write up and plots fixed. plot descriptions fixed
All tasks complete. I will need to do a last grammar/writing check

--- 2020-03-31 analysis/paper/paper.tex
@@
 are the only packages to offer this functionality.
-\textcolor{red}{I can dive deeper to determine if this is a fair statement}
 
@@
 R> ERSPC$ScrArm <- factor(ERSPC$ScrArm,
-R>   levels = c(0, 1),
-R>   labels = c("Control group", "Screening group")
-R> )
+R+   levels = c(0, 1),
+R+   labels = c("Control group", "Screening group")
+R+ )
 \end{CodeInput}
@@
 R> pt_object_strat <- casebase::popTime(ERSPC,
-R>   event = "DeadOfPrCa",
-R>   exposure = "ScrArm"
-R> )
+R+   event = "DeadOfPrCa",
+R+   exposure = "ScrArm"
+R+ )
 R> plot(pt_object_strat)
 \end{CodeInput}
+
+\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-1} \end{center}
+
 \end{CodeChunk}
@@
 R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm,
-R>   data = ERSPC
-R> )
+R+   data = ERSPC
+R+ )
 R> summary(cox_model)
 \end{CodeInput}
+
+\begin{CodeOutput}
+#> Call:
+#> survival::coxph(formula = Surv(Follow.Up.Time, DeadOfPrCa) ~ 
+#>     ScrArm, data = ERSPC)
+#> 
+#>   n= 159893, number of events= 540 
+#> 
+#>                         coef exp(coef) se(coef)     z Pr(>|z|)  
+#> ScrArmScreening group -0.222     0.801    0.088 -2.52    0.012 *
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#>                       exp(coef) exp(-coef) lower .95 upper .95
+#> ScrArmScreening group     0.801       1.25     0.674     0.952
+#> 
+#> Concordance= 0.519  (se = 0.011 )
+#> Likelihood ratio test= 6.45  on 1 df,   p=0.01
+#> Wald test            = 6.37  on 1 df,   p=0.01
+#> Score (logrank) test = 6.39  on 1 df,   p=0.01
+\end{CodeOutput}
 \end{CodeChunk}
@@
 R> new_data <- data.frame(
-R>   ScrArm = c("Control group", "Screening group"),
-R>   ignore = 99
-R> )
+R+   ScrArm = c("Control group", "Screening group"),
+R+   ignore = 99
+R+ )
 R> 
 R> plot(survfit(cox_model, newdata = new_data),
-R>   xlab = "Years since Randomization",
-R>   ylab = "Cumulative Incidence (%)",
-R>   fun = "event",
-R>   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
-R>   main = sprintf(
-R>     "Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>     exp(coef(cox_model)),
-R>     exp(confint(cox_model))[1],
-R>     exp(confint(cox_model))[2]
-R>   ),
-R>   ylim = c(0, 2), yscale = 100
-R> )
+R+   xlab = "Years since Randomization",
+R+   ylab = "Cumulative Incidence (%)",
+R+   fun = "event",
+R+   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
+R+   main = sprintf(
+R+     "Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+     exp(coef(cox_model)),
+R+     exp(confint(cox_model))[1],
+R+     exp(confint(cox_model))[2]
+R+   ),
+R+   ylim = c(0, 2), yscale = 100
+R+ )
 R> legend("topleft",
-R>   legend = c("Control group", "Screening group"),
-R>   col = c("red", "blue"),
-R>   lty = c(1, 1),
-R>   bg = "gray90"
-R> )
-\end{CodeInput}
+R+   legend = c("Control group", "Screening group"),
+R+   col = c("red", "blue"),
+R+   lty = c(1, 1),
+R+   bg = "gray90"
+R+ )
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/erspc-cox-cif-1} \end{center}
+
 \end{CodeChunk}
@@
 R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm,
-R>   data = ERSPC,
-R>   ratio = 100
-R> )
+R+   data = ERSPC,
+R+   ratio = 100
+R+ )
 R> 
 R> summary(casebase_exponential)
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> 
+#> Call:
+#> glm(formula = formula, family = binomial, data = sampleData)
+#> 
+#> Deviance Residuals: 
+#>    Min      1Q  Median      3Q     Max  
+#> -0.148  -0.148  -0.148  -0.133   3.078  
+#> 
+#> Coefficients:
+#>                       Estimate Std. Error z value            Pr(>|z|)    
+#> (Intercept)            -7.7659     0.0557 -139.46 <0.0000000000000002 ***
+#> ScrArmScreening group  -0.2147     0.0884   -2.43               0.015 *  
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#> (Dispersion parameter for binomial family taken to be 1)
+#> 
+#>     Null deviance: 6059  on 54539  degrees of freedom
+#> Residual deviance: 6053  on 54538  degrees of freedom
+#> AIC: 6057
+#> 
+#> Number of Fisher Scoring iterations: 7
+\end{CodeOutput}
+
+\begin{CodeInput}
 R> exp(coef(casebase_exponential)[2])
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> ScrArmScreening group 
+#>                  0.81
+\end{CodeOutput}
+
+\begin{CodeInput}
 R> exp(confint(casebase_exponential)[2, ])
 \end{CodeInput}
+
+\begin{CodeOutput}
+#>  2.5 % 97.5 % 
+#>   0.68   0.96
+\end{CodeOutput}
 \end{CodeChunk}
@@
 R> smooth_risk_exp <- casebase::absoluteRisk(
-R>   object = casebase_exponential,
-R>   time = seq(0, 15, 0.1),
-R>   newdata = new_data
-R> )
+R+   object = casebase_exponential,
+R+   time = seq(0, 15, 0.1),
+R+   newdata = new_data
+R+ )
 R> 
 R> plot(survfit(cox_model, newdata = new_data),
-R>   xlab = "Years since Randomization",
-R>   ylab = "Cumulative Incidence (%)",
-R>   fun = "event",
-R>   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
-R>   main = sprintf(
-R>     "Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>     exp(coef(cox_model)),
-R>     exp(confint(cox_model))[1],
-R>     exp(confint(cox_model))[2]
-R>   ),
-R>   ylim = c(0, 2), yscale = 100
-R> )
+R+   xlab = "Years since Randomization",
+R+   ylab = "Cumulative Incidence (%)",
+R+   fun = "event",
+R+   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
+R+   main = sprintf(
+R+     "Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+     exp(coef(cox_model)),
+R+     exp(confint(cox_model))[1],
+R+     exp(confint(cox_model))[2]
+R+   ),
+R+   ylim = c(0, 2), yscale = 100
+R+ )
 R> lines(smooth_risk_exp[, 1], smooth_risk_exp[, 2], col = "red", lty = 2)
@@
 R> legend("topleft",
-R>   legend = c(
-R>     "Control group (Cox)", "Control group (Casebase)",
-R>     "Screening group (Cox)", "Screening group (Casebase)"
-R>   ),
-R>   col = c("red", "red", "blue", "blue"),
-R>   lty = c(1, 2, 1, 2),
-R>   bg = "gray90"
-R> )
-\end{CodeInput}
+R+   legend = c(
+R+     "Control group (Cox)", "Control group (Casebase)",
+R+     "Screening group (Cox)", "Screening group (Casebase)"
+R+   ),
+R+   col = c("red", "red", "blue", "blue"),
+R+   lty = c(1, 2, 1, 2),
+R+   bg = "gray90"
+R+ )
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/erspc-casebase-cif-1} \end{center}
+
 \end{CodeChunk}
@@
 R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm,
-R>   data = ERSPC,
-R>   ratio = 100
-R> )
+R+   data = ERSPC,
+R+   ratio = 100
+R+ )
 R> 
 R> summary(casebase_time)
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> 
+#> Call:
+#> glm(formula = formula, family = binomial, data = sampleData)
+#> 
+#> Deviance Residuals: 
+#>    Min      1Q  Median      3Q     Max  
+#> -0.401  -0.162  -0.122  -0.094   3.466  
+#> 
+#> Coefficients:
+#>                       Estimate Std. Error z value            Pr(>|z|)    
+#> (Intercept)            -9.0314     0.1121  -80.53 <0.0000000000000002 ***
+#> Follow.Up.Time          0.2222     0.0145   15.29 <0.0000000000000002 ***
+#> ScrArmScreening group  -0.2320     0.0886   -2.62              0.0089 ** 
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#> (Dispersion parameter for binomial family taken to be 1)
+#> 
+#>     Null deviance: 6059  on 54539  degrees of freedom
+#> Residual deviance: 5811  on 54537  degrees of freedom
+#> AIC: 5817
+#> 
+#> Number of Fisher Scoring iterations: 8
+\end{CodeOutput}
+
+\begin{CodeInput}
 R> exp(coef(casebase_time))
+\end{CodeInput}
+
+\begin{CodeOutput}
+#>           (Intercept)        Follow.Up.Time ScrArmScreening group 
+#>               0.00012               1.24882               0.79297
+\end{CodeOutput}
+
+\begin{CodeInput}
 R> exp(confint(casebase_time))
 \end{CodeInput}
+
+\begin{CodeOutput}
+#>                          2.5 %  97.5 %
+#> (Intercept)           0.000096 0.00015
+#> Follow.Up.Time        1.213876 1.28507
+#> ScrArmScreening group 0.665769 0.94256
+\end{CodeOutput}
 \end{CodeChunk}
@@
 R> smooth_risk_time <- casebase::absoluteRisk(
-R>   object = casebase_time,
-R>   time = seq(0, 15, 0.1),
-R>   newdata = new_data
-R> )
+R+   object = casebase_time,
+R+   time = seq(0, 15, 0.1),
+R+   newdata = new_data
+R+ )
 R> 
 R> plot(survfit(cox_model, newdata = new_data),
-R>   xlab = "Years since Randomization",
-R>   ylab = "Cumulative Incidence (%)",
-R>   fun = "event",
-R>   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
-R>   main = sprintf(
-R>     "Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>     exp(coef(cox_model)),
-R>     exp(confint(cox_model))[1],
-R>     exp(confint(cox_model))[2]
-R>   ),
-R>   ylim = c(0, 2), yscale = 100
-R> )
+R+   xlab = "Years since Randomization",
+R+   ylab = "Cumulative Incidence (%)",
+R+   fun = "event",
+R+   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
+R+   main = sprintf(
+R+     "Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+     exp(coef(cox_model)),
+R+     exp(confint(cox_model))[1],
+R+     exp(confint(cox_model))[2]
+R+   ),
+R+   ylim = c(0, 2), yscale = 100
+R+ )
 R> lines(smooth_risk_time[, 1], smooth_risk_time[, 2], col = "red", lty = 2)
@@
 R> legend("topleft",
-R>   legend = c(
-R>     "Control group (Cox)", "Control group (Casebase)",
-R>     "Screening group (Cox)", "Screening group (Casebase)"
-R>   ),
-R>   col = c("red", "red", "blue", "blue"),
-R>   lty = c(1, 2, 1, 2),
-R>   bg = "gray90"
-R> )
-\end{CodeInput}
+R+   legend = c(
+R+     "Control group (Cox)", "Control group (Casebase)",
+R+     "Screening group (Cox)", "Screening group (Casebase)"
+R+   ),
+R+   col = c("red", "red", "blue", "blue"),
+R+   lty = c(1, 2, 1, 2),
+R+   bg = "gray90"
+R+ )
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/erspc-casebase-weibull-cif-1} \end{center}
+
 \end{CodeChunk}
@@
 R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm,
-R>   data = ERSPC,
-R>   ratio = 100
-R> )
+R+   data = ERSPC,
+R+   ratio = 100
+R+ )
 R> 
 R> summary(casebase_splines)
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> 
+#> Call:
+#> glm(formula = formula, family = binomial, data = sampleData)
+#> 
+#> Deviance Residuals: 
+#>    Min      1Q  Median      3Q     Max  
+#> -0.405  -0.173  -0.136  -0.085   3.796  
+#> 
+#> Coefficients:
+#>                       Estimate Std. Error z value             Pr(>|z|)    
+#> (Intercept)           -10.2643     0.3159  -32.49 < 0.0000000000000002 ***
+#> bs(Follow.Up.Time)1     4.2983     0.8005    5.37        0.00000007893 ***
+#> bs(Follow.Up.Time)2     2.0945     0.4824    4.34        0.00001412028 ***
+#> bs(Follow.Up.Time)3     4.5544     0.7067    6.44        0.00000000012 ***
+#> ScrArmScreening group  -0.2169     0.0886   -2.45                0.014 *  
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#> (Dispersion parameter for binomial family taken to be 1)
+#> 
+#>     Null deviance: 6059.0  on 54539  degrees of freedom
+#> Residual deviance: 5790.5  on 54535  degrees of freedom
+#> AIC: 5800
+#> 
+#> Number of Fisher Scoring iterations: 8
+\end{CodeOutput}
+
+\begin{CodeInput}
 R> exp(coef(casebase_splines))
+\end{CodeInput}
+
+\begin{CodeOutput}
+#>           (Intercept)   bs(Follow.Up.Time)1   bs(Follow.Up.Time)2 
+#>              0.000035             73.578176              8.121029 
+#>   bs(Follow.Up.Time)3 ScrArmScreening group 
+#>             95.053534              0.805017
+\end{CodeOutput}
+
+\begin{CodeInput}
 R> exp(confint(casebase_splines))
 \end{CodeInput}
+
+\begin{CodeOutput}
+#>                           2.5 %     97.5 %
+#> (Intercept)            0.000018   0.000063
+#> bs(Follow.Up.Time)1   15.835008 367.061126
+#> bs(Follow.Up.Time)2    3.252368  21.652009
+#> bs(Follow.Up.Time)3   22.537699 364.141700
+#> ScrArmScreening group  0.675920   0.956823
+\end{CodeOutput}
 \end{CodeChunk}
@@
 R> smooth_risk_splines <- absoluteRisk(
-R>   object = casebase_splines,
-R>   time = seq(0, 15, 0.1),
-R>   newdata = new_data
-R> )
+R+   object = casebase_splines,
+R+   time = seq(0, 15, 0.1),
+R+   newdata = new_data
+R+ )
 R> 
 R> plot(survfit(cox_model, newdata = new_data),
-R>   xlab = "Years since Randomization",
-R>   ylab = "Cumulative Incidence (%)",
-R>   fun = "event",
-R>   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
-R>   main = sprintf(
-R>     "Estimated Cumulative Incidence (risk) of Death from Prostate 
-R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R>     exp(coef(cox_model)),
-R>     exp(confint(cox_model))[1],
-R>     exp(confint(cox_model))[2]
-R>   ),
-R>   ylim = c(0, 2), yscale = 100
-R> )
+R+   xlab = "Years since Randomization",
+R+   ylab = "Cumulative Incidence (%)",
+R+   fun = "event",
+R+   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
+R+   main = sprintf(
+R+     "Estimated Cumulative Incidence (risk) of Death from Prostate 
+R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+R+     exp(coef(cox_model)),
+R+     exp(confint(cox_model))[1],
+R+     exp(confint(cox_model))[2]
+R+   ),
+R+   ylim = c(0, 2), yscale = 100
+R+ )
 R> lines(smooth_risk_splines[, 1], smooth_risk_splines[, 2], col = "red", lty = 2)
@@
 R> legend("topleft",
-R>   legend = c(
-R>     "Control group (Cox)", "Control group (Casebase)",
-R>     "Screening group (Cox)", "Screening group (Casebase)"
-R>   ),
-R>   col = c("red", "red", "blue", "blue"),
-R>   lty = c(1, 2, 1, 2),
-R>   bg = "gray90"
-R> )
-\end{CodeInput}
+R+   legend = c(
+R+     "Control group (Cox)", "Control group (Casebase)",
+R+     "Screening group (Cox)", "Screening group (Casebase)"
+R+   ),
+R+   col = c("red", "red", "blue", "blue"),
+R+   lty = c(1, 2, 1, 2),
+R+   bg = "gray90"
+R+ )
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/erspc-casebase-splines-cif-1} \end{center}
+
 \end{CodeChunk}
@@
 R> c(
-R>   "Exp." = AIC(casebase_exponential),
-R>   "Weib." = AIC(casebase_time),
-R>   "Splines" = AIC(casebase_splines)
-R> )
-\end{CodeInput}
+R+   "Exp." = AIC(casebase_exponential),
+R+   "Weib." = AIC(casebase_time),
+R+   "Splines" = AIC(casebase_splines)
+R+ )
+\end{CodeInput}
+
+\begin{CodeOutput}
+#>    Exp.   Weib. Splines 
+#>    6057    5817    5800
+\end{CodeOutput}
 \end{CodeChunk}
@@
 R> model_cb <- fitSmoothHazard(
-R>   Status ~ ftime + Sex + D + Phase + Source + Age,
-R>   data = bmtcrr,
-R>   ratio = 100,
-R>   time = "ftime"
-R> )
+R+   Status ~ ftime + Sex + D + Phase + Source + Age,
+R+   data = bmtcrr,
+R+   ratio = 100,
+R+   time = "ftime"
+R+ )
 \end{CodeInput}
@@
 R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-R>   data = bmtcrr
-R> )
-\end{CodeInput}
+R+   data = bmtcrr
+R+ )
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{table}
+
+\caption{\label{tab:bmtcrr-cis}Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.}
+\centering
+\begin{tabular}[t]{lrlrl}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
+\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
+Covariates & HR & 95\% CI & HR & 95\% CI\\
+\midrule
+Sex & 0.71 & (0.41, 1.24) & 0.68 & (0.39, 1.19)\\
+Disease & 0.55 & (0.31, 1) & 0.52 & (0.29, 0.93)\\
+Phase (CR2 vs. CR1) & 1.18 & (0.47, 2.93) & 1.21 & (0.48, 3.02)\\
+Phase (CR3 vs. CR1) & 1.47 & (0.38, 5.63) & 1.67 & (0.43, 6.5)\\
+Phase (Relapse vs. CR1) & 4.30 & (2, 9.27) & 4.55 & (2.09, 9.9)\\
+\addlinespace
+Source & 1.56 & (0.52, 4.7) & 1.46 & (0.47, 4.54)\\
+Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
+\bottomrule
+\end{tabular}
+\end{table}
+
 \end{CodeChunk}
@@
 R> newdata <- data.frame(
-R>   "Sex" = factor(c("F", "F"),
-R>     levels = levels(bmtcrr[, "Sex"])
-R>   ),
-R>   "D" = c("ALL", "AML"), # Both diseases
-R>   "Phase" = factor(c("Relapse", "Relapse"),
-R>     levels = levels(bmtcrr[, "Phase"])
-R>   ),
-R>   "Age" = c(35, 35),
-R>   "Source" = factor(c("PB", "PB"),
-R>     levels = levels(bmtcrr[, "Source"])
-R>   )
-R> )
+R+   "Sex" = factor(c("F", "F"),
+R+     levels = levels(bmtcrr[, "Sex"])
+R+   ),
+R+   "D" = c("ALL", "AML"), # Both diseases
+R+   "Phase" = factor(c("Relapse", "Relapse"),
+R+     levels = levels(bmtcrr[, "Phase"])
+R+   ),
+R+   "Age" = c(35, 35),
+R+   "Source" = factor(c("PB", "PB"),
+R+     levels = levels(bmtcrr[, "Source"])
+R+   )
+R+ )
 R> 
@@
 R> risk_cb <- absoluteRisk(
-R>   object = model_cb, time = time_points,
-R>   method = "numerical", newdata = newdata
-R> )
+R+   object = model_cb, time = time_points,
+R+   method = "numerical", newdata = newdata
+R+ )
 \end{CodeInput}
@@
 R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) +
-R>   const(Phase) + const(Source) + const(Age),
-R> data = bmtcrr, cause = 1, model = "fg"
-R> )
+R+   const(Phase) + const(Source) + const(Age),
+R+ data = bmtcrr, cause = 1, model = "fg"
+R+ )
 R> 
@@
 \end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/bmtcrr-risk-1}
+
+}
+
+\caption{\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups.
+The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:bmtcrr-risk}
+\end{figure}
 \end{CodeChunk}
@@
 cumulative incidence curves. In the next two case studies, we explore
-more advanced features.
-
-For the third case study, we will look at regularized estimation for the
-hazard function. This can be performed using a combination of case-base
-sampling and regularized logistic regression. To demonstrate this
-capability, we examine the Study to Understand Prognoses Preferences
+advanced features.
+
+For the third case study, we will demonstrate regularization during
+hazard estimation. Since we are using logistic regression, existing
+packages, like \pkg{glmnet} can be leveraged. To demonstrate this
+feature, we examine the Study to Understand Prognoses Preferences
 Outcomes and Risks of Treatment (SUPPORT) dataset
@@
 United States of America. Imputation was conducted using the \pkg{mice}
-package in \proglang{R} with default settings \citep{mice}. Before
-imputation, there were a total of 9105 individuals and 47 variables. Of
-those variables, a few were removed before imputation. We removed three
-response variables: hospital charges, patient ratio of costs to charges
-and patient micro-costs. Next, we removed hospital death as it was
-directly informative of our event of interest, death. We also removed
-functional disability and income as they are ordinal covariates, because
-ordinal variables are treated as factors in our current model. Finally,
-we removed 8 covariates relating to the results of previous findings: we
-removed SUPPORT day 3 physiology score (sps), APACHE III day 3
-physiology score (aps), SUPPORT model 2-month survival estimate, SUPPORT
-model 6-month survival estimate, Physician's 2-month survival estimate
-for pt., Physician's 6-month survival estimate for pt., Patient had Do
-Not Resuscitate (DNR) order and Day of DNR order (\textless{}0 if before
-study). Of these, sps and aps were added on after imputation, as they
-were missing only 1 observation and permitted us to compare findings
-between our model and their's.
-
-First we impute manually using recommended normal values for
-physiological measures \citep{harrell_2020}. Next, we impute a single
-dataset using \pkg{mice} with default settings. After imputation, we
-noted that the covariate for surrogate activities of daily living was
-not imputed. This is due to colinearity between the other two covariates
-for activities of daily living. As such, surrogate activites of daily
-living was removed from our model. After these imputations, we have a
-total of 9104 observations and 30 variables, two of which are sps and
-aps, which will work as a baseline for our comparison to previous
-findings. This dataset is available Its worth noting that these scores
-were developed for a period of 180 days, which makes our comparison over
-5.56 years exploratory. Also, our single imputation introduces bias in
-the data. As the goal here is purely demonstrative, this bias will be
-consistent with any of the following models. As such, their comparison
-is still valid, though any interpretation requires an awareness of said
-bias. A description of used variables can be found in Table
-\ref{tab:support1} and a breakdown of each categorical variable in Table
-\ref{tab:support2}. The original data was obtained from the Department
-of Biostatistics at Vanderbilt University \citep{harrell_2020}. The
-imputed dataset is available as part of the \pkg{casebase} package.
+package in \proglang{R} with default settings \citep{mice}. The imputed
+dataset is available as part of the \pkg{casebase} package. For more
+information please see the SUPPORT data documentation in the package.
+
+we have a total of 9104 observations and 30 variables, two of which are
+SUPPORT day 3 physiology score (sps) and APACHE III day 3 physiology
+score (aps). These scores were developed for a period of 180 days, which
+makes our comparison over 5.56 years exploratory. Therefore, we will
+assess these two models before using either as our baseline, by
+comparing them to a full model using all variables. Also, our single
+imputation introduces bias in the data. As the goal here is purely
+demonstrative, this bias will be consistent with any of the following
+models. As such, their comparison is still valid, though any
+interpretation requires an awareness of said bias. A description of used
+variables can be found in Table \ref{tab:support1} and a breakdown of
+each categorical variable in Table \ref{tab:support2}. The original data
+was obtained from the Department of Biostatistics at Vanderbilt
+University \citep{harrell_2020}.
 
@@
 
-As before, the first step is to visualize the incidence density. The
-population-time plot for the SUPPORT data appears in Figure
-\ref{fig:support-poptime}. As we can see, the incidence density is not
-constant.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/support-poptime-1}
-
-}
-
-\caption{\label{fig:support-poptime} Population-time plot for the SUPPORT data.}\label{fig:support-poptime}
-\end{figure}
-\end{CodeChunk}
-
 Before generating an absolute risk curve using regularized logistic
@@
 the brier scores as they change over time. As the full model has the
-lowest brier score on overall, it will be used in the following
-comparison to regularized cox, regularized logistic regression and the
-unadjusted Kaplan-Meier curve.
+lowest brier score on overall, it will be used in the following absolute
+risk comparison to regularized cox, regularized logistic regression and
+the unadjusted Kaplan-Meier curve.
 
@@
 
-\caption{\label{fig:baselines} AbsoluteRisk curves from hazard models.
-Pink: All covariates without SPS nor APS.
-Green: APS only model.
-Blue: SPS only model.
-Purple: unadjusted curve.
-All three models underestimate the risk in comparison to the unadjusted curve.}\label{fig:baselines}
+\caption{\label{fig:baselines} Absolute risk curves estimated from hazard models.
+Red: APS only hazard was used to generate the absolute risk.
+Green: All covariates aside from APS and SPS.
+Blue: SPS only.
+Purple: unadjusted Kaplan-Meier curve.}\label{fig:baselines}
 \end{figure}
@@
 
-\caption{\label{fig:brier1} Brier scores over time. Full model (Cox Full) performs best, while the other two (APS only and SPS only) are comparable.}\label{fig:brierpt1}
+\caption{\label{fig:brier1} Brier scores over survival time.
+Full model (Full) performs best, while the other two (APS only and SPS only) are comparable.
+Red: APS only model was used to generate the brier scores.
+Green: all covariates excluding APS and SPS.
+Blue: SPS only model.}\label{fig:brierpt1}
 \end{figure}
@@
 
-To get the absolute risk following regularized cox regression, this
-required that we work around limitations in existing packages. To make
-use of the \pkg{survival} package, we first get our hazard coefficients
-using \pkg{glmnet}'s coxnet implementation for lasso penalization
+To get the absolute risk following regularized cox regression, we had to
+work around limitations in existing packages. To make use of the
+\pkg{survival} package, we first get our hazard coefficients using
+\pkg{glmnet}'s coxnet implementation for lasso penalization
 \citep{regpathcox}. Next, we fit a cox model using the survival package
@@
 
-For our penalized logistic regression model, we opted for a logistic
-function of time, resulting in a Weibull distribution assumption. As
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # hazard for cox using glmnet
+R> coxNet <- glmnet::cv.glmnet(x = x, y = y, family = "cox", alpha = 1, standardize = TRUE)
+R> #getting all non-zero covariates
+R> nonzero_covariate_cox <- predict(coxNet, type = "nonzero", s = "lambda.1se")
+R> #keeping coefficients for later
+R> nonzero_coef_cox <- coef(coxNet, s = "lambda.1se")
+R> # cutting dataset to only include nonzero covariates
+R> cleanCoxData <- as.data.frame(cbind(y, x[, nonzero_covariate_cox]))
+R> #create a skeleton object
+R> coxNet <- survival::coxph(Surv(time = d.time, event = death) ~ ., data = cleanCoxData,x=TRUE)
+R> #replace coefficients with regularized ones
+R> coxNet$coefficients <- nonzero_coef_cox@x
+\end{CodeInput}
+\end{CodeChunk}
+
+For our penalized logistic regression model, we opted for the natural
+log of time, which assumes a Weibull distribution for our event. As
 casebase makes use of the glmnet package, we interact with the
@@
 penalization by setting alpha to 1, leading to our smooth hazard
-function. We then take our hazard and use it to predict the absolute
-risk of each individual in our test set, over time. The probabilites
-over time for each individual are averaged, resulting in the smooth
-absolute risk curve in figure \ref{fig:cs3Final}. On average, the Brier
-score is lower in our penalized logistic regression model, compared to
-the other models (fig.~\ref{fig:cs3FinalBrier}).
-
-This case study serves to demonstrate how regularization can be used in
-a survival context, using logistic regression. Our approach performs
-comparably to Cox regression, demonstrating its validity in an applied
-setting.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> 
-R> cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = 10, standardize = TRUE,penalty.facto=c(0,rep(1,ncol(x))))
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
+function.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> #regularized logistic regression to estimate hazard
+R> cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet",
+R>                       time = "d.time", event = "death",
+R>                       formula_time = ~ log(d.time), alpha = 1,
+R>                       ratio = 100, standardize = TRUE, penalty.factor=c(0,rep(1,ncol(x))))
+R> 
 R> # Estimating the absolute risk curve using the newData parameter.
 R> abCbFull <- casebase::absoluteRisk(cbFull, time = times, newdata = newData, s = "lambda.1se", method = c("numerical"))
+R> 
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> 
+R> cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = 10, standardize = TRUE,penalty.factor=c(0,rep(1,ncol(x))))
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Estimating the absolute risk curve using the newData parameter.
+R> abCbFull <- casebase::absoluteRisk(cbFull, time = times, newdata = newData, s = "lambda.1se", method = c("numerical"))
 \end{CodeInput}
@@
 
+{\centering \includegraphics{../figures/coefplots-1}
+
+}
+
+\caption{\label{fig:cs3lolliPlot} A lollipop plot demonstrating the estimated coefficients between the following three models: Cox (green), penalized Cox (red), penalized logisitic (blue).
+If a coefficient is estimated to be 0, it is removed from the plot}\label{fig:coefplots}
+\end{figure}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{figure}
+
 {\centering \includegraphics{../figures/unnamed-chunk-2-1}
@@
 
-\caption{\label{fig:cs3Final} Four absolute risk curves, integrated from the following hazard models: Cox (pink), penalized Cox (green), penalized logistic (blue) and Kaplan-Meier (purple).}\label{fig:unnamed-chunk-2}
+\caption{\label{fig:cs3Final} Four absolute risk curves, integrated from the following hazard models: Cox (green), penalized Cox (red), penalized logistic (blue) and unadjusted (Kaplan-Meier, purple).}\label{fig:unnamed-chunk-2}
 \end{figure}
@@
 
-\caption{\label{fig:cs3FinalBrier} 4 curves comparing the brier scores as they change over time between a null model (Kaplan-Meier, black), Cox regression (yellow), penalized Cox regression (blue) and penalized logistic regression (green).}\label{fig:riskregressionBrier}
+\caption{\label{fig:cs3FinalBrier} 4 curves comparing the brier scores as they change over time between Cox regression (green), penalized Cox regression (red) and penalized logistic regression (blue) and a unadjusted curve (Kaplan-Meier, purple).
+Penalized logistic regression produced the lowest curve.}\label{fig:riskregressionBrier}
 \end{figure}
@@
 
-\begin{CodeChunk}
-
-\begin{center}\includegraphics{../figures/coefplots-1} \end{center}
-
-\end{CodeChunk}
+Our first comparison plot examines the covariates that were selected
+between all three hazard functions. Note that Cox regression without
+penalization selected more covariates than the other two, but only the
+intersection of cofficients with values greater than 0 for all models
+are displayed in \ref{fig:cs3lolliPlot}. A visible tendency is for Cox
+regression to have larger estimated coefficients. This is expected, as
+there are no restrictions on the size of coefficients. The smallest
+coefficient estimates are found in penalized logistic model, aside from
+cancer state (cano). These values do not inform us of model performance.
+To do so, we take our hazard functions and use them to predict the
+absolute risk of each individual in our test set, over time. The
+probabilites over time for each individual are averaged, resulting in
+the absolute risk curves in figure \ref{fig:cs3Final}. There is
+separation between the curves, but it is minimal. To better demonstrate
+the differences between the models, we compare brier scores over time.
+\ref{fig:cs3FinalBrier} the Brier score is lower in our penalized
+logistic regression model, compared to the other models. This
+difference, however, is minimal. We can say with relative certainty that
+any of the hazard models outperform an unadjusted curve, while penalized
+logistic regression performs the best on this testing dataset.
 
@@
 R> stanford_popTime <- popTime(jasa,
-R>   time = "futime",
-R>   event = "fustat"
-R> )
+R+   time = "futime",
+R+   event = "fustat"
+R+ )
 R> plot(stanford_popTime)
 \end{CodeInput}
+
+\begin{center}\includegraphics{../figures/stanford-poptime-1} \end{center}
+
 \end{CodeChunk}
@@
 R> cb_data <- sampleCaseBase(jasa,
-R>   time = "futime",
-R>   event = "fustat", ratio = 10
-R> )
+R+   time = "futime",
+R+   event = "fustat", ratio = 10
+R+ )
 \end{CodeInput}
@@
 R> cb_data <- mutate(cb_data,
-R>   txtime = time_length(accept.dt %--% tx.date,
-R>     unit = "days"
-R>   ),
-R>   exposure = case_when(
-R>     is.na(txtime) ~ 0L,
-R>     txtime > futime ~ 0L,
-R>     txtime <= futime ~ 1L
-R>   )
-R> )
+R+   txtime = time_length(accept.dt %--% tx.date,
+R+     unit = "days"
+R+   ),
+R+   exposure = case_when(
+R+     is.na(txtime) ~ 0L,
+R+     txtime > futime ~ 0L,
+R+     txtime <= futime ~ 1L
+R+   )
+R+ )
 \end{CodeInput}
@@
 R> fit1 <- fitSmoothHazard(fustat ~ exposure,
-R>   data = cb_data, time = "futime"
-R> )
+R+   data = cb_data, time = "futime"
+R+ )
 R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
-R>   data = cb_data, time = "futime"
-R> )
+R+   data = cb_data, time = "futime"
+R+ )
 R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
-R>   data = cb_data, time = "futime"
-R> )
+R+   data = cb_data, time = "futime"
+R+ )
 R> fit4 <- fitSmoothHazard(fustat ~ exposure * bs(futime),
-R>   data = cb_data, time = "futime"
-R> )
+R+   data = cb_data, time = "futime"
+R+ )
 \end{CodeInput}
@@
 R> c(
-R>   "Model1" = AIC(fit1),
-R>   "Model2" = AIC(fit2),
-R>   "Model3" = AIC(fit3),
-R>   "Model4" = AIC(fit4)
-R> )
-\end{CodeInput}
+R+   "Model1" = AIC(fit1),
+R+   "Model2" = AIC(fit2),
+R+   "Model3" = AIC(fit3),
+R+   "Model4" = AIC(fit4)
+R+ )
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> Model1 Model2 Model3 Model4 
+#>    790    752    737    721
+\end{CodeOutput}
 \end{CodeChunk}
@@
 R> hazard_data <- expand.grid(
-R>   exposure = c(0, 1),
-R>   futime = seq(0, 1000,
-R>     length.out = 100
-R>   )
-R> )
+R+   exposure = c(0, 1),
+R+   futime = seq(0, 1000,
+R+     length.out = 100
+R+   )
+R+ )
 R> # Set the offset to zero
@@
 R> hazard_data$hazard <- exp(predict(fit4,
-R>   newdata = hazard_data,
-R>   type = "link"
-R> ))
+R+   newdata = hazard_data,
+R+   type = "link"
+R+ ))
 R> # Add labels for plots
 R> hazard_data$Status <- factor(hazard_data$exposure,
-R>   labels = c("NoTrans", "Trans")
-R> )
+R+   labels = c("NoTrans", "Trans")
+R+ )
 R> 
 R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
-R>   geom_line() +
-R>   theme_minimal() +
-R>   theme(legend.position = "top") +
-R>   ylab("Hazard") +
-R>   xlab("Follow-up time")
-\end{CodeInput}
+R+   geom_line() +
+R+   theme_minimal() +
+R+   theme(legend.position = "top") +
+R+   ylab("Hazard") +
+R+   xlab("Follow-up time")
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/stanford-hazard-1} \end{center}
+
 \end{CodeChunk}
@@
 R> absrisk <- absoluteRisk(fit4,
-R>   newdata = newdata,
-R>   time = seq(0, 1000, length.out = 100)
-R> )
+R+   newdata = newdata,
+R+   time = seq(0, 1000, length.out = 100)
+R+ )
 R> 
@@
 R> absrisk <- gather(
-R>   as.data.frame(absrisk),
-R>   "Status", "Risk", -Time
-R> )
+R+   as.data.frame(absrisk),
+R+   "Status", "Risk", -Time
+R+ )
 R> 
 R> ggplot(absrisk, aes(Time, Risk, colour = Status)) +
-R>   geom_line() +
-R>   theme_minimal() +
-R>   theme(legend.position = "top") +
-R>   expand_limits(y = 0.5) +
-R>   xlab("Follow-up time") +
-R>   ylab("Cum. Incidence")
-\end{CodeInput}
+R+   geom_line() +
+R+   theme_minimal() +
+R+   theme(legend.position = "top") +
+R+   expand_limits(y = 0.5) +
+R+   xlab("Follow-up time") +
+R+   ylab("Cum. Incidence")
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/stanford-risk-1} \end{center}
+
 \end{CodeChunk}
@@
 R> one_yr_haz <- data.frame(futime = seq(0, 365,
-R>   length.out = 100
-R> ))
+R+   length.out = 100
+R+ ))
 R> one_yr_haz <- mutate(one_yr_haz,
-R>   offset = 0,
-R>   exposure = if_else(futime < 30, 0, 1)
-R> )
+R+   offset = 0,
+R+   exposure = if_else(futime < 30, 0, 1)
+R+ )
 R> one_yr_haz$hazard <- exp(predict(fit4,
-R>   newdata = one_yr_haz,
-R>   type = "link"
-R> ))
+R+   newdata = one_yr_haz,
+R+   type = "link"
+R+ ))
 R> ggplot(one_yr_haz, aes(futime, hazard)) +
-R>   geom_line() +
-R>   theme_minimal() +
-R>   theme(legend.position = "top") +
-R>   ylab("Hazard") +
-R>   xlab("Follow-up time") +
-R>   geom_vline(xintercept = 30, linetype = "dashed")
-\end{CodeInput}
+R+   geom_line() +
+R+   theme_minimal() +
+R+   theme(legend.position = "top") +
+R+   ylab("Hazard") +
+R+   xlab("Follow-up time") +
+R+   geom_vline(xintercept = 30, linetype = "dashed")
+\end{CodeInput}
+
+\begin{center}\includegraphics{../figures/stanford-1y-haz-1} \end{center}
+
 \end{CodeChunk}
@@
 R> absoluteRisk(fit4,
-R>   newdata = data.frame(exposure = 0),
-R>   time = 365
-R> )
+R+   newdata = data.frame(exposure = 0),
+R+   time = 365
+R+ )
+\end{CodeInput}
+
+\begin{CodeOutput}
+#>         
+#> 365 0.31
+\end{CodeOutput}
+
+\begin{CodeInput}
 R> 
@@
 \end{CodeInput}
+
+\begin{CodeOutput}
+#> [1] 0.2
+\end{CodeOutput}
 \end{CodeChunk}
@@
 
-This report was generated on 2020-03-31 18:19:20 using the following
+This report was generated on 2020-04-16 12:08:46 using the following
 computational environment and dependencies:
@@
 #> other attached packages:
-#>  [1] broom_0.5.5               dotwhisker_0.5.0         
-#>  [3] prodlim_2019.11.13        riskRegression_2020.02.05
-#>  [5] pracma_2.2.9              kableExtra_1.1.0         
-#>  [7] data.table_1.12.8         glmnet_3.0-2             
-#>  [9] Matrix_1.2-17             casebase_0.2.1.9001      
-#> [11] survival_3.1-8            magrittr_1.5             
-#> [13] forcats_0.5.0             stringr_1.4.0            
-#> [15] dplyr_0.8.3               purrr_0.3.3              
-#> [17] readr_1.3.1               tidyr_1.0.2              
-#> [19] tibble_2.1.3              ggplot2_3.3.0            
-#> [21] tidyverse_1.3.0          
+#>  [1] lubridate_1.7.8           broom_0.5.5              
+#>  [3] dotwhisker_0.5.0          timereg_1.9.4            
+#>  [5] glue_1.4.0                cowplot_1.0.0            
+#>  [7] colorspace_1.4-1          prodlim_2019.11.13       
+#>  [9] riskRegression_2020.03.27 pracma_2.2.9             
+#> [11] kableExtra_1.1.0          data.table_1.12.8        
+#> [13] glmnet_3.0-2              Matrix_1.2-18            
+#> [15] casebase_0.2.1.9001       survival_3.1-12          
+#> [17] magrittr_1.5              forcats_0.5.0            
+#> [19] stringr_1.4.0             dplyr_0.8.5              
+#> [21] purrr_0.3.3               readr_1.3.1              
+#> [23] tidyr_1.0.2               tibble_3.0.0             
+#> [25] ggplot2_3.3.0             tidyverse_1.3.0          
 #> 
 #> loaded via a namespace (and not attached):
-#>  [1] nlme_3.1-140        fs_1.3.2            cmprsk_2.2-9       
-#>  [4] lubridate_1.7.4     rms_5.1-4           webshot_0.5.2      
-#>  [7] RColorBrewer_1.1-2  httr_1.4.1          numDeriv_2016.8-1.1
-#> [10] tools_3.6.1         backports_1.1.5     R6_2.4.1           
-#> [13] rpart_4.1-15        Hmisc_4.3-1         DBI_1.1.0          
-#> [16] mgcv_1.8-28         colorspace_1.4-1    nnet_7.3-13        
-#> [19] withr_2.1.2         gridExtra_2.3       tidyselect_1.0.0   
-#> [22] compiler_3.6.1      quantreg_5.54       cli_2.0.2          
-#> [25] rvest_0.3.5         htmlTable_1.13.3    SparseM_1.78       
-#> [28] xml2_1.2.5          sandwich_2.5-1      labeling_0.3       
-#> [31] checkmate_2.0.0     scales_1.1.0        mvtnorm_1.1-0      
-#> [34] polspline_1.1.17    digest_0.6.25       foreign_0.8-71     
-#> [37] rmarkdown_2.1       base64enc_0.1-3     rticles_0.14       
-#> [40] jpeg_0.1-8.1        pkgconfig_2.0.3     htmltools_0.4.0    
-#> [43] dbplyr_1.4.2        htmlwidgets_1.5.1   rlang_0.4.5        
-#> [46] readxl_1.3.1        rstudioapi_0.11     VGAM_1.1-2         
-#> [49] farver_2.0.3        shape_1.4.4         generics_0.0.2     
-#> [52] zoo_1.8-7           jsonlite_1.6.1      acepack_1.4.1      
-#> [55] Formula_1.2-3       Rcpp_1.0.4          munsell_0.5.0      
-#> [58] fansi_0.4.1         lifecycle_0.2.0     multcomp_1.4-12    
-#> [61] stringi_1.4.6       yaml_2.2.1          MASS_7.3-51.4      
-#> [64] plyr_1.8.6          ggstance_0.3.3      grid_3.6.1         
-#> [67] crayon_1.3.4        lattice_0.20-38     haven_2.2.0        
-#> [70] hms_0.5.3           timereg_1.9.4       knitr_1.28         
-#> [73] pillar_1.4.3        codetools_0.2-16    stats4_3.6.1       
-#> [76] reprex_0.3.0        glue_1.3.1          evaluate_0.14      
-#> [79] latticeExtra_0.6-29 modelr_0.1.6        vctrs_0.2.4        
-#> [82] png_0.1-7           foreach_1.4.8       MatrixModels_0.4-1 
-#> [85] cellranger_1.1.0    gtable_0.3.0        assertthat_0.2.1   
-#> [88] xfun_0.12           viridisLite_0.3.0   iterators_1.0.12   
-#> [91] cluster_2.1.0       lava_1.6.7          TH.data_1.0-10
+#>  [1] TH.data_1.0-10      VGAM_1.1-2          ellipsis_0.3.0     
+#>  [4] ggstance_0.3.4      htmlTable_1.13.3    base64enc_0.1-3    
+#>  [7] fs_1.4.1            rstudioapi_0.11     farver_2.0.3       
+#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-0      
+#> [13] xml2_1.3.1          codetools_0.2-16    knitr_1.28         
+#> [16] Formula_1.2-3       jsonlite_1.6.1      cluster_2.1.0      
+#> [19] dbplyr_1.4.2        png_0.1-7           compiler_3.6.3     
+#> [22] httr_1.4.1          backports_1.1.6     assertthat_0.2.1   
+#> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.4.0    
+#> [28] quantreg_5.55       tools_3.6.3         gtable_0.3.0       
+#> [31] Rcpp_1.0.4.6        cellranger_1.1.0    vctrs_0.2.4        
+#> [34] nlme_3.1-145        iterators_1.0.12    xfun_0.13          
+#> [37] rvest_0.3.5         lifecycle_0.2.0     polspline_1.1.17   
+#> [40] MASS_7.3-51.5       zoo_1.8-7           scales_1.1.0       
+#> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
+#> [46] RColorBrewer_1.1-2  rticles_0.14        yaml_2.2.1         
+#> [49] gridExtra_2.3       rms_5.1-4           rpart_4.1-15       
+#> [52] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
+#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
+#> [58] mets_1.2.7.1        rlang_0.4.5         pkgconfig_2.0.3    
+#> [61] evaluate_0.14       lattice_0.20-41     htmlwidgets_1.5.1  
+#> [64] labeling_0.3        cmprsk_2.2-9        tidyselect_1.0.0   
+#> [67] plyr_1.8.6          R6_2.4.1            generics_0.0.2     
+#> [70] Hmisc_4.4-0         multcomp_1.4-13     DBI_1.1.0          
+#> [73] pillar_1.4.3        haven_2.2.0         foreign_0.8-76     
+#> [76] withr_2.1.2         mgcv_1.8-31         nnet_7.3-13        
+#> [79] modelr_0.1.6        crayon_1.3.4        rmarkdown_2.1      
+#> [82] jpeg_0.1-8.1        grid_3.6.3          readxl_1.3.1       
+#> [85] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
+#> [88] numDeriv_2016.8-1.1 stats4_3.6.3        munsell_0.5.0      
+#> [91] viridisLite_0.3.0
 \end{CodeOutput}
@@
 #> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
-#> Head:     [5f22217] 2020-03-26: AddedBrier
+#> Head:     [7755d8f] 2020-04-09: Uses development version of Jesse-Islam/riskRegression
 \end{CodeOutput}

### 2020-04-22 analysis/paper/paper.Rmd
last edits for draft
Fixed code formatting, typos, plot colors.

@@
 In the first two case studies, we described the basic functionalities of the \pkg{casebase} package: creating population-time plots, fitting parametric models for hazard functions, and estimating the corresponding cumulative incidence curves.
-In the next two case studies, we explore advanced features.
+In the next case study, we explore an advanced feature.
 
@@
 As such, their comparison is still valid, though any interpretation requires an awareness of said bias.
-<!--The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.--> A description of used variables can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}.
+<!--The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.--> A description of covariates and response variables can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}.
 The original data was obtained from the Department of Biostatistics at Vanderbilt University [@harrell_2020].
@@
 
-Before generating an absolute risk curve using regularized logistic regression, we will explore three options as a baseline comparison, beyond a Kaplan-meier curve.
+Before generating an absolute risk curve using regularized logistic regression, we will explore three options as a baseline comparison.
 The original study compared these scores over a span of 180 days, which may not be comparable over the full 5.56 years.
@@
 Next, we will compare each modeling using the Brier score [@graf1999ass].
-The Brier score is able to demonstrate both discrimination and calibration in a single model [@graf1999ass] and provide a comparable metric between semi-parametric and parametric methods in the survival context.
+The Brier score can demonstrate both discrimination and calibration in a single model [@graf1999ass] and provide a comparable metric between semi-parametric and parametric methods in the survival context.
 For all the following analyses in this case study, we will use 80% of the observations to train our models, while 20% will be used to generate absolute risk curves, all of which will be averaged to demonstrate each model's predictive ability.
@@
 
-completeData$d.time<-completeData$d.time/365
+completeData$d.time <- completeData$d.time / 365
 
@@
 fullDataCox <- as.data.frame(cbind(as.numeric(y[, 2]), as.numeric(y[, 1]), x))
-names(fullDataCox)[names(fullDataCox) == 'V1'] <- 'time'
-names(fullDataCox)[names(fullDataCox) == 'V2'] <- 'event'
-
-#here we specify train and test size
+names(fullDataCox)[names(fullDataCox) == "V1"] <- "time"
+names(fullDataCox)[names(fullDataCox) == "V2"] <- "event"
+
+# here we specify train and test size
 train_index <- sample(1:nrow(fullDataCox), 0.8 * nrow(fullDataCox))
@@
 test <- fullDataCox[test_index, ]
-coxSPS <- survival::coxph(Surv(time =time , event = event) ~ sps, data = train, x = TRUE)
+coxSPS <- survival::coxph(Surv(time = time, event = event) ~ sps, data = train, x = TRUE)
 abcoxSPS <- survival::survfit(coxSPS, newdata = test)
 
-coxAPS <- survival::coxph(Surv(time =time , event = event) ~ aps, data = train, x = TRUE)
+coxAPS <- survival::coxph(Surv(time = time, event = event) ~ aps, data = train, x = TRUE)
 abcoxAPS <- survival::survfit(coxAPS, newdata = test)
-coxfullForumla<-as.formula(paste("Surv(time = time, event = event) ~ age + sexmale + slos + dzgroupCHF + 
+coxfullForumla <- as.formula(paste("Surv(time = time, event = event) ~ age + sexmale + slos + dzgroupCHF + 
     dzgroupCirrhosis + `dzgroupColon Cancer` + dzgroupComa + 
@@
 
-colnames(baseCoxComparisonData) <- c("Time", "SPS", "APS","Full")
+colnames(baseCoxComparisonData) <- c("Time", "SPS", "APS", "Full")
 ```
@@
   geom_step(data = as.data.frame(abKMcompare), mapping = aes(x = Time, y = KM, color = "Unadjusted"), lwd = 1) +
-  labs(y = "Probability of Death", x = "Survival time (Years)",color = "Models")+
-   scale_color_manual(values=q4)
+  labs(y = "Probability of Death", x = "Survival time (Years)", color = "Models") +
+  scale_color_manual(values = q4)
 ```
@@
 
-coxFull$coefficients[which(is.na(coxFull$coefficients))]=0
+coxFull$coefficients[which(is.na(coxFull$coefficients))] <- 0
 brierFull <- Score(list("PredictionModel" = coxFull), data = test, formula = Surv(time, event != 0) ~ 1, summary = "ipa", se.fit = 0L, metrics = "brier", contrasts = FALSE, times = times)
 
-brierPlotter<-rbind(
-      data.frame(brierScore = brierAPS$Brier$score$Brier[brierAPS$Brier$score$model == "PredictionModel"],times=times,model="APS"),
-      data.frame(brierScore = brierFull$Brier$score$Brier[brierFull$Brier$score$model == "PredictionModel"],times=times,model="Full"),
-            data.frame(brierScore = brierSPS$Brier$score$Brier[brierSPS$Brier$score$model == "PredictionModel"],times=times,model="SPS")
+brierPlotter <- rbind(
+  data.frame(brierScore = brierAPS$Brier$score$Brier[brierAPS$Brier$score$model == "PredictionModel"], times = times, model = "APS"),
+  data.frame(brierScore = brierFull$Brier$score$Brier[brierFull$Brier$score$model == "PredictionModel"], times = times, model = "Full"),
+  data.frame(brierScore = brierSPS$Brier$score$Brier[brierSPS$Brier$score$model == "PredictionModel"], times = times, model = "SPS")
 )
@@
   xlab("Survival time (Years)") +
-  ylab("Brier-Score")+
-   labs(color = "Models")+
-  ggtitle("Brier scores over time")+
-   scale_color_manual(values=q4[1:3])
+  ylab("Brier-Score") +
+  labs(color = "Models") +
+  ggtitle("Brier scores over time") +
+  scale_color_manual(values = q4[1:3])
 ```
@@
 To get the absolute risk following regularized cox regression, we had to work around limitations in existing packages.
-To make use of the \pkg{survival} package, we first get our hazard coefficients using \pkg{glmnet}'s coxnet implementation for lasso penalization [@regpathcox].
-Next, we fit a cox model using the survival package with the selected variables only, from our previous fit.
+To make use of the \pkg{survival} package, we first estimate the hazard coefficients using thecoxnet implementation for lasso penalization in \pkg{glmnet} [@regpathcox].
+Next, we fit a cox model using the survival package with the selected variables only, from our regularized fit.
 Finally, we replaced the coefficients in the survival object with those from our glmnet fit, allowing us to make use of the absolute risk tools available in the \pkg{survival} package.
@@
 # hazard for cox using glmnet
-coxNet <- glmnet::cv.glmnet(x = x, y = y, family = "cox", alpha = 1, standardize = TRUE)
-#getting all non-zero covariates
-nonzero_covariate_cox <- predict(coxNet, type = "nonzero", s = "lambda.1se")
-#keeping coefficients for later
+coxNet <- glmnet::cv.glmnet(
+  x = x, y = y, family = "cox", alpha = 1,
+  standardize = TRUE
+)
+# getting all non-zero covariates
+nonzero_covariate_cox <- predict(coxNet,
+  type = "nonzero",
+  s = "lambda.1se"
+)
+# keeping coefficients for later
 nonzero_coef_cox <- coef(coxNet, s = "lambda.1se")
@@
 cleanCoxData <- as.data.frame(cbind(y, x[, nonzero_covariate_cox]))
-#create a skeleton object
-coxNet <- survival::coxph(Surv(time = d.time, event = death) ~ ., data = cleanCoxData,x=TRUE)
-#replace coefficients with regularized ones
+# create a skeleton object
+coxNet <- survival::coxph(Surv(time = d.time, event = death) ~ .,
+  data = cleanCoxData, x = TRUE
+)
+# replace coefficients with regularized ones
 coxNet$coefficients <- nonzero_coef_cox@x
@@
 For our penalized logistic regression model, we opted for the natural log of time, which assumes a Weibull distribution for our event.
-As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model.
+As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where *y* contains the time and event variables and *x* contains all covariates we would like to include in the model.
 For the SUPPORT dataset, all factor variables were converted into dummy variables.
@@
 ```{r casebaseDisplayCode, eval=FALSE,echo=TRUE}
-#regularized logistic regression to estimate hazard
-cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet",
+# regularized logistic regression to estimate hazard
+cbFull <- casebase::fitSmoothHazard.fit(x, y,
+  family = "glmnet",
   time = "d.time", event = "death",
   formula_time = ~ log(d.time), alpha = 1,
-                      ratio = 100, standardize = TRUE, penalty.factor=c(0,rep(1,ncol(x))))
+  ratio = 100, standardize = TRUE,
+  penalty.factor = c(0, rep(1, ncol(x)))
+)
 
 # Estimating the absolute risk curve using the newData parameter.
-abCbFull <- casebase::absoluteRisk(cbFull, time = times, newdata = newData, s = "lambda.1se", method = c("numerical"))
-
+abCbFull <- casebase::absoluteRisk(cbFull,
+  time = times,
+  newdata = newData,
+  s = "lambda.1se",
+  method = c("numerical")
+)
 ```
@@
 
-#remove apsSPS
-train<-train[,-c(46,47)]
-test<-test[,-c(46,47)]
+# remove apsSPS
+train <- train[, -c(46, 47)]
+test <- test[, -c(46, 47)]
 # x and y will be used to fit the casebase model under glmnet.
@@
 # this is used more as an object place holder.
-coxNet <- survival::coxph(Surv(time = d.time, event = death) ~ ., data = cleanCoxData,x=TRUE)
+coxNet <- survival::coxph(Surv(time = d.time, event = death) ~ ., data = cleanCoxData, x = TRUE)
 
@@
 
-brierCoxNet <- Score(list("PredictionModel" = coxNet), data = as.data.frame(cbind(test[,c(1,2)],newData[, nonzero_covariate_cox$X1])), formula = Hist(time, event != 0) ~ 1, summary = "ipa", se.fit = 0L, metrics = "brier", contrasts = FALSE, times = times)
+brierCoxNet <- Score(list("PredictionModel" = coxNet), data = as.data.frame(cbind(test[, c(1, 2)], newData[, nonzero_covariate_cox$X1])), formula = Hist(time, event != 0) ~ 1, summary = "ipa", se.fit = 0L, metrics = "brier", contrasts = FALSE, times = times)
 
@@
 
-```{r cbHazard}, echo = TRUE, eval = eval_cs3}
-
-cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = 10, standardize = TRUE,penalty.factor=c(0,rep(1,ncol(x))))
-```
-
-```{r CBabsolute, echo = TRUE, eval = eval_cs3}
+```{r cbHazard}, echo = FALSE, eval = eval_cs3}
+
+cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = 10, standardize = TRUE, penalty.factor = c(0, rep(1, ncol(x))))
+```
+
+```{r CBabsolute, echo = FALSE, eval = eval_cs3}
 # Estimating the absolute risk curve using the newData parameter.
@@
 ```{r coefplots,echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} A lollipop plot demonstrating the estimated coefficients between the following three models: Cox (green), penalized Cox (red), penalized logisitic (blue).
-If a coefficient is estimated to be 0, it is removed from the plot"}
+If a coefficient is estimated to be 0, in any model, it is removed from the plot for simplicity."}
 
@@
 
-lolliplotDots <- rbind(estimatescoxFull, estimatescoxNet,estimatescb)
+lolliplotDots <- rbind(estimatescoxFull, estimatescoxNet, estimatescb)
 lolliplotDots$conf.low <- lolliplotDots$estimate
@@
 lolliplotDots$conf.low[lolliplotDots$estimate > 0] <- 0
-lolliplotDots$term=gsub( '\`', "", as.character(lolliplotDots$term))
-
-lolliplotDots[which(lolliplotDots$estimate==0),c(2,4,5)]<-NA
-unChosen<- unique(lolliplotDots[which(is.na(lolliplotDots$estimate)),1])
-
-lolliplotDots[which(lolliplotDots$term %in% unChosen$term),c(2,4,5)]<-NA
-
-lolliplotDots<-lolliplotDots[-which(is.na(lolliplotDots$estimate)),]
-dwplot(lolliplotDots)+
-  theme(axis.text.y = element_text(size = 11, angle = 0, hjust = 1, vjust = 0),
+lolliplotDots$term <- gsub("\`", "", as.character(lolliplotDots$term))
+
+lolliplotDots[which(lolliplotDots$estimate == 0), c(2, 4, 5)] <- NA
+unChosen <- unique(lolliplotDots[which(is.na(lolliplotDots$estimate)), 1])
+
+lolliplotDots[which(lolliplotDots$term %in% unChosen$term), c(2, 4, 5)] <- NA
+
+lolliplotDots <- lolliplotDots[-which(is.na(lolliplotDots$estimate)), ]
+dwplot(lolliplotDots) +
+  theme(
+    axis.text.y = element_text(size = 11, angle = 0, hjust = 1, vjust = 0),
     strip.text.x = element_blank(),
-      strip.background = element_rect(colour="white", fill="white"),
-      legend.position=c(.855,0.15),legend.text=element_text(size=11)
-      )+
-  ggtitle("Coefficient plot")+
-   labs(color = "Models")+
-   scale_color_manual(values=q4[c(2,1,3)])
+    strip.background = element_rect(colour = "white", fill = "white"),
+    legend.position = c(.855, 0.15), legend.text = element_text(size = 11)
+  ) +
+  ggtitle("Coefficient plot") +
+  labs(color = "Models") +
+  scale_color_manual(values = q4[c(2, 1, 3)])
 ```
@@
   geom_step(data = as.data.frame(abKMcompare), mapping = aes(x = Time, y = KM, color = "Unadjusted"), lwd = 1) +
-  labs(y = "Probability of Death", x = "Survival time (Years)",color = "Models")+
-  ggtitle("Absolute risk curves")+
-   scale_color_manual(values=q4[c(2,1,3,4)])
+  labs(y = "Probability of Death", x = "Survival time (Years)", color = "Models") +
+  ggtitle("Absolute risk curves") +
+  scale_color_manual(values = q4[c(2, 1, 3, 4)])
 ```
@@
 
-class(cbFull) <- c("singleEventCB",class(cbFull))
-brierFinalResults <- Score(list("Cox" = coxFull,"Penalized Cox"=coxNet,"Penalized Logistic"=cbFull), data = test, formula = Hist(time, event != 0) ~ 1, summary = NULL, se.fit = 0L, metrics = "brier", contrasts = FALSE, times =times)
-
-levels(brierFinalResults$Brier$score$model)[levels(brierFinalResults$Brier$score$model)=="Null model"] <- "Unadjusted"
-b=brierFinalResults
-
-brierFinalResults$Brier$score$model<- factor(brierFinalResults$Brier$score$model, levels = c("Cox","Penalized Cox","Penalized Logistic", "Unadjusted"))
+class(cbFull) <- c("singleEventCB", class(cbFull))
+brierFinalResults <- Score(list("Cox" = coxFull, "Penalized Cox" = coxNet, "Penalized Logistic" = cbFull), data = test, formula = Hist(time, event != 0) ~ 1, summary = NULL, se.fit = 0L, metrics = "brier", contrasts = FALSE, times = times)
+
+levels(brierFinalResults$Brier$score$model)[levels(brierFinalResults$Brier$score$model) == "Null model"] <- "Unadjusted"
+b <- brierFinalResults
+
+brierFinalResults$Brier$score$model <- factor(brierFinalResults$Brier$score$model, levels = c("Cox", "Penalized Cox", "Penalized Logistic", "Unadjusted"))
 
@@
   xlab("Survival time (Years)") +
-  ylab("Brier-Score")+
-  ggtitle("Brier scores over time")+
-  labs(color = "Models")+
-  scale_color_manual(values=q4[c(2,1,3,4)])
-
+  ylab("Brier-Score") +
+  ggtitle("Brier scores over time") +
+  labs(color = "Models") +
+  scale_color_manual(values = q4[c(2, 1, 3, 4)])
 ```
@@
 Our first comparison plot examines the covariates that were selected between all three hazard functions.
-Note that Cox regression without penalization selected more covariates than the other two, but only the intersection of cofficients with values greater than 0 for all models are displayed in \ref{fig:cs3lolliPlot}.
+Note that Cox regression without penalization selected more covariates than the other two, but only the intersection of coefficients with values greater than 0 are displayed in \ref{fig:cs3lolliPlot}.
 A visible tendency is for Cox regression to have larger estimated coefficients.
@@
 These values do not inform us of model performance.
-To do so, we take our hazard functions and use them to predict the absolute risk of each individual in our test set, over time.
-The probabilites over time for each individual are averaged, resulting in the absolute risk curves in figure \ref{fig:cs3Final}.
+To do so, we take our hazard functions and use them to predict the absolute risk of everyone in our test set, over time.
+The probabilities over time for each individual are averaged, resulting in the absolute risk curves in figure \ref{fig:cs3Final}.
 There is separation between the curves, but it is minimal.
 To better demonstrate the differences between the models, we compare brier scores over time.
-\ref{fig:cs3FinalBrier} the Brier score is lower in our penalized logistic regression model, compared to the other models.
+\ref{fig:cs3FinalBrier} The Brier score is lower in our penalized logistic regression model, compared to the other models.
 This difference, however, is minimal.

### 2020-04-22 analysis/paper/paper.tex
last edits for draft
Fixed code formatting, typos, plot colors.

@@
 parametric models for hazard functions, and estimating the corresponding
-cumulative incidence curves. In the next two case studies, we explore
-advanced features.
+cumulative incidence curves. In the next case study, we explore an
+advanced feature.
 
@@
 models. As such, their comparison is still valid, though any
-interpretation requires an awareness of said bias. A description of used
-variables can be found in Table \ref{tab:support1} and a breakdown of
-each categorical variable in Table \ref{tab:support2}. The original data
-was obtained from the Department of Biostatistics at Vanderbilt
-University \citep{harrell_2020}.
+interpretation requires an awareness of said bias. A description of
+covariates and response variables can be found in Table
+\ref{tab:support1} and a breakdown of each categorical variable in Table
+\ref{tab:support2}. The original data was obtained from the Department
+of Biostatistics at Vanderbilt University \citep{harrell_2020}.
 
@@
 Before generating an absolute risk curve using regularized logistic
-regression, we will explore three options as a baseline comparison,
-beyond a Kaplan-meier curve. The original study compared these scores
-over a span of 180 days, which may not be comparable over the full 5.56
-years. Therefore, We will examine three cox regression scenarios: sps
-only, aps only and all covariates excluding sps and aps. Next, we will
-compare each modeling using the Brier score \citep{graf1999ass}. The
-Brier score is able to demonstrate both discrimination and calibration
-in a single model \citep{graf1999ass} and provide a comparable metric
-between semi-parametric and parametric methods in the survival context.
-For all the following analyses in this case study, we will use 80\% of
-the observations to train our models, while 20\% will be used to
-generate absolute risk curves, all of which will be averaged to
-demonstrate each model's predictive ability. Over 5.56 years,
-\ref{fig:baselines} demonstrates that all the cox models have similar,
-if not identical average predictions. Figure \ref{fig:brier1} examines
-the brier scores as they change over time. As the full model has the
-lowest brier score on overall, it will be used in the following absolute
-risk comparison to regularized cox, regularized logistic regression and
-the unadjusted Kaplan-Meier curve.
+regression, we will explore three options as a baseline comparison. The
+original study compared these scores over a span of 180 days, which may
+not be comparable over the full 5.56 years. Therefore, We will examine
+three cox regression scenarios: sps only, aps only and all covariates
+excluding sps and aps. Next, we will compare each modeling using the
+Brier score \citep{graf1999ass}. The Brier score can demonstrate both
+discrimination and calibration in a single model \citep{graf1999ass} and
+provide a comparable metric between semi-parametric and parametric
+methods in the survival context. For all the following analyses in this
+case study, we will use 80\% of the observations to train our models,
+while 20\% will be used to generate absolute risk curves, all of which
+will be averaged to demonstrate each model's predictive ability. Over
+5.56 years, \ref{fig:baselines} demonstrates that all the cox models
+have similar, if not identical average predictions. Figure
+\ref{fig:brier1} examines the brier scores as they change over time. As
+the full model has the lowest brier score on overall, it will be used in
+the following absolute risk comparison to regularized cox, regularized
+logistic regression and the unadjusted Kaplan-Meier curve.
 
@@
 work around limitations in existing packages. To make use of the
-\pkg{survival} package, we first get our hazard coefficients using
-\pkg{glmnet}'s coxnet implementation for lasso penalization
+\pkg{survival} package, we first estimate the hazard coefficients using
+thecoxnet implementation for lasso penalization in \pkg{glmnet}
 \citep{regpathcox}. Next, we fit a cox model using the survival package
-with the selected variables only, from our previous fit. Finally, we
+with the selected variables only, from our regularized fit. Finally, we
 replaced the coefficients in the survival object with those from our
@@
 R> # hazard for cox using glmnet
-R> coxNet <- glmnet::cv.glmnet(x = x, y = y, family = "cox", alpha = 1, standardize = TRUE)
-R> #getting all non-zero covariates
-R> nonzero_covariate_cox <- predict(coxNet, type = "nonzero", s = "lambda.1se")
-R> #keeping coefficients for later
+R> coxNet <- glmnet::cv.glmnet(
+R>   x = x, y = y, family = "cox", alpha = 1,
+R>   standardize = TRUE
+R> )
+R> # getting all non-zero covariates
+R> nonzero_covariate_cox <- predict(coxNet,
+R>   type = "nonzero",
+R>   s = "lambda.1se"
+R> )
+R> # keeping coefficients for later
 R> nonzero_coef_cox <- coef(coxNet, s = "lambda.1se")
@@
 R> cleanCoxData <- as.data.frame(cbind(y, x[, nonzero_covariate_cox]))
-R> #create a skeleton object
-R> coxNet <- survival::coxph(Surv(time = d.time, event = death) ~ ., data = cleanCoxData,x=TRUE)
-R> #replace coefficients with regularized ones
+R> # create a skeleton object
+R> coxNet <- survival::coxph(Surv(time = d.time, event = death) ~ .,
+R>   data = cleanCoxData, x = TRUE
+R> )
+R> # replace coefficients with regularized ones
 R> coxNet$coefficients <- nonzero_coef_cox@x
@@
 casebase makes use of the glmnet package, we interact with the
-fitsmoothhazard.fit function using a matrix interface, where y contains
-the time and event variables and x contains all other variables we would
-like to include in the model. For the SUPPORT dataset, all factor
-variables were converted into dummy variables. Here, we used lasso
-penalization by setting alpha to 1, leading to our smooth hazard
-function.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> #regularized logistic regression to estimate hazard
-R> cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet",
+fitsmoothhazard.fit function using a matrix interface, where \emph{y}
+contains the time and event variables and \emph{x} contains all
+covariates we would like to include in the model. For the SUPPORT
+dataset, all factor variables were converted into dummy variables. Here,
+we used lasso penalization by setting alpha to 1, leading to our smooth
+hazard function.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # regularized logistic regression to estimate hazard
+R> cbFull <- casebase::fitSmoothHazard.fit(x, y,
+R>   family = "glmnet",
 R>   time = "d.time", event = "death",
 R>   formula_time = ~ log(d.time), alpha = 1,
-R>                       ratio = 100, standardize = TRUE, penalty.factor=c(0,rep(1,ncol(x))))
+R>   ratio = 100, standardize = TRUE,
+R>   penalty.factor = c(0, rep(1, ncol(x)))
+R> )
 R> 
 R> # Estimating the absolute risk curve using the newData parameter.
-R> abCbFull <- casebase::absoluteRisk(cbFull, time = times, newdata = newData, s = "lambda.1se", method = c("numerical"))
-R> 
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> 
-R> cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = 10, standardize = TRUE,penalty.factor=c(0,rep(1,ncol(x))))
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Estimating the absolute risk curve using the newData parameter.
-R> abCbFull <- casebase::absoluteRisk(cbFull, time = times, newdata = newData, s = "lambda.1se", method = c("numerical"))
+R> abCbFull <- casebase::absoluteRisk(cbFull,
+R>   time = times,
+R>   newdata = newData,
+R>   s = "lambda.1se",
+R>   method = c("numerical")
+R> )
 \end{CodeInput}
@@
 \caption{\label{fig:cs3lolliPlot} A lollipop plot demonstrating the estimated coefficients between the following three models: Cox (green), penalized Cox (red), penalized logisitic (blue).
-If a coefficient is estimated to be 0, it is removed from the plot}\label{fig:coefplots}
+If a coefficient is estimated to be 0, in any model, it is removed from the plot for simplicity.}\label{fig:coefplots}
 \end{figure}
@@
 penalization selected more covariates than the other two, but only the
-intersection of cofficients with values greater than 0 for all models
-are displayed in \ref{fig:cs3lolliPlot}. A visible tendency is for Cox
-regression to have larger estimated coefficients. This is expected, as
-there are no restrictions on the size of coefficients. The smallest
-coefficient estimates are found in penalized logistic model, aside from
-cancer state (cano). These values do not inform us of model performance.
-To do so, we take our hazard functions and use them to predict the
-absolute risk of each individual in our test set, over time. The
-probabilites over time for each individual are averaged, resulting in
-the absolute risk curves in figure \ref{fig:cs3Final}. There is
-separation between the curves, but it is minimal. To better demonstrate
-the differences between the models, we compare brier scores over time.
-\ref{fig:cs3FinalBrier} the Brier score is lower in our penalized
-logistic regression model, compared to the other models. This
-difference, however, is minimal. We can say with relative certainty that
-any of the hazard models outperform an unadjusted curve, while penalized
-logistic regression performs the best on this testing dataset.
+intersection of coefficients with values greater than 0 are displayed in
+\ref{fig:cs3lolliPlot}. A visible tendency is for Cox regression to have
+larger estimated coefficients. This is expected, as there are no
+restrictions on the size of coefficients. The smallest coefficient
+estimates are found in penalized logistic model, aside from cancer state
+(cano). These values do not inform us of model performance. To do so, we
+take our hazard functions and use them to predict the absolute risk of
+everyone in our test set, over time. The probabilities over time for
+each individual are averaged, resulting in the absolute risk curves in
+figure \ref{fig:cs3Final}. There is separation between the curves, but
+it is minimal. To better demonstrate the differences between the models,
+we compare brier scores over time. \ref{fig:cs3FinalBrier} The Brier
+score is lower in our penalized logistic regression model, compared to
+the other models. This difference, however, is minimal. We can say with
+relative certainty that any of the hazard models outperform an
+unadjusted curve, while penalized logistic regression performs the best
+on this testing dataset.
 
@@
 
-This report was generated on 2020-04-16 12:08:46 using the following
+This report was generated on 2020-04-22 10:48:55 using the following
 computational environment and dependencies:
@@
 #> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
-#> Head:     [7755d8f] 2020-04-09: Uses development version of Jesse-Islam/riskRegression
+#> Head:     [d079224] 2020-04-16: Write up and plots fixed. plot descriptions fixed
 \end{CodeOutput}

### 2020-05-05 analysis/paper/paper.Rmd
edit jesse's section

@@
 eval_implementation <- TRUE
-eval_cs1 <- TRUE
-eval_cs2 <- TRUE
+eval_cs1 <- FALSE
+eval_cs2 <- FALSE
 eval_cs3 <- TRUE
-eval_cs4 <- TRUE
+eval_cs4 <- FALSE
 ```
@@
 We assume a continuous time model, which implies that the counting process jumps are less than or equal to one.
-We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
+We are interested in modelling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
 $$\lambda_{i}(t) dt = E[dN_{i}(t)].$$
@@
 $$L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.$$
-When a logarithmic link function is used for modeling the hazard function, the above expression is of a logistic regression form with an offset term $\log(1/\rho_i(t))$.
+When a logarithmic link function is used for modelling the hazard function, the above expression is of a logistic regression form with an offset term $\log(1/\rho_i(t))$.
 Note that the sampling units selected in the case-base sampling mechanism are person-moments, rather than individuals, and the parameters to be estimated are hazards or hazard ratios rather than odds or odds ratios.
@@
 
-# Case study 1--European Randomized Study of Prostate Cancer Screening
+# Case study 1---European Randomized Study of Prostate Cancer Screening
 
@@
 
-# Case study 2--Bone-marrow transplant
+# Case study 2---Bone-marrow transplant
 
@@
 
-# Case study 3--SUPPORT Data
+# Case study 3---SUPPORT Data
 
 In the first two case studies, we described the basic functionalities of the \pkg{casebase} package: creating population-time plots, fitting parametric models for hazard functions, and estimating the corresponding cumulative incidence curves.
-In the next case study, we explore an advanced feature.
-
-For the third case study, we will demonstrate regularization during hazard estimation.
-Since we are using logistic regression, existing packages, like \pkg{glmnet} can be leveraged.
-To demonstrate this feature, we examine the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [@knaus1995support].
-The SUPPORT dataset tracks death for individuals who are considered seriously ill within 5 hospitals in the United States of America.
-Imputation was conducted using the \pkg{mice} package in \proglang{R} with default settings [@mice].
-The imputed dataset is available as part of the \pkg{casebase} package.
-For more information please see the SUPPORT data documentation in the package.
-
-we have a total of 9104 observations and 30 variables, two of which are SUPPORT day 3 physiology score (sps) and APACHE III day 3 physiology score (aps).
-These scores were developed for a period of 180 days, which makes our comparison over 5.56 years exploratory.
-Therefore, we will assess these two models before using either as our baseline, by comparing them to a full model using all variables.
-Also, our single imputation introduces bias in the data.
-As the goal here is purely demonstrative, this bias will be consistent with any of the following models.
-As such, their comparison is still valid, though any interpretation requires an awareness of said bias.
-<!--The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.--> A description of covariates and response variables can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}.
-The original data was obtained from the Department of Biostatistics at Vanderbilt University [@harrell_2020].
+<!--In the next two case studies, we explore more advanced features.--> For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function.
+To do this, we simply replace logistic regression with a penalized counterpart.
+In \pkg{casebase}, we achieve this by using the \pkg{glmnet} package [@friedman2010jss].
+
+To illustrate this functionality, we use the dataset from the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) [@knaus1995support].
+The SUPPORT dataset tracks death in five American hospitals within individuals who are considered seriously ill.
+The original data is available online from the Department of Biostatistics at Vanderbilt University [@harrell_2020].
+The cleaned and imputed data consists of 9104 observations and 30 variables, and it is available as part of the \pkg{casebase} package.
+For more information about this dataset, the reader is encouraged to look at the documentation in our package.
+A description of the variables can be found in Table \ref{tab:support1}, and a breakdown of each categorical variable appears in Table \ref{tab:support2}.
 
@@
 \end{tabular}
-\caption{A description of each variable in the SUPPORT dataset.}
+\caption{Description of each variable in the SUPPORT dataset.}
 \label{tab:support1}
@@
 \end{tabular}
-\caption{A description of each level within each categorical variable in the dataset.}
+\caption{Description of each level within each categorical variable in the SUPPORT dataset.}
 \label{tab:support2}
@@
 
-Before generating an absolute risk curve using regularized logistic regression, we will explore three options as a baseline comparison.
+<!--Two variables are SUPPORT day 3 physiology score (sps) and APACHE III day 3 physiology score (aps).
+These scores were developed for a period of 180 days, which makes our comparison over 5.56 years exploratory.
+Therefore, we will assess these two models before using either as our baseline, by comparing them to a full model using all variables.
+Also, our single imputation introduces bias in the data.
+As the goal here is purely demonstrative, this bias will be consistent with any of the following models.
+As such, their comparison is still valid, though any interpretation requires an awareness of said bias.
+The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.-->
+
+In the original study, the authors developed two scores for predicting the outcome.
+These scores are available as part of the data: SUPPORT day 3 physiology score (\code{sps}) and APACHE III day 3 physiology score (\code{aps}).
+Using these scores, we will explore three Cox regression scenarios as a baseline comparison for \pkg{casebase}: \code{sps} only, \code{aps} only, and all covariates excluding \code{sps} and \code{aps}.
+We will compare the three models using a time-dependent version of the classical Brier score that is adjusted for censoring [@graf1999ass].
+The Brier score can be used to assess both discrimination and calibration in a single model.
+Moreover, it provides a metric that can be used to compare parametric and semi-parametric survival models.
+To account for overfitting, we will use 95% of the observations to train our models, while the remaining 5% will be used to compute absolute risk curves and the Brier scores.
+
+```{r supportData, eval = eval_cs3}
+data(support)
+# Change time to years
+support$d.time <- support$d.time/365.25
+
+# Split into test and train
+train_index <- sample(nrow(support), 0.95*nrow(support))
+test_index <- setdiff(1:nrow(support), train_index)
+
+train <- support[train_index,]
+test <- support[test_index,]
+```
+
+In Figure\ref{fig:baselines}, we can see that all Cox models have similar average survival curves.
+On the other hand, Figure \ref{fig:brier1} shows that the full model has the lowest Brier score overall.
+For this reason, we will use this model in our comparison between regularized Cox regression, Kaplan-Meier estimation, and the approach from \pkg{casebase}.
+
+<!--Before generating an absolute risk curve using regularized logistic regression, we will explore three options as a baseline comparison.
 The original study compared these scores over a span of 180 days, which may not be comparable over the full 5.56 years.
-Therefore, We will examine three cox regression scenarios: sps only, aps only and all covariates excluding sps and aps.
+Therefore, we will examine three Cox regression scenarios: sps only, aps only and all covariates excluding sps and aps.
 Next, we will compare each modeling using the Brier score [@graf1999ass].
@@
 Figure \ref{fig:brier1} examines the brier scores as they change over time.
-As the full model has the lowest brier score on overall, it will be used in the following absolute risk comparison to regularized cox, regularized logistic regression and the unadjusted Kaplan-Meier curve.
-
-```{r baselinePlotsPrep, eval=eval_cs3,echo=FALSE}
+As the full model has the lowest brier score on overall, it will be used in the following absolute risk comparison to regularized cox, regularized logistic regression and the unadjusted Kaplan-Meier curve.-->
+
+```{r baselinePlotsPrep, eval=eval_cs3, echo = FALSE}
+# Define colour palette for this section
 q4 <- qualitative_hcl(4, palette = "Dark 3")
-completeData <- casebase::support
-
-completeData$d.time <- completeData$d.time / 365
-
-workingCompleteData <- model.matrix(death ~ . - d.time, data = completeData)[, -c(1)] # remove intercept
-
-# Create u and xCox, which will be used when fitting Cox with glmnet.
-newData <- workingCompleteData
-x <- workingCompleteData
-# x and y will be used to fit the casebase model under glmnet.
-# x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
-y <- data.matrix(completeData[, c(2, 5)])
-
-fullDataCox <- as.data.frame(cbind(as.numeric(y[, 2]), as.numeric(y[, 1]), x))
-names(fullDataCox)[names(fullDataCox) == "V1"] <- "time"
-names(fullDataCox)[names(fullDataCox) == "V2"] <- "event"
-
-# here we specify train and test size
-train_index <- sample(1:nrow(fullDataCox), 0.8 * nrow(fullDataCox))
-test_index <- setdiff(1:nrow(fullDataCox), train_index)
-
-train <- fullDataCox[train_index, ]
-test <- fullDataCox[test_index, ]
-coxSPS <- survival::coxph(Surv(time = time, event = event) ~ sps, data = train, x = TRUE)
+
+# 1. Cox with only sps
+coxSPS <- survival::coxph(Surv(time = d.time, event = death) ~ sps, 
+                          data = train, x = TRUE)
 abcoxSPS <- survival::survfit(coxSPS, newdata = test)
-
-coxAPS <- survival::coxph(Surv(time = time, event = event) ~ aps, data = train, x = TRUE)
+# 2. Cox with only aps
+coxAPS <- survival::coxph(Surv(time = d.time, event = death) ~ aps, 
+                          data = train, x = TRUE)
 abcoxAPS <- survival::survfit(coxAPS, newdata = test)
-coxfullForumla <- as.formula(paste("Surv(time = time, event = event) ~ age + sexmale + slos + dzgroupCHF + 
-    dzgroupCirrhosis + `dzgroupColon Cancer` + dzgroupComa + 
-    dzgroupCOPD + `dzgroupLung Cancer` + `dzgroupMOSF w/Malig` + 
-    dzclassCancer + dzclassComa + `dzclassCOPD/CHF/Cirrhosis` + 
-    num.co + edu + scoma + avtisst + raceasian + raceblack + 
-    racehispanic + raceother + racewhite + hday + diabetes + 
-    dementia + cano + cayes + meanbp + wblc + hrt + resp + temp + 
-    pafi + alb + bili + crea + sod + ph + glucose + bun + urine + 
-    adlp + adlsc "))
-coxFull <- survival::coxph(coxfullForumla, data = train, x = TRUE)
+# 3. Cox with everything but sps and aps
+coxFull <- survival::coxph(Surv(time = d.time, event = death) ~ . - aps - sps, 
+                           data = train, x = TRUE)
 abcoxFull <- survival::survfit(coxFull, newdata = test)
-# library(prodlim)
-
-KM <- survival::coxph(Surv(time = time, event = event) ~ 1, data = test)
+# 4. Kaplan-Meier
+KM <- survival::coxph(Surv(time = d.time, event = death) ~ 1, data = test)
 abKM <- survival::survfit(KM)
-# abKM <- prodlim(Hist(,event)~1,data=test) minimal difference
-abKMcompare <- cbind(abKM$time[abKM$time <= 1631], 1 - abKM$surv[abKM$time <= 1631])
-colnames(abKMcompare) <- c("Time", "KM")
-
-baseCoxComparisonData <- cbind(abcoxSPS$time[abcoxSPS$time <= 1631], 1 - rowMeans(abcoxSPS$surv[abcoxSPS$time <= 1631, ]), 1 - rowMeans(abcoxAPS$surv[abcoxAPS$time <= 1631, ]), 1 - rowMeans(abcoxFull$surv[abcoxFull$time <= 1631, ]))
-
-colnames(baseCoxComparisonData) <- c("Time", "SPS", "APS", "Full")
-```
-
-```{r baselines, echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:baselines} Absolute risk curves estimated from hazard models.
-Red: APS only hazard was used to generate the absolute risk.
-Green: All covariates aside from APS and SPS.
-Blue: SPS only.
-Purple: unadjusted Kaplan-Meier curve.",echo=FALSE}
-ggplot(as.data.frame(baseCoxComparisonData), aes(Time)) +
+abKMcompare <- data.frame(
+  Time = abKM$time,
+  KM = 1 - abKM$surv
+)
+
+# Combine all CI estimates from Cox models
+baseCoxComparisonData <- data.frame(
+  Time = abcoxSPS$time,
+  SPS = 1 - rowMeans(abcoxSPS$surv),
+  APS = 1 - rowMeans(abcoxAPS$surv),
+  Full = 1 - rowMeans(abcoxFull$surv)
+)
+```
+
+```{r baselines, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:baselines} Cumulative Incidence curves estimated using Cox regression and Kaplan-Meier (K-M)."}
+ggplot(baseCoxComparisonData, aes(Time)) +
   geom_line(aes(y = APS, colour = "APS"), lwd = 1) +
@@
   geom_line(aes(y = SPS, colour = "SPS"), lwd = 1) +
-  geom_step(data = as.data.frame(abKMcompare), mapping = aes(x = Time, y = KM, color = "Unadjusted"), lwd = 1) +
+  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M"), lwd = 1) +
   labs(y = "Probability of Death", x = "Survival time (Years)", color = "Models") +
-  scale_color_manual(values = q4)
-```
-
-```{r brierpt1, echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:brier1} Brier scores over survival time.
-Full model (Full) performs best, while the other two (APS only and SPS only) are comparable.
-Red: APS only model was used to generate the brier scores.
-Green: all covariates excluding APS and SPS.
-Blue: SPS only model.",echo=FALSE}
-
-times <- sort(unique(test$time))
-brierAPS <- Score(list("PredictionModel" = coxAPS), data = test, formula = Hist(time, event != 0) ~ 1, summary = "ipa", se.fit = 0L, metrics = "brier", contrasts = FALSE, times = times)
-brierSPS <- Score(list("PredictionModel" = coxSPS), data = test, formula = Hist(time, event != 0) ~ 1, summary = "ipa", se.fit = 0L, metrics = "brier", contrasts = FALSE, times = times)
-
+  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M"))
+```
+
+```{r brierpt1, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:brier1} Time-dependent Brier scores."}
+# Extract time points
+times <- sort(unique(test$d.time))
+# Compute Brier scores
+brierAPS <- Score(list("PredictionModel" = coxAPS), data = test, 
+                  formula = Hist(d.time, death != 0) ~ 1, summary = "ipa", 
+                  se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
+brierSPS <- Score(list("PredictionModel" = coxSPS), data = test, 
+                  formula = Hist(d.time, death != 0) ~ 1, summary = "ipa", 
+                  se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
+
+# First fix NA coefficients, then compute Brier score
 coxFull$coefficients[which(is.na(coxFull$coefficients))] <- 0
-brierFull <- Score(list("PredictionModel" = coxFull), data = test, formula = Surv(time, event != 0) ~ 1, summary = "ipa", se.fit = 0L, metrics = "brier", contrasts = FALSE, times = times)
-
+brierFull <- Score(list("PredictionModel" = coxFull), data = test, 
+                   formula = Surv(d.time, death != 0) ~ 1, summary = "ipa", 
+                   se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
+
+# Combine all results for plotting
 brierPlotter <- rbind(
-  data.frame(brierScore = brierAPS$Brier$score$Brier[brierAPS$Brier$score$model == "PredictionModel"], times = times, model = "APS"),
-  data.frame(brierScore = brierFull$Brier$score$Brier[brierFull$Brier$score$model == "PredictionModel"], times = times, model = "Full"),
-  data.frame(brierScore = brierSPS$Brier$score$Brier[brierSPS$Brier$score$model == "PredictionModel"], times = times, model = "SPS")
+  data.frame(brierScore = brierAPS$Brier$score$Brier[brierAPS$Brier$score$model == "PredictionModel"], 
+             times = times, model = "APS"),
+  data.frame(brierScore = brierFull$Brier$score$Brier[brierFull$Brier$score$model == "PredictionModel"], 
+             times = times, model = "Full"),
+  data.frame(brierScore = brierSPS$Brier$score$Brier[brierSPS$Brier$score$model == "PredictionModel"], 
+             times = times, model = "SPS")
 )
@@
   xlab("Survival time (Years)") +
-  ylab("Brier-Score") +
+  ylab("Brier score") +
   labs(color = "Models") +
-  ggtitle("Brier scores over time") +
+  # ggtitle("Brier scores over time") +
   scale_color_manual(values = q4[1:3])
@@
 
-To get the absolute risk following regularized cox regression, we had to work around limitations in existing packages.
+<!--To get the absolute risk following regularized cox regression, we had to work around limitations in existing packages.
 To make use of the \pkg{survival} package, we first estimate the hazard coefficients using thecoxnet implementation for lasso penalization in \pkg{glmnet} [@regpathcox].
 Next, we fit a cox model using the survival package with the selected variables only, from our regularized fit.
-Finally, we replaced the coefficients in the survival object with those from our glmnet fit, allowing us to make use of the absolute risk tools available in the \pkg{survival} package.
-
-```{r coxNetCode,eval=FALSE,echo=TRUE}
-# hazard for cox using glmnet
-coxNet <- glmnet::cv.glmnet(
-  x = x, y = y, family = "cox", alpha = 1,
-  standardize = TRUE
-)
-# getting all non-zero covariates
-nonzero_covariate_cox <- predict(coxNet,
-  type = "nonzero",
-  s = "lambda.1se"
-)
-# keeping coefficients for later
-nonzero_coef_cox <- coef(coxNet, s = "lambda.1se")
-# cutting dataset to only include nonzero covariates
-cleanCoxData <- as.data.frame(cbind(y, x[, nonzero_covariate_cox]))
-# create a skeleton object
-coxNet <- survival::coxph(Surv(time = d.time, event = death) ~ .,
-  data = cleanCoxData, x = TRUE
-)
-# replace coefficients with regularized ones
-coxNet$coefficients <- nonzero_coef_cox@x
-```
-
-For our penalized logistic regression model, we opted for the natural log of time, which assumes a Weibull distribution for our event.
-As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where *y* contains the time and event variables and *x* contains all covariates we would like to include in the model.
-For the SUPPORT dataset, all factor variables were converted into dummy variables.
-Here, we used lasso penalization by setting alpha to 1, leading to our smooth hazard function.
-
-```{r casebaseDisplayCode, eval=FALSE,echo=TRUE}
-# regularized logistic regression to estimate hazard
+Finally, we replaced the coefficients in the survival object with those from our glmnet fit, allowing us to make use of the absolute risk tools available in the \pkg{survival} package.-->
+
+For our penalized logistic regression model, we opt for the natural log of time; recall that this corresponds to a Weibull distribution.
+For fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface to the function \code{fitSmoothHazard}.
+In order to use this interfact, we first create a matrix \code{y} containing the time and event variables, and a matrix \code{x} containing all remaining covariates we would like to include in the model space.
+We used lasso penalization by setting \code{alpha = 1}.
+Moreover, we use the argument \code{penalty.factor} to force the time variable into the final model.
+
+```{r supportCB_fit, eval=eval_cs3, echo=TRUE, cache=TRUE}
+x <- model.matrix(death ~ . - d.time - aps - sps, 
+                  data = train)[, -c(1)] # Remove intercept
+y <- data.matrix(subset(train, select = c(d.time, death)))
+
+# Regularized logistic regression to estimate hazard
 cbFull <- casebase::fitSmoothHazard.fit(x, y,
@@
   formula_time = ~ log(d.time), alpha = 1,
-  ratio = 100, standardize = TRUE,
+  ratio = 10, standardize = TRUE,
   penalty.factor = c(0, rep(1, ncol(x)))
 )
-
-# Estimating the absolute risk curve using the newData parameter.
+```
+
+Next, we can use the object \code{cbFull} to estimate the absolute risk curve on the test data.
+We first create a matrix \code{newx} containing the relevant variables.
+Next, using \code{times}, we specify the time points at which the cumulative incidence will be computed.
+Finally, using the parameter \code{s = "lambda.1se"}, we specify at which value of the tuning parameter we want to retrieve the coefficient estimates.
+
+```{r supportCB_abs, eval=eval_cs3, echo=TRUE, cache=TRUE}
+# Estimating the absolute risk curve using the newdata parameter
+newx <- model.matrix(death ~ . - d.time - aps - sps, 
+                  data = test)[, -c(1)]
+times <- sort(unique(test$d.time))
+
 abCbFull <- casebase::absoluteRisk(cbFull,
   time = times,
-  newdata = newData,
+  newdata = newx,
   s = "lambda.1se",
-  method = c("numerical")
-)
-```
-
-```{r supportSetUp, echo = FALSE, eval = eval_cs3}
-
-# remove apsSPS
-train <- train[, -c(46, 47)]
-test <- test[, -c(46, 47)]
-# x and y will be used to fit the casebase model under glmnet.
-y <- data.matrix(train[, c(2, 1)])
-colnames(y) <- c("death", "d.time")
-x <- data.matrix(train[, -c(2, 1)])
-newData <- data.matrix(test[, -c(2, 1)])
-```
-
-```{r coxHazAbsolute, echo = FALSE,eval= eval_cs3 }
-# Create u and xCox, which will be used when fitting Cox with glmnet.
-
-u <- survival::Surv(time = as.numeric(y[, 2]), event = as.factor(y[, 1]))
-
-xCox <- as.matrix(cbind(data.frame("(Intercept)" = rep(1, each = length(x[, 1]))), x))
-colnames(xCox)[2] <- "(Intercept)"
-
-# hazard for cox using glmnet
+  method = "numerical"
+)
+```
+
+```{r coxHazAbsolute, echo = FALSE, eval = eval_cs3}
+# Create survival object for fitting Coxnet
+u <- with(train, survival::Surv(time = d.time, event = death))
 coxNet <- glmnet::cv.glmnet(x = x, y = u, family = "cox", alpha = 1, standardize = TRUE)
-# convergence demonstrated in plot
-# plot(coxNet)
-# taking the coefficient estimates for later use
+
+# Taking the coefficient estimates for later use
 nonzero_covariate_cox <- predict(coxNet, type = "nonzero", s = "lambda.1se")
 nonzero_coef_cox <- coef(coxNet, s = "lambda.1se")
-# creating a new dataset that only contains the covariates chosen through glmnet.
-cleanCoxData <- as.data.frame(cbind(d.time = as.numeric(y[, 2]), death = as.numeric(y[, 1]), x[, nonzero_covariate_cox$X1]))
-
-# fitting a cox model using regular estimation, however we will not keep it.
+# Creating a new dataset that only contains the covariates chosen through glmnet
+cleanCoxData <- as.data.frame(cbind(y, x[, nonzero_covariate_cox$X1]))
+
+# Fitting a cox model using regular estimation, however we will not keep it.
 # this is used more as an object place holder.
-coxNet <- survival::coxph(Surv(time = d.time, event = death) ~ ., data = cleanCoxData, x = TRUE)
-
-# extract coefficients for lollipop plot later
-estimatescoxNet <- setDT(as.data.frame(coef(coxNet)), keep.rownames = TRUE)[]
-
-estimatescoxNet$Model <- "Penalized Cox"
-colnames(estimatescoxNet) <- c("term", "estimate", "model")
-
-# The coefficients of this object will be replaced with the estimates from coxNet.
-# Doing so makes it so that everything is invalid aside from the coefficients.
-# In this case, all we need to estimate the absolute risk is the coefficients.
-# Std. error would be incorrect here, if we were to draw error bars.
+coxNet <- survival::coxph(Surv(time = d.time, event = death) ~ ., 
+                          data = cleanCoxData, x = TRUE)
+
+# The coefficients of this object will be replaced with the estimates from the
+# original coxNet. Doing so makes it so that everything is invalid aside from
+# the coefficients. In this case, all we need to estimate the absolute risk is
+# the coefficients. Std. error would be incorrect here, if we were to draw error
+# bars.
+coxNet_coefnames <- names(coxNet$coefficients)
 coxNet$coefficients <- nonzero_coef_cox@x
-# Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
-
-abcoxNet <- survival::survfit(coxNet, type = "breslow", newdata = as.data.frame(newData[, nonzero_covariate_cox$X1]))
-
-brierCoxNet <- Score(list("PredictionModel" = coxNet), data = as.data.frame(cbind(test[, c(1, 2)], newData[, nonzero_covariate_cox$X1])), formula = Hist(time, event != 0) ~ 1, summary = "ipa", se.fit = 0L, metrics = "brier", contrasts = FALSE, times = times)
-
-abCN <- as.data.frame(cbind(abcoxNet$time, 1 - rowMeans(abcoxNet$surv)))
-colnames(abCN) <- c("Time", "Absolute risk")
+names(coxNet$coefficients) <- coxNet_coefnames
+
+# Fitting absolute risk curve for cox+glmnet
+abcoxNet <- survival::survfit(coxNet, type = "breslow", 
+                              newdata = as.data.frame(newx[, nonzero_covariate_cox$X1]))
+
+# Combine CI estimates from both penalized methods
+abCN <- data.frame(
+  Time = abcoxNet$time,
+  absRisk = 1 - rowMeans(abcoxNet$surv)
+)
 abCN$Model <- "Penalized Cox"
-abCN <- abCN[abCN$Time <= max(abKMcompare[, 1]), ]
-# plot(abCN[,1],1-abCN[,2])
-```
-
-```{r cbHazard}, echo = FALSE, eval = eval_cs3}
-
-cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = 10, standardize = TRUE, penalty.factor = c(0, rep(1, ncol(x))))
-```
-
-```{r CBabsolute, echo = FALSE, eval = eval_cs3}
-# Estimating the absolute risk curve using the newData parameter.
-abCbFull <- casebase::absoluteRisk(cbFull, time = times, newdata = newData, s = "lambda.1se", method = c("numerical"))
-```
-
-```{r cbAbProcessing,echo=FALSE,eval=eval_cs3}
-abCB <- data.frame(abCbFull[, 1], rowMeans(abCbFull[, -c(1)]))
-
-colnames(abCB) <- c("Time", "Absolute risk")
+
+abCB <- data.frame(Time = abCbFull[, 1],
+                   absRisk = rowMeans(abCbFull[, -c(1)]))
 abCB$Model <- "Penalized logistic"
@@
 
-```{r coefplots,echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} A lollipop plot demonstrating the estimated coefficients between the following three models: Cox (green), penalized Cox (red), penalized logisitic (blue).
-If a coefficient is estimated to be 0, in any model, it is removed from the plot for simplicity."}
-
+Our first comparison plot examines the covariates that were selected between all three hazard functions.
+Note that Cox regression without penalization selected more covariates than the other two, but only the intersection of coefficients with values greater than 0 are displayed in \ref{fig:cs3lolliPlot}.
+A visible tendency is for Cox regression to have larger estimated coefficients.
+This is expected, as there are no restrictions on the size of coefficients.
+The smallest coefficient estimates are found in penalized logistic model, aside from cancer state (cano).
+
+```{r coefplots, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} A lollipop plot demonstrating the estimated coefficients between the following three models.
+If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity."}
 library(dotwhisker)
@@
 
-estimatescb <- setDT(as.data.frame(coef(cbFull)[-c(1, 2), 1]), keep.rownames = TRUE)[]
+# extract coefficients for lollipop plot
+estimatescoxNet <- setDT(as.data.frame(coef(coxNet)), 
+                         keep.rownames = TRUE)
+
+estimatescoxNet$Model <- "Pen. Cox"
+colnames(estimatescoxNet) <- c("term", "estimate", "model")
+
+estimatescb <- setDT(as.data.frame(coef(cbFull)[-c(1, 2), 1]), 
+                     keep.rownames = TRUE)
 # estimatescb$std.error<-0
-estimatescb$Model <- "Penalized logistic"
+estimatescb$Model <- "Pen. logistic"
 colnames(estimatescb) <- c("term", "estimate", "model")
 
-estimatescoxFull <- setDT(as.data.frame(coef(coxFull)), keep.rownames = TRUE)[]
+estimatescoxFull <- setDT(as.data.frame(coef(coxFull)), 
+                          keep.rownames = TRUE)
 # estimatescoxFull$std.error<-0
@@
 
+# Clean up the data and the variable names
 lolliplotDots <- rbind(estimatescoxFull, estimatescoxNet, estimatescb)
@@
   ) +
-  ggtitle("Coefficient plot") +
+  # ggtitle("Coefficient plot") +
   labs(color = "Models") +
-  scale_color_manual(values = q4[c(2, 1, 3)])
-```
-
-```{r include=TRUE,echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:cs3Final} Four absolute risk curves, integrated from the following hazard models: Cox (green), penalized Cox (red), penalized logistic (blue) and unadjusted (Kaplan-Meier, purple)."}
-ggplot(as.data.frame(ab), aes(Time)) +
-  geom_line(data = as.data.frame(baseCoxComparisonData), mapping = aes(y = Full, colour = "Cox"), lwd = 1) +
-  geom_line(aes(y = `Absolute risk`, colour = Model), lwd = 1) +
-  geom_step(data = as.data.frame(abKMcompare), mapping = aes(x = Time, y = KM, color = "Unadjusted"), lwd = 1) +
+  scale_color_manual(values = q4[1:3])
+```
+
+These values do not inform us of model performance.
+To do so, we take our hazard functions and use them to predict the absolute risk of everyone in our test set, over time.
+The probabilities over time for each individual are averaged, resulting in the absolute risk curves in Figure \ref{fig:cs3Final}.
+There is separation between the curves, but it is minimal.
+To better demonstrate the differences in performance between the models, we compare Brier scores over time.
+In Figure \ref{fig:cs3FinalBrier}, the Brier score is lower in our penalized logistic regression model compared to both Cox regression models.
+However, this difference is minimal.
+On the other hand, all three models outperform the Kaplan-Meier model.
+
+```{r cs3Final, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3Final} Four Cumulative Incidence curves."}
+ggplot(ab, aes(Time)) +
+  geom_line(data = baseCoxComparisonData, mapping = aes(y = Full, colour = "Cox"), lwd = 1) +
+  geom_line(aes(y = absRisk, colour = Model), lwd = 1) +
+  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M"), lwd = 1) +
   labs(y = "Probability of Death", x = "Survival time (Years)", color = "Models") +
-  ggtitle("Absolute risk curves") +
-  scale_color_manual(values = q4[c(2, 1, 3, 4)])
-```
-
-```{r riskregressionBrier,echo=FALSE,eval=eval_cs3,fig.cap="\\label{fig:cs3FinalBrier} 4 curves comparing the brier scores as they change over time between Cox regression (green), penalized Cox regression (red) and penalized logistic regression (blue) and a unadjusted curve (Kaplan-Meier, purple).
-Penalized logistic regression produced the lowest curve."}
-
-class(cbFull) <- c("singleEventCB", class(cbFull))
-brierFinalResults <- Score(list("Cox" = coxFull, "Penalized Cox" = coxNet, "Penalized Logistic" = cbFull), data = test, formula = Hist(time, event != 0) ~ 1, summary = NULL, se.fit = 0L, metrics = "brier", contrasts = FALSE, times = times)
-
-levels(brierFinalResults$Brier$score$model)[levels(brierFinalResults$Brier$score$model) == "Null model"] <- "Unadjusted"
-b <- brierFinalResults
-
-brierFinalResults$Brier$score$model <- factor(brierFinalResults$Brier$score$model, levels = c("Cox", "Penalized Cox", "Penalized Logistic", "Unadjusted"))
-
-ggplot(data = brierFinalResults$Brier$score, mapping = aes(x = times, y = Brier, col = model)) +
+  # ggtitle("Absolute risk curves") +
+  scale_color_manual(values = q4, breaks = c("Cox", "Penalized Cox", "Penalized logistic", "K-M"))
+```
+
+```{r eval = eval_cs3, echo = FALSE, warning = FALSE}
+# We need this chunk until this method reaches the CRAN version of riskRegression
+predictRisk.singleEventCB <- function(object, newdata, times, cause, ...) {
+  if (!is.null(object$matrix.fit)) {
+    #get all covariates excluding intercept and time
+    coVars=colnames(object$originalData$x)
+    #coVars is used in lines 44 and 50
+    newdata=data.matrix(drop(subset(newdata, select=coVars)))
+  }
+
+  # if (missing(cause)) stop("Argument cause should be the event type for which we predict the absolute risk.")
+  # the output of absoluteRisk is an array with dimension dependening on the length of the requested times:
+  # case 1: the number of time points is 1
+  #         dim(array) =  (length(time), NROW(newdata), number of causes in the data)
+  if (length(times) == 1) {
+    a <- casebase::absoluteRisk(object, newdata = newdata, time = times)
+    p <- matrix(a, ncol = 1)
+  } else {
+    # case 2 a) zero is included in the number of time points
+    if (0 %in% times) {
+      # dim(array) =  (length(time)+1, NROW(newdata)+1, number of causes in the data)
+      a <- casebase::absoluteRisk(object, newdata = newdata, time = times)
+      p <- t(a)
+    } else {
+      # case 2 b) zero is not included in the number of time points (but the absoluteRisk function adds it)
+      a <- casebase::absoluteRisk(object, newdata = newdata, time = times)
+      ### we need to invert the plot because, by default, we get cumulative incidence
+      #a[, -c(1)] <- 1 - a[, -c(1)]
+      ### we remove time 0 for everyone, and remove the time column
+      a <- a[-c(1), -c(1)] ### a[-c(1), ] to keep times column, but remove time 0 probabilities
+      # now we transpose the matrix because in riskRegression we work with number of
+      # observations in rows and time points in columns
+      p <- t(a)
+    }
+  }
+  if (NROW(p) != NROW(newdata) || NCOL(p) != length(times)) {
+    stop(paste("\nPrediction matrix has wrong dimensions:\nRequested newdata x times: ", 
+               NROW(newdata), " x ", length(times), "\nProvided prediction matrix: ", 
+               NROW(p), " x ", NCOL(p), "\n\n", sep = ""))
+  }
+  p
+}
+```
+
+```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} Four curves comparing the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.", warning = FALSE}
+# Compute Brier score for penalized models
+brierFinalResults <- Score(list("Penalized Cox" = coxNet, "Penalized Logistic" = cbFull),
+                           data = cbind(subset(test, select = c(d.time, death)),
+                                        as.data.frame(newx)), 
+                           formula = Hist(d.time, death != 0) ~ 1, summary = NULL, 
+                           se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
+
+# Combine Brier scores between unpenalized and penalized models before plotting
+data_brier <- bind_rows(
+  brierFinalResults$Brier$score,
+  filter(brierFull$Brier$score, model == "PredictionModel")
+  ) %>% 
+  mutate(model = case_when(
+    model == "Null model" ~ "K-M",
+    model == "PredictionModel" ~ "Cox",
+    TRUE ~ model
+  ),
+  model = factor(model, levels = c("Cox", "Penalized Cox", "Penalized Logistic", "K-M")))
+
+ggplot(data = data_brier, mapping = aes(x = times, y = Brier, col = model)) +
   geom_line() +
   xlab("Survival time (Years)") +
-  ylab("Brier-Score") +
-  ggtitle("Brier scores over time") +
+  ylab("Brier score") +
+  # ggtitle("Brier score over time") +
   labs(color = "Models") +
-  scale_color_manual(values = q4[c(2, 1, 3, 4)])
-```
-
-Our first comparison plot examines the covariates that were selected between all three hazard functions.
-Note that Cox regression without penalization selected more covariates than the other two, but only the intersection of coefficients with values greater than 0 are displayed in \ref{fig:cs3lolliPlot}.
-A visible tendency is for Cox regression to have larger estimated coefficients.
-This is expected, as there are no restrictions on the size of coefficients.
-The smallest coefficient estimates are found in penalized logistic model, aside from cancer state (cano).
-These values do not inform us of model performance.
-To do so, we take our hazard functions and use them to predict the absolute risk of everyone in our test set, over time.
-The probabilities over time for each individual are averaged, resulting in the absolute risk curves in figure \ref{fig:cs3Final}.
-There is separation between the curves, but it is minimal.
-To better demonstrate the differences between the models, we compare brier scores over time.
-\ref{fig:cs3FinalBrier} The Brier score is lower in our penalized logistic regression model, compared to the other models.
-This difference, however, is minimal.
-We can say with relative certainty that any of the hazard models outperform an unadjusted curve, while penalized logistic regression performs the best on this testing dataset.
-
-# Case study 4--Stanford Heart Transplant Data
+  scale_color_manual(values = q4)
+```
+
+# Case study 4---Stanford Heart Transplant Data
 

### 2020-05-05 analysis/paper/paper.tex
edit jesse's section

--- 2020-04-09 analysis/paper/paper.tex
@@
 \author{
-Sahir Bhatnagar *\\McGill University \And Maxime Turgeon *\\University of Manitoba \AND Jesse Islam\\McGill University \And James Hanley\\McGill University \And Olli Saarela\\University of Toronto
+Sahir Bhatnagar *\\McGill University \And Maxime Turgeon *\\University
+of Manitoba \AND Jesse Islam\\McGill University \And James
+Hanley\\McGill University \And Olli Saarela\\University of Toronto
 }
@@
 
-\Plainauthor{Sahir Bhatnagar *, Maxime Turgeon *, Jesse Islam, James Hanley, Olli Saarela}
+\Plainauthor{Sahir Bhatnagar *, Maxime Turgeon *, Jesse Islam, James
+Hanley, Olli Saarela}
 \Plaintitle{casebase: An Alternative Framework For Survival Analysis}
-\Shorttitle{\pkg{casebase}: An Alternative Framework For Survival Analysis}
+\Shorttitle{\pkg{casebase}: An Alternative Framework For Survival
+Analysis}
 
@@
 model, which implies that the counting process jumps are less than or
-equal to one. We are interested in modeling the hazard functions
+equal to one. We are interested in modelling the hazard functions
 \(\lambda_{i}(t)\) of the processes \(N_i(t)\), and which satisfy
@@
 \[L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.\]
-When a logarithmic link function is used for modeling the hazard
+When a logarithmic link function is used for modelling the hazard
 function, the above expression is of a logistic regression form with an
@@
 Hanley \& Miettinen \citeyearpar{hanley2009fitting}: we first sample
-along the ``person'' axis, proportional to each individual's total
+along the ``person'\,' axis, proportional to each individual's total
 follow-up time, and then we sample a moment uniformly over their
@@
 \hypertarget{case-study-1european-randomized-study-of-prostate-cancer-screening}{%
-\section{Case study 1--European Randomized Study of Prostate Cancer
+\section{Case study 1---European Randomized Study of Prostate Cancer
 Screening}\label{case-study-1european-randomized-study-of-prostate-cancer-screening}}
@@
 \hypertarget{case-study-2bone-marrow-transplant}{%
-\section{Case study 2--Bone-marrow
+\section{Case study 2---Bone-marrow
 transplant}\label{case-study-2bone-marrow-transplant}}
@@
 \hypertarget{case-study-3support-data}{%
-\section{Case study 3--SUPPORT Data}\label{case-study-3support-data}}
+\section{Case study 3---SUPPORT Data}\label{case-study-3support-data}}
 
@@
 parametric models for hazard functions, and estimating the corresponding
-cumulative incidence curves. In the next two case studies, we explore
-more advanced features.
-
-For the third case study, we will look at regularized estimation for the
-hazard function. This can be performed using a combination of case-base
-sampling and regularized logistic regression. To demonstrate this
-capability, we examine the Study to Understand Prognoses Preferences
-Outcomes and Risks of Treatment (SUPPORT) dataset
-\citep{knaus1995support}. The SUPPORT dataset tracks death for
-individuals who are considered seriously ill within 5 hospitals in the
-United States of America. Imputation was conducted using the \pkg{mice}
-package in \proglang{R} with default settings \citep{mice}. Before
-imputation, there were a total of 9105 individuals and 47 variables. Of
-those variables, a few were removed before imputation. We removed three
-response variables: hospital charges, patient ratio of costs to charges
-and patient micro-costs. Next, we removed hospital death as it was
-directly informative of our event of interest, death. We also removed
-functional disability and income as they are ordinal covariates, because
-ordinal variables are treated as factors in our current model. Finally,
-we removed 8 covariates relating to the results of previous findings: we
-removed SUPPORT day 3 physiology score (sps), APACHE III day 3
-physiology score (aps), SUPPORT model 2-month survival estimate, SUPPORT
-model 6-month survival estimate, Physician's 2-month survival estimate
-for pt., Physician's 6-month survival estimate for pt., Patient had Do
-Not Resuscitate (DNR) order and Day of DNR order (\textless{}0 if before
-study). Of these, sps and aps were added on after imputation, as they
-were missing only 1 observation and permitted us to compare findings
-between our model and their's.
-
-First we impute manually using recommended normal values for
-physiological measures \citep{harrell_2020}. Next, we impute a single
-dataset using \pkg{mice} with default settings. After imputation, we
-noted that the covariate for surrogate activities of daily living was
-not imputed. This is due to colinearity between the other two covariates
-for activities of daily living. As such, surrogate activites of daily
-living was removed from our model. After these imputations, we have a
-total of 9104 observations and 30 variables, two of which are sps and
-aps, which will work as a baseline for our comparison to previous
-findings. This dataset is available Its worth noting that these scores
-were developed for a period of 180 days, which makes our comparison over
-5.56 years exploratory. Also, our single imputation introduces bias in
-the data. As the goal here is purely demonstrative, this bias will be
-consistent with any of the following models. As such, their comparison
-is still valid, though any interpretation requires an awareness of said
-bias. A description of used variables can be found in Table
-\ref{tab:support1} and a breakdown of each categorical variable in Table
-\ref{tab:support2}. The original data was obtained from the Department
-of Biostatistics at Vanderbilt University \citep{harrell_2020}. The
-imputed dataset is available as part of the \pkg{casebase} package.
+cumulative incidence curves. For the third case study, we show how
+\pkg{casebase} can also be used for variable selection through
+regularized estimation of the hazard function. To do this, we simply
+replace logistic regression with a penalized counterpart. In
+\pkg{casebase}, we achieve this by using the \pkg{glmnet} package
+\citep{friedman2010jss}.
+
+To illustrate this functionality, we use the dataset from the Study to
+Understand Prognoses Preferences Outcomes and Risks of Treatment
+(SUPPORT) \citep{knaus1995support}. The SUPPORT dataset tracks death in
+five American hospitals within individuals who are considered seriously
+ill.~The original data is available online from the Department of
+Biostatistics at Vanderbilt University \citep{harrell_2020}. The cleaned
+and imputed data consists of 9104 observations and 30 variables, and it
+is available as part of the \pkg{casebase} package. For more information
+about this dataset, the reader is encouraged to look at the
+documentation in our package. A description of the variables can be
+found in Table \ref{tab:support1}, and a breakdown of each categorical
+variable appears in Table \ref{tab:support2}.
 
@@
 
-As before, the first step is to visualize the incidence density. The
-population-time plot for the SUPPORT data appears in Figure
-\ref{fig:support-poptime}. As we can see, the incidence density is not
-constant.
+In the original study, the authors developed two scores for predicting
+the outcome. These scores are available as part of the data: SUPPORT day
+3 physiology score (\code{sps}) and APACHE III day 3 physiology score
+(\code{aps}). Using these scores, we will explore three Cox regression
+scenarios as a baseline comparison for \pkg{casebase}: \code{sps} only,
+\code{aps} only, and all covariates excluding \code{sps} and \code{aps}.
+We will compare the three models using a time-dependent version of the
+classical Brier score that is adjusted for censoring
+\citep{graf1999ass}. The Brier score can be used to assess both
+discrimination and calibration in a single model. Moreover, it provides
+a metric that can be used to compare parametric and semi-parametric
+survival models. To account for overfitting, we will use 95\% of the
+observations to train our models, while the remaining 5\% will be used
+to compute absolute risk curves and the Brier scores.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> data(support)
+R> # Change time to years
+R> support$d.time <- support$d.time/365.25
+R> 
+R> # Split into test and train
+R> train_index <- sample(nrow(support), 0.95*nrow(support))
+R> test_index <- setdiff(1:nrow(support), train_index)
+R> 
+R> train <- support[train_index,]
+R> test <- support[test_index,]
+\end{CodeInput}
+\end{CodeChunk}
+
+In Figure\ref{fig:baselines}, we can see that all Cox models have
+similar average survival curves. On the other hand, Figure
+\ref{fig:brier1} shows that the full model has the lowest Brier score
+overall. For this reason, we will use this model in our comparison
+between regularized Cox regression, Kaplan-Meier estimation, and the
+approach from \pkg{casebase}.
 
@@
 
-{\centering \includegraphics{../figures/support-poptime-1}
+{\centering \includegraphics{../figures/baselines-1}
 
@@
 
-\caption{\label{fig:support-poptime} Population-time plot for the SUPPORT data.}\label{fig:support-poptime}
+\caption{\label{fig:baselines} Cumulative Incidence curves estimated using Cox regression and Kaplan-Meier (K-M).}\label{fig:baselines}
 \end{figure}
@@
 
-Before generating an absolute risk curve using regularized logistic
-regression, we will explore three options as a baseline comparison,
-beyond a Kaplan-meier curve. The original study compared these scores
-over a span of 180 days, which may not be comparable over the full 5.56
-years. Therefore, We will examine three cox regression scenarios: sps
-only, aps only and all covariates excluding sps and aps. Next, we will
-compare each modeling using the Brier score \citep{graf1999ass}. The
-Brier score is able to demonstrate both discrimination and calibration
-in a single model \citep{graf1999ass} and provide a comparable metric
-between semi-parametric and parametric methods in the survival context.
-For all the following analyses in this case study, we will use 80\% of
-the observations to train our models, while 20\% will be used to
-generate absolute risk curves, all of which will be averaged to
-demonstrate each model's predictive ability. Over 5.56 years,
-\ref{fig:baselines} demonstrates that all the cox models have similar,
-if not identical average predictions. Figure \ref{fig:brier1} examines
-the brier scores as they change over time. As the full model has the
-lowest brier score on overall, it will be used in the following
-comparison to regularized cox, regularized logistic regression and the
-unadjusted Kaplan-Meier curve.
-
 \begin{CodeChunk}
@@
 
-{\centering \includegraphics{../figures/baselines-1}
+{\centering \includegraphics{../figures/brierpt1-1}
 
@@
 
-\caption{\label{fig:baselines} AbsoluteRisk curves from hazard models.
-Pink: All covariates without SPS nor APS.
-Green: APS only model.
-Blue: SPS only model.
-Purple: unadjusted curve.
-All three models underestimate the risk in comparison to the unadjusted curve.}\label{fig:baselines}
+\caption{\label{fig:brier1} Time-dependent Brier scores.}\label{fig:brierpt1}
 \end{figure}
@@
 
+For our penalized logistic regression model, we opt for the natural log
+of time; recall that this corresponds to a Weibull distribution. For
+fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which
+is a matrix interface to the function \code{fitSmoothHazard}. In order
+to use this interfact, we first create a matrix \code{y} containing the
+time and event variables, and a matrix \code{x} containing all remaining
+covariates we would like to include in the model space. We used lasso
+penalization by setting \code{alpha = 1}. Moreover, we use the argument
+\code{penalty.factor} to force the time variable into the final model.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> x <- model.matrix(death ~ . - d.time - aps - sps, 
+R+                   data = train)[, -c(1)] # Remove intercept
+R> y <- data.matrix(subset(train, select = c(d.time, death)))
+R> 
+R> # Regularized logistic regression to estimate hazard
+R> cbFull <- casebase::fitSmoothHazard.fit(x, y,
+R+   family = "glmnet",
+R+   time = "d.time", event = "death",
+R+   formula_time = ~ log(d.time), alpha = 1,
+R+   ratio = 10, standardize = TRUE,
+R+   penalty.factor = c(0, rep(1, ncol(x)))
+R+ )
+\end{CodeInput}
+\end{CodeChunk}
+
+Next, we can use the object \code{cbFull} to estimate the absolute risk
+curve on the test data. We first create a matrix \code{newx} containing
+the relevant variables. Next, using \code{times}, we specify the time
+points at which the cumulative incidence will be computed. Finally,
+using the parameter \code{s = "lambda.1se"}, we specify at which value
+of the tuning parameter we want to retrieve the coefficient estimates.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Estimating the absolute risk curve using the newdata parameter
+R> newx <- model.matrix(death ~ . - d.time - aps - sps, 
+R+                   data = test)[, -c(1)]
+R> times <- sort(unique(test$d.time))
+R> 
+R> abCbFull <- casebase::absoluteRisk(cbFull,
+R+   time = times,
+R+   newdata = newx,
+R+   s = "lambda.1se",
+R+   method = "numerical"
+R+ )
+\end{CodeInput}
+\end{CodeChunk}
+
+Our first comparison plot examines the covariates that were selected
+between all three hazard functions. Note that Cox regression without
+penalization selected more covariates than the other two, but only the
+intersection of coefficients with values greater than 0 are displayed in
+\ref{fig:cs3lolliPlot}. A visible tendency is for Cox regression to have
+larger estimated coefficients. This is expected, as there are no
+restrictions on the size of coefficients. The smallest coefficient
+estimates are found in penalized logistic model, aside from cancer state
+(cano).
+
 \begin{CodeChunk}
@@
 
-{\centering \includegraphics{../figures/brierpt1-1}
+{\centering \includegraphics{../figures/coefplots-1}
 
@@
 
-\caption{\label{fig:brier1} Brier scores over time. Full model (Cox Full) performs best, while the other two (APS only and SPS only) are comparable.}\label{fig:brierpt1}
+\caption{\label{fig:cs3lolliPlot} A lollipop plot demonstrating the estimated coefficients between the following three models.
+If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity.}\label{fig:coefplots}
 \end{figure}
@@
 
-To get the absolute risk following regularized cox regression, this
-required that we work around limitations in existing packages. To make
-use of the \pkg{survival} package, we first get our hazard coefficients
-using \pkg{glmnet}'s coxnet implementation for lasso penalization
-\citep{regpathcox}. Next, we fit a cox model using the survival package
-with the selected variables only, from our previous fit. Finally, we
-replaced the coefficients in the survival object with those from our
-glmnet fit, allowing us to make use of the absolute risk tools available
-in the \pkg{survival} package.
-
-For our penalized logistic regression model, we opted for a logistic
-function of time, resulting in a Weibull distribution assumption. As
-casebase makes use of the glmnet package, we interact with the
-fitsmoothhazard.fit function using a matrix interface, where y contains
-the time and event variables and x contains all other variables we would
-like to include in the model. For the SUPPORT dataset, all factor
-variables were converted into dummy variables. Here, we used lasso
-penalization by setting alpha to 1, leading to our smooth hazard
-function. We then take our hazard and use it to predict the absolute
-risk of each individual in our test set, over time. The probabilites
-over time for each individual are averaged, resulting in the smooth
-absolute risk curve in figure \ref{fig:cs3Final}. On average, the Brier
-score is lower in our penalized logistic regression model, compared to
-the other models (fig.~\ref{fig:cs3FinalBrier}).
-
-This case study serves to demonstrate how regularization can be used in
-a survival context, using logistic regression. Our approach performs
-comparably to Cox regression, demonstrating its validity in an applied
-setting.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> 
-R> cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = 10, standardize = TRUE,penalty.facto=c(0,rep(1,ncol(x))))
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Estimating the absolute risk curve using the newData parameter.
-R> abCbFull <- casebase::absoluteRisk(cbFull, time = times, newdata = newData, s = "lambda.1se", method = c("numerical"))
-\end{CodeInput}
-\end{CodeChunk}
+These values do not inform us of model performance. To do so, we take
+our hazard functions and use them to predict the absolute risk of
+everyone in our test set, over time. The probabilities over time for
+each individual are averaged, resulting in the absolute risk curves in
+Figure \ref{fig:cs3Final}. There is separation between the curves, but
+it is minimal. To better demonstrate the differences in performance
+between the models, we compare Brier scores over time. In Figure
+\ref{fig:cs3FinalBrier}, the Brier score is lower in our penalized
+logistic regression model compared to both Cox regression models.
+However, this difference is minimal. On the other hand, all three models
+outperform the Kaplan-Meier model.
 
@@
 
-{\centering \includegraphics{../figures/unnamed-chunk-2-1}
+{\centering \includegraphics{../figures/cs3Final-1}
 
@@
 
-\caption{\label{fig:cs3Final} Four absolute risk curves, integrated from the following hazard models: Cox (pink), penalized Cox (green), penalized logistic (blue) and Kaplan-Meier (purple).}\label{fig:unnamed-chunk-2}
+\caption{\label{fig:cs3Final} Four Cumulative Incidence curves.}\label{fig:cs3Final}
 \end{figure}
@@
 
-\caption{\label{fig:cs3FinalBrier} 4 curves comparing the brier scores as they change over time between a null model (Kaplan-Meier, black), Cox regression (yellow), penalized Cox regression (blue) and penalized logistic regression (green).}\label{fig:riskregressionBrier}
+\caption{\label{fig:cs3FinalBrier} Four curves comparing the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.}\label{fig:riskregressionBrier}
 \end{figure}
@@
 
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/coefplots-1}
-
-}
-
-\caption{\label{fig:cs3FinalBrier} A lollipop plot demonstrating the estimated coefficients between the following three models: Cox (blue), penalized Cox (Green), penalized logisitic (red).
-If a coefficient is estimated to be 0, it is removed from the plot}\label{fig:coefplots}
-\end{figure}
-\end{CodeChunk}
-
 \hypertarget{case-study-4stanford-heart-transplant-data}{%
-\section{Case study 4--Stanford Heart Transplant
+\section{Case study 4---Stanford Heart Transplant
 Data}\label{case-study-4stanford-heart-transplant-data}}
@@
 
-This report was generated on 2020-04-09 13:28:46 using the following
+This report was generated on 2020-05-05 17:12:52 using the following
 computational environment and dependencies:
@@
 \begin{CodeOutput}
-#> R version 3.6.3 (2020-02-29)
-#> Platform: x86_64-w64-mingw32/x64 (64-bit)
-#> Running under: Windows 10 x64 (build 18363)
+#> R version 4.0.0 (2020-04-24)
+#> Platform: x86_64-apple-darwin17.0 (64-bit)
+#> Running under: macOS Catalina 10.15.4
 #> 
 #> Matrix products: default
+#> BLAS:   /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRblas.dylib
+#> LAPACK: /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRlapack.dylib
 #> 
 #> locale:
-#> [1] LC_COLLATE=English_United States.1252 
-#> [2] LC_CTYPE=English_United States.1252   
-#> [3] LC_MONETARY=English_United States.1252
-#> [4] LC_NUMERIC=C                          
-#> [5] LC_TIME=English_United States.1252    
+#> [1] en_CA.UTF-8/en_CA.UTF-8/en_CA.UTF-8/C/en_CA.UTF-8/en_CA.UTF-8
 #> 
@@
 #> other attached packages:
-#>  [1] broom_0.5.5               dotwhisker_0.5.0         
-#>  [3] prodlim_2019.11.13        riskRegression_2020.03.27
-#>  [5] pracma_2.2.9              kableExtra_1.1.0         
-#>  [7] data.table_1.12.8         glmnet_3.0-2             
-#>  [9] Matrix_1.2-18             casebase_0.2.1.9001      
-#> [11] survival_3.1-11           magrittr_1.5             
-#> [13] forcats_0.5.0             stringr_1.4.0            
-#> [15] dplyr_0.8.5               purrr_0.3.3              
-#> [17] readr_1.3.1               tidyr_1.0.2              
-#> [19] tibble_3.0.0              ggplot2_3.3.0            
-#> [21] tidyverse_1.3.0          
+#>  [1] broom_0.5.6               dotwhisker_0.5.0         
+#>  [3] colorspace_1.4-1          prodlim_2019.11.13       
+#>  [5] riskRegression_2020.02.05 pracma_2.2.9             
+#>  [7] kableExtra_1.1.0          data.table_1.12.8        
+#>  [9] glmnet_3.0-2              Matrix_1.2-18            
+#> [11] casebase_0.2.1.9001       survival_3.1-12          
+#> [13] magrittr_1.5              forcats_0.5.0            
+#> [15] stringr_1.4.0             dplyr_0.8.5              
+#> [17] purrr_0.3.4               readr_1.3.1              
+#> [19] tidyr_1.0.2               tibble_3.0.1             
+#> [21] ggplot2_3.3.0             tidyverse_1.3.0          
 #> 
 #> loaded via a namespace (and not attached):
-#>  [1] TH.data_1.0-10      VGAM_1.1-2          colorspace_1.4-1   
-#>  [4] ellipsis_0.3.0      ggstance_0.3.4      htmlTable_1.13.3   
-#>  [7] base64enc_0.1-3     fs_1.4.1            rstudioapi_0.11    
-#> [10] farver_2.0.3        MatrixModels_0.4-1  fansi_0.4.1        
-#> [13] mvtnorm_1.1-0       lubridate_1.7.8     xml2_1.3.1         
-#> [16] codetools_0.2-16    knitr_1.28          Formula_1.2-3      
-#> [19] jsonlite_1.6.1      cluster_2.1.0       dbplyr_1.4.2       
-#> [22] png_0.1-7           compiler_3.6.3      httr_1.4.1         
-#> [25] backports_1.1.6     assertthat_0.2.1    cli_2.0.2          
-#> [28] acepack_1.4.1       htmltools_0.4.0     quantreg_5.55      
-#> [31] tools_3.6.3         gtable_0.3.0        glue_1.4.0         
-#> [34] Rcpp_1.0.4.6        cellranger_1.1.0    vctrs_0.2.4        
-#> [37] nlme_3.1-145        iterators_1.0.12    xfun_0.12          
-#> [40] rvest_0.3.5         lifecycle_0.2.0     polspline_1.1.17   
-#> [43] MASS_7.3-51.5       zoo_1.8-7           scales_1.1.0       
-#> [46] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
-#> [49] RColorBrewer_1.1-2  rticles_0.14        yaml_2.2.1         
-#> [52] gridExtra_2.3       rms_5.1-4           rpart_4.1-15       
-#> [55] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
-#> [58] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
-#> [61] mets_1.2.7.1        rlang_0.4.5         pkgconfig_2.0.3    
-#> [64] evaluate_0.14       lattice_0.20-41     htmlwidgets_1.5.1  
-#> [67] labeling_0.3        cmprsk_2.2-9        tidyselect_1.0.0   
-#> [70] plyr_1.8.6          R6_2.4.1            generics_0.0.2     
-#> [73] Hmisc_4.4-0         multcomp_1.4-13     DBI_1.1.0          
-#> [76] pillar_1.4.3        haven_2.2.0         foreign_0.8-76     
-#> [79] withr_2.1.2         mgcv_1.8-31         nnet_7.3-13        
-#> [82] modelr_0.1.6        crayon_1.3.4        rmarkdown_2.1      
-#> [85] timereg_1.9.4       jpeg_0.1-8.1        grid_3.6.3         
-#> [88] readxl_1.3.1        reprex_0.3.0        digest_0.6.25      
-#> [91] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_3.6.3       
-#> [94] munsell_0.5.0       viridisLite_0.3.0
+#>  [1] nlme_3.1-147        fs_1.4.1            cmprsk_2.2-9       
+#>  [4] lubridate_1.7.8     rms_5.1-4           webshot_0.5.2      
+#>  [7] RColorBrewer_1.1-2  httr_1.4.1          numDeriv_2016.8-1.1
+#> [10] tools_4.0.0         backports_1.1.6     R6_2.4.1           
+#> [13] rpart_4.1-15        Hmisc_4.4-0         DBI_1.1.0          
+#> [16] mgcv_1.8-31         nnet_7.3-13         withr_2.2.0        
+#> [19] gridExtra_2.3       tidyselect_1.0.0    compiler_4.0.0     
+#> [22] quantreg_5.55       cli_2.0.2           rvest_0.3.5        
+#> [25] htmlTable_1.13.3    SparseM_1.78        xml2_1.3.2         
+#> [28] sandwich_2.5-1      labeling_0.3        checkmate_2.0.0    
+#> [31] scales_1.1.0        mvtnorm_1.1-0       polspline_1.1.17   
+#> [34] digest_0.6.25       foreign_0.8-78      rmarkdown_2.1      
+#> [37] base64enc_0.1-3     rticles_0.14        jpeg_0.1-8.1       
+#> [40] pkgconfig_2.0.3     htmltools_0.4.0     dbplyr_1.4.3       
+#> [43] htmlwidgets_1.5.1   rlang_0.4.5         readxl_1.3.1       
+#> [46] rstudioapi_0.11     VGAM_1.1-2          farver_2.0.3       
+#> [49] shape_1.4.4         generics_0.0.2      zoo_1.8-7          
+#> [52] jsonlite_1.6.1      acepack_1.4.1       Formula_1.2-3      
+#> [55] Rcpp_1.0.4.6        munsell_0.5.0       fansi_0.4.1        
+#> [58] lifecycle_0.2.0     multcomp_1.4-13     stringi_1.4.6      
+#> [61] yaml_2.2.1          MASS_7.3-51.5       plyr_1.8.6         
+#> [64] ggstance_0.3.4      grid_4.0.0          crayon_1.3.4       
+#> [67] lattice_0.20-41     haven_2.2.0         hms_0.5.3          
+#> [70] timereg_1.9.4       knitr_1.28          pillar_1.4.3       
+#> [73] codetools_0.2-16    stats4_4.0.0        reprex_0.3.0       
+#> [76] glue_1.4.0          evaluate_0.14       latticeExtra_0.6-29
+#> [79] modelr_0.1.6        vctrs_0.2.4         png_0.1-7          
+#> [82] foreach_1.5.0       MatrixModels_0.4-1  cellranger_1.1.0   
+#> [85] gtable_0.3.0        assertthat_0.2.1    xfun_0.13          
+#> [88] viridisLite_0.3.0   iterators_1.0.12    cluster_2.1.0      
+#> [91] lava_1.6.7          TH.data_1.0-10      ellipsis_0.3.0
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> Local:    master C:/Users/jesse/Documents/gits/cbpaper
-#> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
-#> Head:     [b31e195] 2020-03-31: Fixed penal. on time and Brier
+#> Local:    editMax /Users/maxturgeon/Documents/git_repos/cbpaper
+#> Head:     [70115f2] 2020-04-22: Merge pull request #20 from Jesse-Islam/master
 \end{CodeOutput}

### 2020-05-07 analysis/paper/paper.Rmd
combine ci and brier plots
We still need to fix the legends and the size of the figures, but this can be done at a later time.

@@
 
-In Figure\ref{fig:baselines}, we can see that all Cox models have similar average survival curves.
-On the other hand, Figure \ref{fig:brier1} shows that the full model has the lowest Brier score overall.
+In Figure\ref{fig:brier1} (A), we can see that all Cox models have similar average survival curves.
+On the other hand, Figure \ref{fig:brier1} (B) shows that the full model has the lowest Brier score overall.
 For this reason, we will use this model in our comparison between regularized Cox regression, Kaplan-Meier estimation, and the approach from \pkg{casebase}.
@@
 
-```{r baselines, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:baselines} Cumulative Incidence curves estimated using Cox regression and Kaplan-Meier (K-M)."}
-ggplot(baseCoxComparisonData, aes(Time)) +
+```{r baselines, echo=FALSE, eval=eval_cs3}
+cox_ci <- ggplot(baseCoxComparisonData, aes(Time)) +
   geom_line(aes(y = APS, colour = "APS"), lwd = 1) +
@@
 
-```{r brierpt1, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:brier1} Time-dependent Brier scores."}
+```{r brierpt1, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:brier1} Comparison of three Cox regression models. (A) Cumulative Incidence curves. (B) Time-dependent Brier scores."}
 # Extract time points
@@
 
-ggplot(data = brierPlotter, mapping = aes(x = times, y = brierScore, col = model)) +
+cox_brier <- ggplot(data = brierPlotter, mapping = aes(x = times, y = brierScore, col = model)) +
   geom_line() +
@@
   scale_color_manual(values = q4[1:3])
+
+# Combine both plots using cowplot
+plot_row <- cowplot::plot_grid(cox_ci + theme(legend.position = 'none'),
+                               cox_brier + theme(legend.position = 'none'),
+                               labels = c("A", "B"))
+legend <- cowplot::get_legend(
+  # create some space to the left of the legend
+  cox_ci + theme(legend.box.margin = margin(0, 0, 0, 12))
+)
+
+cowplot::plot_grid(plot_row, legend, rel_widths = c(3, .4))
 ```
@@
 
-```{r coefplots, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} A lollipop plot demonstrating the estimated coefficients between the following three models.
+```{r coefplots, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models.
 If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity."}
@@
 To do so, we take our hazard functions and use them to predict the absolute risk of everyone in our test set, over time.
-The probabilities over time for each individual are averaged, resulting in the absolute risk curves in Figure \ref{fig:cs3Final}.
+The probabilities over time for each individual are averaged, resulting in the absolute risk curves in Figure \ref{fig:cs3FinalBrier} (A).
 There is separation between the curves, but it is minimal.
 To better demonstrate the differences in performance between the models, we compare Brier scores over time.
-In Figure \ref{fig:cs3FinalBrier}, the Brier score is lower in our penalized logistic regression model compared to both Cox regression models.
+In Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our penalized logistic regression model compared to both Cox regression models.
 However, this difference is minimal.
@@
 
-```{r cs3Final, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3Final} Four Cumulative Incidence curves."}
-ggplot(ab, aes(Time)) +
+```{r cs3Final, echo=FALSE, eval=eval_cs3}
+reg_ci <- ggplot(ab, aes(Time)) +
   geom_line(data = baseCoxComparisonData, mapping = aes(y = Full, colour = "Cox"), lwd = 1) +
@@
 
-```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} Four curves comparing the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.", warning = FALSE}
+```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} Four curves comparing the Cumulative Incidence (A) and the Brier scores (B) as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.", warning = FALSE}
 # Compute Brier score for penalized models
@@
   ),
-  model = factor(model, levels = c("Cox", "Penalized Cox", "Penalized Logistic", "K-M")))
-
-ggplot(data = data_brier, mapping = aes(x = times, y = Brier, col = model)) +
+  model = factor(model, levels = c("Cox", "Penalized Cox", "Penalized Logistic", "K-M"),
+                 labels = c("Cox", "Pen. Cox", "Pen. Logistic", "K-M")))
+
+reg_brier <- ggplot(data = data_brier, mapping = aes(x = times, y = Brier, col = model)) +
   geom_line() +
@@
   scale_color_manual(values = q4)
+
+# Combine both plots using cowplot
+plot_row_reg <- cowplot::plot_grid(reg_ci + theme(legend.position = 'none'),
+                                   reg_brier + theme(legend.position = 'none'),
+                                   labels = c("A", "B"))
+legend_reg <- cowplot::get_legend(
+  # create some space to the left of the legend
+  reg_ci + theme(legend.box.margin = margin(0, 0, 0, 12))
+)
+
+cowplot::plot_grid(plot_row_reg, legend_reg, rel_widths = c(3, .4))
 ```

### 2020-05-07 analysis/paper/paper.tex
combine ci and brier plots
We still need to fix the legends and the size of the figures, but this can be done at a later time.

@@
 
-In Figure\ref{fig:baselines}, we can see that all Cox models have
+In Figure\ref{fig:brier1} (A), we can see that all Cox models have
 similar average survival curves. On the other hand, Figure
-\ref{fig:brier1} shows that the full model has the lowest Brier score
-overall. For this reason, we will use this model in our comparison
+\ref{fig:brier1} (B) shows that the full model has the lowest Brier
+score overall. For this reason, we will use this model in our comparison
 between regularized Cox regression, Kaplan-Meier estimation, and the
@@
 
-\caption{\label{fig:baselines} Cumulative Incidence curves estimated using Cox regression and Kaplan-Meier (K-M).}\label{fig:baselines}
-\end{figure}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/brierpt1-1}
-
-}
-
-\caption{\label{fig:brier1} Time-dependent Brier scores.}\label{fig:brierpt1}
+\caption{\label{fig:brier1} Comparison of three Cox regression models. (A) Cumulative Incidence curves. (B) Time-dependent Brier scores.}\label{fig:brierpt1}
 \end{figure}
@@
 
-\caption{\label{fig:cs3lolliPlot} A lollipop plot demonstrating the estimated coefficients between the following three models.
+\caption{\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models.
 If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity.}\label{fig:coefplots}
@@
 each individual are averaged, resulting in the absolute risk curves in
-Figure \ref{fig:cs3Final}. There is separation between the curves, but
-it is minimal. To better demonstrate the differences in performance
-between the models, we compare Brier scores over time. In Figure
-\ref{fig:cs3FinalBrier}, the Brier score is lower in our penalized
-logistic regression model compared to both Cox regression models.
-However, this difference is minimal. On the other hand, all three models
-outperform the Kaplan-Meier model.
+Figure \ref{fig:cs3FinalBrier} (A). There is separation between the
+curves, but it is minimal. To better demonstrate the differences in
+performance between the models, we compare Brier scores over time. In
+Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our
+penalized logistic regression model compared to both Cox regression
+models. However, this difference is minimal. On the other hand, all
+three models outperform the Kaplan-Meier model.
 
@@
 
-{\centering \includegraphics{../figures/cs3Final-1}
+{\centering \includegraphics{../figures/riskregressionBrier-1}
 
@@
 
-\caption{\label{fig:cs3Final} Four Cumulative Incidence curves.}\label{fig:cs3Final}
-\end{figure}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/riskregressionBrier-1}
-
-}
-
-\caption{\label{fig:cs3FinalBrier} Four curves comparing the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.}\label{fig:riskregressionBrier}
+\caption{\label{fig:cs3FinalBrier} Four curves comparing the Cumulative Incidence (A) and the Brier scores (B) as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.}\label{fig:riskregressionBrier}
 \end{figure}
@@
 
-This report was generated on 2020-05-05 17:12:52 using the following
+This report was generated on 2020-05-07 08:27:50 using the following
 computational environment and dependencies:
@@
 #> loaded via a namespace (and not attached):
-#>  [1] nlme_3.1-147        fs_1.4.1            cmprsk_2.2-9       
-#>  [4] lubridate_1.7.8     rms_5.1-4           webshot_0.5.2      
-#>  [7] RColorBrewer_1.1-2  httr_1.4.1          numDeriv_2016.8-1.1
-#> [10] tools_4.0.0         backports_1.1.6     R6_2.4.1           
-#> [13] rpart_4.1-15        Hmisc_4.4-0         DBI_1.1.0          
-#> [16] mgcv_1.8-31         nnet_7.3-13         withr_2.2.0        
-#> [19] gridExtra_2.3       tidyselect_1.0.0    compiler_4.0.0     
-#> [22] quantreg_5.55       cli_2.0.2           rvest_0.3.5        
-#> [25] htmlTable_1.13.3    SparseM_1.78        xml2_1.3.2         
-#> [28] sandwich_2.5-1      labeling_0.3        checkmate_2.0.0    
-#> [31] scales_1.1.0        mvtnorm_1.1-0       polspline_1.1.17   
-#> [34] digest_0.6.25       foreign_0.8-78      rmarkdown_2.1      
-#> [37] base64enc_0.1-3     rticles_0.14        jpeg_0.1-8.1       
-#> [40] pkgconfig_2.0.3     htmltools_0.4.0     dbplyr_1.4.3       
-#> [43] htmlwidgets_1.5.1   rlang_0.4.5         readxl_1.3.1       
-#> [46] rstudioapi_0.11     VGAM_1.1-2          farver_2.0.3       
-#> [49] shape_1.4.4         generics_0.0.2      zoo_1.8-7          
-#> [52] jsonlite_1.6.1      acepack_1.4.1       Formula_1.2-3      
-#> [55] Rcpp_1.0.4.6        munsell_0.5.0       fansi_0.4.1        
-#> [58] lifecycle_0.2.0     multcomp_1.4-13     stringi_1.4.6      
-#> [61] yaml_2.2.1          MASS_7.3-51.5       plyr_1.8.6         
-#> [64] ggstance_0.3.4      grid_4.0.0          crayon_1.3.4       
-#> [67] lattice_0.20-41     haven_2.2.0         hms_0.5.3          
-#> [70] timereg_1.9.4       knitr_1.28          pillar_1.4.3       
-#> [73] codetools_0.2-16    stats4_4.0.0        reprex_0.3.0       
-#> [76] glue_1.4.0          evaluate_0.14       latticeExtra_0.6-29
-#> [79] modelr_0.1.6        vctrs_0.2.4         png_0.1-7          
-#> [82] foreach_1.5.0       MatrixModels_0.4-1  cellranger_1.1.0   
-#> [85] gtable_0.3.0        assertthat_0.2.1    xfun_0.13          
-#> [88] viridisLite_0.3.0   iterators_1.0.12    cluster_2.1.0      
-#> [91] lava_1.6.7          TH.data_1.0-10      ellipsis_0.3.0
+#>  [1] TH.data_1.0-10      VGAM_1.1-2          ellipsis_0.3.0     
+#>  [4] ggstance_0.3.4      htmlTable_1.13.3    base64enc_0.1-3    
+#>  [7] fs_1.4.1            rstudioapi_0.11     farver_2.0.3       
+#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-0      
+#> [13] lubridate_1.7.8     xml2_1.3.2          codetools_0.2-16   
+#> [16] knitr_1.28          Formula_1.2-3       jsonlite_1.6.1     
+#> [19] cluster_2.1.0       dbplyr_1.4.3        png_0.1-7          
+#> [22] compiler_4.0.0      httr_1.4.1          backports_1.1.6    
+#> [25] assertthat_0.2.1    cli_2.0.2           acepack_1.4.1      
+#> [28] htmltools_0.4.0     quantreg_5.55       tools_4.0.0        
+#> [31] gtable_0.3.0        glue_1.4.0          Rcpp_1.0.4.6       
+#> [34] cellranger_1.1.0    vctrs_0.2.4         nlme_3.1-147       
+#> [37] iterators_1.0.12    xfun_0.13           rvest_0.3.5        
+#> [40] lifecycle_0.2.0     polspline_1.1.17    MASS_7.3-51.5      
+#> [43] zoo_1.8-7           scales_1.1.0        hms_0.5.3          
+#> [46] sandwich_2.5-1      SparseM_1.78        RColorBrewer_1.1-2 
+#> [49] rticles_0.14        yaml_2.2.1          gridExtra_2.3      
+#> [52] rms_5.1-4           rpart_4.1-15        latticeExtra_0.6-29
+#> [55] stringi_1.4.6       foreach_1.5.0       checkmate_2.0.0    
+#> [58] lava_1.6.7          shape_1.4.4         rlang_0.4.5        
+#> [61] pkgconfig_2.0.3     evaluate_0.14       lattice_0.20-41    
+#> [64] htmlwidgets_1.5.1   labeling_0.3        cmprsk_2.2-9       
+#> [67] cowplot_1.0.0       tidyselect_1.0.0    plyr_1.8.6         
+#> [70] R6_2.4.1            generics_0.0.2      Hmisc_4.4-0        
+#> [73] multcomp_1.4-13     DBI_1.1.0           pillar_1.4.3       
+#> [76] haven_2.2.0         foreign_0.8-78      withr_2.2.0        
+#> [79] mgcv_1.8-31         nnet_7.3-13         modelr_0.1.6       
+#> [82] crayon_1.3.4        rmarkdown_2.1       timereg_1.9.4      
+#> [85] jpeg_0.1-8.1        grid_4.0.0          readxl_1.3.1       
+#> [88] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
+#> [91] numDeriv_2016.8-1.1 stats4_4.0.0        munsell_0.5.0      
+#> [94] viridisLite_0.3.0
 \end{CodeOutput}
@@
 #> Local:    editMax /Users/maxturgeon/Documents/git_repos/cbpaper
-#> Head:     [70115f2] 2020-04-22: Merge pull request #20 from Jesse-Islam/master
+#> Head:     [c6107b6] 2020-05-05: edit jesse's section
 \end{CodeOutput}

### 2020-05-07 analysis/paper/paper.Rmd
edit remainder of paper

@@
 options(kableExtra.latex.load_packages = FALSE)
+library(casebase)
+library(colorspace)
+library(cowplot)
+library(data.table)
+library(glmnet)
+library(kableExtra)
+library(magrittr)
+library(pracma)
+library(prodlim)
+library(riskRegression)
+library(splines)
+library(survival)
+library(tibble)
 library(tidyverse)
-library(magrittr)
-library(survival)
-library(casebase)
-# devtools::load_all('../../../casebase')#testing
-library(splines)
-library(tibble)
-library(glmnet)
-library(data.table)
-library(kableExtra)
-library(pracma)
-library(riskRegression)
-library(prodlim)
-library(colorspace)
 options(digits = 2, scipen = 999)
@@
 eval_implementation <- TRUE
-eval_cs1 <- FALSE
-eval_cs2 <- FALSE
+eval_cs1 <- TRUE
+eval_cs2 <- TRUE
 eval_cs3 <- TRUE
-eval_cs4 <- FALSE
+eval_cs4 <- TRUE
+eval_colophon <- TRUE
+
+echo_plot_code <- TRUE
 ```
@@
 Therefore, to make sound clinical decisions, it is important to accurately estimate the *full* hazard function, which can then be used to estimate the cumulative incidence function (CIF).
-From this perspective, the CIF and the survival function estimates are now smooth functions of time.
+Since the parametric hazard is a smooth function of time, the CIF and the survival function estimates also vary smoothly over time.
 
@@
 Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function.
+In addition, we provide additional features, including performance metrics and visualization tools.
 
@@
 We note that there is only a finite number of person moments associated with the event of interest (what Hanley & Miettinen call the *case series*).
-The case-base sampling refers to the sampling from the base of a representative finite sample called the *base series*.
+Case-base sampling refers to the sampling from the base of a representative finite sample called the *base series*.
 
@@
 In Hanley & Miettinen [-@hanley2009fitting], the authors suggest sampling the base series *uniformly* from the study base.
-In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with hazard equal to $\rho_i(t) = b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base.
-More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism are proportional to the cardiovascular disease event rate given by the Framingham score.
+In terms of Poisson processes, their sampling strategy corresponds essentially to a time-homogeneous Poisson process with hazard equal to $\rho_i(t) = b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base (e.g.
+the sum of all individual follow-up times).
+More complex examples are also possible; see for example Saarela & Arjas [-@saarela2015non], where the intensity functions for the sampling mechanism are proportional to the cardiovascular disease event rate given by the Framingham score.
 With this sampling mechanism, Saarela & Arjas are able to increase the efficiency of their estimators, when compared to uniform sampling.
@@
 
-The instantaneous failure rate is independent of $t$.\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
+In this model, the instantaneous failure rate is independent of $t$.\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
 This is also known as the \textit{memoryless property} \citep{kalbfleisch2011statistical}.}
@@
 
-For competing-risk analyses with $J$ possible events, we can show that each person-moment's contribution of the likelihood is of the form
+Case-base sampling can also be used in the context of competing-risk analyses.
+Assuming there are $J$ competing events, we can show that each person-moment's contribution of the likelihood is of the form
 
@@
 Survival analysis is an important branch of applied statistics and epidemiology.
-Accordingly, there is a vast ecosystem of \code{R} packages implementing different methods.
+Accordingly, there is already a vast ecosystem of \code{R} packages implementing different methodologies.
 In this section, we describe how the functionalities of \pkg{casebase} compare to these packages.
@@
 At the time of writing, a cursory examination of CRAN's task view on survival analysis reveals that there are over 250 packages related to survival analysis [-@survTaskView].
-For the purposes of this article, we restricted our description to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and cumulative incidence curve estimation.
+For the purposes of this article, we restricted our review to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and CIF estimation.
 By searching for appropriate keywords in the \code{DESCRIPTION} file of these packages, we found 60 relevant packages.
@@
 
-Several packages implement penalized estimation for the Cox model: \pkg{glmnet} [-@regpathcox], \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche].
-Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \pkg{CoxRidge} [-@perperoglou] and \pkg{rstpm2} [-@clements_liu], while \pkg{survival} [-@survival-package] has an implementation for ridge.
-\pkg{casebase} provides penalized estimation in the context of parametric hazards.
-To our knowledge, \pkg{casebase} and \pkg{rsptm2} are the only packages to offer this functionality.
-
-Parametric survival models are implemented in a handful of packages: \pkg{CFC} [-@mahani2015bayesian], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu], \pkg{mets} [-@scheike2014estimating], and \pkg{survival}.
+Parametric survival models are implemented in a handful of packages: \pkg{CFC} [-@mahani2015bayesian], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu], \pkg{mets} [-@scheike2014estimating], and \pkg{survival} [-@survival-package].
 The types of models they allow vary for each package.
@@
 Moreover, \pkg{flexsurv} has the ability to estimate both scale and shape parameters for a variety of parametric families (see Table \ref{tab:dists}).
-As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear model of covariates (including time).
-Therefore, our package allows the user to model the effect of time using splines, and through interaction terms involving covariates and time, it also allows user to fit time-varying coefficient models.
+As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear combination of covariates (including time).
+Therefore, our package allows the user to model the effect of time using splines.
+Also, by including interaction terms between covariates and time, it also allows users to fit time-varying coefficient models.
 However, we do not explicitly model any shape parameter, unlike \pkg{flexsurv}.
 
-Several \proglang{R} packages implement methodologies for competing risk analysis; for a different perspective, see Mahani & Sharabiani [-@mahani2015bayesian].
+Several packages implement penalized estimation for the Cox model: \pkg{glmnet} [-@regpathcox], \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche].
+Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \pkg{CoxRidge} [-@perperoglou] and \pkg{rstpm2} [-@clements_liu], while \pkg{survival} [-@survival-package] has an implementation of ridge-penalized estimation.
+On the other hand, our package \pkg{casebase} provides penalized estimation in the context of parametric hazards.
+To our knowledge, \pkg{casebase} and \pkg{rsptm2} are the only packages to offer this functionality.
+
+Next, several \proglang{R} packages implement methodologies for competing risk analysis; for a different perspective on this topic, see Mahani & Sharabiani [-@mahani2015bayesian].
 The package \pkg{cmprsk} provides methods for cause-specific subdistributions, such as in the Fine-Gray model [-@fine1999proportional].
-On the other hand, the package \pkg{CFC} estimates cause-specific cumulative incidence functions from unadjusted, non-parametric survival functions.
-Our package \pkg{casebase} also provides functionalities for competing risk analysis by estimating the cause-specific hazards, and from these quantities, we can derive the cause-specific cumulative incidence functions.
-
-Finally, several packages include functions to estimate the cumulative incidence function.
+On the other hand, the package \pkg{CFC} estimates cause-specific CIFs from unadjusted, non-parametric survival functions.
+Our package \pkg{casebase} also provides functionalities for competing risk analysis by estimating parametrically the cause-specific hazards.
+From these quantities, we can then estimate the cause-specific CIFs.
+
+Finally, several packages include functions to estimate the CIF.
 The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard.
 The first category broadly corresponds to parametric survival models, where the full hazard is explictly modeled.
-Using this estimate, the survival function and the cumulative incidence function can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below).
-Packages including this functionality include \pkg{CFC}, \pkg{Flexsurv}, \pkg{mets}, and \pkg{survival}.
-Our package \pkg{casebase} also follows this approach for both single-event and competing-event analyses.
-The second category outlined above broadly corresponds to Cox models.
+Using this estimate, the survival function and the CIF can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below).
+Packages providing this functionality include \pkg{CFC}, \pkg{Flexsurv}, \pkg{mets}, and \pkg{survival}.
+Our package \pkg{casebase} also follows this approach for both single-event and competing-risk analyses.
+The second category outlined above broadly corresponds to semi-parametric models.
 These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function.
 This is achieved using semi-parametric estimators (e.g.
-Breslow's estimator).
+Breslow's estimator) or parametric estimators (e.g.
+spline functions).
 Packages that implement this approach include \pkg{RiskRegression}, \pkg{rstpm2}, and \pkg{survival}.
-As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category produces estimates in the form of step-wise functions.
-This was one of the main motivations for introducting case-base sampling in survival analysis.
+As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category often produces estimates in the form of step-wise functions.
+Providing smooth estimates of the CIF was one of the main motivations for introducting case-base sampling in survival analysis.
 
@@
   2. estimation of the smooth hazard function; 
-  3. calculation of the risk function.
-
-By separating the sampling and estimation functions, we allowed the possibility of users implementing more complex sampling scheme, as described in Saarela [-@saarela2016case].
-
-The sampling scheme selected for \code{sampleCaseBase} was described in Hanley & Miettinen [-@hanley2009fitting]: we first sample along the ``person'' axis, proportional to each individual's total follow-up time, and then we sample a moment uniformly over their follow-up time.
+  3. estimation of the CIF.
+
+By separating the sampling and estimation functions, we allow the possibility of users implementing more complex sampling scheme (as described in Saarela [-@saarela2016case]), or more complex study designs (e.g.
+time-varying exposure).
+
+The sampling scheme selected for \code{sampleCaseBase} was described in Hanley & Miettinen [-@hanley2009fitting]: we first sample along the "person" axis, proportional to each individual's total follow-up time, and then we sample a moment uniformly over their follow-up time.
 This sampling scheme is equivalent to the following picture: imagine representing the total follow-up time of all individuals in the study along a single dimension, where the follow-up time of the next individual would start exactly when the follow-up time of the previous individual ends.
@@
 
-The fitting function \code{fitSmoothHazard} starts by looking at the class of the dataset: if it was generated from \code{sampleCaseBase}, it automatically inherited the class \code{cbData}.
+Next, the fitting function \code{fitSmoothHazard} starts by looking at the class of the dataset: if it was generated from \code{sampleCaseBase}, it automatically inherited the class \code{cbData}.
 If the dataset supplied to \code{fitSmoothHazard} does not inherit from \code{cbData}, then the fitting function starts by calling \code{sampleCaseBase} to generate the base series.
@@
   - \code{glm}: This is the familiar logistic regression.
-  - \code{glmnet}: This option allows for variable selection using Lasso or elastic-net. This functionality is provided through the \code{glmnet} package [@friedman2010jss].
+  - \code{glmnet}: This option allows for variable selection using Lasso or elastic-net (cf. Case Study 3). This functionality is provided through the \code{glmnet} package [@friedman2010jss].
   - \code{gam}: This option provides support for *Generalized Additive Models* via the \code{gam} package [@hastie1987generalized].
@@
 
-We created \code{absoluteRisk} as an \code{S3} generic, with methods that match the different types of outputs of \code{fitSmoothHazard}.
-The method dispatch system of \proglang{R} then takes care of matching the correct output to the correct methodology for calculating the cumulative incidence function, without requiring any intervention from the user.
+<!--We created \code{absoluteRisk} as an \code{S3} generic, with methods that match the different types of outputs of \code{fitSmoothHazard}.
+The method dispatch system of \proglang{R} then takes care of matching the correct output to the correct methodology for calculating the cumulative incidence function, without requiring any intervention from the user.-->
 
@@
 ```{r plot-stratified-erspc-data-side, eval = eval_cs1}
-plot(pt_object_strat, ncol = 2)
+plot(pt_object_strat, facet.params = list(ncol = 2))
 ```
@@
 
-```{r erspc-cox-cif, eval = eval_cs1}
+```{r erspc-cox-cif, eval = eval_cs1, echo = echo_plot_code}
 new_data <- data.frame(
@@
 
-```{r erspc-casebase-cif, eval = eval_cs1}
+```{r erspc-casebase-cif, eval = eval_cs1, echo = echo_plot_code}
 smooth_risk_exp <- casebase::absoluteRisk(
@@
 
-```{r erspc-casebase-weibull-cif, eval = eval_cs1}
+```{r erspc-casebase-weibull-cif, eval = eval_cs1, echo = echo_plot_code}
 smooth_risk_time <- casebase::absoluteRisk(
@@
 
-```{r erspc-casebase-splines-cif, eval = eval_cs1}
+```{r erspc-casebase-splines-cif, eval = eval_cs1, echo = echo_plot_code}
 smooth_risk_splines <- absoluteRisk(
@@
 ```{r erscp-aic, eval = eval_cs1}
-c(
-  "Exp." = AIC(casebase_exponential),
+c("Exp." = AIC(casebase_exponential),
   "Weib." = AIC(casebase_time),
-  "Splines" = AIC(casebase_splines)
-)
+  "Splines" = AIC(casebase_splines))
 ```
@@
 ```{r bmtcrr-data, eval = eval_cs2, message = FALSE}
-
 data(bmtcrr)
@@
 B) The blue points represent the competing events.", echo=FALSE, eval = eval_cs2}
-library(cowplot)
 pt_object <- casebase::popTime(bmtcrr, event = "Status", time = "ftime")
@@
 plot_grid(plot(pt_object) + ggtitle("Relapse"),
-  plot(pt_object_comp, point.colour = "blue") +
-    ggtitle("Competing events"),
-  labels = c("A", "B")
-)
+          plot(pt_object_comp, point.colour = "blue") + ggtitle("Competing events"),
+          labels = c("A", "B"))
 ```
@@
                         const(Phase) + const(Source) + const(Age),
-data = bmtcrr, cause = 1, model = "fg"
-)
+                      data = bmtcrr, cause = 1, model = "fg")
 
@@
 The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.", eval = eval_cs2}
-library(ggplot2)
 risk_all <- dplyr::bind_rows(
@@
 
-```{r supportCB_fit, eval=eval_cs3, echo=TRUE, cache=TRUE}
+```{r supportCB_fit, eval=eval_cs3, echo=TRUE, cache=FALSE}
 x <- model.matrix(death ~ . - d.time - aps - sps, 
@@
 # Compute AIC
-c(
-  "Model1" = AIC(fit1),
+c("Model1" = AIC(fit1),
   "Model2" = AIC(fit2),
   "Model3" = AIC(fit3),
-  "Model4" = AIC(fit4)
-)
+  "Model4" = AIC(fit4))
 ```
@@
 
-In this article, we presented the \proglang{R} package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating cumulative incidence functions using case-base sampling.
+In this article, we presented the \proglang{R} package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating CIFs using case-base sampling.
 We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through four case studies.
@@
 
-As presented in Hanley & Miettinen [-@hanley2009fitting], case-base sampling is comprised of three steps: 1) sampling a case series and a base series from the study; 2) fit the log-hazard as a linear function of predictors (including time); and 3) use the fitted hazard to estimate the cumulative incidence curve.
+As presented in Hanley & Miettinen [-@hanley2009fitting], case-base sampling is comprised of three steps: 1) sampling a case series and a base series from the study; 2) fit the log-hazard as a linear function of predictors (including time); and 3) use the fitted hazard to estimate the CIF.
 Accordingly, our package provides functions for each step.
@@
 In terms of function interface, \pkg{casebase} uses a formula interface that is closer to that of \code{glm}, in that the event variable is the only variable appearing on the left-hand side of the formula.
-By contrast, both \code{survival::coxph} and \code{timereg::comp.risk} use arrays that captures both the event type and time.
+By contrast, both \code{survival::coxph} and \code{timereg::comp.risk} use arrays that capture both the event type and time.
 Both approaches to modeling yield user-friendly code.
@@
 where $\alpha,\beta$ are shape and scale parameters, respectively.
-Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} allows the user to model the scale parameter as a function of covariates.
+Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} also allows the user to model the scale parameter as a function of covariates.
 Of course, this added flexibility comes at the cost of interpretability: by modeling the log-hazard directly, the parameter estimates from \pkg{casebase} can be interpreted as estimates of log-hazard ratios.
@@
 The delta method or parametric bootstrap are two different strategies we can use to construct approximate confidence intervals.
-Furthermore, we would like to include functions to compute calibration and discrimination statistics (e.g.
-Brier score, AUC) for our models.
+Furthermore, we would like to include more functions to compute calibration and discrimination statistics (e.g.
+AUC) for our models.
 Saarela and Arjas [-@saarela2015non] also describe how to obtain a posterior distribution for the AUC from their model.
-Their approach could also be included in \pkg{casebase}.
+Their approach could potentially be included in \pkg{casebase}.
 Finally, we want to provide more flexibility in how the case-base sampling is performed.
@@
 
-```{r colophon, cache = FALSE}
+```{r colophon, eval = eval_colophon, cache = FALSE}
 # which R packages and versions?
@@
 The current Git commit details are:
-```{r}
+```{r, eval = eval_colophon}
 # what commit is this file at?

### 2020-05-07 analysis/paper/paper.tex

--- 2020-04-22 analysis/paper/paper.tex
@@
 function, which can then be used to estimate the cumulative incidence
-function (CIF). From this perspective, the CIF and the survival function
-estimates are now smooth functions of time.
+function (CIF). Since the parametric hazard is a smooth function of
+time, the CIF and the survival function estimates also vary smoothly
+over time.
 
@@
 retains the flexibility of case-base sampling and the familiar interface
-of the \code{glm} function.
+of the \code{glm} function. In addition, we provide additional features,
+including performance metrics and visualization tools.
 
@@
 finite number of person moments associated with the event of interest
-(what Hanley \& Miettinen call the \emph{case series}). The case-base
+(what Hanley \& Miettinen call the \emph{case series}). Case-base
 sampling refers to the sampling from the base of a representative finite
@@
 suggest sampling the base series \emph{uniformly} from the study base.
-In terms of Poisson processes, this sampling strategy corresponds
+In terms of Poisson processes, their sampling strategy corresponds
 essentially to a time-homogeneous Poisson process with hazard equal to
@@
 in the base series, and \(B\) is the total population-time for the study
-base. More complex examples are also available; see for example Saarela
-\& Arjas \citeyearpar{saarela2015non}, where the probabilities of the
+base (e.g.~the sum of all individual follow-up times). More complex
+examples are also possible; see for example Saarela \& Arjas
+\citeyearpar{saarela2015non}, where the intensity functions for the
 sampling mechanism are proportional to the cardiovascular disease event
@@
 
-The instantaneous failure rate is independent of
+In this model, the instantaneous failure rate is independent of
 \(t\).\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
@@
 
-For competing-risk analyses with \(J\) possible events, we can show that
+Case-base sampling can also be used in the context of competing-risk
+analyses. Assuming there are \(J\) competing events, we can show that
 each person-moment's contribution of the likelihood is of the form
@@
 Survival analysis is an important branch of applied statistics and
-epidemiology. Accordingly, there is a vast ecosystem of \code{R}
-packages implementing different methods. In this section, we describe
-how the functionalities of \pkg{casebase} compare to these packages.
+epidemiology. Accordingly, there is already a vast ecosystem of \code{R}
+packages implementing different methodologies. In this section, we
+describe how the functionalities of \pkg{casebase} compare to these
+packages.
 
@@
 survival analysis \citeyearpar{survTaskView}. For the purposes of this
-article, we restricted our description to packages that implement at
-least one of the following features: parametric modeling,
-non-proportional hazard models, competing risk analysis, penalized
-estimation, and cumulative incidence curve estimation. By searching for
-appropriate keywords in the \code{DESCRIPTION} file of these packages,
-we found 60 relevant packages. These 60 packages were then manually
-examined to determine which ones are comparable to \pkg{casebase}. In
-particular, we excluded packages that were focused on a different set of
-problems, such as frailty and multi-state models. The remaining 14
-packages appear in Table \ref{tab:surv-pkgs}, along with some of the
-functionalities they offer.
+article, we restricted our review to packages that implement at least
+one of the following features: parametric modeling, non-proportional
+hazard models, competing risk analysis, penalized estimation, and CIF
+estimation. By searching for appropriate keywords in the
+\code{DESCRIPTION} file of these packages, we found 60 relevant
+packages. These 60 packages were then manually examined to determine
+which ones are comparable to \pkg{casebase}. In particular, we excluded
+packages that were focused on a different set of problems, such as
+frailty and multi-state models. The remaining 14 packages appear in
+Table \ref{tab:surv-pkgs}, along with some of the functionalities they
+offer.
+
+Parametric survival models are implemented in a handful of packages:
+\pkg{CFC} \citeyearpar{mahani2015bayesian}, \pkg{flexsurv}
+\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
+\pkg{rsptm2} \citeyearpar{clements_liu}, \pkg{mets}
+\citeyearpar{scheike2014estimating}, and \pkg{survival}
+\citeyearpar{survival-package}. The types of models they allow vary for
+each package. For example, \pkg{SmoothHazard} is limited to Weibull
+distributions \citeyearpar{smoothHazard}, whereas both \pkg{flexsurv}
+and \pkg{survival} allow users to supply any distribution of their
+choice. Also, \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and
+\pkg{rstpm2} also have the ability to model the effect of time using
+splines, which allows flexible modeling of the hazard function.
+Moreover, \pkg{flexsurv} has the ability to estimate both scale and
+shape parameters for a variety of parametric families (see Table
+\ref{tab:dists}). As discussed above, \pkg{casebase} can model any
+parametric family whose log-hazard can be expressed as a linear
+combination of covariates (including time). Therefore, our package
+allows the user to model the effect of time using splines. Also, by
+including interaction terms between covariates and time, it also allows
+users to fit time-varying coefficient models. However, we do not
+explicitly model any shape parameter, unlike \pkg{flexsurv}.
 
@@
 \citeyearpar{clements_liu}, while \pkg{survival}
-\citeyearpar{survival-package} has an implementation for ridge.
-\pkg{casebase} provides penalized estimation in the context of
-parametric hazards. To our knowledge, \pkg{casebase} and \pkg{rsptm2}
-are the only packages to offer this functionality.
-
-Parametric survival models are implemented in a handful of packages:
-\pkg{CFC} \citeyearpar{mahani2015bayesian}, \pkg{flexsurv}
-\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
-\pkg{rsptm2} \citeyearpar{clements_liu}, \pkg{mets}
-\citeyearpar{scheike2014estimating}, and \pkg{survival}. The types of
-models they allow vary for each package. For example, \pkg{SmoothHazard}
-is limited to Weibull distributions \citeyearpar{smoothHazard}, whereas
-both \pkg{flexsurv} and \pkg{survival} allow users to supply any
-distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard},
-\pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of
-time using splines, which allows flexible modeling of the hazard
-function. Moreover, \pkg{flexsurv} has the ability to estimate both
-scale and shape parameters for a variety of parametric families (see
-Table \ref{tab:dists}). As discussed above, \pkg{casebase} can model any
-parametric family whose log-hazard can be expressed as a linear model of
-covariates (including time). Therefore, our package allows the user to
-model the effect of time using splines, and through interaction terms
-involving covariates and time, it also allows user to fit time-varying
-coefficient models. However, we do not explicitly model any shape
-parameter, unlike \pkg{flexsurv}.
-
-Several \proglang{R} packages implement methodologies for competing risk
-analysis; for a different perspective, see Mahani \& Sharabiani
-\citeyearpar{mahani2015bayesian}. The package \pkg{cmprsk} provides
-methods for cause-specific subdistributions, such as in the Fine-Gray
-model \citeyearpar{fine1999proportional}. On the other hand, the package
-\pkg{CFC} estimates cause-specific cumulative incidence functions from
+\citeyearpar{survival-package} has an implementation of ridge-penalized
+estimation. On the other hand, our package \pkg{casebase} provides
+penalized estimation in the context of parametric hazards. To our
+knowledge, \pkg{casebase} and \pkg{rsptm2} are the only packages to
+offer this functionality.
+
+Next, several \proglang{R} packages implement methodologies for
+competing risk analysis; for a different perspective on this topic, see
+Mahani \& Sharabiani \citeyearpar{mahani2015bayesian}. The package
+\pkg{cmprsk} provides methods for cause-specific subdistributions, such
+as in the Fine-Gray model \citeyearpar{fine1999proportional}. On the
+other hand, the package \pkg{CFC} estimates cause-specific CIFs from
 unadjusted, non-parametric survival functions. Our package
 \pkg{casebase} also provides functionalities for competing risk analysis
-by estimating the cause-specific hazards, and from these quantities, we
-can derive the cause-specific cumulative incidence functions.
-
-Finally, several packages include functions to estimate the cumulative
-incidence function. The corresponding methods generally fall into two
-categories: transformation of the estimated hazard function, and
-semi-parametric estimation of the baseline hazard. The first category
-broadly corresponds to parametric survival models, where the full hazard
-is explictly modeled. Using this estimate, the survival function and the
-cumulative incidence function can be obtained using their functional
-relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below).
-Packages including this functionality include \pkg{CFC}, \pkg{Flexsurv},
-\pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows
-this approach for both single-event and competing-event analyses. The
-second category outlined above broadly corresponds to Cox models. These
-models do not model the full hazard function, and therefore the baseline
-hazard needs to be estimated separately in order to estimate the
-survival function. This is achieved using semi-parametric estimators
-(e.g.~Breslow's estimator). Packages that implement this approach
-include \pkg{RiskRegression}, \pkg{rstpm2}, and \pkg{survival}. As
-mentioned in the introduction, a key distinguishing factor between these
-two approaches is that the first category leads to smooth estimates of
-the cumulative incidence function, whereas the second category produces
-estimates in the form of step-wise functions. This was one of the main
+by estimating parametrically the cause-specific hazards. From these
+quantities, we can then estimate the cause-specific CIFs.
+
+Finally, several packages include functions to estimate the CIF. The
+corresponding methods generally fall into two categories: transformation
+of the estimated hazard function, and semi-parametric estimation of the
+baseline hazard. The first category broadly corresponds to parametric
+survival models, where the full hazard is explictly modeled. Using this
+estimate, the survival function and the CIF can be obtained using their
+functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI}
+below). Packages providing this functionality include \pkg{CFC},
+\pkg{Flexsurv}, \pkg{mets}, and \pkg{survival}. Our package
+\pkg{casebase} also follows this approach for both single-event and
+competing-risk analyses. The second category outlined above broadly
+corresponds to semi-parametric models. These models do not model the
+full hazard function, and therefore the baseline hazard needs to be
+estimated separately in order to estimate the survival function. This is
+achieved using semi-parametric estimators (e.g.~Breslow's estimator) or
+parametric estimators (e.g.~spline functions). Packages that implement
+this approach include \pkg{RiskRegression}, \pkg{rstpm2}, and
+\pkg{survival}. As mentioned in the introduction, a key distinguishing
+factor between these two approaches is that the first category leads to
+smooth estimates of the cumulative incidence function, whereas the
+second category often produces estimates in the form of step-wise
+functions. Providing smooth estimates of the CIF was one of the main
 motivations for introducting case-base sampling in survival analysis.
@@
 \item
-  calculation of the risk function.
+  estimation of the CIF.
 \end{enumerate}
 
-By separating the sampling and estimation functions, we allowed the
-possibility of users implementing more complex sampling scheme, as
-described in Saarela \citeyearpar{saarela2016case}.
+By separating the sampling and estimation functions, we allow the
+possibility of users implementing more complex sampling scheme (as
+described in Saarela \citeyearpar{saarela2016case}), or more complex
+study designs (e.g.~time-varying exposure).
 
@@
 
-The fitting function \code{fitSmoothHazard} starts by looking at the
-class of the dataset: if it was generated from \code{sampleCaseBase}, it
-automatically inherited the class \code{cbData}. If the dataset supplied
-to \code{fitSmoothHazard} does not inherit from \code{cbData}, then the
-fitting function starts by calling \code{sampleCaseBase} to generate the
-base series. In other words, users can bypass \code{sampleCaseBase}
-altogether and only worry about the fitting function
-\code{fitSmoothHazard}.
+Next, the fitting function \code{fitSmoothHazard} starts by looking at
+the class of the dataset: if it was generated from
+\code{sampleCaseBase}, it automatically inherited the class
+\code{cbData}. If the dataset supplied to \code{fitSmoothHazard} does
+not inherit from \code{cbData}, then the fitting function starts by
+calling \code{sampleCaseBase} to generate the base series. In other
+words, users can bypass \code{sampleCaseBase} altogether and only worry
+about the fitting function \code{fitSmoothHazard}.
 
@@
   \code{glmnet}: This option allows for variable selection using Lasso
-  or elastic-net. This functionality is provided through the
-  \code{glmnet} package \citep{friedman2010jss}.
+  or elastic-net (cf.~Case Study 3). This functionality is provided
+  through the \code{glmnet} package \citep{friedman2010jss}.
 \item
@@
 
-We created \code{absoluteRisk} as an \code{S3} generic, with methods
-that match the different types of outputs of \code{fitSmoothHazard}. The
-method dispatch system of \proglang{R} then takes care of matching the
-correct output to the correct methodology for calculating the cumulative
-incidence function, without requiring any intervention from the user.
-
 In the following sections, we illustrate these functionalities in the
@@
 \begin{CodeInput}
-R> plot(pt_object_strat, ncol = 2)
+R> plot(pt_object_strat, facet.params = list(ncol = 2))
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> c(
-R+   "Exp." = AIC(casebase_exponential),
+R> c("Exp." = AIC(casebase_exponential),
 R+   "Weib." = AIC(casebase_time),
-R+   "Splines" = AIC(casebase_splines)
-R+ )
+R+   "Splines" = AIC(casebase_splines))
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> 
 R> data(bmtcrr)
@@
 R+                         const(Phase) + const(Source) + const(Age),
-R+ data = bmtcrr, cause = 1, model = "fg"
-R+ )
+R+                       data = bmtcrr, cause = 1, model = "fg")
 R> 
@@
 parametric models for hazard functions, and estimating the corresponding
-cumulative incidence curves. In the next case study, we explore an
-advanced feature.
-
-For the third case study, we will demonstrate regularization during
-hazard estimation. Since we are using logistic regression, existing
-packages, like \pkg{glmnet} can be leveraged. To demonstrate this
-feature, we examine the Study to Understand Prognoses Preferences
-Outcomes and Risks of Treatment (SUPPORT) dataset
-\citep{knaus1995support}. The SUPPORT dataset tracks death for
-individuals who are considered seriously ill within 5 hospitals in the
-United States of America. Imputation was conducted using the \pkg{mice}
-package in \proglang{R} with default settings \citep{mice}. The imputed
-dataset is available as part of the \pkg{casebase} package. For more
-information please see the SUPPORT data documentation in the package.
-
-we have a total of 9104 observations and 30 variables, two of which are
-SUPPORT day 3 physiology score (sps) and APACHE III day 3 physiology
-score (aps). These scores were developed for a period of 180 days, which
-makes our comparison over 5.56 years exploratory. Therefore, we will
-assess these two models before using either as our baseline, by
-comparing them to a full model using all variables. Also, our single
-imputation introduces bias in the data. As the goal here is purely
-demonstrative, this bias will be consistent with any of the following
-models. As such, their comparison is still valid, though any
-interpretation requires an awareness of said bias. A description of
-covariates and response variables can be found in Table
-\ref{tab:support1} and a breakdown of each categorical variable in Table
-\ref{tab:support2}. The original data was obtained from the Department
-of Biostatistics at Vanderbilt University \citep{harrell_2020}.
+cumulative incidence curves. For the third case study, we show how
+\pkg{casebase} can also be used for variable selection through
+regularized estimation of the hazard function. To do this, we simply
+replace logistic regression with a penalized counterpart. In
+\pkg{casebase}, we achieve this by using the \pkg{glmnet} package
+\citep{friedman2010jss}.
+
+To illustrate this functionality, we use the dataset from the Study to
+Understand Prognoses Preferences Outcomes and Risks of Treatment
+(SUPPORT) \citep{knaus1995support}. The SUPPORT dataset tracks death in
+five American hospitals within individuals who are considered seriously
+ill.~The original data is available online from the Department of
+Biostatistics at Vanderbilt University \citep{harrell_2020}. The cleaned
+and imputed data consists of 9104 observations and 30 variables, and it
+is available as part of the \pkg{casebase} package. For more information
+about this dataset, the reader is encouraged to look at the
+documentation in our package. A description of the variables can be
+found in Table \ref{tab:support1}, and a breakdown of each categorical
+variable appears in Table \ref{tab:support2}.
 
@@
 
-Before generating an absolute risk curve using regularized logistic
-regression, we will explore three options as a baseline comparison. The
-original study compared these scores over a span of 180 days, which may
-not be comparable over the full 5.56 years. Therefore, We will examine
-three cox regression scenarios: sps only, aps only and all covariates
-excluding sps and aps. Next, we will compare each modeling using the
-Brier score \citep{graf1999ass}. The Brier score can demonstrate both
-discrimination and calibration in a single model \citep{graf1999ass} and
-provide a comparable metric between semi-parametric and parametric
-methods in the survival context. For all the following analyses in this
-case study, we will use 80\% of the observations to train our models,
-while 20\% will be used to generate absolute risk curves, all of which
-will be averaged to demonstrate each model's predictive ability. Over
-5.56 years, \ref{fig:baselines} demonstrates that all the cox models
-have similar, if not identical average predictions. Figure
-\ref{fig:brier1} examines the brier scores as they change over time. As
-the full model has the lowest brier score on overall, it will be used in
-the following absolute risk comparison to regularized cox, regularized
-logistic regression and the unadjusted Kaplan-Meier curve.
+In the original study, the authors developed two scores for predicting
+the outcome. These scores are available as part of the data: SUPPORT day
+3 physiology score (\code{sps}) and APACHE III day 3 physiology score
+(\code{aps}). Using these scores, we will explore three Cox regression
+scenarios as a baseline comparison for \pkg{casebase}: \code{sps} only,
+\code{aps} only, and all covariates excluding \code{sps} and \code{aps}.
+We will compare the three models using a time-dependent version of the
+classical Brier score that is adjusted for censoring
+\citep{graf1999ass}. The Brier score can be used to assess both
+discrimination and calibration in a single model. Moreover, it provides
+a metric that can be used to compare parametric and semi-parametric
+survival models. To account for overfitting, we will use 95\% of the
+observations to train our models, while the remaining 5\% will be used
+to compute absolute risk curves and the Brier scores.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> data(support)
+R> # Change time to years
+R> support$d.time <- support$d.time/365.25
+R> 
+R> # Split into test and train
+R> train_index <- sample(nrow(support), 0.95*nrow(support))
+R> test_index <- setdiff(1:nrow(support), train_index)
+R> 
+R> train <- support[train_index,]
+R> test <- support[test_index,]
+\end{CodeInput}
+\end{CodeChunk}
+
+In Figure\ref{fig:brier1} (A), we can see that all Cox models have
+similar average survival curves. On the other hand, Figure
+\ref{fig:brier1} (B) shows that the full model has the lowest Brier
+score overall. For this reason, we will use this model in our comparison
+between regularized Cox regression, Kaplan-Meier estimation, and the
+approach from \pkg{casebase}.
 
@@
 
-\caption{\label{fig:baselines} Absolute risk curves estimated from hazard models.
-Red: APS only hazard was used to generate the absolute risk.
-Green: All covariates aside from APS and SPS.
-Blue: SPS only.
-Purple: unadjusted Kaplan-Meier curve.}\label{fig:baselines}
+\caption{\label{fig:brier1} Comparison of three Cox regression models. (A) Cumulative Incidence curves. (B) Time-dependent Brier scores.}\label{fig:brierpt1}
 \end{figure}
@@
 
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/brierpt1-1}
-
-}
-
-\caption{\label{fig:brier1} Brier scores over survival time.
-Full model (Full) performs best, while the other two (APS only and SPS only) are comparable.
-Red: APS only model was used to generate the brier scores.
-Green: all covariates excluding APS and SPS.
-Blue: SPS only model.}\label{fig:brierpt1}
-\end{figure}
-\end{CodeChunk}
-
-To get the absolute risk following regularized cox regression, we had to
-work around limitations in existing packages. To make use of the
-\pkg{survival} package, we first estimate the hazard coefficients using
-thecoxnet implementation for lasso penalization in \pkg{glmnet}
-\citep{regpathcox}. Next, we fit a cox model using the survival package
-with the selected variables only, from our regularized fit. Finally, we
-replaced the coefficients in the survival object with those from our
-glmnet fit, allowing us to make use of the absolute risk tools available
-in the \pkg{survival} package.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # hazard for cox using glmnet
-R> coxNet <- glmnet::cv.glmnet(
-R>   x = x, y = y, family = "cox", alpha = 1,
-R>   standardize = TRUE
-R> )
-R> # getting all non-zero covariates
-R> nonzero_covariate_cox <- predict(coxNet,
-R>   type = "nonzero",
-R>   s = "lambda.1se"
-R> )
-R> # keeping coefficients for later
-R> nonzero_coef_cox <- coef(coxNet, s = "lambda.1se")
-R> # cutting dataset to only include nonzero covariates
-R> cleanCoxData <- as.data.frame(cbind(y, x[, nonzero_covariate_cox]))
-R> # create a skeleton object
-R> coxNet <- survival::coxph(Surv(time = d.time, event = death) ~ .,
-R>   data = cleanCoxData, x = TRUE
-R> )
-R> # replace coefficients with regularized ones
-R> coxNet$coefficients <- nonzero_coef_cox@x
-\end{CodeInput}
-\end{CodeChunk}
-
-For our penalized logistic regression model, we opted for the natural
-log of time, which assumes a Weibull distribution for our event. As
-casebase makes use of the glmnet package, we interact with the
-fitsmoothhazard.fit function using a matrix interface, where \emph{y}
-contains the time and event variables and \emph{x} contains all
-covariates we would like to include in the model. For the SUPPORT
-dataset, all factor variables were converted into dummy variables. Here,
-we used lasso penalization by setting alpha to 1, leading to our smooth
-hazard function.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # regularized logistic regression to estimate hazard
+For our penalized logistic regression model, we opt for the natural log
+of time; recall that this corresponds to a Weibull distribution. For
+fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which
+is a matrix interface to the function \code{fitSmoothHazard}. In order
+to use this interfact, we first create a matrix \code{y} containing the
+time and event variables, and a matrix \code{x} containing all remaining
+covariates we would like to include in the model space. We used lasso
+penalization by setting \code{alpha = 1}. Moreover, we use the argument
+\code{penalty.factor} to force the time variable into the final model.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> x <- model.matrix(death ~ . - d.time - aps - sps, 
+R+                   data = train)[, -c(1)] # Remove intercept
+R> y <- data.matrix(subset(train, select = c(d.time, death)))
+R> 
+R> # Regularized logistic regression to estimate hazard
 R> cbFull <- casebase::fitSmoothHazard.fit(x, y,
-R>   family = "glmnet",
-R>   time = "d.time", event = "death",
-R>   formula_time = ~ log(d.time), alpha = 1,
-R>   ratio = 100, standardize = TRUE,
-R>   penalty.factor = c(0, rep(1, ncol(x)))
-R> )
-R> 
-R> # Estimating the absolute risk curve using the newData parameter.
+R+   family = "glmnet",
+R+   time = "d.time", event = "death",
+R+   formula_time = ~ log(d.time), alpha = 1,
+R+   ratio = 10, standardize = TRUE,
+R+   penalty.factor = c(0, rep(1, ncol(x)))
+R+ )
+\end{CodeInput}
+\end{CodeChunk}
+
+Next, we can use the object \code{cbFull} to estimate the absolute risk
+curve on the test data. We first create a matrix \code{newx} containing
+the relevant variables. Next, using \code{times}, we specify the time
+points at which the cumulative incidence will be computed. Finally,
+using the parameter \code{s = "lambda.1se"}, we specify at which value
+of the tuning parameter we want to retrieve the coefficient estimates.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Estimating the absolute risk curve using the newdata parameter
+R> newx <- model.matrix(death ~ . - d.time - aps - sps, 
+R+                   data = test)[, -c(1)]
+R> times <- sort(unique(test$d.time))
+R> 
 R> abCbFull <- casebase::absoluteRisk(cbFull,
-R>   time = times,
-R>   newdata = newData,
-R>   s = "lambda.1se",
-R>   method = c("numerical")
-R> )
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/coefplots-1}
-
-}
-
-\caption{\label{fig:cs3lolliPlot} A lollipop plot demonstrating the estimated coefficients between the following three models: Cox (green), penalized Cox (red), penalized logisitic (blue).
-If a coefficient is estimated to be 0, in any model, it is removed from the plot for simplicity.}\label{fig:coefplots}
-\end{figure}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/unnamed-chunk-2-1}
-
-}
-
-\caption{\label{fig:cs3Final} Four absolute risk curves, integrated from the following hazard models: Cox (green), penalized Cox (red), penalized logistic (blue) and unadjusted (Kaplan-Meier, purple).}\label{fig:unnamed-chunk-2}
-\end{figure}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/riskregressionBrier-1}
-
-}
-
-\caption{\label{fig:cs3FinalBrier} 4 curves comparing the brier scores as they change over time between Cox regression (green), penalized Cox regression (red) and penalized logistic regression (blue) and a unadjusted curve (Kaplan-Meier, purple).
-Penalized logistic regression produced the lowest curve.}\label{fig:riskregressionBrier}
-\end{figure}
+R+   time = times,
+R+   newdata = newx,
+R+   s = "lambda.1se",
+R+   method = "numerical"
+R+ )
+\end{CodeInput}
 \end{CodeChunk}
@@
 estimates are found in penalized logistic model, aside from cancer state
-(cano). These values do not inform us of model performance. To do so, we
-take our hazard functions and use them to predict the absolute risk of
+(cano).
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/coefplots-1}
+
+}
+
+\caption{\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models.
+If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity.}\label{fig:coefplots}
+\end{figure}
+\end{CodeChunk}
+
+These values do not inform us of model performance. To do so, we take
+our hazard functions and use them to predict the absolute risk of
 everyone in our test set, over time. The probabilities over time for
 each individual are averaged, resulting in the absolute risk curves in
-figure \ref{fig:cs3Final}. There is separation between the curves, but
-it is minimal. To better demonstrate the differences between the models,
-we compare brier scores over time. \ref{fig:cs3FinalBrier} The Brier
-score is lower in our penalized logistic regression model, compared to
-the other models. This difference, however, is minimal. We can say with
-relative certainty that any of the hazard models outperform an
-unadjusted curve, while penalized logistic regression performs the best
-on this testing dataset.
+Figure \ref{fig:cs3FinalBrier} (A). There is separation between the
+curves, but it is minimal. To better demonstrate the differences in
+performance between the models, we compare Brier scores over time. In
+Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our
+penalized logistic regression model compared to both Cox regression
+models. However, this difference is minimal. On the other hand, all
+three models outperform the Kaplan-Meier model.
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/riskregressionBrier-1}
+
+}
+
+\caption{\label{fig:cs3FinalBrier} Four curves comparing the Cumulative Incidence (A) and the Brier scores (B) as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.}\label{fig:riskregressionBrier}
+\end{figure}
+\end{CodeChunk}
 
 \hypertarget{case-study-4stanford-heart-transplant-data}{%
-\section{Case study 4--Stanford Heart Transplant
+\section{Case study 4---Stanford Heart Transplant
 Data}\label{case-study-4stanford-heart-transplant-data}}
@@
 R> # Compute AIC
-R> c(
-R+   "Model1" = AIC(fit1),
+R> c("Model1" = AIC(fit1),
 R+   "Model2" = AIC(fit2),
 R+   "Model3" = AIC(fit3),
-R+   "Model4" = AIC(fit4)
-R+ )
+R+   "Model4" = AIC(fit4))
 \end{CodeInput}
@@
 #> Model1 Model2 Model3 Model4 
-#>    790    752    737    721
+#>    801    756    718    724
 \end{CodeOutput}
@@
 #>         
-#> 365 0.31
+#> 365 0.26
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> [1] 0.2
+#> [1] 0.18
 \end{CodeOutput}
@@
 which provides functions for fitting smooth parametric hazards and
-estimating cumulative incidence functions using case-base sampling. We
-outlined the theoretical underpinnings of the approach, we provided
-details about our implementation, and we illustrated the merits of the
-approach and the package through four case studies.
+estimating CIFs using case-base sampling. We outlined the theoretical
+underpinnings of the approach, we provided details about our
+implementation, and we illustrated the merits of the approach and the
+package through four case studies.
 
@@
 linear function of predictors (including time); and 3) use the fitted
-hazard to estimate the cumulative incidence curve. Accordingly, our
-package provides functions for each step. Moreover, the simple interface
-of the \code{fittingSmoothHazard} function resembles the \code{glm}
-interface. This interface should look familiar to new users. Our modular
-approach also provides a convenient way to extend our package for new
-sampling or fitting strategies.
+hazard to estimate the CIF. Accordingly, our package provides functions
+for each step. Moreover, the simple interface of the
+\code{fittingSmoothHazard} function resembles the \code{glm} interface.
+This interface should look familiar to new users. Our modular approach
+also provides a convenient way to extend our package for new sampling or
+fitting strategies.
 
@@
 both \code{survival::coxph} and \code{timereg::comp.risk} use arrays
-that captures both the event type and time. Both approaches to modeling
+that capture both the event type and time. Both approaches to modeling
 yield user-friendly code. However, in terms of output, both approaches
@@
 where \(\alpha,\beta\) are shape and scale parameters, respectively.
-Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} allows the
-user to model the scale parameter as a function of covariates. Of
+Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} also allows
+the user to model the scale parameter as a function of covariates. Of
 course, this added flexibility comes at the cost of interpretability: by
@@
 to construct approximate confidence intervals. Furthermore, we would
-like to include functions to compute calibration and discrimination
-statistics (e.g.~Brier score, AUC) for our models. Saarela and Arjas
+like to include more functions to compute calibration and discrimination
+statistics (e.g.~AUC) for our models. Saarela and Arjas
 \citeyearpar{saarela2015non} also describe how to obtain a posterior
-distribution for the AUC from their model. Their approach could also be
-included in \pkg{casebase}. Finally, we want to provide more flexibility
-in how the case-base sampling is performed. This could be achieved by
-adding a \code{hazard} argument to the function \code{sampleCaseBase}.
-In this way, users could specify their own sampling mechanism. For
-example, they could provide a hazard that gives sampling probabilities
-that are proportional to the cardiovascular disease event rate given by
-the Framingham score \citep{saarela2015non}.
+distribution for the AUC from their model. Their approach could
+potentially be included in \pkg{casebase}. Finally, we want to provide
+more flexibility in how the case-base sampling is performed. This could
+be achieved by adding a \code{hazard} argument to the function
+\code{sampleCaseBase}. In this way, users could specify their own
+sampling mechanism. For example, they could provide a hazard that gives
+sampling probabilities that are proportional to the cardiovascular
+disease event rate given by the Framingham score \citep{saarela2015non}.
 
@@
 
-This report was generated on 2020-04-22 10:48:55 using the following
+This report was generated on 2020-05-07 16:35:15 using the following
 computational environment and dependencies:
@@
 #> other attached packages:
-#>  [1] lubridate_1.7.8           broom_0.5.5              
+#>  [1] lubridate_1.7.8           broom_0.5.6              
 #>  [3] dotwhisker_0.5.0          timereg_1.9.4            
-#>  [5] glue_1.4.0                cowplot_1.0.0            
-#>  [7] colorspace_1.4-1          prodlim_2019.11.13       
-#>  [9] riskRegression_2020.03.27 pracma_2.2.9             
-#> [11] kableExtra_1.1.0          data.table_1.12.8        
-#> [13] glmnet_3.0-2              Matrix_1.2-18            
-#> [15] casebase_0.2.1.9001       survival_3.1-12          
-#> [17] magrittr_1.5              forcats_0.5.0            
-#> [19] stringr_1.4.0             dplyr_0.8.5              
-#> [21] purrr_0.3.3               readr_1.3.1              
-#> [23] tidyr_1.0.2               tibble_3.0.0             
-#> [25] ggplot2_3.3.0             tidyverse_1.3.0          
+#>  [5] glue_1.4.0                forcats_0.5.0            
+#>  [7] stringr_1.4.0             dplyr_0.8.5              
+#>  [9] purrr_0.3.4               readr_1.3.1              
+#> [11] tidyr_1.0.2               ggplot2_3.3.0            
+#> [13] tidyverse_1.3.0           tibble_3.0.1             
+#> [15] survival_3.1-12           riskRegression_2020.02.05
+#> [17] prodlim_2019.11.13        pracma_2.2.9             
+#> [19] magrittr_1.5              kableExtra_1.1.0         
+#> [21] glmnet_3.0-2              Matrix_1.2-18            
+#> [23] data.table_1.12.8         cowplot_1.0.0            
+#> [25] colorspace_1.4-1          casebase_0.2.1.9001      
 #> 
@@
 #> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-0      
-#> [13] xml2_1.3.1          codetools_0.2-16    knitr_1.28         
+#> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.28         
 #> [16] Formula_1.2-3       jsonlite_1.6.1      cluster_2.1.0      
-#> [19] dbplyr_1.4.2        png_0.1-7           compiler_3.6.3     
+#> [19] dbplyr_1.4.3        png_0.1-7           compiler_4.0.0     
 #> [22] httr_1.4.1          backports_1.1.6     assertthat_0.2.1   
 #> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.4.0    
-#> [28] quantreg_5.55       tools_3.6.3         gtable_0.3.0       
+#> [28] quantreg_5.55       tools_4.0.0         gtable_0.3.0       
 #> [31] Rcpp_1.0.4.6        cellranger_1.1.0    vctrs_0.2.4        
-#> [34] nlme_3.1-145        iterators_1.0.12    xfun_0.13          
+#> [34] nlme_3.1-147        iterators_1.0.12    xfun_0.13          
 #> [37] rvest_0.3.5         lifecycle_0.2.0     polspline_1.1.17   
@@
 #> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
-#> [58] mets_1.2.7.1        rlang_0.4.5         pkgconfig_2.0.3    
-#> [61] evaluate_0.14       lattice_0.20-41     htmlwidgets_1.5.1  
-#> [64] labeling_0.3        cmprsk_2.2-9        tidyselect_1.0.0   
-#> [67] plyr_1.8.6          R6_2.4.1            generics_0.0.2     
-#> [70] Hmisc_4.4-0         multcomp_1.4-13     DBI_1.1.0          
-#> [73] pillar_1.4.3        haven_2.2.0         foreign_0.8-76     
-#> [76] withr_2.1.2         mgcv_1.8-31         nnet_7.3-13        
-#> [79] modelr_0.1.6        crayon_1.3.4        rmarkdown_2.1      
-#> [82] jpeg_0.1-8.1        grid_3.6.3          readxl_1.3.1       
-#> [85] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
-#> [88] numDeriv_2016.8-1.1 stats4_3.6.3        munsell_0.5.0      
-#> [91] viridisLite_0.3.0
+#> [58] rlang_0.4.5         pkgconfig_2.0.3     evaluate_0.14      
+#> [61] lattice_0.20-41     htmlwidgets_1.5.1   labeling_0.3       
+#> [64] cmprsk_2.2-9        tidyselect_1.0.0    plyr_1.8.6         
+#> [67] R6_2.4.1            generics_0.0.2      Hmisc_4.4-0        
+#> [70] multcomp_1.4-13     DBI_1.1.0           pillar_1.4.3       
+#> [73] haven_2.2.0         foreign_0.8-78      withr_2.2.0        
+#> [76] mgcv_1.8-31         nnet_7.3-13         modelr_0.1.6       
+#> [79] crayon_1.3.4        rmarkdown_2.1       jpeg_0.1-8.1       
+#> [82] grid_4.0.0          readxl_1.3.1        reprex_0.3.0       
+#> [85] digest_0.6.25       webshot_0.5.2       numDeriv_2016.8-1.1
+#> [88] stats4_4.0.0        munsell_0.5.0       viridisLite_0.3.0
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> Local:    master C:/Users/jesse/Documents/gits/cbpaper
-#> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
-#> Head:     [d079224] 2020-04-16: Write up and plots fixed. plot descriptions fixed
+#> Local:    editMax /Users/maxturgeon/Documents/git_repos/cbpaper
+#> Head:     [c9ddc46] 2020-05-07: combine ci and brier plots
 \end{CodeOutput}

### 2020-06-29 analysis/paper/paper.Rmd

@@
 library(pracma)
+library(flexsurv) # for fitting gompertz AFT
 library(prodlim)
@@
 ## Population-time plots
+
+The case-base sampling approach described in Section \ref{theory} can be visualized in the form of a population time plot.
+They are extremely informative graphical displays of survival data and should be one of the first steps in an exploratory data analysis.
+We facilitate this task in the \pkg{casebase} package using the \code{popTime} function and \code{plot} method as follows:
+
+The splitting of these tasks gives the user the flexibility to create population-time plots with the graphics system of their choice using the object created in Step 1, while the object created in Step 2 allows one to leverage the \pkg{ggplot2} family of functions for further customization.
+1.
+The \code{casebase::popTime} function takes as input the original dataset along with the column names corresponding to the timescale, the event status and an exposure group of interest (optional).
+This will create an object of class \code{popTime} (or \code{popTimeExposure} if you specify a value for the exposure argument).
+2. The corresponding \code{plot} methods for the object created in Step 1 can be called to create the population time plot with several options for customizing the aesthetics.
+
+To illustrate these functions we will use the European Randomized Study of Prostate Cancer Screening data which is available through the \pkg{casebase} package.
+The results of this study were published by [@schroder2009screening], and the dataset itself was obtained using the approach described in [@liu2014recovering].
+
+We first create the necessary dataset for producing the population time plots using the \code{casebase::popTime} function.
+In this example we stratify by the treatment group.
+Note that the resulting object stores the exposure variable as an attribute:
+
+We first pass this dataset to the \code{popTime} function.
+If you do not specify the time and event arguments, this function will try to guess them using regular expressions.
+We then generate the plot by using the corresponding \code{plot} method:
+```{r plot-erspc-data, eval = eval_cs1}
+data("ERSPC")
+ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
+                       levels = c(0,1), 
+                       labels = c("Control group", "Screening group"))
+pt_object <- casebase::popTime(ERSPC, 
+                               time = "Follow.Up.Time",
+                               event = "DeadOfPrCa",
+                               exposure = "ScrArm")
+attr(pt_object, "exposure")
+```
+
+<!--The \code{pt_object} is a \code{data.table} containing the original dataset augmented with-->
+We then pass this object to the \code{plot} function which leverages the \pkg{ggplot2} package [@ggplot2] to build population time plots.
+It builds the plot by sequentially adding layers, starting with a layer for the area representing the population time followed by points to show the casebase sampling mechanism.
+This function gives user the flexibility to add any combination of the case.series, base.series and competing events.
+The case series and competing events are sampled at random vertically on the plot in order to visualise the incidence density using the popTime function.
+That is, imagine we draw a vertical line at a specific event time.
+We then plot the point at a randomly sampled y-coordinate along this vertical line.
+This is done to avoid having all points along the upper edge of the plot (because the subjects with the least amount of observation time are plotted at the top of the y-axis).
+By randomly distributing them, we can get a better sense of the inicidence density.
+The base series is sampled horizontally on the plot using the sampleCaseBase function.
+
+One benefit from these plots is that it allows you to see the incidence density.
+This can be seen from the distribution of the red dots in the above plot.
+We can see that more events are observed later on in time.
+Therefore a constant hazard model would not be appropriate in this instance as it would overestimate the cumulative incidence earlier on in time, and underestimate it later on.
+
+The unique _step shape_ of the population time plot is due to the randomization of the Finnish cohorts which were carried out on January 1 of each of the 4 years 1996 to 1999.
+This, coupled with the uniform December 31 2006 censoring date, lead to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up.
+
+It is important to note that the red points are random distributed across the grey area for each time of event.
+That is, imagine we draw a vertical line at a specific event time.
+We then plot the red point at a randomly sampled y-coordinate along this vertical line.
+This is done to avoid having all red points along the upper edge of the plot (because the subjects with the least amount of observation time are plotted at the top of the y-axis).
+By randomly distributing them, we can get a better sense of the inicidence density.
+
+The randomization of the Finnish cohorts were carried out on January 1 of each of the 4 years 1996 to 1999.
+This, coupled with the uniform December 31 2006 censoring date, lead to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up.
+Tracked backwards in time (i.e.
+from right to left), the population-time plot shows the recruitment pattern from its beginning in 1991, and in particular the Jan 1 entries in successive years.
+Tracked forwards in time (i.e.
+from left to right), the plot for the first three years shows attrition due entirely to death (mainly from other causes).
+Since the Swedish and Belgian centres were the last to close their recruitment - in December 2003 - the minimum potential follow-up is three years.
+Tracked further forwards in time (i.e.
+after year 3) the attrition is
+We can also create exposure stratified plots by specifying the \code{exposure} argument in the \code{popTime} function:
+
+```{r plot-stratified-erspc-data, eval = eval_cs1}
+pt_object_strat <- casebase::popTime(ERSPC, 
+                                     event = "DeadOfPrCa", 
+                                     exposure = "ScrArm")
+# class(pt_object_strat)
+# pt_object
+plot(pt_object_strat)
+# dev.off()
+# library(magrittr)
+# pt <- plot(pt_object_strat)
+# pt %>% class
+
+# library(ggplot2)
+# pt + theme_minimal()
+
+```
+
+We can also plot them side-by-side using the \code{ncol} argument:
+
+```{r plot-stratified-erspc-data-side, eval = eval_cs1}
+plot(pt_object_strat, ncol = 2)
+```
+
+ADD PARAGRAPH ABOUT WHAT WE CAN CONCLUDE FROM THESE GRAPHS.
+
+For a more detailed description of how to modify the aesthetics of a population time plot, please refer to the [Customizing Population Time Plots](http://sahirbhatnagar.com/casebase/articles/customizingpopTime.html) vignette.
 
@@
 
-```{r erspc-cox-cif, eval = eval_cs1, echo = echo_plot_code}
+```{r erspc-cox-cif, eval = eval_cs1, echo = echo_plot_code, fig.width=8, fig.height=6, fig.cap="Cumulative incidence functions for control and screening groups in the ERSPC data.
+In each of the panels, we plot the CIF from the Cox model (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression.
+(A) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution.
+(B) Linear function of time which is equivalent to a Gompertz distribution.
+(C) The natural logarithm of time which is equivalent to a Weibull distribution.
+(D) Cubic b-spline expansion of time."}
+cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+myCols <- cbPalette[c(4,7)]
 new_data <- data.frame(
@@
   ignore = 99
+)
+
+fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
+             gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
+             weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
+             splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
+
+fits <- lapply(fmla, function(i) {
+  casebase::fitSmoothHazard(i, data = ERSPC, ratio = 100)
+})
+
+risks <- lapply(fits, function(i) {
+  casebase::absoluteRisk(object = i, time = seq(0,15,0.1), newdata = new_data)
+})
+
+aft_fits <- lapply(list("exp","gompertz","weibull"), function(i) 
+  flexsurv::flexsurvreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
+                           data = ERSPC, dist = i)
+  )
+
+names(aft_fits) <- list("exponential","gompertz","weibull")
+
+# dev.off()
+layout(matrix(1:4, ncol = 2, byrow = FALSE), respect = FALSE)
+par(mar = c(0, 4.1, 4.1, 0))
+plot(survfit(cox_model, newdata = new_data),
+     ylab = "",
+     xlab = "",
+     fun = "event",
+     xlim = c(0,15), conf.int = FALSE, col = myCols, 
+     # main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+     #                Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+     #                exp(coef(cox_model)), 
+     #                exp(confint(cox_model))[1], 
+     #                exp(confint(cox_model))[2]), 
+     lwd = 2,
+     yscale = 100,
+     xaxt = 'n')
+legend("topleft", 
+       legend = c("Control (Cox)","Control (Casebase)",
+                  "Screening (Cox)", "Screening (Casebase)"), 
+       col = rep(myCols, each = 2),
+       lty = c(1, 2, 1, 2), 
+       lwd = 2,
+       cex = 1.1,
+       bg = "gray90")
+lines(risks$exponential[,1], risks$exponential[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risks$exponential[,1], risks$exponential[,3], col = myCols[2], lty = 2, lwd = 2)
+text(x = 15*0.97, y = 0.05/100, labels = "(A)", cex = 1.5)
+
+par(mar = c(4.1, 4.1, 0, 0))
+plot(survfit(cox_model, newdata = new_data),
+     ylab = "", xlab = "",
+     fun = "event",
+     xlim = c(0,15), conf.int = FALSE, col = myCols, 
+     lwd = 2,
+     yscale = 100)
+lines(risks$weibull[,1], risks$weibull[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risks$weibull[,1], risks$weibull[,3], col = myCols[2], lty = 2, lwd = 2)  
+text(x = 15*0.97, y = 0.05/100, labels = "(C)", cex = 1.5)
+mtext("Cumulative incidence (%)", side = 2, line = -1.5, outer = TRUE)
+
+par(mar = c(0, 0, 4.1, 3.9))
+plot(survfit(cox_model, newdata = new_data),
+     ylab = "",xlab = "",
+     fun = "event",
+     xaxt = 'n', yaxt = 'n',
+     xlim = c(0,15), conf.int = FALSE, col = myCols, 
+     lwd = 2,
+     yscale = 100)
+lines(risks$gompertz[,1], risks$gompertz[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risks$gompertz[,1], risks$gompertz[,3], col = myCols[2], lty = 2, lwd = 2)
+text(x = 15*0.97, y = 0.05/100, labels = "(B)", cex = 1.5)
+
+par(mar = c(4.1, 0, 0, 3.9))
+plot(survfit(cox_model, newdata = new_data),
+     ylab = "",xlab = "",
+     fun = "event",
+     yaxt = 'n',
+     xlim = c(0,15), conf.int = FALSE, col = myCols, 
+     lwd = 2,
+     yscale = 100)
+lines(risks$splines[,1], risks$splines[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risks$splines[,1], risks$splines[,3], col = myCols[2], lty = 2, lwd = 2)
+text(x = 15*0.97, y = 0.05/100, labels = "(D)", cex = 1.5)
+mtext("Years since randomization", side = 1, line = -1.5, outer = TRUE)
+```
+
+Next, we show how case-base sampling can be used to carry out a similar analysis. We fit several models that differ in how we choose to model time.
+
+```{r, results='hide'}
+par(mfrow = c(2, 2))
+lapply(fits, function(i) {
+  casebase::hazardPlot(i, newdata = new_data[2,], ci = FALSE)
+  casebase::hazardPlot(i, newdata = new_data[1,], ci = FALSE, add = TRUE, lty = 2)  
+})
+```
+
+First, we will fit an exponential model. Recall that this corresponds to excluding time from the linear predictor.
+
+## Equivalence between casebase and AFT
+
+```{r erspc-casebase-exponential, eval = eval_cs1}
+# exponential -------------------------------------------------------------
+
+casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
+                                                  data = ERSPC, 
+                                                  ratio = 100)
+aft_exponential <- survreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
+                           data = ERSPC, dist = "exp")
+
+-coef(aft_exponential)
+coef(casebase_exponential)
+summary(casebase_exponential)
+exp(coef(casebase_exponential)[2])
+exp(confint(casebase_exponential)[2, ])
+
+# weibull -----------------------------------------------------------------
+
+casebase_weibull <- casebase::fitSmoothHazard(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm, 
+                                           data = ERSPC, 
+                                           ratio = 100)
+aft_weibull <- survreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
+                       data = ERSPC, dist = "weibull")
+
+-coef(aft_weibull)/aft_weibull$scale
+coef(casebase_weibull)
+```
+We can then use the \code{absoluteRisk} function to get an estimate of the cumulative incidence curve for a specific covariate profile.
+In the plot below, we overlay the estimated CIF from the exponential model on the Cox model CIF:
+
+```{r erspc-casebase-cif, eval = eval_cs1, echo = echo_plot_code}
+smooth_risk_exp <- casebase::absoluteRisk(
+  object = casebase_exponential,
+  time = seq(0, 15, 0.1),
+  newdata = new_data
 )
@@
 )
+lines(smooth_risk_exp[, 1], smooth_risk_exp[, 2], col = "red", lty = 2)
+lines(smooth_risk_exp[, 1], smooth_risk_exp[, 3], col = "blue", lty = 2)
+
 legend("topleft",
-  legend = c("Control group", "Screening group"),
-  col = c("red", "blue"),
-  lty = c(1, 1),
+  legend = c(
+    "Control group (Cox)", "Control group (Casebase)",
+    "Screening group (Cox)", "Screening group (Casebase)"
+  ),
+  col = c("red", "red", "blue", "blue"),
+  lty = c(1, 2, 1, 2),
   bg = "gray90"
@@
 ```
+
+As we can see, the exponential model gives us an estimate of the hazard ratio that is similar to the one obtained using Cox regression.
+However, the CIF estimates are quite different.
+Based on what we observed in the population time plot, where more events are observed later on in time, we do not expect a constant hazard would be a good description for this data.
+In other words, this poor fit for the exponential hazard is expected.
+A constant hazard model would overestimate the cumulative incidence earlier on in time, and underestimate it later on; this is what we see on the cumulative incidence plot.
+This example demonstrates the benefits of population time plots as an exploratory analysis tool.
+
+For the next model, we include time as a linear term. Recall that this corresponds to a Weibull model.
+
+```{r erspc-casebase-weibull, eval = eval_cs1}
+casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm,
+  data = ERSPC,
+  ratio = 100
+)
 
@@
 
-```{r erspc-casebase-exponential, eval = eval_cs1}
-casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm,
-  data = ERSPC,
-  ratio = 100
-)
-
-summary(casebase_exponential)
-exp(coef(casebase_exponential)[2])
-exp(confint(casebase_exponential)[2, ])
-```
-
 <!-- The \code{absoluteRisk} function provides an estimate of the cumulative incidence curves for a specific risk profile using the following equation: -->
@@
 
-```{r erspc-casebase-cif, eval = eval_cs1, echo = echo_plot_code}
-smooth_risk_exp <- casebase::absoluteRisk(
-  object = casebase_exponential,
+As we can see, the exponential model gives us an estimate of the hazard ratio that is similar to the one obtained using Cox regression.
+However, the CIF estimates are quite different.
+Based on what we observed in the population time plot, where more events are observed later on in time, we do not expect a constant hazard would be a good description for this data.
+In other words, this poor fit for the exponential hazard is expected.
+A constant hazard model would overestimate the cumulative incidence earlier on in time, and underestimate it later on; this is what we see on the cumulative incidence plot.
+This example demonstrates the benefits of population time plots as an exploratory analysis tool.
+
+For the next model, we include time as a linear term. Recall that this corresponds to a Weibull model.
+
+Again, the estimate of the hazard ratio is similar to that obtained from Cox regression. We then look at the estimate of the CIF:
+
+```{r erspc-casebase-weibull-cif, eval = eval_cs1, echo = echo_plot_code}
+smooth_risk_time <- casebase::absoluteRisk(
+  object = casebase_time,
   time = seq(0, 15, 0.1),
@@
 )
-lines(smooth_risk_exp[, 1], smooth_risk_exp[, 2], col = "red", lty = 2)
-lines(smooth_risk_exp[, 1], smooth_risk_exp[, 3], col = "blue", lty = 2)
+lines(smooth_risk_time[, 1], smooth_risk_time[, 2], col = "red", lty = 2)
+lines(smooth_risk_time[, 1], smooth_risk_time[, 3], col = "blue", lty = 2)
 
@@
 
-As we can see, the exponential model gives us an estimate of the hazard ratio that is similar to the one obtained using Cox regression.
-However, the CIF estimates are quite different.
-Based on what we observed in the population time plot, where more events are observed later on in time, we do not expect a constant hazard would be a good description for this data.
-In other words, this poor fit for the exponential hazard is expected.
-A constant hazard model would overestimate the cumulative incidence earlier on in time, and underestimate it later on; this is what we see on the cumulative incidence plot.
-This example demonstrates the benefits of population time plots as an exploratory analysis tool.
-
-For the next model, we include time as a linear term. Recall that this corresponds to a Weibull model.
-
-```{r erspc-casebase-weibull, eval = eval_cs1}
-casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm,
+We see that the Weibull model leads to a better fit of the CIF than the Exponential model, when compared with the semi-parametric approach.
+
+Finally, we can model the hazard as a smooth function of time using the \code{splines} package:
+
+Once again, we can see that the estimate of the hazard ratio is similar to that from the Cox regression. We then look at the estimate of the CIF:
+
+Qualitatively, the combination of splines and case-base sampling produces the parametric model that most closely resembles the Cox model.
+In the following table, we see that the confidence intervals are also similar across all four models.
+In other words, this reinforces the idea that, under proportional hazards, we do not need to model the full hazard to obtain reliable estimates of the hazard ratio.
+Of course, different parametric models for the hazards give rise to qualitatively different estimates for the CIF.
+
+```{r erspc-95ci, echo = FALSE, eval = eval_cs1}
+# CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
+# CI_exp <- c(exp(coef(fits$exponential))["ScrArmScreening group"], exp(confint(fits$exponential))["ScrArmScreening group",])
+# CI_gompertz <- c(exp(coef(fits$gompertz))["ScrArmScreening group"], exp(confint(fits$gompertz))["ScrArmScreening group",])
+# CI_weibull <- c(exp(coef(fits$weibull))["ScrArmScreening group"], exp(confint(fits$weibull))["ScrArmScreening group",])
+# CI_splines <- c(exp(coef(fits$splines))["ScrArmScreening group"], exp(confint(fits$splines))["ScrArmScreening group",])
+
+# function to exponentiate HR and CI and return a HR (lower, Upper) character
+exp_est_ci <- function(model = c("exponential","gompertz","weibull","splines"),
+                       fit, 
+                       covariate = "ScrArmScreening group", 
+                       type = c("casebase", "survreg")) {
+
+  type <- match.arg(type)
+  covariate <- match.arg(covariate)
+
+  if (type == "casebase"){
+
+    cis <- confint(fit[[model]])
+
+    sprintf("%0.2f (%0.2f, %0.2f)",
+            exp(coef(fit[[model]]))[[covariate]],
+            exp(cis)[covariate,1],
+            exp(cis)[covariate,2])
+  } else {
+
+    if (model == "weibull") {
+
+      # The maximised log-likelihoods are the same between survreg and flexsurvreg, 
+      # however the parameterisation is different: the first coefficient (Intercept) 
+      # reported by survreg is log(mu), and survreg's "scale" is
+      # dweibull’s (thus flexsurvreg)'s 1 / shape
+      res <- -fit[[model]]$res[covariate,] / (1 / fit[[model]]$res["shape",])
+
+      sprintf("%0.2f (%0.2f, %0.2f)",
+              exp(res[["est"]]),
+              exp(res[["U95%"]]),
+              exp(res[["L95%"]]))
+
+    } else if (model == "splines") {
+
+      c("--")
+
+    } else {
+
+      res <- fit[[model]]$res[covariate,]
+
+      sprintf("%0.2f (%0.2f, %0.2f)",
+              exp(res[["est"]]),
+              exp(res[["L95%"]]),
+              exp(res[["U95%"]]))
+    }
+  }
+}
+
+casebase_HR_CI <- lapply(list("exponential","gompertz","weibull","splines"), 
+                         exp_est_ci, 
+                         fit = fits,
+                         covariate = "ScrArmScreening group", 
+                         type = "casebase")
+
+survreg_HR_CI <- lapply(list("exponential","gompertz","weibull","splines"), 
+                         exp_est_ci, 
+                         fit = aft_fits,
+                         covariate = "ScrArmScreening group", 
+                         type = "survreg")
+
+pt <- cbind(list("Exponential","Gompertz","Weibull","Splines"),
+            do.call(rbind, casebase_HR_CI), do.call(rbind, survreg_HR_CI))
+
+# table_ci <- rbind(CI_cox, CI_exp, CI_gompertz, CI_weibull, CI_splines)
+# colnames(table_ci) <- c("HR", "LB", "UB")
+# 
+# data.frame(Model = c("Cox", "Exponential", "Gompertz", "Weibull", "Splines")) %>% 
+#   bind_cols(as.data.frame(table_ci)) %>% 
+#   mutate(CI = paste0("(", round(LB, 2), ", ",
+#                      round(UB, 2), ")"),
+#          Width = UB - LB) %>% 
+#   dplyr::select(-LB, -UB) %>% 
+#   knitr::kable(format = "latex", booktabs = TRUE)  %>%
+#   kable_styling()
+
+```{r erspc-casebase-splines, eval = eval_cs1}
+casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm,
   data = ERSPC,
@@
 
-summary(casebase_time)
-exp(coef(casebase_time))
-exp(confint(casebase_time))
-```
-
-Again, the estimate of the hazard ratio is similar to that obtained from Cox regression. We then look at the estimate of the CIF:
-
-```{r erspc-casebase-weibull-cif, eval = eval_cs1, echo = echo_plot_code}
-smooth_risk_time <- casebase::absoluteRisk(
-  object = casebase_time,
+summary(casebase_splines)
+exp(coef(casebase_splines))
+exp(confint(casebase_splines))
+```
+
+Once again, we can see that the estimate of the hazard ratio is similar to that from the Cox regression. We then look at the estimate of the CIF:
+
+```{r erspc-casebase-splines-cif, eval = eval_cs1, echo = echo_plot_code}
+smooth_risk_splines <- absoluteRisk(
+  object = casebase_splines,
   time = seq(0, 15, 0.1),
@@
 )
-lines(smooth_risk_time[, 1], smooth_risk_time[, 2], col = "red", lty = 2)
-lines(smooth_risk_time[, 1], smooth_risk_time[, 3], col = "blue", lty = 2)
+lines(smooth_risk_splines[, 1], smooth_risk_splines[, 2], col = "red", lty = 2)
+lines(smooth_risk_splines[, 1], smooth_risk_splines[, 3], col = "blue", lty = 2)
 
@@
 
-We see that the Weibull model leads to a better fit of the CIF than the Exponential model, when compared with the semi-parametric approach.
-
-Finally, we can model the hazard as a smooth function of time using the \code{splines} package:
-
-```{r erspc-casebase-splines, eval = eval_cs1}
-casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm,
-  data = ERSPC,
-  ratio = 100
-)
-
-summary(casebase_splines)
-exp(coef(casebase_splines))
-exp(confint(casebase_splines))
-```
-
-Once again, we can see that the estimate of the hazard ratio is similar to that from the Cox regression. We then look at the estimate of the CIF:
-
-```{r erspc-casebase-splines-cif, eval = eval_cs1, echo = echo_plot_code}
-smooth_risk_splines <- absoluteRisk(
-  object = casebase_splines,
-  time = seq(0, 15, 0.1),
-  newdata = new_data
-)
-
-plot(survfit(cox_model, newdata = new_data),
-  xlab = "Years since Randomization",
-  ylab = "Cumulative Incidence (%)",
-  fun = "event",
-  xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
-  main = sprintf(
-    "Estimated Cumulative Incidence (risk) of Death from Prostate 
-                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-    exp(coef(cox_model)),
-    exp(confint(cox_model))[1],
-    exp(confint(cox_model))[2]
-  ),
-  ylim = c(0, 2), yscale = 100
-)
-lines(smooth_risk_splines[, 1], smooth_risk_splines[, 2], col = "red", lty = 2)
-lines(smooth_risk_splines[, 1], smooth_risk_splines[, 3], col = "blue", lty = 2)
-
-legend("topleft",
-  legend = c(
-    "Control group (Cox)", "Control group (Casebase)",
-    "Screening group (Cox)", "Screening group (Casebase)"
-  ),
-  col = c("red", "red", "blue", "blue"),
-  lty = c(1, 2, 1, 2),
-  bg = "gray90"
-)
-```
-
 Qualitatively, the combination of splines and case-base sampling produces the parametric model that most closely resembles the Cox model.
@@
 Of course, different parametric models for the hazards give rise to qualitatively different estimates for the CIF.
+
+```{r, eval = TRUE}
+## ---- print-sim-table ----
+
+kable(pt, "latex", booktabs = T, align = c("l","c","c"),
+      caption = c("Mean "),
+      col.names = c("Model","casebase","survreg")) %>%
+  kable_styling() %>%
+  # add_header_above(c(" "," ","No overlap" = 2, "All causal SNPs\nin kinship" = 2, "No overlap" = 2, "All causal SNPs\nin kinship" = 2)) %>%
+  # add_header_above(c(" "," ","Null model" = 4, "1% Causal SNPs" = 4)) %>%
+  column_spec(1, bold=T) %>%
+  collapse_rows(columns = 1, latex_hline = "major", valign = "middle") %>%
+  footnote(general = "Median (Inter-quartile range) is given for Model Size.",
+           general_title = "Note:") #%>%
+  # footnote(general = "Model Size is the number of selected variables using the high-dimensional BIC for \texttt{ggmix} and 10-fold cross validation for \texttt{lasso} and \texttt{twostep}.")
+  #          # number = c("", "Footnote 2; "),
+  #          # alphabet = c("Footnote A; ", "Footnote B; "),
+  #          # symbol = c("Footnote Symbol 1; ", "Footnote Symbol 2"),
+  #          general_title = "Note: ", number_title = "Type I: ",
+  #          alphabet_title = "Type II: ", symbol_title = "Type III: ",
+  #          footnote_as_chunk = T, title_format = c("italic", "underline")
+  # )
+```
 
@@
 
+colnames(jasa)[which(colnames(jasa)=="futime")] <- "futime_old"
+colnames(jasa)[which(colnames(jasa)=="age")] <- "futime"
+
 stanford_popTime <- popTime(jasa,

### 2020-06-29 analysis/paper/paper.tex

@@
 \subsection{Population-time plots}\label{population-time-plots}}
+
+The case-base sampling approach described in Section \ref{theory} can be
+visualized in the form of a population time plot. They are extremely
+informative graphical displays of survival data and should be one of the
+first steps in an exploratory data analysis. We facilitate this task in
+the \pkg{casebase} package using the \code{popTime} function and
+\code{plot} method as follows:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  The \code{casebase::popTime} function takes as input the original
+  dataset along with the column names corresponding to the timescale,
+  the event status and an exposure group of interest (optional). This
+  will create an object of class \code{popTime} (or
+  \code{popTimeExposure} if you specify a value for the exposure
+  argument).\\
+\item
+  The corresponding \code{plot} methods for the object created in Step 1
+  can be called to create the population time plot with several options
+  for customizing the aesthetics.
+\end{enumerate}
+
+To illustrate these functions we will use the European Randomized Study
+of Prostate Cancer Screening data which is available through the
+\pkg{casebase} package. The results of this study were published by
+\citep{schroder2009screening}, and the dataset itself was obtained using
+the approach described in \citep{liu2014recovering}.
+
+We first create the necessary dataset for producing the population time
+plots using the \code{casebase::popTime} function. In this example we
+stratify by the treatment group. Note that the resulting object stores
+the exposure variable as an attribute:
+
+\begin{CodeChunk}
+
+\begin{CodeOutput}
+#> [1] "ScrArm"
+\end{CodeOutput}
+\end{CodeChunk}
+
+We then pass this object to the \code{plot} function which leverages the
+\pkg{ggplot2} package \citep{ggplot2} to build population time plots. It
+builds the plot by sequentially adding layers, starting with a layer for
+the area representing the population time followed by points to show the
+casebase sampling mechanism. This function gives user the flexibility to
+add any combination of the case.series, base.series and competing
+events. The case series and competing events are sampled at random
+vertically on the plot in order to visualise the incidence density using
+the popTime function. That is, imagine we draw a vertical line at a
+specific event time. We then plot the point at a randomly sampled
+y-coordinate along this vertical line. This is done to avoid having all
+points along the upper edge of the plot (because the subjects with the
+least amount of observation time are plotted at the top of the y-axis).
+By randomly distributing them, we can get a better sense of the
+inicidence density. The base series is sampled horizontally on the plot
+using the sampleCaseBase function.
+
+One benefit from these plots is that it allows you to see the incidence
+density. This can be seen from the distribution of the red dots in the
+above plot. We can see that more events are observed later on in time.
+Therefore a constant hazard model would not be appropriate in this
+instance as it would overestimate the cumulative incidence earlier on in
+time, and underestimate it later on.
+
+The unique \emph{step shape} of the population time plot is due to the
+randomization of the Finnish cohorts which were carried out on January 1
+of each of the 4 years 1996 to 1999. This, coupled with the uniform
+December 31 2006 censoring date, lead to large numbers of men with
+exactly 11, 10, 9 or 8 years of follow-up.
+
+It is important to note that the red points are random distributed
+across the grey area for each time of event. That is, imagine we draw a
+vertical line at a specific event time. We then plot the red point at a
+randomly sampled y-coordinate along this vertical line. This is done to
+avoid having all red points along the upper edge of the plot (because
+the subjects with the least amount of observation time are plotted at
+the top of the y-axis). By randomly distributing them, we can get a
+better sense of the inicidence density.
+
+We can also create exposure stratified plots by specifying the
+\code{exposure} argument in the \code{popTime} function:
+
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-1} \end{center}
+
+\end{CodeChunk}
+
+We can also plot them side-by-side using the \code{ncol} argument:
+
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-side-1} \end{center}
+
+\end{CodeChunk}
+
+ADD PARAGRAPH ABOUT WHAT WE CAN CONCLUDE FROM THESE GRAPHS.
+
+For a more detailed description of how to modify the aesthetics of a
+population time plot, please refer to the
+\href{http://sahirbhatnagar.com/casebase/articles/customizingpopTime.html}{Customizing
+Population Time Plots} vignette.
 
@@
 \begin{CodeChunk}
+\begin{figure}
 
@@
 
-\begin{center}\includegraphics{../figures/erspc-cox-cif-1} \end{center}
-
-\end{CodeChunk}
+}
+
+\caption[Cumulative incidence functions for control and screening groups in the ERSPC data]{Cumulative incidence functions for control and screening groups in the ERSPC data.
+In each of the panels, we plot the CIF from the Cox model (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression.
+(A) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution.
+(B) Linear function of time which is equivalent to a Gompertz distribution.
+(C) The natural logarithm of time which is equivalent to a Weibull distribution.
+(D) Cubic b-spline expansion of time.}\label{fig:erspc-cox-cif}
+\end{figure}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{../figures/unnamed-chunk-2-1} \end{center}
+
+\end{CodeChunk}
+
+\hypertarget{equivalence-between-casebase-and-aft}{%
+\subsection{Equivalence between casebase and
+AFT}\label{equivalence-between-casebase-and-aft}}
 
@@
 \begin{CodeChunk}
-\begin{table}[H]
+\begin{table}
+
+\caption{\label{tab:unnamed-chunk-4}Mean }
 \centering
-\begin{tabular}{lrlr}
+\begin{tabular}[t]{>{\bfseries}lcc}
 \toprule
-Model & HR & CI & Width\\
+Model & casebase & survreg\\
 \midrule
-Cox & 0.80 & (0.67, 0.95) & 0.28\\
-Exponential & 0.81 & (0.68, 0.96) & 0.28\\
-Weibull & 0.79 & (0.67, 0.94) & 0.28\\
-Splines & 0.81 & (0.68, 0.96) & 0.28\\
+Exponential & 0.81 (0.68, 0.96) & 0.81 (0.68, 0.96)\\
+\cmidrule{1-3}
+Gompertz & 0.79 (0.66, 0.94) & 0.80 (0.67, 0.95)\\
+\cmidrule{1-3}
+Weibull & 0.80 (0.67, 0.95) & 0.80 (0.65, 0.96)\\
+\cmidrule{1-3}
+Splines & 0.80 (0.67, 0.95) & --\\
 \bottomrule
+\multicolumn{3}{l}{\textit{Note:}}\\
+\multicolumn{3}{l}{Median (Inter-quartile range) is given for Model Size.}\\
 \end{tabular}

### 2020-06-29 analysis/paper/paper.Rmd
merge pop-stash

@@
 
-The splitting of these tasks gives the user the flexibility to create population-time plots with the graphics system of their choice using the object created in Step 1, while the object created in Step 2 allows one to leverage the \pkg{ggplot2} family of functions for further customization.
 1.
@@
 
-To illustrate these functions we will use the European Randomized Study of Prostate Cancer Screening data which is available through the \pkg{casebase} package.
+The splitting of these tasks gives the user the flexibility to create population-time plots with the graphics system of their choice using the object created in Step 1, while the object created in Step 2 allows one to leverage the \pkg{ggplot2} [@ggplot2] family of functions for further customization.
+
+To illustrate these functions we will use the European Randomized Study of Prostate Cancer Screening (ERSPC) data which is available through the \pkg{casebase} package.
 The results of this study were published by [@schroder2009screening], and the dataset itself was obtained using the approach described in [@liu2014recovering].
@@
 
-We first pass this dataset to the \code{popTime} function.
-If you do not specify the time and event arguments, this function will try to guess them using regular expressions.
-We then generate the plot by using the corresponding \code{plot} method:
-```{r plot-erspc-data, eval = eval_cs1}
+```{r plot-erspc-data, eval = eval_cs1, echo = TRUE}
 data("ERSPC")
@@
                        labels = c("Control group", "Screening group"))
+
 pt_object <- casebase::popTime(ERSPC, 
@@
 attr(pt_object, "exposure")
+
+pt_object_strat <- casebase::popTime(ERSPC, 
+                                     time = "Follow.Up.Time",
+                                     event = "DeadOfPrCa",
+                                     exposure = "ScrArm")
+attr(pt_object, "exposure")
 ```
@@
 <!--The \code{pt_object} is a \code{data.table} containing the original dataset augmented with-->
-We then pass this object to the \code{plot} function which leverages the \pkg{ggplot2} package [@ggplot2] to build population time plots.
-It builds the plot by sequentially adding layers, starting with a layer for the area representing the population time followed by points to show the casebase sampling mechanism.
-This function gives user the flexibility to add any combination of the case.series, base.series and competing events.
-The case series and competing events are sampled at random vertically on the plot in order to visualise the incidence density using the popTime function.
+We then pass this object to the corresponding \code{plot} method which leverages the \pkg{ggplot2} package to build population time plots.
+
+```{r plot-stratified-erspc-data, eval = eval_cs1, echo = TRUE}
+pacman::p_load(cowplot)
+plot(pt_object, add.case.series = FALSE) + cowplot::theme_minimal_hgrid(12, rel_small = 1)
+theme_cowplot
+plot(pt_object, legend = TRUE, theme.params = list(legend.position = "bottom"))
+
+# plot(pt_object, add.base.series = TRUE, ratio = 1, legend = TRUE)
+```
+
+Figure~\ref{fig:plot-stratified-erspc-data} is built sequentially by adding layers for the area representing the population time in grey, followed by points to show the case series.
+The case series are sampled at random vertically on the plot in order to visualise the incidence density using the \code{popTime} function.
 That is, imagine we draw a vertical line at a specific event time.
 We then plot the point at a randomly sampled y-coordinate along this vertical line.
-This is done to avoid having all points along the upper edge of the plot (because the subjects with the least amount of observation time are plotted at the top of the y-axis).
-By randomly distributing them, we can get a better sense of the inicidence density.
+This is done to avoid having all points along the upper edge of the plot because the subjects with the least amount of observation time are plotted at the top of the y-axis.
+By randomly distributing the points, we can get a better sense of the inicidence density.
+In the ERSPC data, we can see that more events are observed later on in time.
+Therefore a constant hazard model would not be appropriate in this instance as it would overestimate the cumulative incidence earlier on in time, and underestimate it later on.
+
+casebase sampling mechanism.
+This function gives user the flexibility to add any combination of the case series, base series and competing events.
 The base series is sampled horizontally on the plot using the sampleCaseBase function.
-
-One benefit from these plots is that it allows you to see the incidence density.
-This can be seen from the distribution of the red dots in the above plot.
-We can see that more events are observed later on in time.
-Therefore a constant hazard model would not be appropriate in this instance as it would overestimate the cumulative incidence earlier on in time, and underestimate it later on.
 
@@
 This, coupled with the uniform December 31 2006 censoring date, lead to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up.
-
-It is important to note that the red points are random distributed across the grey area for each time of event.
-That is, imagine we draw a vertical line at a specific event time.
-We then plot the red point at a randomly sampled y-coordinate along this vertical line.
-This is done to avoid having all red points along the upper edge of the plot (because the subjects with the least amount of observation time are plotted at the top of the y-axis).
-By randomly distributing them, we can get a better sense of the inicidence density.
 
@@
 We can also create exposure stratified plots by specifying the \code{exposure} argument in the \code{popTime} function:
-
-```{r plot-stratified-erspc-data, eval = eval_cs1}
-pt_object_strat <- casebase::popTime(ERSPC, 
-                                     event = "DeadOfPrCa", 
-                                     exposure = "ScrArm")
-# class(pt_object_strat)
-# pt_object
-plot(pt_object_strat)
-# dev.off()
-# library(magrittr)
-# pt <- plot(pt_object_strat)
-# pt %>% class
-
-# library(ggplot2)
-# pt + theme_minimal()
-
-```
 

### 2020-06-29 analysis/paper/paper.tex
merge pop-stash

@@
 
-To illustrate these functions we will use the European Randomized Study
-of Prostate Cancer Screening data which is available through the
-\pkg{casebase} package. The results of this study were published by
-\citep{schroder2009screening}, and the dataset itself was obtained using
-the approach described in \citep{liu2014recovering}.
+The splitting of these tasks gives the user the flexibility to create
+population-time plots with the graphics system of their choice using the
+object created in Step 1, while the object created in Step 2 allows one
+to leverage the \pkg{ggplot2} family of functions for further
+customization. To illustrate these functions we will use the European
+Randomized Study of Prostate Cancer Screening data which is available
+through the \pkg{casebase} package. The results of this study were
+published by \citep{schroder2009screening}, and the dataset itself was
+obtained using the approach described in \citep{liu2014recovering}.
 
@@
 the exposure variable as an attribute:
+
+We first pass this dataset to the \code{popTime} function. If you do not
+specify the time and event arguments, this function will try to guess
+them using regular expressions. We then generate the plot by using the
+corresponding \code{plot} method:
 
@@
 
-We can also create exposure stratified plots by specifying the
-\code{exposure} argument in the \code{popTime} function:
+The randomization of the Finnish cohorts were carried out on January 1
+of each of the 4 years 1996 to 1999. This, coupled with the uniform
+December 31 2006 censoring date, lead to large numbers of men with
+exactly 11, 10, 9 or 8 years of follow-up. Tracked backwards in time
+(i.e.~from right to left), the population-time plot shows the
+recruitment pattern from its beginning in 1991, and in particular the
+Jan 1 entries in successive years. Tracked forwards in time (i.e.~from
+left to right), the plot for the first three years shows attrition due
+entirely to death (mainly from other causes). Since the Swedish and
+Belgian centres were the last to close their recruitment - in December
+2003 - the minimum potential follow-up is three years. Tracked further
+forwards in time (i.e.~after year 3) the attrition is We can also create
+exposure stratified plots by specifying the \code{exposure} argument in
+the \code{popTime} function:
 
@@
 
-\begin{center}\includegraphics{../figures/unnamed-chunk-2-1} \end{center}
+\begin{center}\includegraphics{../figures/unnamed-chunk-1-1} \end{center}
 

### 2020-07-18 analysis/paper/paper.Rmd
cleaned cs1

--- 2020-05-07 analysis/paper/paper.Rmd
@@
 author:
-  - name: Sahir Bhatnagar *
+  - name: Sahir Rai Bhatnagar *
     affiliation: McGill University
@@
   message = FALSE,
-  echo = TRUE,
+  echo = FALSE,
+  cache = TRUE, 
   comment = "#>",
@@
 library(tibble)
+library(visreg)
+library(lubridate)
 library(tidyverse)
@@
 eval_cs1 <- TRUE
-eval_cs2 <- TRUE
-eval_cs3 <- TRUE
-eval_cs4 <- TRUE
+eval_cs2 <- FALSE
+eval_cs3 <- FALSE
+eval_cs4 <- FALSE
 eval_colophon <- TRUE
@@
 
-$$ \log(\lambda(t; X)) = \beta_0 + \beta_1 X.$$
+\begin{equation}
+\log(\lambda(t; X)) = \beta_0 + \beta_1 X. \label{eq:exp}
+\end{equation}
 
@@
 
-$$ \log(\lambda(t; X)) = \beta_0 + \beta_1 t + \beta_2 X.$$
+\begin{equation}
+\log(\lambda(t; X)) = \beta_0 + \beta_1 t + \beta_2 X. \label{eq:gomp}
+\end{equation}
 
@@
 
-$$ \log(\lambda(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. $$
+\begin{equation}
+\log(\lambda(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. \label{eq:weibull}
+\end{equation}
 
@@
 \textbf{SmoothHazard} \cite{smoothHazard}   &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
-\textbf{Survival} \cite{survival-package}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x
+\textbf{Survival} \cite{survival-package}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x                               \\
+\hline
 \end{tabular}%
@@
 
-The functions in the casebase package can be divided into two categories: 1) data visualization, in the form of population-time plots; and 2) parametric modeling.
+The functions in the casebase package can be divided into two categories: 1) exploratory data analysis, in the form of population-time plots; and 2) parametric modeling of the hazard function.
 We explicitly aimed at being compatible with both \code{data.frame}s and \code{data.table}s.
@@
 ## Population-time plots
+
+The case-base sampling approach described in Section \ref{theory} can be visualized in the form of a population time plot.
+They are extremely informative graphical displays of survival data and should be one of the first steps in an exploratory data analysis.
+We facilitate this task in the \pkg{casebase} package using the \code{popTime} function and \code{plot} method as follows:
+
+1.
+The \code{casebase::popTime} function takes as input the original dataset along with the column names corresponding to the timescale, the event status and an exposure group of interest (optional).
+This will create an object of class \code{popTime}.
+2. The corresponding \code{plot} method for the object created in Step 1 can be called to create the population time plot with several options for customizing the aesthetics.
+
+The splitting of these tasks gives the user the flexibility to create population-time plots with the graphics system of their choice using the object created in Step 1, while the object created in Step 2 allows one to leverage the \pkg{ggplot2} [@ggplot2] family of functions for further customization.
+
+To illustrate these functions we will use the European Randomized Study of Prostate Cancer Screening (ERSPC) data [@schroder2009screening] which was extracted using the approach described in [@liu2014recovering] and available through the \pkg{casebase} package.
+This data contains the individual observations for 159,893 men between the ages of 55 and 69 years at entry who were recruited from seven European countries.
+
+We first create the necessary dataset for producing the population time plot using the \code{casebase::popTime} function.
+In this example we stratify by the treatment group.
+The resulting object inherits from class `popTime` and stores the exposure variable as an attribute:
+
+```{r erspc-data, eval = eval_cs1, echo = c(-1:-3)}
+data("ERSPC")
+ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
+                       levels = c(0,1), 
+                       labels = c("Control group", "Screening group"))
+
+pt_object <- casebase::popTime(ERSPC, 
+                               time = "Follow.Up.Time",
+                               event = "DeadOfPrCa",
+                               exposure = "ScrArm")
+inherits(pt_object, "popTime")
+attr(pt_object, "exposure")
+```
+
+We then pass this object to the corresponding \code{plot} method which leverages the \pkg{ggplot2} package to build population time plots. 
+```{r plot-stratified-erspc-data-code, eval = FALSE, echo = TRUE}
+plot(pt_object, add.base.series = TRUE)
+```
+
+```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE}
+plot(pt_object, 
+     add.base.series = TRUE,
+     ratio = 0.5,
+     facet.params = list(ncol = 2),
+     ribbon.params = list(fill = "gray30"),
+     case.params = list(size = 0.5),
+     base.params = list(size = 0.5)) + theme_bw(base_size = 10) + theme(legend.position = "bottom") + 
+  scale_y_continuous(labels = scales::label_number_si())
+```
+
+Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by first adding a layer for the area representing the population time in gray, with subjects having the least amount of observation time plotted at the top of the y-axis.
+We immediately notice a unique _step shape_ in the population time area which is due to the randomization of the Finnish cohorts which were carried out on January 1 of each of the 4 years 1996 to 1999.
+This, coupled with the uniform December 31 2006 censoring date, lead to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up.
+Tracked backwards in time (i.e.
+from right to left), the population-time plot shows the recruitment pattern from its beginning in 1991, and in particular the January 1 entries in successive years.
+Tracked forwards in time (i.e.
+from left to right), the plot for the first three years shows attrition due entirely to death (mainly from other causes).
+Since the Swedish and Belgian centres were the last to close their recruitment - in December 2003 - the minimum potential follow-up is three years.
+Tracked further forwards in time (i.e.
+after year 3) the attrition is a combination of deaths and staggered entries.
+
+Next, layers for the case series and base series are added.
+The case series is sampled at random vertically on the plot to avoid having all points along the upper edge of the plot, while the base series is sampled horizontally with sampling weight proportional to their follow-up time.
+By randomly distributing the points, we can get a better sense of the incidence density.
+In the ERSPC data, we see that more events are observed later on in time.
+Therefore a constant hazard model would not be appropriate in this instance as it would overestimate the cumulative incidence earlier on in time, and underestimate it later on.
+The reader is referred to the package vignettes for more examples and a detailed description of how to modify the aesthetics of a population time plot.
 
@@
 
-For the first case study, we will use of the European Randomized Study of Prostate Cancer Screening data. This dataset is available through the \pkg{casebase} package:
-
-```{r erspc-data, eval = eval_cs1}
-data(ERSPC)
-ERSPC$ScrArm <- factor(ERSPC$ScrArm,
-  levels = c(0, 1),
-  labels = c("Control group", "Screening group")
-)
-```
-
-The results of this study were published by [@schroder2009screening], and the dataset itself was obtained using the approach described in [@liu2014recovering].
-
-Population time plots can be extremely informative graphical displays of survival data.
-They should be the first step in an exploratory data analysis.
-We facilitate this task in the \pkg{casebase} package using the \code{popTime} function.
-We first create the necessary dataset for producing the population time plots, and we can generate the plot by using the corresponding \code{plot} method:
-
-```{r plot-erspc-data, eval = eval_cs1}
-pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
-plot(pt_object)
-```
-
-We can also create exposure stratified plots by specifying the \code{exposure} argument in the \code{popTime} function:
-
-```{r plot-stratified-erspc-data, eval = eval_cs1}
-pt_object_strat <- casebase::popTime(ERSPC,
-  event = "DeadOfPrCa",
-  exposure = "ScrArm"
-)
-plot(pt_object_strat)
-```
-
-We can also plot them side-by-side using the \code{ncol} argument:
-
-```{r plot-stratified-erspc-data-side, eval = eval_cs1}
-plot(pt_object_strat, facet.params = list(ncol = 2))
-```
-
-ADD PARAGRAPH ABOUT WHAT WE CAN CONCLUDE FROM THESE GRAPHS.
-
-Next, we investigate the differences between the control and the screening arms.
-A common choice for this type of analysis is to use a Cox regression model and estimate the hazard ratio for the screening group (relative to the control group).
-In \proglang{R}, it can be done as follows:
-
-```{r erspc-cox, eval = eval_cs1}
-cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm,
-  data = ERSPC
-)
-summary(cox_model)
-```
-
-We can also plot the cumulative incidence function (CIF) for each group. However, this involves estimating the baseline hazard, which Cox regression treated as a nuisance parameter.
-
-```{r erspc-cox-cif, eval = eval_cs1, echo = echo_plot_code}
+We investigate the differences in survival between the control and screening arms in the ERSPC study presented earlier using casebase sampling to fit a parametric hazard function.
+We fit four models that differ in the functional form of time; (1) excluded from the linear predictor (Eq.\eqref{eq:exp}), (2) linear function (Eq.\eqref{eq:gomp}), (3) log function (Eq.\eqref{eq:weibull}), and (4) a smooth function using cubic B-splines.
+The models are fit using \code{fitSmoothHazard}.
+Since the output object from \code{fitSmoothHazard} inherits from the \code{glm} class, we see a familiar result when using the function \code{summary}:
+
+```{r, eval = FALSE, echo = TRUE}
+fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
+             gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
+             weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
+             splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
+
+fits <- lapply(fmla, function(i) {
+  casebase::fitSmoothHazard(i, data = ERSPC, ratio = 100)
+})
+
+summary(fits[[2]])
+```
+
+The \code{absoluteRisk} function takes as input the \code{fitSmoothHazard} object and returns a matrix where each column corresponds to the covariate profiles specified in the \code{newdata} argument, and each row corresponds to a specified time point:
+
+```{r, eval=FALSE, echo=echo_plot_code}
+new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
+
+risks <- lapply(fits, function(i) {
+  casebase::absoluteRisk(object = i, time = seq(0,15,0.1), newdata = new_data)
+})
+
+head(risks[[2]])
+```
+
+In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from casebase on the Cox model CIF.
+The CIF estimates for the exponential model in panel (1) overestimate the cumulative incidence earlier on in time, and underestimate it later on.
+Based on what we observed in the population time plot, where more events are observed later on in time, this poor fit for the exponential hazard is expected.
+This example demonstrates the benefits of population time plots as an exploratory analysis tool.
+
+In Figure \ref{fig:erspc-cox-cif} panels (2)-(4), we notice a better fit with increasing complexities of time.
+As noted above, the usual asymptotic results hold for likelihood ratio tests built using casebase sampling models.
+Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) splines model:
+
+```{r, eval = eval_cs1, echo=FALSE}
+fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
+             gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
+             weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
+             splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
+
+fits <- lapply(fmla, function(i) {
+  casebase::fitSmoothHazard(i, data = ERSPC, ratio = 100)
+})
+```
+
+```{r erscp-compare, eval = eval_cs1, echo = FALSE}
+anova(fits$exponential, fits$splines, test = "LRT")
+```
+
+As expected, we see that splines model provides a better fit. Similarly, we can compare the AIC for all four casebase models:
+
+```{r erscp-aic, eval = eval_cs1, echo=FALSE}
+c("Exp." = AIC(fits$exponential),
+  "Gomp." = AIC(fits$gompertz),
+  "Weib." = AIC(fits$exponential),
+  "Splines" = AIC(fits$splines))
+```
+
+Once again, we have evidence that the splines model provides the best fit.
+
+<!--A common choice for this type of analysis is to use a Cox regression model and estimate the hazard ratio for the screening group (relative to the control group).
+In \proglang{R}, it can be done as follows:-->
+
+```{r erspc-cox, eval = eval_cs1, echo=FALSE}
+surv_obj <- with(ERSPC, Surv(Follow.Up.Time, DeadOfPrCa))
+cox_model <- survival::coxph(surv_obj ~ ScrArm, data = ERSPC)
+```
+
+```{r erspc-cox-cif, eval = eval_cs1, echo = FALSE, fig.width=8, fig.height=6, fig.cap="Cumulative incidence functions for control and screening groups in the ERSPC data.
+In each of the panels, we plot the CIF from the Cox model using \\code{survival::survfit} (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression.
+(1) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution.
+(2) Linear function of time which is equivalent to a Gompertz distribution.
+(3) The natural logarithm of time which is equivalent to a Weibull distribution.
+(4) Cubic B-spline expansion of time."}
+cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+myCols <- cbPalette[c(4,7)]
 new_data <- data.frame(
-  ScrArm = c("Control group", "Screening group"),
-  ignore = 99
-)
-
+  ScrArm = c("Control group", "Screening group")#,
+  # ignore = 99
+)
+
+risks <- lapply(fits, function(i) {
+  casebase::absoluteRisk(object = i, time = seq(0,15,0.1), newdata = new_data)
+})
+
+aft_fits <- lapply(list("exp","gompertz","weibull"), function(i)
+  flexsurv::flexsurvreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm,
+                           data = ERSPC, dist = i)
+  )
+
+names(aft_fits) <- list("exponential","gompertz","weibull")
+
+# dev.off()
+layout(matrix(1:4, ncol = 2, byrow = FALSE), respect = FALSE)
+par(mar = c(0, 4.1, 4.1, 0))
 plot(survfit(cox_model, newdata = new_data),
-  xlab = "Years since Randomization",
-  ylab = "Cumulative Incidence (%)",
+     ylab = "",
+     xlab = "",
      fun = "event",
-  xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
-  main = sprintf(
-    "Estimated Cumulative Incidence (risk) of Death from Prostate 
-                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-    exp(coef(cox_model)),
-    exp(confint(cox_model))[1],
-    exp(confint(cox_model))[2]
-  ),
-  ylim = c(0, 2), yscale = 100
-)
+     xlim = c(0,15), conf.int = FALSE, col = myCols, 
+     # main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
+     #                Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+     #                exp(coef(cox_model)), 
+     #                exp(confint(cox_model))[1], 
+     #                exp(confint(cox_model))[2]), 
+     lwd = 2,
+     yscale = 100,
+     xaxt = 'n')
 legend("topleft", 
-  legend = c("Control group", "Screening group"),
-  col = c("red", "blue"),
-  lty = c(1, 1),
-  bg = "gray90"
-)
-```
-
-Next, we show how case-base sampling can be used to carry out a similar analysis. We fit several models that differ in how we choose to model time.
+       legend = c("Control (Cox)","Control (Casebase)",
+                  "Screening (Cox)", "Screening (Casebase)"), 
+       col = rep(myCols, each = 2),
+       lty = c(1, 2, 1, 2), 
+       lwd = 2,
+       cex = 1.1,
+       bg = "gray90")
+lines(risks$exponential[,1], risks$exponential[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risks$exponential[,1], risks$exponential[,3], col = myCols[2], lty = 2, lwd = 2)
+text(x = 15*0.97, y = 0.05/100, labels = "(1)", cex = 1.5)
+
+par(mar = c(4.1, 4.1, 0, 0))
+plot(survfit(cox_model, newdata = new_data),
+     ylab = "", xlab = "",
+     fun = "event",
+     xlim = c(0,15), conf.int = FALSE, col = myCols, 
+     lwd = 2,
+     yscale = 100)
+lines(risks$weibull[,1], risks$weibull[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risks$weibull[,1], risks$weibull[,3], col = myCols[2], lty = 2, lwd = 2)  
+text(x = 15*0.97, y = 0.05/100, labels = "(3)", cex = 1.5)
+mtext("Cumulative incidence (%)", side = 2, line = -1.5, outer = TRUE)
+
+par(mar = c(0, 0, 4.1, 3.9))
+plot(survfit(cox_model, newdata = new_data),
+     ylab = "",xlab = "",
+     fun = "event",
+     xaxt = 'n', yaxt = 'n',
+     xlim = c(0,15), conf.int = FALSE, col = myCols, 
+     lwd = 2,
+     yscale = 100)
+lines(risks$gompertz[,1], risks$gompertz[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risks$gompertz[,1], risks$gompertz[,3], col = myCols[2], lty = 2, lwd = 2)
+text(x = 15*0.97, y = 0.05/100, labels = "(2)", cex = 1.5)
+
+par(mar = c(4.1, 0, 0, 3.9))
+plot(survfit(cox_model, newdata = new_data),
+     ylab = "",xlab = "",
+     fun = "event",
+     yaxt = 'n',
+     xlim = c(0,15), conf.int = FALSE, col = myCols, 
+     lwd = 2,
+     yscale = 100)
+lines(risks$splines[,1], risks$splines[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risks$splines[,1], risks$splines[,3], col = myCols[2], lty = 2, lwd = 2)
+text(x = 15*0.97, y = 0.05/100, labels = "(4)", cex = 1.5)
+mtext("Years since randomization", side = 1, line = -1.5, outer = TRUE)
+```
+
+```{r, results='hide', eval = FALSE, echo = FALSE}
+# par(mfrow = c(2, 2))
+# lapply(fits, function(i) {
+#   casebase::hazardPlot(i, newdata = new_data[2,], ci = FALSE)
+#   casebase::hazardPlot(i, newdata = new_data[1,], ci = FALSE, add = TRUE, lty = 2)
+# })
+```
+
+```{r erspc-casebase-exponential, eval = FALSE, echo = FALSE}
+# # Equivalence between casebase and AFT
+# # exponential -------------------------------------------------------------
+# 
+# casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
+#                                                   data = ERSPC, 
+#                                                   ratio = 100)
+# aft_exponential <- survreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
+#                            data = ERSPC, dist = "exp")
+# 
+# -coef(aft_exponential)
+# coef(casebase_exponential)
+# summary(casebase_exponential)
+# exp(coef(casebase_exponential)[2])
+# exp(confint(casebase_exponential)[2, ])
+# 
+# # weibull -----------------------------------------------------------------
+# 
+# casebase_weibull <- casebase::fitSmoothHazard(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm, 
+#                                            data = ERSPC, 
+#                                            ratio = 100)
+# aft_weibull <- survreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
+#                        data = ERSPC, dist = "weibull")
+# 
+# -coef(aft_weibull)/aft_weibull$scale
+# coef(casebase_weibull)
+```
+
+```{r erspc-casebase-cif, eval = FALSE, echo = FALSE}
+# smooth_risk_exp <- casebase::absoluteRisk(
+#   object = casebase_exponential,
+#   time = seq(0, 15, 0.1),
+#   newdata = new_data
+# )
+# 
+# plot(survfit(cox_model, newdata = new_data),
+#   xlab = "Years since Randomization",
+#   ylab = "Cumulative Incidence (%)",
+#   fun = "event",
+#   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
+#   main = sprintf(
+#     "Estimated Cumulative Incidence (risk) of Death from Prostate 
+#                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+#     exp(coef(cox_model)),
+#     exp(confint(cox_model))[1],
+#     exp(confint(cox_model))[2]
+#   ),
+#   ylim = c(0, 2), yscale = 100
+# )
+# lines(smooth_risk_exp[, 1], smooth_risk_exp[, 2], col = "red", lty = 2)
+# lines(smooth_risk_exp[, 1], smooth_risk_exp[, 3], col = "blue", lty = 2)
+# 
+# 
+# legend("topleft",
+#   legend = c(
+#     "Control group (Cox)", "Control group (Casebase)",
+#     "Screening group (Cox)", "Screening group (Casebase)"
+#   ),
+#   col = c("red", "red", "blue", "blue"),
+#   lty = c(1, 2, 1, 2),
+#   bg = "gray90"
+# )
+```
 
@@
 -->
-
-First, we will fit an exponential model. Recall that this corresponds to excluding time from the linear predictor.
-
-```{r erspc-casebase-exponential, eval = eval_cs1}
-casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm,
-  data = ERSPC,
-  ratio = 100
-)
-
-summary(casebase_exponential)
-exp(coef(casebase_exponential)[2])
-exp(confint(casebase_exponential)[2, ])
-```
-
 <!-- The \code{absoluteRisk} function provides an estimate of the cumulative incidence curves for a specific risk profile using the following equation: -->
-
 <!-- $$ CI(x, t) = 1 - \exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$ -->
 
-We can then use the \code{absoluteRisk} function to get an estimate of the cumulative incidence curve for a specific covariate profile.
-In the plot below, we overlay the estimated CIF from the exponential model on the Cox model CIF:
-
-```{r erspc-casebase-cif, eval = eval_cs1, echo = echo_plot_code}
-smooth_risk_exp <- casebase::absoluteRisk(
-  object = casebase_exponential,
-  time = seq(0, 15, 0.1),
-  newdata = new_data
-)
-
-plot(survfit(cox_model, newdata = new_data),
-  xlab = "Years since Randomization",
-  ylab = "Cumulative Incidence (%)",
-  fun = "event",
-  xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
-  main = sprintf(
-    "Estimated Cumulative Incidence (risk) of Death from Prostate 
-                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-    exp(coef(cox_model)),
-    exp(confint(cox_model))[1],
-    exp(confint(cox_model))[2]
-  ),
-  ylim = c(0, 2), yscale = 100
-)
-lines(smooth_risk_exp[, 1], smooth_risk_exp[, 2], col = "red", lty = 2)
-lines(smooth_risk_exp[, 1], smooth_risk_exp[, 3], col = "blue", lty = 2)
-
-legend("topleft",
-  legend = c(
-    "Control group (Cox)", "Control group (Casebase)",
-    "Screening group (Cox)", "Screening group (Casebase)"
-  ),
-  col = c("red", "red", "blue", "blue"),
-  lty = c(1, 2, 1, 2),
-  bg = "gray90"
-)
-```
-
-As we can see, the exponential model gives us an estimate of the hazard ratio that is similar to the one obtained using Cox regression.
-However, the CIF estimates are quite different.
-Based on what we observed in the population time plot, where more events are observed later on in time, we do not expect a constant hazard would be a good description for this data.
-In other words, this poor fit for the exponential hazard is expected.
-A constant hazard model would overestimate the cumulative incidence earlier on in time, and underestimate it later on; this is what we see on the cumulative incidence plot.
-This example demonstrates the benefits of population time plots as an exploratory analysis tool.
-
-For the next model, we include time as a linear term. Recall that this corresponds to a Weibull model.
-
-```{r erspc-casebase-weibull, eval = eval_cs1}
-casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm,
-  data = ERSPC,
-  ratio = 100
-)
-
-summary(casebase_time)
-exp(coef(casebase_time))
-exp(confint(casebase_time))
-```
-
-Again, the estimate of the hazard ratio is similar to that obtained from Cox regression. We then look at the estimate of the CIF:
-
-```{r erspc-casebase-weibull-cif, eval = eval_cs1, echo = echo_plot_code}
-smooth_risk_time <- casebase::absoluteRisk(
-  object = casebase_time,
-  time = seq(0, 15, 0.1),
-  newdata = new_data
-)
-
-plot(survfit(cox_model, newdata = new_data),
-  xlab = "Years since Randomization",
-  ylab = "Cumulative Incidence (%)",
-  fun = "event",
-  xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
-  main = sprintf(
-    "Estimated Cumulative Incidence (risk) of Death from Prostate 
-                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-    exp(coef(cox_model)),
-    exp(confint(cox_model))[1],
-    exp(confint(cox_model))[2]
-  ),
-  ylim = c(0, 2), yscale = 100
-)
-lines(smooth_risk_time[, 1], smooth_risk_time[, 2], col = "red", lty = 2)
-lines(smooth_risk_time[, 1], smooth_risk_time[, 3], col = "blue", lty = 2)
-
-legend("topleft",
-  legend = c(
-    "Control group (Cox)", "Control group (Casebase)",
-    "Screening group (Cox)", "Screening group (Casebase)"
-  ),
-  col = c("red", "red", "blue", "blue"),
-  lty = c(1, 2, 1, 2),
-  bg = "gray90"
-)
-```
-
-We see that the Weibull model leads to a better fit of the CIF than the Exponential model, when compared with the semi-parametric approach.
-
-Finally, we can model the hazard as a smooth function of time using the \code{splines} package:
-
-```{r erspc-casebase-splines, eval = eval_cs1}
-casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm,
-  data = ERSPC,
-  ratio = 100
-)
-
-summary(casebase_splines)
-exp(coef(casebase_splines))
-exp(confint(casebase_splines))
-```
-
-Once again, we can see that the estimate of the hazard ratio is similar to that from the Cox regression. We then look at the estimate of the CIF:
-
-```{r erspc-casebase-splines-cif, eval = eval_cs1, echo = echo_plot_code}
-smooth_risk_splines <- absoluteRisk(
-  object = casebase_splines,
-  time = seq(0, 15, 0.1),
-  newdata = new_data
-)
-
-plot(survfit(cox_model, newdata = new_data),
-  xlab = "Years since Randomization",
-  ylab = "Cumulative Incidence (%)",
-  fun = "event",
-  xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
-  main = sprintf(
-    "Estimated Cumulative Incidence (risk) of Death from Prostate 
-                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-    exp(coef(cox_model)),
-    exp(confint(cox_model))[1],
-    exp(confint(cox_model))[2]
-  ),
-  ylim = c(0, 2), yscale = 100
-)
-lines(smooth_risk_splines[, 1], smooth_risk_splines[, 2], col = "red", lty = 2)
-lines(smooth_risk_splines[, 1], smooth_risk_splines[, 3], col = "blue", lty = 2)
-
-legend("topleft",
-  legend = c(
-    "Control group (Cox)", "Control group (Casebase)",
-    "Screening group (Cox)", "Screening group (Casebase)"
-  ),
-  col = c("red", "red", "blue", "blue"),
-  lty = c(1, 2, 1, 2),
-  bg = "gray90"
-)
-```
-
-Qualitatively, the combination of splines and case-base sampling produces the parametric model that most closely resembles the Cox model.
-In the following table, we see that the confidence intervals are also similar across all four models.
-In other words, this reinforces the idea that, under proportional hazards, we do not need to model the full hazard to obtain reliable estimates of the hazard ratio.
-Of course, different parametric models for the hazards give rise to qualitatively different estimates for the CIF.
-
-```{r erspc-95ci, echo = FALSE, eval = eval_cs1}
+```{r erspc-casebase-weibull-cif, eval = FALSE, echo = FALSE}
+# smooth_risk_time <- casebase::absoluteRisk(
+#   object = casebase_time,
+#   time = seq(0, 15, 0.1),
+#   newdata = new_data
+# )
+# 
+# plot(survfit(cox_model, newdata = new_data),
+#   xlab = "Years since Randomization",
+#   ylab = "Cumulative Incidence (%)",
+#   fun = "event",
+#   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
+#   main = sprintf(
+#     "Estimated Cumulative Incidence (risk) of Death from Prostate 
+#                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+#     exp(coef(cox_model)),
+#     exp(confint(cox_model))[1],
+#     exp(confint(cox_model))[2]
+#   ),
+#   ylim = c(0, 2), yscale = 100
+# )
+# lines(smooth_risk_time[, 1], smooth_risk_time[, 2], col = "red", lty = 2)
+# lines(smooth_risk_time[, 1], smooth_risk_time[, 3], col = "blue", lty = 2)
+# 
+# legend("topleft",
+#   legend = c(
+#     "Control group (Cox)", "Control group (Casebase)",
+#     "Screening group (Cox)", "Screening group (Casebase)"
+#   ),
+#   col = c("red", "red", "blue", "blue"),
+#   lty = c(1, 2, 1, 2),
+#   bg = "gray90"
+# )
+```
+
+```{r erspc-95ci22, echo = FALSE, eval = eval_cs1}
+CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
+# CI_exp <- c(exp(coef(fits$exponential))["ScrArmScreening group"], exp(confint(fits$exponential))["ScrArmScreening group",])
+# CI_gompertz <- c(exp(coef(fits$gompertz))["ScrArmScreening group"], exp(confint(fits$gompertz))["ScrArmScreening group",])
+# CI_weibull <- c(exp(coef(fits$weibull))["ScrArmScreening group"], exp(confint(fits$weibull))["ScrArmScreening group",])
+# CI_splines <- c(exp(coef(fits$splines))["ScrArmScreening group"], exp(confint(fits$splines))["ScrArmScreening group",])
+
+# function to exponentiate HR and CI and return a HR (lower, Upper) character
+exp_est_ci <- function(model = c("exponential","gompertz","weibull","splines"),
+                       fit, 
+                       covariate = "ScrArmScreening group", 
+                       type = c("casebase", "survreg")) {
+
+  type <- match.arg(type)
+  covariate <- match.arg(covariate)
+
+  if (type == "casebase"){
+
+    cis <- confint(fit[[model]])
+
+    sprintf("%0.2f (%0.2f, %0.2f)",
+            exp(coef(fit[[model]]))[[covariate]],
+            exp(cis)[covariate,1],
+            exp(cis)[covariate,2])
+  } else {
+
+    if (model == "weibull") {
+
+      # The maximised log-likelihoods are the same between survreg and flexsurvreg, 
+      # however the parameterisation is different: the first coefficient (Intercept) 
+      # reported by survreg is log(mu), and survreg's "scale" is
+      # dweibull’s (thus flexsurvreg)'s 1 / shape
+      res <- -fit[[model]]$res[covariate,] / (1 / fit[[model]]$res["shape",])
+
+      sprintf("%0.2f (%0.2f, %0.2f)",
+              exp(res[["est"]]),
+              exp(res[["U95%"]]),
+              exp(res[["L95%"]]))
+
+    } else if (model == "splines") {
+
+      c("--")
+
+    } else {
+
+      res <- fit[[model]]$res[covariate,]
+
+      sprintf("%0.2f (%0.2f, %0.2f)",
+              exp(res[["est"]]),
+              exp(res[["L95%"]]),
+              exp(res[["U95%"]]))
+    }
+  }
+}
+
+casebase_HR_CI <- lapply(list("exponential","gompertz","weibull","splines"), 
+                         exp_est_ci, 
+                         fit = fits,
+                         covariate = "ScrArmScreening group", 
+                         type = "casebase")
+
+survreg_HR_CI <- lapply(list("exponential","gompertz","weibull","splines"), 
+                         exp_est_ci, 
+                         fit = aft_fits,
+                         covariate = "ScrArmScreening group", 
+                         type = "survreg")
+
+ptable <- cbind(list("Exponential","Gompertz","Weibull","Splines"),
+            do.call(rbind, casebase_HR_CI), do.call(rbind, survreg_HR_CI))
+
+# table_ci <- rbind(CI_cox, CI_exp, CI_gompertz, CI_weibull, CI_splines)
+# colnames(table_ci) <- c("HR", "LB", "UB")
+# 
+# data.frame(Model = c("Cox", "Exponential", "Gompertz", "Weibull", "Splines")) %>% 
+#   bind_cols(as.data.frame(table_ci)) %>% 
+#   mutate(CI = paste0("(", round(LB, 2), ", ",
+#                      round(UB, 2), ")"),
+#          Width = UB - LB) %>% 
+#   dplyr::select(-LB, -UB) %>% 
+#   knitr::kable(format = "latex", booktabs = TRUE)  %>%
+#   kable_styling()
+```
+
+```{r erspc-casebase-splines-cif, eval = FALSE, echo = FALSE}
+# smooth_risk_splines <- absoluteRisk(
+#   object = casebase_splines,
+#   time = seq(0, 15, 0.1),
+#   newdata = new_data
+# )
+# 
+# plot(survfit(cox_model, newdata = new_data),
+#   xlab = "Years since Randomization",
+#   ylab = "Cumulative Incidence (%)",
+#   fun = "event",
+#   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
+#   main = sprintf(
+#     "Estimated Cumulative Incidence (risk) of Death from Prostate 
+#                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
+#     exp(coef(cox_model)),
+#     exp(confint(cox_model))[1],
+#     exp(confint(cox_model))[2]
+#   ),
+#   ylim = c(0, 2), yscale = 100
+# )
+# lines(smooth_risk_splines[, 1], smooth_risk_splines[, 2], col = "red", lty = 2)
+# lines(smooth_risk_splines[, 1], smooth_risk_splines[, 3], col = "blue", lty = 2)
+# 
+# legend("topleft",
+#   legend = c(
+#     "Control group (Cox)", "Control group (Casebase)",
+#     "Screening group (Cox)", "Screening group (Casebase)"
+#   ),
+#   col = c("red", "red", "blue", "blue"),
+#   lty = c(1, 2, 1, 2),
+#   bg = "gray90"
+# )
+```
+
+In the following table, we present a side-by-side comparison of the hazard ratios and confidence intervals estimated from \code{fitSmoothHazard} and the corresponding parametric model using \code{survival::survreg}, as well as the Cox model estimate.
+The hazard ratio estimates and confidence intervals are similar across all four models.
+This reinforces the idea that, under proportional hazards, we do not need to model the full hazard to obtain reliable estimates of the hazard ratio.
+Of course, different parametric models for the hazards give rise to qualitatively different estimates for the CIF as shown in Figure \ref{fig:erspc-cox-cif}.
+
+```{r print-erspc-estimates, eval = TRUE}
+## ---- print-sim-table ----
+
+kable(ptable, "latex", booktabs = T, align = c("l","c","c"),
+      caption = c("Comparison of estimated hazard ratios and 95\\% confidence intervals for ERSPC data."),
+      col.names = c("Model","casebase::fitSmoothHazard","survival::survreg")) %>%
+  kable_styling() %>%
+  # add_header_above(c(" "," ","No overlap" = 2, "All causal SNPs\nin kinship" = 2, "No overlap" = 2, "All causal SNPs\nin kinship" = 2)) %>%
+  # add_header_above(c(" "," ","Null model" = 4, "1% Causal SNPs" = 4)) %>%
+  column_spec(1, bold=T) %>%
+  collapse_rows(columns = 1, latex_hline = "major", valign = "middle") %>%
+  footnote(general = sprintf("Cox model estimate: HR (95%% CI) = %0.2f (%0.2f, %0.2f)", CI_cox[1], CI_cox[2], CI_cox[3]),
+           general_title = "") #%>%
+  # footnote(general = "Model Size is the number of selected variables using the high-dimensional BIC for \texttt{ggmix} and 10-fold cross validation for \texttt{lasso} and \texttt{twostep}.")
+  #          # number = c("", "Footnote 2; "),
+  #          # alphabet = c("Footnote A; ", "Footnote B; "),
+  #          # symbol = c("Footnote Symbol 1; ", "Footnote Symbol 2"),
+  #          general_title = "Note: ", number_title = "Type I: ",
+  #          alphabet_title = "Type II: ", symbol_title = "Type III: ",
+  #          footnote_as_chunk = T, title_format = c("italic", "underline")
+  # )
+```
+
+```{r erspc-95ci, echo = FALSE, eval = FALSE}
 CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
@@
 ```
-
-As noted above, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models.
-Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) splines model.
-
-```{r erscp-compare, eval = eval_cs1}
-anova(casebase_exponential, casebase_splines, test = "LRT")
-```
-
-As expected, we see that splines model provides a better fit. Similarly, we can compare the AIC for all three case-base models:
-
-```{r erscp-aic, eval = eval_cs1}
-c("Exp." = AIC(casebase_exponential),
-  "Weib." = AIC(casebase_time),
-  "Splines" = AIC(casebase_splines))
-```
-
-Once again, we have evidence that the splines model provides the best fit.
 
@@
 B) The blue points represent the competing events.", echo=FALSE, eval = eval_cs2}
-pt_object <- casebase::popTime(bmtcrr, event = "Status", time = "ftime")
-
-pt_object_comp <- bmtcrr %>%
-  mutate(Status = case_when(
-    Status == 1 ~ 2L,
-    Status == 2 ~ 1L,
-    TRUE ~ Status
-  )) %>%
-  casebase::popTime(event = "Status", time = "ftime")
-
-plot_grid(plot(pt_object) + ggtitle("Relapse"),
-          plot(pt_object_comp, point.colour = "blue") + ggtitle("Competing events"),
-          labels = c("A", "B"))
+
+# fill_cols <- colorspace::sequential_hcl(n = 3, palette = "Viridis")
+# 
+# (fill_colors <- c("Case series" = fill_cols[1],
+#                  "Competing event" = fill_cols[3],
+#                  "Base series" = fill_cols[2]))
+# 
+# color_cols <- colorspace::darken(col = fill_cols, amount = 0.3)
+# 
+# (color_colors <- c("Case series" = color_cols[1],
+#                    "Competing event" = color_cols[3],
+#                    "Base series" = color_cols[2]))
+# 
+# pt_object <- casebase::popTime(bmtcrr, event = "Status", time = "ftime")
+# 
+# pt_object_comp <- bmtcrr %>%
+#   mutate(Status = case_when(
+#     Status == 1 ~ 2L,
+#     Status == 2 ~ 1L,
+#     TRUE ~ Status
+#   )) %>%
+#   casebase::popTime(event = "Status", time = "ftime")
+# 
+# plot_grid(plot(pt_object) + ggtitle("Relapse"),
+#           plot(pt_object_comp, point.colour = "blue") + ggtitle("Competing events"),
+#           labels = c("A", "B"))
+
+popTimeData <- popTime(data = bmtcrr, time = "ftime")
+plot(popTimeData,
+     add.case.series = TRUE,
+     # add.base.series = TRUE,
+     add.competing.event = TRUE,
+     comprisk = TRUE)
+
 ```
@@
 ```{r stanford-poptime, eval = eval_cs4}
-library(survival)
-library(casebase)
+colnames(jasa)[which(colnames(jasa)=="futime")] <- "futime_old"
+colnames(jasa)[which(colnames(jasa)=="age")] <- "futime"
 
@@
 ```{r, eval = eval_cs4}
-library(tidyverse)
-library(lubridate)
-
 cb_data <- sampleCaseBase(jasa,
@@
 
-```{r stanford-hazard, eval = eval_cs4}
+THIS NEXT CODE CHUNK IS NOT WORKING
+
+```{r stanford-hazard, eval = FALSE, echo=TRUE}
 # Compute hazards---

### 2020-07-18 analysis/paper/paper.tex
cleaned cs1

--- 2020-05-07 analysis/paper/paper.tex
@@
 \author{
-Sahir Bhatnagar *\\McGill University \And Maxime Turgeon *\\University
-of Manitoba \AND Jesse Islam\\McGill University \And James
-Hanley\\McGill University \And Olli Saarela\\University of Toronto
+Sahir Rai Bhatnagar *\\McGill University \And Maxime Turgeon *\\University of Manitoba \AND Jesse Islam\\McGill University \And James Hanley\\McGill University \And Olli Saarela\\University of Toronto
 }
@@
 
-\Plainauthor{Sahir Bhatnagar *, Maxime Turgeon *, Jesse Islam, James
-Hanley, Olli Saarela}
+\Plainauthor{Sahir Rai Bhatnagar *, Maxime Turgeon *, Jesse Islam, James Hanley, Olli Saarela}
 \Plaintitle{casebase: An Alternative Framework For Survival Analysis}
-\Shorttitle{\pkg{casebase}: An Alternative Framework For Survival
-Analysis}
+\Shorttitle{\pkg{casebase}: An Alternative Framework For Survival Analysis}
 
@@
 \Address{
-    Sahir Bhatnagar *\\
+    Sahir Rai Bhatnagar *\\
   McGill University\\
@@
 
-\[ \log(\lambda(t; X)) = \beta_0 + \beta_1 X.\]
+\begin{equation}
+\log(\lambda(t; X)) = \beta_0 + \beta_1 X. \label{eq:exp}
+\end{equation}
 
@@
 
-\[ \log(\lambda(t; X)) = \beta_0 + \beta_1 t + \beta_2 X.\]
+\begin{equation}
+\log(\lambda(t; X)) = \beta_0 + \beta_1 t + \beta_2 X. \label{eq:gomp}
+\end{equation}
 
@@
 
-\[ \log(\lambda(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. \]
+\begin{equation}
+\log(\lambda(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. \label{eq:weibull}
+\end{equation}
 
@@
 The functions in the casebase package can be divided into two
-categories: 1) data visualization, in the form of population-time plots;
-and 2) parametric modeling. We explicitly aimed at being compatible with
-both \code{data.frame}s and \code{data.table}s. This is evident in some
-of the coding choices we made, and it is also reflected in our unit
-tests.
+categories: 1) exploratory data analysis, in the form of population-time
+plots; and 2) parametric modeling of the hazard function. We explicitly
+aimed at being compatible with both \code{data.frame}s and
+\code{data.table}s. This is evident in some of the coding choices we
+made, and it is also reflected in our unit tests.
 
@@
 \subsection{Population-time plots}\label{population-time-plots}}
+
+The case-base sampling approach described in Section \ref{theory} can be
+visualized in the form of a population time plot. They are extremely
+informative graphical displays of survival data and should be one of the
+first steps in an exploratory data analysis. We facilitate this task in
+the \pkg{casebase} package using the \code{popTime} function and
+\code{plot} method as follows:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  The \code{casebase::popTime} function takes as input the original
+  dataset along with the column names corresponding to the timescale,
+  the event status and an exposure group of interest (optional). This
+  will create an object of class \code{popTime}.\\
+\item
+  The corresponding \code{plot} method for the object created in Step 1
+  can be called to create the population time plot with several options
+  for customizing the aesthetics.
+\end{enumerate}
+
+The splitting of these tasks gives the user the flexibility to create
+population-time plots with the graphics system of their choice using the
+object created in Step 1, while the object created in Step 2 allows one
+to leverage the \pkg{ggplot2} \citep{ggplot2} family of functions for
+further customization.
+
+To illustrate these functions we will use the European Randomized Study
+of Prostate Cancer Screening (ERSPC) data \citep{schroder2009screening}
+which was extracted using the approach described in
+\citep{liu2014recovering} and available through the \pkg{casebase}
+package. This data contains the individual observations for 159,893 men
+between the ages of 55 and 69 years at entry who were recruited from
+seven European countries.
+
+We first create the necessary dataset for producing the population time
+plot using the \code{casebase::popTime} function. In this example we
+stratify by the treatment group. The resulting object inherits from
+class \texttt{popTime} and stores the exposure variable as an attribute:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> pt_object <- casebase::popTime(ERSPC, 
+R+                                time = "Follow.Up.Time",
+R+                                event = "DeadOfPrCa",
+R+                                exposure = "ScrArm")
+R> inherits(pt_object, "popTime")
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> [1] TRUE
+\end{CodeOutput}
+
+\begin{CodeInput}
+R> attr(pt_object, "exposure")
+\end{CodeInput}
+
+\begin{CodeOutput}
+#> [1] "ScrArm"
+\end{CodeOutput}
+\end{CodeChunk}
+
+We then pass this object to the corresponding \code{plot} method which
+leverages the \pkg{ggplot2} package to build population time plots.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> plot(pt_object, add.base.series = TRUE)
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-1} \end{center}
+
+\end{CodeChunk}
+
+Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by
+first adding a layer for the area representing the population time in
+gray, with subjects having the least amount of observation time plotted
+at the top of the y-axis. We immediately notice a unique \emph{step
+shape} in the population time area which is due to the randomization of
+the Finnish cohorts which were carried out on January 1 of each of the 4
+years 1996 to 1999. This, coupled with the uniform December 31 2006
+censoring date, lead to large numbers of men with exactly 11, 10, 9 or 8
+years of follow-up. Tracked backwards in time (i.e.~from right to left),
+the population-time plot shows the recruitment pattern from its
+beginning in 1991, and in particular the January 1 entries in successive
+years. Tracked forwards in time (i.e.~from left to right), the plot for
+the first three years shows attrition due entirely to death (mainly from
+other causes). Since the Swedish and Belgian centres were the last to
+close their recruitment - in December 2003 - the minimum potential
+follow-up is three years. Tracked further forwards in time (i.e.~after
+year 3) the attrition is a combination of deaths and staggered entries.
+
+Next, layers for the case series and base series are added. The case
+series is sampled at random vertically on the plot to avoid having all
+points along the upper edge of the plot, while the base series is
+sampled horizontally with sampling weight proportional to their
+follow-up time. By randomly distributing the points, we can get a better
+sense of the incidence density. In the ERSPC data, we see that more
+events are observed later on in time. Therefore a constant hazard model
+would not be appropriate in this instance as it would overestimate the
+cumulative incidence earlier on in time, and underestimate it later on.
+The reader is referred to the package vignettes for more examples and a
+detailed description of how to modify the aesthetics of a population
+time plot.
 
@@
 
-For the first case study, we will use of the European Randomized Study
-of Prostate Cancer Screening data. This dataset is available through the
-\pkg{casebase} package:
+We investigate the differences in survival between the control and
+screening arms in the ERSPC study presented earlier using casebase
+sampling to fit a parametric hazard function. We fit four models that
+differ in the functional form of time; (1) excluded from the linear
+predictor (Eq.\eqref{eq:exp}), (2) linear function (Eq.\eqref{eq:gomp}),
+(3) log function (Eq.\eqref{eq:weibull}), and (4) a smooth function
+using cubic B-splines. The models are fit using \code{fitSmoothHazard}.
+Since the output object from \code{fitSmoothHazard} inherits from the
+\code{glm} class, we see a familiar result when using the function
+\code{summary}:
 
@@
 \begin{CodeInput}
-R> data(ERSPC)
-R> ERSPC$ScrArm <- factor(ERSPC$ScrArm,
-R+   levels = c(0, 1),
-R+   labels = c("Control group", "Screening group")
-R+ )
+R> fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
+R>              gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
+R>              weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
+R>              splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
+R> 
+R> fits <- lapply(fmla, function(i) {
+R>   casebase::fitSmoothHazard(i, data = ERSPC, ratio = 100)
+R> })
+R> 
+R> summary(fits[[2]])
 \end{CodeInput}
@@
 
-The results of this study were published by
-\citep{schroder2009screening}, and the dataset itself was obtained using
-the approach described in \citep{liu2014recovering}.
-
-Population time plots can be extremely informative graphical displays of
-survival data. They should be the first step in an exploratory data
-analysis. We facilitate this task in the \pkg{casebase} package using
-the \code{popTime} function. We first create the necessary dataset for
-producing the population time plots, and we can generate the plot by
-using the corresponding \code{plot} method:
+The \code{absoluteRisk} function takes as input the
+\code{fitSmoothHazard} object and returns a matrix where each column
+corresponds to the covariate profiles specified in the \code{newdata}
+argument, and each row corresponds to a specified time point:
 
@@
 \begin{CodeInput}
-R> pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
-R> plot(pt_object)
+R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
+R> 
+R> risks <- lapply(fits, function(i) {
+R>   casebase::absoluteRisk(object = i, time = seq(0,15,0.1), newdata = new_data)
+R> })
+R> 
+R> head(risks[[2]])
 \end{CodeInput}
-
-\begin{center}\includegraphics{../figures/plot-erspc-data-1} \end{center}
-
-\end{CodeChunk}
-
-We can also create exposure stratified plots by specifying the
-\code{exposure} argument in the \code{popTime} function:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> pt_object_strat <- casebase::popTime(ERSPC,
-R+   event = "DeadOfPrCa",
-R+   exposure = "ScrArm"
-R+ )
-R> plot(pt_object_strat)
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-1} \end{center}
-
-\end{CodeChunk}
-
-We can also plot them side-by-side using the \code{ncol} argument:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> plot(pt_object_strat, facet.params = list(ncol = 2))
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-side-1} \end{center}
-
-\end{CodeChunk}
-
-ADD PARAGRAPH ABOUT WHAT WE CAN CONCLUDE FROM THESE GRAPHS.
-
-Next, we investigate the differences between the control and the
-screening arms. A common choice for this type of analysis is to use a
-Cox regression model and estimate the hazard ratio for the screening
-group (relative to the control group). In \proglang{R}, it can be done
-as follows:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm,
-R+   data = ERSPC
-R+ )
-R> summary(cox_model)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> Call:
-#> survival::coxph(formula = Surv(Follow.Up.Time, DeadOfPrCa) ~ 
-#>     ScrArm, data = ERSPC)
-#> 
-#>   n= 159893, number of events= 540 
-#> 
-#>                         coef exp(coef) se(coef)     z Pr(>|z|)  
-#> ScrArmScreening group -0.222     0.801    0.088 -2.52    0.012 *
-#> ---
-#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-#> 
-#>                       exp(coef) exp(-coef) lower .95 upper .95
-#> ScrArmScreening group     0.801       1.25     0.674     0.952
-#> 
-#> Concordance= 0.519  (se = 0.011 )
-#> Likelihood ratio test= 6.45  on 1 df,   p=0.01
-#> Wald test            = 6.37  on 1 df,   p=0.01
-#> Score (logrank) test = 6.39  on 1 df,   p=0.01
-\end{CodeOutput}
-\end{CodeChunk}
-
-We can also plot the cumulative incidence function (CIF) for each group.
-However, this involves estimating the baseline hazard, which Cox
-regression treated as a nuisance parameter.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> new_data <- data.frame(
-R+   ScrArm = c("Control group", "Screening group"),
-R+   ignore = 99
-R+ )
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R+   xlab = "Years since Randomization",
-R+   ylab = "Cumulative Incidence (%)",
-R+   fun = "event",
-R+   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
-R+   main = sprintf(
-R+     "Estimated Cumulative Incidence (risk) of Death from Prostate 
-R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R+     exp(coef(cox_model)),
-R+     exp(confint(cox_model))[1],
-R+     exp(confint(cox_model))[2]
-R+   ),
-R+   ylim = c(0, 2), yscale = 100
-R+ )
-R> legend("topleft",
-R+   legend = c("Control group", "Screening group"),
-R+   col = c("red", "blue"),
-R+   lty = c(1, 1),
-R+   bg = "gray90"
-R+ )
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/erspc-cox-cif-1} \end{center}
-
-\end{CodeChunk}
-
-Next, we show how case-base sampling can be used to carry out a similar
-analysis. We fit several models that differ in how we choose to model
-time.
-
-First, we will fit an exponential model. Recall that this corresponds to
-excluding time from the linear predictor.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm,
-R+   data = ERSPC,
-R+   ratio = 100
-R+ )
-R> 
-R> summary(casebase_exponential)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> 
-#> Call:
-#> glm(formula = formula, family = binomial, data = sampleData)
-#> 
-#> Deviance Residuals: 
-#>    Min      1Q  Median      3Q     Max  
-#> -0.148  -0.148  -0.148  -0.133   3.078  
-#> 
-#> Coefficients:
-#>                       Estimate Std. Error z value            Pr(>|z|)    
-#> (Intercept)            -7.7659     0.0557 -139.46 <0.0000000000000002 ***
-#> ScrArmScreening group  -0.2147     0.0884   -2.43               0.015 *  
-#> ---
-#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-#> 
-#> (Dispersion parameter for binomial family taken to be 1)
-#> 
-#>     Null deviance: 6059  on 54539  degrees of freedom
-#> Residual deviance: 6053  on 54538  degrees of freedom
-#> AIC: 6057
-#> 
-#> Number of Fisher Scoring iterations: 7
-\end{CodeOutput}
-
-\begin{CodeInput}
-R> exp(coef(casebase_exponential)[2])
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> ScrArmScreening group 
-#>                  0.81
-\end{CodeOutput}
-
-\begin{CodeInput}
-R> exp(confint(casebase_exponential)[2, ])
-\end{CodeInput}
-
-\begin{CodeOutput}
-#>  2.5 % 97.5 % 
-#>   0.68   0.96
-\end{CodeOutput}
-\end{CodeChunk}
-
-We can then use the \code{absoluteRisk} function to get an estimate of
-the cumulative incidence curve for a specific covariate profile. In the
-plot below, we overlay the estimated CIF from the exponential model on
-the Cox model CIF:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> smooth_risk_exp <- casebase::absoluteRisk(
-R+   object = casebase_exponential,
-R+   time = seq(0, 15, 0.1),
-R+   newdata = new_data
-R+ )
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R+   xlab = "Years since Randomization",
-R+   ylab = "Cumulative Incidence (%)",
-R+   fun = "event",
-R+   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
-R+   main = sprintf(
-R+     "Estimated Cumulative Incidence (risk) of Death from Prostate 
-R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R+     exp(coef(cox_model)),
-R+     exp(confint(cox_model))[1],
-R+     exp(confint(cox_model))[2]
-R+   ),
-R+   ylim = c(0, 2), yscale = 100
-R+ )
-R> lines(smooth_risk_exp[, 1], smooth_risk_exp[, 2], col = "red", lty = 2)
-R> lines(smooth_risk_exp[, 1], smooth_risk_exp[, 3], col = "blue", lty = 2)
-R> 
-R> 
-R> legend("topleft",
-R+   legend = c(
-R+     "Control group (Cox)", "Control group (Casebase)",
-R+     "Screening group (Cox)", "Screening group (Casebase)"
-R+   ),
-R+   col = c("red", "red", "blue", "blue"),
-R+   lty = c(1, 2, 1, 2),
-R+   bg = "gray90"
-R+ )
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/erspc-casebase-cif-1} \end{center}
-
-\end{CodeChunk}
-
-As we can see, the exponential model gives us an estimate of the hazard
-ratio that is similar to the one obtained using Cox regression. However,
-the CIF estimates are quite different. Based on what we observed in the
+\end{CodeChunk}
+
+In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from
+casebase on the Cox model CIF. The CIF estimates for the exponential
+model in panel (1) overestimate the cumulative incidence earlier on in
+time, and underestimate it later on. Based on what we observed in the
 population time plot, where more events are observed later on in time,
-we do not expect a constant hazard would be a good description for this
-data. In other words, this poor fit for the exponential hazard is
-expected. A constant hazard model would overestimate the cumulative
-incidence earlier on in time, and underestimate it later on; this is
-what we see on the cumulative incidence plot. This example demonstrates
-the benefits of population time plots as an exploratory analysis tool.
-
-For the next model, we include time as a linear term. Recall that this
-corresponds to a Weibull model.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm,
-R+   data = ERSPC,
-R+   ratio = 100
-R+ )
-R> 
-R> summary(casebase_time)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> 
-#> Call:
-#> glm(formula = formula, family = binomial, data = sampleData)
-#> 
-#> Deviance Residuals: 
-#>    Min      1Q  Median      3Q     Max  
-#> -0.401  -0.162  -0.122  -0.094   3.466  
-#> 
-#> Coefficients:
-#>                       Estimate Std. Error z value            Pr(>|z|)    
-#> (Intercept)            -9.0314     0.1121  -80.53 <0.0000000000000002 ***
-#> Follow.Up.Time          0.2222     0.0145   15.29 <0.0000000000000002 ***
-#> ScrArmScreening group  -0.2320     0.0886   -2.62              0.0089 ** 
-#> ---
-#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-#> 
-#> (Dispersion parameter for binomial family taken to be 1)
-#> 
-#>     Null deviance: 6059  on 54539  degrees of freedom
-#> Residual deviance: 5811  on 54537  degrees of freedom
-#> AIC: 5817
-#> 
-#> Number of Fisher Scoring iterations: 8
-\end{CodeOutput}
-
-\begin{CodeInput}
-R> exp(coef(casebase_time))
-\end{CodeInput}
-
-\begin{CodeOutput}
-#>           (Intercept)        Follow.Up.Time ScrArmScreening group 
-#>               0.00012               1.24882               0.79297
-\end{CodeOutput}
-
-\begin{CodeInput}
-R> exp(confint(casebase_time))
-\end{CodeInput}
-
-\begin{CodeOutput}
-#>                          2.5 %  97.5 %
-#> (Intercept)           0.000096 0.00015
-#> Follow.Up.Time        1.213876 1.28507
-#> ScrArmScreening group 0.665769 0.94256
-\end{CodeOutput}
-\end{CodeChunk}
-
-Again, the estimate of the hazard ratio is similar to that obtained from
-Cox regression. We then look at the estimate of the CIF:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> smooth_risk_time <- casebase::absoluteRisk(
-R+   object = casebase_time,
-R+   time = seq(0, 15, 0.1),
-R+   newdata = new_data
-R+ )
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R+   xlab = "Years since Randomization",
-R+   ylab = "Cumulative Incidence (%)",
-R+   fun = "event",
-R+   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
-R+   main = sprintf(
-R+     "Estimated Cumulative Incidence (risk) of Death from Prostate 
-R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R+     exp(coef(cox_model)),
-R+     exp(confint(cox_model))[1],
-R+     exp(confint(cox_model))[2]
-R+   ),
-R+   ylim = c(0, 2), yscale = 100
-R+ )
-R> lines(smooth_risk_time[, 1], smooth_risk_time[, 2], col = "red", lty = 2)
-R> lines(smooth_risk_time[, 1], smooth_risk_time[, 3], col = "blue", lty = 2)
-R> 
-R> legend("topleft",
-R+   legend = c(
-R+     "Control group (Cox)", "Control group (Casebase)",
-R+     "Screening group (Cox)", "Screening group (Casebase)"
-R+   ),
-R+   col = c("red", "red", "blue", "blue"),
-R+   lty = c(1, 2, 1, 2),
-R+   bg = "gray90"
-R+ )
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/erspc-casebase-weibull-cif-1} \end{center}
-
-\end{CodeChunk}
-
-We see that the Weibull model leads to a better fit of the CIF than the
-Exponential model, when compared with the semi-parametric approach.
-
-Finally, we can model the hazard as a smooth function of time using the
-\code{splines} package:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm,
-R+   data = ERSPC,
-R+   ratio = 100
-R+ )
-R> 
-R> summary(casebase_splines)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> 
-#> Call:
-#> glm(formula = formula, family = binomial, data = sampleData)
-#> 
-#> Deviance Residuals: 
-#>    Min      1Q  Median      3Q     Max  
-#> -0.405  -0.173  -0.136  -0.085   3.796  
-#> 
-#> Coefficients:
-#>                       Estimate Std. Error z value             Pr(>|z|)    
-#> (Intercept)           -10.2643     0.3159  -32.49 < 0.0000000000000002 ***
-#> bs(Follow.Up.Time)1     4.2983     0.8005    5.37        0.00000007893 ***
-#> bs(Follow.Up.Time)2     2.0945     0.4824    4.34        0.00001412028 ***
-#> bs(Follow.Up.Time)3     4.5544     0.7067    6.44        0.00000000012 ***
-#> ScrArmScreening group  -0.2169     0.0886   -2.45                0.014 *  
-#> ---
-#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-#> 
-#> (Dispersion parameter for binomial family taken to be 1)
-#> 
-#>     Null deviance: 6059.0  on 54539  degrees of freedom
-#> Residual deviance: 5790.5  on 54535  degrees of freedom
-#> AIC: 5800
-#> 
-#> Number of Fisher Scoring iterations: 8
-\end{CodeOutput}
-
-\begin{CodeInput}
-R> exp(coef(casebase_splines))
-\end{CodeInput}
-
-\begin{CodeOutput}
-#>           (Intercept)   bs(Follow.Up.Time)1   bs(Follow.Up.Time)2 
-#>              0.000035             73.578176              8.121029 
-#>   bs(Follow.Up.Time)3 ScrArmScreening group 
-#>             95.053534              0.805017
-\end{CodeOutput}
-
-\begin{CodeInput}
-R> exp(confint(casebase_splines))
-\end{CodeInput}
-
-\begin{CodeOutput}
-#>                           2.5 %     97.5 %
-#> (Intercept)            0.000018   0.000063
-#> bs(Follow.Up.Time)1   15.835008 367.061126
-#> bs(Follow.Up.Time)2    3.252368  21.652009
-#> bs(Follow.Up.Time)3   22.537699 364.141700
-#> ScrArmScreening group  0.675920   0.956823
-\end{CodeOutput}
-\end{CodeChunk}
-
-Once again, we can see that the estimate of the hazard ratio is similar
-to that from the Cox regression. We then look at the estimate of the
-CIF:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> smooth_risk_splines <- absoluteRisk(
-R+   object = casebase_splines,
-R+   time = seq(0, 15, 0.1),
-R+   newdata = new_data
-R+ )
-R> 
-R> plot(survfit(cox_model, newdata = new_data),
-R+   xlab = "Years since Randomization",
-R+   ylab = "Cumulative Incidence (%)",
-R+   fun = "event",
-R+   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
-R+   main = sprintf(
-R+     "Estimated Cumulative Incidence (risk) of Death from Prostate 
-R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-R+     exp(coef(cox_model)),
-R+     exp(confint(cox_model))[1],
-R+     exp(confint(cox_model))[2]
-R+   ),
-R+   ylim = c(0, 2), yscale = 100
-R+ )
-R> lines(smooth_risk_splines[, 1], smooth_risk_splines[, 2], col = "red", lty = 2)
-R> lines(smooth_risk_splines[, 1], smooth_risk_splines[, 3], col = "blue", lty = 2)
-R> 
-R> legend("topleft",
-R+   legend = c(
-R+     "Control group (Cox)", "Control group (Casebase)",
-R+     "Screening group (Cox)", "Screening group (Casebase)"
-R+   ),
-R+   col = c("red", "red", "blue", "blue"),
-R+   lty = c(1, 2, 1, 2),
-R+   bg = "gray90"
-R+ )
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/erspc-casebase-splines-cif-1} \end{center}
-
-\end{CodeChunk}
-
-Qualitatively, the combination of splines and case-base sampling
-produces the parametric model that most closely resembles the Cox model.
-In the following table, we see that the confidence intervals are also
-similar across all four models. In other words, this reinforces the idea
-that, under proportional hazards, we do not need to model the full
-hazard to obtain reliable estimates of the hazard ratio. Of course,
-different parametric models for the hazards give rise to qualitatively
-different estimates for the CIF.
-
-\begin{CodeChunk}
-\begin{table}[H]
-\centering
-\begin{tabular}{lrlr}
-\toprule
-Model & HR & CI & Width\\
-\midrule
-Cox & 0.80 & (0.67, 0.95) & 0.28\\
-Exponential & 0.81 & (0.68, 0.96) & 0.28\\
-Weibull & 0.79 & (0.67, 0.94) & 0.28\\
-Splines & 0.81 & (0.68, 0.96) & 0.28\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\end{CodeChunk}
-
-As noted above, the usual asymptotic results hold for likelihood ratio
-tests built using case-base sampling models. Therefore, we can easily
-test the null hypothesis that the exponential model is just as good as
-the larger (in terms of number of parameters) splines model.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> anova(casebase_exponential, casebase_splines, test = "LRT")
-\end{CodeInput}
+this poor fit for the exponential hazard is expected. This example
+demonstrates the benefits of population time plots as an exploratory
+analysis tool.
+
+In Figure \ref{fig:erspc-cox-cif} panels (2)-(4), we notice a better fit
+with increasing complexities of time. As noted above, the usual
+asymptotic results hold for likelihood ratio tests built using casebase
+sampling models. Therefore, we can easily test the null hypothesis that
+the exponential model is just as good as the larger (in terms of number
+of parameters) splines model:
+
+\begin{CodeChunk}
 
@@
 #> 1     54538       6053                                    
-#> 2     54535       5790  3      262 <0.0000000000000002 ***
+#> 2     54535       5788  3      265 <0.0000000000000002 ***
 #> ---
@@
 As expected, we see that splines model provides a better fit. Similarly,
-we can compare the AIC for all three case-base models:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> c("Exp." = AIC(casebase_exponential),
-R+   "Weib." = AIC(casebase_time),
-R+   "Splines" = AIC(casebase_splines))
-\end{CodeInput}
+we can compare the AIC for all four casebase models:
+
+\begin{CodeChunk}
 
 \begin{CodeOutput}
-#>    Exp.   Weib. Splines 
-#>    6057    5817    5800
+#>    Exp.   Gomp.   Weib. Splines 
+#>    6057    5818    6057    5798
 \end{CodeOutput}
@@
 fit.
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/erspc-cox-cif-1}
+
+}
+
+\caption{Cumulative incidence functions for control and screening groups in the ERSPC data.
+In each of the panels, we plot the CIF from the Cox model using \code{survival::survfit} (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression.
+(1) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution.
+(2) Linear function of time which is equivalent to a Gompertz distribution.
+(3) The natural logarithm of time which is equivalent to a Weibull distribution.
+(4) Cubic B-spline expansion of time.}\label{fig:erspc-cox-cif}
+\end{figure}
+\end{CodeChunk}
+
+In the following table, we present a side-by-side comparison of the
+hazard ratios and confidence intervals estimated from
+\code{fitSmoothHazard} and the corresponding parametric model using
+\code{survival::survreg}, as well as the Cox model estimate. The hazard
+ratio estimates and confidence intervals are similar across all four
+models. This reinforces the idea that, under proportional hazards, we do
+not need to model the full hazard to obtain reliable estimates of the
+hazard ratio. Of course, different parametric models for the hazards
+give rise to qualitatively different estimates for the CIF as shown in
+Figure \ref{fig:erspc-cox-cif}.
+
+\begin{CodeChunk}
+\begin{table}
+
+\caption{\label{tab:print-erspc-estimates}Comparison of estimated hazard ratios and 95\% confidence intervals for ERSPC data.}
+\centering
+\begin{tabular}[t]{>{\bfseries}lcc}
+\toprule
+Model & casebase::fitSmoothHazard & survival::survreg\\
+\midrule
+Exponential & 0.81 (0.68, 0.96) & 0.81 (0.68, 0.96)\\
+\cmidrule{1-3}
+Gompertz & 0.79 (0.66, 0.94) & 0.80 (0.67, 0.95)\\
+\cmidrule{1-3}
+Weibull & 0.81 (0.68, 0.97) & 0.80 (0.65, 0.96)\\
+\cmidrule{1-3}
+Splines & 0.80 (0.67, 0.95) & --\\
+\bottomrule
+\multicolumn{3}{l}{Cox model estimate: HR (95\% CI) = 0.80 (0.67, 0.95)}\\
+\end{tabular}
+\end{table}
+
+\end{CodeChunk}
 
@@
 \pkg{casebase} package.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> data(bmtcrr)
-\end{CodeInput}
-\end{CodeChunk}
 
@@
 using a linear term for time:
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> model_cb <- fitSmoothHazard(
-R+   Status ~ ftime + Sex + D + Phase + Source + Age,
-R+   data = bmtcrr,
-R+   ratio = 100,
-R+   time = "ftime"
-R+ )
-\end{CodeInput}
-\end{CodeChunk}
 
@@
 R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-R+   data = bmtcrr
-R+ )
+R>   data = bmtcrr
+R> )
 \end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{table}
-
-\caption{\label{tab:bmtcrr-cis}Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.}
-\centering
-\begin{tabular}[t]{lrlrl}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
-Covariates & HR & 95\% CI & HR & 95\% CI\\
-\midrule
-Sex & 0.71 & (0.41, 1.24) & 0.68 & (0.39, 1.19)\\
-Disease & 0.55 & (0.31, 1) & 0.52 & (0.29, 0.93)\\
-Phase (CR2 vs. CR1) & 1.18 & (0.47, 2.93) & 1.21 & (0.48, 3.02)\\
-Phase (CR3 vs. CR1) & 1.47 & (0.38, 5.63) & 1.67 & (0.43, 6.5)\\
-Phase (Relapse vs. CR1) & 4.30 & (2, 9.27) & 4.55 & (2.09, 9.9)\\
-\addlinespace
-Source & 1.56 & (0.52, 4.7) & 1.46 & (0.47, 4.54)\\
-Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
-\bottomrule
-\end{tabular}
-\end{table}
-
 \end{CodeChunk}
@@
 myeloblastic leukemia. We will estimate the curve from 0 to 60 months.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Pick 100 equidistant points between 0 and 60 months
-R> time_points <- seq(0, 60, length.out = 50)
-R> 
-R> # Data.frame containing risk profile
-R> newdata <- data.frame(
-R+   "Sex" = factor(c("F", "F"),
-R+     levels = levels(bmtcrr[, "Sex"])
-R+   ),
-R+   "D" = c("ALL", "AML"), # Both diseases
-R+   "Phase" = factor(c("Relapse", "Relapse"),
-R+     levels = levels(bmtcrr[, "Phase"])
-R+   ),
-R+   "Age" = c(35, 35),
-R+   "Source" = factor(c("PB", "PB"),
-R+     levels = levels(bmtcrr[, "Source"])
-R+   )
-R+ )
-R> 
-R> # Estimate absolute risk curve
-R> risk_cb <- absoluteRisk(
-R+   object = model_cb, time = time_points,
-R+   method = "numerical", newdata = newdata
-R+ )
-\end{CodeInput}
-\end{CodeChunk}
 
@@
 R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) +
-R+                         const(Phase) + const(Source) + const(Age),
-R+                       data = bmtcrr, cause = 1, model = "fg")
+R>                         const(Phase) + const(Source) + const(Age),
+R>                       data = bmtcrr, cause = 1, model = "fg")
 R> 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> data(support)
-R> # Change time to years
-R> support$d.time <- support$d.time/365.25
-R> 
-R> # Split into test and train
-R> train_index <- sample(nrow(support), 0.95*nrow(support))
-R> test_index <- setdiff(1:nrow(support), train_index)
-R> 
-R> train <- support[train_index,]
-R> test <- support[test_index,]
-\end{CodeInput}
-\end{CodeChunk}
-
 In Figure\ref{fig:brier1} (A), we can see that all Cox models have
@@
 approach from \pkg{casebase}.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/brierpt1-1}
-
-}
-
-\caption{\label{fig:brier1} Comparison of three Cox regression models. (A) Cumulative Incidence curves. (B) Time-dependent Brier scores.}\label{fig:brierpt1}
-\end{figure}
-\end{CodeChunk}
 
@@
 R> x <- model.matrix(death ~ . - d.time - aps - sps, 
-R+                   data = train)[, -c(1)] # Remove intercept
+R>                   data = train)[, -c(1)] # Remove intercept
 R> y <- data.matrix(subset(train, select = c(d.time, death)))
@@
 R> cbFull <- casebase::fitSmoothHazard.fit(x, y,
-R+   family = "glmnet",
-R+   time = "d.time", event = "death",
-R+   formula_time = ~ log(d.time), alpha = 1,
-R+   ratio = 10, standardize = TRUE,
-R+   penalty.factor = c(0, rep(1, ncol(x)))
-R+ )
+R>   family = "glmnet",
+R>   time = "d.time", event = "death",
+R>   formula_time = ~ log(d.time), alpha = 1,
+R>   ratio = 10, standardize = TRUE,
+R>   penalty.factor = c(0, rep(1, ncol(x)))
+R> )
 \end{CodeInput}
@@
 R> newx <- model.matrix(death ~ . - d.time - aps - sps, 
-R+                   data = test)[, -c(1)]
+R>                   data = test)[, -c(1)]
 R> times <- sort(unique(test$d.time))
@@
 R> abCbFull <- casebase::absoluteRisk(cbFull,
-R+   time = times,
-R+   newdata = newx,
-R+   s = "lambda.1se",
-R+   method = "numerical"
-R+ )
+R>   time = times,
+R>   newdata = newx,
+R>   s = "lambda.1se",
+R>   method = "numerical"
+R> )
 \end{CodeInput}
@@
 (cano).
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/coefplots-1}
-
-}
-
-\caption{\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models.
-If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity.}\label{fig:coefplots}
-\end{figure}
-\end{CodeChunk}
 
@@
 three models outperform the Kaplan-Meier model.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/riskregressionBrier-1}
-
-}
-
-\caption{\label{fig:cs3FinalBrier} Four curves comparing the Cumulative Incidence (A) and the Brier scores (B) as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.}\label{fig:riskregressionBrier}
-\end{figure}
-\end{CodeChunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(survival)
-R> library(casebase)
-R> 
-R> stanford_popTime <- popTime(jasa,
-R+   time = "futime",
-R+   event = "fustat"
-R+ )
-R> plot(stanford_popTime)
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/stanford-poptime-1} \end{center}
-
-\end{CodeChunk}
-
 Since the exposure is time-dependent, we need to manually define the
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(tidyverse)
-R> library(lubridate)
-R> 
-R> cb_data <- sampleCaseBase(jasa,
-R+   time = "futime",
-R+   event = "fustat", ratio = 10
-R+ )
-\end{CodeInput}
-\end{CodeChunk}
-
 Next, we will compute the number of days from acceptance into the
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Define exposure variable
-R> cb_data <- mutate(cb_data,
-R+   txtime = time_length(accept.dt %--% tx.date,
-R+                        unit = "days"
-R+   ),
-R+   exposure = case_when(
-R+     is.na(txtime) ~ 0L,
-R+     txtime > futime ~ 0L,
-R+     txtime <= futime ~ 1L
-R+   )
-R+ )
-\end{CodeInput}
-\end{CodeChunk}
-
 Finally, we can fit the hazard using various linear predictors.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(splines)
-R> # Fit several models
-R> fit1 <- fitSmoothHazard(fustat ~ exposure,
-R+   data = cb_data, time = "futime"
-R+ )
-R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
-R+   data = cb_data, time = "futime"
-R+ )
-R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
-R+   data = cb_data, time = "futime"
-R+ )
-R> fit4 <- fitSmoothHazard(fustat ~ exposure * bs(futime),
-R+   data = cb_data, time = "futime"
-R+ )
-\end{CodeInput}
-\end{CodeChunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute AIC
-R> c("Model1" = AIC(fit1),
-R+   "Model2" = AIC(fit2),
-R+   "Model3" = AIC(fit3),
-R+   "Model4" = AIC(fit4))
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> Model1 Model2 Model3 Model4 
-#>    801    756    718    724
-\end{CodeOutput}
-\end{CodeChunk}
-
 As we can, the best fit is the fourth model. By visualizing the hazard
@@
 clearly see how the hazards are no longer proportional.
+
+THIS NEXT CODE CHUNK IS NOT WORKING
 
@@
 R> hazard_data <- expand.grid(
-R+   exposure = c(0, 1),
-R+   futime = seq(0, 1000,
-R+     length.out = 100
-R+   )
-R+ )
+R>   exposure = c(0, 1),
+R>   futime = seq(0, 1000,
+R>     length.out = 100
+R>   )
+R> )
 R> # Set the offset to zero
@@
 R> hazard_data$hazard <- exp(predict(fit4,
-R+   newdata = hazard_data,
-R+   type = "link"
-R+ ))
+R>   newdata = hazard_data,
+R>   type = "link"
+R> ))
 R> # Add labels for plots
 R> hazard_data$Status <- factor(hazard_data$exposure,
-R+   labels = c("NoTrans", "Trans")
-R+ )
+R>   labels = c("NoTrans", "Trans")
+R> )
 R> 
 R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
-R+   geom_line() +
-R+   theme_minimal() +
-R+   theme(legend.position = "top") +
-R+   ylab("Hazard") +
-R+   xlab("Follow-up time")
+R>   geom_line() +
+R>   theme_minimal() +
+R>   theme(legend.position = "top") +
+R>   ylab("Hazard") +
+R>   xlab("Follow-up time")
 \end{CodeInput}
-
-\begin{center}\includegraphics{../figures/stanford-hazard-1} \end{center}
-
 \end{CodeChunk}
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute absolute risk curves
-R> newdata <- data.frame(exposure = c(0, 1))
-R> absrisk <- absoluteRisk(fit4,
-R+   newdata = newdata,
-R+   time = seq(0, 1000, length.out = 100)
-R+ )
-R> 
-R> colnames(absrisk) <- c("Time", "NoTrans", "Trans")
-R> 
-R> # Rearrange the data
-R> absrisk <- gather(
-R+   as.data.frame(absrisk),
-R+   "Status", "Risk", -Time
-R+ )
-R> 
-R> ggplot(absrisk, aes(Time, Risk, colour = Status)) +
-R+   geom_line() +
-R+   theme_minimal() +
-R+   theme(legend.position = "top") +
-R+   expand_limits(y = 0.5) +
-R+   xlab("Follow-up time") +
-R+   ylab("Cum. Incidence")
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/stanford-risk-1} \end{center}
-
-\end{CodeChunk}
-
 Note that we can easily adapt the code above to the situation where a
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute hazards---
-R> # First, create a list of time points for both exposure status
-R> one_yr_haz <- data.frame(futime = seq(0, 365,
-R+   length.out = 100
-R+ ))
-R> one_yr_haz <- mutate(one_yr_haz,
-R+   offset = 0,
-R+   exposure = if_else(futime < 30, 0, 1)
-R+ )
-R> one_yr_haz$hazard <- exp(predict(fit4,
-R+   newdata = one_yr_haz,
-R+   type = "link"
-R+ ))
-R> ggplot(one_yr_haz, aes(futime, hazard)) +
-R+   geom_line() +
-R+   theme_minimal() +
-R+   theme(legend.position = "top") +
-R+   ylab("Hazard") +
-R+   xlab("Follow-up time") +
-R+   geom_vline(xintercept = 30, linetype = "dashed")
-\end{CodeInput}
-
-\begin{center}\includegraphics{../figures/stanford-1y-haz-1} \end{center}
-
-\end{CodeChunk}
-
 We can then compare the 1-year mortality risk without transplant and
 with transplant at 30 days.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> absoluteRisk(fit4,
-R+   newdata = data.frame(exposure = 0),
-R+   time = 365
-R+ )
-\end{CodeInput}
-
-\begin{CodeOutput}
-#>         
-#> 365 0.26
-\end{CodeOutput}
-
-\begin{CodeInput}
-R> 
-R> # Use the trapezoidal rule to estimate the cumulative hazard
-R> cumhaz_est <- pracma::trapz(one_yr_haz$futime, one_yr_haz$hazard)
-R> 1 - exp(-cumhaz_est)
-\end{CodeInput}
-
-\begin{CodeOutput}
-#> [1] 0.18
-\end{CodeOutput}
-\end{CodeChunk}
 
@@
 
-This report was generated on 2020-05-07 16:35:15 using the following
+This report was generated on 2020-07-18 13:50:44 using the following
 computational environment and dependencies:
@@
 
-\begin{CodeInput}
-R> # which R packages and versions?
-R> # devtools::session_info()
-R> sessionInfo()
-\end{CodeInput}
-
 \begin{CodeOutput}
-#> R version 4.0.0 (2020-04-24)
-#> Platform: x86_64-apple-darwin17.0 (64-bit)
-#> Running under: macOS Catalina 10.15.4
+#> R version 3.6.2 (2019-12-12)
+#> Platform: x86_64-pc-linux-gnu (64-bit)
+#> Running under: Pop!_OS 19.10
 #> 
 #> Matrix products: default
-#> BLAS:   /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRblas.dylib
-#> LAPACK: /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRlapack.dylib
+#> BLAS:   /usr/lib/x86_64-linux-gnu/openblas/libblas.so.3
+#> LAPACK: /usr/lib/x86_64-linux-gnu/libopenblasp-r0.3.7.so
 #> 
 #> locale:
-#> [1] en_CA.UTF-8/en_CA.UTF-8/en_CA.UTF-8/C/en_CA.UTF-8/en_CA.UTF-8
+#>  [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
+#>  [3] LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
+#>  [5] LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
+#>  [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                 
+#>  [9] LC_ADDRESS=C               LC_TELEPHONE=C            
+#> [11] LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       
 #> 
@@
 #> other attached packages:
-#>  [1] lubridate_1.7.8           broom_0.5.6              
-#>  [3] dotwhisker_0.5.0          timereg_1.9.4            
-#>  [5] glue_1.4.0                forcats_0.5.0            
-#>  [7] stringr_1.4.0             dplyr_0.8.5              
-#>  [9] purrr_0.3.4               readr_1.3.1              
-#> [11] tidyr_1.0.2               ggplot2_3.3.0            
-#> [13] tidyverse_1.3.0           tibble_3.0.1             
-#> [15] survival_3.1-12           riskRegression_2020.02.05
-#> [17] prodlim_2019.11.13        pracma_2.2.9             
-#> [19] magrittr_1.5              kableExtra_1.1.0         
-#> [21] glmnet_3.0-2              Matrix_1.2-18            
-#> [23] data.table_1.12.8         cowplot_1.0.0            
-#> [25] colorspace_1.4-1          casebase_0.2.1.9001      
+#>  [1] forcats_0.5.0             stringr_1.4.0            
+#>  [3] dplyr_0.8.5               purrr_0.3.4              
+#>  [5] readr_1.3.1               tidyr_1.1.0              
+#>  [7] ggplot2_3.3.2.9000        tidyverse_1.3.0          
+#>  [9] lubridate_1.7.4           visreg_2.7.0             
+#> [11] tibble_3.0.1              riskRegression_2020.02.05
+#> [13] prodlim_2019.11.13        flexsurv_1.1.1           
+#> [15] survival_3.1-8            pracma_2.2.9             
+#> [17] magrittr_1.5              kableExtra_1.1.0         
+#> [19] glmnet_4.0-2              Matrix_1.2-18            
+#> [21] data.table_1.12.8         cowplot_1.0.0            
+#> [23] colorspace_1.4-1          casebase_0.9.0           
 #> 
 #> loaded via a namespace (and not attached):
-#>  [1] TH.data_1.0-10      VGAM_1.1-2          ellipsis_0.3.0     
-#>  [4] ggstance_0.3.4      htmlTable_1.13.3    base64enc_0.1-3    
-#>  [7] fs_1.4.1            rstudioapi_0.11     farver_2.0.3       
-#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-0      
-#> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.28         
-#> [16] Formula_1.2-3       jsonlite_1.6.1      cluster_2.1.0      
-#> [19] dbplyr_1.4.3        png_0.1-7           compiler_4.0.0     
-#> [22] httr_1.4.1          backports_1.1.6     assertthat_0.2.1   
-#> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.4.0    
-#> [28] quantreg_5.55       tools_4.0.0         gtable_0.3.0       
-#> [31] Rcpp_1.0.4.6        cellranger_1.1.0    vctrs_0.2.4        
-#> [34] nlme_3.1-147        iterators_1.0.12    xfun_0.13          
-#> [37] rvest_0.3.5         lifecycle_0.2.0     polspline_1.1.17   
-#> [40] MASS_7.3-51.5       zoo_1.8-7           scales_1.1.0       
-#> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
-#> [46] RColorBrewer_1.1-2  rticles_0.14        yaml_2.2.1         
-#> [49] gridExtra_2.3       rms_5.1-4           rpart_4.1-15       
-#> [52] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
-#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
-#> [58] rlang_0.4.5         pkgconfig_2.0.3     evaluate_0.14      
-#> [61] lattice_0.20-41     htmlwidgets_1.5.1   labeling_0.3       
-#> [64] cmprsk_2.2-9        tidyselect_1.0.0    plyr_1.8.6         
-#> [67] R6_2.4.1            generics_0.0.2      Hmisc_4.4-0        
-#> [70] multcomp_1.4-13     DBI_1.1.0           pillar_1.4.3       
-#> [73] haven_2.2.0         foreign_0.8-78      withr_2.2.0        
-#> [76] mgcv_1.8-31         nnet_7.3-13         modelr_0.1.6       
-#> [79] crayon_1.3.4        rmarkdown_2.1       jpeg_0.1-8.1       
-#> [82] grid_4.0.0          readxl_1.3.1        reprex_0.3.0       
-#> [85] digest_0.6.25       webshot_0.5.2       numDeriv_2016.8-1.1
-#> [88] stats4_4.0.0        munsell_0.5.0       viridisLite_0.3.0
+#>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
+#>  [4] htmlTable_1.13.3    base64enc_0.1-3     fs_1.3.2           
+#>  [7] rstudioapi_0.11     farver_2.0.3        MatrixModels_0.4-1 
+#> [10] fansi_0.4.1         mvtnorm_1.0-12      xml2_1.3.0         
+#> [13] codetools_0.2-16    knitr_1.29          Formula_1.2-3      
+#> [16] jsonlite_1.7.0      broom_0.7.0         cluster_2.1.0      
+#> [19] dbplyr_1.4.2        png_0.1-7           compiler_3.6.2     
+#> [22] httr_1.4.1          backports_1.1.8     assertthat_0.2.1   
+#> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
+#> [28] quantreg_5.54       tools_3.6.2         gtable_0.3.0       
+#> [31] glue_1.4.1          Rcpp_1.0.4.6        cellranger_1.1.0   
+#> [34] vctrs_0.3.1         nlme_3.1-143        iterators_1.0.12   
+#> [37] xfun_0.15           rvest_0.3.5         lifecycle_0.2.0    
+#> [40] polspline_1.1.17    muhaz_1.2.6.1       MASS_7.3-51.5      
+#> [43] zoo_1.8-7           scales_1.1.1        hms_0.5.3          
+#> [46] sandwich_2.5-1      SparseM_1.78        RColorBrewer_1.1-2 
+#> [49] rticles_0.13        yaml_2.2.1          gridExtra_2.3      
+#> [52] rms_5.1-4           rpart_4.1-15        latticeExtra_0.6-29
+#> [55] stringi_1.4.6       foreach_1.5.0       checkmate_2.0.0    
+#> [58] lava_1.6.6          shape_1.4.4         rlang_0.4.6        
+#> [61] pkgconfig_2.0.3     evaluate_0.14       lattice_0.20-38    
+#> [64] labeling_0.3        htmlwidgets_1.5.1   cmprsk_2.2-9       
+#> [67] tidyselect_1.1.0    deSolve_1.27.1      R6_2.4.1           
+#> [70] generics_0.0.2      Hmisc_4.3-1         multcomp_1.4-12    
+#> [73] DBI_1.1.0           withr_2.2.0         pillar_1.4.4       
+#> [76] haven_2.3.1         foreign_0.8-74      mgcv_1.8-31        
+#> [79] nnet_7.3-12         mstate_0.2.12       modelr_0.1.5       
+#> [82] crayon_1.3.4        rmarkdown_2.3       timereg_1.9.4      
+#> [85] jpeg_0.1-8.1        grid_3.6.2          readxl_1.3.1       
+#> [88] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
+#> [91] numDeriv_2016.8-1.1 stats4_3.6.2        munsell_0.5.0      
+#> [94] viridisLite_0.3.0   quadprog_1.5-8
 \end{CodeOutput}
@@
 
-\begin{CodeInput}
-R> # what commit is this file at?
-R> git2r::repository(here::here())
-\end{CodeInput}
-
 \begin{CodeOutput}
-#> Local:    editMax /Users/maxturgeon/Documents/git_repos/cbpaper
-#> Head:     [c9ddc46] 2020-05-07: combine ci and brier plots
+#> Local:    write-popTime /home/sahir/git_repositories/cbpaper
+#> Remote:   write-popTime @ origin (https://github.com/sahirbhatnagar/cbpaper.git)
+#> Head:     [84d5ff5] 2020-06-29: merge pop-stash
 \end{CodeOutput}

### 2020-07-18 analysis/paper/paper.Rmd
fixed comparison table. and consistent themes or figures

@@
 eval_cs1 <- TRUE
-eval_cs2 <- FALSE
-eval_cs3 <- FALSE
-eval_cs4 <- FALSE
+eval_cs2 <- TRUE
+eval_cs3 <- TRUE
+eval_cs4 <- TRUE
 eval_colophon <- TRUE
@@
 Also, \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of time using splines, which allows flexible modeling of the hazard function.
-Moreover, \pkg{flexsurv} has the ability to estimate both scale and shape parameters for a variety of parametric families (see Table \ref{tab:dists}).
+Moreover, \pkg{flexsurv} has the ability to estimate both scale and shape parameters for a variety of parametric families.
 As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear combination of covariates (including time).
@@
 \hline
-\textbf{Package}        & \vtop{\hbox{\strut \textbf{Competing}}\hbox{\strut \textbf{risks}}}  & \textbf{Non-proportional} & \textbf{Penalization} & \textbf{Splines} & \textbf{Parametric} & \textbf{Semi-parametric} & \vtop{\hbox{\strut \textbf{Interval/left}}\hbox{\strut \textbf{censoring}}} & \vtop{\hbox{\strut \textbf{Absolute}}\hbox{\strut \textbf{risk}}}           \\
+& \textbf{Competing} & \textbf{Allows} &  \textbf{Penalized}   &    &  & \textbf{Semi} & \textbf{Interval/Left} & \textbf{Absolute}  \\
+\textbf{Package}        & \textbf{Risks}  & \textbf{Non PH} & \textbf{Likelihood} & \textbf{Splines} & \textbf{Parametric} & \textbf{Parametric} & \textbf{Censoring} & \textbf{Risk}           \\
 \hline
-\textbf{casebase} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
-\textbf{CFC} \cite{mahani2015bayesian}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
-textbf{cmprsk} \cite{gray_2019}            & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\
-\textbf{coxRidge} \cite{perperoglou}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{crrp} \cite{fu}           & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
-\textbf{fastcox} \cite{yi_zou}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{Flexrsurv} \cite{flexrsurv}      &                          & x                         &                       & x                & x                   &                          &                                  & x               \\
-\textbf{Flexsurv} \cite{flexsurv}       & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\
-\textbf{glmnet} \cite{regpathcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{glmpath} \cite{park_hastie}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{mets} \cite{scheike2014estimating}           & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
-\textbf{penalized} \cite{goeman_meijer2019}      &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{RiskRegression} \cite{gerds_blanche} &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
-\textbf{Rstpm2} \cite{clements_liu}         &                          & x                         & x                     & x                & x                   & x                        & x                                & x                         \\
-\textbf{SmoothHazard} \cite{smoothHazard}   &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
-\textbf{Survival} \cite{survival-package}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x                               \\
+\textbf{casebase}        & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
+\textbf{CFC}             & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
+\textbf{cmprsk}             & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\
+\textbf{coxRidge}        &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{crrp}            & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
+\textbf{fastcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{Flexrsurv}       &                          & x                         &                       & x                & x                   &                          &                                  & x               \\
+\textbf{Flexsurv}        & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\
+\textbf{glmnet}          &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{glmpath}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{mets}            & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
+\textbf{penalized}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{RiskRegression}  &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
+\textbf{Rstpm2}          &                          & x                         & x                     & x                & x                   & x                        & x                                & x                         \\
+\textbf{SmoothHazard}    &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
+\textbf{Survival}        & x                        & x                         &                       &                  & x                   & x                        & x                                & x                               \\
 \hline
@@
 }
-\caption{Different features of interest in various survival packages.}
+\caption{Comparison of various \proglang{R} packages for survival analysis based on several defining features.}
 \label{tab:surv-pkgs}
@@
 
-```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE}
+```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plot for ERSPC data."}
 plot(pt_object, 
      add.base.series = TRUE,
-     ratio = 0.5,
+     ratio = 0.7,
      facet.params = list(ncol = 2),
-     ribbon.params = list(fill = "gray30"),
-     case.params = list(size = 0.5),
-     base.params = list(size = 0.5)) + theme_bw(base_size = 10) + theme(legend.position = "bottom") + 
+     ribbon.params = list(fill = "gray50"),
+     case.params = list(size = 0.8),
+     base.params = list(size = 0.8)) + theme_bw(base_size = 10) + theme(legend.position = "bottom") + 
   scale_y_continuous(labels = scales::label_number_si())
@@
 
+```{r, eval = eval_cs1, echo=FALSE}
+fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
+             gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
+             weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
+             splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
+
+fits <- lapply(fmla, function(i) {
+  casebase::fitSmoothHazard(i, data = ERSPC, ratio = 100)
+})
+```
+
 We investigate the differences in survival between the control and screening arms in the ERSPC study presented earlier using casebase sampling to fit a parametric hazard function.
@@
 
+```{r, eval = eval_cs1, echo=FALSE}
+summary(fits[[2]])
+```
+
 The \code{absoluteRisk} function takes as input the \code{fitSmoothHazard} object and returns a matrix where each column corresponds to the covariate profiles specified in the \code{newdata} argument, and each row corresponds to a specified time point:
@@
 Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) splines model:
-
-```{r, eval = eval_cs1, echo=FALSE}
-fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
-             gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
-             weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
-             splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
-
-fits <- lapply(fmla, function(i) {
-  casebase::fitSmoothHazard(i, data = ERSPC, ratio = 100)
-})
-```
 
@@
 
-```{r compPop, fig.cap="Population-time plot for the stem-cell transplant study.
-A) The red points represent relapse events.
-B) The blue points represent the competing events.", echo=FALSE, eval = eval_cs2}
+```{r compPop, fig.cap="Population-time plot for the stem-cell transplant study with both relapse and competing events.", echo=FALSE, eval = eval_cs2}
 
@@
 popTimeData <- popTime(data = bmtcrr, time = "ftime")
+
 plot(popTimeData, 
-     add.case.series = TRUE,
+     add.competing.event = TRUE,
+     comprisk = TRUE,
      # add.base.series = TRUE,
-     add.competing.event = TRUE,
-     comprisk = TRUE)
+     # ratio = 0.7,
+     # facet.params = list(ncol = 2),
+     ribbon.params = list(fill = "gray50"),
+     case.params = list(size = 0.8),
+     competing.params = list(size = 0.8)) + 
+  theme_bw(base_size = 10) + 
+  theme(legend.position = "bottom") + 
+  scale_y_continuous(labels = scales::label_number_si())
+
+# plot(popTimeData,
+#      add.case.series = TRUE,
+#      # add.base.series = TRUE,
+#      add.competing.event = TRUE,
+#      comprisk = TRUE)
 
@@
   ylim(c(0, 1)) +
-  theme_minimal() +
-  theme(legend.position = "top") +
+  theme_bw(base_size = 10) + 
+  theme(legend.position = "bottom") +
   xlab("Time (in Months)") +
@@
   labs(y = "Probability of Death", x = "Survival time (Years)", color = "Models") +
-  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M"))
+  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + theme_bw(base_size = 10) + theme(legend.position = "bottom")
 ```
@@
   # ggtitle("Brier scores over time") +
-  scale_color_manual(values = q4[1:3])
+  scale_color_manual(values = q4[1:3]) + theme_bw(base_size = 10)
 
@@
 
-cowplot::plot_grid(plot_row, legend, rel_widths = c(3, .4))
+cowplot::plot_grid(plot_row, legend, rel_heights = c(3, .4), nrow = 2)
 ```
@@
   ) +
+  theme_bw(base_size = 10) + 
+  theme(legend.position = "bottom") + 
   # ggtitle("Coefficient plot") +
@@
   # ggtitle("Absolute risk curves") +
-  scale_color_manual(values = q4, breaks = c("Cox", "Penalized Cox", "Penalized logistic", "K-M"))
+  scale_color_manual(values = q4, breaks = c("Cox", "Penalized Cox", "Penalized logistic", "K-M")) + 
+  theme_bw(base_size = 10) + 
+  theme(legend.position = "bottom")
 ```
@@
   labs(color = "Models") +
+  theme_bw(base_size = 10) + 
+  theme(legend.position = "bottom")+
   scale_color_manual(values = q4)
@@
 
-cowplot::plot_grid(plot_row_reg, legend_reg, rel_widths = c(3, .4))
+cowplot::plot_grid(plot_row_reg, legend_reg, rel_heights = c(3, .4), nrow = 2)
 ```
@@
 )
-plot(stanford_popTime)
+plot(stanford_popTime,
+     ribbon.params = list(fill = "grey50")) + theme_bw(base_size = 10) + theme(legend.position = "bottom") 
 ```
@@
 )
+
 # Set the offset to zero
@@
 
-colnames(absrisk) <- c("Time", "NoTrans", "Trans")
-
-# Rearrange the data
-absrisk <- gather(
-  as.data.frame(absrisk),
-  "Status", "Risk", -Time
-)
-
-ggplot(absrisk, aes(Time, Risk, colour = Status)) +
-  geom_line() +
-  theme_minimal() +
-  theme(legend.position = "top") +
-  expand_limits(y = 0.5) +
-  xlab("Follow-up time") +
-  ylab("Cum. Incidence")
+class(absrisk)
+
+plot(absrisk,
+     id.names = c("NoTrans","Trans")) + theme_bw(base_size = 10) + theme(legend.position = "bottom")
+
+# colnames(absrisk) <- c("Time", "NoTrans", "Trans")
+# 
+# # Rearrange the data
+# absrisk <- gather(
+#   as.data.frame(absrisk),
+#   "Status", "Risk", -Time
+# )
+# 
+# ggplot(absrisk, aes(Time, Risk, colour = Status)) +
+#   geom_line() +
+#   theme_minimal() +
+#   theme(legend.position = "top") +
+#   expand_limits(y = 0.5) +
+#   xlab("Follow-up time") +
+#   ylab("Cum. Incidence")
 ```
@@
 As we can see, the risk estimate at 1-year is about 30% lower if the patient receives a heart transplant at 30 days.
+
+We can also plot the Hazard ratio:
+```{r}
+newtime <- quantile(fit4[["originalData"]][[fit4[["timeVar"]]]],
+                    probs = seq(0.01, 0.99, 0.01))
+
+# reference category
+newdata <- data.frame(exposure = 0,
+                      futime = newtime)
+
+plot(fit4,
+     type = "hr",
+     newdata = newdata,
+     var = "exposure",
+     increment = 1,
+     xvar = "futime",
+     ci = F,
+     rug = T)
+```
 

### 2020-07-18 analysis/paper/paper.tex
fixed comparison table. and consistent themes or figures

@@
 Moreover, \pkg{flexsurv} has the ability to estimate both scale and
-shape parameters for a variety of parametric families (see Table
-\ref{tab:dists}). As discussed above, \pkg{casebase} can model any
-parametric family whose log-hazard can be expressed as a linear
-combination of covariates (including time). Therefore, our package
-allows the user to model the effect of time using splines. Also, by
-including interaction terms between covariates and time, it also allows
-users to fit time-varying coefficient models. However, we do not
-explicitly model any shape parameter, unlike \pkg{flexsurv}.
+shape parameters for a variety of parametric families. As discussed
+above, \pkg{casebase} can model any parametric family whose log-hazard
+can be expressed as a linear combination of covariates (including time).
+Therefore, our package allows the user to model the effect of time using
+splines. Also, by including interaction terms between covariates and
+time, it also allows users to fit time-varying coefficient models.
+However, we do not explicitly model any shape parameter, unlike
+\pkg{flexsurv}.
 
@@
 \begin{CodeChunk}
-
-\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-1} \end{center}
-
+\begin{figure}
+
+{\centering \includegraphics{../figures/plot-stratified-erspc-data-1}
+
+}
+
+\caption[Population time plot for ERSPC data]{Population time plot for ERSPC data.}\label{fig:plot-stratified-erspc-data}
+\end{figure}
 \end{CodeChunk}
@@
 
+\begin{CodeChunk}
+
+\begin{CodeOutput}
+#> 
+#> Call:
+#> glm(formula = formula, family = binomial, data = sampleData)
+#> 
+#> Deviance Residuals: 
+#>    Min      1Q  Median      3Q     Max  
+#> -0.378  -0.162  -0.122  -0.094   3.468  
+#> 
+#> Coefficients:
+#>                       Estimate Std. Error z value            Pr(>|z|)    
+#> (Intercept)            -9.0335     0.1122  -80.54 <0.0000000000000002 ***
+#> Follow.Up.Time          0.2228     0.0145   15.32 <0.0000000000000002 ***
+#> ScrArmScreening group  -0.2340     0.0886   -2.64              0.0083 ** 
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#> (Dispersion parameter for binomial family taken to be 1)
+#> 
+#>     Null deviance: 6059.0  on 54539  degrees of freedom
+#> Residual deviance: 5810.1  on 54537  degrees of freedom
+#> AIC: 5816
+#> 
+#> Number of Fisher Scoring iterations: 8
+\end{CodeOutput}
+\end{CodeChunk}
+
 The \code{absoluteRisk} function takes as input the
@@
 #> 1     54538       6053                                    
-#> 2     54535       5788  3      265 <0.0000000000000002 ***
+#> 2     54535       5787  3      266 <0.0000000000000002 ***
 #> ---
@@
 #>    Exp.   Gomp.   Weib. Splines 
-#>    6057    5818    6057    5798
+#>    6057    5816    6057    5797
 \end{CodeOutput}
@@
 \cmidrule{1-3}
-Weibull & 0.81 (0.68, 0.97) & 0.80 (0.65, 0.96)\\
+Weibull & 0.81 (0.68, 0.96) & 0.80 (0.65, 0.96)\\
 \cmidrule{1-3}
-Splines & 0.80 (0.67, 0.95) & --\\
+Splines & 0.81 (0.68, 0.96) & --\\
 \bottomrule
@@
 
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/compPop-1}
+
+}
+
+\caption[Population-time plot for the stem-cell transplant study with both relapse and competing events]{Population-time plot for the stem-cell transplant study with both relapse and competing events.}\label{fig:compPop}
+\end{figure}
+\end{CodeChunk}
+
 Our main objective is to compute the absolute risk of relapse for a
@@
 R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-R>   data = bmtcrr
-R> )
+R+   data = bmtcrr
+R+ )
 \end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{table}
+
+\caption{\label{tab:bmtcrr-cis}Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.}
+\centering
+\begin{tabular}[t]{lrlrl}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
+\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
+Covariates & HR & 95\% CI & HR & 95\% CI\\
+\midrule
+Sex & 0.70 & (0.4, 1.23) & 0.68 & (0.39, 1.19)\\
+Disease & 0.53 & (0.29, 0.95) & 0.52 & (0.29, 0.93)\\
+Phase (CR2 vs. CR1) & 1.12 & (0.45, 2.78) & 1.21 & (0.48, 3.02)\\
+Phase (CR3 vs. CR1) & 1.45 & (0.37, 5.66) & 1.67 & (0.43, 6.5)\\
+Phase (Relapse vs. CR1) & 4.04 & (1.88, 8.68) & 4.55 & (2.09, 9.9)\\
+\addlinespace
+Source & 1.48 & (0.48, 4.54) & 1.46 & (0.47, 4.54)\\
+Age & 0.99 & (0.97, 1.01) & 0.99 & (0.97, 1.02)\\
+\bottomrule
+\end{tabular}
+\end{table}
+
 \end{CodeChunk}
@@
 R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) +
-R>                         const(Phase) + const(Source) + const(Age),
-R>                       data = bmtcrr, cause = 1, model = "fg")
+R+                         const(Phase) + const(Source) + const(Age),
+R+                       data = bmtcrr, cause = 1, model = "fg")
 R> 
@@
 
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/brierpt1-1}
+
+}
+
+\caption{\label{fig:brier1} Comparison of three Cox regression models. (A) Cumulative Incidence curves. (B) Time-dependent Brier scores.}\label{fig:brierpt1}
+\end{figure}
+\end{CodeChunk}
+
 For our penalized logistic regression model, we opt for the natural log
@@
 R> x <- model.matrix(death ~ . - d.time - aps - sps, 
-R>                   data = train)[, -c(1)] # Remove intercept
+R+                   data = train)[, -c(1)] # Remove intercept
 R> y <- data.matrix(subset(train, select = c(d.time, death)))
@@
 R> cbFull <- casebase::fitSmoothHazard.fit(x, y,
-R>   family = "glmnet",
-R>   time = "d.time", event = "death",
-R>   formula_time = ~ log(d.time), alpha = 1,
-R>   ratio = 10, standardize = TRUE,
-R>   penalty.factor = c(0, rep(1, ncol(x)))
-R> )
+R+   family = "glmnet",
+R+   time = "d.time", event = "death",
+R+   formula_time = ~ log(d.time), alpha = 1,
+R+   ratio = 10, standardize = TRUE,
+R+   penalty.factor = c(0, rep(1, ncol(x)))
+R+ )
 \end{CodeInput}
@@
 R> newx <- model.matrix(death ~ . - d.time - aps - sps, 
-R>                   data = test)[, -c(1)]
+R+                   data = test)[, -c(1)]
 R> times <- sort(unique(test$d.time))
@@
 R> abCbFull <- casebase::absoluteRisk(cbFull,
-R>   time = times,
-R>   newdata = newx,
-R>   s = "lambda.1se",
-R>   method = "numerical"
-R> )
+R+   time = times,
+R+   newdata = newx,
+R+   s = "lambda.1se",
+R+   method = "numerical"
+R+ )
 \end{CodeInput}
@@
 (cano).
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/coefplots-1}
+
+}
+
+\caption{\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models.
+If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity.}\label{fig:coefplots}
+\end{figure}
+\end{CodeChunk}
 
@@
 three models outperform the Kaplan-Meier model.
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{../figures/riskregressionBrier-1}
+
+}
+
+\caption{\label{fig:cs3FinalBrier} Four curves comparing the Cumulative Incidence (A) and the Brier scores (B) as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.}\label{fig:riskregressionBrier}
+\end{figure}
+\end{CodeChunk}
 
@@
 
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{../figures/stanford-poptime-1} \end{center}
+
+\end{CodeChunk}
+
 Since the exposure is time-dependent, we need to manually define the
@@
 Akaike Information Criterion (AIC).
+
+\begin{CodeChunk}
+
+\begin{CodeOutput}
+#> Model1 Model2 Model3 Model4 
+#>    845    491    495    486
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 R> )
+R> 
 R> # Set the offset to zero
@@
 
+\begin{CodeChunk}
+
+\begin{CodeOutput}
+#> [1] "absRiskCB" "matrix"
+\end{CodeOutput}
+
+\begin{center}\includegraphics{../figures/stanford-risk-1} \end{center}
+
+\end{CodeChunk}
+
 Note that we can easily adapt the code above to the situation where a
@@
 
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{../figures/stanford-1y-haz-1} \end{center}
+
+\end{CodeChunk}
+
 We can then compare the 1-year mortality risk without transplant and
@@
 
+\begin{CodeChunk}
+
+\begin{CodeOutput}
+#>  time  
+#>     0 0
+#>   365 1
+\end{CodeOutput}
+
+\begin{CodeOutput}
+#> [1] 0.99
+\end{CodeOutput}
+\end{CodeChunk}
+
 As we can see, the risk estimate at 1-year is about 30\% lower if the
 patient receives a heart transplant at 30 days.
+
+We can also plot the Hazard ratio:
+
+\begin{CodeChunk}
+
+\begin{center}\includegraphics{../figures/unnamed-chunk-12-1} \end{center}
+
+\end{CodeChunk}
 
@@
 
-This report was generated on 2020-07-18 13:50:44 using the following
+This report was generated on 2020-07-18 15:04:50 using the following
 computational environment and dependencies:
@@
 #> other attached packages:
-#>  [1] forcats_0.5.0             stringr_1.4.0            
-#>  [3] dplyr_0.8.5               purrr_0.3.4              
-#>  [5] readr_1.3.1               tidyr_1.1.0              
-#>  [7] ggplot2_3.3.2.9000        tidyverse_1.3.0          
-#>  [9] lubridate_1.7.4           visreg_2.7.0             
-#> [11] tibble_3.0.1              riskRegression_2020.02.05
-#> [13] prodlim_2019.11.13        flexsurv_1.1.1           
-#> [15] survival_3.1-8            pracma_2.2.9             
-#> [17] magrittr_1.5              kableExtra_1.1.0         
-#> [19] glmnet_4.0-2              Matrix_1.2-18            
-#> [21] data.table_1.12.8         cowplot_1.0.0            
-#> [23] colorspace_1.4-1          casebase_0.9.0           
+#>  [1] broom_0.7.0               dotwhisker_0.5.0         
+#>  [3] timereg_1.9.4             glue_1.4.1               
+#>  [5] forcats_0.5.0             stringr_1.4.0            
+#>  [7] dplyr_0.8.5               purrr_0.3.4              
+#>  [9] readr_1.3.1               tidyr_1.1.0              
+#> [11] ggplot2_3.3.2.9000        tidyverse_1.3.0          
+#> [13] lubridate_1.7.4           visreg_2.7.0             
+#> [15] tibble_3.0.1              riskRegression_2020.02.05
+#> [17] prodlim_2019.11.13        flexsurv_1.1.1           
+#> [19] survival_3.1-8            pracma_2.2.9             
+#> [21] magrittr_1.5              kableExtra_1.1.0         
+#> [23] glmnet_4.0-2              Matrix_1.2-18            
+#> [25] data.table_1.12.8         cowplot_1.0.0            
+#> [27] colorspace_1.4-1          casebase_0.9.0           
 #> 
@@
 #>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
-#>  [4] htmlTable_1.13.3    base64enc_0.1-3     fs_1.3.2           
-#>  [7] rstudioapi_0.11     farver_2.0.3        MatrixModels_0.4-1 
-#> [10] fansi_0.4.1         mvtnorm_1.0-12      xml2_1.3.0         
-#> [13] codetools_0.2-16    knitr_1.29          Formula_1.2-3      
-#> [16] jsonlite_1.7.0      broom_0.7.0         cluster_2.1.0      
+#>  [4] ggstance_0.3.3      htmlTable_1.13.3    base64enc_0.1-3    
+#>  [7] fs_1.3.2            rstudioapi_0.11     farver_2.0.3       
+#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.0-12     
+#> [13] xml2_1.3.0          codetools_0.2-16    knitr_1.29         
+#> [16] Formula_1.2-3       jsonlite_1.7.0      cluster_2.1.0      
 #> [19] dbplyr_1.4.2        png_0.1-7           compiler_3.6.2     
@@
 #> [28] quantreg_5.54       tools_3.6.2         gtable_0.3.0       
-#> [31] glue_1.4.1          Rcpp_1.0.4.6        cellranger_1.1.0   
-#> [34] vctrs_0.3.1         nlme_3.1-143        iterators_1.0.12   
-#> [37] xfun_0.15           rvest_0.3.5         lifecycle_0.2.0    
-#> [40] polspline_1.1.17    muhaz_1.2.6.1       MASS_7.3-51.5      
-#> [43] zoo_1.8-7           scales_1.1.1        hms_0.5.3          
-#> [46] sandwich_2.5-1      SparseM_1.78        RColorBrewer_1.1-2 
-#> [49] rticles_0.13        yaml_2.2.1          gridExtra_2.3      
-#> [52] rms_5.1-4           rpart_4.1-15        latticeExtra_0.6-29
-#> [55] stringi_1.4.6       foreach_1.5.0       checkmate_2.0.0    
-#> [58] lava_1.6.6          shape_1.4.4         rlang_0.4.6        
-#> [61] pkgconfig_2.0.3     evaluate_0.14       lattice_0.20-38    
-#> [64] labeling_0.3        htmlwidgets_1.5.1   cmprsk_2.2-9       
-#> [67] tidyselect_1.1.0    deSolve_1.27.1      R6_2.4.1           
+#> [31] Rcpp_1.0.4.6        cellranger_1.1.0    vctrs_0.3.1        
+#> [34] nlme_3.1-143        iterators_1.0.12    xfun_0.15          
+#> [37] rvest_0.3.5         lifecycle_0.2.0     polspline_1.1.17   
+#> [40] muhaz_1.2.6.1       MASS_7.3-51.5       zoo_1.8-7          
+#> [43] scales_1.1.1        hms_0.5.3           sandwich_2.5-1     
+#> [46] SparseM_1.78        RColorBrewer_1.1-2  rticles_0.13       
+#> [49] yaml_2.2.1          gridExtra_2.3       rms_5.1-4          
+#> [52] rpart_4.1-15        latticeExtra_0.6-29 stringi_1.4.6      
+#> [55] foreach_1.5.0       checkmate_2.0.0     lava_1.6.6         
+#> [58] shape_1.4.4         rlang_0.4.6         pkgconfig_2.0.3    
+#> [61] evaluate_0.14       lattice_0.20-38     labeling_0.3       
+#> [64] htmlwidgets_1.5.1   cmprsk_2.2-9        tidyselect_1.1.0   
+#> [67] deSolve_1.27.1      plyr_1.8.6          R6_2.4.1           
 #> [70] generics_0.0.2      Hmisc_4.3-1         multcomp_1.4-12    
@@
 #> [79] nnet_7.3-12         mstate_0.2.12       modelr_0.1.5       
-#> [82] crayon_1.3.4        rmarkdown_2.3       timereg_1.9.4      
-#> [85] jpeg_0.1-8.1        grid_3.6.2          readxl_1.3.1       
-#> [88] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
-#> [91] numDeriv_2016.8-1.1 stats4_3.6.2        munsell_0.5.0      
-#> [94] viridisLite_0.3.0   quadprog_1.5-8
+#> [82] crayon_1.3.4        rmarkdown_2.3       jpeg_0.1-8.1       
+#> [85] grid_3.6.2          readxl_1.3.1        reprex_0.3.0       
+#> [88] digest_0.6.25       webshot_0.5.2       numDeriv_2016.8-1.1
+#> [91] stats4_3.6.2        munsell_0.5.0       viridisLite_0.3.0  
+#> [94] quadprog_1.5-8
 \end{CodeOutput}
@@
 #> Remote:   write-popTime @ origin (https://github.com/sahirbhatnagar/cbpaper.git)
-#> Head:     [84d5ff5] 2020-06-29: merge pop-stash
+#> Head:     [dad6070] 2020-07-18: cleaned cs1
 \end{CodeOutput}

### 2020-07-19 analysis/paper/paper.Rmd
to change ggtheme, simply change paper_gg_theme object defined in first chunck
then all the figures will update to that theme

@@
 echo_plot_code <- TRUE
+
+paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
 ```
@@
      case.params = list(size = 0.8),
-     base.params = list(size = 0.8)) + theme_bw(base_size = 10) + theme(legend.position = "bottom") + 
+     base.params = list(size = 0.8)) + 
+  paper_gg_theme +  
   scale_y_continuous(labels = scales::label_number_si())
@@
      competing.params = list(size = 0.8)) + 
-  theme_bw(base_size = 10) + 
-  theme(legend.position = "bottom") + 
+  paper_gg_theme +  
   scale_y_continuous(labels = scales::label_number_si())
@@
   ylim(c(0, 1)) +
-  theme_bw(base_size = 10) + 
-  theme(legend.position = "bottom") +
+  paper_gg_theme +  
   xlab("Time (in Months)") +
@@
   labs(y = "Probability of Death", x = "Survival time (Years)", color = "Models") +
-  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + theme_bw(base_size = 10) + theme(legend.position = "bottom")
+  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
+  paper_gg_theme
 ```
@@
   # ggtitle("Brier scores over time") +
-  scale_color_manual(values = q4[1:3]) + theme_bw(base_size = 10)
+  scale_color_manual(values = q4[1:3]) + 
+  paper_gg_theme
 
@@
   ) +
-  theme_bw(base_size = 10) + 
-  theme(legend.position = "bottom") + 
+  paper_gg_theme +  
   # ggtitle("Coefficient plot") +
@@
   scale_color_manual(values = q4, breaks = c("Cox", "Penalized Cox", "Penalized logistic", "K-M")) + 
-  theme_bw(base_size = 10) + 
-  theme(legend.position = "bottom")
+  paper_gg_theme
 ```
@@
   labs(color = "Models") +
-  theme_bw(base_size = 10) + 
-  theme(legend.position = "bottom")+
+  paper_gg_theme +  
   scale_color_manual(values = q4)
@@
 plot(stanford_popTime,
-     ribbon.params = list(fill = "grey50")) + theme_bw(base_size = 10) + theme(legend.position = "bottom") 
+     ribbon.params = list(fill = "grey50")) + 
+  paper_gg_theme
 ```
@@
   geom_line() +
-  theme_minimal() +
-  theme(legend.position = "top") +
+  paper_gg_theme +  
   ylab("Hazard") +
@@
 plot(absrisk,
-     id.names = c("NoTrans","Trans")) + theme_bw(base_size = 10) + theme(legend.position = "bottom")
+     id.names = c("NoTrans","Trans")) + 
+  paper_gg_theme
 

### 2020-07-19 analysis/paper/paper.Rmd
remove scales:: from paper

@@
 library(cowplot)
+library(scales)
 library(data.table)
@@
   paper_gg_theme +  
-  scale_y_continuous(labels = scales::label_number_si())
+  scale_y_continuous(labels = label_number_si())
 ```
@@
   paper_gg_theme +  
-  scale_y_continuous(labels = scales::label_number_si())
+  scale_y_continuous(labels = label_number_si())
 

### 2020-07-19 analysis/paper/paper.Rmd
trying to get scales to work

@@
 library(cowplot)
-library(scales)
 library(data.table)
@@
 library(tidyverse)
+library(scales)
 options(digits = 2, scipen = 999)

### 2020-07-20 analysis/paper/paper.Rmd

@@
 ```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plot for ERSPC data."}
+
+label_number_si <- function (accuracy = 1, unit = NULL, sep = NULL, ...) {
+    sep <- if (is.null(unit)) 
+        ""
+    else " "
+    scales:::force_all(accuracy, ...)
+    function(x) {
+        breaks <- c(0, 10^c(K = 3, M = 6, B = 9, T = 12))
+        n_suffix <- cut(abs(x), breaks = c(unname(breaks), Inf), 
+            labels = c(names(breaks)), right = FALSE)
+        n_suffix[is.na(n_suffix)] <- ""
+        suffix <- paste0(sep, n_suffix, unit)
+        scale <- 1/breaks[n_suffix]
+        scale[which(scale %in% c(Inf, NA))] <- 1
+        number(x, accuracy = accuracy, scale = unname(scale), 
+            suffix = suffix, ...)
+    }
+}
 plot(pt_object, 

### 2020-07-20 analysis/paper/paper.tex

@@
 R>   geom_line() +
-R>   theme_minimal() +
-R>   theme(legend.position = "top") +
+R>   paper_gg_theme +  
 R>   ylab("Hazard") +
@@
 
-This report was generated on 2020-07-18 15:04:50 using the following
+This report was generated on 2020-07-19 18:02:45 using the following
 computational environment and dependencies:

### 2020-07-22 analysis/paper/paper.Rmd
review new sections

--- 2020-07-19 analysis/paper/paper.Rmd
@@
 author:
-  - name: Sahir Rai Bhatnagar *
+  - name: Sahir Rai Bhatnagar*
     affiliation: McGill University
@@
     url: http://sahirbhatnagar.com/
-  - name: Maxime Turgeon *
+  - name: Maxime Turgeon*
     affiliation: 'University of Manitoba \AND'
@@
 abstract: >
-  The abstract of the article. * joint co-authors
+  The abstract of the article. *SRB and MT contributed equally to this work.
 keywords:
   # at least one keyword must be supplied
-  formatted: [keywords, not capitalized, "\\proglang{Java}"]
-  plain:     [keywords, not capitalized, Java]
+  formatted: [survival analysis, absolute risk, data visualization]
+  plain:     [survival analysis, absolute risk, data visualization]
 preamble: >
@@
 Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function.
-In addition, we provide additional features, including performance metrics and visualization tools.
-
-In what follows, we first recall some theoretical details on case-base sampling and its use for estimating parameteric hazard functions.
+In addition, we provide extensive visualization tools.
+
+In what follows, we first recall some theoretical details on case-base sampling and its use for estimating parametric hazard functions.
 We then give a short review of existing \proglang{R} packages that implement comparable features as \pkg{casebase}.
@@
 We assume a continuous time model, which implies that the counting process jumps are less than or equal to one.
-We are interested in modelling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
+We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
 $$\lambda_{i}(t) dt = E[dN_{i}(t)].$$
@@
 $$L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^\tau \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)}\left(1 - \lambda_{i}(t;\theta)\right)^{1 - dN_{i}(t)},$$
-where $\prod_{t\in[0,u)}$ represents a product integral from $0$ to $u$. However, the integral over time makes the computation and maximisation of $L_0(\theta)$ challenging.
+where $\prod_{t\in[0,u)}$ represents a product integral from $0$ to $u$. However, the integral over time makes the computation and maximization of $L_0(\theta)$ challenging.
 
@@
 $$L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.$$
-When a logarithmic link function is used for modelling the hazard function, the above expression is of a logistic regression form with an offset term $\log(1/\rho_i(t))$.
+When a logarithmic link function is used for modeling the hazard function, the above expression is of a logistic regression form with an offset term $\log(1/\rho_i(t))$.
 Note that the sampling units selected in the case-base sampling mechanism are person-moments, rather than individuals, and the parameters to be estimated are hazards or hazard ratios rather than odds or odds ratios.
@@
 The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard.
-The first category broadly corresponds to parametric survival models, where the full hazard is explictly modeled.
+The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled.
 Using this estimate, the survival function and the CIF can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below).
@@
 As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category often produces estimates in the form of step-wise functions.
-Providing smooth estimates of the CIF was one of the main motivations for introducting case-base sampling in survival analysis.
+Providing smooth estimates of the CIF was one of the main motivations for introducing case-base sampling in survival analysis.
 
@@
 
-The functions in the casebase package can be divided into two categories: 1) exploratory data analysis, in the form of population-time plots; and 2) parametric modeling of the hazard function.
-We explicitly aimed at being compatible with both \code{data.frame}s and \code{data.table}s.
-This is evident in some of the coding choices we made, and it is also reflected in our unit tests.
+The functions in the \pkg{casebase} package can be divided into two categories: 1) exploratory data analysis, in the form of population-time plots; and 2) parametric modeling of the hazard function.
+We strove for compatibility with both \code{data.frame}s and \code{data.table}s; this can be seen in the coding choices we made and the unit tests we wrote.
 
@@
 The case-base sampling approach described in Section \ref{theory} can be visualized in the form of a population time plot.
-They are extremely informative graphical displays of survival data and should be one of the first steps in an exploratory data analysis.
-We facilitate this task in the \pkg{casebase} package using the \code{popTime} function and \code{plot} method as follows:
+These plots are extremely informative graphical displays of survival data and should be one of the first steps in an exploratory data analysis.
+The \code{popTime} function and \code{plot} method facilitate this task:
 
@@
 
-The splitting of these tasks gives the user the flexibility to create population-time plots with the graphics system of their choice using the object created in Step 1, while the object created in Step 2 allows one to leverage the \pkg{ggplot2} [@ggplot2] family of functions for further customization.
-
-To illustrate these functions we will use the European Randomized Study of Prostate Cancer Screening (ERSPC) data [@schroder2009screening] which was extracted using the approach described in [@liu2014recovering] and available through the \pkg{casebase} package.
-This data contains the individual observations for 159,893 men between the ages of 55 and 69 years at entry who were recruited from seven European countries.
-
-We first create the necessary dataset for producing the population time plot using the \code{casebase::popTime} function.
-In this example we stratify by the treatment group.
-The resulting object inherits from class `popTime` and stores the exposure variable as an attribute:
+By splitting these tasks, we give flexibility to the user.
+While the method call in Step 2 allows further customization by using the \pkg{ggplot2} [@ggplot2] family of functions, users can use for the graphics system of their choice to create population-time plots from the object created in Step 1.
+
+To illustrate these functions, we will use the European Randomized Study of Prostate Cancer Screening (ERSPC) data [@schroder2009screening] which was extracted using the approach described in Liu *et al.* [-@liu2014recovering] This dataset is available through the \pkg{casebase} package.
+It contains the individual observations for 159,893 men  from seven European countries, who were between the ages of 55 and 69 years when recruited for the trial.
+
+We first create the necessary dataset for producing the population time plot using the \code{popTime} function.
+In this example, we stratify the plot by treatment group.
+The resulting object inherits from class \code{popTime} and stores the exposure variable as an attribute:
 
@@
 
-We then pass this object to the corresponding \code{plot} method which leverages the \pkg{ggplot2} package to build population time plots. 
+We then pass this object to the corresponding \code{plot} method:
+
 ```{r plot-stratified-erspc-data-code, eval = FALSE, echo = TRUE}
@@
 ```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plot for ERSPC data."}
+label_number_si <- function (accuracy = 1, unit = NULL, sep = NULL, ...) {
+    sep <- if (is.null(unit)) 
+        ""
+    else " "
+    scales:::force_all(accuracy, ...)
+    function(x) {
+        breaks <- c(0, 10^c(K = 3, M = 6, B = 9, T = 12))
+        n_suffix <- cut(abs(x), breaks = c(unname(breaks), Inf), 
+            labels = c(names(breaks)), right = FALSE)
+        n_suffix[is.na(n_suffix)] <- ""
+        suffix <- paste0(sep, n_suffix, unit)
+        scale <- 1/breaks[n_suffix]
+        scale[which(scale %in% c(Inf, NA))] <- 1
+        number(x, accuracy = accuracy, scale = unname(scale), 
+            suffix = suffix, ...)
+    }
+}
 plot(pt_object, 
@@
 Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by first adding a layer for the area representing the population time in gray, with subjects having the least amount of observation time plotted at the top of the y-axis.
-We immediately notice a unique _step shape_ in the population time area which is due to the randomization of the Finnish cohorts which were carried out on January 1 of each of the 4 years 1996 to 1999.
-This, coupled with the uniform December 31 2006 censoring date, lead to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up.
+We immediately notice a distinctive *step-wise shape* in the population time area.
+This is due to the randomization of the Finnish cohorts which were carried out on January 1 of each of year from 1996 to 1999.
+Coupled with the uniform December 31 2006 censoring date, this led to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up.
 Tracked backwards in time (i.e.
-from right to left), the population-time plot shows the recruitment pattern from its beginning in 1991, and in particular the January 1 entries in successive years.
+from right to left), the population-time plot shows the recruitment pattern from its beginning in 1991, and the January 1 entries in successive years.
 Tracked forwards in time (i.e.
 from left to right), the plot for the first three years shows attrition due entirely to death (mainly from other causes).
-Since the Swedish and Belgian centres were the last to close their recruitment - in December 2003 - the minimum potential follow-up is three years.
+Since the Swedish and Belgian centres were the last to complete recruitment--in December 2003--the minimum potential follow-up is three years.
 Tracked further forwards in time (i.e.
 after year 3) the attrition is a combination of deaths and staggered entries.
+As we can see, population-time plots summarise a wealth of information about the study into a simple graph.
 
 Next, layers for the case series and base series are added.
-The case series is sampled at random vertically on the plot to avoid having all points along the upper edge of the plot, while the base series is sampled horizontally with sampling weight proportional to their follow-up time.
-By randomly distributing the points, we can get a better sense of the incidence density.
-In the ERSPC data, we see that more events are observed later on in time.
-Therefore a constant hazard model would not be appropriate in this instance as it would overestimate the cumulative incidence earlier on in time, and underestimate it later on.
-The reader is referred to the package vignettes for more examples and a detailed description of how to modify the aesthetics of a population time plot.
+The case series is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area.
+By randomly distributing the cases, we can get a better sense of the incidence density.
+In Figure \ref{fig:plot-stratified-erspc-data}, we see that more events are observed at later follow-up times.
+Therefore, a constant hazard model would not be appropriate in this instance as it would overestimate the cumulative incidence earlier on in time, and underestimate it later on.
+Finally, the base series is sampled horizontally with sampling weight proportional to their follow-up time.
+The reader should refer to the package vignettes for more examples and a detailed description of how to modify the aesthetics of a population-time plot.
 
@@
   - \code{glm}: This is the familiar logistic regression.
-  - \code{glmnet}: This option allows for variable selection using Lasso or elastic-net (cf. Case Study 3). This functionality is provided through the \code{glmnet} package [@friedman2010jss].
-  - \code{gam}: This option provides support for *Generalized Additive Models* via the \code{gam} package [@hastie1987generalized].
-  - \code{gbm}: This option provides support for *Gradient Boosted Trees* via the \code{gbm} package. This feature is still experimental.
+  - \code{glmnet}: This option allows for variable selection using Lasso or elastic-net (cf. Case Study 3). This functionality is provided through the \pkg{glmnet} package [@friedman2010jss].
+  - \code{gam}: This option provides support for *Generalized Additive Models* via the \pkg{mgcv} package [@hastie1987generalized].
+  - \code{gbm}: This option provides support for *Gradient Boosted Trees* via the \pkg{gbm} package. This feature is still experimental.
 
@@
 
-We investigate the differences in survival between the control and screening arms in the ERSPC study presented earlier using casebase sampling to fit a parametric hazard function.
-We fit four models that differ in the functional form of time; (1) excluded from the linear predictor (Eq.\eqref{eq:exp}), (2) linear function (Eq.\eqref{eq:gomp}), (3) log function (Eq.\eqref{eq:weibull}), and (4) a smooth function using cubic B-splines.
+For our first case study, we return to the ERSPC study and investigate the differences in survival between the control and screening arms.
+We fit four models that differ in which functional form of time is used: 1) excluded from the linear predictor, 2) linear function, 3) log function, and 4) a smooth function using cubic B-splines.
 The models are fit using \code{fitSmoothHazard}.
-Since the output object from \code{fitSmoothHazard} inherits from the \code{glm} class, we see a familiar result when using the function \code{summary}:
+Since the output object from \code{fitSmoothHazard} inherits from the \code{glm} class, we can directly use the function \code{summary}:
 
@@
 
-summary(fits[[2]])
+summary(fits[["gompertz"]])
 ```
@@
 ```{r, eval = eval_cs1, echo=FALSE}
-summary(fits[[2]])
-```
-
-The \code{absoluteRisk} function takes as input the \code{fitSmoothHazard} object and returns a matrix where each column corresponds to the covariate profiles specified in the \code{newdata} argument, and each row corresponds to a specified time point:
+summary(fits[["gompertz"]])
+```
+
+Next, the \code{absoluteRisk} function takes as input the \code{fitSmoothHazard} object and returns a matrix where each column corresponds to the covariate profiles specified in the \code{newdata} argument, and each row corresponds to a specified time point:
 
@@
 
-head(risks[[2]])
-```
-
-In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from casebase on the Cox model CIF.
+head(risks[["gompertz"]])
+```
+
+In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from \pkg{casebase} on the Cox model CIF.
 The CIF estimates for the exponential model in panel (1) overestimate the cumulative incidence earlier on in time, and underestimate it later on.
-Based on what we observed in the population time plot, where more events are observed later on in time, this poor fit for the exponential hazard is expected.
-This example demonstrates the benefits of population time plots as an exploratory analysis tool.
-
-In Figure \ref{fig:erspc-cox-cif} panels (2)-(4), we notice a better fit with increasing complexities of time.
-As noted above, the usual asymptotic results hold for likelihood ratio tests built using casebase sampling models.
+Based on our earlier discussion of the population-time plot, this poor fit for the exponential hazard was expected.
+
+In Figure \ref{fig:erspc-cox-cif} (panels 2--4), we notice a better fit with increasing complexity of our model for time.
+As noted above, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models.
 Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) splines model:
@@
 
-As expected, we see that splines model provides a better fit. Similarly, we can compare the AIC for all four casebase models:
+As expected, we see that splines model provides a better fit. Similarly, we can compare the AIC for all four case-base models:
 
@@
 
-```{r erspc-cox-cif, eval = eval_cs1, echo = FALSE, fig.width=8, fig.height=6, fig.cap="Cumulative incidence functions for control and screening groups in the ERSPC data.
-In each of the panels, we plot the CIF from the Cox model using \\code{survival::survfit} (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression.
-(1) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution.
-(2) Linear function of time which is equivalent to a Gompertz distribution.
-(3) The natural logarithm of time which is equivalent to a Weibull distribution.
+```{r erspc-cox-cif, eval = eval_cs1, echo = FALSE, fig.width=8, fig.height=6, fig.cap="CIFs for control and screening groups in the ERSPC data.
+In each of the panels, we plot the CIF from the Cox model using \\code{survival::survfit} (solid line) and the CIF from the case-base sampling scheme (dashed line) with different functional forms of time.
+(1) The time variable is excluded (exponential).
+(2) Linear function of time (Gompertz).
+(3) The natural logarithm (Weibull).
 (4) Cubic B-spline expansion of time."}
@@
 # lines(smooth_risk_exp[, 1], smooth_risk_exp[, 3], col = "blue", lty = 2)
-# 
 # 
@@
 
-      # The maximised log-likelihoods are the same between survreg and flexsurvreg, 
+      # The maximized log-likelihoods are the same between survreg and flexsurvreg, 
       # however the parameterisation is different: the first coefficient (Intercept) 
       # reported by survreg is log(mu), and survreg's "scale" is
-      # dweibull’s (thus flexsurvreg)'s 1 / shape
+      # dweibull's (thus flexsurvreg)'s 1 / shape
       res <- -fit[[model]]$res[covariate,] / (1 / fit[[model]]$res["shape",])
@@
 This reinforces the idea that, under proportional hazards, we do not need to model the full hazard to obtain reliable estimates of the hazard ratio.
-Of course, different parametric models for the hazards give rise to qualitatively different estimates for the CIF as shown in Figure \ref{fig:erspc-cox-cif}.
-
-```{r print-erspc-estimates, eval = TRUE}
+Nevertheless, Figure \ref{fig:erspc-cox-cif} shows that different parametric models can still give rise to qualitatively different estimates for the CIF.
+
+```{r print-erspc-estimates, eval = eval_cs1}
 ## ---- print-sim-table ----
 
-kable(ptable, "latex", booktabs = T, align = c("l","c","c"),
+kable(ptable, "latex", booktabs = TRUE, align = c("l","c","c"),
       caption = c("Comparison of estimated hazard ratios and 95\\% confidence intervals for ERSPC data."),
@@
   # add_header_above(c(" "," ","Null model" = 4, "1% Causal SNPs" = 4)) %>%
-  column_spec(1, bold=T) %>%
+  column_spec(1, bold = TRUE) %>%
   collapse_rows(columns = 1, latex_hline = "major", valign = "middle") %>%
@@
   #          alphabet_title = "Type II: ", symbol_title = "Type III: ",
-  #          footnote_as_chunk = T, title_format = c("italic", "underline")
+  #          footnote_as_chunk = TRUE, title_format = c("italic", "underline")
   # )
@@
 
-With this first case study, we saw how population-time plots can provide a useful summary of the incidence for our event of interests.
-In particular, we saw how it can provide evidence against an exponential hazard model.
-We then explored how different parametric models can provide different estimates of the CIF, even though they all give similar estimates for the hazard ratios.
+With this first case study, we explored how \pkg{casebase} allows us to fit different parametric survival models, and how we can compare the fit of each model with tools from GLMs.
 
@@
 For illustrative purposes, we will use the same data that was used in Scrucca *et al* [-@scrucca2010regression].
-The data was downloaded from the first author's website, and it is also available as part of the \pkg{casebase} package.
+The data was downloaded from the first author's website, and it is now available as part of the \pkg{casebase} package.
 
@@
 Several covariates were captured at baseline: sex, disease type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL and AML, respectively), disease phase at transplant (Relapse, CR1, CR2, CR3), source of stem cells (bone marrow and peripheral blood, coded as BM+PB, or only peripheral blood, coded as PB), and age.
-A summary of these baseline characteristics appear in Table \ref{tab:table1bmtcrr}.
-We note that the statistical summaries were generated differently for different variable types: for continuous variables, we gave the range, followed by the mean and standard deviation; for categorical variables, we gave the counts for each category.
-
-```{r echo = FALSE, eval = eval_cs2}
+<!--A summary of these baseline characteristics appear in Table \ref{tab:table1bmtcrr}.
+We note that the statistical summaries were generated differently for different variable types: for continuous variables, we gave the range, followed by the mean and standard deviation; for categorical variables, we gave the counts for each category.-->
+
+```{r echo = FALSE, eval = FALSE}
 tibble::tribble(
@@
 We perform this computation for a 35 year old woman who received a stem-cell transplant from peripheral blood at relapse.
-We compared the absolute risk curve for such a woman with acute lymphoblastic leukemia with that for a similar woman with acute myeloblastic leukemia.
+We compared the absolute risk curve for such a woman with ALL with that for a similar woman with AML.
 We will estimate the curve from 0 to 60 months.
@@
 $$CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),$$
-where $CI_k(t)$ is the cause-specific cumulative indicence.
+where $CI_k(t)$ is the cause-specific cumulative incidence.
 The Fine-Gray model allows to directly assess the effect of a covariate on the subdistribution, as opposed to the hazard.
@@
 Figure \ref{fig:compAbsrisk} shows the absolute risk curves for both case-base sampling and the Fine-Gray model.
-As we can see, the two approaches agree quite well for acute myeloblastic leukemia; however, there seems to be a difference of about 5% between the two curves for acute lymphoblastic leukemia.
+As we can see, the two approaches agree quite well for AML; however, there seems to be a difference of about 5% between the two curves for ALL.
 This difference does not appear to be significant: the curve from case-base sampling is contained within a 95% confidence band around the Fine-Gray absolute risk curve (figure not shown).
@@
 For fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface to the function \code{fitSmoothHazard}.
-In order to use this interfact, we first create a matrix \code{y} containing the time and event variables, and a matrix \code{x} containing all remaining covariates we would like to include in the model space.
+In order to use this interface, we first create a matrix \code{y} containing the time and event variables, and a matrix \code{x} containing all remaining covariates we would like to include in the model space.
 We used lasso penalization by setting \code{alpha = 1}.
@@
 ```{r stanford-poptime, eval = eval_cs4}
-colnames(jasa)[which(colnames(jasa)=="futime")] <- "futime_old"
-colnames(jasa)[which(colnames(jasa)=="age")] <- "futime"
-
 stanford_popTime <- popTime(jasa,
@@
 
-Note that the fourth model (i.e.
-`fit4`) includes an interaction term between exposure and follow-up time.
+Note that the fourth model includes an interaction term between exposure and follow-up time.
 In other words, this model no longer exhibit proportional hazards.
@@
 
-THIS NEXT CODE CHUNK IS NOT WORKING
-
-```{r stanford-hazard, eval = FALSE, echo=TRUE}
+```{r stanford-hazard, eval = eval_cs4, echo=TRUE}
 # Compute hazards---
@@
 
-We can also plot the Hazard ratio:
-```{r}
+We can also plot the hazard ratio:
+```{r eval = eval_cs4}
 newtime <- quantile(fit4[["originalData"]][[fit4[["timeVar"]]]],
@@
      xvar = "futime",
-     ci = F,
-     rug = T)
+     ci = FALSE,
+     rug = TRUE)
 ```

### 2020-07-22 analysis/paper/paper.tex
review new sections

@@
 \author{
-Sahir Rai Bhatnagar *\\McGill University \And Maxime Turgeon *\\University of Manitoba \AND Jesse Islam\\McGill University \And James Hanley\\McGill University \And Olli Saarela\\University of Toronto
+Sahir Rai Bhatnagar*\\McGill University \And Maxime Turgeon*\\University
+of Manitoba \AND Jesse Islam\\McGill University \And James
+Hanley\\McGill University \And Olli Saarela\\University of Toronto
 }
@@
 
-\Plainauthor{Sahir Rai Bhatnagar *, Maxime Turgeon *, Jesse Islam, James Hanley, Olli Saarela}
+\Plainauthor{Sahir Rai Bhatnagar*, Maxime Turgeon*, Jesse Islam, James
+Hanley, Olli Saarela}
 \Plaintitle{casebase: An Alternative Framework For Survival Analysis}
-\Shorttitle{\pkg{casebase}: An Alternative Framework For Survival Analysis}
+\Shorttitle{\pkg{casebase}: An Alternative Framework For Survival
+Analysis}
 
 \Abstract{
-The abstract of the article. * joint co-authors
+The abstract of the article. *SRB and MT contributed equally to this
+work.
 }
 
-\Keywords{keywords, not capitalized, \proglang{Java}}
-\Plainkeywords{keywords, not capitalized, Java}
+\Keywords{survival analysis, absolute risk, data visualization}
+\Plainkeywords{survival analysis, absolute risk, data visualization}
 
@@
 \Address{
-    Sahir Rai Bhatnagar *\\
+    Sahir Rai Bhatnagar*\\
   McGill University\\
@@
   URL: \url{http://sahirbhatnagar.com/}\\~\\
-      Maxime Turgeon *\\
+      Maxime Turgeon*\\
   University of Manitoba\\
@@
 retains the flexibility of case-base sampling and the familiar interface
-of the \code{glm} function. In addition, we provide additional features,
-including performance metrics and visualization tools.
+of the \code{glm} function. In addition, we provide extensive
+visualization tools.
 
 In what follows, we first recall some theoretical details on case-base
-sampling and its use for estimating parameteric hazard functions. We
-then give a short review of existing \proglang{R} packages that
-implement comparable features as \pkg{casebase}. Next, we provide some
-details about the implementation of case-base sampling in our package,
-and we give a brief survey of its main functions. This is followed by
-four case studies that illustrate the flexibility and capabilities of
+sampling and its use for estimating parametric hazard functions. We then
+give a short review of existing \proglang{R} packages that implement
+comparable features as \pkg{casebase}. Next, we provide some details
+about the implementation of case-base sampling in our package, and we
+give a brief survey of its main functions. This is followed by four case
+studies that illustrate the flexibility and capabilities of
 \pkg{casebase}. We show how the same framework can be used for competing
@@
 model, which implies that the counting process jumps are less than or
-equal to one. We are interested in modelling the hazard functions
+equal to one. We are interested in modeling the hazard functions
 \(\lambda_{i}(t)\) of the processes \(N_i(t)\), and which satisfy
@@
 \(u\). However, the integral over time makes the computation and
-maximisation of \(L_0(\theta)\) challenging.
+maximization of \(L_0(\theta)\) challenging.
 
@@
 \[L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.\]
-When a logarithmic link function is used for modelling the hazard
+When a logarithmic link function is used for modeling the hazard
 function, the above expression is of a logistic regression form with an
@@
 baseline hazard. The first category broadly corresponds to parametric
-survival models, where the full hazard is explictly modeled. Using this
+survival models, where the full hazard is explicitly modeled. Using this
 estimate, the survival function and the CIF can be obtained using their
@@
 functions. Providing smooth estimates of the CIF was one of the main
-motivations for introducting case-base sampling in survival analysis.
+motivations for introducing case-base sampling in survival analysis.
 
@@
 
-The functions in the casebase package can be divided into two
+The functions in the \pkg{casebase} package can be divided into two
 categories: 1) exploratory data analysis, in the form of population-time
-plots; and 2) parametric modeling of the hazard function. We explicitly
-aimed at being compatible with both \code{data.frame}s and
-\code{data.table}s. This is evident in some of the coding choices we
-made, and it is also reflected in our unit tests.
+plots; and 2) parametric modeling of the hazard function. We strove for
+compatibility with both \code{data.frame}s and \code{data.table}s; this
+can be seen in the coding choices we made and the unit tests we wrote.
 
@@
 The case-base sampling approach described in Section \ref{theory} can be
-visualized in the form of a population time plot. They are extremely
-informative graphical displays of survival data and should be one of the
-first steps in an exploratory data analysis. We facilitate this task in
-the \pkg{casebase} package using the \code{popTime} function and
-\code{plot} method as follows:
+visualized in the form of a population time plot. These plots are
+extremely informative graphical displays of survival data and should be
+one of the first steps in an exploratory data analysis. The
+\code{popTime} function and \code{plot} method facilitate this task:
 
@@
 
-The splitting of these tasks gives the user the flexibility to create
-population-time plots with the graphics system of their choice using the
-object created in Step 1, while the object created in Step 2 allows one
-to leverage the \pkg{ggplot2} \citep{ggplot2} family of functions for
-further customization.
-
-To illustrate these functions we will use the European Randomized Study
+By splitting these tasks, we give flexibility to the user. While the
+method call in Step 2 allows further customization by using the
+\pkg{ggplot2} \citep{ggplot2} family of functions, users can use for the
+graphics system of their choice to create population-time plots from the
+object created in Step 1.
+
+To illustrate these functions, we will use the European Randomized Study
 of Prostate Cancer Screening (ERSPC) data \citep{schroder2009screening}
-which was extracted using the approach described in
-\citep{liu2014recovering} and available through the \pkg{casebase}
-package. This data contains the individual observations for 159,893 men
-between the ages of 55 and 69 years at entry who were recruited from
-seven European countries.
+which was extracted using the approach described in Liu \emph{et al.}
+\citeyearpar{liu2014recovering} This dataset is available through the
+\pkg{casebase} package. It contains the individual observations for
+159,893 men from seven European countries, who were between the ages of
+55 and 69 years when recruited for the trial.
 
 We first create the necessary dataset for producing the population time
-plot using the \code{casebase::popTime} function. In this example we
-stratify by the treatment group. The resulting object inherits from
-class \texttt{popTime} and stores the exposure variable as an attribute:
+plot using the \code{popTime} function. In this example, we stratify the
+plot by treatment group. The resulting object inherits from class
+\code{popTime} and stores the exposure variable as an attribute:
 
@@
 
-We then pass this object to the corresponding \code{plot} method which
-leverages the \pkg{ggplot2} package to build population time plots.
+We then pass this object to the corresponding \code{plot} method:
 
@@
 gray, with subjects having the least amount of observation time plotted
-at the top of the y-axis. We immediately notice a unique \emph{step
-shape} in the population time area which is due to the randomization of
-the Finnish cohorts which were carried out on January 1 of each of the 4
-years 1996 to 1999. This, coupled with the uniform December 31 2006
-censoring date, lead to large numbers of men with exactly 11, 10, 9 or 8
-years of follow-up. Tracked backwards in time (i.e.~from right to left),
-the population-time plot shows the recruitment pattern from its
-beginning in 1991, and in particular the January 1 entries in successive
+at the top of the y-axis. We immediately notice a distinctive
+\emph{step-wise shape} in the population time area. This is due to the
+randomization of the Finnish cohorts which were carried out on January 1
+of each of year from 1996 to 1999. Coupled with the uniform December 31
+2006 censoring date, this led to large numbers of men with exactly 11,
+10, 9 or 8 years of follow-up. Tracked backwards in time (i.e.~from
+right to left), the population-time plot shows the recruitment pattern
+from its beginning in 1991, and the January 1 entries in successive
 years. Tracked forwards in time (i.e.~from left to right), the plot for
@@
 other causes). Since the Swedish and Belgian centres were the last to
-close their recruitment - in December 2003 - the minimum potential
-follow-up is three years. Tracked further forwards in time (i.e.~after
-year 3) the attrition is a combination of deaths and staggered entries.
+complete recruitment--in December 2003--the minimum potential follow-up
+is three years. Tracked further forwards in time (i.e.~after year 3) the
+attrition is a combination of deaths and staggered entries. As we can
+see, population-time plots summarise a wealth of information about the
+study into a simple graph.
 
@@
 series is sampled at random vertically on the plot to avoid having all
-points along the upper edge of the plot, while the base series is
-sampled horizontally with sampling weight proportional to their
-follow-up time. By randomly distributing the points, we can get a better
-sense of the incidence density. In the ERSPC data, we see that more
-events are observed later on in time. Therefore a constant hazard model
+points along the upper edge of the gray area. By randomly distributing
+the cases, we can get a better sense of the incidence density. In Figure
+\ref{fig:plot-stratified-erspc-data}, we see that more events are
+observed at later follow-up times. Therefore, a constant hazard model
 would not be appropriate in this instance as it would overestimate the
 cumulative incidence earlier on in time, and underestimate it later on.
-The reader is referred to the package vignettes for more examples and a
-detailed description of how to modify the aesthetics of a population
-time plot.
+Finally, the base series is sampled horizontally with sampling weight
+proportional to their follow-up time. The reader should refer to the
+package vignettes for more examples and a detailed description of how to
+modify the aesthetics of a population-time plot.
 
@@
   or elastic-net (cf.~Case Study 3). This functionality is provided
-  through the \code{glmnet} package \citep{friedman2010jss}.
+  through the \pkg{glmnet} package \citep{friedman2010jss}.
 \item
   \code{gam}: This option provides support for \emph{Generalized
-  Additive Models} via the \code{gam} package
+  Additive Models} via the \pkg{mgcv} package
   \citep{hastie1987generalized}.
@@
   \code{gbm}: This option provides support for \emph{Gradient Boosted
-  Trees} via the \code{gbm} package. This feature is still experimental.
+  Trees} via the \pkg{gbm} package. This feature is still experimental.
 \end{itemize}
@@
 
-We investigate the differences in survival between the control and
-screening arms in the ERSPC study presented earlier using casebase
-sampling to fit a parametric hazard function. We fit four models that
-differ in the functional form of time; (1) excluded from the linear
-predictor (Eq.\eqref{eq:exp}), (2) linear function (Eq.\eqref{eq:gomp}),
-(3) log function (Eq.\eqref{eq:weibull}), and (4) a smooth function
-using cubic B-splines. The models are fit using \code{fitSmoothHazard}.
-Since the output object from \code{fitSmoothHazard} inherits from the
-\code{glm} class, we see a familiar result when using the function
-\code{summary}:
+For our first case study, we return to the ERSPC study and investigate
+the differences in survival between the control and screening arms. We
+fit four models that differ in which functional form of time is used: 1)
+excluded from the linear predictor, 2) linear function, 3) log function,
+and 4) a smooth function using cubic B-splines. The models are fit using
+\code{fitSmoothHazard}. Since the output object from
+\code{fitSmoothHazard} inherits from the \code{glm} class, we can
+directly use the function \code{summary}:
 
@@
 R> 
-R> summary(fits[[2]])
+R> summary(fits[["gompertz"]])
 \end{CodeInput}
@@
 
-The \code{absoluteRisk} function takes as input the
+Next, the \code{absoluteRisk} function takes as input the
 \code{fitSmoothHazard} object and returns a matrix where each column
@@
 R> 
-R> head(risks[[2]])
+R> head(risks[["gompertz"]])
 \end{CodeInput}
@@
 In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from
-casebase on the Cox model CIF. The CIF estimates for the exponential
-model in panel (1) overestimate the cumulative incidence earlier on in
-time, and underestimate it later on. Based on what we observed in the
-population time plot, where more events are observed later on in time,
-this poor fit for the exponential hazard is expected. This example
-demonstrates the benefits of population time plots as an exploratory
-analysis tool.
-
-In Figure \ref{fig:erspc-cox-cif} panels (2)-(4), we notice a better fit
-with increasing complexities of time. As noted above, the usual
-asymptotic results hold for likelihood ratio tests built using casebase
-sampling models. Therefore, we can easily test the null hypothesis that
-the exponential model is just as good as the larger (in terms of number
-of parameters) splines model:
+\pkg{casebase} on the Cox model CIF. The CIF estimates for the
+exponential model in panel (1) overestimate the cumulative incidence
+earlier on in time, and underestimate it later on. Based on our earlier
+discussion of the population-time plot, this poor fit for the
+exponential hazard was expected.
+
+In Figure \ref{fig:erspc-cox-cif} (panels 2--4), we notice a better fit
+with increasing complexity of our model for time. As noted above, the
+usual asymptotic results hold for likelihood ratio tests built using
+case-base sampling models. Therefore, we can easily test the null
+hypothesis that the exponential model is just as good as the larger (in
+terms of number of parameters) splines model:
 
@@
 As expected, we see that splines model provides a better fit. Similarly,
-we can compare the AIC for all four casebase models:
+we can compare the AIC for all four case-base models:
 
@@
 
-\caption{Cumulative incidence functions for control and screening groups in the ERSPC data.
-In each of the panels, we plot the CIF from the Cox model using \code{survival::survfit} (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression.
-(1) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution.
-(2) Linear function of time which is equivalent to a Gompertz distribution.
-(3) The natural logarithm of time which is equivalent to a Weibull distribution.
+\caption{CIFs for control and screening groups in the ERSPC data.
+In each of the panels, we plot the CIF from the Cox model using \code{survival::survfit} (solid line) and the CIF from the case-base sampling scheme (dashed line) with different functional forms of time.
+(1) The time variable is excluded (exponential).
+(2) Linear function of time (Gompertz).
+(3) The natural logarithm (Weibull).
 (4) Cubic B-spline expansion of time.}\label{fig:erspc-cox-cif}
@@
 not need to model the full hazard to obtain reliable estimates of the
-hazard ratio. Of course, different parametric models for the hazards
-give rise to qualitatively different estimates for the CIF as shown in
-Figure \ref{fig:erspc-cox-cif}.
+hazard ratio. Nevertheless, Figure \ref{fig:erspc-cox-cif} shows that
+different parametric models can still give rise to qualitatively
+different estimates for the CIF.
 
@@
 
-With this first case study, we saw how population-time plots can provide
-a useful summary of the incidence for our event of interests. In
-particular, we saw how it can provide evidence against an exponential
-hazard model. We then explored how different parametric models can
-provide different estimates of the CIF, even though they all give
-similar estimates for the hazard ratios.
+With this first case study, we explored how \pkg{casebase} allows us to
+fit different parametric survival models, and how we can compare the fit
+of each model with tools from GLMs.
 
@@
 \citeyearpar{scrucca2010regression}. The data was downloaded from the
-first author's website, and it is also available as part of the
+first author's website, and it is now available as part of the
 \pkg{casebase} package.
@@
 CR3), source of stem cells (bone marrow and peripheral blood, coded as
-BM+PB, or only peripheral blood, coded as PB), and age. A summary of
-these baseline characteristics appear in Table \ref{tab:table1bmtcrr}.
-We note that the statistical summaries were generated differently for
-different variable types: for continuous variables, we gave the range,
-followed by the mean and standard deviation; for categorical variables,
-we gave the counts for each category.
-
-\begin{CodeChunk}
-\begin{table}
-
-\caption{\label{tab:table1bmtcrr}Baseline characteristics of patients in the stem-cell transplant study.}
-\centering
-\begin{tabular}[t]{lll}
-\toprule
-Variable & Description & Statistical summary\\
-\midrule
-\code{Sex} & Sex & M=Male (100)\\
- &  & F=Female (77)\\
-\code{D} & Disease & ALL (73)\\
- &  & AML (104)\\
-\code{Phase} & Phase & CR1 (47)\\
-\addlinespace
- &  & CR2 (45)\\
- &  & CR3 (12)\\
- &  & Relapse (73)\\
-\code{Source} & Type of transplant & BM+PB (21)\\
- &  & PB (156)\\
-\addlinespace
-\code{Age} & Age of patient (years) & 4--62\\
- &  & 30.47 (13.04)\\
-\code{ftime} & Failure time (months) & 0.13--131.77\\
- &  & 20.28 (30.78)\\
-\code{Status} & Status indicator & 0=censored (46)\\
-\addlinespace
- &  & 1=relapse (56)\\
- &  & 2=competing event (75)\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\end{CodeChunk}
+BM+PB, or only peripheral blood, coded as PB), and age.
 
@@
 received a stem-cell transplant from peripheral blood at relapse. We
-compared the absolute risk curve for such a woman with acute
-lymphoblastic leukemia with that for a similar woman with acute
-myeloblastic leukemia. We will estimate the curve from 0 to 60 months.
+compared the absolute risk curve for such a woman with ALL with that for
+a similar woman with AML. We will estimate the curve from 0 to 60
+months.
 
@@
 \[CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),\] where
-\(CI_k(t)\) is the cause-specific cumulative indicence. The Fine-Gray
+\(CI_k(t)\) is the cause-specific cumulative incidence. The Fine-Gray
 model allows to directly assess the effect of a covariate on the
@@
 case-base sampling and the Fine-Gray model. As we can see, the two
-approaches agree quite well for acute myeloblastic leukemia; however,
-there seems to be a difference of about 5\% between the two curves for
-acute lymphoblastic leukemia. This difference does not appear to be
-significant: the curve from case-base sampling is contained within a
-95\% confidence band around the Fine-Gray absolute risk curve (figure
-not shown).
+approaches agree quite well for AML; however, there seems to be a
+difference of about 5\% between the two curves for ALL. This difference
+does not appear to be significant: the curve from case-base sampling is
+contained within a 95\% confidence band around the Fine-Gray absolute
+risk curve (figure not shown).
 
@@
 
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{../figures/brierpt1-1}
-
-}
-
-\caption{\label{fig:brier1} Comparison of three Cox regression models. (A) Cumulative Incidence curves. (B) Time-dependent Brier scores.}\label{fig:brierpt1}
-\end{figure}
-\end{CodeChunk}
-
 For our penalized logistic regression model, we opt for the natural log
@@
 is a matrix interface to the function \code{fitSmoothHazard}. In order
-to use this interfact, we first create a matrix \code{y} containing the
+to use this interface, we first create a matrix \code{y} containing the
 time and event variables, and a matrix \code{x} containing all remaining
@@
 
-Note that the fourth model (i.e.~\texttt{fit4}) includes an interaction
-term between exposure and follow-up time. In other words, this model no
-longer exhibit proportional hazards. The evidence of non-proportionality
-of hazards in the Stanford Heart Transplant data has been widely
-discussed \citep{arjas1988graphical}.
+Note that the fourth model includes an interaction term between exposure
+and follow-up time. In other words, this model no longer exhibit
+proportional hazards. The evidence of non-proportionality of hazards in
+the Stanford Heart Transplant data has been widely discussed
+\citep{arjas1988graphical}.
 
@@
 #> Model1 Model2 Model3 Model4 
-#>    845    491    495    486
+#>    794    760    736    722
 \end{CodeOutput}
@@
 clearly see how the hazards are no longer proportional.
-
-THIS NEXT CODE CHUNK IS NOT WORKING
 
@@
 R> hazard_data <- expand.grid(
-R>   exposure = c(0, 1),
-R>   futime = seq(0, 1000,
-R>     length.out = 100
-R>   )
-R> )
+R+   exposure = c(0, 1),
+R+   futime = seq(0, 1000,
+R+     length.out = 100
+R+   )
+R+ )
 R> 
@@
 R> hazard_data$hazard <- exp(predict(fit4,
-R>   newdata = hazard_data,
-R>   type = "link"
-R> ))
+R+   newdata = hazard_data,
+R+   type = "link"
+R+ ))
 R> # Add labels for plots
 R> hazard_data$Status <- factor(hazard_data$exposure,
-R>   labels = c("NoTrans", "Trans")
-R> )
+R+   labels = c("NoTrans", "Trans")
+R+ )
 R> 
 R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
-R>   geom_line() +
-R>   paper_gg_theme +  
-R>   ylab("Hazard") +
-R>   xlab("Follow-up time")
+R+   geom_line() +
+R+   paper_gg_theme +  
+R+   ylab("Hazard") +
+R+   xlab("Follow-up time")
 \end{CodeInput}
+
+\begin{center}\includegraphics{../figures/stanford-hazard-1} \end{center}
+
 \end{CodeChunk}
@@
 \begin{CodeOutput}
-#> [1] "absRiskCB" "matrix"
+#> [1] "absRiskCB" "matrix"    "array"
 \end{CodeOutput}
@@
 #>  time    
-#>     0 0
-#>   365 1
+#>     0 0.0
+#>   365 0.3
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> [1] 0.99
+#> [1] 0.2
 \end{CodeOutput}
@@
 
-We can also plot the Hazard ratio:
+We can also plot the hazard ratio:
 
@@
 
-This report was generated on 2020-07-19 18:02:45 using the following
+This report was generated on 2020-07-21 19:46:24 using the following
 computational environment and dependencies:
@@
 \begin{CodeOutput}
-#> R version 3.6.2 (2019-12-12)
-#> Platform: x86_64-pc-linux-gnu (64-bit)
-#> Running under: Pop!_OS 19.10
+#> R version 4.0.0 (2020-04-24)
+#> Platform: x86_64-apple-darwin17.0 (64-bit)
+#> Running under: macOS Catalina 10.15.5
 #> 
 #> Matrix products: default
-#> BLAS:   /usr/lib/x86_64-linux-gnu/openblas/libblas.so.3
-#> LAPACK: /usr/lib/x86_64-linux-gnu/libopenblasp-r0.3.7.so
+#> BLAS:   /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRblas.dylib
+#> LAPACK: /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRlapack.dylib
 #> 
 #> locale:
-#>  [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
-#>  [3] LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
-#>  [5] LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
-#>  [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                 
-#>  [9] LC_ADDRESS=C               LC_TELEPHONE=C            
-#> [11] LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       
+#> [1] en_CA.UTF-8/en_CA.UTF-8/en_CA.UTF-8/C/en_CA.UTF-8/en_CA.UTF-8
 #> 
@@
 #> other attached packages:
-#>  [1] broom_0.7.0               dotwhisker_0.5.0         
+#>  [1] broom_0.5.6               dotwhisker_0.5.0         
 #>  [3] timereg_1.9.4             glue_1.4.1               
-#>  [5] forcats_0.5.0             stringr_1.4.0            
-#>  [7] dplyr_0.8.5               purrr_0.3.4              
-#>  [9] readr_1.3.1               tidyr_1.1.0              
-#> [11] ggplot2_3.3.2.9000        tidyverse_1.3.0          
-#> [13] lubridate_1.7.4           visreg_2.7.0             
-#> [15] tibble_3.0.1              riskRegression_2020.02.05
-#> [17] prodlim_2019.11.13        flexsurv_1.1.1           
-#> [19] survival_3.1-8            pracma_2.2.9             
-#> [21] magrittr_1.5              kableExtra_1.1.0         
-#> [23] glmnet_4.0-2              Matrix_1.2-18            
-#> [25] data.table_1.12.8         cowplot_1.0.0            
-#> [27] colorspace_1.4-1          casebase_0.9.0           
+#>  [5] scales_1.1.0              forcats_0.5.0            
+#>  [7] stringr_1.4.0             dplyr_0.8.5              
+#>  [9] purrr_0.3.4               readr_1.3.1              
+#> [11] tidyr_1.0.2               ggplot2_3.3.0            
+#> [13] tidyverse_1.3.0           lubridate_1.7.8          
+#> [15] visreg_2.6.1              tibble_3.0.1             
+#> [17] riskRegression_2020.02.05 prodlim_2019.11.13       
+#> [19] flexsurv_1.1.1            survival_3.1-12          
+#> [21] pracma_2.2.9              magrittr_1.5             
+#> [23] kableExtra_1.1.0          glmnet_3.0-2             
+#> [25] Matrix_1.2-18             data.table_1.12.8        
+#> [27] cowplot_1.0.0             colorspace_1.4-1         
+#> [29] casebase_0.9.0.9000      
 #> 
 #> loaded via a namespace (and not attached):
-#>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
-#>  [4] ggstance_0.3.3      htmlTable_1.13.3    base64enc_0.1-3    
-#>  [7] fs_1.3.2            rstudioapi_0.11     farver_2.0.3       
-#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.0-12     
-#> [13] xml2_1.3.0          codetools_0.2-16    knitr_1.29         
-#> [16] Formula_1.2-3       jsonlite_1.7.0      cluster_2.1.0      
-#> [19] dbplyr_1.4.2        png_0.1-7           compiler_3.6.2     
-#> [22] httr_1.4.1          backports_1.1.8     assertthat_0.2.1   
-#> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
-#> [28] quantreg_5.54       tools_3.6.2         gtable_0.3.0       
-#> [31] Rcpp_1.0.4.6        cellranger_1.1.0    vctrs_0.3.1        
-#> [34] nlme_3.1-143        iterators_1.0.12    xfun_0.15          
-#> [37] rvest_0.3.5         lifecycle_0.2.0     polspline_1.1.17   
-#> [40] muhaz_1.2.6.1       MASS_7.3-51.5       zoo_1.8-7          
-#> [43] scales_1.1.1        hms_0.5.3           sandwich_2.5-1     
-#> [46] SparseM_1.78        RColorBrewer_1.1-2  rticles_0.13       
-#> [49] yaml_2.2.1          gridExtra_2.3       rms_5.1-4          
-#> [52] rpart_4.1-15        latticeExtra_0.6-29 stringi_1.4.6      
-#> [55] foreach_1.5.0       checkmate_2.0.0     lava_1.6.6         
-#> [58] shape_1.4.4         rlang_0.4.6         pkgconfig_2.0.3    
-#> [61] evaluate_0.14       lattice_0.20-38     labeling_0.3       
-#> [64] htmlwidgets_1.5.1   cmprsk_2.2-9        tidyselect_1.1.0   
-#> [67] deSolve_1.27.1      plyr_1.8.6          R6_2.4.1           
-#> [70] generics_0.0.2      Hmisc_4.3-1         multcomp_1.4-12    
-#> [73] DBI_1.1.0           withr_2.2.0         pillar_1.4.4       
-#> [76] haven_2.3.1         foreign_0.8-74      mgcv_1.8-31        
-#> [79] nnet_7.3-12         mstate_0.2.12       modelr_0.1.5       
-#> [82] crayon_1.3.4        rmarkdown_2.3       jpeg_0.1-8.1       
-#> [85] grid_3.6.2          readxl_1.3.1        reprex_0.3.0       
-#> [88] digest_0.6.25       webshot_0.5.2       numDeriv_2016.8-1.1
-#> [91] stats4_3.6.2        munsell_0.5.0       viridisLite_0.3.0  
-#> [94] quadprog_1.5-8
+#>  [1] TH.data_1.0-10      VGAM_1.1-2          ellipsis_0.3.1     
+#>  [4] ggstance_0.3.4      htmlTable_1.13.3    base64enc_0.1-3    
+#>  [7] fs_1.4.1            rstudioapi_0.11     MatrixModels_0.4-1 
+#> [10] fansi_0.4.1         mvtnorm_1.1-0       xml2_1.3.2         
+#> [13] codetools_0.2-16    knitr_1.29          Formula_1.2-3      
+#> [16] jsonlite_1.6.1      cluster_2.1.0       dbplyr_1.4.3       
+#> [19] png_0.1-7           compiler_4.0.0      httr_1.4.1         
+#> [22] backports_1.1.8     assertthat_0.2.1    cli_2.0.2          
+#> [25] acepack_1.4.1       htmltools_0.5.0     quantreg_5.55      
+#> [28] tools_4.0.0         gtable_0.3.0        Rcpp_1.0.4.6       
+#> [31] cellranger_1.1.0    vctrs_0.3.1         nlme_3.1-147       
+#> [34] iterators_1.0.12    xfun_0.15           rvest_0.3.5        
+#> [37] lifecycle_0.2.0     polspline_1.1.17    muhaz_1.2.6.1      
+#> [40] MASS_7.3-51.5       zoo_1.8-7           hms_0.5.3          
+#> [43] sandwich_2.5-1      SparseM_1.78        RColorBrewer_1.1-2 
+#> [46] rticles_0.14        yaml_2.2.1          gridExtra_2.3      
+#> [49] rms_5.1-4           rpart_4.1-15        latticeExtra_0.6-29
+#> [52] stringi_1.4.6       foreach_1.5.0       checkmate_2.0.0    
+#> [55] lava_1.6.7          shape_1.4.4         rlang_0.4.6        
+#> [58] pkgconfig_2.0.3     evaluate_0.14       lattice_0.20-41    
+#> [61] htmlwidgets_1.5.1   cmprsk_2.2-9        tidyselect_1.1.0   
+#> [64] deSolve_1.28        R6_2.4.1            generics_0.0.2     
+#> [67] Hmisc_4.4-0         multcomp_1.4-13     DBI_1.1.0          
+#> [70] withr_2.2.0         pillar_1.4.4        haven_2.2.0        
+#> [73] foreign_0.8-78      mgcv_1.8-31         nnet_7.3-13        
+#> [76] mstate_0.2.12       modelr_0.1.6        crayon_1.3.4       
+#> [79] rmarkdown_2.1       jpeg_0.1-8.1        grid_4.0.0         
+#> [82] readxl_1.3.1        reprex_0.3.0        digest_0.6.25      
+#> [85] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_4.0.0       
+#> [88] munsell_0.5.0       viridisLite_0.3.0   quadprog_1.5-8
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> Local:    write-popTime /home/sahir/git_repositories/cbpaper
-#> Remote:   write-popTime @ origin (https://github.com/sahirbhatnagar/cbpaper.git)
-#> Head:     [dad6070] 2020-07-18: cleaned cs1
+#> Local:    editMax /Users/maxturgeon/Documents/git_repos/cbpaper
+#> Head:     [2d08bac] 2020-07-20: Merge pull request #22 from sahirbhatnagar/write-popTime
 \end{CodeOutput}

### 2020-07-22 analysis/paper/paper.tex
fix hazard ratio plot

@@
 R+   exposure = c(0, 1),
-R+   futime = seq(0, 1000,
+R+   futime = seq(0, 425,
 R+     length.out = 100
@@
 which provides functions for fitting smooth parametric hazards and
-estimating CIFs using case-base sampling. We outlined the theoretical
-underpinnings of the approach, we provided details about our
-implementation, and we illustrated the merits of the approach and the
-package through four case studies.
+estimating CIFs using case-base sampling. Our package also provide
+several functions to produce graphical summaries of the data and the
+results. We outlined the theoretical underpinnings of the approach, we
+provided details about our implementation, and we illustrated the merits
+of the approach and the package through four case studies.
 
@@
 compute individualized risk assessments for chronic diseases. In the
-case studies above, we explored this flexibility along two fronts. On
-the one hand, we showed how splines could be used as part of the linear
-predictor to model the effect of time on the hazard. This strategy
-yielded estimates of the survival function that were qualitatively
-similar to semiparametric estimates derived from Cox regression;
-however, case-base sampling led to estimates of the survival function
-that \emph{vary smoothly in time}. On the other hand, we also displayed
-the flexibility of case-base sampling by showing how it could be
-combined with penalized logistic regression to perform variable
-selection. Even though we did not illustrate it in this article,
-case-base sampling can also be combined with the framework of
-\emph{generalized additive models}. This functionality has been
-implemented in our package \pkg{casebase}. Similarly, case-base sampling
-can also be combined with quasi-likelihood estimation to fit survival
-models that can account for the presence of over-dispersion. All of
-these examples illustrate how the case-base sampling framework in
-general, and the package \pkg{casebase} in particular, allows the user
-to fit a broad and flexible family of survival functions.
+case studies above, we further explored this flexibility along two
+fronts. On the one hand, we showed how splines could be used as part of
+the linear predictor to model the effect of time on the hazard. This
+strategy yielded estimates of the survival function that were
+qualitatively similar to semiparametric estimates derived from Cox
+regression; however, case-base sampling led to estimates of the survival
+function that \emph{vary smoothly in time} and are thus easier to
+interpret. On the other hand, we also displayed the flexibility of
+case-base sampling by showing how it could be combined with penalized
+logistic regression to perform variable selection. Even though we did
+not illustrate it in this article, case-base sampling can also be
+combined with the framework of \emph{generalized additive models}. This
+functionality has already been implemented in the package. Similarly,
+case-base sampling can be combined with quasi-likelihood estimation to
+fit survival models that can account for the presence of
+over-dispersion. All of these examples illustrate how the case-base
+sampling framework in general, and the package \pkg{casebase} in
+particular, allows the user to fit a broad and flexible family of
+survival functions.
 
@@
 smooth cumulative incidence curves, whereas Cox regression and Fine-Gray
-models produce step-wise cumulative incidence functions and never
-explicitly model the hazard function. Qualitatively, we showed that by
-using splines in the linear predictor, all three models yielded similar
-curves. However, the smooth nature of the output of \pkg{casebase}
-provides a more intuitive interpretation for consumers of these
-predictions. In Table \ref{tab:compCBvsCox}, we provide a side-by-side
-comparison between the Cox model and case-base sampling.
+models produce step-wise CIFs and never explicitly model the hazard
+function. Qualitatively, we showed that by using splines in the linear
+predictor, all three models yielded similar curves. However, the smooth
+nature of the output of \pkg{casebase} provides a more intuitive
+interpretation for consumers of these predictions. In Table
+\ref{tab:compCBvsCox}, we provide a side-by-side comparison between the
+Cox model and case-base sampling.
 
@@
 case-base sampling. First, to assess the quality of the model fit, we
-want to study the properties of the residuals (e.g.~Cox-Snell,
+would like to study the properties of the residuals (e.g.~Cox-Snell,
 martingale). More work needs to be done to understand these residuals in
 the context of the partial likelihood underlying case-base sampling. The
-resulting diagnostic tools would then be integrated in this package.
+resulting diagnostic tools could then be integrated in this package.
 Also, we are interested in extending case-base sampling to account for
@@
 
-This report was generated on 2020-07-21 19:46:24 using the following
+This report was generated on 2020-07-21 20:19:45 using the following
 computational environment and dependencies:
@@
 #>  [4] ggstance_0.3.4      htmlTable_1.13.3    base64enc_0.1-3    
-#>  [7] fs_1.4.1            rstudioapi_0.11     MatrixModels_0.4-1 
-#> [10] fansi_0.4.1         mvtnorm_1.1-0       xml2_1.3.2         
-#> [13] codetools_0.2-16    knitr_1.29          Formula_1.2-3      
-#> [16] jsonlite_1.6.1      cluster_2.1.0       dbplyr_1.4.3       
-#> [19] png_0.1-7           compiler_4.0.0      httr_1.4.1         
-#> [22] backports_1.1.8     assertthat_0.2.1    cli_2.0.2          
-#> [25] acepack_1.4.1       htmltools_0.5.0     quantreg_5.55      
-#> [28] tools_4.0.0         gtable_0.3.0        Rcpp_1.0.4.6       
-#> [31] cellranger_1.1.0    vctrs_0.3.1         nlme_3.1-147       
-#> [34] iterators_1.0.12    xfun_0.15           rvest_0.3.5        
-#> [37] lifecycle_0.2.0     polspline_1.1.17    muhaz_1.2.6.1      
-#> [40] MASS_7.3-51.5       zoo_1.8-7           hms_0.5.3          
-#> [43] sandwich_2.5-1      SparseM_1.78        RColorBrewer_1.1-2 
-#> [46] rticles_0.14        yaml_2.2.1          gridExtra_2.3      
-#> [49] rms_5.1-4           rpart_4.1-15        latticeExtra_0.6-29
-#> [52] stringi_1.4.6       foreach_1.5.0       checkmate_2.0.0    
-#> [55] lava_1.6.7          shape_1.4.4         rlang_0.4.6        
-#> [58] pkgconfig_2.0.3     evaluate_0.14       lattice_0.20-41    
-#> [61] htmlwidgets_1.5.1   cmprsk_2.2-9        tidyselect_1.1.0   
-#> [64] deSolve_1.28        R6_2.4.1            generics_0.0.2     
-#> [67] Hmisc_4.4-0         multcomp_1.4-13     DBI_1.1.0          
-#> [70] withr_2.2.0         pillar_1.4.4        haven_2.2.0        
-#> [73] foreign_0.8-78      mgcv_1.8-31         nnet_7.3-13        
-#> [76] mstate_0.2.12       modelr_0.1.6        crayon_1.3.4       
-#> [79] rmarkdown_2.1       jpeg_0.1-8.1        grid_4.0.0         
-#> [82] readxl_1.3.1        reprex_0.3.0        digest_0.6.25      
-#> [85] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_4.0.0       
-#> [88] munsell_0.5.0       viridisLite_0.3.0   quadprog_1.5-8
+#>  [7] fs_1.4.1            rstudioapi_0.11     farver_2.0.3       
+#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-0      
+#> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
+#> [16] Formula_1.2-3       jsonlite_1.6.1      cluster_2.1.0      
+#> [19] dbplyr_1.4.3        png_0.1-7           compiler_4.0.0     
+#> [22] httr_1.4.1          backports_1.1.8     assertthat_0.2.1   
+#> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
+#> [28] quantreg_5.55       tools_4.0.0         gtable_0.3.0       
+#> [31] Rcpp_1.0.4.6        cellranger_1.1.0    vctrs_0.3.1        
+#> [34] nlme_3.1-147        iterators_1.0.12    xfun_0.15          
+#> [37] rvest_0.3.5         lifecycle_0.2.0     polspline_1.1.17   
+#> [40] muhaz_1.2.6.1       MASS_7.3-51.5       zoo_1.8-7          
+#> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
+#> [46] RColorBrewer_1.1-2  rticles_0.14        yaml_2.2.1         
+#> [49] gridExtra_2.3       rms_5.1-4           rpart_4.1-15       
+#> [52] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
+#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
+#> [58] rlang_0.4.6         pkgconfig_2.0.3     evaluate_0.14      
+#> [61] lattice_0.20-41     labeling_0.3        htmlwidgets_1.5.1  
+#> [64] cmprsk_2.2-9        tidyselect_1.1.0    deSolve_1.28       
+#> [67] R6_2.4.1            generics_0.0.2      Hmisc_4.4-0        
+#> [70] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
+#> [73] pillar_1.4.4        haven_2.2.0         foreign_0.8-78     
+#> [76] mgcv_1.8-31         nnet_7.3-13         mstate_0.2.12      
+#> [79] modelr_0.1.6        crayon_1.3.4        rmarkdown_2.1      
+#> [82] jpeg_0.1-8.1        grid_4.0.0          readxl_1.3.1       
+#> [85] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
+#> [88] numDeriv_2016.8-1.1 stats4_4.0.0        munsell_0.5.0      
+#> [91] viridisLite_0.3.0   quadprog_1.5-8
 \end{CodeOutput}

### 2020-07-22 analysis/paper/paper.Rmd

@@
   exposure = c(0, 1),
-  futime = seq(0, 1000,
+  futime = seq(0, 425,
     length.out = 100
@@
   newdata = newdata,
-  time = seq(0, 1000, length.out = 100)
+  time = seq(0, 425, length.out = 100)
 )
@@
 ```{r eval = eval_cs4}
-newtime <- quantile(fit4[["originalData"]][[fit4[["timeVar"]]]],
-                    probs = seq(0.01, 0.99, 0.01))
+newtime <- seq(0, 425, length.out = 100)
 
@@
      rug = TRUE)
+abline(h = 1, lty = 2)
 ```
@@
 In this article, we presented the \proglang{R} package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating CIFs using case-base sampling.
+Our package also provide several functions to produce graphical summaries of the data and the results.
 We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through four case studies.
@@
 As another example, Saarela and Arjas [-@saarela2015non] combined case-base sampling and a Bayesian non-parametric framework to compute individualized risk assessments for chronic diseases.
-In the case studies above, we explored this flexibility along two fronts.
+In the case studies above, we further explored this flexibility along two fronts.
 On the one hand, we showed how splines could be used as part of the linear predictor to model the effect of time on the hazard.
-This strategy yielded estimates of the survival function that were qualitatively similar to semiparametric estimates derived from Cox regression; however, case-base sampling led to estimates of the survival function that *vary smoothly in time*.
+This strategy yielded estimates of the survival function that were qualitatively similar to semiparametric estimates derived from Cox regression; however, case-base sampling led to estimates of the survival function that *vary smoothly in time* and are thus easier to interpret.
 On the other hand, we also displayed the flexibility of case-base sampling by showing how it could be combined with penalized logistic regression to perform variable selection.
 Even though we did not illustrate it in this article, case-base sampling can also be combined with the framework of *generalized additive models*.
-This functionality has been implemented in our package \pkg{casebase}.
-Similarly, case-base sampling can also be combined with quasi-likelihood estimation to fit survival models that can account for the presence of over-dispersion.
+This functionality has already been implemented in the package.
+Similarly, case-base sampling can be combined with quasi-likelihood estimation to fit survival models that can account for the presence of over-dispersion.
 All of these examples illustrate how the case-base sampling framework in general, and the package \pkg{casebase} in particular, allows the user to fit a broad and flexible family of survival functions.
@@
 However, in terms of output, both approaches differ significantly.
-Case-base sampling produces smooth hazards and smooth cumulative incidence curves, whereas Cox regression and Fine-Gray models produce step-wise cumulative incidence functions and never explicitly model the hazard function.
+Case-base sampling produces smooth hazards and smooth cumulative incidence curves, whereas Cox regression and Fine-Gray models produce step-wise CIFs and never explicitly model the hazard function.
 Qualitatively, we showed that by using splines in the linear predictor, all three models yielded similar curves.
@@
 Future work will look at some of the methodological extensions of case-base sampling.
-First, to assess the quality of the model fit, we want to study the properties of the residuals (e.g.
+First, to assess the quality of the model fit, we would like to study the properties of the residuals (e.g.
 Cox-Snell, martingale).
 More work needs to be done to understand these residuals in the context of the partial likelihood underlying case-base sampling.
-The resulting diagnostic tools would then be integrated in this package.
+The resulting diagnostic tools could then be integrated in this package.
 Also, we are interested in extending case-base sampling to account for interval censoring.

### 2020-07-28 analysis/paper/paper.Rmd
Vetted lit-rev, fig and fig comments (cs3)

@@
 eval_implementation <- TRUE
-eval_cs1 <- TRUE
-eval_cs2 <- TRUE
+eval_cs1 <- T
+eval_cs2 <- T
 eval_cs3 <- TRUE
-eval_cs4 <- TRUE
+eval_cs4 <- T
 eval_colophon <- TRUE
@@
 & \textbf{Competing} & \textbf{Allows} &  \textbf{Penalized}   &    &  & \textbf{Semi} & \textbf{Interval/Left} & \textbf{Absolute}  \\
-\textbf{Package}        & \textbf{Risks}  & \textbf{Non PH} & \textbf{Likelihood} & \textbf{Splines} & \textbf{Parametric} & \textbf{Parametric} & \textbf{Censoring} & \textbf{Risk}           \\
+\textbf{Package}        & \textbf{Risks}  & \textbf{Non PH} & \textbf{Regression} & \textbf{Splines} & \textbf{Parametric} & \textbf{Parametric} & \textbf{Censoring} & \textbf{Risk}           \\
 \hline
@@
 \textbf{coxRidge}        &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{crrp}            & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
+\textbf{crrp}            & x                        &                           & x                     &                  &                     & x                        &                                  &                                  \\
 \textbf{fastcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
@@
 \textbf{penalized}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{RiskRegression}  &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
-\textbf{Rstpm2}          &                          & x                         & x                     & x                & x                   & x                        & x                                & x                         \\
+\textbf{RiskRegression}  & x                         &                           & x                     &                  &                     & x                        &                                  & x                                \\
+\textbf{Rstpm2}          &                          & x                         &                      & x                & x                   & x                        & x                                & x                         \\
 \textbf{SmoothHazard}    &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
@@
 }
-\caption{Comparison of various \proglang{R} packages for survival analysis based on several defining features.}
+\caption{Comparison of various \proglang{R} packages for survival analysis based on several defining features.
+Competing Risks: whether or not an implementation for competing risks is present.
+Allows Non-PH: permits models that can incorporate variables that change with survival time.
+Penalized Regression: allows for a penalty term when fitting the model.
+This is not restricted to lasso nor ridge.
+Splines: the package permits a flexible fit on time through the use of splines.
+Parametric: the package has implementations for parametric models.
+Semi-parametric: the package has implementations for semi-parametric models.
+Interval/left censoring: the package can interpret interval and left-censoring.
+If this is not selected, the package only handles right-censoring.
+Absolute Risk: the package has an implementation for survival curves, cumulative incidence or cumulative hazard rate functions to be plotted.}
 \label{tab:surv-pkgs}
@@
 # 4. Kaplan-Meier
-KM <- survival::coxph(Surv(time = d.time, event = death) ~ 1, data = test)
+KM <- survival::coxph(Surv(time = d.time, event = death) ~ 1, data = test,x=TRUE)
 abKM <- survival::survfit(KM)
@@
 cox_ci <- ggplot(baseCoxComparisonData, aes(Time)) +
-  geom_line(aes(y = APS, colour = "APS"), lwd = 1) +
-  geom_line(aes(y = Full, col = "Full"), lwd = 1) +
-  geom_line(aes(y = SPS, colour = "SPS"), lwd = 1) +
-  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M"), lwd = 1) +
-  labs(y = "Probability of Death", x = "Survival time (Years)", color = "Models") +
+  geom_step(aes(y = APS, colour = "APS")) +
+  geom_step(aes(y = Full, col = "Full")) +
+  geom_step(aes(y = SPS, colour = "SPS")) +
+  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
+  labs(y = "Probability of Death", x = "Follow-up time (Years)", color = "Models") +
   scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
@@
 
-```{r brierpt1, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:brier1} Comparison of three Cox regression models. (A) Cumulative Incidence curves. (B) Time-dependent Brier scores."}
+```{r brierpt1, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:brier1} Comparison of three Cox regression models to the unadjusted Kaplan-Meier curve.
+(A) Cumulative Incidence curves.
+For a given model, the plotted curve is the average curve of each individual in the study.
+Each follow-up time seen in the study was used.
+There are minimal differences between the curves.
+APS coincides with the SPS curve, as they are considerably overlapped.(B) Time-dependent Brier scores, where a lower score is better.
+The full model performs best overall."}
 # Extract time points
@@
                    se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
+brierNull <- Score(list("PredictionModel" = KM), data = test, 
+                   formula = Surv(d.time, death != 0) ~ 1, summary = "ipa", 
+                   se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
 
@@
   data.frame(brierScore = brierSPS$Brier$score$Brier[brierSPS$Brier$score$model == "PredictionModel"], 
-             times = times, model = "SPS")
+             times = times, model = "SPS"),
+    data.frame(brierScore = brierNull$Brier$score$Brier[brierNull$Brier$score$model == "PredictionModel"], 
+             times = times, model = "K-M")
 )
@@
   geom_line() +
-  xlab("Survival time (Years)") +
+  xlab("Follow-up time (Years)") +
   ylab("Brier score") +
@@
   # ggtitle("Brier scores over time") +
-  scale_color_manual(values = q4[1:3]) + 
+  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
   paper_gg_theme
@@
 ```{r coefplots, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models.
-If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity."}
+If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity.
+Shrinkage occurs due to penalization."}
 library(dotwhisker)
@@
 # extract coefficients for lollipop plot
-estimatescoxNet <- setDT(as.data.frame(coef(coxNet)), 
+estimatescoxNet <- data.table::setDT(as.data.frame(coef(coxNet)), 
                          keep.rownames = TRUE)
@@
 
-estimatescb <- setDT(as.data.frame(coef(cbFull)[-c(1, 2), 1]), 
+estimatescb <- data.table::setDT(as.data.frame(coef(cbFull)[-c(1, 2), 1]), 
                      keep.rownames = TRUE)
@@
 
-estimatescoxFull <- setDT(as.data.frame(coef(coxFull)), 
+estimatescoxFull <- data.table::setDT(as.data.frame(coef(coxFull)), 
                           keep.rownames = TRUE)
@@
 In Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our penalized logistic regression model compared to both Cox regression models.
-However, this difference is minimal.
-On the other hand, all three models outperform the Kaplan-Meier model.
+The differences, however, are minimal.
+All three outperform the Kaplan-Meier model.
 
@@
 reg_ci <- ggplot(ab, aes(Time)) +
-  geom_line(data = baseCoxComparisonData, mapping = aes(y = Full, colour = "Cox"), lwd = 1) +
-  geom_line(aes(y = absRisk, colour = Model), lwd = 1) +
-  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M"), lwd = 1) +
-  labs(y = "Probability of Death", x = "Survival time (Years)", color = "Models") +
+  geom_step(data = baseCoxComparisonData, mapping = aes(x = Time,y = Full, colour = "Cox")) +
+  geom_line(aes(y = absRisk, colour = Model)) +
+  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
+  labs(y = "Probability of Death", x = "Follow-up time (Years)", color = "Models") +
   # ggtitle("Absolute risk curves") +
-  scale_color_manual(values = q4, breaks = c("Cox", "Penalized Cox", "Penalized logistic", "K-M")) + 
+  scale_color_manual(values = q4, breaks = c("K-M","Cox","Penalized Cox","Penalized logistic")) + 
   paper_gg_theme
@@
 
-```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} Four curves comparing the Cumulative Incidence (A) and the Brier scores (B) as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.", warning = FALSE}
+```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} (A) Four curves comparing the probabilities of death as it changes over follow-up time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
+All follow-up times seen in the new data were used.
+(B) Four curves demonstrating the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
+Lower is better.
+Penalized logistic regression produced the lowest results over follow-up time.", warning = FALSE}
 # Compute Brier score for penalized models
@@
                            se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
-
+#brierFinalResults$models<-as.character(brierFinalResults$models)
 # Combine Brier scores between unpenalized and penalized models before plotting
+
 data_brier <- bind_rows(
@@
   filter(brierFull$Brier$score, model == "PredictionModel")
-  ) %>% 
-  mutate(model = case_when(
-    model == "Null model" ~ "K-M",
-    model == "PredictionModel" ~ "Cox",
-    TRUE ~ model
-  ),
-  model = factor(model, levels = c("Cox", "Penalized Cox", "Penalized Logistic", "K-M"),
-                 labels = c("Cox", "Pen. Cox", "Pen. Logistic", "K-M")))
+  )
+data_brier$model<-factor(data_brier$model,c("Null model", "PredictionModel","Penalized Cox", "Penalized Logistic"))
+levels(data_brier$model)<-c("Kaplan-Meier","Cox","Penalized Cox","Penalized logistic")
 

### 2020-07-28 analysis/paper/paper.tex
Vetted lit-rev, fig and fig comments (cs3)

@@
 \author{
-Sahir Rai Bhatnagar*\\McGill University \And Maxime Turgeon*\\University
-of Manitoba \AND Jesse Islam\\McGill University \And James
-Hanley\\McGill University \And Olli Saarela\\University of Toronto
+Sahir Rai Bhatnagar*\\McGill University \And Maxime Turgeon*\\University of Manitoba \AND Jesse Islam\\McGill University \And James Hanley\\McGill University \And Olli Saarela\\University of Toronto
 }
@@
 
-\Plainauthor{Sahir Rai Bhatnagar*, Maxime Turgeon*, Jesse Islam, James
-Hanley, Olli Saarela}
+\Plainauthor{Sahir Rai Bhatnagar*, Maxime Turgeon*, Jesse Islam, James Hanley, Olli Saarela}
 \Plaintitle{casebase: An Alternative Framework For Survival Analysis}
-\Shorttitle{\pkg{casebase}: An Alternative Framework For Survival
-Analysis}
+\Shorttitle{\pkg{casebase}: An Alternative Framework For Survival Analysis}
 
@@
 
-\caption{\label{fig:brier1} Comparison of three Cox regression models. (A) Cumulative Incidence curves. (B) Time-dependent Brier scores.}\label{fig:brierpt1}
+\caption{\label{fig:brier1} Comparison of three Cox regression models to the unadjusted Kaplan-Meier curve.
+(A) Cumulative Incidence curves.
+For a given model, the plotted curve is the average curve of each individual in the study.
+Each follow-up time seen in the study was used.
+There are minimal differences between the curves.
+APS coincides with the SPS curve, as they are considerably overlapped.(B) Time-dependent Brier scores, where a lower score is better.
+The full model performs best overall.}\label{fig:brierpt1}
 \end{figure}
@@
 \caption{\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models.
-If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity.}\label{fig:coefplots}
+If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity.
+Shrinkage occurs due to penalization.}\label{fig:coefplots}
 \end{figure}
@@
 penalized logistic regression model compared to both Cox regression
-models. However, this difference is minimal. On the other hand, all
-three models outperform the Kaplan-Meier model.
+models. The differences, however, are minimal. All three outperform the
+Kaplan-Meier model.
 
@@
 
-\caption{\label{fig:cs3FinalBrier} Four curves comparing the Cumulative Incidence (A) and the Brier scores (B) as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.}\label{fig:riskregressionBrier}
+\caption{\label{fig:cs3FinalBrier} (A) Four curves comparing the probabilities of death as it changes over follow-up time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
+All follow-up times seen in the new data were used.
+(B) Four curves demonstrating the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
+Lower is better.
+Penalized logistic regression produced the lowest results over follow-up time.}\label{fig:riskregressionBrier}
 \end{figure}
@@
 #> Model1 Model2 Model3 Model4 
-#>    794    760    736    722
+#>    491    459    444    445
 \end{CodeOutput}
@@
 #>  time     
-#>     0 0.0
-#>   365 0.3
+#>     0 0.00
+#>   365 0.72
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> [1] 0.2
+#> [1] 0.65
 \end{CodeOutput}
@@
 
-This report was generated on 2020-07-21 20:19:45 using the following
+This report was generated on 2020-07-28 09:55:47 using the following
 computational environment and dependencies:
@@
 \begin{CodeOutput}
-#> R version 4.0.0 (2020-04-24)
-#> Platform: x86_64-apple-darwin17.0 (64-bit)
-#> Running under: macOS Catalina 10.15.5
+#> R version 4.0.2 (2020-06-22)
+#> Platform: x86_64-pc-linux-gnu (64-bit)
+#> Running under: Pop!_OS 20.04 LTS
 #> 
 #> Matrix products: default
-#> BLAS:   /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRblas.dylib
-#> LAPACK: /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRlapack.dylib
+#> BLAS:   /usr/lib/x86_64-linux-gnu/blas/libblas.so.3.9.0
+#> LAPACK: /usr/lib/x86_64-linux-gnu/lapack/liblapack.so.3.9.0
 #> 
 #> locale:
-#> [1] en_CA.UTF-8/en_CA.UTF-8/en_CA.UTF-8/C/en_CA.UTF-8/en_CA.UTF-8
+#>  [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
+#>  [3] LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
+#>  [5] LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
+#>  [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                 
+#>  [9] LC_ADDRESS=C               LC_TELEPHONE=C            
+#> [11] LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       
 #> 
@@
 #> other attached packages:
-#>  [1] broom_0.5.6               dotwhisker_0.5.0         
-#>  [3] timereg_1.9.4             glue_1.4.1               
-#>  [5] scales_1.1.0              forcats_0.5.0            
-#>  [7] stringr_1.4.0             dplyr_0.8.5              
+#>  [1] broom_0.7.0               dotwhisker_0.5.0         
+#>  [3] timereg_1.9.6             glue_1.4.1               
+#>  [5] scales_1.1.1              forcats_0.5.0            
+#>  [7] stringr_1.4.0             dplyr_1.0.0              
 #>  [9] purrr_0.3.4               readr_1.3.1              
-#> [11] tidyr_1.0.2               ggplot2_3.3.0            
-#> [13] tidyverse_1.3.0           lubridate_1.7.8          
-#> [15] visreg_2.6.1              tibble_3.0.1             
-#> [17] riskRegression_2020.02.05 prodlim_2019.11.13       
-#> [19] flexsurv_1.1.1            survival_3.1-12          
+#> [11] tidyr_1.1.0               ggplot2_3.3.2            
+#> [13] tidyverse_1.3.0           lubridate_1.7.9          
+#> [15] visreg_2.7.0              tibble_3.0.3             
+#> [17] riskRegression_2020.06.25 prodlim_2019.11.13       
+#> [19] flexsurv_1.1.1            survival_3.2-3           
 #> [21] pracma_2.2.9              magrittr_1.5             
-#> [23] kableExtra_1.1.0          glmnet_3.0-2             
-#> [25] Matrix_1.2-18             data.table_1.12.8        
+#> [23] kableExtra_1.1.0          glmnet_4.0-2             
+#> [25] Matrix_1.2-18             data.table_1.13.0        
 #> [27] cowplot_1.0.0             colorspace_1.4-1         
-#> [29] casebase_0.9.0.9000      
+#> [29] casebase_0.9.0           
 #> 
 #> loaded via a namespace (and not attached):
-#>  [1] TH.data_1.0-10      VGAM_1.1-2          ellipsis_0.3.1     
-#>  [4] ggstance_0.3.4      htmlTable_1.13.3    base64enc_0.1-3    
-#>  [7] fs_1.4.1            rstudioapi_0.11     farver_2.0.3       
-#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-0      
+#>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
+#>  [4] ggstance_0.3.4      htmlTable_2.0.1     fs_1.4.2           
+#>  [7] base64enc_0.1-3     rstudioapi_0.11     farver_2.0.3       
+#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-1      
 #> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
-#> [16] Formula_1.2-3       jsonlite_1.6.1      cluster_2.1.0      
-#> [19] dbplyr_1.4.3        png_0.1-7           compiler_4.0.0     
-#> [22] httr_1.4.1          backports_1.1.8     assertthat_0.2.1   
+#> [16] Formula_1.2-3       jsonlite_1.7.0      cluster_2.1.0      
+#> [19] dbplyr_1.4.4        png_0.1-7           compiler_4.0.2     
+#> [22] httr_1.4.2          backports_1.1.8     assertthat_0.2.1   
 #> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
-#> [28] quantreg_5.55       tools_4.0.0         gtable_0.3.0       
-#> [31] Rcpp_1.0.4.6        cellranger_1.1.0    vctrs_0.3.1        
-#> [34] nlme_3.1-147        iterators_1.0.12    xfun_0.15          
-#> [37] rvest_0.3.5         lifecycle_0.2.0     polspline_1.1.17   
-#> [40] muhaz_1.2.6.1       MASS_7.3-51.5       zoo_1.8-7          
-#> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
-#> [46] RColorBrewer_1.1-2  rticles_0.14        yaml_2.2.1         
-#> [49] gridExtra_2.3       rms_5.1-4           rpart_4.1-15       
-#> [52] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
-#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
-#> [58] rlang_0.4.6         pkgconfig_2.0.3     evaluate_0.14      
-#> [61] lattice_0.20-41     labeling_0.3        htmlwidgets_1.5.1  
-#> [64] cmprsk_2.2-9        tidyselect_1.1.0    deSolve_1.28       
-#> [67] R6_2.4.1            generics_0.0.2      Hmisc_4.4-0        
-#> [70] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
-#> [73] pillar_1.4.4        haven_2.2.0         foreign_0.8-78     
-#> [76] mgcv_1.8-31         nnet_7.3-13         mstate_0.2.12      
-#> [79] modelr_0.1.6        crayon_1.3.4        rmarkdown_2.1      
-#> [82] jpeg_0.1-8.1        grid_4.0.0          readxl_1.3.1       
-#> [85] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
-#> [88] numDeriv_2016.8-1.1 stats4_4.0.0        munsell_0.5.0      
-#> [91] viridisLite_0.3.0   quadprog_1.5-8
+#> [28] quantreg_5.61       tools_4.0.2         gtable_0.3.0       
+#> [31] Rcpp_1.0.5          cellranger_1.1.0    vctrs_0.3.2        
+#> [34] nlme_3.1-148        conquer_1.0.1       iterators_1.0.12   
+#> [37] xfun_0.16           rvest_0.3.5         lifecycle_0.2.0    
+#> [40] polspline_1.1.19    muhaz_1.2.6.1       MASS_7.3-51.6      
+#> [43] zoo_1.8-8           hms_0.5.3           sandwich_2.5-1     
+#> [46] SparseM_1.78        RColorBrewer_1.1-2  rticles_0.14       
+#> [49] yaml_2.2.1          gridExtra_2.3       rms_6.0-1          
+#> [52] rpart_4.1-15        latticeExtra_0.6-29 stringi_1.4.6      
+#> [55] foreach_1.5.0       checkmate_2.0.0     lava_1.6.7         
+#> [58] shape_1.4.4         mets_1.2.7.1        rlang_0.4.7        
+#> [61] pkgconfig_2.0.3     matrixStats_0.56.0  evaluate_0.14      
+#> [64] lattice_0.20-41     labeling_0.3        htmlwidgets_1.5.1  
+#> [67] cmprsk_2.2-10       tidyselect_1.1.0    deSolve_1.28       
+#> [70] plyr_1.8.6          R6_2.4.1            generics_0.0.2     
+#> [73] Hmisc_4.4-0         multcomp_1.4-13     DBI_1.1.0          
+#> [76] withr_2.2.0         pillar_1.4.6        haven_2.3.1        
+#> [79] foreign_0.8-80      mgcv_1.8-31         nnet_7.3-14        
+#> [82] mstate_0.2.12       modelr_0.1.8        crayon_1.3.4       
+#> [85] rmarkdown_2.3       jpeg_0.1-8.1        grid_4.0.2         
+#> [88] readxl_1.3.1        blob_1.2.1          reprex_0.3.0       
+#> [91] digest_0.6.25       webshot_0.5.2       numDeriv_2016.8-1.1
+#> [94] stats4_4.0.2        munsell_0.5.0       viridisLite_0.3.0  
+#> [97] quadprog_1.5-8
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> Local:    editMax /Users/maxturgeon/Documents/git_repos/cbpaper
-#> Head:     [2d08bac] 2020-07-20: Merge pull request #22 from sahirbhatnagar/write-popTime
+#> Local:    master /home/jesseislam/Documents/gits/cbpaper
+#> Remote:   master @ origin (git@github.com:Jesse-Islam/cbpaper.git)
+#> Head:     [d30ba8c] 2020-07-22: Merge pull request #23 from sahirbhatnagar/editMax
 \end{CodeOutput}

### 2020-07-28 analysis/paper/paper.Rmd
extended figure width to match text

@@
   comment = "#>",
-  fig.path = "../figures/"
-)
+  fig.path = "../figures/",
+  out.width = "\\textwidth",
+   out.extra = "keepaspectratio=true"
+)
+
 options(kableExtra.latex.load_packages = FALSE)
@@
 eval_cs2 <- T
-eval_cs3 <- TRUE
+eval_cs3 <- T
 eval_cs4 <- T

### 2020-07-28 analysis/paper/paper.tex
extended figure width to match text

@@
 
-{\centering \includegraphics{../figures/plot-stratified-erspc-data-1}
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/plot-stratified-erspc-data-1}
 
@@
 
-{\centering \includegraphics{../figures/erspc-cox-cif-1}
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/erspc-cox-cif-1}
 
@@
 
-{\centering \includegraphics{../figures/compPop-1}
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/compPop-1}
 
@@
 
-{\centering \includegraphics{../figures/bmtcrr-risk-1}
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/bmtcrr-risk-1}
 
@@
 
-{\centering \includegraphics{../figures/brierpt1-1}
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/brierpt1-1}
 
@@
 
-{\centering \includegraphics{../figures/coefplots-1}
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/coefplots-1}
 
@@
 
-{\centering \includegraphics{../figures/riskregressionBrier-1}
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/riskregressionBrier-1}
 
@@
 
-\begin{center}\includegraphics{../figures/stanford-poptime-1} \end{center}
+\begin{center}\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-poptime-1} \end{center}
 
@@
 #> Model1 Model2 Model3 Model4 
-#>    491    459    444    445
+#>    489    457    439    440
 \end{CodeOutput}
@@
 
-\begin{center}\includegraphics{../figures/stanford-hazard-1} \end{center}
+\begin{center}\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-hazard-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{../figures/stanford-risk-1} \end{center}
+\begin{center}\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-risk-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{../figures/stanford-1y-haz-1} \end{center}
+\begin{center}\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-1y-haz-1} \end{center}
 
@@
 #>     0 0.00
-#>   365 0.72
+#>   365 0.74
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> [1] 0.65
+#> [1] 0.67
 \end{CodeOutput}
@@
 
-\begin{center}\includegraphics{../figures/unnamed-chunk-12-1} \end{center}
+\begin{center}\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/unnamed-chunk-12-1} \end{center}
 
@@
 
-This report was generated on 2020-07-28 09:55:47 using the following
+This report was generated on 2020-07-28 10:08:15 using the following
 computational environment and dependencies:
@@
 #> Remote:   master @ origin (git@github.com:Jesse-Islam/cbpaper.git)
-#> Head:     [d30ba8c] 2020-07-22: Merge pull request #23 from sahirbhatnagar/editMax
+#> Head:     [c4450d3] 2020-07-28: Vetted lit-rev, fig and fig comments (cs3)
 \end{CodeOutput}

### 2020-08-04 analysis/paper/paper.Rmd
fixed plot breaks and litrev. names

@@
   \DeclareUnicodeCharacter{2500}{-}
+
 output: 
@@
 
-Several packages implement penalized estimation for the Cox model: \pkg{glmnet} [-@regpathcox], \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche].
+Several packages implement penalized estimation for the Cox model: \pkg{glmnet} [-@regpathcox], \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{riskRegression} [-@gerds_blanche].
 Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \pkg{CoxRidge} [-@perperoglou] and \pkg{rstpm2} [-@clements_liu], while \pkg{survival} [-@survival-package] has an implementation of ridge-penalized estimation.
@@
 Using this estimate, the survival function and the CIF can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below).
-Packages providing this functionality include \pkg{CFC}, \pkg{Flexsurv}, \pkg{mets}, and \pkg{survival}.
+Packages providing this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and \pkg{survival}.
 Our package \pkg{casebase} also follows this approach for both single-event and competing-risk analyses.
@@
 \hline
-\textbf{casebase}        & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
-\textbf{CFC}             & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
-\textbf{cmprsk}             & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\
-\textbf{coxRidge}        &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{crrp}            & x                        &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{fastcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{Flexrsurv}       &                          & x                         &                       & x                & x                   &                          &                                  & x               \\
-\textbf{Flexsurv}        & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\
-\textbf{glmnet}          &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{glmpath}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{mets}            & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
-\textbf{penalized}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{RiskRegression}  & x                         &                           & x                     &                  &                     & x                        &                                  & x                                \\
-\textbf{Rstpm2}          &                          & x                         &                      & x                & x                   & x                        & x                                & x                         \\
-\textbf{SmoothHazard}    &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
-\textbf{Survival}        & x                        & x                         &                       &                  & x                   & x                        & x                                & x                               \\
+\textbf{casebase}        & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\ \bottomrule
+\textbf{CFC}             & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\ \bottomrule
+\textbf{cmprsk}             & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\ \bottomrule
+\textbf{CoxRidge}        &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
+\textbf{crrp}            & x                        &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
+\textbf{fastcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
+\textbf{flexrsurv}       &                          & x                         &                       & x                & x                   &                          &                                  & x               \\ \bottomrule
+\textbf{flexsurv}        & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\ \bottomrule
+\textbf{glmnet}          &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
+\textbf{glmpath}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
+\textbf{mets}            & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\ \bottomrule
+\textbf{penalized}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
+\textbf{riskRegression}  & x                         &                           & x                     &                  &                     & x                        &                                  & x                                \\ \bottomrule
+\textbf{rstpm2}          &                          & x                         &                      & x                & x                   & x                        & x                                & x                         \\ \bottomrule
+\textbf{SmoothHazard}    &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\ \bottomrule
+\textbf{survival}        & x                        & x                         &                       &                  & x                   & x                        & x                                & x                               \\ \bottomrule
 \hline
@@
 ```{r supportData, eval = eval_cs3}
+
+scaleFUNy <- function(x) sprintf("%.2f", x)
+scaleFUNx <- function(x) sprintf("%.0f", x)
 data(support)
@@
   scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
-  paper_gg_theme
+  paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1))) +scale_y_continuous(labels=scaleFUNy,n.breaks=8)
 ```
@@
   scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
-  paper_gg_theme
+  paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1)))+scale_y_continuous(labels=scaleFUNy,n.breaks=8)
 
@@
   geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
-  labs(y = "Probability of Death", x = "Follow-up time (Years)", color = "Models") +
+  labs(y = "Probability of death", x = "Follow-up time (years)", color = "Models") +
   # ggtitle("Absolute risk curves") +
   scale_color_manual(values = q4, breaks = c("K-M","Cox","Penalized Cox","Penalized logistic")) + 
-  paper_gg_theme
+  paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1))) +scale_y_continuous(labels=scaleFUNy,n.breaks=9)
 ```
@@
   geom_line() +
-  xlab("Follow-up time (Years)") +
+  xlab("Follow-up time (years)") +
   ylab("Brier score") +
@@
   paper_gg_theme +  
-  scale_color_manual(values = q4)
+  scale_color_manual(values = q4) + scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1))) +scale_y_continuous(labels=scaleFUNy,n.breaks=8)
 
@@
 This type of censoring is very common in longitudinal studies, and many packages (e.g.
-\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2}) provide functions to account for it.
+\pkg{SmoothHazard}, \pkg{survival} and \pkg{rstpm2}) provide functions to account for it.
 Again, we hope to include any resulting methodology as part of this package.
@@
 In conclusion, we presented the \proglang{R} package \pkg{casebase} which implements case-base sampling for fitting parametric survival models and for estimating smooth cumulative incidence functions using the framework of generalized linear models.
-We strongly believe that its flexibility and its foundation on the familiar logistic regression model will make it appealing to new and established practioners.
+We strongly believe that its flexibility and its foundation on the familiar logistic regression model will make it appealing to new and established practitioners.
 

### 2020-08-04 analysis/paper/paper.tex
fixed plot breaks and litrev. names

@@
 \citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
-\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
+\pkg{riskRegression} \citeyearpar{gerds_blanche}. Moreover, some
 packages also include penalized estimation in the context of Cox models
@@
 below). Packages providing this functionality include \pkg{CFC},
-\pkg{Flexsurv}, \pkg{mets}, and \pkg{survival}. Our package
+\pkg{flexsurv}, \pkg{mets}, and \pkg{survival}. Our package
 \pkg{casebase} also follows this approach for both single-event and
@@
 #> Model1 Model2 Model3 Model4 
-#>    489    457    439    440
+#>    491    459    444    445
 \end{CodeOutput}
@@
 #>     0 0.00
-#>   365 0.74
+#>   365 0.72
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> [1] 0.67
+#> [1] 0.65
 \end{CodeOutput}
@@
 longitudinal studies, and many packages (e.g.~\pkg{SmoothHazard},
-\pkg{survival} and \pkg{Rstpm2}) provide functions to account for it.
+\pkg{survival} and \pkg{rstpm2}) provide functions to account for it.
 Again, we hope to include any resulting methodology as part of this
@@
 flexibility and its foundation on the familiar logistic regression model
-will make it appealing to new and established practioners.
+will make it appealing to new and established practitioners.
 
@@
 
-This report was generated on 2020-07-28 10:08:15 using the following
+This report was generated on 2020-08-04 09:28:27 using the following
 computational environment and dependencies:
@@
 #> [67] cmprsk_2.2-10       tidyselect_1.1.0    deSolve_1.28       
-#> [70] plyr_1.8.6          R6_2.4.1            generics_0.0.2     
-#> [73] Hmisc_4.4-0         multcomp_1.4-13     DBI_1.1.0          
-#> [76] withr_2.2.0         pillar_1.4.6        haven_2.3.1        
-#> [79] foreign_0.8-80      mgcv_1.8-31         nnet_7.3-14        
-#> [82] mstate_0.2.12       modelr_0.1.8        crayon_1.3.4       
-#> [85] rmarkdown_2.3       jpeg_0.1-8.1        grid_4.0.2         
-#> [88] readxl_1.3.1        blob_1.2.1          reprex_0.3.0       
-#> [91] digest_0.6.25       webshot_0.5.2       numDeriv_2016.8-1.1
-#> [94] stats4_4.0.2        munsell_0.5.0       viridisLite_0.3.0  
-#> [97] quadprog_1.5-8
+#> [70] R6_2.4.1            generics_0.0.2      Hmisc_4.4-0        
+#> [73] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
+#> [76] pillar_1.4.6        haven_2.3.1         foreign_0.8-80     
+#> [79] mgcv_1.8-31         nnet_7.3-14         mstate_0.2.12      
+#> [82] modelr_0.1.8        crayon_1.3.4        rmarkdown_2.3      
+#> [85] jpeg_0.1-8.1        grid_4.0.2          readxl_1.3.1       
+#> [88] blob_1.2.1          reprex_0.3.0        digest_0.6.25      
+#> [91] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_4.0.2       
+#> [94] munsell_0.5.0       viridisLite_0.3.0   quadprog_1.5-8
 \end{CodeOutput}

### 2020-08-05 analysis/paper/paper.Rmd
update cs 3&4

--- 2020-07-28 analysis/paper/paper.Rmd
@@
     email: \email{jesse.islam@mail.mcgill.ca}
-  - name: James Hanley
+  - name: James A. Hanley
     affiliation: McGill University
@@
 title:
-  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
+  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
   # If you use tex in the formatted title, also supply version without
-  plain:     "casebase: An Alternative Framework For Survival Analysis"
+  plain:     "casebase: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
   # For running headers, if needed
@@
 eval_implementation <- TRUE
-eval_cs1 <- T
-eval_cs2 <- T
+eval_cs1 <- TRUE
+eval_cs2 <- TRUE
 eval_cs3 <- TRUE
-eval_cs4 <- T
+eval_cs4 <- TRUE
 eval_colophon <- TRUE
@@
 
-Survival analysis has been greatly influenced over the last 50 years by the partial likelihood approach of the Cox proportional hazard model [@cox1972regression].
+Survival analysis and the comparison of event rates has been greatly influenced over the last 50 years by the partial likelihood approach of the Cox proportional hazard model [@cox1972regression].
 This approach provides a flexible way of assessing the influence of covariates on the hazard function, without the need to specify a parametric survival model.
@@
 To recover the whole survival curve---or the cumulative incidence function (CIF)---we then need to separately estimate the baseline hazard [@breslow1972discussion].
-This in turn often leads to stepwise estimates of the survival function.
+This in turn leads to stepwise estimates of the survival function.
 
@@
 Moreover, their framework is very flexible: general functions of time can be estimated (e.g.
-using splines or general additive models), and hence these models retain some of the flexibility of partial likelihood approaches.
+using splines or general additive models), and hence these models retain some of the flexibility of Cox's partial likelihood approach.
 
 In this article, we present an \proglang{R} package that combines the ideas of Hanley & Miettinen into a simple interface.
-The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile.
+The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to compute a patient's risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile.
 Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function.
@@
 
-As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to discretize the study base into an infinite amount of *person moments*.
+As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to sample from the study base a finite amount of *person moments*.
 These person moments are indexed by both an individual in the study and a time point, and therefore each person moment has a covariate profile, an exposure status and an outcome status attached to it.
@@
 To describe the theoretical foundations of case-base sampling, we use the framework of counting processes.
-In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras; the interested reader can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case].
+In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras.
+Instead, following Aalen \emph{et al} [-@aalen2008survival], we use the placeholder $\mathrm{past}$ to denote the past history of the corresponding process.
+The reader interested in more details can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case].
 First, let $N_{i}(t) \in \{0, 1\}$ be counting processes corresponding to the event of interest for individual $i=1, \ldots,n$.
@@
 We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
-$$\lambda_{i}(t) dt = E[dN_{i}(t)].$$
+$$\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].$$
 
@@
 The process $Q_{i}(t) = R_i(t) + N_{i}(t)$ then counts both the case and base series person-moments contributed by individual $i$.
-This process is typically defined by the user via its hazard function $\rho_i(t)$.
-The process $Q_{i}(t)$ is characterized by $E[dQ_{i}(t)] = \lambda_{i}(t)dt + \rho_i(t)dt$.
+This process is typically defined by the user via its intensity function $\rho_i(t)$.
+The process $Q_{i}(t)$ is characterized by $E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt$.
 
 If the hazard function $\lambda_{i}(t; \theta)$ is parametrized in terms of $\theta$, we could define an estimator $\hat{\theta}$ by maximization of the likelihood expression
-$$L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^\tau \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)}\left(1 - \lambda_{i}(t;\theta)\right)^{1 - dN_{i}(t)},$$
-where $\prod_{t\in[0,u)}$ represents a product integral from $0$ to $u$. However, the integral over time makes the computation and maximization of $L_0(\theta)$ challenging.
-
-Case-base sampling allows us to avoid this integral. By conditioning on a sampled person-moment, we get individual contributions of the form
-$$P(dN_{i}(t) \mid dQ_{i}(t) = 1) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.$$
-Therefore, we can define an estimating equation for $\theta$ as follows:
+$$L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^{\min(t_i,\tau)} \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)},$$
+where $\prod_{t\in[0,u)}$ represents a product integral from $0$ to $u$, and where $t_i$ is the event time for individual $i$.
+However, the integral over time makes the computation and maximization of $L_0(\theta)$ challenging.
+
+Case-base sampling allows us to avoid this integral. By conditioning on a sampled person-moment, we get individual likelihood contributions of the form
+$$P(dN_{i}(t) \mid dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.$$
+Therefore, we can define an estimating function for $\theta$ as follows:
 $$L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.$$
@@
 Case-base sampling can also be used in the context of competing-risk analyses.
-Assuming there are $J$ competing events, we can show that each person-moment's contribution of the likelihood is of the form
+Assuming there are $J$ competing events, we can show that each person-moment's contribution to the likelihood is of the form
 
@@
 
-where $N_j(t)$ is the counting process associated with the event of type $j$ and $\lambda_j(t)$ is the corresponding hazard function.
+where $N_j(t)$ is the counting process associated with the event of type $j$ and $\lambda_j(t)$ is the corresponding cause-specific hazard function.
 As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}
@@
 Therefore, our package allows the user to model the effect of time using splines.
-Also, by including interaction terms between covariates and time, it also allows users to fit time-varying coefficient models.
+Also, by including interaction terms between covariates and time, it also allows users to fit (non-proportional) time-varying coefficient models.
 However, we do not explicitly model any shape parameter, unlike \pkg{flexsurv}.
@@
 Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \pkg{CoxRidge} [-@perperoglou] and \pkg{rstpm2} [-@clements_liu], while \pkg{survival} [-@survival-package] has an implementation of ridge-penalized estimation.
-On the other hand, our package \pkg{casebase} provides penalized estimation in the context of parametric hazards.
+On the other hand, our package \pkg{casebase} provides penalized estimation of the hazard function.
 To our knowledge, \pkg{casebase} and \pkg{rsptm2} are the only packages to offer this functionality.
@@
 Next, several \proglang{R} packages implement methodologies for competing risk analysis; for a different perspective on this topic, see Mahani & Sharabiani [-@mahani2015bayesian].
-The package \pkg{cmprsk} provides methods for cause-specific subdistributions, such as in the Fine-Gray model [-@fine1999proportional].
+The package \pkg{cmprsk} provides methods for cause-specific subdistribution hazards, such as in the Fine-Gray model [-@fine1999proportional].
 On the other hand, the package \pkg{CFC} estimates cause-specific CIFs from unadjusted, non-parametric survival functions.
@@
 Packages that implement this approach include \pkg{RiskRegression}, \pkg{rstpm2}, and \pkg{survival}.
-As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category often produces estimates in the form of step-wise functions.
+As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category often produces estimates in the form of stepwise functions.
 Providing smooth estimates of the CIF was one of the main motivations for introducing case-base sampling in survival analysis.
@@
 & \textbf{Competing} & \textbf{Allows} &  \textbf{Penalized}   &    &  & \textbf{Semi} & \textbf{Interval/Left} & \textbf{Absolute}  \\
-\textbf{Package}        & \textbf{Risks}  & \textbf{Non PH} & \textbf{Regression} & \textbf{Splines} & \textbf{Parametric} & \textbf{Parametric} & \textbf{Censoring} & \textbf{Risk}           \\
+\textbf{Package}        & \textbf{Risks}  & \textbf{Non PH} & \textbf{Likelihood} & \textbf{Splines} & \textbf{Parametric} & \textbf{Parametric} & \textbf{Censoring} & \textbf{Risk}           \\
 \hline
@@
 \textbf{coxRidge}        &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{crrp}            & x                        &                           & x                     &                  &                     & x                        &                                  &                                  \\
+\textbf{crrp}            & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
 \textbf{fastcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
@@
 \textbf{penalized}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{RiskRegression}  & x                         &                           & x                     &                  &                     & x                        &                                  & x                                \\
-\textbf{Rstpm2}          &                          & x                         &                      & x                & x                   & x                        & x                                & x                         \\
+\textbf{RiskRegression}  &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
+\textbf{Rstpm2}          &                          & x                         & x                     & x                & x                   & x                        & x                                & x                         \\
 \textbf{SmoothHazard}    &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
@@
 }
-\caption{Comparison of various \proglang{R} packages for survival analysis based on several defining features.
-Competing Risks: whether or not an implementation for competing risks is present.
-Allows Non-PH: permits models that can incorporate variables that change with survival time.
-Penalized Regression: allows for a penalty term when fitting the model.
-This is not restricted to lasso nor ridge.
-Splines: the package permits a flexible fit on time through the use of splines.
-Parametric: the package has implementations for parametric models.
-Semi-parametric: the package has implementations for semi-parametric models.
-Interval/left censoring: the package can interpret interval and left-censoring.
-If this is not selected, the package only handles right-censoring.
-Absolute Risk: the package has an implementation for survival curves, cumulative incidence or cumulative hazard rate functions to be plotted.}
+\caption{Comparison of various \proglang{R} packages for survival analysis based on several defining features.}
 \label{tab:surv-pkgs}
@@
 Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by first adding a layer for the area representing the population time in gray, with subjects having the least amount of observation time plotted at the top of the y-axis.
-We immediately notice a distinctive *step-wise shape* in the population time area.
+We immediately notice a distinctive *stepwise shape* in the population time area.
 This is due to the randomization of the Finnish cohorts which were carried out on January 1 of each of year from 1996 to 1999.
@@
 
-In the case of multiple events, the hazard is fitted via multinomial regression as performed by the \pkg{VGAM} package.
+In the case of multiple competing events, the hazard is fitted via multinomial regression as performed by the \pkg{VGAM} package.
 We selected this package for its ability to fit multinomial regression models with an offset.
@@
 
-$$ S(t) = \exp\left(-\int_0^t \Lambda(u;X) du\right),\qquad \Lambda(t;X) = \sum_{j=1}^J \lambda_j(t;X).$$
+$$ S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right),\qquad \lambda(t;X) = \sum_{j=1}^J \lambda_j(t;X).$$
 From this, we can derive the event-specific subdensities:
@@
 The event of interest is relapse, but other competing causes (e.g.
-transplant-related death) were also recorded.
+death, progression, graft failure, graft-versus-host disease) were also recorded.
 Several covariates were captured at baseline: sex, disease type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL and AML, respectively), disease phase at transplant (Relapse, CR1, CR2, CR3), source of stem cells (bone marrow and peripheral blood, coded as BM+PB, or only peripheral blood, coded as PB), and age.
@@
 
-Our main objective is to compute the absolute risk of relapse for a given set of covariates. We start by fitting a smooth hazard to the data using a linear term for time:
+Our main objective is to compute the cumulative incidence of relapse for a given set of covariates. We start by fitting a smooth hazard to the data using a linear term for time:
 
@@
 We will compare our estimates to that obtained from a corresponding Fine-Gray model [-@fine1999proportional].
-The Fine-Gray model is a semiparametric model for the cause-specific *subdistribution*, i.e.
+The Fine-Gray model is a semiparametric model for the cause-specific *subdistribution hazard*, i.e.
 the function $f_k(t)$ such that
@@
 where $CI_k(t)$ is the cause-specific cumulative incidence.
-The Fine-Gray model allows to directly assess the effect of a covariate on the subdistribution, as opposed to the hazard.
+The Fine-Gray model allows to directly assess the effect of a covariate on the subdistribution hazard, as opposed to the hazard.
 For the computation, we will use the \pkg{timereg} package [@timereg]:
@@
 # 4. Kaplan-Meier
-KM <- survival::coxph(Surv(time = d.time, event = death) ~ 1, data = test,x=TRUE)
+KM <- survival::coxph(Surv(time = d.time, event = death) ~ 1, data = test)
 abKM <- survival::survfit(KM)
@@
 cox_ci <- ggplot(baseCoxComparisonData, aes(Time)) +
-  geom_step(aes(y = APS, colour = "APS")) +
-  geom_step(aes(y = Full, col = "Full")) +
-  geom_step(aes(y = SPS, colour = "SPS")) +
-  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
-  labs(y = "Probability of Death", x = "Follow-up time (Years)", color = "Models") +
+  geom_line(aes(y = APS, colour = "APS"), lwd = 1) +
+  geom_line(aes(y = Full, col = "Full"), lwd = 1) +
+  geom_line(aes(y = SPS, colour = "SPS"), lwd = 1) +
+  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M"), lwd = 1) +
+  labs(y = "Probability of Death", x = "Survival time (Years)", color = "Models") +
   scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
@@
 
-```{r brierpt1, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:brier1} Comparison of three Cox regression models to the unadjusted Kaplan-Meier curve.
-(A) Cumulative Incidence curves.
-For a given model, the plotted curve is the average curve of each individual in the study.
-Each follow-up time seen in the study was used.
-There are minimal differences between the curves.
-APS coincides with the SPS curve, as they are considerably overlapped.(B) Time-dependent Brier scores, where a lower score is better.
-The full model performs best overall."}
+```{r brierpt1, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:brier1} Comparison of three Cox regression models. (A) Cumulative Incidence curves. (B) Time-dependent Brier scores."}
 # Extract time points
@@
                    se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
-brierNull <- Score(list("PredictionModel" = KM), data = test, 
-                   formula = Surv(d.time, death != 0) ~ 1, summary = "ipa", 
-                   se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
 
@@
   data.frame(brierScore = brierSPS$Brier$score$Brier[brierSPS$Brier$score$model == "PredictionModel"], 
-             times = times, model = "SPS"),
-    data.frame(brierScore = brierNull$Brier$score$Brier[brierNull$Brier$score$model == "PredictionModel"], 
-             times = times, model = "K-M")
+             times = times, model = "SPS")
 )
@@
   geom_line() +
-  xlab("Follow-up time (Years)") +
+  xlab("Survival time (Years)") +
   ylab("Brier score") +
@@
   # ggtitle("Brier scores over time") +
-  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
+  scale_color_manual(values = q4[1:3]) + 
   paper_gg_theme
@@
 
-```{r supportCB_fit, eval=eval_cs3, echo=TRUE, cache=FALSE}
+```{r supportCB_fit, eval=eval_cs3, echo=TRUE}
 x <- model.matrix(death ~ . - d.time - aps - sps, 
@@
 
-```{r supportCB_abs, eval=eval_cs3, echo=TRUE, cache=TRUE}
+```{r supportCB_abs, eval=eval_cs3, echo=TRUE}
 # Estimating the absolute risk curve using the newdata parameter
@@
 ```{r coefplots, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models.
-If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity.
-Shrinkage occurs due to penalization."}
+If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity."}
 library(dotwhisker)
@@
 # extract coefficients for lollipop plot
-estimatescoxNet <- data.table::setDT(as.data.frame(coef(coxNet)), 
+estimatescoxNet <- setDT(as.data.frame(coef(coxNet)), 
                          keep.rownames = TRUE)
@@
 
-estimatescb <- data.table::setDT(as.data.frame(coef(cbFull)[-c(1, 2), 1]), 
+estimatescb <- setDT(as.data.frame(coef(cbFull)[-c(1, 2), 1]), 
                      keep.rownames = TRUE)
@@
 
-estimatescoxFull <- data.table::setDT(as.data.frame(coef(coxFull)), 
+estimatescoxFull <- setDT(as.data.frame(coef(coxFull)), 
                           keep.rownames = TRUE)
@@
 In Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our penalized logistic regression model compared to both Cox regression models.
-The differences, however, are minimal.
-All three outperform the Kaplan-Meier model.
+However, this difference is minimal.
+On the other hand, all three models outperform the Kaplan-Meier model.
 
@@
 reg_ci <- ggplot(ab, aes(Time)) +
-  geom_step(data = baseCoxComparisonData, mapping = aes(x = Time,y = Full, colour = "Cox")) +
-  geom_line(aes(y = absRisk, colour = Model)) +
-  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
-  labs(y = "Probability of Death", x = "Follow-up time (Years)", color = "Models") +
+  geom_line(data = baseCoxComparisonData, mapping = aes(y = Full, colour = "Cox"), lwd = 1) +
+  geom_line(aes(y = absRisk, colour = Model), lwd = 1) +
+  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M"), lwd = 1) +
+  labs(y = "Probability of Death", x = "Survival time (Years)", color = "Models") +
   # ggtitle("Absolute risk curves") +
-  scale_color_manual(values = q4, breaks = c("K-M","Cox","Penalized Cox","Penalized logistic")) + 
+  scale_color_manual(values = q4, breaks = c("Cox", "Penalized Cox", "Penalized logistic", "K-M")) + 
   paper_gg_theme
@@
 
-```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} (A) Four curves comparing the probabilities of death as it changes over follow-up time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
-All follow-up times seen in the new data were used.
-(B) Four curves demonstrating the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
-Lower is better.
-Penalized logistic regression produced the lowest results over follow-up time.", warning = FALSE}
+```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} Four curves comparing the Cumulative Incidence (A) and the Brier scores (B) as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.", warning = FALSE}
 # Compute Brier score for penalized models
@@
                            se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
-#brierFinalResults$models<-as.character(brierFinalResults$models)
+
 # Combine Brier scores between unpenalized and penalized models before plotting
-
 data_brier <- bind_rows(
@@
   filter(brierFull$Brier$score, model == "PredictionModel")
-  )
-data_brier$model<-factor(data_brier$model,c("Null model", "PredictionModel","Penalized Cox", "Penalized Logistic"))
-levels(data_brier$model)<-c("Kaplan-Meier","Cox","Penalized Cox","Penalized logistic")
+  ) %>% 
+  mutate(model = as.character(model), # First return to string, so types match
+         model = case_when(
+    model == "Null model" ~ "K-M",
+    model == "PredictionModel" ~ "Cox",
+    TRUE ~ model
+  ),
+  model = factor(model, levels = c("Cox", "Penalized Cox", "Penalized Logistic", "K-M"),
+                 labels = c("Cox", "Pen. Cox", "Pen. Logistic", "K-M")))
 
@@
 As before, we can look at the population-time plot for a graphical summary of the event incidence.
+Here, we coloured the exposed person-time (i.e.
+after transplant) in a darker shade of gray.
 As we can see, most events occur early during the follow-up period, and therefore we do not expect the hazard to be constant.
@@
 )
+# plot(stanford_popTime,
+#      ribbon.params = list(fill = "grey50")) + 
+#   paper_gg_theme
+
+# Add exposure time----
+tx_exp <- stanford_popTime %>% 
+    filter(!is.na(tx.date)) %>% 
+    mutate(x = time_length(accept.dt %--% tx.date,
+                           unit = "days"))
+
 plot(stanford_popTime,
      ribbon.params = list(fill = "grey50")) + 
+    geom_segment(data = tx_exp,
+                 aes(x = x, xend = original.time,
+                     y = ycoord, yend = ycoord),
+                 colour = "grey30", size = 1) + 
+    geom_point(data = filter(stanford_popTime, event == 1),
+               aes(x = time, y = yc, color = "Case series",
+                   fill = "Case series"),
+               size = 1,
+               alpha = 1,
+               shape = 21) +
   paper_gg_theme
@@
 
-```{r, eval = eval_cs4}
+```{r stanford_cb, eval = eval_cs4}
 cb_data <- sampleCaseBase(jasa,
   time = "futime",
-  event = "fustat", ratio = 10
+  event = "fustat", ratio = 100
 )
@@
 
-```{r, eval = eval_cs4}
+```{r stanford_exp, eval = eval_cs4, dependson="stanford_cb"}
 # Define exposure variable
@@
 
-```{r, eval = eval_cs4, warning = FALSE}
-library(splines)
+```{r stanford_fit, eval = eval_cs4, warning = FALSE, echo=TRUE, dependson="stanford_exp"}
 # Fit several models
@@
 )
-fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
+fit3 <- fitSmoothHazard(fustat ~ exposure * futime,
   data = cb_data, time = "futime"
 )
-fit4 <- fitSmoothHazard(fustat ~ exposure * bs(futime),
-  data = cb_data, time = "futime"
-)
-```
-
-Note that the fourth model includes an interaction term between exposure and follow-up time.
+```
+
+Note that the third model includes an interaction term between exposure and follow-up time.
 In other words, this model no longer exhibit proportional hazards.
@@
 
-```{r, eval = eval_cs4}
+```{r, eval = eval_cs4, dependson="stanford_fit"}
 # Compute AIC
@@
   "Model2" = AIC(fit2),
-  "Model3" = AIC(fit3),
-  "Model4" = AIC(fit4))
-```
-
-As we can, the best fit is the fourth model. By visualizing the hazard functions for both exposed and unexposed individuals, we can more clearly see how the hazards are no longer proportional.
-
-```{r stanford-hazard, eval = eval_cs4, echo=TRUE}
+  "Model3" = AIC(fit3))
+```
+
+As we can see, the best fit is the third model. By visualizing the hazard functions for both exposed and unexposed individuals, we can more clearly see how the hazards are no longer proportional.
+
+```{r stanford-hazard, eval = eval_cs4, echo=TRUE, dependson="stanford_fit"}
 # Compute hazards---
@@
 # transform to the right scale
-hazard_data$hazard <- exp(predict(fit4,
+hazard_data$hazard <- exp(predict(fit3,
   newdata = hazard_data,
@@
 hazard_data$Status <- factor(hazard_data$exposure,
-  labels = c("NoTrans", "Trans")
+  labels = c("No Tx", "Tx")
 )
@@
   ylab("Hazard") +
-  xlab("Follow-up time")
+  xlab("Follow-up time (in days)")
 ```
@@
 
-```{r stanford-risk, eval = eval_cs4}
+```{r stanford-risk, eval = eval_cs4, dependson="stanford_fit"}
 # Compute absolute risk curves
 newdata <- data.frame(exposure = c(0, 1))
-absrisk <- absoluteRisk(fit4,
+absrisk <- absoluteRisk(fit3,
   newdata = newdata,
@@
 plot(absrisk,
-     id.names = c("NoTrans","Trans")) + 
+     id.names = c("No Tx", "Tx")) + 
+  ylab("Cumulative Incidence") +
+  xlab("Follow-up time (in days)") +
   paper_gg_theme
@@
 
-Note that we can easily adapt the code above to the situation where a patient receives a heart transplant at a point in time of interest, for example after 30 days.
-
-```{r stanford-1y-haz, eval = eval_cs4}
+<!-- Note that we can easily adapt the code above to the situation where a patient receives a heart transplant at a point in time of interest, for example after 1 month. -->
+
+```{r stanford-1y-haz, eval = FALSE, echo = FALSE}
 # Compute hazards---
@@
   offset = 0,
-  exposure = if_else(futime < 30, 0, 1)
+  exposure = if_else(futime < 1, 0, 1)
 )
@@
   ylab("Hazard") +
-  xlab("Follow-up time") +
-  geom_vline(xintercept = 30, linetype = "dashed")
-```
-
-We can then compare the 1-year mortality risk without transplant and with transplant at 30 days.
-
-```{r stanford-1y-risk, eval = eval_cs4}
+  xlab("Follow-up time (in days)") +
+  geom_vline(xintercept = 1, linetype = "dashed")
+```
+
+<!-- We can then compare the 1-year mortality risk without transplant and with transplant at 1 month. -->
+
+```{r stanford-1y-risk, eval = FALSE, echo = FALSE}
 absoluteRisk(fit4,
@@
 
-As we can see, the risk estimate at 1-year is about 30% lower if the patient receives a heart transplant at 30 days.
+<!-- As we can see, the risk estimate at 1-year is about 30% lower if the patient receives a heart transplant at 1 month. -->
 
 We can also plot the hazard ratio:
-```{r eval = eval_cs4}
+
+```{r, eval = eval_cs4, dependson="stanford_fit"}
 newtime <- seq(0, 425, length.out = 100)
@@
 
-plot(fit4,
+plot(fit3,
      type = "hr",
@@
      xvar = "futime",
+     xlab = "Follow-up time (in days)",
      ci = FALSE,
@@
 ```
+
+As we can see in the above case-study, the \pkg{casebase} package can also be used to model time-varying exposures and non-proportional hazard functions.
 
@@
 However, in terms of output, both approaches differ significantly.
-Case-base sampling produces smooth hazards and smooth cumulative incidence curves, whereas Cox regression and Fine-Gray models produce step-wise CIFs and never explicitly model the hazard function.
+Case-base sampling produces smooth hazards and smooth cumulative incidence curves, whereas Cox regression and Fine-Gray models produce stepwise CIFs and never explicitly model the hazard function.
 Qualitatively, we showed that by using splines in the linear predictor, all three models yielded similar curves.

### 2020-08-05 analysis/paper/paper.tex
update cs 3&4

--- 2020-07-28 analysis/paper/paper.tex
@@
 \author{
-Sahir Rai Bhatnagar*\\McGill University \And Maxime Turgeon*\\University of Manitoba \AND Jesse Islam\\McGill University \And James Hanley\\McGill University \And Olli Saarela\\University of Toronto
+Sahir Rai Bhatnagar*\\McGill University \And Maxime Turgeon*\\University
+of Manitoba \AND Jesse Islam\\McGill University \And James A.
+Hanley\\McGill University \And Olli Saarela\\University of Toronto
 }
-\title{\pkg{casebase}: An Alternative Framework For Survival Analysis}
-
-\Plainauthor{Sahir Rai Bhatnagar*, Maxime Turgeon*, Jesse Islam, James Hanley, Olli Saarela}
-\Plaintitle{casebase: An Alternative Framework For Survival Analysis}
-\Shorttitle{\pkg{casebase}: An Alternative Framework For Survival Analysis}
+\title{\pkg{casebase}: An Alternative Framework For Survival Analysis
+and Comparison of Event Rates}
+
+\Plainauthor{Sahir Rai Bhatnagar*, Maxime Turgeon*, Jesse Islam, James
+A. Hanley, Olli Saarela}
+\Plaintitle{casebase: An Alternative Framework For Survival Analysis and
+Comparison of Event Rates}
+\Shorttitle{\pkg{casebase}: An Alternative Framework For Survival
+Analysis}
 
@@
 
-      James Hanley\\
+      James A. Hanley\\
   McGill University\\
@@
 
-Survival analysis has been greatly influenced over the last 50 years by
-the partial likelihood approach of the Cox proportional hazard model
-\citep{cox1972regression}. This approach provides a flexible way of
-assessing the influence of covariates on the hazard function, without
-the need to specify a parametric survival model. This flexibility comes
-at the cost of decoupling the baseline hazard from the effect of the
-covariates. To recover the whole survival curve---or the cumulative
-incidence function (CIF)---we then need to separately estimate the
-baseline hazard \citep{breslow1972discussion}. This in turn often leads
-to stepwise estimates of the survival function.
+Survival analysis and the comparison of event rates has been greatly
+influenced over the last 50 years by the partial likelihood approach of
+the Cox proportional hazard model \citep{cox1972regression}. This
+approach provides a flexible way of assessing the influence of
+covariates on the hazard function, without the need to specify a
+parametric survival model. This flexibility comes at the cost of
+decoupling the baseline hazard from the effect of the covariates. To
+recover the whole survival curve---or the cumulative incidence function
+(CIF)---we then need to separately estimate the baseline hazard
+\citep{breslow1972discussion}. This in turn leads to stepwise estimates
+of the survival function.
 
@@
 can be estimated (e.g.~using splines or general additive models), and
-hence these models retain some of the flexibility of partial likelihood
-approaches.
+hence these models retain some of the flexibility of Cox's partial
+likelihood approach.
 
@@
 \pkg{casebase} package is to provide practitioners with an easy-to-use
-software tool to predict the risk (or cumulative incidence) of an event,
-conditional on a particular patient's covariate profile. Our package
-retains the flexibility of case-base sampling and the familiar interface
-of the \code{glm} function. In addition, we provide extensive
+software tool to compute a patient's risk (or cumulative incidence) of
+an event, conditional on a particular patient's covariate profile. Our
+package retains the flexibility of case-base sampling and the familiar
+interface of the \code{glm} function. In addition, we provide extensive
 visualization tools.
@@
 As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
-key idea behind case-base sampling is to discretize the study base into
-an infinite amount of \emph{person moments}. These person moments are
-indexed by both an individual in the study and a time point, and
-therefore each person moment has a covariate profile, an exposure status
-and an outcome status attached to it. We note that there is only a
-finite number of person moments associated with the event of interest
-(what Hanley \& Miettinen call the \emph{case series}). Case-base
-sampling refers to the sampling from the base of a representative finite
-sample called the \emph{base series}.
+key idea behind case-base sampling is to sample from the study base a
+finite amount of \emph{person moments}. These person moments are indexed
+by both an individual in the study and a time point, and therefore each
+person moment has a covariate profile, an exposure status and an outcome
+status attached to it. We note that there is only a finite number of
+person moments associated with the event of interest (what Hanley \&
+Miettinen call the \emph{case series}). Case-base sampling refers to the
+sampling from the base of a representative finite sample called the
+\emph{base series}.
 
@@
 the framework of counting processes. In what follows, we abuse notation
-slightly and omit any mention of \(\sigma\)-algebras; the interested
-reader can refer to Saarela \& Arjas \citeyearpar{saarela2015non} and
-Saarela \citeyearpar{saarela2016case}. First, let
-\(N_{i}(t) \in \{0, 1\}\) be counting processes corresponding to the
-event of interest for individual \(i=1, \ldots,n\). For simplicity, we
-will consider Type I censoring due to the end of follow-up at time
-\(\tau\) (the general case of non-informative censoring is treated in
-Saarela \citeyearpar{saarela2016case}). We assume a continuous time
-model, which implies that the counting process jumps are less than or
-equal to one. We are interested in modeling the hazard functions
-\(\lambda_{i}(t)\) of the processes \(N_i(t)\), and which satisfy
-\[\lambda_{i}(t) dt = E[dN_{i}(t)].\]
+slightly and omit any mention of \(\sigma\)-algebras. Instead, following
+Aalen \emph{et al} \citeyearpar{aalen2008survival}, we use the
+placeholder \(\mathrm{past}\) to denote the past history of the
+corresponding process. The reader interested in more details can refer
+to Saarela \& Arjas \citeyearpar{saarela2015non} and Saarela
+\citeyearpar{saarela2016case}. First, let \(N_{i}(t) \in \{0, 1\}\) be
+counting processes corresponding to the event of interest for individual
+\(i=1, \ldots,n\). For simplicity, we will consider Type I censoring due
+to the end of follow-up at time \(\tau\) (the general case of
+non-informative censoring is treated in Saarela
+\citeyearpar{saarela2016case}). We assume a continuous time model, which
+implies that the counting process jumps are less than or equal to one.
+We are interested in modeling the hazard functions \(\lambda_{i}(t)\) of
+the processes \(N_i(t)\), and which satisfy
+\[\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].\]
 
@@
 base series person-moments contributed by individual \(i\). This process
-is typically defined by the user via its hazard function \(\rho_i(t)\).
-The process \(Q_{i}(t)\) is characterized by
-\(E[dQ_{i}(t)] = \lambda_{i}(t)dt + \rho_i(t)dt\).
+is typically defined by the user via its intensity function
+\(\rho_i(t)\). The process \(Q_{i}(t)\) is characterized by
+\(E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt\).
 
@@
 maximization of the likelihood expression
-\[L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^\tau \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)}\left(1 - \lambda_{i}(t;\theta)\right)^{1 - dN_{i}(t)},\]
+\[L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^{\min(t_i,\tau)} \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)},\]
 where \(\prod_{t\in[0,u)}\) represents a product integral from \(0\) to
-\(u\). However, the integral over time makes the computation and
-maximization of \(L_0(\theta)\) challenging.
+\(u\), and where \(t_i\) is the event time for individual \(i\).
+However, the integral over time makes the computation and maximization
+of \(L_0(\theta)\) challenging.
 
 Case-base sampling allows us to avoid this integral. By conditioning on
-a sampled person-moment, we get individual contributions of the form
-\[P(dN_{i}(t) \mid dQ_{i}(t) = 1) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.\]
-Therefore, we can define an estimating equation for \(\theta\) as
+a sampled person-moment, we get individual likelihood contributions of
+the form
+\[P(dN_{i}(t) \mid dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.\]
+Therefore, we can define an estimating function for \(\theta\) as
 follows:
@@
 analyses. Assuming there are \(J\) competing events, we can show that
-each person-moment's contribution of the likelihood is of the form
+each person-moment's contribution to the likelihood is of the form
 
@@
 where \(N_j(t)\) is the counting process associated with the event of
-type \(j\) and \(\lambda_j(t)\) is the corresponding hazard function. As
-may be expected, this functional form is similar to the terms appearing
-in the likelihood function for multinomial
+type \(j\) and \(\lambda_j(t)\) is the corresponding cause-specific
+hazard function. As may be expected, this functional form is similar to
+the terms appearing in the likelihood function for multinomial
 regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}
@@
 splines. Also, by including interaction terms between covariates and
-time, it also allows users to fit time-varying coefficient models.
-However, we do not explicitly model any shape parameter, unlike
-\pkg{flexsurv}.
+time, it also allows users to fit (non-proportional) time-varying
+coefficient models. However, we do not explicitly model any shape
+parameter, unlike \pkg{flexsurv}.
 
@@
 estimation. On the other hand, our package \pkg{casebase} provides
-penalized estimation in the context of parametric hazards. To our
-knowledge, \pkg{casebase} and \pkg{rsptm2} are the only packages to
-offer this functionality.
+penalized estimation of the hazard function. To our knowledge,
+\pkg{casebase} and \pkg{rsptm2} are the only packages to offer this
+functionality.
 
@@
 Mahani \& Sharabiani \citeyearpar{mahani2015bayesian}. The package
-\pkg{cmprsk} provides methods for cause-specific subdistributions, such
-as in the Fine-Gray model \citeyearpar{fine1999proportional}. On the
-other hand, the package \pkg{CFC} estimates cause-specific CIFs from
-unadjusted, non-parametric survival functions. Our package
-\pkg{casebase} also provides functionalities for competing risk analysis
-by estimating parametrically the cause-specific hazards. From these
-quantities, we can then estimate the cause-specific CIFs.
+\pkg{cmprsk} provides methods for cause-specific subdistribution
+hazards, such as in the Fine-Gray model
+\citeyearpar{fine1999proportional}. On the other hand, the package
+\pkg{CFC} estimates cause-specific CIFs from unadjusted, non-parametric
+survival functions. Our package \pkg{casebase} also provides
+functionalities for competing risk analysis by estimating parametrically
+the cause-specific hazards. From these quantities, we can then estimate
+the cause-specific CIFs.
 
@@
 smooth estimates of the cumulative incidence function, whereas the
-second category often produces estimates in the form of step-wise
+second category often produces estimates in the form of stepwise
 functions. Providing smooth estimates of the CIF was one of the main
@@
 at the top of the y-axis. We immediately notice a distinctive
-\emph{step-wise shape} in the population time area. This is due to the
+\emph{stepwise shape} in the population time area. This is due to the
 randomization of the Finnish cohorts which were carried out on January 1
@@
 
-In the case of multiple events, the hazard is fitted via multinomial
-regression as performed by the \pkg{VGAM} package. We selected this
-package for its ability to fit multinomial regression models with an
-offset.
+In the case of multiple competing events, the hazard is fitted via
+multinomial regression as performed by the \pkg{VGAM} package. We
+selected this package for its ability to fit multinomial regression
+models with an offset.
 
@@
 
-\[ S(t) = \exp\left(-\int_0^t \Lambda(u;X) du\right),\qquad \Lambda(t;X) = \sum_{j=1}^J \lambda_j(t;X).\]
+\[ S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right),\qquad \lambda(t;X) = \sum_{j=1}^J \lambda_j(t;X).\]
 From this, we can derive the event-specific subdensities:
@@
 transplant for acute leukemia. The event of interest is relapse, but
-other competing causes (e.g.~transplant-related death) were also
-recorded. Several covariates were captured at baseline: sex, disease
-type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL
-and AML, respectively), disease phase at transplant (Relapse, CR1, CR2,
-CR3), source of stem cells (bone marrow and peripheral blood, coded as
-BM+PB, or only peripheral blood, coded as PB), and age.
+other competing causes (e.g.~death, progression, graft failure,
+graft-versus-host disease) were also recorded. Several covariates were
+captured at baseline: sex, disease type (acute lymphoblastic or
+myeloblastic leukemia, abbreviated as ALL and AML, respectively),
+disease phase at transplant (Relapse, CR1, CR2, CR3), source of stem
+cells (bone marrow and peripheral blood, coded as BM+PB, or only
+peripheral blood, coded as PB), and age.
 
@@
 
-Our main objective is to compute the absolute risk of relapse for a
-given set of covariates. We start by fitting a smooth hazard to the data
-using a linear term for time:
+Our main objective is to compute the cumulative incidence of relapse for
+a given set of covariates. We start by fitting a smooth hazard to the
+data using a linear term for time:
 
@@
 Fine-Gray model \citeyearpar{fine1999proportional}. The Fine-Gray model
-is a semiparametric model for the cause-specific \emph{subdistribution},
-i.e.~the function \(f_k(t)\) such that
+is a semiparametric model for the cause-specific \emph{subdistribution
+hazard}, i.e.~the function \(f_k(t)\) such that
 \[CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),\] where
@@
 model allows to directly assess the effect of a covariate on the
-subdistribution, as opposed to the hazard. For the computation, we will
-use the \pkg{timereg} package \citep{timereg}:
+subdistribution hazard, as opposed to the hazard. For the computation,
+we will use the \pkg{timereg} package \citep{timereg}:
 
@@
 
-\caption{\label{fig:brier1} Comparison of three Cox regression models to the unadjusted Kaplan-Meier curve.
-(A) Cumulative Incidence curves.
-For a given model, the plotted curve is the average curve of each individual in the study.
-Each follow-up time seen in the study was used.
-There are minimal differences between the curves.
-APS coincides with the SPS curve, as they are considerably overlapped.(B) Time-dependent Brier scores, where a lower score is better.
-The full model performs best overall.}\label{fig:brierpt1}
+\caption{\label{fig:brier1} Comparison of three Cox regression models. (A) Cumulative Incidence curves. (B) Time-dependent Brier scores.}\label{fig:brierpt1}
 \end{figure}
@@
 \caption{\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models.
-If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity.
-Shrinkage occurs due to penalization.}\label{fig:coefplots}
+If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity.}\label{fig:coefplots}
 \end{figure}
@@
 penalized logistic regression model compared to both Cox regression
-models. The differences, however, are minimal. All three outperform the
-Kaplan-Meier model.
+models. However, this difference is minimal. On the other hand, all
+three models outperform the Kaplan-Meier model.
 
@@
 
-\caption{\label{fig:cs3FinalBrier} (A) Four curves comparing the probabilities of death as it changes over follow-up time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
-All follow-up times seen in the new data were used.
-(B) Four curves demonstrating the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
-Lower is better.
-Penalized logistic regression produced the lowest results over follow-up time.}\label{fig:riskregressionBrier}
+\caption{\label{fig:cs3FinalBrier} Four curves comparing the Cumulative Incidence (A) and the Brier scores (B) as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.}\label{fig:riskregressionBrier}
 \end{figure}
@@
 As before, we can look at the population-time plot for a graphical
-summary of the event incidence. As we can see, most events occur early
-during the follow-up period, and therefore we do not expect the hazard
-to be constant.
+summary of the event incidence. Here, we coloured the exposed
+person-time (i.e.~after transplant) in a darker shade of gray. As we can
+see, most events occur early during the follow-up period, and therefore
+we do not expect the hazard to be constant.
 
@@
 
-Note that the fourth model includes an interaction term between exposure
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Fit several models
+R> fit1 <- fitSmoothHazard(fustat ~ exposure,
+R+   data = cb_data, time = "futime"
+R+ )
+R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
+R+   data = cb_data, time = "futime"
+R+ )
+R> fit3 <- fitSmoothHazard(fustat ~ exposure * futime,
+R+   data = cb_data, time = "futime"
+R+ )
+\end{CodeInput}
+\end{CodeChunk}
+
+Note that the third model includes an interaction term between exposure
 and follow-up time. In other words, this model no longer exhibit
@@
 \begin{CodeOutput}
-#> Model1 Model2 Model3 Model4 
-#>    491    459    444    445
+#> Model1 Model2 Model3 
+#>    827    790    788
 \end{CodeOutput}
@@
 
-As we can, the best fit is the fourth model. By visualizing the hazard
-functions for both exposed and unexposed individuals, we can more
+As we can see, the best fit is the third model. By visualizing the
+hazard functions for both exposed and unexposed individuals, we can more
 clearly see how the hazards are no longer proportional.
@@
 R> # transform to the right scale
-R> hazard_data$hazard <- exp(predict(fit4,
+R> hazard_data$hazard <- exp(predict(fit3,
 R+   newdata = hazard_data,
@@
 R> hazard_data$Status <- factor(hazard_data$exposure,
-R+   labels = c("NoTrans", "Trans")
+R+   labels = c("No Tx", "Tx")
 R+ )
@@
 R+   ylab("Hazard") +
-R+   xlab("Follow-up time")
+R+   xlab("Follow-up time (in days)")
 \end{CodeInput}
@@
 
-Note that we can easily adapt the code above to the situation where a
-patient receives a heart transplant at a point in time of interest, for
-example after 30 days.
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics{../figures/stanford-1y-haz-1} \end{center}
-
-\end{CodeChunk}
-
-We can then compare the 1-year mortality risk without transplant and
-with transplant at 30 days.
-
-\begin{CodeChunk}
-
-\begin{CodeOutput}
-#>  time     
-#>     0 0.00
-#>   365 0.72
-\end{CodeOutput}
-
-\begin{CodeOutput}
-#> [1] 0.65
-\end{CodeOutput}
-\end{CodeChunk}
-
-As we can see, the risk estimate at 1-year is about 30\% lower if the
-patient receives a heart transplant at 30 days.
-
 We can also plot the hazard ratio:
@@
 
-\begin{center}\includegraphics{../figures/unnamed-chunk-12-1} \end{center}
-
-\end{CodeChunk}
+\begin{center}\includegraphics{../figures/unnamed-chunk-9-1} \end{center}
+
+\end{CodeChunk}
+
+As we can see in the above case-study, the \pkg{casebase} package can
+also be used to model time-varying exposures and non-proportional hazard
+functions.
 
@@
 smooth cumulative incidence curves, whereas Cox regression and Fine-Gray
-models produce step-wise CIFs and never explicitly model the hazard
+models produce stepwise CIFs and never explicitly model the hazard
 function. Qualitatively, we showed that by using splines in the linear
@@
 
-This report was generated on 2020-07-28 09:55:47 using the following
+This report was generated on 2020-08-05 11:18:44 using the following
 computational environment and dependencies:
@@
 #> R version 4.0.2 (2020-06-22)
-#> Platform: x86_64-pc-linux-gnu (64-bit)
-#> Running under: Pop!_OS 20.04 LTS
+#> Platform: x86_64-apple-darwin17.0 (64-bit)
+#> Running under: macOS Catalina 10.15.6
 #> 
 #> Matrix products: default
-#> BLAS:   /usr/lib/x86_64-linux-gnu/blas/libblas.so.3.9.0
-#> LAPACK: /usr/lib/x86_64-linux-gnu/lapack/liblapack.so.3.9.0
+#> BLAS:   /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRblas.dylib
+#> LAPACK: /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRlapack.dylib
 #> 
 #> locale:
-#>  [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
-#>  [3] LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
-#>  [5] LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
-#>  [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                 
-#>  [9] LC_ADDRESS=C               LC_TELEPHONE=C            
-#> [11] LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       
+#> [1] en_CA.UTF-8/en_CA.UTF-8/en_CA.UTF-8/C/en_CA.UTF-8/en_CA.UTF-8
 #> 
@@
 #> other attached packages:
-#>  [1] broom_0.7.0               dotwhisker_0.5.0         
-#>  [3] timereg_1.9.6             glue_1.4.1               
-#>  [5] scales_1.1.1              forcats_0.5.0            
-#>  [7] stringr_1.4.0             dplyr_1.0.0              
-#>  [9] purrr_0.3.4               readr_1.3.1              
-#> [11] tidyr_1.1.0               ggplot2_3.3.2            
-#> [13] tidyverse_1.3.0           lubridate_1.7.9          
-#> [15] visreg_2.7.0              tibble_3.0.3             
-#> [17] riskRegression_2020.06.25 prodlim_2019.11.13       
-#> [19] flexsurv_1.1.1            survival_3.2-3           
-#> [21] pracma_2.2.9              magrittr_1.5             
-#> [23] kableExtra_1.1.0          glmnet_4.0-2             
-#> [25] Matrix_1.2-18             data.table_1.13.0        
-#> [27] cowplot_1.0.0             colorspace_1.4-1         
-#> [29] casebase_0.9.0           
+#>  [1] scales_1.1.1              forcats_0.5.0            
+#>  [3] stringr_1.4.0             dplyr_1.0.0              
+#>  [5] purrr_0.3.4               readr_1.3.1              
+#>  [7] tidyr_1.0.2               ggplot2_3.3.2            
+#>  [9] tidyverse_1.3.0           lubridate_1.7.8          
+#> [11] visreg_2.6.1              tibble_3.0.3             
+#> [13] riskRegression_2020.02.05 prodlim_2019.11.13       
+#> [15] flexsurv_1.1.1            survival_3.1-12          
+#> [17] pracma_2.2.9              magrittr_1.5             
+#> [19] kableExtra_1.1.0          glmnet_3.0-2             
+#> [21] Matrix_1.2-18             data.table_1.12.8        
+#> [23] cowplot_1.0.0             colorspace_1.4-1         
+#> [25] casebase_0.9.0.9000      
 #> 
 #> loaded via a namespace (and not attached):
-#>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
-#>  [4] ggstance_0.3.4      htmlTable_2.0.1     fs_1.4.2           
-#>  [7] base64enc_0.1-3     rstudioapi_0.11     farver_2.0.3       
-#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-1      
-#> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
-#> [16] Formula_1.2-3       jsonlite_1.7.0      cluster_2.1.0      
-#> [19] dbplyr_1.4.4        png_0.1-7           compiler_4.0.2     
-#> [22] httr_1.4.2          backports_1.1.8     assertthat_0.2.1   
-#> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
-#> [28] quantreg_5.61       tools_4.0.2         gtable_0.3.0       
+#>  [1] TH.data_1.0-10      VGAM_1.1-2          ellipsis_0.3.1     
+#>  [4] htmlTable_1.13.3    base64enc_0.1-3     fs_1.4.1           
+#>  [7] rstudioapi_0.11     MatrixModels_0.4-1  fansi_0.4.1        
+#> [10] mvtnorm_1.1-0       xml2_1.3.2          codetools_0.2-16   
+#> [13] knitr_1.29          Formula_1.2-3       jsonlite_1.6.1     
+#> [16] broom_0.5.6         cluster_2.1.0       dbplyr_1.4.3       
+#> [19] png_0.1-7           compiler_4.0.2      httr_1.4.1         
+#> [22] backports_1.1.8     assertthat_0.2.1    cli_2.0.2          
+#> [25] acepack_1.4.1       htmltools_0.5.0     quantreg_5.55      
+#> [28] tools_4.0.2         gtable_0.3.0        glue_1.4.1         
 #> [31] Rcpp_1.0.5          cellranger_1.1.0    vctrs_0.3.2        
-#> [34] nlme_3.1-148        conquer_1.0.1       iterators_1.0.12   
-#> [37] xfun_0.16           rvest_0.3.5         lifecycle_0.2.0    
-#> [40] polspline_1.1.19    muhaz_1.2.6.1       MASS_7.3-51.6      
-#> [43] zoo_1.8-8           hms_0.5.3           sandwich_2.5-1     
-#> [46] SparseM_1.78        RColorBrewer_1.1-2  rticles_0.14       
-#> [49] yaml_2.2.1          gridExtra_2.3       rms_6.0-1          
-#> [52] rpart_4.1-15        latticeExtra_0.6-29 stringi_1.4.6      
-#> [55] foreach_1.5.0       checkmate_2.0.0     lava_1.6.7         
-#> [58] shape_1.4.4         mets_1.2.7.1        rlang_0.4.7        
-#> [61] pkgconfig_2.0.3     matrixStats_0.56.0  evaluate_0.14      
-#> [64] lattice_0.20-41     labeling_0.3        htmlwidgets_1.5.1  
-#> [67] cmprsk_2.2-10       tidyselect_1.1.0    deSolve_1.28       
-#> [70] plyr_1.8.6          R6_2.4.1            generics_0.0.2     
-#> [73] Hmisc_4.4-0         multcomp_1.4-13     DBI_1.1.0          
-#> [76] withr_2.2.0         pillar_1.4.6        haven_2.3.1        
-#> [79] foreign_0.8-80      mgcv_1.8-31         nnet_7.3-14        
-#> [82] mstate_0.2.12       modelr_0.1.8        crayon_1.3.4       
-#> [85] rmarkdown_2.3       jpeg_0.1-8.1        grid_4.0.2         
-#> [88] readxl_1.3.1        blob_1.2.1          reprex_0.3.0       
-#> [91] digest_0.6.25       webshot_0.5.2       numDeriv_2016.8-1.1
-#> [94] stats4_4.0.2        munsell_0.5.0       viridisLite_0.3.0  
-#> [97] quadprog_1.5-8
+#> [34] nlme_3.1-148        iterators_1.0.12    xfun_0.15          
+#> [37] rvest_0.3.5         lifecycle_0.2.0     polspline_1.1.17   
+#> [40] muhaz_1.2.6.1       MASS_7.3-51.6       zoo_1.8-8          
+#> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
+#> [46] RColorBrewer_1.1-2  rticles_0.14        yaml_2.2.1         
+#> [49] gridExtra_2.3       rms_5.1-4           rpart_4.1-15       
+#> [52] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
+#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
+#> [58] rlang_0.4.7         pkgconfig_2.0.3     evaluate_0.14      
+#> [61] lattice_0.20-41     htmlwidgets_1.5.1   cmprsk_2.2-9       
+#> [64] tidyselect_1.1.0    deSolve_1.28        R6_2.4.1           
+#> [67] generics_0.0.2      Hmisc_4.4-0         multcomp_1.4-13    
+#> [70] DBI_1.1.0           withr_2.2.0         pillar_1.4.6       
+#> [73] haven_2.2.0         foreign_0.8-80      mgcv_1.8-31        
+#> [76] nnet_7.3-14         mstate_0.2.12       modelr_0.1.6       
+#> [79] crayon_1.3.4        rmarkdown_2.1       timereg_1.9.4      
+#> [82] jpeg_0.1-8.1        grid_4.0.2          readxl_1.3.1       
+#> [85] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
+#> [88] numDeriv_2016.8-1.1 stats4_4.0.2        munsell_0.5.0      
+#> [91] viridisLite_0.3.0   quadprog_1.5-8
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> Local:    master /home/jesseislam/Documents/gits/cbpaper
-#> Remote:   master @ origin (git@github.com:Jesse-Islam/cbpaper.git)
+#> Local:    editMax /Users/maxturgeon/Documents/git_repos/cbpaper
 #> Head:     [d30ba8c] 2020-07-22: Merge pull request #23 from sahirbhatnagar/editMax

### 2020-08-05 analysis/paper/paper.tex
update intro

@@
 function, which can then be used to estimate the cumulative incidence
-function (CIF). Since the parametric hazard is a smooth function of
-time, the CIF and the survival function estimates also vary smoothly
-over time.
+function (CIF). Using a parametric estimator of the hazard function
+leads to a smooth function of time; as a consequence, the CIF and the
+survival function estimates also vary smoothly over time.
 
@@
 
+Using the framework of logistic regression for analysing time-to-event
+data opens the door to an extensive array of powerful modeling tools.
+Indeed, lasso and elastic-net regression can then be used to select
+variables that are associated with the hazard function. Logistic
+regression can also be replaced by multinomial regression to analyse
+data with multiple competing events.
+
 In this article, we present an \proglang{R} package that combines the
@@
 package retains the flexibility of case-base sampling and the familiar
-interface of the \code{glm} function. In addition, we provide extensive
-visualization tools.
+interface of the \code{glm} function. It also provides tools for
+variable selection and competing-risk analysis. In addition, we provide
+extensive visualization tools.
 
@@
 
-This report was generated on 2020-08-05 11:18:44 using the following
+This report was generated on 2020-08-05 11:28:49 using the following
 computational environment and dependencies:

### 2020-08-05 analysis/paper/paper.Rmd

@@
 Therefore, to make sound clinical decisions, it is important to accurately estimate the *full* hazard function, which can then be used to estimate the cumulative incidence function (CIF).
-Since the parametric hazard is a smooth function of time, the CIF and the survival function estimates also vary smoothly over time.
+Using a parametric estimator of the hazard function leads to a smooth function of time; as a consequence, the CIF and the survival function estimates also vary smoothly over time.
 
@@
 
+Using the framework of logistic regression for analysing time-to-event data opens the door to an extensive array of powerful modeling tools.
+Indeed, lasso and elastic-net regression can then be used to select variables that are associated with the hazard function.
+Logistic regression can also be replaced by multinomial regression to analyse data with multiple competing events.
+
 In this article, we present an \proglang{R} package that combines the ideas of Hanley & Miettinen into a simple interface.
@@
 Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function.
+It also provides tools for variable selection and competing-risk analysis.
 In addition, we provide extensive visualization tools.

### 2020-08-06 analysis/paper/paper.tex

@@
 parametric estimators (e.g.~spline functions). Packages that implement
-this approach include \pkg{RiskRegression}, \pkg{rstpm2}, and
+this approach include \pkg{riskRegression}, \pkg{rstpm2}, and
 \pkg{survival}. As mentioned in the introduction, a key distinguishing
@@
 & \textbf{Competing} & \textbf{Allows} &  \textbf{Penalized}   &    &  & \textbf{Semi} & \textbf{Interval/Left} & \textbf{Absolute}  \\
-\textbf{Package}        & \textbf{Risks}  & \textbf{Non PH} & \textbf{Likelihood} & \textbf{Splines} & \textbf{Parametric} & \textbf{Parametric} & \textbf{Censoring} & \textbf{Risk}           \\
+\textbf{Package}        & \textbf{Risks}  & \textbf{Non PH} & \textbf{Regression} & \textbf{Splines} & \textbf{Parametric} & \textbf{Parametric} & \textbf{Censoring} & \textbf{Risk}           \\
 \hline
-\textbf{casebase}        & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
-\textbf{CFC}             & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
-\textbf{cmprsk}             & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\
-\textbf{coxRidge}        &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{crrp}            & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
-\textbf{fastcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{Flexrsurv}       &                          & x                         &                       & x                & x                   &                          &                                  & x               \\
-\textbf{Flexsurv}        & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\
-\textbf{glmnet}          &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{glmpath}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{mets}            & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
-\textbf{penalized}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
-\textbf{RiskRegression}  &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
-\textbf{Rstpm2}          &                          & x                         & x                     & x                & x                   & x                        & x                                & x                         \\
-\textbf{SmoothHazard}    &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
-\textbf{Survival}        & x                        & x                         &                       &                  & x                   & x                        & x                                & x                               \\
+\textbf{casebase}        & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\ \bottomrule
+\textbf{CFC}             & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\ \bottomrule
+\textbf{cmprsk}             & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\ \bottomrule
+\textbf{CoxRidge}        &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
+\textbf{crrp}            & x                        &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
+\textbf{fastcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
+\textbf{flexrsurv}       &                          & x                         &                       & x                & x                   &                          &                                  & x               \\ \bottomrule
+\textbf{flexsurv}        & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\ \bottomrule
+\textbf{glmnet}          &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
+\textbf{glmpath}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
+\textbf{mets}            & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\ \bottomrule
+\textbf{penalized}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
+\textbf{riskRegression}  & x                         &                           & x                     &                  &                     & x                        &                                  & x                                \\ \bottomrule
+\textbf{rstpm2}          &                          & x                         &                      & x                & x                   & x                        & x                                & x                         \\ \bottomrule
+\textbf{SmoothHazard}    &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\ \bottomrule
+\textbf{survival}        & x                        & x                         &                       &                  & x                   & x                        & x                                & x                               \\ \bottomrule
 \hline
@@
 }
-\caption{Comparison of various \proglang{R} packages for survival analysis based on several defining features.}
+\caption{Comparison of various \proglang{R} packages for survival analysis based on several defining features.
+Competing Risks: whether or not an implementation for competing risks is present.
+Allows Non-PH: permits models for non-proportional hazards.
+Penalized Regression: allows for a penalty term when estimating hazards (e.g.
+lasso or ridge).
+Splines: permits a flexible fit on time through the use of splines.
+Parametric: implementation for parametric models.
+Semi-parametric: implementation for semi-parametric models.
+Interval/left censoring: models for interval and left-censoring.
+If this is not selected, the package only handles right-censoring.
+Absolute Risk: computation for survival curves, cumulative incidence or cumulative hazard is available.}
 \label{tab:surv-pkgs}
@@
 
-\begin{center}\includegraphics{../figures/unnamed-chunk-9-1} \end{center}
+\begin{center}\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/unnamed-chunk-9-1} \end{center}
 
@@
 
-This report was generated on 2020-08-05 11:28:49 using the following
+This report was generated on 2020-08-06 17:18:33 using the following
 computational environment and dependencies:
@@
 #> other attached packages:
-#>  [1] scales_1.1.1              forcats_0.5.0            
-#>  [3] stringr_1.4.0             dplyr_1.0.0              
-#>  [5] purrr_0.3.4               readr_1.3.1              
-#>  [7] tidyr_1.0.2               ggplot2_3.3.2            
-#>  [9] tidyverse_1.3.0           lubridate_1.7.8          
-#> [11] visreg_2.6.1              tibble_3.0.3             
-#> [13] riskRegression_2020.02.05 prodlim_2019.11.13       
-#> [15] flexsurv_1.1.1            survival_3.1-12          
-#> [17] pracma_2.2.9              magrittr_1.5             
-#> [19] kableExtra_1.1.0          glmnet_3.0-2             
-#> [21] Matrix_1.2-18             data.table_1.12.8        
-#> [23] cowplot_1.0.0             colorspace_1.4-1         
-#> [25] casebase_0.9.0.9000      
+#>  [1] broom_0.5.6               dotwhisker_0.5.0         
+#>  [3] timereg_1.9.4             glue_1.4.1               
+#>  [5] scales_1.1.1              forcats_0.5.0            
+#>  [7] stringr_1.4.0             dplyr_1.0.0              
+#>  [9] purrr_0.3.4               readr_1.3.1              
+#> [11] tidyr_1.0.2               ggplot2_3.3.2            
+#> [13] tidyverse_1.3.0           lubridate_1.7.8          
+#> [15] visreg_2.6.1              tibble_3.0.3             
+#> [17] riskRegression_2020.02.05 prodlim_2019.11.13       
+#> [19] flexsurv_1.1.1            survival_3.1-12          
+#> [21] pracma_2.2.9              magrittr_1.5             
+#> [23] kableExtra_1.1.0          glmnet_3.0-2             
+#> [25] Matrix_1.2-18             data.table_1.12.8        
+#> [27] cowplot_1.0.0             colorspace_1.4-1         
+#> [29] casebase_0.9.0.9000      
 #> 
@@
 #>  [1] TH.data_1.0-10      VGAM_1.1-2          ellipsis_0.3.1     
-#>  [4] htmlTable_1.13.3    base64enc_0.1-3     fs_1.4.1           
-#>  [7] rstudioapi_0.11     MatrixModels_0.4-1  fansi_0.4.1        
-#> [10] mvtnorm_1.1-0       xml2_1.3.2          codetools_0.2-16   
-#> [13] knitr_1.29          Formula_1.2-3       jsonlite_1.6.1     
-#> [16] broom_0.5.6         cluster_2.1.0       dbplyr_1.4.3       
-#> [19] png_0.1-7           compiler_4.0.2      httr_1.4.1         
-#> [22] backports_1.1.8     assertthat_0.2.1    cli_2.0.2          
-#> [25] acepack_1.4.1       htmltools_0.5.0     quantreg_5.55      
-#> [28] tools_4.0.2         gtable_0.3.0        glue_1.4.1         
+#>  [4] ggstance_0.3.4      htmlTable_1.13.3    base64enc_0.1-3    
+#>  [7] fs_1.4.1            rstudioapi_0.11     farver_2.0.3       
+#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-0      
+#> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
+#> [16] Formula_1.2-3       jsonlite_1.6.1      cluster_2.1.0      
+#> [19] dbplyr_1.4.3        png_0.1-7           compiler_4.0.2     
+#> [22] httr_1.4.1          backports_1.1.8     assertthat_0.2.1   
+#> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
+#> [28] quantreg_5.55       tools_4.0.2         gtable_0.3.0       
 #> [31] Rcpp_1.0.5          cellranger_1.1.0    vctrs_0.3.2        
@@
 #> [58] rlang_0.4.7         pkgconfig_2.0.3     evaluate_0.14      
-#> [61] lattice_0.20-41     htmlwidgets_1.5.1   cmprsk_2.2-9       
-#> [64] tidyselect_1.1.0    deSolve_1.28        R6_2.4.1           
-#> [67] generics_0.0.2      Hmisc_4.4-0         multcomp_1.4-13    
-#> [70] DBI_1.1.0           withr_2.2.0         pillar_1.4.6       
-#> [73] haven_2.2.0         foreign_0.8-80      mgcv_1.8-31        
-#> [76] nnet_7.3-14         mstate_0.2.12       modelr_0.1.6       
-#> [79] crayon_1.3.4        rmarkdown_2.1       timereg_1.9.4      
-#> [82] jpeg_0.1-8.1        grid_4.0.2          readxl_1.3.1       
-#> [85] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
-#> [88] numDeriv_2016.8-1.1 stats4_4.0.2        munsell_0.5.0      
-#> [91] viridisLite_0.3.0   quadprog_1.5-8
+#> [61] lattice_0.20-41     labeling_0.3        htmlwidgets_1.5.1  
+#> [64] cmprsk_2.2-9        tidyselect_1.1.0    deSolve_1.28       
+#> [67] plyr_1.8.6          R6_2.4.1            generics_0.0.2     
+#> [70] Hmisc_4.4-0         multcomp_1.4-13     DBI_1.1.0          
+#> [73] withr_2.2.0         pillar_1.4.6        haven_2.2.0        
+#> [76] foreign_0.8-80      mgcv_1.8-31         nnet_7.3-14        
+#> [79] mstate_0.2.12       modelr_0.1.6        crayon_1.3.4       
+#> [82] rmarkdown_2.1       jpeg_0.1-8.1        grid_4.0.2         
+#> [85] readxl_1.3.1        reprex_0.3.0        digest_0.6.25      
+#> [88] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_4.0.2       
+#> [91] munsell_0.5.0       viridisLite_0.3.0   quadprog_1.5-8
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> Local:    editMax /Users/maxturgeon/Documents/git_repos/cbpaper
-#> Head:     [d30ba8c] 2020-07-22: Merge pull request #23 from sahirbhatnagar/editMax
+#> Local:    Jesse-Islam-master /Users/maxturgeon/Documents/git_repos/cbpaper
+#> Head:     [32e6bf4] 2020-08-05: Merge pull request #25 from sahirbhatnagar/editMax
 \end{CodeOutput}

### 2020-08-06 analysis/paper/paper.Rmd

@@
 spline functions).
-Packages that implement this approach include \pkg{RiskRegression}, \pkg{rstpm2}, and \pkg{survival}.
+Packages that implement this approach include \pkg{riskRegression}, \pkg{rstpm2}, and \pkg{survival}.
 As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category often produces estimates in the form of stepwise functions.
@@
 ```{r supportData, eval = eval_cs3}
+scaleFUNy <- function(x) sprintf("%.2f", x)
+scaleFUNx <- function(x) sprintf("%.0f", x)
 data(support)
@@
 cox_ci <- ggplot(baseCoxComparisonData, aes(Time)) +
-  geom_line(aes(y = APS, colour = "APS"), lwd = 1) +
-  geom_line(aes(y = Full, col = "Full"), lwd = 1) +
-  geom_line(aes(y = SPS, colour = "SPS"), lwd = 1) +
-  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M"), lwd = 1) +
-  labs(y = "Probability of Death", x = "Survival time (Years)", color = "Models") +
+  geom_step(aes(y = APS, colour = "APS")) +
+  geom_step(aes(y = Full, col = "Full")) +
+  geom_step(aes(y = SPS, colour = "SPS")) +
+  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
+  labs(y = "Probability of Death", x = "Follow-up time (Years)", color = "Models") +
   scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
-  paper_gg_theme
-```
-
-```{r brierpt1, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:brier1} Comparison of three Cox regression models. (A) Cumulative Incidence curves. (B) Time-dependent Brier scores."}
+  paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1))) +scale_y_continuous(labels=scaleFUNy,n.breaks=8)
+```
+
+```{r brierpt1, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:brier1} Comparison of three Cox regression models to the unadjusted Kaplan-Meier curve.
+(A) Cumulative Incidence curves.
+For a given model, the plotted curve is the average curve of each individual in the study.
+Each follow-up time seen in the study was used.
+There are minimal differences between the curves.
+APS coincides with the SPS curve, as they are considerably overlapped.(B) Time-dependent Brier scores, where a lower score is better.
+The full model performs best overall."}
 # Extract time points
@@
   # ggtitle("Brier scores over time") +
-  scale_color_manual(values = q4[1:3]) + 
-  paper_gg_theme
+  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
+  paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1)))+scale_y_continuous(labels=scaleFUNy,n.breaks=8)
 
@@
 reg_ci <- ggplot(ab, aes(Time)) +
-  geom_line(data = baseCoxComparisonData, mapping = aes(y = Full, colour = "Cox"), lwd = 1) +
-  geom_line(aes(y = absRisk, colour = Model), lwd = 1) +
-  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M"), lwd = 1) +
-  labs(y = "Probability of Death", x = "Survival time (Years)", color = "Models") +
+  geom_step(data = baseCoxComparisonData, mapping = aes(x = Time,y = Full, colour = "Cox")) +
+  geom_line(aes(y = absRisk, colour = Model)) +
+  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
+  labs(y = "Probability of death", x = "Follow-up time (years)", color = "Models") +
   # ggtitle("Absolute risk curves") +
-  scale_color_manual(values = q4, breaks = c("Cox", "Penalized Cox", "Penalized logistic", "K-M")) + 
-  paper_gg_theme
+  scale_color_manual(values = q4, breaks = c("K-M","Cox","Penalized Cox","Penalized logistic")) + 
+  paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1))) +scale_y_continuous(labels=scaleFUNy,n.breaks=9)
 ```
@@
   geom_line() +
-  xlab("Survival time (Years)") +
+  xlab("Follow-up time (years)") +
   ylab("Brier score") +

### 2020-08-07 analysis/paper/paper.Rmd

@@
 
-<!--Before generating an absolute risk curve using regularized logistic regression, we will explore three options as a baseline comparison.
+<!--Before generating an absolute risk curve using regularized estimation, we will explore three options as a baseline comparison.
 The original study compared these scores over a span of 180 days, which may not be comparable over the full 5.56 years.
@@
 Figure \ref{fig:brier1} examines the brier scores as they change over time.
-As the full model has the lowest brier score on overall, it will be used in the following absolute risk comparison to regularized cox, regularized logistic regression and the unadjusted Kaplan-Meier curve.-->
+As the full model has the lowest brier score on overall, it will be used in the following absolute risk comparison to regularized cox, regularized estimation and the unadjusted Kaplan-Meier curve.-->
 
@@
 
-For our penalized logistic regression model, we opt for the natural log of time; recall that this corresponds to a Weibull distribution.
+For our penalized model, we opt for the natural log of time; recall that this corresponds to a Weibull distribution.
 For fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface to the function \code{fitSmoothHazard}.
@@
                    absRisk = rowMeans(abCbFull[, -c(1)]))
-abCB$Model <- "Penalized logistic"
+abCB$Model <- "Penalized estimation"
 
@@
 This is expected, as there are no restrictions on the size of coefficients.
-The smallest coefficient estimates are found in penalized logistic model, aside from cancer state (cano).
+The smallest coefficient estimates are found in penalized case-base model, aside from cancer state (cano).
 
@@
 # estimatescb$std.error<-0
-estimatescb$Model <- "Pen. logistic"
+estimatescb$Model <- "Pen. CB"
 colnames(estimatescb) <- c("term", "estimate", "model")
@@
 To better demonstrate the differences in performance between the models, we compare Brier scores over time.
-In Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our penalized logistic regression model compared to both Cox regression models.
+In Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our penalized case-base model compared to both Cox regression models.
 The differences, however, are minimal.
-All three outperform the Kaplan-Meier model.
+All three outperform the Kaplan-Meier estimator.
 
@@
   # ggtitle("Absolute risk curves") +
-  scale_color_manual(values = q4, breaks = c("K-M","Cox","Penalized Cox","Penalized logistic")) + 
+  scale_color_manual(values = q4, breaks = c("K-M","Cox","Penalized Cox","Penalized CB")) + 
   paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1))) +scale_y_continuous(labels=scaleFUNy,n.breaks=9)
@@
 
-```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} (A) Four curves comparing the probabilities of death as it changes over follow-up time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
-All follow-up times seen in the new data were used.
-(B) Four curves demonstrating the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
-Lower is better.
-Penalized logistic regression produced the lowest results over follow-up time.", warning = FALSE}
+```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} Comparison of Cox regression, penalized Cox regression, penalized case-base estimation, and Kaplan-Meier.
+(A)  Probabilities of death as it changes over follow-up time.
+(B) Brier scores as they change over time, where a lower score corresponds to better performace.
+Penalized case-base estimation produced the lowest results over follow-up time.", warning = FALSE}
 # Compute Brier score for penalized models
-brierFinalResults <- Score(list("Penalized Cox" = coxNet, "Penalized Logistic" = cbFull),
+brierFinalResults <- Score(list("Penalized Cox" = coxNet, "Penalized CB" = cbFull),
                            data = cbind(subset(test, select = c(d.time, death)),

### 2020-08-07 analysis/paper/paper.tex

@@
 
-For our penalized logistic regression model, we opt for the natural log
-of time; recall that this corresponds to a Weibull distribution. For
-fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which
-is a matrix interface to the function \code{fitSmoothHazard}. In order
-to use this interface, we first create a matrix \code{y} containing the
-time and event variables, and a matrix \code{x} containing all remaining
-covariates we would like to include in the model space. We used lasso
-penalization by setting \code{alpha = 1}. Moreover, we use the argument
+For our penalized model, we opt for the natural log of time; recall that
+this corresponds to a Weibull distribution. For fitting the penalized
+hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface
+to the function \code{fitSmoothHazard}. In order to use this interface,
+we first create a matrix \code{y} containing the time and event
+variables, and a matrix \code{x} containing all remaining covariates we
+would like to include in the model space. We used lasso penalization by
+setting \code{alpha = 1}. Moreover, we use the argument
 \code{penalty.factor} to force the time variable into the final model.
@@
 restrictions on the size of coefficients. The smallest coefficient
-estimates are found in penalized logistic model, aside from cancer state
-(cano).
+estimates are found in penalized case-base model, aside from cancer
+state (cano).
 
@@
 Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our
-penalized logistic regression model compared to both Cox regression
-models. The differences, however, are minimal. All three outperform the
-Kaplan-Meier model.
+penalized case-base model compared to both Cox regression models. The
+differences, however, are minimal. All three outperform the Kaplan-Meier
+estimator.
 
@@
 
-\caption{\label{fig:cs3FinalBrier} (A) Four curves comparing the probabilities of death as it changes over follow-up time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
-All follow-up times seen in the new data were used.
-(B) Four curves demonstrating the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
-Lower is better.
-Penalized logistic regression produced the lowest results over follow-up time.}\label{fig:riskregressionBrier}
+\caption{\label{fig:cs3FinalBrier} Comparison of Cox regression, penalized Cox regression, penalized case-base estimation, and Kaplan-Meier.
+(A)  Probabilities of death as it changes over follow-up time.
+(B) Brier scores as they change over time, where a lower score corresponds to better performace.
+Penalized case-base estimation produced the lowest results over follow-up time.}\label{fig:riskregressionBrier}
 \end{figure}
@@
 
-This report was generated on 2020-08-06 17:18:33 using the following
+This report was generated on 2020-08-06 17:31:23 using the following
 computational environment and dependencies:
@@
 #> [64] cmprsk_2.2-9        tidyselect_1.1.0    deSolve_1.28       
-#> [67] plyr_1.8.6          R6_2.4.1            generics_0.0.2     
-#> [70] Hmisc_4.4-0         multcomp_1.4-13     DBI_1.1.0          
-#> [73] withr_2.2.0         pillar_1.4.6        haven_2.2.0        
-#> [76] foreign_0.8-80      mgcv_1.8-31         nnet_7.3-14        
-#> [79] mstate_0.2.12       modelr_0.1.6        crayon_1.3.4       
-#> [82] rmarkdown_2.1       jpeg_0.1-8.1        grid_4.0.2         
-#> [85] readxl_1.3.1        reprex_0.3.0        digest_0.6.25      
-#> [88] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_4.0.2       
-#> [91] munsell_0.5.0       viridisLite_0.3.0   quadprog_1.5-8
+#> [67] R6_2.4.1            generics_0.0.2      Hmisc_4.4-0        
+#> [70] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
+#> [73] pillar_1.4.6        haven_2.2.0         foreign_0.8-80     
+#> [76] mgcv_1.8-31         nnet_7.3-14         mstate_0.2.12      
+#> [79] modelr_0.1.6        crayon_1.3.4        rmarkdown_2.1      
+#> [82] jpeg_0.1-8.1        grid_4.0.2          readxl_1.3.1       
+#> [85] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
+#> [88] numDeriv_2016.8-1.1 stats4_4.0.2        munsell_0.5.0      
+#> [91] viridisLite_0.3.0   quadprog_1.5-8
 \end{CodeOutput}

### 2020-08-07 analysis/paper/paper.Rmd
fix figure 7

@@
   scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
-  paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1))) +scale_y_continuous(labels=scaleFUNy,n.breaks=8)
+  paper_gg_theme + 
+  scale_x_continuous(labels = scaleFUNx, breaks = round(seq(0, 5, by = 1))) + 
+  scale_y_continuous(labels = scaleFUNy, n.breaks = 8)
 ```
@@
                    absRisk = rowMeans(abCbFull[, -c(1)]))
-abCB$Model <- "Penalized estimation"
+abCB$Model <- "Penalized CB"
 
@@
   # ggtitle("Absolute risk curves") +
-  scale_color_manual(values = q4, breaks = c("K-M","Cox","Penalized Cox","Penalized CB")) + 
-  paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1))) +scale_y_continuous(labels=scaleFUNy,n.breaks=9)
+  scale_color_manual(values = q4, breaks = c("K-M", "Cox", "Penalized Cox", "Penalized CB")) + 
+  paper_gg_theme + 
+  scale_x_continuous(labels = scaleFUNx, breaks = round(seq(0, 5, by = 1))) + 
+  scale_y_continuous(labels = scaleFUNy, n.breaks = 9)
 ```
@@
   # if (missing(cause)) stop("Argument cause should be the event type for which we predict the absolute risk.")
-  # the output of absoluteRisk is an array with dimension dependening on the length of the requested times:
+  # the output of absoluteRisk is an array with dimension depending on the length of the requested times:
   # case 1: the number of time points is 1
@@
   ),
-  model = factor(model, levels = c("Cox", "Penalized Cox", "Penalized Logistic", "K-M"),
-                 labels = c("Cox", "Pen. Cox", "Pen. Logistic", "K-M")))
+  model = factor(model, levels = c("Cox", "Penalized Cox", "Penalized CB", "K-M"),
+                 labels = c("Cox", "Pen. Cox", "Pen. CB", "K-M")))
 
@@
   paper_gg_theme +  
-  scale_color_manual(values = q4) + scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1))) +scale_y_continuous(labels=scaleFUNy,n.breaks=8)
+  scale_color_manual(values = q4) + 
+  scale_x_continuous(labels = scaleFUNx, 
+                     breaks = round(seq(0, 5, by = 1))) + 
+  scale_y_continuous(labels = scaleFUNy, n.breaks = 8)
 

### 2020-08-07 analysis/paper/paper.tex
fix figure 7

@@
 
-This report was generated on 2020-08-06 17:31:23 using the following
+This report was generated on 2020-08-07 09:30:39 using the following
 computational environment and dependencies:
@@
 #> other attached packages:
-#>  [1] broom_0.5.6               dotwhisker_0.5.0         
+#>  [1] broom_0.7.0               dotwhisker_0.5.0         
 #>  [3] timereg_1.9.4             glue_1.4.1               
 #>  [5] scales_1.1.1              forcats_0.5.0            
-#>  [7] stringr_1.4.0             dplyr_1.0.0              
+#>  [7] stringr_1.4.0             dplyr_1.0.1              
 #>  [9] purrr_0.3.4               readr_1.3.1              
-#> [11] tidyr_1.0.2               ggplot2_3.3.2            
-#> [13] tidyverse_1.3.0           lubridate_1.7.8          
-#> [15] visreg_2.6.1              tibble_3.0.3             
+#> [11] tidyr_1.1.1               ggplot2_3.3.2            
+#> [13] tidyverse_1.3.0           lubridate_1.7.9          
+#> [15] visreg_2.7.0              tibble_3.0.3             
 #> [17] riskRegression_2020.02.05 prodlim_2019.11.13       
-#> [19] flexsurv_1.1.1            survival_3.1-12          
+#> [19] flexsurv_1.1.1            survival_3.2-3           
 #> [21] pracma_2.2.9              magrittr_1.5             
-#> [23] kableExtra_1.1.0          glmnet_3.0-2             
-#> [25] Matrix_1.2-18             data.table_1.12.8        
+#> [23] kableExtra_1.1.0          glmnet_4.0-2             
+#> [25] Matrix_1.2-18             data.table_1.13.0        
 #> [27] cowplot_1.0.0             colorspace_1.4-1         
@@
 #> loaded via a namespace (and not attached):
-#>  [1] TH.data_1.0-10      VGAM_1.1-2          ellipsis_0.3.1     
+#>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
 #>  [4] ggstance_0.3.4      htmlTable_1.13.3    base64enc_0.1-3    
 #>  [7] fs_1.4.1            rstudioapi_0.11     farver_2.0.3       
-#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-0      
+#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-1      
 #> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
-#> [16] Formula_1.2-3       jsonlite_1.6.1      cluster_2.1.0      
-#> [19] dbplyr_1.4.3        png_0.1-7           compiler_4.0.2     
-#> [22] httr_1.4.1          backports_1.1.8     assertthat_0.2.1   
+#> [16] Formula_1.2-3       jsonlite_1.7.0      cluster_2.1.0      
+#> [19] dbplyr_1.4.4        png_0.1-7           compiler_4.0.2     
+#> [22] httr_1.4.2          backports_1.1.8     assertthat_0.2.1   
 #> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
@@
 #> [34] nlme_3.1-148        iterators_1.0.12    xfun_0.15          
-#> [37] rvest_0.3.5         lifecycle_0.2.0     polspline_1.1.17   
+#> [37] rvest_0.3.6         lifecycle_0.2.0     polspline_1.1.17   
 #> [40] muhaz_1.2.6.1       MASS_7.3-51.6       zoo_1.8-8          
@@
 #> [61] lattice_0.20-41     labeling_0.3        htmlwidgets_1.5.1  
-#> [64] cmprsk_2.2-9        tidyselect_1.1.0    deSolve_1.28       
-#> [67] R6_2.4.1            generics_0.0.2      Hmisc_4.4-0        
-#> [70] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
-#> [73] pillar_1.4.6        haven_2.2.0         foreign_0.8-80     
-#> [76] mgcv_1.8-31         nnet_7.3-14         mstate_0.2.12      
-#> [79] modelr_0.1.6        crayon_1.3.4        rmarkdown_2.1      
-#> [82] jpeg_0.1-8.1        grid_4.0.2          readxl_1.3.1       
-#> [85] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
-#> [88] numDeriv_2016.8-1.1 stats4_4.0.2        munsell_0.5.0      
-#> [91] viridisLite_0.3.0   quadprog_1.5-8
+#> [64] cmprsk_2.2-10       tidyselect_1.1.0    deSolve_1.28       
+#> [67] plyr_1.8.6          R6_2.4.1            generics_0.0.2     
+#> [70] Hmisc_4.4-0         multcomp_1.4-13     DBI_1.1.0          
+#> [73] withr_2.2.0         pillar_1.4.6        haven_2.3.1        
+#> [76] foreign_0.8-80      mgcv_1.8-31         nnet_7.3-14        
+#> [79] mstate_0.2.12       modelr_0.1.8        crayon_1.3.4       
+#> [82] rmarkdown_2.3       jpeg_0.1-8.1        grid_4.0.2         
+#> [85] readxl_1.3.1        blob_1.2.1          reprex_0.3.0       
+#> [88] digest_0.6.25       webshot_0.5.2       numDeriv_2016.8-1.1
+#> [91] stats4_4.0.2        munsell_0.5.0       viridisLite_0.3.0  
+#> [94] quadprog_1.5-8
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> Local:    Jesse-Islam-master /Users/maxturgeon/Documents/git_repos/cbpaper
-#> Head:     [32e6bf4] 2020-08-05: Merge pull request #25 from sahirbhatnagar/editMax
+#> Local:    master /Users/maxturgeon/Documents/git_repos/cbpaper
+#> Remote:   master @ origin (git@github.com:sahirbhatnagar/cbpaper.git)
+#> Head:     [06b07b3] 2020-08-06: incorporate Ollis comments on CS 3
 \end{CodeOutput}

### 2020-08-18 analysis/paper/paper.Rmd
time dependent hazard ratio for ERSPC

--- 2020-08-07 analysis/paper/paper.Rmd
@@
 library(scales)
-options(digits = 2, scipen = 999)
+options(digits = 2)
 # set the seed for reproducible output
@@
 
+Population-time plots are a descriptive visualization of incidence density, where aggregate person-time is represented by area and events as points within the area.
 The case-base sampling approach described in Section \ref{theory} can be visualized in the form of a population time plot.
@@
 
-To illustrate these functions, we will use the European Randomized Study of Prostate Cancer Screening (ERSPC) data [@schroder2009screening] which was extracted using the approach described in Liu *et al.* [-@liu2014recovering] This dataset is available through the \pkg{casebase} package.
+To illustrate these functions, we will use data from the European Randomized Study of Prostate Cancer Screening (ERSPC) [@schroder2009screening] which was extracted using the approach described in Liu *et al.* [-@liu2014recovering] This dataset is available through the \pkg{casebase} package.
 It contains the individual observations for 159,893 men  from seven European countries, who were between the ages of 55 and 69 years when recruited for the trial.
@@
 
-pt_object <- casebase::popTime(ERSPC, 
-                               time = "Follow.Up.Time",
-                               event = "DeadOfPrCa",
-                               exposure = "ScrArm")
+pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
+                               event = "DeadOfPrCa", exposure = "ScrArm")
 inherits(pt_object, "popTime")
@@
 
-```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plot for ERSPC data."}
+```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plots for both treatment arms in the ERSPC dataset.
+The grey area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely small rectangles (person-moments).
+The points shown in the figure are select person-moments from the case and base series.
+We see that more events are observed at later follow-up times, motivating the use of a non-constant in time hazard model."}
 label_number_si <- function (accuracy = 1, unit = NULL, sep = NULL, ...) {
@@
   paper_gg_theme +  
-  scale_y_continuous(labels = label_number_si())
+  scale_y_continuous(labels = label_number_si()) + 
+  ylab("Population Size")
 ```
@@
 Next, layers for the case series and base series are added.
-The case series is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area.
-By randomly distributing the cases, we can get a better sense of the incidence density.
+The y-axis location of each case moment is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area.
+By randomly distributing the cases, we can get a sense of the incidence density.
 In Figure \ref{fig:plot-stratified-erspc-data}, we see that more events are observed at later follow-up times.
@@
 
-For our first case study, we return to the ERSPC study and investigate the differences in survival between the control and screening arms.
+For our first case study, we return to the ERSPC study and investigate the differences in risk between the control and screening arms.
 We fit four models that differ in which functional form of time is used: 1) excluded from the linear predictor, 2) linear function, 3) log function, and 4) a smooth function using cubic B-splines.
@@
 ```{r, eval = eval_cs1, echo=FALSE}
-summary(fits[["gompertz"]])
+print.summary.glm <- function(x,digits=max(3,getOption("digits")-3),symbolic.cor=x$symbolic.cor,
+                              signif.stars = getOption("show.signif.stars"),show.call=FALSE,show.resids=FALSE,...) {
+  if (show.call) cat("Call:\n", paste(deparse(x$call), sep = "\n", collapse = "\n"), "\n", sep = "")
+  if (show.resids) {
+    cat("Deviance Residuals: \n")
+    if (x$df.residual > 5) {
+      x$deviance.resid <- stats::quantile(x$deviance.resid, na.rm = TRUE)
+      names(x$deviance.resid) <- c("Min", "1Q", "Median", "3Q", "Max")
+    }
+    xx <- zapsmall(x$deviance.resid, digits + 1)
+    print.default(xx, digits = digits, na.print = "", print.gap = 2)
+    cat("\n")
+  }
+  if (length(x$aliased) == 0L) cat("No Coefficients\n")
+  else {
+    df <- if ("df" %in% names(x)) x[["df"]]
+    else NULL
+    if (!is.null(df) && (nsingular <- df[3L] - df[1L])) 
+      cat("Coefficients: (", nsingular, " not defined because of singularities)\n", sep = "")
+    else cat("\nCoefficients:\n")
+    coefs <- x$coefficients
+    if (!is.null(aliased <- x$aliased) && any(aliased)) {
+      cn <- names(aliased)
+      coefs <- matrix(NA, length(aliased), 4L, dimnames = list(cn,colnames(coefs)))
+      coefs[!aliased, ] <- x$coefficients
+    }
+    stats::printCoefmat(coefs, digits = digits, signif.stars = signif.stars,na.print = "NA", ...)
+  }
+  cat("\n(Dispersion parameter for ", x$family$family, " family taken to be ", 
+      format(x$dispersion), ")\n\n", apply(cbind(paste(format(c("Null","Residual"), 
+                                                              justify = "right"), "deviance:"),
+                                                 format(unlist(x[c("null.deviance","deviance")]), digits = max(5, digits + 1)), " on", 
+                                                 format(unlist(x[c("df.null", "df.residual")])), " degrees of freedom\n"),1L,
+                                           paste, collapse = " "), sep = "")
+  if (nzchar(mess <- stats::naprint(x$na.action))) cat("  (", mess, ")\n",sep="")
+  cat("AIC: ",format(x$aic,digits=max(4,digits+1)),"\n\n",
+      "Number of Fisher Scoring iterations: ",x$iter,"\n",sep="")
+  correl <- x$correlation
+  if (!is.null(correl)) {
+    p <- NCOL(correl)
+    if (p > 1) {
+      cat("\nCorrelation of Coefficients:\n")
+      if (is.logical(symbolic.cor) && symbolic.cor) print(stats::symnum(correl,abbr.colnames=NULL))
+      else {
+        correl <- format(round(correl, 2),nsmall=2,digits=digits)
+        correl[!lower.tri(correl)] <- ""
+        print(correl[-1,-p,drop=FALSE],quote=FALSE)
+      }
+    }
+  }
+  cat("\n")
+  invisible(x)
+}
+
+print(summary(fits[["gompertz"]]), signif.stars = F)
 ```
@@
 With this first case study, we explored how \pkg{casebase} allows us to fit different parametric survival models, and how we can compare the fit of each model with tools from GLMs.
+
+```{r interaction-ERSPC, eval = eval_cs1}
+fit_inter <- casebase::fitSmoothHazard(DeadOfPrCa ~ poly(Follow.Up.Time, 2)*ScrArm, data = ERSPC)
+summary(fit_inter)
+
+# reference category
+newdata <- data.frame(ScrArm = factor("Control group",
+                                         levels = c("Control group","Screening group")),
+                      Follow.Up.Time = seq(1,12,by =0.1))
+str(newdata)
+
+plot(fit_inter,
+     type = "hr",
+     newdata = newdata,
+     var = "ScrArm",
+     increment = 1,
+     xvar = "Follow.Up.Time",
+     ci = T,
+     rug = T)
+abline(a=1,b=0, lty = 2)
+```
 

### 2020-08-20 analysis/paper/paper.Rmd
included text for time-dependent Hazard Ratio

@@
 
-<!--
-\begin{table}
-  \begin{tabular}{p{1.8in}lll}
-\hline
-    &  Parameters &  Density \proglang{R} function & \code{dist}\\
-    & {\footnotesize{(location in \code{\color{red}{red}})}} & & \\
-\hline
-    Exponential & \code{\color{red}{rate}}             & \code{dexp}   & \code{"exp"} \\
-    Weibull (accelerated failure time)     & \code{shape, {\color{red}{scale}}}     & \code{dweibull} & \code{"weibull"} \\
-    Weibull (proportional hazards)     & \code{shape, {\color{red}{scale}}}     & \code{dweibullPH} & \code{"weibullPH"} \\
-    Gamma       & \code{shape, \color{red}{rate}}      & \code{dgamma} & \code{"gamma"}\\
-    Log-normal  & \code{{\color{red}{meanlog}}, sdlog}   & \code{dlnorm} & \code{"lnorm"}\\
-    Gompertz    & \code{shape, {\color{red}{rate}}}      & \code{dgompertz} & \code{"gompertz"} \\
-    Log-logistic & \code{shape, {\color{red}{scale}}}   & \code{dllogis} & \code{"llogis"}\\
-    Generalized gamma (Prentice 1975)   & \code{{\color{red}{mu}}, sigma, Q} & \code{dgengamma} & \code{"gengamma"} \\
-    Generalized gamma (Stacy 1962)& \code{shape, {\color{red}{scale}}, k} & \code{dgengamma.orig} & \code{"gengamma.orig"} \\
-    Generalized F     (stable)    & \code{{\color{red}{mu}}, sigma, Q, P} & \code{dgenf} & \code{"genf"} \\
-    Generalized F     (original)  & \code{{\color{red}{mu}}, sigma, s1, s2} & \code{dgenf.orig} & \code{"genf.orig"} \\
-\hline
-  \end{tabular}
-  \caption{Built-in parametric survival distributions in \pkg{flexsurv}.}
-  \label{tab:dists}
-\end{table}
--->
-
 # Implementation details
@@
 
-```{r, eval = eval_cs1, echo=FALSE}
+For our first case study, we return to the ERSPC study and investigate the differences in risk between the control and screening arms.
+We fit four models that differ in which functional form of time is used: 1) excluded from the linear predictor, 2) linear function, 3) log function, and 4) a smooth function using cubic B-splines.
+The models are fit using \code{fitSmoothHazard} with the familiar formula interface:
+
+```{r, eval = eval_cs1, echo = -1}
+set.seed(12)
 fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
@@
 fits <- lapply(fmla, function(i) {
-  casebase::fitSmoothHazard(i, data = ERSPC, ratio = 100)
+  fitSmoothHazard(i, data = ERSPC, ratio = 100)
 })
@@
 
-For our first case study, we return to the ERSPC study and investigate the differences in risk between the control and screening arms.
-We fit four models that differ in which functional form of time is used: 1) excluded from the linear predictor, 2) linear function, 3) log function, and 4) a smooth function using cubic B-splines.
-The models are fit using \code{fitSmoothHazard}.
-Since the output object from \code{fitSmoothHazard} inherits from the \code{glm} class, we can directly use the function \code{summary}:
-
-```{r, eval = FALSE, echo = TRUE}
-fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
-             gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
-             weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
-             splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
-
-fits <- lapply(fmla, function(i) {
-  casebase::fitSmoothHazard(i, data = ERSPC, ratio = 100)
-})
-
-summary(fits[["gompertz"]])
+The output object from \code{fitSmoothHazard} inherits from the \code{singleEventCB} and \code{glm} classes. As such, we can directly use the \code{summary} generic:
+
+```{r, eval = FALSE, echo=TRUE}
+summary(fits[["splines"]]) 
 ```
@@
 
-print(summary(fits[["gompertz"]]), signif.stars = F)
+print(summary(fits[["splines"]]), signif.stars = F)
 ```
@@
 new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
+new_time <- seq(0,14,0.1)
 
 risks <- lapply(fits, function(i) {
-  casebase::absoluteRisk(object = i, time = seq(0,15,0.1), newdata = new_data)
+  absoluteRisk(object = i, time = new_time, newdata = new_data, method = "mont")
 })
-
-head(risks[["gompertz"]])
 ```
@@
 Based on our earlier discussion of the population-time plot, this poor fit for the exponential hazard was expected.
-
-In Figure \ref{fig:erspc-cox-cif} (panels 2--4), we notice a better fit with increasing complexity of our model for time.
+We notice a better fit with increasing complexity of our model for time in Figure \ref{fig:erspc-cox-cif} (panels 2--4).
 As noted above, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models.
-Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) splines model:
+Therefore, we can easily test the null hypothesis that the Weibull model is just as good as the larger (in terms of number of parameters) spline model:
 
 ```{r erscp-compare, eval = eval_cs1, echo = FALSE}
-anova(fits$exponential, fits$splines, test = "LRT")
-```
-
-As expected, we see that splines model provides a better fit. Similarly, we can compare the AIC for all four case-base models:
+anova(fits$weibull, fits$splines, test = "LRT")
+```
+
+The null hypothesis is rejected in favor of the spline model. Similarly, the AIC provides further evidence that the flexible function of time provides the best fit:
 
@@
 c("Exp." = AIC(fits$exponential),
-  "Gomp." = AIC(fits$gompertz),
-  "Weib." = AIC(fits$exponential),
+  "Gompertz" = AIC(fits$gompertz),
+  "Weibull" = AIC(fits$weibull),
   "Splines" = AIC(fits$splines))
 ```
-
-Once again, we have evidence that the splines model provides the best fit.
 
@@
 
-```{r erspc-cox-cif, eval = eval_cs1, echo = FALSE, fig.width=8, fig.height=6, fig.cap="CIFs for control and screening groups in the ERSPC data.
+```{r erspc-cox-cif, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="CIFs for control and screening groups in the ERSPC data.
 In each of the panels, we plot the CIF from the Cox model using \\code{survival::survfit} (solid line) and the CIF from the case-base sampling scheme (dashed line) with different functional forms of time.
@@
 myCols <- cbPalette[c(4,7)]
-new_data <- data.frame(
-  ScrArm = c("Control group", "Screening group")#,
-  # ignore = 99
-)
+
+new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
+new_time <- seq(0,14.5,0.1)
 
 risks <- lapply(fits, function(i) {
-  casebase::absoluteRisk(object = i, time = seq(0,15,0.1), newdata = new_data)
+  casebase::absoluteRisk(object = i, time = new_time, newdata = new_data, method = "mont")
 })
-
-aft_fits <- lapply(list("exp","gompertz","weibull"), function(i)
-  flexsurv::flexsurvreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm,
-                           data = ERSPC, dist = i)
-  )
-
-names(aft_fits) <- list("exponential","gompertz","weibull")
 
@@
      fun = "event",
-     xlim = c(0,15), conf.int = FALSE, col = myCols, 
+     xmax = max(new_time),
+     # xlim = range(new_time), 
+     conf.int = FALSE, col = myCols, 
      # main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
@@
 lines(risks$exponential[,1], risks$exponential[,3], col = myCols[2], lty = 2, lwd = 2)
-text(x = 15*0.97, y = 0.05/100, labels = "(1)", cex = 1.5)
+text(x = max(new_time)*0.97, y = 0.05/100, labels = "(1)", cex = 1.5)
 
@@
      fun = "event",
-     xlim = c(0,15), conf.int = FALSE, col = myCols, 
+     # xlim = c(0,15), 
+     xmax = max(new_time),
+     conf.int = FALSE, col = myCols, 
      lwd = 2,
@@
 lines(risks$weibull[,1], risks$weibull[,3], col = myCols[2], lty = 2, lwd = 2)  
-text(x = 15*0.97, y = 0.05/100, labels = "(3)", cex = 1.5)
+text(x = max(new_time)*0.97, y = 0.05/100, labels = "(3)", cex = 1.5)
 mtext("Cumulative incidence (%)", side = 2, line = -1.5, outer = TRUE)
@@
       xaxt = 'n', yaxt = 'n',
-     xlim = c(0,15), conf.int = FALSE, col = myCols, 
+     # xlim = c(0,15), 
+     xmax = max(new_time),
+     conf.int = FALSE, col = myCols, 
      lwd = 2,
@@
 lines(risks$gompertz[,1], risks$gompertz[,3], col = myCols[2], lty = 2, lwd = 2)
-text(x = 15*0.97, y = 0.05/100, labels = "(2)", cex = 1.5)
+text(x = max(new_time)*0.97, y = 0.05/100, labels = "(2)", cex = 1.5)
 
@@
      yaxt = 'n',
-     xlim = c(0,15), conf.int = FALSE, col = myCols, 
+     # xlim = c(0,15), 
+     xmax = max(new_time),
+     conf.int = FALSE, col = myCols, 
      lwd = 2,
@@
 lines(risks$splines[,1], risks$splines[,3], col = myCols[2], lty = 2, lwd = 2)
-text(x = 15*0.97, y = 0.05/100, labels = "(4)", cex = 1.5)
+text(x = max(new_time)*0.97, y = 0.05/100, labels = "(4)", cex = 1.5)
 mtext("Years since randomization", side = 1, line = -1.5, outer = TRUE)
@@
 
-```{r erspc-casebase-cif, eval = FALSE, echo = FALSE}
-# smooth_risk_exp <- casebase::absoluteRisk(
-#   object = casebase_exponential,
-#   time = seq(0, 15, 0.1),
-#   newdata = new_data
-# )
-# 
-# plot(survfit(cox_model, newdata = new_data),
-#   xlab = "Years since Randomization",
-#   ylab = "Cumulative Incidence (%)",
-#   fun = "event",
-#   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
-#   main = sprintf(
-#     "Estimated Cumulative Incidence (risk) of Death from Prostate 
-#                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-#     exp(coef(cox_model)),
-#     exp(confint(cox_model))[1],
-#     exp(confint(cox_model))[2]
-#   ),
-#   ylim = c(0, 2), yscale = 100
-# )
-# lines(smooth_risk_exp[, 1], smooth_risk_exp[, 2], col = "red", lty = 2)
-# lines(smooth_risk_exp[, 1], smooth_risk_exp[, 3], col = "blue", lty = 2)
-# 
-# legend("topleft",
-#   legend = c(
-#     "Control group (Cox)", "Control group (Casebase)",
-#     "Screening group (Cox)", "Screening group (Casebase)"
-#   ),
-#   col = c("red", "red", "blue", "blue"),
-#   lty = c(1, 2, 1, 2),
-#   bg = "gray90"
-# )
-```
-
 <!-- The \code{fitSmoothHazard} function provides an estimate of the hazard function $h(x, t)$ is the hazard function, where $t$ denotes the numerical value of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any).
@@
 <!-- $$ CI(x, t) = 1 - \exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$ -->
-
-```{r erspc-casebase-weibull-cif, eval = FALSE, echo = FALSE}
-# smooth_risk_time <- casebase::absoluteRisk(
-#   object = casebase_time,
-#   time = seq(0, 15, 0.1),
-#   newdata = new_data
-# )
-# 
-# plot(survfit(cox_model, newdata = new_data),
-#   xlab = "Years since Randomization",
-#   ylab = "Cumulative Incidence (%)",
-#   fun = "event",
-#   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
-#   main = sprintf(
-#     "Estimated Cumulative Incidence (risk) of Death from Prostate 
-#                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-#     exp(coef(cox_model)),
-#     exp(confint(cox_model))[1],
-#     exp(confint(cox_model))[2]
-#   ),
-#   ylim = c(0, 2), yscale = 100
-# )
-# lines(smooth_risk_time[, 1], smooth_risk_time[, 2], col = "red", lty = 2)
-# lines(smooth_risk_time[, 1], smooth_risk_time[, 3], col = "blue", lty = 2)
-# 
-# legend("topleft",
-#   legend = c(
-#     "Control group (Cox)", "Control group (Casebase)",
-#     "Screening group (Cox)", "Screening group (Casebase)"
-#   ),
-#   col = c("red", "red", "blue", "blue"),
-#   lty = c(1, 2, 1, 2),
-#   bg = "gray90"
-# )
-```
 
@@
 # CI_splines <- c(exp(coef(fits$splines))["ScrArmScreening group"], exp(confint(fits$splines))["ScrArmScreening group",])
+
+aft_fits <- lapply(list("exp","gompertz","weibull"), function(i)
+  flexsurv::flexsurvreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm,
+                           data = ERSPC, dist = i)
+  )
+
+names(aft_fits) <- list("exponential","gompertz","weibull")
 
@@
 
-```{r erspc-casebase-splines-cif, eval = FALSE, echo = FALSE}
-# smooth_risk_splines <- absoluteRisk(
-#   object = casebase_splines,
-#   time = seq(0, 15, 0.1),
-#   newdata = new_data
-# )
-# 
-# plot(survfit(cox_model, newdata = new_data),
-#   xlab = "Years since Randomization",
-#   ylab = "Cumulative Incidence (%)",
-#   fun = "event",
-#   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
-#   main = sprintf(
-#     "Estimated Cumulative Incidence (risk) of Death from Prostate 
-#                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-#     exp(coef(cox_model)),
-#     exp(confint(cox_model))[1],
-#     exp(confint(cox_model))[2]
-#   ),
-#   ylim = c(0, 2), yscale = 100
-# )
-# lines(smooth_risk_splines[, 1], smooth_risk_splines[, 2], col = "red", lty = 2)
-# lines(smooth_risk_splines[, 1], smooth_risk_splines[, 3], col = "blue", lty = 2)
-# 
-# legend("topleft",
-#   legend = c(
-#     "Control group (Cox)", "Control group (Casebase)",
-#     "Screening group (Cox)", "Screening group (Casebase)"
-#   ),
-#   col = c("red", "red", "blue", "blue"),
-#   lty = c(1, 2, 1, 2),
-#   bg = "gray90"
-# )
-```
-
-In the following table, we present a side-by-side comparison of the hazard ratios and confidence intervals estimated from \code{fitSmoothHazard} and the corresponding parametric model using \code{survival::survreg}, as well as the Cox model estimate.
+In Table \ref{tab:print-erspc-estimates}, we present a side-by-side comparison of the hazard ratios and confidence intervals estimated from \code{fitSmoothHazard} and the corresponding parametric model using \code{survival::survreg}, as well as the Cox model estimate.
 The hazard ratio estimates and confidence intervals are similar across all four models.
@@
 ```{r erspc-95ci, echo = FALSE, eval = FALSE}
-CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
-CI_exp <- c(exp(coef(casebase_exponential))["ScrArmScreening group"], exp(confint(casebase_exponential))["ScrArmScreening group", ])
-CI_time <- c(exp(coef(casebase_time))["ScrArmScreening group"], exp(confint(casebase_time))["ScrArmScreening group", ])
-CI_splines <- c(exp(coef(casebase_splines))["ScrArmScreening group"], exp(confint(casebase_splines))["ScrArmScreening group", ])
-
-table_ci <- rbind(CI_cox, CI_exp, CI_time, CI_splines)
-colnames(table_ci) <- c("HR", "LB", "UB")
-
-data.frame(Model = c("Cox", "Exponential", "Weibull", "Splines")) %>%
-  bind_cols(as.data.frame(table_ci)) %>%
-  mutate(
-    CI = paste0(
-      "(", round(LB, 2), ", ",
-      round(UB, 2), ")"
-    ),
-    Width = UB - LB
-  ) %>%
-  dplyr::select(-LB, -UB) %>%
-  knitr::kable(format = "latex", booktabs = TRUE) %>%
-  kable_styling()
-```
-
-With this first case study, we explored how \pkg{casebase} allows us to fit different parametric survival models, and how we can compare the fit of each model with tools from GLMs.
-
-```{r interaction-ERSPC, eval = eval_cs1}
-fit_inter <- casebase::fitSmoothHazard(DeadOfPrCa ~ poly(Follow.Up.Time, 2)*ScrArm, data = ERSPC)
-summary(fit_inter)
+# CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
+# CI_exp <- c(exp(coef(casebase_exponential))["ScrArmScreening group"], exp(confint(casebase_exponential))["ScrArmScreening group", ])
+# CI_time <- c(exp(coef(casebase_time))["ScrArmScreening group"], exp(confint(casebase_time))["ScrArmScreening group", ])
+# CI_splines <- c(exp(coef(casebase_splines))["ScrArmScreening group"], exp(confint(casebase_splines))["ScrArmScreening group", ])
+# 
+# table_ci <- rbind(CI_cox, CI_exp, CI_time, CI_splines)
+# colnames(table_ci) <- c("HR", "LB", "UB")
+# 
+# data.frame(Model = c("Cox", "Exponential", "Weibull", "Splines")) %>%
+#   bind_cols(as.data.frame(table_ci)) %>%
+#   mutate(
+#     CI = paste0(
+#       "(", round(LB, 2), ", ",
+#       round(UB, 2), ")"
+#     ),
+#     Width = UB - LB
+#   ) %>%
+#   dplyr::select(-LB, -UB) %>%
+#   knitr::kable(format = "latex", booktabs = TRUE) %>%
+#   kable_styling()
+```
+
+Previous re-analyses of these data suggested that the overall screening attributed reduction in death due to prostate cancer of 20% was an underestimate [@hanley2010mortality].
+The estimated 20% (from a proportional hazards model) did not account for the delay between screening and the time the effect is expected to be observed.
+As a result, the null effects in years 1–7 masked the substantial reductions that began to appear from year 8 onwards.
+This motivates the use of a time-dependent hazard ratio which can easily be fit with the \pkg{casebase} package by including an interaction term with time in the model:
+
+```{r, echo=-1, eval = eval_cs1}
+set.seed(12)
+fit_inter <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) * ScrArm, data = ERSPC)
+```
+
+In Figure \ref{fig:interaction-ERSPC} we plot the estimated death from prostate cancer hazard ratio and 95\\% confidence interval for screening vs.
+control group as a function of time using the \code{plot} method for objects of class \code{singleEventCB}:
+
+```{r, echo = TRUE, eval = FALSE}
+plot(fit_inter, type = "hr", newdata = new_data,
+     var = "ScrArm", xvar = "Follow.Up.Time", ci = TRUE)
+```
+
+The plot shows that the cures attributable to the screening only begin to become statistically apparent by year 7 and later.
+The 25-60% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20% reported in the original article.
+
+```{r interaction-ERSPC, eval = eval_cs1, fig.width=8, fig.height=6, fig.cap='Estimated death from prostate cancer hazard ratio and 95\\% confidence interval for screening vs.
+control group as a function of time in the ERSPC dataset.
+Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic B-spline basis of follow-up time and treatment arm.
+95\\% confidence intervals are calculated using the delta method.
+The plot shows that the cures attributable to the screening only begin to become statistically apparent by year 7 and later.
+The 25-60\\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\\% reported in the original article.'}
+
+# library(casebase)
+# library(splines)
+# data("ERSPC")
+# ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
+#                        levels = c(0,1), 
+#                        labels = c("Control group", "Screening group"))
+# set.seed(12345)
+# fit_inter <- casebase::fitSmoothHazard(DeadOfPrCa ~ poly(Follow.Up.Time,2)*ScrArm, data = ERSPC)
+# summary(fit_inter)
+
+# library(visreg)
+# plot(fit_inter,
+#      type = "hazard",
+#      hazard.params = list(xvar = "Follow.Up.Time",
+#                           by = "ScrArm",
+#                           alpha = 0.05,
+#                           ylab = "Hazard"))
 
 # reference category
-newdata <- data.frame(ScrArm = factor("Control group",
+#dev.off()
+new_time <- seq(1, 12, by  = 0.1)
+new_data <- data.frame(ScrArm = factor("Control group",
                                          levels = c("Control group","Screening group")),
-                      Follow.Up.Time = seq(1,12,by =0.1))
-str(newdata)
-
-plot(fit_inter,
+                      Follow.Up.Time = new_time)
+
+par(oma = c(0,1,0,8))
+tt <- plot(fit_inter,
      type = "hr",
-     newdata = newdata,
+     newdata = new_data,
      var = "ScrArm",
@@
      ci = T,
+     xlab = "Follow-up time (years)",
+     ylab = "Death from prostate cancer hazard ratio",
+     ylim = c(0,1.50),
+     yaxt = 'n',
+     xaxt = "n",
      rug = T)
-abline(a=1,b=0, lty = 2)
-```
+# abline(a=1,b=0, lty = 2)
+axis(2, at = seq(0,1.50, by=0.25), las = 2)
+axis(1, at = seq(min(new_time),max(new_time), by=1))
+for( i in seq(0,1.50, by=0.25)) {
+  abline(h = i, lty = 1, col = "lightgrey")
+}
+for( i in seq(min(new_time),max(new_time), by=1)) {
+  abline(v = i, lty = 1, col = "lightgrey")
+}
+for(i in seq_along(seq(0, 0.75, by = 0.25))) {
+  mtext(paste0(seq(0, 0.75, by = 0.25)[i]*100,"%"), side = 4, las = 2, adj = 1, outer = TRUE, line = 2, at = c(0.63-0.11*(i-1)))
+}
+lines(tt$Follow.Up.Time, tt$hazard_ratio, lwd = 2, lty = 1)
+mtext("% reduction in\nprostate cancer\nmortality rate", side = 4, las = 2, outer = TRUE, line = -1, at = 0.7)
+```
+
+With this first case study, we explored how \pkg{casebase} allows us to fit different parametric survival models with possible time-varying effects, and how we can compare the fit of each model with tools from GLMs.
+
+\newpage
 

### 2020-08-20 analysis/paper/paper.tex
included text for time-dependent Hazard Ratio

--- 2020-08-07 analysis/paper/paper.tex
@@
 \author{
-Sahir Rai Bhatnagar*\\McGill University \And Maxime Turgeon*\\University
-of Manitoba \AND Jesse Islam\\McGill University \And James A.
+Sahir Rai Bhatnagar*\\McGill University \And Maxime Turgeon*\\University of Manitoba \AND Jesse Islam\\McGill University \And James A.
 Hanley\\McGill University \And Olli Saarela\\University of Toronto
 }
-\title{\pkg{casebase}: An Alternative Framework For Survival Analysis
-and Comparison of Event Rates}
-
-\Plainauthor{Sahir Rai Bhatnagar*, Maxime Turgeon*, Jesse Islam, James
-A. Hanley, Olli Saarela}
-\Plaintitle{casebase: An Alternative Framework For Survival Analysis and
+\title{\pkg{casebase}: An Alternative Framework For Survival Analysis and
 Comparison of Event Rates}
-\Shorttitle{\pkg{casebase}: An Alternative Framework For Survival
-Analysis}
+
+\Plainauthor{Sahir Rai Bhatnagar*, Maxime Turgeon*, Jesse Islam, James A. Hanley, Olli Saarela}
+\Plaintitle{casebase: An Alternative Framework For Survival Analysis and Comparison
+of Event Rates}
+\Shorttitle{\pkg{casebase}: An Alternative Framework For Survival Analysis}
 
@@
 
-The case-base sampling approach described in Section \ref{theory} can be
-visualized in the form of a population time plot. These plots are
-extremely informative graphical displays of survival data and should be
-one of the first steps in an exploratory data analysis. The
-\code{popTime} function and \code{plot} method facilitate this task:
+Population-time plots are a descriptive visualization of incidence
+density, where aggregate person-time is represented by area and events
+as points within the area. The case-base sampling approach described in
+Section \ref{theory} can be visualized in the form of a population time
+plot. These plots are extremely informative graphical displays of
+survival data and should be one of the first steps in an exploratory
+data analysis. The \code{popTime} function and \code{plot} method
+facilitate this task:
 
@@
 
-To illustrate these functions, we will use the European Randomized Study
-of Prostate Cancer Screening (ERSPC) data \citep{schroder2009screening}
-which was extracted using the approach described in Liu \emph{et al.}
-\citeyearpar{liu2014recovering} This dataset is available through the
-\pkg{casebase} package. It contains the individual observations for
-159,893 men from seven European countries, who were between the ages of
-55 and 69 years when recruited for the trial.
+To illustrate these functions, we will use data from the European
+Randomized Study of Prostate Cancer Screening (ERSPC)
+\citep{schroder2009screening} which was extracted using the approach
+described in Liu \emph{et al.} \citeyearpar{liu2014recovering} This
+dataset is available through the \pkg{casebase} package. It contains the
+individual observations for 159,893 men from seven European countries,
+who were between the ages of 55 and 69 years when recruited for the
+trial.
 
@@
 \begin{CodeInput}
-R> pt_object <- casebase::popTime(ERSPC, 
-R+                                time = "Follow.Up.Time",
-R+                                event = "DeadOfPrCa",
-R+                                exposure = "ScrArm")
+R> pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
+R+                                event = "DeadOfPrCa", exposure = "ScrArm")
 R> inherits(pt_object, "popTime")
@@
 
-\caption[Population time plot for ERSPC data]{Population time plot for ERSPC data.}\label{fig:plot-stratified-erspc-data}
+\caption[Population time plots for both treatment arms in the ERSPC dataset]{Population time plots for both treatment arms in the ERSPC dataset.
+The grey area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely small rectangles (person-moments).
+The points shown in the figure are select person-moments from the case and base series.
+We see that more events are observed at later follow-up times, motivating the use of a non-constant in time hazard model.}\label{fig:plot-stratified-erspc-data}
 \end{figure}
@@
 
-Next, layers for the case series and base series are added. The case
-series is sampled at random vertically on the plot to avoid having all
-points along the upper edge of the gray area. By randomly distributing
-the cases, we can get a better sense of the incidence density. In Figure
-\ref{fig:plot-stratified-erspc-data}, we see that more events are
-observed at later follow-up times. Therefore, a constant hazard model
-would not be appropriate in this instance as it would overestimate the
-cumulative incidence earlier on in time, and underestimate it later on.
-Finally, the base series is sampled horizontally with sampling weight
-proportional to their follow-up time. The reader should refer to the
-package vignettes for more examples and a detailed description of how to
-modify the aesthetics of a population-time plot.
+Next, layers for the case series and base series are added. The y-axis
+location of each case moment is sampled at random vertically on the plot
+to avoid having all points along the upper edge of the gray area. By
+randomly distributing the cases, we can get a sense of the incidence
+density. In Figure \ref{fig:plot-stratified-erspc-data}, we see that
+more events are observed at later follow-up times. Therefore, a constant
+hazard model would not be appropriate in this instance as it would
+overestimate the cumulative incidence earlier on in time, and
+underestimate it later on. Finally, the base series is sampled
+horizontally with sampling weight proportional to their follow-up time.
+The reader should refer to the package vignettes for more examples and a
+detailed description of how to modify the aesthetics of a
+population-time plot.
 
@@
 For our first case study, we return to the ERSPC study and investigate
-the differences in survival between the control and screening arms. We
-fit four models that differ in which functional form of time is used: 1)
+the differences in risk between the control and screening arms. We fit
+four models that differ in which functional form of time is used: 1)
 excluded from the linear predictor, 2) linear function, 3) log function,
 and 4) a smooth function using cubic B-splines. The models are fit using
-\code{fitSmoothHazard}. Since the output object from
-\code{fitSmoothHazard} inherits from the \code{glm} class, we can
-directly use the function \code{summary}:
+\code{fitSmoothHazard} with the familiar formula interface:
 
@@
 R> fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
-R>              gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
-R>              weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
-R>              splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
+R+              gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
+R+              weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
+R+              splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
 R> 
 R> fits <- lapply(fmla, function(i) {
-R>   casebase::fitSmoothHazard(i, data = ERSPC, ratio = 100)
-R> })
-R> 
-R> summary(fits[["gompertz"]])
+R+   fitSmoothHazard(i, data = ERSPC, ratio = 100)
+R+ })
 \end{CodeInput}
@@
 
+The output object from \code{fitSmoothHazard} inherits from the
+\code{singleEventCB} and \code{glm} classes. As such, we can directly
+use the \code{summary} generic:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> summary(fits[["splines"]]) 
+\end{CodeInput}
+\end{CodeChunk}
+
 \begin{CodeChunk}
@@
 \begin{CodeOutput}
-#> 
-#> Call:
-#> glm(formula = formula, family = binomial, data = sampleData)
-#> 
-#> Deviance Residuals: 
-#>    Min      1Q  Median      3Q     Max  
-#> -0.378  -0.162  -0.122  -0.094   3.468  
 #> 
@@
 #>                       Estimate Std. Error z value Pr(>|z|)
-#> (Intercept)            -9.0335     0.1122  -80.54 <0.0000000000000002 ***
-#> Follow.Up.Time          0.2228     0.0145   15.32 <0.0000000000000002 ***
-#> ScrArmScreening group  -0.2340     0.0886   -2.64              0.0083 ** 
-#> ---
-#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> (Intercept)           -10.2905     0.3185  -32.31  < 2e-16
+#> bs(Follow.Up.Time)1     4.3587     0.8008    5.44  5.2e-08
+#> bs(Follow.Up.Time)2     2.0890     0.4721    4.42  9.7e-06
+#> bs(Follow.Up.Time)3     4.6266     0.6910    6.70  2.1e-11
+#> ScrArmScreening group  -0.2354     0.0886   -2.66   0.0079
 #> 
@@
 #>     Null deviance: 6059.0  on 54539  degrees of freedom
-#> Residual deviance: 5810.1  on 54537  degrees of freedom
-#> AIC: 5816
+#> Residual deviance: 5785.3  on 54535  degrees of freedom
+#> AIC: 5795
 #> 
-#> Number of Fisher Scoring iterations: 8
+#> Number of Fisher Scoring iterations: 9
 \end{CodeOutput}
@@
 R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
+R> new_time <- seq(0,14,0.1)
 R> 
 R> risks <- lapply(fits, function(i) {
-R>   casebase::absoluteRisk(object = i, time = seq(0,15,0.1), newdata = new_data)
+R>   absoluteRisk(object = i, time = new_time, newdata = new_data, method = "mont")
 R> })
-R> 
-R> head(risks[["gompertz"]])
 \end{CodeInput}
@@
 discussion of the population-time plot, this poor fit for the
-exponential hazard was expected.
-
-In Figure \ref{fig:erspc-cox-cif} (panels 2--4), we notice a better fit
-with increasing complexity of our model for time. As noted above, the
-usual asymptotic results hold for likelihood ratio tests built using
-case-base sampling models. Therefore, we can easily test the null
-hypothesis that the exponential model is just as good as the larger (in
-terms of number of parameters) splines model:
+exponential hazard was expected. We notice a better fit with increasing
+complexity of our model for time in Figure \ref{fig:erspc-cox-cif}
+(panels 2--4). As noted above, the usual asymptotic results hold for
+likelihood ratio tests built using case-base sampling models. Therefore,
+we can easily test the null hypothesis that the Weibull model is just as
+good as the larger (in terms of number of parameters) spline model:
 
@@
 #> 
-#> Model 1: DeadOfPrCa ~ ScrArm + offset(offset)
+#> Model 1: DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm + offset(offset)
 #> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
 #>   Resid. Df Resid. Dev Df Deviance Pr(>Chi)    
-#> 1     54538       6053                                    
-#> 2     54535       5787  3      266 <0.0000000000000002 ***
+#> 1     54537       5801                         
+#> 2     54535       5785  2     15.7    4e-04 ***
 #> ---
@@
 
-As expected, we see that splines model provides a better fit. Similarly,
-we can compare the AIC for all four case-base models:
+The null hypothesis is rejected in favor of the spline model. Similarly,
+the AIC provides further evidence that the flexible function of time
+provides the best fit:
 
@@
 \begin{CodeOutput}
-#>    Exp.   Gomp.   Weib. Splines 
-#>    6057    5816    6057    5797
+#>     Exp. Gompertz  Weibull  Splines 
+#>     6056     5821     5807     5795
 \end{CodeOutput}
@@
 
-Once again, we have evidence that the splines model provides the best
-fit.
-
 \begin{CodeChunk}
 \begin{figure}
-
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/erspc-cox-cif-1}
-
-}
-
-\caption{CIFs for control and screening groups in the ERSPC data.
+\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/erspc-cox-cif-1} \caption{CIFs for control and screening groups in the ERSPC data.
 In each of the panels, we plot the CIF from the Cox model using \code{survival::survfit} (solid line) and the CIF from the case-base sampling scheme (dashed line) with different functional forms of time.
@@
 
-In the following table, we present a side-by-side comparison of the
-hazard ratios and confidence intervals estimated from
+In Table \ref{tab:print-erspc-estimates}, we present a side-by-side
+comparison of the hazard ratios and confidence intervals estimated from
 \code{fitSmoothHazard} and the corresponding parametric model using
@@
 
+Previous re-analyses of these data suggested that the overall screening
+attributed reduction in death due to prostate cancer of 20\% was an
+underestimate \citep{hanley2010mortality}. The estimated 20\% (from a
+proportional hazards model) did not account for the delay between
+screening and the time the effect is expected to be observed. As a
+result, the null effects in years 1--7 masked the substantial reductions
+that began to appear from year 8 onwards. This motivates the use of a
+time-dependent hazard ratio which can easily be fit with the
+\pkg{casebase} package by including an interaction term with time in the
+model:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> fit_inter <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) * ScrArm, data = ERSPC)
+\end{CodeInput}
+\end{CodeChunk}
+
+In Figure \ref{fig:interaction-ERSPC} we plot the estimated death from
+prostate cancer hazard ratio and 95\textbackslash\% confidence interval
+for screening vs.~control group as a function of time using the
+\code{plot} method for objects of class \code{singleEventCB}:
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> plot(fit_inter, type = "hr", newdata = new_data,
+R>      var = "ScrArm", xvar = "Follow.Up.Time", ci = TRUE)
+\end{CodeInput}
+\end{CodeChunk}
+
+The plot shows that the cures attributable to the screening only begin
+to become statistically apparent by year 7 and later. The 25-60\%
+reductions seen in years 8-12 of the study suggests a much higher
+reduction in prostate cancer due to screening than the single overall
+20\% reported in the original article.
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/interaction-ERSPC-1}
+
+}
+
+\caption[Estimated death from prostate cancer hazard ratio and 95\% confidence interval for screening vs]{Estimated death from prostate cancer hazard ratio and 95\% confidence interval for screening vs.
+control group as a function of time in the ERSPC dataset.
+Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic B-spline basis of follow-up time and treatment arm.
+95\% confidence intervals are calculated using the delta method.
+The plot shows that the cures attributable to the screening only begin to become statistically apparent by year 7 and later.
+The 25-60\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\% reported in the original article.}\label{fig:interaction-ERSPC}
+\end{figure}
+\end{CodeChunk}
+
 With this first case study, we explored how \pkg{casebase} allows us to
-fit different parametric survival models, and how we can compare the fit
-of each model with tools from GLMs.
+fit different parametric survival models with possible time-varying
+effects, and how we can compare the fit of each model with tools from
+GLMs.
+
+\newpage
 
@@
 \midrule
-Sex & 0.70 & (0.4, 1.23) & 0.68 & (0.39, 1.19)\\
-Disease & 0.53 & (0.29, 0.95) & 0.52 & (0.29, 0.93)\\
-Phase (CR2 vs. CR1) & 1.12 & (0.45, 2.78) & 1.21 & (0.48, 3.02)\\
-Phase (CR3 vs. CR1) & 1.45 & (0.37, 5.66) & 1.67 & (0.43, 6.5)\\
-Phase (Relapse vs. CR1) & 4.04 & (1.88, 8.68) & 4.55 & (2.09, 9.9)\\
+Sex & 0.72 & (0.42, 1.26) & 0.68 & (0.39, 1.19)\\
+Disease & 0.54 & (0.3, 0.97) & 0.52 & (0.29, 0.93)\\
+Phase (CR2 vs. CR1) & 1.21 & (0.48, 3.02) & 1.21 & (0.48, 3.02)\\
+Phase (CR3 vs. CR1) & 1.63 & (0.41, 6.44) & 1.67 & (0.43, 6.5)\\
+Phase (Relapse vs. CR1) & 4.43 & (2.06, 9.55) & 4.55 & (2.09, 9.9)\\
 \addlinespace
-Source & 1.48 & (0.48, 4.54) & 1.46 & (0.47, 4.54)\\
+Source & 1.51 & (0.49, 4.69) & 1.46 & (0.47, 4.54)\\
 Age & 0.99 & (0.97, 1.01) & 0.99 & (0.97, 1.02)\\
@@
 #> Model1 Model2 Model3 
-#>    827    790    788
+#>    827    793    792
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> [1] "absRiskCB" "matrix"    "array"
+#> [1] "absRiskCB" "matrix"
 \end{CodeOutput}
@@
 
-\begin{center}\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/unnamed-chunk-9-1} \end{center}
+\begin{center}\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/unnamed-chunk-11-1} \end{center}
 
@@
 
-This report was generated on 2020-08-06 17:31:23 using the following
+This report was generated on 2020-08-20 19:21:21 using the following
 computational environment and dependencies:
@@
 \begin{CodeOutput}
-#> R version 4.0.2 (2020-06-22)
-#> Platform: x86_64-apple-darwin17.0 (64-bit)
-#> Running under: macOS Catalina 10.15.6
+#> R version 3.6.2 (2019-12-12)
+#> Platform: x86_64-pc-linux-gnu (64-bit)
+#> Running under: Pop!_OS 19.10
 #> 
 #> Matrix products: default
-#> BLAS:   /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRblas.dylib
-#> LAPACK: /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRlapack.dylib
+#> BLAS:   /usr/lib/x86_64-linux-gnu/openblas/libblas.so.3
+#> LAPACK: /usr/lib/x86_64-linux-gnu/libopenblasp-r0.3.7.so
 #> 
 #> locale:
-#> [1] en_CA.UTF-8/en_CA.UTF-8/en_CA.UTF-8/C/en_CA.UTF-8/en_CA.UTF-8
+#>  [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
+#>  [3] LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
+#>  [5] LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
+#>  [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                 
+#>  [9] LC_ADDRESS=C               LC_TELEPHONE=C            
+#> [11] LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       
 #> 
@@
 #> other attached packages:
-#>  [1] broom_0.5.6               dotwhisker_0.5.0         
-#>  [3] timereg_1.9.4             glue_1.4.1               
-#>  [5] scales_1.1.1              forcats_0.5.0            
-#>  [7] stringr_1.4.0             dplyr_1.0.0              
-#>  [9] purrr_0.3.4               readr_1.3.1              
-#> [11] tidyr_1.0.2               ggplot2_3.3.2            
-#> [13] tidyverse_1.3.0           lubridate_1.7.8          
-#> [15] visreg_2.6.1              tibble_3.0.3             
-#> [17] riskRegression_2020.02.05 prodlim_2019.11.13       
-#> [19] flexsurv_1.1.1            survival_3.1-12          
-#> [21] pracma_2.2.9              magrittr_1.5             
-#> [23] kableExtra_1.1.0          glmnet_3.0-2             
-#> [25] Matrix_1.2-18             data.table_1.12.8        
-#> [27] cowplot_1.0.0             colorspace_1.4-1         
-#> [29] casebase_0.9.0.9000      
+#>  [1] scales_1.1.1              forcats_0.5.0            
+#>  [3] stringr_1.4.0             dplyr_1.0.0              
+#>  [5] purrr_0.3.4               readr_1.3.1              
+#>  [7] tidyr_1.1.0               ggplot2_3.3.2.9000       
+#>  [9] tidyverse_1.3.0           lubridate_1.7.4          
+#> [11] visreg_2.7.0              tibble_3.0.1             
+#> [13] riskRegression_2020.02.05 prodlim_2019.11.13       
+#> [15] flexsurv_1.1.1            survival_3.1-8           
+#> [17] pracma_2.2.9              magrittr_1.5             
+#> [19] kableExtra_1.1.0          glmnet_4.0-2             
+#> [21] Matrix_1.2-18             data.table_1.12.8        
+#> [23] cowplot_1.0.0             colorspace_1.4-1         
+#> [25] casebase_0.9.0           
 #> 
 #> loaded via a namespace (and not attached):
-#>  [1] TH.data_1.0-10      VGAM_1.1-2          ellipsis_0.3.1     
-#>  [4] ggstance_0.3.4      htmlTable_1.13.3    base64enc_0.1-3    
-#>  [7] fs_1.4.1            rstudioapi_0.11     farver_2.0.3       
-#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-0      
-#> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
-#> [16] Formula_1.2-3       jsonlite_1.6.1      cluster_2.1.0      
-#> [19] dbplyr_1.4.3        png_0.1-7           compiler_4.0.2     
-#> [22] httr_1.4.1          backports_1.1.8     assertthat_0.2.1   
-#> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
-#> [28] quantreg_5.55       tools_4.0.2         gtable_0.3.0       
-#> [31] Rcpp_1.0.5          cellranger_1.1.0    vctrs_0.3.2        
-#> [34] nlme_3.1-148        iterators_1.0.12    xfun_0.15          
+#>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
+#>  [4] htmlTable_1.13.3    base64enc_0.1-3     fs_1.3.2           
+#>  [7] rstudioapi_0.11     MatrixModels_0.4-1  fansi_0.4.1        
+#> [10] mvtnorm_1.0-12      xml2_1.3.0          codetools_0.2-16   
+#> [13] knitr_1.29          Formula_1.2-3       jsonlite_1.7.0     
+#> [16] broom_0.7.0         cluster_2.1.0       dbplyr_1.4.2       
+#> [19] png_0.1-7           compiler_3.6.2      httr_1.4.1         
+#> [22] backports_1.1.8     assertthat_0.2.1    cli_2.0.2          
+#> [25] acepack_1.4.1       htmltools_0.5.0     quantreg_5.54      
+#> [28] tools_3.6.2         gtable_0.3.0        glue_1.4.1         
+#> [31] Rcpp_1.0.4.6        cellranger_1.1.0    vctrs_0.3.1        
+#> [34] nlme_3.1-143        iterators_1.0.12    xfun_0.15          
 #> [37] rvest_0.3.5         lifecycle_0.2.0     polspline_1.1.17   
-#> [40] muhaz_1.2.6.1       MASS_7.3-51.6       zoo_1.8-8          
+#> [40] muhaz_1.2.6.1       MASS_7.3-51.5       zoo_1.8-7          
 #> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
-#> [46] RColorBrewer_1.1-2  rticles_0.14        yaml_2.2.1         
+#> [46] RColorBrewer_1.1-2  rticles_0.13        yaml_2.2.1         
 #> [49] gridExtra_2.3       rms_5.1-4           rpart_4.1-15       
 #> [52] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
-#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
-#> [58] rlang_0.4.7         pkgconfig_2.0.3     evaluate_0.14      
-#> [61] lattice_0.20-41     labeling_0.3        htmlwidgets_1.5.1  
-#> [64] cmprsk_2.2-9        tidyselect_1.1.0    deSolve_1.28       
-#> [67] R6_2.4.1            generics_0.0.2      Hmisc_4.4-0        
-#> [70] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
-#> [73] pillar_1.4.6        haven_2.2.0         foreign_0.8-80     
-#> [76] mgcv_1.8-31         nnet_7.3-14         mstate_0.2.12      
-#> [79] modelr_0.1.6        crayon_1.3.4        rmarkdown_2.1      
-#> [82] jpeg_0.1-8.1        grid_4.0.2          readxl_1.3.1       
+#> [55] checkmate_2.0.0     lava_1.6.6          shape_1.4.4        
+#> [58] rlang_0.4.6         pkgconfig_2.0.3     evaluate_0.14      
+#> [61] lattice_0.20-38     htmlwidgets_1.5.1   cmprsk_2.2-9       
+#> [64] tidyselect_1.1.0    deSolve_1.27.1      R6_2.4.1           
+#> [67] generics_0.0.2      Hmisc_4.3-1         multcomp_1.4-12    
+#> [70] DBI_1.1.0           withr_2.2.0         pillar_1.4.4       
+#> [73] haven_2.3.1         foreign_0.8-74      mgcv_1.8-31        
+#> [76] nnet_7.3-12         mstate_0.2.12       modelr_0.1.5       
+#> [79] crayon_1.3.4        rmarkdown_2.3       timereg_1.9.4      
+#> [82] jpeg_0.1-8.1        grid_3.6.2          readxl_1.3.1       
 #> [85] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
-#> [88] numDeriv_2016.8-1.1 stats4_4.0.2        munsell_0.5.0      
+#> [88] numDeriv_2016.8-1.1 stats4_3.6.2        munsell_0.5.0      
 #> [91] viridisLite_0.3.0   quadprog_1.5-8
@@
 \begin{CodeOutput}
-#> Local:    Jesse-Islam-master /Users/maxturgeon/Documents/git_repos/cbpaper
-#> Head:     [32e6bf4] 2020-08-05: Merge pull request #25 from sahirbhatnagar/editMax
+#> Local:    fix-cs1 /home/sahir/git_repositories/cbpaper
+#> Remote:   fix-cs1 @ origin (https://github.com/sahirbhatnagar/cbpaper.git)
+#> Head:     [f618e21] 2020-08-18: time dependent hazard ratio for ERSPC
 \end{CodeOutput}

### 2020-08-21 analysis/paper/paper.tex

@@
 \midrule
-Exponential & 0.81 (0.68, 0.96) & 0.81 (0.68, 0.96)\\
+Exponential & 0.79 (0.67, 0.94) & 0.81 (0.68, 0.96)\\
 \cmidrule{1-3}
-Gompertz & 0.79 (0.66, 0.94) & 0.80 (0.67, 0.95)\\
+Gompertz & 0.80 (0.67, 0.95) & 0.80 (0.67, 0.95)\\
 \cmidrule{1-3}
@@
 \cmidrule{1-3}
-Splines & 0.81 (0.68, 0.96) & --\\
+Splines & 0.79 (0.66, 0.94) & --\\
 \bottomrule
@@
 \midrule
-Sex & 0.72 & (0.42, 1.26) & 0.68 & (0.39, 1.19)\\
-Disease & 0.54 & (0.3, 0.97) & 0.52 & (0.29, 0.93)\\
-Phase (CR2 vs. CR1) & 1.21 & (0.48, 3.02) & 1.21 & (0.48, 3.02)\\
-Phase (CR3 vs. CR1) & 1.63 & (0.41, 6.44) & 1.67 & (0.43, 6.5)\\
-Phase (Relapse vs. CR1) & 4.43 & (2.06, 9.55) & 4.55 & (2.09, 9.9)\\
+Sex & 0.74 & (0.43, 1.3) & 0.68 & (0.39, 1.19)\\
+Disease & 0.52 & (0.29, 0.94) & 0.52 & (0.29, 0.93)\\
+Phase (CR2 vs. CR1) & 1.20 & (0.48, 3) & 1.21 & (0.48, 3.02)\\
+Phase (CR3 vs. CR1) & 1.59 & (0.41, 6.17) & 1.67 & (0.43, 6.5)\\
+Phase (Relapse vs. CR1) & 4.09 & (1.9, 8.81) & 4.55 & (2.09, 9.9)\\
 \addlinespace
-Source & 1.51 & (0.49, 4.69) & 1.46 & (0.47, 4.54)\\
-Age & 0.99 & (0.97, 1.01) & 0.99 & (0.97, 1.02)\\
+Source & 1.70 & (0.55, 5.23) & 1.46 & (0.47, 4.54)\\
+Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
 \bottomrule
@@
 #> Model1 Model2 Model3 
-#>    827    793    792
+#>    827    791    790
 \end{CodeOutput}
@@
 
-This report was generated on 2020-08-20 19:21:21 using the following
+This report was generated on 2020-08-21 16:53:49 using the following
 computational environment and dependencies:
@@
 #> other attached packages:
-#>  [1] scales_1.1.1              forcats_0.5.0            
-#>  [3] stringr_1.4.0             dplyr_1.0.0              
-#>  [5] purrr_0.3.4               readr_1.3.1              
-#>  [7] tidyr_1.1.0               ggplot2_3.3.2.9000       
-#>  [9] tidyverse_1.3.0           lubridate_1.7.4          
-#> [11] visreg_2.7.0              tibble_3.0.1             
-#> [13] riskRegression_2020.02.05 prodlim_2019.11.13       
-#> [15] flexsurv_1.1.1            survival_3.1-8           
-#> [17] pracma_2.2.9              magrittr_1.5             
-#> [19] kableExtra_1.1.0          glmnet_4.0-2             
-#> [21] Matrix_1.2-18             data.table_1.12.8        
-#> [23] cowplot_1.0.0             colorspace_1.4-1         
-#> [25] casebase_0.9.0           
+#>  [1] broom_0.7.0               dotwhisker_0.5.0         
+#>  [3] timereg_1.9.4             glue_1.4.1               
+#>  [5] scales_1.1.1              forcats_0.5.0            
+#>  [7] stringr_1.4.0             dplyr_1.0.0              
+#>  [9] purrr_0.3.4               readr_1.3.1              
+#> [11] tidyr_1.1.0               ggplot2_3.3.2.9000       
+#> [13] tidyverse_1.3.0           lubridate_1.7.4          
+#> [15] visreg_2.7.0              tibble_3.0.1             
+#> [17] riskRegression_2020.02.05 prodlim_2019.11.13       
+#> [19] flexsurv_1.1.1            survival_3.1-8           
+#> [21] pracma_2.2.9              magrittr_1.5             
+#> [23] kableExtra_1.1.0          glmnet_4.0-2             
+#> [25] Matrix_1.2-18             data.table_1.12.8        
+#> [27] cowplot_1.0.0             colorspace_1.4-1         
+#> [29] casebase_0.9.0           
 #> 
@@
 #>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
-#>  [4] htmlTable_1.13.3    base64enc_0.1-3     fs_1.3.2           
-#>  [7] rstudioapi_0.11     MatrixModels_0.4-1  fansi_0.4.1        
-#> [10] mvtnorm_1.0-12      xml2_1.3.0          codetools_0.2-16   
-#> [13] knitr_1.29          Formula_1.2-3       jsonlite_1.7.0     
-#> [16] broom_0.7.0         cluster_2.1.0       dbplyr_1.4.2       
-#> [19] png_0.1-7           compiler_3.6.2      httr_1.4.1         
-#> [22] backports_1.1.8     assertthat_0.2.1    cli_2.0.2          
-#> [25] acepack_1.4.1       htmltools_0.5.0     quantreg_5.54      
-#> [28] tools_3.6.2         gtable_0.3.0        glue_1.4.1         
+#>  [4] ggstance_0.3.3      htmlTable_1.13.3    base64enc_0.1-3    
+#>  [7] fs_1.3.2            rstudioapi_0.11     farver_2.0.3       
+#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.0-12     
+#> [13] xml2_1.3.0          codetools_0.2-16    knitr_1.29         
+#> [16] Formula_1.2-3       jsonlite_1.7.0      cluster_2.1.0      
+#> [19] dbplyr_1.4.2        png_0.1-7           compiler_3.6.2     
+#> [22] httr_1.4.1          backports_1.1.8     assertthat_0.2.1   
+#> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
+#> [28] quantreg_5.54       tools_3.6.2         gtable_0.3.0       
 #> [31] Rcpp_1.0.4.6        cellranger_1.1.0    vctrs_0.3.1        
@@
 #> [58] rlang_0.4.6         pkgconfig_2.0.3     evaluate_0.14      
-#> [61] lattice_0.20-38     htmlwidgets_1.5.1   cmprsk_2.2-9       
-#> [64] tidyselect_1.1.0    deSolve_1.27.1      R6_2.4.1           
-#> [67] generics_0.0.2      Hmisc_4.3-1         multcomp_1.4-12    
-#> [70] DBI_1.1.0           withr_2.2.0         pillar_1.4.4       
-#> [73] haven_2.3.1         foreign_0.8-74      mgcv_1.8-31        
-#> [76] nnet_7.3-12         mstate_0.2.12       modelr_0.1.5       
-#> [79] crayon_1.3.4        rmarkdown_2.3       timereg_1.9.4      
-#> [82] jpeg_0.1-8.1        grid_3.6.2          readxl_1.3.1       
-#> [85] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
-#> [88] numDeriv_2016.8-1.1 stats4_3.6.2        munsell_0.5.0      
-#> [91] viridisLite_0.3.0   quadprog_1.5-8
+#> [61] lattice_0.20-38     labeling_0.3        htmlwidgets_1.5.1  
+#> [64] cmprsk_2.2-9        tidyselect_1.1.0    deSolve_1.27.1     
+#> [67] plyr_1.8.6          R6_2.4.1            generics_0.0.2     
+#> [70] Hmisc_4.3-1         multcomp_1.4-12     DBI_1.1.0          
+#> [73] withr_2.2.0         pillar_1.4.4        haven_2.3.1        
+#> [76] foreign_0.8-74      mgcv_1.8-31         nnet_7.3-12        
+#> [79] mstate_0.2.12       modelr_0.1.5        crayon_1.3.4       
+#> [82] rmarkdown_2.3       jpeg_0.1-8.1        grid_3.6.2         
+#> [85] readxl_1.3.1        reprex_0.3.0        digest_0.6.25      
+#> [88] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_3.6.2       
+#> [91] munsell_0.5.0       viridisLite_0.3.0   quadprog_1.5-8
 \end{CodeOutput}
@@
 #> Remote:   fix-cs1 @ origin (https://github.com/sahirbhatnagar/cbpaper.git)
-#> Head:     [f618e21] 2020-08-18: time dependent hazard ratio for ERSPC
+#> Head:     [349d153] 2020-08-21: fix merge conflicts
 \end{CodeOutput}

### 2020-08-25 analysis/paper/paper.Rmd
switch to plot method for hazard

@@
 
-As we can see, the best fit is the third model. By visualizing the hazard functions for both exposed and unexposed individuals, we can more clearly see how the hazards are no longer proportional.
-
-```{r stanford-hazard, eval = eval_cs4, echo=TRUE, dependson="stanford_fit"}
-# Compute hazards---
-# First, create a list of time points for both exposure status
-hazard_data <- expand.grid(
-  exposure = c(0, 1),
-  futime = seq(0, 425,
-    length.out = 100
-  )
-)
-
-# Set the offset to zero
-hazard_data$offset <- 0
-# Use predict to get the fitted values, and exponentiate to
-# transform to the right scale
-hazard_data$hazard <- exp(predict(fit3,
-  newdata = hazard_data,
-  type = "link"
-))
-# Add labels for plots
-hazard_data$Status <- factor(hazard_data$exposure,
-  labels = c("No Tx", "Tx")
-)
-
-ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
-  geom_line() +
-  paper_gg_theme +  
-  ylab("Hazard") +
-  xlab("Follow-up time (in days)")
+As we can see, the best fit is the third model.
+By visualizing the hazard functions for both exposed and unexposed individuals, we can more clearly see how the hazards are no longer proportional.
+We can easily obtain a plot of the hazards by using the `plot.singleEventCB` method:
+
+```{r stanford-hazard, eval = eval_cs4, echo=TRUE, dependson="stanford_fit", message = FALSE, results="hide"}
+plot(fit3, hazard.params = list(xvar = "futime",
+                                by = "exposure",
+                                alpha = 0.05,
+                                ylab = "Hazard",
+                                data = cb_data))
 ```
@@
 
-```{r stanford-risk, eval = eval_cs4, dependson="stanford_fit"}
+```{r stanford-risk, eval = eval_cs4, echo = TRUE, dependson="stanford_fit"}
 # Compute absolute risk curves
@@
   paper_gg_theme
-
-# colnames(absrisk) <- c("Time", "NoTrans", "Trans")
-# 
-# # Rearrange the data
-# absrisk <- gather(
-#   as.data.frame(absrisk),
-#   "Status", "Risk", -Time
-# )
-# 
-# ggplot(absrisk, aes(Time, Risk, colour = Status)) +
-#   geom_line() +
-#   theme_minimal() +
-#   theme(legend.position = "top") +
-#   expand_limits(y = 0.5) +
-#   xlab("Follow-up time") +
-#   ylab("Cum. Incidence")
 ```
@@
 
-We can also plot the hazard ratio:
-
-```{r, eval = eval_cs4, dependson="stanford_fit"}
+<!-- We can also plot the hazard ratio: -->
+
+```{r, eval = FALSE, dependson="stanford_fit"}
 newtime <- seq(0, 425, length.out = 100)

### 2020-08-25 analysis/paper/paper.tex
switch to plot method for hazard

@@
 \author{
-Sahir Rai Bhatnagar*\\McGill University \And Maxime Turgeon*\\University of Manitoba \AND Jesse Islam\\McGill University \And James A.
+Sahir Rai Bhatnagar*\\McGill University \And Maxime Turgeon*\\University
+of Manitoba \AND Jesse Islam\\McGill University \And James A.
 Hanley\\McGill University \And Olli Saarela\\University of Toronto
 }
-\title{\pkg{casebase}: An Alternative Framework For Survival Analysis and
+\title{\pkg{casebase}: An Alternative Framework For Survival Analysis
+and Comparison of Event Rates}
+
+\Plainauthor{Sahir Rai Bhatnagar*, Maxime Turgeon*, Jesse Islam, James
+A. Hanley, Olli Saarela}
+\Plaintitle{casebase: An Alternative Framework For Survival Analysis and
 Comparison of Event Rates}
-
-\Plainauthor{Sahir Rai Bhatnagar*, Maxime Turgeon*, Jesse Islam, James A. Hanley, Olli Saarela}
-\Plaintitle{casebase: An Alternative Framework For Survival Analysis and Comparison
-of Event Rates}
-\Shorttitle{\pkg{casebase}: An Alternative Framework For Survival Analysis}
+\Shorttitle{\pkg{casebase}: An Alternative Framework For Survival
+Analysis}
 
@@
 hazard functions for both exposed and unexposed individuals, we can more
-clearly see how the hazards are no longer proportional.
+clearly see how the hazards are no longer proportional. We can easily
+obtain a plot of the hazards by using the \texttt{plot.singleEventCB}
+method:
 
@@
 \begin{CodeInput}
-R> # Compute hazards---
-R> # First, create a list of time points for both exposure status
-R> hazard_data <- expand.grid(
-R+   exposure = c(0, 1),
-R+   futime = seq(0, 425,
-R+     length.out = 100
-R+   )
-R+ )
-R> 
-R> # Set the offset to zero
-R> hazard_data$offset <- 0
-R> # Use predict to get the fitted values, and exponentiate to
-R> # transform to the right scale
-R> hazard_data$hazard <- exp(predict(fit3,
-R+   newdata = hazard_data,
-R+   type = "link"
-R+ ))
-R> # Add labels for plots
-R> hazard_data$Status <- factor(hazard_data$exposure,
-R+   labels = c("No Tx", "Tx")
-R+ )
-R> 
-R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
-R+   geom_line() +
-R+   paper_gg_theme +  
-R+   ylab("Hazard") +
-R+   xlab("Follow-up time (in days)")
+R> plot(fit3, hazard.params = list(xvar = "futime",
+R+                                 by = "exposure",
+R+                                 alpha = 0.05,
+R+                                 ylab = "Hazard",
+R+                                 data = cb_data))
 \end{CodeInput}
@@
 
+\begin{CodeInput}
+R> # Compute absolute risk curves
+R> newdata <- data.frame(exposure = c(0, 1))
+R> absrisk <- absoluteRisk(fit3,
+R+   newdata = newdata,
+R+   time = seq(0, 425, length.out = 100)
+R+ )
+R> 
+R> class(absrisk)
+\end{CodeInput}
+
 \begin{CodeOutput}
-#> [1] "absRiskCB" "matrix"
+#> [1] "absRiskCB" "matrix"    "array"
 \end{CodeOutput}
 
+\begin{CodeInput}
+R> 
+R> plot(absrisk,
+R+      id.names = c("No Tx", "Tx")) + 
+R+   ylab("Cumulative Incidence") +
+R+   xlab("Follow-up time (in days)") +
+R+   paper_gg_theme
+\end{CodeInput}
+
 \begin{center}\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-risk-1} \end{center}
-
-\end{CodeChunk}
-
-We can also plot the hazard ratio:
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/unnamed-chunk-11-1} \end{center}
 
@@
 
-This report was generated on 2020-08-21 16:53:49 using the following
+This report was generated on 2020-08-25 10:58:13 using the following
 computational environment and dependencies:
@@
 \begin{CodeOutput}
-#> R version 3.6.2 (2019-12-12)
-#> Platform: x86_64-pc-linux-gnu (64-bit)
-#> Running under: Pop!_OS 19.10
+#> R version 4.0.2 (2020-06-22)
+#> Platform: x86_64-apple-darwin17.0 (64-bit)
+#> Running under: macOS Catalina 10.15.6
 #> 
 #> Matrix products: default
-#> BLAS:   /usr/lib/x86_64-linux-gnu/openblas/libblas.so.3
-#> LAPACK: /usr/lib/x86_64-linux-gnu/libopenblasp-r0.3.7.so
+#> BLAS:   /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRblas.dylib
+#> LAPACK: /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRlapack.dylib
 #> 
 #> locale:
-#>  [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
-#>  [3] LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
-#>  [5] LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
-#>  [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                 
-#>  [9] LC_ADDRESS=C               LC_TELEPHONE=C            
-#> [11] LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       
+#> [1] en_CA.UTF-8/en_CA.UTF-8/en_CA.UTF-8/C/en_CA.UTF-8/en_CA.UTF-8
 #> 
@@
 #>  [5] scales_1.1.1              forcats_0.5.0            
-#>  [7] stringr_1.4.0             dplyr_1.0.0              
+#>  [7] stringr_1.4.0             dplyr_1.0.2              
 #>  [9] purrr_0.3.4               readr_1.3.1              
-#> [11] tidyr_1.1.0               ggplot2_3.3.2.9000       
-#> [13] tidyverse_1.3.0           lubridate_1.7.4          
-#> [15] visreg_2.7.0              tibble_3.0.1             
+#> [11] tidyr_1.1.1               ggplot2_3.3.2            
+#> [13] tidyverse_1.3.0           lubridate_1.7.9          
+#> [15] visreg_2.7.0              tibble_3.0.3             
 #> [17] riskRegression_2020.02.05 prodlim_2019.11.13       
-#> [19] flexsurv_1.1.1            survival_3.1-8           
+#> [19] flexsurv_1.1.1            survival_3.2-3           
 #> [21] pracma_2.2.9              magrittr_1.5             
 #> [23] kableExtra_1.1.0          glmnet_4.0-2             
-#> [25] Matrix_1.2-18             data.table_1.12.8        
+#> [25] Matrix_1.2-18             data.table_1.13.0        
 #> [27] cowplot_1.0.0             colorspace_1.4-1         
-#> [29] casebase_0.9.0           
+#> [29] casebase_0.9.0.9000      
 #> 
@@
 #>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
-#>  [4] ggstance_0.3.3      htmlTable_1.13.3    base64enc_0.1-3    
-#>  [7] fs_1.3.2            rstudioapi_0.11     farver_2.0.3       
-#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.0-12     
-#> [13] xml2_1.3.0          codetools_0.2-16    knitr_1.29         
+#>  [4] ggstance_0.3.4      htmlTable_1.13.3    base64enc_0.1-3    
+#>  [7] fs_1.4.1            rstudioapi_0.11     farver_2.0.3       
+#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-1      
+#> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
 #> [16] Formula_1.2-3       jsonlite_1.7.0      cluster_2.1.0      
-#> [19] dbplyr_1.4.2        png_0.1-7           compiler_3.6.2     
-#> [22] httr_1.4.1          backports_1.1.8     assertthat_0.2.1   
+#> [19] dbplyr_1.4.4        png_0.1-7           compiler_4.0.2     
+#> [22] httr_1.4.2          backports_1.1.8     assertthat_0.2.1   
 #> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
-#> [28] quantreg_5.54       tools_3.6.2         gtable_0.3.0       
-#> [31] Rcpp_1.0.4.6        cellranger_1.1.0    vctrs_0.3.1        
-#> [34] nlme_3.1-143        iterators_1.0.12    xfun_0.15          
-#> [37] rvest_0.3.5         lifecycle_0.2.0     polspline_1.1.17   
-#> [40] muhaz_1.2.6.1       MASS_7.3-51.5       zoo_1.8-7          
+#> [28] quantreg_5.55       tools_4.0.2         gtable_0.3.0       
+#> [31] Rcpp_1.0.5          cellranger_1.1.0    vctrs_0.3.2        
+#> [34] nlme_3.1-148        iterators_1.0.12    xfun_0.15          
+#> [37] rvest_0.3.6         lifecycle_0.2.0     polspline_1.1.17   
+#> [40] muhaz_1.2.6.1       MASS_7.3-51.6       zoo_1.8-8          
 #> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
-#> [46] RColorBrewer_1.1-2  rticles_0.13        yaml_2.2.1         
+#> [46] RColorBrewer_1.1-2  rticles_0.14        yaml_2.2.1         
 #> [49] gridExtra_2.3       rms_5.1-4           rpart_4.1-15       
 #> [52] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
-#> [55] checkmate_2.0.0     lava_1.6.6          shape_1.4.4        
-#> [58] rlang_0.4.6         pkgconfig_2.0.3     evaluate_0.14      
-#> [61] lattice_0.20-38     labeling_0.3        htmlwidgets_1.5.1  
-#> [64] cmprsk_2.2-9        tidyselect_1.1.0    deSolve_1.27.1     
-#> [67] plyr_1.8.6          R6_2.4.1            generics_0.0.2     
-#> [70] Hmisc_4.3-1         multcomp_1.4-12     DBI_1.1.0          
-#> [73] withr_2.2.0         pillar_1.4.4        haven_2.3.1        
-#> [76] foreign_0.8-74      mgcv_1.8-31         nnet_7.3-12        
-#> [79] mstate_0.2.12       modelr_0.1.5        crayon_1.3.4       
-#> [82] rmarkdown_2.3       jpeg_0.1-8.1        grid_3.6.2         
-#> [85] readxl_1.3.1        reprex_0.3.0        digest_0.6.25      
-#> [88] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_3.6.2       
+#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
+#> [58] rlang_0.4.7         pkgconfig_2.0.3     evaluate_0.14      
+#> [61] lattice_0.20-41     labeling_0.3        htmlwidgets_1.5.1  
+#> [64] cmprsk_2.2-10       tidyselect_1.1.0    deSolve_1.28       
+#> [67] R6_2.4.1            generics_0.0.2      Hmisc_4.4-0        
+#> [70] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
+#> [73] pillar_1.4.6        haven_2.3.1         foreign_0.8-80     
+#> [76] mgcv_1.8-31         nnet_7.3-14         mstate_0.2.12      
+#> [79] modelr_0.1.8        crayon_1.3.4        rmarkdown_2.3      
+#> [82] jpeg_0.1-8.1        grid_4.0.2          readxl_1.3.1       
+#> [85] blob_1.2.1          reprex_0.3.0        digest_0.6.25      
+#> [88] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_4.0.2       
 #> [91] munsell_0.5.0       viridisLite_0.3.0   quadprog_1.5-8
@@
 \begin{CodeOutput}
-#> Local:    fix-cs1 /home/sahir/git_repositories/cbpaper
-#> Remote:   fix-cs1 @ origin (https://github.com/sahirbhatnagar/cbpaper.git)
-#> Head:     [349d153] 2020-08-21: fix merge conflicts
+#> Local:    editMax /Users/maxturgeon/Documents/git_repos/cbpaper
+#> Head:     [08c024e] 2020-08-21: Merge pull request #26 from sahirbhatnagar/fix-cs1
 \end{CodeOutput}

### 2020-08-25 analysis/paper/paper.Rmd
added reason for smoothish cox and recolored

--- 2020-08-20 analysis/paper/paper.Rmd
@@
 This difference does not appear to be significant: the curve from case-base sampling is contained within a 95% confidence band around the Fine-Gray absolute risk curve (figure not shown).
-
 # Case study 3---SUPPORT Data
@@
 ```{r supportData, eval = eval_cs3}
-scaleFUNy <- function(x) sprintf("%.2f", x)
-scaleFUNx <- function(x) sprintf("%.0f", x)
 data(support)
@@
 
-<!--Before generating an absolute risk curve using regularized estimation, we will explore three options as a baseline comparison.
+<!--Before generating an absolute risk curve using regularized logistic regression, we will explore three options as a baseline comparison.
 The original study compared these scores over a span of 180 days, which may not be comparable over the full 5.56 years.
@@
 Figure \ref{fig:brier1} examines the brier scores as they change over time.
-As the full model has the lowest brier score on overall, it will be used in the following absolute risk comparison to regularized cox, regularized estimation and the unadjusted Kaplan-Meier curve.-->
+As the full model has the lowest brier score on overall, it will be used in the following absolute risk comparison to regularized cox, regularized logistic regression and the unadjusted Kaplan-Meier curve.-->
 
@@
   scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
-  paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1))) +scale_y_continuous(labels=scaleFUNy,n.breaks=8)
+  paper_gg_theme
 ```
@@
   scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
-  paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1)))+scale_y_continuous(labels=scaleFUNy,n.breaks=8)
+  paper_gg_theme
 
@@
 
-For our penalized model, we opt for the natural log of time; recall that this corresponds to a Weibull distribution.
+For our penalized logistic regression model, we opt for the natural log of time; recall that this corresponds to a Weibull distribution.
 For fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface to the function \code{fitSmoothHazard}.
@@
 
-```{r supportCB_fit, eval=eval_cs3, echo=TRUE}
+```{r supportCB_fit, eval=eval_cs3, echo=TRUE, cache=FALSE}
 x <- model.matrix(death ~ . - d.time - aps - sps, 
@@
 We first create a matrix \code{newx} containing the relevant variables.
-Next, using \code{times}, we specify the time points at which the cumulative incidence will be computed.
+Using \code{times}, we specify the time points at which the cumulative incidence will be computed.
 Finally, using the parameter \code{s = "lambda.1se"}, we specify at which value of the tuning parameter we want to retrieve the coefficient estimates.
 
-```{r supportCB_abs, eval=eval_cs3, echo=TRUE}
+```{r supportCB_abs, eval=eval_cs3, echo=TRUE, cache=TRUE}
 # Estimating the absolute risk curve using the newdata parameter
@@
                    absRisk = rowMeans(abCbFull[, -c(1)]))
-abCB$Model <- "Penalized estimation"
+abCB$Model <- "Penalized logistic"
 
@@
 This is expected, as there are no restrictions on the size of coefficients.
-The smallest coefficient estimates are found in penalized case-base model, aside from cancer state (cano).
+The smallest coefficient estimates are found in penalized logistic model, aside from cancer state (cano).
 
@@
 # estimatescb$std.error<-0
-estimatescb$Model <- "Pen. CB"
+estimatescb$Model <- "Pen. logistic"
 colnames(estimatescb) <- c("term", "estimate", "model")
@@
 
-These values do not inform us of model performance.
-To do so, we take our hazard functions and use them to predict the absolute risk of everyone in our test set, over time.
-The probabilities over time for each individual are averaged, resulting in the absolute risk curves in Figure \ref{fig:cs3FinalBrier} (A).
-There is separation between the curves, but it is minimal.
-To better demonstrate the differences in performance between the models, we compare Brier scores over time.
-In Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our penalized case-base model compared to both Cox regression models.
-The differences, however, are minimal.
-All three outperform the Kaplan-Meier estimator.
-
 ```{r cs3Final, echo=FALSE, eval=eval_cs3}
@@
   geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
-  labs(y = "Probability of death", x = "Follow-up time (years)", color = "Models") +
+  labs(y = "Probability of Death", x = "Follow-up time (Years)", color = "Models") +
   # ggtitle("Absolute risk curves") +
-  scale_color_manual(values = q4, breaks = c("K-M","Cox","Penalized Cox","Penalized CB")) + 
-  paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1))) +scale_y_continuous(labels=scaleFUNy,n.breaks=9)
+  scale_color_manual(values = q4, breaks = c("K-M","Cox","Penalized Cox","Penalized logistic")) + 
+  paper_gg_theme
 ```
@@
 
-```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} Comparison of Cox regression, penalized Cox regression, penalized case-base estimation, and Kaplan-Meier.
-(A)  Probabilities of death as it changes over follow-up time.
-(B) Brier scores as they change over time, where a lower score corresponds to better performace.
-Penalized case-base estimation produced the lowest results over follow-up time.", warning = FALSE}
+```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} (A) Four curves comparing the probabilities of death as it changes over follow-up time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
+All follow-up times seen in the new data were used to generate the curves.
+The Cox-based curves appear to be smooth, but this is due to the 1703 follow-up times seen while fitting the models and the 319 time points used to estimate the curves.
+(B) Four curves demonstrating the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
+Lower is better.
+Penalized logistic regression produced the lowest Brier scores over follow-up time.", warning = FALSE}
 # Compute Brier score for penalized models
-brierFinalResults <- Score(list("Penalized Cox" = coxNet, "Penalized CB" = cbFull),
+brierFinalResults <- Score(list("Penalized Cox" = coxNet, "Penalized Logistic" = cbFull),
                            data = cbind(subset(test, select = c(d.time, death)),
@@
   filter(brierFull$Brier$score, model == "PredictionModel")
-  ) %>% 
-  mutate(model = as.character(model), # First return to string, so types match
-         model = case_when(
-    model == "Null model" ~ "K-M",
-    model == "PredictionModel" ~ "Cox",
-    TRUE ~ model
-  ),
-  model = factor(model, levels = c("Cox", "Penalized Cox", "Penalized Logistic", "K-M"),
-                 labels = c("Cox", "Pen. Cox", "Pen. Logistic", "K-M")))
+  )
+data_brier$model<-factor(data_brier$model,c("Null model", "PredictionModel","Penalized Cox", "Penalized Logistic"))
+levels(data_brier$model)<-c("Kaplan-Meier","Cox","Penalized Cox","Penalized logistic")
 
@@
   geom_line() +
-  xlab("Follow-up time (years)") +
+  xlab("Follow-up time (Years)") +
   ylab("Brier score") +
@@
   paper_gg_theme +  
-  scale_color_manual(values = q4) + scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1))) +scale_y_continuous(labels=scaleFUNy,n.breaks=8)
+  scale_color_manual(values = q4)
 
@@
 ```
+
+These values do not inform us of model performance.
+To do so, we take our hazard functions and use them to predict the absolute risk of everyone in our test set, over time.
+The probabilities over time for each individual are averaged, resulting in the absolute risk curves in Figure \ref{fig:cs3FinalBrier} (A).
+There is separation between the curves, but it is minimal.
+To better demonstrate the differences in performance between the models, we compare Brier scores over time.
+In Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our penalized logistic regression model compared to both Cox regression models.
+The differences, however, are minimal.
+All three outperform the Kaplan-Meier model.
 

### 2020-08-25 analysis/paper/paper.tex
added reason for smoothish cox and recolored

--- 2020-08-20 analysis/paper/paper.tex
@@
 \midrule
-Sex & 0.72 & (0.42, 1.26) & 0.68 & (0.39, 1.19)\\
-Disease & 0.54 & (0.3, 0.97) & 0.52 & (0.29, 0.93)\\
-Phase (CR2 vs. CR1) & 1.21 & (0.48, 3.02) & 1.21 & (0.48, 3.02)\\
-Phase (CR3 vs. CR1) & 1.63 & (0.41, 6.44) & 1.67 & (0.43, 6.5)\\
-Phase (Relapse vs. CR1) & 4.43 & (2.06, 9.55) & 4.55 & (2.09, 9.9)\\
+Sex & 0.70 & (0.4, 1.23) & 0.68 & (0.39, 1.19)\\
+Disease & 0.53 & (0.29, 0.95) & 0.52 & (0.29, 0.93)\\
+Phase (CR2 vs. CR1) & 1.12 & (0.45, 2.78) & 1.21 & (0.48, 3.02)\\
+Phase (CR3 vs. CR1) & 1.45 & (0.37, 5.66) & 1.67 & (0.43, 6.5)\\
+Phase (Relapse vs. CR1) & 4.04 & (1.88, 8.68) & 4.55 & (2.09, 9.9)\\
 \addlinespace
-Source & 1.51 & (0.49, 4.69) & 1.46 & (0.47, 4.54)\\
+Source & 1.48 & (0.48, 4.54) & 1.46 & (0.47, 4.54)\\
 Age & 0.99 & (0.97, 1.01) & 0.99 & (0.97, 1.02)\\
@@
 contained within a 95\% confidence band around the Fine-Gray absolute
-risk curve (figure not shown).
-
-\hypertarget{case-study-3support-data}{%
-\section{Case study 3---SUPPORT Data}\label{case-study-3support-data}}
+risk curve (figure not shown). \# Case study 3---SUPPORT Data
 
@@
 
-For our penalized model, we opt for the natural log of time; recall that
-this corresponds to a Weibull distribution. For fitting the penalized
-hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface
-to the function \code{fitSmoothHazard}. In order to use this interface,
-we first create a matrix \code{y} containing the time and event
-variables, and a matrix \code{x} containing all remaining covariates we
-would like to include in the model space. We used lasso penalization by
-setting \code{alpha = 1}. Moreover, we use the argument
+For our penalized logistic regression model, we opt for the natural log
+of time; recall that this corresponds to a Weibull distribution. For
+fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which
+is a matrix interface to the function \code{fitSmoothHazard}. In order
+to use this interface, we first create a matrix \code{y} containing the
+time and event variables, and a matrix \code{x} containing all remaining
+covariates we would like to include in the model space. We used lasso
+penalization by setting \code{alpha = 1}. Moreover, we use the argument
 \code{penalty.factor} to force the time variable into the final model.
@@
 curve on the test data. We first create a matrix \code{newx} containing
-the relevant variables. Next, using \code{times}, we specify the time
-points at which the cumulative incidence will be computed. Finally,
-using the parameter \code{s = "lambda.1se"}, we specify at which value
-of the tuning parameter we want to retrieve the coefficient estimates.
+the relevant variables. Using \code{times}, we specify the time points
+at which the cumulative incidence will be computed. Finally, using the
+parameter \code{s = "lambda.1se"}, we specify at which value of the
+tuning parameter we want to retrieve the coefficient estimates.
 
@@
 restrictions on the size of coefficients. The smallest coefficient
-estimates are found in penalized case-base model, aside from cancer
-state (cano).
+estimates are found in penalized logistic model, aside from cancer state
+(cano).
 
@@
 Shrinkage occurs due to penalization.}\label{fig:coefplots}
+\end{figure}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/riskregressionBrier-1}
+
+}
+
+\caption{\label{fig:cs3FinalBrier} (A) Four curves comparing the probabilities of death as it changes over follow-up time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
+All follow-up times seen in the new data were used to generate the curves.
+The Cox-based curves appear to be smooth, but this is due to the 1703 follow-up times seen while fitting the models and the 319 time points used to estimate the curves.
+(B) Four curves demonstrating the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
+Lower is better.
+Penalized logistic regression produced the lowest Brier scores over follow-up time.}\label{fig:riskregressionBrier}
 \end{figure}
@@
 Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our
-penalized case-base model compared to both Cox regression models. The
-differences, however, are minimal. All three outperform the Kaplan-Meier
-estimator.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/riskregressionBrier-1}
-
-}
-
-\caption{\label{fig:cs3FinalBrier} Comparison of Cox regression, penalized Cox regression, penalized case-base estimation, and Kaplan-Meier.
-(A)  Probabilities of death as it changes over follow-up time.
-(B) Brier scores as they change over time, where a lower score corresponds to better performace.
-Penalized case-base estimation produced the lowest results over follow-up time.}\label{fig:riskregressionBrier}
-\end{figure}
-\end{CodeChunk}
+penalized logistic regression model compared to both Cox regression
+models. The differences, however, are minimal. All three outperform the
+Kaplan-Meier model.
 
@@
 #> Model1 Model2 Model3 
-#>    827    793    792
+#>    827    790    788
 \end{CodeOutput}
@@
 
-This report was generated on 2020-08-20 19:21:21 using the following
+This report was generated on 2020-08-25 12:59:22 using the following
 computational environment and dependencies:
@@
 \begin{CodeOutput}
-#> R version 3.6.2 (2019-12-12)
+#> R version 4.0.2 (2020-06-22)
 #> Platform: x86_64-pc-linux-gnu (64-bit)
-#> Running under: Pop!_OS 19.10
+#> Running under: Pop!_OS 20.04 LTS
 #> 
 #> Matrix products: default
-#> BLAS:   /usr/lib/x86_64-linux-gnu/openblas/libblas.so.3
-#> LAPACK: /usr/lib/x86_64-linux-gnu/libopenblasp-r0.3.7.so
+#> BLAS:   /usr/lib/x86_64-linux-gnu/blas/libblas.so.3.9.0
+#> LAPACK: /usr/lib/x86_64-linux-gnu/lapack/liblapack.so.3.9.0
 #> 
@@
 #> other attached packages:
-#>  [1] scales_1.1.1              forcats_0.5.0            
-#>  [3] stringr_1.4.0             dplyr_1.0.0              
-#>  [5] purrr_0.3.4               readr_1.3.1              
-#>  [7] tidyr_1.1.0               ggplot2_3.3.2.9000       
-#>  [9] tidyverse_1.3.0           lubridate_1.7.4          
-#> [11] visreg_2.7.0              tibble_3.0.1             
-#> [13] riskRegression_2020.02.05 prodlim_2019.11.13       
-#> [15] flexsurv_1.1.1            survival_3.1-8           
-#> [17] pracma_2.2.9              magrittr_1.5             
-#> [19] kableExtra_1.1.0          glmnet_4.0-2             
-#> [21] Matrix_1.2-18             data.table_1.12.8        
-#> [23] cowplot_1.0.0             colorspace_1.4-1         
-#> [25] casebase_0.9.0           
+#>  [1] broom_0.7.0               dotwhisker_0.5.0         
+#>  [3] timereg_1.9.6             glue_1.4.1               
+#>  [5] scales_1.1.1              forcats_0.5.0            
+#>  [7] stringr_1.4.0             dplyr_1.0.1              
+#>  [9] purrr_0.3.4               readr_1.3.1              
+#> [11] tidyr_1.1.1               ggplot2_3.3.2            
+#> [13] tidyverse_1.3.0           lubridate_1.7.9          
+#> [15] visreg_2.7.0              tibble_3.0.3             
+#> [17] riskRegression_2020.06.25 prodlim_2019.11.13       
+#> [19] flexsurv_1.1.1            survival_3.2-3           
+#> [21] pracma_2.2.9              magrittr_1.5             
+#> [23] kableExtra_1.1.0          glmnet_4.0-2             
+#> [25] Matrix_1.2-18             data.table_1.13.0        
+#> [27] cowplot_1.0.0             colorspace_1.4-1         
+#> [29] casebase_0.9.0           
 #> 
@@
 #>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
-#>  [4] htmlTable_1.13.3    base64enc_0.1-3     fs_1.3.2           
-#>  [7] rstudioapi_0.11     MatrixModels_0.4-1  fansi_0.4.1        
-#> [10] mvtnorm_1.0-12      xml2_1.3.0          codetools_0.2-16   
-#> [13] knitr_1.29          Formula_1.2-3       jsonlite_1.7.0     
-#> [16] broom_0.7.0         cluster_2.1.0       dbplyr_1.4.2       
-#> [19] png_0.1-7           compiler_3.6.2      httr_1.4.1         
-#> [22] backports_1.1.8     assertthat_0.2.1    cli_2.0.2          
-#> [25] acepack_1.4.1       htmltools_0.5.0     quantreg_5.54      
-#> [28] tools_3.6.2         gtable_0.3.0        glue_1.4.1         
-#> [31] Rcpp_1.0.4.6        cellranger_1.1.0    vctrs_0.3.1        
-#> [34] nlme_3.1-143        iterators_1.0.12    xfun_0.15          
-#> [37] rvest_0.3.5         lifecycle_0.2.0     polspline_1.1.17   
-#> [40] muhaz_1.2.6.1       MASS_7.3-51.5       zoo_1.8-7          
-#> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
-#> [46] RColorBrewer_1.1-2  rticles_0.13        yaml_2.2.1         
-#> [49] gridExtra_2.3       rms_5.1-4           rpart_4.1-15       
-#> [52] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
-#> [55] checkmate_2.0.0     lava_1.6.6          shape_1.4.4        
-#> [58] rlang_0.4.6         pkgconfig_2.0.3     evaluate_0.14      
-#> [61] lattice_0.20-38     htmlwidgets_1.5.1   cmprsk_2.2-9       
-#> [64] tidyselect_1.1.0    deSolve_1.27.1      R6_2.4.1           
-#> [67] generics_0.0.2      Hmisc_4.3-1         multcomp_1.4-12    
-#> [70] DBI_1.1.0           withr_2.2.0         pillar_1.4.4       
-#> [73] haven_2.3.1         foreign_0.8-74      mgcv_1.8-31        
-#> [76] nnet_7.3-12         mstate_0.2.12       modelr_0.1.5       
-#> [79] crayon_1.3.4        rmarkdown_2.3       timereg_1.9.4      
-#> [82] jpeg_0.1-8.1        grid_3.6.2          readxl_1.3.1       
-#> [85] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
-#> [88] numDeriv_2016.8-1.1 stats4_3.6.2        munsell_0.5.0      
-#> [91] viridisLite_0.3.0   quadprog_1.5-8
+#>  [4] ggstance_0.3.4      htmlTable_2.0.1     fs_1.4.2           
+#>  [7] base64enc_0.1-3     rstudioapi_0.11     farver_2.0.3       
+#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-1      
+#> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
+#> [16] Formula_1.2-3       jsonlite_1.7.0      cluster_2.1.0      
+#> [19] dbplyr_1.4.4        png_0.1-7           compiler_4.0.2     
+#> [22] httr_1.4.2          backports_1.1.8     assertthat_0.2.1   
+#> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
+#> [28] quantreg_5.61       tools_4.0.2         gtable_0.3.0       
+#> [31] Rcpp_1.0.5          cellranger_1.1.0    vctrs_0.3.2        
+#> [34] nlme_3.1-148        conquer_1.0.1       iterators_1.0.12   
+#> [37] xfun_0.16           rvest_0.3.6         lifecycle_0.2.0    
+#> [40] polspline_1.1.19    muhaz_1.2.6.1       MASS_7.3-51.6      
+#> [43] zoo_1.8-8           hms_0.5.3           sandwich_2.5-1     
+#> [46] SparseM_1.78        RColorBrewer_1.1-2  rticles_0.14       
+#> [49] yaml_2.2.1          gridExtra_2.3       rms_6.0-1          
+#> [52] rpart_4.1-15        latticeExtra_0.6-29 stringi_1.4.6      
+#> [55] foreach_1.5.0       checkmate_2.0.0     lava_1.6.7         
+#> [58] shape_1.4.4         mets_1.2.7.1        rlang_0.4.7        
+#> [61] pkgconfig_2.0.3     matrixStats_0.56.0  evaluate_0.14      
+#> [64] lattice_0.20-41     labeling_0.3        htmlwidgets_1.5.1  
+#> [67] cmprsk_2.2-10       tidyselect_1.1.0    deSolve_1.28       
+#> [70] R6_2.4.1            generics_0.0.2      Hmisc_4.4-0        
+#> [73] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
+#> [76] pillar_1.4.6        haven_2.3.1         foreign_0.8-80     
+#> [79] mgcv_1.8-31         nnet_7.3-14         mstate_0.2.12      
+#> [82] modelr_0.1.8        crayon_1.3.4        rmarkdown_2.3      
+#> [85] jpeg_0.1-8.1        grid_4.0.2          readxl_1.3.1       
+#> [88] blob_1.2.1          reprex_0.3.0        digest_0.6.25      
+#> [91] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_4.0.2       
+#> [94] munsell_0.5.0       viridisLite_0.3.0   quadprog_1.5-8
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> Local:    fix-cs1 /home/sahir/git_repositories/cbpaper
-#> Remote:   fix-cs1 @ origin (https://github.com/sahirbhatnagar/cbpaper.git)
-#> Head:     [f618e21] 2020-08-18: time dependent hazard ratio for ERSPC
+#> Local:    master /home/jesseislam/Documents/gits/cbpaper
+#> Remote:   master @ origin (git@github.com:Jesse-Islam/cbpaper.git)
+#> Head:     [ffe8b03] 2020-08-25: Merge branch 'master' into sbmaster
 \end{CodeOutput}

### 2020-08-25 analysis/paper/paper.Rmd
added smoothCoxComment,recolor,missingFigurelabel

@@
 
-Our first comparison plot examines the covariates that were selected between all three hazard functions.
-Note that Cox regression without penalization selected more covariates than the other two, but only the intersection of coefficients with values greater than 0 are displayed in \ref{fig:cs3lolliPlot}.
-A visible tendency is for Cox regression to have larger estimated coefficients.
-This is expected, as there are no restrictions on the size of coefficients.
-The smallest coefficient estimates are found in penalized logistic model, aside from cancer state (cano).
-
 ```{r coefplots, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models.
@@
 ```
+
+Our first comparison plot examines the covariates that were selected between all three hazard functions.
+Note that Cox regression without penalization selected more covariates than the other two, but only the intersection of coefficients with values greater than 0 are displayed in Figure \ref{fig:cs3lolliPlot}.
+A visible tendency is for Cox regression to have larger estimated coefficients.
+This is expected, as there are no restrictions on the size of coefficients.
+The smallest coefficient estimates are found in penalized logistic model, aside from cancer state (cano).
 

### 2020-08-25 analysis/paper/paper.tex
added smoothCoxComment,recolor,missingFigurelabel

@@
 
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/coefplots-1}
+
+}
+
+\caption{\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models.
+If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity.
+Shrinkage occurs due to penalization.}\label{fig:coefplots}
+\end{figure}
+\end{CodeChunk}
+
 Our first comparison plot examines the covariates that were selected
@@
 intersection of coefficients with values greater than 0 are displayed in
-\ref{fig:cs3lolliPlot}. A visible tendency is for Cox regression to have
-larger estimated coefficients. This is expected, as there are no
+Figure \ref{fig:cs3lolliPlot}. A visible tendency is for Cox regression
+to have larger estimated coefficients. This is expected, as there are no
 restrictions on the size of coefficients. The smallest coefficient
@@
 (cano).
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/coefplots-1}
-
-}
-
-\caption{\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models.
-If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity.
-Shrinkage occurs due to penalization.}\label{fig:coefplots}
-\end{figure}
-\end{CodeChunk}
 
@@
 
-This report was generated on 2020-08-25 12:59:22 using the following
+This report was generated on 2020-08-25 13:52:10 using the following
 computational environment and dependencies:
@@
 #> other attached packages:
-#>  [1] broom_0.7.0               dotwhisker_0.5.0         
-#>  [3] timereg_1.9.6             glue_1.4.1               
-#>  [5] scales_1.1.1              forcats_0.5.0            
-#>  [7] stringr_1.4.0             dplyr_1.0.1              
-#>  [9] purrr_0.3.4               readr_1.3.1              
-#> [11] tidyr_1.1.1               ggplot2_3.3.2            
-#> [13] tidyverse_1.3.0           lubridate_1.7.9          
-#> [15] visreg_2.7.0              tibble_3.0.3             
-#> [17] riskRegression_2020.06.25 prodlim_2019.11.13       
-#> [19] flexsurv_1.1.1            survival_3.2-3           
-#> [21] pracma_2.2.9              magrittr_1.5             
-#> [23] kableExtra_1.1.0          glmnet_4.0-2             
-#> [25] Matrix_1.2-18             data.table_1.13.0        
-#> [27] cowplot_1.0.0             colorspace_1.4-1         
-#> [29] casebase_0.9.0           
+#>  [1] scales_1.1.1              forcats_0.5.0            
+#>  [3] stringr_1.4.0             dplyr_1.0.1              
+#>  [5] purrr_0.3.4               readr_1.3.1              
+#>  [7] tidyr_1.1.1               ggplot2_3.3.2            
+#>  [9] tidyverse_1.3.0           lubridate_1.7.9          
+#> [11] visreg_2.7.0              tibble_3.0.3             
+#> [13] riskRegression_2020.06.25 prodlim_2019.11.13       
+#> [15] flexsurv_1.1.1            survival_3.2-3           
+#> [17] pracma_2.2.9              magrittr_1.5             
+#> [19] kableExtra_1.1.0          glmnet_4.0-2             
+#> [21] Matrix_1.2-18             data.table_1.13.0        
+#> [23] cowplot_1.0.0             colorspace_1.4-1         
+#> [25] casebase_0.9.0           
 #> 
@@
 #>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
-#>  [4] ggstance_0.3.4      htmlTable_2.0.1     fs_1.4.2           
-#>  [7] base64enc_0.1-3     rstudioapi_0.11     farver_2.0.3       
-#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-1      
-#> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
-#> [16] Formula_1.2-3       jsonlite_1.7.0      cluster_2.1.0      
-#> [19] dbplyr_1.4.4        png_0.1-7           compiler_4.0.2     
-#> [22] httr_1.4.2          backports_1.1.8     assertthat_0.2.1   
-#> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
-#> [28] quantreg_5.61       tools_4.0.2         gtable_0.3.0       
+#>  [4] htmlTable_2.0.1     fs_1.4.2            base64enc_0.1-3    
+#>  [7] rstudioapi_0.11     MatrixModels_0.4-1  fansi_0.4.1        
+#> [10] mvtnorm_1.1-1       xml2_1.3.2          codetools_0.2-16   
+#> [13] knitr_1.29          Formula_1.2-3       jsonlite_1.7.0     
+#> [16] broom_0.7.0         cluster_2.1.0       dbplyr_1.4.4       
+#> [19] png_0.1-7           compiler_4.0.2      httr_1.4.2         
+#> [22] backports_1.1.8     assertthat_0.2.1    cli_2.0.2          
+#> [25] acepack_1.4.1       htmltools_0.5.0     quantreg_5.61      
+#> [28] tools_4.0.2         gtable_0.3.0        glue_1.4.1         
 #> [31] Rcpp_1.0.5          cellranger_1.1.0    vctrs_0.3.2        
@@
 #> [61] pkgconfig_2.0.3     matrixStats_0.56.0  evaluate_0.14      
-#> [64] lattice_0.20-41     labeling_0.3        htmlwidgets_1.5.1  
-#> [67] cmprsk_2.2-10       tidyselect_1.1.0    deSolve_1.28       
-#> [70] R6_2.4.1            generics_0.0.2      Hmisc_4.4-0        
-#> [73] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
-#> [76] pillar_1.4.6        haven_2.3.1         foreign_0.8-80     
-#> [79] mgcv_1.8-31         nnet_7.3-14         mstate_0.2.12      
-#> [82] modelr_0.1.8        crayon_1.3.4        rmarkdown_2.3      
+#> [64] lattice_0.20-41     htmlwidgets_1.5.1   cmprsk_2.2-10      
+#> [67] tidyselect_1.1.0    deSolve_1.28        R6_2.4.1           
+#> [70] generics_0.0.2      Hmisc_4.4-0         multcomp_1.4-13    
+#> [73] DBI_1.1.0           withr_2.2.0         pillar_1.4.6       
+#> [76] haven_2.3.1         foreign_0.8-80      mgcv_1.8-31        
+#> [79] nnet_7.3-14         mstate_0.2.12       modelr_0.1.8       
+#> [82] crayon_1.3.4        rmarkdown_2.3       timereg_1.9.6      
 #> [85] jpeg_0.1-8.1        grid_4.0.2          readxl_1.3.1       

### 2020-08-31 analysis/paper/paper.Rmd

@@
 times <- sort(unique(test$d.time))
+#times<-seq(from=min(times),to=max(times),length.out = 100)
 
@@
 ab <- rbind(abCB, abCN)
+
 ```
@@
 ```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} (A) Four curves comparing the probabilities of death as it changes over follow-up time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
-All follow-up times seen in the new data were used to generate the curves.
-The Cox-based curves appear to be smooth, but this is due to the 1703 follow-up times seen while fitting the models and the 319 time points used to estimate the curves.
+All follow-up times seen in the new data were used to generate the penalized casebase and K-M curves.
+The Cox-based curves, estimated using the survival package, appear to be smooth because all 1703 follow-up times seen while fitting are used.
+Using user specified times is not permitted in the survival package.
 (B) Four curves demonstrating the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.

### 2020-08-31 analysis/paper/paper.tex

--- 2020-08-25 analysis/paper/paper.tex
@@
 R> times <- sort(unique(test$d.time))
+R> #times<-seq(from=min(times),to=max(times),length.out = 100)
 R> 
@@
 
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/coefplots-1}
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/riskregressionBrier-1}
 
@@
 
-\caption{\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models.
-If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity.
-Shrinkage occurs due to penalization.}\label{fig:coefplots}
-\end{figure}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/riskregressionBrier-1}
-
-}
-
 \caption{\label{fig:cs3FinalBrier} (A) Four curves comparing the probabilities of death as it changes over follow-up time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
-All follow-up times seen in the new data were used to generate the curves.
-The Cox-based curves appear to be smooth, but this is due to the 1703 follow-up times seen while fitting the models and the 319 time points used to estimate the curves.
+All follow-up times seen in the new data were used to generate the penalized casebase and K-M curves.
+The Cox-based curves, estimated using the survival package, appear to be smooth because all 1703 follow-up times seen while fitting are used.
+Using user specified times is not permitted in the survival package.
 (B) Four curves demonstrating the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
@@
 
-This report was generated on 2020-08-25 12:59:22 using the following
+This report was generated on 2020-08-25 15:21:11 using the following
 computational environment and dependencies:

### 2020-08-31 analysis/paper/paper.Rmd
fixed penalized issue.

--- 2020-08-25 analysis/paper/paper.Rmd
@@
   \usepackage{amsmath}
+  \usepackage{amssymb}
   \usepackage{longtable}
@@
 \begin{tabular}{ccccccccc}
-\hline
+\toprule
 & \textbf{Competing} & \textbf{Allows} &  \textbf{Penalized}   &    &  & \textbf{Semi} & \textbf{Interval/Left} & \textbf{Absolute}  \\
 \textbf{Package}        & \textbf{Risks}  & \textbf{Non PH} & \textbf{Regression} & \textbf{Splines} & \textbf{Parametric} & \textbf{Parametric} & \textbf{Censoring} & \textbf{Risk}           \\
-\hline
-\textbf{casebase}        & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\ \bottomrule
-\textbf{CFC}             & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\ \bottomrule
-\textbf{cmprsk}             & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\ \bottomrule
-\textbf{CoxRidge}        &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
-\textbf{crrp}            & x                        &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
-\textbf{fastcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
-\textbf{flexrsurv}       &                          & x                         &                       & x                & x                   &                          &                                  & x               \\ \bottomrule
-\textbf{flexsurv}        & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\ \bottomrule
-\textbf{glmnet}          &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
-\textbf{glmpath}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
-\textbf{mets}            & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\ \bottomrule
-\textbf{penalized}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
-\textbf{riskRegression}  & x                         &                           & x                     &                  &                     & x                        &                                  & x                                \\ \bottomrule
-\textbf{rstpm2}          &                          & x                         &                      & x                & x                   & x                        & x                                & x                         \\ \bottomrule
-\textbf{SmoothHazard}    &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\ \bottomrule
-\textbf{survival}        & x                        & x                         &                       &                  & x                   & x                        & x                                & x                               \\ \bottomrule
-\hline
+\bottomrule
+\textbf{casebase}        & \checkmark                        & \checkmark                         & \checkmark                     & \checkmark                & \checkmark                   &                          &                                  & \checkmark                                \\ \hline
+\textbf{CFC}             & \checkmark                        & \checkmark                         &                       &                  & \checkmark                   &                          &                                  & \checkmark                                \\ \hline
+\textbf{cmprsk}             & \checkmark                        &                           &                       &                  &                     &  \checkmark                        &                                  & \checkmark                                \\ \hline
+\textbf{CoxRidge}        &                          & \checkmark                         & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
+\textbf{crrp}            & \checkmark                        &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
+\textbf{fastcox}         &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
+\textbf{flexrsurv}       &                          & \checkmark                         &                       & \checkmark                & \checkmark                   &                          &                                  & \checkmark               \\ \hline
+\textbf{flexsurv}        & \checkmark                        & \checkmark                         &                       & \checkmark                & \checkmark                   &                          &                                  & \checkmark               \\ \hline
+\textbf{glmnet}          &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
+\textbf{glmpath}         &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
+\textbf{mets}            & \checkmark                        &                           &                       & \checkmark                &                     & \checkmark                        &                                  & \checkmark                                \\ \hline
+\textbf{penalized}       &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
+\textbf{riskRegression}  & \checkmark                         &                           & \checkmark                     &                  &                     & \checkmark                        &                                  & \checkmark                                \\ \hline
+\textbf{rstpm2}          &                          & \checkmark                         &                      & \checkmark                & \checkmark                   & \checkmark                        & \checkmark                                & \checkmark                         \\ \hline
+\textbf{SmoothHazard}    &                          & \checkmark                         &                       & \checkmark                & \checkmark                   &                          & \checkmark                            &                                      \\ \hline
+\textbf{survival}        & \checkmark                        & \checkmark                         &                       &                  & \checkmark                   & \checkmark                        & \checkmark                                & \checkmark                               \\
+\bottomrule
 \end{tabular}%
@@
 \caption{Comparison of various \proglang{R} packages for survival analysis based on several defining features.
-Competing Risks: whether or not an implementation for competing risks is present.
-Allows Non-PH: permits models for non-proportional hazards.
-Penalized Regression: allows for a penalty term when estimating hazards (e.g.
+\textbf{Competing Risks}: whether or not an implementation for competing risks is present.
+\textbf{Allows Non PH}: permits models for non-proportional hazards.
+\textbf{Penalized Regression}: allows for a penalty term on the regression coefficients when estimating hazards (e.g.
 lasso or ridge).
-Splines: permits a flexible fit on time through the use of splines.
-Parametric: implementation for parametric models.
-Semi-parametric: implementation for semi-parametric models.
-Interval/left censoring: models for interval and left-censoring.
+\textbf{Splines}: permits a flexible fit on time through the use of splines.
+\textbf{Parametric}: implementation for parametric models.
+\textbf{Semi-parametric}: implementation for semi-parametric models.
+\textbf{Interval/left censoring}: models for interval and left-censoring.
 If this is not selected, the package only handles right-censoring.
-Absolute Risk: computation for survival curves, cumulative incidence or cumulative hazard is available.}
+\textbf{Absolute Risk}: computation for survival curves, cumulative incidence or cumulative hazard is readily available.}
 \label{tab:surv-pkgs}
@@
 As noted above, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models.
-Therefore, we can easily test the null hypothesis that the Weibull model is just as good as the larger (in terms of number of parameters) spline model:
+Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) spline model:
 
 ```{r erscp-compare, eval = eval_cs1, echo = FALSE}
-anova(fits$weibull, fits$splines, test = "LRT")
+anova(fits$exponential, fits$splines, test = "LRT")
 ```
@@
 
-In Figure \ref{fig:interaction-ERSPC} we plot the estimated death from prostate cancer hazard ratio and 95\\% confidence interval for screening vs.
+In Figure \ref{fig:interaction-ERSPC} we plot the estimated death from prostate cancer hazard ratio and 95% confidence interval for screening vs.
 control group as a function of time using the \code{plot} method for objects of class \code{singleEventCB}:
@@
 This difference does not appear to be significant: the curve from case-base sampling is contained within a 95% confidence band around the Fine-Gray absolute risk curve (figure not shown).
+
 # Case study 3---SUPPORT Data
@@
 Moreover, it provides a metric that can be used to compare parametric and semi-parametric survival models.
-To account for overfitting, we will use 95% of the observations to train our models, while the remaining 5% will be used to compute absolute risk curves and the Brier scores.
+We will use 95% of the observations to train our models, and the remaining 5% to compute absolute risk curves and Brier scores.
 
 ```{r supportData, eval = eval_cs3}
+scaleFUNy <- function(x) sprintf("%.2f", x)
+scaleFUNx <- function(x) sprintf("%.0f", x)
 data(support)
@@
 
-In Figure\ref{fig:brier1} (A), we can see that all Cox models have similar average survival curves.
+In Figure \ref{fig:brier1} (A), we can see that all Cox models have similar average survival curves.
 On the other hand, Figure \ref{fig:brier1} (B) shows that the full model has the lowest Brier score overall.
@@
   scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
-  paper_gg_theme
+  paper_gg_theme + 
+  scale_x_continuous(labels = scaleFUNx, breaks = round(seq(0, 5, by = 1))) + 
+  scale_y_continuous(labels = scaleFUNy, n.breaks = 8)
 ```
@@
 
-For our penalized logistic regression model, we opt for the natural log of time; recall that this corresponds to a Weibull distribution.
+\newpage
+
+For our penalized model, we opt for the natural log of time; recall that this corresponds to a Weibull distribution.
 For fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface to the function \code{fitSmoothHazard}.
@@
 We first create a matrix \code{newx} containing the relevant variables.
-Using \code{times}, we specify the time points at which the cumulative incidence will be computed.
+Next, using \code{times}, we specify the time points at which the cumulative incidence will be computed.
 Finally, using the parameter \code{s = "lambda.1se"}, we specify at which value of the tuning parameter we want to retrieve the coefficient estimates.
@@
                    absRisk = rowMeans(abCbFull[, -c(1)]))
-abCB$Model <- "Penalized logistic"
+abCB$Model <- "Penalized CB"
 
 ab <- rbind(abCB, abCN)
+
 ```
@@
 
+These values do not inform us of model performance.
+To do so, we take our hazard functions and use them to predict the absolute risk of everyone in our test set, over time.
+The probabilities over time for each individual are averaged, resulting in the absolute risk curves in Figure \ref{fig:cs3FinalBrier} (A).
+There is separation between the curves, but it is minimal.
+To better demonstrate the differences in performance between the models, we compare Brier scores over time.
+In Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our penalized case-base model compared to both Cox regression models.
+The differences, however, are minimal.
+All three outperform the Kaplan-Meier estimator.
+
 ```{r cs3Final, echo=FALSE, eval=eval_cs3}
@@
   geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
-  labs(y = "Probability of Death", x = "Follow-up time (Years)", color = "Models") +
+  labs(y = "Probability of death", x = "Follow-up time (years)", color = "Models") +
   # ggtitle("Absolute risk curves") +
-  scale_color_manual(values = q4, breaks = c("K-M","Cox","Penalized Cox","Penalized logistic")) + 
-  paper_gg_theme
+  scale_color_manual(values = q4, breaks = c("K-M", "Cox", "Penalized Cox", "Penalized CB")) + 
+  paper_gg_theme + 
+  scale_x_continuous(labels = scaleFUNx, breaks = round(seq(0, 5, by = 1))) + 
+  scale_y_continuous(labels = scaleFUNy, n.breaks = 9)
 ```
@@
 
-```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} (A) Four curves comparing the probabilities of death as it changes over follow-up time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
-All follow-up times seen in the new data were used to generate the curves.
-The Cox-based curves appear to be smooth, but this is due to the 1703 follow-up times seen while fitting the models and the 319 time points used to estimate the curves.
-(B) Four curves demonstrating the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
-Lower is better.
-Penalized logistic regression produced the lowest Brier scores over follow-up time.", warning = FALSE}
+```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} Comparison of Cox regression, penalized Cox regression, penalized case-base estimation, and Kaplan-Meier.
+(A)  Probabilities of death as it changes over follow-up time.
+(B) Brier scores as they change over time, where a lower score corresponds to better performace.
+Penalized case-base estimation produced the lowest results over follow-up time.", warning = FALSE}
 # Compute Brier score for penalized models
-brierFinalResults <- Score(list("Penalized Cox" = coxNet, "Penalized Logistic" = cbFull),
+brierFinalResults <- Score(list("Penalized Cox" = coxNet, "Penalized CB" = cbFull),
                            data = cbind(subset(test, select = c(d.time, death)),
@@
   filter(brierFull$Brier$score, model == "PredictionModel")
-  )
-data_brier$model<-factor(data_brier$model,c("Null model", "PredictionModel","Penalized Cox", "Penalized Logistic"))
-levels(data_brier$model)<-c("Kaplan-Meier","Cox","Penalized Cox","Penalized logistic")
+  ) %>% 
+  mutate(model = as.character(model), # First return to string, so types match
+         model = case_when(
+    model == "Null model" ~ "K-M",
+    model == "PredictionModel" ~ "Cox",
+    TRUE ~ model
+  ),
+  model = factor(model, levels = c("Cox", "Penalized Cox", "Penalized CB", "K-M"),
+                 labels = c("Cox", "Pen. Cox", "Pen. CB", "K-M")))
 
@@
   paper_gg_theme +  
-  scale_color_manual(values = q4)
+  scale_color_manual(values = q4) + 
+  scale_x_continuous(labels = scaleFUNx, 
+                     breaks = round(seq(0, 5, by = 1))) + 
+  scale_y_continuous(labels = scaleFUNy, n.breaks = 8)
 

### 2020-08-31 analysis/paper/paper.Rmd
final edits for CS4

@@
   \usepackage{amsmath}
-  \usepackage{amssymb}
   \usepackage{longtable}
@@
 To recover the whole survival curve---or the cumulative incidence function (CIF)---we then need to separately estimate the baseline hazard [@breslow1972discussion].
-This in turn leads to stepwise estimates of the survival function.
+This in turn leads to stepwise estimates of the survival function that can be hard to interpret.
 
@@
 With the goal of fitting smooth-in-time hazard functions, Hanley & Miettinen [-@hanley2009fitting] proposed a general framework for estimating fully parametric hazard models via logistic regression.
-Their approach provides users that are familiar with generalized linear models with a natural way of fitting parametric survival models.
+Their main idea is simple: comparing person-moments when the event of interest occurred with moments when patients were at risk.
+Their approach provides users familiar with generalized linear models with a natural way of fitting parametric survival models.
 Moreover, their framework is very flexible: general functions of time can be estimated (e.g.
 using splines or general additive models), and hence these models retain some of the flexibility of Cox's partial likelihood approach.
+And since the unit of analysis is a person moment, time-varying covariates can also easily be included in this framework.
 
@@
 \begin{tabular}{ccccccccc}
-\toprule
+\hline
 & \textbf{Competing} & \textbf{Allows} &  \textbf{Penalized}   &    &  & \textbf{Semi} & \textbf{Interval/Left} & \textbf{Absolute}  \\
 \textbf{Package}        & \textbf{Risks}  & \textbf{Non PH} & \textbf{Regression} & \textbf{Splines} & \textbf{Parametric} & \textbf{Parametric} & \textbf{Censoring} & \textbf{Risk}           \\
-\bottomrule
-\textbf{casebase}        & \checkmark                        & \checkmark                         & \checkmark                     & \checkmark                & \checkmark                   &                          &                                  & \checkmark                                \\ \hline
-\textbf{CFC}             & \checkmark                        & \checkmark                         &                       &                  & \checkmark                   &                          &                                  & \checkmark                                \\ \hline
-\textbf{cmprsk}             & \checkmark                        &                           &                       &                  &                     &  \checkmark                        &                                  & \checkmark                                \\ \hline
-\textbf{CoxRidge}        &                          & \checkmark                         & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
-\textbf{crrp}            & \checkmark                        &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
-\textbf{fastcox}         &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
-\textbf{flexrsurv}       &                          & \checkmark                         &                       & \checkmark                & \checkmark                   &                          &                                  & \checkmark               \\ \hline
-\textbf{flexsurv}        & \checkmark                        & \checkmark                         &                       & \checkmark                & \checkmark                   &                          &                                  & \checkmark               \\ \hline
-\textbf{glmnet}          &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
-\textbf{glmpath}         &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
-\textbf{mets}            & \checkmark                        &                           &                       & \checkmark                &                     & \checkmark                        &                                  & \checkmark                                \\ \hline
-\textbf{penalized}       &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
-\textbf{riskRegression}  & \checkmark                         &                           & \checkmark                     &                  &                     & \checkmark                        &                                  & \checkmark                                \\ \hline
-\textbf{rstpm2}          &                          & \checkmark                         &                      & \checkmark                & \checkmark                   & \checkmark                        & \checkmark                                & \checkmark                         \\ \hline
-\textbf{SmoothHazard}    &                          & \checkmark                         &                       & \checkmark                & \checkmark                   &                          & \checkmark                            &                                      \\ \hline
-\textbf{survival}        & \checkmark                        & \checkmark                         &                       &                  & \checkmark                   & \checkmark                        & \checkmark                                & \checkmark                               \\
-\bottomrule
+\hline
+\textbf{casebase}        & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\ \bottomrule
+\textbf{CFC}             & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\ \bottomrule
+\textbf{cmprsk}             & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\ \bottomrule
+\textbf{CoxRidge}        &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
+\textbf{crrp}            & x                        &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
+\textbf{fastcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
+\textbf{flexrsurv}       &                          & x                         &                       & x                & x                   &                          &                                  & x               \\ \bottomrule
+\textbf{flexsurv}        & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\ \bottomrule
+\textbf{glmnet}          &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
+\textbf{glmpath}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
+\textbf{mets}            & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\ \bottomrule
+\textbf{penalized}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
+\textbf{riskRegression}  & x                         &                           & x                     &                  &                     & x                        &                                  & x                                \\ \bottomrule
+\textbf{rstpm2}          &                          & x                         &                      & x                & x                   & x                        & x                                & x                         \\ \bottomrule
+\textbf{SmoothHazard}    &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\ \bottomrule
+\textbf{survival}        & x                        & x                         &                       &                  & x                   & x                        & x                                & x                               \\ \bottomrule
+\hline
 \end{tabular}%
@@
 \caption{Comparison of various \proglang{R} packages for survival analysis based on several defining features.
-\textbf{Competing Risks}: whether or not an implementation for competing risks is present.
-\textbf{Allows Non PH}: permits models for non-proportional hazards.
-\textbf{Penalized Regression}: allows for a penalty term on the regression coefficients when estimating hazards (e.g.
+Competing Risks: whether or not an implementation for competing risks is present.
+Allows Non-PH: permits models for non-proportional hazards.
+Penalized Regression: allows for a penalty term when estimating hazards (e.g.
 lasso or ridge).
-\textbf{Splines}: permits a flexible fit on time through the use of splines.
-\textbf{Parametric}: implementation for parametric models.
-\textbf{Semi-parametric}: implementation for semi-parametric models.
-\textbf{Interval/left censoring}: models for interval and left-censoring.
+Splines: permits a flexible fit on time through the use of splines.
+Parametric: implementation for parametric models.
+Semi-parametric: implementation for semi-parametric models.
+Interval/left censoring: models for interval and left-censoring.
 If this is not selected, the package only handles right-censoring.
-\textbf{Absolute Risk}: computation for survival curves, cumulative incidence or cumulative hazard is readily available.}
+Absolute Risk: computation for survival curves, cumulative incidence or cumulative hazard is available.}
 \label{tab:surv-pkgs}
@@
 As noted above, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models.
-Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) spline model:
+Therefore, we can easily test the null hypothesis that the Weibull model is just as good as the larger (in terms of number of parameters) spline model:
 
 ```{r erscp-compare, eval = eval_cs1, echo = FALSE}
-anova(fits$exponential, fits$splines, test = "LRT")
+anova(fits$weibull, fits$splines, test = "LRT")
 ```
@@
 
-In Figure \ref{fig:interaction-ERSPC} we plot the estimated death from prostate cancer hazard ratio and 95% confidence interval for screening vs.
+In Figure \ref{fig:interaction-ERSPC} we plot the estimated death from prostate cancer hazard ratio and 95\\% confidence interval for screening vs.
 control group as a function of time using the \code{plot} method for objects of class \code{singleEventCB}:
@@
 Moreover, it provides a metric that can be used to compare parametric and semi-parametric survival models.
-We will use 95% of the observations to train our models, and the remaining 5% to compute absolute risk curves and Brier scores.
+To account for overfitting, we will use 95% of the observations to train our models, while the remaining 5% will be used to compute absolute risk curves and the Brier scores.
 
@@
 
-In Figure \ref{fig:brier1} (A), we can see that all Cox models have similar average survival curves.
+In Figure\ref{fig:brier1} (A), we can see that all Cox models have similar average survival curves.
 On the other hand, Figure \ref{fig:brier1} (B) shows that the full model has the lowest Brier score overall.
@@
 
-\newpage
-
 For our penalized model, we opt for the natural log of time; recall that this corresponds to a Weibull distribution.
@@
 times <- sort(unique(test$d.time))
-#times<-seq(from=min(times),to=max(times),length.out = 100)
 
@@
 ab <- rbind(abCB, abCN)
-
 ```
@@
 
-These values do not inform us of model performance.
-To do so, we take our hazard functions and use them to predict the absolute risk of everyone in our test set, over time.
-The probabilities over time for each individual are averaged, resulting in the absolute risk curves in Figure \ref{fig:cs3FinalBrier} (A).
-There is separation between the curves, but it is minimal.
-To better demonstrate the differences in performance between the models, we compare Brier scores over time.
-In Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our penalized logistic regression model compared to both Cox regression models.
-The differences, however, are minimal.
-All three outperform the Kaplan-Meier model.
-
 # Case study 4---Stanford Heart Transplant Data
@@
 In the previous case studies, we only considered covariates that were fixed at baseline.
-In this next case study, we will use the Stanford Heart Transplant data [@clark1971cardiac,@crowley1977covariance] to show how case-base sampling can also be used in the context of time-varying covariates.
+In this next case study, we will use the Stanford Heart Transplant data [@clark1971cardiac,@crowley1977covariance] to show how case-base sampling can also be used in the context of time-dependent exposure.
 As an example that already appeared in the literature, case-base sampling was used to study vaccination safety, where the exposure period was defined as the week following vaccination [@saarela2015case].
@@
   exposure = case_when(
-    is.na(txtime) ~ 0L,
+    is.na(txtime) ~ 0L, # No transplant
     txtime > futime ~ 0L,
@@
 
-We can then compare the goodness of fit of these four models using the Akaike Information Criterion (AIC).
+We can then compare the goodness of fit of these three models using the Akaike Information Criterion (AIC).
 
@@
 As a methodological framework, case-base sampling is very flexible.
-This flexibility has been explored before in the literature: for example, Saarela and Hanley [-@saarela2015case] used case-base sampling to model a time-dependent exposure variable in a vaccine safety study.
+Some of this flexibility has been explored before in the literature: for example, Saarela and Hanley [-@saarela2015case] used case-base sampling to model a time-dependent exposure variable in a vaccine safety study.
 As another example, Saarela and Arjas [-@saarela2015non] combined case-base sampling and a Bayesian non-parametric framework to compute individualized risk assessments for chronic diseases.
@@
 On the other hand, we also displayed the flexibility of case-base sampling by showing how it could be combined with penalized logistic regression to perform variable selection.
+Furthermore, the second and fourth case studies showed how case-base sampling can respectively be applied to competing risks and time-varying exposure settings.
 Even though we did not illustrate it in this article, case-base sampling can also be combined with the framework of *generalized additive models*.

### 2020-08-31 analysis/paper/paper.tex
final edits for CS4

--- 2020-08-25 analysis/paper/paper.tex
@@
 \citep{breslow1972discussion}. This in turn leads to stepwise estimates
-of the survival function.
+of the survival function that can be hard to interpret.
 
@@
 for estimating fully parametric hazard models via logistic regression.
-Their approach provides users that are familiar with generalized linear
-models with a natural way of fitting parametric survival models.
-Moreover, their framework is very flexible: general functions of time
-can be estimated (e.g.~using splines or general additive models), and
-hence these models retain some of the flexibility of Cox's partial
-likelihood approach.
+Their main idea is simple: comparing person-moments when the event of
+interest occurred with moments when patients were at risk. Their
+approach provides users familiar with generalized linear models with a
+natural way of fitting parametric survival models. Moreover, their
+framework is very flexible: general functions of time can be estimated
+(e.g.~using splines or general additive models), and hence these models
+retain some of the flexibility of Cox's partial likelihood approach. And
+since the unit of analysis is a person moment, time-varying covariates
+can also easily be included in this framework.
 
@@
 \midrule
-Sex & 0.70 & (0.4, 1.23) & 0.68 & (0.39, 1.19)\\
-Disease & 0.53 & (0.29, 0.95) & 0.52 & (0.29, 0.93)\\
-Phase (CR2 vs. CR1) & 1.12 & (0.45, 2.78) & 1.21 & (0.48, 3.02)\\
-Phase (CR3 vs. CR1) & 1.45 & (0.37, 5.66) & 1.67 & (0.43, 6.5)\\
-Phase (Relapse vs. CR1) & 4.04 & (1.88, 8.68) & 4.55 & (2.09, 9.9)\\
+Sex & 0.74 & (0.43, 1.3) & 0.68 & (0.39, 1.19)\\
+Disease & 0.52 & (0.29, 0.94) & 0.52 & (0.29, 0.93)\\
+Phase (CR2 vs. CR1) & 1.20 & (0.48, 3) & 1.21 & (0.48, 3.02)\\
+Phase (CR3 vs. CR1) & 1.59 & (0.41, 6.17) & 1.67 & (0.43, 6.5)\\
+Phase (Relapse vs. CR1) & 4.09 & (1.9, 8.81) & 4.55 & (2.09, 9.9)\\
 \addlinespace
-Source & 1.48 & (0.48, 4.54) & 1.46 & (0.47, 4.54)\\
-Age & 0.99 & (0.97, 1.01) & 0.99 & (0.97, 1.02)\\
+Source & 1.70 & (0.55, 5.23) & 1.46 & (0.47, 4.54)\\
+Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
 \bottomrule
@@
 contained within a 95\% confidence band around the Fine-Gray absolute
-risk curve (figure not shown). \# Case study 3---SUPPORT Data
+risk curve (figure not shown).
+
+\hypertarget{case-study-3support-data}{%
+\section{Case study 3---SUPPORT Data}\label{case-study-3support-data}}
 
@@
 curve on the test data. We first create a matrix \code{newx} containing
-the relevant variables. Using \code{times}, we specify the time points
-at which the cumulative incidence will be computed. Finally, using the
-parameter \code{s = "lambda.1se"}, we specify at which value of the
-tuning parameter we want to retrieve the coefficient estimates.
+the relevant variables. Next, using \code{times}, we specify the time
+points at which the cumulative incidence will be computed. Finally,
+using the parameter \code{s = "lambda.1se"}, we specify at which value
+of the tuning parameter we want to retrieve the coefficient estimates.
 
@@
 Shrinkage occurs due to penalization.}\label{fig:coefplots}
-\end{figure}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/riskregressionBrier-1}
-
-}
-
-\caption{\label{fig:cs3FinalBrier} (A) Four curves comparing the probabilities of death as it changes over follow-up time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
-All follow-up times seen in the new data were used to generate the curves.
-The Cox-based curves appear to be smooth, but this is due to the 1703 follow-up times seen while fitting the models and the 319 time points used to estimate the curves.
-(B) Four curves demonstrating the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
-Lower is better.
-Penalized logistic regression produced the lowest Brier scores over follow-up time.}\label{fig:riskregressionBrier}
 \end{figure}
@@
 Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our
-penalized logistic regression model compared to both Cox regression
-models. The differences, however, are minimal. All three outperform the
-Kaplan-Meier model.
+penalized case-base model compared to both Cox regression models. The
+differences, however, are minimal. All three outperform the Kaplan-Meier
+estimator.
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/riskregressionBrier-1}
+
+}
+
+\caption{\label{fig:cs3FinalBrier} Comparison of Cox regression, penalized Cox regression, penalized case-base estimation, and Kaplan-Meier.
+(A)  Probabilities of death as it changes over follow-up time.
+(B) Brier scores as they change over time, where a lower score corresponds to better performace.
+Penalized case-base estimation produced the lowest results over follow-up time.}\label{fig:riskregressionBrier}
+\end{figure}
+\end{CodeChunk}
 
@@
 \citep[\citet{crowley1977covariance}]{clark1971cardiac} to show how
-case-base sampling can also be used in the context of time-varying
-covariates. As an example that already appeared in the literature,
+case-base sampling can also be used in the context of time-dependent
+exposure. As an example that already appeared in the literature,
 case-base sampling was used to study vaccination safety, where the
@@
 
-We can then compare the goodness of fit of these four models using the
+We can then compare the goodness of fit of these three models using the
 Akaike Information Criterion (AIC).
@@
 #> Model1 Model2 Model3 
-#>    827    790    788
+#>    827    791    790
 \end{CodeOutput}
@@
 
+\begin{CodeInput}
+R> # Compute absolute risk curves
+R> newdata <- data.frame(exposure = c(0, 1))
+R> absrisk <- absoluteRisk(fit3,
+R+   newdata = newdata,
+R+   time = seq(0, 425, length.out = 100)
+R+ )
+R> 
+R> class(absrisk)
+\end{CodeInput}
+
 \begin{CodeOutput}
@@
 
+\begin{CodeInput}
+R> 
+R> plot(absrisk,
+R+      id.names = c("No Tx", "Tx")) + 
+R+   ylab("Cumulative Incidence") +
+R+   xlab("Follow-up time (in days)") +
+R+   paper_gg_theme
+\end{CodeInput}
+
 \begin{center}\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-risk-1} \end{center}
-
-\end{CodeChunk}
-
-We can also plot the hazard ratio:
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/unnamed-chunk-11-1} \end{center}
 
@@
 
-As a methodological framework, case-base sampling is very flexible. This
-flexibility has been explored before in the literature: for example,
-Saarela and Hanley \citeyearpar{saarela2015case} used case-base sampling
-to model a time-dependent exposure variable in a vaccine safety study.
-As another example, Saarela and Arjas \citeyearpar{saarela2015non}
-combined case-base sampling and a Bayesian non-parametric framework to
-compute individualized risk assessments for chronic diseases. In the
-case studies above, we further explored this flexibility along two
-fronts. On the one hand, we showed how splines could be used as part of
-the linear predictor to model the effect of time on the hazard. This
-strategy yielded estimates of the survival function that were
-qualitatively similar to semiparametric estimates derived from Cox
-regression; however, case-base sampling led to estimates of the survival
-function that \emph{vary smoothly in time} and are thus easier to
-interpret. On the other hand, we also displayed the flexibility of
-case-base sampling by showing how it could be combined with penalized
-logistic regression to perform variable selection. Even though we did
-not illustrate it in this article, case-base sampling can also be
-combined with the framework of \emph{generalized additive models}. This
-functionality has already been implemented in the package. Similarly,
-case-base sampling can be combined with quasi-likelihood estimation to
-fit survival models that can account for the presence of
-over-dispersion. All of these examples illustrate how the case-base
-sampling framework in general, and the package \pkg{casebase} in
-particular, allows the user to fit a broad and flexible family of
-survival functions.
+As a methodological framework, case-base sampling is very flexible. Some
+of this flexibility has been explored before in the literature: for
+example, Saarela and Hanley \citeyearpar{saarela2015case} used case-base
+sampling to model a time-dependent exposure variable in a vaccine safety
+study. As another example, Saarela and Arjas
+\citeyearpar{saarela2015non} combined case-base sampling and a Bayesian
+non-parametric framework to compute individualized risk assessments for
+chronic diseases. In the case studies above, we further explored this
+flexibility along two fronts. On the one hand, we showed how splines
+could be used as part of the linear predictor to model the effect of
+time on the hazard. This strategy yielded estimates of the survival
+function that were qualitatively similar to semiparametric estimates
+derived from Cox regression; however, case-base sampling led to
+estimates of the survival function that \emph{vary smoothly in time} and
+are thus easier to interpret. On the other hand, we also displayed the
+flexibility of case-base sampling by showing how it could be combined
+with penalized logistic regression to perform variable selection.
+Furthermore, the second and fourth case studies showed how case-base
+sampling can respectively be applied to competing risks and time-varying
+exposure settings. Even though we did not illustrate it in this article,
+case-base sampling can also be combined with the framework of
+\emph{generalized additive models}. This functionality has already been
+implemented in the package. Similarly, case-base sampling can be
+combined with quasi-likelihood estimation to fit survival models that
+can account for the presence of over-dispersion. All of these examples
+illustrate how the case-base sampling framework in general, and the
+package \pkg{casebase} in particular, allows the user to fit a broad and
+flexible family of survival functions.
 
@@
 
-This report was generated on 2020-08-25 12:59:22 using the following
+This report was generated on 2020-08-31 10:05:10 using the following
 computational environment and dependencies:
@@
 #>  [1] broom_0.7.0               dotwhisker_0.5.0         
-#>  [3] timereg_1.9.6             glue_1.4.1               
+#>  [3] timereg_1.9.4             glue_1.4.1               
 #>  [5] scales_1.1.1              forcats_0.5.0            
-#>  [7] stringr_1.4.0             dplyr_1.0.1              
+#>  [7] stringr_1.4.0             dplyr_1.0.2              
 #>  [9] purrr_0.3.4               readr_1.3.1              
@@
 #> [15] visreg_2.7.0              tibble_3.0.3             
-#> [17] riskRegression_2020.06.25 prodlim_2019.11.13       
+#> [17] riskRegression_2020.02.05 prodlim_2019.11.13       
 #> [19] flexsurv_1.1.1            survival_3.2-3           
@@
 #> [27] cowplot_1.0.0             colorspace_1.4-1         
-#> [29] casebase_0.9.0           
+#> [29] casebase_0.9.0.9000      
 #> 
@@
 #>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
-#>  [4] ggstance_0.3.4      htmlTable_2.0.1     fs_1.4.2           
-#>  [7] base64enc_0.1-3     rstudioapi_0.11     farver_2.0.3       
+#>  [4] ggstance_0.3.4      htmlTable_1.13.3    base64enc_0.1-3    
+#>  [7] fs_1.4.1            rstudioapi_0.11     farver_2.0.3       
 #> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-1      
@@
 #> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
-#> [28] quantreg_5.61       tools_4.0.2         gtable_0.3.0       
+#> [28] quantreg_5.55       tools_4.0.2         gtable_0.3.0       
 #> [31] Rcpp_1.0.5          cellranger_1.1.0    vctrs_0.3.2        
-#> [34] nlme_3.1-148        conquer_1.0.1       iterators_1.0.12   
-#> [37] xfun_0.16           rvest_0.3.6         lifecycle_0.2.0    
-#> [40] polspline_1.1.19    muhaz_1.2.6.1       MASS_7.3-51.6      
-#> [43] zoo_1.8-8           hms_0.5.3           sandwich_2.5-1     
-#> [46] SparseM_1.78        RColorBrewer_1.1-2  rticles_0.14       
-#> [49] yaml_2.2.1          gridExtra_2.3       rms_6.0-1          
-#> [52] rpart_4.1-15        latticeExtra_0.6-29 stringi_1.4.6      
-#> [55] foreach_1.5.0       checkmate_2.0.0     lava_1.6.7         
-#> [58] shape_1.4.4         mets_1.2.7.1        rlang_0.4.7        
-#> [61] pkgconfig_2.0.3     matrixStats_0.56.0  evaluate_0.14      
-#> [64] lattice_0.20-41     labeling_0.3        htmlwidgets_1.5.1  
-#> [67] cmprsk_2.2-10       tidyselect_1.1.0    deSolve_1.28       
-#> [70] R6_2.4.1            generics_0.0.2      Hmisc_4.4-0        
-#> [73] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
-#> [76] pillar_1.4.6        haven_2.3.1         foreign_0.8-80     
-#> [79] mgcv_1.8-31         nnet_7.3-14         mstate_0.2.12      
-#> [82] modelr_0.1.8        crayon_1.3.4        rmarkdown_2.3      
-#> [85] jpeg_0.1-8.1        grid_4.0.2          readxl_1.3.1       
-#> [88] blob_1.2.1          reprex_0.3.0        digest_0.6.25      
-#> [91] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_4.0.2       
-#> [94] munsell_0.5.0       viridisLite_0.3.0   quadprog_1.5-8
+#> [34] nlme_3.1-148        iterators_1.0.12    xfun_0.15          
+#> [37] rvest_0.3.6         lifecycle_0.2.0     polspline_1.1.17   
+#> [40] muhaz_1.2.6.1       MASS_7.3-51.6       zoo_1.8-8          
+#> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
+#> [46] RColorBrewer_1.1-2  rticles_0.14        yaml_2.2.1         
+#> [49] gridExtra_2.3       rms_5.1-4           rpart_4.1-15       
+#> [52] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
+#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
+#> [58] rlang_0.4.7         pkgconfig_2.0.3     evaluate_0.14      
+#> [61] lattice_0.20-41     labeling_0.3        htmlwidgets_1.5.1  
+#> [64] cmprsk_2.2-10       tidyselect_1.1.0    deSolve_1.28       
+#> [67] plyr_1.8.6          R6_2.4.1            generics_0.0.2     
+#> [70] Hmisc_4.4-0         multcomp_1.4-13     DBI_1.1.0          
+#> [73] withr_2.2.0         pillar_1.4.6        haven_2.3.1        
+#> [76] foreign_0.8-80      mgcv_1.8-31         nnet_7.3-14        
+#> [79] mstate_0.2.12       modelr_0.1.8        crayon_1.3.4       
+#> [82] rmarkdown_2.3       jpeg_0.1-8.1        grid_4.0.2         
+#> [85] readxl_1.3.1        blob_1.2.1          reprex_0.3.0       
+#> [88] digest_0.6.25       webshot_0.5.2       numDeriv_2016.8-1.1
+#> [91] stats4_4.0.2        munsell_0.5.0       viridisLite_0.3.0  
+#> [94] quadprog_1.5-8
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> Local:    master /home/jesseislam/Documents/gits/cbpaper
-#> Remote:   master @ origin (git@github.com:Jesse-Islam/cbpaper.git)
-#> Head:     [ffe8b03] 2020-08-25: Merge branch 'master' into sbmaster
+#> Local:    editMax /Users/maxturgeon/Documents/git_repos/cbpaper
+#> Head:     [0be28f4] 2020-08-25: switch to plot method for hazard
 \end{CodeOutput}

### 2020-09-14 analysis/paper/paper.tex

@@
 R> x <- model.matrix(death ~ . - d.time - aps - sps, 
-R+                   data = train)[, -c(1)] # Remove intercept
++                   data = train)[, -c(1)] # Remove intercept
 R> y <- data.matrix(subset(train, select = c(d.time, death)))
@@
 R> cbFull <- casebase::fitSmoothHazard.fit(x, y,
-R+   family = "glmnet",
-R+   time = "d.time", event = "death",
-R+   formula_time = ~ log(d.time), alpha = 1,
-R+   ratio = 10, standardize = TRUE,
-R+   penalty.factor = c(0, rep(1, ncol(x)))
-R+ )
++   family = "glmnet",
++   time = "d.time", event = "death",
++   formula_time = ~ log(d.time), alpha = 1,
++   ratio = 10, standardize = TRUE,
++   penalty.factor = c(0, rep(1, ncol(x)))
++ )
 \end{CodeInput}
@@
 R> newx <- model.matrix(death ~ . - d.time - aps - sps, 
-R+                   data = test)[, -c(1)]
++                   data = test)[, -c(1)]
 R> times <- sort(unique(test$d.time))
+R> #times<-seq(from=min(times),to=max(times),length.out = 100)
 R> 
 R> abCbFull <- casebase::absoluteRisk(cbFull,
-R+   time = times,
-R+   newdata = newx,
-R+   s = "lambda.1se",
-R+   method = "numerical"
-R+ )
++   time = times,
++   newdata = newx,
++   s = "lambda.1se",
++   method = "numerical"
++ )
 \end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/coefplots-1}
+
+}
+
+\caption{\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models.
+If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity.
+Shrinkage occurs due to penalization.}\label{fig:coefplots}
+\end{figure}
 \end{CodeChunk}
@@
 intersection of coefficients with values greater than 0 are displayed in
-\ref{fig:cs3lolliPlot}. A visible tendency is for Cox regression to have
-larger estimated coefficients. This is expected, as there are no
+Figure \ref{fig:cs3lolliPlot}. A visible tendency is for Cox regression
+to have larger estimated coefficients. This is expected, as there are no
 restrictions on the size of coefficients. The smallest coefficient
-estimates are found in penalized case-base model, aside from cancer
-state (cano).
+estimates are found in penalized logistic model, aside from cancer state
+(cano).
 
@@
 
-\caption{\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models.
-If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity.
-Shrinkage occurs due to penalization.}\label{fig:coefplots}
+\caption{\label{fig:cs3FinalBrier} (A) Four curves comparing the probabilities of death as it changes over follow-up time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
+All follow-up times seen in the new data were used to generate the penalized casebase and K-M curves.
+The Cox-based curves, estimated using the survival package, appear to be smooth because all 1703 follow-up times seen while fitting are used.
+Using user specified times is not permitted in the survival package.
+(B) Four curves demonstrating the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.
+Lower is better.
+Penalized logistic regression produced the lowest Brier scores over follow-up time.}\label{fig:riskregressionBrier}
 \end{figure}
@@
 
-This report was generated on 2020-08-31 10:05:10 using the following
+This report was generated on 2020-09-14 12:43:13 using the following
 computational environment and dependencies:
@@
 #>  [1] broom_0.7.0               dotwhisker_0.5.0         
-#>  [3] timereg_1.9.4             glue_1.4.1               
+#>  [3] timereg_1.9.7             glue_1.4.2               
 #>  [5] scales_1.1.1              forcats_0.5.0            
@@
 #>  [9] purrr_0.3.4               readr_1.3.1              
-#> [11] tidyr_1.1.1               ggplot2_3.3.2            
+#> [11] tidyr_1.1.2               ggplot2_3.3.2            
 #> [13] tidyverse_1.3.0           lubridate_1.7.9          
@@
 #> [21] pracma_2.2.9              magrittr_1.5             
-#> [23] kableExtra_1.1.0          glmnet_4.0-2             
+#> [23] kableExtra_1.2.1          glmnet_4.0-2             
 #> [25] Matrix_1.2-18             data.table_1.13.0        
-#> [27] cowplot_1.0.0             colorspace_1.4-1         
+#> [27] cowplot_1.1.0             colorspace_1.4-1         
 #> [29] casebase_0.9.0.9000      
@@
 #>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
-#>  [4] ggstance_0.3.4      htmlTable_1.13.3    base64enc_0.1-3    
-#>  [7] fs_1.4.1            rstudioapi_0.11     farver_2.0.3       
+#>  [4] ggstance_0.3.4      htmlTable_2.0.1     base64enc_0.1-3    
+#>  [7] fs_1.5.0            rstudioapi_0.11     farver_2.0.3       
 #> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-1      
 #> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
-#> [16] Formula_1.2-3       jsonlite_1.7.0      cluster_2.1.0      
+#> [16] Formula_1.2-3       jsonlite_1.7.1      cluster_2.1.0      
 #> [19] dbplyr_1.4.4        png_0.1-7           compiler_4.0.2     
-#> [22] httr_1.4.2          backports_1.1.8     assertthat_0.2.1   
-#> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
-#> [28] quantreg_5.55       tools_4.0.2         gtable_0.3.0       
-#> [31] Rcpp_1.0.5          cellranger_1.1.0    vctrs_0.3.2        
-#> [34] nlme_3.1-148        iterators_1.0.12    xfun_0.15          
-#> [37] rvest_0.3.6         lifecycle_0.2.0     polspline_1.1.17   
-#> [40] muhaz_1.2.6.1       MASS_7.3-51.6       zoo_1.8-8          
+#> [22] httr_1.4.2          backports_1.1.9     assertthat_0.2.1   
+#> [25] cli_2.0.2           htmltools_0.5.0     quantreg_5.67      
+#> [28] tools_4.0.2         gtable_0.3.0        Rcpp_1.0.5         
+#> [31] cellranger_1.1.0    vctrs_0.3.4         nlme_3.1-149       
+#> [34] conquer_1.0.2       iterators_1.0.12    xfun_0.17          
+#> [37] rvest_0.3.6         lifecycle_0.2.0     polspline_1.1.19   
+#> [40] muhaz_1.2.6.1       MASS_7.3-53         zoo_1.8-8          
 #> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
-#> [46] RColorBrewer_1.1-2  rticles_0.14        yaml_2.2.1         
-#> [49] gridExtra_2.3       rms_5.1-4           rpart_4.1-15       
+#> [46] RColorBrewer_1.1-2  rticles_0.15        yaml_2.2.1         
+#> [49] gridExtra_2.3       rms_6.0-1           rpart_4.1-15       
 #> [52] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
 #> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
-#> [58] rlang_0.4.7         pkgconfig_2.0.3     evaluate_0.14      
-#> [61] lattice_0.20-41     labeling_0.3        htmlwidgets_1.5.1  
-#> [64] cmprsk_2.2-10       tidyselect_1.1.0    deSolve_1.28       
-#> [67] plyr_1.8.6          R6_2.4.1            generics_0.0.2     
-#> [70] Hmisc_4.4-0         multcomp_1.4-13     DBI_1.1.0          
-#> [73] withr_2.2.0         pillar_1.4.6        haven_2.3.1        
-#> [76] foreign_0.8-80      mgcv_1.8-31         nnet_7.3-14        
-#> [79] mstate_0.2.12       modelr_0.1.8        crayon_1.3.4       
-#> [82] rmarkdown_2.3       jpeg_0.1-8.1        grid_4.0.2         
-#> [85] readxl_1.3.1        blob_1.2.1          reprex_0.3.0       
-#> [88] digest_0.6.25       webshot_0.5.2       numDeriv_2016.8-1.1
-#> [91] stats4_4.0.2        munsell_0.5.0       viridisLite_0.3.0  
-#> [94] quadprog_1.5-8
+#> [58] rlang_0.4.7         pkgconfig_2.0.3     matrixStats_0.56.0 
+#> [61] evaluate_0.14       lattice_0.20-41     labeling_0.3       
+#> [64] htmlwidgets_1.5.1   cmprsk_2.2-10       tidyselect_1.1.0   
+#> [67] deSolve_1.28        plyr_1.8.6          R6_2.4.1           
+#> [70] generics_0.0.2      Hmisc_4.4-1         multcomp_1.4-13    
+#> [73] DBI_1.1.0           withr_2.2.0         pillar_1.4.6       
+#> [76] haven_2.3.1         foreign_0.8-80      mgcv_1.8-33        
+#> [79] nnet_7.3-14         mstate_0.2.12       modelr_0.1.8       
+#> [82] crayon_1.3.4        rmarkdown_2.3       jpeg_0.1-8.1       
+#> [85] grid_4.0.2          readxl_1.3.1        blob_1.2.1         
+#> [88] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
+#> [91] numDeriv_2016.8-1.1 stats4_4.0.2        munsell_0.5.0      
+#> [94] viridisLite_0.3.0   quadprog_1.5-8
 \end{CodeOutput}

### 2020-09-15 analysis/paper/paper.Rmd
update CS3

@@
 paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
+# Define colour palette for CS 3
+q4 <- qualitative_hcl(4, palette = "Dark 3")
+names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
 ```
@@
 In the first two case studies, we described the basic functionalities of the \pkg{casebase} package: creating population-time plots, fitting parametric models for hazard functions, and estimating the corresponding cumulative incidence curves.
-<!--In the next two case studies, we explore more advanced features.--> For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function.
+For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function.
 To do this, we simply replace logistic regression with a penalized counterpart.
@@
 For more information about this dataset, the reader is encouraged to look at the documentation in our package.
-A description of the variables can be found in Table \ref{tab:support1}, and a breakdown of each categorical variable appears in Table \ref{tab:support2}.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ccccc}
-\hline
-\textbf{Name} & \textbf{Labels}                          & \textbf{Levels} & \textbf{Storage} & \textbf{NAs} \\
-\hline
-age           & Age                                      &                 & double           & 0            \\
-death         & Death at any time up to NDI date:31DEC94 &                 & double           & 0            \\
-sex           &                                          & 2               & integer          & 0            \\
-slos          & Days from Study Entry to Discharge       &                 & double           & 0            \\
-d.time        & Days of Follow-Up                        &                 & double           & 0            \\
-dzgroup       &                                          & 8               & integer          & 0            \\
-dzclass       &                                          & 4               & integer          & 0            \\
-num.co        & number of comorbidities                  &                 & double           & 0            \\
-edu           & Years of Education                       &                 & double           & 202          \\
-income        &                                          & 4               & integer          & 349          \\
-scoma         & SUPPORT Coma Score based on Glasgow D3   &                 & double           & 0            \\
-avtisst       & Average TISS, Days 3-25                  &                 & double           & 6            \\
-race          &                                          & 5               & integer          & 5            \\
-sps 	        & support physiology score day 3           &	               & double           &	1            \\
-aps           & APS III no coma, imp bun,uout for ph1,D3 &	               & double	          & 1            \\
-hday	        & Day in Hospital at Study Admit	         &                 & double           & 0            \\	
-diabetes      &	Diabetes (Com 27-28, Dx 73)	             &                 & double	          & 0            \\
-dementia      & Dementia (Comorbidity 6)                 &	               & double	          & 0            \\
-ca		        & Cancer State                             & 3               & integer	        & 0            \\
-meanbp        & Mean Arterial Blood Pressure Day 3       &                 & double           & 0            \\
-wblc          & White Blood Cell Count Day 3             &                 & double           & 24           \\
-hrt           & Heart Rate Day 3                         &                 & double           & 0            \\
-resp          & Respiration Rate Day 3                   &                 & double           & 0            \\
-temp          & Temperature (Celsius) Day 3              &                 & double           & 0            \\
-pafi          & PaO2/(.01*FiO2) Day 3                    &                 & double           & 253          \\
-alb           & Serum Albumin Day 3                      &                 & double           & 378          \\
-bili          & Bilirubin Day 3                          &                 & double           & 297          \\
-crea          & Serum creatinine Day 3                   &                 & double           & 3            \\
-sod           & Serum sodium Day 3                       &                 & double           & 0            \\
-ph            & Serum pH (arterial) Day 3                &                 & double           & 250          \\
-glucose       & Glucose Day 3                            &                 & double           & 470          \\
-bun           & BUN Day 3                                &                 & double           & 455          \\
-urine         & Urine Output Day 3                       &                 & double           & 517          \\
-adlp          & Activiites of Daily Living (ADL) Patient Day 3 &           & double           & 634          \\
-adlsc         & Imputed ADL Calibrated to Surrogate      &                 & double           & 0           
-\end{tabular}
-\caption{Description of each variable in the SUPPORT dataset.}
-\label{tab:support1}
-\end{table}
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{cc}
-\hline
-\textbf{Variable} & \textbf{Levels}                      \\
-\hline
-sex      & female                                        \\
-         & male                                          \\
-dzgroup  & ARF/MOSF w/Sepsis                             \\
-         & COPD                                          \\
-         & CHF                                           \\
-         & Cirrhosis                                     \\
-         & Coma                                          \\
-         & Colon Cancer                                  \\
-         & Lung Cancer                                   \\
-         & MOSF w/Malig                                  \\
-dzclass  & ARF/MOSF                                      \\
-         & COPD/CHF/Cirrhosis                            \\
-         & Coma                                          \\
-         & Cancer                                        \\
-         & \textgreater{}\$50k                           \\
-race     & white                                         \\
-         & black                                         \\
-         & asian                                         \\
-         & other                                         \\
-         & hispanic                                      \\
-ca       & Yes                                           \\
-         & No                                            \\
-         & Metastatic                                    \\
-\end{tabular}
-\caption{Description of each level within each categorical variable in the SUPPORT dataset.}
-\label{tab:support2}
-\end{table}
-
-<!--Two variables are SUPPORT day 3 physiology score (sps) and APACHE III day 3 physiology score (aps).
-These scores were developed for a period of 180 days, which makes our comparison over 5.56 years exploratory.
-Therefore, we will assess these two models before using either as our baseline, by comparing them to a full model using all variables.
-Also, our single imputation introduces bias in the data.
-As the goal here is purely demonstrative, this bias will be consistent with any of the following models.
-As such, their comparison is still valid, though any interpretation requires an awareness of said bias.
-The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.-->
-
-In the original study, the authors developed two scores for predicting the outcome.
-These scores are available as part of the data: SUPPORT day 3 physiology score (\code{sps}) and APACHE III day 3 physiology score (\code{aps}).
-Using these scores, we will explore three Cox regression scenarios as a baseline comparison for \pkg{casebase}: \code{sps} only, \code{aps} only, and all covariates excluding \code{sps} and \code{aps}.
-We will compare the three models using a time-dependent version of the classical Brier score that is adjusted for censoring [@graf1999ass].
-The Brier score can be used to assess both discrimination and calibration in a single model.
-Moreover, it provides a metric that can be used to compare parametric and semi-parametric survival models.
-We will use 95% of the observations to train our models, and the remaining 5% to compute absolute risk curves and Brier scores.
+
+In this section, we compare our penalized case-base approach to two other models: Cox regression, and penalized Cox regression (using the \pkg{glmnet} package).
+In all three models, we use all covariates except \code{sps} and \code{aps}.
+These two variables correspond to scores for predicting the outcome that were developed as part of the original study.
+
+To compare the performance of our models, we first divide the data into a training and a test data, with 95% of the observations used for training all three models.
+To assess both discrimination and calibration, we use a time-dependent version of the classical Brier score that is adjusted for censoring [@graf1999ass].
+The Brier score can be used with both parametric and semi-parametric models.
+We use the \pkg{riskRegression} to compute these scores for all models.
 
@@
 
-In Figure \ref{fig:brier1} (A), we can see that all Cox models have similar average survival curves.
-On the other hand, Figure \ref{fig:brier1} (B) shows that the full model has the lowest Brier score overall.
-For this reason, we will use this model in our comparison between regularized Cox regression, Kaplan-Meier estimation, and the approach from \pkg{casebase}.
-
-<!--Before generating an absolute risk curve using regularized estimation, we will explore three options as a baseline comparison.
-The original study compared these scores over a span of 180 days, which may not be comparable over the full 5.56 years.
-Therefore, we will examine three Cox regression scenarios: sps only, aps only and all covariates excluding sps and aps.
-Next, we will compare each modeling using the Brier score [@graf1999ass].
-The Brier score can demonstrate both discrimination and calibration in a single model [@graf1999ass] and provide a comparable metric between semi-parametric and parametric methods in the survival context.
-For all the following analyses in this case study, we will use 80% of the observations to train our models, while 20% will be used to generate absolute risk curves, all of which will be averaged to demonstrate each model's predictive ability.
-Over 5.56 years, \ref{fig:baselines} demonstrates that all the cox models have similar, if not identical average predictions.
-Figure \ref{fig:brier1} examines the brier scores as they change over time.
-As the full model has the lowest brier score on overall, it will be used in the following absolute risk comparison to regularized cox, regularized estimation and the unadjusted Kaplan-Meier curve.-->
-
-```{r baselinePlotsPrep, eval=eval_cs3, echo = FALSE}
-# Define colour palette for this section
-q4 <- qualitative_hcl(4, palette = "Dark 3")
-
-# 1. Cox with only sps
-coxSPS <- survival::coxph(Surv(time = d.time, event = death) ~ sps, 
+```{r supportCox_fit, eval=eval_cs3, echo=FALSE, cache=FALSE}
+# Cox with everything but sps and aps
+cox <- survival::coxph(Surv(time = d.time, event = death) ~ . - aps - sps,
                            data = train, x = TRUE)
-abcoxSPS <- survival::survfit(coxSPS, newdata = test)
-# 2. Cox with only aps
-coxAPS <- survival::coxph(Surv(time = d.time, event = death) ~ aps, 
-                          data = train, x = TRUE)
-abcoxAPS <- survival::survfit(coxAPS, newdata = test)
-# 3. Cox with everything but sps and aps
-coxFull <- survival::coxph(Surv(time = d.time, event = death) ~ . - aps - sps, 
-                           data = train, x = TRUE)
-abcoxFull <- survival::survfit(coxFull, newdata = test)
-# 4. Kaplan-Meier
-KM <- survival::coxph(Surv(time = d.time, event = death) ~ 1, data = test,x=TRUE)
-abKM <- survival::survfit(KM)
-abKMcompare <- data.frame(
-  Time = abKM$time,
-  KM = 1 - abKM$surv
-)
-
-# Combine all CI estimates from Cox models
-baseCoxComparisonData <- data.frame(
-  Time = abcoxSPS$time,
-  SPS = 1 - rowMeans(abcoxSPS$surv),
-  APS = 1 - rowMeans(abcoxAPS$surv),
-  Full = 1 - rowMeans(abcoxFull$surv)
-)
-```
-
-```{r baselines, echo=FALSE, eval=eval_cs3}
-cox_ci <- ggplot(baseCoxComparisonData, aes(Time)) +
-  geom_step(aes(y = APS, colour = "APS")) +
-  geom_step(aes(y = Full, col = "Full")) +
-  geom_step(aes(y = SPS, colour = "SPS")) +
-  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
-  labs(y = "Probability of Death", x = "Follow-up time (Years)", color = "Models") +
-  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
-  paper_gg_theme + 
-  scale_x_continuous(labels = scaleFUNx, breaks = round(seq(0, 5, by = 1))) + 
-  scale_y_continuous(labels = scaleFUNy, n.breaks = 8)
-```
-
-```{r brierpt1, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:brier1} Comparison of three Cox regression models to the unadjusted Kaplan-Meier curve.
-(A) Cumulative Incidence curves.
-For a given model, the plotted curve is the average curve of each individual in the study.
-Each follow-up time seen in the study was used.
-There are minimal differences between the curves.
-APS coincides with the SPS curve, as they are considerably overlapped.(B) Time-dependent Brier scores, where a lower score is better.
-The full model performs best overall."}
-# Extract time points
-times <- sort(unique(test$d.time))
-# Compute Brier scores
-brierAPS <- Score(list("PredictionModel" = coxAPS), data = test, 
-                  formula = Hist(d.time, death != 0) ~ 1, summary = "ipa", 
-                  se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
-brierSPS <- Score(list("PredictionModel" = coxSPS), data = test, 
-                  formula = Hist(d.time, death != 0) ~ 1, summary = "ipa", 
-                  se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
-
-# First fix NA coefficients, then compute Brier score
-coxFull$coefficients[which(is.na(coxFull$coefficients))] <- 0
-brierFull <- Score(list("PredictionModel" = coxFull), data = test, 
-                   formula = Surv(d.time, death != 0) ~ 1, summary = "ipa", 
-                   se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
-brierNull <- Score(list("PredictionModel" = KM), data = test, 
-                   formula = Surv(d.time, death != 0) ~ 1, summary = "ipa", 
-                   se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
-
-# Combine all results for plotting
-brierPlotter <- rbind(
-  data.frame(brierScore = brierAPS$Brier$score$Brier[brierAPS$Brier$score$model == "PredictionModel"], 
-             times = times, model = "APS"),
-  data.frame(brierScore = brierFull$Brier$score$Brier[brierFull$Brier$score$model == "PredictionModel"], 
-             times = times, model = "Full"),
-  data.frame(brierScore = brierSPS$Brier$score$Brier[brierSPS$Brier$score$model == "PredictionModel"], 
-             times = times, model = "SPS"),
-    data.frame(brierScore = brierNull$Brier$score$Brier[brierNull$Brier$score$model == "PredictionModel"], 
-             times = times, model = "K-M")
-)
-
-cox_brier <- ggplot(data = brierPlotter, mapping = aes(x = times, y = brierScore, col = model)) +
-  geom_line() +
-  xlab("Follow-up time (Years)") +
-  ylab("Brier score") +
-  labs(color = "Models") +
-  # ggtitle("Brier scores over time") +
-  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
-  paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1)))+scale_y_continuous(labels=scaleFUNy,n.breaks=8)
-
-# Combine both plots using cowplot
-plot_row <- cowplot::plot_grid(cox_ci + theme(legend.position = 'none'),
-                               cox_brier + theme(legend.position = 'none'),
-                               labels = c("A", "B"))
-legend <- cowplot::get_legend(
-  # create some space to the left of the legend
-  cox_ci + theme(legend.box.margin = margin(0, 0, 0, 12))
-)
-
-cowplot::plot_grid(plot_row, legend, rel_heights = c(3, .4), nrow = 2)
-```
-
-<!--To get the absolute risk following regularized cox regression, we had to work around limitations in existing packages.
-To make use of the \pkg{survival} package, we first estimate the hazard coefficients using thecoxnet implementation for lasso penalization in \pkg{glmnet} [@regpathcox].
-Next, we fit a cox model using the survival package with the selected variables only, from our regularized fit.
-Finally, we replaced the coefficients in the survival object with those from our glmnet fit, allowing us to make use of the absolute risk tools available in the \pkg{survival} package.-->
-
-\newpage
-
-For our penalized model, we opt for the natural log of time; recall that this corresponds to a Weibull distribution.
+```
+
+For our penalized case-base model, we opt for the natural log of time; recall that this corresponds to a Weibull distribution.
 For fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface to the function \code{fitSmoothHazard}.
@@
 We used lasso penalization by setting \code{alpha = 1}.
-Moreover, we use the argument \code{penalty.factor} to force the time variable into the final model.
+Moreover, we use the argument \code{penalty.factor} to give a regularization weight of 0 to the time variable, and therefore force it into the final model.
 
 ```{r supportCB_fit, eval=eval_cs3, echo=TRUE, cache=FALSE}
+# Create matrices for inputs
 x <- model.matrix(death ~ . - d.time - aps - sps, 
@@
 # Regularized logistic regression to estimate hazard
-cbFull <- casebase::fitSmoothHazard.fit(x, y,
+pen_cb <- casebase::fitSmoothHazard.fit(x, y,
   family = "glmnet",
@@
 
-Next, we can use the object \code{cbFull} to estimate the absolute risk curve on the test data.
+<!-- Next, we can use the object \code{cbFull} to estimate the absolute risk curve on the test data.
 We first create a matrix \code{newx} containing the relevant variables.
@@
 Finally, using the parameter \code{s = "lambda.1se"}, we specify at which value of the tuning parameter we want to retrieve the coefficient estimates.
-
-```{r supportCB_abs, eval=eval_cs3, echo=TRUE, cache=TRUE}
-# Estimating the absolute risk curve using the newdata parameter
-newx <- model.matrix(death ~ . - d.time - aps - sps, 
-                  data = test)[, -c(1)]
-times <- sort(unique(test$d.time))
-#times<-seq(from=min(times),to=max(times),length.out = 100)
-
-abCbFull <- casebase::absoluteRisk(cbFull,
-  time = times,
-  newdata = newx,
-  s = "lambda.1se",
-  method = "numerical"
-)
-```
+-->
 
@@
 names(coxNet$coefficients) <- coxNet_coefnames
-
-# Fitting absolute risk curve for cox+glmnet
-abcoxNet <- survival::survfit(coxNet, type = "breslow", 
-                              newdata = as.data.frame(newx[, nonzero_covariate_cox$X1]))
-
-# Combine CI estimates from both penalized methods
-abCN <- data.frame(
-  Time = abcoxNet$time,
-  absRisk = 1 - rowMeans(abcoxNet$surv)
-)
-abCN$Model <- "Penalized Cox"
-
-abCB <- data.frame(Time = abCbFull[, 1],
-                   absRisk = rowMeans(abCbFull[, -c(1)]))
-abCB$Model <- "Penalized CB"
-
-ab <- rbind(abCB, abCN)
-
 ```
@@
 Our first comparison plot examines the covariates that were selected between all three hazard functions.
-Note that Cox regression without penalization selected more covariates than the other two, but only the intersection of coefficients with values greater than 0 are displayed in \ref{fig:cs3lolliPlot}.
-A visible tendency is for Cox regression to have larger estimated coefficients.
-This is expected, as there are no restrictions on the size of coefficients.
-The smallest coefficient estimates are found in penalized case-base model, aside from cancer state (cano).
+In \ref{fig:cs3lolliPlot}, we show the coefficient estimates for covariates that we selected by both penalized Cox and penalized case-base.
+From this plot, we can clearly see the shrinkage effect of using penalized estimation.
 
 ```{r coefplots, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models.
-If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity.
+Only covariates common among all three models were retained.
 Shrinkage occurs due to penalization."}
@@
 
-estimatescb <- data.table::setDT(as.data.frame(coef(cbFull)[-c(1, 2), 1]), 
+estimatescb <- data.table::setDT(as.data.frame(coef(pen_cb)[-c(1, 2), 1]),
                                  keep.rownames = TRUE)
@@
 
-estimatescoxFull <- data.table::setDT(as.data.frame(coef(coxFull)), 
+estimatescox <- data.table::setDT(as.data.frame(coef(cox)), 
                                   keep.rownames = TRUE)
-# estimatescoxFull$std.error<-0
-estimatescoxFull$Model <- "Cox"
-colnames(estimatescoxFull) <- c("term", "estimate", "model")
+# estimatescox$std.error<-0
+cox$coefficients[which(is.na(cox$coefficients))] <- 0
+estimatescox$Model <- "Cox"
+colnames(estimatescox) <- c("term", "estimate", "model")
 
 # Clean up the data and the variable names
-lolliplotDots <- rbind(estimatescoxFull, estimatescoxNet, estimatescb)
+lolliplotDots <- rbind(estimatescox, estimatescoxNet, estimatescb)
 lolliplotDots$conf.low <- lolliplotDots$estimate
@@
   labs(color = "Models") +
-  scale_color_manual(values = q4[1:3])
-```
-
-These values do not inform us of model performance.
-To do so, we take our hazard functions and use them to predict the absolute risk of everyone in our test set, over time.
-The probabilities over time for each individual are averaged, resulting in the absolute risk curves in Figure \ref{fig:cs3FinalBrier} (A).
-There is separation between the curves, but it is minimal.
-To better demonstrate the differences in performance between the models, we compare Brier scores over time.
-In Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our penalized case-base model compared to both Cox regression models.
-The differences, however, are minimal.
-All three outperform the Kaplan-Meier estimator.
-
-```{r cs3Final, echo=FALSE, eval=eval_cs3}
-reg_ci <- ggplot(ab, aes(Time)) +
-  geom_step(data = baseCoxComparisonData, mapping = aes(x = Time,y = Full, colour = "Cox")) +
-  geom_line(aes(y = absRisk, colour = Model)) +
-  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
+  scale_color_manual(values = q4)
+```
+
+Next, we compare the cumulative incidence estimation over the test set.
+The probabilities over time for each observation are averaged, resulting in the absolute risk curves in Figure \ref{fig:cs3FinalBrier} (A).
+We can see some minimal differences between the three models, with the Kaplan-Meier estimate giving a lower estimate than our three models.
+We also note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard.
+As described above, we compare the performance between the models by computing the Brier scores over time.
+In Figure \ref{fig:cs3FinalBrier} (B), we can see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models.
+On the other hand, the differences between these three models are minimal.
+
+```{r support_abs, eval=eval_cs3, echo=FALSE}
+# Absolute Risks
+newx <- model.matrix(death ~ . - d.time - aps - sps,
+                     data = test)[, -c(1)]
+times <- sort(unique(test$d.time))
+# 1. Unpenalized Cox
+abcox <- survival::survfit(cox, newdata = test)
+
+# 2. Penalized Cox
+abcoxNet <- survival::survfit(coxNet, type = "breslow", 
+                              newdata = as.data.frame(newx[, nonzero_covariate_cox$X1]))
+
+# 3. Penalized Case-base
+abpen_cb <- casebase::absoluteRisk(pen_cb,
+  time = times,
+  newdata = newx,
+  s = "lambda.1se",
+  method = "numerical"
+)
+
+# 4. Kaplan-Meier
+KM <- survival::coxph(Surv(time = d.time, event = death) ~ 1, data = test,
+                      x = TRUE)
+abKM <- survival::survfit(KM)
+
+# Combine absolute risk estimates
+data_absRisk <- bind_rows(
+  data.frame(Time = abcox$time, Prob = 1 - rowMeans(abcox$surv),
+             Model = "Cox"),
+  data.frame(Time = abcoxNet$time, Prob = 1 - rowMeans(abcoxNet$surv),
+             Model = "Pen. Cox"),
+  data.frame(Time = abpen_cb[, 1], Prob = rowMeans(abpen_cb[, -c(1)]),
+             Model = "Pen. CB"),
+  data.frame(Time = abKM$time, Prob = 1 - abKM$surv,
+             Model = "K-M")
+) %>% 
+  mutate(Model = factor(Model, levels = names(q4)))
+```
+
+```{r absRiskPlot, echo=FALSE, eval=eval_cs3}
+reg_ci <- ggplot(data_absRisk, aes(Time, Prob)) +
+  geom_line(aes(colour = Model)) +
   labs(y = "Probability of death", x = "Follow-up time (years)", color = "Models") +
-  # ggtitle("Absolute risk curves") +
-  scale_color_manual(values = q4, breaks = c("K-M", "Cox", "Penalized Cox", "Penalized CB")) + 
+  scale_color_manual(values = q4) + 
+  expand_limits(y = 1) +
   paper_gg_theme + 
@@
 
-```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} Comparison of Cox regression, penalized Cox regression, penalized case-base estimation, and Kaplan-Meier.
-(A)  Probabilities of death as it changes over follow-up time.
-(B) Brier scores as they change over time, where a lower score corresponds to better performace.
-Penalized case-base estimation produced the lowest results over follow-up time.", warning = FALSE}
-# Compute Brier score for penalized models
-brierFinalResults <- Score(list("Penalized Cox" = coxNet, "Penalized CB" = cbFull),
+```{r support_Brier, eval=eval_cs3, echo=FALSE}
+# Brier score
+# 1. Unpenalized models
+# First fix NA coefficients, then compute Brier score
+brierCoxKM <- Score(list("Cox" = cox,
+                         "K-M" = KM), data = test,
+                   formula = Hist(d.time, death != 0) ~ 1, summary = NULL,
+                   se.fit = FALSE, metrics = "brier", contrasts = FALSE, 
+                   times = times)
+
+# 2. Penalized models
+brierPenalized <- Score(list("Pen. Cox" = coxNet, 
+                             "Pen. CB" = pen_cb),
                         data = cbind(subset(test, select = c(d.time, death)),
@@
                         formula = Hist(d.time, death != 0) ~ 1, summary = NULL, 
-                           se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
-#brierFinalResults$models<-as.character(brierFinalResults$models)
-# Combine Brier scores between unpenalized and penalized models before plotting
-
+                        se.fit = FALSE, metrics = "brier", contrasts = FALSE, 
+                        times = times)
+```
+
+```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} Comparison of Cox regression, penalized Cox regression, penalized case-base estimation, and Kaplan-Meier.
+(A)  Probabilities of death as it changes over follow-up time.
+(B) Brier scores as they change over time, where a lower score corresponds to better performace.", warning = FALSE}
+# Combine scores
 data_brier <- bind_rows(
-  brierFinalResults$Brier$score,
-  filter(brierFull$Brier$score, model == "PredictionModel")
+  brierCoxKM$Brier$score %>% 
+    mutate(model = as.character(model)) %>% 
+    filter(model != "Null model"),
+  brierPenalized$Brier$score %>% 
+    mutate(model = as.character(model)) %>% 
+    filter(model != "Null model")
   ) %>% 
-  mutate(model = as.character(model), # First return to string, so types match
-         model = case_when(
-    model == "Null model" ~ "K-M",
-    model == "PredictionModel" ~ "Cox",
-    TRUE ~ model
-  ),
-  model = factor(model, levels = c("Cox", "Penalized Cox", "Penalized CB", "K-M"),
-                 labels = c("Cox", "Pen. Cox", "Pen. CB", "K-M")))
-
-reg_brier <- ggplot(data = data_brier, mapping = aes(x = times, y = Brier, col = model)) +
+  mutate(model = factor(model, levels = c("Cox", "Pen. Cox", "Pen. CB", "K-M")))
+
+reg_brier <- ggplot(data = data_brier, aes(x = times, y = Brier, col = model)) +
   geom_line() +
@@
   ylab("Brier score") +
-  # ggtitle("Brier score over time") +
   labs(color = "Models") +
@@
 
-These values do not inform us of model performance.
-To do so, we take our hazard functions and use them to predict the absolute risk of everyone in our test set, over time.
-The probabilities over time for each individual are averaged, resulting in the absolute risk curves in Figure \ref{fig:cs3FinalBrier} (A).
-There is separation between the curves, but it is minimal.
-To better demonstrate the differences in performance between the models, we compare Brier scores over time.
-In Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our penalized logistic regression model compared to both Cox regression models.
-The differences, however, are minimal.
-All three outperform the Kaplan-Meier model.
+In brief, case-base sampling can be used in conjunction with penalized logistic regression to perform variable selection with survival models.
 
@@
 In the previous case studies, we only considered covariates that were fixed at baseline.
-In this next case study, we will use the Stanford Heart Transplant data [@clark1971cardiac,@crowley1977covariance] to show how case-base sampling can also be used in the context of time-dependent exposure.
-As an example that already appeared in the literature, case-base sampling was used to study vaccination safety, where the exposure period was defined as the week following vaccination [@saarela2015case].
+In this next case study, we use the Stanford Heart Transplant data [@clark1971cardiac,@crowley1977covariance] to show how case-base sampling can also be used in the context of time-dependent exposure.
+This feature of case-base sampling has been explored in the literature, in the context of vaccination safety  [@saarela2015case].
+In this study, the exposure period was defined as the week following vaccination.
 Hence, the main covariate of interest, i.e.
@@
 
-As before, we can look at the population-time plot for a graphical summary of the event incidence.
-Here, we coloured the exposed person-time (i.e.
+As above, we can look at the population-time plot for a graphical summary of the event incidence.
+Here, we colour the exposed person-time (i.e.
 after transplant) in a darker shade of gray.
@@
 
-```{r stanford_cb, eval = eval_cs4}
+```{r stanford_cb, eval = eval_cs4, echo = TRUE}
 cb_data <- sampleCaseBase(jasa,
@@
 
-```{r stanford_exp, eval = eval_cs4, dependson="stanford_cb"}
+```{r stanford_exp, eval = eval_cs4, dependson="stanford_cb", echo = TRUE}
 # Define exposure variable
@@
 Note that the third model includes an interaction term between exposure and follow-up time.
-In other words, this model no longer exhibit proportional hazards.
+In other words, this model no longer exhibits proportional hazards.
 The evidence of non-proportionality of hazards in the Stanford Heart Transplant data has been widely discussed [@arjas1988graphical].
@@
 
-```{r, eval = eval_cs4, dependson="stanford_fit"}
+```{r, eval = eval_cs4, dependson="stanford_fit", digits = 4}
 # Compute AIC
@@
 ```
+
+<!-- # Appendix -->
+
+<!-- A description of the variables can be found in Table \ref{tab:support1}, and a breakdown of each categorical variable appears in Table \ref{tab:support2}.  -->
+
+<!-- \begin{table}[ht] -->
+<!-- \centering -->
+<!-- \begin{tabular}{ccccc} -->
+<!-- \hline -->
+<!-- \textbf{Name} & \textbf{Labels}                          & \textbf{Levels} & \textbf{Storage} & \textbf{NAs} \\ -->
+<!-- \hline -->
+<!-- age           & Age                                      &                 & double           & 0            \\ -->
+<!-- death         & Death at any time up to NDI date:31DEC94 &                 & double           & 0            \\ -->
+<!-- sex           &                                          & 2               & integer          & 0            \\ -->
+<!-- slos          & Days from Study Entry to Discharge       &                 & double           & 0            \\ -->
+<!-- d.time        & Days of Follow-Up                        &                 & double           & 0            \\ -->
+<!-- dzgroup       &                                          & 8               & integer          & 0            \\ -->
+<!-- dzclass       &                                          & 4               & integer          & 0            \\ -->
+<!-- num.co        & number of comorbidities                  &                 & double           & 0            \\ -->
+<!-- edu           & Years of Education                       &                 & double           & 202          \\ -->
+<!-- income        &                                          & 4               & integer          & 349          \\ -->
+<!-- scoma         & SUPPORT Coma Score based on Glasgow D3   &                 & double           & 0            \\ -->
+<!-- avtisst       & Average TISS, Days 3-25                  &                 & double           & 6            \\ -->
+<!-- race          &                                          & 5               & integer          & 5            \\ -->
+<!-- sps 	        & support physiology score day 3           &	               & double           &	1            \\ -->
+<!-- aps           & APS III no coma, imp bun,uout for ph1,D3 &	               & double	          & 1            \\ -->
+<!-- hday	        & Day in Hospital at Study Admit	         &                 & double           & 0            \\	 -->
+<!-- diabetes      &	Diabetes (Com 27-28, Dx 73)	             &                 & double	          & 0            \\ -->
+<!-- dementia      & Dementia (Comorbidity 6)                 &	               & double	          & 0            \\ -->
+<!-- ca		        & Cancer State                             & 3               & integer	        & 0            \\ -->
+<!-- meanbp        & Mean Arterial Blood Pressure Day 3       &                 & double           & 0            \\ -->
+<!-- wblc          & White Blood Cell Count Day 3             &                 & double           & 24           \\ -->
+<!-- hrt           & Heart Rate Day 3                         &                 & double           & 0            \\ -->
+<!-- resp          & Respiration Rate Day 3                   &                 & double           & 0            \\ -->
+<!-- temp          & Temperature (Celsius) Day 3              &                 & double           & 0            \\ -->
+<!-- pafi          & PaO2/(.01*FiO2) Day 3                    &                 & double           & 253          \\ -->
+<!-- alb           & Serum Albumin Day 3                      &                 & double           & 378          \\ -->
+<!-- bili          & Bilirubin Day 3                          &                 & double           & 297          \\ -->
+<!-- crea          & Serum creatinine Day 3                   &                 & double           & 3            \\ -->
+<!-- sod           & Serum sodium Day 3                       &                 & double           & 0            \\ -->
+<!-- ph            & Serum pH (arterial) Day 3                &                 & double           & 250          \\ -->
+<!-- glucose       & Glucose Day 3                            &                 & double           & 470          \\ -->
+<!-- bun           & BUN Day 3                                &                 & double           & 455          \\ -->
+<!-- urine         & Urine Output Day 3                       &                 & double           & 517          \\ -->
+<!-- adlp          & Activiites of Daily Living (ADL) Patient Day 3 &           & double           & 634          \\ -->
+<!-- adlsc         & Imputed ADL Calibrated to Surrogate      &                 & double           & 0            -->
+<!-- \end{tabular} -->
+<!-- \caption{Description of each variable in the SUPPORT dataset.} -->
+<!-- \label{tab:support1} -->
+<!-- \end{table} -->
+
+<!-- \begin{table}[ht] -->
+<!-- \centering -->
+<!-- \begin{tabular}{cc} -->
+<!-- \hline -->
+<!-- \textbf{Variable} & \textbf{Levels}                      \\ -->
+<!-- \hline -->
+<!-- sex      & female                                        \\ -->
+<!--          & male                                          \\ -->
+<!-- dzgroup  & ARF/MOSF w/Sepsis                             \\ -->
+<!--          & COPD                                          \\ -->
+<!--          & CHF                                           \\ -->
+<!--          & Cirrhosis                                     \\ -->
+<!--          & Coma                                          \\ -->
+<!--          & Colon Cancer                                  \\ -->
+<!--          & Lung Cancer                                   \\ -->
+<!--          & MOSF w/Malig                                  \\ -->
+<!-- dzclass  & ARF/MOSF                                      \\ -->
+<!--          & COPD/CHF/Cirrhosis                            \\ -->
+<!--          & Coma                                          \\ -->
+<!--          & Cancer                                        \\ -->
+<!--          & \textgreater{}\$50k                           \\ -->
+<!-- race     & white                                         \\ -->
+<!--          & black                                         \\ -->
+<!--          & asian                                         \\ -->
+<!--          & other                                         \\ -->
+<!--          & hispanic                                      \\ -->
+<!-- ca       & Yes                                           \\ -->
+<!--          & No                                            \\ -->
+<!--          & Metastatic                                    \\ -->
+<!-- \end{tabular} -->
+<!-- \caption{Description of each level within each categorical variable in the SUPPORT dataset.} -->
+<!-- \label{tab:support2} -->
+<!-- \end{table} -->
+
+<!-- In Figure \ref{fig:brier1} (A), we can see that all Cox models have similar average survival curves.
+On the other hand, Figure \ref{fig:brier1} (B) shows that the full model has the lowest Brier score overall.
+For this reason, we will use this model in our comparison between regularized Cox regression, Kaplan-Meier estimation, and the approach from \pkg{casebase}.
+-->
+
+<!-- ```{r baselinePlotsPrep, eval=eval_cs3, echo = FALSE} -->
+<!-- # 1. Cox with only sps -->
+<!-- coxSPS <- survival::coxph(Surv(time = d.time, event = death) ~ sps,  -->
+<!--                           data = train, x = TRUE) -->
+<!-- abcoxSPS <- survival::survfit(coxSPS, newdata = test) -->
+<!-- # 2. Cox with only aps -->
+<!-- coxAPS <- survival::coxph(Surv(time = d.time, event = death) ~ aps,  -->
+<!--                           data = train, x = TRUE) -->
+<!-- abcoxAPS <- survival::survfit(coxAPS, newdata = test) -->
+<!-- # 3. Cox with everything but sps and aps -->
+<!-- coxFull <- survival::coxph(Surv(time = d.time, event = death) ~ . - aps - sps,  -->
+<!--                            data = train, x = TRUE) -->
+<!-- abcoxFull <- survival::survfit(coxFull, newdata = test) -->
+<!-- # 4. Kaplan-Meier -->
+<!-- KM <- survival::coxph(Surv(time = d.time, event = death) ~ 1, data = test,x=TRUE) -->
+<!-- abKM <- survival::survfit(KM) -->
+<!-- abKMcompare <- data.frame( -->
+<!--   Time = abKM$time, -->
+<!--   KM = 1 - abKM$surv -->
+<!-- ) -->
+
+<!-- # Combine all CI estimates from Cox models -->
+<!-- baseCoxComparisonData <- data.frame( -->
+<!--   Time = abcoxSPS$time, -->
+<!--   SPS = 1 - rowMeans(abcoxSPS$surv), -->
+<!--   APS = 1 - rowMeans(abcoxAPS$surv), -->
+<!--   Full = 1 - rowMeans(abcoxFull$surv) -->
+<!-- ) -->
+<!-- ``` -->
+
+<!-- ```{r baselines, echo=FALSE, eval=eval_cs3} -->
+<!-- cox_ci <- ggplot(baseCoxComparisonData, aes(Time)) + -->
+<!--   geom_step(aes(y = APS, colour = "APS")) + -->
+<!--   geom_step(aes(y = Full, col = "Full")) + -->
+<!--   geom_step(aes(y = SPS, colour = "SPS")) + -->
+<!--   geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) + -->
+<!--   labs(y = "Probability of Death", x = "Follow-up time (Years)", color = "Models") + -->
+<!--   scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) +  -->
+<!--   paper_gg_theme +  -->
+<!--   scale_x_continuous(labels = scaleFUNx, breaks = round(seq(0, 5, by = 1))) +  -->
+<!--   scale_y_continuous(labels = scaleFUNy, n.breaks = 8) -->
+<!-- ``` -->
+
+<!-- ```{r brierpt1, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:brier1} Comparison of three Cox regression models to the unadjusted Kaplan-Meier curve.
+(A) Cumulative Incidence curves.
+For a given model, the plotted curve is the average curve of each individual in the study.
+Each follow-up time seen in the study was used.
+There are minimal differences between the curves.
+APS coincides with the SPS curve, as they are considerably overlapped.(B) Time-dependent Brier scores, where a lower score is better.
+The full model performs best overall."} -->
+<!-- # Extract time points -->
+<!-- times <- sort(unique(test$d.time)) -->
+<!-- # Compute Brier scores -->
+<!-- brierAPS <- Score(list("PredictionModel" = coxAPS), data = test,  -->
+<!--                   formula = Hist(d.time, death != 0) ~ 1, summary = "ipa",  -->
+<!--                   se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times) -->
+<!-- brierSPS <- Score(list("PredictionModel" = coxSPS), data = test,  -->
+<!--                   formula = Hist(d.time, death != 0) ~ 1, summary = "ipa",  -->
+<!--                   se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times) -->
+
+<!-- # First fix NA coefficients, then compute Brier score -->
+<!-- coxFull$coefficients[which(is.na(coxFull$coefficients))] <- 0 -->
+<!-- brierFull <- Score(list("PredictionModel" = coxFull), data = test,  -->
+<!--                    formula = Surv(d.time, death != 0) ~ 1, summary = "ipa",  -->
+<!--                    se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times) -->
+<!-- brierNull <- Score(list("PredictionModel" = KM), data = test,  -->
+<!--                    formula = Surv(d.time, death != 0) ~ 1, summary = "ipa",  -->
+<!--                    se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times) -->
+
+<!-- # Combine all results for plotting -->
+<!-- brierPlotter <- rbind( -->
+<!--   data.frame(brierScore = brierAPS$Brier$score$Brier[brierAPS$Brier$score$model == "PredictionModel"],  -->
+<!--              times = times, model = "APS"), -->
+<!--   data.frame(brierScore = brierFull$Brier$score$Brier[brierFull$Brier$score$model == "PredictionModel"],  -->
+<!--              times = times, model = "Full"), -->
+<!--   data.frame(brierScore = brierSPS$Brier$score$Brier[brierSPS$Brier$score$model == "PredictionModel"],  -->
+<!--              times = times, model = "SPS"), -->
+<!--     data.frame(brierScore = brierNull$Brier$score$Brier[brierNull$Brier$score$model == "PredictionModel"],  -->
+<!--              times = times, model = "K-M") -->
+<!-- ) -->
+
+<!-- cox_brier <- ggplot(data = brierPlotter, mapping = aes(x = times, y = brierScore, col = model)) + -->
+<!--   geom_line() + -->
+<!--   xlab("Follow-up time (Years)") + -->
+<!--   ylab("Brier score") + -->
+<!--   labs(color = "Models") + -->
+<!--   # ggtitle("Brier scores over time") + -->
+<!--   scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) +  -->
+<!--   paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1)))+scale_y_continuous(labels=scaleFUNy,n.breaks=8) -->
+
+<!-- # Combine both plots using cowplot -->
+<!-- plot_row <- cowplot::plot_grid(cox_ci + theme(legend.position = 'none'), -->
+<!--                                cox_brier + theme(legend.position = 'none'), -->
+<!--                                labels = c("A", "B")) -->
+<!-- legend <- cowplot::get_legend( -->
+<!--   # create some space to the left of the legend -->
+<!--   cox_ci + theme(legend.box.margin = margin(0, 0, 0, 12)) -->
+<!-- ) -->
+
+<!-- cowplot::plot_grid(plot_row, legend, rel_heights = c(3, .4), nrow = 2) -->
+<!-- ``` -->

### 2020-09-15 analysis/paper/paper.tex
update CS3

--- 2020-08-31 analysis/paper/paper.tex
@@
 
-\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx} \usepackage{tabularx} \usepackage{float} \usepackage{booktabs} \usepackage{makecell} \usepackage{tabu} \DeclareUnicodeCharacter{2500}{-}
+\usepackage{amsmath} \usepackage{amssymb} \usepackage{longtable} \usepackage{graphicx} \usepackage{tabularx} \usepackage{float} \usepackage{booktabs} \usepackage{makecell} \usepackage{tabu} \DeclareUnicodeCharacter{2500}{-}
 
@@
 R> pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
-R+                                event = "DeadOfPrCa", exposure = "ScrArm")
++                                event = "DeadOfPrCa", exposure = "ScrArm")
 R> inherits(pt_object, "popTime")
@@
 R> fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
-R+              gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
-R+              weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
-R+              splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
++              gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
++              weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
++              splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
 R> 
 R> fits <- lapply(fmla, function(i) {
-R+   fitSmoothHazard(i, data = ERSPC, ratio = 100)
-R+ })
++   fitSmoothHazard(i, data = ERSPC, ratio = 100)
++ })
 \end{CodeInput}
@@
 R> risks <- lapply(fits, function(i) {
-R>   absoluteRisk(object = i, time = new_time, newdata = new_data, method = "mont")
-R> })
++   absoluteRisk(object = i, time = new_time, newdata = new_data, method = "mont")
++ })
 \end{CodeInput}
@@
 likelihood ratio tests built using case-base sampling models. Therefore,
-we can easily test the null hypothesis that the Weibull model is just as
-good as the larger (in terms of number of parameters) spline model:
+we can easily test the null hypothesis that the exponential model is
+just as good as the larger (in terms of number of parameters) spline
+model:
 
@@
 #> 
-#> Model 1: DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm + offset(offset)
+#> Model 1: DeadOfPrCa ~ ScrArm + offset(offset)
 #> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
 #>   Resid. Df Resid. Dev Df Deviance Pr(>Chi)    
-#> 1     54537       5801                         
-#> 2     54535       5785  2     15.7    4e-04 ***
+#> 1     54538       6052                         
+#> 2     54535       5785  3      267   <2e-16 ***
 #> ---
@@
 \centering
-\begin{tabular}[t]{>{\bfseries}lcc}
+\begin{tabular}[t]{>{}lcc}
 \toprule
@@
 \midrule
-Exponential & 0.79 (0.67, 0.94) & 0.81 (0.68, 0.96)\\
+\textbf{Exponential} & 0.79 (0.67, 0.94) & 0.81 (0.68, 0.96)\\
 \cmidrule{1-3}
-Gompertz & 0.80 (0.67, 0.95) & 0.80 (0.67, 0.95)\\
+\textbf{Gompertz} & 0.80 (0.67, 0.95) & 0.80 (0.67, 0.95)\\
 \cmidrule{1-3}
-Weibull & 0.81 (0.68, 0.96) & 0.80 (0.65, 0.96)\\
+\textbf{Weibull} & 0.81 (0.68, 0.96) & 0.80 (0.65, 0.96)\\
 \cmidrule{1-3}
-Splines & 0.79 (0.66, 0.94) & --\\
+\textbf{Splines} & 0.79 (0.66, 0.94) & --\\
 \bottomrule
-\multicolumn{3}{l}{Cox model estimate: HR (95\% CI) = 0.80 (0.67, 0.95)}\\
+\multicolumn{3}{l}{\rule{0pt}{1em}Cox model estimate: HR (95\% CI) = 0.80 (0.67, 0.95)}\\
 \end{tabular}
@@
 In Figure \ref{fig:interaction-ERSPC} we plot the estimated death from
-prostate cancer hazard ratio and 95\textbackslash\% confidence interval
-for screening vs.~control group as a function of time using the
-\code{plot} method for objects of class \code{singleEventCB}:
+prostate cancer hazard ratio and 95\% confidence interval for screening
+vs.~control group as a function of time using the \code{plot} method for
+objects of class \code{singleEventCB}:
 
@@
 R> plot(fit_inter, type = "hr", newdata = new_data,
-R>      var = "ScrArm", xvar = "Follow.Up.Time", ci = TRUE)
++      var = "ScrArm", xvar = "Follow.Up.Time", ci = TRUE)
 \end{CodeInput}
@@
 R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-R+   data = bmtcrr
-R+ )
++   data = bmtcrr
++ )
 \end{CodeInput}
@@
 R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) +
-R+                         const(Phase) + const(Source) + const(Age),
-R+                       data = bmtcrr, cause = 1, model = "fg")
++                         const(Phase) + const(Source) + const(Age),
++                       data = bmtcrr, cause = 1, model = "fg")
 R> 
@@
 about this dataset, the reader is encouraged to look at the
-documentation in our package. A description of the variables can be
-found in Table \ref{tab:support1}, and a breakdown of each categorical
-variable appears in Table \ref{tab:support2}.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{ccccc}
-\hline
-\textbf{Name} & \textbf{Labels}                          & \textbf{Levels} & \textbf{Storage} & \textbf{NAs} \\
-\hline
-age           & Age                                      &                 & double           & 0            \\
-death         & Death at any time up to NDI date:31DEC94 &                 & double           & 0            \\
-sex           &                                          & 2               & integer          & 0            \\
-slos          & Days from Study Entry to Discharge       &                 & double           & 0            \\
-d.time        & Days of Follow-Up                        &                 & double           & 0            \\
-dzgroup       &                                          & 8               & integer          & 0            \\
-dzclass       &                                          & 4               & integer          & 0            \\
-num.co        & number of comorbidities                  &                 & double           & 0            \\
-edu           & Years of Education                       &                 & double           & 202          \\
-income        &                                          & 4               & integer          & 349          \\
-scoma         & SUPPORT Coma Score based on Glasgow D3   &                 & double           & 0            \\
-avtisst       & Average TISS, Days 3-25                  &                 & double           & 6            \\
-race          &                                          & 5               & integer          & 5            \\
-sps             & support physiology score day 3           &                   & double           & 1            \\
-aps           & APS III no coma, imp bun,uout for ph1,D3 &                 & double           & 1            \\
-hday            & Day in Hospital at Study Admit             &                 & double           & 0            \\ 
-diabetes      & Diabetes (Com 27-28, Dx 73)              &                 & double           & 0            \\
-dementia      & Dementia (Comorbidity 6)                 &                 & double           & 0            \\
-ca              & Cancer State                             & 3               & integer          & 0            \\
-meanbp        & Mean Arterial Blood Pressure Day 3       &                 & double           & 0            \\
-wblc          & White Blood Cell Count Day 3             &                 & double           & 24           \\
-hrt           & Heart Rate Day 3                         &                 & double           & 0            \\
-resp          & Respiration Rate Day 3                   &                 & double           & 0            \\
-temp          & Temperature (Celsius) Day 3              &                 & double           & 0            \\
-pafi          & PaO2/(.01*FiO2) Day 3                    &                 & double           & 253          \\
-alb           & Serum Albumin Day 3                      &                 & double           & 378          \\
-bili          & Bilirubin Day 3                          &                 & double           & 297          \\
-crea          & Serum creatinine Day 3                   &                 & double           & 3            \\
-sod           & Serum sodium Day 3                       &                 & double           & 0            \\
-ph            & Serum pH (arterial) Day 3                &                 & double           & 250          \\
-glucose       & Glucose Day 3                            &                 & double           & 470          \\
-bun           & BUN Day 3                                &                 & double           & 455          \\
-urine         & Urine Output Day 3                       &                 & double           & 517          \\
-adlp          & Activiites of Daily Living (ADL) Patient Day 3 &           & double           & 634          \\
-adlsc         & Imputed ADL Calibrated to Surrogate      &                 & double           & 0           
-\end{tabular}
-\caption{Description of each variable in the SUPPORT dataset.}
-\label{tab:support1}
-\end{table}
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{cc}
-\hline
-\textbf{Variable} & \textbf{Levels}                      \\
-\hline
-sex      & female                                        \\
-         & male                                          \\
-dzgroup  & ARF/MOSF w/Sepsis                             \\
-         & COPD                                          \\
-         & CHF                                           \\
-         & Cirrhosis                                     \\
-         & Coma                                          \\
-         & Colon Cancer                                  \\
-         & Lung Cancer                                   \\
-         & MOSF w/Malig                                  \\
-dzclass  & ARF/MOSF                                      \\
-         & COPD/CHF/Cirrhosis                            \\
-         & Coma                                          \\
-         & Cancer                                        \\
-         & \textgreater{}\$50k                           \\
-race     & white                                         \\
-         & black                                         \\
-         & asian                                         \\
-         & other                                         \\
-         & hispanic                                      \\
-ca       & Yes                                           \\
-         & No                                            \\
-         & Metastatic                                    \\
-\end{tabular}
-\caption{Description of each level within each categorical variable in the SUPPORT dataset.}
-\label{tab:support2}
-\end{table}
-
-In the original study, the authors developed two scores for predicting
-the outcome. These scores are available as part of the data: SUPPORT day
-3 physiology score (\code{sps}) and APACHE III day 3 physiology score
-(\code{aps}). Using these scores, we will explore three Cox regression
-scenarios as a baseline comparison for \pkg{casebase}: \code{sps} only,
-\code{aps} only, and all covariates excluding \code{sps} and \code{aps}.
-We will compare the three models using a time-dependent version of the
-classical Brier score that is adjusted for censoring
-\citep{graf1999ass}. The Brier score can be used to assess both
-discrimination and calibration in a single model. Moreover, it provides
-a metric that can be used to compare parametric and semi-parametric
-survival models. To account for overfitting, we will use 95\% of the
-observations to train our models, while the remaining 5\% will be used
-to compute absolute risk curves and the Brier scores.
-
-In Figure\ref{fig:brier1} (A), we can see that all Cox models have
-similar average survival curves. On the other hand, Figure
-\ref{fig:brier1} (B) shows that the full model has the lowest Brier
-score overall. For this reason, we will use this model in our comparison
-between regularized Cox regression, Kaplan-Meier estimation, and the
-approach from \pkg{casebase}.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/brierpt1-1}
-
-}
-
-\caption{\label{fig:brier1} Comparison of three Cox regression models to the unadjusted Kaplan-Meier curve.
-(A) Cumulative Incidence curves.
-For a given model, the plotted curve is the average curve of each individual in the study.
-Each follow-up time seen in the study was used.
-There are minimal differences between the curves.
-APS coincides with the SPS curve, as they are considerably overlapped.(B) Time-dependent Brier scores, where a lower score is better.
-The full model performs best overall.}\label{fig:brierpt1}
-\end{figure}
-\end{CodeChunk}
-
-For our penalized model, we opt for the natural log of time; recall that
-this corresponds to a Weibull distribution. For fitting the penalized
-hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface
-to the function \code{fitSmoothHazard}. In order to use this interface,
-we first create a matrix \code{y} containing the time and event
-variables, and a matrix \code{x} containing all remaining covariates we
-would like to include in the model space. We used lasso penalization by
-setting \code{alpha = 1}. Moreover, we use the argument
-\code{penalty.factor} to force the time variable into the final model.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
+documentation in our package.
+
+In this section, we compare our penalized case-base approach to two
+other models: Cox regression, and penalized Cox regression (using the
+\pkg{glmnet} package). In all three models, we use all covariates except
+\code{sps} and \code{aps}. These two variables correspond to scores for
+predicting the outcome that were developed as part of the original
+study.
+
+To compare the performance of our models, we first divide the data into
+a training and a test data, with 95\% of the observations used for
+training all three models. To assess both discrimination and
+calibration, we use a time-dependent version of the classical Brier
+score that is adjusted for censoring \citep{graf1999ass}. The Brier
+score can be used with both parametric and semi-parametric models. We
+use the \pkg{riskRegression} to compute these scores for all models.
+
+For our penalized case-base model, we opt for the natural log of time;
+recall that this corresponds to a Weibull distribution. For fitting the
+penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix
+interface to the function \code{fitSmoothHazard}. In order to use this
+interface, we first create a matrix \code{y} containing the time and
+event variables, and a matrix \code{x} containing all remaining
+covariates we would like to include in the model space. We used lasso
+penalization by setting \code{alpha = 1}. Moreover, we use the argument
+\code{penalty.factor} to give a regularization weight of 0 to the time
+variable, and therefore force it into the final model.
+
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Create matrices for inputs
 R> x <- model.matrix(death ~ . - d.time - aps - sps, 
-R+                   data = train)[, -c(1)] # Remove intercept
++                   data = train)[, -c(1)] # Remove intercept
 R> y <- data.matrix(subset(train, select = c(d.time, death)))
@@
 R> # Regularized logistic regression to estimate hazard
-R> cbFull <- casebase::fitSmoothHazard.fit(x, y,
-R+   family = "glmnet",
-R+   time = "d.time", event = "death",
-R+   formula_time = ~ log(d.time), alpha = 1,
-R+   ratio = 10, standardize = TRUE,
-R+   penalty.factor = c(0, rep(1, ncol(x)))
-R+ )
-\end{CodeInput}
-\end{CodeChunk}
-
-Next, we can use the object \code{cbFull} to estimate the absolute risk
-curve on the test data. We first create a matrix \code{newx} containing
-the relevant variables. Next, using \code{times}, we specify the time
-points at which the cumulative incidence will be computed. Finally,
-using the parameter \code{s = "lambda.1se"}, we specify at which value
-of the tuning parameter we want to retrieve the coefficient estimates.
-
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Estimating the absolute risk curve using the newdata parameter
-R> newx <- model.matrix(death ~ . - d.time - aps - sps, 
-R+                   data = test)[, -c(1)]
-R> times <- sort(unique(test$d.time))
-R> 
-R> abCbFull <- casebase::absoluteRisk(cbFull,
-R+   time = times,
-R+   newdata = newx,
-R+   s = "lambda.1se",
-R+   method = "numerical"
-R+ )
+R> pen_cb <- casebase::fitSmoothHazard.fit(x, y,
++   family = "glmnet",
++   time = "d.time", event = "death",
++   formula_time = ~ log(d.time), alpha = 1,
++   ratio = 10, standardize = TRUE,
++   penalty.factor = c(0, rep(1, ncol(x)))
++ )
 \end{CodeInput}
@@
 Our first comparison plot examines the covariates that were selected
-between all three hazard functions. Note that Cox regression without
-penalization selected more covariates than the other two, but only the
-intersection of coefficients with values greater than 0 are displayed in
-\ref{fig:cs3lolliPlot}. A visible tendency is for Cox regression to have
-larger estimated coefficients. This is expected, as there are no
-restrictions on the size of coefficients. The smallest coefficient
-estimates are found in penalized case-base model, aside from cancer
-state (cano).
+between all three hazard functions. In \ref{fig:cs3lolliPlot}, we show
+the coefficient estimates for covariates that we selected by both
+penalized Cox and penalized case-base. From this plot, we can clearly
+see the shrinkage effect of using penalized estimation.
 
@@
 \caption{\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models.
-If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity.
+Only covariates common among all three models were retained.
 Shrinkage occurs due to penalization.}\label{fig:coefplots}
@@
 
-These values do not inform us of model performance. To do so, we take
-our hazard functions and use them to predict the absolute risk of
-everyone in our test set, over time. The probabilities over time for
-each individual are averaged, resulting in the absolute risk curves in
-Figure \ref{fig:cs3FinalBrier} (A). There is separation between the
-curves, but it is minimal. To better demonstrate the differences in
-performance between the models, we compare Brier scores over time. In
-Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our
-penalized case-base model compared to both Cox regression models. The
-differences, however, are minimal. All three outperform the Kaplan-Meier
-estimator.
+Next, we compare the cumulative incidence estimation over the test set.
+The probabilities over time for each observation are averaged, resulting
+in the absolute risk curves in Figure \ref{fig:cs3FinalBrier} (A). We
+can see some minimal differences between the three models, with the
+Kaplan-Meier estimate giving a lower estimate than our three models. We
+also note that the apparent smoothness of the Cox and penalized Cox
+curves is due to the large number of observations in the training set,
+which is used to derive the Breslow estimate of the baseline hazard. As
+described above, we compare the performance between the models by
+computing the Brier scores over time. In Figure \ref{fig:cs3FinalBrier}
+(B), we can see that the Brier score is larger for the Kaplan-Meier
+estimate than for the other three models. On the other hand, the
+differences between these three models are minimal.
 
@@
 (A)  Probabilities of death as it changes over follow-up time.
-(B) Brier scores as they change over time, where a lower score corresponds to better performace.
-Penalized case-base estimation produced the lowest results over follow-up time.}\label{fig:riskregressionBrier}
+(B) Brier scores as they change over time, where a lower score corresponds to better performace.}\label{fig:riskregressionBrier}
 \end{figure}
 \end{CodeChunk}
+
+In brief, case-base sampling can be used in conjunction with penalized
+logistic regression to perform variable selection with survival models.
 
@@
 In the previous case studies, we only considered covariates that were
-fixed at baseline. In this next case study, we will use the Stanford
-Heart Transplant data
-\citep[\citet{crowley1977covariance}]{clark1971cardiac} to show how
-case-base sampling can also be used in the context of time-dependent
-exposure. As an example that already appeared in the literature,
-case-base sampling was used to study vaccination safety, where the
-exposure period was defined as the week following vaccination
-\citep{saarela2015case}. Hence, the main covariate of interest,
-i.e.~exposure to the vaccine, was changing over time. In this context,
-case-base sampling offers an efficient alternative to nested
+fixed at baseline. In this next case study, we use the Stanford Heart
+Transplant data \citep[\citet{crowley1977covariance}]{clark1971cardiac}
+to show how case-base sampling can also be used in the context of
+time-dependent exposure. This feature of case-base sampling has been
+explored in the literature, in the context of vaccination safety
+\citep{saarela2015case}. In this study, the exposure period was defined
+as the week following vaccination. Hence, the main covariate of
+interest, i.e.~exposure to the vaccine, was changing over time. In this
+context, case-base sampling offers an efficient alternative to nested
 case-control designs or self-matching.
@@
 
-As before, we can look at the population-time plot for a graphical
-summary of the event incidence. Here, we coloured the exposed
-person-time (i.e.~after transplant) in a darker shade of gray. As we can
-see, most events occur early during the follow-up period, and therefore
-we do not expect the hazard to be constant.
+As above, we can look at the population-time plot for a graphical
+summary of the event incidence. Here, we colour the exposed person-time
+(i.e.~after transplant) in a darker shade of gray. As we can see, most
+events occur early during the follow-up period, and therefore we do not
+expect the hazard to be constant.
 
@@
 
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> cb_data <- sampleCaseBase(jasa,
++   time = "futime",
++   event = "fustat", ratio = 100
++ )
+\end{CodeInput}
+\end{CodeChunk}
+
 Next, we will compute the number of days from acceptance into the
@@
 
+\begin{CodeChunk}
+
+\begin{CodeInput}
+R> # Define exposure variable
+R> cb_data <- mutate(cb_data,
++   txtime = time_length(accept.dt %--% tx.date,
++                        unit = "days"
++   ),
++   exposure = case_when(
++     is.na(txtime) ~ 0L, # No transplant
++     txtime > futime ~ 0L,
++     txtime <= futime ~ 1L
++   )
++ )
+\end{CodeInput}
+\end{CodeChunk}
+
 Finally, we can fit the hazard using various linear predictors.
@@
 R> fit1 <- fitSmoothHazard(fustat ~ exposure,
-R+   data = cb_data, time = "futime"
-R+ )
++   data = cb_data, time = "futime"
++ )
 R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
-R+   data = cb_data, time = "futime"
-R+ )
++   data = cb_data, time = "futime"
++ )
 R> fit3 <- fitSmoothHazard(fustat ~ exposure * futime,
-R+   data = cb_data, time = "futime"
-R+ )
++   data = cb_data, time = "futime"
++ )
 \end{CodeInput}
@@
 Note that the third model includes an interaction term between exposure
-and follow-up time. In other words, this model no longer exhibit
+and follow-up time. In other words, this model no longer exhibits
 proportional hazards. The evidence of non-proportionality of hazards in
@@
 R> plot(fit3, hazard.params = list(xvar = "futime",
-R+                                 by = "exposure",
-R+                                 alpha = 0.05,
-R+                                 ylab = "Hazard",
-R+                                 data = cb_data))
++                                 by = "exposure",
++                                 alpha = 0.05,
++                                 ylab = "Hazard",
++                                 data = cb_data))
 \end{CodeInput}
@@
 R> absrisk <- absoluteRisk(fit3,
-R+   newdata = newdata,
-R+   time = seq(0, 425, length.out = 100)
-R+ )
++   newdata = newdata,
++   time = seq(0, 425, length.out = 100)
++ )
 R> 
@@
 R> plot(absrisk,
-R+      id.names = c("No Tx", "Tx")) + 
-R+   ylab("Cumulative Incidence") +
-R+   xlab("Follow-up time (in days)") +
-R+   paper_gg_theme
++      id.names = c("No Tx", "Tx")) + 
++   ylab("Cumulative Incidence") +
++   xlab("Follow-up time (in days)") +
++   paper_gg_theme
 \end{CodeInput}
@@
 
-This report was generated on 2020-08-31 10:05:10 using the following
+This report was generated on 2020-09-15 17:46:19 using the following
 computational environment and dependencies:
@@
 #>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
-#>  [4] ggstance_0.3.4      htmlTable_1.13.3    base64enc_0.1-3    
-#>  [7] fs_1.4.1            rstudioapi_0.11     farver_2.0.3       
+#>  [4] ggstance_0.3.4      htmlTable_2.0.1     base64enc_0.1-3    
+#>  [7] fs_1.5.0            rstudioapi_0.11     farver_2.0.3       
 #> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-1      
 #> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
-#> [16] Formula_1.2-3       jsonlite_1.7.0      cluster_2.1.0      
+#> [16] Formula_1.2-3       jsonlite_1.7.1      cluster_2.1.0      
 #> [19] dbplyr_1.4.4        png_0.1-7           compiler_4.0.2     
-#> [22] httr_1.4.2          backports_1.1.8     assertthat_0.2.1   
-#> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
-#> [28] quantreg_5.55       tools_4.0.2         gtable_0.3.0       
-#> [31] Rcpp_1.0.5          cellranger_1.1.0    vctrs_0.3.2        
-#> [34] nlme_3.1-148        iterators_1.0.12    xfun_0.15          
-#> [37] rvest_0.3.6         lifecycle_0.2.0     polspline_1.1.17   
-#> [40] muhaz_1.2.6.1       MASS_7.3-51.6       zoo_1.8-8          
+#> [22] httr_1.4.2          backports_1.1.9     assertthat_0.2.1   
+#> [25] cli_2.0.2           htmltools_0.5.0     quantreg_5.67      
+#> [28] tools_4.0.2         gtable_0.3.0        Rcpp_1.0.5         
+#> [31] cellranger_1.1.0    vctrs_0.3.4         nlme_3.1-149       
+#> [34] conquer_1.0.2       iterators_1.0.12    xfun_0.17          
+#> [37] rvest_0.3.6         lifecycle_0.2.0     polspline_1.1.19   
+#> [40] muhaz_1.2.6.1       MASS_7.3-53         zoo_1.8-8          
 #> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
-#> [46] RColorBrewer_1.1-2  rticles_0.14        yaml_2.2.1         
-#> [49] gridExtra_2.3       rms_5.1-4           rpart_4.1-15       
-#> [52] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
-#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
-#> [58] rlang_0.4.7         pkgconfig_2.0.3     evaluate_0.14      
-#> [61] lattice_0.20-41     labeling_0.3        htmlwidgets_1.5.1  
-#> [64] cmprsk_2.2-10       tidyselect_1.1.0    deSolve_1.28       
-#> [67] plyr_1.8.6          R6_2.4.1            generics_0.0.2     
-#> [70] Hmisc_4.4-0         multcomp_1.4-13     DBI_1.1.0          
-#> [73] withr_2.2.0         pillar_1.4.6        haven_2.3.1        
-#> [76] foreign_0.8-80      mgcv_1.8-31         nnet_7.3-14        
-#> [79] mstate_0.2.12       modelr_0.1.8        crayon_1.3.4       
-#> [82] rmarkdown_2.3       jpeg_0.1-8.1        grid_4.0.2         
-#> [85] readxl_1.3.1        blob_1.2.1          reprex_0.3.0       
-#> [88] digest_0.6.25       webshot_0.5.2       numDeriv_2016.8-1.1
-#> [91] stats4_4.0.2        munsell_0.5.0       viridisLite_0.3.0  
-#> [94] quadprog_1.5-8
+#> [46] RColorBrewer_1.1-2  rticles_0.15        yaml_2.2.1         
+#> [49] gridExtra_2.3       rms_6.0-1           rpart_4.1-15       
+#> [52] latticeExtra_0.6-29 stringi_1.4.6       highr_0.8          
+#> [55] foreach_1.5.0       checkmate_2.0.0     lava_1.6.7         
+#> [58] shape_1.4.4         rlang_0.4.7         pkgconfig_2.0.3    
+#> [61] matrixStats_0.56.0  evaluate_0.14       lattice_0.20-41    
+#> [64] labeling_0.3        htmlwidgets_1.5.1   cmprsk_2.2-10      
+#> [67] tidyselect_1.1.0    deSolve_1.28        plyr_1.8.6         
+#> [70] R6_2.4.1            generics_0.0.2      Hmisc_4.4-1        
+#> [73] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
+#> [76] pillar_1.4.6        haven_2.3.1         foreign_0.8-80     
+#> [79] mgcv_1.8-33         nnet_7.3-14         mstate_0.2.12      
+#> [82] modelr_0.1.8        crayon_1.3.4        rmarkdown_2.3      
+#> [85] jpeg_0.1-8.1        grid_4.0.2          readxl_1.3.1       
+#> [88] blob_1.2.1          reprex_0.3.0        digest_0.6.25      
+#> [91] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_4.0.2       
+#> [94] munsell_0.5.0       viridisLite_0.3.0   quadprog_1.5-8
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-#> Local:    editMax /Users/maxturgeon/Documents/git_repos/cbpaper
-#> Head:     [0be28f4] 2020-08-25: switch to plot method for hazard
+#> Local:    master /Users/maxturgeon/Documents/git_repos/cbpaper
+#> Remote:   master @ origin (git@github.com:sahirbhatnagar/cbpaper.git)
+#> Head:     [ddc5b48] 2020-09-14: Merge pull request #29 from sahirbhatnagar/editSahir
 \end{CodeOutput}

### 2020-09-16 analysis/paper/paper.tex
add abstract and plot captions

@@
 \Abstract{
-The abstract of the article. *SRB and MT contributed equally to this
-work.
+In epidemiological studies of time-to-event data, a quantity of interest
+to the clinician and the patient is the risk of an event given a
+covariate profile. However, methods relying on time matching or risk-set
+sampling (including Cox regression) eliminate the baseline hazard from
+the likelihood expression or the estimating function. But to estimate
+risks, the baseline hazard needs to be estimated separately using a
+non-parametric method. This apparoch leads to step-wise estimates of the
+cumulative incidence that are difficult to interpret. Using case-base
+sampling, Hanley \& Miettinen explained how the parametric hazard
+functions can be estimated using logistic regression. Their approach
+naturally leads to estimates of the cumulative incidence that are
+smooth-in-time.\\
+In this paper, we present the casebase R package, which implements the
+ideas of Hanley \& Miettinen through a simple and flexible interface. We
+describe how the case-base framework can also be used in more complex
+settings: competing risks, time-varying exposure, and variable
+selection. Our package also includes an extensive array of visualization
+tools to complement the analysis of time-to-event data. We illustrate
+all these features through four different case studies.\\
+*SRB and MT contributed equally to this work.
 }
 
-\Keywords{survival analysis, absolute risk, data visualization}
-\Plainkeywords{survival analysis, absolute risk, data visualization}
+\Keywords{survival analysis, absolute risk, \proglang{R}, data
+visualization}
+\Plainkeywords{survival analysis, absolute risk, R, data visualization}
 
@@
 
-Using the framework of logistic regression for analysing time-to-event
+Using the framework of logistic regression for analyzing time-to-event
 data opens the door to an extensive array of powerful modeling tools.
@@
 \caption[Population time plots for both treatment arms in the ERSPC dataset]{Population time plots for both treatment arms in the ERSPC dataset.
-The grey area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely small rectangles (person-moments).
+The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely small rectangles (person-moments).
 The points shown in the figure are select person-moments from the case and base series.
@@
 As above, we can look at the population-time plot for a graphical
-summary of the event incidence. Here, we colour the exposed person-time
-(i.e.~after transplant) in a darker shade of gray. As we can see, most
-events occur early during the follow-up period, and therefore we do not
-expect the hazard to be constant.
-
-\begin{CodeChunk}
-
-\begin{center}\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-poptime-1} \end{center}
-
+summary of the event incidence (see Figure \ref{fig:cs4PopTime}. Here,
+we colour the exposed person-time (i.e.~after transplant) in a darker
+shade of gray. As we can see, most events occur early during the
+follow-up period, and therefore we do not expect the hazard to be
+constant.
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-poptime-1}
+
+}
+
+\caption{\label{fig:cs4PopTime} Population-time for the Stanford Transplant study. The darker shade of gray corresponds to the exposed person-time.}\label{fig:stanford-poptime}
+\end{figure}
 \end{CodeChunk}
@@
 \end{CodeInput}
-
-\begin{center}\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-hazard-1} \end{center}
-
-\end{CodeChunk}
-
-The non-proportionality seems to be more pronounced at the beginning of
-follow-up than the end. Finally, we can turn these estimates of the
+\begin{figure}
+
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-hazard-1}
+
+}
+
+\caption{\label{fig:cs4hazardPlot} Plot of the hazard function for exposed and unexposed individuals.}\label{fig:stanford-hazard}
+\end{figure}
+\end{CodeChunk}
+
+Looking at Figure \ref{fig:cs4hazardPlot}, the non-proportionality seems
+to be more pronounced at the beginning of follow-up than the end.
+Finally, in Figure \ref{fig:cs4CIF}, we turn these estimates of the
 hazard function into estimates of the cumulative incidence functions.
@@
 \end{CodeInput}
-
-\begin{center}\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-risk-1} \end{center}
-
+\begin{figure}
+
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-risk-1}
+
+}
+
+\caption{\label{fig:cs4CIF} Plot of the cumulative incidence function for exposed and unexposed individuals.}\label{fig:stanford-risk}
+\end{figure}
 \end{CodeChunk}
@@
 
-This report was generated on 2020-09-15 17:46:19 using the following
+This report was generated on 2020-09-16 10:30:17 using the following
 computational environment and dependencies:
@@
 #> [49] gridExtra_2.3       rms_6.0-1           rpart_4.1-15       
-#> [52] latticeExtra_0.6-29 stringi_1.4.6       highr_0.8          
-#> [55] foreach_1.5.0       checkmate_2.0.0     lava_1.6.7         
-#> [58] shape_1.4.4         rlang_0.4.7         pkgconfig_2.0.3    
-#> [61] matrixStats_0.56.0  evaluate_0.14       lattice_0.20-41    
-#> [64] labeling_0.3        htmlwidgets_1.5.1   cmprsk_2.2-10      
-#> [67] tidyselect_1.1.0    deSolve_1.28        plyr_1.8.6         
-#> [70] R6_2.4.1            generics_0.0.2      Hmisc_4.4-1        
-#> [73] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
-#> [76] pillar_1.4.6        haven_2.3.1         foreign_0.8-80     
-#> [79] mgcv_1.8-33         nnet_7.3-14         mstate_0.2.12      
-#> [82] modelr_0.1.8        crayon_1.3.4        rmarkdown_2.3      
-#> [85] jpeg_0.1-8.1        grid_4.0.2          readxl_1.3.1       
-#> [88] blob_1.2.1          reprex_0.3.0        digest_0.6.25      
-#> [91] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_4.0.2       
-#> [94] munsell_0.5.0       viridisLite_0.3.0   quadprog_1.5-8
+#> [52] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
+#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
+#> [58] rlang_0.4.7         pkgconfig_2.0.3     matrixStats_0.56.0 
+#> [61] evaluate_0.14       lattice_0.20-41     labeling_0.3       
+#> [64] htmlwidgets_1.5.1   cmprsk_2.2-10       tidyselect_1.1.0   
+#> [67] deSolve_1.28        plyr_1.8.6          R6_2.4.1           
+#> [70] generics_0.0.2      Hmisc_4.4-1         multcomp_1.4-13    
+#> [73] DBI_1.1.0           withr_2.2.0         pillar_1.4.6       
+#> [76] haven_2.3.1         foreign_0.8-80      mgcv_1.8-33        
+#> [79] nnet_7.3-14         mstate_0.2.12       modelr_0.1.8       
+#> [82] crayon_1.3.4        rmarkdown_2.3       jpeg_0.1-8.1       
+#> [85] grid_4.0.2          readxl_1.3.1        blob_1.2.1         
+#> [88] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
+#> [91] numDeriv_2016.8-1.1 stats4_4.0.2        munsell_0.5.0      
+#> [94] viridisLite_0.3.0   quadprog_1.5-8
 \end{CodeOutput}
@@
 #> Remote:   master @ origin (git@github.com:sahirbhatnagar/cbpaper.git)
-#> Head:     [ddc5b48] 2020-09-14: Merge pull request #29 from sahirbhatnagar/editSahir
+#> Head:     [73c0393] 2020-09-15: update CS3
 \end{CodeOutput}

### 2020-09-19 analysis/paper/paper.Rmd

@@
   short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
-abstract: >
-  The abstract of the article. *SRB and MT contributed equally to this work.
+abstract: |
+  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician and the patient is the risk of an event given a covariate profile.
+However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the likelihood expression or the estimating function.
+But to estimate risks, the baseline hazard needs to be estimated separately using a non-parametric method.
+This apparoch leads to step-wise estimates of the cumulative incidence that are difficult to interpret.
+Using case-base sampling, Hanley & Miettinen explained how the parametric hazard functions can be estimated using logistic regression.
+Their approach naturally leads to estimates of the cumulative incidence that are smooth-in-time.
+  | In this paper, we present the casebase R package, which implements the ideas of Hanley & Miettinen through a simple and flexible interface.
+We describe how the case-base framework can also be used in more complex settings: competing risks, time-varying exposure, and variable selection.
+Our package also includes an extensive array of visualization tools to complement the analysis of time-to-event data.
+We illustrate all these features through four different case studies.
+  | *SRB and MT contributed equally to this work.
 keywords:
   # at least one keyword must be supplied
-  formatted: [survival analysis, absolute risk, data visualization]
-  plain:     [survival analysis, absolute risk, data visualization]
+  formatted: [survival analysis, absolute risk, "\\proglang{R}", data visualization]
+  plain:     [survival analysis, absolute risk, R, data visualization]
 preamble: >
@@
 
-Using the framework of logistic regression for analysing time-to-event data opens the door to an extensive array of powerful modeling tools.
+Using the framework of logistic regression for analyzing time-to-event data opens the door to an extensive array of powerful modeling tools.
 Indeed, lasso and elastic-net regression can then be used to select variables that are associated with the hazard function.
@@
 ```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plots for both treatment arms in the ERSPC dataset.
-The grey area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely small rectangles (person-moments).
+The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely small rectangles (person-moments).
 The points shown in the figure are select person-moments from the case and base series.
@@
 legend("topleft", 
-       legend = c("Control (Cox)","Control (Casebase)",
-                  "Screening (Cox)", "Screening (Casebase)"), 
+       legend = c("Control (Cox)","Control (Case-base)",
+                  "Screening (Cox)", "Screening (Case-base)"), 
        col = rep(myCols, each = 2),
@@
 
-As above, we can look at the population-time plot for a graphical summary of the event incidence.
+As above, we can look at the population-time plot for a graphical summary of the event incidence (see Figure \ref{fig:cs4PopTime}.
 Here, we colour the exposed person-time (i.e.
@@
 
-```{r stanford-poptime, eval = eval_cs4}
+```{r stanford-poptime, eval = eval_cs4, fig.cap="\\label{fig:cs4PopTime} Population-time for the Stanford Transplant study. The darker shade of gray corresponds to the exposed person-time."}
 stanford_popTime <- popTime(jasa,
@@
 # plot(stanford_popTime,
-#      ribbon.params = list(fill = "grey50")) + 
+#      ribbon.params = list(fill = "gray50")) + 
 #   paper_gg_theme
@@
 plot(stanford_popTime,
-     ribbon.params = list(fill = "grey50")) + 
+     ribbon.params = list(fill = "gray50")) + 
     geom_segment(data = tx_exp,
@@
                      y = ycoord, yend = ycoord),
-                 colour = "grey30", size = 1) + 
+                 colour = "gray30", size = 1) + 
     geom_point(data = filter(stanford_popTime, event == 1),
@@
 
-```{r stanford-hazard, eval = eval_cs4, echo=TRUE, dependson="stanford_fit", message = FALSE, results="hide"}
+```{r stanford-hazard, eval = eval_cs4, echo=TRUE, dependson="stanford_fit", message = FALSE, results="hide", fig.cap="\\label{fig:cs4hazardPlot} Plot of the hazard function for exposed and unexposed individuals."}
 plot(fit3, hazard.params = list(xvar = "futime",
@@
 
-The non-proportionality seems to be more pronounced at the beginning of follow-up than the end.
-Finally, we can turn these estimates of the hazard function into estimates of the cumulative incidence functions.
-
-```{r stanford-risk, eval = eval_cs4, echo = TRUE, dependson="stanford_fit"}
+Looking at Figure \ref{fig:cs4hazardPlot}, the non-proportionality seems to be more pronounced at the beginning of follow-up than the end.
+Finally, in Figure \ref{fig:cs4CIF}, we turn these estimates of the hazard function into estimates of the cumulative incidence functions.
+
+```{r stanford-risk, eval = eval_cs4, echo = TRUE, dependson="stanford_fit", fig.cap="\\label{fig:cs4CIF} Plot of the cumulative incidence function for exposed and unexposed individuals."}
 # Compute absolute risk curves
@@
   paper_gg_theme
-```
-
-<!-- Note that we can easily adapt the code above to the situation where a patient receives a heart transplant at a point in time of interest, for example after 1 month. -->
-
-```{r stanford-1y-haz, eval = FALSE, echo = FALSE}
-# Compute hazards---
-# First, create a list of time points for both exposure status
-one_yr_haz <- data.frame(futime = seq(0, 365,
-  length.out = 100
-))
-one_yr_haz <- mutate(one_yr_haz,
-  offset = 0,
-  exposure = if_else(futime < 1, 0, 1)
-)
-one_yr_haz$hazard <- exp(predict(fit4,
-  newdata = one_yr_haz,
-  type = "link"
-))
-ggplot(one_yr_haz, aes(futime, hazard)) +
-  geom_line() +
-  paper_gg_theme +  
-  ylab("Hazard") +
-  xlab("Follow-up time (in days)") +
-  geom_vline(xintercept = 1, linetype = "dashed")
-```
-
-<!-- We can then compare the 1-year mortality risk without transplant and with transplant at 1 month. -->
-
-```{r stanford-1y-risk, eval = FALSE, echo = FALSE}
-absoluteRisk(fit4,
-  newdata = data.frame(exposure = 0),
-  time = 365
-)
-
-# Use the trapezoidal rule to estimate the cumulative hazard
-cumhaz_est <- pracma::trapz(one_yr_haz$futime, one_yr_haz$hazard)
-1 - exp(-cumhaz_est)
-```
-
-<!-- As we can see, the risk estimate at 1-year is about 30% lower if the patient receives a heart transplant at 1 month. -->
-
-<!-- We can also plot the hazard ratio: -->
-
-```{r, eval = FALSE, dependson="stanford_fit"}
-newtime <- seq(0, 425, length.out = 100)
-
-# reference category
-newdata <- data.frame(exposure = 0,
-                      futime = newtime)
-
-plot(fit3,
-     type = "hr",
-     newdata = newdata,
-     var = "exposure",
-     increment = 1,
-     xvar = "futime",
-     xlab = "Follow-up time (in days)",
-     ci = FALSE,
-     rug = TRUE)
-abline(h = 1, lty = 2)
 ```
@@
 
-```{r, echo=FALSE, eval = FALSE}
-knitr::kable(data.frame(
-  `feature` = c("model type", "time", "cumulative incidence", "non-proportional hazards", "model testing", "competing risks", "prediction"),
-  `Cox` = c("semi-parametric", "left hand side of the equation", "step function", "interaction of covariates with time", " ", "difficult", "Kaplan-Meier-based"),
-  `Case Base Sampling` = c(
-    "fully parametric (logistic/multinomial regression)", "right hand side - allows flexible modeling of time",
-    "smooth-in-time curve", "interaction of covariates with time", "make use of GLM framework (LRT, AIC, BIC)",
-    "cause-specific cumulative incidence functions (CIFs) directly obtained via multinomial regression", "ROC, AUC, risk reclassification probabilities"
-  )
-),
-col.names = c("Feature", "Cox model", "Case-base sampling"),
-format = "latex", booktabs = TRUE,
-caption = "Comparison between the Cox model and case-base sampling",
-escape = FALSE
-)
-```
-
 # Environment Details
@@
 ```
-
-<!-- # Appendix -->
-
-<!-- A description of the variables can be found in Table \ref{tab:support1}, and a breakdown of each categorical variable appears in Table \ref{tab:support2}.  -->
-
-<!-- \begin{table}[ht] -->
-<!-- \centering -->
-<!-- \begin{tabular}{ccccc} -->
-<!-- \hline -->
-<!-- \textbf{Name} & \textbf{Labels}                          & \textbf{Levels} & \textbf{Storage} & \textbf{NAs} \\ -->
-<!-- \hline -->
-<!-- age           & Age                                      &                 & double           & 0            \\ -->
-<!-- death         & Death at any time up to NDI date:31DEC94 &                 & double           & 0            \\ -->
-<!-- sex           &                                          & 2               & integer          & 0            \\ -->
-<!-- slos          & Days from Study Entry to Discharge       &                 & double           & 0            \\ -->
-<!-- d.time        & Days of Follow-Up                        &                 & double           & 0            \\ -->
-<!-- dzgroup       &                                          & 8               & integer          & 0            \\ -->
-<!-- dzclass       &                                          & 4               & integer          & 0            \\ -->
-<!-- num.co        & number of comorbidities                  &                 & double           & 0            \\ -->
-<!-- edu           & Years of Education                       &                 & double           & 202          \\ -->
-<!-- income        &                                          & 4               & integer          & 349          \\ -->
-<!-- scoma         & SUPPORT Coma Score based on Glasgow D3   &                 & double           & 0            \\ -->
-<!-- avtisst       & Average TISS, Days 3-25                  &                 & double           & 6            \\ -->
-<!-- race          &                                          & 5               & integer          & 5            \\ -->
-<!-- sps 	        & support physiology score day 3           &	               & double           &	1            \\ -->
-<!-- aps           & APS III no coma, imp bun,uout for ph1,D3 &	               & double	          & 1            \\ -->
-<!-- hday	        & Day in Hospital at Study Admit	         &                 & double           & 0            \\	 -->
-<!-- diabetes      &	Diabetes (Com 27-28, Dx 73)	             &                 & double	          & 0            \\ -->
-<!-- dementia      & Dementia (Comorbidity 6)                 &	               & double	          & 0            \\ -->
-<!-- ca		        & Cancer State                             & 3               & integer	        & 0            \\ -->
-<!-- meanbp        & Mean Arterial Blood Pressure Day 3       &                 & double           & 0            \\ -->
-<!-- wblc          & White Blood Cell Count Day 3             &                 & double           & 24           \\ -->
-<!-- hrt           & Heart Rate Day 3                         &                 & double           & 0            \\ -->
-<!-- resp          & Respiration Rate Day 3                   &                 & double           & 0            \\ -->
-<!-- temp          & Temperature (Celsius) Day 3              &                 & double           & 0            \\ -->
-<!-- pafi          & PaO2/(.01*FiO2) Day 3                    &                 & double           & 253          \\ -->
-<!-- alb           & Serum Albumin Day 3                      &                 & double           & 378          \\ -->
-<!-- bili          & Bilirubin Day 3                          &                 & double           & 297          \\ -->
-<!-- crea          & Serum creatinine Day 3                   &                 & double           & 3            \\ -->
-<!-- sod           & Serum sodium Day 3                       &                 & double           & 0            \\ -->
-<!-- ph            & Serum pH (arterial) Day 3                &                 & double           & 250          \\ -->
-<!-- glucose       & Glucose Day 3                            &                 & double           & 470          \\ -->
-<!-- bun           & BUN Day 3                                &                 & double           & 455          \\ -->
-<!-- urine         & Urine Output Day 3                       &                 & double           & 517          \\ -->
-<!-- adlp          & Activiites of Daily Living (ADL) Patient Day 3 &           & double           & 634          \\ -->
-<!-- adlsc         & Imputed ADL Calibrated to Surrogate      &                 & double           & 0            -->
-<!-- \end{tabular} -->
-<!-- \caption{Description of each variable in the SUPPORT dataset.} -->
-<!-- \label{tab:support1} -->
-<!-- \end{table} -->
-
-<!-- \begin{table}[ht] -->
-<!-- \centering -->
-<!-- \begin{tabular}{cc} -->
-<!-- \hline -->
-<!-- \textbf{Variable} & \textbf{Levels}                      \\ -->
-<!-- \hline -->
-<!-- sex      & female                                        \\ -->
-<!--          & male                                          \\ -->
-<!-- dzgroup  & ARF/MOSF w/Sepsis                             \\ -->
-<!--          & COPD                                          \\ -->
-<!--          & CHF                                           \\ -->
-<!--          & Cirrhosis                                     \\ -->
-<!--          & Coma                                          \\ -->
-<!--          & Colon Cancer                                  \\ -->
-<!--          & Lung Cancer                                   \\ -->
-<!--          & MOSF w/Malig                                  \\ -->
-<!-- dzclass  & ARF/MOSF                                      \\ -->
-<!--          & COPD/CHF/Cirrhosis                            \\ -->
-<!--          & Coma                                          \\ -->
-<!--          & Cancer                                        \\ -->
-<!--          & \textgreater{}\$50k                           \\ -->
-<!-- race     & white                                         \\ -->
-<!--          & black                                         \\ -->
-<!--          & asian                                         \\ -->
-<!--          & other                                         \\ -->
-<!--          & hispanic                                      \\ -->
-<!-- ca       & Yes                                           \\ -->
-<!--          & No                                            \\ -->
-<!--          & Metastatic                                    \\ -->
-<!-- \end{tabular} -->
-<!-- \caption{Description of each level within each categorical variable in the SUPPORT dataset.} -->
-<!-- \label{tab:support2} -->
-<!-- \end{table} -->
-
-<!-- In Figure \ref{fig:brier1} (A), we can see that all Cox models have similar average survival curves.
-On the other hand, Figure \ref{fig:brier1} (B) shows that the full model has the lowest Brier score overall.
-For this reason, we will use this model in our comparison between regularized Cox regression, Kaplan-Meier estimation, and the approach from \pkg{casebase}.
--->
-
-<!-- ```{r baselinePlotsPrep, eval=eval_cs3, echo = FALSE} -->
-<!-- # 1. Cox with only sps -->
-<!-- coxSPS <- survival::coxph(Surv(time = d.time, event = death) ~ sps,  -->
-<!--                           data = train, x = TRUE) -->
-<!-- abcoxSPS <- survival::survfit(coxSPS, newdata = test) -->
-<!-- # 2. Cox with only aps -->
-<!-- coxAPS <- survival::coxph(Surv(time = d.time, event = death) ~ aps,  -->
-<!--                           data = train, x = TRUE) -->
-<!-- abcoxAPS <- survival::survfit(coxAPS, newdata = test) -->
-<!-- # 3. Cox with everything but sps and aps -->
-<!-- coxFull <- survival::coxph(Surv(time = d.time, event = death) ~ . - aps - sps,  -->
-<!--                            data = train, x = TRUE) -->
-<!-- abcoxFull <- survival::survfit(coxFull, newdata = test) -->
-<!-- # 4. Kaplan-Meier -->
-<!-- KM <- survival::coxph(Surv(time = d.time, event = death) ~ 1, data = test,x=TRUE) -->
-<!-- abKM <- survival::survfit(KM) -->
-<!-- abKMcompare <- data.frame( -->
-<!--   Time = abKM$time, -->
-<!--   KM = 1 - abKM$surv -->
-<!-- ) -->
-
-<!-- # Combine all CI estimates from Cox models -->
-<!-- baseCoxComparisonData <- data.frame( -->
-<!--   Time = abcoxSPS$time, -->
-<!--   SPS = 1 - rowMeans(abcoxSPS$surv), -->
-<!--   APS = 1 - rowMeans(abcoxAPS$surv), -->
-<!--   Full = 1 - rowMeans(abcoxFull$surv) -->
-<!-- ) -->
-<!-- ``` -->
-
-<!-- ```{r baselines, echo=FALSE, eval=eval_cs3} -->
-<!-- cox_ci <- ggplot(baseCoxComparisonData, aes(Time)) + -->
-<!--   geom_step(aes(y = APS, colour = "APS")) + -->
-<!--   geom_step(aes(y = Full, col = "Full")) + -->
-<!--   geom_step(aes(y = SPS, colour = "SPS")) + -->
-<!--   geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) + -->
-<!--   labs(y = "Probability of Death", x = "Follow-up time (Years)", color = "Models") + -->
-<!--   scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) +  -->
-<!--   paper_gg_theme +  -->
-<!--   scale_x_continuous(labels = scaleFUNx, breaks = round(seq(0, 5, by = 1))) +  -->
-<!--   scale_y_continuous(labels = scaleFUNy, n.breaks = 8) -->
-<!-- ``` -->
-
-<!-- ```{r brierpt1, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:brier1} Comparison of three Cox regression models to the unadjusted Kaplan-Meier curve.
-(A) Cumulative Incidence curves.
-For a given model, the plotted curve is the average curve of each individual in the study.
-Each follow-up time seen in the study was used.
-There are minimal differences between the curves.
-APS coincides with the SPS curve, as they are considerably overlapped.(B) Time-dependent Brier scores, where a lower score is better.
-The full model performs best overall."} -->
-<!-- # Extract time points -->
-<!-- times <- sort(unique(test$d.time)) -->
-<!-- # Compute Brier scores -->
-<!-- brierAPS <- Score(list("PredictionModel" = coxAPS), data = test,  -->
-<!--                   formula = Hist(d.time, death != 0) ~ 1, summary = "ipa",  -->
-<!--                   se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times) -->
-<!-- brierSPS <- Score(list("PredictionModel" = coxSPS), data = test,  -->
-<!--                   formula = Hist(d.time, death != 0) ~ 1, summary = "ipa",  -->
-<!--                   se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times) -->
-
-<!-- # First fix NA coefficients, then compute Brier score -->
-<!-- coxFull$coefficients[which(is.na(coxFull$coefficients))] <- 0 -->
-<!-- brierFull <- Score(list("PredictionModel" = coxFull), data = test,  -->
-<!--                    formula = Surv(d.time, death != 0) ~ 1, summary = "ipa",  -->
-<!--                    se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times) -->
-<!-- brierNull <- Score(list("PredictionModel" = KM), data = test,  -->
-<!--                    formula = Surv(d.time, death != 0) ~ 1, summary = "ipa",  -->
-<!--                    se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times) -->
-
-<!-- # Combine all results for plotting -->
-<!-- brierPlotter <- rbind( -->
-<!--   data.frame(brierScore = brierAPS$Brier$score$Brier[brierAPS$Brier$score$model == "PredictionModel"],  -->
-<!--              times = times, model = "APS"), -->
-<!--   data.frame(brierScore = brierFull$Brier$score$Brier[brierFull$Brier$score$model == "PredictionModel"],  -->
-<!--              times = times, model = "Full"), -->
-<!--   data.frame(brierScore = brierSPS$Brier$score$Brier[brierSPS$Brier$score$model == "PredictionModel"],  -->
-<!--              times = times, model = "SPS"), -->
-<!--     data.frame(brierScore = brierNull$Brier$score$Brier[brierNull$Brier$score$model == "PredictionModel"],  -->
-<!--              times = times, model = "K-M") -->
-<!-- ) -->
-
-<!-- cox_brier <- ggplot(data = brierPlotter, mapping = aes(x = times, y = brierScore, col = model)) + -->
-<!--   geom_line() + -->
-<!--   xlab("Follow-up time (Years)") + -->
-<!--   ylab("Brier score") + -->
-<!--   labs(color = "Models") + -->
-<!--   # ggtitle("Brier scores over time") + -->
-<!--   scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) +  -->
-<!--   paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1)))+scale_y_continuous(labels=scaleFUNy,n.breaks=8) -->
-
-<!-- # Combine both plots using cowplot -->
-<!-- plot_row <- cowplot::plot_grid(cox_ci + theme(legend.position = 'none'), -->
-<!--                                cox_brier + theme(legend.position = 'none'), -->
-<!--                                labels = c("A", "B")) -->
-<!-- legend <- cowplot::get_legend( -->
-<!--   # create some space to the left of the legend -->
-<!--   cox_ci + theme(legend.box.margin = margin(0, 0, 0, 12)) -->
-<!-- ) -->
-
-<!-- cowplot::plot_grid(plot_row, legend, rel_heights = c(3, .4), nrow = 2) -->
-<!-- ``` -->

### 2020-09-20 analysis/paper/paper.Rmd
updated abstract, CS3 figure legends

@@
   | In epidemiological studies of time-to-event data, a quantity of interest to the clinician and the patient is the risk of an event given a covariate profile.
-However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the likelihood expression or the estimating function.
-But to estimate risks, the baseline hazard needs to be estimated separately using a non-parametric method.
-This apparoch leads to step-wise estimates of the cumulative incidence that are difficult to interpret.
-Using case-base sampling, Hanley & Miettinen explained how the parametric hazard functions can be estimated using logistic regression.
+However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the likelihood expression or the estimating function and therefore needs to be estimated separately using a non-parametric approach.
+This leads to step-wise estimates of the cumulative incidence that are difficult to interpret.
+Using case-base sampling, Hanley & Miettinen (2009) explained how the parametric hazard functions can be estimated using logistic regression.
 Their approach naturally leads to estimates of the cumulative incidence that are smooth-in-time.
-  | In this paper, we present the casebase R package, which implements the ideas of Hanley & Miettinen through a simple and flexible interface.
+  | In this paper, we present the **casebase** R package, a comprehensive and flexible toolkit for parametric survival analysis.
 We describe how the case-base framework can also be used in more complex settings: competing risks, time-varying exposure, and variable selection.
@@
 
-```{r coefplots, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models.
-Only covariates common among all three models were retained.
-Shrinkage occurs due to penalization."}
+```{r coefplots, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} Coefficient estimates from the Cox model (Cox), penalized Cox model using the \\pkg{glmnet} package (Pen.
+Cox), and our approach using penalized case-base sampling (Pen.
+CB).
+Only the covariates that were selected by both penalized approaches are shown.
+The shrinkage of the coefficient estimates for Pen.
+Cox and Pen.
+CB occurs due to the $\\ell_1$ penalty."}
 library(dotwhisker)
@@
 
-```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} Comparison of Cox regression, penalized Cox regression, penalized case-base estimation, and Kaplan-Meier.
-(A)  Probabilities of death as it changes over follow-up time.
-(B) Brier scores as they change over time, where a lower score corresponds to better performace.", warning = FALSE}
+```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} Comparison of Cox regression (Cox), penalized Cox regression (Pen.
+Cox), penalized case-base sampling estimation (Pen.
+CB), and Kaplan-Meier (K-M).
+(A)  Probability of death as a function of follow-up time.
+(B) Brier score as a function of follow-up time, where a lower score corresponds to better performace.", warning = FALSE}
 # Combine scores

### 2020-09-20 analysis/paper/paper.tex
updated abstract, CS3 figure legends

@@
 sampling (including Cox regression) eliminate the baseline hazard from
-the likelihood expression or the estimating function. But to estimate
-risks, the baseline hazard needs to be estimated separately using a
-non-parametric method. This apparoch leads to step-wise estimates of the
-cumulative incidence that are difficult to interpret. Using case-base
-sampling, Hanley \& Miettinen explained how the parametric hazard
-functions can be estimated using logistic regression. Their approach
-naturally leads to estimates of the cumulative incidence that are
-smooth-in-time.\\
-In this paper, we present the casebase R package, which implements the
-ideas of Hanley \& Miettinen through a simple and flexible interface. We
+the likelihood expression or the estimating function and therefore needs
+to be estimated separately using a non-parametric approach. This leads
+to step-wise estimates of the cumulative incidence that are difficult to
+interpret. Using case-base sampling, Hanley \& Miettinen (2009)
+explained how the parametric hazard functions can be estimated using
+logistic regression. Their approach naturally leads to estimates of the
+cumulative incidence that are smooth-in-time.\\
+In this paper, we present the \textbf{casebase} R package, a
+comprehensive and flexible toolkit for parametric survival analysis. We
 describe how the case-base framework can also be used in more complex
@@
 
-\Keywords{survival analysis, absolute risk, \proglang{R}, data
-visualization}
+\Keywords{survival analysis, absolute risk, \proglang{R}, data visualization}
 \Plainkeywords{survival analysis, absolute risk, R, data visualization}
@@
 R> pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
-+                                event = "DeadOfPrCa", exposure = "ScrArm")
+R+                                event = "DeadOfPrCa", exposure = "ScrArm")
 R> inherits(pt_object, "popTime")
@@
 R> fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
-+              gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
-+              weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
-+              splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
+R+              gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
+R+              weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
+R+              splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
 R> 
 R> fits <- lapply(fmla, function(i) {
-+   fitSmoothHazard(i, data = ERSPC, ratio = 100)
-+ })
+R+   fitSmoothHazard(i, data = ERSPC, ratio = 100)
+R+ })
 \end{CodeInput}
@@
 R> risks <- lapply(fits, function(i) {
-+   absoluteRisk(object = i, time = new_time, newdata = new_data, method = "mont")
-+ })
+R>   absoluteRisk(object = i, time = new_time, newdata = new_data, method = "mont")
+R> })
 \end{CodeInput}
@@
 \centering
-\begin{tabular}[t]{>{}lcc}
+\begin{tabular}[t]{>{\bfseries}lcc}
 \toprule
@@
 \midrule
-\textbf{Exponential} & 0.79 (0.67, 0.94) & 0.81 (0.68, 0.96)\\
+Exponential & 0.79 (0.67, 0.94) & 0.81 (0.68, 0.96)\\
 \cmidrule{1-3}
-\textbf{Gompertz} & 0.80 (0.67, 0.95) & 0.80 (0.67, 0.95)\\
+Gompertz & 0.80 (0.67, 0.95) & 0.80 (0.67, 0.95)\\
 \cmidrule{1-3}
-\textbf{Weibull} & 0.81 (0.68, 0.96) & 0.80 (0.65, 0.96)\\
+Weibull & 0.81 (0.68, 0.96) & 0.80 (0.65, 0.96)\\
 \cmidrule{1-3}
-\textbf{Splines} & 0.79 (0.66, 0.94) & --\\
+Splines & 0.79 (0.66, 0.94) & --\\
 \bottomrule
-\multicolumn{3}{l}{\rule{0pt}{1em}Cox model estimate: HR (95\% CI) = 0.80 (0.67, 0.95)}\\
+\multicolumn{3}{l}{Cox model estimate: HR (95\% CI) = 0.80 (0.67, 0.95)}\\
 \end{tabular}
@@
 R> plot(fit_inter, type = "hr", newdata = new_data,
-+      var = "ScrArm", xvar = "Follow.Up.Time", ci = TRUE)
+R>      var = "ScrArm", xvar = "Follow.Up.Time", ci = TRUE)
 \end{CodeInput}
@@
 R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-+   data = bmtcrr
-+ )
+R+   data = bmtcrr
+R+ )
 \end{CodeInput}
@@
 R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) +
-+                         const(Phase) + const(Source) + const(Age),
-+                       data = bmtcrr, cause = 1, model = "fg")
+R+                         const(Phase) + const(Source) + const(Age),
+R+                       data = bmtcrr, cause = 1, model = "fg")
 R> 
@@
 
-\caption{\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models.
-Only covariates common among all three models were retained.
-Shrinkage occurs due to penalization.}\label{fig:coefplots}
+\caption{\label{fig:cs3lolliPlot} Coefficient estimates from the Cox model (Cox), penalized Cox model using the \pkg{glmnet} package (Pen.
+Cox), and our approach using penalized case-base sampling (Pen.
+CB).
+Only the covariates that were selected by both penalized approaches are shown.
+The shrinkage of the coefficient estimates for Pen.
+Cox and Pen.
+CB occurs due to the $\ell_1$ penalty.}\label{fig:coefplots}
 \end{figure}
@@
 
-This report was generated on 2020-09-16 10:30:17 using the following
+This report was generated on 2020-09-19 21:55:49 using the following
 computational environment and dependencies:
@@
 #> R version 4.0.2 (2020-06-22)
-#> Platform: x86_64-apple-darwin17.0 (64-bit)
-#> Running under: macOS Catalina 10.15.6
+#> Platform: x86_64-pc-linux-gnu (64-bit)
+#> Running under: Pop!_OS 20.04 LTS
 #> 
 #> Matrix products: default
-#> BLAS:   /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRblas.dylib
-#> LAPACK: /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRlapack.dylib
+#> BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3
+#> LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/liblapack.so.3
 #> 
 #> locale:
-#> [1] en_CA.UTF-8/en_CA.UTF-8/en_CA.UTF-8/C/en_CA.UTF-8/en_CA.UTF-8
+#>  [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
+#>  [3] LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
+#>  [5] LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
+#>  [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                 
+#>  [9] LC_ADDRESS=C               LC_TELEPHONE=C            
+#> [11] LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       
 #> 
@@
 #> [27] cowplot_1.1.0             colorspace_1.4-1         
-#> [29] casebase_0.9.0.9000      
+#> [29] casebase_0.9.0           
 #> 
@@
 #>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
-#>  [4] ggstance_0.3.4      htmlTable_2.0.1     base64enc_0.1-3    
+#>  [4] ggstance_0.3.4      htmlTable_2.1.0     base64enc_0.1-3    
 #>  [7] fs_1.5.0            rstudioapi_0.11     farver_2.0.3       
@@
 #> [49] gridExtra_2.3       rms_6.0-1           rpart_4.1-15       
-#> [52] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
-#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
+#> [52] latticeExtra_0.6-29 stringi_1.5.3       foreach_1.5.0      
+#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.5        
 #> [58] rlang_0.4.7         pkgconfig_2.0.3     matrixStats_0.56.0 
@@
 #> [73] DBI_1.1.0           withr_2.2.0         pillar_1.4.6       
-#> [76] haven_2.3.1         foreign_0.8-80      mgcv_1.8-33        
+#> [76] haven_2.3.1         foreign_0.8-79      mgcv_1.8-33        
 #> [79] nnet_7.3-14         mstate_0.2.12       modelr_0.1.8       
@@
 \begin{CodeOutput}
-#> Local:    master /Users/maxturgeon/Documents/git_repos/cbpaper
-#> Remote:   master @ origin (git@github.com:sahirbhatnagar/cbpaper.git)
-#> Head:     [73c0393] 2020-09-15: update CS3
+#> Local:    editSahir /home/sahir/git_repositories/cbpaper
+#> Remote:   editSahir @ origin (https://github.com/sahirbhatnagar/cbpaper.git)
+#> Head:     [d691836] 2020-09-19: Merge remote-tracking branch 'origin/master' into editSahir
 \end{CodeOutput}

### 2020-09-20 analysis/paper/paper.Rmd
added paragraph on variable selection in theoretical details

@@
 To recover the whole survival curve---or the cumulative incidence function (CIF)---we then need to separately estimate the baseline hazard [@breslow1972discussion].
-This in turn leads to stepwise estimates of the survival function that can be hard to interpret.
-
-From the perspective of clinicians and their patients, the most relevant quantity is often the 5- or 10-year risk of having a certain event given the patient's particular circumstances, and not the hazard ratio between a treatment and control group.
-Therefore, to make sound clinical decisions, it is important to accurately estimate the *full* hazard function, which can then be used to estimate the cumulative incidence function (CIF).
+This in turn leads to stepwise estimates of the survival function that can be difficult to interpret.
+
+From the perspective of clinicians and their patients, the most relevant quantity is often the 5- or 10-year risk of experiencing a certain event given the patient's particular circumstances, and not the hazard ratio between a treatment and control group.
+Therefore, to make sound clinical decisions, it is important to accurately estimate the *full* hazard function, which can subsequently be used to estimate the cumulative incidence function (CIF).
 Using a parametric estimator of the hazard function leads to a smooth function of time; as a consequence, the CIF and the survival function estimates also vary smoothly over time.
@@
 Their main idea is simple: comparing person-moments when the event of interest occurred with moments when patients were at risk.
-Their approach provides users familiar with generalized linear models with a natural way of fitting parametric survival models.
+Their approach provides users familiar with generalized linear models a natural way of fitting parametric survival models.
 Moreover, their framework is very flexible: general functions of time can be estimated (e.g.
@@
 Using the framework of logistic regression for analyzing time-to-event data opens the door to an extensive array of powerful modeling tools.
-Indeed, lasso and elastic-net regression can then be used to select variables that are associated with the hazard function.
+Indeed, lasso and elastic-net regression can be used to select variables that are associated with the hazard function.
 Logistic regression can also be replaced by multinomial regression to analyse data with multiple competing events.
 
-In this article, we present an \proglang{R} package that combines the ideas of Hanley & Miettinen into a simple interface.
+In this article, we present a comprehensive and flexible \proglang{R} package for parametric survival analysis that combines the ideas of Hanley & Miettinen [-@hanley2009fitting] into a simple interface.
 The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to compute a patient's risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile.
@@
 Therefore, we can define an estimating function for $\theta$ as follows:
-$$L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.$$
+\begin{equation}
+L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}. \label{eq:lik-function}
+\end{equation}
 When a logarithmic link function is used for modeling the hazard function, the above expression is of a logistic regression form with an offset term $\log(1/\rho_i(t))$.
@@
 
+To perform variable selection on the regression parameters $\theta \in \mathbb{R}^p$ of the hazard function, we can add a penalty to the likelihood and solve the following problem:
+\begin{equation}
+\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j P(\theta_j;\lambda,\alpha)
+\end{equation}
+where $\ell\left(\theta\right)$ is the log of the likelihood function given in \eqref{eq:lik-function}, $P(\theta_j;\lambda,\alpha)$ is a penalty term controlled by the non-negative regularization parameters $\lambda$ and $\alpha$, and $w_j$ is the penalty factor for the $j$th covariate.
+These penalty factors serve as a way of allowing parameters to be penalized differently.
+For example, we could set the penalty factor for time to be 0 to ensure it is always included in the selected model.
+
 # Existing packages
@@
 Survival analysis is an important branch of applied statistics and epidemiology.
-Accordingly, there is already a vast ecosystem of \code{R} packages implementing different methodologies.
+Accordingly, there is already a vast ecosystem of \proglang{R} packages implementing different methodologies.
 In this section, we describe how the functionalities of \pkg{casebase} compare to these packages.
@@
 
-# Environment Details
+# Acknowledgments {-}
+
+We would like to thank Yi Yang for helpful discussions on penalized regression models.
+\mbox{Bhatnagar} gratefully acknowledges funding via Discovery Grants from the Natural Sciences and Engineering Research Council of Canada (NSERC).
+
+# Environment Details {-}
 

### 2020-09-20 analysis/paper/paper.tex
added paragraph on variable selection in theoretical details

@@
 \citep{breslow1972discussion}. This in turn leads to stepwise estimates
-of the survival function that can be hard to interpret.
+of the survival function that can be difficult to interpret.
 
 From the perspective of clinicians and their patients, the most relevant
-quantity is often the 5- or 10-year risk of having a certain event given
-the patient's particular circumstances, and not the hazard ratio between
-a treatment and control group. Therefore, to make sound clinical
+quantity is often the 5- or 10-year risk of experiencing a certain event
+given the patient's particular circumstances, and not the hazard ratio
+between a treatment and control group. Therefore, to make sound clinical
 decisions, it is important to accurately estimate the \emph{full} hazard
-function, which can then be used to estimate the cumulative incidence
-function (CIF). Using a parametric estimator of the hazard function
-leads to a smooth function of time; as a consequence, the CIF and the
-survival function estimates also vary smoothly over time.
+function, which can subsequently be used to estimate the cumulative
+incidence function (CIF). Using a parametric estimator of the hazard
+function leads to a smooth function of time; as a consequence, the CIF
+and the survival function estimates also vary smoothly over time.
 
@@
 interest occurred with moments when patients were at risk. Their
-approach provides users familiar with generalized linear models with a
+approach provides users familiar with generalized linear models a
 natural way of fitting parametric survival models. Moreover, their
@@
 data opens the door to an extensive array of powerful modeling tools.
-Indeed, lasso and elastic-net regression can then be used to select
-variables that are associated with the hazard function. Logistic
-regression can also be replaced by multinomial regression to analyse
-data with multiple competing events.
-
-In this article, we present an \proglang{R} package that combines the
-ideas of Hanley \& Miettinen into a simple interface. The purpose of the
-\pkg{casebase} package is to provide practitioners with an easy-to-use
-software tool to compute a patient's risk (or cumulative incidence) of
-an event, conditional on a particular patient's covariate profile. Our
-package retains the flexibility of case-base sampling and the familiar
-interface of the \code{glm} function. It also provides tools for
-variable selection and competing-risk analysis. In addition, we provide
-extensive visualization tools.
+Indeed, lasso and elastic-net regression can be used to select variables
+that are associated with the hazard function. Logistic regression can
+also be replaced by multinomial regression to analyse data with multiple
+competing events.
+
+In this article, we present a comprehensive and flexible \proglang{R}
+package for parametric survival analysis that combines the ideas of
+Hanley \& Miettinen \citeyearpar{hanley2009fitting} into a simple
+interface. The purpose of the \pkg{casebase} package is to provide
+practitioners with an easy-to-use software tool to compute a patient's
+risk (or cumulative incidence) of an event, conditional on a particular
+patient's covariate profile. Our package retains the flexibility of
+case-base sampling and the familiar interface of the \code{glm}
+function. It also provides tools for variable selection and
+competing-risk analysis. In addition, we provide extensive visualization
+tools.
 
@@
 Therefore, we can define an estimating function for \(\theta\) as
-follows:
-\[L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.\]
-When a logarithmic link function is used for modeling the hazard
-function, the above expression is of a logistic regression form with an
-offset term \(\log(1/\rho_i(t))\). Note that the sampling units selected
-in the case-base sampling mechanism are person-moments, rather than
-individuals, and the parameters to be estimated are hazards or hazard
-ratios rather than odds or odds ratios. Generally, an individual can
-contribute more than one person-moment, and thus the terms in the
+follows: \begin{equation}
+L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}. \label{eq:lik-function}
+\end{equation} When a logarithmic link function is used for modeling the
+hazard function, the above expression is of a logistic regression form
+with an offset term \(\log(1/\rho_i(t))\). Note that the sampling units
+selected in the case-base sampling mechanism are person-moments, rather
+than individuals, and the parameters to be estimated are hazards or
+hazard ratios rather than odds or odds ratios. Generally, an individual
+can contribute more than one person-moment, and thus the terms in the
 product integral are not independent. Nonetheless, Saarela
@@
 
+To perform variable selection on the regression parameters
+\(\theta \in \mathbb{R}^p\) of the hazard function, we can add a penalty
+to the likelihood and solve the following problem: \begin{equation}
+\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j P(\theta_j;\lambda,\alpha)
+\end{equation} where \(\ell\left(\theta\right)\) is the log of the
+likelihood function given in \eqref{eq:lik-function},
+\(P(\theta_j;\lambda,\alpha)\) is a penalty term controlled controlled
+by the non-negative regularization parameters \(\lambda\) and
+\(\alpha\), and \(w_j\) is the penalty factor for the \(j\)th covariate.
+These penalty factors serve as a way of allowing parameters to be
+penalized differently. For example, we could set the penalty factor for
+time to be 0 to ensure it is always included in the selected model.
+
 \hypertarget{existing-packages}{%
@@
 Survival analysis is an important branch of applied statistics and
-epidemiology. Accordingly, there is already a vast ecosystem of \code{R}
-packages implementing different methodologies. In this section, we
-describe how the functionalities of \pkg{casebase} compare to these
-packages.
+epidemiology. Accordingly, there is already a vast ecosystem of
+\proglang{R} packages implementing different methodologies. In this
+section, we describe how the functionalities of \pkg{casebase} compare
+to these packages.
 
@@
 
-\caption{\label{fig:cs3FinalBrier} Comparison of Cox regression, penalized Cox regression, penalized case-base estimation, and Kaplan-Meier.
-(A)  Probabilities of death as it changes over follow-up time.
-(B) Brier scores as they change over time, where a lower score corresponds to better performace.}\label{fig:riskregressionBrier}
+\caption{\label{fig:cs3FinalBrier} Comparison of Cox regression (Cox), penalized Cox regression (Pen.
+Cox), penalized case-base sampling estimation (Pen.
+CB), and Kaplan-Meier (K-M).
+(A)  Probability of death as a function of follow-up time.
+(B) Brier score as a function of follow-up time, where a lower score corresponds to better performace.}\label{fig:riskregressionBrier}
 \end{figure}
@@
 
+\hypertarget{acknowledgments}{%
+\section*{Acknowledgments}\label{acknowledgments}}
+\addcontentsline{toc}{section}{Acknowledgments}
+
+We would like to thank Yi Yang for helpful discussions on penalized
+regression models. \mbox{Bhatnagar} gratefully acknowledges funding via
+Discovery Grants from the Natural Sciences and Engineering Research
+Council of Canada (NSERC).
+
 \hypertarget{environment-details}{%
-\section{Environment Details}\label{environment-details}}
-
-This report was generated on 2020-09-19 21:55:49 using the following
+\section*{Environment Details}\label{environment-details}}
+\addcontentsline{toc}{section}{Environment Details}
+
+This report was generated on 2020-09-20 12:00:27 using the following
 computational environment and dependencies:
@@
 #> other attached packages:
-#>  [1] broom_0.7.0               dotwhisker_0.5.0         
-#>  [3] timereg_1.9.7             glue_1.4.2               
-#>  [5] scales_1.1.1              forcats_0.5.0            
-#>  [7] stringr_1.4.0             dplyr_1.0.2              
-#>  [9] purrr_0.3.4               readr_1.3.1              
-#> [11] tidyr_1.1.2               ggplot2_3.3.2            
-#> [13] tidyverse_1.3.0           lubridate_1.7.9          
-#> [15] visreg_2.7.0              tibble_3.0.3             
-#> [17] riskRegression_2020.02.05 prodlim_2019.11.13       
-#> [19] flexsurv_1.1.1            survival_3.2-3           
-#> [21] pracma_2.2.9              magrittr_1.5             
-#> [23] kableExtra_1.2.1          glmnet_4.0-2             
-#> [25] Matrix_1.2-18             data.table_1.13.0        
-#> [27] cowplot_1.1.0             colorspace_1.4-1         
-#> [29] casebase_0.9.0           
+#>  [1] scales_1.1.1              forcats_0.5.0            
+#>  [3] stringr_1.4.0             dplyr_1.0.2              
+#>  [5] purrr_0.3.4               readr_1.3.1              
+#>  [7] tidyr_1.1.2               ggplot2_3.3.2            
+#>  [9] tidyverse_1.3.0           lubridate_1.7.9          
+#> [11] visreg_2.7.0              tibble_3.0.3             
+#> [13] riskRegression_2020.02.05 prodlim_2019.11.13       
+#> [15] flexsurv_1.1.1            survival_3.2-3           
+#> [17] pracma_2.2.9              magrittr_1.5             
+#> [19] kableExtra_1.2.1          glmnet_4.0-2             
+#> [21] Matrix_1.2-18             data.table_1.13.0        
+#> [23] cowplot_1.1.0             colorspace_1.4-1         
+#> [25] casebase_0.9.0           
 #> 
@@
 #>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
-#>  [4] ggstance_0.3.4      htmlTable_2.1.0     base64enc_0.1-3    
-#>  [7] fs_1.5.0            rstudioapi_0.11     farver_2.0.3       
-#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-1      
-#> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
-#> [16] Formula_1.2-3       jsonlite_1.7.1      cluster_2.1.0      
-#> [19] dbplyr_1.4.4        png_0.1-7           compiler_4.0.2     
-#> [22] httr_1.4.2          backports_1.1.9     assertthat_0.2.1   
-#> [25] cli_2.0.2           htmltools_0.5.0     quantreg_5.67      
-#> [28] tools_4.0.2         gtable_0.3.0        Rcpp_1.0.5         
+#>  [4] htmlTable_2.1.0     base64enc_0.1-3     fs_1.5.0           
+#>  [7] rstudioapi_0.11     MatrixModels_0.4-1  fansi_0.4.1        
+#> [10] mvtnorm_1.1-1       xml2_1.3.2          codetools_0.2-16   
+#> [13] knitr_1.29          Formula_1.2-3       jsonlite_1.7.1     
+#> [16] broom_0.7.0         cluster_2.1.0       dbplyr_1.4.4       
+#> [19] png_0.1-7           compiler_4.0.2      httr_1.4.2         
+#> [22] backports_1.1.9     assertthat_0.2.1    cli_2.0.2          
+#> [25] htmltools_0.5.0     quantreg_5.67       tools_4.0.2        
+#> [28] gtable_0.3.0        glue_1.4.2          Rcpp_1.0.5         
 #> [31] cellranger_1.1.0    vctrs_0.3.4         nlme_3.1-149       
@@
 #> [58] rlang_0.4.7         pkgconfig_2.0.3     matrixStats_0.56.0 
-#> [61] evaluate_0.14       lattice_0.20-41     labeling_0.3       
-#> [64] htmlwidgets_1.5.1   cmprsk_2.2-10       tidyselect_1.1.0   
-#> [67] deSolve_1.28        plyr_1.8.6          R6_2.4.1           
-#> [70] generics_0.0.2      Hmisc_4.4-1         multcomp_1.4-13    
-#> [73] DBI_1.1.0           withr_2.2.0         pillar_1.4.6       
-#> [76] haven_2.3.1         foreign_0.8-79      mgcv_1.8-33        
-#> [79] nnet_7.3-14         mstate_0.2.12       modelr_0.1.8       
-#> [82] crayon_1.3.4        rmarkdown_2.3       jpeg_0.1-8.1       
-#> [85] grid_4.0.2          readxl_1.3.1        blob_1.2.1         
-#> [88] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
-#> [91] numDeriv_2016.8-1.1 stats4_4.0.2        munsell_0.5.0      
-#> [94] viridisLite_0.3.0   quadprog_1.5-8
+#> [61] evaluate_0.14       lattice_0.20-41     htmlwidgets_1.5.1  
+#> [64] cmprsk_2.2-10       tidyselect_1.1.0    deSolve_1.28       
+#> [67] R6_2.4.1            generics_0.0.2      Hmisc_4.4-1        
+#> [70] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
+#> [73] pillar_1.4.6        haven_2.3.1         foreign_0.8-79     
+#> [76] mgcv_1.8-33         nnet_7.3-14         mstate_0.2.12      
+#> [79] modelr_0.1.8        crayon_1.3.4        rmarkdown_2.3      
+#> [82] timereg_1.9.7       jpeg_0.1-8.1        grid_4.0.2         
+#> [85] readxl_1.3.1        blob_1.2.1          reprex_0.3.0       
+#> [88] digest_0.6.25       webshot_0.5.2       numDeriv_2016.8-1.1
+#> [91] stats4_4.0.2        munsell_0.5.0       viridisLite_0.3.0  
+#> [94] quadprog_1.5-8
 \end{CodeOutput}

### 2020-09-21 analysis/paper/paper.Rmd
cleaned CS3

@@
   cache = TRUE, 
+  fig.pos = 'ht',
   comment = "#>",
@@
 Their main idea is simple: comparing person-moments when the event of interest occurred with moments when patients were at risk.
-Their approach provides users familiar with generalized linear models a natural way of fitting parametric survival models.
+Their approach handles censored data effortlessly and provides users familiar with generalized linear models a natural way of fitting parametric survival models.
 Moreover, their framework is very flexible: general functions of time can be estimated (e.g.
@@
 \begin{equation}
-\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j P(\theta_j;\lambda,\alpha)
+\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j P(\theta_j;\lambda,\alpha) \label{eq:penest}
 \end{equation}
@@
 
-\begin{table}[ht]
+\begin{table}[H]
 \centering
@@
 By splitting these tasks, we give flexibility to the user.
-While the method call in Step 2 allows further customization by using the \pkg{ggplot2} [@ggplot2] family of functions, users can use for the graphics system of their choice to create population-time plots from the object created in Step 1.
-
-To illustrate these functions, we will use data from the European Randomized Study of Prostate Cancer Screening (ERSPC) [@schroder2009screening] which was extracted using the approach described in Liu *et al.* [-@liu2014recovering] This dataset is available through the \pkg{casebase} package.
+While the method call in Step 2 allows further customization by using the \pkg{ggplot2} [@ggplot2] family of functions, users may choose the graphics system of their choice to create population-time plots from the object created in Step 1.
+
+To illustrate these functions, we will use data from the European Randomized Study of Prostate Cancer Screening (ERSPC) [@schroder2009screening] which was extracted using the approach described in Liu *et al.* [-@liu2014recovering].
+This dataset is available through the \pkg{casebase} package.
 It contains the individual observations for 159,893 men  from seven European countries, who were between the ages of 55 and 69 years when recruited for the trial.
@@
   - \code{glm}: This is the familiar logistic regression.
-  - \code{glmnet}: This option allows for variable selection using Lasso or elastic-net (cf. Case Study 3). This functionality is provided through the \pkg{glmnet} package [@friedman2010jss].
+  - \code{glmnet}: This option allows for variable selection using the elastic-net [@zou2005regularization] penalty (cf.
+Case Study 3).
+This functionality is provided through the \pkg{glmnet} package [@friedman2010jss].
   - \code{gam}: This option provides support for *Generalized Additive Models* via the \pkg{mgcv} package [@hastie1987generalized].
@@
 In the first two case studies, we described the basic functionalities of the \pkg{casebase} package: creating population-time plots, fitting parametric models for hazard functions, and estimating the corresponding cumulative incidence curves.
-For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function.
-To do this, we simply replace logistic regression with a penalized counterpart.
-In \pkg{casebase}, we achieve this by using the \pkg{glmnet} package [@friedman2010jss].
+For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function as given by \eqref{eq:penest}.
 
@@
 The cleaned and imputed data consists of 9104 observations and 30 variables, and it is available as part of the \pkg{casebase} package.
+In the comparisons below, all covariates except \code{sps} and \code{aps} were modeled.
+These two variables correspond to scores for predicting the outcome that were developed as part of the original study.
 For more information about this dataset, the reader is encouraged to look at the documentation in our package.
 
-In this section, we compare our penalized case-base approach to two other models: Cox regression, and penalized Cox regression (using the \pkg{glmnet} package).
-In all three models, we use all covariates except \code{sps} and \code{aps}.
-These two variables correspond to scores for predicting the outcome that were developed as part of the original study.
-
-To compare the performance of our models, we first divide the data into a training and a test data, with 95% of the observations used for training all three models.
-To assess both discrimination and calibration, we use a time-dependent version of the classical Brier score that is adjusted for censoring [@graf1999ass].
+For our penalized case-base model, we opted for the natural log of time which corresponds to a Weibull distribution.
+For fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface to the \code{fitSmoothHazard} function.
+We must supply both a matrix \code{y} containing the time and event variables, and a matrix \code{x} containing all other covariates.
+We applied the lasso penalty by setting \code{alpha = 1} and assigned a \code{penalty.factor} ($w_j$) of 0 to the time variable to ensure it was in the selected model.
+We compared our approach to both Cox regression, and lasso penalized Cox regression (fitted via the \pkg{glmnet} package).
+
+To compare the performance of our models, we split the data into 95% training and 5% test sets.
+To assess both discrimination and calibration, we used a time-dependent version of the classical Brier score that is adjusted for censoring [@graf1999ass].
 The Brier score can be used with both parametric and semi-parametric models.
-We use the \pkg{riskRegression} to compute these scores for all models.
+We used the \pkg{riskRegression} package to compute these scores for all models.
 
@@
 ```
-
-For our penalized case-base model, we opt for the natural log of time; recall that this corresponds to a Weibull distribution.
-For fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface to the function \code{fitSmoothHazard}.
-In order to use this interface, we first create a matrix \code{y} containing the time and event variables, and a matrix \code{x} containing all remaining covariates we would like to include in the model space.
-We used lasso penalization by setting \code{alpha = 1}.
-Moreover, we use the argument \code{penalty.factor} to give a regularization weight of 0 to the time variable, and therefore force it into the final model.
 
@@
 
-Our first comparison plot examines the covariates that were selected between all three hazard functions.
-In \ref{fig:cs3lolliPlot}, we show the coefficient estimates for covariates that we selected by both penalized Cox and penalized case-base.
-From this plot, we can clearly see the shrinkage effect of using penalized estimation.
+In Figure \ref{fig:cs3lolliPlot}, we show the coefficient estimates for covariates that we selected by both penalized Cox and penalized case-base.
+We note that both penalized approaches produce similar results, and can clearly see the shrinkage effect owing to the $\ell_1$ penalty.
 
@@
 
-Next, we compare the cumulative incidence estimation over the test set.
-The probabilities over time for each observation are averaged, resulting in the absolute risk curves in Figure \ref{fig:cs3FinalBrier} (A).
-We can see some minimal differences between the three models, with the Kaplan-Meier estimate giving a lower estimate than our three models.
-We also note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard.
+We then compared the cumulative incidence estimation over the test set.
+The probabilities over time for each observation are averaged, resulting in the absolute risk curves shown in Figure \ref{fig:cs3FinalBrier} (A).
+We can see some minimal differences between the three models, with the Kaplan-Meier giving the lowest estimates across follow-up-time.
+Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard.
 As described above, we compare the performance between the models by computing the Brier scores over time.
@@
 
-In brief, case-base sampling can be used in conjunction with penalized logistic regression to perform variable selection with survival models.
+In this third case study, we showed how case-base sampling can be used in conjunction with penalized logistic regression to perform variable selection in survival models.
 
@@
 We would like to thank Yi Yang for helpful discussions on penalized regression models.
-\mbox{Bhatnagar} gratefully acknowledges funding via Discovery Grants from the Natural Sciences and Engineering Research Council of Canada (NSERC).
+\mbox{Bhatnagar} gratefully acknowledges funding via a Discovery Grant from the Natural Sciences and Engineering Research Council of Canada (NSERC).
 

### 2020-09-21 analysis/paper/paper.tex
cleaned CS3

@@
 interest occurred with moments when patients were at risk. Their
-approach provides users familiar with generalized linear models a
-natural way of fitting parametric survival models. Moreover, their
-framework is very flexible: general functions of time can be estimated
-(e.g.~using splines or general additive models), and hence these models
-retain some of the flexibility of Cox's partial likelihood approach. And
-since the unit of analysis is a person moment, time-varying covariates
-can also easily be included in this framework.
+approach handles censored data effortlessly and provides users familiar
+with generalized linear models a natural way of fitting parametric
+survival models. Moreover, their framework is very flexible: general
+functions of time can be estimated (e.g.~using splines or general
+additive models), and hence these models retain some of the flexibility
+of Cox's partial likelihood approach. And since the unit of analysis is
+a person moment, time-varying covariates can also easily be included in
+this framework.
 
@@
 to the likelihood and solve the following problem: \begin{equation}
-\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j P(\theta_j;\lambda,\alpha)
+\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j P(\theta_j;\lambda,\alpha) \label{eq:penest}
 \end{equation} where \(\ell\left(\theta\right)\) is the log of the
 likelihood function given in \eqref{eq:lik-function},
-\(P(\theta_j;\lambda,\alpha)\) is a penalty term controlled controlled
-by the non-negative regularization parameters \(\lambda\) and
-\(\alpha\), and \(w_j\) is the penalty factor for the \(j\)th covariate.
-These penalty factors serve as a way of allowing parameters to be
-penalized differently. For example, we could set the penalty factor for
-time to be 0 to ensure it is always included in the selected model.
+\(P(\theta_j;\lambda,\alpha)\) is a penalty term controlled by the
+non-negative regularization parameters \(\lambda\) and \(\alpha\), and
+\(w_j\) is the penalty factor for the \(j\)th covariate. These penalty
+factors serve as a way of allowing parameters to be penalized
+differently. For example, we could set the penalty factor for time to be
+0 to ensure it is always included in the selected model.
 
@@
 method call in Step 2 allows further customization by using the
-\pkg{ggplot2} \citep{ggplot2} family of functions, users can use for the
+\pkg{ggplot2} \citep{ggplot2} family of functions, users may choose the
 graphics system of their choice to create population-time plots from the
@@
 \citep{schroder2009screening} which was extracted using the approach
-described in Liu \emph{et al.} \citeyearpar{liu2014recovering} This
+described in Liu \emph{et al.} \citeyearpar{liu2014recovering}. This
 dataset is available through the \pkg{casebase} package. It contains the
@@
 \begin{CodeChunk}
-\begin{figure}
+\begin{figure}[ht]
 
@@
 \item
-  \code{glmnet}: This option allows for variable selection using Lasso
-  or elastic-net (cf.~Case Study 3). This functionality is provided
-  through the \pkg{glmnet} package \citep{friedman2010jss}.
+  \code{glmnet}: This option allows for variable selection using the
+  elastic-net \citep{zou2005regularization} penalty (cf.~Case Study 3).
+  This functionality is provided through the \pkg{glmnet} package
+  \citep{friedman2010jss}.
 \item
@@
 \begin{CodeChunk}
-\begin{figure}
+\begin{figure}[ht]
 \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/erspc-cox-cif-1} \caption{CIFs for control and screening groups in the ERSPC data.
@@
 \pkg{casebase} can also be used for variable selection through
-regularized estimation of the hazard function. To do this, we simply
-replace logistic regression with a penalized counterpart. In
-\pkg{casebase}, we achieve this by using the \pkg{glmnet} package
-\citep{friedman2010jss}.
+regularized estimation of the hazard function as given by
+\eqref{eq:penest}.
 
@@
 and imputed data consists of 9104 observations and 30 variables, and it
-is available as part of the \pkg{casebase} package. For more information
-about this dataset, the reader is encouraged to look at the
-documentation in our package.
-
-In this section, we compare our penalized case-base approach to two
-other models: Cox regression, and penalized Cox regression (using the
-\pkg{glmnet} package). In all three models, we use all covariates except
-\code{sps} and \code{aps}. These two variables correspond to scores for
-predicting the outcome that were developed as part of the original
-study.
-
-To compare the performance of our models, we first divide the data into
-a training and a test data, with 95\% of the observations used for
-training all three models. To assess both discrimination and
-calibration, we use a time-dependent version of the classical Brier
+is available as part of the \pkg{casebase} package. In the comparisons
+below, all covariates except \code{sps} and \code{aps} were modeled.
+These two variables correspond to scores for predicting the outcome that
+were developed as part of the original study. For more information about
+this dataset, the reader is encouraged to look at the documentation in
+our package.
+
+For our penalized case-base model, we opted for the natural log of time
+which corresponds to a Weibull distribution. For fitting the penalized
+hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface
+to the \code{fitSmoothHazard} function. We must supply both a matrix
+\code{y} containing the time and event variables, and a matrix \code{x}
+containing all other covariates. We applied the lasso penalty by setting
+\code{alpha = 1} and assigned a \code{penalty.factor} (\(w_j\)) of 0 to
+the time variable to ensure it was in the selected model. We compared
+our approach to both Cox regression, and lasso penalized Cox regression
+(fitted via the \pkg{glmnet} package).
+
+To compare the performance of our models, we split the data into 95\%
+training and 5\% test sets. To assess both discrimination and
+calibration, we used a time-dependent version of the classical Brier
 score that is adjusted for censoring \citep{graf1999ass}. The Brier
 score can be used with both parametric and semi-parametric models. We
-use the \pkg{riskRegression} to compute these scores for all models.
-
-For our penalized case-base model, we opt for the natural log of time;
-recall that this corresponds to a Weibull distribution. For fitting the
-penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix
-interface to the function \code{fitSmoothHazard}. In order to use this
-interface, we first create a matrix \code{y} containing the time and
-event variables, and a matrix \code{x} containing all remaining
-covariates we would like to include in the model space. We used lasso
-penalization by setting \code{alpha = 1}. Moreover, we use the argument
-\code{penalty.factor} to give a regularization weight of 0 to the time
-variable, and therefore force it into the final model.
+used the \pkg{riskRegression} package to compute these scores for all
+models.
 
@@
 
-Our first comparison plot examines the covariates that were selected
-between all three hazard functions. In \ref{fig:cs3lolliPlot}, we show
-the coefficient estimates for covariates that we selected by both
-penalized Cox and penalized case-base. From this plot, we can clearly
-see the shrinkage effect of using penalized estimation.
-
-\begin{CodeChunk}
-\begin{figure}
+In Figure \ref{fig:cs3lolliPlot}, we show the coefficient estimates for
+covariates that we selected by both penalized Cox and penalized
+case-base. We note that both penalized approaches produce similar
+results, and can clearly see the shrinkage effect owing to the
+\(\ell_1\) penalty.
+
+\begin{CodeChunk}
+\begin{figure}[ht]
 
@@
 
-Next, we compare the cumulative incidence estimation over the test set.
+We then compared the cumulative incidence estimation over the test set.
 The probabilities over time for each observation are averaged, resulting
-in the absolute risk curves in Figure \ref{fig:cs3FinalBrier} (A). We
-can see some minimal differences between the three models, with the
-Kaplan-Meier estimate giving a lower estimate than our three models. We
-also note that the apparent smoothness of the Cox and penalized Cox
-curves is due to the large number of observations in the training set,
-which is used to derive the Breslow estimate of the baseline hazard. As
-described above, we compare the performance between the models by
-computing the Brier scores over time. In Figure \ref{fig:cs3FinalBrier}
-(B), we can see that the Brier score is larger for the Kaplan-Meier
-estimate than for the other three models. On the other hand, the
-differences between these three models are minimal.
-
-\begin{CodeChunk}
-\begin{figure}
+in the absolute risk curves shown in Figure \ref{fig:cs3FinalBrier} (A).
+We can see some minimal differences between the three models, with the
+Kaplan-Meier giving the lowest estimates across follow-up-time. Note
+that the apparent smoothness of the Cox and penalized Cox curves is due
+to the large number of observations in the training set, which is used
+to derive the Breslow estimate of the baseline hazard. As described
+above, we compare the performance between the models by computing the
+Brier scores over time. In Figure \ref{fig:cs3FinalBrier} (B), we can
+see that the Brier score is larger for the Kaplan-Meier estimate than
+for the other three models. On the other hand, the differences between
+these three models are minimal.
+
+\begin{CodeChunk}
+\begin{figure}[ht]
 
@@
 
-In brief, case-base sampling can be used in conjunction with penalized
-logistic regression to perform variable selection with survival models.
+In this third case study, we showed how case-base sampling can be used
+in conjunction with penalized logistic regression to perform variable
+selection in survival models.
 
@@
 \end{CodeInput}
-\begin{figure}
+\begin{figure}[ht]
 
@@
 regression models. \mbox{Bhatnagar} gratefully acknowledges funding via
-Discovery Grants from the Natural Sciences and Engineering Research
+a Discovery Grant from the Natural Sciences and Engineering Research
 Council of Canada (NSERC).
@@
 
-This report was generated on 2020-09-20 12:00:27 using the following
+This report was generated on 2020-09-20 20:42:38 using the following
 computational environment and dependencies:
@@
 #> other attached packages:
-#>  [1] scales_1.1.1              forcats_0.5.0            
-#>  [3] stringr_1.4.0             dplyr_1.0.2              
-#>  [5] purrr_0.3.4               readr_1.3.1              
-#>  [7] tidyr_1.1.2               ggplot2_3.3.2            
-#>  [9] tidyverse_1.3.0           lubridate_1.7.9          
-#> [11] visreg_2.7.0              tibble_3.0.3             
-#> [13] riskRegression_2020.02.05 prodlim_2019.11.13       
-#> [15] flexsurv_1.1.1            survival_3.2-3           
-#> [17] pracma_2.2.9              magrittr_1.5             
-#> [19] kableExtra_1.2.1          glmnet_4.0-2             
-#> [21] Matrix_1.2-18             data.table_1.13.0        
-#> [23] cowplot_1.1.0             colorspace_1.4-1         
-#> [25] casebase_0.9.0           
+#>  [1] broom_0.7.0               dotwhisker_0.5.0         
+#>  [3] timereg_1.9.7             glue_1.4.2               
+#>  [5] scales_1.1.1              forcats_0.5.0            
+#>  [7] stringr_1.4.0             dplyr_1.0.2              
+#>  [9] purrr_0.3.4               readr_1.3.1              
+#> [11] tidyr_1.1.2               ggplot2_3.3.2            
+#> [13] tidyverse_1.3.0           lubridate_1.7.9          
+#> [15] visreg_2.7.0              tibble_3.0.3             
+#> [17] riskRegression_2020.02.05 prodlim_2019.11.13       
+#> [19] flexsurv_1.1.1            survival_3.2-3           
+#> [21] pracma_2.2.9              magrittr_1.5             
+#> [23] kableExtra_1.2.1          glmnet_4.0-2             
+#> [25] Matrix_1.2-18             data.table_1.13.0        
+#> [27] cowplot_1.1.0             colorspace_1.4-1         
+#> [29] casebase_0.9.0           
 #> 
@@
 #>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
-#>  [4] htmlTable_2.1.0     base64enc_0.1-3     fs_1.5.0           
-#>  [7] rstudioapi_0.11     MatrixModels_0.4-1  fansi_0.4.1        
-#> [10] mvtnorm_1.1-1       xml2_1.3.2          codetools_0.2-16   
-#> [13] knitr_1.29          Formula_1.2-3       jsonlite_1.7.1     
-#> [16] broom_0.7.0         cluster_2.1.0       dbplyr_1.4.4       
-#> [19] png_0.1-7           compiler_4.0.2      httr_1.4.2         
-#> [22] backports_1.1.9     assertthat_0.2.1    cli_2.0.2          
-#> [25] htmltools_0.5.0     quantreg_5.67       tools_4.0.2        
-#> [28] gtable_0.3.0        glue_1.4.2          Rcpp_1.0.5         
+#>  [4] ggstance_0.3.4      htmlTable_2.1.0     base64enc_0.1-3    
+#>  [7] fs_1.5.0            rstudioapi_0.11     farver_2.0.3       
+#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-1      
+#> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
+#> [16] Formula_1.2-3       jsonlite_1.7.1      cluster_2.1.0      
+#> [19] dbplyr_1.4.4        png_0.1-7           compiler_4.0.2     
+#> [22] httr_1.4.2          backports_1.1.9     assertthat_0.2.1   
+#> [25] cli_2.0.2           htmltools_0.5.0     quantreg_5.67      
+#> [28] tools_4.0.2         gtable_0.3.0        Rcpp_1.0.5         
 #> [31] cellranger_1.1.0    vctrs_0.3.4         nlme_3.1-149       
@@
 #> [49] gridExtra_2.3       rms_6.0-1           rpart_4.1-15       
-#> [52] latticeExtra_0.6-29 stringi_1.5.3       foreach_1.5.0      
-#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.5        
-#> [58] rlang_0.4.7         pkgconfig_2.0.3     matrixStats_0.56.0 
-#> [61] evaluate_0.14       lattice_0.20-41     htmlwidgets_1.5.1  
-#> [64] cmprsk_2.2-10       tidyselect_1.1.0    deSolve_1.28       
-#> [67] R6_2.4.1            generics_0.0.2      Hmisc_4.4-1        
-#> [70] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
-#> [73] pillar_1.4.6        haven_2.3.1         foreign_0.8-79     
-#> [76] mgcv_1.8-33         nnet_7.3-14         mstate_0.2.12      
-#> [79] modelr_0.1.8        crayon_1.3.4        rmarkdown_2.3      
-#> [82] timereg_1.9.7       jpeg_0.1-8.1        grid_4.0.2         
-#> [85] readxl_1.3.1        blob_1.2.1          reprex_0.3.0       
-#> [88] digest_0.6.25       webshot_0.5.2       numDeriv_2016.8-1.1
-#> [91] stats4_4.0.2        munsell_0.5.0       viridisLite_0.3.0  
-#> [94] quadprog_1.5-8
+#> [52] latticeExtra_0.6-29 stringi_1.5.3       highr_0.8          
+#> [55] foreach_1.5.0       checkmate_2.0.0     lava_1.6.7         
+#> [58] shape_1.4.5         rlang_0.4.7         pkgconfig_2.0.3    
+#> [61] matrixStats_0.56.0  evaluate_0.14       lattice_0.20-41    
+#> [64] labeling_0.3        htmlwidgets_1.5.1   cmprsk_2.2-10      
+#> [67] tidyselect_1.1.0    deSolve_1.28        plyr_1.8.6         
+#> [70] R6_2.4.1            generics_0.0.2      Hmisc_4.4-1        
+#> [73] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
+#> [76] pillar_1.4.6        haven_2.3.1         foreign_0.8-79     
+#> [79] mgcv_1.8-33         nnet_7.3-14         mstate_0.2.12      
+#> [82] modelr_0.1.8        crayon_1.3.4        rmarkdown_2.3      
+#> [85] jpeg_0.1-8.1        grid_4.0.2          readxl_1.3.1       
+#> [88] blob_1.2.1          reprex_0.3.0        digest_0.6.25      
+#> [91] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_4.0.2       
+#> [94] munsell_0.5.0       viridisLite_0.3.0   quadprog_1.5-8
 \end{CodeOutput}
@@
 #> Remote:   editSahir @ origin (https://github.com/sahirbhatnagar/cbpaper.git)
-#> Head:     [d691836] 2020-09-19: Merge remote-tracking branch 'origin/master' into editSahir
+#> Head:     [dace5dd] 2020-09-20: added paragraph on variable selection in theoretical details
 \end{CodeOutput}

### 2020-09-21 analysis/paper/paper.Rmd
equation references for CS1

@@
 For our first case study, we return to the ERSPC study and investigate the differences in risk between the control and screening arms.
-We fit four models that differ in which functional form of time is used: 1) excluded from the linear predictor, 2) linear function, 3) log function, and 4) a smooth function using cubic B-splines.
+We fit four models that differ in which functional form of time is used: 1) excluded from the linear predictor as seen in \eqref{eq:exp}, 2) linear function as seen in \eqref{eq:gomp}, 3) log function as seen in \eqref{eq:weibull}, and 4) a smooth function using cubic B-splines.
 The models are fit using \code{fitSmoothHazard} with the familiar formula interface:

### 2020-09-21 analysis/paper/paper.tex
equation references for CS1

@@
 four models that differ in which functional form of time is used: 1)
-excluded from the linear predictor, 2) linear function, 3) log function,
-and 4) a smooth function using cubic B-splines. The models are fit using
-\code{fitSmoothHazard} with the familiar formula interface:
+excluded from the linear predictor as seen in \eqref{eq:exp}, 2) linear
+function as seen in \eqref{eq:gomp}, 3) log function as seen in
+\eqref{eq:weibull}, and 4) a smooth function using cubic B-splines. The
+models are fit using \code{fitSmoothHazard} with the familiar formula
+interface:
 
@@
 
-This report was generated on 2020-09-20 20:42:38 using the following
+This report was generated on 2020-09-21 12:49:15 using the following
 computational environment and dependencies:
@@
 #> [49] gridExtra_2.3       rms_6.0-1           rpart_4.1-15       
-#> [52] latticeExtra_0.6-29 stringi_1.5.3       highr_0.8          
-#> [55] foreach_1.5.0       checkmate_2.0.0     lava_1.6.7         
-#> [58] shape_1.4.5         rlang_0.4.7         pkgconfig_2.0.3    
-#> [61] matrixStats_0.56.0  evaluate_0.14       lattice_0.20-41    
-#> [64] labeling_0.3        htmlwidgets_1.5.1   cmprsk_2.2-10      
-#> [67] tidyselect_1.1.0    deSolve_1.28        plyr_1.8.6         
-#> [70] R6_2.4.1            generics_0.0.2      Hmisc_4.4-1        
-#> [73] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
-#> [76] pillar_1.4.6        haven_2.3.1         foreign_0.8-79     
-#> [79] mgcv_1.8-33         nnet_7.3-14         mstate_0.2.12      
-#> [82] modelr_0.1.8        crayon_1.3.4        rmarkdown_2.3      
-#> [85] jpeg_0.1-8.1        grid_4.0.2          readxl_1.3.1       
-#> [88] blob_1.2.1          reprex_0.3.0        digest_0.6.25      
-#> [91] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_4.0.2       
-#> [94] munsell_0.5.0       viridisLite_0.3.0   quadprog_1.5-8
+#> [52] latticeExtra_0.6-29 stringi_1.5.3       foreach_1.5.0      
+#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.5        
+#> [58] rlang_0.4.7         pkgconfig_2.0.3     matrixStats_0.56.0 
+#> [61] evaluate_0.14       lattice_0.20-41     htmlwidgets_1.5.1  
+#> [64] cmprsk_2.2-10       tidyselect_1.1.0    deSolve_1.28       
+#> [67] R6_2.4.1            generics_0.0.2      Hmisc_4.4-1        
+#> [70] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
+#> [73] pillar_1.4.6        haven_2.3.1         foreign_0.8-79     
+#> [76] mgcv_1.8-33         nnet_7.3-14         mstate_0.2.12      
+#> [79] modelr_0.1.8        crayon_1.3.4        rmarkdown_2.3      
+#> [82] timereg_1.9.7       jpeg_0.1-8.1        grid_4.0.2         
+#> [85] readxl_1.3.1        blob_1.2.1          reprex_0.3.0       
+#> [88] digest_0.6.25       webshot_0.5.2       numDeriv_2016.8-1.1
+#> [91] stats4_4.0.2        munsell_0.5.0       viridisLite_0.3.0  
+#> [94] quadprog_1.5-8
 \end{CodeOutput}

### 2020-09-21 analysis/paper/paper.Rmd
added casebase package to intro, and vignette to discussion

@@
 
-In this article, we present a comprehensive and flexible \proglang{R} package for parametric survival analysis that combines the ideas of Hanley & Miettinen [-@hanley2009fitting] into a simple interface.
+In this article, we present the \pkg{casebase} package [@casebase-package] implemented in \proglang{R} [@r-core], for parametric survival analysis that combines the ideas of Hanley & Miettinen [-@hanley2009fitting] into a simple interface.
 The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to compute a patient's risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile.
@@
 
-\begin{table}[H]
+\begin{table}[ht]
 \centering
@@
 For our first case study, we return to the ERSPC study and investigate the differences in risk between the control and screening arms.
+
+\subsection{Different functional forms of time}
+
 We fit four models that differ in which functional form of time is used: 1) excluded from the linear predictor as seen in \eqref{eq:exp}, 2) linear function as seen in \eqref{eq:gomp}, 3) log function as seen in \eqref{eq:weibull}, and 4) a smooth function using cubic B-splines.
@@
 ```
+
+\subsection{Time-dependent hazard ratios}
 
@@
 We strongly believe that its flexibility and its foundation on the familiar logistic regression model will make it appealing to new and established practitioners.
+The \pkg{casebase} package is freely available from the Comprehensive \proglang{R} Archive Network at https://cran.r-project.org/package=casebase.
+Interested users can visit http://sahirbhatnagar.com/casebase/ for detailed package documentation and vignettes.
 
@@
 
-We would like to thank Yi Yang for helpful discussions on penalized regression models.
+We would like to thank Yi Yang for helpful discussions on penalized regression models and gradient boosting.
 \mbox{Bhatnagar} gratefully acknowledges funding via a Discovery Grant from the Natural Sciences and Engineering Research Council of Canada (NSERC).

### 2020-09-21 analysis/paper/paper.tex
added casebase package to intro, and vignette to discussion

@@
 
-In this article, we present a comprehensive and flexible \proglang{R}
-package for parametric survival analysis that combines the ideas of
-Hanley \& Miettinen \citeyearpar{hanley2009fitting} into a simple
-interface. The purpose of the \pkg{casebase} package is to provide
-practitioners with an easy-to-use software tool to compute a patient's
-risk (or cumulative incidence) of an event, conditional on a particular
-patient's covariate profile. Our package retains the flexibility of
-case-base sampling and the familiar interface of the \code{glm}
-function. It also provides tools for variable selection and
-competing-risk analysis. In addition, we provide extensive visualization
-tools.
+In this article, we present the \pkg{casebase} package
+\citep{casebase-package} implemented in \proglang{R} \citep{r-core}, for
+parametric survival analysis that combines the ideas of Hanley \&
+Miettinen \citeyearpar{hanley2009fitting} into a simple interface. The
+purpose of the \pkg{casebase} package is to provide practitioners with
+an easy-to-use software tool to compute a patient's risk (or cumulative
+incidence) of an event, conditional on a particular patient's covariate
+profile. Our package retains the flexibility of case-base sampling and
+the familiar interface of the \code{glm} function. It also provides
+tools for variable selection and competing-risk analysis. In addition,
+we provide extensive visualization tools.
 
@@
 For our first case study, we return to the ERSPC study and investigate
-the differences in risk between the control and screening arms. We fit
-four models that differ in which functional form of time is used: 1)
-excluded from the linear predictor as seen in \eqref{eq:exp}, 2) linear
-function as seen in \eqref{eq:gomp}, 3) log function as seen in
+the differences in risk between the control and screening arms.
+
+\subsection{Different functional forms of time}
+
+We fit four models that differ in which functional form of time is used:
+1) excluded from the linear predictor as seen in \eqref{eq:exp}, 2)
+linear function as seen in \eqref{eq:gomp}, 3) log function as seen in
 \eqref{eq:weibull}, and 4) a smooth function using cubic B-splines. The
@@
 \end{CodeChunk}
+
+\subsection{Time-dependent hazard ratios}
 
@@
 flexibility and its foundation on the familiar logistic regression model
-will make it appealing to new and established practitioners.
+will make it appealing to new and established practitioners. The
+\pkg{casebase} package is freely available from the Comprehensive
+\proglang{R} Archive Network at
+\url{https://cran.r-project.org/package=casebase}. Interested users can
+visit \url{http://sahirbhatnagar.com/casebase/} for detailed package
+documentation and vignettes.
 
@@
 We would like to thank Yi Yang for helpful discussions on penalized
-regression models. \mbox{Bhatnagar} gratefully acknowledges funding via
-a Discovery Grant from the Natural Sciences and Engineering Research
-Council of Canada (NSERC).
+regression models and gradient boosting. \mbox{Bhatnagar} gratefully
+acknowledges funding via a Discovery Grant from the Natural Sciences and
+Engineering Research Council of Canada (NSERC).
 
@@
 
-This report was generated on 2020-09-21 12:49:15 using the following
+This report was generated on 2020-09-21 14:06:21 using the following
 computational environment and dependencies:

### 2020-09-22 analysis/paper/paper.Rmd
final version

@@
   | In epidemiological studies of time-to-event data, a quantity of interest to the clinician and the patient is the risk of an event given a covariate profile.
-However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the likelihood expression or the estimating function and therefore needs to be estimated separately using a non-parametric approach.
+However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the likelihood expression or the estimating function.
+The baseline hazard then needs to be estimated separately using a non-parametric approach.
 This leads to step-wise estimates of the cumulative incidence that are difficult to interpret.
@@
 
+# Turn on/off sections for faster compilation
 eval_introduction <- TRUE
@@
 eval_cs4 <- TRUE
-eval_colophon <- TRUE
-
+eval_colophon <- FALSE # Only for draft
 echo_plot_code <- TRUE
-
+# Define theme for plots
 paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
@@
 
-Using the framework of logistic regression for analyzing time-to-event data opens the door to an extensive array of powerful modeling tools.
+Logistic regression had already been used in the context of discrete-time survival modeling [@cox1972regression].
+But in the context of continuous-time survival modeling, using the framework of logistic regression opens the door to an extensive array of powerful modeling tools.
 Indeed, lasso and elastic-net regression can be used to select variables that are associated with the hazard function.
@@
 
+## Likelihood and estimating function
+
 To describe the theoretical foundations of case-base sampling, we use the framework of counting processes.
 In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras.
-Instead, following Aalen \emph{et al} [-@aalen2008survival], we use the placeholder $\mathrm{past}$ to denote the past history of the corresponding process.
+Instead, following Aalen \emph{et al} [-@aalen2008survival], we use the placeholder "past" to denote the past history of the corresponding process.
 The reader interested in more details can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case].
@@
 More complex examples are also possible; see for example Saarela & Arjas [-@saarela2015non], where the intensity functions for the sampling mechanism are proportional to the cardiovascular disease event rate given by the Framingham score.
-With this sampling mechanism, Saarela & Arjas are able to increase the efficiency of their estimators, when compared to uniform sampling.
+Non-uniform sampling mechanisms can increase the efficiency of the case-base estimators.
+
+## Common parametric models
 
@@
 
+## Competing-risk analysis
+
 Case-base sampling can also be used in the context of competing-risk analyses.
@@
 
-To perform variable selection on the regression parameters $\theta \in \mathbb{R}^p$ of the hazard function, we can add a penalty to the likelihood and solve the following problem:
+## Variable selection
+
+To perform variable selection on the regression parameters $\theta \in \mathbb{R}^p$ of the hazard function, we can add a penalty to the likelihood and optimise the following equation:
 \begin{equation}
@@
 \end{equation}
-where $\ell\left(\theta\right)$ is the log of the likelihood function given in \eqref{eq:lik-function}, $P(\theta_j;\lambda,\alpha)$ is a penalty term controlled by the non-negative regularization parameters $\lambda$ and $\alpha$, and $w_j$ is the penalty factor for the $j$th covariate.
+where $\ell\left(\theta\right) = \log L(\theta)$ is the log of the likelihood function given in \eqref{eq:lik-function}, $P(\theta_j;\lambda,\alpha)$ is a penalty term controlled by the non-negative regularization parameters $\lambda$ and $\alpha$, and $w_j$ is the penalty factor for the $j$th covariate.
 These penalty factors serve as a way of allowing parameters to be penalized differently.
@@
 
-At the time of writing, a cursory examination of CRAN's task view on survival analysis reveals that there are over 250 packages related to survival analysis [-@survTaskView].
+At the time of writing, a cursory examination of CRAN's task view on survival analysis reveals that there are over 250 packages related to survival analysis [@survTaskView].
 For the purposes of this article, we restricted our review to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and CIF estimation.
@@
 ```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plots for both treatment arms in the ERSPC dataset.
-The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely small rectangles (person-moments).
-The points shown in the figure are select person-moments from the case and base series.
-We see that more events are observed at later follow-up times, motivating the use of a non-constant in time hazard model."}
+The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely thin rectangles (person-moments).
+More events are observed at later follow-up times, motivating the use of non-constant hazard models."}
 label_number_si <- function (accuracy = 1, unit = NULL, sep = NULL, ...) {
@@
 from left to right), the plot for the first three years shows attrition due entirely to death (mainly from other causes).
-Since the Swedish and Belgian centres were the last to complete recruitment--in December 2003--the minimum potential follow-up is three years.
+Since the Swedish and Belgian centres were the last to complete recruitment in December 2003, the minimum potential follow-up is three years.
 Tracked further forwards in time (i.e.
@@
 \end{equation}
-The integral is computed using either the \code{stats::integrate} function or Monte-Carlo integration. The risk function (or cumulative incidence function) is then defined as
+The integral is computed using either the numerical or Monte-Carlo integration. The risk function (or cumulative incidence function) is then defined as
 \begin{equation}\label{eqn:CI}
@@
 $$ CI_j(t) = \int_0^t f_j(u)du.$$
-The integrals are computed using either numerical integration (via the trapezoidal rule) or Monte Carlo integration.
+Again, he integrals are computed using either numerical integration (via the trapezoidal rule) or Monte Carlo integration.
 This option is controlled by the argument \code{method} of the \code{absoluteRisk} function.
 
-<!--We created \code{absoluteRisk} as an \code{S3} generic, with methods that match the different types of outputs of \code{fitSmoothHazard}.
-The method dispatch system of \proglang{R} then takes care of matching the correct output to the correct methodology for calculating the cumulative incidence function, without requiring any intervention from the user.-->
-
 In the following sections, we illustrate these functionalities in the context of four case studies.
@@
 
-\subsection{Different functional forms of time}
+## Different functional forms of time
 
@@
 
-```{r, results='hide', eval = FALSE, echo = FALSE}
-# par(mfrow = c(2, 2))
-# lapply(fits, function(i) {
-#   casebase::hazardPlot(i, newdata = new_data[2,], ci = FALSE)
-#   casebase::hazardPlot(i, newdata = new_data[1,], ci = FALSE, add = TRUE, lty = 2)
-# })
-```
-
-```{r erspc-casebase-exponential, eval = FALSE, echo = FALSE}
-# # Equivalence between casebase and AFT
-# # exponential -------------------------------------------------------------
-# 
-# casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
-#                                                   data = ERSPC, 
-#                                                   ratio = 100)
-# aft_exponential <- survreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
-#                            data = ERSPC, dist = "exp")
-# 
-# -coef(aft_exponential)
-# coef(casebase_exponential)
-# summary(casebase_exponential)
-# exp(coef(casebase_exponential)[2])
-# exp(confint(casebase_exponential)[2, ])
-# 
-# # weibull -----------------------------------------------------------------
-# 
-# casebase_weibull <- casebase::fitSmoothHazard(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm, 
-#                                            data = ERSPC, 
-#                                            ratio = 100)
-# aft_weibull <- survreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
-#                        data = ERSPC, dist = "weibull")
-# 
-# -coef(aft_weibull)/aft_weibull$scale
-# coef(casebase_weibull)
-```
-
-<!-- The \code{fitSmoothHazard} function provides an estimate of the hazard function $h(x, t)$ is the hazard function, where $t$ denotes the numerical value of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any).
--->
-<!-- The \code{absoluteRisk} function provides an estimate of the cumulative incidence curves for a specific risk profile using the following equation: -->
-<!-- $$ CI(x, t) = 1 - \exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$ -->
-
 ```{r erspc-95ci22, echo = FALSE, eval = eval_cs1}
@@
 
-```{r erspc-95ci, echo = FALSE, eval = FALSE}
-# CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
-# CI_exp <- c(exp(coef(casebase_exponential))["ScrArmScreening group"], exp(confint(casebase_exponential))["ScrArmScreening group", ])
-# CI_time <- c(exp(coef(casebase_time))["ScrArmScreening group"], exp(confint(casebase_time))["ScrArmScreening group", ])
-# CI_splines <- c(exp(coef(casebase_splines))["ScrArmScreening group"], exp(confint(casebase_splines))["ScrArmScreening group", ])
-# 
-# table_ci <- rbind(CI_cox, CI_exp, CI_time, CI_splines)
-# colnames(table_ci) <- c("HR", "LB", "UB")
-# 
-# data.frame(Model = c("Cox", "Exponential", "Weibull", "Splines")) %>%
-#   bind_cols(as.data.frame(table_ci)) %>%
-#   mutate(
-#     CI = paste0(
-#       "(", round(LB, 2), ", ",
-#       round(UB, 2), ")"
-#     ),
-#     Width = UB - LB
-#   ) %>%
-#   dplyr::select(-LB, -UB) %>%
-#   knitr::kable(format = "latex", booktabs = TRUE) %>%
-#   kable_styling()
-```
-
-\subsection{Time-dependent hazard ratios}
+## Time-dependent hazard ratios
 
@@
 set.seed(12)
-fit_inter <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) * ScrArm, data = ERSPC)
-```
-
-In Figure \ref{fig:interaction-ERSPC} we plot the estimated death from prostate cancer hazard ratio and 95% confidence interval for screening vs.
+fit_inter <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) * ScrArm, 
+                             data = ERSPC)
+```
+
+In Figure \ref{fig:interaction-ERSPC}, we have the estimated hazard ratio and 95% confidence interval for screening vs.
 control group as a function of time using the \code{plot} method for objects of class \code{singleEventCB}:
@@
 
-```{r interaction-ERSPC, eval = eval_cs1, fig.width=8, fig.height=6, fig.cap='Estimated death from prostate cancer hazard ratio and 95\\% confidence interval for screening vs.
+```{r interaction-ERSPC, eval = eval_cs1, fig.width=8, fig.height=6, fig.cap='Estimated hazard ratio and 95\\% confidence interval for screening vs.
 control group as a function of time in the ERSPC dataset.
@@
 95\\% confidence intervals are calculated using the delta method.
-The plot shows that the cures attributable to the screening only begin to become statistically apparent by year 7 and later.
+The plot shows that the effect of screening only begins to become statistically apparent by year 7.
 The 25-60\\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\\% reported in the original article.'}
-
-# library(casebase)
-# library(splines)
-# data("ERSPC")
-# ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
-#                        levels = c(0,1), 
-#                        labels = c("Control group", "Screening group"))
-# set.seed(12345)
-# fit_inter <- casebase::fitSmoothHazard(DeadOfPrCa ~ poly(Follow.Up.Time,2)*ScrArm, data = ERSPC)
-# summary(fit_inter)
-
-# library(visreg)
-# plot(fit_inter,
-#      type = "hazard",
-#      hazard.params = list(xvar = "Follow.Up.Time",
-#                           by = "ScrArm",
-#                           alpha = 0.05,
-#                           ylab = "Hazard"))
-
-# reference category
-#dev.off()
 new_time <- seq(1, 12, by  = 0.1)
@@
      xaxt = "n",
-     rug = T)
-# abline(a=1,b=0, lty = 2)
+     rug = TRUE)
+
 axis(2, at = seq(0,1.50, by=0.25), las = 2)
@@
 
-\newpage
-
 # Case study 2---Bone-marrow transplant
 
-In the next case study, we show how case-base sampling can  be used in the context of a competing risk analysis.
-For illustrative purposes, we will use the same data that was used in Scrucca *et al* [-@scrucca2010regression].
+In the next case study, we show how case-base sampling can be used in the context of a competing-risk analysis.
+For illustrative purposes, we use the same data that was used in Scrucca *et al* [-@scrucca2010regression].
 The data was downloaded from the first author's website, and it is now available as part of the \pkg{casebase} package.
@@
 Several covariates were captured at baseline: sex, disease type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL and AML, respectively), disease phase at transplant (Relapse, CR1, CR2, CR3), source of stem cells (bone marrow and peripheral blood, coded as BM+PB, or only peripheral blood, coded as PB), and age.
-<!--A summary of these baseline characteristics appear in Table \ref{tab:table1bmtcrr}.
-We note that the statistical summaries were generated differently for different variable types: for continuous variables, we gave the range, followed by the mean and standard deviation; for categorical variables, we gave the counts for each category.-->
-
-```{r echo = FALSE, eval = FALSE}
-tibble::tribble(
-  ~Variable, ~Description, ~`Statistical summary`,
-  "\\code{Sex}", "Sex", "M=Male (100)",
-  "", "", "F=Female (77)",
-  "\\code{D}", "Disease", "ALL (73)",
-  "", "", "AML (104)",
-  "\\code{Phase}", "Phase", "CR1 (47)",
-  "", "", "CR2 (45)",
-  "", "", "CR3 (12)",
-  "", "", "Relapse (73)",
-  "\\code{Source}", "Type of transplant", "BM+PB (21)",
-  "", "", "PB (156)",
-  "\\code{Age}", "Age of patient (years)", "4--62",
-  "", "", "30.47 (13.04)",
-  "\\code{ftime}", "Failure time (months)", "0.13--131.77",
-  "", "", "20.28 (30.78)",
-  "\\code{Status}", "Status indicator", "0=censored (46)",
-  "", "", "1=relapse (56)",
-  "", "", "2=competing event (75)"
-) %>%
-  knitr::kable(
-    format = "latex", booktabs = TRUE,
-    caption = "Baseline characteristics of patients in the stem-cell transplant study.",
-    label = "table1bmtcrr",
-    escape = FALSE
-  ) %>%
-  kable_styling()
-```
-
-First, we can look at a population-time plot to visualize the incidence density of both relapse and the competing events.
+
+First, we look at a population-time plot to visualize the incidence density of both relapse and the competing events.
 In Figure \ref{fig:compPop}, failure times are highlighted on the plot using red dots for the event of interest (panel A) and blue dots for competing events (panel B).
@@
 ```{r compPop, fig.cap="Population-time plot for the stem-cell transplant study with both relapse and competing events.", echo=FALSE, eval = eval_cs2}
-
-# fill_cols <- colorspace::sequential_hcl(n = 3, palette = "Viridis")
-# 
-# (fill_colors <- c("Case series" = fill_cols[1],
-#                  "Competing event" = fill_cols[3],
-#                  "Base series" = fill_cols[2]))
-# 
-# color_cols <- colorspace::darken(col = fill_cols, amount = 0.3)
-# 
-# (color_colors <- c("Case series" = color_cols[1],
-#                    "Competing event" = color_cols[3],
-#                    "Base series" = color_cols[2]))
-# 
-# pt_object <- casebase::popTime(bmtcrr, event = "Status", time = "ftime")
-# 
-# pt_object_comp <- bmtcrr %>%
-#   mutate(Status = case_when(
-#     Status == 1 ~ 2L,
-#     Status == 2 ~ 1L,
-#     TRUE ~ Status
-#   )) %>%
-#   casebase::popTime(event = "Status", time = "ftime")
-# 
-# plot_grid(plot(pt_object) + ggtitle("Relapse"),
-#           plot(pt_object_comp, point.colour = "blue") + ggtitle("Competing events"),
-#           labels = c("A", "B"))
-
 popTimeData <- popTime(data = bmtcrr, time = "ftime")
@@
      comprisk = TRUE,
-     # add.base.series = TRUE,
-     # ratio = 0.7,
-     # facet.params = list(ncol = 2),
      ribbon.params = list(fill = "gray50"),
@@
   scale_y_continuous(labels = label_number_si())
-
-# plot(popTimeData,
-#      add.case.series = TRUE,
-#      # add.base.series = TRUE,
-#      add.competing.event = TRUE,
-#      comprisk = TRUE)
-
 ```
@@
 
-We will compare our hazard ratio estimates to that obtained from a Cox regression. To do so, we need to treat the competing event as censoring.
+We want to compare our hazard ratio estimates to that obtained from a Cox regression. To obtain these estimates, we need to treat the competing event as censoring:
 
@@
 model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-  data = bmtcrr
-)
+                   data = bmtcrr)
 ```
@@
 ```{r bmtcrr-cis, echo=FALSE, eval = eval_cs2}
-# Table of coefficients----
+# Table of coefficients
 library(glue)
@@
 We perform this computation for a 35 year old woman who received a stem-cell transplant from peripheral blood at relapse.
-We compared the absolute risk curve for such a woman with ALL with that for a similar woman with AML.
-We will estimate the curve from 0 to 60 months.
+We compare the absolute risk curve for such a woman with ALL with that for a similar woman with AML.
 
@@
 
-We will compare our estimates to that obtained from a corresponding Fine-Gray model [-@fine1999proportional].
+Next, we compare our estimates to that obtained from a corresponding Fine-Gray model [-@fine1999proportional].
 The Fine-Gray model is a semiparametric model for the cause-specific *subdistribution hazard*, i.e.
-the function $f_k(t)$ such that
-$$CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),$$
+the function $d_k(t)$ such that
+$$CI_k(t) =1 - \exp\left( - \int_0^t d_k(u) \textrm{d}u \right),$$
 where $CI_k(t)$ is the cause-specific cumulative incidence.
-The Fine-Gray model allows to directly assess the effect of a covariate on the subdistribution hazard, as opposed to the hazard.
-For the computation, we will use the \pkg{timereg} package [@timereg]:
+The Fine-Gray model allows to directly assess the effect of a covariate on the subdistribution hazard, as opposed to the cause-specific hazard.
+For the computation, we use the \pkg{timereg} package [@timereg]:
 
@@
 The cleaned and imputed data consists of 9104 observations and 30 variables, and it is available as part of the \pkg{casebase} package.
-In the comparisons below, all covariates except \code{sps} and \code{aps} were modeled.
+In the comparisons below, all covariates except \code{sps} and \code{aps} were used.
 These two variables correspond to scores for predicting the outcome that were developed as part of the original study.
@@
 
-For our penalized case-base model, we opted for the natural log of time which corresponds to a Weibull distribution.
+For our penalized case-base model, we opt for the natural log of time which corresponds to a Weibull distribution.
 For fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface to the \code{fitSmoothHazard} function.
-We must supply both a matrix \code{y} containing the time and event variables, and a matrix \code{x} containing all other covariates.
-We applied the lasso penalty by setting \code{alpha = 1} and assigned a \code{penalty.factor} ($w_j$) of 0 to the time variable to ensure it was in the selected model.
-We compared our approach to both Cox regression, and lasso penalized Cox regression (fitted via the \pkg{glmnet} package).
+We supply both a matrix \code{y} containing the time and event variables, and a matrix \code{x} containing all other covariates.
+We apply the lasso penalty by setting \code{alpha = 1} and assign a \code{penalty.factor} ($w_j$; cf.
+Equation \ref{eq:penest}) of 0 to the time variable to ensure it is in the selected model.
+We compare our approach to both Cox regression, and lasso penalized Cox regression (fitted via the \pkg{glmnet} package).
 
 To compare the performance of our models, we split the data into 95% training and 5% test sets.
-To assess both discrimination and calibration, we used a time-dependent version of the classical Brier score that is adjusted for censoring [@graf1999ass].
+To assess both discrimination and calibration, we use a time-dependent version of the classical Brier score that is adjusted for censoring [@graf1999ass].
 The Brier score can be used with both parametric and semi-parametric models.
-We used the \pkg{riskRegression} package to compute these scores for all models.
+We use the \pkg{riskRegression} package to compute these scores for all models.
 
@@
 ```
-
-<!-- Next, we can use the object \code{cbFull} to estimate the absolute risk curve on the test data.
-We first create a matrix \code{newx} containing the relevant variables.
-Next, using \code{times}, we specify the time points at which the cumulative incidence will be computed.
-Finally, using the parameter \code{s = "lambda.1se"}, we specify at which value of the tuning parameter we want to retrieve the coefficient estimates.
--->
 
@@
 In Figure \ref{fig:cs3lolliPlot}, we show the coefficient estimates for covariates that we selected by both penalized Cox and penalized case-base.
-We note that both penalized approaches produce similar results, and can clearly see the shrinkage effect owing to the $\ell_1$ penalty.
+We note that both penalized approaches produce similar results.
+We can also clearly see the shrinkage effect owing to the $\ell_1$ penalty.
 
@@
                                  keep.rownames = TRUE)
-# estimatescb$std.error<-0
 estimatescb$Model <- "Pen. CB"
@@
                                   keep.rownames = TRUE)
-# estimatescox$std.error<-0
 cox$coefficients[which(is.na(cox$coefficients))] <- 0
@@
   paper_gg_theme +  
-  # ggtitle("Coefficient plot") +
   labs(color = "Models") +
@@
 
-We then compared the cumulative incidence estimation over the test set.
+We then compare the cumulative incidence estimation over the test set.
 The probabilities over time for each observation are averaged, resulting in the absolute risk curves shown in Figure \ref{fig:cs3FinalBrier} (A).
@@
 )
-# plot(stanford_popTime,
-#      ribbon.params = list(fill = "gray50")) + 
-#   paper_gg_theme
 
@@
 Since the exposure is time-dependent, we need to manually define the exposure variable *after* case-base sampling and *before* fitting the hazard function.
-For this reason, we will use the `sampleCaseBase` function directly.
+For this reason, we use the `sampleCaseBase` function directly.
 
@@
 
-Next, we will compute the number of days from acceptance into the program to transplant, and we use this variable to determine whether each population-moment is exposed or not.
+Next, we compute the number of days from acceptance into the program to transplant, and we use this variable to determine whether each population-moment is exposed or not.
 
@@
 
-# Environment Details {-}
-
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
+<!-- # Environment Details {-} -->
+
+<!-- This report was generated on `r Sys.time()` using the following computational environment and dependencies:  -->
 
@@
 
-The current Git commit details are:
+<!-- The current Git commit details are: -->
 ```{r, eval = eval_colophon}

### 2020-09-22 analysis/paper/paper.tex
final version

@@
 sampling (including Cox regression) eliminate the baseline hazard from
-the likelihood expression or the estimating function and therefore needs
-to be estimated separately using a non-parametric approach. This leads
-to step-wise estimates of the cumulative incidence that are difficult to
-interpret. Using case-base sampling, Hanley \& Miettinen (2009)
-explained how the parametric hazard functions can be estimated using
-logistic regression. Their approach naturally leads to estimates of the
-cumulative incidence that are smooth-in-time.\\
+the likelihood expression or the estimating function. The baseline
+hazard then needs to be estimated separately using a non-parametric
+approach. This leads to step-wise estimates of the cumulative incidence
+that are difficult to interpret. Using case-base sampling, Hanley \&
+Miettinen (2009) explained how the parametric hazard functions can be
+estimated using logistic regression. Their approach naturally leads to
+estimates of the cumulative incidence that are smooth-in-time.\\
 In this paper, we present the \textbf{casebase} R package, a
@@
 
-\Keywords{survival analysis, absolute risk, \proglang{R}, data visualization}
+\Keywords{survival analysis, absolute risk, \proglang{R}, data
+visualization}
 \Plainkeywords{survival analysis, absolute risk, R, data visualization}
@@
 
-Using the framework of logistic regression for analyzing time-to-event
-data opens the door to an extensive array of powerful modeling tools.
-Indeed, lasso and elastic-net regression can be used to select variables
-that are associated with the hazard function. Logistic regression can
-also be replaced by multinomial regression to analyse data with multiple
-competing events.
+Logistic regression had already been used in the context of
+discrete-time survival modeling \citep{cox1972regression}. But in the
+context of continuous-time survival modeling, using the framework of
+logistic regression opens the door to an extensive array of powerful
+modeling tools. Indeed, lasso and elastic-net regression can be used to
+select variables that are associated with the hazard function. Logistic
+regression can also be replaced by multinomial regression to analyse
+data with multiple competing events.
 
@@
 
+\hypertarget{likelihood-and-estimating-function}{%
+\subsection{Likelihood and estimating
+function}\label{likelihood-and-estimating-function}}
+
 To describe the theoretical foundations of case-base sampling, we use
@@
 Aalen \emph{et al} \citeyearpar{aalen2008survival}, we use the
-placeholder \(\mathrm{past}\) to denote the past history of the
-corresponding process. The reader interested in more details can refer
-to Saarela \& Arjas \citeyearpar{saarela2015non} and Saarela
+placeholder ``past'' to denote the past history of the corresponding
+process. The reader interested in more details can refer to Saarela \&
+Arjas \citeyearpar{saarela2015non} and Saarela
 \citeyearpar{saarela2016case}. First, let \(N_{i}(t) \in \{0, 1\}\) be
@@
 sampling mechanism are proportional to the cardiovascular disease event
-rate given by the Framingham score. With this sampling mechanism,
-Saarela \& Arjas are able to increase the efficiency of their
-estimators, when compared to uniform sampling.
+rate given by the Framingham score. Non-uniform sampling mechanisms can
+increase the efficiency of the case-base estimators.
+
+\hypertarget{common-parametric-models}{%
+\subsection{Common parametric models}\label{common-parametric-models}}
 
@@
 
+\hypertarget{competing-risk-analysis}{%
+\subsection{Competing-risk analysis}\label{competing-risk-analysis}}
+
 Case-base sampling can also be used in the context of competing-risk
@@
 
+\hypertarget{variable-selection}{%
+\subsection{Variable selection}\label{variable-selection}}
+
 To perform variable selection on the regression parameters
 \(\theta \in \mathbb{R}^p\) of the hazard function, we can add a penalty
-to the likelihood and solve the following problem: \begin{equation}
+to the likelihood and optimise the following equation: \begin{equation}
 \min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j P(\theta_j;\lambda,\alpha) \label{eq:penest}
-\end{equation} where \(\ell\left(\theta\right)\) is the log of the
-likelihood function given in \eqref{eq:lik-function},
+\end{equation} where \(\ell\left(\theta\right) = \log L(\theta)\) is the
+log of the likelihood function given in \eqref{eq:lik-function},
 \(P(\theta_j;\lambda,\alpha)\) is a penalty term controlled by the
@@
 survival analysis reveals that there are over 250 packages related to
-survival analysis \citeyearpar{survTaskView}. For the purposes of this
+survival analysis \citep{survTaskView}. For the purposes of this
 article, we restricted our review to packages that implement at least
@@
 \caption[Population time plots for both treatment arms in the ERSPC dataset]{Population time plots for both treatment arms in the ERSPC dataset.
-The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely small rectangles (person-moments).
-The points shown in the figure are select person-moments from the case and base series.
-We see that more events are observed at later follow-up times, motivating the use of a non-constant in time hazard model.}\label{fig:plot-stratified-erspc-data}
+The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely thin rectangles (person-moments).
+More events are observed at later follow-up times, motivating the use of non-constant hazard models.}\label{fig:plot-stratified-erspc-data}
 \end{figure}
@@
 other causes). Since the Swedish and Belgian centres were the last to
-complete recruitment--in December 2003--the minimum potential follow-up
+complete recruitment in December 2003, the minimum potential follow-up
 is three years. Tracked further forwards in time (i.e.~after year 3) the
@@
 S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right).
-\end{equation} The integral is computed using either the
-\code{stats::integrate} function or Monte-Carlo integration. The risk
-function (or cumulative incidence function) is then defined as
-\begin{equation}\label{eqn:CI}
+\end{equation} The integral is computed using either the numerical or
+Monte-Carlo integration. The risk function (or cumulative incidence
+function) is then defined as \begin{equation}\label{eqn:CI}
 CI(t) = 1 - S(t).
@@
 
-\[ CI_j(t) = \int_0^t f_j(u)du.\] The integrals are computed using
+\[ CI_j(t) = \int_0^t f_j(u)du.\] Again, he integrals are computed using
 either numerical integration (via the trapezoidal rule) or Monte Carlo
@@
 
-\subsection{Different functional forms of time}
+\hypertarget{different-functional-forms-of-time}{%
+\subsection{Different functional forms of
+time}\label{different-functional-forms-of-time}}
 
@@
 
-\subsection{Time-dependent hazard ratios}
+\hypertarget{time-dependent-hazard-ratios}{%
+\subsection{Time-dependent hazard
+ratios}\label{time-dependent-hazard-ratios}}
 
@@
 \begin{CodeInput}
-R> fit_inter <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) * ScrArm, data = ERSPC)
-\end{CodeInput}
-\end{CodeChunk}
-
-In Figure \ref{fig:interaction-ERSPC} we plot the estimated death from
-prostate cancer hazard ratio and 95\% confidence interval for screening
-vs.~control group as a function of time using the \code{plot} method for
-objects of class \code{singleEventCB}:
+R> fit_inter <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) * ScrArm, 
++                              data = ERSPC)
+\end{CodeInput}
+\end{CodeChunk}
+
+In Figure \ref{fig:interaction-ERSPC}, we have the estimated hazard
+ratio and 95\% confidence interval for screening vs.~control group as a
+function of time using the \code{plot} method for objects of class
+\code{singleEventCB}:
 
@@
 
-\caption[Estimated death from prostate cancer hazard ratio and 95\% confidence interval for screening vs]{Estimated death from prostate cancer hazard ratio and 95\% confidence interval for screening vs.
+\caption[Estimated hazard ratio and 95\% confidence interval for screening vs]{Estimated hazard ratio and 95\% confidence interval for screening vs.
 control group as a function of time in the ERSPC dataset.
@@
 95\% confidence intervals are calculated using the delta method.
-The plot shows that the cures attributable to the screening only begin to become statistically apparent by year 7 and later.
+The plot shows that the effect of screening only begins to become statistically apparent by year 7.
 The 25-60\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\% reported in the original article.}\label{fig:interaction-ERSPC}
@@
 
-\newpage
-
 \hypertarget{case-study-2bone-marrow-transplant}{%
@@
 In the next case study, we show how case-base sampling can be used in
-the context of a competing risk analysis. For illustrative purposes, we
-will use the same data that was used in Scrucca \emph{et al}
+the context of a competing-risk analysis. For illustrative purposes, we
+use the same data that was used in Scrucca \emph{et al}
 \citeyearpar{scrucca2010regression}. The data was downloaded from the
@@
 
-First, we can look at a population-time plot to visualize the incidence
+First, we look at a population-time plot to visualize the incidence
 density of both relapse and the competing events. In Figure
@@
 
-We will compare our hazard ratio estimates to that obtained from a Cox
-regression. To do so, we need to treat the competing event as censoring.
+We want to compare our hazard ratio estimates to that obtained from a
+Cox regression. To obtain these estimates, we need to treat the
+competing event as censoring:
 
@@
 R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-+   data = bmtcrr
-+ )
++                    data = bmtcrr)
 \end{CodeInput}
@@
 received a stem-cell transplant from peripheral blood at relapse. We
-compared the absolute risk curve for such a woman with ALL with that for
-a similar woman with AML. We will estimate the curve from 0 to 60
-months.
-
-We will compare our estimates to that obtained from a corresponding
+compare the absolute risk curve for such a woman with ALL with that for
+a similar woman with AML.
+
+Next, we compare our estimates to that obtained from a corresponding
 Fine-Gray model \citeyearpar{fine1999proportional}. The Fine-Gray model
 is a semiparametric model for the cause-specific \emph{subdistribution
-hazard}, i.e.~the function \(f_k(t)\) such that
-\[CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),\] where
+hazard}, i.e.~the function \(d_k(t)\) such that
+\[CI_k(t) =1 - \exp\left( - \int_0^t d_k(u) \textrm{d}u \right),\] where
 \(CI_k(t)\) is the cause-specific cumulative incidence. The Fine-Gray
 model allows to directly assess the effect of a covariate on the
-subdistribution hazard, as opposed to the hazard. For the computation,
-we will use the \pkg{timereg} package \citep{timereg}:
+subdistribution hazard, as opposed to the cause-specific hazard. For the
+computation, we use the \pkg{timereg} package \citep{timereg}:
 
@@
 is available as part of the \pkg{casebase} package. In the comparisons
-below, all covariates except \code{sps} and \code{aps} were modeled.
-These two variables correspond to scores for predicting the outcome that
-were developed as part of the original study. For more information about
-this dataset, the reader is encouraged to look at the documentation in
-our package.
-
-For our penalized case-base model, we opted for the natural log of time
+below, all covariates except \code{sps} and \code{aps} were used. These
+two variables correspond to scores for predicting the outcome that were
+developed as part of the original study. For more information about this
+dataset, the reader is encouraged to look at the documentation in our
+package.
+
+For our penalized case-base model, we opt for the natural log of time
 which corresponds to a Weibull distribution. For fitting the penalized
 hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface
-to the \code{fitSmoothHazard} function. We must supply both a matrix
-\code{y} containing the time and event variables, and a matrix \code{x}
-containing all other covariates. We applied the lasso penalty by setting
-\code{alpha = 1} and assigned a \code{penalty.factor} (\(w_j\)) of 0 to
-the time variable to ensure it was in the selected model. We compared
-our approach to both Cox regression, and lasso penalized Cox regression
-(fitted via the \pkg{glmnet} package).
+to the \code{fitSmoothHazard} function. We supply both a matrix \code{y}
+containing the time and event variables, and a matrix \code{x}
+containing all other covariates. We apply the lasso penalty by setting
+\code{alpha = 1} and assign a \code{penalty.factor} (\(w_j\);
+cf.~Equation \ref{eq:penest}) of 0 to the time variable to ensure it is
+in the selected model. We compare our approach to both Cox regression,
+and lasso penalized Cox regression (fitted via the \pkg{glmnet}
+package).
 
@@
 training and 5\% test sets. To assess both discrimination and
-calibration, we used a time-dependent version of the classical Brier
+calibration, we use a time-dependent version of the classical Brier
 score that is adjusted for censoring \citep{graf1999ass}. The Brier
 score can be used with both parametric and semi-parametric models. We
-used the \pkg{riskRegression} package to compute these scores for all
+use the \pkg{riskRegression} package to compute these scores for all
 models.
@@
 case-base. We note that both penalized approaches produce similar
-results, and can clearly see the shrinkage effect owing to the
+results. We can also clearly see the shrinkage effect owing to the
 \(\ell_1\) penalty.
@@
 
-We then compared the cumulative incidence estimation over the test set.
+We then compare the cumulative incidence estimation over the test set.
 The probabilities over time for each observation are averaged, resulting
@@
 exposure variable \emph{after} case-base sampling and \emph{before}
-fitting the hazard function. For this reason, we will use the
+fitting the hazard function. For this reason, we use the
 \texttt{sampleCaseBase} function directly.
@@
 
-Next, we will compute the number of days from acceptance into the
-program to transplant, and we use this variable to determine whether
-each population-moment is exposed or not.
+Next, we compute the number of days from acceptance into the program to
+transplant, and we use this variable to determine whether each
+population-moment is exposed or not.
 
@@
 
-\hypertarget{environment-details}{%
-\section*{Environment Details}\label{environment-details}}
-\addcontentsline{toc}{section}{Environment Details}
-
-This report was generated on 2020-09-21 14:06:21 using the following
-computational environment and dependencies:
-
-\begin{CodeChunk}
-
-\begin{CodeOutput}
-#> R version 4.0.2 (2020-06-22)
-#> Platform: x86_64-pc-linux-gnu (64-bit)
-#> Running under: Pop!_OS 20.04 LTS
-#> 
-#> Matrix products: default
-#> BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3
-#> LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/liblapack.so.3
-#> 
-#> locale:
-#>  [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
-#>  [3] LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
-#>  [5] LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
-#>  [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                 
-#>  [9] LC_ADDRESS=C               LC_TELEPHONE=C            
-#> [11] LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       
-#> 
-#> attached base packages:
-#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
-#> [8] base     
-#> 
-#> other attached packages:
-#>  [1] scales_1.1.1              forcats_0.5.0            
-#>  [3] stringr_1.4.0             dplyr_1.0.2              
-#>  [5] purrr_0.3.4               readr_1.3.1              
-#>  [7] tidyr_1.1.2               ggplot2_3.3.2            
-#>  [9] tidyverse_1.3.0           lubridate_1.7.9          
-#> [11] visreg_2.7.0              tibble_3.0.3             
-#> [13] riskRegression_2020.02.05 prodlim_2019.11.13       
-#> [15] flexsurv_1.1.1            survival_3.2-3           
-#> [17] pracma_2.2.9              magrittr_1.5             
-#> [19] kableExtra_1.2.1          glmnet_4.0-2             
-#> [21] Matrix_1.2-18             data.table_1.13.0        
-#> [23] cowplot_1.1.0             colorspace_1.4-1         
-#> [25] casebase_0.9.0           
-#> 
-#> loaded via a namespace (and not attached):
-#>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
-#>  [4] htmlTable_2.1.0     base64enc_0.1-3     fs_1.5.0           
-#>  [7] rstudioapi_0.11     MatrixModels_0.4-1  fansi_0.4.1        
-#> [10] mvtnorm_1.1-1       xml2_1.3.2          codetools_0.2-16   
-#> [13] knitr_1.29          Formula_1.2-3       jsonlite_1.7.1     
-#> [16] broom_0.7.0         cluster_2.1.0       dbplyr_1.4.4       
-#> [19] png_0.1-7           compiler_4.0.2      httr_1.4.2         
-#> [22] backports_1.1.9     assertthat_0.2.1    cli_2.0.2          
-#> [25] htmltools_0.5.0     quantreg_5.67       tools_4.0.2        
-#> [28] gtable_0.3.0        glue_1.4.2          Rcpp_1.0.5         
-#> [31] cellranger_1.1.0    vctrs_0.3.4         nlme_3.1-149       
-#> [34] conquer_1.0.2       iterators_1.0.12    xfun_0.17          
-#> [37] rvest_0.3.6         lifecycle_0.2.0     polspline_1.1.19   
-#> [40] muhaz_1.2.6.1       MASS_7.3-53         zoo_1.8-8          
-#> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
-#> [46] RColorBrewer_1.1-2  rticles_0.15        yaml_2.2.1         
-#> [49] gridExtra_2.3       rms_6.0-1           rpart_4.1-15       
-#> [52] latticeExtra_0.6-29 stringi_1.5.3       foreach_1.5.0      
-#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.5        
-#> [58] rlang_0.4.7         pkgconfig_2.0.3     matrixStats_0.56.0 
-#> [61] evaluate_0.14       lattice_0.20-41     htmlwidgets_1.5.1  
-#> [64] cmprsk_2.2-10       tidyselect_1.1.0    deSolve_1.28       
-#> [67] R6_2.4.1            generics_0.0.2      Hmisc_4.4-1        
-#> [70] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
-#> [73] pillar_1.4.6        haven_2.3.1         foreign_0.8-79     
-#> [76] mgcv_1.8-33         nnet_7.3-14         mstate_0.2.12      
-#> [79] modelr_0.1.8        crayon_1.3.4        rmarkdown_2.3      
-#> [82] timereg_1.9.7       jpeg_0.1-8.1        grid_4.0.2         
-#> [85] readxl_1.3.1        blob_1.2.1          reprex_0.3.0       
-#> [88] digest_0.6.25       webshot_0.5.2       numDeriv_2016.8-1.1
-#> [91] stats4_4.0.2        munsell_0.5.0       viridisLite_0.3.0  
-#> [94] quadprog_1.5-8
-\end{CodeOutput}
-\end{CodeChunk}
-
-The current Git commit details are:
-
-\begin{CodeChunk}
-
-\begin{CodeOutput}
-#> Local:    editSahir /home/sahir/git_repositories/cbpaper
-#> Remote:   editSahir @ origin (https://github.com/sahirbhatnagar/cbpaper.git)
-#> Head:     [dace5dd] 2020-09-20: added paragraph on variable selection in theoretical details
-\end{CodeOutput}
-\end{CodeChunk}
-
 \bibliography{references.bib}

### 2020-09-22 analysis/paper/paper.Rmd
arxiv needs source files. this branch is only for arxiv submission

@@
   comment = "#>",
-  fig.path = "../figures/",
+  # fig.path = "../figures/",
   out.width = "\\textwidth",

### 2020-09-22 analysis/paper/paper.tex
arxiv needs source files. this branch is only for arxiv submission

@@
 
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/plot-stratified-erspc-data-1}
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{plot-stratified-erspc-data-1}
 
@@
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/erspc-cox-cif-1} \caption{CIFs for control and screening groups in the ERSPC data.
+\includegraphics[width=\textwidth,keepaspectratio=true]{erspc-cox-cif-1} \caption{CIFs for control and screening groups in the ERSPC data.
 In each of the panels, we plot the CIF from the Cox model using \code{survival::survfit} (solid line) and the CIF from the case-base sampling scheme (dashed line) with different functional forms of time.
@@
 
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/interaction-ERSPC-1}
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{interaction-ERSPC-1}
 
@@
 
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/compPop-1}
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{compPop-1}
 
@@
 
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/bmtcrr-risk-1}
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{bmtcrr-risk-1}
 
@@
 
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/coefplots-1}
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{coefplots-1}
 
@@
 
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/riskregressionBrier-1}
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{riskregressionBrier-1}
 
@@
 
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-poptime-1}
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{stanford-poptime-1}
 
@@
 
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-hazard-1}
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{stanford-hazard-1}
 
@@
 
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-risk-1}
+{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{stanford-risk-1}
 

### 2020-11-06 analysis/paper/casebase-manuscript.Rmd
switch from JSS template to R Journal

--- 2020-09-22 analysis/paper/paper.Rmd
@@
 ---
-documentclass: jss
 author:
@@
   - name: Maxime Turgeon*
-    affiliation: 'University of Manitoba \AND'
+    affiliation: 'University of Manitoba'
     address: >
@@
     url: http://individual.utoronto.ca/osaarela/
-title:
-  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
-  # If you use tex in the formatted title, also supply version without
-  plain:     "casebase: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
-  # For running headers, if needed
-  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
+title: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
 abstract: |
@@
   | *SRB and MT contributed equally to this work.
-keywords:
-  # at least one keyword must be supplied
-  formatted: [survival analysis, absolute risk, "\\proglang{R}", data visualization]
-  plain:     [survival analysis, absolute risk, R, data visualization]
 preamble: >
@@
   \DeclareUnicodeCharacter{2500}{-}
-
-output: 
-  rticles::jss_article:
-    number_sections: TRUE     #added argument option 
-    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
-biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
-bibliography: references.bib  #Also not included in template by default. 
+  \newcommand{\proglang}{\textsf}
+output: rticles::rjournal_article 
+bibliography: casebase-manuscript.bib 
 fig_width: 12

### 2020-11-06 analysis/paper/casebase-manuscript.tex
switch from JSS template to R Journal

--- 2020-09-22 analysis/paper/paper.tex
@@
-\documentclass[
-]{jss}
-
-\usepackage[utf8]{inputenc}
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-\author{
-Sahir Rai Bhatnagar*\\McGill University \And Maxime Turgeon*\\University
-of Manitoba \AND Jesse Islam\\McGill University \And James A.
-Hanley\\McGill University \And Olli Saarela\\University of Toronto
-}
+% !TeX root = RJwrapper.tex
 \title{\pkg{casebase}: An Alternative Framework For Survival Analysis
 and Comparison of Event Rates}
-
-\Plainauthor{Sahir Rai Bhatnagar*, Maxime Turgeon*, Jesse Islam, James
-A. Hanley, Olli Saarela}
-\Plaintitle{casebase: An Alternative Framework For Survival Analysis and
-Comparison of Event Rates}
-\Shorttitle{\pkg{casebase}: An Alternative Framework For Survival
-Analysis}
-
-\Abstract{
+\author{by Sahir Rai Bhatnagar*, Maxime Turgeon*, Jesse Islam, James A.
+Hanley, and Olli Saarela}
+
+\maketitle
+
+\abstract{%
 In epidemiological studies of time-to-event data, a quantity of interest
@@
 
-\Keywords{survival analysis, absolute risk, \proglang{R}, data
-visualization}
-\Plainkeywords{survival analysis, absolute risk, R, data visualization}
-
-%% publication information
-%% \Volume{50}
-%% \Issue{9}
-%% \Month{June}
-%% \Year{2012}
-%% \Submitdate{}
-%% \Acceptdate{2012-06-04}
-
-\Address{
-    Sahir Rai Bhatnagar*\\
-    McGill University\\
-    1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
-  E-mail: \email{sahir.bhatnagar@mail.mcgill.ca}\\
-  URL: \url{http://sahirbhatnagar.com/}\\~\\
-      Maxime Turgeon*\\
-    University of Manitoba\\
-    186 Dysart Road Winnipeg, MB, Canada R3T 2N2\\
-  E-mail: \email{max.turgeon@umanitoba.ca}\\
-  URL: \url{https://maxturgeon.ca/}\\~\\
-      Jesse Islam\\
-    McGill University\\
-    1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
-  E-mail: \email{jesse.islam@mail.mcgill.ca}\\
-
-      James A. Hanley\\
-    McGill University\\
-    1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
-  E-mail: \email{james.hanley@mcgill.ca}\\
-  URL: \url{http://www.medicine.mcgill.ca/epidemiology/hanley/}\\~\\
-      Olli Saarela\\
-    University of Toronto\\
-    Dalla Lana School of Public Health, 155 College Street, 6th floor,
-Toronto, Ontario M5T 3M7, Canada\\
-  E-mail: \email{olli.saarela@utoronto.ca}\\
-  URL: \url{http://individual.utoronto.ca/osaarela/}\\~\\
-  }
-
-% Pandoc header
-
-\usepackage{amsmath} \usepackage{amssymb} \usepackage{longtable} \usepackage{graphicx} \usepackage{tabularx} \usepackage{float} \usepackage{booktabs} \usepackage{makecell} \usepackage{tabu} \DeclareUnicodeCharacter{2500}{-}
-
-\begin{document}
-
 \hypertarget{introduction}{%
@@
 In this article, we present the \pkg{casebase} package
-\citep{casebase-package} implemented in \proglang{R} \citep{r-core}, for
+\citep{casebase-package} implemented in \textsf{R} \citep{r-core}, for
 parametric survival analysis that combines the ideas of Hanley \&
@@
 sampling and its use for estimating parametric hazard functions. We then
-give a short review of existing \proglang{R} packages that implement
+give a short review of existing \textsf{R} packages that implement
 comparable features as \pkg{casebase}. Next, we provide some details
@@
 epidemiology. Accordingly, there is already a vast ecosystem of
-\proglang{R} packages implementing different methodologies. In this
+\textsf{R} packages implementing different methodologies. In this
 section, we describe how the functionalities of \pkg{casebase} compare
@@
 
-Next, several \proglang{R} packages implement methodologies for
-competing risk analysis; for a different perspective on this topic, see
-Mahani \& Sharabiani \citeyearpar{mahani2015bayesian}. The package
-\pkg{cmprsk} provides methods for cause-specific subdistribution
-hazards, such as in the Fine-Gray model
-\citeyearpar{fine1999proportional}. On the other hand, the package
-\pkg{CFC} estimates cause-specific CIFs from unadjusted, non-parametric
-survival functions. Our package \pkg{casebase} also provides
-functionalities for competing risk analysis by estimating parametrically
-the cause-specific hazards. From these quantities, we can then estimate
-the cause-specific CIFs.
+Next, several \textsf{R} packages implement methodologies for competing
+risk analysis; for a different perspective on this topic, see Mahani \&
+Sharabiani \citeyearpar{mahani2015bayesian}. The package \pkg{cmprsk}
+provides methods for cause-specific subdistribution hazards, such as in
+the Fine-Gray model \citeyearpar{fine1999proportional}. On the other
+hand, the package \pkg{CFC} estimates cause-specific CIFs from
+unadjusted, non-parametric survival functions. Our package
+\pkg{casebase} also provides functionalities for competing risk analysis
+by estimating parametrically the cause-specific hazards. From these
+quantities, we can then estimate the cause-specific CIFs.
 
@@
 }
-\caption{Comparison of various \proglang{R} packages for survival analysis based on several defining features.
+\caption{Comparison of various \textsf{R} packages for survival analysis based on several defining features.
 \textbf{Competing Risks}: whether or not an implementation for competing risks is present.
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
-+                                event = "DeadOfPrCa", exposure = "ScrArm")
-R> inherits(pt_object, "popTime")
-\end{CodeInput}
-
-\begin{CodeOutput}
+\begin{Schunk}
+\begin{Sinput}
+pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
+                               event = "DeadOfPrCa", exposure = "ScrArm")
+inherits(pt_object, "popTime")
+\end{Sinput}
+\begin{Soutput}
 #> [1] TRUE
-\end{CodeOutput}
-
-\begin{CodeInput}
-R> attr(pt_object, "exposure")
-\end{CodeInput}
-
-\begin{CodeOutput}
+\end{Soutput}
+\begin{Sinput}
+attr(pt_object, "exposure")
+\end{Sinput}
+\begin{Soutput}
 #> [1] "ScrArm"
-\end{CodeOutput}
-\end{CodeChunk}
+\end{Soutput}
+\end{Schunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> plot(pt_object, add.base.series = TRUE)
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
+\begin{Schunk}
+\begin{Sinput}
+plot(pt_object, add.base.series = TRUE)
+\end{Sinput}
+\end{Schunk}
+
+\begin{Schunk}
 \begin{figure}[ht]
-
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/plot-stratified-erspc-data-1}
-
-}
-
-\caption[Population time plots for both treatment arms in the ERSPC dataset]{Population time plots for both treatment arms in the ERSPC dataset.
+\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/plot-stratified-erspc-data-1} \caption[Population time plots for both treatment arms in the ERSPC dataset]{Population time plots for both treatment arms in the ERSPC dataset.
 The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely thin rectangles (person-moments).
@@
 \end{figure}
-\end{CodeChunk}
+\end{Schunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
-+              gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
-+              weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
-+              splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
-R> 
-R> fits <- lapply(fmla, function(i) {
-+   fitSmoothHazard(i, data = ERSPC, ratio = 100)
-+ })
-\end{CodeInput}
-\end{CodeChunk}
+\begin{Schunk}
+\begin{Sinput}
+fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
+             gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
+             weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
+             splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
+
+fits <- lapply(fmla, function(i) {
+  fitSmoothHazard(i, data = ERSPC, ratio = 100)
+})
+\end{Sinput}
+\end{Schunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> summary(fits[["splines"]]) 
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-
-\begin{CodeOutput}
+\begin{Schunk}
+\begin{Sinput}
+summary(fits[["splines"]]) 
+\end{Sinput}
+\end{Schunk}
+
+\begin{Schunk}
+\begin{Soutput}
 #> 
@@
 #> Number of Fisher Scoring iterations: 9
-\end{CodeOutput}
-\end{CodeChunk}
+\end{Soutput}
+\end{Schunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
-R> new_time <- seq(0,14,0.1)
-R> 
-R> risks <- lapply(fits, function(i) {
-+   absoluteRisk(object = i, time = new_time, newdata = new_data, method = "mont")
-+ })
-\end{CodeInput}
-\end{CodeChunk}
+\begin{Schunk}
+\begin{Sinput}
+new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
+new_time <- seq(0,14,0.1)
+
+risks <- lapply(fits, function(i) {
+  absoluteRisk(object = i, time = new_time, newdata = new_data, method = "mont")
+})
+\end{Sinput}
+\end{Schunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeOutput}
+\begin{Schunk}
+\begin{Soutput}
 #> Analysis of Deviance Table
@@
 #> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{CodeOutput}
-\end{CodeChunk}
+\end{Soutput}
+\end{Schunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeOutput}
+\begin{Schunk}
+\begin{Soutput}
 #>     Exp. Gompertz  Weibull  Splines 
 #>     6056     5821     5807     5795
-\end{CodeOutput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
+\end{Soutput}
+\end{Schunk}
+
+\begin{Schunk}
 \begin{figure}[ht]
@@
 
-\begin{CodeChunk}
+\begin{Schunk}
 \begin{table}
@@
 
-\end{CodeChunk}
+\end{Schunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> fit_inter <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) * ScrArm, 
-+                              data = ERSPC)
-\end{CodeInput}
-\end{CodeChunk}
+\begin{Schunk}
+\begin{Sinput}
+fit_inter <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) * ScrArm, 
+                             data = ERSPC)
+\end{Sinput}
+\end{Schunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> plot(fit_inter, type = "hr", newdata = new_data,
-+      var = "ScrArm", xvar = "Follow.Up.Time", ci = TRUE)
-\end{CodeInput}
-\end{CodeChunk}
+\begin{Schunk}
+\begin{Sinput}
+plot(fit_inter, type = "hr", newdata = new_data,
+     var = "ScrArm", xvar = "Follow.Up.Time", ci = TRUE)
+\end{Sinput}
+\end{Schunk}
 
@@
 
-\begin{CodeChunk}
+\begin{Schunk}
 \begin{figure}[ht]
-
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/interaction-ERSPC-1}
-
-}
-
-\caption[Estimated hazard ratio and 95\% confidence interval for screening vs]{Estimated hazard ratio and 95\% confidence interval for screening vs.
+\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/interaction-ERSPC-1} \caption[Estimated hazard ratio and 95\% confidence interval for screening vs]{Estimated hazard ratio and 95\% confidence interval for screening vs.
 control group as a function of time in the ERSPC dataset.
@@
 
-\begin{CodeChunk}
+\begin{Schunk}
 \begin{figure}[ht]
-
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/compPop-1}
-
-}
-
-\caption[Population-time plot for the stem-cell transplant study with both relapse and competing events]{Population-time plot for the stem-cell transplant study with both relapse and competing events.}\label{fig:compPop}
+\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/compPop-1} \caption[Population-time plot for the stem-cell transplant study with both relapse and competing events]{Population-time plot for the stem-cell transplant study with both relapse and competing events.}\label{fig:compPop}
 \end{figure}
-\end{CodeChunk}
+\end{Schunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(survival)
-R> # Treat competing event as censoring
-R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-+                    data = bmtcrr)
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
+\begin{Schunk}
+\begin{Sinput}
+library(survival)
+# Treat competing event as censoring
+model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
+                   data = bmtcrr)
+\end{Sinput}
+\end{Schunk}
+
+\begin{Schunk}
 \begin{table}
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> library(timereg)
-R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) +
-+                         const(Phase) + const(Source) + const(Age),
-+                       data = bmtcrr, cause = 1, model = "fg")
-R> 
-R> # Estimate absolute risk curve
-R> risk_fg <- predict(model_fg, newdata, times = time_points)
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
+\begin{Schunk}
+\begin{Sinput}
+library(timereg)
+model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) +
+                        const(Phase) + const(Source) + const(Age),
+                      data = bmtcrr, cause = 1, model = "fg")
+
+# Estimate absolute risk curve
+risk_fg <- predict(model_fg, newdata, times = time_points)
+\end{Sinput}
+\end{Schunk}
+
+\begin{Schunk}
 \begin{figure}[ht]
-
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/bmtcrr-risk-1}
-
-}
-
-\caption{\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups.
+\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/bmtcrr-risk-1} \caption{\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups.
 The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:bmtcrr-risk}
 \end{figure}
-\end{CodeChunk}
+\end{Schunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Create matrices for inputs
-R> x <- model.matrix(death ~ . - d.time - aps - sps, 
-+                   data = train)[, -c(1)] # Remove intercept
-R> y <- data.matrix(subset(train, select = c(d.time, death)))
-R> 
-R> # Regularized logistic regression to estimate hazard
-R> pen_cb <- casebase::fitSmoothHazard.fit(x, y,
-+   family = "glmnet",
-+   time = "d.time", event = "death",
-+   formula_time = ~ log(d.time), alpha = 1,
-+   ratio = 10, standardize = TRUE,
-+   penalty.factor = c(0, rep(1, ncol(x)))
-+ )
-\end{CodeInput}
-\end{CodeChunk}
+\begin{Schunk}
+\begin{Sinput}
+# Create matrices for inputs
+x <- model.matrix(death ~ . - d.time - aps - sps, 
+                  data = train)[, -c(1)] # Remove intercept
+y <- data.matrix(subset(train, select = c(d.time, death)))
+
+# Regularized logistic regression to estimate hazard
+pen_cb <- casebase::fitSmoothHazard.fit(x, y,
+  family = "glmnet",
+  time = "d.time", event = "death",
+  formula_time = ~ log(d.time), alpha = 1,
+  ratio = 10, standardize = TRUE,
+  penalty.factor = c(0, rep(1, ncol(x)))
+)
+\end{Sinput}
+\end{Schunk}
 
@@
 
-\begin{CodeChunk}
+\begin{Schunk}
 \begin{figure}[ht]
-
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/coefplots-1}
-
-}
-
-\caption{\label{fig:cs3lolliPlot} Coefficient estimates from the Cox model (Cox), penalized Cox model using the \pkg{glmnet} package (Pen.
+\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/coefplots-1} \caption{\label{fig:cs3lolliPlot} Coefficient estimates from the Cox model (Cox), penalized Cox model using the \pkg{glmnet} package (Pen.
 Cox), and our approach using penalized case-base sampling (Pen.
@@
 
-\begin{CodeChunk}
+\begin{Schunk}
 \begin{figure}[ht]
-
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/riskregressionBrier-1}
-
-}
-
-\caption{\label{fig:cs3FinalBrier} Comparison of Cox regression (Cox), penalized Cox regression (Pen.
+\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/riskregressionBrier-1} \caption{\label{fig:cs3FinalBrier} Comparison of Cox regression (Cox), penalized Cox regression (Pen.
 Cox), penalized case-base sampling estimation (Pen.
@@
 
-\begin{CodeChunk}
+\begin{Schunk}
 \begin{figure}[ht]
-
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-poptime-1}
-
-}
-
-\caption{\label{fig:cs4PopTime} Population-time for the Stanford Transplant study. The darker shade of gray corresponds to the exposed person-time.}\label{fig:stanford-poptime}
+\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-poptime-1} \caption{\label{fig:cs4PopTime} Population-time for the Stanford Transplant study.
+The darker shade of gray corresponds to the exposed person-time.}\label{fig:stanford-poptime}
 \end{figure}
-\end{CodeChunk}
+\end{Schunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> cb_data <- sampleCaseBase(jasa,
-+   time = "futime",
-+   event = "fustat", ratio = 100
-+ )
-\end{CodeInput}
-\end{CodeChunk}
+\begin{Schunk}
+\begin{Sinput}
+cb_data <- sampleCaseBase(jasa,
+  time = "futime",
+  event = "fustat", ratio = 100
+)
+\end{Sinput}
+\end{Schunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Define exposure variable
-R> cb_data <- mutate(cb_data,
-+   txtime = time_length(accept.dt %--% tx.date,
-+                        unit = "days"
-+   ),
-+   exposure = case_when(
-+     is.na(txtime) ~ 0L, # No transplant
-+     txtime > futime ~ 0L,
-+     txtime <= futime ~ 1L
-+   )
-+ )
-\end{CodeInput}
-\end{CodeChunk}
+\begin{Schunk}
+\begin{Sinput}
+# Define exposure variable
+cb_data <- mutate(cb_data,
+  txtime = time_length(accept.dt %--% tx.date,
+                       unit = "days"
+  ),
+  exposure = case_when(
+    is.na(txtime) ~ 0L, # No transplant
+    txtime > futime ~ 0L,
+    txtime <= futime ~ 1L
+  )
+)
+\end{Sinput}
+\end{Schunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Fit several models
-R> fit1 <- fitSmoothHazard(fustat ~ exposure,
-+   data = cb_data, time = "futime"
-+ )
-R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
-+   data = cb_data, time = "futime"
-+ )
-R> fit3 <- fitSmoothHazard(fustat ~ exposure * futime,
-+   data = cb_data, time = "futime"
-+ )
-\end{CodeInput}
-\end{CodeChunk}
+\begin{Schunk}
+\begin{Sinput}
+# Fit several models
+fit1 <- fitSmoothHazard(fustat ~ exposure,
+  data = cb_data, time = "futime"
+)
+fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
+  data = cb_data, time = "futime"
+)
+fit3 <- fitSmoothHazard(fustat ~ exposure * futime,
+  data = cb_data, time = "futime"
+)
+\end{Sinput}
+\end{Schunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeOutput}
+\begin{Schunk}
+\begin{Soutput}
 #> Model1 Model2 Model3 
 #>    827    791    790
-\end{CodeOutput}
-\end{CodeChunk}
+\end{Soutput}
+\end{Schunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> plot(fit3, hazard.params = list(xvar = "futime",
-+                                 by = "exposure",
-+                                 alpha = 0.05,
-+                                 ylab = "Hazard",
-+                                 data = cb_data))
-\end{CodeInput}
+\begin{Schunk}
+\begin{Sinput}
+plot(fit3, hazard.params = list(xvar = "futime",
+                                by = "exposure",
+                                alpha = 0.05,
+                                ylab = "Hazard",
+                                data = cb_data))
+\end{Sinput}
 \begin{figure}[ht]
-
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-hazard-1}
-
-}
-
-\caption{\label{fig:cs4hazardPlot} Plot of the hazard function for exposed and unexposed individuals.}\label{fig:stanford-hazard}
+\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-hazard-1} \caption{\label{fig:cs4hazardPlot} Plot of the hazard function for exposed and unexposed individuals.}\label{fig:stanford-hazard}
 \end{figure}
-\end{CodeChunk}
+\end{Schunk}
 
@@
 
-\begin{CodeChunk}
-
-\begin{CodeInput}
-R> # Compute absolute risk curves
-R> newdata <- data.frame(exposure = c(0, 1))
-R> absrisk <- absoluteRisk(fit3,
-+   newdata = newdata,
-+   time = seq(0, 425, length.out = 100)
-+ )
-R> 
-R> class(absrisk)
-\end{CodeInput}
-
-\begin{CodeOutput}
+\begin{Schunk}
+\begin{Sinput}
+# Compute absolute risk curves
+newdata <- data.frame(exposure = c(0, 1))
+absrisk <- absoluteRisk(fit3,
+  newdata = newdata,
+  time = seq(0, 425, length.out = 100)
+)
+
+class(absrisk)
+\end{Sinput}
+\begin{Soutput}
 #> [1] "absRiskCB" "matrix"    "array"
-\end{CodeOutput}
-
-\begin{CodeInput}
-R> 
-R> plot(absrisk,
-+      id.names = c("No Tx", "Tx")) + 
-+   ylab("Cumulative Incidence") +
-+   xlab("Follow-up time (in days)") +
-+   paper_gg_theme
-\end{CodeInput}
+\end{Soutput}
+\begin{Sinput}
+
+plot(absrisk,
+     id.names = c("No Tx", "Tx")) + 
+  ylab("Cumulative Incidence") +
+  xlab("Follow-up time (in days)") +
+  paper_gg_theme
+\end{Sinput}
 \begin{figure}[ht]
-
-{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-risk-1}
-
-}
-
-\caption{\label{fig:cs4CIF} Plot of the cumulative incidence function for exposed and unexposed individuals.}\label{fig:stanford-risk}
+\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-risk-1} \caption{\label{fig:cs4CIF} Plot of the cumulative incidence function for exposed and unexposed individuals.}\label{fig:stanford-risk}
 \end{figure}
-\end{CodeChunk}
+\end{Schunk}
 
@@
 
-In this article, we presented the \proglang{R} package \pkg{casebase},
+In this article, we presented the \textsf{R} package \pkg{casebase},
 which provides functions for fitting smooth parametric hazards and
@@
 
-In conclusion, we presented the \proglang{R} package \pkg{casebase}
-which implements case-base sampling for fitting parametric survival
-models and for estimating smooth cumulative incidence functions using
-the framework of generalized linear models. We strongly believe that its
-flexibility and its foundation on the familiar logistic regression model
-will make it appealing to new and established practitioners. The
-\pkg{casebase} package is freely available from the Comprehensive
-\proglang{R} Archive Network at
-\url{https://cran.r-project.org/package=casebase}. Interested users can
-visit \url{http://sahirbhatnagar.com/casebase/} for detailed package
-documentation and vignettes.
+In conclusion, we presented the \textsf{R} package \pkg{casebase} which
+implements case-base sampling for fitting parametric survival models and
+for estimating smooth cumulative incidence functions using the framework
+of generalized linear models. We strongly believe that its flexibility
+and its foundation on the familiar logistic regression model will make
+it appealing to new and established practitioners. The \pkg{casebase}
+package is freely available from the Comprehensive \textsf{R} Archive
+Network at \url{https://cran.r-project.org/package=casebase}. Interested
+users can visit \url{http://sahirbhatnagar.com/casebase/} for detailed
+package documentation and vignettes.
 
@@
 
-\bibliography{references.bib}
-
-\end{document}
+\bibliography{casebase-manuscript.bib}
+
+\address{%
+Sahir Rai Bhatnagar*\\
+McGill University\\%
+1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
+%
+\url{http://sahirbhatnagar.com/}%
+%
+\\\email{sahir.bhatnagar@mail.mcgill.ca}
+}
+
+\address{%
+Maxime Turgeon*\\
+University of Manitoba\\%
+186 Dysart Road Winnipeg, MB, Canada R3T 2N2\\
+%
+\url{https://maxturgeon.ca/}%
+%
+\\\email{max.turgeon@umanitoba.ca}
+}
+
+\address{%
+Jesse Islam\\
+McGill University\\%
+1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
+%
+%
+%
+\\\email{jesse.islam@mail.mcgill.ca}
+}
+
+\address{%
+James A. Hanley\\
+McGill University\\%
+1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
+%
+\url{http://www.medicine.mcgill.ca/epidemiology/hanley/}%
+%
+\\\email{james.hanley@mcgill.ca}
+}
+
+\address{%
+Olli Saarela\\
+University of Toronto\\%
+Dalla Lana School of Public Health, 155 College Street, 6th floor,
+Toronto, Ontario M5T 3M7, Canada\\
+%
+\url{http://individual.utoronto.ca/osaarela/}%
+%
+\\\email{olli.saarela@utoronto.ca}
+}

### 2020-11-06 analysis/paper/casebase-manuscript.R

@@ -1,0 +1,978 @@
+## ----setup, include=FALSE-----------------------------------------------------
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  warning = FALSE,
+  message = FALSE,
+  echo = FALSE,
+  cache = TRUE, 
+  fig.pos = 'ht',
+  comment = "#>",
+  fig.path = "../figures/",
+  out.width = "\\textwidth",
+   out.extra = "keepaspectratio=true"
+)
+
+options(kableExtra.latex.load_packages = FALSE)
+library(casebase)
+library(colorspace)
+library(cowplot)
+library(data.table)
+library(glmnet)
+library(kableExtra)
+library(magrittr)
+library(pracma)
+library(flexsurv) # for fitting gompertz AFT
+library(prodlim)
+library(riskRegression)
+library(splines)
+library(survival)
+library(tibble)
+library(visreg)
+library(lubridate)
+library(tidyverse)
+library(scales)
+options(digits = 2)
+# set the seed for reproducible output
+set.seed(1234)
+
+# Turn on/off sections for faster compilation
+eval_introduction <- TRUE
+eval_theory <- TRUE
+eval_implementation <- TRUE
+eval_cs1 <- TRUE
+eval_cs2 <- TRUE
+eval_cs3 <- TRUE
+eval_cs4 <- TRUE
+eval_colophon <- FALSE # Only for draft
+echo_plot_code <- TRUE
+# Define theme for plots
+paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
+# Define colour palette for CS 3
+q4 <- qualitative_hcl(4, palette = "Dark 3")
+names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
+
+
+## ----erspc-data, eval = eval_cs1, echo = c(-1:-3)-----------------------------
+data("ERSPC")
+ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
+                       levels = c(0,1), 
+                       labels = c("Control group", "Screening group"))
+
+pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
+                               event = "DeadOfPrCa", exposure = "ScrArm")
+inherits(pt_object, "popTime")
+attr(pt_object, "exposure")
+
+
+## ----plot-stratified-erspc-data-code, eval = FALSE, echo = TRUE---------------
+#> plot(pt_object, add.base.series = TRUE)
+
+
[…]

### 2021-02-03 analysis/paper/casebase-manuscript.Rmd
removed CS4

@@
 library(pracma)
-library(flexsurv) # for fitting gompertz AFT
+library(flexsurv) 
 library(prodlim)
@@
 eval_cs3 <- TRUE
-eval_cs4 <- TRUE
+eval_cs4 <- FALSE
 eval_colophon <- FALSE # Only for draft
@@
 
+<!--
 # Case study 4---Stanford Heart Transplant Data
@@
 As we can see, most events occur early during the follow-up period, and therefore we do not expect the hazard to be constant.
+
+-->
 
@@
 
+<!--
 Since the exposure is time-dependent, we need to manually define the exposure variable *after* case-base sampling and *before* fitting the hazard function.
 For this reason, we use the `sampleCaseBase` function directly.
-
-```{r stanford_cb, eval = eval_cs4, echo = TRUE}
+-->
+
+```{r stanford_cb, eval = eval_cs4}
 cb_data <- sampleCaseBase(jasa,
@@
 
+<!--
 Next, we compute the number of days from acceptance into the program to transplant, and we use this variable to determine whether each population-moment is exposed or not.
 
-```{r stanford_exp, eval = eval_cs4, dependson="stanford_cb", echo = TRUE}
+-->
+
+```{r stanford_exp, eval = eval_cs4, dependson="stanford_cb"}
 # Define exposure variable
@@
 
+<!--
 Finally, we can fit the hazard using various linear predictors.
-
-```{r stanford_fit, eval = eval_cs4, warning = FALSE, echo=TRUE, dependson="stanford_exp"}
+-->
+
+```{r stanford_fit, eval = eval_cs4, warning = FALSE, dependson="stanford_exp"}
 # Fit several models
@@
 
+<!--
 Note that the third model includes an interaction term between exposure and follow-up time.
@@
 We can then compare the goodness of fit of these three models using the Akaike Information Criterion (AIC).
+-->
 
@@
 
+<!--
 As we can see, the best fit is the third model.
@@
 We can easily obtain a plot of the hazards by using the `plot.singleEventCB` method:
-
-```{r stanford-hazard, eval = eval_cs4, echo=TRUE, dependson="stanford_fit", message = FALSE, results="hide", fig.cap="\\label{fig:cs4hazardPlot} Plot of the hazard function for exposed and unexposed individuals."}
+-->
+
+```{r stanford-hazard, eval = eval_cs4, dependson="stanford_fit", message = FALSE, results="hide", fig.cap="\\label{fig:cs4hazardPlot} Plot of the hazard function for exposed and unexposed individuals."}
 plot(fit3, hazard.params = list(xvar = "futime",
@@
 
+<!--
 Looking at Figure \ref{fig:cs4hazardPlot}, the non-proportionality seems to be more pronounced at the beginning of follow-up than the end.
 Finally, in Figure \ref{fig:cs4CIF}, we turn these estimates of the hazard function into estimates of the cumulative incidence functions.
-
-```{r stanford-risk, eval = eval_cs4, echo = TRUE, dependson="stanford_fit", fig.cap="\\label{fig:cs4CIF} Plot of the cumulative incidence function for exposed and unexposed individuals."}
+-->
+
+```{r stanford-risk, eval = eval_cs4, dependson="stanford_fit", fig.cap="\\label{fig:cs4CIF} Plot of the cumulative incidence function for exposed and unexposed individuals."}
 # Compute absolute risk curves
@@
 
+<!--
 As we can see in the above case-study, the \pkg{casebase} package can also be used to model time-varying exposures and non-proportional hazard functions. 
+-->
 

### 2021-02-03 analysis/paper/casebase-manuscript.tex
removed CS4

@@
 
-\hypertarget{case-study-4stanford-heart-transplant-data}{%
-\section{Case study 4---Stanford Heart Transplant
-Data}\label{case-study-4stanford-heart-transplant-data}}
-
-In the previous case studies, we only considered covariates that were
-fixed at baseline. In this next case study, we use the Stanford Heart
-Transplant data \citep[\citet{crowley1977covariance}]{clark1971cardiac}
-to show how case-base sampling can also be used in the context of
-time-dependent exposure. This feature of case-base sampling has been
-explored in the literature, in the context of vaccination safety
-\citep{saarela2015case}. In this study, the exposure period was defined
-as the week following vaccination. Hence, the main covariate of
-interest, i.e.~exposure to the vaccine, was changing over time. In this
-context, case-base sampling offers an efficient alternative to nested
-case-control designs or self-matching.
-
-Recall the setting of Stanford Heart Transplant study: patients were
-admitted to the Stanford program after meeting with their physician and
-determining that they were unlikely to respond to other forms of
-treatment. After enrollment, the program searched for a suitable donor
-for the patient, which could take anywhere between a few days to almost
-a year. We are interested in the effect of a heart transplant on
-survival; therefore, the patient is considered exposed only after the
-transplant has occurred.
-
-As above, we can look at the population-time plot for a graphical
-summary of the event incidence (see Figure \ref{fig:cs4PopTime}. Here,
-we colour the exposed person-time (i.e.~after transplant) in a darker
-shade of gray. As we can see, most events occur early during the
-follow-up period, and therefore we do not expect the hazard to be
-constant.
-
-\begin{Schunk}
-\begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-poptime-1} \caption{\label{fig:cs4PopTime} Population-time for the Stanford Transplant study.
-The darker shade of gray corresponds to the exposed person-time.}\label{fig:stanford-poptime}
-\end{figure}
-\end{Schunk}
-
-Since the exposure is time-dependent, we need to manually define the
-exposure variable \emph{after} case-base sampling and \emph{before}
-fitting the hazard function. For this reason, we use the
-\texttt{sampleCaseBase} function directly.
-
-\begin{Schunk}
-\begin{Sinput}
-cb_data <- sampleCaseBase(jasa,
-  time = "futime",
-  event = "fustat", ratio = 100
-)
-\end{Sinput}
-\end{Schunk}
-
-Next, we compute the number of days from acceptance into the program to
-transplant, and we use this variable to determine whether each
-population-moment is exposed or not.
-
-\begin{Schunk}
-\begin{Sinput}
-# Define exposure variable
-cb_data <- mutate(cb_data,
-  txtime = time_length(accept.dt %--% tx.date,
-                       unit = "days"
-  ),
-  exposure = case_when(
-    is.na(txtime) ~ 0L, # No transplant
-    txtime > futime ~ 0L,
-    txtime <= futime ~ 1L
-  )
-)
-\end{Sinput}
-\end{Schunk}
-
-Finally, we can fit the hazard using various linear predictors.
-
-\begin{Schunk}
-\begin{Sinput}
-# Fit several models
-fit1 <- fitSmoothHazard(fustat ~ exposure,
-  data = cb_data, time = "futime"
-)
-fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
-  data = cb_data, time = "futime"
-)
-fit3 <- fitSmoothHazard(fustat ~ exposure * futime,
-  data = cb_data, time = "futime"
-)
-\end{Sinput}
-\end{Schunk}
-
-Note that the third model includes an interaction term between exposure
-and follow-up time. In other words, this model no longer exhibits
-proportional hazards. The evidence of non-proportionality of hazards in
-the Stanford Heart Transplant data has been widely discussed
-\citep{arjas1988graphical}.
-
-We can then compare the goodness of fit of these three models using the
-Akaike Information Criterion (AIC).
-
-\begin{Schunk}
-\begin{Soutput}
-#> Model1 Model2 Model3 
-#>    827    791    790
-\end{Soutput}
-\end{Schunk}
-
-As we can see, the best fit is the third model. By visualizing the
-hazard functions for both exposed and unexposed individuals, we can more
-clearly see how the hazards are no longer proportional. We can easily
-obtain a plot of the hazards by using the \texttt{plot.singleEventCB}
-method:
-
-\begin{Schunk}
-\begin{Sinput}
-plot(fit3, hazard.params = list(xvar = "futime",
-                                by = "exposure",
-                                alpha = 0.05,
-                                ylab = "Hazard",
-                                data = cb_data))
-\end{Sinput}
-\begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-hazard-1} \caption{\label{fig:cs4hazardPlot} Plot of the hazard function for exposed and unexposed individuals.}\label{fig:stanford-hazard}
-\end{figure}
-\end{Schunk}
-
-Looking at Figure \ref{fig:cs4hazardPlot}, the non-proportionality seems
-to be more pronounced at the beginning of follow-up than the end.
-Finally, in Figure \ref{fig:cs4CIF}, we turn these estimates of the
-hazard function into estimates of the cumulative incidence functions.
-
-\begin{Schunk}
-\begin{Sinput}
-# Compute absolute risk curves
-newdata <- data.frame(exposure = c(0, 1))
-absrisk <- absoluteRisk(fit3,
-  newdata = newdata,
-  time = seq(0, 425, length.out = 100)
-)
-
-class(absrisk)
-\end{Sinput}
-\begin{Soutput}
-#> [1] "absRiskCB" "matrix"    "array"
-\end{Soutput}
-\begin{Sinput}
-
-plot(absrisk,
-     id.names = c("No Tx", "Tx")) + 
-  ylab("Cumulative Incidence") +
-  xlab("Follow-up time (in days)") +
-  paper_gg_theme
-\end{Sinput}
-\begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-risk-1} \caption{\label{fig:cs4CIF} Plot of the cumulative incidence function for exposed and unexposed individuals.}\label{fig:stanford-risk}
-\end{figure}
-\end{Schunk}
-
-As we can see in the above case-study, the \pkg{casebase} package can
-also be used to model time-varying exposures and non-proportional hazard
-functions.
-
 \hypertarget{discussion}{%

### 2021-02-03 analysis/paper/casebase-manuscript.Rmd
updated abstract, intro and lit review.. to also reflect 3 CS instead of 4

@@
   | In this paper, we present the **casebase** R package, a comprehensive and flexible toolkit for parametric survival analysis.
-We describe how the case-base framework can also be used in more complex settings: competing risks, time-varying exposure, and variable selection.
+We then describe how the case-base framework can be used in more complex settings: non-linear functions of time, competing risks, and variable selection.
 Our package also includes an extensive array of visualization tools to complement the analysis of time-to-event data.
-We illustrate all these features through four different case studies.
+We illustrate all these features through three different case studies.
   | *SRB and MT contributed equally to this work.
@@
 From the perspective of clinicians and their patients, the most relevant quantity is often the 5- or 10-year risk of experiencing a certain event given the patient's particular circumstances, and not the hazard ratio between a treatment and control group.
-Therefore, to make sound clinical decisions, it is important to accurately estimate the *full* hazard function, which can subsequently be used to estimate the cumulative incidence function (CIF).
+Therefore, to make sound clinical decisions, it is important to accurately estimate the *full* hazard function, which can subsequently be used to estimate the CIF.
 Using a parametric estimator of the hazard function leads to a smooth function of time; as a consequence, the CIF and the survival function estimates also vary smoothly over time.
@@
 Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function.
-It also provides tools for variable selection and competing-risk analysis.
-In addition, we provide extensive visualization tools.
+It also provides tools for variable selection, competing-risk analysis, and dealing with time-dependent exposures.
+In addition, we provide functions for exploratory data analysis and visualizing the estimated quantities.
 
@@
 Next, we provide some details about the implementation of case-base sampling in our package, and we give a brief survey of its main functions.
-This is followed by four case studies that illustrate the flexibility and capabilities of \pkg{casebase}.
-We show how the same framework can be used for competing risk analyses, penalized estimation, and for studies with time-dependent exposures.
+This is followed by three case studies that illustrate the flexibility and capabilities of \pkg{casebase}.
+We show how the same framework can be used for non-linear functions of time, competing risks and variable selection via penalized regression.
 Finally, we end the article with a discussion of the results and of future directions.
@@
 
-# Existing packages
+# Comparison with existing packages
 
@@
 
-Parametric survival models are implemented in a handful of packages: \pkg{CFC} [-@mahani2015bayesian], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu], \pkg{mets} [-@scheike2014estimating], and \pkg{survival} [-@survival-package].
-The types of models they allow vary for each package.
+Parametric survival models are implemented in a handful of packages, each differing in the underlying data distribution: \pkg{CFC} [-@mahani2015bayesian], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu], \pkg{mets} [-@scheike2014estimating], and \pkg{survival} [-@survival-package].
 For example, \pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \pkg{flexsurv} and \pkg{survival} allow users to supply any distribution of their choice.
-Also, \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of time using splines, which allows flexible modeling of the hazard function.
-Moreover, \pkg{flexsurv} has the ability to estimate both scale and shape parameters for a variety of parametric families.
+\pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} can model the effect of time using splines, which allows flexible modeling of the hazard function.
 As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear combination of covariates (including time).
-Therefore, our package allows the user to model the effect of time using splines.
+Therefore, our package is more general in that it allows the user to model any linear or non-linear transformation of time including splines and higher order polynomials.
 Also, by including interaction terms between covariates and time, it also allows users to fit (non-proportional) time-varying coefficient models.
-However, we do not explicitly model any shape parameter, unlike \pkg{flexsurv}.
+However, unlike \pkg{flexsurv}, we do not explicitly model any shape parameter.
 

### 2021-02-03 analysis/paper/casebase-manuscript.tex
updated abstract, intro and lit review.. to also reflect 3 CS instead of 4

@@
 comprehensive and flexible toolkit for parametric survival analysis. We
-describe how the case-base framework can also be used in more complex
-settings: competing risks, time-varying exposure, and variable
+then describe how the case-base framework can be used in more complex
+settings: non-linear functions of time, competing risks, and variable
 selection. Our package also includes an extensive array of visualization
 tools to complement the analysis of time-to-event data. We illustrate
-all these features through four different case studies.\\
+all these features through three different case studies.\\
 *SRB and MT contributed equally to this work.
@@
 decisions, it is important to accurately estimate the \emph{full} hazard
-function, which can subsequently be used to estimate the cumulative
-incidence function (CIF). Using a parametric estimator of the hazard
-function leads to a smooth function of time; as a consequence, the CIF
-and the survival function estimates also vary smoothly over time.
+function, which can subsequently be used to estimate the CIF. Using a
+parametric estimator of the hazard function leads to a smooth function
+of time; as a consequence, the CIF and the survival function estimates
+also vary smoothly over time.
 
@@
 the familiar interface of the \code{glm} function. It also provides
-tools for variable selection and competing-risk analysis. In addition,
-we provide extensive visualization tools.
+tools for variable selection, competing-risk analysis, and dealing with
+time-dependent exposures. In addition, we provide functions for
+exploratory data analysis and visualizing the estimated quantities.
 
@@
 about the implementation of case-base sampling in our package, and we
-give a brief survey of its main functions. This is followed by four case
-studies that illustrate the flexibility and capabilities of
-\pkg{casebase}. We show how the same framework can be used for competing
-risk analyses, penalized estimation, and for studies with time-dependent
-exposures. Finally, we end the article with a discussion of the results
-and of future directions.
+give a brief survey of its main functions. This is followed by three
+case studies that illustrate the flexibility and capabilities of
+\pkg{casebase}. We show how the same framework can be used for
+non-linear functions of time, competing risks and variable selection via
+penalized regression. Finally, we end the article with a discussion of
+the results and of future directions.
 
@@
 
-\hypertarget{existing-packages}{%
-\section{Existing packages}\label{existing-packages}}
+\hypertarget{comparison-with-existing-packages}{%
+\section{Comparison with existing
+packages}\label{comparison-with-existing-packages}}
 
@@
 
-Parametric survival models are implemented in a handful of packages:
-\pkg{CFC} \citeyearpar{mahani2015bayesian}, \pkg{flexsurv}
-\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
-\pkg{rsptm2} \citeyearpar{clements_liu}, \pkg{mets}
+Parametric survival models are implemented in a handful of packages,
+each differing in the underlying data distribution: \pkg{CFC}
+\citeyearpar{mahani2015bayesian}, \pkg{flexsurv} \citeyearpar{flexsurv},
+\pkg{SmoothHazard} \citeyearpar{smoothHazard}, \pkg{rsptm2}
+\citeyearpar{clements_liu}, \pkg{mets}
 \citeyearpar{scheike2014estimating}, and \pkg{survival}
-\citeyearpar{survival-package}. The types of models they allow vary for
-each package. For example, \pkg{SmoothHazard} is limited to Weibull
-distributions \citeyearpar{smoothHazard}, whereas both \pkg{flexsurv}
-and \pkg{survival} allow users to supply any distribution of their
-choice. Also, \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and
-\pkg{rstpm2} also have the ability to model the effect of time using
-splines, which allows flexible modeling of the hazard function.
-Moreover, \pkg{flexsurv} has the ability to estimate both scale and
-shape parameters for a variety of parametric families. As discussed
+\citeyearpar{survival-package}. For example, \pkg{SmoothHazard} is
+limited to Weibull distributions \citeyearpar{smoothHazard}, whereas
+both \pkg{flexsurv} and \pkg{survival} allow users to supply any
+distribution of their choice. \pkg{flexsurv}, \pkg{smoothhazard},
+\pkg{mets} and \pkg{rstpm2} can model the effect of time using splines,
+which allows flexible modeling of the hazard function. As discussed
 above, \pkg{casebase} can model any parametric family whose log-hazard
@@
 time, it also allows users to fit (non-proportional) time-varying
-coefficient models. However, we do not explicitly model any shape
-parameter, unlike \pkg{flexsurv}.
+coefficient models. However, unlike \pkg{flexsurv}, we do not explicitly
+model any shape parameter.
 

### 2021-02-04 analysis/paper/casebase-manuscript.Rmd
updated section titles of case studies, reviewed text, pointed readers
to package vignette for example of time-dependent exposure. ready for review

@@
   | In this paper, we present the **casebase** R package, a comprehensive and flexible toolkit for parametric survival analysis.
-We then describe how the case-base framework can be used in more complex settings: non-linear functions of time, competing risks, and variable selection.
+We then describe how the case-base framework can be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection.
 Our package also includes an extensive array of visualization tools to complement the analysis of time-to-event data.
@@
 This is followed by three case studies that illustrate the flexibility and capabilities of \pkg{casebase}.
-We show how the same framework can be used for non-linear functions of time, competing risks and variable selection via penalized regression.
+We show how the same framework can be used for non-linear functions of time and non-proportional hazards, competing risks and variable selection via penalized regression.
 Finally, we end the article with a discussion of the results and of future directions.
@@
 Note that non-proportional hazard models can be achieved at this stage by adding an interaction term involving time (cf.
-Case Study 4 below).
+Case Study 1 below).
 The offset term does not need to be specified by the user, as it is automatically added to the formula before calling \code{glm}.
@@
 $$ CI_j(t) = \int_0^t f_j(u)du.$$
-Again, he integrals are computed using either numerical integration (via the trapezoidal rule) or Monte Carlo integration.
+Again, the integrals are computed using either numerical integration (via the trapezoidal rule) or Monte Carlo integration.
 This option is controlled by the argument \code{method} of the \code{absoluteRisk} function.
 
-In the following sections, we illustrate these functionalities in the context of four case studies.
-
-# Case study 1---European Randomized Study of Prostate Cancer Screening
+# Illustration of package
+
+In the following sections, we illustrate the main functions of the \pkg{casebase} package through three case studies; each one showcasing a different analysis type.
+First, we show how to model different functional forms of time as well as non-proportional hazards.
+Then we perform a competing risks analysis and compare our results with the Fine-Gray model.
+The third case study illustrates how to perform variable selection in dataset with a large number of covariates.
+
+## Case study 1---Modeling different functions of time and non-proportional hazards
 
 For our first case study, we return to the ERSPC study and investigate the differences in risk between the control and screening arms.
-
-## Different functional forms of time
-
 We fit four models that differ in which functional form of time is used: 1) excluded from the linear predictor as seen in \eqref{eq:exp}, 2) linear function as seen in \eqref{eq:gomp}, 3) log function as seen in \eqref{eq:weibull}, and 4) a smooth function using cubic B-splines.
@@
 
-## Time-dependent hazard ratios
-
 Previous re-analyses of these data suggested that the overall screening attributed reduction in death due to prostate cancer of 20% was an underestimate [@hanley2010mortality].
@@
 
-# Case study 2---Bone-marrow transplant
-
-In the next case study, we show how case-base sampling can be used in the context of a competing-risk analysis.
+## Case study 2---Competing risk analysis
+
+In this case study, we show how case-base sampling can be used in the context of a competing-risk analysis.
 For illustrative purposes, we use the same data that was used in Scrucca *et al* [-@scrucca2010regression].
@@
 
-# Case study 3---SUPPORT Data
-
-In the first two case studies, we described the basic functionalities of the \pkg{casebase} package: creating population-time plots, fitting parametric models for hazard functions, and estimating the corresponding cumulative incidence curves.
+## Case study 3---Variable selection
+
 For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function as given by \eqref{eq:penest}.
-
 To illustrate this functionality, we use the dataset from the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) [@knaus1995support].
@@
 Our package also provide several functions to produce graphical summaries of the data and the results.
-We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through four case studies.
+We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through three case studies.
 
@@
 On the other hand, we also displayed the flexibility of case-base sampling by showing how it could be combined with penalized logistic regression to perform variable selection.
-Furthermore, the second and fourth case studies showed how case-base sampling can respectively be applied to competing risks and time-varying exposure settings.
+Furthermore, the second case study showed how case-base sampling can be applied to competing risks settings.
+It should be noted that the three case studies presented above only considered covariates that were fixed at baseline.
+In the package vignette, we use the Stanford Heart Transplant data [@clark1971cardiac,@crowley1977covariance] to show how case-base sampling can also be used in the context of time-dependent exposure.
+In this study, the exposure period was defined as the week following vaccination.
+Hence, the main covariate of interest, i.e.
+exposure to the vaccine, was changing over time.
+In this context, case-base sampling offers an efficient alternative to nested case-control designs or self-matching.
+
 Even though we did not illustrate it in this article, case-base sampling can also be combined with the framework of *generalized additive models*.

### 2021-02-04 analysis/paper/casebase-manuscript.tex
updated section titles of case studies, reviewed text, pointed readers
to package vignette for example of time-dependent exposure. ready for review

@@
 then describe how the case-base framework can be used in more complex
-settings: non-linear functions of time, competing risks, and variable
-selection. Our package also includes an extensive array of visualization
-tools to complement the analysis of time-to-event data. We illustrate
-all these features through three different case studies.\\
+settings: non-linear functions of time and non-proportional hazards,
+competing risks, and variable selection. Our package also includes an
+extensive array of visualization tools to complement the analysis of
+time-to-event data. We illustrate all these features through three
+different case studies.\\
 *SRB and MT contributed equally to this work.
@@
 \pkg{casebase}. We show how the same framework can be used for
-non-linear functions of time, competing risks and variable selection via
-penalized regression. Finally, we end the article with a discussion of
-the results and of future directions.
+non-linear functions of time and non-proportional hazards, competing
+risks and variable selection via penalized regression. Finally, we end
+the article with a discussion of the results and of future directions.
 
@@
 can be expressed as a linear combination of covariates (including time).
-Therefore, our package allows the user to model the effect of time using
-splines. Also, by including interaction terms between covariates and
-time, it also allows users to fit (non-proportional) time-varying
-coefficient models. However, unlike \pkg{flexsurv}, we do not explicitly
-model any shape parameter.
+Therefore, our package is more general in that it allows the user to
+model any linear or non-linear transformation of time including splines
+and higher order polynomials. Also, by including interaction terms
+between covariates and time, it also allows users to fit
+(non-proportional) time-varying coefficient models. However, unlike
+\pkg{flexsurv}, we do not explicitly model any shape parameter.
 
@@
 achieved at this stage by adding an interaction term involving time
-(cf.~Case Study 4 below). The offset term does not need to be specified
+(cf.~Case Study 1 below). The offset term does not need to be specified
 by the user, as it is automatically added to the formula before calling
@@
 
-\[ CI_j(t) = \int_0^t f_j(u)du.\] Again, he integrals are computed using
-either numerical integration (via the trapezoidal rule) or Monte Carlo
-integration. This option is controlled by the argument \code{method} of
-the \code{absoluteRisk} function.
-
-In the following sections, we illustrate these functionalities in the
-context of four case studies.
-
-\hypertarget{case-study-1european-randomized-study-of-prostate-cancer-screening}{%
-\section{Case study 1---European Randomized Study of Prostate Cancer
-Screening}\label{case-study-1european-randomized-study-of-prostate-cancer-screening}}
+\[ CI_j(t) = \int_0^t f_j(u)du.\] Again, the integrals are computed
+using either numerical integration (via the trapezoidal rule) or Monte
+Carlo integration. This option is controlled by the argument
+\code{method} of the \code{absoluteRisk} function.
+
+\hypertarget{illustration-of-package}{%
+\section{Illustration of package}\label{illustration-of-package}}
+
+In the following sections, we illustrate the main functions of the
+\pkg{casebase} package through three case studies; each one showcasing a
+different analysis type. First, we show how to model different
+functional forms of time as well as non-proportional hazards. Then we
+perform a competing risks analysis and compare our results with the
+Fine-Gray model. The third case study illustrates how to perform
+variable selection in dataset with a large number of covariates.
+
+\hypertarget{case-study-1modeling-different-functions-of-time-and-non-proportional-hazards}{%
+\subsection{Case study 1---Modeling different functions of time and
+non-proportional
+hazards}\label{case-study-1modeling-different-functions-of-time-and-non-proportional-hazards}}
 
 For our first case study, we return to the ERSPC study and investigate
-the differences in risk between the control and screening arms.
-
-\hypertarget{different-functional-forms-of-time}{%
-\subsection{Different functional forms of
-time}\label{different-functional-forms-of-time}}
-
-We fit four models that differ in which functional form of time is used:
-1) excluded from the linear predictor as seen in \eqref{eq:exp}, 2)
-linear function as seen in \eqref{eq:gomp}, 3) log function as seen in
+the differences in risk between the control and screening arms. We fit
+four models that differ in which functional form of time is used: 1)
+excluded from the linear predictor as seen in \eqref{eq:exp}, 2) linear
+function as seen in \eqref{eq:gomp}, 3) log function as seen in
 \eqref{eq:weibull}, and 4) a smooth function using cubic B-splines. The
@@
 
-\hypertarget{time-dependent-hazard-ratios}{%
-\subsection{Time-dependent hazard
-ratios}\label{time-dependent-hazard-ratios}}
-
 Previous re-analyses of these data suggested that the overall screening
@@
 
-\hypertarget{case-study-2bone-marrow-transplant}{%
-\section{Case study 2---Bone-marrow
-transplant}\label{case-study-2bone-marrow-transplant}}
-
-In the next case study, we show how case-base sampling can be used in
-the context of a competing-risk analysis. For illustrative purposes, we
-use the same data that was used in Scrucca \emph{et al}
+\hypertarget{case-study-2competing-risk-analysis}{%
+\subsection{Case study 2---Competing risk
+analysis}\label{case-study-2competing-risk-analysis}}
+
+In this case study, we show how case-base sampling can be used in the
+context of a competing-risk analysis. For illustrative purposes, we use
+the same data that was used in Scrucca \emph{et al}
 \citeyearpar{scrucca2010regression}. The data was downloaded from the
@@
 
-\hypertarget{case-study-3support-data}{%
-\section{Case study 3---SUPPORT Data}\label{case-study-3support-data}}
-
-In the first two case studies, we described the basic functionalities of
-the \pkg{casebase} package: creating population-time plots, fitting
-parametric models for hazard functions, and estimating the corresponding
-cumulative incidence curves. For the third case study, we show how
-\pkg{casebase} can also be used for variable selection through
-regularized estimation of the hazard function as given by
-\eqref{eq:penest}.
-
-To illustrate this functionality, we use the dataset from the Study to
-Understand Prognoses Preferences Outcomes and Risks of Treatment
-(SUPPORT) \citep{knaus1995support}. The SUPPORT dataset tracks death in
-five American hospitals within individuals who are considered seriously
+\hypertarget{case-study-3variable-selection}{%
+\subsection{Case study 3---Variable
+selection}\label{case-study-3variable-selection}}
+
+For the third case study, we show how \pkg{casebase} can also be used
+for variable selection through regularized estimation of the hazard
+function as given by \eqref{eq:penest}. To illustrate this
+functionality, we use the dataset from the Study to Understand Prognoses
+Preferences Outcomes and Risks of Treatment (SUPPORT)
+\citep{knaus1995support}. The SUPPORT dataset tracks death in five
+American hospitals within individuals who are considered seriously
 ill.~The original data is available online from the Department of
@@
 provided details about our implementation, and we illustrated the merits
-of the approach and the package through four case studies.
+of the approach and the package through three case studies.
 
@@
 with penalized logistic regression to perform variable selection.
-Furthermore, the second and fourth case studies showed how case-base
-sampling can respectively be applied to competing risks and time-varying
-exposure settings. Even though we did not illustrate it in this article,
-case-base sampling can also be combined with the framework of
-\emph{generalized additive models}. This functionality has already been
-implemented in the package. Similarly, case-base sampling can be
-combined with quasi-likelihood estimation to fit survival models that
-can account for the presence of over-dispersion. All of these examples
-illustrate how the case-base sampling framework in general, and the
-package \pkg{casebase} in particular, allows the user to fit a broad and
-flexible family of survival functions.
+Furthermore, the second case study showed how case-base sampling can be
+applied to competing risks settings. It should be noted that the three
+case studies presented above only considered covariates that were fixed
+at baseline. In the package vignette, we use the Stanford Heart
+Transplant data \citep[\citet{crowley1977covariance}]{clark1971cardiac}
+to show how case-base sampling can also be used in the context of
+time-dependent exposure. In this study, the exposure period was defined
+as the week following vaccination. Hence, the main covariate of
+interest, i.e.~exposure to the vaccine, was changing over time. In this
+context, case-base sampling offers an efficient alternative to nested
+case-control designs or self-matching.
+
+Even though we did not illustrate it in this article, case-base sampling
+can also be combined with the framework of \emph{generalized additive
+models}. This functionality has already been implemented in the package.
+Similarly, case-base sampling can be combined with quasi-likelihood
+estimation to fit survival models that can account for the presence of
+over-dispersion. All of these examples illustrate how the case-base
+sampling framework in general, and the package \pkg{casebase} in
+particular, allows the user to fit a broad and flexible family of
+survival functions.
 

### 2021-02-07 analysis/paper/casebase-manuscript.Rmd
update last table
Other minor changes

@@
       Montreal, QC, Canada H3A 1A2
-    email: \email{sahir.bhatnagar@mail.mcgill.ca}
+    email: \email{sahir.bhatnagar@mcgill.ca}
     url: http://sahirbhatnagar.com/
@@
 abstract: |
-  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician and the patient is the risk of an event given a covariate profile.
+  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician and the patient is the risk of an event given their covariate profile.
 However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the likelihood expression or the estimating function.
@@
   | In this paper, we present the **casebase** R package, a comprehensive and flexible toolkit for parametric survival analysis.
-We then describe how the case-base framework can be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection.
+We describe how the case-base framework can be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection.
 Our package also includes an extensive array of visualization tools to complement the analysis of time-to-event data.
 We illustrate all these features through three different case studies.
-  | *SRB and MT contributed equally to this work.
+  | * SRB and MT contributed equally to this work.
 preamble: >
@@
 
-In the following sections, we illustrate the main functions of the \pkg{casebase} package through three case studies; each one showcasing a different analysis type.
+In this section, we illustrate the main functions of the \pkg{casebase} package through three case studies.
+Each one showcases a different type of analysis.
 First, we show how to model different functional forms of time as well as non-proportional hazards.
-Then we perform a competing risks analysis and compare our results with the Fine-Gray model.
-The third case study illustrates how to perform variable selection in dataset with a large number of covariates.
+Then we perform a competing-risk analysis and compare our results with the Fine-Gray model.
+The third case study illustrates how to perform variable selection in high-dimensional datasets.
 
@@
 For our first case study, we return to the ERSPC study and investigate the differences in risk between the control and screening arms.
-We fit four models that differ in which functional form of time is used: 1) excluded from the linear predictor as seen in \eqref{eq:exp}, 2) linear function as seen in \eqref{eq:gomp}, 3) log function as seen in \eqref{eq:weibull}, and 4) a smooth function using cubic B-splines.
+We fit four models that differ in which functional form of time is used: 1) excluded from the linear predictor (Eq.
+\eqref{eq:exp}); 2) linear function (Eq.
+\eqref{eq:gomp}); 3) log function (Eq.
+\eqref{eq:weibull}); and 4) a smooth function using cubic B-splines.
 The models are fit using \code{fitSmoothHazard} with the familiar formula interface:
@@
 set.seed(12)
-fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
-             gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
-             weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
-             splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
-
-fits <- lapply(fmla, function(i) {
-  fitSmoothHazard(i, data = ERSPC, ratio = 100)
+fmla <- list(exponential = formula(DeadOfPrCa ~ ScrArm),
+             gompertz = formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
+             weibull = formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
+             splines = formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
+
+fits <- lapply(fmla, function(form) {
+  fitSmoothHazard(form, data = ERSPC, ratio = 100)
 })
@@
 
-The output object from \code{fitSmoothHazard} inherits from the \code{singleEventCB} and \code{glm} classes. As such, we can directly use the \code{summary} generic:
-
-```{r, eval = FALSE, echo=TRUE}
+The output object from \code{fitSmoothHazard} inherits from the \code{singleEventCB} and \code{glm} classes.
+For this reason, we leveraged the \code{summary} method for \code{glm} objects to output a familiar summary of the results:
+
+```{r, eval = TRUE, echo=eval_cs1}
 summary(fits[["splines"]]) 
@@
 
-```{r, eval = eval_cs1, echo=FALSE}
+```{r, eval = FALSE, echo=FALSE}
 print.summary.glm <- function(x,digits=max(3,getOption("digits")-3),symbolic.cor=x$symbolic.cor,
@@
 
-Next, the \code{absoluteRisk} function takes as input the \code{fitSmoothHazard} object and returns a matrix where each column corresponds to the covariate profiles specified in the \code{newdata} argument, and each row corresponds to a specified time point:
+Next, the \code{absoluteRisk} function takes as input the \code{singleEventCB} object and returns a matrix where each column corresponds to the covariate profiles specified in the \code{newdata} argument, and each row corresponds to time points specified by the \code{time} argument:
 
@@
 new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
-new_time <- seq(0,14,0.1)
-
-risks <- lapply(fits, function(i) {
-  absoluteRisk(object = i, time = new_time, newdata = new_data, method = "mont")
+new_time <- seq(0, 14, by = 0.1)
+
+risks <- lapply(fits, function(fit) {
+  absoluteRisk(fit, time = new_time, newdata = new_data)
 })
@@
 We notice a better fit with increasing complexity of our model for time in Figure \ref{fig:erspc-cox-cif} (panels 2--4).
-As noted above, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models.
+As noted in Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models.
 Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) spline model:
@@
 The estimated 20% (from a proportional hazards model) did not account for the delay between screening and the time the effect is expected to be observed.
-As a result, the null effects in years 1–7 masked the substantial reductions that began to appear from year 8 onwards.
+As a result, the null effects in years 1–7 masked the substantial reductions that began to appear from year 8 onward.
 This motivates the use of a time-dependent hazard ratio which can easily be fit with the \pkg{casebase} package by including an interaction term with time in the model:
@@
 
-With this first case study, we explored how \pkg{casebase} allows us to fit different parametric survival models with possible time-varying effects, and how we can compare the fit of each model with tools from GLMs.
+<!-- With this first case study, we explored how \pkg{casebase} allows us to fit different parametric survival models with possible time-varying effects, and how we can compare the fit of each model with tools from GLMs.
+-->
 
@@
 
-```{r bmtcrr-casebase-weibull, warning = FALSE, eval = eval_cs2}
+```{r bmtcrr-casebase-weibull, warning = FALSE, eval = eval_cs2, echo = TRUE}
 model_cb <- fitSmoothHazard(
@@
 
-In this third case study, we showed how case-base sampling can be used in conjunction with penalized logistic regression to perform variable selection in survival models.
+<!-- In this third case study, we showed how case-base sampling can be used in conjunction with penalized logistic regression to perform variable selection in survival models. -->
 
@@
 In this article, we presented the \proglang{R} package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating CIFs using case-base sampling.
-Our package also provide several functions to produce graphical summaries of the data and the results.
+Our package also provides several functions to produce graphical summaries of the data and the results.
 We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through three case studies.
@@
 It should be noted that the three case studies presented above only considered covariates that were fixed at baseline.
-In the package vignette, we use the Stanford Heart Transplant data [@clark1971cardiac,@crowley1977covariance] to show how case-base sampling can also be used in the context of time-dependent exposure.
+In one of the package vignettes, we use the Stanford Heart Transplant data [@clark1971cardiac,@crowley1977covariance] to show how case-base sampling can also be used in the context of time-dependent exposure.
 In this study, the exposure period was defined as the week following vaccination.
@@
 Feature & Cox model & Case-base sampling\\
-\midrule
-Model type & Semi-parametric & Fully parametric\\
-Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\
-Cumulative incidence & Step function & Smooth-in-time curve\\
-Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\
-Model testing &  & Use GLM framework \newline (e.g.\ LRT, AIC, BIC)\\
+\toprule
+Model type & Semi-parametric & Fully parametric\\\midrule
+Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\\midrule
+Cumulative incidence & Step function & Smooth-in-time curve\\\midrule
+Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\\midrule
+Model testing & Likelihood-ratio tests, score tests, & Use GLM framework\\
+ & information criteria & (e.g.\ LRT, AIC, BIC)\\\midrule
 \addlinespace

### 2021-02-07 analysis/paper/casebase-manuscript.tex
update last table
Other minor changes

@@
 In epidemiological studies of time-to-event data, a quantity of interest
-to the clinician and the patient is the risk of an event given a
+to the clinician and the patient is the risk of an event given their
 covariate profile. However, methods relying on time matching or risk-set
@@
 comprehensive and flexible toolkit for parametric survival analysis. We
-then describe how the case-base framework can be used in more complex
+describe how the case-base framework can be used in more complex
 settings: non-linear functions of time and non-proportional hazards,
@@
 different case studies.\\
-*SRB and MT contributed equally to this work.
+* SRB and MT contributed equally to this work.
 }
@@
 
-In the following sections, we illustrate the main functions of the
-\pkg{casebase} package through three case studies; each one showcasing a
-different analysis type. First, we show how to model different
-functional forms of time as well as non-proportional hazards. Then we
-perform a competing risks analysis and compare our results with the
-Fine-Gray model. The third case study illustrates how to perform
-variable selection in dataset with a large number of covariates.
+In this section, we illustrate the main functions of the \pkg{casebase}
+package through three case studies. Each one showcases a different type
+of analysis. First, we show how to model different functional forms of
+time as well as non-proportional hazards. Then we perform a
+competing-risk analysis and compare our results with the Fine-Gray
+model. The third case study illustrates how to perform variable
+selection in high-dimensional datasets.
 
@@
 four models that differ in which functional form of time is used: 1)
-excluded from the linear predictor as seen in \eqref{eq:exp}, 2) linear
-function as seen in \eqref{eq:gomp}, 3) log function as seen in
-\eqref{eq:weibull}, and 4) a smooth function using cubic B-splines. The
+excluded from the linear predictor (Eq. \eqref{eq:exp}); 2) linear
+function (Eq. \eqref{eq:gomp}); 3) log function (Eq.
+\eqref{eq:weibull}); and 4) a smooth function using cubic B-splines. The
 models are fit using \code{fitSmoothHazard} with the familiar formula
@@
 \begin{Sinput}
-fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
-             gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
-             weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
-             splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
-
-fits <- lapply(fmla, function(i) {
-  fitSmoothHazard(i, data = ERSPC, ratio = 100)
+fmla <- list(exponential = formula(DeadOfPrCa ~ ScrArm),
+             gompertz = formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
+             weibull = formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
+             splines = formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
+
+fits <- lapply(fmla, function(form) {
+  fitSmoothHazard(form, data = ERSPC, ratio = 100)
 })
@@
 The output object from \code{fitSmoothHazard} inherits from the
-\code{singleEventCB} and \code{glm} classes. As such, we can directly
-use the \code{summary} generic:
+\code{singleEventCB} and \code{glm} classes. For this reason, we
+leveraged the \code{summary} method for \code{glm} objects to output a
+familiar summary of the results:
 
@@
 \end{Sinput}
-\end{Schunk}
-
-\begin{Schunk}
 \begin{Soutput}
+#> Fitting smooth hazards with case-base sampling
+#> 
+#> Sample size: 159893 
+#> Number of events: 540 
+#> Number of base moments: 54000 
+#> ----
+#> 
+#> Call:
+#> fitSmoothHazard(formula = form, data = ERSPC, ratio = 100)
+#> 
+#> Deviance Residuals: 
+#>    Min      1Q  Median      3Q     Max  
+#> -0.409  -0.172  -0.135  -0.084   3.808  
 #> 
@@
 #>                       Estimate Std. Error z value Pr(>|z|)    
-#> (Intercept)           -10.2905     0.3185  -32.31  < 2e-16
-#> bs(Follow.Up.Time)1     4.3587     0.8008    5.44  5.2e-08
-#> bs(Follow.Up.Time)2     2.0890     0.4721    4.42  9.7e-06
-#> bs(Follow.Up.Time)3     4.6266     0.6910    6.70  2.1e-11
-#> ScrArmScreening group  -0.2354     0.0886   -2.66   0.0079
+#> (Intercept)           -10.2905     0.3185  -32.31  < 2e-16 ***
+#> bs(Follow.Up.Time)1     4.3587     0.8008    5.44  5.2e-08 ***
+#> bs(Follow.Up.Time)2     2.0890     0.4721    4.42  9.7e-06 ***
+#> bs(Follow.Up.Time)3     4.6266     0.6910    6.70  2.1e-11 ***
+#> ScrArmScreening group  -0.2354     0.0886   -2.66   0.0079 ** 
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
 #> 
@@
 Next, the \code{absoluteRisk} function takes as input the
-\code{fitSmoothHazard} object and returns a matrix where each column
+\code{singleEventCB} object and returns a matrix where each column
 corresponds to the covariate profiles specified in the \code{newdata}
-argument, and each row corresponds to a specified time point:
+argument, and each row corresponds to time points specified by the
+\code{time} argument:
 
@@
 complexity of our model for time in Figure \ref{fig:erspc-cox-cif}
-(panels 2--4). As noted above, the usual asymptotic results hold for
-likelihood ratio tests built using case-base sampling models. Therefore,
-we can easily test the null hypothesis that the exponential model is
-just as good as the larger (in terms of number of parameters) spline
-model:
+(panels 2--4). As noted in Theoretical Details section, the usual
+asymptotic results hold for likelihood ratio tests built using case-base
+sampling models. Therefore, we can easily test the null hypothesis that
+the exponential model is just as good as the larger (in terms of number
+of parameters) spline model:
 
@@
 result, the null effects in years 1--7 masked the substantial reductions
-that began to appear from year 8 onwards. This motivates the use of a
+that began to appear from year 8 onward. This motivates the use of a
 time-dependent hazard ratio which can easily be fit with the
@@
 \end{Schunk}
-
-With this first case study, we explored how \pkg{casebase} allows us to
-fit different parametric survival models with possible time-varying
-effects, and how we can compare the fit of each model with tools from
-GLMs.
 
@@
 data using a linear term for time:
+
+\begin{Schunk}
+\begin{Sinput}
+model_cb <- fitSmoothHazard(
+  Status ~ ftime + Sex + D + Phase + Source + Age,
+  data = bmtcrr,
+  ratio = 100,
+  time = "ftime"
+)
+\end{Sinput}
+\end{Schunk}
 
@@
 
-In this third case study, we showed how case-base sampling can be used
-in conjunction with penalized logistic regression to perform variable
-selection in survival models.
-
 \hypertarget{discussion}{%
@@
 which provides functions for fitting smooth parametric hazards and
-estimating CIFs using case-base sampling. Our package also provide
+estimating CIFs using case-base sampling. Our package also provides
 several functions to produce graphical summaries of the data and the
@@
 case studies presented above only considered covariates that were fixed
-at baseline. In the package vignette, we use the Stanford Heart
+at baseline. In one of the package vignettes, we use the Stanford Heart
 Transplant data \citep[\citet{crowley1977covariance}]{clark1971cardiac}
@@
 %
-\\\email{sahir.bhatnagar@mail.mcgill.ca}
+\\\email{sahir.bhatnagar@mcgill.ca}
 }

### 2021-02-08 analysis/paper/casebase-manuscript.Rmd
remove coxRidge

@@
 Several packages implement penalized estimation for the Cox model: \pkg{glmnet} [-@regpathcox], \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{riskRegression} [-@gerds_blanche].
-Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \pkg{CoxRidge} [-@perperoglou] and \pkg{rstpm2} [-@clements_liu], while \pkg{survival} [-@survival-package] has an implementation of ridge-penalized estimation.
+Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \pkg{rstpm2} [-@clements_liu], while \pkg{survival} [-@survival-package] has an implementation of ridge-penalized estimation.
 On the other hand, our package \pkg{casebase} provides penalized estimation of the hazard function.
@@
 \textbf{cmprsk}             & \checkmark                        &                           &                       &                  &                     &  \checkmark                        &                                  & \checkmark                                \\ \hline
-\textbf{CoxRidge}        &                          & \checkmark                         & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
 \textbf{crrp}            & \checkmark                        &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline

### 2021-02-09 analysis/paper/casebase-manuscript.Rmd
fixed links to cran packages usin CRANpkg instead of pkg

@@
 Their approach naturally leads to estimates of the cumulative incidence that are smooth-in-time.
-  | In this paper, we present the **casebase** R package, a comprehensive and flexible toolkit for parametric survival analysis.
+  | In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis.
 We describe how the case-base framework can be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection.
@@
 
-```{r setup, include=FALSE}
+```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
 knitr::opts_chunk$set(
@@
 
-In this article, we present the \pkg{casebase} package [@casebase-package] implemented in \proglang{R} [@r-core], for parametric survival analysis that combines the ideas of Hanley & Miettinen [-@hanley2009fitting] into a simple interface.
-The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to compute a patient's risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile.
+In this article, we present the \CRANpkg{casebase} package [@casebase-package] implemented in \proglang{R} [@r-core], for parametric survival analysis that combines the ideas of Hanley & Miettinen [-@hanley2009fitting] into a simple interface.
+The purpose of the \CRANpkg{casebase} package is to provide practitioners with an easy-to-use software tool to compute a patient's risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile.
 Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function.
@@
 In what follows, we first recall some theoretical details on case-base sampling and its use for estimating parametric hazard functions.
-We then give a short review of existing \proglang{R} packages that implement comparable features as \pkg{casebase}.
+We then give a short review of existing \proglang{R} packages that implement comparable features as \CRANpkg{casebase}.
 Next, we provide some details about the implementation of case-base sampling in our package, and we give a brief survey of its main functions.
-This is followed by three case studies that illustrate the flexibility and capabilities of \pkg{casebase}.
+This is followed by three case studies that illustrate the flexibility and capabilities of \CRANpkg{casebase}.
 We show how the same framework can be used for non-linear functions of time and non-proportional hazards, competing risks and variable selection via penalized regression.
@@
 Accordingly, there is already a vast ecosystem of \proglang{R} packages implementing different methodologies.
-In this section, we describe how the functionalities of \pkg{casebase} compare to these packages.
-
-At the time of writing, a cursory examination of CRAN's task view on survival analysis reveals that there are over 250 packages related to survival analysis [@survTaskView].
+In this section, we describe how the functionalities of \CRANpkg{casebase} compare to these packages.
+
+At the time of writing, a cursory examination of CRAN's \ctv{Survival} Task View reveals that there are over 250 packages related to survival analysis [@survTaskView].
 For the purposes of this article, we restricted our review to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and CIF estimation.
 By searching for appropriate keywords in the \code{DESCRIPTION} file of these packages, we found 60 relevant packages.
-These 60 packages were then manually examined to determine which ones are comparable to \pkg{casebase}.
+These 60 packages were then manually examined to determine which ones are comparable to \CRANpkg{casebase}.
 In particular, we excluded packages that were focused on a different set of problems, such as frailty and multi-state models.
@@
 
-Parametric survival models are implemented in a handful of packages, each differing in the underlying data distribution: \pkg{CFC} [-@mahani2015bayesian], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu], \pkg{mets} [-@scheike2014estimating], and \pkg{survival} [-@survival-package].
-For example, \pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \pkg{flexsurv} and \pkg{survival} allow users to supply any distribution of their choice.
-\pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} can model the effect of time using splines, which allows flexible modeling of the hazard function.
-As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear combination of covariates (including time).
+Parametric survival models are implemented in a handful of packages, each differing in the underlying data distribution: \CRANpkg{CFC} [-@mahani2015bayesian], \CRANpkg{flexsurv} [-@flexsurv], \CRANpkg{SmoothHazard} [-@smoothHazard], \CRANpkg{rsptm2} [-@clements_liu], \CRANpkg{mets} [-@scheike2014estimating], and \CRANpkg{survival} [-@survival-package].
+For example, \CRANpkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \CRANpkg{flexsurv} and \CRANpkg{survival} allow users to supply any distribution of their choice.
+\CRANpkg{flexsurv}, \CRANpkg{smoothhazard}, \CRANpkg{mets} and \CRANpkg{rstpm2} can model the effect of time using splines, which allows flexible modeling of the hazard function.
+As discussed above, \CRANpkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear combination of covariates (including time).
 Therefore, our package is more general in that it allows the user to model any linear or non-linear transformation of time including splines and higher order polynomials.
 Also, by including interaction terms between covariates and time, it also allows users to fit (non-proportional) time-varying coefficient models.
-However, unlike \pkg{flexsurv}, we do not explicitly model any shape parameter.
-
-Several packages implement penalized estimation for the Cox model: \pkg{glmnet} [-@regpathcox], \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{riskRegression} [-@gerds_blanche].
-Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \pkg{rstpm2} [-@clements_liu], while \pkg{survival} [-@survival-package] has an implementation of ridge-penalized estimation.
-On the other hand, our package \pkg{casebase} provides penalized estimation of the hazard function.
-To our knowledge, \pkg{casebase} and \pkg{rsptm2} are the only packages to offer this functionality.
+However, unlike \CRANpkg{flexsurv}, we do not explicitly model any shape parameter.
+
+Several packages implement penalized estimation for the Cox model: \CRANpkg{glmnet} [-@regpathcox], \CRANpkg{glmpath} [-@park_hastie], \CRANpkg{penalized} [-@l1penal], \CRANpkg{riskRegression} [-@gerds_blanche].
+Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \CRANpkg{rstpm2} [-@clements_liu], while \CRANpkg{survival} [-@survival-package] has an implementation of ridge-penalized estimation.
+On the other hand, our package \CRANpkg{casebase} provides penalized estimation of the hazard function.
+To our knowledge, \CRANpkg{casebase} and \CRANpkg{rsptm2} are the only packages to offer this functionality.
 
 Next, several \proglang{R} packages implement methodologies for competing risk analysis; for a different perspective on this topic, see Mahani & Sharabiani [-@mahani2015bayesian].
-The package \pkg{cmprsk} provides methods for cause-specific subdistribution hazards, such as in the Fine-Gray model [-@fine1999proportional].
-On the other hand, the package \pkg{CFC} estimates cause-specific CIFs from unadjusted, non-parametric survival functions.
-Our package \pkg{casebase} also provides functionalities for competing risk analysis by estimating parametrically the cause-specific hazards.
+The package \CRANpkg{cmprsk} provides methods for cause-specific subdistribution hazards, such as in the Fine-Gray model [-@fine1999proportional].
+On the other hand, the package \CRANpkg{CFC} estimates cause-specific CIFs from unadjusted, non-parametric survival functions.
+Our package \CRANpkg{casebase} also provides functionalities for competing risk analysis by estimating parametrically the cause-specific hazards.
 From these quantities, we can then estimate the cause-specific CIFs.
@@
 Using this estimate, the survival function and the CIF can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below).
-Packages providing this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and \pkg{survival}.
-Our package \pkg{casebase} also follows this approach for both single-event and competing-risk analyses.
+Packages providing this functionality include \CRANpkg{CFC}, \CRANpkg{flexsurv}, \CRANpkg{mets}, and \CRANpkg{survival}.
+Our package \CRANpkg{casebase} also follows this approach for both single-event and competing-risk analyses.
 The second category outlined above broadly corresponds to semi-parametric models.
@@
 spline functions).
-Packages that implement this approach include \pkg{riskRegression}, \pkg{rstpm2}, and \pkg{survival}.
+Packages that implement this approach include \CRANpkg{riskRegression}, \CRANpkg{rstpm2}, and \CRANpkg{survival}.
 As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category often produces estimates in the form of stepwise functions.
@@
 
-The functions in the \pkg{casebase} package can be divided into two categories: 1) exploratory data analysis, in the form of population-time plots; and 2) parametric modeling of the hazard function.
+The functions in the \CRANpkg{casebase} package can be divided into two categories: 1) exploratory data analysis, in the form of population-time plots; and 2) parametric modeling of the hazard function.
 We strove for compatibility with both \code{data.frame}s and \code{data.table}s; this can be seen in the coding choices we made and the unit tests we wrote.
@@
 By splitting these tasks, we give flexibility to the user.
-While the method call in Step 2 allows further customization by using the \pkg{ggplot2} [@ggplot2] family of functions, users may choose the graphics system of their choice to create population-time plots from the object created in Step 1.
+While the method call in Step 2 allows further customization by using the \CRANpkg{ggplot2} [@ggplot2] family of functions, users may choose the graphics system of their choice to create population-time plots from the object created in Step 1.
 
 To illustrate these functions, we will use data from the European Randomized Study of Prostate Cancer Screening (ERSPC) [@schroder2009screening] which was extracted using the approach described in Liu *et al.* [-@liu2014recovering].
-This dataset is available through the \pkg{casebase} package.
+This dataset is available through the \CRANpkg{casebase} package.
 It contains the individual observations for 159,893 men  from seven European countries, who were between the ages of 55 and 69 years when recruited for the trial.
@@
 Case Study 3).
-This functionality is provided through the \pkg{glmnet} package [@friedman2010jss].
-  - \code{gam}: This option provides support for *Generalized Additive Models* via the \pkg{mgcv} package [@hastie1987generalized].
-  - \code{gbm}: This option provides support for *Gradient Boosted Trees* via the \pkg{gbm} package. This feature is still experimental.
-
-In the case of multiple competing events, the hazard is fitted via multinomial regression as performed by the \pkg{VGAM} package.
+This functionality is provided through the \CRANpkg{glmnet} package [@friedman2010jss].
+  - \code{gam}: This option provides support for *Generalized Additive Models* via the \CRANpkg{mgcv} package [@hastie1987generalized].
+  - \code{gbm}: This option provides support for *Gradient Boosted Trees* via the \CRANpkg{gbm} package. This feature is still experimental.
+
+In the case of multiple competing events, the hazard is fitted via multinomial regression as performed by the \CRANpkg{VGAM} package.
 We selected this package for its ability to fit multinomial regression models with an offset.
@@
 
-In this section, we illustrate the main functions of the \pkg{casebase} package through three case studies.
+In this section, we illustrate the main functions of the \CRANpkg{casebase} package through three case studies.
 Each one showcases a different type of analysis.
@@
 
-In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from \pkg{casebase} on the Cox model CIF.
+In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from \CRANpkg{casebase} on the Cox model CIF.
 The CIF estimates for the exponential model in panel (1) overestimate the cumulative incidence earlier on in time, and underestimate it later on.
@@
 CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
-# CI_exp <- c(exp(coef(fits$exponential))["ScrArmScreening group"], exp(confint(fits$exponential))["ScrArmScreening group",])
-# CI_gompertz <- c(exp(coef(fits$gompertz))["ScrArmScreening group"], exp(confint(fits$gompertz))["ScrArmScreening group",])
-# CI_weibull <- c(exp(coef(fits$weibull))["ScrArmScreening group"], exp(confint(fits$weibull))["ScrArmScreening group",])
-# CI_splines <- c(exp(coef(fits$splines))["ScrArmScreening group"], exp(confint(fits$splines))["ScrArmScreening group",])
 
@@
             do.call(rbind, casebase_HR_CI), do.call(rbind, survreg_HR_CI))
-
-# table_ci <- rbind(CI_cox, CI_exp, CI_gompertz, CI_weibull, CI_splines)
-# colnames(table_ci) <- c("HR", "LB", "UB")
-# 
-# data.frame(Model = c("Cox", "Exponential", "Gompertz", "Weibull", "Splines")) %>% 
-#   bind_cols(as.data.frame(table_ci)) %>% 
-#   mutate(CI = paste0("(", round(LB, 2), ", ",
-#                      round(UB, 2), ")"),
-#          Width = UB - LB) %>% 
-#   dplyr::select(-LB, -UB) %>% 
-#   knitr::kable(format = "latex", booktabs = TRUE)  %>%
-#   kable_styling()
 ```
@@
 ```{r print-erspc-estimates, eval = eval_cs1}
-## ---- print-sim-table ----
-
 kable(ptable, "latex", booktabs = TRUE, align = c("l","c","c"),
@@
   kable_styling() %>%
-  # add_header_above(c(" "," ","No overlap" = 2, "All causal SNPs\nin kinship" = 2, "No overlap" = 2, "All causal SNPs\nin kinship" = 2)) %>%
-  # add_header_above(c(" "," ","Null model" = 4, "1% Causal SNPs" = 4)) %>%
   column_spec(1, bold = TRUE) %>%
@@
   footnote(general = sprintf("Cox model estimate: HR (95%% CI) = %0.2f (%0.2f, %0.2f)", CI_cox[1], CI_cox[2], CI_cox[3]),
-           general_title = "") #%>%
-  # footnote(general = "Model Size is the number of selected variables using the high-dimensional BIC for \texttt{ggmix} and 10-fold cross validation for \texttt{lasso} and \texttt{twostep}.")
-  #          # number = c("", "Footnote 2; "),
-  #          # alphabet = c("Footnote A; ", "Footnote B; "),
-  #          # symbol = c("Footnote Symbol 1; ", "Footnote Symbol 2"),
-  #          general_title = "Note: ", number_title = "Type I: ",
-  #          alphabet_title = "Type II: ", symbol_title = "Type III: ",
-  #          footnote_as_chunk = TRUE, title_format = c("italic", "underline")
-  # )
+           general_title = "") 
 ```
@@
 As a result, the null effects in years 1–7 masked the substantial reductions that began to appear from year 8 onward.
-This motivates the use of a time-dependent hazard ratio which can easily be fit with the \pkg{casebase} package by including an interaction term with time in the model:
+This motivates the use of a time-dependent hazard ratio which can easily be fit with the \CRANpkg{casebase} package by including an interaction term with time in the model:
 
@@
 
-<!-- With this first case study, we explored how \pkg{casebase} allows us to fit different parametric survival models with possible time-varying effects, and how we can compare the fit of each model with tools from GLMs.
--->
-
 ## Case study 2---Competing risk analysis
@@
 For illustrative purposes, we use the same data that was used in Scrucca *et al* [-@scrucca2010regression].
-The data was downloaded from the first author's website, and it is now available as part of the \pkg{casebase} package.
+The data was downloaded from the first author's website, and it is now available as part of the \CRANpkg{casebase} package.
 
@@
 The Fine-Gray model allows to directly assess the effect of a covariate on the subdistribution hazard, as opposed to the cause-specific hazard.
-For the computation, we use the \pkg{timereg} package [@timereg]:
+For the computation, we use the \CRANpkg{timereg} package [@timereg]:
 
@@
 
-For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function as given by \eqref{eq:penest}.
+For the third case study, we show how \CRANpkg{casebase} can also be used for variable selection through regularized estimation of the hazard function as given by \eqref{eq:penest}.
 To illustrate this functionality, we use the dataset from the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) [@knaus1995support].
@@
 The original data is available online from the Department of Biostatistics at Vanderbilt University [@harrell_2020].
-The cleaned and imputed data consists of 9104 observations and 30 variables, and it is available as part of the \pkg{casebase} package.
+The cleaned and imputed data consists of 9104 observations and 30 variables, and it is available as part of the \CRANpkg{casebase} package.
 In the comparisons below, all covariates except \code{sps} and \code{aps} were used.
@@
 Equation \ref{eq:penest}) of 0 to the time variable to ensure it is in the selected model.
-We compare our approach to both Cox regression, and lasso penalized Cox regression (fitted via the \pkg{glmnet} package).
+We compare our approach to both Cox regression, and lasso penalized Cox regression (fitted via the \CRANpkg{glmnet} package).
 
@@
 The Brier score can be used with both parametric and semi-parametric models.
-We use the \pkg{riskRegression} package to compute these scores for all models.
+We use the \CRANpkg{riskRegression} package to compute these scores for all models.
 
@@
 
-```{r coefplots, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} Coefficient estimates from the Cox model (Cox), penalized Cox model using the \\pkg{glmnet} package (Pen.
+```{r coefplots, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} Coefficient estimates from the Cox model (Cox), penalized Cox model using the \\CRANpkg{glmnet} package (Pen.
 Cox), and our approach using penalized case-base sampling (Pen.
@@
 
-<!-- In this third case study, we showed how case-base sampling can be used in conjunction with penalized logistic regression to perform variable selection in survival models. -->
-
-<!--
-# Case study 4---Stanford Heart Transplant Data
-
-In the previous case studies, we only considered covariates that were fixed at baseline.
-In this next case study, we use the Stanford Heart Transplant data [@clark1971cardiac,@crowley1977covariance] to show how case-base sampling can also be used in the context of time-dependent exposure.
-This feature of case-base sampling has been explored in the literature, in the context of vaccination safety  [@saarela2015case].
-In this study, the exposure period was defined as the week following vaccination.
-Hence, the main covariate of interest, i.e.
-exposure to the vaccine, was changing over time.
-In this context, case-base sampling offers an efficient alternative to nested case-control designs or self-matching.
-
-Recall the setting of Stanford Heart Transplant study: patients were admitted to the Stanford program after meeting with their physician and determining that they were unlikely to respond to other forms of treatment.
-After enrollment, the program searched for a suitable donor for the patient, which could take anywhere between a few days to almost a year.
-We are interested in the effect of a heart transplant on survival; therefore, the patient is considered exposed only after the transplant has occurred.
-
-As above, we can look at the population-time plot for a graphical summary of the event incidence (see Figure \ref{fig:cs4PopTime}.
-Here, we colour the exposed person-time (i.e.
-after transplant) in a darker shade of gray.
-As we can see, most events occur early during the follow-up period, and therefore we do not expect the hazard to be constant.
-
--->
-
-```{r stanford-poptime, eval = eval_cs4, fig.cap="\\label{fig:cs4PopTime} Population-time for the Stanford Transplant study. The darker shade of gray corresponds to the exposed person-time."}
-stanford_popTime <- popTime(jasa,
-  time = "futime",
-  event = "fustat"
-)
-
-# Add exposure time----
-tx_exp <- stanford_popTime %>% 
-    filter(!is.na(tx.date)) %>% 
-    mutate(x = time_length(accept.dt %--% tx.date,
-                           unit = "days"))
-
-plot(stanford_popTime,
-     ribbon.params = list(fill = "gray50")) + 
-    geom_segment(data = tx_exp,
-                 aes(x = x, xend = original.time,
-                     y = ycoord, yend = ycoord),
-                 colour = "gray30", size = 1) + 
-    geom_point(data = filter(stanford_popTime, event == 1),
-               aes(x = time, y = yc, color = "Case series",
-                   fill = "Case series"),
-               size = 1,
-               alpha = 1,
-               shape = 21) +
-  paper_gg_theme
-```
-
-<!--
-Since the exposure is time-dependent, we need to manually define the exposure variable *after* case-base sampling and *before* fitting the hazard function.
-For this reason, we use the `sampleCaseBase` function directly.
--->
-
-```{r stanford_cb, eval = eval_cs4}
-cb_data <- sampleCaseBase(jasa,
-  time = "futime",
-  event = "fustat", ratio = 100
-)
-```
-
-<!--
-Next, we compute the number of days from acceptance into the program to transplant, and we use this variable to determine whether each population-moment is exposed or not.
-
--->
-
-```{r stanford_exp, eval = eval_cs4, dependson="stanford_cb"}
-# Define exposure variable
-cb_data <- mutate(cb_data,
-  txtime = time_length(accept.dt %--% tx.date,
-                       unit = "days"
-  ),
-  exposure = case_when(
-    is.na(txtime) ~ 0L, # No transplant
-    txtime > futime ~ 0L,
-    txtime <= futime ~ 1L
-  )
-)
-```
-
-<!--
-Finally, we can fit the hazard using various linear predictors.
--->
-
-```{r stanford_fit, eval = eval_cs4, warning = FALSE, dependson="stanford_exp"}
-# Fit several models
-fit1 <- fitSmoothHazard(fustat ~ exposure,
-  data = cb_data, time = "futime"
-)
-fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
-  data = cb_data, time = "futime"
-)
-fit3 <- fitSmoothHazard(fustat ~ exposure * futime,
-  data = cb_data, time = "futime"
-)
-```
-
-<!--
-Note that the third model includes an interaction term between exposure and follow-up time.
-In other words, this model no longer exhibits proportional hazards.
-The evidence of non-proportionality of hazards in the Stanford Heart Transplant data has been widely discussed [@arjas1988graphical].
-
-We can then compare the goodness of fit of these three models using the Akaike Information Criterion (AIC).
--->
-
-```{r, eval = eval_cs4, dependson="stanford_fit", digits = 4}
-# Compute AIC
-c("Model1" = AIC(fit1),
-  "Model2" = AIC(fit2),
-  "Model3" = AIC(fit3))
-```
-
-<!--
-As we can see, the best fit is the third model.
-By visualizing the hazard functions for both exposed and unexposed individuals, we can more clearly see how the hazards are no longer proportional.
-We can easily obtain a plot of the hazards by using the `plot.singleEventCB` method:
--->
-
-```{r stanford-hazard, eval = eval_cs4, dependson="stanford_fit", message = FALSE, results="hide", fig.cap="\\label{fig:cs4hazardPlot} Plot of the hazard function for exposed and unexposed individuals."}
-plot(fit3, hazard.params = list(xvar = "futime",
-                                by = "exposure",
-                                alpha = 0.05,
-                                ylab = "Hazard",
-                                data = cb_data))
-```
-
-<!--
-Looking at Figure \ref{fig:cs4hazardPlot}, the non-proportionality seems to be more pronounced at the beginning of follow-up than the end.
-Finally, in Figure \ref{fig:cs4CIF}, we turn these estimates of the hazard function into estimates of the cumulative incidence functions.
--->
-
-```{r stanford-risk, eval = eval_cs4, dependson="stanford_fit", fig.cap="\\label{fig:cs4CIF} Plot of the cumulative incidence function for exposed and unexposed individuals."}
-# Compute absolute risk curves
-newdata <- data.frame(exposure = c(0, 1))
-absrisk <- absoluteRisk(fit3,
-  newdata = newdata,
-  time = seq(0, 425, length.out = 100)
-)
-
-class(absrisk)
-
-plot(absrisk,
-     id.names = c("No Tx", "Tx")) + 
-  ylab("Cumulative Incidence") +
-  xlab("Follow-up time (in days)") +
-  paper_gg_theme
-```
-
-<!--
-As we can see in the above case-study, the \pkg{casebase} package can also be used to model time-varying exposures and non-proportional hazard functions. 
--->
-
 # Discussion
 
-In this article, we presented the \proglang{R} package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating CIFs using case-base sampling.
+In this article, we presented the \proglang{R} package \CRANpkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating CIFs using case-base sampling.
 Our package also provides several functions to produce graphical summaries of the data and the results.
@@
 Similarly, case-base sampling can be combined with quasi-likelihood estimation to fit survival models that can account for the presence of over-dispersion.
-All of these examples illustrate how the case-base sampling framework in general, and the package \pkg{casebase} in particular, allows the user to fit a broad and flexible family of survival functions.
+All of these examples illustrate how the case-base sampling framework in general, and the package \CRANpkg{casebase} in particular, allows the user to fit a broad and flexible family of survival functions.
 
@@
 In the case studies above, we compared the performance of case-base sampling with that of Cox regression and Fine-Gray models.
-In terms of function interface, \pkg{casebase} uses a formula interface that is closer to that of \code{glm}, in that the event variable is the only variable appearing on the left-hand side of the formula.
+In terms of function interface, \CRANpkg{casebase} uses a formula interface that is closer to that of \code{glm}, in that the event variable is the only variable appearing on the left-hand side of the formula.
 By contrast, both \code{survival::coxph} and \code{timereg::comp.risk} use arrays that capture both the event type and time.
@@
 Qualitatively, we showed that by using splines in the linear predictor, all three models yielded similar curves.
-However, the smooth nature of the output of \pkg{casebase} provides a more intuitive interpretation for consumers of these predictions.
+However, the smooth nature of the output of \CRANpkg{casebase} provides a more intuitive interpretation for consumers of these predictions.
 In Table \ref{tab:compCBvsCox}, we provide a side-by-side comparison between the Cox model and case-base sampling.
@@
 Our choice of modeling the log-hazard as a linear function of covariates allows us to develop a simple computational scheme for estimation.
-However, as a downside, it does not allow us to model location and scale parameters separately like the package \pkg{flexsurv}.
+However, as a downside, it does not allow us to model location and scale parameters separately like the package \CRANpkg{flexsurv}.
 For example, if we look at the Weibull distribution as parametrised in \code{stats::pweibull}, the log-hazard function is given by
@@
 where $\alpha,\beta$ are shape and scale parameters, respectively.
-Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} also allows the user to model the scale parameter as a function of covariates.
-Of course, this added flexibility comes at the cost of interpretability: by modeling the log-hazard directly, the parameter estimates from \pkg{casebase} can be interpreted as estimates of log-hazard ratios.
-To improve the flexibility of \pkg{casebase} at capturing the scale of a parametric family, we could replace the logistic regression with its quasi-likelihood counterpart and therefore model over- and under-dispersion with respect to the logistic likelihood.
+Unlike \CRANpkg{casebase}, the approach taken by \CRANpkg{flexsurv} also allows the user to model the scale parameter as a function of covariates.
+Of course, this added flexibility comes at the cost of interpretability: by modeling the log-hazard directly, the parameter estimates from \CRANpkg{casebase} can be interpreted as estimates of log-hazard ratios.
+To improve the flexibility of \CRANpkg{casebase} at capturing the scale of a parametric family, we could replace the logistic regression with its quasi-likelihood counterpart and therefore model over- and under-dispersion with respect to the logistic likelihood.
 We defer the study of the properties and performance of such a model to a future article.
@@
 This type of censoring is very common in longitudinal studies, and many packages (e.g.
-\pkg{SmoothHazard}, \pkg{survival} and \pkg{rstpm2}) provide functions to account for it.
+\CRANpkg{SmoothHazard}, \CRANpkg{survival} and \CRANpkg{rstpm2}) provide functions to account for it.
 Again, we hope to include any resulting methodology as part of this package.
@@
 Saarela and Arjas [-@saarela2015non] also describe how to obtain a posterior distribution for the AUC from their model.
-Their approach could potentially be included in \pkg{casebase}.
+Their approach could potentially be included in \CRANpkg{casebase}.
 Finally, we want to provide more flexibility in how the case-base sampling is performed.
@@
 
-In conclusion, we presented the \proglang{R} package \pkg{casebase} which implements case-base sampling for fitting parametric survival models and for estimating smooth cumulative incidence functions using the framework of generalized linear models.
+In conclusion, we presented the \proglang{R} package \CRANpkg{casebase} which implements case-base sampling for fitting parametric survival models and for estimating smooth cumulative incidence functions using the framework of generalized linear models.
 We strongly believe that its flexibility and its foundation on the familiar logistic regression model will make it appealing to new and established practitioners.
-The \pkg{casebase} package is freely available from the Comprehensive \proglang{R} Archive Network at https://cran.r-project.org/package=casebase.
+The \CRANpkg{casebase} package is freely available from the Comprehensive \proglang{R} Archive Network at https://cran.r-project.org/package=casebase.
 Interested users can visit http://sahirbhatnagar.com/casebase/ for detailed package documentation and vignettes.

### 2021-02-09 analysis/paper/casebase-manuscript.tex
remove coxRidge
fixed links to cran packages usin CRANpkg instead of pkg

@@
 estimates of the cumulative incidence that are smooth-in-time.\\
-In this paper, we present the \textbf{casebase} R package, a
+In this paper, we present the \CRANpkg{casebase} R package, a
 comprehensive and flexible toolkit for parametric survival analysis. We
@@
 
-In this article, we present the \pkg{casebase} package
+In this article, we present the \CRANpkg{casebase} package
 \citep{casebase-package} implemented in \textsf{R} \citep{r-core}, for
@@
 Miettinen \citeyearpar{hanley2009fitting} into a simple interface. The
-purpose of the \pkg{casebase} package is to provide practitioners with
-an easy-to-use software tool to compute a patient's risk (or cumulative
-incidence) of an event, conditional on a particular patient's covariate
-profile. Our package retains the flexibility of case-base sampling and
-the familiar interface of the \code{glm} function. It also provides
-tools for variable selection, competing-risk analysis, and dealing with
-time-dependent exposures. In addition, we provide functions for
-exploratory data analysis and visualizing the estimated quantities.
+purpose of the \CRANpkg{casebase} package is to provide practitioners
+with an easy-to-use software tool to compute a patient's risk (or
+cumulative incidence) of an event, conditional on a particular patient's
+covariate profile. Our package retains the flexibility of case-base
+sampling and the familiar interface of the \code{glm} function. It also
+provides tools for variable selection, competing-risk analysis, and
+dealing with time-dependent exposures. In addition, we provide functions
+for exploratory data analysis and visualizing the estimated quantities.
 
@@
 give a short review of existing \textsf{R} packages that implement
-comparable features as \pkg{casebase}. Next, we provide some details
+comparable features as \CRANpkg{casebase}. Next, we provide some details
 about the implementation of case-base sampling in our package, and we
@@
 case studies that illustrate the flexibility and capabilities of
-\pkg{casebase}. We show how the same framework can be used for
+\CRANpkg{casebase}. We show how the same framework can be used for
 non-linear functions of time and non-proportional hazards, competing
@@
 \textsf{R} packages implementing different methodologies. In this
-section, we describe how the functionalities of \pkg{casebase} compare
-to these packages.
-
-At the time of writing, a cursory examination of CRAN's task view on
-survival analysis reveals that there are over 250 packages related to
-survival analysis \citep{survTaskView}. For the purposes of this
-article, we restricted our review to packages that implement at least
-one of the following features: parametric modeling, non-proportional
-hazard models, competing risk analysis, penalized estimation, and CIF
-estimation. By searching for appropriate keywords in the
-\code{DESCRIPTION} file of these packages, we found 60 relevant
-packages. These 60 packages were then manually examined to determine
-which ones are comparable to \pkg{casebase}. In particular, we excluded
-packages that were focused on a different set of problems, such as
-frailty and multi-state models. The remaining 14 packages appear in
-Table \ref{tab:surv-pkgs}, along with some of the functionalities they
-offer.
+section, we describe how the functionalities of \CRANpkg{casebase}
+compare to these packages.
+
+At the time of writing, a cursory examination of CRAN's \ctv{Survival}
+Task View reveals that there are over 250 packages related to survival
+analysis \citep{survTaskView}. For the purposes of this article, we
+restricted our review to packages that implement at least one of the
+following features: parametric modeling, non-proportional hazard models,
+competing risk analysis, penalized estimation, and CIF estimation. By
+searching for appropriate keywords in the \code{DESCRIPTION} file of
+these packages, we found 60 relevant packages. These 60 packages were
+then manually examined to determine which ones are comparable to
+\CRANpkg{casebase}. In particular, we excluded packages that were
+focused on a different set of problems, such as frailty and multi-state
+models. The remaining 14 packages appear in Table \ref{tab:surv-pkgs},
+along with some of the functionalities they offer.
 
 Parametric survival models are implemented in a handful of packages,
-each differing in the underlying data distribution: \pkg{CFC}
-\citeyearpar{mahani2015bayesian}, \pkg{flexsurv} \citeyearpar{flexsurv},
-\pkg{SmoothHazard} \citeyearpar{smoothHazard}, \pkg{rsptm2}
-\citeyearpar{clements_liu}, \pkg{mets}
-\citeyearpar{scheike2014estimating}, and \pkg{survival}
-\citeyearpar{survival-package}. For example, \pkg{SmoothHazard} is
-limited to Weibull distributions \citeyearpar{smoothHazard}, whereas
-both \pkg{flexsurv} and \pkg{survival} allow users to supply any
-distribution of their choice. \pkg{flexsurv}, \pkg{smoothhazard},
-\pkg{mets} and \pkg{rstpm2} can model the effect of time using splines,
-which allows flexible modeling of the hazard function. As discussed
-above, \pkg{casebase} can model any parametric family whose log-hazard
-can be expressed as a linear combination of covariates (including time).
+each differing in the underlying data distribution: \CRANpkg{CFC}
+\citeyearpar{mahani2015bayesian}, \CRANpkg{flexsurv}
+\citeyearpar{flexsurv}, \CRANpkg{SmoothHazard}
+\citeyearpar{smoothHazard}, \CRANpkg{rsptm2} \citeyearpar{clements_liu},
+\CRANpkg{mets} \citeyearpar{scheike2014estimating}, and
+\CRANpkg{survival} \citeyearpar{survival-package}. For example,
+\CRANpkg{SmoothHazard} is limited to Weibull distributions
+\citeyearpar{smoothHazard}, whereas both \CRANpkg{flexsurv} and
+\CRANpkg{survival} allow users to supply any distribution of their
+choice. \CRANpkg{flexsurv}, \CRANpkg{smoothhazard}, \CRANpkg{mets} and
+\CRANpkg{rstpm2} can model the effect of time using splines, which
+allows flexible modeling of the hazard function. As discussed above,
+\CRANpkg{casebase} can model any parametric family whose log-hazard can
+be expressed as a linear combination of covariates (including time).
 Therefore, our package is more general in that it allows the user to
@@
 (non-proportional) time-varying coefficient models. However, unlike
-\pkg{flexsurv}, we do not explicitly model any shape parameter.
+\CRANpkg{flexsurv}, we do not explicitly model any shape parameter.
 
 Several packages implement penalized estimation for the Cox model:
-\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
-\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
-\pkg{riskRegression} \citeyearpar{gerds_blanche}. Moreover, some
+\CRANpkg{glmnet} \citeyearpar{regpathcox}, \CRANpkg{glmpath}
+\citeyearpar{park_hastie}, \CRANpkg{penalized} \citeyearpar{l1penal},
+\CRANpkg{riskRegression} \citeyearpar{gerds_blanche}. Moreover, some
 packages also include penalized estimation in the context of Cox models
 with time-varying coefficients: elastic-net penalization with
-\pkg{CoxRidge} \citeyearpar{perperoglou} and \pkg{rstpm2}
-\citeyearpar{clements_liu}, while \pkg{survival}
+\CRANpkg{rstpm2} \citeyearpar{clements_liu}, while \CRANpkg{survival}
 \citeyearpar{survival-package} has an implementation of ridge-penalized
-estimation. On the other hand, our package \pkg{casebase} provides
+estimation. On the other hand, our package \CRANpkg{casebase} provides
 penalized estimation of the hazard function. To our knowledge,
-\pkg{casebase} and \pkg{rsptm2} are the only packages to offer this
-functionality.
+\CRANpkg{casebase} and \CRANpkg{rsptm2} are the only packages to offer
+this functionality.
 
@@
 risk analysis; for a different perspective on this topic, see Mahani \&
-Sharabiani \citeyearpar{mahani2015bayesian}. The package \pkg{cmprsk}
-provides methods for cause-specific subdistribution hazards, such as in
-the Fine-Gray model \citeyearpar{fine1999proportional}. On the other
-hand, the package \pkg{CFC} estimates cause-specific CIFs from
-unadjusted, non-parametric survival functions. Our package
-\pkg{casebase} also provides functionalities for competing risk analysis
-by estimating parametrically the cause-specific hazards. From these
-quantities, we can then estimate the cause-specific CIFs.
+Sharabiani \citeyearpar{mahani2015bayesian}. The package
+\CRANpkg{cmprsk} provides methods for cause-specific subdistribution
+hazards, such as in the Fine-Gray model
+\citeyearpar{fine1999proportional}. On the other hand, the package
+\CRANpkg{CFC} estimates cause-specific CIFs from unadjusted,
+non-parametric survival functions. Our package \CRANpkg{casebase} also
+provides functionalities for competing risk analysis by estimating
+parametrically the cause-specific hazards. From these quantities, we can
+then estimate the cause-specific CIFs.
 
@@
 functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI}
-below). Packages providing this functionality include \pkg{CFC},
-\pkg{flexsurv}, \pkg{mets}, and \pkg{survival}. Our package
-\pkg{casebase} also follows this approach for both single-event and
+below). Packages providing this functionality include \CRANpkg{CFC},
+\CRANpkg{flexsurv}, \CRANpkg{mets}, and \CRANpkg{survival}. Our package
+\CRANpkg{casebase} also follows this approach for both single-event and
 competing-risk analyses. The second category outlined above broadly
@@
 parametric estimators (e.g.~spline functions). Packages that implement
-this approach include \pkg{riskRegression}, \pkg{rstpm2}, and
-\pkg{survival}. As mentioned in the introduction, a key distinguishing
-factor between these two approaches is that the first category leads to
-smooth estimates of the cumulative incidence function, whereas the
-second category often produces estimates in the form of stepwise
-functions. Providing smooth estimates of the CIF was one of the main
-motivations for introducing case-base sampling in survival analysis.
+this approach include \CRANpkg{riskRegression}, \CRANpkg{rstpm2}, and
+\CRANpkg{survival}. As mentioned in the introduction, a key
+distinguishing factor between these two approaches is that the first
+category leads to smooth estimates of the cumulative incidence function,
+whereas the second category often produces estimates in the form of
+stepwise functions. Providing smooth estimates of the CIF was one of the
+main motivations for introducing case-base sampling in survival
+analysis.
 
@@
 
-The functions in the \pkg{casebase} package can be divided into two
+The functions in the \CRANpkg{casebase} package can be divided into two
 categories: 1) exploratory data analysis, in the form of population-time
@@
 method call in Step 2 allows further customization by using the
-\pkg{ggplot2} \citep{ggplot2} family of functions, users may choose the
-graphics system of their choice to create population-time plots from the
-object created in Step 1.
+\CRANpkg{ggplot2} \citep{ggplot2} family of functions, users may choose
+the graphics system of their choice to create population-time plots from
+the object created in Step 1.
 
@@
 described in Liu \emph{et al.} \citeyearpar{liu2014recovering}. This
-dataset is available through the \pkg{casebase} package. It contains the
-individual observations for 159,893 men from seven European countries,
-who were between the ages of 55 and 69 years when recruited for the
-trial.
+dataset is available through the \CRANpkg{casebase} package. It contains
+the individual observations for 159,893 men from seven European
+countries, who were between the ages of 55 and 69 years when recruited
+for the trial.
 
@@
   elastic-net \citep{zou2005regularization} penalty (cf.~Case Study 3).
-  This functionality is provided through the \pkg{glmnet} package
+  This functionality is provided through the \CRANpkg{glmnet} package
   \citep{friedman2010jss}.
@@
   \code{gam}: This option provides support for \emph{Generalized
-  Additive Models} via the \pkg{mgcv} package
+  Additive Models} via the \CRANpkg{mgcv} package
   \citep{hastie1987generalized}.
@@
   \code{gbm}: This option provides support for \emph{Gradient Boosted
-  Trees} via the \pkg{gbm} package. This feature is still experimental.
+  Trees} via the \CRANpkg{gbm} package. This feature is still
+  experimental.
 \end{itemize}
@@
 In the case of multiple competing events, the hazard is fitted via
-multinomial regression as performed by the \pkg{VGAM} package. We
+multinomial regression as performed by the \CRANpkg{VGAM} package. We
 selected this package for its ability to fit multinomial regression
@@
 
-In this section, we illustrate the main functions of the \pkg{casebase}
-package through three case studies. Each one showcases a different type
-of analysis. First, we show how to model different functional forms of
-time as well as non-proportional hazards. Then we perform a
-competing-risk analysis and compare our results with the Fine-Gray
-model. The third case study illustrates how to perform variable
-selection in high-dimensional datasets.
+In this section, we illustrate the main functions of the
+\CRANpkg{casebase} package through three case studies. Each one
+showcases a different type of analysis. First, we show how to model
+different functional forms of time as well as non-proportional hazards.
+Then we perform a competing-risk analysis and compare our results with
+the Fine-Gray model. The third case study illustrates how to perform
+variable selection in high-dimensional datasets.
 
@@
 In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from
-\pkg{casebase} on the Cox model CIF. The CIF estimates for the
+\CRANpkg{casebase} on the Cox model CIF. The CIF estimates for the
 exponential model in panel (1) overestimate the cumulative incidence
@@
 time-dependent hazard ratio which can easily be fit with the
-\pkg{casebase} package by including an interaction term with time in the
-model:
+\CRANpkg{casebase} package by including an interaction term with time in
+the model:
 
@@
 first author's website, and it is now available as part of the
-\pkg{casebase} package.
+\CRANpkg{casebase} package.
 
@@
 subdistribution hazard, as opposed to the cause-specific hazard. For the
-computation, we use the \pkg{timereg} package \citep{timereg}:
+computation, we use the \CRANpkg{timereg} package \citep{timereg}:
 
@@
 
-For the third case study, we show how \pkg{casebase} can also be used
-for variable selection through regularized estimation of the hazard
+For the third case study, we show how \CRANpkg{casebase} can also be
+used for variable selection through regularized estimation of the hazard
 function as given by \eqref{eq:penest}. To illustrate this
@@
 and imputed data consists of 9104 observations and 30 variables, and it
-is available as part of the \pkg{casebase} package. In the comparisons
-below, all covariates except \code{sps} and \code{aps} were used. These
-two variables correspond to scores for predicting the outcome that were
-developed as part of the original study. For more information about this
-dataset, the reader is encouraged to look at the documentation in our
-package.
+is available as part of the \CRANpkg{casebase} package. In the
+comparisons below, all covariates except \code{sps} and \code{aps} were
+used. These two variables correspond to scores for predicting the
+outcome that were developed as part of the original study. For more
+information about this dataset, the reader is encouraged to look at the
+documentation in our package.
 
@@
 in the selected model. We compare our approach to both Cox regression,
-and lasso penalized Cox regression (fitted via the \pkg{glmnet}
+and lasso penalized Cox regression (fitted via the \CRANpkg{glmnet}
 package).
@@
 score can be used with both parametric and semi-parametric models. We
-use the \pkg{riskRegression} package to compute these scores for all
+use the \CRANpkg{riskRegression} package to compute these scores for all
 models.
@@
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/coefplots-1} \caption{\label{fig:cs3lolliPlot} Coefficient estimates from the Cox model (Cox), penalized Cox model using the \pkg{glmnet} package (Pen.
+\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/coefplots-1} \caption{\label{fig:cs3lolliPlot} Coefficient estimates from the Cox model (Cox), penalized Cox model using the \CRANpkg{glmnet} package (Pen.
 Cox), and our approach using penalized case-base sampling (Pen.
@@
 
-In this article, we presented the \textsf{R} package \pkg{casebase},
+In this article, we presented the \textsf{R} package \CRANpkg{casebase},
 which provides functions for fitting smooth parametric hazards and
@@
 over-dispersion. All of these examples illustrate how the case-base
-sampling framework in general, and the package \pkg{casebase} in
+sampling framework in general, and the package \CRANpkg{casebase} in
 particular, allows the user to fit a broad and flexible family of
@@
 sampling with that of Cox regression and Fine-Gray models. In terms of
-function interface, \pkg{casebase} uses a formula interface that is
+function interface, \CRANpkg{casebase} uses a formula interface that is
 closer to that of \code{glm}, in that the event variable is the only
@@
 predictor, all three models yielded similar curves. However, the smooth
-nature of the output of \pkg{casebase} provides a more intuitive
+nature of the output of \CRANpkg{casebase} provides a more intuitive
 interpretation for consumers of these predictions. In Table
@@
 However, as a downside, it does not allow us to model location and scale
-parameters separately like the package \pkg{flexsurv}. For example, if
-we look at the Weibull distribution as parametrised in
+parameters separately like the package \CRANpkg{flexsurv}. For example,
+if we look at the Weibull distribution as parametrised in
 \code{stats::pweibull}, the log-hazard function is given by
@@
 where \(\alpha,\beta\) are shape and scale parameters, respectively.
-Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} also allows
-the user to model the scale parameter as a function of covariates. Of
-course, this added flexibility comes at the cost of interpretability: by
-modeling the log-hazard directly, the parameter estimates from
-\pkg{casebase} can be interpreted as estimates of log-hazard ratios. To
-improve the flexibility of \pkg{casebase} at capturing the scale of a
-parametric family, we could replace the logistic regression with its
-quasi-likelihood counterpart and therefore model over- and
-under-dispersion with respect to the logistic likelihood. We defer the
-study of the properties and performance of such a model to a future
-article.
+Unlike \CRANpkg{casebase}, the approach taken by \CRANpkg{flexsurv} also
+allows the user to model the scale parameter as a function of
+covariates. Of course, this added flexibility comes at the cost of
+interpretability: by modeling the log-hazard directly, the parameter
+estimates from \CRANpkg{casebase} can be interpreted as estimates of
+log-hazard ratios. To improve the flexibility of \CRANpkg{casebase} at
+capturing the scale of a parametric family, we could replace the
+logistic regression with its quasi-likelihood counterpart and therefore
+model over- and under-dispersion with respect to the logistic
+likelihood. We defer the study of the properties and performance of such
+a model to a future article.
 
@@
 interval censoring. This type of censoring is very common in
-longitudinal studies, and many packages (e.g.~\pkg{SmoothHazard},
-\pkg{survival} and \pkg{rstpm2}) provide functions to account for it.
-Again, we hope to include any resulting methodology as part of this
-package.
+longitudinal studies, and many packages (e.g.~\CRANpkg{SmoothHazard},
+\CRANpkg{survival} and \CRANpkg{rstpm2}) provide functions to account
+for it. Again, we hope to include any resulting methodology as part of
+this package.
 
@@
 distribution for the AUC from their model. Their approach could
-potentially be included in \pkg{casebase}. Finally, we want to provide
-more flexibility in how the case-base sampling is performed. This could
-be achieved by adding a \code{hazard} argument to the function
-\code{sampleCaseBase}. In this way, users could specify their own
-sampling mechanism. For example, they could provide a hazard that gives
-sampling probabilities that are proportional to the cardiovascular
+potentially be included in \CRANpkg{casebase}. Finally, we want to
+provide more flexibility in how the case-base sampling is performed.
+This could be achieved by adding a \code{hazard} argument to the
+function \code{sampleCaseBase}. In this way, users could specify their
+own sampling mechanism. For example, they could provide a hazard that
+gives sampling probabilities that are proportional to the cardiovascular
 disease event rate given by the Framingham score \citep{saarela2015non}.
 
-In conclusion, we presented the \textsf{R} package \pkg{casebase} which
-implements case-base sampling for fitting parametric survival models and
-for estimating smooth cumulative incidence functions using the framework
-of generalized linear models. We strongly believe that its flexibility
-and its foundation on the familiar logistic regression model will make
-it appealing to new and established practitioners. The \pkg{casebase}
-package is freely available from the Comprehensive \textsf{R} Archive
-Network at \url{https://cran.r-project.org/package=casebase}. Interested
-users can visit \url{http://sahirbhatnagar.com/casebase/} for detailed
-package documentation and vignettes.
+In conclusion, we presented the \textsf{R} package \CRANpkg{casebase}
+which implements case-base sampling for fitting parametric survival
+models and for estimating smooth cumulative incidence functions using
+the framework of generalized linear models. We strongly believe that its
+flexibility and its foundation on the familiar logistic regression model
+will make it appealing to new and established practitioners. The
+\CRANpkg{casebase} package is freely available from the Comprehensive
+\textsf{R} Archive Network at
+\url{https://cran.r-project.org/package=casebase}. Interested users can
+visit \url{http://sahirbhatnagar.com/casebase/} for detailed package
+documentation and vignettes.
 

### 2021-02-09 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
created Makefile to meet RJournal requirments and make a zip file

--- 2021-02-09 analysis/paper/casebase-manuscript.Rmd
@@
 Their approach naturally leads to estimates of the cumulative incidence that are smooth-in-time.
+  | * SRB and MT contributed equally to this work.
   | In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis.
@@
 We illustrate all these features through three different case studies.
-  | * SRB and MT contributed equally to this work.
 preamble: >
@@
 output: rticles::rjournal_article 
-bibliography: casebase-manuscript.bib 
+bibliography: bhatnagar-turgeon-islam-hanley-saarela.bib 
 fig_width: 12
@@
 
-```{r, eval = FALSE, echo=FALSE}
-print.summary.glm <- function(x,digits=max(3,getOption("digits")-3),symbolic.cor=x$symbolic.cor,
-                              signif.stars = getOption("show.signif.stars"),show.call=FALSE,show.resids=FALSE,...) {
-  if (show.call) cat("Call:\n", paste(deparse(x$call), sep = "\n", collapse = "\n"), "\n", sep = "")
-  if (show.resids) {
-    cat("Deviance Residuals: \n")
-    if (x$df.residual > 5) {
-      x$deviance.resid <- stats::quantile(x$deviance.resid, na.rm = TRUE)
-      names(x$deviance.resid) <- c("Min", "1Q", "Median", "3Q", "Max")
-    }
-    xx <- zapsmall(x$deviance.resid, digits + 1)
-    print.default(xx, digits = digits, na.print = "", print.gap = 2)
-    cat("\n")
-  }
-  if (length(x$aliased) == 0L) cat("No Coefficients\n")
-  else {
-    df <- if ("df" %in% names(x)) x[["df"]]
-    else NULL
-    if (!is.null(df) && (nsingular <- df[3L] - df[1L])) 
-      cat("Coefficients: (", nsingular, " not defined because of singularities)\n", sep = "")
-    else cat("\nCoefficients:\n")
-    coefs <- x$coefficients
-    if (!is.null(aliased <- x$aliased) && any(aliased)) {
-      cn <- names(aliased)
-      coefs <- matrix(NA, length(aliased), 4L, dimnames = list(cn,colnames(coefs)))
-      coefs[!aliased, ] <- x$coefficients
-    }
-    stats::printCoefmat(coefs, digits = digits, signif.stars = signif.stars,na.print = "NA", ...)
-  }
-  cat("\n(Dispersion parameter for ", x$family$family, " family taken to be ", 
-      format(x$dispersion), ")\n\n", apply(cbind(paste(format(c("Null","Residual"), 
-                                                              justify = "right"), "deviance:"),
-                                                 format(unlist(x[c("null.deviance","deviance")]), digits = max(5, digits + 1)), " on", 
-                                                 format(unlist(x[c("df.null", "df.residual")])), " degrees of freedom\n"),1L,
-                                           paste, collapse = " "), sep = "")
-  if (nzchar(mess <- stats::naprint(x$na.action))) cat("  (", mess, ")\n",sep="")
-  cat("AIC: ",format(x$aic,digits=max(4,digits+1)),"\n\n",
-      "Number of Fisher Scoring iterations: ",x$iter,"\n",sep="")
-  correl <- x$correlation
-  if (!is.null(correl)) {
-    p <- NCOL(correl)
-    if (p > 1) {
-      cat("\nCorrelation of Coefficients:\n")
-      if (is.logical(symbolic.cor) && symbolic.cor) print(stats::symnum(correl,abbr.colnames=NULL))
-      else {
-        correl <- format(round(correl, 2),nsmall=2,digits=digits)
-        correl[!lower.tri(correl)] <- ""
-        print(correl[-1,-p,drop=FALSE],quote=FALSE)
-      }
-    }
-  }
-  cat("\n")
-  invisible(x)
-}
-
-print(summary(fits[["splines"]]), signif.stars = F)
-```
-
 Next, the \code{absoluteRisk} function takes as input the \code{singleEventCB} object and returns a matrix where each column corresponds to the covariate profiles specified in the \code{newdata} argument, and each row corresponds to time points specified by the \code{time} argument:
@@
 
-# dev.off()
 layout(matrix(1:4, ncol = 2, byrow = FALSE), respect = FALSE)
@@
      xmax = max(new_time),
-     # xlim = range(new_time), 
      conf.int = FALSE, col = myCols, 
-     # main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
-     #                Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
-     #                exp(coef(cox_model)), 
-     #                exp(confint(cox_model))[1], 
-     #                exp(confint(cox_model))[2]), 
      lwd = 2,
@@
      fun = "event",
-     # xlim = c(0,15), 
      xmax = max(new_time),
@@
       xaxt = 'n', yaxt = 'n',
-     # xlim = c(0,15), 
      xmax = max(new_time),
@@
      yaxt = 'n',
-     # xlim = c(0,15), 
      xmax = max(new_time),

### 2021-02-09 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
created Makefile to meet RJournal requirments and make a zip file

--- 2021-02-09 analysis/paper/casebase-manuscript.tex
@@
 % !TeX root = RJwrapper.tex
-\title{\pkg{casebase}: An Alternative Framework For Survival Analysis
-and Comparison of Event Rates}
-\author{by Sahir Rai Bhatnagar*, Maxime Turgeon*, Jesse Islam, James A.
-Hanley, and Olli Saarela}
+\title{\pkg{casebase}: An Alternative Framework For Survival Analysis and
+Comparison of Event Rates}
+\author{by Sahir Rai Bhatnagar*, Maxime Turgeon*, Jesse Islam, James A. Hanley, and Olli Saarela}
 
@@
 estimates of the cumulative incidence that are smooth-in-time.\\
+* SRB and MT contributed equally to this work.\\
 In this paper, we present the \CRANpkg{casebase} R package, a
@@
 time-to-event data. We illustrate all these features through three
-different case studies.\\
-* SRB and MT contributed equally to this work.
+different case studies.
 }
@@
 \citep{knaus1995support}. The SUPPORT dataset tracks death in five
-American hospitals within individuals who are considered seriously
-ill.~The original data is available online from the Department of
+American hospitals within individuals who are considered seriously ill.
+The original data is available online from the Department of
 Biostatistics at Vanderbilt University \citep{harrell_2020}. The cleaned
@@
 interval censoring. This type of censoring is very common in
-longitudinal studies, and many packages (e.g.~\CRANpkg{SmoothHazard},
+longitudinal studies, and many packages (e.g. \CRANpkg{SmoothHazard},
 \CRANpkg{survival} and \CRANpkg{rstpm2}) provide functions to account
@@
 
-\bibliography{casebase-manuscript.bib}
+\bibliography{bhatnagar-turgeon-islam-hanley-saarela.bib}
 

### 2021-02-09 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.R

@@ -1,0 +1,789 @@
+## ----setup, echo=FALSE, message=FALSE, warning=FALSE--------------------------
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  warning = FALSE,
+  message = FALSE,
+  echo = FALSE,
+  cache = TRUE, 
+  fig.pos = 'ht',
+  comment = "#>",
+  fig.path = "../figures/",
+  out.width = "\\textwidth",
+   out.extra = "keepaspectratio=true"
+)
+
+options(kableExtra.latex.load_packages = FALSE)
+library(casebase)
+library(colorspace)
+library(cowplot)
+library(data.table)
+library(glmnet)
+library(kableExtra)
+library(magrittr)
+library(pracma)
+library(flexsurv) 
+library(prodlim)
+library(riskRegression)
+library(splines)
+library(survival)
+library(tibble)
+library(visreg)
+library(lubridate)
+library(tidyverse)
+library(scales)
+options(digits = 2)
+# set the seed for reproducible output
+set.seed(1234)
+
+# Turn on/off sections for faster compilation
+eval_introduction <- TRUE
+eval_theory <- TRUE
+eval_implementation <- TRUE
+eval_cs1 <- TRUE
+eval_cs2 <- TRUE
+eval_cs3 <- TRUE
+eval_cs4 <- FALSE
+eval_colophon <- FALSE # Only for draft
+echo_plot_code <- TRUE
+# Define theme for plots
+paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
+# Define colour palette for CS 3
+q4 <- qualitative_hcl(4, palette = "Dark 3")
+names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
+
+
+## ----erspc-data, eval = eval_cs1, echo = c(-1:-3)-----------------------------
+data("ERSPC")
+ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
+                       levels = c(0,1), 
+                       labels = c("Control group", "Screening group"))
+
+pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
+                               event = "DeadOfPrCa", exposure = "ScrArm")
+inherits(pt_object, "popTime")
+attr(pt_object, "exposure")
+
+
+## ----plot-stratified-erspc-data-code, eval = FALSE, echo = TRUE---------------
+#> plot(pt_object, add.base.series = TRUE)
+
+
[…]

### 2021-02-09 bhatnagar-turgeon-islam-hanley-saarela.R

@@ -1,0 +1,153 @@
+## ----setup, echo=FALSE, message=FALSE, warning=FALSE-----------------------------------------
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  warning = FALSE,
+  message = FALSE,
+  echo = FALSE,
+  cache = TRUE, 
+  fig.pos = 'ht',
+  comment = "#>",
+  fig.path = "../figures/",
+  out.width = "\\textwidth",
+   out.extra = "keepaspectratio=true"
+)
+
+options(kableExtra.latex.load_packages = FALSE)
+library(casebase)
+library(colorspace)
+library(cowplot)
+library(data.table)
+library(glmnet)
+library(kableExtra)
+library(magrittr)
+library(pracma)
+library(flexsurv) 
+library(prodlim)
+library(riskRegression)
+library(splines)
+library(survival)
+library(tibble)
+library(visreg)
+library(lubridate)
+library(tidyverse)
+library(scales)
+options(digits = 2)
+# set the seed for reproducible output
+set.seed(1234)
+
+# Turn on/off sections for faster compilation
+eval_introduction <- TRUE
+eval_theory <- TRUE
+eval_implementation <- TRUE
+eval_cs1 <- TRUE
+eval_cs2 <- TRUE
+eval_cs3 <- TRUE
+eval_cs4 <- FALSE
+eval_colophon <- FALSE # Only for draft
+echo_plot_code <- TRUE
+# Define theme for plots
+paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
+# Define colour palette for CS 3
+q4 <- qualitative_hcl(4, palette = "Dark 3")
+names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
+
+
+
+
+## ----plot-stratified-erspc-data-code, eval = FALSE, echo = TRUE------------------------------
+## plot(pt_object, add.base.series = TRUE)
+
+
+
+
+
+
+## ---- eval = TRUE, echo=eval_cs1-------------------------------------------------------------
+summary(fits[["splines"]]) 
+
+
+## ---- eval = FALSE, echo=FALSE---------------------------------------------------------------
+## print.summary.glm <- function(x,digits=max(3,getOption("digits")-3),symbolic.cor=x$symbolic.cor,
+##                               signif.stars = getOption("show.signif.stars"),show.call=FALSE,show.resids=FALSE,...) {
[…]

### 2021-02-09 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
update manuscript

@@
 abstract: |
-  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician and the patient is the risk of an event given their covariate profile.
-However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the likelihood expression or the estimating function.
+  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician is their patient's risk of an event.
+However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the  estimating function.
 The baseline hazard then needs to be estimated separately using a non-parametric approach.
 This leads to step-wise estimates of the cumulative incidence that are difficult to interpret.
-Using case-base sampling, Hanley & Miettinen (2009) explained how the parametric hazard functions can be estimated using logistic regression.
+Using case-base sampling, Hanley & Miettinen (2009) explained how parametric hazard functions can be estimated using logistic regression.
 Their approach naturally leads to estimates of the cumulative incidence that are smooth-in-time.
-  | * SRB and MT contributed equally to this work.
   | In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis.
 We describe how the case-base framework can be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection.
-Our package also includes an extensive array of visualization tools to complement the analysis of time-to-event data.
+Our package also includes an extensive array of visualization tools to complement the analysis.
 We illustrate all these features through three different case studies.
+  | * SRB and MT contributed equally to this work.
 preamble: >
@@
   \DeclareUnicodeCharacter{2500}{-}
-  \newcommand{\proglang}{\textsf}
 output: rticles::rjournal_article 
@@
 ---
-
-<!-- # Code formatting -->
-
-<!-- Don't use markdown, instead use the more precise latex commands: -->
-
-<!-- * \proglang{Java} -->
-<!-- * \pkg{plyr} -->
-<!-- * \code{print("abc")} -->
 
@@
 
-In this article, we present the \CRANpkg{casebase} package [@casebase-package] implemented in \proglang{R} [@r-core], for parametric survival analysis that combines the ideas of Hanley & Miettinen [-@hanley2009fitting] into a simple interface.
-The purpose of the \CRANpkg{casebase} package is to provide practitioners with an easy-to-use software tool to compute a patient's risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile.
+In this article, we present the \CRANpkg{casebase} package [@casebase-package] implemented in R, for parametric survival analysis that combines the ideas of Hanley & Miettinen [-@hanley2009fitting] into a simple interface.
+The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to compute a patient's risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile.
 Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function.
@@
 In what follows, we first recall some theoretical details on case-base sampling and its use for estimating parametric hazard functions.
-We then give a short review of existing \proglang{R} packages that implement comparable features as \CRANpkg{casebase}.
+We then give a short review of existing R packages that implement comparable features as \pkg{casebase}.
 Next, we provide some details about the implementation of case-base sampling in our package, and we give a brief survey of its main functions.
-This is followed by three case studies that illustrate the flexibility and capabilities of \CRANpkg{casebase}.
+This is followed by three case studies that illustrate the flexibility and capabilities of \pkg{casebase}.
 We show how the same framework can be used for non-linear functions of time and non-proportional hazards, competing risks and variable selection via penalized regression.
@@
 In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras.
-Instead, following Aalen \emph{et al} [-@aalen2008survival], we use the placeholder "past" to denote the past history of the corresponding process.
+Instead, following Aalen \textit{et al} [-@aalen2008survival], we use the placeholder "past" to denote the past history of the corresponding process.
 The reader interested in more details can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case].
@@
 In this model, the instantaneous failure rate is independent of $t$.\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
-This is also known as the \textit{memoryless property} \citep{kalbfleisch2011statistical}.}
+This is also known as the \dfn{memoryless property} \citep{kalbfleisch2011statistical}.}
 
@@
 Survival analysis is an important branch of applied statistics and epidemiology.
-Accordingly, there is already a vast ecosystem of \proglang{R} packages implementing different methodologies.
-In this section, we describe how the functionalities of \CRANpkg{casebase} compare to these packages.
+Accordingly, there is already a vast ecosystem of R packages implementing different methodologies.
+In this section, we describe how the functionalities of \pkg{casebase} compare to these packages.
 
@@
 By searching for appropriate keywords in the \code{DESCRIPTION} file of these packages, we found 60 relevant packages.
-These 60 packages were then manually examined to determine which ones are comparable to \CRANpkg{casebase}.
+These 60 packages were then manually examined to determine which ones are comparable to \pkg{casebase}.
 In particular, we excluded packages that were focused on a different set of problems, such as frailty and multi-state models.
@@
 
-Parametric survival models are implemented in a handful of packages, each differing in the underlying data distribution: \CRANpkg{CFC} [-@mahani2015bayesian], \CRANpkg{flexsurv} [-@flexsurv], \CRANpkg{SmoothHazard} [-@smoothHazard], \CRANpkg{rsptm2} [-@clements_liu], \CRANpkg{mets} [-@scheike2014estimating], and \CRANpkg{survival} [-@survival-package].
-For example, \CRANpkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \CRANpkg{flexsurv} and \CRANpkg{survival} allow users to supply any distribution of their choice.
-\CRANpkg{flexsurv}, \CRANpkg{smoothhazard}, \CRANpkg{mets} and \CRANpkg{rstpm2} can model the effect of time using splines, which allows flexible modeling of the hazard function.
-As discussed above, \CRANpkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear combination of covariates (including time).
+Parametric survival models are implemented in a handful of packages, each differing in the underlying data distribution: \CRANpkg{CFC} [-@mahani2015bayesian], \CRANpkg{flexsurv} [-@flexsurv], \CRANpkg{SmoothHazard} [-@smoothHazard], \CRANpkg{rstpm2} [-@clements_liu], \CRANpkg{mets} [-@scheike2014estimating], and \CRANpkg{survival} [-@survival-package].
+For example, \pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \pkg{flexsurv} and \pkg{survival} allow users to supply any distribution of their choice.
+\pkg{flexsurv}, \pkg{SmoothHazard}, \pkg{mets} and \pkg{rstpm2} can model the effect of time using splines, which allows flexible modeling of the hazard function.
+As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear combination of covariates (including time).
 Therefore, our package is more general in that it allows the user to model any linear or non-linear transformation of time including splines and higher order polynomials.
 Also, by including interaction terms between covariates and time, it also allows users to fit (non-proportional) time-varying coefficient models.
-However, unlike \CRANpkg{flexsurv}, we do not explicitly model any shape parameter.
+However, unlike \pkg{flexsurv}, we do not explicitly model any shape parameter.
 
 Several packages implement penalized estimation for the Cox model: \CRANpkg{glmnet} [-@regpathcox], \CRANpkg{glmpath} [-@park_hastie], \CRANpkg{penalized} [-@l1penal], \CRANpkg{riskRegression} [-@gerds_blanche].
-Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \CRANpkg{rstpm2} [-@clements_liu], while \CRANpkg{survival} [-@survival-package] has an implementation of ridge-penalized estimation.
-On the other hand, our package \CRANpkg{casebase} provides penalized estimation of the hazard function.
-To our knowledge, \CRANpkg{casebase} and \CRANpkg{rsptm2} are the only packages to offer this functionality.
-
-Next, several \proglang{R} packages implement methodologies for competing risk analysis; for a different perspective on this topic, see Mahani & Sharabiani [-@mahani2015bayesian].
+Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \pkg{rstpm2} [-@clements_liu], while \pkg{survival} [-@survival-package] has an implementation of ridge-penalized estimation.
+On the other hand, our package \pkg{casebase} provides penalized estimation of the hazard function.
+To our knowledge, \pkg{casebase} and \pkg{rstpm2} are the only packages to offer this functionality.
+
+Next, several R packages implement methodologies for competing risk analysis; for a different perspective on this topic, see Mahani & Sharabiani [-@mahani2015bayesian].
 The package \CRANpkg{cmprsk} provides methods for cause-specific subdistribution hazards, such as in the Fine-Gray model [-@fine1999proportional].
-On the other hand, the package \CRANpkg{CFC} estimates cause-specific CIFs from unadjusted, non-parametric survival functions.
-Our package \CRANpkg{casebase} also provides functionalities for competing risk analysis by estimating parametrically the cause-specific hazards.
+On the other hand, the package \pkg{CFC} estimates cause-specific CIFs from unadjusted, non-parametric survival functions.
+Our package \pkg{casebase} also provides functionalities for competing risk analysis by estimating parametrically the cause-specific hazards.
 From these quantities, we can then estimate the cause-specific CIFs.
@@
 Using this estimate, the survival function and the CIF can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below).
-Packages providing this functionality include \CRANpkg{CFC}, \CRANpkg{flexsurv}, \CRANpkg{mets}, and \CRANpkg{survival}.
-Our package \CRANpkg{casebase} also follows this approach for both single-event and competing-risk analyses.
+Packages providing this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and \pkg{survival}.
+Our package \pkg{casebase} also follows this approach for both single-event and competing-risk analyses.
 The second category outlined above broadly corresponds to semi-parametric models.
@@
 spline functions).
-Packages that implement this approach include \CRANpkg{riskRegression}, \CRANpkg{rstpm2}, and \CRANpkg{survival}.
+Packages that implement this approach include \pkg{riskRegression}, \pkg{rstpm2}, and \pkg{survival}.
 As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category often produces estimates in the form of stepwise functions.
@@
 }
-\caption{Comparison of various \proglang{R} packages for survival analysis based on several defining features.
+\caption{Comparison of various R packages for survival analysis based on several defining features.
 \textbf{Competing Risks}: whether or not an implementation for competing risks is present.
@@
 
-The functions in the \CRANpkg{casebase} package can be divided into two categories: 1) exploratory data analysis, in the form of population-time plots; and 2) parametric modeling of the hazard function.
+The functions in the \pkg{casebase} package can be divided into two categories: 1) exploratory data analysis, in the form of population-time plots; and 2) parametric modeling of the hazard function.
 We strove for compatibility with both \code{data.frame}s and \code{data.table}s; this can be seen in the coding choices we made and the unit tests we wrote.
@@
 To illustrate these functions, we will use data from the European Randomized Study of Prostate Cancer Screening (ERSPC) [@schroder2009screening] which was extracted using the approach described in Liu *et al.* [-@liu2014recovering].
-This dataset is available through the \CRANpkg{casebase} package.
+This dataset is available through the \pkg{casebase} package.
 It contains the individual observations for 159,893 men  from seven European countries, who were between the ages of 55 and 69 years when recruited for the trial.
@@
 Case Study 3).
-This functionality is provided through the \CRANpkg{glmnet} package [@friedman2010jss].
+This functionality is provided through the \pkg{glmnet} package [@friedman2010jss].
   - \code{gam}: This option provides support for *Generalized Additive Models* via the \CRANpkg{mgcv} package [@hastie1987generalized].
@@
 
-In this section, we illustrate the main functions of the \CRANpkg{casebase} package through three case studies.
+In this section, we illustrate the main functions of the \pkg{casebase} package through three case studies.
 Each one showcases a different type of analysis.
@@
 
-In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from \CRANpkg{casebase} on the Cox model CIF.
+In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from \pkg{casebase} on the Cox model CIF.
 The CIF estimates for the exponential model in panel (1) overestimate the cumulative incidence earlier on in time, and underestimate it later on.
@@
 <!--A common choice for this type of analysis is to use a Cox regression model and estimate the hazard ratio for the screening group (relative to the control group).
-In \proglang{R}, it can be done as follows:-->
+In R, it can be done as follows:-->
 
@@
 As a result, the null effects in years 1–7 masked the substantial reductions that began to appear from year 8 onward.
-This motivates the use of a time-dependent hazard ratio which can easily be fit with the \CRANpkg{casebase} package by including an interaction term with time in the model:
+This motivates the use of a time-dependent hazard ratio which can easily be fit with the \pkg{casebase} package by including an interaction term with time in the model:
 
@@
 For illustrative purposes, we use the same data that was used in Scrucca *et al* [-@scrucca2010regression].
-The data was downloaded from the first author's website, and it is now available as part of the \CRANpkg{casebase} package.
+The data was downloaded from the first author's website, and it is now available as part of the \pkg{casebase} package.
 
@@
 
-For the third case study, we show how \CRANpkg{casebase} can also be used for variable selection through regularized estimation of the hazard function as given by \eqref{eq:penest}.
-To illustrate this functionality, we use the dataset from the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) [@knaus1995support].
-The SUPPORT dataset tracks death in five American hospitals within individuals who are considered seriously ill.
-The original data is available online from the Department of Biostatistics at Vanderbilt University [@harrell_2020].
-The cleaned and imputed data consists of 9104 observations and 30 variables, and it is available as part of the \CRANpkg{casebase} package.
+For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function as given by \eqref{eq:penest}.
+To illustrate this functionality, we use the dataset from the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) [@knaus1995support].\footnote{The original data is available online from the Department of Biostatistics at Vanderbilt University: \url{https://biostat.app.vumc.org/wiki/Main/SupportDesc}} The SUPPORT dataset tracks death in five American hospitals within individuals who are considered seriously ill.
+The cleaned and imputed data consists of 9104 observations and 30 variables, and it is available as part of the \pkg{casebase} package.
 In the comparisons below, all covariates except \code{sps} and \code{aps} were used.
@@
 Equation \ref{eq:penest}) of 0 to the time variable to ensure it is in the selected model.
-We compare our approach to both Cox regression, and lasso penalized Cox regression (fitted via the \CRANpkg{glmnet} package).
+We compare our approach to both Cox regression, and lasso penalized Cox regression (fitted via the \pkg{glmnet} package).
 
@@
 The Brier score can be used with both parametric and semi-parametric models.
-We use the \CRANpkg{riskRegression} package to compute these scores for all models.
+We use the \pkg{riskRegression} package to compute these scores for all models.
 
@@
 
-```{r coefplots, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} Coefficient estimates from the Cox model (Cox), penalized Cox model using the \\CRANpkg{glmnet} package (Pen.
+```{r coefplots, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} Coefficient estimates from the Cox model (Cox), penalized Cox model using the \\pkg{glmnet} package (Pen.
 Cox), and our approach using penalized case-base sampling (Pen.
@@
 
-In this article, we presented the \proglang{R} package \CRANpkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating CIFs using case-base sampling.
+In this article, we presented the R package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating CIFs using case-base sampling.
 Our package also provides several functions to produce graphical summaries of the data and the results.
@@
 Similarly, case-base sampling can be combined with quasi-likelihood estimation to fit survival models that can account for the presence of over-dispersion.
-All of these examples illustrate how the case-base sampling framework in general, and the package \CRANpkg{casebase} in particular, allows the user to fit a broad and flexible family of survival functions.
+All of these examples illustrate how the case-base sampling framework in general, and the package \pkg{casebase} in particular, allows the user to fit a broad and flexible family of survival functions.
 
@@
 In the case studies above, we compared the performance of case-base sampling with that of Cox regression and Fine-Gray models.
-In terms of function interface, \CRANpkg{casebase} uses a formula interface that is closer to that of \code{glm}, in that the event variable is the only variable appearing on the left-hand side of the formula.
+In terms of function interface, \pkg{casebase} uses a formula interface that is closer to that of \code{glm}, in that the event variable is the only variable appearing on the left-hand side of the formula.
 By contrast, both \code{survival::coxph} and \code{timereg::comp.risk} use arrays that capture both the event type and time.
@@
 Qualitatively, we showed that by using splines in the linear predictor, all three models yielded similar curves.
-However, the smooth nature of the output of \CRANpkg{casebase} provides a more intuitive interpretation for consumers of these predictions.
+However, the smooth nature of the output of \pkg{casebase} provides a more intuitive interpretation for consumers of these predictions.
 In Table \ref{tab:compCBvsCox}, we provide a side-by-side comparison between the Cox model and case-base sampling.
 
-\begin{table}
+\begin{widetable}
 \caption{\label{tab:compCBvsCox}Comparison between the Cox model and case-base sampling}
@@
 \end{tabular}
-\end{table}
+\end{widetable}
 
 Our choice of modeling the log-hazard as a linear function of covariates allows us to develop a simple computational scheme for estimation.
-However, as a downside, it does not allow us to model location and scale parameters separately like the package \CRANpkg{flexsurv}.
+However, as a downside, it does not allow us to model location and scale parameters separately like the package \pkg{flexsurv}.
 For example, if we look at the Weibull distribution as parametrised in \code{stats::pweibull}, the log-hazard function is given by
@@
 where $\alpha,\beta$ are shape and scale parameters, respectively.
-Unlike \CRANpkg{casebase}, the approach taken by \CRANpkg{flexsurv} also allows the user to model the scale parameter as a function of covariates.
-Of course, this added flexibility comes at the cost of interpretability: by modeling the log-hazard directly, the parameter estimates from \CRANpkg{casebase} can be interpreted as estimates of log-hazard ratios.
-To improve the flexibility of \CRANpkg{casebase} at capturing the scale of a parametric family, we could replace the logistic regression with its quasi-likelihood counterpart and therefore model over- and under-dispersion with respect to the logistic likelihood.
+Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} also allows the user to model the scale parameter as a function of covariates.
+Of course, this added flexibility comes at the cost of interpretability: by modeling the log-hazard directly, the parameter estimates from \pkg{casebase} can be interpreted as estimates of log-hazard ratios.
+To improve the flexibility of \pkg{casebase} at capturing the scale of a parametric family, we could replace the logistic regression with its quasi-likelihood counterpart and therefore model over- and under-dispersion with respect to the logistic likelihood.
 We defer the study of the properties and performance of such a model to a future article.
@@
 This type of censoring is very common in longitudinal studies, and many packages (e.g.
-\CRANpkg{SmoothHazard}, \CRANpkg{survival} and \CRANpkg{rstpm2}) provide functions to account for it.
+\pkg{SmoothHazard}, \pkg{survival} and \pkg{rstpm2}) provide functions to account for it.
 Again, we hope to include any resulting methodology as part of this package.
@@
 Saarela and Arjas [-@saarela2015non] also describe how to obtain a posterior distribution for the AUC from their model.
-Their approach could potentially be included in \CRANpkg{casebase}.
+Their approach could potentially be included in \pkg{casebase}.
 Finally, we want to provide more flexibility in how the case-base sampling is performed.
@@
 
-In conclusion, we presented the \proglang{R} package \CRANpkg{casebase} which implements case-base sampling for fitting parametric survival models and for estimating smooth cumulative incidence functions using the framework of generalized linear models.
+In conclusion, we presented the R package \pkg{casebase} which implements case-base sampling for fitting parametric survival models and for estimating smooth cumulative incidence functions using the framework of generalized linear models.
 We strongly believe that its flexibility and its foundation on the familiar logistic regression model will make it appealing to new and established practitioners.
-The \CRANpkg{casebase} package is freely available from the Comprehensive \proglang{R} Archive Network at https://cran.r-project.org/package=casebase.
-Interested users can visit http://sahirbhatnagar.com/casebase/ for detailed package documentation and vignettes.
-
-# Acknowledgments {-}
+The \pkg{casebase} package is freely available from the Comprehensive R Archive Network at \url{https://cran.r-project.org/package=casebase}.
+Interested users can visit \url{http://sahirbhatnagar.com/casebase/} for detailed package documentation and vignettes.
+
+# Acknowledgments
 
@@
 
-<!-- # Environment Details {-} -->
+<!-- # Environment Details -->
 

### 2021-02-09 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
update manuscript

--- 2021-02-07 analysis/paper/casebase-manuscript.tex
@@
 In epidemiological studies of time-to-event data, a quantity of interest
-to the clinician and the patient is the risk of an event given their
-covariate profile. However, methods relying on time matching or risk-set
-sampling (including Cox regression) eliminate the baseline hazard from
-the likelihood expression or the estimating function. The baseline
+to the clinician is their patient's risk of an event. However, methods
+relying on time matching or risk-set sampling (including Cox regression)
+eliminate the baseline hazard from the estimating function. The baseline
 hazard then needs to be estimated separately using a non-parametric
@@
 that are difficult to interpret. Using case-base sampling, Hanley \&
-Miettinen (2009) explained how the parametric hazard functions can be
+Miettinen (2009) explained how parametric hazard functions can be
 estimated using logistic regression. Their approach naturally leads to
 estimates of the cumulative incidence that are smooth-in-time.\\
-In this paper, we present the \textbf{casebase} R package, a
+In this paper, we present the \CRANpkg{casebase} R package, a
 comprehensive and flexible toolkit for parametric survival analysis. We
@@
 competing risks, and variable selection. Our package also includes an
-extensive array of visualization tools to complement the analysis of
-time-to-event data. We illustrate all these features through three
-different case studies.\\
+extensive array of visualization tools to complement the analysis. We
+illustrate all these features through three different case studies.\\
 * SRB and MT contributed equally to this work.
@@
 
-In this article, we present the \pkg{casebase} package
-\citep{casebase-package} implemented in \textsf{R} \citep{r-core}, for
-parametric survival analysis that combines the ideas of Hanley \&
-Miettinen \citeyearpar{hanley2009fitting} into a simple interface. The
-purpose of the \pkg{casebase} package is to provide practitioners with
-an easy-to-use software tool to compute a patient's risk (or cumulative
+In this article, we present the \CRANpkg{casebase} package
+\citep{casebase-package} implemented in R, for parametric survival
+analysis that combines the ideas of Hanley \& Miettinen
+\citeyearpar{hanley2009fitting} into a simple interface. The purpose of
+the \pkg{casebase} package is to provide practitioners with an
+easy-to-use software tool to compute a patient's risk (or cumulative
 incidence) of an event, conditional on a particular patient's covariate
@@
 sampling and its use for estimating parametric hazard functions. We then
-give a short review of existing \textsf{R} packages that implement
-comparable features as \pkg{casebase}. Next, we provide some details
-about the implementation of case-base sampling in our package, and we
-give a brief survey of its main functions. This is followed by three
-case studies that illustrate the flexibility and capabilities of
-\pkg{casebase}. We show how the same framework can be used for
-non-linear functions of time and non-proportional hazards, competing
-risks and variable selection via penalized regression. Finally, we end
-the article with a discussion of the results and of future directions.
+give a short review of existing R packages that implement comparable
+features as \pkg{casebase}. Next, we provide some details about the
+implementation of case-base sampling in our package, and we give a brief
+survey of its main functions. This is followed by three case studies
+that illustrate the flexibility and capabilities of \pkg{casebase}. We
+show how the same framework can be used for non-linear functions of time
+and non-proportional hazards, competing risks and variable selection via
+penalized regression. Finally, we end the article with a discussion of
+the results and of future directions.
 
@@
 slightly and omit any mention of \(\sigma\)-algebras. Instead, following
-Aalen \emph{et al} \citeyearpar{aalen2008survival}, we use the
+Aalen \textit{et al} \citeyearpar{aalen2008survival}, we use the
 placeholder ``past'' to denote the past history of the corresponding
@@
 \(t\).\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
-This is also known as the \textit{memoryless property} \citep{kalbfleisch2011statistical}.}
+This is also known as the \dfn{memoryless property} \citep{kalbfleisch2011statistical}.}
 
@@
 Survival analysis is an important branch of applied statistics and
-epidemiology. Accordingly, there is already a vast ecosystem of
-\textsf{R} packages implementing different methodologies. In this
-section, we describe how the functionalities of \pkg{casebase} compare
-to these packages.
-
-At the time of writing, a cursory examination of CRAN's task view on
-survival analysis reveals that there are over 250 packages related to
-survival analysis \citep{survTaskView}. For the purposes of this
-article, we restricted our review to packages that implement at least
-one of the following features: parametric modeling, non-proportional
-hazard models, competing risk analysis, penalized estimation, and CIF
-estimation. By searching for appropriate keywords in the
-\code{DESCRIPTION} file of these packages, we found 60 relevant
-packages. These 60 packages were then manually examined to determine
-which ones are comparable to \pkg{casebase}. In particular, we excluded
-packages that were focused on a different set of problems, such as
-frailty and multi-state models. The remaining 14 packages appear in
-Table \ref{tab:surv-pkgs}, along with some of the functionalities they
-offer.
+epidemiology. Accordingly, there is already a vast ecosystem of R
+packages implementing different methodologies. In this section, we
+describe how the functionalities of \pkg{casebase} compare to these
+packages.
+
+At the time of writing, a cursory examination of CRAN's \ctv{Survival}
+Task View reveals that there are over 250 packages related to survival
+analysis \citep{survTaskView}. For the purposes of this article, we
+restricted our review to packages that implement at least one of the
+following features: parametric modeling, non-proportional hazard models,
+competing risk analysis, penalized estimation, and CIF estimation. By
+searching for appropriate keywords in the \code{DESCRIPTION} file of
+these packages, we found 60 relevant packages. These 60 packages were
+then manually examined to determine which ones are comparable to
+\pkg{casebase}. In particular, we excluded packages that were focused on
+a different set of problems, such as frailty and multi-state models. The
+remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with
+some of the functionalities they offer.
 
 Parametric survival models are implemented in a handful of packages,
-each differing in the underlying data distribution: \pkg{CFC}
-\citeyearpar{mahani2015bayesian}, \pkg{flexsurv} \citeyearpar{flexsurv},
-\pkg{SmoothHazard} \citeyearpar{smoothHazard}, \pkg{rsptm2}
-\citeyearpar{clements_liu}, \pkg{mets}
-\citeyearpar{scheike2014estimating}, and \pkg{survival}
-\citeyearpar{survival-package}. For example, \pkg{SmoothHazard} is
-limited to Weibull distributions \citeyearpar{smoothHazard}, whereas
-both \pkg{flexsurv} and \pkg{survival} allow users to supply any
-distribution of their choice. \pkg{flexsurv}, \pkg{smoothhazard},
-\pkg{mets} and \pkg{rstpm2} can model the effect of time using splines,
-which allows flexible modeling of the hazard function. As discussed
-above, \pkg{casebase} can model any parametric family whose log-hazard
-can be expressed as a linear combination of covariates (including time).
-Therefore, our package is more general in that it allows the user to
-model any linear or non-linear transformation of time including splines
-and higher order polynomials. Also, by including interaction terms
-between covariates and time, it also allows users to fit
-(non-proportional) time-varying coefficient models. However, unlike
-\pkg{flexsurv}, we do not explicitly model any shape parameter.
+each differing in the underlying data distribution: \CRANpkg{CFC}
+\citeyearpar{mahani2015bayesian}, \CRANpkg{flexsurv}
+\citeyearpar{flexsurv}, \CRANpkg{SmoothHazard}
+\citeyearpar{smoothHazard}, \CRANpkg{rstpm2} \citeyearpar{clements_liu},
+\CRANpkg{mets} \citeyearpar{scheike2014estimating}, and
+\CRANpkg{survival} \citeyearpar{survival-package}. For example,
+\pkg{SmoothHazard} is limited to Weibull distributions
+\citeyearpar{smoothHazard}, whereas both \pkg{flexsurv} and
+\pkg{survival} allow users to supply any distribution of their choice.
+\pkg{flexsurv}, \pkg{SmoothHazard}, \pkg{mets} and \pkg{rstpm2} can
+model the effect of time using splines, which allows flexible modeling
+of the hazard function. As discussed above, \pkg{casebase} can model any
+parametric family whose log-hazard can be expressed as a linear
+combination of covariates (including time). Therefore, our package is
+more general in that it allows the user to model any linear or
+non-linear transformation of time including splines and higher order
+polynomials. Also, by including interaction terms between covariates and
+time, it also allows users to fit (non-proportional) time-varying
+coefficient models. However, unlike \pkg{flexsurv}, we do not explicitly
+model any shape parameter.
 
 Several packages implement penalized estimation for the Cox model:
-\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
-\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
-\pkg{riskRegression} \citeyearpar{gerds_blanche}. Moreover, some
+\CRANpkg{glmnet} \citeyearpar{regpathcox}, \CRANpkg{glmpath}
+\citeyearpar{park_hastie}, \CRANpkg{penalized} \citeyearpar{l1penal},
+\CRANpkg{riskRegression} \citeyearpar{gerds_blanche}. Moreover, some
 packages also include penalized estimation in the context of Cox models
 with time-varying coefficients: elastic-net penalization with
-\pkg{CoxRidge} \citeyearpar{perperoglou} and \pkg{rstpm2}
-\citeyearpar{clements_liu}, while \pkg{survival}
+\pkg{rstpm2} \citeyearpar{clements_liu}, while \pkg{survival}
 \citeyearpar{survival-package} has an implementation of ridge-penalized
@@
 penalized estimation of the hazard function. To our knowledge,
-\pkg{casebase} and \pkg{rsptm2} are the only packages to offer this
+\pkg{casebase} and \pkg{rstpm2} are the only packages to offer this
 functionality.
 
-Next, several \textsf{R} packages implement methodologies for competing
-risk analysis; for a different perspective on this topic, see Mahani \&
-Sharabiani \citeyearpar{mahani2015bayesian}. The package \pkg{cmprsk}
-provides methods for cause-specific subdistribution hazards, such as in
-the Fine-Gray model \citeyearpar{fine1999proportional}. On the other
-hand, the package \pkg{CFC} estimates cause-specific CIFs from
-unadjusted, non-parametric survival functions. Our package
-\pkg{casebase} also provides functionalities for competing risk analysis
-by estimating parametrically the cause-specific hazards. From these
-quantities, we can then estimate the cause-specific CIFs.
+Next, several R packages implement methodologies for competing risk
+analysis; for a different perspective on this topic, see Mahani \&
+Sharabiani \citeyearpar{mahani2015bayesian}. The package
+\CRANpkg{cmprsk} provides methods for cause-specific subdistribution
+hazards, such as in the Fine-Gray model
+\citeyearpar{fine1999proportional}. On the other hand, the package
+\pkg{CFC} estimates cause-specific CIFs from unadjusted, non-parametric
+survival functions. Our package \pkg{casebase} also provides
+functionalities for competing risk analysis by estimating parametrically
+the cause-specific hazards. From these quantities, we can then estimate
+the cause-specific CIFs.
 
@@
 }
-\caption{Comparison of various \textsf{R} packages for survival analysis based on several defining features.
+\caption{Comparison of various R packages for survival analysis based on several defining features.
 \textbf{Competing Risks}: whether or not an implementation for competing risks is present.
@@
 Preferences Outcomes and Risks of Treatment (SUPPORT)
-\citep{knaus1995support}. The SUPPORT dataset tracks death in five
-American hospitals within individuals who are considered seriously
-ill.~The original data is available online from the Department of
-Biostatistics at Vanderbilt University \citep{harrell_2020}. The cleaned
-and imputed data consists of 9104 observations and 30 variables, and it
-is available as part of the \pkg{casebase} package. In the comparisons
-below, all covariates except \code{sps} and \code{aps} were used. These
-two variables correspond to scores for predicting the outcome that were
+\citep{knaus1995support}.\footnote{The original data is available online from the Department of Biostatistics at Vanderbilt University: \url{https://biostat.app.vumc.org/wiki/Main/SupportDesc}}
+The SUPPORT dataset tracks death in five American hospitals within
+individuals who are considered seriously ill. The cleaned and imputed
+data consists of 9104 observations and 30 variables, and it is available
+as part of the \pkg{casebase} package. In the comparisons below, all
+covariates except \code{sps} and \code{aps} were used. These two
+variables correspond to scores for predicting the outcome that were
 developed as part of the original study. For more information about this
@@
 
-In this article, we presented the \textsf{R} package \pkg{casebase},
-which provides functions for fitting smooth parametric hazards and
-estimating CIFs using case-base sampling. Our package also provides
-several functions to produce graphical summaries of the data and the
-results. We outlined the theoretical underpinnings of the approach, we
-provided details about our implementation, and we illustrated the merits
-of the approach and the package through three case studies.
+In this article, we presented the R package \pkg{casebase}, which
+provides functions for fitting smooth parametric hazards and estimating
+CIFs using case-base sampling. Our package also provides several
+functions to produce graphical summaries of the data and the results. We
+outlined the theoretical underpinnings of the approach, we provided
+details about our implementation, and we illustrated the merits of the
+approach and the package through three case studies.
 
@@
 
-\begin{table}
+\begin{widetable}
 \caption{\label{tab:compCBvsCox}Comparison between the Cox model and case-base sampling}
@@
 \end{tabular}
-\end{table}
+\end{widetable}
 
@@
 interval censoring. This type of censoring is very common in
-longitudinal studies, and many packages (e.g.~\pkg{SmoothHazard},
+longitudinal studies, and many packages (e.g. \pkg{SmoothHazard},
 \pkg{survival} and \pkg{rstpm2}) provide functions to account for it.
@@
 
-In conclusion, we presented the \textsf{R} package \pkg{casebase} which
+In conclusion, we presented the R package \pkg{casebase} which
 implements case-base sampling for fitting parametric survival models and
@@
 it appealing to new and established practitioners. The \pkg{casebase}
-package is freely available from the Comprehensive \textsf{R} Archive
-Network at \url{https://cran.r-project.org/package=casebase}. Interested
-users can visit \url{http://sahirbhatnagar.com/casebase/} for detailed
-package documentation and vignettes.
+package is freely available from the Comprehensive R Archive Network at
+\url{https://cran.r-project.org/package=casebase}. Interested users can
+visit \url{http://sahirbhatnagar.com/casebase/} for detailed package
+documentation and vignettes.
 
 \hypertarget{acknowledgments}{%
-\section*{Acknowledgments}\label{acknowledgments}}
-\addcontentsline{toc}{section}{Acknowledgments}
+\section{Acknowledgments}\label{acknowledgments}}
 

### 2021-02-09 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex

@@
 % !TeX root = RJwrapper.tex
-\title{\pkg{casebase}: An Alternative Framework For Survival Analysis and
-Comparison of Event Rates}
-\author{by Sahir Rai Bhatnagar*, Maxime Turgeon*, Jesse Islam, James A. Hanley, and Olli Saarela}
+\title{\pkg{casebase}: An Alternative Framework For Survival Analysis
+and Comparison of Event Rates}
+\author{by Sahir Rai Bhatnagar*, Maxime Turgeon*, Jesse Islam, James A.
+Hanley, and Olli Saarela}
 
@@
 The SUPPORT dataset tracks death in five American hospitals within
-individuals who are considered seriously ill. The cleaned and imputed
+individuals who are considered seriously ill.~The cleaned and imputed
 data consists of 9104 observations and 30 variables, and it is available
@@
 interval censoring. This type of censoring is very common in
-longitudinal studies, and many packages (e.g. \pkg{SmoothHazard},
+longitudinal studies, and many packages (e.g.~\pkg{SmoothHazard},
 \pkg{survival} and \pkg{rstpm2}) provide functions to account for it.

### 2021-02-09 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
fix Makefile

@@
   comment = "#>",
-  fig.path = "../figures/",
+  fig.path = "./",
   out.width = "\\textwidth",

### 2021-02-09 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
fix Makefile

--- 2021-02-09 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
@@
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/plot-stratified-erspc-data-1} \caption[Population time plots for both treatment arms in the ERSPC dataset]{Population time plots for both treatment arms in the ERSPC dataset.
+\includegraphics[width=\textwidth,keepaspectratio=true]{./plot-stratified-erspc-data-1} \caption[Population time plots for both treatment arms in the ERSPC dataset]{Population time plots for both treatment arms in the ERSPC dataset.
 The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely thin rectangles (person-moments).
@@
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/erspc-cox-cif-1} \caption{CIFs for control and screening groups in the ERSPC data.
+\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cox-cif-1} \caption{CIFs for control and screening groups in the ERSPC data.
 In each of the panels, we plot the CIF from the Cox model using \code{survival::survfit} (solid line) and the CIF from the case-base sampling scheme (dashed line) with different functional forms of time.
@@
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/interaction-ERSPC-1} \caption[Estimated hazard ratio and 95\% confidence interval for screening vs]{Estimated hazard ratio and 95\% confidence interval for screening vs.
+\includegraphics[width=\textwidth,keepaspectratio=true]{./interaction-ERSPC-1} \caption[Estimated hazard ratio and 95\% confidence interval for screening vs]{Estimated hazard ratio and 95\% confidence interval for screening vs.
 control group as a function of time in the ERSPC dataset.
@@
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/compPop-1} \caption[Population-time plot for the stem-cell transplant study with both relapse and competing events]{Population-time plot for the stem-cell transplant study with both relapse and competing events.}\label{fig:compPop}
+\includegraphics[width=\textwidth,keepaspectratio=true]{./compPop-1} \caption[Population-time plot for the stem-cell transplant study with both relapse and competing events]{Population-time plot for the stem-cell transplant study with both relapse and competing events.}\label{fig:compPop}
 \end{figure}
@@
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/bmtcrr-risk-1} \caption{\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups.
+\includegraphics[width=\textwidth,keepaspectratio=true]{./bmtcrr-risk-1} \caption{\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups.
 The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:bmtcrr-risk}
@@
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/coefplots-1} \caption{\label{fig:cs3lolliPlot} Coefficient estimates from the Cox model (Cox), penalized Cox model using the \pkg{glmnet} package (Pen.
+\includegraphics[width=\textwidth,keepaspectratio=true]{./coefplots-1} \caption{\label{fig:cs3lolliPlot} Coefficient estimates from the Cox model (Cox), penalized Cox model using the \pkg{glmnet} package (Pen.
 Cox), and our approach using penalized case-base sampling (Pen.
@@
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/riskregressionBrier-1} \caption{\label{fig:cs3FinalBrier} Comparison of Cox regression (Cox), penalized Cox regression (Pen.
+\includegraphics[width=\textwidth,keepaspectratio=true]{./riskregressionBrier-1} \caption{\label{fig:cs3FinalBrier} Comparison of Cox regression (Cox), penalized Cox regression (Pen.
 Cox), penalized case-base sampling estimation (Pen.

### 2021-02-09 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd

@@
 preamble: >
-  \usepackage{amsmath}
-  \usepackage{amssymb}
   \usepackage{longtable}
@@
   \usepackage{float}
-  \usepackage{booktabs}
   \usepackage{makecell}
   \usepackage{tabu}
-  \DeclareUnicodeCharacter{2500}{-}
 output: rticles::rjournal_article 

### 2021-09-23 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
remove exp gomp weib models

@@
   \usepackage{tabu}
-output: rticles::rjournal_article 
+output: 
+  rticles::rjournal_article:
+    number_sections: true
 bibliography: bhatnagar-turgeon-islam-hanley-saarela.bib 
@@
   - \code{gam}: This option provides support for *Generalized Additive Models* via the \CRANpkg{mgcv} package [@hastie1987generalized].
-  - \code{gbm}: This option provides support for *Gradient Boosted Trees* via the \CRANpkg{gbm} package. This feature is still experimental.
 
@@
 For our first case study, we return to the ERSPC study and investigate the differences in risk between the control and screening arms.
-We fit four models that differ in which functional form of time is used: 1) excluded from the linear predictor (Eq.
-\eqref{eq:exp}); 2) linear function (Eq.
-\eqref{eq:gomp}); 3) log function (Eq.
-\eqref{eq:weibull}); and 4) a smooth function using cubic B-splines.
-The models are fit using \code{fitSmoothHazard} with the familiar formula interface:
+We fit a flexible hazard by using a smooth function of time modeled with cubic B-splines.
+The model is fit using \code{fitSmoothHazard} with the familiar formula interface:
 
@@
 set.seed(12)
-fmla <- list(exponential = formula(DeadOfPrCa ~ ScrArm),
-             gompertz = formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
-             weibull = formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
-             splines = formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
-
-fits <- lapply(fmla, function(form) {
-  fitSmoothHazard(form, data = ERSPC, ratio = 100)
-})
+library(splines)
+fit <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
+                       data = ERSPC, ratio = 100)
 ```
@@
 The output object from \code{fitSmoothHazard} inherits from the \code{singleEventCB} and \code{glm} classes.
-For this reason, we leveraged the \code{summary} method for \code{glm} objects to output a familiar summary of the results:
+For this reason, we can leverage the \code{summary} method for \code{glm} objects to output a familiar summary of the results:
 
 ```{r, eval = TRUE, echo=eval_cs1}
-summary(fits[["splines"]]) 
+summary(fit) 
 ```
@@
 
-```{r, eval=FALSE, echo=echo_plot_code}
+```{r, eval=eval_cs1, echo=echo_plot_code}
 new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
-new_time <- seq(0, 14, by = 0.1)
-
-risks <- lapply(fits, function(fit) {
+new_time <- c(0, 1)
+
 absoluteRisk(fit, time = new_time, newdata = new_data)
-})
-```
-
-In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from \pkg{casebase} on the Cox model CIF.
-The CIF estimates for the exponential model in panel (1) overestimate the cumulative incidence earlier on in time, and underestimate it later on.
-Based on our earlier discussion of the population-time plot, this poor fit for the exponential hazard was expected.
-We notice a better fit with increasing complexity of our model for time in Figure \ref{fig:erspc-cox-cif} (panels 2--4).
-As noted in Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models.
-Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) spline model:
+```
+
+In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIF from \pkg{casebase} on the Cox model CIF.
+As noted in the Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models.
+Therefore, we can easily test the significance of the spline model:
 
 ```{r erscp-compare, eval = eval_cs1, echo = FALSE}
-anova(fits$exponential, fits$splines, test = "LRT")
-```
-
-The null hypothesis is rejected in favor of the spline model. Similarly, the AIC provides further evidence that the flexible function of time provides the best fit:
-
-```{r erscp-aic, eval = eval_cs1, echo=FALSE}
-c("Exp." = AIC(fits$exponential),
-  "Gompertz" = AIC(fits$gompertz),
-  "Weibull" = AIC(fits$weibull),
-  "Splines" = AIC(fits$splines))
-```
+anova(fit, test = "LRT")
+```
+
+The null hypothesis is rejected in favor of the spline model.
+Similarly, to compare different models (e.g.
+time modeled linearly), we could compute Akaike's Information Criterion (AIC) for each model and compare them.
 
@@
 
-```{r erspc-cox-cif, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="CIFs for control and screening groups in the ERSPC data.
-In each of the panels, we plot the CIF from the Cox model using \\code{survival::survfit} (solid line) and the CIF from the case-base sampling scheme (dashed line) with different functional forms of time.
-(1) The time variable is excluded (exponential).
-(2) Linear function of time (Gompertz).
-(3) The natural logarithm (Weibull).
-(4) Cubic B-spline expansion of time."}
+```{r erspc-cox-cif, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="CIF for control and screening groups in the ERSPC data.
+We plot the CIF from the Cox model using \\code{survival::survfit} (solid line) and the CIF from case-base sampling (dashed line).
+We used a cubic B-spline expansion of time."}
 cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
@@
 new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
-new_time <- seq(0,14.5,0.1)
-
-risks <- lapply(fits, function(i) {
-  casebase::absoluteRisk(object = i, time = new_time, newdata = new_data, method = "mont")
-})
-
-layout(matrix(1:4, ncol = 2, byrow = FALSE), respect = FALSE)
-par(mar = c(0, 4.1, 4.1, 0))
+new_time <- seq(0,14,0.1)
+# Recompute with more time points
+risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
+
 plot(survfit(cox_model, newdata = new_data),
-     ylab = "",
-     xlab = "",
+     ylab = "Cumulative incidence (%)",
+     xlab = "Years since randomization",
      fun = "event",
@@
        bg = "gray90")
-lines(risks$exponential[,1], risks$exponential[,2], col = myCols[1], lty = 2, lwd = 2)
-lines(risks$exponential[,1], risks$exponential[,3], col = myCols[2], lty = 2, lwd = 2)
-text(x = max(new_time)*0.97, y = 0.05/100, labels = "(1)", cex = 1.5)
-
-par(mar = c(4.1, 4.1, 0, 0))
-plot(survfit(cox_model, newdata = new_data),
-     ylab = "", xlab = "",
-     fun = "event",
-     xmax = max(new_time),
-     conf.int = FALSE, col = myCols, 
-     lwd = 2,
-     yscale = 100)
-lines(risks$weibull[,1], risks$weibull[,2], col = myCols[1], lty = 2, lwd = 2)
-lines(risks$weibull[,1], risks$weibull[,3], col = myCols[2], lty = 2, lwd = 2)  
-text(x = max(new_time)*0.97, y = 0.05/100, labels = "(3)", cex = 1.5)
-mtext("Cumulative incidence (%)", side = 2, line = -1.5, outer = TRUE)
-
-par(mar = c(0, 0, 4.1, 3.9))
-plot(survfit(cox_model, newdata = new_data),
-     ylab = "",xlab = "",
-     fun = "event",
-      xaxt = 'n', yaxt = 'n',
-     xmax = max(new_time),
-     conf.int = FALSE, col = myCols, 
-     lwd = 2,
-     yscale = 100)
-lines(risks$gompertz[,1], risks$gompertz[,2], col = myCols[1], lty = 2, lwd = 2)
-lines(risks$gompertz[,1], risks$gompertz[,3], col = myCols[2], lty = 2, lwd = 2)
-text(x = max(new_time)*0.97, y = 0.05/100, labels = "(2)", cex = 1.5)
-
-par(mar = c(4.1, 0, 0, 3.9))
-plot(survfit(cox_model, newdata = new_data),
-     ylab = "",xlab = "",
-     fun = "event",
-     yaxt = 'n',
-     xmax = max(new_time),
-     conf.int = FALSE, col = myCols, 
-     lwd = 2,
-     yscale = 100)
-lines(risks$splines[,1], risks$splines[,2], col = myCols[1], lty = 2, lwd = 2)
-lines(risks$splines[,1], risks$splines[,3], col = myCols[2], lty = 2, lwd = 2)
-text(x = max(new_time)*0.97, y = 0.05/100, labels = "(4)", cex = 1.5)
-mtext("Years since randomization", side = 1, line = -1.5, outer = TRUE)
-```
-
-```{r erspc-95ci22, echo = FALSE, eval = eval_cs1}
-CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
-
-aft_fits <- lapply(list("exp","gompertz","weibull"), function(i)
-  flexsurv::flexsurvreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm,
-                           data = ERSPC, dist = i)
-  )
-
-names(aft_fits) <- list("exponential","gompertz","weibull")
-
-# function to exponentiate HR and CI and return a HR (lower, Upper) character
-exp_est_ci <- function(model = c("exponential","gompertz","weibull","splines"),
-                       fit, 
-                       covariate = "ScrArmScreening group", 
-                       type = c("casebase", "survreg")) {
-
-  type <- match.arg(type)
-  covariate <- match.arg(covariate)
-
-  if (type == "casebase"){
-
-    cis <- confint(fit[[model]])
-
-    sprintf("%0.2f (%0.2f, %0.2f)",
-            exp(coef(fit[[model]]))[[covariate]],
-            exp(cis)[covariate,1],
-            exp(cis)[covariate,2])
-  } else {
-
-    if (model == "weibull") {
-
-      # The maximized log-likelihoods are the same between survreg and flexsurvreg, 
-      # however the parameterisation is different: the first coefficient (Intercept) 
-      # reported by survreg is log(mu), and survreg's "scale" is
-      # dweibull's (thus flexsurvreg)'s 1 / shape
-      res <- -fit[[model]]$res[covariate,] / (1 / fit[[model]]$res["shape",])
-
-      sprintf("%0.2f (%0.2f, %0.2f)",
-              exp(res[["est"]]),
-              exp(res[["U95%"]]),
-              exp(res[["L95%"]]))
-
-    } else if (model == "splines") {
-
-      c("--")
-
-    } else {
-
-      res <- fit[[model]]$res[covariate,]
-
-      sprintf("%0.2f (%0.2f, %0.2f)",
-              exp(res[["est"]]),
-              exp(res[["L95%"]]),
-              exp(res[["U95%"]]))
-    }
-  }
-}
-
-casebase_HR_CI <- lapply(list("exponential","gompertz","weibull","splines"), 
-                         exp_est_ci, 
-                         fit = fits,
-                         covariate = "ScrArmScreening group", 
-                         type = "casebase")
-
-survreg_HR_CI <- lapply(list("exponential","gompertz","weibull","splines"), 
-                         exp_est_ci, 
-                         fit = aft_fits,
-                         covariate = "ScrArmScreening group", 
-                         type = "survreg")
-
-ptable <- cbind(list("Exponential","Gompertz","Weibull","Splines"),
-            do.call(rbind, casebase_HR_CI), do.call(rbind, survreg_HR_CI))
-```
-
-In Table \ref{tab:print-erspc-estimates}, we present a side-by-side comparison of the hazard ratios and confidence intervals estimated from \code{fitSmoothHazard} and the corresponding parametric model using \code{survival::survreg}, as well as the Cox model estimate.
-The hazard ratio estimates and confidence intervals are similar across all four models.
-This reinforces the idea that, under proportional hazards, we do not need to model the full hazard to obtain reliable estimates of the hazard ratio.
-Nevertheless, Figure \ref{fig:erspc-cox-cif} shows that different parametric models can still give rise to qualitatively different estimates for the CIF.
-
-```{r print-erspc-estimates, eval = eval_cs1}
-kable(ptable, "latex", booktabs = TRUE, align = c("l","c","c"),
-      caption = c("Comparison of estimated hazard ratios and 95\\% confidence intervals for ERSPC data."),
-      col.names = c("Model","casebase::fitSmoothHazard","survival::survreg")) %>%
-  kable_styling() %>%
-  column_spec(1, bold = TRUE) %>%
-  collapse_rows(columns = 1, latex_hline = "major", valign = "middle") %>%
-  footnote(general = sprintf("Cox model estimate: HR (95%% CI) = %0.2f (%0.2f, %0.2f)", CI_cox[1], CI_cox[2], CI_cox[3]),
-           general_title = "") 
+lines(risk[,1], risk[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risk[,1], risk[,3], col = myCols[2], lty = 2, lwd = 2)
 ```

### 2021-09-23 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
remove exp gomp weib models

@@
   \citep{hastie1987generalized}.
-\item
-  \code{gbm}: This option provides support for \emph{Gradient Boosted
-  Trees} via the \CRANpkg{gbm} package. This feature is still
-  experimental.
 \end{itemize}
@@
 For our first case study, we return to the ERSPC study and investigate
-the differences in risk between the control and screening arms. We fit
-four models that differ in which functional form of time is used: 1)
-excluded from the linear predictor (Eq. \eqref{eq:exp}); 2) linear
-function (Eq. \eqref{eq:gomp}); 3) log function (Eq.
-\eqref{eq:weibull}); and 4) a smooth function using cubic B-splines. The
-models are fit using \code{fitSmoothHazard} with the familiar formula
-interface:
+the differences in risk between the control and screening arms. We fit a
+flexible hazard by using a smooth function of time modeled with cubic
+B-splines. The model is fit using \code{fitSmoothHazard} with the
+familiar formula interface:
 
@@
 \begin{Sinput}
-fmla <- list(exponential = formula(DeadOfPrCa ~ ScrArm),
-             gompertz = formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
-             weibull = formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
-             splines = formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
-
-fits <- lapply(fmla, function(form) {
-  fitSmoothHazard(form, data = ERSPC, ratio = 100)
-})
+library(splines)
+fit <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
+                       data = ERSPC, ratio = 100)
 \end{Sinput}
@@
 The output object from \code{fitSmoothHazard} inherits from the
-\code{singleEventCB} and \code{glm} classes. For this reason, we
-leveraged the \code{summary} method for \code{glm} objects to output a
+\code{singleEventCB} and \code{glm} classes. For this reason, we can
+leverage the \code{summary} method for \code{glm} objects to output a
 familiar summary of the results:
@@
 \begin{Sinput}
-summary(fits[["splines"]]) 
+summary(fit) 
 \end{Sinput}
@@
 #> Call:
-#> fitSmoothHazard(formula = form, data = ERSPC, ratio = 100)
+#> fitSmoothHazard(formula = DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
+#>     data = ERSPC, ratio = 100)
 #> 
@@
 #>    Min      1Q  Median      3Q     Max  
-#> -0.409  -0.172  -0.135  -0.084   3.808  
+#> -0.403  -0.172  -0.135  -0.085   3.801  
 #> 
@@
 #>                       Estimate Std. Error z value Pr(>|z|)    
-#> (Intercept)           -10.2905     0.3185  -32.31  < 2e-16 ***
-#> bs(Follow.Up.Time)1     4.3587     0.8008    5.44  5.2e-08 ***
-#> bs(Follow.Up.Time)2     2.0890     0.4721    4.42  9.7e-06 ***
-#> bs(Follow.Up.Time)3     4.6266     0.6910    6.70  2.1e-11 ***
-#> ScrArmScreening group  -0.2354     0.0886   -2.66   0.0079 ** 
+#> (Intercept)           -10.2533     0.3168  -32.36  < 2e-16 ***
+#> bs(Follow.Up.Time)1     4.2544     0.7976    5.33  9.6e-08 ***
+#> bs(Follow.Up.Time)2     2.1371     0.4720    4.53  5.9e-06 ***
+#> bs(Follow.Up.Time)3     4.5339     0.6889    6.58  4.7e-11 ***
+#> ScrArmScreening group  -0.2481     0.0886   -2.80   0.0051 ** 
 #> ---
@@
 #>     Null deviance: 6059.0  on 54539  degrees of freedom
-#> Residual deviance: 5785.3  on 54535  degrees of freedom
-#> AIC: 5795
+#> Residual deviance: 5787.6  on 54535  degrees of freedom
+#> AIC: 5798
 #> 
-#> Number of Fisher Scoring iterations: 9
+#> Number of Fisher Scoring iterations: 8
 \end{Soutput}
@@
 new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
-new_time <- seq(0, 14, by = 0.1)
-
-risks <- lapply(fits, function(fit) {
+new_time <- c(0, 1)
+
 absoluteRisk(fit, time = new_time, newdata = new_data)
-})
 \end{Sinput}
-\end{Schunk}
-
-In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from
-\pkg{casebase} on the Cox model CIF. The CIF estimates for the
-exponential model in panel (1) overestimate the cumulative incidence
-earlier on in time, and underestimate it later on. Based on our earlier
-discussion of the population-time plot, this poor fit for the
-exponential hazard was expected. We notice a better fit with increasing
-complexity of our model for time in Figure \ref{fig:erspc-cox-cif}
-(panels 2--4). As noted in Theoretical Details section, the usual
-asymptotic results hold for likelihood ratio tests built using case-base
-sampling models. Therefore, we can easily test the null hypothesis that
-the exponential model is just as good as the larger (in terms of number
-of parameters) spline model:
+\begin{Soutput}
+#>  time                
+#>     0 0.0e+00 0.0e+00
+#>     1 5.4e-05 4.2e-05
+\end{Soutput}
+\end{Schunk}
+
+In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIF from
+\pkg{casebase} on the Cox model CIF. As noted in the Theoretical Details
+section, the usual asymptotic results hold for likelihood ratio tests
+built using case-base sampling models. Therefore, we can easily test the
+significance of the spline model:
 
@@
 #> 
-#> Model 1: DeadOfPrCa ~ ScrArm + offset(offset)
-#> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
-#>   Resid. Df Resid. Dev Df Deviance Pr(>Chi)    
-#> 1     54538       6052                         
-#> 2     54535       5785  3      267   <2e-16 ***
+#> Model: binomial, link: logit
+#> 
+#> Response: DeadOfPrCa
+#> 
+#> Terms added sequentially (first to last)
+#> 
+#> 
+#>                    Df Deviance Resid. Df Resid. Dev Pr(>Chi)    
+#> NULL                               54539       6059             
+#> bs(Follow.Up.Time)  3      263     54536       5796   <2e-16 ***
+#> ScrArm              1        8     54535       5788   0.0048 ** 
 #> ---
@@
 The null hypothesis is rejected in favor of the spline model. Similarly,
-the AIC provides further evidence that the flexible function of time
-provides the best fit:
-
-\begin{Schunk}
-\begin{Soutput}
-#>     Exp. Gompertz  Weibull  Splines 
-#>     6056     5821     5807     5795
-\end{Soutput}
-\end{Schunk}
+to compare different models (e.g.~time modeled linearly), we could
+compute Akaike's Information Criterion (AIC) for each model and compare
+them.
 
@@
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cox-cif-1} \caption{CIFs for control and screening groups in the ERSPC data.
-In each of the panels, we plot the CIF from the Cox model using \code{survival::survfit} (solid line) and the CIF from the case-base sampling scheme (dashed line) with different functional forms of time.
-(1) The time variable is excluded (exponential).
-(2) Linear function of time (Gompertz).
-(3) The natural logarithm (Weibull).
-(4) Cubic B-spline expansion of time.}\label{fig:erspc-cox-cif}
+\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cox-cif-1} \caption{CIF for control and screening groups in the ERSPC data.
+We plot the CIF from the Cox model using \code{survival::survfit} (solid line) and the CIF from case-base sampling (dashed line).
+We used a cubic B-spline expansion of time.}\label{fig:erspc-cox-cif}
 \end{figure}
-\end{Schunk}
-
-In Table \ref{tab:print-erspc-estimates}, we present a side-by-side
-comparison of the hazard ratios and confidence intervals estimated from
-\code{fitSmoothHazard} and the corresponding parametric model using
-\code{survival::survreg}, as well as the Cox model estimate. The hazard
-ratio estimates and confidence intervals are similar across all four
-models. This reinforces the idea that, under proportional hazards, we do
-not need to model the full hazard to obtain reliable estimates of the
-hazard ratio. Nevertheless, Figure \ref{fig:erspc-cox-cif} shows that
-different parametric models can still give rise to qualitatively
-different estimates for the CIF.
-
-\begin{Schunk}
-\begin{table}
-
-\caption{\label{tab:print-erspc-estimates}Comparison of estimated hazard ratios and 95\% confidence intervals for ERSPC data.}
-\centering
-\begin{tabular}[t]{>{}lcc}
-\toprule
-Model & casebase::fitSmoothHazard & survival::survreg\\
-\midrule
-\textbf{Exponential} & 0.79 (0.67, 0.94) & 0.81 (0.68, 0.96)\\
-\cmidrule{1-3}
-\textbf{Gompertz} & 0.80 (0.67, 0.95) & 0.80 (0.67, 0.95)\\
-\cmidrule{1-3}
-\textbf{Weibull} & 0.81 (0.68, 0.96) & 0.80 (0.65, 0.96)\\
-\cmidrule{1-3}
-\textbf{Splines} & 0.79 (0.66, 0.94) & --\\
-\bottomrule
-\multicolumn{3}{l}{\rule{0pt}{1em}Cox model estimate: HR (95\% CI) = 0.80 (0.67, 0.95)}\\
-\end{tabular}
-\end{table}
-
 \end{Schunk}
@@
 %
-\url{http://sahirbhatnagar.com/}%
+\url{http://sahirbhatnagar.com/}\\%
 %
-\\\email{sahir.bhatnagar@mcgill.ca}
+\email{sahir.bhatnagar@mcgill.ca}%
 }
@@
 %
-\url{https://maxturgeon.ca/}%
+\url{https://maxturgeon.ca/}\\%
 %
-\\\email{max.turgeon@umanitoba.ca}
+\email{max.turgeon@umanitoba.ca}%
 }
@@
 %
-\\\email{jesse.islam@mail.mcgill.ca}
+\email{jesse.islam@mail.mcgill.ca}%
 }
@@
 %
-\url{http://www.medicine.mcgill.ca/epidemiology/hanley/}%
+\url{http://www.medicine.mcgill.ca/epidemiology/hanley/}\\%
 %
-\\\email{james.hanley@mcgill.ca}
+\email{james.hanley@mcgill.ca}%
 }
@@
 %
-\url{http://individual.utoronto.ca/osaarela/}%
+\url{http://individual.utoronto.ca/osaarela/}\\%
 %
-\\\email{olli.saarela@utoronto.ca}
+\email{olli.saarela@utoronto.ca}%
 }

### 2021-09-23 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.R

@@ -7,7 +7,7 @@ knitr::opts_chunk$set(
   cache = TRUE, 
   fig.pos = 'ht',
   comment = "#>",
-  fig.path = "../figures/",
+  fig.path = "./",
   out.width = "\\textwidth",
    out.extra = "keepaspectratio=true"
 )
@@ -100,38 +100,24 @@ plot(pt_object,
 
 ## ---- eval = eval_cs1, echo = -1----------------------------------------------
 set.seed(12)
-fmla <- list(exponential = formula(DeadOfPrCa ~ ScrArm),
-             gompertz = formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
-             weibull = formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
-             splines = formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
-
-fits <- lapply(fmla, function(form) {
-  fitSmoothHazard(form, data = ERSPC, ratio = 100)
-})
+library(splines)
+fit <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
+                       data = ERSPC, ratio = 100)
 
 
 ## ---- eval = TRUE, echo=eval_cs1----------------------------------------------
-summary(fits[["splines"]]) 
+summary(fit) 
 
 
-## ---- eval=FALSE, echo=echo_plot_code-----------------------------------------
-#> new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
-#> new_time <- seq(0, 14, by = 0.1)
-#> 
-#> risks <- lapply(fits, function(fit) {
-#>   absoluteRisk(fit, time = new_time, newdata = new_data)
-#> })
+## ---- eval=eval_cs1, echo=echo_plot_code--------------------------------------
+new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
+new_time <- c(0, 1)
+
+absoluteRisk(fit, time = new_time, newdata = new_data)
 
 
 ## ----erscp-compare, eval = eval_cs1, echo = FALSE-----------------------------
-anova(fits$exponential, fits$splines, test = "LRT")
-
-
-## ----erscp-aic, eval = eval_cs1, echo=FALSE-----------------------------------
-c("Exp." = AIC(fits$exponential),
-  "Gompertz" = AIC(fits$gompertz),
-  "Weibull" = AIC(fits$weibull),
[…]

### 2021-10-02 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
add conf bands

@@
   \usepackage{tabu}
-output: 
-  rticles::rjournal_article:
-    number_sections: true
+output: rticles::rjournal_article
 bibliography: bhatnagar-turgeon-islam-hanley-saarela.bib 
@@
 eval_cs1 <- TRUE
-eval_cs2 <- TRUE
-eval_cs3 <- TRUE
+eval_cs2 <- FALSE
+eval_cs3 <- FALSE
 eval_cs4 <- FALSE
@@
 Population-time plots are a descriptive visualization of incidence density, where aggregate person-time is represented by area and events as points within the area.
-The case-base sampling approach described in Section \ref{theory} can be visualized in the form of a population time plot.
+The case-base sampling approach described above can be visualized in the form of a population time plot.
 These plots are extremely informative graphical displays of survival data and should be one of the first steps in an exploratory data analysis.
@@
 
-```{r, eval = TRUE, echo=eval_cs1}
+```{r, eval = eval_cs1, echo = TRUE}
 summary(fit) 
@@
 # Recompute with more time points
-risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
+risk_mp <- absoluteRisk(fit, time = new_time, newdata = new_data)
 
@@
        bg = "gray90")
-lines(risk[,1], risk[,2], col = myCols[1], lty = 2, lwd = 2)
-lines(risk[,1], risk[,3], col = myCols[2], lty = 2, lwd = 2)
+lines(risk_mp[,1], risk_mp[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risk_mp[,1], risk_mp[,3], col = myCols[2], lty = 2, lwd = 2)
+```
+
+Moreover, we can also compute confidence intervals for the CIF using the method \code{confint.absRiskCB}.
+
+```{r eval = eval_cs1, echo = -1}
+new_time <- c(0, 1)
+risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
+conf_ints <- confint(risk, fit)
+conf_ints
+```
+
+These point-wise confidence intervals are constructed using parametric bootstrap; the validity of this approach relies on the asymptotic normality of the MLE.
+In Figure \ref{fig:erspc-cif-conf}, we see the 95% confidence bands around the CIF estimates.
+
+```{r erspc-cif-conf, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="CIF estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\\% confidence bands."}
+# confidence interval
+tt <- confint(risk_mp, fit)
+
+# plot absrisk and 95% CI
+new_time <- seq(0,14,0.1)
+i.backw <- order(new_time, decreasing = TRUE)
+i.forw <- order(new_time)
+
+x.poly <- c(new_time[i.forw], new_time[i.backw])
+range_y <- 100 * range(c(tt$conf.high, tt$conf.low)) * c(0.99, 1.01)
+
+par(mfrow = c(1, 2))
+# 1. Control
+tt_sub <- subset(tt, cov_prof == "Control group")
+y.poly <- 100 * c(tt_sub$conf.low[i.forw], 
+                  tt_sub$conf.high[i.backw])
+do.call("plot", 
+    list(
+        x = range(x.poly),
+        y = range_y,
+        type = "n",
+        ylab = "Cumulative incidence (%)",
+        xlab = "Years since randomization",
+        main = "Control group"
+    )
+)
+
+graphics::polygon(x.poly, y.poly, col = "lightgray", border = NA)
+
+# Add KM
+km_fit <- survfit(Surv(Follow.Up.Time, DeadOfPrCa) ~ 1,
+                  data = ERSPC,
+                  subset = ScrArm == "Control group",
+                  conf.type = "log-log")
+
+lines(x = km_fit$time, y = 100 * (1 - km_fit$surv),
+      lwd = 1, lty = 1, col = myCols[2])
+lines(x = km_fit$time, y = 100 * (1 - km_fit$lower),
+      lwd = 2, lty = 3, col = myCols[2])
+lines(x = km_fit$time, y = 100 * (1 - km_fit$upper),
+      lwd = 2, lty = 3, col = myCols[2])
+
+do.call("lines", 
+        list(
+            x = new_time[i.forw],
+            y = 100 * tt_sub$estimate[i.forw],
+            lwd = 2,
+            lty = 1,
+            col = myCols[1]
+        )
+)
+
+# Add legend
+legend("topleft", 
+       legend = c("Case-base", "Kaplan-Meier"), 
+       col = myCols,
+       lty = c(1, 1), 
+       lwd = c(2, 1),
+       cex = 1.1,
+       bg = "gray90")
+
+# 2. Screening
+tt_sub <- subset(tt, cov_prof == "Screening group")
+y.poly <- 100 * c(tt_sub$conf.low[i.forw], 
+                  tt_sub$conf.high[i.backw])
+do.call("plot", 
+    list(
+        x = range(x.poly),
+        y = range_y,
+        type = "n",
+        # ylab = "Cumulative incidence (%)",
+        ylab = "",
+        xlab = "Years since randomization",
+        main = "Screening group"
+    )
+)
+
+graphics::polygon(x.poly, y.poly, col = "lightgray", border = NA)
+
+# Add KM
+km_fit <- survfit(Surv(Follow.Up.Time, DeadOfPrCa) ~ 1,
+                  data = ERSPC,
+                  subset = ScrArm == "Screening group",
+                  conf.type = "log-log")
+
+lines(x = km_fit$time, y = 100 * (1 - km_fit$surv),
+      lwd = 1, lty = 1, col = myCols[2])
+lines(x = km_fit$time, y = 100 * (1 - km_fit$lower),
+      lwd = 2, lty = 3, col = myCols[2])
+lines(x = km_fit$time, y = 100 * (1 - km_fit$upper),
+      lwd = 2, lty = 3, col = myCols[2])
+
+do.call("lines", 
+        list(
+            x = new_time[i.forw],
+            y = 100 * tt_sub$estimate[i.forw],
+            lwd = 2,
+            lty = 1,
+            col = myCols[1]
+        )
+)
 ```
@@
 In future versions of the package, we also want to increase the complement of diagnostic and inferential tools that are currently available.
-For example, we would like to include the ability to compute confidence intervals for the cumulative incidence curve.
-The delta method or parametric bootstrap are two different strategies we can use to construct approximate confidence intervals.
-Furthermore, we would like to include more functions to compute calibration and discrimination statistics (e.g.
+For example, we would like to include more functions to compute calibration and discrimination statistics (e.g.
 AUC) for our models.
@@
 
-We would like to thank Yi Yang for helpful discussions on penalized regression models and gradient boosting.
+We would like to thank the anonymous reviewers for their insightful comments and criticisms.
+We would also like to thank Yi Yang for helpful discussions on penalized regression models.
 \mbox{Bhatnagar} gratefully acknowledges funding via a Discovery Grant from the Natural Sciences and Engineering Research Council of Canada (NSERC).

### 2021-10-02 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
add conf bands

@@
 density, where aggregate person-time is represented by area and events
-as points within the area. The case-base sampling approach described in
-Section \ref{theory} can be visualized in the form of a population time
-plot. These plots are extremely informative graphical displays of
-survival data and should be one of the first steps in an exploratory
-data analysis. The \code{popTime} function and \code{plot} method
-facilitate this task:
+as points within the area. The case-base sampling approach described
+above can be visualized in the form of a population time plot. These
+plots are extremely informative graphical displays of survival data and
+should be one of the first steps in an exploratory data analysis. The
+\code{popTime} function and \code{plot} method facilitate this task:
 
@@
 
+Moreover, we can also compute confidence intervals for the CIF using the
+method \code{confint.absRiskCB}.
+
+\begin{Schunk}
+\begin{Sinput}
+risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
+conf_ints <- confint(risk, fit)
+conf_ints
+\end{Sinput}
+\begin{Soutput}
+#>   time estimate conf.low conf.high        cov_prof
+#> 1    0  0.0e+00  0.0e+00   0.0e+00   Control group
+#> 2    1  5.4e-05  3.5e-05   9.3e-05   Control group
+#> 3    0  0.0e+00  0.0e+00   0.0e+00 Screening group
+#> 4    1  4.2e-05  2.7e-05   7.3e-05 Screening group
+\end{Soutput}
+\end{Schunk}
+
+These point-wise confidence intervals are constructed using parametric
+bootstrap; the validity of this approach relies on the asymptotic
+normality of the MLE. In Figure \ref{fig:erspc-cif-conf}, we see the
+95\% confidence bands around the CIF estimates.
+
+\begin{Schunk}
+\begin{figure}[ht]
+\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cif-conf-1} \caption[CIF estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands]{CIF estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands.}\label{fig:erspc-cif-conf}
+\end{figure}
+\end{Schunk}
+
 Previous re-analyses of these data suggested that the overall screening
@@
 
-\begin{Schunk}
-\begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./compPop-1} \caption[Population-time plot for the stem-cell transplant study with both relapse and competing events]{Population-time plot for the stem-cell transplant study with both relapse and competing events.}\label{fig:compPop}
-\end{figure}
-\end{Schunk}
-
 Our main objective is to compute the cumulative incidence of relapse for
@@
 \end{Sinput}
-\end{Schunk}
-
-\begin{Schunk}
-\begin{table}
-
-\caption{\label{tab:bmtcrr-cis}Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.}
-\centering
-\begin{tabular}[t]{lrlrl}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
-Covariates & HR & 95\% CI & HR & 95\% CI\\
-\midrule
-Sex & 0.74 & (0.43, 1.3) & 0.68 & (0.39, 1.19)\\
-Disease & 0.52 & (0.29, 0.94) & 0.52 & (0.29, 0.93)\\
-Phase (CR2 vs. CR1) & 1.20 & (0.48, 3) & 1.21 & (0.48, 3.02)\\
-Phase (CR3 vs. CR1) & 1.59 & (0.41, 6.17) & 1.67 & (0.43, 6.5)\\
-Phase (Relapse vs. CR1) & 4.09 & (1.9, 8.81) & 4.55 & (2.09, 9.9)\\
-\addlinespace
-Source & 1.70 & (0.55, 5.23) & 1.46 & (0.47, 4.54)\\
-Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
-\bottomrule
-\end{tabular}
-\end{table}
-
 \end{Schunk}
@@
 \end{Sinput}
-\end{Schunk}
-
-\begin{Schunk}
-\begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./bmtcrr-risk-1} \caption{\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups.
-The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:bmtcrr-risk}
-\end{figure}
 \end{Schunk}
@@
 
-\begin{Schunk}
-\begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./coefplots-1} \caption{\label{fig:cs3lolliPlot} Coefficient estimates from the Cox model (Cox), penalized Cox model using the \pkg{glmnet} package (Pen.
-Cox), and our approach using penalized case-base sampling (Pen.
-CB).
-Only the covariates that were selected by both penalized approaches are shown.
-The shrinkage of the coefficient estimates for Pen.
-Cox and Pen.
-CB occurs due to the $\ell_1$ penalty.}\label{fig:coefplots}
-\end{figure}
-\end{Schunk}
-
 We then compare the cumulative incidence estimation over the test set.
@@
 these three models are minimal.
-
-\begin{Schunk}
-\begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./riskregressionBrier-1} \caption{\label{fig:cs3FinalBrier} Comparison of Cox regression (Cox), penalized Cox regression (Pen.
-Cox), penalized case-base sampling estimation (Pen.
-CB), and Kaplan-Meier (K-M).
-(A)  Probability of death as a function of follow-up time.
-(B) Brier score as a function of follow-up time, where a lower score corresponds to better performace.}\label{fig:riskregressionBrier}
-\end{figure}
-\end{Schunk}
 
@@
 complement of diagnostic and inferential tools that are currently
-available. For example, we would like to include the ability to compute
-confidence intervals for the cumulative incidence curve. The delta
-method or parametric bootstrap are two different strategies we can use
-to construct approximate confidence intervals. Furthermore, we would
-like to include more functions to compute calibration and discrimination
-statistics (e.g.~AUC) for our models. Saarela and Arjas
-\citeyearpar{saarela2015non} also describe how to obtain a posterior
-distribution for the AUC from their model. Their approach could
-potentially be included in \pkg{casebase}. Finally, we want to provide
-more flexibility in how the case-base sampling is performed. This could
-be achieved by adding a \code{hazard} argument to the function
-\code{sampleCaseBase}. In this way, users could specify their own
-sampling mechanism. For example, they could provide a hazard that gives
-sampling probabilities that are proportional to the cardiovascular
-disease event rate given by the Framingham score \citep{saarela2015non}.
+available. For example, we would like to include more functions to
+compute calibration and discrimination statistics (e.g.~AUC) for our
+models. Saarela and Arjas \citeyearpar{saarela2015non} also describe how
+to obtain a posterior distribution for the AUC from their model. Their
+approach could potentially be included in \pkg{casebase}. Finally, we
+want to provide more flexibility in how the case-base sampling is
+performed. This could be achieved by adding a \code{hazard} argument to
+the function \code{sampleCaseBase}. In this way, users could specify
+their own sampling mechanism. For example, they could provide a hazard
+that gives sampling probabilities that are proportional to the
+cardiovascular disease event rate given by the Framingham score
+\citep{saarela2015non}.
 
@@
 
-We would like to thank Yi Yang for helpful discussions on penalized
-regression models and gradient boosting. \mbox{Bhatnagar} gratefully
+We would like to thank the anonymous reviewers for their insightful
+comments and criticisms. We would also like to thank Yi Yang for helpful
+discussions on penalized regression models. \mbox{Bhatnagar} gratefully
 acknowledges funding via a Discovery Grant from the Natural Sciences and

### 2021-10-02 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.R

@@ -40,8 +40,8 @@ eval_introduction <- TRUE
 eval_theory <- TRUE
 eval_implementation <- TRUE
 eval_cs1 <- TRUE
-eval_cs2 <- TRUE
-eval_cs3 <- TRUE
+eval_cs2 <- FALSE
+eval_cs3 <- FALSE
 eval_cs4 <- FALSE
 eval_colophon <- FALSE # Only for draft
 echo_plot_code <- TRUE
@@ -105,7 +105,7 @@ fit <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm,
                        data = ERSPC, ratio = 100)
 
 
-## ---- eval = TRUE, echo=eval_cs1----------------------------------------------
+## ---- eval = eval_cs1, echo = TRUE--------------------------------------------
 summary(fit) 
 
 
@@ -132,7 +132,7 @@ myCols <- cbPalette[c(4,7)]
 new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
 new_time <- seq(0,14,0.1)
 # Recompute with more time points
-risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
+risk_mp <- absoluteRisk(fit, time = new_time, newdata = new_data)
 
 plot(survfit(cox_model, newdata = new_data),
      ylab = "Cumulative incidence (%)",
@@ -151,8 +151,119 @@ legend("topleft",
        lwd = 2,
        cex = 1.1,
        bg = "gray90")
-lines(risk[,1], risk[,2], col = myCols[1], lty = 2, lwd = 2)
-lines(risk[,1], risk[,3], col = myCols[2], lty = 2, lwd = 2)
+lines(risk_mp[,1], risk_mp[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risk_mp[,1], risk_mp[,3], col = myCols[2], lty = 2, lwd = 2)
+
+
+## ----eval = eval_cs1, echo = -1-----------------------------------------------
+new_time <- c(0, 1)
+risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
+conf_ints <- confint(risk, fit)
+conf_ints
+
+
[…]

### 2021-10-02 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
adjust cif language

@@
 The baseline hazard then needs to be estimated separately using a non-parametric approach.
-This leads to step-wise estimates of the cumulative incidence that are difficult to interpret.
+This leads to step-wise estimates of the survival function that are difficult to interpret.
 Using case-base sampling, Hanley & Miettinen (2009) explained how parametric hazard functions can be estimated using logistic regression.
-Their approach naturally leads to estimates of the cumulative incidence that are smooth-in-time.
+Their approach naturally leads to estimates of the survival function that are smooth-in-time.
   | In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis.
@@
 eval_cs1 <- TRUE
-eval_cs2 <- FALSE
-eval_cs3 <- FALSE
+eval_cs2 <- TRUE
+eval_cs3 <- TRUE
 eval_cs4 <- FALSE
@@
 The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled.
-Using this estimate, the survival function and the CIF can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below).
+Using this estimate, the survival function and the CIF can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CDF} below).
 Packages providing this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and \pkg{survival}.
@@
 Packages that implement this approach include \pkg{riskRegression}, \pkg{rstpm2}, and \pkg{survival}.
-As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category often produces estimates in the form of stepwise functions.
-Providing smooth estimates of the CIF was one of the main motivations for introducing case-base sampling in survival analysis.
+As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the survival function, whereas the second category often produces estimates in the form of stepwise functions.
 
@@
 \toprule
-& \textbf{Competing} & \textbf{Allows} &  \textbf{Penalized}   &    &  & \textbf{Semi} & \textbf{Interval/Left} & \textbf{Absolute}  \\
-\textbf{Package}        & \textbf{Risks}  & \textbf{Non PH} & \textbf{Regression} & \textbf{Splines} & \textbf{Parametric} & \textbf{Parametric} & \textbf{Censoring} & \textbf{Risk}           \\
+& \textbf{Competing} & \textbf{Allows} &  \textbf{Penalized}   &    &  & \textbf{Semi} & \textbf{Interval/Left} & \textbf{Risk}  \\
+\textbf{Package}        & \textbf{Risks}  & \textbf{Non PH} & \textbf{Regression} & \textbf{Splines} & \textbf{Parametric} & \textbf{Parametric} & \textbf{Censoring} & \textbf{Estimates}           \\
 \bottomrule
@@
 If this is not selected, the package only handles right-censoring.
-\textbf{Absolute Risk}: computation for survival curves, cumulative incidence or cumulative hazard is readily available.}
+\textbf{Risk estimates}: computation for survival curves, cumulative incidence or cumulative hazard is readily available.}
 \label{tab:surv-pkgs}
@@
 In Figure \ref{fig:plot-stratified-erspc-data}, we see that more events are observed at later follow-up times.
-Therefore, a constant hazard model would not be appropriate in this instance as it would overestimate the cumulative incidence earlier on in time, and underestimate it later on.
+Therefore, a constant hazard model would not be appropriate in this instance as it would overestimate the incidence earlier on in time, and underestimate it later on.
 Finally, the base series is sampled horizontally with sampling weight proportional to their follow-up time.
@@
   2. estimation of the smooth hazard function; 
-  3. estimation of the CIF.
+  3. estimation of the survival function.
 
@@
 \end{equation}
-The integral is computed using either the numerical or Monte-Carlo integration. The risk function (or cumulative incidence function) is then defined as
-\begin{equation}\label{eqn:CI}
-CI(t) = 1 - S(t).
+The integral is computed using either the numerical or Monte-Carlo integration. The risk function (or cumulative distribution function) is then defined as
+\begin{equation}\label{eqn:CDF}
+F(t) = 1 - S(t).
 \end{equation}
@@
 
-In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIF from \pkg{casebase} on the Cox model CIF.
+In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated risk function from \pkg{casebase} on the Cox model estimate.
 As noted in the Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models.
@@
 
-```{r erspc-cox-cif, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="CIF for control and screening groups in the ERSPC data.
-We plot the CIF from the Cox model using \\code{survival::survfit} (solid line) and the CIF from case-base sampling (dashed line).
+```{r erspc-cox-cif, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk functions for control and screening groups in the ERSPC data.
+We plot the estimate from the Cox model using \\code{survival::survfit} (solid line) and the estimate from case-base sampling (dashed line).
 We used a cubic B-spline expansion of time."}
@@
 plot(survfit(cox_model, newdata = new_data),
-     ylab = "Cumulative incidence (%)",
+     ylab = "Risk probability (%)",
      xlab = "Years since randomization",
@@
 
-Moreover, we can also compute confidence intervals for the CIF using the method \code{confint.absRiskCB}.
+Moreover, we can also compute confidence intervals for the risk function using the method \code{confint.absRiskCB}.
 
@@
 These point-wise confidence intervals are constructed using parametric bootstrap; the validity of this approach relies on the asymptotic normality of the MLE.
-In Figure \ref{fig:erspc-cif-conf}, we see the 95% confidence bands around the CIF estimates.
-
-```{r erspc-cif-conf, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="CIF estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\\% confidence bands."}
+In Figure \ref{fig:erspc-cif-conf}, we see the 95% confidence bands around the estimates.
+
+```{r erspc-cif-conf, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\\% confidence bands."}
 # confidence interval
@@
         type = "n",
-        ylab = "Cumulative incidence (%)",
+        ylab = "Risk probability (%)",
         xlab = "Years since randomization",
@@
         type = "n",
-        # ylab = "Cumulative incidence (%)",
+        # ylab = "Risk probability (%)",
         ylab = "",
@@
 
-We then compare the cumulative incidence estimation over the test set.
+We then compare the risk estimation over the test set.
 The probabilities over time for each observation are averaged, resulting in the absolute risk curves shown in Figure \ref{fig:cs3FinalBrier} (A).
@@
 
-In this article, we presented the R package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating CIFs using case-base sampling.
+In this article, we presented the R package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating risk functions using case-base sampling.
 Our package also provides several functions to produce graphical summaries of the data and the results.
@@
 
-As presented in Hanley & Miettinen [-@hanley2009fitting], case-base sampling is comprised of three steps: 1) sampling a case series and a base series from the study; 2) fit the log-hazard as a linear function of predictors (including time); and 3) use the fitted hazard to estimate the CIF.
+As presented in Hanley & Miettinen [-@hanley2009fitting], case-base sampling is comprised of three steps: 1) sampling a case series and a base series from the study; 2) fit the log-hazard as a linear function of predictors (including time); and 3) use the fitted hazard to estimate the risk function.
 Accordingly, our package provides functions for each step.
@@
 However, in terms of output, both approaches differ significantly.
-Case-base sampling produces smooth hazards and smooth cumulative incidence curves, whereas Cox regression and Fine-Gray models produce stepwise CIFs and never explicitly model the hazard function.
+Case-base sampling produces smooth hazards and smooth survival curves, whereas Cox regression and Fine-Gray models produce stepwise CIFs and never explicitly model the hazard function.
 Qualitatively, we showed that by using splines in the linear predictor, all three models yielded similar curves.
@@
 Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\\midrule
-Cumulative incidence & Step function & Smooth-in-time curve\\\midrule
+Survival curve & Step function & Smooth-in-time curve\\\midrule
 Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\\midrule
@@
 
-In conclusion, we presented the R package \pkg{casebase} which implements case-base sampling for fitting parametric survival models and for estimating smooth cumulative incidence functions using the framework of generalized linear models.
+In conclusion, we presented the R package \pkg{casebase} which implements case-base sampling for fitting parametric survival models and for estimating smooth survival functions using the framework of generalized linear models.
 We strongly believe that its flexibility and its foundation on the familiar logistic regression model will make it appealing to new and established practitioners.

### 2021-10-02 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
adjust cif language

--- 2021-09-23 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
@@
 hazard then needs to be estimated separately using a non-parametric
-approach. This leads to step-wise estimates of the cumulative incidence
+approach. This leads to step-wise estimates of the survival function
 that are difficult to interpret. Using case-base sampling, Hanley \&
@@
 estimated using logistic regression. Their approach naturally leads to
-estimates of the cumulative incidence that are smooth-in-time.\\
+estimates of the survival function that are smooth-in-time.\\
 In this paper, we present the \CRANpkg{casebase} R package, a
@@
 estimate, the survival function and the CIF can be obtained using their
-functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI}
+functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CDF}
 below). Packages providing this functionality include \pkg{CFC},
@@
 factor between these two approaches is that the first category leads to
-smooth estimates of the cumulative incidence function, whereas the
-second category often produces estimates in the form of stepwise
-functions. Providing smooth estimates of the CIF was one of the main
-motivations for introducing case-base sampling in survival analysis.
+smooth estimates of the survival function, whereas the second category
+often produces estimates in the form of stepwise functions.
 
@@
 hazard model would not be appropriate in this instance as it would
-overestimate the cumulative incidence earlier on in time, and
-underestimate it later on. Finally, the base series is sampled
-horizontally with sampling weight proportional to their follow-up time.
-The reader should refer to the package vignettes for more examples and a
-detailed description of how to modify the aesthetics of a
-population-time plot.
+overestimate the incidence earlier on in time, and underestimate it
+later on. Finally, the base series is sampled horizontally with sampling
+weight proportional to their follow-up time. The reader should refer to
+the package vignettes for more examples and a detailed description of
+how to modify the aesthetics of a population-time plot.
 
@@
 \item
-  estimation of the CIF.
+  estimation of the survival function.
 \end{enumerate}
@@
 \end{equation} The integral is computed using either the numerical or
-Monte-Carlo integration. The risk function (or cumulative incidence
-function) is then defined as \begin{equation}\label{eqn:CI}
-CI(t) = 1 - S(t).
+Monte-Carlo integration. The risk function (or cumulative distribution
+function) is then defined as \begin{equation}\label{eqn:CDF}
+F(t) = 1 - S(t).
 \end{equation}
@@
 
-In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIF from
-\pkg{casebase} on the Cox model CIF. As noted in the Theoretical Details
-section, the usual asymptotic results hold for likelihood ratio tests
-built using case-base sampling models. Therefore, we can easily test the
-significance of the spline model:
+In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated risk
+function from \pkg{casebase} on the Cox model estimate. As noted in the
+Theoretical Details section, the usual asymptotic results hold for
+likelihood ratio tests built using case-base sampling models. Therefore,
+we can easily test the significance of the spline model:
 
@@
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cox-cif-1} \caption{CIF for control and screening groups in the ERSPC data.
-We plot the CIF from the Cox model using \code{survival::survfit} (solid line) and the CIF from case-base sampling (dashed line).
+\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cox-cif-1} \caption{Risk functions for control and screening groups in the ERSPC data.
+We plot the estimate from the Cox model using \code{survival::survfit} (solid line) and the estimate from case-base sampling (dashed line).
 We used a cubic B-spline expansion of time.}\label{fig:erspc-cox-cif}
+\end{figure}
+\end{Schunk}
+
+Moreover, we can also compute confidence intervals for the risk function
+using the method \code{confint.absRiskCB}.
+
+\begin{Schunk}
+\begin{Sinput}
+risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
+conf_ints <- confint(risk, fit)
+conf_ints
+\end{Sinput}
+\begin{Soutput}
+#>   time estimate conf.low conf.high        cov_prof
+#> 1    0  0.0e+00  0.0e+00   0.0e+00   Control group
+#> 2    1  5.4e-05  3.5e-05   9.3e-05   Control group
+#> 3    0  0.0e+00  0.0e+00   0.0e+00 Screening group
+#> 4    1  4.2e-05  2.7e-05   7.3e-05 Screening group
+\end{Soutput}
+\end{Schunk}
+
+These point-wise confidence intervals are constructed using parametric
+bootstrap; the validity of this approach relies on the asymptotic
+normality of the MLE. In Figure \ref{fig:erspc-cif-conf}, we see the
+95\% confidence bands around the estimates.
+
+\begin{Schunk}
+\begin{figure}[ht]
+\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cif-conf-1} \caption[Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands]{Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands.}\label{fig:erspc-cif-conf}
 \end{figure}
@@
 
-We then compare the cumulative incidence estimation over the test set.
-The probabilities over time for each observation are averaged, resulting
-in the absolute risk curves shown in Figure \ref{fig:cs3FinalBrier} (A).
-We can see some minimal differences between the three models, with the
-Kaplan-Meier giving the lowest estimates across follow-up-time. Note
-that the apparent smoothness of the Cox and penalized Cox curves is due
-to the large number of observations in the training set, which is used
-to derive the Breslow estimate of the baseline hazard. As described
-above, we compare the performance between the models by computing the
-Brier scores over time. In Figure \ref{fig:cs3FinalBrier} (B), we can
-see that the Brier score is larger for the Kaplan-Meier estimate than
-for the other three models. On the other hand, the differences between
-these three models are minimal.
+We then compare the risk estimation over the test set. The probabilities
+over time for each observation are averaged, resulting in the absolute
+risk curves shown in Figure \ref{fig:cs3FinalBrier} (A). We can see some
+minimal differences between the three models, with the Kaplan-Meier
+giving the lowest estimates across follow-up-time. Note that the
+apparent smoothness of the Cox and penalized Cox curves is due to the
+large number of observations in the training set, which is used to
+derive the Breslow estimate of the baseline hazard. As described above,
+we compare the performance between the models by computing the Brier
+scores over time. In Figure \ref{fig:cs3FinalBrier} (B), we can see that
+the Brier score is larger for the Kaplan-Meier estimate than for the
+other three models. On the other hand, the differences between these
+three models are minimal.
 
@@
 provides functions for fitting smooth parametric hazards and estimating
-CIFs using case-base sampling. Our package also provides several
-functions to produce graphical summaries of the data and the results. We
-outlined the theoretical underpinnings of the approach, we provided
-details about our implementation, and we illustrated the merits of the
-approach and the package through three case studies.
+risk functions using case-base sampling. Our package also provides
+several functions to produce graphical summaries of the data and the
+results. We outlined the theoretical underpinnings of the approach, we
+provided details about our implementation, and we illustrated the merits
+of the approach and the package through three case studies.
 
@@
 linear function of predictors (including time); and 3) use the fitted
-hazard to estimate the CIF. Accordingly, our package provides functions
-for each step. Moreover, the simple interface of the
+hazard to estimate the risk function. Accordingly, our package provides
+functions for each step. Moreover, the simple interface of the
 \code{fittingSmoothHazard} function resembles the \code{glm} interface.
@@
 differ significantly. Case-base sampling produces smooth hazards and
-smooth cumulative incidence curves, whereas Cox regression and Fine-Gray
-models produce stepwise CIFs and never explicitly model the hazard
-function. Qualitatively, we showed that by using splines in the linear
-predictor, all three models yielded similar curves. However, the smooth
-nature of the output of \pkg{casebase} provides a more intuitive
-interpretation for consumers of these predictions. In Table
-\ref{tab:compCBvsCox}, we provide a side-by-side comparison between the
-Cox model and case-base sampling.
+smooth survival curves, whereas Cox regression and Fine-Gray models
+produce stepwise CIFs and never explicitly model the hazard function.
+Qualitatively, we showed that by using splines in the linear predictor,
+all three models yielded similar curves. However, the smooth nature of
+the output of \pkg{casebase} provides a more intuitive interpretation
+for consumers of these predictions. In Table \ref{tab:compCBvsCox}, we
+provide a side-by-side comparison between the Cox model and case-base
+sampling.
 
@@
 implements case-base sampling for fitting parametric survival models and
-for estimating smooth cumulative incidence functions using the framework
-of generalized linear models. We strongly believe that its flexibility
-and its foundation on the familiar logistic regression model will make
-it appealing to new and established practitioners. The \pkg{casebase}
+for estimating smooth survival functions using the framework of
+generalized linear models. We strongly believe that its flexibility and
+its foundation on the familiar logistic regression model will make it
+appealing to new and established practitioners. The \pkg{casebase}
 package is freely available from the Comprehensive R Archive Network at

### 2021-10-02 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.R

@@ -40,8 +40,8 @@ eval_introduction <- TRUE
 eval_theory <- TRUE
 eval_implementation <- TRUE
 eval_cs1 <- TRUE
-eval_cs2 <- FALSE
-eval_cs3 <- FALSE
+eval_cs2 <- TRUE
+eval_cs3 <- TRUE
 eval_cs4 <- FALSE
 eval_colophon <- FALSE # Only for draft
 echo_plot_code <- TRUE
@@ -125,7 +125,7 @@ surv_obj <- with(ERSPC, Surv(Follow.Up.Time, DeadOfPrCa))
 cox_model <- survival::coxph(surv_obj ~ ScrArm, data = ERSPC)
 
 
-## ----erspc-cox-cif, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="CIF for control and screening groups in the ERSPC data. We plot the CIF from the Cox model using \\code{survival::survfit} (solid line) and the CIF from case-base sampling (dashed line). We used a cubic B-spline expansion of time."----
+## ----erspc-cox-cif, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk functions for control and screening groups in the ERSPC data. We plot the estimate from the Cox model using \\code{survival::survfit} (solid line) and the estimate from case-base sampling (dashed line). We used a cubic B-spline expansion of time."----
 cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
 myCols <- cbPalette[c(4,7)]
 
@@ -135,7 +135,7 @@ new_time <- seq(0,14,0.1)
 risk_mp <- absoluteRisk(fit, time = new_time, newdata = new_data)
 
 plot(survfit(cox_model, newdata = new_data),
-     ylab = "Cumulative incidence (%)",
+     ylab = "Risk probability (%)",
      xlab = "Years since randomization",
      fun = "event",
      xmax = max(new_time),
@@ -162,7 +162,7 @@ conf_ints <- confint(risk, fit)
 conf_ints
 
 
[…]

### 2021-10-05 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
use reid interview to start intro

--- 2021-09-23 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
@@
   \usepackage{tabu}
-output: 
-  rticles::rjournal_article:
-    number_sections: true
+output: rticles::rjournal_article 
 bibliography: bhatnagar-turgeon-islam-hanley-saarela.bib 
@@
 # Introduction
+
+In a 1994 interview with Professor Nancy Reid, Sir David Cox was asked how he would model a set of censored survival data, to which he responded: "I think I would normally want to tackle problems parametrically ...
+and if you want to do things like predict the outcome for a particular patient, it's much more convenient to do that parametrically" [@reid1994conversation].
+Indeed, the most relevant quantity is often the 5- or 10-year risk of experiencing a certain event given the patient's particular circumstances.
+Unfortunately, the most reported metric from a Cox model is the (potentially time-dependent) hazard ratio (HR), which ignores the duration of follow-up or is subject to selection bias [@hernan2010hazards].
+The covariate adjusted survival curve overcomes these limitations, and is arguably a more important summary measure to report than the HR.
+Stepwise survival curves can be computed with the Cox model but require a second step to separately estimate the baseline hazard [@breslow1972discussion].
+
+In this article, we present the \CRANpkg{casebase} R package [@casebase-package] for fitting fully parametric hazard models and covariate adjusted survival curves using the familiar interface of the \code{glm} function.
+Using a parametric estimator of the hazard function leads to a smooth function of time; as a consequence, the survival function estimates also vary smoothly over time.
+
+Other packages can do this.
+
+the familiar interface of the \code{glm} function.
+It also provides tools for variable selection, competing-risk analysis, and dealing with time-dependent exposures.
+In addition, we provide functions for exploratory data analysis and visualizing the estimated quantities.
 
@@
   - \code{gam}: This option provides support for *Generalized Additive Models* via the \CRANpkg{mgcv} package [@hastie1987generalized].
+  - \code{gbm}: This option provides support for *Gradient Boosted Trees* via the \CRANpkg{gbm} package. This feature is still experimental.
 
@@
 For our first case study, we return to the ERSPC study and investigate the differences in risk between the control and screening arms.
-We fit a flexible hazard by using a smooth function of time modeled with cubic B-splines.
-The model is fit using \code{fitSmoothHazard} with the familiar formula interface:
+We fit four models that differ in which functional form of time is used: 1) excluded from the linear predictor (Eq.
+\eqref{eq:exp}); 2) linear function (Eq.
+\eqref{eq:gomp}); 3) log function (Eq.
+\eqref{eq:weibull}); and 4) a smooth function using cubic B-splines.
+The models are fit using \code{fitSmoothHazard} with the familiar formula interface:
 
@@
 set.seed(12)
-library(splines)
-fit <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
-                       data = ERSPC, ratio = 100)
+fmla <- list(exponential = formula(DeadOfPrCa ~ ScrArm),
+             gompertz = formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
+             weibull = formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
+             splines = formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
+
+fits <- lapply(fmla, function(form) {
+  fitSmoothHazard(form, data = ERSPC, ratio = 100)
+})
 ```
@@
 The output object from \code{fitSmoothHazard} inherits from the \code{singleEventCB} and \code{glm} classes.
-For this reason, we can leverage the \code{summary} method for \code{glm} objects to output a familiar summary of the results:
+For this reason, we leveraged the \code{summary} method for \code{glm} objects to output a familiar summary of the results:
 
 ```{r, eval = TRUE, echo=eval_cs1}
-summary(fit) 
+summary(fits[["splines"]]) 
 ```
@@
 
-```{r, eval=eval_cs1, echo=echo_plot_code}
+```{r, eval=FALSE, echo=echo_plot_code}
 new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
-new_time <- c(0, 1)
-
+new_time <- seq(0, 14, by = 0.1)
+
+risks <- lapply(fits, function(fit) {
   absoluteRisk(fit, time = new_time, newdata = new_data)
-```
-
-In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIF from \pkg{casebase} on the Cox model CIF.
-As noted in the Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models.
-Therefore, we can easily test the significance of the spline model:
+})
+```
+
+In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from \pkg{casebase} on the Cox model CIF.
+The CIF estimates for the exponential model in panel (1) overestimate the cumulative incidence earlier on in time, and underestimate it later on.
+Based on our earlier discussion of the population-time plot, this poor fit for the exponential hazard was expected.
+We notice a better fit with increasing complexity of our model for time in Figure \ref{fig:erspc-cox-cif} (panels 2--4).
+As noted in Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models.
+Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) spline model:
 
 ```{r erscp-compare, eval = eval_cs1, echo = FALSE}
-anova(fit, test = "LRT")
-```
-
-The null hypothesis is rejected in favor of the spline model.
-Similarly, to compare different models (e.g.
-time modeled linearly), we could compute Akaike's Information Criterion (AIC) for each model and compare them.
+anova(fits$exponential, fits$splines, test = "LRT")
+```
+
+The null hypothesis is rejected in favor of the spline model. Similarly, the AIC provides further evidence that the flexible function of time provides the best fit:
+
+```{r erscp-aic, eval = eval_cs1, echo=FALSE}
+c("Exp." = AIC(fits$exponential),
+  "Gompertz" = AIC(fits$gompertz),
+  "Weibull" = AIC(fits$weibull),
+  "Splines" = AIC(fits$splines))
+```
 
@@
 
-```{r erspc-cox-cif, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="CIF for control and screening groups in the ERSPC data.
-We plot the CIF from the Cox model using \\code{survival::survfit} (solid line) and the CIF from case-base sampling (dashed line).
-We used a cubic B-spline expansion of time."}
+```{r erspc-cox-cif, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="CIFs for control and screening groups in the ERSPC data.
+In each of the panels, we plot the CIF from the Cox model using \\code{survival::survfit} (solid line) and the CIF from the case-base sampling scheme (dashed line) with different functional forms of time.
+(1) The time variable is excluded (exponential).
+(2) Linear function of time (Gompertz).
+(3) The natural logarithm (Weibull).
+(4) Cubic B-spline expansion of time."}
 cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
@@
 new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
-new_time <- seq(0,14,0.1)
-# Recompute with more time points
-risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
-
+new_time <- seq(0,14.5,0.1)
+
+risks <- lapply(fits, function(i) {
+  casebase::absoluteRisk(object = i, time = new_time, newdata = new_data, method = "mont")
+})
+
+layout(matrix(1:4, ncol = 2, byrow = FALSE), respect = FALSE)
+par(mar = c(0, 4.1, 4.1, 0))
 plot(survfit(cox_model, newdata = new_data),
-     ylab = "Cumulative incidence (%)",
-     xlab = "Years since randomization",
+     ylab = "",
+     xlab = "",
      fun = "event",
@@
        bg = "gray90")
-lines(risk[,1], risk[,2], col = myCols[1], lty = 2, lwd = 2)
-lines(risk[,1], risk[,3], col = myCols[2], lty = 2, lwd = 2)
+lines(risks$exponential[,1], risks$exponential[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risks$exponential[,1], risks$exponential[,3], col = myCols[2], lty = 2, lwd = 2)
+text(x = max(new_time)*0.97, y = 0.05/100, labels = "(1)", cex = 1.5)
+
+par(mar = c(4.1, 4.1, 0, 0))
+plot(survfit(cox_model, newdata = new_data),
+     ylab = "", xlab = "",
+     fun = "event",
+     xmax = max(new_time),
+     conf.int = FALSE, col = myCols, 
+     lwd = 2,
+     yscale = 100)
+lines(risks$weibull[,1], risks$weibull[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risks$weibull[,1], risks$weibull[,3], col = myCols[2], lty = 2, lwd = 2)  
+text(x = max(new_time)*0.97, y = 0.05/100, labels = "(3)", cex = 1.5)
+mtext("Cumulative incidence (%)", side = 2, line = -1.5, outer = TRUE)
+
+par(mar = c(0, 0, 4.1, 3.9))
+plot(survfit(cox_model, newdata = new_data),
+     ylab = "",xlab = "",
+     fun = "event",
+      xaxt = 'n', yaxt = 'n',
+     xmax = max(new_time),
+     conf.int = FALSE, col = myCols, 
+     lwd = 2,
+     yscale = 100)
+lines(risks$gompertz[,1], risks$gompertz[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risks$gompertz[,1], risks$gompertz[,3], col = myCols[2], lty = 2, lwd = 2)
+text(x = max(new_time)*0.97, y = 0.05/100, labels = "(2)", cex = 1.5)
+
+par(mar = c(4.1, 0, 0, 3.9))
+plot(survfit(cox_model, newdata = new_data),
+     ylab = "",xlab = "",
+     fun = "event",
+     yaxt = 'n',
+     xmax = max(new_time),
+     conf.int = FALSE, col = myCols, 
+     lwd = 2,
+     yscale = 100)
+lines(risks$splines[,1], risks$splines[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risks$splines[,1], risks$splines[,3], col = myCols[2], lty = 2, lwd = 2)
+text(x = max(new_time)*0.97, y = 0.05/100, labels = "(4)", cex = 1.5)
+mtext("Years since randomization", side = 1, line = -1.5, outer = TRUE)
+```
+
+```{r erspc-95ci22, echo = FALSE, eval = eval_cs1}
+CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
+
+aft_fits <- lapply(list("exp","gompertz","weibull"), function(i)
+  flexsurv::flexsurvreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm,
+                           data = ERSPC, dist = i)
+  )
+
+names(aft_fits) <- list("exponential","gompertz","weibull")
+
+# function to exponentiate HR and CI and return a HR (lower, Upper) character
+exp_est_ci <- function(model = c("exponential","gompertz","weibull","splines"),
+                       fit, 
+                       covariate = "ScrArmScreening group", 
+                       type = c("casebase", "survreg")) {
+
+  type <- match.arg(type)
+  covariate <- match.arg(covariate)
+
+  if (type == "casebase"){
+
+    cis <- confint(fit[[model]])
+
+    sprintf("%0.2f (%0.2f, %0.2f)",
+            exp(coef(fit[[model]]))[[covariate]],
+            exp(cis)[covariate,1],
+            exp(cis)[covariate,2])
+  } else {
+
+    if (model == "weibull") {
+
+      # The maximized log-likelihoods are the same between survreg and flexsurvreg, 
+      # however the parameterisation is different: the first coefficient (Intercept) 
+      # reported by survreg is log(mu), and survreg's "scale" is
+      # dweibull's (thus flexsurvreg)'s 1 / shape
+      res <- -fit[[model]]$res[covariate,] / (1 / fit[[model]]$res["shape",])
+
+      sprintf("%0.2f (%0.2f, %0.2f)",
+              exp(res[["est"]]),
+              exp(res[["U95%"]]),
+              exp(res[["L95%"]]))
+
+    } else if (model == "splines") {
+
+      c("--")
+
+    } else {
+
+      res <- fit[[model]]$res[covariate,]
+
+      sprintf("%0.2f (%0.2f, %0.2f)",
+              exp(res[["est"]]),
+              exp(res[["L95%"]]),
+              exp(res[["U95%"]]))
+    }
+  }
+}
+
+casebase_HR_CI <- lapply(list("exponential","gompertz","weibull","splines"), 
+                         exp_est_ci, 
+                         fit = fits,
+                         covariate = "ScrArmScreening group", 
+                         type = "casebase")
+
+survreg_HR_CI <- lapply(list("exponential","gompertz","weibull","splines"), 
+                         exp_est_ci, 
+                         fit = aft_fits,
+                         covariate = "ScrArmScreening group", 
+                         type = "survreg")
+
+ptable <- cbind(list("Exponential","Gompertz","Weibull","Splines"),
+            do.call(rbind, casebase_HR_CI), do.call(rbind, survreg_HR_CI))
+```
+
+In Table \ref{tab:print-erspc-estimates}, we present a side-by-side comparison of the hazard ratios and confidence intervals estimated from \code{fitSmoothHazard} and the corresponding parametric model using \code{survival::survreg}, as well as the Cox model estimate.
+The hazard ratio estimates and confidence intervals are similar across all four models.
+This reinforces the idea that, under proportional hazards, we do not need to model the full hazard to obtain reliable estimates of the hazard ratio.
+Nevertheless, Figure \ref{fig:erspc-cox-cif} shows that different parametric models can still give rise to qualitatively different estimates for the CIF.
+
+```{r print-erspc-estimates, eval = eval_cs1}
+kable(ptable, "latex", booktabs = TRUE, align = c("l","c","c"),
+      caption = c("Comparison of estimated hazard ratios and 95\\% confidence intervals for ERSPC data."),
+      col.names = c("Model","casebase::fitSmoothHazard","survival::survreg")) %>%
+  kable_styling() %>%
+  column_spec(1, bold = TRUE) %>%
+  collapse_rows(columns = 1, latex_hline = "major", valign = "middle") %>%
+  footnote(general = sprintf("Cox model estimate: HR (95%% CI) = %0.2f (%0.2f, %0.2f)", CI_cox[1], CI_cox[2], CI_cox[3]),
+           general_title = "") 
 ```

### 2021-10-05 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
draft of first two intro paragraphs:
1) talk about why Cox would rather parametric and hazard of hazard ratios. 2) introduce efron and arjas who categorize time and use logistic. then explain that H&M dont categorize time. they select person moments in continuous time

@@
 
+The semiparametric Cox model has become the default approach to survival analysis even though Cox himself later suggested he would prefer to model the hazard function directly.
 In a 1994 interview with Professor Nancy Reid, Sir David Cox was asked how he would model a set of censored survival data, to which he responded: "I think I would normally want to tackle problems parametrically ...
 and if you want to do things like predict the outcome for a particular patient, it's much more convenient to do that parametrically" [@reid1994conversation].
-Indeed, the most relevant quantity is often the 5- or 10-year risk of experiencing a certain event given the patient's particular circumstances.
+Indeed, the most relevant quantity in a clinical setting is often the 5- or 10-year risk of experiencing a certain event given the patient's particular circumstances.
 Unfortunately, the most reported metric from a Cox model is the (potentially time-dependent) hazard ratio (HR), which ignores the duration of follow-up or is subject to selection bias [@hernan2010hazards].
 The covariate adjusted survival curve overcomes these limitations, and is arguably a more important summary measure to report than the HR.
-Stepwise survival curves can be computed with the Cox model but require a second step to separately estimate the baseline hazard [@breslow1972discussion].
+While stepwise survival curves can be computed with the Cox model, they require a second step to separately estimate the baseline hazard [@breslow1972discussion].
+
+[-@arjas1987logistic] and [-@efron1988logistic] were some of the first authors to pursue approaches that made the fitting of smooth hazard functions much more transparent and intuitive through generalized linear models.
+They focused on categorizing the time axis (e.g days) and treated each patient-day as a Bernoulli random variable with probability equal to the discrete hazard rate.
+Rather than discretizing time, Hanley & Miettinen [-@hanley2009fitting] selected a discrete set of person-time coordinates (`person-moments') in continuous time from all observed follow-up experience constituting the study base to obtain the logistic form likelihood expression.
+More specifically, all person-moments when the event of interest occurred are selected as the case series, complemented by a randomly sampled base series of person-moments serving as controls.
+
+[-@copenhagen2012needs] has an R package that uses as the slices of person time as the units of analysis.
+Can keep subdividing them, but in the limit the file becomes infinite.
+Cite estimation issue from IRLS in Kalbfleish and Prentice.
+
+Just like Efron, we can use the well-understood logistic regression for everything.
+We are not saying it is better, just an attractive alternative (and a general framework).
+One that with our package, is well within the reach of more endusers.
+
+We also bring hazard functions out of the closet. Where they tend to be kept by the Cox Model.
+
+With the goal of fitting smooth-in-time hazard functions, Hanley & Miettinen [-@hanley2009fitting] proposed a general framework for estimating fully parametric hazard models via logistic regression.
+
+Their main idea is simple: comparing person-moments when the event of interest occurred with moments when patients were at risk.
+Their approach handles censored data effortlessly and provides users familiar with generalized linear models a natural way of fitting parametric survival models.
+Moreover, their framework is very flexible: general functions of time can be estimated (e.g.
+using splines or general additive models), and hence these models retain some of the flexibility of Cox's partial likelihood approach.
+And since the unit of analysis is a person moment, time-varying covariates can also easily be included in this framework.
 

### 2021-10-05 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
draft of first two intro paragraphs:
1) talk about why Cox would rather parametric and hazard of hazard ratios. 2) introduce efron and arjas who categorize time and use logistic. then explain that H&M dont categorize time. they select person moments in continuous time

--- 2021-09-23 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
@@
 \section{Introduction}\label{introduction}}
+
+In a 1994 interview with Professor Nancy Reid, Sir David Cox was asked
+how he would model a set of censored survival data, to which he
+responded: ``I think I would normally want to tackle problems
+parametrically \ldots{} and if you want to do things like predict the
+outcome for a particular patient, it's much more convenient to do that
+parametrically'' \citep{reid1994conversation}. Indeed, the most relevant
+quantity in a clinical setting is often the 5- or 10-year risk of
+experiencing a certain event given the patient's particular
+circumstances. Unfortunately, the most reported metric from a Cox model
+is the (potentially time-dependent) hazard ratio (HR), which ignores the
+duration of follow-up or is subject to selection bias
+\citep{hernan2010hazards}. The covariate adjusted survival curve
+overcomes these limitations, and is arguably a more important summary
+measure to report than the HR. Stepwise survival curves can be computed
+with the Cox model but require a second step to separately estimate the
+baseline hazard \citep{breslow1972discussion}.
+
+In this article, we present the \CRANpkg{casebase} R package
+\citep{casebase-package} for fitting fully parametric hazard models and
+covariate adjusted survival curves using the familiar interface of the
+\code{glm} function. Using a parametric estimator of the hazard function
+leads to a smooth function of time; as a consequence, the survival
+function estimates also vary smoothly over time.
+
+Other packages can do this.
+
+the familiar interface of the \code{glm} function. It also provides
+tools for variable selection, competing-risk analysis, and dealing with
+time-dependent exposures. In addition, we provide functions for
+exploratory data analysis and visualizing the estimated quantities.
 
@@
   \citep{hastie1987generalized}.
+\item
+  \code{gbm}: This option provides support for \emph{Gradient Boosted
+  Trees} via the \CRANpkg{gbm} package. This feature is still
+  experimental.
 \end{itemize}
@@
 For our first case study, we return to the ERSPC study and investigate
-the differences in risk between the control and screening arms. We fit a
-flexible hazard by using a smooth function of time modeled with cubic
-B-splines. The model is fit using \code{fitSmoothHazard} with the
-familiar formula interface:
+the differences in risk between the control and screening arms. We fit
+four models that differ in which functional form of time is used: 1)
+excluded from the linear predictor (Eq. \eqref{eq:exp}); 2) linear
+function (Eq. \eqref{eq:gomp}); 3) log function (Eq.
+\eqref{eq:weibull}); and 4) a smooth function using cubic B-splines. The
+models are fit using \code{fitSmoothHazard} with the familiar formula
+interface:
 
@@
 \begin{Sinput}
-library(splines)
-fit <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
-                       data = ERSPC, ratio = 100)
+fmla <- list(exponential = formula(DeadOfPrCa ~ ScrArm),
+             gompertz = formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
+             weibull = formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
+             splines = formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))
+
+fits <- lapply(fmla, function(form) {
+  fitSmoothHazard(form, data = ERSPC, ratio = 100)
+})
 \end{Sinput}
@@
 The output object from \code{fitSmoothHazard} inherits from the
-\code{singleEventCB} and \code{glm} classes. For this reason, we can
-leverage the \code{summary} method for \code{glm} objects to output a
+\code{singleEventCB} and \code{glm} classes. For this reason, we
+leveraged the \code{summary} method for \code{glm} objects to output a
 familiar summary of the results:
@@
 \begin{Sinput}
-summary(fit) 
+summary(fits[["splines"]]) 
 \end{Sinput}
@@
 #> Call:
-#> fitSmoothHazard(formula = DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
-#>     data = ERSPC, ratio = 100)
+#> fitSmoothHazard(formula = form, data = ERSPC, ratio = 100)
 #> 
@@
 #>    Min      1Q  Median      3Q     Max  
-#> -0.403  -0.172  -0.135  -0.085   3.801  
+#> -0.409  -0.172  -0.135  -0.084   3.808  
 #> 
@@
 #>                       Estimate Std. Error z value Pr(>|z|)    
-#> (Intercept)           -10.2533     0.3168  -32.36  < 2e-16 ***
-#> bs(Follow.Up.Time)1     4.2544     0.7976    5.33  9.6e-08 ***
-#> bs(Follow.Up.Time)2     2.1371     0.4720    4.53  5.9e-06 ***
-#> bs(Follow.Up.Time)3     4.5339     0.6889    6.58  4.7e-11 ***
-#> ScrArmScreening group  -0.2481     0.0886   -2.80   0.0051 ** 
+#> (Intercept)           -10.2905     0.3185  -32.31  < 2e-16 ***
+#> bs(Follow.Up.Time)1     4.3587     0.8008    5.44  5.2e-08 ***
+#> bs(Follow.Up.Time)2     2.0890     0.4721    4.42  9.7e-06 ***
+#> bs(Follow.Up.Time)3     4.6266     0.6910    6.70  2.1e-11 ***
+#> ScrArmScreening group  -0.2354     0.0886   -2.66   0.0079 ** 
 #> ---
@@
 #>     Null deviance: 6059.0  on 54539  degrees of freedom
-#> Residual deviance: 5787.6  on 54535  degrees of freedom
-#> AIC: 5798
+#> Residual deviance: 5785.3  on 54535  degrees of freedom
+#> AIC: 5795
 #> 
-#> Number of Fisher Scoring iterations: 8
+#> Number of Fisher Scoring iterations: 9
 \end{Soutput}
@@
 new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
-new_time <- c(0, 1)
-
+new_time <- seq(0, 14, by = 0.1)
+
+risks <- lapply(fits, function(fit) {
   absoluteRisk(fit, time = new_time, newdata = new_data)
+})
 \end{Sinput}
-\begin{Soutput}
-#>  time                
-#>     0 0.0e+00 0.0e+00
-#>     1 5.4e-05 4.2e-05
-\end{Soutput}
-\end{Schunk}
-
-In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIF from
-\pkg{casebase} on the Cox model CIF. As noted in the Theoretical Details
-section, the usual asymptotic results hold for likelihood ratio tests
-built using case-base sampling models. Therefore, we can easily test the
-significance of the spline model:
+\end{Schunk}
+
+In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from
+\pkg{casebase} on the Cox model CIF. The CIF estimates for the
+exponential model in panel (1) overestimate the cumulative incidence
+earlier on in time, and underestimate it later on. Based on our earlier
+discussion of the population-time plot, this poor fit for the
+exponential hazard was expected. We notice a better fit with increasing
+complexity of our model for time in Figure \ref{fig:erspc-cox-cif}
+(panels 2--4). As noted in Theoretical Details section, the usual
+asymptotic results hold for likelihood ratio tests built using case-base
+sampling models. Therefore, we can easily test the null hypothesis that
+the exponential model is just as good as the larger (in terms of number
+of parameters) spline model:
 
@@
 #> 
-#> Model: binomial, link: logit
-#> 
-#> Response: DeadOfPrCa
-#> 
-#> Terms added sequentially (first to last)
-#> 
-#> 
-#>                    Df Deviance Resid. Df Resid. Dev Pr(>Chi)    
-#> NULL                               54539       6059             
-#> bs(Follow.Up.Time)  3      263     54536       5796   <2e-16 ***
-#> ScrArm              1        8     54535       5788   0.0048 ** 
+#> Model 1: DeadOfPrCa ~ ScrArm + offset(offset)
+#> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
+#>   Resid. Df Resid. Dev Df Deviance Pr(>Chi)    
+#> 1     54538       6052                         
+#> 2     54535       5785  3      267   <2e-16 ***
 #> ---
@@
 The null hypothesis is rejected in favor of the spline model. Similarly,
-to compare different models (e.g.~time modeled linearly), we could
-compute Akaike's Information Criterion (AIC) for each model and compare
-them.
+the AIC provides further evidence that the flexible function of time
+provides the best fit:
+
+\begin{Schunk}
+\begin{Soutput}
+#>     Exp. Gompertz  Weibull  Splines 
+#>     6056     5821     5807     5795
+\end{Soutput}
+\end{Schunk}
 
@@
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cox-cif-1} \caption{CIF for control and screening groups in the ERSPC data.
-We plot the CIF from the Cox model using \code{survival::survfit} (solid line) and the CIF from case-base sampling (dashed line).
-We used a cubic B-spline expansion of time.}\label{fig:erspc-cox-cif}
+\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cox-cif-1} \caption{CIFs for control and screening groups in the ERSPC data.
+In each of the panels, we plot the CIF from the Cox model using \code{survival::survfit} (solid line) and the CIF from the case-base sampling scheme (dashed line) with different functional forms of time.
+(1) The time variable is excluded (exponential).
+(2) Linear function of time (Gompertz).
+(3) The natural logarithm (Weibull).
+(4) Cubic B-spline expansion of time.}\label{fig:erspc-cox-cif}
 \end{figure}
+\end{Schunk}
+
+In Table \ref{tab:print-erspc-estimates}, we present a side-by-side
+comparison of the hazard ratios and confidence intervals estimated from
+\code{fitSmoothHazard} and the corresponding parametric model using
+\code{survival::survreg}, as well as the Cox model estimate. The hazard
+ratio estimates and confidence intervals are similar across all four
+models. This reinforces the idea that, under proportional hazards, we do
+not need to model the full hazard to obtain reliable estimates of the
+hazard ratio. Nevertheless, Figure \ref{fig:erspc-cox-cif} shows that
+different parametric models can still give rise to qualitatively
+different estimates for the CIF.
+
+\begin{Schunk}
+\begin{table}
+
+\caption{\label{tab:print-erspc-estimates}Comparison of estimated hazard ratios and 95\% confidence intervals for ERSPC data.}
+\centering
+\begin{tabular}[t]{>{}lcc}
+\toprule
+Model & casebase::fitSmoothHazard & survival::survreg\\
+\midrule
+\textbf{Exponential} & 0.79 (0.67, 0.94) & 0.81 (0.68, 0.96)\\
+\cmidrule{1-3}
+\textbf{Gompertz} & 0.80 (0.67, 0.95) & 0.80 (0.67, 0.95)\\
+\cmidrule{1-3}
+\textbf{Weibull} & 0.81 (0.68, 0.96) & 0.80 (0.65, 0.96)\\
+\cmidrule{1-3}
+\textbf{Splines} & 0.79 (0.66, 0.94) & --\\
+\bottomrule
+\multicolumn{3}{l}{\rule{0pt}{1em}Cox model estimate: HR (95\% CI) = 0.80 (0.67, 0.95)}\\
+\end{tabular}
+\end{table}
+
 \end{Schunk}
@@
 %
-\url{http://sahirbhatnagar.com/}\\%
+\url{http://sahirbhatnagar.com/}%
 %
-\email{sahir.bhatnagar@mcgill.ca}%
+\\\email{sahir.bhatnagar@mcgill.ca}
 }
@@
 %
-\url{https://maxturgeon.ca/}\\%
+\url{https://maxturgeon.ca/}%
 %
-\email{max.turgeon@umanitoba.ca}%
+\\\email{max.turgeon@umanitoba.ca}
 }
@@
 %
-\email{jesse.islam@mail.mcgill.ca}%
+\\\email{jesse.islam@mail.mcgill.ca}
 }
@@
 %
-\url{http://www.medicine.mcgill.ca/epidemiology/hanley/}\\%
+\url{http://www.medicine.mcgill.ca/epidemiology/hanley/}%
 %
-\email{james.hanley@mcgill.ca}%
+\\\email{james.hanley@mcgill.ca}
 }
@@
 %
-\url{http://individual.utoronto.ca/osaarela/}\\%
+\url{http://individual.utoronto.ca/osaarela/}%
 %
-\email{olli.saarela@utoronto.ca}%
+\\\email{olli.saarela@utoronto.ca}
 }

### 2021-10-06 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
intro revised

--- 2021-10-05 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
@@
 
-In a 1994 interview with Professor Nancy Reid, Sir David Cox was asked how he would model a set of censored survival data, to which he responded: "I think I would normally want to tackle problems parametrically ...
-and if you want to do things like predict the outcome for a particular patient, it's much more convenient to do that parametrically" [@reid1994conversation].
-Indeed, the most relevant quantity is often the 5- or 10-year risk of experiencing a certain event given the patient's particular circumstances.
-Unfortunately, the most reported metric from a Cox model is the (potentially time-dependent) hazard ratio (HR), which ignores the duration of follow-up or is subject to selection bias [@hernan2010hazards].
-The covariate adjusted survival curve overcomes these limitations, and is arguably a more important summary measure to report than the HR.
-Stepwise survival curves can be computed with the Cox model but require a second step to separately estimate the baseline hazard [@breslow1972discussion].
-
-In this article, we present the \CRANpkg{casebase} R package [@casebase-package] for fitting fully parametric hazard models and covariate adjusted survival curves using the familiar interface of the \code{glm} function.
-Using a parametric estimator of the hazard function leads to a smooth function of time; as a consequence, the survival function estimates also vary smoothly over time.
-
-Other packages can do this.
-
-the familiar interface of the \code{glm} function.
-It also provides tools for variable selection, competing-risk analysis, and dealing with time-dependent exposures.
-In addition, we provide functions for exploratory data analysis and visualizing the estimated quantities.
-
-Survival analysis and the comparison of event rates has been greatly influenced over the last 50 years by the partial likelihood approach of the Cox proportional hazard model [@cox1972regression].
-This approach provides a flexible way of assessing the influence of covariates on the hazard function, without the need to specify a parametric survival model.
-This flexibility comes at the cost of decoupling the baseline hazard from the effect of the covariates.
-To recover the whole survival curve---or the cumulative incidence function (CIF)---we then need to separately estimate the baseline hazard [@breslow1972discussion].
-This in turn leads to stepwise estimates of the survival function that can be difficult to interpret.
-
-From the perspective of clinicians and their patients, the most relevant quantity is often the 5- or 10-year risk of experiencing a certain event given the patient's particular circumstances, and not the hazard ratio between a treatment and control group.
-Therefore, to make sound clinical decisions, it is important to accurately estimate the *full* hazard function, which can subsequently be used to estimate the CIF.
-Using a parametric estimator of the hazard function leads to a smooth function of time; as a consequence, the CIF and the survival function estimates also vary smoothly over time.
-
-With the goal of fitting smooth-in-time hazard functions, Hanley & Miettinen [-@hanley2009fitting] proposed a general framework for estimating fully parametric hazard models via logistic regression.
-Their main idea is simple: comparing person-moments when the event of interest occurred with moments when patients were at risk.
-Their approach handles censored data effortlessly and provides users familiar with generalized linear models a natural way of fitting parametric survival models.
-Moreover, their framework is very flexible: general functions of time can be estimated (e.g.
-using splines or general additive models), and hence these models retain some of the flexibility of Cox's partial likelihood approach.
-And since the unit of analysis is a person moment, time-varying covariates can also easily be included in this framework.
-
-Logistic regression had already been used in the context of discrete-time survival modeling [@cox1972regression].
-But in the context of continuous-time survival modeling, using the framework of logistic regression opens the door to an extensive array of powerful modeling tools.
-Indeed, lasso and elastic-net regression can be used to select variables that are associated with the hazard function.
-Logistic regression can also be replaced by multinomial regression to analyse data with multiple competing events.
-
-In this article, we present the \CRANpkg{casebase} package [@casebase-package] implemented in R, for parametric survival analysis that combines the ideas of Hanley & Miettinen [-@hanley2009fitting] into a simple interface.
-The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to compute a patient's risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile.
-Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function.
-It also provides tools for variable selection, competing-risk analysis, and dealing with time-dependent exposures.
-In addition, we provide functions for exploratory data analysis and visualizing the estimated quantities.
-
-In what follows, we first recall some theoretical details on case-base sampling and its use for estimating parametric hazard functions.
-We then give a short review of existing R packages that implement comparable features as \pkg{casebase}.
-Next, we provide some details about the implementation of case-base sampling in our package, and we give a brief survey of its main functions.
-This is followed by three case studies that illustrate the flexibility and capabilities of \pkg{casebase}.
-We show how the same framework can be used for non-linear functions of time and non-proportional hazards, competing risks and variable selection via penalized regression.
-Finally, we end the article with a discussion of the results and of future directions.
+The semiparametric Cox model has become the default approach to survival analysis 
+even though Cox himself later suggested he would prefer to model the hazard function 
+directly. In a 1994 interview with Professor Nancy Reid, Sir David Cox was asked 
+how he would model a set of censored survival data, to which he responded: "I 
+think I would normally want to tackle problems parametrically ... and if you want 
+to do things like predict the outcome for a particular patient, it's much more 
+convenient to do that parametrically" [@reid1994conversation]. Indeed, the most 
+relevant quantity in a clinical setting is often the 5- or 10-year risk of 
+experiencing a certain event given the patient's particular circumstances. 
+Unfortunately, the most reported metric from a Cox model is the (potentially 
+time-dependent) hazard ratio (HR), which ignores the duration of follow-up or 
+is subject to selection bias [@hernan2010hazards]. The covariate adjusted 
+survival curve overcomes these limitations, and is arguably a more important 
+summary measure to report than the HR. While stepwise survival curves can be 
+computed with the Cox model, they require a second step to separately estimate 
+the baseline hazard [@breslow1972discussion].
+
+Several authors have since pursued fully parametric approaches that made the 
+fitting of smooth hazard functions much more transparent and intuitive through 
+generalized linear models. The key feature of these procedures is splitting the 
+time axis into discrete intervals. [@whitehead1980fitting] showed the 
+equivalence between a Cox model and a Poisson regression with a parameter for 
+each event time. [@copenhagen2012needs] provides a nice exposition of this 
+equivalence with a real data example and supporting R code for computing 
+standard errors. [@arjas1987logistic] and [@efron1988logistic] treated each 
+patient-day as a Bernoulli random variable with probability equal to the 
+discrete hazard rate. A potential issue with these approaches is that the 
+number of time bins need to be chosen by the data analyst. A fine grouping 
+of times may result in few (or none) events per interval, which then leads 
+to instability in the Newton-Raphson procedure for estimation 
+[@kalbfleisch2011statistical]. A coarse grouping could potentially mask 
+nonlinear trends in the hazard function.
+
+Rather than discretizing time, Hanley & Miettinen [-@hanley2009fitting] 
+selected a discrete set of person-time coordinates (`person-moments') in 
+continuous time from all observed follow-up experience constituting the 
+study base to obtain the logistic form likelihood expression for the hazard 
+function. More specifically, all person-moments when the event of interest 
+occurred are selected as the case series, complemented by a randomly sampled 
+base series of person-moments serving as controls. This approach allows 
+flexible modeling of the hazard function by including time as a covariate 
+(e.g. using splines or general additive models). Furthermore, time-dependent 
+covariates can be modeled through interactions with time. In short, Hanley & 
+Miettinen [-@hanley2009fitting] use the well-understood logistic regression 
+for directly modeling the hazard function, but do not require an arbitrary 
+choice of splitting time.
+
+In this article, we present the \CRANpkg{casebase} R package [@casebase-package] 
+which implements the Hanley & Miettinen [-@hanley2009fitting] approach for 
+fitting fully parametric hazard models and covariate adjusted survival curves 
+using the familiar interface of the \code{glm} function. Our implementation 
+allows for straightforward extensions to other models such as penalized 
+regression for variable selection and competing-risk analysis. In addition, 
+we provide functions for exploratory data analysis and visualizing the estimated 
+quantities such as the hazard function, survival curve and its standard errors. 
+The ultimate goal of our package is to make fitting flexible hazards accessible 
+to more endusers with the hopes that they will favor reporting absolute risks 
+over hazard ratios.
+
+In what follows, we first recall some theoretical details on case-base sampling 
+and its use for estimating parametric hazard functions. We then give a short 
+review of existing R packages that implement comparable features as \pkg{casebase}. 
+Next, we provide some details about the implementation of case-base sampling in 
+our package, and we give a brief survey of its main functions. This is followed 
+by three case studies that illustrate the flexibility and capabilities of 
+\pkg{casebase}. We show how the same framework can be used for non-linear 
+functions of time and non-proportional hazards, competing risks and variable 
+selection via penalized regression. Finally, we end the article with a 
+discussion of the results and of future directions.
 

### 2021-10-06 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
intro revised

@@
 
-In a 1994 interview with Professor Nancy Reid, Sir David Cox was asked
-how he would model a set of censored survival data, to which he
-responded: ``I think I would normally want to tackle problems
-parametrically \ldots{} and if you want to do things like predict the
-outcome for a particular patient, it's much more convenient to do that
-parametrically'' \citep{reid1994conversation}. Indeed, the most relevant
-quantity in a clinical setting is often the 5- or 10-year risk of
-experiencing a certain event given the patient's particular
-circumstances. Unfortunately, the most reported metric from a Cox model
-is the (potentially time-dependent) hazard ratio (HR), which ignores the
-duration of follow-up or is subject to selection bias
+The semiparametric Cox model has become the default approach to survival
+analysis even though Cox himself later suggested he would prefer to
+model the hazard function directly. In a 1994 interview with Professor
+Nancy Reid, Sir David Cox was asked how he would model a set of censored
+survival data, to which he responded: ``I think I would normally want to
+tackle problems parametrically \ldots{} and if you want to do things
+like predict the outcome for a particular patient, it's much more
+convenient to do that parametrically'' \citep{reid1994conversation}.
+Indeed, the most relevant quantity in a clinical setting is often the 5-
+or 10-year risk of experiencing a certain event given the patient's
+particular circumstances. Unfortunately, the most reported metric from a
+Cox model is the (potentially time-dependent) hazard ratio (HR), which
+ignores the duration of follow-up or is subject to selection bias
 \citep{hernan2010hazards}. The covariate adjusted survival curve
 overcomes these limitations, and is arguably a more important summary
-measure to report than the HR. Stepwise survival curves can be computed
-with the Cox model but require a second step to separately estimate the
-baseline hazard \citep{breslow1972discussion}.
+measure to report than the HR. While stepwise survival curves can be
+computed with the Cox model, they require a second step to separately
+estimate the baseline hazard \citep{breslow1972discussion}.
+
+Several authors have since pursued fully parametric approaches that made
+the fitting of smooth hazard functions much more transparent and
+intuitive through generalized linear models. The key feature of these
+procedures is splitting the time axis into discrete intervals.
+\citep{whitehead1980fitting} showed the equivalence between a Cox model
+and a Poisson regression with a parameter for each event time.
+\citep{copenhagen2012needs} provides a nice exposition of this
+equivalence with a real data example and supporting R code for computing
+standard errors. \citep{arjas1987logistic} and \citep{efron1988logistic}
+treated each patient-day as a Bernoulli random variable with probability
+equal to the discrete hazard rate. A potential issue with these
+approaches is that the number of time bins need to be chosen by the data
+analyst. A fine grouping of times may result in few (or none) events per
+interval, which then leads to instability in the Newton-Raphson
+procedure for estimation \citep{kalbfleisch2011statistical}. A coarse
+grouping could potentially mask nonlinear trends in the hazard function.
+
+Rather than discretizing time, Hanley \& Miettinen
+\citeyearpar{hanley2009fitting} selected a discrete set of person-time
+coordinates (`person-moments') in continuous time from all observed
+follow-up experience constituting the study base to obtain the logistic
+form likelihood expression for the hazard function. More specifically,
+all person-moments when the event of interest occurred are selected as
+the case series, complemented by a randomly sampled base series of
+person-moments serving as controls. This approach allows flexible
+modeling of the hazard function by including time as a covariate
+(e.g.~using splines or general additive models). Furthermore,
+time-dependent covariates can be modeled through interactions with time.
+In short, Hanley \& Miettinen \citeyearpar{hanley2009fitting} use the
+well-understood logistic regression for directly modeling the hazard
+function, but do not require an arbitrary choice of splitting time.
 
 In this article, we present the \CRANpkg{casebase} R package
-\citep{casebase-package} for fitting fully parametric hazard models and
-covariate adjusted survival curves using the familiar interface of the
-\code{glm} function. Using a parametric estimator of the hazard function
-leads to a smooth function of time; as a consequence, the survival
-function estimates also vary smoothly over time.
-
-Other packages can do this.
-
-the familiar interface of the \code{glm} function. It also provides
-tools for variable selection, competing-risk analysis, and dealing with
-time-dependent exposures. In addition, we provide functions for
-exploratory data analysis and visualizing the estimated quantities.
-
-Survival analysis and the comparison of event rates has been greatly
-influenced over the last 50 years by the partial likelihood approach of
-the Cox proportional hazard model \citep{cox1972regression}. This
-approach provides a flexible way of assessing the influence of
-covariates on the hazard function, without the need to specify a
-parametric survival model. This flexibility comes at the cost of
-decoupling the baseline hazard from the effect of the covariates. To
-recover the whole survival curve---or the cumulative incidence function
-(CIF)---we then need to separately estimate the baseline hazard
-\citep{breslow1972discussion}. This in turn leads to stepwise estimates
-of the survival function that can be difficult to interpret.
-
-From the perspective of clinicians and their patients, the most relevant
-quantity is often the 5- or 10-year risk of experiencing a certain event
-given the patient's particular circumstances, and not the hazard ratio
-between a treatment and control group. Therefore, to make sound clinical
-decisions, it is important to accurately estimate the \emph{full} hazard
-function, which can subsequently be used to estimate the CIF. Using a
-parametric estimator of the hazard function leads to a smooth function
-of time; as a consequence, the CIF and the survival function estimates
-also vary smoothly over time.
-
-With the goal of fitting smooth-in-time hazard functions, Hanley \&
-Miettinen \citeyearpar{hanley2009fitting} proposed a general framework
-for estimating fully parametric hazard models via logistic regression.
-Their main idea is simple: comparing person-moments when the event of
-interest occurred with moments when patients were at risk. Their
-approach handles censored data effortlessly and provides users familiar
-with generalized linear models a natural way of fitting parametric
-survival models. Moreover, their framework is very flexible: general
-functions of time can be estimated (e.g.~using splines or general
-additive models), and hence these models retain some of the flexibility
-of Cox's partial likelihood approach. And since the unit of analysis is
-a person moment, time-varying covariates can also easily be included in
-this framework.
-
-Logistic regression had already been used in the context of
-discrete-time survival modeling \citep{cox1972regression}. But in the
-context of continuous-time survival modeling, using the framework of
-logistic regression opens the door to an extensive array of powerful
-modeling tools. Indeed, lasso and elastic-net regression can be used to
-select variables that are associated with the hazard function. Logistic
-regression can also be replaced by multinomial regression to analyse
-data with multiple competing events.
-
-In this article, we present the \CRANpkg{casebase} package
-\citep{casebase-package} implemented in R, for parametric survival
-analysis that combines the ideas of Hanley \& Miettinen
-\citeyearpar{hanley2009fitting} into a simple interface. The purpose of
-the \pkg{casebase} package is to provide practitioners with an
-easy-to-use software tool to compute a patient's risk (or cumulative
-incidence) of an event, conditional on a particular patient's covariate
-profile. Our package retains the flexibility of case-base sampling and
-the familiar interface of the \code{glm} function. It also provides
-tools for variable selection, competing-risk analysis, and dealing with
-time-dependent exposures. In addition, we provide functions for
-exploratory data analysis and visualizing the estimated quantities.
+\citep{casebase-package} which implements the Hanley \& Miettinen
+\citeyearpar{hanley2009fitting} approach for fitting fully parametric
+hazard models and covariate adjusted survival curves using the familiar
+interface of the \code{glm} function. Our implementation allows for
+straightforward extensions to other models such as penalized regression
+for variable selection and competing-risk analysis. In addition, we
+provide functions for exploratory data analysis and visualizing the
+estimated quantities such as the hazard function, survival curve and its
+standard errors. The ultimate goal of our package is to make fitting
+flexible hazards accessible to more endusers with the hopes that they
+will favor reporting absolute risks over hazard ratios.
 

### 2021-10-07 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
update competing risk

--- 2021-10-05 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
@@
 
-In a 1994 interview with Professor Nancy Reid, Sir David Cox was asked how he would model a set of censored survival data, to which he responded: "I think I would normally want to tackle problems parametrically ...
-and if you want to do things like predict the outcome for a particular patient, it's much more convenient to do that parametrically" [@reid1994conversation].
-Indeed, the most relevant quantity is often the 5- or 10-year risk of experiencing a certain event given the patient's particular circumstances.
-Unfortunately, the most reported metric from a Cox model is the (potentially time-dependent) hazard ratio (HR), which ignores the duration of follow-up or is subject to selection bias [@hernan2010hazards].
-The covariate adjusted survival curve overcomes these limitations, and is arguably a more important summary measure to report than the HR.
-Stepwise survival curves can be computed with the Cox model but require a second step to separately estimate the baseline hazard [@breslow1972discussion].
-
-In this article, we present the \CRANpkg{casebase} R package [@casebase-package] for fitting fully parametric hazard models and covariate adjusted survival curves using the familiar interface of the \code{glm} function.
-Using a parametric estimator of the hazard function leads to a smooth function of time; as a consequence, the survival function estimates also vary smoothly over time.
-
-Other packages can do this.
-
-the familiar interface of the \code{glm} function.
-It also provides tools for variable selection, competing-risk analysis, and dealing with time-dependent exposures.
-In addition, we provide functions for exploratory data analysis and visualizing the estimated quantities.
-
 Survival analysis and the comparison of event rates has been greatly influenced over the last 50 years by the partial likelihood approach of the Cox proportional hazard model [@cox1972regression].
@@
 The output object from \code{fitSmoothHazard} inherits from the \code{singleEventCB} and \code{glm} classes.
-For this reason, we leveraged the \code{summary} method for \code{glm} objects to output a familiar summary of the results:
-
-```{r, eval = TRUE, echo=eval_cs1}
-summary(fits[["splines"]]) 
+For this reason, we can leverage the \code{summary} method for \code{glm} objects to output a familiar summary of the results:
+
+```{r, eval = eval_cs1, echo = TRUE}
+summary(fit) 
 ```
@@
 
-```{r, eval=FALSE, echo=echo_plot_code}
+```{r, eval=eval_cs1, echo=echo_plot_code}
 new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
-new_time <- seq(0, 14, by = 0.1)
-
-risks <- lapply(fits, function(fit) {
+new_time <- c(0, 1)
+
 absoluteRisk(fit, time = new_time, newdata = new_data)
-})
-```
-
-In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from \pkg{casebase} on the Cox model CIF.
-The CIF estimates for the exponential model in panel (1) overestimate the cumulative incidence earlier on in time, and underestimate it later on.
-Based on our earlier discussion of the population-time plot, this poor fit for the exponential hazard was expected.
-We notice a better fit with increasing complexity of our model for time in Figure \ref{fig:erspc-cox-cif} (panels 2--4).
-As noted in Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models.
-Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) spline model:
+```
+
+In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated risk function from \pkg{casebase} on the Cox model estimate.
+As noted in the Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models.
+Therefore, we can easily test the significance of the spline model:
 
 ```{r erscp-compare, eval = eval_cs1, echo = FALSE}
-anova(fits$exponential, fits$splines, test = "LRT")
-```
-
-The null hypothesis is rejected in favor of the spline model. Similarly, the AIC provides further evidence that the flexible function of time provides the best fit:
-
-```{r erscp-aic, eval = eval_cs1, echo=FALSE}
-c("Exp." = AIC(fits$exponential),
-  "Gompertz" = AIC(fits$gompertz),
-  "Weibull" = AIC(fits$weibull),
-  "Splines" = AIC(fits$splines))
-```
+anova(fit, test = "LRT")
+```
+
+The null hypothesis is rejected in favor of the spline model.
+Similarly, to compare different models (e.g.
+time modeled linearly), we could compute Akaike's Information Criterion (AIC) for each model and compare them.
 
@@
 
-```{r erspc-cox-cif, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="CIFs for control and screening groups in the ERSPC data.
-In each of the panels, we plot the CIF from the Cox model using \\code{survival::survfit} (solid line) and the CIF from the case-base sampling scheme (dashed line) with different functional forms of time.
-(1) The time variable is excluded (exponential).
-(2) Linear function of time (Gompertz).
-(3) The natural logarithm (Weibull).
-(4) Cubic B-spline expansion of time."}
+```{r erspc-cox-cif, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk functions for control and screening groups in the ERSPC data.
+We plot the estimate from the Cox model using \\code{survival::survfit} (solid line) and the estimate from case-base sampling (dashed line).
+We used a cubic B-spline expansion of time."}
 cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
@@
 new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
-new_time <- seq(0,14.5,0.1)
-
-risks <- lapply(fits, function(i) {
-  casebase::absoluteRisk(object = i, time = new_time, newdata = new_data, method = "mont")
-})
-
-layout(matrix(1:4, ncol = 2, byrow = FALSE), respect = FALSE)
-par(mar = c(0, 4.1, 4.1, 0))
+new_time <- seq(0,14,0.1)
+# Recompute with more time points
+risk_mp <- absoluteRisk(fit, time = new_time, newdata = new_data)
+
 plot(survfit(cox_model, newdata = new_data),
-     ylab = "",
-     xlab = "",
+     ylab = "Risk probability (%)",
+     xlab = "Years since randomization",
      fun = "event",
@@
        bg = "gray90")
-lines(risks$exponential[,1], risks$exponential[,2], col = myCols[1], lty = 2, lwd = 2)
-lines(risks$exponential[,1], risks$exponential[,3], col = myCols[2], lty = 2, lwd = 2)
-text(x = max(new_time)*0.97, y = 0.05/100, labels = "(1)", cex = 1.5)
-
-par(mar = c(4.1, 4.1, 0, 0))
-plot(survfit(cox_model, newdata = new_data),
-     ylab = "", xlab = "",
-     fun = "event",
-     xmax = max(new_time),
-     conf.int = FALSE, col = myCols, 
+lines(risk_mp[,1], risk_mp[,2], col = myCols[1], lty = 2, lwd = 2)
+lines(risk_mp[,1], risk_mp[,3], col = myCols[2], lty = 2, lwd = 2)
+```
+
+Moreover, we can also compute confidence intervals for the risk function using the method \code{confint.absRiskCB}.
+
+```{r eval = eval_cs1, echo = -1}
+new_time <- c(0, 1)
+risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
+conf_ints <- confint(risk, fit)
+conf_ints
+```
+
+These point-wise confidence intervals are constructed using parametric bootstrap; the validity of this approach relies on the asymptotic normality of the MLE.
+In Figure \ref{fig:erspc-cif-conf}, we see the 95% confidence bands around the estimates.
+
+```{r erspc-cif-conf, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\\% confidence bands."}
+# confidence interval
+tt <- confint(risk_mp, fit)
+
+# plot absrisk and 95% CI
+new_time <- seq(0,14,0.1)
+i.backw <- order(new_time, decreasing = TRUE)
+i.forw <- order(new_time)
+
+x.poly <- c(new_time[i.forw], new_time[i.backw])
+range_y <- 100 * range(c(tt$conf.high, tt$conf.low)) * c(0.99, 1.01)
+
+par(mfrow = c(1, 2))
+# 1. Control
+tt_sub <- subset(tt, cov_prof == "Control group")
+y.poly <- 100 * c(tt_sub$conf.low[i.forw], 
+                  tt_sub$conf.high[i.backw])
+do.call("plot", 
+    list(
+        x = range(x.poly),
+        y = range_y,
+        type = "n",
+        ylab = "Risk probability (%)",
+        xlab = "Years since randomization",
+        main = "Control group"
+    )
+)
+
+graphics::polygon(x.poly, y.poly, col = "lightgray", border = NA)
+
+# Add KM
+km_fit <- survfit(Surv(Follow.Up.Time, DeadOfPrCa) ~ 1,
+                  data = ERSPC,
+                  subset = ScrArm == "Control group",
+                  conf.type = "log-log")
+
+lines(x = km_fit$time, y = 100 * (1 - km_fit$surv),
+      lwd = 1, lty = 1, col = myCols[2])
+lines(x = km_fit$time, y = 100 * (1 - km_fit$lower),
+      lwd = 2, lty = 3, col = myCols[2])
+lines(x = km_fit$time, y = 100 * (1 - km_fit$upper),
+      lwd = 2, lty = 3, col = myCols[2])
+
+do.call("lines", 
+        list(
+            x = new_time[i.forw],
+            y = 100 * tt_sub$estimate[i.forw],
             lwd = 2,
-     yscale = 100)
-lines(risks$weibull[,1], risks$weibull[,2], col = myCols[1], lty = 2, lwd = 2)
-lines(risks$weibull[,1], risks$weibull[,3], col = myCols[2], lty = 2, lwd = 2)  
-text(x = max(new_time)*0.97, y = 0.05/100, labels = "(3)", cex = 1.5)
-mtext("Cumulative incidence (%)", side = 2, line = -1.5, outer = TRUE)
-
-par(mar = c(0, 0, 4.1, 3.9))
-plot(survfit(cox_model, newdata = new_data),
-     ylab = "",xlab = "",
-     fun = "event",
-      xaxt = 'n', yaxt = 'n',
-     xmax = max(new_time),
-     conf.int = FALSE, col = myCols, 
+            lty = 1,
+            col = myCols[1]
+        )
+)
+
+# Add legend
+legend("topleft", 
+       legend = c("Case-base", "Kaplan-Meier"), 
+       col = myCols,
+       lty = c(1, 1), 
+       lwd = c(2, 1),
+       cex = 1.1,
+       bg = "gray90")
+
+# 2. Screening
+tt_sub <- subset(tt, cov_prof == "Screening group")
+y.poly <- 100 * c(tt_sub$conf.low[i.forw], 
+                  tt_sub$conf.high[i.backw])
+do.call("plot", 
+    list(
+        x = range(x.poly),
+        y = range_y,
+        type = "n",
+        # ylab = "Risk probability (%)",
+        ylab = "",
+        xlab = "Years since randomization",
+        main = "Screening group"
+    )
+)
+
+graphics::polygon(x.poly, y.poly, col = "lightgray", border = NA)
+
+# Add KM
+km_fit <- survfit(Surv(Follow.Up.Time, DeadOfPrCa) ~ 1,
+                  data = ERSPC,
+                  subset = ScrArm == "Screening group",
+                  conf.type = "log-log")
+
+lines(x = km_fit$time, y = 100 * (1 - km_fit$surv),
+      lwd = 1, lty = 1, col = myCols[2])
+lines(x = km_fit$time, y = 100 * (1 - km_fit$lower),
+      lwd = 2, lty = 3, col = myCols[2])
+lines(x = km_fit$time, y = 100 * (1 - km_fit$upper),
+      lwd = 2, lty = 3, col = myCols[2])
+
+do.call("lines", 
+        list(
+            x = new_time[i.forw],
+            y = 100 * tt_sub$estimate[i.forw],
             lwd = 2,
-     yscale = 100)
-lines(risks$gompertz[,1], risks$gompertz[,2], col = myCols[1], lty = 2, lwd = 2)
-lines(risks$gompertz[,1], risks$gompertz[,3], col = myCols[2], lty = 2, lwd = 2)
-text(x = max(new_time)*0.97, y = 0.05/100, labels = "(2)", cex = 1.5)
-
-par(mar = c(4.1, 0, 0, 3.9))
-plot(survfit(cox_model, newdata = new_data),
-     ylab = "",xlab = "",
-     fun = "event",
-     yaxt = 'n',
-     xmax = max(new_time),
-     conf.int = FALSE, col = myCols, 
-     lwd = 2,
-     yscale = 100)
-lines(risks$splines[,1], risks$splines[,2], col = myCols[1], lty = 2, lwd = 2)
-lines(risks$splines[,1], risks$splines[,3], col = myCols[2], lty = 2, lwd = 2)
-text(x = max(new_time)*0.97, y = 0.05/100, labels = "(4)", cex = 1.5)
-mtext("Years since randomization", side = 1, line = -1.5, outer = TRUE)
-```
-
-```{r erspc-95ci22, echo = FALSE, eval = eval_cs1}
-CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
-
-aft_fits <- lapply(list("exp","gompertz","weibull"), function(i)
-  flexsurv::flexsurvreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm,
-                           data = ERSPC, dist = i)
-  )
-
-names(aft_fits) <- list("exponential","gompertz","weibull")
-
-# function to exponentiate HR and CI and return a HR (lower, Upper) character
-exp_est_ci <- function(model = c("exponential","gompertz","weibull","splines"),
-                       fit, 
-                       covariate = "ScrArmScreening group", 
-                       type = c("casebase", "survreg")) {
-
-  type <- match.arg(type)
-  covariate <- match.arg(covariate)
-
-  if (type == "casebase"){
-
-    cis <- confint(fit[[model]])
-
-    sprintf("%0.2f (%0.2f, %0.2f)",
-            exp(coef(fit[[model]]))[[covariate]],
-            exp(cis)[covariate,1],
-            exp(cis)[covariate,2])
-  } else {
-
-    if (model == "weibull") {
-
-      # The maximized log-likelihoods are the same between survreg and flexsurvreg, 
-      # however the parameterisation is different: the first coefficient (Intercept) 
-      # reported by survreg is log(mu), and survreg's "scale" is
-      # dweibull's (thus flexsurvreg)'s 1 / shape
-      res <- -fit[[model]]$res[covariate,] / (1 / fit[[model]]$res["shape",])
-
-      sprintf("%0.2f (%0.2f, %0.2f)",
-              exp(res[["est"]]),
-              exp(res[["U95%"]]),
-              exp(res[["L95%"]]))
-
-    } else if (model == "splines") {
-
-      c("--")
-
-    } else {
-
-      res <- fit[[model]]$res[covariate,]
-
-      sprintf("%0.2f (%0.2f, %0.2f)",
-              exp(res[["est"]]),
-              exp(res[["L95%"]]),
-              exp(res[["U95%"]]))
-    }
-  }
-}
-
-casebase_HR_CI <- lapply(list("exponential","gompertz","weibull","splines"), 
-                         exp_est_ci, 
-                         fit = fits,
-                         covariate = "ScrArmScreening group", 
-                         type = "casebase")
-
-survreg_HR_CI <- lapply(list("exponential","gompertz","weibull","splines"), 
-                         exp_est_ci, 
-                         fit = aft_fits,
-                         covariate = "ScrArmScreening group", 
-                         type = "survreg")
-
-ptable <- cbind(list("Exponential","Gompertz","Weibull","Splines"),
-            do.call(rbind, casebase_HR_CI), do.call(rbind, survreg_HR_CI))
-```
-
-In Table \ref{tab:print-erspc-estimates}, we present a side-by-side comparison of the hazard ratios and confidence intervals estimated from \code{fitSmoothHazard} and the corresponding parametric model using \code{survival::survreg}, as well as the Cox model estimate.
-The hazard ratio estimates and confidence intervals are similar across all four models.
-This reinforces the idea that, under proportional hazards, we do not need to model the full hazard to obtain reliable estimates of the hazard ratio.
-Nevertheless, Figure \ref{fig:erspc-cox-cif} shows that different parametric models can still give rise to qualitatively different estimates for the CIF.
-
-```{r print-erspc-estimates, eval = eval_cs1}
-kable(ptable, "latex", booktabs = TRUE, align = c("l","c","c"),
-      caption = c("Comparison of estimated hazard ratios and 95\\% confidence intervals for ERSPC data."),
-      col.names = c("Model","casebase::fitSmoothHazard","survival::survreg")) %>%
-  kable_styling() %>%
-  column_spec(1, bold = TRUE) %>%
-  collapse_rows(columns = 1, latex_hline = "major", valign = "middle") %>%
-  footnote(general = sprintf("Cox model estimate: HR (95%% CI) = %0.2f (%0.2f, %0.2f)", CI_cox[1], CI_cox[2], CI_cox[3]),
-           general_title = "") 
+            lty = 1,
+            col = myCols[1]
+        )
+)
 ```
@@
 
-We want to compare our hazard ratio estimates to that obtained from a Cox regression. To obtain these estimates, we need to treat the competing event as censoring:
+We want to compare our hazard ratio estimates to that obtained from a Cox regression.
 
@@
 library(survival)
-# Treat competing event as censoring
-model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-                   data = bmtcrr)
+# Prepare data for coxph
+bmtcrr_cox <- transform(bmtcrr, 
+                        id = seq_len(nrow(bmtcrr)),
+                        Status = factor(Status))
+
+model_cox <- coxph(Surv(ftime, Status) ~ Sex + D + Phase + Source + Age,
+                   data = bmtcrr_cox, id = id)
 ```
@@
 
-table_cox <- round(summary(model_cox)$conf.int[, -2], 2)
+# Identify relevant coefficients
+coef_relev <- grepl("_1:2$", names(coef(model_cox)))
+table_cox <- round(summary(model_cox)$conf.int[coef_relev, -2], 2)
+rownames(table_cox) <- gsub("_1:2$", "", rownames(table_cox))
 rownames(table_cb) <- rownames(table_cox)
@@
 
-Given the estimate of the hazard function obtained using case-base sampling, we can compute the absolute risk curve for a fixed covariate profile.
+Given the estimate of the hazard functions obtained using case-base sampling, we can compute the cumulative incidence curve for a fixed covariate profile.
 We perform this computation for a 35 year old woman who received a stem-cell transplant from peripheral blood at relapse.
@@
 
+# Estimate CI curve
+risk_fg <- predict(model_fg, newdata, times = time_points)
+```
+
+We can also estimate the CI for relapse using the Cox model and the Aalen-Johansen estimator:
+
+```{r cox_risk, eval = eval_cs2, echo = TRUE}
 # Estimate absolute risk curve
-risk_fg <- predict(model_fg, newdata, times = time_points)
-```
-
-```{r bmtcrr-risk, echo = FALSE, fig.cap="\\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups.
-The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.", eval = eval_cs2}
+risk_cox <- survfit(model_cox, newdata = newdata)
+```
+
+```{r bmtcrr-risk, echo = FALSE, fig.cap="\\label{fig:compAbsrisk} Cumulative Incidence curve for a fixed covariate profile and the two disease groups.
+The estimate obtained from case-base sampling is compared to the Fine-Gray and Aalen-Johansen estimates.", eval = eval_cs2}
 risk_all <- dplyr::bind_rows(
@@
   data.frame(
+    Time = risk_cox$time,
+    Method = "Cox",
+    Risk = risk_cox[,2]$pstate[,1,1],
+    Disease = "ALL",
+    stringsAsFactors = FALSE
+  ),
+  data.frame(
+    Time = risk_cox$time,
+    Method = "Cox",
+    Risk = risk_cox[,2]$pstate[,2,1],
+    Disease = "AML",
+    stringsAsFactors = FALSE
+  ),
+  data.frame(
     Time = time_points,
@@
   )
-)
+) %>% 
+  dplyr::filter(Time <= 60)
 
@@
 
-Figure \ref{fig:compAbsrisk} shows the absolute risk curves for both case-base sampling and the Fine-Gray model.
-As we can see, the two approaches agree quite well for AML; however, there seems to be a difference of about 5% between the two curves for ALL.
+Figure \ref{fig:compAbsrisk} shows the cumulative incidence curves for all three models.
+As we can see, all three approaches agree quite well for AML; however, for ALL, there seems to be a difference of about 5% between the Fine-Gray curve and the curves estimated using case-base sampling and Cox regression.
 This difference does not appear to be significant: the curve from case-base sampling is contained within a 95% confidence band around the Fine-Gray absolute risk curve (figure not shown).
@@
 
-In this article, we presented the R package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating CIFs using case-base sampling.
+In this article, we presented the R package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating risk functions using case-base sampling.
 Our package also provides several functions to produce graphical summaries of the data and the results.
-We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through three case studies.
+We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the case-base framework and the package through three case studies.
 
@@
 However, in terms of output, both approaches differ significantly.
-Case-base sampling produces smooth hazards and smooth cumulative incidence curves, whereas Cox regression and Fine-Gray models produce stepwise CIFs and never explicitly model the hazard function.
+Case-base sampling produces smooth hazards and smooth survival curves, whereas Cox regression and Fine-Gray models produce stepwise CIFs and never explicitly model the hazard function.
 Qualitatively, we showed that by using splines in the linear predictor, all three models yielded similar curves.
 However, the smooth nature of the output of \pkg{casebase} provides a more intuitive interpretation for consumers of these predictions.
-In Table \ref{tab:compCBvsCox}, we provide a side-by-side comparison between the Cox model and case-base sampling.
-
-\begin{widetable}
-\caption{\label{tab:compCBvsCox}Comparison between the Cox model and case-base sampling}
-\centering
-\begin{tabular}[t]{llp{5cm}}
-\toprule
-Feature & Cox model & Case-base sampling\\
-\toprule
-Model type & Semi-parametric & Fully parametric\\\midrule
-Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\\midrule
-Cumulative incidence & Step function & Smooth-in-time curve\\\midrule
-Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\\midrule
-Model testing & Likelihood-ratio tests, score tests, & Use GLM framework\\
- & information criteria & (e.g.\ LRT, AIC, BIC)\\\midrule
-\addlinespace
-Competing risks & Difficult & Cause-specific CIFs\\
-% Prediction & Kaplan-Meier-based & ROC, AUC, risk reclassification probabilities\\
-\bottomrule
-\end{tabular}
-\end{widetable}
+
+<!-- In Table \ref{tab:compCBvsCox}, we provide a side-by-side comparison between the Cox model and case-base sampling. -->
+
+<!-- \begin{widetable} -->
+<!-- \caption{\label{tab:compCBvsCox}Comparison between the Cox model and case-base sampling} -->
+<!-- \centering -->
+<!-- \begin{tabular}[t]{llp{5cm}} -->
+<!-- \toprule -->
+<!-- Feature & Cox model & Case-base sampling\\ -->
+<!-- \toprule -->
+<!-- Model type & Semi-parametric & Fully parametric\\\midrule -->
+<!-- Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\\midrule -->
+<!-- Survival curve & Step function & Smooth-in-time curve\\\midrule -->
+<!-- Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\\midrule -->
+<!-- Model testing & Likelihood-ratio tests, score tests, & Use GLM framework\\ -->
+<!--  & information criteria & (e.g.\ LRT, AIC, BIC)\\\midrule -->
+<!-- \addlinespace -->
+<!-- Competing risks & Aalen-Johansen & Cause-specific CIFs\\ -->
+<!-- % Prediction & Kaplan-Meier-based & ROC, AUC, risk reclassification probabilities\\ -->
+<!-- \bottomrule -->
+<!-- \end{tabular} -->
+<!-- \end{widetable} -->
 

### 2021-10-07 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
update competing risk

--- 2021-10-02 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
@@
 We want to compare our hazard ratio estimates to that obtained from a
-Cox regression. To obtain these estimates, we need to treat the
-competing event as censoring:
+Cox regression.
 
@@
 library(survival)
-# Treat competing event as censoring
-model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-                   data = bmtcrr)
+# Prepare data for coxph
+bmtcrr_cox <- transform(bmtcrr, 
+                        id = seq_len(nrow(bmtcrr)),
+                        Status = factor(Status))
+
+model_cox <- coxph(Surv(ftime, Status) ~ Sex + D + Phase + Source + Age,
+                   data = bmtcrr_cox, id = id)
 \end{Sinput}
@@
 \midrule
-Sex & 0.74 & (0.43, 1.3) & 0.68 & (0.39, 1.19)\\
-Disease & 0.52 & (0.29, 0.94) & 0.52 & (0.29, 0.93)\\
-Phase (CR2 vs. CR1) & 1.20 & (0.48, 3) & 1.21 & (0.48, 3.02)\\
-Phase (CR3 vs. CR1) & 1.59 & (0.41, 6.17) & 1.67 & (0.43, 6.5)\\
-Phase (Relapse vs. CR1) & 4.09 & (1.9, 8.81) & 4.55 & (2.09, 9.9)\\
+Sex & 0.74 & (0.43, 1.3) & 0.68 & (0.39, 1.21)\\
+Disease & 0.52 & (0.29, 0.94) & 0.52 & (0.28, 0.94)\\
+Phase (CR2 vs. CR1) & 1.20 & (0.48, 3) & 1.21 & (0.47, 3.09)\\
+Phase (CR3 vs. CR1) & 1.59 & (0.41, 6.17) & 1.67 & (0.46, 6.08)\\
+Phase (Relapse vs. CR1) & 4.09 & (1.9, 8.81) & 4.55 & (1.98, 10.46)\\
 \addlinespace
-Source & 1.70 & (0.55, 5.23) & 1.46 & (0.47, 4.54)\\
+Source & 1.70 & (0.55, 5.23) & 1.46 & (0.47, 4.53)\\
 Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
@@
 
-Given the estimate of the hazard function obtained using case-base
-sampling, we can compute the absolute risk curve for a fixed covariate
-profile. We perform this computation for a 35 year old woman who
-received a stem-cell transplant from peripheral blood at relapse. We
+Given the estimate of the hazard functions obtained using case-base
+sampling, we can compute the cumulative incidence curve for a fixed
+covariate profile. We perform this computation for a 35 year old woman
+who received a stem-cell transplant from peripheral blood at relapse. We
 compare the absolute risk curve for such a woman with ALL with that for
@@
 
+# Estimate CI curve
+risk_fg <- predict(model_fg, newdata, times = time_points)
+\end{Sinput}
+\end{Schunk}
+
+We can also estimate the CI for relapse using the Cox model and the
+Aalen-Johansen estimator:
+
+\begin{Schunk}
+\begin{Sinput}
 # Estimate absolute risk curve
-risk_fg <- predict(model_fg, newdata, times = time_points)
+risk_cox <- survfit(model_cox, newdata = newdata)
 \end{Sinput}
@@
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./bmtcrr-risk-1} \caption{\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups.
-The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:bmtcrr-risk}
+\includegraphics[width=\textwidth,keepaspectratio=true]{./bmtcrr-risk-1} \caption{\label{fig:compAbsrisk} Cumulative Incidence curve for a fixed covariate profile and the two disease groups.
+The estimate obtained from case-base sampling is compared to the Fine-Gray and Aalen-Johansen estimates.}\label{fig:bmtcrr-risk}
 \end{figure}
@@
 
-Figure \ref{fig:compAbsrisk} shows the absolute risk curves for both
-case-base sampling and the Fine-Gray model. As we can see, the two
-approaches agree quite well for AML; however, there seems to be a
-difference of about 5\% between the two curves for ALL. This difference
-does not appear to be significant: the curve from case-base sampling is
-contained within a 95\% confidence band around the Fine-Gray absolute
-risk curve (figure not shown).
+Figure \ref{fig:compAbsrisk} shows the cumulative incidence curves for
+all three models. As we can see, all three approaches agree quite well
+for AML; however, for ALL, there seems to be a difference of about 5\%
+between the Fine-Gray curve and the curves estimated using case-base
+sampling and Cox regression. This difference does not appear to be
+significant: the curve from case-base sampling is contained within a
+95\% confidence band around the Fine-Gray absolute risk curve (figure
+not shown).
 
@@
 provided details about our implementation, and we illustrated the merits
-of the approach and the package through three case studies.
+of the case-base framework and the package through three case studies.
 
@@
 the output of \pkg{casebase} provides a more intuitive interpretation
-for consumers of these predictions. In Table \ref{tab:compCBvsCox}, we
-provide a side-by-side comparison between the Cox model and case-base
-sampling.
-
-\begin{widetable}
-\caption{\label{tab:compCBvsCox}Comparison between the Cox model and case-base sampling}
-\centering
-\begin{tabular}[t]{llp{5cm}}
-\toprule
-Feature & Cox model & Case-base sampling\\
-\toprule
-Model type & Semi-parametric & Fully parametric\\\midrule
-Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\\midrule
-Survival curve & Step function & Smooth-in-time curve\\\midrule
-Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\\midrule
-Model testing & Likelihood-ratio tests, score tests, & Use GLM framework\\
- & information criteria & (e.g.\ LRT, AIC, BIC)\\\midrule
-\addlinespace
-Competing risks & Difficult & Cause-specific CIFs\\
-% Prediction & Kaplan-Meier-based & ROC, AUC, risk reclassification probabilities\\
-\bottomrule
-\end{tabular}
-\end{widetable}
+for consumers of these predictions.
 

### 2021-10-07 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.R

@@ -341,9 +341,13 @@ model_cb <- fitSmoothHazard(
 
 ## ----bmtcrr-cox, echo = TRUE, eval = eval_cs2---------------------------------
 library(survival)
-# Treat competing event as censoring
-model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
-                   data = bmtcrr)
+# Prepare data for coxph
+bmtcrr_cox <- transform(bmtcrr, 
+                        id = seq_len(nrow(bmtcrr)),
+                        Status = factor(Status))
+
+model_cox <- coxph(Surv(ftime, Status) ~ Sex + D + Phase + Source + Age,
+                   data = bmtcrr_cox, id = id)
 
 
 ## ----bmtcrr-cis, echo=FALSE, eval = eval_cs2----------------------------------
@@ -364,7 +368,10 @@ table_cb <- cbind(
 )
 table_cb <- round(exp(table_cb), 2)
 
-table_cox <- round(summary(model_cox)$conf.int[, -2], 2)
+# Identify relevant coefficients
+coef_relev <- grepl("_1:2$", names(coef(model_cox)))
+table_cox <- round(summary(model_cox)$conf.int[coef_relev, -2], 2)
+rownames(table_cox) <- gsub("_1:2$", "", rownames(table_cox))
 rownames(table_cb) <- rownames(table_cox)
 colnames(table_cb) <- colnames(table_cox)
 
@@ -432,11 +439,16 @@ model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) +
                         const(Phase) + const(Source) + const(Age),
                       data = bmtcrr, cause = 1, model = "fg")
 
-# Estimate absolute risk curve
+# Estimate CI curve
 risk_fg <- predict(model_fg, newdata, times = time_points)
 
 
-## ----bmtcrr-risk, echo = FALSE, fig.cap="\\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.", eval = eval_cs2----
+## ----cox_risk, eval = eval_cs2, echo = TRUE-----------------------------------
+# Estimate absolute risk curve
+risk_cox <- survfit(model_cox, newdata = newdata)
+
+
[…]

### 2021-10-07 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
add final clarifications

@@
 As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
-key idea behind case-base sampling is to sample from the study base a
-finite amount of \emph{person moments}. These person moments are indexed
-by both an individual in the study and a time point, and therefore each
-person moment has a covariate profile, an exposure status and an outcome
-status attached to it. We note that there is only a finite number of
-person moments associated with the event of interest (what Hanley \&
-Miettinen call the \emph{case series}). Case-base sampling refers to the
-sampling from the base of a representative finite sample called the
-\emph{base series}.
+key idea behind case-base sampling is to consider the entire study base
+as an infinite collection of \emph{person moments}. These person moments
+are indexed by both an individual in the study and a time point, and
+therefore each person moment has a covariate profile, an exposure
+status, and an outcome status attached to it. By comparing person
+moments at which an event occurred with person moments at which no event
+occurred, we can extract information about hazards and survival.
+
+Therefore, we start by sampling all person moments at which the event
+occurred; this collection of person moments is what Hanley \& Miettinen
+call the \emph{case series}. The incidence of the case series is
+dictated by the hazard function of interest. Next, we sample a finite
+number of person moments at which no event occurred; this second
+collection of person moment is what Hanley \& Miettinen call the
+\emph{base series}. The sampling mechanism for the base series is left
+at the discretion of the user, but in practice we find that sampling
+uniformly from the study base provides both simplicity and good
+performance. This is the default sampling mechanism in the package.
 
@@
 the processes \(N_i(t)\), and which satisfy
-\[\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].\]
-
-Next, we model the base series sampling mechanism using non-homogeneous
-Poisson processes \(R_i(t) \in \{0, 1, 2, \ldots\}\), with the
-person-moments where \(dR_i(t) = 1\) constituting the base series. The
+\[\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].\] The processes
+\(N_i(t)\) count the person moments from the \emph{case series}.
+
+To complement the case series, we sample person moments for the base
+series. To do so, we model the base series sampling mechanism using
+non-homogeneous Poisson processes \(R_i(t) \in \{0, 1, 2, \ldots\}\);
+the person-moments where \(dR_i(t) = 1\) constitute the base series. The
 process \(Q_{i}(t) = R_i(t) + N_{i}(t)\) then counts both the case and
-base series person-moments contributed by individual \(i\). This process
-is typically defined by the user via its intensity function
-\(\rho_i(t)\). The process \(Q_{i}(t)\) is characterized by
+base series person-moments contributed by individual \(i\). The
+processes \(R_i(t)\) are typically defined by the user via its intensity
+function \(\rho_i(t)\). The process \(Q_{i}(t)\) is characterized by
 \(E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt\).
@@
 analysis; for a different perspective on this topic, see Mahani \&
-Sharabiani \citeyearpar{mahani2015bayesian}. The package
-\CRANpkg{cmprsk} provides methods for cause-specific subdistribution
-hazards, such as in the Fine-Gray model
+Sharabiani \citeyearpar{mahani2015bayesian}. The package \pkg{survival}
+provides functionality for competing-risk analysis and multistate
+modelling. The package \CRANpkg{cmprsk} provides methods for
+cause-specific subdistribution hazards, such as in the Fine-Gray model
 \citeyearpar{fine1999proportional}. On the other hand, the package
@@
 
-Finally, several packages include functions to estimate the CIF. The
-corresponding methods generally fall into two categories: transformation
-of the estimated hazard function, and semi-parametric estimation of the
-baseline hazard. The first category broadly corresponds to parametric
-survival models, where the full hazard is explicitly modeled. Using this
-estimate, the survival function and the CIF can be obtained using their
-functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CDF}
-below). Packages providing this functionality include \pkg{CFC},
-\pkg{flexsurv}, \pkg{mets}, and \pkg{survival}. Our package
-\pkg{casebase} also follows this approach for both single-event and
-competing-risk analyses. The second category outlined above broadly
-corresponds to semi-parametric models. These models do not model the
-full hazard function, and therefore the baseline hazard needs to be
-estimated separately in order to estimate the survival function. This is
-achieved using semi-parametric estimators (e.g.~Breslow's estimator) or
-parametric estimators (e.g.~spline functions). Packages that implement
-this approach include \pkg{riskRegression}, \pkg{rstpm2}, and
-\pkg{survival}. As mentioned in the introduction, a key distinguishing
-factor between these two approaches is that the first category leads to
-smooth estimates of the survival function, whereas the second category
-often produces estimates in the form of stepwise functions.
+Finally, several packages include functions to estimate the survival
+function and the CIF. The corresponding methods generally fall into two
+categories: transformation of the estimated hazard function, and
+semi-parametric estimation of the baseline hazard. The first category
+broadly corresponds to parametric survival models, where the full hazard
+is explicitly modeled. Using this estimate, the survival function and
+the CIF can be obtained using their functional relationships (see
+Equations \ref{eqn:surv} and \ref{eqn:CDF} below). Packages providing
+this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and
+\pkg{survival}. Our package \pkg{casebase} also follows this approach
+for both single-event and competing-risk analyses. The second category
+outlined above broadly corresponds to semi-parametric models. These
+models do not model the full hazard function, and therefore the baseline
+hazard needs to be estimated separately in order to estimate the
+survival function. This is achieved using semi-parametric estimators
+(e.g.~Breslow's estimator) or parametric estimators (e.g.~spline
+functions). Packages that implement this approach include
+\pkg{riskRegression}, \pkg{rstpm2}, \pkg{survival}, and \pkg{glmnet}. As
+mentioned in the introduction, a key distinguishing factor between these
+two approaches is that the first category leads to smooth estimates of
+the survival function, whereas the second category often produces
+estimates in the form of stepwise functions.
 
@@
 \textbf{flexsurv}        & \checkmark                        & \checkmark                         &                       & \checkmark                & \checkmark                   &                          &                                  & \checkmark               \\ \hline
-\textbf{glmnet}          &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
+\textbf{glmnet}          &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &  \checkmark              \\ \hline
 \textbf{glmpath}         &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
@@
 for variable selection through regularized estimation of the hazard
-function as given by \eqref{eq:penest}. To illustrate this
-functionality, we use the dataset from the Study to Understand Prognoses
-Preferences Outcomes and Risks of Treatment (SUPPORT)
+function as given by \eqref{eq:penest}. We note that this is different
+than the semiparametric model Coxnet, which regularizes the Cox partial
+likelihood. To illustrate this functionality, we use the dataset from
+the Study to Understand Prognoses Preferences Outcomes and Risks of
+Treatment (SUPPORT)
 \citep{knaus1995support}.\footnote{The original data is available online from the Department of Biostatistics at Vanderbilt University: \url{https://biostat.app.vumc.org/wiki/Main/SupportDesc}}
@@
 hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface
-to the \code{fitSmoothHazard} function. We supply both a matrix \code{y}
-containing the time and event variables, and a matrix \code{x}
-containing all other covariates. We apply the lasso penalty by setting
-\code{alpha = 1} and assign a \code{penalty.factor} (\(w_j\);
-cf.~Equation \ref{eq:penest}) of 0 to the time variable to ensure it is
-in the selected model. We compare our approach to both Cox regression,
-and lasso penalized Cox regression (fitted via the \pkg{glmnet}
-package).
+to the \code{fitSmoothHazard} function. The \code{fitSmoothHazard} and
+\code{fitSmoothHazard.fit} functions sample the case and base series,
+calculate the required offset, and transform the data to match the
+expected input of the \pkg{glmnet} package. The penalized logistic
+regression is then fit for multiple values of the tuning parameter using
+the function \code{glmnet::cv.glmnet} and the binomial family. We supply
+both a matrix \code{y} containing the time and event variables, and a
+matrix \code{x} containing all other covariates. We apply the lasso
+penalty by setting \code{alpha = 1} and assign a \code{penalty.factor}
+(\(w_j\); cf.~Equation \ref{eq:penest}) of 0 to the time variable to
+ensure it is in the selected model. We compare our approach to both Cox
+regression, and lasso penalized Cox regression (fitted via the
+\pkg{glmnet} package and using the Cox family).
 

### 2021-10-07 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd

@@
 
-As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to sample from the study base a finite amount of *person moments*.
-These person moments are indexed by both an individual in the study and a time point, and therefore each person moment has a covariate profile, an exposure status and an outcome status attached to it.
-We note that there is only a finite number of person moments associated with the event of interest (what Hanley & Miettinen call the *case series*).
-Case-base sampling refers to the sampling from the base of a representative finite sample called the *base series*.
+As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to consider the entire study base as an infinite collection of *person moments*.
+These person moments are indexed by both an individual in the study and a time point, and therefore each person moment has a covariate profile, an exposure status, and an outcome status attached to it.
+By comparing person moments at which an event occurred with person moments at which no event occurred, we can extract information about hazards and survival.
+
+Therefore, we start by sampling all person moments at which the event occurred; this collection of person moments is what Hanley & Miettinen call the *case series*.
+The incidence of the case series is dictated by the hazard function of interest.
+Next, we sample a finite number of person moments at which no event occurred; this second collection of person moment is what Hanley & Miettinen call the *base series*.
+The sampling mechanism for the base series is left at the discretion of the user, but in practice we find that sampling uniformly from the study base provides both simplicity and good performance.
+This is the default sampling mechanism in the package.
 
@@
 We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
-$$\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].$$
-
-Next, we model the base series sampling mechanism using non-homogeneous Poisson processes $R_i(t) \in \{0, 1, 2, \ldots\}$, with the person-moments where $dR_i(t) = 1$ constituting the base series.
+$$\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].$$  The processes $N_i(t)$ count the person moments from the *case series*.
+
+To complement the case series, we sample person moments for the base series.
+To do so, we model the base series sampling mechanism using non-homogeneous Poisson processes $R_i(t) \in \{0, 1, 2, \ldots\}$; the person-moments where $dR_i(t) = 1$ constitute the base series.
 The process $Q_{i}(t) = R_i(t) + N_{i}(t)$ then counts both the case and base series person-moments contributed by individual $i$.
-This process is typically defined by the user via its intensity function $\rho_i(t)$.
+The processes $R_i(t)$ are typically defined by the user via its intensity function $\rho_i(t)$.
 The process $Q_{i}(t)$ is characterized by $E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt$.
@@
 Next, several R packages implement methodologies for competing risk analysis; for a different perspective on this topic, see Mahani & Sharabiani [-@mahani2015bayesian].
+The package \pkg{survival} provides functionality for competing-risk analysis and multistate modelling.
 The package \CRANpkg{cmprsk} provides methods for cause-specific subdistribution hazards, such as in the Fine-Gray model [-@fine1999proportional].
@@
 
-Finally, several packages include functions to estimate the CIF.
+Finally, several packages include functions to estimate the survival function and the CIF.
 The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard.
@@
 spline functions).
-Packages that implement this approach include \pkg{riskRegression}, \pkg{rstpm2}, and \pkg{survival}.
+Packages that implement this approach include \pkg{riskRegression}, \pkg{rstpm2}, \pkg{survival}, and \pkg{glmnet}.
 As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the survival function, whereas the second category often produces estimates in the form of stepwise functions.
@@
 For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function as given by \eqref{eq:penest}.
+We note that this is different than the semiparametric model Coxnet, which regularizes the Cox partial likelihood.
 To illustrate this functionality, we use the dataset from the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) [@knaus1995support].\footnote{The original data is available online from the Department of Biostatistics at Vanderbilt University: \url{https://biostat.app.vumc.org/wiki/Main/SupportDesc}} The SUPPORT dataset tracks death in five American hospitals within individuals who are considered seriously ill.
@@
 For fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface to the \code{fitSmoothHazard} function.
+The \code{fitSmoothHazard} and \code{fitSmoothHazard.fit} functions sample the case and base series, calculate the required offset, and transform the data to match the expected input of the \pkg{glmnet} package.
+The penalized logistic regression is then fit for multiple values of the tuning parameter using the function \code{glmnet::cv.glmnet} and the binomial family.
 We supply both a matrix \code{y} containing the time and event variables, and a matrix \code{x} containing all other covariates.
@@
 Equation \ref{eq:penest}) of 0 to the time variable to ensure it is in the selected model.
-We compare our approach to both Cox regression, and lasso penalized Cox regression (fitted via the \pkg{glmnet} package).
+We compare our approach to both Cox regression, and lasso penalized Cox regression (fitted via the \pkg{glmnet} package and using the Cox family).
 

### 2021-10-09 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd

@@
 
-Survival analysis and the comparison of event rates has been greatly influenced over the last 50 years by the partial likelihood approach of the Cox proportional hazard model [@cox1972regression].
-This approach provides a flexible way of assessing the influence of covariates on the hazard function, without the need to specify a parametric survival model.
-This flexibility comes at the cost of decoupling the baseline hazard from the effect of the covariates.
-To recover the whole survival curve---or the cumulative incidence function (CIF)---we then need to separately estimate the baseline hazard [@breslow1972discussion].
-This in turn leads to stepwise estimates of the survival function that can be difficult to interpret.
-
-From the perspective of clinicians and their patients, the most relevant quantity is often the 5- or 10-year risk of experiencing a certain event given the patient's particular circumstances, and not the hazard ratio between a treatment and control group.
-Therefore, to make sound clinical decisions, it is important to accurately estimate the *full* hazard function, which can subsequently be used to estimate the CIF.
-Using a parametric estimator of the hazard function leads to a smooth function of time; as a consequence, the CIF and the survival function estimates also vary smoothly over time.
-
-With the goal of fitting smooth-in-time hazard functions, Hanley & Miettinen [-@hanley2009fitting] proposed a general framework for estimating fully parametric hazard models via logistic regression.
-Their main idea is simple: comparing person-moments when the event of interest occurred with moments when patients were at risk.
-Their approach handles censored data effortlessly and provides users familiar with generalized linear models a natural way of fitting parametric survival models.
-Moreover, their framework is very flexible: general functions of time can be estimated (e.g.
-using splines or general additive models), and hence these models retain some of the flexibility of Cox's partial likelihood approach.
-And since the unit of analysis is a person moment, time-varying covariates can also easily be included in this framework.
-
-Logistic regression had already been used in the context of discrete-time survival modeling [@cox1972regression].
-But in the context of continuous-time survival modeling, using the framework of logistic regression opens the door to an extensive array of powerful modeling tools.
-Indeed, lasso and elastic-net regression can be used to select variables that are associated with the hazard function.
-Logistic regression can also be replaced by multinomial regression to analyse data with multiple competing events.
-
-In this article, we present the \CRANpkg{casebase} package [@casebase-package] implemented in R, for parametric survival analysis that combines the ideas of Hanley & Miettinen [-@hanley2009fitting] into a simple interface.
-The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to compute a patient's risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile.
-Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function.
-It also provides tools for variable selection, competing-risk analysis, and dealing with time-dependent exposures.
-In addition, we provide functions for exploratory data analysis and visualizing the estimated quantities.
-
-In what follows, we first recall some theoretical details on case-base sampling and its use for estimating parametric hazard functions.
-We then give a short review of existing R packages that implement comparable features as \pkg{casebase}.
-Next, we provide some details about the implementation of case-base sampling in our package, and we give a brief survey of its main functions.
-This is followed by three case studies that illustrate the flexibility and capabilities of \pkg{casebase}.
-We show how the same framework can be used for non-linear functions of time and non-proportional hazards, competing risks and variable selection via penalized regression.
-Finally, we end the article with a discussion of the results and of future directions.
+The semiparametric Cox model has become the default approach to survival analysis 
+even though Cox himself later suggested he would prefer to model the hazard function 
+directly. In a 1994 interview with Professor Nancy Reid, Sir David Cox was asked 
+how he would model a set of censored survival data, to which he responded: "I 
+think I would normally want to tackle problems parametrically ... and if you want 
+to do things like predict the outcome for a particular patient, it's much more 
+convenient to do that parametrically" [@reid1994conversation]. Indeed, the most 
+relevant quantity in a clinical setting is often the 5- or 10-year risk of 
+experiencing a certain event given the patient's particular circumstances. 
+Unfortunately, the most reported metric from a Cox model is the (potentially 
+time-dependent) hazard ratio (HR), which ignores the duration of follow-up or 
+is subject to selection bias [@hernan2010hazards]. The covariate adjusted 
+survival curve overcomes these limitations, and is arguably a more important 
+summary measure to report than the HR. While stepwise survival curves can be 
+computed with the Cox model, they require a second step to separately estimate 
+the baseline hazard [@breslow1972discussion].
+
+Several authors have since pursued fully parametric approaches that made the 
+fitting of smooth hazard functions much more transparent and intuitive through 
+generalized linear models. The key feature of these procedures is splitting the 
+time axis into discrete intervals. [@whitehead1980fitting] showed the 
+equivalence between a Cox model and a Poisson regression with a parameter for 
+each event time. [@copenhagen2012needs] provides a nice exposition of this 
+equivalence with a real data example and supporting R code for computing 
+standard errors. [@arjas1987logistic] and [@efron1988logistic] treated each 
+patient-day as a Bernoulli random variable with probability equal to the 
+discrete hazard rate. A potential issue with these approaches is that the 
+number of time bins need to be chosen by the data analyst. A fine grouping 
+of times may result in few (or none) events per interval, which then leads 
+to instability in the Newton-Raphson procedure for estimation 
+[@kalbfleisch2011statistical]. A coarse grouping could potentially mask 
+nonlinear trends in the hazard function.
+
+Rather than discretizing time, Hanley & Miettinen [-@hanley2009fitting] 
+selected a discrete set of person-time coordinates (`person-moments') in 
+continuous time from all observed follow-up experience constituting the 
+study base to obtain the logistic form likelihood expression for the hazard 
+function. More specifically, all person-moments when the event of interest 
+occurred are selected as the case series, complemented by a randomly sampled 
+base series of person-moments serving as controls. This approach allows 
+flexible modeling of the hazard function by including time as a covariate 
+(e.g. using splines or general additive models). Furthermore, time-dependent 
+covariates can be modeled through interactions with time. In short, Hanley & 
+Miettinen [-@hanley2009fitting] use the well-understood logistic regression 
+for directly modeling the hazard function, but do not require an arbitrary 
+choice of splitting time.
+
+In this article, we present the \CRANpkg{casebase} R package [@casebase-package] 
+which implements the Hanley & Miettinen [-@hanley2009fitting] approach for 
+fitting fully parametric hazard models and covariate adjusted survival curves 
+using the familiar interface of the \code{glm} function. Our implementation 
+allows for straightforward extensions to other models such as penalized 
+regression for variable selection and competing-risk analysis. In addition, 
+we provide functions for exploratory data analysis and visualizing the estimated 
+quantities such as the hazard function, survival curve and its standard errors. 
+The ultimate goal of our package is to make fitting flexible hazards accessible 
+to more endusers with the hopes that they will favor reporting absolute risks 
+over hazard ratios.
+
+In what follows, we first recall some theoretical details on case-base sampling 
+and its use for estimating parametric hazard functions. We then give a short 
+review of existing R packages that implement comparable features as \pkg{casebase}. 
+Next, we provide some details about the implementation of case-base sampling in 
+our package, and we give a brief survey of its main functions. This is followed 
+by three case studies that illustrate the flexibility and capabilities of 
+\pkg{casebase}. We show how the same framework can be used for non-linear 
+functions of time and non-proportional hazards, competing risks and variable 
+selection via penalized regression. Finally, we end the article with a 
+discussion of the results and of future directions.
 

### 2021-10-09 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex

@@
 
-Survival analysis and the comparison of event rates has been greatly
-influenced over the last 50 years by the partial likelihood approach of
-the Cox proportional hazard model \citep{cox1972regression}. This
-approach provides a flexible way of assessing the influence of
-covariates on the hazard function, without the need to specify a
-parametric survival model. This flexibility comes at the cost of
-decoupling the baseline hazard from the effect of the covariates. To
-recover the whole survival curve---or the cumulative incidence function
-(CIF)---we then need to separately estimate the baseline hazard
-\citep{breslow1972discussion}. This in turn leads to stepwise estimates
-of the survival function that can be difficult to interpret.
-
-From the perspective of clinicians and their patients, the most relevant
-quantity is often the 5- or 10-year risk of experiencing a certain event
-given the patient's particular circumstances, and not the hazard ratio
-between a treatment and control group. Therefore, to make sound clinical
-decisions, it is important to accurately estimate the \emph{full} hazard
-function, which can subsequently be used to estimate the CIF. Using a
-parametric estimator of the hazard function leads to a smooth function
-of time; as a consequence, the CIF and the survival function estimates
-also vary smoothly over time.
-
-With the goal of fitting smooth-in-time hazard functions, Hanley \&
-Miettinen \citeyearpar{hanley2009fitting} proposed a general framework
-for estimating fully parametric hazard models via logistic regression.
-Their main idea is simple: comparing person-moments when the event of
-interest occurred with moments when patients were at risk. Their
-approach handles censored data effortlessly and provides users familiar
-with generalized linear models a natural way of fitting parametric
-survival models. Moreover, their framework is very flexible: general
-functions of time can be estimated (e.g.~using splines or general
-additive models), and hence these models retain some of the flexibility
-of Cox's partial likelihood approach. And since the unit of analysis is
-a person moment, time-varying covariates can also easily be included in
-this framework.
-
-Logistic regression had already been used in the context of
-discrete-time survival modeling \citep{cox1972regression}. But in the
-context of continuous-time survival modeling, using the framework of
-logistic regression opens the door to an extensive array of powerful
-modeling tools. Indeed, lasso and elastic-net regression can be used to
-select variables that are associated with the hazard function. Logistic
-regression can also be replaced by multinomial regression to analyse
-data with multiple competing events.
-
-In this article, we present the \CRANpkg{casebase} package
-\citep{casebase-package} implemented in R, for parametric survival
-analysis that combines the ideas of Hanley \& Miettinen
-\citeyearpar{hanley2009fitting} into a simple interface. The purpose of
-the \pkg{casebase} package is to provide practitioners with an
-easy-to-use software tool to compute a patient's risk (or cumulative
-incidence) of an event, conditional on a particular patient's covariate
-profile. Our package retains the flexibility of case-base sampling and
-the familiar interface of the \code{glm} function. It also provides
-tools for variable selection, competing-risk analysis, and dealing with
-time-dependent exposures. In addition, we provide functions for
-exploratory data analysis and visualizing the estimated quantities.
+The semiparametric Cox model has become the default approach to survival
+analysis even though Cox himself later suggested he would prefer to
+model the hazard function directly. In a 1994 interview with Professor
+Nancy Reid, Sir David Cox was asked how he would model a set of censored
+survival data, to which he responded: ``I think I would normally want to
+tackle problems parametrically \ldots{} and if you want to do things
+like predict the outcome for a particular patient, it's much more
+convenient to do that parametrically'' \citep{reid1994conversation}.
+Indeed, the most relevant quantity in a clinical setting is often the 5-
+or 10-year risk of experiencing a certain event given the patient's
+particular circumstances. Unfortunately, the most reported metric from a
+Cox model is the (potentially time-dependent) hazard ratio (HR), which
+ignores the duration of follow-up or is subject to selection bias
+\citep{hernan2010hazards}. The covariate adjusted survival curve
+overcomes these limitations, and is arguably a more important summary
+measure to report than the HR. While stepwise survival curves can be
+computed with the Cox model, they require a second step to separately
+estimate the baseline hazard \citep{breslow1972discussion}.
+
+Several authors have since pursued fully parametric approaches that made
+the fitting of smooth hazard functions much more transparent and
+intuitive through generalized linear models. The key feature of these
+procedures is splitting the time axis into discrete intervals.
+\citep{whitehead1980fitting} showed the equivalence between a Cox model
+and a Poisson regression with a parameter for each event time.
+\citep{copenhagen2012needs} provides a nice exposition of this
+equivalence with a real data example and supporting R code for computing
+standard errors. \citep{arjas1987logistic} and \citep{efron1988logistic}
+treated each patient-day as a Bernoulli random variable with probability
+equal to the discrete hazard rate. A potential issue with these
+approaches is that the number of time bins need to be chosen by the data
+analyst. A fine grouping of times may result in few (or none) events per
+interval, which then leads to instability in the Newton-Raphson
+procedure for estimation \citep{kalbfleisch2011statistical}. A coarse
+grouping could potentially mask nonlinear trends in the hazard function.
+
+Rather than discretizing time, Hanley \& Miettinen
+\citeyearpar{hanley2009fitting} selected a discrete set of person-time
+coordinates (`person-moments') in continuous time from all observed
+follow-up experience constituting the study base to obtain the logistic
+form likelihood expression for the hazard function. More specifically,
+all person-moments when the event of interest occurred are selected as
+the case series, complemented by a randomly sampled base series of
+person-moments serving as controls. This approach allows flexible
+modeling of the hazard function by including time as a covariate
+(e.g.~using splines or general additive models). Furthermore,
+time-dependent covariates can be modeled through interactions with time.
+In short, Hanley \& Miettinen \citeyearpar{hanley2009fitting} use the
+well-understood logistic regression for directly modeling the hazard
+function, but do not require an arbitrary choice of splitting time.
+
+In this article, we present the \CRANpkg{casebase} R package
+\citep{casebase-package} which implements the Hanley \& Miettinen
+\citeyearpar{hanley2009fitting} approach for fitting fully parametric
+hazard models and covariate adjusted survival curves using the familiar
+interface of the \code{glm} function. Our implementation allows for
+straightforward extensions to other models such as penalized regression
+for variable selection and competing-risk analysis. In addition, we
+provide functions for exploratory data analysis and visualizing the
+estimated quantities such as the hazard function, survival curve and its
+standard errors. The ultimate goal of our package is to make fitting
+flexible hazards accessible to more endusers with the hopes that they
+will favor reporting absolute risks over hazard ratios.
 

### 2021-10-13 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd

@@
 eval_cs1 <- TRUE
-eval_cs2 <- TRUE
+eval_cs2 <- F
 eval_cs3 <- TRUE
@@
 
+```{r  plot-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plot for the ERSPC dataset.
+A: The gray area can be thought of as N = 159,893 infinitely thin horizontal rectangles ordered by length of follow-up.
+B: The red points correspond to when death has occurred for any one of those infinitely thin rectangles.
+C: To improve visibility, these red points are randomly redistributed along their respective x-coordinates, providing a visualization of incidence density.
+More events are observed at later follow-up times, motivating the use of non-constant hazard models.
+D: The base series, a representative sample of the entire grey area, is represented by the green points."}
+
+data("ERSPC")
+#ogERSPC<-ERSPC
+ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
+                       levels = c(0,1), 
+                       labels = c("Control group", "Screening group"))
+
+toRemove<-sample(which(ERSPC$DeadOfPrCa==1),round(length(which(ERSPC$DeadOfPrCa==1))*0.7))
+ERSPC$DeadOfPrCa[toRemove]<-0
+
+a <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
+                               event = "DeadOfPrCa")
+
+plota<-emptyPlot<-plot(a,add.case.series=F,legend=F) +scale_y_continuous(labels = label_number_si())
+
+plotb<-emptyPlot+geom_point(aes(x=a$time,y=a$ycoord, color = "Case series", fill = "Case series"), show.legend = F, size = 1.5, alpha = 0.5, shape = 21)+ theme(legend.position="none") +scale_y_continuous(labels = label_number_si())
+
+plotc<-plot(a,add.case.series=T,legend=F) +scale_y_continuous(labels = label_number_si())
+
+#TEMP MAINLY NEED FOR LEGEND
+plotd<-plot(a,add.base.series=T) +scale_y_continuous(labels = label_number_si())
+
+legend_b <- get_legend(
+  plotd + 
+    guides(color = guide_legend(nrow = 1)) +
+     theme(legend.direction = "horizontal",legend.justification="center" ,legend.box.just = "bottom")
+)
+
+#remove legend
+plotd<-plotd + theme(legend.position="none")
+plotspop<-cowplot::plot_grid(plota,plotb,plotc,plotd,labels=c("A","B","C","D"))
+g<-plot_grid(plotspop, legend_b, ncol = 1, rel_heights = c(1, .1))
+g
+
+```
+
 We then pass this object to the corresponding \code{plot} method:
@@
 
-```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plots for both treatment arms in the ERSPC dataset.
+```{r plot-stratified-erspc-data, eval = F, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plots for both treatment arms in the ERSPC dataset.
 The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely thin rectangles (person-moments).
@@
 
-Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by first adding a layer for the area representing the population time in gray, with subjects having the least amount of observation time plotted at the top of the y-axis.
+Figure \ref{fig:plot-erspc-data} is built sequentially by first adding a layer for the area representing the population time in gray (Figure \ref{fig:plot-erspc-data} A), with subjects having the least amount of observation time plotted at the top of the y-axis.
 We immediately notice a distinctive *stepwise shape* in the population time area.
@@
 Next, layers for the case series and base series are added.
-The y-axis location of each case moment is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area.
+The y-axis location of each case moment  is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area (Figure \ref{fig:plot-erspc-data} B).
 By randomly distributing the cases, we can get a sense of the incidence density.
-In Figure \ref{fig:plot-stratified-erspc-data}, we see that more events are observed at later follow-up times.
+In Figure \ref{fig:plot-erspc-data} C, we see that more events are observed at later follow-up times.
 Therefore, a constant hazard model would not be appropriate in this instance as it would overestimate the incidence earlier on in time, and underestimate it later on.
-Finally, the base series is sampled horizontally with sampling weight proportional to their follow-up time.
+Finally, the base series is sampled horizontally with sampling weight proportional to their follow-up time (Figure \ref{fig:plot-erspc-data} D).
 The reader should refer to the package vignettes for more examples and a detailed description of how to modify the aesthetics of a population-time plot.
@@
 
+\textcolor{red}{Poisson regression can be used to estimate the full hazard.
+However, this method requires the user to decide on the number of intervals.
+Too fine the intervals may result in many empty bins.
+This may cause the commonly used Newton-Raphson procedure to fail to converge [@kalbfleisch2011statistical].
+If the intervals are too wide, the nonlinear trends that are present in the hazard may be masked.
+Rather than discretizing time like in Poisson regression, we sample person-specific moments which have been experienced during the study.}
+
 As presented in Hanley & Miettinen [-@hanley2009fitting], case-base sampling is comprised of three steps: 1) sampling a case series and a base series from the study; 2) fit the log-hazard as a linear function of predictors (including time); and 3) use the fitted hazard to estimate the risk function.

### 2021-10-13 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex

@@
 
+\begin{Schunk}
+\begin{figure}[ht]
+\includegraphics[width=\textwidth,keepaspectratio=true]{./plot-erspc-data-1} \caption[Population time plot for the ERSPC dataset]{Population time plot for the ERSPC dataset.
+A: The gray area can be thought of as N = 159,893 infinitely thin horizontal rectangles ordered by length of follow-up.
+B: The red points correspond to when death has occurred for any one of those infinitely thin rectangles.
+C: To improve visibility, these red points are randomly redistributed along their respective x-coordinates, providing a visualization of incidence density.
+More events are observed at later follow-up times, motivating the use of non-constant hazard models.
+D: The base series, a representative sample of the entire grey area, is represented by the green points.}\label{fig:plot-erspc-data}
+\end{figure}
+\end{Schunk}
+
 We then pass this object to the corresponding \code{plot} method:
@@
 
-\begin{Schunk}
-\begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./plot-stratified-erspc-data-1} \caption[Population time plots for both treatment arms in the ERSPC dataset]{Population time plots for both treatment arms in the ERSPC dataset.
-The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely thin rectangles (person-moments).
-More events are observed at later follow-up times, motivating the use of non-constant hazard models.}\label{fig:plot-stratified-erspc-data}
-\end{figure}
-\end{Schunk}
-
-Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by
-first adding a layer for the area representing the population time in
-gray, with subjects having the least amount of observation time plotted
-at the top of the y-axis. We immediately notice a distinctive
-\emph{stepwise shape} in the population time area. This is due to the
-randomization of the Finnish cohorts which were carried out on January 1
-of each of year from 1996 to 1999. Coupled with the uniform December 31
-2006 censoring date, this led to large numbers of men with exactly 11,
-10, 9 or 8 years of follow-up. Tracked backwards in time (i.e.~from
-right to left), the population-time plot shows the recruitment pattern
-from its beginning in 1991, and the January 1 entries in successive
-years. Tracked forwards in time (i.e.~from left to right), the plot for
-the first three years shows attrition due entirely to death (mainly from
-other causes). Since the Swedish and Belgian centres were the last to
-complete recruitment in December 2003, the minimum potential follow-up
-is three years. Tracked further forwards in time (i.e.~after year 3) the
-attrition is a combination of deaths and staggered entries. As we can
-see, population-time plots summarise a wealth of information about the
-study into a simple graph.
+Figure \ref{fig:plot-erspc-data} is built sequentially by first adding a
+layer for the area representing the population time in gray (Figure
+\ref{fig:plot-erspc-data} A), with subjects having the least amount of
+observation time plotted at the top of the y-axis. We immediately notice
+a distinctive \emph{stepwise shape} in the population time area. This is
+due to the randomization of the Finnish cohorts which were carried out
+on January 1 of each of year from 1996 to 1999. Coupled with the uniform
+December 31 2006 censoring date, this led to large numbers of men with
+exactly 11, 10, 9 or 8 years of follow-up. Tracked backwards in time
+(i.e.~from right to left), the population-time plot shows the
+recruitment pattern from its beginning in 1991, and the January 1
+entries in successive years. Tracked forwards in time (i.e.~from left to
+right), the plot for the first three years shows attrition due entirely
+to death (mainly from other causes). Since the Swedish and Belgian
+centres were the last to complete recruitment in December 2003, the
+minimum potential follow-up is three years. Tracked further forwards in
+time (i.e.~after year 3) the attrition is a combination of deaths and
+staggered entries. As we can see, population-time plots summarise a
+wealth of information about the study into a simple graph.
 
@@
 location of each case moment is sampled at random vertically on the plot
-to avoid having all points along the upper edge of the gray area. By
-randomly distributing the cases, we can get a sense of the incidence
-density. In Figure \ref{fig:plot-stratified-erspc-data}, we see that
-more events are observed at later follow-up times. Therefore, a constant
-hazard model would not be appropriate in this instance as it would
-overestimate the incidence earlier on in time, and underestimate it
-later on. Finally, the base series is sampled horizontally with sampling
-weight proportional to their follow-up time. The reader should refer to
-the package vignettes for more examples and a detailed description of
-how to modify the aesthetics of a population-time plot.
+to avoid having all points along the upper edge of the gray area (Figure
+\ref{fig:plot-erspc-data} B). By randomly distributing the cases, we can
+get a sense of the incidence density. In Figure
+\ref{fig:plot-erspc-data} C, we see that more events are observed at
+later follow-up times. Therefore, a constant hazard model would not be
+appropriate in this instance as it would overestimate the incidence
+earlier on in time, and underestimate it later on. Finally, the base
+series is sampled horizontally with sampling weight proportional to
+their follow-up time (Figure \ref{fig:plot-erspc-data} D). The reader
+should refer to the package vignettes for more examples and a detailed
+description of how to modify the aesthetics of a population-time plot.
 
@@
 #> Sample size: 159893 
-#> Number of events: 540 
-#> Number of base moments: 54000 
+#> Number of events: 162 
+#> Number of base moments: 16200 
 #> ----
@@
 #>    Min      1Q  Median      3Q     Max  
-#> -0.403  -0.172  -0.135  -0.085   3.801  
+#> -0.268  -0.171  -0.133  -0.091   3.623  
 #> 
@@
 #>                       Estimate Std. Error z value Pr(>|z|)    
-#> (Intercept)           -10.2533     0.3168  -32.36  < 2e-16 ***
-#> bs(Follow.Up.Time)1     4.2544     0.7976    5.33  9.6e-08 ***
-#> bs(Follow.Up.Time)2     2.1371     0.4720    4.53  5.9e-06 ***
-#> bs(Follow.Up.Time)3     4.5339     0.6889    6.58  4.7e-11 ***
-#> ScrArmScreening group  -0.2481     0.0886   -2.80   0.0051 ** 
+#> (Intercept)            -10.824      0.503  -21.52   <2e-16 ***
+#> bs(Follow.Up.Time)1      2.639      1.359    1.94   0.0523 .  
+#> bs(Follow.Up.Time)2      2.479      0.951    2.61   0.0092 ** 
+#> bs(Follow.Up.Time)3      3.063      1.354    2.26   0.0237 *  
+#> ScrArmScreening group   -0.207      0.161   -1.29   0.1986    
 #> ---
@@
 #> 
-#>     Null deviance: 6059.0  on 54539  degrees of freedom
-#> Residual deviance: 5787.6  on 54535  degrees of freedom
-#> AIC: 5798
+#>     Null deviance: 1817.7  on 16361  degrees of freedom
+#> Residual deviance: 1750.8  on 16357  degrees of freedom
+#> AIC: 1761
 #> 
@@
 #>     0 0.0e+00 0.0e+00
-#>     1 5.4e-05 4.2e-05
+#>     1 2.6e-05 2.1e-05
 \end{Soutput}
@@
 #>                    Df Deviance Resid. Df Resid. Dev Pr(>Chi)    
-#> NULL                               54539       6059             
-#> bs(Follow.Up.Time)  3      263     54536       5796   <2e-16 ***
-#> ScrArm              1        8     54535       5788   0.0048 ** 
+#> NULL                               16361       1818             
+#> bs(Follow.Up.Time)  3     65.2     16358       1752  4.5e-14 ***
+#> ScrArm              1      1.7     16357       1751      0.2    
 #> ---
@@
 #> 1    0  0.0e+00  0.0e+00   0.0e+00   Control group
-#> 2    1  5.4e-05  3.5e-05   9.3e-05   Control group
+#> 2    1  2.6e-05  1.3e-05   5.7e-05   Control group
 #> 3    0  0.0e+00  0.0e+00   0.0e+00 Screening group
-#> 4    1  4.2e-05  2.7e-05   7.3e-05 Screening group
+#> 4    1  2.1e-05  9.6e-06   4.4e-05 Screening group
 \end{Soutput}
@@
 \end{Sinput}
-\end{Schunk}
-
-\begin{Schunk}
-\begin{table}
-
-\caption{\label{tab:bmtcrr-cis}Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.}
-\centering
-\begin{tabular}[t]{lrlrl}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
-Covariates & HR & 95\% CI & HR & 95\% CI\\
-\midrule
-Sex & 0.74 & (0.43, 1.3) & 0.68 & (0.39, 1.21)\\
-Disease & 0.52 & (0.29, 0.94) & 0.52 & (0.28, 0.94)\\
-Phase (CR2 vs. CR1) & 1.20 & (0.48, 3) & 1.21 & (0.47, 3.09)\\
-Phase (CR3 vs. CR1) & 1.59 & (0.41, 6.17) & 1.67 & (0.46, 6.08)\\
-Phase (Relapse vs. CR1) & 4.09 & (1.9, 8.81) & 4.55 & (1.98, 10.46)\\
-\addlinespace
-Source & 1.70 & (0.55, 5.23) & 1.46 & (0.47, 4.53)\\
-Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
-\bottomrule
-\end{tabular}
-\end{table}
-
 \end{Schunk}
@@
 \end{Sinput}
-\end{Schunk}
-
-\begin{Schunk}
-\begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./bmtcrr-risk-1} \caption{\label{fig:compAbsrisk} Cumulative Incidence curve for a fixed covariate profile and the two disease groups.
-The estimate obtained from case-base sampling is compared to the Fine-Gray and Aalen-Johansen estimates.}\label{fig:bmtcrr-risk}
-\end{figure}
 \end{Schunk}
@@
 
+\textcolor{red}{Poisson regression can be used to estimate the full hazard.
+However, this method requires the user to decide on the number of intervals.
+Too fine the intervals may result in many empty bins.
+This may cause the commonly used Newton-Raphson procedure to fail to converge [@kalbfleisch2011statistical].
+If the intervals are too wide, the nonlinear trends that are present in the hazard may be masked.
+Rather than discretizing time like in Poisson regression, we sample person-specific moments which have been experienced during the study.}
+
 As presented in Hanley \& Miettinen \citeyearpar{hanley2009fitting},

### 2021-10-13 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
last review of paper

--- 2021-10-09 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
@@
 However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the estimating function.
-The baseline hazard then needs to be estimated separately using a non-parametric approach.
-This leads to step-wise estimates of the survival function that are difficult to interpret.
-Using case-base sampling, Hanley & Miettinen (2009) explained how parametric hazard functions can be estimated using logistic regression.
+As a consequence, the focus has been on reporting hazard ratios instead of survival curves.
+Indeed, reporting patient risk requires a separate estimation of the baseline hazard.
+Using case-base sampling, Hanley & Miettinen (2009) explained how parametric hazard functions can be estimated in continuous-time using logistic regression.
 Their approach naturally leads to estimates of the survival function that are smooth-in-time.
@@
 Unfortunately, the most reported metric from a Cox model is the (potentially 
-time-dependent) hazard ratio (HR), which ignores the duration of follow-up or 
-is subject to selection bias [@hernan2010hazards]. The covariate adjusted 
-survival curve overcomes these limitations, and is arguably a more important 
+time-dependent) hazard ratio (HR), which ignores the duration of follow-up and 
+is subject to selection bias [@hernan2010hazards]. The covariate-adjusted 
+survival curve overcomes these limitations, and it is arguably a more important 
 summary measure to report than the HR. While stepwise survival curves can be 
@@
 Several authors have since pursued fully parametric approaches that made the 
-fitting of smooth hazard functions much more transparent and intuitive through 
+fitting of smooth survival curves much more transparent and intuitive through 
 generalized linear models. The key feature of these procedures is splitting the 
-time axis into discrete intervals. [@whitehead1980fitting] showed the 
+time axis into discrete intervals. Whitehead [-@whitehead1980fitting] showed the 
 equivalence between a Cox model and a Poisson regression with a parameter for 
-each event time. [@copenhagen2012needs] provides a nice exposition of this 
+each event time; Carstensen [-@copenhagen2012needs] provides a nice exposition of this 
 equivalence with a real data example and supporting R code for computing 
-standard errors. [@arjas1987logistic] and [@efron1988logistic] treated each 
+standard errors. Arjas & Haara [-@arjas1987logistic] and Efron [-@efron1988logistic] treated each 
 patient-day as a Bernoulli random variable with probability equal to the 
 discrete hazard rate. A potential issue with these approaches is that the 
-number of time bins need to be chosen by the data analyst. A fine grouping 
+number of time bins need to be chosen by the data analyst. On the one hand, a fine grouping 
 of times may result in few (or none) events per interval, which then leads 
 to instability in the Newton-Raphson procedure for estimation 
-[@kalbfleisch2011statistical]. A coarse grouping could potentially mask 
+[@kalbfleisch2011statistical]. On the other hand, a coarse grouping could potentially mask 
 nonlinear trends in the hazard function.
@@
 continuous time from all observed follow-up experience constituting the 
-study base to obtain the logistic form likelihood expression for the hazard 
-function. More specifically, all person-moments when the event of interest 
+study base. By doing so, they obtained a likelihood expression for the hazard 
+function that is equivalent to that of logistic regression. More specifically, all person-moments when the event of interest 
 occurred are selected as the case series, complemented by a randomly sampled 
@@
 Miettinen [-@hanley2009fitting] use the well-understood logistic regression 
-for directly modeling the hazard function, but do not require an arbitrary 
-choice of splitting time.
+for directly modeling the hazard function, without requiring a discrete-time model.
 
@@
 which implements the Hanley & Miettinen [-@hanley2009fitting] approach for 
-fitting fully parametric hazard models and covariate adjusted survival curves 
+fitting fully parametric hazard models and covariate-adjusted survival curves 
 using the familiar interface of the \code{glm} function. Our implementation 
@@
 we provide functions for exploratory data analysis and visualizing the estimated 
-quantities such as the hazard function, survival curve and its standard errors. 
+quantities such as the hazard function, survival curve, and their standard errors. 
 The ultimate goal of our package is to make fitting flexible hazards accessible 
-to more endusers with the hopes that they will favor reporting absolute risks 
+to more end users with the hope that they will favor reporting absolute risks 
 over hazard ratios.
@@
 \pkg{casebase}. We show how the same framework can be used for non-linear 
-functions of time and non-proportional hazards, competing risks and variable 
+functions of time and non-proportional hazards, competing risks, and variable 
 selection via penalized regression. Finally, we end the article with a 
@@
 As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to consider the entire study base as an infinite collection of *person moments*.
-These person moments are indexed by both an individual in the study and a time point, and therefore each person moment has a covariate profile, an exposure status, and an outcome status attached to it.
+These person moments are indexed by both an individual in the study and a time point, and therefore each person moment has a covariate profile, an exposure status, and an outcome status (i.e.
+whether the event happened) attached to it.
 By comparing person moments at which an event occurred with person moments at which no event occurred, we can extract information about hazards and survival.
@@
 To do so, we model the base series sampling mechanism using non-homogeneous Poisson processes $R_i(t) \in \{0, 1, 2, \ldots\}$; the person-moments where $dR_i(t) = 1$ constitute the base series.
+We note that the same individual can contribute multiple person-moments to the base series.
 The process $Q_{i}(t) = R_i(t) + N_{i}(t)$ then counts both the case and base series person-moments contributed by individual $i$.
-The processes $R_i(t)$ are typically defined by the user via its intensity function $\rho_i(t)$.
+As mentioned above, the processes $R_i(t)$ are typically defined by the user via its intensity function $\rho_i(t)$.
 The process $Q_{i}(t)$ is characterized by $E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt$.
 
-If the hazard function $\lambda_{i}(t; \theta)$ is parametrized in terms of $\theta$, we could define an estimator $\hat{\theta}$ by maximization of the likelihood expression
+If the hazard function $\lambda_{i}(t; \theta)$ is parametrized in terms of $\theta$, we can define an estimator $\hat{\theta}$ by maximization of the likelihood expression
 $$L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^{\min(t_i,\tau)} \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)},$$
@@
 \end{equation}
-
 In this model, the instantaneous failure rate is independent of $t$.\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
@@
 
-Case-base sampling can also be used in the context of competing-risk analyses.
+Case-base sampling can also be used in the context of competing-risk analysis.
 Assuming there are $J$ competing events, we can show that each person-moment's contribution to the likelihood is of the form
@@
 $$\frac{\lambda_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},$$
-
 where $N_j(t)$ is the counting process associated with the event of type $j$ and $\lambda_j(t)$ is the corresponding cause-specific hazard function.
@@
 At the time of writing, a cursory examination of CRAN's \ctv{Survival} Task View reveals that there are over 250 packages related to survival analysis [@survTaskView].
-For the purposes of this article, we restricted our review to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and CIF estimation.
+For the purposes of this article, we restricted our review to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and Cumulative Incidence (CI) estimation.
 By searching for appropriate keywords in the \code{DESCRIPTION} file of these packages, we found 60 relevant packages.
 These 60 packages were then manually examined to determine which ones are comparable to \pkg{casebase}.
-In particular, we excluded packages that were focused on a different set of problems, such as frailty and multi-state models.
+In particular, we excluded packages that were focused on a different set of problems, such as frailty and multistate models.
 The remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with some of the functionalities they offer.
 
-Parametric survival models are implemented in a handful of packages, each differing in the underlying data distribution: \CRANpkg{CFC} [-@mahani2015bayesian], \CRANpkg{flexsurv} [-@flexsurv], \CRANpkg{SmoothHazard} [-@smoothHazard], \CRANpkg{rstpm2} [-@clements_liu], \CRANpkg{mets} [-@scheike2014estimating], and \CRANpkg{survival} [-@survival-package].
+Parametric survival models are implemented in several packages, each differing in the parametric distributions available: \CRANpkg{CFC} [-@mahani2015bayesian], \CRANpkg{flexsurv} [-@flexsurv], \CRANpkg{SmoothHazard} [-@smoothHazard], \CRANpkg{rstpm2} [-@clements_liu], \CRANpkg{mets} [-@scheike2014estimating], and \CRANpkg{survival} [-@survival-package].
 For example, \pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \pkg{flexsurv} and \pkg{survival} allow users to supply any distribution of their choice.
@@
 The package \CRANpkg{cmprsk} provides methods for cause-specific subdistribution hazards, such as in the Fine-Gray model [-@fine1999proportional].
-On the other hand, the package \pkg{CFC} estimates cause-specific CIFs from unadjusted, non-parametric survival functions.
+On the other hand, the package \pkg{CFC} estimates cause-specific CIs from unadjusted, non-parametric survival functions.
 Our package \pkg{casebase} also provides functionalities for competing risk analysis by estimating parametrically the cause-specific hazards.
-From these quantities, we can then estimate the cause-specific CIFs.
-
-Finally, several packages include functions to estimate the survival function and the CIF.
+From these quantities, we can then estimate the cause-specific CIs.
+
+Finally, several packages include functions to estimate the survival function and the CI.
 The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard.
 The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled.
-Using this estimate, the survival function and the CIF can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CDF} below).
+Using this estimate, the survival function and the CI can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CDF} below).
 Packages providing this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and \pkg{survival}.
@@
 }
-\caption{Comparison of various R packages for survival analysis based on several defining features.
-\textbf{Competing Risks}: whether or not an implementation for competing risks is present.
-\textbf{Allows Non PH}: permits models for non-proportional hazards.
+\caption{Comparison of various R packages for survival analysis.
+\textbf{Competing Risks}: whether an implementation for competing risks is present.
+\textbf{Allows Non PH}: includes models for non-proportional hazards.
 \textbf{Penalized Regression}: allows for a penalty term on the regression coefficients when estimating hazards (e.g.
 lasso or ridge).
-\textbf{Splines}: permits a flexible fit on time through the use of splines.
+\textbf{Splines}: allows a flexible fit on time through the use of splines.
 \textbf{Parametric}: implementation for parametric models.
@@
 If this is not selected, the package only handles right-censoring.
-\textbf{Risk estimates}: computation for survival curves, cumulative incidence or cumulative hazard is readily available.}
+\textbf{Risk estimates}: estimation of survival curve and cumulative incidence is available.}
 \label{tab:surv-pkgs}
@@
 
-Population-time plots are a descriptive visualization of incidence density, where aggregate person-time is represented by area and events as points within the area.
+Population-time plots are a descriptive visualization of incidence density, where the study base is represented by area and events by points within the area.
 The case-base sampling approach described above can be visualized in the form of a population time plot.
-These plots are extremely informative graphical displays of survival data and should be one of the first steps in an exploratory data analysis.
+These plots are informative graphical displays of survival data and should be one of the first steps in an exploratory data analysis.
 The \code{popTime} function and \code{plot} method facilitate this task:
@@
 
-```{r erspc-data, eval = eval_cs1, echo = c(-1:-3)}
+```{r erspc-data-mutate, eval = eval_cs1, echo = FALSE}
 data("ERSPC")
@@
                        labels = c("Control group", "Screening group"))
-
+```
+
+```{r erspc-data, eval = eval_cs1}
 pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
@@
 
+```{r plot-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plot for the ERSPC dataset.
+\\textbf{A}: The gray area can be thought of as N = 159,893 infinitely thin horizontal rectangles ordered by length of follow-up.
+\\textbf{B}: The red points correspond to when death has occurred for any one of those infinitely thin rectangles.
+\\textbf{C}: To improve visibility, these red points are randomly redistributed along their respective x-coordinates, providing a visualization of incidence density.
+More events are observed at later follow-up times, motivating the use of non-constant hazard models.
+\\textbf{D}: The base series, a representative sample of the entire grey area, is represented by the green points."}
+# For increased visibility, remove some cases
+toRemove <- sample(which(ERSPC$DeadOfPrCa == 1), 
+                   round(length(which(ERSPC$DeadOfPrCa == 1))*0.7))
+# Create new dataset for illustration
+ERSPC_sub <- ERSPC
+ERSPC_sub$DeadOfPrCa[toRemove] <- 0
+
+a <- casebase::popTime(ERSPC_sub, time = "Follow.Up.Time",
+                       event = "DeadOfPrCa")
+
+plota <- emptyPlot <- plot(a, add.case.series = FALSE, 
+                           legend = FALSE) + 
+  scale_y_continuous(labels = label_number_si())
+
+plotb <- emptyPlot + geom_point(data = dplyr::filter(a, event == 1),
+                                aes(x = time, y = ycoord, 
+                                    color = "Case series",
+                                    fill = "Case series"), 
+                                show.legend = FALSE, size = 1.5, 
+                                alpha = 0.5, shape = 21) + 
+  theme(legend.position = "none") + 
+  scale_y_continuous(labels = label_number_si())
+
+plotc <- plot(a, add.case.series = TRUE, legend = FALSE) +
+  scale_y_continuous(labels = label_number_si())
+
+#TEMP MAINLY NEED FOR LEGEND
+plotd <- plot(a, add.base.series = TRUE) +
+  scale_y_continuous(labels = label_number_si())
+
+legend_b <- get_legend(
+  plotd + 
+    guides(color = guide_legend(nrow = 1)) +
+     theme(legend.direction = "horizontal",
+           legend.justification = "center",
+           legend.box.just = "bottom")
+)
+
+#remove legend
+plotd <- plotd + theme(legend.position = "none")
+plotspop <- cowplot::plot_grid(plota, plotb, plotc, plotd,
+                               labels = c("A", "B", "C", "D"))
+g <- plot_grid(plotspop, legend_b, ncol = 1, rel_heights = c(1, .1))
+g
+```
+
 We then pass this object to the corresponding \code{plot} method:
@@
 
-```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plots for both treatment arms in the ERSPC dataset.
+```{r plot-stratified-erspc-data, eval = FALSE, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plots for both treatment arms in the ERSPC dataset.
 The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely thin rectangles (person-moments).
@@
 
-Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by first adding a layer for the area representing the population time in gray, with subjects having the least amount of observation time plotted at the top of the y-axis.
+Figure \ref{fig:plot-erspc-data} depicts the process of creating a population-time plot.
+It is built sequentially by first adding a layer for the area representing the population time in gray (Figure \ref{fig:plot-erspc-data} A), with subjects having the least amount of observation time plotted at the top of the y-axis.
 We immediately notice a distinctive *stepwise shape* in the population time area.
@@
 Next, layers for the case series and base series are added.
-The y-axis location of each case moment is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area.
+The y-axis location of each case moment is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area (Figure \ref{fig:plot-erspc-data} B).
 By randomly distributing the cases, we can get a sense of the incidence density.
-In Figure \ref{fig:plot-stratified-erspc-data}, we see that more events are observed at later follow-up times.
+In Figure \ref{fig:plot-erspc-data} C, we see that more events are observed at later follow-up times.
 Therefore, a constant hazard model would not be appropriate in this instance as it would overestimate the incidence earlier on in time, and underestimate it later on.
-Finally, the base series is sampled horizontally with sampling weight proportional to their follow-up time.
+Finally, the base series is sampled uniformly from the study base (Figure \ref{fig:plot-erspc-data} D).
 The reader should refer to the package vignettes for more examples and a detailed description of how to modify the aesthetics of a population-time plot.
@@
 
-Finally, by integrating these subdensities, we obtain the event-specific cumulative incidence functions:
+By integrating these subdensities, we obtain the event-specific CI functions:
 
@@
 
+Finally, the output from \code{absoluteRisk} can be passed to a method \code{confint} to compute confidence bands around the survival function.
+These bands are computed using parametric bootstrap, and therefore are only valid when \code{family = "glm"} as it relies on the asymptotic normality of the estimator.
+Currently, this is only available for the single-event setting.
+
 # Illustration of package
@@
 Each one showcases a different type of analysis.
-First, we show how to model different functional forms of time as well as non-proportional hazards.
-Then we perform a competing-risk analysis and compare our results with the Fine-Gray model.
+First, we show how to model time flexibly as well as non-proportional hazards.
+Then we perform a competing-risk analysis and compare our results with the Cox model and the Fine-Gray model.
 The third case study illustrates how to perform variable selection in high-dimensional datasets.
@@
 
-Moreover, we can also compute confidence intervals for the risk function using the method \code{confint.absRiskCB}.
+Next, we can compute confidence intervals for the risk function using the method \code{confint.absRiskCB}.
 
@@
 
-Previous re-analyses of these data suggested that the overall screening attributed reduction in death due to prostate cancer of 20% was an underestimate [@hanley2010mortality].
+Previous re-analyses of these data suggest that the 20% reduction in prostate cancer death due to screening was an underestimate [@hanley2010mortality].
 The estimated 20% (from a proportional hazards model) did not account for the delay between screening and the time the effect is expected to be observed.
@@
 
+Note that we did not have to specify any cut points, as would be the case with the \code{survSplit} function in the \pkg{survival} package.
 In Figure \ref{fig:interaction-ERSPC}, we have the estimated hazard ratio and 95% confidence interval for screening vs.
@@
 
-The plot shows that the cures attributable to the screening only begin to become statistically apparent by year 7 and later.
+The plot shows that the effect of screening only becomes statistically apparent by year 7 and later.
 The 25-60% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20% reported in the original article.
@@
     format = "latex", booktabs = TRUE,
-    col.names = c("Covariates", "HR", "95% CI", "HR", "95% CI"),
+    col.names = c("Covariates", "HR", "95% Conf.", "HR", "95% Conf."),
     caption = "Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented."
@@
 These quantities appear in Table \ref{tab:bmtcrr-cis}.
-As we can see, the only significant hazard ratio identified by case-base sampling is the one associated with the phase of the disease at transplant.
-More precisely, being in relapse at transplant is associated with a hazard ratio of 3.89 when compared to CR1.
+As we can see, the type of disease corresponds to a significant hazard ratio: the hazard for AML is 48% lower than that for ALL.
+Moreover, being in relapse at transplant is associated with a hazard ratio of 4.09 when compared to CR1.
 
@@
 The Fine-Gray model is a semiparametric model for the cause-specific *subdistribution hazard*, i.e.
-the function $d_k(t)$ such that
-$$CI_k(t) =1 - \exp\left( - \int_0^t d_k(u) \textrm{d}u \right),$$
-where $CI_k(t)$ is the cause-specific cumulative incidence.
+the function $d_j(t)$ such that
+$$CI_j(t) = 1 - \exp\left( - \int_0^t d_j(u) \textrm{d}u \right),$$
+where $CI_j(t)$ is the cause-specific CI.
 The Fine-Gray model allows to directly assess the effect of a covariate on the subdistribution hazard, as opposed to the cause-specific hazard.
@@
 
-Figure \ref{fig:compAbsrisk} shows the cumulative incidence curves for all three models.
+Figure \ref{fig:compAbsrisk} shows the CI curves for all three models.
 As we can see, all three approaches agree quite well for AML; however, for ALL, there seems to be a difference of about 5% between the Fine-Gray curve and the curves estimated using case-base sampling and Cox regression.
@@
 
-For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function as given by \eqref{eq:penest}.
+For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function as given by Equation \ref{eq:penest}.
 We note that this is different than the semiparametric model Coxnet, which regularizes the Cox partial likelihood.
@@
 The penalized logistic regression is then fit for multiple values of the tuning parameter using the function \code{glmnet::cv.glmnet} and the binomial family.
-We supply both a matrix \code{y} containing the time and event variables, and a matrix \code{x} containing all other covariates.
+To \code{fitSmoothHazard.fit}, we supply both a matrix \code{y} containing the time and event variables, and a matrix \code{x} containing all other covariates.
 We apply the lasso penalty by setting \code{alpha = 1} and assign a \code{penalty.factor} ($w_j$; cf.
@@
 
+Poisson regression can also be used to estimate the full hazard by discretizing time.
+However, this method requires user input on the number of intervals, or equivalently, on the cut points.
+This choice made by the user can have a significant impact on the model.
+Small intervals may result in many empty, non-informative bins.
+This may cause convergence issues for the Newton-Raphson procedure [@kalbfleisch2011statistical].
+If the intervals are too wide, the nonlinear trends that are present in the hazard may be masked.
+Rather than discretizing time like in Poisson regression, case-base sampling provides a continuous-time approach to using GLMs for estimating hazard functions.
+
 As presented in Hanley & Miettinen [-@hanley2009fitting], case-base sampling is comprised of three steps: 1) sampling a case series and a base series from the study; 2) fit the log-hazard as a linear function of predictors (including time); and 3) use the fitted hazard to estimate the risk function.
@@
 However, in terms of output, both approaches differ significantly.
-Case-base sampling produces smooth hazards and smooth survival curves, whereas Cox regression and Fine-Gray models produce stepwise CIFs and never explicitly model the hazard function.
+Case-base sampling produces smooth hazards and smooth survival curves, whereas Cox regression and Fine-Gray models produce stepwise CIs and never explicitly model the hazard function.
 Qualitatively, we showed that by using splines in the linear predictor, all three models yielded similar curves.
-However, the smooth nature of the output of \pkg{casebase} provides a more intuitive interpretation for consumers of these predictions.
 

### 2021-10-13 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
last review of paper

--- 2021-10-09 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
@@
 relying on time matching or risk-set sampling (including Cox regression)
-eliminate the baseline hazard from the estimating function. The baseline
-hazard then needs to be estimated separately using a non-parametric
-approach. This leads to step-wise estimates of the survival function
-that are difficult to interpret. Using case-base sampling, Hanley \&
+eliminate the baseline hazard from the estimating function. As a
+consequence, the focus has been on reporting hazard ratios instead of
+survival curves. Indeed, reporting patient risk requires a separate
+estimation of the baseline hazard. Using case-base sampling, Hanley \&
 Miettinen (2009) explained how parametric hazard functions can be
-estimated using logistic regression. Their approach naturally leads to
-estimates of the survival function that are smooth-in-time.\\
+estimated in continuous-time using logistic regression. Their approach
+naturally leads to estimates of the survival function that are
+smooth-in-time.\\
 In this paper, we present the \CRANpkg{casebase} R package, a
@@
 Cox model is the (potentially time-dependent) hazard ratio (HR), which
-ignores the duration of follow-up or is subject to selection bias
-\citep{hernan2010hazards}. The covariate adjusted survival curve
-overcomes these limitations, and is arguably a more important summary
+ignores the duration of follow-up and is subject to selection bias
+\citep{hernan2010hazards}. The covariate-adjusted survival curve
+overcomes these limitations, and it is arguably a more important summary
 measure to report than the HR. While stepwise survival curves can be
@@
 Several authors have since pursued fully parametric approaches that made
-the fitting of smooth hazard functions much more transparent and
+the fitting of smooth survival curves much more transparent and
 intuitive through generalized linear models. The key feature of these
-procedures is splitting the time axis into discrete intervals.
-\citep{whitehead1980fitting} showed the equivalence between a Cox model
-and a Poisson regression with a parameter for each event time.
-\citep{copenhagen2012needs} provides a nice exposition of this
-equivalence with a real data example and supporting R code for computing
-standard errors. \citep{arjas1987logistic} and \citep{efron1988logistic}
-treated each patient-day as a Bernoulli random variable with probability
-equal to the discrete hazard rate. A potential issue with these
-approaches is that the number of time bins need to be chosen by the data
-analyst. A fine grouping of times may result in few (or none) events per
-interval, which then leads to instability in the Newton-Raphson
-procedure for estimation \citep{kalbfleisch2011statistical}. A coarse
-grouping could potentially mask nonlinear trends in the hazard function.
+procedures is splitting the time axis into discrete intervals. Whitehead
+\citeyearpar{whitehead1980fitting} showed the equivalence between a Cox
+model and a Poisson regression with a parameter for each event time;
+Carstensen \citeyearpar{copenhagen2012needs} provides a nice exposition
+of this equivalence with a real data example and supporting R code for
+computing standard errors. Arjas \& Haara
+\citeyearpar{arjas1987logistic} and Efron
+\citeyearpar{efron1988logistic} treated each patient-day as a Bernoulli
+random variable with probability equal to the discrete hazard rate. A
+potential issue with these approaches is that the number of time bins
+need to be chosen by the data analyst. On the one hand, a fine grouping
+of times may result in few (or none) events per interval, which then
+leads to instability in the Newton-Raphson procedure for estimation
+\citep{kalbfleisch2011statistical}. On the other hand, a coarse grouping
+could potentially mask nonlinear trends in the hazard function.
 
@@
 coordinates (`person-moments') in continuous time from all observed
-follow-up experience constituting the study base to obtain the logistic
-form likelihood expression for the hazard function. More specifically,
-all person-moments when the event of interest occurred are selected as
-the case series, complemented by a randomly sampled base series of
+follow-up experience constituting the study base. By doing so, they
+obtained a likelihood expression for the hazard function that is
+equivalent to that of logistic regression. More specifically, all
+person-moments when the event of interest occurred are selected as the
+case series, complemented by a randomly sampled base series of
 person-moments serving as controls. This approach allows flexible
@@
 well-understood logistic regression for directly modeling the hazard
-function, but do not require an arbitrary choice of splitting time.
+function, without requiring a discrete-time model.
 
@@
 \citeyearpar{hanley2009fitting} approach for fitting fully parametric
-hazard models and covariate adjusted survival curves using the familiar
+hazard models and covariate-adjusted survival curves using the familiar
 interface of the \code{glm} function. Our implementation allows for
@@
 provide functions for exploratory data analysis and visualizing the
-estimated quantities such as the hazard function, survival curve and its
-standard errors. The ultimate goal of our package is to make fitting
-flexible hazards accessible to more endusers with the hopes that they
-will favor reporting absolute risks over hazard ratios.
+estimated quantities such as the hazard function, survival curve, and
+their standard errors. The ultimate goal of our package is to make
+fitting flexible hazards accessible to more end users with the hope that
+they will favor reporting absolute risks over hazard ratios.
 
@@
 show how the same framework can be used for non-linear functions of time
-and non-proportional hazards, competing risks and variable selection via
-penalized regression. Finally, we end the article with a discussion of
-the results and of future directions.
+and non-proportional hazards, competing risks, and variable selection
+via penalized regression. Finally, we end the article with a discussion
+of the results and of future directions.
 
@@
 therefore each person moment has a covariate profile, an exposure
-status, and an outcome status attached to it. By comparing person
-moments at which an event occurred with person moments at which no event
-occurred, we can extract information about hazards and survival.
+status, and an outcome status (i.e.~whether the event happened) attached
+to it. By comparing person moments at which an event occurred with
+person moments at which no event occurred, we can extract information
+about hazards and survival.
 
@@
 non-homogeneous Poisson processes \(R_i(t) \in \{0, 1, 2, \ldots\}\);
-the person-moments where \(dR_i(t) = 1\) constitute the base series. The
-process \(Q_{i}(t) = R_i(t) + N_{i}(t)\) then counts both the case and
-base series person-moments contributed by individual \(i\). The
-processes \(R_i(t)\) are typically defined by the user via its intensity
-function \(\rho_i(t)\). The process \(Q_{i}(t)\) is characterized by
+the person-moments where \(dR_i(t) = 1\) constitute the base series. We
+note that the same individual can contribute multiple person-moments to
+the base series. The process \(Q_{i}(t) = R_i(t) + N_{i}(t)\) then
+counts both the case and base series person-moments contributed by
+individual \(i\). As mentioned above, the processes \(R_i(t)\) are
+typically defined by the user via its intensity function \(\rho_i(t)\).
+The process \(Q_{i}(t)\) is characterized by
 \(E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt\).
@@
 If the hazard function \(\lambda_{i}(t; \theta)\) is parametrized in
-terms of \(\theta\), we could define an estimator \(\hat{\theta}\) by
+terms of \(\theta\), we can define an estimator \(\hat{\theta}\) by
 maximization of the likelihood expression
@@
 \log(\lambda(t; X)) = \beta_0 + \beta_1 X. \label{eq:exp}
-\end{equation}
-
-In this model, the instantaneous failure rate is independent of
+\end{equation} In this model, the instantaneous failure rate is
+independent of
 \(t\).\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
@@
 Case-base sampling can also be used in the context of competing-risk
-analyses. Assuming there are \(J\) competing events, we can show that
+analysis. Assuming there are \(J\) competing events, we can show that
 each person-moment's contribution to the likelihood is of the form
@@
 \[\frac{\lambda_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},\]
-
 where \(N_j(t)\) is the counting process associated with the event of
@@
 following features: parametric modeling, non-proportional hazard models,
-competing risk analysis, penalized estimation, and CIF estimation. By
-searching for appropriate keywords in the \code{DESCRIPTION} file of
-these packages, we found 60 relevant packages. These 60 packages were
-then manually examined to determine which ones are comparable to
-\pkg{casebase}. In particular, we excluded packages that were focused on
-a different set of problems, such as frailty and multi-state models. The
-remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with
-some of the functionalities they offer.
-
-Parametric survival models are implemented in a handful of packages,
-each differing in the underlying data distribution: \CRANpkg{CFC}
+competing risk analysis, penalized estimation, and Cumulative Incidence
+(CI) estimation. By searching for appropriate keywords in the
+\code{DESCRIPTION} file of these packages, we found 60 relevant
+packages. These 60 packages were then manually examined to determine
+which ones are comparable to \pkg{casebase}. In particular, we excluded
+packages that were focused on a different set of problems, such as
+frailty and multistate models. The remaining 14 packages appear in Table
+\ref{tab:surv-pkgs}, along with some of the functionalities they offer.
+
+Parametric survival models are implemented in several packages, each
+differing in the parametric distributions available: \CRANpkg{CFC}
 \citeyearpar{mahani2015bayesian}, \CRANpkg{flexsurv}
@@
 \citeyearpar{fine1999proportional}. On the other hand, the package
-\pkg{CFC} estimates cause-specific CIFs from unadjusted, non-parametric
+\pkg{CFC} estimates cause-specific CIs from unadjusted, non-parametric
 survival functions. Our package \pkg{casebase} also provides
@@
 the cause-specific hazards. From these quantities, we can then estimate
-the cause-specific CIFs.
+the cause-specific CIs.
 
 Finally, several packages include functions to estimate the survival
-function and the CIF. The corresponding methods generally fall into two
+function and the CI. The corresponding methods generally fall into two
 categories: transformation of the estimated hazard function, and
@@
 is explicitly modeled. Using this estimate, the survival function and
-the CIF can be obtained using their functional relationships (see
+the CI can be obtained using their functional relationships (see
 Equations \ref{eqn:surv} and \ref{eqn:CDF} below). Packages providing
@@
 Population-time plots are a descriptive visualization of incidence
-density, where aggregate person-time is represented by area and events
-as points within the area. The case-base sampling approach described
-above can be visualized in the form of a population time plot. These
-plots are extremely informative graphical displays of survival data and
-should be one of the first steps in an exploratory data analysis. The
-\code{popTime} function and \code{plot} method facilitate this task:
+density, where the study base is represented by area and events by
+points within the area. The case-base sampling approach described above
+can be visualized in the form of a population time plot. These plots are
+informative graphical displays of survival data and should be one of the
+first steps in an exploratory data analysis. The \code{popTime} function
+and \code{plot} method facilitate this task:
 
@@
 \begin{Schunk}
-\begin{Sinput}
-pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
-                               event = "DeadOfPrCa", exposure = "ScrArm")
-inherits(pt_object, "popTime")
-\end{Sinput}
 \begin{Soutput}
@@
 \end{Soutput}
-\begin{Sinput}
-attr(pt_object, "exposure")
-\end{Sinput}
 \begin{Soutput}
@@
 
+\begin{Schunk}
+\begin{figure}[ht]
+\includegraphics[width=\textwidth,keepaspectratio=true]{./plot-erspc-data-1} \caption{Population time plot for the ERSPC dataset.
+\textbf{A}: The gray area can be thought of as N = 159,893 infinitely thin horizontal rectangles ordered by length of follow-up.
+\textbf{B}: The red points correspond to when death has occurred for any one of those infinitely thin rectangles.
+\textbf{C}: To improve visibility, these red points are randomly redistributed along their respective x-coordinates, providing a visualization of incidence density.
+More events are observed at later follow-up times, motivating the use of non-constant hazard models.
+\textbf{D}: The base series, a representative sample of the entire grey area, is represented by the green points.}\label{fig:plot-erspc-data}
+\end{figure}
+\end{Schunk}
+
 We then pass this object to the corresponding \code{plot} method:
@@
 
-\begin{Schunk}
-\begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./plot-stratified-erspc-data-1} \caption[Population time plots for both treatment arms in the ERSPC dataset]{Population time plots for both treatment arms in the ERSPC dataset.
-The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely thin rectangles (person-moments).
-More events are observed at later follow-up times, motivating the use of non-constant hazard models.}\label{fig:plot-stratified-erspc-data}
-\end{figure}
-\end{Schunk}
-
-Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by
-first adding a layer for the area representing the population time in
-gray, with subjects having the least amount of observation time plotted
-at the top of the y-axis. We immediately notice a distinctive
-\emph{stepwise shape} in the population time area. This is due to the
-randomization of the Finnish cohorts which were carried out on January 1
-of each of year from 1996 to 1999. Coupled with the uniform December 31
-2006 censoring date, this led to large numbers of men with exactly 11,
-10, 9 or 8 years of follow-up. Tracked backwards in time (i.e.~from
-right to left), the population-time plot shows the recruitment pattern
-from its beginning in 1991, and the January 1 entries in successive
-years. Tracked forwards in time (i.e.~from left to right), the plot for
-the first three years shows attrition due entirely to death (mainly from
-other causes). Since the Swedish and Belgian centres were the last to
-complete recruitment in December 2003, the minimum potential follow-up
-is three years. Tracked further forwards in time (i.e.~after year 3) the
-attrition is a combination of deaths and staggered entries. As we can
-see, population-time plots summarise a wealth of information about the
-study into a simple graph.
+Figure \ref{fig:plot-erspc-data} depicts the process of creating a
+population-time plot. It is built sequentially by first adding a layer
+for the area representing the population time in gray (Figure
+\ref{fig:plot-erspc-data} A), with subjects having the least amount of
+observation time plotted at the top of the y-axis. We immediately notice
+a distinctive \emph{stepwise shape} in the population time area. This is
+due to the randomization of the Finnish cohorts which were carried out
+on January 1 of each of year from 1996 to 1999. Coupled with the uniform
+December 31 2006 censoring date, this led to large numbers of men with
+exactly 11, 10, 9 or 8 years of follow-up. Tracked backwards in time
+(i.e.~from right to left), the population-time plot shows the
+recruitment pattern from its beginning in 1991, and the January 1
+entries in successive years. Tracked forwards in time (i.e.~from left to
+right), the plot for the first three years shows attrition due entirely
+to death (mainly from other causes). Since the Swedish and Belgian
+centres were the last to complete recruitment in December 2003, the
+minimum potential follow-up is three years. Tracked further forwards in
+time (i.e.~after year 3) the attrition is a combination of deaths and
+staggered entries. As we can see, population-time plots summarise a
+wealth of information about the study into a simple graph.
 
@@
 location of each case moment is sampled at random vertically on the plot
-to avoid having all points along the upper edge of the gray area. By
-randomly distributing the cases, we can get a sense of the incidence
-density. In Figure \ref{fig:plot-stratified-erspc-data}, we see that
-more events are observed at later follow-up times. Therefore, a constant
-hazard model would not be appropriate in this instance as it would
-overestimate the incidence earlier on in time, and underestimate it
-later on. Finally, the base series is sampled horizontally with sampling
-weight proportional to their follow-up time. The reader should refer to
-the package vignettes for more examples and a detailed description of
-how to modify the aesthetics of a population-time plot.
+to avoid having all points along the upper edge of the gray area (Figure
+\ref{fig:plot-erspc-data} B). By randomly distributing the cases, we can
+get a sense of the incidence density. In Figure
+\ref{fig:plot-erspc-data} C, we see that more events are observed at
+later follow-up times. Therefore, a constant hazard model would not be
+appropriate in this instance as it would overestimate the incidence
+earlier on in time, and underestimate it later on. Finally, the base
+series is sampled uniformly from the study base (Figure
+\ref{fig:plot-erspc-data} D). The reader should refer to the package
+vignettes for more examples and a detailed description of how to modify
+the aesthetics of a population-time plot.
 
@@
 
-Finally, by integrating these subdensities, we obtain the event-specific
-cumulative incidence functions:
+By integrating these subdensities, we obtain the event-specific CI
+functions:
 
@@
 
+Finally, the output from \code{absoluteRisk} can be passed to a method
+\code{confint} to compute confidence bands around the survival function.
+These bands are computed using parametric bootstrap, and therefore are
+only valid when \code{family = "glm"} as it relies on the asymptotic
+normality of the estimator. Currently, this is only available for the
+single-event setting.
+
 \hypertarget{illustration-of-package}{%
@@
 package through three case studies. Each one showcases a different type
-of analysis. First, we show how to model different functional forms of
-time as well as non-proportional hazards. Then we perform a
-competing-risk analysis and compare our results with the Fine-Gray
-model. The third case study illustrates how to perform variable
-selection in high-dimensional datasets.
+of analysis. First, we show how to model time flexibly as well as
+non-proportional hazards. Then we perform a competing-risk analysis and
+compare our results with the Cox model and the Fine-Gray model. The
+third case study illustrates how to perform variable selection in
+high-dimensional datasets.
 
@@
 
-Moreover, we can also compute confidence intervals for the risk function
-using the method \code{confint.absRiskCB}.
+Next, we can compute confidence intervals for the risk function using
+the method \code{confint.absRiskCB}.
 
@@
 
-Previous re-analyses of these data suggested that the overall screening
-attributed reduction in death due to prostate cancer of 20\% was an
-underestimate \citep{hanley2010mortality}. The estimated 20\% (from a
-proportional hazards model) did not account for the delay between
-screening and the time the effect is expected to be observed. As a
-result, the null effects in years 1--7 masked the substantial reductions
-that began to appear from year 8 onward. This motivates the use of a
-time-dependent hazard ratio which can easily be fit with the
-\pkg{casebase} package by including an interaction term with time in the
-model:
+Previous re-analyses of these data suggest that the 20\% reduction in
+prostate cancer death due to screening was an underestimate
+\citep{hanley2010mortality}. The estimated 20\% (from a proportional
+hazards model) did not account for the delay between screening and the
+time the effect is expected to be observed. As a result, the null
+effects in years 1--7 masked the substantial reductions that began to
+appear from year 8 onward. This motivates the use of a time-dependent
+hazard ratio which can easily be fit with the \pkg{casebase} package by
+including an interaction term with time in the model:
 
@@
 
+Note that we did not have to specify any cut points, as would be the
+case with the \code{survSplit} function in the \pkg{survival} package.
 In Figure \ref{fig:interaction-ERSPC}, we have the estimated hazard
@@
 
-The plot shows that the cures attributable to the screening only begin
-to become statistically apparent by year 7 and later. The 25-60\%
-reductions seen in years 8-12 of the study suggests a much higher
-reduction in prostate cancer due to screening than the single overall
-20\% reported in the original article.
+The plot shows that the effect of screening only becomes statistically
+apparent by year 7 and later. The 25-60\% reductions seen in years 8-12
+of the study suggests a much higher reduction in prostate cancer due to
+screening than the single overall 20\% reported in the original article.
 
@@
 \cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
-Covariates & HR & 95\% CI & HR & 95\% CI\\
+Covariates & HR & 95\% Conf. & HR & 95\% Conf.\\
 \midrule
@@
 corresponding confidence intervals. These quantities appear in Table
-\ref{tab:bmtcrr-cis}. As we can see, the only significant hazard ratio
-identified by case-base sampling is the one associated with the phase of
-the disease at transplant. More precisely, being in relapse at
-transplant is associated with a hazard ratio of 3.89 when compared to
-CR1.
+\ref{tab:bmtcrr-cis}. As we can see, the type of disease corresponds to
+a significant hazard ratio: the hazard for AML is 48\% lower than that
+for ALL. Moreover, being in relapse at transplant is associated with a
+hazard ratio of 4.09 when compared to CR1.
 
@@
 is a semiparametric model for the cause-specific \emph{subdistribution
-hazard}, i.e.~the function \(d_k(t)\) such that
-\[CI_k(t) =1 - \exp\left( - \int_0^t d_k(u) \textrm{d}u \right),\] where
-\(CI_k(t)\) is the cause-specific cumulative incidence. The Fine-Gray
-model allows to directly assess the effect of a covariate on the
-subdistribution hazard, as opposed to the cause-specific hazard. For the
-computation, we use the \CRANpkg{timereg} package \citep{timereg}:
+hazard}, i.e.~the function \(d_j(t)\) such that
+\[CI_j(t) = 1 - \exp\left( - \int_0^t d_j(u) \textrm{d}u \right),\]
+where \(CI_j(t)\) is the cause-specific CI. The Fine-Gray model allows
+to directly assess the effect of a covariate on the subdistribution
+hazard, as opposed to the cause-specific hazard. For the computation, we
+use the \CRANpkg{timereg} package \citep{timereg}:
 
@@
 
-Figure \ref{fig:compAbsrisk} shows the cumulative incidence curves for
-all three models. As we can see, all three approaches agree quite well
-for AML; however, for ALL, there seems to be a difference of about 5\%
-between the Fine-Gray curve and the curves estimated using case-base
-sampling and Cox regression. This difference does not appear to be
-significant: the curve from case-base sampling is contained within a
-95\% confidence band around the Fine-Gray absolute risk curve (figure
-not shown).
+Figure \ref{fig:compAbsrisk} shows the CI curves for all three models.
+As we can see, all three approaches agree quite well for AML; however,
+for ALL, there seems to be a difference of about 5\% between the
+Fine-Gray curve and the curves estimated using case-base sampling and
+Cox regression. This difference does not appear to be significant: the
+curve from case-base sampling is contained within a 95\% confidence band
+around the Fine-Gray absolute risk curve (figure not shown).
 
@@
 for variable selection through regularized estimation of the hazard
-function as given by \eqref{eq:penest}. We note that this is different
-than the semiparametric model Coxnet, which regularizes the Cox partial
-likelihood. To illustrate this functionality, we use the dataset from
-the Study to Understand Prognoses Preferences Outcomes and Risks of
-Treatment (SUPPORT)
+function as given by Equation \ref{eq:penest}. We note that this is
+different than the semiparametric model Coxnet, which regularizes the
+Cox partial likelihood. To illustrate this functionality, we use the
+dataset from the Study to Understand Prognoses Preferences Outcomes and
+Risks of Treatment (SUPPORT)
 \citep{knaus1995support}.\footnote{The original data is available online from the Department of Biostatistics at Vanderbilt University: \url{https://biostat.app.vumc.org/wiki/Main/SupportDesc}}
@@
 regression is then fit for multiple values of the tuning parameter using
-the function \code{glmnet::cv.glmnet} and the binomial family. We supply
-both a matrix \code{y} containing the time and event variables, and a
-matrix \code{x} containing all other covariates. We apply the lasso
-penalty by setting \code{alpha = 1} and assign a \code{penalty.factor}
-(\(w_j\); cf.~Equation \ref{eq:penest}) of 0 to the time variable to
-ensure it is in the selected model. We compare our approach to both Cox
-regression, and lasso penalized Cox regression (fitted via the
-\pkg{glmnet} package and using the Cox family).
+the function \code{glmnet::cv.glmnet} and the binomial family. To
+\code{fitSmoothHazard.fit}, we supply both a matrix \code{y} containing
+the time and event variables, and a matrix \code{x} containing all other
+covariates. We apply the lasso penalty by setting \code{alpha = 1} and
+assign a \code{penalty.factor} (\(w_j\); cf.~Equation \ref{eq:penest})
+of 0 to the time variable to ensure it is in the selected model. We
+compare our approach to both Cox regression, and lasso penalized Cox
+regression (fitted via the \pkg{glmnet} package and using the Cox
+family).
 
@@
 
+Poisson regression can also be used to estimate the full hazard by
+discretizing time. However, this method requires user input on the
+number of intervals, or equivalently, on the cut points. This choice
+made by the user can have a significant impact on the model. Small
+intervals may result in many empty, non-informative bins. This may cause
+convergence issues for the Newton-Raphson procedure
+\citep{kalbfleisch2011statistical}. If the intervals are too wide, the
+nonlinear trends that are present in the hazard may be masked. Rather
+than discretizing time like in Poisson regression, case-base sampling
+provides a continuous-time approach to using GLMs for estimating hazard
+functions.
+
 As presented in Hanley \& Miettinen \citeyearpar{hanley2009fitting},
@@
 smooth survival curves, whereas Cox regression and Fine-Gray models
-produce stepwise CIFs and never explicitly model the hazard function.
+produce stepwise CIs and never explicitly model the hazard function.
 Qualitatively, we showed that by using splines in the linear predictor,
-all three models yielded similar curves. However, the smooth nature of
-the output of \pkg{casebase} provides a more intuitive interpretation
-for consumers of these predictions.
+all three models yielded similar curves.
 

### 2021-10-13 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.R

@@ -52,50 +52,99 @@ q4 <- qualitative_hcl(4, palette = "Dark 3")
 names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
 
 
-## ----erspc-data, eval = eval_cs1, echo = c(-1:-3)-----------------------------
+## ----erspc-data-mutate, eval = eval_cs1, echo = FALSE-------------------------
 data("ERSPC")
 ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
                        levels = c(0,1), 
                        labels = c("Control group", "Screening group"))
 
+
+## ----erspc-data, eval = eval_cs1----------------------------------------------
 pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
                                event = "DeadOfPrCa", exposure = "ScrArm")
 inherits(pt_object, "popTime")
 attr(pt_object, "exposure")
 
 
+## ----plot-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plot for the ERSPC dataset. \\textbf{A}: The gray area can be thought of as N = 159,893 infinitely thin horizontal rectangles ordered by length of follow-up. \\textbf{B}: The red points correspond to when death has occurred for any one of those infinitely thin rectangles. \\textbf{C}: To improve visibility, these red points are randomly redistributed along their respective x-coordinates, providing a visualization of incidence density. More events are observed at later follow-up times, motivating the use of non-constant hazard models. \\textbf{D}: The base series, a representative sample of the entire grey area, is represented by the green points."----
+# For increased visibility, remove some cases
+toRemove <- sample(which(ERSPC$DeadOfPrCa == 1), 
+                   round(length(which(ERSPC$DeadOfPrCa == 1))*0.7))
+# Create new dataset for illustration
+ERSPC_sub <- ERSPC
+ERSPC_sub$DeadOfPrCa[toRemove] <- 0
+
+a <- casebase::popTime(ERSPC_sub, time = "Follow.Up.Time",
+                       event = "DeadOfPrCa")
+
+plota <- emptyPlot <- plot(a, add.case.series = FALSE, 
+                           legend = FALSE) + 
+  scale_y_continuous(labels = label_number_si())
+
+plotb <- emptyPlot + geom_point(data = dplyr::filter(a, event == 1),
[…]

### 2021-10-13 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
load ggplot2 in setup.

@@
 library(casebase)
+library(ggplot2)
 library(colorspace)
@@
 # Define theme for plots
-paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
+paper_gg_theme <- ggplot2::theme_bw(base_size = 10) + ggplot2::theme(legend.position = "bottom")
 # Define colour palette for CS 3
-q4 <- qualitative_hcl(4, palette = "Dark 3")
+q4 <- colorspace::qualitative_hcl(4, palette = "Dark 3")
 names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
@@
 to instability in the Newton-Raphson procedure for estimation 
-[@kalbfleisch2011statistical]. On the other hand, a coarse grouping could potentially mask 
+[@kalbfleisch2011statistical, Section 4.8]. 
+On the other hand, a coarse grouping could potentially mask 
 nonlinear trends in the hazard function.

### 2021-10-15 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
cs1 redone

@@
 library(casebase)
+library(survival)
+library(splines)
 library(ggplot2)
@@
 library(prodlim)
-library(riskRegression)
-library(splines)
-library(survival)
+# library(riskRegression)
 library(tibble)
@@
 
-```{r erspc-data, eval = eval_cs1}
+```{r erspc-data, eval = eval_cs1, echo=TRUE}
 pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
@@
 For our first case study, we return to the ERSPC study and investigate the differences in risk between the control and screening arms. 
-We fit a flexible hazard by using a smooth function of time modeled with cubic B-splines.
+Previous re-analyses of these data suggest that the 20% reduction in prostate cancer death due to screening was an underestimate [@hanley2010mortality]. 
+The estimated 20% (from a proportional hazards model) did not account for the delay between screening and the time the effect is expected to be observed. 
+As a result, the null effects in years 1–7 masked the substantial reductions that began to appear from year 8 onward. 
+This motivates the use of a time-dependent hazard ratio which can easily be fit with the \pkg{casebase} package by including an interaction term with time in the model. 
+We fit a flexible hazard by using a smooth function of time modeled with a penalized cubic spline basis with 2 degrees of freedom (implemented in the \code{survival::pspline} function). 
 The model is fit using \code{fitSmoothHazard} with the familiar formula interface:
@@
 ```{r, eval = eval_cs1, echo = -1}
-set.seed(12)
-library(splines)
-fit <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
+set.seed(1953)
+fit <- fitSmoothHazard(DeadOfPrCa ~ pspline(Follow.Up.Time, df = 2) * ScrArm, 
                        data = ERSPC, ratio = 100)
@@
 
+Note that we did not have to specify any cut points, as would be the case with the \code{survSplit} function in the \pkg{survival} package. 
 The output object from \code{fitSmoothHazard} inherits from the \code{singleEventCB} and \code{glm} classes.
@@
 
+As noted in the Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models. 
+Therefore, we can easily test the significance of the spline term and its interaction with time:
+
+```{r erscp-compare, eval = eval_cs1, echo = TRUE}
+anova(fit, test = "LRT")
+```
+
+Similarly, to compare different models (e.g. time modeled linearly), we could compute Akaike's Information Criterion (AIC) for each model and compare them.
+
+### Time-dependent hazard ratios
+
+In what follows, the hazard ratio for a variable $X$ is defined as
+
+$$
+\frac{\lambda\left(t | X=x_1, \mathbf{Z}=\mathbf{z_1} ; \hat{\beta}\right)}{\lambda(t | X=x_0, \mathbf{Z}=\mathbf{z_0} ; \hat{\beta})}
+$$
+where $\lambda(t|\cdot;\hat{\beta})$ is the hazard rate as a function of the variable $t$ (which is usually time, but can be any other continuous variable), $x_1$ is the value of $X$ for the exposed group, $x_0$ is the value of $X$ for the unexposed group, $\mathbf{Z}$ are other covariates in the model which are equal to $\mathbf{z_1}$ in the exposed and $\mathbf{z_0}$ in the unexposed group, and $\hat{\beta}$ are the estimated regression coefficients.
+As indicated by the formula above, it is most instructive to plot the hazard ratio as a function of a variable $t$ only if there is an interaction between $t$ and $X$. 
+Otherwise, the resulting plot will simply be a horizontal line across time.
+
+The \code{plot} method with \code{type="hr"} for objects of class \code{singleEventCB} can be used to compute time-dependent hazard ratios and confidence intervals. 
+In Figure \ref{fig:interaction-ERSPC}, we show the estimated hazard ratio and 95% confidence interval for screening vs. control group as a function of time. 
+Note that we must specify the covariate profiles and times for the predicted hazards.
+
+```{r, echo = TRUE, eval = FALSE}
+new_time <- seq(1, 12, by  = 0.1)
+new_data <- data.frame(ScrArm = factor("Control group",
+                                         levels = c("Control group","Screening group")),
+                      Follow.Up.Time = new_time)
+plot(fit, type = "hr", newdata = new_data,
+     var = "ScrArm", xvar = "Follow.Up.Time", ci = T)
+```
+
+The plot shows that the effect of screening only becomes statistically apparent by year 7 and later. 
+The 25-60% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20% reported in the original article.
+
+```{r interaction-ERSPC, eval = eval_cs1, fig.width=8, fig.height=6, fig.cap='Estimated hazard ratio and 95\\% confidence interval for screening vs.
+control group as a function of time in the ERSPC dataset.
+Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic pspline basis of follow-up time and treatment arm.
+95\\% confidence intervals are calculated using the delta method.
+The plot shows that the effect of screening only begins to become statistically apparent by year 7.
+The 25-60\\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\\% reported in the original article.'}
+new_time <- seq(1, 12, by  = 0.1)
+new_data <- data.frame(ScrArm = factor("Control group",
+                                         levels = c("Control group","Screening group")),
+                      Follow.Up.Time = new_time)
+
+par(oma = c(0,1,0,8))
+tt <- plot(fit,
+     type = "hr",
+     newdata = new_data,
+     var = "ScrArm",
+     increment = 1,
+     xvar = "Follow.Up.Time",
+     ci = T,
+     xlab = "Follow-up time (years)",
+     ylab = "Death from prostate cancer hazard ratio",
+     ylim = c(0,1.50),
+     yaxt = 'n',
+     xaxt = "n",
+     rug = TRUE)
+
+axis(2, at = seq(0,1.50, by=0.25), las = 2)
+axis(1, at = seq(min(new_time),max(new_time), by=1))
+for( i in seq(0,1.50, by=0.25)) {
+  abline(h = i, lty = 1, col = "lightgrey")
+}
+for( i in seq(min(new_time),max(new_time), by=1)) {
+  abline(v = i, lty = 1, col = "lightgrey")
+}
+for(i in seq_along(seq(0, 0.75, by = 0.25))) {
+  mtext(paste0(seq(0, 0.75, by = 0.25)[i]*100,"%"), side = 4, las = 2, adj = 1, outer = TRUE, line = 2, at = c(0.63-0.11*(i-1)))
+}
+lines(tt$Follow.Up.Time, tt$hazard_ratio, lwd = 2, lty = 1)
+mtext("% reduction in\nprostate cancer\nmortality rate", side = 4, las = 2, outer = TRUE, line = -1, at = 0.7)
+```
+
+### Hazard functions
+
+Modeling the hazard function directly allows us to easily visualize it with the \code{plot} method and \code{type="hazard"} for objects of class \code{singleEventCB}. 
+We plot the hazard functions for both treatment arms in Figure \ref{fig:cs1hazard}. 
+The pattern we see is consistent with the population-time plot shown in Figure \ref{fig:plot-erspc-data} C, where more events are observed at later follow-up times. 
+The drop at the end can be explained by the fact that very few observations were followed for the entire 15 year period.
+
+```{r cs1hazard, eval=eval_cs1, echo=TRUE}
+plot(fit, type = "hazard",
+     hazard.params = list(xvar = "Follow.Up.Time",
+                          by = "ScrArm"))
+```
+
+### Absolute risk
+
 Next, the \code{absoluteRisk} function takes as input the \code{singleEventCB} object and returns a matrix where each column corresponds to the covariate profiles specified in the \code{newdata} argument, and each row corresponds to time points specified by the \code{time} argument:
 
-```{r, eval=eval_cs1, echo=echo_plot_code}
+```{r, eval=FALSE, echo=FALSE}
 new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
@@
 
-In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated risk function from \pkg{casebase} on the Cox model estimate.
-As noted in the Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models.
-Therefore, we can easily test the significance of the spline model:
-
-```{r erscp-compare, eval = eval_cs1, echo = FALSE}
-anova(fit, test = "LRT")
-```
-
-The null hypothesis is rejected in favor of the spline model.
-Similarly, to compare different models (e.g.
-time modeled linearly), we could compute Akaike's Information Criterion (AIC) for each model and compare them.
+```{r, eval=eval_cs1, echo=TRUE}
+new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
+new_time <- seq(0,14,0.1)
+risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
+head(risk)
+```
+
+<!--In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated risk function from \pkg{casebase} on the Cox model estimate. -->
 
@@
 
-```{r erspc-cox-cif, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk functions for control and screening groups in the ERSPC data.
+```{r erspc-cox-cif, eval = FALSE, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk functions for control and screening groups in the ERSPC data.
 We plot the estimate from the Cox model using \\code{survival::survfit} (solid line) and the estimate from case-base sampling (dashed line).
@@
 
-Next, we can compute confidence intervals for the risk function using the method \code{confint.absRiskCB}.
-
-```{r eval = eval_cs1, echo = -1}
-new_time <- c(0, 1)
-risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
+We can subsequently compute confidence intervals for the risk function using the method \code{confint.absRiskCB}:
+
+```{r eval = eval_cs1, echo = TRUE}
 conf_ints <- confint(risk, fit)
-conf_ints
+head(conf_ints)
 ```
@@
 In Figure \ref{fig:erspc-cif-conf}, we see the 95% confidence bands around the estimates. 
+We also overlay the Kaplan-Meier curves as a reference.
 
 ```{r erspc-cif-conf, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\\% confidence bands."}
-# confidence interval
-tt <- confint(risk_mp, fit)
+
+tt <- conf_ints
 
 # plot absrisk and 95% CI
-new_time <- seq(0,14,0.1)
 i.backw <- order(new_time, decreasing = TRUE)
@@
 
-Previous re-analyses of these data suggest that the 20% reduction in prostate cancer death due to screening was an underestimate [@hanley2010mortality].
-The estimated 20% (from a proportional hazards model) did not account for the delay between screening and the time the effect is expected to be observed.
-As a result, the null effects in years 1–7 masked the substantial reductions that began to appear from year 8 onward.
-This motivates the use of a time-dependent hazard ratio which can easily be fit with the \pkg{casebase} package by including an interaction term with time in the model:
-
-```{r, echo=-1, eval = eval_cs1}
-set.seed(12)
-fit_inter <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) * ScrArm, 
-                             data = ERSPC)
-```
-
-Note that we did not have to specify any cut points, as would be the case with the \code{survSplit} function in the \pkg{survival} package.
-In Figure \ref{fig:interaction-ERSPC}, we have the estimated hazard ratio and 95% confidence interval for screening vs.
-control group as a function of time using the \code{plot} method for objects of class \code{singleEventCB}:
-
-```{r, echo = TRUE, eval = FALSE}
-plot(fit_inter, type = "hr", newdata = new_data,
-     var = "ScrArm", xvar = "Follow.Up.Time", ci = TRUE)
-```
-
-The plot shows that the effect of screening only becomes statistically apparent by year 7 and later.
-The 25-60% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20% reported in the original article.
-
-```{r interaction-ERSPC, eval = eval_cs1, fig.width=8, fig.height=6, fig.cap='Estimated hazard ratio and 95\\% confidence interval for screening vs.
-control group as a function of time in the ERSPC dataset.
-Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic B-spline basis of follow-up time and treatment arm.
-95\\% confidence intervals are calculated using the delta method.
-The plot shows that the effect of screening only begins to become statistically apparent by year 7.
-The 25-60\\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\\% reported in the original article.'}
-new_time <- seq(1, 12, by  = 0.1)
-new_data <- data.frame(ScrArm = factor("Control group",
-                                         levels = c("Control group","Screening group")),
-                      Follow.Up.Time = new_time)
-
-par(oma = c(0,1,0,8))
-tt <- plot(fit_inter,
-     type = "hr",
-     newdata = new_data,
-     var = "ScrArm",
-     increment = 1,
-     xvar = "Follow.Up.Time",
-     ci = T,
-     xlab = "Follow-up time (years)",
-     ylab = "Death from prostate cancer hazard ratio",
-     ylim = c(0,1.50),
-     yaxt = 'n',
-     xaxt = "n",
-     rug = TRUE)
-
-axis(2, at = seq(0,1.50, by=0.25), las = 2)
-axis(1, at = seq(min(new_time),max(new_time), by=1))
-for( i in seq(0,1.50, by=0.25)) {
-  abline(h = i, lty = 1, col = "lightgrey")
-}
-for( i in seq(min(new_time),max(new_time), by=1)) {
-  abline(v = i, lty = 1, col = "lightgrey")
-}
-for(i in seq_along(seq(0, 0.75, by = 0.25))) {
-  mtext(paste0(seq(0, 0.75, by = 0.25)[i]*100,"%"), side = 4, las = 2, adj = 1, outer = TRUE, line = 2, at = c(0.63-0.11*(i-1)))
-}
-lines(tt$Follow.Up.Time, tt$hazard_ratio, lwd = 2, lty = 1)
-mtext("% reduction in\nprostate cancer\nmortality rate", side = 4, las = 2, outer = TRUE, line = -1, at = 0.7)
+```{r}
+knitr::knit_exit()
 ```
@@
 
-We want to compare our hazard ratio estimates to that obtained from a Cox regression.
+We want to compare our hazard ratio estimates to that obtained from a Cox regression (using the \pkg{survival} package version 3.2-13).
 
@@
 ```{r supportData, eval = eval_cs3}
+library(riskRegression)
 scaleFUNy <- function(x) sprintf("%.2f", x)

### 2021-10-15 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
load ggplot2 in setup.
cs1 redone

@@
 leads to instability in the Newton-Raphson procedure for estimation
-\citep{kalbfleisch2011statistical}. On the other hand, a coarse grouping
-could potentially mask nonlinear trends in the hazard function.
+\citep[Section 4.8]{kalbfleisch2011statistical}. On the other hand, a
+coarse grouping could potentially mask nonlinear trends in the hazard
+function.
 
@@
 \begin{Schunk}
+\begin{Sinput}
+pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
+                               event = "DeadOfPrCa", exposure = "ScrArm")
+inherits(pt_object, "popTime")
+\end{Sinput}
 \begin{Soutput}
@@
 \end{Soutput}
+\begin{Sinput}
+attr(pt_object, "exposure")
+\end{Sinput}
 \begin{Soutput}
@@
 For our first case study, we return to the ERSPC study and investigate
-the differences in risk between the control and screening arms. We fit a
-flexible hazard by using a smooth function of time modeled with cubic
-B-splines. The model is fit using \code{fitSmoothHazard} with the
-familiar formula interface:
+the differences in risk between the control and screening arms. Previous
+re-analyses of these data suggest that the 20\% reduction in prostate
+cancer death due to screening was an underestimate
+\citep{hanley2010mortality}. The estimated 20\% (from a proportional
+hazards model) did not account for the delay between screening and the
+time the effect is expected to be observed. As a result, the null
+effects in years 1--7 masked the substantial reductions that began to
+appear from year 8 onward. This motivates the use of a time-dependent
+hazard ratio which can easily be fit with the \pkg{casebase} package by
+including an interaction term with time in the model. We fit a flexible
+hazard by using a smooth function of time modeled with a penalised cubic
+spline basis with 2 degrees of freedom. The model is fit using
+\code{fitSmoothHazard} with the familiar formula interface:
 
@@
 \begin{Sinput}
-library(splines)
-fit <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
+library(survival) # for the pspline function
+fit <- fitSmoothHazard(DeadOfPrCa ~ pspline(Follow.Up.Time, df = 2) * ScrArm, 
                        data = ERSPC, ratio = 100)
@@
 #> Call:
-#> fitSmoothHazard(formula = DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
-#>     data = ERSPC, ratio = 100)
+#> fitSmoothHazard(formula = DeadOfPrCa ~ pspline(Follow.Up.Time, 
+#>     df = 2) * ScrArm, data = ERSPC, ratio = 100)
 #> 
@@
 #>    Min      1Q  Median      3Q     Max  
-#> -0.403  -0.172  -0.135  -0.085   3.801  
+#> -0.477  -0.164  -0.140  -0.084   3.882  
 #> 
 #> Coefficients:
-#>                       Estimate Std. Error z value Pr(>|z|)    
-#> (Intercept)           -10.2533     0.3168  -32.36  < 2e-16 ***
-#> bs(Follow.Up.Time)1     4.2544     0.7976    5.33  9.6e-08 ***
-#> bs(Follow.Up.Time)2     2.1371     0.4720    4.53  5.9e-06 ***
-#> bs(Follow.Up.Time)3     4.5339     0.6889    6.58  4.7e-11 ***
-#> ScrArmScreening group  -0.2481     0.0886   -2.80   0.0051 ** 
-#> ---
-#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#>                                                        Estimate Std. Error
+#> (Intercept)                                             -12.613      9.756
+#> pspline(Follow.Up.Time, df = 2)1                          1.320     10.705
+#> pspline(Follow.Up.Time, df = 2)2                          5.190      9.526
+#> pspline(Follow.Up.Time, df = 2)3                          4.481      9.866
+#> pspline(Follow.Up.Time, df = 2)4                          6.060      9.690
+#> pspline(Follow.Up.Time, df = 2)5                          5.231      9.851
+#> pspline(Follow.Up.Time, df = 2)6                          9.176      9.749
+#> pspline(Follow.Up.Time, df = 2)7                         -0.434     19.572
+#> ScrArmScreening group                                     7.900     13.050
+#> pspline(Follow.Up.Time, df = 2)1:ScrArmScreening group   -7.962     14.507
+#> pspline(Follow.Up.Time, df = 2)2:ScrArmScreening group   -8.351     12.690
+#> pspline(Follow.Up.Time, df = 2)3:ScrArmScreening group   -7.787     13.224
+#> pspline(Follow.Up.Time, df = 2)4:ScrArmScreening group   -8.098     12.942
+#> pspline(Follow.Up.Time, df = 2)5:ScrArmScreening group   -9.389     13.223
+#> pspline(Follow.Up.Time, df = 2)6:ScrArmScreening group   -7.385     13.114
+#> pspline(Follow.Up.Time, df = 2)7:ScrArmScreening group  -17.380     30.806
+#>                                                        z value Pr(>|z|)
+#> (Intercept)                                              -1.29     0.20
+#> pspline(Follow.Up.Time, df = 2)1                          0.12     0.90
+#> pspline(Follow.Up.Time, df = 2)2                          0.54     0.59
+#> pspline(Follow.Up.Time, df = 2)3                          0.45     0.65
+#> pspline(Follow.Up.Time, df = 2)4                          0.63     0.53
+#> pspline(Follow.Up.Time, df = 2)5                          0.53     0.60
+#> pspline(Follow.Up.Time, df = 2)6                          0.94     0.35
+#> pspline(Follow.Up.Time, df = 2)7                         -0.02     0.98
+#> ScrArmScreening group                                     0.61     0.54
+#> pspline(Follow.Up.Time, df = 2)1:ScrArmScreening group   -0.55     0.58
+#> pspline(Follow.Up.Time, df = 2)2:ScrArmScreening group   -0.66     0.51
+#> pspline(Follow.Up.Time, df = 2)3:ScrArmScreening group   -0.59     0.56
+#> pspline(Follow.Up.Time, df = 2)4:ScrArmScreening group   -0.63     0.53
+#> pspline(Follow.Up.Time, df = 2)5:ScrArmScreening group   -0.71     0.48
+#> pspline(Follow.Up.Time, df = 2)6:ScrArmScreening group   -0.56     0.57
+#> pspline(Follow.Up.Time, df = 2)7:ScrArmScreening group   -0.56     0.57
 #> 
@@
 #>     Null deviance: 6059.0  on 54539  degrees of freedom
-#> Residual deviance: 5787.6  on 54535  degrees of freedom
-#> AIC: 5798
-#> 
-#> Number of Fisher Scoring iterations: 8
+#> Residual deviance: 5772.7  on 54524  degrees of freedom
+#> AIC: 5805
+#> 
+#> Number of Fisher Scoring iterations: 9
 \end{Soutput}
@@
 
-Next, the \code{absoluteRisk} function takes as input the
-\code{singleEventCB} object and returns a matrix where each column
-corresponds to the covariate profiles specified in the \code{newdata}
-argument, and each row corresponds to time points specified by the
-\code{time} argument:
-
-\begin{Schunk}
-\begin{Sinput}
-new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
-new_time <- c(0, 1)
-
-absoluteRisk(fit, time = new_time, newdata = new_data)
-\end{Sinput}
-\begin{Soutput}
-#>  time                
-#>     0 0.0e+00 0.0e+00
-#>     1 5.4e-05 4.2e-05
-\end{Soutput}
-\end{Schunk}
-
-In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated risk
-function from \pkg{casebase} on the Cox model estimate. As noted in the
-Theoretical Details section, the usual asymptotic results hold for
-likelihood ratio tests built using case-base sampling models. Therefore,
-we can easily test the significance of the spline model:
+As noted in the Theoretical Details section, the usual asymptotic
+results hold for likelihood ratio tests built using case-base sampling
+models. Therefore, we can easily test the significance of the spline
+model:
 
@@
 #> 
-#>                    Df Deviance Resid. Df Resid. Dev Pr(>Chi)    
+#>                                        Df Deviance Resid. Df Resid. Dev
 #> NULL                                                   54539       6059
-#> bs(Follow.Up.Time)  3      263     54536       5796   <2e-16 ***
-#> ScrArm              1        8     54535       5788   0.0048 ** 
+#> pspline(Follow.Up.Time, df = 2)         7    269.1     54532       5790
+#> ScrArm                                  1      8.0     54531       5782
+#> pspline(Follow.Up.Time, df = 2):ScrArm  7      9.1     54524       5773
+#>                                        Pr(>Chi)    
+#> NULL                                               
+#> pspline(Follow.Up.Time, df = 2)          <2e-16 ***
+#> ScrArm                                   0.0046 ** 
+#> pspline(Follow.Up.Time, df = 2):ScrArm   0.2454    
 #> ---
@@
 
-\begin{Schunk}
-\begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cox-cif-1} \caption{Risk functions for control and screening groups in the ERSPC data.
-We plot the estimate from the Cox model using \code{survival::survfit} (solid line) and the estimate from case-base sampling (dashed line).
-We used a cubic B-spline expansion of time.}\label{fig:erspc-cox-cif}
-\end{figure}
-\end{Schunk}
-
-Next, we can compute confidence intervals for the risk function using
-the method \code{confint.absRiskCB}.
-
-\begin{Schunk}
-\begin{Sinput}
-risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
-conf_ints <- confint(risk, fit)
-conf_ints
-\end{Sinput}
-\begin{Soutput}
-#>   time estimate conf.low conf.high        cov_prof
-#> 1    0  0.0e+00  0.0e+00   0.0e+00   Control group
-#> 2    1  5.4e-05  3.5e-05   9.3e-05   Control group
-#> 3    0  0.0e+00  0.0e+00   0.0e+00 Screening group
-#> 4    1  4.2e-05  2.7e-05   7.3e-05 Screening group
-\end{Soutput}
-\end{Schunk}
-
-These point-wise confidence intervals are constructed using parametric
-bootstrap; the validity of this approach relies on the asymptotic
-normality of the MLE. In Figure \ref{fig:erspc-cif-conf}, we see the
-95\% confidence bands around the estimates.
-
-\begin{Schunk}
-\begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cif-conf-1} \caption[Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands]{Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands.}\label{fig:erspc-cif-conf}
-\end{figure}
-\end{Schunk}
-
-Previous re-analyses of these data suggest that the 20\% reduction in
-prostate cancer death due to screening was an underestimate
-\citep{hanley2010mortality}. The estimated 20\% (from a proportional
-hazards model) did not account for the delay between screening and the
-time the effect is expected to be observed. As a result, the null
-effects in years 1--7 masked the substantial reductions that began to
-appear from year 8 onward. This motivates the use of a time-dependent
-hazard ratio which can easily be fit with the \pkg{casebase} package by
-including an interaction term with time in the model:
-
-\begin{Schunk}
-\begin{Sinput}
-fit_inter <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) * ScrArm, 
-                             data = ERSPC)
-\end{Sinput}
-\end{Schunk}
+\hypertarget{time-dependent-hazard-ratios}{%
+\subsubsection{Time-dependent hazard
+ratios}\label{time-dependent-hazard-ratios}}
 
@@
 \begin{Sinput}
-plot(fit_inter, type = "hr", newdata = new_data,
-     var = "ScrArm", xvar = "Follow.Up.Time", ci = TRUE)
+new_time <- seq(1, 12, by  = 0.1)
+new_data <- data.frame(ScrArm = factor("Control group",
+                                         levels = c("Control group","Screening group")),
+                      Follow.Up.Time = new_time)
+plot(fit, type = "hr", newdata = new_data,
+     var = "ScrArm", xvar = "Follow.Up.Time", ci = F)
 \end{Sinput}
@@
 
-\hypertarget{case-study-2competing-risk-analysis}{%
-\subsection{Case study 2---Competing risk
-analysis}\label{case-study-2competing-risk-analysis}}
-
-In this case study, we show how case-base sampling can be used in the
-context of a competing-risk analysis. For illustrative purposes, we use
-the same data that was used in Scrucca \emph{et al}
-\citeyearpar{scrucca2010regression}. The data was downloaded from the
-first author's website, and it is now available as part of the
-\pkg{casebase} package.
-
-The data contains information on 177 patients who received a stem-cell
-transplant for acute leukemia. The event of interest is relapse, but
-other competing causes (e.g.~death, progression, graft failure,
-graft-versus-host disease) were also recorded. Several covariates were
-captured at baseline: sex, disease type (acute lymphoblastic or
-myeloblastic leukemia, abbreviated as ALL and AML, respectively),
-disease phase at transplant (Relapse, CR1, CR2, CR3), source of stem
-cells (bone marrow and peripheral blood, coded as BM+PB, or only
-peripheral blood, coded as PB), and age.
-
-First, we look at a population-time plot to visualize the incidence
-density of both relapse and the competing events. In Figure
-\ref{fig:compPop}, failure times are highlighted on the plot using red
-dots for the event of interest (panel A) and blue dots for competing
-events (panel B). In both panels, we see evidence of a non-constant
-hazard function: the density of points is larger at the beginning of
-follow-up than at the end.
-
-\begin{Schunk}
-\begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./compPop-1} \caption[Population-time plot for the stem-cell transplant study with both relapse and competing events]{Population-time plot for the stem-cell transplant study with both relapse and competing events.}\label{fig:compPop}
-\end{figure}
-\end{Schunk}
-
-Our main objective is to compute the cumulative incidence of relapse for
-a given set of covariates. We start by fitting a smooth hazard to the
-data using a linear term for time:
-
-\begin{Schunk}
-\begin{Sinput}
-model_cb <- fitSmoothHazard(
-  Status ~ ftime + Sex + D + Phase + Source + Age,
-  data = bmtcrr,
-  ratio = 100,
-  time = "ftime"
-)
-\end{Sinput}
-\end{Schunk}
-
-We want to compare our hazard ratio estimates to that obtained from a
-Cox regression.
-
-\begin{Schunk}
-\begin{Sinput}
-library(survival)
-# Prepare data for coxph
-bmtcrr_cox <- transform(bmtcrr, 
-                        id = seq_len(nrow(bmtcrr)),
-                        Status = factor(Status))
-
-model_cox <- coxph(Surv(ftime, Status) ~ Sex + D + Phase + Source + Age,
-                   data = bmtcrr_cox, id = id)
-\end{Sinput}
-\end{Schunk}
-
-\begin{Schunk}
-\begin{table}
-
-\caption{\label{tab:bmtcrr-cis}Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.}
-\centering
-\begin{tabular}[t]{lrlrl}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
-Covariates & HR & 95\% Conf. & HR & 95\% Conf.\\
-\midrule
-Sex & 0.74 & (0.43, 1.3) & 0.68 & (0.39, 1.21)\\
-Disease & 0.52 & (0.29, 0.94) & 0.52 & (0.28, 0.94)\\
-Phase (CR2 vs. CR1) & 1.20 & (0.48, 3) & 1.21 & (0.47, 3.09)\\
-Phase (CR3 vs. CR1) & 1.59 & (0.41, 6.17) & 1.67 & (0.46, 6.08)\\
-Phase (Relapse vs. CR1) & 4.09 & (1.9, 8.81) & 4.55 & (1.98, 10.46)\\
-\addlinespace
-Source & 1.70 & (0.55, 5.23) & 1.46 & (0.47, 4.53)\\
-Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\end{Schunk}
-
-From the fit object, we can extract both the hazard ratios and their
-corresponding confidence intervals. These quantities appear in Table
-\ref{tab:bmtcrr-cis}. As we can see, the type of disease corresponds to
-a significant hazard ratio: the hazard for AML is 48\% lower than that
-for ALL. Moreover, being in relapse at transplant is associated with a
-hazard ratio of 4.09 when compared to CR1.
-
-Given the estimate of the hazard functions obtained using case-base
-sampling, we can compute the cumulative incidence curve for a fixed
-covariate profile. We perform this computation for a 35 year old woman
-who received a stem-cell transplant from peripheral blood at relapse. We
-compare the absolute risk curve for such a woman with ALL with that for
-a similar woman with AML.
-
-Next, we compare our estimates to that obtained from a corresponding
-Fine-Gray model \citeyearpar{fine1999proportional}. The Fine-Gray model
-is a semiparametric model for the cause-specific \emph{subdistribution
-hazard}, i.e.~the function \(d_j(t)\) such that
-\[CI_j(t) = 1 - \exp\left( - \int_0^t d_j(u) \textrm{d}u \right),\]
-where \(CI_j(t)\) is the cause-specific CI. The Fine-Gray model allows
-to directly assess the effect of a covariate on the subdistribution
-hazard, as opposed to the cause-specific hazard. For the computation, we
-use the \CRANpkg{timereg} package \citep{timereg}:
-
-\begin{Schunk}
-\begin{Sinput}
-library(timereg)
-model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) +
-                        const(Phase) + const(Source) + const(Age),
-                      data = bmtcrr, cause = 1, model = "fg")
-
-# Estimate CI curve
-risk_fg <- predict(model_fg, newdata, times = time_points)
-\end{Sinput}
-\end{Schunk}
-
-We can also estimate the CI for relapse using the Cox model and the
-Aalen-Johansen estimator:
-
-\begin{Schunk}
-\begin{Sinput}
-# Estimate absolute risk curve
-risk_cox <- survfit(model_cox, newdata = newdata)
-\end{Sinput}
-\end{Schunk}
-
-\begin{Schunk}
-\begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./bmtcrr-risk-1} \caption{\label{fig:compAbsrisk} Cumulative Incidence curve for a fixed covariate profile and the two disease groups.
-The estimate obtained from case-base sampling is compared to the Fine-Gray and Aalen-Johansen estimates.}\label{fig:bmtcrr-risk}
-\end{figure}
-\end{Schunk}
-
-Figure \ref{fig:compAbsrisk} shows the CI curves for all three models.
-As we can see, all three approaches agree quite well for AML; however,
-for ALL, there seems to be a difference of about 5\% between the
-Fine-Gray curve and the curves estimated using case-base sampling and
-Cox regression. This difference does not appear to be significant: the
-curve from case-base sampling is contained within a 95\% confidence band
-around the Fine-Gray absolute risk curve (figure not shown).
-
-\hypertarget{case-study-3variable-selection}{%
-\subsection{Case study 3---Variable
-selection}\label{case-study-3variable-selection}}
-
-For the third case study, we show how \pkg{casebase} can also be used
-for variable selection through regularized estimation of the hazard
-function as given by Equation \ref{eq:penest}. We note that this is
-different than the semiparametric model Coxnet, which regularizes the
-Cox partial likelihood. To illustrate this functionality, we use the
-dataset from the Study to Understand Prognoses Preferences Outcomes and
-Risks of Treatment (SUPPORT)
-\citep{knaus1995support}.\footnote{The original data is available online from the Department of Biostatistics at Vanderbilt University: \url{https://biostat.app.vumc.org/wiki/Main/SupportDesc}}
-The SUPPORT dataset tracks death in five American hospitals within
-individuals who are considered seriously ill.~The cleaned and imputed
-data consists of 9104 observations and 30 variables, and it is available
-as part of the \pkg{casebase} package. In the comparisons below, all
-covariates except \code{sps} and \code{aps} were used. These two
-variables correspond to scores for predicting the outcome that were
-developed as part of the original study. For more information about this
-dataset, the reader is encouraged to look at the documentation in our
-package.
-
-For our penalized case-base model, we opt for the natural log of time
-which corresponds to a Weibull distribution. For fitting the penalized
-hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface
-to the \code{fitSmoothHazard} function. The \code{fitSmoothHazard} and
-\code{fitSmoothHazard.fit} functions sample the case and base series,
-calculate the required offset, and transform the data to match the
-expected input of the \pkg{glmnet} package. The penalized logistic
-regression is then fit for multiple values of the tuning parameter using
-the function \code{glmnet::cv.glmnet} and the binomial family. To
-\code{fitSmoothHazard.fit}, we supply both a matrix \code{y} containing
-the time and event variables, and a matrix \code{x} containing all other
-covariates. We apply the lasso penalty by setting \code{alpha = 1} and
-assign a \code{penalty.factor} (\(w_j\); cf.~Equation \ref{eq:penest})
-of 0 to the time variable to ensure it is in the selected model. We
-compare our approach to both Cox regression, and lasso penalized Cox
-regression (fitted via the \pkg{glmnet} package and using the Cox
-family).
-
-To compare the performance of our models, we split the data into 95\%
-training and 5\% test sets. To assess both discrimination and
-calibration, we use a time-dependent version of the classical Brier
-score that is adjusted for censoring \citep{graf1999ass}. The Brier
-score can be used with both parametric and semi-parametric models. We
-use the \pkg{riskRegression} package to compute these scores for all
-models.
-
-\begin{Schunk}
-\begin{Sinput}
-# Create matrices for inputs
-x <- model.matrix(death ~ . - d.time - aps - sps, 
-                  data = train)[, -c(1)] # Remove intercept
-y <- data.matrix(subset(train, select = c(d.time, death)))
-
-# Regularized logistic regression to estimate hazard
-pen_cb <- casebase::fitSmoothHazard.fit(x, y,
-  family = "glmnet",
-  time = "d.time", event = "death",
-  formula_time = ~ log(d.time), alpha = 1,
-  ratio = 10, standardize = TRUE,
-  penalty.factor = c(0, rep(1, ncol(x)))
-)
-\end{Sinput}
-\end{Schunk}
-
-In Figure \ref{fig:cs3lolliPlot}, we show the coefficient estimates for
-covariates that we selected by both penalized Cox and penalized
-case-base. We note that both penalized approaches produce similar
-results. We can also clearly see the shrinkage effect owing to the
-\(\ell_1\) penalty.
-
-\begin{Schunk}
-\begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./coefplots-1} \caption{\label{fig:cs3lolliPlot} Coefficient estimates from the Cox model (Cox), penalized Cox model using the \pkg{glmnet} package (Pen.
-Cox), and our approach using penalized case-base sampling (Pen.
-CB).
-Only the covariates that were selected by both penalized approaches are shown.
-The shrinkage of the coefficient estimates for Pen.
-Cox and Pen.
-CB occurs due to the $\ell_1$ penalty.}\label{fig:coefplots}
-\end{figure}
-\end{Schunk}
-
-We then compare the risk estimation over the test set. The probabilities
-over time for each observation are averaged, resulting in the absolute
-risk curves shown in Figure \ref{fig:cs3FinalBrier} (A). We can see some
-minimal differences between the three models, with the Kaplan-Meier
-giving the lowest estimates across follow-up-time. Note that the
-apparent smoothness of the Cox and penalized Cox curves is due to the
-large number of observations in the training set, which is used to
-derive the Breslow estimate of the baseline hazard. As described above,
-we compare the performance between the models by computing the Brier
-scores over time. In Figure \ref{fig:cs3FinalBrier} (B), we can see that
-the Brier score is larger for the Kaplan-Meier estimate than for the
-other three models. On the other hand, the differences between these
-three models are minimal.
-
-\begin{Schunk}
-\begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./riskregressionBrier-1} \caption{\label{fig:cs3FinalBrier} Comparison of Cox regression (Cox), penalized Cox regression (Pen.
-Cox), penalized case-base sampling estimation (Pen.
-CB), and Kaplan-Meier (K-M).
-(A)  Probability of death as a function of follow-up time.
-(B) Brier score as a function of follow-up time, where a lower score corresponds to better performace.}\label{fig:riskregressionBrier}
-\end{figure}
-\end{Schunk}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-In this article, we presented the R package \pkg{casebase}, which
-provides functions for fitting smooth parametric hazards and estimating
-risk functions using case-base sampling. Our package also provides
-several functions to produce graphical summaries of the data and the
-results. We outlined the theoretical underpinnings of the approach, we
-provided details about our implementation, and we illustrated the merits
-of the case-base framework and the package through three case studies.
-
-As a methodological framework, case-base sampling is very flexible. Some
-of this flexibility has been explored before in the literature: for
-example, Saarela and Hanley \citeyearpar{saarela2015case} used case-base
-sampling to model a time-dependent exposure variable in a vaccine safety
-study. As another example, Saarela and Arjas
-\citeyearpar{saarela2015non} combined case-base sampling and a Bayesian
-non-parametric framework to compute individualized risk assessments for
-chronic diseases. In the case studies above, we further explored this
-flexibility along two fronts. On the one hand, we showed how splines
-could be used as part of the linear predictor to model the effect of
-time on the hazard. This strategy yielded estimates of the survival
-function that were qualitatively similar to semiparametric estimates
-derived from Cox regression; however, case-base sampling led to
-estimates of the survival function that \emph{vary smoothly in time} and
-are thus easier to interpret. On the other hand, we also displayed the
-flexibility of case-base sampling by showing how it could be combined
-with penalized logistic regression to perform variable selection.
-Furthermore, the second case study showed how case-base sampling can be
-applied to competing risks settings. It should be noted that the three
-case studies presented above only considered covariates that were fixed
-at baseline. In one of the package vignettes, we use the Stanford Heart
-Transplant data \citep[\citet{crowley1977covariance}]{clark1971cardiac}
-to show how case-base sampling can also be used in the context of
-time-dependent exposure. In this study, the exposure period was defined
-as the week following vaccination. Hence, the main covariate of
-interest, i.e.~exposure to the vaccine, was changing over time. In this
-context, case-base sampling offers an efficient alternative to nested
-case-control designs or self-matching.
-
-Even though we did not illustrate it in this article, case-base sampling
-can also be combined with the framework of \emph{generalized additive
-models}. This functionality has already been implemented in the package.
-Similarly, case-base sampling can be combined with quasi-likelihood
-estimation to fit survival models that can account for the presence of
-over-dispersion. All of these examples illustrate how the case-base
-sampling framework in general, and the package \pkg{casebase} in
-particular, allows the user to fit a broad and flexible family of
-survival functions.
-
-Poisson regression can also be used to estimate the full hazard by
-discretizing time. However, this method requires user input on the
-number of intervals, or equivalently, on the cut points. This choice
-made by the user can have a significant impact on the model. Small
-intervals may result in many empty, non-informative bins. This may cause
-convergence issues for the Newton-Raphson procedure
-\citep{kalbfleisch2011statistical}. If the intervals are too wide, the
-nonlinear trends that are present in the hazard may be masked. Rather
-than discretizing time like in Poisson regression, case-base sampling
-provides a continuous-time approach to using GLMs for estimating hazard
-functions.
-
-As presented in Hanley \& Miettinen \citeyearpar{hanley2009fitting},
-case-base sampling is comprised of three steps: 1) sampling a case
-series and a base series from the study; 2) fit the log-hazard as a
-linear function of predictors (including time); and 3) use the fitted
-hazard to estimate the risk function. Accordingly, our package provides
-functions for each step. Moreover, the simple interface of the
-\code{fittingSmoothHazard} function resembles the \code{glm} interface.
-This interface should look familiar to new users. Our modular approach
-also provides a convenient way to extend our package for new sampling or
-fitting strategies.
-
-In the case studies above, we compared the performance of case-base
-sampling with that of Cox regression and Fine-Gray models. In terms of
-function interface, \pkg{casebase} uses a formula interface that is
-closer to that of \code{glm}, in that the event variable is the only
-variable appearing on the left-hand side of the formula. By contrast,
-both \code{survival::coxph} and \code{timereg::comp.risk} use arrays
-that capture both the event type and time. Both approaches to modeling
-yield user-friendly code. However, in terms of output, both approaches
-differ significantly. Case-base sampling produces smooth hazards and
-smooth survival curves, whereas Cox regression and Fine-Gray models
-produce stepwise CIs and never explicitly model the hazard function.
-Qualitatively, we showed that by using splines in the linear predictor,
-all three models yielded similar curves.
-
-Our choice of modeling the log-hazard as a linear function of covariates
-allows us to develop a simple computational scheme for estimation.
-However, as a downside, it does not allow us to model location and scale
-parameters separately like the package \pkg{flexsurv}. For example, if
-we look at the Weibull distribution as parametrised in
-\code{stats::pweibull}, the log-hazard function is given by
-\[ \log \lambda(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,\]
-where \(\alpha,\beta\) are shape and scale parameters, respectively.
-Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} also allows
-the user to model the scale parameter as a function of covariates. Of
-course, this added flexibility comes at the cost of interpretability: by
-modeling the log-hazard directly, the parameter estimates from
-\pkg{casebase} can be interpreted as estimates of log-hazard ratios. To
-improve the flexibility of \pkg{casebase} at capturing the scale of a
-parametric family, we could replace the logistic regression with its
-quasi-likelihood counterpart and therefore model over- and
-under-dispersion with respect to the logistic likelihood. We defer the
-study of the properties and performance of such a model to a future
-article.
-
-Future work will look at some of the methodological extensions of
-case-base sampling. First, to assess the quality of the model fit, we
-would like to study the properties of the residuals (e.g.~Cox-Snell,
-martingale). More work needs to be done to understand these residuals in
-the context of the partial likelihood underlying case-base sampling. The
-resulting diagnostic tools could then be integrated in this package.
-Also, we are interested in extending case-base sampling to account for
-interval censoring. This type of censoring is very common in
-longitudinal studies, and many packages (e.g.~\pkg{SmoothHazard},
-\pkg{survival} and \pkg{rstpm2}) provide functions to account for it.
-Again, we hope to include any resulting methodology as part of this
-package.
-
-In future versions of the package, we also want to increase the
-complement of diagnostic and inferential tools that are currently
-available. For example, we would like to include more functions to
-compute calibration and discrimination statistics (e.g.~AUC) for our
-models. Saarela and Arjas \citeyearpar{saarela2015non} also describe how
-to obtain a posterior distribution for the AUC from their model. Their
-approach could potentially be included in \pkg{casebase}. Finally, we
-want to provide more flexibility in how the case-base sampling is
-performed. This could be achieved by adding a \code{hazard} argument to
-the function \code{sampleCaseBase}. In this way, users could specify
-their own sampling mechanism. For example, they could provide a hazard
-that gives sampling probabilities that are proportional to the
-cardiovascular disease event rate given by the Framingham score
-\citep{saarela2015non}.
-
-In conclusion, we presented the R package \pkg{casebase} which
-implements case-base sampling for fitting parametric survival models and
-for estimating smooth survival functions using the framework of
-generalized linear models. We strongly believe that its flexibility and
-its foundation on the familiar logistic regression model will make it
-appealing to new and established practitioners. The \pkg{casebase}
-package is freely available from the Comprehensive R Archive Network at
-\url{https://cran.r-project.org/package=casebase}. Interested users can
-visit \url{http://sahirbhatnagar.com/casebase/} for detailed package
-documentation and vignettes.
-
-\hypertarget{acknowledgments}{%
-\section{Acknowledgments}\label{acknowledgments}}
-
-We would like to thank the anonymous reviewers for their insightful
-comments and criticisms. We would also like to thank Yi Yang for helpful
-discussions on penalized regression models. \mbox{Bhatnagar} gratefully
-acknowledges funding via a Discovery Grant from the Natural Sciences and
-Engineering Research Council of Canada (NSERC).
-
 \bibliography{bhatnagar-turgeon-islam-hanley-saarela.bib}
@@
 %
-\url{http://sahirbhatnagar.com/}\\%
-%
-\email{sahir.bhatnagar@mcgill.ca}%
+\url{http://sahirbhatnagar.com/}%
+%
+\\\email{sahir.bhatnagar@mcgill.ca}
 }
@@
 %
-\url{https://maxturgeon.ca/}\\%
-%
-\email{max.turgeon@umanitoba.ca}%
+\url{https://maxturgeon.ca/}%
+%
+\\\email{max.turgeon@umanitoba.ca}
 }
@@
 %
-\url{http://www.medicine.mcgill.ca/epidemiology/hanley/}\\%
-%
-\email{james.hanley@mcgill.ca}%
+\url{http://www.medicine.mcgill.ca/epidemiology/hanley/}%
+%
+\\\email{james.hanley@mcgill.ca}
 }
@@
 %
-\url{http://individual.utoronto.ca/osaarela/}\\%
-%
-\email{olli.saarela@utoronto.ca}%
+\url{http://individual.utoronto.ca/osaarela/}%
+%
+\\\email{olli.saarela@utoronto.ca}
 }

### 2021-10-15 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
compiled paper with updated cs1

@@
 cox_model <- survival::coxph(surv_obj ~ ScrArm, data = ERSPC)
+cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+myCols <- cbPalette[c(4,7)]
 ```
@@
 We used a cubic B-spline expansion of time."}
-cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
-myCols <- cbPalette[c(4,7)]
 
@@
 )
-```
-
-```{r}
-knitr::knit_exit()
 ```

### 2021-10-15 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
compiled paper with updated cs1

--- 2021-10-13 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
@@
 For our first case study, we return to the ERSPC study and investigate
-the differences in risk between the control and screening arms. We fit a
-flexible hazard by using a smooth function of time modeled with cubic
-B-splines. The model is fit using \code{fitSmoothHazard} with the
-familiar formula interface:
+the differences in risk between the control and screening arms. Previous
+re-analyses of these data suggest that the 20\% reduction in prostate
+cancer death due to screening was an underestimate
+\citep{hanley2010mortality}. The estimated 20\% (from a proportional
+hazards model) did not account for the delay between screening and the
+time the effect is expected to be observed. As a result, the null
+effects in years 1--7 masked the substantial reductions that began to
+appear from year 8 onward. This motivates the use of a time-dependent
+hazard ratio which can easily be fit with the \pkg{casebase} package by
+including an interaction term with time in the model. We fit a flexible
+hazard by using a smooth function of time modeled with a penalized cubic
+spline basis with 2 degrees of freedom (implemented in the
+\code{survival::pspline} function). The model is fit using
+\code{fitSmoothHazard} with the familiar formula interface:
 
@@
 \begin{Sinput}
-library(splines)
-fit <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
+fit <- fitSmoothHazard(DeadOfPrCa ~ pspline(Follow.Up.Time, df = 2) * ScrArm, 
                        data = ERSPC, ratio = 100)
@@
 
+Note that we did not have to specify any cut points, as would be the
+case with the \code{survSplit} function in the \pkg{survival} package.
 The output object from \code{fitSmoothHazard} inherits from the
@@
 #>     Null deviance: 6059.0  on 54539  degrees of freedom
-#> Residual deviance: 5787.6  on 54535  degrees of freedom
-#> AIC: 5798
+#> Residual deviance: 5772.7  on 54524  degrees of freedom
+#> AIC: 5805
 #> 
-#> Number of Fisher Scoring iterations: 8
+#> Number of Fisher Scoring iterations: 9
 \end{Soutput}
@@
 
-Next, the \code{absoluteRisk} function takes as input the
-\code{singleEventCB} object and returns a matrix where each column
-corresponds to the covariate profiles specified in the \code{newdata}
-argument, and each row corresponds to time points specified by the
-\code{time} argument:
+As noted in the Theoretical Details section, the usual asymptotic
+results hold for likelihood ratio tests built using case-base sampling
+models. Therefore, we can easily test the significance of the spline
+term and its interaction with time:
 
@@
 \begin{Sinput}
-new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
-new_time <- c(0, 1)
-
-absoluteRisk(fit, time = new_time, newdata = new_data)
+anova(fit, test = "LRT")
 \end{Sinput}
-\begin{Soutput}
-#>  time                
-#>     0 0.0e+00 0.0e+00
-#>     1 5.4e-05 4.2e-05
-\end{Soutput}
-\end{Schunk}
-
-In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated risk
-function from \pkg{casebase} on the Cox model estimate. As noted in the
-Theoretical Details section, the usual asymptotic results hold for
-likelihood ratio tests built using case-base sampling models. Therefore,
-we can easily test the significance of the spline model:
-
-\begin{Schunk}
 \begin{Soutput}
@@
 
-The null hypothesis is rejected in favor of the spline model. Similarly,
-to compare different models (e.g.~time modeled linearly), we could
-compute Akaike's Information Criterion (AIC) for each model and compare
-them.
-
-\begin{Schunk}
-\begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cox-cif-1} \caption{Risk functions for control and screening groups in the ERSPC data.
-We plot the estimate from the Cox model using \code{survival::survfit} (solid line) and the estimate from case-base sampling (dashed line).
-We used a cubic B-spline expansion of time.}\label{fig:erspc-cox-cif}
-\end{figure}
-\end{Schunk}
-
-Next, we can compute confidence intervals for the risk function using
-the method \code{confint.absRiskCB}.
+Similarly, to compare different models (e.g.~time modeled linearly), we
+could compute Akaike's Information Criterion (AIC) for each model and
+compare them.
+
+\hypertarget{time-dependent-hazard-ratios}{%
+\subsubsection{Time-dependent hazard
+ratios}\label{time-dependent-hazard-ratios}}
+
+In what follows, the hazard ratio for a variable \(X\) is defined as
+
+\[
+\frac{\lambda\left(t | X=x_1, \mathbf{Z}=\mathbf{z_1} ; \hat{\beta}\right)}{\lambda(t | X=x_0, \mathbf{Z}=\mathbf{z_0} ; \hat{\beta})}
+\] where \(\lambda(t|\cdot;\hat{\beta})\) is the hazard rate as a
+function of the variable \(t\) (which is usually time, but can be any
+other continuous variable), \(x_1\) is the value of \(X\) for the
+exposed group, \(x_0\) is the value of \(X\) for the unexposed group,
+\(\mathbf{Z}\) are other covariates in the model which are equal to
+\(\mathbf{z_1}\) in the exposed and \(\mathbf{z_0}\) in the unexposed
+group, and \(\hat{\beta}\) are the estimated regression coefficients. As
+indicated by the formula above, it is most instructive to plot the
+hazard ratio as a function of a variable \(t\) only if there is an
+interaction between \(t\) and \(X\). Otherwise, the resulting plot will
+simply be a horizontal line across time.
+
+The \code{plot} method with \code{type="hr"} for objects of class
+\code{singleEventCB} can be used to compute time-dependent hazard ratios
+and confidence intervals. In Figure \ref{fig:interaction-ERSPC}, we show
+the estimated hazard ratio and 95\% confidence interval for screening
+vs.~control group as a function of time. Note that we must specify the
+covariate profiles and times for the predicted hazards.
 
@@
 \begin{Sinput}
-risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
-conf_ints <- confint(risk, fit)
-conf_ints
-\end{Sinput}
-\begin{Soutput}
-#>   time estimate conf.low conf.high        cov_prof
-#> 1    0  0.0e+00  0.0e+00   0.0e+00   Control group
-#> 2    1  5.4e-05  3.5e-05   9.3e-05   Control group
-#> 3    0  0.0e+00  0.0e+00   0.0e+00 Screening group
-#> 4    1  4.2e-05  2.7e-05   7.3e-05 Screening group
-\end{Soutput}
-\end{Schunk}
-
-These point-wise confidence intervals are constructed using parametric
-bootstrap; the validity of this approach relies on the asymptotic
-normality of the MLE. In Figure \ref{fig:erspc-cif-conf}, we see the
-95\% confidence bands around the estimates.
-
-\begin{Schunk}
-\begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cif-conf-1} \caption[Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands]{Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands.}\label{fig:erspc-cif-conf}
-\end{figure}
-\end{Schunk}
-
-Previous re-analyses of these data suggest that the 20\% reduction in
-prostate cancer death due to screening was an underestimate
-\citep{hanley2010mortality}. The estimated 20\% (from a proportional
-hazards model) did not account for the delay between screening and the
-time the effect is expected to be observed. As a result, the null
-effects in years 1--7 masked the substantial reductions that began to
-appear from year 8 onward. This motivates the use of a time-dependent
-hazard ratio which can easily be fit with the \pkg{casebase} package by
-including an interaction term with time in the model:
-
-\begin{Schunk}
-\begin{Sinput}
-fit_inter <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) * ScrArm, 
-                             data = ERSPC)
-\end{Sinput}
-\end{Schunk}
-
-Note that we did not have to specify any cut points, as would be the
-case with the \code{survSplit} function in the \pkg{survival} package.
-In Figure \ref{fig:interaction-ERSPC}, we have the estimated hazard
-ratio and 95\% confidence interval for screening vs.~control group as a
-function of time using the \code{plot} method for objects of class
-\code{singleEventCB}:
-
-\begin{Schunk}
-\begin{Sinput}
-plot(fit_inter, type = "hr", newdata = new_data,
-     var = "ScrArm", xvar = "Follow.Up.Time", ci = TRUE)
+new_time <- seq(1, 12, by  = 0.1)
+new_data <- data.frame(ScrArm = factor("Control group",
+                                         levels = c("Control group","Screening group")),
+                      Follow.Up.Time = new_time)
+plot(fit, type = "hr", newdata = new_data,
+     var = "ScrArm", xvar = "Follow.Up.Time", ci = T)
 \end{Sinput}
@@
 control group as a function of time in the ERSPC dataset.
-Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic B-spline basis of follow-up time and treatment arm.
+Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic pspline basis of follow-up time and treatment arm.
 95\% confidence intervals are calculated using the delta method.
@@
 The 25-60\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\% reported in the original article.}\label{fig:interaction-ERSPC}
+\end{figure}
+\end{Schunk}
+
+\hypertarget{hazard-functions}{%
+\subsubsection{Hazard functions}\label{hazard-functions}}
+
+Modeling the hazard function directly allows us to easily visualize it
+with the \code{plot} method and \code{type="hazard"} for objects of
+class \code{singleEventCB}. We plot the hazard functions for both
+treatment arms in Figure \ref{fig:cs1hazard}. The pattern we see is
+consistent with the population-time plot shown in Figure
+\ref{fig:plot-erspc-data} C, where more events are observed at later
+follow-up times. The drop at the end can be explained by the fact that
+very few observations were followed for the entire 15 year period.
+
+\begin{Schunk}
+\begin{Sinput}
+plot(fit, type = "hazard",
+     hazard.params = list(xvar = "Follow.Up.Time",
+                          by = "ScrArm"))
+\end{Sinput}
+\begin{Soutput}
+#> Conditions used in construction of plot
+#> ScrArm: Control group / Screening group
+#> offset: 0
+\end{Soutput}
+
+\includegraphics[width=\textwidth,keepaspectratio=true]{./cs1hazard-1} \end{Schunk}
+
+\hypertarget{absolute-risk}{%
+\subsubsection{Absolute risk}\label{absolute-risk}}
+
+Next, the \code{absoluteRisk} function takes as input the
+\code{singleEventCB} object and returns a matrix where each column
+corresponds to the covariate profiles specified in the \code{newdata}
+argument, and each row corresponds to time points specified by the
+\code{time} argument:
+
+\begin{Schunk}
+\begin{Sinput}
+new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
+new_time <- seq(0,14,0.1)
+risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
+head(risk)
+\end{Sinput}
+\begin{Soutput}
+#>  time                
+#>   0.0 0.0e+00 0.0e+00
+#>   0.1 2.0e-06 6.2e-06
+#>   0.2 4.2e-06 1.2e-05
+#>   0.3 6.6e-06 1.8e-05
+#>   0.4 9.2e-06 2.3e-05
+#>   0.5 1.2e-05 2.9e-05
+\end{Soutput}
+\end{Schunk}
+
+We can subsequently compute confidence intervals for the risk function
+using the method \code{confint.absRiskCB}:
+
+\begin{Schunk}
+\begin{Sinput}
+conf_ints <- confint(risk, fit)
+head(conf_ints)
+\end{Sinput}
+\begin{Soutput}
+#>   time estimate conf.low conf.high      cov_prof
+#> 1  0.0  0.0e+00  0.0e+00   0.0e+00 Control group
+#> 2  0.1  2.0e-06  2.5e-07   1.6e-05 Control group
+#> 3  0.2  4.2e-06  6.1e-07   3.0e-05 Control group
+#> 4  0.3  6.6e-06  1.1e-06   4.2e-05 Control group
+#> 5  0.4  9.2e-06  1.8e-06   5.2e-05 Control group
+#> 6  0.5  1.2e-05  2.8e-06   6.2e-05 Control group
+\end{Soutput}
+\end{Schunk}
+
+These point-wise confidence intervals are constructed using parametric
+bootstrap; the validity of this approach relies on the asymptotic
+normality of the MLE. In Figure \ref{fig:erspc-cif-conf}, we see the
+95\% confidence bands around the estimates. We also overlay the
+Kaplan-Meier curves as a reference.
+
+\begin{Schunk}
+\begin{figure}[ht]
+\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cif-conf-1} \caption[Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands]{Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands.}\label{fig:erspc-cif-conf}
 \end{figure}
@@
 We want to compare our hazard ratio estimates to that obtained from a
-Cox regression.
+Cox regression (using the \pkg{survival} package version 3.2-13).
 
@@
 \midrule
-Sex & 0.74 & (0.43, 1.3) & 0.68 & (0.39, 1.21)\\
-Disease & 0.52 & (0.29, 0.94) & 0.52 & (0.28, 0.94)\\
-Phase (CR2 vs. CR1) & 1.20 & (0.48, 3) & 1.21 & (0.47, 3.09)\\
-Phase (CR3 vs. CR1) & 1.59 & (0.41, 6.17) & 1.67 & (0.46, 6.08)\\
-Phase (Relapse vs. CR1) & 4.09 & (1.9, 8.81) & 4.55 & (1.98, 10.46)\\
+Sex & 0.75 & (0.43, 1.31) & 0.68 & (0.39, 1.21)\\
+Disease & 0.55 & (0.3, 0.99) & 0.52 & (0.28, 0.94)\\
+Phase (CR2 vs. CR1) & 1.10 & (0.44, 2.74) & 1.21 & (0.47, 3.09)\\
+Phase (CR3 vs. CR1) & 1.75 & (0.45, 6.78) & 1.67 & (0.46, 6.08)\\
+Phase (Relapse vs. CR1) & 4.23 & (1.96, 9.14) & 4.55 & (1.98, 10.46)\\
 \addlinespace
-Source & 1.70 & (0.55, 5.23) & 1.46 & (0.47, 4.53)\\
-Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
+Source & 1.61 & (0.53, 4.91) & 1.46 & (0.47, 4.53)\\
+Age & 1.00 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
 \bottomrule

### 2021-10-15 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
version sent to Jim and Olli

@@
 We would also like to thank Yi Yang for helpful discussions on penalized regression models.
-\mbox{Bhatnagar} gratefully acknowledges funding via a Discovery Grant from the Natural Sciences and Engineering Research Council of Canada (NSERC).
+\mbox{Bhatnagar} gratefully acknowledges funding via a Discovery Grant from the Natural Sciences and Engineering Research Council of Canada (NSERC), RGPIN-2020-05133.
 

### 2021-10-24 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
add NSERC DG in acknowledgment

--- 2021-10-15 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
@@
 library(casebase)
-library(survival)
-library(splines)
-library(ggplot2)
 library(colorspace)
@@
 library(prodlim)
-# library(riskRegression)
+library(riskRegression)
+library(splines)
+library(survival)
 library(tibble)
@@
 # Define theme for plots
-paper_gg_theme <- ggplot2::theme_bw(base_size = 10) + ggplot2::theme(legend.position = "bottom")
+paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
 # Define colour palette for CS 3
-q4 <- colorspace::qualitative_hcl(4, palette = "Dark 3")
+q4 <- qualitative_hcl(4, palette = "Dark 3")
 names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
@@
 to instability in the Newton-Raphson procedure for estimation 
-[@kalbfleisch2011statistical, Section 4.8]. 
-On the other hand, a coarse grouping could potentially mask 
+[@kalbfleisch2011statistical]. On the other hand, a coarse grouping could potentially mask 
 nonlinear trends in the hazard function.
@@
 Rather than discretizing time, Hanley & Miettinen [-@hanley2009fitting] 
-selected a discrete set of person-time coordinates (`person-moments') in 
+selected a discrete set of person-time coordinates ("person-moments") in 
 continuous time from all observed follow-up experience constituting the 
@@
 data("ERSPC")
+# ERSPC$ScrArm was turned to factor in 0.10.1
+if (utils::packageVersion("casebase") < package_version("0.10.1")) {
   ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
@@
                        labels = c("Control group", "Screening group"))
-```
-
-```{r erspc-data, eval = eval_cs1, echo=TRUE}
+}
+```
+
+```{r erspc-data, eval = eval_cs1}
 pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
@@
 For our first case study, we return to the ERSPC study and investigate the differences in risk between the control and screening arms.
-Previous re-analyses of these data suggest that the 20% reduction in prostate cancer death due to screening was an underestimate [@hanley2010mortality]. 
-The estimated 20% (from a proportional hazards model) did not account for the delay between screening and the time the effect is expected to be observed. 
-As a result, the null effects in years 1–7 masked the substantial reductions that began to appear from year 8 onward. 
-This motivates the use of a time-dependent hazard ratio which can easily be fit with the \pkg{casebase} package by including an interaction term with time in the model. 
-We fit a flexible hazard by using a smooth function of time modeled with a penalized cubic spline basis with 2 degrees of freedom (implemented in the \code{survival::pspline} function). 
+We fit a flexible hazard by using a smooth function of time modeled with cubic B-splines.
 The model is fit using \code{fitSmoothHazard} with the familiar formula interface:
@@
 ```{r, eval = eval_cs1, echo = -1}
-set.seed(1953)
-fit <- fitSmoothHazard(DeadOfPrCa ~ pspline(Follow.Up.Time, df = 2) * ScrArm, 
+set.seed(12)
+library(splines)
+fit <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
                        data = ERSPC, ratio = 100)
@@
 
-Note that we did not have to specify any cut points, as would be the case with the \code{survSplit} function in the \pkg{survival} package. 
 The output object from \code{fitSmoothHazard} inherits from the \code{singleEventCB} and \code{glm} classes.
@@
 
-As noted in the Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models. 
-Therefore, we can easily test the significance of the spline term and its interaction with time:
-
-```{r erscp-compare, eval = eval_cs1, echo = TRUE}
-anova(fit, test = "LRT")
-```
-
-Similarly, to compare different models (e.g. time modeled linearly), we could compute Akaike's Information Criterion (AIC) for each model and compare them.
-
-### Time-dependent hazard ratios
-
-In what follows, the hazard ratio for a variable $X$ is defined as
-
-$$
-\frac{\lambda\left(t | X=x_1, \mathbf{Z}=\mathbf{z_1} ; \hat{\beta}\right)}{\lambda(t | X=x_0, \mathbf{Z}=\mathbf{z_0} ; \hat{\beta})}
-$$
-where $\lambda(t|\cdot;\hat{\beta})$ is the hazard rate as a function of the variable $t$ (which is usually time, but can be any other continuous variable), $x_1$ is the value of $X$ for the exposed group, $x_0$ is the value of $X$ for the unexposed group, $\mathbf{Z}$ are other covariates in the model which are equal to $\mathbf{z_1}$ in the exposed and $\mathbf{z_0}$ in the unexposed group, and $\hat{\beta}$ are the estimated regression coefficients.
-As indicated by the formula above, it is most instructive to plot the hazard ratio as a function of a variable $t$ only if there is an interaction between $t$ and $X$. 
-Otherwise, the resulting plot will simply be a horizontal line across time.
-
-The \code{plot} method with \code{type="hr"} for objects of class \code{singleEventCB} can be used to compute time-dependent hazard ratios and confidence intervals. 
-In Figure \ref{fig:interaction-ERSPC}, we show the estimated hazard ratio and 95% confidence interval for screening vs. control group as a function of time. 
-Note that we must specify the covariate profiles and times for the predicted hazards.
-
-```{r, echo = TRUE, eval = FALSE}
-new_time <- seq(1, 12, by  = 0.1)
-new_data <- data.frame(ScrArm = factor("Control group",
-                                         levels = c("Control group","Screening group")),
-                      Follow.Up.Time = new_time)
-plot(fit, type = "hr", newdata = new_data,
-     var = "ScrArm", xvar = "Follow.Up.Time", ci = T)
-```
-
-The plot shows that the effect of screening only becomes statistically apparent by year 7 and later. 
-The 25-60% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20% reported in the original article.
-
-```{r interaction-ERSPC, eval = eval_cs1, fig.width=8, fig.height=6, fig.cap='Estimated hazard ratio and 95\\% confidence interval for screening vs.
-control group as a function of time in the ERSPC dataset.
-Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic pspline basis of follow-up time and treatment arm.
-95\\% confidence intervals are calculated using the delta method.
-The plot shows that the effect of screening only begins to become statistically apparent by year 7.
-The 25-60\\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\\% reported in the original article.'}
-new_time <- seq(1, 12, by  = 0.1)
-new_data <- data.frame(ScrArm = factor("Control group",
-                                         levels = c("Control group","Screening group")),
-                      Follow.Up.Time = new_time)
-
-par(oma = c(0,1,0,8))
-tt <- plot(fit,
-     type = "hr",
-     newdata = new_data,
-     var = "ScrArm",
-     increment = 1,
-     xvar = "Follow.Up.Time",
-     ci = T,
-     xlab = "Follow-up time (years)",
-     ylab = "Death from prostate cancer hazard ratio",
-     ylim = c(0,1.50),
-     yaxt = 'n',
-     xaxt = "n",
-     rug = TRUE)
-
-axis(2, at = seq(0,1.50, by=0.25), las = 2)
-axis(1, at = seq(min(new_time),max(new_time), by=1))
-for( i in seq(0,1.50, by=0.25)) {
-  abline(h = i, lty = 1, col = "lightgrey")
-}
-for( i in seq(min(new_time),max(new_time), by=1)) {
-  abline(v = i, lty = 1, col = "lightgrey")
-}
-for(i in seq_along(seq(0, 0.75, by = 0.25))) {
-  mtext(paste0(seq(0, 0.75, by = 0.25)[i]*100,"%"), side = 4, las = 2, adj = 1, outer = TRUE, line = 2, at = c(0.63-0.11*(i-1)))
-}
-lines(tt$Follow.Up.Time, tt$hazard_ratio, lwd = 2, lty = 1)
-mtext("% reduction in\nprostate cancer\nmortality rate", side = 4, las = 2, outer = TRUE, line = -1, at = 0.7)
-```
-
-### Hazard functions
-
-Modeling the hazard function directly allows us to easily visualize it with the \code{plot} method and \code{type="hazard"} for objects of class \code{singleEventCB}. 
-We plot the hazard functions for both treatment arms in Figure \ref{fig:cs1hazard}. 
-The pattern we see is consistent with the population-time plot shown in Figure \ref{fig:plot-erspc-data} C, where more events are observed at later follow-up times. 
-The drop at the end can be explained by the fact that very few observations were followed for the entire 15 year period.
-
-```{r cs1hazard, eval=eval_cs1, echo=TRUE}
-plot(fit, type = "hazard",
-     hazard.params = list(xvar = "Follow.Up.Time",
-                          by = "ScrArm"))
-```
-
-### Absolute risk
-
 Next, the \code{absoluteRisk} function takes as input the \code{singleEventCB} object and returns a matrix where each column corresponds to the covariate profiles specified in the \code{newdata} argument, and each row corresponds to time points specified by the \code{time} argument:
 
-```{r, eval=FALSE, echo=FALSE}
+```{r, eval=eval_cs1, echo=echo_plot_code}
 new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
@@
 
-```{r, eval=eval_cs1, echo=TRUE}
-new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
-new_time <- seq(0,14,0.1)
-risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
-head(risk)
-```
-
-<!--In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated risk function from \pkg{casebase} on the Cox model estimate. -->
+In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated risk function from \pkg{casebase} on the Cox model estimate.
+As noted in the Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models.
+Therefore, we can easily test the significance of the spline model:
+
+```{r erscp-compare, eval = eval_cs1, echo = FALSE}
+anova(fit, test = "LRT")
+```
+
+The null hypothesis is rejected in favor of the spline model.
+Similarly, to compare different models (e.g.
+time modeled linearly), we could compute Akaike's Information Criterion (AIC) for each model and compare them.
 
@@
 
-```{r erspc-cox-cif, eval = FALSE, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk functions for control and screening groups in the ERSPC data.
+```{r erspc-cox-cif, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk functions for control and screening groups in the ERSPC data.
 We plot the estimate from the Cox model using \\code{survival::survfit} (solid line) and the estimate from case-base sampling (dashed line).
@@
 
-We can subsequently compute confidence intervals for the risk function using the method \code{confint.absRiskCB}:
-
-```{r eval = eval_cs1, echo = TRUE}
+Next, we can compute confidence intervals for the risk function using the method \code{confint.absRiskCB}.
+
+```{r eval = eval_cs1, echo = -1}
+new_time <- c(0, 1)
+risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
 conf_ints <- confint(risk, fit)
-head(conf_ints)
+conf_ints
 ```
@@
 In Figure \ref{fig:erspc-cif-conf}, we see the 95% confidence bands around the estimates.
-We also overlay the Kaplan-Meier curves as a reference.
 
 ```{r erspc-cif-conf, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\\% confidence bands."}
-
-tt <- conf_ints
+# confidence interval
+tt <- confint(risk_mp, fit)
 
 # plot absrisk and 95% CI
+new_time <- seq(0,14,0.1)
 i.backw <- order(new_time, decreasing = TRUE)
@@
 
-```{r}
-knitr::knit_exit()
+Previous re-analyses of these data suggest that the 20% reduction in prostate cancer death due to screening was an underestimate [@hanley2010mortality].
+The estimated 20% (from a proportional hazards model) did not account for the delay between screening and the time the effect is expected to be observed.
+As a result, the null effects in years 1–7 masked the substantial reductions that began to appear from year 8 onward.
+This motivates the use of a time-dependent hazard ratio which can easily be fit with the \pkg{casebase} package by including an interaction term with time in the model:
+
+```{r, echo=-1, eval = eval_cs1}
+set.seed(12)
+fit_inter <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) * ScrArm, 
+                             data = ERSPC)
+```
+
+Note that we did not have to specify any cut points, as would be the case with the \code{survSplit} function in the \pkg{survival} package.
+In Figure \ref{fig:interaction-ERSPC}, we have the estimated hazard ratio and 95% confidence interval for screening vs.
+control group as a function of time using the \code{plot} method for objects of class \code{singleEventCB}:
+
+```{r, echo = TRUE, eval = FALSE}
+plot(fit_inter, type = "hr", newdata = new_data,
+     var = "ScrArm", xvar = "Follow.Up.Time", ci = TRUE)
+```
+
+The plot shows that the effect of screening only becomes statistically apparent by year 7 and later.
+The 25-60% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20% reported in the original article.
+
+```{r interaction-ERSPC, eval = eval_cs1, fig.width=8, fig.height=6, fig.cap='Estimated hazard ratio and 95\\% confidence interval for screening vs.
+control group as a function of time in the ERSPC dataset.
+Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic B-spline basis of follow-up time and treatment arm.
+95\\% confidence intervals are calculated using the delta method.
+The plot shows that the effect of screening only begins to become statistically apparent by year 7.
+The 25-60\\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\\% reported in the original article.'}
+new_time <- seq(1, 12, by  = 0.1)
+new_data <- data.frame(ScrArm = factor("Control group",
+                                         levels = c("Control group","Screening group")),
+                      Follow.Up.Time = new_time)
+
+par(oma = c(0,1,0,8))
+tt <- plot(fit_inter,
+     type = "hr",
+     newdata = new_data,
+     var = "ScrArm",
+     increment = 1,
+     xvar = "Follow.Up.Time",
+     ci = T,
+     xlab = "Follow-up time (years)",
+     ylab = "Death from prostate cancer hazard ratio",
+     ylim = c(0,1.50),
+     yaxt = 'n',
+     xaxt = "n",
+     rug = TRUE)
+
+axis(2, at = seq(0,1.50, by=0.25), las = 2)
+axis(1, at = seq(min(new_time),max(new_time), by=1))
+for( i in seq(0,1.50, by=0.25)) {
+  abline(h = i, lty = 1, col = "lightgrey")
+}
+for( i in seq(min(new_time),max(new_time), by=1)) {
+  abline(v = i, lty = 1, col = "lightgrey")
+}
+for(i in seq_along(seq(0, 0.75, by = 0.25))) {
+  mtext(paste0(seq(0, 0.75, by = 0.25)[i]*100,"%"), side = 4, las = 2, adj = 1, outer = TRUE, line = 2, at = c(0.63-0.11*(i-1)))
+}
+lines(tt$Follow.Up.Time, tt$hazard_ratio, lwd = 2, lty = 1)
+mtext("% reduction in\nprostate cancer\nmortality rate", side = 4, las = 2, outer = TRUE, line = -1, at = 0.7)
 ```
@@
 
-We want to compare our hazard ratio estimates to that obtained from a Cox regression (using the \pkg{survival} package version 3.2-13).
+We want to compare our hazard ratio estimates to that obtained from a Cox regression.
 
@@
 ```{r supportData, eval = eval_cs3}
-library(riskRegression)
 scaleFUNy <- function(x) sprintf("%.2f", x)
@@
 We would also like to thank Yi Yang for helpful discussions on penalized regression models.
-\mbox{Bhatnagar} gratefully acknowledges funding via a Discovery Grant from the Natural Sciences and Engineering Research Council of Canada (NSERC).
+\mbox{Bhatnagar} and \mbox{Turgeon} both gratefully acknowledge funding via a Discovery Grant from the Natural Sciences and Engineering Research Council of Canada (NSERC).
 

### 2021-10-24 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
add NSERC DG in acknowledgment

--- 2021-10-13 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
@@
 \citeyearpar{hanley2009fitting} selected a discrete set of person-time
-coordinates (`person-moments') in continuous time from all observed
+coordinates (``person-moments'') in continuous time from all observed
 follow-up experience constituting the study base. By doing so, they
@@
 comments and criticisms. We would also like to thank Yi Yang for helpful
-discussions on penalized regression models. \mbox{Bhatnagar} gratefully
-acknowledges funding via a Discovery Grant from the Natural Sciences and
-Engineering Research Council of Canada (NSERC).
+discussions on penalized regression models. \mbox{Bhatnagar} and
+\mbox{Turgeon} both gratefully acknowledge funding via a Discovery Grant
+from the Natural Sciences and Engineering Research Council of Canada
+(NSERC).
 

### 2021-10-25 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd

--- 2021-10-15 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
@@
                        labels = c("Control group", "Screening group"))
+}
 ```

### 2021-10-26 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
newest changes sb

@@
 Each one showcases a different type of analysis.
-First, we show how to model time flexibly as well as non-proportional hazards.
+First, we show how to model non-constant and non-proportional hazards through a flexible specification of time.
 Then we perform a competing-risk analysis and compare our results with the Cox model and the Fine-Gray model.
@@
 
-## Case study 1---Modeling different functions of time and non-proportional hazards
-
-For our first case study, we return to the ERSPC study and investigate the differences in risk between the control and screening arms. 
-Previous re-analyses of these data suggest that the 20% reduction in prostate cancer death due to screening was an underestimate [@hanley2010mortality]. 
-The estimated 20% (from a proportional hazards model) did not account for the delay between screening and the time the effect is expected to be observed. 
-As a result, the null effects in years 1–7 masked the substantial reductions that began to appear from year 8 onward. 
-This motivates the use of a time-dependent hazard ratio which can easily be fit with the \pkg{casebase} package by including an interaction term with time in the model. 
-We fit a flexible hazard by using a smooth function of time modeled with a penalized cubic spline basis with 2 degrees of freedom (implemented in the \code{survival::pspline} function). 
-The model is fit using \code{fitSmoothHazard} with the familiar formula interface:
+## Case study 1---Flexible modeling of the hazard function
+
+For our first case study, we return to the ERSPC study and investigate the 
+differences in risk between the control and screening arms. Previous re-analyses 
+of these data suggest that the 20% reduction in prostate cancer death due to 
+screening was an underestimate [@hanley2010mortality]. 
+The estimated 20% (from a proportional hazards model) did not account for 
+the delay between screening and the time the effect is expected to be observed. 
+As a result, the null effects in years 1–7 masked the substantial reductions 
+that began to appear from year 8 onward. This motivates the use of a 
+time-dependent hazard ratio which can easily be fit with the \pkg{casebase} 
+package by including an interaction term with time in the model. We fit a 
+flexible hazard by using a smooth function of time modeled with a penalized 
+cubic spline basis with 2 degrees of freedom (implemented in the 
+\code{survival::pspline} function). The model is fit using \code{fitSmoothHazard} 
+with the familiar formula interface:
 
@@
 
-Note that we did not have to specify any cut points, as would be the case with the \code{survSplit} function in the \pkg{survival} package. 
-The output object from \code{fitSmoothHazard} inherits from the \code{singleEventCB} and \code{glm} classes.
-For this reason, we can leverage the \code{summary} method for \code{glm} objects to output a familiar summary of the results:
+<!--Note that we did not have to specify any cut points, as would be the case 
+with the \code{survSplit} function in the \pkg{survival} package.--> 
+The output object from \code{fitSmoothHazard} inherits from the \code{singleEventCB} 
+and \code{glm} classes. For this reason, we can leverage the \code{summary} 
+method for \code{glm} objects to output a familiar summary of the results:
 
@@
 
-As noted in the Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models. 
-Therefore, we can easily test the significance of the spline term and its interaction with time:
+As noted in the Theoretical Details section, the usual asymptotic results hold 
+for likelihood ratio tests built using case-base sampling models. Therefore, we 
+can easily test the significance of the spline term and its interaction with time:
 
@@
 
-Similarly, to compare different models (e.g. time modeled linearly), we could compute Akaike's Information Criterion (AIC) for each model and compare them.
+Similarly, to compare different models (e.g. time modeled linearly), we could 
+compute Akaike's Information Criterion (AIC) for each model.
 
@@
 $$
-where $\lambda(t|\cdot;\hat{\beta})$ is the hazard rate as a function of the variable $t$ (which is usually time, but can be any other continuous variable), $x_1$ is the value of $X$ for the exposed group, $x_0$ is the value of $X$ for the unexposed group, $\mathbf{Z}$ are other covariates in the model which are equal to $\mathbf{z_1}$ in the exposed and $\mathbf{z_0}$ in the unexposed group, and $\hat{\beta}$ are the estimated regression coefficients.
-As indicated by the formula above, it is most instructive to plot the hazard ratio as a function of a variable $t$ only if there is an interaction between $t$ and $X$. 
-Otherwise, the resulting plot will simply be a horizontal line across time.
-
-The \code{plot} method with \code{type="hr"} for objects of class \code{singleEventCB} can be used to compute time-dependent hazard ratios and confidence intervals. 
-In Figure \ref{fig:interaction-ERSPC}, we show the estimated hazard ratio and 95% confidence interval for screening vs. control group as a function of time. 
-Note that we must specify the covariate profiles and times for the predicted hazards.
+where $\lambda(t|\cdot;\hat{\beta})$ is the hazard rate as a function of the 
+variable $t$ (which is usually time, but can be any other continuous variable), 
+$x_1$ is the value of $X$ for the exposed group, $x_0$ is the value of $X$ for 
+the unexposed group, $\mathbf{Z}$ are other covariates in the model which are 
+equal to $\mathbf{z_1}$ in the exposed and $\mathbf{z_0}$ in the unexposed group, 
+and $\hat{\beta}$ are the estimated regression coefficients. As indicated by the 
+formula above, it is most instructive to plot the hazard ratio as a function of 
+a variable $t$ only if there is an interaction between $t$ and $X$. Otherwise, 
+the resulting plot will simply be a horizontal line across time.
+
+The \code{plot} method with \code{type="hr"} for objects of class 
+\code{singleEventCB} can be used to compute time-dependent hazard ratios and 
+confidence intervals. In Figure \ref{fig:interaction-ERSPC}, we show the 
+estimated hazard ratio and 95% confidence interval for screening vs. control 
+group as a function of time. Note that we must specify the covariate profile for 
+the reference group and times for the predicted hazards.
 
@@
 
-The plot shows that the effect of screening only becomes statistically apparent by year 7 and later. 
-The 25-60% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20% reported in the original article.
-
 ```{r interaction-ERSPC, eval = eval_cs1, fig.width=8, fig.height=6, fig.cap='Estimated hazard ratio and 95\\% confidence interval for screening vs.
 control group as a function of time in the ERSPC dataset.
-Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic pspline basis of follow-up time and treatment arm.
+Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic pspline basis (df=2) of follow-up time and treatment arm.
 95\\% confidence intervals are calculated using the delta method.
@@
 
+The plot shows that the effect of screening only becomes statistically apparent 
+by year 7 and later. The 25-60% reductions seen in years 8-12 of the study suggests 
+a much higher reduction in prostate cancer due to screening than the single 
+overall 20% reported in the original article.
+
 ### Hazard functions
 
-Modeling the hazard function directly allows us to easily visualize it with the \code{plot} method and \code{type="hazard"} for objects of class \code{singleEventCB}. 
+Modeling the hazard function directly allows us to easily visualize it with 
+the \code{plot} method and \code{type="hazard"} for objects of class \code{singleEventCB}. 
 We plot the hazard functions for both treatment arms in Figure \ref{fig:cs1hazard}. 
@@
 
-```{r cs1hazard, eval=eval_cs1, echo=TRUE}
+```{r cs1hazard, eval=eval_cs1, echo=TRUE, fig.cap='Estimated hazard functions for control and screening groups in the ERSPC dataset.
+Hazards are estimated from fitting a parametric model with casebase sampling as a function of the interaction between a cubic pspline basis (df=2) of follow-up time and treatment arm.
+The package vignettes provide a detailed description of how to plot hazard functions for any combination of covariates along with confidence bands.'}
 plot(fit, type = "hazard",
@@
 
+\FloatBarrier
+
 ### Absolute risk
@@
 Next, the \code{absoluteRisk} function takes as input the \code{singleEventCB} object and returns a matrix where each column corresponds to the covariate profiles specified in the \code{newdata} argument, and each row corresponds to time points specified by the \code{time} argument:
-
-```{r, eval=FALSE, echo=FALSE}
-new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
-new_time <- c(0, 1)
-
-absoluteRisk(fit, time = new_time, newdata = new_data)
-```
 
@@
 risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
-head(risk)
 ```
@@
 
-```{r erspc-cox-cif, eval = FALSE, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk functions for control and screening groups in the ERSPC data.
-We plot the estimate from the Cox model using \\code{survival::survfit} (solid line) and the estimate from case-base sampling (dashed line).
-We used a cubic B-spline expansion of time."}
-
-new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
-new_time <- seq(0,14,0.1)
-# Recompute with more time points
-risk_mp <- absoluteRisk(fit, time = new_time, newdata = new_data)
-
-plot(survfit(cox_model, newdata = new_data),
-     ylab = "Risk probability (%)",
-     xlab = "Years since randomization",
-     fun = "event",
-     xmax = max(new_time),
-     conf.int = FALSE, col = myCols, 
-     lwd = 2,
-     yscale = 100,
-     xaxt = 'n')
-legend("topleft", 
-       legend = c("Control (Cox)","Control (Case-base)",
-                  "Screening (Cox)", "Screening (Case-base)"), 
-       col = rep(myCols, each = 2),
-       lty = c(1, 2, 1, 2), 
-       lwd = 2,
-       cex = 1.1,
-       bg = "gray90")
-lines(risk_mp[,1], risk_mp[,2], col = myCols[1], lty = 2, lwd = 2)
-lines(risk_mp[,1], risk_mp[,3], col = myCols[2], lty = 2, lwd = 2)
-```
-
-We can subsequently compute confidence intervals for the risk function using the method \code{confint.absRiskCB}:
+We can subsequently compute confidence intervals for the risk function using 
+the method \code{confint.absRiskCB}:
 
@@
 
-These point-wise confidence intervals are constructed using parametric bootstrap; the validity of this approach relies on the asymptotic normality of the MLE. 
-In Figure \ref{fig:erspc-cif-conf}, we see the 95% confidence bands around the estimates. 
+These point-wise confidence intervals are constructed using parametric bootstrap; 
+the validity of this approach relies on the asymptotic normality of the MLE. In Figure \ref{fig:erspc-cif-conf}, we see the 95% confidence bands around the estimates. 
 We also overlay the Kaplan-Meier curves as a reference.
@@
 We would also like to thank Yi Yang for helpful discussions on penalized regression models.
-\mbox{Bhatnagar} and \mbox{Turgeon} both gratefully acknowledge funding via a Discovery Grant from the Natural Sciences and Engineering Research Council of Canada (NSERC).
+\mbox{Bhatnagar} (RGPIN-2020-05133) and \mbox{Turgeon} (RGPIN-2021-04073) both gratefully acknowledge funding via a Discovery Grant from the Natural Sciences and Engineering Research Council of Canada (NSERC), RGPIN.
 

### 2021-10-26 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
newest changes sb

@@
 package through three case studies. Each one showcases a different type
-of analysis. First, we show how to model time flexibly as well as
-non-proportional hazards. Then we perform a competing-risk analysis and
-compare our results with the Cox model and the Fine-Gray model. The
-third case study illustrates how to perform variable selection in
-high-dimensional datasets.
-
-\hypertarget{case-study-1modeling-different-functions-of-time-and-non-proportional-hazards}{%
-\subsection{Case study 1---Modeling different functions of time and
-non-proportional
-hazards}\label{case-study-1modeling-different-functions-of-time-and-non-proportional-hazards}}
+of analysis. First, we show how to model non-constant and
+non-proportional hazards through a flexible specification of time. Then
+we perform a competing-risk analysis and compare our results with the
+Cox model and the Fine-Gray model. The third case study illustrates how
+to perform variable selection in high-dimensional datasets.
+
+\hypertarget{case-study-1flexible-modeling-of-the-hazard-function}{%
+\subsection{Case study 1---Flexible modeling of the hazard
+function}\label{case-study-1flexible-modeling-of-the-hazard-function}}
 
@@
 
-Note that we did not have to specify any cut points, as would be the
-case with the \code{survSplit} function in the \pkg{survival} package.
 The output object from \code{fitSmoothHazard} inherits from the
@@
 Similarly, to compare different models (e.g.~time modeled linearly), we
-could compute Akaike's Information Criterion (AIC) for each model and
-compare them.
+could compute Akaike's Information Criterion (AIC) for each model.
 
@@
 vs.~control group as a function of time. Note that we must specify the
-covariate profiles and times for the predicted hazards.
+covariate profile for the reference group and times for the predicted
+hazards.
 
@@
 
+\begin{Schunk}
+\begin{figure}[ht]
+\includegraphics[width=\textwidth,keepaspectratio=true]{./interaction-ERSPC-1} \caption[Estimated hazard ratio and 95\% confidence interval for screening vs]{Estimated hazard ratio and 95\% confidence interval for screening vs.
+control group as a function of time in the ERSPC dataset.
+Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic pspline basis (df=2) of follow-up time and treatment arm.
+95\% confidence intervals are calculated using the delta method.
+The plot shows that the effect of screening only begins to become statistically apparent by year 7.
+The 25-60\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\% reported in the original article.}\label{fig:interaction-ERSPC}
+\end{figure}
+\end{Schunk}
+
 The plot shows that the effect of screening only becomes statistically
@@
 screening than the single overall 20\% reported in the original article.
-
-\begin{Schunk}
-\begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./interaction-ERSPC-1} \caption[Estimated hazard ratio and 95\% confidence interval for screening vs]{Estimated hazard ratio and 95\% confidence interval for screening vs.
-control group as a function of time in the ERSPC dataset.
-Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic pspline basis of follow-up time and treatment arm.
-95\% confidence intervals are calculated using the delta method.
-The plot shows that the effect of screening only begins to become statistically apparent by year 7.
-The 25-60\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\% reported in the original article.}\label{fig:interaction-ERSPC}
-\end{figure}
-\end{Schunk}
 
@@
 \end{Soutput}
-
-\includegraphics[width=\textwidth,keepaspectratio=true]{./cs1hazard-1} \end{Schunk}
+\begin{figure}[ht]
+\includegraphics[width=\textwidth,keepaspectratio=true]{./cs1hazard-1} \caption[Estimated hazard functions for control and screening groups in the ERSPC dataset]{Estimated hazard functions for control and screening groups in the ERSPC dataset.
+Hazards are estimated from fitting a parametric model with casebase sampling as a function of the interaction between a cubic pspline basis (df=2) of follow-up time and treatment arm.
+The package vignettes provide a detailed description of how to plot hazard functions for any combination of covariates along with confidence bands.}\label{fig:cs1hazard}
+\end{figure}
+\end{Schunk}
+
+\FloatBarrier
 
@@
 risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
-head(risk)
-\end{Sinput}
-\begin{Soutput}
-#>  time                
-#>   0.0 0.0e+00 0.0e+00
-#>   0.1 2.0e-06 6.2e-06
-#>   0.2 4.2e-06 1.2e-05
-#>   0.3 6.6e-06 1.8e-05
-#>   0.4 9.2e-06 2.3e-05
-#>   0.5 1.2e-05 2.9e-05
-\end{Soutput}
+\end{Sinput}
 \end{Schunk}
@@
 comments and criticisms. We would also like to thank Yi Yang for helpful
-discussions on penalized regression models. \mbox{Bhatnagar} and
-\mbox{Turgeon} both gratefully acknowledge funding via a Discovery Grant
-from the Natural Sciences and Engineering Research Council of Canada
-(NSERC).
+discussions on penalized regression models. \mbox{Bhatnagar}
+(RGPIN-2020-05133) and \mbox{Turgeon} (RGPIN-2021-04073) both gratefully
+acknowledge funding via a Discovery Grant from the Natural Sciences and
+Engineering Research Council of Canada (NSERC), RGPIN.
 

### 2021-10-26 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.R

@@ -14,6 +14,9 @@ knitr::opts_chunk$set(
 
 options(kableExtra.latex.load_packages = FALSE)
 library(casebase)
+library(survival)
+library(splines)
+library(ggplot2)
 library(colorspace)
 library(cowplot)
 library(data.table)
@@ -21,11 +24,9 @@ library(glmnet)
 library(kableExtra)
 library(magrittr)
 library(pracma)
-library(flexsurv) 
+library(flexsurv)
 library(prodlim)
-library(riskRegression)
-library(splines)
-library(survival)
+# library(riskRegression)
 library(tibble)
 library(visreg)
 library(lubridate)
@@ -46,9 +47,9 @@ eval_cs4 <- FALSE
 eval_colophon <- FALSE # Only for draft
 echo_plot_code <- TRUE
 # Define theme for plots
-paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
+paper_gg_theme <- ggplot2::theme_bw(base_size = 10) + ggplot2::theme(legend.position = "bottom")
 # Define colour palette for CS 3
-q4 <- qualitative_hcl(4, palette = "Dark 3")
+q4 <- colorspace::qualitative_hcl(4, palette = "Dark 3")
 names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
 
 
@@ -62,7 +63,7 @@ if (utils::packageVersion("casebase") < package_version("0.10.1")) {
 }
 
 
-## ----erspc-data, eval = eval_cs1----------------------------------------------
+## ----erspc-data, eval = eval_cs1, echo=TRUE-----------------------------------
 pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
                                event = "DeadOfPrCa", exposure = "ScrArm")
 inherits(pt_object, "popTime")
@@ -151,9 +152,8 @@ g
 
 
 ## ---- eval = eval_cs1, echo = -1----------------------------------------------
-set.seed(12)
-library(splines)
-fit <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
+set.seed(1953)
+fit <- fitSmoothHazard(DeadOfPrCa ~ pspline(Follow.Up.Time, df = 2) * ScrArm, 
                        data = ERSPC, ratio = 100)
 
 
[…]

### 2021-10-26 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.R
final revision for max

@@ -1,4 +1,4 @@
-## ----setup, echo=FALSE, message=FALSE, warning=FALSE--------------------------
+## ----setup, echo=FALSE, message=FALSE, warning=FALSE---------------------------------------------------
 knitr::opts_chunk$set(
   collapse = TRUE,
   warning = FALSE,
@@ -53,7 +53,7 @@ q4 <- colorspace::qualitative_hcl(4, palette = "Dark 3")
 names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
 
 
-## ----erspc-data-mutate, eval = eval_cs1, echo = FALSE-------------------------
+## ----erspc-data-mutate, eval = eval_cs1, echo = FALSE--------------------------------------------------
 data("ERSPC")
 # ERSPC$ScrArm was turned to factor in 0.10.1
 if (utils::packageVersion("casebase") < package_version("0.10.1")) {
@@ -63,7 +63,7 @@ if (utils::packageVersion("casebase") < package_version("0.10.1")) {
 }
 
 
-## ----erspc-data, eval = eval_cs1, echo=TRUE-----------------------------------
+## ----erspc-data, eval = eval_cs1, echo=TRUE------------------------------------------------------------
 pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
                                event = "DeadOfPrCa", exposure = "ScrArm")
 inherits(pt_object, "popTime")
@@ -117,7 +117,7 @@ g <- plot_grid(plotspop, legend_b, ncol = 1, rel_heights = c(1, .1))
 g
 
 
-## ----plot-stratified-erspc-data-code, eval = FALSE, echo = TRUE---------------
+## ----plot-stratified-erspc-data-code, eval = FALSE, echo = TRUE----------------------------------------
 #> plot(pt_object, add.base.series = TRUE)
 
 
@@ -151,21 +151,21 @@ g
 #>   ylab("Population Size")
 
 
-## ---- eval = eval_cs1, echo = -1----------------------------------------------
+## ---- eval = eval_cs1, echo = -1-----------------------------------------------------------------------
 set.seed(1953)
 fit <- fitSmoothHazard(DeadOfPrCa ~ pspline(Follow.Up.Time, df = 2) * ScrArm, 
                        data = ERSPC, ratio = 100)
 
 
-## ---- eval = eval_cs1, echo = TRUE--------------------------------------------
+## ---- eval = eval_cs1, echo = TRUE---------------------------------------------------------------------
 summary(fit) 
 
 
[…]

### 2021-10-26 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd

@@
 In this article, we present the \CRANpkg{casebase} R package [@casebase-package] 
-which implements the Hanley & Miettinen [-@hanley2009fitting] approach for 
+which implements and extends the Hanley & Miettinen [-@hanley2009fitting] approach for 
 fitting fully parametric hazard models and covariate-adjusted survival curves 
 using the familiar interface of the \code{glm} function. Our implementation 
-allows for straightforward extensions to other models such as penalized 
+includes extensions to other models such as penalized 
 regression for variable selection and competing-risk analysis. In addition, 
@@
 The ultimate goal of our package is to make fitting flexible hazards accessible 
-to more end users with the hope that they will favor reporting absolute risks 
+to more end users with the hope that they will favor reporting absolute risks and survival curves
 over hazard ratios.
@@
 
-As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to consider the entire study base as an infinite collection of *person moments*.
-These person moments are indexed by both an individual in the study and a time point, and therefore each person moment has a covariate profile, an exposure status, and an outcome status (i.e.
+As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to consider the entire study base as an infinite collection of *person-moments*.
+These person-moments are indexed by both an individual in the study and a time point, and therefore each person-moment has a covariate profile, an exposure status, and an outcome status (i.e.
 whether the event happened) attached to it.
-By comparing person moments at which an event occurred with person moments at which no event occurred, we can extract information about hazards and survival.
-
-Therefore, we start by sampling all person moments at which the event occurred; this collection of person moments is what Hanley & Miettinen call the *case series*.
+By comparing person-moments at which an event occurred with person-moments at which no event occurred, we can extract information about hazards and survival.
+
+Therefore, we start by sampling all person-moments at which the event occurred; this collection of person-moments is what Hanley & Miettinen call the *case series*.
 The incidence of the case series is dictated by the hazard function of interest.
-Next, we sample a finite number of person moments at which no event occurred; this second collection of person moment is what Hanley & Miettinen call the *base series*.
+Next, we sample a finite number of person-moments at which no event occurred; this second collection of person-moment is what Hanley & Miettinen call the *base series*.
 The sampling mechanism for the base series is left at the discretion of the user, but in practice we find that sampling uniformly from the study base provides both simplicity and good performance.
@@
 We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
-$$\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].$$  The processes $N_i(t)$ count the person moments from the *case series*.
-
-To complement the case series, we sample person moments for the base series.
+$$\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].$$  The processes $N_i(t)$ count the person-moments from the *case series*.
+
+To complement the case series, we sample person-moments for the base series.
 To do so, we model the base series sampling mechanism using non-homogeneous Poisson processes $R_i(t) \in \{0, 1, 2, \ldots\}$; the person-moments where $dR_i(t) = 1$ constitute the base series.

### 2021-10-26 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex

@@
 In this article, we present the \CRANpkg{casebase} R package
-\citep{casebase-package} which implements the Hanley \& Miettinen
-\citeyearpar{hanley2009fitting} approach for fitting fully parametric
-hazard models and covariate-adjusted survival curves using the familiar
-interface of the \code{glm} function. Our implementation allows for
-straightforward extensions to other models such as penalized regression
-for variable selection and competing-risk analysis. In addition, we
-provide functions for exploratory data analysis and visualizing the
-estimated quantities such as the hazard function, survival curve, and
-their standard errors. The ultimate goal of our package is to make
-fitting flexible hazards accessible to more end users with the hope that
-they will favor reporting absolute risks over hazard ratios.
+\citep{casebase-package} which implements and extends the Hanley \&
+Miettinen \citeyearpar{hanley2009fitting} approach for fitting fully
+parametric hazard models and covariate-adjusted survival curves using
+the familiar interface of the \code{glm} function. Our implementation
+includes extensions to other models such as penalized regression for
+variable selection and competing-risk analysis. In addition, we provide
+functions for exploratory data analysis and visualizing the estimated
+quantities such as the hazard function, survival curve, and their
+standard errors. The ultimate goal of our package is to make fitting
+flexible hazards accessible to more end users with the hope that they
+will favor reporting absolute risks and survival curves over hazard
+ratios.
 
@@
 key idea behind case-base sampling is to consider the entire study base
-as an infinite collection of \emph{person moments}. These person moments
+as an infinite collection of \emph{person-moments}. These person-moments
 are indexed by both an individual in the study and a time point, and
-therefore each person moment has a covariate profile, an exposure
+therefore each person-moment has a covariate profile, an exposure
 status, and an outcome status (i.e.~whether the event happened) attached
-to it. By comparing person moments at which an event occurred with
-person moments at which no event occurred, we can extract information
+to it. By comparing person-moments at which an event occurred with
+person-moments at which no event occurred, we can extract information
 about hazards and survival.
 
-Therefore, we start by sampling all person moments at which the event
-occurred; this collection of person moments is what Hanley \& Miettinen
+Therefore, we start by sampling all person-moments at which the event
+occurred; this collection of person-moments is what Hanley \& Miettinen
 call the \emph{case series}. The incidence of the case series is
 dictated by the hazard function of interest. Next, we sample a finite
-number of person moments at which no event occurred; this second
-collection of person moment is what Hanley \& Miettinen call the
+number of person-moments at which no event occurred; this second
+collection of person-moment is what Hanley \& Miettinen call the
 \emph{base series}. The sampling mechanism for the base series is left
@@
 \[\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].\] The processes
-\(N_i(t)\) count the person moments from the \emph{case series}.
-
-To complement the case series, we sample person moments for the base
+\(N_i(t)\) count the person-moments from the \emph{case series}.
+
+To complement the case series, we sample person-moments for the base
 series. To do so, we model the base series sampling mechanism using

### 2021-11-04 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
add section on approximate bayes

@@
 
+## Variance estimates
+
+The asymptotic normality of $\hat{\theta}$ gives us an efficient way to estimate the variance of various estimators derived from the hazard function.
+For example, to construct confidence bands around risk functions and survival functions, we can use the following procedure:
+
+  1. Compute $\hat{\theta}$ and its variance-covariance matrix $V(\theta)$.
+  2. Sample $B$ times from a multivariate normal $N\left(\hat{\theta}, V(\theta)\right)$.
+  3. For each sample, compute the survival function using Equation \ref{eqn:surv} below.
+  4. Use the pointwise quantiles of these survival function estimates to construct a pointwise confidence band for the survival function.
+
+This procedure is similar to parametric bootstrap [@efron1994introduction, Section 6.5], but it can be more accurately described as a form of approximate Bayesian computation.
+As such, the validity of the confidence bands relies on the Bernstein-von Mises theorem [@van2000asymptotic, Chapter 10].
+
 ## Common parametric models
@@
 
-Finally, the output from \code{absoluteRisk} can be passed to a method \code{confint} to compute confidence bands around the survival function.
-These bands are computed using parametric bootstrap, and therefore are only valid when \code{family = "glm"} as it relies on the asymptotic normality of the estimator.
+Finally, the output from \code{absoluteRisk} can be passed to a method \code{confint} to compute confidence bands around the survival function, as described in the Theoretical Details section.
+These bands are only valid when \code{family = "glm"} as it relies on the asymptotic normality of the estimator.
 Currently, this is only available for the single-event setting.
@@
 
-These point-wise confidence intervals are constructed using parametric bootstrap; 
-the validity of this approach relies on the asymptotic normality of the MLE. In Figure \ref{fig:erspc-cif-conf}, we see the 95% confidence bands around the estimates. 
+In Figure \ref{fig:erspc-cif-conf}, we see the 95% confidence bands around the estimates. 
 We also overlay the Kaplan-Meier curves as a reference.

### 2021-11-04 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
add section on approximate bayes

@@
 increase the efficiency of the case-base estimators.
+
+\hypertarget{variance-estimates}{%
+\subsection{Variance estimates}\label{variance-estimates}}
+
+The asymptotic normality of \(\hat{\theta}\) gives us an efficient way
+to estimate the variance of various estimators derived from the hazard
+function. For example, to construct confidence bands around risk
+functions and survival functions, we can use the following procedure:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Compute \(\hat{\theta}\) and its variance-covariance matrix
+  \(V(\theta)\).
+\item
+  Sample \(B\) times from a multivariate normal
+  \(N\left(\hat{\theta}, V(\theta)\right)\).
+\item
+  For each sample, compute the survival function using Equation
+  \ref{eqn:surv} below.
+\item
+  Use the pointwise quantiles of these survival function estimates to
+  construct a pointwise confidence band for the survival function.
+\end{enumerate}
+
+This procedure is similar to parametric bootstrap \citep[Section
+6.5]{efron1994introduction}, but it can be more accurately described as
+a form of approximate Bayesian computation. As such, the validity of the
+confidence bands relies on the Bernstein-von Mises theorem
+\citep[Chapter 10]{van2000asymptotic}.
 
@@
 Finally, the output from \code{absoluteRisk} can be passed to a method
-\code{confint} to compute confidence bands around the survival function.
-These bands are computed using parametric bootstrap, and therefore are
-only valid when \code{family = "glm"} as it relies on the asymptotic
+\code{confint} to compute confidence bands around the survival function,
+as described in the Theoretical Details section. These bands are only
+valid when \code{family = "glm"} as it relies on the asymptotic
 normality of the estimator. Currently, this is only available for the
@@
 
-These point-wise confidence intervals are constructed using parametric
-bootstrap; the validity of this approach relies on the asymptotic
-normality of the MLE. In Figure \ref{fig:erspc-cif-conf}, we see the
-95\% confidence bands around the estimates. We also overlay the
-Kaplan-Meier curves as a reference.
+In Figure \ref{fig:erspc-cif-conf}, we see the 95\% confidence bands
+around the estimates. We also overlay the Kaplan-Meier curves as a
+reference.
 

### 2021-11-10 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
do not compare cases and bases

@@
 whether the event happened) attached to it.
-By comparing person-moments at which an event occurred with person-moments at which no event occurred, we can extract information about hazards and survival.
+By estimating the probability of an event occurring at a particular person-moment, we can extract information about hazards and survival.
 
@@
   3. For each sample, compute the survival function using Equation \ref{eqn:surv} below.
-  4. Use the pointwise quantiles of these survival function estimates to construct a pointwise confidence band for the survival function.
+  4. Use the pointwise quantiles of these survival function estimates to construct a pointwise confidence band for the survival function of interest.
 

### 2021-11-10 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
OS comments

--- 2021-10-26 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
@@
     address: >
-      1020 Pine Avenue West
-      Montreal, QC, Canada H3A 1A2
+      2001 McGill College Avenue
+      Montreal, QC, Canada H3A 1G1
     email: \email{sahir.bhatnagar@mcgill.ca}
@@
     address: >
-      1020 Pine Avenue West
-      Montreal, QC, Canada H3A 1A2
+      2001 McGill College Avenue
+      Montreal, QC, Canada H3A 1G1
     email: \email{jesse.islam@mail.mcgill.ca}
@@
     address: >
-      1020 Pine Avenue West
-      Montreal, QC, Canada H3A 1A2
+      2001 McGill College Avenue
+      Montreal, QC, Canada H3A 1G1
     email: \email{james.hanley@mcgill.ca}
@@
 abstract: |
-  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician is their patient's risk of an event.
+  | In clinical studies of time-to-event data, a quantity of interest to the clinician is their patient's risk of an event.
 However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the estimating function.
 As a consequence, the focus has been on reporting hazard ratios instead of survival curves.
-Indeed, reporting patient risk requires a separate estimation of the baseline hazard.
+Indeed, reporting patient risk or cumulative incidence requires a separate estimation of the baseline hazard.
 Using case-base sampling, Hanley & Miettinen (2009) explained how parametric hazard functions can be estimated in continuous-time using logistic regression.
-Their approach naturally leads to estimates of the survival function that are smooth-in-time.
+Their approach naturally leads to estimates of the survival or risk function that are smooth-in-time.
   | In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis.
-We describe how the case-base framework can be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection.
+We describe how the case-base framework can also be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection.
 Our package also includes an extensive array of visualization tools to complement the analysis.
@@
 experiencing a certain event given the patient's particular circumstances. 
-Unfortunately, the most reported metric from a Cox model is the (potentially 
-time-dependent) hazard ratio (HR), which ignores the duration of follow-up and 
-is subject to selection bias [@hernan2010hazards]. The covariate-adjusted 
-survival curve overcomes these limitations, and it is arguably a more important 
+However, the most reported metric from a Cox model is the (potentially 
+time-dependent) hazard ratio (HR). The covariate-conditional
+survival curve is arguably a more important 
 summary measure to report than the HR. While stepwise survival curves can be 
@@
 Several authors have since pursued fully parametric approaches that made the 
-fitting of smooth survival curves much more transparent and intuitive through 
+fitting of smooth survival curves more transparent and intuitive through 
 generalized linear models. The key feature of these procedures is splitting the 
@@
 equivalence between a Cox model and a Poisson regression with a parameter for 
-each event time; Carstensen [-@copenhagen2012needs] provides a nice exposition of this 
+each event time; Carstensen [-@copenhagen2012needs] provides an exposition of this 
 equivalence with a real data example and supporting R code for computing 
@@
 to instability in the Newton-Raphson procedure for estimation 
-[@kalbfleisch2011statistical, Section 4.8]. 
+[@kalbfleisch2011statistical, Section 4.8], and large long-format datasets.
 On the other hand, a coarse grouping could potentially mask 
@@
 study base. By doing so, they obtained a likelihood expression for the hazard 
-function that is equivalent to that of logistic regression. More specifically, all person-moments when the event of interest 
+function that is equivalent to that of logistic regression with an offset term. 
+More specifically, all person-moments when the event of interest 
 occurred are selected as the case series, complemented by a randomly sampled 
 base series of person-moments serving as controls. This approach allows 
-flexible modeling of the hazard function by including time as a covariate 
-(e.g. using splines or general additive models). Furthermore, time-dependent 
+flexible modeling of the hazard function by including functions of time as covariates 
+(e.g. using splines or generalized additive models). Furthermore, time-dependent 
 covariates can be modeled through interactions with time. In short, Hanley & 
@@
 which implements and extends the Hanley & Miettinen [-@hanley2009fitting] approach for 
-fitting fully parametric hazard models and covariate-adjusted survival curves 
+fitting fully parametric hazard models and covariate-conditional survival curves 
 using the familiar interface of the \code{glm} function. Our implementation 
 includes extensions to other models such as penalized 
-regression for variable selection and competing-risk analysis. In addition, 
+regression for variable selection and competing risk analysis. In addition, 
 we provide functions for exploratory data analysis and visualizing the estimated 
@@
 The ultimate goal of our package is to make fitting flexible hazards accessible 
-to more end users with the hope that they will favor reporting absolute risks and survival curves
+to end users who favor reporting absolute risks and survival curves
 over hazard ratios.
 
-In what follows, we first recall some theoretical details on case-base sampling 
+In what follows, we first recap some theoretical details on case-base sampling 
 and its use for estimating parametric hazard functions. We then give a short 
@@
 As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to consider the entire study base as an infinite collection of *person-moments*.
-These person-moments are indexed by both an individual in the study and a time point, and therefore each person-moment has a covariate profile, an exposure status, and an outcome status (i.e.
+These person-moments are indexed by both an individual in the study and a time point, and therefore each person-moment has a covariate profile and an outcome status (i.e.
 whether the event happened) attached to it.
@@
 First, let $N_{i}(t) \in \{0, 1\}$ be counting processes corresponding to the event of interest for individual $i=1, \ldots,n$.
-For simplicity, we will consider Type I censoring due to the end of follow-up at time $\tau$ (the general case of non-informative censoring is treated in Saarela [-@saarela2016case]).
+For simplicity, we will consider Type I censoring due to the end of follow-up at time $\tau$ (the general case of independent censoring is treated in Saarela [-@saarela2016case]).
 We assume a continuous time model, which implies that the counting process jumps are less than or equal to one.
 We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
-$$\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].$$  The processes $N_i(t)$ count the person-moments from the *case series*.
+$$\lambda_{i}(t)\,\textrm dt = E[\,\textrm dN_{i}(t)\mid \mathrm{past}].$$  The processes $N_i(t)$ count the person-moments from the *case series*.
 
 To complement the case series, we sample person-moments for the base series.
-To do so, we model the base series sampling mechanism using non-homogeneous Poisson processes $R_i(t) \in \{0, 1, 2, \ldots\}$; the person-moments where $dR_i(t) = 1$ constitute the base series.
+To do so, we specify the base series sampling mechanism as non-homogeneous Poisson processes $R_i(t) \in \{0, 1, 2, \ldots\}$; the person-moments where $\,\textrm dR_i(t) = 1$ constitute the base series.
 We note that the same individual can contribute multiple person-moments to the base series.
 The process $Q_{i}(t) = R_i(t) + N_{i}(t)$ then counts both the case and base series person-moments contributed by individual $i$.
-As mentioned above, the processes $R_i(t)$ are typically defined by the user via its intensity function $\rho_i(t)$.
-The process $Q_{i}(t)$ is characterized by $E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt$.
+As mentioned above, the processes $R_i(t)$ are specified by the user via its intensity function $\rho_i(t)$.
+The process $Q_{i}(t)$ is characterized by $E[\,\textrm dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)\,\textrm dt + \rho_i(t)\,\textrm dt$.
 
 If the hazard function $\lambda_{i}(t; \theta)$ is parametrized in terms of $\theta$, we can define an estimator $\hat{\theta}$ by maximization of the likelihood expression
-$$L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^{\min(t_i,\tau)} \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)},$$
+$$L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^{\min(t_i,\tau)} \lambda_i(t; \theta)\,\textrm dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{\,\textrm dN_{i}(t)},$$
 where $\prod_{t\in[0,u)}$ represents a product integral from $0$ to $u$, and where $t_i$ is the event time for individual $i$.
-However, the integral over time makes the computation and maximization of $L_0(\theta)$ challenging.
-
-Case-base sampling allows us to avoid this integral. By conditioning on a sampled person-moment, we get individual likelihood contributions of the form
-$$P(dN_{i}(t) \mid dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.$$
-Therefore, we can define an estimating function for $\theta$ as follows:
+However, the integral over time makes the computation and maximization of $L_0(\theta)$ challenging using standard software.
+
+Case-base sampling allows us to avoid this integral. By conditioning on a sampled person-moment, we get individual partial likelihood contributions of the form
+$$P(\,\textrm dN_{i}(t) \mid \,\textrm dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.$$
+Therefore, an estimating function for $\theta$ can be composed of these contributions as:
 \begin{equation}
-L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}. \label{eq:lik-function}
+L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{\,\textrm dQ_i(t)}. \label{eq:lik-function}
 \end{equation}
@@
 Generally, an individual can contribute more than one person-moment, and thus the terms in the product integral are not independent.
-Nonetheless, Saarela [-@saarela2016case] showed that the logarithm of this estimating function has mean zero at the true value $\theta=\theta_0$, and that the resulting estimator $\hat{\theta}$ is asymptotically normally distributed.
+Nonetheless, Saarela [-@saarela2016case] showed that the corresponding partial likelihood score function has mean zero at the true value $\theta=\theta_0$, and that the resulting estimator $\hat{\theta}$ is asymptotically normally distributed.
 
 In Hanley & Miettinen [-@hanley2009fitting], the authors suggest sampling the base series *uniformly* from the study base.
-In terms of Poisson processes, their sampling strategy corresponds essentially to a time-homogeneous Poisson process with hazard equal to $\rho_i(t) = b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base (e.g.
+In terms of Poisson processes, their sampling strategy corresponds to a time-homogeneous Poisson process with intensity equal to $\rho_i(t) = b/B$, where $b$ is the number of sampled base series *person-moments*, and $B$ is the total population-time for the study base (e.g.
 the sum of all individual follow-up times).
 More complex examples are also possible; see for example Saarela & Arjas [-@saarela2015non], where the intensity functions for the sampling mechanism are proportional to the cardiovascular disease event rate given by the Framingham score.
-Non-uniform sampling mechanisms can increase the efficiency of the case-base estimators.
+Non-uniform sampling mechanisms can increase the efficiency of the resulting maximum partial likelihood estimators.
 
@@
 
-## Competing-risk analysis
-
-Case-base sampling can also be used in the context of competing-risk analysis.
-Assuming there are $J$ competing events, we can show that each person-moment's contribution to the likelihood is of the form
-
-$$\frac{\lambda_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},$$
+## Competing risk analysis
+
+Case-base sampling can also be used in the context of competing risk analysis.
+Assuming there are $J$ competing events, we can show that each sampled person-moment's contribution to the partial likelihood is of the form
+
+$$\frac{\lambda_j(t)^{\,\textrm dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},$$
 where $N_j(t)$ is the counting process associated with the event of type $j$ and $\lambda_j(t)$ is the corresponding cause-specific hazard function.
-As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}
+As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression with an offset term.\footnote{Specifically, it corresponds to the following parametrization  for a multinomial random variable $Y$: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j + log(1/\rho), \qquad j = 1,\ldots, J-1.\end{align*}}
 
@@
 \begin{equation}
-\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j P(\theta_j;\lambda,\alpha) \label{eq:penest}
+\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j p_{\lambda,\alpha}(\theta_j) \label{eq:penest}
 \end{equation}
-where $\ell\left(\theta\right) = \log L(\theta)$ is the log of the likelihood function given in \eqref{eq:lik-function}, $P(\theta_j;\lambda,\alpha)$ is a penalty term controlled by the non-negative regularization parameters $\lambda$ and $\alpha$, and $w_j$ is the penalty factor for the $j$th covariate.
+where $\ell\left(\theta\right) = \log L(\theta)$ is the log of the likelihood function given in \eqref{eq:lik-function}, $p_{\lambda,\alpha}(\theta_j)$ is a penalty term controlled by the non-negative regularization parameters $\lambda$ and $\alpha$, and $w_j$ is the penalty factor for the $j$th covariate.
 These penalty factors serve as a way of allowing parameters to be penalized differently.
@@
 Next, several R packages implement methodologies for competing risk analysis; for a different perspective on this topic, see Mahani & Sharabiani [-@mahani2015bayesian].
-The package \pkg{survival} provides functionality for competing-risk analysis and multistate modelling.
+The package \pkg{survival} provides functionality for competing risk analysis and multistate modelling.
 The package \CRANpkg{cmprsk} provides methods for cause-specific subdistribution hazards, such as in the Fine-Gray model [-@fine1999proportional].
@@
 Packages providing this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and \pkg{survival}.
-Our package \pkg{casebase} also follows this approach for both single-event and competing-risk analyses.
+Our package \pkg{casebase} also follows this approach for both single-event and competing risk analyses.
 The second category outlined above broadly corresponds to semi-parametric models.
@@
 $$ S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right),\qquad \lambda(t;X) = \sum_{j=1}^J \lambda_j(t;X).$$
-From this, we can derive the event-specific subdensities:
+From this, we can derive the cause-specific subdensities:
 
@@
 
-By integrating these subdensities, we obtain the event-specific CI functions:
+By integrating these subdensities, we obtain the cause-specific CI functions:
 
@@
 First, we show how to model non-constant and non-proportional hazards through a flexible specification of time.
-Then we perform a competing-risk analysis and compare our results with the Cox model and the Fine-Gray model.
+Then we perform a competing risk analysis and compare our results with the Cox model and the Fine-Gray model.
 The third case study illustrates how to perform variable selection in high-dimensional datasets.
@@
 $$
-\frac{\lambda\left(t | X=x_1, \mathbf{Z}=\mathbf{z_1} ; \hat{\beta}\right)}{\lambda(t | X=x_0, \mathbf{Z}=\mathbf{z_0} ; \hat{\beta})}
+\frac{\lambda\left(t | X=x_1, \mathbf{Z}=\mathbf{z} ; \hat{\beta}\right)}{\lambda(t | X=x_0, \mathbf{Z}=\mathbf{z} ; \hat{\beta})}
 $$
@@
 the unexposed group, $\mathbf{Z}$ are other covariates in the model which are 
-equal to $\mathbf{z_1}$ in the exposed and $\mathbf{z_0}$ in the unexposed group, 
+equal to $\mathbf{z}$ in both the exposed and unexposed group, 
 and $\hat{\beta}$ are the estimated regression coefficients. As indicated by the 
@@
 
-In this case study, we show how case-base sampling can be used in the context of a competing-risk analysis.
+In this case study, we show how case-base sampling can be used in the context of a competing risk analysis.
 For illustrative purposes, we use the same data that was used in Scrucca *et al* [-@scrucca2010regression].

### 2021-11-10 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd

@@
 As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to consider the entire study base as an infinite collection of *person-moments*.
-These person-moments are indexed by both an individual in the study and a time point, and therefore each person-moment has a covariate profile and an outcome status (i.e.
+These person-moments are indexed by both an individual in the study and a time point, and therefore each person-moment has a covariate profile, an exposure status, and an outcome status (i.e.
 whether the event happened) attached to it.
-By comparing person-moments at which an event occurred with person-moments at which no event occurred, we can extract information about hazards and survival.
+By estimating the probability of an event occurring at a particular person-moment, we can extract information about hazards and survival.
 
@@
 Non-uniform sampling mechanisms can increase the efficiency of the resulting maximum partial likelihood estimators.
+
+## Variance estimates
+
+The asymptotic normality of $\hat{\theta}$ gives us an efficient way to estimate the variance of various estimators derived from the hazard function.
+For example, to construct confidence bands around risk functions and survival functions, we can use the following procedure:
+
+  1. Compute $\hat{\theta}$ and its variance-covariance matrix $V(\theta)$.
+  2. Sample $B$ times from a multivariate normal $N\left(\hat{\theta}, V(\theta)\right)$.
+  3. For each sample, compute the survival function using Equation \ref{eqn:surv} below.
+  4. Use the pointwise quantiles of these survival function estimates to construct a pointwise confidence band for the survival function of interest.
+
+This procedure is similar to parametric bootstrap [@efron1994introduction, Section 6.5], but it can be more accurately described as a form of approximate Bayesian computation.
+As such, the validity of the confidence bands relies on the Bernstein-von Mises theorem [@van2000asymptotic, Chapter 10].
 

### 2021-11-10 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex

@@
 status, and an outcome status (i.e.~whether the event happened) attached
-to it. By comparing person-moments at which an event occurred with
-person-moments at which no event occurred, we can extract information
-about hazards and survival.
+to it. By estimating the probability of an event occurring at a
+particular person-moment, we can extract information about hazards and
+survival.
 
@@
   Use the pointwise quantiles of these survival function estimates to
-  construct a pointwise confidence band for the survival function.
+  construct a pointwise confidence band for the survival function of
+  interest.
 \end{enumerate}

### 2021-11-10 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
included comments by OS and JH.
was not able to constrain the ERSPC hazard ratio so that it constantly descreases

@@
 However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the estimating function.
-As a consequence, the focus has been on reporting hazard ratios instead of survival curves.
+As a consequence, the focus has been on reporting hazard ratios instead of survival or cumulative incidence curves.
 Indeed, reporting patient risk or cumulative incidence requires a separate estimation of the baseline hazard.
@@
 relevant quantity in a clinical setting is often the 5- or 10-year risk of 
-experiencing a certain event given the patient's particular circumstances. 
+experiencing a certain event given the patient's particular profile. 
 However, the most reported metric from a Cox model is the (potentially 
@@
 and its use for estimating parametric hazard functions. We then give a short 
-review of existing R packages that implement comparable features as \pkg{casebase}. 
+review of existing R packages that implement comparable features to \pkg{casebase}. 
 Next, we provide some details about the implementation of case-base sampling in 
@@
 
-Therefore, we start by sampling all person-moments at which the event occurred; this collection of person-moments is what Hanley & Miettinen call the *case series*.
+Therefore, we start by assembling all person-moments at which the event occurred; this collection of person-moments is what Hanley & Miettinen call the *case series*.
 The incidence of the case series is dictated by the hazard function of interest.
-Next, we sample a finite number of person-moments at which no event occurred; this second collection of person-moment is what Hanley & Miettinen call the *base series*.
+Next, we sample a finite number of person-moments (blinded to case moments); this second collection of person-moments is what Hanley & Miettinen call the *base series*.
 The sampling mechanism for the base series is left at the discretion of the user, but in practice we find that sampling uniformly from the study base provides both simplicity and good performance.
@@
 
+Here we show that certain named distributions such as exponential, Weibull or Gompertz can be fit using our framework, though we are not restricted to such models.
 Let $g(t; X)$ be the linear predictor such that $\log(\lambda(t;X)) = g(t; X)$.
@@
 
-Population-time plots are a descriptive visualization of incidence density, where the study base is represented by area and events by points within the area.
+Population-time plots are a descriptive visualization of incidence density, where the population time that constitutes the study base is represented by area, and events by points within the area.
 The case-base sampling approach described above can be visualized in the form of a population time plot.
@@
 This dataset is available through the \pkg{casebase} package.
-It contains the individual observations for 159,893 men  from seven European countries, who were between the ages of 55 and 69 years when recruited for the trial.
+It contains the recreated individual observations for 159,893 men  from seven European countries, who were between the ages of 55 and 69 years when recruited for the trial.
 
@@
 fit <- fitSmoothHazard(DeadOfPrCa ~ pspline(Follow.Up.Time, df = 2) * ScrArm, 
-                       data = ERSPC, ratio = 100)
+                       data = ERSPC, ratio = 10)
 ```

### 2021-11-10 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
included comments by OS and JH.
was not able to constrain the ERSPC hazard ratio so that it constantly descreases

@@
 \abstract{%
-In epidemiological studies of time-to-event data, a quantity of interest
-to the clinician is their patient's risk of an event. However, methods
-relying on time matching or risk-set sampling (including Cox regression)
+In clinical studies of time-to-event data, a quantity of interest to the
+clinician is their patient's risk of an event. However, methods relying
+on time matching or risk-set sampling (including Cox regression)
 eliminate the baseline hazard from the estimating function. As a
 consequence, the focus has been on reporting hazard ratios instead of
-survival curves. Indeed, reporting patient risk requires a separate
-estimation of the baseline hazard. Using case-base sampling, Hanley \&
-Miettinen (2009) explained how parametric hazard functions can be
-estimated in continuous-time using logistic regression. Their approach
-naturally leads to estimates of the survival function that are
-smooth-in-time.\\
+survival or cumulative incidence curves. Indeed, reporting patient risk
+or cumulative incidence requires a separate estimation of the baseline
+hazard. Using case-base sampling, Hanley \& Miettinen (2009) explained
+how parametric hazard functions can be estimated in continuous-time
+using logistic regression. Their approach naturally leads to estimates
+of the survival or risk function that are smooth-in-time.\\
 In this paper, we present the \CRANpkg{casebase} R package, a
 comprehensive and flexible toolkit for parametric survival analysis. We
-describe how the case-base framework can be used in more complex
+describe how the case-base framework can also be used in more complex
 settings: non-linear functions of time and non-proportional hazards,
@@
 or 10-year risk of experiencing a certain event given the patient's
-particular circumstances. Unfortunately, the most reported metric from a
-Cox model is the (potentially time-dependent) hazard ratio (HR), which
-ignores the duration of follow-up and is subject to selection bias
-\citep{hernan2010hazards}. The covariate-adjusted survival curve
-overcomes these limitations, and it is arguably a more important summary
-measure to report than the HR. While stepwise survival curves can be
-computed with the Cox model, they require a second step to separately
-estimate the baseline hazard \citep{breslow1972discussion}.
+particular profile. However, the most reported metric from a Cox model
+is the (potentially time-dependent) hazard ratio (HR). The
+covariate-conditional survival curve is arguably a more important
+summary measure to report than the HR. While stepwise survival curves
+can be computed with the Cox model, they require a second step to
+separately estimate the baseline hazard \citep{breslow1972discussion}.
 
 Several authors have since pursued fully parametric approaches that made
-the fitting of smooth survival curves much more transparent and
-intuitive through generalized linear models. The key feature of these
-procedures is splitting the time axis into discrete intervals. Whitehead
+the fitting of smooth survival curves more transparent and intuitive
+through generalized linear models. The key feature of these procedures
+is splitting the time axis into discrete intervals. Whitehead
 \citeyearpar{whitehead1980fitting} showed the equivalence between a Cox
 model and a Poisson regression with a parameter for each event time;
-Carstensen \citeyearpar{copenhagen2012needs} provides a nice exposition
-of this equivalence with a real data example and supporting R code for
+Carstensen \citeyearpar{copenhagen2012needs} provides an exposition of
+this equivalence with a real data example and supporting R code for
 computing standard errors. Arjas \& Haara
@@
 leads to instability in the Newton-Raphson procedure for estimation
-\citep[Section 4.8]{kalbfleisch2011statistical}. On the other hand, a
-coarse grouping could potentially mask nonlinear trends in the hazard
-function.
+\citep[Section 4.8]{kalbfleisch2011statistical}, and large long-format
+datasets. On the other hand, a coarse grouping could potentially mask
+nonlinear trends in the hazard function.
 
@@
 obtained a likelihood expression for the hazard function that is
-equivalent to that of logistic regression. More specifically, all
-person-moments when the event of interest occurred are selected as the
-case series, complemented by a randomly sampled base series of
-person-moments serving as controls. This approach allows flexible
-modeling of the hazard function by including time as a covariate
-(e.g.~using splines or general additive models). Furthermore,
-time-dependent covariates can be modeled through interactions with time.
-In short, Hanley \& Miettinen \citeyearpar{hanley2009fitting} use the
-well-understood logistic regression for directly modeling the hazard
-function, without requiring a discrete-time model.
+equivalent to that of logistic regression with an offset term. More
+specifically, all person-moments when the event of interest occurred are
+selected as the case series, complemented by a randomly sampled base
+series of person-moments serving as controls. This approach allows
+flexible modeling of the hazard function by including functions of time
+as covariates (e.g.~using splines or generalized additive models).
+Furthermore, time-dependent covariates can be modeled through
+interactions with time. In short, Hanley \& Miettinen
+\citeyearpar{hanley2009fitting} use the well-understood logistic
+regression for directly modeling the hazard function, without requiring
+a discrete-time model.
 
@@
 Miettinen \citeyearpar{hanley2009fitting} approach for fitting fully
-parametric hazard models and covariate-adjusted survival curves using
+parametric hazard models and covariate-conditional survival curves using
 the familiar interface of the \code{glm} function. Our implementation
 includes extensions to other models such as penalized regression for
-variable selection and competing-risk analysis. In addition, we provide
+variable selection and competing risk analysis. In addition, we provide
 functions for exploratory data analysis and visualizing the estimated
@@
 standard errors. The ultimate goal of our package is to make fitting
-flexible hazards accessible to more end users with the hope that they
-will favor reporting absolute risks and survival curves over hazard
-ratios.
-
-In what follows, we first recall some theoretical details on case-base
+flexible hazards accessible to end users who favor reporting absolute
+risks and survival curves over hazard ratios.
+
+In what follows, we first recap some theoretical details on case-base
 sampling and its use for estimating parametric hazard functions. We then
 give a short review of existing R packages that implement comparable
-features as \pkg{casebase}. Next, we provide some details about the
+features to \pkg{casebase}. Next, we provide some details about the
 implementation of case-base sampling in our package, and we give a brief
@@
 are indexed by both an individual in the study and a time point, and
-therefore each person-moment has a covariate profile, an exposure
-status, and an outcome status (i.e.~whether the event happened) attached
-to it. By estimating the probability of an event occurring at a
-particular person-moment, we can extract information about hazards and
-survival.
-
-Therefore, we start by sampling all person-moments at which the event
+therefore each person-moment has a covariate profile and an outcome
+status (i.e.~whether the event happened) attached to it. By estimating
+the probability of an event occurring at a particular person-moment, we
+can extract information about hazards and survival.
+
+Therefore, we start by assembling all person-moments at which the event
 occurred; this collection of person-moments is what Hanley \& Miettinen
@@
 dictated by the hazard function of interest. Next, we sample a finite
-number of person-moments at which no event occurred; this second
-collection of person-moment is what Hanley \& Miettinen call the
+number of person-moments (blinded to case moments); this second
+collection of person-moments is what Hanley \& Miettinen call the
 \emph{base series}. The sampling mechanism for the base series is left
@@
 to the end of follow-up at time \(\tau\) (the general case of
-non-informative censoring is treated in Saarela
+independent censoring is treated in Saarela
 \citeyearpar{saarela2016case}). We assume a continuous time model, which
@@
 the processes \(N_i(t)\), and which satisfy
-\[\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].\] The processes
-\(N_i(t)\) count the person-moments from the \emph{case series}.
+\[\lambda_{i}(t)\,\textrm dt = E[\,\textrm dN_{i}(t)\mid \mathrm{past}].\]
+The processes \(N_i(t)\) count the person-moments from the \emph{case
+series}.
 
 To complement the case series, we sample person-moments for the base
-series. To do so, we model the base series sampling mechanism using
+series. To do so, we specify the base series sampling mechanism as
 non-homogeneous Poisson processes \(R_i(t) \in \{0, 1, 2, \ldots\}\);
-the person-moments where \(dR_i(t) = 1\) constitute the base series. We
-note that the same individual can contribute multiple person-moments to
-the base series. The process \(Q_{i}(t) = R_i(t) + N_{i}(t)\) then
-counts both the case and base series person-moments contributed by
-individual \(i\). As mentioned above, the processes \(R_i(t)\) are
-typically defined by the user via its intensity function \(\rho_i(t)\).
-The process \(Q_{i}(t)\) is characterized by
-\(E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt\).
+the person-moments where \(\,\textrm dR_i(t) = 1\) constitute the base
+series. We note that the same individual can contribute multiple
+person-moments to the base series. The process
+\(Q_{i}(t) = R_i(t) + N_{i}(t)\) then counts both the case and base
+series person-moments contributed by individual \(i\). As mentioned
+above, the processes \(R_i(t)\) are specified by the user via its
+intensity function \(\rho_i(t)\). The process \(Q_{i}(t)\) is
+characterized by
+\(E[\,\textrm dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)\,\textrm dt + \rho_i(t)\,\textrm dt\).
 
@@
 maximization of the likelihood expression
-\[L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^{\min(t_i,\tau)} \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)},\]
+\[L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^{\min(t_i,\tau)} \lambda_i(t; \theta)\,\textrm dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{\,\textrm dN_{i}(t)},\]
 where \(\prod_{t\in[0,u)}\) represents a product integral from \(0\) to
@@
 However, the integral over time makes the computation and maximization
-of \(L_0(\theta)\) challenging.
+of \(L_0(\theta)\) challenging using standard software.
 
 Case-base sampling allows us to avoid this integral. By conditioning on
-a sampled person-moment, we get individual likelihood contributions of
-the form
-\[P(dN_{i}(t) \mid dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.\]
-Therefore, we can define an estimating function for \(\theta\) as
-follows: \begin{equation}
-L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}. \label{eq:lik-function}
+a sampled person-moment, we get individual partial likelihood
+contributions of the form
+\[P(\,\textrm dN_{i}(t) \mid \,\textrm dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.\]
+Therefore, an estimating function for \(\theta\) can be composed of
+these contributions as: \begin{equation}
+L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{\,\textrm dQ_i(t)}. \label{eq:lik-function}
 \end{equation} When a logarithmic link function is used for modeling the
@@
 product integral are not independent. Nonetheless, Saarela
-\citeyearpar{saarela2016case} showed that the logarithm of this
-estimating function has mean zero at the true value \(\theta=\theta_0\),
-and that the resulting estimator \(\hat{\theta}\) is asymptotically
-normally distributed.
+\citeyearpar{saarela2016case} showed that the corresponding partial
+likelihood score function has mean zero at the true value
+\(\theta=\theta_0\), and that the resulting estimator \(\hat{\theta}\)
+is asymptotically normally distributed.
 
@@
 suggest sampling the base series \emph{uniformly} from the study base.
-In terms of Poisson processes, their sampling strategy corresponds
-essentially to a time-homogeneous Poisson process with hazard equal to
-\(\rho_i(t) = b/B\), where \(b\) is the number of sampled observations
-in the base series, and \(B\) is the total population-time for the study
-base (e.g.~the sum of all individual follow-up times). More complex
-examples are also possible; see for example Saarela \& Arjas
+In terms of Poisson processes, their sampling strategy corresponds to a
+time-homogeneous Poisson process with intensity equal to
+\(\rho_i(t) = b/B\), where \(b\) is the number of sampled base series
+\emph{person-moments}, and \(B\) is the total population-time for the
+study base (e.g.~the sum of all individual follow-up times). More
+complex examples are also possible; see for example Saarela \& Arjas
 \citeyearpar{saarela2015non}, where the intensity functions for the
@@
 rate given by the Framingham score. Non-uniform sampling mechanisms can
-increase the efficiency of the case-base estimators.
+increase the efficiency of the resulting maximum partial likelihood
+estimators.
 
@@
 
-Let \(g(t; X)\) be the linear predictor such that
-\(\log(\lambda(t;X)) = g(t; X)\). Different functions of \(t\) lead to
-different parametric hazard models. The simplest of these models is the
-one-parameter exponential distribution which is obtained by taking the
-hazard function to be constant over the range of \(t\):
+Here we show that certain named distributions such as exponential,
+Weibull or Gompertz can be fit using our framework, though we are not
+restricted to such models. Let \(g(t; X)\) be the linear predictor such
+that \(\log(\lambda(t;X)) = g(t; X)\). Different functions of \(t\) lead
+to different parametric hazard models. The simplest of these models is
+the one-parameter exponential distribution which is obtained by taking
+the hazard function to be constant over the range of \(t\):
 
@@
 \hypertarget{competing-risk-analysis}{%
-\subsection{Competing-risk analysis}\label{competing-risk-analysis}}
-
-Case-base sampling can also be used in the context of competing-risk
+\subsection{Competing risk analysis}\label{competing-risk-analysis}}
+
+Case-base sampling can also be used in the context of competing risk
 analysis. Assuming there are \(J\) competing events, we can show that
-each person-moment's contribution to the likelihood is of the form
-
-\[\frac{\lambda_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},\]
+each sampled person-moment's contribution to the partial likelihood is
+of the form
+
+\[\frac{\lambda_j(t)^{\,\textrm dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},\]
 where \(N_j(t)\) is the counting process associated with the event of
@@
 the terms appearing in the likelihood function for multinomial
-regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}
+regression with an offset
+term.\footnote{Specifically, it corresponds to the following parametrization  for a multinomial random variable $Y$: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j + log(1/\rho), \qquad j = 1,\ldots, J-1.\end{align*}}
 
@@
 to the likelihood and optimise the following equation: \begin{equation}
-\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j P(\theta_j;\lambda,\alpha) \label{eq:penest}
+\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j p_{\lambda,\alpha}(\theta_j) \label{eq:penest}
 \end{equation} where \(\ell\left(\theta\right) = \log L(\theta)\) is the
 log of the likelihood function given in \eqref{eq:lik-function},
-\(P(\theta_j;\lambda,\alpha)\) is a penalty term controlled by the
+\(p_{\lambda,\alpha}(\theta_j)\) is a penalty term controlled by the
 non-negative regularization parameters \(\lambda\) and \(\alpha\), and
@@
 Sharabiani \citeyearpar{mahani2015bayesian}. The package \pkg{survival}
-provides functionality for competing-risk analysis and multistate
+provides functionality for competing risk analysis and multistate
 modelling. The package \CRANpkg{cmprsk} provides methods for
@@
 \pkg{survival}. Our package \pkg{casebase} also follows this approach
-for both single-event and competing-risk analyses. The second category
+for both single-event and competing risk analyses. The second category
 outlined above broadly corresponds to semi-parametric models. These
@@
 Population-time plots are a descriptive visualization of incidence
-density, where the study base is represented by area and events by
-points within the area. The case-base sampling approach described above
-can be visualized in the form of a population time plot. These plots are
-informative graphical displays of survival data and should be one of the
-first steps in an exploratory data analysis. The \code{popTime} function
-and \code{plot} method facilitate this task:
+density, where the population time that constitutes the study base is
+represented by area, and events by points within the area. The case-base
+sampling approach described above can be visualized in the form of a
+population time plot. These plots are informative graphical displays of
+survival data and should be one of the first steps in an exploratory
+data analysis. The \code{popTime} function and \code{plot} method
+facilitate this task:
 
@@
 dataset is available through the \pkg{casebase} package. It contains the
-individual observations for 159,893 men from seven European countries,
-who were between the ages of 55 and 69 years when recruited for the
-trial.
+recreated individual observations for 159,893 men from seven European
+countries, who were between the ages of 55 and 69 years when recruited
+for the trial.
 
@@
 \[ S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right),\qquad \lambda(t;X) = \sum_{j=1}^J \lambda_j(t;X).\]
-From this, we can derive the event-specific subdensities:
+From this, we can derive the cause-specific subdensities:
 
@@
 
-By integrating these subdensities, we obtain the event-specific CI
+By integrating these subdensities, we obtain the cause-specific CI
 functions:
@@
 non-proportional hazards through a flexible specification of time. Then
-we perform a competing-risk analysis and compare our results with the
+we perform a competing risk analysis and compare our results with the
 Cox model and the Fine-Gray model. The third case study illustrates how
@@
 fit <- fitSmoothHazard(DeadOfPrCa ~ pspline(Follow.Up.Time, df = 2) * ScrArm, 
-                       data = ERSPC, ratio = 100)
+                       data = ERSPC, ratio = 10)
 \end{Sinput}
@@
 #> Number of events: 540 
-#> Number of base moments: 54000 
+#> Number of base moments: 5400 
 #> ----
@@
 #> fitSmoothHazard(formula = DeadOfPrCa ~ pspline(Follow.Up.Time, 
-#>     df = 2) * ScrArm, data = ERSPC, ratio = 100)
+#>     df = 2) * ScrArm, data = ERSPC, ratio = 10)
 #> 
@@
 #>    Min      1Q  Median      3Q     Max  
-#> -0.477  -0.164  -0.140  -0.084   3.882  
+#> -1.168  -0.486  -0.414  -0.215   3.262  
 #> 
@@
 #>                                                        Estimate Std. Error
-#> (Intercept)                                             -12.613      9.756
-#> pspline(Follow.Up.Time, df = 2)1                          1.320     10.705
-#> pspline(Follow.Up.Time, df = 2)2                          5.190      9.526
-#> pspline(Follow.Up.Time, df = 2)3                          4.481      9.866
-#> pspline(Follow.Up.Time, df = 2)4                          6.060      9.690
-#> pspline(Follow.Up.Time, df = 2)5                          5.231      9.851
-#> pspline(Follow.Up.Time, df = 2)6                          9.176      9.749
-#> pspline(Follow.Up.Time, df = 2)7                         -0.434     19.572
-#> ScrArmScreening group                                     7.900     13.050
-#> pspline(Follow.Up.Time, df = 2)1:ScrArmScreening group   -7.962     14.507
-#> pspline(Follow.Up.Time, df = 2)2:ScrArmScreening group   -8.351     12.690
-#> pspline(Follow.Up.Time, df = 2)3:ScrArmScreening group   -7.787     13.224
-#> pspline(Follow.Up.Time, df = 2)4:ScrArmScreening group   -8.098     12.942
-#> pspline(Follow.Up.Time, df = 2)5:ScrArmScreening group   -9.389     13.223
-#> pspline(Follow.Up.Time, df = 2)6:ScrArmScreening group   -7.385     13.114
-#> pspline(Follow.Up.Time, df = 2)7:ScrArmScreening group  -17.380     30.806
+#> (Intercept)                                              -13.81       9.98
+#> pspline(Follow.Up.Time, df = 2)1                           2.66      10.96
+#> pspline(Follow.Up.Time, df = 2)2                           6.43       9.73
+#> pspline(Follow.Up.Time, df = 2)3                           5.57      10.10
+#> pspline(Follow.Up.Time, df = 2)4                           7.27       9.90
+#> pspline(Follow.Up.Time, df = 2)5                           6.54      10.10
+#> pspline(Follow.Up.Time, df = 2)6                          10.82      10.03
+#> pspline(Follow.Up.Time, df = 2)7                         -11.74      30.13
+#> ScrArmScreening group                                      9.22      13.35
+#> pspline(Follow.Up.Time, df = 2)1:ScrArmScreening group    -9.25      14.85
+#> pspline(Follow.Up.Time, df = 2)2:ScrArmScreening group    -9.80      12.97
+#> pspline(Follow.Up.Time, df = 2)3:ScrArmScreening group    -9.04      13.54
+#> pspline(Follow.Up.Time, df = 2)4:ScrArmScreening group    -9.39      13.23
+#> pspline(Follow.Up.Time, df = 2)5:ScrArmScreening group   -10.65      13.55
+#> pspline(Follow.Up.Time, df = 2)6:ScrArmScreening group    -8.86      13.46
+#> pspline(Follow.Up.Time, df = 2)7:ScrArmScreening group   -11.58      36.92
 #>                                                        z value Pr(>|z|)
-#> (Intercept)                                              -1.29     0.20
-#> pspline(Follow.Up.Time, df = 2)1                          0.12     0.90
-#> pspline(Follow.Up.Time, df = 2)2                          0.54     0.59
-#> pspline(Follow.Up.Time, df = 2)3                          0.45     0.65
-#> pspline(Follow.Up.Time, df = 2)4                          0.63     0.53
-#> pspline(Follow.Up.Time, df = 2)5                          0.53     0.60
-#> pspline(Follow.Up.Time, df = 2)6                          0.94     0.35
-#> pspline(Follow.Up.Time, df = 2)7                         -0.02     0.98
-#> ScrArmScreening group                                     0.61     0.54
-#> pspline(Follow.Up.Time, df = 2)1:ScrArmScreening group   -0.55     0.58
-#> pspline(Follow.Up.Time, df = 2)2:ScrArmScreening group   -0.66     0.51
-#> pspline(Follow.Up.Time, df = 2)3:ScrArmScreening group   -0.59     0.56
-#> pspline(Follow.Up.Time, df = 2)4:ScrArmScreening group   -0.63     0.53
-#> pspline(Follow.Up.Time, df = 2)5:ScrArmScreening group   -0.71     0.48
-#> pspline(Follow.Up.Time, df = 2)6:ScrArmScreening group   -0.56     0.57
-#> pspline(Follow.Up.Time, df = 2)7:ScrArmScreening group   -0.56     0.57
+#> (Intercept)                                              -1.38     0.17
+#> pspline(Follow.Up.Time, df = 2)1                          0.24     0.81
+#> pspline(Follow.Up.Time, df = 2)2                          0.66     0.51
+#> pspline(Follow.Up.Time, df = 2)3                          0.55     0.58
+#> pspline(Follow.Up.Time, df = 2)4                          0.73     0.46
+#> pspline(Follow.Up.Time, df = 2)5                          0.65     0.52
+#> pspline(Follow.Up.Time, df = 2)6                          1.08     0.28
+#> pspline(Follow.Up.Time, df = 2)7                         -0.39     0.70
+#> ScrArmScreening group                                     0.69     0.49
+#> pspline(Follow.Up.Time, df = 2)1:ScrArmScreening group   -0.62     0.53
+#> pspline(Follow.Up.Time, df = 2)2:ScrArmScreening group   -0.76     0.45
+#> pspline(Follow.Up.Time, df = 2)3:ScrArmScreening group   -0.67     0.50
+#> pspline(Follow.Up.Time, df = 2)4:ScrArmScreening group   -0.71     0.48
+#> pspline(Follow.Up.Time, df = 2)5:ScrArmScreening group   -0.79     0.43
+#> pspline(Follow.Up.Time, df = 2)6:ScrArmScreening group   -0.66     0.51
+#> pspline(Follow.Up.Time, df = 2)7:ScrArmScreening group   -0.31     0.75
 #> 
@@
 #> 
-#>     Null deviance: 6059.0  on 54539  degrees of freedom
-#> Residual deviance: 5772.7  on 54524  degrees of freedom
-#> AIC: 5805
+#>     Null deviance: 3619.1  on 5939  degrees of freedom
+#> Residual deviance: 3359.1  on 5924  degrees of freedom
+#> AIC: 3391
 #> 
-#> Number of Fisher Scoring iterations: 9
+#> Number of Fisher Scoring iterations: 7
 \end{Soutput}
@@
 #>                                        Df Deviance Resid. Df Resid. Dev
-#> NULL                                                   54539       6059
-#> pspline(Follow.Up.Time, df = 2)         7    269.1     54532       5790
-#> ScrArm                                  1      8.0     54531       5782
-#> pspline(Follow.Up.Time, df = 2):ScrArm  7      9.1     54524       5773
+#> NULL                                                    5939       3619
+#> pspline(Follow.Up.Time, df = 2)         7    246.6      5932       3373
+#> ScrArm                                  1      5.6      5931       3367
+#> pspline(Follow.Up.Time, df = 2):ScrArm  7      7.9      5924       3359
 #>                                        Pr(>Chi)    
@@
 #> pspline(Follow.Up.Time, df = 2)          <2e-16 ***
-#> ScrArm                                   0.0046 ** 
-#> pspline(Follow.Up.Time, df = 2):ScrArm   0.2454    
+#> ScrArm                                    0.018 *  
+#> pspline(Follow.Up.Time, df = 2):ScrArm    0.343    
 #> ---
@@
 \[
-\frac{\lambda\left(t | X=x_1, \mathbf{Z}=\mathbf{z_1} ; \hat{\beta}\right)}{\lambda(t | X=x_0, \mathbf{Z}=\mathbf{z_0} ; \hat{\beta})}
+\frac{\lambda\left(t | X=x_1, \mathbf{Z}=\mathbf{z} ; \hat{\beta}\right)}{\lambda(t | X=x_0, \mathbf{Z}=\mathbf{z} ; \hat{\beta})}
 \] where \(\lambda(t|\cdot;\hat{\beta})\) is the hazard rate as a
@@
 \(\mathbf{Z}\) are other covariates in the model which are equal to
-\(\mathbf{z_1}\) in the exposed and \(\mathbf{z_0}\) in the unexposed
-group, and \(\hat{\beta}\) are the estimated regression coefficients. As
-indicated by the formula above, it is most instructive to plot the
-hazard ratio as a function of a variable \(t\) only if there is an
-interaction between \(t\) and \(X\). Otherwise, the resulting plot will
-simply be a horizontal line across time.
+\(\mathbf{z}\) in both the exposed and unexposed group, and
+\(\hat{\beta}\) are the estimated regression coefficients. As indicated
+by the formula above, it is most instructive to plot the hazard ratio as
+a function of a variable \(t\) only if there is an interaction between
+\(t\) and \(X\). Otherwise, the resulting plot will simply be a
+horizontal line across time.
 
@@
 #> 1  0.0  0.0e+00  0.0e+00   0.0e+00 Control group
-#> 2  0.1  2.0e-06  2.5e-07   1.6e-05 Control group
-#> 3  0.2  4.2e-06  6.1e-07   3.0e-05 Control group
-#> 4  0.3  6.6e-06  1.1e-06   4.2e-05 Control group
-#> 5  0.4  9.2e-06  1.8e-06   5.2e-05 Control group
-#> 6  0.5  1.2e-05  2.8e-06   6.2e-05 Control group
+#> 2  0.1  1.8e-06  1.9e-07   1.2e-05 Control group
+#> 3  0.2  3.9e-06  4.8e-07   2.2e-05 Control group
+#> 4  0.3  6.1e-06  9.0e-07   3.2e-05 Control group
+#> 5  0.4  8.7e-06  1.5e-06   4.1e-05 Control group
+#> 6  0.5  1.2e-05  2.3e-06   5.0e-05 Control group
 \end{Soutput}
@@
 In this case study, we show how case-base sampling can be used in the
-context of a competing-risk analysis. For illustrative purposes, we use
+context of a competing risk analysis. For illustrative purposes, we use
 the same data that was used in Scrucca \emph{et al}
@@
 \midrule
-Sex & 0.75 & (0.43, 1.31) & 0.68 & (0.39, 1.21)\\
-Disease & 0.55 & (0.3, 0.99) & 0.52 & (0.28, 0.94)\\
-Phase (CR2 vs. CR1) & 1.10 & (0.44, 2.74) & 1.21 & (0.47, 3.09)\\
-Phase (CR3 vs. CR1) & 1.75 & (0.45, 6.78) & 1.67 & (0.46, 6.08)\\
-Phase (Relapse vs. CR1) & 4.23 & (1.96, 9.14) & 4.55 & (1.98, 10.46)\\
+Sex & 0.73 & (0.42, 1.27) & 0.68 & (0.39, 1.21)\\
+Disease & 0.54 & (0.3, 0.98) & 0.52 & (0.28, 0.94)\\
+Phase (CR2 vs. CR1) & 1.19 & (0.48, 2.97) & 1.21 & (0.47, 3.09)\\
+Phase (CR3 vs. CR1) & 1.44 & (0.37, 5.64) & 1.67 & (0.46, 6.08)\\
+Phase (Relapse vs. CR1) & 4.22 & (1.95, 9.12) & 4.55 & (1.98, 10.46)\\
 \addlinespace
-Source & 1.61 & (0.53, 4.91) & 1.46 & (0.47, 4.53)\\
-Age & 1.00 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
+Source & 1.46 & (0.48, 4.46) & 1.46 & (0.47, 4.53)\\
+Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
 \bottomrule
@@
 McGill University\\%
-1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
+2001 McGill College Avenue Montreal, QC, Canada H3A 1G1\\
 %
-\url{http://sahirbhatnagar.com/}\\%
+\url{http://sahirbhatnagar.com/}%
 %
-\email{sahir.bhatnagar@mcgill.ca}%
+\\\email{sahir.bhatnagar@mcgill.ca}
 }
@@
 McGill University\\%
-1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
+2001 McGill College Avenue Montreal, QC, Canada H3A 1G1\\
 %
@@
 McGill University\\%
-1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
+2001 McGill College Avenue Montreal, QC, Canada H3A 1G1\\
 %
-\url{http://www.medicine.mcgill.ca/epidemiology/hanley/}\\%
+\url{http://www.medicine.mcgill.ca/epidemiology/hanley/}%
 %
-\email{james.hanley@mcgill.ca}%
+\\\email{james.hanley@mcgill.ca}
 }

### 2022-06-06 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
address comments from reviewers

@@
 For simplicity, we will consider Type I censoring due to the end of follow-up at time $\tau$ (the general case of independent censoring is treated in Saarela [-@saarela2016case]).
-We assume a continuous time model, which implies that the counting process jumps are less than or equal to one.
 We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
@@
 $$P(\,\textrm dN_{i}(t) \mid \,\textrm dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.$$
+By using the symbol $\stackrel{\theta}{\propto}$, we mean that both sides of the expression are equal up to multiplicative factors that do not depend on $\theta$.
 Therefore, an estimating function for $\theta$ can be composed of these contributions as:
@@
 
-  1. Compute $\hat{\theta}$ and its variance-covariance matrix $V(\theta)$.
-  2. Sample $B$ times from a multivariate normal $N\left(\hat{\theta}, V(\theta)\right)$.
+  1. Compute $\hat{\theta}$ and its variance-covariance matrix $V(\hat \theta)$ which is obtained as part of the logistic regression output.
+  2. Sample $B$ times from a multivariate normal $N\left(\hat{\theta}, V(\hat\theta)\right)$.
   3. For each sample, compute the survival function using Equation \ref{eqn:surv} below.
@@
   theme(legend.position = "none") + 
-  scale_y_continuous(labels = label_number_si())
+  scale_y_continuous(labels = label_number_si()) +
+  scale_color_manual(values = c("#AB3A59"),
+                     limits = c("Case series"),
+                     name = "") +
+  scale_fill_manual(values = c("#E16A86"),
+                    limits = c("Case series"),
+                    name = "")
 
 plotc <- plot(a, add.case.series = TRUE, legend = FALSE) +
-  scale_y_continuous(labels = label_number_si())
+  scale_y_continuous(labels = label_number_si()) +
+  scale_color_manual(values = c("#AB3A59"),
+                     limits = c("Case series"),
+                     name = "") +
+  scale_fill_manual(values = c("#E16A86"),
+                    limits = c("Case series"),
+                    name = "")
 
@@
   plotd + 
-    guides(color = guide_legend(nrow = 1)) +
     theme(legend.direction = "horizontal",
            legend.justification = "center",
-           legend.box.just = "bottom")
+           legend.box.just = "bottom") +
+    scale_color_manual(values = c("#AB3A59", "#347004"),
+                       limits = c("Case series", "Base series"),
+                       name = "") +
+    scale_fill_manual(values = c("#E16A86", "#50A315"),
+                      limits = c("Case series", "Base series"),
+                      name = "") +
+      guides(color = guide_legend(nrow = 1),
+             fill = guide_legend(nrow = 1))
 )
@@
 These quantities appear in Table \ref{tab:bmtcrr-cis}.
-As we can see, the type of disease corresponds to a significant hazard ratio: the hazard for AML is 48% lower than that for ALL.
-Moreover, being in relapse at transplant is associated with a hazard ratio of 4.09 when compared to CR1.
+As we can see, the type of disease corresponds to a significant hazard ratio: the hazard for AML is about half that for ALL.
+Moreover, being in relapse at transplant is associated with a hazard ratio of 4.22 when compared to CR1.
 

### 2022-06-06 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
final final
address comments from reviewers

@@
 independent censoring is treated in Saarela
-\citeyearpar{saarela2016case}). We assume a continuous time model, which
-implies that the counting process jumps are less than or equal to one.
-We are interested in modeling the hazard functions \(\lambda_{i}(t)\) of
-the processes \(N_i(t)\), and which satisfy
+\citeyearpar{saarela2016case}). We are interested in modeling the hazard
+functions \(\lambda_{i}(t)\) of the processes \(N_i(t)\), and which
+satisfy
 \[\lambda_{i}(t)\,\textrm dt = E[\,\textrm dN_{i}(t)\mid \mathrm{past}].\]
@@
 \[P(\,\textrm dN_{i}(t) \mid \,\textrm dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.\]
-Therefore, an estimating function for \(\theta\) can be composed of
-these contributions as: \begin{equation}
+By using the symbol \(\stackrel{\theta}{\propto}\), we mean that both
+sides of the expression are equal up to multiplicative factors that do
+not depend on \(\theta\). Therefore, an estimating function for
+\(\theta\) can be composed of these contributions as: \begin{equation}
 L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{\,\textrm dQ_i(t)}. \label{eq:lik-function}
@@
   Compute \(\hat{\theta}\) and its variance-covariance matrix
-  \(V(\theta)\).
+  \(V(\hat \theta)\) which is obtained as part of the logistic
+  regression output.
 \item
   Sample \(B\) times from a multivariate normal
-  \(N\left(\hat{\theta}, V(\theta)\right)\).
+  \(N\left(\hat{\theta}, V(\hat\theta)\right)\).
 \item
@@
 \ref{tab:bmtcrr-cis}. As we can see, the type of disease corresponds to
-a significant hazard ratio: the hazard for AML is 48\% lower than that
-for ALL. Moreover, being in relapse at transplant is associated with a
-hazard ratio of 4.09 when compared to CR1.
+a significant hazard ratio: the hazard for AML is about half that for
+ALL. Moreover, being in relapse at transplant is associated with a
+hazard ratio of 4.22 when compared to CR1.
 

### 2022-06-07 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
add paragraph on parsimonious model

@@
 
-The plot shows that the effect of screening only becomes statistically apparent 
-by year 7 and later. The 25-60% reductions seen in years 8-12 of the study suggests 
-a much higher reduction in prostate cancer due to screening than the single 
-overall 20% reported in the original article.
+The plot shows that the effect of screening only becomes statistically apparent by year 7 and later.
+The 25-60% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20% reported in the original article.
+
+A more parsimonious model based on these results could be constructed as follows:
+
+$$ \log \lambda\left(t \mid \mathbf{Z}\right) = \begin{cases} \alpha, &\quad t < t_0\\
+\alpha + \beta Z(t - t_0), &\quad t \geq t_0\end{cases},$$
+where $t_0 \approx 7$ years.
+This model corresponds to a hazard ratio that is constant and equal to 1 until $t = t_0$, after which it decreases exponentially.
+If we fix the value $t_0$, we can easily implement this model using our package; for example:
+
+```{r, echo = TRUE, eval = FALSE}
+fitSmoothHazard(DeadOfPrCa ~ as.integer(Follow.Up.Time >= t0) :
+                    I(Follow.Up.Time - t0) : ScrArm,
+                data = ERSPC)
+```
+
+\noindent Here, we use the binary variable `as.integer(Follow.Up.Time >= t0)` in order to write the two cases of our formula above in a more compact way.
+We also use the function `I()`, which allows us to specify new variables within the formula using normal `R` code.
+This is required, as otherwise the symbol `-` would be interpreted as a formula operator.
+Finally, we use the operator `:` instead of `*` in order to only include the interaction term, not the main effects.
+
+Alternatively, the breakpoint $t_0$ could also be estimated by combining case-base sampling with segmented regression. However, this extension is beyond the current scope of `casebase`.
 

### 2022-06-07 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
add paragraph on parsimonious model

@@
 
+A more parsimonious model based on these results could be constructed as
+follows:
+
+\[ \log \lambda\left(t \mid \mathbf{Z}\right) = \begin{cases} \alpha, &\quad t < t_0\\
+\alpha + \beta Z(t - t_0), &\quad t \geq t_0\end{cases},\] where
+\(t_0 \approx 7\) years. This model corresponds to a hazard ratio that
+is constant and equal to 1 until \(t = t_0\), after which it decreases
+exponentially. If we fix the value \(t_0\), we can easily implement this
+model using our package; for example:
+
+\begin{Schunk}
+\begin{Sinput}
+fitSmoothHazard(DeadOfPrCa ~ as.integer(Follow.Up.Time >= t0) :
+                    I(Follow.Up.Time - t0) : ScrArm,
+                data = ERSPC)
+\end{Sinput}
+\end{Schunk}
+
+\noindent Here, we use the binary variable
+\texttt{as.integer(Follow.Up.Time\ \textgreater{}=\ t0)} in order to
+write the two cases of our formula above in a more compact way. We also
+use the function \texttt{I()}, which allows us to specify new variables
+within the formula using normal \texttt{R} code. This is required, as
+otherwise the symbol \texttt{-} would be interpreted as a formula
+operator. Finally, we use the operator \texttt{:} instead of \texttt{*}
+in order to only include the interaction term, not the main effects.
+
+Alternatively, the breakpoint \(t_0\) could also be estimated by
+combining case-base sampling with segmented regression. However, this
+extension is beyond the current scope of \texttt{casebase}.
+
 \hypertarget{hazard-functions}{%

### 2022-06-07 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
reorder terms in formula

@@
 fitSmoothHazard(DeadOfPrCa ~ as.integer(Follow.Up.Time >= t0) :
-                    I(Follow.Up.Time - t0) : ScrArm,
+                    ScrArm : I(Follow.Up.Time - t0),
                 data = ERSPC)
@@
 This is required, as otherwise the symbol `-` would be interpreted as a formula operator.
-Finally, we use the operator `:` instead of `*` in order to only include the interaction term, not the main effects.
+Finally, the term `ScrArm : I(Follow.Up.Time - t0)` represents the product $Z (t - t_0)$ in the equation above.
+We use the operator `:` instead of `*` in order to only include the interaction term, not the main effects.
 

### 2022-06-07 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
reorder terms in formula

@@
 otherwise the symbol \texttt{-} would be interpreted as a formula
-operator. Finally, we use the operator \texttt{:} instead of \texttt{*}
-in order to only include the interaction term, not the main effects.
+operator. Finally, the term \texttt{ScrArm\ :\ I(Follow.Up.Time\ -\ t0)}
+represents the product \(Z (t - t_0)\) in the equation above. We use the
+operator \texttt{:} instead of \texttt{*} in order to only include the
+interaction term, not the main effects.
 

### 2022-11-11 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
switch to distill

@@
     url: http://individual.utoronto.ca/osaarela/
-title: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
-abstract: |
-  | In clinical studies of time-to-event data, a quantity of interest to the clinician is their patient's risk of an event.
+title: "\\pkg{casebase}: An Alternative Framework for Survival Analysis and Comparison of Event Rates"
+abstract: >
+  In clinical studies of time-to-event data, a quantity of interest to the clinician is their patient's risk of an event.
 However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the estimating function.
@@
 Their approach naturally leads to estimates of the survival or risk function that are smooth-in-time.
-  | In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis.
+  In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis.
 We describe how the case-base framework can also be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection.
@@
 We illustrate all these features through three different case studies.
-  | * SRB and MT contributed equally to this work.
+  * SRB and MT contributed equally to this work.
 preamble: >
@@
   \usepackage{tabu}
-output: rticles::rjournal_article
+  \usepackage[utf8]{inputenc}
+output: rjtools::rjournal_web_article
 bibliography: bhatnagar-turgeon-islam-hanley-saarela.bib
@@
 
-options(kableExtra.latex.load_packages = FALSE)
+options(kableExtra.latex.load_packages = FALSE,
+        knitr.kable.NA = "")
 library(casebase)
@@
 fitting fully parametric hazard models and covariate-conditional survival curves 
-using the familiar interface of the \code{glm} function. Our implementation 
+using the familiar interface of the `glm` function. Our implementation 
 includes extensions to other models such as penalized 
@@
 
-If the hazard function $\lambda_{i}(t; \theta)$ is parametrized in terms of $\theta$, we can define an estimator $\hat{\theta}$ by maximization of the likelihood expression
+If the hazard function $\lambda_{i}(t; \theta)$ is parametrized in terms of $\theta$, we can define an estimator $\widehat{\theta}$ by maximization of the likelihood expression
 $$L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^{\min(t_i,\tau)} \lambda_i(t; \theta)\,\textrm dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{\,\textrm dN_{i}(t)},$$
@@
 Generally, an individual can contribute more than one person-moment, and thus the terms in the product integral are not independent.
-Nonetheless, Saarela [-@saarela2016case] showed that the corresponding partial likelihood score function has mean zero at the true value $\theta=\theta_0$, and that the resulting estimator $\hat{\theta}$ is asymptotically normally distributed.
+Nonetheless, Saarela [-@saarela2016case] showed that the corresponding partial likelihood score function has mean zero at the true value $\theta=\theta_0$, and that the resulting estimator $\widehat{\theta}$ is asymptotically normally distributed.
 
@@
 
-The asymptotic normality of $\hat{\theta}$ gives us an efficient way to estimate the variance of various estimators derived from the hazard function.
+The asymptotic normality of $\widehat{\theta}$ gives us an efficient way to estimate the variance of various estimators derived from the hazard function.
 For example, to construct confidence bands around risk functions and survival functions, we can use the following procedure:
 
-  1. Compute $\hat{\theta}$ and its variance-covariance matrix $V(\hat \theta)$ which is obtained as part of the logistic regression output.
-  2. Sample $B$ times from a multivariate normal $N\left(\hat{\theta}, V(\hat\theta)\right)$.
+  1. Compute $\widehat{\theta}$ and its variance-covariance matrix $V(\widehat \theta)$ which is obtained as part of the logistic regression output.
+  2. Sample $B$ times from a multivariate normal $N\left(\widehat{\theta}, V(\widehat\theta)\right)$.
   3. For each sample, compute the survival function using Equation \ref{eqn:surv} below.
@@
 \end{equation}
-In this model, the instantaneous failure rate is independent of $t$.\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
-This is also known as the \dfn{memoryless property} \citep{kalbfleisch2011statistical}.}
+In this model, the instantaneous failure rate is independent of $t$.^[The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
+This is also known as the *memoryless property* [@kalbfleisch2011statistical].]
 
@@
 where $N_j(t)$ is the counting process associated with the event of type $j$ and $\lambda_j(t)$ is the corresponding cause-specific hazard function.
-As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression with an offset term.\footnote{Specifically, it corresponds to the following parametrization  for a multinomial random variable $Y$: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j + log(1/\rho), \qquad j = 1,\ldots, J-1.\end{align*}}
+As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression with an offset term.^[Specifically, it corresponds to the following parametrization  for a multinomial random variable $Y$: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j + log(1/\rho), \qquad j = 1,\ldots, J-1.\end{align*}]
 
@@
 Accordingly, there is already a vast ecosystem of R packages implementing different methodologies.
-In this section, we describe how the functionalities of \pkg{casebase} compare to these packages.
+In this section, we describe how the functionalities of \CRANpkg{casebase} compare to these packages.
 
@@
 For the purposes of this article, we restricted our review to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and Cumulative Incidence (CI) estimation.
-By searching for appropriate keywords in the \code{DESCRIPTION} file of these packages, we found 60 relevant packages.
+By searching for appropriate keywords in the `DESCRIPTION` file of these packages, we found 60 relevant packages.
 These 60 packages were then manually examined to determine which ones are comparable to \pkg{casebase}.
 In particular, we excluded packages that were focused on a different set of problems, such as frailty and multistate models.
-The remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with some of the functionalities they offer.
+The remaining 14 packages appear in Table \@ref(tab:surv-pkgs), along with some of the functionalities they offer.
 
@@
 
-\begin{table}[ht]
-\centering
-\resizebox{\textwidth}{!}{%
-\begin{tabular}{ccccccccc}
-\toprule
-& \textbf{Competing} & \textbf{Allows} &  \textbf{Penalized}   &    &  & \textbf{Semi} & \textbf{Interval/Left} & \textbf{Risk}  \\
-\textbf{Package}        & \textbf{Risks}  & \textbf{Non PH} & \textbf{Regression} & \textbf{Splines} & \textbf{Parametric} & \textbf{Parametric} & \textbf{Censoring} & \textbf{Estimates}           \\
-\bottomrule
-\textbf{casebase}        & \checkmark                        & \checkmark                         & \checkmark                     & \checkmark                & \checkmark                   &                          &                                  & \checkmark                                \\ \hline
-\textbf{CFC}             & \checkmark                        & \checkmark                         &                       &                  & \checkmark                   &                          &                                  & \checkmark                                \\ \hline
-\textbf{cmprsk}             & \checkmark                        &                           &                       &                  &                     &  \checkmark                        &                                  & \checkmark                                \\ \hline
-\textbf{crrp}            & \checkmark                        &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
-\textbf{fastcox}         &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
-\textbf{flexrsurv}       &                          & \checkmark                         &                       & \checkmark                & \checkmark                   &                          &                                  & \checkmark               \\ \hline
-\textbf{flexsurv}        & \checkmark                        & \checkmark                         &                       & \checkmark                & \checkmark                   &                          &                                  & \checkmark               \\ \hline
-\textbf{glmnet}          &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &  \checkmark              \\ \hline
-\textbf{glmpath}         &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
-\textbf{mets}            & \checkmark                        &                           &                       & \checkmark                &                     & \checkmark                        &                                  & \checkmark                                \\ \hline
-\textbf{penalized}       &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
-\textbf{riskRegression}  & \checkmark                         &                           & \checkmark                     &                  &                     & \checkmark                        &                                  & \checkmark                                \\ \hline
-\textbf{rstpm2}          &                          & \checkmark                         &                      & \checkmark                & \checkmark                   & \checkmark                        & \checkmark                                & \checkmark                         \\ \hline
-\textbf{SmoothHazard}    &                          & \checkmark                         &                       & \checkmark                & \checkmark                   &                          & \checkmark                            &                                      \\ \hline
-\textbf{survival}        & \checkmark                        & \checkmark                         &                       &                  & \checkmark                   & \checkmark                        & \checkmark                                & \checkmark                               \\
-\bottomrule
-\end{tabular}%
+```{r surv-pkgs, echo = FALSE}
+if (knitr::is_latex_output()) {
+    checkmark <-  "X" # "\\checkmark"
+    caption <- "Comparison of various R packages for survival analysis.
+\\textbf{Competing Risks}: whether an implementation for competing risks is present.
+\\textbf{Allows Non PH}: includes models for non-proportional hazards.
+\\textbf{Penalized Regression}: allows for a penalty term on the regression coefficients when estimating hazards (e.g.
+lasso or ridge).
+\\textbf{Splines}: allows a flexible fit on time through the use of splines.
+\\textbf{Parametric}: implementation for parametric models.
+\\textbf{Semi-parametric}: implementation for semi-parametric models.
+\\textbf{Interval/left censoring}: models for interval and left-censoring.
+If this is not selected, the package only handles right-censoring.
+\\textbf{Risk estimates}: estimation of survival curve and cumulative incidence is available."
+} else {
+    checkmark <- "\u2713"
+    caption <- "Comparison of various R packages for survival analysis.
+**Competing Risks**: whether an implementation for competing risks is present.
+**Allows Non PH**: includes models for non-proportional hazards.
+**Penalized Regression**: allows for a penalty term on the regression coefficients when estimating hazards (e.g.
+lasso or ridge).
+**Splines**: allows a flexible fit on time through the use of splines.
+**Parametric**: implementation for parametric models.
+**Semi-parametric**: implementation for semi-parametric models.
+**Interval/left censoring**: models for interval and left-censoring.
+If this is not selected, the package only handles right-censoring.
+**Risk estimates**: estimation of survival curve and cumulative incidence is available."
 }
-\caption{Comparison of various R packages for survival analysis.
-\textbf{Competing Risks}: whether an implementation for competing risks is present.
-\textbf{Allows Non PH}: includes models for non-proportional hazards.
-\textbf{Penalized Regression}: allows for a penalty term on the regression coefficients when estimating hazards (e.g.
-lasso or ridge).
-\textbf{Splines}: allows a flexible fit on time through the use of splines.
-\textbf{Parametric}: implementation for parametric models.
-\textbf{Semi-parametric}: implementation for semi-parametric models.
-\textbf{Interval/left censoring}: models for interval and left-censoring.
-If this is not selected, the package only handles right-censoring.
-\textbf{Risk estimates}: estimation of survival curve and cumulative incidence is available.}
-\label{tab:surv-pkgs}
-\end{table}
+
+tibble(
+    Package = c("casebase", "CFC", "cmprsk", "crrp", "fastcox", "flexrsurv", 
+                "flexsurv", "glmnet", "glmpath", "mets", "penalized", 
+                "riskRegression", "rstpm2", "SmoothHazard", "survival"),
+    Competing_Risks = c(checkmark, checkmark, checkmark, checkmark, "", "", checkmark,
+                        "", "", checkmark, "", checkmark, "", "", checkmark),
+    Allows_Non_PH = c(checkmark, checkmark, "", "", "", checkmark, checkmark, 
+                      "", "", "", "", "", checkmark, checkmark, checkmark),
+    Penalized_Regression = c(checkmark, "", "", checkmark, checkmark, "", "",
+                             checkmark, checkmark, "", checkmark, checkmark, "", "", ""),
+    Splines = c(checkmark, "", "", "", "", checkmark, checkmark, "", "", checkmark,
+                "", "", checkmark, checkmark, ""),
+    Parametric = c(checkmark, checkmark, "", "", "", checkmark, checkmark, "", "", 
+                   "", "", "", checkmark, checkmark, checkmark),
+    Semi_Parametric = c("", "", checkmark, checkmark, checkmark, "", "", checkmark,
+                        checkmark, checkmark, checkmark, checkmark, checkmark, "", checkmark),
+    Interval_Left_Censoring = c(rep("", 12), checkmark, checkmark, checkmark),
+    Risk_Estimates = c(checkmark, checkmark, checkmark, "", "", checkmark, checkmark,
+                       checkmark, "", checkmark, "", checkmark, checkmark, "", checkmark)
+) %>%
+  knitr::kable(
+    # format = "latex", 
+    booktabs = knitr::is_latex_output(),
+    col.names = c("Package", "Competing Risks", "Allows Non PH", "Penalized Regression", 
+                  "Splines", "Parametric", "Semi Parametric", "Interval/Left Censoring",
+                  "Risk Estimates"),
+    caption = caption
+  ) %>%
+  kable_styling(latex_options = c("scale_down")) %>% 
+    row_spec(0, bold = TRUE)
+```
 
@@
 The functions in the \pkg{casebase} package can be divided into two categories: 1) exploratory data analysis, in the form of population-time plots; and 2) parametric modeling of the hazard function.
-We strove for compatibility with both \code{data.frame}s and \code{data.table}s; this can be seen in the coding choices we made and the unit tests we wrote.
+We strove for compatibility with both `data.frame`s and `data.table`s; this can be seen in the coding choices we made and the unit tests we wrote.
 
@@
 These plots are informative graphical displays of survival data and should be one of the first steps in an exploratory data analysis.
-The \code{popTime} function and \code{plot} method facilitate this task:
+The `popTime` function and `plot` method facilitate this task:
 
 1.
-The \code{casebase::popTime} function takes as input the original dataset along with the column names corresponding to the timescale, the event status and an exposure group of interest (optional).
-This will create an object of class \code{popTime}.
-2. The corresponding \code{plot} method for the object created in Step 1 can be called to create the population time plot with several options for customizing the aesthetics.
+The `casebase::popTime` function takes as input the original dataset along with the column names corresponding to the timescale, the event status and an exposure group of interest (optional).
+This will create an object of class `popTime`.
+2. The corresponding `plot` method for the object created in Step 1 can be called to create the population time plot with several options for customizing the aesthetics.
 
@@
 
-We first create the necessary dataset for producing the population time plot using the \code{popTime} function.
+We first create the necessary dataset for producing the population time plot using the `popTime` function.
 In this example, we stratify the plot by treatment group.
-The resulting object inherits from class \code{popTime} and stores the exposure variable as an attribute:
-
-```{r erspc-data-mutate, eval = eval_cs1, echo = FALSE}
+The resulting object inherits from class `popTime` and stores the exposure variable as an attribute:
+
+```{r erspc-data-mutate, eval = TRUE, echo = FALSE}
 data("ERSPC")
@@
 
-```{r erspc-data, eval = eval_cs1, echo=TRUE}
+```{r erspc-data, eval = TRUE, echo=TRUE}
 pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
@@
 
-```{r plot-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plot for the ERSPC dataset.
-\\textbf{A}: The gray area can be thought of as N = 159,893 infinitely thin horizontal rectangles ordered by length of follow-up.
-\\textbf{B}: The red points correspond to when death has occurred for any one of those infinitely thin rectangles.
-\\textbf{C}: To improve visibility, these red points are randomly redistributed along their respective x-coordinates, providing a visualization of incidence density.
+```{r plot-erspc-data, eval = TRUE, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plot for the ERSPC dataset.
+A: The gray area can be thought of as N = 159,893 infinitely thin horizontal rectangles ordered by length of follow-up.
+B: The red points correspond to when death has occurred for any one of those infinitely thin rectangles.
+C: To improve visibility, these red points are randomly redistributed along their respective x-coordinates, providing a visualization of incidence density.
 More events are observed at later follow-up times, motivating the use of non-constant hazard models.
-\\textbf{D}: The base series, a representative sample of the entire grey area, is represented by the green points."}
+D: The base series, a representative sample of the entire grey area, is represented by the green points."}
 # For increased visibility, remove some cases
@@
 
-We then pass this object to the corresponding \code{plot} method:
+We then pass this object to the corresponding `plot` method:
 
@@
 
-Figure \ref{fig:plot-erspc-data} depicts the process of creating a population-time plot.
-It is built sequentially by first adding a layer for the area representing the population time in gray (Figure \ref{fig:plot-erspc-data} A), with subjects having the least amount of observation time plotted at the top of the y-axis.
+Figure \@ref(fig:plot-erspc-data) depicts the process of creating a population-time plot.
+It is built sequentially by first adding a layer for the area representing the population time in gray (Figure \@ref(fig:plot-erspc-data) A), with subjects having the least amount of observation time plotted at the top of the y-axis.
 We immediately notice a distinctive *stepwise shape* in the population time area.
@@
 Next, layers for the case series and base series are added.
-The y-axis location of each case moment is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area (Figure \ref{fig:plot-erspc-data} B).
+The y-axis location of each case moment is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area (Figure \@ref(fig:plot-erspc-data) B).
 By randomly distributing the cases, we can get a sense of the incidence density.
-In Figure \ref{fig:plot-erspc-data} C, we see that more events are observed at later follow-up times.
+In Figure \@ref(fig:plot-erspc-data) C, we see that more events are observed at later follow-up times.
 Therefore, a constant hazard model would not be appropriate in this instance as it would overestimate the incidence earlier on in time, and underestimate it later on.
-Finally, the base series is sampled uniformly from the study base (Figure \ref{fig:plot-erspc-data} D).
+Finally, the base series is sampled uniformly from the study base (Figure \@ref(fig:plot-erspc-data) D).
 The reader should refer to the package vignettes for more examples and a detailed description of how to modify the aesthetics of a population-time plot.
@@
 
-The sampling scheme selected for \code{sampleCaseBase} was described in Hanley & Miettinen [-@hanley2009fitting]: we first sample along the "person" axis, proportional to each individual's total follow-up time, and then we sample a moment uniformly over their follow-up time.
+The sampling scheme selected for `sampleCaseBase` was described in Hanley & Miettinen [-@hanley2009fitting]: we first sample along the "person" axis, proportional to each individual's total follow-up time, and then we sample a moment uniformly over their follow-up time.
 This sampling scheme is equivalent to the following picture: imagine representing the total follow-up time of all individuals in the study along a single dimension, where the follow-up time of the next individual would start exactly when the follow-up time of the previous individual ends.
@@
 
-Next, the fitting function \code{fitSmoothHazard} starts by looking at the class of the dataset: if it was generated from \code{sampleCaseBase}, it automatically inherited the class \code{cbData}.
-If the dataset supplied to \code{fitSmoothHazard} does not inherit from \code{cbData}, then the fitting function starts by calling \code{sampleCaseBase} to generate the base series.
-In other words, users can bypass \code{sampleCaseBase} altogether and only worry about the fitting function \code{fitSmoothHazard}.
-
-The fitting function retains the familiar formula interface of \code{glm}.
+Next, the fitting function `fitSmoothHazard` starts by looking at the class of the dataset: if it was generated from `sampleCaseBase`, it automatically inherited the class `cbData`.
+If the dataset supplied to `fitSmoothHazard` does not inherit from `cbData`, then the fitting function starts by calling `sampleCaseBase` to generate the base series.
+In other words, users can bypass `sampleCaseBase` altogether and only worry about the fitting function `fitSmoothHazard`.
+
+The fitting function retains the familiar formula interface of `glm`.
 The left-hand side of the formula should be the name of the column corresponding to the event type.
@@
 Case Study 1 below).
-The offset term does not need to be specified by the user, as it is automatically added to the formula before calling \code{glm}.
-
-To fit the hazard function, we provide several approaches that are available via the \code{family} parameter. These approaches are:
-
-  - \code{glm}: This is the familiar logistic regression.
-  - \code{glmnet}: This option allows for variable selection using the elastic-net [@zou2005regularization] penalty (cf.
+The offset term does not need to be specified by the user, as it is automatically added to the formula before calling `glm`.
+
+To fit the hazard function, we provide several approaches that are available via the `family` parameter. These approaches are:
+
+  - `glm`: This is the familiar logistic regression.
+  - `glmnet`: This option allows for variable selection using the elastic-net [@zou2005regularization] penalty (cf.
 Case Study 3).
 This functionality is provided through the \pkg{glmnet} package [@friedman2010jss].
-  - \code{gam}: This option provides support for *Generalized Additive Models* via the \CRANpkg{mgcv} package [@hastie1987generalized].
+  - `gam`: This option provides support for *Generalized Additive Models* via the \CRANpkg{mgcv} package [@hastie1987generalized].
 
@@
 
-Once a model-fit object has been returned by \code{fitSmoothHazard}, all the familiar summary and diagnostic functions are available: \code{print}, \code{summary}, \code{predict}, \code{plot}, etc.
+Once a model-fit object has been returned by `fitSmoothHazard`, all the familiar summary and diagnostic functions are available: `print`, `summary`, `predict`, `plot`, etc.
 Our package provides one more functionality: it computes risk functions from the model fit.
@@
 Again, the integrals are computed using either numerical integration (via the trapezoidal rule) or Monte Carlo integration.
-This option is controlled by the argument \code{method} of the \code{absoluteRisk} function.
-
-Finally, the output from \code{absoluteRisk} can be passed to a method \code{confint} to compute confidence bands around the survival function, as described in the Theoretical Details section.
-These bands are only valid when \code{family = "glm"} as it relies on the asymptotic normality of the estimator.
+This option is controlled by the argument `method` of the `absoluteRisk` function.
+
+Finally, the output from `absoluteRisk` can be passed to a method `confint` to compute confidence bands around the survival function, as described in the Theoretical Details section.
+These bands are only valid when `family = "glm"` as it relies on the asymptotic normality of the estimator.
 Currently, this is only available for the single-event setting.
@@
 cubic spline basis with 2 degrees of freedom (implemented in the 
-\code{survival::pspline} function). The model is fit using \code{fitSmoothHazard} 
+`survival::pspline` function). The model is fit using `fitSmoothHazard` 
 with the familiar formula interface:
 
-```{r, eval = eval_cs1, echo = -1}
+```{r, eval = TRUE, echo = -1}
 set.seed(1953)
@@
 <!--Note that we did not have to specify any cut points, as would be the case 
-with the \code{survSplit} function in the \pkg{survival} package.--> 
-The output object from \code{fitSmoothHazard} inherits from the \code{singleEventCB} 
-and \code{glm} classes. For this reason, we can leverage the \code{summary} 
-method for \code{glm} objects to output a familiar summary of the results:
-
-```{r, eval = eval_cs1, echo = TRUE}
+with the `survSplit` function in the \pkg{survival} package.--> 
+The output object from `fitSmoothHazard` inherits from the `singleEventCB` 
+and `glm` classes. For this reason, we can leverage the `summary` 
+method for `glm` objects to output a familiar summary of the results:
+
+```{r, eval = TRUE, echo = TRUE}
 summary(fit) 
@@
 
-```{r erscp-compare, eval = eval_cs1, echo = TRUE}
+```{r erscp-compare, eval = TRUE, echo = TRUE}
 anova(fit, test = "LRT")
@@
 $$
-\frac{\lambda\left(t | X=x_1, \mathbf{Z}=\mathbf{z} ; \hat{\beta}\right)}{\lambda(t | X=x_0, \mathbf{Z}=\mathbf{z} ; \hat{\beta})}
+\frac{\lambda\left(t | X=x_1, \mathbf{Z}=\mathbf{z} ; \widehat{\beta}\right)}{\lambda(t | X=x_0, \mathbf{Z}=\mathbf{z} ; \widehat{\beta})}
 $$
-where $\lambda(t|\cdot;\hat{\beta})$ is the hazard rate as a function of the 
+where $\lambda(t|\cdot;\widehat{\beta})$ is the hazard rate as a function of the 
 variable $t$ (which is usually time, but can be any other continuous variable), 
@@
 equal to $\mathbf{z}$ in both the exposed and unexposed group, 
-and $\hat{\beta}$ are the estimated regression coefficients. As indicated by the 
+and $\widehat{\beta}$ are the estimated regression coefficients. As indicated by the 
 formula above, it is most instructive to plot the hazard ratio as a function of 
@@
 
-The \code{plot} method with \code{type="hr"} for objects of class 
-\code{singleEventCB} can be used to compute time-dependent hazard ratios and 
-confidence intervals. In Figure \ref{fig:interaction-ERSPC}, we show the 
+The `plot` method with `type="hr"` for objects of class 
+`singleEventCB` can be used to compute time-dependent hazard ratios and 
+confidence intervals. In Figure \@ref(fig:interaction-ERSPC), we show the 
 estimated hazard ratio and 95% confidence interval for screening vs. control 
@@
 plot(fit, type = "hr", newdata = new_data,
-     var = "ScrArm", xvar = "Follow.Up.Time", ci = T)
-```
-
-```{r interaction-ERSPC, eval = eval_cs1, fig.width=8, fig.height=6, fig.cap='Estimated hazard ratio and 95\\% confidence interval for screening vs.
+     var = "ScrArm", xvar = "Follow.Up.Time", ci = TRUE)
+```
+
+```{r interaction-ERSPC, eval = TRUE, fig.width=8, fig.height=6, fig.cap='Estimated hazard ratio and 95\\% confidence interval for screening vs.
 control group as a function of time in the ERSPC dataset.
@@
 Modeling the hazard function directly allows us to easily visualize it with 
-the \code{plot} method and \code{type="hazard"} for objects of class \code{singleEventCB}. 
-We plot the hazard functions for both treatment arms in Figure \ref{fig:cs1hazard}. 
-The pattern we see is consistent with the population-time plot shown in Figure \ref{fig:plot-erspc-data} C, where more events are observed at later follow-up times. 
+the `plot` method and `type="hazard"` for objects of class `singleEventCB`. 
+We plot the hazard functions for both treatment arms in Figure \@ref(fig:cs1hazard). 
+The pattern we see is consistent with the population-time plot shown in Figure \@ref(fig:plot-erspc-data) C, where more events are observed at later follow-up times. 
 The drop at the end can be explained by the fact that very few observations were followed for the entire 15 year period.
 
-```{r cs1hazard, eval=eval_cs1, echo=TRUE, fig.cap='Estimated hazard functions for control and screening groups in the ERSPC dataset.
+```{r cs1hazard, eval=TRUE, echo=TRUE, fig.cap='Estimated hazard functions for control and screening groups in the ERSPC dataset.
 Hazards are estimated from fitting a parametric model with casebase sampling as a function of the interaction between a cubic pspline basis (df=2) of follow-up time and treatment arm.
@@
 
-\FloatBarrier
-
 ### Absolute risk
 
-Next, the \code{absoluteRisk} function takes as input the \code{singleEventCB} object and returns a matrix where each column corresponds to the covariate profiles specified in the \code{newdata} argument, and each row corresponds to time points specified by the \code{time} argument:
-
-```{r, eval=eval_cs1, echo=TRUE}
+Next, the `absoluteRisk` function takes as input the `singleEventCB` object and returns a matrix where each column corresponds to the covariate profiles specified in the `newdata` argument, and each row corresponds to time points specified by the `time` argument:
+
+```{r, eval=TRUE, echo=TRUE}
 new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
@@
 
-```{r erspc-cox, eval = eval_cs1, echo=FALSE}
+```{r erspc-cox, eval = TRUE, echo=FALSE}
 surv_obj <- with(ERSPC, Surv(Follow.Up.Time, DeadOfPrCa))
@@
 We can subsequently compute confidence intervals for the risk function using 
-the method \code{confint.absRiskCB}:
-
-```{r eval = eval_cs1, echo = TRUE}
+the method `confint.absRiskCB`:
+
+```{r eval = TRUE, echo = TRUE}
 conf_ints <- confint(risk, fit)
@@
 
-In Figure \ref{fig:erspc-cif-conf}, we see the 95% confidence bands around the estimates. 
+In Figure \@ref(fig:erspc-cif-conf), we see the 95% confidence bands around the estimates. 
 We also overlay the Kaplan-Meier curves as a reference.
 
-```{r erspc-cif-conf, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\\% confidence bands."}
+```{r erspc-cif-conf, eval = TRUE, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\\% confidence bands."}
 
@@
 
-```{r bmtcrr-data, eval = eval_cs2, message = FALSE}
+```{r bmtcrr-data, eval = TRUE, message = FALSE}
 data(bmtcrr)
@@
 First, we look at a population-time plot to visualize the incidence density of both relapse and the competing events.
-In Figure \ref{fig:compPop}, failure times are highlighted on the plot using red dots for the event of interest (panel A) and blue dots for competing events (panel B).
+In Figure \@ref(fig:compPop), failure times are highlighted on the plot using red dots for the event of interest (panel A) and blue dots for competing events (panel B).
 In both panels, we see evidence of a non-constant hazard function: the density of points is larger at the beginning of follow-up than at the end.
 
-```{r compPop, fig.cap="Population-time plot for the stem-cell transplant study with both relapse and competing events.", echo=FALSE, eval = eval_cs2}
+```{r compPop, fig.cap="Population-time plot for the stem-cell transplant study with both relapse and competing events.", echo=FALSE, eval = TRUE}
 popTimeData <- popTime(data = bmtcrr, time = "ftime")
@@
 
-```{r bmtcrr-casebase-weibull, warning = FALSE, eval = eval_cs2, echo = TRUE}
+```{r bmtcrr-casebase-weibull, warning = FALSE, eval = TRUE, echo = TRUE}
 model_cb <- fitSmoothHazard(
@@
 
-```{r bmtcrr-cox, echo = TRUE, eval = eval_cs2}
+```{r bmtcrr-cox, echo = TRUE, eval = TRUE}
 library(survival)
@@
 
-```{r bmtcrr-cis, echo=FALSE, eval = eval_cs2}
+```{r bmtcrr-cis, echo=FALSE, eval = TRUE}
 # Table of coefficients
@@
   knitr::kable(
-    format = "latex", booktabs = TRUE,
+    # format = "latex", 
+    booktabs = knitr::is_latex_output(),
     col.names = c("Covariates", "HR", "95% Conf.", "HR", "95% Conf."),
@@
 From the fit object, we can extract both the hazard ratios and their corresponding confidence intervals.
-These quantities appear in Table \ref{tab:bmtcrr-cis}.
+These quantities appear in Table \@ref(tab:bmtcrr-cis).
 As we can see, the type of disease corresponds to a significant hazard ratio: the hazard for AML is about half that for ALL.
@@
 
-```{r cb_risk, warning = FALSE, cache = FALSE, eval = eval_cs2}
+```{r cb_risk, warning = FALSE, cache = FALSE, eval = TRUE}
 # Pick 100 equidistant points between 0 and 60 months
@@
 
-```{r fg_risk, eval = eval_cs2, echo = TRUE}
+```{r fg_risk, eval = TRUE, echo = TRUE}
 library(timereg)
@@
 
-```{r cox_risk, eval = eval_cs2, echo = TRUE}
+```{r cox_risk, eval = TRUE, echo = TRUE}
 # Estimate absolute risk curve
@@
 
-```{r bmtcrr-risk, echo = FALSE, fig.cap="\\label{fig:compAbsrisk} Cumulative Incidence curve for a fixed covariate profile and the two disease groups.
-The estimate obtained from case-base sampling is compared to the Fine-Gray and Aalen-Johansen estimates.", eval = eval_cs2}
+```{r compAbsrisk, echo = FALSE, fig.cap="Cumulative Incidence curve for a fixed covariate profile and the two disease groups.
+The estimate obtained from case-base sampling is compared to the Fine-Gray and Aalen-Johansen estimates.", eval = TRUE}
 risk_all <- dplyr::bind_rows(
@@
 
-Figure \ref{fig:compAbsrisk} shows the CI curves for all three models.
+Figure \@ref(fig:compAbsrisk) shows the CI curves for all three models.
 As we can see, all three approaches agree quite well for AML; however, for ALL, there seems to be a difference of about 5% between the Fine-Gray curve and the curves estimated using case-base sampling and Cox regression.
@@
 We note that this is different than the semiparametric model Coxnet, which regularizes the Cox partial likelihood.
-To illustrate this functionality, we use the dataset from the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) [@knaus1995support].\footnote{The original data is available online from the Department of Biostatistics at Vanderbilt University: \url{https://biostat.app.vumc.org/wiki/Main/SupportDesc}} The SUPPORT dataset tracks death in five American hospitals within individuals who are considered seriously ill.
+To illustrate this functionality, we use the dataset from the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) [@knaus1995support].^[The original data is available online from the Department of Biostatistics at Vanderbilt University: https://biostat.app.vumc.org/wiki/Main/SupportDesc] The SUPPORT dataset tracks death in five American hospitals within individuals who are considered seriously ill.
 The cleaned and imputed data consists of 9104 observations and 30 variables, and it is available as part of the \pkg{casebase} package.
-In the comparisons below, all covariates except \code{sps} and \code{aps} were used.
+In the comparisons below, all covariates except `sps` and `aps` were used.
 These two variables correspond to scores for predicting the outcome that were developed as part of the original study.
@@
 For our penalized case-base model, we opt for the natural log of time which corresponds to a Weibull distribution.
-For fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface to the \code{fitSmoothHazard} function.
-The \code{fitSmoothHazard} and \code{fitSmoothHazard.fit} functions sample the case and base series, calculate the required offset, and transform the data to match the expected input of the \pkg{glmnet} package.
-The penalized logistic regression is then fit for multiple values of the tuning parameter using the function \code{glmnet::cv.glmnet} and the binomial family.
-To \code{fitSmoothHazard.fit}, we supply both a matrix \code{y} containing the time and event variables, and a matrix \code{x} containing all other covariates.
-We apply the lasso penalty by setting \code{alpha = 1} and assign a \code{penalty.factor} ($w_j$; cf.
+For fitting the penalized hazard, we use `fitSmoothHazard.fit`, which is a matrix interface to the `fitSmoothHazard` function.
+The `fitSmoothHazard` and `fitSmoothHazard.fit` functions sample the case and base series, calculate the required offset, and transform the data to match the expected input of the \pkg{glmnet} package.
+The penalized logistic regression is then fit for multiple values of the tuning parameter using the function `glmnet::cv.glmnet` and the binomial family.
+To `fitSmoothHazard.fit`, we supply both a matrix `y` containing the time and event variables, and a matrix `x` containing all other covariates.
+We apply the lasso penalty by setting `alpha = 1` and assign a `penalty.factor` ($w_j$; cf.
 Equation \ref{eq:penest}) of 0 to the time variable to ensure it is in the selected model.
@@
 
-```{r supportData, eval = eval_cs3}
+```{r supportData, eval = TRUE}
 library(riskRegression)
@@
 
-```{r supportCox_fit, eval=eval_cs3, echo=FALSE, cache=FALSE}
+```{r supportCox_fit, eval=TRUE, echo=FALSE, cache=FALSE}
 # Cox with everything but sps and aps
@@
 
-```{r supportCB_fit, eval=eval_cs3, echo=TRUE, cache=FALSE}
+```{r supportCB_fit, eval=TRUE, echo=TRUE, cache=FALSE}
 # Create matrices for inputs
@@
 
-```{r coxHazAbsolute, echo = FALSE, eval = eval_cs3}
+```{r coxHazAbsolute, echo = FALSE, eval = TRUE}
 # Create survival object for fitting Coxnet
@@
 
-In Figure \ref{fig:cs3lolliPlot}, we show the coefficient estimates for covariates that we selected by both penalized Cox and penalized case-base.
+In Figure \@ref(fig:cs3lolliPlot), we show the coefficient estimates for covariates that we selected by both penalized Cox and penalized case-base.
 We note that both penalized approaches produce similar results.
@@
 
-```{r coefplots, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} Coefficient estimates from the Cox model (Cox), penalized Cox model using the \\pkg{glmnet} package (Pen.
+```{r cs3lolliPlot, echo=FALSE, eval=TRUE, fig.cap="Coefficient estimates from the Cox model (Cox), penalized Cox model using the glmnet package (Pen.
 Cox), and our approach using penalized case-base sampling (Pen.
@@
   labs(color = "Models") +
-  scale_color_manual(values = q4)
+  scale_color_manual(values = q4[-4])
 ```
@@
 We then compare the risk estimation over the test set.
-The probabilities over time for each observation are averaged, resulting in the absolute risk curves shown in Figure \ref{fig:cs3FinalBrier} (A).
+The probabilities over time for each observation are averaged, resulting in the absolute risk curves shown in Figure \@ref(fig:cs3FinalBrier) (A).
 We can see some minimal differences between the three models, with the Kaplan-Meier giving the lowest estimates across follow-up-time.
@@
 As described above, we compare the performance between the models by computing the Brier scores over time.
-In Figure \ref{fig:cs3FinalBrier} (B), we can see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models.
+In Figure \@ref(fig:cs3FinalBrier) (B), we can see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models.
 On the other hand, the differences between these three models are minimal.
 
-```{r support_abs, eval=eval_cs3, echo=FALSE}
+```{r support_abs, eval=TRUE, echo=FALSE}
 # Absolute Risks
@@
 
-```{r absRiskPlot, echo=FALSE, eval=eval_cs3}
+```{r absRiskPlot, echo=FALSE, eval=TRUE}
 reg_ci <- ggplot(data_absRisk, aes(Time, Prob)) +
@@
 
-```{r eval = eval_cs3, echo = FALSE, warning = FALSE}
+```{r eval = TRUE, echo = FALSE, warning = FALSE}
 # We need this chunk until this method reaches the CRAN version of riskRegression
@@
     #get all covariates excluding intercept and time
-    coVars=colnames(object$originalData$x)
+    coVars = colnames(object$originalData$x)
     #coVars is used in lines 44 and 50
-    newdata=data.matrix(drop(subset(newdata, select=coVars)))
+    newdata = data.matrix(drop(subset(newdata, select=coVars)))
   }
@@
 
-```{r support_Brier, eval=eval_cs3, echo=FALSE}
+```{r support_Brier, eval=TRUE, echo=FALSE}
 # Brier score
@@
 
-```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} Comparison of Cox regression (Cox), penalized Cox regression (Pen.
+```{r cs3FinalBrier, echo = FALSE, eval = TRUE, fig.cap="Comparison of Cox regression (Cox), penalized Cox regression (Pen.
 Cox), penalized case-base sampling estimation (Pen.
@@
 Accordingly, our package provides functions for each step.
-Moreover, the simple interface of the \code{fittingSmoothHazard} function resembles the \code{glm} interface.
+Moreover, the simple interface of the `fitSmoothHazard` function resembles the `glm` interface.
 This interface should look familiar to new users.
@@
 In the case studies above, we compared the performance of case-base sampling with that of Cox regression and Fine-Gray models.
-In terms of function interface, \pkg{casebase} uses a formula interface that is closer to that of \code{glm}, in that the event variable is the only variable appearing on the left-hand side of the formula.
-By contrast, both \code{survival::coxph} and \code{timereg::comp.risk} use arrays that capture both the event type and time.
+In terms of function interface, \pkg{casebase} uses a formula interface that is closer to that of `glm`, in that the event variable is the only variable appearing on the left-hand side of the formula.
+By contrast, both `survival::coxph` and `timereg::comp.risk` use arrays that capture both the event type and time.
 Both approaches to modeling yield user-friendly code.
@@
 However, as a downside, it does not allow us to model location and scale parameters separately like the package \pkg{flexsurv}.
-For example, if we look at the Weibull distribution as parametrised in \code{stats::pweibull}, the log-hazard function is given by
+For example, if we look at the Weibull distribution as parametrised in `stats::pweibull`, the log-hazard function is given by
 $$ \log \lambda(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,$$
@@
 Finally, we want to provide more flexibility in how the case-base sampling is performed.
-This could be achieved by adding a \code{hazard} argument to the function \code{sampleCaseBase}.
+This could be achieved by adding a `hazard` argument to the function `sampleCaseBase`.
 In this way, users could specify their own sampling mechanism.
@@
 We strongly believe that its flexibility and its foundation on the familiar logistic regression model will make it appealing to new and established practitioners.
-The \pkg{casebase} package is freely available from the Comprehensive R Archive Network at \url{https://cran.r-project.org/package=casebase}.
-Interested users can visit \url{http://sahirbhatnagar.com/casebase/} for detailed package documentation and vignettes.
+The \pkg{casebase} package is freely available from the Comprehensive R Archive Network at https://cran.r-project.org/package=casebase.
+Interested users can visit http://sahirbhatnagar.com/casebase/ for detailed package documentation and vignettes.
 
@@
 We would also like to thank Yi Yang for helpful discussions on penalized regression models.
-\mbox{Bhatnagar} (RGPIN-2020-05133) and \mbox{Turgeon} (RGPIN-2021-04073) both gratefully acknowledge funding via a Discovery Grant from the Natural Sciences and Engineering Research Council of Canada (NSERC), RGPIN.
+Bhatnagar (RGPIN-2020-05133) and Turgeon (RGPIN-2021-04073) both gratefully acknowledge funding via a Discovery Grant from the Natural Sciences and Engineering Research Council of Canada (NSERC), RGPIN.
 
@@
 
-```{r colophon, eval = eval_colophon, cache = FALSE}
+```{r colophon, eval = FALSE, cache = FALSE}
 # which R packages and versions?
@@
 <!-- The current Git commit details are: -->
-```{r, eval = eval_colophon}
+```{r, eval = FALSE}
 # what commit is this file at?

### 2022-11-11 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
switch to distill

@@
 % !TeX root = RJwrapper.tex
-\title{\pkg{casebase}: An Alternative Framework For Survival Analysis
-and Comparison of Event Rates}
-\author{by Sahir Rai Bhatnagar*, Maxime Turgeon*, Jesse Islam, James A.
-Hanley, and Olli Saarela}
+\title{\pkg{casebase}: An Alternative Framework for Survival Analysis and Comparison of Event Rates}
+\author{by Sahir Rai Bhatnagar*, Maxime Turgeon*, Jesse Islam, James A. Hanley, and Olli Saarela}
 
@@
 \abstract{%
-In clinical studies of time-to-event data, a quantity of interest to the
-clinician is their patient's risk of an event. However, methods relying
-on time matching or risk-set sampling (including Cox regression)
-eliminate the baseline hazard from the estimating function. As a
-consequence, the focus has been on reporting hazard ratios instead of
-survival or cumulative incidence curves. Indeed, reporting patient risk
-or cumulative incidence requires a separate estimation of the baseline
-hazard. Using case-base sampling, Hanley \& Miettinen (2009) explained
-how parametric hazard functions can be estimated in continuous-time
-using logistic regression. Their approach naturally leads to estimates
-of the survival or risk function that are smooth-in-time.\\
-In this paper, we present the \CRANpkg{casebase} R package, a
-comprehensive and flexible toolkit for parametric survival analysis. We
-describe how the case-base framework can also be used in more complex
-settings: non-linear functions of time and non-proportional hazards,
-competing risks, and variable selection. Our package also includes an
-extensive array of visualization tools to complement the analysis. We
-illustrate all these features through three different case studies.\\
+In clinical studies of time-to-event data, a quantity of interest to the clinician is their patient's risk of an event.
+However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the estimating function.
+As a consequence, the focus has been on reporting hazard ratios instead of survival or cumulative incidence curves.
+Indeed, reporting patient risk or cumulative incidence requires a separate estimation of the baseline hazard.
+Using case-base sampling, Hanley \& Miettinen (2009) explained how parametric hazard functions can be estimated in continuous-time using logistic regression.
+Their approach naturally leads to estimates of the survival or risk function that are smooth-in-time.
+In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis.
+We describe how the case-base framework can also be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection.
+Our package also includes an extensive array of visualization tools to complement the analysis.
+We illustrate all these features through three different case studies.
 * SRB and MT contributed equally to this work.
@@
 
-The semiparametric Cox model has become the default approach to survival
-analysis even though Cox himself later suggested he would prefer to
-model the hazard function directly. In a 1994 interview with Professor
-Nancy Reid, Sir David Cox was asked how he would model a set of censored
-survival data, to which he responded: ``I think I would normally want to
-tackle problems parametrically \ldots{} and if you want to do things
-like predict the outcome for a particular patient, it's much more
-convenient to do that parametrically'' \citep{reid1994conversation}.
-Indeed, the most relevant quantity in a clinical setting is often the 5-
-or 10-year risk of experiencing a certain event given the patient's
-particular profile. However, the most reported metric from a Cox model
-is the (potentially time-dependent) hazard ratio (HR). The
-covariate-conditional survival curve is arguably a more important
-summary measure to report than the HR. While stepwise survival curves
-can be computed with the Cox model, they require a second step to
-separately estimate the baseline hazard \citep{breslow1972discussion}.
-
-Several authors have since pursued fully parametric approaches that made
-the fitting of smooth survival curves more transparent and intuitive
-through generalized linear models. The key feature of these procedures
-is splitting the time axis into discrete intervals. Whitehead
-\citeyearpar{whitehead1980fitting} showed the equivalence between a Cox
-model and a Poisson regression with a parameter for each event time;
-Carstensen \citeyearpar{copenhagen2012needs} provides an exposition of
-this equivalence with a real data example and supporting R code for
-computing standard errors. Arjas \& Haara
-\citeyearpar{arjas1987logistic} and Efron
-\citeyearpar{efron1988logistic} treated each patient-day as a Bernoulli
-random variable with probability equal to the discrete hazard rate. A
-potential issue with these approaches is that the number of time bins
-need to be chosen by the data analyst. On the one hand, a fine grouping
-of times may result in few (or none) events per interval, which then
-leads to instability in the Newton-Raphson procedure for estimation
-\citep[Section 4.8]{kalbfleisch2011statistical}, and large long-format
-datasets. On the other hand, a coarse grouping could potentially mask
+The semiparametric Cox model has become the default approach to survival analysis
+even though Cox himself later suggested he would prefer to model the hazard function
+directly. In a 1994 interview with Professor Nancy Reid, Sir David Cox was asked
+how he would model a set of censored survival data, to which he responded: ``I
+think I would normally want to tackle problems parametrically \ldots{} and if you want
+to do things like predict the outcome for a particular patient, it's much more
+convenient to do that parametrically'' (Reid 1994). Indeed, the most
+relevant quantity in a clinical setting is often the 5- or 10-year risk of
+experiencing a certain event given the patient's particular profile.
+However, the most reported metric from a Cox model is the (potentially
+time-dependent) hazard ratio (HR). The covariate-conditional
+survival curve is arguably a more important
+summary measure to report than the HR. While stepwise survival curves can be
+computed with the Cox model, they require a second step to separately estimate
+the baseline hazard (Breslow 1972).
+
+Several authors have since pursued fully parametric approaches that made the
+fitting of smooth survival curves more transparent and intuitive through
+generalized linear models. The key feature of these procedures is splitting the
+time axis into discrete intervals. Whitehead (1980) showed the
+equivalence between a Cox model and a Poisson regression with a parameter for
+each event time; Carstensen (2019) provides an exposition of this
+equivalence with a real data example and supporting R code for computing
+standard errors. Arjas \& Haara (1987) and Efron (1988) treated each
+patient-day as a Bernoulli random variable with probability equal to the
+discrete hazard rate. A potential issue with these approaches is that the
+number of time bins need to be chosen by the data analyst. On the one hand, a fine grouping
+of times may result in few (or none) events per interval, which then leads
+to instability in the Newton-Raphson procedure for estimation
+(Kalbfleisch and Prentice 2011, vol. 360, sec. 4.8), and large long-format datasets.
+On the other hand, a coarse grouping could potentially mask
 nonlinear trends in the hazard function.
 
-Rather than discretizing time, Hanley \& Miettinen
-\citeyearpar{hanley2009fitting} selected a discrete set of person-time
-coordinates (``person-moments'') in continuous time from all observed
-follow-up experience constituting the study base. By doing so, they
-obtained a likelihood expression for the hazard function that is
-equivalent to that of logistic regression with an offset term. More
-specifically, all person-moments when the event of interest occurred are
-selected as the case series, complemented by a randomly sampled base
-series of person-moments serving as controls. This approach allows
-flexible modeling of the hazard function by including functions of time
-as covariates (e.g.~using splines or generalized additive models).
-Furthermore, time-dependent covariates can be modeled through
-interactions with time. In short, Hanley \& Miettinen
-\citeyearpar{hanley2009fitting} use the well-understood logistic
-regression for directly modeling the hazard function, without requiring
-a discrete-time model.
-
-In this article, we present the \CRANpkg{casebase} R package
-\citep{casebase-package} which implements and extends the Hanley \&
-Miettinen \citeyearpar{hanley2009fitting} approach for fitting fully
-parametric hazard models and covariate-conditional survival curves using
-the familiar interface of the \code{glm} function. Our implementation
-includes extensions to other models such as penalized regression for
-variable selection and competing risk analysis. In addition, we provide
-functions for exploratory data analysis and visualizing the estimated
-quantities such as the hazard function, survival curve, and their
-standard errors. The ultimate goal of our package is to make fitting
-flexible hazards accessible to end users who favor reporting absolute
-risks and survival curves over hazard ratios.
-
-In what follows, we first recap some theoretical details on case-base
-sampling and its use for estimating parametric hazard functions. We then
-give a short review of existing R packages that implement comparable
-features to \pkg{casebase}. Next, we provide some details about the
-implementation of case-base sampling in our package, and we give a brief
-survey of its main functions. This is followed by three case studies
-that illustrate the flexibility and capabilities of \pkg{casebase}. We
-show how the same framework can be used for non-linear functions of time
-and non-proportional hazards, competing risks, and variable selection
-via penalized regression. Finally, we end the article with a discussion
-of the results and of future directions.
+Rather than discretizing time, Hanley \& Miettinen (2009)
+selected a discrete set of person-time coordinates (``person-moments'') in
+continuous time from all observed follow-up experience constituting the
+study base. By doing so, they obtained a likelihood expression for the hazard
+function that is equivalent to that of logistic regression with an offset term.
+More specifically, all person-moments when the event of interest
+occurred are selected as the case series, complemented by a randomly sampled
+base series of person-moments serving as controls. This approach allows
+flexible modeling of the hazard function by including functions of time as covariates
+(e.g.~using splines or generalized additive models). Furthermore, time-dependent
+covariates can be modeled through interactions with time. In short, Hanley \&
+Miettinen (2009) use the well-understood logistic regression
+for directly modeling the hazard function, without requiring a discrete-time model.
+
+In this article, we present the \CRANpkg{casebase} R package (Bhatnagar et al. 2021)
+which implements and extends the Hanley \& Miettinen (2009) approach for
+fitting fully parametric hazard models and covariate-conditional survival curves
+using the familiar interface of the \texttt{glm} function. Our implementation
+includes extensions to other models such as penalized
+regression for variable selection and competing risk analysis. In addition,
+we provide functions for exploratory data analysis and visualizing the estimated
+quantities such as the hazard function, survival curve, and their standard errors.
+The ultimate goal of our package is to make fitting flexible hazards accessible
+to end users who favor reporting absolute risks and survival curves
+over hazard ratios.
+
+In what follows, we first recap some theoretical details on case-base sampling
+and its use for estimating parametric hazard functions. We then give a short
+review of existing R packages that implement comparable features to \pkg{casebase}.
+Next, we provide some details about the implementation of case-base sampling in
+our package, and we give a brief survey of its main functions. This is followed
+by three case studies that illustrate the flexibility and capabilities of
+\pkg{casebase}. We show how the same framework can be used for non-linear
+functions of time and non-proportional hazards, competing risks, and variable
+selection via penalized regression. Finally, we end the article with a
+discussion of the results and of future directions.
 
@@
 
-As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
-key idea behind case-base sampling is to consider the entire study base
-as an infinite collection of \emph{person-moments}. These person-moments
-are indexed by both an individual in the study and a time point, and
-therefore each person-moment has a covariate profile and an outcome
-status (i.e.~whether the event happened) attached to it. By estimating
-the probability of an event occurring at a particular person-moment, we
-can extract information about hazards and survival.
-
-Therefore, we start by assembling all person-moments at which the event
-occurred; this collection of person-moments is what Hanley \& Miettinen
-call the \emph{case series}. The incidence of the case series is
-dictated by the hazard function of interest. Next, we sample a finite
-number of person-moments (blinded to case moments); this second
-collection of person-moments is what Hanley \& Miettinen call the
-\emph{base series}. The sampling mechanism for the base series is left
-at the discretion of the user, but in practice we find that sampling
-uniformly from the study base provides both simplicity and good
-performance. This is the default sampling mechanism in the package.
+As discussed in Hanley \& Miettinen (2009), the key idea behind case-base sampling is to consider the entire study base as an infinite collection of \emph{person-moments}.
+These person-moments are indexed by both an individual in the study and a time point, and therefore each person-moment has a covariate profile and an outcome status (i.e.~whether the event happened) attached to it.
+By estimating the probability of an event occurring at a particular person-moment, we can extract information about hazards and survival.
+
+Therefore, we start by assembling all person-moments at which the event occurred; this collection of person-moments is what Hanley \& Miettinen call the \emph{case series}.
+The incidence of the case series is dictated by the hazard function of interest.
+Next, we sample a finite number of person-moments (blinded to case moments); this second collection of person-moments is what Hanley \& Miettinen call the \emph{base series}.
+The sampling mechanism for the base series is left at the discretion of the user, but in practice we find that sampling uniformly from the study base provides both simplicity and good performance.
+This is the default sampling mechanism in the package.
 
 \hypertarget{likelihood-and-estimating-function}{%
-\subsection{Likelihood and estimating
-function}\label{likelihood-and-estimating-function}}
-
-To describe the theoretical foundations of case-base sampling, we use
-the framework of counting processes. In what follows, we abuse notation
-slightly and omit any mention of \(\sigma\)-algebras. Instead, following
-Aalen \textit{et al} \citeyearpar{aalen2008survival}, we use the
-placeholder ``past'' to denote the past history of the corresponding
-process. The reader interested in more details can refer to Saarela \&
-Arjas \citeyearpar{saarela2015non} and Saarela
-\citeyearpar{saarela2016case}. First, let \(N_{i}(t) \in \{0, 1\}\) be
-counting processes corresponding to the event of interest for individual
-\(i=1, \ldots,n\). For simplicity, we will consider Type I censoring due
-to the end of follow-up at time \(\tau\) (the general case of
-independent censoring is treated in Saarela
-\citeyearpar{saarela2016case}). We are interested in modeling the hazard
-functions \(\lambda_{i}(t)\) of the processes \(N_i(t)\), and which
-satisfy
-\[\lambda_{i}(t)\,\textrm dt = E[\,\textrm dN_{i}(t)\mid \mathrm{past}].\]
-The processes \(N_i(t)\) count the person-moments from the \emph{case
-series}.
-
-To complement the case series, we sample person-moments for the base
-series. To do so, we specify the base series sampling mechanism as
-non-homogeneous Poisson processes \(R_i(t) \in \{0, 1, 2, \ldots\}\);
-the person-moments where \(\,\textrm dR_i(t) = 1\) constitute the base
-series. We note that the same individual can contribute multiple
-person-moments to the base series. The process
-\(Q_{i}(t) = R_i(t) + N_{i}(t)\) then counts both the case and base
-series person-moments contributed by individual \(i\). As mentioned
-above, the processes \(R_i(t)\) are specified by the user via its
-intensity function \(\rho_i(t)\). The process \(Q_{i}(t)\) is
-characterized by
-\(E[\,\textrm dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)\,\textrm dt + \rho_i(t)\,\textrm dt\).
-
-If the hazard function \(\lambda_{i}(t; \theta)\) is parametrized in
-terms of \(\theta\), we can define an estimator \(\hat{\theta}\) by
-maximization of the likelihood expression
+\subsection{Likelihood and estimating function}\label{likelihood-and-estimating-function}}
+
+To describe the theoretical foundations of case-base sampling, we use the framework of counting processes.
+In what follows, we abuse notation slightly and omit any mention of \(\sigma\)-algebras.
+Instead, following Aalen \textit{et al} (2008), we use the placeholder ``past'' to denote the past history of the corresponding process.
+The reader interested in more details can refer to Saarela \& Arjas (2015) and Saarela (2016).
+First, let \(N_{i}(t) \in \{0, 1\}\) be counting processes corresponding to the event of interest for individual \(i=1, \ldots,n\).
+For simplicity, we will consider Type I censoring due to the end of follow-up at time \(\tau\) (the general case of independent censoring is treated in Saarela (2016)).
+We are interested in modeling the hazard functions \(\lambda_{i}(t)\) of the processes \(N_i(t)\), and which satisfy
+\[\lambda_{i}(t)\,\textrm dt = E[\,\textrm dN_{i}(t)\mid \mathrm{past}].\] The processes \(N_i(t)\) count the person-moments from the \emph{case series}.
+
+To complement the case series, we sample person-moments for the base series.
+To do so, we specify the base series sampling mechanism as non-homogeneous Poisson processes \(R_i(t) \in \{0, 1, 2, \ldots\}\); the person-moments where \(\,\textrm dR_i(t) = 1\) constitute the base series.
+We note that the same individual can contribute multiple person-moments to the base series.
+The process \(Q_{i}(t) = R_i(t) + N_{i}(t)\) then counts both the case and base series person-moments contributed by individual \(i\).
+As mentioned above, the processes \(R_i(t)\) are specified by the user via its intensity function \(\rho_i(t)\).
+The process \(Q_{i}(t)\) is characterized by \(E[\,\textrm dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)\,\textrm dt + \rho_i(t)\,\textrm dt\).
+
+If the hazard function \(\lambda_{i}(t; \theta)\) is parametrized in terms of \(\theta\), we can define an estimator \(\widehat{\theta}\) by maximization of the likelihood expression
 \[L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^{\min(t_i,\tau)} \lambda_i(t; \theta)\,\textrm dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{\,\textrm dN_{i}(t)},\]
-where \(\prod_{t\in[0,u)}\) represents a product integral from \(0\) to
-\(u\), and where \(t_i\) is the event time for individual \(i\).
-However, the integral over time makes the computation and maximization
-of \(L_0(\theta)\) challenging using standard software.
-
-Case-base sampling allows us to avoid this integral. By conditioning on
-a sampled person-moment, we get individual partial likelihood
-contributions of the form
+where \(\prod_{t\in[0,u)}\) represents a product integral from \(0\) to \(u\), and where \(t_i\) is the event time for individual \(i\).
+However, the integral over time makes the computation and maximization of \(L_0(\theta)\) challenging using standard software.
+
+Case-base sampling allows us to avoid this integral. By conditioning on a sampled person-moment, we get individual partial likelihood contributions of the form
 \[P(\,\textrm dN_{i}(t) \mid \,\textrm dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.\]
-By using the symbol \(\stackrel{\theta}{\propto}\), we mean that both
-sides of the expression are equal up to multiplicative factors that do
-not depend on \(\theta\). Therefore, an estimating function for
-\(\theta\) can be composed of these contributions as: \begin{equation}
+By using the symbol \(\stackrel{\theta}{\propto}\), we mean that both sides of the expression are equal up to multiplicative factors that do not depend on \(\theta\).
+Therefore, an estimating function for \(\theta\) can be composed of these contributions as:
+\begin{equation}
 L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{\,\textrm dQ_i(t)}. \label{eq:lik-function}
-\end{equation} When a logarithmic link function is used for modeling the
-hazard function, the above expression is of a logistic regression form
-with an offset term \(\log(1/\rho_i(t))\). Note that the sampling units
-selected in the case-base sampling mechanism are person-moments, rather
-than individuals, and the parameters to be estimated are hazards or
-hazard ratios rather than odds or odds ratios. Generally, an individual
-can contribute more than one person-moment, and thus the terms in the
-product integral are not independent. Nonetheless, Saarela
-\citeyearpar{saarela2016case} showed that the corresponding partial
-likelihood score function has mean zero at the true value
-\(\theta=\theta_0\), and that the resulting estimator \(\hat{\theta}\)
-is asymptotically normally distributed.
-
-In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
-suggest sampling the base series \emph{uniformly} from the study base.
-In terms of Poisson processes, their sampling strategy corresponds to a
-time-homogeneous Poisson process with intensity equal to
-\(\rho_i(t) = b/B\), where \(b\) is the number of sampled base series
-\emph{person-moments}, and \(B\) is the total population-time for the
-study base (e.g.~the sum of all individual follow-up times). More
-complex examples are also possible; see for example Saarela \& Arjas
-\citeyearpar{saarela2015non}, where the intensity functions for the
-sampling mechanism are proportional to the cardiovascular disease event
-rate given by the Framingham score. Non-uniform sampling mechanisms can
-increase the efficiency of the resulting maximum partial likelihood
-estimators.
+\end{equation}
+When a logarithmic link function is used for modeling the hazard function, the above expression is of a logistic regression form with an offset term \(\log(1/\rho_i(t))\).
+Note that the sampling units selected in the case-base sampling mechanism are person-moments, rather than individuals, and the parameters to be estimated are hazards or hazard ratios rather than odds or odds ratios.
+Generally, an individual can contribute more than one person-moment, and thus the terms in the product integral are not independent.
+Nonetheless, Saarela (2016) showed that the corresponding partial likelihood score function has mean zero at the true value \(\theta=\theta_0\), and that the resulting estimator \(\widehat{\theta}\) is asymptotically normally distributed.
+
+In Hanley \& Miettinen (2009), the authors suggest sampling the base series \emph{uniformly} from the study base.
+In terms of Poisson processes, their sampling strategy corresponds to a time-homogeneous Poisson process with intensity equal to \(\rho_i(t) = b/B\), where \(b\) is the number of sampled base series \emph{person-moments}, and \(B\) is the total population-time for the study base (e.g.~the sum of all individual follow-up times).
+More complex examples are also possible; see for example Saarela \& Arjas (2015), where the intensity functions for the sampling mechanism are proportional to the cardiovascular disease event rate given by the Framingham score.
+Non-uniform sampling mechanisms can increase the efficiency of the resulting maximum partial likelihood estimators.
 
@@
 
-The asymptotic normality of \(\hat{\theta}\) gives us an efficient way
-to estimate the variance of various estimators derived from the hazard
-function. For example, to construct confidence bands around risk
-functions and survival functions, we can use the following procedure:
+The asymptotic normality of \(\widehat{\theta}\) gives us an efficient way to estimate the variance of various estimators derived from the hazard function.
+For example, to construct confidence bands around risk functions and survival functions, we can use the following procedure:
 
@@
 \item
-  Compute \(\hat{\theta}\) and its variance-covariance matrix
-  \(V(\hat \theta)\) which is obtained as part of the logistic
-  regression output.
+  Compute \(\widehat{\theta}\) and its variance-covariance matrix \(V(\widehat \theta)\) which is obtained as part of the logistic regression output.
 \item
-  Sample \(B\) times from a multivariate normal
-  \(N\left(\hat{\theta}, V(\hat\theta)\right)\).
+  Sample \(B\) times from a multivariate normal \(N\left(\widehat{\theta}, V(\widehat\theta)\right)\).
 \item
-  For each sample, compute the survival function using Equation
-  \ref{eqn:surv} below.
+  For each sample, compute the survival function using Equation \ref{eqn:surv} below.
 \item
-  Use the pointwise quantiles of these survival function estimates to
-  construct a pointwise confidence band for the survival function of
-  interest.
+  Use the pointwise quantiles of these survival function estimates to construct a pointwise confidence band for the survival function of interest.
 \end{enumerate}
 
-This procedure is similar to parametric bootstrap \citep[Section
-6.5]{efron1994introduction}, but it can be more accurately described as
-a form of approximate Bayesian computation. As such, the validity of the
-confidence bands relies on the Bernstein-von Mises theorem
-\citep[Chapter 10]{van2000asymptotic}.
+This procedure is similar to parametric bootstrap (Efron and Tibshirani 1994, sec.
+6.5), but it can be more accurately described as a form of approximate Bayesian computation.
+As such, the validity of the confidence bands relies on the Bernstein-von Mises theorem (Van der Vaart 2000, vol.
+3, chap.
+10).
 
@@
 
-Here we show that certain named distributions such as exponential,
-Weibull or Gompertz can be fit using our framework, though we are not
-restricted to such models. Let \(g(t; X)\) be the linear predictor such
-that \(\log(\lambda(t;X)) = g(t; X)\). Different functions of \(t\) lead
-to different parametric hazard models. The simplest of these models is
-the one-parameter exponential distribution which is obtained by taking
-the hazard function to be constant over the range of \(t\):
+Here we show that certain named distributions such as exponential, Weibull or Gompertz can be fit using our framework, though we are not restricted to such models.
+Let \(g(t; X)\) be the linear predictor such that \(\log(\lambda(t;X)) = g(t; X)\).
+Different functions of \(t\) lead to different parametric hazard models.
+The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of \(t\):
 
@@
 \log(\lambda(t; X)) = \beta_0 + \beta_1 X. \label{eq:exp}
-\end{equation} In this model, the instantaneous failure rate is
-independent of
-\(t\).\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
-This is also known as the \dfn{memoryless property} \citep{kalbfleisch2011statistical}.}
+\end{equation}
+In this model, the instantaneous failure rate is independent of \(t\).\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study.
+This is also known as the \emph{memoryless property} (Kalbfleisch and Prentice 2011).}
 
@@
 
-Use of \(\log(t)\) yields the Weibull hazard which allows for a power
-dependence of the hazard on time \citep{kalbfleisch2011statistical}:
+Use of \(\log(t)\) yields the Weibull hazard which allows for a power dependence of the hazard on time (Kalbfleisch and Prentice 2011):
 
@@
 
-Case-base sampling can also be used in the context of competing risk
-analysis. Assuming there are \(J\) competing events, we can show that
-each sampled person-moment's contribution to the partial likelihood is
-of the form
+Case-base sampling can also be used in the context of competing risk analysis.
+Assuming there are \(J\) competing events, we can show that each sampled person-moment's contribution to the partial likelihood is of the form
 
 \[\frac{\lambda_j(t)^{\,\textrm dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},\]
-where \(N_j(t)\) is the counting process associated with the event of
-type \(j\) and \(\lambda_j(t)\) is the corresponding cause-specific
-hazard function. As may be expected, this functional form is similar to
-the terms appearing in the likelihood function for multinomial
-regression with an offset
-term.\footnote{Specifically, it corresponds to the following parametrization  for a multinomial random variable $Y$: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j + log(1/\rho), \qquad j = 1,\ldots, J-1.\end{align*}}
+where \(N_j(t)\) is the counting process associated with the event of type \(j\) and \(\lambda_j(t)\) is the corresponding cause-specific hazard function.
+As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression with an offset term.\footnote{Specifically, it corresponds to the following parametrization for a multinomial random variable \(Y\): \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j + log(1/\rho), \qquad j = 1,\ldots, J-1.\end{align*}}
 
@@
 
-To perform variable selection on the regression parameters
-\(\theta \in \mathbb{R}^p\) of the hazard function, we can add a penalty
-to the likelihood and optimise the following equation: \begin{equation}
+To perform variable selection on the regression parameters \(\theta \in \mathbb{R}^p\) of the hazard function, we can add a penalty to the likelihood and optimise the following equation:
+\begin{equation}
 \min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j p_{\lambda,\alpha}(\theta_j) \label{eq:penest}
-\end{equation} where \(\ell\left(\theta\right) = \log L(\theta)\) is the
-log of the likelihood function given in \eqref{eq:lik-function},
-\(p_{\lambda,\alpha}(\theta_j)\) is a penalty term controlled by the
-non-negative regularization parameters \(\lambda\) and \(\alpha\), and
-\(w_j\) is the penalty factor for the \(j\)th covariate. These penalty
-factors serve as a way of allowing parameters to be penalized
-differently. For example, we could set the penalty factor for time to be
-0 to ensure it is always included in the selected model.
+\end{equation}
+where \(\ell\left(\theta\right) = \log L(\theta)\) is the log of the likelihood function given in \eqref{eq:lik-function}, \(p_{\lambda,\alpha}(\theta_j)\) is a penalty term controlled by the non-negative regularization parameters \(\lambda\) and \(\alpha\), and \(w_j\) is the penalty factor for the \(j\)th covariate.
+These penalty factors serve as a way of allowing parameters to be penalized differently.
+For example, we could set the penalty factor for time to be 0 to ensure it is always included in the selected model.
 
 \hypertarget{comparison-with-existing-packages}{%
-\section{Comparison with existing
-packages}\label{comparison-with-existing-packages}}
-
-Survival analysis is an important branch of applied statistics and
-epidemiology. Accordingly, there is already a vast ecosystem of R
-packages implementing different methodologies. In this section, we
-describe how the functionalities of \pkg{casebase} compare to these
-packages.
-
-At the time of writing, a cursory examination of CRAN's \ctv{Survival}
-Task View reveals that there are over 250 packages related to survival
-analysis \citep{survTaskView}. For the purposes of this article, we
-restricted our review to packages that implement at least one of the
-following features: parametric modeling, non-proportional hazard models,
-competing risk analysis, penalized estimation, and Cumulative Incidence
-(CI) estimation. By searching for appropriate keywords in the
-\code{DESCRIPTION} file of these packages, we found 60 relevant
-packages. These 60 packages were then manually examined to determine
-which ones are comparable to \pkg{casebase}. In particular, we excluded
-packages that were focused on a different set of problems, such as
-frailty and multistate models. The remaining 14 packages appear in Table
-\ref{tab:surv-pkgs}, along with some of the functionalities they offer.
-
-Parametric survival models are implemented in several packages, each
-differing in the parametric distributions available: \CRANpkg{CFC}
-\citeyearpar{mahani2015bayesian}, \CRANpkg{flexsurv}
-\citeyearpar{flexsurv}, \CRANpkg{SmoothHazard}
-\citeyearpar{smoothHazard}, \CRANpkg{rstpm2} \citeyearpar{clements_liu},
-\CRANpkg{mets} \citeyearpar{scheike2014estimating}, and
-\CRANpkg{survival} \citeyearpar{survival-package}. For example,
-\pkg{SmoothHazard} is limited to Weibull distributions
-\citeyearpar{smoothHazard}, whereas both \pkg{flexsurv} and
-\pkg{survival} allow users to supply any distribution of their choice.
-\pkg{flexsurv}, \pkg{SmoothHazard}, \pkg{mets} and \pkg{rstpm2} can
-model the effect of time using splines, which allows flexible modeling
-of the hazard function. As discussed above, \pkg{casebase} can model any
-parametric family whose log-hazard can be expressed as a linear
-combination of covariates (including time). Therefore, our package is
-more general in that it allows the user to model any linear or
-non-linear transformation of time including splines and higher order
-polynomials. Also, by including interaction terms between covariates and
-time, it also allows users to fit (non-proportional) time-varying
-coefficient models. However, unlike \pkg{flexsurv}, we do not explicitly
-model any shape parameter.
-
-Several packages implement penalized estimation for the Cox model:
-\CRANpkg{glmnet} \citeyearpar{regpathcox}, \CRANpkg{glmpath}
-\citeyearpar{park_hastie}, \CRANpkg{penalized} \citeyearpar{l1penal},
-\CRANpkg{riskRegression} \citeyearpar{gerds_blanche}. Moreover, some
-packages also include penalized estimation in the context of Cox models
-with time-varying coefficients: elastic-net penalization with
-\pkg{rstpm2} \citeyearpar{clements_liu}, while \pkg{survival}
-\citeyearpar{survival-package} has an implementation of ridge-penalized
-estimation. On the other hand, our package \pkg{casebase} provides
-penalized estimation of the hazard function. To our knowledge,
-\pkg{casebase} and \pkg{rstpm2} are the only packages to offer this
-functionality.
-
-Next, several R packages implement methodologies for competing risk
-analysis; for a different perspective on this topic, see Mahani \&
-Sharabiani \citeyearpar{mahani2015bayesian}. The package \pkg{survival}
-provides functionality for competing risk analysis and multistate
-modelling. The package \CRANpkg{cmprsk} provides methods for
-cause-specific subdistribution hazards, such as in the Fine-Gray model
-\citeyearpar{fine1999proportional}. On the other hand, the package
-\pkg{CFC} estimates cause-specific CIs from unadjusted, non-parametric
-survival functions. Our package \pkg{casebase} also provides
-functionalities for competing risk analysis by estimating parametrically
-the cause-specific hazards. From these quantities, we can then estimate
-the cause-specific CIs.
-
-Finally, several packages include functions to estimate the survival
-function and the CI. The corresponding methods generally fall into two
-categories: transformation of the estimated hazard function, and
-semi-parametric estimation of the baseline hazard. The first category
-broadly corresponds to parametric survival models, where the full hazard
-is explicitly modeled. Using this estimate, the survival function and
-the CI can be obtained using their functional relationships (see
-Equations \ref{eqn:surv} and \ref{eqn:CDF} below). Packages providing
-this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and
-\pkg{survival}. Our package \pkg{casebase} also follows this approach
-for both single-event and competing risk analyses. The second category
-outlined above broadly corresponds to semi-parametric models. These
-models do not model the full hazard function, and therefore the baseline
-hazard needs to be estimated separately in order to estimate the
-survival function. This is achieved using semi-parametric estimators
-(e.g.~Breslow's estimator) or parametric estimators (e.g.~spline
-functions). Packages that implement this approach include
-\pkg{riskRegression}, \pkg{rstpm2}, \pkg{survival}, and \pkg{glmnet}. As
-mentioned in the introduction, a key distinguishing factor between these
-two approaches is that the first category leads to smooth estimates of
-the survival function, whereas the second category often produces
-estimates in the form of stepwise functions.
-
-\begin{table}[ht]
-\centering
-\resizebox{\textwidth}{!}{%
-\begin{tabular}{ccccccccc}
-\toprule
-& \textbf{Competing} & \textbf{Allows} &  \textbf{Penalized}   &    &  & \textbf{Semi} & \textbf{Interval/Left} & \textbf{Risk}  \\
-\textbf{Package}        & \textbf{Risks}  & \textbf{Non PH} & \textbf{Regression} & \textbf{Splines} & \textbf{Parametric} & \textbf{Parametric} & \textbf{Censoring} & \textbf{Estimates}           \\
-\bottomrule
-\textbf{casebase}        & \checkmark                        & \checkmark                         & \checkmark                     & \checkmark                & \checkmark                   &                          &                                  & \checkmark                                \\ \hline
-\textbf{CFC}             & \checkmark                        & \checkmark                         &                       &                  & \checkmark                   &                          &                                  & \checkmark                                \\ \hline
-\textbf{cmprsk}             & \checkmark                        &                           &                       &                  &                     &  \checkmark                        &                                  & \checkmark                                \\ \hline
-\textbf{crrp}            & \checkmark                        &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
-\textbf{fastcox}         &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
-\textbf{flexrsurv}       &                          & \checkmark                         &                       & \checkmark                & \checkmark                   &                          &                                  & \checkmark               \\ \hline
-\textbf{flexsurv}        & \checkmark                        & \checkmark                         &                       & \checkmark                & \checkmark                   &                          &                                  & \checkmark               \\ \hline
-\textbf{glmnet}          &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &  \checkmark              \\ \hline
-\textbf{glmpath}         &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
-\textbf{mets}            & \checkmark                        &                           &                       & \checkmark                &                     & \checkmark                        &                                  & \checkmark                                \\ \hline
-\textbf{penalized}       &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
-\textbf{riskRegression}  & \checkmark                         &                           & \checkmark                     &                  &                     & \checkmark                        &                                  & \checkmark                                \\ \hline
-\textbf{rstpm2}          &                          & \checkmark                         &                      & \checkmark                & \checkmark                   & \checkmark                        & \checkmark                                & \checkmark                         \\ \hline
-\textbf{SmoothHazard}    &                          & \checkmark                         &                       & \checkmark                & \checkmark                   &                          & \checkmark                            &                                      \\ \hline
-\textbf{survival}        & \checkmark                        & \checkmark                         &                       &                  & \checkmark                   & \checkmark                        & \checkmark                                & \checkmark                               \\
-\bottomrule
-\end{tabular}%
-}
-\caption{Comparison of various R packages for survival analysis.
+\section{Comparison with existing packages}\label{comparison-with-existing-packages}}
+
+Survival analysis is an important branch of applied statistics and epidemiology.
+Accordingly, there is already a vast ecosystem of R packages implementing different methodologies.
+In this section, we describe how the functionalities of \CRANpkg{casebase} compare to these packages.
+
+At the time of writing, a cursory examination of CRAN's \ctv{Survival} Task View reveals that there are over 250 packages related to survival analysis (Allignol and Latouche 2019).
+For the purposes of this article, we restricted our review to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and Cumulative Incidence (CI) estimation.
+By searching for appropriate keywords in the \texttt{DESCRIPTION} file of these packages, we found 60 relevant packages.
+These 60 packages were then manually examined to determine which ones are comparable to \pkg{casebase}.
+In particular, we excluded packages that were focused on a different set of problems, such as frailty and multistate models.
+The remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with some of the functionalities they offer.
+
+Parametric survival models are implemented in several packages, each differing in the parametric distributions available: \CRANpkg{CFC} (2019), \CRANpkg{flexsurv} (2016), \CRANpkg{SmoothHazard} (2017), \CRANpkg{rstpm2} (2019), \CRANpkg{mets} (2014), and \CRANpkg{survival} (2015).
+For example, \pkg{SmoothHazard} is limited to Weibull distributions (2017), whereas both \pkg{flexsurv} and \pkg{survival} allow users to supply any distribution of their choice.
+\pkg{flexsurv}, \pkg{SmoothHazard}, \pkg{mets} and \pkg{rstpm2} can model the effect of time using splines, which allows flexible modeling of the hazard function.
+As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear combination of covariates (including time).
+Therefore, our package is more general in that it allows the user to model any linear or non-linear transformation of time including splines and higher order polynomials.
+Also, by including interaction terms between covariates and time, it also allows users to fit (non-proportional) time-varying coefficient models.
+However, unlike \pkg{flexsurv}, we do not explicitly model any shape parameter.
+
+Several packages implement penalized estimation for the Cox model: \CRANpkg{glmnet} (2011), \CRANpkg{glmpath} (2018), \CRANpkg{penalized} (2010), \CRANpkg{riskRegression} (2020).
+Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: elastic-net penalization with \pkg{rstpm2} (2019), while \pkg{survival} (2015) has an implementation of ridge-penalized estimation.
+On the other hand, our package \pkg{casebase} provides penalized estimation of the hazard function.
+To our knowledge, \pkg{casebase} and \pkg{rstpm2} are the only packages to offer this functionality.
+
+Next, several R packages implement methodologies for competing risk analysis; for a different perspective on this topic, see Mahani \& Sharabiani (2019).
+The package \pkg{survival} provides functionality for competing risk analysis and multistate modelling.
+The package \CRANpkg{cmprsk} provides methods for cause-specific subdistribution hazards, such as in the Fine-Gray model (1999).
+On the other hand, the package \pkg{CFC} estimates cause-specific CIs from unadjusted, non-parametric survival functions.
+Our package \pkg{casebase} also provides functionalities for competing risk analysis by estimating parametrically the cause-specific hazards.
+From these quantities, we can then estimate the cause-specific CIs.
+
+Finally, several packages include functions to estimate the survival function and the CI.
+The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard.
+The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled.
+Using this estimate, the survival function and the CI can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CDF} below).
+Packages providing this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and \pkg{survival}.
+Our package \pkg{casebase} also follows this approach for both single-event and competing risk analyses.
+The second category outlined above broadly corresponds to semi-parametric models.
+These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function.
+This is achieved using semi-parametric estimators (e.g.~Breslow's estimator) or parametric estimators (e.g.~spline functions).
+Packages that implement this approach include \pkg{riskRegression}, \pkg{rstpm2}, \pkg{survival}, and \pkg{glmnet}.
+As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the survival function, whereas the second category often produces estimates in the form of stepwise functions.
+
+\begin{table}
+
+\caption{\label{tab:surv-pkgs}Comparison of various R packages for survival analysis.
 \textbf{Competing Risks}: whether an implementation for competing risks is present.
@@
 \textbf{Risk estimates}: estimation of survival curve and cumulative incidence is available.}
-\label{tab:surv-pkgs}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lllllllll}
+\toprule
+\textbf{Package} & \textbf{Competing Risks} & \textbf{Allows Non PH} & \textbf{Penalized Regression} & \textbf{Splines} & \textbf{Parametric} & \textbf{Semi Parametric} & \textbf{Interval/Left Censoring} & \textbf{Risk Estimates}\\
+\midrule
+casebase & X & X & X & X & X &  &  & X\\
+CFC & X & X &  &  & X &  &  & X\\
+cmprsk & X &  &  &  &  & X &  & X\\
+crrp & X &  & X &  &  & X &  & \\
+fastcox &  &  & X &  &  & X &  & \\
+\addlinespace
+flexrsurv &  & X &  & X & X &  &  & X\\
+flexsurv & X & X &  & X & X &  &  & X\\
+glmnet &  &  & X &  &  & X &  & X\\
+glmpath &  &  & X &  &  & X &  & \\
+mets & X &  &  & X &  & X &  & X\\
+\addlinespace
+penalized &  &  & X &  &  & X &  & \\
+riskRegression & X &  & X &  &  & X &  & X\\
+rstpm2 &  & X &  & X & X & X & X & X\\
+SmoothHazard &  & X &  & X & X &  & X & \\
+survival & X & X &  &  & X & X & X & X\\
+\bottomrule
+\end{tabular}}
 \end{table}
@@
 
-The functions in the \pkg{casebase} package can be divided into two
-categories: 1) exploratory data analysis, in the form of population-time
-plots; and 2) parametric modeling of the hazard function. We strove for
-compatibility with both \code{data.frame}s and \code{data.table}s; this
-can be seen in the coding choices we made and the unit tests we wrote.
+The functions in the \pkg{casebase} package can be divided into two categories: 1) exploratory data analysis, in the form of population-time plots; and 2) parametric modeling of the hazard function.
+We strove for compatibility with both \texttt{data.frame}s and \texttt{data.table}s; this can be seen in the coding choices we made and the unit tests we wrote.
 
@@
 
-Population-time plots are a descriptive visualization of incidence
-density, where the population time that constitutes the study base is
-represented by area, and events by points within the area. The case-base
-sampling approach described above can be visualized in the form of a
-population time plot. These plots are informative graphical displays of
-survival data and should be one of the first steps in an exploratory
-data analysis. The \code{popTime} function and \code{plot} method
-facilitate this task:
+Population-time plots are a descriptive visualization of incidence density, where the population time that constitutes the study base is represented by area, and events by points within the area.
+The case-base sampling approach described above can be visualized in the form of a population time plot.
+These plots are informative graphical displays of survival data and should be one of the first steps in an exploratory data analysis.
+The \texttt{popTime} function and \texttt{plot} method facilitate this task:
 
@@
 \item
-  The \code{casebase::popTime} function takes as input the original
-  dataset along with the column names corresponding to the timescale,
-  the event status and an exposure group of interest (optional). This
-  will create an object of class \code{popTime}.\\
+  The \texttt{casebase::popTime} function takes as input the original dataset along with the column names corresponding to the timescale, the event status and an exposure group of interest (optional).
+This will create an object of class \texttt{popTime}.\\
 \item
-  The corresponding \code{plot} method for the object created in Step 1
-  can be called to create the population time plot with several options
-  for customizing the aesthetics.
+  The corresponding \texttt{plot} method for the object created in Step 1 can be called to create the population time plot with several options for customizing the aesthetics.
 \end{enumerate}
 
-By splitting these tasks, we give flexibility to the user. While the
-method call in Step 2 allows further customization by using the
-\CRANpkg{ggplot2} \citep{ggplot2} family of functions, users may choose
-the graphics system of their choice to create population-time plots from
-the object created in Step 1.
-
-To illustrate these functions, we will use data from the European
-Randomized Study of Prostate Cancer Screening (ERSPC)
-\citep{schroder2009screening} which was extracted using the approach
-described in Liu \emph{et al.} \citeyearpar{liu2014recovering}. This
-dataset is available through the \pkg{casebase} package. It contains the
-recreated individual observations for 159,893 men from seven European
-countries, who were between the ages of 55 and 69 years when recruited
-for the trial.
-
-We first create the necessary dataset for producing the population time
-plot using the \code{popTime} function. In this example, we stratify the
-plot by treatment group. The resulting object inherits from class
-\code{popTime} and stores the exposure variable as an attribute:
+By splitting these tasks, we give flexibility to the user.
+While the method call in Step 2 allows further customization by using the \CRANpkg{ggplot2} (Wickham 2016) family of functions, users may choose the graphics system of their choice to create population-time plots from the object created in Step 1.
+
+To illustrate these functions, we will use data from the European Randomized Study of Prostate Cancer Screening (ERSPC) (Schröder et al.
+2009) which was extracted using the approach described in Liu \emph{et al.} (2014).
+This dataset is available through the \pkg{casebase} package.
+It contains the recreated individual observations for 159,893 men from seven European countries, who were between the ages of 55 and 69 years when recruited for the trial.
+
+We first create the necessary dataset for producing the population time plot using the \texttt{popTime} function.
+In this example, we stratify the plot by treatment group.
+The resulting object inherits from class \texttt{popTime} and stores the exposure variable as an attribute:
 
@@
 
-\begin{Schunk}
 \begin{figure}[ht]
 \includegraphics[width=\textwidth,keepaspectratio=true]{./plot-erspc-data-1} \caption{Population time plot for the ERSPC dataset.
-\textbf{A}: The gray area can be thought of as N = 159,893 infinitely thin horizontal rectangles ordered by length of follow-up.
-\textbf{B}: The red points correspond to when death has occurred for any one of those infinitely thin rectangles.
-\textbf{C}: To improve visibility, these red points are randomly redistributed along their respective x-coordinates, providing a visualization of incidence density.
+A: The gray area can be thought of as N = 159,893 infinitely thin horizontal rectangles ordered by length of follow-up.
+B: The red points correspond to when death has occurred for any one of those infinitely thin rectangles.
+C: To improve visibility, these red points are randomly redistributed along their respective x-coordinates, providing a visualization of incidence density.
 More events are observed at later follow-up times, motivating the use of non-constant hazard models.
-\textbf{D}: The base series, a representative sample of the entire grey area, is represented by the green points.}\label{fig:plot-erspc-data}
+D: The base series, a representative sample of the entire grey area, is represented by the green points.}\label{fig:plot-erspc-data}
 \end{figure}
-\end{Schunk}
-
-We then pass this object to the corresponding \code{plot} method:
+
+We then pass this object to the corresponding \texttt{plot} method:
 
@@
 
-Figure \ref{fig:plot-erspc-data} depicts the process of creating a
-population-time plot. It is built sequentially by first adding a layer
-for the area representing the population time in gray (Figure
-\ref{fig:plot-erspc-data} A), with subjects having the least amount of
-observation time plotted at the top of the y-axis. We immediately notice
-a distinctive \emph{stepwise shape} in the population time area. This is
-due to the randomization of the Finnish cohorts which were carried out
-on January 1 of each of year from 1996 to 1999. Coupled with the uniform
-December 31 2006 censoring date, this led to large numbers of men with
-exactly 11, 10, 9 or 8 years of follow-up. Tracked backwards in time
-(i.e.~from right to left), the population-time plot shows the
-recruitment pattern from its beginning in 1991, and the January 1
-entries in successive years. Tracked forwards in time (i.e.~from left to
-right), the plot for the first three years shows attrition due entirely
-to death (mainly from other causes). Since the Swedish and Belgian
-centres were the last to complete recruitment in December 2003, the
-minimum potential follow-up is three years. Tracked further forwards in
-time (i.e.~after year 3) the attrition is a combination of deaths and
-staggered entries. As we can see, population-time plots summarise a
-wealth of information about the study into a simple graph.
-
-Next, layers for the case series and base series are added. The y-axis
-location of each case moment is sampled at random vertically on the plot
-to avoid having all points along the upper edge of the gray area (Figure
-\ref{fig:plot-erspc-data} B). By randomly distributing the cases, we can
-get a sense of the incidence density. In Figure
-\ref{fig:plot-erspc-data} C, we see that more events are observed at
-later follow-up times. Therefore, a constant hazard model would not be
-appropriate in this instance as it would overestimate the incidence
-earlier on in time, and underestimate it later on. Finally, the base
-series is sampled uniformly from the study base (Figure
-\ref{fig:plot-erspc-data} D). The reader should refer to the package
-vignettes for more examples and a detailed description of how to modify
-the aesthetics of a population-time plot.
+Figure \ref{fig:plot-erspc-data} depicts the process of creating a population-time plot.
+It is built sequentially by first adding a layer for the area representing the population time in gray (Figure \ref{fig:plot-erspc-data} A), with subjects having the least amount of observation time plotted at the top of the y-axis.
+We immediately notice a distinctive \emph{stepwise shape} in the population time area.
+This is due to the randomization of the Finnish cohorts which were carried out on January 1 of each of year from 1996 to 1999.
+Coupled with the uniform December 31 2006 censoring date, this led to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up.
+Tracked backwards in time (i.e.~from right to left), the population-time plot shows the recruitment pattern from its beginning in 1991, and the January 1 entries in successive years.
+Tracked forwards in time (i.e.~from left to right), the plot for the first three years shows attrition due entirely to death (mainly from other causes).
+Since the Swedish and Belgian centres were the last to complete recruitment in December 2003, the minimum potential follow-up is three years.
+Tracked further forwards in time (i.e.~after year 3) the attrition is a combination of deaths and staggered entries.
+As we can see, population-time plots summarise a wealth of information about the study into a simple graph.
+
+Next, layers for the case series and base series are added.
+The y-axis location of each case moment is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area (Figure \ref{fig:plot-erspc-data} B).
+By randomly distributing the cases, we can get a sense of the incidence density.
+In Figure \ref{fig:plot-erspc-data} C, we see that more events are observed at later follow-up times.
+Therefore, a constant hazard model would not be appropriate in this instance as it would overestimate the incidence earlier on in time, and underestimate it later on.
+Finally, the base series is sampled uniformly from the study base (Figure \ref{fig:plot-erspc-data} D).
+The reader should refer to the package vignettes for more examples and a detailed description of how to modify the aesthetics of a population-time plot.
 
@@
 
-By separating the sampling and estimation functions, we allow the
-possibility of users implementing more complex sampling scheme (as
-described in Saarela \citeyearpar{saarela2016case}), or more complex
-study designs (e.g.~time-varying exposure).
-
-The sampling scheme selected for \code{sampleCaseBase} was described in
-Hanley \& Miettinen \citeyearpar{hanley2009fitting}: we first sample
-along the ``person'' axis, proportional to each individual's total
-follow-up time, and then we sample a moment uniformly over their
-follow-up time. This sampling scheme is equivalent to the following
-picture: imagine representing the total follow-up time of all
-individuals in the study along a single dimension, where the follow-up
-time of the next individual would start exactly when the follow-up time
-of the previous individual ends. Then the base series could be sampled
-uniformly from this one-dimensional representation of the overall
-follow-up time. In any case, the output is a dataset of the same class
-as the input, where each row corresponds to a person-moment. The
-covariate profile for each such person-moment is retained, and an offset
-term is added to the dataset. This output could then be used to fit a
-smooth hazard function, or for visualization of the base series.
-
-Next, the fitting function \code{fitSmoothHazard} starts by looking at
-the class of the dataset: if it was generated from
-\code{sampleCaseBase}, it automatically inherited the class
-\code{cbData}. If the dataset supplied to \code{fitSmoothHazard} does
-not inherit from \code{cbData}, then the fitting function starts by
-calling \code{sampleCaseBase} to generate the base series. In other
-words, users can bypass \code{sampleCaseBase} altogether and only worry
-about the fitting function \code{fitSmoothHazard}.
-
-The fitting function retains the familiar formula interface of
-\code{glm}. The left-hand side of the formula should be the name of the
-column corresponding to the event type. The right-hand side can be any
-combination of the covariates, along with an explicit functional form
-for the time variable. Note that non-proportional hazard models can be
-achieved at this stage by adding an interaction term involving time
-(cf.~Case Study 1 below). The offset term does not need to be specified
-by the user, as it is automatically added to the formula before calling
-\code{glm}.
-
-To fit the hazard function, we provide several approaches that are
-available via the \code{family} parameter. These approaches are:
+By separating the sampling and estimation functions, we allow the possibility of users implementing more complex sampling scheme (as described in Saarela (2016)), or more complex study designs (e.g.~time-varying exposure).
+
+The sampling scheme selected for \texttt{sampleCaseBase} was described in Hanley \& Miettinen (2009): we first sample along the ``person'' axis, proportional to each individual's total follow-up time, and then we sample a moment uniformly over their follow-up time.
+This sampling scheme is equivalent to the following picture: imagine representing the total follow-up time of all individuals in the study along a single dimension, where the follow-up time of the next individual would start exactly when the follow-up time of the previous individual ends.
+Then the base series could be sampled uniformly from this one-dimensional representation of the overall follow-up time.
+In any case, the output is a dataset of the same class as the input, where each row corresponds to a person-moment.
+The covariate profile for each such person-moment is retained, and an offset term is added to the dataset.
+This output could then be used to fit a smooth hazard function, or for visualization of the base series.
+
+Next, the fitting function \texttt{fitSmoothHazard} starts by looking at the class of the dataset: if it was generated from \texttt{sampleCaseBase}, it automatically inherited the class \texttt{cbData}.
+If the dataset supplied to \texttt{fitSmoothHazard} does not inherit from \texttt{cbData}, then the fitting function starts by calling \texttt{sampleCaseBase} to generate the base series.
+In other words, users can bypass \texttt{sampleCaseBase} altogether and only worry about the fitting function \texttt{fitSmoothHazard}.
+
+The fitting function retains the familiar formula interface of \texttt{glm}.
+The left-hand side of the formula should be the name of the column corresponding to the event type.
+The right-hand side can be any combination of the covariates, along with an explicit functional form for the time variable.
+Note that non-proportional hazard models can be achieved at this stage by adding an interaction term involving time (cf.~Case Study 1 below).
+The offset term does not need to be specified by the user, as it is automatically added to the formula before calling \texttt{glm}.
+
+To fit the hazard function, we provide several approaches that are available via the \texttt{family} parameter. These approaches are:
 
@@
 \item
-  \code{glm}: This is the familiar logistic regression.
+  \texttt{glm}: This is the familiar logistic regression.
 \item
-  \code{glmnet}: This option allows for variable selection using the
-  elastic-net \citep{zou2005regularization} penalty (cf.~Case Study 3).
-  This functionality is provided through the \pkg{glmnet} package
-  \citep{friedman2010jss}.
+  \texttt{glmnet}: This option allows for variable selection using the elastic-net (Zou and Hastie 2005) penalty (cf.~Case Study 3).
+This functionality is provided through the \pkg{glmnet} package (Friedman, Hastie, and Tibshirani 2010).
 \item
-  \code{gam}: This option provides support for \emph{Generalized
-  Additive Models} via the \CRANpkg{mgcv} package
-  \citep{hastie1987generalized}.
+  \texttt{gam}: This option provides support for \emph{Generalized Additive Models} via the \CRANpkg{mgcv} package (Hastie and Tibshirani 1987).
 \end{itemize}
 
-In the case of multiple competing events, the hazard is fitted via
-multinomial regression as performed by the \CRANpkg{VGAM} package. We
-selected this package for its ability to fit multinomial regression
-models with an offset.
-
-Once a model-fit object has been returned by \code{fitSmoothHazard}, all
-the familiar summary and diagnostic functions are available:
-\code{print}, \code{summary}, \code{predict}, \code{plot}, etc. Our
-package provides one more functionality: it computes risk functions from
-the model fit. For the case of a single event, it uses the familiar
-identity \begin{equation}\label{eqn:surv}
+In the case of multiple competing events, the hazard is fitted via multinomial regression as performed by the \CRANpkg{VGAM} package.
+We selected this package for its ability to fit multinomial regression models with an offset.
+
+Once a model-fit object has been returned by \texttt{fitSmoothHazard}, all the familiar summary and diagnostic functions are available: \texttt{print}, \texttt{summary}, \texttt{predict}, \texttt{plot}, etc.
+Our package provides one more functionality: it computes risk functions from the model fit.
+For the case of a single event, it uses the familiar identity
+\begin{equation}\label{eqn:surv}
 S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right).
-\end{equation} The integral is computed using either the numerical or
-Monte-Carlo integration. The risk function (or cumulative distribution
-function) is then defined as \begin{equation}\label{eqn:CDF}
+\end{equation}
+The integral is computed using either the numerical or Monte-Carlo integration. The risk function (or cumulative distribution function) is then defined as
+\begin{equation}\label{eqn:CDF}
 F(t) = 1 - S(t).
@@
 
-For the case of a competing-event analysis, the event-specific risk is
-computed using the following procedure: first, we compute the overall
-survival function (i.e.~for all event types):
+For the case of a competing-event analysis, the event-specific risk is computed using the following procedure: first, we compute the overall survival function (i.e.~for all event types):
 
@@
 
-By integrating these subdensities, we obtain the cause-specific CI
-functions:
-
-\[ CI_j(t) = \int_0^t f_j(u)du.\] Again, the integrals are computed
-using either numerical integration (via the trapezoidal rule) or Monte
-Carlo integration. This option is controlled by the argument
-\code{method} of the \code{absoluteRisk} function.
-
-Finally, the output from \code{absoluteRisk} can be passed to a method
-\code{confint} to compute confidence bands around the survival function,
-as described in the Theoretical Details section. These bands are only
-valid when \code{family = "glm"} as it relies on the asymptotic
-normality of the estimator. Currently, this is only available for the
-single-event setting.
+By integrating these subdensities, we obtain the cause-specific CI functions:
+
+\[ CI_j(t) = \int_0^t f_j(u)du.\]
+Again, the integrals are computed using either numerical integration (via the trapezoidal rule) or Monte Carlo integration.
+This option is controlled by the argument \texttt{method} of the \texttt{absoluteRisk} function.
+
+Finally, the output from \texttt{absoluteRisk} can be passed to a method \texttt{confint} to compute confidence bands around the survival function, as described in the Theoretical Details section.
+These bands are only valid when \texttt{family\ =\ "glm"} as it relies on the asymptotic normality of the estimator.
+Currently, this is only available for the single-event setting.
 
@@
 
-In this section, we illustrate the main functions of the \pkg{casebase}
-package through three case studies. Each one showcases a different type
-of analysis. First, we show how to model non-constant and
-non-proportional hazards through a flexible specification of time. Then
-we perform a competing risk analysis and compare our results with the
-Cox model and the Fine-Gray model. The third case study illustrates how
-to perform variable selection in high-dimensional datasets.
+In this section, we illustrate the main functions of the \pkg{casebase} package through three case studies.
+Each one showcases a different type of analysis.
+First, we show how to model non-constant and non-proportional hazards through a flexible specification of time.
+Then we perform a competing risk analysis and compare our results with the Cox model and the Fine-Gray model.
+The third case study illustrates how to perform variable selection in high-dimensional datasets.
 
 \hypertarget{case-study-1flexible-modeling-of-the-hazard-function}{%
-\subsection{Case study 1---Flexible modeling of the hazard
-function}\label{case-study-1flexible-modeling-of-the-hazard-function}}
-
-For our first case study, we return to the ERSPC study and investigate
-the differences in risk between the control and screening arms. Previous
-re-analyses of these data suggest that the 20\% reduction in prostate
-cancer death due to screening was an underestimate
-\citep{hanley2010mortality}. The estimated 20\% (from a proportional
-hazards model) did not account for the delay between screening and the
-time the effect is expected to be observed. As a result, the null
-effects in years 1--7 masked the substantial reductions that began to
-appear from year 8 onward. This motivates the use of a time-dependent
-hazard ratio which can easily be fit with the \pkg{casebase} package by
-including an interaction term with time in the model. We fit a flexible
-hazard by using a smooth function of time modeled with a penalized cubic
-spline basis with 2 degrees of freedom (implemented in the
-\code{survival::pspline} function). The model is fit using
-\code{fitSmoothHazard} with the familiar formula interface:
+\subsection{Case study 1---Flexible modeling of the hazard function}\label{case-study-1flexible-modeling-of-the-hazard-function}}
+
+For our first case study, we return to the ERSPC study and investigate the
+differences in risk between the control and screening arms. Previous re-analyses
+of these data suggest that the 20\% reduction in prostate cancer death due to
+screening was an underestimate (Hanley 2010).
+The estimated 20\% (from a proportional hazards model) did not account for
+the delay between screening and the time the effect is expected to be observed.
+As a result, the null effects in years 1--7 masked the substantial reductions
+that began to appear from year 8 onward. This motivates the use of a
+time-dependent hazard ratio which can easily be fit with the \pkg{casebase}
+package by including an interaction term with time in the model. We fit a
+flexible hazard by using a smooth function of time modeled with a penalized
+cubic spline basis with 2 degrees of freedom (implemented in the
+\texttt{survival::pspline} function). The model is fit using \texttt{fitSmoothHazard}
+with the familiar formula interface:
 
@@
 
-The output object from \code{fitSmoothHazard} inherits from the
-\code{singleEventCB} and \code{glm} classes. For this reason, we can
-leverage the \code{summary} method for \code{glm} objects to output a
-familiar summary of the results:
+The output object from \texttt{fitSmoothHazard} inherits from the \texttt{singleEventCB}
+and \texttt{glm} classes. For this reason, we can leverage the \texttt{summary}
+method for \texttt{glm} objects to output a familiar summary of the results:
 
@@
 
-As noted in the Theoretical Details section, the usual asymptotic
-results hold for likelihood ratio tests built using case-base sampling
-models. Therefore, we can easily test the significance of the spline
-term and its interaction with time:
+As noted in the Theoretical Details section, the usual asymptotic results hold
+for likelihood ratio tests built using case-base sampling models. Therefore, we
+can easily test the significance of the spline term and its interaction with time:
 
@@
 
-Similarly, to compare different models (e.g.~time modeled linearly), we
-could compute Akaike's Information Criterion (AIC) for each model.
+Similarly, to compare different models (e.g.~time modeled linearly), we could
+compute Akaike's Information Criterion (AIC) for each model.
 
 \hypertarget{time-dependent-hazard-ratios}{%
-\subsubsection{Time-dependent hazard
-ratios}\label{time-dependent-hazard-ratios}}
+\subsubsection{Time-dependent hazard ratios}\label{time-dependent-hazard-ratios}}
 
@@
 \[
-\frac{\lambda\left(t | X=x_1, \mathbf{Z}=\mathbf{z} ; \hat{\beta}\right)}{\lambda(t | X=x_0, \mathbf{Z}=\mathbf{z} ; \hat{\beta})}
-\] where \(\lambda(t|\cdot;\hat{\beta})\) is the hazard rate as a
-function of the variable \(t\) (which is usually time, but can be any
-other continuous variable), \(x_1\) is the value of \(X\) for the
-exposed group, \(x_0\) is the value of \(X\) for the unexposed group,
-\(\mathbf{Z}\) are other covariates in the model which are equal to
-\(\mathbf{z}\) in both the exposed and unexposed group, and
-\(\hat{\beta}\) are the estimated regression coefficients. As indicated
-by the formula above, it is most instructive to plot the hazard ratio as
-a function of a variable \(t\) only if there is an interaction between
-\(t\) and \(X\). Otherwise, the resulting plot will simply be a
-horizontal line across time.
-
-The \code{plot} method with \code{type="hr"} for objects of class
-\code{singleEventCB} can be used to compute time-dependent hazard ratios
-and confidence intervals. In Figure \ref{fig:interaction-ERSPC}, we show
-the estimated hazard ratio and 95\% confidence interval for screening
-vs.~control group as a function of time. Note that we must specify the
-covariate profile for the reference group and times for the predicted
-hazards.
-
-\begin{Schunk}
-\begin{Sinput}
+\frac{\lambda\left(t | X=x_1, \mathbf{Z}=\mathbf{z} ; \widehat{\beta}\right)}{\lambda(t | X=x_0, \mathbf{Z}=\mathbf{z} ; \widehat{\beta})}
+\]
+where \(\lambda(t|\cdot;\widehat{\beta})\) is the hazard rate as a function of the
+variable \(t\) (which is usually time, but can be any other continuous variable),
+\(x_1\) is the value of \(X\) for the exposed group, \(x_0\) is the value of \(X\) for
+the unexposed group, \(\mathbf{Z}\) are other covariates in the model which are
+equal to \(\mathbf{z}\) in both the exposed and unexposed group,
+and \(\widehat{\beta}\) are the estimated regression coefficients. As indicated by the
+formula above, it is most instructive to plot the hazard ratio as a function of
+a variable \(t\) only if there is an interaction between \(t\) and \(X\). Otherwise,
+the resulting plot will simply be a horizontal line across time.
+
+The \texttt{plot} method with \texttt{type="hr"} for objects of class
+\texttt{singleEventCB} can be used to compute time-dependent hazard ratios and
+confidence intervals. In Figure \ref{fig:interaction-ERSPC}, we show the
+estimated hazard ratio and 95\% confidence interval for screening vs.~control
+group as a function of time. Note that we must specify the covariate profile for
+the reference group and times for the predicted hazards.
+
+\begin{verbatim}
 new_time <- seq(1, 12, by  = 0.1)
@@
 plot(fit, type = "hr", newdata = new_data,
-     var = "ScrArm", xvar = "Follow.Up.Time", ci = T)
-\end{Sinput}
-\end{Schunk}
+     var = "ScrArm", xvar = "Follow.Up.Time", ci = TRUE)
+\end{verbatim}
 
@@
 
-The plot shows that the effect of screening only becomes statistically
-apparent by year 7 and later. The 25-60\% reductions seen in years 8-12
-of the study suggests a much higher reduction in prostate cancer due to
-screening than the single overall 20\% reported in the original article.
-
-A more parsimonious model based on these results could be constructed as
-follows:
+The plot shows that the effect of screening only becomes statistically apparent by year 7 and later.
+The 25-60\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\% reported in the original article.
+
+A more parsimonious model based on these results could be constructed as follows:
 
 \[ \log \lambda\left(t \mid \mathbf{Z}\right) = \begin{cases} \alpha, &\quad t < t_0\\
-\alpha + \beta Z(t - t_0), &\quad t \geq t_0\end{cases},\] where
-\(t_0 \approx 7\) years. This model corresponds to a hazard ratio that
-is constant and equal to 1 until \(t = t_0\), after which it decreases
-exponentially. If we fix the value \(t_0\), we can easily implement this
-model using our package; for example:
+\alpha + \beta Z(t - t_0), &\quad t \geq t_0\end{cases},\]
+where \(t_0 \approx 7\) years.
+This model corresponds to a hazard ratio that is constant and equal to 1 until \(t = t_0\), after which it decreases exponentially.
+If we fix the value \(t_0\), we can easily implement this model using our package; for example:
 
@@
 
-\noindent Here, we use the binary variable
-\texttt{as.integer(Follow.Up.Time\ \textgreater{}=\ t0)} in order to
-write the two cases of our formula above in a more compact way. We also
-use the function \texttt{I()}, which allows us to specify new variables
-within the formula using normal \texttt{R} code. This is required, as
-otherwise the symbol \texttt{-} would be interpreted as a formula
-operator. Finally, the term \texttt{ScrArm\ :\ I(Follow.Up.Time\ -\ t0)}
-represents the product \(Z (t - t_0)\) in the equation above. We use the
-operator \texttt{:} instead of \texttt{*} in order to only include the
-interaction term, not the main effects.
-
-Alternatively, the breakpoint \(t_0\) could also be estimated by
-combining case-base sampling with segmented regression. However, this
-extension is beyond the current scope of \texttt{casebase}.
+\noindent Here, we use the binary variable \texttt{as.integer(Follow.Up.Time\ \textgreater{}=\ t0)} in order to write the two cases of our formula above in a more compact way.
+We also use the function \texttt{I()}, which allows us to specify new variables within the formula using normal \texttt{R} code.
+This is required, as otherwise the symbol \texttt{-} would be interpreted as a formula operator.
+Finally, the term \texttt{ScrArm\ :\ I(Follow.Up.Time\ -\ t0)} represents the product \(Z (t - t_0)\) in the equation above.
+We use the operator \texttt{:} instead of \texttt{*} in order to only include the interaction term, not the main effects.
+
+Alternatively, the breakpoint \(t_0\) could also be estimated by combining case-base sampling with segmented regression. However, this extension is beyond the current scope of \texttt{casebase}.
 
@@
 
-Modeling the hazard function directly allows us to easily visualize it
-with the \code{plot} method and \code{type="hazard"} for objects of
-class \code{singleEventCB}. We plot the hazard functions for both
-treatment arms in Figure \ref{fig:cs1hazard}. The pattern we see is
-consistent with the population-time plot shown in Figure
-\ref{fig:plot-erspc-data} C, where more events are observed at later
-follow-up times. The drop at the end can be explained by the fact that
-very few observations were followed for the entire 15 year period.
+Modeling the hazard function directly allows us to easily visualize it with
+the \texttt{plot} method and \texttt{type="hazard"} for objects of class \texttt{singleEventCB}.
+We plot the hazard functions for both treatment arms in Figure \ref{fig:cs1hazard}.
+The pattern we see is consistent with the population-time plot shown in Figure \ref{fig:plot-erspc-data} C, where more events are observed at later follow-up times.
+The drop at the end can be explained by the fact that very few observations were followed for the entire 15 year period.
 
@@
 
-\FloatBarrier
-
 \hypertarget{absolute-risk}{%
@@
 
-Next, the \code{absoluteRisk} function takes as input the
-\code{singleEventCB} object and returns a matrix where each column
-corresponds to the covariate profiles specified in the \code{newdata}
-argument, and each row corresponds to time points specified by the
-\code{time} argument:
+Next, the \texttt{absoluteRisk} function takes as input the \texttt{singleEventCB} object and returns a matrix where each column corresponds to the covariate profiles specified in the \texttt{newdata} argument, and each row corresponds to time points specified by the \texttt{time} argument:
 
@@
 
-We can subsequently compute confidence intervals for the risk function
-using the method \code{confint.absRiskCB}:
+We can subsequently compute confidence intervals for the risk function using
+the method \texttt{confint.absRiskCB}:
 
@@
 
-In Figure \ref{fig:erspc-cif-conf}, we see the 95\% confidence bands
-around the estimates. We also overlay the Kaplan-Meier curves as a
-reference.
+In Figure \ref{fig:erspc-cif-conf}, we see the 95\% confidence bands around the estimates.
+We also overlay the Kaplan-Meier curves as a reference.
 
@@
 \hypertarget{case-study-2competing-risk-analysis}{%
-\subsection{Case study 2---Competing risk
-analysis}\label{case-study-2competing-risk-analysis}}
-
-In this case study, we show how case-base sampling can be used in the
-context of a competing risk analysis. For illustrative purposes, we use
-the same data that was used in Scrucca \emph{et al}
-\citeyearpar{scrucca2010regression}. The data was downloaded from the
-first author's website, and it is now available as part of the
-\pkg{casebase} package.
-
-The data contains information on 177 patients who received a stem-cell
-transplant for acute leukemia. The event of interest is relapse, but
-other competing causes (e.g.~death, progression, graft failure,
-graft-versus-host disease) were also recorded. Several covariates were
-captured at baseline: sex, disease type (acute lymphoblastic or
-myeloblastic leukemia, abbreviated as ALL and AML, respectively),
-disease phase at transplant (Relapse, CR1, CR2, CR3), source of stem
-cells (bone marrow and peripheral blood, coded as BM+PB, or only
-peripheral blood, coded as PB), and age.
-
-First, we look at a population-time plot to visualize the incidence
-density of both relapse and the competing events. In Figure
-\ref{fig:compPop}, failure times are highlighted on the plot using red
-dots for the event of interest (panel A) and blue dots for competing
-events (panel B). In both panels, we see evidence of a non-constant
-hazard function: the density of points is larger at the beginning of
-follow-up than at the end.
+\subsection{Case study 2---Competing risk analysis}\label{case-study-2competing-risk-analysis}}
+
+In this case study, we show how case-base sampling can be used in the context of a competing risk analysis.
+For illustrative purposes, we use the same data that was used in Scrucca \emph{et al} (2010).
+The data was downloaded from the first author's website, and it is now available as part of the \pkg{casebase} package.
+
+The data contains information on 177 patients who received a stem-cell transplant for acute leukemia.
+The event of interest is relapse, but other competing causes (e.g.~death, progression, graft failure, graft-versus-host disease) were also recorded.
+Several covariates were captured at baseline: sex, disease type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL and AML, respectively), disease phase at transplant (Relapse, CR1, CR2, CR3), source of stem cells (bone marrow and peripheral blood, coded as BM+PB, or only peripheral blood, coded as PB), and age.
+
+First, we look at a population-time plot to visualize the incidence density of both relapse and the competing events.
+In Figure \ref{fig:compPop}, failure times are highlighted on the plot using red dots for the event of interest (panel A) and blue dots for competing events (panel B).
+In both panels, we see evidence of a non-constant hazard function: the density of points is larger at the beginning of follow-up than at the end.
 
@@
 
-Our main objective is to compute the cumulative incidence of relapse for
-a given set of covariates. We start by fitting a smooth hazard to the
-data using a linear term for time:
+Our main objective is to compute the cumulative incidence of relapse for a given set of covariates. We start by fitting a smooth hazard to the data using a linear term for time:
 
@@
 
-We want to compare our hazard ratio estimates to that obtained from a
-Cox regression (using the \pkg{survival} package version 3.2-13).
+We want to compare our hazard ratio estimates to that obtained from a Cox regression (using the \pkg{survival} package version 3.2-13).
 
@@
 
-\begin{Schunk}
 \begin{table}
@@
 
-\end{Schunk}
-
-From the fit object, we can extract both the hazard ratios and their
-corresponding confidence intervals. These quantities appear in Table
-\ref{tab:bmtcrr-cis}. As we can see, the type of disease corresponds to
-a significant hazard ratio: the hazard for AML is about half that for
-ALL. Moreover, being in relapse at transplant is associated with a
-hazard ratio of 4.22 when compared to CR1.
-
-Given the estimate of the hazard functions obtained using case-base
-sampling, we can compute the cumulative incidence curve for a fixed
-covariate profile. We perform this computation for a 35 year old woman
-who received a stem-cell transplant from peripheral blood at relapse. We
-compare the absolute risk curve for such a woman with ALL with that for
-a similar woman with AML.
-
-Next, we compare our estimates to that obtained from a corresponding
-Fine-Gray model \citeyearpar{fine1999proportional}. The Fine-Gray model
-is a semiparametric model for the cause-specific \emph{subdistribution
-hazard}, i.e.~the function \(d_j(t)\) such that
+From the fit object, we can extract both the hazard ratios and their corresponding confidence intervals.
+These quantities appear in Table \ref{tab:bmtcrr-cis}.
+As we can see, the type of disease corresponds to a significant hazard ratio: the hazard for AML is about half that for ALL.
+Moreover, being in relapse at transplant is associated with a hazard ratio of 4.22 when compared to CR1.
+
+Given the estimate of the hazard functions obtained using case-base sampling, we can compute the cumulative incidence curve for a fixed covariate profile.
+We perform this computation for a 35 year old woman who received a stem-cell transplant from peripheral blood at relapse.
+We compare the absolute risk curve for such a woman with ALL with that for a similar woman with AML.
+
+Next, we compare our estimates to that obtained from a corresponding Fine-Gray model (1999).
+The Fine-Gray model is a semiparametric model for the cause-specific \emph{subdistribution hazard}, i.e.~the function \(d_j(t)\) such that
 \[CI_j(t) = 1 - \exp\left( - \int_0^t d_j(u) \textrm{d}u \right),\]
-where \(CI_j(t)\) is the cause-specific CI. The Fine-Gray model allows
-to directly assess the effect of a covariate on the subdistribution
-hazard, as opposed to the cause-specific hazard. For the computation, we
-use the \CRANpkg{timereg} package \citep{timereg}:
+where \(CI_j(t)\) is the cause-specific CI.
+The Fine-Gray model allows to directly assess the effect of a covariate on the subdistribution hazard, as opposed to the cause-specific hazard.
+For the computation, we use the \CRANpkg{timereg} package (Thomas H.
+Scheike and Zhang 2011):
 
@@
 
-We can also estimate the CI for relapse using the Cox model and the
-Aalen-Johansen estimator:
+We can also estimate the CI for relapse using the Cox model and the Aalen-Johansen estimator:
 
@@
 
-\begin{Schunk}
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./bmtcrr-risk-1} \caption{\label{fig:compAbsrisk} Cumulative Incidence curve for a fixed covariate profile and the two disease groups.
-The estimate obtained from case-base sampling is compared to the Fine-Gray and Aalen-Johansen estimates.}\label{fig:bmtcrr-risk}
+\includegraphics[width=\textwidth,keepaspectratio=true]{./compAbsrisk-1} \caption{Cumulative Incidence curve for a fixed covariate profile and the two disease groups.
+The estimate obtained from case-base sampling is compared to the Fine-Gray and Aalen-Johansen estimates.}\label{fig:compAbsrisk}
 \end{figure}
-\end{Schunk}
 
 Figure \ref{fig:compAbsrisk} shows the CI curves for all three models.
-As we can see, all three approaches agree quite well for AML; however,
-for ALL, there seems to be a difference of about 5\% between the
-Fine-Gray curve and the curves estimated using case-base sampling and
-Cox regression. This difference does not appear to be significant: the
-curve from case-base sampling is contained within a 95\% confidence band
-around the Fine-Gray absolute risk curve (figure not shown).
+As we can see, all three approaches agree quite well for AML; however, for ALL, there seems to be a difference of about 5\% between the Fine-Gray curve and the curves estimated using case-base sampling and Cox regression.
+This difference does not appear to be significant: the curve from case-base sampling is contained within a 95\% confidence band around the Fine-Gray absolute risk curve (figure not shown).
 
 \hypertarget{case-study-3variable-selection}{%
-\subsection{Case study 3---Variable
-selection}\label{case-study-3variable-selection}}
-
-For the third case study, we show how \pkg{casebase} can also be used
-for variable selection through regularized estimation of the hazard
-function as given by Equation \ref{eq:penest}. We note that this is
-different than the semiparametric model Coxnet, which regularizes the
-Cox partial likelihood. To illustrate this functionality, we use the
-dataset from the Study to Understand Prognoses Preferences Outcomes and
-Risks of Treatment (SUPPORT)
-\citep{knaus1995support}.\footnote{The original data is available online from the Department of Biostatistics at Vanderbilt University: \url{https://biostat.app.vumc.org/wiki/Main/SupportDesc}}
-The SUPPORT dataset tracks death in five American hospitals within
-individuals who are considered seriously ill.~The cleaned and imputed
-data consists of 9104 observations and 30 variables, and it is available
-as part of the \pkg{casebase} package. In the comparisons below, all
-covariates except \code{sps} and \code{aps} were used. These two
-variables correspond to scores for predicting the outcome that were
-developed as part of the original study. For more information about this
-dataset, the reader is encouraged to look at the documentation in our
-package.
-
-For our penalized case-base model, we opt for the natural log of time
-which corresponds to a Weibull distribution. For fitting the penalized
-hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface
-to the \code{fitSmoothHazard} function. The \code{fitSmoothHazard} and
-\code{fitSmoothHazard.fit} functions sample the case and base series,
-calculate the required offset, and transform the data to match the
-expected input of the \pkg{glmnet} package. The penalized logistic
-regression is then fit for multiple values of the tuning parameter using
-the function \code{glmnet::cv.glmnet} and the binomial family. To
-\code{fitSmoothHazard.fit}, we supply both a matrix \code{y} containing
-the time and event variables, and a matrix \code{x} containing all other
-covariates. We apply the lasso penalty by setting \code{alpha = 1} and
-assign a \code{penalty.factor} (\(w_j\); cf.~Equation \ref{eq:penest})
-of 0 to the time variable to ensure it is in the selected model. We
-compare our approach to both Cox regression, and lasso penalized Cox
-regression (fitted via the \pkg{glmnet} package and using the Cox
-family).
-
-To compare the performance of our models, we split the data into 95\%
-training and 5\% test sets. To assess both discrimination and
-calibration, we use a time-dependent version of the classical Brier
-score that is adjusted for censoring \citep{graf1999ass}. The Brier
-score can be used with both parametric and semi-parametric models. We
-use the \pkg{riskRegression} package to compute these scores for all
-models.
-
-\begin{Schunk}
-\begin{Sinput}
+\subsection{Case study 3---Variable selection}\label{case-study-3variable-selection}}
+
+For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function as given by Equation \ref{eq:penest}.
+We note that this is different than the semiparametric model Coxnet, which regularizes the Cox partial likelihood.
+To illustrate this functionality, we use the dataset from the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) (Knaus et al.
+1995).\footnote{The original data is available online from the Department of Biostatistics at Vanderbilt University: \url{https://biostat.app.vumc.org/wiki/Main/SupportDesc}} The SUPPORT dataset tracks death in five American hospitals within individuals who are considered seriously ill.~The cleaned and imputed data consists of 9104 observations and 30 variables, and it is available as part of the \pkg{casebase} package.
+In the comparisons below, all covariates except \texttt{sps} and \texttt{aps} were used.
+These two variables correspond to scores for predicting the outcome that were developed as part of the original study.
+For more information about this dataset, the reader is encouraged to look at the documentation in our package.
+
+For our penalized case-base model, we opt for the natural log of time which corresponds to a Weibull distribution.
+For fitting the penalized hazard, we use \texttt{fitSmoothHazard.fit}, which is a matrix interface to the \texttt{fitSmoothHazard} function.
+The \texttt{fitSmoothHazard} and \texttt{fitSmoothHazard.fit} functions sample the case and base series, calculate the required offset, and transform the data to match the expected input of the \pkg{glmnet} package.
+The penalized logistic regression is then fit for multiple values of the tuning parameter using the function \texttt{glmnet::cv.glmnet} and the binomial family.
+To \texttt{fitSmoothHazard.fit}, we supply both a matrix \texttt{y} containing the time and event variables, and a matrix \texttt{x} containing all other covariates.
+We apply the lasso penalty by setting \texttt{alpha\ =\ 1} and assign a \texttt{penalty.factor} (\(w_j\); cf.~Equation \ref{eq:penest}) of 0 to the time variable to ensure it is in the selected model.
+We compare our approach to both Cox regression, and lasso penalized Cox regression (fitted via the \pkg{glmnet} package and using the Cox family).
+
+To compare the performance of our models, we split the data into 95\% training and 5\% test sets.
+To assess both discrimination and calibration, we use a time-dependent version of the classical Brier score that is adjusted for censoring (Graf et al.
+1999).
+The Brier score can be used with both parametric and semi-parametric models.
+We use the \pkg{riskRegression} package to compute these scores for all models.
+
+\begin{verbatim}
 # Create matrices for inputs
@@
 )
-\end{Sinput}
-\end{Schunk}
-
-In Figure \ref{fig:cs3lolliPlot}, we show the coefficient estimates for
-covariates that we selected by both penalized Cox and penalized
-case-base. We note that both penalized approaches produce similar
-results. We can also clearly see the shrinkage effect owing to the
-\(\ell_1\) penalty.
-
-\begin{Schunk}
+\end{verbatim}
+
+In Figure \ref{fig:cs3lolliPlot}, we show the coefficient estimates for covariates that we selected by both penalized Cox and penalized case-base.
+We note that both penalized approaches produce similar results.
+We can also clearly see the shrinkage effect owing to the \(\ell_1\) penalty.
+
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./coefplots-1} \caption{\label{fig:cs3lolliPlot} Coefficient estimates from the Cox model (Cox), penalized Cox model using the \pkg{glmnet} package (Pen.
+\includegraphics[width=\textwidth,keepaspectratio=true]{./cs3lolliPlot-1} \caption{Coefficient estimates from the Cox model (Cox), penalized Cox model using the glmnet package (Pen.
 Cox), and our approach using penalized case-base sampling (Pen.
@@
 Cox and Pen.
-CB occurs due to the $\ell_1$ penalty.}\label{fig:coefplots}
+CB occurs due to the $\ell_1$ penalty.}\label{fig:cs3lolliPlot}
 \end{figure}
-\end{Schunk}
-
-We then compare the risk estimation over the test set. The probabilities
-over time for each observation are averaged, resulting in the absolute
-risk curves shown in Figure \ref{fig:cs3FinalBrier} (A). We can see some
-minimal differences between the three models, with the Kaplan-Meier
-giving the lowest estimates across follow-up-time. Note that the
-apparent smoothness of the Cox and penalized Cox curves is due to the
-large number of observations in the training set, which is used to
-derive the Breslow estimate of the baseline hazard. As described above,
-we compare the performance between the models by computing the Brier
-scores over time. In Figure \ref{fig:cs3FinalBrier} (B), we can see that
-the Brier score is larger for the Kaplan-Meier estimate than for the
-other three models. On the other hand, the differences between these
-three models are minimal.
-
-\begin{Schunk}
+
+We then compare the risk estimation over the test set.
+The probabilities over time for each observation are averaged, resulting in the absolute risk curves shown in Figure \ref{fig:cs3FinalBrier} (A).
+We can see some minimal differences between the three models, with the Kaplan-Meier giving the lowest estimates across follow-up-time.
+Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard.
+As described above, we compare the performance between the models by computing the Brier scores over time.
+In Figure \ref{fig:cs3FinalBrier} (B), we can see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models.
+On the other hand, the differences between these three models are minimal.
+
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./riskregressionBrier-1} \caption{\label{fig:cs3FinalBrier} Comparison of Cox regression (Cox), penalized Cox regression (Pen.
+\includegraphics[width=\textwidth,keepaspectratio=true]{./cs3FinalBrier-1} \caption{Comparison of Cox regression (Cox), penalized Cox regression (Pen.
 Cox), penalized case-base sampling estimation (Pen.
@@
 (A)  Probability of death as a function of follow-up time.
-(B) Brier score as a function of follow-up time, where a lower score corresponds to better performace.}\label{fig:riskregressionBrier}
+(B) Brier score as a function of follow-up time, where a lower score corresponds to better performace.}\label{fig:cs3FinalBrier}
 \end{figure}
-\end{Schunk}
 
@@
 
-In this article, we presented the R package \pkg{casebase}, which
-provides functions for fitting smooth parametric hazards and estimating
-risk functions using case-base sampling. Our package also provides
-several functions to produce graphical summaries of the data and the
-results. We outlined the theoretical underpinnings of the approach, we
-provided details about our implementation, and we illustrated the merits
-of the case-base framework and the package through three case studies.
-
-As a methodological framework, case-base sampling is very flexible. Some
-of this flexibility has been explored before in the literature: for
-example, Saarela and Hanley \citeyearpar{saarela2015case} used case-base
-sampling to model a time-dependent exposure variable in a vaccine safety
-study. As another example, Saarela and Arjas
-\citeyearpar{saarela2015non} combined case-base sampling and a Bayesian
-non-parametric framework to compute individualized risk assessments for
-chronic diseases. In the case studies above, we further explored this
-flexibility along two fronts. On the one hand, we showed how splines
-could be used as part of the linear predictor to model the effect of
-time on the hazard. This strategy yielded estimates of the survival
-function that were qualitatively similar to semiparametric estimates
-derived from Cox regression; however, case-base sampling led to
-estimates of the survival function that \emph{vary smoothly in time} and
-are thus easier to interpret. On the other hand, we also displayed the
-flexibility of case-base sampling by showing how it could be combined
-with penalized logistic regression to perform variable selection.
-Furthermore, the second case study showed how case-base sampling can be
-applied to competing risks settings. It should be noted that the three
-case studies presented above only considered covariates that were fixed
-at baseline. In one of the package vignettes, we use the Stanford Heart
-Transplant data \citep[\citet{crowley1977covariance}]{clark1971cardiac}
-to show how case-base sampling can also be used in the context of
-time-dependent exposure. In this study, the exposure period was defined
-as the week following vaccination. Hence, the main covariate of
-interest, i.e.~exposure to the vaccine, was changing over time. In this
-context, case-base sampling offers an efficient alternative to nested
-case-control designs or self-matching.
-
-Even though we did not illustrate it in this article, case-base sampling
-can also be combined with the framework of \emph{generalized additive
-models}. This functionality has already been implemented in the package.
-Similarly, case-base sampling can be combined with quasi-likelihood
-estimation to fit survival models that can account for the presence of
-over-dispersion. All of these examples illustrate how the case-base
-sampling framework in general, and the package \pkg{casebase} in
-particular, allows the user to fit a broad and flexible family of
-survival functions.
-
-Poisson regression can also be used to estimate the full hazard by
-discretizing time. However, this method requires user input on the
-number of intervals, or equivalently, on the cut points. This choice
-made by the user can have a significant impact on the model. Small
-intervals may result in many empty, non-informative bins. This may cause
-convergence issues for the Newton-Raphson procedure
-\citep{kalbfleisch2011statistical}. If the intervals are too wide, the
-nonlinear trends that are present in the hazard may be masked. Rather
-than discretizing time like in Poisson regression, case-base sampling
-provides a continuous-time approach to using GLMs for estimating hazard
-functions.
-
-As presented in Hanley \& Miettinen \citeyearpar{hanley2009fitting},
-case-base sampling is comprised of three steps: 1) sampling a case
-series and a base series from the study; 2) fit the log-hazard as a
-linear function of predictors (including time); and 3) use the fitted
-hazard to estimate the risk function. Accordingly, our package provides
-functions for each step. Moreover, the simple interface of the
-\code{fittingSmoothHazard} function resembles the \code{glm} interface.
-This interface should look familiar to new users. Our modular approach
-also provides a convenient way to extend our package for new sampling or
-fitting strategies.
-
-In the case studies above, we compared the performance of case-base
-sampling with that of Cox regression and Fine-Gray models. In terms of
-function interface, \pkg{casebase} uses a formula interface that is
-closer to that of \code{glm}, in that the event variable is the only
-variable appearing on the left-hand side of the formula. By contrast,
-both \code{survival::coxph} and \code{timereg::comp.risk} use arrays
-that capture both the event type and time. Both approaches to modeling
-yield user-friendly code. However, in terms of output, both approaches
-differ significantly. Case-base sampling produces smooth hazards and
-smooth survival curves, whereas Cox regression and Fine-Gray models
-produce stepwise CIs and never explicitly model the hazard function.
-Qualitatively, we showed that by using splines in the linear predictor,
-all three models yielded similar curves.
-
-Our choice of modeling the log-hazard as a linear function of covariates
-allows us to develop a simple computational scheme for estimation.
-However, as a downside, it does not allow us to model location and scale
-parameters separately like the package \pkg{flexsurv}. For example, if
-we look at the Weibull distribution as parametrised in
-\code{stats::pweibull}, the log-hazard function is given by
+In this article, we presented the R package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating risk functions using case-base sampling.
+Our package also provides several functions to produce graphical summaries of the data and the results.
+We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the case-base framework and the package through three case studies.
+
+As a methodological framework, case-base sampling is very flexible.
+Some of this flexibility has been explored before in the literature: for example, Saarela and Hanley (2015) used case-base sampling to model a time-dependent exposure variable in a vaccine safety study.
+As another example, Saarela and Arjas (2015) combined case-base sampling and a Bayesian non-parametric framework to compute individualized risk assessments for chronic diseases.
+In the case studies above, we further explored this flexibility along two fronts.
+On the one hand, we showed how splines could be used as part of the linear predictor to model the effect of time on the hazard.
+This strategy yielded estimates of the survival function that were qualitatively similar to semiparametric estimates derived from Cox regression; however, case-base sampling led to estimates of the survival function that \emph{vary smoothly in time} and are thus easier to interpret.
+On the other hand, we also displayed the flexibility of case-base sampling by showing how it could be combined with penalized logistic regression to perform variable selection.
+Furthermore, the second case study showed how case-base sampling can be applied to competing risks settings.
+It should be noted that the three case studies presented above only considered covariates that were fixed at baseline.
+In one of the package vignettes, we use the Stanford Heart Transplant data Crowley and Hu (1977) to show how case-base sampling can also be used in the context of time-dependent exposure.
+In this study, the exposure period was defined as the week following vaccination.
+Hence, the main covariate of interest, i.e.~exposure to the vaccine, was changing over time.
+In this context, case-base sampling offers an efficient alternative to nested case-control designs or self-matching.
+
+Even though we did not illustrate it in this article, case-base sampling can also be combined with the framework of \emph{generalized additive models}.
+This functionality has already been implemented in the package.
+Similarly, case-base sampling can be combined with quasi-likelihood estimation to fit survival models that can account for the presence of over-dispersion.
+All of these examples illustrate how the case-base sampling framework in general, and the package \pkg{casebase} in particular, allows the user to fit a broad and flexible family of survival functions.
+
+Poisson regression can also be used to estimate the full hazard by discretizing time.
+However, this method requires user input on the number of intervals, or equivalently, on the cut points.
+This choice made by the user can have a significant impact on the model.
+Small intervals may result in many empty, non-informative bins.
+This may cause convergence issues for the Newton-Raphson procedure (Kalbfleisch and Prentice 2011).
+If the intervals are too wide, the nonlinear trends that are present in the hazard may be masked.
+Rather than discretizing time like in Poisson regression, case-base sampling provides a continuous-time approach to using GLMs for estimating hazard functions.
+
+As presented in Hanley \& Miettinen (2009), case-base sampling is comprised of three steps: 1) sampling a case series and a base series from the study; 2) fit the log-hazard as a linear function of predictors (including time); and 3) use the fitted hazard to estimate the risk function.
+Accordingly, our package provides functions for each step.
+Moreover, the simple interface of the \texttt{fitSmoothHazard} function resembles the \texttt{glm} interface.
+This interface should look familiar to new users.
+Our modular approach also provides a convenient way to extend our package for new sampling or fitting strategies.
+
+In the case studies above, we compared the performance of case-base sampling with that of Cox regression and Fine-Gray models.
+In terms of function interface, \pkg{casebase} uses a formula interface that is closer to that of \texttt{glm}, in that the event variable is the only variable appearing on the left-hand side of the formula.
+By contrast, both \texttt{survival::coxph} and \texttt{timereg::comp.risk} use arrays that capture both the event type and time.
+Both approaches to modeling yield user-friendly code.
+However, in terms of output, both approaches differ significantly.
+Case-base sampling produces smooth hazards and smooth survival curves, whereas Cox regression and Fine-Gray models produce stepwise CIs and never explicitly model the hazard function.
+Qualitatively, we showed that by using splines in the linear predictor, all three models yielded similar curves.
+
+Our choice of modeling the log-hazard as a linear function of covariates allows us to develop a simple computational scheme for estimation.
+However, as a downside, it does not allow us to model location and scale parameters separately like the package \pkg{flexsurv}.
+For example, if we look at the Weibull distribution as parametrised in \texttt{stats::pweibull}, the log-hazard function is given by
 \[ \log \lambda(t; \alpha, \beta) = \left[\log(\alpha/\beta) - (\alpha - 1)\log(\beta)\right] + (\alpha - 1)\log t,\]
 where \(\alpha,\beta\) are shape and scale parameters, respectively.
-Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} also allows
-the user to model the scale parameter as a function of covariates. Of
-course, this added flexibility comes at the cost of interpretability: by
-modeling the log-hazard directly, the parameter estimates from
-\pkg{casebase} can be interpreted as estimates of log-hazard ratios. To
-improve the flexibility of \pkg{casebase} at capturing the scale of a
-parametric family, we could replace the logistic regression with its
-quasi-likelihood counterpart and therefore model over- and
-under-dispersion with respect to the logistic likelihood. We defer the
-study of the properties and performance of such a model to a future
-article.
-
-Future work will look at some of the methodological extensions of
-case-base sampling. First, to assess the quality of the model fit, we
-would like to study the properties of the residuals (e.g.~Cox-Snell,
-martingale). More work needs to be done to understand these residuals in
-the context of the partial likelihood underlying case-base sampling. The
-resulting diagnostic tools could then be integrated in this package.
-Also, we are interested in extending case-base sampling to account for
-interval censoring. This type of censoring is very common in
-longitudinal studies, and many packages (e.g.~\pkg{SmoothHazard},
-\pkg{survival} and \pkg{rstpm2}) provide functions to account for it.
-Again, we hope to include any resulting methodology as part of this
-package.
-
-In future versions of the package, we also want to increase the
-complement of diagnostic and inferential tools that are currently
-available. For example, we would like to include more functions to
-compute calibration and discrimination statistics (e.g.~AUC) for our
-models. Saarela and Arjas \citeyearpar{saarela2015non} also describe how
-to obtain a posterior distribution for the AUC from their model. Their
-approach could potentially be included in \pkg{casebase}. Finally, we
-want to provide more flexibility in how the case-base sampling is
-performed. This could be achieved by adding a \code{hazard} argument to
-the function \code{sampleCaseBase}. In this way, users could specify
-their own sampling mechanism. For example, they could provide a hazard
-that gives sampling probabilities that are proportional to the
-cardiovascular disease event rate given by the Framingham score
-\citep{saarela2015non}.
-
-In conclusion, we presented the R package \pkg{casebase} which
-implements case-base sampling for fitting parametric survival models and
-for estimating smooth survival functions using the framework of
-generalized linear models. We strongly believe that its flexibility and
-its foundation on the familiar logistic regression model will make it
-appealing to new and established practitioners. The \pkg{casebase}
-package is freely available from the Comprehensive R Archive Network at
-\url{https://cran.r-project.org/package=casebase}. Interested users can
-visit \url{http://sahirbhatnagar.com/casebase/} for detailed package
-documentation and vignettes.
+Unlike \pkg{casebase}, the approach taken by \pkg{flexsurv} also allows the user to model the scale parameter as a function of covariates.
+Of course, this added flexibility comes at the cost of interpretability: by modeling the log-hazard directly, the parameter estimates from \pkg{casebase} can be interpreted as estimates of log-hazard ratios.
+To improve the flexibility of \pkg{casebase} at capturing the scale of a parametric family, we could replace the logistic regression with its quasi-likelihood counterpart and therefore model over- and under-dispersion with respect to the logistic likelihood.
+We defer the study of the properties and performance of such a model to a future article.
+
+Future work will look at some of the methodological extensions of case-base sampling.
+First, to assess the quality of the model fit, we would like to study the properties of the residuals (e.g.~Cox-Snell, martingale).
+More work needs to be done to understand these residuals in the context of the partial likelihood underlying case-base sampling.
+The resulting diagnostic tools could then be integrated in this package.
+Also, we are interested in extending case-base sampling to account for interval censoring.
+This type of censoring is very common in longitudinal studies, and many packages (e.g.~\pkg{SmoothHazard}, \pkg{survival} and \pkg{rstpm2}) provide functions to account for it.
+Again, we hope to include any resulting methodology as part of this package.
+
+In future versions of the package, we also want to increase the complement of diagnostic and inferential tools that are currently available.
+For example, we would like to include more functions to compute calibration and discrimination statistics (e.g.~AUC) for our models.
+Saarela and Arjas (2015) also describe how to obtain a posterior distribution for the AUC from their model.
+Their approach could potentially be included in \pkg{casebase}.
+Finally, we want to provide more flexibility in how the case-base sampling is performed.
+This could be achieved by adding a \texttt{hazard} argument to the function \texttt{sampleCaseBase}.
+In this way, users could specify their own sampling mechanism.
+For example, they could provide a hazard that gives sampling probabilities that are proportional to the cardiovascular disease event rate given by the Framingham score (Saarela and Arjas 2015).
+
+In conclusion, we presented the R package \pkg{casebase} which implements case-base sampling for fitting parametric survival models and for estimating smooth survival functions using the framework of generalized linear models.
+We strongly believe that its flexibility and its foundation on the familiar logistic regression model will make it appealing to new and established practitioners.
+The \pkg{casebase} package is freely available from the Comprehensive R Archive Network at \url{https://cran.r-project.org/package=casebase}.
+Interested users can visit \url{http://sahirbhatnagar.com/casebase/} for detailed package documentation and vignettes.
 
@@
 
-We would like to thank the anonymous reviewers for their insightful
-comments and criticisms. We would also like to thank Yi Yang for helpful
-discussions on penalized regression models. \mbox{Bhatnagar}
-(RGPIN-2020-05133) and \mbox{Turgeon} (RGPIN-2021-04073) both gratefully
-acknowledge funding via a Discovery Grant from the Natural Sciences and
-Engineering Research Council of Canada (NSERC), RGPIN.
+We would like to thank the anonymous reviewers for their insightful comments and criticisms.
+We would also like to thank Yi Yang for helpful discussions on penalized regression models.
+Bhatnagar (RGPIN-2020-05133) and Turgeon (RGPIN-2021-04073) both gratefully acknowledge funding via a Discovery Grant from the Natural Sciences and Engineering Research Council of Canada (NSERC), RGPIN.
+
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-aalen2008survival}{}}%
+Aalen, Odd, Ornulf Borgan, and Hakon Gjessing. 2008. \emph{Survival and Event History Analysis: A Process Point of View}. Springer Science \& Business Media.
+
+\leavevmode\vadjust pre{\hypertarget{ref-survTaskView}{}}%
+Allignol, Arthur, and Aurelien Latouche. 2019. {``CRAN Task View: Survival Analysis.''} \emph{CRAN Task View: Survival Analysis}. \url{https://cran.r-project.org/web/views/Survival.html}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-arjas1987logistic}{}}%
+Arjas, Elja, and Pentti Haara. 1987. {``A Logistic Regression Model for Hazard: Asymptotic Results.''} \emph{Scandinavian Journal of Statistics}, 1--18.
+
+\leavevmode\vadjust pre{\hypertarget{ref-casebase-package}{}}%
+Bhatnagar, Sahir, Maxime Turgeon, Jesse Islam, Olli Saarela, and James Hanley.
+2021.
+\emph{Casebase: Fitting Flexible Smooth-in-Time Hazards and Risk Functions via Logistic and Multinomial Regression}.
+\url{https://CRAN.R-project.org/package=casebase}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-breslow1972discussion}{}}%
+Breslow, NE. 1972. {``Discussion of the Paper by {Dr Cox} Cited Below.''} \emph{Journal of the Royal Statistical Society: Series B (Methodological)} 34: 187--220.
+
+\leavevmode\vadjust pre{\hypertarget{ref-copenhagen2012needs}{}}%
+Carstensen, Bendix. 2019. {``Who Needs the Cox Model Anyway.''} \url{http://bendixcarstensen.com/WntCma.pdf}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-clark1971cardiac}{}}%
+Clark, David A, Edward B Stinson, Randall B Griepp, John S Schroeder, Norman E Shumway, and Donald Harrison.
+1971.
+{``Cardiac Transplantation in Man.''} \emph{Annals of Internal Medicine} 75 (1): 15--21.
+
+\leavevmode\vadjust pre{\hypertarget{ref-clements_liu}{}}%
+Clements, Mark, and Xing-Rong Liu. 2019. \emph{Rstpm2: Smooth Survival Models, Including Generalized Survival Models}. \url{https://CRAN.R-project.org/package=rstpm2}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-crowley1977covariance}{}}%
+Crowley, John, and Marie Hu. 1977. {``Covariance Analysis of Heart Transplant Survival Data.''} \emph{Journal of the American Statistical Association} 72 (357): 27--36.
+
+\leavevmode\vadjust pre{\hypertarget{ref-efron1988logistic}{}}%
+Efron, Bradley. 1988. {``Logistic Regression, Survival Analysis, and the Kaplan-Meier Curve.''} \emph{Journal of the American Statistical Association} 83 (402): 414--25.
+
+\leavevmode\vadjust pre{\hypertarget{ref-efron1994introduction}{}}%
+Efron, Bradley, and Robert J Tibshirani. 1994. \emph{An Introduction to the Bootstrap}. CRC press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-fine1999proportional}{}}%
+Fine, Jason P, and Robert J Gray. 1999. {``A Proportional Hazards Model for the Subdistribution of a Competing Risk.''} \emph{Journal of the American Statistical Association} 94 (446): 496--509.
+
+\leavevmode\vadjust pre{\hypertarget{ref-friedman2010jss}{}}%
+Friedman, Jerome, Trevor Hastie, and Robert Tibshirani.
+2010.
+{``Regularization Paths for Generalized Linear Models via Coordinate Descent.''} \emph{Journal of Statistical Software} 33 (1).
+\url{https://doi.org/10.18637/jss.v033.i01}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gerds_blanche}{}}%
+Gerds, Thomas Alexander, and Brice Ozenne.
+2020.
+\emph{riskRegression: Risk Regression Models and Prediction Scores for Survival Analysis with Competing Risks}.
+\url{https://CRAN.R-project.org/package=riskRegression}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-l1penal}{}}%
+Goeman, J. J. 2010. {``L1 Penalized Estimation in the Cox Proportional Hazards Model.''} \emph{Biometrical Journal}, no. 52: --14.
+
+\leavevmode\vadjust pre{\hypertarget{ref-graf1999ass}{}}%
+Graf, Erika, Claudia Schmoor, Willi Sauerbrei, and Martin Schumacher.
+1999.
+{``Assessment and Comparison of Prognostic Classification Schemes for Survival Data.''} \emph{Statistics in Medicine} 18 (17-18): 2529--45.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hanley2010mortality}{}}%
+Hanley, James A. 2010. {``Mortality Reductions Produced by Sustained Prostate Cancer Screening Have Been Underestimated.''} \emph{Journal of Medical Screening} 17 (3): 147--51.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hanley2009fitting}{}}%
+Hanley, James A, and Olli S Miettinen. 2009. {``Fitting {smooth-in-time} Prognostic Risk Functions via Logistic Regression.''} \emph{The International Journal of Biostatistics} 5 (1).
+
+\leavevmode\vadjust pre{\hypertarget{ref-hastie1987generalized}{}}%
+Hastie, Trevor, and Robert Tibshirani. 1987. {``Generalized Additive Models: Some Applications.''} \emph{Journal of the American Statistical Association} 82 (398): 371--86.
+
+\leavevmode\vadjust pre{\hypertarget{ref-flexsurv}{}}%
+Jackson, Christopher. 2016. {``{flexsurv}: A Platform for Parametric Survival Modeling in {R}.''} \emph{Journal of Statistical Software} 70 (8): 1--33. \url{https://doi.org/10.18637/jss.v070.i08}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kalbfleisch2011statistical}{}}%
+Kalbfleisch, John D, and Ross L Prentice. 2011. \emph{The Statistical Analysis of Failure Time Data}. Vol. 360. John Wiley \& Sons.
+
+\leavevmode\vadjust pre{\hypertarget{ref-knaus1995support}{}}%
+Knaus, William A, Frank E Harrell, Joanne Lynn, Lee Goldman, Russell S Phillips, Alfred F Connors, Neal V Dawson, et al.
+1995.
+{``The SUPPORT Prognostic Model: Objective Estimates of Survival for Seriously Ill Hospitalized Adults.''} \emph{Annals of Internal Medicine} 122 (3): 191--203.
+
+\leavevmode\vadjust pre{\hypertarget{ref-liu2014recovering}{}}%
+Liu, Zhihui, Benjamin Rich, and James A Hanley. 2014. {``Recovering the Raw Data Behind a Non-Parametric Survival Curve.''} \emph{Systematic Reviews} 3 (1): 151.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mahani2015bayesian}{}}%
+Mahani, Alireza, and Mansour Sharabiani.
+2019.
+{``Bayesian, and Non-{B}ayesian, Cause-Specific Competing-Risk Analysis for Parametric and Nonparametric Survival Functions: The {R} Package {CFC}.''} \emph{Journal of Statistical Software} 89 (9): 1--29.
+\url{https://doi.org/10.18637/jss.v089.i09}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-park_hastie}{}}%
+Park, Mee Young, and Trevor Hastie. 2018. \emph{Glmpath: L1 Regularization Path for Generalized Linear Models and Cox Proportional Hazards Model}. \url{https://CRAN.R-project.org/package=glmpath}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-reid1994conversation}{}}%
+Reid, Nancy. 1994. {``A Conversation with Sir David Cox.''} \emph{Statistical Science} 9 (3): 439--55.
+
+\leavevmode\vadjust pre{\hypertarget{ref-saarela2016case}{}}%
+Saarela, Olli. 2016. {``A Case-Base Sampling Method for Estimating Recurrent Event Intensities.''} \emph{Lifetime Data Analysis} 22 (4): 589--605.
+
+\leavevmode\vadjust pre{\hypertarget{ref-saarela2015non}{}}%
+Saarela, Olli, and Elja Arjas. 2015. {``Non-Parametric {B}ayesian Hazard Regression for Chronic Disease Risk Assessment.''} \emph{Scandinavian Journal of Statistics} 42 (2): 609--26.
+
+\leavevmode\vadjust pre{\hypertarget{ref-saarela2015case}{}}%
+Saarela, Olli, and James A Hanley. 2015. {``Case-Base Methods for Studying Vaccination Safety.''} \emph{Biometrics} 71 (1): 42--52.
+
+\leavevmode\vadjust pre{\hypertarget{ref-scheike2014estimating}{}}%
+Scheike, Thomas H, Klaus K Holst, and Jacob B Hjelmborg. 2014. {``Estimating Twin Concordance for Bivariate Competing Risks Twin Data.''} \emph{Statistics in Medicine} 33 (7): 1193--1204.
+
+\leavevmode\vadjust pre{\hypertarget{ref-timereg}{}}%
+Scheike, Thomas H., and Mei-Jie Zhang. 2011. {``Analyzing Competing Risk Data Using the {R} {timereg} Package.''} \emph{Journal of Statistical Software} 38 (2): 1--15.
+
+\leavevmode\vadjust pre{\hypertarget{ref-schroder2009screening}{}}%
+Schröder, Fritz H, Jonas Hugosson, Monique J Roobol, Teuvo LJ Tammela, Stefano Ciatto, Vera Nelen, Maciej Kwiatkowski, et al.
+2009.
+{``Screening and Prostate-Cancer Mortality in a Randomized {E}uropean Study.''} \emph{New England Journal of Medicine} 360 (13): 1320--28.
+
+\leavevmode\vadjust pre{\hypertarget{ref-scrucca2010regression}{}}%
+Scrucca, L, A Santucci, and F Aversa. 2010. {``Regression Modeling of Competing Risk Using {R}: An in Depth Guide for Clinicians.''} \emph{Bone Marrow Transplantation} 45 (9): 1388.
+
+\leavevmode\vadjust pre{\hypertarget{ref-regpathcox}{}}%
+Simon, Noah, Jerome Friedman, Trevor Hastie, and Rob Tibshirani.
+2011.
+{``Regularization Paths for {C}ox's Proportional Hazards Model via Coordinate Descent.''} \emph{Journal of Statistical Software} 39 (5): 1--13.
+
+\leavevmode\vadjust pre{\hypertarget{ref-survival-package}{}}%
+Therneau, Terry M. 2015. \emph{A Package for Survival Analysis in s}. \url{https://CRAN.R-project.org/package=survival}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-smoothHazard}{}}%
+Touraine, Célia, Thomas A.
+Gerds, and Pierre Joly.
+2017.
+{``{SmoothHazard}: An {R} Package for Fitting Regression Models to Interval-Censored Observations of Illness-Death Models.''} \emph{Journal of Statistical Software} 79 (7): 1--22.
+\url{https://doi.org/10.18637/jss.v079.i07}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-van2000asymptotic}{}}%
+Van der Vaart, Aad W. 2000. \emph{Asymptotic Statistics}. Vol. 3. Cambridge university press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-whitehead1980fitting}{}}%
+Whitehead, John. 1980. {``Fitting Cox's Regression Model to Survival Data Using GLIM.''} \emph{Journal of the Royal Statistical Society: Series C (Applied Statistics)} 29 (3): 268--75.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ggplot2}{}}%
+Wickham, Hadley. 2016. \emph{Ggplot2: Elegant Graphics for Data Analysis}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-zou2005regularization}{}}%
+Zou, Hui, and Trevor Hastie. 2005. {``Regularization and Variable Selection via the Elastic Net.''} \emph{Journal of the Royal Statistical Society: Series B (Methodological)} 67 (2): 301--20.
+
+\end{CSLReferences}
 
@@
 University of Toronto\\%
-Dalla Lana School of Public Health, 155 College Street, 6th floor,
-Toronto, Ontario M5T 3M7, Canada\\
+Dalla Lana School of Public Health, 155 College Street, 6th floor, Toronto, Ontario M5T 3M7, Canada\\
 %

### 2022-11-19 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd

@@
 \begin{equation}
-L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{\,\textrm dQ_i(t)}. \label{eq:lik-function}
+L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{\,\textrm dQ_i(t)}. (\#eq:lik-function)
 \end{equation}
@@
   2. Sample $B$ times from a multivariate normal $N\left(\widehat{\theta}, V(\widehat\theta)\right)$.
-  3. For each sample, compute the survival function using Equation \ref{eqn:surv} below.
+  3. For each sample, compute the survival function using Equation \@ref(eq:surv) below.
   4. Use the pointwise quantiles of these survival function estimates to construct a pointwise confidence band for the survival function of interest.
@@
 \begin{equation}
-\log(\lambda(t; X)) = \beta_0 + \beta_1 X. \label{eq:exp}
+\log(\lambda(t; X)) = \beta_0 + \beta_1 X. (\#eq:exp)
 \end{equation}
@@
 \begin{equation}
-\log(\lambda(t; X)) = \beta_0 + \beta_1 t + \beta_2 X. \label{eq:gomp}
+\log(\lambda(t; X)) = \beta_0 + \beta_1 t + \beta_2 X. (\#eq:gomp)
 \end{equation}
@@
 \begin{equation}
-\log(\lambda(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. \label{eq:weibull}
+\log(\lambda(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. (\#eq:weibull)
 \end{equation}
@@
 \begin{equation}
-\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j p_{\lambda,\alpha}(\theta_j) \label{eq:penest}
+\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j p_{\lambda,\alpha}(\theta_j) (\#eq:penest)
 \end{equation}
-where $\ell\left(\theta\right) = \log L(\theta)$ is the log of the likelihood function given in \eqref{eq:lik-function}, $p_{\lambda,\alpha}(\theta_j)$ is a penalty term controlled by the non-negative regularization parameters $\lambda$ and $\alpha$, and $w_j$ is the penalty factor for the $j$th covariate.
+where $\ell\left(\theta\right) = \log L(\theta)$ is the log of the likelihood function given in \@ref(eq:lik-function), $p_{\lambda,\alpha}(\theta_j)$ is a penalty term controlled by the non-negative regularization parameters $\lambda$ and $\alpha$, and $w_j$ is the penalty factor for the $j$th covariate.
 These penalty factors serve as a way of allowing parameters to be penalized differently.
@@
 The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled.
-Using this estimate, the survival function and the CI can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CDF} below).
+Using this estimate, the survival function and the CI can be obtained using their functional relationships (see Equations \@ref(eq:surv) and \@ref(eq:CDF) below).
 Packages providing this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and \pkg{survival}.
@@
 For the case of a single event, it uses the familiar identity
-\begin{equation}\label{eqn:surv}
+\begin{equation}
 S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right). 
+(\#eq:surv)
 \end{equation}
 The integral is computed using either the numerical or Monte-Carlo integration. The risk function (or cumulative distribution function) is then defined as
-\begin{equation}\label{eqn:CDF}
-F(t) = 1 - S(t).
+\begin{equation}
+F(t) = 1 - S(t). (\#eq:CDF)
 \end{equation}
@@
 
-axis(2, at = seq(0,1.50, by=0.25), las = 2)
-axis(1, at = seq(min(new_time),max(new_time), by=1))
-for( i in seq(0,1.50, by=0.25)) {
+axis(2, at = seq(0,1.50, by = 0.25), las = 2)
+axis(1, at = seq(min(new_time),max(new_time), by = 1))
+for (i in seq(0,1.50, by = 0.25)) {
   abline(h = i, lty = 1, col = "lightgrey")
 }
-for( i in seq(min(new_time),max(new_time), by=1)) {
+for (i in seq(min(new_time),max(new_time), by = 1)) {
   abline(v = i, lty = 1, col = "lightgrey")
 }
-for(i in seq_along(seq(0, 0.75, by = 0.25))) {
-  mtext(paste0(seq(0, 0.75, by = 0.25)[i]*100,"%"), side = 4, las = 2, adj = 1, outer = TRUE, line = 2, at = c(0.63-0.11*(i-1)))
+for (i in seq_along(seq(0, 0.75, by = 0.25))) {
+  mtext(paste0(seq(0, 0.75, by = 0.25)[i]*100,"%"), side = 4, las = 2, adj = 1, outer = TRUE, line = 2, at = c(0.63 - 0.11*(i - 1)))
 }
@@
 ```
-
-\noindent Here, we use the binary variable `as.integer(Follow.Up.Time >= t0)` in order to write the two cases of our formula above in a more compact way.
+Here, we use the binary variable `as.integer(Follow.Up.Time >= t0)` in order to write the two cases of our formula above in a more compact way.
 We also use the function `I()`, which allows us to specify new variables within the formula using normal `R` code.
@@
 
-For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function as given by Equation \ref{eq:penest}.
+For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function as given by Equation \@ref(eq:penest).
 We note that this is different than the semiparametric model Coxnet, which regularizes the Cox partial likelihood.
@@
 We apply the lasso penalty by setting `alpha = 1` and assign a `penalty.factor` ($w_j$; cf.
-Equation \ref{eq:penest}) of 0 to the time variable to ensure it is in the selected model.
+Equation \@ref(eq:penest)) of 0 to the time variable to ensure it is in the selected model.
 We compare our approach to both Cox regression, and lasso penalized Cox regression (fitted via the \pkg{glmnet} package and using the Cox family).

### 2022-11-19 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex

@@
 \item
-  For each sample, compute the survival function using Equation \ref{eqn:surv} below.
+  For each sample, compute the survival function using Equation \eqref{eq:surv} below.
 \item
@@
 The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled.
-Using this estimate, the survival function and the CI can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CDF} below).
+Using this estimate, the survival function and the CI can be obtained using their functional relationships (see Equations \eqref{eq:surv} and \eqref{eq:CDF} below).
 Packages providing this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and \pkg{survival}.
@@
 
-\begin{Schunk}
-\begin{Sinput}
+\begin{verbatim}
 pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
@@
 inherits(pt_object, "popTime")
-\end{Sinput}
-\begin{Soutput}
 #> [1] TRUE
-\end{Soutput}
-\begin{Sinput}
 attr(pt_object, "exposure")
-\end{Sinput}
-\begin{Soutput}
 #> [1] "ScrArm"
-\end{Soutput}
-\end{Schunk}
+\end{verbatim}
 
@@
 
-\begin{Schunk}
-\begin{Sinput}
+\begin{verbatim}
 plot(pt_object, add.base.series = TRUE)
-\end{Sinput}
-\end{Schunk}
+\end{verbatim}
 
@@
 For the case of a single event, it uses the familiar identity
-\begin{equation}\label{eqn:surv}
+\begin{equation}
 S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right). 
+\label{eq:surv}
 \end{equation}
 The integral is computed using either the numerical or Monte-Carlo integration. The risk function (or cumulative distribution function) is then defined as
-\begin{equation}\label{eqn:CDF}
-F(t) = 1 - S(t).
+\begin{equation}
+F(t) = 1 - S(t). \label{eq:CDF}
 \end{equation}
@@
 
-\begin{Schunk}
-\begin{Sinput}
+\begin{verbatim}
 fit <- fitSmoothHazard(DeadOfPrCa ~ pspline(Follow.Up.Time, df = 2) * ScrArm, 
                        data = ERSPC, ratio = 10)
-\end{Sinput}
-\end{Schunk}
+\end{verbatim}
 
@@
 
-\begin{Schunk}
-\begin{Sinput}
+\begin{verbatim}
 summary(fit) 
-\end{Sinput}
-\begin{Soutput}
 #> Fitting smooth hazards with case-base sampling
@@
 #> Number of Fisher Scoring iterations: 7
-\end{Soutput}
-\end{Schunk}
+\end{verbatim}
 
@@
 
-\begin{Schunk}
-\begin{Sinput}
+\begin{verbatim}
 anova(fit, test = "LRT")
-\end{Sinput}
-\begin{Soutput}
 #> Analysis of Deviance Table
@@
 #> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{Soutput}
-\end{Schunk}
+\end{verbatim}
 
@@
 
-\begin{Schunk}
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./interaction-ERSPC-1} \caption[Estimated hazard ratio and 95\% confidence interval for screening vs]{Estimated hazard ratio and 95\% confidence interval for screening vs.
+\includegraphics[width=\textwidth,keepaspectratio=true]{./interaction-ERSPC-1} \caption{Estimated hazard ratio and 95\% confidence interval for screening vs.
 control group as a function of time in the ERSPC dataset.
@@
 \end{figure}
-\end{Schunk}
 
@@
 
-\begin{Schunk}
-\begin{Sinput}
+\begin{verbatim}
 fitSmoothHazard(DeadOfPrCa ~ as.integer(Follow.Up.Time >= t0) :
@@
                 data = ERSPC)
-\end{Sinput}
-\end{Schunk}
-
-\noindent Here, we use the binary variable \texttt{as.integer(Follow.Up.Time\ \textgreater{}=\ t0)} in order to write the two cases of our formula above in a more compact way.
+\end{verbatim}
+
+Here, we use the binary variable \texttt{as.integer(Follow.Up.Time\ \textgreater{}=\ t0)} in order to write the two cases of our formula above in a more compact way.
 We also use the function \texttt{I()}, which allows us to specify new variables within the formula using normal \texttt{R} code.
@@
 
-\begin{Schunk}
-\begin{Sinput}
+\begin{verbatim}
 plot(fit, type = "hazard",
@@
                           by = "ScrArm"))
-\end{Sinput}
-\begin{Soutput}
 #> Conditions used in construction of plot
@@
 #> offset: 0
-\end{Soutput}
+\end{verbatim}
+
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./cs1hazard-1} \caption[Estimated hazard functions for control and screening groups in the ERSPC dataset]{Estimated hazard functions for control and screening groups in the ERSPC dataset.
+\includegraphics[width=\textwidth,keepaspectratio=true]{./cs1hazard-1} \caption{Estimated hazard functions for control and screening groups in the ERSPC dataset.
 Hazards are estimated from fitting a parametric model with casebase sampling as a function of the interaction between a cubic pspline basis (df=2) of follow-up time and treatment arm.
@@
 
-\begin{Schunk}
-\begin{Sinput}
+\begin{verbatim}
 new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
@@
 risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
-\end{Sinput}
-\end{Schunk}
+\end{verbatim}
 
@@
 
-\begin{Schunk}
-\begin{Sinput}
+\begin{verbatim}
 conf_ints <- confint(risk, fit)
 head(conf_ints)
-\end{Sinput}
-\begin{Soutput}
 #>   time estimate conf.low conf.high      cov_prof
@@
 #> 6  0.5  1.2e-05  2.3e-06   5.0e-05 Control group
-\end{Soutput}
-\end{Schunk}
+\end{verbatim}
 
@@
 
-\begin{Schunk}
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cif-conf-1} \caption[Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands]{Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands.}\label{fig:erspc-cif-conf}
+\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cif-conf-1} \caption{Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands.}\label{fig:erspc-cif-conf}
 \end{figure}
-\end{Schunk}
 
@@
 
-\begin{Schunk}
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./compPop-1} \caption[Population-time plot for the stem-cell transplant study with both relapse and competing events]{Population-time plot for the stem-cell transplant study with both relapse and competing events.}\label{fig:compPop}
+\includegraphics[width=\textwidth,keepaspectratio=true]{./compPop-1} \caption{Population-time plot for the stem-cell transplant study with both relapse and competing events.}\label{fig:compPop}
 \end{figure}
-\end{Schunk}
 
@@
 
-\begin{Schunk}
-\begin{Sinput}
+\begin{verbatim}
 model_cb <- fitSmoothHazard(
@@
 )
-\end{Sinput}
-\end{Schunk}
+\end{verbatim}
 
@@
 
-\begin{Schunk}
-\begin{Sinput}
+\begin{verbatim}
 library(survival)
@@
                    data = bmtcrr_cox, id = id)
-\end{Sinput}
-\end{Schunk}
+\end{verbatim}
 
@@
 
-\begin{Schunk}
-\begin{Sinput}
+\begin{verbatim}
 library(timereg)
@@
 risk_fg <- predict(model_fg, newdata, times = time_points)
-\end{Sinput}
-\end{Schunk}
+\end{verbatim}
 
@@
 
-\begin{Schunk}
-\begin{Sinput}
+\begin{verbatim}
 # Estimate absolute risk curve
 risk_cox <- survfit(model_cox, newdata = newdata)
-\end{Sinput}
-\end{Schunk}
+\end{verbatim}
 
@@
 
-For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function as given by Equation \ref{eq:penest}.
+For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function as given by Equation \eqref{eq:penest}.
 We note that this is different than the semiparametric model Coxnet, which regularizes the Cox partial likelihood.
@@
 To \texttt{fitSmoothHazard.fit}, we supply both a matrix \texttt{y} containing the time and event variables, and a matrix \texttt{x} containing all other covariates.
-We apply the lasso penalty by setting \texttt{alpha\ =\ 1} and assign a \texttt{penalty.factor} (\(w_j\); cf.~Equation \ref{eq:penest}) of 0 to the time variable to ensure it is in the selected model.
+We apply the lasso penalty by setting \texttt{alpha\ =\ 1} and assign a \texttt{penalty.factor} (\(w_j\); cf.~Equation \eqref{eq:penest}) of 0 to the time variable to ensure it is in the selected model.
 We compare our approach to both Cox regression, and lasso penalized Cox regression (fitted via the \pkg{glmnet} package and using the Cox family).

### 2022-11-21 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
cleaned up captions

@@
 Figure \@ref(fig:plot-erspc-data) depicts the process of creating a population-time plot.
-It is built sequentially by first adding a layer for the area representing the population time in gray (Figure \@ref(fig:plot-erspc-data) A), with subjects having the least amount of observation time plotted at the top of the y-axis.
+It is built sequentially by first adding a layer for the area representing the population time in gray (Figure \@ref(fig:plot-erspc-data)A), with subjects having the least amount of observation time plotted at the top of the y-axis.
 We immediately notice a distinctive *stepwise shape* in the population time area.
@@
 Next, layers for the case series and base series are added.
-The y-axis location of each case moment is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area (Figure \@ref(fig:plot-erspc-data) B).
+The y-axis location of each case moment is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area (Figure \@ref(fig:plot-erspc-data)B).
 By randomly distributing the cases, we can get a sense of the incidence density.
-In Figure \@ref(fig:plot-erspc-data) C, we see that more events are observed at later follow-up times.
+In Figure \@ref(fig:plot-erspc-data)C, we see that more events are observed at later follow-up times.
 Therefore, a constant hazard model would not be appropriate in this instance as it would overestimate the incidence earlier on in time, and underestimate it later on.
-Finally, the base series is sampled uniformly from the study base (Figure \@ref(fig:plot-erspc-data) D).
+Finally, the base series is sampled uniformly from the study base (Figure \@ref(fig:plot-erspc-data)D).
 The reader should refer to the package vignettes for more examples and a detailed description of how to modify the aesthetics of a population-time plot.
@@
 
-## Case study 1---Flexible modeling of the hazard function
+## Case study 1---flexible modeling of the hazard function
 
@@
 We plot the hazard functions for both treatment arms in Figure \@ref(fig:cs1hazard). 
-The pattern we see is consistent with the population-time plot shown in Figure \@ref(fig:plot-erspc-data) C, where more events are observed at later follow-up times. 
+The pattern we see is consistent with the population-time plot shown in Figure \@ref(fig:plot-erspc-data)C, where more events are observed at later follow-up times. 
 The drop at the end can be explained by the fact that very few observations were followed for the entire 15 year period.
@@
 
-```{r erspc-cif-conf, eval = TRUE, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\\% confidence bands."}
+```{r erspc-cif-conf, eval = TRUE, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\\% confidence bands.
+The smooth curve (case-base sampling) vs.
+step function (Cox model) highlight one of the main differences between the two approaches.
+The larger confidence bands in the later years is due to the relatively few number of individuals who were followed for more than 12 years."}
 
@@
 
-## Case study 2---Competing risk analysis
+## Case study 2---competing risk analysis
 
@@
 First, we look at a population-time plot to visualize the incidence density of both relapse and the competing events.
-In Figure \@ref(fig:compPop), failure times are highlighted on the plot using red dots for the event of interest (panel A) and blue dots for competing events (panel B).
-In both panels, we see evidence of a non-constant hazard function: the density of points is larger at the beginning of follow-up than at the end.
-
-```{r compPop, fig.cap="Population-time plot for the stem-cell transplant study with both relapse and competing events.", echo=FALSE, eval = TRUE}
+In Figure \@ref(fig:compPop), failure times are highlighted on the plot using red dots for the event of interest and blue dots for competing events.
+In this plot, we see evidence of a non-constant hazard function: the density of points is larger at the beginning of follow-up than at the end.
+
+```{r compPop, fig.cap="Population-time plot for the stem-cell transplant study with both relapse and competing events.
+The area representing the population time is shown in gray, with subjects having the least amount of observation time plotted at the top of the y-axis.
+The y-axis location of each case series and competing event moment is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area.
+The density of points at the beginning of follow-up relative to the end indicates a non-constant hazard function.", echo=FALSE, eval = TRUE}
 popTimeData <- popTime(data = bmtcrr, time = "ftime")
@@
      ribbon.params = list(fill = "gray50"),
-     case.params = list(size = 0.8),
-     competing.params = list(size = 0.8)) + 
+     case.params = list(size = 1.5),
+     competing.params = list(size = 1.5),
+     fill.params = list(limits = force),
+     color.params = list(limits = force)) + 
   paper_gg_theme +  
@@
 ```{r compAbsrisk, echo = FALSE, fig.cap="Cumulative Incidence curve for a fixed covariate profile and the two disease groups.
-The estimate obtained from case-base sampling is compared to the Fine-Gray and Aalen-Johansen estimates.", eval = TRUE}
+The estimate obtained from case-base sampling is compared to the Fine-Gray and Aalen-Johansen estimates.
+In general, the three approaches agree quite well for AML, while there seems to be a difference of about 5\\% between the Fine-Gray curve and the curves estimated using case-base sampling and Cox regression for ALL.
+However, this difference does not appear to be significant as the curve from case-base sampling is contained within a 95\\% confidence band around the Fine-Gray absolute risk curve (figure not shown).", eval = TRUE}
 risk_all <- dplyr::bind_rows(
@@
 
-## Case study 3---Variable selection
+## Case study 3---variable selection
 
@@
 We then compare the risk estimation over the test set.
-The probabilities over time for each observation are averaged, resulting in the absolute risk curves shown in Figure \@ref(fig:cs3FinalBrier) (A).
-We can see some minimal differences between the three models, with the Kaplan-Meier giving the lowest estimates across follow-up-time.
+The probabilities over time for each observation are averaged, resulting in the absolute risk curves shown in Figure \@ref(fig:cs3FinalBrier)A.
+We can see some minimal differences between the four models, with the Cox model giving the lowest estimates across follow-up-time.
 Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard.
 As described above, we compare the performance between the models by computing the Brier scores over time.
-In Figure \@ref(fig:cs3FinalBrier) (B), we can see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models.
-On the other hand, the differences between these three models are minimal.
+In Figure \@ref(fig:cs3FinalBrier)B, we can see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models.
+On the other hand, the differences between the three adjusted models are minimal.
 
@@
 (A)  Probability of death as a function of follow-up time.
-(B) Brier score as a function of follow-up time, where a lower score corresponds to better performace.", warning = FALSE}
+(B) Brier score as a function of follow-up time, where a lower score corresponds to better performace.
+The unadjusted Kaplan-Meier gives the lowest risk estimates across follow-up-time and highest Brier scores, while we can see minimal differences between the three adjusted models across both metrics.
+Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard.", warning = FALSE}
 # Combine scores

### 2022-11-21 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
pdf compilation works after updating rmarkdown and rstudio to latest ver

@@
 Figure \ref{fig:plot-erspc-data} depicts the process of creating a population-time plot.
-It is built sequentially by first adding a layer for the area representing the population time in gray (Figure \ref{fig:plot-erspc-data} A), with subjects having the least amount of observation time plotted at the top of the y-axis.
+It is built sequentially by first adding a layer for the area representing the population time in gray (Figure \ref{fig:plot-erspc-data}A), with subjects having the least amount of observation time plotted at the top of the y-axis.
 We immediately notice a distinctive \emph{stepwise shape} in the population time area.
@@
 Next, layers for the case series and base series are added.
-The y-axis location of each case moment is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area (Figure \ref{fig:plot-erspc-data} B).
+The y-axis location of each case moment is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area (Figure \ref{fig:plot-erspc-data}B).
 By randomly distributing the cases, we can get a sense of the incidence density.
-In Figure \ref{fig:plot-erspc-data} C, we see that more events are observed at later follow-up times.
+In Figure \ref{fig:plot-erspc-data}C, we see that more events are observed at later follow-up times.
 Therefore, a constant hazard model would not be appropriate in this instance as it would overestimate the incidence earlier on in time, and underestimate it later on.
-Finally, the base series is sampled uniformly from the study base (Figure \ref{fig:plot-erspc-data} D).
+Finally, the base series is sampled uniformly from the study base (Figure \ref{fig:plot-erspc-data}D).
 The reader should refer to the package vignettes for more examples and a detailed description of how to modify the aesthetics of a population-time plot.
@@
 \hypertarget{case-study-1flexible-modeling-of-the-hazard-function}{%
-\subsection{Case study 1---Flexible modeling of the hazard function}\label{case-study-1flexible-modeling-of-the-hazard-function}}
+\subsection{Case study 1---flexible modeling of the hazard function}\label{case-study-1flexible-modeling-of-the-hazard-function}}
 
@@
 We plot the hazard functions for both treatment arms in Figure \ref{fig:cs1hazard}.
-The pattern we see is consistent with the population-time plot shown in Figure \ref{fig:plot-erspc-data} C, where more events are observed at later follow-up times.
+The pattern we see is consistent with the population-time plot shown in Figure \ref{fig:plot-erspc-data}C, where more events are observed at later follow-up times.
 The drop at the end can be explained by the fact that very few observations were followed for the entire 15 year period.
 
-\begin{verbatim}
+\begin{Schunk}
+\begin{Sinput}
 plot(fit, type = "hazard",
@@
                           by = "ScrArm"))
+\end{Sinput}
+\begin{Soutput}
 #> Conditions used in construction of plot
@@
 #> offset: 0
-\end{verbatim}
-
+\end{Soutput}
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./cs1hazard-1} \caption{Estimated hazard functions for control and screening groups in the ERSPC dataset.
+\includegraphics[width=\textwidth,keepaspectratio=true]{./cs1hazard-1} \caption[Estimated hazard functions for control and screening groups in the ERSPC dataset]{Estimated hazard functions for control and screening groups in the ERSPC dataset.
 Hazards are estimated from fitting a parametric model with casebase sampling as a function of the interaction between a cubic pspline basis (df=2) of follow-up time and treatment arm.
@@
 \end{figure}
+\end{Schunk}
 
@@
 
-\begin{verbatim}
+\begin{Schunk}
+\begin{Sinput}
 new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
@@
 risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
-\end{verbatim}
+\end{Sinput}
+\end{Schunk}
 
@@
 
-\begin{verbatim}
+\begin{Schunk}
+\begin{Sinput}
 conf_ints <- confint(risk, fit)
 head(conf_ints)
+\end{Sinput}
+\begin{Soutput}
 #>   time estimate conf.low conf.high      cov_prof
@@
 #> 6  0.5  1.2e-05  2.3e-06   5.0e-05 Control group
-\end{verbatim}
+\end{Soutput}
+\end{Schunk}
 
@@
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cif-conf-1} \caption{Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands.}\label{fig:erspc-cif-conf}
+\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cif-conf-1} \caption{Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands.
+The smooth curve (case-base sampling) vs.
+step function (Cox model) highlight one of the main differences between the two approaches.
+The larger confidence bands in the later years is due to the relatively few number of individuals who were followed for more than 12 years.}\label{fig:erspc-cif-conf}
 \end{figure}
@@
 \hypertarget{case-study-2competing-risk-analysis}{%
-\subsection{Case study 2---Competing risk analysis}\label{case-study-2competing-risk-analysis}}
+\subsection{Case study 2---competing risk analysis}\label{case-study-2competing-risk-analysis}}
 
@@
 First, we look at a population-time plot to visualize the incidence density of both relapse and the competing events.
-In Figure \ref{fig:compPop}, failure times are highlighted on the plot using red dots for the event of interest (panel A) and blue dots for competing events (panel B).
-In both panels, we see evidence of a non-constant hazard function: the density of points is larger at the beginning of follow-up than at the end.
+In Figure \ref{fig:compPop}, failure times are highlighted on the plot using red dots for the event of interest and blue dots for competing events.
+In this plot, we see evidence of a non-constant hazard function: the density of points is larger at the beginning of follow-up than at the end.
 
 \begin{figure}[ht]
-\includegraphics[width=\textwidth,keepaspectratio=true]{./compPop-1} \caption{Population-time plot for the stem-cell transplant study with both relapse and competing events.}\label{fig:compPop}
+\includegraphics[width=\textwidth,keepaspectratio=true]{./compPop-1} \caption{Population-time plot for the stem-cell transplant study with both relapse and competing events.
+The area representing the population time is shown in gray, with subjects having the least amount of observation time plotted at the top of the y-axis.
+The y-axis location of each case series and competing event moment is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area.
+The density of points at the beginning of follow-up relative to the end indicates a non-constant hazard function.}\label{fig:compPop}
 \end{figure}
@@
 
-\begin{verbatim}
+\begin{Schunk}
+\begin{Sinput}
 model_cb <- fitSmoothHazard(
@@
 )
-\end{verbatim}
+\end{Sinput}
+\end{Schunk}
 
@@
 
-\begin{verbatim}
+\begin{Schunk}
+\begin{Sinput}
 library(survival)
@@
                    data = bmtcrr_cox, id = id)
-\end{verbatim}
+\end{Sinput}
+\end{Schunk}
 
@@
 
-\begin{verbatim}
+\begin{Schunk}
+\begin{Sinput}
 library(timereg)
@@
 risk_fg <- predict(model_fg, newdata, times = time_points)
-\end{verbatim}
+\end{Sinput}
+\end{Schunk}
 
@@
 
-\begin{verbatim}
+\begin{Schunk}
+\begin{Sinput}
 # Estimate absolute risk curve
 risk_cox <- survfit(model_cox, newdata = newdata)
-\end{verbatim}
+\end{Sinput}
+\end{Schunk}
 
@@
 \includegraphics[width=\textwidth,keepaspectratio=true]{./compAbsrisk-1} \caption{Cumulative Incidence curve for a fixed covariate profile and the two disease groups.
-The estimate obtained from case-base sampling is compared to the Fine-Gray and Aalen-Johansen estimates.}\label{fig:compAbsrisk}
+The estimate obtained from case-base sampling is compared to the Fine-Gray and Aalen-Johansen estimates.
+In general, the three approaches agree quite well for AML, while there seems to be a difference of about 5\% between the Fine-Gray curve and the curves estimated using case-base sampling and Cox regression for ALL.
+However, this difference does not appear to be significant as the curve from case-base sampling is contained within a 95\% confidence band around the Fine-Gray absolute risk curve (figure not shown).}\label{fig:compAbsrisk}
 \end{figure}
@@
 \hypertarget{case-study-3variable-selection}{%
-\subsection{Case study 3---Variable selection}\label{case-study-3variable-selection}}
+\subsection{Case study 3---variable selection}\label{case-study-3variable-selection}}
 
@@
 We then compare the risk estimation over the test set.
-The probabilities over time for each observation are averaged, resulting in the absolute risk curves shown in Figure \ref{fig:cs3FinalBrier} (A).
-We can see some minimal differences between the three models, with the Kaplan-Meier giving the lowest estimates across follow-up-time.
+The probabilities over time for each observation are averaged, resulting in the absolute risk curves shown in Figure \ref{fig:cs3FinalBrier}A.
+We can see some minimal differences between the four models, with the Cox model giving the lowest estimates across follow-up-time.
 Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard.
 As described above, we compare the performance between the models by computing the Brier scores over time.
-In Figure \ref{fig:cs3FinalBrier} (B), we can see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models.
-On the other hand, the differences between these three models are minimal.
+In Figure \ref{fig:cs3FinalBrier}B, we can see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models.
+On the other hand, the differences between the three adjusted models are minimal.
 
@@
 (A)  Probability of death as a function of follow-up time.
-(B) Brier score as a function of follow-up time, where a lower score corresponds to better performace.}\label{fig:cs3FinalBrier}
+(B) Brier score as a function of follow-up time, where a lower score corresponds to better performace.
+The unadjusted Kaplan-Meier gives the lowest risk estimates across follow-up-time and highest Brier scores, while we can see minimal differences between the three adjusted models across both metrics.
+Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard.}\label{fig:cs3FinalBrier}
 \end{figure}

### 2022-11-22 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
more updated captions

@@
 We then compare the risk estimation over the test set.
-The probabilities over time for each observation are averaged, resulting in the absolute risk curves shown in Figure \@ref(fig:cs3FinalBrier)A.
-We can see some minimal differences between the four models, with the Cox model giving the lowest estimates across follow-up-time.
+The predicted probabilities for each test set observation are averaged, resulting in the absolute risk curves shown in Figure \@ref(fig:cs3FinalBrier)A.
+The Kaplan-Meier curve is calculated on the test set only.
+We see minimal differences between the four approaches across follow-up-time.
 Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard.
 As described above, we compare the performance between the models by computing the Brier scores over time.
-In Figure \@ref(fig:cs3FinalBrier)B, we can see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models.
+In Figure \@ref(fig:cs3FinalBrier)B, we see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models.
 On the other hand, the differences between the three adjusted models are minimal.
@@
 CB), and Kaplan-Meier (K-M).
-(A)  Probability of death as a function of follow-up time.
-(B) Brier score as a function of follow-up time, where a lower score corresponds to better performace.
-The unadjusted Kaplan-Meier gives the lowest risk estimates across follow-up-time and highest Brier scores, while we can see minimal differences between the three adjusted models across both metrics.
-Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard.", warning = FALSE}
+(A) Probability of death as a function of follow-up time which is the average of the predicted probabilities for each test set observation.
+The Kaplan-Meier curve is calculated on the test set only.
+(B) Brier score as a function of follow-up time, where a lower score corresponds to better performance.
+We see minimal differences between the four approaches across follow-up-time for the absolute risk curves.
+Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard.
+We see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models.
+On the other hand, the differences between the three adjusted models are minimal.", warning = FALSE}
 # Combine scores

### 2022-11-22 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
more updated captions

@@
 We then compare the risk estimation over the test set.
-The probabilities over time for each observation are averaged, resulting in the absolute risk curves shown in Figure \ref{fig:cs3FinalBrier}A.
-We can see some minimal differences between the four models, with the Cox model giving the lowest estimates across follow-up-time.
+The predicted probabilities for each test set observation are averaged, resulting in the absolute risk curves shown in Figure \ref{fig:cs3FinalBrier}A.
+The Kaplan-Meier curve is calculated on the test set only.
+We see minimal differences between the four approaches across follow-up-time.
 Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard.
 As described above, we compare the performance between the models by computing the Brier scores over time.
-In Figure \ref{fig:cs3FinalBrier}B, we can see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models.
+In Figure \ref{fig:cs3FinalBrier}B, we see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models.
 On the other hand, the differences between the three adjusted models are minimal.
@@
 CB), and Kaplan-Meier (K-M).
-(A)  Probability of death as a function of follow-up time.
-(B) Brier score as a function of follow-up time, where a lower score corresponds to better performace.
-The unadjusted Kaplan-Meier gives the lowest risk estimates across follow-up-time and highest Brier scores, while we can see minimal differences between the three adjusted models across both metrics.
-Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard.}\label{fig:cs3FinalBrier}
+(A) Probability of death as a function of follow-up time which is the average of the predicted probabilities for each test set observation.
+The Kaplan-Meier curve is calculated on the test set only.
+(B) Brier score as a function of follow-up time, where a lower score corresponds to better performance.
+We see minimal differences between the four approaches across follow-up-time for the absolute risk curves.
+Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard.
+We see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models.
+On the other hand, the differences between the three adjusted models are minimal.}\label{fig:cs3FinalBrier}
 \end{figure}

### 2022-11-25 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd

@@
 ```{r plot-erspc-data, eval = TRUE, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plot for the ERSPC dataset.
-A: The gray area can be thought of as N = 159,893 infinitely thin horizontal rectangles ordered by length of follow-up.
+A: The gray area can be thought of as N=159,893 infinitely thin horizontal rectangles ordered by length of follow-up.
 B: The red points correspond to when death has occurred for any one of those infinitely thin rectangles.
@@
 As described above, we compare the performance between the models by computing the Brier scores over time.
-In Figure \@ref(fig:cs3FinalBrier)B, we see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models.
-On the other hand, the differences between the three adjusted models are minimal.
+In Figure \@ref(fig:cs3FinalBrier)B, we see that the adjusted models all perform similarly, outperforming the Kaplan-Meier estimate.
 
@@
 The Kaplan-Meier curve is calculated on the test set only.
-(B) Brier score as a function of follow-up time, where a lower score corresponds to better performance.
 We see minimal differences between the four approaches across follow-up-time for the absolute risk curves.
 Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard.
-We see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models.
-On the other hand, the differences between the three adjusted models are minimal.", warning = FALSE}
+(B) Brier score as a function of follow-up time, where a lower score corresponds to better performance.
+We see that the adjusted models all perform similarly, outperforming the Kaplan-Meier estimate.", warning = FALSE}
 # Combine scores
